### 2020-07-11 main.tex
Add title

\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts
% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\begin{document}

\title{On the Use and Removal of Eval from R}

\author{\IEEEauthorblockN{1\textsuperscript{st} Given Name Surname}
\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
\textit{name of organization (of Aff.)}\\
City, Country \\
email address or ORCID}
\and
\IEEEauthorblockN{2\textsuperscript{nd} Given Name Surname}
\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
\textit{name of organization (of Aff.)}\\
City, Country \\
email address or ORCID}
\and
\IEEEauthorblockN{3\textsuperscript{rd} Given Name Surname}
\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
\textit{name of organization (of Aff.)}\\
City, Country \\
email address or ORCID}
\and
\IEEEauthorblockN{4\textsuperscript{th} Given Name Surname}
\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
\textit{name of organization (of Aff.)}\\
City, Country \\
email address or ORCID}
\and
\IEEEauthorblockN{5\textsuperscript{th} Given Name Surname}
\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
\textit{name of organization (of Aff.)}\\
City, Country \\
email address or ORCID}
\and
\IEEEauthorblockN{6\textsuperscript{th} Given Name Surname}
\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
\textit{name of organization (of Aff.)}\\
City, Country \\
email address or ORCID}
}

\maketitle

\begin{abstract}
This document is a model and instructions for \LaTeX.
This and the IEEEtran.cls file define the components of your paper [title, text, heads, etc.]. *CRITICAL: Do Not Use Symbols, Special Characters, Footnotes, 
or Math in Paper Title or Abstract.
\end{abstract}

\begin{IEEEkeywords}
component, formatting, style, styling, insert
\end{IEEEkeywords}

\section{Introduction}
This document is a model and instructions for \LaTeX.
Please observe the conference page limits.

\section{Ease of Use}

\subsection{Maintaining the Integrity of the Specifications}

The IEEEtran class file is used to format your paper and style the text. All margins, 
column widths, line spaces, and text fonts are prescribed; please do not 
alter them. You may note peculiarities. For example, the head margin
measures proportionately more than is customary. This measurement 
and others are deliberate, using specifications that anticipate your paper 
as one part of the entire proceedings, and not as an independent document. 
Please do not revise any of the current designations.

\section{Prepare Your Paper Before Styling}
Before you begin to format your paper, first write and save the content as a 
separate text file. Complete all content and organizational editing before 
formatting. Please note sections \ref{AA}--\ref{SCM} below for more information on 
proofreading, spelling and grammar.

Keep your text and graphic files separate until after the text has been 
formatted and styled. Do not number text heads---{\LaTeX} will do that 
for you.

\subsection{Abbreviations and Acronyms}\label{AA}
Define abbreviations and acronyms the first time they are used in the text, 
even after they have been defined in the abstract. Abbreviations such as 
IEEE, SI, MKS, CGS, ac, dc, and rms do not have to be defined. Do not use 
abbreviations in the title or heads unless they are unavoidable.

\subsection{Units}
\begin{itemize}
\item Use either SI (MKS) or CGS as primary units. (SI units are encouraged.) English units may be used as secondary units (in parentheses). An exception would be the use of English units as identifiers in trade, such as ``3.5-inch disk drive''.
\item Avoid combining SI and CGS units, such as current in amperes and magnetic field in oersteds. This often leads to confusion because equations do not balance dimensionally. If you must use mixed units, clearly state the units for each quantity that you use in an equation.
\item Do not mix complete spellings and abbreviations of units: ``Wb/m\textsuperscript{2}'' or ``webers per square meter'', not ``webers/m\textsuperscript{2}''. Spell out units when they appear in text: ``. . . a few henries'', not ``. . . a few H''.
\item Use a zero before decimal points: ``0.25'', not ``.25''. Use ``cm\textsuperscript{3}'', not ``cc''.)
\end{itemize}

\subsection{Equations}
Number equations consecutively. To make your 
equations more compact, you may use the solidus (~/~), the exp function, or 
appropriate exponents. Italicize Roman symbols for quantities and variables, 
but not Greek symbols. Use a long dash rather than a hyphen for a minus 
sign. Punctuate equations with commas or periods when they are part of a 
sentence, as in:
\begin{equation}
a+b=\gamma\label{eq}
\end{equation}

Be sure that the 
symbols in your equation have been defined before or immediately following 
the equation. Use ``\eqref{eq}'', not ``Eq.~\eqref{eq}'' or ``equation \eqref{eq}'', except at 
the beginning of a sentence: ``Equation \eqref{eq} is . . .''

\subsection{\LaTeX-Specific Advice}

Please use ``soft'' (e.g., \verb|\eqref{Eq}|) cross references instead
of ``hard'' references (e.g., \verb|(1)|). That will make it possible
to combine sections, add equations, or change the order of figures or
citations without having to go through the file line by line.

Please don't use the \verb|{eqnarray}| equation environment. Use
\verb|{align}| or \verb|{IEEEeqnarray}| instead. The \verb|{eqnarray}|
environment leaves unsightly spaces around relation symbols.

Please note that the \verb|{subequations}| environment in {\LaTeX}
will increment the main equation counter even when there are no
equation numbers displayed. If you forget that, you might write an
article in which the equation numbers skip from (17) to (20), causing
the copy editors to wonder if you've discovered a new method of
counting.

{\BibTeX} does not work by magic. It doesn't get the bibliographic
data from thin air but from .bib files. If you use {\BibTeX} to produce a
bibliography you must send the .bib files.

{\LaTeX} can't read your mind. If you assign the same label to a
subsubsection and a table, you might find that Table I has been cross
referenced as Table IV-B3.

{\LaTeX} does not have precognitive abilities. If you put a
\verb|\label| command before the command that updates the counter it's
supposed to be using, the label will pick up the last counter to be
cross referenced instead. In particular, a \verb|\label| command
should not go before the caption of a figure or a table.

Do not use \verb|\nonumber| inside the \verb|{array}| environment. It
will not stop equation numbers inside \verb|{array}| (there won't be
any anyway) and it might stop a wanted equation number in the
surrounding equation.

\subsection{Some Common Mistakes}\label{SCM}
\begin{itemize}
\item The word ``data'' is plural, not singular.
\item The subscript for the permeability of vacuum $\mu_{0}$, and other common scientific constants, is zero with subscript formatting, not a lowercase letter ``o''.
\item In American English, commas, semicolons, periods, question and exclamation marks are located within quotation marks only when a complete thought or name is cited, such as a title or full quotation. When quotation marks are used, instead of a bold or italic typeface, to highlight a word or phrase, punctuation should appear outside of the quotation marks. A parenthetical phrase or statement at the end of a sentence is punctuated outside of the closing parenthesis (like this). (A parenthetical sentence is punctuated within the parentheses.)
\item A graph within a graph is an ``inset'', not an ``insert''. The word alternatively is preferred to the word ``alternately'' (unless you really mean something that alternates).
\item Do not use the word ``essentially'' to mean ``approximately'' or ``effectively''.
\item In your paper title, if the words ``that uses'' can accurately replace the word ``using'', capitalize the ``u''; if not, keep using lower-cased.
\item Be aware of the different meanings of the homophones ``affect'' and ``effect'', ``complement'' and ``compliment'', ``discreet'' and ``discrete'', ``principal'' and ``principle''.
\item Do not confuse ``imply'' and ``infer''.
\item The prefix ``non'' is not a word; it should be joined to the word it modifies, usually without a hyphen.
\item There is no period after the ``et'' in the Latin abbreviation ``et al.''.
\item The abbreviation ``i.e.'' means ``that is'', and the abbreviation ``e.g.'' means ``for example''.
\end{itemize}
An excellent style manual for science writers is \cite{b7}.

\subsection{Authors and Affiliations}
\textbf{The class file is designed for, but not limited to, six authors.} A 
minimum of one author is required for all conference articles. Author names 
should be listed starting from left to right and then moving down to the 
next line. This is the author sequence that will be used in future citations 
and by indexing services. Names should not be listed in columns nor group by 
affiliation. Please keep your affiliations as succinct as possible (for 
example, do not differentiate among departments of the same organization).

\subsection{Identify the Headings}
Headings, or heads, are organizational devices that guide the reader through 
your paper. There are two types: component heads and text heads.

Component heads identify the different components of your paper and are not 
topically subordinate to each other. Examples include Acknowledgments and 
References and, for these, the correct style to use is ``Heading 5''. Use 
``figure caption'' for your Figure captions, and ``table head'' for your 
table title. Run-in heads, such as ``Abstract'', will require you to apply a 
style (in this case, italic) in addition to the style provided by the drop 
down menu to differentiate the head from the text.

Text heads organize the topics on a relational, hierarchical basis. For 
example, the paper title is the primary text head because all subsequent 
material relates and elaborates on this one topic. If there are two or more 
sub-topics, the next level head (uppercase Roman numerals) should be used 
and, conversely, if there are not at least two sub-topics, then no subheads 
should be introduced.

\subsection{Figures and Tables}
\paragraph{Positioning Figures and Tables} Place figures and tables at the top and 
bottom of columns. Avoid placing them in the middle of columns. Large 
figures and tables may span across both columns. Figure captions should be 
below the figures; table heads should appear above the tables. Insert 
figures and tables after they are cited in the text. Use the abbreviation 
``Fig.~\ref{fig}'', even at the beginning of a sentence.

\begin{table}[htbp]
\caption{Table Type Styles}
\begin{center}
\begin{tabular}{|c|c|c|c|}
\hline
\textbf{Table}&\multicolumn{3}{|c|}{\textbf{Table Column Head}} \\
\cline{2-4} 
\textbf{Head} & \textbf{\textit{Table column subhead}}& \textbf{\textit{Subhead}}& \textbf{\textit{Subhead}} \\
\hline
copy& More table copy$^{\mathrm{a}}$& &  \\
\hline
\multicolumn{4}{l}{$^{\mathrm{a}}$Sample of a Table footnote.}
\end{tabular}
\label{tab1}
\end{center}
\end{table}

\begin{figure}[htbp]
\centerline{\includegraphics{fig1.png}}
\caption{Example of a figure caption.}
\label{fig}
\end{figure}

Figure Labels: Use 8 point Times New Roman for Figure labels. Use words 
rather than symbols or abbreviations when writing Figure axis labels to 
avoid confusing the reader. As an example, write the quantity 
``Magnetization'', or ``Magnetization, M'', not just ``M''. If including 
units in the label, present them within parentheses. Do not label axes only 
with units. In the example, write ``Magnetization (A/m)'' or ``Magnetization 
\{A[m(1)]\}'', not just ``A/m''. Do not label axes with a ratio of 
quantities and units. For example, write ``Temperature (K)'', not 
``Temperature/K''.

\section*{Acknowledgment}

The preferred spelling of the word ``acknowledgment'' in America is without 
an ``e'' after the ``g''. Avoid the stilted expression ``one of us (R. B. 
G.) thanks $\ldots$''. Instead, try ``R. B. G. thanks$\ldots$''. Put sponsor 
acknowledgments in the unnumbered footnote on the first page.

\section*{References}

Please number citations consecutively within brackets \cite{b1}. The 
sentence punctuation follows the bracket \cite{b2}. Refer simply to the reference 
number, as in \cite{b3}---do not use ``Ref. \cite{b3}'' or ``reference \cite{b3}'' except at 
the beginning of a sentence: ``Reference \cite{b3} was the first $\ldots$''

Number footnotes separately in superscripts. Place the actual footnote at 
the bottom of the column in which it was cited. Do not put footnotes in the 
abstract or reference list. Use letters for table footnotes.

Unless there are six authors or more give all authors' names; do not use 
``et al.''. Papers that have not been published, even if they have been 
submitted for publication, should be cited as ``unpublished'' \cite{b4}. Papers 
that have been accepted for publication should be cited as ``in press'' \cite{b5}. 
Capitalize only the first word in a paper title, except for proper nouns and 
element symbols.

For papers published in translation journals, please give the English 
citation first, followed by the original foreign-language citation \cite{b6}.

\begin{thebibliography}{00}
\bibitem{b1} G. Eason, B. Noble, and I. N. Sneddon, ``On certain integrals of Lipschitz-Hankel type involving products of Bessel functions,'' Phil. Trans. Roy. Soc. London, vol. A247, pp. 529--551, April 1955.
\bibitem{b2} J. Clerk Maxwell, A Treatise on Electricity and Magnetism, 3rd ed., vol. 2. Oxford: Clarendon, 1892, pp.68--73.
\bibitem{b3} I. S. Jacobs and C. P. Bean, ``Fine particles, thin films and exchange anisotropy,'' in Magnetism, vol. III, G. T. Rado and H. Suhl, Eds. New York: Academic, 1963, pp. 271--350.
\bibitem{b4} K. Elissa, ``Title of paper if known,'' unpublished.
\bibitem{b5} R. Nicole, ``Title of paper with only first word capitalized,'' J. Name Stand. Abbrev., in press.
\bibitem{b6} Y. Yorozu, M. Hirano, K. Oka, and Y. Tagawa, ``Electron spectroscopy studies on magneto-optical media and plastic substrate interface,'' IEEE Transl. J. Magn. Japan, vol. 2, pp. 740--741, August 1987 [Digests 9th Annual Conf. Magnetics Japan, p. 301, 1982].
\bibitem{b7} M. Young, The Technical Writer's Handbook. Mill Valley, CA: University Science, 1989.
\end{thebibliography}
\vspace{12pt}
\color{red}
IEEE conference templates contain guidance text for composing and formatting conference papers. Please ensure that all template text is removed from your conference paper prior to submission to the conference. Failure to remove the template text from your paper may result in your paper not being published.

\end{document}

### 2020-07-13 main.tex
Add rough paper outline

@@
 
-\title{On the Use and Removal of Eval from R}
+\title{On the Design, Use and Removal of Eval in R}
 
@@
 \section{Introduction}
-This document is a model and instructions for \LaTeX.
-Please observe the conference page limits.
+\textbf{*}
 
-\section{Ease of Use}
+\section{Background on R}
+\textbf{Aviral}
 
-\subsection{Maintaining the Integrity of the Specifications}
+\section{The Design of Eval}
+\textbf{Aviral}
 
-The IEEEtran class file is used to format your paper and style the text. All margins, 
-column widths, line spaces, and text fonts are prescribed; please do not 
-alter them. You may note peculiarities. For example, the head margin
-measures proportionately more than is customary. This measurement 
-and others are deliberate, using specifications that anticipate your paper 
-as one part of the entire proceedings, and not as an independent document. 
-Please do not revise any of the current designations.
+\section{Infrastructure}
+\textbf{Aviral + Filip}
 
-\section{Prepare Your Paper Before Styling}
-Before you begin to format your paper, first write and save the content as a 
-separate text file. Complete all content and organizational editing before 
-formatting. Please note sections \ref{AA}--\ref{SCM} below for more information on 
-proofreading, spelling and grammar.
+\section{Corpus of R Programs}
+\textbf{Filip}
 
-Keep your text and graphic files separate until after the text has been 
-formatted and styled. Do not number text heads---{\LaTeX} will do that 
-for you.
+\section{Threats to Validity}
+\textbf{Filip}
 
-\subsection{Abbreviations and Acronyms}\label{AA}
-Define abbreviations and acronyms the first time they are used in the text, 
-even after they have been defined in the abstract. Abbreviations such as 
-IEEE, SI, MKS, CGS, ac, dc, and rms do not have to be defined. Do not use 
-abbreviations in the title or heads unless they are unavoidable.
+\section{Analyzing Eval Usage Patterns}
+This section presents the results of our empirical study of eval in the R language.
+\subsection{How many evals}
+\textbf{Pierre}
 
-\subsection{Units}
-\begin{itemize}
-\item Use either SI (MKS) or CGS as primary units.
-(SI units are encouraged.) English units may be used as secondary units (in parentheses).
-An exception would be the use of English units as identifiers in trade, such as ``3.5-inch disk drive''.
-\item Avoid combining SI and CGS units, such as current in amperes and magnetic field in oersteds.
-This often leads to confusion because equations do not balance dimensionally.
-If you must use mixed units, clearly state the units for each quantity that you use in an equation.
-\item Do not mix complete spellings and abbreviations of units: ``Wb/m\textsuperscript{2}'' or ``webers per square meter'', not ``webers/m\textsuperscript{2}''.
-Spell out units when they appear in text: ``.
-.
-.
-a few henries'', not ``.
-.
-.
-a few H''.
-\item Use a zero before decimal points: ``0.25'', not ``.25''. Use ``cm\textsuperscript{3}'', not ``cc''.)
-\end{itemize}
+\subsection{How are they used}
+\textbf{Pierre}
 
-\subsection{Equations}
-Number equations consecutively. To make your 
-equations more compact, you may use the solidus (~/~), the exp function, or 
-appropriate exponents. Italicize Roman symbols for quantities and variables, 
-but not Greek symbols. Use a long dash rather than a hyphen for a minus 
-sign. Punctuate equations with commas or periods when they are part of a 
-sentence, as in:
-\begin{equation}
-a+b=\gamma\label{eq}
-\end{equation}
+\subsection{How to replace evals}
+\textbf{Aviral}
 
-Be sure that the 
-symbols in your equation have been defined before or immediately following 
-the equation. Use ``\eqref{eq}'', not ``Eq.~\eqref{eq}'' or ``equation \eqref{eq}'', except at 
-the beginning of a sentence: ``Equation \eqref{eq} is . . .''
+\subsection{How many evals can be replaced}
+\textbf{Aviral}
 
-\subsection{\LaTeX-Specific Advice}
+\section{Related Work}
+\textbf{Aviral}
 
-Please use ``soft'' (e.g., \verb|\eqref{Eq}|) cross references instead
-of ``hard'' references (e.g., \verb|(1)|). That will make it possible
-to combine sections, add equations, or change the order of figures or
-citations without having to go through the file line by line.
-
-Please don't use the \verb|{eqnarray}| equation environment. Use
-\verb|{align}| or \verb|{IEEEeqnarray}| instead. The \verb|{eqnarray}|
-environment leaves unsightly spaces around relation symbols.
-
-Please note that the \verb|{subequations}| environment in {\LaTeX}
-will increment the main equation counter even when there are no
-equation numbers displayed. If you forget that, you might write an
-article in which the equation numbers skip from (17) to (20), causing
-the copy editors to wonder if you've discovered a new method of
-counting.
-
-{\BibTeX} does not work by magic. It doesn't get the bibliographic
-data from thin air but from .bib files. If you use {\BibTeX} to produce a
-bibliography you must send the .bib files.
-
-{\LaTeX} can't read your mind. If you assign the same label to a
-subsubsection and a table, you might find that Table I has been cross
-referenced as Table IV-B3.
-
-{\LaTeX} does not have precognitive abilities. If you put a
-\verb|\label| command before the command that updates the counter it's
-supposed to be using, the label will pick up the last counter to be
-cross referenced instead. In particular, a \verb|\label| command
-should not go before the caption of a figure or a table.
-
-Do not use \verb|\nonumber| inside the \verb|{array}| environment. It
-will not stop equation numbers inside \verb|{array}| (there won't be
-any anyway) and it might stop a wanted equation number in the
-surrounding equation.
-
-\subsection{Some Common Mistakes}\label{SCM}
-\begin{itemize}
-\item The word ``data'' is plural, not singular.
-\item The subscript for the permeability of vacuum $\mu_{0}$, and other common scientific constants, is zero with subscript formatting, not a lowercase letter ``o''.
-\item In American English, commas, semicolons, periods, question and exclamation marks are located within quotation marks only when a complete thought or name is cited, such as a title or full quotation.
-When quotation marks are used, instead of a bold or italic typeface, to highlight a word or phrase, punctuation should appear outside of the quotation marks.
-A parenthetical phrase or statement at the end of a sentence is punctuated outside of the closing parenthesis (like this).
-(A parenthetical sentence is punctuated within the parentheses.)
-\item A graph within a graph is an ``inset'', not an ``insert''. The word alternatively is preferred to the word ``alternately'' (unless you really mean something that alternates).
-\item Do not use the word ``essentially'' to mean ``approximately'' or ``effectively''.
-\item In your paper title, if the words ``that uses'' can accurately replace the word ``using'', capitalize the ``u''; if not, keep using lower-cased.
-\item Be aware of the different meanings of the homophones ``affect'' and ``effect'', ``complement'' and ``compliment'', ``discreet'' and ``discrete'', ``principal'' and ``principle''.
-\item Do not confuse ``imply'' and ``infer''.
-\item The prefix ``non'' is not a word; it should be joined to the word it modifies, usually without a hyphen.
-\item There is no period after the ``et'' in the Latin abbreviation ``et al.''.
-\item The abbreviation ``i.e.'' means ``that is'', and the abbreviation ``e.g.'' means ``for example''.
-\end{itemize}
-An excellent style manual for science writers is \cite{b7}.
-
-\subsection{Authors and Affiliations}
-\textbf{The class file is designed for, but not limited to, six authors.} A 
-minimum of one author is required for all conference articles. Author names 
-should be listed starting from left to right and then moving down to the 
-next line. This is the author sequence that will be used in future citations 
-and by indexing services. Names should not be listed in columns nor group by 
-affiliation. Please keep your affiliations as succinct as possible (for 
-example, do not differentiate among departments of the same organization).
-
-\subsection{Identify the Headings}
-Headings, or heads, are organizational devices that guide the reader through 
-your paper. There are two types: component heads and text heads.
-
-Component heads identify the different components of your paper and are not 
-topically subordinate to each other. Examples include Acknowledgments and 
-References and, for these, the correct style to use is ``Heading 5''. Use 
-``figure caption'' for your Figure captions, and ``table head'' for your 
-table title. Run-in heads, such as ``Abstract'', will require you to apply a 
-style (in this case, italic) in addition to the style provided by the drop 
-down menu to differentiate the head from the text.
-
-Text heads organize the topics on a relational, hierarchical basis. For 
-example, the paper title is the primary text head because all subsequent 
-material relates and elaborates on this one topic. If there are two or more 
-sub-topics, the next level head (uppercase Roman numerals) should be used 
-and, conversely, if there are not at least two sub-topics, then no subheads 
-should be introduced.
-
-\subsection{Figures and Tables}
-\paragraph{Positioning Figures and Tables} Place figures and tables at the top and 
-bottom of columns. Avoid placing them in the middle of columns. Large 
-figures and tables may span across both columns. Figure captions should be 
-below the figures; table heads should appear above the tables. Insert 
-figures and tables after they are cited in the text. Use the abbreviation 
-``Fig.~\ref{fig}'', even at the beginning of a sentence.
-
-\begin{table}[htbp]
-\caption{Table Type Styles}
-\begin{center}
-\begin{tabular}{|c|c|c|c|}
-\hline
-\textbf{Table}&\multicolumn{3}{|c|}{\textbf{Table Column Head}} \\
-\cline{2-4} 
-\textbf{Head} & \textbf{\textit{Table column subhead}}& \textbf{\textit{Subhead}}& \textbf{\textit{Subhead}} \\
-\hline
-copy& More table copy$^{\mathrm{a}}$& &  \\
-\hline
-\multicolumn{4}{l}{$^{\mathrm{a}}$Sample of a Table footnote.}
-\end{tabular}
-\label{tab1}
-\end{center}
-\end{table}
-
-\begin{figure}[htbp]
-\centerline{\includegraphics{fig1.png}}
-\caption{Example of a figure caption.}
-\label{fig}
-\end{figure}
-
-Figure Labels: Use 8 point Times New Roman for Figure labels. Use words 
-rather than symbols or abbreviations when writing Figure axis labels to 
-avoid confusing the reader. As an example, write the quantity 
-``Magnetization'', or ``Magnetization, M'', not just ``M''. If including 
-units in the label, present them within parentheses. Do not label axes only 
-with units. In the example, write ``Magnetization (A/m)'' or ``Magnetization 
-\{A[m(1)]\}'', not just ``A/m''. Do not label axes with a ratio of 
-quantities and units. For example, write ``Temperature (K)'', not 
-``Temperature/K''.
+\section{Conclusion}
+\textbf{*}
 
@@
 
-The preferred spelling of the word ``acknowledgment'' in America is without 
-an ``e'' after the ``g''. Avoid the stilted expression ``one of us (R. B. 
-G.) thanks $\ldots$''. Instead, try ``R. B. G. thanks$\ldots$''. Put sponsor 
-acknowledgments in the unnumbered footnote on the first page.
-
 \section*{References}
 
-Please number citations consecutively within brackets \cite{b1}. The 
-sentence punctuation follows the bracket \cite{b2}. Refer simply to the reference 
-number, as in \cite{b3}---do not use ``Ref. \cite{b3}'' or ``reference \cite{b3}'' except at 
-the beginning of a sentence: ``Reference \cite{b3} was the first $\ldots$''
-
-Number footnotes separately in superscripts. Place the actual footnote at 
-the bottom of the column in which it was cited. Do not put footnotes in the 
-abstract or reference list. Use letters for table footnotes.
-
-Unless there are six authors or more give all authors' names; do not use 
-``et al.''. Papers that have not been published, even if they have been 
-submitted for publication, should be cited as ``unpublished'' \cite{b4}. Papers 
-that have been accepted for publication should be cited as ``in press'' \cite{b5}. 
-Capitalize only the first word in a paper title, except for proper nouns and 
-element symbols.
-
-For papers published in translation journals, please give the English 
-citation first, followed by the original foreign-language citation \cite{b6}.
-
-\begin{thebibliography}{00}
-\bibitem{b1} G.
-Eason, B.
-Noble, and I.
-N.
-Sneddon, ``On certain integrals of Lipschitz-Hankel type involving products of Bessel functions,'' Phil.
-Trans.
-Roy.
-Soc.
-London, vol.
-A247, pp.
-529--551, April 1955.
-\bibitem{b2} J. Clerk Maxwell, A Treatise on Electricity and Magnetism, 3rd ed., vol. 2. Oxford: Clarendon, 1892, pp.68--73.
-\bibitem{b3} I. S. Jacobs and C. P. Bean, ``Fine particles, thin films and exchange anisotropy,'' in Magnetism, vol. III, G. T. Rado and H. Suhl, Eds. New York: Academic, 1963, pp. 271--350.
-\bibitem{b4} K. Elissa, ``Title of paper if known,'' unpublished.
-\bibitem{b5} R. Nicole, ``Title of paper with only first word capitalized,'' J. Name Stand. Abbrev., in press.
-\bibitem{b6} Y.
-Yorozu, M.
-Hirano, K.
-Oka, and Y.
-Tagawa, ``Electron spectroscopy studies on magneto-optical media and plastic substrate interface,'' IEEE Transl.
-J.
-Magn.
-Japan, vol.
-2, pp.
-740--741, August 1987 [Digests 9th Annual Conf.
-Magnetics Japan, p.
-301, 1982].
-\bibitem{b7} M. Young, The Technical Writer's Handbook. Mill Valley, CA: University Science, 1989.
-\end{thebibliography}
-\vspace{12pt}
-\color{red}
-IEEE conference templates contain guidance text for composing and formatting conference papers.
-Please ensure that all template text is removed from your conference paper prior to submission to the conference.
-Failure to remove the template text from your paper may result in your paper not being published.
-
 \end{document}

### 2020-08-10 analysis/analysis_stable.Rmd
Adding the analysis notebook.

---
title: "analysis_stable"
output: 
  html_document:
    toc: true
    toc_float: true
    df_print: paged
date: "`r substr(Sys.time(), 1, 10)`"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.align = "center",
  fig.retina = 2,
  fig.width = 10,
  cache.lazy = FALSE
  )

library(tidyverse)
library(stringr)
library(viridis)
library(vctrs)

library(DT)

library(fst)

library(bench)

source("insights.R", local = knitr::knit_global())
```

All insights about eval calls are to put here, after the computations are stabilized (so use another notebook to explore).

# Setup

```{r eval_calls_file}
eval_calls_file <- "/var/lib/R/project-evalR/run/package-evals-traced-all.1/calls.fst"
```

```{r load_file}
eval_calls <- read_fst(eval_calls_file)
eval_calls <- as_tibble(eval_calls) 
```

```{r resolve_types}
eval_calls <- eval_calls %>% resolve_sexp_name(expr_expression_type) %>% 
                             resolve_sexp_name(expr_resolved_type) %>%
                             resolve_sexp_name(enclos_type) %>%
                             resolve_sexp_name(envir_type)
```

# Usage metrics

## R code size

## Number of `eval`

For the number of call sites, we either use the `srcref` when there is one. When there is not (such as in *base*), we use a combination of file and `eval_expr_expression`.

```{r}
nb_call_sites <- eval_calls %>% select(eval_call_srcref) %>% n_distinct(na.rm = TRUE)  + eval_calls %>% filter(is.na(eval_call_srcref)) %>% select(file, eval_call_expression)%>% n_distinct()
```

There are `r nb_call_sites` different `eval` call sites.

## Number of calls to `eval`

```{r}
nb_eval_calls <- eval_calls %>% nrow()
```

There are `r nb_eval_calls` eval calls.

## Size of expressions as arguments to `eval`

## Amount of computations performed in `eval`

## Aliases of `eval`

# A taxonomy of `eval`

## The operation mix

## Scope (environments)

## Patterns

## Provenance

## Consistence

# Other faces of `eval`

# Case studies

### 2020-08-10 analysis/insights.R

@@ -1,0 +1,179 @@
+
+get_expr <- function(eval_call)
+{
+  # Special case for expressions starting with _ (such as _inherit in ggproto)
+  escaped_eval_call <- if(startsWith(eval_call, "_"))
+  {
+    paste0("`", eval_call, "`")
+  }
+  else
+  {
+    eval_call
+  }
+  
+  exp <- NA
+  # Would fail for instance for 
+  # "`$<-`(new(\"C++Field\", .xData = <environment>), \"read_only\", TRUE)" (classInt package)
+  # exp <- tryCatch(
+  #   parse(text = escaped_eval_call)[[1]],
+  #   error = function(e) {return(NA)})
+  try(exp <- parse(text = escaped_eval_call)[[1]], silent = TRUE)
+
+  
+  return(exp)
+}
+
+function_name <- function(eval_call)
+{
+  exp <- get_expr(eval_call)
+  if(is.call(exp))
+  {
+    return(paste(deparse(exp[[1]]), collapse = "\n"))
+  }
+  return(NA)
+}
+
+
+function_arguments <- function(eval_call)
+{
+  exp <- get_expr(eval_call)
+  if(is.call(exp))
+  {
+    return(map_chr(as.list(exp[-1]), function(chr) { paste(deparse(chr), collapse = "\n")}))
+  }
+  return(NA)
+}
+
+# See extract_inner_exp which takes a str
+# This one takes an expression
+# We assume that eval_expr_call is a call
+extract_inner_exp_aux <- function(eval_expr_call)
+{
+  args <- as.list(eval_expr_call[-1])
+  leaves <- list()
+  no_calls <- TRUE
+  for(arg in args)
+  {
+    if(!missing(arg)) # we cannot have the two conditions in the same if
+    {
+      if(is.call(arg))
+      {
+        leaves <-  c(leaves, extract_inner_exp_aux(arg))
+        no_calls <- FALSE
+      }
+    }
+  }
+  if(no_calls)
+  {
+    return(c(leaves, deparse(eval_expr_call)))
+  }
+  return(leaves)
+}
+
+# Extract the inner calls and arguments
+# For instance, in f(g(h(e))), we would get h(e)
+# For f(g(e), t), we will get g(e)
+extract_inner_exp <- function(eval_call)
+{
+  #cat(eval_call, "\n")
+  exp <- get_expr(eval_call)
+  if(is.call(exp))
+  {
[…]

### 2020-08-10 main.tex
Added some comments to the paper

@@
 \section{Introduction}
+- Why eval is used in R
+- What the community says about using eval
+- What existing tools do about eval
+- Do people use eval for obvious things like looking up variable, if yes, it is
+because they are unaware of alternatives?
+- Why automating the transformation is a hard problem.
+- What is the goal of the work? To replace eval with code for static analyssi or
+for consumption by the users of the language?
+- We only permit transformations that preserve code behavior?
+- Do we accept overly complex code as the output?
+
+Why do we do this dynamically? Because it raises a chicken-and-egg problem,
+doing it statically would require us to reason about eval based code which may
+not be possible in general.
+
+What are the challenges? Why is it not trivial?
+- Do we handle aliasing of eval?
+- Simple cases can be subtle? (laziness for ex.
+
+Why can't we replace eval(code) by code?
+
+Intended user of the technique? Which version of R is targeted.
+
+Contributions:
+- Describe the use of eval in R
+- Develop a tool to replace evals guided by dynamic traces
+
+- cannot replace with (cite study of with construct).
 \textbf{*}
@@
 \textbf{Pierre}
+- classify evals based on what they can be replaced with
 
@@
 
+\section{Evaluation}
+
 \section{Related Work}
+\eval has been widely discussed in the research literature for the challenges it
+imposes for code comprehension and automated code analyses.
+Perhaps, \citeit{EvalMenDo} is the most popular exposition of \eval and its
+challenges for JavaScript.
+
 \textbf{Aviral}
+- javascript
+- ruby
+- php
 

### 2020-08-10 inst/paper/main.tex
Reorganize as R package

--- 2020-08-10 main.tex
@@
 \section{Related Work}
-\eval has been widely discussed in the research literature for the challenges it
+eval has been widely discussed in the research literature for the challenges it
 imposes for code comprehension and automated code analyses.
-Perhaps, \citeit{EvalMenDo} is the most popular exposition of \eval and its
+Perhaps, EvalMenDo is the most popular exposition of eval and its
 challenges for JavaScript.

### 2020-08-11 inst/analysis/corpus-analysis.Rmd
Add corpus-analysis

---
title: "Corpus Analysis"
output: 
  html_document: 
    theme: united
    toc: yes
    toc_float: yes
editor_options: 
  chunk_output_type: console
params:
  data_dir: /var/lib/R/project-evalR/revalstudy/data
  num_packages: 500
  output_dir: /var/lib/R/project-evalR/revalstudy/data
  which_dependencies:
    - Depends
    - Imports
---

```{r setup, include=FALSE}
for (x in c(
  "dplyr", 
  "DT", 
  "fs", 
  "readr",
  "purrr"
  )) {
  suppressPackageStartupMessages(library(x, character.only=TRUE))
}

knitr::opts_chunk$set(echo = TRUE)

# TODO: move into the package
source("inc/paths.R")
source("inc/setup.R")
source("inc/latextags.R")

create_tags(path(TAGS_DIR, "corpus.tex"), prefix="", default=TRUE)
```

## Description

The aim of this notebook is to prepare the corpus of R packages to be included in the paper.

### Input

Nothing

### Output

- ``r CORPUS_FILE`` - the corpus in the corpus file (one package file line)
- ``r CORPUS_REVDEPS`` - CSV of package / reverse dependency pairs
- ``r CORPUS_REVDEPS_FILE`` - all unique reverse dependencies in the corpus file format

## Prepare the corpus

We are interested in TOP `r params$num_packages` packages based on the number of reverse dependencies.

```{r get all available packages}
# TODO this should be installed.packages() once they are installed
cran_packages_raw <- available.packages()
cran_packages <- as_tibble(cran_packages_raw) %>%
  select(package=Package, version=Version)
```

### Package reverse dependencies

There are two functions `package_dependencies` and `dependsOnPkgs`.
Not sure what is the difference, but the `dependsOnPkgs` is very slow.

```{r compute reverse dependencies}
pkgs_revdeps_raw <- 
  tools::package_dependencies(
    cran_packages$package,
    db=cran_packages_raw,
    which=params$which_dependencies,
    reverse=TRUE,
    recursive=FALSE
  ) %>% 
  imap_dfr(~tibble(package=.y, revdep=.x))

pkgs_revdeps <- 
  pkgs_revdeps_raw %>%
  count(package) %>%
  rename(revdeps=n)
```

### TOP packages based on number of reverse dependencies

```{r compute TOP n packages}
pkgs_top_revdeps <-
  pkgs_revdeps %>%
  top_n(n = params$num_packages, wt=revdeps) %>%
  arrange(desc(revdeps))

# cup the number of reverse dependencies to the required number
pkgs_top_revdeps <- head(pkgs_top_revdeps, min(nrow(pkgs_top_revdeps), params$num_packages))
```

### Missing packages from TOP 100 downloads

```{r get last month most downloaded packages}
pkgs_top_downloaded <-
  cranlogs::cran_top_downloads("last-month", 100)
```

```{r compute missing packages from the most downloaded ones}
missing <- setdiff(pkgs_top_downloaded$package, pkgs_top_revdeps$package)
missing_revdeps <- filter(pkgs_revdeps, package %in% missing)
missing_revdeps %>%
  left_join(pkgs_top_downloaded, by="package") %>%
  datatable()
```

### Corpus

```{r render selected packages}
pkgs_top_revdeps %>%
  left_join(pkgs_top_downloaded, by="package") %>%
  arrange(desc(revdeps)) %>%
  datatable()
```

```{r write down corpus}
write_lines(pkgs_top_revdeps$package, CORPUS_FILE)
write_csv(pkgs_revdeps, CORPUS_REVDEPS)
write_lines(unique(pkgs_revdeps$revdep), CORPUS_REVDEPS_FILE)
```

### 2020-08-11 inst/analysis/inc/paths.R

@@ -1,0 +1,6 @@
+## corpus only
+CORPUS_FILE                  <- path(params$output_dir, "corpus.txt")
+CORPUS_REVDEPS               <- path(params$output_dir, "corpus-revdeps.csv")
+CORPUS_REVDEPS_FILE          <- path(params$output_dir, "corpus-revdeps.txt")
+
+TAGS_DIR <- params$output_dir

### 2020-08-11 inst/analysis/inc/setup.R

@@ -1,0 +1,7 @@
+is_outlier_min <- function(x, m=1.5) quantile(x, 0.25) - m * IQR(x)
+
+is_outlier_max <- function(x, m=1.5) quantile(x, 0.75) + m * IQR(x)
+
+is_outlier <- function(x, m=1.5) {
+  (x < is_outlier_min(x, m)) | (x > is_outlier_max(x, m))
+}

### 2020-08-11 inst/analysis/analysis_stable.Rmd
Number of evals, of eval call sites, eval argument size

---
title: "analysis_stable"
output: 
  html_document:
    toc: true
    toc_float: true
    df_print: paged
date: "`r substr(Sys.time(), 1, 10)`"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.align = "center",
  fig.retina = 2,
  fig.width = 10,
  cache.lazy = FALSE
  )

library(tidyverse)
library(stringr)
library(viridis)
library(vctrs)

library(DT)

library(fst)

library(bench)

source("insights.R", local = knitr::knit_global())
```

All insights about eval calls are to put here, after the computations are stabilized (so use another notebook to explore).

# Setup

```{r eval_calls_file}
eval_calls_file <- "/var/lib/R/project-evalR/run/package-evals-traced-all.1/calls.fst"
```

```{r load_file}
eval_calls <- read_fst(eval_calls_file)
eval_calls <- as_tibble(eval_calls) 
```

```{r resolve_types}
eval_calls <- eval_calls %>% resolve_sexp_name(expr_expression_type) %>% 
                             resolve_sexp_name(expr_resolved_type) %>%
                             resolve_sexp_name(enclos_type) %>%
                             resolve_sexp_name(envir_type)
```

# Usage metrics

## Number of `eval`

For the number of call sites, we either use the `srcref` when there is one. When there is not (such as in *base*), we use a combination of `caller_package` and `caller_function`.

```{r}
nb_call_sites <- eval_calls %>% select(eval_call_srcref) %>% n_distinct(na.rm = TRUE)  + eval_calls %>% filter(is.na(eval_call_srcref)) %>% select(caller_package, caller_function)%>% n_distinct()
```

<details>
  <summary>More information about when there is no <code>srcref</code> </summary>

  We assume that:

  - the combination `caller_package` and `caller_function` is unique
  - there is only one call site per such `caller_function`

  It means that we get a lower bound.
</details>

There are `r nb_call_sites` different `eval` call sites.

### Per package

```{r eval_call_sites_base}
eval_calls %>% filter(caller_package == "base") %>% select(caller_function) %>% n_distinct(na.rm = TRUE)
```

<details>
  <summary> Comparing with <code>grep</code></summary>
  Grepping *base* to find eval call sites yields 40 `eval` call sites.
</details>

```{r eval_call_sites_per_package}
eval_calls %>% filter(caller_package != "base") %>% group_by(caller_package) %>% summarize(nb_call_sites =  n_distinct(eval_call_srcref, na.rm = TRUE)) %>% arrange(desc(nb_call_sites))
```

We probably forget the generated ones here, but should we count them in the _call sites_?

## Number of calls to `eval`

```{r}
nb_eval_calls <- eval_calls %>% nrow()
```

There are `r nb_eval_calls` eval calls.

```{r nb_eval_calls_per_package}
nb_eval_calls_per_package <- eval_calls %>% group_by(caller_package) %>% summarise(nb_eval_calls_package = n(), per_cent = nb_eval_calls_package / nb_eval_calls * 100) %>% arrange(desc(nb_eval_calls_package))
```

*base* has the highest number of eval calls, i.e. `r nb_eval_calls_per_package["per_cent", 1]`% of them.

```{r}
nb_eval_calls_per_package  %>% datatable()
```

## Size of expressions as arguments to `eval`

```{r expression_size}

eval_calls <- eval_calls %>% mutate(expr_expression_size = str_length(expr_expression), expr_resolved_size = str_length(expr_resolved), envir_expression_size = str_length(envir_expression), enclos_expression_size = str_length(enclos_expression))

```

```{r}
eval_calls %>% ggplot() +
  geom_histogram(aes(x = expr_expression_size), na.rm = TRUE, binwidth=1) +
  scale_x_log10() +  scale_y_sqrt()
```

If we remove the outliers:
```{r}
eval_calls %>% filter(expr_expression_size < 25000) %>% ggplot() +
  geom_histogram(aes(x = expr_expression_size), na.rm = TRUE, binwidth=2)  + scale_x_sqrt() +  scale_y_sqrt() 
```

<details>
  <summary>More details</summary>

What about the code size of expressions that seem to be generated?
An approximation of *generated* will be no `srcref` and not in `base`.

```{r code_size_srcref}
eval_calls %>% filter(is.na(eval_call_srcref), caller_package != "base") %>% ggplot() +
  geom_histogram(aes(x = expr_expression_size), na.rm = TRUE, binwidth=1) + scale_x_sqrt() + scale_y_sqrt()
```

And the average size per package:
```{r average_size_package}
eval_calls %>% group_by(caller_package) %>% summarize(average_expr_size = mean(expr_expression_size, na.rm = TRUE)) %>% arrange(desc(average_expr_size))
```

If we look at the maxima, the largest ones are in *base*:
```{r max_size_package}
eval_calls %>% group_by(caller_package) %>% summarize(max_expr_size = max(expr_expression_size, na.rm = TRUE)) %>% arrange(desc(max_expr_size))
```

</details>

## Amount of computations performed in `eval`

## Aliases of `eval`

# A taxonomy of `eval`

## The operation mix

## Scope (environments)

## Patterns

## Provenance

Some `eval` result from parsing text (directly or through a file) with `parse`.

```{r eval_from_parse}
parsed_evals <- eval_calls %>% filter(!is.na(expr_parsed))
```

There are `r parsed_evals %>% nrow()` such calls, `r (parsed_evals %>% nrow()) / nb_eval_calls * 100`%.

```{r, hide = TRUE}
nb_parsed_call_sites <- nb_eval_call_sites(parsed_evals)
```

In terms of call sites, there are `r nb_parsed_call_sites` i.e. `r nb_parsed_call_sites / nb_call_sites * 100`.

<details>
  <summary>More about <code>parse</code> </summary>

  We can differentiate further the provenance, i.e. whether the evaluated string comes from `file` or from `text`.

```{r parsed_evals_source}
parsed_evals <- parsed_evals %>% mutate(parse_args = map(expr_expression, function_arguments)) %>% unnest_wider(parse_args, names_repair = "unique")

```

</details>

## Consistence

# Other faces of `eval`

# Case studies

### 2020-08-14 inst/analysis/analysis_stable.Rmd
New dataset, expression size and so on

@@
 ```{r eval_calls_file}
-eval_calls_file <- "/var/lib/R/project-evalR/run/package-evals-traced-all.1/calls.fst"
+# All CRAN
+#eval_calls_file <- "/var/lib/R/project-evalR/run/package-evals-traced-all.1/calls.fst"
+
+# Subset (500 packages)
+eval_calls_file <- "/var/lib/R/project-evalR/run/package-evals-traced-corpus.1/calls.fst"
 ```
@@
 
-```{r expression_size}
-
-eval_calls <- eval_calls %>% mutate(expr_expression_size = str_length(expr_expression), expr_resolved_size = str_length(expr_resolved), envir_expression_size = str_length(envir_expression), enclos_expression_size = str_length(enclos_expression))
-
-```
-
 ```{r}
 eval_calls %>% ggplot() +
-  geom_histogram(aes(x = expr_expression_size), na.rm = TRUE, binwidth=1) +
+  geom_histogram(aes(x = expr_expression_length), na.rm = TRUE, binwidth=1) +
   scale_x_log10() +  scale_y_sqrt()
@@
 ```{r}
-eval_calls %>% filter(expr_expression_size < 25000) %>% ggplot() +
-  geom_histogram(aes(x = expr_expression_size), na.rm = TRUE, binwidth=2)  + scale_x_sqrt() +  scale_y_sqrt() 
+eval_calls %>% filter(expr_expression_length < 25000) %>% ggplot() +
+  geom_histogram(aes(x = expr_expression_length), na.rm = TRUE, binwidth=2)  + scale_x_sqrt() +  scale_y_sqrt() 
 ```
@@
 eval_calls %>% filter(is.na(eval_call_srcref), caller_package != "base") %>% ggplot() +
-  geom_histogram(aes(x = expr_expression_size), na.rm = TRUE, binwidth=1) + scale_x_sqrt() + scale_y_sqrt()
+  geom_histogram(aes(x = expr_expression_length), na.rm = TRUE, binwidth=1) + scale_x_sqrt() + scale_y_sqrt()
 ```
@@
 ```{r average_size_package}
-eval_calls %>% group_by(caller_package) %>% summarize(average_expr_size = mean(expr_expression_size, na.rm = TRUE)) %>% arrange(desc(average_expr_size))
+eval_calls %>% group_by(caller_package) %>% summarize(average_expr_size = mean(expr_expression_length, na.rm = TRUE)) %>% arrange(desc(average_expr_size))
 ```
@@
 ```{r max_size_package}
-eval_calls %>% group_by(caller_package) %>% summarize(max_expr_size = max(expr_expression_size, na.rm = TRUE)) %>% arrange(desc(max_expr_size))
+eval_calls %>% group_by(caller_package) %>% summarize(max_expr_size = max(expr_expression_length, na.rm = TRUE)) %>% arrange(desc(max_expr_size))
 ```
@@
 ```{r eval_from_parse}
-parsed_evals <- eval_calls %>% filter(!is.na(expr_parsed))
+parsed_evals <- eval_calls %>% filter(!is.na(expr_parsed_expression))
+nb_parsed_evals <- parsed_evals %>% nrow()
 ```
 
-There are `r parsed_evals %>% nrow()` such calls, `r (parsed_evals %>% nrow()) / nb_eval_calls * 100`%.
+There are `r nb_parsed_evals` such calls, `r nb_parsed_evals / nb_eval_calls * 100`%.
 
@@
 
-  We can differentiate further the provenance, i.e. whether the evaluated string comes from `file` or from `text`.
+  We can differentiate further the provenance, i.e.
+whether the evaluated string comes from `file` or from `text`.
+However, we can only do it for `parse`, `str2lang` and `str2expression` at the top level of `expr_expression`, not buried in a function call.
 
 ```{r parsed_evals_source}
-parsed_evals <- parsed_evals %>% mutate(parse_args = map(expr_expression, function_arguments)) %>% unnest_wider(parse_args, names_repair = "unique")
+parsed_evals_args <- parsed_evals %>% filter(str_detect(expr_parsed_expression, "(parse|str2lang|str2expression)\\(")) %>% mutate(parse_args = map(expr_parsed_expression, extract_args_parse)) %>% unnest_wider(parse_args,  names_sep = "_")
+```
 
+```{r}
+nb_file <- parsed_evals_args %>% filter(!is.na(parse_args_file)) %>% nrow()
 ```
+
+```{r}
+nb_text <- parsed_evals_args %>% filter(!is.na(parse_args_text)) %>% nrow()
+```
+
+There are `r nb_file` calls with the `file` argument, and `r nb_text` with the `text` argument, i.e. `r  nb_file / nb_parsed_evals * 100`%, and `r nb_text / nb_parsed_evals *100`% respectively.
 

### 2020-08-14 inst/analysis/insights.R

@@ -1,3 +1,16 @@
+library(rlang)
+
+
+nb_eval_call_sites <- function(eval_calls)
+{
+  # Two cases: with srcref, without srcref
+  
+  nb_with_srcref <- eval_calls %>% select(eval_call_srcref) %>% n_distinct(na.rm = TRUE)
+  nb_without_srcref <- eval_calls %>% filter(is.na(eval_call_srcref)) %>% select(caller_package, caller_function) %>% n_distinct(na.rm = TRUE)
+  
+  return(nb_with_srcref + nb_without_srcref)
+}
+
 
 get_expr <- function(eval_call)
 {
@@ -42,6 +55,20 @@ function_arguments <- function(eval_call)
     return(map_chr(as.list(exp[-1]), function(chr) { paste(deparse(chr), collapse = "\n")}))
   }
   return(NA)
+}
+
+# placeholder for the `parse_only` function of library xfun
+parse_only <- function(code) {}
+.myparse <- function(text) {}
+parse_all <- function(x, filename,allow_error) {}
+
+extract_args_parse <- function(eval_call) 
+{
+  exp <- get_expr(eval_call)
+  exp <- call_standardise(exp)
+  args <- map_chr(as.list(exp[-1]), function(chr) { paste(deparse(chr), collapse = "\n")})
+  
+  return(args[str_detect(names(args), "file|text|n|s|keep.source|srcfile")])# "file|text|n|s|prompt|keep.source|srcfile|code"
 }
 
 # See extract_inner_exp which takes a str

### 2020-08-14 inst/analysis/analysis_stable.Rmd
Call sites with parse, classification per type

@@
 
+### Classifying according to the type of the `expr` argument
+
+```{r types}
+per_expr_type <- eval_calls %>% count(expr_expression_type)
+```
+
+```{r}
+ per_expr_type %>% mutate(per_cent = n / nb_eval_calls * 100 ) %>% arrange(desc(n)) %>% knitr::kable()
+```
+
+```{r}
+per_expr_type %>% ggplot() +
+  geom_col(aes(x = fct_reorder(expr_expression_type, n), y = n)) +
+  scale_y_sqrt() + 
+  coord_flip() +
+  labs(title = "Calls per type", y = "n", x = "type" ) 
+```
+
+<details> 
+  <summary>Expressions left unchanged by <code>eval</code></summary>
+
+Expressions of types other than `call`,  `expression`, `name`, `bytecode` and `promise` are just returned unchanged.
+Here `language` refers to `call` , `name`to `symbol`.
+`bytecode` refers to bytecode objects, that are generated by the `compile`function:
+
+```{r}
+typeof(compiler::compile(1+1))
+```
+
+```{r not_unchanged_eval}
+changed_evals <- eval_calls %>% mutate(changed =  fct_collapse(expr_expression_type, base_types = c("LANGSXP", "EXPRSXP", "SYMSXP", "PROMSXP", "BCODESXP"))) %>% count(changed) %>% mutate(per_cent = n / nb_eval_calls * 100 )
+```
+
+We can remove `eval` and leave there the expression in `r changed_evals[2, "per_cent"][[1]]`% of the eval calls (i.e. `r changed_evals[2, "n"][[1]]`).
+
+</details>
+
 ## Provenance
@@
 
-In terms of call sites, there are `r nb_parsed_call_sites` i.e. `r nb_parsed_call_sites / nb_call_sites * 100`.
+In terms of call sites, there are `r nb_parsed_call_sites` i.e. `r nb_parsed_call_sites / nb_call_sites * 100`%.
 
@@
 ```{r}
-nb_file <- parsed_evals_args %>% filter(!is.na(parse_args_file)) %>% nrow()
+parsed_files <- parsed_evals_args %>% filter(!is.na(parse_args_file))
+nb_file <- parsed_files %>% nrow()
+nb_file_call_sites <- nb_eval_call_sites(parsed_files)
 ```
@@
 ```{r}
-nb_text <- parsed_evals_args %>% filter(!is.na(parse_args_text)) %>% nrow()
+parsed_texts <- parsed_evals_args %>% filter(!is.na(parse_args_text) | !is.na(parse_args_s))
+nb_text <-  parsed_texts %>% nrow()
+nb_text_call_sites <- nb_eval_call_sites(parsed_texts)
 ```
@@
 There are `r nb_file` calls with the `file` argument, and `r nb_text` with the `text` argument, i.e. `r  nb_file / nb_parsed_evals * 100`%, and `r nb_text / nb_parsed_evals *100`% respectively.
+
+There are `r nb_file_call_sites` call sites with the `file` argument, and `r nb_text_call_sites` with the `text` argument, i.e.
+`r  nb_file_call_sites / nb_parsed_call_sites * 100`%, and `r nb_text_call_sites / nb_parsed_call_sites *100`% respectively.
 

### 2020-08-14 inst/analysis/analysis_stable.Rmd
Description of column names

@@
                              resolve_sexp_name(envir_type)
+```
+
+# Description of the columns
+
+```{r, echo = FALSE}
+c_names <- names(eval_calls)
+column_names <- setNames(data.frame(matrix(ncol = length(c_names), nrow = 0)), c_names)
+
+column_names <- column_names %>% add_row()
+
+column_names["file"] <- "File where the `eval` is executed"
+column_names["eval_call_id"] <- "Unique id of a call" 
+column_names["expr_parsed_expression"] <- "Contains the expression with a call to `parse` if there is one in the `expr` argument somewhere in the call stack."
+
+# Pivoting
+
+column_names <- column_names %>% rownames_to_column() %>% 
+  pivot_longer(-rowname) %>%
+  pivot_wider(names_from=rowname, values_from=value)
+
+column_names <- column_names %>%  rename("Description" = `1`) %>% filter(!is.na(Description))
+
+```
+
+```{r, echo = FALSE}
+column_names %>% knitr::kable()
+
 ```

### 2020-08-14 inst/analysis/analysis_stable.Rmd
Update notebook header and organize package imports

@@
 ---
-title: "analysis_stable"
+title: "Empirical Evaluation of eval in R"
 output:
-  html_document:
-    toc: true
-    toc_float: true
+  rmdformats::readthedown:
+        code_folding: hide
+        lightbox: true
+        gallery: false
         df_print: paged
-date: "`r substr(Sys.time(), 1, 10)`"
+        toc_depth: 3
+date: "`r Sys.Date()`"
 editor_options: 
   chunk_output_type: console
+params:
+    data_filepath: "/var/lib/R/project-evalR/run/package-evals-traced-all.1/calls.fst"
 ---
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(
-  echo = TRUE,
+knitr::opts_chunk$set(echo = TRUE,
                       fig.align = "center",
@@
                       fig.width = 10,
-  cache.lazy = FALSE
-  )
+                      cache.lazy = FALSE)
 
@@
 library(vctrs)
-
 library(DT)
-
 library(fst)
-
-library(bench)
 
@@
 
-All insights about eval calls are to put here, after the computations are stabilized (so use another notebook to explore).
+# Data
 
-# Setup
-
-```{r eval_calls_file}
-# All CRAN
-#eval_calls_file <- "/var/lib/R/project-evalR/run/package-evals-traced-all.1/calls.fst"
-
-# Subset (500 packages)
-eval_calls_file <- "/var/lib/R/project-evalR/run/package-evals-traced-corpus.1/calls.fst"
-```
-
-```{r load_file}
-eval_calls <- read_fst(eval_calls_file)
-eval_calls <- as_tibble(eval_calls) 
-```
-
-```{r resolve_types}
-eval_calls <- eval_calls %>% resolve_sexp_name(expr_expression_type) %>% 
+```{r load-data, echo=FALSE}
+eval_calls <-
+    read_fst(params$data_filepath) %>%
+    as_tibble() %>%
+    resolve_sexp_name(expr_expression_type) %>%
     resolve_sexp_name(expr_resolved_type) %>%
@@
     resolve_sexp_name(envir_type)
-```
 
-# Description of the columns
-
-```{r, echo = FALSE}
 c_names <- names(eval_calls)
@@
 
+column_names %>% knitr::kable()
 ```
 
-```{r, echo = FALSE}
-column_names %>% knitr::kable()
+# Research Questions
 
-```
-
-# Usage metrics
-
-## Number of `eval`
+## RQ1: How many evals are there?
 

### 2020-08-14 inst/analysis/analysis_stable.Rmd
Add eval count and callsite distribution analysis

@@
 library(fst)
+library(fs)
 
@@
 # Data
+
+The analysis is being performed on dataset read from file `r params$data_filepath`.
+
+```{r file-info, echo=FALSE}
+file_info(params$data_filepath)
+```
 
@@
 
+# Corpus
+
+TODO
+
 # Research Questions
@@
 
-For the number of call sites, we either use the `srcref` when there is one. When there is not (such as in *base*), we use a combination of `caller_package` and `caller_function`.
+There are four kinds of `eval` in the `base` package. Their definitions are shown below.
 
-```{r}
-nb_call_sites <- eval_calls %>% select(eval_call_srcref) %>% n_distinct(na.rm = TRUE)  + eval_calls %>% filter(is.na(eval_call_srcref)) %>% select(caller_package, caller_function)%>% n_distinct()
+```{r, eval-definition, eval=FALSE}
+base::eval <-
+    function(expr, 
+             envir = parent.frame(), 
+             enclos = if (is.list(envir)||is.pairlist(envir)) parent.frame() else baseenv())
+        .Internal(eval(expr, envir, enclos))
+
+base::evalq <-
+    function (expr, 
+              envir = parent.frame(), 
+              enclos = if (is.list(envir) || is.pairlist(envir)) parent.frame() else baseenv())
+        .Internal(eval(substitute(expr), envir, enclos))
+
+base::eval.parent <-
+    function (expr, n = 1) {
+        p <- parent.frame(n + 1)
+        eval(expr, p)
+    }
+
+base::local <-
+    function (expr, envir = new.env())
+        eval.parent(substitute(eval(quote(expr), envir)))
 ```
 
-<details>
-  <summary>More information about when there is no <code>srcref</code> </summary>
+```{r eval-count-table}
+callsites_with_srcref <-
+    eval_calls %>% 
+    select(eval_call_srcref) %>% 
+    n_distinct(na.rm = TRUE)
 
+callsites_without_srcref <-
+    eval_calls %>% 
+    filter(is.na(eval_call_srcref)) %>% 
+    select(caller_package, caller_function) %>% 
+    n_distinct()
+
+eval_calls <-
+    eval_calls %>%
+    mutate(eval_call_srcref = coalesce(eval_call_srcref, 
+                                       str_c("missing", "::", caller_package, "::", caller_function)))
+eval_count_table <-
+    eval_calls %>%
+    group_by(eval_function, eval_call_srcref) %>%
+    summarize(count = n()) %>%
+    summarize(call_count = sum(count), callsite_count = n())
+
+nb_eval_calls <- total_call_count <- sum(eval_count_table$call_count)
+
+nb_call_sites <- total_callsite_count <- sum(eval_count_table$callsite_count)
+
+eval_count_table <-
+    eval_count_table %>%
+    mutate(call_proportion = 100 * call_count / total_call_count,
+           callsite_proportion = 100 * callsite_count / total_callsite_count) %>%
+    arrange(desc(call_count))
+
+eval_call_proportion <- 
+    eval_count_table %>%
+    filter(eval_function == "eval") %>%
+    pull(call_proportion)
+```
+
+There are `r NA` R function calls with `r total_call_count` calls to the four kinds of eval which accounts for `r NA`% of all R function calls.
+The table below summarizes the proportion of calls to the different kinds of evals.
+
+```{}
+eval_count_table %>%
+select(eval_function, call_count, call_proportion)
+datatable()
+```
+
+We observe that `r eval_call_proportion`% of the calls are made to the `eval` function.
+
+These `eval` calls originate from `r total_callsite_count` call sites.
+For the number of call sites, we use line number information attached to the AST by the parser.
+When it is not available (such as in *base*), we use a combination of `caller_package` and `caller_function`.
 We assume that:
-
   - the combination `caller_package` and `caller_function` is unique
   - there is only one call site per such `caller_function`
+This assumption results in a lower bound for the number of unique callsites for eval.
 
-  It means that we get a lower bound.
-</details>
+```{r call-site-table}
+eval_callsite_table <-
+    eval_calls %>%
+    count(eval_function, caller_package, eval_call_srcref, name = "call_count") %>%
+    group_by(eval_function, caller_package) %>%
+    summarize(call_count = sum(call_count), callsite_count = n()) %>%
+    ungroup()
 
-There are `r nb_call_sites` different `eval` call sites.
+eval_callsite_table %>%
+datatable()
+```
 
-### Per package
+```{r call-site-distribution}
+package_callsite_distribution <-
+    eval_callsite_table %>%
+    mutate(callsite_count = if_else(callsite_count <= 5, as.character(callsite_count), "> 5")) %>%
+    group_by(eval_function, callsite_count) %>%
+    summarize(package_count = n())
+
+package_callsite_distribution %>%
+datatable()
+
+p <-
+    package_callsite_distribution %>%
+    ggplot(aes(callsite_count, package_count)) +
+    geom_col() +
+    facet_wrap(vars(eval_function))
+p
+```
 
@@
 We probably forget the generated ones here, but should we count them in the _call sites_?
-
-## Number of calls to `eval`
-
-```{r}
-nb_eval_calls <- eval_calls %>% nrow()
-```
-
-There are `r nb_eval_calls` eval calls.
 

### 2020-08-17 inst/analysis/analysis_stable.Rmd
Add tag to code block
Documenting columns. Fix error line 159 (missing %>%)

@@
 
+# A flexible way to document the dataset: it is quick and easy to document a new column
 c_names <- names(eval_calls)
@@
 column_names["expr_parsed_expression"] <- "Contains the expression with a call to `parse` if there is one in the `expr` argument somewhere in the call stack."
+column_names["caller_function"] <- "Function in which the `eval` is called."
+column_names["caller_package"] <- "Package in which the `eval` is called."
+column_names["eval_function"] <- "One of `eval`, `evalq`, `eval.parent`, and `local`."
+column_names["eval_call_expression"] <- "The full expression with `eval` that has been traced."
+column_names["eval_call_srcref"] <- "File name and line where the `eval` call is. Might ne NA, especially for calls originating from package base."
+column_names["caller_expression"] <- "Expression in which the `eval` is called."
+column_names["caller_srcref"] <- "File name and line where the caller of the `eval` is located. Might be NA (especially for callers in base)."
+column_names["caller_srcref2"] <- "File name and line for some calls for which the caller seemed to be itself (mainly Rcpp and C calls)."
 
@@
 
-```{}
+```{r }
 eval_count_table %>%
-select(eval_function, call_count, call_proportion)
+select(eval_function, call_count, call_proportion) %>% 
 datatable()
@@
 We assume that:
+
   - the combination `caller_package` and `caller_function` is unique

### 2020-08-17 inst/analysis/analysis_stable.Rmd
Fix duplicated chunck name
Use the 500 package dataset

@@
 params:
-    data_filepath: "/var/lib/R/project-evalR/run/package-evals-traced-all.1/calls.fst"
+    data_filepath: "/var/lib/R/project-evalR/run/package-evals-traced-corpus.1/calls.fst"
+    #data_filepath: "/var/lib/R/project-evalR/run/package-evals-traced-all.1/calls.fst"
+    # The trace data on all CRAN does not have all the columns of the smaller corpus. The analysis will fail for the big corpus.
 ---
@@
 column_names["caller_srcref2"] <- "File name and line for some calls for which the caller seemed to be itself (mainly Rcpp and C calls)."
+# To continue
 
@@
 
-<details> 
-  <summary>Expressions left unchanged by <code>eval</code></summary>
-
 Expressions of types other than `call`,  `expression`, `name`, `bytecode` and `promise` are just returned unchanged.
@@
 
-</details>
-
 ## Provenance
@@
 
-```{r, hide = TRUE}
+```{r, echo = FALSE}
 nb_parsed_call_sites <- nb_eval_call_sites(parsed_evals)
@@
 In terms of call sites, there are `r nb_parsed_call_sites` i.e. `r nb_parsed_call_sites / nb_call_sites * 100`%.
-
-<details>
-  <summary>More about <code>parse</code> </summary>
 
@@
 
-</details>
-
 ## Consistence

### 2020-08-17 inst/analysis/analysis_stable.Rmd
The case of `base`

@@
 
+To asses how precise the call site estimate is for calls that do not have `srcref`, we compare it on *base* with the actual count given by `grep`.
+
 ```{r eval_call_sites_base}
-eval_calls %>% filter(caller_package == "base") %>% select(caller_function) %>% n_distinct(na.rm = TRUE)
+nb_call_site_base <- eval_calls %>% filter(caller_package == "base") %>% select(caller_function) %>% n_distinct(na.rm = TRUE)
 ```
+The estimate for *base* is `r nb_call_site_base` call sites, while `grep` yields 40 `eval` call sites.
 
-<details>
-  <summary> Comparing with <code>grep</code></summary>
-  Grepping *base* to find eval call sites yields 40 `eval` call sites.
-</details>
-
+For the other packages, ranked by call sites:
 ```{r eval_call_sites_per_package}
@@
 
+What are the packages with the highest number of `eval` calls?
+
 ```{r nb_eval_calls_per_package}
@@
 
-*base* has the highest number of eval calls, i.e. `r nb_eval_calls_per_package["per_cent", 1]`% of them.
+*base* has the highest number of eval calls, i.e. `r nb_eval_calls_per_package[1, "per_cent"][[1]]`% of them.
 
@@
 nb_eval_calls_per_package  %>% datatable()
+```
+
+### Excluding *base* from the data set
+
+We tried to replace the 40 `eval` call sites in *base* and could do it in about 50% of the cases. The rest of the cases are legitimate uses of `eval`, basically variations of `eval`.
+
+TODO: show some examples, replacements and non replacements
+
+We argue that `eval` in *base* should be considered apart from the other `eval` calls.
+
+`eval` in *base* does not use `parse`, so does not take strings potentially coming from users as input,  and thus are deemed to be _safe_.
+
+```{r parse_in_base}
+nb_parse_base <- eval_calls %>% filter(!is.na(expr_parsed_expression), caller_package == "base")
 ```

### 2020-08-18 inst/analysis/analysis_stable.Rmd
Htmnl is now generated in the html/ subfolder
Extract package name from srcref for the eval call expression

@@
 params:
-    data_filepath: "/var/lib/R/project-evalR/run/package-evals-traced-corpus.1/calls.fst"
+    data_filepath: "/var/lib/R/project-evalR/run/package-evals-traced-corpus.2/calls.fst"
     #data_filepath: "/var/lib/R/project-evalR/run/package-evals-traced-all.1/calls.fst"
     # The trace data on all CRAN does not have all the columns of the smaller corpus. The analysis will fail for the big corpus.
+
+# This makes it possible to generate the html file in the html subfolder
+knit: (function(inputFile, encoding) {
+  rmarkdown::render(inputFile, encoding = encoding, output_dir = "html") })
 ---
@@
 column_names["caller_srcref"] <- "File name and line where the caller of the `eval` is located. Might be NA (especially for callers in base)."
-column_names["caller_srcref2"] <- "File name and line for some calls for which the caller seemed to be itself (mainly Rcpp and C calls)."
+# The following ones were in package-evals-traced-corpus.1/calls.fst
+#column_names["caller_srcref2"] <- "File name and line for some calls for which the caller seemed to be itself (mainly Rcpp and C calls)."
+#column_names["caller_expression_2"] <- "Caller expression when the caller seems to call itself (see `caller_srcref2`)."
+#column_names["caller_expression_2"] <- "Same as `caller_expression_2` but uses `format` to print, not `expr_to_string`."
+column_names["caller_stack_expression"] <-"When `eval_function` == `caller_function`, climb up the stack to find the first different caller."
+column_names["caller_stack_expression_raw"] <- "See `caller_stack_expression`. Does not use `expr_to_str` but `format`."
+column_names["caller_stack_expression_srcref"] <- "srcref for `caller_stack_expression."
 # To continue
@@
 
+```
+
+The columns of the dataset are the following ones:
+```{r column_names}
 column_names %>% knitr::kable()
@@
 ```{r parse_in_base}
-nb_parse_base <- eval_calls %>% filter(!is.na(expr_parsed_expression), caller_package == "base")
+parsed_base <- eval_calls %>% filter(!is.na(expr_parsed_expression), caller_package == "base")
+
+nb_call_site_parse_base <-  nb_eval_call_sites(parsed_base)
+```
+
+There are `r nrow(parsed_base)` `eval` calls with `parse` in *base*, which is `r nrow(parsed_base) / nrow(filter(eval_calls, caller_package == "base"))`%.
+In terms of call sites, there are `r nb_call_site_parse_base` such ones, i.e.
+`r nb_call_site_parse_base / nb_call_site_base * 100`.
+
+They are the following ones:
+
+```{r call_site_parse_base}
+parsed_base %>% count(file, caller_function, caller_expression, expr_parsed_expression) %>% 
+  arrange(desc(n)) %>% knitr::kable()
 ```

### 2020-08-18 inst/analysis/insights.R

@@ -204,3 +204,38 @@ resolve_sexp_name <- function(df, var) {
     select(-!!en_var) %>%
     rename(!!en_var:=name)
 }
+
+
+extract_package_name <- function(src_ref)
+{
+  # There are 5 possibilities for a srcref:
+  # - NA
+  # - /tmp/Rtmp..../R.INSTALL....../packagename/R/file:linenumbers
+  # - /mnt/nvme0/R/project-evalR/library/4.0/instrumentr/srcref/packagename/4.0.4/file:linenumbers
+  # - /R/* : core packages (we cannot distinguish between them yet so we write core for the package name)
+  # - /testit/... or /testthat/... : it is the testit or testthat packages
+  if(is.na(src_ref))
+  {
+    return("base")
+  }
+  else if(str_starts(src_ref, fixed("./R/")))
+  {
+    return("core")
+  }
+  else if(str_starts(src_ref, fixed("/tmp/")))
+  {
+    return(str_match(src_ref, "/tmp/Rtmp[^/]*/R\\.INSTALL[^/]*/([^/]+)/.*")[[2]])
+  }
+  else if(str_starts(src_ref, fixed("/mnt/nvme0/")))
+  {
+    return(str_match(src_ref, "/mnt/nvme0/R/project-evalR/library/4.0/instrumentr/srcref/([^/]*)/.*")[[2]])
+  }
+  else if(str_starts(src_ref, fixed("test")))
+  {
+    return(str_match(src_ref, "([^/]*)/.*")[[2]])
+  }
+  else
+  {
+    return("unknown")
+  }
+}

### 2020-08-18 inst/analysis/analysis_stable.Rmd
Better description of `eval` in base ( and occurrences of `parse`)

@@
     as_tibble() %>%
+    mutate(eval_call_package = map_chr(eval_call_srcref, extract_package_name)) %>% 
     resolve_sexp_name(expr_expression_type) %>%
@@
 column_names["caller_stack_expression_raw"] <- "See `caller_stack_expression`. Does not use `expr_to_str` but `format`."
-column_names["caller_stack_expression_srcref"] <- "srcref for `caller_stack_expression."
+column_names["caller_stack_expression_srcref"] <- "srcref for `caller_stack_expression`."
 # To continue
@@
 
-We tried to replace the 40 `eval` call sites in *base* and could do it in about 50% of the cases. The rest of the cases are legitimate uses of `eval`, basically variations of `eval`.
+When using `grep` , we find 30 call sites that use `eval` in *base*.
+We tried to replace the 30 `eval` call sites in *base* and could do it in  50% of the cases.
 
-TODO: show some examples, replacements and non replacements
+We reckon that `eval` in *base* should be considered apart from the other `eval` calls.
 
-We argue that `eval` in *base* should be considered apart from the other `eval` calls.
+`eval` in *base* does not consume user input, i.e. input coming from `parse`, except in 5 legitimate cases, such as the `sys.source` function that is used to `eval` a file.
 
-`eval` in *base* does not use `parse`, so does not take strings potentially coming from users as input,  and thus are deemed to be _safe_.
+| Function name              | File           | Line | Parse ? | Comments                                                                           |
+|----------------------------|----------------|------|---------|------------------------------------------------------------------------------------|
+| autoload                   | autoload       | 25   |         |                                                                                    |
+| autoloader                 | autoload       | 43   |         | To execute a call to library                                                       |
+| autoloader                 | autoload       | 49   |         |                                                                                    |
+| bquote                     | backquote      | 33   |         | In subfunction unquote. To eval what is in .() . Metaprogramming                   |
+| bquote                     | backquote      | 56   |         | idem (in unquote.list, defined inside bquote)                                      |
+| by.default                 | by             | 35   |         | eval(substitute pattern?                                                           |
+| by.data.frame              | by             | 50   |         | idem                                                                               |
+| invokeRestartInteractively | conditions     | 245  | yes     | prompt                                                                             |
+| invokeRestartInteractively | conditions     | 249  | yes     | prompt                                                                             |
+| Ops.data.frame             | dataframe      | 1678 |         | for arithmetic operators on data frames ; do.call could be probably used here also |
+| dget                       | dput           | 32   | yes     | from file                                                                          |
+| eval                       | eval           | 26   |         | Call Internal eval                                                                 |
+| eval.parent                | eval           | 30   |         |                                                                                    |
+| evalq                      | eval           | 36   |         | evalq is basically eval(substitute                                                 |
+| local                      | eval           | 49   |         |                                                                                    |
+| with.default               | eval           | 57   |         |                                                                                    |
+| within.data.frame          | eval           | 63   |         |                                                                                    |
+| within.list                | eval           | 77   |         | evalq                                                                              |
+| replicate                  | replicate      | 21   |         |                                                                                    |
+| subset.data.frame          | frametools     | 25   |         |                                                                                    |
+| subset.data.frame          | frametools     | 34   |         |                                                                                    |
+| subset.matrix              | frametools     | 56   |         |                                                                                    |
+| transform.data.frame       | frametools     | 69   |         |                                                                                    |
+| match.arg                  | match          | 45   |         |                                                                                    |
+| char.expand                | match          | 79   |         |                                                                                    |
+| max.col                    | max            | 22   |         | To access formals                                                                  |
+| parseNamespaceFile         | namespace      | 1315 | yes     | To parse directives                                                                |
+| idem                       | namespace      | 1319 | yes     |                                                                                    |
+| idem                       | namespace      | 1442 | yes     | evaluate a call from a parse                                                       |
+| source                     | source         | 123  | yes     | Use the internal parse function (here it is a comment)                             |
+| source                     | source         | 170  | yes     | comment                                                                            |
+| source                     | source         | 219  | yes     | the actual eval                                                                    |
+| sys.source                 | source         | 277  | yes     | the non-internal parse is used                                                     |
+| stopifnot                  | stop           | 54   |         | change of caller pattern                                                           |
+| as.data.frame.table        | table          | 244  |         | Build a call?                                                                      |
+| match.fun                  | match.fun      | 28   |         | eval.parent                                                                        |
+| trace                      | methodsSupport | 38   |         | eval.parent                                                                        |
+| untrace                    | methodsSupport | 53   |         | eval.parent                                                                        |
+| .doTrace                   | methodsSupport | 78   |         | eval.parent                                                                        |
 
 ```{r parse_in_base}
-parsed_base <- eval_calls %>% filter(!is.na(expr_parsed_expression), caller_package == "base")
+parsed_base <- eval_calls %>% filter(!is.na(expr_parsed_expression), eval_call_package == "base")
 
@@
 
-There are `r nrow(parsed_base)` `eval` calls with `parse` in *base*, which is `r nrow(parsed_base) / nrow(filter(eval_calls, caller_package == "base"))`%.
-In terms of call sites, there are `r nb_call_site_parse_base` such ones, i.e.
-`r nb_call_site_parse_base / nb_call_site_base * 100`.
-
-They are the following ones:
-
-```{r call_site_parse_base}
-parsed_base %>% count(file, caller_function, caller_expression, expr_parsed_expression) %>% 
-  arrange(desc(n)) %>% knitr::kable()
-```
+There are `r nrow(parsed_base)` `eval` calls with `parse` in *base*, which is `r nrow(parsed_base) / nrow(filter(eval_calls, eval_call_package == "base"))`%.
 

### 2020-08-18 inst/analysis/analysis_stable.Rmd
Classification of language expr (!st argument of eval)

@@
 
+We look at the first argument of `eval`, the `expr` argument.
+It is the expression to be evaluated.
+It can be of different types, mainly usual numeric types, vectors, strings, and language types: calls, symbols, promise, bytecode and expressions.
+
+`bytecode` refers to bytecode objects, that are generated by the `compile`function:
+
+```{r}
+typeof(compiler::compile(1+1))
+```
+
+Type `expression` is a sequence of calls and symbols.
+
 ```{r types}
@@
 Expressions of types other than `call`,  `expression`, `name`, `bytecode` and `promise` are just returned unchanged.
-Here `language` refers to `call` , `name`to `symbol`.
-`bytecode` refers to bytecode objects, that are generated by the `compile`function:
-
-```{r}
-typeof(compiler::compile(1+1))
-```
 
@@
 We can remove `eval` and leave there the expression in `r changed_evals[2, "per_cent"][[1]]`% of the eval calls (i.e. `r changed_evals[2, "n"][[1]]`).
+
+### `expr` argument when it is a call
+
+```{r expr_call_arg}
+expr_call_arg <- eval_calls %>% filter(expr_expression_type == "LANGSXP") %>% 
+  mutate(function_expr = map_chr(expr_expression, function_name))
+```
+
+The following top level functions are used:
+
+```{r top_level_function_expr}
+expr_call_arg %>% count(function_expr) %>% arrange(desc(n))
+```
+
+We can distinguish a few groups:
+
+- assignments
+- reading (in a data frame, a vector)
+- parsing
+- defusing (`quote` and so on)
+- conversion or creation of new values
+- building calls and calling
+- working with environments
+- class assignments, reads and definitions
+- building logical values
+- formula
+- meta programming on functions (getting access to the body of the function, to the call stack)
+
+```{r group_expr_functions}
+group_expr_functions <- expr_call_arg %>% mutate(group_function_expr = fct_collapse(function_expr, 
+                          assignment = c("<-", "[[<-", "[<-", "$<-"),
+                          read = c("[", "[[", "$"),
+                          parsing = c("parse", "str2lang", "str2expression", "parse_only", "base::parse"),
+                          defusing = c("quote", "substitute", "bquote", "expr", "quo"),
+                          conversion = c("as.list", "expression", "as.name", "as.expression", "list", "as.data.frame", "as.character"),
+                          calling = c("call", "as.call", "do.call"),
+                          environment = c("environment", "as.environment"),
+                          classAssign = c("@<-"),
+                          classRead = c("slot", "@"),
+                          classDef = c(".Primitive(\"class\")"),
+                          logical = c("<", ">", "is.element"),
+                          scope = c("{"),
+                          choice = c("if"),
+                          function_meta = c("sys.calls", "function", "body"),
+                          formula = c("~"), # Add mathematical operators
+                          other_level = "other" ))
+                          #TODO: add anonymous function applications
+```
+
+```{r}
+group_expr_functions %>% count(group_function_expr) %>% arrange(desc(n)) %>% knitr::kable()
+```
+
+```{r}
+group_expr_functions %>% ggplot() +
+  geom_bar(aes(x = fct_rev(fct_infreq(group_function_expr)))) +
+  scale_y_sqrt(n.breaks = 15) +
+  coord_flip() +
+  labs(title = "Category of calls", y = "n", x = "category" ) 
+```
 

### 2020-08-18 inst/analysis/analysis_stable.Rmd
Facet wrap on eval_function for the classification

--- 2020-08-18 inst/analysis/analysis_stable.Rmd
@@
 We can remove `eval` and leave there the expression in `r changed_evals[2, "per_cent"][[1]]`% of the eval calls (i.e. `r changed_evals[2, "n"][[1]]`).
+
+### `expr` argument when it is a call
+
+```{r expr_call_arg}
+expr_call_arg <- eval_calls %>% filter(expr_expression_type == "LANGSXP") %>% 
+  mutate(function_expr = map_chr(expr_expression, function_name))
+```
+
+The following top level functions are used:
+
+```{r top_level_function_expr}
+expr_call_arg %>% count(function_expr) %>% arrange(desc(n))
+```
+
+We can distinguish a few groups:
+
+- assignments
+- reading (in a data frame, a vector)
+- parsing
+- defusing (`quote` and so on)
+- conversion or creation of new values
+- building calls and calling
+- working with environments
+- class assignments, reads and definitions
+- building logical values
+- formula
+- meta programming on functions (getting access to the body of the function, to the call stack)
+- 'NA' corresponds to no function calls (or that it was not deparsed successfully)
+
+```{r group_expr_functions}
+group_expr_functions <- expr_call_arg %>% mutate(group_function_expr = fct_collapse(function_expr, 
+                          assignment = c("<-", "[[<-", "[<-", "$<-"),
+                          read = c("[", "[[", "$"),
+                          parsing = c("parse", "str2lang", "str2expression", "parse_only", "base::parse"),
+                          defusing = c("quote", "substitute", "bquote", "expr", "quo"),
+                          expressions = c("as.list", "expression", "as.name", "as.expression", "list", "as.data.frame", "as.character", "{", "modify_lang"),
+                          calling = c("call", "as.call", "do.call"),
+                          environment = c("environment", "as.environment"),
+                          classAssign = c("@<-"),
+                          classRead = c("slot", "@"),
+                          classDef = c(".Primitive(\"class\")"),
+                          logical = c("<", ">", "is.element"),
+                          choice = c("if"),
+                          function_meta = c("sys.calls", "function", "body"),
+                          formula = c("~", "deriv", "D", "model.frame", "rootftfunc", "contest.lmerModLmerTest", "JacFunc", "gamlss.family", "contest1D.lmerModLmerTest", "contestMD.lmerModLmerTest", "reexpr"),
+                          other_level = "other" ))
+                          #TODO: add anonymous function applications
+```
+
+```{r}
+group_expr_functions %>% count(group_function_expr) %>% arrange(desc(n)) %>% knitr::kable()
+```
+
+```{r}
+group_expr_functions %>% ggplot() +
+  geom_bar(aes(x = fct_rev(fct_infreq(group_function_expr)))) +
+  scale_y_sqrt() +
+  coord_flip() +
+  labs(title = "Category of calls", y = "n", x = "category" ) 
+```
+
+```{r }
+group_expr_functions %>% ggplot() +
+  geom_bar(aes(x = fct_rev(fct_infreq(group_function_expr)))) +
+  scale_y_sqrt() +
+  coord_flip() +
+  labs(title = "Category of calls", y = "n", x = "category" ) +
+  facet_wrap(vars(eval_function))
+```
 

### 2020-08-19 inst/analysis/analysis_stable.Rmd
Type classification for expr_resolved

@@
 ```{r not_unchanged_eval}
-changed_evals <- eval_calls %>% mutate(changed =  fct_collapse(expr_expression_type, base_types = c("LANGSXP", "EXPRSXP", "SYMSXP", "PROMSXP", "BCODESXP"))) %>% count(changed) %>% mutate(per_cent = n / nb_eval_calls * 100 )
+changed_evals <- eval_calls %>% mutate(changed =  fct_collapse(expr_expression_type, lang_types = c("LANGSXP", "EXPRSXP", "SYMSXP", "PROMSXP", "BCODESXP"), other_level = "other_types")) %>% count(changed) %>% mutate(per_cent = n / nb_eval_calls * 100 )
 ```
@@
 - meta programming on functions (getting access to the body of the function, to the call stack)
-- 'NA' corresponds to no function calls (or that it was not deparsed successfully)
+- 'NA' corresponds to no function calls (or that it was not parsed successfully)
 
@@
 ```
+
+### For resolved `expr`
+
+We can do the same for resolved expressions:
+
+```{r types_resolved}
+per_expr_resolved_type <- eval_calls %>% count(expr_resolved_type)
+```
+
+```{r}
+ per_expr_resolved_type %>% mutate(per_cent = n / nb_eval_calls * 100 ) %>% arrange(desc(n)) %>% knitr::kable()
+```
+
+```{r}
+per_expr_resolved_type %>% ggplot() +
+  geom_col(aes(x = fct_reorder(expr_resolved_type, n), y = n)) +
+  scale_y_sqrt() + 
+  coord_flip() +
+  labs(title = "Calls per type", y = "n", x = "type" ) 
+```
+
+There is a much larger variety of types for the resolved expressions.
+
+```{r not_unchanged_resolved_eval}
+changed_resolved_evals <- eval_calls %>% mutate(changed =  fct_collapse(expr_resolved_type, lang_types = c("LANGSXP", "EXPRSXP", "SYMSXP", "PROMSXP", "BCODESXP"), other_level = "other_types")) %>% count(changed) %>% mutate(per_cent = n / nb_eval_calls * 100 )
+```
+
+We can remove `eval` and leave there the expression in `r changed_resolved_evals[2, "per_cent"][[1]]`% of the eval calls (i.e. `r changed_resolved_evals[2, "n"][[1]]`).
 

### 2020-08-19 inst/analysis/analysis_stable.Rmd
Classification of calls on expr_resolved

@@
 params:
-    data_filepath: "/var/lib/R/project-evalR/run/package-evals-traced-corpus.2/calls.fst"
+    data_filepath: "/var/lib/R/project-evalR/run/package-evals-traced-corpus.3/calls.fst"
     #data_filepath: "/var/lib/R/project-evalR/run/package-evals-traced-all.1/calls.fst"
@@
 - 'NA' corresponds to no function calls (or that it was not parsed successfully)
+- anonymous functions
 
 ```{r group_expr_functions}
-group_expr_functions <- expr_call_arg %>% mutate(group_function_expr = fct_collapse(function_expr, 
+group_expr_functions <- expr_call_arg %>%
+                            mutate(function_expr2 = map_chr(detect_anonymous_func, function_expr)) %>%
+                            mutate(group_function_expr = fct_collapse(function_expr2, 
                               assignment = c("<-", "[[<-", "[<-", "$<-"),
@@
                               choice = c("if"),
+                              anonymous = c("anonymous"),
                               function_meta = c("sys.calls", "function", "body"),
@@
                               other_level = "other" ))
-                          #TODO: add anonymous function applications
 ```
@@
 
+We can do the same analysis as for the non-resolved expressions:
+
+```{r expr_resolved__call_arg}
+expr_resolved_call_arg <- eval_calls %>% filter(expr_resolved_type == "LANGSXP") %>% 
+  mutate(function_expr = map_chr(expr_resolved, function_name))
+```
+
+The following top level functions are used:
+
+```{r top_level_function_resolved_expr}
+expr_resolved_call_arg %>% count(function_expr) %>% arrange(desc(n))
+```
+
+```{r group_expr_resolved_functions}
+group_expr_resolved_functions <- expr_resolved_call_arg %>%
+                            mutate(function_expr2 = map_chr(detect_anonymous_func, function_expr)) %>%
+                            mutate(group_function_expr = fct_collapse(function_expr2, 
+                              assignment = c("<-", "[[<-", "[<-", "$<-"),
+                              read = c("[", "[[", "$"),
+                              parsing = c("parse", "str2lang", "str2expression", "parse_only", "base::parse"),
+                              defusing = c("quote", "substitute", "bquote", "expr", "quo"),
+                              expressions = c("as.list", "expression", "as.name", "as.expression", "list", "as.data.frame", "as.character", "{", "modify_lang"),
+                              calling = c("call", "as.call", "do.call"),
+                              environment = c("environment", "as.environment"),
+                              classAssign = c("@<-"),
+                              classRead = c("slot", "@"),
+                              classDef = c(".Primitive(\"class\")"),
+                              logical = c("<", ">", "is.element"),
+                              choice = c("if"),
+                              anonymous = c("anonymous"),
+                              function_meta = c("sys.calls", "function", "body"),
+                              formula = c("~", "deriv", "D", "model.frame", "rootftfunc", "contest.lmerModLmerTest", "JacFunc", "gamlss.family", "contest1D.lmerModLmerTest", "contestMD.lmerModLmerTest", "reexpr"),
+                              other_level = "other" ))
+```
+
+```{r}
+group_expr_resolved_functions %>% count(group_function_expr) %>% arrange(desc(n)) %>% knitr::kable()
+```
+
+```{r}
+group_expr_resolved_functions %>% ggplot() +
+  geom_bar(aes(x = fct_rev(fct_infreq(group_function_expr)))) +
+  scale_y_sqrt() +
+  coord_flip() +
+  labs(title = "Category of calls", y = "n", x = "category" ) 
+```
+
+```{r }
+group_expr_functions %>% ggplot() +
+  geom_bar(aes(x = fct_rev(fct_infreq(group_function_expr)))) +
+  scale_y_sqrt() +
+  coord_flip() +
+  labs(title = "Category of calls", y = "n", x = "category" ) +
+  facet_wrap(vars(eval_function))
+```
+
 ## Provenance

### 2020-08-19 inst/analysis/enrich_dataset.Rmd
A notebook to add long computing columns to the dataset

--- 2020-08-18 inst/analysis/analysis_stable.Rmd
@@
 ---
-title: "Empirical Evaluation of eval in R"
+title: "Enrich the datasets with new columns."
 output:
@@
 params:
-    data_filepath: "/var/lib/R/project-evalR/run/package-evals-traced-corpus.2/calls.fst"
-    #data_filepath: "/var/lib/R/project-evalR/run/package-evals-traced-all.1/calls.fst"
-    # The trace data on all CRAN does not have all the columns of the smaller corpus. The analysis will fail for the big corpus.
-
-# This makes it possible to generate the html file in the html subfolder
+  data_dirpath: "/var/lib/R/project-evalR/run/package-evals-traced-corpus.3/"
+  data_filepath: "/var/lib/R/project-evalR/run/package-evals-traced-corpus.3/calls.fst"
 knit: (function(inputFile, encoding) {
@@
 source("insights.R", local = knitr::knit_global())
+
+source("inc/latextags.R")
+create_tags(path("/var/lib/R/project-evalR/revalstudy/data", "analysisStable.tex"), prefix="", default=TRUE)
+```
+
+```{r setup, include=FALSE}
+knitr::opts_chunk$set(echo = TRUE)
 ```
@@
 
-# A flexible way to document the dataset: it is quick and easy to document a new column
-c_names <- names(eval_calls)
-column_names <- setNames(data.frame(matrix(ncol = length(c_names), nrow = 0)), c_names)
-
-column_names <- column_names %>% add_row()
-
-column_names["file"] <- "File where the `eval` is executed"
-column_names["eval_call_id"] <- "Unique id of a call" 
-column_names["expr_parsed_expression"] <- "Contains the expression with a call to `parse` if there is one in the `expr` argument somewhere in the call stack."
-column_names["caller_function"] <- "Function in which the `eval` is called."
-column_names["caller_package"] <- "Package in which the `eval` is called."
-column_names["eval_function"] <- "One of `eval`, `evalq`, `eval.parent`, and `local`."
-column_names["eval_call_expression"] <- "The full expression with `eval` that has been traced."
-column_names["eval_call_srcref"] <- "File name and line where the `eval` call is. Might ne NA, especially for calls originating from package base."
-column_names["caller_expression"] <- "Expression in which the `eval` is called."
-column_names["caller_srcref"] <- "File name and line where the caller of the `eval` is located. Might be NA (especially for callers in base)."
-# The following ones were in package-evals-traced-corpus.1/calls.fst
-#column_names["caller_srcref2"] <- "File name and line for some calls for which the caller seemed to be itself (mainly Rcpp and C calls)."
-#column_names["caller_expression_2"] <- "Caller expression when the caller seems to call itself (see `caller_srcref2`)."
-#column_names["caller_expression_2"] <- "Same as `caller_expression_2` but uses `format` to print, not `expr_to_string`."
-column_names["caller_stack_expression"] <-"When `eval_function` == `caller_function`, climb up the stack to find the first different caller."
-column_names["caller_stack_expression_raw"] <- "See `caller_stack_expression`. Does not use `expr_to_str` but `format`."
-column_names["caller_stack_expression_srcref"] <- "srcref for `caller_stack_expression`."
-# To continue
-
-# Pivoting
-
-column_names <- column_names %>% rownames_to_column() %>% 
-  pivot_longer(-rowname) %>%
-  pivot_wider(names_from=rowname, values_from=value)
-
-column_names <- column_names %>%  rename("Description" = `1`) %>% filter(!is.na(Description))
-
 ```
-
-The columns of the dataset are the following ones:
-```{r column_names}
-column_names %>% knitr::kable()
-```
-
-# Corpus
-
-TODO
-
-# Research Questions
-
-## RQ1: How many evals are there?
-
-There are four kinds of `eval` in the `base` package. Their definitions are shown below.
-
-```{r, eval-definition, eval=FALSE}
-base::eval <-
-    function(expr, 
-             envir = parent.frame(), 
-             enclos = if (is.list(envir)||is.pairlist(envir)) parent.frame() else baseenv())
-        .Internal(eval(expr, envir, enclos))
-
-base::evalq <-
-    function (expr, 
-              envir = parent.frame(), 
-              enclos = if (is.list(envir) || is.pairlist(envir)) parent.frame() else baseenv())
-        .Internal(eval(substitute(expr), envir, enclos))
-
-base::eval.parent <-
-    function (expr, n = 1) {
-        p <- parent.frame(n + 1)
-        eval(expr, p)
-    }
-
-base::local <-
-    function (expr, envir = new.env())
-        eval.parent(substitute(eval(quote(expr), envir)))
-```
-
-```{r eval-count-table}
-callsites_with_srcref <-
-    eval_calls %>% 
-    select(eval_call_srcref) %>% 
-    n_distinct(na.rm = TRUE)
-
-callsites_without_srcref <-
-    eval_calls %>% 
-    filter(is.na(eval_call_srcref)) %>% 
-    select(caller_package, caller_function) %>% 
-    n_distinct()
-
-eval_calls <-
-    eval_calls %>%
-    mutate(eval_call_srcref = coalesce(eval_call_srcref, 
-                                       str_c("missing", "::", caller_package, "::", caller_function)))
-eval_count_table <-
-    eval_calls %>%
-    group_by(eval_function, eval_call_srcref) %>%
-    summarize(count = n()) %>%
-    summarize(call_count = sum(count), callsite_count = n())
-
-nb_eval_calls <- total_call_count <- sum(eval_count_table$call_count)
-
-nb_call_sites <- total_callsite_count <- sum(eval_count_table$callsite_count)
-
-eval_count_table <-
-    eval_count_table %>%
-    mutate(call_proportion = 100 * call_count / total_call_count,
-           callsite_proportion = 100 * callsite_count / total_callsite_count) %>%
-    arrange(desc(call_count))
-
-eval_call_proportion <- 
-    eval_count_table %>%
-    filter(eval_function == "eval") %>%
-    pull(call_proportion)
-```
-
-There are `r NA` R function calls with `r total_call_count` calls to the four kinds of eval which accounts for `r NA`% of all R function calls.
-The table below summarizes the proportion of calls to the different kinds of evals.
-
-```{r }
-eval_count_table %>%
-select(eval_function, call_count, call_proportion) %>% 
-datatable()
-```
-
-We observe that `r eval_call_proportion`% of the calls are made to the `eval` function.
-
-These `eval` calls originate from `r total_callsite_count` call sites.
-For the number of call sites, we use line number information attached to the AST by the parser.
-When it is not available (such as in *base*), we use a combination of `caller_package` and `caller_function`.
-We assume that:
-
-  - the combination `caller_package` and `caller_function` is unique
-  - there is only one call site per such `caller_function`
-This assumption results in a lower bound for the number of unique callsites for eval.
-
-```{r call-site-table}
-eval_callsite_table <-
-    eval_calls %>%
-    count(eval_function, caller_package, eval_call_srcref, name = "call_count") %>%
-    group_by(eval_function, caller_package) %>%
-    summarize(call_count = sum(call_count), callsite_count = n()) %>%
-    ungroup()
-
-eval_callsite_table %>%
-datatable()
-```
-
-```{r call-site-distribution}
-package_callsite_distribution <-
-    eval_callsite_table %>%
-    mutate(callsite_count = if_else(callsite_count <= 5, as.character(callsite_count), "> 5")) %>%
-    group_by(eval_function, callsite_count) %>%
-    summarize(package_count = n())
-
-package_callsite_distribution %>%
-datatable()
-
-p <-
-    package_callsite_distribution %>%
-    ggplot(aes(callsite_count, package_count)) +
-    geom_col() +
-    facet_wrap(vars(eval_function))
-p
-```
-
-To asses how precise the call site estimate is for calls that do not have `srcref`, we compare it on *base* with the actual count given by `grep`.
-
-```{r eval_call_sites_base}
-nb_call_site_base <- eval_calls %>% filter(caller_package == "base") %>% select(caller_function) %>% n_distinct(na.rm = TRUE)
-```
-The estimate for *base* is `r nb_call_site_base` call sites, while `grep` yields 40 `eval` call sites.
-
-For the other packages, ranked by call sites:
-```{r eval_call_sites_per_package}
-eval_calls %>% filter(caller_package != "base") %>% group_by(caller_package) %>% summarize(nb_call_sites =  n_distinct(eval_call_srcref, na.rm = TRUE)) %>% arrange(desc(nb_call_sites))
-```
-
-We probably forget the generated ones here, but should we count them in the _call sites_?
-
-What are the packages with the highest number of `eval` calls?
-
-```{r nb_eval_calls_per_package}
-nb_eval_calls_per_package <- eval_calls %>% group_by(caller_package) %>% summarise(nb_eval_calls_package = n(), per_cent = nb_eval_calls_package / nb_eval_calls * 100) %>% arrange(desc(nb_eval_calls_package))
-```
-
-*base* has the highest number of eval calls, i.e. `r nb_eval_calls_per_package[1, "per_cent"][[1]]`% of them.
-
-```{r}
-nb_eval_calls_per_package  %>% datatable()
-```
-
-### Excluding *base* from the data set
-
-When using `grep` , we find 30 call sites that use `eval` in *base*.
-We tried to replace the 30 `eval` call sites in *base* and could do it in  50% of the cases.
-
-We reckon that `eval` in *base* should be considered apart from the other `eval` calls.
-
-`eval` in *base* does not consume user input, i.e. input coming from `parse`, except in 5 legitimate cases, such as the `sys.source` function that is used to `eval` a file.
-
-| Function name              | File           | Line | Parse ? | Comments                                                                           |
-|----------------------------|----------------|------|---------|------------------------------------------------------------------------------------|
-| autoload                   | autoload       | 25   |         |                                                                                    |
-| autoloader                 | autoload       | 43   |         | To execute a call to library                                                       |
-| autoloader                 | autoload       | 49   |         |                                                                                    |
-| bquote                     | backquote      | 33   |         | In subfunction unquote. To eval what is in .() . Metaprogramming                   |
-| bquote                     | backquote      | 56   |         | idem (in unquote.list, defined inside bquote)                                      |
-| by.default                 | by             | 35   |         | eval(substitute pattern?                                                           |
-| by.data.frame              | by             | 50   |         | idem                                                                               |
-| invokeRestartInteractively | conditions     | 245  | yes     | prompt                                                                             |
-| invokeRestartInteractively | conditions     | 249  | yes     | prompt                                                                             |
-| Ops.data.frame             | dataframe      | 1678 |         | for arithmetic operators on data frames ; do.call could be probably used here also |
-| dget                       | dput           | 32   | yes     | from file                                                                          |
-| eval                       | eval           | 26   |         | Call Internal eval                                                                 |
-| eval.parent                | eval           | 30   |         |                                                                                    |
-| evalq                      | eval           | 36   |         | evalq is basically eval(substitute                                                 |
-| local                      | eval           | 49   |         |                                                                                    |
-| with.default               | eval           | 57   |         |                                                                                    |
-| within.data.frame          | eval           | 63   |         |                                                                                    |
-| within.list                | eval           | 77   |         | evalq                                                                              |
-| replicate                  | replicate      | 21   |         |                                                                                    |
-| subset.data.frame          | frametools     | 25   |         |                                                                                    |
-| subset.data.frame          | frametools     | 34   |         |                                                                                    |
-| subset.matrix              | frametools     | 56   |         |                                                                                    |
-| transform.data.frame       | frametools     | 69   |         |                                                                                    |
-| match.arg                  | match          | 45   |         |                                                                                    |
-| char.expand                | match          | 79   |         |                                                                                    |
-| max.col                    | max            | 22   |         | To access formals                                                                  |
-| parseNamespaceFile         | namespace      | 1315 | yes     | To parse directives                                                                |
-| idem                       | namespace      | 1319 | yes     |                                                                                    |
-| idem                       | namespace      | 1442 | yes     | evaluate a call from a parse                                                       |
-| source                     | source         | 123  | yes     | Use the internal parse function (here it is a comment)                             |
-| source                     | source         | 170  | yes     | comment                                                                            |
-| source                     | source         | 219  | yes     | the actual eval                                                                    |
-| sys.source                 | source         | 277  | yes     | the non-internal parse is used                                                     |
-| stopifnot                  | stop           | 54   |         | change of caller pattern                                                           |
-| as.data.frame.table        | table          | 244  |         | Build a call?                                                                      |
-| match.fun                  | match.fun      | 28   |         | eval.parent                                                                        |
-| trace                      | methodsSupport | 38   |         | eval.parent                                                                        |
-| untrace                    | methodsSupport | 53   |         | eval.parent                                                                        |
-| .doTrace                   | methodsSupport | 78   |         | eval.parent                                                                        |
-
-```{r parse_in_base}
-parsed_base <- eval_calls %>% filter(!is.na(expr_parsed_expression), eval_call_package == "base")
-
-nb_call_site_parse_base <-  nb_eval_call_sites(parsed_base)
-```
-
-There are `r nrow(parsed_base)` `eval` calls with `parse` in *base*, which is `r nrow(parsed_base) / nrow(filter(eval_calls, eval_call_package == "base"))`%.
-
-## Size of expressions as arguments to `eval`
-
-```{r}
-eval_calls %>% ggplot() +
-  geom_histogram(aes(x = expr_expression_length), na.rm = TRUE, binwidth=1) +
-  scale_x_log10() +  scale_y_sqrt()
-```
-
-If we remove the outliers:
-```{r}
-eval_calls %>% filter(expr_expression_length < 25000) %>% ggplot() +
-  geom_histogram(aes(x = expr_expression_length), na.rm = TRUE, binwidth=2)  + scale_x_sqrt() +  scale_y_sqrt() 
-```
-
-<details>
-  <summary>More details</summary>
-
-What about the code size of expressions that seem to be generated?
-An approximation of *generated* will be no `srcref` and not in `base`.
-
-```{r code_size_srcref}
-eval_calls %>% filter(is.na(eval_call_srcref), caller_package != "base") %>% ggplot() +
-  geom_histogram(aes(x = expr_expression_length), na.rm = TRUE, binwidth=1) + scale_x_sqrt() + scale_y_sqrt()
-```
-
-And the average size per package:
-```{r average_size_package}
-eval_calls %>% group_by(caller_package) %>% summarize(average_expr_size = mean(expr_expression_length, na.rm = TRUE)) %>% arrange(desc(average_expr_size))
-```
-
-If we look at the maxima, the largest ones are in *base*:
-```{r max_size_package}
-eval_calls %>% group_by(caller_package) %>% summarize(max_expr_size = max(expr_expression_length, na.rm = TRUE)) %>% arrange(desc(max_expr_size))
-```
-
-</details>
-
-## Amount of computations performed in `eval`
-
-## Aliases of `eval`
-
-# A taxonomy of `eval`
-
-## The operation mix
-
-## Scope (environments)
-
-## Patterns
-
-### Classifying according to the type of the `expr` argument
-
-We look at the first argument of `eval`, the `expr` argument.
-It is the expression to be evaluated.
-It can be of different types, mainly usual numeric types, vectors, strings, and language types: calls, symbols, promise, bytecode and expressions.
-
-`bytecode` refers to bytecode objects, that are generated by the `compile`function:
-
-```{r}
-typeof(compiler::compile(1+1))
-```
-
-Type `expression` is a sequence of calls and symbols.
-
-```{r types}
-per_expr_type <- eval_calls %>% count(expr_expression_type)
-```
-
-```{r}
- per_expr_type %>% mutate(per_cent = n / nb_eval_calls * 100 ) %>% arrange(desc(n)) %>% knitr::kable()
-```
-
-```{r}
-per_expr_type %>% ggplot() +
-  geom_col(aes(x = fct_reorder(expr_expression_type, n), y = n)) +
-  scale_y_sqrt() + 
-  coord_flip() +
-  labs(title = "Calls per type", y = "n", x = "type" ) 
-```
-
-Expressions of types other than `call`,  `expression`, `name`, `bytecode` and `promise` are just returned unchanged.
-
-```{r not_unchanged_eval}
-changed_evals <- eval_calls %>% mutate(changed =  fct_collapse(expr_expression_type, base_types = c("LANGSXP", "EXPRSXP", "SYMSXP", "PROMSXP", "BCODESXP"))) %>% count(changed) %>% mutate(per_cent = n / nb_eval_calls * 100 )
-```
-
-We can remove `eval` and leave there the expression in `r changed_evals[2, "per_cent"][[1]]`% of the eval calls (i.e. `r changed_evals[2, "n"][[1]]`).
-
-### `expr` argument when it is a call
-
-```{r expr_call_arg}
-expr_call_arg <- eval_calls %>% filter(expr_expression_type == "LANGSXP") %>% 
-  mutate(function_expr = map_chr(expr_expression, function_name))
-```
-
-The following top level functions are used:
-
-```{r top_level_function_expr}
-expr_call_arg %>% count(function_expr) %>% arrange(desc(n))
-```
-
-We can distinguish a few groups:
-
-- assignments
-- reading (in a data frame, a vector)
-- parsing
-- defusing (`quote` and so on)
-- conversion or creation of new values
-- building calls and calling
-- working with environments
-- class assignments, reads and definitions
-- building logical values
-- formula
-- meta programming on functions (getting access to the body of the function, to the call stack)
-- 'NA' corresponds to no function calls (or that it was not deparsed successfully)
-
-```{r group_expr_functions}
-group_expr_functions <- expr_call_arg %>% mutate(group_function_expr = fct_collapse(function_expr, 
-                          assignment = c("<-", "[[<-", "[<-", "$<-"),
-                          read = c("[", "[[", "$"),
-                          parsing = c("parse", "str2lang", "str2expression", "parse_only", "base::parse"),
-                          defusing = c("quote", "substitute", "bquote", "expr", "quo"),
-                          expressions = c("as.list", "expression", "as.name", "as.expression", "list", "as.data.frame", "as.character", "{", "modify_lang"),
-                          calling = c("call", "as.call", "do.call"),
-                          environment = c("environment", "as.environment"),
-                          classAssign = c("@<-"),
-                          classRead = c("slot", "@"),
-                          classDef = c(".Primitive(\"class\")"),
-                          logical = c("<", ">", "is.element"),
-                          choice = c("if"),
-                          function_meta = c("sys.calls", "function", "body"),
-                          formula = c("~", "deriv", "D", "model.frame", "rootftfunc", "contest.lmerModLmerTest", "JacFunc", "gamlss.family", "contest1D.lmerModLmerTest", "contestMD.lmerModLmerTest", "reexpr"),
-                          other_level = "other" ))
-                          #TODO: add anonymous function applications
-```
-
-```{r}
-group_expr_functions %>% count(group_function_expr) %>% arrange(desc(n)) %>% knitr::kable()
-```
-
-```{r}
-group_expr_functions %>% ggplot() +
-  geom_bar(aes(x = fct_rev(fct_infreq(group_function_expr)))) +
-  scale_y_sqrt() +
-  coord_flip() +
-  labs(title = "Category of calls", y = "n", x = "category" ) 
-```
-
-```{r }
-group_expr_functions %>% ggplot() +
-  geom_bar(aes(x = fct_rev(fct_infreq(group_function_expr)))) +
-  scale_y_sqrt() +
-  coord_flip() +
-  labs(title = "Category of calls", y = "n", x = "category" ) +
-  facet_wrap(vars(eval_function))
-```
-
-## Provenance
-
-Some `eval` result from parsing text (directly or through a file) with `parse`.
-
-```{r eval_from_parse}
-parsed_evals <- eval_calls %>% filter(!is.na(expr_parsed_expression))
-nb_parsed_evals <- parsed_evals %>% nrow()
-```
-
-There are `r nb_parsed_evals` such calls, `r nb_parsed_evals / nb_eval_calls * 100`%.
-
-```{r, echo = FALSE}
-nb_parsed_call_sites <- nb_eval_call_sites(parsed_evals)
-```
-
-In terms of call sites, there are `r nb_parsed_call_sites` i.e. `r nb_parsed_call_sites / nb_call_sites * 100`%.
-
-  We can differentiate further the provenance, i.e.
-whether the evaluated string comes from `file` or from `text`.
-However, we can only do it for `parse`, `str2lang` and `str2expression` at the top level of `expr_expression`, not buried in a function call.
-
-```{r parsed_evals_source}
-parsed_evals_args <- parsed_evals %>% filter(str_detect(expr_parsed_expression, "(parse|str2lang|str2expression)\\(")) %>% mutate(parse_args = map(expr_parsed_expression, extract_args_parse)) %>% unnest_wider(parse_args,  names_sep = "_")
-```
-
-```{r}
-parsed_files <- parsed_evals_args %>% filter(!is.na(parse_args_file))
-nb_file <- parsed_files %>% nrow()
-nb_file_call_sites <- nb_eval_call_sites(parsed_files)
-```
-
-```{r}
-parsed_texts <- parsed_evals_args %>% filter(!is.na(parse_args_text) | !is.na(parse_args_s))
-nb_text <-  parsed_texts %>% nrow()
-nb_text_call_sites <- nb_eval_call_sites(parsed_texts)
-```
-
-There are `r nb_file` calls with the `file` argument, and `r nb_text` with the `text` argument, i.e. `r  nb_file / nb_parsed_evals * 100`%, and `r nb_text / nb_parsed_evals *100`% respectively.
-
-There are `r nb_file_call_sites` call sites with the `file` argument, and `r nb_text_call_sites` with the `text` argument, i.e.
-`r  nb_file_call_sites / nb_parsed_call_sites * 100`%, and `r nb_text_call_sites / nb_parsed_call_sites *100`% respectively.
-
-## Consistence
-
-# Other faces of `eval`
-
-# Case studies

### 2020-08-19 inst/analysis/insights.R
Vectorized extract_package_name function

@@ -206,7 +206,7 @@ resolve_sexp_name <- function(df, var) {
 }
 
 
-extract_package_name <- function(src_ref)
+extract_package_name <- function(src_ref, file)
 {
   # There are 5 possibilities for a srcref:
   # - NA
@@ -214,28 +214,18 @@ extract_package_name <- function(src_ref)
   # - /mnt/nvme0/R/project-evalR/library/4.0/instrumentr/srcref/packagename/4.0.4/file:linenumbers
   # - /R/* : core packages (we cannot distinguish between them yet so we write core for the package name)
   # - /testit/... or /testthat/... : it is the testit or testthat packages
-  if(is.na(src_ref))
-  {
-    return("base")
-  }
-  else if(str_starts(src_ref, fixed("./R/")))
-  {
-    return("core")
-  }
-  else if(str_starts(src_ref, fixed("/tmp/")))
-  {
-    return(str_match(src_ref, "/tmp/Rtmp[^/]*/R\\.INSTALL[^/]*/([^/]+)/.*")[[2]])
-  }
-  else if(str_starts(src_ref, fixed("/mnt/nvme0/")))
-  {
-    return(str_match(src_ref, "/mnt/nvme0/R/project-evalR/library/4.0/instrumentr/srcref/([^/]*)/.*")[[2]])
-  }
-  else if(str_starts(src_ref, fixed("test")))
-  {
-    return(str_match(src_ref, "([^/]*)/.*")[[2]])
-  }
-  else
-  {
-    return("unknown")
-  }
+  # - :/R : extract the package name from the file path in the column path
+  case_when
+  (
+    is.na(src_ref) ~ "base",
+    str_starts(src_ref, fixed("./R/")) ~ "core",
+    str_starts(src_ref, fixed("/tmp/")) ~ str_match(src_ref, "/tmp/Rtmp[^/]*/R\\.INSTALL[^/]*/([^/]+)/.*")[[2]],
+    str_starts(src_ref, fixed("/mnt/nvme0/")) ~ str_match(src_ref, "/mnt/nvme0/R/project-evalR/library/4.0/instrumentr/srcref/([^/]*)/.*")[[2]],
+    str_starts(src_ref, fixed("test")) ~ str_match(src_ref, "([^/]*)/.*")[[2]],
+    str_starts(src_ref, ":/") ~ str_match(file, "[^/]*/[^/]*/([^/]*)/.*")[[2]],
+    TRUE ~ "unknown"
+  )
 }
+
+
+

### 2020-08-19 inst/analysis/insights.R
Vectorized extract_package_name

@@ -215,8 +215,7 @@ extract_package_name <- function(src_ref, file)
   # - /R/* : core packages (we cannot distinguish between them yet so we write core for the package name)
   # - /testit/... or /testthat/... : it is the testit or testthat packages
   # - :/R : extract the package name from the file path in the column path
-  case_when
-  (
+  case_when(
     is.na(src_ref) ~ "base",
     str_starts(src_ref, fixed("./R/")) ~ "core",
     str_starts(src_ref, fixed("/tmp/")) ~ str_match(src_ref, "/tmp/Rtmp[^/]*/R\\.INSTALL[^/]*/([^/]+)/.*")[[2]],

### 2020-08-19 inst/analysis/analysis_stable.Rmd
Latex tags

@@
 source("insights.R", local = knitr::knit_global())
+
+source("inc/latextags.R")
+create_tags(path("/var/lib/R/project-evalR/revalstudy/data", "analysisStable.tex"), prefix="", default=TRUE)
 ```
@@
     as_tibble() %>%
-    mutate(eval_call_package = map_chr(eval_call_srcref, extract_package_name)) %>% 
+    mutate(eval_call_package = extract_package_name(eval_call_srcref, file)) %>% 
     resolve_sexp_name(expr_expression_type) %>%
@@
 
-<details>
-  <summary>More details</summary>
-
 What about the code size of expressions that seem to be generated?
@@
 ```
-
-</details>
 
@@
 - parsing
-- defusing (`quote` and so on)
+- defusing (`quote`, `substitute` and so on)
 - conversion or creation of new values
@@
 group_expr_functions <- expr_call_arg %>%
-                            mutate(function_expr2 = map_chr(detect_anonymous_func, function_expr)) %>%
+                            mutate(function_expr2 = if_else(str_starts(function_expr, fixed("(function(")), "anonymous", function_expr))
+
+group_expr_functions <-  group_expr_functions %>% 
                               mutate(group_function_expr = fct_collapse(function_expr2, 
-                              assignment = c("<-", "[[<-", "[<-", "$<-"),
+                              assignment = c("<-", "[[<-", "[<-", "$<-", "<<-", "="),
                               read = c("[", "[[", "$"),
@@
 group_expr_resolved_functions <- expr_resolved_call_arg %>%
-                            mutate(function_expr2 = map_chr(detect_anonymous_func, function_expr)) %>%
+                            mutate(function_expr2 = if_else(str_starts(function_expr, fixed("(function(")), "anonymous", function_expr)) %>%
                             mutate(group_function_expr = fct_collapse(function_expr2, 
-                              assignment = c("<-", "[[<-", "[<-", "$<-"),
+                              assignment = c("<-", "[[<-", "[<-", "$<-", "<<-", "="),
                               read = c("[", "[[", "$"),
@@
                               defusing = c("quote", "substitute", "bquote", "expr", "quo"),
-                              expressions = c("as.list", "expression", "as.name", "as.expression", "list", "as.data.frame", "as.character", "{", "modify_lang"),
+                              expressions = c("as.list", "expression", "as.name", "as.expression", "list", "c", "as.data.frame", "as.character", "{", "modify_lang"),
                               calling = c("call", "as.call", "do.call"),
@@
                               anonymous = c("anonymous"),
-                              function_meta = c("sys.calls", "function", "body"),
+                              function_meta = c("sys.calls", "function", "body", "missing"),
                               formula = c("~", "deriv", "D", "model.frame", "rootftfunc", "contest.lmerModLmerTest", "JacFunc", "gamlss.family", "contest1D.lmerModLmerTest", "contestMD.lmerModLmerTest", "reexpr"),

### 2020-08-19 inst/analysis/enrich_dataset.Rmd
Remove the eval originating from functions in base

@@
 
-The analysis is being performed on dataset read from file `r params$data_filepath`.
+The analysis is being performed on a dataset read from file `r params$data_filepath`.
 
@@
 
+```{r eval_in_base}
+eval_base_functions <- c("autoload", "autoloader", "bquote", "by.default", "by.data.frame", "invokeRestartInteractively", "Ops.data.frame", "dget", 
+                         "eval", "eval.parent", "evalq", "local", "with.default", "within.data.frame", "within.list", "replicate", "subset.data.frame",
+                         "subset.matrix", "transform.data.frame", "match.arg", "char.expand", "max.col", "parseNamespaceFile", "source", "sys.source",
+                         "stopfinot", "as.data.frane.table", "match.fun", "trace", "untrace", ".doTrace")
+```
+
 ```{r load-data, echo=FALSE}
@@
     read_fst(params$data_filepath) %>%
+    filter(!caller_function %in% eval_base_functions) %>%
     as_tibble() %>%
@@
     resolve_sexp_name(envir_type)
+```
 
+```{r}
+eval_calls <-  eval_calls %>% mutate(function_expr = if_else(expr_expression_type == "LANGSXP", map_chr(expr_expression, function_name), NA), function_expr_resolved = if_else(expr_resolved_type == "LANGSXP", map_chr(expr_resolved, function_name), NA))
 ```
+
+```{r}
+eval_calls <- eval_calls %>% mutate(parse_args = if_else(is.na(expr_parsed_expression), NA, map(expr_parsed_expression, extract_args_parse)))
+```
+
+```{r write_back}
+write_fst(eval_calls, paste0(params$data_dirpath, "enriched_calls.fst"))
+```

### 2020-08-19 inst/analysis/inc/latextags.R
Tags

@@ -17,9 +17,9 @@ my_datatable <- function(df, page_size=20, round=TRUE, ...) {
   }
   
   table <- datatable(df, options=options, ...)
-  if (round) {
-    table <- formatRound(table, sapply(df, is.numeric), 2)
-  }
+  #if (round) {
+  #  table <- formatRound(table, sapply(df, is.numeric), 2)
+  #}
   table
 }
 
@@ -54,6 +54,15 @@ r.Duration <- function(name, value, ...) {
 
   r(name, v)
 }
+
+r.difftime <- function(name, value, ...) {
+  v <- as.numeric(value)
+  class(v) <- "num_with_suffix"
+  attr(v, "suffix") <- paste0(" ", attr(value, "units"))
+
+  r(name, v)
+}
+
 
 r.default <- function(name, value, ...) {
   stopifnot(length(value) == 1)
@@ -205,14 +214,11 @@ latex_escape <- function(s) {
 }
 
 generate_latex_command_name <- function(name, prefix="") {
-  if (nchar(prefix)) {
-    prefix <- stringr::str_c(prefix, " ")
-  }
-  
-  name <- stringr::str_c(prefix, name)
   name <- latex_command_name(name)
+  name <- latex_escape(name)
+  prefix <- latex_escape(prefix)
   
-  latex_escape(name)
+  stringr::str_c(prefix, name)
 }
 
 generate_latex_command <- function(command_name, value) {

### 2020-08-19 inst/analysis/inc/paths.R

@@ -1,6 +1,14 @@
+## INPUTs
+
+PACKAGE_METADATA     <- path(params$data_dir, "metadata.csv")
+PACKAGE_SLOC         <- path(params$data_dir, "sloc.csv")
+
+## OUTPUTs
+
 ## corpus only
 CORPUS_FILE                  <- path(params$output_dir, "corpus.txt")
 CORPUS_REVDEPS               <- path(params$output_dir, "corpus-revdeps.csv")
 CORPUS_REVDEPS_FILE          <- path(params$output_dir, "corpus-revdeps.txt")
+CORPUS_ALL_PACKAGES_FILE     <- path(params$output_dir, "corpus-all.txt")
 
 TAGS_DIR <- params$output_dir

### 2020-08-19 inst/analysis/enrich_dataset.Rmd
Keeping on decreasing the size of the dataset

@@
   rmdformats::readthedown:
-        code_folding: hide
         lightbox: true
@@
 ```{r load-data, echo=FALSE}
-eval_calls <-
+eval_calls_raw <-
     read_fst(params$data_filepath) %>%
-    filter(!caller_function %in% eval_base_functions) %>%
-    as_tibble() %>%
-    mutate(eval_call_package = map_chr(eval_call_srcref, extract_package_name)) %>% 
+    as_tibble()
+
+eval_calls <- eval_calls_raw %>% filter(!caller_function %in% eval_base_functions)
+```
+
+There are so many calls that are similar:
+```{r}
+eval_calls_base <- eval_calls_raw %>% filter(caller_function %in% eval_base_functions) 
+```
+
+Let's remove the duplicate ones:
+
+```{r}
+eval_calls_base <- eval_calls_base %>% select(-eval_call_id, -starts_with("caller_stack")) %>% distinct()
+
+eval_calls <- eval_calls_base %>% select(-eval_call_id) %>% distinct()
+```
+
+```{r}
+eval_calls_base <- eval_calls_base %>% 
+    resolve_sexp_name(expr_expression_type) %>%
+    resolve_sexp_name(expr_resolved_type) %>%
+    resolve_sexp_name(enclos_type) %>%
+    resolve_sexp_name(envir_type)
+
+eval_calls <- eval_calls %>% 
+    mutate(eval_call_package = extract_package_name(eval_call_srcref, file)) %>% 
     resolve_sexp_name(expr_expression_type) %>%
@@
 ```{r}
+# We are doing some work once again! Parsing what was deparsed previously...
 eval_calls <-  eval_calls %>% mutate(function_expr = if_else(expr_expression_type == "LANGSXP", map_chr(expr_expression, function_name), NA), function_expr_resolved = if_else(expr_resolved_type == "LANGSXP", map_chr(expr_resolved, function_name), NA))
@@
 ```{r}
-eval_calls <- eval_calls %>% mutate(parse_args = if_else(is.na(expr_parsed_expression), NA, map(expr_parsed_expression, extract_args_parse)))
+eval_calls <- eval_calls %>% mutate(parse_args = if_else(is.na(expr_parsed_expression), NA, map_chr(expr_parsed_expression, extract_args_parse)))
 ```
@@
 write_fst(eval_calls, paste0(params$data_dirpath, "enriched_calls.fst"))
+
+write_dst(eval_calls_base, paste0(params$data_dirpath, "base_calls.fst"))
 ```

### 2020-08-19 inst/in-docker.sh
Updated docker image and first tasks

@@ -1,0 +1,26 @@
+#!/bin/bash
+
+DOCKER_IMAGE_NAME="prlprg/project-evalr"
+
+script_dir="$( cd "$(dirname "${BASH_SOURCE[0]}")" ; pwd -P )"
+base_dir=$(dirname "$script_dir")
+docker_working_dir="/home/rstudio/evalR"
+local_working_dir="$base_dir/inst"
+
+docker run \
+    -ti \
+    --rm \
+    -e ROOT=TRUE \
+    -e DISABLE_AUTH=true \
+    -e USERID=$(id -u) \
+    -e GROUPID=$(id -g) \
+    -v "$local_working_dir:$docker_working_dir" \
+    -w "$docker_working_dir" \
+    "$DOCKER_IMAGE_NAME" \
+    "$@"
+
+# docker run \
+#     --rm \
+#     -v "$local_working_dir:$docker_working_dir" \
+#     "$DOCKER_IMAGE_NAME" \
+#     chown -R $(id -u):$(id -g) "$docker_working_dir"

### 2020-08-20 inst/analysis/enrich_dataset.Rmd
Add the number of calls of each deduplicated calls.

@@
 params:
-  data_dirpath: "/var/lib/R/project-evalR/run/package-evals-traced-corpus.3/"
-  data_filepath: "/var/lib/R/project-evalR/run/package-evals-traced-corpus.3/calls.fst"
+  data_dirpath: "/var/lib/R/project-evalR/run/package-evals-traced-corpus.4/"
+  data_filepath: "/var/lib/R/project-evalR/run/package-evals-traced-corpus.4/calls.fst"
 knit: (function(inputFile, encoding) {
@@
 
-Let's remove the duplicate ones:
+Let's remove the duplicate ones (but keep:
 
 ```{r}
-eval_calls_base <- eval_calls_base %>% select(-eval_call_id, -starts_with("caller_stack")) %>% distinct()
-
-eval_calls <- eval_calls_base %>% select(-eval_call_id) %>% distinct()
+eval_calls_base <- eval_calls_base %>% count(across(c(-eval_call_id, -starts_with("caller_stack"))), name = "nb_eval_calls") 
+eval_calls <- eval_calls %>% count(across(c(-eval_call_id)), name = "nb_eval_calls")
 ```
@@
 ```{r}
-# We are doing some work once again! Parsing what was deparsed previously...
-eval_calls <-  eval_calls %>% mutate(function_expr = if_else(expr_expression_type == "LANGSXP", map_chr(expr_expression, function_name), NA), function_expr_resolved = if_else(expr_resolved_type == "LANGSXP", map_chr(expr_resolved, function_name), NA))
-```
-
-```{r}
 eval_calls <- eval_calls %>% mutate(parse_args = if_else(is.na(expr_parsed_expression), NA, map_chr(expr_parsed_expression, extract_args_parse)))
@@
 
-write_dst(eval_calls_base, paste0(params$data_dirpath, "base_calls.fst"))
+write_fst(eval_calls_base, paste0(params$data_dirpath, "base_calls.fst"))
 ```

### 2020-08-20 inst/analysis/analysis_stable.Rmd
Compute the arity of the functions in exp arguments when they have language type

@@
 params:
-    data_filepath: "/var/lib/R/project-evalR/run/package-evals-traced-corpus.3/calls.fst"
+    data_filepath: "/var/lib/R/project-evalR/run/package-evals-traced-corpus.4/calls.fst"
     #data_filepath: "/var/lib/R/project-evalR/run/package-evals-traced-all.1/calls.fst"
@@
 ```{r expr_call_arg}
-expr_call_arg <- eval_calls %>% filter(expr_expression_type == "LANGSXP") %>% 
-  mutate(function_expr = map_chr(expr_expression, function_name))
+expr_call_arg <- eval_calls %>% filter(expr_expression_type == "LANGSXP")
 ```
@@
 ```{r top_level_function_expr}
-expr_call_arg %>% count(function_expr) %>% arrange(desc(n))
+expr_call_arg %>% count(expr_expression_function, sort = TRUE) 
 ```
@@
 group_expr_functions <- expr_call_arg %>%
-                            mutate(function_expr2 = if_else(str_starts(function_expr, fixed("(function(")), "anonymous", function_expr))
+                            mutate(function_expr2 = if_else(str_starts(expr_expression_function, fixed("(function(")), "anonymous", expr_expression_function))
 
@@
 
+We also look at the arity of the top-level function:
+
+```{r arity_expr_expression}
+expr_call_arg %>% count(expr_expression_args_num, sort = TRUE) %>% knitr::kable()
+```
+
+```{r arity_expr_expression_plot}
+expr_call_arg %>% ggplot() + 
+  geom_bar(aes(x = fct_rev(fct_infreq(expr_expression_args_num)))) +
+  coord_flip() + 
+  labs(title = "Arity of calls", x = "arity", y = "n")
+
+```
+
+```{r arity_expr_expression_plot_per_eval}
+expr_call_arg %>% ggplot() + 
+  geom_bar(aes(x = fct_rev(fct_infreq(expr_expression_args_num)))) +
+  coord_flip() + 
+  labs(title = "Arity of calls", x = "arity", y = "n") +
+  facet_wrap(vars(eval_function))
+```
+
 ### For resolved `expr`
@@
 ```{r expr_resolved__call_arg}
-expr_resolved_call_arg <- eval_calls %>% filter(expr_resolved_type == "LANGSXP") %>% 
-  mutate(function_expr = map_chr(expr_resolved, function_name))
+expr_resolved_call_arg <- eval_calls %>% filter(expr_resolved_type == "LANGSXP")
 ```
@@
 ```{r top_level_function_resolved_expr}
-expr_resolved_call_arg %>% count(function_expr) %>% arrange(desc(n))
+expr_resolved_call_arg %>% count(expr_resolved_function) %>% arrange(desc(n))
 ```
@@
 group_expr_resolved_functions <- expr_resolved_call_arg %>%
-                            mutate(function_expr2 = if_else(str_starts(function_expr, fixed("(function(")), "anonymous", function_expr)) %>%
+                            mutate(function_expr2 = if_else(str_starts(expr_resolved_function, fixed("(function(")), "anonymous", expr_resolved_function)) %>%
                             mutate(group_function_expr = fct_collapse(function_expr2, 
@@
   labs(title = "Category of calls", y = "n", x = "category" ) +
+  facet_wrap(vars(eval_function))
+```
+
+We also look at the arity of the top-level function:
+
+```{r arity_expr_resolved}
+expr_resolved_call_arg %>% count(expr_expression_args_num, sort = TRUE) %>% knitr::kable()
+```
+
+```{r arity_expr_resolved_plot}
+expr_resolved_call_arg %>% ggplot() + 
+  geom_bar(aes(x = fct_rev(fct_infreq(expr_resolved_args_num)))) +
+  coord_flip() + 
+  labs(title = "Arity of calls", x = "arity", y = "n")
+
+```
+
+```{r arity_expr_resolved_plot_per_eval}
+expr_resolved_call_arg %>% ggplot() + 
+  geom_bar(aes(x = fct_rev(fct_infreq(expr_resolved_args_num)))) +
+  coord_flip() + 
+  labs(title = "Arity of calls", x = "arity", y = "n") +
   facet_wrap(vars(eval_function))

### 2020-08-20 inst/analysis/insights.R
New category: primitive

@@ -227,4 +227,13 @@ extract_package_name <- function(src_ref, file)
 }
 
 
+groupify_function <- function(expr_function)
+{
+  case_when(
+    str_starts(expr_function, fixed("(function(")) ~ "anonymous",
+    str_starts(expr_function, fixed(".Primitive(")) ~ "primitive",
+    TRUE ~ expr_function
+  )
+}
+
 

### 2020-08-20 inst/analysis/insights.R
is_repleaceable AST traversing function

@@ -226,5 +226,38 @@ extract_package_name <- function(src_ref, file)
   )
 }
 
+replaceable_functions <- c("+", "*", "/", "-", "%%", 
+                           "<-", "[[<-", "[<-", "$<-", "<<-", "=",
+                           "[", "[[", "$",
+                           "@<-",
+                           "slot", "@",
+                           "<", ">", "==", "<-", ">=",
+                           "if")
+
+
+is_replaceable <- function(expr)
+{
+  # We are going to traverse the ast recursively
+  # We cannot replace in 2 cases: the function name in a call is too complex (we could do a do.call though), it is bytecode
+  if(is.call(expr))
+  {
+    function_name <- expr[[1]]
+    function_args <- expr[-1]
+    return(as.character(function_name) %in% replaceable_functions && every(function_args, is_replaceable))
+    
+  }
+  else if(is.expression(expr))
+  {
+    return(every(expr, is_replaceable))
+  }
+  else if(typeof(expr) == "bytecode")
+  {
+    return(FALSE)
+  }
+  else # Symbols, promises, vector types, closure and builtins, S4 objects. There should not be "..." here
+  {
+    return(TRUE)
+  }
+}
 
 

### 2020-08-20 inst/analysis/analysis_stable.Rmd

@@
 group_expr_functions <- expr_call_arg %>%
-                            mutate(function_expr2 = if_else(str_starts(expr_expression_function, fixed("(function(")), "anonymous", expr_expression_function))
+                            #mutate(function_expr2 = if_else(str_starts(expr_expression_function, fixed("(function(")), "anonymous", expr_expression_function))
+                            mutate(function_expr2 = groupify_function(expr_expression_function)) %>%
 
@@
 expr_call_arg %>% ggplot() + 
-  geom_bar(aes(x = fct_rev(fct_infreq(expr_expression_args_num)))) +
+  geom_bar(aes(x = expr_expression_args_num)) +
+  scale_y_sqrt() +
   coord_flip() + 
@@
 group_expr_resolved_functions <- expr_resolved_call_arg %>%
-                            mutate(function_expr2 = if_else(str_starts(expr_resolved_function, fixed("(function(")), "anonymous", expr_resolved_function)) %>%
+                            #mutate(function_expr2 = if_else(str_starts(expr_resolved_function, fixed("(function(")), "anonymous", expr_resolved_function)) %>%
+                            mutate(function_expr2 = groupify_function(expr_resolved_function)) %>%
                             mutate(group_function_expr = fct_collapse(function_expr2, 
@@
                               anonymous = c("anonymous"),
+                              primitive = c("primitive"),
                               function_meta = c("sys.calls", "function", "body", "missing"),
@@
 expr_resolved_call_arg %>% ggplot() + 
-  geom_bar(aes(x = fct_rev(fct_infreq(expr_resolved_args_num)))) +
+  geom_bar(aes(x = expr_resolved_args_num)) +
+  scale_y_sqrt() +
   coord_flip() + 
@@
 expr_resolved_call_arg %>% ggplot() + 
-  geom_bar(aes(x = fct_rev(fct_infreq(expr_resolved_args_num)))) +
+  geom_bar(aes(x = expr_resolved_args_num)) + 
+  scale_y_sqrt() +
   coord_flip() + 

### 2020-08-20 inst/analysis/enrich_dataset.Rmd

@@
 
-Let's remove the duplicate ones (but keep:
+Let's remove the duplicate ones (but keep their numbers):
 
@@
 ```{r}
-eval_calls <- eval_calls %>% mutate(parse_args = if_else(is.na(expr_parsed_expression), NA, map_chr(expr_parsed_expression, extract_args_parse)))
+#eval_calls <- eval_calls %>% mutate(parse_args = if_else(is.na(expr_parsed_expression), NA, map_chr(expr_parsed_expression, extract_args_parse)))
 ```
@@
 ```{r write_back}
-write_fst(eval_calls, paste0(params$data_dirpath, "enriched_calls.fst"))
+write_fst(eval_calls, paste0(params$data_dirpath, "deduplicated_calls.fst"))
 

### 2020-08-20 inst/paper/main.tex
Add some details

@@
 - cannot replace with (cite study of with construct).
+- 
 \textbf{*}
@@
 \section{Background on R}
-\textbf{Aviral}
+- Talk about environments
+- How is it different from JS (takes language, env in R but string in JS).
 
 \section{The Design of Eval}
-\textbf{Aviral}
 
 \section{Infrastructure}
-\textbf{Aviral + Filip}
 
 \section{Corpus of R Programs}
-\textbf{Filip}
 
 \section{Threats to Validity}
-\textbf{Filip}
 
-\section{Analyzing Eval Usage Patterns}
+\section{Research Question}
 This section presents the results of our empirical study of eval in the R language.
 \subsection{How many evals}
-\textbf{Pierre}
 
 \subsection{How are they used}
-\textbf{Pierre}
+
 - classify evals based on what they can be replaced with
@@
 \subsection{How to replace evals}
-\textbf{Aviral}
+
+\subsection{How dangerous is eval}
 
 \subsection{How many evals can be replaced}
-\textbf{Aviral}
-
-\section{Evaluation}
 
@@
 \end{document}
+
+Eval is evil, we would like to replace it.
+Research Questions
+How often is eval used? 
+How often it is used in base / other packages?
+blocking: cannot find out precisely what comes from base
+What is the source of evals?
+Provenance (source arguments to eval)
+How many evals are in libraries / vignettes
+What is eval used for?
+??Eval in the tidyverse
+TODO: Classification of evals in  packages (Pierre)
+TODO: look at the first function to get
+binary / unary / assignment / other function
+does it have {
+evals that do nothing (they get values)
+...
+TODO: In the case of base package (Filip)
+access default values of formals
+simply implementation
+weird replacement of calls (write.csv -> write.table)
+accessing variable from another environment (instead of get)
+evaling user input
+Change of caller pattern: 
+m <- match.call()
+ m$name <- NULL
+ m[[1L]] <- as.name("library")
+eval(m, .GlobalEnv)
+     - 	TODO: How dangerous is eval? (Aviral)
+Does eval generate new code (create new functions)?
+environment writes (reads)
+system environment variables
+connections
+random number generator - setting seed
+Does eval do dynamic code loading?
+Does eval argument reflect on the call stack?
+Calls to C Code
+TODO: Could the eval be made safer?
+TODO: Could some of the eval calls be replaced?

### 2020-08-20 inst/paper/main.tex
Outline for The Design of Eval

@@
 
+There are four different kinds of \texttt{eval} in the \emph{base} package:  \texttt{eval}, \texttt{evalq}, \texttt{eval.parent},  \texttt{local}.
+The most important one is \texttt{eval}:
+
+- the four eval functions in base
+- the arguments, expr and envir, enclos
+- which expressions are evaluated, which ones are left unchanged
+
 \section{Infrastructure}

### 2020-08-20 inst/paper/main.tex
Case study section: base
Basic text for number of evals

@@
 \usepackage{xcolor}
+\usepackage{xspace}
 \def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
     T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
+
+\newcommand{\missingNumber}{\textcolor{red}{XX}\xspace}
+\newcommand{\missingPercentage}{\textcolor{red}{XX\%}\xspace}
+\newcommand{\missingTable}{\textcolor{red}{XXTable}\xspace}
+\newcommand{\missingGraph}{\textcolor{red}{XXGraph}\xspace}
 \begin{document}
@@
 
+In our corpus eval is called \missingNumber times out of \missingNumber
+total calls to all R functions. This is merely \missingPercentage of the total
+number of calls. \missingTable breaks down the number of calls and callsites for
+the four kinds of eval functions from the three source; core R, packages and
+Kaggle code.
+
+The core R packages are the biggest consumers of eval. Majority of calls,
+\missingPercentage, originate from the core R packages. Kaggle code, which is
+end-user code, has merely \missingNumber calls to eval.
+
+Of the \missingNumber packages in our corpus, only \missingNumber packages call
+eval. \missingTable shows the number of calls to eval made by the ten most
+frequent callers to eval. \missingNumber packages account for over
+\missingPercentage calls to eval.
+
+\missingGraph shows the distribution of number of callsites to eval across these
+packages. \missingNumber packages have a single eval callsite.
+
 \subsection{How are they used}
@@
 - classify evals based on what they can be replaced with
+
+\subsubsection{A taxonomy of \texttt{eval}}
+
+We first classify \texttt{eval} depending on the resolved expression, \emph{i.e.} the first argument of \texttt{eval} after all possible function calls and symbol resolution in it have been executed.
 
@@
 \subsection{How many evals can be replaced}
+
+\section{Case studies}
+
+\subsection{\texttt{base}}
 

### 2020-08-21 inst/paper/main.tex
minor

@@
 \documentclass[conference]{IEEEtran}
-\IEEEoverridecommandlockouts
-% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
 \usepackage{cite}
@@
 \usepackage{textcomp}
-\usepackage{xcolor}
-\usepackage{xspace}
-\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
-    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
-
 \newcommand{\missingNumber}{\textcolor{red}{XX}\xspace}
@@
 \newcommand{\missingGraph}{\textcolor{red}{XXGraph}\xspace}
+\usepackage{xcolor,xspace}
 \begin{document}
 
-\title{On the Design, Use and Removal of Eval in R}
+\title{A Study of the Usage of Eval in R}
 
-\author{\IEEEauthorblockN{1\textsuperscript{st} Given Name Surname}
-\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
-\textit{name of organization (of Aff.)}\\
-City, Country \\
-email address or ORCID}
-\and
-\IEEEauthorblockN{2\textsuperscript{nd} Given Name Surname}
-\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
-\textit{name of organization (of Aff.)}\\
-City, Country \\
-email address or ORCID}
-\and
-\IEEEauthorblockN{3\textsuperscript{rd} Given Name Surname}
-\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
-\textit{name of organization (of Aff.)}\\
-City, Country \\
-email address or ORCID}
-\and
-\IEEEauthorblockN{4\textsuperscript{th} Given Name Surname}
-\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
-\textit{name of organization (of Aff.)}\\
-City, Country \\
-email address or ORCID}
-\and
-\IEEEauthorblockN{5\textsuperscript{th} Given Name Surname}
-\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
-\textit{name of organization (of Aff.)}\\
-City, Country \\
-email address or ORCID}
-\and
-\IEEEauthorblockN{6\textsuperscript{th} Given Name Surname}
-\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
-\textit{name of organization (of Aff.)}\\
-City, Country \\
-email address or ORCID}
+\author{\IEEEauthorblockN{For double blind review}%
+  %\IEEEauthorblockA{ dept} \\ \and
 }  
-
 \maketitle
 
+\renewcommand{\c}[1]{{\small\sf{#1}}\xspace}
+
 \begin{abstract}
-This document is a model and instructions for \LaTeX.
-This and the IEEEtran.cls file define the components of your paper [title, text, heads, etc.]. *CRITICAL: Do Not Use Symbols, Special Characters, Footnotes, 
-or Math in Paper Title or Abstract.
+
 \end{abstract}
 
-\begin{IEEEkeywords}
-component, formatting, style, styling, insert
-\end{IEEEkeywords}
+\section{Introduction}
 
-\section{Introduction}
+The \c{eval} function has a bad reputation. This function, under one name or
+another and with various exact semantics, is available in many dynamic
+languages. It is used to transform text into code and execute that code.
+
 - Why eval is used in R
@@
 
-\section{Background on R}
+\section{Related Work}
+
+\section{Eval in R}
 - Talk about environments

### 2020-08-21 inst/paper/main.tex
stuff

@@
 
-\title{A Study of the Usage of Eval in R}
-
-\author{\IEEEauthorblockN{For double blind review}%
+\title{A Large-Scale Study of the Use of Eval in R}
+\author{\vspace{-.8cm}\IEEEauthorblockN{\it\Large The Eval that Data Scientist Do}%
   %\IEEEauthorblockA{ dept} \\ \and
@@
 \begin{abstract}
-
+  The eval function turns text into code and thus causes difficulties for
+  reasoning about the behavior of software that uses it. This paper is a
+  study of the usage of eval in R, a widely used data science programming
+  language. We combine static and dynamic analysis on a large and
+  representative corpus of R programs. Our results show...
 \end{abstract}
@@
 another and with various exact semantics, is available in many dynamic
-languages. It is used to transform text into code and execute that code.
+languages. It is used to transform text into code and execute that code.  It
+earned its reputation because of the challenges it poses for reasoning about
+code that uses it.  The problem with eval is that the behavior of a system
+departs from the source text of the program and becomes dependent on the
+program's inputs. This brings challenge to program analysis and security.
+
+While the worst case is indeed pretty bad, previous work has found that
+in many cases eval can be fairly inoccuous or at least much less dangerous
+that the worst case.
+
+There has been research in understanding the behavior of eval in the context
+of web programming and in particular the JavaScript language.
+
+This paper looks at eval in the context of data science. We look at the use
+of eval in the R programming language. And in particular we look at its use
+in three different contexts.  The first context is the implementation of the
+language, how is eval used by the core developers of R in the packages that
+are part of the R virtual machine.  The secoond context is how it is used in
+libraries. And the third context is how it is used by end users of the
+language.
+
+There are reasons to believe that usage patterns may be different. The core
+R code is extremely stable, it has only been maintained for the last 20
+years with very few far reaching changes. The libraries represent a more
+lively ecosystem with new libraries added each day. Typically they are
+developped by experience R developer.  Finally, end user code is often
+written by non-programmers, data scientists who may or may not understand R.
+
+To set up our experiements we have created a corpus composed of three
+distinct data sets.  We have set out the 8 packages that are included with
+the R virtual machine and refer to these as `` base'' packages.  We have
+selected the 400 most widely included libraries and obtained a number of
+client programs to exercises them.  Finally we have obtained 400 end users
+programs from the Kaggle competition.
+
+To understand the behavior of eval we perform a combination of static
+analysis, dynamic analysis and manual inspection. The results presented
+here summarize our findings.
+
+Our infrastructure is publicly available, released in open source, and our
+data, and code will be submitted to artifact evaluation.
 
@@
 - Do we accept overly complex code as the output?
-
-Why do we do this dynamically? Because it raises a chicken-and-egg problem,
-doing it statically would require us to reason about eval based code which may
-not be possible in general.
 
@@
 - cannot replace with (cite study of with construct).
-- 
-\textbf{*}
 
@@
 
-\section{Eval in R}
+Previous work looked at eval in JavaScript
+
+\section{Shade of Eval}
+
 - Talk about environments

### 2020-08-21 inst/paper/main.tex
more stuff

@@
 
-- Why eval is used in R
-- What the community says about using eval
-- What existing tools do about eval
-- Do people use eval for obvious things like looking up variable, if yes, it is
-because they are unaware of alternatives?
-- Why automating the transformation is a hard problem.
-- What is the goal of the work? To replace eval with code for static analyssi or
-for consumption by the users of the language?
-- We only permit transformations that preserve code behavior?
-- Do we accept overly complex code as the output?
-
-What are the challenges? Why is it not trivial?
-- Do we handle aliasing of eval?
-- Simple cases can be subtle? (laziness for ex.
-
-Why can't we replace eval(code) by code?
-
-Intended user of the technique? Which version of R is targeted.
-
-Contributions:
-- Describe the use of eval in R
-- Develop a tool to replace evals guided by dynamic traces
-
-- cannot replace with (cite study of with construct).
-
 \section{Related Work}
 
+\cite{ecoop11}
+
 Previous work looked at eval in JavaScript
+
+There is some work on
 
@@
 
-There are four different kinds of \texttt{eval} in the \emph{base} package:  \texttt{eval}, \texttt{evalq}, \texttt{eval.parent},  \texttt{local}.
-The most important one is \texttt{eval}:
+There are four different kinds of \c{eval} in the \emph{base} package:  \c{eval}, \c{evalq}, \c{eval.parent},  \c{local}.
+The most important one is \c{eval}:
 
@@
 
-\subsubsection{A taxonomy of \texttt{eval}}
+\subsubsection{A taxonomy of \c{eval}}
 
-We first classify \texttt{eval} depending on the resolved expression, \emph{i.e.} the first argument of \texttt{eval} after all possible function calls and symbol resolution in it have been executed.
+We first classify \c{eval} depending on the resolved expression, \emph{i.e.}
+the first argument of \c{eval} after all possible function calls and symbol
+resolution in it have been executed.
 
@@
 
-\subsection{\texttt{base}}
-
-\section{Related Work}
-eval has been widely discussed in the research literature for the challenges it
-imposes for code comprehension and automated code analyses.
-Perhaps, EvalMenDo is the most popular exposition of eval and its
-challenges for JavaScript.
-
-\textbf{Aviral}
-- javascript
-- ruby
-- php
+\subsection{base}
 
 \section{Conclusion}
-\textbf{*}
 
-\section*{Acknowledgment}
-
-\section*{References}
+\bibliographystyle{IEEEtran}
+\bibliography{bib/bibliography,bib/jv}
 

### 2020-08-21 inst/paper/main.tex
abstract

@@
 
-\renewcommand{\c}[1]{{\small\sf{#1}}\xspace}
+\renewcommand{\c}[1]{{\texttt{#1}}\xspace}
+\newcommand{\miss}[1]{{\textcolor{red}{#1}}\xspace}
 
 \begin{abstract}
-  The eval function turns text into code and thus causes difficulties for
-  reasoning about the behavior of software that uses it. This paper is a
-  study of the usage of eval in R, a widely used data science programming
-  language. We combine static and dynamic analysis on a large and
-  representative corpus of R programs. Our results show...
+  The \c{eval} function turns text into code. The widespread use of that
+  capability hinders the static reasoning about software systems. This paper
+  is a large-scale study of a corpus of \miss{3} million lines of R code.
+  We performed static and dynamic analysis of our corpus and report that
+  usage of \c{eval} is indeed wide spread. It is used by expert core
+  developers of the language, experienced library developers, and end-users
+  alike. The patterns of usage range from meta-programming to
+  misunderstandings of what \c{eval} is good for. Case studies give example
+  use-cases and some evidence that automated tools could be used to refactor
+  some of the uses of \c{eval}.
 \end{abstract}
@@
 
-There is some work on
+Othere languages
 
-\section{Shade of Eval}
+Dynamic analysis
+
+\section{Shades of Eval}
 

### 2020-08-21 inst/paper/main.tex
tpyo
Syntax highlighting for R

@@
 \usepackage{textcomp}
+\usepackage{listings}
 \newcommand{\missingNumber}{\textcolor{red}{XX}\xspace}
@@
 \newcommand{\missingGraph}{\textcolor{red}{XXGraph}\xspace}
-\usepackage{xcolor,xspace}
+\usepackage[usenames,dvipsnames]{xcolor} 
+\usepackage{xspace}
+
+\lstset{ 
+	language=R,                     % the language of the code
+	basicstyle=\tiny\ttfamily, % the size of the fonts that are used for the code
+	numbers=left,                   % where to put the line-numbers
+	numberstyle=\tiny\color{Blue},  % the style that is used for the line-numbers
+	stepnumber=1,                   % the step between two line-numbers. If it is 1, each line
+	% will be numbered
+	numbersep=5pt,                  % how far the line-numbers are from the code
+	backgroundcolor=\color{white},  % choose the background color. You must add \usepackage{color}
+	showspaces=false,               % show spaces adding particular underscores
+	showstringspaces=false,         % underline spaces within strings
+	showtabs=false,                 % show tabs within strings adding particular underscores
+	%frame=single,                   % adds a frame around the code
+	rulecolor=\color{black},        % if not set, the frame-color may be changed on line-breaks within not-black text (e.g. commens (green here))
+	tabsize=2,                      % sets default tabsize to 2 spaces
+	captionpos=b,                   % sets the caption-position to bottom
+	breaklines=true,                % sets automatic line breaking
+	breakatwhitespace=false,        % sets if automatic breaks should only happen at whitespace
+	keywordstyle=\color{RoyalBlue},      % keyword style
+	commentstyle=\color{YellowGreen},   % comment style
+	stringstyle=\color{ForestGreen}      % string literal style
+} 
 \begin{document}
@@
 \title{A Large-Scale Study of the Use of Eval in R}
-\author{\vspace{-.8cm}\IEEEauthorblockN{\it\Large The Eval that Data Scientist Do}%
+\author{\vspace{-.8cm}\IEEEauthorblockN{
+    \it\Large The Eval that Data Scientists Do}%
   %\IEEEauthorblockA{ dept} \\ \and
@@
 There are four different kinds of \c{eval} in the \emph{base} package:  \c{eval}, \c{evalq}, \c{eval.parent},  \c{local}.
-The most important one is \c{eval}:
+The most important one is \c{eval}: \lstinline|eval(expr, envir, enclos)|
 
 - the four eval functions in base
+
 - the arguments, expr and envir, enclos
+
 - which expressions are evaluated, which ones are left unchanged

### 2020-08-21 inst/paper/main.tex
The four eval functions

@@
 	language=R,                     % the language of the code
-	basicstyle=\tiny\ttfamily, % the size of the fonts that are used for the code
+	basicstyle=\ttfamily, % the size of the fonts that are used for the code
 	numbers=left,                   % where to put the line-numbers
-	numberstyle=\tiny\color{Blue},  % the style that is used for the line-numbers
+	numberstyle=\color{Blue},  % the style that is used for the line-numbers
 	stepnumber=1,                   % the step between two line-numbers. If it is 1, each line
 	% will be numbered
-	numbersep=5pt,                  % how far the line-numbers are from the code
+	numbersep=2pt,                  % how far the line-numbers are from the code
 	backgroundcolor=\color{white},  % choose the background color. You must add \usepackage{color}
@@
 
-\renewcommand{\c}[1]{{\texttt{#1}}\xspace}
+\renewcommand{\c}[1]{\lstinline{#1}\xspace}
 \newcommand{\miss}[1]{{\textcolor{red}{#1}}\xspace}
+\newcommand{\eval}{\c{eval}}
 
@@
 
-There are four different kinds of \c{eval} in the \emph{base} package:  \c{eval}, \c{evalq}, \c{eval.parent},  \c{local}.
-The most important one is \c{eval}: \lstinline|eval(expr, envir, enclos)|
+There are four different kinds of \eval in the \emph{base} package:  \c{eval}, \c{evalq}, \c{eval.parent},  \c{local}.
+The most important one is \eval:
 
-- the four eval functions in base
+\begin{lstlisting}
+eval <- function (expr, envir = parent.frame(), enclos = if (is.list(envir) || 
+is.pairlist(envir)) parent.frame() else baseenv()) 
+.Internal(eval(expr, envir, enclos))
+\end{lstlisting}
+
+The three other functions are defined using \lstinline|eval|.
+
+\begin{lstlisting}
+evalq <- function (expr, envir = parent.frame(), enclos = if (is.list(envir) || 	is.pairlist(envir)) parent.frame() else baseenv()) 
+	.Internal(eval(substitute(expr), envir, enclos))
+
+eval.parent <- function (expr, n = 1) 
+{
+p <- parent.frame(n + 1)
+eval(expr, p)
+}
+
+local <- function (expr, envir = new.env()) eval.parent(substitute(eval(quote(expr), envir)))
+\end{lstlisting}
 

### 2020-08-21 inst/analysis/insights.R
get_expr does not failt anymore on NA

@@ -14,6 +14,10 @@ nb_eval_call_sites <- function(eval_calls)
 
 get_expr <- function(eval_call)
 {
+  if(is.na(eval_call))
+  {
+    return(NA)
+  }
   # Special case for expressions starting with _ (such as _inherit in ggproto)
   escaped_eval_call <- if(startsWith(eval_call, "_"))
   {
@@ -65,10 +69,14 @@ parse_all <- function(x, filename,allow_error) {}
 extract_args_parse <- function(eval_call) 
 {
   exp <- get_expr(eval_call)
-  exp <- call_standardise(exp)
+  exp <- tryCatch(call_standardise(exp),
+                  error = function(c) 
+                    if(exp[[2]] == "...") { names (exp)[[2]] <- "dots" ; return(exp)}
+                    else stop(paste0("extract_arg failed with: ", eval_call))
+  )
   args <- map_chr(as.list(exp[-1]), function(chr) { paste(deparse(chr), collapse = "\n")})
   
-  return(args[str_detect(names(args), "file|text|n|s|keep.source|srcfile")])# "file|text|n|s|prompt|keep.source|srcfile|code"
+  return(args[str_detect(names(args), "file|text|n|s|keep.source|srcfile|dots")])# "file|text|n|s|prompt|keep.source|srcfile|code"
 }
 
 # See extract_inner_exp which takes a str
@@ -217,11 +225,11 @@ extract_package_name <- function(src_ref, file)
   # - :/R : extract the package name from the file path in the column path
   case_when(
     is.na(src_ref) ~ "base",
-    str_starts(src_ref, fixed("./R/")) ~ "core",
+    str_starts(src_ref, fixed("./R/")) ~ str_match(file, "[^/]*/[^/]*/([^/]*)/.*")[[2]],
     str_starts(src_ref, fixed("/tmp/")) ~ str_match(src_ref, "/tmp/Rtmp[^/]*/R\\.INSTALL[^/]*/([^/]+)/.*")[[2]],
     str_starts(src_ref, fixed("/mnt/nvme0/")) ~ str_match(src_ref, "/mnt/nvme0/R/project-evalR/library/4.0/instrumentr/srcref/([^/]*)/.*")[[2]],
     str_starts(src_ref, fixed("test")) ~ str_match(src_ref, "([^/]*)/.*")[[2]],
-    str_starts(src_ref, ":/") ~ str_match(file, "[^/]*/[^/]*/([^/]*)/.*")[[2]],
[…]

### 2020-08-21 inst/paper/main.tex

@@
 \begin{lstlisting}
-eval <- function (expr, envir = parent.frame(), enclos = if (is.list(envir) || 
-is.pairlist(envir)) parent.frame() else baseenv()) 
+eval <- function (expr, envir = parent.frame(), enclos = if (is.list(envir) || is.pairlist(envir)) parent.frame() else baseenv())  
 	.Internal(eval(expr, envir, enclos))
@@
 
-- the arguments, expr and envir, enclos
+\c{expr} is the expression to be evaluated. It is not a string as in javascript. In order to evaluate a string, it should be first parsed using \c(parse(text = str)) to get an expression.
 
-- which expressions are evaluated, which ones are left unchanged
+\eval makes it possible to specify the environment in which to evaluate the expression.
+\c{enclos} is an additional environment used for lookup when \c{envir} is a list, a pairlist or a dataframe.
+In that case, \c{envir} is copied into a temporary environment enclosed in \c{enclos}.
+
+The \c{expr} argument of \eval is first evaluated in the current scope, then passed to the \eval.
+Quoting that argument prevents if from being evaluated in the current environment.
+\c{evalq} is a shortcut for \c{eval(quote(expr), ...)}.
+In the following code snippet, variable \c{r} is first looked up in the current scope with \eval whereas it is looked up in \c{env} with \c{evalq}.
+
+\begin{lstlisting}
+> r <- 5
+> env <- new.env()
+> env$r <- 9
+> eval(r, env)
+[1] 5
+> evalq(r, envir = env)
+[1] 9
+\end{lstlisting}
+
+\eval will return unchanged basic types such as vectors, functions, strings and environments.
+Promises are forced by \eval.
+Symbols are looked up in the current environment.
+Calls and expressions are evaluated.
 

### 2020-08-21 inst/analysis/enrich_dataset.Rmd
Optimizations

@@
 params:
-  data_dirpath: "/var/lib/R/project-evalR/run/package-evals-traced-corpus.4/"
-  data_filepath: "/var/lib/R/project-evalR/run/package-evals-traced-corpus.4/calls.fst"
+  data_dirpath: "/var/lib/R/project-evalR/run/package-evals-traced-corpus.5/"
+  data_filepath: "/var/lib/R/project-evalR/run/package-evals-traced-corpus.5/calls.fst"
 knit: (function(inputFile, encoding) {
@@
 ```{r}
-eval_calls_base <- eval_calls_base %>% count(across(c(-eval_call_id, -starts_with("caller_stack"))), name = "nb_eval_calls") 
-eval_calls <- eval_calls %>% count(across(c(-eval_call_id)), name = "nb_eval_calls")
-```
-
-```{r}
 eval_calls_base <- eval_calls_base %>% 
-    resolve_sexp_name(expr_expression_type) %>%
-    resolve_sexp_name(expr_resolved_type) %>%
-    resolve_sexp_name(enclos_type) %>%
-    resolve_sexp_name(envir_type)
-
-eval_calls <- eval_calls %>% 
-    mutate(eval_call_package = extract_package_name(eval_call_srcref, file)) %>% 
     resolve_sexp_name(expr_expression_type) %>%
@@
 ```{r}
-#eval_calls <- eval_calls %>% mutate(parse_args = if_else(is.na(expr_parsed_expression), NA, map_chr(expr_parsed_expression, extract_args_parse)))
+eval_calls_base <- eval_calls_base %>% count(across(c(-eval_call_id, -starts_with("caller_stack"))), name = "nb_eval_calls") 
+eval_calls_dedup <- eval_calls %>% count(across(c(-eval_call_id)), name = "nb_eval_calls")
+eval_calls_dedup_more <- eval_calls %>% count(across(c(-eval_call_id, -starts_with("caller_stack"))), name = "nb_eval_calls") 
+```
+
+```{r}
+eval_calls <- eval_calls_dedup_more %>% mutate(parse_args = if_else(is.na(expr_parsed_expression), NA, map_chr(expr_parsed_expression, extract_args_parse)))
+```
+
+```{r}
+eval_calls_dedup_more <- eval_calls_dedup_more %>% mutate(eval_call_package = map2_chr(eval_call_srcref, file, extract_package_name)) 
+```
+
+```{r}
+eval_calls_dedup_more <- eval_calls_dedup_more %>% 
+    resolve_sexp_name(expr_expression_type) %>%
+    resolve_sexp_name(expr_resolved_type) %>%
+    resolve_sexp_name(enclos_type) %>%
+    resolve_sexp_name(envir_type)
 ```
@@
 ```{r write_back}
-write_fst(eval_calls, paste0(params$data_dirpath, "deduplicated_calls.fst"))
+write_fst(eval_calls_dedup, paste0(params$data_dirpath, "deduplicated_calls.fst"))
+
+write_fst(eval_calls_dedup_more, paste0(params$data_dirpath, "deduplicated_more_calls.fst"))
 
 write_fst(eval_calls_base, paste0(params$data_dirpath, "base_calls.fst"))
+
+write_fst(eval_calls, paste0(params$data_dirpath, "enriched_calls.fst"))
 ```

### 2020-08-21 inst/analysis/enrich_dataset.Rmd
Right eval_call_package

@@
 ```{r}
-eval_calls_base <- eval_calls_base %>% count(across(c(-eval_call_id, -starts_with("caller_stack"))), name = "nb_eval_calls") 
+eval_calls_base <- eval_calls_base %>% count(across(c(-eval_call_id, -starts_with("caller_stack"))), name = "nb_ev_calls") 
 eval_calls_dedup <- eval_calls %>% count(across(c(-eval_call_id)), name = "nb_eval_calls")
-eval_calls_dedup_more <- eval_calls %>% count(across(c(-eval_call_id, -starts_with("caller_stack"))), name = "nb_eval_calls") 
+eval_calls_dedup_more <- eval_calls %>% count(across(c(-eval_call_id, -starts_with("caller_stack"))), name = "nb_ev_calls") 
 ```
@@
 ```{r}
-eval_calls <- eval_calls_dedup_more %>% mutate(parse_args = if_else(is.na(expr_parsed_expression), NA, map_chr(expr_parsed_expression, extract_args_parse)))
+eval_calls_dedup_more <- eval_calls_dedup_more %>% mutate(parse_args = map(expr_parsed_expression, function(e) { if(!is.na(e) && str_starts(e, "(parse|str2lang|str2expression)\\("))   extract_args_parse(e) else list() }))  %>% unnest_wider(parse_args)
 ```
@@
 
+```{r}
+eval_calls <- eval_calls %>% select(-ends_with("_type")) %>% left_join(eval_calls_dedup_more)
+eval_calls_dedup <- eval_calls_dedup %>% select(-ends_with("_type"))%>% left_join(eval_calls_dedup_more)
+```
+
 ```{r write_back}

### 2020-08-21 inst/analysis/insights.R

@@ -71,12 +71,13 @@ extract_args_parse <- function(eval_call)
   exp <- get_expr(eval_call)
   exp <- tryCatch(call_standardise(exp),
                   error = function(c) 
-                    if(exp[[2]] == "...") { names (exp)[[2]] <- "dots" ; return(exp)}
+                    if(length(exp) >= 2 && exp[[2]] == "...") { names(exp)[[2]] <- "dots" ; return(exp)}
                     else stop(paste0("extract_arg failed with: ", eval_call))
   )
   args <- map_chr(as.list(exp[-1]), function(chr) { paste(deparse(chr), collapse = "\n")})
+  names(args) <- map_chr(names(args), function(chr) paste0("parse_args_", chr))
   
-  return(args[str_detect(names(args), "file|text|n|s|keep.source|srcfile|dots")])# "file|text|n|s|prompt|keep.source|srcfile|code"
+  return(args[str_detect(names(args), "^parse_args_(file_name|text|n|s|keep\\.source|srcfile|dots)$")])# "file|text|n|s|prompt|keep.source|srcfile|code"
 }
 
 # See extract_inner_exp which takes a str
@@ -213,6 +214,30 @@ resolve_sexp_name <- function(df, var) {
     rename(!!en_var:=name)
 }
 
+R_LIB_SRC <- "/var/lib/R/project-evalR/R-4.0.2/src/library"
+CORE_PACKAGES <- c(
+  "compiler",
+  "graphics",
+  "grDevices",
+  "grid",
+  "methods",
+  "parallel",
+  "profile",
+  "splines",
+  "stats",
+  "stats4",
+  "tcltk",
+  "tools",
+  "utils"
+)
+core_package_files <- map_dfr(CORE_PACKAGES, function(x) {
+  p <- file.path(R_LIB_SRC, x, "R")
+  f <- list.files(p, pattern="\\.R$", recursive=FALSE)
+  f <- f[!str_ends(f, "all\\.R")]
+  f <- file.path("./R", f)
+  tibble(package=x, file=f)
+})
+
 
 extract_package_name <- function(src_ref, file)
 {
@@ -224,12 +249,12 @@ extract_package_name <- function(src_ref, file)
   # - /testit/... or /testthat/... : it is the testit or testthat packages
[…]

### 2020-08-21 inst/analysis/analysis_stable.Rmd
Handle deduplicated datasets (wt, weight arguments)

--- 2020-08-19 inst/analysis/analysis_stable.Rmd
@@
 params:
-    data_filepath: "/var/lib/R/project-evalR/run/package-evals-traced-corpus.3/calls.fst"
+    data_filepath: "/var/lib/R/project-evalR/run/package-evals-traced-corpus.5/deduplicated_more_calls.fst"
     #data_filepath: "/var/lib/R/project-evalR/run/package-evals-traced-all.1/calls.fst"
@@
     as_tibble() %>%
-    mutate(eval_call_package = extract_package_name(eval_call_srcref, file)) %>% 
-    resolve_sexp_name(expr_expression_type) %>%
-    resolve_sexp_name(expr_resolved_type) %>%
-    resolve_sexp_name(enclos_type) %>%
-    resolve_sexp_name(envir_type)
+    mutate(eval_call_package = extract_package_name(eval_call_srcref, file))
 
@@
     group_by(eval_function, eval_call_srcref) %>%
-    summarize(count = n()) %>%
+    summarize(count = sum(nb_ev_calls)) %>%
     summarize(call_count = sum(count), callsite_count = n())
@@
 ```{r nb_eval_calls_per_package}
-nb_eval_calls_per_package <- eval_calls %>% group_by(caller_package) %>% summarise(nb_eval_calls_package = n(), per_cent = nb_eval_calls_package / nb_eval_calls * 100) %>% arrange(desc(nb_eval_calls_package))
+nb_eval_calls_per_package <- eval_calls %>% group_by(caller_package) %>% summarise(nb_eval_calls_package = sum(nb_ev_calls), per_cent = nb_eval_calls_package / nb_eval_calls * 100) %>% arrange(desc(nb_eval_calls_package))
 ```
@@
 
-There are `r nrow(parsed_base)` `eval` calls with `parse` in *base*, which is `r nrow(parsed_base) / nrow(filter(eval_calls, eval_call_package == "base"))`%.
+There are `r count(parsed_base, wt = nb_ev_calls)` `eval` calls with `parse` in *base*, which is `r count(parsed_base, wt = nb_ev_calls) / count(filter(eval_calls, eval_call_package == "base"), wt=nb_ev_calls)`%.
 
@@
 eval_calls %>% ggplot() +
-  geom_histogram(aes(x = expr_expression_length), na.rm = TRUE, binwidth=1) +
+  geom_histogram(aes(x = expr_expression_length, weight = nb_ev_calls), na.rm = TRUE, binwidth=1) +
   scale_x_log10() +  scale_y_sqrt()
@@
 eval_calls %>% filter(expr_expression_length < 25000) %>% ggplot() +
-  geom_histogram(aes(x = expr_expression_length), na.rm = TRUE, binwidth=2)  + scale_x_sqrt() +  scale_y_sqrt() 
+  geom_histogram(aes(x = expr_expression_length, weight = nb_ev_calls), na.rm = TRUE, binwidth=2)  + scale_x_sqrt() +  scale_y_sqrt() 
 ```
@@
 eval_calls %>% filter(is.na(eval_call_srcref), caller_package != "base") %>% ggplot() +
-  geom_histogram(aes(x = expr_expression_length), na.rm = TRUE, binwidth=1) + scale_x_sqrt() + scale_y_sqrt()
+  geom_histogram(aes(x = expr_expression_length, weight = nb_ev_calls), na.rm = TRUE, binwidth=1) + scale_x_sqrt() + scale_y_sqrt()
 ```
@@
 
-### Classifying according to the type of the `expr` argument
-
-We look at the first argument of `eval`, the `expr` argument.
+We look at the first argument of `eval`, the `expr` argument, after it is resolved.
 It is the expression to be evaluated.
@@
 
-```{r types}
-per_expr_type <- eval_calls %>% count(expr_expression_type)
+```{r types_resolved}
+per_expr_resolved_type <- eval_calls %>% count(expr_resolved_type, wt = nb_ev_calls)
 ```
@@
 ```{r}
- per_expr_type %>% mutate(per_cent = n / nb_eval_calls * 100 ) %>% arrange(desc(n)) %>% knitr::kable()
+ per_expr_resolved_type %>% mutate(per_cent = n / nb_eval_calls * 100 ) %>% arrange(desc(n)) %>% knitr::kable()
 ```
@@
 ```{r}
-per_expr_type %>% ggplot() +
-  geom_col(aes(x = fct_reorder(expr_expression_type, n), y = n)) +
+per_expr_resolved_type %>% ggplot() +
+  geom_col(aes(x = fct_reorder(expr_resolved_type, n), y = n)) +
   scale_y_sqrt() + 
@@
 
-Expressions of types other than `call`,  `expression`, `name`, `bytecode` and `promise` are just returned unchanged.
+There is a much larger variety of types for the resolved expressions.
 
-```{r not_unchanged_eval}
-changed_evals <- eval_calls %>% mutate(changed =  fct_collapse(expr_expression_type, lang_types = c("LANGSXP", "EXPRSXP", "SYMSXP", "PROMSXP", "BCODESXP"), other_level = "other_types")) %>% count(changed) %>% mutate(per_cent = n / nb_eval_calls * 100 )
+```{r not_unchanged_resolved_eval}
+changed_resolved_evals <- eval_calls %>% mutate(changed =  fct_collapse(expr_resolved_type, lang_types = c("LANGSXP", "EXPRSXP", "SYMSXP", "PROMSXP", "BCODESXP"), other_level = "other_types")) %>% count(changed, wt = nb_eval_calls) %>% mutate(per_cent = n / nb_eval_calls * 100 )
 ```
 
-We can remove `eval` and leave there the expression in `r changed_evals[2, "per_cent"][[1]]`% of the eval calls (i.e. `r changed_evals[2, "n"][[1]]`).
+We can remove `eval` and leave there the expression in `r changed_resolved_evals[2, "per_cent"][[1]]`% of the eval calls (i.e. `r changed_resolved_evals[2, "n"][[1]]`).
 
-### `expr` argument when it is a call
+We can do the same analysis as for the non-resolved expressions:
 
-```{r expr_call_arg}
-expr_call_arg <- eval_calls %>% filter(expr_expression_type == "LANGSXP") %>% 
-  mutate(function_expr = map_chr(expr_expression, function_name))
+```{r expr_resolved__call_arg}
+expr_resolved_call_arg <- eval_calls %>% filter(expr_resolved_type == "LANGSXP")
 ```
@@
 
-```{r top_level_function_expr}
-expr_call_arg %>% count(function_expr) %>% arrange(desc(n))
+```{r top_level_function_resolved_expr}
+expr_resolved_call_arg %>% count(expr_resolved_function, wt=nb_ev_calls) %>% arrange(desc(n))
 ```
@@
 
-```{r group_expr_functions}
-group_expr_functions <- expr_call_arg %>%
-                            mutate(function_expr2 = if_else(str_starts(function_expr, fixed("(function(")), "anonymous", function_expr))
-
-group_expr_functions <-  group_expr_functions %>% 
-                              mutate(group_function_expr = fct_collapse(function_expr2, 
-                              assignment = c("<-", "[[<-", "[<-", "$<-", "<<-", "="),
-                              read = c("[", "[[", "$"),
-                              parsing = c("parse", "str2lang", "str2expression", "parse_only", "base::parse"),
-                              defusing = c("quote", "substitute", "bquote", "expr", "quo"),
-                              expressions = c("as.list", "expression", "as.name", "as.expression", "list", "as.data.frame", "as.character", "{", "modify_lang"),
-                              calling = c("call", "as.call", "do.call"),
-                              environment = c("environment", "as.environment"),
-                              classAssign = c("@<-"),
-                              classRead = c("slot", "@"),
-                              classDef = c(".Primitive(\"class\")"),
-                              logical = c("<", ">", "is.element"),
-                              choice = c("if"),
-                              anonymous = c("anonymous"),
-                              function_meta = c("sys.calls", "function", "body"),
-                              formula = c("~", "deriv", "D", "model.frame", "rootftfunc", "contest.lmerModLmerTest", "JacFunc", "gamlss.family", "contest1D.lmerModLmerTest", "contestMD.lmerModLmerTest", "reexpr"),
-                              other_level = "other" ))
-```
-
-```{r}
-group_expr_functions %>% count(group_function_expr) %>% arrange(desc(n)) %>% knitr::kable()
-```
-
-```{r}
-group_expr_functions %>% ggplot() +
-  geom_bar(aes(x = fct_rev(fct_infreq(group_function_expr)))) +
-  scale_y_sqrt() +
-  coord_flip() +
-  labs(title = "Category of calls", y = "n", x = "category" ) 
-```
-
-```{r }
-group_expr_functions %>% ggplot() +
-  geom_bar(aes(x = fct_rev(fct_infreq(group_function_expr)))) +
-  scale_y_sqrt() +
-  coord_flip() +
-  labs(title = "Category of calls", y = "n", x = "category" ) +
-  facet_wrap(vars(eval_function))
-```
-
-### For resolved `expr`
-
-We can do the same for resolved expressions:
-
-```{r types_resolved}
-per_expr_resolved_type <- eval_calls %>% count(expr_resolved_type)
-```
-
-```{r}
- per_expr_resolved_type %>% mutate(per_cent = n / nb_eval_calls * 100 ) %>% arrange(desc(n)) %>% knitr::kable()
-```
-
-```{r}
-per_expr_resolved_type %>% ggplot() +
-  geom_col(aes(x = fct_reorder(expr_resolved_type, n), y = n)) +
-  scale_y_sqrt() + 
-  coord_flip() +
-  labs(title = "Calls per type", y = "n", x = "type" ) 
-```
-
-There is a much larger variety of types for the resolved expressions.
-
-```{r not_unchanged_resolved_eval}
-changed_resolved_evals <- eval_calls %>% mutate(changed =  fct_collapse(expr_resolved_type, lang_types = c("LANGSXP", "EXPRSXP", "SYMSXP", "PROMSXP", "BCODESXP"), other_level = "other_types")) %>% count(changed) %>% mutate(per_cent = n / nb_eval_calls * 100 )
-```
-
-We can remove `eval` and leave there the expression in `r changed_resolved_evals[2, "per_cent"][[1]]`% of the eval calls (i.e. `r changed_resolved_evals[2, "n"][[1]]`).
-
-We can do the same analysis as for the non-resolved expressions:
-
-```{r expr_resolved__call_arg}
-expr_resolved_call_arg <- eval_calls %>% filter(expr_resolved_type == "LANGSXP") %>% 
-  mutate(function_expr = map_chr(expr_resolved, function_name))
-```
-
-The following top level functions are used:
-
-```{r top_level_function_resolved_expr}
-expr_resolved_call_arg %>% count(function_expr) %>% arrange(desc(n))
-```
-
 ```{r group_expr_resolved_functions}
 group_expr_resolved_functions <- expr_resolved_call_arg %>%
-                            mutate(function_expr2 = if_else(str_starts(function_expr, fixed("(function(")), "anonymous", function_expr)) %>%
+                            mutate(function_expr2 = groupify_function(expr_resolved_function)) %>%
                             mutate(group_function_expr = fct_collapse(function_expr2, 
@@
 ```{r}
-group_expr_resolved_functions %>% count(group_function_expr) %>% arrange(desc(n)) %>% knitr::kable()
+group_expr_resolved_functions %>% count(group_function_expr, wt = nb_ev_calls) %>% arrange(desc(n)) %>% knitr::kable()
 ```
@@
 group_expr_resolved_functions %>% ggplot() +
-  geom_bar(aes(x = fct_rev(fct_infreq(group_function_expr)))) +
+  geom_bar(aes(x = fct_rev(fct_infreq(group_function_expr)), weight = nb_ev_calls)) +
   scale_y_sqrt() +
@@
 ```{r }
-group_expr_functions %>% ggplot() +
-  geom_bar(aes(x = fct_rev(fct_infreq(group_function_expr)))) +
+group_expr_resolved_functions %>% ggplot() +
+  geom_bar(aes(x = fct_rev(fct_infreq(group_function_expr)), weight = nb_ev_calls)) +
   scale_y_sqrt() +
@@
   labs(title = "Category of calls", y = "n", x = "category" ) +
+  facet_wrap(vars(eval_function))
+```
+
+We also look at the arity of the top-level function:
+
+```{r arity_expr_resolved}
+expr_resolved_call_arg %>% count(expr_expression_args_num, sort = TRUE) %>% knitr::kable()
+```
+
+```{r arity_expr_resolved_plot}
+expr_resolved_call_arg %>% ggplot() + 
+  geom_bar(aes(x = expr_resolved_args_num, weight = nb_ev_calls)) +
+  scale_y_sqrt() +
+  coord_flip() + 
+  labs(title = "Arity of calls", x = "arity", y = "n")
+
+```
+
+```{r arity_expr_resolved_plot_per_eval}
+expr_resolved_call_arg %>% ggplot() + 
+  geom_bar(aes(x = expr_resolved_args_num, weight = nb_ev_calls)) + 
+  scale_y_sqrt() +
+  coord_flip() + 
+  labs(title = "Arity of calls", x = "arity", y = "n") +
   facet_wrap(vars(eval_function))
@@
 parsed_evals <- eval_calls %>% filter(!is.na(expr_parsed_expression))
-nb_parsed_evals <- parsed_evals %>% nrow()
+nb_parsed_evals <- parsed_evals %>% count(wt = nb_ev_calls)
 ```
@@
 
-```{r parsed_evals_source}
-parsed_evals_args <- parsed_evals %>% filter(str_detect(expr_parsed_expression, "(parse|str2lang|str2expression)\\(")) %>% mutate(parse_args = map(expr_parsed_expression, extract_args_parse)) %>% unnest_wider(parse_args,  names_sep = "_")
+```{r}
+if("parse_args_file" %in% names(parsed_evals))
+{
+  parsed_files <- parsed_evals %>% filter(!is.na(parse_args_file))
+  nb_file <- parsed_files %>% count(wt = nb_ev_calls)
+  nb_file_call_sites <- nb_eval_call_sites(parsed_files)
+} else
+{
+  nb_file <- 0
+  nb_file_call_sites <- 0
+}
+
 ```
@@
 ```{r}
-parsed_files <- parsed_evals_args %>% filter(!is.na(parse_args_file))
-nb_file <- parsed_files %>% nrow()
-nb_file_call_sites <- nb_eval_call_sites(parsed_files)
-```
-
-```{r}
-parsed_texts <- parsed_evals_args %>% filter(!is.na(parse_args_text) | !is.na(parse_args_s))
-nb_text <-  parsed_texts %>% nrow()
+parsed_texts <- parsed_evals %>% filter(!is.na(parse_args_text) | !is.na(parse_args_s))
+nb_text <-  parsed_texts %>% count(wt = nb_ev_calls)
 nb_text_call_sites <- nb_eval_call_sites(parsed_texts)

### 2020-08-21 inst/analysis/corpus-analysis.Rmd
Split corpus-creation and corpus-analysis

@@
 output: 
-  html_document: 
-    theme: united
-    toc: yes
-    toc_float: yes
+  rmdformats::readthedown:
+        code_folding: hide
+        lightbox: true
+        gallery: false
+        df_print: paged
+        toc_depth: 3
+date: "`r Sys.Date()`"
 editor_options: 
@@
 params:
-  data_dir: /var/lib/R/project-evalR/revalstudy/data
-  num_packages: 500
-  output_dir: /var/lib/R/project-evalR/revalstudy/data
-  which_dependencies:
-    - Depends
-    - Imports
+knit: (function(inputFile, encoding) {
+  rmarkdown::render(inputFile, encoding = encoding, output_dir = "html") })
 ---
-
-```{r setup, include=FALSE}
-for (x in c(
-  "dplyr", 
-  "DT", 
-  "fs", 
-  "readr",
-  "purrr"
-  )) {
-  suppressPackageStartupMessages(library(x, character.only=TRUE))
-}
-
-knitr::opts_chunk$set(echo = TRUE)
-
-# TODO: move into the package
-source("inc/paths.R")
-source("inc/setup.R")
-source("inc/latextags.R")
-
-create_tags(path(TAGS_DIR, "corpus.tex"), prefix="", default=TRUE)
-```
-
-## Description
-
-The aim of this notebook is to prepare the corpus of R packages to be included in the paper.
-
-### Input
-
-Nothing
-
-### Output
-
-- ``r CORPUS_FILE`` - the corpus in the corpus file (one package file line)
-- ``r CORPUS_REVDEPS`` - CSV of package / reverse dependency pairs
-- ``r CORPUS_REVDEPS_FILE`` - all unique reverse dependencies in the corpus file format
-
-## Prepare the corpus
-
-We are interested in TOP `r params$num_packages` packages based on the number of reverse dependencies.
-
-```{r get all available packages}
-# TODO this should be installed.packages() once they are installed
-cran_packages_raw <- available.packages()
-cran_packages <- as_tibble(cran_packages_raw) %>%
-  select(package=Package, version=Version)
-```
-
-### Package reverse dependencies
-
-There are two functions `package_dependencies` and `dependsOnPkgs`.
-Not sure what is the difference, but the `dependsOnPkgs` is very slow.
-
-```{r compute reverse dependencies}
-pkgs_revdeps_raw <- 
-  tools::package_dependencies(
-    cran_packages$package,
-    db=cran_packages_raw,
-    which=params$which_dependencies,
-    reverse=TRUE,
-    recursive=FALSE
-  ) %>% 
-  imap_dfr(~tibble(package=.y, revdep=.x))
-
-pkgs_revdeps <- 
-  pkgs_revdeps_raw %>%
-  count(package) %>%
-  rename(revdeps=n)
-```
-
-### TOP packages based on number of reverse dependencies
-
-```{r compute TOP n packages}
-pkgs_top_revdeps <-
-  pkgs_revdeps %>%
-  top_n(n = params$num_packages, wt=revdeps) %>%
-  arrange(desc(revdeps))
-
-# cup the number of reverse dependencies to the required number
-pkgs_top_revdeps <- head(pkgs_top_revdeps, min(nrow(pkgs_top_revdeps), params$num_packages))
-```
-
-### Missing packages from TOP 100 downloads
-
-```{r get last month most downloaded packages}
-pkgs_top_downloaded <-
-  cranlogs::cran_top_downloads("last-month", 100)
-```
-
-```{r compute missing packages from the most downloaded ones}
-missing <- setdiff(pkgs_top_downloaded$package, pkgs_top_revdeps$package)
-missing_revdeps <- filter(pkgs_revdeps, package %in% missing)
-missing_revdeps %>%
-  left_join(pkgs_top_downloaded, by="package") %>%
-  datatable()
-```
-
-### Corpus
-
-```{r render selected packages}
-pkgs_top_revdeps %>%
-  left_join(pkgs_top_downloaded, by="package") %>%
-  arrange(desc(revdeps)) %>%
-  datatable()
-```
-
-```{r write down corpus}
-write_lines(pkgs_top_revdeps$package, CORPUS_FILE)
-write_csv(pkgs_revdeps, CORPUS_REVDEPS)
-write_lines(unique(pkgs_revdeps$revdep), CORPUS_REVDEPS_FILE)
-```

### 2020-08-21 inst/analysis/corpus-creation.Rmd
Split corpus-creation and corpus-analysis

---
title: "Corpus Creation"
output: 
  html_document: 
    theme: united
    toc: yes
    toc_float: yes
editor_options: 
  chunk_output_type: console
params:
  data_dir: /var/lib/R/project-evalR/revalstudy/data
  num_packages: 500
  output_dir: /var/lib/R/project-evalR/revalstudy/data
  which_dependencies:
    - Depends
    - Imports
---

```{r setup, include=FALSE}
for (x in c(
  "dplyr", 
  "DT", 
  "fs", 
  "readr",
  "purrr"
  )) {
  suppressPackageStartupMessages(library(x, character.only=TRUE))
}

knitr::opts_chunk$set(echo = TRUE)

# TODO: move into the package
source("inc/paths.R")
source("inc/setup.R")
source("inc/latextags.R")

create_tags(path(TAGS_DIR, "corpus.tex"), prefix="", default=TRUE)
```

## Description

The aim of this notebook is to prepare the corpus of R packages to be included in the paper.

### Output

- ``r CORPUS_FILE`` - the corpus in the corpus file (one package file line)
- ``r CORPUS_REVDEPS`` - CSV of package / reverse dependency pairs
- ``r CORPUS_REVDEPS_FILE`` - all unique reverse dependencies in the corpus file format

## Prepare the corpus

We are interested in TOP `r params$num_packages` packages based on the number of reverse dependencies.

```{r get all available packages}
# TODO this should be installed.packages() once they are installed
cran_packages_raw <- available.packages()
cran_packages <- as_tibble(cran_packages_raw) %>%
  select(package=Package, version=Version)
```

### Package reverse dependencies

There are two functions `package_dependencies` and `dependsOnPkgs`.
Not sure what is the difference, but the `dependsOnPkgs` is very slow.

```{r compute reverse dependencies}
pkgs_revdeps_raw <- 
  tools::package_dependencies(
    cran_packages$package,
    db=cran_packages_raw,
    which=params$which_dependencies,
    reverse=TRUE,
    recursive=FALSE
  ) %>% 
  imap_dfr(~tibble(package=.y, revdep=.x))

pkgs_revdeps <- 
  pkgs_revdeps_raw %>%
  count(package) %>%
  rename(revdeps=n)
```

### TOP packages based on number of reverse dependencies

```{r compute TOP n packages}
pkgs_top_revdeps <-
  pkgs_revdeps %>%
  top_n(n = params$num_packages, wt=revdeps) %>%
  arrange(desc(revdeps))

# cup the number of reverse dependencies to the required number
pkgs_top_revdeps <- head(pkgs_top_revdeps, min(nrow(pkgs_top_revdeps), params$num_packages))
```

### Missing packages from TOP 100 downloads

```{r get last month most downloaded packages}
pkgs_top_downloaded <-
  cranlogs::cran_top_downloads("last-month", 100)
```

```{r compute missing packages from the most downloaded ones}
missing <- setdiff(pkgs_top_downloaded$package, pkgs_top_revdeps$package)
missing_revdeps <- filter(pkgs_revdeps, package %in% missing)
missing_revdeps %>%
  left_join(pkgs_top_downloaded, by="package") %>%
  datatable()
```

### Corpus

Putting it all together and outputting the results.

```{r render selected packages}
pkgs_top_revdeps %>%
  left_join(pkgs_top_downloaded, by="package") %>%
  arrange(desc(revdeps)) %>%
  datatable()
```

```{r write down corpus}
pkgs_revdeps

write_lines(pkgs_top_revdeps$package, CORPUS_FILE)
write_csv(pkgs_revdeps, CORPUS_REVDEPS)
write_lines(unique(pkgs_revdeps_raw$revdep), CORPUS_REVDEPS_FILE)
write_lines(unique(), CORPUS_ALL_PACKAGES_FILE)
```

### 2020-08-21 inst/analysis/trace-run-analysis.Rmd
Add trace run analysis

---
title: "Trace run analysis"
output: 
  html_document: 
    theme: united
    toc: yes
    toc_float: yes
editor_options: 
  chunk_output_type: console
params:
  data_dir: /var/lib/R/project-evalR/run/package-evals-traced-corpus.3
  output_dir: /var/lib/R/project-evalR/revalstudy/data
  paper_dir: /var/lib/R/project-evalR/revalstudy/inst/paper
  base_url: http://prl3:8003/file_show?path=
---

```{r setup, include=FALSE}
for (x in c(
  "dplyr", 
  "DT", 
  "fs", 
  "purrr",
  "readr",
  "runr",
  "stringr"
  )) {
  suppressPackageStartupMessages(library(x, character.only=TRUE))
}

knitr::opts_chunk$set(echo = TRUE)

# TODO: move into the package
source("inc/paths.R")
source("inc/setup.R")
source("inc/latextags.R")

create_tags(path(TAGS_DIR, "trace-run.tex"), prefix="TC", default=TRUE)

PARALLEL_LOG_FILE <- path(params$data_dir, "parallel.log")
```

## Description

The aim of this notebook is to analyze the run of the tracer.

### Output

- ``r PARALLEL_LOG_FILE`` - GNU parallel log file

## Load data

```{r}
parallel_log <- 
  read_parallel_log(PARALLEL_LOG_FILE) %>%
  mutate(command=str_replace(command, ".*/(.*)$", "\\1")) %>%
  select(package=command, everything()) %>%
  select(-seq, -host, -send, -receive)
```

## Overview

```{r}
failed_packages <- filter(parallel_log, exitval != 0)
errored_packages <- filter(parallel_log, exitval == 1)
timeouted_packages <- filter(parallel_log, exitval == -1)

overview_table(
  r("Run packages", parallel_log),
  r("Succeeded packages", filter(parallel_log, exitval == 0)),
  r("Failed packages",    failed_packages),
  r("Failed packages ratio", ratio(failed_packages, parallel_log)),
  r("Timeouted packages", timeouted_packages),
  r("Errored packages", errored_packages),
  r("Other errored packages",   filter(parallel_log, !(exitval %in% c(-1, 0, 1)))),
  r("Duration", max(parallel_log$starttime + parallel_log$jobruntime) - min(parallel_log$starttime))
)
```

## Failed packages

### Errored packages

Running tracer on these packages resulted in exit code `1`.

```{r}
errored_packages %>%
  mutate(
    stdout=paste("<a href=", URLencode(file.path(params$base_url, params$data_dir, package, "stdout")), ">stdout</a>"),
    stderr=paste("<a href=", URLencode(file.path(params$base_url, params$data_dir, package, "stderr")), ">stderr</a>")
  ) %>%
  my_datatable(escape=FALSE)
```

### Timeouted packages

The GNU parallel killed the tracing process.

```{r}
timeouted_packages %>%
  my_datatable()
```

### 2020-08-21 inst/analysis/inc/paths.R
Update pipeline makefiles

@@ -1,14 +1,16 @@
+DATA_DIR  <- path(params$base_dir, "data")
+PAPER_DIR <- path(params$base_dir, "paper")
+TAGS_DIR  <- path(PAPER_DIR, "tags")
+
 ## INPUTs
 
-PACKAGE_METADATA     <- path(params$data_dir, "metadata.csv")
-PACKAGE_SLOC         <- path(params$data_dir, "sloc.csv")
+PACKAGE_METADATA     <- path(DATA_DIR, "metadata.csv")
+PACKAGE_SLOC         <- path(DATA_DIR, "sloc.csv")
 
 ## OUTPUTs
 
-## corpus only
-CORPUS_FILE                  <- path(params$output_dir, "corpus.txt")
-CORPUS_REVDEPS               <- path(params$output_dir, "corpus-revdeps.csv")
-CORPUS_REVDEPS_FILE          <- path(params$output_dir, "corpus-revdeps.txt")
-CORPUS_ALL_PACKAGES_FILE     <- path(params$output_dir, "corpus-all.txt")
+CORPUS_FILE                  <- path(DATA_DIR, "corpus.txt")
+CORPUS_REVDEPS               <- path(DATA_DIR, "corpus-revdeps.csv")
+CORPUS_REVDEPS_FILE          <- path(DATA_DIR, "corpus-revdeps.txt")
+CORPUS_ALL_PACKAGES_FILE     <- path(DATA_DIR, "corpus-all.txt")
 
-TAGS_DIR <- params$output_dir

### 2020-08-21 inst/run.sh

@@ -1,0 +1,54 @@
+#!/bin/bash
+
+DEF_PORT=8787
+DEF_CMD=/init
+
+DOCKER_IMAGE_NAME=prlprg/project-evalr
+CONTAINER_NAME=project-eval-rstudio
+
+function show_help() {
+    echo "Usage: $(basename $0) [-p NUM ] [CMD]"
+    echo
+    echo "where:"
+    echo
+    echo "  -p NUM      (optional) port for RStudio (defaults to $DEF_PORT)"
+    echo "  CMD         (optional) command to run (defaults to RStudio)"
+    echo
+}
+
+port=$DEF_PORT
+cmd=$DEF_CMD
+
+while getopts "hp:" opt; do
+    case "$opt" in
+    h)
+        show_help
+        exit 0
+        ;;
+    p)  port=$OPTARG
+        ;;
+    esac
+done
+
+shift $((OPTIND -1))
+
+[ $# -gt 0 ] && cmd="$@"
+
+docker run \
+  -ti \
+  --rm \
+  --name "$CONTAINER_NAME" \
+  -e ROOT=TRUE \
+  -e DISABLE_AUTH=true \
+  -e USERID=$(id -u) \
+  -e GROUPID=$(id -g) \
+  -e USER=rstudio \
+  -p "$port:8787" \
+  -v "$(pwd):/home/rstudio/evalR" \
+  "$DOCKER_IMAGE_NAME" \
+  bash -c "source /etc/cont-init.d/userconf && exec /usr/lib/rstudio-server/bin/rserver --auth-none 1 --auth-timeout-minutes 0 --www-port 8787 --server-daemonize 0"
+
+  # -v "$(pwd)/README.Rmd:/home/rstudio/README.Rmd" \
+  # -v "$(pwd)/README.html:/home/rstudio/README.html" \
+  # -v "$(pwd)/README.md:/home/rstudio/README.md" \
+  # -v "$(pwd)/.Rprofile:/home/rstudio/.Rprofile" \

### 2020-08-22 inst/paper/main.tex
intro

@@
 
+\newcommand{\eval}{\texttt{eval}\xspace}
+
 \renewcommand{\c}[1]{\lstinline{#1}\xspace}
 \newcommand{\miss}[1]{{\textcolor{red}{#1}}\xspace}
-\newcommand{\eval}{\c{eval}}
 
 \begin{abstract}
-  The \c{eval} function turns text into code. The widespread use of that
-  capability hinders the static reasoning about software systems. This paper
-  is a large-scale study of a corpus of \miss{3} million lines of R code.
-  We performed static and dynamic analysis of our corpus and report that
-  usage of \c{eval} is indeed wide spread. It is used by expert core
-  developers of the language, experienced library developers, and end-users
-  alike. The patterns of usage range from meta-programming to
-  misunderstandings of what \c{eval} is good for. Case studies give example
-  use-cases and some evidence that automated tools could be used to refactor
-  some of the uses of \c{eval}.
+  The \eval function turns text into code. From the point of view of
+  reasoning about software system, uses of \eval turn the program's code
+  from a statically known quantity that can be reasoned about and analyzed
+  into one of the inputs to the program. Thus, widespread use of \eval
+  hinders our ability to provide safety and security guarantees for
+  software. Understanding how \eval is used in practice is key to finding
+  ways to mitigate its damage. One question that has not been studied is
+  whether the usage of \eval is dependent on the application domain or other
+  features of the language. This paper is a large-scale study of a corpus of
+  \miss{3} million lines of data science libraries and end-user written in
+  the R language. We performed a simple static analysis and more
+  sophisticated dynamic analysis to confirm that \eval is indeed in wide
+  spread use, and to categorized patters of usage. We found that \eval in
+  this code base is more dangerous in some ways, and safer in others, than
+  what was previously reported for JavaScript.
 \end{abstract}

### 2020-08-22 inst/paper/main.tex
intro

@@
 \newcommand{\eval}{\texttt{eval}\xspace}
-
 \renewcommand{\c}[1]{\lstinline{#1}\xspace}
@@
 
-The \c{eval} function has a bad reputation. This function, under one name or
-another and with various exact semantics, is available in many dynamic
-languages. It is used to transform text into code and execute that code.  It
-earned its reputation because of the challenges it poses for reasoning about
-code that uses it.  The problem with eval is that the behavior of a system
-departs from the source text of the program and becomes dependent on the
-program's inputs. This brings challenge to program analysis and security.
+Traditional program analysis techniques are based on sound reasoning about
+program behavior, a program analyzer computes an over-approximation of the
+set of possible operations performed by the system being
+analyzed~\cite{cc77}.  This is predicated on knowing the program's code, and
+computing that code's behavior for all possible inputs. The presence of
+\eval entails that the program's input, any string variable, may be
+reflected into its code giving rise to \emph{any} behavior allowed by the
+language semantics. In many dynamic languages, \eval can redefine most
+user-defined and built-in functions resulting in a complete loss of
+precision for any operation following \eval. A group of influential
+resaerchers argued to give up on soundness and, instead, to
+under-approximate dynamic features~\cite{soundy} In their words ``a
+practical analysis, therefore, may pretend that \eval does nothing, unless
+it can precisely resolve its string argument at compile time.''  Assuming
+that \eval does not have side-effects may be too optimistic. We rather
+propose to study what \eval does in practice, in real-world code, so as to
+give researchers data about how to better under-approximate uses of that
+feature.
+
+While there has been previous work investigating the usage of \eval in web
+programming and JavaScript, there are no other studies that shed light on
+the usage of \eval in other programming languages and application domains.
+It is thus reasonable to wonder if the results from previous work can be
+carried over to other contexts. For this paper, we have picked the data
+science language R~\cite{r} as our object of study. R is both different from
+JavaScript in linguistic terms and in terms of its user base. JavaScript was
+designed to run untrusted code in browser, while R was design for
+statistical computing on a desktop. JavaScript is used as a general purpose
+language by a wide community of programmers; while R is used for scientific
+computing by data scientist and domain experts with, often, limited
+programming experience.
+
+This paper sets out to highlight the differences in usage patterns between
+these two languages and user communities. Our results are broadly applicable
+as they increase our understanding of the range of uses of \eval, present
+new patterns, and hopefully provide insights into reasonable
+under-approximations of \eval. More narrowly, for researchers interested in
+static analysis of data science code, we provide a publicly available corpus
+of programs and a toolchain that can be customized to deliver different
+types of information about the behavior of \eval.
 

### 2020-08-22 inst/paper/main.tex
intro

@@
 While there has been previous work investigating the usage of \eval in web
-programming and JavaScript, there are no other studies that shed light on
-the usage of \eval in other programming languages and application domains.
-It is thus reasonable to wonder if the results from previous work can be
-carried over to other contexts. For this paper, we have picked the data
-science language R~\cite{r} as our object of study. R is both different from
-JavaScript in linguistic terms and in terms of its user base. JavaScript was
-designed to run untrusted code in browser, while R was design for
-statistical computing on a desktop. JavaScript is used as a general purpose
-language by a wide community of programmers; while R is used for scientific
-computing by data scientist and domain experts with, often, limited
-programming experience.
+programming and JavaScript~\cite{ecoop11}, there are no other studies that
+shed light on the usage of \eval in other programming languages and
+application domains.  It is thus reasonable to wonder if the results from
+previous work can be carried over to other contexts. For this paper, we have
+picked the data science language R~\cite{r} as our object of study. R is
+both different from JavaScript in linguistic terms and in terms of its user
+base. JavaScript was designed to run untrusted code in browser, while R was
+design for statistical computing on a desktop. JavaScript is used as a
+general purpose language by a wide community of programmers; while R is used
+for scientific computing by data scientist and domain experts with, often,
+limited programming experience.
 
-This paper sets out to highlight the differences in usage patterns between
-these two languages and user communities. Our results are broadly applicable
-as they increase our understanding of the range of uses of \eval, present
-new patterns, and hopefully provide insights into reasonable
-under-approximations of \eval. More narrowly, for researchers interested in
-static analysis of data science code, we provide a publicly available corpus
-of programs and a toolchain that can be customized to deliver different
-types of information about the behavior of \eval.
+This paper sets out to highlight the differences in usage patterns of \eval
+between these two languages and user communities. Our results are broadly
+applicable as they increase our understanding of the range of uses of \eval,
+present new use-cases, and, hopefully, provide insights into reasonable
+under-approximations. More narrowly, for researchers interested in static
+analysis of data science code, we provide a publicly available corpus of
+programs and a toolchain that can be customized to deliver fine-grained
+information about the usage and impact of \eval.
 
-While the worst case is indeed pretty bad, previous work has found that
-in many cases eval can be fairly inoccuous or at least much less dangerous
-that the worst case.
+Our methodology for this work bows to the fact that static analysis of R is
+difficult, instead we focus on dynamic analysis. To observe \eval we have
+built a two-level monitoring infrastructure: we are able to monitor R
+programs by instrumentation -- this gives us access to many user-visible
+properties of R programs -- and we also can monitor the inner working of the
+R interpreter -- this allows us to capture details of execution that are not
+exposed at the source level.
 
-There has been research in understanding the behavior of eval in the context
-of web programming and in particular the JavaScript language.
+Our corpus has been constructed to reflect the different level of
+sophistication in the R community. We distinguish between the core
+implementers of the language, developers with both extensive programming
+experience and keen understanding of the language semantics, library
+implementers, developers with some programming experience and a working
+knowledge of R, and end-users, who are typically not expert programmers and
+often have only a cursory knowledge of the language.\footnote{To illustrate
+  the above, R is a lazy language, like Haskell, the authors have informally
+  surveyed end-users, some of them computer scientists, and did not find a
+  single one aware of that fact. Library developers, on the other hand, tend
+  to know about laziness and program defensively around it.} Thus, our
+corpus distinguished between ,....
 

### 2020-08-22 inst/paper/main.tex
intro

@@
 often have only a cursory knowledge of the language.\footnote{To illustrate
-  the above, R is a lazy language, like Haskell, the authors have informally
-  surveyed end-users, some of them computer scientists, and did not find a
-  single one aware of that fact. Library developers, on the other hand, tend
-  to know about laziness and program defensively around it.} Thus, our
-corpus distinguished between ,....
-
-This paper looks at eval in the context of data science. We look at the use
-of eval in the R programming language. And in particular we look at its use
-in three different contexts.  The first context is the implementation of the
-language, how is eval used by the core developers of R in the packages that
-are part of the R virtual machine.  The secoond context is how it is used in
-libraries. And the third context is how it is used by end users of the
-language.
+  the above consider that R uses lazy evaluation like Haskell. The authors
+  have informally surveyed end-users, some of them computer scientists, and
+  did not find a single one aware of that fact. Library developers, on the
+  other hand, tend to know about laziness and program defensively around
+  it.}  Thus, our corpus distinguished between \emph{core packages} (there
+are \miss{8} such packages that are loaded together with the R virtual
+machine), \emph{CRAN packages} (these are curated packages that pass
+stringent community quality checks and are equipped with tests and sample
+data), \emph{scripts} (end-user written code that performs a particular
+machine learning or data analysis task).
 

### 2020-08-22 inst/paper/main.tex
intro

@@
 are \miss{8} such packages that are loaded together with the R virtual
-machine), \emph{CRAN packages} (these are curated packages that pass
-stringent community quality checks and are equipped with tests and sample
-data), \emph{scripts} (end-user written code that performs a particular
-machine learning or data analysis task).
-
-There are reasons to believe that usage patterns may be different. The core
-R code is extremely stable, it has only been maintained for the last 20
+machine), \emph{CRAN packages} (we selected \miss{500} curated packages that
+pass stringent community quality checks and are equipped with tests and
+sample data), \emph{scripts} (we used \miss{400} end-user written programs
+that performs a particular data analysis task).  There are reasons to
+believe that \eval usage patterns may differ between these datasets. The
+core R code is extremely stable, it has only been maintained for the last 20
 years with very few far reaching changes. The libraries represent a more
-lively ecosystem with new libraries added each day. Typically they are
-developped by experience R developer.  Finally, end user code is often
-written by non-programmers, data scientists who may or may not understand R.
-
-To set up our experiements we have created a corpus composed of three
-distinct data sets.  We have set out the 8 packages that are included with
-the R virtual machine and refer to these as `` base'' packages.  We have
-selected the 400 most widely included libraries and obtained a number of
-client programs to exercises them.  Finally we have obtained 400 end users
-programs from the Kaggle competition.
+lively ecosystem with new libraries added each day.  Finally, end user code
+is often one-use scripts thrown together and sometimes never revisited.
 

### 2020-08-22 inst/docker-image/get-installed-packages.R
Moved to custom install instead whole CRAN install

@@ -1,0 +1,36 @@
+#!/usr/bin/env Rscript
+options(repos=Sys.getenv("CRAN_LOCAL_MIRROR", "https://cloud.r-project.org"))
+
+# pins the current version of the corpus packages as well as
+# their dependencies
+
+BASE_PACKAGES <- c(
+  "base",
+  "compiler",
+  "datasets",
+  "graphics",
+  "grDevices",
+  "grid",
+  "methods",
+  "parallel",
+  "splines",
+  "stats",
+  "stats4",
+  "tcltk",
+  "tools",
+  "utils"
+)
+
+packages_corpus <- readLines("../data/corpus.txt")
+packages_kaggle <- readLines("kaggle-dependencies.txt")
+packages_base <- readLines("our-packages-dependencies.txt")
+packages <- unique(c(packages_corpus, packages_kaggle, packages_base))
+
+dependencies <- unlist(tools::package_dependencies(packages, recursive=T))
+packages <- unique(c(packages, dependencies))
+packages <- sort(setdiff(packages, BASE_PACKAGES))
+
+versions <- sapply(packages, function(x) packageDescription(x)$Version)
+
+df <- data.frame(packages, versions)
+write.table(df, "CRAN-packages.txt", row.names=FALSE, col.names=FALSE, quote=FALSE)

### 2020-08-22 inst/analysis/corpus-creation.Rmd
Updated corpus blacklisting ggm and shinyBS

@@
 params:
-  data_dir: /var/lib/R/project-evalR/revalstudy/data
+  base_dir: /home/rstudio/evalR
+  hostname: prl2
+  port: 8788
   num_packages: 500
-  output_dir: /var/lib/R/project-evalR/revalstudy/data
   which_dependencies:
@@
 
+```{r package blacklist}
+blacklist <- tribble(
+  ~package,
+  "ggm",      # requires bioc
+  "shinyBS"   # runs shiny server as part of its tests
+)
+```
+
 ```{r get all available packages}
 # TODO this should be installed.packages() once they are installed
-cran_packages_raw <- available.packages()
+#cran_packages_raw <- available.packages()
+cran_packages_raw <- installed.packages()
 cran_packages <- as_tibble(cran_packages_raw) %>%
-  select(package=Package, version=Version)
+  select(package=Package, version=Version) %>%
+  anti_join(blacklist, by="package")
 ```
@@
 
-write_lines(pkgs_top_revdeps$package, CORPUS_FILE)
+write_lines(sort(pkgs_top_revdeps$package), CORPUS_FILE)
 write_csv(pkgs_revdeps, CORPUS_REVDEPS)
-write_lines(unique(pkgs_revdeps_raw$revdep), CORPUS_REVDEPS_FILE)
-write_lines(unique(), CORPUS_ALL_PACKAGES_FILE)
+write_lines(sort(unique(pkgs_revdeps_raw$revdep)), CORPUS_REVDEPS_FILE)
 ```

### 2020-08-22 inst/analysis/enrich_dataset.Rmd
Splitting and summarizing the dataset correctly

--- 2020-08-20 inst/analysis/enrich_dataset.Rmd
@@
 params:
-  data_dirpath: "/var/lib/R/project-evalR/run/package-evals-traced-corpus.4/"
-  data_filepath: "/var/lib/R/project-evalR/run/package-evals-traced-corpus.4/calls.fst"
+  data_dirpath: "/var/lib/R/project-evalR/run/package-evals-traced-corpus.5/"
+  data_filepath: "/var/lib/R/project-evalR/run/package-evals-traced-corpus.5/calls.fst"
+  data_kagglepath: "/var/lib/R/project-evalR/kaggle-korpus/run/titanic/calls.fst"
 knit: (function(inputFile, encoding) {
@@
 
-source("inc/latextags.R")
-create_tags(path("/var/lib/R/project-evalR/revalstudy/data", "analysisStable.tex"), prefix="", default=TRUE)
+source("preprocess.R", local = knitr::knit_global())
 ```
@@
 
-```{r eval_in_base}
-eval_base_functions <- c("autoload", "autoloader", "bquote", "by.default", "by.data.frame", "invokeRestartInteractively", "Ops.data.frame", "dget", 
-                         "eval", "eval.parent", "evalq", "local", "with.default", "within.data.frame", "within.list", "replicate", "subset.data.frame",
-                         "subset.matrix", "transform.data.frame", "match.arg", "char.expand", "max.col", "parseNamespaceFile", "source", "sys.source",
-                         "stopfinot", "as.data.frane.table", "match.fun", "trace", "untrace", ".doTrace")
-```
-
 ```{r load-data, echo=FALSE}
@@
     read_fst(params$data_filepath) %>%
-    as_tibble()
+    as_tibble() %>%
+    mutate(corpus = "cran")
 
-eval_calls <- eval_calls_raw %>% filter(!caller_function %in% eval_base_functions)
+eval_calls_kaggle_raw <- 
+  read_fst(params$data_kagglepath) %>%
+  as_tibble() %>% 
+  mutate(corpus = "kaggle")
+
+eval_calls_raw <- bind_rows(eval_calls_raw, eval_calls_kaggle_raw)
 ```
 
-There are so many calls that are similar:
-```{r}
-eval_calls_base <- eval_calls_raw %>% filter(caller_function %in% eval_base_functions) 
+```{r deduplicate}
+eval_calls <- eval_calls_raw %>% deduplicate()
 ```
 
-Let's remove the duplicate ones (but keep their numbers):
+```{r add_types}
+eval_calls <- eval_calls %>% add_types()
+```
 
-```{r}
-eval_calls_base <- eval_calls_base %>% count(across(c(-eval_call_id, -starts_with("caller_stack"))), name = "nb_eval_calls") 
-eval_calls <- eval_calls %>% count(across(c(-eval_call_id)), name = "nb_eval_calls")
+```{r add_source}
+eval_calls <- eval_calls %>% add_eval_source() %>% add_eval_source_type()
+```
+
+```{r parse_args}
+eval_calls <- eval_calls %>% add_parse_args()
+```
+
+```{r separate_datasets}
+eval_calls_core <- eval_calls %>% filter(eval_source_type == "core")
+eval_calls_packages <- eval_calls %>% filter(eval_source_type %in% c("package", "<undefined>") )
+eval_calls_kaggle <- eval_calls %>% filter(eval_source_type == "kaggle")
 ```
@@
 ```{r}
-eval_calls_base <- eval_calls_base %>% 
-    resolve_sexp_name(expr_expression_type) %>%
-    resolve_sexp_name(expr_resolved_type) %>%
-    resolve_sexp_name(enclos_type) %>%
-    resolve_sexp_name(envir_type)
-
-eval_calls <- eval_calls %>% 
-    mutate(eval_call_package = extract_package_name(eval_call_srcref, file)) %>% 
-    resolve_sexp_name(expr_expression_type) %>%
-    resolve_sexp_name(expr_resolved_type) %>%
-    resolve_sexp_name(enclos_type) %>%
-    resolve_sexp_name(envir_type)
-```
-
-```{r}
-#eval_calls <- eval_calls %>% mutate(parse_args = if_else(is.na(expr_parsed_expression), NA, map_chr(expr_parsed_expression, extract_args_parse)))
+eval_calls_raw <- eval_calls_raw %>% select(-ends_with("_type")) %>% left_join(eval_calls)
 ```
@@
 ```{r write_back}
-write_fst(eval_calls, paste0(params$data_dirpath, "deduplicated_calls.fst"))
+write_fst(eval_calls, paste0(params$data_dirpath, "raws.fst"))
 
-write_fst(eval_calls_base, paste0(params$data_dirpath, "base_calls.fst"))
+write_fst(eval_calls_core, paste0(params$data_dirpath, "summarized-core.fst"))
+
+write_fst(eval_calls_packages, paste0(params$data_dirpath, "summarized-packages.fst"))
+
+write_fst(eval_calls_kaggle, paste0(params$data_dirpath, "summarized-kaggle.fst"))
 ```

### 2020-08-22 inst/analysis/insights.R

@@ -248,6 +248,7 @@ extract_package_name <- function(src_ref, file)
   # - /R/* : core packages (we cannot distinguish between them yet so we write core for the package name)
   # - /testit/... or /testthat/... : it is the testit or testthat packages
   # - :/R : extract the package name from the file path in the column path
+  # - .../kaggle-run/<id>/run.R:...
   case_when(
     is.na(src_ref) ~ "base?",
     str_starts(src_ref, fixed("./R/")) ~ "core",

### 2020-08-22 inst/analysis/preprocess.R

@@ -1,0 +1,65 @@
+# Preprocess the dataset:
+# - remove duplicates
+# - add some columns (such as `eval_call_package`)
+# - cut the dataset into smaller ones
+
+# Datasets
+# Input: raw.fst
+# Outputs:
+#   - summarized-core.fst
+#   - summarized-packages.fst
+
+source("insights.R")
+
+deduplicate <- function(dataset)
+{
+  return(dataset %>% count(across(c(-eval_call_id, -starts_with("caller_stack"))), name = "nb_ev_calls"))
+}
+
+add_types <- function(dataset)
+{
+  return(dataset %>%
+           resolve_sexp_name(expr_expression_type) %>%
+           resolve_sexp_name(expr_resolved_type) %>%
+           resolve_sexp_name(enclos_type) %>%
+           resolve_sexp_name(envir_type))
+}
+
+add_parse_args <- function(dataset)
+{
+  return(dataset %>% mutate(parse_args = map(expr_parsed_expression, function(e) 
+    { if(!is.na(e) && str_starts(e, "(parse|str2lang|str2expression)\\(")) 
+        extract_args_parse(e) 
+      else list() }))  %>% unnest_wider(parse_args))
+}
+
+eval_base_functions <- c("autoload", "autoloader", "bquote", "by.default", "by.data.frame", "invokeRestartInteractively", "Ops.data.frame", "dget", 
+                         "eval", "eval.parent", "evalq", "local", "with.default", "within.data.frame", "within.list", "replicate", "subset.data.frame",
+                         "subset.matrix", "transform.data.frame", "match.arg", "char.expand", "max.col", "parseNamespaceFile", "source", "sys.source",
+                         "stopfinot", "as.data.frane.table", "match.fun", "trace", "untrace", ".doTrace")
+
+# TODO: add support for package name extraction from caller_stack_expression for when 
+# eval_call_srcref is NA
+find_package_name <- function(caller_function, srcref, file)
+{
+  if(caller_function %in% eval_base_functions)
+    return("base")
+  else
+    return(extract_package_name(srcref, file))
+}
+
+add_eval_source <- function(dataset)
+{
+  return(dataset %>% 
+           mutate(eval_source = pmap_chr(list(caller_function, eval_call_srcref, file),
+                                                         find_package_name)) )
+}
+
+add_eval_source_type <- function(dataset)
+{
+  return(dataset %>% 
+           mutate(eval_source_type = case_when(eval_source %in% c("base", "core") ~ "core",
[…]

### 2020-08-22 inst/analysis/analysis_stable.Rmd
Degree of polymorphism

@@
 params:
-    data_filepath: "/var/lib/R/project-evalR/run/package-evals-traced-corpus.5/deduplicated_more_calls.fst"
+    data_filepath: "/var/lib/R/project-evalR/run/package-evals-traced-corpus.5/summarized-packages.fst"
     #data_filepath: "/var/lib/R/project-evalR/run/package-evals-traced-all.1/calls.fst"
@@
 
+Another interesting figure is the degreee of polymorphism of the `expr` argument: how many different types (for `expr_resolved`) are there for one call sites?
+
+```{r polymorphism_callsite}
+expr_resolved_polymorphism <- eval_calls %>% group_by(eval_call_srcref) %>%
+        summarise(polym_degree = vec_unique_count(expr_resolved_type))
+```
+
+```{r}
+expr_resolved_polymorphism %>% ggplot() +
+  geom_bar(aes(x = polym_degree)) +
+  scale_x_continuous(breaks=c(0, 1, 5, 10)) +
+  labs(title = "Degree of polymorphism", x = "degree", y = "n") 
+```
+
 ## Provenance

### 2020-08-22 inst/paper/main.tex
Degree of polymorphism

@@
 \usepackage{xspace}
+
+\graphicspath{{img/}}
 
@@
 
+We look at how \emph{polymorphic} the \c{expr} argument of \eval can be, \emph{i.e.} how many different types of the resolved \c{expr} there are per call sites, in Figure~\ref{fig:polymorphism}.
+89\% of the call sites are \emph{monomorphic}.
+
+\begin{figure}
+    \includegraphics[width=\columnwidth]{polymorphism-1}
+    \caption{Degree of polymorphism of \c{expr} per call sites.} \label{fig:polymorphism}
+\end{figure}
+
 \subsection{How dangerous is eval}

### 2020-08-23 inst/paper/main.tex
minor

@@
  language=R,              % the language of the code
-	basicstyle=\ttfamily, % the size of the fonts that are used for the code
-	numbers=left,                   % where to put the line-numbers
-	numberstyle=\color{Blue},  % the style that is used for the line-numbers
-	stepnumber=1,                   % the step between two line-numbers. If it is 1, each line
-	% will be numbered
+ basicstyle=\small\ttfamily,    % the size of the fonts
+% numbers=left,            % where to put the line-numbers
+ numberstyle=\color{Blue},% the style that is used for the line-numbers
+ stepnumber=1,            % the step between two line-numbers. If it is 1,
+                          % each line will be numbered
  numbersep=2pt,           % how far the line-numbers are from the code
-	backgroundcolor=\color{white},  % choose the background color. You must add \usepackage{color}
+ backgroundcolor=\color{white},  % choose the background color.
  showspaces=false,        % show spaces adding particular underscores
  showstringspaces=false,  % underline spaces within strings
-	showtabs=false,                 % show tabs within strings adding particular underscores
+ showtabs=false,          % show tabs within strings adding particular 
 %frame=single,            % adds a frame around the code
-	rulecolor=\color{black},        % if not set, the frame-color may be changed on line-breaks within not-black text (e.g. commens (green here))
+ rulecolor=\color{black}, % if not set, the frame-color may be changed on
+                          % line-breaks within not-black text (e.g. commens
+                          % (green here))
  tabsize=2,               % sets default tabsize to 2 spaces
@@
  breaklines=true,         % sets automatic line breaking
-	breakatwhitespace=false,        % sets if automatic breaks should only happen at whitespace
-	keywordstyle=\color{RoyalBlue},      % keyword style
+ breakatwhitespace=false, % sets if automatic breaks should only happen at 
+ keywordstyle=\color{Blue},    % keyword style
  commentstyle=\color{YellowGreen},  % comment style
@@
   the above consider that R uses lazy evaluation like Haskell. The authors
-  have informally surveyed end-users, some of them computer scientists, and
-  did not find a single one aware of that fact. Library developers, on the
-  other hand, tend to know about laziness and program defensively around
-  it.}  Thus, our corpus distinguished between \emph{core packages} (there
-are \miss{8} such packages that are loaded together with the R virtual
-machine), \emph{CRAN packages} (we selected \miss{500} curated packages that
-pass stringent community quality checks and are equipped with tests and
-sample data), \emph{scripts} (we used \miss{400} end-user written programs
-that performs a particular data analysis task).  There are reasons to
-believe that \eval usage patterns may differ between these datasets. The
-core R code is extremely stable, it has only been maintained for the last 20
-years with very few far reaching changes. The libraries represent a more
-lively ecosystem with new libraries added each day.  Finally, end user code
-is often one-use scripts thrown together and sometimes never revisited.
+  have informally surveyed end-users, including computer scientists, and did
+  not find a single one aware of this fact. Library developers, on the other
+  hand, know about laziness and program defensively around it.}  Thus, our
+corpus distinguished between \emph{core packages} (there are \miss{8} such
+packages that are loaded together with the R virtual machine), \emph{CRAN
+  packages} (we selected \miss{500} curated packages that pass stringent
+community quality checks and are equipped with tests and sample data),
+\emph{scripts} (we used \miss{400} end-user written programs that performs a
+particular data analysis task).  There are reasons to believe that \eval
+usage patterns may differ between these datasets. The core R code is
+extremely stable, it has only been maintained for the last 20 years with
+very few far reaching changes. The libraries represent a more lively
+ecosystem with new libraries added each day.  Finally, end user code is
+often one-use scripts thrown together and sometimes never revisited.
 

### 2020-08-23 inst/paper/main.tex
related

@@
 \title{A Large-Scale Study of the Use of Eval in R}
-\author{\vspace{-.8cm}\IEEEauthorblockN{
-    \it\Large The Eval that Data Scientists Do}%
-  %\IEEEauthorblockA{ dept} \\ \and
-}  
+\author{\vspace{-.8cm}\IEEEauthorblockN{~}} %\IEEEauthorblockA{ dept} \\ \and
 \maketitle
@@
 
-\section{Related Work}
+\section{Background and Related Work}
 
-\cite{ecoop11}
+The origins of \eval go back to the first implementation of the Lisp
+programming language~\cite{lisp}. Over the years, \eval has been included in
+many dynamic languages, with some variability in expressive power and
+nomenclature. In all of its incarnation \eval allow developpers to
+dynamically extend the code of their program at run-time.  Originally, \eval
+was limited to interpreted languages, but with the advent of just-in-time
+compilation this restriction was gradually lifted.
+
+Some languages have chosen to restrict \eval. The essential difference
+between languages relates to scoping of variables that can be accessed from
+code executed by \eval.  Java and Julia are among the more restrictive
+languages. In Java, \eval must be synthesized as a combination dynamic
+loading and reflection.\footnote{While Java does not have \eval built in,
+  one can take the input to \eval, wrap it in a static method of an
+  anonymous class, generate bytecode for that class, invoke the Java class
+  loader to dynamically install it, and then use reflection to call the new
+  method.} Julia provides \eval more directly but limits its scope to top
+level variables and methods. JavaScript support both a, so-called, global
+\eval, and a local \eval. Where the latter has acces to local variables of
+the function being executed (as well as variables of lexically enclosing
+scopes). Finally languages such as R allow \eval to manipulate variables
+defined in the caller environment of the function in which \eval is invoked.
+
+The difference between these design choices has an impact on the ability of
+compilers to optimize, and of analysis tools to reason about, code. The
+impact of a top level \eval is that any global definition may be affected
+but the current function can continue to execute unchanged. A local \eval
+may change the value of local variables, their types, or even their
+presence. This is a much more invasive change for reasoning about code and
+may invalidate common optimizations such as constant folding or copy
+propagation. The designers of Julia were aware of this difference and chose
+to restrict \eval to preserve their ability to compile
+code~\cite{oopsla18a}. They even added a versioning mechanism to function
+definitions to ensure that new code added during an \eval would not
+invalidate optimizations such as inlining.  R's ability to affect any one of
+the callers of a function is even more invasive as the impact of \eval is
+non-local. Such unfettered power is difficult to deal with.
 
@@
 
-\section{Shades of Eval}
-
-- Talk about environments
-- How is it different from JS (takes language, env in R but string in JS).
-
 \section{The Design of Eval}
 
-There are four different kinds of \eval in the \emph{base} package:  \c{eval}, \c{evalq}, \c{eval.parent},  \c{local}.
-The most important one is \eval:
+There are four different kinds of \eval in the \emph{base} package:
+\c{eval}, \c{evalq}, \c{eval.parent}, \c{local}.  The most important one is
+\eval:
 
 \begin{lstlisting}
-eval <- function (expr, envir = parent.frame(), enclos = if (is.list(envir) || is.pairlist(envir)) parent.frame() else baseenv())  
+  eval <- function(expr, envir=parent.frame(),
+                   enclos=if (is.list(envir)||is.pairlist(envir)) parent.frame() else baseenv())  
 	.Internal(eval(expr, envir, enclos))

### 2020-08-23 inst/paper/main.tex
rw

@@
 
+The intuition behind the soundiness manifesto~\cite{soundy} is that while
+the use of a single \eval may invalidate every single invariant computed by
+a static analysis or optimizer, this worst case usually does not happen.  In
+many cases, \eval is used carefully and has limited effect. This is a
+hypothesis that must be validated and, to be useful, we should have the
+means to provide some reasonable guesses as to which side-effects will
+result from any particular \eval call.
+
 Previous work looked at eval in JavaScript

### 2020-08-23 inst/analysis/corpus-analysis.Rmd
Prototype how could it be in docker

@@
 params:
+  base_dir: /home/rstudio/evalR
+  hostname: prl2
+  port: 8788
 knit: (function(inputFile, encoding) {
@@
 ---
+
+```{r setup, include=FALSE}
+for (x in c(
+  "dplyr", 
+  "DT", 
+  "fs",
+  "lubridate",
+  "purrr",
+  "readr",
+  "rjson",
+  "stringr",
+  "tibble"
+  )) {
+  suppressPackageStartupMessages(library(x, character.only=TRUE))
+}
+
+knitr::opts_chunk$set(echo = TRUE)
+
+source("inc/paths.R")
+source("inc/setup.R")
+source("inc/latextags.R")
+
+create_tags(path(TAGS_DIR, "corpus.tex"), prefix="", default=TRUE)
+
+# TODO
+PACKAGE_COVERAGE_LOG <- "/home/rstudio/evalR/run/package-coverage/parallel.csv"
+```
+
+## Description
+
+The aim of this notebook is to analyze the corpus of R packages.
+
+## Loading data
+
+This notebook requires the following data files:
+
+- `r PACKAGE_COVERAGE_LOG` - package coverage log
+- `r PACKAGE_COVERAGE` - package coverage
+
+```{r load data - packages, include=FALSE}
+coverage_log <- read_runr_log(PACKAGE_COVERAGE_LOG)
+```
+
+```{r}
+failed_coverage <- filter(coverage_log, exitval != 0)
+```
+
+```{r}
+failed_coverage %>%
+  mutate(
+    run_path=show_url(run_path)
+  ) %>%
+  my_datatable(escape=FALSE)
+```
+
+```{r load data - revdep, include=FALSE}
+pkgs_revdeps_raw <- read_csv(
+  CRAN_REVDEPS, 
+  col_types=cols(
+    package = col_character(),
+    revdep = col_character()
+  )
+)
+```
+
+## Thresholds
+
+```{r thresholds}
+overview_table(
+  r("threshold code coverage", percent(params$min_coverage)),
+  r("threshold revdeps", params$min_revdeps),
+  r("package size outier", params$package_size_outlier)
+)
+```
+
+## CRAN Overview
+
+### Table
+
+```{r summary}
+summary_corpus <- function(df, prefix) {
+  w_coverage <- filter(df, coverage_expr > 0)
+  w_revdeps <- filter(df, revdeps > 0)
+  w_runnable <- filter(df, !is.na(runnable_time))
+
+  overview_table(
+    r(paste(prefix, "cran"), df),
+    r(paste(prefix, "installed"), filter(df, installed)),
+    r(paste(prefix, "loadable"), filter(df, loadable)),
+
+    r(paste(prefix, "with runnable code"), filter(df, runnable_code > 0)),
+    r(paste(prefix, "runnable code"), sum(df$runnable_code, na.rm=T)),
+    r(paste(prefix, "examples code"), sum(df$runnable_code_examples, na.rm=T)),
+    r(paste(prefix, "tests code"), sum(df$runnable_code_tests, na.rm=T)),
+    r(paste(prefix, "vignettes code"), sum(df$runnable_code_vignettes , na.rm=T)),
+
+    r(paste(prefix, "with coverage"), w_coverage),
+    r(paste(prefix, "mean expr coverage"), percent(mean(w_coverage$coverage_expr))),
+    r(paste(prefix, "median expr coverage"), percent(median(w_coverage$coverage_expr))),
+    r(paste(prefix, "covered functions ratio"), ratio(sum(w_coverage$funs_with_coverage),sum(w_coverage$funs))),
+
+    r(paste(prefix, "with R code"), filter(df, package_r_code > 0)),
+    r(paste(prefix, "R code"), sum(df$package_r_code, na.rm=T)),
+    r(paste(prefix, "with native code"), filter(df, package_native_code > 0)),
+    r(paste(prefix, "native code"), sum(df$package_native_code, na.rm=T)),
+
+    r(paste(prefix, "functions"), sum(df$funs, na.rm=T)),
+    r(paste(prefix, "public functions"), sum(df$funs_public, na.rm=T)),
+    r(paste(prefix, "private functions"), sum(df$funs_private, na.rm=T)),
+    r(paste(prefix, "S three functions"), sum(df$funs_s3, na.rm=T)),
+    r(paste(prefix, "S three classes"), sum(df$s3_classes, na.rm=T)),
+
+    r(paste(prefix, "w revdes"), w_revdeps),
+    r(paste(prefix, "revdes"), sum(w_revdeps$revdeps)),
+    r(paste(prefix, "mean revdes"), mean(w_revdeps$revdeps)),
+    r(paste(prefix, "median revdes"), median(w_revdeps$revdeps)),
+    r(paste(prefix, "at least min revdep"), filter(w_revdeps, revdeps >= params$min_revdeps)),
+    r(paste(prefix, "revdep runnable code"), sum(w_revdeps$revdeps_runnable_code, na.rm=T)),
+    r(paste(prefix, "revdep mean code coverage"), percent(mean(w_revdeps$revdep_coverage_expr)/100)),
+    r(paste(prefix, "revdep median code coverage"), percent(median(w_revdeps$revdep_coverage_expr)/100)),
+
+    r(paste(prefix, "runtime"), as.duration(sum(w_runnable$runnable_time))),
+    r(paste(prefix, "mean runtime"), as.duration(mean(w_runnable$runnable_time))),
+    r(paste(prefix, "median runtime"), as.duration(median(w_runnable$runnable_time))),
+
+    r(paste(prefix, "using roxygen"), sum(df$uses_roxygen, na.rm=T)),
+  )
+}
+
+summary_corpus(pkgs, "All")
+```
+
+### Plots
+
+```{r all considered packages}
+pkgs_considered <- filter(pkgs, loadable, package_r_code > 0, coverage_expr > 0)
+```
+
+```{r plot all considered packages}
+plot_all <- function(df) {
+  coverage_mean <- mean(df$coverage_expr)
+  coverage_median <- median(df$coverage_expr)
+  revdeps_mean <- mean(df$revdeps)
+  revdeps_median <- median(df$revdeps)
+
+  df %>%
+    mutate(
+      outlier=package_code > 5e4,
+      label=if_else(outlier, str_glue("{package}"), as.character(NA))
+    ) %>%
+    ggplot(
+      aes(
+        x=coverage_expr, 
+        y=revdeps,
+        label=label,
+        size=package_code
+      )
+    ) +
+    geom_point(alpha=.6) +
+    geom_text(size=3, check_overlap = T, vjust=1.5, na.rm = TRUE) +
+
+    geom_hline(aes(yintercept=revdeps_mean), linetype="dashed", color="black", size=0.2) +
+    geom_vline(aes(xintercept=coverage_mean), linetype="dashed", color="black", size=0.2) +
+    geom_vline(aes(xintercept=params$min_coverage),linetype="solid", color="red", size=0.4) +
+    geom_hline(aes(yintercept=params$min_revdeps),linetype="solid", color="red", size=0.4) +
+
+    scale_x_continuous(labels = scales::percent) + 
+    scale_y_log10(labels = scales::comma) +
+    scale_size(range = c(.1, 5), labels = scales::comma) +
+      theme(
+      legend.position=c(0.2, 0.85),
+      legend.box="horizontal",
+      legend.box.background = element_rect(fill="white", size=0.1)
+    ) +
+    labs(
+      x="Code coverage", 
+      y="Number of reverse dependencies (log)", 
+      size="Code size",
+      linetype=""
+    ) +
+    guides(
+      size=guide_legend(order=1),
+      color=guide_legend(order=2),
+      linetype=F
+    )
+}
+
+plot_all(pkgs_considered)
+```
+
+## Corpus overview
+
+We only consider packages that have at least `r params$min_revdeps` reverse dependencies and `r params$min_coverage*100`% expression coverage.
+
+```{r project corpus}
+corpus <- 
+  filter(
+    pkgs, 
+    loadable, 
+    runnable_code > 0, 
+    revdeps >= params$min_revdeps, 
+    package_r_code > 0, 
+    coverage_expr >= params$min_coverage
+  )
+
+write_lines(corpus$package, CORPUS_FILE)
+write_csv(corpus, CORPUS_DETAILS)
+```
+
+### Reverse dependencies
+
+```{r revdeps}
+corpus_revdeps <- pkgs_revdeps_raw %>% semi_join(corpus, by="package")
+write_csv(pkgs_revdeps_raw, CORPUS_REVDEPS)
+write_lines(unique(corpus_revdeps$revdep), CORPUS_REVDEPS_UNIQUE)
+```
+
+### Table
+
+```{r corpus summary}
+summary_corpus(corpus, "corpus")
+
+write_csv(corpus, CORPUS_DETAILS)
+write_lines(corpus$package, CORPUS_FILE)
+```
+
+### Plot
+
+```{r plot corpus}
+plot_corpus <- function(df, code_outlier) {
+  coverage_mean <- mean(df$coverage_expr)
+  coverage_median <- median(df$coverage_expr)
+  revdeps_mean <- mean(df$revdeps)
+  revdeps_median <- median(df$revdeps)
+
+  df %>%
+    mutate(
+      outlier=package_code > code_outlier,
+      label=if_else(outlier, str_glue("{package}"), as.character(NA))
+    ) %>%
+    ggplot(
+      aes(
+        x=coverage_expr, 
+        y=revdeps,
+        label=label,
+        size=package_code
+      )
+    ) +
+    geom_point(alpha=.6) +
+    geom_text(size=3, check_overlap = T, vjust=1.5, na.rm = TRUE) +
+
+    geom_hline(aes(yintercept=revdeps_mean), linetype="dashed", color="red", size=0.2) +
+    geom_vline(aes(xintercept=coverage_mean), linetype="dashed", color="red", size=0.2) +
+
+    scale_x_continuous(limits = c(params$min_coverage, 1), labels = scales::percent) + 
+    scale_y_log10(labels = scales::comma) +
+    scale_size(range = c(.1, 3), labels = scales::comma) +
+      theme(
+      legend.position=c(0.2, 0.85),
+      legend.box="horizontal",
+      legend.box.background = element_rect(fill="white", size=0.1)
+    ) +
+    labs(
+      x="Code coverage", 
+      y="Number of reverse dependencies (log)", 
+      size="Code size",
+      linetype=""
+    ) +
+    guides(
+      size=guide_legend(order=1),
+      color=guide_legend(order=2),
+      linetype=F
+    )
+}
+
+plot_corpus(corpus, params$package_size_outlier)
+ggsave(CORPUS_PLOT)
+```
+
+### List of corpus packages
+
+The following table is a detailed view of all the packages included in the corpus.
+
+```{r list of packages}
+datatable(
+  corpus, 
+  filter="top", 
+  options=list(
+    pageLength=50,
+    autoWidth=TRUE
+  )
+)
+```

### 2020-08-23 inst/analysis/corpus-creation.Rmd
Prototype how could it be in docker

@@
 
+```{r consts}
+BASE_PACKAGES <- tribble(
+  ~package,
+  "base",
+  "compiler",
+  "datasets",
+  "graphics",
+  "grDevices",
+  "grid",
+  "methods",
+  "parallel",
+  "splines",
+  "stats",
+  "stats4",
+  "tcltk",
+  "tools",
+  "utils"
+)
+```
+
 ## Description
@@
   "ggm",       # requires bioc
-  "shinyBS"   # runs shiny server as part of its tests
+  "future",    # does not work with OMP thread setting
+  "shinyBS",   # runs shiny server as part of its tests
+  "spatstat",  # takes too long
+  "survival"   # takes too long
 )
@@
   select(package=Package, version=Version) %>%
-  anti_join(blacklist, by="package")
+  anti_join(blacklist, by="package") %>%
+  anti_join(BASE_PACKAGES, by="package")
 ```
@@
 
-```{r get last month most downloaded packages}
+```{r get last month most downloaded packages, eval=F}
 pkgs_top_downloaded <-
@@
 
-```{r compute missing packages from the most downloaded ones}
+```{r compute missing packages from the most downloaded ones, eval=F}
 missing <- setdiff(pkgs_top_downloaded$package, pkgs_top_revdeps$package)

### 2020-08-23 inst/analysis/kaggle.Rmd
Prototype how could it be in docker

---
title: "Kaggle"
output: html_document
params:
  base_dir: /home/rstudio/evalR
  hostname: prl2
  port: 8788
  competitions:
    - titanic
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
for (x in c(
  "dplyr", 
  "DT", 
  "fs",
  "fst",
  "purrr",
  "readr",
  "rjson",
  "runr",
  "stringr",
  "tibble"
  )) {
  suppressPackageStartupMessages(library(x, character.only=TRUE))
}

knitr::opts_chunk$set(echo = TRUE)

# TODO: move into the package
source("inc/paths.R")
source("inc/setup.R")
source("inc/latextags.R")

create_tags(path(TAGS_DIR, "kaggle.tex"), prefix="Kaggle", default=TRUE)

KAGGLE_KERNELS_DIR  <- path(params$base_dir, "run", "kaggle-kernels", "notebooks", "r", "kernels")
KAGGLE_DATASETS_DIR <- path(params$base_dir, "run", "kaggle-datasets")
KAGGLE_RUN_DIR      <- path(params$base_dir, "run", "kaggle-run")

KERNELS_FILE      <- path(KAGGLE_RUN_DIR, "kernels.csv")
SCRIPTS_FILE      <- path(KAGGLE_RUN_DIR, "scripts.txt")
PARALLEL_LOG_FILE <- path(KAGGLE_RUN_DIR, "parallel.csv")
```

```{r check}
stopifnot(dir_exists(KAGGLE_KERNELS_DIR))
stopifnot(dir_exists(path(KAGGLE_DATASETS_DIR, params$competitions)))
```

```{r aux functions}
wrap <- function(file, body) {
  str_glue(
    "evil::write_eval_traces(",
    "  evil::trace_eval(",
    "    quote = TRUE,",
    "    code = {{",
    "      {gsub('\n', '\n      ', body)}",
    "    }}",
    "  ),",
    "  datadir=getwd()",
    ")",
    .sep = "\n"
  )
}

process <- function(x) {
  run_path <- path(KAGGLE_RUN_DIR, x$competition, basename(x$path))
  code_file <- path(x$path, "script", x$code_file)
  run_code_file <- path(run_path, "run.R")

  dir_create(run_path)

  if (str_ends(code_file, "\\.[rR]$")) {
    file_copy(code_file, run_code_file, overwrite=TRUE)
  } else if (str_ends(code_file, "\\.Rmd")) {
    knitr::purl(code_file, run_code_file, quiet=TRUE)
  } else if (str_ends(code_file, "\\.irnb") || str_ends(code_file, "\\.ipynb")) {
    tmp <- tempfile(fileext = "Rmd")
    rmarkdown:::convert_ipynb(code_file, tmp)
    knitr::purl(tmp, run_code_file, quiet=TRUE)
  }

  if (file_exists(run_code_file)) {
    tmp <- read_file(run_code_file)
    tmp <- wrap(run_code_file, tmp)
    write_file(tmp, run_code_file)
  }

  run_code_file
}
```

```{r load kaggle metadata}
files <- list.files(KAGGLE_KERNELS_DIR, "kernel-metadata\\.json$", full.names=TRUE, recursive=TRUE)

metadata_json <- map(files, ~fromJSON(file = .))
metadata <- map_dfr(
  metadata_json, 
  ~tibble(id=str_replace(.$id, "/", "-"), language=.$language, kernel_type=.$kernel_type, competition=.$competition_sources, code_file=.$code_file)
)

kernels <- 
  metadata %>%
  mutate(
    path=path(KAGGLE_KERNELS_DIR, id)
  ) %>% 
  filter(dir_exists(path))
```

```{r filter known competitions}
known_competitions <- semi_join(kernels, tibble(competition=params$competitions), by="competition")
```

```{r delete run dir}
if (dir_exists(KAGGLE_RUN_DIR)) dir_delete(KAGGLE_RUN_DIR)
```

```{r extract kernels}
extraction <- 
  known_competitions %>%
  rowwise() %>%
  do({
    tryCatch({
      tibble(id=.$id, run_file=process(.))
    }, error=function(e) {
      tibble(id=.$id, error=e$message)
    })
  })
```

## Prepare for run

```{r make a kernels_supported set}
kernels_supported <- 
  known_competitions %>%
  left_join(extraction, by="id") %>%
  filter(file_exists(run_file))
```

```{r compute sloc}
sloc <- map_dfr(path(KAGGLE_RUN_DIR, params$competitions), function(x) {
  out <- system2("cloc", c("--by-file-by-lang", "-q", "--csv", shQuote(x)), stdout = TRUE)  
  out <- out[2:(length(out)-4)]
  out[1] <- "language,filename,blank,comment,code"
  df <- read_csv(out)
}) %>%
  mutate(
    kernel=basename(dirname(filename))
  ) %>%
  select(kernel, code)

stopifnot(any(!duplicated(sloc$kernel)))
```

```{r add sloc to the kernels_supported set}
kernels_supported <- left_join(kernels_supported, sloc, by=c("id"="kernel"))
```

```{r kernels with no R code or duplication}
kernels_duplicated <- filter(kernels_supported, is.na(code) | code == 0)
kernels_runnable <- filter(kernels_supported, code > 0)
```

```{r display duplicated or empty kernels}
kernels_duplicated %>% 
  mutate(
    code_file=show_url(file.path(KAGGLE_KERNELS_DIR, id, "script", code_file)),
    path=show_url(path),
    run_file=show_url(run_file)
  ) %>%
  select(id, competition, path, run_file, language, code_file) %>%
  my_datatable(escape=FALSE)
```

```{r}
write_csv(kernels_runnable, KERNELS_FILE)
write_lines(dirname(kernels_runnable$run_file), SCRIPTS_FILE)
```

## Run

```{r run}
Sys.unsetenv("R_HOME")
run <- system2(
  "parallel",
  c(
    "-a", SCRIPTS_FILE,
    "--bar",
    "--jobs", parallel::detectCores(),
    "--results", PARALLEL_LOG_FILE,
    "--tagstring", "{/}",
    "--timeout", "1h",
    "--workdir", "{1}/",
    file.path(R_DIR, "bin", "R"), "CMD", "BATCH", "{1}/run.R", "{1}/run.R.out"
  )
)
```

```{r load results}
calls <- kernels_runnable %>%
  do({
    bd <- basename(.$run_file)
    tryCatch({
      df <- as_tibble(read_fst(path(bd, "calls.fst")))
      add_column(df, id=.$id, .before=1)
    }, error=function(e) {
      tibble(id=.$id, error=e$message)
    })
  })
```

```{r}
parallel_log <- 
  read_csv(PARALLEL_LOG_FILE)

parallel_log %>%
  mutate(
    output=show_url(path(V1, "run.R.out"))
  ) %>%
  my_datatable(escape=FALSE)
```

## Errors

### 2020-08-23 inst/analysis/inc/paths.R

@@ -2,10 +2,15 @@ DATA_DIR  <- path(params$base_dir, "data")
 PAPER_DIR <- path(params$base_dir, "paper")
 TAGS_DIR  <- path(PAPER_DIR, "tags")
 
+
+R_DIR <- "/R/R-4.0.2"
+
 ## INPUTs
 
-PACKAGE_METADATA     <- path(DATA_DIR, "metadata.csv")
-PACKAGE_SLOC         <- path(DATA_DIR, "sloc.csv")
+PACKAGE_METADATA_FILE     <- path(DATA_DIR, "metadata.csv")
+PACKAGE_SLOC_FILE         <- path(DATA_DIR, "sloc.csv")
+PACKAGE_COVERAGE_FILE     <- path(DATA_DIR, "coverage.csv")
+PACKAGE_COVERAGE_LOG      <- path(DATA_DIR, "coverage.log")
 
 ## OUTPUTs
 

### 2020-08-23 inst/analysis/inc/setup.R

@@ -5,3 +5,61 @@ is_outlier_max <- function(x, m=1.5) quantile(x, 0.75) + m * IQR(x)
 is_outlier <- function(x, m=1.5) {
   (x < is_outlier_min(x, m)) | (x > is_outlier_max(x, m))
 }
+
+show_url <- Vectorize(function(path, name=basename(path), hostname=params$hostname, port=params$port) {
+  str_glue('<a href="http://{hostname}:{port}/{URLencode(path)}">{name}</a>')
+}, vectorize.args = "path")
+
+read_runr_log <- function(path) {
+  log <- read_csv(
+    path,
+    col_types = cols(
+      Seq = col_double(),
+      Host = col_character(),
+      Starttime = col_double(),
+      JobRuntime = col_double(),
+      Send = col_double(),
+      Receive = col_double(),
+      Exitval = col_double(),
+      Signal = col_double(),
+      Command = col_character(),
+      V1 = col_character(),
+      Stdout = col_character(),
+      Stderr = col_logical()
+    )
+  ) %>%
+    rename_all(tolower) %>%
+    mutate(
+      package=basename(v1),
+      run_path=path(dirname(path), package),
+      output_file=path(run_path, "task-output.txt")
+    ) %>%
+    select(
+      package,
+      run_path
+    )
+  
+  task_status <- map2_dfr(log$package, log$run_path, function(package, run_path) {
+    tryCatch({
+      read_csv(
+        path(run_path, "task-stats.csv"), 
+        col_types=cols(
+          exitval=col_character(),
+          hostname=col_character(),
+          start_time=col_integer(),
+          end_time=col_integer(),
+          command=col_character()
+        )
+      ) %>%
+      add_column(package, .before=1)
+    }, error=function(e) tibble(package, start_time=NA, end_time=NA, error=e$message))
+  })
+  
+  left_join(log, task_status, by="package") %>%
+    mutate(
+      start_time=as_datetime(start_time),
+      end_time=as_datetime(end_time),
+      elapsed=end_time-start_time
+    ) %>%
+    select(-hostname)
+}

### 2020-08-23 inst/docker-image/get-installed-packages.R

@@ -1,5 +1,5 @@
 #!/usr/bin/env Rscript
-options(repos=Sys.getenv("CRAN_LOCAL_MIRROR", "https://cloud.r-project.org"))
+options(repos="https://cloud.r-project.org")
 
 # pins the current version of the corpus packages as well as
 # their dependencies
@@ -26,7 +26,13 @@ packages_kaggle <- readLines("kaggle-dependencies.txt")
 packages_base <- readLines("our-packages-dependencies.txt")
 packages <- unique(c(packages_corpus, packages_kaggle, packages_base))
 
-dependencies <- unlist(tools::package_dependencies(packages, recursive=T))
+dependencies <- unlist(
+  tools::package_dependencies(
+    packages,
+    which = c("Depends", "Imports", "LinkingTo", "Suggests"),
+    recursive=T
+  )
+)
 packages <- unique(c(packages, dependencies))
 packages <- sort(setdiff(packages, BASE_PACKAGES))
 

### 2020-08-23 inst/in-docker.sh

@@ -6,6 +6,8 @@ script_dir="$( cd "$(dirname "${BASH_SOURCE[0]}")" ; pwd -P )"
 base_dir=$(dirname "$script_dir")
 docker_working_dir="/home/rstudio/evalR"
 local_working_dir="$base_dir/inst"
+cmd="bash"
+[ $# -gt 0 ] && cmd="$@"
 
 docker run \
     -ti \
@@ -17,7 +19,7 @@ docker run \
     -v "$local_working_dir:$docker_working_dir" \
     -w "$docker_working_dir" \
     "$DOCKER_IMAGE_NAME" \
-    "$@"
+    $cmd
 
 # docker run \
 #     --rm \

### 2020-08-23 inst/run.sh

@@ -46,7 +46,9 @@ docker run \
   -p "$port:8787" \
   -v "$(pwd):/home/rstudio/evalR" \
   "$DOCKER_IMAGE_NAME" \
-  bash -c "source /etc/cont-init.d/userconf && exec /usr/lib/rstudio-server/bin/rserver --auth-none 1 --auth-timeout-minutes 0 --www-port 8787 --server-daemonize 0"
+  $cmd
+
+  #bash -c "source /etc/cont-init.d/userconf && exec /usr/lib/rstudio-server/bin/rserver --auth-none 1 --auth-timeout-minutes 0 --www-port 8787 --server-daemonize 0"
 
   # -v "$(pwd)/README.Rmd:/home/rstudio/README.Rmd" \
   # -v "$(pwd)/README.html:/home/rstudio/README.html" \

### 2020-08-23 inst/analysis/corpus-stage1.Rmd
Add corpus data

---
title: "Corpus Stage 1"
output: html_document
params:
  base_dir: /var/lib/R/project-evalR
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
local(for (x in c(
  "dplyr", 
  "DT", 
  "fs", 
  "fst",
  "readr",
  "purrr"
  )) {
  suppressPackageStartupMessages(library(x, character.only=TRUE))
})

knitr::opts_chunk$set(echo = TRUE)

source("inc/paths.R")
source("inc/setup.R")
source("inc/latextags.R")
```

## Description

This notebook selects TOP `r fmt(CORPUS_S1_NUM_PKGS)` packages into `r CORPUS_S1_FILE` based on the number of reverse dependencies.
We then run all the tasks on this set.
Finally, we select TOP `r fmt(CORPUS_NUM_PKGS)` packages for the paper for the packages that ran successfully.
The reason is that not all the code in packages is runnable in all environments.
Especially, the way `covr` runs to get code coverage does not work on all.

### Output

- ``r CORPUS_S1_FILE`` - selected packages based on

## Prepare the corpus

```{r set local mirror}
options(repos=CRAN_MIRROR_LOCAL_URL)
```

```{r get all available packages}
cran_packages_raw <- available.packages()
cran_packages <- as_tibble(cran_packages_raw) %>%
  select(package=Package, version=Version)
```

### Package reverse dependencies

There are two functions `package_dependencies` and `dependsOnPkgs`.
Not sure what is the difference, but the `dependsOnPkgs` is very slow.

```{r compute reverse dependencies}
pkgs_revdeps_raw <- 
  tools::package_dependencies(
    cran_packages$package,
    db=cran_packages_raw,
    which=WHICH_DEPENDENCIES,
    reverse=TRUE,
    recursive=FALSE
  ) %>% 
  imap_dfr(~tibble(package=.y, revdep=.x))

pkgs_revdeps <- 
  pkgs_revdeps_raw %>%
  count(package) %>%
  rename(revdeps=n)
```

### TOP packages based on number of reverse dependencies

```{r compute TOP n packages}
pkgs_top_revdeps <-
  pkgs_revdeps %>%
  top_n(n = CORPUS_S1_NUM_PKGS, wt=revdeps) %>%
  arrange(desc(revdeps))

# cup the number of reverse dependencies to the required number
pkgs_top_revdeps <- head(pkgs_top_revdeps, min(nrow(pkgs_top_revdeps), CORPUS_S1_NUM_PKGS))
```

### Missing packages from TOP 100 downloads

```{r get last month most downloaded packages}
pkgs_top_downloaded <-
  cranlogs::cran_top_downloads("last-month", 100)
```

```{r compute missing packages from the most downloaded ones}
missing <- setdiff(pkgs_top_downloaded$package, pkgs_top_revdeps$package)
missing_revdeps <- filter(pkgs_revdeps, package %in% missing)
missing_revdeps %>%
  left_join(pkgs_top_downloaded, by="package") %>%
  datatable()
```

### Corpus

Putting it all together and outputting the results.

```{r corpus s1}
corpus_s1 <- 
  pkgs_top_revdeps %>%
  left_join(pkgs_top_downloaded, by="package") %>%
  arrange(desc(revdeps))
```

```{r render selected packages}
corpus_s1 %>%
  datatable()
```

```{r write down corpus}
write_lines(corpus_s1$package, CORPUS_S1_FILE)
write_fst(select(corpus_s1, package, revdeps, rank), CORPUS_S1_DETAILS_FILE)
```

### 2020-08-23 inst/analysis/inc/paths.R

@@ -1,14 +1,47 @@
+################################################################################
+# PATHS
+################################################################################
+
+DATA_DIR  <- path(params$base_dir, "revalstudy", "inst", "data")
+PAPER_DIR <- path(params$base_dir, "revalstudy", "inst", "paper")
+RUN_DIR   <- path(params$base_dir, "run")
+TAGS_DIR  <- path(PAPER_DIR, "tags")
+
+R_DIR      <- path(params$base_dir, "R-4.0.2")
+R_LIBS_DIR <- path(params$base_dir, "library", "4.0")
+
 ## INPUTs
 
-PACKAGE_METADATA     <- path(params$data_dir, "metadata.csv")
-PACKAGE_SLOC         <- path(params$data_dir, "sloc.csv")
+PACKAGE_EVALS_FILE         <- path(DATA_DIR, "evals.fst")
+PACKAGE_COVERAGE_FILE      <- path(DATA_DIR, "coverage.fst")
+PACKAGE_METADATA_FILE      <- path(DATA_DIR, "metadata.fst")
+PACKAGE_SLOC_FILE          <- path(DATA_DIR, "sloc.fst")
+PACKAGE_REVDEPS_FILE       <- path(DATA_DIR, "revdeps.fst")
+PACKAGE_FUNCTIONS_FILE     <- path(DATA_DIR, "functions.fst")
+PACKAGE_RUNNABLE_CODE_FILE <- path(DATA_DIR, "runnable-code-metadata.fst")
+PACKAGE_RUNS_FILE          <- path(DATA_DIR, "run.fst")
+PACKAGE_TRACE_RUNS_FILE    <- path(DATA_DIR, "run-trace.fst")
+PACKAGE_TRACE_LOG_FILE     <- path(DATA_DIR, "parallel-trace.fst")
+
+KAGGLE_KERNELS_FILE <- path(DATA_DIR, "kaggle-kernels.fst")
+KAGGLE_LOG_FILE     <- path(DATA_DIR, "parallel-kaggle.fst")
 
 ## OUTPUTs
 
-## corpus only
-CORPUS_FILE                  <- path(params$output_dir, "corpus.txt")
-CORPUS_REVDEPS               <- path(params$output_dir, "corpus-revdeps.csv")
-CORPUS_REVDEPS_FILE          <- path(params$output_dir, "corpus-revdeps.txt")
-CORPUS_ALL_PACKAGES_FILE     <- path(params$output_dir, "corpus-all.txt")
+# TODO: rename to corpus-stage1.*
[…]

### 2020-08-23 inst/analysis/inc/setup.R

@@ -5,3 +5,7 @@ is_outlier_max <- function(x, m=1.5) quantile(x, 0.75) + m * IQR(x)
 is_outlier <- function(x, m=1.5) {
   (x < is_outlier_min(x, m)) | (x > is_outlier_max(x, m))
 }
+
+show_url <- Vectorize(function(path, name=basename(path), hostname=params$hostname, port=params$port) {
+  str_glue('<a href="http://{hostname}:{port}/{URLencode(path)}">{name}</a>')
+}, vectorize.args = "path")

### 2020-08-23 inst/in-docker.sh

@@ -1,26 +1,0 @@
-#!/bin/bash
-
-DOCKER_IMAGE_NAME="prlprg/project-evalr"
-
-script_dir="$( cd "$(dirname "${BASH_SOURCE[0]}")" ; pwd -P )"
-base_dir=$(dirname "$script_dir")
-docker_working_dir="/home/rstudio/evalR"
-local_working_dir="$base_dir/inst"
-
-docker run \
-    -ti \
-    --rm \
-    -e ROOT=TRUE \
-    -e DISABLE_AUTH=true \
-    -e USERID=$(id -u) \
-    -e GROUPID=$(id -g) \
-    -v "$local_working_dir:$docker_working_dir" \
-    -w "$docker_working_dir" \
-    "$DOCKER_IMAGE_NAME" \
-    "$@"
-
-# docker run \
-#     --rm \
-#     -v "$local_working_dir:$docker_working_dir" \
-#     "$DOCKER_IMAGE_NAME" \
-#     chown -R $(id -u):$(id -g) "$docker_working_dir"

### 2020-08-23 inst/paper/main.tex
Add basic eval numbers

@@
 \usepackage{xspace}
+
+\input{counts.tex}
 
@@
 
-In our corpus eval is called \missingNumber times out of \missingNumber
+In our corpus eval is called \AllAllCallCountRnd{} times out of \missingNumber
 total calls to all R functions. This is merely \missingPercentage of the total
@@
 Kaggle code.
+
+\begin{tabular}{c|c|c|c|c|c}
+  Corpus  & eval & evalq & evalParent & local & all \\
+  Core & \CoreEvalCallCountRnd & \CoreEvalqCallCountRnd & \CoreEvalParentCallCountRnd & \CoreLocalCallCountRnd & \CoreAllCallCountRnd \\
+  Package & \PackageEvalCallCountRnd & \PackageEvalqCallCount & \PackageEvalParentCallCountRnd & \PackageLocalCallCountRnd & \PackageAllCallCountRnd \\
+\end{tabular}
 

### 2020-08-23 inst/analysis/inc/latextags.R
Fix latextags to generate Rnd variables

@@ -34,7 +34,7 @@ r.numeric <- function(name, value, ...) {
   raw_str <- fmt(value, ...)
   rnd_str <- fmt(oom(value), ...)
   
-  if (!is.na(value) && raw_str != rnd_str) {
+  if (!is.na(value)) {
     tag(str_c(name, " rnd"), rnd_str)
   }
   

### 2020-08-23 inst/paper/main.tex
Add tables for call counts and call sites

--- 2020-08-23 inst/paper/main.tex
@@
 
+\begin{tabular}{c|c|c|c|c|c}
+  Function  & Core & Packages & Kaggle & All \\
+  eval & \CoreEvalCallCountRnd & \PackageEvalCallCountRnd & \KaggleEvalCallCountRnd & \AllEvalCallCountRnd \\
+  evalParent & \CoreEvalParentCallCountRnd & \PackageEvalParentCallCountRnd & \KaggleEvalParentCallCountRnd & \AllEvalParentCallCountRnd \\
+  evalq & \CoreEvalqCallCountRnd & \PackageEvalqCallCountRnd & \KaggleEvalqCallCountRnd & \AllEvalqCallCountRnd \\
+  local & \CoreLocalCallCountRnd & \PackageLocalCallCountRnd & \KaggleLocalCallCountRnd & \AllLocalCallCountRnd \\
+  all & \CoreAllCallCountRnd & \PackageAllCallCountRnd & \KaggleAllCallCountRnd & \AllAllCallCountRnd
+\end{tabular}
+
+\begin{tabular}{c|c|c|c|c|c}
+  Function  & Core & Packages & Kaggle & All \\
+  eval & \CoreEvalSiteCountRnd & \PackageEvalSiteCountRnd & \KaggleEvalSiteCountRnd & \AllEvalSiteCountRnd \\
+  evalParent & \CoreEvalParentSiteCountRnd & \PackageEvalParentSiteCountRnd & \KaggleEvalParentSiteCountRnd & \AllEvalParentSiteCountRnd \\
+  evalq & \CoreEvalqSiteCountRnd & \PackageEvalqSiteCountRnd & \KaggleEvalqSiteCountRnd & \AllEvalqSiteCountRnd \\
+  local & \CoreLocalSiteCountRnd & \PackageLocalSiteCountRnd & \KaggleLocalSiteCountRnd & \AllLocalSiteCountRnd \\
+  all & \CoreAllSiteCountRnd & \PackageAllSiteCountRnd & \KaggleAllSiteCountRnd & \AllAllSiteCountRnd
+\end{tabular}
+
 The core R packages are the biggest consumers of eval. Majority of calls,
-\missingPercentage, originate from the core R packages. Kaggle code, which is
-end-user code, has merely \missingNumber calls to eval.
+\CoreAllCallCountRnd, originate from the core R packages. Kaggle code, which is
+end-user code, has merely \AllKaggleCallCountRnd calls to eval.
 

### 2020-08-23 inst/paper/main.tex
Fix
Add run data

@@
 \CoreAllCallCountRnd, originate from the core R packages. Kaggle code, which is
-end-user code, has merely \AllKaggleCallCountRnd calls to eval.
+end-user code, has merely \KaggleAllCallCountRnd calls to eval.
 
@@
 \missingPercentage calls to eval.
+
+All the \TotalFileCount files in our corpus call eval. \CoreEvalFileCount
+(\CoreEvalFilePerc) call only core evals, \AllEvalFileCount (\AllEvalFilePerc)
+call both package and core evals; and, \PackageEvalFileCount
+(\PackageEvalFilePerc) call only package evals. Of the \AllEvalFileCount
+(\AllEvalFilePerc) files that call both package and base eval,
+\EightyCoreEvalFilePerc (\EightyCoreEvalFileCount) files make over 80\% eval
+calls to core evals.
 

### 2020-08-23 inst/analysis/analysis_stable.Rmd
Classification

--- 2020-08-21 inst/analysis/analysis_stable.Rmd
@@
         toc_depth: 3
+        self_contained: no
 date: "`r Sys.Date()`"
@@
 source("inc/latextags.R")
-create_tags(path("/var/lib/R/project-evalR/revalstudy/data", "analysisStable.tex"), prefix="", default=TRUE)
+create_tags(path("/var/lib/R/project-evalR/revalstudy/inst/data", "analysisStable.tex"), prefix="", default=TRUE)
 ```
@@
                             mutate(group_function_expr = fct_collapse(function_expr2, 
-                              assignment = c("<-", "[[<-", "[<-", "$<-", "<<-", "="),
-                              read = c("[", "[[", "$"),
+                              assignment = c("<-", "[[<-", "[<-", "$<-", "<<-", "=", "@<-"),
+                              slot_access = c("[", "[[", "$", "slot", "@", "::", ":::"),
                               parsing = c("parse", "str2lang", "str2expression", "parse_only", "base::parse"),
@@
                               environment = c("environment", "as.environment"),
-                              classAssign = c("@<-"),
-                              classRead = c("slot", "@"),
-                              classDef = c(".Primitive(\"class\")"),
-                              logical = c("<", ">", "is.element"),
+                              logical = c("!", "&", "|", "&&", "||"),
+                              relation = c("<", ">", "is.element", "<=", "=>", "!="),
                               choice = c("if"),
@@
                               function_meta = c("sys.calls", "function", "body", "missing"),
+                              arithmetic = c("+", "-", "*", "/", "%%", "^"),
                               formula = c("~", "deriv", "D", "model.frame", "rootftfunc", "contest.lmerModLmerTest", "JacFunc", "gamlss.family", "contest1D.lmerModLmerTest", "contestMD.lmerModLmerTest", "reexpr"),
@@
   facet_wrap(vars(eval_function))
+```
+
+Another interesting figure is the degreee of polymorphism of the `expr` argument: how many different types (for `expr_resolved`) are there for one call sites?
+
+```{r polymorphism_callsite}
+expr_resolved_polymorphism <- eval_calls %>% group_by(eval_call_srcref) %>%
+        summarise(polym_degree = vec_unique_count(expr_resolved_type))
+
+expr_resolved_monomorphic <- expr_resolved_polymorphism %>% filter(polym_degree == 1)
+nb_monomorphic <- expr_resolved_monomorphic %>% nrow()
+```
+
+```{r polymorphism}
+expr_resolved_polymorphism %>% ggplot() +
+  geom_bar(aes(x = polym_degree)) +
+  scale_x_continuous(breaks=c(0, 1, 5, 10)) +
+  labs(title = "Degree of polymorphism", x = "degree", y = "n") 
+```
+
+`r nb_monomorphic` call sites are not polymorphic, which represents `r nb_monomorphic / nb_call_sites * 100`% of the call sites.
+
+From the monomorphic functions, we can deduce how many are _replaceable_:
+
+```{r}
+expr_resolved_replaceable <- expr_resolved_monomorphic %>% left_join(eval_calls) %>% mutate(can_replace = map_lgl(expr_resolved,  is_replaceable_str))
 ```

### 2020-08-23 inst/analysis/enrich_dataset.Rmd
Classification

@@
 params:
-  data_dirpath: "/var/lib/R/project-evalR/run/package-evals-traced-corpus.5/"
-  data_filepath: "/var/lib/R/project-evalR/run/package-evals-traced-corpus.5/calls.fst"
+  data_dirpath: "/var/lib/R/project-evalR/run/package-evals-traced/"
+  data_filepath: "/var/lib/R/project-evalR/run/package-evals-traced/calls-final.fst"
   data_kagglepath: "/var/lib/R/project-evalR/kaggle-korpus/run/titanic/calls.fst"

### 2020-08-23 inst/analysis/insights.R

@@ -265,6 +265,7 @@ replaceable_functions <- c("+", "*", "/", "-", "%%",
                            "[", "[[", "$",
                            "@<-",
                            "slot", "@",
+                           "&", "&&", "|", "||", "!",
                            "<", ">", "==", "<-", ">=", "!=",
                            "if")
 

### 2020-08-23 inst/analysis/analysis_stable.Rmd
Add makefile for the pipeline

@@
 params:
-    data_filepath: "/var/lib/R/project-evalR/run/package-evals-traced-corpus.5/summarized-packages.fst"
-    #data_filepath: "/var/lib/R/project-evalR/run/package-evals-traced-all.1/calls.fst"
-    # The trace data on all CRAN does not have all the columns of the smaller corpus. The analysis will fail for the big corpus.
+  base_dir: /var/lib/R/project-evalR
 
@@
 
+source("inc/paths.R")
 source("inc/latextags.R")
-create_tags(path("/var/lib/R/project-evalR/revalstudy/inst/data", "analysisStable.tex"), prefix="", default=TRUE)
+create_tags(path(TAGS_DIR, "analysisStable.tex"), prefix="", default=TRUE)
+
+data_filepath <- EVALS_SUM_PKGS_FILE
+#data_filepath <- "/var/lib/R/project-evalR/run/package-evals-traced-all.1/calls.fst"
+# The trace data on all CRAN does not have all the columns of the smaller corpus. The analysis will fail for the big corpus.
 ```
@@
 
-The analysis is being performed on dataset read from file `r params$data_filepath`.
+The analysis is being performed on dataset read from file `r data_filepath`.
 
 ```{r file-info, echo=FALSE}
-file_info(params$data_filepath)
+file_info(data_filepath)
 ```
@@
 eval_calls <-
-    read_fst(params$data_filepath) %>%
+    read_fst(data_filepath) %>%
     as_tibble() %>%

### 2020-08-23 inst/analysis/corpus-analysis.Rmd
Add makefile for the pipeline

--- 2020-08-21 inst/analysis/corpus-analysis.Rmd
@@
 title: "Corpus Analysis"
-output: 
-  rmdformats::readthedown:
-        code_folding: hide
-        lightbox: true
-        gallery: false
-        df_print: paged
-        toc_depth: 3
-date: "`r Sys.Date()`"
+output: html_document
+params:
+  base_dir: /var/lib/R/project-evalR
 editor_options: 
   chunk_output_type: console
-params:
-knit: (function(inputFile, encoding) {
-  rmarkdown::render(inputFile, encoding = encoding, output_dir = "html") })
 ---
+
+```{r setup, include=FALSE}
+for (x in c(
+  "dplyr", 
+  "DT",
+  "fs",
+  "fst",
+  "lubridate",
+  "readr",
+  "purrr",
+  "stringr",
+  "tidyr"
+  )) {
+  suppressPackageStartupMessages(library(x, character.only=TRUE))
+}
+
+knitr::opts_chunk$set(echo = TRUE)
+
+# TODO: move into the package
+source("inc/paths.R")
+source("inc/setup.R")
+source("inc/latextags.R")
+
+create_tags(path(TAGS_DIR, "corpus.tex"), prefix="", default=TRUE)
+```
+
+## Load data
+
+```{r}
+
+```
+
+## Corpus Overview
+
+### Table
+
+```{r summary}
+overview_table(
+  r("corpus", corpus),
+  r("runnable code",        sum(corpus$runnable_code, na.rm=T)),
+  r("examples code",        sum(corpus$runnable_code_examples, na.rm=T)),
+  r("tests code",           sum(corpus$runnable_code_tests, na.rm=T)),
+  r("vignettes code",       sum(corpus$runnable_code_vignettes , na.rm=T)),
+  r("mean expr coverage",   percent(mean(corpus$coverage_expr))),
+  r("median expr coverage", percent(median(corpus$coverage_expr))),
+  r("with R code",          filter(corpus, package_r_code > 0)),
+  r("R code",               sum(corpus$package_r_code, na.rm=T)),
+  r("with native code",     filter(corpus, package_native_code > 0)),
+  r("native code",          sum(corpus$package_native_code, na.rm=T)),
+  r("functions",            sum(corpus$funs, na.rm=T)),
+  r("public functions",     sum(corpus$funs_public, na.rm=T)),
+  r("private functions",    sum(corpus$funs_private, na.rm=T)),
+  r("revdes",               sum(corpus$revdeps)),
+  r("mean revdes",          mean(corpus$revdeps)),
+  r("median revdes",        median(corpus$revdeps)),
+  r("runtime",              as.duration(sum(corpus$runnable_time))),
+  r("mean runtime",         as.duration(mean(corpus$runnable_time))),
+  r("median runtime",       as.duration(median(corpus$runnable_time)))
+)
+```
+
+### Plots
+
+```{r plot corpus}
+coverage_mean <- mean(df$coverage_expr)
+coverage_median <- median(df$coverage_expr)
+revdeps_mean <- mean(df$revdeps)
+revdeps_median <- median(df$revdeps)
+
+df %>%
+  mutate(
+    outlier=package_code > PACKAGE_SIZE_OUTLIER,
+    label=if_else(outlier, str_glue("{package}"), as.character(NA))
+  ) %>%
+  ggplot(
+    aes(
+      x=coverage_expr, 
+      y=revdeps,
+      label=label,
+      size=package_code
+    )
+  ) +
+  geom_point(alpha=.6) +
+  geom_text(size=3, check_overlap = T, vjust=1.5, na.rm = TRUE) +
+
+  geom_hline(aes(yintercept=revdeps_mean), linetype="dashed", color="red", size=0.2) +
+  geom_vline(aes(xintercept=coverage_mean), linetype="dashed", color="red", size=0.2) +
+
+  scale_x_continuous(limits = c(params$min_coverage, 1), labels = scales::percent) + 
+  scale_y_log10(labels = scales::comma) +
+  scale_size(range = c(.1, 3), labels = scales::comma) +
+    theme(
+    legend.position=c(0.2, 0.85),
+    legend.box="horizontal",
+    legend.box.background = element_rect(fill="white", size=0.1)
+  ) +
+  labs(
+    x="Code coverage", 
+    y="Number of reverse dependencies (log)", 
+    size="Code size",
+    linetype=""
+  ) +
+  guides(
+    size=guide_legend(order=1),
+    color=guide_legend(order=2),
+    linetype=F
+  )
+
+ggsave(CORPUS_PLOT)
+```
+
+### List of corpus packages
+
+```{r list of packages, eval=FALSE}
+datatable(
+  corpus, 
+  filter="top", 
+  options=list(
+    pageLength=50,
+    autoWidth=TRUE
+  )
+)
+```
+
+## Kaggle
+
+```{r, eval=FALSE}
+parallel_log <- 
+  read_csv(PARALLEL_LOG_FILE)
+
+parallel_log %>%
+  mutate(
+    output=show_url(path(V1, "run.R.out"))
+  ) %>%
+  my_datatable(escape=FALSE)
+```

### 2020-08-23 inst/analysis/corpus-stage2.Rmd
Add makefile for the pipeline

---
title: "Corpus Analysis"
output: html_document
params:
  base_dir: /var/lib/R/project-evalR
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
for (x in c(
  "dplyr", 
  "DT",
  "fs",
  "fst",
  "lubridate",
  "readr",
  "purrr",
  "stringr",
  "tidyr"
  )) {
  suppressPackageStartupMessages(library(x, character.only=TRUE))
}

knitr::opts_chunk$set(echo = TRUE)

# TODO: move into the package
source("inc/paths.R")
source("inc/setup.R")
```

```{r aux functions}
read_task_result <- function(path) {
  read_fst(path) %>% as_tibble() %>% mutate(package=basename(package))
}

read_parallel_log <- function(path) {
  read_fst(path) %>% 
    as_tibble() %>% 
    rename_all(tolower) %>%
    mutate(
      package=map_chr(command, ~basename(str_split(., " ")[[1]][2])),
      starttime=as_datetime(starttime),
      endtime=as_datetime(starttime + jobruntime),
      runtime=endtime-starttime,
      run_path=map_chr(command, ~dirname(str_split(., " ")[[1]][3]))
    ) %>%
    select(package, exitval, starttime, endtime, runtime, command, run_path)
}
```

## Load data

```{r loading raw data}
corpus_s1_raw     <- read_task_result(CORPUS_S1_DETAILS_FILE)
evals_raw         <- read_task_result(PACKAGE_EVALS_FILE)
coverage_raw      <- read_task_result(PACKAGE_COVERAGE_FILE)
metadata_raw      <- read_task_result(PACKAGE_METADATA_FILE)
functions_raw     <- read_task_result(PACKAGE_FUNCTIONS_FILE)
sloc_raw          <- read_task_result(PACKAGE_SLOC_FILE)
runnable_code_raw <- read_task_result(PACKAGE_RUNNABLE_CODE_FILE)
runs_raw          <- read_task_result(PACKAGE_RUNS_FILE)
trace_runs_raw    <- read_task_result(PACKAGE_TRACE_RUNS_FILE)

trace_log_raw     <- read_parallel_log(PACKAGE_TRACE_LOG_FILE)
```

## Process rawdata

### Metadata

```{r process metadata}
metadata <- select(metadata_raw, -name)
```

```{r process sloc}
sloc <- 
  sloc_raw %>%
  filter(path == "R" | path == "src") %>%
  select(-blank, -comment, -files, -path) %>%
  mutate(language=case_when(
    language == "R" ~ "package_r_code",
    language == "C" ~ "package_native_code",
    language == "C/C++ Header" ~ "package_native_code",
    language == "C++" ~ "package_native_code",
    startsWith(language, "Fortran") ~ "package_native_code",
    TRUE ~ as.character(NA)
  )) %>%
  filter(!is.na(language)) %>%
  group_by(package, language) %>%
  summarise(code=sum(code)) %>%
  spread(key=language, value=code, fill=0) %>%
  mutate(package_code=package_native_code + package_r_code)
```

```{r process functions}
functions <-
  functions_raw %>%
  count(package, exported) %>%
  mutate(exported=ifelse(exported, "funs_public", "funs_private")) %>%
  pivot_wider(
    names_from=exported, 
    values_from=n, 
    values_fill=list(n=0)
  )
```

### Runnable code

Next to the code itself, R packages have examples, tests and vignettes.
These can include snippets of R code which calls the package functions.
We extract this code and run it to narrow down the list of packages that can be considered into the final corpus.

```{r process runnable code}
runnable_code <- 
  runnable_code_raw %>%
  filter(language=="R") %>%
  select(package, path, code, files) %>%
  pivot_wider(
    names_from=path, 
    values_from=c(code, files), 
    values_fill=list(code=0, files=0)
  ) %>%
  # TODO we have old version of tidyr, names_glue is missing, so have to rename manually
  rename(
    runnable_code_examples=code_examples,
    runnable_code_tests=code_tests,
    runnable_code_vignettes=code_vignettes,
    runnable_files_examples=files_examples,
    runnable_files_tests=files_tests,
    runnable_files_vignettes=files_vignettes,
  ) %>%
  mutate(
    runnable_code=runnable_code_examples+runnable_code_tests+runnable_code_vignettes,
    runnable_files=runnable_files_examples+runnable_files_tests+runnable_files_vignettes
  ) %>%
  select(
    package,
    runnable_code,
    runnable_files,
    everything()
  )
```

### Coverage

```{r process coverage}
coverage <- coverage_raw %>%
  filter(type=="all", is.na(error), !is.na(coverage_expression)) %>% 
  select(-type, -error, coverage_expr=coverage_expression) %>%
  mutate(coverage_expr=coverage_expr/100, coverage_line=coverage_line/100)
```

### Package runs

```{r process package runs}
runs <-
  runs_raw %>%
  group_by(package) %>%
  summarise(runnable_success=sum(is.na(error)), runnable_error=sum(!is.na(error)), runnable_time=sum(time, na.rm=TRUE)) %>%
  mutate(runnable_rate=runnable_success/(runnable_success+runnable_error))
```

### Trace runs

```{r process trace runs}
trace_runs <-
  trace_runs_raw %>%
  group_by(package) %>%
  summarise(trace_runnable_success=sum(is.na(error)), trace_runnable_error=sum(!is.na(error)), trace_runnable_time=sum(time, na.rm=TRUE)) %>%
  mutate(trace_runnable_rate=trace_runnable_success/(trace_runnable_success+trace_runnable_error))
```

```{r process trace log}
trace_log <-
  trace_log_raw %>%
  select(package, trace_exitval=exitval, trace_start=starttime, trace_end=endtime, trace_runtime=runtime)
```

### Package evals

```{r process evals}
evals <- 
  evals_raw %>%
  count(package, fun_name) %>%
  group_by(package) %>%
  summarise(funs_with_eval=length(unique(fun_name)), evals=sum(n))
```

### Join

```{r joining data}
corpus_s1 <-
  corpus_s1_raw %>%
  left_join(metadata, by="package") %>%
  left_join(sloc, by="package") %>%
  left_join(functions, by="package") %>%
  left_join(runnable_code, by="package") %>%
  left_join(coverage, by="package") %>%
  left_join(runs, by="package") %>%
  left_join(trace_runs, by="package") %>%
  left_join(trace_log, by="package") %>%
  left_join(evals, by="package")
```

### Select corpus

```{r compute corpus}
corpus <-
  corpus_s1 %>%
  filter(
    trace_exitval == 0,
    coverage_expr > 0
  ) %>%
  arrange(desc(revdeps), desc(coverage_expr)) %>%
  head(CORPUS_NUM_PKGS)
```

```{r export corpus}
write_lines(corpus$package, CORPUS_FILE)
write_fst(corpus, CORPUS_DETAILS_FILE)
```

### 2020-08-23 inst/analysis/enrich_dataset.Rmd
Add makefile for the pipeline

@@
 params:
-  data_dirpath: "/var/lib/R/project-evalR/run/package-evals-traced/"
-  data_filepath: "/var/lib/R/project-evalR/run/package-evals-traced/calls-final.fst"
-  data_kagglepath: "/var/lib/R/project-evalR/kaggle-korpus/run/titanic/calls.fst"
+  base_dir: /var/lib/R/project-evalR
 knit: (function(inputFile, encoding) {
@@
 
+source("inc/paths.R")
 source("insights.R", local = knitr::knit_global())
-
 source("preprocess.R", local = knitr::knit_global())
-```
-
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
 ```
@@
 
-The analysis is being performed on a dataset read from file `r params$data_filepath`.
+The analysis is being performed on a dataset read from file `r CALLS_FINAL_FILE`.
 
 ```{r file-info, echo=FALSE}
-file_info(params$data_filepath)
+file_info(CALLS_FINAL_FILE)
 ```
@@
 eval_calls_raw <-
-    read_fst(params$data_filepath) %>%
+    read_fst(CALLS_FINAL_FILE) %>%
     as_tibble() %>%
@@
 eval_calls_kaggle_raw <- 
-  read_fst(params$data_kagglepath) %>%
+  read_fst(KAGGLE_CALLS_FINAL_FILE) %>%
   as_tibble() %>% 
@@
 ```{r write_back}
-write_fst(eval_calls, paste0(params$data_dirpath, "raws.fst"))
+write_fst(eval_calls, EVALS_RAW_FILE)
 
-write_fst(eval_calls_core, paste0(params$data_dirpath, "summarized-core.fst"))
+write_fst(eval_calls_core, EVALS_SUM_CORE_FILE)
 
-write_fst(eval_calls_packages, paste0(params$data_dirpath, "summarized-packages.fst"))
+write_fst(eval_calls_packages, EVALS_SUM_PKGS_FILE)
 
-write_fst(eval_calls_kaggle, paste0(params$data_dirpath, "summarized-kaggle.fst"))
+write_fst(eval_calls_kaggle, EVALS_SUM_KAGGLE_FILE)
 ```

### 2020-08-23 inst/analysis/kaggle.Rmd
Add makefile for the pipeline

@@
 params:
-  base_dir: /home/rstudio/evalR
-  hostname: prl2
+  base_dir: /var/lib/R/project-evalR
+  hostname: prl3
   port: 8788
@@
 for (x in c(
+  "callr",
   "dplyr", 
@@
   "fst",
+  "knitr",
   "purrr",
@@
   "rjson",
+  "rmarkdown",
   "runr",
@@
 source("inc/latextags.R")
+```
 
-create_tags(path(TAGS_DIR, "kaggle.tex"), prefix="Kaggle", default=TRUE)
-
-KAGGLE_KERNELS_DIR  <- path(params$base_dir, "run", "kaggle-kernels", "notebooks", "r", "kernels")
-KAGGLE_DATASETS_DIR <- path(params$base_dir, "run", "kaggle-datasets")
-KAGGLE_RUN_DIR      <- path(params$base_dir, "run", "kaggle-run")
+```{r consts}
+KAGGLE_KERNELS_DIR  <- path(RUN_DIR, "kaggle-kernels", "notebooks", "r", "kernels")
+KAGGLE_DATASETS_DIR <- path(RUN_DIR, "kaggle-datasets")
+KAGGLE_RUN_DIR      <- path(RUN_DIR, "kaggle-run")
 
@@
 SCRIPTS_FILE      <- path(KAGGLE_RUN_DIR, "scripts.txt")
-PARALLEL_LOG_FILE <- path(KAGGLE_RUN_DIR, "parallel.csv")
+PARALLEL_LOG_FILE <- path(KAGGLE_RUN_DIR, "parallel.log")
 ```
@@
 
-process <- function(x) {
-  run_path <- path(KAGGLE_RUN_DIR, x$competition, basename(x$path))
-  code_file <- path(x$path, "script", x$code_file)
+extract_kernel_code <- function(path, code_file, competition) {
+  run_path <- path(KAGGLE_RUN_DIR, competition, basename(path))
+  code_file <- path(path, "script", code_file)
   run_code_file <- path(run_path, "run.R")
 
-  dir_create(run_path)
+  if (!dir_exists(run_path)) dir_create(run_path)
 
@@
 ```
+
+## Description
+
+This notebook prepares the kaggle kernels to be run with the eval tracing framework.
+
+### Input
+
+- `r KAGGLE_DATASETS_DIR` - directory with kaggle datasets
+- `r KAGGLE_KERNELS_DIR` - directory with kaggle kernels
+
+### Output
+
+- `r KAGGLE_RUN_DIR` - directory with prepared kernels to be run
+- command to run the datasets
+
+### Load kernels
 
@@
 ```{r delete run dir}
-if (dir_exists(KAGGLE_RUN_DIR)) dir_delete(KAGGLE_RUN_DIR)
+if (dir_exists(KAGGLE_RUN_DIR)) {
+  warning("*** ", KAGGLE_RUN_DIR, " exists")
+}
 ```
@@
 ```{r extract kernels}
-extraction <- 
-  known_competitions %>%
-  rowwise() %>%
-  do({
+extraction_lst <- pbapply::pbapply(known_competitions, 1, cl=16, function(x) {
     tryCatch({
-      tibble(id=.$id, run_file=process(.))
+      tibble(id=x["id"], run_file=extract_kernel_code(x["path"], x["code_file"], x["competition"]))
     }, error=function(e) {
-      tibble(id=.$id, error=e$message)
+      tibble(id=x["id"], error=e$message)
     })
 })
+extraction <- bind_rows(extraction_lst)
 ```
@@
 sloc <- map_dfr(path(KAGGLE_RUN_DIR, params$competitions), function(x) {
-  out <- system2("cloc", c("--by-file-by-lang", "-q", "--csv", shQuote(x)), stdout = TRUE)  
-  out <- out[2:(length(out)-4)]
-  out[1] <- "language,filename,blank,comment,code"
-  df <- read_csv(out)
+  out <- system2("cloc", c("--include-lang=R", "--by-file-by-lang", "-q", "--csv", shQuote(x)), stdout = TRUE)  
+  out_r <- out[startsWith(out, "R,")]
+  out_csv <- c("language,filename,blank,comment,code", out_r)
+  df <- read_csv(out_csv)
 }) %>%
@@
 ```{r add sloc to the kernels_supported set}
-kernels_supported <- left_join(kernels_supported, sloc, by=c("id"="kernel"))
-```
-
-```{r kernels with no R code or duplication}
-kernels_duplicated <- filter(kernels_supported, is.na(code) | code == 0)
-kernels_runnable <- filter(kernels_supported, code > 0)
+kernels_supported <- 
+  kernels_supported %>%
+  left_join(sloc, by=c("id"="kernel")) %>%
+  mutate(runnable=ifelse(is.na(code), FALSE, code > 0))
 ```
@@
 ```{r display duplicated or empty kernels}
-kernels_duplicated %>% 
+filter(kernels_supported, !runnable) %>% 
   mutate(
@@
   ) %>%
-  select(id, competition, path, run_file, language, code_file) %>%
+  select(id, competition, path, code, run_file, language, code_file) %>%
   my_datatable(escape=FALSE)
@@
 ```{r}
-write_csv(kernels_runnable, KERNELS_FILE)
+kernels_runnable <- filter(kernels_supported, runnable)
+write_csv(kernels_supported, KERNELS_FILE)
 write_lines(dirname(kernels_runnable$run_file), SCRIPTS_FILE)
+```
+
+```{r}
+count(kernels_supported, runnable)
 ```
@@
 
-```{r run}
-Sys.unsetenv("R_HOME")
-run <- system2(
-  "parallel",
-  c(
-    "-a", SCRIPTS_FILE,
+Run the following command from the `r params$base_dir` directory:
+
+```{r}
+cat("cd", shQuote(params$base_dir))
+
+cat("parallel",
+    "-a", shQuote(SCRIPTS_FILE),
     "--bar",
     "--jobs", parallel::detectCores(),
-    "--results", PARALLEL_LOG_FILE,
-    "--tagstring", "{/}",
+    "--results", shQuote(PARALLEL_LOG_FILE),
+    "--tagstring", shQuote("{/}"),
     "--timeout", "1h",
-    "--workdir", "{1}/",
-    file.path(R_DIR, "bin", "R"), "CMD", "BATCH", "{1}/run.R", "{1}/run.R.out"
+    "--workdir", shQuote("{1}/"),
+    path(R_DIR, "bin", "R"), "CMD", "BATCH", shQuote("{1}/run.R"), shQuote("{1}/run.R.out"), 
+    "\n"
 )
-)
+
+cat("runr/inst/merge-fst.R", "run/kaggle-run", "calls.fst", "ERROR.fst", "\n")
 ```
-
-```{r load results}
-calls <- kernels_runnable %>%
-  do({
-    bd <- basename(.$run_file)
-    tryCatch({
-      df <- as_tibble(read_fst(path(bd, "calls.fst")))
-      add_column(df, id=.$id, .before=1)
-    }, error=function(e) {
-      tibble(id=.$id, error=e$message)
-    })
-  })
-```
-
-```{r}
-parallel_log <- 
-  read_csv(PARALLEL_LOG_FILE)
-
-parallel_log %>%
-  mutate(
-    output=show_url(path(V1, "run.R.out"))
-  ) %>%
-  my_datatable(escape=FALSE)
-```
-
-## Errors

### 2020-08-23 inst/analysis/inc/paths.R

@@ -5,7 +5,11 @@
 DATA_DIR  <- path(params$base_dir, "revalstudy", "inst", "data")
 PAPER_DIR <- path(params$base_dir, "revalstudy", "inst", "paper")
 RUN_DIR   <- path(params$base_dir, "run")
+PLOT_DIR  <- path(PAPER_DIR, "imgs")
 TAGS_DIR  <- path(PAPER_DIR, "tags")
+
+if (!dir_exists(PLOT_DIR)) dir_create(PLOT_DIR)
+if (!dir_exists(TAGS_DIR)) dir_create(TAGS_DIR)
 
 R_DIR      <- path(params$base_dir, "R-4.0.2")
 R_LIBS_DIR <- path(params$base_dir, "library", "4.0")
@@ -26,12 +30,25 @@ PACKAGE_TRACE_LOG_FILE     <- path(DATA_DIR, "parallel-trace.fst")
 KAGGLE_KERNELS_FILE <- path(DATA_DIR, "kaggle-kernels.fst")
 KAGGLE_LOG_FILE     <- path(DATA_DIR, "parallel-kaggle.fst")
 
+RUN_TRACE_DIR  <- path(RUN_DIR, "package-evals-traced")
+RUN_KAGGLE_DIR <- path(RUN_DIR, "kaggle-run")
+
+CALLS_FINAL_FILE        <- path(RUN_TRACE_DIR, "calls-final.fst")
+KAGGLE_CALLS_FINAL_FILE <- path(RUN_KAGGLE_DIR, "calls-final.fst")
+
 ## OUTPUTs
 
 # TODO: rename to corpus-stage1.*
 CORPUS_S1_FILE         <- path(DATA_DIR, "corpus-1000.txt")
 CORPUS_S1_DETAILS_FILE <- path(DATA_DIR, "corpus-1000.fst")
 CORPUS_FILE            <- path(DATA_DIR, "corpus.txt")
+CORPUS_DETAILS_FILE    <- path(DATA_DIR, "corpus.fst")
+CORPUS_PLOT            <- path(PLOT_DIR, "corpus.pdf")
+
+EVALS_RAW_FILE        <- path(RUN_TRACE_DIR, "raws.fst")
+EVALS_SUM_CORE_FILE   <- path(RUN_TRACE_DIR, "summarized-core.fst")
+EVALS_SUM_PKGS_FILE   <- path(RUN_TRACE_DIR, "summarized-packages.fst")
[…]

### 2020-08-23 inst/analysis/kaggle.R

@@ -1,0 +1,143 @@
+#!/usr/bin/env Rscript
+
+for (x in c(
+  "dplyr", 
+  "DT", 
+  "fs",
+  "fst",
+  "knitr",
+  "purrr",
+  "pbapply",
+  "readr",
+  "rjson",
+  "rmarkdown",
+  "runr",
+  "stringr",
+  "tibble"
+)) {
+  suppressPackageStartupMessages(library(x, character.only=TRUE))
+}
+
+pboptions(type="txt")
+
+# simulate from Rmd
+params <- list(
+  base_dir=commandArgs(trailingOnly=TRUE)[1],
+  competitions=c("titanic")
+)
+
+source("inc/paths.R")
+
+KAGGLE_KERNELS_DIR  <- path(RUN_DIR, "kaggle-kernels", "notebooks", "r", "kernels")
+KAGGLE_DATASETS_DIR <- path(RUN_DIR, "kaggle-datasets")
+KAGGLE_RUN_DIR      <- path(RUN_DIR, "kaggle-run")
+
+KERNELS_FILE      <- path(KAGGLE_RUN_DIR, "kernels.csv")
+SCRIPTS_FILE      <- path(KAGGLE_RUN_DIR, "scripts.txt")
+PARALLEL_LOG_FILE <- path(KAGGLE_RUN_DIR, "parallel.log")
+
+cat("- kaggle kernels:", KAGGLE_KERNELS_DIR, "\n")
+cat("- kaggle datasets:", KAGGLE_DATASETS_DIR, "\n")
+
+stopifnot(dir_exists(KAGGLE_KERNELS_DIR))
+stopifnot(dir_exists(path(KAGGLE_DATASETS_DIR, params$competitions)))
+
+wrap <- function(file, body) {
+  str_glue(
+    "evil::write_eval_traces(",
+    "  evil::trace_eval(",
+    "    quote = TRUE,",
+    "    code = {{",
+    "      {gsub('\n', '\n      ', body)}",
+    "    }}",
+    "  ),",
+    "  datadir=getwd()",
+    ")",
+    .sep = "\n"
+  )
+}
+
+extract_kernel_code <- function(path, code_file, competition) {
+  run_path <- path(KAGGLE_RUN_DIR, competition, basename(path))
+  code_file <- path(path, "script", code_file)
+  run_code_file <- path(run_path, "run.R")
+  
+  if (!dir_exists(run_path)) dir_create(run_path)
+  
[…]

### 2020-08-23 inst/analysis/corpus-analysis.Rmd
First paragraph in 5

--- 2020-08-21 inst/analysis/corpus-analysis.Rmd
@@
 title: "Corpus Analysis"
-output: 
-  rmdformats::readthedown:
-        code_folding: hide
-        lightbox: true
-        gallery: false
-        df_print: paged
-        toc_depth: 3
-date: "`r Sys.Date()`"
+output: html_document
+params:
+  base_dir: ../../../
 editor_options: 
   chunk_output_type: console
-params:
-knit: (function(inputFile, encoding) {
-  rmarkdown::render(inputFile, encoding = encoding, output_dir = "html") })
 ---
+
+```{r setup, include=FALSE}
+for (x in c(
+  "dplyr", 
+  "DT",
+  "fs",
+  "fst",
+  "ggplot2",
+  "ggthemes",
+  "lubridate",
+  "readr",
+  "purrr",
+  "stringr",
+  "tidyr"
+  )) {
+  suppressPackageStartupMessages(library(x, character.only=TRUE))
+}
+
+knitr::opts_chunk$set(
+  echo = TRUE,
+  fig.align = "center",
+  fig.retina = 2,
+  fig.width = 10
+)
+
+# TODO: move into the package
+source("inc/paths.R")
+source("inc/setup.R")
+source("inc/latextags.R")
+
+options(repos="https://cran.r-project.org")
+theme_set(theme_minimal())
+create_tags(path(TAGS_DIR, "corpus.tex"), prefix="Corpus", default=TRUE)
+```
+
+## Load data
+
+```{r}
+corpus <- as_tibble(read_fst(CORPUS_DETAILS_FILE))
+```
+
+## Corpus Overview
+
+### Table
+
+```{r summary}
+overview_table(
+  r("packages",             corpus),
+  r("runnable code",        sum(corpus$runnable_code, na.rm=T)),
+  r("examples code",        sum(corpus$runnable_code_examples, na.rm=T)),
+  r("tests code",           sum(corpus$runnable_code_tests, na.rm=T)),
+  r("vignettes code",       sum(corpus$runnable_code_vignettes , na.rm=T)),
+  r("mean expr coverage",   percent(mean(corpus$coverage_expr))),
+  r("median expr coverage", percent(median(corpus$coverage_expr))),
+  r("with R code",          filter(corpus, package_r_code > 0)),
+  r("R code",               sum(corpus$package_r_code, na.rm=T)),
+  r("with native code",     filter(corpus, package_native_code > 0)),
+  r("native code",          sum(corpus$package_native_code, na.rm=T)),
+  r("functions",            sum(corpus$funs_public+corpus$funs_private, na.rm=T)),
+  r("public functions",     sum(corpus$funs_public, na.rm=T)),
+  r("private functions",    sum(corpus$funs_private, na.rm=T)),
+  r("ranked in cranlogs",   filter(corpus, !is.na(rank))),
+  r("mean revdes",          mean(corpus$revdeps)),
+  r("median revdes",        median(corpus$revdeps)),
+  r("runtime",              as.duration(sum(corpus$runnable_time))),
+  r("mean runtime",         as.duration(mean(corpus$runnable_time))),
+  r("median runtime",       as.duration(median(corpus$runnable_time))),
+  r("function with eval",   sum(corpus$funs_with_eval, na.rm=T))
+)
+```
+
+### CRAN
+
+```{r}
+overview_table(
+  r("all cran", dim(available.packages())[1])
+)
+```
+
+### Plots
+
+```{r plot corpus}
+coverage_mean <- mean(corpus$coverage_expr)
+coverage_median <- median(corpus$coverage_expr)
+revdeps_mean <- mean(corpus$revdeps)
+revdeps_median <- median(corpus$revdeps)
+
+corpus %>%
+  mutate(
+    outlier=package_code > PACKAGE_SIZE_OUTLIER,
+    label=ifelse(rank <= 25 | outlier, package, NA)  #if_else(outlier, str_glue("{package}"), as.character(NA))
+  ) %>%
+  ggplot(
+    aes(
+      x=coverage_expr, 
+      y=revdeps,
+      label=label,
+      size=package_code
+    )
+  ) +
+  geom_point(alpha=.6) +
+  geom_text(size=2, check_overlap = T, vjust=1.5, na.rm = TRUE) +
+
+  geom_hline(aes(yintercept=revdeps_mean), linetype="dashed", color="red", size=0.2) +
+  geom_vline(aes(xintercept=coverage_mean), linetype="dashed", color="red", size=0.2) +
+
+  scale_x_continuous(labels = scales::percent) + 
+  scale_y_log10(labels = scales::comma) +
+  scale_size(range = c(.1, 3), labels = scales::comma) +
+  theme(
+    legend.position=c(0.2, 0.85),
+    legend.box="horizontal",
+    legend.box.background = element_rect(fill="white", size=0.1)
+  ) +
+  labs(
+    x="Code coverage", 
+    y="Number of reverse dependencies (log)", 
+    size="Code size (R + native)",
+    linetype=""
+  ) +
+  guides(
+    size=guide_legend(order=1),
+    color=guide_legend(order=2),
+    linetype=F
+  )
+
+ggsave(CORPUS_PLOT)
+```
+
+### List of corpus packages
+
+```{r list of packages, eval=FALSE}
+datatable(
+  corpus, 
+  filter="top", 
+  options=list(
+    pageLength=50,
+    autoWidth=TRUE
+  )
+)
+```
+
+## Kaggle
+
+```{r, eval=FALSE}
+parallel_log <- 
+  read_csv(PARALLEL_LOG_FILE)
+
+parallel_log %>%
+  mutate(
+    output=show_url(path(V1, "run.R.out"))
+  ) %>%
+  my_datatable(escape=FALSE)
+```

### 2020-08-23 inst/paper/main.tex
First paragraph in 5

@@
 \usepackage{xspace}
+\usepackage{url}
 
 \input{counts.tex}
+\input{tag/corpus.tex}
 
@@
 \maketitle
+
+\newcommand{\eg}{\emph{e.g.},\xspace}
+\newcommand{\ie}{\emph{i.e.},\xspace}
+\newcommand{\cf}{\emph{cf.}\xspace}
 
@@
 
+We have selected \CorpusPackages packages consisting of \CorpusRCodeRnd lines of
+R code and \CorpusNativeCodeRnd lines of native code (C/C++/Fortran).
+Figure~\ref{fig:corpus} shows these packages, the size of the dots reflects the
+project's size in lines of code including both R and native code\footnote{Lines
+  of source code reported excludes comments and blank lines, counted by
+  \emph{cloc}, \cf \url{https://github.com/AlDanial/cloc}}, the x-axis indicates
+the expression code coverage in percents and the y-axis gives the number of
+reverse dependencies in log scale. Dotted lines indicate means. Large and
+popular packages are annotated.
+
+\begin{figure*}[!tb]\centering\includegraphics[width=.9\linewidth]
+  {corpus.pdf}\caption{Corpus}\label{fig:corpus}
+\end{figure*}
+
+These packages come from the Comprehensive R Archive Network
+(CRAN\footnote{\url{http://cran.r-project.org}}), the largest repository of R
+code with over \CorpusAllCranRnd packages\footnote{CRAN receives about 6 new
+  package submissions a day~\cite{Ligges2017}}. Unlike other open code
+repositories such as GitHub, CRAN is a curated repository. Each submitted
+package must abide to a number of well-formedness rules that are automatically
+checked asserting certain quality. Most relevant for this work is that all of
+the runnable code (including code snippets from examples and vignettes) is
+tested and only a successfully running package is admitted in the archive.
+
+% FILIP: WIP
+% We have downloaded and installed all available CRAN packages. Out of the
+% \AllCranRnd packages, we managed to install \AllLoadableRnd. The main reason for
+% this is that \rdt is based on GNU-R 3.5.0 and some of the packages are no longer
+% compatible with this version. Some packages also require extra native
+% dependencies which were not present on our servers. We defined two criteria for
+% including a package into the corpus:
+% \begin{inparaenum}[(1)]
+%   \item have a runnable code that covers a significant part of the package
+%   source code from which type signatures could be inferred, and
+%   \item have some reverse dependencies that will allows us to evaluate the
+%   inferred types using the runnable code from these dependencies.
+% \end{inparaenum}
+% The concrete thresholds used were: at least \ThresholdCodeCoverage of expression
+% coverage and at minimum \ThresholdRevdeps reverse dependencies. The code
+% coverage was computed for each package using
+% \covr\footnote{\url{https://github.com/r-lib/covr}}, the R code coverage R tool.
+% The reverse package dependencies were extracted from the package metadata using
+% builtin function.
+
+% The \CorpusLoadable selected packages contain \CorpusRunnableCodeRnd lines of
+% runnable code in examples (\CorpusExamplesCodeRnd), tests (\CorpusTestsCodeRnd)
+% and vignettes (\CorpusVignettesCodeRnd). Running this code results on average in
+% \CorpusMeanExprCoverage package code coverage (CRAN average is
+% \AllMeanExprCoverage). Together, there is \CorpusRevdesRnd (on average
+% \CorpusMeanRevdes, median \CorpusMedianRevdes; CRAN average is \AllMeanRevdes,
+% median \AllMedianRevdes) reverse dependencies with \CorpusRevdepRunnableCodeRnd
+% runnable lines of code resulting in \CorpusRevdepMeanCodeCoverage coverage (on
+% average) of the corpus packages.
+% Together there are \CorpusFunctionsRnd defined R functions
+% (\CorpusPublicFunctionsRnd are from packages' public API).
+% \CorpusSThreeFunctionsRnd are S3 functions. Packages in the corpus define
+% \CorpusSThreeClasses S3 classes.
+% The code coverage was computed for each package using
+% \covr\footnote{\url{https://github.com/r-lib/covr}}, the R code coverage R
+% tool.  The reverse package dependencies were extracted from the package
+% metadata using builtin function.
+
 \section{Threats to Validity}

### 2020-08-23 inst/analysis/inc/paths.R

@@ -5,8 +5,8 @@
 DATA_DIR  <- path(params$base_dir, "revalstudy", "inst", "data")
 PAPER_DIR <- path(params$base_dir, "revalstudy", "inst", "paper")
 RUN_DIR   <- path(params$base_dir, "run")
-PLOT_DIR  <- path(PAPER_DIR, "imgs")
-TAGS_DIR  <- path(PAPER_DIR, "tags")
+PLOT_DIR  <- path(PAPER_DIR, "img")
+TAGS_DIR  <- path(PAPER_DIR, "tag")
 
 if (!dir_exists(PLOT_DIR)) dir_create(PLOT_DIR)
 if (!dir_exists(TAGS_DIR)) dir_create(TAGS_DIR)
@@ -63,4 +63,4 @@ WHICH_DEPENDENCIES <- c("Depends", "Imports")
 # local cran mirror
 CRAN_MIRROR_LOCAL_URL <- paste0("file://", params$base_dir, "/CRAN")
 # size of a package to be considered as outlier
-PACKAGE_SIZE_OUTLIER <- 5000
+PACKAGE_SIZE_OUTLIER <- 50000

### 2020-08-23 inst/paper/main.tex
Fix table

@@
 total calls to all R functions. This is merely \missingPercentage of the total
-number of calls. \missingTable breaks down the number of calls and callsites for
-the four kinds of eval functions from the three source; core R, packages and
-Kaggle code.
-
-\begin{tabular}{c|c|c|c|c|c}
-  Function  & Core & Packages & Kaggle & All \\
-  eval & \CoreEvalCallCountRnd & \PackageEvalCallCountRnd & \KaggleEvalCallCountRnd & \AllEvalCallCountRnd \\
-  evalParent & \CoreEvalParentCallCountRnd & \PackageEvalParentCallCountRnd & \KaggleEvalParentCallCountRnd & \AllEvalParentCallCountRnd \\
-  evalq & \CoreEvalqCallCountRnd & \PackageEvalqCallCountRnd & \KaggleEvalqCallCountRnd & \AllEvalqCallCountRnd \\
-  local & \CoreLocalCallCountRnd & \PackageLocalCallCountRnd & \KaggleLocalCallCountRnd & \AllLocalCallCountRnd \\
-  all & \CoreAllCallCountRnd & \PackageAllCallCountRnd & \KaggleAllCallCountRnd & \AllAllCallCountRnd
-\end{tabular}
-
-\begin{tabular}{c|c|c|c|c|c}
-  Function  & Core & Packages & Kaggle & All \\
-  eval & \CoreEvalSiteCountRnd & \PackageEvalSiteCountRnd & \KaggleEvalSiteCountRnd & \AllEvalSiteCountRnd \\
-  evalParent & \CoreEvalParentSiteCountRnd & \PackageEvalParentSiteCountRnd & \KaggleEvalParentSiteCountRnd & \AllEvalParentSiteCountRnd \\
-  evalq & \CoreEvalqSiteCountRnd & \PackageEvalqSiteCountRnd & \KaggleEvalqSiteCountRnd & \AllEvalqSiteCountRnd \\
-  local & \CoreLocalSiteCountRnd & \PackageLocalSiteCountRnd & \KaggleLocalSiteCountRnd & \AllLocalSiteCountRnd \\
-  all & \CoreAllSiteCountRnd & \PackageAllSiteCountRnd & \KaggleAllSiteCountRnd & \AllAllSiteCountRnd
-\end{tabular}
-
+number of calls. Table \ref{table:eval-count-summary} shows the number of
+calls and callsites for the four kinds of eval functions from core and packages.
 The core R packages are the biggest consumers of eval. Majority of calls,
@@
 
-Of the \missingNumber packages in our corpus, only \missingNumber packages call
-eval. \missingTable shows the number of calls to eval made by the ten most
-frequent callers to eval. \missingNumber packages account for over
-\missingPercentage calls to eval.
+\begin{table}[ht]
+  \resizebox{\columnwidth}{!}{%
+    \begin{tabular}{l|rr|r||rr|r}
+      \hline
+      \multicolumn{1}{c}{Function}& \multicolumn{3}{c}{Calls} & \multicolumn{3}{c}{Callsites}\\
+      \cline{1-1} \cline{2-4} \cline{5-7}
+                                  & Core & Packages & Total & Core & Packages & Total\\
+      \hline
+      \c{eval} & \CoreEvalCallCountRnd & \PackageEvalCallCountRnd  & \AllEvalCallCountRnd & \CoreEvalSiteCountRnd & \PackageEvalSiteCountRnd &  \AllEvalSiteCountRnd\\
+      \c{eval.parent} & \CoreEvalParentCallCountRnd & \PackageEvalParentCallCountRnd  & \AllEvalParentCallCountRnd & \CoreEvalParentSiteCountRnd & \PackageEvalParentSiteCountRnd & \AllEvalParentSiteCountRnd\\
+      \c{evalq} & \CoreEvalqCallCountRnd & \PackageEvalqCallCountRnd  & \AllEvalqCallCountRnd & \CoreEvalqSiteCountRnd & \PackageEvalqSiteCountRnd & \AllEvalqSiteCountRnd \\
+      \c{local} & \CoreLocalCallCountRnd & \PackageLocalCallCountRnd  & \AllLocalCallCountRnd & \CoreLocalSiteCountRnd & \PackageLocalSiteCountRnd & \AllLocalSiteCountRnd \\
+      \hline
+      Total & \CoreAllCallCountRnd & \PackageAllCallCountRnd  & \AllAllCallCountRnd & \CoreAllSiteCountRnd & \PackageAllSiteCountRnd & \AllAllSiteCountRnd \\
+      \hline
+    \end{tabular}
+  }
+  \label{table:eval-count-summary}
+  \caption{Distribution of eval calls and callsites in the corpus}
+\end{table}
 

### 2020-08-23 inst/paper/main.tex
Add package counts

@@
 
+Of the \missingNumber packages in our corpus, only \missingNumber packages call
+eval. \ref{table:top-ten-package-summary} shows the number of calls to eval made
+by the ten most frequent callers to eval. These packages together have
+\missingNumber callsites and account for \missingPercentage calls to eval.
+Furthermore, the top ten callsites account for \missingPercentage of package
+eval calls. A single callsite of \TopTenPackageNameA accounts for over
+\missingPercentage calls to eval.
+
+\begin{table}[ht]
+  \resizebox{\columnwidth}{!}{%
+    \begin{tabular}{l|rr||l|rr}
+      \hline
+      \multicolumn{3}{c}{}& \multicolumn{3}{c}{} \\
+      \cline{1-3} \cline{4-6}
+      Package & Callsites & Calls & Package & Callsites & Call\\
+      \hline
+      \TopTenPackageNameA & \TopTenPackageCallsiteCountA & \TopTenPackageCallCountA (\TopTenPackageCallPercA) & \TopTenPackageNameF & \TopTenPackageCallsiteCountF & \TopTenPackageCallCountF (\TopTenPackageCallPercF) \\
+      \TopTenPackageNameB & \TopTenPackageCallsiteCountB & \TopTenPackageCallCountB (\TopTenPackageCallPercB) & \TopTenPackageNameG & \TopTenPackageCallsiteCountG & \TopTenPackageCallCountG (\TopTenPackageCallPercG) \\
+      \TopTenPackageNameC & \TopTenPackageCallsiteCountC & \TopTenPackageCallCountC (\TopTenPackageCallPercC) & \TopTenPackageNameH & \TopTenPackageCallsiteCountH & \TopTenPackageCallCountH (\TopTenPackageCallPercH) \\
+      \TopTenPackageNameD & \TopTenPackageCallsiteCountD & \TopTenPackageCallCountD (\TopTenPackageCallPercD) & \TopTenPackageNameI & \TopTenPackageCallsiteCountI & \TopTenPackageCallCountI (\TopTenPackageCallPercI) \\
+      \TopTenPackageNameE & \TopTenPackageCallsiteCountE & \TopTenPackageCallCountE (\TopTenPackageCallPercE) & \TopTenPackageNameJ & \TopTenPackageCallsiteCountJ & \TopTenPackageCallCountJ (\TopTenPackageCallPercJ) \\
+      \hline
+    \end{tabular}
+  }
+  \label{table:top-ten-package-summary}
+  \caption{Distribution of eval calls and callsites in the top ten packages}
+\end{table}
+
 \missingGraph shows the distribution of number of callsites to eval across these

### 2020-08-24 inst/analysis/counts.Rmd
Add tags and call site summary

---
title: "Empirical Evaluation of eval in R"
output:
  rmdformats::readthedown:
        code_folding: hide
        lightbox: true
        gallery: false
        df_print: paged
        toc_depth: 3
date: "`r Sys.Date()`"
editor_options: 
  chunk_output_type: console
params:
    base_dir: /home/aviral/projects/revalstudy2
    kaggle_datafile: "summarized-kaggle.fst"
    core_datafile: "summarized-core.fst"
    packages_datafile: "summarized-packages.fst"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.align = "center",
                      fig.retina = 2,
                      fig.width = 10,
                      cache.lazy = FALSE)

library(tidyverse)
library(stringr)
library(viridis)
library(vctrs)
library(DT)
library(fst)
library(fs)
library(purrr)
library(readr)
library(xtable)

source("inc/paths.R")
source("inc/setup.R")
source("inc/latextags.R")
create_tags(path("../paper/tag", "counts.tex"), prefix="", default=TRUE)
```

```{r eval=TRUE}
core_data <- read_fst(params$core_datafile)
packages_data <- 
    read_fst(params$packages_datafile) %>%
    mutate(eval_source_type = "package")
kaggle_data <- 
    read_fst(params$kaggle_datafile)
    #mutate(eval_source_type == "core")

all_data <- bind_rows(core_data, packages_data, kaggle_data)
all_data <-
    all_data %>%
    mutate(eval_call_srcref = 
          coalesce(eval_call_srcref, 
                   str_c("missing", "::", caller_package, "::", caller_function)))

count_summary <-
    all_data %>%
    group_by(eval_source_type, eval_function) %>%
    summarize(call_count = sum(nb_ev_calls),
              site_count = length(unique(eval_call_srcref))) %>%
    ungroup() %>%
    add_row(eval_source_type = "kaggle", eval_function = "eval", call_count = 0, site_count = 0) %>%
    add_row(eval_source_type = "kaggle", eval_function = "eval.parent", call_count = 0, site_count = 0) %>%
    add_row(eval_source_type = "kaggle", eval_function = "evalq", call_count = 0, site_count = 0) %>%
    add_row(eval_source_type = "kaggle", eval_function = "local", call_count = 0, site_count = 0)

eval_source_type_all_summary <-
    count_summary %>%
    group_by(eval_source_type) %>%
    summarize(call_count = sum(call_count),
              site_count = sum(site_count)) %>%
    ungroup() %>%
    mutate(eval_function = "all")

all_eval_function_summary <-
    count_summary %>%
    group_by(eval_function) %>%
    summarize(call_count = sum(call_count),
              site_count = sum(site_count)) %>%
    ungroup() %>%
    mutate(eval_source_type = "all")

all_all_summary <- tibble(
    eval_source_type = "all",
    eval_function = "all",
    call_count = sum(count_summary$call_count),
    site_count = sum(count_summary$site_count)
)

count_summary <-
    bind_rows(count_summary, 
              eval_source_type_all_summary, 
              all_eval_function_summary, 
              all_all_summary)

print(count_summary)

count_summary %>%
pmap_dfr(function(eval_source_type, eval_function, call_count, site_count) {
    eval_function <- str_replace(eval_function, "\\.", " ")
    name <- str_c(eval_source_type, " ", eval_function, " ", "call count")
    r(name, call_count)
    name <- str_c(eval_source_type, " ", eval_function, " ", "site count")
    r(name, site_count)
})
```

```{r run-data, eval=TRUE}
run_eval_proportion <-
    all_data %>%
    group_by(file, eval_source_type) %>%
    summarize(eval_count = sum(nb_ev_calls)) %>%
    ungroup() %>%
    pivot_wider(names_from = "eval_source_type", values_from = "eval_count", values_fill = 0) %>%
    mutate(all = core + package)

total_file_count <- length(unique(run_eval_proportion$file))

no_eval_file_count <-
    run_eval_proportion %>%
    filter(core == 0 & package == 0)

core_eval_file_count <-
    run_eval_proportion %>%
    filter(core != 0 & package == 0)

package_eval_file_count <-
    run_eval_proportion %>%
    filter(core == 0 & package != 0)

all_eval_file_count <-
    run_eval_proportion %>%
    filter(core != 0 & package != 0)

r("total file count", total_file_count)
r("no eval file count", no_eval_file_count)
r("no eval file perc", ratio(no_eval_file_count, total_file_count))
r("core eval file count", core_eval_file_count)
r("core eval file perc", ratio(core_eval_file_count, total_file_count))
r("package eval file count", package_eval_file_count)
r("package eval file perc", ratio(package_eval_file_count, total_file_count))
r("all eval file count", all_eval_file_count)
r("all eval file perc", ratio(all_eval_file_count, total_file_count))

run_eval_proportion <-
    run_eval_proportion %>%
    filter(core != 0 & package != 0) %>%
    mutate(core_perc = round(core * 100 / all)) %>%
    group_by(core_perc) %>%
    summarize(file_count = n()) %>%
    ungroup() %>%
    arrange(desc(core_perc)) %>%
    mutate(file_perc = round(100 * file_count / sum(file_count), 1)) %>%
    mutate(cumsum_file_perc = cumsum(file_perc)) %>%
    mutate(cumsum_file_count = cumsum(file_count))

eighty_core_eval_file_count <-
    run_eval_proportion %>%
    filter(core_perc == 80) %>%
    pull(cumsum_file_count)

r("eighty core eval file count", eighty_core_eval_file_count)
r("eighty core eval file perc", ratio(eighty_core_eval_file_count, all_eval_file_count))

run_eval_proportion
```

```{r package-counts, eval=TRUE}
package_calls <-
    packages_data %>%
    filter(eval_source != "base?") %>%
    group_by(eval_source) %>%
    summarize(site_count = length(unique(eval_call_srcref)), 
              call_count = sum(nb_ev_calls)) %>%
    ungroup() %>%
    arrange(desc(call_count)) %>%
    mutate(call_perc = round(100 * call_count / sum(call_count), 1)) %>%
    mutate(cumsum_call_perc = cumsum(call_perc))

top_ten_package_calls <-
    package_calls %>%
    slice(n = 1:10)

datatable(top_ten_package_calls)

top_ten_package_call_count <- sum(top_ten_package_calls$call_count)
top_ten_package_site_count <- sum(top_ten_package_calls$site_count)
total_package_call_count <- sum(package_calls$call_count)

r("top ten package call count", top_ten_package_call_count)
r("top ten package call perc", ratio(top_ten_package_call_count, total_package_call_count))

r("top ten package site count", top_ten_package_site_count)
r("top ten package site perc", ratio(top_ten_package_site_count, sum(package_calls$site_count)))

top_ten_package_calls %>%
add_column(index = toupper(letters)[1:nrow(top_ten_package_calls)]) %>%
pmap_dfr(function(eval_source, site_count, call_count, call_perc, cumsum_call_perc, index) {
    r(str_c("top ten package name ", index), eval_source)
    r(str_c("top ten package callsite count ", index), site_count)
    r(str_c("top ten package call count ", index), call_count)
    r(str_c("top ten package call perc ", index), ratio(call_count, total_package_call_count))
})
```

```{r callsite-distribution, eval=TRUE}
```{r call-site-table}
eval_callsite_table <-
    package_calls %>%
    mutate(site_count = if_else(site_count <= 10, as.character(site_count), 
                        if_else(site_count <= 50, "11--50",
                        if_else(site_count <= 100, "51--100",
                        if_else(site_count <= 150, "101--150",
                        if_else(site_count <= 200, "151--200",
                        "> 200")))))) %>%
    group_by(site_count) %>%
    summarize(package_count = n()) %>%
    ungroup() %>%
    arrange(match(site_count, c(1:10,"11-50", "51-100", "101-150", "151-200")))

datatable(eval_callsite_table)

eval_callsite_table %>%
add_column(index = toupper(letters)[1:nrow(eval_callsite_table)]) %>%
pmap_dfr(function(site_count, package_count, index) {
    r(str_c("site summary site count ", index), site_count)
    r(str_c("site summary package count ", index), package_count)
})
```

### 2020-08-24 inst/paper/main.tex
Add tags and call site summary

--- 2020-08-23 inst/paper/main.tex
@@
 
-\input{counts.tex}
+\input{tag/counts.tex}
 \input{tag/corpus.tex}
@@
 
+Of the \missingNumber packages in our corpus, only \missingNumber packages call
+eval. \ref{table:top-ten-package-summary} shows the number of calls to eval made
+by the ten most frequent callers to eval. These packages together have
+\missingNumber callsites and account for \missingPercentage calls to eval.
+Furthermore, the top ten callsites account for \missingPercentage of package
+eval calls. A single callsite of \TopTenPackageNameA accounts for over
+\missingPercentage calls to eval.
+
+\begin{table}[ht]
+  \resizebox{\columnwidth}{!}{%
+    \begin{tabular}{l|rrr||l|rrr}
+      \hline
+      \multicolumn{4}{c}{}& \multicolumn{4}{c}{} \\
+      \cline{1-4} \cline{5-8}
+      Package & Callsites & Calls & Calls \% & Package & Callsites & Calls & Calls \%\\
+      \hline
+      \TopTenPackageNameA & \TopTenPackageCallsiteCountA & \TopTenPackageCallCountA &\TopTenPackageCallPercA & \TopTenPackageNameF & \TopTenPackageCallsiteCountF & \TopTenPackageCallCountF & \TopTenPackageCallPercF \\
+      \TopTenPackageNameB & \TopTenPackageCallsiteCountB & \TopTenPackageCallCountB &\TopTenPackageCallPercB & \TopTenPackageNameG & \TopTenPackageCallsiteCountG & \TopTenPackageCallCountG & \TopTenPackageCallPercG \\
+      \TopTenPackageNameC & \TopTenPackageCallsiteCountC & \TopTenPackageCallCountC &\TopTenPackageCallPercC & \TopTenPackageNameH & \TopTenPackageCallsiteCountH & \TopTenPackageCallCountH & \TopTenPackageCallPercH \\
+      \TopTenPackageNameD & \TopTenPackageCallsiteCountD & \TopTenPackageCallCountD &\TopTenPackageCallPercD & \TopTenPackageNameI & \TopTenPackageCallsiteCountI & \TopTenPackageCallCountI & \TopTenPackageCallPercI \\
+      \TopTenPackageNameE & \TopTenPackageCallsiteCountE & \TopTenPackageCallCountE &\TopTenPackageCallPercE & \TopTenPackageNameJ & \TopTenPackageCallsiteCountJ & \TopTenPackageCallCountJ & \TopTenPackageCallPercJ \\
+      \hline
+    \end{tabular}
+  }
+  \label{table:top-ten-package-summary}
+  \caption{Distribution of eval calls and callsites in the top ten packages}
+\end{table}
+
 \missingGraph shows the distribution of number of callsites to eval across these
 packages. \missingNumber packages have a single eval callsite.
+
+\begin{table}[ht]
+  \resizebox{\columnwidth}{!}{%
+    \begin{tabular}{rr|rr|rr}
+      \hline
+      Callsites & Packages & Callsites & Packages & Callsites & Packages \\
+      \hline
+      \SiteSummarySiteCountA & \SiteSummaryPackageCountA & \SiteSummarySiteCountF & \SiteSummaryPackageCountF & \SiteSummarySiteCountK & \SiteSummaryPackageCountK \\
+      \SiteSummarySiteCountB & \SiteSummaryPackageCountB & \SiteSummarySiteCountG & \SiteSummaryPackageCountG & \SiteSummarySiteCountL & \SiteSummaryPackageCountL \\
+      \SiteSummarySiteCountC & \SiteSummaryPackageCountC & \SiteSummarySiteCountH & \SiteSummaryPackageCountH & \SiteSummarySiteCountM & \SiteSummaryPackageCountM \\
+      \SiteSummarySiteCountD & \SiteSummaryPackageCountD & \SiteSummarySiteCountI & \SiteSummaryPackageCountI & \SiteSummarySiteCountN & \SiteSummaryPackageCountN \\
+      \SiteSummarySiteCountE & \SiteSummaryPackageCountE & \SiteSummarySiteCountJ & \SiteSummaryPackageCountJ & \SiteSummarySiteCountO & \SiteSummaryPackageCountO \\ 
+      \hline
+    \end{tabular}
+  }
+  \label{table:site-package-summary}
+  \caption{Distribution of callsites across packages}
+\end{table}
 

### 2020-08-24 inst/paper/main.tex
wip

@@
 
-We have selected \CorpusPackages packages consisting of \CorpusRCodeRnd lines of
-R code and \CorpusNativeCodeRnd lines of native code (C/C++/Fortran).
-Figure~\ref{fig:corpus} shows these packages, the size of the dots reflects the
-project's size in lines of code including both R and native code\footnote{Lines
-  of source code reported excludes comments and blank lines, counted by
-  \emph{cloc}, \cf \url{https://github.com/AlDanial/cloc}}, the x-axis indicates
-the expression code coverage in percents and the y-axis gives the number of
+We have selected \CorpusPackages packages consisting of \CorpusRCodeRnd
+lines of R code and \CorpusNativeCodeRnd lines of native code
+(C/C++/Fortran).  Figure~\ref{fig:corpus} shows these packages, the size of
+the dots reflects the project's size in lines of code including both R and
+native code\footnote{Lines of source code reported excludes comments and
+  blank lines, counted by \emph{cloc}, \cf
+  \url{https://github.com/AlDanial/cloc}}, the x-axis indicates the
+expression code coverage in percents and the y-axis gives the number of
 reverse dependencies in log scale. Dotted lines indicate means. Large and
@@
 
-\section{Threats to Validity}
+\section{Usage Metrics}
 
-\section{Research Question}
-This section presents the results of our empirical study of eval in the R language.
-\subsection{How many evals}
+This section presents an overiew of the usage of \eval in our corpus.
+Previous work reported that 82\% of JavaScript-enabled web pages used
+\eval~\cite{ecoop11}, the case for R is more striking at \miss{100\%} of the
+programs use that feature.
 
-In our corpus eval is called \AllAllCallCountRnd{} times out of \missingNumber
-total calls to all R functions. This is merely \missingPercentage of the total
-number of calls. Table \ref{table:eval-count-summary} shows the number of
-calls and callsites for the four kinds of eval functions from core and packages.
-The core R packages are the biggest consumers of eval. Majority of calls,
-\CoreAllCallCountRnd, originate from the core R packages. Kaggle code, which is
-end-user code, has merely \KaggleAllCallCountRnd calls to eval.
+This strikingly high number is due to the fact that the language
+implementation, the core packages, use \eval extensively. So, in our
+analysis, we differentiate \eval based on where they originate from. A {\bf
+  core} \eval is one that occurs in the code of a core package.  A {\bf
+  dependent} \eval is one that occurs in a package that was included by the
+target package. An {\bf own} \eval is one that occurs in the source code of
+the target package. We label an \eval occurence as {\bf unknown} when we are
+not able to deterimine its origin.\footnote{One source of unknown evals is
+  when the \eval function passed as an argument to a higher-order function
+  such as \c{lapply}, another is when the code lacks source references.}
+
+Over all, we observed \AllAllCallCountRnd call to \eval in \miss{3,000} runs
+of our corpus. The mean number of calls per run was \miss{20} (with a min of
+\c{1} and a max of \c{10,000}). Table~\ref{A} details the usage of \eval
+across the corpus.  The format is T (m/M/x) where T is the total number of
+calls observed for all runs of the corpus, m and x are the minimum and
+maximum number of calls for any given run, and A is the mean number of calls
+per run.  We can observe \miss{???}. {\bf <The following is a place
+  holder. Real text will come once we have the table.>} The majority of
+calls come from the core packages.  Kaggle does not have any own calls, any
+\eval in those happens in dependent packages and core.
+
+\begin{verbatim}
+Table A
+
+         core      dependent   own
+----------------------------------------
+CRAN   10M?(1/2/3) 1M?(1/2/3) 1K?(1/2/3)
+Kaggle 2K?(1/2/3)  1K?(1/2/3) 0 (0/0/0)
+\end{verbatim}
+
+The following table will be split up:
 
@@
 
+\section{Threats to Validity}
+
 \section{Conclusion}
@@
 \end{document}
-
-Eval is evil, we would like to replace it.
-Research Questions
-How often is eval used? 
-How often it is used in base / other packages?
-blocking: cannot find out precisely what comes from base
-What is the source of evals?
-Provenance (source arguments to eval)
-How many evals are in libraries / vignettes
-What is eval used for?
-??Eval in the tidyverse
-TODO: Classification of evals in  packages (Pierre)
-TODO: look at the first function to get
-binary / unary / assignment / other function
-does it have {
-evals that do nothing (they get values)
-...
-TODO: In the case of base package (Filip)
-access default values of formals
-simply implementation
-weird replacement of calls (write.csv -> write.table)
-accessing variable from another environment (instead of get)
-evaling user input
-Change of caller pattern: 
-m <- match.call()
- m$name <- NULL
- m[[1L]] <- as.name("library")
-eval(m, .GlobalEnv)
-     - 	TODO: How dangerous is eval? (Aviral)
-Does eval generate new code (create new functions)?
-environment writes (reads)
-system environment variables
-connections
-random number generator - setting seed
-Does eval do dynamic code loading?
-Does eval argument reflect on the call stack?
-Calls to C Code
-TODO: Could the eval be made safer?
-TODO: Could some of the eval calls be replaced?

### 2020-08-24 inst/paper/main.tex
wip

@@
 calls come from the core packages.  Kaggle does not have any own calls, any
-\eval in those happens in dependent packages and core.
+\eval in those happens in dependent packages and core.  {\bf <The following
+  will be changed. The file level stuff is less useful than package level.>}
+All the \TotalFileCount files in our corpus call eval. \CoreEvalFileCount
+(\CoreEvalFilePerc) call only core evals, \AllEvalFileCount
+(\AllEvalFilePerc) call both package and core evals; and,
+\PackageEvalFileCount (\PackageEvalFilePerc) call only package evals. Of the
+\AllEvalFileCount (\AllEvalFilePerc) files that call both package and base
+eval, \EightyCoreEvalFilePerc (\EightyCoreEvalFileCount) files make over
+80\% eval calls to core evals. Of the \missingNumber packages in our corpus,
+only \missingNumber packages call eval. \ref{table:top-ten-package-summary}
+shows the number of calls to eval made by the ten most frequent callers to
+eval. These packages together have \missingNumber callsites and account for
+\missingPercentage calls to eval.  Furthermore, the top ten callsites
+account for \missingPercentage of package eval calls. A single callsite of
+\TopTenPackageNameA accounts for over \missingPercentage calls to eval.
 
@@
 
-The following table will be split up:
+The following table will be replaced. {\bf <At this point separating between
+  the different kinds of evals is distracting. We have not argued why it is
+  important.>}
 
@@
 \end{table}
-
-All the \TotalFileCount files in our corpus call eval. \CoreEvalFileCount
-(\CoreEvalFilePerc) call only core evals, \AllEvalFileCount (\AllEvalFilePerc)
-call both package and core evals; and, \PackageEvalFileCount
-(\PackageEvalFilePerc) call only package evals. Of the \AllEvalFileCount
-(\AllEvalFilePerc) files that call both package and base eval,
-\EightyCoreEvalFilePerc (\EightyCoreEvalFileCount) files make over 80\% eval
-calls to core evals.
-
-Of the \missingNumber packages in our corpus, only \missingNumber packages call
-eval. \ref{table:top-ten-package-summary} shows the number of calls to eval made
-by the ten most frequent callers to eval. These packages together have
-\missingNumber callsites and account for \missingPercentage calls to eval.
-Furthermore, the top ten callsites account for \missingPercentage of package
-eval calls. A single callsite of \TopTenPackageNameA accounts for over
-\missingPercentage calls to eval.
 

### 2020-08-24 inst/paper/main.tex
minor

--- 2020-08-24 inst/paper/main.tex
@@
 the dots reflects the project's size in lines of code including both R and
-native code\footnote{Lines of source code reported excludes comments and
-  blank lines, counted by \emph{cloc}, \cf
-  \url{https://github.com/AlDanial/cloc}}, the x-axis indicates the
-expression code coverage in percents and the y-axis gives the number of
-reverse dependencies in log scale. Dotted lines indicate means. Large and
-popular packages are annotated.
+native code\footnote{Excluding comments and blank lines,
+  \url{github.com/AlDanial/cloc}}, the x-axis indicates the expression code
+coverage in percents and the y-axis gives the number of reverse dependencies
+in log scale. Dotted lines indicate means. Large and popular packages are
+annotated.
 
@@
 \eval~\cite{ecoop11}, the case for R is more striking at \miss{100\%} of the
-programs use that feature.
+programs use that feature. In this section, a call to \eval denotes a call
+to any one of the functions that provide \eval-like
+functionality.\footnote{These include \c{eval}, \c{eval.parent}, \c{evalq}
+  and \c{local}.}
 
@@
 the target package. We label an \eval occurence as {\bf unknown} when we are
-not able to deterimine its origin.\footnote{One source of unknown evals is
-  when the \eval function passed as an argument to a higher-order function
-  such as \c{lapply}, another is when the code lacks source references.}
+not able to deterimine its origin.\footnote{A source of unknowns is when the
+  \eval function is passed as an argument to a function such as \c{lapply},
+  another is when the code lacks source references.}
 
@@
 of our corpus. The mean number of calls per run was \miss{20} (with a min of
-\c{1} and a max of \c{10,000}). Table~\ref{A} details the usage of \eval
-across the corpus.  The format is T (m/M/x) where T is the total number of
-calls observed for all runs of the corpus, m and x are the minimum and
-maximum number of calls for any given run, and A is the mean number of calls
-per run.  We can observe \miss{???}. {\bf <The following is a place
+\miss{1} and a max of \miss{10,000}). Table~\ref{A} details the usage of
+\eval across the corpus.  The format is T (m/M/x) where T is the total
+number of calls observed for all runs of the corpus, m and x are the minimum
+and maximum number of calls for any given run, and A is the mean number of
+calls per run.  We can observe \miss{???}. {\bf <The following is a place
   holder. Real text will come once we have the table.>} The majority of
 calls come from the core packages.  Kaggle does not have any own calls, any
-\eval in those happens in dependent packages and core.
+\eval in those happens in dependent packages and core.  Of the
+\missingNumber packages in our corpus, only \missingNumber packages call
+eval.
 
@@
 
-All the \TotalFileCount files in our corpus call eval. \CoreEvalFileCount
-(\CoreEvalFilePerc) call only core evals, \AllEvalFileCount (\AllEvalFilePerc)
-call both package and core evals; and, \PackageEvalFileCount
-(\PackageEvalFilePerc) call only package evals. Of the \AllEvalFileCount
-(\AllEvalFilePerc) files that call both package and base eval,
-\EightyCoreEvalFilePerc (\EightyCoreEvalFileCount) files make over 80\% eval
-calls to core evals.
-
-Of the \missingNumber packages in our corpus, only \missingNumber packages call
-eval. \ref{table:top-ten-package-summary} shows the number of calls to eval made
-by the ten most frequent callers to eval. These packages together have
+\ref{table:top-ten-package-summary} shows the number of calls to eval
+made by the ten most frequent callers to eval. These packages together have
 \missingNumber callsites and account for \missingPercentage calls to eval.

### 2020-08-24 inst/analysis/counts.Rmd
Fix callsite range formatting

@@
                         if_else(site_count <= 200, "151--200",
-                        "> 200")))))) %>%
+                        if_else(site_count <= 250, "201--250",
+                        if_else(site_count <= 300, "251--300",
+                        "> 300")))))))) %>%
     group_by(site_count) %>%
@@
     ungroup() %>%
-    arrange(match(site_count, c(1:10,"11-50", "51-100", "101-150", "151-200")))
+    arrange(match(site_count, c(1:10,"11--50", "51--100", "101--150", "151--200", "201--250", "251--300", "> 300")))
 
@@
 ```
+
+```{r core-dependent-own-eval}
+
+```

### 2020-08-24 inst/paper/main.tex
AST size (on random sample for now)

@@
 
+\subsubsection{The size of \eval}
+
+In Figure~\ref{fig:ast-size}, we look at the the size of the AST, in number of nodes, of the \c(expr) argument of \eval, after it is resolved.
+
+\begin{figure}[htbp]
+	\includegraphics[width=\columnwidth]{ast_sizes}
+	\caption{Size of the \c{expr} argument of \eval in terms of AST nodes, for the \emph{core} dataset and the \emph{packages} dataset.} \label{fig:ast-size}
+\end{figure}
+
+Most of these expressions have very few nodes and about 50\% have only one. It is because \c{expr} is often reduced to just a value or a symbol.
+
+\miss{This is on a random sample of 10000 unique rows of the summarized data set as ST size takes a lof of time to compute. To update when it finishes computing...}
+
 \subsection{How are they used}
@@
 
+\subsubsection{\eval and environments}
+
+One of the main differences of the R \eval with the javascript one is how it interacts with \emph{scope}, or environments in R.
+In Javascript, \c{eval} can access the local scope and the global scope.
+In R, \c{eval} has an argument \c{envir} to indicate in which environment its first argument, after being resolved, must be evaluated.
+
+By default, it is equal to \c{parent.frame()}, which is the parent environment of the \c{eval} call.
+
 \subsection{How to replace evals}
 
-We look at how \emph{polymorphic} the \c{expr} argument of \eval can be, \emph{i.e.} how many different types of the resolved \c{expr} there are per call sites, in Figure~\ref{fig:polymorphism}.
-89\% of the call sites are \emph{monomorphic}.
+We look at how \emph{polymorphic} the \c{expr} argument of \eval can be, \ie how many different types of the resolved \c{expr} there are per call sites, in Figure~\ref{fig:polymorphism}.
+\DegreeMonomorphism of the call sites are \emph{monomorphic}.
 
 \begin{figure}
-    \includegraphics[width=\columnwidth]{polymorphism-1}
+    \includegraphics[width=\columnwidth]{polymorphism}
     \caption{Degree of polymorphism of \c{expr} per call sites.} \label{fig:polymorphism}

### 2020-08-24 inst/analysis/corpus-stage2.Rmd
Update the runnable-code-metadata to include per program data

@@
   filter(language=="R") %>%
-  select(package, path, code, files) %>%
+  mutate(path=case_when(
+    str_starts(filename, "./examples") ~ "examples",
+    str_starts(filename, "./tests") ~ "tests",
+    str_starts(filename, "./vignettes") ~ "vignettes"
+  )) %>%
+  group_by(package, path) %>%
+  summarise(files=n(), code=sum(code)) %>%
   pivot_wider(

### 2020-08-24 inst/analysis/counts.Rmd
Add own, dependent and core eval calls from CRAN and Kaggle

@@
 ```{r core-dependent-own-eval}
+all_data <-
+    bind_rows(core_data, packages_data, kaggle_data)
+    ##filter(is.na(eval_call_srcref))
 
+all_data$package <- map_chr(path_split(all_data$file), function(l) l[[length(l) - 2]])
+
+all_data <-
+    all_data %>%
+    mutate(core_own_dependent = if_else(eval_source == "core", "core",
+                                if_else(eval_source == "base", "core",
+                                if_else(package == caller_package, "own",
+                                        "dependent"))),
+          eval_source)
+
+core_own_dependent <-
+    all_data %>%
+        group_by(file, core_own_dependent) %>%
+        summarize(corpus = first(corpus), eval_calls = sum(nb_ev_calls)) %>%
+        ungroup() %>%
+        group_by(corpus, core_own_dependent) %>%
+        summarize(total_eval_calls = sum(eval_calls),
+                  max_eval_calls = max(eval_calls),
+                  min_eval_calls = min(eval_calls),
+                  avg_eval_calls = mean(eval_calls))
+
+core_own_dependent %>%
+pmap_dfr(function(corpus, core_own_dependent, total_eval_calls, max_eval_calls, min_eval_calls, avg_eval_calls) {
+    r(str_c(corpus, " ", core_own_dependent, " total eval calls"), total_eval_calls)
+    r(str_c(corpus, " ", core_own_dependent, " max eval calls"), max_eval_calls)
+    r(str_c(corpus, " ", core_own_dependent, " min eval calls"), min_eval_calls)
+    r(str_c(corpus, " ", core_own_dependent, " avg eval calls"), avg_eval_calls)
+})
+
+datatable(core_own_dependent)
 ```

### 2020-08-24 inst/analysis/corpus-analysis.Rmd
Regenerated data for V

@@
   "stringr",
-  "tidyr"
+  "tidyr",
+  "viridis"
   )) {
@@
 
-```{r}
-corpus <- as_tibble(read_fst(CORPUS_DETAILS_FILE))
-```
-
-## Corpus Overview
-
-### Table
-
-```{r summary}
-overview_table(
-  r("packages",             corpus),
-  r("runnable code",        sum(corpus$runnable_code, na.rm=T)),
-  r("examples code",        sum(corpus$runnable_code_examples, na.rm=T)),
-  r("tests code",           sum(corpus$runnable_code_tests, na.rm=T)),
-  r("vignettes code",       sum(corpus$runnable_code_vignettes , na.rm=T)),
-  r("mean expr coverage",   percent(mean(corpus$coverage_expr))),
-  r("median expr coverage", percent(median(corpus$coverage_expr))),
-  r("with R code",          filter(corpus, package_r_code > 0)),
-  r("R code",               sum(corpus$package_r_code, na.rm=T)),
-  r("with native code",     filter(corpus, package_native_code > 0)),
-  r("native code",          sum(corpus$package_native_code, na.rm=T)),
-  r("functions",            sum(corpus$funs_public+corpus$funs_private, na.rm=T)),
-  r("public functions",     sum(corpus$funs_public, na.rm=T)),
-  r("private functions",    sum(corpus$funs_private, na.rm=T)),
-  r("ranked in cranlogs",   filter(corpus, !is.na(rank))),
-  r("mean revdes",          mean(corpus$revdeps)),
-  r("median revdes",        median(corpus$revdeps)),
-  r("runtime",              as.duration(sum(corpus$runnable_time))),
-  r("mean runtime",         as.duration(mean(corpus$runnable_time))),
-  r("median runtime",       as.duration(median(corpus$runnable_time))),
-  r("function with eval",   sum(corpus$funs_with_eval, na.rm=T))
+```{r load corpus}
+corpus <- 
+  read_fst(CORPUS_DETAILS_FILE) %>%
+  as_tibble() %>%
+  mutate_at(
+    vars(
+      starts_with("runnable_code"),
+      starts_with("runnable_files"),
+      starts_with("package_"),
+      starts_with("funs_"),
+    ), 
+    replace_na, 
+    0
   )
@@
 
-### CRAN
+```{r load kaggle}
+kaggle_kernels <- as_tibble(read_fst(KAGGLE_KERNELS_FILE))
+kaggle_log <- as_tibble(read_fst(KAGGLE_LOG_FILE)) %>%
+  transmute(
+    id=str_replace(Command, ".*/kaggle-run/[^/]+/([^/]+)/.*", "\\1"),
+    exitval=Exitval,
+    runtime=as.duration(JobRuntime)
+  )
 
-```{r}
-overview_table(
-  r("all cran", dim(available.packages())[1])
+kaggle <- left_join(
+  kaggle_kernels,
+  kaggle_log,
+  by="id"
 )
@@
 
-### Plots
+```{r load runnable code metadata}
+runnable_code_metadata <- read_fst(PACKAGE_RUNNABLE_CODE_FILE) %>%
+  as_tibble() %>%
+  mutate(package=basename(package)) %>%
+  select(package, file=filename, code)
+
+# TODO semi_join with traced-programs.fst
+```
+
+```{r load run metadata}
+runs <- read_fst(PACKAGE_RUNS_FILE) %>%
+  as_tibble() %>%
+  mutate(
+    package=basename(package),
+    file=as.character(path("./", basename(dirname(file)), basename(file)))
+  ) %>%
+  semi_join(corpus, by="package")
+```
+
+## CRAN packages
+
+```{r summary}
+corpus_runtime  <- corpus$runnable_time / corpus$runnable_files
+runnable_code   <- corpus$runnable_code
+corpus_programs_files <- corpus$runnable_files
+
+overview_table(
+  r("packages",             corpus),
+  r("in cranlogs",          filter(corpus, !is.na(rank))),
+  r("package programs",     sum(corpus_programs_files)),
+
+  r("runnable code",        sum(runnable_code)),
+  r("runnable code mean",   mean(runnable_code)),
+  r("runnable code median", median(runnable_code)),
+
+  r("examples code",        sum(corpus$runnable_code_examples)),
+  r("tests code",           sum(corpus$runnable_code_tests)),
+  r("vignettes code",       sum(corpus$runnable_code_vignettes)),
+
+  r("mean expr coverage",   percent(mean(corpus$coverage_expr))),
+  r("median expr coverage", percent(median(corpus$coverage_expr))),
+
+  r("with R code",          filter(corpus, package_r_code > 0)),
+  r("with native code",     filter(corpus, package_native_code > 0)),
+
+  r("code",                 sum(corpus$package_code)),
+  r("R code",               sum(corpus$package_r_code)),
+  r("native code",          sum(corpus$package_native_code)),
+
+  r("functions",            sum(corpus$funs_public+corpus$funs_private)),
+  r("function with eval",   sum(corpus$funs_with_eval)),
+  r("public functions",     sum(corpus$funs_public)),
+  r("private functions",    sum(corpus$funs_private)),
+
+  r("mean revdes",          mean(corpus$revdeps)),
+  r("median revdes",        median(corpus$revdeps)),
+
+  r("runtime",              as.duration(sum(corpus$runnable_time))),
+  r("mean runtime",         as.duration(mean(corpus_runtime))),
+  r("median runtime",       as.duration(median(corpus_runtime)))
+)
+```
+
+```{r generals}
+overview_table(
+  r("all cran", dim(available.packages())[1]),
+  r("core packages", length(CORE_PACKAGES))
+)
+```
+
+### Plot
 
@@
 
-```{r, eval=FALSE}
-parallel_log <- 
-  read_csv(PARALLEL_LOG_FILE)
+```{r}
+kaggle_runnable <- filter(kaggle, runnable)
+kaggle_finished <- filter(kaggle, exitval==0)
+kaggle_code     <- kaggle_finished$code
+kaggle_runtime  <- kaggle_finished$runtime
 
-parallel_log %>%
-  mutate(
-    output=show_url(path(V1, "run.R.out"))
+overview_table(
+  r("kaggle",                 kaggle),
+  r("runnable kaggle",        kaggle_runnable),
+  r("finished kaggle",        kaggle_finished),
+  r("finished kaggle code",   sum(kaggle_code)),
+  r("finished kaggle mean",   mean(kaggle_code)),
+  r("finished kaggle median", median(kaggle_code)),
+  r("kaggle runtime",         as.duration(sum(kaggle_runtime))),
+  r("kaggle mean",            as.duration(mean(kaggle_runtime))),
+  r("kaggle median",          as.duration(median(kaggle_runtime)))
+)
+```
+
+## Combined
+
+```{r}
+overview_table(
+  r("all programs", nrow(kaggle_finished) + sum(corpus_programs_files)),
+  r("all runnbale code", sum(kaggle_code) + sum(runnable_code))
+)
+```
+
+```{r}
+corpus_programs <-
+  left_join(
+    filter(runs, exitval==0) %>% select(package, file, time), 
+    runnable_code_metadata %>% select(package, file, code),
+    by=c("package", "file")
   ) %>%
-  my_datatable(escape=FALSE)
+  transmute(id=paste(package, ":", file), code, runtime=time, corpus="packages")
+
+kaggle_programs <-
+  kaggle %>%
+  filter(exitval==0) %>%
+  transmute(id, code, runtime=as.numeric(runtime, units="seconds"), corpus="kaggle")
+
+programs <- bind_rows(corpus_programs, kaggle_programs)
 ```
+
+```{r}
+programs %>%
+  ggplot(aes(y=code, x=corpus, fill=corpus)) + 
+  geom_flat_violin(trim=FALSE, width=.5) +
+  geom_dotplot(
+    binaxis="y", 
+    dotsize=0.1, 
+    stackdir="down", 
+    binwidth=0.01, 
+    position=position_nudge(-0.025)) +
+  scale_y_log10(labels = scales::comma) +
+  scale_x_discrete(labels=c("kaggle"="Kaggle", "packages"="Packages")) +
+  scale_fill_viridis(discrete = TRUE) +
+  guides(fill = FALSE) +
+  labs(y="Code size (log)") +
+  theme(
+    legend.box="horizontal",
+    axis.title.x=element_blank()
+  )
+
+ggsave(path(PLOT_DIR, "programs-code.pdf"), width=7, height=5.5)
+```
+
+```{r}
+programs %>%
+  ggplot(aes(y=runtime, x=corpus, fill=corpus)) + 
+  geom_flat_violin(trim=FALSE, width=.5) +
+  geom_dotplot(
+    binaxis="y", 
+    dotsize=0.1, 
+    stackdir="down", 
+    binwidth=0.01, 
+    position=position_nudge(-0.025)) +
+  scale_y_log10(labels = function(x, ...) {
+    min <- as.integer(x/60L)
+    sec <- as.integer(x%%60L)
+    sprintf("%.2d:%.2d", min, sec)
+  }) +
+  scale_x_discrete(labels=c("kaggle"="Kaggle", "packages"="Packages")) +
+  scale_fill_viridis(discrete = TRUE) +
+  guides(fill = FALSE) +
+  labs(y="Running time [mm:ss] (log)") +
+  theme(
+    legend.box="horizontal",
+    axis.title.x=element_blank()
+  )
+ggsave(path(PLOT_DIR, "programs-runtime.pdf"), width=3.5, height=5.0)
+```

### 2020-08-24 inst/paper/main.tex
Add own, dependent and core eval calls from CRAN and Kaggle
Regenerated data for V

@@
 
+We have assembled a corpus that consists of \CorpusPackages R packages and user programs.
+
 We have selected \CorpusPackages packages consisting of \CorpusRCodeRnd
@@
 tested and only a successfully running package is admitted in the archive.
+
+\begin{figure}[!tb]\centering\includegraphics[width=.9\linewidth]
+  {programs-code.pdf}\caption{Corpus}\label{fig:corpus}
+\end{figure}
+\begin{figure}[!tb]\centering\includegraphics[width=.9\linewidth]
+  {programs-runtime.pdf}\caption{Corpus}\label{fig:corpus}
+\end{figure}
 
@@
 \end{verbatim}
+
+\begin{table}[ht]
+  \resizebox{\columnwidth}{!}{%
+    \begin{tabular}{r|rrr|}
+      \hline
+      Source & Core & Dependent & Own \\
+      \hline
+      CRAN & \CranCoreTotalEvalCallsRnd (\CranCoreMinEvalCallsRnd/\CranCoreAvgEvalCallsRnd/\CranCoreMaxEvalCallsRnd) & \CranDependentTotalEvalCallsRnd (\CranDependentMinEvalCallsRnd/\CranDependentAvgEvalCallsRnd/\CranDependentMaxEvalCallsRnd) & \CranOwnTotalEvalCallsRnd (\CranOwnMinEvalCallsRnd/\CranOwnAvgEvalCallsRnd/\CranOwnMaxEvalCallsRnd) \\
+      Kaggle & \KaggleCoreTotalEvalCallsRnd (\KaggleCoreMinEvalCallsRnd/\KaggleCoreAvgEvalCallsRnd/\KaggleCoreMaxEvalCallsRnd) & \KaggleDependentTotalEvalCallsRnd (\KaggleDependentMinEvalCallsRnd/\KaggleDependentAvgEvalCallsRnd/\KaggleDependentMaxEvalCallsRnd) & 0(0/0/0) \\
+      \hline
+    \end{tabular}
+  }
+  \label{table:source-eval-summary}
+  \caption{Distribution of eval calls in CRAN and Kaggle}
+\end{table}
 

### 2020-08-24 inst/analysis/inc/setup.R

@@ -9,3 +9,67 @@ is_outlier <- function(x, m=1.5) {
 show_url <- Vectorize(function(path, name=basename(path), hostname=params$hostname, port=params$port) {
   str_glue('<a href="http://{hostname}:{port}/{URLencode(path)}">{name}</a>')
 }, vectorize.args = "path")
+
+
+"%||%" <- function(a, b) {
+  if (!is.null(a)) a else b
+}
+
+geom_flat_violin <- function(mapping = NULL, data = NULL, stat = "ydensity",
+                             position = "dodge", trim = TRUE, scale = "area",
+                             show.legend = NA, inherit.aes = TRUE, ...) {
+  layer(
+    data = data,
+    mapping = mapping,
+    stat = stat,
+    geom = GeomFlatViolin,
+    position = position,
+    show.legend = show.legend,
+    inherit.aes = inherit.aes,
+    params = list(
+      trim = trim,
+      scale = scale,
+      ...
+    )
+  )
+}
+
+GeomFlatViolin <-
+  ggproto("GeomFlatViolin", Geom,
+          setup_data = function(data, params) {
+            data$width <- data$width %||%
+              params$width %||% (resolution(data$x, FALSE) * 0.9)
+            
+            # ymin, ymax, xmin, and xmax define the bounding rectangle for each group
+            data %>%
+              group_by(group) %>%
+              mutate(ymin = min(y),
+                     ymax = max(y),
+                     xmin = x,
+                     xmax = x + width / 2
+              )
+          },
+  
+  draw_group = function(data, panel_scales, coord) {
+    # Find the points for the line to go all the way around
+    data <- transform(data, xminv = x,
+                      xmaxv = x + violinwidth * (xmax - x))
+    
+    # Make sure it's sorted properly to draw the outline
+    newdata <- rbind(plyr::arrange(transform(data, x = xminv), y),
+                     plyr::arrange(transform(data, x = xmaxv), -y))
+    
+    # Close the polygon: set first and last point the same
+    # Needed for coord_polar and such
+    newdata <- rbind(newdata, newdata[1,])
+    
+    ggplot2:::ggname("geom_flat_violin", GeomPolygon$draw_panel(newdata, panel_scales, coord))
+  },
+  
+  draw_key = draw_key_polygon,
+  
+  default_aes = aes(weight = 1, colour = "grey20", fill = "white", size = 0.5,
+                    alpha = NA, linetype = "solid"),
+  
+  required_aes = c("x", "y")
+)

### 2020-08-24 inst/analysis/corpus-analysis.Rmd
Filled some misses

@@
   r("all programs", nrow(kaggle_finished) + sum(corpus_programs_files)),
-  r("all runnbale code", sum(kaggle_code) + sum(runnable_code))
+  r("all runnbale code", sum(kaggle_code) + sum(runnable_code)),
+  r("all code", sum(kaggle_code) + sum(corpus_code) + sum(runnable_code))
 )

### 2020-08-24 inst/paper/main.tex
Filled some misses

@@
   features of the language. This paper is a large-scale study of a corpus of
-  \miss{3} million lines of data science libraries and end-user written in
+  \CorpusAllCodeRnd lines of data science libraries and end-user written in
   the R language. We performed a simple static analysis and more
@@
   hand, know about laziness and program defensively around it.}  Thus, our
-corpus distinguished between \emph{core packages} (there are \miss{8} such
-packages that are loaded together with the R virtual machine), \emph{CRAN
-  packages} (we selected \miss{500} curated packages that pass stringent
+corpus distinguished between \emph{core packages} (there are \CorpusCorePackages such
+packages that are distributed together with the R virtual machine), \emph{CRAN
+  packages} (we selected \CorpusPackages curated packages that pass stringent
 community quality checks and are equipped with tests and sample data),
-\emph{scripts} (we used \miss{400} end-user written programs that performs a
+\emph{Kaggle scripts} (we used \CorpusFinishedKaggle end-user written programs that performs a
 particular data analysis task). There are reasons to believe that \eval
@@
 
-Over all, we observed \AllAllCallCountRnd call to \eval in \miss{3,000} runs
-of our corpus. The mean number of calls per run was \miss{20} (with a min of
-\miss{1} and a max of \miss{10,000}). Table~\ref{A} details the usage of
-\eval across the corpus.  The format is T (m/M/x) where T is the total
-number of calls observed for all runs of the corpus, m and x are the minimum
-and maximum number of calls for any given run, and A is the mean number of
-calls per run.  We can observe \miss{???}. {\bf <The following is a place
-  holder. Real text will come once we have the table.>} The majority of
-calls come from the core packages.  Kaggle does not have any own calls, any
-\eval in those happens in dependent packages and core.  Of the
-\missingNumber packages in our corpus, only \missingNumber packages call
-eval.
+Over all, we observed \AllAllCallCountRnd call to \eval in \CorpusAllProgramsRnd
+programs in our corpus. The mean number of calls per run was \miss{20} (with a
+min of \miss{1} and a max of \miss{10,000}). Table~\ref{A} details the usage of
+\eval across the corpus. The format is T (m/M/x) where T is the total number of
+calls observed for all runs of the corpus, m and x are the minimum and maximum
+number of calls for any given run, and A is the mean number of calls per run. We
+can observe \miss{???}. {\bf <The following is a place holder. Real text will
+  come once we have the table.>} The majority of calls come from the core
+packages. Kaggle does not have any own calls, any \eval in those happens in
+dependent packages and core. Of the \missingNumber packages in our corpus, only
+\missingNumber packages call eval.
 

### 2020-08-24 inst/analysis/corpus-analysis.Rmd
V.A

--- 2020-08-24 inst/analysis/corpus-analysis.Rmd
@@
   r("R code",               sum(corpus$package_r_code)),
+  r("R code mean",          mean(corpus$package_r_code)),
+  r("R code median",        median(corpus$package_r_code)),
   r("native code",          sum(corpus$package_native_code)),
+  r("native code mean",     mean(corpus$package_native_code)),
+  r("native code median",   median(corpus$package_native_code)),
 
@@
   r("all cran", dim(available.packages())[1]),
-  r("core packages", length(CORE_PACKAGES))
+  r("core packages", length(CORE_PACKAGES)),
+  r("most downloaded treshold", MOST_DOWNLOADED),
+  r("large package treshold", PACKAGE_SIZE_OUTLIER)
 )
@@
     outlier=package_code > PACKAGE_SIZE_OUTLIER,
-    label=ifelse(rank <= 25 | outlier, package, NA)  #if_else(outlier, str_glue("{package}"), as.character(NA))
+    label=ifelse(rank <= MOST_DOWNLOADED | outlier, package, NA)  #if_else(outlier, str_glue("{package}"), as.character(NA))
   ) %>%
@@
 
-```{r}
+```{r kaggle summary}
 kaggle_runnable <- filter(kaggle, runnable)
@@
 
-```{r}
+```{r all summary}
 overview_table(
   r("all programs", nrow(kaggle_finished) + sum(corpus_programs_files)),
-  r("all runnbale code", sum(kaggle_code) + sum(runnable_code))
+  r("all runnbale code", sum(kaggle_code) + sum(runnable_code)),
+  r("all code", sum(kaggle_code) + sum(corpus$package_code) + sum(runnable_code))
 )
@@
 
-```{r}
+```{r compute programs}
 corpus_programs <-

### 2020-08-24 inst/paper/main.tex
V.A

@@
 
-We have assembled a corpus that consists of \CorpusPackages R packages and user programs.
+We have assembled a corpus of \CorpusPackages R packages and
+\CorpusFinishedKaggle user-written data-analysis scripts. Together it contains
+\CorpusAllCodeRnd lines of source code\footnote{Excluding comments and blank lines using \url{github.com/AlDanial/cloc}}.
 
-We have selected \CorpusPackages packages consisting of \CorpusRCodeRnd
-lines of R code and \CorpusNativeCodeRnd lines of native code
-(C/C++/Fortran).  Figure~\ref{fig:corpus} shows these packages, the size of
-the dots reflects the project's size in lines of code including both R and
-native code\footnote{Excluding comments and blank lines,
-  \url{github.com/AlDanial/cloc}}, the x-axis indicates the expression code
-coverage in percents and the y-axis gives the number of reverse dependencies
-in log scale. Dotted lines indicate means. Large and popular packages are
-annotated.
+\subsection{CRAN Packages}
+
+We have selected the TOP \CorpusPackages packages based on the number of reverse
+dependencies, \ie how many packages depend on them\footnote{extracted from
+  package metadata using builtin function}. Using the number of reverse
+dependencies helps us to choose popular packages which in general should be of a
+better quality and should have better code coverage. The resulting set consists
+of \CorpusRCodeRnd (median=\CorpusRCodeMedian) lines of R code and
+\CorpusNativeCodeRnd (median=\CorpusNativeCodeMedian) lines of native code
+(C/C++/Fortran). Figure~\ref{fig:corpus} shows these packages, the size of the
+dots reflects the project's size in lines of code including both R and native
+code, the x-axis indicates the expression code coverage in percents and the
+y-axis gives the number of reverse dependencies in log scale. Dotted lines
+indicate means. Large (over \CorpusLargePackageTresholdRnd lines of code) and popular (in \CorpusMostDownloadedTreshold most downloaded) packages are annotated.
 
@@
 package must abide to a number of well-formedness rules that are automatically
-checked asserting certain quality. Most relevant for this work is that all of
-the runnable code (including code snippets from examples and vignettes) is
-tested and only a successfully running package is admitted in the archive.
+checked asserting certain quality.
+
+Next to code and tests, R packages also contain examples and vignettes. Examples
+are runnable snippets extracted from documentation. Vignettes are tutorial-like
+guides into the package functionality. They are written in the literate
+programming style using either Latex or Rmarkdown format with embedded code
+snippets. Since CRAN is a curated repository, upon submission, it checks all the
+runnable code in tests, examples and vignettes and only a successfully running
+package is admitted to the archive. We use the same mechanism to extract
+runnable code and run it for the dynamic analysis.
+
+In the selected packages, there are \CorpusRunnableCodeRnd lines of runnable
+code in examples (\CorpusExamplesCodeRnd), tests (\CorpusTestsCodeRnd) and
+vignettes (\CorpusVignettesCodeRnd). Running this code results on average in
+\CorpusMeanExprCoverage package code coverage. The code coverage was computed
+for each package using
 
 \begin{figure}[!tb]\centering\includegraphics[width=.9\linewidth]
-  {programs-code.pdf}\caption{Corpus}\label{fig:corpus}
+  {programs-code.pdf}\caption{Programs code size}\label{fig:programs-code}
 \end{figure}
 \begin{figure}[!tb]\centering\includegraphics[width=.9\linewidth]
-  {programs-runtime.pdf}\caption{Corpus}\label{fig:corpus}
+  {programs-runtime.pdf}\caption{Programs execution time}\label{fig:programs-runtime}
 \end{figure}
 
-% FILIP: WIP
-% We have downloaded and installed all available CRAN packages. Out of the
-% \AllCranRnd packages, we managed to install \AllLoadableRnd. The main reason for
-% this is that \rdt is based on GNU-R 3.5.0 and some of the packages are no longer
-% compatible with this version. Some packages also require extra native
-% dependencies which were not present on our servers. We defined two criteria for
-% including a package into the corpus:
-% \begin{inparaenum}[(1)]
-%   \item have a runnable code that covers a significant part of the package
-%   source code from which type signatures could be inferred, and
-%   \item have some reverse dependencies that will allows us to evaluate the
-%   inferred types using the runnable code from these dependencies.
-% \end{inparaenum}
-% The concrete thresholds used were: at least \ThresholdCodeCoverage of expression
-% coverage and at minimum \ThresholdRevdeps reverse dependencies. The code
-% coverage was computed for each package using
-% \covr\footnote{\url{https://github.com/r-lib/covr}}, the R code coverage R tool.
-% The reverse package dependencies were extracted from the package metadata using
-% builtin function.
-
-% The \CorpusLoadable selected packages contain \CorpusRunnableCodeRnd lines of
-% runnable code in examples (\CorpusExamplesCodeRnd), tests (\CorpusTestsCodeRnd)
-% and vignettes (\CorpusVignettesCodeRnd). Running this code results on average in
-% \CorpusMeanExprCoverage package code coverage (CRAN average is
-% \AllMeanExprCoverage). Together, there is \CorpusRevdesRnd (on average
-% \CorpusMeanRevdes, median \CorpusMedianRevdes; CRAN average is \AllMeanRevdes,
-% median \AllMedianRevdes) reverse dependencies with \CorpusRevdepRunnableCodeRnd
-% runnable lines of code resulting in \CorpusRevdepMeanCodeCoverage coverage (on
-% average) of the corpus packages.
-% Together there are \CorpusFunctionsRnd defined R functions
-% (\CorpusPublicFunctionsRnd are from packages' public API).
-% \CorpusSThreeFunctionsRnd are S3 functions. Packages in the corpus define
-% \CorpusSThreeClasses S3 classes.
-% The code coverage was computed for each package using
-% \covr\footnote{\url{https://github.com/r-lib/covr}}, the R code coverage R
-% tool.  The reverse package dependencies were extracted from the package
-% metadata using builtin function.
+\subsection{Kaggle sripts}
+% contrast to CRAN - duplication
 

### 2020-08-24 inst/analysis/inc/paths.R

@@ -64,6 +64,8 @@ WHICH_DEPENDENCIES <- c("Depends", "Imports")
 CRAN_MIRROR_LOCAL_URL <- paste0("file://", params$base_dir, "/CRAN")
 # size of a package to be considered as outlier
 PACKAGE_SIZE_OUTLIER <- 50000
+# 
+MOST_DOWNLOADED <- 25
 # R packages distributed with vanilla R
 CORE_PACKAGES <- c(
   "base",

### 2020-08-25 inst/paper/main.tex
usage metrics

@@
 
+\begin{figure*}[!tb]\centering\includegraphics[width=.8\linewidth]
+  {corpus.pdf}\caption{Corpus}\label{fig:corpus}
+\end{figure*}
+
 \section{Corpus of R Programs}
@@
 
-\begin{figure*}[!tb]\centering\includegraphics[width=.9\linewidth]
-  {corpus.pdf}\caption{Corpus}\label{fig:corpus}
-\end{figure*}
+\begin{figure}[!htb]\centering\includegraphics[width=.9\linewidth]
+  {programs-code.pdf}\caption{Programs code size}\label{fig:programs-code}
+\end{figure}
+
+\begin{figure}[!htb]\centering\includegraphics[width=.9\linewidth]
+  {programs-runtime.pdf}\caption{Programs execution time}\label{fig:programs-runtime}
+\end{figure}
 
@@
 
-\begin{figure}[!tb]\centering\includegraphics[width=.9\linewidth]
-  {programs-code.pdf}\caption{Programs code size}\label{fig:programs-code}
-\end{figure}
-\begin{figure}[!tb]\centering\includegraphics[width=.9\linewidth]
-  {programs-runtime.pdf}\caption{Programs execution time}\label{fig:programs-runtime}
-\end{figure}
-
 \subsection{Kaggle sripts}
@@
 Previous work reported that 82\% of JavaScript-enabled web pages used
-\eval~\cite{ecoop11}, the case for R is more striking at \miss{100\%} of the
-programs use that feature. In this section, a call to \eval denotes a call
-to any one of the functions that provide \eval-like
-functionality.\footnote{These include \c{eval}, \c{eval.parent}, \c{evalq}
-  and \c{local}.}
+\eval~\cite{ecoop11}. The case for R is more striking with 100\% of the
+programs using that \eval.\footnote{In this section, we combine calls to
+  \c{eval}, \c{eval.parent}, \c{evalq} and \c{local}.}  This high number is
+explained by the fact that the R implementation, its core packages, use
+\eval extensively. In our analysis, we differentiate uses of \eval based on
+where they originate from. A {\bf core} \eval is one that occurs in the code
+of a core package.  A {\bf dependent} \eval is one that occurs in a package
+that was included by the target package. An {\bf own} \eval is one that
+occurs in the source code of the target package.
 
-This strikingly high number is due to the fact that the language
-implementation, the core packages, use \eval extensively. So, in our
-analysis, we differentiate \eval based on where they originate from. A {\bf
-  core} \eval is one that occurs in the code of a core package.  A {\bf
-  dependent} \eval is one that occurs in a package that was included by the
-target package. An {\bf own} \eval is one that occurs in the source code of
-the target package. We label an \eval occurence as {\bf unknown} when we are
-not able to deterimine its origin.\footnote{A source of unknowns is when the
-  \eval function is passed as an argument to a function such as \c{lapply},
-  another is when the code lacks source references.}
-
-Over all, we observed \AllAllCallCountRnd call to \eval in \CorpusAllProgramsRnd
-programs in our corpus. The mean number of calls per run was \miss{20} (with a
-min of \miss{1} and a max of \miss{10,000}). Table~\ref{A} details the usage of
-\eval across the corpus. The format is T (m/M/x) where T is the total number of
-calls observed for all runs of the corpus, m and x are the minimum and maximum
-number of calls for any given run, and A is the mean number of calls per run. We
-can observe \miss{???}. {\bf <The following is a place holder. Real text will
-  come once we have the table.>} The majority of calls come from the core
-packages. Kaggle does not have any own calls, any \eval in those happens in
-dependent packages and core. Of the \missingNumber packages in our corpus, only
-\missingNumber packages call eval.
-
-\begin{verbatim}
-Table A
-
-         core      dependent   own
-----------------------------------------
-CRAN   10M?(1/2/3) 1M?(1/2/3) 1K?(1/2/3)
-Kaggle 2K?(1/2/3)  1K?(1/2/3) 0 (0/0/0)
-\end{verbatim}
+Over all, we observed \AllAllCallCountRnd calls to \eval in the
+\CorpusAllProgramsRnd programs in our corpus.  Table~\ref{A} details this
+run-time information. The format is T (m/M/x) where T is the total number of
+calls observed for all runs of the corpus, m and x are the minimum and
+maximum number of calls for any given run, and M is the mean number of calls
+per run. We can observe that an average run of CRAN package will trigger 581
+{\eval}s in core package, 532 in dependent packages and 292 in the package
+itself.  The Kaggle data set has small programs, and, these programs only
+trigger \eval in core and dependent packages. Thus, an average Kaggle run
+has 412 core calls, 96 dependent calls, and no own calls.
 
@@
 \resizebox{\columnwidth}{!}{%
-    \begin{tabular}{r|rrr|}
-      \hline
-      Source & Core & Dependent & Own \\
-      \hline
-      CRAN & \CranCoreTotalEvalCallsRnd (\CranCoreMinEvalCallsRnd/\CranCoreAvgEvalCallsRnd/\CranCoreMaxEvalCallsRnd) & \CranDependentTotalEvalCallsRnd (\CranDependentMinEvalCallsRnd/\CranDependentAvgEvalCallsRnd/\CranDependentMaxEvalCallsRnd) & \CranOwnTotalEvalCallsRnd (\CranOwnMinEvalCallsRnd/\CranOwnAvgEvalCallsRnd/\CranOwnMaxEvalCallsRnd) \\
-      Kaggle & \KaggleCoreTotalEvalCallsRnd (\KaggleCoreMinEvalCallsRnd/\KaggleCoreAvgEvalCallsRnd/\KaggleCoreMaxEvalCallsRnd) & \KaggleDependentTotalEvalCallsRnd (\KaggleDependentMinEvalCallsRnd/\KaggleDependentAvgEvalCallsRnd/\KaggleDependentMaxEvalCallsRnd) & 0(0/0/0) \\
+\begin{tabular}{@{}r|ccc@{}} \hline
+Source & Core & Dependent & Own \\\hline
+      %      CRAN & \CranCoreTotalEvalCallsRnd (\CranCoreMinEvalCallsRnd/\CranCoreAvgEvalCallsRnd/\CranCoreMaxEvalCallsRnd) & \CranDependentTotalEvalCallsRnd (\CranDependentMinEvalCallsRnd/\CranDependentAvgEvalCallsRnd/\CranDependentMaxEvalCallsRnd) & \CranOwnTotalEvalCallsRnd (\CranOwnMinEvalCallsRnd/\CranOwnAvgEvalCallsRnd/\CranOwnMaxEvalCallsRnd) \\
+CRAN & \CranCoreTotalEvalCallsRnd (1 / 581 / 416K)  & \CranDependentTotalEvalCallsRnd (1 / 532 / 183K) & \CranOwnTotalEvalCallsRnd (1 / 292 / 144K)\\
+Kaggle & \KaggleCoreTotalEvalCallsRnd (1 / 412 / 2.8K) & \KaggleDependentTotalEvalCallsRnd (3 / 96 / 1.1K) & 0 (0 / 0 / 0) \\
+%      Kaggle & \KaggleCoreTotalEvalCallsRnd (\KaggleCoreMinEvalCallsRnd/\KaggleCoreAvgEvalCallsRnd/\KaggleCoreMaxEvalCallsRnd) & \KaggleDependentTotalEvalCallsRnd (\KaggleDependentMinEvalCallsRnd/\KaggleDependentAvgEvalCallsRnd/\KaggleDependentMaxEvalCallsRnd) & 0(0/0/0) \\
       \hline
@@
   }
-  \label{table:source-eval-summary}
-  \caption{Distribution of eval calls in CRAN and Kaggle}
+  \label{A}
+  \caption{Eval calls in CRAN and Kaggle}
 \end{table}
 
-The following table will be replaced. {\bf <At this point separating between
-  the different kinds of evals is distracting. We have not argued why it is
-  important.>}
+%% The following table will be replaced. {\bf <At this point separating between
+%%   the different kinds of evals is distracting. We have not argued why it is
+%%   important.>}
 
-\begin{table}[ht]
-  \resizebox{\columnwidth}{!}{%
-    \begin{tabular}{l|rr|r||rr|r}
-      \hline
-      \multicolumn{1}{c}{Function}& \multicolumn{3}{c}{Calls} & \multicolumn{3}{c}{Callsites}\\
-      \cline{1-1} \cline{2-4} \cline{5-7}
-                                  & Core & Packages & Total & Core & Packages & Total\\
-      \hline
-      \c{eval} & \CoreEvalCallCountRnd & \PackageEvalCallCountRnd  & \AllEvalCallCountRnd & \CoreEvalSiteCountRnd & \PackageEvalSiteCountRnd &  \AllEvalSiteCountRnd\\
-      \c{eval.parent} & \CoreEvalParentCallCountRnd & \PackageEvalParentCallCountRnd  & \AllEvalParentCallCountRnd & \CoreEvalParentSiteCountRnd & \PackageEvalParentSiteCountRnd & \AllEvalParentSiteCountRnd\\
-      \c{evalq} & \CoreEvalqCallCountRnd & \PackageEvalqCallCountRnd  & \AllEvalqCallCountRnd & \CoreEvalqSiteCountRnd & \PackageEvalqSiteCountRnd & \AllEvalqSiteCountRnd \\
-      \c{local} & \CoreLocalCallCountRnd & \PackageLocalCallCountRnd  & \AllLocalCallCountRnd & \CoreLocalSiteCountRnd & \PackageLocalSiteCountRnd & \AllLocalSiteCountRnd \\
-      \hline
-      Total & \CoreAllCallCountRnd & \PackageAllCallCountRnd  & \AllAllCallCountRnd & \CoreAllSiteCountRnd & \PackageAllSiteCountRnd & \AllAllSiteCountRnd \\
-      \hline
-    \end{tabular}
-  }
-  \label{table:eval-count-summary}
-  \caption{Distribution of eval calls and callsites in the corpus}
-\end{table}
+%% \begin{table}[ht]\resizebox{\columnwidth}{!}{\begin{tabular}{l|rr|r||rr|r}\hline
+%% \multicolumn{1}{c}{Function}& \multicolumn{3}{c}{Calls} & \multicolumn{3}{c}{Callsites}\\
+%% \cline{1-1} \cline{2-4} \cline{5-7}
+%% & Core & Packages & Total & Core & Packages & Total\\\hline
+%% \c{eval} & \CoreEvalCallCountRnd & \PackageEvalCallCountRnd  & \AllEvalCallCountRnd & \CoreEvalSiteCountRnd & \PackageEvalSiteCountRnd &  \AllEvalSiteCountRnd\\
+%% \c{eval.parent} & \CoreEvalParentCallCountRnd & \PackageEvalParentCallCountRnd  & \AllEvalParentCallCountRnd & \CoreEvalParentSiteCountRnd & \PackageEvalParentSiteCountRnd & \AllEvalParentSiteCountRnd\\
+%% \c{evalq} & \CoreEvalqCallCountRnd & \PackageEvalqCallCountRnd  & \AllEvalqCallCountRnd & \CoreEvalqSiteCountRnd & \PackageEvalqSiteCountRnd & \AllEvalqSiteCountRnd \\
+%% \c{local} & \CoreLocalCallCountRnd & \PackageLocalCallCountRnd  & \AllLocalCallCountRnd & \CoreLocalSiteCountRnd & \PackageLocalSiteCountRnd & \AllLocalSiteCountRnd \\\hline
+%% Total & \CoreAllCallCountRnd & \PackageAllCallCountRnd  & \AllAllCallCountRnd & \CoreAllSiteCountRnd & \PackageAllSiteCountRnd & \AllAllSiteCountRnd \\  \hline
+%% \end{tabular} } \label{table:eval-count-summary}
+%%  \caption{Distribution of eval calls and callsites in the corpus}
+%% \end{table}
 
-\ref{table:top-ten-package-summary} shows the number of calls to eval
-made by the ten most frequent callers to eval. These packages together have
-\missingNumber callsites and account for \missingPercentage calls to eval.
-Furthermore, the top ten callsites account for \missingPercentage of package
-eval calls. A single callsite of \TopTenPackageNameA accounts for over
-\missingPercentage calls to eval.
+%% \ref{table:top-ten-package-summary} shows the number of calls to eval
+%% made by the ten most frequent callers to eval. These packages together have
+%% \missingNumber callsites and account for \missingPercentage calls to eval.
+%% Furthermore, the top ten callsites account for \missingPercentage of package
+%% eval calls. A single callsite of \TopTenPackageNameA accounts for over
+%% \missingPercentage calls to eval.
 
-\begin{table}[ht]
-  \resizebox{\columnwidth}{!}{%
-    \begin{tabular}{l|rrr||l|rrr}
-      \hline
-      \multicolumn{4}{c}{}& \multicolumn{4}{c}{} \\
-      \cline{1-4} \cline{5-8}
-      Package & Callsites & Calls & Calls \% & Package & Callsites & Calls & Calls \%\\
-      \hline
-      \TopTenPackageNameA & \TopTenPackageCallsiteCountA & \TopTenPackageCallCountA &\TopTenPackageCallPercA & \TopTenPackageNameF & \TopTenPackageCallsiteCountF & \TopTenPackageCallCountF & \TopTenPackageCallPercF \\
-      \TopTenPackageNameB & \TopTenPackageCallsiteCountB & \TopTenPackageCallCountB &\TopTenPackageCallPercB & \TopTenPackageNameG & \TopTenPackageCallsiteCountG & \TopTenPackageCallCountG & \TopTenPackageCallPercG \\
-      \TopTenPackageNameC & \TopTenPackageCallsiteCountC & \TopTenPackageCallCountC &\TopTenPackageCallPercC & \TopTenPackageNameH & \TopTenPackageCallsiteCountH & \TopTenPackageCallCountH & \TopTenPackageCallPercH \\
-      \TopTenPackageNameD & \TopTenPackageCallsiteCountD & \TopTenPackageCallCountD &\TopTenPackageCallPercD & \TopTenPackageNameI & \TopTenPackageCallsiteCountI & \TopTenPackageCallCountI & \TopTenPackageCallPercI \\
-      \TopTenPackageNameE & \TopTenPackageCallsiteCountE & \TopTenPackageCallCountE &\TopTenPackageCallPercE & \TopTenPackageNameJ & \TopTenPackageCallsiteCountJ & \TopTenPackageCallCountJ & \TopTenPackageCallPercJ \\
-      \hline
-    \end{tabular}
-  }
-  \label{table:top-ten-package-summary}
-  \caption{Distribution of eval calls and callsites in the top ten packages}
-\end{table}
+%% \begin{table}[ht]
+%%   \resizebox{\columnwidth}{!}{%
+%%     \begin{tabular}{l|rrr||l|rrr}
+%%       \hline
+%%       \multicolumn{4}{c}{}& \multicolumn{4}{c}{} \\
+%%       \cline{1-4} \cline{5-8}
+%%       Package & Callsites & Calls & Calls \% & Package & Callsites & Calls & Calls \%\\
+%%       \hline
+%%       \TopTenPackageNameA & \TopTenPackageCallsiteCountA & \TopTenPackageCallCountA &\TopTenPackageCallPercA & \TopTenPackageNameF & \TopTenPackageCallsiteCountF & \TopTenPackageCallCountF & \TopTenPackageCallPercF \\
+%%       \TopTenPackageNameB & \TopTenPackageCallsiteCountB & \TopTenPackageCallCountB &\TopTenPackageCallPercB & \TopTenPackageNameG & \TopTenPackageCallsiteCountG & \TopTenPackageCallCountG & \TopTenPackageCallPercG \\
+%%       \TopTenPackageNameC & \TopTenPackageCallsiteCountC & \TopTenPackageCallCountC &\TopTenPackageCallPercC & \TopTenPackageNameH & \TopTenPackageCallsiteCountH & \TopTenPackageCallCountH & \TopTenPackageCallPercH \\
+%%       \TopTenPackageNameD & \TopTenPackageCallsiteCountD & \TopTenPackageCallCountD &\TopTenPackageCallPercD & \TopTenPackageNameI & \TopTenPackageCallsiteCountI & \TopTenPackageCallCountI & \TopTenPackageCallPercI \\
+%%       \TopTenPackageNameE & \TopTenPackageCallsiteCountE & \TopTenPackageCallCountE &\TopTenPackageCallPercE & \TopTenPackageNameJ & \TopTenPackageCallsiteCountJ & \TopTenPackageCallCountJ & \TopTenPackageCallPercJ \\
+%%       \hline
+%%     \end{tabular}
+%%   }
+%%   \label{table:top-ten-package-summary}
+%%   \caption{Distribution of eval calls and callsites in the top ten packages}
+%% \end{table}
 
-\missingGraph shows the distribution of number of callsites to eval across these
-packages. \missingNumber packages have a single eval callsite.
+Table \ref{table:site-package-summary} shows the results of static analysis
+for the CRAN packages in our corpus. 89 packages have a single call site,
+and one package has \miss{252} call sites of \eval. \miss{XXX} packages have
+no calls to \eval. The core packages have \miss{41} callsites of eval and
+Kaggle has none. These static numbers are under approximations as \eval may
+be called reflectively, through aliases, or passed to a higher-order function.
 
-\begin{table}[ht]
-  \resizebox{\columnwidth}{!}{%
-    \begin{tabular}{rr|rr|rr}
-      \hline
-      Callsites & Packages & Callsites & Packages & Callsites & Packages \\
-      \hline
+\begin{table}[ht]\resizebox{\columnwidth}{!}{%
+\begin{tabular}{rr|rr|rr}\hline
+Callsites & Packages & Callsites & Packages & Callsites & Packages \\ \hline
 \SiteSummarySiteCountA & \SiteSummaryPackageCountA & \SiteSummarySiteCountF & \SiteSummaryPackageCountF & \SiteSummarySiteCountK & \SiteSummaryPackageCountK \\
@@
 \SiteSummarySiteCountD & \SiteSummaryPackageCountD & \SiteSummarySiteCountI & \SiteSummaryPackageCountI & \SiteSummarySiteCountN & \SiteSummaryPackageCountN \\
-      \SiteSummarySiteCountE & \SiteSummaryPackageCountE & \SiteSummarySiteCountJ & \SiteSummaryPackageCountJ & \SiteSummarySiteCountO & \SiteSummaryPackageCountO \\ 
-      \hline
-    \end{tabular}
-  }
-  \label{table:site-package-summary}
-  \caption{Distribution of callsites across packages}
+      \SiteSummarySiteCountE & \SiteSummaryPackageCountE & \SiteSummarySiteCountJ & \SiteSummaryPackageCountJ & \SiteSummarySiteCountO & \SiteSummaryPackageCountO \\    \hline
+\end{tabular}} \label{table:site-package-summary}
+\caption{Distribution of callsites in CRAN}
 \end{table}
 
-\subsubsection{The size of \eval}
-
-In Figure~\ref{fig:ast-size}, we look at the the size of the AST, in number of nodes, of the \c(expr) argument of \eval, after it is resolved.
+In Figure~\ref{fig:ast-size}, we look at the the size of the AST, in number
+of nodes, of the \c(expr) argument of \eval, after it is resolved.
 

### 2020-08-25 inst/paper/main.tex
usage metrics (done)

@@
 
-In Figure~\ref{fig:ast-size}, we look at the the size of the AST, in number
-of nodes, of the \c(expr) argument of \eval, after it is resolved.
+To estimate the complexity of the code being evaluated,
+Fig.~\ref{fig:ast-size} shows the distribution of the abstract syntax tree
+size of the \c{expr} argument to \eval. The majority of inputs across core
+and packages (\miss{XX}\%) is 1. That is to say, the code being evaluated
+consists of a single variable name or constant. A function call needs at
+least three nodes. Inputs smaller than \c{X} nodes account for 95\% of all
+calls. Sizes are roughly similar between core and packages.
 
@@
 \includegraphics[width=\columnwidth]{ast_sizes}
-	\caption{Size of the \c{expr} argument of \eval in terms of AST nodes, for the \emph{core} dataset and the \emph{packages} dataset.} \label{fig:ast-size}
+\caption{Input sizes} \label{fig:ast-size}
 \end{figure}
 
-Most of these expressions have very few nodes and about 50\% have only one. It is because \c{expr} is often reduced to just a value or a symbol.
-
-\miss{This is on a random sample of 10000 unique rows of the summarized data set as ST size takes a lof of time to compute. To update when it finishes computing...}
+\medskip\noindent\emph{Discussion:} Comparing the use of \eval in R and JavaScript is
+instructive. The use of \eval is more widespread in data science code than
+it was in JavaScript. Any run of an R program seems to trigger \eval, more
+over package programmers use \eval liberally. On the other hand, the authors
+of Kaggle packages do not use the feature at all.  This point to a
+bifurcated user community, experts who write package know how to use \eval
+while end-users do not. An alternative explanation is that end-users do not
+need \eval for simple data analysis tasks. This bifurcation was not observed
+in the JavaScript community, but this may have changed with the advent of
+Node.js ecosystem.  The distribution of input sizes is similar to that found
+in JavaScript, the majority of inputs being rather small.
 
@@
 
-\begin{figure}
+\begin{figure}[!h]
     \includegraphics[width=\columnwidth]{polymorphism}

### 2020-08-25 inst/paper/main.tex
strucutre

@@
 
-To estimate the complexity of the code being evaluated,
+To roughly estimate the complexity of the code being evaluated,
 Fig.~\ref{fig:ast-size} shows the distribution of the abstract syntax tree
@@
 
-\medskip\noindent\emph{Discussion:} Comparing the use of \eval in R and JavaScript is
-instructive. The use of \eval is more widespread in data science code than
-it was in JavaScript. Any run of an R program seems to trigger \eval, more
-over package programmers use \eval liberally. On the other hand, the authors
-of Kaggle packages do not use the feature at all.  This point to a
-bifurcated user community, experts who write package know how to use \eval
-while end-users do not. An alternative explanation is that end-users do not
-need \eval for simple data analysis tasks. This bifurcation was not observed
-in the JavaScript community, but this may have changed with the advent of
-Node.js ecosystem.  The distribution of input sizes is similar to that found
-in JavaScript, the majority of inputs being rather small.
+\medskip\noindent\emph{Discussion:} Comparing the use of \eval in R and
+JavaScript is instructive. The use of \eval is more widespread in data
+science code than it was in JavaScript. Any run of an R program seems to
+trigger \eval, more over package programmers use \eval liberally. On the
+other hand, the authors of Kaggle packages do not use the feature at all.
+This point to a bifurcated user community, experts who write package know
+how to use \eval while end-users do not. An alternative explanation is that
+end-users do not need \eval for simple data analysis tasks. This bifurcation
+was not observed in the JavaScript community, but this may have changed with
+the advent of Node.js ecosystem.  The distribution of input sizes is similar
+to that found in JavaScript, the majority of inputs being rather small.
 
-\subsection{How are they used}
+{\bf Missing from this section is the measure of the number of events
+  happening in an eval}
 
-- classify evals based on what they can be replaced with
-
-\subsubsection{A taxonomy of \c{eval}}
+\section{A taxonomy of eval}
 
@@
 
-\subsubsection{\eval and environments}
+\subsection{Operation mix}
 
-One of the main differences of the R \eval with the javascript one is how it interacts with \emph{scope}, or environments in R.
-In Javascript, \c{eval} can access the local scope and the global scope.
-In R, \c{eval} has an argument \c{envir} to indicate in which environment its first argument, after being resolved, must be evaluated.
+Here we should show some information about what happens during an eval. What kinds of operations...
 
-By default, it is equal to \c{parent.frame()}, which is the parent environment of the \c{eval} call.
+One of the main differences of the R \eval with the javascript one is how it
+interacts with \emph{scope}, or environments in R. In Javascript, \c{eval}
+can access the local scope and the global scope. In R, \c{eval} has an
+argument \c{envir} to indicate in which environment its first argument,
+after being resolved, must be evaluated.
 
-\subsection{How to replace evals}
+By default, it is equal to \c{parent.frame()}, which is the parent
+environment of the \c{eval} call.
 
-We look at how \emph{polymorphic} the \c{expr} argument of \eval can be, \ie how many different types of the resolved \c{expr} there are per call sites, in Figure~\ref{fig:polymorphism}.
-\DegreeMonomorphism of the call sites are \emph{monomorphic}.
+\subsection{Scope}
+
+Here we talk about scopes and envs
+
+\subsection{Patterns}
+
+Here we categorize evals in some common groups
+
+\subsection{Provenance}
+
+Here we talk about where the input of an eval comes from
+
+\subsection{Consistency}
+
+Here we talk about whether the same call site sees the same pattern
+
+{\bf Polymorphic may not be the right word.}
+
+We look at how \emph{polymorphic} the \c{expr} argument of \eval can be, \ie
+how many different types of the resolved \c{expr} there are per call sites,
+in Figure~\ref{fig:polymorphism}. \DegreeMonomorphism of the call sites are
+\emph{monomorphic}.
 
@@
 \end{figure}
-
-\subsection{How dangerous is eval}
-
-\subsection{How many evals can be replaced}
 

### 2020-08-25 inst/analysis/analysis_stable.Rmd
With tags

@@
 # This makes it possible to generate the html file in the html subfolder
-knit: (function(inputFile, encoding) {
-  rmarkdown::render(inputFile, encoding = encoding, output_dir = "html") })
+# knit: (function(inputFile, encoding) {
+#   rmarkdown::render(inputFile, encoding = encoding, output_dir = "html") })
 ---
@@
 library(fs)
+library(ggthemes)
 
@@
 source("inc/latextags.R")
+source("inc/setup.R")
+theme_set(theme_minimal())
 create_tags(path(TAGS_DIR, "analysisStable.tex"), prefix="", default=TRUE)
 
-data_filepath <- EVALS_SUM_PKGS_FILE
-#data_filepath <- "/var/lib/R/project-evalR/run/package-evals-traced-all.1/calls.fst"
-# The trace data on all CRAN does not have all the columns of the smaller corpus. The analysis will fail for the big corpus.
+packages_filepath <- EVALS_SUM_PKGS_FILE
+core_filepath <- EVALS_SUM_CORE_FILE
 ```
@@
 
-The analysis is being performed on dataset read from file `r data_filepath`.
+The analysis is being performed on dataset read from files `r packages_filepath` and `core_filepath`.
 
-```{r file-info, echo=FALSE}
-file_info(data_filepath)
+```{r file-info-packages, echo=FALSE}
+file_info(packages_filepath)
+```
+
+```{r file-info-core, echo=FALSE}
+file_info(core_filepath)
 ```
@@
 ```{r load-data, echo=FALSE}
-eval_calls <-
-    read_fst(data_filepath) %>%
+eval_calls_packages <-
+    read_fst(packages_filepath) %>%
     as_tibble()  %>% 
-    mutate(eval_call_package = extract_package_name(eval_call_srcref, file))
+    mutate(dataset = "packages")
+
+eval_call_core <- 
+  read_fst(packages_filepath) %>%
+  as_tibble() %>%
+  mutate(dataset = "core")
+
+eval_calls <- bind_rows(eval_calls_packages, eval_call_core)
 
@@
 ```{r parse_in_base}
-parsed_base <- eval_calls %>% filter(!is.na(expr_parsed_expression), eval_call_package == "base")
+parsed_base <- eval_calls %>% filter(!is.na(expr_parsed_expression), eval_source == "base")
 
@@
 
-There are `r count(parsed_base, wt = nb_ev_calls)` `eval` calls with `parse` in *base*, which is `r count(parsed_base, wt = nb_ev_calls) / count(filter(eval_calls, eval_call_package == "base"), wt=nb_ev_calls)`%.
+There are `r count(parsed_base, wt = nb_ev_calls)` `eval` calls with `parse` in *base*, which is `r count(parsed_base, wt = nb_ev_calls) / count(filter(eval_calls, eval_source == "base"), wt=nb_ev_calls)`%.
 
 ## Size of expressions as arguments to `eval`
+
+### String size
 
@@
 
+### AST size
+
+```{r}
+ast_sizes <- eval_calls %>% slice_sample(n=10000) %>% mutate(expr_resolved_ast_size = map_int(expr_resolved, expr_size_str))
+```
+
+```{r}
+# ast_sizes %>% ggplot(aes(x = expr_resolved_ast_size, weight = nb_ev_calls, group = dataset)) +
+#   geom_histogram(aes(y = stat(width*density)), na.rm = TRUE, binwidth = 0.1) +
+#   scale_x_log10() +  scale_y_sqrt(labels=scales::percent) +
+#   labs(x = "AST size", y = "relative frequencies")
+
+# ast_sizes %>% 
+#   ggplot(aes(x = dataset, y = expr_resolved_ast_size, fill = dataset, weight = nb_ev_calls)) +
+#   geom_flat_violin(trim=FALSE, width=.5) +
+#   geom_dotplot(
+#     binaxis="y", 
+#     dotsize=0.01, 
+#     stackdir="down", 
+#     binwidth=0.01, 
+#     position=position_nudge(-0.025)) +
+#   scale_y_log10(labels = scales::comma) +
+#   labs(y="AST size (log)") +
+#   theme(
+#     legend.box="horizontal",
+#     axis.title.x=element_blank()
+#   )
+
+ast_sizes %>%
+  ggplot(aes(x = dataset, y = expr_resolved_ast_size, fill = dataset, weight = nb_ev_calls)) +
+  geom_violin(trim = FALSE, width=.5)+
+  geom_boxplot(width=.05) +
+  scale_y_log10(labels = scales::comma) +
+  labs(y="AST size (log)") +
+  theme(
+    legend.box="horizontal",
+    axis.title.x=element_blank()
+  )
+
+ggsave(path(PLOT_DIR, "ast_sizes.pdf"))
+```
+
 ## Amount of computations performed in `eval`
@@
 ## Scope (environments)
+
+One of the main differences of the R \eval with the javascript one is how it interacts with \emph{scope}, or environments in R.
+In Javascript, \c{eval} can access the local scope and the global scope.
+In R, \c{eval} has an argument \c{envir} to indicate in which environment its first argument, after being resolved, must be evaluated.
+
+By default, it is equal to \c{parent.frame()}, which is the parent environment of the \c{eval} call.
+
+```{r default_environment}
+default_envir <- eval_calls %>% filter(is.na(envir_expression)) 
+nb_default_envir <- default_envir %>% count(wt = nb_ev_calls)
+nb_call_sites_default_envir <- nb_eval_call_sites(default_envir)
+```
+
+`r ratio(nb_default_envir, nb_eval_calls)`% of the calls use the default environment, which corresponds to `r ratio(nb_call_sites_default_envir, nb_call_sites)`% of the call sites.
+
+We also classify environments in the following way:
+
+- `base`: environment of a primitive function of *base* package
+- `empty`: empty environment
+- `callee`: same environment (it seems it only happens with `local` )
+- `global`: ` .GlobalEnv` 
+- `package:package_name`: package environment
+- `loop`: there was a loop in the frame
+
+```{r environment_hierarchy}
+# classified_envir <- eval_calls %>% filter(!is.na(envir_expression)) %>%
+#       mutate(env_class = pmap(list(environment_class, envir_type, envir_expression), extract_envir)) %>% 
+#       unnest_wider(env_class)
+
+#TODO
+```
 
@@
 nb_monomorphic <- expr_resolved_monomorphic %>% nrow()
+
+r("nb monomorphic", nb_monomorphic)
+r("percent monomorphic", ratio(nb_monomorphic, nb_call_sites))
 ```
@@
   labs(title = "Degree of polymorphism", x = "degree", y = "n")
+
+ggsave(path(PLOT_DIR, "polymorphism.pdf"))
 ```
@@
 nb_parsed_evals <- parsed_evals %>% count(wt = nb_ev_calls)
+
+r("percent parsed evals",  ratio(nb_parsed_evals, nb_eval_calls))
 ```
@@
 nb_parsed_call_sites <- nb_eval_call_sites(parsed_evals)
+
+r("percent parsed call sites", ratio(nb_parsed_call_sites, nb_call_sites))
 ```
@@
 
-  We can differentiate further the provenance, i.e.
-whether the evaluated string comes from `file` or from `text`.
-However, we can only do it for `parse`, `str2lang` and `str2expression` at the top level of `expr_expression`, not buried in a function call.
+  We can differentiate further the provenance, i.e. whether the evaluated string comes from `file` or from `text`.
 
@@
 
+We also look to the occurences of `paste` and `str_c` in the `parse` argument. It is an indication that meta-pogramming happens.
+
+```{r parse_meta_prog}
+meta_parsed <- parsed_texts %>% filter(str_detect(expr_parsed_expression, "paste|str_c"))
+
+nb_meta_parsed <- meta_parsed %>% count(wt = nb_ev_calls)
+nb_meta_parsed_call_sites <- nb_eval_call_sites(meta_parsed)
+
+r("paste parsed evals", ratio(nb_meta_parsed, nb_parsed_evals))
+r("paste parsed call site", ratio(nb_meta_parsed_call_sites, nb_parsed_call_sites))
+```
+
 ## Consistence

### 2020-08-25 inst/analysis/enrich_dataset.Rmd
With tags

@@
 title: "Enrich the datasets with new columns."
-output:
-  rmdformats::readthedown:
-        lightbox: true
-        gallery: false
-        df_print: paged
-        toc_depth: 3
+output: html_document
 date: "`r Sys.Date()`"
@@
   base_dir: /var/lib/R/project-evalR
-knit: (function(inputFile, encoding) {
-  rmarkdown::render(inputFile, encoding = encoding, output_dir = "html") })
 ---
@@
 library(fs)
+library(parallel)
 
@@
 
+```{r ast_size}
+eval_calls <- eval_calls %>% add_ast_size()
+```
+
+```{r add_package}
+eval_calls <- eval_calls %>% add_package()
+```
+
 ```{r separate_datasets}

### 2020-08-25 inst/analysis/insights.R

@@ -325,6 +325,28 @@ expr_depth_str <- function(expr)
   return(expr_depth(e))
 }
 
+expr_size <- function(expr)
+{
+  if(is.call(expr))
+  {
+    return(1L + sum(map_int(expr[-1], expr_size)))
+  }
+  else if(is.expression(expr))
+  {
+    return(sum(map_int(expr, expr_size)))
+  }
+  else
+  {
+    return(1L)
+  }
+}
+
+expr_size_str <- function(expr)
+{
+  e <- get_expr(expr)
+  return(expr_size(e))
+}
+
 groupify_function <- function(expr_function)
 {
   case_when(
@@ -334,4 +356,25 @@ groupify_function <- function(expr_function)
   )
 }
 
+extract_write_envir <- function(env_class)
+{
+  return(str_split(env_class, fixed("+"), n = 2)[[1]])
+}
+
+extract_envir <- function(env_class, envir_type, envir_expression)
+{
+  # First element is write, second is read
+  case_when(
+    is.na(env_class) && envir_type == "VECSXP" ~ c("list", "enclos"),
+    is.na(env_class) && envir_type == "NILSXP" ~ c("NULL", "enclos"),
+    is.na(env_class) && envir_type == "INTSXP" ~ {e <- paste0("sys.call(", envir_expression, ")"); c(e, e)},
+    str_ends(env_class, fixed("global")) ~ c(extract_write_envir(env_class), "global"),
+    str_ends(env_class, fixed("base")) ~ c(extract_write_envir(env_class), "base"),
+    str_ends(env_class, fixed("callee")) ~ c(extract_write_envir(env_class), "callee"),
+    str_ends(env_class, fixed("empty")) ~ c(extract_write_envir(env_class), "empty"),
+    str_ends(env_class, "caller-.*") ~ c(extract_write_envir(env_class), str_extract(env_class, "caller-[0-9]*")), 
+    str_ends(env_class, "loop") ~ c("loop", "loop"),
+    str_ends(env_class, "package:.*") ~ c(extract_write_envir(env_class), str_extract(env_class, "package:.*"))
+  )
+}
 

### 2020-08-25 inst/analysis/preprocess.R

@@ -11,6 +11,9 @@
 
 source("insights.R")
 
+library(fs)
+library(multidplyr)
+
 deduplicate <- function(dataset)
 {
   return(dataset %>% count(across(c(-eval_call_id, -starts_with("caller_stack"))), name = "nb_ev_calls"))
@@ -36,7 +39,7 @@ add_parse_args <- function(dataset)
 eval_base_functions <- c("autoload", "autoloader", "bquote", "by.default", "by.data.frame", "invokeRestartInteractively", "Ops.data.frame", "dget", 
                          "eval", "eval.parent", "evalq", "local", "with.default", "within.data.frame", "within.list", "replicate", "subset.data.frame",
                          "subset.matrix", "transform.data.frame", "match.arg", "char.expand", "max.col", "parseNamespaceFile", "source", "sys.source",
-                         "stopfinot", "as.data.frane.table", "match.fun", "trace", "untrace", ".doTrace")
+                         "stopfinot", "as.data.frane.table", "match.fun", "trace", "untrace", ".doTrace", "Vectorize")
 
 # TODO: add support for package name extraction from caller_stack_expression for when 
 # eval_call_srcref is NA
@@ -63,3 +66,25 @@ add_eval_source_type <- function(dataset)
                                                TRUE ~ "package" )))
 }
 
+
+add_ast_size <- function(dataset)
+{
+  cluster <- new_cluster(parallel::detectCores() - 10)
+  cluster_copy(cluster, "get_expr")
+  cluster_copy(cluster, "expr_size")
+  cluster_copy(cluster, "expr_size_str")
+  cluster_library(cluster, "tidyverse")
+  dataset_c <- dataset %>% group_by(eval_call_srcref) %>% partition(cluster)
+  return(dataset_c %>% 
+           mutate(expr_resolved_ast_size = map_int(expr_resolved, expr_size_str)) %>%
+           collect()
+         )
+}
+
+add_package <- function(dataset)
+{
+  mutate(
+    dataset, 
+    package = basename(dirname(dirname(file)))
+  )
+}

### 2020-08-25 inst/analysis/analysis_stable.Rmd
Fix usage of ratio

@@
 
-r("percent parsed evals",  ratio(nb_parsed_evals, nb_eval_calls))
+r("percent parsed evals",  ratio(nb_parsed_evals$n, nb_eval_calls))
 ```
@@
 
-r("paste parsed evals", ratio(nb_meta_parsed, nb_parsed_evals))
+r("paste parsed evals", ratio(nb_meta_parsed$n,  nb_parsed_evals$n))
 r("paste parsed call site", ratio(nb_meta_parsed_call_sites,  nb_parsed_call_sites))

### 2020-08-25 inst/paper/main.tex
Provenance: parse and metaprogramming

@@
 \input{tag/corpus.tex}
+\input{tag/analysisStable.tex}
 
@@
 
+\subsection{Scope}
+
+Here we talk about scopes and envs
+
 One of the main differences of the R \eval with the javascript one is how it
@@
 
-\subsection{Scope}
-
-Here we talk about scopes and envs
-
 \subsection{Patterns}
@@
 
-Here we talk about where the input of an eval comes from
+Here we talk about where the input of an eval comes from.
+
+Although \c{eval} takes an R expression, it can evaluate a string if this string is first parsed with \c{parse}, \c{str2expression} or \c{str2lang}.
+This happens for \PercentParsedEvals of the eval calls, and \PercentParsedCallSites of the call sites.
+
+This corresponds to strings loaded from a file or to metaprogramming, \ie building new variables or functions by concatenating strings together.
+To detect this kind of metaprogramming, we look at the usage of the main concatenating functions in R, \c{paste}, \c{paste0} and \c{str_c}, and \PasteParsedEvals of the evals and \PasteParsedCallSite of the call sites using a parse function actually call a concatenating function.
 
@@
 how many different types of the resolved \c{expr} there are per call sites,
-in Figure~\ref{fig:polymorphism}. \DegreeMonomorphism of the call sites are
+in Figure~\ref{fig:polymorphism}. \PercentMonomorphic of the call sites are
 \emph{monomorphic}.

### 2020-08-25 inst/paper/main.tex
Adding an example of paste in parse

@@
 
-This corresponds to strings loaded from a file or to metaprogramming, \ie building new variables or functions by concatenating strings together.
+This corresponds to strings loaded from a file or to metaprogramming, \ie building new variables or functions by concatenating strings together, as in Listing~\ref{code:parsepaste}.
 To detect this kind of metaprogramming, we look at the usage of the main concatenating functions in R, \c{paste}, \c{paste0} and \c{str_c}, and \PasteParsedEvals of the evals and \PasteParsedCallSite of the call sites using a parse function actually call a concatenating function.
+
+\begin{lstlisting}[caption={Using \c{eval} and \c{parse} to do meta-programming in package \emph{data.table}, to build the options at load time.}, label=code:parsepaste]
+eval(parse(text=paste0("options(",i,"=",opts[i],")")))
+\end{lstlisting}
 

### 2020-08-25 inst/analysis/corpus-analysis.Rmd
V.A

--- 2020-08-24 inst/analysis/corpus-analysis.Rmd
@@
 ```{r setup, include=FALSE}
-for (x in c(
+local(for (x in c(
   "dplyr", 
@@
   suppressPackageStartupMessages(library(x, character.only=TRUE))
-}
+})
 
@@
       starts_with("funs_"),
+      "evals"
     ), 
@@
 
+```{r compute corpus programs}
+corpus_programs <-
+  left_join(
+    filter(runs, exitval==0) %>% select(package, file, time), 
+    runnable_code_metadata %>% select(package, file, code),
+    by=c("package", "file")
+  ) %>%
+  transmute(
+    package,
+    file,
+    type=basename(dirname(file)),
+    code, 
+    runtime=time
+  )
+```
+
+There will be some programs that were not strictly part of examples, tests or vignettes, or they do not have
+counts for code size or for execution time.
+
+```{r compute program inconsistencies}
+corpus_programs_inconsistent <-
+  corpus_programs %>%
+  filter(
+    !(type %in% c("examples", "tests", "vignettes")) |
+    is.na(code) | 
+    is.na(runtime) |
+    code == 0
+  )
+
+corpus_programs_inconsistent %>% 
+    my_datatable()
+```
+
+Since, there are just a few of them, we ignore them.
+
+```{r filter out inconsistent programs}
+corpus_programs <- anti_join(corpus_programs, corpus_programs_inconsistent, by=c("package", "file"))
+```
+
 ```{r summary}
-corpus_runtime  <- corpus$runnable_time / corpus$runnable_files
-runnable_code   <- corpus$runnable_code
-corpus_programs_files <- corpus$runnable_files
+runnable_code <- 
+  corpus_programs %>%
+  mutate(files=1) %>%
+  pivot_wider(names_from=type, values_from=c(files, code)) %>%
+  select(-file) %>%
+  group_by(package) %>%
+  summarise_all(sum, na.rm=TRUE) %>%
+  group_by(package) %>%
+  mutate(
+    files=files_examples + files_tests + files_vignettes,
+    code=code_examples + code_tests + code_vignettes
+  )
+
+rs <- function(name, value) {
+  stopifnot(is.numeric(value))
+  bind_rows(
+    r(name, sum(value)),
+    r(paste(name, "mean"), mean(value)),
+    r(paste(name, "median"), median(value)),
+    r(paste(name, "min"), min(value)),
+    r(paste(name, "max"), max(value))
+  )
+}
+
+corpus_program_examples <- filter(corpus_programs, type=="examples")
+corpus_programs_tests <- filter(corpus_programs, type=="tests")
+corpus_programs_vignettes <- filter(corpus_programs, type=="vignettes")
 
@@
   r("packages",    corpus),
+  rs("revdeps", corpus$revdeps),
   r("in cranlogs", filter(corpus, !is.na(rank))),
-  r("package programs",     sum(corpus_programs_files)),
 
-  r("runnable code",        sum(runnable_code)),
-  r("runnable code mean",   mean(runnable_code)),
-  r("runnable code median", median(runnable_code)),
+  rs("package programs",         runnable_code$files),
+  rs("package programs code",    runnable_code$code),
+  rs("package programs runtime", as.duration(runnable_code$runtime)),
 
-  r("examples code",        sum(corpus$runnable_code_examples)),
-  r("tests code",           sum(corpus$runnable_code_tests)),
-  r("vignettes code",       sum(corpus$runnable_code_vignettes)),
+  rs("examples programs",  runnable_code$files_examples),
+  rs("tests programs",     runnable_code$files_tests),
+  rs("vignettes programs", runnable_code$files_vignettes),
+
+  rs("examples code",  corpus_program_examples$code),
+  rs("tests code",     corpus_programs_tests$code),
+  rs("vignettes code", corpus_programs_vignettes$code),
+
+  rs("examples runtime",  corpus_program_examples$runtime),
+  rs("tests runtime",     corpus_programs_tests$runtime),
+  rs("vignettes runtime", corpus_programs_vignettes$runtime),
 
@@
 
-  r("code",                 sum(corpus$package_code)),
-  r("R code",               sum(corpus$package_r_code)),
-  r("native code",          sum(corpus$package_native_code)),
+  rs("code",        corpus$package_code),
+  rs("R code",      corpus$package_r_code),
+  rs("native code", corpus$package_native_code),
 
-  r("functions",            sum(corpus$funs_public+corpus$funs_private)),
-  r("function with eval",   sum(corpus$funs_with_eval)),
-  r("public functions",     sum(corpus$funs_public)),
-  r("private functions",    sum(corpus$funs_private)),
+  rs("functions",          corpus$funs_public + corpus$funs_private),
+  rs("function with eval", corpus$funs_with_eval),
+  rs("public functions",   corpus$funs_public),
+  rs("private functions",  corpus$funs_private),
 
-  r("mean revdes",          mean(corpus$revdeps)),
-  r("median revdes",        median(corpus$revdeps)),
-
-  r("runtime",              as.duration(sum(corpus$runnable_time))),
-  r("mean runtime",         as.duration(mean(corpus_runtime))),
-  r("median runtime",       as.duration(median(corpus_runtime)))
+  rs("revdes", corpus$revdeps)
 )
@@
 revdeps_median  <- median(corpus$revdeps)
+funs_with_eval_mean    <- mean(corpus$funs_with_eval)
+funs_with_eval_median  <- median(corpus$funs_with_eval)
+
+size_bins       <- c(0L, 1e3L, 5e3L, 1e4L, 5e4L, 1e5L)
+size_bins_lasb  <- c("< 1000", "1,000 - 4,999", "5,000 - 9,999", "10,000 - 49,999", "> 50,000")
 
@@
   mutate(
-    outlier=package_code > PACKAGE_SIZE_OUTLIER,
-    label=ifelse(rank <= 25 | outlier, package, NA)  #if_else(outlier, str_glue("{package}"), as.character(NA))
+    outlier=funs_with_eval > 30 | package_r_code > PACKAGE_SIZE_OUTLIER,
+    label=ifelse(rank <= MOST_DOWNLOADED | outlier, package, NA),
+    code_size=cut(package_r_code, size_bins, labels=size_bins_lasb, include.lowest=TRUE)
   ) %>%
@@
       x=coverage_expr, 
-      y=revdeps,
+#      y=funs_with_eval,
+      y=evals,
       label=label,
-      size=package_code
+      size=code_size
     )
   ) +
-  geom_point(alpha=.6) +
-  geom_text(size=2, check_overlap = T, vjust=1.5, na.rm = TRUE) +
-
-  geom_hline(aes(yintercept=revdeps_mean), linetype="dashed", color="red", size=0.2) +
+  geom_point(alpha=.4) +
+  geom_text(size=3, check_overlap = T, vjust=-1.5, na.rm = TRUE) + 
+  geom_hline(aes(yintercept=funs_with_eval_mean), linetype="dashed", color="red", size=0.2) +
   geom_vline(aes(xintercept=coverage_mean), linetype="dashed", color="red", size=0.2) +
-
   scale_x_continuous(labels = scales::percent) + 
-  scale_y_log10(labels = scales::comma) +
-  scale_size(range = c(.1, 3), labels = scales::comma) +
+  scale_y_continuous(trans=scales::pseudo_log_trans(), breaks=c(0, 1, 2, 5, 10, 20, 50, 100, 200, 500)) +
+  scale_size_discrete(range = c(1, 6)) +
   theme(
@@
     x="Code coverage", 
-    y="Number of reverse dependencies (log)", 
-    size="Code size (R + native)",
+    y="Number of eval call sites", 
+    size="Lines of R code",
     linetype=""
@@
     size=guide_legend(order=1),
-    color=guide_legend(order=2),
-    linetype=F
+    color=FALSE,
+    linetype=FALSE
   )
@@
 
-```{r}
-kaggle_runnable <- filter(kaggle, runnable)
-kaggle_finished <- filter(kaggle, exitval==0)
-kaggle_code     <- kaggle_finished$code
-kaggle_runtime  <- kaggle_finished$runtime
+```{r compute kaggle programs}
+kaggle_programs <-
+  kaggle %>%
+  filter(exitval==0) %>%
+  transmute(
+    id, 
+    code, 
+    runtime=as.numeric(runtime, units="seconds")
+  )
+```
+
+```{r kaggle summary}
+kaggle_code     <- kaggle_programs$code
+kaggle_runtime  <- kaggle_programs$runtime
 
@@
   r("kaggle",                 kaggle),
-  r("runnable kaggle",        kaggle_runnable),
-  r("finished kaggle",        kaggle_finished),
+  r("runnable kaggle",        filter(kaggle, runnable)),
+  r("finished kaggle",        kaggle_programs),
   r("finished kaggle code",   sum(kaggle_code)),
@@
 
-```{r}
-overview_table(
-  r("all programs", nrow(kaggle_finished) + sum(corpus_programs_files)),
-  r("all runnbale code", sum(kaggle_code) + sum(runnable_code))
+```{r compute programs}
+programs <- bind_rows(
+  corpus_programs %>% mutate(corpus="packages") %>% select(code, runtime, corpus),
+  kaggle_programs %>% mutate(corpus="kaggle") %>% select(code, runtime, corpus)
 )
@@
 
-```{r}
-corpus_programs <-
-  left_join(
-    filter(runs, exitval==0) %>% select(package, file, time), 
-    runnable_code_metadata %>% select(package, file, code),
-    by=c("package", "file")
-  ) %>%
-  transmute(id=paste(package, ":", file), code, runtime=time, corpus="packages")
-
-kaggle_programs <-
-  kaggle %>%
-  filter(exitval==0) %>%
-  transmute(id, code, runtime=as.numeric(runtime, units="seconds"), corpus="kaggle")
-
-programs <- bind_rows(corpus_programs, kaggle_programs)
+```{r all summary}
+overview_table(
+  r("all programs",      nrow(kaggle_programs) + nrow(corpus_programs)),
+  r("all runnbale code", sum(kaggle_code) + sum(runnable_code$code)),
+  r("all code",          sum(kaggle_code) + sum(corpus$package_code) + sum(runnable_code$code))
+)
 ```
 
-```{r}
+```{r programs-code, eval=F}
 programs %>%
   ggplot(aes(y=code, x=corpus, fill=corpus)) + 
-  geom_flat_violin(trim=FALSE, width=.5) +
-  geom_dotplot(
-    binaxis="y", 
-    dotsize=0.1, 
-    stackdir="down", 
-    binwidth=0.01, 
-    position=position_nudge(-0.025)) +
+  geom_violin() +
   scale_y_log10(labels = scales::comma) +
@@
 
-```{r}
+```{r programs-runtime, eval=F}
 programs %>%

### 2020-08-25 inst/paper/main.tex
V.A

@@
 \usepackage[usenames,dvipsnames]{xcolor}
+\usepackage{paralist}
 \usepackage{xspace}
@@
 \newcommand{\cf}{\emph{cf.}\xspace}
+\newcommand{\summary}[1]{({\csname #1Min\endcsname} / {\csname #1Mean\endcsname} / {\csname #1Max\endcsname})}
 
@@
 
-\begin{figure*}[!tb]\centering\includegraphics[width=.8\linewidth]
+\begin{figure*}[!tb]\centering\includegraphics[width=\linewidth]
   {corpus.pdf}\caption{Corpus}\label{fig:corpus}
@@
 We have selected the TOP \CorpusPackages packages based on the number of reverse
-dependencies, \ie how many packages depend on them\footnote{extracted from
-  package metadata using builtin function}. Using the number of reverse
-dependencies helps us to choose popular packages which in general should be of a
-better quality and should have better code coverage. The resulting set consists
-of \CorpusRCodeRnd (median=\CorpusRCodeMedian) lines of R code and
-\CorpusNativeCodeRnd (median=\CorpusNativeCodeMedian) lines of native code
-(C/C++/Fortran). Figure~\ref{fig:corpus} shows these packages, the size of the
-dots reflects the project's size in lines of code including both R and native
-code, the x-axis indicates the expression code coverage in percents and the
-y-axis gives the number of reverse dependencies in log scale. Dotted lines
-indicate means. Large (over \CorpusLargePackageTresholdRnd lines of code) and popular (in \CorpusMostDownloadedTreshold most downloaded) packages are annotated.
-
-\begin{figure}[!htb]\centering\includegraphics[width=.9\linewidth]
-  {programs-code.pdf}\caption{Programs code size}\label{fig:programs-code}
-\end{figure}
-
-\begin{figure}[!htb]\centering\includegraphics[width=.9\linewidth]
-  {programs-runtime.pdf}\caption{Programs execution time}\label{fig:programs-runtime}
-\end{figure}
+dependencies, \ie how many other packages depend on them\footnote{Extracted from
+  package metadata using builtin function}. Using this criterion helps us to
+choose popular packages which in general should be of a better quality and
+should have better code coverage. The resulting set has \summary{CorpusRevdeps}
+reverse dependencies (min / mean / max) and consists of \CorpusRCodeRnd lines of
+R code and \CorpusNativeCodeRnd lines of native code (C/C++/Fortran).
+Figure~\ref{fig:corpus} shows these packages, the size of the dots reflects the
+project's size in lines of code including both R and native code, the x-axis
+indicates the expression code coverage in percents and the y-axis gives the
+number of reverse dependencies in log scale. Dotted lines indicate means. Large
+(over \CorpusLargePackageTresholdRnd lines of code) and popular (in
+\CorpusMostDownloadedTreshold most downloaded) packages are annotated.
 
@@
 package must abide to a number of well-formedness rules that are automatically
-checked asserting certain quality.
+checked asserting certain quality. Most relevant for this work is that all of
+the \emph{runnable code} is tested and only a successfully running package is
+admitted in the archive. This is important as we use this runnable code for the
+dynamic analysis of eval calls.
 
-Next to code and tests, R packages also contain examples and vignettes. Examples
-are runnable snippets extracted from documentation. Vignettes are tutorial-like
-guides into the package functionality. They are written in the literate
-programming style using either Latex or Rmarkdown format with embedded code
-snippets. Since CRAN is a curated repository, upon submission, it checks all the
-runnable code in tests, examples and vignettes and only a successfully running
-package is admitted to the archive. We use the same mechanism to extract
-runnable code and run it for the dynamic analysis.
+There are three sources of runnable code in a R package. Next to the package
+\emph{tests}, there are: \emph{examples} and \emph{vignettes}. Examples are R
+code snippets extracted from package documentation into scripts file. Vignettes
+are long-form guides into package functionality written using
+literate-programming style in \LaTeX\xspace or Rmarkdown with chunks of R code
+that can be similarly extracted into R scripts. R provides builtin functions for
+tangling the embedded code into files.
 
-In the selected packages, there are \CorpusRunnableCodeRnd lines of runnable
-code in examples (\CorpusExamplesCodeRnd), tests (\CorpusTestsCodeRnd) and
-vignettes (\CorpusVignettesCodeRnd). Running this code results on average in
-\CorpusMeanExprCoverage package code coverage. The code coverage was computed
-for each package using
+The selected packages contain \CorpusPackagePrograms programs with
+\CorpusPackageProgramsCodeRnd lines of R code. Most programs are examples
+(\CorpusExamplesPrograms) with \CorpusExamplesCodeRnd lines of code, followed by
+tests (\CorpusTestsPrograms) with \CorpusTestsCodeRnd lines of code and
+vignettes (\CorpusVignettesPrograms) with \CorpusVignettesCodeRnd lines of code.
+
+Running this code results on average in \CorpusMeanExprCoverage package code
+coverage. The code coverage was computed for each package using
+Covr\footnote{\url{ https://github.com/r-lib/covr}}, the R code coverage tool.
 

### 2020-08-25 inst/paper/main.tex
V.B

@@
 \newcommand{\summary}[1]{({\csname #1Min\endcsname} / {\csname #1Mean\endcsname} / {\csname #1Max\endcsname})}
+\newcommand{\summaryrnd}[1]{({\csname #1MinRnd\endcsname} / {\csname #1MeanRnd\endcsname} / {\csname #1MaxRnd\endcsname})}
 
@@
 indicates the expression code coverage in percents and the y-axis gives the
-number of reverse dependencies in log scale. Dotted lines indicate means. Large
-(over \CorpusLargePackageTresholdRnd lines of code) and popular (in
-\CorpusMostDownloadedTreshold most downloaded) packages are annotated.
+number of reverse dependencies in log scale. Dotted lines indicate means.
+Packages with over \CorpusEvalsPackageTreshold eval call sites are annotated.
 
-These packages come from the Comprehensive R Archive Network
+All included packages come from the Comprehensive R Archive Network
 (CRAN\footnote{\url{http://cran.r-project.org}}), the largest repository of R
@@
 
-\subsection{Kaggle sripts}
-% contrast to CRAN - duplication
+\subsection{Kaggle Scripts}
+
+To represent user-written code in the corpus, we truned to
+Kaggle\footnote{\url{https://www.kaggle.com}}, an online platform for
+data-science and machine-learning practitioner. Among others, the website allow
+people to share data-science competitions, a data-analysis problem together with
+data, for which users try to find the best solution. The solutions, called
+\emph{kernels}, are then shared back to the platform in the form of either plain
+scripts or as Rmarkdown / Jupyter notebooks written in either Julia, Python, R
+or SQLite.
+
+One of the most popular competition is about predicting passenger survival on
+Titanic\footnote{\url{https://www.kaggle.com/c/titanic}} with \CorpusKaggle
+kernels in R (over 1/4 of all available R kernels) which we used for our corpus.
+
+Unlike CRAN, Kaggle is not a curated repository and therefore there are no
+guarantees about the quality of the code. After downloading all of the
+\CorpusKaggle kernels and extracting the R code from the various
+formats\footnote{We use \texttt{rmarkdown} package to convert from Jupyer
+  notebook to Rmarkdown as well as to convert from Rmarkdown to plain R}, we
+found that \CorpusDuplicatedKaggle were whole-file duplicates. From
+\CorpusRunnableKaggle kernels, \CorpusFailedKaggle failed to execute. Next to
+various runtime exceptions, common problems were parse errors and misspelled
+package names.
+
+The final set contains \CorpusFinishedKaggle kernels with
+\CorpusFinishedKaggleCodeRnd \summaryrnd{CorpusFinishedKaggleCode} lines of R
+code.
 

### 2020-08-25 inst/analysis/inc/paths.R

@@ -64,7 +64,7 @@ WHICH_DEPENDENCIES <- c("Depends", "Imports")
 CRAN_MIRROR_LOCAL_URL <- paste0("file://", params$base_dir, "/CRAN")
 # size of a package to be considered as outlier
 PACKAGE_SIZE_OUTLIER <- 50000
-# 
+PACKAGE_EVALS_OUTLIER <- 50
 MOST_DOWNLOADED <- 25
 # R packages distributed with vanilla R
 CORE_PACKAGES <- c(

### 2020-08-25 inst/paper/main.tex
Precised the provenance

--- 2020-08-25 inst/paper/main.tex
@@
 
-\begin{lstlisting}[caption={Using \c{eval} and \c{parse} to do meta-programming in package \emph{data.table}, to build the options at load time.}, label=code:parsepaste]
+\begin{lstlisting}[caption={Using \c{eval} and \c{parse} to do meta-programming in package \emph{data.table}, to build the options at load time (in \emph{onLoad.R}).}, label=code:parsepaste]
 eval(parse(text=paste0("options(",i,"=",opts[i],")")))
 \end{lstlisting}
+
+\textbf{What about a table of packages ranked by usage of parse (and/or paste...)?}
 

### 2020-08-25 inst/analysis/corpus-analysis.Rmd

@@
   r("most downloaded treshold", MOST_DOWNLOADED),
-  r("large package treshold", PACKAGE_SIZE_OUTLIER)
+  r("large package treshold", PACKAGE_SIZE_OUTLIER),
+  r("evals package treshold", PACKAGE_EVALS_OUTLIER)
 )
@@
   mutate(
-    outlier=funs_with_eval > 30 | package_r_code > PACKAGE_SIZE_OUTLIER,
-    label=ifelse(rank <= MOST_DOWNLOADED | outlier, package, NA),
+    outlier=evals > PACKAGE_EVALS_OUTLIER,# | package_r_code > PACKAGE_SIZE_OUTLIER,
+    label=ifelse(outlier, package, NA),
     code_size=cut(package_r_code, size_bins, labels=size_bins_lasb, include.lowest=TRUE)
@@
   r("kaggle",                 kaggle),
+  r("duplicated kaggle",      nrow(kaggle)-nrow(filter(kaggle, runnable))),
   r("runnable kaggle",        filter(kaggle, runnable)),
   r("finished kaggle",        kaggle_programs),
-  r("finished kaggle code",   sum(kaggle_code)),
-  r("finished kaggle mean",   mean(kaggle_code)),
-  r("finished kaggle median", median(kaggle_code)),
-  r("kaggle runtime",         as.duration(sum(kaggle_runtime))),
-  r("kaggle mean",            as.duration(mean(kaggle_runtime))),
-  r("kaggle median",          as.duration(median(kaggle_runtime)))
+  r("failed kaggle",          nrow(filter(kaggle, runnable))-nrow(kaggle_programs)),
+  rs("finished kaggle code",   kaggle_code),
+  rs("kaggle runtime",         as.duration(kaggle_runtime)),
 )

### 2020-08-25 inst/paper/main.tex
V

@@
 choose popular packages which in general should be of a better quality and
-should have better code coverage. The resulting set has \summary{CorpusRevdeps}
-reverse dependencies (min / mean / max) and consists of \CorpusRCodeRnd lines of
-R code and \CorpusNativeCodeRnd lines of native code (C/C++/Fortran).
+should have better code coverage. The resulting set has
+\summary{CorpusRevdeps}\footnote{Such triples represent ($min$ / $mean$ /
+  $max$)} reverse dependencies and consists of \CorpusRCodeRnd lines of R code
+and \CorpusNativeCodeRnd lines of native code (C/C++/Fortran).
 Figure~\ref{fig:corpus} shows these packages, the size of the dots reflects the
@@
 There are three sources of runnable code in a R package. Next to the package
-\emph{tests}, there are: \emph{examples} and \emph{vignettes}. Examples are R
-code snippets extracted from package documentation into scripts file. Vignettes
-are long-form guides into package functionality written using
+\emph{tests}, there are also \emph{examples} and \emph{vignettes}. Examples are
+R code snippets extracted from package documentation into scripts file.
+Vignettes are long-form description of package functionality written using
 literate-programming style in \LaTeX\xspace or Rmarkdown with chunks of R code
@@
 The selected packages contain \CorpusPackagePrograms programs with
-\CorpusPackageProgramsCodeRnd lines of R code. Most programs are examples
-(\CorpusExamplesPrograms) with \CorpusExamplesCodeRnd lines of code, followed by
-tests (\CorpusTestsPrograms) with \CorpusTestsCodeRnd lines of code and
-vignettes (\CorpusVignettesPrograms) with \CorpusVignettesCodeRnd lines of code.
+\CorpusPackageProgramsCodeRnd lines of R code:
+%
+\begin{compactitem}[$-$]
+  \item \CorpusExamplesProgramsRnd examples with \CorpusExamplesCodeRnd
+  \summaryrnd{CorpusExamplesCode} lines of code,
+  \item \CorpusTestsProgramsRnd tests with \CorpusTestsCodeRnd
+  \summaryrnd{CorpusTestsCode} lines of code, and
+  \item \CorpusVignettesProgramsRnd vignettes with \CorpusVignettesCodeRnd
+  \summaryrnd{CorpusVignettesCode} lines of code.
+\end{compactitem}
 

### 2020-08-25 inst/paper/main.tex
Add glue to the paste functions

@@
 This corresponds to strings loaded from a file or to metaprogramming, \ie building new variables or functions by concatenating strings together, as in Listing~\ref{code:parsepaste}.
-To detect this kind of metaprogramming, we look at the usage of the main concatenating functions in R, \c{paste}, \c{paste0} and \c{str_c}, and \PasteParsedEvals of the evals and \PasteParsedCallSite of the call sites using a parse function actually call a concatenating function.
+To detect this kind of metaprogramming, we look at the usage of the main concatenating functions in R, \c{paste}, \c{paste0}, \c{glue} and \c{str_c}, and \PasteParsedEvals of the evals and \PasteParsedCallSite of the call sites using a parse function actually call a concatenating function.
 

### 2020-08-25 inst/paper/main.tex
Updated the code coverage metric

@@
 
-Running this code results on average in \CorpusMeanExprCoverage package code
-coverage. The code coverage was computed for each package using
-Covr\footnote{\url{ https://github.com/r-lib/covr}}, the R code coverage tool.
+For each package we use its runnable code to compute the package code
+coverage\footnote{Computed using \url{ https://github.com/r-lib/covr}}. On
+average, across the \CorpusPackages packages, it is \CorpusMeanExprCoverage.
 

### 2020-08-25 inst/paper/main.tex
Changes in V

@@
 
-\section{Corpus of R Programs}
+\section{Corpus}
 
-We have assembled a corpus of \CorpusPackages R packages and
-\CorpusFinishedKaggle user-written data-analysis scripts. Together it contains
-\CorpusAllCodeRnd lines of source code\footnote{Excluding comments and blank lines using \url{github.com/AlDanial/cloc}}.
+We now describe the corpus used in this study.
+
+\begin{figure*}[!tb]\centering\includegraphics[width=\linewidth]
+  {corpus.pdf}\caption{CRAN packages}\label{fig:corpus}
+\end{figure*}
 
@@
   $max$)} reverse dependencies and consists of \CorpusRCodeRnd lines of R code
-and \CorpusNativeCodeRnd lines of native code (C/C++/Fortran).
+and \CorpusNativeCodeRnd lines of native code (C/C++/Fortran)\footnote{Excluding comments and blank lines using \url{github.com/AlDanial/cloc}}.
 Figure~\ref{fig:corpus} shows these packages, the size of the dots reflects the
@@
 code.
+
+\subsection{Summary}
+
+Together, we have assembled a corpus of \CorpusPackages R packages and
+\CorpusFinishedKaggle user-written data-analysis scripts. Together it contains
+\CorpusAllCodeRnd lines of source code (both and R and native code). Out of
+this, \CorpusAllRunnbaleCode lines R code spread across \CorpusAllPrograms R
+scripts is the code that we run during the dynamic analysis.
 

### 2020-08-25 inst/paper/main.tex
Updates

@@
 
+This section describes the infrastructure we used to run the dynamic program
+analysis to record the calls to evals.
+
 \begin{figure*}[!tb]\centering\includegraphics[width=\linewidth]
-  {corpus.pdf}\caption{Corpus}\label{fig:corpus}
+  {pipeline.pdf}\caption{Pipeline}\label{fig:pipeline}
 \end{figure*}
@@
 
-We now describe the corpus used in this study.
+The corpus for this study includes R programs extracted from popular R packages
+as well as complete user-written data-analysis scripts. In this section we
+describe the methodology used to collect them.
 
-\begin{figure*}[!tb]\centering\includegraphics[width=\linewidth]
-  {corpus.pdf}\caption{CRAN packages}\label{fig:corpus}
-\end{figure*}
-
-\subsection{CRAN Packages}
+\subsection{CRAN Packages}\label{sec:corpus}
 
@@
   package metadata using builtin function}. Using this criterion helps us to
-choose popular packages which in general should be of a better quality and
-should have better code coverage. The resulting set has
+choose popular packages which in general should be of a better quality and have
+better code coverage and therefore more code to run. The resulting set has
 \summary{CorpusRevdeps}\footnote{Such triples represent ($min$ / $mean$ /
   $max$)} reverse dependencies and consists of \CorpusRCodeRnd lines of R code
-and \CorpusNativeCodeRnd lines of native code (C/C++/Fortran)\footnote{Excluding comments and blank lines using \url{github.com/AlDanial/cloc}}.
+and \CorpusNativeCodeRnd lines of native code (C/C++/Fortran)\footnote{Excluding
+  comments and blank lines using \url{github.com/AlDanial/cloc}}.
 Figure~\ref{fig:corpus} shows these packages, the size of the dots reflects the
@@
 Packages with over \CorpusEvalsPackageTreshold eval call sites are annotated.
+
+\begin{figure*}[!tb]\centering\includegraphics[width=\linewidth]
+  {corpus.pdf}\caption{CRAN packages}\label{fig:corpus}
+\end{figure*}
 

### 2020-08-26 inst/analysis/counts.Rmd
Add potentially useful stuff

@@
 ```{r eval=TRUE}
+run_trace_data <- read_fst(PACKAGE_TRACE_RUNS_FILE)
+parallel_kaggle_data <- read_fst(KAGGLE_LOG_FILE)
+
 core_data <- read_fst(params$core_datafile)
@@
                                 if_else(eval_source == "base", "core",
-                                if_else(package == caller_package, "own",
+                                if_else(package == eval_source, "own",
                                         "dependent"))),
@@
 
+#all_package_programs <- 
+#    run_trace_data %>%
+#    filter(Exitval != 0)
+#    
+#
+#all_kaggle_programs <-
+#    parallel_kaggle_data %>% 
+#    filter(Exitval == 0) %>% 
+#    pull(Command) %>% 
+#    map_chr(function(str) { 
+#        script <- str_split(str, " ")[[1]][4]; 
+#        comps <- path_split(script)[[1]]; 
+#        str_c(comps[8], "/", comps[9]) 
+#    })
+#    
+#all_files <-
+#    rbind(tibble(file = run_trace_data$file, core_own_dependent = "core", corpus = "cran", nb_ev_calls = 0),
+#          tibble(file = run_trace_data$file, core_own_dependent = "own", corpus = "cran", nb_ev_calls = 0),
+#          tibble(file = run_trace_data$file, core_own_dependent = "dependent", corpus = "cran", nb_ev_calls = 0)) %>%
+#    rbind(tibble(file = all_kaggle_programs, core_own_dependent = "core", corpus = "kaggle", nb_ev_calls = 0),
+#          tibble(file = all_kaggle_programs, core_own_dependent = "own", corpus = "kaggle", nb_ev_calls = 0),
+#          tibble(file = all_kaggle_programs, core_own_dependent = "dependent", corpus = "kaggle", nb_ev_calls = 0))
+
 core_own_dependent <-
     all_data %>%
+#        rbind(all_files) %>%
+#        select(file, core_own_dependent, corpus, nb_ev_calls) %>%
         group_by(file, core_own_dependent) %>%

### 2020-08-26 inst/analysis/analysis_stable.Rmd
Reduce the number of unknown evals

@@
         toc_depth: 3
-        self_contained: no
 date: "`r Sys.Date()`"
@@
 eval_call_core <- 
-  read_fst(packages_filepath) %>%
+  read_fst(core_filepath) %>%
   as_tibble() %>%
@@
 ```
+
+```{r unique_calls}
+nb_unique_calls <- nrow(eval_calls)
+
+r("unique calls", nb_unique_calls)
+```
+
+There are `r nb_unique_calls` unique calls;
 
@@
 nb_eval_calls_per_package  %>% datatable()
+```
+
+### Unknown evals
+
+Some `eval` do not have srcref information attached and are not in *base*. It is often because they are located in a function that is created by anotehr function.
+
+```{r unknown_evals}
+undefined_evals <- eval_calls %>% filter(eval_source_type == "<undefined>")
+
+nb_undefined <- undefined_evals %>% count(wt = nb_ev_calls)
+
+nb_different_call_expr <- undefined_evals %>% select(eval_call_expression) %>% n_distinct()
+
+r("undefined evals", nb_undefined$n)
+r("percent undefined eval", ratio(nb_undefined$n, nb_eval_calls))
+r("undefined eval expressions", nb_different_call_expr)
 ```
@@
 
-We also look to the occurences of `paste` and `str_c` in the `parse` argument. It is an indication that meta-pogramming happens.
+We also look to the occurrences of `paste` (and `parse0`), `glue` and `str_c` in the `parse` argument. It is an indication that meta-pogramming happens.
 
 ```{r parse_meta_prog}
-meta_parsed <- parsed_texts %>% filter(str_detect(expr_parsed_expression, "paste|str_c"))
+meta_parsed <- parsed_texts %>% filter(str_detect(expr_parsed_expression, "paste|str_c|glue"))
 
@@
 
+We can look at the most common meta-programming examples:
+
+```{r parse_meta_prog_ex}
+meta_parsed %>% count(expr_parsed_expression, wt = nb_ev_calls, sort = TRUE)
+```
+
 ## Consistence

### 2020-08-26 inst/analysis/insights.R

@@ -155,22 +155,6 @@ check_call <- function(arg)
 }
 
 
-# Transforms a call in a character string in eval(parse(text = str)) into a do.call 
-text_call_to_do_call <- function(eval_call)
-{
-  exp <- get_expr(eval_call)
-  stopifnot(exp[[1]] == "eval")
-  args <- exp[-1]
-  stopifnot(is.call(args[[1]]) & args[[1]][[1]] == "parse")
-  text <- args[[1]]$text
-  
-  # TODO: handle environment
-
-  f_name <- function_name(text)
-  f_args <- function_arguments(text)
-  return(paste0("do.call(", f_name, ", list(", paste0(f_args, collapse = ","), "))"))
-}
-
 SEXP_TYPES <- tribble(~sexp_type, ~name,
                       0, "NILSXP",
                       1, "SYMSXP",
@@ -238,6 +222,28 @@ core_package_files <- map_dfr(CORE_PACKAGES, function(x) {
   tibble(package=x, file=f)
 })
 
+# match eval that are not called (but passed to a higher order function)
+is_eval <- function(s)
+{
+  return(!is.na(s) && str_detect(s, "(eval(q|\\.parent)?|local)[^\\(_\\.q]"))
+}
+
+# To use if srcref is NA and caller_function is not one of the base ones
+# Will not always work but should work for lapply ones
+package_name_from_call_stack <- function(caller_stack_expr, caller_stack_expr_srcref)
+{
+  stack_expr <- str_split(caller_stack_expr, fixed("\n"))
+  stack_srcref <- str_split(caller_stack_expr_srcref, fixed("\n"))
+  
+  eval_pos <- detect_index(stack_expr[[1]], is_eval)
+  if(eval_pos != 0)
+  {
+    srcref <- stack_srcref[[1]][[eval_pos]]
+    return(if(srcref == "NA") "base?" else srcref)
+  }
+  
+  return("base?")
+}
 
 extract_package_name <- function(src_ref, file)
 {
@@ -360,6 +366,7 @@ extract_write_envir <- function(env_class)
 {
   return(str_split(env_class, fixed("+"), n = 2)[[1]])
 }
+
 
 extract_envir <- function(env_class, envir_type, envir_expression)
 {

### 2020-08-26 inst/analysis/preprocess.R

@@ -39,23 +39,59 @@ add_parse_args <- function(dataset)
 eval_base_functions <- c("autoload", "autoloader", "bquote", "by.default", "by.data.frame", "invokeRestartInteractively", "Ops.data.frame", "dget", 
                          "eval", "eval.parent", "evalq", "local", "with.default", "within.data.frame", "within.list", "replicate", "subset.data.frame",
                          "subset.matrix", "transform.data.frame", "match.arg", "char.expand", "max.col", "parseNamespaceFile", "source", "sys.source",
-                         "stopfinot", "as.data.frane.table", "match.fun", "trace", "untrace", ".doTrace", "Vectorize")
+                         "stopifnot", "as.data.frane.table", "match.fun", "trace", "untrace", ".doTrace", "Vectorize")
 
-# TODO: add support for package name extraction from caller_stack_expression for when 
-# eval_call_srcref is NA
-find_package_name <- function(caller_function, srcref, file)
+
+find_package_name <- function(caller_function, caller_package, caller_expression, srcref, file)
 {
   if(caller_function %in% eval_base_functions)
     return("base")
   else
-    return(extract_package_name(srcref, file))
+  {
+    tempPack <- extract_package_name(srcref, file)
+    if(tempPack == "base?") # It means the srcref was NA
+    {
+       if(caller_package == "foreach" && caller_expression == "e$fun(obj, substitute(ex), parent.frame(), e$data)")
+       {
+         # This eval is defined in function doSEQ in do.R of package foreach
+         return("foreach")
+       }
+      else
+      {
+        return("base?")
+      }
+    }
+    else
+      return(tempPack)
+  }
 }
 
-add_eval_source <- function(dataset)
+find_package_name_second_chance <- function(file, caller_stack_expr, caller_stack_expr_srcref)
 {
-  return(dataset %>% 
-           mutate(eval_source = pmap_chr(list(caller_function, eval_call_srcref, file),
-                                                         find_package_name)) )
+  new_srcref <- package_name_from_call_stack(caller_stack_expr, caller_stack_expr_srcref)
+  return(extract_package_name(new_srcref, file))
+}
+
+add_eval_source <- function(dataset, dataset_with_stacks)
+{ 
+  dataset_c <- dataset %>% 
[…]

### 2020-08-26 inst/analysis/corpus-stage2.Rmd
Updates in the pipeline

@@
 source("inc/setup.R")
-```
-
-```{r aux functions}
-read_task_result <- function(path) {
-  read_fst(path) %>% as_tibble() %>% mutate(package=basename(package))
-}
-
-read_parallel_log <- function(path) {
-  read_fst(path) %>% 
-    as_tibble() %>% 
-    rename_all(tolower) %>%
-    mutate(
-      package=map_chr(command, ~basename(str_split(., " ")[[1]][2])),
-      starttime=as_datetime(starttime),
-      endtime=as_datetime(starttime + jobruntime),
-      runtime=endtime-starttime,
-      run_path=map_chr(command, ~dirname(str_split(., " ")[[1]][3]))
-    ) %>%
-    select(package, exitval, starttime, endtime, runtime, command, run_path)
-}
 ```

### 2020-08-26 inst/analysis/inc/paths.R

@@ -16,6 +16,7 @@ R_LIBS_DIR <- path(params$base_dir, "library", "4.0")
 
 ## INPUTs
 
+PACKAGE_EVALS_DYNAMIC_FILE <- path(DATA_DIR, "evals-dynamic.fst")
 PACKAGE_EVALS_FILE         <- path(DATA_DIR, "evals.fst")
 PACKAGE_COVERAGE_FILE      <- path(DATA_DIR, "coverage.fst")
 PACKAGE_METADATA_FILE      <- path(DATA_DIR, "metadata.fst")
@@ -33,6 +34,7 @@ KAGGLE_LOG_FILE     <- path(DATA_DIR, "parallel-kaggle.fst")
 RUN_TRACE_DIR  <- path(RUN_DIR, "package-evals-traced")
 RUN_KAGGLE_DIR <- path(RUN_DIR, "kaggle-run")
 
+CALLS_FINAL_FILE        <- path(RUN_TRACE_DIR, "calls.fst")
 CALLS_FINAL_FILE        <- path(RUN_TRACE_DIR, "calls-final.fst")
 KAGGLE_CALLS_FINAL_FILE <- path(RUN_KAGGLE_DIR, "calls-final.fst")
 

### 2020-08-26 inst/analysis/inc/setup.R

@@ -10,66 +10,83 @@ show_url <- Vectorize(function(path, name=basename(path), hostname=params$hostna
   str_glue('<a href="http://{hostname}:{port}/{URLencode(path)}">{name}</a>')
 }, vectorize.args = "path")
 
-
-"%||%" <- function(a, b) {
-  if (!is.null(a)) a else b
+read_task_result <- function(path) {
+  read_fst(path) %>% as_tibble() %>% mutate(package=basename(package))
 }
 
-geom_flat_violin <- function(mapping = NULL, data = NULL, stat = "ydensity",
-                             position = "dodge", trim = TRUE, scale = "area",
-                             show.legend = NA, inherit.aes = TRUE, ...) {
-  layer(
-    data = data,
-    mapping = mapping,
-    stat = stat,
-    geom = GeomFlatViolin,
-    position = position,
-    show.legend = show.legend,
-    inherit.aes = inherit.aes,
-    params = list(
-      trim = trim,
-      scale = scale,
-      ...
-    )
-  )
+read_parallel_log <- function(path) {
+  read_fst(path) %>% 
+    as_tibble() %>% 
+    rename_all(tolower) %>%
+    mutate(
+      package=map_chr(command, ~basename(str_split(., " ")[[1]][2])),
+      starttime=as_datetime(starttime),
+      endtime=as_datetime(starttime + jobruntime),
+      runtime=endtime-starttime,
+      run_path=map_chr(command, ~dirname(str_split(., " ")[[1]][3]))
+    ) %>%
+    select(package, exitval, starttime, endtime, runtime, command, run_path)
 }
 
-GeomFlatViolin <-
-  ggproto("GeomFlatViolin", Geom,
-          setup_data = function(data, params) {
-            data$width <- data$width %||%
-              params$width %||% (resolution(data$x, FALSE) * 0.9)
+## "%||%" <- function(a, b) {
+##   if (!is.null(a)) a else b
+## }
+
+## geom_flat_violin <- function(mapping = NULL, data = NULL, stat = "ydensity",
+##                              position = "dodge", trim = TRUE, scale = "area",
+##                              show.legend = NA, inherit.aes = TRUE, ...) {
+##   layer(
+##     data = data,
+##     mapping = mapping,
+##     stat = stat,
+##     geom = GeomFlatViolin,
+##     position = position,
+##     show.legend = show.legend,
+##     inherit.aes = inherit.aes,
+##     params = list(
[…]

### 2020-08-26 inst/paper/main.tex
Updated the intro text of V

@@
 
-The corpus for this study includes R programs extracted from popular R packages
-as well as complete user-written data-analysis scripts. In this section we
-describe the methodology used to collect them.
+In this study, we report on eval usage in a corpus of \CorpusCorePackages R core
+and \CorpusPackages popular CRAN packages. To quantify the eval use, we run
+\CorpusAllProgramsRnd of R programs that are extracted form those packages as
+well as from the user-written data-analyses shared on the Kaggle platform. In
+this section we describe this corpus and methodology used to assemble it.
 
-\subsection{CRAN Packages}\label{sec:corpus}
+\subsection{CRAN Packages}
 
@@
 
-To represent user-written code in the corpus, we truned to
+To represent user-written code in the corpus, we turned to
 Kaggle\footnote{\url{https://www.kaggle.com}}, an online platform for
@@
   }
+  \caption{Eval calls in CRAN and Kaggle}
   \label{A}
-  \caption{Eval calls in CRAN and Kaggle}
 \end{table}

### 2020-08-26 inst/analysis/inc/paths.R

@@ -28,6 +28,7 @@ PACKAGE_RUNS_FILE          <- path(DATA_DIR, "run.fst")
 PACKAGE_TRACE_RUNS_FILE    <- path(DATA_DIR, "run-trace.fst")
 PACKAGE_TRACE_LOG_FILE     <- path(DATA_DIR, "parallel-trace.fst")
 
+KAGGLE_DEPENDENCIES_FILE <- path(DATA_DIR, "kaggle-dependencies.txt")
 KAGGLE_KERNELS_FILE <- path(DATA_DIR, "kaggle-kernels.fst")
 KAGGLE_LOG_FILE     <- path(DATA_DIR, "parallel-kaggle.fst")
 

### 2020-08-26 inst/analysis/inc/paths.R
Add program file variable

@@ -48,6 +48,7 @@ CORPUS_FILE            <- path(DATA_DIR, "corpus.txt")
 CORPUS_DETAILS_FILE    <- path(DATA_DIR, "corpus.fst")
 CORPUS_PLOT            <- path(PLOT_DIR, "corpus.pdf")
 
+PROGRAM_FILE          <- path(RUN_TRACE_DIR, "program.fst")
 EVALS_RAW_FILE        <- path(RUN_TRACE_DIR, "raws.fst")
 EVALS_SUM_CORE_FILE   <- path(RUN_TRACE_DIR, "summarized-core.fst")
 EVALS_SUM_PKGS_FILE   <- path(RUN_TRACE_DIR, "summarized-packages.fst")

### 2020-08-26 inst/paper/main.tex
add case study outline

@@
 
+\begin{itemize}
+\item Escape Hatch: \c{eval(unlock_binding ... )} to prevent R CMD check warnings
+\item Accessing data frame columns like local variables: \c{eval(x + y + z, data.frame(x, y z))}
+\item Poor man's sandboxing: local function
+\item Assembling call strings from components passed by callers: eval(substitute(f(x, y, z)))
+\item Convenience: 
+\item Dynamic code loading: \c{eval(parse(...))}.
+The textual input in parse can come from three places: a string, a file or in interactive R sessions directly from user input.
+Next to \c{parse}, there are also \c{str2expression} and \c{str2lang} which are a specialized versions of parse that do not keep source references and thus are a bit faster.
+\item bypassing R CMD check in the case one accesses dangerous functions such as unlockBinding or private symbols.
+\item evaling recorded calls in model-fitting functions or passing most of the current call to another function.
+Both ``patterns'' use the \c{match.call} to capture current call.
+The latter is just for laziness as it is the same as calling the new function with explicitly selecting the arguments.
+\end{itemize}
+
 \subsection{base}

### 2020-08-26 inst/paper/main.tex
Case study on base and environments

--- 2020-08-25 inst/paper/main.tex
@@
 > env$r <- 9
-> eval(r, env)
+> eval(r, envir = env)
 [1] 5
@@
 argument \c{envir} to indicate in which environment its first argument,
-after being resolved, must be evaluated.
+after being resolved, must be evaluated. By default, it is equal to \c{parent.frame()}, which is the parent environment of the \c{eval} call.
 
-By default, it is equal to \c{parent.frame()}, which is the parent
-environment of the \c{eval} call.
+ This argument can also be \c{NULL}, a list or a dataframe.
+In that case, \c{eval} copies \c{envir} and encloses it the environment defined by \c{enclos}.
+Finally, the argument can also be a number $n$.
+It means that the environment in which the expressions is evaluated will be the results of \c{sys.call(n)} where $n$ refers to the $n$-th call frame after the global environment.
+
+\textbf{Do we speak more about what an environment is, \c{.GlobalEnv}, \c{emptyenv()}, environment navigation here, or in the Background section?}
+
+We classify the environment from two points of view:
+\begin{itemize}
+	\item in which environment \c{eval} writes new bindings 
+	\item in which \emph{existing} environment \c{eval} reads. It is the first non new environment. \c{eval} will also be able to read in all parent environment of that environment. 
+\end{itemize}
+
+% Variability: how many different envir class, and different envir expr, for one call site?
 
@@
 
+There are 30 functions in \emph{base} that use \eval.
+Only five of them use \c{parse}, for legitimate reasons, such \c{source} and \c{sys.source} which load a file and then evaluate it, or \c{invokeRestartInteractively} which prompts the user for arguments to restart execution after an error.
+
+We endeavored  to write these functions without \eval.
+\c{match.arg} appears in 60.1\% of the eval calls.
+It is used to get the arguments, possibly abbreviated, of a function.
+With one argument, as \c{match.arg(arg)}, it uses \c{eval} to infer them from a parent function's formal arguments.
+We can rewrite it using its version with two arguments, \c{match.arg(arg, choices)} where \c{choices} are the possible formal arguments of the function, as shown in Listing~\ref{code:matcharg}.
+
+\begin{lstlisting}[caption={Rewriting \c{match.arg} without \eval}, label=code:matcharg]
+f <- function(type=c("examples", "tests", "vignettes")) {
+actual_type = match.arg(type)
+...
+}
+
+f <- function(type=c("examples", "tests", "vignettes")) {
+actual_type = match.arg(type, choices=c("examples", "tests", "vignettes"))
+...
+}
+\end{lstlisting}
+
 \section{Threats to Validity}

### 2020-08-26 inst/analysis/enrich_dataset.Rmd

@@
 knitr::opts_chunk$set(echo = TRUE,
-                      fig.align = "center",
                       fig.retina = 2,
@@
                       cache.lazy = FALSE)
+
+now <- Sys.time()
 
@@
 ```{r add_source}
-eval_calls <- eval_calls %>% add_eval_source() %>% add_eval_source_type()
+eval_calls <- eval_calls %>% add_eval_source(eval_calls_raw) %>% add_eval_source_type()
 ```
@@
 ```{r ast_size}
-eval_calls <- eval_calls %>% add_ast_size()
+#eval_calls <- eval_calls %>% add_ast_size()
 ```
@@
 
+There are `r nrow(eval_calls)` unique rows.
+
 ```{r separate_datasets}
@@
 ```
+
+The core dataset has `r nrow(eval_calls_core)` unique calls, the packages dataset, `r nrow(eval_calls_packages)` and the kaggle one, `r nrow(eval_calls_kaggle)`.
 
@@
 ```
+
+Notebook execution was `r difftime(Sys.time(), now)`.

### 2020-08-26 inst/analysis/corpus-analysis.Rmd

@@
 ```{r load kaggle}
+kaggle_dependencies <- tibble(package=read_lines(KAGGLE_DEPENDENCIES_FILE))
 kaggle_kernels <- as_tibble(read_fst(KAGGLE_KERNELS_FILE))
@@
 
+```{r}
+setdiff(kaggle_dependencies$package, corpus$package)
+```
+
 ## Combined

### 2020-08-26 inst/analysis/analysis_stable.Rmd

@@
 nb_eval_calls_per_package  %>% datatable()
-```
-
-### Unknown evals
-
-Some `eval` do not have srcref information attached and are not in *base*. It is often because they are located in a function that is created by anotehr function.
-
-```{r unknown_evals}
-undefined_evals <- eval_calls %>% filter(eval_source_type == "<undefined>")
-
-nb_undefined <- undefined_evals %>% count(wt = nb_ev_calls)
-
-nb_different_call_expr <- undefined_evals %>% select(eval_call_expression) %>% n_distinct()
-
-r("undefined evals", nb_undefined$n)
-r("percent undefined eval", ratio(nb_undefined$n, nb_eval_calls))
-r("undefined eval expressions", nb_different_call_expr)
 ```

### 2020-08-26 inst/analysis/enrich_dataset.Rmd

@@
 source("inc/paths.R")
+source("inc/latextags.R")
+source("inc/setup.R")
+theme_set(theme_minimal())
+create_tags(path(TAGS_DIR, "enrichDataset.tex"), prefix="", default=TRUE)
+
 source("insights.R", local = knitr::knit_global())
@@
 
+```{r shrink_to_corpus}
+corpus_files <- read_lines(CORPUS_FILE)
+eval_calls <- eval_calls %>% keep_only_corpus(corpus_files)
+```
+
+```{r fake_srcref}
+eval_calls <- eval_calls %>% add_fake_srcref()
+```
+
 ```{r parse_args}
@@
 eval_calls_core <- eval_calls %>% filter(eval_source_type == "core")
-eval_calls_packages <- eval_calls %>% filter(eval_source_type %in% c("package", "<undefined>") )
+eval_calls_packages <- eval_calls %>% filter(eval_source_type == "package" )
 eval_calls_kaggle <- eval_calls %>% filter(eval_source_type == "kaggle")
@@
 ```{r}
-eval_calls_raw <- eval_calls_raw %>% select(-ends_with("_type")) %>% left_join(eval_calls)
+#eval_calls_raw <- eval_calls_raw %>% select(-ends_with("_type")) %>% left_join(eval_calls)
+
 ```
 
+# Unknown evals
+
+Some `eval` do not have srcref information attached and are not in *base*. It is often because they are located in a function that is created by another function.
+
+```{r unknown_evals}
+
+nb_eval_calls <- count(eval_calls, wt = nb_ev_calls)$n
+
+undefined_evals <- eval_calls %>% filter(eval_source_type == "<undefined>")
+
+nb_undefined <- undefined_evals %>% count(wt = nb_ev_calls)
+
+nb_different_call_expr <- undefined_evals %>% select(eval_call_expression) %>% n_distinct()
+
+r("undefined evals", nb_undefined$n)
+r("percent undefined eval", ratio(nb_undefined$n, nb_eval_calls))
+r("undefined eval expressions", nb_different_call_expr)
+```
+
+How many different undefined eval call sites do we have per package run?
+
+```{r undefined_per_package}
+undefined_per_package <- undefined_evals %>% group_by(package) %>% summarize(n = n_distinct(eval_call_expression)) 
+known_packages <- setdiff(eval_calls$package, undefined_per_package$package) %>% as_tibble_col(column_name = "package")
+known_packages <- known_packages %>% mutate(n = 0)
+
+undefined_per_package <- bind_rows(undefined_per_package, known_packages) %>% arrange(desc(n))
+
+```
+
+# Write
+
 ```{r write_back}
+write_fst(undefined_per_package, PACKAGE_EVALS_DYNAMIC_FILE)
+
 write_fst(eval_calls, EVALS_RAW_FILE)
@@
 
-Notebook execution was `r difftime(Sys.time(), now)`.
+```{r}
+duration <- difftime(Sys.time(), now)
+```
+
+Notebook execution was `r duration` `r units(duration)`.

### 2020-08-26 inst/analysis/analysis_stable.Rmd
More data on ast size

@@
 
+What are the most used caller functions?
+
+```{r caller_functions_used}
+caller_function_ranked <- eval_calls %>% count(caller_function, wt = nb_ev_calls, sort = TRUE) %>% mutate(proportion = n / sum(n))
+
+caller_function_ranked %>% datatable()
+
+```
+
+The most used function is `r caller_function_ranked[1, "caller_function"][[1]]`, as much as `r 100 * caller_function_ranked[1, "proportion"][[1]]`%.
+
+```{r match_arg}
+match_arg_r <- caller_function_ranked %>% filter(caller_function == "match.arg")
+
+r("match arg number", match_arg_r[1, "n"][[1]])
+r("match arg percent", percent(match_arg_r[1, "proportion"][[1]]))
+```
+
 ### Excluding *base* from the data set
@@
 ggsave(path(PLOT_DIR, "ast_sizes.pdf"))
+
+ast_sizes_prop <- ast_sizes %>% count(expr_resolved_ast_size, wt = nb_eval_calls) %>%
+                      arrange(desc(expr_resolved_ast_size)) %>% 
+                      mutate(proportion = n / nb_eval_calls) %>% 
+                      mutate(cumulative = cumsum(proportion))
+
+ast_size_one <- ast_sizes %>% filter(expr_resolved_ast_size == 1) 
+nb_ast_size_one <- count(ast_size_one, wt = nb_ev_calls)$n
+nb_ast_size_one_call_sites <- nb_eval_call_sites(ast_size_one)
+r("nb ast one percent", ratio(nb_ast_size_one, nb_eval_calls))
+r("nb ast one call site percent", ratio(nb_ast_size_one_call_sites, nb_call_sites))
+
+nb_nodes_many <- ast_sizes_prop %>% filter(cumulative >= 95) %>% .[1,]
+
+r("ast size 95", nb_nodes_many$expr_resolved_ast_size)
+r("ast size 95 exact percent", percent(nb_nodes_many$proportion))
+
 ```

### 2020-08-26 inst/paper/main.tex
P
Add basic infrastructure details

@@
 This section describes the infrastructure we used to run the dynamic program
-analysis to record the calls to evals.
+analysis to record the calls to evals. Our analysis pipeline starts with
+downloading and installing the open source R packages from CRAN. The Kaggle
+kernels are downloaded from Kaggle.com by a crawler and handled separately.
+Next, we extract code size and coverage metrics from the installed packages
+using the covr package.
+This is followed by extraction of runnable programs from R packages: tests,
+examples and vignettes. The extracted programs are then executed through our
+dynamic analyzer, instrumentr, which obtains relevant information from the
+program runs. instrumentr is an R package that enables us to write our tracing
+logic in R. It internally uses R-dyntrace which is a modified R interpreter that
+exposes hooks for events of interest. instrumentr provides an API to execute R
+callbacks from these hooks.
+Finally, we use RMarkdown notebooks to analyze and visualize this data.
+Our infrastructure contains three
+components:
+Figure~\ref{fig:pipeline} shows the main steps of our pipeline.
 
@@
 We endeavored  to write these functions without \eval.
-\c{match.arg} appears in 60.1\% of the eval calls.
+\c{match.arg} appears in \MatchArgPercent of the \eval calls.
 It is used to get the arguments, possibly abbreviated, of a function.

### 2020-08-26 inst/paper/main.tex
Fix infrastructure

--- 2020-08-26 inst/paper/main.tex
@@
 \usepackage{listings}
+\usepackage{hyperref}
 \newcommand{\missingNumber}{\textcolor{red}{XX}\xspace}
@@
 \newcommand{\miss}[1]{{\textcolor{red}{#1}}\xspace}
+\newcommand{\evil}{\emph{evil}\xspace}
+\newcommand{\instrumentr}{\emph{instrumentr}\xspace}
+\newcommand{\rdyntrace}{\emph{R-dyntrace}\xspace}
+\newcommand{\covr}{\emph{covr}\xspace}
+\newcommand{\runr}{\emph{runr}\xspace}
 
@@
 
-This section describes the infrastructure we used to run the dynamic program
-analysis to record the calls to evals.
+Our infrastructure consists of a dynamic analysis pipeline to record the calls
+to evals and RMarkdown notebooks for data analysis and visualization.
+
+Figure~\ref{fig:pipeline} shows the main steps of the pipeline. We provide the
+runtime, input and output data size for each step.
 
@@
 \end{figure*}
+
+The pipeline begins with downloading and installing open source R packages from
+CRAN for the corpus along with their dependencies. The Kaggle programs are
+downloaded from \href{http://www.kaggle.com}{Kaggle} separately by a crawler.
+Next, code size and coverage metrics from the installed packages are extracted
+using the \href{ https://github.com/r-lib/covr}{\covr} package. This is followed
+by extraction of runnable programs from R packages: tests, examples and
+vignettes. The extracted programs are then run through \emph{evil} (\emph{Ev}al
+\emph{I}nspection \emph{L}ibrary), a dynamic analyzer that collects information
+about eval calls from program runs. This dynamic analyzer builds upon our
+dynamic analysis framework, \instrumentr which is another R package that enables
+us to write our dynamic analysis logic in R. It internally uses a modified R
+interpreter, \rdyntrace~\cite{oopsla19a}, that exposes hooks from within the
+interpreter implementation for events of interest. \instrumentr serves as an
+intermediary between \rdyntrace and \evil, it intercepts the hooks exposed by
+\rdyntrace and attaches R functions exported by \evil as callbacks. The \evil
+callbacks execute on corresponding interpreter events. All steps of this
+pipeline are parallelized. To schedule and parallelize extraction and analysis
+of programs, we use our homegrown library \runr. Furthermore, \runr gracefully
+handles and reports failures across large-scale program runs which greatly aids
+debugging of the analysis pipeline. The data extracted by \evil from each
+program is concatenated, cleaned and summarized in the post-processing phase by
+custom R scripts. Finally, the summarized data is analyzed in RMarkdown
+notebooks to gather insights. Apart from the figures, the data points included
+in the paper are also generated by RMarkdown notebooks as latex macros.
 

### 2020-08-26 inst/paper/main.tex
Add threats to validity

@@
 \section{Threats to Validity}
+The primary threat to validity is the issue of code coverage, a cause of concern
+for any dynamic approach. We mitigate this by including only those R packages in
+our corpus that have many reverse dependencies and thus high code coverage.
+Secondly, we turn off the bytecode compiler for this study. The bytecode
+compiler can also call eval. Thirdly, we do not get source locations for
+\missingNumber eval calls. These are the calls in which eval is passed as an
+argument to a higher-order functions and the R parser does not retain location
+information for them. However, this is a meager \missingPercentage of all eval
+calls and is unlikely to affect our analysis.
 

### 2020-08-26 inst/paper/main.tex
Fix references to eval

@@
 Our infrastructure consists of a dynamic analysis pipeline to record the calls
-to evals and RMarkdown notebooks for data analysis and visualization.
+to \eval and RMarkdown notebooks for data analysis and visualization.
 
@@
 \emph{I}nspection \emph{L}ibrary), a dynamic analyzer that collects information
-about eval calls from program runs. This dynamic analyzer builds upon our
+about \eval calls from program runs. This dynamic analyzer builds upon our
 dynamic analysis framework, \instrumentr which is another R package that enables
@@
 
-In this study, we report on eval usage in a corpus of \CorpusCorePackages R core
+In this study, we report on \eval usage in a corpus of \CorpusCorePackages R core
 and \CorpusPackages popular CRAN packages. To quantify the eval use, we run

### 2020-08-26 inst/paper/main.tex

@@
 \item Escape Hatch: \c{eval(unlock_binding ... )} to prevent R CMD check warnings
-\item Accessing data frame columns like local variables: \c{eval(x + y + z, data.frame(x, y z))}
+  \subsubsection{Evaluation using Data Fields}
+  A common use of \eval is to evaluate expressions that access data frame
+columns or list (heterogeneous vectors) elements by name. Core R provides two
+functions, \c{with} and \c{within} that implement this pattern. These functions
+construct a temporary environment with data frame columns or list elements as
+the bindings. Side-effects performed by the evaluated expression take place in
+this temporary environment and are ignored by the \c{with} function. The
+\c{within} function copies these modifications to a local copy of the input data
+and returns it. This pattern is used quite often for statistical modeling. The
+modeling expressions is constructed from variable names whose values are
+supplied by a data frame or list.
+
 \item Poor man's sandboxing: local function

### 2020-08-26 inst/paper/main.tex
Include tag
Add sandboxing case study

@@
 \input{tag/analysisStable.tex}
+\input{tag/enrichDataset.tex}
 
@@
 \begin{itemize}
-\item Escape Hatch: \c{eval(unlock_binding ... )} to prevent R CMD check warnings
-  \subsubsection{Evaluation using Data Fields}
-  A common use of \eval is to evaluate expressions that access data frame
-columns or list (heterogeneous vectors) elements by name. Core R provides two
-functions, \c{with} and \c{within} that implement this pattern. These functions
-construct a temporary environment with data frame columns or list elements as
-the bindings. Side-effects performed by the evaluated expression take place in
-this temporary environment and are ignored by the \c{with} function. The
-\c{within} function copies these modifications to a local copy of the input data
-and returns it. This pattern is used quite often for statistical modeling. The
-modeling expressions is constructed from variable names whose values are
-supplied by a data frame or list.
+\item Escape Hatch: \c{eval(unlock_binding ... )} to prevent R CMD check
+  warnings
+  \subsubsection{Evaluation using Data Fields} A common use of \eval is to
+  evaluate expressions that access data frame columns or list (heterogeneous
+  vectors) elements by name. Core R provides two functions, \c{with} and
+  \c{within} that implement this pattern. These functions construct a temporary
+  environment with data frame columns or list elements as the bindings.
+  Side-effects performed by the evaluated expression take place in this temporary
+  environment and are ignored by the \c{with} function. The \c{within} function
+  copies these modifications to a local copy of the input data and returns it.
+  This pattern is used quite often for statistical modeling. The modeling
+  expressions is constructed from variable names whose values are supplied by a
+  data frame or list.
 
-\item Poor man's sandboxing: local function
+  \subsubsection{Sandboxing} The combination of \eval and first class
+  environments can be used to construct a sandbox for evaluating side-effecting
+  expressions. The newly constructed environment can use the current environment
+  for its lexical scope. This provides the evaluated expression access to all the
+  bindings in the current scope but restricts the side-effect to the newly
+  constructed throwaway environment. However, this sandboxing is limited because
+  the expression can reflectively access any environment in the call stack or use
+  the super assignment operator(\c{<<-}) to write to parent scope. The \c{local}
+  function which is part of Core R is an example of this usage pattern.
+
 \item Assembling call strings from components passed by callers: eval(substitute(f(x, y, z)))

### 2020-08-26 inst/paper/main.tex
Add dynamic code loading explanation

@@
 \newcommand{\eval}{\texttt{eval}\xspace}
+\newcommand{\parse}{\texttt{parse}\xspace}
+\newcommand{\source}{\texttt{source}\xspace}
 \renewcommand{\c}[1]{\lstinline{#1}\xspace}
@@
 
+  \subsubsection{Dynamic Code Loading} The combination of \parse and \eval is
+  used to parse text from string or file and evaluate in a custom environment.
+  This pattern is used by the \source Core R function to load R code from a file
+  in the current workspace. This is quite useful for loading functions for
+  interactive data analysis.
+
 \item Assembling call strings from components passed by callers: eval(substitute(f(x, y, z)))

### 2020-08-26 inst/paper/main.tex
Move case study to usage patterns

@@
 
-Here we categorize evals in some common groups
+  \subsubsection{Evaluation using Data Fields} A common use of \eval is to
+  evaluate expressions that access data frame columns or list (heterogeneous
+  vectors) elements by name. Core R provides two functions, \c{with} and
+  \c{within} that implement this pattern. These functions construct a temporary
+  environment with data frame columns or list elements as the bindings.
+  Side-effects performed by the evaluated expression take place in this temporary
+  environment and are ignored by the \c{with} function. The \c{within} function
+  copies these modifications to a local copy of the input data and returns it.
+  This pattern is used quite often for statistical modeling. The modeling
+  expressions is constructed from variable names whose values are supplied by a
+  data frame or list.
+
+  \subsubsection{Sandboxing} The combination of \eval and first class
+  environments can be used to construct a sandbox for evaluating side-effecting
+  expressions. The newly constructed environment can use the current environment
+  for its lexical scope. This provides the evaluated expression access to all the
+  bindings in the current scope but restricts the side-effect to the newly
+  constructed throwaway environment. However, this sandboxing is limited because
+  the expression can reflectively access any environment in the call stack or use
+  the super assignment operator(\c{<<-}) to write to parent scope. The \c{local}
+  function which is part of Core R is an example of this usage pattern.
+
+  \subsubsection{Dynamic Code Loading} The combination of \parse and \eval is
+  used to parse text from string or file and evaluate in a custom environment.
+  This pattern is used by the \source Core R function to load R code from a file
+  in the current workspace. This is quite useful for loading functions for
+  interactive data analysis.
+
+  \subsubsection{Metaprogramming} \eval in combination with \substitute is used
+  to
+  Assembling call strings from components passed by callers: eval(substitute(f(x, y, z)))
+
+  \subsubsection{Escape Hatch}: \c{eval(unlock_binding ... )} to prevent R CMD check
+  warnings
+
+  bypassing R CMD check in the case one accesses dangerous functions such as unlockBinding or private symbols.
+ evaling recorded calls in model-fitting functions or passing most of the current call to another function.
+Both ``patterns'' use the \c{match.call} to capture current call.
+The latter is just for laziness as it is the same as calling the new function with explicitly selecting the arguments.
 
@@
 \end{figure}
-
-\section{Case studies}
-
-\begin{itemize}
-\item Escape Hatch: \c{eval(unlock_binding ... )} to prevent R CMD check
-  warnings
-  \subsubsection{Evaluation using Data Fields} A common use of \eval is to
-  evaluate expressions that access data frame columns or list (heterogeneous
-  vectors) elements by name. Core R provides two functions, \c{with} and
-  \c{within} that implement this pattern. These functions construct a temporary
-  environment with data frame columns or list elements as the bindings.
-  Side-effects performed by the evaluated expression take place in this temporary
-  environment and are ignored by the \c{with} function. The \c{within} function
-  copies these modifications to a local copy of the input data and returns it.
-  This pattern is used quite often for statistical modeling. The modeling
-  expressions is constructed from variable names whose values are supplied by a
-  data frame or list.
-
-  \subsubsection{Sandboxing} The combination of \eval and first class
-  environments can be used to construct a sandbox for evaluating side-effecting
-  expressions. The newly constructed environment can use the current environment
-  for its lexical scope. This provides the evaluated expression access to all the
-  bindings in the current scope but restricts the side-effect to the newly
-  constructed throwaway environment. However, this sandboxing is limited because
-  the expression can reflectively access any environment in the call stack or use
-  the super assignment operator(\c{<<-}) to write to parent scope. The \c{local}
-  function which is part of Core R is an example of this usage pattern.
-
-  \subsubsection{Dynamic Code Loading} The combination of \parse and \eval is
-  used to parse text from string or file and evaluate in a custom environment.
-  This pattern is used by the \source Core R function to load R code from a file
-  in the current workspace. This is quite useful for loading functions for
-  interactive data analysis.
-
-\item Assembling call strings from components passed by callers: eval(substitute(f(x, y, z)))
-\item Convenience: 
-\item Dynamic code loading: \c{eval(parse(...))}.
-The textual input in parse can come from three places: a string, a file or in interactive R sessions directly from user input.
-Next to \c{parse}, there are also \c{str2expression} and \c{str2lang} which are a specialized versions of parse that do not keep source references and thus are a bit faster.
-\item bypassing R CMD check in the case one accesses dangerous functions such as unlockBinding or private symbols.
-\item evaling recorded calls in model-fitting functions or passing most of the current call to another function.
-Both ``patterns'' use the \c{match.call} to capture current call.
-The latter is just for laziness as it is the same as calling the new function with explicitly selecting the arguments.
-\end{itemize}
 

### 2020-08-26 inst/analysis/analysis_stable.Rmd
Fixing tags

@@
 ast_sizes_prop <- ast_sizes %>% count(expr_resolved_ast_size, wt = nb_eval_calls) %>%
-                      arrange(desc(expr_resolved_ast_size)) %>% 
-                      mutate(proportion = n / nb_eval_calls) %>% 
+                      mutate(proportion = n / sum(n)) %>% 
+                      arrange(expr_resolved_ast_size) %>% 
                       mutate(cumulative = cumsum(proportion))
@@
 nb_ast_size_one_call_sites <- nb_eval_call_sites(ast_size_one)
-r("nb ast one percent", ratio(nb_ast_size_one, nb_eval_calls))
-r("nb ast one call site percent", ratio(nb_ast_size_one_call_sites, nb_call_sites))
+r("nb ast one percent", ratio(nb_ast_size_one, sum(ast_sizes_prop$n)))
+r("nb ast one call site percent", ratio(nb_ast_size_one_call_sites, nb_eval_call_sites(ast_sizes)))
 
@@
 
-r("ast size 95", nb_nodes_many$expr_resolved_ast_size)
-r("ast size 95 exact percent", percent(nb_nodes_many$proportion))
+r("ast size nine five", nb_nodes_many$expr_resolved_ast_size)
+r("ast size nine five exact percent", percent(nb_nodes_many$proportion))
 

### 2020-08-26 inst/analysis/enrich_dataset.Rmd
Fixing tags

@@
 
-```{r shrink_to_corpus}
-corpus_files <- read_lines(CORPUS_FILE)
-eval_calls <- eval_calls %>% keep_only_corpus(corpus_files)
-```
-
 ```{r fake_srcref}
@@
 
-There are `r nrow(eval_calls)` unique rows.
+```{r shrink_to_corpus}
+corpus_files <- read_lines(CORPUS_FILE)
+eval_calls_corpus <- eval_calls %>% keep_only_corpus(corpus_files)
+
+eval_calls_externals <- eval_calls %>% get_externals(corpus_files)
+```
+
+There are `r nrow(eval_calls)` unique rows. The file with dependent packages has `r nrow(eval_calls_externals)` rows and the one with only the packages in the corpus, `r nrow(eval_calls_corpus)`.
 
 ```{r separate_datasets}
-eval_calls_core <- eval_calls %>% filter(eval_source_type == "core")
-eval_calls_packages <- eval_calls %>% filter(eval_source_type == "package" )
-eval_calls_kaggle <- eval_calls %>% filter(eval_source_type == "kaggle")
+eval_calls_core <- eval_calls_corpus %>% filter(eval_source_type == "core")
+eval_calls_packages <- eval_calls_corpus %>% filter(eval_source_type == "package" )
+eval_calls_kaggle <- eval_calls_corpus %>% filter(eval_source_type == "kaggle")
 ```
@@
 
+nb_out_corpus <- undefined_evals %>% semi_join(eval_calls_externals) %>%
+    count(wt = nb_ev_calls)
+
 nb_different_call_expr <- undefined_evals %>% select(eval_call_expression) %>% n_distinct()
@@
 r("undefined eval expressions", nb_different_call_expr)
+r("undefined out of corpus", nb_out_corpus)
+r("undefined out of coprus percent", ratio(nb_out_corpus, nb_undefined))
 ```
@@
 write_fst(eval_calls_kaggle, EVALS_SUM_KAGGLE_FILE)
+
+write_fst(eval_calls_externals, EVALS_SUM_EXTERNALS_FILE)
 ```

### 2020-08-26 inst/analysis/inc/paths.R

@@ -27,10 +27,12 @@ PACKAGE_RUNNABLE_CODE_FILE <- path(DATA_DIR, "runnable-code-metadata.fst")
 PACKAGE_RUNS_FILE          <- path(DATA_DIR, "run.fst")
 PACKAGE_TRACE_RUNS_FILE    <- path(DATA_DIR, "run-trace.fst")
 PACKAGE_TRACE_LOG_FILE     <- path(DATA_DIR, "parallel-trace.fst")
+CRAN_PROGRAMS_FILE         <- path(DATA_DIR, "cran-programs.fst")
 
 KAGGLE_DEPENDENCIES_FILE <- path(DATA_DIR, "kaggle-dependencies.txt")
-KAGGLE_KERNELS_FILE <- path(DATA_DIR, "kaggle-kernels.fst")
-KAGGLE_LOG_FILE     <- path(DATA_DIR, "parallel-kaggle.fst")
+KAGGLE_KERNELS_FILE      <- path(DATA_DIR, "kaggle-kernels.fst")
+KAGGLE_LOG_FILE          <- path(DATA_DIR, "parallel-kaggle.fst")
+KAGGLE_PROGRAMS_FILE     <- path(DATA_DIR, "kaggle-programs.fst")
 
 RUN_TRACE_DIR  <- path(RUN_DIR, "package-evals-traced")
 RUN_KAGGLE_DIR <- path(RUN_DIR, "kaggle-run")
@@ -53,6 +55,7 @@ EVALS_RAW_FILE        <- path(RUN_TRACE_DIR, "raws.fst")
 EVALS_SUM_CORE_FILE   <- path(RUN_TRACE_DIR, "summarized-core.fst")
 EVALS_SUM_PKGS_FILE   <- path(RUN_TRACE_DIR, "summarized-packages.fst")
 EVALS_SUM_KAGGLE_FILE <- path(RUN_TRACE_DIR, "summarized-kaggle.fst")
+EVALS_SUM_EXTERNALS_FILE <- path(RUN_TRACE_DIR, "summarized-externals.fst")
 
 ################################################################################
 # GLOBALS

### 2020-08-26 inst/analysis/preprocess.R

@@ -147,3 +147,9 @@ keep_only_corpus <- function(dataset, corpus_files)
            filter(eval_source %in% c(corpus_files, "core", "base", "base?")))
 }
 
+get_externals <- function(dataset, corpus_files)
+{
+  return(dataset %>% 
+           filter(!eval_source %in% c(corpus_files, "core", "base", "base?")))
+}
+

### 2020-08-26 inst/paper/main.tex
Add substitute macro
Percentage of unknown

@@
 \newcommand{\source}{\texttt{source}\xspace}
+\newcommand{\substitute}{\texttt{substitute}\xspace}
 \renewcommand{\c}[1]{\lstinline{#1}\xspace}
@@
 compiler can also call eval. Thirdly, we do not get source locations for
-\missingNumber eval calls. These are the calls in which eval is passed as an
-argument to a higher-order functions and the R parser does not retain location
-information for them. However, this is a meager \missingPercentage of all eval
+\UndefinedEvalsRnd eval calls. These are the calls in which eval is passed as an
+argument to a higher-order functions or is defined in a function returned by a higher-order function and the R parser does not retain location
+information for them. However, this is a meager \PercentUndefinedEval of all eval
 calls and is unlikely to affect our analysis.

### 2020-08-26 inst/paper/main.tex
Add section on metaprogramming

@@
 
-  \subsubsection{Metaprogramming} \eval in combination with \substitute is used
-  to
-  Assembling call strings from components passed by callers: eval(substitute(f(x, y, z)))
+  \subsubsection{Metaprogramming} The combination of \eval and \substitute is
+  used to do metaprogramming in R. \substitute can be used to synthesize ASTs by
+  stitching together user inputs. \substitute walks over its argument AST and
+  replaces the symbols with their bindings in the specified environment and
+  returns this new AST which can then be evaluated in custom environments using
+  \eval.
 

### 2020-08-26 inst/paper/main.tex
Add description of miscellaneous eval usage

@@
 \newcommand{\source}{\texttt{source}\xspace}
+\newcommand{\unlockBinding}{\texttt{unlockBinding}\xspace}
 \newcommand{\substitute}{\texttt{substitute}\xspace}
+\newcommand{\datatable}{\texttt{data.table}\xspace}
+\newcommand{\base}{\texttt{base}\xspace}
 \renewcommand{\c}[1]{\lstinline{#1}\xspace}
@@
 
-  \subsubsection{Escape Hatch}: \c{eval(unlock_binding ... )} to prevent R CMD check
-  warnings
+  \subsubsection{Convenience}
+  evaling recorded calls in model-fitting functions or passing most of the
+  current call to another function. Both ``patterns'' use the \c{match.call} to
+  capture current call. The latter is just for laziness as it is the same as
+  calling the new function with explicitly selecting the arguments.
 
-  bypassing R CMD check in the case one accesses dangerous functions such as unlockBinding or private symbols.
- evaling recorded calls in model-fitting functions or passing most of the current call to another function.
-Both ``patterns'' use the \c{match.call} to capture current call.
-The latter is just for laziness as it is the same as calling the new function with explicitly selecting the arguments.
+  \subsubsection{Miscellaneous} \eval is also used in R packages to get around
+  various limitations. For example, in \datatable package, \eval is used to modify
+  a binding in the \base package of R. Bindings in the \base package are locked
+  for modification and require explicit unlocking using \unlockBinding function.
+  However using this function directly triggers a warning during package building
+  process which statically analyzes the package for use of such restricted
+  functions. The \datatable gets around this limitation by enclosing the call to
+  \unlockBinding inside \eval.
 

### 2020-08-26 inst/paper/main.tex
Minor fix

@@
   for modification and require explicit unlocking using \unlockBinding function.
-  However using this function directly triggers a warning during package building
+  However, directly using this function triggers a warning during package building
   process which statically analyzes the package for use of such restricted

### 2020-08-26 inst/paper/main.tex
Merge

--- 2020-08-26 inst/paper/main.tex
@@
 \newcommand{\eval}{\texttt{eval}\xspace}
+\newcommand{\substitute}{\texttt{substitute}\xspace}
 \newcommand{\parse}{\texttt{parse}\xspace}
 \newcommand{\source}{\texttt{source}\xspace}
-\newcommand{\substitute}{\texttt{substitute}\xspace}
 \renewcommand{\c}[1]{\lstinline{#1}\xspace}
@@
 
-  \subsubsection{Metaprogramming} The combination of \eval and \substitute is
-  used to do metaprogramming in R. \substitute can be used to synthesize ASTs by
-  stitching together user inputs. \substitute walks over its argument AST and
-  replaces the symbols with their bindings in the specified environment and
-  returns this new AST which can then be evaluated in custom environments using
-  \eval.
+  \subsubsection{Metaprogramming} \eval in combination with \substitute is used
+  to
+  Assembling call strings from components passed by callers: eval(substitute(f(x, y, z)))
 

### 2020-08-26 inst/paper/main.tex
Add introduction to classification section

--- 2020-08-26 inst/paper/main.tex
@@
 
-This section presents an overiew of the usage of \eval in our corpus.
+This section presents an overview of the usage of \eval in our corpus.
 Previous work reported that 82\% of JavaScript-enabled web pages used
@@
 
+This section presents the common patterns of \eval usage in our corpus. We
+randomly selected 20 packages from the corpus and studied how they use eval.
+We have grouped those uses in the following categories.
+
   \subsubsection{Evaluation using Data Fields} A common use of \eval is to

### 2020-08-26 inst/paper/main.tex
Fix threats

@@
 
-\section{Threats to Validity}
-The primary threat to validity is the issue of code coverage, a cause of concern
-for any dynamic approach. We mitigate this by including only those R packages in
-our corpus that have many reverse dependencies and thus high code coverage.
-Secondly, we turn off the bytecode compiler for this study. The bytecode
-compiler can also call eval. Thirdly, we do not get source locations for
-\UndefinedEvalsRnd eval calls. These are the calls in which eval is passed as an
-argument to a higher-order functions or is defined in a function returned by a higher-order function and the R parser does not retain location
-information for them. However, this is a meager \PercentUndefinedEval of all eval
-calls and is unlikely to affect our analysis.
+\section{Threats to Validity} The primary threat to validity is the issue of
+code coverage, a cause of concern for any dynamic approach. We mitigate this by
+including only those R packages in our corpus that have many reverse
+dependencies and thus high code coverage. Secondly, we turn off the bytecode
+compiler for this study. The bytecode compiler can also call eval. Thirdly, we
+do not get source locations for \UndefinedEvalsRnd eval calls. In these cases
+\eval is either passed as an argument to a higher-order functions or is defined
+in a function returned by a higher-order function and the R parser does not
+retain location information for \eval. However, this is a meager
+\PercentUndefinedEval of all eval calls and is unlikely to affect our analysis.
 

### 2020-08-27 inst/analysis/side-effects.Rmd
Add side-effect data

---
title: "Side effects"
output: html_document
params:
  base_dir: ../../../
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
for (x in c(
  "dplyr", 
  "DT",
  "fs",
  "fst",
  "ggplot2",
  "ggthemes",
  "lubridate",
  "readr",
  "purrr",
  "stringr",
  "tidyr"
  )) {
  suppressPackageStartupMessages(library(x, character.only=TRUE))
}

knitr::opts_chunk$set(
  echo = TRUE,
  fig.align = "center",
  fig.retina = 2,
  fig.width = 10
)

# TODO: move into the package
source("inc/paths.R")
source("inc/setup.R")
source("inc/latextags.R")

options(repos="https://cran.r-project.org")
theme_set(theme_minimal())
create_tags(path(TAGS_DIR, "side-effects.tex"), prefix="", default=TRUE)
```

```{r}
core_data <- read_fst(EVALS_SUM_CORE_FILE)
package_data <- read_fst(EVALS_SUM_PKGS_FILE)
kaggle_data <- read_fst(EVALS_SUM_KAGGLE_FILE)
program_data <- read_fst(PROGRAM_FILE)

all_data <- rbind(core_data, package_data, kaggle_data)
```

## Side-Effects and Eval

Each entry in `call.fst` contains `direct_writes` and `indirect_writes` (both
`0` for `eval.parent` and `local`). `direct_writes` are the side-effects
directly caused by the `eval` in question and `indirect_writes` are the
side-effects caused by all children evals of the eval. All application
assignments can be found in `direct_writes` field of `program.fst` and all
`eval_assignments` can be found in `indirect_writes` field of `program.fst`. All
non-local assignments can be found by simply summing up the `direct_writes`
field of all eval calls.

```{r preprocess-write-data}
eval_writes_data <-
    all_data %>%
    filter(eval_function == "eval" | eval_function == "evalq") %>%
    select(eval_source_type, eval_source, eval_call_srcref, direct_writes, indirect_writes, nb_ev_calls) %>%
    mutate(all_writes = direct_writes + indirect_writes, side_effecting = as.logical(all_writes))

program_writes_data <-
    program_data %>%
    select(direct_writes, indirect_writes)
```

```{r compute-side-effecting-write-counts}
all_writes <- sum(program_writes_data$direct_writes)
eval_writes <- sum(program_writes_data$indirect_writes)

r("All Writes", all_writes)
r("Eval Writes", eval_writes)

eval_side_effecting_writes <- sum(eval_writes_data$direct_writes * eval_writes_data$nb_ev_calls)

r("Eval Side Effecting Writes", eval_side_effecting_writes)

r("Eval Side Effecting Writes Eval Perc", ratio(eval_side_effecting_writes, eval_writes))
r("Eval Side Effecting Writes All Perc", ratio(eval_side_effecting_writes, all_writes))
```

```{r compute-side-effecting-call-counts}
eval_calls_by_source_writes <-
    eval_writes_data %>%
    count(eval_source_type, side_effecting, wt = nb_ev_calls, name = "call_count") %>%
    group_by(eval_source_type) %>%
    mutate(total_call_count = sum(call_count)) %>%
    ungroup() %>%
    mutate(call_perc = round(100 * call_count/total_call_count, 1)) %>%
    filter(side_effecting)

datatable(eval_calls_by_source_writes)

eval_calls_by_source_writes %>%
pmap(function(eval_source_type, side_effecting, call_count, total_call_count, call_perc) {
    r(str_c("Side Effecting ", eval_source_type, " call perc"), str_c(call_perc, "%"))
})

eval_calls_by_package_writes <-
    eval_writes_data %>%
    filter(eval_source_type != "core") %>%
    count(eval_source, side_effecting, wt = nb_ev_calls, name = "call_count") %>%
    arrange(desc(call_count))

datatable(eval_calls_by_package_writes)
```

The call site computation is incorrect.

```{r compute-side-effecting-site-counts, eval = FALSE}
eval_sites_by_source_writes <-
    eval_writes_data %>%
    group_by(eval_source_type, eval_call_srcref) %>%
    summarize(side_effecting = any(side_effecting)) %>%
    ungroup() %>%
    count(eval_source_type, side_effecting, name = "site_count") %>%
    group_by(eval_source_type) %>%
    mutate(total_site_count = sum(site_count)) %>%
    ungroup() %>%
    mutate(site_perc = round(100 * site_count / total_site_count, 1)) %>%
    filter(side_effecting)

datatable(eval_sites_by_source_writes)

eval_sites_by_source_writes %>%
pmap(function(eval_source_type, side_effecting, site_count, total_site_count, site_perc) {
    r(str_c("SideEffecting ", eval_source_type, " site perc"), str_c(site_perc, "%"))
})

eval_sites_by_package_writes <-
    eval_writes_data %>%
    filter(eval_source_type != "core") %>%
    group_by(eval_source, eval_call_srcref) %>%
    summarize(side_effecting = any(side_effecting)) %>%
    ungroup() %>%
    count(eval_source, side_effecting, name = "site_count") %>%
    group_by(eval_source, side_effecting) %>%
    mutate(site_perc = round(100 * site_count / sum(site_count), 1)) %>%
    ungroup() %>%
    arrange(side_effecting, desc(site_perc))

datatable(eval_sites_by_package_writes)
```

### 2020-08-27 inst/paper/main.tex
Add side-effect data

@@
 \input{tag/corpus.tex}
+\input{tag/side-effects.tex}
 \input{tag/analysisStable.tex}
@@
 
+\section{Side-Effects}
+
+In the corpus we observe \AllWritesRnd writes to variables of which
+\EvalWritesRnd writes happen inside \eval. However, all writes are not
+dangerous. Only writes to environments not local to the computation spawned
+by \eval are side-effecting. These writes outlive the computation and hence are
+visible outside the computation. The remaining writes are local to the
+computation. We observe that \EvalSideEffectingWritesRnd writes inside evals are
+side-effecting. This is only \EvalSideEffectingWritesEvalPerc of all variable
+writes inside \eval and \EvalSideEffectingWritesAllPerc of all variable writes
+in the corpus.
+
+An \eval is considered side-effecting if it performs a side-effecting write to a
+variable, directly or indirectly. Only \SideEffectingCoreCallPerc \eval calls in
+Core R are side-effecting and \SideEffectingPackageCallPerc \eval calls in CRAN
+packages are side-effecting.
+
 \section{A taxonomy of eval}

### 2020-08-27 inst/paper/main.tex
Updated 3

@@
 by extraction of runnable programs from R packages: tests, examples and
-vignettes. The extracted programs are then run through \emph{evil} (\emph{Ev}al
-\emph{I}nspection \emph{L}ibrary), a dynamic analyzer that collects information
-about \eval calls from program runs. This dynamic analyzer builds upon our
-dynamic analysis framework, \instrumentr which is another R package that enables
-us to write our dynamic analysis logic in R. It internally uses a modified R
-interpreter, \rdyntrace~\cite{oopsla19a}, that exposes hooks from within the
-interpreter implementation for events of interest. \instrumentr serves as an
-intermediary between \rdyntrace and \evil, it intercepts the hooks exposed by
-\rdyntrace and attaches R functions exported by \evil as callbacks. The \evil
-callbacks execute on corresponding interpreter events. All steps of this
-pipeline are parallelized. To schedule and parallelize extraction and analysis
-of programs, we use our homegrown library \runr. Furthermore, \runr gracefully
-handles and reports failures across large-scale program runs which greatly aids
-debugging of the analysis pipeline. The data extracted by \evil from each
-program is concatenated, cleaned and summarized in the post-processing phase by
-custom R scripts. Finally, the summarized data is analyzed in RMarkdown
-notebooks to gather insights. Apart from the figures, the data points included
-in the paper are also generated by RMarkdown notebooks as latex macros.
+vignettes as well as from Kaggle. The code of each extracted program is wrapped
+into a call to dynamic analyzer \emph{evil} (\emph{Ev}al \emph{I}nspection
+\emph{L}ibrary) that collect eval usage. The reason for this is that we only
+want to record eval usage for the extracted code. Without this, the data would
+include eval calls from the unit testing frameworks as well as from
+bootstrapping R virtual machine itself. To avoid any interference, each program
+is run in its own R instance. The \emph{evil} framework implemented as R package
+in 2K lines of R and 400 lines of C++ code.
+
+The dynamic analyzer builds upon our dynamic analysis framework, \instrumentr
+that we have implemented to enable us to write dynamic analysis logic in R. It
+is an R package implemented in 2.5K lines of R and 6K lines of C++ code. It
+internally uses a modified R interpreter, \rdyntrace~\cite{oopsla19a}, that
+exposes hooks from within the interpreter implementation for events of interest.
+\instrumentr serves as an intermediary between \rdyntrace and \evil, it
+intercepts the hooks exposed by \rdyntrace and attaches R functions exported by
+\evil as callbacks. The \evil callbacks execute on corresponding interpreter
+events.
+
+All steps of this pipeline are parallelized using GNU
+parallel~\cite{GNUparallel} and orchestrated by GNU make. To schedule and
+parallelize extraction and analysis of programs, we use our homegrown library
+\runr. Furthermore, \runr gracefully handles and reports failures across
+large-scale program runs which greatly aids debugging of the analysis pipeline.
+The data extracted by \evil from each program is concatenated, cleaned and
+summarized in the post-processing phase by custom R scripts. Finally, the
+summarized data is analyzed in RMarkdown notebooks to gather insights. Apart
+from the figures, the data points included in the paper are also generated by
+RMarkdown notebooks as latex macros.
+
+All runs are done on a single server Intel Xeon 6140, 2.30GHz with 72 cores and
+256GB of RAM.
 

### 2020-08-27 inst/analysis/side-effects.Rmd
Percentages for AST size

@@
 theme_set(theme_minimal())
-create_tags(path(TAGS_DIR, "side-effects.tex"), prefix="", default=TRUE)
+#create_tags(path(TAGS_DIR, "side-effects.tex"), prefix="SE", default=TRUE)
 ```
@@
 ```{r}
+# FIXME: path!
+calls_raw <- read_fst(print(path(params$base_dir, "run", "package-evals-traced.1", "calls.fst")))
+calls <- 
+  calls_raw %>%
+  mutate(package=basename(dirname(dirname(file))))
+
 core_data <- read_fst(EVALS_SUM_CORE_FILE)
@@
 
-all_data <- rbind(core_data, package_data, kaggle_data)
+all_data <- rbind(core_data,
+                  package_data, 
+                  kaggle_data)
+
+head(all_data)
+
 ```
 
-## Side-Effects and Eval
+## Amount of computation via eval
 
-Each entry in `call.fst` contains `direct_writes` and `indirect_writes` (both
-`0` for `eval.parent` and `local`). `direct_writes` are the side-effects
-directly caused by the `eval` in question and `indirect_writes` are the
-side-effects caused by all children evals of the eval. All application
-assignments can be found in `direct_writes` field of `program.fst` and all
-`eval_assignments` can be found in `indirect_writes` field of `program.fst`. All
-non-local assignments can be found by simply summing up the `direct_writes`
-field of all eval calls.
-
-```{r preprocess-write-data}
-eval_writes_data <-
+```{r writes-data}
+writes_data <-
     all_data %>%
+    select(eval_source_type, eval_source, eval_call_srcref, direct_writes, indirect_writes) %>%
     filter(eval_function == "eval" | eval_function == "evalq") %>%
-    select(eval_source_type, eval_source, eval_call_srcref, direct_writes, indirect_writes, nb_ev_calls) %>%
     mutate(all_writes = direct_writes + indirect_writes, side_effecting = as.logical(all_writes))
-
-program_writes_data <-
-    program_data %>%
-    select(direct_writes, indirect_writes)
+# eval_events <-
+#  calls %>%
+#  filter(!is.na(eval_call_srcref)) %>%
+#  filter(eval_function == "eval" | eval_function == "evalq") %>%
+#  mutate(all_writes = direct_writes + indirect_writes) %>%
+#  select(package, file, eval_call_srcref, eval_call_id, builtin, special, closure, interpreter_eval, c_call, allocation, direct_writes, indirect_writes, all_writes) %>%
+#  pivot_longer(cols=c(builtin, special, closure, interpreter_eval, c_call, allocation, direct_writes, indirect_writes, all_writes), names_to="event", values_to="event_count")
 ```
 
-```{r compute-side-effecting-write-counts}
-all_writes <- sum(program_writes_data$direct_writes)
-eval_writes <- sum(program_writes_data$indirect_writes)
-
-r("All Writes", all_writes)
-r("Eval Writes", eval_writes)
-
-eval_side_effecting_writes <- sum(eval_writes_data$direct_writes * eval_writes_data$nb_ev_calls)
-
-r("Eval Side Effecting Writes", eval_side_effecting_writes)
-
-r("Eval Side Effecting Writes Eval Perc", ratio(eval_side_effecting_writes, eval_writes))
-r("Eval Side Effecting Writes All Perc", ratio(eval_side_effecting_writes, all_writes))
-```
-
-```{r compute-side-effecting-call-counts}
+```{r side-effects-analysis}
 eval_calls_by_source_writes <-
-    eval_writes_data %>%
-    count(eval_source_type, side_effecting, wt = nb_ev_calls, name = "call_count") %>%
-    group_by(eval_source_type) %>%
-    mutate(total_call_count = sum(call_count)) %>%
-    ungroup() %>%
-    mutate(call_perc = round(100 * call_count/total_call_count, 1)) %>%
-    filter(side_effecting)
+    writes_data %>%
+    count(eval_source_type, side_effecting, wt = nb_ev_calls, name = call_count)
 
@@
 
-eval_calls_by_source_writes %>%
-pmap(function(eval_source_type, side_effecting, call_count, total_call_count, call_perc) {
-    r(str_c("Side Effecting ", eval_source_type, " call perc"), str_c(call_perc, "%"))
-})
-
 eval_calls_by_package_writes <-
-    eval_writes_data %>%
+    writes_data %>%
     filter(eval_source_type != "core") %>%
-    count(eval_source, side_effecting, wt = nb_ev_calls, name = "call_count") %>%
+    count(eval_source, side_effecting, wt = nb_ev_calls, name = call_count) %>%
     arrange(desc(call_count))
@@
 datatable(eval_calls_by_package_writes)
-```
 
-The call site computation is incorrect.
-
-```{r compute-side-effecting-site-counts, eval = FALSE}
 eval_sites_by_source_writes <-
-    eval_writes_data %>%
+    writes_data %>%
     group_by(eval_source_type, eval_call_srcref) %>%
@@
     ungroup() %>%
-    count(eval_source_type, side_effecting, name = "site_count") %>%
-    group_by(eval_source_type) %>%
-    mutate(total_site_count = sum(site_count)) %>%
-    ungroup() %>%
-    mutate(site_perc = round(100 * site_count / total_site_count, 1)) %>%
-    filter(side_effecting)
+    count(eval_source_type, side_effecting, name = site_count)
 
@@
 
-eval_sites_by_source_writes %>%
-pmap(function(eval_source_type, side_effecting, site_count, total_site_count, site_perc) {
-    r(str_c("SideEffecting ", eval_source_type, " site perc"), str_c(site_perc, "%"))
-})
-
 eval_sites_by_package_writes <-
-    eval_writes_data %>%
+    writes_data %>%
     filter(eval_source_type != "core") %>%
@@
     ungroup() %>%
-    count(eval_source, side_effecting, name = "site_count") %>%
+    count(eval_source, side_effecting, name = site_count) %>%
     group_by(eval_source, side_effecting) %>%
@@
 datatable(eval_sites_by_package_writes)
+
 ```
+
+Each entry in `call.fst` contains `direct_writes` and `indirect_writes` (both `0` for `eval.parent` and `local`). 
+`direct_writes` are the side-effects directly caused by this `eval`. 
+`indirect_writes` are the side-effects caused by all children evals of this eval.
+
+For graphing, you may want to sum them up for each eval.
+I compute them separately because it is more informative (should we need this information for debugging or qualitative insights).
+Now, for normalization you have two quantities: all assignments done by the application and all assignments done by all evals (non-local or local).
+all application assignments can be found in `direct_writes` field of `program.fst` and all `eval_assignments` can be found in `indirect_writes` field of `program.fst`.
+You can also find all non-local assignments by simply summing up the `direct_writes` field of all eval calls.
+
+```{r}
+
+```

### 2020-08-27 inst/analysis/analysis_stable.Rmd
Fixes

--- 2020-08-26 inst/analysis/analysis_stable.Rmd
@@
 
-eval_calls <- bind_rows(eval_calls_packages, eval_call_core)
+eval_calls_raw <- bind_rows(eval_calls_packages, eval_call_core)
+```
 
+We remove a few columns we do not need here and deduplicate.
+```{r deduplicate}
+eval_calls <- eval_calls_raw %>% count(across(-c(builtin:indirect_allocation)), name = "nb_ev_calls2") %>%
+    mutate(nb_ev_calls = nb_ev_calls * nb_ev_calls2) %>% 
+    select(-nb_ev_calls2)
+```
+
+```{r document}
 # A flexible way to document the dataset: it is quick and easy to document a new column
@@
 ```{r}
-ast_sizes <- eval_calls %>% slice_sample(n=10000) %>% mutate(expr_resolved_ast_size = map_int(expr_resolved, expr_size_str))
+ast_sizes <- eval_calls %>% slice_sample(n=10000) %>% mutate(expr_resolved_nodes = map_int(expr_resolved, expr_size_str))
+# Now precomputed during the tracing
+#ast_sizes <- eval_calls
 ```
@@
 ```{r}
-# ast_sizes %>% ggplot(aes(x = expr_resolved_ast_size, weight = nb_ev_calls, group = dataset)) +
+# ast_sizes %>% ggplot(aes(x = expr_resolved_nodes, weight = nb_ev_calls, group = dataset)) +
 #   geom_histogram(aes(y = stat(width*density)), na.rm = TRUE, binwidth = 0.1) +
@@
 # ast_sizes %>% 
-#   ggplot(aes(x = dataset, y = expr_resolved_ast_size, fill = dataset, weight = nb_ev_calls)) +
+#   ggplot(aes(x = dataset, y = expr_resolved_nodes, fill = dataset, weight = nb_ev_calls)) +
 #   geom_flat_violin(trim=FALSE, width=.5) +
@@
 ast_sizes %>%
-  ggplot(aes(x = dataset, y = expr_resolved_ast_size, fill = dataset, weight = nb_ev_calls)) +
+  ggplot(aes(x = dataset, y = expr_resolved_nodes, fill = dataset, weight = nb_ev_calls)) +
   geom_violin(trim = FALSE, width=.5)+
@@
 ggsave(path(PLOT_DIR, "ast_sizes.pdf"))
+
+ast_sizes_prop <- ast_sizes %>% count(expr_resolved_nodes, wt = nb_ev_calls) %>%
+                      mutate(proportion = n / sum(n)) %>% 
+                      arrange(expr_resolved_nodes) %>% 
+                      mutate(cumulative = cumsum(proportion))
+
+ast_size_one <- ast_sizes %>% filter(expr_resolved_nodes == 1) 
+nb_ast_size_one <- count(ast_size_one, wt = nb_ev_calls)$n
+nb_ast_size_one_call_sites <- nb_eval_call_sites(ast_size_one)
+r("nb ast one percent", ratio(nb_ast_size_one, sum(ast_sizes_prop$n)))
+r("nb ast one call site percent", ratio(nb_ast_size_one_call_sites, nb_eval_call_sites(ast_sizes)))
+
+nb_nodes_many <- ast_sizes_prop %>% filter(cumulative >= 0.95) %>% .[1,]
+
+r("ast size nine five", nb_nodes_many$expr_resolved_nodes)
+r("ast size nine five exact percent", percent(nb_nodes_many$cumulative))
+
 ```
@@
 # Case studies
+
+```{r}
+duration <- difftime(Sys.time(), now)
+```
+
+Notebook execution was `r duration` `r units(duration)`.

### 2020-08-27 inst/analysis/inc/paths.R

@@ -34,7 +34,7 @@ KAGGLE_KERNELS_FILE      <- path(DATA_DIR, "kaggle-kernels.fst")
 KAGGLE_LOG_FILE          <- path(DATA_DIR, "parallel-kaggle.fst")
 KAGGLE_PROGRAMS_FILE     <- path(DATA_DIR, "kaggle-programs.fst")
 
-RUN_TRACE_DIR  <- path(RUN_DIR, "package-evals-traced")
+RUN_TRACE_DIR  <- path(RUN_DIR, "package-evals-traced.2")
 RUN_KAGGLE_DIR <- path(RUN_DIR, "kaggle-run")
 
 CALLS_FINAL_FILE        <- path(RUN_TRACE_DIR, "calls.fst")

### 2020-08-27 inst/analysis/insights.R

@@ -77,7 +77,7 @@ extract_args_parse <- function(eval_call)
   args <- map_chr(as.list(exp[-1]), function(chr) { paste(deparse(chr), collapse = "\n")})
   names(args) <- map_chr(names(args), function(chr) paste0("parse_args_", chr))
   
-  return(args[str_detect(names(args), "^parse_args_(file_name|text|n|s|keep\\.source|srcfile|dots)$")])# "file|text|n|s|prompt|keep.source|srcfile|code"
+  return(args[str_detect(names(args), "^parse_args_(file|text|n|s|keep\\.source|srcfile|dots)$")])# "file|text|n|s|prompt|keep.source|srcfile|code"
 }
 
 # See extract_inner_exp which takes a str

### 2020-08-27 inst/paper/main.tex
AST size percentages

--- 2020-08-27 inst/paper/main.tex
@@
 size of the \c{expr} argument to \eval. The majority of inputs across core
-and packages (\miss{XX}\%) is 1. That is to say, the code being evaluated
+and packages (\NbAstOnePercent of \eval calls and \NbAstOneCallSitePercent of call sites) is 1. That is to say, the code being evaluated
 consists of a single variable name or constant. A function call needs at
-least three nodes. Inputs smaller than \c{X} nodes account for 95\% of all
+least three nodes. Inputs smaller than \AstSizeNineFive nodes account for 95\% of all
 calls. Sizes are roughly similar between core and packages.

### 2020-08-27 inst/analysis/enrich_dataset.Rmd

@@
 r("undefined eval expressions", nb_different_call_expr)
-r("undefined out of corpus", nb_out_corpus)
-r("undefined out of coprus percent", ratio(nb_out_corpus, nb_undefined))
+r("undefined out of corpus", nb_out_corpus$n)
+r("undefined out of coprus percent", ratio(nb_out_corpus$n, nb_undefined$n))
 ```

### 2020-08-27 inst/paper/main.tex
Parse

@@
 > env <- new.env()
-> env$r <- 9
+> env\$r <- 9
 > eval(r, envir = env)
@@
 
-Although \c{eval} takes an R expression, it can evaluate a string if this string is first parsed with \c{parse}, \c{str2expression} or \c{str2lang}.
-This happens for \PercentParsedEvals of the eval calls, and \PercentParsedCallSites of the call sites.
+\c{eval} takes an R expression. To evaluate a string, it first has to be parse
+using either \c{parse}, \c{str2expression} or \c{str2lang} functions. The
+difference between them is that the latter two are specialized versions of
+\c{parse} that do not keep source references (line and column numbers). This
+happens for \PercentParsedEvals of the eval calls, and \PercentParsedCallSites
+of the call sites. The input to \c{parse} can come from either a string, a file
+or directly a from a user input in interactive REPL session\footnote{in the
+  corpus, there is only one call site to eval that takes parse expression from a
+  user input, \c{invokeRestartInteractively} that prompts user for new argument
+  values when a function invocation fails.}.
 
-This corresponds to strings loaded from a file or to metaprogramming, \ie building new variables or functions by concatenating strings together, as in Listing~\ref{code:parsepaste}.
-To detect this kind of metaprogramming, we look at the usage of the main concatenating functions in R, \c{paste}, \c{paste0}, \c{glue} and \c{str_c}, and \PasteParsedEvals of the evals and \PasteParsedCallSite of the call sites using a parse function actually call a concatenating function.
+This corresponds to strings loaded from a file or to metaprogramming, \ie
+building new variables or functions by concatenating strings together, as in
+Listing~\ref{code:parsepaste}. To detect this kind of metaprogramming, we look
+at the usage of the main concatenating functions in R, \c{paste}, \c{paste0},
+\c{glue} and \c{str_c}, and \PasteParsedEvals of the evals and
+\PasteParsedCallSite of the call sites using a parse function actually call a
+concatenating function.
 

### 2020-08-27 inst/analysis/corpus-stage2.Rmd
Added programs counting

@@
 ```
+
+### Programs
+
+```{r}
+package_runs <- trace_runs_raw %>%
+  mutate(
+    dir=as.character(path("../../run/package-evals-traced", package, basename(file))),
+    file=as.character(path("./", basename(dirname(file)), basename(file))),
+  ) %>%
+  semi_join(corpus, by="package")
+
+package_programs <-
+  left_join(
+    filter(package_runs, exitval==0) %>% select(package, file, dir, time), 
+    runnable_code_raw %>% select(package, file=filename, code),
+    by=c("package", "file")
+  ) %>%
+  filter(!endsWith(file, "tests/testthat.R"), !endsWith(file, "tests/run-all.R")) %>% # testthat drivers are not traced
+  transmute(
+    type=basename(dirname(file)),
+    file=as.character(path(dir, "calls.fst")),
+    code, 
+    runtime=time,
+    corpus="cran"
+  )
+
+package_programs_errors <- 
+  package_programs %>%
+  mutate(
+    error_file=path(dirname(str_replace(file, fixed("../.."), params$base_dir)), "calls-ERROR"), 
+    error=file.exists(error_file)
+  ) %>%
+  filter(error) %>%
+  rowwise() %>%
+  do(mutate(read_fst(.$error_file), file=.$file))
+
+package_programs <-
+  left_join(
+    package_programs, 
+    transmute(package_programs_errors, error=paste(source, message, call), file), 
+    by="file"
+  ) %>%
+  mutate(
+    success=
+      type %in% c("examples", "tests", "vignettes") & 
+      !is.na(code) &
+      !is.na(runtime) &
+      code > 0 &
+      is.na(error)
+  )
+```
+
+```{r}
+write_fst(package_programs, CRAN_PROGRAMS_FILE)
+```

### 2020-08-27 inst/analysis/inc/paths.R

@@ -50,12 +50,11 @@ CORPUS_FILE            <- path(DATA_DIR, "corpus.txt")
 CORPUS_DETAILS_FILE    <- path(DATA_DIR, "corpus.fst")
 CORPUS_PLOT            <- path(PLOT_DIR, "corpus.pdf")
 
-RUN_TRACE_DIR <- path(params$base_dir, "inst", "analysis")
-PROGRAM_FILE          <- path(RUN_TRACE_DIR, "program.fst")
-EVALS_RAW_FILE        <- path(RUN_TRACE_DIR, "raws.fst")
-EVALS_SUM_CORE_FILE   <- path(RUN_TRACE_DIR, "summarized-core.fst")
-EVALS_SUM_PKGS_FILE   <- path(RUN_TRACE_DIR, "summarized-packages.fst")
-EVALS_SUM_KAGGLE_FILE <- path(RUN_TRACE_DIR, "summarized-kaggle.fst")
+PROGRAM_FILE             <- path(RUN_TRACE_DIR, "program.fst")
+EVALS_RAW_FILE           <- path(RUN_TRACE_DIR, "raws.fst")
+EVALS_SUM_CORE_FILE      <- path(RUN_TRACE_DIR, "summarized-core.fst")
+EVALS_SUM_PKGS_FILE      <- path(RUN_TRACE_DIR, "summarized-packages.fst")
+EVALS_SUM_KAGGLE_FILE    <- path(RUN_TRACE_DIR, "summarized-kaggle.fst")
 EVALS_SUM_EXTERNALS_FILE <- path(RUN_TRACE_DIR, "summarized-externals.fst")
 
 ################################################################################

### 2020-08-27 inst/paper/main.tex
Add implementation usage scenario

@@
 
+In the corpus, we observe \AllEventCountRnd events. Only \EvalEventCountRnd
+events occur inside eval. This is just \EvalEventAllPerc of all events.
+Figure~\ref{fig:event-distribution} shows the distribution of events for
+core and package \eval calls.
+
 {\bf Missing from this section is the measure of the number of events
@@
 
+  \subsubsection{Implementation}
+  The package installation process adds an extra file to the package directory
+  that contains the code to load the package code inside a \local block. It
+  creates a namespace object and injects the package bindings.
+
   \subsubsection{Miscellaneous} \eval is also used in R packages to get around

### 2020-08-27 inst/analysis/enrich_dataset.Rmd
Getting rid of the -finals.fst

@@
 
-The analysis is being performed on a dataset read from file `r CALLS_FINAL_FILE`.
+The analysis is being performed on a dataset read from file `r CALLS_FILE`.
 
 ```{r file-info, echo=FALSE}
-file_info(CALLS_FINAL_FILE)
+file_info(CALLS_FILE)
 ```
@@
 ```{r load-data, echo=FALSE}
+corpus <- read_fst(CORPUS_DETAILS_FILE)
+
 eval_calls_raw <-
-    read_fst(CALLS_FINAL_FILE) %>%
+    read_fst(CALLS_FILE) %>%
     as_tibble() %>%
-    mutate(corpus = "cran")
+    mutate(
+      package=basename(dirname(dirname(file))),
+      corpus="cran"
+    ) %>%
+    semi_join(corpus, by="package") # There might be more packages traced than what is in the corpus
 
 eval_calls_kaggle_raw <- 
-  read_fst(KAGGLE_CALLS_FINAL_FILE) %>%
+  read_fst(KAGGLE_CALLS_FILE) %>%
   as_tibble() %>% 
-  mutate(corpus = "kaggle")
+  mutate(
+    package=basename(dirname(file)),
+    corpus="kaggle"
+  )
 
@@
 ```{r add_package}
-eval_calls <- eval_calls %>% add_package()
+# The package is already added when the files are loaded
+#eval_calls <- eval_calls %>% add_package()
 ```

### 2020-08-27 inst/analysis/inc/paths.R

@@ -29,17 +29,14 @@ PACKAGE_TRACE_RUNS_FILE    <- path(DATA_DIR, "run-trace.fst")
 PACKAGE_TRACE_LOG_FILE     <- path(DATA_DIR, "parallel-trace.fst")
 CRAN_PROGRAMS_FILE         <- path(DATA_DIR, "cran-programs.fst")
 
+RUN_TRACE_DIR  <- path(RUN_DIR, "package-evals-traced.2")
+RUN_KAGGLE_DIR <- path(RUN_DIR, "kaggle-run")
+
 KAGGLE_DEPENDENCIES_FILE <- path(DATA_DIR, "kaggle-dependencies.txt")
 KAGGLE_KERNELS_FILE      <- path(DATA_DIR, "kaggle-kernels.fst")
 KAGGLE_LOG_FILE          <- path(DATA_DIR, "parallel-kaggle.fst")
 KAGGLE_PROGRAMS_FILE     <- path(DATA_DIR, "kaggle-programs.fst")
-
-RUN_TRACE_DIR  <- path(RUN_DIR, "package-evals-traced.2")
-RUN_KAGGLE_DIR <- path(RUN_DIR, "kaggle-run")
-
-CALLS_FINAL_FILE        <- path(RUN_TRACE_DIR, "calls.fst")
-CALLS_FINAL_FILE        <- path(RUN_TRACE_DIR, "calls-final.fst")
-KAGGLE_CALLS_FINAL_FILE <- path(RUN_KAGGLE_DIR, "calls-final.fst")
+KAGGLE_CALLS_FILE        <- path(RUN_KAGGLE_DIR, "calls.fst")
 
 ## OUTPUTs
 
@@ -50,7 +47,9 @@ CORPUS_FILE            <- path(DATA_DIR, "corpus.txt")
 CORPUS_DETAILS_FILE    <- path(DATA_DIR, "corpus.fst")
 CORPUS_PLOT            <- path(PLOT_DIR, "corpus.pdf")
 
+CALLS_FILE               <- path(RUN_TRACE_DIR, "calls.fst")
 PROGRAM_FILE             <- path(RUN_TRACE_DIR, "program.fst")
+# TODO: rename to SUMMARIZED_...
 EVALS_RAW_FILE           <- path(RUN_TRACE_DIR, "raws.fst")
 EVALS_SUM_CORE_FILE      <- path(RUN_TRACE_DIR, "summarized-core.fst")
[…]

### 2020-08-27 inst/analysis/analysis_stable.Rmd

@@
 theme_set(theme_minimal())
+theme_update(text = element_text(rel(1.5)))
+
 create_tags(path(TAGS_DIR, "analysisStable.tex"), prefix="", default=TRUE)
@@
   theme(
+    #text=element_text(rel=2),
     legend.box="horizontal",
@@
 ```{r environment_hierarchy}
-# classified_envir <- eval_calls %>% filter(!is.na(envir_expression)) %>%
-#       mutate(env_class = pmap(list(environment_class, envir_type, envir_expression), extract_envir)) %>%
-#       unnest_wider(env_class)
+classified_envir <- eval_calls %>% filter(!is.na(envir_expression)) %>%
+      mutate(env_class = pmap(list(environment_class, envir_type, envir_expression), extract_envir)) %>%
+      hoist(env_class, read_env = 1, write_env = 2, .remove=FALSE) #Quicker when not removing
 
-#TODO
+```
+
+```{r write_env}
+
+classified_envir %>% ggplot() +
+  geom_bar(aes(x = write_env, weight = nb_ev_calls, group = dataset)) +
+  labs(x = "Environment write")
+```
+
+```{r read_env}
+
+classified_envir %>% ggplot() +
+  geom_bar(aes(x = read_env, weight = nb_ev_calls, group = dataset)) +
+  labs(x = "Environment reads")
 ```

### 2020-08-27 inst/analysis/kaggle-stage2.Rmd

---
title: "Kaggle stage2"
output: html_document
params:
  base_dir: /var/lib/R/project-evalR
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
for (x in c(
  "dplyr", 
  "DT",
  "fs",
  "fst",
  "lubridate",
  "readr",
  "purrr",
  "stringr",
  "tidyr"
  )) {
  suppressPackageStartupMessages(library(x, character.only=TRUE))
}

knitr::opts_chunk$set(echo = TRUE)

# TODO: move into the package
source("inc/paths.R")
source("inc/setup.R")
```

## Load data

```{r loading raw data}
kaggle_kernels <- as_tibble(read_fst(KAGGLE_KERNELS_FILE))
kaggle_log <- as_tibble(read_fst(KAGGLE_LOG_FILE)) %>%
  transmute(
    id=str_replace(Command, ".*/kaggle-run/[^/]+/([^/]+)/.*", "\\1"),
    exitval=Exitval,
    runtime=as.duration(JobRuntime)
  )
```

### Programs

```{r}
kaggle_programs <- left_join(
  kaggle_kernels,
  kaggle_log,
  by="id"
) %>%
  mutate(
    file=path("../../run/kaggle-run", competition, id, "calls.fst"),
    corpus="kaggle"
  ) %>%
  select(
    file,
    id,
    language,
    kernel_type,
    competition,
    runtime,
    code,
    exitval,
    corpus
  )

kaggle_programs_errors <- 
  kaggle_programs %>%
  mutate(
    error_file=path(dirname(str_replace(file, fixed("../.."), params$base_dir)), "calls-ERROR"), 
    error=file.exists(error_file)
  ) %>%
  filter(error) %>%
  rowwise() %>%
  do(mutate(read_fst(.$error_file), file=.$file))

kaggle_programs <-
    left_join(
      kaggle_programs, 
      if (nrow(kaggle_programs_errors) > 0) {
        transmute(kaggle_programs_errors, error=paste(source, message, call), file)
      } else {
        transmute(kaggle_programs, error=NA, file)
      }, 
      by="file"
    ) %>%
    mutate(
      success=
        exitval==0 &
        !is.na(code) &
        !is.na(runtime) &
        code > 0 &
        is.na(error)
    )
```

```{r}
write_fst(kaggle_programs, KAGGLE_PROGRAMS_FILE)
```

### 2020-08-27 inst/paper/main.tex
Add tags and comment event description
precise default write environment

--- 2020-08-27 inst/paper/main.tex
@@
 \newcommand{\source}{\texttt{source}\xspace}
+\newcommand{\local}{\texttt{local}\xspace}
 \newcommand{\unlockBinding}{\texttt{unlockBinding}\xspace}
@@
 
+%%In the corpus, we observe \AllEventCountRnd events. Only \EvalEventCountRnd
+%%events occur inside eval. This is just \EvalEventAllPerc of all events.
+%%Figure~\ref{fig:event-distribution} shows the distribution of events for
+%%core and package \eval calls.
+
 {\bf Missing from this section is the measure of the number of events
@@
 \begin{itemize}
-	\item in which environment \c{eval} writes new bindings 
+	\item the \emph{default} assignment environment \c{eval} writes new bindings.\footnote{It is the default one but the expression in \eval can always write in any other environments using \c{assign} or in the parent environment using \c{<<-}.}
 	\item in which \emph{existing} environment \c{eval} reads. It is the first non new environment. \c{eval} will also be able to read in all parent environment of that environment. 
@@
   calling the new function with explicitly selecting the arguments.
+
+  \subsubsection{Implementation}
+  The package installation process adds an extra file to the package directory
+  that contains the code to load the package code inside a \local block. It
+  creates a namespace object and injects the package bindings.
 

### 2020-08-27 inst/analysis/analysis_stable.Rmd
Environments

--- 2020-08-27 inst/analysis/analysis_stable.Rmd
@@
 theme_set(theme_minimal())
+theme_update(text = element_text(size = 20))
+
 create_tags(path(TAGS_DIR, "analysisStable.tex"), prefix="", default=TRUE)
 
+# Temp 
+# RUN_DIR   <- path(params$base_dir, "run")
+# RUN_TRACE_DIR  <- path(RUN_DIR, "package-evals-traced")
+# EVALS_RAW_FILE           <- path(RUN_TRACE_DIR, "raws.fst")
+# EVALS_SUM_CORE_FILE      <- path(RUN_TRACE_DIR, "summarized-core.fst")
+# EVALS_SUM_PKGS_FILE      <- path(RUN_TRACE_DIR, "summarized-packages.fst")
+# EVALS_SUM_KAGGLE_FILE    <- path(RUN_TRACE_DIR, "summarized-kaggle.fst")
+# EVALS_SUM_EXTERNALS_FILE <- path(RUN_TRACE_DIR, "summarized-externals.fst")
 packages_filepath <- EVALS_SUM_PKGS_FILE
@@
 ```{r deduplicate}
-eval_calls <- eval_calls_raw %>% count(across(-c(builtin:indirect_allocation)), name = "nb_ev_calls2") %>%
-    mutate(nb_ev_calls = nb_ev_calls * nb_ev_calls2) %>% 
-    select(-nb_ev_calls2)
+# eval_calls <- eval_calls_raw %>% count(across(-c(builtin:indirect_allocation)), name = "nb_ev_calls2") %>%
+#     mutate(nb_ev_calls = nb_ev_calls * nb_ev_calls2) %>% 
+#     select(-nb_ev_calls2)
+
+#
+eval_calls <- eval_calls_raw
 ```
@@
 nb_call_sites_default_envir <- nb_eval_call_sites(default_envir)
+
+r("default envir expr", nb_default_envir$n)
+r("default envir expr percent", ratio(nb_default_envir$n, nb_eval_calls))
+r("default envir  call sites percent", ratio(nb_call_sites_default_envir, nb_call_sites))
 ```
 
-`r ratio(nb_default_envir, nb_eval_calls)`% of the calls use the default environment, which corresponds to `r ratio(nb_call_sites_default_envir, nb_call_sites)`% of the call sites.
+`r ratio(nb_default_envir$n, nb_eval_calls)`% of the calls use the default environment, which corresponds to `r ratio(nb_call_sites_default_envir, nb_call_sites)`% of the call sites.
 
@@
 #       mutate(env_class = pmap(list(environment_class, envir_type, envir_expression), extract_envir)) %>%
-#       unnest_wider(env_class)
+#       hoist(env_class, read_env = 1, write_env = 2, .remove=FALSE) #Quicker when not removing
 
-#TODO
+classified_envir <- eval_calls %>% filter(!is.na(envir_expression)) %>% 
+  mutate(env_class = pmap_chr(list(environment_class, envir_type, envir_expression), simplify_envir)) %>% 
+  mutate(env_class_read = map_chr(env_class, ~ str_match(., "([^\\+]*)\\+?")[[2]]))
+```
+
+```{r envirall}
+classified_envir %>% ggplot() +
+  geom_bar(aes(x =fct_rev(fct_infreq(env_class)), weight = nb_ev_calls )) +
+  coord_flip() +
+  scale_y_log10() +
+  labs(x = "Environments") + 
+  facet_wrap(~dataset)
+
+ggsave(path(PLOT_DIR, "environments.pdf"))
+```
+
+```{r envir_read}
+classified_envir %>% ggplot() +
+  geom_bar(aes(x =fct_rev(fct_infreq(env_class_read)), weight = nb_ev_calls )) +
+  coord_flip() +
+  scale_y_log10() +
+  labs(x = "Environments") + 
+  facet_wrap(~dataset)
+
+ggsave(path(PLOT_DIR, "environments-read.pdf"))
 ```

### 2020-08-27 inst/analysis/corpus-stage2.Rmd
Environments

@@
 runnable_code_raw <- read_task_result(PACKAGE_RUNNABLE_CODE_FILE)
-runs_raw          <- read_task_result(PACKAGE_RUNS_FILE)
 trace_runs_raw    <- read_task_result(PACKAGE_TRACE_RUNS_FILE)
@@
     runnable_code_examples=code_examples,
-    runnable_code_tests=code_tests,
+#    runnable_code_tests=code_tests,
     runnable_code_vignettes=code_vignettes,
     runnable_files_examples=files_examples,
-    runnable_files_tests=files_tests,
+#    runnable_files_tests=files_tests,
     runnable_files_vignettes=files_vignettes,
@@
   mutate(
-    runnable_code=runnable_code_examples+runnable_code_tests+runnable_code_vignettes,
-    runnable_files=runnable_files_examples+runnable_files_tests+runnable_files_vignettes
+#    runnable_code=runnable_code_examples+runnable_code_tests+runnable_code_vignettes,
+#    runnable_files=runnable_files_examples+runnable_files_tests+runnable_files_vignettes
+    runnable_code=runnable_code_examples+runnable_code_vignettes,
+    runnable_files=runnable_files_examples+runnable_files_vignettes
   ) %>%
@@
   mutate(coverage_expr=coverage_expr/100, coverage_line=coverage_line/100)
-```
-
-### Package runs
-
-```{r process package runs}
-runs <-
-  runs_raw %>%
-  group_by(package) %>%
-  summarise(runnable_success=sum(is.na(error)), runnable_error=sum(!is.na(error)), runnable_time=sum(time, na.rm=TRUE)) %>%
-  mutate(runnable_rate=runnable_success/(runnable_success+runnable_error))
 ```
@@
   left_join(coverage, by="package") %>%
-  left_join(runs, by="package") %>%
   left_join(trace_runs, by="package") %>%

### 2020-08-27 inst/analysis/counts.Rmd
Environments

--- 2020-08-24 inst/analysis/counts.Rmd
@@
 title: "Empirical Evaluation of eval in R"
-output:
-  rmdformats::readthedown:
-        code_folding: hide
-        lightbox: true
-        gallery: false
-        df_print: paged
-        toc_depth: 3
-date: "`r Sys.Date()`"
+output: html_document
+params:
+  base_dir: /var/lib/R/project-evalR
 editor_options: 
   chunk_output_type: console
-params:
-    base_dir: /home/aviral/projects/revalstudy2
-    kaggle_datafile: "summarized-kaggle.fst"
-    core_datafile: "summarized-core.fst"
-    packages_datafile: "summarized-packages.fst"
 ---
@@
 source("inc/latextags.R")
-create_tags(path("../paper/tag", "counts.tex"), prefix="", default=TRUE)
+
+create_tags(path(TAGS_DIR, "counts.tex"), prefix="", default=TRUE)
 ```
@@
 ```{r eval=TRUE}
-core_data <- read_fst(params$core_datafile)
+run_trace_data <- read_fst(PACKAGE_TRACE_RUNS_FILE)
+parallel_kaggle_data <- read_fst(KAGGLE_LOG_FILE)
+
+core_data <- read_fst(EVALS_SUM_CORE_FILE)
 packages_data <- 
-    read_fst(params$packages_datafile) %>%
+    read_fst(EVALS_SUM_PKGS_FILE) %>%
     mutate(eval_source_type = "package")
+
+# kaggle should be always empty
 kaggle_data <- 
-    read_fst(params$kaggle_datafile)
+    read_fst(EVALS_SUM_KAGGLE_FILE)
     #mutate(eval_source_type == "core")
 
-all_data <- bind_rows(core_data, packages_data, kaggle_data)
+all_data <- bind_rows(core_data, packages_data, kaggle_data) %>% as_tibble()
 all_data <-
@@
 
-r("total file count", total_file_count)
-r("no eval file count", no_eval_file_count)
-r("no eval file perc", ratio(no_eval_file_count, total_file_count))
-r("core eval file count", core_eval_file_count)
-r("core eval file perc", ratio(core_eval_file_count, total_file_count))
-r("package eval file count", package_eval_file_count)
-r("package eval file perc", ratio(package_eval_file_count, total_file_count))
-r("all eval file count", all_eval_file_count)
+overview_table(    
+  r("total file count", total_file_count),
+  r("no eval file count", no_eval_file_count),
+  r("no eval file perc", ratio(no_eval_file_count, total_file_count)),
+  r("core eval file count", core_eval_file_count),
+  r("core eval file perc", ratio(core_eval_file_count, total_file_count)),
+  r("package eval file count", package_eval_file_count),
+  r("package eval file perc", ratio(package_eval_file_count, total_file_count)),
+  r("all eval file count", all_eval_file_count),
   r("all eval file perc", ratio(all_eval_file_count, total_file_count))
+)
 
@@
 ```{r callsite-distribution, eval=TRUE}
-```{r call-site-table}
 eval_callsite_table <-
@@
 ```{r core-dependent-own-eval}
+all_data <- bind_rows(core_data, packages_data, kaggle_data) %>%
+  as_tibble() %>%
+  mutate(
+    package=basename(dirname(dirname(file))),
+    core_own_dependent=case_when(
+      eval_source == "core"  ~ "core",
+      eval_source == "base"  ~ "core",
+      package == eval_source ~ "own",
+      TRUE ~ "dependent"
+    )
+  )
 
+# TODO - we need to summarize the data better
+# all paths must start with this
+stopifnot(nrow(filter(all_data, !startsWith(file, "../../run/"))) == 0)
 ```
+
+```{r}
+cran_programs <- read_fst(CRAN_PROGRAMS_FILE) %>% 
+  as_tibble() %>%
+  filter(success) %>%
+  select(file, corpus)
+
+kaggle_programs <- read_fst(KAGGLE_PROGRAMS_FILE) %>% 
+  as_tibble() %>%
+  filter(success) %>%
+  select(file, corpus)
+
+all_programs <- bind_rows(cran_programs, kaggle_programs)
+```
+
+How many programs are missing in the summarized data
+
+```{r}
+anti_join(all_programs, all_data, by="file")
+```
+
+```{r}
+all_programs_cod <-
+  map_dfr(
+    c("core", "own", "dependent"), 
+    ~mutate(
+      all_programs,
+      core_own_dependent=.,
+      nb_ev_calls=0
+    )
+  )
+
+all_data_summary <- 
+  all_data %>%
+  select(file, core_own_dependent, corpus, nb_ev_calls) %>%
+  right_join(all_programs_cod, by=c("file", "corpus", "core_own_dependent")) %>%
+  mutate(nb_ev_calls=ifelse(is.na(nb_ev_calls.x), nb_ev_calls.y, nb_ev_calls.x)) %>%
+  select(file, core_own_dependent, corpus, nb_ev_calls)
+
+core_own_dependent <-
+  all_data_summary %>%
+  group_by(file, core_own_dependent) %>%
+  summarize(corpus=first(corpus), eval_calls = sum(nb_ev_calls)) %>%
+  ungroup()
+
+# %>%
+#   group_by(corpus, core_own_dependent) %>%
+#   summarize(
+#     total_eval_calls = sum(eval_calls),
+#     max_eval_calls = max(eval_calls),
+#     min_eval_calls = min(eval_calls),
+#     avg_eval_calls = mean(eval_calls)
+#   )
+
+vars <- expand.grid(corpus=c("cran", "kaggle"), cod=c("core", "own", "dependent"), stringsAsFactors = FALSE)
+map2_dfr(vars$corpus, vars$cod, function(crps, cod) {
+  print(filter(core_own_dependent, corpus==crps, core_own_dependent==cod))
+  rs(paste(crps, cod, "eval calls"), filter(core_own_dependent, corpus==crps, core_own_dependent==cod)$eval_calls)
+}) %>%
+  overview_table()
+
+# core_own_dependent  %>%
+#   group_by(corpus, core_own_dependent) %>%
+#   pivot_longer(cols=c(total_eval_calls, max_eval_calls, min_eval_calls, avg_eval_calls))
+# 
+#   xtable::xtable(align="llrrrr") %>%
+#   print(
+#     tabular.environment = "tabularx", 
+#     width = "\\textwidth",
+#     floating=F,
+# #    file=fname,
+#     include.rownames=F
+#   )
+
+# core_own_dependent %>%
+# pmap_dfr(function(corpus, core_own_dependent, total_eval_calls, max_eval_calls, min_eval_calls, avg_eval_calls) {
+#     r(str_c(corpus, " ", core_own_dependent, " total eval calls"), total_eval_calls)
+#     r(str_c(corpus, " ", core_own_dependent, " max eval calls"), max_eval_calls)
+#     r(str_c(corpus, " ", core_own_dependent, " min eval calls"), min_eval_calls)
+#     r(str_c(corpus, " ", core_own_dependent, " avg eval calls"), avg_eval_calls)
+# })
+# 
+# datatable(core_own_dependent)
+# 
+#     xtable::xtable(align="lXrrrr") %>%
+#     print(
+#       tabular.environment = "tabularx", 
+#       width = "\\textwidth",
+#       floating=F,
+#       file=fname,
+#       include.rownames=F
+#     )
+```

### 2020-08-27 inst/analysis/enrich_dataset.Rmd
Environments

@@
 ```{r setup, include=FALSE}
+knitr::knit_hooks$set(time_it = local({
+  now <- NULL
+  function(before, options) {
+    if (before) {
+      # record the current time before each chunk
+      now <<- Sys.time()
+    } else {
+      # calculate the time difference after a chunk
+      res <- difftime(Sys.time(), now)
+      # return a character string to show the time
+      paste("Time for this code chunk to run:", res)
+    }
+  }
+}))
+
 knitr::opts_chunk$set(echo = TRUE,
@@
                       fig.width = 10,
-                      cache.lazy = FALSE)
+                      cache.lazy = FALSE,
+                      time_it = TRUE)
 

### 2020-08-27 inst/analysis/inc/paths.R

@@ -24,7 +24,6 @@ PACKAGE_SLOC_FILE          <- path(DATA_DIR, "sloc.fst")
 PACKAGE_REVDEPS_FILE       <- path(DATA_DIR, "revdeps.fst")
 PACKAGE_FUNCTIONS_FILE     <- path(DATA_DIR, "functions.fst")
 PACKAGE_RUNNABLE_CODE_FILE <- path(DATA_DIR, "runnable-code-metadata.fst")
-PACKAGE_RUNS_FILE          <- path(DATA_DIR, "run.fst")
 PACKAGE_TRACE_RUNS_FILE    <- path(DATA_DIR, "run-trace.fst")
 PACKAGE_TRACE_LOG_FILE     <- path(DATA_DIR, "parallel-trace.fst")
 CRAN_PROGRAMS_FILE         <- path(DATA_DIR, "cran-programs.fst")

### 2020-08-27 inst/analysis/insights.R

@@ -362,9 +362,26 @@ groupify_function <- function(expr_function)
   )
 }
 
+build_env_rep <- function(env_class)
+{
+  splitted <- str_split(env_class, fixed("+"))[[1]]
+  return(paste0(splitted[[length(splitted)]], if(length(splitted) > 1) "+" else ""))
+}
+
+simplify_envir <- function(env_class, envir_type, envir_expression)
+{
+  # Add a + to the last environment if the first is new
+  case_when(
+    is.na(env_class) & envir_type == "VECSXP" ~ "list",
+    is.na(env_class) & envir_type == "NILSXP" ~ "NULL",
+    is.na(env_class) & envir_type == "INTSXP" ~ paste0("sys.call(", envir_expression, ")"),
+    TRUE ~ build_env_rep(env_class)
+    )
+}
+
 extract_write_envir <- function(env_class)
 {
-  return(str_split(env_class, fixed("+"), n = 2)[[1]])
+  return(str_split(env_class, fixed("+"), n = 2)[[1]][[1]])
 }
 
 
@@ -372,9 +389,9 @@ extract_envir <- function(env_class, envir_type, envir_expression)
 {
   # First element is write, second is read
   case_when(
-    is.na(env_class) && envir_type == "VECSXP" ~ c("list", "enclos"),
-    is.na(env_class) && envir_type == "NILSXP" ~ c("NULL", "enclos"),
-    is.na(env_class) && envir_type == "INTSXP" ~ {e <- paste0("sys.call(", envir_expression, ")"); c(e, e)},
+    is.na(env_class) & envir_type == "VECSXP" ~ c("list", "enclos"),
+    is.na(env_class) & envir_type == "NILSXP" ~ c("NULL", "enclos"),
+    is.na(env_class) & envir_type == "INTSXP" ~ {e <- paste0("sys.call(", envir_expression, ")"); c(e, e)},
     str_ends(env_class, fixed("global")) ~ c(extract_write_envir(env_class), "global"),
     str_ends(env_class, fixed("base")) ~ c(extract_write_envir(env_class), "base"),
     str_ends(env_class, fixed("callee")) ~ c(extract_write_envir(env_class), "callee"),
@@ -384,4 +401,5 @@ extract_envir <- function(env_class, envir_type, envir_expression)
     str_ends(env_class, "package:.*") ~ c(extract_write_envir(env_class), str_extract(env_class, "package:.*"))
   )
 }
+
 

### 2020-08-27 inst/analysis/corpus-stage2.Rmd
Fix missing package in cran-programs.fst

@@
   transmute(
+    package,
     type=basename(dirname(file)),

### 2020-08-27 inst/analysis/corpus-stage2.Rmd
Fixes in pipeline

@@
   mutate(
-    dir=as.character(path("../../run/package-evals-traced", package, basename(file))),
+    dir=as.character(path("../../run/package-evals-traced.2", package, basename(file))),
     file=as.character(path("./", basename(dirname(file)), basename(file))),
   ) %>%
-  semi_join(corpus, by="package")
+  semi_join(corpus, by="package") %>%
+  mutate(
+    call_file=path(paste0(tools::file_path_sans_ext(out_file), ".R"), "calls.fst")
+  ) %>%
+  filter(file.exists(call_file))
 
@@
   )
+
+package_programs <-
+  package_programs %>%
+  filter(file_exists(path(params$base_dir, str_replace(file, fixed("../../"), ""))))
 ```
@@
 ```
+
+```{r remove tests, eval=FALSE}
+df <- read_csv(path(params$base_dir, "runnable-code-metadata.csv"))
+df <- mutate(df, file=path(params$base_dir, "run", "package-evals-traced.2", basename(package), basename(filename)), type=basename(dirname(filename)), file_e=dir.exists(file))
+df <- filter(df, !(type %in% c("examples", "vignettes")), file_e)
+```
+
+```{r sanity check, eval=FALSE}
+x <- read_fst("/var/lib/R/project-evalR/run/package-evals-traced.2/calls.fst")
+y <- x %>% mutate(package=basename(dirname(dirname(file)))) %>% semi_join(corpus, by="package")
+files <- count(as_tibble(y), file)
+a <- anti_join(files, package_programs, by="file")
+a
+a$file %>% str_replace(fixed("../.."), params$base_dir) %>% dirname() %>% dir_exists()
+```

### 2020-08-27 inst/paper/main.tex

@@
 \begin{figure}[htbp]
+	\centering
 \includegraphics[width=\columnwidth]{ast_sizes}
@@
 argument \c{envir} to indicate in which environment its first argument,
-after being resolved, must be evaluated. By default, it is equal to \c{parent.frame()}, which is the parent environment of the \c{eval} call.
+after being resolved, must be evaluated.
+By default, it is equal to \c{parent.frame()}, which is the parent environment of the \c{eval} call.
+This default argument happens in \miss{XX\%} of the eval calls.
 
@@
 
-We classify the environment from two points of view:
+There are two main environments affected by \eval:
 \begin{itemize}
-	\item the \emph{default} assignment environment \c{eval} writes new bindings.\footnote{It is the default one but the expression in \eval can always write in any other environments using \c{assign} or in the parent environment using \c{<<-}.}
-	\item in which \emph{existing} environment \c{eval} reads. It is the first non new environment. \c{eval} will also be able to read in all parent environment of that environment. 
+	\item the \emph{existing} environment where \c{eval} reads.
+It is the first non new environment.
+\c{eval} will also be able to read in all parent environment of that environment.
+It can be the global environment, a package environment, an empty environment, the callee environment, or a parent environment of the callee.
+	\item the \emph{default} assignment environment \c{eval} writes new bindings.\footnote{It is the default one but the expression in \eval can always write in any other environments using \c{assign} or in the parent environment using \c{<<-}.} If it is different from the read environment and is a new one that inherits from the read environment, we add a symbol \texttt{+} after the read environment.
 \end{itemize}
 
+As shown in Figure~\ref{fig:environments}, \eval in R is often performed in the global environment.  It is also used a lot to evaluate in the context of a dataframe (\emph{list} environment).
+
 % Variability: how many different envir class, and different envir expr, for one call site?
+\begin{figure*}[htbp]
+	\centering
+	\includegraphics[width=\linewidth]{environments.pdf}
+	\caption{The environments in which the expression is evaluated. \c{NA} corresponds to the default environment.} \label{fig:environments}
+\end{figure*}
 
@@
 \begin{figure}[!h]
+	\centering
     \includegraphics[width=\columnwidth]{polymorphism}

### 2020-08-27 inst/analysis/corpus-stage2.Rmd
Fix forgotten kaggle stage2

@@
     package_programs,
-    transmute(package_programs_errors, error=paste(source, message, call), file), 
+    if (nrow(package_programs_errors) > 0) {
+      transmute(package_programs_errors, error=paste(source, message, call), file)
+    } else {
+      transmute(package_programs, error=NA, file)
+    }, 
     by="file"

### 2020-08-27 inst/analysis/analysis_stable.Rmd
Figures with legend
Environments table

@@
 
+save_table <- function(df, fname) {
+  df %>%
+    xtable::xtable(align="lXrrrr") %>%
+    print(
+      tabular.environment = "tabularx", 
+      width = "\\textwidth",
+      floating=F,
+      file=fname,
+      include.rownames=F
+    )
+}
+
 source("insights.R", local = knitr::knit_global())
@@
 theme_set(theme_minimal())
-theme_update(text = element_text(size = 20))
+#theme_update(text = element_text(size = 16))
 
@@
 ```{r envirall}
-classified_envir %>% ggplot() +
+classified_envir %>% 
+  ggplot() +
   geom_bar(aes(x =fct_rev(fct_infreq(env_class)), weight = nb_ev_calls )) +
   coord_flip() +
-  scale_y_log10() +
   labs(x = "Environments") + 
@@
   coord_flip() +
-  scale_y_log10() +
   labs(x = "Environments") + 
@@
 ggsave(path(PLOT_DIR, "environments-read.pdf"))
+```
+
+```{r}
+classified_envir_prop_core <- classified_envir %>% 
+  filter(dataset == "core") %>%
+  count(env_class, wt = nb_ev_calls, sort = TRUE) %>%
+  mutate(proportion = n / sum(n))
+
+classified_envir_prop_packages <- classified_envir %>% 
+  filter(dataset == "packages") %>%
+  count(env_class, wt = nb_ev_calls, sort = TRUE) %>%
+  mutate(proportion = n / sum(n))
+
+classified_envir_prop_core %>%
+  head(5) %>%
+  add_column(index = toupper(letters)[1:5]) %>%
+  pmap_dfr(function(env_class, n, proportion, index)
+  {
+    r(paste("core", "env class", index), env_class)
+    r(paste("core", "count", index), n)
+    r(paste("core", "proportion", index), percent(proportion))
+  })
+
+classified_envir_prop_packages %>%
+  head(5) %>%
+  add_column(index = toupper(letters)[1:5]) %>%
+  pmap_dfr(function(env_class, n, proportion, index)
+  {
+    r(paste("packages", "env class", index), env_class)
+    r(paste("packages", "count", index), n)
+    r(paste("packages", "proportion", index), percent(proportion))
+  })
 ```

### 2020-08-27 inst/analysis/enrich_dataset.Rmd
Environments table

--- 2020-08-27 inst/analysis/enrich_dataset.Rmd
@@
 ```{r setup, include=FALSE}
+knitr::knit_hooks$set(time_it = local({
+  now <- NULL
+  function(before, options) {
+    if (before) {
+      # record the current time before each chunk
+      now <<- Sys.time()
+    } else {
+      # calculate the time difference after a chunk
+      res <- difftime(Sys.time(), now)
+      # return a character string to show the time
+      paste("Time for this code chunk to run:", res, unit(res))
+    }
+  }
+}))
+
 knitr::opts_chunk$set(echo = TRUE,

### 2020-08-27 inst/analysis/insights.R

@@ -1,6 +1,8 @@
 library(rlang)
 
 
+
+
 nb_eval_call_sites <- function(eval_calls)
 {
   # Two cases: with srcref, without srcref
@@ -365,7 +367,9 @@ groupify_function <- function(expr_function)
 build_env_rep <- function(env_class)
 {
   splitted <- str_split(env_class, fixed("+"))[[1]]
-  return(paste0(splitted[[length(splitted)]], if(length(splitted) > 1) "+" else ""))
+  env_cl <- splitted[[length(splitted)]]
+  env_cl <- replace_na(str_match(env_cl, "caller-([:digit:]+)")[[2]], env_cl)
+  return(paste0(env_cl, if(length(splitted) > 1) "+" else ""))
 }
 
 simplify_envir <- function(env_class, envir_type, envir_expression)

### 2020-08-27 inst/paper/main.tex
Add events information

@@
 \input{tag/side-effects.tex}
+\input{tag/events.tex}
 \input{tag/analysisStable.tex}
@@
 \newcommand{\datatable}{\texttt{data.table}\xspace}
+\newcommand{\mlogit}{\texttt{mlogit}\xspace}
+\newcommand{\mboost}{\texttt{mboost}\xspace}
+\newcommand{\metafor}{\texttt{metafor}\xspace}
+\newcommand{\lavaan}{\texttt{lavaan}\xspace}
+\newcommand{\mclust}{\texttt{mclust}\xspace}
+\newcommand{\gamlss}{\texttt{gamlss}\xspace}
+
 \newcommand{\base}{\texttt{base}\xspace}
@@
 
-%%In the corpus, we observe \AllEventCountRnd events. Only \EvalEventCountRnd
-%%events occur inside eval. This is just \EvalEventAllPerc of all events.
-%%Figure~\ref{fig:event-distribution} shows the distribution of events for
-%%core and package \eval calls.
+We compare the number of events intercepted during execution of \eval across the
+whole corpus. For this, we use the number of interpreter's evaluator loops as a
+proxy for events. In the corpus, we observe \AllEventCountRnd events. Only
+\EvalEventCountRnd events occur inside eval. This is just \EvalEventAllPerc of
+all events. Figure~\ref{fig:event-distribution} shows the distribution of events
+caused by \eval calls from R Core and CRAN packages.
 
-{\bf Missing from this section is the measure of the number of events
-  happening in an eval}
+\begin{table}[ht]\resizebox{\columnwidth}{!}{%
+\begin{tabular}{lll}\hline
+Events & Core Eval Call \% & Package Eval Call \% \\ \hline
+\EventsMinRangeA--\EventsMaxRangeA & \EventsCoreEvalPercA & \EventsPackageEvalPercA \\
+\EventsMinRangeB--\EventsMaxRangeB & \EventsCoreEvalPercB & \EventsPackageEvalPercB \\
+\EventsMinRangeC--\EventsMaxRangeC & \EventsCoreEvalPercC & \EventsPackageEvalPercC \\
+\end{tabular}} \label{table:event-distribution}
+\caption{Distribution of events in \eval calls across Core and Package}
+\end{table}
+
+We observe a very wide spread in the number of events generated by \eval
+calls. The largest number of events generated by an \eval call is \EventsMaxCount.
+However majority of \eval calls perform up to \EventsMaxRangeA events.
+It is interesting to note that a higher proportion of Core R \eval calls perform
+side effects in the \EventsMinRangeB--\EventsMaxRangeB range.
+Only \EventsCoreEvalCountC Core R \eval calls and \EventsPackageEvalCountC
+package \eval call generate \EventsMinRangeC to \EventsMaxRangeC events.
+These \eval calls originate from statistical modeling packages such as \mlogit,
+\mboost, \metafor, \lavaan, \mclust and \gamlss.
 

### 2020-08-27 inst/analysis/analysis_stable.Rmd
Tags

@@
 
-eval_call_core <- 
+eval_calls_core <- 
   read_fst(core_filepath) %>%
@@
 
-eval_calls_raw <- bind_rows(eval_calls_packages, eval_call_core)
+eval_calls_raw <- bind_rows(eval_calls_packages, eval_calls_core)
 ```
@@
 
+nb_default_envir_core <- count(default_envir %>% filter(dataset == "core"), wt = nb_ev_calls)
+
+nb_eval_calls_core <- count(eval_calls_core, wt = nb_ev_calls)$n
+nb_eval_calls_packages <- count(eval_calls_packages, wt = nb_ev_calls)$n
+
+nb_default_envir_packages <- count(default_envir %>% filter(dataset == "packages"), wt = nb_ev_calls)
+
 r("default envir expr", nb_default_envir$n)
@@
 r("default envir  call sites percent", ratio(nb_call_sites_default_envir, nb_call_sites))
+
+r("default envir core expr percent", ratio(nb_default_envir_core$n, nb_eval_calls_core))
+r("default envir packages expr percent", ratio(nb_default_envir_packages$n, nb_eval_calls_packages))
 ```
@@
 
+nb_top_envir <- 8
+
 classified_envir_prop_core %>%
-  head(5) %>%
-  add_column(index = toupper(letters)[1:5]) %>%
+  head(nb_top_envir) %>%
+  add_column(index = toupper(letters)[1:nb_top_envir]) %>%
   pmap_dfr(function(env_class, n, proportion, index)
@@
 classified_envir_prop_packages %>%
-  head(5) %>%
-  add_column(index = toupper(letters)[1:5]) %>%
+  head(nb_top_envir) %>%
+  add_column(index = toupper(letters)[1:nb_top_envir]) %>%
   pmap_dfr(function(env_class, n, proportion, index)
@@
   })
+
+envir_enclos <- classified_envir %>% filter(env_class %in% c("NULL", "list"))
+nb_enclos <- count(envir_enclos, wt = nb_ev_calls)$n
+r("nb enclos envir", nb_enclos)
+r("enclos envir percent", ratio(nb_enclos, nb_eval_calls))
+
+envir_sys_call <- classified_envir %>% filter(str_detect(env_class, fixed("sys.class")))
+nb_sys_call <- count(envir_sys_call, wt = nb_ev_calls)$n
+r("nb sys call envir", nb_sys_call)
+r("sys call envir percent", ratio(nb_sys_call, nb_eval_calls))
 ```

### 2020-08-27 inst/analysis/events.Rmd

---
title: "Events"
output: html_document
params:
  base_dir: /home/aviral/projects/revalstudy2
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
for (x in c(
  "dplyr", 
  "DT",
  "fs",
  "fst",
  "ggplot2",
  "ggthemes",
  "lubridate",
  "readr",
  "purrr",
  "stringr",
  "tidyr",
  "tibble"
  )) {
  suppressPackageStartupMessages(library(x, character.only=TRUE))
}

knitr::opts_chunk$set(
  echo = TRUE,
  fig.align = "center",
  fig.retina = 2,
  fig.width = 10
)

# TODO: move into the package
source("inc/paths.R")
source("inc/setup.R")
source("inc/latextags.R")

options(repos="https://cran.r-project.org")
theme_set(theme_minimal())
create_tags(path(TAGS_DIR, "events.tex"), prefix="", default=TRUE)
```

```{r}
core_data <- read_fst(EVALS_SUM_CORE_FILE)
package_data <- read_fst(EVALS_SUM_PKGS_FILE)
kaggle_data <- read_fst(EVALS_SUM_KAGGLE_FILE)
program_data <- read_fst(PROGRAM_FILE)

all_data <- rbind(core_data, package_data, kaggle_data) %>% filter(eval_source != "testthat")
```

```{r preprocess-write-data}

eval_events_data <-
    all_data %>%
    filter(eval_function == "eval" | eval_function == "evalq") %>%
    select(eval_source_type, eval_source, direct_interpreter_eval, indirect_interpreter_eval, nb_ev_calls) %>%
    mutate(all_interpreter_eval = direct_interpreter_eval + indirect_interpreter_eval)

program_events_data <-
    program_data %>%
    select(direct_interpreter_eval, indirect_interpreter_eval)
```

```{r compute-side-effecting-write-counts}
all_events <- sum(program_events_data$direct_interpreter_eval + program_events_data$indirect_interpreter_eval)
eval_events <- sum(eval_events_data$direct_interpreter_eval * eval_events_data$nb_ev_calls)

r("All Event Count", all_events)
r("Eval Event Count", eval_events)

r("events max count", max(eval_events_data$all_interpreter_eval))

r("Eval Event All Perc", ratio(eval_events, all_events))
```

```{r compute-side-effecting-call-counts}
eval_events_data %>%
arrange(desc(all_interpreter_eval)) %>%
slice(1:500)

eval_events_by_source_type <-
    eval_events_data %>%
    group_by(eval_source_type, eval_source, all_interpreter_eval) %>%
    summarize(eval_count = sum(nb_ev_calls)) %>%
    ungroup()

datatable(eval_events_by_source_type)

eval_events_by_source_type %>%
arrange(desc(all_interpreter_eval)) %>%
datatable()

eval_events_by_source_type_summarized <-
    eval_events_by_source_type %>%
    mutate(all_interpreter_eval = if_else(all_interpreter_eval <= 9, as.character(all_interpreter_eval), "> 9")) %>%
    group_by(eval_source_type, all_interpreter_eval) %>%
    summarize(eval_count = sum(eval_count)) %>%
    ungroup()

datatable(eval_events_by_source_type_summarized)

eval_events_by_source_type_summarized_by_package <-
    eval_events_by_source_type %>%
    group_by(eval_source_type, eval_source) %>%
    summarize(total_interpreter_eval = sum(all_interpreter_eval * eval_count),
              max_interpreter_eval = max(all_interpreter_eval),
              average_interpreter_eval = total_interpreter_eval / sum(eval_count)) %>%
    ungroup()

datatable(eval_events_by_source_type_summarized_by_package)

p <- eval_events_by_source_type %>%
    ggplot(aes(x = eval_source_type, y = all_interpreter_eval, weight = eval_count)) +
    geom_boxplot() + 
    scale_y_log10()

p
```

```{r}
right <- c(100L, 10000L)
left <- right + 1L
right <- c(right, 2 * 10L^7L)
left <- c(0L, left)
package_counts <- integer(length(right))
core_counts <- integer(length(right))

incrementer <- function(eval_source_type, all_interpreter_eval, eval_count) {

    index <- 1
    while(all_interpreter_eval > right[index]) {
        index <- index + 1
    }
    if(eval_source_type == "core") {
        core_counts[index] <<- core_counts[index] + eval_count
    }
    else {
        package_counts[index] <<- package_counts[index] + eval_count
    }
}

ignore_result <-
    eval_events_by_source_type %>%
    group_by(eval_source_type, all_interpreter_eval) %>%
    summarize(eval_count = sum(eval_count)) %>%
    ungroup() %>%
    pmap(incrementer)

eval_events_table <-
    tibble(left = left, right = right, core_eval_count = core_counts, package_eval_count = package_counts) %>%
    mutate(core_eval_perc = round(100 * core_eval_count / sum(core_eval_count), 1),
           package_eval_perc = round(100 * package_eval_count / sum(package_eval_count), 1))

eval_events_table %>%
add_column(index = toupper(letters)[1:nrow(eval_events_table)]) %>%
pmap(function(left, right, core_eval_count, package_eval_count, core_eval_perc, package_eval_perc, index) {
    r(str_c("Events min range ", index), left)
    r(str_c("Events max range ", index), right)
    r(str_c("Events core eval count ", index), core_eval_count)
    r(str_c("Events core eval perc ", index), ratio(core_eval_perc, 100))
    r(str_c("Events package eval count ", index), package_eval_count)
    r(str_c("Events package eval perc ", index), ratio(package_eval_perc, 100))

})

datatable(eval_events_table)
```

mlogit, mboost, metafor, lavaan, foreach, gamlss, mclust, magrittr, VGAM

0-5
5-10
10-100
101-1000
1001-10000
10001-1000000

### 2020-08-27 inst/paper/main.tex
More env examples

--- 2020-08-27 inst/paper/main.tex
@@
 By default, it is equal to \c{parent.frame()}, which is the parent environment of the \c{eval} call.
-This default argument happens in \miss{XX\%} of the eval calls.
+This default argument happens in \DefaultEnvirExprPercent of the eval calls.
 
@@
 \c{eval} will also be able to read in all parent environment of that environment.
-It can be the global environment, a package environment, an empty environment, the callee environment, or a parent environment of the callee.
+It can be the global environment, a package environment, an empty environment, the callee environment, or a parent environment of the callee: $1$ is the parent, $2$ the grand-parent, and so on.
 	\item the \emph{default} assignment environment \c{eval} writes new bindings.\footnote{It is the default one but the expression in \eval can always write in any other environments using \c{assign} or in the parent environment using \c{<<-}.} If it is different from the read environment and is a new one that inherits from the read environment, we add a symbol \texttt{+} after the read environment.
@@
 
-As shown in Figure~\ref{fig:environments}, \eval in R is often performed in the global environment.  It is also used a lot to evaluate in the context of a dataframe (\emph{list} environment).
+To give some examples,  global+ refers to a new scope that inherits from the global scope, while 3 refers to the great grand parent of the \eval scope.
+
+Table~\ref{tab:environments} shows the environments when they are \emph{explicitly specified}.
+\eval in R is often performed in a new environment inherited from the global environment for packages, while for core, it is mostly done in the environment of the grand-parent.
+It is also used a lot to evaluate in the context of a dataframe (\emph{list} environment).
+\emph{loop} \textbf{To explain...?
+But we do not see it here anyway}
+
+The \c{sys.call} option is not used a lot, only \SysCallEnvirPercent as a whole, and always with \c{sys.parent()} as argument.
 
 % Variability: how many different envir class, and different envir expr, for one call site?
-\begin{figure*}[htbp]
+%\begin{figure*}[htbp]
+%	\centering
+%	\includegraphics[width=\linewidth]{environments.pdf}
+%	\caption{The environments in which the expression is evaluated. \c{NA} corresponds to the default environment.} \label{fig:environments}
+%\end{figure*}
+
+\begin{table}[htbp]
     \centering
-	\includegraphics[width=\linewidth]{environments.pdf}
-	\caption{The environments in which the expression is evaluated. \c{NA} corresponds to the default environment.} \label{fig:environments}
-\end{figure*}
+    \begin{tabular}{|c|c||c|c|}
+        \hline
+        \multicolumn{2}{|c|}{Core} &  \multicolumn{2}{|c|}{Packages}  \\
+        \hline
+        Environment & Proportion & Environment & Proportion \\
+        \hline
+        \CoreEnvClassA &  \CoreProportionA & \PackagesEnvClassA &  \PackagesProportionA\\
+        \hline
+        \CoreEnvClassB &  \CoreProportionB & \PackagesEnvClassB &  \PackagesProportionB \\
+        \hline
+        \CoreEnvClassC & \CoreProportionC  & \PackagesEnvClassC &  \PackagesProportionC \\
+        \hline
+       \CoreEnvClassD  &  \CoreProportionD & \PackagesEnvClassD &  \PackagesProportionD \\
+        \hline
+        \CoreEnvClassE  & \CoreProportionE & \PackagesEnvClassE & \PackagesProportionE \\
+        \hline
+    \end{tabular} \label{tab:environments}
+\caption{Environments in terms of eval calls.}
+\end{table}
+
+\c{match.arg} (see Listing~\ref{code:matchargenvir}) is the function that contributes the most to the eval calls (\MatchArgPercent) and has environment class 2.
+\c{sys.parent()} refers to the parent environment of \c{match.arg} and so to the grand-parent environment of \c{eval}, \ie what we denote 2.
+
+\begin{lstlisting}[label=code:matchargenvir, caption={\c{match.arg} using \eval with a custom environment.}]
+formal.args <- formals(sys.function(sysP <- sys.parent()))
+choices <- eval(formal.args[[as.character(substitute(arg))]], envir = sys.frame(sysP))
+\end{lstlisting}
+
+In package \emph{gamlss}, function \c{get.K} evaluates its expression in its parent environment, the one of the body of \c{get.K} returned from \c{environment()}.
+
+\begin{lstlisting}
+eval(parse(text = paste(\"fam$dld\", ii, sep = \"\")), envir = environment())
+\end{lstlisting}
+
+The \emph{glue} library is used to interpolate string litterals with variables, such as in \c{glue('My name is {name}.')} where \c{name} will be subtituted by its value.
+\emph{glue} makes it possible to define transformers that process those interpolated variables.
+They are built upon the \c{identity_transformer}, which uses \eval.
+The interpretation environment for \eval is often global here, as the string litterals are often used to print variabels at the top level.
+
+\begin{lstlisting}
+identity_transformer <- function(text, envir) {
+    eval(parse(text = text, keep.source = FALSE), envir)
+}
+\end{lstlisting}
+
+The \c{imchange} function of package \c{imager} makes it possible to modify images using a dedicated formula syntax using \c{~}.\footnote{Inspired by \c{map} in package \emph{purr}.}  Here, \eval is evaluated in \c{newenv}, which creates a new environment that inherits from \c{parent.frame()} by default.
+We classify it as 1+.
+
+\begin{lstlisting}
+newenv <- new.env()
+...
+fo <- parse(text=as.character(fo)[2])
+im[where] <- eval(fo,envir=newenv,enclos=env)
+\end{lstlisting}
 

### 2020-08-27 inst/paper/main.tex

@@
 
-%%In the corpus, we observe \AllEventCountRnd events. Only \EvalEventCountRnd
-%%events occur inside eval. This is just \EvalEventAllPerc of all events.
-%%Figure~\ref{fig:event-distribution} shows the distribution of events for
-%%core and package \eval calls.
+We compare the number of events intercepted during execution of \eval across the
+whole corpus. For this, we use the number of interpreter's evaluator loops as a
+proxy for events. In the corpus, we observe \AllEventCountRnd events. Only
+\EvalEventCountRnd events occur inside eval. This is just \EvalEventAllPerc of
+all events. Figure~\ref{fig:event-distribution} shows the distribution of events
+caused by \eval calls from R Core and CRAN packages.
 
-{\bf Missing from this section is the measure of the number of events
-  happening in an eval}
+\begin{table}[ht]\resizebox{\columnwidth}{!}{%
+\begin{tabular}{lll}\hline
+Events & Core Eval Call \% & Package Eval Call \% \\ \hline
+\EventsMinRangeA--\EventsMaxRangeA & \EventsCoreEvalPercA & \EventsPackageEvalPercA \\
+\EventsMinRangeB--\EventsMaxRangeB & \EventsCoreEvalPercB & \EventsPackageEvalPercB \\
+\EventsMinRangeC--\EventsMaxRangeC & \EventsCoreEvalPercC & \EventsPackageEvalPercC \\
+\end{tabular}} \label{table:event-distribution}
+\caption{Distribution of events in \eval calls across Core and Package}
+\end{table}
+
+We observe a very wide spread in the number of events generated by \eval
+calls. The largest number of events generated by an \eval call is \EventsMaxCount.
+However majority of \eval calls perform up to \EventsMaxRangeA events suggesting
+that most expressions passed to \eval are small.
+It is interesting to note that a higher proportion of Core R \eval calls perform
+side effects in the \EventsMinRangeB--\EventsMaxRangeB range.
+Only \EventsCoreEvalCountC Core R \eval calls and \EventsPackageEvalCountC
+package \eval call generate \EventsMinRangeC to \EventsMaxRangeC events.
+These \eval calls originate from statistical modeling packages such as \mlogit,
+\mboost, \metafor, \lavaan, \mclust and \gamlss.
 

### 2020-08-27 inst/paper/main.tex
reorder

@@
 }
+
+% https://tex.stackexchange.com/questions/351435/table-name-placment-in-tabular-documentclass-ieeetran-conference
+\makeatletter
+\def\@IEEEtablestring{figure}
+\makeatother
+
 \begin{document}
@@
 \newcommand{\cf}{\emph{cf.}\xspace}
-\newcommand{\summary}[1]{({\csname #1Min\endcsname} / {\csname #1Mean\endcsname} / {\csname #1Max\endcsname})}
-\newcommand{\summaryrnd}[1]{({\csname #1MinRnd\endcsname} / {\csname #1MeanRnd\endcsname} / {\csname #1MaxRnd\endcsname})}
+\newcommand{\summary}[1]{{\csname #1\endcsname} ({\csname #1Min\endcsname} / {\csname #1Mean\endcsname} / {\csname #1Max\endcsname})}
+\newcommand{\summaryrnd}[1]{{\csname #1Rnd\endcsname} ({\csname #1MinRnd\endcsname} / {\csname #1MeanRnd\endcsname} / {\csname #1MaxRnd\endcsname})}
 
@@
 choose popular packages which in general should be of a better quality and have
-better code coverage and therefore more code to run. The resulting set has
-\summary{CorpusRevdeps}\footnote{Such triples represent ($min$ / $mean$ /
-  $max$)} reverse dependencies and consists of \CorpusRCodeRnd lines of R code
-and \CorpusNativeCodeRnd lines of native code (C/C++/Fortran)\footnote{Excluding
-  comments and blank lines using \url{github.com/AlDanial/cloc}}.
-Figure~\ref{fig:corpus} shows these packages, the size of the dots reflects the
-project's size in lines of code including both R and native code, the x-axis
-indicates the expression code coverage in percents and the y-axis gives the
-number of reverse dependencies in log scale. Dotted lines indicate means.
-Packages with over \CorpusEvalsPackageTreshold eval call sites are annotated.
+better code coverage and therefore more code to run. The resulting set contains
+\CorpusRCodeRnd lines of R code and \CorpusNativeCodeRnd lines of native code
+(C/C++/Fortran)\footnote{Excluding comments and blank lines using
+  \url{github.com/AlDanial/cloc}}. Figure~\ref{fig:corpus} shows these packages,
+the size of the dots reflects the project's size in lines of code including both
+R and native code, the x-axis indicates the expression code coverage in percents
+and the y-axis gives the number of reverse dependencies in log scale. Dotted
+lines indicate means. Packages with over \CorpusEvalsPackageTreshold eval call
+sites are annotated.
 
@@
 \begin{compactitem}[$-$]
-  \item \CorpusExamplesProgramsRnd examples with \CorpusExamplesCodeRnd
-  \summaryrnd{CorpusExamplesCode} lines of code,
-  \item \CorpusTestsProgramsRnd tests with \CorpusTestsCodeRnd
-  \summaryrnd{CorpusTestsCode} lines of code, and
-  \item \CorpusVignettesProgramsRnd vignettes with \CorpusVignettesCodeRnd
+  \item \CorpusExamplesProgramsRnd examples with
+  \summaryrnd{CorpusExamplesCode}\footnote{This notation represents $sum$ ($min$
+    / $mean$ / $max$)} lines of code,
+  \item \CorpusVignettesProgramsRnd vignettes with
   \summaryrnd{CorpusVignettesCode} lines of code.
@@
 The final set contains \CorpusFinishedKaggle kernels with
-\CorpusFinishedKaggleCodeRnd \summaryrnd{CorpusFinishedKaggleCode} lines of R
-code.
+\summaryrnd{CorpusFinishedKaggleCode} lines of R code.
 
@@
 
-\begin{table}[ht]
-\resizebox{\columnwidth}{!}{%
-\begin{tabular}{@{}r|ccc@{}} \hline
-Source & Core & Dependent & Own \\\hline
-      %      CRAN & \CranCoreTotalEvalCallsRnd (\CranCoreMinEvalCallsRnd/\CranCoreAvgEvalCallsRnd/\CranCoreMaxEvalCallsRnd) & \CranDependentTotalEvalCallsRnd (\CranDependentMinEvalCallsRnd/\CranDependentAvgEvalCallsRnd/\CranDependentMaxEvalCallsRnd) & \CranOwnTotalEvalCallsRnd (\CranOwnMinEvalCallsRnd/\CranOwnAvgEvalCallsRnd/\CranOwnMaxEvalCallsRnd) \\
-CRAN & \CranCoreTotalEvalCallsRnd (1 / 581 / 416K)  & \CranDependentTotalEvalCallsRnd (1 / 532 / 183K) & \CranOwnTotalEvalCallsRnd (1 / 292 / 144K)\\
-Kaggle & \KaggleCoreTotalEvalCallsRnd (1 / 412 / 2.8K) & \KaggleDependentTotalEvalCallsRnd (3 / 96 / 1.1K) & 0 (0 / 0 / 0) \\
-%      Kaggle & \KaggleCoreTotalEvalCallsRnd (\KaggleCoreMinEvalCallsRnd/\KaggleCoreAvgEvalCallsRnd/\KaggleCoreMaxEvalCallsRnd) & \KaggleDependentTotalEvalCallsRnd (\KaggleDependentMinEvalCallsRnd/\KaggleDependentAvgEvalCallsRnd/\KaggleDependentMaxEvalCallsRnd) & 0(0/0/0) \\
-      \hline
+ \begin{table*}[tb]
+   \centering
+   \begin{tabular}{r|c|c|c@{}} \hline
+     Corpus / Source & Core                              & Dependent                              & Own \\\hline
+     CRAN   & \summaryrnd{CranCoreEvalCalls}    & \summaryrnd{CranDependentEvalCalls}    & \summaryrnd{CranOwnEvalCalls} \\
+     Kaggle & \summaryrnd{KaggleCoreEvalCalls}  & \summaryrnd{KaggleDependentEvalCalls}  & \summaryrnd{KaggleOwnEvalCalls} \\\hline
    \end{tabular}
-  }
    \caption{Eval calls in CRAN and Kaggle}
    \label{A}
-\end{table}
+ \end{table*}
 
@@
 \begin{tabular}{rr|rr|rr}\hline
-Callsites & Packages & Callsites & Packages & Callsites & Packages \\ \hline
-\SiteSummarySiteCountA & \SiteSummaryPackageCountA & \SiteSummarySiteCountF & \SiteSummaryPackageCountF & \SiteSummarySiteCountK & \SiteSummaryPackageCountK \\
-\SiteSummarySiteCountB & \SiteSummaryPackageCountB & \SiteSummarySiteCountG & \SiteSummaryPackageCountG & \SiteSummarySiteCountL & \SiteSummaryPackageCountL \\
-\SiteSummarySiteCountC & \SiteSummaryPackageCountC & \SiteSummarySiteCountH & \SiteSummaryPackageCountH & \SiteSummarySiteCountM & \SiteSummaryPackageCountM \\
-\SiteSummarySiteCountD & \SiteSummaryPackageCountD & \SiteSummarySiteCountI & \SiteSummaryPackageCountI & \SiteSummarySiteCountN & \SiteSummaryPackageCountN \\
-      \SiteSummarySiteCountE & \SiteSummaryPackageCountE & \SiteSummarySiteCountJ & \SiteSummaryPackageCountJ & \SiteSummarySiteCountO & \SiteSummaryPackageCountO \\    \hline
-\end{tabular}} \label{table:site-package-summary}
+ Callsites & Packages & Callsites & Packages & Callsites & Packages \\\hline
+\input{tag/table-site-package-summary.tex}
+\end{tabular}}
 \caption{Distribution of callsites in CRAN}
+\label{table:site-package-summary}
 \end{table}
@@
 
-\section{Side-Effects}
+\section{A taxonomy of eval}
+
+We first classify \c{eval} depending on the resolved expression, \emph{i.e.}
+the first argument of \c{eval} after all possible function calls and symbol
+resolution in it have been executed.
+
+\subsection{Operation Mix}
+
+Unlike Javascript, R packages also call native code for access to efficient
+numerical and statistical routines. We intercept all calls to R functions and
+native functions from the programs and \eval calls. In the whole corpus there
+are \AllClosureCallCount calls to R functions and \AllNativeCallCount to native
+code. Of those, \eval contribute to only \EvalClosureCallCount calls to R
+functions and \EvalNativeCallCount to native code.
+
+\subsection{Side-Effects}
 
@@
 packages are side-effecting.
-
-\section{A taxonomy of eval}
-
-We first classify \c{eval} depending on the resolved expression, \emph{i.e.}
-the first argument of \c{eval} after all possible function calls and symbol
-resolution in it have been executed.
-
-\subsection{Operation mix}
-
-Here we should show some information about what happens during an eval. What kinds of operations...
 

### 2020-08-27 inst/paper/main.tex
Table 1

@@
 
-\section{A taxonomy of eval}
-
-We first classify \c{eval} depending on the resolved expression, \emph{i.e.}
-the first argument of \c{eval} after all possible function calls and symbol
-resolution in it have been executed.
-
-\subsection{Operation Mix}
-
-Unlike Javascript, R packages also call native code for access to efficient
-numerical and statistical routines. We intercept all calls to R functions and
-native functions from the programs and \eval calls. In the whole corpus there
-are \AllClosureCallCount calls to R functions and \AllNativeCallCount to native
-code. Of those, \eval contribute to only \EvalClosureCallCount calls to R
-functions and \EvalNativeCallCount to native code.
-
-\subsection{Side-Effects}
+\section{Side-Effects}
 
@@
 packages are side-effecting.
+
+\section{A taxonomy of eval}
+
+We first classify \c{eval} depending on the resolved expression, \emph{i.e.}
+the first argument of \c{eval} after all possible function calls and symbol
+resolution in it have been executed.
+
+\subsection{Operation mix}
+
+Here we should show some information about what happens during an eval. What kinds of operations...
 

### 2020-08-27 inst/paper/main.tex
Fix escaping

--- 2020-08-27 inst/paper/main.tex
@@
 
-The \c{imchange} function of package \c{imager} makes it possible to modify images using a dedicated formula syntax using \c{~}.\footnote{Inspired by \c{map} in package \emph{purr}.}  Here, \eval is evaluated in \c{newenv}, which creates a new environment that inherits from \c{parent.frame()} by default.
-We classify it as 1+.
+The \c{imchange} function of package \c{imager} makes it possible to modify images using a dedicated formula syntax using \c{\~}.\footnote{Inspired by \c{map} in package \emph{purr}.}
+Here, \eval is evaluated in \c{newenv}, which creates a new environment that inherits from \c{parent.frame()} by default. We classify it as 1+.
 

### 2020-08-27 inst/paper/main.tex
Comment explanation

@@
 
-Unlike Javascript, R packages also call native code for access to efficient
-numerical and statistical routines. We intercept all calls to R functions and
-native functions from the programs and \eval calls. In the whole corpus there
-are \AllClosureCallCount calls to R functions and \AllNativeCallCount to native
-code. Of those, \eval contribute to only \EvalClosureCallCount calls to R
-functions and \EvalNativeCallCount to native code.
+%Unlike Javascript, R packages also call native code for access to efficient
+%numerical and statistical routines. We intercept all calls to R functions and
+%native functions from the programs and \eval calls. In the whole corpus there
+%are \AllClosureCallCount calls to R functions and \AllNativeCallCount to native
+%code. Of those, \eval contribute to only \EvalClosureCallCount calls to R
+%functions and \EvalNativeCallCount to native code.
 

### 2020-08-27 inst/paper/main.tex
Even more examples of environments.

--- 2020-08-27 inst/paper/main.tex
@@
 } 
-
-% https://tex.stackexchange.com/questions/351435/table-name-placment-in-tabular-documentclass-ieeetran-conference
-\makeatletter
-\def\@IEEEtablestring{figure}
-\makeatother
-
 \begin{document}
@@
 \newcommand{\cf}{\emph{cf.}\xspace}
-\newcommand{\summary}[1]{{\csname #1\endcsname} ({\csname #1Min\endcsname} / {\csname #1Mean\endcsname} / {\csname #1Max\endcsname})}
-\newcommand{\summaryrnd}[1]{{\csname #1Rnd\endcsname} ({\csname #1MinRnd\endcsname} / {\csname #1MeanRnd\endcsname} / {\csname #1MaxRnd\endcsname})}
+\newcommand{\summary}[1]{({\csname #1Min\endcsname} / {\csname #1Mean\endcsname} / {\csname #1Max\endcsname})}
+\newcommand{\summaryrnd}[1]{({\csname #1MinRnd\endcsname} / {\csname #1MeanRnd\endcsname} / {\csname #1MaxRnd\endcsname})}
 
@@
 choose popular packages which in general should be of a better quality and have
-better code coverage and therefore more code to run. The resulting set contains
-\CorpusRCodeRnd lines of R code and \CorpusNativeCodeRnd lines of native code
-(C/C++/Fortran)\footnote{Excluding comments and blank lines using
-  \url{github.com/AlDanial/cloc}}. Figure~\ref{fig:corpus} shows these packages,
-the size of the dots reflects the project's size in lines of code including both
-R and native code, the x-axis indicates the expression code coverage in percents
-and the y-axis gives the number of reverse dependencies in log scale. Dotted
-lines indicate means. Packages with over \CorpusEvalsPackageTreshold eval call
-sites are annotated.
+better code coverage and therefore more code to run. The resulting set has
+\summary{CorpusRevdeps}\footnote{Such triples represent ($min$ / $mean$ /
+  $max$)} reverse dependencies and consists of \CorpusRCodeRnd lines of R code
+and \CorpusNativeCodeRnd lines of native code (C/C++/Fortran)\footnote{Excluding
+  comments and blank lines using \url{github.com/AlDanial/cloc}}.
+Figure~\ref{fig:corpus} shows these packages, the size of the dots reflects the
+project's size in lines of code including both R and native code, the x-axis
+indicates the expression code coverage in percents and the y-axis gives the
+number of reverse dependencies in log scale. Dotted lines indicate means.
+Packages with over \CorpusEvalsPackageTreshold eval call sites are annotated.
 
@@
 \begin{compactitem}[$-$]
-  \item \CorpusExamplesProgramsRnd examples with
-  \summaryrnd{CorpusExamplesCode}\footnote{This notation represents $sum$ ($min$
-    / $mean$ / $max$)} lines of code,
-  \item \CorpusVignettesProgramsRnd vignettes with
+  \item \CorpusExamplesProgramsRnd examples with \CorpusExamplesCodeRnd
+  \summaryrnd{CorpusExamplesCode} lines of code,
+  \item \CorpusTestsProgramsRnd tests with \CorpusTestsCodeRnd
+  \summaryrnd{CorpusTestsCode} lines of code, and
+  \item \CorpusVignettesProgramsRnd vignettes with \CorpusVignettesCodeRnd
   \summaryrnd{CorpusVignettesCode} lines of code.
@@
 The final set contains \CorpusFinishedKaggle kernels with
-\summaryrnd{CorpusFinishedKaggleCode} lines of R code.
+\CorpusFinishedKaggleCodeRnd \summaryrnd{CorpusFinishedKaggleCode} lines of R
+code.
 
@@
 
- \begin{table*}[tb]
-   \centering
-   \begin{tabular}{r|c|c|c@{}} \hline
-     Corpus / Source & Core                              & Dependent                              & Own \\\hline
-     CRAN   & \summaryrnd{CranCoreEvalCalls}    & \summaryrnd{CranDependentEvalCalls}    & \summaryrnd{CranOwnEvalCalls} \\
-     Kaggle & \summaryrnd{KaggleCoreEvalCalls}  & \summaryrnd{KaggleDependentEvalCalls}  & \summaryrnd{KaggleOwnEvalCalls} \\\hline
+\begin{table}[ht]
+\resizebox{\columnwidth}{!}{%
+\begin{tabular}{@{}r|ccc@{}} \hline
+Source & Core & Dependent & Own \\\hline
+      %      CRAN & \CranCoreTotalEvalCallsRnd (\CranCoreMinEvalCallsRnd/\CranCoreAvgEvalCallsRnd/\CranCoreMaxEvalCallsRnd) & \CranDependentTotalEvalCallsRnd (\CranDependentMinEvalCallsRnd/\CranDependentAvgEvalCallsRnd/\CranDependentMaxEvalCallsRnd) & \CranOwnTotalEvalCallsRnd (\CranOwnMinEvalCallsRnd/\CranOwnAvgEvalCallsRnd/\CranOwnMaxEvalCallsRnd) \\
+CRAN & \CranCoreTotalEvalCallsRnd (1 / 581 / 416K)  & \CranDependentTotalEvalCallsRnd (1 / 532 / 183K) & \CranOwnTotalEvalCallsRnd (1 / 292 / 144K)\\
+Kaggle & \KaggleCoreTotalEvalCallsRnd (1 / 412 / 2.8K) & \KaggleDependentTotalEvalCallsRnd (3 / 96 / 1.1K) & 0 (0 / 0 / 0) \\
+%      Kaggle & \KaggleCoreTotalEvalCallsRnd (\KaggleCoreMinEvalCallsRnd/\KaggleCoreAvgEvalCallsRnd/\KaggleCoreMaxEvalCallsRnd) & \KaggleDependentTotalEvalCallsRnd (\KaggleDependentMinEvalCallsRnd/\KaggleDependentAvgEvalCallsRnd/\KaggleDependentMaxEvalCallsRnd) & 0(0/0/0) \\
+      \hline
     \end{tabular}
+  }
   \caption{Eval calls in CRAN and Kaggle}
   \label{A}
- \end{table*}
+\end{table}
 
@@
 \begin{tabular}{rr|rr|rr}\hline
- Callsites & Packages & Callsites & Packages & Callsites & Packages \\\hline
-\input{tag/table-site-package-summary.tex}
-\end{tabular}}
+Callsites & Packages & Callsites & Packages & Callsites & Packages \\ \hline
+\SiteSummarySiteCountA & \SiteSummaryPackageCountA & \SiteSummarySiteCountF & \SiteSummaryPackageCountF & \SiteSummarySiteCountK & \SiteSummaryPackageCountK \\
+\SiteSummarySiteCountB & \SiteSummaryPackageCountB & \SiteSummarySiteCountG & \SiteSummaryPackageCountG & \SiteSummarySiteCountL & \SiteSummaryPackageCountL \\
+\SiteSummarySiteCountC & \SiteSummaryPackageCountC & \SiteSummarySiteCountH & \SiteSummaryPackageCountH & \SiteSummarySiteCountM & \SiteSummaryPackageCountM \\
+\SiteSummarySiteCountD & \SiteSummaryPackageCountD & \SiteSummarySiteCountI & \SiteSummaryPackageCountI & \SiteSummarySiteCountN & \SiteSummaryPackageCountN \\
+      \SiteSummarySiteCountE & \SiteSummaryPackageCountE & \SiteSummarySiteCountJ & \SiteSummaryPackageCountJ & \SiteSummarySiteCountO & \SiteSummaryPackageCountO \\    \hline
+\end{tabular}} \label{table:site-package-summary}
 \caption{Distribution of callsites in CRAN}
-\label{table:site-package-summary}
 \end{table}
@@
 
-The \c{imchange} function of package \c{imager} makes it possible to modify images using a dedicated formula syntax using \c{~}.\footnote{Inspired by \c{map} in package \emph{purr}.}  Here, \eval is evaluated in \c{newenv}, which creates a new environment that inherits from \c{parent.frame()} by default.
+The \c{imchange} function of package \c{imager} makes it possible to modify images using a dedicated formula syntax using \c{\~}.\footnote{Inspired by \c{map} in package \emph{purr}.}  Here, \eval is evaluated in \c{newenv}, which creates a new environment that inherits from \c{parent.frame()} by default.
 We classify it as 1+.
@@
 im[where] <- eval(fo,envir=newenv,enclos=env)
+\end{lstlisting}
+
+We finally look at the case where \c{envir} is not an environment, but a list, such as this \c{h2} function, defined inside function \c{adjCoef} in package \emph{actuar}.
+
+\begin{lstlisting}
+h2 <- function(x, y)
+    eval(parse(text = fcall),
+    envir = list(x = x, y = y),
+    enclos = parent.frame(2))
 \end{lstlisting}

### 2020-08-27 inst/analysis/counts.Rmd

@@
 params:
-  base_dir: /var/lib/R/project-evalR
+  base_dir: ../../../
 editor_options: 
@@
 
+```{r load programs}
+cran_programs <- read_fst(CRAN_PROGRAMS_FILE) %>% as_tibble() %>% filter(success)
+kaggle_programs <- read_fst(KAGGLE_PROGRAMS_FILE) %>% as_tibble() %>% filter(success)
+programs <- bind_rows(
+  select(cran_programs, file, corpus),
+  select(kaggle_programs, file, corpus)
+)
+```
+
 ```{r eval=TRUE}
-run_trace_data <- read_fst(PACKAGE_TRACE_RUNS_FILE)
-parallel_kaggle_data <- read_fst(KAGGLE_LOG_FILE)
+core_data <- 
+  read_fst(EVALS_SUM_CORE_FILE) %>%
+  as_tibble() %>%
+  semi_join(programs, by="file")
 
-core_data <- read_fst(EVALS_SUM_CORE_FILE)
 packages_data <- 
   read_fst(EVALS_SUM_PKGS_FILE) %>%
-    mutate(eval_source_type = "package")
+  as_tibble() %>%
+  semi_join(programs, by="file")
 
@@
 kaggle_data <- 
-    read_fst(EVALS_SUM_KAGGLE_FILE)
-    #mutate(eval_source_type == "core")
+  read_fst(EVALS_SUM_KAGGLE_FILE) %>%
+  as_tibble() %>%
+  semi_join(programs, by="file")
 
-all_data <- bind_rows(core_data, packages_data, kaggle_data) %>% as_tibble()
+all_data <- 
+  bind_rows(core_data, packages_data, kaggle_data)
+
 all_data <-
@@
 
+# add the programs that have no evals
+run_eval_proportion <-
+  bind_rows(
+    run_eval_proportion,
+    anti_join(programs, run_eval_proportion, by="file") %>% mutate(core=0, package=0, all=0)
+  )
+
 total_file_count <- length(unique(run_eval_proportion$file))
+stopifnot(total_file_count == nrow(programs))
+# anti_join(run_eval_proportion, programs, by="file")
 
@@
 
-r("top ten package call count", top_ten_package_call_count)
-r("top ten package call perc", ratio(top_ten_package_call_count, total_package_call_count))
+overview_table(
+  r("top ten package call count", top_ten_package_call_count),
+  r("top ten package call perc", ratio(top_ten_package_call_count, total_package_call_count)),
 
-r("top ten package site count", top_ten_package_site_count)
+  r("top ten package site count", top_ten_package_site_count),
   r("top ten package site perc", ratio(top_ten_package_site_count, sum(package_calls$site_count)))
+)
 
@@
   package_calls %>%
-    mutate(site_count = if_else(site_count <= 10, as.character(site_count), 
-                        if_else(site_count <= 50, "11--50",
-                        if_else(site_count <= 100, "51--100",
-                        if_else(site_count <= 150, "101--150",
-                        if_else(site_count <= 200, "151--200",
-                        if_else(site_count <= 250, "201--250",
-                        if_else(site_count <= 300, "251--300",
-                        "> 300")))))))) %>%
+  mutate(
+    site_count=cut(
+      site_count, 
+      breaks=c(0:10, 50, 100, 150, 200, 250, 300, Inf), 
+      labels=c(1:10, "11--50", "51--100", "101--150", "151--200", "201--250", "251--300", "> 300"),
+      include.lowest = T
+    )
+  ) %>%
   group_by(site_count) %>%
-    summarize(package_count = n()) %>%
-    ungroup() %>%
-    arrange(match(site_count, c(1:10,"11--50", "51--100", "101--150", "151--200", "201--250", "251--300", "> 300")))
+  summarize(package_count = n())
 
-datatable(eval_callsite_table)
+eval_callsite_table_row <- ceiling(nrow(eval_callsite_table)/3)
 
-eval_callsite_table %>%
-add_column(index = toupper(letters)[1:nrow(eval_callsite_table)]) %>%
-pmap_dfr(function(site_count, package_count, index) {
-    r(str_c("site summary site count ", index), site_count)
-    r(str_c("site summary package count ", index), package_count)
+eval_callsite_table3 <- map_dfc(seq(1, nrow(eval_callsite_table), eval_callsite_table_row), function(i) {
+  x <- tibble(
+    eval_callsite_table[i:(i+eval_callsite_table_row), "site_count"],
+    eval_callsite_table[i:(i+eval_callsite_table_row), "package_count"]
+  )
+  colnames(x) <- paste0(colnames(x), i)
+  x
 })
+rownames(eval_callsite_table3) <- NULL
+
+eval_callsite_table3 %>%
+  xtable() %>%
+  print(
+    only.contents=TRUE, 
+    include.colnames=FALSE, 
+    include.rownames=FALSE,
+    hline.after=nrow(eval_callsite_table3),
+    file=path(TAGS_DIR, "table-site-package-summary.tex")
+  )
+
+datatable(eval_callsite_table3)
 ```
@@
 ```{r core-dependent-own-eval}
-all_data <- bind_rows(core_data, packages_data, kaggle_data) %>%
-  as_tibble() %>%
+cod_data <- 
+  all_data %>%
   mutate(
@@
   )
-
-# TODO - we need to summarize the data better
-# all paths must start with this
-stopifnot(nrow(filter(all_data, !startsWith(file, "../../run/"))) == 0)
 ```
@@
 ```{r}
-cran_programs <- read_fst(CRAN_PROGRAMS_FILE) %>% 
-  as_tibble() %>%
-  filter(success) %>%
-  select(file, corpus)
-
-kaggle_programs <- read_fst(KAGGLE_PROGRAMS_FILE) %>% 
-  as_tibble() %>%
-  filter(success) %>%
-  select(file, corpus)
-
-all_programs <- bind_rows(cran_programs, kaggle_programs)
-```
-
-How many programs are missing in the summarized data
-
-```{r}
-anti_join(all_programs, all_data, by="file")
-```
-
-```{r}
-all_programs_cod <-
+programs_cod <-
   map_dfr(
@@
     ~mutate(
-      all_programs,
+      programs,
       core_own_dependent=.,
@@
 
-all_data_summary <- 
-  all_data %>%
+cod_data_summary <- 
+  cod_data %>%
   select(file, core_own_dependent, corpus, nb_ev_calls) %>%
-  right_join(all_programs_cod, by=c("file", "corpus", "core_own_dependent")) %>%
+  right_join(programs_cod, by=c("file", "corpus", "core_own_dependent")) %>%
   mutate(nb_ev_calls=ifelse(is.na(nb_ev_calls.x), nb_ev_calls.y, nb_ev_calls.x)) %>%
@@
 core_own_dependent <-
-  all_data_summary %>%
+  cod_data_summary %>%
   group_by(file, core_own_dependent) %>%
@@
 
-# %>%
-#   group_by(corpus, core_own_dependent) %>%
-#   summarize(
-#     total_eval_calls = sum(eval_calls),
-#     max_eval_calls = max(eval_calls),
-#     min_eval_calls = min(eval_calls),
-#     avg_eval_calls = mean(eval_calls)
-#   )
-
 vars <- expand.grid(corpus=c("cran", "kaggle"), cod=c("core", "own", "dependent"), stringsAsFactors = FALSE)
 map2_dfr(vars$corpus, vars$cod, function(crps, cod) {
-  print(filter(core_own_dependent, corpus==crps, core_own_dependent==cod))
   rs(paste(crps, cod, "eval calls"), filter(core_own_dependent, corpus==crps, core_own_dependent==cod)$eval_calls)

### 2020-08-27 inst/paper/main.tex
Add operation mix table

--- 2020-08-27 inst/paper/main.tex
@@
 \EventsMinRangeB--\EventsMaxRangeB & \EventsCoreEvalPercB & \EventsPackageEvalPercB \\
-\EventsMinRangeC--\EventsMaxRangeC & \EventsCoreEvalPercC & \EventsPackageEvalPercC \\
+\EventsMinRangeC--\EventsMaxRangeC & \EventsCoreEvalPercC & \EventsPackageEvalPercC \\ \hline
 \end{tabular}} \label{table:event-distribution}
@@
 
-Unlike Javascript, R packages also call native code for access to efficient
-numerical and statistical routines. We intercept all calls to R functions and
-native functions from the programs and \eval calls. In the whole corpus there
-are \AllClosureCallCount calls to R functions and \AllNativeCallCount to native
-code. Of those, \eval contribute to only \EvalClosureCallCount calls to R
-functions and \EvalNativeCallCount to native code.
+%%Unlike Javascript, R packages also call native code for access to efficient
+%%numerical and statistical routines. We intercept all calls to R functions and
+%%native functions from the programs and \eval calls. In the whole corpus there
+%%are \AllClosureCallCount calls to R functions and \AllNativeCallCount to native
+%%code. Of those, \eval contribute to only \EvalClosureCallCount calls to R
+%%functions and \EvalNativeCallCount to native code.
+%%
+%%\begin{table}[ht]\resizebox{\columnwidth}{!}{%
+%%    \begin{tabular}{lll}\hline
+%%      Source & Primitive Functions & R Functions & Native Functions \\ \hline
+%%      Core & \CorePrimitiveFunctionCount & \CoreRFunctionCount & \CoreNativeFunctionCount \\
+%%      CRAN & \PackagePrimitiveFunctionCount & \PackageRFunctionCount & \PackageNativeFunctionCount \\
+%%      All  & \AllPrimitiveFunctionCount & \AllRFunctionCount & \AllNativeFunctionCount \\
+%%    \end{tabular}} \label{table:operation-distribution}
+%%  \caption{Distribution of operation calls to \eval calls across Core and Package}
+%%\end{table}
 
@@
 
-The \c{imchange} function of package \c{imager} makes it possible to modify images using a dedicated formula syntax using \c{\~}.\footnote{Inspired by \c{map} in package \emph{purr}.}
-Here, \eval is evaluated in \c{newenv}, which creates a new environment that inherits from \c{parent.frame()} by default. We classify it as 1+.
+The \c{imchange} function of package \c{imager} makes it possible to modify images using a dedicated formula syntax using \c{\~}.\footnote{Inspired by \c{map} in package \emph{purr}.}  Here, \eval is evaluated in \c{newenv}, which creates a new environment that inherits from \c{parent.frame()} by default.
+We classify it as 1+.
 

### 2020-08-28 inst/analysis/corpus-analysis.Rmd
Updated corpus

--- 2020-08-25 inst/analysis/corpus-analysis.Rmd
@@
 ```{r load corpus}
+evals_dynamic <- read_fst(PACKAGE_EVALS_DYNAMIC_FILE) %>% as_tibble() %>% rename(evals_dyn=n)
+
 corpus <- 
@@
   as_tibble() %>%
+  left_join(evals_dynamic, by="package") %>%
   mutate_at(
@@
       starts_with("funs_"),
-      "evals"
+      "evals",
+      "evals_dyn"
     ), 
@@
 
-```{r load kaggle}
-kaggle_kernels <- as_tibble(read_fst(KAGGLE_KERNELS_FILE))
-kaggle_log <- as_tibble(read_fst(KAGGLE_LOG_FILE)) %>%
-  transmute(
-    id=str_replace(Command, ".*/kaggle-run/[^/]+/([^/]+)/.*", "\\1"),
-    exitval=Exitval,
-    runtime=as.duration(JobRuntime)
-  )
-
-kaggle <- left_join(
-  kaggle_kernels,
-  kaggle_log,
-  by="id"
-)
-```
-
-```{r load runnable code metadata}
-runnable_code_metadata <- read_fst(PACKAGE_RUNNABLE_CODE_FILE) %>%
-  as_tibble() %>%
-  mutate(package=basename(package)) %>%
-  select(package, file=filename, code)
-
-# TODO semi_join with traced-programs.fst
-```
-
-```{r load run metadata}
-runs <- read_fst(PACKAGE_RUNS_FILE) %>%
-  as_tibble() %>%
-  mutate(
-    package=basename(package),
-    file=as.character(path("./", basename(dirname(file)), basename(file)))
-  ) %>%
-  semi_join(corpus, by="package")
-```
-
 ## CRAN packages
 
-```{r compute corpus programs}
+```{r filter out inconsistent programs}
 corpus_programs <-
-  left_join(
-    filter(runs, exitval==0) %>% select(package, file, time), 
-    runnable_code_metadata %>% select(package, file, code),
-    by=c("package", "file")
-  ) %>%
-  transmute(
-    package,
-    file,
-    type=basename(dirname(file)),
-    code, 
-    runtime=time
-  )
-```
-
-There will be some programs that were not strictly part of examples, tests or vignettes, or they do not have
-counts for code size or for execution time.
-
-```{r compute program inconsistencies}
-corpus_programs_inconsistent <-
-  corpus_programs %>%
-  filter(
-    !(type %in% c("examples", "tests", "vignettes")) |
-    is.na(code) | 
-    is.na(runtime) |
-    code == 0
-  )
-
-corpus_programs_inconsistent %>% 
-    my_datatable()
-```
-
-Since, there are just a few of them, we ignore them.
-
-```{r filter out inconsistent programs}
-corpus_programs <- anti_join(corpus_programs, corpus_programs_inconsistent, by=c("package", "file"))
+  read_fst(CRAN_PROGRAMS_FILE) %>%
+  as_tibble()
 ```
@@
   pivot_wider(names_from=type, values_from=c(files, code)) %>%
-  select(-file) %>%
+  select(-file, -runtime, -error, -success, -corpus) %>%
   group_by(package) %>%
@@
   mutate(
-    files=files_examples + files_tests + files_vignettes,
-    code=code_examples + code_tests + code_vignettes
-  )
-
-rs <- function(name, value) {
-  stopifnot(is.numeric(value))
-  bind_rows(
-    r(name, sum(value)),
-    r(paste(name, "mean"), mean(value)),
-    r(paste(name, "median"), median(value)),
-    r(paste(name, "min"), min(value)),
-    r(paste(name, "max"), max(value))
-  )
-}
+    files=files_examples + files_vignettes,
+    code=code_examples + code_vignettes
+  ) %>%
+  ungroup()
 
@@
   rs("package programs code",    runnable_code$code),
-  rs("package programs runtime", as.duration(runnable_code$runtime)),
 
   rs("examples programs",  runnable_code$files_examples),
-  rs("tests programs",     runnable_code$files_tests),
   rs("vignettes programs", runnable_code$files_vignettes),
@@
   rs("examples code",  corpus_program_examples$code),
-  rs("tests code",     corpus_programs_tests$code),
   rs("vignettes code", corpus_programs_vignettes$code),
-
-  rs("examples runtime",  corpus_program_examples$runtime),
-  rs("tests runtime",     corpus_programs_tests$runtime),
-  rs("vignettes runtime", corpus_programs_vignettes$runtime),
 
@@
   mutate(
-    outlier=evals > PACKAGE_EVALS_OUTLIER,# | package_r_code > PACKAGE_SIZE_OUTLIER,
+    outlier=evals_dyn > PACKAGE_EVALS_DYN_OUTLIER,# | package_r_code > PACKAGE_SIZE_OUTLIER,
     label=ifelse(outlier, package, NA),
@@
 #      y=funs_with_eval,
-      y=evals,
+      y=evals_dyn,
       label=label,
@@
 
+```{r kaggle dependencies}
+kaggle_dependencies <- tibble(package=read_lines(KAGGLE_DEPENDENCIES_FILE))
+```
+
 ```{r compute kaggle programs}
+kaggle <- read_fst(KAGGLE_KERNELS_FILE) %>% as_tibble()
 kaggle_programs <-
-  kaggle %>%
-  filter(exitval==0) %>%
-  transmute(
-    id, 
-    code, 
-    runtime=as.numeric(runtime, units="seconds")
-  )
+  read_fst(KAGGLE_PROGRAMS_FILE) %>% as_tibble() %>% filter(success)
 ```
@@
 kaggle_code     <- kaggle_programs$code
-kaggle_runtime  <- kaggle_programs$runtime
 
@@
   r("failed kaggle",          nrow(filter(kaggle, runnable))-nrow(kaggle_programs)),
-  rs("finished kaggle code",   kaggle_code),
-  rs("kaggle runtime",         as.duration(kaggle_runtime)),
+  rs("finished kaggle code",  kaggle_code)
 )
+```
+
+```{r}
+setdiff(kaggle_dependencies$package, corpus$package)
 ```
@@
 programs <- bind_rows(
-  corpus_programs %>% mutate(corpus="packages") %>% select(code, runtime, corpus),
-  kaggle_programs %>% mutate(corpus="kaggle") %>% select(code, runtime, corpus)
+  corpus_programs %>% select(code, runtime, corpus),
+  kaggle_programs %>% select(code, runtime, corpus)
 )

### 2020-08-28 inst/analysis/inc/paths.R

@@ -70,6 +70,7 @@ CRAN_MIRROR_LOCAL_URL <- paste0("file://", params$base_dir, "/CRAN")
 # size of a package to be considered as outlier
 PACKAGE_SIZE_OUTLIER <- 50000
 PACKAGE_EVALS_OUTLIER <- 50
+PACKAGE_EVALS_DYN_OUTLIER <- 10
 MOST_DOWNLOADED <- 25
 # R packages distributed with vanilla R
 CORE_PACKAGES <- c(

### 2020-08-28 inst/paper/main.tex
Updated corpus
comment

--- 2020-08-27 inst/paper/main.tex
@@
 \begin{figure*}[!tb]\centering\includegraphics[width=\linewidth]
-{pipeline.pdf}\caption{Pipeline}\label{fig:pipeline}
+{pipeline.pdf}\caption{Pipeline}label{fig:pipeline}
 \end{figure*}
@@
 
-%Unlike Javascript, R packages also call native code for access to efficient
-%numerical and statistical routines. We intercept all calls to R functions and
-%native functions from the programs and \eval calls. In the whole corpus there
-%are \AllClosureCallCount calls to R functions and \AllNativeCallCount to native
-%code. Of those, \eval contribute to only \EvalClosureCallCount calls to R
-%functions and \EvalNativeCallCount to native code.
+##Unlike Javascript, R packages also call native code for access to efficient
+##numerical and statistical routines. We intercept all calls to R functions and
+##native functions from the programs and \eval calls. In the whole corpus there
+##are \AllClosureCallCount calls to R functions and \AllNativeCallCount to native
+##code. Of those, \eval contribute to only \EvalClosureCallCount calls to R
+##functions and \EvalNativeCallCount to native code.
+##
+##\begin{table}[ht]\resizebox{\columnwidth}{!}{%
+##    \begin{tabular}{lll}\hline
+##      Source & Primitive Functions & R Functions & Native Functions \\ \hline
+##      Core & \CorePrimitiveFunctionCount & \CoreRFunctionCount & \CoreNativeFunctionCount \\
+##      CRAN & \PackagePrimitiveFunctionCount & \PackageRFunctionCount & \PackageNativeFunctionCount \\
+##      All  & \AllPrimitiveFunctionCount & \AllRFunctionCount & \AllNativeFunctionCount \\
+##    \end{tabular}} \label{table:operation-distribution}
+##  \caption{Distribution of operation calls to \eval calls across Core and Package}
+##\end{table}
 

### 2020-08-28 inst/paper/main.tex
Add operation mix introduction

--- 2020-08-27 inst/paper/main.tex
@@
 \input{tag/side-effects.tex}
+\input{tag/operations.tex}
 \input{tag/events.tex}
@@
 \newcommand{\gamlss}{\texttt{gamlss}\xspace}
+\newcommand{\ggproto}{\texttt{ggproto}\xspace}
+\newcommand{\ggplot}{\texttt{ggplot2}\xspace}
 
@@
 
-We first classify \c{eval} depending on the resolved expression, \emph{i.e.}
-the first argument of \c{eval} after all possible function calls and symbol
-resolution in it have been executed.
+\subsection{Operation Mix}
 
-\subsection{Operation Mix}
+In this section we turn our attention to the expression passed for evaluation to
+\eval. \eval accepts all types of R objects as inputs. In our corpus, we observe
+that \AllValueInputEvalCallPerc of all inputs to \eval are values, i.e., they evaluate to
+themselves. The remaining \AllExpressionInputEvalCallPerc inputs are expressions
+which are reduced to values by \eval.
+% Furthermore, the proportion of values
+% passed to core \eval call is \CoreEvalValueInputPerc compared to a significantly
+% higher \PackageEvalValueInputPerc for package \eval calls.
+To investigate this, we look at the type of expressions accepted by \eval
+callsites in our corpus. We find that only \AllExpressionInputEvalSitePerc call
+sites exclusively receive expressions as arguments across all program runs. The
+remaining \AllPolymorphicInputEvalSitePerc call sites don't always accept expressions
+as inputs. The top three of these call sites contribute to over
+~80\% of all calls to evals. The first one is inside \c{match.arg} from core.
+Given a vector of names as strings, this function finds the default values for
+parameters with those names in the caller function. Since the default value for
+a function parameter can be a complex expression or a benign value such as
+\c{NULL}, \c{match.arg} evaluates it through a call to \eval. This makes the
+\eval inside \c{match.arg} ``polymorhpic''. The second such \eval call
+originates from the popular \ggplot package of R which is a DSL for plotting.
+\ggplot has it's own object system, \ggproto, which uses \eval inside the
+\ggproto function to determine the super-class from which it inherits. The
+default value of the superclass is \c{NULL} but in general could be an
+arbitrarily complex expression. The third \eval call is in core R \c{str}
+function. This function is used to print a summary of R objects. It internally
+maps \eval function on the bindings of an environment which results in inputs of
+multiples types being passed to \eval.
+
+%%We further investigate the type of input passed to \eval in
+%%Figure~\cite{fig:eval-expr-type} by the type of input and
+%%the source of \eval. In this figure, \emph{Expression} is the type of program text,
+%%\emph{Environmment} is the type of scopes or environments used for binding,
+%%\emph{Null} is the type of the unique \c{NULL} object, \emph{Vector} is the type
+%%assigned to vectors of strings, booleans, integers, reals and bytes;
+%%\emph{Symbol} is the type of unevaluated symbols, and \emph{Function} is the
+%%type of functions.
+%%
+%%
+%%From the figure, we observe that majority of inputs to Core \eval functions are
+%%expressions in contrast to package \eval calls which receive mostly environments
+%%as inputs. In fact, \PackageEnvironmentEvalPerc of inputs to package evals are
+%%environments.
+%%
+%%Now, we turn our attention to the expressions passed as input to \eval.
 

### 2020-08-28 inst/paper/main.tex
Add more numbers

@@
 To investigate this, we look at the type of expressions accepted by \eval
-callsites in our corpus. We find that only \AllExpressionInputEvalSitePerc call
-sites exclusively receive expressions as arguments across all program runs. The
-remaining \AllPolymorphicInputEvalSitePerc call sites don't always accept expressions
-as inputs. The top three of these call sites contribute to over
+callsites in our corpus. We find that only \AllValueInputEvalSitePerc receive
+only values as inputs, \AllExpressionInputEvalSitePerc call
+sites exclusively receive expressions as inputs and a very large proportion,
+\AllPolymorphicInputEvalSitePerc, receive both expressions and values as inputs across all program runs.
+We inspected the eval callsites that accept only values or both values and expression as inputs.
+The top three of these call sites contribute to over
 ~80\% of all calls to evals. The first one is inside \c{match.arg} from core.

### 2020-08-28 inst/paper/main.tex
Add figure

@@
 \eval. \eval accepts all types of R objects as inputs. In our corpus, we observe
-that \AllValueInputEvalCallPerc of all inputs to \eval are values, i.e., they evaluate to
-themselves. The remaining \AllExpressionInputEvalCallPerc inputs are expressions
-which are reduced to values by \eval.
-% Furthermore, the proportion of values
-% passed to core \eval call is \CoreEvalValueInputPerc compared to a significantly
-% higher \PackageEvalValueInputPerc for package \eval calls.
-To investigate this, we look at the type of expressions accepted by \eval
-callsites in our corpus. We find that only \AllValueInputEvalSitePerc receive
-only values as inputs, \AllExpressionInputEvalSitePerc call
-sites exclusively receive expressions as inputs and a very large proportion,
-\AllPolymorphicInputEvalSitePerc, receive both expressions and values as inputs across all program runs.
-We inspected the eval callsites that accept only values or both values and expression as inputs.
-The top three of these call sites contribute to over
-~80\% of all calls to evals. The first one is inside \c{match.arg} from core.
-Given a vector of names as strings, this function finds the default values for
-parameters with those names in the caller function. Since the default value for
-a function parameter can be a complex expression or a benign value such as
-\c{NULL}, \c{match.arg} evaluates it through a call to \eval. This makes the
-\eval inside \c{match.arg} ``polymorhpic''. The second such \eval call
-originates from the popular \ggplot package of R which is a DSL for plotting.
-\ggplot has it's own object system, \ggproto, which uses \eval inside the
-\ggproto function to determine the super-class from which it inherits. The
-default value of the superclass is \c{NULL} but in general could be an
-arbitrarily complex expression. The third \eval call is in core R \c{str}
+that \AllValueInputEvalCallPerc of all inputs to \eval are values, i.e., they
+evaluate to themselves. The remaining \AllExpressionInputEvalCallPerc inputs are
+expressions which are reduced to values by \eval. % Furthermore, the proportion
+of values % passed to core \eval call is \CoreEvalValueInputPerc compared to a
+significantly % higher \PackageEvalValueInputPerc for package \eval calls. To
+investigate this, we look at the type of expressions accepted by \eval callsites
+in our corpus. We find that only \AllValueInputEvalSitePerc receive only values
+as inputs, \AllExpressionInputEvalSitePerc call sites exclusively receive
+expressions as inputs and a very large proportion,
+\AllPolymorphicInputEvalSitePerc, receive both expressions and values as inputs
+across all program runs. We inspected the eval callsites that accept only values
+or both values and expression as inputs. The top three of these call sites
+contribute to over ~80\% of all calls to evals. The first one is inside
+\c{match.arg} from core. Given a vector of names as strings, this function finds
+the default values for parameters with those names in the caller function. Since
+the default value for a function parameter can be a complex expression or a
+benign value such as \c{NULL}, \c{match.arg} evaluates it through a call to
+\eval. This makes the \eval inside \c{match.arg} ``polymorhpic''. The second
+such \eval call originates from the popular \ggplot package of R which is a DSL
+for plotting. \ggplot has it's own object system, \ggproto, which uses \eval
+inside the \ggproto function to determine the super-class from which it
+inherits. The default value of the superclass is \c{NULL} but in general could
+be an arbitrarily complex expression. The third \eval call is in core R \c{str}
 function. This function is used to print a summary of R objects. It internally
@@
 
-%%We further investigate the type of input passed to \eval in
-%%Figure~\cite{fig:eval-expr-type} by the type of input and
-%%the source of \eval. In this figure, \emph{Expression} is the type of program text,
-%%\emph{Environmment} is the type of scopes or environments used for binding,
-%%\emph{Null} is the type of the unique \c{NULL} object, \emph{Vector} is the type
-%%assigned to vectors of strings, booleans, integers, reals and bytes;
-%%\emph{Symbol} is the type of unevaluated symbols, and \emph{Function} is the
-%%type of functions.
-%%
-%%
-%%From the figure, we observe that majority of inputs to Core \eval functions are
-%%expressions in contrast to package \eval calls which receive mostly environments
-%%as inputs. In fact, \PackageEnvironmentEvalPerc of inputs to package evals are
-%%environments.
-%%
-%%Now, we turn our attention to the expressions passed as input to \eval.
+We further investigate the type of input passed to \eval in
+Figure~\cite{fig:eval-expression-kind} by the type of input and
+the source of \eval. In this figure, \emph{Expression} is the type of program text,
+\emph{Environmment} is the type of scopes or environments used for binding,
+\emph{Null} is the type of the unique \c{NULL} object, \emph{Vector} is the type
+assigned to vectors of strings, booleans, integers, reals and bytes;
+\emph{Symbol} is the type of unevaluated symbols, and \emph{Function} is the
+type of functions.
+
+\begin{figure}[!h]
+	\centering
+  \includegraphics[width=\columnwidth]{eval-expression-kind}
+  \caption{Type of Expression passed to \eval} \label{fig:eval-expression-kind}
+\end{figure}
+
+From the figure, we observe that majority of inputs to Core \eval functions are
+expressions in contrast to package \eval calls which receive mostly environments
+as inputs. In fact, \PackageEnvironmentInputEvalCallPerc of \eval calls in
+packages receive environments as input expressions.
+
+Now, we turn our attention to the expressions passed as input to \eval.
 

### 2020-08-28 inst/paper/main.tex
Add callsite description

@@
 as inputs. In fact, \PackageEnvironmentInputEvalCallPerc of \eval calls in
-packages receive environments as input expressions.
+packages receive environments as input expressions. There are only four
+callsites in package evals that receive environments as expressions.
+\begin{itemize}
+\item \c{ggplot2::ggproto} contributes to 99.9\% of package \eval calls that
+receive environments as input.
+\item \c{R6::generator_funs} function contributes to only 936 \eval calls and
+implements the same functionality as \c{ggplot2::ggproto} but for the \emph{R6}
+OOP system.
+\item \c{future::backtrace} function applies \eval to a future object which is
+implemented as an environment. This is called only once.
+\item \c{RModel::str.RMmodel} function overloads the core \c{str} method for its
+\c{RModel} objects and maps \eval exactly like \c{str}. This is called only
+once.
+\end{itemize}
 
-Now, we turn our attention to the expressions passed as input to \eval.
+\subsubsection{Provenance}
+
+In this section we turn our attention to the \AllExpressionInputEvalCallPerc
+expressions passed as input to \eval. There are multiple ways to obtain
+expressions in R.
+
+\c{eval} takes an R expression. To evaluate a string, it first has to be parse
+using either \c{parse}, \c{str2expression} or \c{str2lang} functions. The
+difference between them is that the latter two are specialized versions of
+\c{parse} that do not keep source references (line and column numbers). This
+happens for \PercentParsedEvals of the eval calls, and \PercentParsedCallSites
+of the call sites. The input to \c{parse} can come from either a string, a file
+or directly a from a user input in interactive REPL session\footnote{in the
+  corpus, there is only one call site to eval that takes parse expression from a
+  user input, \c{invokeRestartInteractively} that prompts user for new argument
+  values when a function invocation fails.}.
+
+This corresponds to strings loaded from a file or to metaprogramming, \ie
+building new variables or functions by concatenating strings together, as in
+Listing~\ref{code:parsepaste}. To detect this kind of metaprogramming, we look
+at the usage of the main concatenating functions in R, \c{paste}, \c{paste0},
+\c{glue} and \c{str_c}, and \PasteParsedEvals of the evals and
+\PasteParsedCallSite of the call sites using a parse function actually call a
+concatenating function.
+
+\begin{lstlisting}[caption={Using \c{eval} and \c{parse} to do meta-programming in package \emph{data.table}, to build the options at load time (in \emph{onLoad.R}).}, label=code:parsepaste]
+eval(parse(text=paste0("options(",i,"=",opts[i],")")))
+\end{lstlisting}
+
+\textbf{What about a table of packages ranked by usage of parse (and/or paste...)?}
 
@@
 
-\subsection{Provenance}
-
-Here we talk about where the input of an eval comes from.
-
-\c{eval} takes an R expression. To evaluate a string, it first has to be parse
-using either \c{parse}, \c{str2expression} or \c{str2lang} functions. The
-difference between them is that the latter two are specialized versions of
-\c{parse} that do not keep source references (line and column numbers). This
-happens for \PercentParsedEvals of the eval calls, and \PercentParsedCallSites
-of the call sites. The input to \c{parse} can come from either a string, a file
-or directly a from a user input in interactive REPL session\footnote{in the
-  corpus, there is only one call site to eval that takes parse expression from a
-  user input, \c{invokeRestartInteractively} that prompts user for new argument
-  values when a function invocation fails.}.
-
-This corresponds to strings loaded from a file or to metaprogramming, \ie
-building new variables or functions by concatenating strings together, as in
-Listing~\ref{code:parsepaste}. To detect this kind of metaprogramming, we look
-at the usage of the main concatenating functions in R, \c{paste}, \c{paste0},
-\c{glue} and \c{str_c}, and \PasteParsedEvals of the evals and
-\PasteParsedCallSite of the call sites using a parse function actually call a
-concatenating function.
-
-\begin{lstlisting}[caption={Using \c{eval} and \c{parse} to do meta-programming in package \emph{data.table}, to build the options at load time (in \emph{onLoad.R}).}, label=code:parsepaste]
-eval(parse(text=paste0("options(",i,"=",opts[i],")")))
-\end{lstlisting}
-
-\textbf{What about a table of packages ranked by usage of parse (and/or paste...)?}
-
 \subsection{Consistency}

### 2020-08-28 inst/paper/main.tex
Add stuff

@@
 
+We looked at the top ten expressions passed to core and
+package \eval calls.
+For package \eval calls, the ten most frequently evaluated expressions account
+for 80\% of all package eval calls. The most frequently occurring call is to the
+\eval in \c{ggplot2::ggproto} which happens over 68\% of the times. The
+remaining calls
+
+%%eval(quote(function(value) freduce(value, `_function_list`)),
+%%env, env), c("fseq", "function")
 \subsubsection{Provenance}
@@
 In this section we turn our attention to the \AllExpressionInputEvalCallPerc
-expressions passed as input to \eval. There are multiple ways to obtain
+expressions passed as input to \eval. There are two ways to obtain
 expressions in R.
+\begin{itemize}
+  \item \c{substitute} function can be used to synthesize ASTs from
+  expressions by replacing symbols with their bindings in the specified
+  environment.
+  \item \c{expression} function can be used to create a vector of expression
+    objects from text.
+  \item \c{parse} function and its variants \c{str2expression} and \c{str2lang}
+    can be used to turn strings or text from file into R expressions.
+\end{itemize}
 
-\c{eval} takes an R expression. To evaluate a string, it first has to be parse
-using either \c{parse}, \c{str2expression} or \c{str2lang} functions. The
-difference between them is that the latter two are specialized versions of
-\c{parse} that do not keep source references (line and column numbers). This
-happens for \PercentParsedEvals of the eval calls, and \PercentParsedCallSites
-of the call sites. The input to \c{parse} can come from either a string, a file
-or directly a from a user input in interactive REPL session\footnote{in the
-  corpus, there is only one call site to eval that takes parse expression from a
-  user input, \c{invokeRestartInteractively} that prompts user for new argument
-  values when a function invocation fails.}.
+In our corpus, we observe 5.2\% cases where \eval directly evaluates output of
+\c{substitute}, 0.7\% cases where output of \c{parse} and its variants is read
+directly and only 704 cases for \c{expression}. Majority of expression objects
+consumed by \eval are generated by their transitive uses by other functions.
 
-This corresponds to strings loaded from a file or to metaprogramming, \ie
-building new variables or functions by concatenating strings together, as in
-Listing~\ref{code:parsepaste}. To detect this kind of metaprogramming, we look
-at the usage of the main concatenating functions in R, \c{paste}, \c{paste0},
-\c{glue} and \c{str_c}, and \PasteParsedEvals of the evals and
-\PasteParsedCallSite of the call sites using a parse function actually call a
-concatenating function.
-
-\begin{lstlisting}[caption={Using \c{eval} and \c{parse} to do meta-programming in package \emph{data.table}, to build the options at load time (in \emph{onLoad.R}).}, label=code:parsepaste]
-eval(parse(text=paste0("options(",i,"=",opts[i],")")))
-\end{lstlisting}
-
-\textbf{What about a table of packages ranked by usage of parse (and/or paste...)?}
+The pattern \c{eval(parse(...))} can be used for dynamic code loading. This
+forms the core of \c{source} and \c{sys.source} functions in R that are commonly
+used for loading code in R files in interactive settings. We investigated the
+number of cases in which the output of \c{parse} and its variants is passed to
+\eval, directly or transitively by tainting their output. We observed that
+\PercentParsedEvals of the eval calls consume the result of calling \c{parse} on
+a file and \PasteParsedEvals of the eval calls consume the result of calling
+\c{parse} on a string. This corresponds to \PercentParsedCallSites and
+\PasteParsedCallSite of the total \eval call sites. We also identified one
+function in core R, \c{invokeRestartInteractively} that prompts the user for
+input, parses it, and passes it to \eval.
 

### 2020-08-28 inst/paper/main.tex
Add more threats

@@
 including only those R packages in our corpus that have many reverse
-dependencies and thus high code coverage. Secondly, we turn off the bytecode
-compiler for this study. The bytecode compiler can also call eval. Thirdly, we
-do not get source locations for \UndefinedEvalsRnd eval calls. In these cases
-\eval is either passed as an argument to a higher-order functions or is defined
-in a function returned by a higher-order function and the R parser does not
-retain location information for \eval. However, this is a meager
-\PercentUndefinedEval of all eval calls and is unlikely to affect our analysis.
+dependencies and thus high code coverage. We turn off the bytecode compiler for
+this study. The bytecode compiler can also call eval. We do not get source
+locations for \UndefinedEvalsRnd eval calls. In these cases \eval is either
+passed as an argument to a higher-order functions or is defined in a function
+returned by a higher-order function and the R parser does not retain location
+information for \eval. However, this is a meager \PercentUndefinedEval of all
+eval calls and is unlikely to affect our analysis. We ignore calls to the native
+\eval function exposed by R. We also ignore the \c{rlang::tidy_eval} function
+which uses native \eval internally because \c{rlang} is used to implement a DSL
+for data analysis in R. It introduces a new first-class promise object called
+\c{quosure} for which it implements special evaluation support in \c{tidy_eval}.
 

### 2020-08-28 inst/paper/main.tex
CRAN calls

@@
 For package \eval calls, the ten most frequently evaluated expressions account
-for 80\% of all package eval calls. The most frequently occurring call is to the
+for 80\% of the calls. The most frequently occurring call is to the
 \eval in \c{ggplot2::ggproto} which happens over 68\% of the times. The
-remaining calls
+remaining expressions occur with a frequency of less than 4\%. They originate
+from \c{plyr}, \c{VGAM}, \c{magrittr}, \c{??}
 
-%%eval(quote(function(value) freduce(value, `_function_list`)),
-%%env, env), c("fseq", "function")
+\begin{table*}[tb]
+  \centering
+  \begin{tabular}{l|rr} \hline
+    Expression & Eval Call & Eval Call \% \\\hline
+    \c{c("auto", "shell", "radix")} & 1987105 & 29.3\%\\
+    \c{c("auto", "shell", "quick", "radix")} & 1593169  & 23.5\%\\
+    \c{local(\{info <- loadingNamespaceInfo(...} & 1008632 &       14.9\%\\
+    \c{c("onLoad", "attach", "detach", "onUnload")}   & 470566 &      6.9\%\\
+    \c{c("append", "prepend", "replace")} &              261587&       3.9\% \\
+    \c{c("left", "right", "centre", "none")} & 162086     & 2.4\%\\
+    \c{c("no", "ifany", "always")}   &                71580 &       1.1\%\\
+    \c{c("pearson", "kendall", "spearman")}  & 72962 &      1.1\%\\
+    \c{NULL}& 75330  &      1.1\% \\
+    \c{Integer Vector}&                 66279&       1\%\\
+  \end{tabular}
+  \caption{Top ten eval calls in CRAN}
+  \label{A}
+\end{table*}
+
 \subsubsection{Provenance}

### 2020-08-28 inst/paper/main.tex
Add expressions for CRAN packages

@@
 
+The most frequent ten expressions to eval calls from core R contribute to
+85.2\% of all \eval calls.
 \begin{table*}[tb]
@@
     \c{c("pearson", "kendall", "spearman")}  & 72962 &      1.1\%\\
-    \c{NULL}& 75330  &      1.1\% \\
-    \c{Integer Vector}&                 66279&       1\%\\
+    \c{\<NULL\>}& 75330  &      1.1\% \\
+    \c{\<Integer Vector\>}&                 66279&       1\%\\\hline
+  \end{tabular}
+  \caption{Top ten eval calls in Core}
+  \label{A}
+\end{table*}
+
+The most frequent ten expressions to eval calls from CRAN packages contribute to
+90\% of all \eval calls.
+\begin{table*}[tb]
+  \centering
+  \begin{tabular}{l|rr} \hline
+    Expression & Eval Call & Eval Call \% \\\hline
+    \c{\<Environment\>} &                                                                                                                           1091474   &  68.2\%\\
+    \c{column[rows] <<- what} &                                                                                                         55677    &   3.5\%\\
+    \c{function(value) freduce(value, `_function_list`)} &                                                                              37251   &    2.3\%\\
+    \c{c("default", "default2012", "default2011" ...}&      20610      & 1.3\%\\
+    \c{force(..1)}                                        &                                                                             20461     &  1.3\%\\
+    \c{alist(`_spec`)}                                   &                                                                              18532    &   1.2\%\\
+    \c{inner}                                           &                                                                               18530   &    1.2\%\\
+    \c{slot(family, "fini")}                           &                                                                                 7288   &    0.5\%\\
+    \c{\~fitted(.)}                                    &                                                                                  4733   &    0.3\%\\
+    \c{function(.) purrr::keep(., is.function)}      &                                                                                   4344  &     0.3\%\\\hline
   \end{tabular}

### 2020-08-28 inst/paper/main.tex
explain

@@
 
-We looked at the top ten expressions passed to core and
-package \eval calls. 
-For package \eval calls, the ten most frequently evaluated expressions account
-for 80\% of the calls. The most frequently occurring call is to the
-\eval in \c{ggplot2::ggproto} which happens over 68\% of the times. The
-remaining expressions occur with a frequency of less than 4\%. They originate
-from \c{plyr}, \c{VGAM}, \c{magrittr}, \c{??}
-
-The most frequent ten expressions to eval calls from core R contribute to
-85.2\% of all \eval calls.
+We looked at the top ten expressions passed to core and package \eval calls. The
+most frequent ten expressions to eval calls from core R contribute to 85.2\% of
+all \eval calls.
 \begin{table*}[tb]
@@
     \c{c("auto", "shell", "quick", "radix")} & 1593169  & 23.5\%\\
-    \c{local(\{info <- loadingNamespaceInfo(...} & 1008632 &       14.9\%\\
+    \c{\{info <- loadingNamespaceInfo(...} & 1008632 &       14.9\%\\
     \c{c("onLoad", "attach", "detach", "onUnload")}   & 470566 &      6.9\%\\
@@
     \c{c("pearson", "kendall", "spearman")}  & 72962 &      1.1\%\\
-    \c{\<NULL\>}& 75330  &      1.1\% \\
-    \c{\<Integer Vector\>}&                 66279&       1\%\\\hline
+    \c{NULL}& 75330  &      1.1\% \\
+    \c{Symbol}&                 66279&       1\%\\\hline
   \end{tabular}
@@
 \end{table*}
+
+The expression \c{\{info <- loadingNamespaceInfo(...} is added by core R
+to a package directory during installation. To load the package, this code is
+executed. It creates a namespace for the package, injects the package bindings,
+and attaches the namespace to the program search path. The \c{NULL} comes from a
+call to \c{substitute(subset)} in \c{stats::model.frame.default} function which
+has a default value of \c{subset} as \c{NULL}. The \c{Symbol} arises from a
+call to \c{as.name} in \c{base::str} function that returns a symbol that is
+looked up by evaluating it in a specific environment. The remaining cases arise
+from calls to \c{match.arg} which is used to look up the default choices for a
+variable and match against the choice passed by the caller.
 
@@
 \end{table*}
+The most frequently occurring call is to the
+\eval in \c{ggplot2::ggproto} which happens over 68\% of the times. The
+remaining expressions occur with a frequency of less than 4\%. They originate
+from \c{plyr}, \c{VGAM}, \c{magrittr}, \c{??}
 

### 2020-08-28 inst/paper/main.tex
Add explanation for CRAN callsites

@@
 The most frequent ten expressions to eval calls from CRAN packages contribute to
-90\% of all \eval calls.
+77.1\% of all \eval calls.
 \begin{table*}[tb]
@@
     Expression & Eval Call & Eval Call \% \\\hline
-    \c{\<Environment\>} &                                                          1091474  & 68.2\%\\
+    \c{Environment} &                                                              989302   & 61.8\%\\
     \c{column[rows] <<- what} &                                                    55677    & 3.5\%\\
     \c{function(value) freduce(value, `_function_list`)} &                         37251    & 2.3\%\\
+    \c{NULL} &                         32005    & 2\%\\
+    \c{List} &                         22293    & 1.4\%\\
     \c{c("default", "default2012", "default2011" ...}&                             20610    & 1.3\%\\
@@
     \c{inner}                                           &                          18530    & 1.2\%\\
-    \c{slot(family, "fini")}                           &                           7288     & 0.5\%\\
-    \c{\~fitted(.)}                                    &                           4733     & 0.3\%\\
-    \c{function(.) purrr::keep(., is.function)}      &                             4344     & 0.3\%\\\hline
+    \c{String Vector}                           &                           17487     & 1.1\%\\
   \end{tabular}
@@
 \end{table*}
-The most frequently occurring call is to the
-\eval in \c{ggplot2::ggproto} which happens over 68\% of the times. The
-remaining expressions occur with a frequency of less than 4\%. They originate
-from \c{plyr}, \c{VGAM}, \c{magrittr}, \c{??}
+
+The expression \c{Environment} occurs because of the four callsites explained
+above, \c{ggplot2::ggproto}, \c{R6::generator_funs}, \c{future::backtrace} and
+\c{RModel::str.RMmodel}. The next expression, \c{column[rows] <<- what}, is used
+inside the \c{plyr::rbind.fill} function to merge data frames by assigning
+concatenated vectors to rows. The \c{<<-} operator is interesting in that it
+skips the current scope and assigns in a parent scope in which the variable is
+already present. In our corpus, all these \eval calls contribute to a single
+side-effect. The expression \c{function(value) freduce(value, `_function_list`)}
+arises from the \c{magrittr::\%>\%} function which is a pipe operator that pipes
+the output of previous command to the next one. The expression is evaluated in a
+custom environment to create a function binding for evaluating the components of
+the pipe.\c{String Vector} and \c{List} also arise from the same function when a
+string or a list is piped using the \c{\%>\%} function into the next expression.
+The \c{NULL} arises from \c{R6::generator_funs} function when the \eval is
+passed a \c{NULL} argument by the \c{DataMask_generator} package. The
+\c{c("default", "default2012", "default2011" ...} pattern arises from
+\c{copula::polyG} where it reflectively access the default expression for its
+formal parameter and evaluates it. The \c{force(..1)} and \c{alist(`_spec`)}
+patterns occur in \c{glue::glue_data} function which concatenates and
+interpolates strings. The two patterns occur because the function captures
+unevaluated unnamed arguments and maps the evaluation of \c{force(..1)} on them.
+The \c{force} function forces promises~\cite{oopsla19a} and returns the result
+of evaluation. The \c{inner} pattern arises from \c{glue::identity_transformer}
+which enables the creation of custom transformation functions for affecting the
+interpolation and concatenation of input by the \c{glue} package.
 

### 2020-08-28 inst/paper/main.tex
Comment irrelevant stuff

@@
 
-\subsection{Consistency}
-
-Here we talk about whether the same call site sees the same pattern
-
-{\bf Polymorphic may not be the right word.}
-
-We look at how \emph{polymorphic} the \c{expr} argument of \eval can be, \ie
-how many different types of the resolved \c{expr} there are per call sites,
-in Figure~\ref{fig:polymorphism}. \PercentMonomorphic of the call sites are
-\emph{monomorphic}.
-
-\begin{figure}[!h]
-	\centering
-    \includegraphics[width=\columnwidth]{polymorphism}
-    \caption{Degree of polymorphism of \c{expr} per call sites.} \label{fig:polymorphism}
-\end{figure}
-
-\subsection{base}
-
-There are 30 functions in \emph{base} that use \eval.
+%%\subsection{Consistency}
+%%Here we talk about whether the same call site sees the same pattern
+%%
+%%{\bf Polymorphic may not be the right word.}
+%%
+%%We look at how \emph{polymorphic} the \c{expr} argument of \eval can be, \ie
+%%how many different types of the resolved \c{expr} there are per call sites,
+%%in Figure~\ref{fig:polymorphism}. \PercentMonomorphic of the call sites are
+%%\emph{monomorphic}.
+%%
+%%\begin{figure}[!h]
+%%	\centering
+%%    \includegraphics[width=\columnwidth]{polymorphism}
+%%    \caption{Degree of polymorphism of \c{expr} per call sites.} \label{fig:polymorphism}
+%%\end{figure}
+%%
+%%\subsection{base}
+%%
+%%There are 30 functions in \emph{base} that use \eval.
 Only five of them use \c{parse}, for legitimate reasons, such \c{source} and \c{sys.source} which load a file and then evaluate it, or \c{invokeRestartInteractively} which prompts the user for arguments to restart execution after an error.
-
-We endeavored  to write these functions without \eval.
+%%
+%%We endeavored  to write these functions without \eval.
 \c{match.arg} appears in \MatchArgPercent of the \eval calls.
@@
 We can rewrite it using its version with two arguments, \c{match.arg(arg, choices)} where \c{choices} are the possible formal arguments of the function, as shown in Listing~\ref{code:matcharg}.
-
-\begin{lstlisting}[caption={Rewriting \c{match.arg} without \eval}, label=code:matcharg]
-f <- function(type=c("examples", "tests", "vignettes")) {
-actual_type = match.arg(type)
-...
-}
-
-f <- function(type=c("examples", "tests", "vignettes")) {
-actual_type = match.arg(type, choices=c("examples", "tests", "vignettes"))
-...
-}
-\end{lstlisting}
+%%
+%%\begin{lstlisting}[caption={Rewriting \c{match.arg} without \eval}, label=code:matcharg]
+%%f <- function(type=c("examples", "tests", "vignettes")) {
+%%actual_type = match.arg(type)
+%%...
+%%}
+%%
+%%f <- function(type=c("examples", "tests", "vignettes")) {
+%%actual_type = match.arg(type, choices=c("examples", "tests", "vignettes"))
+%%...
+%%}
+%%\end{lstlisting}
 

### 2020-08-28 inst/paper/main.tex
Improve environments

--- 2020-08-28 inst/paper/main.tex
@@
 
-The \c{sys.call} option is not used a lot, only \SysCallEnvirPercent as a whole, and always with \c{sys.parent()} as argument.
+%The \c{sys.call} option is not used a lot, only \SysCallEnvirPercent as a whole, and always with \c{sys.parent()} as argument.
 
@@
 
-The \emph{glue} library is used to interpolate string litterals with variables, such as in \c{glue('My name is {name}.')} where \c{name} will be subtituted by its value.
+The \emph{glue} library is used to interpolate string litterals with variables, such as in \c{glue('My name is \{name\}.')} where \c{name} will be subtituted by its value.
 \emph{glue} makes it possible to define transformers that process those interpolated variables.
@@
 
-We finally look at the case where \c{envir} is not an environment, but a list, such as this \c{h2} function, defined inside function \c{adjCoef} in package \emph{actuar}.
+\c{adjCoef} in package \emph{actuar} find the root of an equation defined by a function \c{h} whose arguments must be named \c{x} and \c{y}.
+\c{h} is transformed into an auxiliary function \c{h2} that can be optimized.
+Here, the list used for \c{envir} ensures the correspondance between the textual arguments of \c{h} and the arguments of \c{h2}.
 
 \begin{lstlisting}
+sh <- substitute(h)
+fcall <- paste(sh, "(x, y)")
+...
 h2 <- function(x, y)

### 2020-08-28 inst/paper/main.tex
Fix indentation
Add operation mix

@@
 \section{A taxonomy of eval}
-
-\subsection{Operation Mix}
 
@@
 
+\subsection{Operation Mix}
+In this section, we look at the operations executed by the expressions and
+symbols on evaluation. R is a lazy language, so symbols can be bound to promises
+that can evaluate arbitrary code. All operations in R are function calls, even
+primitive operations such as addition and assignment are treated as calls to
+functions implemented in the interpreter. In our corpus, we observe 2.9G calls
+to R functions of which 125.5M happen inside \eval. Furthermore, unlike
+Javascript, R code also calls into peformant native implementations of numerical
+and statistical routines. In the corpus there are 463.8M calls to native code of
+which only 31.2M happens inside \eval.
+Qualitatively, we observe all kinds of operations being executed inside \eval
+calls, from numerical computing for statistical modeling to reflection and
+metaprogramming for implementing DSLs. Even testing frameworks such \c{genthat}
+heavily rely upon \eval to execute test code snippets. The core R implementation
+is rife with uses of \eval for mundane things like accessing default values of
+formal parameters and in statistical packages for computation.
+
 \subsubsection{Provenance}
@@
 \begin{lstlisting}
-identity_transformer <- function(text, envir) {
+identity_transformer <- function(text, envir) 
+{
     eval(parse(text = text, keep.source = FALSE), envir)

### 2020-08-28 inst/paper/main.tex
add explanation of environment class

@@
 
-Here we talk about scopes and envs
-
 One of the main differences of the R \eval with the javascript one is how it
-interacts with \emph{scope}, or environments in R. In Javascript, \c{eval}
-can access the local scope and the global scope. In R, \c{eval} has an
-argument \c{envir} to indicate in which environment its first argument,
-after being resolved, must be evaluated.
-By default, it is equal to \c{parent.frame()}, which is the parent environment of the \c{eval} call.
-This default argument happens in \DefaultEnvirExprPercent of the eval calls.
-
-This argument can also be \c{NULL}, a list or a dataframe.
-In that case, \c{eval} copies \c{envir} and encloses it the environment defined by \c{enclos}.
-Finally, the argument can also be a number $n$.
-It means that the environment in which the expressions is evaluated will be the results of \c{sys.call(n)} where $n$ refers to the $n$-th call frame after the global environment.
-
-\textbf{Do we speak more about what an environment is, \c{.GlobalEnv}, \c{emptyenv()}, environment navigation here, or in the Background section?}
-
-There are two main environments affected by \eval:
-\begin{itemize}
-	\item the \emph{existing} environment where \c{eval} reads.
-It is the first non new environment.
-\c{eval} will also be able to read in all parent environment of that environment.
-It can be the global environment, a package environment, an empty environment, the callee environment, or a parent environment of the callee: $1$ is the parent, $2$ the grand-parent, and so on.
-	\item the \emph{default} assignment environment \c{eval} writes new bindings.\footnote{It is the default one but the expression in \eval can always write in any other environments using \c{assign} or in the parent environment using \c{<<-}.} If it is different from the read environment and is a new one that inherits from the read environment, we add a symbol \texttt{+} after the read environment.
-\end{itemize}
-
-To give some examples,  global+ refers to a new scope that inherits from the global scope, while 3 refers to the great grand parent of the \eval scope.
-
-Table~\ref{tab:environments} shows the environments when they are \emph{explicitly specified}.
-\eval in R is often performed in a new environment inherited from the global environment for packages, while for core, it is mostly done in the environment of the grand-parent.
-It is also used a lot to evaluate in the context of a dataframe (\emph{list} environment).
-\emph{loop} \textbf{To explain...?
-But we do not see it here anyway}
-
-%The \c{sys.call} option is not used a lot, only \SysCallEnvirPercent as a whole, and always with \c{sys.parent()} as argument.
-
-% Variability: how many different envir class, and different envir expr, for one call site?
-%\begin{figure*}[htbp]
-%	\centering
-%	\includegraphics[width=\linewidth]{environments.pdf}
-%	\caption{The environments in which the expression is evaluated. \c{NA} corresponds to the default environment.} \label{fig:environments}
-%\end{figure*}
+interacts with \emph{scope}, or environments in R. In Javascript, \c{eval} can
+access the local scope and the global scope. In R, \c{eval} has an argument
+\c{envir} to indicate in which environment its first argument must be evaluated.
+By default, it is equal to \c{parent.frame()}, 
+which is the parent environment of \c{eval} call.This default argument
+happens in \DefaultEnvirExprPercent of the eval calls.
+
+This argument can also be \c{NULL}, a \c{List} or a data frame. This happens in
+3.5\% of the cases.
+\c{eval} copies the fields of the list or data frame and creates bindings for
+them in a new environment. This pattern is used to evaluate formulas which can
+directly refer to the fields of the data.
+The \c{envir} argument can also be a number $n$. It means that the environment in
+which the expressions is evaluated will be the result of \c{sys.call(n)} where
+$n$ refers to the $n$-th stack frame.
+
+The top-level environment in R is called the global environment. New
+environments can be created using \c{new.env}. They can be provided a parent
+environment which becomes the enclosing scope of the new environment.
+
+We looked at the environments passed to all the \eval calls in our corpus.
+Table~\ref{tab:environments} summarizes the results. A numeric environment class
+\c{n} denotes the environment of the $n$-th call stack frame from the current
+function. \c{global} denotes the top-level environment and \c{list} denotes a list
+passed for evaluation of formulas. Environment classes of the form $n+$ denote
+the $n$-th environment extended with a new environment. The new environment
+provides a limited form of sandboxing. All assignments using the \c{\<-}
+function occur inside it and prevent the extended environment from
+mutation. However, it is still possible to mutate the extended environment
+using the \c{\<\<-} function; but, that happens rarely.
 
@@
     \centering
-    \begin{tabular}{|c|c||c|c|}
+    \begin{tabular}{c|c|c|c}
         \hline
-        \multicolumn{2}{|c|}{Core} &  \multicolumn{2}{|c|}{Packages}  \\
+        \multicolumn{2}{c|}{Core} & \multicolumn{2}{|c}{Packages}  \\
         \hline
-        Environment & Proportion & Environment & Proportion \\
+        Environment Class & Eval Calls \% & Environment Class & Eval Calls \% \\
         \hline
         \CoreEnvClassA & \CoreProportionA & \PackagesEnvClassA &  \PackagesProportionA\\
-        \hline
-        \CoreEnvClassB &  \CoreProportionB & \PackagesEnvClassB &  \PackagesProportionB \\
-        \hline
-        \CoreEnvClassC & \CoreProportionC  & \PackagesEnvClassC &  \PackagesProportionC \\
-        \hline
-       \CoreEnvClassD  &  \CoreProportionD & \PackagesEnvClassD &  \PackagesProportionD \\
-        \hline
-        \CoreEnvClassE  & \CoreProportionE & \PackagesEnvClassE & \PackagesProportionE \\
+        \CoreEnvClassB & \CoreProportionB & \PackagesEnvClassB &  \PackagesProportionB\\
+        \CoreEnvClassC & \CoreProportionC & \PackagesEnvClassC &  \PackagesProportionC\\
+        \CoreEnvClassD & \CoreProportionD & \PackagesEnvClassD &  \PackagesProportionD\\
+        \CoreEnvClassE & \CoreProportionE & \PackagesEnvClassE &  \PackagesProportionE\\
         \hline
@@
 
-\c{match.arg} (see Listing~\ref{code:matchargenvir}) is the function that contributes the most to the eval calls (\MatchArgPercent) and has environment class 2.
-\c{sys.parent()} refers to the parent environment of \c{match.arg} and so to the grand-parent environment of \c{eval}, \ie what we denote 2.
-
-\begin{lstlisting}[label=code:matchargenvir, caption={\c{match.arg} using \eval with a custom environment.}]
-formal.args <- formals(sys.function(sysP <- sys.parent()))
-choices <- eval(formal.args[[as.character(substitute(arg))]], envir = sys.frame(sysP))
-\end{lstlisting}
-
-In package \emph{gamlss}, function \c{get.K} evaluates its expression in its parent environment, the one of the body of \c{get.K} returned from \c{environment()}.
-
-\begin{lstlisting}
-eval(parse(text = paste(\"fam$dld\", ii, sep = \"\")), envir = environment())
-\end{lstlisting}
-
-The \emph{glue} library is used to interpolate string litterals with variables, such as in \c{glue('My name is \{name\}.')} where \c{name} will be subtituted by its value.
+We observe that a disproportionately high core R \eval calls access the
+caller's caller's environment. This is because many core R functions call
+functions that pass the result of \c{parent.frame()} to an \eval.
+A disproportionately high calls to \eval happen in an extended
+top-level environment. This can be explained by the fact that many packages
+evaluate code passed from the user's workspace in the top-level environment to
+access the bindings.
+
+%%Below, we provide some code snippets to show how some of these environment
+%%classes manifest.
+
+%%The \c{match.arg} (see Listing~\ref{code:matchargenvir}) function has an
+%%environment class 2. \c{sys.parent()} refers to the parent environment of
+%%\c{match.arg} and so to the grand-parent environment of \c{eval} which we denote
+%%as $2$.
+%%
+%%\begin{lstlisting}[label=code:matchargenvir, caption={\c{match.arg} using \eval with a custom environment.}]
+%%formal.args <- formals(sys.function(sysP <- sys.parent()))
+%%choices <- eval(formal.args[[as.character(substitute(arg))]], envir = sys.frame(sysP))
+%%\end{lstlisting}
+
+%%In package \emph{gamlss::get.K} function \c{get.K} evaluates its expression in its
+%%parent environment, the one of the body of \c{get.K} returned from
+%%\c{environment()}.
+%%
+%%\begin{lstlisting}
+%%eval(parse(text = paste(\"fam$dld\", ii, sep = \"\")), envir = environment())
+%%\end{lstlisting}
+%%
+%%
+%%The \emph{glue} library is used to interpolate string litterals with variables, such as in \c{glue('My name is \{name\}.')} where \c{name} will be subtituted by its value.
 \emph{glue} makes it possible to define transformers that process those interpolated variables.
@@
 The interpretation environment for \eval is often global here, as the string litterals are often used to print variabels at the top level.
-
-\begin{lstlisting}
-identity_transformer <- function(text, envir) 
-{
-    eval(parse(text = text, keep.source = FALSE), envir)
-}
-\end{lstlisting}
-
-The \c{imchange} function of package \c{imager} makes it possible to modify images using a dedicated formula syntax using \c{\~}.\footnote{Inspired by \c{map} in package \emph{purr}.}  Here, \eval is evaluated in \c{newenv}, which creates a new environment that inherits from \c{parent.frame()} by default.
+%%
+%%\begin{lstlisting}
+%%identity_transformer <- function(text, envir) 
+%%{
+%%    eval(parse(text = text, keep.source = FALSE), envir)
+%%}
+%%\end{lstlisting}
+%%
+%%
+%%The \c{imchange} function of package \c{imager} makes it possible to modify images using a dedicated formula syntax using \c{\~}.\footnote{Inspired by \c{map} in package \emph{purr}.}  Here, \eval is evaluated in \c{newenv}, which creates a new environment that inherits from \c{parent.frame()} by default.
 We classify it as 1+.
-
-\begin{lstlisting}
-newenv <- new.env()
-...
-fo <- parse(text=as.character(fo)[2])
-im[where] <- eval(fo,envir=newenv,enclos=env)
-\end{lstlisting}
-
-\c{adjCoef} in package \emph{actuar} find the root of an equation defined by a function \c{h} whose arguments must be named \c{x} and \c{y}.
-\c{h} is transformed into an auxiliary function \c{h2} that can be optimized.
+%%
+%%
+%%\begin{lstlisting}
+%%newenv <- new.env()
+%%...
+%%fo <- parse(text=as.character(fo)[2])
+%%im[where] <- eval(fo,envir=newenv,enclos=env)
+%%\end{lstlisting}
+%%
+%%\c{adjCoef} in package \emph{actuar} find the root of an equation defined by a function \c{h} whose arguments must be named \c{x} and \c{y}.
+%%\c{h} is transformed into an auxiliary function \c{h2} that can be optimized.
 Here, the list used for \c{envir} ensures the correspondance between the textual arguments of \c{h} and the arguments of \c{h2}.
-
-\begin{lstlisting}
-sh <- substitute(h)
-fcall <- paste(sh, "(x, y)")
-...
-h2 <- function(x, y)
-    eval(parse(text = fcall),
-    envir = list(x = x, y = y),
-    enclos = parent.frame(2))
-\end{lstlisting}
+%%
+%%
+%%\begin{lstlisting}
+%%sh <- substitute(h)
+%%fcall <- paste(sh, "(x, y)")
+%%...
+%%h2 <- function(x, y)
+%%    eval(parse(text = fcall),
+%%    envir = list(x = x, y = y),
+%%    enclos = parent.frame(2))
+%%\end{lstlisting}
 

### 2020-08-28 inst/analysis/analysis_stable.Rmd
AST size more readable

@@
 source("inc/setup.R")
-theme_set(theme_minimal())
+theme_set(theme_minimal(base_size = 25))
+
 #theme_update(text = element_text(size = 16))
@@
 
-```{r}
+```{r subsample_for_ast}
 ast_sizes <- eval_calls %>% slice_sample(n=10000) %>% mutate(expr_resolved_nodes = map_int(expr_resolved, expr_size_str))
 # Now precomputed during the tracing
-#ast_sizes <- eval_calls
+#ast_sizes <- read_fst(NODE_COUNT_FILE)
 ```
@@
   geom_violin(trim = FALSE, width=.5)+
-  geom_boxplot(width=.05) +
+  #geom_boxplot(width=.05) +
   scale_y_log10(labels = scales::comma) +
@@
 
-envir_sys_call <- classified_envir %>% filter(str_detect(env_class, fixed("sys.class")))
+envir_sys_call <- classified_envir %>% filter(str_detect(env_class, fixed("sys.call")))
 nb_sys_call <- count(envir_sys_call, wt = nb_ev_calls)$n
@@
                               parsing = c("parse", "str2lang", "str2expression", "parse_only", "base::parse"),
-                              defusing = c("quote", "substitute", "bquote", "expr", "quo"),
-                              expressions = c("as.list", "expression", "as.name", "as.expression", "list", "c", "as.data.frame", "as.character", "{", "modify_lang"),
+                              defusing = c("quote", "substitute", "bquote", "expr", "quo", "enquo", expression),
+                              expressions = c("as.list", "as.name", "as.expression", "list", "c", "as.data.frame", "as.character", "{", "modify_lang"),
                               calling = c("call", "as.call", "do.call"),

### 2020-08-28 inst/paper/main.tex

@@
   hand, know about laziness and program defensively around it.}  Thus, our
-corpus distinguished between \emph{core packages} (there are \CorpusCorePackages such
-packages that are distributed together with the R virtual machine), \emph{CRAN
-  packages} (we selected \CorpusPackages curated packages that pass stringent
-community quality checks and are equipped with tests and sample data),
-\emph{Kaggle scripts} (we used \CorpusFinishedKaggle end-user written programs that performs a
-particular data analysis task). There are reasons to believe that \eval
-usage patterns may differ between these datasets. The core R code is
-extremely stable, it has only been maintained for the last 20 years with
-very few far reaching changes. The libraries represent a more lively
-ecosystem with new libraries added each day.  Finally, end user code is
-often one-use scripts thrown together and sometimes never revisited.
+corpus distinguished between \emph{core packages} (there are
+\CorpusCorePackages such packages that are distributed together with the R
+virtual machine), \emph{CRAN packages} (we selected \CorpusPackages curated
+packages that pass stringent community quality checks and are equipped with
+tests and sample data), \emph{Kaggle scripts} (we used \CorpusFinishedKaggle
+end-user written programs that performs a particular data analysis
+task). There are reasons to believe that \eval usage patterns may differ
+between these datasets. The core R code is extremely stable, it has only
+been maintained for the last 20 years with very few far reaching
+changes. The libraries represent a more lively ecosystem with new libraries
+added each day.  Finally, end user code is often one-use scripts thrown
+together and sometimes never revisited.
 
 To understand the behavior of eval we perform a combination of static
-analysis, dynamic analysis and manual inspection. The results presented
+analysis, dynamic analysis and manual code inspection. The results presented
 here summarize our findings.
@@
 languages. In Java, \eval must be synthesized as a combination dynamic
-loading and reflection.\footnote{While Java does not have \eval built in,
+loading and reflection.\footnote{While Java does not have \eval built-in,
   one can take the input to \eval, wrap it in a static method of an
   anonymous class, generate bytecode for that class, invoke the Java class
-  loader to dynamically install it, and then use reflection to call the new
-  method.} Julia provides \eval more directly but limits its scope to top
-level variables and methods. JavaScript support both a, so-called, global
-\eval, and a local \eval. Where the latter has acces to local variables of
-the function being executed (as well as variables of lexically enclosing
-scopes). Finally languages such as R allow \eval to manipulate variables
-defined in the caller environment of the function in which \eval is invoked.
+  loader to dynamically install the new code, and then use reflection to
+  call the new method.} Julia provides \eval more directly but limits its
+scope to top level variables and methods. JavaScript support both a,
+so-called, global \eval, and a local \eval. Where the latter has acces to
+local variables of the function being executed (as well as variables of
+lexically enclosing scopes). Finally languages such as R allow \eval to
+manipulate variables defined in the caller environment of the function in
+which \eval is invoked.
 

### 2020-08-28 inst/paper/main.tex

@@
 \usepackage{hyperref}
+\newcommand{\missingTag}[1]{\textcolor{red}{#1}\xspace}
 \newcommand{\missingNumber}{\textcolor{red}{XX}\xspace}
@@
 In this section we turn our attention to the expression passed for evaluation to
-\eval. \eval accepts all types of R objects as inputs. In our corpus, we observe
-that \AllValueInputEvalCallPerc of all inputs to \eval are values, i.e., they
-evaluate to themselves. The remaining \AllExpressionInputEvalCallPerc inputs are
-expressions which are reduced to values by \eval. % Furthermore, the proportion
-of values % passed to core \eval call is \CoreEvalValueInputPerc compared to a
-significantly % higher \PackageEvalValueInputPerc for package \eval calls. To
-investigate this, we look at the type of expressions accepted by \eval callsites
-in our corpus. We find that only \AllValueInputEvalSitePerc receive only values
-as inputs, \AllExpressionInputEvalSitePerc call sites exclusively receive
-expressions as inputs and a very large proportion,
-\AllPolymorphicInputEvalSitePerc, receive both expressions and values as inputs
+\eval. The \eval function accepts all types of R objects as inputs, but in the
+majority of cases (\AllExpressionInputEvalCallPerc), it is called with
+expressions. The remaining \AllValueInputEvalCallPerc are values, \ie, they
+evaluate to themselves.
+%
+%Furthermore, the proportion of values passed to core
+%\eval call is \missingTag{CoreEvalValueInputPerc} compared to a significantly higher
+%\missingTag{PackageEvalValueInputPerc} for package \eval calls.
+%
+Passing a value into an eval is effectively an no-op so we look closely at the
+expression types the call sites pass to eval across all the runs.
+\AllValueInputEvalSitePerc of all call sites always pass values,
+\AllExpressionInputEvalSitePerc pass expressions and a very large proportion,
+\AllPolymorphicInputEvalSitePerc, pass both expressions and values as inputs
 across all program runs. We inspected the eval callsites that accept only values
@@
 callsites in package evals that receive environments as expressions.
-\begin{itemize}
+\begin{compactitem}[$-$]
 \item \c{ggplot2::ggproto} contributes to 99.9\% of package \eval calls that
@@
 once.
-\end{itemize}
+\end{compactitem}
 

### 2020-08-28 inst/paper/main.tex
related

@@
   \CorpusAllCodeRnd lines of data science libraries and end-user written in
-  the R language. We performed a simple static analysis and more
-  sophisticated dynamic analysis to confirm that \eval is indeed in wide
-  spread use, and to categorized patters of usage. We found that \eval in
-  this code base is more dangerous in some ways, and safer in others, than
-  what was previously reported for JavaScript.
+  the R language. We performed a sophisticated dynamic analysis to confirm
+  that \eval is indeed in wide spread use, and to categorized patters of
+  usage. We found that \eval in this code base is more dangerous in some
+  ways, and safer in others, than what was previously reported for
+  JavaScript.
 \end{abstract}
@@
 
-To understand the behavior of eval we perform a combination of static
-analysis, dynamic analysis and manual code inspection. The results presented
-here summarize our findings.
+To understand the behavior of eval we perform a dynamic analysis and manual
+code inspection. The results presented here summarize our findings.
 
@@
 The difference between these design choices has an impact on the ability of
-compilers to optimize, and of analysis tools to reason about, code. The
-impact of a top level \eval is that any global definition may be affected
-but the current function can continue to execute unchanged. A local \eval
-may change the value of local variables, their types, or even their
-presence. This is a much more invasive change for reasoning about code and
-may invalidate common optimizations such as constant folding or copy
-propagation. The designers of Julia were aware of this difference and chose
-to restrict \eval to preserve their ability to compile
-code~\cite{oopsla18a}. They even added a versioning mechanism to function
-definitions to ensure that new code added during an \eval would not
-invalidate optimizations such as inlining.  R's ability to affect any one of
-the callers of a function is even more invasive as the impact of \eval is
-non-local. Such unfettered power is difficult to deal with.
+compilers to optimize, and of analysis tools to reason about code. The
+impact of a top level \eval is that any global variable may be modified, but
+currently running functions can continue to execute unchanged. A local
+\eval, may change the values of local variables, their types, or even their
+presence. This is a much more invasive change, one that may invalidate
+currently executing functions as it may change the results of common
+optimizations such as constant folding or copy propagation. The designers of
+the Julia programming language were aware of this difference and
+purposefully chose to restrict \eval to the top level in order to preserve
+their ability to optimize code~\cite{oopsla18a}. They even added a
+versioning mechanism to function definitions to ensure that code added
+during an \eval would not invalidate optimizations such as inlining.  R's
+ability to affect any one of the callers of a function is even more invasive
+as the impact of \eval is non-local. Such unfettered power is difficult to
+deal with.
 
@@
 the use of a single \eval may invalidate every single invariant computed by
-a static analysis or optimizer, this worst case usually does not happen.  In
-many cases, \eval is used carefully and has limited effect. This is a
-hypothesis that must be validated and, to be useful, we should have the
-means to provide some reasonable guesses as to which side-effects will
-result from any particular \eval call.
-
-Previous work looked at eval in JavaScript
-
-Othere languages
-
-Dynamic analysis
+a static analyzer, this worst case usually does not happen.  In many cases,
+\eval is used carefully to limited effect. This hypothesis must be validated
+and, for it to be useful, we need to have a way to make some educated
+guesses as to which side-effects will result from any particular \eval.
+
+Richards et al.~\cite{ecoop11} provided the first large-scale study of the
+runtime behavior of \eval in JavaScript. They analyzed a corpus of the
+10,000 most popular websites. They used an instrumented web browser to
+gather execution traces and perform a dynamic analysis.  Their results show
+that \eval is pervasive with 82\% of the most popular websites using it. The
+reasons for using \eval include the desire to load code on demand (to reduce
+the time it takes to display a page), the deserialization of JSON data
+(about 37\% of all calls) and lightweight meta-programming to customize web
+pages.  While many uses \eval were legitimate, many were unnecessary and
+could be replaced with equivalent and safer code.  The reported on a
+categorization of inputs to \eval that covers the vast majority of input
+strings.  Restricting themselves to \eval in which all named variables refer
+to the global scope, many patterns can be replaced by more disciplined
+code~\cite{oopsla12b}.  The limitation of that work are that there was no
+measure of code coverage, so the numbers presented can be a lower bound on
+the possible behaviors.  Furthermore in 2011, JavaScript usage is quite
+different from today. Their work captured very few libraries and none of the
+more recent Node.js ecosystem.
 
@@
 
+\begin{figure*}[!tb]\centering\includegraphics[width=.8\linewidth]
+{pipeline.pdf}\caption{Analysis Pipeline}\label{fig:pipeline}
+\end{figure*}
+
 \section{Infrastructure}
@@
 runtime, input and output data size for each step.
-
-\begin{figure*}[!tb]\centering\includegraphics[width=\linewidth]
-{pipeline.pdf}\caption{Pipeline}label{fig:pipeline}
-\end{figure*}
 
@@
 
-\begin{figure*}[!tb]\centering\includegraphics[width=\linewidth]
+\begin{figure*}[!tb]\centering\includegraphics[width=.8\linewidth]
   {corpus.pdf}\caption{CRAN packages}\label{fig:corpus}

### 2020-08-28 inst/analysis/analysis_stable.Rmd
Bette rplot

@@
 ```{r subsample_for_ast}
-ast_sizes <- eval_calls %>% slice_sample(n=10000) %>% mutate(expr_resolved_nodes = map_int(expr_resolved, expr_size_str))
+#ast_sizes <- eval_calls %>% slice_sample(n=10000) %>% mutate(expr_resolved_nodes = map_int(expr_resolved, expr_size_str))
 # Now precomputed during the tracing
-#ast_sizes <- read_fst(NODE_COUNT_FILE)
+ast_sizes <- read_fst(path(RUN_DIR, "package-evals-traced.3/summarized-ast-size.fst")) %>% as_tibble() %>% 
+      filter(eval_source_type != "<undefined>") %>%
+      rename(dataset = eval_source_type) 
 ```
@@
 ast_sizes %>%
-  ggplot(aes(x = dataset, y = expr_resolved_nodes, fill = dataset, weight = nb_ev_calls)) +
+  ggplot(aes(x = dataset, y = expr_resolved_nodes, fill = dataset, weight = nb_ev_calls), na.rm = TRUE) +
   geom_violin(trim = FALSE, width=.5)+
-  #geom_boxplot(width=.05) +
+  geom_boxplot(width=.05) +
   scale_y_log10(labels = scales::comma) +
@@
   theme(
+     legend.position=c(0.2, 0.85),
+    legend.box.background = element_rect(fill="white", size=0.1),
     legend.box="horizontal",
@@
                               parsing = c("parse", "str2lang", "str2expression", "parse_only", "base::parse"),
-                              defusing = c("quote", "substitute", "bquote", "expr", "quo", "enquo", expression),
-                              expressions = c("as.list", "as.name", "as.expression", "list", "c", "as.data.frame", "as.character", "{", "modify_lang"),
+                              defusing = c("quote", "substitute", "bquote", "expr", "quo", "enquo", "expression", "as.name", "as.expression"),
+                              expressions = c("as.list",  "list", "c", "as.data.frame", "as.character", "{", "modify_lang"),
                               calling = c("call", "as.call", "do.call"),
@@
   facet_wrap(vars(eval_function))
-```
-
-Another interesting figure is the degreee of polymorphism of the `expr` argument: how many different types (for `expr_resolved`) are there for one call sites?
-
-```{r polymorphism_callsite}
-expr_resolved_polymorphism <- eval_calls %>% group_by(eval_call_srcref) %>%
-        summarise(polym_degree = vec_unique_count(expr_resolved_type))
-
-expr_resolved_monomorphic <- expr_resolved_polymorphism %>% filter(polym_degree == 1)
-nb_monomorphic <- expr_resolved_monomorphic %>% nrow()
-
-r("nb monomorphic", nb_monomorphic)
-r("percent monomorphic", ratio(nb_monomorphic, nb_call_sites))
-```
-
-```{r polymorphism}
-expr_resolved_polymorphism %>% ggplot() +
-  geom_bar(aes(x = polym_degree)) +
-  scale_x_continuous(breaks=c(0, 1, 5, 10)) +
-  labs(title = "Degree of polymorphism", x = "degree", y = "n")
-
-ggsave(path(PLOT_DIR, "polymorphism.pdf"))
-```
-
-`r nb_monomorphic` call sites are not polymorphic, which represents `r nb_monomorphic / nb_call_sites * 100`% of the call sites.
-
-From the monomorphic functions, we can deduce how many are _replaceable_:
-
-```{r}
-expr_resolved_replaceable <- expr_resolved_monomorphic %>% left_join(eval_calls) %>% mutate(can_replace = map_lgl(expr_resolved,  is_replaceable_str))
 ```
@@
 
+Another interesting figure is the degree of consistency of the `expr` argument: how many different types (for `expr_resolved`) are there for one call sites?
+
+```{r polymorphism_callsite}
+expr_resolved_polymorphism <- eval_calls %>% group_by(eval_call_srcref) %>%
+        summarise(polym_degree = vec_unique_count(expr_resolved_type))
+
+expr_resolved_monomorphic <- expr_resolved_polymorphism %>% filter(polym_degree == 1)
+nb_monomorphic <- expr_resolved_monomorphic %>% nrow()
+
+r("nb monomorphic", nb_monomorphic)
+r("percent monomorphic", ratio(nb_monomorphic, nb_call_sites))
+
+nb
+```
+
+```{r polymorphism}
+expr_resolved_polymorphism %>% ggplot() +
+  geom_bar(aes(x = polym_degree)) +
+  scale_x_continuous(breaks=c(0, 1, 5, 10)) +
+  labs( x = "Number of different types", y = "n")
+
+ggsave(path(PLOT_DIR, "polymorphism.pdf"))
+```
+
+`r nb_monomorphic` call sites are not polymorphic, which represents `r nb_monomorphic / nb_call_sites * 100`% of the call sites.
+
+From the monomorphic functions, we can deduce how many are _replaceable_:
+
+```{r}
+expr_resolved_replaceable <- expr_resolved_monomorphic %>% left_join(eval_calls) %>% mutate(can_replace = map_lgl(expr_resolved,  is_replaceable_str))
+```
+
 # Other faces of `eval`

### 2020-08-28 inst/analysis/analysis_stable.Rmd

@@
 ast_sizes <- read_fst(path(RUN_DIR, "package-evals-traced.3/summarized-ast-size.fst")) %>% as_tibble() %>% 
-      filter(eval_source_type != "<undefined>") %>%
-      rename(dataset = eval_source_type) 
+      filter(eval_source_type != "<undefined>")
 ```
@@
 ast_sizes %>%
-  ggplot(aes(x = dataset, y = expr_resolved_nodes, fill = dataset, weight = nb_ev_calls), na.rm = TRUE) +
+  ggplot(aes(x = eval_source_type, y = expr_resolved_nodes, fill = eval_source_type, weight = nb_ev_calls)) +
   geom_violin(trim = FALSE, width=.5)+
-  geom_boxplot(width=.05) +
+  #geom_boxplot(width=.05) +
   scale_y_log10(labels = scales::comma) +
@@
   theme(
-     legend.position=c(0.2, 0.85),
-    legend.box.background = element_rect(fill="white", size=0.1),
     legend.box="horizontal",
@@
 
+Another interesting figure is the degreee of polymorphism of the `expr` argument: how many different types (for `expr_resolved`) are there for one call sites?
+
+```{r polymorphism_callsite}
+expr_resolved_polymorphism <- eval_calls %>% group_by(eval_call_srcref) %>%
+        summarise(polym_degree = vec_unique_count(expr_resolved_type))
+
+expr_resolved_monomorphic <- expr_resolved_polymorphism %>% filter(polym_degree == 1)
+nb_monomorphic <- expr_resolved_monomorphic %>% nrow()
+
+r("nb monomorphic", nb_monomorphic)
+r("percent monomorphic", ratio(nb_monomorphic, nb_call_sites))
+```
+
+```{r polymorphism}
+expr_resolved_polymorphism %>% ggplot() +
+  geom_bar(aes(x = polym_degree)) +
+  scale_x_continuous(breaks=c(0, 1, 5, 10)) +
+  labs( x = "Number of different types", y = "n")
+
+ggsave(path(PLOT_DIR, "polymorphism.pdf"))
+```
+
+`r nb_monomorphic` call sites are not polymorphic, which represents `r nb_monomorphic / nb_call_sites * 100`% of the call sites.
+
+From the monomorphic functions, we can deduce how many are _replaceable_:
+
+```{r}
+expr_resolved_replaceable <- expr_resolved_monomorphic %>% left_join(eval_calls) %>% mutate(can_replace = map_lgl(expr_resolved,  is_replaceable_str))
+```
+
 ## Provenance
@@
 
-Another interesting figure is the degree of consistency of the `expr` argument: how many different types (for `expr_resolved`) are there for one call sites?
-
-```{r polymorphism_callsite}
-expr_resolved_polymorphism <- eval_calls %>% group_by(eval_call_srcref) %>%
-        summarise(polym_degree = vec_unique_count(expr_resolved_type))
-
-expr_resolved_monomorphic <- expr_resolved_polymorphism %>% filter(polym_degree == 1)
-nb_monomorphic <- expr_resolved_monomorphic %>% nrow()
-
-r("nb monomorphic", nb_monomorphic)
-r("percent monomorphic", ratio(nb_monomorphic, nb_call_sites))
-
-nb
-```
-
-```{r polymorphism}
-expr_resolved_polymorphism %>% ggplot() +
-  geom_bar(aes(x = polym_degree)) +
-  scale_x_continuous(breaks=c(0, 1, 5, 10)) +
-  labs( x = "Number of different types", y = "n")
-
-ggsave(path(PLOT_DIR, "polymorphism.pdf"))
-```
-
-`r nb_monomorphic` call sites are not polymorphic, which represents `r nb_monomorphic / nb_call_sites * 100`% of the call sites.
-
-From the monomorphic functions, we can deduce how many are _replaceable_:
-
-```{r}
-expr_resolved_replaceable <- expr_resolved_monomorphic %>% left_join(eval_calls) %>% mutate(can_replace = map_lgl(expr_resolved,  is_replaceable_str))
-```
-
 # Other faces of `eval`

### 2020-08-28 inst/analysis/analysis_stable.Rmd
Typos. Consistency WIP

@@
 params:
-  base_dir: /var/lib/R/project-evalR
+  base_dir: /home/pierre/Documents/Articles/ISCE2021
 
@@
 ast_sizes %>%
-  ggplot(aes(x = eval_source_type, y = expr_resolved_nodes, fill = eval_source_type, weight = nb_ev_calls)) +
+  ggplot(aes(x = eval_source_type, y = expr_resolved_nodes, fill = eval_source_type, weight = nb_ev_calls), na.rm = TRUE) +
   geom_violin(trim = FALSE, width=.5)+
-  #geom_boxplot(width=.05) +
+  geom_boxplot(width=.05) +
   scale_y_log10(labels = scales::comma) +
@@
   theme(
+    legend.position=c(0.5, 0.85),
+    legend.box.background = element_rect(fill="white", size=0.1),
     legend.box="horizontal",
@@
 
+```{r}
+eval_calls %>% semi_join(expr_resolved_polymorphism %>% filter(polym_degree == max(polym_degree)), by ="eval_call_srcref")
+```
+
 From the monomorphic functions, we can deduce how many are _replaceable_:
@@
 ```{r}
-expr_resolved_replaceable <- expr_resolved_monomorphic %>% left_join(eval_calls) %>% mutate(can_replace = map_lgl(expr_resolved,  is_replaceable_str))
+#expr_resolved_replaceable <- expr_resolved_monomorphic %>% left_join(eval_calls) %>% mutate(can_replace = map_lgl(expr_resolved,  is_replaceable_str))
 ```

### 2020-08-28 inst/paper/main.tex
B
Typos. Consistency WIP

@@
 
-The three other functions are defined using \lstinline|eval|.
+The three other functions are defined using \eval.
 
@@
 
-\c{expr} is the expression to be evaluated. It is not a string as in javascript. In order to evaluate a string, it should be first parsed using \c(parse(text = str)) to get an expression.
+\c{expr} is the expression to be evaluated. It is not a string as in javascript. In order to evaluate a string, it should be first parsed using \c{parse(text = str)} to get an expression.
 
@@
 Calls and expressions are evaluated.
+
+\begin{lstlisting}
+> eval(2)
+[1] 2
+> eval(expression(1+2))
+[1] 3
+> eval(parse(text = "1 + 2"))
+[1] 3
+\end{lstlisting}
 
@@
 
-%%\subsection{Consistency}
-%%Here we talk about whether the same call site sees the same pattern
-%%
-%%{\bf Polymorphic may not be the right word.}
-%%
-%%We look at how \emph{polymorphic} the \c{expr} argument of \eval can be, \ie
-%%how many different types of the resolved \c{expr} there are per call sites,
-%%in Figure~\ref{fig:polymorphism}. \PercentMonomorphic of the call sites are
-%%\emph{monomorphic}.
-%%
-%%\begin{figure}[!h]
-%%	\centering
-%%    \includegraphics[width=\columnwidth]{polymorphism}
-%%    \caption{Degree of polymorphism of \c{expr} per call sites.} \label{fig:polymorphism}
-%%\end{figure}
-%%
-%%\subsection{base}
-%%
+\subsection{Consistency}
+
+We look at how \emph{polymorphic} the \c{expr} argument of \eval can be, \ie
+how many different types of the resolved \c{expr} there are per call sites,
+in Figure~\ref{fig:polymorphism}. Most of the call sites, \ie \PercentMonomorphic,  are
+\emph{consistent}. However, a few ones are \emph{polymorphic}, such as:
+
+\begin{figure}[!h]
+	\centering
+    \includegraphics[width=\columnwidth]{polymorphism}
+    \caption{Number of different types  of \c{expr} per call sites.} \label{fig:polymorphism}
+\end{figure}
+%
+%\subsection{base}
+%
 %%There are 30 functions in \emph{base} that use \eval.

### 2020-08-28 inst/paper/main.tex
related

@@
 different from today. Their work captured very few libraries and none of the
-more recent Node.js ecosystem.
+more recent Node.js ecosystem.  More details about dynamic analysis of
+JavaScript can be found in~\cite{liang}.
+
+Wang et al.~\cite{wang} analyzed usage of dynamic features in 18 Python
+programs include dynamic code generation to find if they affects file
+change-proneness.  They reported that files with dynamic features are
+significantly more likely to be the subject of changes, than other files.
+Chen et al. looked at the correlation between code changes and dynamic
+features, including \eval, in 17 Python programs~\cite{chen}. They did not
+observe many uses of \eval.  Callau et al. performed an empirical study of
+the usage of dynamic features in the 1,000 largest Smalltalk projects in the
+Squeaksource repository, accounting for more than 4 million lines of
+code~\cite{oscar}. While \eval itself is not present, Smalltalk has a rich
+reflective interface that gives programmer great flexibility. Their work was
+based on a static analysis.
+
+Morandat et al.~\cite{ecoop12} had a short section on the usage of \eval in
+R. They found the \eval function to be widely used in R code with 8500 call
+sites in CRAN and 2.2 million dynamic calls. The 15 most frequent call sites
+account for 88\% of eval. The \c{match.arg} function is the highest user
+with 54\% of all calls. In the other call sites, they saw two uses
+cases. The most common is the evaluation of the source code of a promise
+retrieved by \c{substitute} in a new environment. This is done in the
+\c{with} function. The other use case is the invocation of a function whose
+name or arguments are determined dynamically. For this purpose, R provides
+\c{do.call} and thus using \eval is overkill.
 

### 2020-08-28 inst/paper/main.tex
design

@@
 \section{The Design of Eval}
+
+The R Project is a key tool for data analysis.  At the heart of the R
+project is a \emph{vectorized, dynamic, lazy, functional, object-oriented}
+programming language with a rather unusual combination of
+features~\cite{ecoop12} designed to ease learning by non-programmer and
+enable rapid development of new statistical methods.  The language was
+designed~\cite{R96} as a successor to S~\cite{S88}.
+
+In R, most data types are vectorized. Values are constructed by the
+\c{c(...)} function: \c{c("hi","ho")} creates a vector of two strings. The
+language does not differentiate scalars from vectors, thus \c{1==c(1)}. In
+order to enable equational reasoning, R copies most values accessible
+through multiple variables if they are written to.  Every linguistic
+construct is desugared to a function call, even control flow statements,
+assignments, and bracketing. Furthermore, all functions can be redefined in
+packages or user code. This makes R both flexible and challenging to
+compile~\cite{dls19}. In R, arguments to a user-defined function are bundled
+into a thunk called a promise. Logically, a promise combines an expression's
+code, its environment, and its value.  To access the value of a promise, one
+must force it. Forcing a promise triggers evaluation and the computed value
+is captured for future reference.  Values can be tagged by user-defined
+attributes. For instance, one can attach the attribute \c{dim} to the value
+\c{x<-c(1,2,3,4)} by evaluating \c{attr(x,"dim")<-c(2,2)}.  Once done,
+arithmetic functions will treat \c x as a \c{2x2} matrix. Another attribute
+is \c{class} which can be bound to a list of names. For instance,
+\c{class(x)<-"human"}, sets the class of \c{x} to \c{human}.  Attributes are
+used for object-oriented dispatch.
+
+R supports reflection and meta-programming. The \c{substitute(e,envir)}
+function yields the parse tree of the expression \c{e} after performing
+substitutions defined by the bindings in \c{envir}.
+
+\begin{lstlisting}
+> substitute(expression(a + b), list(a = 1)))  
+expression(1 + b)
+\end{lstlisting}
+\vspace{2mm}
+
+\noindent
+R allows programmatic manipulation of parse trees, which are themselves
+first class objects. They are evaluated using the \c{eval(e,envir)}
+function. Environment on the call stack can also be accessed.
 

### 2020-08-28 inst/paper/main.tex
design

@@
 
-The R Project is a key tool for data analysis.  At the heart of the R
-project is a \emph{vectorized, dynamic, lazy, functional, object-oriented}
-programming language with a rather unusual combination of
-features~\cite{ecoop12} designed to ease learning by non-programmer and
-enable rapid development of new statistical methods.  The language was
-designed~\cite{R96} as a successor to S~\cite{S88}.
+The R Project is a tool for data analysis.  At its heart is a {vectorized,
+  dynamic, lazy, functional, object-oriented} programming language with a
+rather unusual combination of features~\cite{ecoop12} designed to ease
+learning by non-programmer and enable rapid development of statistical
+methods~\cite{R96}.  The language was designed as a successor to
+S~\cite{S88}.
 
@@
 \c{c(...)} function: \c{c("hi","ho")} creates a vector of two strings. The
-language does not differentiate scalars from vectors, thus \c{1==c(1)}. In
-order to enable equational reasoning, R copies most values accessible
-through multiple variables if they are written to.  Every linguistic
-construct is desugared to a function call, even control flow statements,
-assignments, and bracketing. Furthermore, all functions can be redefined in
-packages or user code. This makes R both flexible and challenging to
-compile~\cite{dls19}. In R, arguments to a user-defined function are bundled
-into a thunk called a promise. Logically, a promise combines an expression's
-code, its environment, and its value.  To access the value of a promise, one
-must force it. Forcing a promise triggers evaluation and the computed value
-is captured for future reference.  Values can be tagged by user-defined
-attributes. For instance, one can attach the attribute \c{dim} to the value
+language does not differentiate scalars from vectors. To enable equational
+reasoning, R copies values accessible through multiple variables when they
+are written to.  Values can be tagged by user-defined attributes. For
+instance, one can attach the attribute \c{dim} to the value
 \c{x<-c(1,2,3,4)} by evaluating \c{attr(x,"dim")<-c(2,2)}.  Once done,
@@
 \c{class(x)<-"human"}, sets the class of \c{x} to \c{human}.  Attributes are
-used for object-oriented dispatch.
+used for object-oriented dispatch.  Every linguistic construct is desugared
+to a function call, even control flow statements, assignments, and
+bracketing. Furthermore, all functions can be redefined in user code. This
+makes R both flexible and challenging to compile~\cite{dls19}.
+
+Arguments to user-defined functions are bundled into a thunk called a
+promise. Logically, a promise combines an expression's code, its
+environment, and its value.  To access the value of a promise, one must
+force it. Forcing a promise triggers evaluation and the computed value is
+captured for future reference.
 

### 2020-08-28 inst/analysis/analysis_stable.Rmd

@@
 params:
-  base_dir: /home/pierre/Documents/Articles/ISCE2021
+  base_dir: /var/lib/R/project-evalR
 

### 2020-08-28 inst/paper/main.tex
stuff

@@
  language=R,              % the language of the code
- basicstyle=\small\ttfamily,    % the size of the fonts
+ basicstyle=\small\sffamily,    % the size of the fonts
 % numbers=left,            % where to put the line-numbers
@@
 
-\newcommand{\eval}{\texttt{eval}\xspace}
-\newcommand{\parse}{\texttt{parse}\xspace}
-\newcommand{\source}{\texttt{source}\xspace}
-\newcommand{\local}{\texttt{local}\xspace}
-\newcommand{\unlockBinding}{\texttt{unlockBinding}\xspace}
-\newcommand{\substitute}{\texttt{substitute}\xspace}
-\newcommand{\datatable}{\texttt{data.table}\xspace}
-\newcommand{\mlogit}{\texttt{mlogit}\xspace}
-\newcommand{\mboost}{\texttt{mboost}\xspace}
-\newcommand{\metafor}{\texttt{metafor}\xspace}
-\newcommand{\lavaan}{\texttt{lavaan}\xspace}
-\newcommand{\mclust}{\texttt{mclust}\xspace}
-\newcommand{\gamlss}{\texttt{gamlss}\xspace}
-\newcommand{\ggproto}{\texttt{ggproto}\xspace}
-\newcommand{\ggplot}{\texttt{ggplot2}\xspace}
-
-\newcommand{\base}{\texttt{base}\xspace}
+\newcommand{\eval}{\c{eval}\xspace}
+\newcommand{\parse}{\c{parse}\xspace}
+\newcommand{\source}{\c{source}\xspace}
+\newcommand{\local}{\c{local}\xspace}
+\newcommand{\unlockBinding}{\c{unlockBinding}\xspace}
+\newcommand{\substitute}{\c{substitute}\xspace}
+\newcommand{\datatable}{\c{data.table}\xspace}
+\newcommand{\mlogit}{\c{mlogit}\xspace}
+\newcommand{\mboost}{\c{mboost}\xspace}
+\newcommand{\metafor}{\c{metafor}\xspace}
+\newcommand{\lavaan}{\c{lavaan}\xspace}
+\newcommand{\mclust}{\c{mclust}\xspace}
+\newcommand{\gamlss}{\c{gamlss}\xspace}
+\newcommand{\ggproto}{\c{ggproto}\xspace}
+\newcommand{\ggplot}{\c{ggplot2}\xspace}
+
+\newcommand{\base}{\c{base}\xspace}
 \renewcommand{\c}[1]{\lstinline{#1}\xspace}
@@
 
-\section{The Design of Eval}
+\section{R and The Design of Eval}
 
@@
 
-R supports reflection and meta-programming. The \c{substitute(e,envir)}
-function yields the parse tree of the expression \c{e} after performing
-substitutions defined by the bindings in \c{envir}.
+R supports reflection and meta-programming. The function
+\c{substitute(e,envir)} yields the parse tree of the expression \c{e} after
+performing substitutions defined by the bindings in \c{envir}.
 
 \begin{lstlisting}
-> substitute(expression(a + b), list(a = 1)))  
-expression(1 + b)
+> substitute(expression(a+b), list(a=1)))  
+expression(1+b)
 \end{lstlisting}
-\vspace{2mm}
 
@@
 R allows programmatic manipulation of parse trees, which are themselves
-first class objects. They are evaluated using the \c{eval(e,envir)}
-function. Environment on the call stack can also be accessed.
-
-There are four different kinds of \eval in the \emph{base} package:
-\c{eval}, \c{evalq}, \c{eval.parent}, \c{local}.  The most important one is
-\eval:
+first class objects. They are evaluated using the \c{eval()}
+function. There are three alternative forms  of \eval in the \c{base} package:
+\c{evalq}, \c{eval.parent}, \c{local}.
 
 \begin{lstlisting}
-  eval <- function(expr, envir=parent.frame(),
-                   enclos=if (is.list(envir)||is.pairlist(envir)) parent.frame() else baseenv())  
-	.Internal(eval(expr, envir, enclos))
+ eval(expr,
+      envir = parent.frame(),
+      enclos = if(is.list(envir) || is.pairlist(envir))
+               parent.frame() else baseenv())
 \end{lstlisting}
 
-The three other functions are defined using \eval.
-
-\begin{lstlisting}
-evalq <- function (expr, envir = parent.frame(), enclos = if (is.list(envir) || is.pairlist(envir)) parent.frame() else baseenv())  
-	.Internal(eval(substitute(expr), envir, enclos))
-
-eval.parent <- function (expr, n = 1) 
-{
-	p <- parent.frame(n + 1)
-	eval(expr, p)
-}
-
-local <- function (expr, envir = new.env()) eval.parent(substitute(eval(quote(expr), envir)))
-\end{lstlisting}
-
-\c{expr} is the expression to be evaluated. It is not a string as in javascript. In order to evaluate a string, it should be first parsed using \c{parse(text = str)} to get an expression.
-
-\eval makes it possible to specify the environment in which to evaluate the expression.
-\c{enclos} is an additional environment used for lookup when \c{envir} is a list, a pairlist or a dataframe.
-In that case, \c{envir} is copied into a temporary environment enclosed in \c{enclos}.
-
-The \c{expr} argument of \eval is first evaluated in the current scope, then passed to the \eval.
-Quoting that argument prevents if from being evaluated in the current environment.
-\c{evalq} is a shortcut for \c{eval(quote(expr), ...)}.
-In the following code snippet, variable \c{r} is first looked up in the current scope with \eval whereas it is looked up in \c{env} with \c{evalq}.
+The arguments are as follows: \c{expr} is the value to be evaluated,
+\c{envir} is the environment in which \c{expr} is to be evaluated, and
+\c{enclos} is relevant when \c{envir} is a list or a data frame, then it
+specifies where R looks for objects not found in \c{envir}. This can be
+\c{NULL} (interpreted as the base package environment, \c{baseenv()}) or an
+environment.
+
+The semanitcs of \eval is to evaluate \c{expr} in the environment specified
+by \c{envir} and return the computed value. If \c{envir} is not specified,
+then the default is \c{parent.frame()} (the environment where the call to
+\eval was made).  Values to be evaluated can be of types \c{call} (AST for
+function calls) or \c{expression} (AST nodes for arbitrary expressions) or
+\c{name} (when the symbol is looked up in the current scope and its binding
+is evaluated), a promise or any of the basic types such as vectors,
+functions and environments (which are returned unchanged).
+
+The \c{evalq} form is equivalent to \c{eval(quote(expr), ...)}. eval
+evaluates its first argument in the current scope before passing it to the
+evaluator: evalq avoids this.
+
+\c{eval.parent(expr, n)} is a shorthand for \c{eval(expr, parent.frame(n))}.
+
+If envir is a list (such as a data frame) or pairlist, it is copied into a
+temporary environment (with enclosure enclos), and the temporary environment
+is used for evaluation. So if expr changes any of the components named in
+the (pair)list, the changes are lost.
+
+If envir is NULL it is interpreted as an empty list so no values could be
+found in envir and look-up goes directly to enclos.
+
+\c{local} evaluates an expression in a local environment. It is equivalent to
+evalq except that its default argument creates a new, empty
+environment. This is useful to create anonymous recursive functions and as a
+kind of limited namespace feature since variables defined in the environment
+are not visible from the outside.
+
+The \c{expr} argument of \eval is first evaluated in the current scope, then
+passed to the \eval. Quoting that argument prevents if from being evaluated
+in the current environment. \c{evalq} is a shortcut for \c{eval(quote(expr),
+  ...)}. In the following code snippet, variable \c{r} is first looked up in
+the current scope with \eval whereas it is looked up in \c{env} with
+\c{evalq}.
 
@@
 
-\eval will return unchanged basic types such as vectors, functions, strings and environments.
-Promises are forced by \eval.
-Symbols are looked up in the current environment.
-Calls and expressions are evaluated.
+\eval will return unchanged basic types such as vectors, functions, strings
+and environments. Promises are forced by \eval. Symbols are looked up in the
+current environment. Calls and expressions are evaluated.
 
@@
 \CorpusKaggle kernels and extracting the R code from the various
-formats\footnote{We use \texttt{rmarkdown} package to convert from Jupyer
+formats\footnote{We use \c{rmarkdown} package to convert from Jupyer
   notebook to Rmarkdown as well as to convert from Rmarkdown to plain R}, we

### 2020-08-28 inst/analysis/analysis_stable.Rmd
Consistence

@@
 params:
+  #base_dir: /home/pierre/Documents/Articles/ISCE2021
   base_dir: /var/lib/R/project-evalR
@@
 ast_sizes <- read_fst(path(RUN_DIR, "package-evals-traced.3/summarized-ast-size.fst")) %>% as_tibble() %>% 
-      filter(eval_source_type != "<undefined>")
+      filter(eval_source_type != "<undefined>") %>% 
+      rename(dataset = eval_source_type)
 ```
@@
 ast_sizes %>%
-  ggplot(aes(x = eval_source_type, y = expr_resolved_nodes, fill = eval_source_type, weight = nb_ev_calls), na.rm = TRUE) +
+  ggplot(aes(x = dataset, y = expr_resolved_nodes, fill = dataset, weight = nb_ev_calls), na.rm = TRUE) +
   geom_violin(trim = FALSE, width=.5)+
-  geom_boxplot(width=.05) +
+  geom_boxplot(data = slice_sample(ast_sizes, n = 100000), width=.03) + # too slow otherwise
   scale_y_log10(labels = scales::comma) +
@@
 
+The maximum number of types is `r max(expr_resolved_polymorphism$polym_degree)`.
+
 ```{r}
-eval_calls %>% semi_join(expr_resolved_polymorphism %>% filter(polym_degree == max(polym_degree)), by ="eval_call_srcref")
+highly_polymorphic <- eval_calls %>% semi_join(expr_resolved_polymorphism %>% filter(polym_degree == max(polym_degree)), by ="eval_call_srcref")
+```
+
+```{r}
+highly_polymorphic %>% count(caller_function, wt = nb_ev_calls, sort = TRUE)
 ```

### 2020-08-28 inst/paper/main.tex
Consistence

--- 2020-08-28 inst/paper/main.tex
@@
 different from today. Their work captured very few libraries and none of the
-more recent Node.js ecosystem.  More details about dynamic analysis of
-JavaScript can be found in~\cite{liang}.
-
-Wang et al.~\cite{wang} analyzed usage of dynamic features in 18 Python
-programs include dynamic code generation to find if they affects file
-change-proneness.  They reported that files with dynamic features are
-significantly more likely to be the subject of changes, than other files.
-Chen et al. looked at the correlation between code changes and dynamic
-features, including \eval, in 17 Python programs~\cite{chen}. They did not
-observe many uses of \eval.  Callau et al. performed an empirical study of
-the usage of dynamic features in the 1,000 largest Smalltalk projects in the
-Squeaksource repository, accounting for more than 4 million lines of
-code~\cite{oscar}. While \eval itself is not present, Smalltalk has a rich
-reflective interface that gives programmer great flexibility. Their work was
-based on a static analysis.
-
-Morandat et al.~\cite{ecoop12} had a short section on the usage of \eval in
-R. They found the \eval function to be widely used in R code with 8500 call
-sites in CRAN and 2.2 million dynamic calls. The 15 most frequent call sites
-account for 88\% of eval. The \c{match.arg} function is the highest user
-with 54\% of all calls. In the other call sites, they saw two uses
-cases. The most common is the evaluation of the source code of a promise
-retrieved by \c{substitute} in a new environment. This is done in the
-\c{with} function. The other use case is the invocation of a function whose
-name or arguments are determined dynamically. For this purpose, R provides
-\c{do.call} and thus using \eval is overkill.
+more recent Node.js ecosystem.
 
 \section{The Design of Eval}
-
-The R Project is a tool for data analysis.  At its heart is a {vectorized,
-  dynamic, lazy, functional, object-oriented} programming language with a
-rather unusual combination of features~\cite{ecoop12} designed to ease
-learning by non-programmer and enable rapid development of statistical
-methods~\cite{R96}.  The language was designed as a successor to
-S~\cite{S88}.
-
-In R, most data types are vectorized. Values are constructed by the
-\c{c(...)} function: \c{c("hi","ho")} creates a vector of two strings. The
-language does not differentiate scalars from vectors. To enable equational
-reasoning, R copies values accessible through multiple variables when they
-are written to.  Values can be tagged by user-defined attributes. For
-instance, one can attach the attribute \c{dim} to the value
-\c{x<-c(1,2,3,4)} by evaluating \c{attr(x,"dim")<-c(2,2)}.  Once done,
-arithmetic functions will treat \c x as a \c{2x2} matrix. Another attribute
-is \c{class} which can be bound to a list of names. For instance,
-\c{class(x)<-"human"}, sets the class of \c{x} to \c{human}.  Attributes are
-used for object-oriented dispatch.  Every linguistic construct is desugared
-to a function call, even control flow statements, assignments, and
-bracketing. Furthermore, all functions can be redefined in user code. This
-makes R both flexible and challenging to compile~\cite{dls19}.
-
-Arguments to user-defined functions are bundled into a thunk called a
-promise. Logically, a promise combines an expression's code, its
-environment, and its value.  To access the value of a promise, one must
-force it. Forcing a promise triggers evaluation and the computed value is
-captured for future reference.
-
-R supports reflection and meta-programming. The \c{substitute(e,envir)}
-function yields the parse tree of the expression \c{e} after performing
-substitutions defined by the bindings in \c{envir}.
-
-\begin{lstlisting}
-> substitute(expression(a + b), list(a = 1)))  
-expression(1 + b)
-\end{lstlisting}
-\vspace{2mm}
-
-\noindent
-R allows programmatic manipulation of parse trees, which are themselves
-first class objects. They are evaluated using the \c{eval(e,envir)}
-function. Environment on the call stack can also be accessed.
 
@@
 > env <- new.env()
-> env\$r <- 9
+> env$r <- 9
 > eval(r, envir = env)
@@
 size of the \c{expr} argument to \eval. The majority of inputs across core
-and packages (\NbAstOnePercent of \eval calls and \NbAstOneCallSitePercent of call sites) is 1. That is to say, the code being evaluated
+and packages (\NbAstOnePercent of \eval calls and \NbAstOneCallSitePercent of call sites) have size 1. That is to say, the code being evaluated
 consists of a single variable name or constant. A function call needs at
-least three nodes. Inputs smaller than \AstSizeNineFive nodes account for 95\% of all
-calls. Sizes are roughly similar between core and packages.
+least three nodes and a list such as \c{c(1, 2, 3)} counts for 4 nodes. Inputs smaller than \AstSizeNineFive nodes account for 95\% of all
+calls. Sizes in core are larger than in packages. It is due to core
 
@@
 By default, it is equal to \c{parent.frame()}, 
-which is the parent environment of \c{eval} call.This default argument
+which is the parent environment of \c{eval} call. This default argument
 happens in \DefaultEnvirExprPercent of the eval calls.
@@
 mutation. However, it is still possible to mutate the extended environment
-using the \c{\<\<-} function; but, that happens rarely.
+using the \c{\<\<-} or \c{assign} functions; but, that happens rarely.
 
@@
 
-We look at how \emph{polymorphic} the \c{expr} argument of \eval can be, \ie
-how many different types of the resolved \c{expr} there are per call sites,
-in Figure~\ref{fig:polymorphism}. Most of the call sites, \ie \PercentMonomorphic,  are
-\emph{consistent}. However, a few ones are \emph{polymorphic}, such as:
-
-\begin{figure}[!h]
-	\centering
-    \includegraphics[width=\columnwidth]{polymorphism}
-    \caption{Number of different types  of \c{expr} per call sites.} \label{fig:polymorphism}
-\end{figure}
+We look at how \emph{consistent} the \c{expr} argument of \eval can be, \ie
+how many different types of the resolved \c{expr} there are per call sites. Most of the call sites, \ie \PercentMonomorphic,  are
+\emph{consistent}, and this is similar to javascript.
+However, a few ones are highly \emph{polymorphic} (10 different types).
+They are the  pipe  operators \c{\%>\%}, \c{\%<>\%} and \c{\%$\%} in package \emph{magrittr}.
+It is effectively used to compose functions on their first argument, which can be of any type.
+
+%\begin{figure}[!h]
+%	\centering
+%    \includegraphics[width=\columnwidth]{polymorphism}
+%    \caption{Number of different types  of \c{expr} per call sites.} \label{fig:polymorphism}
+%\end{figure}
 %

### 2020-08-28 inst/paper/main.tex
minor

@@
 The R Project is a tool for data analysis.  At its heart is a {vectorized,
-  dynamic, lazy, functional, object-oriented} programming language with a
-rather unusual combination of features~\cite{ecoop12} designed to ease
-learning by non-programmer and enable rapid development of statistical
+  dynamic, lazy, functional, object-oriented} programming language with an
+unusual combination of features~\cite{ecoop12} designed to ease learning by
+non-programmer and enable rapid development of statistical
 methods~\cite{R96}.  The language was designed as a successor to
@@
 
-In R, most data types are vectorized. Values are constructed by the
-\c{c(...)} function: \c{c("hi","ho")} creates a vector of two strings. The
-language does not differentiate scalars from vectors. To enable equational
-reasoning, R copies values accessible through multiple variables when they
-are written to.  Values can be tagged by user-defined attributes. For
-instance, one can attach the attribute \c{dim} to the value
-\c{x<-c(1,2,3,4)} by evaluating \c{attr(x,"dim")<-c(2,2)}.  Once done,
-arithmetic functions will treat \c x as a \c{2x2} matrix. Another attribute
-is \c{class} which can be bound to a list of names. For instance,
-\c{class(x)<-"human"}, sets the class of \c{x} to \c{human}.  Attributes are
-used for object-oriented dispatch.  Every linguistic construct is desugared
-to a function call, even control flow statements, assignments, and
-bracketing. Furthermore, all functions can be redefined in user code. This
-makes R both flexible and challenging to compile~\cite{dls19}.
-
-Arguments to user-defined functions are bundled into a thunk called a
-promise. Logically, a promise combines an expression's code, its
+Most R data types are vectorized. Values are constructed by the \c{c(...)}
+function, e.g., \c{c("hi","ho")} creates a vector of two strings.  R does
+not differentiate scalars from vectors. To enable equational reasoning, R
+copies values accessible through multiple variables when they are updated.
+Values can be tagged by user-defined attributes. For instance, one can
+attach the attribute \c{dim} to the value \c{x<-c(1,2,3,4)} with
+\c{attr(x,"dim")<-c(2,2)}.  This causes arithmetic functions to treat \c x
+as a 2x2 matrix. Another attribute is \c{class} which can be bound to a list
+of names, e.g., \c{class(x)<-"human"}. This sets the class of \c{x} to
+\c{human}, classes are used for object-oriented dispatch.  Every R
+linguistic construct is desugared to a function call, even control flow
+statements, assignments, and bracketing. Furthermore, all functions can be
+redefined in user code. This makes R both flexible and challenging to
+compile~\cite{dls19}.
+
+Arguments to user-defined functions are bundled into thunks called
+promises. Logically, a promise combines an expression's code, its
 environment, and its value.  To access the value of a promise, one must
@@
 
-The arguments are as follows: \c{expr} is the value to be evaluated,
-\c{envir} is the environment in which \c{expr} is to be evaluated, and
-\c{enclos} is relevant when \c{envir} is a list or a data frame, then it
-specifies where R looks for objects not found in \c{envir}. This can be
-\c{NULL} (interpreted as the base package environment, \c{baseenv()}) or an
-environment.
-
-The semanitcs of \eval is to evaluate \c{expr} in the environment specified
-by \c{envir} and return the computed value. If \c{envir} is not specified,
-then the default is \c{parent.frame()} (the environment where the call to
-\eval was made).  Values to be evaluated can be of types \c{call} (AST for
-function calls) or \c{expression} (AST nodes for arbitrary expressions) or
-\c{name} (when the symbol is looked up in the current scope and its binding
-is evaluated), a promise or any of the basic types such as vectors,
-functions and environments (which are returned unchanged).
-
-The \c{evalq} form is equivalent to \c{eval(quote(expr), ...)}. eval
-evaluates its first argument in the current scope before passing it to the
-evaluator: evalq avoids this.
-
-\c{eval.parent(expr, n)} is a shorthand for \c{eval(expr, parent.frame(n))}.
-
-If envir is a list (such as a data frame) or pairlist, it is copied into a
-temporary environment (with enclosure enclos), and the temporary environment
-is used for evaluation. So if expr changes any of the components named in
-the (pair)list, the changes are lost.
-
-If envir is NULL it is interpreted as an empty list so no values could be
-found in envir and look-up goes directly to enclos.
-
-\c{local} evaluates an expression in a local environment. It is equivalent to
-evalq except that its default argument creates a new, empty
-environment. This is useful to create anonymous recursive functions and as a
-kind of limited namespace feature since variables defined in the environment
-are not visible from the outside.
-
-The \c{expr} argument of \eval is first evaluated in the current scope, then
-passed to the \eval. Quoting that argument prevents if from being evaluated
-in the current environment. \c{evalq} is a shortcut for \c{eval(quote(expr),
-  ...)}. In the following code snippet, variable \c{r} is first looked up in
+Arguments are as follows: \c{expr} is the value to be evaluated, \c{envir}
+is the environment in which it is to be evaluated, and \c{enclos} is
+relevant when \c{envir} is a list or a data frame, then it specifies where R
+looks for objects not found in \c{envir}. This can be \c{NULL} (interpreted
+as the base package environment, \c{baseenv()}) or an environment.
+
+When called, \eval evaluates \c{expr} in the environment specified by
+\c{envir} and return the computed value. If \c{envir} is missing, then the
+default is \c{parent.frame()} (the environment where the call to \eval was
+made).  Values to be evaluated can be of types \c{call} (AST for function
+calls) or \c{expression} (AST nodes for arbitrary expressions), \c{name}
+(when the symbol is looked up in the current scope and its binding is
+evaluated), promise, or any of the basic types such as vectors and
+environments (which are returned unchanged).
+
+The \c{evalq} form is equivalent to \c{eval(quote(expr), ...)}.  Quoting
+that argument prevents if from being evaluated in the current environment.
+In the following code snippet, variable \c{r} is first looked up in
 the current scope with \eval whereas it is looked up in \c{env} with
@@
 
-\eval will return unchanged basic types such as vectors, functions, strings
-and environments. Promises are forced by \eval. Symbols are looked up in the
-current environment. Calls and expressions are evaluated.
+\c{eval.parent(expr, n)} is a shorthand for \c{eval(expr, parent.frame(n))}.
+\c{local} evaluates an expression in a local environment. It is equivalent
+to evalq except that its default argument creates a new, empty
+environment. This is useful to create anonymous recursive functions and as a
+kind of limited namespace feature since variables defined in the environment
+are not visible from the outside.
 

### 2020-08-28 inst/paper/main.tex
Precisions about AST size
Correct numbers for parse evals

@@
 least three nodes and a list such as \c{c(1, 2, 3)} counts for 4 nodes. Inputs smaller than \AstSizeNineFive nodes account for 95\% of all
-calls. Sizes in core are larger than in packages. It is due to core
+calls.
+Sizes in core are larger than in packages.
+Most large AST originate from library \emph{VGAM}, which is used for statistical modelling.
+It uses \eval to evaluate large matrices with values in a given environment.
 
@@
 number of cases in which the output of \c{parse} and its variants is passed to
-\eval, directly or transitively by tainting their output. We observed that
-\PercentParsedEvals of the eval calls consume the result of calling \c{parse} on
-a file and \PasteParsedEvals of the eval calls consume the result of calling
-\c{parse} on a string. This corresponds to \PercentParsedCallSites and
-\PasteParsedCallSite of the total \eval call sites. We also identified one
+\eval, directly or transitively by tainting their output. This corresponds to \PercentParsedCallSites of the total \eval call sites and
+\PercentParsedEvals of the eval calls. We observed that
+very few of the eval call(\NbParseFilesRnd in total) consume the result of calling \c{parse} on
+a file. Most of the eval calls consume the result of calling
+\c{parse} on a string.  We also identified one
 function in core R, \c{invokeRestartInteractively} that prompts the user for

### 2020-08-28 inst/paper/main.tex
minor

@@
 
+\begin{figure*}[!t]\centering\includegraphics[width=.8\linewidth]
+{pipeline.pdf}\caption{Analysis Pipeline}\label{fig:pipeline}
+\end{figure*}
+
 Most R data types are vectorized. Values are constructed by the \c{c(...)}
@@
 
-\begin{lstlisting}
-> eval(2)
-[1] 2
-> eval(expression(1+2))
-[1] 3
-> eval(parse(text = "1 + 2"))
-[1] 3
-\end{lstlisting}
+\section{Infrastructure}
 
 \begin{figure*}[!tb]\centering\includegraphics[width=.8\linewidth]
-{pipeline.pdf}\caption{Analysis Pipeline}\label{fig:pipeline}
+  {corpus.pdf}\caption{CRAN packages}\label{fig:corpus}
 \end{figure*}
 
-\section{Infrastructure}
-
-Our infrastructure consists of a dynamic analysis pipeline to record the calls
-to \eval and RMarkdown notebooks for data analysis and visualization.
-
-Figure~\ref{fig:pipeline} shows the main steps of the pipeline. We provide the
-runtime, input and output data size for each step.
-
-The pipeline begins with downloading and installing open source R packages from
-CRAN for the corpus along with their dependencies. The Kaggle programs are
-downloaded from \href{http://www.kaggle.com}{Kaggle} separately by a crawler.
-Next, code size and coverage metrics from the installed packages are extracted
-using the \href{ https://github.com/r-lib/covr}{\covr} package. This is followed
-by extraction of runnable programs from R packages: tests, examples and
-vignettes as well as from Kaggle. The code of each extracted program is wrapped
-into a call to dynamic analyzer \emph{evil} (\emph{Ev}al \emph{I}nspection
-\emph{L}ibrary) that collect eval usage. The reason for this is that we only
-want to record eval usage for the extracted code. Without this, the data would
-include eval calls from the unit testing frameworks as well as from
-bootstrapping R virtual machine itself. To avoid any interference, each program
-is run in its own R instance. The \emph{evil} framework implemented as R package
-in 2K lines of R and 400 lines of C++ code.
-
-The dynamic analyzer builds upon our dynamic analysis framework, \instrumentr
+Our infrastructure consists of a dynamic analysis pipeline to record the
+calls to \eval and RMarkdown notebooks for data analysis and visualization.
+Figure~\ref{fig:pipeline} shows the main steps of the pipeline. We provide
+the runtime, input and output data size for each step. All runs are done on
+a single server Intel Xeon 6140, 2.30GHz with 72 cores and 256GB of RAM.
+
+The pipeline begins with downloading and installing open source R packages
+from CRAN for the corpus along with their dependencies. The Kaggle programs
+are downloaded from \href{http://www.kaggle.com}{Kaggle} separately by a
+crawler.  Next, code size and coverage metrics from the installed packages
+are extracted using the \href{ https://github.com/r-lib/covr}{\covr}
+package. This is followed by extraction of runnable programs from R
+packages: tests, examples and vignettes as well as from Kaggle. The code of
+each extracted program is wrapped into a call to dynamic analyzer
+\emph{evil} (\emph{Ev}al \emph{I}nspection \emph{L}ibrary) that collect eval
+usage. The reason for this is that we only want to record eval usage for the
+extracted code. Without this, the data would include eval calls from the
+unit testing frameworks as well as from bootstrapping R virtual machine
+itself. To avoid any interference, each program is run in its own R
+instance. The \emph{evil} framework implemented as R package in 2K lines of
+R and 400 lines of C++ code.
+
+The dynamic analyzer builds upon the dynamic analysis framework, \instrumentr
 that we have implemented to enable us to write dynamic analysis logic in R. It
@@
 parallel~\cite{GNUparallel} and orchestrated by GNU make. To schedule and
-parallelize extraction and analysis of programs, we use our homegrown library
-\runr. Furthermore, \runr gracefully handles and reports failures across
-large-scale program runs which greatly aids debugging of the analysis pipeline.
-The data extracted by \evil from each program is concatenated, cleaned and
-summarized in the post-processing phase by custom R scripts. Finally, the
-summarized data is analyzed in RMarkdown notebooks to gather insights. Apart
-from the figures, the data points included in the paper are also generated by
-RMarkdown notebooks as latex macros.
-
-All runs are done on a single server Intel Xeon 6140, 2.30GHz with 72 cores and
-256GB of RAM.
+parallelize extraction and analysis of programs, we use the \runr
+package. Furthermore, \runr gracefully handles and reports failures across
+large-scale program runs which greatly aids debugging of the analysis
+pipeline.  The data extracted by \evil from each program is concatenated,
+cleaned and summarized in the post-processing phase by custom R
+scripts. Finally, the summarized data is analyzed in RMarkdown notebooks to
+gather insights. Apart from the figures, the data points included in the
+paper are also generated by RMarkdown notebooks as latex macros.
 
@@
 
-In this study, we report on \eval usage in a corpus of \CorpusCorePackages R core
-and \CorpusPackages popular CRAN packages. To quantify the eval use, we run
-\CorpusAllProgramsRnd of R programs that are extracted form those packages as
-well as from the user-written data-analyses shared on the Kaggle platform. In
-this section we describe this corpus and methodology used to assemble it.
+In this study, we report on a corpus of \CorpusCorePackages R core and
+\CorpusPackages CRAN packages. We run \CorpusAllProgramsRnd programs
+extracted from those packages as well as from the user-written data-analyses
+shared on the Kaggle platform. In this section we describe this corpus and
+methodology used to assemble it.
 
@@
 
-We have selected the TOP \CorpusPackages packages based on the number of reverse
-dependencies, \ie how many other packages depend on them\footnote{Extracted from
-  package metadata using builtin function}. Using this criterion helps us to
-choose popular packages which in general should be of a better quality and have
-better code coverage and therefore more code to run. The resulting set contains
-\CorpusRCodeRnd lines of R code and \CorpusNativeCodeRnd lines of native code
+We have selected the TOP \CorpusPackages packages based on the number of
+reverse dependencies, \ie how many other packages depend on
+them\footnote{Extracted from package metadata using a builtin function}.
+This criterion helps us find packages which should be better written and
+have better code coverage.  The resulting set contains \CorpusRCodeRnd lines
+of R code and \CorpusNativeCodeRnd lines of native code
 (C/C++/Fortran)\footnote{Excluding comments and blank lines using
-  \url{github.com/AlDanial/cloc}}. Figure~\ref{fig:corpus} shows these packages,
-the size of the dots reflects the project's size in lines of code including both
-R and native code, the x-axis indicates the expression code coverage in percents
-and the y-axis gives the number of reverse dependencies in log scale. Dotted
-lines indicate means. Packages with over \CorpusEvalsPackageTreshold eval call
-sites are annotated.
-
-\begin{figure*}[!tb]\centering\includegraphics[width=.8\linewidth]
-  {corpus.pdf}\caption{CRAN packages}\label{fig:corpus}
-\end{figure*}
+  \url{github.com/AlDanial/cloc}}. Figure~\ref{fig:corpus} shows these
+packages, the size of the dots reflects the project's size in lines of code.
+The x-axis indicates code coverage in percents and the y-axis gives the
+number of call sites to \eval in log scale. Dotted lines indicate
+means. Packages with over \CorpusEvalsPackageTreshold eval call sites are
+named.
 
@@
 There are three sources of runnable code in a R package. Next to the package
-\emph{tests}, there are also \emph{examples} and \emph{vignettes}. Examples are
-R code snippets extracted from package documentation into scripts file.
+\emph{tests}, there are also \emph{examples} and \emph{vignettes}. Examples
+are R code snippets extracted from package documentation into scripts files.
 Vignettes are long-form description of package functionality written using
-literate-programming style in \LaTeX\xspace or Rmarkdown with chunks of R code
-that can be similarly extracted into R scripts. R provides builtin functions for
-tangling the embedded code into files.
+literate-programming style in \LaTeX\xspace or Rmarkdown with chunks of R
+code that can be similarly extracted into R scripts. R provides builtin
+functions for tangling the embedded code into files.
 

### 2020-08-28 inst/paper/main.tex
Fixing tables and figures

--- 2020-08-28 inst/paper/main.tex
@@
 
-\begin{lstlisting}
-> eval(2)
-[1] 2
-> eval(expression(1+2))
-[1] 3
-> eval(parse(text = "1 + 2"))
-[1] 3
-\end{lstlisting}
+%\begin{lstlisting}
+%> eval(2)
+%[1] 2
+%> eval(expression(1+2))
+%[1] 3
+%> eval(parse(text = "1 + 2"))
+%[1] 3
+%\end{lstlisting}
 
@@
 \EvalEventCountRnd events occur inside eval. This is just \EvalEventAllPerc of
-all events. Figure~\ref{fig:event-distribution} shows the distribution of events
+all events. Table~\ref{table:event-distribution} shows the distribution of events
 caused by \eval calls from R Core and CRAN packages.
@@
 We further investigate the type of input passed to \eval in
-Figure~\cite{fig:eval-expression-kind} by the type of input and
+Figure~\ref{fig:eval-expression-kind} by the type of input and
 the source of \eval. In this figure, \emph{Expression} is the type of program text,

### 2020-08-28 inst/paper/main.tex
corpus

@@
 In this study, we report on a corpus of \CorpusCorePackages R core and
-\CorpusPackages CRAN packages. We run \CorpusAllProgramsRnd programs
-extracted from those packages as well as from the user-written data-analyses
-shared on the Kaggle platform. In this section we describe this corpus and
-methodology used to assemble it.
+\CorpusPackages CRAN packages with \CorpusAllCodeRnd lines of source code
+(native + R).  We run \CorpusAllProgramsRnd programs extracted from those
+packages as well as from the user-written data-analyses shared on the Kaggle
+platform. In this section we describe this corpus and methodology used to
+assemble it.
 
@@
 
-We have selected the TOP \CorpusPackages packages based on the number of
-reverse dependencies, \ie how many other packages depend on
-them\footnote{Extracted from package metadata using a builtin function}.
-This criterion helps us find packages which should be better written and
-have better code coverage.  The resulting set contains \CorpusRCodeRnd lines
-of R code and \CorpusNativeCodeRnd lines of native code
-(C/C++/Fortran)\footnote{Excluding comments and blank lines using
-  \url{github.com/AlDanial/cloc}}. Figure~\ref{fig:corpus} shows these
-packages, the size of the dots reflects the project's size in lines of code.
-The x-axis indicates code coverage in percents and the y-axis gives the
-number of call sites to \eval in log scale. Dotted lines indicate
-means. Packages with over \CorpusEvalsPackageTreshold eval call sites are
-named.
+We have selected the TOP \CorpusPackages packages based on their reverse
+dependencies, \ie how many other packages depend on them.\footnote{Extracted
+  from package metadata using a builtin function.}  This helps us find
+packages which should be better written and have better code coverage.  The
+resulting set contains \CorpusRCodeRnd lines of R code and
+\CorpusNativeCodeRnd lines of native code.\footnote{Excluding comments and
+  blank lines using \url{github.com/AlDanial/cloc}.}  For each package we
+use its runnable code to compute the package code coverage\footnote{Computed
+  using \url{ https://github.com/r-lib/covr}}. On average, across the
+\CorpusPackages packages, it is \CorpusMeanExprCoverage.
+Figure~\ref{fig:corpus} shows these packages, the size of the dots reflects
+the project's size in lines of code.  The x-axis indicates code coverage in
+percents and the y-axis gives the number of call sites to \eval in log
+scale. Dotted lines indicate means. Packages with over
+\CorpusEvalsPackageTreshold eval call sites are named.
 
@@
 There are three sources of runnable code in a R package. Next to the package
-\emph{tests}, there are also \emph{examples} and \emph{vignettes}. Examples
-are R code snippets extracted from package documentation into scripts files.
+\emph{tests},\footnote{We had to exclude most tests because the {testthat}
+  harness uses \eval and thus causes the entire test to register as an \eval
+  call.} there are also \emph{examples} and \emph{vignettes}. Examples are R
+code snippets extracted from package documentation into scripts files.
 Vignettes are long-form description of package functionality written using
@@
 code that can be similarly extracted into R scripts. R provides builtin
-functions for tangling the embedded code into files.
-
-The selected packages contain \CorpusPackagePrograms programs with
-\CorpusPackageProgramsCodeRnd lines of R code:
-%
-\begin{compactitem}[$-$]
-  \item \CorpusExamplesProgramsRnd examples with
-  \summaryrnd{CorpusExamplesCode}\footnote{This notation represents $sum$ ($min$
-    / $mean$ / $max$)} lines of code,
-  \item \CorpusVignettesProgramsRnd vignettes with
-  \summaryrnd{CorpusVignettesCode} lines of code.
-\end{compactitem}
-
-For each package we use its runnable code to compute the package code
-coverage\footnote{Computed using \url{ https://github.com/r-lib/covr}}. On
-average, across the \CorpusPackages packages, it is \CorpusMeanExprCoverage.
+functions for tangling the embedded code into files.  The selected packages
+contain \CorpusPackagePrograms programs with \CorpusPackageProgramsCodeRnd
+lines of R code: \CorpusExamplesProgramsRnd examples with
+\summaryrnd{CorpusExamplesCode}\footnote{This notation represents $sum$
+  ($min$ / $mean$ / $max$)} lines of code, and \CorpusVignettesProgramsRnd
+vignettes.%%% with \summaryrnd{CorpusVignettesCode} lines of code.
 
@@
 
-To represent user-written code in the corpus, we turned to
+To represent end-user code in the corpus, we turned to
 Kaggle\footnote{\url{https://www.kaggle.com}}, an online platform for
-data-science and machine-learning practitioner. Among others, the website allow
-people to share data-science competitions, a data-analysis problem together with
-data, for which users try to find the best solution. The solutions, called
-\emph{kernels}, are then shared back to the platform in the form of either plain
-scripts or as Rmarkdown / Jupyter notebooks written in either Julia, Python, R
-or SQLite.
-
-One of the most popular competition is about predicting passenger survival on
+data-science and machine-learning. The website allow people to share
+data-science competitions, a data-analysis problem together with data, for
+which users try to find the best solution. The solutions, called {kernels},
+are then shared back to the platform in the form of either plain scripts or
+as notebooks.  One of the most popular competition is about predicting
+passenger survival on
 Titanic\footnote{\url{https://www.kaggle.com/c/titanic}} with \CorpusKaggle
-kernels in R (over 1/4 of all available R kernels) which we used for our corpus.
+kernels in R (over 1/4 of all available R kernels) which we used for our
+corpus.
 
@@
 \CorpusKaggle kernels and extracting the R code from the various
-formats\footnote{We use \c{rmarkdown} package to convert from Jupyer
-  notebook to Rmarkdown as well as to convert from Rmarkdown to plain R}, we
-found that \CorpusDuplicatedKaggle were whole-file duplicates. From
-\CorpusRunnableKaggle kernels, \CorpusFailedKaggle failed to execute. Next to
-various runtime exceptions, common problems were parse errors and misspelled
-package names.
-
-The final set contains \CorpusFinishedKaggle kernels with
-\summaryrnd{CorpusFinishedKaggleCode} lines of R code.
-
-\subsection{Summary}
-
-Together, we have assembled a corpus of \CorpusPackages R packages and
-\CorpusFinishedKaggle user-written data-analysis scripts. Together it contains
-\CorpusAllCodeRnd lines of source code (both and R and native code). Out of
-this, \CorpusAllRunnbaleCode lines R code spread across \CorpusAllPrograms R
-scripts is the code that we run during the dynamic analysis.
+formats,\footnote{We use rmarkdown package to convert from Jupyer notebook
+  to Rmarkdown as well as to convert from Rmarkdown to plain R.} we found
+that \CorpusDuplicatedKaggle were whole-file duplicates. From
+\CorpusRunnableKaggle kernels, \CorpusFailedKaggle failed to execute. Next
+to various runtime exceptions, common problems were parse errors and
+misspelled package names.  The final set contains \CorpusFinishedKaggle
+kernels with \summaryrnd{CorpusFinishedKaggleCode} lines of R code.
 

### 2020-08-28 inst/paper/main.tex
A few examples for scopes

--- 2020-08-28 inst/paper/main.tex
@@
 They are built upon the \c{identity_transformer}, which uses \eval.
-The interpretation environment for \eval is often global here, as the string litterals are often used to print variabels at the top level.
+The interpretation environment for \eval is often global here, as the string litterals are often used to print variables at the top level.
 %%
@@
 %%
-%%The \c{imchange} function of package \c{imager} makes it possible to modify images using a dedicated formula syntax using \c{\~}.\footnote{Inspired by \c{map} in package \emph{purr}.}  Here, \eval is evaluated in \c{newenv}, which creates a new environment that inherits from \c{parent.frame()} by default.
-We classify it as 1+.
-%%
-%%
-%%\begin{lstlisting}
-%%newenv <- new.env()
-%%...
-%%fo <- parse(text=as.character(fo)[2])
-%%im[where] <- eval(fo,envir=newenv,enclos=env)
-%%\end{lstlisting}
-%%
-%%\c{adjCoef} in package \emph{actuar} find the root of an equation defined by a function \c{h} whose arguments must be named \c{x} and \c{y}.
-%%\c{h} is transformed into an auxiliary function \c{h2} that can be optimized.
+
+The \c{imchange} function of package \c{imager} makes it possible to modify images using a dedicated formula syntax using \c{\~}.\footnote{Inspired by \c{map} in package \emph{purr}.}  Here, \eval is evaluated in \c{newenv}, which creates a new environment that inherits from \c{parent.frame()} by default (classified as 1+).
+
+\begin{lstlisting}
+newenv <- new.env()
+...
+fo <- parse(text=as.character(fo)[2])
+im[where] <- eval(fo,envir=newenv,enclos=env)
+\end{lstlisting}
+
+\c{adjCoef} in package \emph{actuar} find the root of an equation defined by a function \c{h} whose arguments must be named \c{x} and \c{y}.
+\c{h} is transformed into an auxiliary function \c{h2} that can be optimized.
 Here, the list used for \c{envir} ensures the correspondance between the textual arguments of \c{h} and the arguments of \c{h2}.
-%%
-%%
-%%\begin{lstlisting}
-%%sh <- substitute(h)
-%%fcall <- paste(sh, "(x, y)")
-%%...
-%%h2 <- function(x, y)
-%%    eval(parse(text = fcall),
-%%    envir = list(x = x, y = y),
-%%    enclos = parent.frame(2))
-%%\end{lstlisting}
+
+\begin{lstlisting}
+sh <- substitute(h)
+fcall <- paste(sh, "(x, y)")
+...
+h2 <- function(x, y)
+    eval(parse(text = fcall),
+    envir = list(x = x, y = y),
+    enclos = parent.frame(2))
+\end{lstlisting}
 

### 2020-08-28 inst/analysis/analysis_stable.Rmd
<<- and assign

@@
   ggplot(aes(x = dataset, y = expr_resolved_nodes, fill = dataset, weight = nb_ev_calls), na.rm = TRUE) +
-  geom_violin(trim = FALSE, width=.5)+
+  geom_violin(trim = FALSE, width=1.7)+
   geom_boxplot(data = slice_sample(ast_sizes, n = 100000), width=.03) + # too slow otherwise
@@
 
+```
+
+```{r}
+ast_sizes %>% filter(expr_resolved_nodes > 1000) %>% count(eval_call_srcref, wt = nb_ev_calls, sort = TRUE)
 ```
@@
 
+```{r env_assign}
+eval_calls_env_assign <- eval_calls %>% filter(!is.na(expr_resolved)) %>% mutate(env_assign = pblapply(expr_resolved, is_assign_str, cl = parallel::detectCores() - 2))
+```
+
 ## Patterns
@@
                             mutate(group_function_expr = fct_collapse(function_expr2, 
-                              assignment = c("<-", "[[<-", "[<-", "$<-", "<<-", "=", "@<-"),
+                              assignment = c("<-", "[[<-", "[<-", "$<-", "<<-", "=", "@<-", "assign"),
                               slot_access = c("[", "[[", "$", "slot", "@", "::", ":::"),
@@
 
+r("nb parse files", nb_file$n)
+r("percent parse file call sites", ratio(nb_file_call_sites, nb_call_sites))
+
 ```

### 2020-08-28 inst/analysis/insights.R

@@ -303,6 +303,32 @@ is_replaceable <- function(expr)
   }
 }
 
+env_assign <- c("<<-", "assign")
+
+is_assign <- function(expr)
+{
+  if(is.call(expr))
+  {
+    function_name <- expr[[1]]
+    function_args <- expr[-1]
+    return(as.character(function_name) %in% env_assign && any(function_args, is_assign))
+  }
+  else if(is.expression(expr))
+  {
+    return(any(expr, is_assign))
+  }
+  else
+  {
+    return(FALSE)
+  }
+}
+
+is_assign_str <- function(expr)
+{
+  e <- get_expr(expr)
+  return(is_assign(e))
+}
+
 
 
 is_replaceable_str <- function(expr)

### 2020-08-28 inst/paper/main.tex
metrics

@@
 
-\newcommand{\eval}{\c{eval}\xspace}
-\newcommand{\parse}{\c{parse}\xspace}
-\newcommand{\source}{\c{source}\xspace}
-\newcommand{\local}{\c{local}\xspace}
-\newcommand{\unlockBinding}{\c{unlockBinding}\xspace}
-\newcommand{\substitute}{\c{substitute}\xspace}
-\newcommand{\datatable}{\c{data.table}\xspace}
-\newcommand{\mlogit}{\c{mlogit}\xspace}
-\newcommand{\mboost}{\c{mboost}\xspace}
-\newcommand{\metafor}{\c{metafor}\xspace}
-\newcommand{\lavaan}{\c{lavaan}\xspace}
-\newcommand{\mclust}{\c{mclust}\xspace}
-\newcommand{\gamlss}{\c{gamlss}\xspace}
-\newcommand{\ggproto}{\c{ggproto}\xspace}
-\newcommand{\ggplot}{\c{ggplot2}\xspace}
-
-\newcommand{\base}{\c{base}\xspace}
+\newcommand{\eval}{\c{eval}}
+\newcommand{\parse}{\c{parse}}
+\newcommand{\source}{\c{source}}
+\newcommand{\local}{\c{local}}
+\newcommand{\unlockBinding}{\c{unlockBinding}}
+\newcommand{\substitute}{\c{substitute}}
+\newcommand{\datatable}{\c{data.table}}
+\newcommand{\mlogit}{\c{mlogit}}
+\newcommand{\mboost}{\c{mboost}}
+\newcommand{\metafor}{\c{metafor}}
+\newcommand{\lavaan}{\c{lavaan}}
+\newcommand{\mclust}{\c{mclust}}
+\newcommand{\gamlss}{\c{gamlss}}
+\newcommand{\ggproto}{\c{ggproto}}
+\newcommand{\ggplot}{\c{ggplot2}}
+
+\newcommand{\base}{\c{base}}
 \renewcommand{\c}[1]{\lstinline{#1}\xspace}
@@
   blank lines using \url{github.com/AlDanial/cloc}.}  For each package we
-use its runnable code to compute the package code coverage\footnote{Computed
-  using \url{ https://github.com/r-lib/covr}}. On average, across the
+use its runnable code to compute the package code coverage.\footnote{Computed
+  using \url{ https://github.com/r-lib/covr}.} On average, across the
 \CorpusPackages packages, it is \CorpusMeanExprCoverage.
@@
 (CRAN\footnote{\url{http://cran.r-project.org}}), the largest repository of R
-code with over \CorpusAllCranRnd packages\footnote{CRAN receives about 6 new
-  package submissions a day~\cite{Ligges2017}}. Unlike other open code
+code with over \CorpusAllCranRnd packages.\footnote{CRAN receives about 6 new
+  package submissions a day~\cite{Ligges2017}.} Unlike other open code
 repositories such as GitHub, CRAN is a curated repository. Each submitted
@@
 To represent end-user code in the corpus, we turned to
-Kaggle\footnote{\url{https://www.kaggle.com}}, an online platform for
+Kaggle,\footnote{\url{https://www.kaggle.com}} an online platform for
 data-science and machine-learning. The website allow people to share
@@
 This section presents an overview of the usage of \eval in our corpus.
-Previous work reported that 82\% of JavaScript-enabled web pages used
+Richards et al. reported 82\% of JavaScript-enabled web pages using
 \eval~\cite{ecoop11}. The case for R is more striking with 100\% of the
-programs using that \eval.\footnote{In this section, we combine calls to
-  \c{eval}, \c{eval.parent}, \c{evalq} and \c{local}.}  This high number is
-explained by the fact that the R implementation, its core packages, use
-\eval extensively. In our analysis, we differentiate uses of \eval based on
-where they originate from. A {\bf core} \eval is one that occurs in the code
-of a core package.  A {\bf dependent} \eval is one that occurs in a package
-that was included by the target package. An {\bf own} \eval is one that
-occurs in the source code of the target package.
-
-Over all, we observed \AllAllCallCountRnd calls to \eval in the
-\CorpusAllProgramsRnd programs in our corpus.  Table~\ref{A} details this
-run-time information. The format is T (m/M/x) where T is the total number of
-calls observed for all runs of the corpus, m and x are the minimum and
-maximum number of calls for any given run, and M is the mean number of calls
-per run. We can observe that an average run of CRAN package will trigger 581
-{\eval}s in core package, 532 in dependent packages and 292 in the package
-itself.  The Kaggle data set has small programs, and, these programs only
+programs call the \eval function.\footnote{We combine calls to \c{eval},
+  \c{eval.parent}, \c{evalq} and \c{local}.}  This high number is explained
+by the fact that the R implementation, its core packages, use \eval
+extensively. We differentiate uses based on their origin. A {\bf core} \eval
+is one that occurs in the code of a core package.  A {\bf dependent} \eval
+occurs in a package imported by the target package. An {\bf own} \eval is
+one that occurs in the source code of the target package.
+
+Over all, we observe \AllAllCallCountRnd calls to \eval in our corpus.
+Table~\ref{A} details this run-time information. The format is T (m/M/x)
+where T is the total number of calls observed for all runs of the corpus, m
+and x are the minimum and maximum number of calls for any given run, and M
+is the mean number of calls per run. An average CRAN package run triggers
+581 {\eval} calls in core package, 532 in dependent packages and 292 in the
+CRAN package itself.  The Kaggle data set has small programs which only
 trigger \eval in core and dependent packages. Thus, an average Kaggle run
@@
 
- \begin{table*}[tb]
-   \centering
-   \begin{tabular}{r|c|c|c@{}} \hline
-     Corpus / Source & Core                              & Dependent                              & Own \\\hline
-     CRAN   & \summaryrnd{CranCoreEvalCalls}    & \summaryrnd{CranDependentEvalCalls}    & \summaryrnd{CranOwnEvalCalls} \\
-     Kaggle & \summaryrnd{KaggleCoreEvalCalls}  & \summaryrnd{KaggleDependentEvalCalls}  & \summaryrnd{KaggleOwnEvalCalls} \\\hline
-   \end{tabular}
-   \caption{Eval calls in CRAN and Kaggle}
-   \label{A}
- \end{table*}
-
-%% The following table will be replaced. {\bf <At this point separating between
-%%   the different kinds of evals is distracting. We have not argued why it is
-%%   important.>}
-
-%% \begin{table}[ht]\resizebox{\columnwidth}{!}{\begin{tabular}{l|rr|r||rr|r}\hline
-%% \multicolumn{1}{c}{Function}& \multicolumn{3}{c}{Calls} & \multicolumn{3}{c}{Callsites}\\
-%% \cline{1-1} \cline{2-4} \cline{5-7}
-%% & Core & Packages & Total & Core & Packages & Total\\\hline
-%% \c{eval} & \CoreEvalCallCountRnd & \PackageEvalCallCountRnd  & \AllEvalCallCountRnd & \CoreEvalSiteCountRnd & \PackageEvalSiteCountRnd &  \AllEvalSiteCountRnd\\
-%% \c{eval.parent} & \CoreEvalParentCallCountRnd & \PackageEvalParentCallCountRnd  & \AllEvalParentCallCountRnd & \CoreEvalParentSiteCountRnd & \PackageEvalParentSiteCountRnd & \AllEvalParentSiteCountRnd\\
-%% \c{evalq} & \CoreEvalqCallCountRnd & \PackageEvalqCallCountRnd  & \AllEvalqCallCountRnd & \CoreEvalqSiteCountRnd & \PackageEvalqSiteCountRnd & \AllEvalqSiteCountRnd \\
-%% \c{local} & \CoreLocalCallCountRnd & \PackageLocalCallCountRnd  & \AllLocalCallCountRnd & \CoreLocalSiteCountRnd & \PackageLocalSiteCountRnd & \AllLocalSiteCountRnd \\\hline
-%% Total & \CoreAllCallCountRnd & \PackageAllCallCountRnd  & \AllAllCallCountRnd & \CoreAllSiteCountRnd & \PackageAllSiteCountRnd & \AllAllSiteCountRnd \\  \hline
-%% \end{tabular} } \label{table:eval-count-summary}
-%%  \caption{Distribution of eval calls and callsites in the corpus}
-%% \end{table}
-
-%% \ref{table:top-ten-package-summary} shows the number of calls to eval
-%% made by the ten most frequent callers to eval. These packages together have
-%% \missingNumber callsites and account for \missingPercentage calls to eval.
-%% Furthermore, the top ten callsites account for \missingPercentage of package
-%% eval calls. A single callsite of \TopTenPackageNameA accounts for over
-%% \missingPercentage calls to eval.
-
-%% \begin{table}[ht]
-%%   \resizebox{\columnwidth}{!}{%
-%%     \begin{tabular}{l|rrr||l|rrr}
-%%       \hline
-%%       \multicolumn{4}{c}{}& \multicolumn{4}{c}{} \\
-%%       \cline{1-4} \cline{5-8}
-%%       Package & Callsites & Calls & Calls \% & Package & Callsites & Calls & Calls \%\\
-%%       \hline
-%%       \TopTenPackageNameA & \TopTenPackageCallsiteCountA & \TopTenPackageCallCountA &\TopTenPackageCallPercA & \TopTenPackageNameF & \TopTenPackageCallsiteCountF & \TopTenPackageCallCountF & \TopTenPackageCallPercF \\
-%%       \TopTenPackageNameB & \TopTenPackageCallsiteCountB & \TopTenPackageCallCountB &\TopTenPackageCallPercB & \TopTenPackageNameG & \TopTenPackageCallsiteCountG & \TopTenPackageCallCountG & \TopTenPackageCallPercG \\
-%%       \TopTenPackageNameC & \TopTenPackageCallsiteCountC & \TopTenPackageCallCountC &\TopTenPackageCallPercC & \TopTenPackageNameH & \TopTenPackageCallsiteCountH & \TopTenPackageCallCountH & \TopTenPackageCallPercH \\
-%%       \TopTenPackageNameD & \TopTenPackageCallsiteCountD & \TopTenPackageCallCountD &\TopTenPackageCallPercD & \TopTenPackageNameI & \TopTenPackageCallsiteCountI & \TopTenPackageCallCountI & \TopTenPackageCallPercI \\
-%%       \TopTenPackageNameE & \TopTenPackageCallsiteCountE & \TopTenPackageCallCountE &\TopTenPackageCallPercE & \TopTenPackageNameJ & \TopTenPackageCallsiteCountJ & \TopTenPackageCallCountJ & \TopTenPackageCallPercJ \\
-%%       \hline
-%%     \end{tabular}
-%%   }
-%%   \label{table:top-ten-package-summary}
-%%   \caption{Distribution of eval calls and callsites in the top ten packages}
-%% \end{table}
-
-Table \ref{table:site-package-summary} shows the results of static analysis
-for the CRAN packages in our corpus. 89 packages have a single call site,
-and one package has \miss{252} call sites of \eval. \miss{XXX} packages have
-no calls to \eval. The core packages have \miss{41} callsites of eval and
-Kaggle has none. These static numbers are under approximations as \eval may
-be called reflectively, through aliases, or passed to a higher-order function.
+%% SORRY  I HAD TO REFORMAT
+\begin{table}[tb] \centering
+ \begin{tabular}{r|c|c|c} \hline
+Source & Core    & Dependent                      & Own \\\hline
+CRAN   & 5.6M &  867K    & 259K \\
+       & (1 / 337 / 46K)& (0 / 52 / 58K ) & (0 / 15 / 15K) \\\hline
+Kaggle &  1.2M &  470K   & 0\\
+ &  (0 / 721 / 18K) & (0 / 290 / 20K) & (0 / 0 / 0)\\\hline
+\end{tabular}\caption{Eval calls in CRAN and Kaggle}\label{A}\end{table}
+
+Table~\ref{table:site-package-summary} shows the number of call sites in
+CRAN packages. 61 packages have a single call site, and one package has 142
+call sites of \eval. 60 packages have no calls to \eval. The core packages
+%%%% FIX THESE NUMBERS
+have \miss{41} callsites of eval and Kaggle has none. These numbers are
+under approximations as \eval may be called reflectively, through aliases,
+or passed to a higher-order function.
 
@@
 
-To roughly estimate the complexity of the code being evaluated,
-Fig.~\ref{fig:ast-size} shows the distribution of the abstract syntax tree
-size of the \c{expr} argument to \eval. The majority of inputs across core
-and packages (\NbAstOnePercent of \eval calls and \NbAstOneCallSitePercent of call sites) have size 1. That is to say, the code being evaluated
-consists of a single variable name or constant. A function call needs at
-least three nodes and a list such as \c{c(1, 2, 3)} counts for 4 nodes. Inputs smaller than \AstSizeNineFive nodes account for 95\% of all
-calls.
-Sizes in core are larger than in packages.
-Most large AST originate from library \emph{VGAM}, which is used for statistical modelling.
-It uses \eval to evaluate large matrices with values in a given environment.
-
-\begin{figure}[htbp]
+\begin{figure}[!b]
 	\centering
@@
 \end{figure}
+
+To roughly estimate the complexity of the code being evaluated,
+Fig.~\ref{fig:ast-size} shows the distribution of the abstract syntax tree
+size of the \c{expr} argument to \eval. The majority of inputs across core
+and packages (\NbAstOnePercent of \eval calls and \NbAstOneCallSitePercent
+of call sites) have size 1. That is to say, the code being evaluated
+consists of a single variable name or constant. A function call needs at
+least three nodes and a list such as \c{c(1, 2, 3)} counts for 4
+nodes. Inputs smaller than \AstSizeNineFive nodes account for 95\% of all
+calls. Sizes in core are larger than in packages. Most large AST originate
+from library \c{VGAM}, which is used for statistical modelling. It uses
+\eval to evaluate large matrices with values in a given environment.
+
+We compare the number of events intercepted during execution of \eval across
+the whole corpus. For this, we use the number of interpreter steps as a
+proxy for events. We observe \AllEventCountRnd events all told, and
+\EvalEventCountRnd events occur inside eval (\EvalEventAllPerc of all
+events). Table~\ref{table:event-distribution} shows the distribution of
+events caused by \eval calls.
+
+\begin{table}[ht]\resizebox{\columnwidth}{!}{%
+\begin{tabular}{lll}\hline
+Events & Core Eval Call \% & Package Eval Call \% \\ \hline
+\EventsMinRangeA--\EventsMaxRangeA & \EventsCoreEvalPercA & \EventsPackageEvalPercA \\
+\EventsMinRangeB--\EventsMaxRangeB & \EventsCoreEvalPercB & \EventsPackageEvalPercB \\
+\EventsMinRangeC+ & \EventsCoreEvalPercC & \EventsPackageEvalPercC \\ \hline
+\end{tabular}} \label{table:event-distribution}
+  \caption{Events in \eval calls}
+\end{table}
+
+We observe a very wide spread in the number of events generated by \eval
+calls. The largest number of events generated by an \eval call is
+\EventsMaxCount.  However majority of \eval calls perform up to
+\EventsMaxRangeA events suggesting that most expressions passed to \eval are
+small.  It is interesting to note that a higher proportion of Core R \eval
+calls perform side effects in the \EventsMinRangeB--\EventsMaxRangeB range.
+Only \EventsCoreEvalCountC Core R \eval calls and \EventsPackageEvalCountC
+package \eval call generate \EventsMinRangeC to \EventsMaxRangeC events.
+These \eval calls originate from statistical modeling packages such as
+\mlogit, \mboost, \metafor, \lavaan, \mclust and \gamlss.
 
@@
 to that found in JavaScript, the majority of inputs being rather small.
-
-We compare the number of events intercepted during execution of \eval across the
-whole corpus. For this, we use the number of interpreter's evaluator loops as a
-proxy for events. In the corpus, we observe \AllEventCountRnd events. Only
-\EvalEventCountRnd events occur inside eval. This is just \EvalEventAllPerc of
-all events. Table~\ref{table:event-distribution} shows the distribution of events
-caused by \eval calls from R Core and CRAN packages.
-
-\begin{table}[ht]\resizebox{\columnwidth}{!}{%
-\begin{tabular}{lll}\hline
-Events & Core Eval Call \% & Package Eval Call \% \\ \hline
-\EventsMinRangeA--\EventsMaxRangeA & \EventsCoreEvalPercA & \EventsPackageEvalPercA \\
-\EventsMinRangeB--\EventsMaxRangeB & \EventsCoreEvalPercB & \EventsPackageEvalPercB \\
-\EventsMinRangeC--\EventsMaxRangeC & \EventsCoreEvalPercC & \EventsPackageEvalPercC \\ \hline
-\end{tabular}} \label{table:event-distribution}
-\caption{Distribution of events in \eval calls across Core and Package}
-\end{table}
-
-We observe a very wide spread in the number of events generated by \eval
-calls. The largest number of events generated by an \eval call is \EventsMaxCount.
-However majority of \eval calls perform up to \EventsMaxRangeA events suggesting
-that most expressions passed to \eval are small.
-It is interesting to note that a higher proportion of Core R \eval calls perform
-side effects in the \EventsMinRangeB--\EventsMaxRangeB range.
-Only \EventsCoreEvalCountC Core R \eval calls and \EventsPackageEvalCountC
-package \eval call generate \EventsMinRangeC to \EventsMaxRangeC events.
-These \eval calls originate from statistical modeling packages such as \mlogit,
-\mboost, \metafor, \lavaan, \mclust and \gamlss.
 

### 2020-08-28 inst/paper/main.tex
Fix typos in I and II

--- 2020-08-28 inst/paper/main.tex
@@
 many dynamic languages, with some variability in expressive power and
-nomenclature. In all of its incarnation \eval allow developpers to
+nomenclature. In all of its incarnation \eval allows developers to
 dynamically extend the code of their program at run-time.  Originally, \eval
@@
 code executed by \eval.  Java and Julia are among the more restrictive
-languages. In Java, \eval must be synthesized as a combination dynamic
+languages. In Java, \eval must be synthesized as a combination of dynamic
 loading and reflection.\footnote{While Java does not have \eval built-in,
@@
 Wang et al.~\cite{wang} analyzed usage of dynamic features in 18 Python
-programs include dynamic code generation to find if they affects file
+programs include dynamic code generation to find if they affect file
 change-proneness.  They reported that files with dynamic features are
@@
 sites in CRAN and 2.2 million dynamic calls. The 15 most frequent call sites
-account for 88\% of eval. The \c{match.arg} function is the highest user
+account for 88\% of eval. The \c{match.arg} function is the highest used one
 with 54\% of all calls. In the other call sites, they saw two uses

### 2020-08-28 inst/paper/main.tex
Fix typo in V

@@
 
-The R Project is a tool for data analysis.  At its heart is a {vectorized,
+The R Project is a tool for data analysis.  At its heart it is a {vectorized,
   dynamic, lazy, functional, object-oriented} programming language with an
@@
 each extracted program is wrapped into a call to dynamic analyzer
-\emph{evil} (\emph{Ev}al \emph{I}nspection \emph{L}ibrary) that collect eval
+\emph{evil} (\emph{Ev}al \emph{I}nspection \emph{L}ibrary) that collects eval
 usage. The reason for this is that we only want to record eval usage for the
@@
 Kaggle,\footnote{\url{https://www.kaggle.com}} an online platform for
-data-science and machine-learning. The website allow people to share
+data-science and machine-learning. The website allows people to share
 data-science competitions, a data-analysis problem together with data, for

### 2020-08-28 inst/paper/main.tex
patterns

@@
 \newcommand{\missingPercentage}{\textcolor{red}{XX\%}\xspace}
-\newcommand{\missingTable}{\textcolor{red}{XXTable}\xspace}
+\newcommand{\missingTable}[1][XX]{\textcolor{red}{Table #1}\xspace}
 \newcommand{\missingGraph}{\textcolor{red}{XXGraph}\xspace}
@@
 %%
+%% \end{lstlisting}
+%%$ <- to fix emacs fontification issues
 
@@
 
-This section presents the common patterns of \eval usage in our corpus. We
-randomly selected 20 packages from the corpus and studied how they use eval.
-We have grouped those uses in the following categories.
+The JavaScript paper identified a number of frequently occurring eval
+patterns. The patterns were detected using simple regular expressions on the
+string passed to \eval. In the case of R look at the AST passed to evals.
+
+\missingTable[JS evals in R]
+
+For the rest of this section we show some of the common use of \eval in R that
+we have seen by looking at 20 randomly selected packages:
 
@@
   \subsubsection{Miscellaneous} \eval is also used in R packages to get around
-  various limitations. For example, in \datatable package, \eval is used to modify
-  a binding in the \base package of R. Bindings in the \base package are locked
-  for modification and require explicit unlocking using \unlockBinding function.
-  However, directly using this function triggers a warning during package building
-  process which statically analyzes the package for use of such restricted
-  functions. The \datatable gets around this limitation by enclosing the call to
-  \unlockBinding inside \eval.
+  various limitations. For example, in \datatable package, \eval is used to
+  modify a binding in the \base package of R. Bindings in the \base package are
+  locked for modification and require explicit unlocking using \unlockBinding
+  function. However, directly using this function triggers a
+  restricted-function-use warning during package build, preventing the package
+  to be accepted by CRAN. The \datatable gets around this limitation by
+  enclosing the call to \unlockBinding inside \eval. From the same reason, other
+  packages use \eval to access private variables and functions from their
+  dependencies.
+  % \AllValueInputEvalCallPerc migth hit on a wong usage
 

### 2020-08-28 inst/analysis/patterns.Rmd
Add pattern table

---
title: "Patterns"
output: html_document
params:
  base_dir: ../../../
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
local(for (x in c(
  "dplyr", 
  "DT",
  "fs",
  "fst",
  "ggplot2",
  "ggthemes",
  "lubridate",
  "readr",
  "purrr",
  "stringr",
  "tidyr",
  "tidytext",
  "viridis",
  "xtable"
  )) {
  suppressPackageStartupMessages(library(x, character.only=TRUE))
})

knitr::opts_chunk$set(
  echo = TRUE,
  fig.align = "center",
  fig.retina = 2,
  fig.width = 10
)

# TODO: move into the package
source("inc/paths.R")
source("inc/setup.R")
source("inc/latextags.R")

options(repos="https://cran.r-project.org")
theme_set(theme_minimal())
create_tags(path(TAGS_DIR, "patterns.tex"), prefix="Pattern", default=TRUE)
```

## Data

```{r}
cran_programs <- read_fst(CRAN_PROGRAMS_FILE) %>% as_tibble() %>% filter(success)
kaggle_programs <- read_fst(KAGGLE_PROGRAMS_FILE) %>% as_tibble() %>% filter(success)
programs <- bind_rows(
  select(cran_programs, file, corpus),
  select(kaggle_programs, file, corpus)
)
```

```{r}
core_data <- 
  read_fst(EVALS_SUM_CORE_FILE) %>%
  as_tibble() %>%
  semi_join(programs, by="file")

packages_data <- 
  read_fst(EVALS_SUM_PKGS_FILE) %>%
  as_tibble() %>%
  semi_join(programs, by="file")

# kaggle should be always empty
kaggle_data <- 
  read_fst(EVALS_SUM_KAGGLE_FILE) %>%
  as_tibble() %>%
  semi_join(programs, by="file")

all_data <- 
  bind_rows(core_data, packages_data, kaggle_data) %>%
  filter(eval_function %in% c("eval", "evalq")) %>%
  select(
    file, 
    expr=expr_resolved, 
    expr_type=expr_resolved_type, 
    expr_fun=expr_resolved_function,
    srcref=eval_call_srcref, 
    caller_expression,
    caller_function,
    caller_package,
    expr_parsed_expression, 
    eval_source_type,
    expr_expression_type,
    envir_expression,
    enclos_expression,
    environment_class,
    nb_ev_calls
  )
```

## JS Patterns

```{r}
calls <- function(x) {
  stopifnot("nb_ev_calls" %in% colnames(x))
  sum(x$nb_ev_calls)
}

ratio_calls <- function(x, y) {
  ratio(calls(x), calls(y))
}
```

### library

```{r}
p_lib <- filter(all_data, str_detect(expr, "<-\\s*function\\("))
p_lib_class <- 
  p_lib %>%
  mutate(
    reason=case_when(
      caller_expression == "eval(ei, envir)"    ~ "source", # base/R/source.R:219
      caller_function == "source"               ~ "source",
      caller_function == "sys.source"           ~ "source",
      str_starts(expr, fixed("eval(quote({\n")) ~ "local",
      TRUE                                      ~ "unknown"
    )
  )

count(p_lib_class, reason)
filter(p_lib_class, reason=="unknown") %>% arrange(desc(nb_ev_calls))
```

```{r}
overview_table(
  r("lib",        calls(p_lib_class)),
  r("lib ratio",  ratio_calls(p_lib, all_data)),
  r("lib source", calls(filter(p_lib_class, reason=="source"))),  
  r("lib local",  calls(filter(p_lib_class, reason=="local")))  
)
```

### Read

resolving variable -- `evalq(x)` or `eval(quote(x))`

```{r}
p_read <-
  all_data %>%
  filter(expr_type=="SYMSXP" | expr_fun %in% c("[", "[[", "$", "slot", "@", "::", ":::"))

p_read %>% 
  count(caller_function, sort=T) %>% print(n=Inf)

p_read_slot_access <- filter(p_read, expr_fun %in% c("[", "[[", "$", "slot", "@", "::", ":::"))

overview_table(
  r("read", calls(p_read)),
  r("read ratio", ratio_calls(p_read, all_data)),
  r("read slot access ratio", ratio_calls(p_read_slot_access, p_read)),
  r("read datatable ratio", ratio_calls(filter(p_read, caller_package=="data.table"), p_read_slot_access))
)
```

### Assign

```{r}
p_assign <-
  all_data %>%
  filter(expr_fun == "=" | str_ends(expr_fun, "<-") | expr_fun %in% c("assign", "delayedAssign"))

p_assign_arrow <- filter(p_assign, expr_fun=="<-")
p_assign_superarrow <- filter(p_assign, expr_fun=="<<-")

overview_table(
  r("assign", calls(p_assign)),
  r("assign ratio", ratio_calls(p_assign, all_data)),
  r("assign arrow ratio", ratio_calls(p_assign_arrow, p_assign)),
  r("assign arrow magrittr ratio", ratio_calls(filter(p_assign_arrow, caller_package=="magrittr"), p_assign_arrow)),
  r("assign superarrow ratio", ratio_calls(p_assign_superarrow, p_assign)),
  r("assign superarrow plyr ratio", ratio_calls(filter(p_assign_superarrow, caller_package=="plyr"), p_assign_superarrow))
)

count(p_assign, expr_fun) %>%
  rowwise() %>%
  do({
    e_f <- .$expr_fun
    e_n <- .$n
    tibble(e_f, e_n, calls=calls(filter(p_assign, expr_fun==e_f)))
  })

count(p_assign, caller_package) %>%
  rowwise() %>%
  do({
    e_p <- .$caller_package
    e_n <- .$n
    tibble(e_p, e_n, calls=calls(filter(p_assign, caller_package==e_p)))
  })
```

### typeof

```{r}
p_typeof <- 
  all_data %>%
  filter(
    expr_fun=="typeof" |
    expr_fun=="class" |
    str_starts(expr_fun, "is\\.")
  )

count(p_typeof, expr_fun) %>%
  rowwise() %>%
  do({
    e_f <- .$expr_fun
    e_n <- .$n
    tibble(e_f, e_n, calls=calls(filter(p_typeof, expr_fun==e_f)))
  })

overview_table(
  r("typeof", calls(p_typeof)),
  r("typeof ratio", ratio_calls(p_assign, all_data)),
  r("assign arrow ratio", ratio_calls(filter(p_assign, expr_fun=="<-"), p_assign)),
  r("assign superarrow ratio", ratio_calls(p_assign_superarrow, p_assign)),
  r("typeof assertthat ratio", ratio_calls(filter(p_assign_superarrow, caller_package=="plyr"), p_assign_superarrow))
)
```

### try

```{r}
p_try <- 
  all_data %>%
  filter(
    expr_fun %in% c("try", "tryCatch")
  )

overview_table(
  r("try", calls(p_try)),
  r("try ratio", ratio_calls(p_try, all_data))
)
```

mostly from methods

### empty

```{r}
p_empty <- all_data %>%
  filter(expr=="" | str_detect(expr, "^[\"']\\s*[\"']$"))

overview_table(
  r("empty", calls(p_empty)),
  r("empty ratio", ratio_calls(p_empty, all_data))
)
```

### call

```{r}
p_call <- all_data %>%
  filter(
    expr_type %in% c("LANGSXP", "EXPRSXP")
  ) %>%
  anti_join(
    bind_rows(
      p_lib, p_read, p_assign, p_typeof, p_try, p_empty
    )
  )

filter(all_data, expr_type == "EXPRSXP", !str_starts(expr, "^.*\\(")) %>% count(expr)

overview_table(
  r("call", calls(p_call)),
  r("call ratio", ratio_calls(p_call, all_data))
)
```

### other

```{r}
p_other <-
  anti_join(
    all_data,
    bind_rows(
      p_lib, p_read, p_assign, p_typeof, p_try, p_call, p_empty
    )
  )

overview_table(
  r("other", calls(p_other)),
  r("other ratio", ratio_calls(p_other, all_data))
)
```

```{r}
merge_patterns <- function(...) {
  vars <- substitute(list(...))
  vars <- as.list(vars[-1])

  map_dfr(
    vars,
    function(x) {
      df <- get(as.character(x))
      mutate(df, pattern=as.character(x))
    }
  )
}

merge_patterns_sum <- function(...) {
  vars <- substitute(list(...))
  vars <- as.list(vars[-1])

  all <- all_data %>% group_by(eval_source_type) %>% summarize(all_evals=sum(nb_ev_calls))

  map_dfr(
    vars,
    function(x) {
      name <- as.character(x)
      df <- get(name)
      if (nrow(df) == 0) {
        tibble(eval_source_type=all$eval_source_type, pattern=name, nb_ev_calls=0, ratio=0)
      } else {
        df %>%
          mutate(pattern=name) %>%
          select(eval_source_type, nb_ev_calls, pattern) %>%
          group_by(eval_source_type, pattern) %>%
          summarize_all(sum) %>%
          left_join(all, by="eval_source_type") %>%
          mutate(ratio=nb_ev_calls/all_evals) %>%
          select(-all_evals)
      }
    }
  )
}
```

```{r}
patterns <- merge_patterns(p_lib, p_read, p_assign, p_typeof, p_try, p_call, p_other)

patterns_sum <- merge_patterns_sum(p_lib, p_read, p_assign, p_typeof, p_try, p_call, p_other)
```

```{r}
patterns_sum %>%
  group_by(eval_source_type) %>%
  summarise(ratio=sum(ratio))
```

```{r}
patterns_sum %>%
  ggplot(aes(x=reorder(pattern, -ratio), y=ratio)) +
  geom_bar(stat="identity") +
  scale_y_continuous(labels = scales::percent) +
  facet_wrap(~eval_source_type)
```

```{r}
patterns_table <- 
  patterns_sum %>%
  mutate(ratio=map_chr(ratio, ~fmt(percent(.))), nb_ev_calls=map_chr(nb_ev_calls, ~fmt(oom(.)))) %>%
  pivot_wider(names_from=eval_source_type, values_from=c(nb_ev_calls, ratio), values_fill=list(nb_ev_calls="0", ratio="0%")) %>%
  left_join(
    tibble(
      name   =c("Library", "Read", "Assign",   "Typeof",   "Try",   "Call",   "Empty",   "Other"),
      pattern=c("p_lib", "p_read", "p_assign", "p_typeof", "p_try", "p_call", "p_empty", "p_other")
    ),
    by="pattern"
  ) %>%
  select(name, nb_ev_calls_core, ratio_core, nb_ev_calls_package, ratio_package)

patterns_table %>%
  arrange(desc(ratio_core)) %>%
  xtable() %>%
  print(
    only.contents=TRUE, 
    include.colnames=FALSE, 
    include.rownames=FALSE,
    hline.after=nrow(patterns_table),
    file=path(TAGS_DIR, "table-js-patterns-in-r.tex")
  )
```

```
p_assign_ast <-
  p_assign %>%
  mutate(expr_ast=map(expr, ~tryCatch(str2lang(.), error=function(e) NA)))

p_assign_ast_len <-
  p_assign_ast %>%
  mutate(expr_ast_length=map_int(expr_ast, length)) %>%
  filter(expr_ast_length==3) %>%
  mutate(fun=map_chr(expr_ast, function(x) {
    if (length(x[[1]]) == 1) as.character(x[[1]]) else NA
  })) %>%
  filter(!is.na(fun))

p_assign_arrow <- 
  filter(p_assign_ast_len, fun %in% c("<-", "<<-"))

p_assign_arrow %>% count(caller_function, sort = TRUE)

overview_table(
  r("assign", p_assign_arrow),
  r("assign ratio", ratio(p_assign_arrow, all_data)),
  r("assign super", ratio(filter(p_assign_ast_len, fun == "<<-"), p_assign_ast_len)),
  r("assign maggit", ratio(filter(p_assign_ast_len, caller_function == "%<>%"), p_assign_ast_len))
)
```

## Patterns

### Value

```{r}

```

```
filter(corpus, evals>0)$package %>% sample(20)
```

```{r}
packages <- c(
  "timeSeries",
  "irlba",
  "assertthat",
  "rgenoud",
  "taxize",
  "prodlim",
  "vars" ,
  "betareg"   ,
  "bookdown"  ,
  "R2HTML"   ,
  "withr",
  "processx" ,
  "modeltools"  ,
  "rprojroot"  ,
  "ggplot2",
  "binom",
  "rmarkdown" ,
  "R.methodsS3" ,
  "FactoMineR" ,
  "PerformanceAnalytics"
)
```

```{r}
evals_static <- read_fst(PACKAGE_EVALS_FILE)
```

### parse

```{r}
all_data %>%
  filter(!is.na(parse_args_text))
```

### timeSeries

```{r}
filter(evals_static, package=="timeSeries")
```

```
as.timeSeries.character - dynamic code loading
    # Load Demo File - Returns a data frame:
    x <- eval(parse(text = eval(x)))

setMethod("rank", "timeSeries",
  function(x,  na.last = TRUE,
      ties.method = eval(formals(rank)$ties.method)) -- incorrect usage

```

```{r}
paste0("scp -r prl3:/var/lib/R/project-evalR/CRAN/extracted/",packages," . &&") %>% cat(sep="\n")
```

### 2020-08-28 inst/paper/main.tex
Add pattern table
Fix numbers of call sites

--- 2020-08-28 inst/paper/main.tex
@@
 Table~\ref{table:site-package-summary} shows the number of call sites in
-CRAN packages. 61 packages have a single call site, and one package has 142
-call sites of \eval. 60 packages have no calls to \eval. The core packages
-%%%% FIX THESE NUMBERS
-have \miss{41} callsites of eval and Kaggle has none. These numbers are
+CRAN packages. 61 packages have a single call site, and one package, \emph{VGAM}, has \MaxCallSitePackage
+call sites of \eval. \NoCallSitePackages packages have no calls to \eval. The core packages
+have \NbCoreEvalCallSites call sites of \eval and Kaggle has none. These numbers are
 under approximations as \eval may be called reflectively, through aliases,

### 2020-08-28 inst/analysis/insights.R

@@ -311,11 +311,11 @@ is_assign <- function(expr)
   {
     function_name <- expr[[1]]
     function_args <- expr[-1]
-    return(as.character(function_name) %in% env_assign && any(function_args, is_assign))
+    return(as.character(function_name) %in% env_assign || some(function_args, is_assign))
   }
   else if(is.expression(expr))
   {
-    return(any(expr, is_assign))
+    return(some(expr, is_assign))
   }
   else
   {

### 2020-08-28 inst/paper/main.tex

--- 2020-08-28 inst/paper/main.tex
@@
 \usepackage{hyperref}
+\usepackage{multirow}
 \newcommand{\missingTag}[1]{\textcolor{red}{#1}\xspace}
@@
 %% \end{lstlisting}
-%%$ <- to fix emacs fontification issues
+%% $$$$ <- to fix emacs fontification issues
 
@@
 
-\missingTable[JS evals in R]
+\begin{table}[ht]%\resizebox{\columnwidth}{!}{%
+  \centering
+\begin{tabular}{r|r|r|r|r}\hline
+\multirow{2}{*}{Pattern} & \multicolumn{2}{c|}{Core} & \multicolumn{2}{c}{Packages} \\
+                         & Calls & Ratio            & Calls & Ratio \\\hline
+\input{tag/table-js-patterns-in-r.tex}
+\end{tabular}
+\caption{JavaScript evals patterns in R}
+\label{table:js-pattens-in-r}
+\end{table}
 

### 2020-08-28 inst/paper/main.tex
AST size and VGAM
Fix introduction and background

@@
   features of the language. This paper is a large-scale study of a corpus of
-  \CorpusAllCodeRnd lines of data science libraries and end-user written in
+  \CorpusAllCodeRnd lines of data science libraries and end-user code written in
   the R language. We performed a sophisticated dynamic analysis to confirm
-  that \eval is indeed in wide spread use, and to categorized patters of
+  that \eval is indeed in wide spread use, and categorized patterns of
   usage. We found that \eval in this code base is more dangerous in some
@@
 base. JavaScript was designed to run untrusted code in browser, while R was
-design for statistical computing on a desktop. JavaScript is used as a
+designed for statistical computing on a desktop. JavaScript is used as a
 general purpose language by a wide community of programmers; while R is used
-for scientific computing by data scientist and domain experts with, often,
+for scientific computing by data scientists and domain experts with, often,
 limited programming experience.
@@
 
-To understand the behavior of eval we perform a dynamic analysis and manual
+To understand the behavior of \eval we perform a dynamic analysis and manual
 code inspection. The results presented here summarize our findings.
@@
 Our infrastructure is publicly available, released in open source, and our
-data, and code will be submitted to artifact evaluation.
+code, and data will be submitted for artifact evaluation.
 
@@
   call the new method.} Julia provides \eval more directly but limits its
-scope to top level variables and methods. JavaScript support both a,
-so-called, global \eval, and a local \eval. Where the latter has acces to
+scope to top level variables and methods. JavaScript supports both a,
+so-called, global \eval, and a local \eval, where the latter has acces to
 local variables of the function being executed (as well as variables of
-lexically enclosing scopes). Finally languages such as R allow \eval to
-manipulate variables defined in the caller environment of the function in
+lexically enclosing scopes). Finally, languages such as R allow \eval to
+even manipulate variables defined in the caller environment of the function in
 which \eval is invoked.
@@
 10,000 most popular websites. They used an instrumented web browser to
-gather execution traces and perform a dynamic analysis.  Their results show
+gather execution traces and performed a dynamic analysis.  Their results show
 that \eval is pervasive with 82\% of the most popular websites using it. The
@@
 pages.  While many uses \eval were legitimate, many were unnecessary and
-could be replaced with equivalent and safer code.  The reported on a
+could be replaced with equivalent and safer code.  They reported on a
 categorization of inputs to \eval that covers the vast majority of input
 strings.  Restricting themselves to \eval in which all named variables refer
-to the global scope, many patterns can be replaced by more disciplined
+to the global scope, many patterns could be replaced by more disciplined
 code~\cite{oopsla12b}.  The limitation of that work are that there was no
 measure of code coverage, so the numbers presented can be a lower bound on
-the possible behaviors.  Furthermore in 2011, JavaScript usage is quite
+the possible behaviors.  Furthermore in 2011, JavaScript usage was quite
 different from today. Their work captured very few libraries and none of the
@@
 features, including \eval, in 17 Python programs~\cite{chen}. They did not
-observe many uses of \eval.  Callau et al. performed an empirical study of
+observe many uses of \eval.  Callau et al.~\cite{oscar} performed an empirical study of
 the usage of dynamic features in the 1,000 largest Smalltalk projects in the
@@
 calls. Sizes in core are larger than in packages. Most large AST originate
-from library \c{VGAM}, which is used for statistical modelling. It uses
+from library \c{VGAM}, which is used for statistical modeling, and also calls functions using \eval from core. It uses
 \eval to evaluate large matrices with values in a given environment.

### 2020-08-28 inst/paper/main.tex
Minor fixes

--- 2020-08-28 inst/paper/main.tex
@@
 itself. To avoid any interference, each program is run in its own R
-instance. The \emph{evil} framework implemented as R package in 2K lines of
+instance. The \emph{evil} framework is implemented as a R package in 2K lines of
 R and 400 lines of C++ code.
@@
 the project's size in lines of code.  The x-axis indicates code coverage in
-percents and the y-axis gives the number of call sites to \eval in log
+percents and the y-axis gives the number of call sites to \eval that were traced, in log
 scale. Dotted lines indicate means. Packages with over

### 2020-08-28 inst/paper/main.tex
Fix R and the Design of Eval

@@
 unusual combination of features~\cite{ecoop12} designed to ease learning by
-non-programmer and enable rapid development of statistical
+non-programmers and enable rapid development of statistical
 methods~\cite{R96}.  The language was designed as a successor to
@@
 \c{attr(x,"dim")<-c(2,2)}.  This causes arithmetic functions to treat \c x
-as a 2x2 matrix. Another attribute is \c{class} which can be bound to a list
+as a $2 \times 2$ matrix. Another attribute is \c{class} which can be bound to a list
 of names, e.g., \c{class(x)<-"human"}. This sets the class of \c{x} to
-\c{human}, classes are used for object-oriented dispatch.  Every R
+\c{human}; classes are used for object-oriented dispatch.  Every R
 linguistic construct is desugared to a function call, even control flow
@@
 
-R supports reflection and meta-programming. The function
-\c{substitute(e,envir)} yields the parse tree of the expression \c{e} after
-performing substitutions defined by the bindings in \c{envir}.
+R supports reflection and meta-programming. The function \c{substitute(e,envir)}
+yields the parse tree of the expression \c{e} after performing substitutions
+defined by the bindings in \c{envir}.
 

### 2020-08-28 inst/paper/main.tex
Minor

--- 2020-08-28 inst/paper/main.tex
@@
 have \NbCoreEvalCallSites call sites of \eval and Kaggle has none. These numbers are
-under approximations as \eval may be called reflectively, through aliases,
+under-approximations as \eval may be called reflectively, through aliases,
 or passed to a higher-order function.
@@
 nodes. Inputs smaller than \AstSizeNineFive nodes account for 95\% of all
-calls. Sizes in core are larger than in packages. Most large AST originate
+calls. Sizes in core are larger than in packages. Most large ASTs originate
 from library \c{VGAM}, which is used for statistical modeling, and also calls functions using \eval from core. It uses
@@
 calls. The largest number of events generated by an \eval call is
-\EventsMaxCount.  However majority of \eval calls perform up to
+\EventsMaxCount.  However, a majority of \eval calls performs up to
 \EventsMaxRangeA events suggesting that most expressions passed to \eval are
@@
 science code than it was in JavaScript. Any run of an R program seems to
-trigger \eval, more over package programmers use \eval liberally. On the
+trigger \eval; moreover package programmers use \eval liberally. On the
 other hand, the authors of Kaggle packages do not use the feature at all.
-This point to a bifurcated user community, experts who write package know
+This point to a bifurcated user community: experts who write package know
 how to use \eval while end-users do not. An alternative explanation is that

### 2020-08-28 inst/analysis/counts.Rmd
fixed table 2

@@
   x <- tibble(
-    eval_callsite_table[i:(i+eval_callsite_table_row), "site_count"],
-    eval_callsite_table[i:(i+eval_callsite_table_row), "package_count"]
+    eval_callsite_table[i:(i+eval_callsite_table_row-1), "site_count"],
+    eval_callsite_table[i:(i+eval_callsite_table_row-1), "package_count"]
   )

### 2020-08-28 inst/paper/main.tex
more patterns

@@
 \input{tag/enrichDataset.tex}
+\input{tag/patterns.tex}
 
@@
 
-The JavaScript paper identified a number of frequently occurring eval
-patterns. The patterns were detected using simple regular expressions on the
-string passed to \eval. In the case of R look at the AST passed to evals.
+The JavaScript paper identified a number of frequently occurring \eval patterns
+which were detected using simple regular expressions on the \eval strings. In
+the case of R we match the patterns on the recorded AST.
+
+Table~\ref{table:js-pattens-in-r} shows the ratio of \eval in each of these
+patterns. We included all the original patterns except for \emph{JSON} as R uses
+its own serialization format and \emph{empty} as there were no empty expression
+passed to \eval:
 
@@
 
-For the rest of this section we show some of the common use of \eval in R that
-we have seen by looking at 20 randomly selected packages:
-
-\subsubsection{Evaluation using Data Fields} A common use of \eval is to
-  evaluate expressions that access data frame columns or list (heterogeneous
-  vectors) elements by name. Core R provides two functions, \c{with} and
-  \c{within} that implement this pattern. These functions construct a temporary
-  environment with data frame columns or list elements as the bindings.
-  Side-effects performed by the evaluated expression take place in this temporary
-  environment and are ignored by the \c{with} function. The \c{within} function
-  copies these modifications to a local copy of the input data and returns it.
-  This pattern is used quite often for statistical modeling. The modeling
-  expressions is constructed from variable names whose values are supplied by a
-  data frame or list.
+%\noindent
+%\begin{compactitem}[$-$]
+  % \item \emph{Library}: in JavaScript, the library pattern captures asynchronous
+  % loading of external external function definitions. In R we have seen only
+  % \PatternLibRnd such calls (\PatternLibRatio) mostly coming from source loading
+  % functions (\ie \c{source} and \c{sys.source}). Most of the external function
+  % definitions in R come from packages and R makes it very easy for users to
+  % install packages from its centralized repository.
+
+\noindent \emph{Library}, \emph{Typeof} and \emph{Try} are all barely used
+mostly because the language provides dedicated features for it. The few calls in
+Library pattern come mostly from file sourcing functions (\ie \c{source} and \c{sys.source}).
+%
+\noindent \emph{Read} pattern in JavaScript represented mostly an incorrect
+access to an object's property or local variable both replaceable by JavaScript
+core functions. In the case of R there are \PatternReadRnd \eval calls
+(\PatternReadRatio). They come from calls like \c{subset(airquality, select =
+  Temp)} where the variable \c{Temp} will be evaluated in the \c{airquality}
+data frame.
+%
+\noindent \emph{Assign} category comprises of assignments to variables within
+an \eval call. We have seen \PatternAssignRnd such calls (\PatternAssignRatio).
+\PatternAssignArrowRatio uses the simple assignment (\c{<-}). The majority of
+cases comes from the \c{magrittr}'s \c{\%<>\%} pipe operator
+(\PatternAssignArrowMagrittrRatio) which forwards an object to a function on rhs
+and update lhs with the result. Far more calls used the super assignment
+(\c{<<-}) operator, but almost all come from the \c{plyr} package which provides
+a convenient API for common data manipulation problems.
+%
+\noindent \emph{Call} and \emph{Other} together represents the vast majority of
+eval calls in R. The \emph{Other} pattern represents all the calls where eval
+was called with a value instead of with an expression. The \emph{Call} pattern
+contains all the calls that did not fall in the previous categories. Without a
+context it is usually difficult to further categorize the intent of these calls.
+Therefore we have randomly chosen \PatternManualPackages packages and manually
+looked at \PatternManualCallsites call sites from which we have observed the following categories:
+
+% this is read pattern
+% \subsubsection{Evaluation using Data Fields} A common use of \eval is to
+%   evaluate expressions that access data frame columns or list (heterogeneous
+%   vectors) elements by name. Core R provides two functions, \c{with} and
+%   \c{within} that implement this pattern. These functions construct a temporary
+%   environment with data frame columns or list elements as the bindings.
+%   Side-effects performed by the evaluated expression take place in this temporary
+%   environment and are ignored by the \c{with} function. The \c{within} function
+%   copies these modifications to a local copy of the input data and returns it.
+%   This pattern is used quite often for statistical modeling. The modeling
+%   expressions is constructed from variable names whose values are supplied by a
+%   data frame or list.
+\subsubsection{Reconding the current call} R provides a \c{match.call} function
+which can return the current call. This is heavily used by the statistical
+model-fitting functions that records the current call that is then reevaluated
+in the different environment. Another pattern is to match the current call,
+adjust some arguments and pass it to another function. For example the
+\c{write.csv} function changes the \c{sep} argument in the matched call and reevaluate itself as \c{write.table}:
+%
+\begin{lstlisting}
+    Call <- match.call(expand.dots = TRUE)
+    Call$sep <- ","; Call[[1L]] <- as.name("write.table")
+    eval.parent(Call)
+\end{lstlisting}
 

### 2020-08-28 inst/paper/main.tex
fix

@@
 
-\input{tag/counts.tex}
+%\input{tag/counts.tex}
+\newcommand{\CoreEvalCallCountRnd}{6.3M\xspace}
+\newcommand{\CoreEvalCallCount}{6,282,429\xspace}
+\newcommand{\CoreEvalSiteCountRnd}{185\xspace}
+\newcommand{\CoreEvalSiteCount}{185\xspace}
+\newcommand{\CoreEvalParentCallCountRnd}{186K\xspace}
+\newcommand{\CoreEvalParentCallCount}{186,275\xspace}
+\newcommand{\CoreEvalParentSiteCountRnd}{13\xspace}
+\newcommand{\CoreEvalParentSiteCount}{13\xspace}
+\newcommand{\CoreEvalqCallCountRnd}{132K\xspace}
+\newcommand{\CoreEvalqCallCount}{132,891\xspace}
+\newcommand{\CoreEvalqSiteCountRnd}{20K\xspace}
+\newcommand{\CoreEvalqSiteCount}{20,460\xspace}
+\newcommand{\CoreLocalCallCountRnd}{175K\xspace}
+\newcommand{\CoreLocalCallCount}{175,079\xspace}
+\newcommand{\CoreLocalSiteCountRnd}{920\xspace}
+\newcommand{\CoreLocalSiteCount}{920\xspace}
+\newcommand{\PackageEvalCallCountRnd}{1.6M\xspace}
+\newcommand{\PackageEvalCallCount}{1,582,738\xspace}
+\newcommand{\PackageEvalSiteCountRnd}{1.5K\xspace}
+\newcommand{\PackageEvalSiteCount}{1,459\xspace}
+\newcommand{\PackageEvalParentCallCountRnd}{13K\xspace}
+\newcommand{\PackageEvalParentCallCount}{13,230\xspace}
+\newcommand{\PackageEvalParentSiteCountRnd}{249\xspace}
+\newcommand{\PackageEvalParentSiteCount}{249\xspace}
+\newcommand{\PackageEvalqCallCountRnd}{2\xspace}
+\newcommand{\PackageEvalqCallCount}{2\xspace}
+\newcommand{\PackageEvalqSiteCountRnd}{1\xspace}
+\newcommand{\PackageEvalqSiteCount}{1\xspace}
+\newcommand{\PackageLocalCallCountRnd}{2.2K\xspace}
+\newcommand{\PackageLocalCallCount}{2,228\xspace}
+\newcommand{\PackageLocalSiteCountRnd}{15\xspace}
+\newcommand{\PackageLocalSiteCount}{15\xspace}
+\newcommand{\KaggleEvalCallCountRnd}{0\xspace}
+\newcommand{\KaggleEvalCallCount}{0\xspace}
+\newcommand{\KaggleEvalSiteCountRnd}{0\xspace}
+\newcommand{\KaggleEvalSiteCount}{0\xspace}
+\newcommand{\KaggleEvalParentCallCountRnd}{0\xspace}
+\newcommand{\KaggleEvalParentCallCount}{0\xspace}
+\newcommand{\KaggleEvalParentSiteCountRnd}{0\xspace}
+\newcommand{\KaggleEvalParentSiteCount}{0\xspace}
+\newcommand{\KaggleEvalqCallCountRnd}{0\xspace}
+\newcommand{\KaggleEvalqCallCount}{0\xspace}
+\newcommand{\KaggleEvalqSiteCountRnd}{0\xspace}
+\newcommand{\KaggleEvalqSiteCount}{0\xspace}
+\newcommand{\KaggleLocalCallCountRnd}{0\xspace}
+\newcommand{\KaggleLocalCallCount}{0\xspace}
+\newcommand{\KaggleLocalSiteCountRnd}{0\xspace}
+\newcommand{\KaggleLocalSiteCount}{0\xspace}
+\newcommand{\CoreAllCallCountRnd}{6.8M\xspace}
+\newcommand{\CoreAllCallCount}{6,776,674\xspace}
+\newcommand{\CoreAllSiteCountRnd}{21.6K\xspace}
+\newcommand{\CoreAllSiteCount}{21,578\xspace}
+\newcommand{\KaggleAllCallCountRnd}{0\xspace}
+\newcommand{\KaggleAllCallCount}{0\xspace}
+\newcommand{\KaggleAllSiteCountRnd}{0\xspace}
+\newcommand{\KaggleAllSiteCount}{0\xspace}
+\newcommand{\PackageAllCallCountRnd}{1.6M\xspace}
+\newcommand{\PackageAllCallCount}{1,598,198\xspace}
+\newcommand{\PackageAllSiteCountRnd}{1.7K\xspace}
+\newcommand{\PackageAllSiteCount}{1,724\xspace}
+\newcommand{\AllEvalCallCountRnd}{7.9M\xspace}
+\newcommand{\AllEvalCallCount}{7,865,167\xspace}
+\newcommand{\AllEvalSiteCountRnd}{1.6K\xspace}
+\newcommand{\AllEvalSiteCount}{1,644\xspace}
+\newcommand{\AllEvalParentCallCountRnd}{199.5K\xspace}
+\newcommand{\AllEvalParentCallCount}{199,505\xspace}
+\newcommand{\AllEvalParentSiteCountRnd}{262\xspace}
+\newcommand{\AllEvalParentSiteCount}{262\xspace}
+\newcommand{\AllEvalqCallCountRnd}{132.9K\xspace}
+\newcommand{\AllEvalqCallCount}{132,893\xspace}
+\newcommand{\AllEvalqSiteCountRnd}{20.5K\xspace}
+\newcommand{\AllEvalqSiteCount}{20,461\xspace}
+\newcommand{\AllLocalCallCountRnd}{177.3K\xspace}
+\newcommand{\AllLocalCallCount}{177,307\xspace}
+\newcommand{\AllLocalSiteCountRnd}{935\xspace}
+\newcommand{\AllLocalSiteCount}{935\xspace}
+\newcommand{\AllAllCallCountRnd}{8.4M\xspace}
+\newcommand{\AllAllCallCount}{8,374,872\xspace}
+\newcommand{\AllAllSiteCountRnd}{23.3K\xspace}
+\newcommand{\AllAllSiteCount}{23,302\xspace}
+\newcommand{\TotalFileCountRnd}{18.3K\xspace}
+\newcommand{\TotalFileCount}{18,258\xspace}
+\newcommand{\NoEvalFileCountRnd}{89\xspace}
+\newcommand{\NoEvalFileCount}{89\xspace}
+\newcommand{\NoEvalFilePerc}{0.5\%\xspace}
+\newcommand{\CoreEvalFileCountRnd}{12.3K\xspace}
+\newcommand{\CoreEvalFileCount}{12,307\xspace}
+\newcommand{\CoreEvalFilePerc}{67.4\%\xspace}
+\newcommand{\PackageEvalFileCountRnd}{0\xspace}
+\newcommand{\PackageEvalFileCount}{0\xspace}
+\newcommand{\PackageEvalFilePerc}{0\%\xspace}
+\newcommand{\AllEvalFileCountRnd}{5.9K\xspace}
+\newcommand{\AllEvalFileCount}{5,862\xspace}
+\newcommand{\AllEvalFilePerc}{32.1\%\xspace}
+\newcommand{\EightyCoreEvalFileCountRnd}{3.8K\xspace}
+\newcommand{\EightyCoreEvalFileCount}{3,769\xspace}
+\newcommand{\EightyCoreEvalFilePerc}{64.3\%\xspace}
+\newcommand{\TopTenPackageCallCountRnd}{1.4M\xspace}
+\newcommand{\TopTenPackageCallCount}{1,425,902\xspace}
+\newcommand{\TopTenPackageCallPerc}{89.2\%\xspace}
+\newcommand{\TopTenPackageSiteCountRnd}{282\xspace}
+\newcommand{\TopTenPackageSiteCount}{282\xspace}
+\newcommand{\TopTenPackageSitePerc}{16\%\xspace}
+\newcommand{\TopTenPackageNameA}{ggplot2\xspace}
+\newcommand{\TopTenPackageCallsiteCountARnd}{2\xspace}
+\newcommand{\TopTenPackageCallsiteCountA}{2\xspace}
+\newcommand{\TopTenPackageCallCountARnd}{987K\xspace}
+\newcommand{\TopTenPackageCallCountA}{987,946\xspace}
+\newcommand{\TopTenPackageCallPercA}{61\%\xspace}
+\newcommand{\TopTenPackageNameB}{magrittr\xspace}
+\newcommand{\TopTenPackageCallsiteCountBRnd}{5\xspace}
+\newcommand{\TopTenPackageCallsiteCountB}{5\xspace}
+\newcommand{\TopTenPackageCallCountBRnd}{139K\xspace}
+\newcommand{\TopTenPackageCallCountB}{139,346\xspace}
+\newcommand{\TopTenPackageCallPercB}{8.7\%\xspace}
+\newcommand{\TopTenPackageNameC}{plyr\xspace}
+\newcommand{\TopTenPackageCallsiteCountCRnd}{11\xspace}
+\newcommand{\TopTenPackageCallsiteCountC}{11\xspace}
+\newcommand{\TopTenPackageCallCountCRnd}{68K\xspace}
+\newcommand{\TopTenPackageCallCountC}{68,671\xspace}
+\newcommand{\TopTenPackageCallPercC}{4.3\%\xspace}
+\newcommand{\TopTenPackageNameD}{glue\xspace}
+\newcommand{\TopTenPackageCallsiteCountDRnd}{5\xspace}
+\newcommand{\TopTenPackageCallsiteCountD}{5\xspace}
+\newcommand{\TopTenPackageCallCountDRnd}{62K\xspace}
+\newcommand{\TopTenPackageCallCountD}{62,564\xspace}
+\newcommand{\TopTenPackageCallPercD}{3.9\%\xspace}
+\newcommand{\TopTenPackageNameE}{data.table\xspace}
+\newcommand{\TopTenPackageCallsiteCountERnd}{31\xspace}
+\newcommand{\TopTenPackageCallsiteCountE}{31\xspace}
+\newcommand{\TopTenPackageCallCountERnd}{47K\xspace}
+\newcommand{\TopTenPackageCallCountE}{47,464\xspace}
+\newcommand{\TopTenPackageCallPercE}{3\%\xspace}
+\newcommand{\TopTenPackageNameF}{statnet.common\xspace}
+\newcommand{\TopTenPackageCallsiteCountFRnd}{12\xspace}
+\newcommand{\TopTenPackageCallsiteCountF}{12\xspace}
+\newcommand{\TopTenPackageCallCountFRnd}{30K\xspace}
+\newcommand{\TopTenPackageCallCountF}{30,581\xspace}
+\newcommand{\TopTenPackageCallPercF}{1.9\%\xspace}
+\newcommand{\TopTenPackageNameG}{VGAM\xspace}
+\newcommand{\TopTenPackageCallsiteCountGRnd}{97\xspace}
+\newcommand{\TopTenPackageCallsiteCountG}{97\xspace}
+\newcommand{\TopTenPackageCallCountGRnd}{23K\xspace}
+\newcommand{\TopTenPackageCallCountG}{23,796\xspace}
+\newcommand{\TopTenPackageCallPercG}{1.5\%\xspace}
+\newcommand{\TopTenPackageNameH}{copula\xspace}
+\newcommand{\TopTenPackageCallsiteCountHRnd}{21\xspace}
+\newcommand{\TopTenPackageCallsiteCountH}{21\xspace}
+\newcommand{\TopTenPackageCallCountHRnd}{22K\xspace}
+\newcommand{\TopTenPackageCallCountH}{22,794\xspace}
+\newcommand{\TopTenPackageCallPercH}{1.4\%\xspace}
+\newcommand{\TopTenPackageNameI}{nlme\xspace}
+\newcommand{\TopTenPackageCallsiteCountIRnd}{97\xspace}
+\newcommand{\TopTenPackageCallsiteCountI}{97\xspace}
+\newcommand{\TopTenPackageCallCountIRnd}{21K\xspace}
+\newcommand{\TopTenPackageCallCountI}{21,483\xspace}
+\newcommand{\TopTenPackageCallPercI}{1.3\%\xspace}
+\newcommand{\TopTenPackageNameJ}{R6\xspace}
+\newcommand{\TopTenPackageCallsiteCountJRnd}{1\xspace}
+\newcommand{\TopTenPackageCallsiteCountJ}{1\xspace}
+\newcommand{\TopTenPackageCallCountJRnd}{21K\xspace}
+\newcommand{\TopTenPackageCallCountJ}{21,257\xspace}
+\newcommand{\TopTenPackageCallPercJ}{1.3\%\xspace}
+
 \input{tag/corpus.tex}
-\input{tag/side-effects.tex}
-\input{tag/operations.tex}
-\input{tag/events.tex}
-\input{tag/analysisStable.tex}
-\input{tag/enrichDataset.tex}
-\input{tag/patterns.tex}
+%\input{tag/side-effects.tex}
+\newcommand{\AllWritesRnd}{9.4G\xspace}
+\newcommand{\AllWrites}{9,427,046,091\xspace}
+\newcommand{\EvalWritesRnd}{1.1G\xspace}
+\newcommand{\EvalWrites}{1,065,212,013\xspace}
+\newcommand{\EvalSideEffectingWritesRnd}{45M\xspace}
+\newcommand{\EvalSideEffectingWrites}{45,701,209\xspace}
+\newcommand{\EvalSideEffectingWritesEvalPerc}{4.3\%\xspace}
+\newcommand{\EvalSideEffectingWritesAllPerc}{.5\%\xspace}
+\newcommand{\SideEffectingCoreCallPerc}{4.4\%\xspace}
+\newcommand{\SideEffectingPackageCallPerc}{7.5\%\xspace}
+
+%\input{tag/operations.tex}
+\newcommand{\AllValueInputEvalCallPerc}{17\%\xspace}
+\newcommand{\AllExpressionInputEvalCallPerc}{82\%\xspace}
+\newcommand{\AllExpressionInputEvalSitePerc}{14\%\xspace}
+\newcommand{\AllValueInputEvalSitePerc}{3\%\xspace}
+\newcommand{\AllPolymorphicInputEvalSitePerc}{82\%\xspace}
+\newcommand{\PackageEnvironmentInputEvalCallPerc}{62\%\xspace}
+\newcommand{\AllClosureCallCountRnd}{2.9G\xspace}
+\newcommand{\AllClosureCallCount}{2,886,992,072\xspace}
+\newcommand{\EvalClosureCallCountRnd}{125M\xspace}
+\newcommand{\EvalClosureCallCount}{125,469,701\xspace}
+\newcommand{\AllNativeCallCountRnd}{463M\xspace}
+\newcommand{\AllNativeCallCount}{463,803,879\xspace}
+\newcommand{\EvalNativeCallCountRnd}{31M\xspace}
+\newcommand{\EvalNativeCallCount}{31,164,219\xspace}
+
+%\input{tag/events.tex}
+\newcommand{\AllEventCountRnd}{39G\xspace}
+\newcommand{\AllEventCount}{38,991,360,760\xspace}
+\newcommand{\EvalEventCountRnd}{2G\xspace}
+\newcommand{\EvalEventCount}{1,973,736,718\xspace}
+\newcommand{\EventsMaxCountRnd}{17M\xspace}
+\newcommand{\EventsMaxCount}{17,656,694\xspace}
+\newcommand{\EvalEventAllPerc}{5\%\xspace}
+\newcommand{\EventsMinRangeARnd}{0\xspace}
+\newcommand{\EventsMinRangeA}{0\xspace}
+\newcommand{\EventsMaxRangeARnd}{100\xspace}
+\newcommand{\EventsMaxRangeA}{100\xspace}
+\newcommand{\EventsCoreEvalCountARnd}{5M\xspace}
+\newcommand{\EventsCoreEvalCountA}{5,659,053\xspace}
+\newcommand{\EventsCoreEvalPercA}{88\%\xspace}
+\newcommand{\EventsPackageEvalCountARnd}{1.5M\xspace}
+\newcommand{\EventsPackageEvalCountA}{1,514,222\xspace}
+\newcommand{\EventsPackageEvalPercA}{93\%\xspace}
+\newcommand{\EventsMinRangeBRnd}{101\xspace}
+\newcommand{\EventsMinRangeB}{101\xspace}
+\newcommand{\EventsMaxRangeBRnd}{10K\xspace}
+\newcommand{\EventsMaxRangeB}{10,000\xspace}
+\newcommand{\EventsCoreEvalCountBRnd}{769K\xspace}
+\newcommand{\EventsCoreEvalCountB}{769,341\xspace}
+\newcommand{\EventsCoreEvalPercB}{12\%\xspace}
+\newcommand{\EventsPackageEvalCountBRnd}{100K\xspace}
+\newcommand{\EventsPackageEvalCountB}{100,141\xspace}
+\newcommand{\EventsPackageEvalPercB}{6\%\xspace}
+\newcommand{\EventsMinRangeCRnd}{10K\xspace}
+\newcommand{\EventsMinRangeC}{10,001\xspace}
+\newcommand{\EventsMaxRangeCRnd}{20M\xspace}
+\newcommand{\EventsMaxRangeC}{20,000,000\xspace}
+\newcommand{\EventsCoreEvalCountCRnd}{3K\xspace}
+\newcommand{\EventsCoreEvalCountC}{2,997\xspace}
+\newcommand{\EventsCoreEvalPercC}{0\%\xspace}
+\newcommand{\EventsPackageEvalCountCRnd}{14K\xspace}
+\newcommand{\EventsPackageEvalCountC}{14,209\xspace}
+\newcommand{\EventsPackageEvalPercC}{.9\%\xspace}
+
+%\input{tag/analysisStable.tex}
+\newcommand{\UniqueCallsRnd}{1.3M\xspace}
+\newcommand{\UniqueCalls}{1,277,948\xspace}
+\newcommand{\MatchArgNumberRnd}{5M\xspace}
+\newcommand{\MatchArgNumber}{5,029,351\xspace}
+\newcommand{\MatchArgPercent}{60\%\xspace}
+\newcommand{\NbAstOnePercent}{15\%\xspace}
+\newcommand{\NbAstOneCallSitePercent}{96\%\xspace}
+\newcommand{\AstSizeNineFiveRnd}{42\xspace}
+\newcommand{\AstSizeNineFive}{42\xspace}
+\newcommand{\AstSizeNineFiveExactPercent}{95\%\xspace}
+\newcommand{\DefaultEnvirExprRnd}{1.1M\xspace}
+\newcommand{\DefaultEnvirExpr}{1,054,124\xspace}
+\newcommand{\DefaultEnvirExprPercent}{12\%\xspace}
+\newcommand{\DefaultEnvirCallSitesPercent}{9.3\%\xspace}
+\newcommand{\DefaultEnvirCoreExprPercent}{10\%\xspace}
+\newcommand{\DefaultEnvirPackagesExprPercent}{20\%\xspace}
+\newcommand{\CoreEnvClassA}{2\xspace}
+\newcommand{\CoreCountARnd}{5M\xspace}
+\newcommand{\CoreCountA}{5,079,433\xspace}
+\newcommand{\CoreProportionA}{83\%\xspace}
+\newcommand{\CoreEnvClassB}{list\xspace}
+\newcommand{\CoreCountBRnd}{229K\xspace}
+\newcommand{\CoreCountB}{229,249\xspace}
+\newcommand{\CoreProportionB}{3.8\%\xspace}
+\newcommand{\CoreEnvClassC}{3\xspace}
+\newcommand{\CoreCountCRnd}{199K\xspace}
+\newcommand{\CoreCountC}{199,497\xspace}
+\newcommand{\CoreProportionC}{3.3\%\xspace}
+\newcommand{\CoreEnvClassD}{global+\xspace}
+\newcommand{\CoreCountDRnd}{183K\xspace}
+\newcommand{\CoreCountD}{183,631\xspace}
+\newcommand{\CoreProportionD}{3\%\xspace}
+\newcommand{\CoreEnvClassE}{1+\xspace}
+\newcommand{\CoreCountERnd}{175K\xspace}
+\newcommand{\CoreCountE}{175,151\xspace}
+\newcommand{\CoreProportionE}{2.9\%\xspace}
+\newcommand{\CoreEnvClassF}{global\xspace}
+\newcommand{\CoreCountFRnd}{124K\xspace}
+\newcommand{\CoreCountF}{124,673\xspace}
+\newcommand{\CoreProportionF}{2.1\%\xspace}
+\newcommand{\CoreEnvClassG}{sys.call(sys.parent())\xspace}
+\newcommand{\CoreCountGRnd}{32K\xspace}
+\newcommand{\CoreCountG}{32,349\xspace}
+\newcommand{\CoreProportionG}{.5\%\xspace}
+\newcommand{\CoreEnvClassH}{loop\xspace}
+\newcommand{\CoreCountHRnd}{24K\xspace}
+\newcommand{\CoreCountH}{24,286\xspace}
+\newcommand{\CoreProportionH}{.4\%\xspace}
+\newcommand{\PackagesEnvClassA}{global+\xspace}
+\newcommand{\PackagesCountARnd}{943K\xspace}
+\newcommand{\PackagesCountA}{943,682\xspace}
+\newcommand{\PackagesProportionA}{74\%\xspace}
+\newcommand{\PackagesEnvClassB}{3\xspace}
+\newcommand{\PackagesCountBRnd}{71K\xspace}
+\newcommand{\PackagesCountB}{71,037\xspace}
+\newcommand{\PackagesProportionB}{5.6\%\xspace}
+\newcommand{\PackagesEnvClassC}{2\xspace}
+\newcommand{\PackagesCountCRnd}{67K\xspace}
+\newcommand{\PackagesCountC}{67,555\xspace}
+\newcommand{\PackagesProportionC}{5.3\%\xspace}
+\newcommand{\PackagesEnvClassD}{5\xspace}
+\newcommand{\PackagesCountDRnd}{49K\xspace}
+\newcommand{\PackagesCountD}{49,782\xspace}
+\newcommand{\PackagesProportionD}{3.9\%\xspace}
+\newcommand{\PackagesEnvClassE}{list\xspace}
+\newcommand{\PackagesCountERnd}{44K\xspace}
+\newcommand{\PackagesCountE}{44,194\xspace}
+\newcommand{\PackagesProportionE}{3.5\%\xspace}
+\newcommand{\PackagesEnvClassF}{2+\xspace}
+\newcommand{\PackagesCountFRnd}{32K\xspace}
+\newcommand{\PackagesCountF}{32,931\xspace}
+\newcommand{\PackagesProportionF}{2.6\%\xspace}
+\newcommand{\PackagesEnvClassG}{global\xspace}
+\newcommand{\PackagesCountGRnd}{22K\xspace}
+\newcommand{\PackagesCountG}{22,825\xspace}
+\newcommand{\PackagesProportionG}{1.8\%\xspace}
+\newcommand{\PackagesEnvClassH}{base+\xspace}
+\newcommand{\PackagesCountHRnd}{18K\xspace}
+\newcommand{\PackagesCountH}{18,538\xspace}
+\newcommand{\PackagesProportionH}{1.5\%\xspace}
+\newcommand{\NbEnclosEnvirRnd}{274K\xspace}
+\newcommand{\NbEnclosEnvir}{274,370\xspace}
+\newcommand{\EnclosEnvirPercent}{3.3\%\xspace}
+\newcommand{\NbSysCallEnvirRnd}{33K\xspace}
+\newcommand{\NbSysCallEnvir}{33,650\xspace}
+\newcommand{\SysCallEnvirPercent}{.4\%\xspace}
+\newcommand{\NbMonomorphicRnd}{23.1K\xspace}
+\newcommand{\NbMonomorphic}{23,069\xspace}
+\newcommand{\PercentMonomorphic}{99\%\xspace}
+\newcommand{\PercentParsedEvals}{.9\%\xspace}
+\newcommand{\PercentParsedCallSites}{1\%\xspace}
+\newcommand{\NbParseFilesRnd}{10\xspace}
+\newcommand{\NbParseFiles}{10\xspace}
+\newcommand{\PercentParseFileCallSites}{.01\%\xspace}
+\newcommand{\PasteParsedEvals}{39\%\xspace}
+\newcommand{\PasteParsedCallSite}{51\%\xspace}
+\newcommand{\NbCoreEvalCallSitesRnd}{310\xspace}
+\newcommand{\NbCoreEvalCallSites}{310\xspace}
+\newcommand{\MaxCallSitePackageRnd}{2.4K\xspace}
+\newcommand{\MaxCallSitePackage}{2,372\xspace}
+\newcommand{\OneCallSitePackageRnd}{48\xspace}
+\newcommand{\OneCallSitePackage}{48\xspace}
+\newcommand{\NoCallSitePackagesRnd}{233\xspace}
+\newcommand{\NoCallSitePackages}{233\xspace}
+
+%\input{tag/enrichDataset.tex}
+\newcommand{\UndefinedEvalsRnd}{285K\xspace}
+\newcommand{\UndefinedEvals}{285,289\xspace}
+\newcommand{\PercentUndefinedEval}{3.3\%\xspace}
+\newcommand{\UndefinedEvalExpressionsRnd}{262\xspace}
+\newcommand{\UndefinedEvalExpressions}{262\xspace}
+\newcommand{\UndefinedOutOfCorpusRnd}{0\xspace}
+\newcommand{\UndefinedOutOfCorpus}{0\xspace}
+\newcommand{\UndefinedOutOfCoprusPercent}{0\%\xspace}
+\newcommand{\MaxCallSitesEnrich}{114\xspace}
+\newcommand{\OneCallSitesEnrich}{61\xspace}
+\newcommand{\NoCallSitesEnrich}{297\xspace}
+%\input{tag/patterns.tex}
+\newcommand{\PatternLibRnd}{4K\xspace}
+\newcommand{\PatternLib}{4,278\xspace}
+\newcommand{\PatternLibRatio}{.05\%\xspace}
+\newcommand{\PatternLibSourceRnd}{370\xspace}
+\newcommand{\PatternLibSource}{370\xspace}
+\newcommand{\PatternLibLocalRnd}{100\xspace}
+\newcommand{\PatternLibLocal}{100\xspace}
+\newcommand{\PatternReadRnd}{59K\xspace}
+\newcommand{\PatternRead}{59,776\xspace}
+\newcommand{\PatternReadRatio}{.7\%\xspace}
+\newcommand{\PatternReadSlotAccessRatio}{88\%\xspace}
+\newcommand{\PatternReadDatatableRatio}{27\%\xspace}
+\newcommand{\PatternAssignRnd}{68K\xspace}
+\newcommand{\PatternAssign}{68,754\xspace}
+\newcommand{\PatternAssignRatio}{.9\%\xspace}
+\newcommand{\PatternAssignArrowRatio}{18\%\xspace}
+\newcommand{\PatternAssignArrowMagrittrRatio}{59\%\xspace}
+\newcommand{\PatternAssignSuperarrowRatio}{81\%\xspace}
+\newcommand{\PatternAssignSuperarrowPlyrRatio}{99\%\xspace}
+\newcommand{\PatternManualPackagesRnd}{20\xspace}
+\newcommand{\PatternManualPackages}{20\xspace}
+\newcommand{\PatternManualCallsitesRnd}{161\xspace}
+\newcommand{\PatternManualCallsites}{161\xspace}
 
@@
 
-\newcommand{\eval}{\c{eval}}
+\newcommand{\eval}{\texttt{eval}}
+\newcommand{\evals}{{\sf eval}s\xspace}
 \newcommand{\parse}{\c{parse}}
@@
 \newcommand{\ggplot}{\c{ggplot2}}
-
 \newcommand{\base}{\c{base}}
@@
 
-\begin{figure*}[!t]\centering\includegraphics[width=.8\linewidth]
-{pipeline.pdf}\caption{Analysis Pipeline}\label{fig:pipeline}
-\end{figure*}
-
 Most R data types are vectorized. Values are constructed by the \c{c(...)}
@@
 yields the parse tree of the expression \c{e} after performing substitutions
-defined by the bindings in \c{envir}.
+defined by the bindings in the \c{envir} parameter.
 
@@
 function. There are three alternative forms  of \eval in the \c{base} package:
-\c{evalq}, \c{eval.parent}, \c{local}.
-
+\c{evalq}, \c{eval.parent} and \c{local}.
+
+\eval function in R takes three arguments as shown below.
 \begin{lstlisting}
@@
 
-Arguments are as follows: \c{expr} is the value to be evaluated, \c{envir}
+\c{expr} is the value to be evaluated, \c{envir}
 is the environment in which it is to be evaluated, and \c{enclos} is
@@
 When called, \eval evaluates \c{expr} in the environment specified by
-\c{envir} and return the computed value. If \c{envir} is missing, then the
+\c{envir} and returns the computed value. If \c{envir} is missing, then the
 default is \c{parent.frame()} (the environment where the call to \eval was
@@
 
+\begin{figure*}[!t]\centering\includegraphics[width=.8\linewidth]
+{pipeline.pdf}\caption{Analysis Pipeline}\label{fig:pipeline}
+\end{figure*}
+
+\medskip\emph{Discussion:~} There are two outstanding differences that set
+the design of \eval in R apart from JavaScript. On the usability side, \eval
+does not take a string, but an AST and an environment. This is slightly more
+involved than just taking a string and may discourage some casual users.
+The second substantial difference is the control that R gives over the
+choice of environment in which the AST is evaluated in. Beside traditional
+choices such as the top level or current lexical environment, R allows
+programmers to choose a fresh, empty, environment or any environment on the
+call stack. This is a significant increase in expressive power as functions
+that are otherwise not aware of the use of \eval may be deeply impacted.
+
 \section{Infrastructure}
-
-\begin{figure*}[!tb]\centering\includegraphics[width=.8\linewidth]
-  {corpus.pdf}\caption{CRAN packages}\label{fig:corpus}
-\end{figure*}
 
@@
 
+\begin{figure}[!h]\centering\includegraphics[width=\linewidth]
+  {corpus.pdf}\caption{CRAN packages}\label{fig:corpus}
+\end{figure}
+
 \subsection{CRAN Packages}
@@
 lines of R code: \CorpusExamplesProgramsRnd examples with
-\summaryrnd{CorpusExamplesCode}\footnote{This notation represents $sum$
-  ($min$ / $mean$ / $max$)} lines of code, and \CorpusVignettesProgramsRnd
-vignettes.%%% with \summaryrnd{CorpusVignettesCode} lines of code.
+\CorpusExamplesCodeRnd lines of code, and \CorpusVignettesProgramsRnd
+vignettes.
 
@@
 
-To represent end-user code in the corpus, we turned to
-Kaggle,\footnote{\url{https://www.kaggle.com}} an online platform for
-data-science and machine-learning. The website allows people to share
-data-science competitions, a data-analysis problem together with data, for
-which users try to find the best solution. The solutions, called {kernels},
-are then shared back to the platform in the form of either plain scripts or
-as notebooks.  One of the most popular competition is about predicting
-passenger survival on
+To represent end-user code in the corpus, we turned to Kaggle, an online
+platform for data-science and machine-learning. The website allows people to
+share data-science competitions, a data-analysis problem together with data,
+for which users try to find the best solution. The solutions, called
+{kernels}, are then shared back to the platform in the form of either plain
+scripts or as notebooks.  One of the most popular competition is about
+predicting passenger survival on
 Titanic\footnote{\url{https://www.kaggle.com/c/titanic}} with \CorpusKaggle
@@
 \CorpusKaggle kernels and extracting the R code from the various
-formats,\footnote{We use rmarkdown package to convert from Jupyer notebook
-  to Rmarkdown as well as to convert from Rmarkdown to plain R.} we found
-that \CorpusDuplicatedKaggle were whole-file duplicates. From
+formats,\footnote{We use {\sf rmarkdown} to convert from notebooks to R.} we
+found that \CorpusDuplicatedKaggle were whole-file duplicates. From
 \CorpusRunnableKaggle kernels, \CorpusFailedKaggle failed to execute. Next
@@
 misspelled package names.  The final set contains \CorpusFinishedKaggle
-kernels with \summaryrnd{CorpusFinishedKaggleCode} lines of R code.
+kernels with \CorpusFinishedKaggleCodeRnd lines of R code.
 
@@
 
-This section presents an overview of the usage of \eval in our corpus.
-Richards et al. reported 82\% of JavaScript-enabled web pages using
-\eval~\cite{ecoop11}. The case for R is more striking with 100\% of the
-programs call the \eval function.\footnote{We combine calls to \c{eval},
-  \c{eval.parent}, \c{evalq} and \c{local}.}  This high number is explained
-by the fact that the R implementation, its core packages, use \eval
-extensively. We differentiate uses based on their origin. A {\bf core} \eval
-is one that occurs in the code of a core package.  A {\bf dependent} \eval
-occurs in a package imported by the target package. An {\bf own} \eval is
-one that occurs in the source code of the target package.
+This section presents an overview of the usage of \eval in our corpus. Richards
+et al. reported 82\% of JavaScript-enabled web pages using \eval~\cite{ecoop11}.
+The case for R is more striking with 100\% of the programs call the \eval
+function.\footnote{We combine calls to {\sf eval}, {\sf eval.parent}, {\sf
+    evalq} and {\sf local}.} This high number is explained by the fact that the
+R implementation, its core packages, use \eval extensively. We differentiate
+three types of eval based on their origin:
+%
+\begin{compactitem}[$-$]
+  \item a {\bf core eval} is one that occurs in the code of a core package,
+  \item a {\bf dependent eval} occurs in a package imported by the target
+  package, and
+  \item an {\bf own eval} is one that occurs in the source code of the target
+  package.
+\end{compactitem}
 
@@
 %% SORRY  I HAD TO REFORMAT
-\begin{table}[tb] \centering
+\begin{table}[!h] \centering
  \begin{tabular}{r|c|c|c} \hline
@@
 Table~\ref{table:site-package-summary} shows the number of call sites in
-CRAN packages. 61 packages have a single call site, and one package, \emph{VGAM}, has \MaxCallSitePackage
-call sites of \eval. \NoCallSitePackages packages have no calls to \eval. The core packages
-have \NbCoreEvalCallSites call sites of \eval and Kaggle has none. These numbers are
-under-approximations as \eval may be called reflectively, through aliases,
-or passed to a higher-order function.
+CRAN packages. \OneCallSitesEnrich{} packages have a single call site, and one
+package, \emph{metafor}, has \MaxCallSitesEnrich{} call sites of
+\eval. \NoCallSitesEnrich{} packages have no calls to \eval. The core packages
+have \NbCoreEvalCallSites{} call sites of \eval and Kaggle has none. These
+numbers are under-approximations as \eval may be called reflectively,
+through aliases, or passed to a higher-order function.
 
@@
 
-In this section we turn our attention to the expression passed for evaluation to
-\eval. The \eval function accepts all types of R objects as inputs, but in the
-majority of cases (\AllExpressionInputEvalCallPerc), it is called with
-expressions. The remaining \AllValueInputEvalCallPerc are values, \ie, they
-evaluate to themselves.
-%
-%Furthermore, the proportion of values passed to core
-%\eval call is \missingTag{CoreEvalValueInputPerc} compared to a significantly higher
-%\missingTag{PackageEvalValueInputPerc} for package \eval calls.
+In this section we turn our attention to the expression passed for
+evaluation to \eval. The \eval function accepts all types of R objects as
+inputs, but in the majority of cases (\AllExpressionInputEvalCallPerc), it
+is called with expressions. The remaining \AllValueInputEvalCallPerc are
+values, \ie, they evaluate to themselves.
 %
@@
 the default value for a function parameter can be a complex expression or a
-benign value such as \c{NULL}, \c{match.arg} evaluates it through a call to
-\eval. This makes the \eval inside \c{match.arg} ``polymorhpic''. The second
-such \eval call originates from the popular \ggplot package of R which is a DSL
-for plotting. \ggplot has it's own object system, \ggproto, which uses \eval
-inside the \ggproto function to determine the super-class from which it
-inherits. The default value of the superclass is \c{NULL} but in general could
-be an arbitrarily complex expression. The third \eval call is in core R \c{str}
-function. This function is used to print a summary of R objects. It internally
-maps \eval function on the bindings of an environment which results in inputs of
-multiples types being passed to \eval.
+benign value such as \c{NULL}, the \c{match.arg} function evaluates it through a
+call to \eval. This makes the \eval inside \c{match.arg} ``polymorhpic''. The
+second such \eval call originates from \ggplot, a popular R package for making
+plots. \ggplot has it's own object system, \ggproto, which uses \eval inside the
+\ggproto function to determine the super-class from which it inherits (the
+default value is \c{NULL} but in general it could be an arbitrarily expression).
+The third \eval call is in core R \c{str} function used to print a summary of R
+objects. It internally maps \eval function on the bindings of an environment
+which results in inputs of multiples types being passed to \eval.
 
 We further investigate the type of input passed to \eval in
-Figure~\ref{fig:eval-expression-kind} by the type of input and
-the source of \eval. In this figure, \emph{Expression} is the type of program text,
+Figure~\ref{fig:eval-expression-kind} by the type of input and the source of
+\eval. In this figure, \emph{Expression} is the type of program text,
 \emph{Environmment} is the type of scopes or environments used for binding,
-\emph{Null} is the type of the unique \c{NULL} object, \emph{Vector} is the type
-assigned to vectors of strings, booleans, integers, reals and bytes;
+\emph{Null} is the type of the unique \c{NULL} object, \emph{Vector} is the
+type assigned to vectors of strings, booleans, integers, reals and bytes;
 \emph{Symbol} is the type of unevaluated symbols, and \emph{Function} is the
@@
 
-From the figure, we observe that majority of inputs to Core \eval functions are
-expressions in contrast to package \eval calls which receive mostly environments
-as inputs. In fact, \PackageEnvironmentInputEvalCallPerc of \eval calls in
-packages receive environments as input expressions. There are only four
-callsites in package evals that receive environments as expressions.
+From the figure, we observe that majority of inputs to Core \eval functions
+are expressions in contrast to package \eval calls which receive mostly
+environments as inputs. In fact, \PackageEnvironmentInputEvalCallPerc of
+\eval calls in packages receive environments as input expressions. There are
+only four callsites in package evals that receive environments as
+expressions.
 \begin{compactitem}[$-$]
@@
 OOP system.
-\item \c{future::backtrace} function applies \eval to a future object which is
-implemented as an environment. This is called only once.
+\item \c{future::backtrace} function applies \eval to a future object which
+  is implemented as an environment. This is called only once.
 \item \c{RModel::str.RMmodel} function overloads the core \c{str} method for its
@@
 
-We looked at the top ten expressions passed to core and package \eval calls. The
-most frequent ten expressions to eval calls from core R contribute to 85.2\% of
-all \eval calls.
-\begin{table*}[tb]
-  \centering
-  \begin{tabular}{l|rr} \hline
-    Expression & Eval Call & Eval Call \% \\\hline
-    \c{c("auto", "shell", "radix")} & 1987105 & 29.3\%\\
-    \c{c("auto", "shell", "quick", "radix")} & 1593169  & 23.5\%\\
-    \c{\{info <- loadingNamespaceInfo(...} & 1008632 &       14.9\%\\
-    \c{c("onLoad", "attach", "detach", "onUnload")}   & 470566 &      6.9\%\\
-    \c{c("append", "prepend", "replace")} &              261587&       3.9\% \\
-    \c{c("left", "right", "centre", "none")} & 162086     & 2.4\%\\
-    \c{c("no", "ifany", "always")}   &                71580 &       1.1\%\\
-    \c{c("pearson", "kendall", "spearman")}  & 72962 &      1.1\%\\
+We looked at the top ten expressions passed to core and package \eval
+calls. The most frequent ten expressions to eval calls from core R
+contribute to 85\% of all \eval calls.
+
+\begin{table}[!h] \centering
+\begin{tabular}{@{}l|rr@{}} \hline
+Expression & Eval Call &  \% \\\hline
+\c{c("auto", "shell", "radix")} & 198,7105 & 29\%\\
+\c{c("auto", "shell", "quick", "radix")} & 1,593,169  & 23\%\\
+\c{\{info <- loadingNamespaceInfo(...} & 1,008,632 &       14\%\\
+\c{c("onLoad", "attach", "detach", "onUnload")}   & 470,566 &      6.9\%\\
+\c{c("append", "prepend", "replace")} &              261,587&       3.9\% \\
+\c{c("left", "right", "centre", "none")} & 162,086     & 2.4\%\\
+\c{c("no", "ifany", "always")}   &                71,580 &       1.1\%\\
+\c{c("pearson", "kendall", "spearman")}  & 72,962 &      1.1\%\\
 \c{NULL}& 75330  &      1.1\% \\
-    \c{Symbol}&                 66279&       1\%\\\hline
+\c{Symbol}&                 6,6279&       1\%\\\hline
 \end{tabular}
-  \caption{Top ten eval calls in Core}
-  \label{A}
-\end{table*}
+\caption{Top ten eval calls in Core}\label{B}
+\end{table}
 
@@
 
-The most frequent ten expressions to eval calls from CRAN packages contribute to
-77.1\% of all \eval calls.
-\begin{table*}[tb]
-  \centering
-  \begin{tabular}{l|rr} \hline
-    Expression & Eval Call & Eval Call \% \\\hline
-    \c{Environment} &                                                              989302   & 61.8\%\\
+The most frequent ten expressions to eval calls from CRAN packages
+contribute to 77.1\% of all \eval calls.
+
+\begin{table}[!h]  \centering
+\begin{tabular}{@{}l@{~}|@{~}r@{~}r@{}} \hline
+Expression & Eval Call &  \% \\\hline
+\c{Environment} &                                  989302   & 61\%\\
 \c{column[rows] <<- what} &                        55677    & 3.5\%\\
-    \c{function(value) freduce(value, `_function_list`)} &                         37251    & 2.3\%\\
+\c{function(value) freduce(value, `_function_list`)} & 37251& 2.3\%\\
 \c{NULL} &                         32005    & 2\%\\
@@
 \c{String Vector}      &                           17487     & 1.1\%\\
-  \end{tabular}
-  \caption{Top ten eval calls in CRAN}
-  \label{A}
-\end{table*}
-
-The expression \c{Environment} occurs because of the four callsites explained
-above, \c{ggplot2::ggproto}, \c{R6::generator_funs}, \c{future::backtrace} and
-\c{RModel::str.RMmodel}. The next expression, \c{column[rows] <<- what}, is used
-inside the \c{plyr::rbind.fill} function to merge data frames by assigning
-concatenated vectors to rows. The \c{<<-} operator is interesting in that it
-skips the current scope and assigns in a parent scope in which the variable is
-already present. In our corpus, all these \eval calls contribute to a single
-side-effect. The expression \c{function(value) freduce(value, `_function_list`)}
-arises from the \c{magrittr::\%>\%} function which is a pipe operator that pipes
-the output of previous command to the next one. The expression is evaluated in a
-custom environment to create a function binding for evaluating the components of
-the pipe.\c{String Vector} and \c{List} also arise from the same function when a
-string or a list is piped using the \c{\%>\%} function into the next expression.
-The \c{NULL} arises from \c{R6::generator_funs} function when the \eval is
-passed a \c{NULL} argument by the \c{DataMask_generator} package. The
-\c{c("default", "default2012", "default2011" ...} pattern arises from
-\c{copula::polyG} where it reflectively access the default expression for its
-formal parameter and evaluates it. The \c{force(..1)} and \c{alist(`_spec`)}
-patterns occur in \c{glue::glue_data} function which concatenates and
-interpolates strings. The two patterns occur because the function captures
-unevaluated unnamed arguments and maps the evaluation of \c{force(..1)} on them.
-The \c{force} function forces promises~\cite{oopsla19a} and returns the result
-of evaluation. The \c{inner} pattern arises from \c{glue::identity_transformer}
-which enables the creation of custom transformation functions for affecting the
-interpolation and concatenation of input by the \c{glue} package.
+\end{tabular}\caption{Top ten eval calls in CRAN} \label{C}
+\end{table}
+
+The expression \c{Environment} occurs because of the four callsites
+explained above, \c{ggplot2::ggproto}, \c{R6::generator_funs},
+\c{future::backtrace} and \c{RModel::str.RMmodel}. The next expression,
+\c{column[rows] <<- what}, is used inside the \c{plyr::rbind.fill} function
+to merge data frames by assigning concatenated vectors to rows. The \c{<<-}
+operator is interesting in that it skips the current scope and assigns in a
+parent scope in which the variable is already present. In our corpus, all
+these \eval calls contribute to a single side-effect. The expression
+\c{function(value) freduce(value, `_function_list`)} arises from the
+\c{magrittr::\%>\%} function which is a pipe operator that pipes the output
+of previous command to the next one. The expression is evaluated in a custom
+environment to create a function binding for evaluating the components of
+the pipe.\c{String Vector} and \c{List} also arise from the same function
+when a string or a list is piped using the \c{\%>\%} function into the next
+expression.  The \c{NULL} arises from \c{R6::generator_funs} function when
+the \eval is passed a \c{NULL} argument by the \c{DataMask_generator}
+package. The \c{c("default", "default2012", "default2011" ...} pattern
+arises from \c{copula::polyG} where it reflectively access the default
+expression for its formal parameter and evaluates it. The \c{force(..1)} and
+\c{alist(`_spec`)} patterns occur in \c{glue::glue_data} function which
+concatenates and interpolates strings. The two patterns occur because the
+function captures unevaluated unnamed arguments and maps the evaluation of
+\c{force(..1)} on them.  The \c{force} function forces promises and returns
+the result of evaluation. The \c{inner} pattern arises from
+\c{glue::identity_transformer} which enables the creation of custom
+transformation functions for affecting the interpolation and concatenation
+of input by the \c{glue} package.
 
 \subsection{Operation Mix}
-In this section, we look at the operations executed by the expressions and
-symbols on evaluation. R is a lazy language, so symbols can be bound to promises
-that can evaluate arbitrary code. All operations in R are function calls, even
-primitive operations such as addition and assignment are treated as calls to
-functions implemented in the interpreter. In our corpus, we observe 2.9G calls
-to R functions of which 125.5M happen inside \eval. Furthermore, unlike
-Javascript, R code also calls into peformant native implementations of numerical
-and statistical routines. In the corpus there are 463.8M calls to native code of
-which only 31.2M happens inside \eval.
-Qualitatively, we observe all kinds of operations being executed inside \eval
-calls, from numerical computing for statistical modeling to reflection and
-metaprogramming for implementing DSLs. Even testing frameworks such \c{genthat}
-heavily rely upon \eval to execute test code snippets. The core R implementation
-is rife with uses of \eval for mundane things like accessing default values of
-formal parameters and in statistical packages for computation.
-
-\subsubsection{Provenance}
-
-In this section we turn our attention to the \AllExpressionInputEvalCallPerc
-expressions passed as input to \eval. There are two ways to obtain
-expressions in R.
+
+The JavaScript paper reported on the operation mix within \evals. This means
+the proportion of read, write, and call bytecodes performed by the
+JavaScript interpreter. They found that some web pages were creating objects
+more frequently and writing into their fields.  In R, nearly every operation
+performed in the interpreter is a function call. We observe 2.9G calls and,
+of those 125M happen inside \eval (there are 463M native calls, of which
+only 31M are in \eval).
+
+{\bf there is no good reason for not counting reads and writes, we can...}
+
+%% ...The following is fluff ...
+
+%% Qualitatively, we observe all kinds of operations being executed inside
+%% \eval calls, from numerical computing for statistical modeling to reflection
+%% and metaprogramming for implementing DSLs. Even testing frameworks such
+%% \c{genthat} heavily rely upon \eval to execute test code snippets. The core
+%% R implementation is rife with uses of \eval for mundane things like
+%% accessing default values of formal parameters and in statistical packages
+%% for computation.
+
+\subsection{Provenance}
+
+In JavaScript, the provenance of the string passed into \eval is a matter of
+concern as \eval can be used to breach the security of websites. The main
+danger comes from string that were obtained from external sources. If they
+are allowed to flow into an \eval, then an attacker could hijack the
+program.
+
+{\bf what have they found...}
+
+There are various ways to obtain expressions:
 \begin{itemize}
-  \item \c{substitute} function can be used to synthesize ASTs from
-  expressions by replacing symbols with their bindings in the specified
-  environment.
-  \item \c{expression} function can be used to create a vector of expression
+  \item \c{substitute} synthesizes ASTs from expressions by replacing
+    symbols with their bindings in the specified environment.
+  \item \c{expression}  creates a vector of expression
     objects from text.
-  \item \c{parse} function and its variants \c{str2expression} and \c{str2lang}
-    can be used to turn strings or text from file into R expressions.
+  \item \c{parse}, \c{str2expression} and \c{str2lang} turn strings into
+    expressions.
 \end{itemize}
 
-In our corpus, we observe 5.2\% cases where \eval directly evaluates output of
-\c{substitute}, 0.7\% cases where output of \c{parse} and its variants is read
-directly and only 704 cases for \c{expression}. Majority of expression objects
-consumed by \eval are generated by their transitive uses by other functions.
-
-The pattern \c{eval(parse(...))} can be used for dynamic code loading. This
-forms the core of \c{source} and \c{sys.source} functions in R that are commonly
-used for loading code in R files in interactive settings. We investigated the
-number of cases in which the output of \c{parse} and its variants is passed to
-\eval, directly or transitively by tainting their output. This corresponds to \PercentParsedCallSites of the total \eval call sites and
-\PercentParsedEvals of the eval calls. We observed that
-very few of the eval call(\NbParseFilesRnd in total) consume the result of calling \c{parse} on
-a file. Most of the eval calls consume the result of calling
-\c{parse} on a string.  We also identified one
-function in core R, \c{invokeRestartInteractively} that prompts the user for
-input, parses it, and passes it to \eval.
-
-%%Unlike Javascript, R packages also call native code for access to efficient
-%%numerical and statistical routines. We intercept all calls to R functions and
-%%native functions from the programs and \eval calls. In the whole corpus there
-%%are \AllClosureCallCount calls to R functions and \AllNativeCallCount to native
-%%code. Of those, \eval contribute to only \EvalClosureCallCount calls to R
-%%functions and \EvalNativeCallCount to native code.
-%%
-%%\begin{table}[ht]\resizebox{\columnwidth}{!}{%
-%%    \begin{tabular}{lll}\hline
-%%      Source & Primitive Functions & R Functions & Native Functions \\ \hline
-%%      Core & \CorePrimitiveFunctionCount & \CoreRFunctionCount & \CoreNativeFunctionCount \\
-%%      CRAN & \PackagePrimitiveFunctionCount & \PackageRFunctionCount & \PackageNativeFunctionCount \\
-%%      All  & \AllPrimitiveFunctionCount & \AllRFunctionCount & \AllNativeFunctionCount \\
-%%    \end{tabular}} \label{table:operation-distribution}
-%%  \caption{Distribution of operation calls to \eval calls across Core and Package}
-%%\end{table}
+We observe 5.2\% cases where \eval directly evaluates the output of
+\c{substitute}, 0.7\% cases where output of \c{parse} is read directly and
+only 704 cases for \c{expression}. Most expressions consumed by \eval are
+generated by other functions.
+
+\c{eval(parse(...))} can be used for dynamic code loading. This forms the
+core of \c{source} and \c{sys.source} functions in R that are commonly used
+for loading code in R files in interactive settings. We investigated the
+number of cases in which the output of \c{parse} and its variants is passed
+to \eval, directly or transitively by tainting their output. This
+corresponds to \PercentParsedCallSites of the total \eval call sites and
+\PercentParsedEvals of the eval calls. We observed that very few of the eval
+call (\NbParseFilesRnd in total) consume the result of calling \c{parse} on
+a file. Most of the eval calls consume the result of calling \c{parse} on a
+string.  We also identified one function in core R,
+\c{invokeRestartInteractively} that prompts the user for input, parses it,
+and passes it to \eval.
 
@@
 dangerous. Only writes to environments not local to the computation spawned
-by \eval are side-effecting. These writes outlive the computation and hence are
-visible outside the computation. The remaining writes are local to the
-computation. We observe that \EvalSideEffectingWritesRnd writes inside evals are
-side-effecting. This is only \EvalSideEffectingWritesEvalPerc of all variable
-writes inside \eval and \EvalSideEffectingWritesAllPerc of all variable writes
-in the corpus.
+by \eval are side-effecting. These writes outlive the computation and hence
+are visible outside it. The remaining writes are local to the
+computation. We observe that \EvalSideEffectingWritesRnd writes inside evals
+are side-effecting. This is only \EvalSideEffectingWritesEvalPerc of all
+variable writes inside \eval and \EvalSideEffectingWritesAllPerc of all
+variable writes in the corpus.
 
@@
 One of the main differences of the R \eval with the javascript one is how it
-interacts with \emph{scope}, or environments in R. In Javascript, \c{eval} can
-access the local scope and the global scope. In R, \c{eval} has an argument
+interacts with \emph{scope}, or environments in R. In Javascript, \eval can
+access the local scope and the global scope. In R, \eval has an argument
 \c{envir} to indicate in which environment its first argument must be evaluated.
 By default, it is equal to \c{parent.frame()}, 
-which is the parent environment of \c{eval} call. This default argument
+which is the parent environment of \eval call. This default argument
 happens in \DefaultEnvirExprPercent of the eval calls.
 
-This argument can also be \c{NULL}, a \c{List} or a data frame. This happens in
-3.5\% of the cases.
-\c{eval} copies the fields of the list or data frame and creates bindings for
-them in a new environment. This pattern is used to evaluate formulas which can
-directly refer to the fields of the data.
-The \c{envir} argument can also be a number $n$. It means that the environment in
-which the expressions is evaluated will be the result of \c{sys.call(n)} where
-$n$ refers to the $n$-th stack frame.
+This argument can also be \c{NULL}, a \c{List} or a data frame. This happens
+in 3.5\% of the cases: \eval copies the fields of the list or data frame
+and creates bindings for them in a new environment. This pattern is used to
+evaluate formulas which can directly refer to the fields of the data.  The
+\c{envir} argument can also be a number $n$. It means that the environment
+in which the expressions is evaluated will be the result of \c{sys.call(n)}
+where $n$ refers to the $n$-th stack frame.
 
@@
 We looked at the environments passed to all the \eval calls in our corpus.
-Table~\ref{tab:environments} summarizes the results. A numeric environment class
-\c{n} denotes the environment of the $n$-th call stack frame from the current
-function. \c{global} denotes the top-level environment and \c{list} denotes a list
-passed for evaluation of formulas. Environment classes of the form $n+$ denote
-the $n$-th environment extended with a new environment. The new environment
-provides a limited form of sandboxing. All assignments using the \c{\<-}
-function occur inside it and prevent the extended environment from
-mutation. However, it is still possible to mutate the extended environment
-using the \c{\<\<-} or \c{assign} functions; but, that happens rarely.
-
-\begin{table}[htbp]
+Table~\ref{tab:environments} summarizes the results. A numeric environment
+class \c{n} denotes the environment of the $n$-th call stack frame from the
+current function. \c{global} denotes the top-level environment and \c{list}
+denotes a list passed for evaluation of formulas. Environment classes of the
+form $n+$ denote the $n$-th environment extended with a new environment. The
+new environment provides a limited form of sandboxing. All assignments using
+the \c{\<-} function occur inside it and prevent the extended environment
+from mutation. However, it is still possible to mutate the extended
+environment using the \c{\<\<-} or \c{assign} functions; but, that happens
+rarely.
+
+\begin{table}[htbp] \resizebox{\columnwidth}{!}{%
     \centering
@@
         \hline
-    \end{tabular} \label{tab:environments}
+    \end{tabular}} \label{tab:environments}
 \caption{Environments in terms of eval calls.}
@@
 
-We observe that a disproportionately high core R \eval calls access the
-caller's caller's environment. This is because many core R functions call
-functions that pass the result of \c{parent.frame()} to an \eval.
-A disproportionately high calls to \eval happen in an extended
-top-level environment. This can be explained by the fact that many packages
-evaluate code passed from the user's workspace in the top-level environment to
-access the bindings.
-
-%%Below, we provide some code snippets to show how some of these environment
-%%classes manifest.
-
-%%The \c{match.arg} (see Listing~\ref{code:matchargenvir}) function has an
-%%environment class 2. \c{sys.parent()} refers to the parent environment of
-%%\c{match.arg} and so to the grand-parent environment of \c{eval} which we denote
-%%as $2$.
-%%
-%%\begin{lstlisting}[label=code:matchargenvir, caption={\c{match.arg} using \eval with a custom environment.}]
-%%formal.args <- formals(sys.function(sysP <- sys.parent()))
-%%choices <- eval(formal.args[[as.character(substitute(arg))]], envir = sys.frame(sysP))
-%%\end{lstlisting}
-
-%%In package \emph{gamlss::get.K} function \c{get.K} evaluates its expression in its
-%%parent environment, the one of the body of \c{get.K} returned from
-%%\c{environment()}.
-%%
-%%\begin{lstlisting}
-%%eval(parse(text = paste(\"fam$dld\", ii, sep = \"\")), envir = environment())
-%%\end{lstlisting}
-%%
-%%
-%%The \emph{glue} library is used to interpolate string litterals with variables, such as in \c{glue('My name is \{name\}.')} where \c{name} will be subtituted by its value.
-\emph{glue} makes it possible to define transformers that process those interpolated variables.
-They are built upon the \c{identity_transformer}, which uses \eval.
-The interpretation environment for \eval is often global here, as the string litterals are often used to print variables at the top level.
-%%
-%%\begin{lstlisting}
-%%identity_transformer <- function(text, envir) 
-%%{
-%%    eval(parse(text = text, keep.source = FALSE), envir)
-%%}
-%%\end{lstlisting}
-%%
-%%
-%% \end{lstlisting}
-%% $$$$ <- to fix emacs fontification issues
-
-The \c{imchange} function of package \c{imager} makes it possible to modify images using a dedicated formula syntax using \c{\~}.\footnote{Inspired by \c{map} in package \emph{purr}.}  Here, \eval is evaluated in \c{newenv}, which creates a new environment that inherits from \c{parent.frame()} by default (classified as 1+).
+We observe that a disproportionately high number of core R \eval calls
+access the caller's caller's environment. This is because many core R
+functions call functions that pass the result of \c{parent.frame()} to an
+\eval.  A disproportionately high number of calls to \eval happens in an
+extended top-level environment. This can be explained by the fact that many
+packages evaluate code passed from the user's workspace in the top-level
+environment to access the bindings.
+
+The \c{imchange} function of package \c{imager} makes it possible to modify
+images using a dedicated formula syntax using \c{\~}.\footnote{Inspired by
+ {map} in package \emph{purr}.}  Here, \eval is evaluated in \c{newenv},
+which creates a new environment that inherits from \c{parent.frame()} by
+default (classified as 1+).
 
@@
 
-\c{adjCoef} in package \emph{actuar} find the root of an equation defined by a function \c{h} whose arguments must be named \c{x} and \c{y}.
-\c{h} is transformed into an auxiliary function \c{h2} that can be optimized.
-Here, the list used for \c{envir} ensures the correspondance between the textual arguments of \c{h} and the arguments of \c{h2}.
+\c{adjCoef} in package \emph{actuar} find the root of an equation defined by
+a function \c{h} whose arguments must be named \c{x} and \c{y}.  \c{h} is
+transformed into an auxiliary function \c{h2} that can be optimized. Here,
+the list used for \c{envir} ensures the correspondance between the textual
+arguments of \c{h} and the arguments of \c{h2}.
 
@@
 
-The JavaScript paper identified a number of frequently occurring \eval patterns
-which were detected using simple regular expressions on the \eval strings. In
-the case of R we match the patterns on the recorded AST.
+The JavaScript paper identified a number of frequently occurring \eval
+patterns which were detected using simple regular expressions on the \eval
+strings. In the case of R, we match the patterns on the resolved AST.
 
@@
 patterns. We included all the original patterns except for \emph{JSON} as R uses
-its own serialization format and \emph{empty} as there were no empty expression
+its own serialization format and for \emph{empty} as there were no empty expression
 passed to \eval:
@@
 
-%\noindent
-%\begin{compactitem}[$-$]
-  % \item \emph{Library}: in JavaScript, the library pattern captures asynchronous
-  % loading of external external function definitions. In R we have seen only
-  % \PatternLibRnd such calls (\PatternLibRatio) mostly coming from source loading
-  % functions (\ie \c{source} and \c{sys.source}). Most of the external function
-  % definitions in R come from packages and R makes it very easy for users to
-  % install packages from its centralized repository.
-
 \noindent \emph{Library}, \emph{Typeof} and \emph{Try} are all barely used
@@
 \noindent \emph{Read} pattern in JavaScript represented mostly an incorrect
-access to an object's property or local variable both replaceable by JavaScript
-core functions. In the case of R there are \PatternReadRnd \eval calls
-(\PatternReadRatio). They come from calls like \c{subset(airquality, select =
-  Temp)} where the variable \c{Temp} will be evaluated in the \c{airquality}
-data frame.
+access to an object's property or local variable both replaceable by
+JavaScript core functions. In the case of R there are \PatternReadRnd read
+\eval calls (\PatternReadRatio). They come from calls like
+\c{subset(airquality, select = Temp)} where the variable \c{Temp} will be
+evaluated in the \c{airquality} data frame.
 %
-\noindent \emph{Assign} category comprises of assignments to variables within
-an \eval call. We have seen \PatternAssignRnd such calls (\PatternAssignRatio).
-\PatternAssignArrowRatio uses the simple assignment (\c{<-}). The majority of
-cases comes from the \c{magrittr}'s \c{\%<>\%} pipe operator
-(\PatternAssignArrowMagrittrRatio) which forwards an object to a function on rhs
-and update lhs with the result. Far more calls used the super assignment
-(\c{<<-}) operator, but almost all come from the \c{plyr} package which provides
-a convenient API for common data manipulation problems.
+\noindent \emph{Assign} category comprises of assignments to variables
+within an \eval call. We have seen \PatternAssignRnd such calls
+(\PatternAssignRatio).  \PatternAssignArrowRatio uses the simple assignment
+(\c{<-}). The majority of cases comes from the \c{magrittr}'s \c{\%<>\%}
+pipe operator (\PatternAssignArrowMagrittrRatio) which forwards an object to
+a function on rhs and update lhs with the result. Far more calls used the
+super assignment (\c{<<-}) operator, but almost all come from the \c{plyr}
+package which provides a convenient API for common data manipulation
+problems.
 %
-\noindent \emph{Call} and \emph{Other} together represents the vast majority of
-eval calls in R. The \emph{Other} pattern represents all the calls where eval
-was called with a value instead of with an expression. The \emph{Call} pattern
-contains all the calls that did not fall in the previous categories. Without a
-context it is usually difficult to further categorize the intent of these calls.
-Therefore we have randomly chosen \PatternManualPackages packages and manually
-looked at \PatternManualCallsites call sites from which we have observed the following categories:
-
-% this is read pattern
-% \subsubsection{Evaluation using Data Fields} A common use of \eval is to
-%   evaluate expressions that access data frame columns or list (heterogeneous
-%   vectors) elements by name. Core R provides two functions, \c{with} and
-%   \c{within} that implement this pattern. These functions construct a temporary
-%   environment with data frame columns or list elements as the bindings.
-%   Side-effects performed by the evaluated expression take place in this temporary
-%   environment and are ignored by the \c{with} function. The \c{within} function
-%   copies these modifications to a local copy of the input data and returns it.
-%   This pattern is used quite often for statistical modeling. The modeling
-%   expressions is constructed from variable names whose values are supplied by a
-%   data frame or list.
-\subsubsection{Reconding the current call} R provides a \c{match.call} function
+\noindent \emph{Call} and \emph{Other} together represents the vast majority
+of eval calls in R. The \emph{Other} pattern represents all the calls where
+eval was called with a value instead of with an expression. The \emph{Call}
+pattern contains all the calls that did not fall in the previous
+categories. Without a context it is usually difficult to further categorize
+the intent of these calls.  Therefore we have randomly chosen
+\PatternManualPackages packages and manually looked at
+\PatternManualCallsites call sites from which we have observed the following
+categories:
+
+\subsubsection{Recording the current call} R provides a \c{match.call} function
 which can return the current call. This is heavily used by the statistical
-model-fitting functions that records the current call that is then reevaluated
-in the different environment. Another pattern is to match the current call,
-adjust some arguments and pass it to another function. For example the
-\c{write.csv} function changes the \c{sep} argument in the matched call and reevaluate itself as \c{write.table}:
+model-fitting functions that records the current call that is then
+reevaluated in the different environment. Another pattern is to match the
+current call, adjust some arguments and pass it to another function. For
+example the \c{write.csv} function changes the \c{sep} argument in the
+matched call and reevaluates itself as \c{write.table}:
 %
@@
 \subsubsection{Sandboxing} The combination of \eval and first class
-  environments can be used to construct a sandbox for evaluating side-effecting
-  expressions. The newly constructed environment can use the current environment
-  for its lexical scope. This provides the evaluated expression access to all the
-  bindings in the current scope but restricts the side-effect to the newly
-  constructed throwaway environment. However, this sandboxing is limited because
-  the expression can reflectively access any environment in the call stack or use
-  the super assignment operator(\c{<<-}) to write to parent scope. The \c{local}
-  function which is part of Core R is an example of this usage pattern.
+environments can be used to construct a sandbox for evaluating
+side-effecting expressions. The newly constructed environment can use the
+current environment for its lexical scope. This provides the evaluated
+expression access to all the bindings in the current scope but restricts the
+side-effect to the newly constructed throwaway environment. However, this
+sandboxing is limited because the expression can reflectively access any
+environment in the call stack or use the super assignment operator(\c{<<-})
+to write to parent scope. The \c{local} function which is part of Core R is
+an example of this usage pattern.
 
@@
 used to parse text from string or file and evaluate in a custom environment.
-  This pattern is used by the \source Core R function to load R code from a file
-  in the current workspace. This is quite useful for loading functions for
-  interactive data analysis.
+This pattern is used by the \source Core R function to load R code from a
+file in the current workspace. This is quite useful for loading functions
+for interactive data analysis.
 
 \subsubsection{Metaprogramming} The combination of \eval and \substitute is
-  used to do metaprogramming in R. \substitute can be used to synthesize ASTs by
-  stitching together user inputs. \substitute walks over its argument AST and
-  replaces the symbols with their bindings in the specified environment and
-  returns this new AST which can then be evaluated in custom environments using
-  \eval.
-
-  \subsubsection{Convenience}
-  evaling recorded calls in model-fitting functions or passing most of the
-  current call to another function. Both ``patterns'' use the \c{match.call} to
-  capture current call. The latter is just for laziness as it is the same as
-  calling the new function with explicitly selecting the arguments.
+ used to do metaprogramming in R. \substitute can be used to synthesize ASTs
+ by stitching together user inputs. \substitute walks over its argument AST
+ and replaces the symbols with their bindings in the specified environment
+ and returns this new AST which can then be evaluated in custom environments
+ using \eval.
+
+  \subsubsection{Convenience} Eval is used to make some programming tasks easier (and/or shorter). For example, in \c{data.table}, it is used to set R options from a character vector
+  \begin{lstlisting}
+opts = c("datatable.verbose"="FALSE", ...)
+for (i in names(opts)) eval(parse(text=paste0("options(",i,"=",opts[i],")")))
+
+  \end{lstlisting}
+  %
+  instead of manually calling \c{options(datatable.verbose=FALSE)}
 
@@
   modify a binding in the \base package of R. Bindings in the \base package are
-  locked for modification and require explicit unlocking using \unlockBinding
-  function. However, directly using this function triggers a
+  locked for modification and require explicit unlocking using 
+  function \unlockBinding. However, directly using this function triggers a
   restricted-function-use warning during package build, preventing the package
@@
   dependencies.
-  % \AllValueInputEvalCallPerc migth hit on a wong usage
+
+  \subsubsection{Unnecessary use of eval}
+  Similarly to JavaScript, there are also unnecessary uses of \eval.
+  For example, the \c{PerformanceAnalytics} package contains a function
+  \c{chart.QQPlot} that uses \eval to resolve a string into function and
+  another to call it and assign its results into a variable:
+  \begin{lstlisting}
+function (R, d="norm", dp, ...) {
+q.f <- eval(parse(text=paste("q",d,sep="")))
+z <- NULL
+eval(parse(text=paste("z<-q.f(",dp,",...)")))
+}
+  \end{lstlisting}
+  In both cases, there is no need for \eval:
+  \begin{lstlisting}
+function (R, d="norm", dp, ...) {
+q.f <- get(paste0("q",d))
+z <- q.f(dp, ...)
+}
+  \end{lstlisting}
+  or even to a oneliner \c{do.call(paste0("q",d), as.list(dp, ...))}.
 
@@
 We look at how \emph{consistent} the \c{expr} argument of \eval can be, \ie
-how many different types of the resolved \c{expr} there are per call sites. Most of the call sites, \ie \PercentMonomorphic,  are
-\emph{consistent}, and this is similar to javascript.
-However, a few ones are highly \emph{polymorphic} (10 different types).
-They are the  pipe  operators \c{\%>\%}, \c{\%<>\%} and \c{\%$\%} in package \emph{magrittr}.
-It is effectively used to compose functions on their first argument, which can be of any type.
-
-%\begin{figure}[!h]
-%	\centering
-%    \includegraphics[width=\columnwidth]{polymorphism}
-%    \caption{Number of different types  of \c{expr} per call sites.} \label{fig:polymorphism}
-%\end{figure}
-%
-%\subsection{base}
-%
-%%There are 30 functions in \emph{base} that use \eval.
-Only five of them use \c{parse}, for legitimate reasons, such \c{source} and \c{sys.source} which load a file and then evaluate it, or \c{invokeRestartInteractively} which prompts the user for arguments to restart execution after an error.
-%%
-%%We endeavored  to write these functions without \eval.
-\c{match.arg} appears in \MatchArgPercent of the \eval calls.
-It is used to get the arguments, possibly abbreviated, of a function.
-With one argument, as \c{match.arg(arg)}, it uses \c{eval} to infer them from a parent function's formal arguments.
-We can rewrite it using its version with two arguments, \c{match.arg(arg, choices)} where \c{choices} are the possible formal arguments of the function, as shown in Listing~\ref{code:matcharg}.
-%%
-%%\begin{lstlisting}[caption={Rewriting \c{match.arg} without \eval}, label=code:matcharg]
-%%f <- function(type=c("examples", "tests", "vignettes")) {
-%%actual_type = match.arg(type)
-%%...
-%%}
-%%
-%%f <- function(type=c("examples", "tests", "vignettes")) {
-%%actual_type = match.arg(type, choices=c("examples", "tests", "vignettes"))
-%%...
-%%}
-%%\end{lstlisting}
+how many different types of the resolved \c{expr} there are per call
+sites. Most of the call sites, \ie \PercentMonomorphic, are
+\emph{consistent}, and this is similar to javascript. However, a few ones
+are highly \emph{polymorphic} (10 different types). They are the pipe
+operators \c{\%>\%}, \c{\%<>\%} and \c{\%$\%} in package \emph{magrittr}. It
+is effectively used to compose functions on their first argument, which can
+be of any type.
 
 \section{Threats to Validity} The primary threat to validity is the issue of
-code coverage, a cause of concern for any dynamic approach. We mitigate this by
-including only those R packages in our corpus that have many reverse
-dependencies and thus high code coverage. We turn off the bytecode compiler for
-this study. The bytecode compiler can also call eval. We do not get source
-locations for \UndefinedEvalsRnd eval calls. In these cases \eval is either
-passed as an argument to a higher-order functions or is defined in a function
-returned by a higher-order function and the R parser does not retain location
-information for \eval. However, this is a meager \PercentUndefinedEval of all
-eval calls and is unlikely to affect our analysis. We ignore calls to the native
-\eval function exposed by R. We also ignore the \c{rlang::tidy_eval} function
-which uses native \eval internally because \c{rlang} is used to implement a DSL
-for data analysis in R. It introduces a new first-class promise object called
-\c{quosure} for which it implements special evaluation support in \c{tidy_eval}.
+code coverage, a cause of concern for any dynamic approach. We mitigate this
+by including focusing on R packages that have high code coverage. We turn
+off the bytecode compiler for this study. The bytecode compiler can also
+call \eval. We do not get source locations for \UndefinedEvalsRnd eval
+calls. In these cases \eval is either passed as an argument to a
+higher-order functions or is defined in a function returned by a
+higher-order function and the R parser does not retain location information
+for \eval. However, this is a meager \PercentUndefinedEval of all eval calls
+and is unlikely to affect our analysis. We ignore calls to the native \eval
+function exposed by R. We also ignore the \c{rlang::tidy_eval} function
+which uses native \eval internally because \c{rlang} is used to implement a
+DSL for data analysis in R. It introduces a new first-class promise object
+called \c{quosure} for which it implements special evaluation support in
+\c{tidy_eval}.
 

### 2020-08-28 inst/analysis/enrich_dataset.Rmd
AST Size on all the dataset!
Fix call sites per package

@@
 
+```{r}
+eval_calls_ast_size <- eval_calls_corpus %>% select(eval_call_srcref, caller_package, caller_function, eval_source_type, expr_resolved_nodes, nb_ev_calls)
+
+write_fst(eval_calls_ast_size, path(RUN_TRACE_DIR, "summarized-ast-size.fst"))
+```
+
 # Unknown evals
@@
 
+# Known call sites per packages
+
+```{r defined_call_sites}
+call_sites_per_package <- eval_calls_corpus %>% 
+    filter(eval_source_type == "package", eval_source %in% corpus_files) %>%
+    group_by(eval_source) %>%
+    summarize(n = n_distinct(eval_call_srcref))
+
+known_packages <- setdiff(corpus_files, call_sites_per_package$eval_source) %>% as_tibble_col(column_name = "eval_source")
+
+known_packages <- known_packages %>% mutate(n = 0)
+
+call_sites_per_package <- bind_rows(call_sites_per_package, known_packages) %>%
+  rename(package = eval_source) %>% 
+  arrange(desc(n))
+
+write_fst(call_sites_per_package, PACKAGE_EVALS_DYNAMIC_FILE)
+
+```
+
+```{r}
+max_call_sites <- max(call_sites_per_package$n)
+r("max call sites enrich")
+```
+
 # Write
@@
 ```{r write_back}
-write_fst(undefined_per_package, PACKAGE_EVALS_DYNAMIC_FILE)
+write_fst(undefined_per_package, PACKAGE_EVALS_UNDEFINED_FILE)
 

### 2020-08-28 inst/analysis/analysis_stable.Rmd

@@
 library(ggthemes)
+library(pbapply)
 
@@
 eval_calls_env_assign <- eval_calls %>% filter(!is.na(expr_resolved)) %>% mutate(env_assign = pblapply(expr_resolved, is_assign_str, cl = parallel::detectCores() - 2))
+
+#eval_calls_env_assign2 <-  eval_calls_env_assign %>% filter(is.logical(env_assign)) %>% hoist("env_assign", env_assign = 1)
 ```
@@
 
+Number of eval calls in packages.
+
+```{r call_sites_core}
+
+corpus_files <- read_lines(CORPUS_FILE)
+corpus_size <- length(corpus_files)
+
+program_call_sites <- read_fst(PACKAGE_EVALS_FILE) %>% as_tibble() %>%
+  filter(package %in% c(corpus_files, CORE_PACKAGES))
+
+in_core <- program_call_sites %>% filter(package %in% CORE_PACKAGES)
+
+nb_in_core <- in_core %>% nrow()
+
+r("nb core eval call sites", nb_in_core)
+
+call_sites_per_package <- program_call_sites %>% count(package, sort = TRUE)
+
+max_call_sites <- max(call_sites_per_package$n)
+
+r("max call site package", max_call_sites)
+
+nb_one_call_site <- call_sites_per_package %>% filter(n == 1) %>% nrow()
+
+r("one call site package", nb_one_call_site)
+
+nb_packages_plus_one <- n_distinct(program_call_sites %>% filter(!package %in% CORE_PACKAGES) %>% select(package))
+
+nb_package_none_eval_call_site <- corpus_size - nb_packages_plus_one
+
+r("no call site packages", nb_package_none_eval_call_site)
+```
+
 Notebook execution was `r duration` `r units(duration)`.

### 2020-08-28 inst/paper/main.tex
Fix number of call sites per traced packages
Typos

--- 2020-08-28 inst/paper/main.tex
@@
 Table~\ref{table:site-package-summary} shows the number of call sites in
-CRAN packages. 61 packages have a single call site, and one package, \emph{VGAM}, has \MaxCallSitePackage
-call sites of \eval. \NoCallSitePackages packages have no calls to \eval. The core packages
+CRAN packages. \OneCallSitesEnrich packages have a single call site, and one package, \emph{metafor}, has \MaxCallSitesEnrich
+call sites of \eval. \NoCallSitesEnrich packages have no calls to \eval. The core packages
 have \NbCoreEvalCallSites call sites of \eval and Kaggle has none. These numbers are
@@
 
-In our corpus, we observe 5.2\% cases where \eval directly evaluates output of
+In our corpus, we observe 5.2\% cases where \eval directly evaluates the output of
 \c{substitute}, 0.7\% cases where output of \c{parse} and its variants is read
@@
 \PercentParsedEvals of the eval calls. We observed that
-very few of the eval call(\NbParseFilesRnd in total) consume the result of calling \c{parse} on
+very few of the eval call (\NbParseFilesRnd in total) consume the result of calling \c{parse} on
 a file. Most of the eval calls consume the result of calling
@@
 by \eval are side-effecting. These writes outlive the computation and hence are
-visible outside the computation. The remaining writes are local to the
+visible outside it. The remaining writes are local to the
 computation. We observe that \EvalSideEffectingWritesRnd writes inside evals are
@@
 This argument can also be \c{NULL}, a \c{List} or a data frame. This happens in
-3.5\% of the cases.
-\c{eval} copies the fields of the list or data frame and creates bindings for
+3.5\% of the cases: \c{eval} copies the fields of the list or data frame and creates bindings for
 them in a new environment. This pattern is used to evaluate formulas which can
@@
 
-\begin{table}[htbp]
+\begin{table}[htbp] \resizebox{\columnwidth}{!}{%
     \centering
@@
 
-We observe that a disproportionately high core R \eval calls access the
+We observe that a disproportionately high number of core R \eval calls access the
 caller's caller's environment. This is because many core R functions call
 functions that pass the result of \c{parent.frame()} to an \eval.
-A disproportionately high calls to \eval happen in an extended
+A disproportionately high number of calls to \eval happens in an extended
 top-level environment. This can be explained by the fact that many packages

### 2020-08-28 inst/paper/main.tex
even more patterns

@@
 
-  \subsubsection{Convenience}
-  evaling recorded calls in model-fitting functions or passing most of the
-  current call to another function. Both ``patterns'' use the \c{match.call} to
-  capture current call. The latter is just for laziness as it is the same as
-  calling the new function with explicitly selecting the arguments.
+  \subsubsection{Convenience} Eval is used to make some programming tasks easier (and/or shorter). For example, in \c{data.table}, it is used to set R options from a character vector
+  \begin{lstlisting}
+opts = c("datatable.verbose"="FALSE", ...)
+for (i in names(opts)) eval(parse(text=paste0("options(",i,"=",opts[i],")")))
+
+  \end{lstlisting}
+  %
+  instead of manually calling \c{options(datatable.verbose=FALSE)}
 
@@
   dependencies.
-  % \AllValueInputEvalCallPerc migth hit on a wong usage
+
+  \subsubsection{Unnecessary use of eval} Similarly to JavaScript, there are also unnecessary uses of \eval.
+  For example, the \c{PerfromanceAnalytics} package contains a function \c{chart.QQPlot} that uses \eval to resolve a string into function and anther to call it and assign its results into a variable:
+  \begin{lstlisting}
+function (R, d="norm", dp, ...) {
+q.f <- eval(parse(text=paste("q",d,sep="")))
+z <- NULL
+eval(parse(text=paste("z<-q.f(",dp,",...)")))
+}
+  \end{lstlisting}
+  In both cases, there is no need for \eval:
+  \begin{lstlisting}
+function (R, d="norm", dp, ...) {
+q.f <- get(paste0("q",d))
+z <- q.f(dp, ...)
+}
+  \end{lstlisting}
+  or even to a oneliner \c{do.call(paste0("q",d), as.list(dp, ...))}.
 

### 2020-08-28 inst/paper/main.tex
Typos

--- 2020-08-28 inst/paper/main.tex
@@
 which were detected using simple regular expressions on the \eval strings. In
-the case of R we match the patterns on the recorded AST.
+the case of R, we match the patterns on the resolved AST.
 
@@
 access to an object's property or local variable both replaceable by JavaScript
-core functions. In the case of R there are \PatternReadRnd \eval calls
+core functions. In the case of R there are \PatternReadRnd read \eval calls
 (\PatternReadRatio). They come from calls like \c{subset(airquality, select =
@@
 adjust some arguments and pass it to another function. For example the
-\c{write.csv} function changes the \c{sep} argument in the matched call and reevaluate itself as \c{write.table}:
+\c{write.csv} function changes the \c{sep} argument in the matched call and reevaluates itself as \c{write.table}:
 %

### 2020-08-28 inst/analysis/patterns.Rmd
more patterns

@@
 ```{r}
-packages <- c(
+rnd_packages <- tibble(package=c(
   "timeSeries",
@@
   "prodlim",
-  "vars" ,
+  "data.table" ,
   "betareg"   ,
@@
   "ggplot2",
-  "binom",
+  "anytime",
   "rmarkdown" ,
@@
   "PerformanceAnalytics"
-)
+))
 ```
@@
 ```{r}
-evals_static <- read_fst(PACKAGE_EVALS_FILE)
+evals_static <- read_fst(PACKAGE_EVALS_FILE) %>% as_tibble() %>% semi_join(rnd_packages, by="package")
 ```
 
-### parse
-
 ```{r}
-all_data %>%
-  filter(!is.na(parse_args_text))
+overview_table(
+  r("Manual packages", nrow(rnd_packages)),
+  r("Manual callsites", nrow(evals_static))
+)
 ```

### 2020-08-28 inst/paper/main.tex
fix patterns

@@
 Table~\ref{table:site-package-summary} shows the number of call sites in
-CRAN packages. \OneCallSitesEnrich packages have a single call site, and one package, \emph{metafor}, has \MaxCallSitesEnrich
-call sites of \eval. \NoCallSitesEnrich packages have no calls to \eval. The core packages
-have \NbCoreEvalCallSites call sites of \eval and Kaggle has none. These numbers are
-under-approximations as \eval may be called reflectively, through aliases,
-or passed to a higher-order function.
+CRAN packages. \OneCallSitesEnrich{} packages have a single call site, and one
+package, \emph{metafor}, has \MaxCallSitesEnrich{} call sites of
+\eval. \NoCallSitesEnrich{} packages have no calls to \eval. The core packages
+have \NbCoreEvalCallSites{} call sites of \eval and Kaggle has none. These
+numbers are under-approximations as \eval may be called reflectively,
+through aliases, or passed to a higher-order function.
 
@@
 \c{String Vector}      &                           17487     & 1.1\%\\
-  \end{tabular}
-  \caption{Top ten eval calls in CRAN}
-  \label{A}
-\end{table*}
-
-The expression \c{Environment} occurs because of the four callsites explained
-above, \c{ggplot2::ggproto}, \c{R6::generator_funs}, \c{future::backtrace} and
-\c{RModel::str.RMmodel}. The next expression, \c{column[rows] <<- what}, is used
-inside the \c{plyr::rbind.fill} function to merge data frames by assigning
-concatenated vectors to rows. The \c{<<-} operator is interesting in that it
-skips the current scope and assigns in a parent scope in which the variable is
-already present. In our corpus, all these \eval calls contribute to a single
-side-effect. The expression \c{function(value) freduce(value, `_function_list`)}
-arises from the \c{magrittr::\%>\%} function which is a pipe operator that pipes
-the output of previous command to the next one. The expression is evaluated in a
-custom environment to create a function binding for evaluating the components of
-the pipe.\c{String Vector} and \c{List} also arise from the same function when a
-string or a list is piped using the \c{\%>\%} function into the next expression.
-The \c{NULL} arises from \c{R6::generator_funs} function when the \eval is
-passed a \c{NULL} argument by the \c{DataMask_generator} package. The
-\c{c("default", "default2012", "default2011" ...} pattern arises from
-\c{copula::polyG} where it reflectively access the default expression for its
-formal parameter and evaluates it. The \c{force(..1)} and \c{alist(`_spec`)}
-patterns occur in \c{glue::glue_data} function which concatenates and
-interpolates strings. The two patterns occur because the function captures
-unevaluated unnamed arguments and maps the evaluation of \c{force(..1)} on them.
-The \c{force} function forces promises~\cite{oopsla19a} and returns the result
-of evaluation. The \c{inner} pattern arises from \c{glue::identity_transformer}
-which enables the creation of custom transformation functions for affecting the
-interpolation and concatenation of input by the \c{glue} package.
+\end{tabular}\caption{Top ten eval calls in CRAN} \label{C}
+\end{table}
+
+The expression \c{Environment} occurs because of the four callsites
+explained above, \c{ggplot2::ggproto}, \c{R6::generator_funs},
+\c{future::backtrace} and \c{RModel::str.RMmodel}. The next expression,
+\c{column[rows] <<- what}, is used inside the \c{plyr::rbind.fill} function
+to merge data frames by assigning concatenated vectors to rows. The \c{<<-}
+operator is interesting in that it skips the current scope and assigns in a
+parent scope in which the variable is already present. In our corpus, all
+these \eval calls contribute to a single side-effect. The expression
+\c{function(value) freduce(value, `_function_list`)} arises from the
+\c{magrittr::\%>\%} function which is a pipe operator that pipes the output
+of previous command to the next one. The expression is evaluated in a custom
+environment to create a function binding for evaluating the components of
+the pipe.\c{String Vector} and \c{List} also arise from the same function
+when a string or a list is piped using the \c{\%>\%} function into the next
+expression.  The \c{NULL} arises from \c{R6::generator_funs} function when
+the \eval is passed a \c{NULL} argument by the \c{DataMask_generator}
+package. The \c{c("default", "default2012", "default2011" ...} pattern
+arises from \c{copula::polyG} where it reflectively access the default
+expression for its formal parameter and evaluates it. The \c{force(..1)} and
+\c{alist(`_spec`)} patterns occur in \c{glue::glue_data} function which
+concatenates and interpolates strings. The two patterns occur because the
+function captures unevaluated unnamed arguments and maps the evaluation of
+\c{force(..1)} on them.  The \c{force} function forces promises and returns
+the result of evaluation. The \c{inner} pattern arises from
+\c{glue::identity_transformer} which enables the creation of custom
+transformation functions for affecting the interpolation and concatenation
+of input by the \c{glue} package.
 
 \subsection{Operation Mix}
-In this section, we look at the operations executed by the expressions and
-symbols on evaluation. R is a lazy language, so symbols can be bound to promises
-that can evaluate arbitrary code. All operations in R are function calls, even
-primitive operations such as addition and assignment are treated as calls to
-functions implemented in the interpreter. In our corpus, we observe 2.9G calls
-to R functions of which 125.5M happen inside \eval. Furthermore, unlike
-Javascript, R code also calls into peformant native implementations of numerical
-and statistical routines. In the corpus there are 463.8M calls to native code of
-which only 31.2M happens inside \eval.
-Qualitatively, we observe all kinds of operations being executed inside \eval
-calls, from numerical computing for statistical modeling to reflection and
-metaprogramming for implementing DSLs. Even testing frameworks such \c{genthat}
-heavily rely upon \eval to execute test code snippets. The core R implementation
-is rife with uses of \eval for mundane things like accessing default values of
-formal parameters and in statistical packages for computation.
-
-\subsubsection{Provenance}
-
-In this section we turn our attention to the \AllExpressionInputEvalCallPerc
-expressions passed as input to \eval. There are two ways to obtain
-expressions in R.
+
+The JavaScript paper reported on the operation mix within \evals. This means
+the proportion of read, write, and call bytecodes performed by the
+JavaScript interpreter. They found that some web pages were creating objects
+more frequently and writing into their fields.  In R, nearly every operation
+performed in the interpreter is a function call. We observe 2.9G calls and,
+of those 125M happen inside \eval (there are 463M native calls, of which
+only 31M are in \eval).
+
+{\bf there is no good reason for not counting reads and writes, we can...}
+
+%% ...The following is fluff ...
+
+%% Qualitatively, we observe all kinds of operations being executed inside
+%% \eval calls, from numerical computing for statistical modeling to reflection
+%% and metaprogramming for implementing DSLs. Even testing frameworks such
+%% \c{genthat} heavily rely upon \eval to execute test code snippets. The core
+%% R implementation is rife with uses of \eval for mundane things like
+%% accessing default values of formal parameters and in statistical packages
+%% for computation.
+
+\subsection{Provenance}
+
+In JavaScript, the provenance of the string passed into \eval is a matter of
+concern as \eval can be used to breach the security of websites. The main
+danger comes from string that were obtained from external sources. If they
+are allowed to flow into an \eval, then an attacker could hijack the
+program.
+
+{\bf what have they found...}
+
+There are various ways to obtain expressions:
 \begin{itemize}
-  \item \c{substitute} function can be used to synthesize ASTs from
-  expressions by replacing symbols with their bindings in the specified
-  environment.
-  \item \c{expression} function can be used to create a vector of expression
+  \item \c{substitute} synthesizes ASTs from expressions by replacing
+    symbols with their bindings in the specified environment.
+  \item \c{expression}  creates a vector of expression
     objects from text.
-  \item \c{parse} function and its variants \c{str2expression} and \c{str2lang}
-    can be used to turn strings or text from file into R expressions.
+  \item \c{parse}, \c{str2expression} and \c{str2lang} turn strings into
+    expressions.
 \end{itemize}
 
-In our corpus, we observe 5.2\% cases where \eval directly evaluates the output of
-\c{substitute}, 0.7\% cases where output of \c{parse} and its variants is read
-directly and only 704 cases for \c{expression}. Majority of expression objects
-consumed by \eval are generated by their transitive uses by other functions.
-
-The pattern \c{eval(parse(...))} can be used for dynamic code loading. This
-forms the core of \c{source} and \c{sys.source} functions in R that are commonly
-used for loading code in R files in interactive settings. We investigated the
-number of cases in which the output of \c{parse} and its variants is passed to
-\eval, directly or transitively by tainting their output. This corresponds to \PercentParsedCallSites of the total \eval call sites and
-\PercentParsedEvals of the eval calls. We observed that
-very few of the eval call (\NbParseFilesRnd in total) consume the result of calling \c{parse} on
-a file. Most of the eval calls consume the result of calling
-\c{parse} on a string.  We also identified one
-function in core R, \c{invokeRestartInteractively} that prompts the user for
-input, parses it, and passes it to \eval.
-
-%%Unlike Javascript, R packages also call native code for access to efficient
-%%numerical and statistical routines. We intercept all calls to R functions and
-%%native functions from the programs and \eval calls. In the whole corpus there
-%%are \AllClosureCallCount calls to R functions and \AllNativeCallCount to native
-%%code. Of those, \eval contribute to only \EvalClosureCallCount calls to R
-%%functions and \EvalNativeCallCount to native code.
-%%
-%%\begin{table}[ht]\resizebox{\columnwidth}{!}{%
-%%    \begin{tabular}{lll}\hline
-%%      Source & Primitive Functions & R Functions & Native Functions \\ \hline
-%%      Core & \CorePrimitiveFunctionCount & \CoreRFunctionCount & \CoreNativeFunctionCount \\
-%%      CRAN & \PackagePrimitiveFunctionCount & \PackageRFunctionCount & \PackageNativeFunctionCount \\
-%%      All  & \AllPrimitiveFunctionCount & \AllRFunctionCount & \AllNativeFunctionCount \\
-%%    \end{tabular}} \label{table:operation-distribution}
-%%  \caption{Distribution of operation calls to \eval calls across Core and Package}
-%%\end{table}
+We observe 5.2\% cases where \eval directly evaluates the output of
+\c{substitute}, 0.7\% cases where output of \c{parse} is read directly and
+only 704 cases for \c{expression}. Most expressions consumed by \eval are
+generated by other functions.
+
+\c{eval(parse(...))} can be used for dynamic code loading. This forms the
+core of \c{source} and \c{sys.source} functions in R that are commonly used
+for loading code in R files in interactive settings. We investigated the
+number of cases in which the output of \c{parse} and its variants is passed
+to \eval, directly or transitively by tainting their output. This
+corresponds to \PercentParsedCallSites of the total \eval call sites and
+\PercentParsedEvals of the eval calls. We observed that very few of the eval
+call (\NbParseFilesRnd in total) consume the result of calling \c{parse} on
+a file. Most of the eval calls consume the result of calling \c{parse} on a
+string.  We also identified one function in core R,
+\c{invokeRestartInteractively} that prompts the user for input, parses it,
+and passes it to \eval.
 
@@
 dangerous. Only writes to environments not local to the computation spawned
-by \eval are side-effecting. These writes outlive the computation and hence are
-visible outside it. The remaining writes are local to the
-computation. We observe that \EvalSideEffectingWritesRnd writes inside evals are
-side-effecting. This is only \EvalSideEffectingWritesEvalPerc of all variable
-writes inside \eval and \EvalSideEffectingWritesAllPerc of all variable writes
-in the corpus.
+by \eval are side-effecting. These writes outlive the computation and hence
+are visible outside it. The remaining writes are local to the
+computation. We observe that \EvalSideEffectingWritesRnd writes inside evals
+are side-effecting. This is only \EvalSideEffectingWritesEvalPerc of all
+variable writes inside \eval and \EvalSideEffectingWritesAllPerc of all
+variable writes in the corpus.
 
@@
 One of the main differences of the R \eval with the javascript one is how it
-interacts with \emph{scope}, or environments in R. In Javascript, \c{eval} can
-access the local scope and the global scope. In R, \c{eval} has an argument
+interacts with \emph{scope}, or environments in R. In Javascript, \eval can
+access the local scope and the global scope. In R, \eval has an argument
 \c{envir} to indicate in which environment its first argument must be evaluated.
 By default, it is equal to \c{parent.frame()}, 
-which is the parent environment of \c{eval} call. This default argument
+which is the parent environment of \eval call. This default argument
 happens in \DefaultEnvirExprPercent of the eval calls.
 
-This argument can also be \c{NULL}, a \c{List} or a data frame. This happens in
-3.5\% of the cases: \c{eval} copies the fields of the list or data frame and creates bindings for
-them in a new environment. This pattern is used to evaluate formulas which can
-directly refer to the fields of the data.
-The \c{envir} argument can also be a number $n$. It means that the environment in
-which the expressions is evaluated will be the result of \c{sys.call(n)} where
-$n$ refers to the $n$-th stack frame.
+This argument can also be \c{NULL}, a \c{List} or a data frame. This happens
+in 3.5\% of the cases: \eval copies the fields of the list or data frame
+and creates bindings for them in a new environment. This pattern is used to
+evaluate formulas which can directly refer to the fields of the data.  The
+\c{envir} argument can also be a number $n$. It means that the environment
+in which the expressions is evaluated will be the result of \c{sys.call(n)}
+where $n$ refers to the $n$-th stack frame.
 
@@
 We looked at the environments passed to all the \eval calls in our corpus.
-Table~\ref{tab:environments} summarizes the results. A numeric environment class
-\c{n} denotes the environment of the $n$-th call stack frame from the current
-function. \c{global} denotes the top-level environment and \c{list} denotes a list
-passed for evaluation of formulas. Environment classes of the form $n+$ denote
-the $n$-th environment extended with a new environment. The new environment
-provides a limited form of sandboxing. All assignments using the \c{\<-}
-function occur inside it and prevent the extended environment from
-mutation. However, it is still possible to mutate the extended environment
-using the \c{\<\<-} or \c{assign} functions; but, that happens rarely.
+Table~\ref{tab:environments} summarizes the results. A numeric environment
+class \c{n} denotes the environment of the $n$-th call stack frame from the
+current function. \c{global} denotes the top-level environment and \c{list}
+denotes a list passed for evaluation of formulas. Environment classes of the
+form $n+$ denote the $n$-th environment extended with a new environment. The
+new environment provides a limited form of sandboxing. All assignments using
+the \c{\<-} function occur inside it and prevent the extended environment
+from mutation. However, it is still possible to mutate the extended
+environment using the \c{\<\<-} or \c{assign} functions; but, that happens
+rarely.
 
@@
 
-We observe that a disproportionately high number of core R \eval calls access the
-caller's caller's environment. This is because many core R functions call
-functions that pass the result of \c{parent.frame()} to an \eval.
-A disproportionately high number of calls to \eval happens in an extended
-top-level environment. This can be explained by the fact that many packages
-evaluate code passed from the user's workspace in the top-level environment to
-access the bindings.
-
-%%Below, we provide some code snippets to show how some of these environment
-%%classes manifest.
-
-%%The \c{match.arg} (see Listing~\ref{code:matchargenvir}) function has an
-%%environment class 2. \c{sys.parent()} refers to the parent environment of
-%%\c{match.arg} and so to the grand-parent environment of \c{eval} which we denote
-%%as $2$.
-%%
-%%\begin{lstlisting}[label=code:matchargenvir, caption={\c{match.arg} using \eval with a custom environment.}]
-%%formal.args <- formals(sys.function(sysP <- sys.parent()))
-%%choices <- eval(formal.args[[as.character(substitute(arg))]], envir = sys.frame(sysP))
-%%\end{lstlisting}
-
-%%In package \emph{gamlss::get.K} function \c{get.K} evaluates its expression in its
-%%parent environment, the one of the body of \c{get.K} returned from
-%%\c{environment()}.
-%%
-%%\begin{lstlisting}
-%%eval(parse(text = paste(\"fam$dld\", ii, sep = \"\")), envir = environment())
-%%\end{lstlisting}
-%%
-%%
-%%The \emph{glue} library is used to interpolate string litterals with variables, such as in \c{glue('My name is \{name\}.')} where \c{name} will be subtituted by its value.
-\emph{glue} makes it possible to define transformers that process those interpolated variables.
-They are built upon the \c{identity_transformer}, which uses \eval.
-The interpretation environment for \eval is often global here, as the string litterals are often used to print variables at the top level.
-%%
-%%\begin{lstlisting}
-%%identity_transformer <- function(text, envir) 
-%%{
-%%    eval(parse(text = text, keep.source = FALSE), envir)
-%%}
-%%\end{lstlisting}
-%%
-%%
-%% \end{lstlisting}
-%% $$$$ <- to fix emacs fontification issues
-
-The \c{imchange} function of package \c{imager} makes it possible to modify images using a dedicated formula syntax using \c{\~}.\footnote{Inspired by \c{map} in package \emph{purr}.}  Here, \eval is evaluated in \c{newenv}, which creates a new environment that inherits from \c{parent.frame()} by default (classified as 1+).
+We observe that a disproportionately high number of core R \eval calls
+access the caller's caller's environment. This is because many core R
+functions call functions that pass the result of \c{parent.frame()} to an
+\eval.  A disproportionately high number of calls to \eval happens in an
+extended top-level environment. This can be explained by the fact that many
+packages evaluate code passed from the user's workspace in the top-level
+environment to access the bindings.
+
+The \c{imchange} function of package \c{imager} makes it possible to modify
+images using a dedicated formula syntax using \c{\~}.\footnote{Inspired by
+ {map} in package \emph{purr}.}  Here, \eval is evaluated in \c{newenv},
+which creates a new environment that inherits from \c{parent.frame()} by
+default (classified as 1+).
 
@@
 
-The JavaScript paper identified a number of frequently occurring \eval patterns
-which were detected using simple regular expressions on the \eval strings. In
-the case of R, we match the patterns on the resolved AST.
+Richards et al.~\cite{ecoop11} identified a number of frequently occurring \eval
+patterns which were detected using simple regular expressions on the \eval
+strings. In the case of R we match the patterns on the recorded AST.
 
@@
 \end{tabular}
-\caption{JavaScript evals patterns in R}
+\caption{JavaScript \eval Patterns in R}
 \label{table:js-pattens-in-r}
@@
 
-%\noindent
-%\begin{compactitem}[$-$]
-  % \item \emph{Library}: in JavaScript, the library pattern captures asynchronous
-  % loading of external external function definitions. In R we have seen only
-  % \PatternLibRnd such calls (\PatternLibRatio) mostly coming from source loading
-  % functions (\ie \c{source} and \c{sys.source}). Most of the external function
-  % definitions in R come from packages and R makes it very easy for users to
-  % install packages from its centralized repository.
-
 \noindent \emph{Library}, \emph{Typeof} and \emph{Try} are all barely used
 mostly because the language provides dedicated features for it. The few calls in
-Library pattern come mostly from file sourcing functions (\ie \c{source} and \c{sys.source}).
+Library pattern come mostly from file loading functions such as \c{source} and
+\c{sys.source}.
 %
 \noindent \emph{Read} pattern in JavaScript represented mostly an incorrect
-access to an object's property or local variable both replaceable by JavaScript
-core functions. In the case of R there are \PatternReadRnd read \eval calls
-(\PatternReadRatio). They come from calls like \c{subset(airquality, select =
-  Temp)} where the variable \c{Temp} will be evaluated in the \c{airquality}
-data frame.
+
+access to an object's property or local variable both replaceable by
+JavaScript core functions. In the case of R there are \PatternReadRnd read
+\eval calls (\PatternReadRatio). They come from calls like
+\c{subset(airquality, select = Temp)} where the variable \c{Temp} will be
+evaluated in the \c{airquality} data frame.
+
 %
-\noindent \emph{Assign} category comprises of assignments to variables within
-an \eval call. We have seen \PatternAssignRnd such calls (\PatternAssignRatio).
-\PatternAssignArrowRatio uses the simple assignment (\c{<-}). The majority of
-cases comes from the \c{magrittr}'s \c{\%<>\%} pipe operator
-(\PatternAssignArrowMagrittrRatio) which forwards an object to a function on rhs
-and update lhs with the result. Far more calls used the super assignment
-(\c{<<-}) operator, but almost all come from the \c{plyr} package which provides
-a convenient API for common data manipulation problems.
+\noindent \emph{Assign} category comprises of assignments to variables
+within an \eval call. We have seen \PatternAssignRnd such calls
+(\PatternAssignRatio).  \PatternAssignArrowRatio uses the simple assignment
+(\c{<-}). The majority of cases comes from the \c{magrittr}'s \c{\%<>\%}
+pipe operator (\PatternAssignArrowMagrittrRatio) which forwards an object to
+a function on rhs and update lhs with the result. Far more calls used the
+super assignment (\c{<<-}) operator, but almost all come from the \c{plyr}
+package which provides a convenient API for common data manipulation
+problems.
 %
-\noindent \emph{Call} and \emph{Other} together represents the vast majority of
-eval calls in R. The \emph{Other} pattern represents all the calls where eval
-was called with a value instead of with an expression. The \emph{Call} pattern
-contains all the calls that did not fall in the previous categories. Without a
-context it is usually difficult to further categorize the intent of these calls.
-Therefore we have randomly chosen \PatternManualPackages packages and manually
-looked at \PatternManualCallsites call sites from which we have observed the following categories:
-
-% this is read pattern
-% \subsubsection{Evaluation using Data Fields} A common use of \eval is to
-%   evaluate expressions that access data frame columns or list (heterogeneous
-%   vectors) elements by name. Core R provides two functions, \c{with} and
-%   \c{within} that implement this pattern. These functions construct a temporary
-%   environment with data frame columns or list elements as the bindings.
-%   Side-effects performed by the evaluated expression take place in this temporary
-%   environment and are ignored by the \c{with} function. The \c{within} function
-%   copies these modifications to a local copy of the input data and returns it.
-%   This pattern is used quite often for statistical modeling. The modeling
-%   expressions is constructed from variable names whose values are supplied by a
-%   data frame or list.
-\subsubsection{Reconding the current call} R provides a \c{match.call} function
+\noindent \emph{Call} and \emph{Other} together represents the vast majority
+of eval calls in R. The \emph{Other} pattern represents all the calls where
+eval was called with a value instead of with an expression. The \emph{Call}
+pattern contains all the calls that did not fall in the previous
+categories. Without a context it is usually difficult to further categorize
+the intent of these calls.  Therefore we have randomly chosen
+\PatternManualPackages packages and manually looked at
+\PatternManualCallsites call sites from which we have observed the following
+categories:
+
+\subsubsection{Recording the current call} R provides a \c{match.call} function
 which can return the current call. This is heavily used by the statistical
-model-fitting functions that records the current call that is then reevaluated
-in the different environment. Another pattern is to match the current call,
-adjust some arguments and pass it to another function. For example the
-\c{write.csv} function changes the \c{sep} argument in the matched call and reevaluates itself as \c{write.table}:
+model-fitting functions that records the current call that is then
+reevaluated in the different environment. Another pattern is to match the
+current call, adjust some arguments and pass it to another function. For
+example the \c{write.csv} function changes the \c{sep} argument in the
+matched call and reevaluates itself as \c{write.table}:
 %
@@
 used to parse text from string or file and evaluate in a custom environment.
-  This pattern is used by the \source Core R function to load R code from a file
-  in the current workspace. This is quite useful for loading functions for
-  interactive data analysis.
+This pattern is used by the \source Core R function to load R code from a
+file in the current workspace. This is quite useful for loading functions
+for interactive data analysis.
 
 \subsubsection{Metaprogramming} The combination of \eval and \substitute is
-  used to do metaprogramming in R. \substitute can be used to synthesize ASTs by
-  stitching together user inputs. \substitute walks over its argument AST and
-  replaces the symbols with their bindings in the specified environment and
-  returns this new AST which can then be evaluated in custom environments using
-  \eval.
+ used to do metaprogramming in R. \substitute can be used to synthesize ASTs
+ by stitching together user inputs. \substitute walks over its argument AST
+ and replaces the symbols with their bindings in the specified environment
+ and returns this new AST which can then be evaluated in custom environments
+ using \eval.
 
@@
 
-  \subsubsection{Unnecessary use of eval} Similarly to JavaScript, there are also unnecessary uses of \eval.
-  For example, the \c{PerfromanceAnalytics} package contains a function \c{chart.QQPlot} that uses \eval to resolve a string into function and anther to call it and assign its results into a variable:
+  \subsubsection{Unnecessary use of eval}
+  Similarly to JavaScript, there are also unnecessary uses of \eval.
+  For example, the \c{PerformanceAnalytics} package contains a function
+  \c{chart.QQPlot} that uses \eval to resolve a string into function and
+  another to call it and assign its results into a variable:
   \begin{lstlisting}

### 2020-08-28 inst/paper/main.tex
Typos

--- 2020-08-28 inst/paper/main.tex
@@
 %   data frame or list.
-\subsubsection{Reconding the current call} R provides a \c{match.call} function
+\subsubsection{Recording the current call} R provides a \c{match.call} function
 which can return the current call. This is heavily used by the statistical
@@
   \subsubsection{Unnecessary use of eval} Similarly to JavaScript, there are also unnecessary uses of \eval.
-  For example, the \c{PerfromanceAnalytics} package contains a function \c{chart.QQPlot} that uses \eval to resolve a string into function and anther to call it and assign its results into a variable:
+  For example, the \c{PerformanceAnalytics} package contains a function \c{chart.QQPlot} that uses \eval to resolve a string into function and another to call it and assign its results into a variable:
   \begin{lstlisting}

### 2020-08-29 inst/paper/main.tex
Tidying
remove dots from numnbers

--- 2020-08-28 inst/paper/main.tex
@@
 \newcommand{\UndefinedOutOfCoprusPercent}{0\%\xspace}
-\newcommand{\MaxCallSitesEnrich}{114\xspace}
-\newcommand{\OneCallSitesEnrich}{61\xspace}
-\newcommand{\NoCallSitesEnrich}{297\xspace}
+\newcommand{\MaxCallSitesEnrich}{114}
+\newcommand{\OneCallSitesEnrich}{61}
+\newcommand{\NoCallSitesEnrich}{297}
 %\input{tag/patterns.tex}
@@
 
-\newcommand{\eval}{\texttt{eval}}
-\newcommand{\evals}{{\sf eval}s\xspace}
+\newcommand{\eval}{\c{eval}}
 \newcommand{\parse}{\c{parse}}
@@
 yields the parse tree of the expression \c{e} after performing substitutions
-defined by the bindings in the \c{envir} parameter.
+defined by the bindings in \c{envir}.
 
@@
 function. There are three alternative forms  of \eval in the \c{base} package:
-\c{evalq}, \c{eval.parent} and \c{local}.
-
-\eval function in R takes three arguments as shown below.
+\c{evalq}, \c{eval.parent}, \c{local}.
+
 \begin{lstlisting}
@@
 
-\c{expr} is the value to be evaluated, \c{envir}
+Arguments are as follows: \c{expr} is the value to be evaluated, \c{envir}
 is the environment in which it is to be evaluated, and \c{enclos} is
@@
 When called, \eval evaluates \c{expr} in the environment specified by
-\c{envir} and returns the computed value. If \c{envir} is missing, then the
+\c{envir} and return the computed value. If \c{envir} is missing, then the
 default is \c{parent.frame()} (the environment where the call to \eval was
@@
 
+\begin{figure*}[!tb]\centering\includegraphics[width=.8\linewidth]
+  {corpus.pdf}\caption{CRAN packages}\label{fig:corpus}
+\end{figure*}
+
 All steps of this pipeline are parallelized using GNU
@@
 assemble it.
-
-\begin{figure}[!h]\centering\includegraphics[width=\linewidth]
-  {corpus.pdf}\caption{CRAN packages}\label{fig:corpus}
-\end{figure}
 
@@
 lines of R code: \CorpusExamplesProgramsRnd examples with
-\CorpusExamplesCodeRnd lines of code, and \CorpusVignettesProgramsRnd
-vignettes.
+\summaryrnd{CorpusExamplesCode}\footnote{This notation represents $sum$
+  ($min$ / $mean$ / $max$)} lines of code, and \CorpusVignettesProgramsRnd
+vignettes.%%% with \summaryrnd{CorpusVignettesCode} lines of code.
 
@@
 
-To represent end-user code in the corpus, we turned to Kaggle, an online
-platform for data-science and machine-learning. The website allows people to
-share data-science competitions, a data-analysis problem together with data,
-for which users try to find the best solution. The solutions, called
-{kernels}, are then shared back to the platform in the form of either plain
-scripts or as notebooks.  One of the most popular competition is about
-predicting passenger survival on
+To represent end-user code in the corpus, we turned to
+Kaggle,\footnote{\url{https://www.kaggle.com}} an online platform for
+data-science and machine-learning. The website allows people to share
+data-science competitions, a data-analysis problem together with data, for
+which users try to find the best solution. The solutions, called {kernels},
+are then shared back to the platform in the form of either plain scripts or
+as notebooks.  One of the most popular competition is about predicting
+passenger survival on
 Titanic\footnote{\url{https://www.kaggle.com/c/titanic}} with \CorpusKaggle
@@
 \CorpusKaggle kernels and extracting the R code from the various
-formats,\footnote{We use {\sf rmarkdown} to convert from notebooks to R.} we
-found that \CorpusDuplicatedKaggle were whole-file duplicates. From
+formats,\footnote{We use rmarkdown package to convert from Jupyer notebook
+  to Rmarkdown as well as to convert from Rmarkdown to plain R.} we found
+that \CorpusDuplicatedKaggle were whole-file duplicates. From
 \CorpusRunnableKaggle kernels, \CorpusFailedKaggle failed to execute. Next
@@
 misspelled package names.  The final set contains \CorpusFinishedKaggle
-kernels with \CorpusFinishedKaggleCodeRnd lines of R code.
+kernels with \summaryrnd{CorpusFinishedKaggleCode} lines of R code.
 
@@
 
-This section presents an overview of the usage of \eval in our corpus. Richards
-et al. reported 82\% of JavaScript-enabled web pages using \eval~\cite{ecoop11}.
-The case for R is more striking with 100\% of the programs call the \eval
-function.\footnote{We combine calls to {\sf eval}, {\sf eval.parent}, {\sf
-    evalq} and {\sf local}.} This high number is explained by the fact that the
-R implementation, its core packages, use \eval extensively. We differentiate
-three types of eval based on their origin:
-%
-\begin{compactitem}[$-$]
-  \item a {\bf core eval} is one that occurs in the code of a core package,
-  \item a {\bf dependent eval} occurs in a package imported by the target
-  package, and
-  \item an {\bf own eval} is one that occurs in the source code of the target
-  package.
-\end{compactitem}
+This section presents an overview of the usage of \eval in our corpus.
+Richards et al. reported 82\% of JavaScript-enabled web pages using
+\eval~\cite{ecoop11}. The case for R is more striking with 100\% of the
+programs call the \eval function.\footnote{We combine calls to \c{eval},
+  \c{eval.parent}, \c{evalq} and \c{local}.}  This high number is explained
+by the fact that the R implementation, its core packages, use \eval
+extensively. We differentiate uses based on their origin. A {\bf core} \eval
+is one that occurs in the code of a core package.  A {\bf dependent} \eval
+occurs in a package imported by the target package. An {\bf own} \eval is
+one that occurs in the source code of the target package.
 
@@
 %% SORRY  I HAD TO REFORMAT
-\begin{table}[!h] \centering
+\begin{table}[tb] \centering
  \begin{tabular}{r|c|c|c} \hline
@@
 Table~\ref{table:site-package-summary} shows the number of call sites in
-CRAN packages. \OneCallSitesEnrich{} packages have a single call site, and one
-package, \emph{metafor}, has \MaxCallSitesEnrich{} call sites of
-\eval. \NoCallSitesEnrich{} packages have no calls to \eval. The core packages
-have \NbCoreEvalCallSites{} call sites of \eval and Kaggle has none. These
-numbers are under-approximations as \eval may be called reflectively,
-through aliases, or passed to a higher-order function.
+CRAN packages. \OneCallSitesEnrich packages have a single call site, and one package, \emph{metafor}, has \MaxCallSitesEnrich
+call sites of \eval. \NoCallSitesEnrich packages have no calls to \eval. The core packages
+have \NbCoreEvalCallSites call sites of \eval and Kaggle has none. These numbers are
+under-approximations as \eval may be called reflectively, through aliases,
+or passed to a higher-order function.
 
@@
 
-In this section we turn our attention to the expression passed for
-evaluation to \eval. The \eval function accepts all types of R objects as
-inputs, but in the majority of cases (\AllExpressionInputEvalCallPerc), it
-is called with expressions. The remaining \AllValueInputEvalCallPerc are
-values, \ie, they evaluate to themselves.
+In this section we turn our attention to the expression passed for evaluation to
+\eval. The \eval function accepts all types of R objects as inputs, but in the
+majority of cases (\AllExpressionInputEvalCallPerc), it is called with
+expressions. The remaining \AllValueInputEvalCallPerc are values, \ie, they
+evaluate to themselves.
+%
+%Furthermore, the proportion of values passed to core
+%\eval call is \missingTag{CoreEvalValueInputPerc} compared to a significantly higher
+%\missingTag{PackageEvalValueInputPerc} for package \eval calls.
 %
@@
 the default value for a function parameter can be a complex expression or a
-benign value such as \c{NULL}, the \c{match.arg} function evaluates it through a
-call to \eval. This makes the \eval inside \c{match.arg} ``polymorhpic''. The
-second such \eval call originates from \ggplot, a popular R package for making
-plots. \ggplot has it's own object system, \ggproto, which uses \eval inside the
-\ggproto function to determine the super-class from which it inherits (the
-default value is \c{NULL} but in general it could be an arbitrarily expression).
-The third \eval call is in core R \c{str} function used to print a summary of R
-objects. It internally maps \eval function on the bindings of an environment
-which results in inputs of multiples types being passed to \eval.
+benign value such as \c{NULL}, \c{match.arg} evaluates it through a call to
+\eval. This makes the \eval inside \c{match.arg} ``polymorhpic''. The second
+such \eval call originates from the popular \ggplot package of R which is a DSL
+for plotting. \ggplot has it's own object system, \ggproto, which uses \eval
+inside the \ggproto function to determine the super-class from which it
+inherits. The default value of the superclass is \c{NULL} but in general could
+be an arbitrarily complex expression. The third \eval call is in core R \c{str}
+function. This function is used to print a summary of R objects. It internally
+maps \eval function on the bindings of an environment which results in inputs of
+multiples types being passed to \eval.
 
 We further investigate the type of input passed to \eval in
-Figure~\ref{fig:eval-expression-kind} by the type of input and the source of
-\eval. In this figure, \emph{Expression} is the type of program text,
+Figure~\ref{fig:eval-expression-kind} by the type of input and
+the source of \eval. In this figure, \emph{Expression} is the type of program text,
 \emph{Environmment} is the type of scopes or environments used for binding,
-\emph{Null} is the type of the unique \c{NULL} object, \emph{Vector} is the
-type assigned to vectors of strings, booleans, integers, reals and bytes;
+\emph{Null} is the type of the unique \c{NULL} object, \emph{Vector} is the type
+assigned to vectors of strings, booleans, integers, reals and bytes;
 \emph{Symbol} is the type of unevaluated symbols, and \emph{Function} is the
@@
 
-From the figure, we observe that majority of inputs to Core \eval functions
-are expressions in contrast to package \eval calls which receive mostly
-environments as inputs. In fact, \PackageEnvironmentInputEvalCallPerc of
-\eval calls in packages receive environments as input expressions. There are
-only four callsites in package evals that receive environments as
-expressions.
+From the figure, we observe that majority of inputs to Core \eval functions are
+expressions in contrast to package \eval calls which receive mostly environments
+as inputs. In fact, \PackageEnvironmentInputEvalCallPerc of \eval calls in
+packages receive environments as input expressions. There are only four
+callsites in package evals that receive environments as expressions.
 \begin{compactitem}[$-$]
@@
 OOP system.
-\item \c{future::backtrace} function applies \eval to a future object which
-  is implemented as an environment. This is called only once.
+\item \c{future::backtrace} function applies \eval to a future object which is
+implemented as an environment. This is called only once.
 \item \c{RModel::str.RMmodel} function overloads the core \c{str} method for its
@@
 
-We looked at the top ten expressions passed to core and package \eval
-calls. The most frequent ten expressions to eval calls from core R
-contribute to 85\% of all \eval calls.
-
-\begin{table}[!h] \centering
-\begin{tabular}{@{}l|rr@{}} \hline
-Expression & Eval Call &  \% \\\hline
-\c{c("auto", "shell", "radix")} & 198,7105 & 29\%\\
-\c{c("auto", "shell", "quick", "radix")} & 1,593,169  & 23\%\\
-\c{\{info <- loadingNamespaceInfo(...} & 1,008,632 &       14\%\\
-\c{c("onLoad", "attach", "detach", "onUnload")}   & 470,566 &      6.9\%\\
-\c{c("append", "prepend", "replace")} &              261,587&       3.9\% \\
-\c{c("left", "right", "centre", "none")} & 162,086     & 2.4\%\\
-\c{c("no", "ifany", "always")}   &                71,580 &       1.1\%\\
-\c{c("pearson", "kendall", "spearman")}  & 72,962 &      1.1\%\\
+We looked at the top ten expressions passed to core and package \eval calls. The
+most frequent ten expressions to eval calls from core R contribute to 85\% of
+all \eval calls.
+\begin{table*}[tb]
+  \centering
+  \begin{tabular}{l|rr} \hline
+    Expression & Eval Call & Eval Call \% \\\hline
+    \c{c("auto", "shell", "radix")} & 1987105 & 29\%\\
+    \c{c("auto", "shell", "quick", "radix")} & 1593169  & 23\%\\
+    \c{\{info <- loadingNamespaceInfo(...} & 1008632 &       14\%\\
+    \c{c("onLoad", "attach", "detach", "onUnload")}   & 470566 &      6.9\%\\
+    \c{c("append", "prepend", "replace")} &              261587&       3.9\% \\
+    \c{c("left", "right", "centre", "none")} & 162086     & 2.4\%\\
+    \c{c("no", "ifany", "always")}   &                71580 &       1.1\%\\
+    \c{c("pearson", "kendall", "spearman")}  & 72962 &      1.1\%\\
     \c{NULL}& 75330  &      1.1\% \\
-\c{Symbol}&                 6,6279&       1\%\\\hline
+    \c{Symbol}&                 66279&       1\%\\\hline
   \end{tabular}
-\caption{Top ten eval calls in Core}\label{B}
-\end{table}
+  \caption{Top ten eval calls in Core}
+  \label{A}
+\end{table*}
 
@@
 
-The most frequent ten expressions to eval calls from CRAN packages
-contribute to 77.1\% of all \eval calls.
-
-\begin{table}[!h]  \centering
-\begin{tabular}{@{}l@{~}|@{~}r@{~}r@{}} \hline
-Expression & Eval Call &  \% \\\hline
+The most frequent ten expressions to eval calls from CRAN packages contribute to
+77.1\% of all \eval calls.
+\begin{table*}[tb]
+  \centering
+  \begin{tabular}{l|rr} \hline
+    Expression & Eval Call & Eval Call \% \\\hline
     \c{Environment} &                                                              989302   & 61\%\\
     \c{column[rows] <<- what} &                                                    55677    & 3.5\%\\
-\c{function(value) freduce(value, `_function_list`)} & 37251& 2.3\%\\
+    \c{function(value) freduce(value, `_function_list`)} &                         37251    & 2.3\%\\
     \c{NULL} &                         32005    & 2\%\\
@@
     \c{String Vector}                           &                           17487     & 1.1\%\\
-\end{tabular}\caption{Top ten eval calls in CRAN} \label{C}
-\end{table}
+  \end{tabular}
+  \caption{Top ten eval calls in CRAN}
+  \label{A}
+\end{table*}
 
@@
 function captures unevaluated unnamed arguments and maps the evaluation of
-\c{force(..1)} on them.  The \c{force} function forces promises and returns
-the result of evaluation. The \c{inner} pattern arises from
-\c{glue::identity_transformer} which enables the creation of custom
-transformation functions for affecting the interpolation and concatenation
-of input by the \c{glue} package.
+\c{force(..1)} on them.  The \c{force} function forces
+promises~\cite{oopsla19a} and returns the result of evaluation. The
+\c{inner} pattern arises from \c{glue::identity_transformer} which enables
+the creation of custom transformation functions for affecting the
+interpolation and concatenation of input by the \c{glue} package.
 
 \subsection{Operation Mix}
-
-The JavaScript paper reported on the operation mix within \evals. This means
-the proportion of read, write, and call bytecodes performed by the
-JavaScript interpreter. They found that some web pages were creating objects
-more frequently and writing into their fields.  In R, nearly every operation
-performed in the interpreter is a function call. We observe 2.9G calls and,
-of those 125M happen inside \eval (there are 463M native calls, of which
-only 31M are in \eval).
-
-{\bf there is no good reason for not counting reads and writes, we can...}
-
-%% ...The following is fluff ...
-
-%% Qualitatively, we observe all kinds of operations being executed inside
-%% \eval calls, from numerical computing for statistical modeling to reflection
-%% and metaprogramming for implementing DSLs. Even testing frameworks such
-%% \c{genthat} heavily rely upon \eval to execute test code snippets. The core
-%% R implementation is rife with uses of \eval for mundane things like
-%% accessing default values of formal parameters and in statistical packages
-%% for computation.
-
-\subsection{Provenance}
-
-In JavaScript, the provenance of the string passed into \eval is a matter of
-concern as \eval can be used to breach the security of websites. The main
-danger comes from string that were obtained from external sources. If they
-are allowed to flow into an \eval, then an attacker could hijack the
-program.
-
-{\bf what have they found...}
-
-There are various ways to obtain expressions:
+In this section, we look at the operations executed by the expressions and
+symbols on evaluation. R is a lazy language, so symbols can be bound to
+promises that can evaluate arbitrary code. All operations in R are function
+calls, even primitive operations such as addition and assignment are treated
+as calls to functions implemented in the interpreter. In our corpus, we
+observe 2.9G calls to R functions of which 125M happen inside
+\eval. Furthermore, unlike Javascript, R code also calls into peformant
+native implementations of numerical and statistical routines. In the corpus
+there are 463M calls to native code of which only 31M happens inside \eval.
+Qualitatively, we observe all kinds of operations being executed inside
+\eval calls, from numerical computing for statistical modeling to reflection
+and metaprogramming for implementing DSLs. Even testing frameworks such
+\c{genthat} heavily rely upon \eval to execute test code snippets. The core
+R implementation is rife with uses of \eval for mundane things like
+accessing default values of formal parameters and in statistical packages
+for computation.
+
+\subsubsection{Provenance}
+
+In this section we turn our attention to the \AllExpressionInputEvalCallPerc
+expressions passed as input to \eval. There are two ways to obtain
+expressions in R.
 \begin{itemize}
-  \item \c{substitute} synthesizes ASTs from expressions by replacing
-    symbols with their bindings in the specified environment.
-  \item \c{expression}  creates a vector of expression
+  \item \c{substitute} function can be used to synthesize ASTs from
+    expressions by replacing symbols with their bindings in the specified
+    environment.
+  \item \c{expression} function can be used to create a vector of expression
     objects from text.
-  \item \c{parse}, \c{str2expression} and \c{str2lang} turn strings into
+  \item \c{parse} function and its variants \c{str2expression} and
+    \c{str2lang} can be used to turn strings or text from file into R
     expressions.
@@
 
-We observe 5.2\% cases where \eval directly evaluates the output of
-\c{substitute}, 0.7\% cases where output of \c{parse} is read directly and
-only 704 cases for \c{expression}. Most expressions consumed by \eval are
-generated by other functions.
-
-\c{eval(parse(...))} can be used for dynamic code loading. This forms the
-core of \c{source} and \c{sys.source} functions in R that are commonly used
-for loading code in R files in interactive settings. We investigated the
-number of cases in which the output of \c{parse} and its variants is passed
-to \eval, directly or transitively by tainting their output. This
-corresponds to \PercentParsedCallSites of the total \eval call sites and
-\PercentParsedEvals of the eval calls. We observed that very few of the eval
-call (\NbParseFilesRnd in total) consume the result of calling \c{parse} on
-a file. Most of the eval calls consume the result of calling \c{parse} on a
-string.  We also identified one function in core R,
+In our corpus, we observe 5.2\% cases where \eval directly evaluates the
+output of \c{substitute}, 0.7\% cases where output of \c{parse} and its
+variants is read directly and only 704 cases for \c{expression}. Majority of
+expression objects consumed by \eval are generated by their transitive uses
+by other functions.
+
+The pattern \c{eval(parse(...))} can be used for dynamic code loading. This
+forms the core of \c{source} and \c{sys.source} functions in R that are
+commonly used for loading code in R files in interactive settings. We
+investigated the number of cases in which the output of \c{parse} and its
+variants is passed to \eval, directly or transitively by tainting their
+output. This corresponds to \PercentParsedCallSites of the total \eval call
+sites and \PercentParsedEvals of the eval calls. We observed that very few
+of the eval call (\NbParseFilesRnd in total) consume the result of calling
+\c{parse} on a file. Most of the eval calls consume the result of calling
+\c{parse} on a string.  We also identified one function in core R,
 \c{invokeRestartInteractively} that prompts the user for input, parses it,
@@
 dangerous. Only writes to environments not local to the computation spawned
-by \eval are side-effecting. These writes outlive the computation and hence
-are visible outside it. The remaining writes are local to the
-computation. We observe that \EvalSideEffectingWritesRnd writes inside evals
-are side-effecting. This is only \EvalSideEffectingWritesEvalPerc of all
-variable writes inside \eval and \EvalSideEffectingWritesAllPerc of all
-variable writes in the corpus.
+by \eval are side-effecting. These writes outlive the computation and hence are
+visible outside it. The remaining writes are local to the
+computation. We observe that \EvalSideEffectingWritesRnd writes inside evals are
+side-effecting. This is only \EvalSideEffectingWritesEvalPerc of all variable
+writes inside \eval and \EvalSideEffectingWritesAllPerc of all variable writes
+in the corpus.
 
@@
 One of the main differences of the R \eval with the javascript one is how it
-interacts with \emph{scope}, or environments in R. In Javascript, \eval can
-access the local scope and the global scope. In R, \eval has an argument
+interacts with \emph{scope}, or environments in R. In Javascript, \c{eval} can
+access the local scope and the global scope. In R, \c{eval} has an argument
 \c{envir} to indicate in which environment its first argument must be evaluated.
 By default, it is equal to \c{parent.frame()}, 
-which is the parent environment of \eval call. This default argument
+which is the parent environment of \c{eval} call. This default argument
 happens in \DefaultEnvirExprPercent of the eval calls.
@@
 This argument can also be \c{NULL}, a \c{List} or a data frame. This happens
-in 3.5\% of the cases: \eval copies the fields of the list or data frame
+in 3.5\% of the cases: \c{eval} copies the fields of the list or data frame
 and creates bindings for them in a new environment. This pattern is used to
@@
 
-Richards et al.~\cite{ecoop11} identified a number of frequently occurring \eval
+The JavaScript paper identified a number of frequently occurring \eval
 patterns which were detected using simple regular expressions on the \eval
-strings. In the case of R we match the patterns on the recorded AST.
+strings. In the case of R, we match the patterns on the resolved AST.
 
@@
 \end{tabular}
-\caption{JavaScript \eval Patterns in R}
+\caption{JavaScript evals patterns in R}
 \label{table:js-pattens-in-r}
@@
 mostly because the language provides dedicated features for it. The few calls in
-Library pattern come mostly from file loading functions such as \c{source} and
-\c{sys.source}.
+Library pattern come mostly from file sourcing functions (\ie \c{source} and \c{sys.source}).
 %
 \noindent \emph{Read} pattern in JavaScript represented mostly an incorrect
-
 access to an object's property or local variable both replaceable by
@@
 evaluated in the \c{airquality} data frame.
-
 %

### 2020-08-29 inst/paper/main.tex
lost the plot

@@
 
-\begin{figure*}[!tb]\centering\includegraphics[width=.8\linewidth]
-  {corpus.pdf}\caption{CRAN packages}\label{fig:corpus}
-\end{figure*}
-
 All steps of this pipeline are parallelized using GNU
@@
 assemble it.
+
+\begin{figure}[!h]\centering\includegraphics[width=\linewidth]
+  {corpus.pdf}\caption{CRAN packages}\label{fig:corpus}
+\end{figure}
 
@@
 \eval~\cite{ecoop11}. The case for R is more striking with 100\% of the
-programs call the \eval function.\footnote{We combine calls to \c{eval},
-  \c{eval.parent}, \c{evalq} and \c{local}.}  This high number is explained
-by the fact that the R implementation, its core packages, use \eval
-extensively. We differentiate uses based on their origin. A {\bf core} \eval
-is one that occurs in the code of a core package.  A {\bf dependent} \eval
-occurs in a package imported by the target package. An {\bf own} \eval is
-one that occurs in the source code of the target package.
+programs call the \eval function.\footnote{We combine calls to {\sf eval},
+  {\sf eval.parent}, {\sf evalq} and {\sf local}.}  This high number is
+explained by the fact that the R implementation, its core packages, use
+\eval extensively. We differentiate uses based on their origin. A {\bf core}
+\eval is one that occurs in the code of a core package.  A {\bf dependent}
+\eval occurs in a package imported by the target package. An {\bf own} \eval
+is one that occurs in the source code of the target package.
 
@@
 Table~\ref{table:site-package-summary} shows the number of call sites in
-CRAN packages. \OneCallSitesEnrich packages have a single call site, and one package, \emph{metafor}, has \MaxCallSitesEnrich
-call sites of \eval. \NoCallSitesEnrich packages have no calls to \eval. The core packages
-have \NbCoreEvalCallSites call sites of \eval and Kaggle has none. These numbers are
-under-approximations as \eval may be called reflectively, through aliases,
-or passed to a higher-order function.
+CRAN packages. \OneCallSitesEnrich packages have a single call site, and one
+package, \emph{metafor}, has \MaxCallSitesEnrich call sites of
+\eval. \NoCallSitesEnrich packages have no calls to \eval. The core packages
+have \NbCoreEvalCallSites call sites of \eval and Kaggle has none. These
+numbers are under-approximations as \eval may be called reflectively,
+through aliases, or passed to a higher-order function.
 
@@
 
-In this section we turn our attention to the expression passed for evaluation to
-\eval. The \eval function accepts all types of R objects as inputs, but in the
-majority of cases (\AllExpressionInputEvalCallPerc), it is called with
-expressions. The remaining \AllValueInputEvalCallPerc are values, \ie, they
-evaluate to themselves.
+In this section we turn our attention to the expression passed for
+evaluation to \eval. The \eval function accepts all types of R objects as
+inputs, but in the majority of cases (\AllExpressionInputEvalCallPerc), it
+is called with expressions. The remaining \AllValueInputEvalCallPerc are
+values, \ie, they evaluate to themselves.
 %
-%Furthermore, the proportion of values passed to core
-%\eval call is \missingTag{CoreEvalValueInputPerc} compared to a significantly higher
-%\missingTag{PackageEvalValueInputPerc} for package \eval calls.
-%
-Passing a value into an eval is effectively an no-op so we look closely at the
-expression types the call sites pass to eval across all the runs.
+Passing a value into an eval is effectively an no-op so we look closely at
+the expression types the call sites pass to eval across all the runs.
 \AllValueInputEvalSitePerc of all call sites always pass values,
-\AllExpressionInputEvalSitePerc pass expressions and a very large proportion,
-\AllPolymorphicInputEvalSitePerc, pass both expressions and values as inputs
-across all program runs. We inspected the eval callsites that accept only values
-or both values and expression as inputs. The top three of these call sites
-contribute to over ~80\% of all calls to evals. The first one is inside
-\c{match.arg} from core. Given a vector of names as strings, this function finds
-the default values for parameters with those names in the caller function. Since
-the default value for a function parameter can be a complex expression or a
-benign value such as \c{NULL}, \c{match.arg} evaluates it through a call to
-\eval. This makes the \eval inside \c{match.arg} ``polymorhpic''. The second
-such \eval call originates from the popular \ggplot package of R which is a DSL
-for plotting. \ggplot has it's own object system, \ggproto, which uses \eval
-inside the \ggproto function to determine the super-class from which it
-inherits. The default value of the superclass is \c{NULL} but in general could
-be an arbitrarily complex expression. The third \eval call is in core R \c{str}
-function. This function is used to print a summary of R objects. It internally
-maps \eval function on the bindings of an environment which results in inputs of
+\AllExpressionInputEvalSitePerc pass expressions and a very large
+proportion, \AllPolymorphicInputEvalSitePerc, pass both expressions and
+values as inputs across all program runs. We inspected the eval callsites
+that accept only values or both values and expression as inputs. The top
+three of these call sites contribute to over ~80\% of all calls to
+evals. The first one is inside \c{match.arg} from core. Given a vector of
+names as strings, this function finds the default values for parameters with
+those names in the caller function. Since the default value for a function
+parameter can be a complex expression or a benign value such as \c{NULL},
+\c{match.arg} evaluates it through a call to \eval. This makes the \eval
+inside \c{match.arg} ``polymorhpic''. The second such \eval call originates
+from the popular \ggplot package of R which is a DSL for plotting. \ggplot
+has it's own object system, \ggproto, which uses \eval inside the \ggproto
+function to determine the super-class from which it inherits. The default
+value of the superclass is \c{NULL} but in general could be an arbitrarily
+complex expression. The third \eval call is in core R \c{str} function. This
+function is used to print a summary of R objects. It internally maps \eval
+function on the bindings of an environment which results in inputs of
 multiples types being passed to \eval.
@@
 We further investigate the type of input passed to \eval in
-Figure~\ref{fig:eval-expression-kind} by the type of input and
-the source of \eval. In this figure, \emph{Expression} is the type of program text,
+Figure~\ref{fig:eval-expression-kind} by the type of input and the source of
+\eval. In this figure, \emph{Expression} is the type of program text,
 \emph{Environmment} is the type of scopes or environments used for binding,
-\emph{Null} is the type of the unique \c{NULL} object, \emph{Vector} is the type
-assigned to vectors of strings, booleans, integers, reals and bytes;
+\emph{Null} is the type of the unique \c{NULL} object, \emph{Vector} is the
+type assigned to vectors of strings, booleans, integers, reals and bytes;
 \emph{Symbol} is the type of unevaluated symbols, and \emph{Function} is the
@@
 
-We looked at the top ten expressions passed to core and package \eval calls. The
-most frequent ten expressions to eval calls from core R contribute to 85\% of
-all \eval calls.
-\begin{table*}[tb]
-  \centering
-  \begin{tabular}{l|rr} \hline
-    Expression & Eval Call & Eval Call \% \\\hline
-    \c{c("auto", "shell", "radix")} & 1987105 & 29\%\\
-    \c{c("auto", "shell", "quick", "radix")} & 1593169  & 23\%\\
-    \c{\{info <- loadingNamespaceInfo(...} & 1008632 &       14\%\\
-    \c{c("onLoad", "attach", "detach", "onUnload")}   & 470566 &      6.9\%\\
-    \c{c("append", "prepend", "replace")} &              261587&       3.9\% \\
-    \c{c("left", "right", "centre", "none")} & 162086     & 2.4\%\\
-    \c{c("no", "ifany", "always")}   &                71580 &       1.1\%\\
-    \c{c("pearson", "kendall", "spearman")}  & 72962 &      1.1\%\\
+We looked at the top ten expressions passed to core and package \eval
+calls. The most frequent ten expressions to eval calls from core R
+contribute to 85\% of all \eval calls.
+
+\begin{table}[!h] \centering
+\begin{tabular}{@{}l|rr@{}} \hline
+Expression & Eval Call &  \% \\\hline
+\c{c("auto", "shell", "radix")} & 198,7105 & 29\%\\
+\c{c("auto", "shell", "quick", "radix")} & 1,593,169  & 23\%\\
+\c{\{info <- loadingNamespaceInfo(...} & 1,008,632 &       14\%\\
+\c{c("onLoad", "attach", "detach", "onUnload")}   & 470,566 &      6.9\%\\
+\c{c("append", "prepend", "replace")} &              261,587&       3.9\% \\
+\c{c("left", "right", "centre", "none")} & 162,086     & 2.4\%\\
+\c{c("no", "ifany", "always")}   &                71,580 &       1.1\%\\
+\c{c("pearson", "kendall", "spearman")}  & 72,962 &      1.1\%\\
 \c{NULL}& 75330  &      1.1\% \\
-    \c{Symbol}&                 66279&       1\%\\\hline
+\c{Symbol}&                 6,6279&       1\%\\\hline
 \end{tabular}
-  \caption{Top ten eval calls in Core}
-  \label{A}
-\end{table*}
+\caption{Top ten eval calls in Core}\label{B}
+\end{table}
 
@@
 
-The most frequent ten expressions to eval calls from CRAN packages contribute to
-77.1\% of all \eval calls.
-\begin{table*}[tb]
-  \centering
-  \begin{tabular}{l|rr} \hline
-    Expression & Eval Call & Eval Call \% \\\hline
+The most frequent ten expressions to eval calls from CRAN packages
+contribute to 77.1\% of all \eval calls.
+
+\begin{table}[!h]  \centering
+\begin{tabular}{@{}l@{~}|@{~}r@{~}r@{}} \hline
+Expression & Eval Call &  \% \\\hline
 \c{Environment} &                                  989302   & 61\%\\
 \c{column[rows] <<- what} &                        55677    & 3.5\%\\
-    \c{function(value) freduce(value, `_function_list`)} &                         37251    & 2.3\%\\
+\c{function(value) freduce(value, `_function_list`)} & 37251& 2.3\%\\
 \c{NULL} &                         32005    & 2\%\\
@@
 \c{String Vector}      &                           17487     & 1.1\%\\
-  \end{tabular}
-  \caption{Top ten eval calls in CRAN}
-  \label{A}
-\end{table*}
+\end{tabular}\caption{Top ten eval calls in CRAN} \label{C}
+\end{table}
 
@@
 function captures unevaluated unnamed arguments and maps the evaluation of
-\c{force(..1)} on them.  The \c{force} function forces
-promises~\cite{oopsla19a} and returns the result of evaluation. The
-\c{inner} pattern arises from \c{glue::identity_transformer} which enables
-the creation of custom transformation functions for affecting the
-interpolation and concatenation of input by the \c{glue} package.
+\c{force(..1)} on them.  The \c{force} function forces promises and returns
+the result of evaluation. The \c{inner} pattern arises from
+\c{glue::identity_transformer} which enables the creation of custom
+transformation functions for affecting the interpolation and concatenation
+of input by the \c{glue} package.
 

### 2020-08-29 inst/paper/main.tex
stuff

@@
 \newcommand{\eval}{\c{eval}}
+\newcommand{\evals}{{\sf eval}s\xspace}
 \newcommand{\parse}{\c{parse}}
@@
 \subsection{Operation Mix}
-In this section, we look at the operations executed by the expressions and
-symbols on evaluation. R is a lazy language, so symbols can be bound to
-promises that can evaluate arbitrary code. All operations in R are function
-calls, even primitive operations such as addition and assignment are treated
-as calls to functions implemented in the interpreter. In our corpus, we
-observe 2.9G calls to R functions of which 125M happen inside
-\eval. Furthermore, unlike Javascript, R code also calls into peformant
-native implementations of numerical and statistical routines. In the corpus
-there are 463M calls to native code of which only 31M happens inside \eval.
-Qualitatively, we observe all kinds of operations being executed inside
-\eval calls, from numerical computing for statistical modeling to reflection
-and metaprogramming for implementing DSLs. Even testing frameworks such
-\c{genthat} heavily rely upon \eval to execute test code snippets. The core
-R implementation is rife with uses of \eval for mundane things like
-accessing default values of formal parameters and in statistical packages
-for computation.
-
-\subsubsection{Provenance}
-
-In this section we turn our attention to the \AllExpressionInputEvalCallPerc
-expressions passed as input to \eval. There are two ways to obtain
-expressions in R.
+
+The JavaScript paper reported on the operation mix within \evals. This means
+the proportion of read, write, and call bytecodes performed by the
+JavaScript interpreter. They found that some web pages were creating objects
+more frequently and writing into their fields.  In R, nearly every operation
+performed in the interpreter is a function call. We observe 2.9G calls and,
+of those 125M happen inside \eval (there are 463M native calls, of which
+only 31M are in \eval).
+
+{\bf there is no good reason for not counting reads and writes, we can...}
+
+%% ...The following is fluff ...
+
+%% Qualitatively, we observe all kinds of operations being executed inside
+%% \eval calls, from numerical computing for statistical modeling to reflection
+%% and metaprogramming for implementing DSLs. Even testing frameworks such
+%% \c{genthat} heavily rely upon \eval to execute test code snippets. The core
+%% R implementation is rife with uses of \eval for mundane things like
+%% accessing default values of formal parameters and in statistical packages
+%% for computation.
+
+\subsection{Provenance}
+
+In JavaScript, the provenance of the string passed into \eval is a matter of
+concern as \eval can be used to breach the security of websites. The main
+danger comes from string that were obtained from external sources. If they
+are allowed to flow into an \eval, then an attacker could hijack the
+program.
+
+{\bf what have they found...}
+
+There are various ways to obtain expressions:
 \begin{itemize}
-  \item \c{substitute} function can be used to synthesize ASTs from
-    expressions by replacing symbols with their bindings in the specified
-    environment.
-  \item \c{expression} function can be used to create a vector of expression
+  \item \c{substitute} synthesizes ASTs from expressions by replacing
+    symbols with their bindings in the specified environment.
+  \item \c{expression}  creates a vector of expression
     objects from text.
-  \item \c{parse} function and its variants \c{str2expression} and
-    \c{str2lang} can be used to turn strings or text from file into R
+  \item \c{parse}, \c{str2expression} and \c{str2lang} turn strings into
     expressions.
@@
 
-In our corpus, we observe 5.2\% cases where \eval directly evaluates the
-output of \c{substitute}, 0.7\% cases where output of \c{parse} and its
-variants is read directly and only 704 cases for \c{expression}. Majority of
-expression objects consumed by \eval are generated by their transitive uses
-by other functions.
-
-The pattern \c{eval(parse(...))} can be used for dynamic code loading. This
-forms the core of \c{source} and \c{sys.source} functions in R that are
-commonly used for loading code in R files in interactive settings. We
-investigated the number of cases in which the output of \c{parse} and its
-variants is passed to \eval, directly or transitively by tainting their
-output. This corresponds to \PercentParsedCallSites of the total \eval call
-sites and \PercentParsedEvals of the eval calls. We observed that very few
-of the eval call (\NbParseFilesRnd in total) consume the result of calling
-\c{parse} on a file. Most of the eval calls consume the result of calling
-\c{parse} on a string.  We also identified one function in core R,
+We observe 5.2\% cases where \eval directly evaluates the output of
+\c{substitute}, 0.7\% cases where output of \c{parse} is read directly and
+only 704 cases for \c{expression}. Most expressions consumed by \eval are
+generated by other functions.
+
+\c{eval(parse(...))} can be used for dynamic code loading. This forms the
+core of \c{source} and \c{sys.source} functions in R that are commonly used
+for loading code in R files in interactive settings. We investigated the
+number of cases in which the output of \c{parse} and its variants is passed
+to \eval, directly or transitively by tainting their output. This
+corresponds to \PercentParsedCallSites of the total \eval call sites and
+\PercentParsedEvals of the eval calls. We observed that very few of the eval
+call (\NbParseFilesRnd in total) consume the result of calling \c{parse} on
+a file. Most of the eval calls consume the result of calling \c{parse} on a
+string.  We also identified one function in core R,
 \c{invokeRestartInteractively} that prompts the user for input, parses it,

### 2020-08-29 inst/paper/main.tex
sruff
Fix commas

@@
 \newcommand{\UndefinedOutOfCoprusPercent}{0\%\xspace}
-\newcommand{\MaxCallSitesEnrich}{114}
-\newcommand{\OneCallSitesEnrich}{61}
-\newcommand{\NoCallSitesEnrich}{297}
+\newcommand{\MaxCallSitesEnrich}{114\xspace}
+\newcommand{\OneCallSitesEnrich}{61\xspace}
+\newcommand{\NoCallSitesEnrich}{297\xspace}
 %\input{tag/patterns.tex}
@@
 
-\newcommand{\eval}{\c{eval}}
+\newcommand{\eval}{\texttt{eval}}
 \newcommand{\evals}{{\sf eval}s\xspace}
@@
 
-This section presents an overview of the usage of \eval in our corpus.
-Richards et al. reported 82\% of JavaScript-enabled web pages using
-\eval~\cite{ecoop11}. The case for R is more striking with 100\% of the
-programs call the \eval function.\footnote{We combine calls to {\sf eval},
-  {\sf eval.parent}, {\sf evalq} and {\sf local}.}  This high number is
-explained by the fact that the R implementation, its core packages, use
-\eval extensively. We differentiate uses based on their origin. A {\bf core}
-\eval is one that occurs in the code of a core package.  A {\bf dependent}
-\eval occurs in a package imported by the target package. An {\bf own} \eval
-is one that occurs in the source code of the target package.
+This section presents an overview of the usage of \eval in our corpus. Richards
+et al. reported 82\% of JavaScript-enabled web pages using \eval~\cite{ecoop11}.
+The case for R is more striking with 100\% of the programs call the \eval
+function.\footnote{We combine calls to {\sf eval}, {\sf eval.parent}, {\sf
+    evalq} and {\sf local}.} This high number is explained by the fact that the
+R implementation, its core packages, use \eval extensively. We differentiate
+three types of eval based on their origin:
+%
+\begin{compactitem}[$-$]
+  \item a {\bf core eval} is one that occurs in the code of a core package,
+  \item a {\bf dependent eval} occurs in a package imported by the target
+  package, and
+  \item an {\bf own eval} is one that occurs in the source code of the target
+  package.
+\end{compactitem}
 
@@
 Table~\ref{table:site-package-summary} shows the number of call sites in
-CRAN packages. \OneCallSitesEnrich packages have a single call site, and one
-package, \emph{metafor}, has \MaxCallSitesEnrich call sites of
-\eval. \NoCallSitesEnrich packages have no calls to \eval. The core packages
-have \NbCoreEvalCallSites call sites of \eval and Kaggle has none. These
+CRAN packages. \OneCallSitesEnrich{} packages have a single call site, and one
+package, \emph{metafor}, has \MaxCallSitesEnrich{} call sites of
+\eval. \NoCallSitesEnrich{} packages have no calls to \eval. The core packages
+have \NbCoreEvalCallSites{} call sites of \eval and Kaggle has none. These
 numbers are under-approximations as \eval may be called reflectively,
@@
 %
-Passing a value into an eval is effectively an no-op so we look closely at
-the expression types the call sites pass to eval across all the runs.
+Passing a value into an eval is effectively an no-op so we look closely at the
+expression types the call sites pass to eval across all the runs.
 \AllValueInputEvalSitePerc of all call sites always pass values,
-\AllExpressionInputEvalSitePerc pass expressions and a very large
-proportion, \AllPolymorphicInputEvalSitePerc, pass both expressions and
-values as inputs across all program runs. We inspected the eval callsites
-that accept only values or both values and expression as inputs. The top
-three of these call sites contribute to over ~80\% of all calls to
-evals. The first one is inside \c{match.arg} from core. Given a vector of
-names as strings, this function finds the default values for parameters with
-those names in the caller function. Since the default value for a function
-parameter can be a complex expression or a benign value such as \c{NULL},
-\c{match.arg} evaluates it through a call to \eval. This makes the \eval
-inside \c{match.arg} ``polymorhpic''. The second such \eval call originates
-from the popular \ggplot package of R which is a DSL for plotting. \ggplot
-has it's own object system, \ggproto, which uses \eval inside the \ggproto
-function to determine the super-class from which it inherits. The default
-value of the superclass is \c{NULL} but in general could be an arbitrarily
-complex expression. The third \eval call is in core R \c{str} function. This
-function is used to print a summary of R objects. It internally maps \eval
-function on the bindings of an environment which results in inputs of
-multiples types being passed to \eval.
+\AllExpressionInputEvalSitePerc pass expressions and a very large proportion,
+\AllPolymorphicInputEvalSitePerc, pass both expressions and values as inputs
+across all program runs. We inspected the eval callsites that accept only values
+or both values and expression as inputs. The top three of these call sites
+contribute to over ~80\% of all calls to evals. The first one is inside
+\c{match.arg} from core. Given a vector of names as strings, this function finds
+the default values for parameters with those names in the caller function. Since
+the default value for a function parameter can be a complex expression or a
+benign value such as \c{NULL}, the \c{match.arg} function evaluates it through a
+call to \eval. This makes the \eval inside \c{match.arg} ``polymorhpic''. The
+second such \eval call originates from \ggplot, a popular R package for making
+plots. \ggplot has it's own object system, \ggproto, which uses \eval inside the
+\ggproto function to determine the super-class from which it inherits (the
+default value is \c{NULL} but in general it could be an arbitrarily expression).
+The third \eval call is in core R \c{str} function used to print a summary of R
+objects. It internally maps \eval function on the bindings of an environment
+which results in inputs of multiples types being passed to \eval.
 
@@
 Expression & Eval Call &  \% \\\hline
-\c{c("auto", "shell", "radix")} & 198,7105 & 29\%\\
+\c{c("auto", "shell", "radix")} & 1,987,105 & 29\%\\
 \c{c("auto", "shell", "quick", "radix")} & 1,593,169  & 23\%\\
@@
 \c{c("pearson", "kendall", "spearman")}  & 72,962 &      1.1\%\\
-\c{NULL}& 75330  &      1.1\% \\
-\c{Symbol}&                 6,6279&       1\%\\\hline
+\c{NULL}& 75,330  &      1.1\% \\
+\c{Symbol}&                 66,279&       1\%\\\hline
 \end{tabular}
@@
 One of the main differences of the R \eval with the javascript one is how it
-interacts with \emph{scope}, or environments in R. In Javascript, \c{eval} can
-access the local scope and the global scope. In R, \c{eval} has an argument
+interacts with \emph{scope}, or environments in R. In Javascript, \eval can
+access the local scope and the global scope. In R, \eval has an argument
 \c{envir} to indicate in which environment its first argument must be evaluated.
 By default, it is equal to \c{parent.frame()}, 
-which is the parent environment of \c{eval} call. This default argument
+which is the parent environment of \eval call. This default argument
 happens in \DefaultEnvirExprPercent of the eval calls.
@@
 This argument can also be \c{NULL}, a \c{List} or a data frame. This happens
-in 3.5\% of the cases: \c{eval} copies the fields of the list or data frame
+in 3.5\% of the cases: \eval copies the fields of the list or data frame
 and creates bindings for them in a new environment. This pattern is used to
@@
 
-The JavaScript paper identified a number of frequently occurring \eval
+Richards et al.~\cite{ecoop11} identified a number of frequently occurring \eval
 patterns which were detected using simple regular expressions on the \eval
-strings. In the case of R, we match the patterns on the resolved AST.
+strings. In the case of R we match the patterns on the recorded AST.
 
@@
 mostly because the language provides dedicated features for it. The few calls in
-Library pattern come mostly from file sourcing functions (\ie \c{source} and \c{sys.source}).
+Library pattern come mostly from file loading functions such as \c{source} and
+\c{sys.source}.
 %
 \noindent \emph{Read} pattern in JavaScript represented mostly an incorrect
+
 access to an object's property or local variable both replaceable by
@@
 evaluated in the \c{airquality} data frame.
+
 %

### 2020-09-14 inst/analysis/analysis_stable.Rmd
Updates

@@
 # Now precomputed during the tracing
-ast_sizes <- read_fst(path(RUN_DIR, "package-evals-traced.3/summarized-ast-size.fst")) %>% as_tibble() %>% 
-      filter(eval_source_type != "<undefined>") %>% 
-      rename(dataset = eval_source_type)
+#ast_sizes <- read_fst(path(RUN_DIR, "package-evals-traced.3/summarized-ast-size.fst")) %>% as_tibble() %>% 
+#      filter(eval_source_type != "<undefined>") %>% 
+#      rename(dataset = eval_source_type)
+
+ast_sizes <- eval_calls 
 ```
@@
 
-```{r env_assign}
-eval_calls_env_assign <- eval_calls %>% filter(!is.na(expr_resolved)) %>% mutate(env_assign = pblapply(expr_resolved, is_assign_str, cl = parallel::detectCores() - 2))
+```{r display_frequent_env_core}
+classified_envir_core <- classified_envir %>% filter(dataset == "core") %>% add_count(env_class, wt=nb_ev_calls, sort = TRUE) %>% group_by(env_class)  %>% slice_max(order_by = nb_ev_calls, n = 10) %>% ungroup()
 
-#eval_calls_env_assign2 <-  eval_calls_env_assign %>% filter(is.logical(env_assign)) %>% hoist("env_assign", env_assign = 1)
+classified_envir_core %>% select(env_class, eval_call_expression, eval_call_srcref, caller_expression, n) %>% distinct() %>% datatable()
+```
+
+```{r display_frequent_env_packages}
+classified_envir_packages <- classified_envir %>% filter(dataset == "packages") %>% add_count(env_class, wt=nb_ev_calls, sort = TRUE) %>% group_by(env_class)  %>% slice_max(order_by = nb_ev_calls, n = 10) %>% ungroup()
+
+classified_envir_packages %>% select(env_class, eval_call_expression, eval_call_srcref, caller_expression, n) %>% distinct() %>% datatable()
 ```

### 2020-09-14 inst/analysis/inc/paths.R

@@ -29,7 +29,7 @@ PACKAGE_TRACE_RUNS_FILE    <- path(DATA_DIR, "run-trace.fst")
 PACKAGE_TRACE_LOG_FILE     <- path(DATA_DIR, "parallel-trace.fst")
 CRAN_PROGRAMS_FILE         <- path(DATA_DIR, "cran-programs.fst")
 
-RUN_TRACE_DIR  <- path(RUN_DIR, "package-evals-traced.2")
+RUN_TRACE_DIR  <- path(RUN_DIR, "package-evals-traced.3")
 RUN_KAGGLE_DIR <- path(RUN_DIR, "kaggle-run")
 
 KAGGLE_DEPENDENCIES_FILE <- path(DATA_DIR, "kaggle-dependencies.txt")

### 2020-09-14 inst/analysis/preprocess.R
Macros and figures

@@ -12,7 +12,8 @@
 source("insights.R")
 
 library(fs)
-#library(multidplyr)
+library(future.apply)
+plan(multiprocess) # multicore (fork) with fallback to multisession (create processes and then copy)
 
 deduplicate <- function(dataset)
 {
@@ -30,6 +31,8 @@ add_types <- function(dataset)
 
 add_parse_args <- function(dataset)
 {
+  
+  print("Cluster created. Copying functiond and libraries.")
   return(dataset %>% mutate(parse_args = map(expr_parsed_expression, function(e) 
     { if(!is.na(e) && str_starts(e, "(parse|str2lang|str2expression)\\(")) 
         extract_args_parse(e) 
@@ -39,7 +42,7 @@ add_parse_args <- function(dataset)
 eval_base_functions <- c("autoload", "autoloader", "bquote", "by.default", "by.data.frame", "invokeRestartInteractively", "Ops.data.frame", "dget", 
                          "eval", "eval.parent", "evalq", "local", "with.default", "within.data.frame", "within.list", "replicate", "subset.data.frame",
                          "subset.matrix", "transform.data.frame", "match.arg", "char.expand", "max.col", "parseNamespaceFile", "source", "sys.source",
-                         "stopifnot", "as.data.frane.table", "match.fun", "trace", "untrace", ".doTrace", "Vectorize")
+                         "stopifnot", "as.data.frame.table", "match.fun", "trace", "untrace", ".doTrace", "Vectorize")
 
 
 find_package_name <- function(caller_function, caller_package, caller_expression, srcref, file)

### 2020-09-14 inst/paper/main.tex
Eval with space
Minors

--- 2020-08-29 inst/paper/main.tex
@@
-\documentclass[conference]{IEEEtran}
-\usepackage{cite}
-\usepackage{amsmath,amssymb,amsfonts}
-\usepackage{algorithmic}
-\usepackage{graphicx}
-\usepackage{textcomp}
+\documentclass[acmsmall=true, screen=true, review=true]{acmart}
+
 \usepackage{listings}
@@
 \usepackage{multirow}
+\usepackage{paralist}
+\usepackage{xspace}
+\usepackage{url}
+
 \newcommand{\missingTag}[1]{\textcolor{red}{#1}\xspace}
@@
 \newcommand{\missingGraph}{\textcolor{red}{XXGraph}\xspace}
-\usepackage[usenames,dvipsnames]{xcolor}
-\usepackage{paralist}
-\usepackage{xspace}
-\usepackage{url}
 
@@
 
-\lstset{ 
- language=R,              % the language of the code
- basicstyle=\small\sffamily,    % the size of the fonts
-% numbers=left,            % where to put the line-numbers
- numberstyle=\color{Blue},% the style that is used for the line-numbers
- stepnumber=1,            % the step between two line-numbers. If it is 1,
-                          % each line will be numbered
- numbersep=2pt,           % how far the line-numbers are from the code
- backgroundcolor=\color{white},  % choose the background color.
- showspaces=false,        % show spaces adding particular underscores
- showstringspaces=false,  % underline spaces within strings
- showtabs=false,          % show tabs within strings adding particular 
-%frame=single,            % adds a frame around the code
- rulecolor=\color{black}, % if not set, the frame-color may be changed on
-                          % line-breaks within not-black text (e.g. commens
-                          % (green here))
- tabsize=2,               % sets default tabsize to 2 spaces
- captionpos=b,            % sets the caption-position to bottom
- breaklines=true,         % sets automatic line breaking
- breakatwhitespace=false, % sets if automatic breaks should only happen at 
- keywordstyle=\color{Blue},    % keyword style
- commentstyle=\color{YellowGreen},  % comment style
- stringstyle=\color{ForestGreen}    % string literal style
+\lstset{language=R}
+
+\definecolor{LightGray}{rgb}{.92,.92,.92}
+\definecolor{Gray}{rgb}{.3,.3,.3}
+\definecolor{DarkGray}{rgb}{.5,.5,.5}
+
+\lstset{ %
+  columns=flexible,
+  captionpos=b,
+  frame=single,
+  framerule=0pt,
+  framexleftmargin=-1mm,
+  framexrightmargin=-1mm,
+  tabsize=2,
+  belowskip=0pt,
+  basicstyle=\small\ttfamily,
+  backgroundcolor=\color{LightGray},
+  emphstyle=\sffamily,
+  keywordstyle=\bfseries,
+  commentstyle=\color{Gray}\em,
+  stringstyle=\color{Gray}
 }
-
-% https://tex.stackexchange.com/questions/351435/table-name-placment-in-tabular-documentclass-ieeetran-conference
-\makeatletter
-\def\@IEEEtablestring{figure}
-\makeatother
-
-\begin{document}
-
-\title{A Large-Scale Study of the Use of Eval in R}
-\author{\vspace{-.8cm}\IEEEauthorblockN{~}} %\IEEEauthorblockA{ dept} \\ \and
-\maketitle
+\lstdefinestyle{R}{ %
+  language=R,
+  deletekeywords={new, env, equal, c, runif, trace, args, exp, t, all, get,
+    names, is, environment, class, substitute, expression, list, null, Internal,
+    sample, diag, length, rep, nrow, stop, offset, pmax, Machine,
+    double, parent, frame, par, methods, end, dir, apply, deparse, missing,
+    plot, as, integer, character, inherits, numeric, paste, eval, quote, call,
+    formula, df, log, sum, c, local, legend, file, scale, round, title, order,
+    drop, which, grid, print, ncol, dim, max, format, sort, rug, matrix, start,
+    unique, mean, df, attr, do, power},
+  otherkeywords={},
+  breaklines=true
+}
 
@@
 
-\newcommand{\eval}{\texttt{eval}}
+\newcommand{\eval}{\texttt{eval}\xspace}
 \newcommand{\evals}{{\sf eval}s\xspace}
@@
 
+\bibliographystyle{ACM-Reference-Format}
+\citestyle{acmauthoryear}   %% For author/year citations
+
+\begin{document}
+
+\title{A Large-Scale Study of the Use of Eval in R}
+\author{Anony Mouse}
+\authorsaddresses{Author' email address: anony@mouse.mouse}
+
 \begin{abstract}
@@
 \end{abstract}
+
+\begin{CCSXML}
+  <ccs2012>
+  <concept>
+  <concept_id>10011007.10011006.10011050.10010517</concept_id>
+  <concept_desc>Software and its engineering~Scripting languages</concept_desc>
+  <concept_significance>500</concept_significance>
+  </concept>
+  <concept>
+  <concept_id>10002944.10011123.10010912</concept_id>
+  <concept_desc>General and reference~Empirical studies</concept_desc>
+  <concept_significance>500</concept_significance>
+  </concept>
+  <concept>
+  <concept_id>10011007.10011006.10011008</concept_id>
+  <concept_desc>Software and its engineering~General programming languages</concept_desc>
+  <concept_significance>500</concept_significance>
+  </concept>
+  </ccs2012>
+\end{CCSXML}
+
+\ccsdesc[500]{Software and its engineering~Scripting languages}
+\ccsdesc[500]{General and reference~Empirical studies}
+\ccsdesc[500]{Software and its engineering~General programming languages}
+
+\maketitle
 
@@
 \end{lstlisting}
+% $
 
@@
 calls. The largest number of events generated by an \eval call is
-\EventsMaxCount.  However, a majority of \eval calls performs up to
+\EventsMaxCountRnd.  However, a majority of \eval calls performs up to
 \EventsMaxRangeA events suggesting that most expressions passed to \eval are
@@
 
-In this section we turn our attention to the expression passed for
-evaluation to \eval. The \eval function accepts all types of R objects as
-inputs, but in the majority of cases (\AllExpressionInputEvalCallPerc), it
-is called with expressions. The remaining \AllValueInputEvalCallPerc are
-values, \ie, they evaluate to themselves.
+% Following the high-level overview of the usage of eval in the previous section,
+% we now look into categorization of the eval behavior. Similarly to the JavaScript paper, we look at:
+% %
+% \begin{inparaenum}[(A)]
+%   \item \textbf{mix} of operations performed by the expressions passed to eval,
+%   \item \textbf{}
+% \end{inparaenum}
+
+In this section we turn our attention to the expression passed for evaluation to
+\eval. The \eval function accepts all types of R objects as inputs, but in the
+majority of cases (\AllExpressionInputEvalCallPerc), it is called with
+expressions. The remaining \AllValueInputEvalCallPerc are values, \ie, they
+evaluate to themselves.
 %
@@
 
-\bibliographystyle{IEEEtran}
 \bibliography{bib/bibliography,bib/jv}

### 2020-11-02 inst/paper/main.tex
Fix R syntax highlighting

@@
     drop, which, grid, print, ncol, dim, max, format, sort, rug, matrix, start,
-    unique, mean, df, attr, do, power},
+    unique, mean, df, attr, do, power, \$},
   otherkeywords={},

### 2020-11-03 inst/analysis/environments.Rmd
Classification of environments

---
title: "Empirical Evaluation of eval in R"
author: "Pierre Donat-Bouillud"
output:
  rmdformats::readthedown:
        code_folding: hide
        lightbox: true
        gallery: false
        df_print: paged
        toc_depth: 3
date: "`r Sys.Date()`"
editor_options: 
  chunk_output_type: console
params:
  base_dir: /var/lib/R/project-evalR
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.align = "center",
                      fig.retina = 2,
                      fig.width = 10,
                      cache.lazy = FALSE)

now <- Sys.time()

library(tidyverse)
library(fst)

source("insights.R", local = knitr::knit_global())
source("preprocess.R", local = knitr::knit_global())
```

# Loading datafiles

```{r loading_fst}
eval_calls <- read_fst("/var/lib/R/project-evalR/run/package-evals-traced-test/calls.fst") %>%
  as_tibble() %>%
  add_types()
```

# Description of the data files

```{r packages}

```

# Classification of environments

We distinguish between 3 main categories of `eval` calls with respect to their `envir` argument:

- `envir` is a list or a data frame
- `envir` comes from an argument of the caller function
- `envir` is absolutely computed (with `parent.frame()` or equivalent)

##  `envir` as a list or data frame

```{r envir_list}
eval_calls_env_list <- eval_calls %>% filter(envir_type == "VECSXP")
```

## `envir` as an argument from the caller function

```{r envir_arg_caller}
eval_calls_arg_caller <- eval_calls %>% filter(!is.na(envir_from_arg))
```

## `envir` absolutely computed

## `envir` absolutely computed

```{r}
duration <- difftime(Sys.time(), now)
```

Notebook execution was `r duration` `r units(duration)`.

### 2020-11-03 inst/analysis/preprocess.R

@@ -12,8 +12,8 @@
 source("insights.R")
 
 library(fs)
-library(future.apply)
-plan(multiprocess) # multicore (fork) with fallback to multisession (create processes and then copy)
+#library(future.apply)
+#plan(multiprocess) # multicore (fork) with fallback to multisession (create processes and then copy)
 
 deduplicate <- function(dataset)
 {
@@ -31,8 +31,6 @@ add_types <- function(dataset)
 
 add_parse_args <- function(dataset)
 {
-  
-  print("Cluster created. Copying functiond and libraries.")
   return(dataset %>% mutate(parse_args = map(expr_parsed_expression, function(e) 
     { if(!is.na(e) && str_starts(e, "(parse|str2lang|str2expression)\\(")) 
         extract_args_parse(e) 

### 2020-11-05 inst/analysis/environments.Rmd
Detecting Rcpp in the case of envir as argument from the caller

@@
   as_tibble() %>%
-  add_types()
+  deduplicate() %>% 
+  add_types() %>%
+  add_eval_source(eval_calls_raw) %>% add_eval_source_type()
 ```
@@
 
+There are `r count(eval_calls, wt=nb_ev_calls)` eval calls and `r eval_calls %>% select(eval_call_srcref) %>% n_distinct()` eval call sites in the dataset.
+
 # Classification of environments
 
-We distinguish between 3 main categories of `eval` calls with respect to their `envir` argument:
+We distinguish between 4 main categories of `eval` calls with respect to their `envir` argument:
 
+- `envir` is not provided explicitly (and thus is the default argument, which semantically belongs to the absolute category)
 - `envir` is a list or a data frame
@@
 - `envir` is absolutely computed (with `parent.frame()` or equivalent)
+
+## `envir` as default argument
+
+```{r envir_default}
+eval_calls_default <- eval_calls %>% filter(is.na(envir_expression))
+
+eval_calls <- eval_calls %>% filter(!is.na(envir_expression))
+```
 
@@
 
+Most of the eval calls directly use the argument, but some seem to modify it a lot:
+
+```{r}
+eval_calls_arg_caller %>% count(envir_from_arg, wt=nb_ev_calls, sort = TRUE) %>% knitr::kable()
+```
+
+But all those strange calls are due to Rcpp:
+
+```{r arg_call_rcpp}
+eval_calls_arg_caller %>% filter(envir_from_arg > 1, !str_detect(expr_expression, "C\\+\\+Constructor|C\\+\\+Field|C\\+\\+OverloadedMethods")) 
+```
+
+In the current dataset, there are no examples of sandbox usage but we should see adistance of 1 in that case.
+
 ## `envir` absolutely computed
 
-## `envir` absolutely computed
+```{r envir_absolute}
+eval_calls_absolute <- eval_calls %>% anti_join(eval_calls_env_list) %>% anti_join(eval_calls_arg_caller)
+```
+
+Let's do a sanity check: is the environment class the same in that category?
+
+```{r}
+eval_calls_absolute <- eval_calls_absolute  %>% add_fake_srcref() %>% group_by(eval_call_srcref) %>%  mutate(envir_distinct = n_distinct(environment_class)) %>% ungroup()
+```
 

### 2020-11-05 inst/paper/main.tex

@@
-\documentclass[acmsmall=true, screen=true, review=true]{acmart}
+\documentclass[a4paper,USenglish,cleveref, autoref, thm-restate]{lipics-v2019}
+% This is a template for producing LIPIcs articles. 
+% See lipics-manual.pdf for further information.
+% for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
+% for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
+% for section-numbered lemmas etc., use "numberwithinsect"
+% for enabling cleveref support, use "cleveref"
+% for enabling autoref support, use "autoref"
+% for anonymousing the authors (e.g. for double-blind review), add "anonymous"
+% for enabling thm-restate support, use "thm-restate"
+
+% \graphicspath{{./graphics/}}%helpful if your graphic files are in another directory
 
@@
 \newcommand{\evals}{{\sf eval}s\xspace}
-\newcommand{\parse}{\c{parse}}
+\renewcommand{\parse}{\c{parse}}
 \newcommand{\source}{\c{source}}
@@
 
-\bibliographystyle{ACM-Reference-Format}
-\citestyle{acmauthoryear}   %% For author/year citations
-
-\begin{document}
+\bibliographystyle{plainurl}
 
 \title{A Large-Scale Study of the Use of Eval in R}
-\author{Anony Mouse}
-\authorsaddresses{Author' email address: anony@mouse.mouse}
-
-\begin{abstract}
-  The \eval function turns text into code. From the point of view of
-  reasoning about software system, uses of \eval turn the program's code
-  from a statically known quantity that can be reasoned about and analyzed
-  into one of the inputs to the program. Thus, widespread use of \eval
-  hinders our ability to provide safety and security guarantees for
-  software. Understanding how \eval is used in practice is key to finding
-  ways to mitigate its damage. One question that has not been studied is
-  whether the usage of \eval is dependent on the application domain or other
-  features of the language. This paper is a large-scale study of a corpus of
-  \CorpusAllCodeRnd lines of data science libraries and end-user code written in
-  the R language. We performed a sophisticated dynamic analysis to confirm
-  that \eval is indeed in wide spread use, and categorized patterns of
-  usage. We found that \eval in this code base is more dangerous in some
-  ways, and safer in others, than what was previously reported for
-  JavaScript.
-\end{abstract}
+
+%\author{Aviral Goel}{Northeastern University, [optional: Address], Country \and My second affiliation, Country \and \url{http://www.myhomepage.edu} }{goel.aviral@gmail.com}{https://orcid.org/0000-0002-0814-5015}{(Optional) author-specific funding acknowledgements}%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty.
+Please provide at least the name of the affiliation and the country.
+The full address is optional
+
+%\author{Joan R.
+Public\footnote{Optional footnote, e.g.
+to mark corresponding author}}{Department of Informatics, Dummy College, [optional: Address], Country}{joanrpublic@dummycollege.org}{[orcid]}{[funding]}
+
+\authorrunning{J.\,Q. Public and J.\,R. Public} %TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'
+
+\Copyright{Aviral Goel, Pierre Donat-Bouillud, Filip Křikava, Christoph Kirsch and Jan Vitek}
+% TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/
 
@@
 
+\keywords{Dummy keyword} %TODO mandatory; please add comma-separated list of keywords
+
+\category{} %optional, e.g. invited paper
+
+\relatedversion{} %optional, e.g. full version hosted on arXiv, HAL, or other respository/website
+%\relatedversion{A full version of the paper is available at \url{...}.}
+
+\supplement{}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...
+
+%\funding{(Optional) general funding statement \dots}%optional, to capture a funding statement, which applies to all authors.
+Please enter author specific funding statements as fifth argument of the \author macro.
+
+\acknowledgements{I want to thank \dots}%optional
+
+%\nolinenumbers %uncomment to disable line numbering
+
+%\hideLIPIcs  %uncomment to remove references to LIPIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{document}
+
 \maketitle
+
+\begin{abstract}
+  The \eval function turns text into code. From the point of view of
+  reasoning about software system, uses of \eval turn the program's code
+  from a statically known quantity that can be reasoned about and analyzed
+  into one of the inputs to the program. Thus, widespread use of \eval
+  hinders our ability to provide safety and security guarantees for
+  software. Understanding how \eval is used in practice is key to finding
+  ways to mitigate its damage. One question that has not been studied is
+  whether the usage of \eval is dependent on the application domain or other
+  features of the language. This paper is a large-scale study of a corpus of
+  \CorpusAllCodeRnd lines of data science libraries and end-user code written in
+  the R language. We performed a sophisticated dynamic analysis to confirm
+  that \eval is indeed in wide spread use, and categorized patterns of
+  usage. We found that \eval in this code base is more dangerous in some
+  ways, and safer in others, than what was previously reported for
+  JavaScript.
+\end{abstract}
 

### 2020-11-05 inst/analysis/environments.Rmd
Analyze environment for the absolute category

@@
 
-eval_calls <- eval_calls %>% filter(!is.na(envir_expression))
+eval_calls_explicit <- eval_calls %>% filter(!is.na(envir_expression))
 ```
@@
 ```{r envir_list}
-eval_calls_env_list <- eval_calls %>% filter(envir_type == "VECSXP")
+eval_calls_env_list <- eval_calls_explicit %>% filter(envir_type == "VECSXP")
 ```
@@
 ```{r envir_arg_caller}
-eval_calls_arg_caller <- eval_calls %>% filter(!is.na(envir_from_arg))
+eval_calls_arg_caller <- eval_calls_explicit %>% filter(!is.na(envir_from_arg))
 ```
@@
 
-In the current dataset, there are no examples of sandbox usage but we should see adistance of 1 in that case.
+In the current dataset, there are no examples of sandbox usage but we should see a distance of 1 in that case.
 
@@
 ```{r envir_absolute}
-eval_calls_absolute <- eval_calls %>% anti_join(eval_calls_env_list) %>% anti_join(eval_calls_arg_caller)
+eval_calls_absolute <- eval_calls_explicit %>% anti_join(eval_calls_env_list) %>% anti_join(eval_calls_arg_caller)
 ```
@@
 ```{r}
-eval_calls_absolute <- eval_calls_absolute  %>% add_fake_srcref() %>% group_by(eval_call_srcref) %>%  mutate(envir_distinct = n_distinct(environment_class)) %>% ungroup()
+eval_calls_absolute <- eval_calls_absolute  %>% add_fake_srcref() %>% group_by(eval_call_srcref) %>%  mutate(envir_distinct = n_distinct(environment_class))
+
+eval_calls_absolute %>% ungroup() %>% count(envir_distinct, sort = TRUE)
+```
+
+We can also see a few examples:
+
+```{r}
+eval_calls_absolute %>% filter(envir_distinct == 2) %>% select(eval_call_expression, environment_class, envir_distinct)
 ```

### 2020-11-05 inst/paper/main.tex
Reorganize sections

@@
 precision for any operation following \eval. A group of influential
-resaerchers argued to give up on soundness and, instead, to
+researchers argued to give up on soundness and, instead, to
 under-approximate dynamic features~\cite{soundy} In their words ``a
@@
 
-\section{Infrastructure}
+\section{Methodology}
+
+\subsection{Infrastructure}
 
@@
 
-\section{Corpus}
+\subsection{Corpus}
 
@@
 
-\subsection{CRAN Packages}
+\subsubsection{CRAN Packages}
 
@@
 
-\subsection{Kaggle Scripts}
+\subsubsection{Kaggle Scripts}
 
@@
 
-\section{Usage Metrics}
+\subsection{Threats to Validity}
+The primary threat to validity is the issue of
+code coverage, a cause of concern for any dynamic approach. We mitigate this
+by including focusing on R packages that have high code coverage. We turn
+off the bytecode compiler for this study. The bytecode compiler can also
+call \eval. We do not get source locations for \UndefinedEvalsRnd eval
+calls. In these cases \eval is either passed as an argument to a
+higher-order functions or is defined in a function returned by a
+higher-order function and the R parser does not retain location information
+for \eval. However, this is a meager \PercentUndefinedEval of all eval calls
+and is unlikely to affect our analysis. We ignore calls to the native \eval
+function exposed by R. We also ignore the \c{rlang::tidy_eval} function
+which uses native \eval internally because \c{rlang} is used to implement a
+DSL for data analysis in R. It introduces a new first-class promise object
+called \c{quosure} for which it implements special evaluation support in
+\c{tidy_eval}.
+
+\section{Analysis of Eval}
+
+This section presents the results of our empirical study of \eval in R as
+answers to four research questions.
+
+\subsection{Use of Eval}
+
+The first research question we address is how and why \eval is used in R code.
+
+\begin{center}{\bfseries RQ1: What are the patterns of usage of \eval in the corpus?}\end{center}
 
@@
 
-\begin{figure}[!b]
-	\centering
-\includegraphics[width=\columnwidth]{ast_sizes}
-\caption{Input sizes} \label{fig:ast-size}
-\end{figure}
-
-To roughly estimate the complexity of the code being evaluated,
-Fig.~\ref{fig:ast-size} shows the distribution of the abstract syntax tree
-size of the \c{expr} argument to \eval. The majority of inputs across core
-and packages (\NbAstOnePercent of \eval calls and \NbAstOneCallSitePercent
-of call sites) have size 1. That is to say, the code being evaluated
-consists of a single variable name or constant. A function call needs at
-least three nodes and a list such as \c{c(1, 2, 3)} counts for 4
-nodes. Inputs smaller than \AstSizeNineFive nodes account for 95\% of all
-calls. Sizes in core are larger than in packages. Most large ASTs originate
-from library \c{VGAM}, which is used for statistical modeling, and also calls functions using \eval from core. It uses
-\eval to evaluate large matrices with values in a given environment.
-
 We compare the number of events intercepted during execution of \eval across
@@
 
-\medskip\noindent\emph{Discussion:} Comparing the use of \eval in R and
-JavaScript is instructive. The use of \eval is more widespread in data
-science code than it was in JavaScript. Any run of an R program seems to
-trigger \eval; moreover package programmers use \eval liberally. On the
-other hand, the authors of Kaggle packages do not use the feature at all.
-This point to a bifurcated user community: experts who write package know
-how to use \eval while end-users do not. An alternative explanation is that
-end-users do not need \eval for simple data analysis tasks. This bifurcation
-was not observed in the JavaScript community, but this may have changed with
-the advent of Node.js ecosystem.  The distribution of input sizes is similar
-to that found in JavaScript, the majority of inputs being rather small.
-
-\section{A taxonomy of eval}
+\subsubsection{A taxonomy of eval}
 
@@
 
-\subsection{Operation Mix}
+\subsubsection{Operation Mix}
 
@@
 
-\subsection{Provenance}
-
-In JavaScript, the provenance of the string passed into \eval is a matter of
-concern as \eval can be used to breach the security of websites. The main
-danger comes from string that were obtained from external sources. If they
-are allowed to flow into an \eval, then an attacker could hijack the
-program.
-
-{\bf what have they found...}
-
-There are various ways to obtain expressions:
-\begin{itemize}
-  \item \c{substitute} synthesizes ASTs from expressions by replacing
-    symbols with their bindings in the specified environment.
-  \item \c{expression}  creates a vector of expression
-    objects from text.
-  \item \c{parse}, \c{str2expression} and \c{str2lang} turn strings into
-    expressions.
-\end{itemize}
-
-We observe 5.2\% cases where \eval directly evaluates the output of
-\c{substitute}, 0.7\% cases where output of \c{parse} is read directly and
-only 704 cases for \c{expression}. Most expressions consumed by \eval are
-generated by other functions.
-
-\c{eval(parse(...))} can be used for dynamic code loading. This forms the
-core of \c{source} and \c{sys.source} functions in R that are commonly used
-for loading code in R files in interactive settings. We investigated the
-number of cases in which the output of \c{parse} and its variants is passed
-to \eval, directly or transitively by tainting their output. This
-corresponds to \PercentParsedCallSites of the total \eval call sites and
-\PercentParsedEvals of the eval calls. We observed that very few of the eval
-call (\NbParseFilesRnd in total) consume the result of calling \c{parse} on
-a file. Most of the eval calls consume the result of calling \c{parse} on a
-string.  We also identified one function in core R,
-\c{invokeRestartInteractively} that prompts the user for input, parses it,
-and passes it to \eval.
-
-\subsection{Side-Effects}
-
-In the corpus we observe \AllWritesRnd writes to variables of which
-\EvalWritesRnd writes happen inside \eval. However, all writes are not
-dangerous. Only writes to environments not local to the computation spawned
-by \eval are side-effecting. These writes outlive the computation and hence
-are visible outside it. The remaining writes are local to the
-computation. We observe that \EvalSideEffectingWritesRnd writes inside evals
-are side-effecting. This is only \EvalSideEffectingWritesEvalPerc of all
-variable writes inside \eval and \EvalSideEffectingWritesAllPerc of all
-variable writes in the corpus.
-
-An \eval is considered side-effecting if it performs a side-effecting write to a
-variable, directly or indirectly. Only \SideEffectingCoreCallPerc \eval calls in
-Core R are side-effecting and \SideEffectingPackageCallPerc \eval calls in CRAN
-packages are side-effecting.
-
-\subsection{Scope}
-
-One of the main differences of the R \eval with the javascript one is how it
-interacts with \emph{scope}, or environments in R. In Javascript, \eval can
-access the local scope and the global scope. In R, \eval has an argument
-\c{envir} to indicate in which environment its first argument must be evaluated.
-By default, it is equal to \c{parent.frame()}, 
-which is the parent environment of \eval call. This default argument
-happens in \DefaultEnvirExprPercent of the eval calls.
-
-This argument can also be \c{NULL}, a \c{List} or a data frame. This happens
-in 3.5\% of the cases: \eval copies the fields of the list or data frame
-and creates bindings for them in a new environment. This pattern is used to
-evaluate formulas which can directly refer to the fields of the data.  The
-\c{envir} argument can also be a number $n$. It means that the environment
-in which the expressions is evaluated will be the result of \c{sys.call(n)}
-where $n$ refers to the $n$-th stack frame.
-
-The top-level environment in R is called the global environment. New
-environments can be created using \c{new.env}. They can be provided a parent
-environment which becomes the enclosing scope of the new environment.
-
-We looked at the environments passed to all the \eval calls in our corpus.
-Table~\ref{tab:environments} summarizes the results. A numeric environment
-class \c{n} denotes the environment of the $n$-th call stack frame from the
-current function. \c{global} denotes the top-level environment and \c{list}
-denotes a list passed for evaluation of formulas. Environment classes of the
-form $n+$ denote the $n$-th environment extended with a new environment. The
-new environment provides a limited form of sandboxing. All assignments using
-the \c{\<-} function occur inside it and prevent the extended environment
-from mutation. However, it is still possible to mutate the extended
-environment using the \c{\<\<-} or \c{assign} functions; but, that happens
-rarely.
-
-\begin{table}[htbp] \resizebox{\columnwidth}{!}{%
-    \centering
-    \begin{tabular}{c|c|c|c}
-        \hline
-        \multicolumn{2}{c|}{Core} & \multicolumn{2}{|c}{Packages}  \\
-        \hline
-        Environment Class & Eval Calls \% & Environment Class & Eval Calls \% \\
-        \hline
-        \CoreEnvClassA & \CoreProportionA & \PackagesEnvClassA &  \PackagesProportionA\\
-        \CoreEnvClassB & \CoreProportionB & \PackagesEnvClassB &  \PackagesProportionB\\
-        \CoreEnvClassC & \CoreProportionC & \PackagesEnvClassC &  \PackagesProportionC\\
-        \CoreEnvClassD & \CoreProportionD & \PackagesEnvClassD &  \PackagesProportionD\\
-        \CoreEnvClassE & \CoreProportionE & \PackagesEnvClassE &  \PackagesProportionE\\
-        \hline
-    \end{tabular}} \label{tab:environments}
-\caption{Environments in terms of eval calls.}
-\end{table}
-
-We observe that a disproportionately high number of core R \eval calls
-access the caller's caller's environment. This is because many core R
-functions call functions that pass the result of \c{parent.frame()} to an
-\eval.  A disproportionately high number of calls to \eval happens in an
-extended top-level environment. This can be explained by the fact that many
-packages evaluate code passed from the user's workspace in the top-level
-environment to access the bindings.
-
-The \c{imchange} function of package \c{imager} makes it possible to modify
-images using a dedicated formula syntax using \c{\~}.\footnote{Inspired by
- {map} in package \emph{purr}.}  Here, \eval is evaluated in \c{newenv},
-which creates a new environment that inherits from \c{parent.frame()} by
-default (classified as 1+).
-
-\begin{lstlisting}
-newenv <- new.env()
-...
-fo <- parse(text=as.character(fo)[2])
-im[where] <- eval(fo,envir=newenv,enclos=env)
-\end{lstlisting}
-
-\c{adjCoef} in package \emph{actuar} find the root of an equation defined by
-a function \c{h} whose arguments must be named \c{x} and \c{y}.  \c{h} is
-transformed into an auxiliary function \c{h2} that can be optimized. Here,
-the list used for \c{envir} ensures the correspondance between the textual
-arguments of \c{h} and the arguments of \c{h2}.
-
-\begin{lstlisting}
-sh <- substitute(h)
-fcall <- paste(sh, "(x, y)")
-...
-h2 <- function(x, y)
-    eval(parse(text = fcall),
-    envir = list(x = x, y = y),
-    enclos = parent.frame(2))
-\end{lstlisting}
-
-\subsection{Patterns}
+\subsubsection{Usage Patterns}
 
@@
 
-\subsection{Consistency}
+\subsection{Comparison with eval in Javascript}
+
+\medskip\noindent\emph{Discussion:} Comparing the use of \eval in R and
+JavaScript is instructive. The use of \eval is more widespread in data
+science code than it was in JavaScript. Any run of an R program seems to
+trigger \eval; moreover package programmers use \eval liberally. On the
+other hand, the authors of Kaggle packages do not use the feature at all.
+This point to a bifurcated user community: experts who write package know
+how to use \eval while end-users do not. An alternative explanation is that
+end-users do not need \eval for simple data analysis tasks. This bifurcation
+was not observed in the JavaScript community, but this may have changed with
+the advent of Node.js ecosystem.  The distribution of input sizes is similar
+to that found in JavaScript, the majority of inputs being rather small.
+
+\begin{figure}[!b]
+	\centering
+  \includegraphics[width=\columnwidth]{ast_sizes}
+  \caption{Input sizes} \label{fig:ast-size}
+\end{figure}
+
+To roughly estimate the complexity of the code being evaluated,
+Fig.~\ref{fig:ast-size} shows the distribution of the abstract syntax tree
+size of the \c{expr} argument to \eval. The majority of inputs across core
+and packages (\NbAstOnePercent of \eval calls and \NbAstOneCallSitePercent
+of call sites) have size 1. That is to say, the code being evaluated
+consists of a single variable name or constant. A function call needs at
+least three nodes and a list such as \c{c(1, 2, 3)} counts for 4
+nodes. Inputs smaller than \AstSizeNineFive nodes account for 95\% of all
+calls. Sizes in core are larger than in packages. Most large ASTs originate
+from library \c{VGAM}, which is used for statistical modeling, and also calls functions using \eval from core. It uses
+\eval to evaluate large matrices with values in a given environment.
+
+\subsubsection{Consistency}
 
@@
 are highly \emph{polymorphic} (10 different types). They are the pipe
-operators \c{\%>\%}, \c{\%<>\%} and \c{\%$\%} in package \emph{magrittr}. It
+operators \c{\%>\%}, \c{\%<>\%} and \c{\%\$\%} in package \emph{magrittr}. It
 is effectively used to compose functions on their first argument, which can
@@
 
-\section{Threats to Validity} The primary threat to validity is the issue of
-code coverage, a cause of concern for any dynamic approach. We mitigate this
-by including focusing on R packages that have high code coverage. We turn
-off the bytecode compiler for this study. The bytecode compiler can also
-call \eval. We do not get source locations for \UndefinedEvalsRnd eval
-calls. In these cases \eval is either passed as an argument to a
-higher-order functions or is defined in a function returned by a
-higher-order function and the R parser does not retain location information
-for \eval. However, this is a meager \PercentUndefinedEval of all eval calls
-and is unlikely to affect our analysis. We ignore calls to the native \eval
-function exposed by R. We also ignore the \c{rlang::tidy_eval} function
-which uses native \eval internally because \c{rlang} is used to implement a
-DSL for data analysis in R. It introduces a new first-class promise object
-called \c{quosure} for which it implements special evaluation support in
-\c{tidy_eval}.
+\subsection{How Dangerous is eval}
+
+\subsubsection{Provenance}
+
+In JavaScript, the provenance of the string passed into \eval is a matter of
+concern as \eval can be used to breach the security of websites. The main
+danger comes from string that were obtained from external sources. If they
+are allowed to flow into an \eval, then an attacker could hijack the
+program.
+
+{\bf what have they found...}
+
+There are various ways to obtain expressions:
+\begin{itemize}
+  \item \c{substitute} synthesizes ASTs from expressions by replacing
+    symbols with their bindings in the specified environment.
+  \item \c{expression}  creates a vector of expression
+    objects from text.
+  \item \c{parse}, \c{str2expression} and \c{str2lang} turn strings into
+    expressions.
+\end{itemize}
+
+We observe 5.2\% cases where \eval directly evaluates the output of
+\c{substitute}, 0.7\% cases where output of \c{parse} is read directly and
+only 704 cases for \c{expression}. Most expressions consumed by \eval are
+generated by other functions.
+
+\c{eval(parse(...))} can be used for dynamic code loading. This forms the
+core of \c{source} and \c{sys.source} functions in R that are commonly used
+for loading code in R files in interactive settings. We investigated the
+number of cases in which the output of \c{parse} and its variants is passed
+to \eval, directly or transitively by tainting their output. This
+corresponds to \PercentParsedCallSites of the total \eval call sites and
+\PercentParsedEvals of the eval calls. We observed that very few of the eval
+call (\NbParseFilesRnd in total) consume the result of calling \c{parse} on
+a file. Most of the eval calls consume the result of calling \c{parse} on a
+string.  We also identified one function in core R,
+\c{invokeRestartInteractively} that prompts the user for input, parses it,
+and passes it to \eval.
+
+\subsubsection{Side-Effects}
+
+In the corpus we observe \AllWritesRnd writes to variables of which
+\EvalWritesRnd writes happen inside \eval. However, all writes are not
+dangerous. Only writes to environments not local to the computation spawned
+by \eval are side-effecting. These writes outlive the computation and hence
+are visible outside it. The remaining writes are local to the
+computation. We observe that \EvalSideEffectingWritesRnd writes inside evals
+are side-effecting. This is only \EvalSideEffectingWritesEvalPerc of all
+variable writes inside \eval and \EvalSideEffectingWritesAllPerc of all
+variable writes in the corpus.
+
+An \eval is considered side-effecting if it performs a side-effecting write to a
+variable, directly or indirectly. Only \SideEffectingCoreCallPerc \eval calls in
+Core R are side-effecting and \SideEffectingPackageCallPerc \eval calls in CRAN
+packages are side-effecting.
+
+\subsubsection{Scope}
+
+One of the main differences of the R \eval with the javascript one is how it
+interacts with \emph{scope}, or environments in R. In Javascript, \eval can
+access the local scope and the global scope. In R, \eval has an argument
+\c{envir} to indicate in which environment its first argument must be evaluated.
+By default, it is equal to \c{parent.frame()}, 
+which is the parent environment of \eval call. This default argument
+happens in \DefaultEnvirExprPercent of the eval calls.
+
+This argument can also be \c{NULL}, a \c{List} or a data frame. This happens
+in 3.5\% of the cases: \eval copies the fields of the list or data frame
+and creates bindings for them in a new environment. This pattern is used to
+evaluate formulas which can directly refer to the fields of the data.  The
+\c{envir} argument can also be a number $n$. It means that the environment
+in which the expressions is evaluated will be the result of \c{sys.call(n)}
+where $n$ refers to the $n$-th stack frame.
+
+The top-level environment in R is called the global environment. New
+environments can be created using \c{new.env}. They can be provided a parent
+environment which becomes the enclosing scope of the new environment.
+
+We looked at the environments passed to all the \eval calls in our corpus.
+Table~\ref{tab:environments} summarizes the results. A numeric environment
+class \c{n} denotes the environment of the $n$-th call stack frame from the
+current function. \c{global} denotes the top-level environment and \c{list}
+denotes a list passed for evaluation of formulas. Environment classes of the
+form $n+$ denote the $n$-th environment extended with a new environment. The
+new environment provides a limited form of sandboxing. All assignments using
+the \c{\<-} function occur inside it and prevent the extended environment
+from mutation. However, it is still possible to mutate the extended
+environment using the \c{\<\<-} or \c{assign} functions; but, that happens
+rarely.
+
+\begin{table}[htbp] \resizebox{\columnwidth}{!}{%
+    \centering
+    \begin{tabular}{c|c|c|c}
+        \hline
+        \multicolumn{2}{c|}{Core} & \multicolumn{2}{|c}{Packages}  \\
+        \hline
+        Environment Class & Eval Calls \% & Environment Class & Eval Calls \% \\
+        \hline
+        \CoreEnvClassA & \CoreProportionA & \PackagesEnvClassA &  \PackagesProportionA\\
+        \CoreEnvClassB & \CoreProportionB & \PackagesEnvClassB &  \PackagesProportionB\\
+        \CoreEnvClassC & \CoreProportionC & \PackagesEnvClassC &  \PackagesProportionC\\
+        \CoreEnvClassD & \CoreProportionD & \PackagesEnvClassD &  \PackagesProportionD\\
+        \CoreEnvClassE & \CoreProportionE & \PackagesEnvClassE &  \PackagesProportionE\\
+        \hline
+    \end{tabular}} \label{tab:environments}
+\caption{Environments in terms of eval calls.}
+\end{table}
+
+We observe that a disproportionately high number of core R \eval calls
+access the caller's caller's environment. This is because many core R
+functions call functions that pass the result of \c{parent.frame()} to an
+\eval.  A disproportionately high number of calls to \eval happens in an
+extended top-level environment. This can be explained by the fact that many
+packages evaluate code passed from the user's workspace in the top-level
+environment to access the bindings.
+
+The \c{imchange} function of package \c{imager} makes it possible to modify
+images using a dedicated formula syntax using \c{\~}.\footnote{Inspired by
+ {map} in package \emph{purr}.}  Here, \eval is evaluated in \c{newenv},
+which creates a new environment that inherits from \c{parent.frame()} by
+default (classified as 1+).
+
+\begin{lstlisting}
+newenv <- new.env()
+...
+fo <- parse(text=as.character(fo)[2])
+im[where] <- eval(fo,envir=newenv,enclos=env)
+\end{lstlisting}
+
+\c{adjCoef} in package \emph{actuar} find the root of an equation defined by
+a function \c{h} whose arguments must be named \c{x} and \c{y}.  \c{h} is
+transformed into an auxiliary function \c{h2} that can be optimized. Here,
+the list used for \c{envir} ensures the correspondance between the textual
+arguments of \c{h} and the arguments of \c{h2}.
+
+\begin{lstlisting}
+sh <- substitute(h)
+fcall <- paste(sh, "(x, y)")
+...
+h2 <- function(x, y)
+    eval(parse(text = fcall),
+    envir = list(x = x, y = y),
+    enclos = parent.frame(2))
+\end{lstlisting}
+
+\subsection{Removing Eval}
 

### 2020-11-05 inst/paper/main.tex
Change heading
Add key points to discuss for how dangerous is eval

@@
 
-\section{Analysis of Eval}
+\section{Evaluation}
 
@@
 
-\subsection{Comparison with eval in Javascript}
-
-\medskip\noindent\emph{Discussion:} Comparing the use of \eval in R and
-JavaScript is instructive. The use of \eval is more widespread in data
-science code than it was in JavaScript. Any run of an R program seems to
-trigger \eval; moreover package programmers use \eval liberally. On the
-other hand, the authors of Kaggle packages do not use the feature at all.
-This point to a bifurcated user community: experts who write package know
-how to use \eval while end-users do not. An alternative explanation is that
-end-users do not need \eval for simple data analysis tasks. This bifurcation
-was not observed in the JavaScript community, but this may have changed with
-the advent of Node.js ecosystem.  The distribution of input sizes is similar
-to that found in JavaScript, the majority of inputs being rather small.
-
-\begin{figure}[!b]
-	\centering
-  \includegraphics[width=\columnwidth]{ast_sizes}
-  \caption{Input sizes} \label{fig:ast-size}
-\end{figure}
-
-To roughly estimate the complexity of the code being evaluated,
-Fig.~\ref{fig:ast-size} shows the distribution of the abstract syntax tree
-size of the \c{expr} argument to \eval. The majority of inputs across core
-and packages (\NbAstOnePercent of \eval calls and \NbAstOneCallSitePercent
-of call sites) have size 1. That is to say, the code being evaluated
-consists of a single variable name or constant. A function call needs at
-least three nodes and a list such as \c{c(1, 2, 3)} counts for 4
-nodes. Inputs smaller than \AstSizeNineFive nodes account for 95\% of all
-calls. Sizes in core are larger than in packages. Most large ASTs originate
-from library \c{VGAM}, which is used for statistical modeling, and also calls functions using \eval from core. It uses
-\eval to evaluate large matrices with values in a given environment.
-
-\subsubsection{Consistency}
-
-We look at how \emph{consistent} the \c{expr} argument of \eval can be, \ie
-how many different types of the resolved \c{expr} there are per call
-sites. Most of the call sites, \ie \PercentMonomorphic, are
-\emph{consistent}, and this is similar to javascript. However, a few ones
-are highly \emph{polymorphic} (10 different types). They are the pipe
-operators \c{\%>\%}, \c{\%<>\%} and \c{\%\$\%} in package \emph{magrittr}. It
-is effectively used to compose functions on their first argument, which can
-be of any type.
-
 \subsection{How Dangerous is eval}
+
+\subsubsection{How much code is executed inside eval}
+- measure number of operations inside eval and outside eval.
+- this is useful because if analyzers treat eval as a nop, they may miss X\% of the code.
+
+\subsubsection{Does eval create observable side-effects}
+- measure number of reads and writes by eval outside of its ``bubble''.
+- Since R is lazy, even reads can cause promise evaluation and hence, side-effects.
+- Look at writes to non-local scopes. Compare with total number of writes for
+reference?
+- Categorize side-effects between env passed to eval and other envs.
+
+\subsubsection{Does eval do reflection}
+\eval can access parent scopes via reflection. Find number of times it does that
+and the depth to which it access parent frames.
+
+\subsubsection{Does eval introduce new bindings}
+Find number of times, eval introduces new bindings. There are many ways -- library,
+load, attach, source, and explicitly introduce bindings using super-assign, assign and define.
+
+\subsubsection{Does eval call C code}
+Find number of times, eval makes call to native code using - .Call, .External,
+etc. Find out if there are native functions that are called only from within eval.
+
+\subsubsection{Does eval add finalizers}
+Eval can attach code to be executed when the object goes out of scope or program
+exits. Find if eval does that.
+
+\subsubsection{Does eval do non-local returns}
+Eval can do non-local returns effective bypassing evaluation of the rest of the
+function. This can be useful for static analyzers.
+
+Conclude with number of evals that are ``pure'', i.e. evals which
+could be ignored by a static analyzer without any problem.
 
@@
 
+\subsection{Comparison with eval in Javascript}
+
+\medskip\noindent\emph{Discussion:} Comparing the use of \eval in R and
+JavaScript is instructive. The use of \eval is more widespread in data
+science code than it was in JavaScript. Any run of an R program seems to
+trigger \eval; moreover package programmers use \eval liberally. On the
+other hand, the authors of Kaggle packages do not use the feature at all.
+This point to a bifurcated user community: experts who write package know
+how to use \eval while end-users do not. An alternative explanation is that
+end-users do not need \eval for simple data analysis tasks. This bifurcation
+was not observed in the JavaScript community, but this may have changed with
+the advent of Node.js ecosystem.  The distribution of input sizes is similar
+to that found in JavaScript, the majority of inputs being rather small.
+
+\begin{figure}[!b]
+	\centering
+  \includegraphics[width=\columnwidth]{ast_sizes}
+  \caption{Input sizes} \label{fig:ast-size}
+\end{figure}
+
+To roughly estimate the complexity of the code being evaluated,
+Fig.~\ref{fig:ast-size} shows the distribution of the abstract syntax tree
+size of the \c{expr} argument to \eval. The majority of inputs across core
+and packages (\NbAstOnePercent of \eval calls and \NbAstOneCallSitePercent
+of call sites) have size 1. That is to say, the code being evaluated
+consists of a single variable name or constant. A function call needs at
+least three nodes and a list such as \c{c(1, 2, 3)} counts for 4
+nodes. Inputs smaller than \AstSizeNineFive nodes account for 95\% of all
+calls. Sizes in core are larger than in packages. Most large ASTs originate
+from library \c{VGAM}, which is used for statistical modeling, and also calls functions using \eval from core. It uses
+\eval to evaluate large matrices with values in a given environment.
+
+\subsubsection{Consistency}
+
+We look at how \emph{consistent} the \c{expr} argument of \eval can be, \ie
+how many different types of the resolved \c{expr} there are per call
+sites. Most of the call sites, \ie \PercentMonomorphic, are
+\emph{consistent}, and this is similar to javascript. However, a few ones
+are highly \emph{polymorphic} (10 different types). They are the pipe
+operators \c{\%>\%}, \c{\%<>\%} and \c{\%\$\%} in package \emph{magrittr}. It
+is effectively used to compose functions on their first argument, which can
+be of any type.
+
 \subsection{Removing Eval}

### 2020-11-09 inst/analysis/enrich_dataset.Rmd
Move to .4 dataset

@@
       # return a character string to show the time
-      paste("Time for this code chunk to run:", res, unit(res))
+      paste("Time for this code chunk to run:", res, units(res))
     }
@@
 now <- Sys.time() 
-
 library(tidyverse)
@@
 
-```{r ast_size}
-#eval_calls <- eval_calls %>% add_ast_size()
-```
-
 ```{r add_package}
@@
 ```{r}
-eval_calls_ast_size <- eval_calls_corpus %>% select(eval_call_srcref, caller_package, caller_function, eval_source_type, expr_resolved_nodes, nb_ev_calls)
+# Used when using dataset 2 which has incorrect AST sizes.
 
-write_fst(eval_calls_ast_size, path(RUN_TRACE_DIR, "summarized-ast-size.fst"))
+#eval_calls_ast_size <- eval_calls_corpus %>% select(eval_call_srcref, caller_package, caller_function, eval_source_type, expr_resolved_nodes, nb_ev_calls)
+
+#write_fst(eval_calls_ast_size, path(RUN_TRACE_DIR, "summarized-ast-size.fst"))
 ```
@@
 max_call_sites <- max(call_sites_per_package$n)
-r("max call sites enrich")
+r("max call sites enrich", max_call_sites)
+
+one_call_sites <- nrow(call_sites_per_package %>% filter(n == 1))
+r("one call sites enrich", one_call_sites)
+
+no_call_sites <- nrow(call_sites_per_package %>% filter(n == 0))
+r("no call sites enrich", no_call_sites)
 ```
@@
 ```{r write_back}
-write_fst(undefined_per_package, PACKAGE_EVALS_UNDEFINED_FILE)
+write_fst(undefined_per_package, EVALS_UNDEFINED_FILE)
 

### 2020-11-25 inst/analysis/preprocess.R
Better formatting for preprocess

@@ -12,102 +12,169 @@
 source("insights.R")
 
 library(fs)
-#library(future.apply)
-#plan(multiprocess) # multicore (fork) with fallback to multisession (create processes and then copy)
+
 
 deduplicate <- function(dataset)
 {
-  return(dataset %>% count(across(c(-eval_call_id, -starts_with("caller_stack"))), name = "nb_ev_calls"))
+  return(dataset %>% count(across(c(
+    -eval_call_id,-starts_with("caller_stack")
+  )), name = "nb_ev_calls"))
 }
 
 add_types <- function(dataset)
 {
-  return(dataset %>%
-           resolve_sexp_name(expr_expression_type) %>%
-           resolve_sexp_name(expr_resolved_type) %>%
-           resolve_sexp_name(enclos_type) %>%
-           resolve_sexp_name(envir_type))
+  return(
+    dataset %>%
+      resolve_sexp_name(expr_expression_type) %>%
+      resolve_sexp_name(expr_resolved_type) %>%
+      resolve_sexp_name(enclos_type) %>%
+      resolve_sexp_name(envir_type)
+  )
 }
 
 add_parse_args <- function(dataset)
 {
-  return(dataset %>% mutate(parse_args = map(expr_parsed_expression, function(e) 
-    { if(!is.na(e) && str_starts(e, "(parse|str2lang|str2expression)\\(")) 
-        extract_args_parse(e) 
-      else list() }))  %>% unnest_wider(parse_args))
-}
-
-eval_base_functions <- c("autoload", "autoloader", "bquote", "by.default", "by.data.frame", "invokeRestartInteractively", "Ops.data.frame", "dget", 
-                         "eval", "eval.parent", "evalq", "local", "with.default", "within.data.frame", "within.list", "replicate", "subset.data.frame",
-                         "subset.matrix", "transform.data.frame", "match.arg", "char.expand", "max.col", "parseNamespaceFile", "source", "sys.source",
-                         "stopifnot", "as.data.frame.table", "match.fun", "trace", "untrace", ".doTrace", "Vectorize")
-
-
-find_package_name <- function(caller_function, caller_package, caller_expression, srcref, file)
-{
-  if(caller_function %in% eval_base_functions)
-    return("core") #"Actually base but we generalize
-  else
+  return(dataset %>% mutate(parse_args = map(expr_parsed_expression, function(e)
   {
-    tempPack <- extract_package_name(srcref, file)
-    if(tempPack == "base?") # It means the srcref was NA
-    {
[…]

### 2020-11-26 inst/paper/main.tex
Reorganize sections
LIPICS cs formatting for ECOOP

--- 2020-11-05 inst/paper/main.tex
@@
-\documentclass[a4paper,USenglish,cleveref, autoref, thm-restate]{lipics-v2019}
-% This is a template for producing LIPIcs articles. 
-% See lipics-manual.pdf for further information.
-% for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
-% for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
-% for section-numbered lemmas etc., use "numberwithinsect"
-% for enabling cleveref support, use "cleveref"
-% for enabling autoref support, use "autoref"
-% for anonymousing the authors (e.g. for double-blind review), add "anonymous"
-% for enabling thm-restate support, use "thm-restate"
-
-% \graphicspath{{./graphics/}}%helpful if your graphic files are in another directory
+\documentclass[USenglish,cleveref, autoref, thm-restate]{lipics-v2019}
 
@@
     drop, which, grid, print, ncol, dim, max, format, sort, rug, matrix, start,
-    unique, mean, df, attr, do, power, \$},
+    unique, mean, df, attr, do, power},
   otherkeywords={},
@@
 \newcommand{\runr}{\emph{runr}\xspace}
-
-\bibliographystyle{plainurl}
+\bibliographystyle{plainurl}% the mandatory bibstyle
+
+\begin{document}
 
 \title{A Large-Scale Study of the Use of Eval in R}
-
-%\author{Aviral Goel}{Northeastern University, [optional: Address], Country \and My second affiliation, Country \and \url{http://www.myhomepage.edu} }{goel.aviral@gmail.com}{https://orcid.org/0000-0002-0814-5015}{(Optional) author-specific funding acknowledgements}%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty.
-Please provide at least the name of the affiliation and the country.
-The full address is optional
-
-%\author{Joan R.
-Public\footnote{Optional footnote, e.g.
-to mark corresponding author}}{Department of Informatics, Dummy College, [optional: Address], Country}{joanrpublic@dummycollege.org}{[orcid]}{[funding]}
-
-\authorrunning{J.\,Q. Public and J.\,R. Public} %TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'
-
-\Copyright{Aviral Goel, Pierre Donat-Bouillud, Filip Křikava, Christoph Kirsch and Jan Vitek}
-% TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/
-
-\begin{CCSXML}
-  <ccs2012>
-  <concept>
-  <concept_id>10011007.10011006.10011050.10010517</concept_id>
-  <concept_desc>Software and its engineering~Scripting languages</concept_desc>
-  <concept_significance>500</concept_significance>
-  </concept>
-  <concept>
-  <concept_id>10002944.10011123.10010912</concept_id>
-  <concept_desc>General and reference~Empirical studies</concept_desc>
-  <concept_significance>500</concept_significance>
-  </concept>
-  <concept>
-  <concept_id>10011007.10011006.10011008</concept_id>
-  <concept_desc>Software and its engineering~General programming languages</concept_desc>
-  <concept_significance>500</concept_significance>
-  </concept>
-  </ccs2012>
-\end{CCSXML}
-
-\ccsdesc[500]{Software and its engineering~Scripting languages}
-\ccsdesc[500]{General and reference~Empirical studies}
-\ccsdesc[500]{Software and its engineering~General programming languages}
-
-\keywords{Dummy keyword} %TODO mandatory; please add comma-separated list of keywords
-
-\category{} %optional, e.g. invited paper
-
-\relatedversion{} %optional, e.g. full version hosted on arXiv, HAL, or other respository/website
-%\relatedversion{A full version of the paper is available at \url{...}.}
-
-\supplement{}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...
-
-%\funding{(Optional) general funding statement \dots}%optional, to capture a funding statement, which applies to all authors.
-Please enter author specific funding statements as fifth argument of the \author macro.
-
-\acknowledgements{I want to thank \dots}%optional
-
-%\nolinenumbers %uncomment to disable line numbering
-
-%\hideLIPIcs  %uncomment to remove references to LIPIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository
+\titlerunning{Eval in R}
+
+\author{Anon}{Under review}{}{}{}
+\authorrunning{Anon}
+\ccsdesc[100]{Something}
+\Copyright{Anon}
+\keywords{R, reflection, eval}
+
+%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%\
+%%%%%%%                                                                         
+\EventEditors{Eds.}
+\EventNoEds{2}
+\EventLongTitle{European Conference on Object Oriented Programming (ECOOP 2021)}
+\EventShortTitle{ECOOP 2021}
+\EventAcronym{ECOOP}
+\EventYear{2021}
+\EventDate{December 24--27, 2016}
+\EventLocation{Little Whinging, United Kingdom}
+\EventLogo{}
+\SeriesVolume{42}
+\ArticleNo{23}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{document}
 

### 2020-11-26 inst/analysis/environments.Rmd

@@
 output:
-  rmdformats::readthedown:
+  html_document:
         code_folding: hide
-        lightbox: true
         gallery: false
-        df_print: paged
+        toc: true
         toc_depth: 3
+        toc_float: true
 date: "`r Sys.Date()`"
@@
 ```{r packages}
+nb_eval_calls <- count(eval_calls, wt = nb_ev_calls)
 
+nb_call_sites <- nb_eval_call_sites(eval_calls)
 ```
@@
 
-We distinguish between 4 main categories of `eval` calls with respect to their `envir` argument:
+We distinguish between 5 main categories of `eval` calls with respect to their `envir` argument:
 
 - `envir` is not provided explicitly (and thus is the default argument, which semantically belongs to the absolute category)
+- `envir` in an eval call generated by Rcpp
 - `envir` is a list or a data frame
@@
 
+nb_default <- count(eval_calls_default, wt = nb_ev_calls)
+nb_default_sites <- nb_eval_call_sites(eval_calls_default)
+
 eval_calls_explicit <- eval_calls %>% filter(!is.na(envir_expression))
+```
+
+# `envir` with Rcpp
+
+```{r envir_rcpp}
+eval_calls_rcpp <- eval_calls_explicit %>% filter(str_detect(expr_expression, "C\\+\\+Constructor|C\\+\\+Field|C\\+\\+OverloadedMethods"))
+
+nb_rcpp <- count(eval_calls_rcpp, wt = nb_ev_calls)
+nb_rcpp_sites <- nb_eval_call_sites(eval_calls_rcpp)
+
+eval_calls_explicit <- eval_calls_explicit %>% anti_join(eval_calls_rcpp)
 ```
@@
 eval_calls_env_list <- eval_calls_explicit %>% filter(envir_type == "VECSXP")
+
+nb_env_list <- count(eval_calls_env_list, wt = nb_ev_calls)
+nb_env_list_sites <- nb_eval_call_sites(eval_calls_env_list)
 ```
@@
 eval_calls_arg_caller <- eval_calls_explicit %>% filter(!is.na(envir_from_arg))
+
+nb_arg_caller <- count(eval_calls_arg_caller, wt = nb_ev_calls)
+nb_arg_caller_sites <- nb_eval_call_sites(eval_calls_arg_caller)
 ```
@@
 
-But all those strange calls are due to Rcpp:
-
-```{r arg_call_rcpp}
-eval_calls_arg_caller %>% filter(envir_from_arg > 1, !str_detect(expr_expression, "C\\+\\+Constructor|C\\+\\+Field|C\\+\\+OverloadedMethods")) 
-```
-
 In the current dataset, there are no examples of sandbox usage but we should see a distance of 1 in that case.
@@
 eval_calls_absolute <- eval_calls_explicit %>% anti_join(eval_calls_env_list) %>% anti_join(eval_calls_arg_caller)
+
+nb_absolute <- count(eval_calls_absolute, wt = nb_ev_calls)
+nb_absolute_sites <- nb_eval_call_sites(eval_calls_absolute)
 ```
@@
 
+## Summary of the categories
+
+Total eval calls: `r nb_eval_calls`
+
+Total eval call sites: `r nb_call_sites`
+
+- Default: `r nb_default`, call sites: `r nb_default_sites`
+- Rcpp: `r nb_rcpp`, call sites: `r nb_rcpp_sites`
+- Argument of caller: `r nb_arg_caller`, call sites: `r nb_arg_caller_sites`
+- Absolute: `r nb_absolute`, call sites: `r nb_absolute_sites`
+
 ```{r}

### 2020-11-26 inst/paper/main.tex

--- 2020-11-05 inst/paper/main.tex
@@
 
+\bibliographystyle{plainurl}% the mandatory bibstyle
+
+\title{A Large-Scale Study of the Use of Eval in R}
+\titlerunning{Eval in R}
+
+\author{Anon}{Under review}{}{}{}
+\authorrunning{Anon}
+\Copyright{Anon}
+\keywords{R, reflection, eval}
+
+%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%\
+%%%%%%%                                                                         
+\EventEditors{Eds.}
+\EventNoEds{2}
+\EventLongTitle{European Conference on Object Oriented Programming (ECOOP 2021)}
+\EventShortTitle{ECOOP 2021}
+\EventAcronym{ECOOP}
+\EventYear{2021}
+\EventDate{December 24--27, 2016}
+\EventLocation{Little Whinging, United Kingdom}
+\EventLogo{}
+\SeriesVolume{42}
+\ArticleNo{23}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \bibliographystyle{plainurl}
-
-\title{A Large-Scale Study of the Use of Eval in R}
-
-%\author{Aviral Goel}{Northeastern University, [optional: Address], Country \and My second affiliation, Country \and \url{http://www.myhomepage.edu} }{goel.aviral@gmail.com}{https://orcid.org/0000-0002-0814-5015}{(Optional) author-specific funding acknowledgements}%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty.
-Please provide at least the name of the affiliation and the country.
-The full address is optional
-
-%\author{Joan R.
-Public\footnote{Optional footnote, e.g.
-to mark corresponding author}}{Department of Informatics, Dummy College, [optional: Address], Country}{joanrpublic@dummycollege.org}{[orcid]}{[funding]}
-
-\authorrunning{J.\,Q. Public and J.\,R. Public} %TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'
-
-\Copyright{Aviral Goel, Pierre Donat-Bouillud, Filip Křikava, Christoph Kirsch and Jan Vitek}
-% TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/
 
@@
 \ccsdesc[500]{Software and its engineering~General programming languages}
-
-\keywords{Dummy keyword} %TODO mandatory; please add comma-separated list of keywords
-
-\category{} %optional, e.g. invited paper
-
-\relatedversion{} %optional, e.g. full version hosted on arXiv, HAL, or other respository/website
-%\relatedversion{A full version of the paper is available at \url{...}.}
-
-\supplement{}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...
-
-%\funding{(Optional) general funding statement \dots}%optional, to capture a funding statement, which applies to all authors.
-Please enter author specific funding statements as fifth argument of the \author macro.
-
-\acknowledgements{I want to thank \dots}%optional
-
-%\nolinenumbers %uncomment to disable line numbering
-
-%\hideLIPIcs  %uncomment to remove references to LIPIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-This section presents the results of our empirical study of \eval in R as
-answers to four research questions.
+This section presents the results of our empirical study of \eval in R in terms
+of four research questions.
 
@@
 
+\subsection{Necessity of Eval}
+
 \subsubsection{Unnecessary use of eval}
@@
 
-\subsection{How Dangerous is eval}
+\subsection{Power of Eval}
 
@@
 
-\subsection{Comparison with eval in Javascript}
+\section{Discussion}
+
+\subsection{Comparison with Javascript}
 
@@
 
-\subsection{Removing Eval}
-
 \section{Conclusion}

### 2020-11-26 inst/paper/main.tex
abstract

@@
 
-\title{A Large-Scale Study of the Use of Eval in R}
+\title{Why do we Eval?\\[2mm]\Large
+A large-scale study of Eval usage in R}
 \titlerunning{Eval in R}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\bibliographystyle{plainurl}
 
@@
 \begin{abstract}
-  The \eval function turns text into code. From the point of view of
-  reasoning about software system, uses of \eval turn the program's code
-  from a statically known quantity that can be reasoned about and analyzed
-  into one of the inputs to the program. Thus, widespread use of \eval
-  hinders our ability to provide safety and security guarantees for
-  software. Understanding how \eval is used in practice is key to finding
-  ways to mitigate its damage. One question that has not been studied is
-  whether the usage of \eval is dependent on the application domain or other
-  features of the language. This paper is a large-scale study of a corpus of
-  \CorpusAllCodeRnd lines of data science libraries and end-user code written in
-  the R language. We performed a sophisticated dynamic analysis to confirm
-  that \eval is indeed in wide spread use, and categorized patterns of
-  usage. We found that \eval in this code base is more dangerous in some
-  ways, and safer in others, than what was previously reported for
-  JavaScript.
+  \noindent Most dynamic languages allow users to turn text into code
+  using various functions, often named \eval, with language-dependent
+  semantics. The widespread use of these reflective functions hinders
+  static analysis and prevents compilers from performing many
+  optimizations. This paper aims to provide a better sense of why
+  programmers use \eval.  Understanding how \eval is used in practice
+  is key to finding ways to mitigate its negative impact. We have
+  reasons to believe that reflective feature usage is language and
+  application domain specific. In this paper, we focus on data science
+  code written in R, and compare our results to previous work that
+  analyzed web programming in JavaScript.  We conduct a large-scale
+  study of a corpus of \CorpusAllCodeRnd lines of libraries and
+  end-user code on which we performed dynamic analysis to confirm that
+  \eval is indeed in widespread use, and categorized patterns of
+  usage. We found that \eval is more dangerous in some ways, and safer
+  in others, than what was previously reported for JavaScript.
 \end{abstract}

### 2020-11-26 inst/paper/main.tex
intro

@@
 \documentclass[USenglish,cleveref, autoref, thm-restate]{lipics-v2019}
-
-\usepackage{listings}
-\usepackage{hyperref}
-\usepackage{multirow}
-\usepackage{paralist}
-\usepackage{xspace}
-\usepackage{url}
-
-\newcommand{\missingTag}[1]{\textcolor{red}{#1}\xspace}
-\newcommand{\missingNumber}{\textcolor{red}{XX}\xspace}
-\newcommand{\missingPercentage}{\textcolor{red}{XX\%}\xspace}
-\newcommand{\missingTable}[1][XX]{\textcolor{red}{Table #1}\xspace}
-\newcommand{\missingGraph}{\textcolor{red}{XXGraph}\xspace}
-
-%\input{tag/counts.tex}
-\newcommand{\CoreEvalCallCountRnd}{6.3M\xspace}
-\newcommand{\CoreEvalCallCount}{6,282,429\xspace}
-\newcommand{\CoreEvalSiteCountRnd}{185\xspace}
-\newcommand{\CoreEvalSiteCount}{185\xspace}
-\newcommand{\CoreEvalParentCallCountRnd}{186K\xspace}
-\newcommand{\CoreEvalParentCallCount}{186,275\xspace}
-\newcommand{\CoreEvalParentSiteCountRnd}{13\xspace}
-\newcommand{\CoreEvalParentSiteCount}{13\xspace}
-\newcommand{\CoreEvalqCallCountRnd}{132K\xspace}
-\newcommand{\CoreEvalqCallCount}{132,891\xspace}
-\newcommand{\CoreEvalqSiteCountRnd}{20K\xspace}
-\newcommand{\CoreEvalqSiteCount}{20,460\xspace}
-\newcommand{\CoreLocalCallCountRnd}{175K\xspace}
-\newcommand{\CoreLocalCallCount}{175,079\xspace}
-\newcommand{\CoreLocalSiteCountRnd}{920\xspace}
-\newcommand{\CoreLocalSiteCount}{920\xspace}
-\newcommand{\PackageEvalCallCountRnd}{1.6M\xspace}
-\newcommand{\PackageEvalCallCount}{1,582,738\xspace}
-\newcommand{\PackageEvalSiteCountRnd}{1.5K\xspace}
-\newcommand{\PackageEvalSiteCount}{1,459\xspace}
-\newcommand{\PackageEvalParentCallCountRnd}{13K\xspace}
-\newcommand{\PackageEvalParentCallCount}{13,230\xspace}
-\newcommand{\PackageEvalParentSiteCountRnd}{249\xspace}
-\newcommand{\PackageEvalParentSiteCount}{249\xspace}
-\newcommand{\PackageEvalqCallCountRnd}{2\xspace}
-\newcommand{\PackageEvalqCallCount}{2\xspace}
-\newcommand{\PackageEvalqSiteCountRnd}{1\xspace}
-\newcommand{\PackageEvalqSiteCount}{1\xspace}
-\newcommand{\PackageLocalCallCountRnd}{2.2K\xspace}
-\newcommand{\PackageLocalCallCount}{2,228\xspace}
-\newcommand{\PackageLocalSiteCountRnd}{15\xspace}
-\newcommand{\PackageLocalSiteCount}{15\xspace}
-\newcommand{\KaggleEvalCallCountRnd}{0\xspace}
-\newcommand{\KaggleEvalCallCount}{0\xspace}
-\newcommand{\KaggleEvalSiteCountRnd}{0\xspace}
-\newcommand{\KaggleEvalSiteCount}{0\xspace}
-\newcommand{\KaggleEvalParentCallCountRnd}{0\xspace}
-\newcommand{\KaggleEvalParentCallCount}{0\xspace}
-\newcommand{\KaggleEvalParentSiteCountRnd}{0\xspace}
-\newcommand{\KaggleEvalParentSiteCount}{0\xspace}
-\newcommand{\KaggleEvalqCallCountRnd}{0\xspace}
-\newcommand{\KaggleEvalqCallCount}{0\xspace}
-\newcommand{\KaggleEvalqSiteCountRnd}{0\xspace}
-\newcommand{\KaggleEvalqSiteCount}{0\xspace}
-\newcommand{\KaggleLocalCallCountRnd}{0\xspace}
-\newcommand{\KaggleLocalCallCount}{0\xspace}
-\newcommand{\KaggleLocalSiteCountRnd}{0\xspace}
-\newcommand{\KaggleLocalSiteCount}{0\xspace}
-\newcommand{\CoreAllCallCountRnd}{6.8M\xspace}
-\newcommand{\CoreAllCallCount}{6,776,674\xspace}
-\newcommand{\CoreAllSiteCountRnd}{21.6K\xspace}
-\newcommand{\CoreAllSiteCount}{21,578\xspace}
-\newcommand{\KaggleAllCallCountRnd}{0\xspace}
-\newcommand{\KaggleAllCallCount}{0\xspace}
-\newcommand{\KaggleAllSiteCountRnd}{0\xspace}
-\newcommand{\KaggleAllSiteCount}{0\xspace}
-\newcommand{\PackageAllCallCountRnd}{1.6M\xspace}
-\newcommand{\PackageAllCallCount}{1,598,198\xspace}
-\newcommand{\PackageAllSiteCountRnd}{1.7K\xspace}
-\newcommand{\PackageAllSiteCount}{1,724\xspace}
-\newcommand{\AllEvalCallCountRnd}{7.9M\xspace}
-\newcommand{\AllEvalCallCount}{7,865,167\xspace}
-\newcommand{\AllEvalSiteCountRnd}{1.6K\xspace}
-\newcommand{\AllEvalSiteCount}{1,644\xspace}
-\newcommand{\AllEvalParentCallCountRnd}{199.5K\xspace}
-\newcommand{\AllEvalParentCallCount}{199,505\xspace}
-\newcommand{\AllEvalParentSiteCountRnd}{262\xspace}
-\newcommand{\AllEvalParentSiteCount}{262\xspace}
-\newcommand{\AllEvalqCallCountRnd}{132.9K\xspace}
-\newcommand{\AllEvalqCallCount}{132,893\xspace}
-\newcommand{\AllEvalqSiteCountRnd}{20.5K\xspace}
-\newcommand{\AllEvalqSiteCount}{20,461\xspace}
-\newcommand{\AllLocalCallCountRnd}{177.3K\xspace}
-\newcommand{\AllLocalCallCount}{177,307\xspace}
-\newcommand{\AllLocalSiteCountRnd}{935\xspace}
-\newcommand{\AllLocalSiteCount}{935\xspace}
-\newcommand{\AllAllCallCountRnd}{8.4M\xspace}
-\newcommand{\AllAllCallCount}{8,374,872\xspace}
-\newcommand{\AllAllSiteCountRnd}{23.3K\xspace}
-\newcommand{\AllAllSiteCount}{23,302\xspace}
-\newcommand{\TotalFileCountRnd}{18.3K\xspace}
-\newcommand{\TotalFileCount}{18,258\xspace}
-\newcommand{\NoEvalFileCountRnd}{89\xspace}
-\newcommand{\NoEvalFileCount}{89\xspace}
-\newcommand{\NoEvalFilePerc}{0.5\%\xspace}
-\newcommand{\CoreEvalFileCountRnd}{12.3K\xspace}
-\newcommand{\CoreEvalFileCount}{12,307\xspace}
-\newcommand{\CoreEvalFilePerc}{67.4\%\xspace}
-\newcommand{\PackageEvalFileCountRnd}{0\xspace}
-\newcommand{\PackageEvalFileCount}{0\xspace}
-\newcommand{\PackageEvalFilePerc}{0\%\xspace}
-\newcommand{\AllEvalFileCountRnd}{5.9K\xspace}
-\newcommand{\AllEvalFileCount}{5,862\xspace}
-\newcommand{\AllEvalFilePerc}{32.1\%\xspace}
-\newcommand{\EightyCoreEvalFileCountRnd}{3.8K\xspace}
-\newcommand{\EightyCoreEvalFileCount}{3,769\xspace}
-\newcommand{\EightyCoreEvalFilePerc}{64.3\%\xspace}
-\newcommand{\TopTenPackageCallCountRnd}{1.4M\xspace}
-\newcommand{\TopTenPackageCallCount}{1,425,902\xspace}
-\newcommand{\TopTenPackageCallPerc}{89.2\%\xspace}
-\newcommand{\TopTenPackageSiteCountRnd}{282\xspace}
-\newcommand{\TopTenPackageSiteCount}{282\xspace}
-\newcommand{\TopTenPackageSitePerc}{16\%\xspace}
-\newcommand{\TopTenPackageNameA}{ggplot2\xspace}
-\newcommand{\TopTenPackageCallsiteCountARnd}{2\xspace}
-\newcommand{\TopTenPackageCallsiteCountA}{2\xspace}
-\newcommand{\TopTenPackageCallCountARnd}{987K\xspace}
-\newcommand{\TopTenPackageCallCountA}{987,946\xspace}
-\newcommand{\TopTenPackageCallPercA}{61\%\xspace}
-\newcommand{\TopTenPackageNameB}{magrittr\xspace}
-\newcommand{\TopTenPackageCallsiteCountBRnd}{5\xspace}
-\newcommand{\TopTenPackageCallsiteCountB}{5\xspace}
-\newcommand{\TopTenPackageCallCountBRnd}{139K\xspace}
-\newcommand{\TopTenPackageCallCountB}{139,346\xspace}
-\newcommand{\TopTenPackageCallPercB}{8.7\%\xspace}
-\newcommand{\TopTenPackageNameC}{plyr\xspace}
-\newcommand{\TopTenPackageCallsiteCountCRnd}{11\xspace}
-\newcommand{\TopTenPackageCallsiteCountC}{11\xspace}
-\newcommand{\TopTenPackageCallCountCRnd}{68K\xspace}
-\newcommand{\TopTenPackageCallCountC}{68,671\xspace}
-\newcommand{\TopTenPackageCallPercC}{4.3\%\xspace}
-\newcommand{\TopTenPackageNameD}{glue\xspace}
-\newcommand{\TopTenPackageCallsiteCountDRnd}{5\xspace}
-\newcommand{\TopTenPackageCallsiteCountD}{5\xspace}
-\newcommand{\TopTenPackageCallCountDRnd}{62K\xspace}
-\newcommand{\TopTenPackageCallCountD}{62,564\xspace}
-\newcommand{\TopTenPackageCallPercD}{3.9\%\xspace}
-\newcommand{\TopTenPackageNameE}{data.table\xspace}
-\newcommand{\TopTenPackageCallsiteCountERnd}{31\xspace}
-\newcommand{\TopTenPackageCallsiteCountE}{31\xspace}
-\newcommand{\TopTenPackageCallCountERnd}{47K\xspace}
-\newcommand{\TopTenPackageCallCountE}{47,464\xspace}
-\newcommand{\TopTenPackageCallPercE}{3\%\xspace}
-\newcommand{\TopTenPackageNameF}{statnet.common\xspace}
-\newcommand{\TopTenPackageCallsiteCountFRnd}{12\xspace}
-\newcommand{\TopTenPackageCallsiteCountF}{12\xspace}
-\newcommand{\TopTenPackageCallCountFRnd}{30K\xspace}
-\newcommand{\TopTenPackageCallCountF}{30,581\xspace}
-\newcommand{\TopTenPackageCallPercF}{1.9\%\xspace}
-\newcommand{\TopTenPackageNameG}{VGAM\xspace}
-\newcommand{\TopTenPackageCallsiteCountGRnd}{97\xspace}
-\newcommand{\TopTenPackageCallsiteCountG}{97\xspace}
-\newcommand{\TopTenPackageCallCountGRnd}{23K\xspace}
-\newcommand{\TopTenPackageCallCountG}{23,796\xspace}
-\newcommand{\TopTenPackageCallPercG}{1.5\%\xspace}
-\newcommand{\TopTenPackageNameH}{copula\xspace}
-\newcommand{\TopTenPackageCallsiteCountHRnd}{21\xspace}
-\newcommand{\TopTenPackageCallsiteCountH}{21\xspace}
-\newcommand{\TopTenPackageCallCountHRnd}{22K\xspace}
-\newcommand{\TopTenPackageCallCountH}{22,794\xspace}
-\newcommand{\TopTenPackageCallPercH}{1.4\%\xspace}
-\newcommand{\TopTenPackageNameI}{nlme\xspace}
-\newcommand{\TopTenPackageCallsiteCountIRnd}{97\xspace}
-\newcommand{\TopTenPackageCallsiteCountI}{97\xspace}
-\newcommand{\TopTenPackageCallCountIRnd}{21K\xspace}
-\newcommand{\TopTenPackageCallCountI}{21,483\xspace}
-\newcommand{\TopTenPackageCallPercI}{1.3\%\xspace}
-\newcommand{\TopTenPackageNameJ}{R6\xspace}
-\newcommand{\TopTenPackageCallsiteCountJRnd}{1\xspace}
-\newcommand{\TopTenPackageCallsiteCountJ}{1\xspace}
-\newcommand{\TopTenPackageCallCountJRnd}{21K\xspace}
-\newcommand{\TopTenPackageCallCountJ}{21,257\xspace}
-\newcommand{\TopTenPackageCallPercJ}{1.3\%\xspace}
-
-\input{tag/corpus.tex}
-%\input{tag/side-effects.tex}
-\newcommand{\AllWritesRnd}{9.4G\xspace}
-\newcommand{\AllWrites}{9,427,046,091\xspace}
-\newcommand{\EvalWritesRnd}{1.1G\xspace}
-\newcommand{\EvalWrites}{1,065,212,013\xspace}
-\newcommand{\EvalSideEffectingWritesRnd}{45M\xspace}
-\newcommand{\EvalSideEffectingWrites}{45,701,209\xspace}
-\newcommand{\EvalSideEffectingWritesEvalPerc}{4.3\%\xspace}
-\newcommand{\EvalSideEffectingWritesAllPerc}{.5\%\xspace}
-\newcommand{\SideEffectingCoreCallPerc}{4.4\%\xspace}
-\newcommand{\SideEffectingPackageCallPerc}{7.5\%\xspace}
-
-%\input{tag/operations.tex}
-\newcommand{\AllValueInputEvalCallPerc}{17\%\xspace}
-\newcommand{\AllExpressionInputEvalCallPerc}{82\%\xspace}
-\newcommand{\AllExpressionInputEvalSitePerc}{14\%\xspace}
-\newcommand{\AllValueInputEvalSitePerc}{3\%\xspace}
-\newcommand{\AllPolymorphicInputEvalSitePerc}{82\%\xspace}
-\newcommand{\PackageEnvironmentInputEvalCallPerc}{62\%\xspace}
-\newcommand{\AllClosureCallCountRnd}{2.9G\xspace}
-\newcommand{\AllClosureCallCount}{2,886,992,072\xspace}
-\newcommand{\EvalClosureCallCountRnd}{125M\xspace}
-\newcommand{\EvalClosureCallCount}{125,469,701\xspace}
-\newcommand{\AllNativeCallCountRnd}{463M\xspace}
-\newcommand{\AllNativeCallCount}{463,803,879\xspace}
-\newcommand{\EvalNativeCallCountRnd}{31M\xspace}
-\newcommand{\EvalNativeCallCount}{31,164,219\xspace}
-
-%\input{tag/events.tex}
-\newcommand{\AllEventCountRnd}{39G\xspace}
-\newcommand{\AllEventCount}{38,991,360,760\xspace}
-\newcommand{\EvalEventCountRnd}{2G\xspace}
-\newcommand{\EvalEventCount}{1,973,736,718\xspace}
-\newcommand{\EventsMaxCountRnd}{17M\xspace}
-\newcommand{\EventsMaxCount}{17,656,694\xspace}
-\newcommand{\EvalEventAllPerc}{5\%\xspace}
-\newcommand{\EventsMinRangeARnd}{0\xspace}
-\newcommand{\EventsMinRangeA}{0\xspace}
-\newcommand{\EventsMaxRangeARnd}{100\xspace}
-\newcommand{\EventsMaxRangeA}{100\xspace}
-\newcommand{\EventsCoreEvalCountARnd}{5M\xspace}
-\newcommand{\EventsCoreEvalCountA}{5,659,053\xspace}
-\newcommand{\EventsCoreEvalPercA}{88\%\xspace}
-\newcommand{\EventsPackageEvalCountARnd}{1.5M\xspace}
-\newcommand{\EventsPackageEvalCountA}{1,514,222\xspace}
-\newcommand{\EventsPackageEvalPercA}{93\%\xspace}
-\newcommand{\EventsMinRangeBRnd}{101\xspace}
-\newcommand{\EventsMinRangeB}{101\xspace}
-\newcommand{\EventsMaxRangeBRnd}{10K\xspace}
-\newcommand{\EventsMaxRangeB}{10,000\xspace}
-\newcommand{\EventsCoreEvalCountBRnd}{769K\xspace}
-\newcommand{\EventsCoreEvalCountB}{769,341\xspace}
-\newcommand{\EventsCoreEvalPercB}{12\%\xspace}
-\newcommand{\EventsPackageEvalCountBRnd}{100K\xspace}
-\newcommand{\EventsPackageEvalCountB}{100,141\xspace}
-\newcommand{\EventsPackageEvalPercB}{6\%\xspace}
-\newcommand{\EventsMinRangeCRnd}{10K\xspace}
-\newcommand{\EventsMinRangeC}{10,001\xspace}
-\newcommand{\EventsMaxRangeCRnd}{20M\xspace}
-\newcommand{\EventsMaxRangeC}{20,000,000\xspace}
-\newcommand{\EventsCoreEvalCountCRnd}{3K\xspace}
-\newcommand{\EventsCoreEvalCountC}{2,997\xspace}
-\newcommand{\EventsCoreEvalPercC}{0\%\xspace}
-\newcommand{\EventsPackageEvalCountCRnd}{14K\xspace}
-\newcommand{\EventsPackageEvalCountC}{14,209\xspace}
-\newcommand{\EventsPackageEvalPercC}{.9\%\xspace}
-
-%\input{tag/analysisStable.tex}
-\newcommand{\UniqueCallsRnd}{1.3M\xspace}
-\newcommand{\UniqueCalls}{1,277,948\xspace}
-\newcommand{\MatchArgNumberRnd}{5M\xspace}
-\newcommand{\MatchArgNumber}{5,029,351\xspace}
-\newcommand{\MatchArgPercent}{60\%\xspace}
-\newcommand{\NbAstOnePercent}{15\%\xspace}
-\newcommand{\NbAstOneCallSitePercent}{96\%\xspace}
-\newcommand{\AstSizeNineFiveRnd}{42\xspace}
-\newcommand{\AstSizeNineFive}{42\xspace}
-\newcommand{\AstSizeNineFiveExactPercent}{95\%\xspace}
-\newcommand{\DefaultEnvirExprRnd}{1.1M\xspace}
-\newcommand{\DefaultEnvirExpr}{1,054,124\xspace}
-\newcommand{\DefaultEnvirExprPercent}{12\%\xspace}
-\newcommand{\DefaultEnvirCallSitesPercent}{9.3\%\xspace}
-\newcommand{\DefaultEnvirCoreExprPercent}{10\%\xspace}
-\newcommand{\DefaultEnvirPackagesExprPercent}{20\%\xspace}
-\newcommand{\CoreEnvClassA}{2\xspace}
-\newcommand{\CoreCountARnd}{5M\xspace}
-\newcommand{\CoreCountA}{5,079,433\xspace}
-\newcommand{\CoreProportionA}{83\%\xspace}
-\newcommand{\CoreEnvClassB}{list\xspace}
-\newcommand{\CoreCountBRnd}{229K\xspace}
-\newcommand{\CoreCountB}{229,249\xspace}
-\newcommand{\CoreProportionB}{3.8\%\xspace}
-\newcommand{\CoreEnvClassC}{3\xspace}
-\newcommand{\CoreCountCRnd}{199K\xspace}
-\newcommand{\CoreCountC}{199,497\xspace}
-\newcommand{\CoreProportionC}{3.3\%\xspace}
-\newcommand{\CoreEnvClassD}{global+\xspace}
-\newcommand{\CoreCountDRnd}{183K\xspace}
-\newcommand{\CoreCountD}{183,631\xspace}
-\newcommand{\CoreProportionD}{3\%\xspace}
-\newcommand{\CoreEnvClassE}{1+\xspace}
-\newcommand{\CoreCountERnd}{175K\xspace}
-\newcommand{\CoreCountE}{175,151\xspace}
-\newcommand{\CoreProportionE}{2.9\%\xspace}
-\newcommand{\CoreEnvClassF}{global\xspace}
-\newcommand{\CoreCountFRnd}{124K\xspace}
-\newcommand{\CoreCountF}{124,673\xspace}
-\newcommand{\CoreProportionF}{2.1\%\xspace}
-\newcommand{\CoreEnvClassG}{sys.call(sys.parent())\xspace}
-\newcommand{\CoreCountGRnd}{32K\xspace}
-\newcommand{\CoreCountG}{32,349\xspace}
-\newcommand{\CoreProportionG}{.5\%\xspace}
-\newcommand{\CoreEnvClassH}{loop\xspace}
-\newcommand{\CoreCountHRnd}{24K\xspace}
-\newcommand{\CoreCountH}{24,286\xspace}
-\newcommand{\CoreProportionH}{.4\%\xspace}
-\newcommand{\PackagesEnvClassA}{global+\xspace}
-\newcommand{\PackagesCountARnd}{943K\xspace}
-\newcommand{\PackagesCountA}{943,682\xspace}
-\newcommand{\PackagesProportionA}{74\%\xspace}
-\newcommand{\PackagesEnvClassB}{3\xspace}
-\newcommand{\PackagesCountBRnd}{71K\xspace}
-\newcommand{\PackagesCountB}{71,037\xspace}
-\newcommand{\PackagesProportionB}{5.6\%\xspace}
-\newcommand{\PackagesEnvClassC}{2\xspace}
-\newcommand{\PackagesCountCRnd}{67K\xspace}
-\newcommand{\PackagesCountC}{67,555\xspace}
-\newcommand{\PackagesProportionC}{5.3\%\xspace}
-\newcommand{\PackagesEnvClassD}{5\xspace}
-\newcommand{\PackagesCountDRnd}{49K\xspace}
-\newcommand{\PackagesCountD}{49,782\xspace}
-\newcommand{\PackagesProportionD}{3.9\%\xspace}
-\newcommand{\PackagesEnvClassE}{list\xspace}
-\newcommand{\PackagesCountERnd}{44K\xspace}
-\newcommand{\PackagesCountE}{44,194\xspace}
-\newcommand{\PackagesProportionE}{3.5\%\xspace}
-\newcommand{\PackagesEnvClassF}{2+\xspace}
-\newcommand{\PackagesCountFRnd}{32K\xspace}
-\newcommand{\PackagesCountF}{32,931\xspace}
-\newcommand{\PackagesProportionF}{2.6\%\xspace}
-\newcommand{\PackagesEnvClassG}{global\xspace}
-\newcommand{\PackagesCountGRnd}{22K\xspace}
-\newcommand{\PackagesCountG}{22,825\xspace}
-\newcommand{\PackagesProportionG}{1.8\%\xspace}
-\newcommand{\PackagesEnvClassH}{base+\xspace}
-\newcommand{\PackagesCountHRnd}{18K\xspace}
-\newcommand{\PackagesCountH}{18,538\xspace}
-\newcommand{\PackagesProportionH}{1.5\%\xspace}
-\newcommand{\NbEnclosEnvirRnd}{274K\xspace}
-\newcommand{\NbEnclosEnvir}{274,370\xspace}
-\newcommand{\EnclosEnvirPercent}{3.3\%\xspace}
-\newcommand{\NbSysCallEnvirRnd}{33K\xspace}
-\newcommand{\NbSysCallEnvir}{33,650\xspace}
-\newcommand{\SysCallEnvirPercent}{.4\%\xspace}
-\newcommand{\NbMonomorphicRnd}{23.1K\xspace}
-\newcommand{\NbMonomorphic}{23,069\xspace}
-\newcommand{\PercentMonomorphic}{99\%\xspace}
-\newcommand{\PercentParsedEvals}{.9\%\xspace}
-\newcommand{\PercentParsedCallSites}{1\%\xspace}
-\newcommand{\NbParseFilesRnd}{10\xspace}
-\newcommand{\NbParseFiles}{10\xspace}
-\newcommand{\PercentParseFileCallSites}{.01\%\xspace}
-\newcommand{\PasteParsedEvals}{39\%\xspace}
-\newcommand{\PasteParsedCallSite}{51\%\xspace}
-\newcommand{\NbCoreEvalCallSitesRnd}{310\xspace}
-\newcommand{\NbCoreEvalCallSites}{310\xspace}
-\newcommand{\MaxCallSitePackageRnd}{2.4K\xspace}
-\newcommand{\MaxCallSitePackage}{2,372\xspace}
-\newcommand{\OneCallSitePackageRnd}{48\xspace}
-\newcommand{\OneCallSitePackage}{48\xspace}
-\newcommand{\NoCallSitePackagesRnd}{233\xspace}
-\newcommand{\NoCallSitePackages}{233\xspace}
-
-%\input{tag/enrichDataset.tex}
-\newcommand{\UndefinedEvalsRnd}{285K\xspace}
-\newcommand{\UndefinedEvals}{285,289\xspace}
-\newcommand{\PercentUndefinedEval}{3.3\%\xspace}
-\newcommand{\UndefinedEvalExpressionsRnd}{262\xspace}
-\newcommand{\UndefinedEvalExpressions}{262\xspace}
-\newcommand{\UndefinedOutOfCorpusRnd}{0\xspace}
-\newcommand{\UndefinedOutOfCorpus}{0\xspace}
-\newcommand{\UndefinedOutOfCoprusPercent}{0\%\xspace}
-\newcommand{\MaxCallSitesEnrich}{114\xspace}
-\newcommand{\OneCallSitesEnrich}{61\xspace}
-\newcommand{\NoCallSitesEnrich}{297\xspace}
-%\input{tag/patterns.tex}
-\newcommand{\PatternLibRnd}{4K\xspace}
-\newcommand{\PatternLib}{4,278\xspace}
-\newcommand{\PatternLibRatio}{.05\%\xspace}
-\newcommand{\PatternLibSourceRnd}{370\xspace}
-\newcommand{\PatternLibSource}{370\xspace}
-\newcommand{\PatternLibLocalRnd}{100\xspace}
-\newcommand{\PatternLibLocal}{100\xspace}
-\newcommand{\PatternReadRnd}{59K\xspace}
-\newcommand{\PatternRead}{59,776\xspace}
-\newcommand{\PatternReadRatio}{.7\%\xspace}
-\newcommand{\PatternReadSlotAccessRatio}{88\%\xspace}
-\newcommand{\PatternReadDatatableRatio}{27\%\xspace}
-\newcommand{\PatternAssignRnd}{68K\xspace}
-\newcommand{\PatternAssign}{68,754\xspace}
-\newcommand{\PatternAssignRatio}{.9\%\xspace}
-\newcommand{\PatternAssignArrowRatio}{18\%\xspace}
-\newcommand{\PatternAssignArrowMagrittrRatio}{59\%\xspace}
-\newcommand{\PatternAssignSuperarrowRatio}{81\%\xspace}
-\newcommand{\PatternAssignSuperarrowPlyrRatio}{99\%\xspace}
-\newcommand{\PatternManualPackagesRnd}{20\xspace}
-\newcommand{\PatternManualPackages}{20\xspace}
-\newcommand{\PatternManualCallsitesRnd}{161\xspace}
-\newcommand{\PatternManualCallsites}{161\xspace}
-
-\graphicspath{{img/}}
-
-\lstset{language=R}
-
-\definecolor{LightGray}{rgb}{.92,.92,.92}
-\definecolor{Gray}{rgb}{.3,.3,.3}
-\definecolor{DarkGray}{rgb}{.5,.5,.5}
-
-\lstset{ %
-  columns=flexible,
-  captionpos=b,
-  frame=single,
-  framerule=0pt,
-  framexleftmargin=-1mm,
-  framexrightmargin=-1mm,
-  tabsize=2,
-  belowskip=0pt,
-  basicstyle=\small\ttfamily,
-  backgroundcolor=\color{LightGray},
-  emphstyle=\sffamily,
-  keywordstyle=\bfseries,
-  commentstyle=\color{Gray}\em,
-  stringstyle=\color{Gray}
-}
-\lstdefinestyle{R}{ %
-  language=R,
-  deletekeywords={new, env, equal, c, runif, trace, args, exp, t, all, get,
-    names, is, environment, class, substitute, expression, list, null, Internal,
-    sample, diag, length, rep, nrow, stop, offset, pmax, Machine,
-    double, parent, frame, par, methods, end, dir, apply, deparse, missing,
-    plot, as, integer, character, inherits, numeric, paste, eval, quote, call,
-    formula, df, log, sum, c, local, legend, file, scale, round, title, order,
-    drop, which, grid, print, ncol, dim, max, format, sort, rug, matrix, start,
-    unique, mean, df, attr, do, power, \$},
-  otherkeywords={},
-  breaklines=true
-}
-
-\newcommand{\eg}{\emph{e.g.},\xspace}
-\newcommand{\ie}{\emph{i.e.},\xspace}
-\newcommand{\cf}{\emph{cf.}\xspace}
-\newcommand{\summary}[1]{{\csname #1\endcsname} ({\csname #1Min\endcsname} / {\csname #1Mean\endcsname} / {\csname #1Max\endcsname})}
-\newcommand{\summaryrnd}[1]{{\csname #1Rnd\endcsname} ({\csname #1MinRnd\endcsname} / {\csname #1MeanRnd\endcsname} / {\csname #1MaxRnd\endcsname})}
-
-\newcommand{\eval}{\texttt{eval}\xspace}
-\newcommand{\evals}{{\sf eval}s\xspace}
-\renewcommand{\parse}{\c{parse}}
-\newcommand{\source}{\c{source}}
-\newcommand{\local}{\c{local}}
-\newcommand{\unlockBinding}{\c{unlockBinding}}
-\newcommand{\substitute}{\c{substitute}}
-\newcommand{\datatable}{\c{data.table}}
-\newcommand{\mlogit}{\c{mlogit}}
-\newcommand{\mboost}{\c{mboost}}
-\newcommand{\metafor}{\c{metafor}}
-\newcommand{\lavaan}{\c{lavaan}}
-\newcommand{\mclust}{\c{mclust}}
-\newcommand{\gamlss}{\c{gamlss}}
-\newcommand{\ggproto}{\c{ggproto}}
-\newcommand{\ggplot}{\c{ggplot2}}
-\newcommand{\base}{\c{base}}
-\renewcommand{\c}[1]{\lstinline{#1}\xspace}
-\newcommand{\miss}[1]{{\textcolor{red}{#1}}\xspace}
-\newcommand{\evil}{\emph{evil}\xspace}
-\newcommand{\instrumentr}{\emph{instrumentr}\xspace}
-\newcommand{\rdyntrace}{\emph{R-dyntrace}\xspace}
-\newcommand{\covr}{\emph{covr}\xspace}
-\newcommand{\runr}{\emph{runr}\xspace}
-
-\bibliographystyle{plainurl}% the mandatory bibstyle
-
-\title{Why do we Eval?\\[2mm]\Large
-A large-scale study of Eval usage in R}
+\usepackage{listings,hyperref,multirow,paralist,xspace,url}
+\input{macro}
+
+\title{Why do we Eval?\\[2mm]\Large A large-scale study of Eval usage in R}
 \titlerunning{Eval in R}
 
-\author{Anon}{Under review}{}{}{}
-\authorrunning{Anon}
-\Copyright{Anon}
-\keywords{R, reflection, eval}
-
-%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%\
-%%%%%%%                                                                         
-\EventEditors{Eds.}
-\EventNoEds{2}
-\EventLongTitle{European Conference on Object Oriented Programming (ECOOP 2021)}
-\EventShortTitle{ECOOP 2021}
-\EventAcronym{ECOOP}
-\EventYear{2021}
-\EventDate{December 24--27, 2016}
-\EventLocation{Little Whinging, United Kingdom}
-\EventLogo{}
-\SeriesVolume{42}
-\ArticleNo{23}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{CCSXML}
-  <ccs2012>
-  <concept>
-  <concept_id>10011007.10011006.10011050.10010517</concept_id>
-  <concept_desc>Software and its engineering~Scripting languages</concept_desc>
-  <concept_significance>500</concept_significance>
-  </concept>
-  <concept>
-  <concept_id>10002944.10011123.10010912</concept_id>
-  <concept_desc>General and reference~Empirical studies</concept_desc>
-  <concept_significance>500</concept_significance>
-  </concept>
-  <concept>
-  <concept_id>10011007.10011006.10011008</concept_id>
-  <concept_desc>Software and its engineering~General programming languages</concept_desc>
-  <concept_significance>500</concept_significance>
-  </concept>
-  </ccs2012>
-\end{CCSXML}
-
-\ccsdesc[500]{Software and its engineering~Scripting languages}
-\ccsdesc[500]{General and reference~Empirical studies}
-\ccsdesc[500]{Software and its engineering~General programming languages}
-
-\begin{document}
-
-\maketitle
+\begin{document}\maketitle
 
@@
   end-user code on which we performed dynamic analysis to confirm that
-  \eval is indeed in widespread use, and categorized patterns of
-  usage. We found that \eval is more dangerous in some ways, and safer
-  in others, than what was previously reported for JavaScript.
+  \eval is indeed in widespread use. We found that \eval is more
+  dangerous in some ways, and safer in others, than what was
+  previously reported for JavaScript.
 \end{abstract}
@@
 
-Traditional program analysis techniques are based on sound reasoning about
-program behavior, a program analyzer computes an over-approximation of the
-set of possible operations performed by the system being
-analyzed~\cite{cc77}.  This is predicated on knowing the program's code, and
-computing that code's behavior for all possible inputs. The presence of
-\eval entails that the program's input, any string variable, may be
-reflected into its code giving rise to \emph{any} behavior allowed by the
-language semantics. In many dynamic languages, \eval can redefine most
-user-defined and built-in functions resulting in a complete loss of
-precision for any operation following \eval. A group of influential
-researchers argued to give up on soundness and, instead, to
-under-approximate dynamic features~\cite{soundy} In their words ``a
-practical analysis, therefore, may pretend that \eval does nothing, unless
-it can precisely resolve its string argument at compile time.''  Assuming
-that \eval does not have side-effects may be too optimistic. We rather
-propose to study what \eval does in practice, in real-world code, so as to
-give researchers data about how to better under-approximate uses of that
-feature.
+Most dynamic languages provide their users with a facility to
+transform unstructured text into executable code, and to evaluate that
+code. In this paper we refer to this reflective facility of a language
+as \eval bowing to its origins in LISP in 1956~\cite{lisp}. \Eval has
+been much maligned over the years. In computing lore, it is as close
+to a boogeyman as it gets. Yet, for McCarthy, \eval was simply the way
+to write down the definition of LISP, he was surprised that someone
+coded it up and offered it end users.  Since then, reflective
+facilities have been used to parameterize programs over code patterns
+that can be provided after the program is written, when it
+executes. The presence of such a feature in a language is a hallmark
+of dynamism as it is a form of delayed binding.
+
+\vspace{2mm}\noindent\emph{Trouble in Paradise.} Reflective facilities
+hinder most attempt to reason about the code using them, as well as to
+apply meaning-preserving program transformations. In practice this
+means that static analysis techniques will loose so much precision as
+to become pointless and that anything but the most trivial, local,
+compiler optimizations will be unsound. Furthermore the addition of
+new code --code that could have been obtained from a network
+connection-- as the program is running is a security vulnerability
+waiting to happen.
+
+To further illustrate the challenges at hand, consider the interaction
+between a static analysis tool and a dynamic language. An abstract
+interpretation-based program analyzer computes an over-approximation
+of the set of possible behaviors exhibited by the system under
+study~\cite{cc77}. The presence of a reflective operation entails that
+the operation may have any behavior that can be expressed in the
+language; as if we replaced it by any legal sequence of instructions.
+As dynamic languages tend to be permissive, the analysis has to assume
+that many (and sometimes all) functions in scope may have been
+redefined, e.g. that \texttt{`+`} now opens an Internet connection or
+something equally surprising. This means that a single occurrence of
+\eval will cause the static analyzer to loose all information about
+the program state and the meaning of identifiers. This loss of
+precision can sometimes be mitigated by analyzing the string passed to
+\eval~\cite{} to bound its possible behavior but when the string comes
+from outside the program not much can be done.  A group researchers
+argued giving up on soundness and, instead, under-approximate dynamic
+features~\cite{soundy} In their words ``a practical analysis,
+therefore, may pretend that \eval does nothing, unless it can
+precisely resolve its string argument at compile time.''  Alas
+assuming that \eval does not have side-effects may be unduly
+optimistic.
 

### 2020-11-26 inst/paper/main.tex
intro

@@
 study~\cite{cc77}. The presence of a reflective operation entails that
-the operation may have any behavior that can be expressed in the
-language; as if we replaced it by any legal sequence of instructions.
-As dynamic languages tend to be permissive, the analysis has to assume
-that many (and sometimes all) functions in scope may have been
-redefined, e.g. that \texttt{`+`} now opens an Internet connection or
-something equally surprising. This means that a single occurrence of
-\eval will cause the static analyzer to loose all information about
-the program state and the meaning of identifiers. This loss of
-precision can sometimes be mitigated by analyzing the string passed to
-\eval~\cite{} to bound its possible behavior but when the string comes
-from outside the program not much can be done.  A group researchers
-argued giving up on soundness and, instead, under-approximate dynamic
-features~\cite{soundy} In their words ``a practical analysis,
+the operation may have \emph{any} behavior that can be expressed in
+the target language; i.e. as if we replaced it by any legal sequence
+of instructions.  As dynamic languages tend to be permissive, the
+analysis has to, for example, assume that many (and sometimes all)
+functions in scope may have been redefined, e.g. that \texttt{`+`} now
+opens an Internet connection or something equally surprising. This
+means that a single occurrence of \eval causes the static analyzer to
+loose all information about the program state and the meaning of
+identifiers. This loss of precision can sometimes be mitigated by
+analyzing the string passed to \eval~\cite{} to bound its possible
+behavior but when the string comes from outside the program not much
+can be done.  A frustrated group researchers argued giving up on
+soundness and, instead, under-approximating dynamic
+features (soundiness)~\cite{soundy}. In their words ``a practical analysis,
 therefore, may pretend that \eval does nothing, unless it can
-precisely resolve its string argument at compile time.''  Alas
-assuming that \eval does not have side-effects may be unduly
-optimistic.
-
-While there has been previous work investigating the usage of \eval in web
-programming and JavaScript~\cite{ecoop11}, there are no other studies that
-shed light on the usage of \eval in other programming languages and
+precisely resolve its string argument at compile time.''  Alas,
+assuming that \eval does not have side-effects, or that side-effects
+will not affect the results of the analysis, may be unduly optimistic.
+
+\vspace{2mm}\noindent\emph{Is Past Prologue?} Previous work
+investigated how \eval is used in web programming, specifically in web
+sites that use JavaScript~\cite{pldi10a}. In 2010, 17 of the largest
+web site used the feature. In 2011, 82\% of the 10,000 most accessed
+sites used \eval~\cite{ecoop11}.  Yet, the strings passed to \eval,
+and their behaviors when executed, are far from random; it was shown
+that when one can observe several calls to \eval, the ``shape'' of
+future calls can be predicted with 97\% accuracy~\cite{oopsla12b}.
+Overall, practical usage suggested that most reflective calls were
+relatively harmless. While this backs up the soundiness squad's
+approach, does it generalize to other application domains than web
+programming and to other languages?
+
 application domains.  It is thus reasonable to wonder if the results from

### 2020-11-26 inst/paper/main.tex
intro

@@
 
-application domains.  It is thus reasonable to wonder if the results from
-previous work can be carried over to other contexts. For this paper, we have
-picked the data science language R~\cite{r} as our object of study. R is
-both different from JavaScript in linguistic terms and in terms of its user
-base. JavaScript was designed to run untrusted code in browser, while R was
-designed for statistical computing on a desktop. JavaScript is used as a
-general purpose language by a wide community of programmers; while R is used
-for scientific computing by data scientists and domain experts with, often,
-limited programming experience.
-
-This paper sets out to highlight the differences in usage patterns of \eval
-between these two languages and user communities. Our results are broadly
-applicable as they increase our understanding of the range of uses of \eval,
-present new use-cases, and, hopefully, provide insights into reasonable
-under-approximations. More narrowly, for researchers interested in static
-analysis of data science code, we provide a publicly available corpus of
-programs and a toolchain that can be customized to deliver fine-grained
-information about the usage and impact of \eval.
+\vspace{2mm}\noindent\emph{The here and now.}  In this study, we
+investigate the usage of \eval in programs written in the R
+programming language. R is language designed by statisticians for
+applications in data science~\cite{r}. What makes looking at R after
+JavaScript interesting is that, while both languages are dynamic, they
+are quite different. While one can program in an object-oriented style
+in R, like in JavaScript, R is primarily a lazy, untyped, functional
+language.  JavaScript was designed to run untrusted code in browser,
+while R is used for statistical computing on desktops. JavaScript is a
+general purpose language used by a wide community of programmers;
+while R is used for scientific computing by data scientists and domain
+experts with, often, limited programming experience. One can
+distinguish between library implementers, developers with some
+programming experience and a working knowledge of R, and end-users,
+who are typically not expert programmers and often have only a cursory
+knowledge of the language.\footnote{Consider that R evaluates function
+argument lazily, just like Haskell. We informally surveyed end-users,
+including computer scientists, and did not find a single user aware of
+this fact. Library developers, on the other hand, know about laziness
+and program defensively around it.} Our goal is thus to highlight the
+differences in usage between JavaScript and R, and try to explain
+those differences in terms of language features, application domain
+and programmer experience.
+
+\vspace{2mm}\noindent\emph{Our contributions.} 
+TODO
 
@@
 Our corpus has been constructed to reflect the different level of
-sophistication in the R community. We distinguish between the core
-implementers of the language, developers with both extensive programming
-experience and keen understanding of the language semantics, library
-implementers, developers with some programming experience and a working
-knowledge of R, and end-users, who are typically not expert programmers and
-often have only a cursory knowledge of the language.\footnote{To illustrate
-  the above consider that R uses lazy evaluation like Haskell. The authors
-  have informally surveyed end-users, including computer scientists, and did
-  not find a single one aware of this fact. Library developers, on the other
-  hand, know about laziness and program defensively around it.}  Thus, our
+sophistication in the R community.
+
+Thus, our
 corpus distinguished between \emph{core packages} (there are

### 2020-11-26 inst/analysis/preprocess.R
Porting enrich.Rmd to preprocess.R

@@ -10,6 +10,7 @@
 #   - summarized-packages.fst
 
 source("insights.R")
+source("inc/paths.R")
 
 library(fs)
 
@@ -219,3 +220,139 @@ get_externals <- function(dataset, corpus_files)
   return(dataset %>%
            filter(!eval_source %in% c(corpus_files, "core", "base", "base?")))
 }
+
+undefined_packages <- function(dataset)
+{
+  undefined_evals <- eval_calls %>% filter(eval_source_type == "<undefined>")
+  undefined_per_package <- undefined_evals %>% group_by(package) %>% summarize(n = n_distinct(eval_call_expression))
+  known_packages <- setdiff(eval_calls$package, undefined_per_package$package) %>% as_tibble_col(column_name = "package")
+  known_packages <- known_packages %>% mutate(n = 0)
+
+  undefined_per_package <- bind_rows(undefined_per_package, known_packages) %>% arrange(desc(n))
+
+  return(undefined_per_package)
+}
+
+
+known_call_sites <- function(eval_calls_corpus, corpus_file)
+{
+  call_sites_per_package <- eval_calls_corpus %>%
+    filter(eval_source_type == "package", eval_source %in% corpus_files) %>%
+    group_by(eval_source) %>%
+    summarize(n = n_distinct(eval_call_srcref))
+
+  known_packages <- setdiff(corpus_files, call_sites_per_package$eval_source) %>% as_tibble_col(column_name = "eval_source")
+
+  known_packages <- known_packages %>% mutate(n = 0)
+
+  call_sites_per_package <- bind_rows(call_sites_per_package, known_packages) %>%
+    rename(package = eval_source) %>%
+    arrange(desc(n))
+
+  return(call_sites_per_package)
+}
+
+### Command line and preprocessing pipeline
+
+usage <- function()
+{
+  print("preprocess corpus_file calls_file kaggle_calls_file package_evals_dynamic_file
+        evals_undefined_file evals_raw_file evals_summarized_core_file evals_summarized_pkgs_file
+        evals_summarized_externals_file.
+
+        The command has 10 arguments. The first 3 ones are input files, the last 7 ones are output files.
+        ")
+}
+
+main <- function(argv) {
+
+  if(length(argv) != 10)
+  {
+    usage()
+  }
+
+  corpus_file <- argv[1]
+  calls_file <- argv[2]
[…]

### 2020-11-26 inst/paper/main.tex

@@
 facilities have been used to parameterize programs over code patterns
-that can be provided after the program is written, when it
-executes. The presence of such a feature in a language is a hallmark
-of dynamism as it is a form of delayed binding.
+that can be provided after the program is written.  The presence of
+such a feature in a language is a hallmark of dynamism as it is a form
+of delayed binding as what any particular call to \eval will do is
+only know at program run-time when that particular call site is
+evaluated.
 
@@
 
-\vspace{2mm}\noindent\emph{Our contributions.} 
-TODO
-
-Our methodology for this work bows to the fact that static analysis of R is
-difficult, instead we focus on dynamic analysis. To observe \eval we have
-built a two-level monitoring infrastructure: we are able to monitor R
-programs by instrumentation -- this gives us access to many user-visible
-properties of R programs -- and we also can monitor the inner working of the
-R interpreter -- this allows us to capture details of execution that are not
-exposed at the source level.
-
-Our corpus has been constructed to reflect the different level of
-sophistication in the R community.
-
-Thus, our
-corpus distinguished between \emph{core packages} (there are
-\CorpusCorePackages such packages that are distributed together with the R
-virtual machine), \emph{CRAN packages} (we selected \CorpusPackages curated
-packages that pass stringent community quality checks and are equipped with
-tests and sample data), \emph{Kaggle scripts} (we used \CorpusFinishedKaggle
+\vspace{2mm}\noindent\emph{The What and How.}  One significant benefit
+of choosing R is that every package in the CRAN repository is curated
+and comes with examples of typical usage. This gives us a large code
+base that we can analyze dynamically.  To observe \eval we built a
+two-level monitoring infrastructure: we can monitor R programs by
+instrumentation --this gives us access to many user-visible properties
+of R programs-- but we can also monitor the inner working of the R
+interpreter --this allows us to capture details of execution that are
+not exposed at the source level. Dynamic analysis is of course limited
+in that it can only observe behaviors that are triggered by the
+particular inputs passed to the program. Luckily, CRAN libraries come
+with multiple input datasets and use-cases.
+
+The choice of corpus is crucial for analyses like ours.  Our corpus
+has been constructed to reflect the levels of sophistication of the R
+community.  We distinguish between \emph{core packages} (there are
+\CorpusCorePackages such packages that are distributed together with
+the R virtual machine), \emph{CRAN packages} (\CorpusPackages curated
+packages that pass stringent quality checks and are equipped with
+tests and sample data), \emph{Kaggle scripts} (\CorpusFinishedKaggle
 end-user written programs that performs a particular data analysis
-task). There are reasons to believe that \eval usage patterns may differ
-between these datasets. The core R code is extremely stable, it has only
-been maintained for the last 20 years with very few far reaching
-changes. The libraries represent a more lively ecosystem with new libraries
-added each day.  Finally, end user code is often one-use scripts thrown
-together and sometimes never revisited.
-
-To understand the behavior of \eval we perform a dynamic analysis and manual
-code inspection. The results presented here summarize our findings.
-
-Our infrastructure is publicly available, released in open source, and our
-code, and data will be submitted for artifact evaluation.
+task). It is reasonable to expect that \eval usage differ between
+these datasets. The core R code is extremely stable, maintained for
+the last 20 years with very few far reaching changes. The libraries
+represent a more lively ecosystem with new libraries added each day.
+Finally, end user code is often one-use scripts thrown together and
+sometimes never revisited.
+
+Our infrastructure is publicly available, released in open source, and
+our code, and data will be submitted for artifact evaluation.
 

### 2020-11-26 inst/analysis/enrich_dataset.Rmd
Polishing preprocessing

@@
 r("undefined out of corpus", nb_out_corpus$n)
-r("undefined out of coprus percent", ratio(nb_out_corpus$n, nb_undefined$n))
+r("undefined out of corpus percent", ratio(nb_out_corpus$n, nb_undefined$n))
 ```

### 2020-11-26 inst/analysis/insights.R

@@ -1,4 +1,4 @@
-library(rlang)
+suppressPackageStartupMessages(library(rlang))
 
 
 
@@ -6,10 +6,10 @@ library(rlang)
 nb_eval_call_sites <- function(eval_calls)
 {
   # Two cases: with srcref, without srcref
-  
+
   nb_with_srcref <- eval_calls %>% select(eval_call_srcref) %>% n_distinct(na.rm = TRUE)
   nb_without_srcref <- eval_calls %>% filter(is.na(eval_call_srcref)) %>% select(caller_package, caller_function) %>% n_distinct(na.rm = TRUE)
-  
+
   return(nb_with_srcref + nb_without_srcref)
 }
 
@@ -29,16 +29,16 @@ get_expr <- function(eval_call)
   {
     eval_call
   }
-  
+
   exp <- NA
-  # Would fail for instance for 
+  # Would fail for instance for
   # "`$<-`(new(\"C++Field\", .xData = <environment>), \"read_only\", TRUE)" (classInt package)
   # exp <- tryCatch(
   #   parse(text = escaped_eval_call)[[1]],
   #   error = function(e) {return(NA)})
   try(exp <- parse(text = escaped_eval_call)[[1]], silent = TRUE)
 
-  
+
   return(exp)
 }
 
@@ -68,17 +68,17 @@ parse_only <- function(code) {}
 .myparse <- function(text) {}
 parse_all <- function(x, filename,allow_error) {}
 
-extract_args_parse <- function(eval_call) 
+extract_args_parse <- function(eval_call)
 {
   exp <- get_expr(eval_call)
   exp <- tryCatch(call_standardise(exp),
-                  error = function(c) 
+                  error = function(c)
                     if(length(exp) >= 2 && exp[[2]] == "...") { names(exp)[[2]] <- "dots" ; return(exp)}
                     else stop(paste0("extract_arg failed with: ", eval_call))
   )
   args <- map_chr(as.list(exp[-1]), function(chr) { paste(deparse(chr), collapse = "\n")})
   names(args) <- map_chr(names(args), function(chr) paste0("parse_args_", chr))
-  
+
   return(args[str_detect(names(args), "^parse_args_(file|text|n|s|keep\\.source|srcfile|dots)$")])# "file|text|n|s|prompt|keep.source|srcfile|code"
 }
 
@@ -236,14 +236,14 @@ package_name_from_call_stack <- function(caller_stack_expr, caller_stack_expr_sr
 {
[…]

### 2020-11-26 inst/analysis/preprocess.R

@@ -1,18 +1,20 @@
+#!/usr/bin/env Rscript
+
 # Preprocess the dataset:
 # - remove duplicates
+# - correct src refs
 # - add some columns (such as `eval_call_package`)
 # - cut the dataset into smaller ones
 
-# Datasets
-# Input: raw.fst
-# Outputs:
-#   - summarized-core.fst
-#   - summarized-packages.fst
 
-source("insights.R")
-source("inc/paths.R")
-
+suppressPackageStartupMessages(library(tidyverse))
 library(fs)
+library(fst)
+
+
+
+# Rather relative source because here it depends on where we launch the script...
+source("revalstudy/inst/analysis/insights.R")
 
 
 deduplicate <- function(dataset)
@@ -256,19 +258,24 @@ known_call_sites <- function(eval_calls_corpus, corpus_file)
 
 usage <- function()
 {
-  print("preprocess corpus_file calls_file kaggle_calls_file package_evals_dynamic_file
-        evals_undefined_file evals_raw_file evals_summarized_core_file evals_summarized_pkgs_file
-        evals_summarized_externals_file.
+  cat("USAGE:
+./preprocess <corpus_file> <calls_file> <kaggle_calls_file> <package_evals_dynamic_file> <evals_undefined_file> <evals_raw_file> <evals_summarized_core_file> <evals_summarized_pkgs_file> <evals_summarized_kaggle_file> <evals_summarized_externals_file>.
 
-        The command has 10 arguments. The first 3 ones are input files, the last 7 ones are output files.
-        ")
+The command has 10 arguments. The first 3 ones are input files, the last 7 ones are output files. All files are assumed to be fst files.
+
+Example:
+./preprocess revalstudy/inst/data/corpus.fst run/package-evals-traced.4/calls.fst run/kaggle-run/calls.fst revalstudy/inst/data/evals-dynamic.fst run/package-evals-traced.4/summarized-evals-undefined.fst run/package-evals-traced.4/raws.fst run/package-evals-traced.4/summarized-core.fst run/package-evals-traced.4/summarized-packages.fst run/package-evals-traced.4/summarized-kaggle.fst run/package-evals-traced.4/summarized-externals.fst
+\n")
 }
 
 main <- function(argv) {
 
+  argv <- argv[-1] # Remove the command name
   if(length(argv) != 10)
   {
[…]

### 2020-11-26 inst/paper/main.tex

@@
 to write down the definition of LISP, he was surprised that someone
-coded it up and offered it end users.  Since then, reflective
+coded it up and offered it to end users.  Since then, reflective
 facilities have been used to parameterize programs over code patterns
@@
 programming language. R is language designed by statisticians for
-applications in data science~\cite{r}. What makes looking at R after
+applications in data science~\cite{r,R96}. What makes looking at R after
 JavaScript interesting is that, while both languages are dynamic, they
@@
 
-\section{Background and Related Work}
-
-The origins of \eval go back to the first implementation of the Lisp
-programming language~\cite{lisp}. Over the years, \eval has been included in
-many dynamic languages, with some variability in expressive power and
-nomenclature. In all of its incarnation \eval allows developers to
-dynamically extend the code of their program at run-time.  Originally, \eval
-was limited to interpreted languages, but with the advent of just-in-time
-compilation this restriction was gradually lifted.
-
-Some languages have chosen to restrict \eval. The essential difference
-between languages relates to scoping of variables that can be accessed from
-code executed by \eval.  Java and Julia are among the more restrictive
-languages. In Java, \eval must be synthesized as a combination of dynamic
-loading and reflection.\footnote{While Java does not have \eval built-in,
-  one can take the input to \eval, wrap it in a static method of an
-  anonymous class, generate bytecode for that class, invoke the Java class
-  loader to dynamically install the new code, and then use reflection to
-  call the new method.} Julia provides \eval more directly but limits its
-scope to top level variables and methods. JavaScript supports both a,
-so-called, global \eval, and a local \eval, where the latter has acces to
-local variables of the function being executed (as well as variables of
-lexically enclosing scopes). Finally, languages such as R allow \eval to
-even manipulate variables defined in the caller environment of the function in
-which \eval is invoked.
-
-The difference between these design choices has an impact on the ability of
-compilers to optimize, and of analysis tools to reason about code. The
-impact of a top level \eval is that any global variable may be modified, but
-currently running functions can continue to execute unchanged. A local
-\eval, may change the values of local variables, their types, or even their
-presence. This is a much more invasive change, one that may invalidate
-currently executing functions as it may change the results of common
-optimizations such as constant folding or copy propagation. The designers of
-the Julia programming language were aware of this difference and
-purposefully chose to restrict \eval to the top level in order to preserve
-their ability to optimize code~\cite{oopsla18a}. They even added a
-versioning mechanism to function definitions to ensure that code added
-during an \eval would not invalidate optimizations such as inlining.  R's
-ability to affect any one of the callers of a function is even more invasive
-as the impact of \eval is non-local. Such unfettered power is difficult to
-deal with.
-
-The intuition behind the soundiness manifesto~\cite{soundy} is that while
-the use of a single \eval may invalidate every single invariant computed by
-a static analyzer, this worst case usually does not happen.  In many cases,
-\eval is used carefully to limited effect. This hypothesis must be validated
-and, for it to be useful, we need to have a way to make some educated
-guesses as to which side-effects will result from any particular \eval.
-
-Richards et al.~\cite{ecoop11} provided the first large-scale study of the
-runtime behavior of \eval in JavaScript. They analyzed a corpus of the
-10,000 most popular websites. They used an instrumented web browser to
-gather execution traces and performed a dynamic analysis.  Their results show
-that \eval is pervasive with 82\% of the most popular websites using it. The
-reasons for using \eval include the desire to load code on demand (to reduce
-the time it takes to display a page), the deserialization of JSON data
-(about 37\% of all calls) and lightweight meta-programming to customize web
-pages.  While many uses \eval were legitimate, many were unnecessary and
-could be replaced with equivalent and safer code.  They reported on a
-categorization of inputs to \eval that covers the vast majority of input
-strings.  Restricting themselves to \eval in which all named variables refer
-to the global scope, many patterns could be replaced by more disciplined
-code~\cite{oopsla12b}.  The limitation of that work are that there was no
-measure of code coverage, so the numbers presented can be a lower bound on
-the possible behaviors.  Furthermore in 2011, JavaScript usage was quite
-different from today. Their work captured very few libraries and none of the
-more recent Node.js ecosystem.  More details about dynamic analysis of
-JavaScript can be found in~\cite{liang}.
-
-Wang et al.~\cite{wang} analyzed usage of dynamic features in 18 Python
-programs include dynamic code generation to find if they affect file
-change-proneness.  They reported that files with dynamic features are
-significantly more likely to be the subject of changes, than other files.
-Chen et al. looked at the correlation between code changes and dynamic
-features, including \eval, in 17 Python programs~\cite{chen}. They did not
-observe many uses of \eval.  Callau et al.~\cite{oscar} performed an empirical study of
-the usage of dynamic features in the 1,000 largest Smalltalk projects in the
-Squeaksource repository, accounting for more than 4 million lines of
-code~\cite{oscar}. While \eval itself is not present, Smalltalk has a rich
-reflective interface that gives programmer great flexibility. Their work was
-based on a static analysis.
-
-Morandat et al.~\cite{ecoop12} had a short section on the usage of \eval in
-R. They found the \eval function to be widely used in R code with 8500 call
-sites in CRAN and 2.2 million dynamic calls. The 15 most frequent call sites
-account for 88\% of eval. The \c{match.arg} function is the highest used one
-with 54\% of all calls. In the other call sites, they saw two uses
-cases. The most common is the evaluation of the source code of a promise
-retrieved by \c{substitute} in a new environment. This is done in the
-\c{with} function. The other use case is the invocation of a function whose
-name or arguments are determined dynamically. For this purpose, R provides
-\c{do.call} and thus using \eval is overkill.
-
-\section{R and The Design of Eval}
-
-The R Project is a tool for data analysis.  At its heart it is a {vectorized,
-  dynamic, lazy, functional, object-oriented} programming language with an
-unusual combination of features~\cite{ecoop12} designed to ease learning by
-non-programmers and enable rapid development of statistical
-methods~\cite{R96}.  The language was designed as a successor to
-S~\cite{S88}.
-
-Most R data types are vectorized. Values are constructed by the \c{c(...)}
-function, e.g., \c{c("hi","ho")} creates a vector of two strings.  R does
-not differentiate scalars from vectors. To enable equational reasoning, R
-copies values accessible through multiple variables when they are updated.
-Values can be tagged by user-defined attributes. For instance, one can
-attach the attribute \c{dim} to the value \c{x<-c(1,2,3,4)} with
-\c{attr(x,"dim")<-c(2,2)}.  This causes arithmetic functions to treat \c x
-as a $2 \times 2$ matrix. Another attribute is \c{class} which can be bound to a list
-of names, e.g., \c{class(x)<-"human"}. This sets the class of \c{x} to
+\section{Background and Previous work}
+
+\subsection{R in a nutshell}
+
+R is a lazy functional programming language~\cite{ecoop12} with
+dynamic features that allow to write object-oriented code.  Most data
+types are vectorized. Values are constructed by the \c{c(...)}
+function, e.g. \c{c("hi","ho")} creates a vector of two strings.  To
+enable equational reasoning, R copies values accessible through
+multiple variables when they are written to.  Values can be tagged by
+user-defined attributes. For instance, one can attach the attribute
+\c{dim} to the value \c{x<-c(1,2,3,4)} with \c{attr(x,"dim")<-c(2,2)}.
+This causes arithmetic functions to treat \c x as a $2 \times 2$
+matrix. Another attribute is \c{class} which can be bound to a list of
+names, e.g., \c{class(x)<-"human"}. This sets the class of \c{x} to
 \c{human}; classes are used for object-oriented dispatch.  Every R
-linguistic construct is desugared to a function call, even control flow
-statements, assignments, and bracketing. Furthermore, all functions can be
-redefined in user code. This makes R both flexible and challenging to
-compile~\cite{dls19}.
-
-Arguments to user-defined functions are bundled into thunks called
-promises. Logically, a promise combines an expression's code, its
-environment, and its value.  To access the value of a promise, one must
-force it. Forcing a promise triggers evaluation and the computed value is
-captured for future reference.
-
-R supports reflection and meta-programming. The function \c{substitute(e,envir)}
-yields the parse tree of the expression \c{e} after performing substitutions
-defined by the bindings in the \c{envir} parameter.
+linguistic construct is desugared to a function call, even control
+flow statements, assignments, and bracketing. Furthermore, all
+functions can be redefined in user code. This makes R both flexible
+and challenging to compile~\cite{dls19}.  Arguments to user-defined
+functions are bundled into thunks called \emph{promises}. Logically, a
+promise combines an expression's code, its environment, and its value.
+To access the value of a promise, one must force it. Forcing a promise
+triggers evaluation and the computed value is captured for future
+reference.
+
+\subsection{Eval Semantics}
+
+R supports reflection and meta-programming. The function
+\c{substitute(e,envir)} yields the parse tree of the expression \c{e}
+after performing substitutions defined by the bindings in the
+\c{envir} parameter.
 
@@
 
-\begin{figure*}[!t]\centering\includegraphics[width=.8\linewidth]
-{pipeline.pdf}\caption{Analysis Pipeline}\label{fig:pipeline}
-\end{figure*}
-
 \medskip\emph{Discussion:~} There are two outstanding differences that set
@@
 that are otherwise not aware of the use of \eval may be deeply impacted.
+
+\subsection{Previous Work}
+Originally, \eval was limited to interpreted languages, but with the
+advent of just-in-time compilation this restriction was gradually
+lifted.
+
+Some languages have chosen to restrict \eval. The essential difference
+between languages relates to scoping of variables that can be accessed
+from code executed by \eval.  Java and Julia are among the more
+restrictive languages. In Java, \eval must be synthesized as a
+combination of dynamic loading and reflection.\footnote{While Java
+does not have \eval built-in, one can take the input to \eval, wrap it
+in a static method of an anonymous class, generate bytecode for that
+class, invoke the Java class loader to dynamically install the new
+code, and then use reflection to call the new method.} Julia provides
+\eval more directly but limits its scope to top level variables and
+methods. JavaScript supports both a, so-called, global \eval, and a
+local \eval, where the latter has acces to local variables of the
+function being executed (as well as variables of lexically enclosing
+scopes). Finally, languages such as R allow \eval to even manipulate
+variables defined in the caller environment of the function in which
+\eval is invoked.
+
+The difference between these design choices has an impact on the
+ability of compilers to optimize, and of analysis tools to reason
+about code. The impact of a top level \eval is that any global
+variable may be modified, but currently running functions can continue
+to execute unchanged. A local \eval, may change the values of local
+variables, their types, or even their presence. This is a much more
+invasive change, one that may invalidate currently executing functions
+as it may change the results of common optimizations such as constant
+folding or copy propagation. The designers of the Julia programming
+language were aware of this difference and purposefully chose to
+restrict \eval to the top level in order to preserve their ability to
+optimize code~\cite{oopsla18a}. They even added a versioning mechanism
+to function definitions to ensure that code added during an \eval
+would not invalidate optimizations such as inlining.  R's ability to
+affect any one of the callers of a function is even more invasive as
+the impact of \eval is non-local. Such unfettered power is difficult
+to deal with.
+
+Richards et al.~\cite{ecoop11} provided the first large-scale study of
+the runtime behavior of \eval in JavaScript. They analyzed a corpus of
+the 10,000 most popular websites. They used an instrumented web
+browser to gather execution traces and performed a dynamic analysis.
+Their results show that \eval is pervasive with 82\% of the most
+popular websites using it. The reasons for using \eval include the
+desire to load code on demand (to reduce the time it takes to display
+a page), the deserialization of JSON data (about 37\% of all calls)
+and lightweight meta-programming to customize web pages.  While many
+uses \eval were legitimate, many were unnecessary and could be
+replaced with equivalent and safer code.  They reported on a
+categorization of inputs to \eval that covers the vast majority of
+input strings.  Restricting themselves to \eval in which all named
+variables refer to the global scope, many patterns could be replaced
+by more disciplined code~\cite{oopsla12b}.  The limitation of that
+work are that there was no measure of code coverage, so the numbers
+presented can be a lower bound on the possible behaviors.  Furthermore
+in 2011, JavaScript usage was quite different from today. Their work
+captured very few libraries and none of the more recent Node.js
+ecosystem.  More details about dynamic analysis of JavaScript can be
+found in~\cite{liang}.
+
+Wang et al.~\cite{wang} analyzed usage of dynamic features in 18
+Python programs include dynamic code generation to find if they affect
+file change-proneness.  They reported that files with dynamic features
+are significantly more likely to be the subject of changes, than other
+files.  Chen et al. looked at the correlation between code changes and
+dynamic features, including \eval, in 17 Python
+programs~\cite{chen}. They did not observe many uses of \eval.  Callau
+et al.~\cite{oscar} performed an empirical study of the usage of
+dynamic features in the 1,000 largest Smalltalk projects in the
+Squeaksource repository, accounting for more than 4 million lines of
+code~\cite{oscar}. While \eval itself is not present, Smalltalk has a
+rich reflective interface that gives programmer great
+flexibility. Their work was based on a static analysis.
+
+Morandat et al.~\cite{ecoop12} had a short section on the usage of
+\eval in R. They found the \eval function to be widely used in R code
+with 8500 call sites in CRAN and 2.2 million dynamic calls. The 15
+most frequent call sites account for 88\% of eval. The \c{match.arg}
+function is the highest used one with 54\% of all calls. In the other
+call sites, they saw two uses cases. The most common is the evaluation
+of the source code of a promise retrieved by \c{substitute} in a new
+environment. This is done in the \c{with} function. The other use case
+is the invocation of a function whose name or arguments are determined
+dynamically. For this purpose, R provides \c{do.call} and thus using
+\eval is overkill.
+
+\begin{figure}[!t]\hspace{-5mm}\includegraphics[width=1.05\linewidth]
+{pipeline.pdf}\caption{Analysis Pipeline}\label{fig:pipeline}
+\end{figure}
 
@@
 
-\begin{figure}[!h]\centering\includegraphics[width=\linewidth]
+\begin{figure}[!h]\centering\includegraphics[width=.7\linewidth]
   {corpus.pdf}\caption{CRAN packages}\label{fig:corpus}

### 2020-11-26 inst/paper/main.tex
background

@@
 \vspace{2mm}\noindent\emph{Is Past Prologue?} Previous work
-investigated how \eval is used in web programming, specifically in web
-sites that use JavaScript~\cite{pldi10a}. In 2010, 17 of the largest
-web site used the feature. In 2011, 82\% of the 10,000 most accessed
-sites used \eval~\cite{ecoop11}.  Yet, the strings passed to \eval,
-and their behaviors when executed, are far from random; it was shown
-that when one can observe several calls to \eval, the ``shape'' of
-future calls can be predicted with 97\% accuracy~\cite{oopsla12b}.
+investigated how \eval is used in web programming, specifically in
+websites that use JavaScript~\cite{pldi10a}. In 2010, 17 of the
+largest website used the feature. In 2011, 82\% of the 10,000 most
+accessed sites used \eval~\cite{ecoop11}.  Yet, the strings passed to
+\eval, and their behaviors when executed, are far from random; it was
+shown that when one can observe several calls to \eval, the ``shape''
+of future calls can be predicted with 97\% accuracy~\cite{oopsla12b}.
 Overall, practical usage suggested that most reflective calls were
@@
 \subsection{Eval Semantics}
-
-R supports reflection and meta-programming. The function
-\c{substitute(e,envir)} yields the parse tree of the expression \c{e}
-after performing substitutions defined by the bindings in the
-\c{envir} parameter.
-
-\begin{lstlisting}
-> substitute(expression(a+b), list(a=1)))  
-expression(1+b)
-\end{lstlisting}
-
-\noindent
-R allows programmatic manipulation of parse trees, which are themselves
-first class objects. They are evaluated using the \c{eval()}
-function. There are three alternative forms  of \eval in the \c{base} package:
-\c{evalq}, \c{eval.parent} and \c{local}.
-
-\eval function in R takes three arguments as shown below.
-\begin{lstlisting}
- eval(expr,
-      envir = parent.frame(),
-      enclos = if(is.list(envir) || is.pairlist(envir))
-               parent.frame() else baseenv())
-\end{lstlisting}
-
-\c{expr} is the value to be evaluated, \c{envir}
-is the environment in which it is to be evaluated, and \c{enclos} is
-relevant when \c{envir} is a list or a data frame, then it specifies where R
-looks for objects not found in \c{envir}. This can be \c{NULL} (interpreted
-as the base package environment, \c{baseenv()}) or an environment.
-
-When called, \eval evaluates \c{expr} in the environment specified by
-\c{envir} and returns the computed value. If \c{envir} is missing, then the
-default is \c{parent.frame()} (the environment where the call to \eval was
-made).  Values to be evaluated can be of types \c{call} (AST for function
-calls) or \c{expression} (AST nodes for arbitrary expressions), \c{name}
-(when the symbol is looked up in the current scope and its binding is
-evaluated), promise, or any of the basic types such as vectors and
-environments (which are returned unchanged).
-
-The \c{evalq} form is equivalent to \c{eval(quote(expr), ...)}.  Quoting
-that argument prevents if from being evaluated in the current environment.
-In the following code snippet, variable \c{r} is first looked up in
-the current scope with \eval whereas it is looked up in \c{env} with
-\c{evalq}.
-
-\begin{lstlisting}
-> r <- 5
-> env <- new.env()
-> env$r <- 9
-> eval(r, envir = env)
-[1] 5
-> evalq(r, envir = env)
-[1] 9
-\end{lstlisting}
-% $
-
-\c{eval.parent(expr, n)} is a shorthand for \c{eval(expr, parent.frame(n))}.
-\c{local} evaluates an expression in a local environment. It is equivalent
-to evalq except that its default argument creates a new, empty
-environment. This is useful to create anonymous recursive functions and as a
-kind of limited namespace feature since variables defined in the environment
-are not visible from the outside.
-
-\medskip\emph{Discussion:~} There are two outstanding differences that set
-the design of \eval in R apart from JavaScript. On the usability side, \eval
-does not take a string, but an AST and an environment. This is slightly more
-involved than just taking a string and may discourage some casual users.
-The second substantial difference is the control that R gives over the
-choice of environment in which the AST is evaluated in. Beside traditional
-choices such as the top level or current lexical environment, R allows
-programmers to choose a fresh, empty, environment or any environment on the
-call stack. This is a significant increase in expressive power as functions
-that are otherwise not aware of the use of \eval may be deeply impacted.
-
-\subsection{Previous Work}
 Originally, \eval was limited to interpreted languages, but with the
@@
 
+R supports reflection and meta-programming. The function
+\c{substitute(e,envir)} yields the parse tree of the expression \c{e}
+after performing substitutions defined by the bindings in the
+\c{envir} parameter.
+
+\begin{lstlisting}
+> substitute(expression(a+b), list(a=1)))  
+expression(1+b)
+\end{lstlisting}
+
+\noindent
+R allows programmatic manipulation of parse trees, which are themselves
+first class objects. They are evaluated using the \c{eval()}
+function. There are three alternative forms  of \eval in the \c{base} package:
+\c{evalq}, \c{eval.parent} and \c{local}.
+
+\eval function in R takes three arguments as shown below.
+\begin{lstlisting}
+ eval(expr,
+      envir = parent.frame(),
+      enclos = if(is.list(envir) || is.pairlist(envir))
+               parent.frame() else baseenv())
+\end{lstlisting}
+
+\c{expr} is the value to be evaluated, \c{envir}
+is the environment in which it is to be evaluated, and \c{enclos} is
+relevant when \c{envir} is a list or a data frame, then it specifies where R
+looks for objects not found in \c{envir}. This can be \c{NULL} (interpreted
+as the base package environment, \c{baseenv()}) or an environment.
+
+When called, \eval evaluates \c{expr} in the environment specified by
+\c{envir} and returns the computed value. If \c{envir} is missing, then the
+default is \c{parent.frame()} (the environment where the call to \eval was
+made).  Values to be evaluated can be of types \c{call} (AST for function
+calls) or \c{expression} (AST nodes for arbitrary expressions), \c{name}
+(when the symbol is looked up in the current scope and its binding is
+evaluated), promise, or any of the basic types such as vectors and
+environments (which are returned unchanged).
+
+The \c{evalq} form is equivalent to \c{eval(quote(expr), ...)}.  Quoting
+that argument prevents if from being evaluated in the current environment.
+In the following code snippet, variable \c{r} is first looked up in
+the current scope with \eval whereas it is looked up in \c{env} with
+\c{evalq}.
+
+\begin{lstlisting}
+> r <- 5
+> env <- new.env()
+> env$r <- 9
+> eval(r, envir = env)
+[1] 5
+> evalq(r, envir = env)
+[1] 9
+\end{lstlisting}
+% $
+
+\c{eval.parent(expr, n)} is a shorthand for \c{eval(expr, parent.frame(n))}.
+\c{local} evaluates an expression in a local environment. It is equivalent
+to evalq except that its default argument creates a new, empty
+environment. This is useful to create anonymous recursive functions and as a
+kind of limited namespace feature since variables defined in the environment
+are not visible from the outside.
+
+\medskip\emph{Discussion:~} There are two outstanding differences that set
+the design of \eval in R apart from JavaScript. On the usability side, \eval
+does not take a string, but an AST and an environment. This is slightly more
+involved than just taking a string and may discourage some casual users.
+The second substantial difference is the control that R gives over the
+choice of environment in which the AST is evaluated in. Beside traditional
+choices such as the top level or current lexical environment, R allows
+programmers to choose a fresh, empty, environment or any environment on the
+call stack. This is a significant increase in expressive power as functions
+that are otherwise not aware of the use of \eval may be deeply impacted.
+
+\subsection{Previous Work}
+
 Richards et al.~\cite{ecoop11} provided the first large-scale study of
-the runtime behavior of \eval in JavaScript. They analyzed a corpus of
-the 10,000 most popular websites. They used an instrumented web
-browser to gather execution traces and performed a dynamic analysis.
-Their results show that \eval is pervasive with 82\% of the most
-popular websites using it. The reasons for using \eval include the
-desire to load code on demand (to reduce the time it takes to display
-a page), the deserialization of JSON data (about 37\% of all calls)
-and lightweight meta-programming to customize web pages.  While many
-uses \eval were legitimate, many were unnecessary and could be
-replaced with equivalent and safer code.  They reported on a
-categorization of inputs to \eval that covers the vast majority of
-input strings.  Restricting themselves to \eval in which all named
-variables refer to the global scope, many patterns could be replaced
-by more disciplined code~\cite{oopsla12b}.  The limitation of that
-work are that there was no measure of code coverage, so the numbers
-presented can be a lower bound on the possible behaviors.  Furthermore
-in 2011, JavaScript usage was quite different from today. Their work
-captured very few libraries and none of the more recent Node.js
-ecosystem.  More details about dynamic analysis of JavaScript can be
-found in~\cite{liang}.
-
-Wang et al.~\cite{wang} analyzed usage of dynamic features in 18
-Python programs include dynamic code generation to find if they affect
-file change-proneness.  They reported that files with dynamic features
-are significantly more likely to be the subject of changes, than other
-files.  Chen et al. looked at the correlation between code changes and
-dynamic features, including \eval, in 17 Python
-programs~\cite{chen}. They did not observe many uses of \eval.  Callau
-et al.~\cite{oscar} performed an empirical study of the usage of
-dynamic features in the 1,000 largest Smalltalk projects in the
-Squeaksource repository, accounting for more than 4 million lines of
-code~\cite{oscar}. While \eval itself is not present, Smalltalk has a
-rich reflective interface that gives programmer great
-flexibility. Their work was based on a static analysis.
+the runtime behavior of \eval in JavaScript. They dynamically analyzed
+a corpus of the 10,000 most popular websites with an instrumented web
+browser to gather execution traces.  They show that \eval is pervasive
+with 82\% of the most popular websites using it. The reasons for its
+use include the desire to load code on demand, deserialization of JSON
+dataq and lightweight meta-programming to customize web pages.  While
+many uses were legitimate, just as many were unnecessary and could be
+replaced with equivalent and safer code.  They categorized inputs to
+\eval so as to cover the vast majority of input strings.  Restricting
+themselves to \eval in which all named variables refer to the global
+scope, many patterns could be replaced by more disciplined
+code~\cite{oopsla12b}.  The work did not measure code coverage, so the
+numbers presented are a lower bound on the possible behaviors.
+Furthermore, JavaScript usage in 2011 is likely different from today,
+e.g. Node.js was not covered by Richards.  More details about dynamic
+analysis of JavaScript can be found in~\cite{liang}.
+
+Wang et al.~\cite{wang} analyzed use of dynamic features in 18 Python
+programs to find if they affect file change-proneness.  Files with
+dynamic features are significantly more likely to be the subject of
+changes, than other files.  Chen et al. looked at the correlation
+between code changes and dynamic features, including \eval, in 17
+Python programs~\cite{chen}. They did not observe many uses of \eval.
+Callau et al.~\cite{oscar} performed an empirical study of the usage
+of dynamic features in 1,000 Smalltalk projects. While \eval itself is
+not present, Smalltalk has a rich reflective interface that gives
+programmer great flexibility. Their work was based on a static
+analysis.  %% TODO:: Say someting meaningfull about OSCAR
 
 Morandat et al.~\cite{ecoop12} had a short section on the usage of
-\eval in R. They found the \eval function to be widely used in R code
-with 8500 call sites in CRAN and 2.2 million dynamic calls. The 15
-most frequent call sites account for 88\% of eval. The \c{match.arg}
-function is the highest used one with 54\% of all calls. In the other
-call sites, they saw two uses cases. The most common is the evaluation
-of the source code of a promise retrieved by \c{substitute} in a new
-environment. This is done in the \c{with} function. The other use case
-is the invocation of a function whose name or arguments are determined
-dynamically. For this purpose, R provides \c{do.call} and thus using
-\eval is overkill.
+\eval in R. They found the it widely used in R code with 8500 call
+sites in CRAN and 2.2 million dynamic calls. The 15 most frequent call
+sites account for 88\% of those. The \c{match.arg} function is the
+highest used one with 54\% of all calls. In the other call sites, they
+saw two uses cases. The most common is the evaluation of the source
+code of a promise retrieved by \c{substitute} in a new environment;
+e.g. as done in the \c{with} function. The other use case is the
+invocation of a function whose name or arguments are determined
+dynamically. For this purpose, R provides \c{do.call} and thus \eval
+is overkill.
 

### 2020-11-26 inst/analysis/environments.Rmd

@@
   as_tibble() %>%
-  deduplicate() %>% 
-  add_types() %>%
-  add_eval_source(eval_calls_raw) %>% add_eval_source_type()
+  deduplicate()
+
+eval_calls <-   add_eval_source(eval_calls) %>% add_eval_source_type()
 ```

### 2020-11-26 inst/analysis/preprocess.R
Preprocessing pipeline in shape

@@ -8,24 +8,87 @@
 
 
 suppressPackageStartupMessages(library(tidyverse))
+suppressPackageStartupMessages(library(rlang))
 library(fs)
 library(fst)
 
 
 
-# Rather relative source because here it depends on where we launch the script...
-source("revalstudy/inst/analysis/insights.R")
+get_expr <- function(eval_call) {
+  if (is.na(eval_call)) {
+    return(NA)
+  }
+  # Special case for expressions starting with _ (such as _inherit in ggproto)
+  escaped_eval_call <- if (startsWith(eval_call, "_")) {
+    paste0("`", eval_call, "`")
+  }
+  else {
+    eval_call
+  }
+
+  exp <- NA
+  # Would fail for instance for
+  # "`$<-`(new(\"C++Field\", .xData = <environment>), \"read_only\", TRUE)" (classInt package)
+  try(exp <- parse(text = escaped_eval_call)[[1]], silent = TRUE)
 
 
-deduplicate <- function(dataset)
-{
+  return(exp)
+}
+
+
+deduplicate <- function(dataset) {
   return(dataset %>% count(across(c(
-    -eval_call_id,-starts_with("caller_stack")
+    -eval_call_id, -starts_with("caller_stack")
   )), name = "nb_ev_calls"))
 }
 
-add_types <- function(dataset)
-{
+
+SEXP_TYPES <- tribble(
+  ~sexp_type, ~name,
+  0, "NILSXP",
+  1, "SYMSXP",
+  2, "LISTSXP",
+  3, "CLOSXP",
+  4, "ENVSXP",
+  5, "PROMSXP",
+  6, "LANGSXP",
+  7, "SPECIALSXP",
+  8, "BUILTINSXP",
+  9, "CHARSXP",
+  10, "LGLSXP",
+  13, "INTSXP",
+  14, "REALSXP",
+  15, "CPLXSXP",
+  16, "STRSXP",
+  17, "DOTSXP",
+  18, "ANYSXP",
+  19, "VECSXP",
+  20, "EXPRSXP",
+  21, "BCODESXP",
+  22, "EXTPTRSXP",
+  23, "WEAKREFSXP",
+  24, "RAWSXP",
+  25, "S4SXP",
+  30, "NEWSXP",
+  31, "FREESXP",
+  99, "FUNSXP"
+)
+
+
+SEXP_TYPES <- SEXP_TYPES %>% mutate(name = factor(name))
+
+resolve_sexp_name <- function(df, var) {
[…]

### 2020-11-26 inst/analysis/preprocess.R
NA -> NA_character_

@@ -16,7 +16,7 @@ library(fst)
 
 get_expr <- function(eval_call) {
   if (is.na(eval_call)) {
-    return(NA)
+    return(NA_character_)
   }
   # Special case for expressions starting with _ (such as _inherit in ggproto)
   escaped_eval_call <- if (startsWith(eval_call, "_")) {
@@ -26,7 +26,7 @@ get_expr <- function(eval_call) {
     eval_call
   }
 
-  exp <- NA
+  exp <- NA_character_
   # Would fail for instance for
   # "`$<-`(new(\"C++Field\", .xData = <environment>), \"read_only\", TRUE)" (classInt package)
   try(exp <- parse(text = escaped_eval_call)[[1]], silent = TRUE)
@@ -471,8 +471,8 @@ main <- function(argv) {
 
   cat("Deduplicating\n")
   eval_calls <- eval_calls_raw %>% deduplicate()
-  cat("Adding types\n")
-  eval_calls <- eval_calls %>% add_types()
+  # cat("Adding types\n")
+  # eval_calls <- eval_calls %>% add_types() # This step is now useless as there are directly strings
   cat("Correcting srcrefs\n")
   eval_calls <-
     eval_calls %>%

### 2020-11-27 inst/paper/main.tex
mijor

@@
 \subsection{Eval Semantics}
+
+There are two outstanding differences that set
+the design of \eval in R apart from JavaScript. On the usability side, \eval
+does not take a string, but an AST and an environment. This is slightly more
+involved than just taking a string and may discourage some casual users.
+The second substantial difference is the control that R gives over the
+choice of environment in which the AST is evaluated in. Beside traditional
+choices such as the top level or current lexical environment, R allows
+programmers to choose a fresh, empty, environment or any environment on the
+call stack. This is a significant increase in expressive power as functions
+that are otherwise not aware of the use of \eval may be deeply impacted.
+
 Originally, \eval was limited to interpreted languages, but with the
@@
 are not visible from the outside.
-
-\medskip\emph{Discussion:~} There are two outstanding differences that set
-the design of \eval in R apart from JavaScript. On the usability side, \eval
-does not take a string, but an AST and an environment. This is slightly more
-involved than just taking a string and may discourage some casual users.
-The second substantial difference is the control that R gives over the
-choice of environment in which the AST is evaluated in. Beside traditional
-choices such as the top level or current lexical environment, R allows
-programmers to choose a fresh, empty, environment or any environment on the
-call stack. This is a significant increase in expressive power as functions
-that are otherwise not aware of the use of \eval may be deeply impacted.
 

### 2020-11-27 inst/paper/main.tex
more bib

@@
 can be done. A frustrated group researchers argued giving up on
-soundness and, instead, under-approximating dynamic
-features (soundiness)~\cite{soundy}. In their words ``a practical analysis,
+soundness and, instead, under-approximating dynamic features
+(soundiness)~\cite{soundy}. In their words ``a practical analysis,
 therefore, may pretend that \eval does nothing, unless it can
@@
 
+This section provides a short introduction to R as it has a few
+surprising features that impact the use of reflective features of the
+language. We then look at the semantics of \eval in R and discuss some
+design choices. Lastly, we put this paper in context of previous work.
+
 \subsection{R in a nutshell}
@@
 types are vectorized. Values are constructed by the \c{c(...)}
-function, e.g. \c{c("hi","ho")} creates a vector of two strings.  To
+function, e.g. \c{c("a","bc")} creates a vector of two strings. To
 enable equational reasoning, R copies values accessible through
@@
 
-There are two outstanding differences that set
-the design of \eval in R apart from JavaScript. On the usability side, \eval
-does not take a string, but an AST and an environment. This is slightly more
-involved than just taking a string and may discourage some casual users.
-The second substantial difference is the control that R gives over the
-choice of environment in which the AST is evaluated in. Beside traditional
-choices such as the top level or current lexical environment, R allows
-programmers to choose a fresh, empty, environment or any environment on the
-call stack. This is a significant increase in expressive power as functions
-that are otherwise not aware of the use of \eval may be deeply impacted.
-
-Originally, \eval was limited to interpreted languages, but with the
-advent of just-in-time compilation this restriction was gradually
-lifted.
-
-Some languages have chosen to restrict \eval. The essential difference
-between languages relates to scoping of variables that can be accessed
-from code executed by \eval.  Java and Julia are among the more
-restrictive languages. In Java, \eval must be synthesized as a
-combination of dynamic loading and reflection.\footnote{While Java
-does not have \eval built-in, one can take the input to \eval, wrap it
-in a static method of an anonymous class, generate bytecode for that
-class, invoke the Java class loader to dynamically install the new
-code, and then use reflection to call the new method.} Julia provides
-\eval more directly but limits its scope to top level variables and
-methods. JavaScript supports both a, so-called, global \eval, and a
-local \eval, where the latter has acces to local variables of the
-function being executed (as well as variables of lexically enclosing
-scopes). Finally, languages such as R allow \eval to even manipulate
-variables defined in the caller environment of the function in which
-\eval is invoked.
-
-The difference between these design choices has an impact on the
-ability of compilers to optimize, and of analysis tools to reason
-about code. The impact of a top level \eval is that any global
-variable may be modified, but currently running functions can continue
-to execute unchanged. A local \eval, may change the values of local
-variables, their types, or even their presence. This is a much more
-invasive change, one that may invalidate currently executing functions
-as it may change the results of common optimizations such as constant
-folding or copy propagation. The designers of the Julia programming
-language were aware of this difference and purposefully chose to
-restrict \eval to the top level in order to preserve their ability to
-optimize code~\cite{oopsla18a}. They even added a versioning mechanism
-to function definitions to ensure that code added during an \eval
-would not invalidate optimizations such as inlining.  R's ability to
-affect any one of the callers of a function is even more invasive as
-the impact of \eval is non-local. Such unfettered power is difficult
-to deal with.
+The expressive power of \eval depends on a number of choices made by
+the programming language designers. The two axes that are the most
+relevant here are:
+\begin{itemize}
+\item {\bf Scoping:} What environment does the string passed to \eval
+  execute in? JavaScript and R evaluate the expression in the current
+  scope. This means that local variables and parameters are exposed to
+  \eval, breaking the abstraction boundary of the enclosing function.
+  Julia takes a more restrictive approach by running \eval at the
+  ``top level'', i.e. in the global environment, outside of any
+  function's environment. JavaScript has a global \eval that behaves
+  like Julia, it also has a ``strict mode'' in which \eval is allowed
+  to access local variables, but fobidden from injecting new variables
+  in the local environment. In Java, one can implement \eval but the
+  expression will execute in new environment.\footnote{While \eval
+  used to be available only in interpreted languages, the advent of
+  just-in-time compilation gradually lifted this restriction. To
+  implement \eval in Java requires wrapping its input in a static
+  method of an anonymous class, generating bytecode for that class,
+  invoking the class loader to install the code, and then using
+  reflection to call the new method. }
+\item {\bf Reflective API:} What other reflective operations are
+  exposed to user code? JavaScript does provides few reflective
+  functions (other than being able to enumerate the properties of an
+  object and to access properties with strings), the string passed to
+  \eval is thus limited to what can be expressed in JavaScript. R, on
+  the other hand, has a rich reflective interface. For example, R
+  allows user code to walk the call stack and arbitrarily add and
+  delete local variables in any environment.
+\end{itemize}
+
+\noindent
+These design choices impact our ability to reason about code. More
+expressive power for \eval translates directly into additional
+restriction for program analysis and transformation tools. The reason
+Julia has restricts \eval to execute in top level environment is that
+they wanted to preserve their compiler's ability to optimize
+functions~\cite{oopsla18a}. What they wanted to prevent was the case
+where \eval changed the type of local variables as this would require
+recompilation of the method (or prevent optimization). Julia even adds
+a versoning mechanism, called world age, to ensure that code added
+during an \eval does not invalidate optimizations such as
+inlining~\cite{oopsla20a}.
+
+R's ability to affect any one of the callers of a function is even
+more invasive as the impact of \eval is non-local. Such unfettered
+power is difficult to deal with.
+
+There are two outstanding differences that set the design of \eval in
+R apart from JavaScript. On the usability side, \eval does not take a
+string, but an AST and an environment. This is slightly more involved
+than just taking a string and may discourage some casual users. The
+second substantial difference is the control that R gives over the
+choice of environment in which the AST is evaluated in. Beside
+traditional choices such as the top level or current lexical
+environment, R allows programmers to choose a fresh, empty,
+environment or any environment on the call stack. This is a
+significant increase in expressive power as functions that are
+otherwise not aware of the use of \eval may be deeply impacted.
 

### 2020-11-27 inst/paper/main.tex
sec 2

@@
 \title{Why do we Eval?\\[2mm]\Large A large-scale study of Eval usage in R}
-\titlerunning{Eval in R}
+\titlerunning{Why Eval (in R)?}
 
@@
 compiler optimizations will be unsound. Furthermore the addition of
-new code --code that could have been obtained from a network
-connection-- as the program is running is a security vulnerability
+new code --- code that could have been obtained from a network
+connection --- as the program is running is a security vulnerability
 waiting to happen.
@@
 two-level monitoring infrastructure: we can monitor R programs by
-instrumentation --this gives us access to many user-visible properties
-of R programs-- but we can also monitor the inner working of the R
-interpreter --this allows us to capture details of execution that are
+instrumentation --- this gives us access to many user-visible properties
+of R programs --- but we can also monitor the inner working of the R
+interpreter --- this allows us to capture details of execution that are
 not exposed at the source level. Dynamic analysis is of course limited
@@
 during an \eval does not invalidate optimizations such as
-inlining~\cite{oopsla20a}.
-
-R's ability to affect any one of the callers of a function is even
-more invasive as the impact of \eval is non-local. Such unfettered
-power is difficult to deal with.
-
-There are two outstanding differences that set the design of \eval in
-R apart from JavaScript. On the usability side, \eval does not take a
-string, but an AST and an environment. This is slightly more involved
-than just taking a string and may discourage some casual users. The
-second substantial difference is the control that R gives over the
+inlining~\cite{oopsla20a}. More permissive languages such as R are
+much harder to deal with from a compiler's perspective.
+
+\medskip\noindent{\it Eval in R.} There are four reflective functions
+that take abstract syntax trees (called expressions), parse it as
+code, and evaluate it. These functions are \c{eval}, \c{evalq},
+\c{eval.parent}, and \c{local}. Unlike other languages, R's \eval does
+not take a string, but rather an expression. As this is slightly more
+involved, it may discourage some casual users. There are two ways to
+obtain an expression. Given a string, the \c{parse()} function retuns
+an unevaluated expression. The other way is to use the function
+\c{substitute(e,envir)}, it yields the parse tree of the expression
+\c{e} after performing substitutions defined by the bindings in the
+\c{envir} parameter.
+
+\begin{lstlisting}
+ > substitute(expression(a+b), list(a=1)))
+ expression(1+b)
+\end{lstlisting}
+
+\noindent
+This function is widely used in conjunction with \eval, as it can
+leverage R's laziness for metaprogramming. R allows programmatic
+manipulation of parse trees, which are themselves first class objects.
+Expressions are evaluated using \eval (or one its variants). The \eval
+function takes three arguments:
+
+\begin{lstlisting}
+ eval(expr,
+      envir = parent.frame(),
+      enclos = if(is.list(envir) || is.pairlist(envir))
+               parent.frame() else baseenv())
+\end{lstlisting}
+
+\noindent
+This code is part of the definition of \eval, it specifies that the
+function takes an \c{expr} which is the value to be evaluated,
+\c{envir} is the environment in which evaluation happes, and
+\c{enclos} is relevant when \c{envir} is a list or a data frame, then
+it specifies where R looks for objects not found in \c{envir}. This
+can be \c{NULL} (interpreted as the base package environment,
+\c{baseenv()}) or an environment. Thus, the expression \c{expr}
+evaluates in the environment \c{envir} and returns the computed value.
+If \c{envir} is missing, then the default is \c{parent.frame()} (the
+environment where the call to \eval was made). Values to be evaluated
+can be of types \c{call} (function calls) or \c{expression} (arbitrary
+expressions), \c{name} (when the symbol is looked up in the current
+scope and its binding is evaluated), promise, or any of the basic
+types such as vectors and environments (which are returned unchanged).
+
+The \c{evalq} form is equivalent to \c{eval(quote(expr), ...)}.
+Quoting that argument prevents if from being evaluated in the current
+environment. \c{eval.parent(expr, n)} is a shorthand for \c{eval(expr,
+  parent.frame(n))}. \c{local} evaluates an expression in a local
+environment. It is equivalent to evalq except that its default
+argument creates a new, empty environment. This is useful to create
+anonymous recursive functions and as a kind of limited namespace
+feature since variables defined in the environment are hidden.
+
+The second substantial difference is the control that R gives over the
 choice of environment in which the AST is evaluated in. Beside
@@
 otherwise not aware of the use of \eval may be deeply impacted.
-
-R supports reflection and meta-programming. The function
-\c{substitute(e,envir)} yields the parse tree of the expression \c{e}
-after performing substitutions defined by the bindings in the
-\c{envir} parameter.
-
-\begin{lstlisting}
-> substitute(expression(a+b), list(a=1)))
-expression(1+b)
-\end{lstlisting}
-
-\noindent
-R allows programmatic manipulation of parse trees, which are themselves
-first class objects. They are evaluated using the \c{eval()}
-function. There are three alternative forms  of \eval in the \c{base} package:
-\c{evalq}, \c{eval.parent} and \c{local}.
-
-\eval function in R takes three arguments as shown below.
-\begin{lstlisting}
- eval(expr,
-      envir = parent.frame(),
-      enclos = if(is.list(envir) || is.pairlist(envir))
-               parent.frame() else baseenv())
-\end{lstlisting}
-
-\c{expr} is the value to be evaluated, \c{envir}
-is the environment in which it is to be evaluated, and \c{enclos} is
-relevant when \c{envir} is a list or a data frame, then it specifies where R
-looks for objects not found in \c{envir}. This can be \c{NULL} (interpreted
-as the base package environment, \c{baseenv()}) or an environment.
-
-When called, \eval evaluates \c{expr} in the environment specified by
-\c{envir} and returns the computed value. If \c{envir} is missing, then the
-default is \c{parent.frame()} (the environment where the call to \eval was
-made).  Values to be evaluated can be of types \c{call} (AST for function
-calls) or \c{expression} (AST nodes for arbitrary expressions), \c{name}
-(when the symbol is looked up in the current scope and its binding is
-evaluated), promise, or any of the basic types such as vectors and
-environments (which are returned unchanged).
-
-The \c{evalq} form is equivalent to \c{eval(quote(expr), ...)}.  Quoting
-that argument prevents if from being evaluated in the current environment.
-In the following code snippet, variable \c{r} is first looked up in
-the current scope with \eval whereas it is looked up in \c{env} with
-\c{evalq}.
-
-\begin{lstlisting}
-> r <- 5
-> env <- new.env()
-> env$r <- 9
-> eval(r, envir = env)
-[1] 5
-> evalq(r, envir = env)
-[1] 9
-\end{lstlisting}
-% $
-
-\c{eval.parent(expr, n)} is a shorthand for \c{eval(expr, parent.frame(n))}.
-\c{local} evaluates an expression in a local environment. It is equivalent
-to evalq except that its default argument creates a new, empty
-environment. This is useful to create anonymous recursive functions and as a
-kind of limited namespace feature since variables defined in the environment
-are not visible from the outside.
 

### 2020-11-27 inst/paper/main.tex
sec 3

@@
 of dynamic features in 1,000 Smalltalk projects. While \eval itself is
-not present, Smalltalk has a rich reflective interface that gives
-programmer great flexibility. Their work was based on a static
-analysis.  %% TODO:: Say someting meaningfull about OSCAR
+not present, Smalltalk has a rich reflective interface. The authors
+found that reflective are used in less than 2\% of methods. The most
+common reflective method is \c{perform:}; it send a message that is
+specified by a string. These features are mostly used in the core
+libraries.
+
+Bodden et al.~\cite{bodden} looked at usage of reflection in the Java
+DaCapo benchmark suite. They found that dynamic loading was triggered
+by the benchmark harness. The harness then executes methods via
+reflection, this caused static analysis tools to generate an incorrect
+call graph for the programs in DaCapo.
 
@@
 
-\begin{figure}[!t]\hspace{-5mm}\includegraphics[width=1.05\linewidth]
-{pipeline.pdf}\caption{Analysis Pipeline}\label{fig:pipeline}
-\end{figure}
-
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Methodology}
 
-\subsection{Infrastructure}
-
-Our infrastructure consists of a dynamic analysis pipeline to record the
-calls to \eval and RMarkdown notebooks for data analysis and visualization.
-Figure~\ref{fig:pipeline} shows the main steps of the pipeline. We provide
-the runtime, input and output data size for each step. All runs are done on
-a single server Intel Xeon 6140, 2.30GHz with 72 cores and 256GB of RAM.
-
-The pipeline begins with downloading and installing open source R packages
-from CRAN for the corpus along with their dependencies. The Kaggle programs
-are downloaded from \href{http://www.kaggle.com}{Kaggle} separately by a
-crawler.  Next, code size and coverage metrics from the installed packages
-are extracted using the \href{ https://github.com/r-lib/covr}{\covr}
-package. This is followed by extraction of runnable programs from R
-packages: tests, examples and vignettes as well as from Kaggle. The code of
-each extracted program is wrapped into a call to dynamic analyzer
-\emph{evil} (\emph{Ev}al \emph{I}nspection \emph{L}ibrary) that collects eval
-usage. The reason for this is that we only want to record eval usage for the
-extracted code. Without this, the data would include eval calls from the
-unit testing frameworks as well as from bootstrapping R virtual machine
-itself. To avoid any interference, each program is run in its own R
-instance. The \emph{evil} framework is implemented as a R package in 2K lines of
-R and 400 lines of C++ code.
-
-The dynamic analyzer builds upon the dynamic analysis framework, \instrumentr
-that we have implemented to enable us to write dynamic analysis logic in R. It
-is an R package implemented in 2.5K lines of R and 6K lines of C++ code. It
-internally uses a modified R interpreter, \rdyntrace~\cite{oopsla19a}, that
-exposes hooks from within the interpreter implementation for events of interest.
-\instrumentr serves as an intermediary between \rdyntrace and \evil, it
-intercepts the hooks exposed by \rdyntrace and attaches R functions exported by
-\evil as callbacks. The \evil callbacks execute on corresponding interpreter
-events.
-
-All steps of this pipeline are parallelized using GNU
-parallel~\cite{GNUparallel} and orchestrated by GNU make. To schedule and
-parallelize extraction and analysis of programs, we use the \runr
-package. Furthermore, \runr gracefully handles and reports failures across
-large-scale program runs which greatly aids debugging of the analysis
-pipeline.  The data extracted by \evil from each program is concatenated,
-cleaned and summarized in the post-processing phase by custom R
-scripts. Finally, the summarized data is analyzed in RMarkdown notebooks to
-gather insights. Apart from the figures, the data points included in the
-paper are also generated by RMarkdown notebooks as latex macros.
+This section explains our methodology for selecting the corpus that
+will be analyzed and how we implemented our dynamic analysis.
 
@@
 
-In this study, we report on a corpus of \CorpusCorePackages R core and
-\CorpusPackages CRAN packages with \CorpusAllCodeRnd lines of source code
-(native + R).  We run \CorpusAllProgramsRnd programs extracted from those
-packages as well as from the user-written data-analyses shared on the Kaggle
-platform. In this section we describe this corpus and methodology used to
-assemble it.
+Our corpus is assembled out of two set of programs, one consisting of
+popular libraries obtained from the Comprehensive R Archive
+Network\footnote{CRAN is the largest repository of R code with over
+\CorpusAllCranRnd packages. It receives about 6 new package
+submissions a day~\cite{Ligges2017}. Unlike other open code
+repositories such as GitHub, CRAN is a curated repository. Each
+submitted package must abide to a number of well-formedness rules that
+are automatically checked asserting certain quality. Most relevant for
+this work is that all of the runnable code is tested and only a
+successfully running package is admitted in the archive.} (we refer to
+it as {\bf CRAN}) and the other from the Kaggle data science
+competition (written {\bf Kaggle}). The intent here is to contrast
+code written by experienced developpers (CRAN) with code authored by
+typical end users of the language (Kaggle).
+
+All the code in our is runnable, i.e. we have scripts that invoke the
+CRAN packages and the Kaggle programs have input data. Clearly code
+coverage will vary; as is the case for any dynamic analysis, the
+quality of results is predicated on how representative the executions
+we can observe are with respect to all possible runs of the corpus. To
+mitigate the threat to validity attached to limited coverage, we
+obtain clients that will generate multiple \emph{runs} of CRAN
+packages. In this context, a run is the execution of a CRAN package or
+Kaggle program with a particular set of input values.
+
+\medskip\noindent\emph{CRAN Packages.} The packages that are included
+in this study are the top \CorpusPackages packages based on their
+reverse dependencies. The set of reverse dependencies for some package
+\emph{P} is transitive closure of packages that import \emph{P}. The
+hypothesis underlying that choice is that packages with a large set of
+reverse depencies are likely to be higher in quality and their tests
+will provide better coverage. The packages are implemented in R with
+some C and Fortran code. Using \c{cloc}, we measure a total
+\CorpusRCodeRnd lines of R code and \CorpusNativeCodeRnd lines of
+native code. For each package, we use its runnable code (tests, etc.)
+to compute code coverage. On average, the cde coverage is
+\CorpusMeanExprCoverage which is acceptable without being exhaustive.
+Figure~\ref{fig:corpus} shows these packages, the size of the dots
+reflects the project's size in lines of code. The x-axis indicates
+code coverage in percents and the y-axis gives the number of call
+sites to \eval that were traced, in log scale. Dotted lines indicate
+means. Packages with over \CorpusEvalsPackageTreshold eval call sites
+are named.
 
@@
 
-\subsubsection{CRAN Packages}
-
-We have selected the TOP \CorpusPackages packages based on their reverse
-dependencies, \ie how many other packages depend on them.\footnote{Extracted
-  from package metadata using a builtin function.}  This helps us find
-packages which should be better written and have better code coverage.  The
-resulting set contains \CorpusRCodeRnd lines of R code and
-\CorpusNativeCodeRnd lines of native code.\footnote{Excluding comments and
-  blank lines using \url{github.com/AlDanial/cloc}.}  For each package we
-use its runnable code to compute the package code coverage.\footnote{Computed
-  using \url{ https://github.com/r-lib/covr}.} On average, across the
-\CorpusPackages packages, it is \CorpusMeanExprCoverage.
-Figure~\ref{fig:corpus} shows these packages, the size of the dots reflects
-the project's size in lines of code.  The x-axis indicates code coverage in
-percents and the y-axis gives the number of call sites to \eval that were traced, in log
-scale. Dotted lines indicate means. Packages with over
-\CorpusEvalsPackageTreshold eval call sites are named.
-
-All included packages come from the Comprehensive R Archive Network
-(CRAN\footnote{\url{http://cran.r-project.org}}), the largest repository of R
-code with over \CorpusAllCranRnd packages.\footnote{CRAN receives about 6 new
-  package submissions a day~\cite{Ligges2017}.} Unlike other open code
-repositories such as GitHub, CRAN is a curated repository. Each submitted
-package must abide to a number of well-formedness rules that are automatically
-checked asserting certain quality. Most relevant for this work is that all of
-the \emph{runnable code} is tested and only a successfully running package is
-admitted in the archive. This is important as we use this runnable code for the
-dynamic analysis of eval calls.
-
-There are three sources of runnable code in a R package. Next to the package
-\emph{tests},\footnote{We had to exclude most tests because the {testthat}
-  harness uses \eval and thus causes the entire test to register as an \eval
-  call.} there are also \emph{examples} and \emph{vignettes}. Examples are R
-code snippets extracted from package documentation into scripts files.
-Vignettes are long-form description of package functionality written using
-literate-programming style in \LaTeX\xspace or Rmarkdown with chunks of R
-code that can be similarly extracted into R scripts. R provides builtin
-functions for tangling the embedded code into files.  The selected packages
-contain \CorpusPackagePrograms programs with \CorpusPackageProgramsCodeRnd
-lines of R code: \CorpusExamplesProgramsRnd examples with
-\CorpusExamplesCodeRnd lines of code, and \CorpusVignettesProgramsRnd
-vignettes.
-
-\subsubsection{Kaggle Scripts}
-
-To represent end-user code in the corpus, we turned to Kaggle, an online
-platform for data-science and machine-learning. The website allows people to
-share data-science competitions, a data-analysis problem together with data,
-for which users try to find the best solution. The solutions, called
-{kernels}, are then shared back to the platform in the form of either plain
-scripts or as notebooks.  One of the most popular competition is about
-predicting passenger survival on
-Titanic\footnote{\url{https://www.kaggle.com/c/titanic}} with \CorpusKaggle
-kernels in R (over 1/4 of all available R kernels) which we used for our
-corpus.
-
-Unlike CRAN, Kaggle is not a curated repository and therefore there are no
-guarantees about the quality of the code. After downloading all of the
-\CorpusKaggle kernels and extracting the R code from the various
-formats,\footnote{We use {\sf rmarkdown} to convert from notebooks to R.} we
-found that \CorpusDuplicatedKaggle were whole-file duplicates. From
-\CorpusRunnableKaggle kernels, \CorpusFailedKaggle failed to execute. Next
-to various runtime exceptions, common problems were parse errors and
-misspelled package names.  The final set contains \CorpusFinishedKaggle
-kernels with \CorpusFinishedKaggleCodeRnd lines of R code.
-
-\subsection{Threats to Validity}
-The primary threat to validity is the issue of
-code coverage, a cause of concern for any dynamic approach. We mitigate this
-by including focusing on R packages that have high code coverage. We turn
-off the bytecode compiler for this study. The bytecode compiler can also
-call \eval. We do not get source locations for \UndefinedEvalsRnd eval
+To analyze a package, we need client code that invokes its methods.
+There are three sources of built-in runnable code that come with each
+CRAN package: \emph{tests, examples} and \emph{vignettes}. They are,
+respectively, traditional unit tests, code snippets from the
+documentation, and long-form use-cases written in Rmarkdown. Examples
+and vignettes are automatically extracted and turned into scripts,
+their input is bundled with the package. Most tests had to be discared
+due to a limitation of our pipeline: the \c{testthat} harness uses
+\eval and thus causes the entire test to register as an \eval call.
+The selected packages are bundled with \CorpusPackagePrograms
+programs; \CorpusExamplesProgramsRnd examples and
+\CorpusVignettesProgramsRnd vignettes.
+
+\medskip\noindent\emph{Kaggle Scripts.} Kaggle is an online platform
+for data-science and machine-learning. The website allows people to
+submit data-analysis problems, users compete to find the best
+solution. The solutions are uploaded to the platform as either plain
+scripts or notebooks. We chose one of the most popular competition,
+predicting the survival of passengers the
+Titanic\footnote{\url{https://www.kaggle.com/c/titanic}}. Unlike CRAN,
+Kaggle is not curated. After downloading the \CorpusKaggle solutions
+and extracting the R code, we found that \CorpusDuplicatedKaggle were
+duplicates. From the remaining \CorpusRunnableKaggle solutions,
+\CorpusFailedKaggle failed to execute. Next to various runtime
+exceptions, common problems were parse errors and misspelled package
+names. The final set contains \CorpusFinishedKaggle programs
+implemented in \CorpusFinishedKaggleCodeRnd lines of R code.
+
+\subsection{Infrastructure}
+
+The results presented in this paper are the result of an automated
+analysis pipeline that acquires the code of packages, extract
+metadata, executes programs, traces their behavior and summarizes the
+observations. Figure~\ref{fig:pipeline} shows the main steps of the
+pipeline along with approximate time to execute each step, the data
+size, and the number of elements manipulated by the stage. Timings are
+for runs on an Intel Xeon 6140, 2.30GHz with 72 cores and 256GB of
+RAM.
+
+\begin{figure}[!h]\hspace{-5mm}\includegraphics[width=1.05\linewidth]
+{pipeline.pdf}\caption{Analysis Pipeline}\label{fig:pipeline}
+\end{figure}
+
+The first step in the pipeline consist of downloading CRAN packages
+along with their dependencies and acquiring Kaggle programs with the
+help of a web crawler. The second step, is to compute code size and
+coverage metrics for the CRAN packages (we use the \covr package for
+coverage). The third step consist in extracting runnable programs from
+packages: i.e. the tests, examples and vignettes. Each extracted
+program is wrapped into a call to our dynamic analyzer --- the tool is
+called \emph{evil} for \underline{ev}al \underline{i}nspection
+\underline{l}ibrary. This step is needed to ensure that we record
+\eval usage only for the target package. Without this, the data would
+include eval calls from the unit testing frameworks as well as from
+bootstrapping R virtual machine itself. To avoid any interference,
+each program is run in its own R instance.
+The fourth step in the pipeline is to perform dynamic analysis for
+each run of a CRAN package or Kaggle program.
+
+The dynamic analyzer builds upon the dynamic analysis framework,
+\instrumentr that we have implemented to enable us to write dynamic
+analysis logic in R. \instrumentr serves as an intermediary between
+\rdyntrace and \evil, it intercepts the hooks exposed by \rdyntrace
+and attaches R functions exported by \evil as callbacks. The \evil
+callbacks execute on corresponding interpreter events.
+
+The data extracted by \evil from each program is
+concatenated, cleaned and summarized in the post-processing phase by
+custom R scripts. Finally, the summarized data is analyzed in
+RMarkdown notebooks to gather insights. Apart from the figures, the
+data points included in the paper are also generated by RMarkdown
+notebooks as latex macros.
+
+The \emph{evil} framework is implemented as a R package in 2K lines of
+R and 400 lines of C++ code. \instrumentr is an R package implemented
+in 2.5K lines of R and 6K lines of C++ code. It internally uses a
+modified R interpreter, \rdyntrace~\cite{oopsla19a}, that exposes
+hooks from within the interpreter implementation for events of
+interest.
+
+All steps of this pipeline are parallelized using GNU
+parallel~\cite{GNUparallel} and orchestrated by GNU make. To schedule
+and parallelize extraction and analysis of programs, we use the \runr
+package. Furthermore, \runr gracefully handles and reports failures
+across large-scale program runs which greatly aids debugging of the
+analysis pipeline.
+
+\medskip\noindent\emph{Limitations.}
+The primary threat to validity is the issue of code coverage, a cause
+of concern for any dynamic approach. We mitigate this by including
+focusing on R packages that have high code coverage. We turn off the
+bytecode compiler for this study. The bytecode compiler can also call
+\eval. We do not get source locations for \UndefinedEvalsRnd \eval
 calls. In these cases \eval is either passed as an argument to a
 higher-order functions or is defined in a function returned by a
-higher-order function and the R parser does not retain location information
-for \eval. However, this is a meager \PercentUndefinedEval of all eval calls
-and is unlikely to affect our analysis. We ignore calls to the native \eval
-function exposed by R. We also ignore the \c{rlang::tidy_eval} function
-which uses native \eval internally because \c{rlang} is used to implement a
-DSL for data analysis in R. It introduces a new first-class promise object
-called \c{quosure} for which it implements special evaluation support in
-\c{tidy_eval}.
+higher-order function and the R parser does not retain location
+information for \eval. However, this is a meager \PercentUndefinedEval
+of all \eval calls and is unlikely to affect our analysis. We ignore
+calls to the native \eval function exposed by R. We also ignore the
+\c{rlang::tidy_eval} function which uses native \eval internally
+because \c{rlang} is used to implement a DSL for data analysis in R.
+It introduces a new first-class promise object called \c{quosure} for
+which it implements special evaluation support in \c{tidy_eval}.
 
@@
 
-This section presents the results of our empirical study of \eval in R in terms
-of four research questions.
+This section presents the results of our empirical study of \eval in R
+in terms of four research questions.
 

### 2020-11-27 inst/paper/main.tex
Environments

--- 2020-11-27 inst/paper/main.tex
@@
 reference.
+
+\textit{Environments.} In R, environments can be accessed, manipulated and created at will.
+Each function has a corresponding environment, which inherits from its calling function.
+In the scope of the function, it can be accessed with \c{environment()}.
+A new environment, which by default inherits from the current environment, is created with \c{new.env}.
+Contrary to other values, environments in R follow a \emph{reference semantics}, not a \emph{copy semantics}.
+The chain of inheriting environments can be climbed up with \c{parent.env}, and ultimately, one arrives at \c{emptyenv}, the empty environment.
+Functions such as \c{parent.frame} and \c{sys.frame} also make it possible to access environments further up in the calling stack, up to the global environment \c{.GlobalEnv}, enabling \emph{dynamic scoping}.
+One can also directly read, modify or create new bindings in any environments. Given an environment \c{e}:
+
+\begin{itemize}
+	\item  \c{e$v} or \c{get("v", envir = e)} reads variable \c{v} in \c{e};
+	\item \c{e$v <- 2} or \c{assign("v", 2, envir = e)} writes value 2 in variable \c{v} in \c{e}.
+\c{v} is created if it did not exist in \c{e}.
+%if argument inherits = TRUE, then if v exists in a parent environment, that binding is modified, instead of creating a new one
+\end{itemize}
+
+It means that for instance, one can directly modify a variable in the scope of its calling function, and there is even a special assignment operator for that purpose, \c{<<-}, \ie \c{v <<- 2} will search for a definition of \c{v} in all the parent environments, redefine the value, or create a new definition in the global environment.
+Explicit environments are used in R as hashtables, or to maintain state across function calls, or to create package namespaces.
 

### 2020-11-27 inst/paper/main.tex
minor

--- 2020-11-27 inst/paper/main.tex
@@
 
-\medskip\noindent\emph{Limitations.}
-The primary threat to validity is the issue of code coverage, a cause
-of concern for any dynamic approach. We mitigate this by including
-focusing on R packages that have high code coverage. We turn off the
+\medskip\noindent\emph{Limitations.} Dynamic analysis can only observe
+calls that are triggered by the program's input. We believe that
+focusing on R packages with high code coverage does mitigate this to
+some extent. The results we report here were obtained with R's
+bytecode compiler turned off, this should not affect the results as
+the compiler does not optimize \eval.
+
+. We turn off the
 bytecode compiler for this study. The bytecode compiler can also call

### 2020-11-27 inst/paper/main.tex
move

@@
 reference.
-
-\textit{Environments.} In R, environments can be accessed, manipulated and created at will.
-Each function has a corresponding environment, which inherits from its calling function.
-In the scope of the function, it can be accessed with \c{environment()}.
-A new environment, which by default inherits from the current environment, is created with \c{new.env}.
-Contrary to other values, environments in R follow a \emph{reference semantics}, not a \emph{copy semantics}.
-The chain of inheriting environments can be climbed up with \c{parent.env}, and ultimately, one arrives at \c{emptyenv}, the empty environment.
-Functions such as \c{parent.frame} and \c{sys.frame} also make it possible to access environments further up in the calling stack, up to the global environment \c{.GlobalEnv}, enabling \emph{dynamic scoping}.
-One can also directly read, modify or create new bindings in any environments. Given an environment \c{e}:
-
-\begin{itemize}
-	\item  \c{e$v} or \c{get("v", envir = e)} reads variable \c{v} in \c{e};
-	\item \c{e$v <- 2} or \c{assign("v", 2, envir = e)} writes value 2 in variable \c{v} in \c{e}.
-\c{v} is created if it did not exist in \c{e}.
-%if argument inherits = TRUE, then if v exists in a parent environment, that binding is modified, instead of creating a new one
-\end{itemize}
-
-It means that for instance, one can directly modify a variable in the scope of its calling function, and there is even a special assignment operator for that purpose, \c{<<-}, \ie \c{v <<- 2} will search for a definition of \c{v} in all the parent environments, redefine the value, or create a new definition in the global environment.
-Explicit environments are used in R as hashtables, or to maintain state across function calls, or to create package namespaces.
 
@@
 
-The second substantial difference is the control that R gives over the
-choice of environment in which the AST is evaluated in. Beside
-traditional choices such as the top level or current lexical
-environment, R allows programmers to choose a fresh, empty,
-environment or any environment on the call stack. This is a
-significant increase in expressive power as functions that are
-otherwise not aware of the use of \eval may be deeply impacted.
+\medskip\noindent\textit{Environments.} In R, environments can be
+accessed, manipulated and created at will. Each function has a
+corresponding environment, which inherits from its calling function.
+In the scope of the function, it can be accessed with
+\c{environment()}. A new environment, which by default inherits from
+the current environment, is created with \c{new.env}. Contrary to
+other values, environments in R follow a \emph{reference semantics},
+not a \emph{copy semantics}. The chain of inheriting environments can
+be climbed up with \c{parent.env}, and ultimately, one arrives at
+\c{emptyenv}, the empty environment. Functions such as
+\c{parent.frame} and \c{sys.frame} also make it possible to access
+environments further up in the calling stack, up to the global
+environment \c{.GlobalEnv}, enabling \emph{dynamic scoping}. One can
+also directly read, modify or create new bindings in any environments.
+Given an environment \c{e}:
+
+\begin{itemize}
+	\item  \c{e$v} or \c{get("v", envir = e)} reads variable \c{v} in \c{e};
+	\item \c{e$v <- 2} or \c{assign("v", 2, envir = e)} writes value 2 in variable \c{v} in \c{e}. \c{v} is created if it did not exist in \c{e}.
+\end{itemize}
+
+It means that for instance, one can directly modify a variable in the
+scope of its calling function, and there is even a special assignment
+operator for that purpose, \c{<<-}, \ie \c{v <<- 2} will search for a
+definition of \c{v} in all the parent environments, redefine the
+value, or create a new definition in the global environment. Explicit
+environments are used in R as hashtables, or to maintain state across
+function calls, or to create package namespaces.
 

### 2020-11-28 inst/paper/main.tex
minor

--- 2020-11-27 inst/paper/main.tex
@@
 
-The second substantial difference is the control that R gives over the
-choice of environment in which the AST is evaluated in. Beside
-traditional choices such as the top level or current lexical
-environment, R allows programmers to choose a fresh, empty,
-environment or any environment on the call stack. This is a
-significant increase in expressive power as functions that are
-otherwise not aware of the use of \eval may be deeply impacted.
+\medskip\noindent\textit{Environments.} In R, environments can be
+manipulated at will. Each function has an environment that can be
+accessed with \c{environment()}. A new environment, which by default
+inherits from the current environment, is created with \c{new.env}.
+Contrary to other values, environments have reference semantics rather
+than copy semantics. The chain of inheriting environments can be
+traversed with \c{parent.env}, ultimately one arrives at \c{emptyenv}.
+\c{parent.frame} and \c{sys.frame} grant access to environments
+further up in the calling stack, up to the global environment
+\c{.GlobalEnv}, and packages. One can also directly read, modify or
+create new bindings in any environments. Given an environment \c{e}:
+
+\begin{itemize}
+\item  \c{e$v} or \c{get("v", envir = e)} reads variable \c{v} in \c{e};
+\item \c{e$v <- 2} or \c{assign("v", 2, envir = e)} writes value 2 in variable \c{v} in \c{e}. \c{v} is created if it did not exist in \c{e}.
+\end{itemize}
+\noindent One can directly modify a variable in the scope of its
+calling function, and there is even a special assignment operator for
+that purpose; \c{v <<- 2} will search for \c{v} in parent
+environments, redefine the value, or create a definition in the
+current environment. Environments are used as hashtables, to maintain
+state across function calls, or to create package namespaces.
 

### 2020-11-28 inst/paper/main.tex
cites to Moller

@@
 Most dynamic languages provide their users with a facility to
-transform unstructured text into executable code, and to evaluate that
-code. In this paper we refer to this reflective facility of a language
-as \eval bowing to its origins in LISP in 1956~\cite{lisp}. \Eval has
-been much maligned over the years. In computing lore, it is as close
-to a boogeyman as it gets. Yet, for McCarthy, \eval was simply the way
-to write down the definition of LISP, he was surprised that someone
-coded it up and offered it to end users.  Since then, reflective
-facilities have been used to parameterize programs over code patterns
-that can be provided after the program is written.  The presence of
-such a feature in a language is a hallmark of dynamism as it is a form
-of delayed binding as what any particular call to \eval will do is
-only know at program run-time when that particular call site is
+transform unstructured text into executable code and evaluate that
+code. We refer to this reflective facility as \eval bowing to its
+origins in LISP, all the way back in 1956~\cite{lisp}. \Eval has been
+much maligned over the years. In computing lore, it is as close to a
+boogeyman as it gets. Yet, for McCarthy, \eval was simply the way to
+write down the definition of LISP, he was surprised that someone coded
+it up and offered it to end users. Since then, reflective facilities
+have been used to parameterize programs over code patterns that can be
+provided after the program is written. The presence of such a feature
+in a language is a hallmark of dynamism; it is a form of delayed
+binding as the behavior of any particular call to \eval will only be
+known when the program is run and that particular call site is
 evaluated.
@@
 \vspace{2mm}\noindent\emph{Trouble in Paradise.} Reflective facilities
-hinder most attempt to reason about the code using them, as well as to
-apply meaning-preserving program transformations. In practice this
-means that static analysis techniques will loose so much precision as
-to become pointless and that anything but the most trivial, local,
-compiler optimizations will be unsound. Furthermore the addition of
-new code --- code that could have been obtained from a network
-connection --- as the program is running is a security vulnerability
-waiting to happen.
-
-To further illustrate the challenges at hand, consider the interaction
-between a static analysis tool and a dynamic language. An abstract
-interpretation-based program analyzer computes an over-approximation
-of the set of possible behaviors exhibited by the system under
-study~\cite{cc77}. The presence of a reflective operation entails that
-the operation may have \emph{any} behavior that can be expressed in
-the target language; i.e. as if we replaced it by any legal sequence
-of instructions. As dynamic languages tend to be permissive, the
-analysis has to, for example, assume that many (and sometimes all)
-functions in scope may have been redefined, e.g. that \texttt{`+`} now
-opens an Internet connection or something equally surprising. This
-means that a single occurrence of \eval causes the static analyzer to
-loose all information about the program state and the meaning of
-identifiers. This loss of precision can sometimes be mitigated by
-analyzing the string passed to \eval~\cite{} to bound its possible
-behavior but when the string comes from outside the program not much
-can be done. A frustrated group researchers argued giving up on
-soundness and, instead, under-approximating dynamic features
+hinder most attempts to reason about, or apply meaning-preserving
+transformation to, the code using them. In practice, \eval causes
+static analysis techniques to loose so much precision as to become
+pointless. For compilers, anything but the most trivial, local,
+optimizations are unsound after a use of \eval. Furthermore, the
+addition of arbitrary code --- code that could have been obtained from
+a network connection --- as a program is running is a security
+vulnerability waiting to happen. To illustrate these challenges,
+consider the interaction of a static analysis tool with a dynamic
+language. An abstract interpretation-based program analyzer computes
+an over-approximation of the set of possible behaviors exhibited by
+the program under study~\cite{cc77}. A reflective may have \emph{any}
+behavior that can be expressed in the target language; i.e. \eval can
+be replaced by any legal sequence of instructions. As dynamic
+languages tend to be permissive, the analysis has to, for example,
+assume that many (or all) functions in scope may have been redefined,
+e.g. that \texttt{`+`} now opens a network connection or something
+equally surprising. A single occurrence of \eval causes the static
+analyzer to loose all information about program state and
+meaning of identifiers. This loss of precision can sometimes be
+mitigated by analyzing the string argument~\cite{moller03} to bound its
+possible behavior but when the string comes from outside the program
+not much can be done. A frustrated group researchers argued giving up
+on soundness and, instead, under-approximating dynamic features
 (soundiness)~\cite{soundy}. In their words ``a practical analysis,
@@
 
-\vspace{2mm}\noindent\emph{The here and now.}  In this study, we
+\vspace{2mm}\noindent\emph{The Here and Now.} In this study, we
 investigate the usage of \eval in programs written in the R
 programming language. R is language designed by statisticians for
-applications in data science~\cite{r,R96}. What makes looking at R after
-JavaScript interesting is that, while both languages are dynamic, they
-are quite different. While one can program in an object-oriented style
-in R, like in JavaScript, R is primarily a lazy, untyped, functional
-language.  JavaScript was designed to run untrusted code in browser,
-while R is used for statistical computing on desktops. JavaScript is a
-general purpose language used by a wide community of programmers;
-while R is used for scientific computing by data scientists and domain
-experts with, often, limited programming experience. One can
-distinguish between library implementers, developers with some
-programming experience and a working knowledge of R, and end-users,
-who are typically not expert programmers and often have only a cursory
-knowledge of the language.\footnote{Consider that R evaluates function
-argument lazily, just like Haskell. We informally surveyed end-users,
-including computer scientists, and did not find a single user aware of
-this fact. Library developers, on the other hand, know about laziness
-and program defensively around it.} Our goal is thus to highlight the
-differences in usage between JavaScript and R, and try to explain
-those differences in terms of language features, application domain
-and programmer experience.
+applications in data science~\cite{r,R96}. What makes looking at R
+after JavaScript interesting is that, while both languages are
+dynamic, they are quite different. While one can program in an
+object-oriented style in R, like in JavaScript, R is primarily a lazy,
+untyped, functional language. JavaScript was designed to run untrusted
+code in browser, while R is used for statistical computing on
+desktops. JavaScript is a general purpose language used by a wide
+community of programmers; while R is used for scientific computing by
+data scientists and domain experts with, often, limited programming
+experience. One can distinguish between library implementers,
+developers with some programming experience and a working knowledge of
+R, and end-users, who are typically not expert programmers and often
+have only a cursory knowledge of the language.\footnote{Consider that
+R evaluates function argument lazily, just like Haskell. We informally
+surveyed end-users, including computer scientists, and did not find a
+single user aware of this fact. Library developers, on the other hand,
+know about laziness and program defensively around it.} Our goal is
+thus to highlight the differences in usage between JavaScript and R,
+and try to explain those differences in terms of language features,
+application domain and programmer experience. Hopefully some of our
+observations will generalize to other languages.
 
@@
 base that we can analyze dynamically. To observe \eval we built a
-two-level monitoring infrastructure: we can monitor R programs by
-instrumentation --- this gives us access to many user-visible properties
-of R programs --- but we can also monitor the inner working of the R
-interpreter --- this allows us to capture details of execution that are
-not exposed at the source level. Dynamic analysis is of course limited
-in that it can only observe behaviors that are triggered by the
-particular inputs passed to the program. Luckily, CRAN libraries come
-with multiple input datasets and use-cases.
+two-level monitoring infrastructure:\footnote{Our infrastructure is
+open source and publicly available, our anonymized code is at {\tt
+  http://url.com}, a more complete artifact will be submitted to the
+artifact evaluation committee. } we can monitor R programs by
+instrumentation --- this gives us access to many user-visible
+properties of R programs --- but we can also monitor the inner working
+of the R interpreter --- this allows us to capture details of
+execution that are not exposed at the source level. Dynamic analysis
+is of course limited in that it can only observe behaviors that are
+triggered by the particular inputs passed to the program. Luckily,
+CRAN libraries come with multiple input datasets and use-cases.
 
@@
 
-Our infrastructure is publicly available, released in open source, and
-our code, and data will be submitted for artifact evaluation.
+\vspace{2mm}\noindent\emph{Why do we Eval?} {\bf A short summary of the
+results should go here.}
 
@@
 
-The expressive power of \eval depends on a number of choices made by
-the programming language designers. The two axes that are the most
-relevant here are:
+The expressive power of \eval depends on choices made by the language
+designers along the following two axes:
 \begin{itemize}
 \item {\bf Scoping:} What environment does the string passed to \eval
-  execute in? JavaScript and R evaluate the expression in the current
-  scope. This means that local variables and parameters are exposed to
-  \eval, breaking the abstraction boundary of the enclosing function.
-  Julia takes a more restrictive approach by running \eval at the
-  ``top level'', i.e. in the global environment, outside of any
-  function's environment. JavaScript has a global \eval that behaves
-  like Julia, it also has a ``strict mode'' in which \eval is allowed
-  to access local variables, but fobidden from injecting new variables
-  in the local environment. In Java, one can implement \eval but the
+  execute in? JavaScript and R evaluate it in the current environment,
+  thus exposing local variables and parameters to \eval and breaking
+  the abstraction boundary of the enclosing function. Julia is more
+  restrictive, it runs \eval at the ``top level'', in the global
+  environment. JavaScript has a global \eval that behaves like Julia,
+  it also has a ``strict mode'' in which \eval is allowed to access
+  local variables, but fobidden from injecting new variables in the
+  local environment. In Java, one can implement \eval but the
   expression will execute in new environment.\footnote{While \eval
-  used to be available only in interpreted languages, the advent of
-  just-in-time compilation gradually lifted this restriction. To
-  implement \eval in Java requires wrapping its input in a static
-  method of an anonymous class, generating bytecode for that class,
-  invoking the class loader to install the code, and then using
-  reflection to call the new method. }
+  used the purview of interpreted languages, the advent of
+  just-in-time compilation lifted this restriction. To implement \eval
+  in Java requires wrapping its input in a static method of a new
+  anonymous class, generating bytecode for that class, invoking the
+  class loader to install that code, and finally reflectively calling
+  the new method. }
 \item {\bf Reflective API:} What other reflective operations are
-  exposed to user code? JavaScript does provides few reflective
-  functions (other than being able to enumerate the properties of an
-  object and to access properties with strings), the string passed to
-  \eval is thus limited to what can be expressed in JavaScript. R, on
-  the other hand, has a rich reflective interface. For example, R
-  allows user code to walk the call stack and arbitrarily add and
-  delete local variables in any environment.
+  exposed to user code? JavaScript provides few reflective functions
+  (other than enumeration of the properties of an object and access to
+  properties with strings), \eval is thus limited to what can be
+  expressed in JavaScript code. R, on the other hand, has a rich
+  reflective interface. For example, R allows user code to walk the
+  call stack and arbitrarily add and delete local variables in any
+  environment. This reflective API means that there are few limits to
+  what can do.
 \end{itemize}
@@
 \noindent
-These design choices impact our ability to reason about code. More
-expressive power for \eval translates directly into additional
-restriction for program analysis and transformation tools. The reason
-Julia has restricts \eval to execute in top level environment is that
-they wanted to preserve their compiler's ability to optimize
-functions~\cite{oopsla18a}. What they wanted to prevent was the case
-where \eval changed the type of local variables as this would require
-recompilation of the method (or prevent optimization). Julia even adds
-a versoning mechanism, called world age, to ensure that code added
-during an \eval does not invalidate optimizations such as
-inlining~\cite{oopsla20a}. More permissive languages such as R are
-much harder to deal with from a compiler's perspective.
+These design choices impact our ability to reason about code when
+reflection is used. A more expressive \eval translates directly into
+additional restrictions for program analysis and transformation tools.
+The reason Julia restricts \eval to the top level environment is to
+preserve the compiler's ability to optimize
+functions~\cite{oopsla18a}. What the Julia designers wanted was to
+prevent \eval from modifying the type of local variables as this would
+require recompilation of the method. Julia even adds a versoning
+mechanism, called world age, to ensure that code added during an \eval
+does not invalidate optimizations such as inlining~\cite{oopsla20a}.
+More permissive languages such as R are much harder to deal with from
+a compiler's perspective.
 
 \medskip\noindent{\it Eval in R.} There are four reflective functions
-that take abstract syntax trees (called expressions), parse it as
-code, and evaluate it. These functions are \c{eval}, \c{evalq},
-\c{eval.parent}, and \c{local}. Unlike other languages, R's \eval does
-not take a string, but rather an expression. As this is slightly more
-involved, it may discourage some casual users. There are two ways to
-obtain an expression. Given a string, the \c{parse()} function retuns
-an unevaluated expression. The other way is to use the function
+that take abstract syntax trees (called expressions), parse them as
+code, and evaluate the result. These functions are \c{eval},
+\c{evalq}, \c{eval.parent}, and \c{local}. Taking an expression rather
+than a string, may discourage some casual users as manipulating
+expressions is slightly more involved.\footnote{The rationale for
+\eval taking expressions is explained by the main use case for \eval,
+namely extending the language by reinterpretation of promises; R gives
+access to the code of a promise as an expression.} There are two ways
+to obtain an expression. Given a string, the \c{parse()} function
+retuns an unevaluated expression. The other way is to use the function
 \c{substitute(e,envir)}, it yields the parse tree of the expression
@@
 scope, many patterns could be replaced by more disciplined
-code~\cite{oopsla12b}.  The work did not measure code coverage, so the
-numbers presented are a lower bound on the possible behaviors.
-Furthermore, JavaScript usage in 2011 is likely different from today,
-e.g. Node.js was not covered by Richards.  More details about dynamic
-analysis of JavaScript can be found in~\cite{liang}.
+code~\cite{oopsla12b, moller12}. The work did not measure code
+coverage, so the numbers presented are a lower bound on the possible
+behaviors. Furthermore, JavaScript usage in 2011 is likely different
+from today, e.g. Node.js was not covered by Richards. More details
+about dynamic analysis of JavaScript can be found in~\cite{liang}.
 

### 2020-11-28 inst/paper/main.tex
intro: removing core

@@
 instrumentation --- this gives us access to many user-visible
-properties of R programs --- but we can also monitor the inner working
-of the R interpreter --- this allows us to capture details of
-execution that are not exposed at the source level. Dynamic analysis
-is of course limited in that it can only observe behaviors that are
-triggered by the particular inputs passed to the program. Luckily,
-CRAN libraries come with multiple input datasets and use-cases.
-
-The choice of corpus is crucial for analyses like ours.  Our corpus
-has been constructed to reflect the levels of sophistication of the R
-community.  We distinguish between \emph{core packages} (there are
-\CorpusCorePackages such packages that are distributed together with
-the R virtual machine), \emph{CRAN packages} (\CorpusPackages curated
-packages that pass stringent quality checks and are equipped with
-tests and sample data), \emph{Kaggle scripts} (\CorpusFinishedKaggle
-end-user written programs that performs a particular data analysis
-task). It is reasonable to expect that \eval usage differ between
-these datasets. The core R code is extremely stable, maintained for
-the last 20 years with very few far reaching changes. The libraries
-represent a more lively ecosystem with new libraries added each day.
-Finally, end user code is often one-use scripts thrown together and
-sometimes never revisited.
+properties of R programs --- but we can also monitor the
+inner-workings of the R interpreter --- this allows us to capture
+details not exposed at the source level. Dynamic analysis is limited,
+it can only observe behaviors triggered by the particular inputs
+passed to a program. Luckily, CRAN libraries come with many tests and
+use-cases. The choice of corpus is crucial. Our corpus has been
+constructed to reflect the levels of sophistication of the R
+community. We distinguish between \emph{CRAN packages}
+(\CorpusPackages curated packages that pass stringent quality checks
+and are equipped with tests and sample data) and \emph{Kaggle scripts}
+(\CorpusFinishedKaggle end-user written programs that performs a
+particular data analysis task). It is reasonable to expect that \eval
+usage differ between these datasets: the libraries represent a lively
+ecosystem with new libraries added each day, while end user code is
+often thrown together, run once, and never revisited.
 
@@
 
-\subsection{Infrastructure}
+\subsection{Analysis Pipeline}
 

### 2020-11-29 inst/paper/main.tex
wip

@@
   semantics. The widespread use of these reflective functions hinders
-  static analysis and prevents compilers from performing many
+  static analysis and prevents compilers from performing
   optimizations. This paper aims to provide a better sense of why
-  programmers use \eval.  Understanding how \eval is used in practice
+  programmers use \eval. Understanding why \eval is used in practice
   is key to finding ways to mitigate its negative impact. We have
   reasons to believe that reflective feature usage is language and
-  application domain specific. In this paper, we focus on data science
-  code written in R, and compare our results to previous work that
-  analyzed web programming in JavaScript.  We conduct a large-scale
-  study of a corpus of \CorpusAllCodeRnd lines of libraries and
-  end-user code on which we performed dynamic analysis to confirm that
-  \eval is indeed in widespread use. We found that \eval is more
-  dangerous in some ways, and safer in others, than what was
-  previously reported for JavaScript.
+  application domain specific; we focus on data science code written
+  in R, and compare our results to previous work that analyzed web
+  programming in JavaScript. Dynamic analysis of a corpus of
+  \CorpusAllCodeRnd lines of libraries and end-user code confirms that
+  \eval is indeed in widespread use; R's \eval is more dangerous in
+  some ways, and safer in others, than what was previously reported
+  for JavaScript.
+  %% TODO ::: Add a sentance about summarizing why users resort to
+  %% eval -- Jan
 \end{abstract}
@@
 
-The expressive power of \eval depends on choices made by the language
-designers along the following two axes:
+The expressive power of \eval depends on design decisions along two
+axes:
 \begin{itemize}
-\item {\bf Scoping:} What environment does the string passed to \eval
-  execute in? JavaScript and R evaluate it in the current environment,
-  thus exposing local variables and parameters to \eval and breaking
-  the abstraction boundary of the enclosing function. Julia is more
-  restrictive, it runs \eval at the ``top level'', in the global
-  environment. JavaScript has a global \eval that behaves like Julia,
-  it also has a ``strict mode'' in which \eval is allowed to access
-  local variables, but fobidden from injecting new variables in the
-  local environment. In Java, one can implement \eval but the
-  expression will execute in new environment.\footnote{While \eval
-  used the purview of interpreted languages, the advent of
-  just-in-time compilation lifted this restriction. To implement \eval
-  in Java requires wrapping its input in a static method of a new
-  anonymous class, generating bytecode for that class, invoking the
-  class loader to install that code, and finally reflectively calling
-  the new method. }
+\item {\bf Scoping:} What environment does \eval executes its argument
+  in? JavaScript and R evaluate it in the current environment, thus
+  exposing local variables and parameters and breaking the caller's
+  abstraction boundary. Julia is more restrictive, \eval runs at the
+  ``top level'', in the global environment. JavaScript has a global
+  \eval that behaves like Julia, and a ``strict mode'' in which \eval
+  may access local variables, but fobidden from injecting new ones in
+  the local environment. In Java, one can implement \eval with
+  Julia-semantics.\footnote{While \eval used to be the purview of
+  interpreted languages, just-in-time compilation lifted this
+  restriction. \Eval takes its input, wraps it in a static method of a
+  new anonymous class, generates bytecode for that class, invokes
+  the class loader to install that code, and finally reflectively
+  calls the method. }
 \item {\bf Reflective API:} What other reflective operations are
   exposed to user code? JavaScript provides few reflective functions
-  (other than enumeration of the properties of an object and access to
-  properties with strings), \eval is thus limited to what can be
-  expressed in JavaScript code. R, on the other hand, has a rich
-  reflective interface. For example, R allows user code to walk the
-  call stack and arbitrarily add and delete local variables in any
-  environment. This reflective API means that there are few limits to
-  what can do.
+  (other than enumeration of an object's properties and string-based
+  indexing), the impact of \eval is thus limited. R, on the other
+  hand, has a rich reflective API. For example, R allows user code to
+  walk the call stack and arbitrarily add and delete local variables.
+  This means that there are few limits to the side-effects of an
+  \eval.
 \end{itemize}
@@
 additional restrictions for program analysis and transformation tools.
-The reason Julia restricts \eval to the top level environment is to
-preserve the compiler's ability to optimize
-functions~\cite{oopsla18a}. What the Julia designers wanted was to
-prevent \eval from modifying the type of local variables as this would
-require recompilation of the method. Julia even adds a versoning
-mechanism, called world age, to ensure that code added during an \eval
-does not invalidate optimizations such as inlining~\cite{oopsla20a}.
-More permissive languages such as R are much harder to deal with from
-a compiler's perspective.
-
-\medskip\noindent{\it Eval in R.} There are four reflective functions
+The reason Julia restricts \eval to the top level is to preserve the
+compiler's ability to optimize code~\cite{oopsla18a}. In particular,
+\eval had to be prevented from changing the type of local variables as
+this would require recompilation of the method. Julia even adds a
+versoning mechanism, called world age, to ensure that code added
+during an \eval does not invalidate inlining
+optimizations~\cite{oopsla20a}. More permissive languages such as R
+are much harder to deal with from a compiler's perspective.
+
+\medskip\noindent{\it Eval in R.} The reflective interface exposed by
+R is somewhat complex. The core library exports four functions with
+slightly different semantics, \c{eval}, \c{evalq}, \c{eval.parent},
+and \c{local}. \Eval is the more general function, and, unlike in
+JavaScript, it takes three arguments, an expression, its environment
+and the enclosing environment. The following discussion simplify some
+details which are not relevant to this paper. For the interested
+reader we recommend the excellent book by Hadley
+Wickham~\cite{hadley}. The definition of \eval starts as follows:
+
+\begin{lstlisting}
+ eval
+   <- function(expr,
+                envir = parent.frame(),
+                enclos = if(is.list(envir)) parent.frame() else baseenv())
+          { ...body... }
+\end{lstlisting}
+The parameters are \c{expr}, the value to be evaluated, \c{envir}, the
+environment in which evaluation happens, and \c{enclos}, the
+environment to look up objects not found in \c{envir} (by default
+either the top level, \c{baseenv} or the environment of the caller).
+The \c{expr} parameter can take values of different type, for our
+purposes we focus on the most common, {\tt expression}s. It is easiest
+to think of an \c{expression} as an abstract syntax tree as returned
+by R's parser. The simplest way to create an \c{expression} is to call
+\c{quote} passing some R expression:
+\begin{lstlisting}
+  > quote(a + b)
+  # a + b
+\end{lstlisting}
+Here the return value is an abstract syntax tree. From a
+string, use  the \c{parse} function:
+\begin{lstlisting}
+  > parse(text="a+b")
+  # expression(a+b)
+\end{lstlisting}
+
+There are four reflective functions
 that take abstract syntax trees (called expressions), parse them as
@@
 function takes three arguments:
-
-\begin{lstlisting}
- eval(expr,
-      envir = parent.frame(),
-      enclos = if(is.list(envir) || is.pairlist(envir))
-               parent.frame() else baseenv())
-\end{lstlisting}
-
-\noindent
-This code is part of the definition of \eval, it specifies that the
-function takes an \c{expr} which is the value to be evaluated,
-\c{envir} is the environment in which evaluation happes, and
-\c{enclos} is relevant when \c{envir} is a list or a data frame, then
-it specifies where R looks for objects not found in \c{envir}. This
-can be \c{NULL} (interpreted as the base package environment,
-\c{baseenv()}) or an environment. Thus, the expression \c{expr}
-evaluates in the environment \c{envir} and returns the computed value.
-If \c{envir} is missing, then the default is \c{parent.frame()} (the
-environment where the call to \eval was made). Values to be evaluated
-can be of types \c{call} (function calls) or \c{expression} (arbitrary
-expressions), \c{name} (when the symbol is looked up in the current
-scope and its binding is evaluated), promise, or any of the basic
-types such as vectors and environments (which are returned unchanged).
 

### 2020-11-30 inst/paper/main.tex
background

@@
 \end{lstlisting}
-
-There are four reflective functions
-that take abstract syntax trees (called expressions), parse them as
-code, and evaluate the result. These functions are \c{eval},
-\c{evalq}, \c{eval.parent}, and \c{local}. Taking an expression rather
-than a string, may discourage some casual users as manipulating
-expressions is slightly more involved.\footnote{The rationale for
-\eval taking expressions is explained by the main use case for \eval,
-namely extending the language by reinterpretation of promises; R gives
-access to the code of a promise as an expression.} There are two ways
-to obtain an expression. Given a string, the \c{parse()} function
-retuns an unevaluated expression. The other way is to use the function
-\c{substitute(e,envir)}, it yields the parse tree of the expression
-\c{e} after performing substitutions defined by the bindings in the
-\c{envir} parameter.
-
+The most common way to create an expression is to extract it from an
+argument. Each argument comes packaged inside a promise which retains
+the source code of the argument. Consider the following function
+definition, \c{f} is a function with a single parameter \c{x}. When
+the call \c{f(a+b)} is evaluated, \c{a+b} is stored inside a newly
+created promise.
 \begin{lstlisting}
- > substitute(expression(a+b), list(a=1)))
- expression(1+b)
+  > f <- function(x) substitute(x), list(a=1)))
+  > f(a+b)
+  # expression(1+b)
 \end{lstlisting}
-
-\noindent
-This function is widely used in conjunction with \eval, as it can
-leverage R's laziness for metaprogramming. R allows programmatic
-manipulation of parse trees, which are themselves first class objects.
-Expressions are evaluated using \eval (or one its variants). The \eval
-function takes three arguments:
-
-The \c{evalq} form is equivalent to \c{eval(quote(expr), ...)}.
-Quoting that argument prevents if from being evaluated in the current
-environment. \c{eval.parent(expr, n)} is a shorthand for \c{eval(expr,
-  parent.frame(n))}. \c{local} evaluates an expression in a local
-environment. It is equivalent to evalq except that its default
-argument creates a new, empty environment. This is useful to create
-anonymous recursive functions and as a kind of limited namespace
-feature since variables defined in the environment are hidden.
-
-\medskip\noindent\textit{Environments.} In R, environments can be
-manipulated at will. Each function has an environment that can be
-accessed with \c{environment()}. A new environment, which by default
-inherits from the current environment, is created with \c{new.env}.
-Contrary to other values, environments have reference semantics rather
-than copy semantics. The chain of inheriting environments can be
-traversed with \c{parent.env}, ultimately one arrives at \c{emptyenv}.
+The call to \c{substitute(x)} extracts, from the promise bound to
+\c{x}, the expression corresponding to that argument, i.e., \c{a+b}.
+The general form of the function takes two arguments, the second
+argument is an object that can be used as an environment. This can be
+either a list of named values, a data frame or an actual environment.
+\begin{lstlisting}
+  > environment()
+  # <environment: R_GlobalEnv>
+\end{lstlisting}
+The above shows how to access the current environment, this can belong
+to the current function or, as here, the top level. Environments are
+nested, with each environment having a parent. This nesting is used
+when looking up names. When calling \c{new.env}, the default parent is
+the current environment. The chain of environments can be traversed
+with \c{parent.env}, ultimately one arrives at \c{emptyenv}.
 \c{parent.frame} and \c{sys.frame} grant access to environments
@@
 \c{.GlobalEnv}, and packages. One can also directly read, modify or
-create new bindings in any environments. Given an environment \c{e}:
-
+create new bindings, given any environment:
 \begin{itemize}
-\item  \c{e$v} or \c{get("v", envir = e)} reads variable \c{v} in \c{e};
-\item \c{e$v <- 2} or \c{assign("v", 2, envir = e)} writes value 2 in variable \c{v} in \c{e}. \c{v} is created if it did not exist in \c{e}.
+\item \c{env$v} and \c{get("v", envir = env)} read variable \c{v} in
+  \c{e};
+\item \c{env$v <- 2} and \c{assign("v", 2, envir = env)} write 2 in
+  \c{v}. If it did not exist, \c{v} is created in \c{env}.
 \end{itemize}
-\noindent One can directly modify a variable in the scope of its
-calling function, and there is even a special assignment operator for
-that purpose; \c{v <<- 2} will search for \c{v} in parent
-environments, redefine the value, or create a definition in the
-current environment. Environments are used as hashtables, to maintain
-state across function calls, or to create package namespaces.
+\noindent Environments are often used as hashtables by programmers as
+they have reference semantics (other are copy-on-write) and have a
+built-in string look up. across function calls, or to create package
+namespaces.
+\begin{lstlisting}
+  evalq <- function(exp,env,enclos) eval(quote(exp),env,enclos)
+  eval.parent <- function(exp,n) eval(exp,parent.frame(n))
+  local <- function(exp,env,enclos) evalq(exp,new.env())
+\end{lstlisting}
+The above code show simplified implementation for the variants of
+\eval. The \c{evalq} form simply quotes the argument to prevent it
+from being evaluated in the current environment. The
+\c{eval.parent(e,n)} form evaluates \c{e} in the environment of the
+  {\tt n}-th caller up the call stack. Finally \c{local} evaluates an
+  expression in a new environment to avoid polluting the current
+  environment.
+
+%% Taking an expression rather
+%% than a string, may discourage some casual users as manipulating
+%% expressions is slightly more involved.\footnote{The rationale for
+%% \eval taking expressions is explained by the main use case for \eval,
+%% namely extending the language by reinterpretation of promises; R gives
+%% access to the code of a promise as an expression.}
 

### 2020-11-30 inst/paper/main.tex
sec 2.3

@@
 The above shows how to access the current environment, this can belong
-to the current function or, as here, the top level. Environments are
+to the current function or, as above, the top level. Environments are
 nested, with each environment having a parent. This nesting is used
@@
 \begin{itemize}
-\item \c{env$v} and \c{get("v", envir = env)} read variable \c{v} in
-  \c{e};
-\item \c{env$v <- 2} and \c{assign("v", 2, envir = env)} write 2 in
-  \c{v}. If it did not exist, \c{v} is created in \c{env}.
+\item \c{env$v} and \c{get("v",envir=env)} read variable \c{v} in
+  environment \c{env};
+\item \c{env$v<-2} and \c{assign("v",2,envir=env)} write 2 in
+  \c{v}. If not found, \c{v} is created in \c{env}.
 \end{itemize}
@@
 \c{eval.parent(e,n)} form evaluates \c{e} in the environment of the
-  {\tt n}-th caller up the call stack. Finally \c{local} evaluates an
+  {\tt n}-th caller up the call stack. Finally, \c{local} evaluates an
   expression in a new environment to avoid polluting the current
@@
 
-%% Taking an expression rather
-%% than a string, may discourage some casual users as manipulating
-%% expressions is slightly more involved.\footnote{The rationale for
-%% \eval taking expressions is explained by the main use case for \eval,
-%% namely extending the language by reinterpretation of promises; R gives
-%% access to the code of a promise as an expression.}
+\subsection{Eval Usage in R}
+
+The R language was intended to be extensible, the combination of lazy
+evaluation, \c{substitute} and \c{eval} are the tools given to
+developers to this end. This API is slightly more complex than just
+passing a string, it is conceivable that this may discourage some
+casual users. \Eval is also being used to reduce boilerplate code and
+provide convenience features for programmers. We now give some
+representative examples of its usage.
 

### 2020-11-30 inst/paper/main.tex
usage example

@@
   # expression(1+b)
+  > substitute(a+b,list(a=1,b=2))
+  # 1+2
 \end{lstlisting}
 The call to \c{substitute(x)} extracts, from the promise bound to
-\c{x}, the expression corresponding to that argument, i.e., \c{a+b}.
-The general form of the function takes two arguments, the second
-argument is an object that can be used as an environment. This can be
-either a list of named values, a data frame or an actual environment.
+\c{x}, the expression passed in the call to \c{f}, i.e., \c{a+b}. The
+\c{substitute(exp,env)} function takes two arguments, the second is an
+object that can be used as an environment. This can be either a list
+of named values, a data frame or an actual environment. \c{substitute}
+will look for occurrences of each symbol from \c{env} in \c{exp} and
+replace them with their value taken from \c{env}.
 \begin{lstlisting}
@@
 \end{lstlisting}
-The above code show simplified implementation for the variants of
-\eval. The \c{evalq} form simply quotes the argument to prevent it
-from being evaluated in the current environment. The
-\c{eval.parent(e,n)} form evaluates \c{e} in the environment of the
-  {\tt n}-th caller up the call stack. Finally, \c{local} evaluates an
-  expression in a new environment to avoid polluting the current
-  environment.
+The three \eval variants can be expressed as calls to \eval. The
+\c{evalq} form quotes the argument to prevent it from being evaluated
+in the current environment. The \c{eval.parent(e,n)} form evaluates
+\c{e} in the environment of the {\tt n}-th caller of this function.
+Finally, \c{local} evaluates an \c{exp} in a new environment to
+avoid polluting the current one.
 
@@
 representative examples of its usage.
+
+\begin{lstlisting}
+  > library("ggplot2")  # load a library from string
+  > library(ggplot2)     # equivalent, except symbol is not in scope
+
+  library <- function(package, ...) {
+    ...
+    if(!is.character(package))
+      package <- as.character(substitute(package))
+    ...
+  }
+\end{lstlisting}
 

### 2020-11-30 inst/analysis/preprocess.R
Preprocess stage uses preprocess.R now

@@ -421,7 +421,6 @@ Example:
 }
 
 main <- function(argv) {
-  argv <- argv[-1] # Remove the command name
   if (length(argv) != 10) {
     cat("Only ", length(argv), " arguments. Missing arguments!\n")
     usage()

### 2020-11-30 inst/paper/main.tex

@@
 
-\begin{lstlisting}
-  > library("ggplot2")  # load a library from string
-  > library(ggplot2)     # equivalent, except symbol is not in scope
-
-  library <- function(package, ...) {
-    ...
-    if(!is.character(package))
-      package <- as.character(substitute(package))
-    ...
-  }
-\end{lstlisting}
-
 \subsection{Previous Work}
@@
 
-\section{Evaluation}
-
-This section presents the results of our empirical study of \eval in R
-in terms of four research questions.
-
-\subsection{Use of Eval}
-
-The first research question we address is how and why \eval is used in R code.
-
-\begin{center}{\bfseries RQ1: What are the patterns of usage of \eval in the corpus?}\end{center}
-
-This section presents an overview of the usage of \eval in our corpus. Richards
-et al. reported 82\% of JavaScript-enabled web pages using \eval~\cite{ecoop11}.
-The case for R is more striking with 100\% of the programs call the \eval
-function.\footnote{We combine calls to {\sf eval}, {\sf eval.parent}, {\sf
-    evalq} and {\sf local}.} This high number is explained by the fact that the
-R implementation, its core packages, use \eval extensively. We differentiate
-three types of eval based on their origin:
+\newcommand{\mypara}[1]{\medskip\noindent\emph{#1}\xspace}
+
+%%%%%
+\section{Usage Metrics}
+
+This section presents a high-level picture of the usage of \eval in R.
+
+(CHECK IF TRUE) As none of the Kaggle programs invoke \eval, the
+remainder of the paper focuses solely on the CRAN data. We will return
+to Kaggle in our conclusion.
+
+We recorded \AllAllCallCountRnd calls to \eval and its variants (XXX001
+for \eval, XXX002 for \c{evalq}, XXX003 for \c{eval.parent} and XXX004 for
+\c{local}).
+
+XXX005 packages in our corpus call \eval at least once.
+
+\mypara{Number of eval call sites.}
+We observed that the number of \eval call sites is small.
+Figure XXX006 shows a violin plot of the distribution of call sites to
+\eval and its three variants. For packages that have call sites, the
+median is XXX007 call sites. \OneCallSitesEnrich{} packages have a
+single call site, and one package, \emph{metafor}, has
+\MaxCallSitesEnrich{} call sites. \NoCallSitesEnrich{} packages have
+no calls to \eval. These numbers are under-approximations as \eval may
+be aliased, called reflectively and given as argument to a
+higher-order function such as \c{map}.
+
+\newcommand{\NOTE}[1]{{\it Note: #1}\xspace}
+
+\mypara{Number of calls to eval.} We observed an average XXX010 calls
+to \eval per run. \NOTE{this is in packages that had any call to eval.
+  How do we compute this? Say we are interested in package P. Do we
+  only count runs that call eval from package P. Or do we also count
+  runs that did not have an eval but executed some code from P? The
+  latter may be easier. Which one is right?} Figure XXX011 shows a
+violin plot of the distribution of calls to eval per run. The largest
+number of calls is XXX012 and comes from the package XXX013.
+
+\mypara{Amount of code loaded by eval.} We use \c{deparse} to turn the
+input to \eval into strings and measure the size of that input. The
+largest input is XXX008 and the median input size is XXX009. Most
+inputs are small, less than XXX014 bytes. The average size is XXX015
+bytes.
+
+\mypara{Amount of computation via eval.}
+We compare the number of events intercepted during execution of \eval
+across the whole corpus. For this, we use the number of interpreter
+steps as a proxy for events. We observe \AllEventCountRnd events all
+told, and \EvalEventCountRnd events occur inside eval
+(\EvalEventAllPerc of all events).
+Figure~XXX015 show a violin plot of events.
+
+We observe a very wide spread in the number of events generated by
+\eval calls. The largest number of events generated by an \eval call
+is \EventsMaxCountRnd. However, a majority of \eval calls performs up
+to \EventsMaxRangeA events suggesting that most expressions passed to
+\eval are small. It is interesting to note that a higher proportion of
+Core R \eval calls perform side effects in the
+\EventsMinRangeB--\EventsMaxRangeB range. Only \EventsCoreEvalCountC
+Core R \eval calls and \EventsPackageEvalCountC package \eval call
+generate \EventsMinRangeC to \EventsMaxRangeC events. These \eval
+calls originate from statistical modeling packages such as \mlogit,
+\mboost, \metafor, \lavaan, \mclust and \gamlss.
+
+\mypara{The case of CORE libraries} ... \NOTE{Give numbers for the
+  same things as above but for the base/core libraries} Compare.
+\NOTE{Add XXX016 to XXX025.}
+
+\section{A Taxonomy of Eval}
+
+In this section we turn our attention to the expression passed for
+evaluation to \eval. The \eval function accepts all types of R objects
+as inputs, but in the majority of cases
+(\AllExpressionInputEvalCallPerc), it is called with expressions. The
+remaining \AllValueInputEvalCallPerc are values, \ie, they evaluate to
+themselves.
 %
-\begin{compactitem}[$-$]
-  \item a {\bf core eval} is one that occurs in the code of a core package,
-  \item a {\bf dependent eval} occurs in a package imported by the target
-  package, and
-  \item an {\bf own eval} is one that occurs in the source code of the target
-  package.
-\end{compactitem}
-
-Over all, we observe \AllAllCallCountRnd calls to \eval in our corpus.
-Table~\ref{A} details this run-time information. The format is T (m/M/x)
-where T is the total number of calls observed for all runs of the corpus, m
-and x are the minimum and maximum number of calls for any given run, and M
-is the mean number of calls per run. An average CRAN package run triggers
-581 {\eval} calls in core package, 532 in dependent packages and 292 in the
-CRAN package itself.  The Kaggle data set has small programs which only
-trigger \eval in core and dependent packages. Thus, an average Kaggle run
-has 412 core calls, 96 dependent calls, and no own calls.
-
-%% SORRY  I HAD TO REFORMAT
-\begin{table}[!h] \centering
- \begin{tabular}{r|c|c|c} \hline
-Source & Core    & Dependent                      & Own \\\hline
-CRAN   & 5.6M &  867K    & 259K \\
-       & (1 / 337 / 46K)& (0 / 52 / 58K ) & (0 / 15 / 15K) \\\hline
-Kaggle &  1.2M &  470K   & 0\\
- &  (0 / 721 / 18K) & (0 / 290 / 20K) & (0 / 0 / 0)\\\hline
-\end{tabular}\caption{Eval calls in CRAN and Kaggle}\label{A}\end{table}
-
-Table~\ref{table:site-package-summary} shows the number of call sites in
-CRAN packages. \OneCallSitesEnrich{} packages have a single call site, and one
-package, \emph{metafor}, has \MaxCallSitesEnrich{} call sites of
-\eval. \NoCallSitesEnrich{} packages have no calls to \eval. The core packages
-have \NbCoreEvalCallSites{} call sites of \eval and Kaggle has none. These
-numbers are under-approximations as \eval may be called reflectively,
-through aliases, or passed to a higher-order function.
-
-\begin{table}[ht]\resizebox{\columnwidth}{!}{%
-\begin{tabular}{rr|rr|rr}\hline
- Callsites & Packages & Callsites & Packages & Callsites & Packages \\\hline
-\input{tag/table-site-package-summary.tex}
-\end{tabular}}
-\caption{Distribution of callsites in CRAN}
-\label{table:site-package-summary}
-\end{table}
-
-We compare the number of events intercepted during execution of \eval across
-the whole corpus. For this, we use the number of interpreter steps as a
-proxy for events. We observe \AllEventCountRnd events all told, and
-\EvalEventCountRnd events occur inside eval (\EvalEventAllPerc of all
-events). Table~\ref{table:event-distribution} shows the distribution of
-events caused by \eval calls.
-
-\begin{table}[ht]\resizebox{\columnwidth}{!}{%
-\begin{tabular}{lll}\hline
-Events & Core Eval Call \% & Package Eval Call \% \\ \hline
-\EventsMinRangeA--\EventsMaxRangeA & \EventsCoreEvalPercA & \EventsPackageEvalPercA \\
-\EventsMinRangeB--\EventsMaxRangeB & \EventsCoreEvalPercB & \EventsPackageEvalPercB \\
-\EventsMinRangeC+ & \EventsCoreEvalPercC & \EventsPackageEvalPercC \\ \hline
-\end{tabular}} \label{table:event-distribution}
-  \caption{Events in \eval calls}
-\end{table}
-
-We observe a very wide spread in the number of events generated by \eval
-calls. The largest number of events generated by an \eval call is
-\EventsMaxCountRnd.  However, a majority of \eval calls performs up to
-\EventsMaxRangeA events suggesting that most expressions passed to \eval are
-small.  It is interesting to note that a higher proportion of Core R \eval
-calls perform side effects in the \EventsMinRangeB--\EventsMaxRangeB range.
-Only \EventsCoreEvalCountC Core R \eval calls and \EventsPackageEvalCountC
-package \eval call generate \EventsMinRangeC to \EventsMaxRangeC events.
-These \eval calls originate from statistical modeling packages such as
-\mlogit, \mboost, \metafor, \lavaan, \mclust and \gamlss.
-
-\subsubsection{A taxonomy of eval}
-
-% Following the high-level overview of the usage of eval in the previous section,
-% we now look into categorization of the eval behavior. Similarly to the JavaScript paper, we look at:
-% %
-% \begin{inparaenum}[(A)]
-%   \item \textbf{mix} of operations performed by the expressions passed to eval,
-%   \item \textbf{}
-% \end{inparaenum}
-
-In this section we turn our attention to the expression passed for evaluation to
-\eval. The \eval function accepts all types of R objects as inputs, but in the
-majority of cases (\AllExpressionInputEvalCallPerc), it is called with
-expressions. The remaining \AllValueInputEvalCallPerc are values, \ie, they
-evaluate to themselves.
-%
-Passing a value into an eval is effectively an no-op so we look closely at the
-expression types the call sites pass to eval across all the runs.
-\AllValueInputEvalSitePerc of all call sites always pass values,
-\AllExpressionInputEvalSitePerc pass expressions and a very large proportion,
-\AllPolymorphicInputEvalSitePerc, pass both expressions and values as inputs
-across all program runs. We inspected the eval callsites that accept only values
-or both values and expression as inputs. The top three of these call sites
-contribute to over ~80\% of all calls to evals. The first one is inside
-\c{match.arg} from core. Given a vector of names as strings, this function finds
-the default values for parameters with those names in the caller function. Since
-the default value for a function parameter can be a complex expression or a
-benign value such as \c{NULL}, the \c{match.arg} function evaluates it through a
-call to \eval. This makes the \eval inside \c{match.arg} ``polymorhpic''. The
-second such \eval call originates from \ggplot, a popular R package for making
-plots. \ggplot has it's own object system, \ggproto, which uses \eval inside the
-\ggproto function to determine the super-class from which it inherits (the
-default value is \c{NULL} but in general it could be an arbitrarily expression).
-The third \eval call is in core R \c{str} function used to print a summary of R
-objects. It internally maps \eval function on the bindings of an environment
-which results in inputs of multiples types being passed to \eval.
+Passing a value into an eval is effectively an no-op so we look
+closely at the expression types the call sites pass to eval across all
+the runs. \AllValueInputEvalSitePerc of all call sites always pass
+values, \AllExpressionInputEvalSitePerc pass expressions and a very
+large proportion, \AllPolymorphicInputEvalSitePerc, pass both
+expressions and values as inputs across all program runs. We inspected
+the eval callsites that accept only values or both values and
+expression as inputs. The top three of these call sites contribute to
+over ~80\% of all calls to evals. The first one is inside
+\c{match.arg} from core. Given a vector of names as strings, this
+function finds the default values for parameters with those names in
+the caller function. Since the default value for a function parameter
+can be a complex expression or a benign value such as \c{NULL}, the
+\c{match.arg} function evaluates it through a call to \eval. This
+makes the \eval inside \c{match.arg} ``polymorhpic''. The second such
+\eval call originates from \ggplot, a popular R package for making
+plots. \ggplot has it's own object system, \ggproto, which uses \eval
+inside the \ggproto function to determine the super-class from which
+it inherits (the default value is \c{NULL} but in general it could be
+an arbitrarily expression). The third \eval call is in core R \c{str}
+function used to print a summary of R objects. It internally maps
+\eval function on the bindings of an environment which results in
+inputs of multiples types being passed to \eval.
 
 We further investigate the type of input passed to \eval in
-Figure~\ref{fig:eval-expression-kind} by the type of input and the source of
-\eval. In this figure, \emph{Expression} is the type of program text,
-\emph{Environmment} is the type of scopes or environments used for binding,
-\emph{Null} is the type of the unique \c{NULL} object, \emph{Vector} is the
-type assigned to vectors of strings, booleans, integers, reals and bytes;
-\emph{Symbol} is the type of unevaluated symbols, and \emph{Function} is the
-type of functions.
+Figure~\ref{fig:eval-expression-kind} by the type of input and the
+source of \eval. In this figure, \emph{Expression} is the type of
+program text, \emph{Environmment} is the type of scopes or
+environments used for binding, \emph{Null} is the type of the unique
+\c{NULL} object, \emph{Vector} is the type assigned to vectors of
+strings, booleans, integers, reals and bytes; \emph{Symbol} is the
+type of unevaluated symbols, and \emph{Function} is the type of
+functions.
 
@@
 
-From the figure, we observe that majority of inputs to Core \eval functions
-are expressions in contrast to package \eval calls which receive mostly
-environments as inputs. In fact, \PackageEnvironmentInputEvalCallPerc of
-\eval calls in packages receive environments as input expressions. There are
-only four callsites in package evals that receive environments as
-expressions.
+From the figure, we observe that majority of inputs to Core \eval
+functions are expressions in contrast to package \eval calls which
+receive mostly environments as inputs. In fact,
+\PackageEnvironmentInputEvalCallPerc of \eval calls in packages
+receive environments as input expressions. There are only four
+callsites in package evals that receive environments as expressions.
 \begin{compactitem}[$-$]
@@
   receive environments as input.
-\item \c{R6::generator_funs} function contributes to only 936 \eval calls and
-implements the same functionality as \c{ggplot2::ggproto} but for the \emph{R6}
-OOP system.
-\item \c{future::backtrace} function applies \eval to a future object which
-  is implemented as an environment. This is called only once.
-\item \c{RModel::str.RMmodel} function overloads the core \c{str} method for its
-\c{RModel} objects and maps \eval exactly like \c{str}. This is called only
-once.
+\item \c{R6::generator_funs} function contributes to only 936 \eval
+  calls and implements the same functionality as \c{ggplot2::ggproto}
+  but for the \emph{R6} OOP system.
+\item \c{future::backtrace} function applies \eval to a future object
+  which is implemented as an environment. This is called only once.
+\item \c{RModel::str.RMmodel} function overloads the core \c{str}
+  method for its \c{RModel} objects and maps \eval exactly like
+  \c{str}. This is called only once.
 \end{compactitem}
@@
 
-The expression \c{\{info <- loadingNamespaceInfo(...} is added by core R
-to a package directory during installation. To load the package, this code is
-executed. It creates a namespace for the package, injects the package bindings,
-and attaches the namespace to the program search path. The \c{NULL} comes from a
-call to \c{substitute(subset)} in \c{stats::model.frame.default} function which
-has a default value of \c{subset} as \c{NULL}. The \c{Symbol} arises from a
-call to \c{as.name} in \c{base::str} function that returns a symbol that is
-looked up by evaluating it in a specific environment. The remaining cases arise
-from calls to \c{match.arg} which is used to look up the default choices for a
-variable and match against the choice passed by the caller.
+The expression \c{\{info <- loadingNamespaceInfo(...} is added by core
+R to a package directory during installation. To load the package,
+this code is executed. It creates a namespace for the package, injects
+the package bindings, and attaches the namespace to the program search
+path. The \c{NULL} comes from a call to \c{substitute(subset)} in
+\c{stats::model.frame.default} function which has a default value of
+\c{subset} as \c{NULL}. The \c{Symbol} arises from a call to
+\c{as.name} in \c{base::str} function that returns a symbol that is
+looked up by evaluating it in a specific environment. The remaining
+cases arise from calls to \c{match.arg} which is used to look up the
+default choices for a variable and match against the choice passed by
+the caller.
 
@@
 explained above, \c{ggplot2::ggproto}, \c{R6::generator_funs},
-\c{future::backtrace} and \c{RModel::str.RMmodel}. The next expression,
-\c{column[rows] <<- what}, is used inside the \c{plyr::rbind.fill} function
-to merge data frames by assigning concatenated vectors to rows. The \c{<<-}
-operator is interesting in that it skips the current scope and assigns in a
-parent scope in which the variable is already present. In our corpus, all
-these \eval calls contribute to a single side-effect. The expression
-\c{function(value) freduce(value, `_function_list`)} arises from the
-\c{magrittr::\%>\%} function which is a pipe operator that pipes the output
-of previous command to the next one. The expression is evaluated in a custom
-environment to create a function binding for evaluating the components of
-the pipe.\c{String Vector} and \c{List} also arise from the same function
-when a string or a list is piped using the \c{\%>\%} function into the next
-expression.  The \c{NULL} arises from \c{R6::generator_funs} function when
-the \eval is passed a \c{NULL} argument by the \c{DataMask_generator}
-package. The \c{c("default", "default2012", "default2011" ...} pattern
-arises from \c{copula::polyG} where it reflectively access the default
-expression for its formal parameter and evaluates it. The \c{force(..1)} and
-\c{alist(`_spec`)} patterns occur in \c{glue::glue_data} function which
-concatenates and interpolates strings. The two patterns occur because the
-function captures unevaluated unnamed arguments and maps the evaluation of
-\c{force(..1)} on them.  The \c{force} function forces promises and returns
-the result of evaluation. The \c{inner} pattern arises from
-\c{glue::identity_transformer} which enables the creation of custom
-transformation functions for affecting the interpolation and concatenation
-of input by the \c{glue} package.
+\c{future::backtrace} and \c{RModel::str.RMmodel}. The next
+expression, \c{column[rows] <<- what}, is used inside the
+\c{plyr::rbind.fill} function to merge data frames by assigning
+concatenated vectors to rows. The \c{<<-} operator is interesting in
+that it skips the current scope and assigns in a parent scope in which
+the variable is already present. In our corpus, all these \eval calls
+contribute to a single side-effect. The expression \c{function(value)
+  freduce(value, `_function_list`)} arises from the
+\c{magrittr::\%>\%} function which is a pipe operator that pipes the
+output of previous command to the next one. The expression is
+evaluated in a custom environment to create a function binding for
+evaluating the components of the pipe.\c{String Vector} and \c{List}
+also arise from the same function when a string or a list is piped
+using the \c{\%>\%} function into the next expression. The \c{NULL}
+arises from \c{R6::generator_funs} function when the \eval is passed a
+\c{NULL} argument by the \c{DataMask_generator} package. The
+\c{c("default", "default2012", "default2011" ...} pattern arises from
+\c{copula::polyG} where it reflectively access the default expression
+for its formal parameter and evaluates it. The \c{force(..1)} and
+\c{alist(`_spec`)} patterns occur in \c{glue::glue_data} function
+which concatenates and interpolates strings. The two patterns occur
+because the function captures unevaluated unnamed arguments and maps
+the evaluation of \c{force(..1)} on them. The \c{force} function
+forces promises and returns the result of evaluation. The \c{inner}
+pattern arises from \c{glue::identity_transformer} which enables the
+creation of custom transformation functions for affecting the
+interpolation and concatenation of input by the \c{glue} package.
 
@@
 
-The JavaScript paper reported on the operation mix within \evals. This means
-the proportion of read, write, and call bytecodes performed by the
-JavaScript interpreter. They found that some web pages were creating objects
-more frequently and writing into their fields.  In R, nearly every operation
-performed in the interpreter is a function call. We observe 2.9G calls and,
-of those 125M happen inside \eval (there are 463M native calls, of which
-only 31M are in \eval).
+The JavaScript paper reported on the operation mix within \evals. This
+means the proportion of read, write, and call bytecodes performed by
+the JavaScript interpreter. They found that some web pages were
+creating objects more frequently and writing into their fields. In R,
+nearly every operation performed in the interpreter is a function
+call. We observe 2.9G calls and, of those 125M happen inside \eval
+(there are 463M native calls, of which only 31M are in \eval).
 
@@
 
-%% Qualitatively, we observe all kinds of operations being executed inside
-%% \eval calls, from numerical computing for statistical modeling to reflection
-%% and metaprogramming for implementing DSLs. Even testing frameworks such
-%% \c{genthat} heavily rely upon \eval to execute test code snippets. The core
-%% R implementation is rife with uses of \eval for mundane things like
-%% accessing default values of formal parameters and in statistical packages
-%% for computation.
+%% Qualitatively, we observe all kinds of operations being executed
+%% inside \eval calls, from numerical computing for statistical
+%% modeling to reflection and metaprogramming for implementing DSLs.
+%% Even testing frameworks such \c{genthat} heavily rely upon \eval to
+%% execute test code snippets. The core R implementation is rife with
+%% uses of \eval for mundane things like accessing default values of
+%% formal parameters and in statistical packages for computation.
 
@@
 
-Richards et al.~\cite{ecoop11} identified a number of frequently occurring \eval
-patterns which were detected using simple regular expressions on the \eval
-strings. In the case of R we match the patterns on the recorded AST.
-
-Table~\ref{table:js-pattens-in-r} shows the ratio of \eval in each of these
-patterns. We included all the original patterns except for \emph{JSON} as R uses
-its own serialization format and for \emph{empty} as there were no empty expression
-passed to \eval:
+Richards et al.~\cite{ecoop11} identified a number of frequently
+occurring \eval patterns which were detected using simple regular
+expressions on the \eval strings. In the case of R we match the
+patterns on the recorded AST.
+
+Table~\ref{table:js-pattens-in-r} shows the ratio of \eval in each of
+these patterns. We included all the original patterns except for
+\emph{JSON} as R uses its own serialization format and for
+\emph{empty} as there were no empty expression passed to \eval:
 
@@
 
-\noindent \emph{Library}, \emph{Typeof} and \emph{Try} are all barely used
-mostly because the language provides dedicated features for it. The few calls in
-Library pattern come mostly from file loading functions such as \c{source} and
-\c{sys.source}.
+\noindent \emph{Library}, \emph{Typeof} and \emph{Try} are all barely
+used mostly because the language provides dedicated features for it.
+The few calls in Library pattern come mostly from file loading
+functions such as \c{source} and \c{sys.source}.
 %
@@
 access to an object's property or local variable both replaceable by
-JavaScript core functions. In the case of R there are \PatternReadRnd read
-\eval calls (\PatternReadRatio). They come from calls like
-\c{subset(airquality, select = Temp)} where the variable \c{Temp} will be
-evaluated in the \c{airquality} data frame.
+JavaScript core functions. In the case of R there are \PatternReadRnd
+read \eval calls (\PatternReadRatio). They come from calls like
+\c{subset(airquality, select = Temp)} where the variable \c{Temp} will
+be evaluated in the \c{airquality} data frame.
 
@@
 within an \eval call. We have seen \PatternAssignRnd such calls
-(\PatternAssignRatio).  \PatternAssignArrowRatio uses the simple assignment
-(\c{<-}). The majority of cases comes from the \c{magrittr}'s \c{\%<>\%}
-pipe operator (\PatternAssignArrowMagrittrRatio) which forwards an object to
-a function on rhs and update lhs with the result. Far more calls used the
-super assignment (\c{<<-}) operator, but almost all come from the \c{plyr}
-package which provides a convenient API for common data manipulation
-problems.
+(\PatternAssignRatio). \PatternAssignArrowRatio uses the simple
+assignment (\c{<-}). The majority of cases comes from the
+\c{magrittr}'s \c{\%<>\%} pipe operator
+(\PatternAssignArrowMagrittrRatio) which forwards an object to a
+function on rhs and update lhs with the result. Far more calls used
+the super assignment (\c{<<-}) operator, but almost all come from the
+\c{plyr} package which provides a convenient API for common data
+manipulation problems.
 %
-\noindent \emph{Call} and \emph{Other} together represents the vast majority
-of eval calls in R. The \emph{Other} pattern represents all the calls where
-eval was called with a value instead of with an expression. The \emph{Call}
-pattern contains all the calls that did not fall in the previous
-categories. Without a context it is usually difficult to further categorize
-the intent of these calls.  Therefore we have randomly chosen
-\PatternManualPackages packages and manually looked at
-\PatternManualCallsites call sites from which we have observed the following
-categories:
+\noindent \emph{Call} and \emph{Other} together represents the vast
+majority of eval calls in R. The \emph{Other} pattern represents all
+the calls where eval was called with a value instead of with an
+expression. The \emph{Call} pattern contains all the calls that did
+not fall in the previous categories. Without a context it is usually
+difficult to further categorize the intent of these calls. Therefore
+we have randomly chosen \PatternManualPackages packages and manually
+looked at \PatternManualCallsites call sites from which we have
+observed the following categories:
 
 \subsubsection{Recording the current call} R provides a \c{match.call} function
-which can return the current call. This is heavily used by the statistical
-model-fitting functions that records the current call that is then
-reevaluated in the different environment. Another pattern is to match the
-current call, adjust some arguments and pass it to another function. For
-example the \c{write.csv} function changes the \c{sep} argument in the
-matched call and reevaluates itself as \c{write.table}:
+which can return the current call. This is heavily used by the
+statistical model-fitting functions that records the current call that
+is then reevaluated in the different environment. Another pattern is
+to match the current call, adjust some arguments and pass it to
+another function. For example the \c{write.csv} function changes the
+\c{sep} argument in the matched call and reevaluates itself as
+\c{write.table}:
 %
@@
 environments can be used to construct a sandbox for evaluating
-side-effecting expressions. The newly constructed environment can use the
-current environment for its lexical scope. This provides the evaluated
-expression access to all the bindings in the current scope but restricts the
-side-effect to the newly constructed throwaway environment. However, this
-sandboxing is limited because the expression can reflectively access any
-environment in the call stack or use the super assignment operator(\c{<<-})
-to write to parent scope. The \c{local} function which is part of Core R is
-an example of this usage pattern.
+side-effecting expressions. The newly constructed environment can use
+the current environment for its lexical scope. This provides the
+evaluated expression access to all the bindings in the current scope
+but restricts the side-effect to the newly constructed throwaway
+environment. However, this sandboxing is limited because the
+expression can reflectively access any environment in the call stack
+or use the super assignment operator(\c{<<-}) to write to parent
+scope. The \c{local} function which is part of Core R is an example of
+this usage pattern.
 
@@
 \subsubsection{Metaprogramming} The combination of \eval and \substitute is
- used to do metaprogramming in R. \substitute can be used to synthesize ASTs
- by stitching together user inputs. \substitute walks over its argument AST
- and replaces the symbols with their bindings in the specified environment
- and returns this new AST which can then be evaluated in custom environments
- using \eval.
+ used to do metaprogramming in R. \substitute can be used to
+ synthesize ASTs by stitching together user inputs. \substitute walks
+ over its argument AST and replaces the symbols with their bindings in
+ the specified environment and returns this new AST which can then be
+ evaluated in custom environments using \eval.
 
@@
   \begin{lstlisting}
-opts = c("datatable.verbose"="FALSE", ...)
-for (i in names(opts)) eval(parse(text=paste0("options(",i,"=",opts[i],")")))
+opts = c("datatable.verbose"="FALSE", ...) for (i in names(opts))
+eval(parse(text=paste0("options(",i,"=",opts[i],")")))
 
@@
   \subsubsection{Implementation}
-  The package installation process adds an extra file to the package directory
-  that contains the code to load the package code inside a \local block. It
-  creates a namespace object and injects the package bindings.
-
-  \subsubsection{Miscellaneous} \eval is also used in R packages to get around
-  various limitations. For example, in \datatable package, \eval is used to
-  modify a binding in the \base package of R. Bindings in the \base package are
+  The package installation process adds an extra file to the package
+  directory that contains the code to load the package code inside a
+  \local block. It creates a namespace object and injects the package
+  bindings.
+
+  \subsubsection{Miscellaneous}
+  \eval is also used in R packages to get around various limitations.
+  For example, in \datatable package, \eval is used to modify a
+  binding in the \base package of R. Bindings in the \base package are
   locked for modification and require explicit unlocking using
-  function \unlockBinding. However, directly using this function triggers a
-  restricted-function-use warning during package build, preventing the package
-  to be accepted by CRAN. The \datatable gets around this limitation by
-  enclosing the call to \unlockBinding inside \eval. From the same reason, other
-  packages use \eval to access private variables and functions from their
-  dependencies.
+  function \unlockBinding. However, directly using this function
+  triggers a restricted-function-use warning during package build,
+  preventing the package to be accepted by CRAN. The \datatable gets
+  around this limitation by enclosing the call to \unlockBinding
+  inside \eval. From the same reason, other packages use \eval to
+  access private variables and functions from their dependencies.
 
@@
 
-\c{eval(parse(...))} can be used for dynamic code loading. This forms the
-core of \c{source} and \c{sys.source} functions in R that are commonly used
-for loading code in R files in interactive settings. We investigated the
-number of cases in which the output of \c{parse} and its variants is passed
-to \eval, directly or transitively by tainting their output. This
-corresponds to \PercentParsedCallSites of the total \eval call sites and
-\PercentParsedEvals of the eval calls. We observed that very few of the eval
-call (\NbParseFilesRnd in total) consume the result of calling \c{parse} on
-a file. Most of the eval calls consume the result of calling \c{parse} on a
-string.  We also identified one function in core R,
-\c{invokeRestartInteractively} that prompts the user for input, parses it,
-and passes it to \eval.
+\c{eval(parse(...))} can be used for dynamic code loading. This forms
+the core of \c{source} and \c{sys.source} functions in R that are
+commonly used for loading code in R files in interactive settings. We
+investigated the number of cases in which the output of \c{parse} and
+its variants is passed to \eval, directly or transitively by tainting
+their output. This corresponds to \PercentParsedCallSites of the total
+\eval call sites and \PercentParsedEvals of the eval calls. We
+observed that very few of the eval call (\NbParseFilesRnd in total)
+consume the result of calling \c{parse} on a file. Most of the eval
+calls consume the result of calling \c{parse} on a string. We also
+identified one function in core R, \c{invokeRestartInteractively} that
+prompts the user for input, parses it, and passes it to \eval.
 
@@
 \EvalWritesRnd writes happen inside \eval. However, all writes are not
-dangerous. Only writes to environments not local to the computation spawned
-by \eval are side-effecting. These writes outlive the computation and hence
-are visible outside it. The remaining writes are local to the
-computation. We observe that \EvalSideEffectingWritesRnd writes inside evals
-are side-effecting. This is only \EvalSideEffectingWritesEvalPerc of all
-variable writes inside \eval and \EvalSideEffectingWritesAllPerc of all
-variable writes in the corpus.
+dangerous. Only writes to environments not local to the computation
+spawned by \eval are side-effecting. These writes outlive the
+computation and hence are visible outside it. The remaining writes are
+local to the computation. We observe that \EvalSideEffectingWritesRnd
+writes inside evals are side-effecting. This is only
+\EvalSideEffectingWritesEvalPerc of all variable writes inside \eval
+and \EvalSideEffectingWritesAllPerc of all variable writes in the
+corpus.
 
@@
 
-One of the main differences of the R \eval with the javascript one is how it
-interacts with \emph{scope}, or environments in R. In Javascript, \eval can
-access the local scope and the global scope. In R, \eval has an argument
-\c{envir} to indicate in which environment its first argument must be evaluated.
-By default, it is equal to \c{parent.frame()},
-which is the parent environment of \eval call. This default argument
-happens in \DefaultEnvirExprPercent of the eval calls.
-
-This argument can also be \c{NULL}, a \c{List} or a data frame. This happens
-in 3.5\% of the cases: \eval copies the fields of the list or data frame
-and creates bindings for them in a new environment. This pattern is used to
-evaluate formulas which can directly refer to the fields of the data.  The
-\c{envir} argument can also be a number $n$. It means that the environment
-in which the expressions is evaluated will be the result of \c{sys.call(n)}
-where $n$ refers to the $n$-th stack frame.
+One of the main differences of the R \eval with the javascript one is
+how it interacts with \emph{scope}, or environments in R. In
+Javascript, \eval can access the local scope and the global scope. In
+R, \eval has an argument \c{envir} to indicate in which environment
+its first argument must be evaluated. By default, it is equal to
+\c{parent.frame()}, which is the parent environment of \eval call.
+This default argument happens in \DefaultEnvirExprPercent of the eval
+calls.
+
+This argument can also be \c{NULL}, a \c{List} or a data frame. This
+happens in 3.5\% of the cases: \eval copies the fields of the list or
+data frame and creates bindings for them in a new environment. This
+pattern is used to evaluate formulas which can directly refer to the
+fields of the data. The \c{envir} argument can also be a number $n$.
+It means that the environment in which the expressions is evaluated
+will be the result of \c{sys.call(n)} where $n$ refers to the $n$-th
+stack frame.
 
 The top-level environment in R is called the global environment. New
-environments can be created using \c{new.env}. They can be provided a parent
-environment which becomes the enclosing scope of the new environment.
-
-We looked at the environments passed to all the \eval calls in our corpus.
-Table~\ref{tab:environments} summarizes the results. A numeric environment
-class \c{n} denotes the environment of the $n$-th call stack frame from the
-current function. \c{global} denotes the top-level environment and \c{list}
-denotes a list passed for evaluation of formulas. Environment classes of the
-form $n+$ denote the $n$-th environment extended with a new environment. The
-new environment provides a limited form of sandboxing. All assignments using
-the \c{\<-} function occur inside it and prevent the extended environment
-from mutation. However, it is still possible to mutate the extended
-environment using the \c{\<\<-} or \c{assign} functions; but, that happens
-rarely.
-
-\begin{table}[htbp] \resizebox{\columnwidth}{!}{%
+environments can be created using \c{new.env}. They can be provided a
+parent environment which becomes the enclosing scope of the new
+environment.
+
+We looked at the environments passed to all the \eval calls in our
+corpus. Table~\ref{tab:environments} summarizes the results. A numeric
+environment class \c{n} denotes the environment of the $n$-th call
+stack frame from the current function. \c{global} denotes the
+top-level environment and \c{list} denotes a list passed for
+evaluation of formulas. Environment classes of the form $n+$ denote
+the $n$-th environment extended with a new environment. The new
+environment provides a limited form of sandboxing. All assignments
+using the \c{\<-} function occur inside it and prevent the extended
+environment from mutation. However, it is still possible to mutate the
+extended environment using the \c{\<\<-} or \c{assign} functions; but,
+that happens rarely.
+
+\begin{table}[htbp]{%} \resizebox{\columnwidth}{!}{%
     \centering
@@
 access the caller's caller's environment. This is because many core R
-functions call functions that pass the result of \c{parent.frame()} to an
-\eval.  A disproportionately high number of calls to \eval happens in an
-extended top-level environment. This can be explained by the fact that many
-packages evaluate code passed from the user's workspace in the top-level
-environment to access the bindings.
-
-The \c{imchange} function of package \c{imager} makes it possible to modify
-images using a dedicated formula syntax using \c{\~}.\footnote{Inspired by
- {map} in package \emph{purr}.}  Here, \eval is evaluated in \c{newenv},
-which creates a new environment that inherits from \c{parent.frame()} by
-default (classified as 1+).
+functions call functions that pass the result of \c{parent.frame()} to
+an \eval. A disproportionately high number of calls to \eval happens
+in an extended top-level environment. This can be explained by the
+fact that many packages evaluate code passed from the user's workspace
+in the top-level environment to access the bindings.
+
+The \c{imchange} function of package \c{imager} makes it possible to
+modify images using a dedicated formula syntax using
+\c{\~}.\footnote{Inspired by {map} in package \emph{purr}.} Here,
+\eval is evaluated in \c{newenv}, which creates a new environment that
+inherits from \c{parent.frame()} by default (classified as 1+).
 
@@
 
-\c{adjCoef} in package \emph{actuar} find the root of an equation defined by
-a function \c{h} whose arguments must be named \c{x} and \c{y}.  \c{h} is
-transformed into an auxiliary function \c{h2} that can be optimized. Here,
-the list used for \c{envir} ensures the correspondance between the textual
-arguments of \c{h} and the arguments of \c{h2}.
+\c{adjCoef} in package \emph{actuar} find the root of an equation
+defined by a function \c{h} whose arguments must be named \c{x} and
+\c{y}. \c{h} is transformed into an auxiliary function \c{h2} that can
+be optimized. Here, the list used for \c{envir} ensures the
+correspondance between the textual arguments of \c{h} and the
+arguments of \c{h2}.
 
@@
 
+\section{Case Studies}
+
+\section{Why do we Eval}
+
+\section{Analysis Results}
+
+\subsubsection{A taxonomy of eval}
+
 \section{Discussion}
@@
 
-\medskip\noindent\emph{Discussion:} Comparing the use of \eval in R and
-JavaScript is instructive. The use of \eval is more widespread in data
-science code than it was in JavaScript. Any run of an R program seems to
-trigger \eval; moreover package programmers use \eval liberally. On the
-other hand, the authors of Kaggle packages do not use the feature at all.
-This point to a bifurcated user community: experts who write package know
-how to use \eval while end-users do not. An alternative explanation is that
-end-users do not need \eval for simple data analysis tasks. This bifurcation
-was not observed in the JavaScript community, but this may have changed with
-the advent of Node.js ecosystem.  The distribution of input sizes is similar
-to that found in JavaScript, the majority of inputs being rather small.
+\medskip\noindent\emph{Discussion:} Comparing the use of \eval in R
+and JavaScript is instructive. The use of \eval is more widespread in
+data science code than it was in JavaScript. Any run of an R program
+seems to trigger \eval; moreover package programmers use \eval
+liberally. On the other hand, the authors of Kaggle packages do not
+use the feature at all. This point to a bifurcated user community:
+experts who write package know how to use \eval while end-users do
+not. An alternative explanation is that end-users do not need \eval
+for simple data analysis tasks. This bifurcation was not observed in
+the JavaScript community, but this may have changed with the advent of
+Node.js ecosystem. The distribution of input sizes is similar to that
+found in JavaScript, the majority of inputs being rather small.
 
@@
 To roughly estimate the complexity of the code being evaluated,
-Fig.~\ref{fig:ast-size} shows the distribution of the abstract syntax tree
-size of the \c{expr} argument to \eval. The majority of inputs across core
-and packages (\NbAstOnePercent of \eval calls and \NbAstOneCallSitePercent
-of call sites) have size 1. That is to say, the code being evaluated
-consists of a single variable name or constant. A function call needs at
-least three nodes and a list such as \c{c(1, 2, 3)} counts for 4
-nodes. Inputs smaller than \AstSizeNineFive nodes account for 95\% of all
-calls. Sizes in core are larger than in packages. Most large ASTs originate
-from library \c{VGAM}, which is used for statistical modeling, and also calls functions using \eval from core. It uses
-\eval to evaluate large matrices with values in a given environment.
+Fig.~\ref{fig:ast-size} shows the distribution of the abstract syntax
+tree size of the \c{expr} argument to \eval. The majority of inputs
+across core and packages (\NbAstOnePercent of \eval calls and
+\NbAstOneCallSitePercent of call sites) have size 1. That is to say,
+the code being evaluated consists of a single variable name or
+constant. A function call needs at least three nodes and a list such
+as \c{c(1, 2, 3)} counts for 4 nodes. Inputs smaller than
+\AstSizeNineFive nodes account for 95\% of all calls. Sizes in core
+are larger than in packages. Most large ASTs originate from library
+\c{VGAM}, which is used for statistical modeling, and also calls
+functions using \eval from core. It uses \eval to evaluate large
+matrices with values in a given environment.
 
@@
 
-We look at how \emph{consistent} the \c{expr} argument of \eval can be, \ie
-how many different types of the resolved \c{expr} there are per call
-sites. Most of the call sites, \ie \PercentMonomorphic, are
-\emph{consistent}, and this is similar to javascript. However, a few ones
-are highly \emph{polymorphic} (10 different types). They are the pipe
-operators \c{\%>\%}, \c{\%<>\%} and \c{\%\$\%} in package \emph{magrittr}. It
-is effectively used to compose functions on their first argument, which can
-be of any type.
+We look at how \emph{consistent} the \c{expr} argument of \eval can
+be, \ie how many different types of the resolved \c{expr} there are
+per call sites. Most of the call sites, \ie \PercentMonomorphic, are
+\emph{consistent}, and this is similar to javascript. However, a few
+ones are highly \emph{polymorphic} (10 different types). They are the
+pipe operators \c{\%>\%}, \c{\%<>\%} and \c{\%\$\%} in package
+\emph{magrittr}. It is effectively used to compose functions on their
+first argument, which can be of any type.
 

### 2020-11-30 inst/paper/main.tex
minor

@@
 
+The previous section gave a quantitative view of the usage of \eval;
+in this section we try to elucidate \emph{what} it does.
+
 In this section we turn our attention to the expression passed for
@@
 \begin{compactitem}[$-$]
-\item \c{ggplot2::ggproto} contributes to 99.9\% of package \eval calls that
-  receive environments as input.
+\item \c{ggplot2::ggproto} contributes to 99.9\% of package \eval
+  calls that receive environments as input.
 \item \c{R6::generator_funs} function contributes to only 936 \eval
@@
 
-\subsubsection{Operation Mix}
-
-The JavaScript paper reported on the operation mix within \evals. This
-means the proportion of read, write, and call bytecodes performed by
-the JavaScript interpreter. They found that some web pages were
-creating objects more frequently and writing into their fields. In R,
-nearly every operation performed in the interpreter is a function
-call. We observe 2.9G calls and, of those 125M happen inside \eval
-(there are 463M native calls, of which only 31M are in \eval).
-
-{\bf there is no good reason for not counting reads and writes, we can...}
-
-%% ...The following is fluff ...
-
-%% Qualitatively, we observe all kinds of operations being executed
-%% inside \eval calls, from numerical computing for statistical
-%% modeling to reflection and metaprogramming for implementing DSLs.
-%% Even testing frameworks such \c{genthat} heavily rely upon \eval to
-%% execute test code snippets. The core R implementation is rife with
-%% uses of \eval for mundane things like accessing default values of
-%% formal parameters and in statistical packages for computation.
-
 \subsubsection{Usage Patterns}
@@
 %
-\noindent \emph{Read} pattern in JavaScript represented mostly an incorrect
-
+\noindent
+\emph{Read} pattern in JavaScript represented mostly an incorrect
 access to an object's property or local variable both replaceable by
@@
 
-\subsubsection{Recording the current call} R provides a \c{match.call} function
-which can return the current call. This is heavily used by the
-statistical model-fitting functions that records the current call that
-is then reevaluated in the different environment. Another pattern is
-to match the current call, adjust some arguments and pass it to
-another function. For example the \c{write.csv} function changes the
-\c{sep} argument in the matched call and reevaluates itself as
-\c{write.table}:
+\subsubsection{Recording the current call}
+R provides a \c{match.call} function which can return the current
+call. This is heavily used by the statistical model-fitting functions
+that records the current call that is then reevaluated in the
+different environment. Another pattern is to match the current call,
+adjust some arguments and pass it to another function. For example the
+\c{write.csv} function changes the \c{sep} argument in the matched
+call and reevaluates itself as \c{write.table}:
 %
@@
 
-\subsubsection{Dynamic Code Loading} The combination of \parse and \eval is
-used to parse text from string or file and evaluate in a custom environment.
-This pattern is used by the \source Core R function to load R code from a
-file in the current workspace. This is quite useful for loading functions
-for interactive data analysis.
-
-\subsubsection{Metaprogramming} The combination of \eval and \substitute is
- used to do metaprogramming in R. \substitute can be used to
- synthesize ASTs by stitching together user inputs. \substitute walks
- over its argument AST and replaces the symbols with their bindings in
- the specified environment and returns this new AST which can then be
- evaluated in custom environments using \eval.
-
-  \subsubsection{Convenience} Eval is used to make some programming tasks easier (and/or shorter). For example, in \c{data.table}, it is used to set R options from a character vector
+\subsubsection{Dynamic Code Loading}
+The combination of \parse and \eval is used to parse text from string
+or file and evaluate in a custom environment. This pattern is used by
+the \source Core R function to load R code from a file in the current
+workspace. This is quite useful for loading functions for interactive
+data analysis.
+
+\subsubsection{Metaprogramming}
+The combination of \eval and \substitute is used to do metaprogramming
+in R. \substitute can be used to synthesize ASTs by stitching together
+user inputs. \substitute walks over its argument AST and replaces the
+symbols with their bindings in the specified environment and returns
+this new AST which can then be evaluated in custom environments using
+\eval.
+
+\subsubsection{Convenience}
+Eval is used to make some programming tasks easier (and/or shorter).
+For example, in \c{data.table}, it is used to set R options from a
+character vector
   \begin{lstlisting}
@@
 \eval is also used in R packages to get around various limitations.
-  For example, in \datatable package, \eval is used to modify a
-  binding in the \base package of R. Bindings in the \base package are
-  locked for modification and require explicit unlocking using
-  function \unlockBinding. However, directly using this function
-  triggers a restricted-function-use warning during package build,
-  preventing the package to be accepted by CRAN. The \datatable gets
-  around this limitation by enclosing the call to \unlockBinding
-  inside \eval. From the same reason, other packages use \eval to
-  access private variables and functions from their dependencies.
+For example, in \datatable package, \eval is used to modify a binding
+in the \base package of R. Bindings in the \base package are locked
+for modification and require explicit unlocking using function
+\unlockBinding. However, directly using this function triggers a
+restricted-function-use warning during package build, preventing the
+package to be accepted by CRAN. The \datatable gets around this
+limitation by enclosing the call to \unlockBinding inside \eval. From
+the same reason, other packages use \eval to access private variables
+and functions from their dependencies.
 
@@
 \subsubsection{Unnecessary use of eval}
-  Similarly to JavaScript, there are also unnecessary uses of \eval.
-  For example, the \c{PerformanceAnalytics} package contains a function
+Similarly to JavaScript, there are also unnecessary uses of \eval. For
+example, the \c{PerformanceAnalytics} package contains a function
 \c{chart.QQPlot} that uses \eval to resolve a string into function and
@@
 - measure number of operations inside eval and outside eval.
-- this is useful because if analyzers treat eval as a nop, they may miss X\% of the code.
+
+- this is useful because if analyzers treat eval as a nop, they may
+miss X\% of the code.
 
@@
 - measure number of reads and writes by eval outside of its ``bubble''.
-- Since R is lazy, even reads can cause promise evaluation and hence, side-effects.
-- Look at writes to non-local scopes. Compare with total number of writes for
-reference?
+
+- Since R is lazy, even reads can cause promise evaluation and hence,
+side-effects.
+
+- Look at writes to non-local scopes. Compare with total number of
+writes for reference?
+
 - Categorize side-effects between env passed to eval and other envs.
@@
 \subsubsection{Does eval do reflection}
-\eval can access parent scopes via reflection. Find number of times it does that
-and the depth to which it access parent frames.
+\eval can access parent scopes via reflection. Find number of times it
+does that and the depth to which it access parent frames.
 
 \subsubsection{Does eval introduce new bindings}
-Find number of times, eval introduces new bindings. There are many ways -- library,
-load, attach, source, and explicitly introduce bindings using super-assign, assign and define.
+Find number of times, eval introduces new bindings. There are many
+ways -- library, load, attach, source, and explicitly introduce
+bindings using super-assign, assign and define.
 
 \subsubsection{Does eval call C code}
-Find number of times, eval makes call to native code using - .Call, .External,
-etc. Find out if there are native functions that are called only from within eval.
+Find number of times, eval makes call to native code using - .Call,
+.External, etc. Find out if there are native functions that are called
+only from within eval.
 
 \subsubsection{Does eval add finalizers}
-Eval can attach code to be executed when the object goes out of scope or program
-exits. Find if eval does that.
+Eval can attach code to be executed when the object goes out of scope
+or program exits. Find if eval does that.
 
 \subsubsection{Does eval do non-local returns}
-Eval can do non-local returns effective bypassing evaluation of the rest of the
-function. This can be useful for static analyzers.
+Eval can do non-local returns effective bypassing evaluation of the
+rest of the function. This can be useful for static analyzers.
 
@@
 \subsubsection{Provenance}
-
-In JavaScript, the provenance of the string passed into \eval is a matter of
-concern as \eval can be used to breach the security of websites. The main
-danger comes from string that were obtained from external sources. If they
-are allowed to flow into an \eval, then an attacker could hijack the
-program.
-
-{\bf what have they found...}
 
@@
 We observe 5.2\% cases where \eval directly evaluates the output of
-\c{substitute}, 0.7\% cases where output of \c{parse} is read directly and
-only 704 cases for \c{expression}. Most expressions consumed by \eval are
-generated by other functions.
+\c{substitute}, 0.7\% cases where output of \c{parse} is read directly
+and only 704 cases for \c{expression}. Most expressions consumed by
+\eval are generated by other functions.
 
@@
 
-An \eval is considered side-effecting if it performs a side-effecting write to a
-variable, directly or indirectly. Only \SideEffectingCoreCallPerc \eval calls in
-Core R are side-effecting and \SideEffectingPackageCallPerc \eval calls in CRAN
-packages are side-effecting.
+An \eval is considered side-effecting if it performs a side-effecting
+write to a variable, directly or indirectly. Only
+\SideEffectingCoreCallPerc \eval calls in Core R are side-effecting
+and \SideEffectingPackageCallPerc \eval calls in CRAN packages are
+side-effecting.
 
@@
 
-\begin{table}[htbp]{%} \resizebox{\columnwidth}{!}{%
-    \centering
-    \begin{tabular}{c|c|c|c}
-        \hline
-        \multicolumn{2}{c|}{Core} & \multicolumn{2}{|c}{Packages}  \\
-        \hline
-        Environment Class & Eval Calls \% & Environment Class & Eval Calls \% \\
-        \hline
+\begin{table}[htbp]{ \centering
+\begin{tabular}{c|c|c|c}\hline
+\multicolumn{2}{c|}{Core} & \multicolumn{2}{|c}{Packages}  \\\hline
+Environment Class & Eval Calls \% & Environment Class & Eval Calls\\\hline
 \CoreEnvClassA & \CoreProportionA & \PackagesEnvClassA &  \PackagesProportionA\\
@@
 \CoreEnvClassD & \CoreProportionD & \PackagesEnvClassD &  \PackagesProportionD\\
-        \CoreEnvClassE & \CoreProportionE & \PackagesEnvClassE &  \PackagesProportionE\\
-        \hline
+\CoreEnvClassE & \CoreProportionE & \PackagesEnvClassE &  \PackagesProportionE\\\hline
 \end{tabular}} \label{tab:environments}
-\caption{Environments in terms of eval calls.}
+\caption{Environments in terms of \eval calls}
 \end{table}

### 2020-11-30 inst/paper/main.tex
minor

@@
 \input{macro}
+\newcommand{\mypara}[1]{\medskip\noindent\emph{#1}\xspace}
+\newcommand{\NOTE}[1]{{\it Note: #1}\xspace}
 
@@
 
-\medskip\noindent{\it Eval in R.} The reflective interface exposed by
+\mypara{Eval in R.} The reflective interface exposed by
 R is somewhat complex. The core library exports four functions with
@@
 
-\medskip\noindent\emph{CRAN Packages.} The packages that are included
+\mypara{CRAN Packages.} The packages that are included
 in this study are the top \CorpusPackages packages based on their
@@
 
-\medskip\noindent\emph{Kaggle Scripts.} Kaggle is an online platform
+\mypara{Kaggle Scripts.} Kaggle is an online platform
 for data-science and machine-learning. The website allows people to
@@
 
-\begin{figure}[!h]\hspace{-5mm}\includegraphics[width=1.05\linewidth]
-{pipeline.pdf}\caption{Analysis Pipeline}\label{fig:pipeline}
+\begin{figure}[!h]\hspace{-5mm}
+  \includegraphics[width=1.05\linewidth]{pipeline.pdf}
+  \caption{Analysis Pipeline}\label{fig:pipeline}
 \end{figure}
@@
 
-\medskip\noindent\emph{Limitations.} Dynamic analysis can only observe
+\mypara{Limitations.} Dynamic analysis can only observe
 calls that are triggered by the program's input. We believe that
@@
 
-\newcommand{\mypara}[1]{\medskip\noindent\emph{#1}\xspace}
-
 %%%%%
@@
 higher-order function such as \c{map}.
-
-\newcommand{\NOTE}[1]{{\it Note: #1}\xspace}
 
@@
 
-\subsubsection{Unnecessary use of eval}
+\subsection{Power of Eval}
+
+\subsubsection{How much code is executed inside eval}
+- measure number of operations inside eval and outside eval.
+
+- this is useful because if analyzers treat eval as a nop, they may
+miss X\% of the code.
+
+\subsubsection{Does eval create observable side-effects}
+- measure number of reads and writes by eval outside of its ``bubble''.
+
+- Since R is lazy, even reads can cause promise evaluation and hence,
+side-effects.
+
+- Look at writes to non-local scopes. Compare with total number of
+writes for reference?
+
+- Categorize side-effects between env passed to eval and other envs.
+
+\subsubsection{Does eval do reflection}
+\eval can access parent scopes via reflection. Find number of times it
+does that and the depth to which it access parent frames.
+
+\subsubsection{Does eval introduce new bindings}
+Find number of times, eval introduces new bindings. There are many
+ways -- library, load, attach, source, and explicitly introduce
+bindings using super-assign, assign and define.
+
+\subsubsection{Does eval call C code}
+Find number of times, eval makes call to native code using - .Call,
+.External, etc. Find out if there are native functions that are called
+only from within eval.
+
+\subsubsection{Does eval add finalizers}
+Eval can attach code to be executed when the object goes out of scope
+or program exits. Find if eval does that.
+
+\subsubsection{Does eval do non-local returns}
+Eval can do non-local returns effective bypassing evaluation of the
+rest of the function. This can be useful for static analyzers.
+
+Conclude with number of evals that are ``pure'', i.e. evals which
+could be ignored by a static analyzer without any problem.
+
+\subsubsection{Provenance}
+
+There are various ways to obtain expressions:
+\begin{itemize}
+  \item \c{substitute} synthesizes ASTs from expressions by replacing
+    symbols with their bindings in the specified environment.
+  \item \c{expression}  creates a vector of expression
+    objects from text.
+  \item \c{parse}, \c{str2expression} and \c{str2lang} turn strings into
+    expressions.
+\end{itemize}
+
+We observe 5.2\% cases where \eval directly evaluates the output of
+\c{substitute}, 0.7\% cases where output of \c{parse} is read directly
+and only 704 cases for \c{expression}. Most expressions consumed by
+\eval are generated by other functions.
+
+\c{eval(parse(...))} can be used for dynamic code loading. This forms
+the core of \c{source} and \c{sys.source} functions in R that are
+commonly used for loading code in R files in interactive settings. We
+investigated the number of cases in which the output of \c{parse} and
+its variants is passed to \eval, directly or transitively by tainting
+their output. This corresponds to \PercentParsedCallSites of the total
+\eval call sites and \PercentParsedEvals of the eval calls. We
+observed that very few of the eval call (\NbParseFilesRnd in total)
+consume the result of calling \c{parse} on a file. Most of the eval
+calls consume the result of calling \c{parse} on a string. We also
+identified one function in core R, \c{invokeRestartInteractively} that
+prompts the user for input, parses it, and passes it to \eval.
+
+\subsubsection{Side-Effects}
+
+In the corpus we observe \AllWritesRnd writes to variables of which
+\EvalWritesRnd writes happen inside \eval. However, all writes are not
+dangerous. Only writes to environments not local to the computation
+spawned by \eval are side-effecting. These writes outlive the
+computation and hence are visible outside it. The remaining writes are
+local to the computation. We observe that \EvalSideEffectingWritesRnd
+writes inside evals are side-effecting. This is only
+\EvalSideEffectingWritesEvalPerc of all variable writes inside \eval
+and \EvalSideEffectingWritesAllPerc of all variable writes in the
+corpus.
+
+An \eval is considered side-effecting if it performs a side-effecting
+write to a variable, directly or indirectly. Only
+\SideEffectingCoreCallPerc \eval calls in Core R are side-effecting
+and \SideEffectingPackageCallPerc \eval calls in CRAN packages are
+side-effecting.
+
+\subsubsection{Scope}
+
+One of the main differences of the R \eval with the javascript one is
+how it interacts with \emph{scope}, or environments in R. In
+Javascript, \eval can access the local scope and the global scope. In
+R, \eval has an argument \c{envir} to indicate in which environment
+its first argument must be evaluated. By default, it is equal to
+\c{parent.frame()}, which is the parent environment of \eval call.
+This default argument happens in \DefaultEnvirExprPercent of the eval
+calls.
+
+This argument can also be \c{NULL}, a \c{List} or a data frame. This
+happens in 3.5\% of the cases: \eval copies the fields of the list or
+data frame and creates bindings for them in a new environment. This
+pattern is used to evaluate formulas which can directly refer to the
+fields of the data. The \c{envir} argument can also be a number $n$.
+It means that the environment in which the expressions is evaluated
+will be the result of \c{sys.call(n)} where $n$ refers to the $n$-th
+stack frame.
+
+The top-level environment in R is called the global environment. New
+environments can be created using \c{new.env}. They can be provided a
+parent environment which becomes the enclosing scope of the new
+environment.
+
+We looked at the environments passed to all the \eval calls in our
+corpus. Table~\ref{tab:environments} summarizes the results. A numeric
+environment class \c{n} denotes the environment of the $n$-th call
+stack frame from the current function. \c{global} denotes the
+top-level environment and \c{list} denotes a list passed for
+evaluation of formulas. Environment classes of the form $n+$ denote
+the $n$-th environment extended with a new environment. The new
+environment provides a limited form of sandboxing. All assignments
+using the \c{\<-} function occur inside it and prevent the extended
+environment from mutation. However, it is still possible to mutate the
+extended environment using the \c{\<\<-} or \c{assign} functions; but,
+that happens rarely.
+
+\begin{table}[htbp]{ \centering
+\begin{tabular}{c|c|c|c}\hline
+\multicolumn{2}{c|}{Core} & \multicolumn{2}{|c}{Packages}  \\\hline
+Environment Class & Eval Calls \% & Environment Class & Eval Calls\\\hline
+\CoreEnvClassA & \CoreProportionA & \PackagesEnvClassA &  \PackagesProportionA\\
+\CoreEnvClassB & \CoreProportionB & \PackagesEnvClassB &  \PackagesProportionB\\
+\CoreEnvClassC & \CoreProportionC & \PackagesEnvClassC &  \PackagesProportionC\\
+\CoreEnvClassD & \CoreProportionD & \PackagesEnvClassD &  \PackagesProportionD\\
+\CoreEnvClassE & \CoreProportionE & \PackagesEnvClassE &  \PackagesProportionE\\\hline
+\end{tabular}} \label{tab:environments}
+\caption{Environments in terms of \eval calls}
+\end{table}
+
+We observe that a disproportionately high number of core R \eval calls
+access the caller's caller's environment. This is because many core R
+functions call functions that pass the result of \c{parent.frame()} to
+an \eval. A disproportionately high number of calls to \eval happens
+in an extended top-level environment. This can be explained by the
+fact that many packages evaluate code passed from the user's workspace
+in the top-level environment to access the bindings.
+
+The \c{imchange} function of package \c{imager} makes it possible to
+modify images using a dedicated formula syntax using
+\c{\~}.\footnote{Inspired by {map} in package \emph{purr}.} Here,
+\eval is evaluated in \c{newenv}, which creates a new environment that
+inherits from \c{parent.frame()} by default (classified as 1+).
+
+\begin{lstlisting}
+newenv <- new.env()
+...
+fo <- parse(text=as.character(fo)[2])
+im[where] <- eval(fo,envir=newenv,enclos=env)
+\end{lstlisting}
+
+\c{adjCoef} in package \emph{actuar} find the root of an equation
+defined by a function \c{h} whose arguments must be named \c{x} and
+\c{y}. \c{h} is transformed into an auxiliary function \c{h2} that can
+be optimized. Here, the list used for \c{envir} ensures the
+correspondance between the textual arguments of \c{h} and the
+arguments of \c{h2}.
+
+\begin{lstlisting}
+sh <- substitute(h)
+fcall <- paste(sh, "(x, y)")
+...
+h2 <- function(x, y)
+    eval(parse(text = fcall),
+    envir = list(x = x, y = y),
+    enclos = parent.frame(2))
+\end{lstlisting}
+
+\section{Case Studies}
+
+5(?) examples from real code.
+
+\section{Why do we Eval}
+
+\subsection{Discussion}
+
+Yeah, why?
+
+\subsection{Can we do without?}
+
+We look at how \emph{consistent} the \c{expr} argument of \eval can
+be, \ie how many different types of the resolved \c{expr} there are
+per call sites. Most of the call sites, \ie \PercentMonomorphic, are
+\emph{consistent}, and this is similar to javascript. However, a few
+ones are highly \emph{polymorphic} (10 different types). They are the
+pipe operators \c{\%>\%}, \c{\%<>\%} and \c{\%\$\%} in package
+\emph{magrittr}. It is effectively used to compose functions on their
+first argument, which can be of any type.
+
 Similarly to JavaScript, there are also unnecessary uses of \eval. For
@@
 
-\subsection{Power of Eval}
-
-\subsubsection{How much code is executed inside eval}
-- measure number of operations inside eval and outside eval.
-
-- this is useful because if analyzers treat eval as a nop, they may
-miss X\% of the code.
-
-\subsubsection{Does eval create observable side-effects}
-- measure number of reads and writes by eval outside of its ``bubble''.
-
-- Since R is lazy, even reads can cause promise evaluation and hence,
-side-effects.
-
-- Look at writes to non-local scopes. Compare with total number of
-writes for reference?
-
-- Categorize side-effects between env passed to eval and other envs.
-
-\subsubsection{Does eval do reflection}
-\eval can access parent scopes via reflection. Find number of times it
-does that and the depth to which it access parent frames.
-
-\subsubsection{Does eval introduce new bindings}
-Find number of times, eval introduces new bindings. There are many
-ways -- library, load, attach, source, and explicitly introduce
-bindings using super-assign, assign and define.
-
-\subsubsection{Does eval call C code}
-Find number of times, eval makes call to native code using - .Call,
-.External, etc. Find out if there are native functions that are called
-only from within eval.
-
-\subsubsection{Does eval add finalizers}
-Eval can attach code to be executed when the object goes out of scope
-or program exits. Find if eval does that.
-
-\subsubsection{Does eval do non-local returns}
-Eval can do non-local returns effective bypassing evaluation of the
-rest of the function. This can be useful for static analyzers.
-
-Conclude with number of evals that are ``pure'', i.e. evals which
-could be ignored by a static analyzer without any problem.
-
-\subsubsection{Provenance}
-
-There are various ways to obtain expressions:
-\begin{itemize}
-  \item \c{substitute} synthesizes ASTs from expressions by replacing
-    symbols with their bindings in the specified environment.
-  \item \c{expression}  creates a vector of expression
-    objects from text.
-  \item \c{parse}, \c{str2expression} and \c{str2lang} turn strings into
-    expressions.
-\end{itemize}
-
-We observe 5.2\% cases where \eval directly evaluates the output of
-\c{substitute}, 0.7\% cases where output of \c{parse} is read directly
-and only 704 cases for \c{expression}. Most expressions consumed by
-\eval are generated by other functions.
-
-\c{eval(parse(...))} can be used for dynamic code loading. This forms
-the core of \c{source} and \c{sys.source} functions in R that are
-commonly used for loading code in R files in interactive settings. We
-investigated the number of cases in which the output of \c{parse} and
-its variants is passed to \eval, directly or transitively by tainting
-their output. This corresponds to \PercentParsedCallSites of the total
-\eval call sites and \PercentParsedEvals of the eval calls. We
-observed that very few of the eval call (\NbParseFilesRnd in total)
-consume the result of calling \c{parse} on a file. Most of the eval
-calls consume the result of calling \c{parse} on a string. We also
-identified one function in core R, \c{invokeRestartInteractively} that
-prompts the user for input, parses it, and passes it to \eval.
-
-\subsubsection{Side-Effects}
-
-In the corpus we observe \AllWritesRnd writes to variables of which
-\EvalWritesRnd writes happen inside \eval. However, all writes are not
-dangerous. Only writes to environments not local to the computation
-spawned by \eval are side-effecting. These writes outlive the
-computation and hence are visible outside it. The remaining writes are
-local to the computation. We observe that \EvalSideEffectingWritesRnd
-writes inside evals are side-effecting. This is only
-\EvalSideEffectingWritesEvalPerc of all variable writes inside \eval
-and \EvalSideEffectingWritesAllPerc of all variable writes in the
-corpus.
-
-An \eval is considered side-effecting if it performs a side-effecting
-write to a variable, directly or indirectly. Only
-\SideEffectingCoreCallPerc \eval calls in Core R are side-effecting
-and \SideEffectingPackageCallPerc \eval calls in CRAN packages are
-side-effecting.
-
-\subsubsection{Scope}
-
-One of the main differences of the R \eval with the javascript one is
-how it interacts with \emph{scope}, or environments in R. In
-Javascript, \eval can access the local scope and the global scope. In
-R, \eval has an argument \c{envir} to indicate in which environment
-its first argument must be evaluated. By default, it is equal to
-\c{parent.frame()}, which is the parent environment of \eval call.
-This default argument happens in \DefaultEnvirExprPercent of the eval
-calls.
-
-This argument can also be \c{NULL}, a \c{List} or a data frame. This
-happens in 3.5\% of the cases: \eval copies the fields of the list or
-data frame and creates bindings for them in a new environment. This
-pattern is used to evaluate formulas which can directly refer to the
-fields of the data. The \c{envir} argument can also be a number $n$.
-It means that the environment in which the expressions is evaluated
-will be the result of \c{sys.call(n)} where $n$ refers to the $n$-th
-stack frame.
-
-The top-level environment in R is called the global environment. New
-environments can be created using \c{new.env}. They can be provided a
-parent environment which becomes the enclosing scope of the new
-environment.
-
-We looked at the environments passed to all the \eval calls in our
-corpus. Table~\ref{tab:environments} summarizes the results. A numeric
-environment class \c{n} denotes the environment of the $n$-th call
-stack frame from the current function. \c{global} denotes the
-top-level environment and \c{list} denotes a list passed for
-evaluation of formulas. Environment classes of the form $n+$ denote
-the $n$-th environment extended with a new environment. The new
-environment provides a limited form of sandboxing. All assignments
-using the \c{\<-} function occur inside it and prevent the extended
-environment from mutation. However, it is still possible to mutate the
-extended environment using the \c{\<\<-} or \c{assign} functions; but,
-that happens rarely.
-
-\begin{table}[htbp]{ \centering
-\begin{tabular}{c|c|c|c}\hline
-\multicolumn{2}{c|}{Core} & \multicolumn{2}{|c}{Packages}  \\\hline
-Environment Class & Eval Calls \% & Environment Class & Eval Calls\\\hline
-\CoreEnvClassA & \CoreProportionA & \PackagesEnvClassA &  \PackagesProportionA\\
-\CoreEnvClassB & \CoreProportionB & \PackagesEnvClassB &  \PackagesProportionB\\
-\CoreEnvClassC & \CoreProportionC & \PackagesEnvClassC &  \PackagesProportionC\\
-\CoreEnvClassD & \CoreProportionD & \PackagesEnvClassD &  \PackagesProportionD\\
-\CoreEnvClassE & \CoreProportionE & \PackagesEnvClassE &  \PackagesProportionE\\\hline
-\end{tabular}} \label{tab:environments}
-\caption{Environments in terms of \eval calls}
-\end{table}
-
-We observe that a disproportionately high number of core R \eval calls
-access the caller's caller's environment. This is because many core R
-functions call functions that pass the result of \c{parent.frame()} to
-an \eval. A disproportionately high number of calls to \eval happens
-in an extended top-level environment. This can be explained by the
-fact that many packages evaluate code passed from the user's workspace
-in the top-level environment to access the bindings.
-
-The \c{imchange} function of package \c{imager} makes it possible to
-modify images using a dedicated formula syntax using
-\c{\~}.\footnote{Inspired by {map} in package \emph{purr}.} Here,
-\eval is evaluated in \c{newenv}, which creates a new environment that
-inherits from \c{parent.frame()} by default (classified as 1+).
-
-\begin{lstlisting}
-newenv <- new.env()
-...
-fo <- parse(text=as.character(fo)[2])
-im[where] <- eval(fo,envir=newenv,enclos=env)
-\end{lstlisting}
-
-\c{adjCoef} in package \emph{actuar} find the root of an equation
-defined by a function \c{h} whose arguments must be named \c{x} and
-\c{y}. \c{h} is transformed into an auxiliary function \c{h2} that can
-be optimized. Here, the list used for \c{envir} ensures the
-correspondance between the textual arguments of \c{h} and the
-arguments of \c{h2}.
-
-\begin{lstlisting}
-sh <- substitute(h)
-fcall <- paste(sh, "(x, y)")
-...
-h2 <- function(x, y)
-    eval(parse(text = fcall),
-    envir = list(x = x, y = y),
-    enclos = parent.frame(2))
-\end{lstlisting}
-
-\section{Case Studies}
-
-\section{Why do we Eval}
-
 \section{Analysis Results}
@@
 
-\medskip\noindent\emph{Discussion:} Comparing the use of \eval in R
-and JavaScript is instructive. The use of \eval is more widespread in
-data science code than it was in JavaScript. Any run of an R program
-seems to trigger \eval; moreover package programmers use \eval
-liberally. On the other hand, the authors of Kaggle packages do not
-use the feature at all. This point to a bifurcated user community:
-experts who write package know how to use \eval while end-users do
-not. An alternative explanation is that end-users do not need \eval
-for simple data analysis tasks. This bifurcation was not observed in
-the JavaScript community, but this may have changed with the advent of
-Node.js ecosystem. The distribution of input sizes is similar to that
-found in JavaScript, the majority of inputs being rather small.
-
-\begin{figure}[!b]
-	\centering
-  \includegraphics[width=\columnwidth]{ast_sizes}
-  \caption{Input sizes} \label{fig:ast-size}
-\end{figure}
-
-To roughly estimate the complexity of the code being evaluated,
-Fig.~\ref{fig:ast-size} shows the distribution of the abstract syntax
-tree size of the \c{expr} argument to \eval. The majority of inputs
-across core and packages (\NbAstOnePercent of \eval calls and
-\NbAstOneCallSitePercent of call sites) have size 1. That is to say,
-the code being evaluated consists of a single variable name or
-constant. A function call needs at least three nodes and a list such
-as \c{c(1, 2, 3)} counts for 4 nodes. Inputs smaller than
-\AstSizeNineFive nodes account for 95\% of all calls. Sizes in core
-are larger than in packages. Most large ASTs originate from library
-\c{VGAM}, which is used for statistical modeling, and also calls
-functions using \eval from core. It uses \eval to evaluate large
-matrices with values in a given environment.
-
-\subsubsection{Consistency}
-
-We look at how \emph{consistent} the \c{expr} argument of \eval can
-be, \ie how many different types of the resolved \c{expr} there are
-per call sites. Most of the call sites, \ie \PercentMonomorphic, are
-\emph{consistent}, and this is similar to javascript. However, a few
-ones are highly \emph{polymorphic} (10 different types). They are the
-pipe operators \c{\%>\%}, \c{\%<>\%} and \c{\%\$\%} in package
-\emph{magrittr}. It is effectively used to compose functions on their
-first argument, which can be of any type.
+Comparing the use of \eval in R and JavaScript is instructive. The use
+of \eval is more widespread in data science code than it was in
+JavaScript. Any run of an R program seems to trigger \eval; moreover
+package programmers use \eval liberally. On the other hand, the
+authors of Kaggle packages do not use the feature at all. This point
+to a bifurcated user community: experts who write package know how to
+use \eval while end-users do not. An alternative explanation is that
+end-users do not need \eval for simple data analysis tasks. This
+bifurcation was not observed in the JavaScript community, but this may
+have changed with the advent of Node.js ecosystem. The distribution of
+input sizes is similar to that found in JavaScript, the majority of
+inputs being rather small.
 

### 2020-11-30 inst/paper/main.tex
random walk

@@
 
+\NOTE{Talk about the percentage of code executed in eval}
+
 \mypara{The case of CORE libraries} ... \NOTE{Give numbers for the
@@
 in this section we try to elucidate \emph{what} it does.
+
+\mypara{Side-effects.}
+XXX026 total number of reads performed by \eval. \NOTE{What is the definition of a read?  An env lookup?}  XXX027 average number of reads.
+XXX028 total number of writes performed by \eval. Average number XXX029.
+
+\NOTE{Aviral wrote''measure number of reads and writes by eval outside
+  of its bubble'' --- what is the bubble here?}
+
+XXX029 reads to caller environment -- where caller is the caller of \eval.
+
+XXX030 reads to other caller environment -- i.e. envs that are on the call stack.
+
+XXX031 reads to caller environment -- where caller is the caller of \eval.
+
+XXX032 reads to other caller environment -- i.e. envs that are on the call stack.
+
+Categorize side-effects between env passed to eval and other envs:
+\begin{itemize}
+\item ephemeral -- read to envs that are created during \eval  XXX033
+\item local -- reads to env of the function in which the call to \eval
+  lexically occurred; XXX034
+\item parent -- reads to parent envs of the local one. XXX035
+\item call-stack -- reads to envs that are on the call stack but not
+  caller. XXX035.
+\end{itemize}
+
+\subsubsection{Does eval introduce new bindings}
+Find number of times, eval introduces new bindings. There are many
+ways -- library, load, attach, source, and explicitly introduce
+bindings using super-assign, assign and define.
+
+\subsubsection{Does eval call C code}
+Find number of times, eval makes call to native code using - .Call,
+.External, etc. Find out if there are native functions that are called
+only from within eval.
+
+\subsubsection{Does eval add finalizers}
+Eval can attach code to be executed when the object goes out of scope
+or program exits. Find if eval does that.
+
+\subsubsection{Does eval do non-local returns}
+Eval can do non-local returns effective bypassing evaluation of the
+rest of the function. This can be useful for static analyzers.
+
+Conclude with number of evals that are ``pure'', i.e. evals which
+could be ignored by a static analyzer without any problem.
+
+\subsection{LEGACY TEXT}
 
@@
 
+\subsubsection{Usage Patterns}
+
+Richards et al.~\cite{ecoop11} identified a number of frequently
+occurring \eval patterns which were detected using simple regular
+expressions on the \eval strings. In the case of R we match the
+patterns on the recorded AST.
+
+Table~\ref{table:js-pattens-in-r} shows the ratio of \eval in each of
+these patterns. We included all the original patterns except for
+\emph{JSON} as R uses its own serialization format and for
+\emph{empty} as there were no empty expression passed to \eval:
+
+\begin{table}[ht]%\resizebox{\columnwidth}{!}{%
+  \centering
+\begin{tabular}{r|r|r|r|r}\hline
+\multirow{2}{*}{Pattern} & \multicolumn{2}{c|}{Core} & \multicolumn{2}{c}{Packages} \\
+                       & Calls & Ratio            & Calls & Ratio \\\hline
+\input{tag/table-js-patterns-in-r.tex}
+\end{tabular}
+\caption{JavaScript \eval Patterns in R}
+\label{table:js-pattens-in-r}
+\end{table}
+
+\noindent \emph{Library}, \emph{Typeof} and \emph{Try} are all barely
+used mostly because the language provides dedicated features for it.
+The few calls in Library pattern come mostly from file loading
+functions such as \c{source} and \c{sys.source}.
+%
+\noindent
+\emph{Read} pattern in JavaScript represented mostly an incorrect
+access to an object's property or local variable both replaceable by
+JavaScript core functions. In the case of R there are \PatternReadRnd
+read \eval calls (\PatternReadRatio). They come from calls like
+\c{subset(airquality, select = Temp)} where the variable \c{Temp} will
+be evaluated in the \c{airquality} data frame.
+
+%
+\noindent \emph{Assign} category comprises of assignments to variables
+within an \eval call. We have seen \PatternAssignRnd such calls
+(\PatternAssignRatio). \PatternAssignArrowRatio uses the simple
+assignment (\c{<-}). The majority of cases comes from the
+\c{magrittr}'s \c{\%<>\%} pipe operator
+(\PatternAssignArrowMagrittrRatio) which forwards an object to a
+function on rhs and update lhs with the result. Far more calls used
+the super assignment (\c{<<-}) operator, but almost all come from the
+\c{plyr} package which provides a convenient API for common data
+manipulation problems.
+%
+\noindent \emph{Call} and \emph{Other} together represents the vast
+majority of eval calls in R. The \emph{Other} pattern represents all
+the calls where eval was called with a value instead of with an
+expression. The \emph{Call} pattern contains all the calls that did
+not fall in the previous categories. Without a context it is usually
+difficult to further categorize the intent of these calls. Therefore
+we have randomly chosen \PatternManualPackages packages and manually
+looked at \PatternManualCallsites call sites from which we have
+observed the following categories:
+
+\subsubsection{Recording the current call}
+R provides a \c{match.call} function which can return the current
+call. This is heavily used by the statistical model-fitting functions
+that records the current call that is then reevaluated in the
+different environment. Another pattern is to match the current call,
+adjust some arguments and pass it to another function. For example the
+\c{write.csv} function changes the \c{sep} argument in the matched
+call and reevaluates itself as \c{write.table}:
+%
+\begin{lstlisting}
+    Call <- match.call(expand.dots = TRUE)
+    Call$sep <- ","; Call[[1L]] <- as.name("write.table")
+    eval.parent(Call)
+\end{lstlisting}
+
+\subsubsection{Sandboxing} The combination of \eval and first class
+environments can be used to construct a sandbox for evaluating
+side-effecting expressions. The newly constructed environment can use
+the current environment for its lexical scope. This provides the
+evaluated expression access to all the bindings in the current scope
+but restricts the side-effect to the newly constructed throwaway
+environment. However, this sandboxing is limited because the
+expression can reflectively access any environment in the call stack
+or use the super assignment operator(\c{<<-}) to write to parent
+scope. The \c{local} function which is part of Core R is an example of
+this usage pattern.
+
+\subsubsection{Dynamic Code Loading}
+The combination of \parse and \eval is used to parse text from string
+or file and evaluate in a custom environment. This pattern is used by
+the \source Core R function to load R code from a file in the current
+workspace. This is quite useful for loading functions for interactive
+data analysis.
+
+\subsubsection{Metaprogramming}
+The combination of \eval and \substitute is used to do metaprogramming
+in R. \substitute can be used to synthesize ASTs by stitching together
+user inputs. \substitute walks over its argument AST and replaces the
+symbols with their bindings in the specified environment and returns
+this new AST which can then be evaluated in custom environments using
+\eval.
+
+\subsubsection{Convenience}
+Eval is used to make some programming tasks easier (and/or shorter).
+For example, in \c{data.table}, it is used to set R options from a
+character vector
+  \begin{lstlisting}
+opts = c("datatable.verbose"="FALSE", ...) for (i in names(opts))
+eval(parse(text=paste0("options(",i,"=",opts[i],")")))
+
+  \end{lstlisting}
+  %
+instead of manually calling \c{options(datatable.verbose=FALSE)}
+
+\subsubsection{Implementation}
+The package installation process adds an extra file to the package
+directory that contains the code to load the package code inside a
+\local block. It creates a namespace object and injects the package
+bindings.
+
+\subsubsection{Miscellaneous}
+\eval is also used in R packages to get around various limitations.
+For example, in \datatable package, \eval is used to modify a binding
+in the \base package of R. Bindings in the \base package are locked
+for modification and require explicit unlocking using function
+\unlockBinding. However, directly using this function triggers a
+restricted-function-use warning during package build, preventing the
+package to be accepted by CRAN. The \datatable gets around this
+limitation by enclosing the call to \unlockBinding inside \eval. From
+the same reason, other packages use \eval to access private variables
+and functions from their dependencies.
+
+\subsection{Necessity of Eval}
+
+\subsubsection{Provenance}
+
+There are various ways to obtain expressions:
+\begin{itemize}
+  \item \c{substitute} synthesizes ASTs from expressions by replacing
+    symbols with their bindings in the specified environment.
+  \item \c{expression}  creates a vector of expression
+    objects from text.
+  \item \c{parse}, \c{str2expression} and \c{str2lang} turn strings into
+    expressions.
+\end{itemize}
+
+We observe 5.2\% cases where \eval directly evaluates the output of
+\c{substitute}, 0.7\% cases where output of \c{parse} is read directly
+and only 704 cases for \c{expression}. Most expressions consumed by
+\eval are generated by other functions.
+
+\c{eval(parse(...))} can be used for dynamic code loading. This forms
+the core of \c{source} and \c{sys.source} functions in R that are
+commonly used for loading code in R files in interactive settings. We
+investigated the number of cases in which the output of \c{parse} and
+its variants is passed to \eval, directly or transitively by tainting
+their output. This corresponds to \PercentParsedCallSites of the total
+\eval call sites and \PercentParsedEvals of the eval calls. We
+observed that very few of the eval call (\NbParseFilesRnd in total)
+consume the result of calling \c{parse} on a file. Most of the eval
+calls consume the result of calling \c{parse} on a string. We also
+identified one function in core R, \c{invokeRestartInteractively} that
+prompts the user for input, parses it, and passes it to \eval.
+
+\subsubsection{Side-Effects}
+
+In the corpus we observe \AllWritesRnd writes to variables of which
+\EvalWritesRnd writes happen inside \eval. However, all writes are not
+dangerous. Only writes to environments not local to the computation
+spawned by \eval are side-effecting. These writes outlive the
+computation and hence are visible outside it. The remaining writes are
+local to the computation. We observe that \EvalSideEffectingWritesRnd
+writes inside evals are side-effecting. This is only
+\EvalSideEffectingWritesEvalPerc of all variable writes inside \eval
+and \EvalSideEffectingWritesAllPerc of all variable writes in the
+corpus.
+
+An \eval is considered side-effecting if it performs a side-effecting
+write to a variable, directly or indirectly. Only
+\SideEffectingCoreCallPerc \eval calls in Core R are side-effecting
+and \SideEffectingPackageCallPerc \eval calls in CRAN packages are
+side-effecting.
+
+\subsubsection{Scope}
+
+One of the main differences of the R \eval with the javascript one is
+how it interacts with \emph{scope}, or environments in R. In
+Javascript, \eval can access the local scope and the global scope. In
+R, \eval has an argument \c{envir} to indicate in which environment
+its first argument must be evaluated. By default, it is equal to
+\c{parent.frame()}, which is the parent environment of \eval call.
+This default argument happens in \DefaultEnvirExprPercent of the eval
+calls.
+
+This argument can also be \c{NULL}, a \c{List} or a data frame. This
+happens in 3.5\% of the cases: \eval copies the fields of the list or
+data frame and creates bindings for them in a new environment. This
+pattern is used to evaluate formulas which can directly refer to the
+fields of the data. The \c{envir} argument can also be a number $n$.
+It means that the environment in which the expressions is evaluated
+will be the result of \c{sys.call(n)} where $n$ refers to the $n$-th
+stack frame.
+
+The top-level environment in R is called the global environment. New
+environments can be created using \c{new.env}. They can be provided a
+parent environment which becomes the enclosing scope of the new
+environment.
+
+We looked at the environments passed to all the \eval calls in our
+corpus. Table~\ref{tab:environments} summarizes the results. A numeric
+environment class \c{n} denotes the environment of the $n$-th call
+stack frame from the current function. \c{global} denotes the
+top-level environment and \c{list} denotes a list passed for
+evaluation of formulas. Environment classes of the form $n+$ denote
+the $n$-th environment extended with a new environment. The new
+environment provides a limited form of sandboxing. All assignments
+using the \c{\<-} function occur inside it and prevent the extended
+environment from mutation. However, it is still possible to mutate the
+extended environment using the \c{\<\<-} or \c{assign} functions; but,
+that happens rarely.
+
+\begin{table}[htbp]{ \centering
+\begin{tabular}{c|c|c|c}\hline
+\multicolumn{2}{c|}{Core} & \multicolumn{2}{|c}{Packages}  \\\hline
+Environment Class & Eval Calls \% & Environment Class & Eval Calls\\\hline
+\CoreEnvClassA & \CoreProportionA & \PackagesEnvClassA &  \PackagesProportionA\\
+\CoreEnvClassB & \CoreProportionB & \PackagesEnvClassB &  \PackagesProportionB\\
+\CoreEnvClassC & \CoreProportionC & \PackagesEnvClassC &  \PackagesProportionC\\
+\CoreEnvClassD & \CoreProportionD & \PackagesEnvClassD &  \PackagesProportionD\\
+\CoreEnvClassE & \CoreProportionE & \PackagesEnvClassE &  \PackagesProportionE\\\hline
+\end{tabular}} \label{tab:environments}
+\caption{Environments in terms of \eval calls}
+\end{table}
+
+We observe that a disproportionately high number of core R \eval calls
+access the caller's caller's environment. This is because many core R
+functions call functions that pass the result of \c{parent.frame()} to
+an \eval. A disproportionately high number of calls to \eval happens
+in an extended top-level environment. This can be explained by the
+fact that many packages evaluate code passed from the user's workspace
+in the top-level environment to access the bindings.
+
+The \c{imchange} function of package \c{imager} makes it possible to
+modify images using a dedicated formula syntax using
+\c{\~}.\footnote{Inspired by {map} in package \emph{purr}.} Here,
+\eval is evaluated in \c{newenv}, which creates a new environment that
+inherits from \c{parent.frame()} by default (classified as 1+).
+
+\begin{lstlisting}
+newenv <- new.env()
+...
+fo <- parse(text=as.character(fo)[2])
+im[where] <- eval(fo,envir=newenv,enclos=env)
+\end{lstlisting}
+
+\c{adjCoef} in package \emph{actuar} find the root of an equation
+defined by a function \c{h} whose arguments must be named \c{x} and
+\c{y}. \c{h} is transformed into an auxiliary function \c{h2} that can
+be optimized. Here, the list used for \c{envir} ensures the
+correspondance between the textual arguments of \c{h} and the
+arguments of \c{h2}.
+
+\begin{lstlisting}
+sh <- substitute(h)
+fcall <- paste(sh, "(x, y)")
+...
+h2 <- function(x, y)
+    eval(parse(text = fcall),
+    envir = list(x = x, y = y),
+    enclos = parent.frame(2))
+\end{lstlisting}
+
+\section{Case Studies}
+
+\NOTE{5(?) examples from real code?}
+
+\subsection{LEGACY TEXT}
+
 We looked at the top ten expressions passed to core and package \eval
@@
 
-\subsubsection{Usage Patterns}
-
-Richards et al.~\cite{ecoop11} identified a number of frequently
-occurring \eval patterns which were detected using simple regular
-expressions on the \eval strings. In the case of R we match the
-patterns on the recorded AST.
-
-Table~\ref{table:js-pattens-in-r} shows the ratio of \eval in each of
-these patterns. We included all the original patterns except for
-\emph{JSON} as R uses its own serialization format and for
-\emph{empty} as there were no empty expression passed to \eval:
-
-\begin{table}[ht]%\resizebox{\columnwidth}{!}{%
-  \centering
-\begin{tabular}{r|r|r|r|r}\hline
-\multirow{2}{*}{Pattern} & \multicolumn{2}{c|}{Core} & \multicolumn{2}{c}{Packages} \\
-                       & Calls & Ratio            & Calls & Ratio \\\hline
-\input{tag/table-js-patterns-in-r.tex}
-\end{tabular}
-\caption{JavaScript \eval Patterns in R}
-\label{table:js-pattens-in-r}
-\end{table}
-
-\noindent \emph{Library}, \emph{Typeof} and \emph{Try} are all barely
-used mostly because the language provides dedicated features for it.
-The few calls in Library pattern come mostly from file loading
-functions such as \c{source} and \c{sys.source}.
-%
-\noindent
-\emph{Read} pattern in JavaScript represented mostly an incorrect
-access to an object's property or local variable both replaceable by
-JavaScript core functions. In the case of R there are \PatternReadRnd
-read \eval calls (\PatternReadRatio). They come from calls like
-\c{subset(airquality, select = Temp)} where the variable \c{Temp} will
-be evaluated in the \c{airquality} data frame.
-
-%
-\noindent \emph{Assign} category comprises of assignments to variables
-within an \eval call. We have seen \PatternAssignRnd such calls
-(\PatternAssignRatio). \PatternAssignArrowRatio uses the simple
-assignment (\c{<-}). The majority of cases comes from the
-\c{magrittr}'s \c{\%<>\%} pipe operator
-(\PatternAssignArrowMagrittrRatio) which forwards an object to a
-function on rhs and update lhs with the result. Far more calls used
-the super assignment (\c{<<-}) operator, but almost all come from the
-\c{plyr} package which provides a convenient API for common data
-manipulation problems.
-%
-\noindent \emph{Call} and \emph{Other} together represents the vast
-majority of eval calls in R. The \emph{Other} pattern represents all
-the calls where eval was called with a value instead of with an
-expression. The \emph{Call} pattern contains all the calls that did
-not fall in the previous categories. Without a context it is usually
-difficult to further categorize the intent of these calls. Therefore
-we have randomly chosen \PatternManualPackages packages and manually
-looked at \PatternManualCallsites call sites from which we have
-observed the following categories:
-
-\subsubsection{Recording the current call}
-R provides a \c{match.call} function which can return the current
-call. This is heavily used by the statistical model-fitting functions
-that records the current call that is then reevaluated in the
-different environment. Another pattern is to match the current call,
-adjust some arguments and pass it to another function. For example the
-\c{write.csv} function changes the \c{sep} argument in the matched
-call and reevaluates itself as \c{write.table}:
-%
-\begin{lstlisting}
-    Call <- match.call(expand.dots = TRUE)
-    Call$sep <- ","; Call[[1L]] <- as.name("write.table")
-    eval.parent(Call)
-\end{lstlisting}
-
-\subsubsection{Sandboxing} The combination of \eval and first class
-environments can be used to construct a sandbox for evaluating
-side-effecting expressions. The newly constructed environment can use
-the current environment for its lexical scope. This provides the
-evaluated expression access to all the bindings in the current scope
-but restricts the side-effect to the newly constructed throwaway
-environment. However, this sandboxing is limited because the
-expression can reflectively access any environment in the call stack
-or use the super assignment operator(\c{<<-}) to write to parent
-scope. The \c{local} function which is part of Core R is an example of
-this usage pattern.
-
-\subsubsection{Dynamic Code Loading}
-The combination of \parse and \eval is used to parse text from string
-or file and evaluate in a custom environment. This pattern is used by
-the \source Core R function to load R code from a file in the current
-workspace. This is quite useful for loading functions for interactive
-data analysis.
-
-\subsubsection{Metaprogramming}
-The combination of \eval and \substitute is used to do metaprogramming
-in R. \substitute can be used to synthesize ASTs by stitching together
-user inputs. \substitute walks over its argument AST and replaces the
-symbols with their bindings in the specified environment and returns
-this new AST which can then be evaluated in custom environments using
-\eval.
-
-\subsubsection{Convenience}
-Eval is used to make some programming tasks easier (and/or shorter).
-For example, in \c{data.table}, it is used to set R options from a
-character vector
-  \begin{lstlisting}
-opts = c("datatable.verbose"="FALSE", ...) for (i in names(opts))
-eval(parse(text=paste0("options(",i,"=",opts[i],")")))
-
-  \end{lstlisting}
-  %
-instead of manually calling \c{options(datatable.verbose=FALSE)}
-
-\subsubsection{Implementation}
-The package installation process adds an extra file to the package
-directory that contains the code to load the package code inside a
-\local block. It creates a namespace object and injects the package
-bindings.
-
-\subsubsection{Miscellaneous}
-\eval is also used in R packages to get around various limitations.
-For example, in \datatable package, \eval is used to modify a binding
-in the \base package of R. Bindings in the \base package are locked
-for modification and require explicit unlocking using function
-\unlockBinding. However, directly using this function triggers a
-restricted-function-use warning during package build, preventing the
-package to be accepted by CRAN. The \datatable gets around this
-limitation by enclosing the call to \unlockBinding inside \eval. From
-the same reason, other packages use \eval to access private variables
-and functions from their dependencies.
-
-\subsection{Necessity of Eval}
-
-\subsection{Power of Eval}
-
-\subsubsection{How much code is executed inside eval}
-- measure number of operations inside eval and outside eval.
-
-- this is useful because if analyzers treat eval as a nop, they may
-miss X\% of the code.
-
-\subsubsection{Does eval create observable side-effects}
-- measure number of reads and writes by eval outside of its ``bubble''.
-
-- Since R is lazy, even reads can cause promise evaluation and hence,
-side-effects.
-
-- Look at writes to non-local scopes. Compare with total number of
-writes for reference?
-
-- Categorize side-effects between env passed to eval and other envs.
-
-\subsubsection{Does eval do reflection}
-\eval can access parent scopes via reflection. Find number of times it
-does that and the depth to which it access parent frames.
-
-\subsubsection{Does eval introduce new bindings}
-Find number of times, eval introduces new bindings. There are many
-ways -- library, load, attach, source, and explicitly introduce
-bindings using super-assign, assign and define.
-
-\subsubsection{Does eval call C code}
-Find number of times, eval makes call to native code using - .Call,
-.External, etc. Find out if there are native functions that are called
-only from within eval.
-
-\subsubsection{Does eval add finalizers}
-Eval can attach code to be executed when the object goes out of scope
-or program exits. Find if eval does that.
-
-\subsubsection{Does eval do non-local returns}
-Eval can do non-local returns effective bypassing evaluation of the
-rest of the function. This can be useful for static analyzers.
-
-Conclude with number of evals that are ``pure'', i.e. evals which
-could be ignored by a static analyzer without any problem.
-
-\subsubsection{Provenance}
-
-There are various ways to obtain expressions:
-\begin{itemize}
-  \item \c{substitute} synthesizes ASTs from expressions by replacing
-    symbols with their bindings in the specified environment.
-  \item \c{expression}  creates a vector of expression
-    objects from text.
-  \item \c{parse}, \c{str2expression} and \c{str2lang} turn strings into
-    expressions.
-\end{itemize}
-
-We observe 5.2\% cases where \eval directly evaluates the output of
-\c{substitute}, 0.7\% cases where output of \c{parse} is read directly
-and only 704 cases for \c{expression}. Most expressions consumed by
-\eval are generated by other functions.
-
-\c{eval(parse(...))} can be used for dynamic code loading. This forms
-the core of \c{source} and \c{sys.source} functions in R that are
-commonly used for loading code in R files in interactive settings. We
-investigated the number of cases in which the output of \c{parse} and
-its variants is passed to \eval, directly or transitively by tainting
-their output. This corresponds to \PercentParsedCallSites of the total
-\eval call sites and \PercentParsedEvals of the eval calls. We
-observed that very few of the eval call (\NbParseFilesRnd in total)
-consume the result of calling \c{parse} on a file. Most of the eval
-calls consume the result of calling \c{parse} on a string. We also
-identified one function in core R, \c{invokeRestartInteractively} that
-prompts the user for input, parses it, and passes it to \eval.
-
-\subsubsection{Side-Effects}
-
-In the corpus we observe \AllWritesRnd writes to variables of which
-\EvalWritesRnd writes happen inside \eval. However, all writes are not
-dangerous. Only writes to environments not local to the computation
-spawned by \eval are side-effecting. These writes outlive the
-computation and hence are visible outside it. The remaining writes are
-local to the computation. We observe that \EvalSideEffectingWritesRnd
-writes inside evals are side-effecting. This is only
-\EvalSideEffectingWritesEvalPerc of all variable writes inside \eval
-and \EvalSideEffectingWritesAllPerc of all variable writes in the
-corpus.
-
-An \eval is considered side-effecting if it performs a side-effecting
-write to a variable, directly or indirectly. Only
-\SideEffectingCoreCallPerc \eval calls in Core R are side-effecting
-and \SideEffectingPackageCallPerc \eval calls in CRAN packages are
-side-effecting.
-
-\subsubsection{Scope}
-
-One of the main differences of the R \eval with the javascript one is
-how it interacts with \emph{scope}, or environments in R. In
-Javascript, \eval can access the local scope and the global scope. In
-R, \eval has an argument \c{envir} to indicate in which environment
-its first argument must be evaluated. By default, it is equal to
-\c{parent.frame()}, which is the parent environment of \eval call.
-This default argument happens in \DefaultEnvirExprPercent of the eval
-calls.
-
-This argument can also be \c{NULL}, a \c{List} or a data frame. This
-happens in 3.5\% of the cases: \eval copies the fields of the list or
-data frame and creates bindings for them in a new environment. This
-pattern is used to evaluate formulas which can directly refer to the
-fields of the data. The \c{envir} argument can also be a number $n$.
-It means that the environment in which the expressions is evaluated
-will be the result of \c{sys.call(n)} where $n$ refers to the $n$-th
-stack frame.
-
-The top-level environment in R is called the global environment. New
-environments can be created using \c{new.env}. They can be provided a
-parent environment which becomes the enclosing scope of the new
-environment.
-
-We looked at the environments passed to all the \eval calls in our
-corpus. Table~\ref{tab:environments} summarizes the results. A numeric
-environment class \c{n} denotes the environment of the $n$-th call
-stack frame from the current function. \c{global} denotes the
-top-level environment and \c{list} denotes a list passed for
-evaluation of formulas. Environment classes of the form $n+$ denote
-the $n$-th environment extended with a new environment. The new
-environment provides a limited form of sandboxing. All assignments
-using the \c{\<-} function occur inside it and prevent the extended
-environment from mutation. However, it is still possible to mutate the
-extended environment using the \c{\<\<-} or \c{assign} functions; but,
-that happens rarely.
-
-\begin{table}[htbp]{ \centering
-\begin{tabular}{c|c|c|c}\hline
-\multicolumn{2}{c|}{Core} & \multicolumn{2}{|c}{Packages}  \\\hline
-Environment Class & Eval Calls \% & Environment Class & Eval Calls\\\hline
-\CoreEnvClassA & \CoreProportionA & \PackagesEnvClassA &  \PackagesProportionA\\
-\CoreEnvClassB & \CoreProportionB & \PackagesEnvClassB &  \PackagesProportionB\\
-\CoreEnvClassC & \CoreProportionC & \PackagesEnvClassC &  \PackagesProportionC\\
-\CoreEnvClassD & \CoreProportionD & \PackagesEnvClassD &  \PackagesProportionD\\
-\CoreEnvClassE & \CoreProportionE & \PackagesEnvClassE &  \PackagesProportionE\\\hline
-\end{tabular}} \label{tab:environments}
-\caption{Environments in terms of \eval calls}
-\end{table}
-
-We observe that a disproportionately high number of core R \eval calls
-access the caller's caller's environment. This is because many core R
-functions call functions that pass the result of \c{parent.frame()} to
-an \eval. A disproportionately high number of calls to \eval happens
-in an extended top-level environment. This can be explained by the
-fact that many packages evaluate code passed from the user's workspace
-in the top-level environment to access the bindings.
-
-The \c{imchange} function of package \c{imager} makes it possible to
-modify images using a dedicated formula syntax using
-\c{\~}.\footnote{Inspired by {map} in package \emph{purr}.} Here,
-\eval is evaluated in \c{newenv}, which creates a new environment that
-inherits from \c{parent.frame()} by default (classified as 1+).
-
-\begin{lstlisting}
-newenv <- new.env()
-...
-fo <- parse(text=as.character(fo)[2])
-im[where] <- eval(fo,envir=newenv,enclos=env)
-\end{lstlisting}
-
-\c{adjCoef} in package \emph{actuar} find the root of an equation
-defined by a function \c{h} whose arguments must be named \c{x} and
-\c{y}. \c{h} is transformed into an auxiliary function \c{h2} that can
-be optimized. Here, the list used for \c{envir} ensures the
-correspondance between the textual arguments of \c{h} and the
-arguments of \c{h2}.
-
-\begin{lstlisting}
-sh <- substitute(h)
-fcall <- paste(sh, "(x, y)")
-...
-h2 <- function(x, y)
-    eval(parse(text = fcall),
-    envir = list(x = x, y = y),
-    enclos = parent.frame(2))
-\end{lstlisting}
-
-\section{Case Studies}
-
-5(?) examples from real code.
-
 \section{Why do we Eval}
@@
 
-\section{Analysis Results}
-
-\subsubsection{A taxonomy of eval}
-
-\section{Discussion}
-
-\subsection{Comparison with Javascript}
+\section{Comparison with Javascript}
 

### 2020-11-30 inst/paper/main.tex
minor

@@
 
-\subsubsection{Does eval introduce new bindings}
-Find number of times, eval introduces new bindings. There are many
-ways -- library, load, attach, source, and explicitly introduce
-bindings using super-assign, assign and define.
-
-\subsubsection{Does eval call C code}
-Find number of times, eval makes call to native code using - .Call,
-.External, etc. Find out if there are native functions that are called
-only from within eval.
-
-\subsubsection{Does eval add finalizers}
-Eval can attach code to be executed when the object goes out of scope
-or program exits. Find if eval does that.
-
-\subsubsection{Does eval do non-local returns}
+In the corpus we observe \AllWritesRnd writes to variables of which
+\EvalWritesRnd writes happen inside \eval. However, all writes are not
+dangerous. Only writes to environments not local to the computation
+spawned by \eval are side-effecting. These writes outlive the
+computation and hence are visible outside it. The remaining writes are
+local to the computation. We observe that \EvalSideEffectingWritesRnd
+writes inside evals are side-effecting. This is only
+\EvalSideEffectingWritesEvalPerc of all variable writes inside \eval
+and \EvalSideEffectingWritesAllPerc of all variable writes in the
+corpus.
+
+An \eval is considered side-effecting if it performs a side-effecting
+write to a variable, directly or indirectly. Only
+\SideEffectingCoreCallPerc \eval calls in Core R are side-effecting
+and \SideEffectingPackageCallPerc \eval calls in CRAN packages are
+side-effecting.
+
+\mypara{New Bindings.} Find number of times, eval introduces new
+bindings. There are many ways -- library, load, attach, source, and
+explicitly introduce bindings using super-assign, assign and define.
+
+\mypara{C Calls.} Find number of times, eval makes call to native code
+using - .Call, .External, etc. Find out if there are native functions
+that are called only from within eval.
+
+\mypara{Non-local returns}
 Eval can do non-local returns effective bypassing evaluation of the
@@
 
+\mypara{Purity.}
 Conclude with number of evals that are ``pure'', i.e. evals which
@@
 
-\subsubsection{Side-Effects}
-
-In the corpus we observe \AllWritesRnd writes to variables of which
-\EvalWritesRnd writes happen inside \eval. However, all writes are not
-dangerous. Only writes to environments not local to the computation
-spawned by \eval are side-effecting. These writes outlive the
-computation and hence are visible outside it. The remaining writes are
-local to the computation. We observe that \EvalSideEffectingWritesRnd
-writes inside evals are side-effecting. This is only
-\EvalSideEffectingWritesEvalPerc of all variable writes inside \eval
-and \EvalSideEffectingWritesAllPerc of all variable writes in the
-corpus.
-
-An \eval is considered side-effecting if it performs a side-effecting
-write to a variable, directly or indirectly. Only
-\SideEffectingCoreCallPerc \eval calls in Core R are side-effecting
-and \SideEffectingPackageCallPerc \eval calls in CRAN packages are
-side-effecting.
-
 \subsubsection{Scope}

### 2020-11-30 inst/paper/main.tex
Add two examples of eval

@@
 
+Recording calls - this one is more on the misuse than doing something useful. I
+really wonder why they do not use \c{do.call(write.table, Call)} (except that it
+will create additional frames).
+\begin{lstlisting}
+write.csv <- function (...)
+{
+  Call <- match.call(expand.dots = TRUE)
+  for (argname in c("append", "col.names", "sep", "dec", "qmethod"))
+    if (!is.null(Call[[argname]]))
+      warning(gettextf("attempt to set '%s' ignored", argname), domain = NA)
+  rn <- eval.parent(Call$row.names)
+  Call$append <- NULL
+  Call$col.names <- if (is.logical(rn) && !rn) TRUE else NA
+  Call$sep <- ","
+  Call$dec <- "."
+  Call$qmethod <- "double"
+  Call[[1L]] <- as.name("write.table")
+  eval.parent(Call)
+}
+\end{lstlisting}
+
+creating and evaling strings - here they assemble the call to plot function. plot is s3 generic which takes
+ridiculous amount of parameters and in addition their plot function also takes parameters.
+This code could be nicely refactored using do.call.
+\begin{lstlisting}
+wsbackfit::plot.sback <- function(x, composed = TRUE, ask = TRUE, select = NULL, ...) {
+ dots <- list(...)
+ grph.opt <- names(dots)
+...
+ stub <- paste(
+  ifelse("xlab" %in% grph.opt, "", paste(", xlab = \"", var, "\"",sep = "")),
+  ifelse("ylab" %in% grph.opt, "", paste(", ylab = \"", p.functions[i], "\"", sep = "")),
+  ifelse("main" %in% grph.opt, "", main.aux),
+  ifelse("type" %in% grph.opt, "", ", type = \"l\""),
+  ifelse("ylim" %in% grph.opt, "", paste(", ylim = c(", min.ylim,",", max.ylim,")", sep = "")), ",...)",
+  sep = ""
+ )
+
+  if(composed) {
+          plot <- paste("plot(x.data, y.composed", stub, sep = "")
+          eval(parse(text = plot))
+          rug(x$data[,var])
+  } else {
+          plot <- paste("plot(x.data, y.data.nl", stub, sep = "")
+          eval(parse(text = plot))
+
+          stub <- paste(ifelse("lty" %in% grph.opt, "", ",lty = 2"), ",...)", sep = "")
+          lines <- paste("lines(x.data, y.data.l", stub, sep = "")
+          eval(parse(text = lines))
+
+          rug(x$data[,var])
+          stub <- paste(ifelse("lty" %in% grph.opt, "", "lty = 1:2"), ", legend = c(\"Non linear\",\"Linear\"), bty = \"n\"", ", cex = ", dots$cex,")", sep = "")
+          legend <- paste("legend(\"topleft\",", stub, sep = "")
+          eval(parse(text = legend))
+  }
+...
+}
+\end{lstlisting}
+
 \subsection{Previous Work}

### 2020-11-30 inst/paper/main.tex
moving around

--- 2020-11-30 inst/paper/main.tex
@@
 
+\mypara{Record Current Call.}
+R provides a \c{match.call} function which can return the current
+call. This is heavily used by the statistical model-fitting functions
+that records the current call that is then reevaluated in the
+different environment. Another pattern is to match the current call,
+adjust some arguments and pass it to another function. For example the
+\c{write.csv} function changes the \c{sep} argument in the matched
+call and reevaluates itself as \c{write.table}:
+%
+\begin{lstlisting}
+    Call <- match.call(expand.dots = TRUE)
+    Call$sep <- ","; Call[[1L]] <- as.name("write.table")
+    eval.parent(Call)
+\end{lstlisting}
+
+\mypara{Sandbox.} The combination of \eval and first class
+environments can be used to construct a sandbox for evaluating
+side-effecting expressions. The newly constructed environment can use
+the current environment for its lexical scope. This provides the
+evaluated expression access to all the bindings in the current scope
+but restricts the side-effect to the newly constructed throwaway
+environment. However, this sandboxing is limited because the
+expression can reflectively access any environment in the call stack
+or use the super assignment operator(\c{<<-}) to write to parent
+scope. The \c{local} function which is part of Core R is an example of
+this usage pattern.
+
+\mypara{Code Loading.}
+The combination of \parse and \eval is used to parse text from string
+or file and evaluate in a custom environment. This pattern is used by
+the \source Core R function to load R code from a file in the current
+workspace. This is quite useful for loading functions for interactive
+data analysis.
+
+\mypara{Metaprogramming.}
+The combination of \eval and \substitute is used to do metaprogramming
+in R. \substitute can be used to synthesize ASTs by stitching together
+user inputs. \substitute walks over its argument AST and replaces the
+symbols with their bindings in the specified environment and returns
+this new AST which can then be evaluated in custom environments using
+\eval.
+
+\mypara{Convenience.}
+Eval is used to make some programming tasks easier (and/or shorter).
+For example, in \c{data.table}, it is used to set R options from a
+character vector
+  \begin{lstlisting}
+opts = c("datatable.verbose"="FALSE", ...) for (i in names(opts))
+eval(parse(text=paste0("options(",i,"=",opts[i],")")))
+
+  \end{lstlisting}
+  %
+instead of manually calling \c{options(datatable.verbose=FALSE)}
+
 \subsection{Previous Work}
@@
 
-Richards et al.~\cite{ecoop11} identified a number of frequently
-occurring \eval patterns which were detected using simple regular
-expressions on the \eval strings. In the case of R we match the
-patterns on the recorded AST.
-
 Table~\ref{table:js-pattens-in-r} shows the ratio of \eval in each of
@@
 observed the following categories:
-
-\subsubsection{Recording the current call}
-R provides a \c{match.call} function which can return the current
-call. This is heavily used by the statistical model-fitting functions
-that records the current call that is then reevaluated in the
-different environment. Another pattern is to match the current call,
-adjust some arguments and pass it to another function. For example the
-\c{write.csv} function changes the \c{sep} argument in the matched
-call and reevaluates itself as \c{write.table}:
-%
-\begin{lstlisting}
-    Call <- match.call(expand.dots = TRUE)
-    Call$sep <- ","; Call[[1L]] <- as.name("write.table")
-    eval.parent(Call)
-\end{lstlisting}
-
-\subsubsection{Sandboxing} The combination of \eval and first class
-environments can be used to construct a sandbox for evaluating
-side-effecting expressions. The newly constructed environment can use
-the current environment for its lexical scope. This provides the
-evaluated expression access to all the bindings in the current scope
-but restricts the side-effect to the newly constructed throwaway
-environment. However, this sandboxing is limited because the
-expression can reflectively access any environment in the call stack
-or use the super assignment operator(\c{<<-}) to write to parent
-scope. The \c{local} function which is part of Core R is an example of
-this usage pattern.
-
-\subsubsection{Dynamic Code Loading}
-The combination of \parse and \eval is used to parse text from string
-or file and evaluate in a custom environment. This pattern is used by
-the \source Core R function to load R code from a file in the current
-workspace. This is quite useful for loading functions for interactive
-data analysis.
-
-\subsubsection{Metaprogramming}
-The combination of \eval and \substitute is used to do metaprogramming
-in R. \substitute can be used to synthesize ASTs by stitching together
-user inputs. \substitute walks over its argument AST and replaces the
-symbols with their bindings in the specified environment and returns
-this new AST which can then be evaluated in custom environments using
-\eval.
-
-\subsubsection{Convenience}
-Eval is used to make some programming tasks easier (and/or shorter).
-For example, in \c{data.table}, it is used to set R options from a
-character vector
-  \begin{lstlisting}
-opts = c("datatable.verbose"="FALSE", ...) for (i in names(opts))
-eval(parse(text=paste0("options(",i,"=",opts[i],")")))
-
-  \end{lstlisting}
-  %
-instead of manually calling \c{options(datatable.verbose=FALSE)}
 

### 2020-11-30 inst/paper/main.tex

--- 2020-11-30 inst/paper/main.tex
@@
 
+\mypara{Record Current Call.}
+R provides a \c{match.call} function which can return the current
+call. This is heavily used by the statistical model-fitting functions
+that records the current call that is then reevaluated in the
+different environment. Another pattern is to match the current call,
+adjust some arguments and pass it to another function. For example the
+\c{write.csv} function changes the \c{sep} argument in the matched
+call and reevaluates itself as \c{write.table}:
+%
+\begin{lstlisting}
+    Call <- match.call(expand.dots = TRUE)
+    Call$sep <- ","; Call[[1L]] <- as.name("write.table")
+    eval.parent(Call)
+\end{lstlisting}
+
+\mypara{Sandbox.} The combination of \eval and first class
+environments can be used to construct a sandbox for evaluating
+side-effecting expressions. The newly constructed environment can use
+the current environment for its lexical scope. This provides the
+evaluated expression access to all the bindings in the current scope
+but restricts the side-effect to the newly constructed throwaway
+environment. However, this sandboxing is limited because the
+expression can reflectively access any environment in the call stack
+or use the super assignment operator(\c{<<-}) to write to parent
+scope. The \c{local} function which is part of Core R is an example of
+this usage pattern.
+
+\mypara{Code Loading.}
+The combination of \parse and \eval is used to parse text from string
+or file and evaluate in a custom environment. This pattern is used by
+the \source Core R function to load R code from a file in the current
+workspace. This is quite useful for loading functions for interactive
+data analysis.
+
+\mypara{Metaprogramming.}
+The combination of \eval and \substitute is used to do metaprogramming
+in R. \substitute can be used to synthesize ASTs by stitching together
+user inputs. \substitute walks over its argument AST and replaces the
+symbols with their bindings in the specified environment and returns
+this new AST which can then be evaluated in custom environments using
+\eval.
+
+\mypara{Convenience.}
+Eval is used to make some programming tasks easier (and/or shorter).
+For example, in \c{data.table}, it is used to set R options from a
+character vector
+  \begin{lstlisting}
+opts = c("datatable.verbose"="FALSE", ...) for (i in names(opts))
+eval(parse(text=paste0("options(",i,"=",opts[i],")")))
+
+  \end{lstlisting}
+  %
+instead of manually calling \c{options(datatable.verbose=FALSE)}
 Recording calls - this one is more on the misuse than doing something useful. I
@@
 
-creating and evaling strings - here they assemble the call to plot function. plot is s3 generic which takes
-ridiculous amount of parameters and in addition their plot function also takes parameters.
+creating and evaling strings - here they assemble the call to plot
+function. plot is s3 generic which takes ridiculous amount of
+parameters and in addition their plot function also takes parameters.
 This code could be nicely refactored using do.call.

### 2020-11-30 inst/paper/main.tex
Add one more

@@
 
-creating and evaling strings - here they assemble the call to plot
-function. plot is s3 generic which takes ridiculous amount of
-parameters and in addition their plot function also takes parameters.
-This code could be nicely refactored using do.call.
+In vcrpart package they use this a lot (following is just one instance):
+
+\begin{lstlisting}
+vcrpart::tvcglm <- function(formula, data, family,
+                   weights, subset, offset, na.action = na.omit,
+                   control = tvcglm_control(), ...) {
+    mc <- match.call()
+    mc[[1L]] <- as.name("tvcm")
+    if (!"control" %in% names(mc))
+        mc$control <- formals(tvcglm)$control
+    if ("fit" %in% names(list(...)))
+        warning("inputs for 'fit' in 'tvcglm' are ignored. ",
+                "'fit' is set to 'glm'.")
+    mc$fit <- "glm"
+    return(eval.parent(mc))
+  }
+\end{lstlisting}
+
+The examples above are related to the use of \c{match.call} whose documentation talks about these use cases:
+
+\begin{verbatim}
+‘match.call’ is most commonly used in two circumstances:
+
+        • To record the call for later re-use: for example most
+          model-fitting functions record the call as element ‘call’ of
+          the list they return.  Here the default ‘expand.dots = TRUE’
+          is appropriate.
+
+        • To pass most of the call to another function, often
+          ‘model.frame’.  Here the common idiom is that ‘expand.dots =
+          FALSE’ is used, and the ‘...’ element of the matched call is
+          removed.  An alternative is to explicitly select the
+          arguments to be passed on, as is done in ‘lm’.
+\end{verbatim}
+
+creating and evaling strings - here they assemble the call to plot function.
+plot is s3 generic which takes ridiculous amount of parameters and in addition
+their plot function also takes parameters. This code could be nicely refactored
+using do.call.
 \begin{lstlisting}

### 2020-12-01 inst/paper/main.tex
use cases

--- 2020-11-30 inst/paper/main.tex
@@
 
-\mypara{Record Current Call.}
-R provides a \c{match.call} function which can return the current
-call. This is heavily used by the statistical model-fitting functions
-that records the current call that is then reevaluated in the
-different environment. Another pattern is to match the current call,
-adjust some arguments and pass it to another function. For example the
-\c{write.csv} function changes the \c{sep} argument in the matched
-call and reevaluates itself as \c{write.table}:
-%
-\begin{lstlisting}
-    Call <- match.call(expand.dots = TRUE)
-    Call$sep <- ","; Call[[1L]] <- as.name("write.table")
-    eval.parent(Call)
+\mypara{Intercession.} A common use case for \eval is to be combined
+with \c{match.call}. What \c{match.call} does is step up the call
+stack and capture the invocation of the currently executing function
+and return it as an unevaluated expression. The general pattern here
+is that a call to some function \c{f} is transformed to a call to
+another function \c{g} with some arguments retained and others
+modified. An illustration taken from the \c{vcpart} package shows a
+function \c{tvcglm} that is translated to the \c{tvcm} function but
+with two modifications to the argument list, argument \c{control}
+can't be missing and argument \c{fit} is forced to have value \c{glm}.
+The function ends with a call to \c{eval.parent} to ensure that the
+rewritten call is evaluted in the same environment the original call
+was.
+\begin{lstlisting}
+ tvcglm <- function(formula, data, family, control=tvcglm_control(), ...) {
+   Call <- match.call()
+   Call[[1L]] <- as.name("tvcm")
+   if (!"control" %in% names(Call)) Call$control<-formals(tvcglm)$control
+   if ("fit" %in% names(list(...))) warning("'fit' is ignored. ")
+   Call$fit <- "glm"
+   return(eval.parent(Call))
+ }
+\end{lstlisting}
+
+\mypara{Code Generation.} The more traditional use of \eval is to
+execute code that was assembled by the programmer into a string. Here
+we show the method \c{plot} for class \c{sback} in package
+\c{wsbackfit}, simplified for explanatory purposes. The function takes
+a long argument list, the names of which are captured in the list
+\c{opt}. The string \c{stub} is composed of a subset of the arguments
+passed to this function; the variable is used to construct a call to
+\c{base::plot} which will draw a plot. \NOTE{This example is a bit
+  messed up. What is var? Does the use of ... in the eval grab the
+  arguments of this call? If yes does it not end up with xlab twice?
+  HELP}
+\begin{lstlisting}
+ plot.sback <- function(x,...) {
+   opt <- names(list(...))
+   stub <- paste(
+    ifelse("xlab" %in% opt,"",paste(",xlab=\"",var,"\"",sep="")),
+    ifelse("main" %in% opt,"",main.aux),
+    ifelse("type" %in% opt,"",",type=\"l\""),
+    sep = "")
+   plot <- paste("plot(x.data,",stub,",...)", sep = "")
+   eval(parse(text=plot))
+   ...
 \end{lstlisting}
@@
 
-\mypara{Side-effects.}
-XXX026 total number of reads performed by \eval. \NOTE{What is the definition of a read?  An env lookup?}  XXX027 average number of reads.
-XXX028 total number of writes performed by \eval. Average number XXX029.
+\mypara{Side-effects.} XXX026 total number of reads performed by
+\eval. \NOTE{What is the definition of a read? An env lookup?} XXX027
+average number of reads. XXX028 total number of writes performed by
+\eval. Average number XXX029.
 
@@
 
-XXX029 reads to caller environment -- where caller is the caller of \eval.
-
-XXX030 reads to other caller environment -- i.e. envs that are on the call stack.
-
-XXX031 reads to caller environment -- where caller is the caller of \eval.
-
-XXX032 reads to other caller environment -- i.e. envs that are on the call stack.
+XXX029 reads to caller environment -- where caller is the caller of
+\eval.
+
+XXX030 reads to other caller environment -- i.e. envs that are on the
+call stack.
+
+XXX031 reads to caller environment -- where caller is the caller of
+\eval.
+
+XXX032 reads to other caller environment -- i.e. envs that are on the
+call stack.
 
@@
 
-\subsection{Necessity of Eval}
-
 \subsubsection{Provenance}

### 2020-12-01 inst/paper/main.tex
2

@@
 \end{lstlisting}
+The \parse and \eval combination is used by the the \source function
+in the base library to load R code from a file in the current
+workspace.
 
@@
 this usage pattern.
-
-\mypara{Code Loading.}
-The combination of \parse and \eval is used to parse text from string
-or file and evaluate in a custom environment. This pattern is used by
-the \source Core R function to load R code from a file in the current
-workspace. This is quite useful for loading functions for interactive
-data analysis.
-
-\mypara{Metaprogramming.}
-The combination of \eval and \substitute is used to do metaprogramming
-in R. \substitute can be used to synthesize ASTs by stitching together
-user inputs. \substitute walks over its argument AST and replaces the
-symbols with their bindings in the specified environment and returns
-this new AST which can then be evaluated in custom environments using
-\eval.
 

### 2020-12-01 inst/paper/main.tex
minor

@@
 \end{lstlisting}
+This pattern is recognizable by the fact that the expression is a call
+and the target environment is that of the parent.
 
@@
   arguments of this call? If yes does it not end up with xlab twice?
-  HELP}
+  HELP} The \parse and \eval combination is used by the the \source
+function in the base library to load R code from a file in the current
+workspace.
 \begin{lstlisting}
@@
     sep = "")
-   plot <- paste("plot(x.data,",stub,",...)", sep = "")
+   plot <- paste("plot(x.data,",stub,",...)",sep="")
    eval(parse(text=plot))
@@
 \end{lstlisting}
-The \parse and \eval combination is used by the the \source function
-in the base library to load R code from a file in the current
-workspace.
+This pattern is recognized by the use of \c{parse} to turn
+a string into an expression.
 
@@
 
-\mypara{Convenience.}
-Eval is used to make some programming tasks easier (and/or shorter).
-For example, in \c{data.table}, it is used to set R options from a
-character vector
-  \begin{lstlisting}
-opts = c("datatable.verbose"="FALSE", ...) for (i in names(opts))
+\mypara{Debloating.} \Eval is often used as a means to reduce
+boilerplate code; simple and repetitive code can easily be replaced
+with judcious use of \eval. For example, the \c{data.table} package
+uses \eval to calls the \c{options} function with named arguments
+taken from a vector of strings. While the benefits are limited in this
+example, it is an attractive tool for programmers.
+\begin{lstlisting}
+  opts = c("datatable.verbose"="FALSE", # ... many others
+  for (i in names(opts))
     eval(parse(text=paste0("options(",i,"=",opts[i],")")))
-
-  \end{lstlisting}
-  %
- instead of manually calling \c{options(datatable.verbose=FALSE)}
+\end{lstlisting}
+This pattern is a special case of code generation, recognizable by the
+fact that \eval is executed in a loop.
+
+\mypara{Trivial.} When values are passed to \eval, they are returned
+unchanged. They are an example of trivial uses of \eval. Another
+trivial use is the empty expression, often found in JavaScript, but
+rare in R.
 

### 2020-12-01 inst/paper/main.tex
2

@@
 \mypara{Intercession.} A common use case for \eval is to be combined
-with \c{match.call}. What \c{match.call} does is step up the call
-stack and capture the invocation of the currently executing function
-and return it as an unevaluated expression. The general pattern here
-is that a call to some function \c{f} is transformed to a call to
-another function \c{g} with some arguments retained and others
-modified. An illustration taken from the \c{vcpart} package shows a
-function \c{tvcglm} that is translated to the \c{tvcm} function but
-with two modifications to the argument list, argument \c{control}
-can't be missing and argument \c{fit} is forced to have value \c{glm}.
-The function ends with a call to \c{eval.parent} to ensure that the
+with \c{match.call}. \c{match.call} walks up the call stack, captures
+the code that invoked the currently executing function, and returns it
+as an unevaluated expression. The pattern is to transform a call to
+some function \c{f} into a call to \c{g} with some arguments retained
+and others modified. As an illustration, consider the \c{vcpart}
+package's function \c{tvcglm} that is translated to a call to \c{tvcm}
+with two modifications to the argument list: argument \c{control}
+can't be missing and \c{fit} is set  \c{"glm"}. The
+function ends with a call to \c{eval.parent} to ensure that the
 rewritten call is evaluted in the same environment the original call
@@
 \end{lstlisting}
-This pattern is recognized by the use of \c{parse} to turn
-a string into an expression.
-
-\mypara{Sandbox.} The combination of \eval and first class
-environments can be used to construct a sandbox for evaluating
-side-effecting expressions. The newly constructed environment can use
-the current environment for its lexical scope. This provides the
-evaluated expression access to all the bindings in the current scope
-but restricts the side-effect to the newly constructed throwaway
-environment. However, this sandboxing is limited because the
-expression can reflectively access any environment in the call stack
-or use the super assignment operator(\c{<<-}) to write to parent
-scope. The \c{local} function which is part of Core R is an example of
-this usage pattern.
+This pattern is recognizable by its use of \c{parse} to turn a string
+into an expression.
 

### 2020-12-01 inst/paper/main.tex
stuff

@@
 
-\mypara{The case of CORE libraries} ... \NOTE{Give numbers for the
-  same things as above but for the base/core libraries} Compare.
-\NOTE{Add XXX016 to XXX025.}
-
-\section{A Taxonomy of Eval}
-
-The previous section gave a quantitative view of the usage of \eval;
-in this section we try to elucidate \emph{what} it does.
-
-\mypara{Side-effects.} XXX026 total number of reads performed by
-\eval. \NOTE{What is the definition of a read? An env lookup?} XXX027
-average number of reads. XXX028 total number of writes performed by
-\eval. Average number XXX029.
-
-\NOTE{Aviral wrote''measure number of reads and writes by eval outside
-  of its bubble'' --- what is the bubble here?}
-
-XXX029 reads to caller environment -- where caller is the caller of
-\eval.
-
-XXX030 reads to other caller environment -- i.e. envs that are on the
-call stack.
-
-XXX031 reads to caller environment -- where caller is the caller of
-\eval.
-
-XXX032 reads to other caller environment -- i.e. envs that are on the
-call stack.
-
-Categorize side-effects between env passed to eval and other envs:
-\begin{itemize}
-\item ephemeral -- read to envs that are created during \eval  XXX033
-\item local -- reads to env of the function in which the call to \eval
-  lexically occurred; XXX034
-\item parent -- reads to parent envs of the local one. XXX035
-\item call-stack -- reads to envs that are on the call stack but not
-  caller. XXX035.
-\end{itemize}
-
-In the corpus we observe \AllWritesRnd writes to variables of which
-\EvalWritesRnd writes happen inside \eval. However, all writes are not
-dangerous. Only writes to environments not local to the computation
-spawned by \eval are side-effecting. These writes outlive the
-computation and hence are visible outside it. The remaining writes are
-local to the computation. We observe that \EvalSideEffectingWritesRnd
-writes inside evals are side-effecting. This is only
-\EvalSideEffectingWritesEvalPerc of all variable writes inside \eval
-and \EvalSideEffectingWritesAllPerc of all variable writes in the
-corpus.
-
-An \eval is considered side-effecting if it performs a side-effecting
-write to a variable, directly or indirectly. Only
-\SideEffectingCoreCallPerc \eval calls in Core R are side-effecting
-and \SideEffectingPackageCallPerc \eval calls in CRAN packages are
-side-effecting.
-
-\mypara{New Bindings.} Find number of times, eval introduces new
-bindings. There are many ways -- library, load, attach, source, and
-explicitly introduce bindings using super-assign, assign and define.
-
-\mypara{C Calls.} Find number of times, eval makes call to native code
-using - .Call, .External, etc. Find out if there are native functions
-that are called only from within eval.
-
-\mypara{Non-local returns}
-Eval can do non-local returns effective bypassing evaluation of the
-rest of the function. This can be useful for static analyzers.
-
-\mypara{Purity.}
-Conclude with number of evals that are ``pure'', i.e. evals which
-could be ignored by a static analyzer without any problem.
-
-\subsection{LEGACY TEXT}
-
-In this section we turn our attention to the expression passed for
-evaluation to \eval. The \eval function accepts all types of R objects
-as inputs, but in the majority of cases
-(\AllExpressionInputEvalCallPerc), it is called with expressions. The
-remaining \AllValueInputEvalCallPerc are values, \ie, they evaluate to
-themselves.
-%
-Passing a value into an eval is effectively an no-op so we look
-closely at the expression types the call sites pass to eval across all
-the runs. \AllValueInputEvalSitePerc of all call sites always pass
-values, \AllExpressionInputEvalSitePerc pass expressions and a very
-large proportion, \AllPolymorphicInputEvalSitePerc, pass both
-expressions and values as inputs across all program runs. We inspected
-the eval callsites that accept only values or both values and
-expression as inputs. The top three of these call sites contribute to
-over ~80\% of all calls to evals. The first one is inside
-\c{match.arg} from core. Given a vector of names as strings, this
-function finds the default values for parameters with those names in
-the caller function. Since the default value for a function parameter
-can be a complex expression or a benign value such as \c{NULL}, the
-\c{match.arg} function evaluates it through a call to \eval. This
-makes the \eval inside \c{match.arg} ``polymorhpic''. The second such
-\eval call originates from \ggplot, a popular R package for making
-plots. \ggplot has it's own object system, \ggproto, which uses \eval
-inside the \ggproto function to determine the super-class from which
-it inherits (the default value is \c{NULL} but in general it could be
-an arbitrarily expression). The third \eval call is in core R \c{str}
-function used to print a summary of R objects. It internally maps
-\eval function on the bindings of an environment which results in
-inputs of multiples types being passed to \eval.
-
-We further investigate the type of input passed to \eval in
-Figure~\ref{fig:eval-expression-kind} by the type of input and the
-source of \eval. In this figure, \emph{Expression} is the type of
-program text, \emph{Environmment} is the type of scopes or
-environments used for binding, \emph{Null} is the type of the unique
-\c{NULL} object, \emph{Vector} is the type assigned to vectors of
-strings, booleans, integers, reals and bytes; \emph{Symbol} is the
-type of unevaluated symbols, and \emph{Function} is the type of
-functions.
-
-\begin{figure}[!h]
-  \centering
-  \includegraphics[width=\columnwidth]{eval-expression-kind}
-  \caption{Type of Expression passed to \eval} \label{fig:eval-expression-kind}
-\end{figure}
-
-From the figure, we observe that majority of inputs to Core \eval
-functions are expressions in contrast to package \eval calls which
-receive mostly environments as inputs. In fact,
-\PackageEnvironmentInputEvalCallPerc of \eval calls in packages
-receive environments as input expressions. There are only four
-callsites in package evals that receive environments as expressions.
+\mypara{Eval Arguments.} Figure~\ref{fig:eval-expression-kind} breaks
+down the expressions passed to \eval by R data type. It separates the
+core packages from CRAN as they have different characteristics. In
+core, XXX036\% of arguments in core are expressions. Whereas, in CRAN
+package XXX037\% are expressions. When the argument is not an
+expression, it is a value. Values are trivial for \eval as they are
+returned unchanged. \PackageEnvironmentInputEvalCallPerc of CRAN
+arguments are environments. This is an anomaly as only four \eval
+call sites dominate this result:
 \begin{compactitem}[$-$]
@@
 \end{compactitem}
+
+\begin{figure}[!h]
+  \centering
+  \includegraphics[width=\columnwidth]{eval-expression-kind}
+  \caption{Type of Expression passed to \eval} \label{fig:eval-expression-kind}
+\end{figure}
+
+\mypara{The case of CORE libraries} ... \NOTE{Give numbers for the
+  same things as above but for the base/core libraries} Compare.
+\NOTE{Add XXX016 to XXX025.}
+
+\section{A Taxonomy of Eval}
+
+The previous section gave a quantitative view of the usage of \eval;
+in this section we try to elucidate \emph{what} it does.
+
+\mypara{Side-effects.} XXX026 total number of reads performed by
+\eval. \NOTE{What is the definition of a read? An env lookup?} XXX027
+average number of reads. XXX028 total number of writes performed by
+\eval. Average number XXX029.
+
+\NOTE{Aviral wrote''measure number of reads and writes by eval outside
+  of its bubble'' --- what is the bubble here?}
+
+XXX029 reads to caller environment -- where caller is the caller of
+\eval.
+
+XXX030 reads to other caller environment -- i.e. envs that are on the
+call stack.
+
+XXX031 reads to caller environment -- where caller is the caller of
+\eval.
+
+XXX032 reads to other caller environment -- i.e. envs that are on the
+call stack.
+
+Categorize side-effects between env passed to eval and other envs:
+\begin{itemize}
+\item ephemeral -- read to envs that are created during \eval  XXX033
+\item local -- reads to env of the function in which the call to \eval
+  lexically occurred; XXX034
+\item parent -- reads to parent envs of the local one. XXX035
+\item call-stack -- reads to envs that are on the call stack but not
+  caller. XXX035.
+\end{itemize}
+
+In the corpus we observe \AllWritesRnd writes to variables of which
+\EvalWritesRnd writes happen inside \eval. However, all writes are not
+dangerous. Only writes to environments not local to the computation
+spawned by \eval are side-effecting. These writes outlive the
+computation and hence are visible outside it. The remaining writes are
+local to the computation. We observe that \EvalSideEffectingWritesRnd
+writes inside evals are side-effecting. This is only
+\EvalSideEffectingWritesEvalPerc of all variable writes inside \eval
+and \EvalSideEffectingWritesAllPerc of all variable writes in the
+corpus.
+
+An \eval is considered side-effecting if it performs a side-effecting
+write to a variable, directly or indirectly. Only
+\SideEffectingCoreCallPerc \eval calls in Core R are side-effecting
+and \SideEffectingPackageCallPerc \eval calls in CRAN packages are
+side-effecting.
+
+\mypara{New Bindings.} Find number of times, eval introduces new
+bindings. There are many ways -- library, load, attach, source, and
+explicitly introduce bindings using super-assign, assign and define.
+
+\mypara{C Calls.} Find number of times, eval makes call to native code
+using - .Call, .External, etc. Find out if there are native functions
+that are called only from within eval.
+
+\mypara{Non-local returns}
+Eval can do non-local returns effective bypassing evaluation of the
+rest of the function. This can be useful for static analyzers.
+
+\mypara{Purity.}
+Conclude with number of evals that are ``pure'', i.e. evals which
+could be ignored by a static analyzer without any problem.
+
+\subsection{LEGACY TEXT}
 

### 2020-12-02 inst/paper/main.tex
delete

@@
 
-\subsection{LEGACY TEXT}
-
-\subsubsection{Usage Patterns}
-
-Table~\ref{table:js-pattens-in-r} shows the ratio of \eval in each of
-these patterns. We included all the original patterns except for
-\emph{JSON} as R uses its own serialization format and for
-\emph{empty} as there were no empty expression passed to \eval:
-
-\begin{table}[ht]%\resizebox{\columnwidth}{!}{%
-  \centering
-\begin{tabular}{r|r|r|r|r}\hline
-\multirow{2}{*}{Pattern} & \multicolumn{2}{c|}{Core} & \multicolumn{2}{c}{Packages} \\
-                       & Calls & Ratio            & Calls & Ratio \\\hline
-\input{tag/table-js-patterns-in-r.tex}
-\end{tabular}
-\caption{JavaScript \eval Patterns in R}
-\label{table:js-pattens-in-r}
-\end{table}
-
-\noindent \emph{Library}, \emph{Typeof} and \emph{Try} are all barely
-used mostly because the language provides dedicated features for it.
-The few calls in Library pattern come mostly from file loading
-functions such as \c{source} and \c{sys.source}.
-%
-\noindent
-\emph{Read} pattern in JavaScript represented mostly an incorrect
-access to an object's property or local variable both replaceable by
-JavaScript core functions. In the case of R there are \PatternReadRnd
-read \eval calls (\PatternReadRatio). They come from calls like
-\c{subset(airquality, select = Temp)} where the variable \c{Temp} will
-be evaluated in the \c{airquality} data frame.
-
-%
-\noindent \emph{Assign} category comprises of assignments to variables
-within an \eval call. We have seen \PatternAssignRnd such calls
-(\PatternAssignRatio). \PatternAssignArrowRatio uses the simple
-assignment (\c{<-}). The majority of cases comes from the
-\c{magrittr}'s \c{\%<>\%} pipe operator
-(\PatternAssignArrowMagrittrRatio) which forwards an object to a
-function on rhs and update lhs with the result. Far more calls used
-the super assignment (\c{<<-}) operator, but almost all come from the
-\c{plyr} package which provides a convenient API for common data
-manipulation problems.
-%
-\noindent \emph{Call} and \emph{Other} together represents the vast
-majority of eval calls in R. The \emph{Other} pattern represents all
-the calls where eval was called with a value instead of with an
-expression. The \emph{Call} pattern contains all the calls that did
-not fall in the previous categories. Without a context it is usually
-difficult to further categorize the intent of these calls. Therefore
-we have randomly chosen \PatternManualPackages packages and manually
-looked at \PatternManualCallsites call sites from which we have
-observed the following categories:
-
-\subsubsection{Implementation}
-The package installation process adds an extra file to the package
-directory that contains the code to load the package code inside a
-\local block. It creates a namespace object and injects the package
-bindings.
-
-\subsubsection{Miscellaneous}
-\eval is also used in R packages to get around various limitations.
-For example, in \datatable package, \eval is used to modify a binding
-in the \base package of R. Bindings in the \base package are locked
-for modification and require explicit unlocking using function
-\unlockBinding. However, directly using this function triggers a
-restricted-function-use warning during package build, preventing the
-package to be accepted by CRAN. The \datatable gets around this
-limitation by enclosing the call to \unlockBinding inside \eval. From
-the same reason, other packages use \eval to access private variables
-and functions from their dependencies.
-
-\subsubsection{Provenance}
-
+\myoara{Source.}
 There are various ways to obtain expressions:
@@
 
-\subsubsection{Scope}
-
-One of the main differences of the R \eval with the javascript one is
-how it interacts with \emph{scope}, or environments in R. In
-Javascript, \eval can access the local scope and the global scope. In
-R, \eval has an argument \c{envir} to indicate in which environment
-its first argument must be evaluated. By default, it is equal to
-\c{parent.frame()}, which is the parent environment of \eval call.
-This default argument happens in \DefaultEnvirExprPercent of the eval
-calls.
-
-This argument can also be \c{NULL}, a \c{List} or a data frame. This
-happens in 3.5\% of the cases: \eval copies the fields of the list or
-data frame and creates bindings for them in a new environment. This
-pattern is used to evaluate formulas which can directly refer to the
-fields of the data. The \c{envir} argument can also be a number $n$.
-It means that the environment in which the expressions is evaluated
-will be the result of \c{sys.call(n)} where $n$ refers to the $n$-th
-stack frame.
+\mypara{Scope.} The environment argument can be \c{NULL}, a \c{List}
+or a data frame. This happens in 3.5\% of the cases: \eval copies the
+fields of the list or data frame and creates bindings for them in a
+new environment. This pattern is used to evaluate formulas which can
+directly refer to the fields of the data. The \c{envir} argument can
+also be a number $n$. It means that the environment in which the
+expressions is evaluated will be the result of \c{sys.call(n)} where
+$n$ refers to the $n$-th stack frame.
 
@@
 
-\section{Comparison with Javascript}
-
-Comparing the use of \eval in R and JavaScript is instructive. The use
-of \eval is more widespread in data science code than it was in
-JavaScript. Any run of an R program seems to trigger \eval; moreover
-package programmers use \eval liberally. On the other hand, the
-authors of Kaggle packages do not use the feature at all. This point
-to a bifurcated user community: experts who write package know how to
-use \eval while end-users do not. An alternative explanation is that
-end-users do not need \eval for simple data analysis tasks. This
-bifurcation was not observed in the JavaScript community, but this may
-have changed with the advent of Node.js ecosystem. The distribution of
-input sizes is similar to that found in JavaScript, the majority of
-inputs being rather small.
+\subsection{Comparison with Javascript}
 

### 2020-12-03 inst/analysis/preprocess.R
Fix regexes that determine eval source

@@ -223,7 +223,7 @@ extract_package_name <- function(src_ref, file) {
   case_when(
     is.na(src_ref) ~ "base?",
     str_starts(src_ref, fixed("./R/")) ~ "core",
-    str_starts(src_ref, fixed("/tmp/")) ~ str_match(src_ref, "/tmp/Rtmp[^/]*/R\\.INSTALL[^/]*/([^/]+)/.*")[[2]],
+    str_starts(src_ref, fixed("/tmp/")) ~ str_match(src_ref, "/tmp.*/Rtmp[^/]*/R\\.INSTALL[^/]*/([^/]+)/R/.*$")[[2]], #path problem here
     str_starts(src_ref, fixed("/mnt/nvme0/")) ~ "base",
     str_starts(src_ref, fixed("test")) ~ str_match(src_ref, "([^/]*)/.*")[[2]],
     str_starts(src_ref, fixed("/:")) ~ str_match(file, "[^/]*/[^/]*/([^/]*)/.*")[[2]],
@@ -243,6 +243,9 @@ find_package_name <-
     } # "Actually base but we generalize
     else {
       tempPack <- extract_package_name(srcref, file)
+      #It would indicate that one of the regex has failed and
+      # so that probably assumptions made on the paths in the srcref are no longer valid
+      stopifnot(!is.na(tempPack))
       if (tempPack == "base?")
       # It means the srcref was NA
         {
@@ -423,8 +426,9 @@ Example:
 main <- function(argv) {
   if (length(argv) != 10) {
     cat("Only ", length(argv), " arguments. Missing arguments!\n")
+    print(argv)
     usage()
-    quit(save = "no", status = 1)
+    return(1)
   }
 
   corpus_file <- argv[1]
@@ -439,8 +443,10 @@ main <- function(argv) {
   evals_summarized_externals_file <- argv[10]
 
 
+
   cat("Reading and validating input files\n")
 
+  now_first <- Sys.time()
   corpus <- read_fst(corpus_file)
   stopifnot(length(corpus) == 29)
   stopifnot("package" %in% names(corpus)) # for preprocess, we mainly care about the package names
@@ -449,10 +455,11 @@ main <- function(argv) {
     read_fst(calls_file) %>%
     as_tibble() %>%
     mutate(
-      package = basename(dirname(dirname(file))),
+      package = basename(dirname(dirname(dirname(file)))),
       corpus = "cran"
-    ) %>%
[…]

### 2020-12-03 inst/analysis/usage_metrics.Rmd
Usage metrics file

---
title: "Usage metrics"
author: "Pierre Donat-Bouillud"
output: html_document
date: "`r Sys.Date()`"
editor_options: 
  chunk_output_type: console
params:
  base_dir: /var/lib/R/project-evalR
  calls_packages_path: !r paste0(params$base_dir, "/run/package-evals-traced.5/summarized-packages.fst")
  corpus_path: !r paste0(params$base_dir, "/revalstudy/inst/data/corpus.fst")
---

```{r setup, include=FALSE}
knitr::knit_hooks$set(time_it = local({
  now <- NULL
  function(before, options) {
    if (before) {
      # record the current time before each chunk
      now <<- Sys.time()
    } else {
      # calculate the time difference after a chunk
      res <- difftime(Sys.time(), now)
      # return a character string to show the time
      paste("Time for this code chunk to run:", res, units(res))
    }
  }
}))

knitr::opts_chunk$set(echo = TRUE,
                      fig.retina = 2,
                      fig.width = 10,
                      cache.lazy = FALSE,
                      time_it = TRUE)

now <- Sys.time() 
library(tidyverse)
library(stringr)
library(fst)
library(fs)

source("inc/paths.R")
source("inc/latextags.R")
source("inc/setup.R")
theme_set(theme_minimal())
create_tags(path(TAGS_DIR, "enrichDataset.tex"), prefix="", default=TRUE)
```

```{r file-info, echo=FALSE}
file_info(params$calls_packages_path)
```

```{r read-data}
eval_calls <- read_fst(params$calls_packages_path) %>% as_tibble()
```

```{r read_corpus}
corpus <- read_fst(params$corpus_path) %>% as_tibble()
```

```{r eval-calls-overview}
own_eval_calls <- eval_calls %>% filter(eval_source == package) # package column is wrong currently
nb_eval_calls <- own_eval_calls %>% count(wt = nb_ev_calls) %>% select(n)

nb_eval_pres_calls <- own_eval_calls %>% filter(eval_function == "eval") %>% count(wt = nb_ev_calls) %>% select(n)
nb_evalq_calls <- own_eval_calls %>% filter(eval_function == "evalq") %>% count(wt = nb_ev_calls) %>% select(n)
nb_eval.parent_calls <- own_eval_calls %>% filter(eval_function == "eval.parent") %>% count(wt = nb_ev_calls) %>% select(n)
nb_local_calls <- own_eval_calls %>% filter(eval_function == "local") %>% count(wt = nb_ev_calls) %>% select(n)

eval_at_least_once <- own_eval_calls %>% count(package, wt = nb_ev_calls) %>% nrow()

r("nb eval and variant calls", nb_eval_calls)
r("nb eval calls", nb_eval_pres_calls)
r("nb evalq calls", nb_evalq_calls)
r("nb eval parent calls", nb_eval.parent_calls)
r("nb local calls", nb_local_calls)
r("nb package eval at least once", eval_at_least_once)
```

# Number of eval call sites

```{r nb-eval-call-sites}
nb_ev_call_sites <- eval_calls %>% select(eval_call_srcref) %>% n_distinct(na.rm = TRUE)

call_sites_per_package <- eval_calls %>% group_by(eval_source) %>% mutate(nb_call_sites = n_distinct(eval_call_srcref, na.rm = TRUE))

nb_packages_one_call_site <- call_sites_per_package %>% filter(nb_call_sites == 1) %>% nrow()

packages_at_least_one_call_site <- call_sites_per_package %>% filter(nb_call_sites >= 1)

median_call_sites <- packages_at_least_one_call_site %>% summarise(med = median(nb_call_sites, na.rm = TRUE))
max_packages <- packages_at_least_one_call_site %>% filter(nb_call_site == max(nb_call_sites, na.rm = TRUE)) %>% select(eval_source, nb_call_site)

nb_no_calls_to_eval <- nrow(corpus) - nrow(packages_at_least_one_call_site)

r("nb eval call sites", nb_ev_call_sites)
r("packages no calls to eval", nb_no_calls_to_eval)
```

```{r call-sites-violin-plot}
call_sites_per_package %>%
 ggplot(aes(x = eval_source, y = nb_call_sites, fill = eval_function), na.rm = TRUE) +
  geom_violin(trim = FALSE, width=1.7)+
  geom_boxplot(data = slice_sample(nb_ev_call_sites, n = 100000), width=.03) + # too slow otherwise
  #scale_y_log10(labels = scales::comma) +
  labs(y="Call sites per package") +
  theme(
    legend.position=c(0.5, 0.85),
    legend.box.background = element_rect(fill="white", size=0.1),
    legend.box="horizontal",
    axis.title.x=element_blank()
  )

ggsave(path(PLOT_DIR, "call_sites_per_packages.pdf"))
```

# Numbers of calls to eval

This is the number of eval calls which have a call site in each package.

```{r nb-eval-calls}
# run = package?
eval_calls_per_package <- own_eval_calls %>% count(eval_source, wt=nb_ev_calls)

average_eval_calls <- eval_calls_per_package %>% filter(n > 0) %>% summarise(, average = mean(n)) %>% .["average"]
max_eval_calls <- eval_calls_per_package %>% filter(n = max(n))

r("average eval calls per package", average_eval_calls) # average on packages with at least one eval call
r("max eval calls", max_eval_calls$n)
r("max eval call package", max_eval_calls$eval_source)
```

```{r eval-calls-violin-plot}
call_sites_per_package %>%
 ggplot(aes(x = eval_source, y = n), na.rm = TRUE) +
  geom_violin(trim = FALSE, width=1.7)+
  geom_boxplot(data = slice_sample(nb_ev_call_sites, n = 100000), width=.03) + # too slow otherwise
  #scale_y_log10(labels = scales::comma) +
  labs(y="Eval calls per package") +
  theme(
    legend.position=c(0.5, 0.85),
    legend.box.background = element_rect(fill="white", size=0.1),
    legend.box="horizontal",
    axis.title.x=element_blank()
  )

ggsave(path(PLOT_DIR, "call_sites_per_packages.pdf"))
```

```{r}
nb_eval_calls_tot <- eval_calls %>% count(wt = nb_ev_calls) %>% select(n)
```

If we don't exclude dependent calls, there are `r nb_eval_calls_tot` eval calls, so the proper eval calls are `r percent(nb_eval_calls, nb_eval_calls_tot)` of the total.

# Amount of code loaded by eval

```{r amout-of-code}

med_size <- own_eval_calls %>% summarise(med = weighted.median(expr_expression_length, na.rm = TRUE, w = nb_ev_calls))
max_size <- own_eval_calls %>% summarize(m = max(expr_expression_length, na.rm = TRUE))
average_size <- own_eval_calls %>% summarize(average = weithed.mean(expr_expression_length, na.rm = TRUE, w = nb_eval_calls))

sizes_prop <- own_eval_calls %>% count(expr_expression_size, wt = nb_ev_calls) %>%
                      mutate(proportion = n / sum(n)) %>% 
                      arrange(expr_expression_size) %>% 
                      mutate(cumulative = cumsum(proportion))

nb_nodes_many <- sizes_prop %>% filter(cumulative >= 0.95) %>% .[1,]

r("size nine five", nb_nodes_many$expr_expression_size)
r("size nine five exact percent", percent(nb_nodes_many$cumulative))

```

# Amount of computations via eval

```{r amount_computations}
nb_events_in_eval <- own_eval_calls %>% summarise(nb_events = sum(direct_interpreter_eval))

max_events_in_eval <- own_eval_calls %>% summarize(maxi = max(direct_interpreter_eval))

less_100_events <- own_eval_calls %>% filter(direct_interpreter_eval <= 100)
nb_less_100_events <- count(less_100_events, wt = nb_ev_calls)$n

r("nb events eval", nb_events_in_eval$nb_events)
r("maxi events eval", max_events_in_eval$maxi)
r("small events eval", percent(count(nb_less_100_events), nb_eval_calls))
```

```{r nb-events-violin-plot}
own_eval_calls %>%
 ggplot(aes(x = eval_source, y = direct_interpreter_eval), na.rm = TRUE) +
  geom_violin(trim = FALSE, width=1.7)+
  geom_boxplot(data = slice_sample(direct_interpreter_eval, n = 100000), width=.03) + # too slow otherwise
  #scale_y_log10(labels = scales::comma) +
  labs(y="Number of events in eval per package") +
  theme(
    legend.position=c(0.5, 0.85),
    legend.box.background = element_rect(fill="white", size=0.1),
    legend.box="horizontal",
    axis.title.x=element_blank()
  )

ggsave(path(PLOT_DIR, "nb_events_per_packages.pdf"))
```

# The case of core libraries

### 2020-12-04 inst/analysis/usage_metrics.Rmd
Approximation of the number of runs per package

@@
   base_dir: /var/lib/R/project-evalR
-  calls_packages_path: !r paste0(params$base_dir, "/run/package-evals-traced.5/summarized-packages.fst")
-  corpus_path: !r paste0(params$base_dir, "/revalstudy/inst/data/corpus.fst")
+  calls_packages_path: /run/package-evals-traced.7/summarized-externals.fst
+  corpus_path: /revalstudy/inst/data/corpus.fst
 ---
@@
 theme_set(theme_minimal())
-create_tags(path(TAGS_DIR, "enrichDataset.tex"), prefix="", default=TRUE)
+create_tags(path(TAGS_DIR, "usage_metrics.tex"), prefix="", default=TRUE)
 ```
@@
 ```{r read-data}
-eval_calls <- read_fst(params$calls_packages_path) %>% as_tibble()
+eval_calls <- read_fst(paste0(params$base_dir, params$calls_packages_path)) %>% as_tibble()
 ```
@@
 ```{r read_corpus}
-corpus <- read_fst(params$corpus_path) %>% as_tibble()
+corpus <- read_fst(paste0(params$base_dir, params$corpus_path)) %>% as_tibble()
 ```
@@
 ```{r eval-calls-overview}
-own_eval_calls <- eval_calls %>% filter(eval_source == package) # package column is wrong currently
-nb_eval_calls <- own_eval_calls %>% count(wt = nb_ev_calls) %>% select(n)
+#own_eval_calls <- eval_calls %>% filter(eval_source == package) # probably to remove
+own_eval_calls <- eval_calls
+nb_eval_calls <- (own_eval_calls %>% count(wt = nb_ev_calls))$n
 
-nb_eval_pres_calls <- own_eval_calls %>% filter(eval_function == "eval") %>% count(wt = nb_ev_calls) %>% select(n)
-nb_evalq_calls <- own_eval_calls %>% filter(eval_function == "evalq") %>% count(wt = nb_ev_calls) %>% select(n)
-nb_eval.parent_calls <- own_eval_calls %>% filter(eval_function == "eval.parent") %>% count(wt = nb_ev_calls) %>% select(n)
-nb_local_calls <- own_eval_calls %>% filter(eval_function == "local") %>% count(wt = nb_ev_calls) %>% select(n)
+nb_eval_pres_calls <- own_eval_calls %>% filter(eval_function == "eval") %>% count(wt = nb_ev_calls)
+nb_evalq_calls <- own_eval_calls %>% filter(eval_function == "evalq") %>% count(wt = nb_ev_calls)
+nb_eval.parent_calls <- own_eval_calls %>% filter(eval_function == "eval.parent") %>% count(wt = nb_ev_calls)
+nb_local_calls <- own_eval_calls %>% filter(eval_function == "local") %>% count(wt = nb_ev_calls)
 
@@
 r("nb eval and variant calls", nb_eval_calls)
-r("nb eval calls", nb_eval_pres_calls)
-r("nb evalq calls", nb_evalq_calls)
-r("nb eval parent calls", nb_eval.parent_calls)
-r("nb local calls", nb_local_calls)
+r("nb eval calls", nb_eval_pres_calls$n)
+r("nb evalq calls", nb_evalq_calls$n)
+r("nb eval parent calls", nb_eval.parent_calls$n)
+r("nb local calls", nb_local_calls$n)
 r("nb package eval at least once", eval_at_least_once)
@@
 median_call_sites <- packages_at_least_one_call_site %>% summarise(med = median(nb_call_sites, na.rm = TRUE))
-max_packages <- packages_at_least_one_call_site %>% filter(nb_call_site == max(nb_call_sites, na.rm = TRUE)) %>% select(eval_source, nb_call_site)
+max_packages <- packages_at_least_one_call_site %>% filter(nb_call_sites == max(nb_call_sites, na.rm = TRUE)) %>% select(eval_source, nb_call_sites)
 
-nb_no_calls_to_eval <- nrow(corpus) - nrow(packages_at_least_one_call_site)
+nb_no_calls_to_eval <- nrow(corpus) - nrow(packages_at_least_one_call_site) # corpus seems to be wrong
 
@@
   geom_violin(trim = FALSE, width=1.7)+
-  geom_boxplot(data = slice_sample(nb_ev_call_sites, n = 100000), width=.03) + # too slow otherwise
+  geom_boxplot(data = slice_sample(call_sites_per_package, n = 100000), width=.03) + # too slow otherwise
   #scale_y_log10(labels = scales::comma) +
@@
 
-ggsave(path(PLOT_DIR, "call_sites_per_packages.pdf"))
+#ggsave(path(PLOT_DIR, "call_sites_per_packages.pdf"))
 ```
@@
 
-This is the number of eval calls which have a call site in each package.
+This is the number of eval calls which have a call site in each package, divided by the number of runs for the package.
+
+We first compute an approximation of the number of runs per package.
+```{r runs-per-package}
+runs_package <- own_eval_calls %>% group_by(eval_source) %>% summarise(nb_runs = n_distinct(file)) %>% arrange(desc(nb_runs))
+```
+
+What are the packages with the most runs?
+
+```{r show_run-per-package}
+runs_package %>% datatable()
+```
 
@@
 # run = package?
-eval_calls_per_package <- own_eval_calls %>% count(eval_source, wt=nb_ev_calls)
+eval_calls_per_package <- own_eval_calls %>% count(eval_source, wt=nb_ev_calls) %>% left_join(runs_package, by = "eval_source") %>% mutate(n = n / nb_runs)
 
-average_eval_calls <- eval_calls_per_package %>% filter(n > 0) %>% summarise(, average = mean(n)) %>% .["average"]
-max_eval_calls <- eval_calls_per_package %>% filter(n = max(n))
+average_eval_calls <- eval_calls_per_package %>% filter(n > 0) %>% summarise(average = mean(n))
+max_eval_calls <- eval_calls_per_package %>% filter(n == max(n))
 
-r("average eval calls per package", average_eval_calls) # average on packages with at least one eval call
+r("average eval calls per package", average_eval_calls$average) # average on packages with at least one eval call
 r("max eval calls", max_eval_calls$n)
@@
 ```{r eval-calls-violin-plot}
-call_sites_per_package %>%
+eval_calls_per_package %>%
  ggplot(aes(x = eval_source, y = n), na.rm = TRUE) +
   geom_violin(trim = FALSE, width=1.7)+
-  geom_boxplot(data = slice_sample(nb_ev_call_sites, n = 100000), width=.03) + # too slow otherwise
+  geom_boxplot(data = slice_sample(eval_calls_per_package, n = 100000), width=.03) + # too slow otherwise
   #scale_y_log10(labels = scales::comma) +
@@
 
-ggsave(path(PLOT_DIR, "call_sites_per_packages.pdf"))
+ggsave(path(PLOT_DIR, "eval_calls_per_packages.pdf"))
 ```
@@
 ```{r}
-nb_eval_calls_tot <- eval_calls %>% count(wt = nb_ev_calls) %>% select(n)
+nb_eval_calls_tot <- eval_calls %>% count(wt = nb_ev_calls)
 ```
 
-If we don't exclude dependent calls, there are `r nb_eval_calls_tot` eval calls, so the proper eval calls are `r percent(nb_eval_calls, nb_eval_calls_tot)` of the total.
+If we don't exclude dependent calls, there are `r nb_eval_calls_tot` eval calls, so the proper eval calls are `r ratio(nb_eval_calls, nb_eval_calls_tot$n)` of the total.
 
@@
 
-med_size <- own_eval_calls %>% summarise(med = weighted.median(expr_expression_length, na.rm = TRUE, w = nb_ev_calls))
+weighted.median <- function(x, w) {
+  w <- w[order(x)]
+  x <- x[order(x)]
+
+  prob <- cumsum(w)/sum(w)
+  ps <- which(abs(prob - .5) == min(abs(prob - .5)))
+  return(x[ps])
+}
+
+med_size <- own_eval_calls %>% summarise(med = median(expr_expression_length, na.rm = TRUE)) # should be weighted by nb_ev_calls
 max_size <- own_eval_calls %>% summarize(m = max(expr_expression_length, na.rm = TRUE))
-average_size <- own_eval_calls %>% summarize(average = weithed.mean(expr_expression_length, na.rm = TRUE, w = nb_eval_calls))
+average_size <- weighted.mean(own_eval_calls$expr_expression_length, na.rm = TRUE, w = own_eval_calls$nb_ev_calls)
 
-sizes_prop <- own_eval_calls %>% count(expr_expression_size, wt = nb_ev_calls) %>%
+sizes_prop <- own_eval_calls %>% count(expr_expression_length, wt = nb_ev_calls) %>%
                       mutate(proportion = n / sum(n)) %>% 
-                      arrange(expr_expression_size) %>% 
+                      arrange(expr_expression_length) %>% 
                       mutate(cumulative = cumsum(proportion))
@@
 
-r("size nine five", nb_nodes_many$expr_expression_size)
+r("median size eval", med_size)
+r("max size eval", max_size)
+r("average size eval", average_size)
+r("size nine five", nb_nodes_many$expr_expression_length)
 r("size nine five exact percent", percent(nb_nodes_many$cumulative))
@@
 r("maxi events eval", max_events_in_eval$maxi)
-r("small events eval", percent(count(nb_less_100_events), nb_eval_calls))
+r("small events eval", ratio(nb_less_100_events, nb_eval_calls))
 ```
@@
   geom_violin(trim = FALSE, width=1.7)+
-  geom_boxplot(data = slice_sample(direct_interpreter_eval, n = 100000), width=.03) + # too slow otherwise
+  geom_boxplot(data = slice_sample(own_eval_calls, n = 100000), width=.03) + # too slow otherwise
   #scale_y_log10(labels = scales::comma) +
@@
 # The case of core libraries
+
+```{r}
+duration <- difftime(Sys.time(), now)
+```
+
+Notebook execution was `r duration` `r units(duration)`.

### 2020-12-04 inst/analysis/preprocess.R

@@ -569,8 +569,8 @@ defaultArgs <- c("data/corpus.fst", "../../run/package-evals-traced.7/calls.fst"
 # Only execute if it is launched as a script
 if (identical(environment(), globalenv())) {
 
-  if(FALSE) {
-    quit(status = commandArgs(trailingOnly = TRUE))
+  if(TRUE) {
+    quit(status = main(commandArgs(trailingOnly = TRUE)))
   }
   else {
     setwd("/var/lib/R/project-evalR/revalstudy/inst/")

### 2020-12-04 inst/analysis/usage_metrics.Rmd
Correct median and max for amount of code given to eval.

@@
 author: "Pierre Donat-Bouillud"
-output: html_document
+output:
+    html_document:
+        gallery: false
+        toc: true
+        toc_depth: 3
+        toc_float: true
+        df_print: paged
 date: "`r Sys.Date()`"
@@
   base_dir: /var/lib/R/project-evalR
-  calls_packages_path: /run/package-evals-traced.7/summarized-externals.fst
+  calls_packages_path: /run/package-evals-traced.9/summarized-externals.fst
   corpus_path: /revalstudy/inst/data/corpus.fst
@@
 library(fs)
+library(DT)
 
@@
 
-nb_no_calls_to_eval <- nrow(corpus) - nrow(packages_at_least_one_call_site) # corpus seems to be wrong
+nb_no_calls_to_eval <- nrow(corpus) - n_distinct(select(packages_at_least_one_call_site, eval_source))
 
@@
 call_sites_per_package %>%
- ggplot(aes(x = eval_source, y = nb_call_sites, fill = eval_function), na.rm = TRUE) +
+ ggplot(aes(x = eval_function, y = nb_call_sites, fill = eval_function), na.rm = TRUE) +
   geom_violin(trim = FALSE, width=1.7)+
-  geom_boxplot(data = slice_sample(call_sites_per_package, n = 100000), width=.03) + # too slow otherwise
-  #scale_y_log10(labels = scales::comma) +
+  #geom_jitter(shape=16, position=position_jitter(0.2)) +
   labs(y="Call sites per package") +
@@
 
-#ggsave(path(PLOT_DIR, "call_sites_per_packages.pdf"))
+ggsave(path(PLOT_DIR, "call_sites_per_packages.pdf"))
 ```
@@
 eval_calls_per_package %>%
- ggplot(aes(x = eval_source, y = n), na.rm = TRUE) +
+ ggplot(aes(x = "", y = n, fill = ""), na.rm = TRUE) +
   geom_violin(trim = FALSE, width=1.7)+
-  geom_boxplot(data = slice_sample(eval_calls_per_package, n = 100000), width=.03) + # too slow otherwise
-  #scale_y_log10(labels = scales::comma) +
   labs(y="Eval calls per package") +
   theme(
-    legend.position=c(0.5, 0.85),
-    legend.box.background = element_rect(fill="white", size=0.1),
-    legend.box="horizontal",
-    axis.title.x=element_blank()
+    legend.position= "none"
   )
@@
 
-```{r}
-nb_eval_calls_tot <- eval_calls %>% count(wt = nb_ev_calls)
-```
+<!-- How is the normalized distribution of eval calls?  -->
 
-If we don't exclude dependent calls, there are `r nb_eval_calls_tot` eval calls, so the proper eval calls are `r ratio(nb_eval_calls, nb_eval_calls_tot$n)` of the total.
+<!-- ```{r distribution-call-sites} -->
+<!-- eval_calls_per_package %>% -->
+<!--   ggplot(aes(x = n), na.rm = TRUE) + -->
+<!--   geom_density(adjust = 10) + -->
+<!--   scale_y_log10()  -->
+<!-- ``` -->
+
+<!-- (Like the violin plot, but smoothed) -->
+(th)
 
@@
 
-weighted.median <- function(x, w) {
+weighted.median <- function(x, w, na.rm=FALSE) {
+  if(na.rm)
+    na.omit(x)
   w <- w[order(x)]
@@
 
-med_size <- own_eval_calls %>% summarise(med = median(expr_expression_length, na.rm = TRUE)) # should be weighted by nb_ev_calls
-max_size <- own_eval_calls %>% summarize(m = max(expr_expression_length, na.rm = TRUE))
+med_size <-  weighted.median(own_eval_calls$expr_expression_length, w = own_eval_calls$nb_ev_calls, na.rm = TRUE)
+max_size <- max(own_eval_calls$expr_expression_length, na.rm = TRUE)
 average_size <- weighted.mean(own_eval_calls$expr_expression_length, na.rm = TRUE, w = own_eval_calls$nb_ev_calls)
@@
 ```{r nb-events-violin-plot}
-own_eval_calls %>%
- ggplot(aes(x = eval_source, y = direct_interpreter_eval), na.rm = TRUE) +
+own_eval_calls %>% filter(direct_interpreter_eval > 0) %>%
+ ggplot(aes(x = "", y = direct_interpreter_eval, fill = ""), na.rm = TRUE) +
   geom_violin(trim = FALSE, width=1.7)+
-  geom_boxplot(data = slice_sample(own_eval_calls, n = 100000), width=.03) + # too slow otherwise
-  #scale_y_log10(labels = scales::comma) +
+  scale_y_log10(labels = scales::comma) +
   labs(y="Number of events in eval per package") +
   theme(
-    legend.position=c(0.5, 0.85),
-    legend.box.background = element_rect(fill="white", size=0.1),
-    legend.box="horizontal",
-    axis.title.x=element_blank()
+    legend.position="none"
   )

### 2020-12-04 inst/analysis/preprocess.R

@@ -215,7 +215,7 @@ extract_package_name <- function(src_ref, file) {
   # There are 5 possibilities for a srcref:
   # - NA
   # - /tmp/Rtmp..../R.INSTALL....../packagename/R/file:linenumbers
-  # - /mnt/nvme0/R/project-evalR/library/4.0/instrumentr/srcref/packagename/4.0.4/file:linenumbers
+  # - mnt/ocfs_vol_00/project-evalR/library/4.0/instrumentr/srcref/packagename/4.0.4/file:linenumbers
   # - /R/* : core packages (we cannot distinguish between them yet so we write core for the package name)
   # - /testit/... or /testthat/... : it is the testit or testthat packages
   # - :/R : extract the package name from the file path in the column path
@@ -224,7 +224,7 @@ extract_package_name <- function(src_ref, file) {
     is.na(src_ref) ~ "base?",
     str_starts(src_ref, fixed("./R/")) ~ "core",
     str_starts(src_ref, fixed("/tmp/")) ~ str_match(src_ref, "/tmp.*/Rtmp[^/]*/R\\.INSTALL[^/]*/([^/]+)/R/.*$")[[2]], #path problem here
-    str_starts(src_ref, fixed("/mnt/nvme0/")) ~ "base",
+    str_starts(src_ref, fixed("/mnt/ocfs_vol")) ~ "base", #depends on where the shared is mounted!
     str_starts(src_ref, fixed("test")) ~ str_match(src_ref, "([^/]*)/.*")[[2]],
     str_starts(src_ref, fixed("/:")) ~ str_match(file, "[^/]*/[^/]*/([^/]*)/.*")[[2]],
     TRUE ~ "unknown"
@@ -448,7 +448,7 @@ main <- function(argv) {
 
   now_first <- Sys.time()
   corpus <- read_fst(corpus_file)
-  stopifnot(length(corpus) == 29)
+  #stopifnot(length(corpus) == 29)
   stopifnot("package" %in% names(corpus)) # for preprocess, we mainly care about the package names
 
   eval_calls_raw <-
@@ -502,8 +502,10 @@ main <- function(argv) {
 
   cat("only keep eval in corpus\n")
   now <- Sys.time()
-  # This seems to be already performed by the semijoin above. To remove?
-  corpus_files <- corpus %>% select(package)
+  # This is probably useless as there already was a filtering at tracing time.
+  # But this is a sanity check...
+  # Keep if quick
+  corpus_files <- corpus %>% select(package) %>% c()
[…]

### 2020-12-07 inst/analysis/preprocess.R
Fix corpus filtering

@@ -363,12 +363,12 @@ add_package <- function(dataset) {
 
 keep_only_corpus <- function(dataset, corpus_files) {
   return(dataset %>%
-    filter(eval_source %in% c(corpus_files, "core", "base", "base?")))
+    filter(eval_source %in% c(corpus_files$package, "core", "base", "base?")))
 }
 
 get_externals <- function(dataset, corpus_files) {
   return(dataset %>%
-    filter(!eval_source %in% c(corpus_files, "core", "base", "base?")))
+    filter(!eval_source %in% c(corpus_files$package, "core", "base", "base?")))
 }
 
 undefined_packages <- function(eval_calls) {

### 2020-12-07 inst/analysis/usage_metrics.Rmd
Fix path for usage_metrics.Rmd

@@
   base_dir: /var/lib/R/project-evalR
-  calls_packages_path: /run/package-evals-traced.9/summarized-externals.fst
+  calls_packages_path: /run/package-evals-traced.9/summarized-packages.fst
   corpus_path: /revalstudy/inst/data/corpus.fst

### 2020-12-07 inst/analysis/usage_metrics.Rmd
Adding some number for usage_metrics

@@
 
-call_sites_per_package <- eval_calls %>% group_by(eval_source) %>% mutate(nb_call_sites = n_distinct(eval_call_srcref, na.rm = TRUE))
+call_sites_per_package <- eval_calls %>% group_by(eval_source) %>% summarize(nb_call_sites = n_distinct(eval_call_srcref, na.rm = TRUE))
 
@@
 r("packages no calls to eval", nb_no_calls_to_eval)
+r("nb median call sites", median_call_sites$med)
+r("max call sites", max_packages$nb_call_sites)
+r("max call sites package", max_packages$eval_source)
+r("at least one call site", nrow(packages_at_least_one_call_site))
 ```
@@
 ```{r call-sites-violin-plot}
-call_sites_per_package %>%
+eval_calls %>% group_by(eval_source) %>% mutate(nb_call_sites = n_distinct(eval_call_srcref, na.rm = TRUE)) %>%
  ggplot(aes(x = eval_function, y = nb_call_sites, fill = eval_function), na.rm = TRUE) +

### 2020-12-07 inst/analysis/usage_metrics.Rmd
Update usage metrics

@@
 r("nb eval call sites", nb_ev_call_sites)
+r("nb one call site", nb_packages_one_call_site)
 r("packages no calls to eval", nb_no_calls_to_eval)

### 2020-12-07 inst/paper/main.tex
typo
Numbers and figures for usage metrics sections now in the .tex

@@
 
-We recorded \AllAllCallCountRnd calls to \eval and its variants (XXX001
-for \eval, XXX002 for \c{evalq}, XXX003 for \c{eval.parent} and XXX004 for
+We recorded \NbEvalAndVariantCallsRnd calls to \eval and its variants (\NbEvalCallsRnd
+for \eval, \NbEvalqCallsRnd for \c{evalq}, \NbEvalParentCallsRnd for \c{eval.parent} and \NbLocalCallsRnd for
 \c{local}).
 
-XXX005 packages in our corpus call \eval at least once.
+\NbPackageEvalAtLeastOnce packages in our corpus call \eval at least once.
 
@@
 We observed that the number of \eval call sites is small.
-Figure XXX006 shows a violin plot of the distribution of call sites to
+Figure~\ref{fig:callSitesPerPackage} shows a violin plot of the distribution of call sites to
 \eval and its three variants. For packages that have call sites, the
-median is XXX007 call sites. \OneCallSitesEnrich{} packages have a
-single call site, and one package, \emph{metafor}, has
-\MaxCallSitesEnrich{} call sites. \NoCallSitesEnrich{} packages have
-no calls to \eval. These numbers are under-approximations as \eval may
+median is \NbMedianCallSites call sites. \NbOneCallSite{} packages have a
+single call site, and one package, \emph{\MaxCallSitesPackage}, has
+\MaxCallSites{} call sites. \PackagesNoCallsToEval{} packages have
+no calls sites to \eval. These numbers are under-approximations as \eval may
 be aliased, called reflectively and given as argument to a
@@
 
-\mypara{Number of calls to eval.} We observed an average XXX010 calls
+\begin{figure}[h]
+	\centering 
+	\includegraphics[width=\textwidth]{call_sites_per_packages}
+	\caption{Violin plot of the call sites per packages per eval function. } \label{fig:callSitesPerPackage}
+\end{figure}
+
+\mypara{Number of calls to eval.} We observed an average \AverageEvalCallsPerPackage{} calls
 to \eval per run. \NOTE{this is in packages that had any call to eval.
@@
   runs that did not have an eval but executed some code from P? The
-  latter may be easier. Which one is right?} Figure XXX011 shows a
+  latter may be easier. Which one is right?} Figure~\ref{fig:evalCallsPerPackage} shows a
 violin plot of the distribution of calls to eval per run. The largest
-number of calls is XXX012 and comes from the package XXX013.
+number of calls per run is \MaxEvalCalls{} and comes from the package \MaxEvalCallPackage.
+
+\begin{figure}[h]
+	\centering 
+	\includegraphics[width=\textwidth]{eval_calls_per_packages}
+	\caption{Violin plot of the normalized number of eval calls per package. } \label{fig:evalCallsPerPackage}
+\end{figure}
 
@@
 input to \eval into strings and measure the size of that input. The
-largest input is XXX008 and the median input size is XXX009. Most
-inputs are small, less than XXX014 bytes. The average size is XXX015
+largest input is \MaxSizeEval and the median input size is \MedianSizeEval. Most
+inputs are small, less than \SizeNineFiveRnd bytes. The average size is \AverageSizeEval
 bytes.
@@
 across the whole corpus. For this, we use the number of interpreter
-steps as a proxy for events. We observe \AllEventCountRnd events all
-told, and \EvalEventCountRnd events occur inside eval
+steps as a proxy for events. We observe \AllEventCountRnd{} \NOTE{Check that number.\quad} events all
+told, and \NbEventsEvalRnd events occur inside eval
 (\EvalEventAllPerc of all events).
-Figure~XXX015 show a violin plot of events.
+Figure~\ref{fig:eventsPerPackage} show a violin plot of events.
+
+\begin{figure}[h]
+	\centering 
+	\includegraphics[width=\textwidth]{nb_events_per_packages}
+	\caption{Violin plot of the number of events per packages (log scale). } \label{fig:eventsPerPackage}
+\end{figure}
 
@@
 \eval calls. The largest number of events generated by an \eval call
-is \EventsMaxCountRnd. However, a majority of \eval calls performs up
-to \EventsMaxRangeA events suggesting that most expressions passed to
+is \MaxiEventsEvalRnd. However, a majority (\SmallEventsEval) of \eval calls performs up
+to 100 events suggesting that most expressions passed to
 \eval are small. It is interesting to note that a higher proportion of
@@
 
-\myoara{Source.}
+\mypara{Source.}
 There are various ways to obtain expressions:

### 2020-12-07 inst/analysis/preprocess.R
FOrmatting

@@ -223,8 +223,8 @@ extract_package_name <- function(src_ref, file) {
   case_when(
     is.na(src_ref) ~ "base?",
     str_starts(src_ref, fixed("./R/")) ~ "core",
-    str_starts(src_ref, fixed("/tmp/")) ~ str_match(src_ref, "/tmp.*/Rtmp[^/]*/R\\.INSTALL[^/]*/([^/]+)/R/.*$")[[2]], #path problem here
-    str_starts(src_ref, fixed("/mnt/ocfs_vol")) ~ "base", #depends on where the shared is mounted!
+    str_starts(src_ref, fixed("/tmp/")) ~ str_match(src_ref, "/tmp.*/Rtmp[^/]*/R\\.INSTALL[^/]*/([^/]+)/R/.*$")[[2]], # path problem here
+    str_starts(src_ref, fixed("/mnt/ocfs_vol")) ~ "base", # depends on where the shared is mounted!
     str_starts(src_ref, fixed("test")) ~ str_match(src_ref, "([^/]*)/.*")[[2]],
     str_starts(src_ref, fixed("/:")) ~ str_match(file, "[^/]*/[^/]*/([^/]*)/.*")[[2]],
     TRUE ~ "unknown"
@@ -243,7 +243,7 @@ find_package_name <-
     } # "Actually base but we generalize
     else {
       tempPack <- extract_package_name(srcref, file)
-      #It would indicate that one of the regex has failed and
+      # It would indicate that one of the regex has failed and
       # so that probably assumptions made on the paths in the srcref are no longer valid
       stopifnot(!is.na(tempPack))
       if (tempPack == "base?")
@@ -448,7 +448,7 @@ main <- function(argv) {
 
   now_first <- Sys.time()
   corpus <- read_fst(corpus_file)
-  #stopifnot(length(corpus) == 29)
+  # stopifnot(length(corpus) == 29)
   stopifnot("package" %in% names(corpus)) # for preprocess, we mainly care about the package names
 
   eval_calls_raw <-
@@ -458,8 +458,8 @@ main <- function(argv) {
       package = basename(dirname(dirname(dirname(file)))),
       corpus = "cran"
     )
-  #%>%
-  #semi_join(corpus, by = "package") # There might be more packages traced than what is in the corpus
+  # %>%
+  # semi_join(corpus, by = "package") # There might be more packages traced than what is in the corpus
   # stopifnot(length(eval_calls_raw) == 52)
 
   eval_calls_kaggle_raw <-
[…]

### 2020-12-07 inst/analysis/usage_metrics.Rmd
Eval calls per call site per un

@@
 
-knitr::opts_chunk$set(echo = TRUE,
+knitr::opts_chunk$set(
+  echo = TRUE,
   fig.retina = 2,
@@
   cache.lazy = FALSE,
-                      time_it = TRUE)
+  time_it = TRUE
+)
 
@@
 theme_set(theme_minimal())
-create_tags(path(TAGS_DIR, "usage_metrics.tex"), prefix="", default=TRUE)
+create_tags(path(TAGS_DIR, "usage_metrics.tex"), prefix = "", default = TRUE)
 ```
@@
 ```{r eval-calls-overview}
-#own_eval_calls <- eval_calls %>% filter(eval_source == package) # probably to remove
+# own_eval_calls <- eval_calls %>% filter(eval_source == package) # probably to remove
 own_eval_calls <- eval_calls
@@
 
-nb_eval_pres_calls <- own_eval_calls %>% filter(eval_function == "eval") %>% count(wt = nb_ev_calls)
-nb_evalq_calls <- own_eval_calls %>% filter(eval_function == "evalq") %>% count(wt = nb_ev_calls)
-nb_eval.parent_calls <- own_eval_calls %>% filter(eval_function == "eval.parent") %>% count(wt = nb_ev_calls)
-nb_local_calls <- own_eval_calls %>% filter(eval_function == "local") %>% count(wt = nb_ev_calls)
+nb_eval_pres_calls <- own_eval_calls %>%
+  filter(eval_function == "eval") %>%
+  count(wt = nb_ev_calls)
+nb_evalq_calls <- own_eval_calls %>%
+  filter(eval_function == "evalq") %>%
+  count(wt = nb_ev_calls)
+nb_eval.parent_calls <- own_eval_calls %>%
+  filter(eval_function == "eval.parent") %>%
+  count(wt = nb_ev_calls)
+nb_local_calls <- own_eval_calls %>%
+  filter(eval_function == "local") %>%
+  count(wt = nb_ev_calls)
 
-eval_at_least_once <- own_eval_calls %>% count(package, wt = nb_ev_calls) %>% nrow()
+eval_at_least_once <- own_eval_calls %>%
+  count(package, wt = nb_ev_calls) %>%
+  nrow()
 
@@
 ```{r nb-eval-call-sites}
-nb_ev_call_sites <- eval_calls %>% select(eval_call_srcref) %>% n_distinct(na.rm = TRUE)
+nb_ev_call_sites <- eval_calls %>%
+  select(eval_call_srcref) %>%
+  n_distinct(na.rm = TRUE)
 
-call_sites_per_package <- eval_calls %>% group_by(eval_source) %>% summarize(nb_call_sites = n_distinct(eval_call_srcref, na.rm = TRUE))
+call_sites_per_package <- eval_calls %>%
+  group_by(eval_source) %>%
+  summarize(nb_call_sites = n_distinct(eval_call_srcref, na.rm = TRUE))
 
-nb_packages_one_call_site <- call_sites_per_package %>% filter(nb_call_sites == 1) %>% nrow()
+nb_packages_one_call_site <- call_sites_per_package %>%
+  filter(nb_call_sites == 1) %>%
+  nrow()
 
@@
 median_call_sites <- packages_at_least_one_call_site %>% summarise(med = median(nb_call_sites, na.rm = TRUE))
-max_packages <- packages_at_least_one_call_site %>% filter(nb_call_sites == max(nb_call_sites, na.rm = TRUE)) %>% select(eval_source, nb_call_sites)
+max_packages <- packages_at_least_one_call_site %>%
+  filter(nb_call_sites == max(nb_call_sites, na.rm = TRUE)) %>%
+  select(eval_source, nb_call_sites)
 
@@
 ```{r call-sites-violin-plot}
-eval_calls %>% group_by(eval_source) %>% mutate(nb_call_sites = n_distinct(eval_call_srcref, na.rm = TRUE)) %>%
+eval_calls %>%
+  group_by(eval_source) %>%
+  mutate(nb_call_sites = n_distinct(eval_call_srcref, na.rm = TRUE)) %>%
   ggplot(aes(x = eval_function, y = nb_call_sites, fill = eval_function), na.rm = TRUE) +
-  geom_violin(trim = FALSE, width=1.7)+
-  #geom_jitter(shape=16, position=position_jitter(0.2)) +
-  labs(y="Call sites per package") +
+  geom_violin(trim = FALSE, width = 1.7) +
+  # geom_jitter(shape=16, position=position_jitter(0.2)) +
+  labs(y = "Call sites per package") +
   theme(
-    legend.position=c(0.5, 0.85),
-    legend.box.background = element_rect(fill="white", size=0.1),
-    legend.box="horizontal",
-    axis.title.x=element_blank()
+    legend.position = c(0.5, 0.85),
+    legend.box.background = element_rect(fill = "white", size = 0.1),
+    legend.box = "horizontal",
+    axis.title.x = element_blank()
   )
@@
 ```{r runs-per-package}
-runs_package <- own_eval_calls %>% group_by(eval_source) %>% summarise(nb_runs = n_distinct(file)) %>% arrange(desc(nb_runs))
+runs_package <- own_eval_calls %>%
+  group_by(eval_source) %>%
+  summarise(nb_runs = n_distinct(file)) %>%
+  arrange(desc(nb_runs))
 ```
@@
 # run = package?
-eval_calls_per_package <- own_eval_calls %>% count(eval_source, wt=nb_ev_calls) %>% left_join(runs_package, by = "eval_source") %>% mutate(n = n / nb_runs)
+eval_calls_per_package <- own_eval_calls %>%
+  count(eval_source, wt = nb_ev_calls) %>%
+  left_join(runs_package, by = "eval_source") %>%
+  mutate(n = n / nb_runs)
 
-average_eval_calls <- eval_calls_per_package %>% filter(n > 0) %>% summarise(average = mean(n))
+average_eval_calls <- eval_calls_per_package %>%
+  filter(n > 0) %>%
+  summarise(average = mean(n))
 max_eval_calls <- eval_calls_per_package %>% filter(n == max(n))
@@
   ggplot(aes(x = "", y = n, fill = ""), na.rm = TRUE) +
-  geom_violin(trim = FALSE, width=1.7)+
-  labs(y="Eval calls per package") +
+  geom_violin(trim = FALSE, width = 1.7) +
+  labs(y = "Eval calls per package") +
   theme(
-    legend.position= "none"
+    legend.position = "none"
   )
@@
 
-weighted.median <- function(x, w, na.rm=FALSE) {
-  if(na.rm)
+weighted.median <- function(x, w, na.rm = FALSE) {
+  if (na.rm) {
     na.omit(x)
+  }
   w <- w[order(x)]
@@
 
-  prob <- cumsum(w)/sum(w)
+  prob <- cumsum(w) / sum(w)
   ps <- which(abs(prob - .5) == min(abs(prob - .5)))
@@
 
-sizes_prop <- own_eval_calls %>% count(expr_expression_length, wt = nb_ev_calls) %>%
+sizes_prop <- own_eval_calls %>%
+  count(expr_expression_length, wt = nb_ev_calls) %>%
   mutate(proportion = n / sum(n)) %>%
@@
 
-nb_nodes_many <- sizes_prop %>% filter(cumulative >= 0.95) %>% .[1,]
+nb_nodes_many <- sizes_prop %>%
+  filter(cumulative >= 0.95) %>%
+  .[1, ]
 
@@
 r("size nine five exact percent", percent(nb_nodes_many$cumulative))
-
 ```
@@
 ```{r nb-events-violin-plot}
-own_eval_calls %>% filter(direct_interpreter_eval > 0) %>%
+own_eval_calls %>%
+  filter(direct_interpreter_eval > 0) %>%
   ggplot(aes(x = "", y = direct_interpreter_eval, fill = ""), na.rm = TRUE) +
-  geom_violin(trim = FALSE, width=1.7)+
+  geom_violin(trim = FALSE, width = 1.7) +
   scale_y_log10(labels = scales::comma) +
-  labs(y="Number of events in eval per package") +
+  labs(y = "Number of events in eval per package") +
   theme(
-    legend.position="none"
+    legend.position = "none"
   )
@@
 ggsave(path(PLOT_DIR, "nb_events_per_packages.pdf"))
+```
+
+# Loops
+
+Here, we try to have an estimation of how often a call site is exercised per run. If it is often, maybe it is a loop!
+
+```{r loops}
+eval_calls_exercised <- eval_calls %>%
+  group_by(file) %>%
+  count(eval_call_srcref, wt = nb_ev_calls)
+
+eval_calls_exercised %>% ggplot(aes(n)) +
+  geom_histogram() +
+  scale_x_log10(labels = scales::comma) +
+  labs(x = "Number of eval calls per srcref per run")
 ```

### 2020-12-08 inst/analysis/preprocess.R
Simplify preprocessing

@@ -363,12 +363,12 @@ add_package <- function(dataset) {
 
 keep_only_corpus <- function(dataset, corpus_files) {
   return(dataset %>%
-    filter(eval_source %in% c(corpus_files$package, "core", "base", "base?")))
+           semi_join(corpus_files, by = c("eval_source" = "package")))
 }
 
 get_externals <- function(dataset, corpus_files) {
   return(dataset %>%
-    filter(!eval_source %in% c(corpus_files$package, "core", "base", "base?")))
+           anti_join(corpus_files, by = c("eval_source" = "package")))
 }
 
 undefined_packages <- function(eval_calls) {
@@ -389,9 +389,10 @@ undefined_packages <- function(eval_calls) {
 }
 
 
+# This is performed directly in usage_metrics.Rmd
 known_call_sites <- function(eval_calls_corpus, corpus_files) {
   call_sites_per_package <- eval_calls_corpus %>%
-    filter(eval_source_type == "package", eval_source %in% corpus_files) %>%
+    filter(eval_source_type == "package", eval_source %in% corpus_files$package) %>%
     group_by(eval_source) %>%
     summarize(n = n_distinct(eval_call_srcref))
 
@@ -505,9 +506,7 @@ main <- function(argv) {
   # This is probably useless as there already was a filtering at tracing time.
   # But this is a sanity check...
   # Keep if quick
-  corpus_files <- corpus %>%
-    select(package) %>%
-    c()
+  corpus_files <- select(corpus, package) %>% bind_rows(tribble(~package, "core", "base", "base?"))
   eval_calls_corpus <- eval_calls %>% keep_only_corpus(corpus_files)
   eval_calls_externals <- eval_calls %>% get_externals(corpus_files)
   res <- difftime(Sys.time(), now)
@@ -532,12 +531,12 @@ main <- function(argv) {
   res <- difftime(Sys.time(), now)
   cat("Done in ", res, units(res), "\n")
 
-  cat("Number of call sites per package\n")
-  now <- Sys.time()
-  calls_site_per_package <-
-    known_call_sites(eval_calls_corpus, corpus_file)
-  res <- difftime(Sys.time(), now)
-  cat("Done in ", res, units(res), "\n")
+  # cat("Number of call sites per package\n")
+  # now <- Sys.time()
+  # calls_site_per_package <-
+  #   known_call_sites(eval_calls_corpus, corpus_files)
+  # res <- difftime(Sys.time(), now)
[…]

### 2020-12-09 inst/analysis/usage_metrics.Rmd

@@
   calls_packages_path: /run/package-evals-traced.9/summarized-packages.fst
-  corpus_path: /revalstudy/inst/data/corpus.fst
+  corpus_path: /run/package-evals-traced.9/corpus.fst
 ---

### 2020-12-09 inst/analysis/usage_metrics.Rmd
Meaning preserving refactoring

@@
 
+The following is a set up chunk. Currently there is an variable "jan" 
+used to run the script from docker on my machine. This should be
+removed. (TODO)
+
 ```{r setup, include=FALSE}
+library(tidyverse)
+library(fst)
+library(fs)
+library(DT)
+
+jan <- FALSE
+
 knitr::knit_hooks$set(time_it = local({
@@
   function(before, options) {
-    if (before) {
-      # record the current time before each chunk
+    if (before) {.       # record the current time before each chunk
       now <<- Sys.time()
-    } else {
-      # calculate the time difference after a chunk
+    } else {             # calculate the time difference after a chunk
       res <- difftime(Sys.time(), now)
@@
       paste("Time for this code chunk to run:", res, units(res))
-    }
-  }
-}))
-
+}}}))
+now <- Sys.time()
 knitr::opts_chunk$set(
@@
   cache.lazy = FALSE,
-  time_it = TRUE
-)
+  time_it = TRUE)
 
-now <- Sys.time()
-library(tidyverse)
-library(stringr)
-library(fst)
-library(fs)
-library(DT)
-
-source("inc/paths.R")
+if (jan == FALSE) { source("inc/paths.R") }
 source("inc/latextags.R")
@@
 theme_set(theme_minimal())
+if (jan) { TAGS_DIR <- PLOT_DIR <- "." }
 create_tags(path(TAGS_DIR, "usage_metrics.tex"), prefix = "", default = TRUE)
+
+base_dir <- params$base_dir
+# Set the paths, and customize for Jan.
+if(jan) { base_dir <- "/mnt/code/eval" }
+eval_calls_path <- paste0(base_dir, params$calls_packages_path)
+corpus_path <- paste0(base_dir, params$corpus_path)
 ```
 
-```{r file-info, echo=FALSE}
-file_info(params$calls_packages_path)
+Read in the summarized calls data and the file that describes
+which package belong to our corpus.
+
+```{r read}
+read_fst(eval_calls_path) %>% as_tibble() -> E_raw
+read_fst(corpus_path)     %>% as_tibble() -> C_raw
 ```
 
-```{r read-data}
-eval_calls <- read_fst(paste0(params$base_dir, params$calls_packages_path)) %>% as_tibble()
+The two data sets are E and C for conciseness.  C is only used here to compute
+the corpus size. E is used throughout. Recall that E_raw has the source data
+with all columns.
+
+```{r}
+E_raw %>% select(package, eval_function, nb_ev_calls,
+                 eval_call_srcref, file, eval_source, 
+                 expr_expression_length,
+                 direct_interpreter_eval)-> E
+C_raw -> C
+corpus_size <- nrow(C)
 ```
 
-```{r read_corpus}
-corpus <- read_fst(paste0(params$base_dir, params$corpus_path)) %>% as_tibble()
-```
+Compute the number of times eval and it variants were called in the entire 
+corpus. This is a dynamic count that aggregates all call sites within
+the corpus.
 
 ```{r eval-calls-overview}
-# own_eval_calls <- eval_calls %>% filter(eval_source == package) # probably to remove
-own_eval_calls <- eval_calls
-nb_eval_calls <- (own_eval_calls %>% count(wt = nb_ev_calls))$n
+nb_calls <- (E %>% count(wt=nb_ev_calls))$n
+eval_calls <- E %>% filter(eval_function=="eval") %>% count(wt=nb_ev_calls)
+evalq_calls <- E %>% filter(eval_function=="evalq") %>% count(wt=nb_ev_calls)
+eval.parent_calls <- E %>% filter(eval_function=="eval.parent") %>%
+  count(wt = nb_ev_calls)
+local_calls <- E %>% filter(eval_function=="local") %>% count(wt = nb_ev_calls)
+eval_at_least_once <- E %>% count(package, wt=nb_ev_calls)
 
-nb_eval_pres_calls <- own_eval_calls %>%
-  filter(eval_function == "eval") %>%
-  count(wt = nb_ev_calls)
-nb_evalq_calls <- own_eval_calls %>%
-  filter(eval_function == "evalq") %>%
-  count(wt = nb_ev_calls)
-nb_eval.parent_calls <- own_eval_calls %>%
-  filter(eval_function == "eval.parent") %>%
-  count(wt = nb_ev_calls)
-nb_local_calls <- own_eval_calls %>%
-  filter(eval_function == "local") %>%
-  count(wt = nb_ev_calls)
-
-eval_at_least_once <- own_eval_calls %>%
-  count(package, wt = nb_ev_calls) %>%
-  nrow()
-
-r("nb eval and variant calls", nb_eval_calls)
-r("nb eval calls", nb_eval_pres_calls$n)
-r("nb evalq calls", nb_evalq_calls$n)
-r("nb eval parent calls", nb_eval.parent_calls$n)
-r("nb local calls", nb_local_calls$n)
-r("nb package eval at least once", eval_at_least_once)
+r("nb eval and variant calls", nb_calls)
+r("nb eval calls", eval_calls$n)
+r("nb evalq calls", evalq_calls$n)
+r("nb eval parent calls", eval.parent_calls$n)
+r("nb local calls", local_calls$n)
+r("nb package eval at least once", nrow(eval_at_least_once))
 ```
@@
 ```{r nb-eval-call-sites}
-nb_ev_call_sites <- eval_calls %>%
+nb_ev_call_sites <- E %>%
   select(eval_call_srcref) %>%
@@
 
-call_sites_per_package <- eval_calls %>%
+call_sites_per_package <- E %>%
   group_by(eval_source) %>%
@@
   nrow()
-
-packages_at_least_one_call_site <- call_sites_per_package %>% filter(nb_call_sites >= 1)
-
-median_call_sites <- packages_at_least_one_call_site %>% summarise(med = median(nb_call_sites, na.rm = TRUE))
+packages_at_least_one_call_site <- call_sites_per_package %>% 
+  filter(nb_call_sites >= 1)
+median_call_sites <- packages_at_least_one_call_site %>% 
+  summarise(med = median(nb_call_sites, na.rm = TRUE))
 max_packages <- packages_at_least_one_call_site %>%
@@
   select(eval_source, nb_call_sites)
-
-nb_no_calls_to_eval <- nrow(corpus) - n_distinct(select(packages_at_least_one_call_site, eval_source))
+nb_no_calls_to_eval <- corpus_size - 
+  n_distinct(select(packages_at_least_one_call_site, eval_source))
 
@@
 ```{r call-sites-violin-plot}
-eval_calls %>%
+E %>%
   group_by(eval_source) %>%
@@
 ```{r runs-per-package}
-runs_package <- own_eval_calls %>%
+runs_package <- E %>%
   group_by(eval_source) %>%
@@
 # run = package?
-eval_calls_per_package <- own_eval_calls %>%
+eval_calls_per_package <- E %>%
   count(eval_source, wt = nb_ev_calls) %>%
@@
   labs(y = "Eval calls per package") +
-  theme(
-    legend.position = "none"
-  )
+  theme(legend.position = "none")
 
@@
 
-med_size <- weighted.median(own_eval_calls$expr_expression_length, w = own_eval_calls$nb_ev_calls, na.rm = TRUE)
-max_size <- max(own_eval_calls$expr_expression_length, na.rm = TRUE)
-average_size <- weighted.mean(own_eval_calls$expr_expression_length, na.rm = TRUE, w = own_eval_calls$nb_ev_calls)
+med_size <- weighted.median(E$expr_expression_length, w = E$nb_ev_calls, na.rm = TRUE)
+max_size <- max(E$expr_expression_length, na.rm = TRUE)
+average_size <- weighted.mean(E$expr_expression_length, na.rm = TRUE, w = E$nb_ev_calls)
 
-sizes_prop <- own_eval_calls %>%
+sizes_prop <- E %>%
   count(expr_expression_length, wt = nb_ev_calls) %>%
@@
 ```{r amount_computations}
-nb_events_in_eval <- own_eval_calls %>% summarise(nb_events = sum(direct_interpreter_eval))
+nb_events_in_eval <- E %>% summarise(nb_events = sum(direct_interpreter_eval))
 
-max_events_in_eval <- own_eval_calls %>% summarize(maxi = max(direct_interpreter_eval))
+max_events_in_eval <- E %>% summarize(maxi = max(direct_interpreter_eval))
 
-less_100_events <- own_eval_calls %>% filter(direct_interpreter_eval <= 100)
+less_100_events <- E %>% filter(direct_interpreter_eval <= 100)
 nb_less_100_events <- count(less_100_events, wt = nb_ev_calls)$n
@@
 r("maxi events eval", max_events_in_eval$maxi)
-r("small events eval", ratio(nb_less_100_events, nb_eval_calls))
+r("small events eval", ratio(nb_less_100_events, nb_calls))
 ```
@@
 ```{r nb-events-violin-plot}
-own_eval_calls %>%
+E %>%
   filter(direct_interpreter_eval > 0) %>%
@@
   labs(y = "Number of events in eval per package") +
-  theme(
-    legend.position = "none"
-  )
+  theme(legend.position = "none")
 
@@
 ```{r loops}
-eval_calls_exercised <- eval_calls %>%
-  group_by(file) %>%
+eval_calls_exercised <- E %>% group_by(file) %>%
   count(eval_call_srcref, wt = nb_ev_calls)
@@
 eval_calls_exercised %>% ggplot(aes(n)) +
-  geom_histogram() +
-  scale_x_log10(labels = scales::comma) +
+  geom_histogram() + scale_x_log10(labels = scales::comma) +
   labs(x = "Number of eval calls per srcref per run")

### 2020-12-09 inst/analysis/inc/latextags.R

@@ -130,7 +130,8 @@ create_tags <- function(filename, prefix="", default=TRUE) {
 }
 
 clear_tags <- function(tags=get_default_tags()) {
-  tags$values <- tibble::tibble(name=character(), value=character(), latex=character())
+  tags$values <- tibble::tibble(name=character(), 
+                                value=character(), latex=character())
 }
 
 tags_filename <- function(tags) {
@@ -242,7 +243,7 @@ generate_latex_command <- function(command_name, value) {
   stopifnot(length(command_name) == length(value))
   
   stringr::str_c(
-    "\\newcommand{\\", command_name, "}{", latex_escape_value(value), "\\xspace}"
+    "\\newcommand{\\", command_name, "}{", latex_escape_value(value),"\\xspace}"
   )
 }
 
@@ -317,7 +318,7 @@ fmt <- function(x, prefix="", suffix="", ...) {
   format(x, big.mark=",", trim=TRUE)
 }
 
-# TODO: make this parameters part of some global options so they can be set globally
+# TODO: make this parameter a global option so it can be set globally
 # TODO: test c(0.0123, 1.123, -0.0123, -1.123)
 .fmt.double <- function(x, digits=1, floor=FALSE, ceiling=FALSE) {
   if (floor) x <- floor(x)

### 2020-12-09 inst/analysis/usage_metrics.Rmd
Dataset .10

--- 2020-12-07 inst/analysis/usage_metrics.Rmd
@@
   base_dir: /var/lib/R/project-evalR
-  calls_packages_path: /run/package-evals-traced.9/summarized-packages.fst
-  corpus_path: /revalstudy/inst/data/corpus.fst
+  calls_packages_path: /run/package-evals-traced.10/summarized-packages.fst
+  corpus_path: /run/package-evals-traced.10/corpus.fst
 ---
@@
 ```{r file-info, echo=FALSE}
-file_info(params$calls_packages_path)
-```
 
-```{r read-data}
-eval_calls <- read_fst(paste0(params$base_dir, params$calls_packages_path)) %>% as_tibble()
+calls_packages_path <- paste0(params$base_dir, params$calls_packages_path)
+file_info(calls_packages_path)
+
+eval_calls <- read_fst(calls_packages_path) %>% as_tibble()
 ```
@@
 ```{r eval-calls-overview}
-# own_eval_calls <- eval_calls %>% filter(eval_source == package) # probably to remove
 own_eval_calls <- eval_calls

### 2020-12-09 inst/analysis/preprocess.R

@@ -509,7 +509,7 @@ main <- function(argv) {
   corpus_files <- select(corpus, package) %>% bind_rows(tribble(~package, "core", "base", "base?"))
   eval_calls_corpus <- eval_calls %>% keep_only_corpus(corpus_files)
   eval_calls_externals <- eval_calls %>% get_externals(corpus_files)
-  res <- difftime(Sys.time(), now)
+  # res <- difftime(Sys.time(), now)
   cat("Done in ", res, units(res), "\n")
 
   # Separate datasets
@@ -553,7 +553,7 @@ main <- function(argv) {
 
   write_fst(eval_calls_externals, evals_summarized_externals_file)
 
-  write_fst(calls_site_per_package, package_evals_dynamic_file)
+  #write_fst(calls_site_per_package, package_evals_dynamic_file)
   res <- difftime(Sys.time(), now)
   cat("Done in ", res, units(res), "\n")
 

### 2020-12-09 inst/analysis/jan_usage_metrics.Rmd

---
title: "Usage metrics"
author: "Pierre Donat-Bouillud"
output:
    html_document:
        gallery: false
        toc: true
        toc_depth: 3
        toc_float: true
        df_print: paged
date: "`r Sys.Date()`"
editor_options: 
  chunk_output_type: console
params:
  base_dir: /Users/jan/mine/code/eval
  calls_packages_path: /run/package-evals-traced.9/summarized-packages.fst
  corpus_path: /revalstudy/inst/data/corpus.fst
---

```{r setup, include=FALSE}
knitr::knit_hooks$set(time_it = local({
  now <- NULL
  function(before, options) {
    if (before) {
      # record the current time before each chunk
      now <<- Sys.time()
    } else {
      # calculate the time difference after a chunk
      res <- difftime(Sys.time(), now)
      # return a character string to show the time
      paste("Time for this code chunk to run:", res, units(res))
    }
  }
}))

knitr::opts_chunk$set(echo = TRUE,
                      fig.retina = 2,
                      fig.width = 10,
                      cache.lazy = FALSE,
                      time_it = TRUE)
now <- Sys.time() 
library(tidyverse)
library(stringr)
library(fst)
library(fs)
library(DT)

source("inc/paths.R")
source("inc/latextags.R")
source("inc/setup.R")
theme_set(theme_minimal())
create_tags(path(TAGS_DIR, "usage_metrics.tex"), prefix="", default=TRUE)
```

```{r file-info, echo=FALSE}
calls_package_path <- paste0(params$base_dir, params$calls_packages_path)
file_info(calls_packages_path)

eval_calls <- read_fst(calls_packages_path) %>% as_tibble()
corpus <- read_fst(paste0(params$base_dir, params$corpus_path)) %>% as_tibble()
```

```{r eval-calls-overview}
nb_eval_calls <- (eval_calls %>% count(wt = nb_ev_calls))$n                                         # eval and all of its variants
nb_eval_pres_calls <- eval_calls %>% filter(eval_function == "eval") %>% count(wt = nb_ev_calls)    # just eval
nb_evalq_calls <- eval_calls %>% filter(eval_function == "evalq") %>% count(wt = nb_ev_calls)       # number of evalq
nb_eval.parent_calls <- eval_calls %>% filter(eval_function == "eval.parent") %>% count(wt = nb_ev_calls) # eval.parent  
nb_local_calls <- eval_calls %>% filter(eval_function == "local") %>% count(wt = nb_ev_calls)       # local
eval_at_least_once <- eval_calls %>% count(package, wt = nb_ev_calls) %>% nrow()                    # number of packages that have at least one call

r("nb eval and variant calls", nb_eval_calls)
r("nb eval calls", nb_eval_pres_calls$n)
r("nb evalq calls", nb_evalq_calls$n)
r("nb eval parent calls", nb_eval.parent_calls$n)
r("nb local calls", nb_local_calls$n)
r("nb package eval at least once", eval_at_least_once)
```

# Number of eval call sites

```{r nb-eval-call-sites}
nb_ev_call_sites <- eval_calls %>% select(eval_call_srcref) %>% n_distinct(na.rm = TRUE)   # underestimation of the potential number of call sites

call_sites_per_package <- eval_calls %>% group_by(eval_source) %>% summarize(nb_call_sites = n_distinct(eval_call_srcref, na.rm = TRUE))

nb_packages_one_call_site <- call_sites_per_package %>% filter(nb_call_sites == 1) %>% nrow()

packages_at_least_one_call_site <- call_sites_per_package %>% filter(nb_call_sites >= 1)

median_call_sites <- packages_at_least_one_call_site %>% summarise(med = median(nb_call_sites, na.rm = TRUE))
max_packages <- packages_at_least_one_call_site %>% filter(nb_call_sites == max(nb_call_sites, na.rm = TRUE)) %>% select(eval_source, nb_call_sites)  #largest num of eval

nb_no_calls_to_eval <- nrow(corpus) - n_distinct(select(packages_at_least_one_call_site, eval_source))  #packages with no eval (or variants) in source code

r("nb eval call sites", nb_ev_call_sites)                                
r("nb one call site", nb_packages_one_call_site)
r("packages no calls to eval", nb_no_calls_to_eval)
r("nb median call sites", median_call_sites$med)
r("max call sites", max_packages$nb_call_sites)
r("max call sites package", max_packages$eval_source)
r("at least one call site", nrow(packages_at_least_one_call_site))
```

```{r call-sites-violin-plot}
eval_calls %>% group_by(eval_source) %>% mutate(nb_call_sites = n_distinct(eval_call_srcref, na.rm = TRUE)) %>%
 ggplot(aes(x = eval_function, y = nb_call_sites, fill = eval_function), na.rm = TRUE) +
  geom_violin(trim = FALSE, width=1.7)+
  #geom_jitter(shape=16, position=position_jitter(0.2)) +
  labs(y="Call sites per package") +
    theme(
    legend.position=c(0.5, 0.85),
    legend.box.background = element_rect(fill="white", size=0.1),
    legend.box="horizontal",
    axis.title.x=element_blank()
  )

ggsave(path(PLOT_DIR, "call_sites_per_packages.pdf"))
```

# Numbers of calls to eval

This is the number of eval calls which have a call site in each package, divided by the number of runs for the package.

We first compute an approximation of the number of runs per package.
```{r runs-per-package}
runs_package <- eval_calls %>% group_by(eval_source) %>% summarise(nb_runs = n_distinct(file)) %>% arrange(desc(nb_runs))
```

What are the packages with the most runs?

```{r show_run-per-package}
runs_package %>% datatable()
```

```{r nb-eval-calls}
# run = package?
eval_calls_per_package <- eval_calls %>% count(eval_source, wt=nb_ev_calls) %>% left_join(runs_package, by = "eval_source") %>% mutate(n = n / nb_runs)

average_eval_calls <- eval_calls_per_package %>% filter(n > 0) %>% summarise(average = mean(n))
max_eval_calls <- eval_calls_per_package %>% filter(n == max(n))

r("average eval calls per package", average_eval_calls$average) # average on packages with at least one eval call
r("max eval calls", max_eval_calls$n)
r("max eval call package", max_eval_calls$eval_source)
```

```{r eval-calls-violin-plot}
eval_calls_per_package %>%
 ggplot(aes(x = "", y = n, fill = ""), na.rm = TRUE) +
  geom_violin(trim = FALSE, width=1.7)+
  labs(y="Eval calls per package") +
  theme(
    legend.position= "none"
  )

ggsave(path(PLOT_DIR, "eval_calls_per_packages.pdf"))
```

<!-- How is the normalized distribution of eval calls?  -->

<!-- ```{r distribution-call-sites} -->
<!-- eval_calls_per_package %>% -->
<!--   ggplot(aes(x = n), na.rm = TRUE) + -->
<!--   geom_density(adjust = 10) + -->
<!--   scale_y_log10()  -->
<!-- ``` -->

<!-- (Like the violin plot, but smoothed) -->
(th)

# Amount of code loaded by eval

```{r amout-of-code}

weighted.median <- function(x, w, na.rm=FALSE) {
  if(na.rm)
    na.omit(x)
  w <- w[order(x)]
  x <- x[order(x)]

  prob <- cumsum(w)/sum(w)
  ps <- which(abs(prob - .5) == min(abs(prob - .5)))
  return(x[ps])
}

med_size <-  weighted.median(eval_calls$expr_expression_length, w = eval_calls$nb_ev_calls, na.rm = TRUE)
max_size <- max(eval_calls$expr_expression_length, na.rm = TRUE)
average_size <- weighted.mean(eval_calls$expr_expression_length, na.rm = TRUE, w = eval_calls$nb_ev_calls)

sizes_prop <- eval_calls %>% count(expr_expression_length, wt = nb_ev_calls) %>%
                      mutate(proportion = n / sum(n)) %>% 
                      arrange(expr_expression_length) %>% 
                      mutate(cumulative = cumsum(proportion))

nb_nodes_many <- sizes_prop %>% filter(cumulative >= 0.95) %>% .[1,]

r("median size eval", med_size)
r("max size eval", max_size)
r("average size eval", average_size)
r("size nine five", nb_nodes_many$expr_expression_length)
r("size nine five exact percent", percent(nb_nodes_many$cumulative))

```

# Amount of computations via eval

```{r amount_computations}
nb_events_in_eval <- eval_calls %>% summarise(nb_events = sum(direct_interpreter_eval))

max_events_in_eval <- eval_calls %>% summarize(maxi = max(direct_interpreter_eval))

less_100_events <- eval_calls %>% filter(direct_interpreter_eval <= 100)
nb_less_100_events <- count(less_100_events, wt = nb_ev_calls)$n

r("nb events eval", nb_events_in_eval$nb_events)
r("maxi events eval", max_events_in_eval$maxi)
r("small events eval", ratio(nb_less_100_events, nb_eval_calls))
```

```{r nb-events-violin-plot}
eval_calls %>% filter(direct_interpreter_eval > 0) %>%
 ggplot(aes(x = "", y = direct_interpreter_eval, fill = ""), na.rm = TRUE) +
  geom_violin(trim = FALSE, width=1.7)+
  scale_y_log10(labels = scales::comma) +
  labs(y="Number of events in eval per package") +
  theme(
    legend.position="none"
  )

ggsave(path(PLOT_DIR, "nb_events_per_packages.pdf"))
```

# The case of core libraries

```{r}
duration <- difftime(Sys.time(), now)
```

Notebook execution was `r duration` `r units(duration)`.

### 2020-12-10 inst/analysis/usage_metrics.Rmd
various
refactor

--- 2020-12-09 inst/analysis/usage_metrics.Rmd
@@
 
-jan <- FALSE
+jan <- TRUE
 
@@
   function(before, options) {
-    if (before) {.       # record the current time before each chunk
+    if (before) { # record the current time before each chunk
       now <<- Sys.time()
@@
   cache.lazy = FALSE,
-  time_it = TRUE)
+  time_it = TRUE
+)
 
-if (jan == FALSE) { source("inc/paths.R") }
+if (jan == FALSE) {
+  source("inc/paths.R")
+}
 source("inc/latextags.R")
@@
 theme_set(theme_minimal())
-if (jan) { TAGS_DIR <- PLOT_DIR <- "." }
+if (jan) {
+  TAGS_DIR <- PLOT_DIR <- "."
+}
 create_tags(path(TAGS_DIR, "usage_metrics.tex"), prefix = "", default = TRUE)
@@
 # Set the paths, and customize for Jan.
-if(jan) { base_dir <- "/mnt/code/eval" }
+if (jan) {
+  base_dir <- "/mnt/code/eval"
+}
 eval_calls_path <- paste0(base_dir, params$calls_packages_path)
@@
 ```{r}
-E_raw %>% select(package, eval_function, nb_ev_calls,
-                 eval_call_srcref, file, eval_source, 
-                 expr_expression_length,
-                 direct_interpreter_eval)-> E
+E_raw %>% select(
+  run_package = package,        # The package which was run to trigger the eval
+  ev_variant = eval_function,   # Which eval variant was called
+  duplicates = nb_ev_calls,     # Number of identical eval calls (weight)
+  src_ref = eval_call_srcref,   # Source ref for the eval call site
+  file,                         # Script file name (name of this run)
+  ev_package =eval_source,      # Package of the call site
+  arg_size = expr_expression_length, # Size in bytes of eval argument string
+  opcodes = direct_interpreter_eval  # Non-recursive AST nodes traversed
+) -> E                               # by the interpreter
+
 C_raw -> C
@@
 ```{r eval-calls-overview}
-nb_calls <- (E %>% count(wt=nb_ev_calls))$n
-eval_calls <- E %>% filter(eval_function=="eval") %>% count(wt=nb_ev_calls)
-evalq_calls <- E %>% filter(eval_function=="evalq") %>% count(wt=nb_ev_calls)
-eval.parent_calls <- E %>% filter(eval_function=="eval.parent") %>%
-  count(wt = nb_ev_calls)
-local_calls <- E %>% filter(eval_function=="local") %>% count(wt = nb_ev_calls)
-eval_at_least_once <- E %>% count(package, wt=nb_ev_calls)
+nb_calls <- (E %>% count(wt = duplicates))$n
+eval_calls <- E %>%
+  filter(ev_variant == "eval") %>%
+  count(wt = duplicates)
+evalq_calls <- E %>%
+  filter(ev_variant == "evalq") %>%
+  count(wt = duplicates)
+eval.parent_calls <- E %>%
+  filter(ev_variant == "eval.parent") %>%
+  count(wt = duplicates)
+local_calls <- E %>%
+  filter(ev_variant == "local") %>%
+  count(wt = duplicates)
 
@@
 r("nb local calls", local_calls$n)
-r("nb package eval at least once", nrow(eval_at_least_once))
 ```
+
+There so mane calls: `r nb_calls`
 
@@
 nb_ev_call_sites <- E %>%
-  select(eval_call_srcref) %>%
+  select(src_ref) %>%
   n_distinct(na.rm = TRUE)
@@
 call_sites_per_package <- E %>%
-  group_by(eval_source) %>%
-  summarize(nb_call_sites = n_distinct(eval_call_srcref, na.rm = TRUE))
+  group_by(ev_package) %>%
+  summarize(nb_call_sites = n_distinct(src_ref, na.rm = TRUE))
 
@@
   filter(nb_call_sites == max(nb_call_sites, na.rm = TRUE)) %>%
-  select(eval_source, nb_call_sites)
+  select(ev_package, nb_call_sites)
 nb_no_calls_to_eval <- corpus_size -
-  n_distinct(select(packages_at_least_one_call_site, eval_source))
+  n_distinct(select(packages_at_least_one_call_site, ev_package))
 
+# sites triggered over the whole corpus
 r("nb eval call sites", nb_ev_call_sites) 
+# How many packages in the corpus have a single triggered call site
 r("nb one call site", nb_packages_one_call_site)
+r("at least one call site", nrow(packages_at_least_one_call_site))
 r("packages no calls to eval", nb_no_calls_to_eval)
@@
 r("max call sites", max_packages$nb_call_sites)
-r("max call sites package", max_packages$eval_source)
-r("at least one call site", nrow(packages_at_least_one_call_site))
+r("max call sites package", max_packages$ev_package)
 ```
@@
 E %>%
-  group_by(eval_source) %>%
-  mutate(nb_call_sites = n_distinct(eval_call_srcref, na.rm = TRUE)) %>%
-  ggplot(aes(x = eval_function, y = nb_call_sites, fill = eval_function), na.rm = TRUE) +
+  group_by(ev_package) %>%
+  mutate(nb_call_sites = n_distinct(src_ref, na.rm = TRUE)) %>%
+  ggplot(aes(x = ev_variant, y = nb_call_sites, fill = ev_variant), na.rm = TRUE) +
   geom_violin(trim = FALSE, width = 1.7) +
@@
 runs_package <- E %>%
-  group_by(eval_source) %>%
+  group_by(ev_package) %>%
   summarise(nb_runs = n_distinct(file)) %>%
@@
 eval_calls_per_package <- E %>%
-  count(eval_source, wt = nb_ev_calls) %>%
-  left_join(runs_package, by = "eval_source") %>%
+  count(ev_package, wt = duplicates) %>%
+  left_join(runs_package, by = "ev_package") %>%
   mutate(n = n / nb_runs)
@@
 r("max eval calls", max_eval_calls$n)
-r("max eval call package", max_eval_calls$eval_source)
+r("max eval call package", max_eval_calls$ev_package)
 ```
@@
 
-med_size <- weighted.median(E$expr_expression_length, w = E$nb_ev_calls, na.rm = TRUE)
-max_size <- max(E$expr_expression_length, na.rm = TRUE)
-average_size <- weighted.mean(E$expr_expression_length, na.rm = TRUE, w = E$nb_ev_calls)
+med_size <- weighted.median(E$arg_size, w = E$duplicates, na.rm = TRUE)
+max_size <- max(E$arg_size, na.rm = TRUE)
+average_size <- weighted.mean(E$arg_size, na.rm = TRUE, w = E$duplicates)
 
 sizes_prop <- E %>%
-  count(expr_expression_length, wt = nb_ev_calls) %>%
+  count(arg_size, wt = duplicates) %>%
   mutate(proportion = n / sum(n)) %>%
-  arrange(expr_expression_length) %>%
+  arrange(arg_size) %>%
   mutate(cumulative = cumsum(proportion))
@@
 r("average size eval", average_size)
-r("size nine five", nb_nodes_many$expr_expression_length)
+r("size nine five", nb_nodes_many$arg_size)
 r("size nine five exact percent", percent(nb_nodes_many$cumulative))
@@
 ```{r amount_computations}
-nb_events_in_eval <- E %>% summarise(nb_events = sum(direct_interpreter_eval))
+nb_events_in_eval <- E %>% summarise(nb_events = sum(opcodes))
 
-max_events_in_eval <- E %>% summarize(maxi = max(direct_interpreter_eval))
+max_events_in_eval <- E %>% summarize(maxi = max(opcodes))
 
-less_100_events <- E %>% filter(direct_interpreter_eval <= 100)
-nb_less_100_events <- count(less_100_events, wt = nb_ev_calls)$n
+less_100_events <- E %>% filter(opcodes <= 100)
+nb_less_100_events <- count(less_100_events, wt = duplicates)$n
 
@@
 E %>%
-  filter(direct_interpreter_eval > 0) %>%
-  ggplot(aes(x = "", y = direct_interpreter_eval, fill = ""), na.rm = TRUE) +
+  filter(opcodes > 0) %>%
+  ggplot(aes(x = "", y = opcodes, fill = ""), na.rm = TRUE) +
   geom_violin(trim = FALSE, width = 1.7) +
@@
 ```{r loops}
-eval_calls_exercised <- E %>% group_by(file) %>%
-  count(eval_call_srcref, wt = nb_ev_calls)
+eval_calls_exercised <- E %>%
+  group_by(file) %>%
+  count(src_ref, wt = duplicates)
 
 eval_calls_exercised %>% ggplot(aes(n)) +
-  geom_histogram() + scale_x_log10(labels = scales::comma) +
+  geom_histogram() +
+  scale_x_log10(labels = scales::comma) +
   labs(x = "Number of eval calls per srcref per run")

### 2020-12-11 inst/analysis/usage_metrics.Rmd
usage metrics

@@
 
-The following is a set up chunk. Currently there is an variable "jan" 
-used to run the script from docker on my machine. This should be
-removed. (TODO)
+The setup chunk has a variable `jan` with setting to run the script on
+docker on Jan's mamchine. This should be removed.
 
@@
       # return a character string to show the time
-      paste("Time for this code chunk to run:", res, units(res))
+      paste("Time for this code chunk to run:", round(res,2), units(res))
 }}}))
@@
 
-Read in the summarized calls data and the file that describes
-which package belong to our corpus.
+Read in the summarized calls data (`E_raw`) and the file that describes
+which package belong to our corpus (`C_raw`). For conciseness, `E` is
+the input data set with only the columns that we use.
 
-```{r read}
+```{r read,echo=F,cache=F}
 read_fst(eval_calls_path) %>% as_tibble() -> E_raw
 read_fst(corpus_path) %>% as_tibble() -> C_raw
-```
+#read_csv(paste0(base_dir, "/run/package-evals-traced.9/package-eval-static.csv")) -> P
 
-The two data sets are E and C for conciseness.  C is only used here to compute
-the corpus size. E is used throughout. Recall that E_raw has the source data
-with all columns.
-
-```{r}
 E_raw %>% select(
@@
 
-Compute the number of times eval and it variants were called in the entire 
-corpus. This is a dynamic count that aggregates all call sites within
-the corpus.
+# Eval Calls Overiew
+The corpus has `r corpus_size` packages, and the data set has 
+`r nrow(E)` distinct calls to the `eval` and its variants.
 
-```{r eval-calls-overview}
-nb_calls <- (E %>% count(wt = duplicates))$n
+```{r eval-calls-overview,echo=F}
+nb_calls <- E %>% 
+  count(wt = duplicates) %>%
+  .$n
 eval_calls <- E %>%
   filter(ev_variant == "eval") %>%
-  count(wt = duplicates)
+  count(wt = duplicates)  %>%
+  .$n
 evalq_calls <- E %>%
   filter(ev_variant == "evalq") %>%
-  count(wt = duplicates)
+  count(wt = duplicates) %>%
+  .$n
 eval.parent_calls <- E %>%
   filter(ev_variant == "eval.parent") %>%
-  count(wt = duplicates)
+  count(wt = duplicates)  %>%
+  .$n
 local_calls <- E %>%
   filter(ev_variant == "local") %>%
-  count(wt = duplicates)
+  count(wt = duplicates) %>%
+  .$n
 
-r("nb eval and variant calls", nb_calls)
-r("nb eval calls", eval_calls$n)
-r("nb evalq calls", evalq_calls$n)
-r("nb eval parent calls", eval.parent_calls$n)
-r("nb local calls", local_calls$n)
+TBL <- overview_table(
+r("nb eval and variant calls", nb_calls),
+r("nb eval calls", eval_calls),
+r("nb evalq calls", evalq_calls),
+r("nb eval parent calls", eval.parent_calls),
+r("nb local calls", local_calls)
+)
 ```
 
-There so mane calls: `r nb_calls`
+We recorded `r nb_calls` to eval, out of these, there were `r evalq_calls`
+(`r fmt(ratio(evalq_calls,nb_calls))`) calls to `evalq`, `r eval.parent_calls` 
+(`r fmt(ratio(evalq_calls,nb_calls))`) calls to `eval.parent`,  and
+`r local_calls` (`r fmt(ratio(evalq_calls,nb_calls))`) calls to `local`.
 
-# Number of eval call sites
+`r TBL`
 
-```{r nb-eval-call-sites}
-nb_ev_call_sites <- E %>%
-  select(src_ref) %>%
-  n_distinct(na.rm = TRUE)
+# Nb eval call sites
 
-call_sites_per_package <- E %>%
-  group_by(ev_package) %>%
-  summarize(nb_call_sites = n_distinct(src_ref, na.rm = TRUE))
+```{r nb-eval-call-sites, echo=F}
+stopifnot(sum(is.na(E$src_ref))==0)  # No NAs in scr_ref!
 
-nb_packages_one_call_site <- call_sites_per_package %>%
-  filter(nb_call_sites == 1) %>%
-  nrow()
-packages_at_least_one_call_site <- call_sites_per_package %>%
-  filter(nb_call_sites >= 1)
-median_call_sites <- packages_at_least_one_call_site %>%
-  summarise(med = median(nb_call_sites, na.rm = TRUE))
-max_packages <- packages_at_least_one_call_site %>%
-  filter(nb_call_sites == max(nb_call_sites, na.rm = TRUE)) %>%
-  select(ev_package, nb_call_sites)
-nb_no_calls_to_eval <- corpus_size -
-  n_distinct(select(packages_at_least_one_call_site, ev_package))
+sites <- E %>%             # src_ref indicates where the call originates from.
+  pull(src_ref) %>% 
+  n_distinct()             # This the number of distinct sites
 
-# sites triggered over the whole corpus
-r("nb eval call sites", nb_ev_call_sites) 
-# How many packages in the corpus have a single triggered call site
-r("nb one call site", nb_packages_one_call_site)
-r("at least one call site", nrow(packages_at_least_one_call_site))
-r("packages no calls to eval", nb_no_calls_to_eval)
-r("nb median call sites", median_call_sites$med)
-r("max call sites", max_packages$nb_call_sites)
-r("max call sites package", max_packages$ev_package)
+sites_per_package <- E %>%
+  count(ev_package, src_ref) %>%
+  count(ev_package)        # All packages have at least one site!
+
+max_packages <- sites_per_package %>%
+  filter(n == max(n)) %>%
+  .$n
+
+TBL <- overview_table(
+# Sites triggered over the whole corpus:
+r("nb eval call sites", sites) ,
+# How many packages in the corpus have a single triggered call site:
+r("nb one call site", sites_per_package %>% filter(n==1)),  
+r("at least one call site", sites_per_package),
+r("packages no calls to eval", corpus_size - nrow(sites_per_package)),
+r("nb median call sites",  median(sites_per_package$n)),
+r("max call sites", max_packages),
+)
 ```
 
-```{r call-sites-violin-plot}
+`r TBL`
+
+```{r call-sites-violin-plot, echo=F}
 E %>%
   group_by(ev_package) %>%
-  mutate(nb_call_sites = n_distinct(src_ref, na.rm = TRUE)) %>%
-  ggplot(aes(x = ev_variant, y = nb_call_sites, fill = ev_variant), na.rm = TRUE) +
-  geom_violin(trim = FALSE, width = 1.7) +
+  mutate(sites=n_distinct(src_ref)) %>%
+  ggplot(aes(x=ev_variant, y=sites, fill=ev_variant), na.rm=TRUE) +
+  geom_violin(trim=FALSE, width=1.7) +
   # geom_jitter(shape=16, position=position_jitter(0.2)) +
-  labs(y = "Call sites per package") +
+  labs(y="Call sites per package") +
   theme(
-    legend.position = c(0.5, 0.85),
-    legend.box.background = element_rect(fill = "white", size = 0.1),
-    legend.box = "horizontal",
-    axis.title.x = element_blank()
+    legend.position=c(0.5, 0.85),
+    legend.box.background=element_rect(fill="white", size=0.1),
+    legend.box="horizontal",
+    axis.title.x=element_blank()
   )
@@
 ```
-
-<!-- How is the normalized distribution of eval calls?  -->
-
-<!-- ```{r distribution-call-sites} -->
-<!-- eval_calls_per_package %>% -->
-<!--   ggplot(aes(x = n), na.rm = TRUE) + -->
-<!--   geom_density(adjust = 10) + -->
-<!--   scale_y_log10()  -->
-<!-- ``` -->
-
-<!-- (Like the violin plot, but smoothed) -->
-(th)
 

### 2020-12-11 inst/analysis/usage_metrics.Rmd
Adding ratio of call sites triggered over the total number of them in the corpus

--- 2020-12-10 inst/analysis/usage_metrics.Rmd
@@
 params:
-  base_dir: /var/lib/R/project-evalR
-  calls_packages_path: /run/package-evals-traced.9/summarized-packages.fst
-  corpus_path: /run/package-evals-traced.9/corpus.fst
+  base_dir: /var/lib/R/project-evalR/run/package-evals-traced.10/
+  calls_packages_path: summarized-packages.fst
+  corpus_path: corpus.fst
+  static_evals_path: package-evals-static.csv
 ---
@@
 
-jan <- TRUE
+jan <- FALSE
 
@@
 corpus_path <- paste0(base_dir, params$corpus_path)
+eval_static_path <- paste0(base_dir, params$static_evals_path)
 ```
@@
 read_fst(corpus_path) %>% as_tibble() -> C_raw
+
+read_csv(eval_static_path) %>% semi_join(C_raw, by = "package") -> E_static
 ```
@@
 the corpus size. E is used throughout. Recall that E_raw has the source data
-with all columns.
+with all columns. E_static corresponds to static information about the call sites.
 
@@
 ggsave(path(PLOT_DIR, "call_sites_per_packages.pdf"))
+```
+
+How many call sites and how many packages did we exercise out of the total number of call sites of the corpus?
+
+```{r exercised-call-sites}
+nb_static_call_sites <- nrow(E_static)
+
+static_call_sites_per_package <- E_static %>%
+  count(package, sort = TRUE)
+
+static_at_least_one_call_site <- static_call_sites_per_package %>%
+  filter(n >= 1)
+
+r("nb static call sites", nb_static_call_sites)
+r("call sites percent", ratio(nb_ev_call_sites, nb_static_call_sites))
+r("nb static at least one call site", nrow(static_at_least_one_call_site))
+r("static at least one call site percent", ratio(nrow(packages_at_least_one_call_site), nrow(static_at_least_one_call_site)))
 ```

### 2020-12-11 inst/analysis/usage_metrics.Rmd

--- 2020-12-11 inst/analysis/usage_metrics.Rmd
@@
 The setup chunk has a variable `jan` with setting to run the script on
-docker on Jan's mamchine. This should be removed.
+docker on Jan's machine. This should be removed.
 
@@
       # return a character string to show the time
-      paste("Time for this code chunk to run:", round(res,2), units(res))
-}}}))
+      paste("Time for this code chunk to run:", round(res, 2), units(res))
+    }
+  }
+}))
 now <- Sys.time()
@@
 read_fst(corpus_path) %>% as_tibble() -> C_raw
-#read_csv(paste0(base_dir, "/run/package-evals-traced.9/package-eval-static.csv")) -> P
 
+read_csv(eval_static_path) %>% semi_join(C_raw, by = "package") -> P
+```
+
+The two data sets are E and C for conciseness.  C is only used here to compute
+the corpus size. E is used throughout. Recall that E_raw has the source data
+with all columns. P corresponds to static information about the call sites.
+
+```{r select-columns,echo=F}
 E_raw %>% select(
@@
   file, # Script file name (name of this run)
-  ev_package =eval_source,      # Package of the call site
+  ev_package = eval_source, # Package of the call site
   arg_size = expr_expression_length, # Size in bytes of eval argument string
@@
 C_raw -> C
-corpus_size <- nrow(C)
+corpus_size <- nrow(P)
 ```
@@
 ```{r nb-eval-call-sites, echo=F}
-stopifnot(sum(is.na(E$src_ref))==0)  # No NAs in scr_ref!
+stopifnot(sum(is.na(E$src_ref)) == 0) # No NAs in scr_ref!
 
@@
   # Sites triggered over the whole corpus:
-r("nb eval call sites", sites) ,
+  r("nb eval call sites", sites),
   # How many packages in the corpus have a single triggered call site:
-r("nb one call site", sites_per_package %>% filter(n==1)),  
+  r("nb one call site", sites_per_package %>% filter(n == 1)),
   r("at least one call site", sites_per_package),
@@
   group_by(ev_package) %>%
-  mutate(sites=n_distinct(src_ref)) %>%
-  ggplot(aes(x=ev_variant, y=sites, fill=ev_variant), na.rm=TRUE) +
-  geom_violin(trim=FALSE, width=1.7) +
+  mutate(sites = n_distinct(src_ref)) %>%
+  ggplot(aes(x = ev_variant, y = sites, fill = ev_variant), na.rm = TRUE) +
+  geom_violin(trim = FALSE, width = 1.7) +
   # geom_jitter(shape=16, position=position_jitter(0.2)) +
-  labs(y="Call sites per package") +
+  labs(y = "Call sites per package") +
   theme(
-    legend.position=c(0.5, 0.85),
-    legend.box.background=element_rect(fill="white", size=0.1),
-    legend.box="horizontal",
-    axis.title.x=element_blank()
+    legend.position = c(0.5, 0.85),
+    legend.box.background = element_rect(fill = "white", size = 0.1),
+    legend.box = "horizontal",
+    axis.title.x = element_blank()
   )
@@
 ggsave(path(PLOT_DIR, "call_sites_per_packages.pdf"))
+```
+
+How many call sites and how many packages did we exercise out of the total number of call sites of the corpus?
+
+```{r exercised-call-sites}
+static_sites <- nrow(P)
+
+static_call_sites_per_package <- P %>%
+  count(package, sort = TRUE)
+
+static_at_least_one_call_site <- static_call_sites_per_package %>%
+  filter(n >= 1)
+
+r("nb static call sites", static_sites)
+r("call sites percent", ratio(sites, static_sites))
+r("nb static at least one call site", nrow(static_at_least_one_call_site))
+r("static at least one call site percent", ratio(nrow(sites_per_package %>% filter(n >= 1)), nrow(static_at_least_one_call_site)))
 ```

### 2020-12-11 inst/analysis/usage_metrics.Rmd
minor

--- 2020-12-11 inst/analysis/usage_metrics.Rmd
@@
 E %>% 
-  group_by(ev_package) %>%
-  mutate(sites=n_distinct(src_ref)) %>%
+  select(ev_package, ev_variant, src_ref) %>%
+  group_by(ev_package, ev_variant) %>%
+  summarize(sites=n_distinct(src_ref)) %>%
   ggplot(aes(x=ev_variant, y=sites, fill=ev_variant), na.rm=TRUE) +
@@
     axis.title.x=element_blank()
-  )
+  ) -> PLOT
 
@@
 ```
+
+`r PLOT`
 

### 2020-12-14 inst/analysis/usage_metrics.Rmd
minor

--- 2020-12-11 inst/analysis/usage_metrics.Rmd
@@
 if (jan) {
-  base_dir <- "/mnt/code/eval"
+  base_dir <- "/mnt/code/eval/run/package-evals-traced.9/"
 }
@@
 
+# Read
+
 Read in the summarized calls data (`E_raw`) and the file that describes
@@
 
-```{r read,echo=F,cache=F}
+```{r read,include=F}
 read_fst(eval_calls_path) %>% as_tibble() -> E_raw
 read_fst(corpus_path) %>% as_tibble() -> C_raw
-
 read_csv(eval_static_path) %>% semi_join(C_raw, by = "package") -> P
@@
 
-The two data sets are E and C for conciseness.  C is only used here to compute
-the corpus size. E is used throughout. Recall that E_raw has the source data
-with all columns. P corresponds to static information about the call sites.
+The two data sets are `E` and `C` for conciseness.  `C` is only used here to 
+compute the corpus size. `E` is used throughout. Recall that `E_raw` has the 
+source data with all columns. `P` corresponds to static information about the
+call sites.
 
@@
   arg_size = expr_expression_length, # Size in bytes of eval argument string
-  opcodes = direct_interpreter_eval # Non-recursive AST nodes traversed
-) -> E # by the interpreter
+  opcodes = direct_interpreter_eval #  Count of the AST nodes traversed by the
+) -> E #                               interpreter during eval (non-recursive).
 
@@
 
-# Eval Calls Overiew
-The corpus has `r corpus_size` packages, and the data set has 
-`r nrow(E)` distinct calls to the `eval` and its variants.
+#  Variants
 
-```{r eval-calls-overview,echo=F}
+The corpus has `r corpus_size` packages, and the data set has `r nrow(E)` 
+distinct calls to the `eval` and its variants. There are 
+`r length(unique(E$ev_package))` corpus packages with `eval` calls (note that
+we may not trigger all `eval`s in a package due to spotty covereage).
+
+```{r variants,echo=F}
 nb_calls <- E %>%
@@
 
-# Nb eval call sites
+# Triggered sites
 
-```{r nb-eval-call-sites, echo=F}
+```{r triggered-sites, include=F}
 stopifnot(sum(is.na(E$src_ref)) == 0) # No NAs in scr_ref!
@@
 
-```{r call-sites-violin-plot, echo=F}
+```{r call-sites-violin-plot, include=F}
 E %>%
@@
     axis.title.x = element_blank()
-  )
+  ) -> PLOT
 
@@
 
-```{r exercised-call-sites}
+```{r exercised-call-sites, include=F}
 static_sites <- nrow(P)
@@
 
-```{r amout-of-code}
+```{r amout-of-code, include=F}
 
@@
 
-```{r amount_computations}
+```{r amount_computations, include=F}
 nb_events_in_eval <- E %>% summarise(nb_events = sum(opcodes))
@@
 
-```{r nb-events-violin-plot}
+```{r nb-events-violin-plot,include=F}
 E %>%

### 2020-12-15 analysis/corpus-stage3.Rmd
changes!

--- 2020-08-27 inst/analysis/kaggle-stage2.Rmd
@@
 ---
-title: "Kaggle stage2"
+title: "Corpus stage 3"
 output: html_document
@@
 
-```{r loading raw data}
-kaggle_kernels <- as_tibble(read_fst(KAGGLE_KERNELS_FILE))
-kaggle_log <- as_tibble(read_fst(KAGGLE_LOG_FILE)) %>%
-  transmute(
-    id=str_replace(Command, ".*/kaggle-run/[^/]+/([^/]+)/.*", "\\1"),
-    exitval=Exitval,
-    runtime=as.duration(JobRuntime)
-  )
+```{r}
+calls       <- read_fst(CALLS_FILE) %>% mutate(package=basename(dirname(dirname(file))))
+corpus      <- read_fst(CORPUS_DETAILS_FILE)
+calls_final <- semi_join(calls, corpus, by="package")
 ```
 
-### Programs
-
-```{r}
-kaggle_programs <- left_join(
-  kaggle_kernels,
-  kaggle_log,
-  by="id"
-) %>%
-  mutate(
-    file=path("../../run/kaggle-run", competition, id, "calls.fst"),
-    corpus="kaggle"
-  ) %>%
-  select(
-    file,
-    id,
-    language,
-    kernel_type,
-    competition,
-    runtime,
-    code,
-    exitval,
-    corpus
-  )
-
-kaggle_programs_errors <- 
-  kaggle_programs %>%
-  mutate(
-    error_file=path(dirname(str_replace(file, fixed("../.."), params$base_dir)), "calls-ERROR"), 
-    error=file.exists(error_file)
-  ) %>%
-  filter(error) %>%
-  rowwise() %>%
-  do(mutate(read_fst(.$error_file), file=.$file))
-
-kaggle_programs <-
-    left_join(
-      kaggle_programs, 
-      if (nrow(kaggle_programs_errors) > 0) {
-        transmute(kaggle_programs_errors, error=paste(source, message, call), file)
-      } else {
-        transmute(kaggle_programs, error=NA, file)
-      }, 
-      by="file"
-    ) %>%
-    mutate(
-      success=
-        exitval==0 &
-        !is.na(code) &
-        !is.na(runtime) &
-        code > 0 &
-        is.na(error)
-    )
+```{r export data}
+write_fst(calls_final, "../../run/package-evals-traced/calls-final.fst")
 ```
-
-```{r}
-write_fst(kaggle_programs, KAGGLE_PROGRAMS_FILE)
-```

### 2020-12-15 analysis/environments.Rmd
changes!

---
title: "Environments and eval in R"
author: "Pierre Donat-Bouillud"
output:
  html_document:
        gallery: false
        toc: true
        toc_depth: 3
        toc_float: true
        df_print: paged
date: "`r Sys.Date()`"
editor_options: 
  chunk_output_type: console
params:
  base_dir: /var/lib/R/project-evalR
  calls_packages_path: /run/package-evals-traced.10/summarized-packages.fst
---

```{r setup, include=FALSE}
knitr::knit_hooks$set(time_it = local({
  now <- NULL
  function(before, options) {
    if (before) { # record the current time before each chunk
      now <<- Sys.time()
    } else { # calculate the time difference after a chunk
      res <- difftime(Sys.time(), now)
      # return a character string to show the time
      paste("Time for this code chunk to run:", res, units(res))
}}}))

knitr::opts_chunk$set(echo = TRUE,
                      fig.align = "center",
                      fig.retina = 2,
                      fig.width = 10,
                      cache.lazy = FALSE)

now <- Sys.time()

library(tidyverse)
library(fst)
library(fs)
library(DT)

source("inc/latextags.R")
source("inc/setup.R")
source("inc/paths.R")
theme_set(theme_minimal())
create_tags(path(TAGS_DIR, "environments.tex"), prefix = "", default = TRUE)

source("insights.R", local = knitr::knit_global())
```

# Loading datafiles

```{r loading_fst}
eval_calls_path <- paste0(params$base_dir, params$calls_packages_path)

eval_calls <- read_fst(eval_calls_path) %>%
  as_tibble() 
```

# Description of the data files

```{r packages}
nb_eval_calls <- count(eval_calls, wt = nb_ev_calls)

nb_call_sites <- nb_eval_call_sites(eval_calls)
```

There are `r count(eval_calls, wt=nb_ev_calls)$n` eval calls and `r eval_calls %>% select(eval_call_srcref) %>% n_distinct()` eval call sites in the dataset.

# Classification of environments

We distinguish between 5 main categories of `eval` calls with respect to their `envir` argument:

- `envir` is not provided explicitly (and thus is the default argument, which semantically belongs to the absolute category)
- `envir` in an eval call generated by Rcpp
- `envir` is a list or a data frame
- `envir` comes from an argument of the caller function
- `envir` is absolutely computed (with `parent.frame()` or equivalent)
- `envir` is an integer. Then `sys.call` will be used and it is similar to the absolute category
- `envir` is `NULL` and then `enclos` is directly used.

## `envir` as default argument

```{r envir_default}
eval_calls_default <- eval_calls %>% filter(is.na(envir_expression))

nb_default <- count(eval_calls_default, wt = nb_ev_calls)
nb_default_sites <- nb_eval_call_sites(eval_calls_default)

eval_calls_explicit <- eval_calls %>% filter(!is.na(envir_expression))
nb_explicit <- count(eval_calls_explicit, wt = nb_ev_calls)
nb_explicit_sites <- nb_eval_call_sites(eval_calls_explicit)
```

There are `r nb_default$n` default calls, which corresponds to `r nb_default_sites` call sites. On the contrary, `r nb_explicit$n` eval calls explicit provide an `envir` argument, which is `r nb_explicit_sites`. The default eval calls are `r ratio(nb_default$n, nb_eval_calls$n)`% of the total and `r ratio(nb_default_sites, nb_call_sites)`% in terms of call sites.

# `envir` with Rcpp

```{r envir_rcpp}
eval_calls_rcpp <- eval_calls_explicit %>% filter(str_detect(expr_expression, "C\\+\\+Constructor|C\\+\\+Field|C\\+\\+OverloadedMethods"))

nb_rcpp <- count(eval_calls_rcpp, wt = nb_ev_calls)
nb_rcpp_sites <- nb_eval_call_sites(eval_calls_rcpp)

eval_calls_explicit <- eval_calls_explicit %>% anti_join(eval_calls_rcpp)
```

##  `envir` as a list or data frame

```{r envir_list}
eval_calls_env_list <- eval_calls_explicit %>% filter(envir_type == "list")

nb_env_list <- count(eval_calls_env_list, wt = nb_ev_calls)
nb_env_list_sites <- nb_eval_call_sites(eval_calls_env_list)
```

## `envir` as an argument from the caller function

```{r envir_arg_caller}
eval_calls_arg_caller <- eval_calls_explicit %>% filter(!is.na(envir_from_arg))

nb_arg_caller <- count(eval_calls_arg_caller, wt = nb_ev_calls)
nb_arg_caller_sites <- nb_eval_call_sites(eval_calls_arg_caller)
```

Most of the eval calls directly use the argument, but some seem to modify it a lot:

```{r}
eval_calls_arg_caller %>% count(envir_from_arg, wt=nb_ev_calls, sort = TRUE) %>% knitr::kable()
```

In the current dataset, there are no examples of sandbox usage but we should see a distance of 1 in that case.

## `envir` absolutely computed

```{r envir_absolute}
eval_calls_absolute <- eval_calls_explicit %>% anti_join(eval_calls_env_list) %>% anti_join(eval_calls_arg_caller)

nb_absolute <- count(eval_calls_absolute, wt = nb_ev_calls)
nb_absolute_sites <- nb_eval_call_sites(eval_calls_absolute)
```

Let's do a sanity check: is the environment class the same in that category?

```{r}
eval_calls_absolute <- eval_calls_absolute %>% group_by(eval_call_srcref) %>%  mutate(envir_distinct = n_distinct(environment_class))

eval_calls_absolute %>% ungroup() %>% count(envir_distinct, sort = TRUE)
```

We can also see a few examples:

```{r}
eval_calls_absolute %>% filter(envir_distinct == 2) %>% select(eval_call_expression, environment_class, envir_distinct)
```

# `envir` is an integer

```{r envir_integer}
eval_calls_integer <- eval_calls_explicit %>% filter(envir_type == "integer")

nb_integer <- count(eval_calls_integer, wt = nb_ev_calls)
nb_integer_sites <- nb_eval_call_sites(eval_calls_integer)
```

# `envir` is `NULL`

```{r envir_null}
eval_calls_null <- eval_calls_explicit %>% filter(envir_type == "NULL")

nb_null <- count(eval_calls_null, wt = nb_ev_calls)
nb_null_sites <- nb_eval_call_sites(eval_calls_null)
```

## Summary of the categories

Total eval calls: `r nb_eval_calls`

Total eval call sites: `r nb_call_sites`

```{r summary}
summary <- tribble(
  ~category, ~calls, ~percent_calls, ~callsites, ~percent_callsites,
  "Default", nb_default$n, ratio(nb_default$n, nb_eval_calls$n), nb_default_sites, ratio(nb_default_sites, nb_call_sites),
  "Rcpp",  nb_rcpp$n, ratio(nb_rcpp$n, nb_eval_calls$n), nb_rcpp_sites, ratio(nb_rcpp_sites, nb_call_sites),
  "List or dataframe", nb_env_list$n, ratio(nb_env_list$n, nb_eval_calls$n), nb_env_list_sites, ratio(nb_env_list_sites, nb_call_sites),
  "Argument of caller", nb_arg_caller$n, ratio(nb_arg_caller$n, nb_eval_calls$n), nb_arg_caller_sites, ratio(nb_arg_caller_sites, nb_call_sites),
  "Absolute", nb_absolute$n, ratio(nb_absolute$n, nb_eval_calls$n), nb_absolute_sites, ratio(nb_absolute_sites, nb_call_sites),
  "Integer", nb_integer$n, ratio(nb_integer$n, nb_eval_calls$n), nb_integer_sites, ratio(nb_integer_sites, nb_call_sites),
  "Null", nb_null$n, ratio(nb_null$n, nb_eval_calls$n), nb_null_sites, ratio(nb_null_sites, nb_call_sites),
)

summary %>% datatable(colnames =  c("", "Eval calls", "%", "Call sites", "%"))
```

```{r}
duration <- difftime(Sys.time(), now)
```

Notebook execution was `r duration` `r units(duration)`.

### 2020-12-15 analysis/side-effects.Rmd
changes!

---
title: "Side effects"
output: html_document
params:
  base_dir: /home/aviral/projects/revalstudy2
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
for (x in c(
  "dplyr", 
  "DT",
  "fs",
  "fst",
  "ggplot2",
  "ggthemes",
  "lubridate",
  "readr",
  "purrr",
  "stringr",
  "tidyr"
  )) {
  suppressPackageStartupMessages(library(x, character.only=TRUE))
}

knitr::opts_chunk$set(
  echo = TRUE,
  fig.align = "center",
  fig.retina = 2,
  fig.width = 10
)

# TODO: move into the package
source("inc/paths.R")
source("inc/setup.R")
source("inc/latextags.R")

options(repos="https://cran.r-project.org")
theme_set(theme_minimal())
create_tags(path(TAGS_DIR, "side-effects.tex"), prefix="", default=TRUE)
```

```{r}
core_data <- read_fst(EVALS_SUM_CORE_FILE)
package_data <- read_fst(EVALS_SUM_PKGS_FILE)
kaggle_data <- read_fst(EVALS_SUM_KAGGLE_FILE)
program_data <- read_fst(PROGRAM_FILE)

all_data <- rbind(core_data, package_data, kaggle_data)
```

## Side-Effects and Eval

Each entry in `call.fst` contains `direct_writes` and `indirect_writes` (both
`0` for `eval.parent` and `local`). `direct_writes` are the side-effects
directly caused by the `eval` in question and `indirect_writes` are the
side-effects caused by all children evals of the eval. All application
assignments can be found in `direct_writes` field of `program.fst` and all
`eval_assignments` can be found in `indirect_writes` field of `program.fst`. All
non-local assignments can be found by simply summing up the `direct_writes`
field of all eval calls.

```{r preprocess-write-data}
eval_writes_data <-
    all_data %>%
    filter(eval_function == "eval" | eval_function == "evalq") %>%
    select(eval_source_type, eval_source, eval_call_srcref, direct_writes, indirect_writes, nb_ev_calls) %>%
    mutate(all_writes = direct_writes + indirect_writes, side_effecting = as.logical(all_writes))

program_writes_data <-
    program_data %>%
    select(direct_writes, indirect_writes)
```

```{r compute-side-effecting-write-counts}
all_writes <- sum(program_writes_data$direct_writes)
eval_writes <- sum(program_writes_data$indirect_writes)

r("All Writes", all_writes)
r("Eval Writes", eval_writes)

eval_side_effecting_writes <- sum(eval_writes_data$direct_writes * eval_writes_data$nb_ev_calls)

r("Eval Side Effecting Writes", eval_side_effecting_writes)

r("Eval Side Effecting Writes Eval Perc", ratio(eval_side_effecting_writes, eval_writes))
r("Eval Side Effecting Writes All Perc", ratio(eval_side_effecting_writes, all_writes))
```

```{r compute-side-effecting-call-counts}
eval_calls_by_source_writes <-
    eval_writes_data %>%
    count(eval_source_type, side_effecting, wt = nb_ev_calls, name = "call_count") %>%
    group_by(eval_source_type) %>%
    mutate(total_call_count = sum(call_count)) %>%
    ungroup() %>%
    mutate(call_perc = round(100 * call_count/total_call_count, 1)) %>%
    filter(side_effecting)

datatable(eval_calls_by_source_writes)

eval_calls_by_source_writes %>%
pmap(function(eval_source_type, side_effecting, call_count, total_call_count, call_perc) {
    r(str_c("Side Effecting ", eval_source_type, " call perc"), str_c(call_perc, "%"))
})

eval_calls_by_package_writes <-
    eval_writes_data %>%
    filter(eval_source_type != "core") %>%
    count(eval_source, side_effecting, wt = nb_ev_calls, name = "call_count") %>%
    arrange(desc(call_count))

datatable(eval_calls_by_package_writes)

eval_calls_by_source_write_count <-
    eval_writes_data %>%
    filter(all_writes != 0) %>%
    mutate(all_writes = if_else(all_writes <= 5, as.character(all_writes), "> 5")) %>%
    count(eval_source_type, all_writes, wt = nb_ev_calls, name = "call_count") %>%
    group_by(eval_source_type) %>%
    mutate(total_call_count = sum(call_count)) %>%
    ungroup() %>%
    mutate(call_perc = round(100 * call_count/total_call_count, 1))

datatable(eval_calls_by_source_write_count)

plot <-
    eval_calls_by_source_write_count %>%
    ggplot(aes(factor(all_writes, level = c(1:5, "> 5")), call_perc)) +
    geom_col(aes(fill = eval_source_type), position = "dodge") +
    labs(title="", x ="Side-Effects", y = "Side-Effecting Eval Calls %", fill = "") +
    theme(text = element_text(size=15), axis.text.x = element_text(size=15))
plot
```

The call site computation is incorrect.

```{r compute-side-effecting-site-counts, eval = FALSE}
eval_sites_by_source_writes <-
    eval_writes_data %>%
    group_by(eval_source_type, eval_call_srcref) %>%
    summarize(side_effecting = any(side_effecting)) %>%
    ungroup() %>%
    count(eval_source_type, side_effecting, name = "site_count") %>%
    group_by(eval_source_type) %>%
    mutate(total_site_count = sum(site_count)) %>%
    ungroup() %>%
    mutate(site_perc = round(100 * site_count / total_site_count, 1)) %>%
    filter(side_effecting)

datatable(eval_sites_by_source_writes)

eval_sites_by_source_writes %>%
pmap(function(eval_source_type, side_effecting, site_count, total_site_count, site_perc) {
    r(str_c("SideEffecting ", eval_source_type, " site perc"), str_c(site_perc, "%"))
})

eval_sites_by_package_writes <-
    eval_writes_data %>%
    filter(eval_source_type != "core") %>%
    group_by(eval_source, eval_call_srcref) %>%
    summarize(side_effecting = any(side_effecting)) %>%
    ungroup() %>%
    count(eval_source, side_effecting, name = "site_count") %>%
    group_by(eval_source, side_effecting) %>%
    mutate(site_perc = round(100 * site_count / sum(site_count), 1)) %>%
    ungroup() %>%
    arrange(side_effecting, desc(site_perc))

datatable(eval_sites_by_package_writes)
```

### 2020-12-15 analysis/inc/latextags.R

@@ -1,0 +1,404 @@
+## Overview
+
+overview <- function(...) {
+  bind_rows(...)
+}
+
+overview_table <- function(...) {
+  overview(...) %>%
+    my_datatable(page_size=Inf)
+}
+
+my_datatable <- function(df, page_size=20, round=TRUE, ...) {
+  options <- if (nrow(df) < page_size) {
+    list(paging=FALSE, searching=FALSE, info=FALSE)
+  } else {
+    list(paging=TRUE, searching=TRUE, info=TRUE, pageLength=page_size)
+  }
+  
+  table <- datatable(df, options=options, ...)
+  #if (round) {
+  #  table <- formatRound(table, sapply(df, is.numeric), 2)
+  #}
+  table
+}
+
+rs <- function(name, value) {
+  stopifnot(is.numeric(value))
+  bind_rows(
+    r(name, sum(value)),
+    r(paste(name, "mean"), mean(value)),
+    r(paste(name, "median"), median(value)),
+    r(paste(name, "min"), min(value)),
+    r(paste(name, "max"), max(value))
+  )
+}
+
+r <- function(name, value, ...) {
+  stopifnot(is.character(name))
+  stopifnot(length(name) == 1)
+  UseMethod("r", value)  
+}
+
+r.numeric <- function(name, value, ...) {
+  raw_str <- fmt(value, ...)
+  rnd_str <- fmt(oom(value), ...)
+  if (!is.na(value)) {
+    tag(str_c(name, " rnd"), rnd_str)
+  }
+  NextMethod("r", value)
+}
+
+r.Duration <- function(name, value, ...) {
+  units <- c("years", "days", "hours", "minutes", "seconds")
+  
+  for (u in units) {
+    v <- as.numeric(value, units=u)
+    class(v) <- "num_with_suffix"
+    attr(v, "suffix") <- paste0(" ", u)
+    
+    if (all(v > 1)) break
+  }
+
+  r(name, v)
+}
+
+r.difftime <- function(name, value, ...) {
+  v <- as.numeric(value)
+  class(v) <- "num_with_suffix"
+  attr(v, "suffix") <- paste0(" ", attr(value, "units"))
+
+  r(name, v)
+}
+
+
+r.default <- function(name, value, ...) {
+  stopifnot(length(value) == 1)
+  
+  value <- fmt(value, ...)
+  tag_name <- tag(name, value)
+  
[…]

### 2020-12-15 analysis/kaggle.R

@@ -1,0 +1,147 @@
+#!/usr/bin/env Rscript
+
+for (x in c(
+  "dplyr", 
+  "DT", 
+  "fs",
+  "fst",
+  "knitr",
+  "purrr",
+  "pbapply",
+  "readr",
+  "rjson",
+  "rmarkdown",
+  "runr",
+  "stringr",
+  "tibble"
+)) {
+  suppressPackageStartupMessages(library(x, character.only=TRUE))
+}
+
+pboptions(type="txt")
+
+# simulate from Rmd
+params <- list(
+  base_dir=commandArgs(trailingOnly=TRUE)[1],
+  competitions=c("titanic")
+)
+
+source("inc/paths.R")
+
+KAGGLE_KERNELS_DIR  <- path(RUN_DIR, "kaggle-kernels", "notebooks", "r", "kernels")
+KAGGLE_DATASETS_DIR <- path(RUN_DIR, "kaggle-datasets")
+KAGGLE_RUN_DIR      <- path(RUN_DIR, "kaggle-run")
+
+KERNELS_FILE      <- path(KAGGLE_RUN_DIR, "kernels.csv")
+SCRIPTS_FILE      <- path(KAGGLE_RUN_DIR, "scripts.txt")
+PARALLEL_LOG_FILE <- path(KAGGLE_RUN_DIR, "parallel.log")
+
+cat("- kaggle kernels:", KAGGLE_KERNELS_DIR, "\n")
+cat("- kaggle datasets:", KAGGLE_DATASETS_DIR, "\n")
+
+stopifnot(dir_exists(KAGGLE_KERNELS_DIR))
+stopifnot(dir_exists(path(KAGGLE_DATASETS_DIR, params$competitions)))
+
+wrap <- function(file, body) {
+  str_glue(
+    "evil::write_eval_traces(",
+    "  evil::trace_eval(",
+    "    quote = TRUE,",
+    "    code = {{",
+    "      {gsub('\n', '\n      ', body)}",
+    "    }}",
+    "  ),",
+    "  datadir=getwd()",
+    ")",
+    .sep = "\n"
+  )
+}
+
+extract_kernel_code <- function(path, code_file, competition) {
+  run_path <- path(KAGGLE_RUN_DIR, competition, basename(path))
+  code_file <- path(path, "script", code_file)
+  run_code_file <- path(run_path, "run.R")
+  
+  if (!dir_exists(run_path)) dir_create(run_path)
+  
[…]

### 2020-12-15 analysis/preprocess.R

@@ -1,0 +1,583 @@
+#!/usr/bin/env Rscript
+
+# Preprocess the dataset:
+# - remove duplicates
+# - correct src refs
+# - add some columns (such as `eval_call_package`)
+# - cut the dataset into smaller ones
+
+
+suppressPackageStartupMessages(library(tidyverse))
+suppressPackageStartupMessages(library(rlang))
+library(fs)
+library(fst)
+
+
+
+get_expr <- function(eval_call) {
+  if (is.na(eval_call)) {
+    return(NA_character_)
+  }
+  # Special case for expressions starting with _ (such as _inherit in ggproto)
+  escaped_eval_call <- if (startsWith(eval_call, "_")) {
+    paste0("`", eval_call, "`")
+  }
+  else {
+    eval_call
+  }
+
+  exp <- NA_character_
+  # Would fail for instance for
+  # "`$<-`(new(\"C++Field\", .xData = <environment>), \"read_only\", TRUE)" (classInt package)
+  try(exp <- parse(text = escaped_eval_call)[[1]], silent = TRUE)
+
+
+  return(exp)
+}
+
+
+deduplicate <- function(dataset) {
+  return(dataset %>% count(across(c(
+    -eval_call_id, -starts_with("caller_stack")
+  )), name = "nb_ev_calls"))
+}
+
+
+SEXP_TYPES <- tribble(
+  ~sexp_type, ~name,
+  0, "NILSXP",
+  1, "SYMSXP",
+  2, "LISTSXP",
+  3, "CLOSXP",
+  4, "ENVSXP",
+  5, "PROMSXP",
+  6, "LANGSXP",
+  7, "SPECIALSXP",
+  8, "BUILTINSXP",
+  9, "CHARSXP",
+  10, "LGLSXP",
+  13, "INTSXP",
+  14, "REALSXP",
+  15, "CPLXSXP",
+  16, "STRSXP",
+  17, "DOTSXP",
+  18, "ANYSXP",
+  19, "VECSXP",
+  20, "EXPRSXP",
+  21, "BCODESXP",
+  22, "EXTPTRSXP",
+  23, "WEAKREFSXP",
+  24, "RAWSXP",
+  25, "S4SXP",
+  30, "NEWSXP",
+  31, "FREESXP",
+  99, "FUNSXP"
+)
+
+
+SEXP_TYPES <- SEXP_TYPES %>% mutate(name = factor(name))
+
+resolve_sexp_name <- function(df, var) {
[…]

### 2020-12-15 inst/analysis/insights.R

@@ -1,332 +1,0 @@
-suppressPackageStartupMessages(library(rlang))
-
-
-
-
-nb_eval_call_sites <- function(eval_calls)
-{
-  # Two cases: with srcref, without srcref
-
-  nb_with_srcref <- eval_calls %>% select(eval_call_srcref) %>% n_distinct(na.rm = TRUE)
-  nb_without_srcref <- eval_calls %>% filter(is.na(eval_call_srcref)) %>% select(caller_package, caller_function) %>% n_distinct(na.rm = TRUE)
-
-  return(nb_with_srcref + nb_without_srcref)
-}
-
-
-get_expr <- function(eval_call)
-{
-  if(is.na(eval_call))
-  {
-    return(NA)
-  }
-  # Special case for expressions starting with _ (such as _inherit in ggproto)
-  escaped_eval_call <- if(startsWith(eval_call, "_"))
-  {
-    paste0("`", eval_call, "`")
-  }
-  else
-  {
-    eval_call
-  }
-
-  exp <- NA
-  # Would fail for instance for
-  # "`$<-`(new(\"C++Field\", .xData = <environment>), \"read_only\", TRUE)" (classInt package)
-  # exp <- tryCatch(
-  #   parse(text = escaped_eval_call)[[1]],
-  #   error = function(e) {return(NA)})
-  try(exp <- parse(text = escaped_eval_call)[[1]], silent = TRUE)
-
-
-  return(exp)
-}
-
-function_name <- function(eval_call)
-{
-  exp <- get_expr(eval_call)
-  if(is.call(exp))
-  {
-    return(paste(deparse(exp[[1]]), collapse = "\n"))
-  }
-  return(NA)
-}
-
-
-function_arguments <- function(eval_call)
-{
-  exp <- get_expr(eval_call)
-  if(is.call(exp))
-  {
-    return(map_chr(as.list(exp[-1]), function(chr) { paste(deparse(chr), collapse = "\n")}))
-  }
-  return(NA)
-}
-
-
-# See extract_inner_exp which takes a str
-# This one takes an expression
-# We assume that eval_expr_call is a call
-extract_inner_exp_aux <- function(eval_expr_call)
-{
-  args <- as.list(eval_expr_call[-1])
-  leaves <- list()
-  no_calls <- TRUE
-  for(arg in args)
-  {
-    if(!missing(arg)) # we cannot have the two conditions in the same if
-    {
-      if(is.call(arg))
-      {
[…]

### 2020-12-15 inst/analysis/preprocess.R

@@ -1,583 +1,0 @@
-#!/usr/bin/env Rscript
-
-# Preprocess the dataset:
-# - remove duplicates
-# - correct src refs
-# - add some columns (such as `eval_call_package`)
-# - cut the dataset into smaller ones
-
-
-suppressPackageStartupMessages(library(tidyverse))
-suppressPackageStartupMessages(library(rlang))
-library(fs)
-library(fst)
-
-
-
-get_expr <- function(eval_call) {
-  if (is.na(eval_call)) {
-    return(NA_character_)
-  }
-  # Special case for expressions starting with _ (such as _inherit in ggproto)
-  escaped_eval_call <- if (startsWith(eval_call, "_")) {
-    paste0("`", eval_call, "`")
-  }
-  else {
-    eval_call
-  }
-
-  exp <- NA_character_
-  # Would fail for instance for
-  # "`$<-`(new(\"C++Field\", .xData = <environment>), \"read_only\", TRUE)" (classInt package)
-  try(exp <- parse(text = escaped_eval_call)[[1]], silent = TRUE)
-
-
-  return(exp)
-}
-
-
-deduplicate <- function(dataset) {
-  return(dataset %>% count(across(c(
-    -eval_call_id, -starts_with("caller_stack")
-  )), name = "nb_ev_calls"))
-}
-
-
-SEXP_TYPES <- tribble(
-  ~sexp_type, ~name,
-  0, "NILSXP",
-  1, "SYMSXP",
-  2, "LISTSXP",
-  3, "CLOSXP",
-  4, "ENVSXP",
-  5, "PROMSXP",
-  6, "LANGSXP",
-  7, "SPECIALSXP",
-  8, "BUILTINSXP",
-  9, "CHARSXP",
-  10, "LGLSXP",
-  13, "INTSXP",
-  14, "REALSXP",
-  15, "CPLXSXP",
-  16, "STRSXP",
-  17, "DOTSXP",
-  18, "ANYSXP",
-  19, "VECSXP",
-  20, "EXPRSXP",
-  21, "BCODESXP",
-  22, "EXTPTRSXP",
-  23, "WEAKREFSXP",
-  24, "RAWSXP",
-  25, "S4SXP",
-  30, "NEWSXP",
-  31, "FREESXP",
-  99, "FUNSXP"
-)
-
-
-SEXP_TYPES <- SEXP_TYPES %>% mutate(name = factor(name))
-
-resolve_sexp_name <- function(df, var) {
[…]

### 2020-12-15 analysis/kaggle.R
more changes

@@ -1,147 +1,0 @@
-#!/usr/bin/env Rscript
-
-for (x in c(
-  "dplyr", 
-  "DT", 
-  "fs",
-  "fst",
-  "knitr",
-  "purrr",
-  "pbapply",
-  "readr",
-  "rjson",
-  "rmarkdown",
-  "runr",
-  "stringr",
-  "tibble"
-)) {
-  suppressPackageStartupMessages(library(x, character.only=TRUE))
-}
-
-pboptions(type="txt")
-
-# simulate from Rmd
-params <- list(
-  base_dir=commandArgs(trailingOnly=TRUE)[1],
-  competitions=c("titanic")
-)
-
-source("inc/paths.R")
-
-KAGGLE_KERNELS_DIR  <- path(RUN_DIR, "kaggle-kernels", "notebooks", "r", "kernels")
-KAGGLE_DATASETS_DIR <- path(RUN_DIR, "kaggle-datasets")
-KAGGLE_RUN_DIR      <- path(RUN_DIR, "kaggle-run")
-
-KERNELS_FILE      <- path(KAGGLE_RUN_DIR, "kernels.csv")
-SCRIPTS_FILE      <- path(KAGGLE_RUN_DIR, "scripts.txt")
-PARALLEL_LOG_FILE <- path(KAGGLE_RUN_DIR, "parallel.log")
-
-cat("- kaggle kernels:", KAGGLE_KERNELS_DIR, "\n")
-cat("- kaggle datasets:", KAGGLE_DATASETS_DIR, "\n")
-
-stopifnot(dir_exists(KAGGLE_KERNELS_DIR))
-stopifnot(dir_exists(path(KAGGLE_DATASETS_DIR, params$competitions)))
-
-wrap <- function(file, body) {
-  str_glue(
-    "evil::write_eval_traces(",
-    "  evil::trace_eval(",
-    "    quote = TRUE,",
-    "    code = {{",
-    "      {gsub('\n', '\n      ', body)}",
-    "    }}",
-    "  ),",
-    "  datadir=getwd()",
-    ")",
-    .sep = "\n"
-  )
-}
-
-extract_kernel_code <- function(path, code_file, competition) {
-  run_path <- path(KAGGLE_RUN_DIR, competition, basename(path))
-  code_file <- path(path, "script", code_file)
-  run_code_file <- path(run_path, "run.R")
-  
-  if (!dir_exists(run_path)) dir_create(run_path)
-  
[…]

### 2020-12-15 analysis/Attic/counts.Rmd
changes

--- 2020-08-28 inst/analysis/counts.Rmd
@@
 ```{r callsite-distribution, eval=TRUE}
+package_evals_dynamic <- read_fst(PACKAGE_EVALS_DYNAMIC_FILE) %>% as_tibble()
 eval_callsite_table <-
-  package_calls %>%
+  package_evals_dynamic %>%
   mutate(
     site_count=cut(
-      site_count, 
+      n, 
       breaks=c(0:10, 50, 100, 150, 200, 250, 300, Inf), 

### 2020-12-15 inst/analysis/preprocess.R
Environments

@@ -211,6 +211,16 @@ package_name_from_call_stack <-
     return("base?")
   }
 
+
+# we try to find a same combination of caller_package, caller_function and eval_call_expression in both datasets
+# if yes, we inditifed a call site
+# data set undefined should be the packages with missing srcref and which caller_function is not in base
+package_name_from_static <- function(dataset_undefined, static_data) {
+  dataset_undefined %>%
+    semi_join(static_data, by = c("eval_call_expression", "caller_package")) %>%
+    count(eval_call_expression, caller_package, caller_function)
+}
+
 extract_package_name <- function(src_ref, file) {
   # There are 5 possibilities for a srcref:
   # - NA
@@ -313,6 +323,7 @@ add_eval_source_type <- function(dataset) {
     ))
 }
 
+
 add_fake_srcref <- function(dataset) {
   return(dataset %>%
     mutate(
@@ -363,12 +374,12 @@ add_package <- function(dataset) {
 
 keep_only_corpus <- function(dataset, corpus_files) {
   return(dataset %>%
-           semi_join(corpus_files, by = c("eval_source" = "package")))
+    semi_join(corpus_files, by = c("eval_source" = "package")))
 }
 
 get_externals <- function(dataset, corpus_files) {
   return(dataset %>%
-           anti_join(corpus_files, by = c("eval_source" = "package")))
+    anti_join(corpus_files, by = c("eval_source" = "package")))
 }
 
 undefined_packages <- function(eval_calls) {
@@ -553,7 +564,7 @@ main <- function(argv) {
 
   write_fst(eval_calls_externals, evals_summarized_externals_file)
 
-  #write_fst(calls_site_per_package, package_evals_dynamic_file)
+  # write_fst(calls_site_per_package, package_evals_dynamic_file)
   res <- difftime(Sys.time(), now)
   cat("Done in ", res, units(res), "\n")
 

### 2020-12-15 analysis/Attic/analysis_stable.Rmd

--- 2020-09-14 inst/analysis/analysis_stable.Rmd
@@
 ```
-
-# Corpus
-
-TODO
 
@@
 
-### String size
-
-```{r}
-eval_calls %>% ggplot() +
-  geom_histogram(aes(x = expr_expression_length, weight = nb_ev_calls), na.rm = TRUE, binwidth=1) +
-  scale_x_log10() +  scale_y_sqrt()
-```
-
-If we remove the outliers:
-```{r}
-eval_calls %>% filter(expr_expression_length < 25000) %>% ggplot() +
-  geom_histogram(aes(x = expr_expression_length, weight = nb_ev_calls), na.rm = TRUE, binwidth=2)  + scale_x_sqrt() +  scale_y_sqrt() 
-```
-
-What about the code size of expressions that seem to be generated?
-An approximation of *generated* will be no `srcref` and not in `base`.
-
-```{r code_size_srcref}
-eval_calls %>% filter(is.na(eval_call_srcref), caller_package != "base") %>% ggplot() +
-  geom_histogram(aes(x = expr_expression_length, weight = nb_ev_calls), na.rm = TRUE, binwidth=1) + scale_x_sqrt() + scale_y_sqrt()
-```
-
-And the average size per package:
-```{r average_size_package}
-eval_calls %>% group_by(caller_package) %>% summarize(average_expr_size = mean(expr_expression_length, na.rm = TRUE)) %>% arrange(desc(average_expr_size))
-```
-
-If we look at the maxima, the largest ones are in *base*:
-```{r max_size_package}
-eval_calls %>% group_by(caller_package) %>% summarize(max_expr_size = max(expr_expression_length, na.rm = TRUE)) %>% arrange(desc(max_expr_size))
-```
-
 ### AST size
@@
 
-classified_envir_packages %>% select(env_class, eval_call_expression, eval_call_srcref, caller_expression, n) %>% distinct() %>% datatable()
+classified_envir_packages %>% select(env_class, eval_call_expression, eval_call_srcref, caller_expression, caller_srcref, n) %>% distinct() %>% datatable()
+```
+
+And per call sites?
+
+```{r frequent_env_per_call_sites}
+classified_envir_prop_core_call_sites <- classified_envir %>% 
+  filter(dataset == "core") %>%
+  group_by(env_class) %>% 
+  summarize(n = vec_unique_count(eval_call_srcref)) %>%
+  arrange(desc(n)) %>% 
+  mutate(proportion = n / sum(n))
+
+classified_envir_prop_packages_call_sites <- classified_envir %>% 
+  filter(dataset == "packages") %>%
+  group_by(env_class) %>% 
+  summarize(n = vec_unique_count(eval_call_srcref)) %>%
+  arrange(desc(n)) %>% 
+  mutate(proportion = n / sum(n))
+
 ```

### 2020-12-15 analysis/corpus.Rmd

--- 2020-08-28 inst/analysis/corpus-analysis.Rmd
@@
   theme(
-    legend.position=c(0.2, 0.85),
+    legend.position=c(0.15, 0.73),
     legend.box="horizontal",
@@
     linetype=FALSE
-  )
+  ) +
+  expand_limits(y = 300)
 

### 2020-12-15 analysis/environments.Rmd

@@
       paste("Time for this code chunk to run:", res, units(res))
-}}}))
+    }
+  }
+}))
 
-knitr::opts_chunk$set(echo = TRUE,
+knitr::opts_chunk$set(
+  echo = TRUE,
   fig.align = "center",
@@
   fig.width = 10,
-                      cache.lazy = FALSE)
+  cache.lazy = FALSE
+)
 
@@
 ```{r}
-eval_calls_arg_caller %>% count(envir_from_arg, wt=nb_ev_calls, sort = TRUE) %>% knitr::kable()
+eval_calls_arg_caller %>%
+  count(envir_from_arg, wt = nb_ev_calls, sort = TRUE) %>%
+  knitr::kable()
 ```
@@
 ```{r envir_absolute}
-eval_calls_absolute <- eval_calls_explicit %>% anti_join(eval_calls_env_list) %>% anti_join(eval_calls_arg_caller)
+eval_calls_absolute <- eval_calls_explicit %>%
+  anti_join(eval_calls_env_list) %>%
+  anti_join(eval_calls_arg_caller)
 
@@
 ```{r}
-eval_calls_absolute <- eval_calls_absolute %>% group_by(eval_call_srcref) %>%  mutate(envir_distinct = n_distinct(environment_class))
+eval_calls_absolute <- eval_calls_absolute %>%
+  group_by(eval_call_srcref) %>%
+  mutate(envir_distinct = n_distinct(environment_class))
 
-eval_calls_absolute %>% ungroup() %>% count(envir_distinct, sort = TRUE)
+eval_calls_absolute %>%
+  ungroup() %>%
+  count(envir_distinct, sort = TRUE)
 ```
@@
 ```{r}
-eval_calls_absolute %>% filter(envir_distinct == 2) %>% select(eval_call_expression, environment_class, envir_distinct)
+eval_calls_absolute %>%
+  filter(envir_distinct == 2) %>%
+  select(eval_call_expression, environment_class, envir_distinct)
 ```

### 2020-12-15 analysis/usage.Rmd

--- 2020-12-14 inst/analysis/usage_metrics.Rmd
@@
 
+```{r eval-calls-summary}
+
+eval_calls_table <- E %>% 
+  count(ev_package, wt=duplicates, sort = TRUE)
+
+breaks <- c(0, 10, 100, 1000, 10000, 100000, 1000000, 10000000, 100000000)
+
+group_tags <- cut(eval_calls_table$n,
+                  breaks = breaks,
+                  include.lowest = TRUE)
+
+ev_summary <- summary(group_tags)
+
+r("bin1", ev_summary[[1]])
+r("bin2", ev_summary[[2]])
+r("bin3", ev_summary[[3]])
+r("bin4", ev_summary[[4]])
+r("bin5", ev_summary[[5]])
+r("bin6", ev_summary[[6]])
+r("bin7", ev_summary[[7]])
+r("bin8", ev_summary[[8]])
+```
+
 # Triggered sites
@@
 r("static at least one call site percent", ratio(nrow(sites_per_package %>% filter(n >= 1)), nrow(static_at_least_one_call_site)))
+```
+
+And now, by `eval` variant:
+
+```{r exercised-call-site-per-variant}
+
+static_eval <- P %>% filter(call_fun_name == "eval") %>% nrow()
+static_evalq <- P %>% filter(call_fun_name == "evalq") %>% nrow()
+static_eval.parent <- P %>% filter(call_fun_name == "eval.parent") %>% nrow()
+static_local <- P %>% filter(call_fun_name == "local") %>% nrow()
+
+sites_per_variant <- E %>% # src_ref indicates where the call originates from.
+  group_by(ev_variant) %>% 
+  summarize( n = n_distinct(src_ref)) # This the number of distinct sites
+
+trig_eval <- sites_per_variant %>% filter(ev_variant == "eval") %>% .$n
+trig_evalq <- sites_per_variant %>% filter(ev_variant == "evalq") %>% .$n
+trig_eval.parent <- sites_per_variant %>% filter(ev_variant == "eval.parent") %>% .$n
+trig_local <- sites_per_variant %>% filter(ev_variant == "local") %>% .$n
+
+r("static eval", static_eval)
+r("static evalq", static_evalq)
+r("static eval.parent", static_eval.parent)
+r("static local", static_local)
+
+r("triggered eval", trig_eval)
+r("triggered evalq", trig_evalq)
+r("triggered eval.parent", trig_eval.parent)
+r("triggered local", trig_local)
+
+r("eval percent call site", ratio(trig_eval, static_eval))
+r("evalq percent call site", ratio(trig_evalq, static_evalq))
+r("eval.parent percent call site", ratio(trig_eval.parent, static_eval.parent))
+r("local percent call site", ratio(trig_local, static_local))
 ```
@@
 
+```{r}
+eval_call_triggered <- E %>% 
+  group_by(src_ref) %>%
+  summarize(n_runs = n_distinct(file), n = sum(duplicates), average = n / n_runs) %>%
+  arrange(desc(average))
+```
+
 # The case of core libraries

### 2020-12-15 analysis/usage.Rmd
Fix cases when one of the variant is absent from the dataset.

@@
 
-trig_eval <- sites_per_variant %>% filter(ev_variant == "eval") %>% .$n
-trig_evalq <- sites_per_variant %>% filter(ev_variant == "evalq") %>% .$n
-trig_eval.parent <- sites_per_variant %>% filter(ev_variant == "eval.parent") %>% .$n
+trig_eval <- sites_per_variant %>% filter(ev_variant == "eval")
+trig_eval <- if(nrow(trig_eval) == 0)
+{
+  0
+} else
+{
+  trig_eval$n
+}
+
+trig_evalq <- sites_per_variant %>% filter(ev_variant == "evalq")
+trig_evalq <- if(nrow(trig_evalq) == 0)
+{
+  0
+} else
+{
+  trig_evalq$n
+}
+trig_eval.parent <- sites_per_variant %>% filter(ev_variant == "eval.parent") 
+trig_eval.parent <- if(nrow(trig_eval.parent) == 0)
+{
+  0
+} else
+{
+  trig_eval.parent$n
+}
 trig_local <- sites_per_variant %>% filter(ev_variant == "local") %>% .$n
+trig_local <- if(nrow(trig_local) == 0)
+{
+  0
+} else
+{
+  trig_local$n
+}
 

### 2020-12-16 analysis/usage.Rmd
More readable bins
Fix corpus size
changes ot usage

@@
 params:
-  base_dir: /var/lib/R/project-evalR/run/package-evals-traced.10/
+  base_dir: Data/
   calls_packages_path: summarized-packages.fst
@@
 library(DT)
-
-jan <- FALSE
 
@@
 
-if (jan == FALSE) {
 source("inc/paths.R")
-}
 source("inc/latextags.R")
@@
 theme_set(theme_minimal())
-if (jan) {
-  TAGS_DIR <- PLOT_DIR <- "."
-}
 create_tags(path(TAGS_DIR, "usage_metrics.tex"), prefix = "", default = TRUE)
@@
 base_dir <- params$base_dir
-# Set the paths, and customize for Jan.
-if (jan) {
-  base_dir <- "/mnt/code/eval/run/package-evals-traced.9/"
-}
 eval_calls_path <- paste0(base_dir, params$calls_packages_path)
@@
 
-```{r read,include=F}
+```{r read, message=TRUE, include=F}
 read_fst(eval_calls_path) %>% as_tibble() -> E_raw
@@
 
-```{r select-columns,echo=F}
+```{r select-columns, message=TRUE, warning=FALSE, include=FALSE}
 E_raw %>% select(
@@
 corpus_size <- nrow(P)
+
+overview_table(
+  r("Corpus",corpus_size)
+) 
 ```
+
+The size of our corpus is `corpus_size` packages.
 
@@
 
-The corpus has `r corpus_size` packages, and the data set has `r nrow(E)` 
-distinct calls to the `eval` and its variants. There are 
-`r length(unique(E$ev_package))` corpus packages with `eval` calls (note that
-we may not trigger all `eval`s in a package due to spotty covereage).
+The data set has `r nrow(E)`distinct calls to the `eval` and its variants. 
+There are `r length(unique(E$ev_package))` corpus packages with `eval` calls
+(note that we may not trigger all `eval`s in a package due to spotty covereage).
 
@@
 
-TBL <- overview_table(
-  r("nb eval and variant calls", nb_calls),
-  r("nb eval calls", eval_calls),
-  r("nb evalq calls", evalq_calls),
-  r("nb eval parent calls", eval.parent_calls),
-  r("nb local calls", local_calls)
-)
+overview_table(
+  r("Allcalls", nb_calls),
+  r("Evals", eval_calls),
+  r("Evalqs", evalq_calls),
+  r("Eparents", eval.parent_calls),
+  r("Locals", local_calls),
+  r("Triggeredpkgs", length(unique(E$ev_package)))
+) -> TBL
 ```
@@
 
-```{r eval-calls-summary}
+# Ev-dist
 
+We compute the frequency of evals by package. This shows that some packages
+hardly use eval while other use it intensily. The numbers are "for all runs",
+this means that a popular package may be triggered by more runs and thus
+have higher frequencies. We will later control for this by looking at individual 
+call sites.
+
+```{r ev-dist, include=FALSE}
 eval_calls_table <- E %>% 
   count(ev_package, wt=duplicates, sort = TRUE)
-
-breaks <- c(0, 10, 100, 1000, 10000, 100000, 1000000, 10000000, 100000000)
-
+breaks <- c(0, 10^seq(1,8))
 group_tags <- cut(eval_calls_table$n,
@@
                   include.lowest = TRUE)
-
 ev_summary <- summary(group_tags)
 
-r("bin1", ev_summary[[1]])
-r("bin2", ev_summary[[2]])
-r("bin3", ev_summary[[3]])
-r("bin4", ev_summary[[4]])
-r("bin5", ev_summary[[5]])
-r("bin6", ev_summary[[6]])
-r("bin7", ev_summary[[7]])
-r("bin8", ev_summary[[8]])
+overview_table(
+r("bina", ev_summary[[1]]),
+r("binb", ev_summary[[2]]),
+r("binc", ev_summary[[3]]),
+r("bind", ev_summary[[4]]),
+r("bine", ev_summary[[5]]),
+r("binf", ev_summary[[6]]),
+r("bing", ev_summary[[7]]),
+r("binh", ev_summary[[8]])) -> TBL
 ```
+
+The distribution of eval calls, each bin represents a power of 10, the number in the bin is the count of packages.
+
+`r TBL`
 

### 2020-12-16 paper/main.tex
changes ot usage

--- 2020-12-07 inst/paper/main.tex
@@
 \documentclass[USenglish,cleveref, autoref, thm-restate]{lipics-v2019}
-\usepackage{listings,hyperref,multirow,paralist,xspace,url}
+\usepackage{listings,hyperref,multirow,paralist,xspace,url,wrapfig}
 \input{macro}
@@
 
-\begin{document}\maketitle
+\begin{document}
+\maketitle
 
@@
 
-This section presents a high-level picture of the usage of \eval in R.
-
-(CHECK IF TRUE) As none of the Kaggle programs invoke \eval, the
-remainder of the paper focuses solely on the CRAN data. We will return
-to Kaggle in our conclusion.
-
-We recorded \NbEvalAndVariantCallsRnd calls to \eval and its variants (\NbEvalCallsRnd
-for \eval, \NbEvalqCallsRnd for \c{evalq}, \NbEvalParentCallsRnd for \c{eval.parent} and \NbLocalCallsRnd for
-\c{local}).
+To provide a picture of \eval usage in R, we recorded \Allcalls
+invocations of \eval and its variants in the CRAN corpus. These calls
+were obtained by XXX runs, with runs that did not trigger \eval
+discarded. Out of the \Corpus CRAN packages with \eval call sites, we
+observed calls in \Triggeredpkgs packages, the difference can be chalk
+down to incomplete code coverage and occasional failures due to our
+pipeline.
+
+\begin{wrapfigure}{r}{5.2cm}
+\begin{tabular}{|r@{\,}r@{\,}lr|r@{\,}r@{\,}lr|} \hline
+\tt 0 &--& \tt 10      & \Bina  & \tt 1K &--&\tt 100K  & \Bine\\
+\tt 11 &--& \tt 100    & \Binb  & \tt 100K &--&\tt 1M  & \Binf\\
+\tt 101 &--& \tt 1K    & \Binc  & \tt 1M &--&\tt 10M   & \Bing\\
+\tt 1K &--& \tt 10K    & \Bind  & \tt 10M &--& \tt 100M & \Binh\\\hline
+\end{tabular}
+\caption{Call frequency}\label{freq}
+\end{wrapfigure}
+
+Figure~\ref{freq} shows the number of \evals generated by each package
+over all runs that invoked functions in that package. Out of the XXX
+packages in the corpus, XXX call have a low frequency of \eval calls,
+with fewer than 100 of them. On the other hand, XXX packages call eval
+more than 1,000 times, with one package (???) calling it XXXX times.
+
+\end{document}
+
+(\EvalsRnd for \eval, \EvalqsRnd for \c{evalq}, \EparentsRnd for
+\c{eval.parent} and \LocalsRnd for \c{local}).
 
@@
 
-\mypara{Number of eval call sites.}
-We observed that the number of \eval call sites is small.
-Figure~\ref{fig:callSitesPerPackage} shows a violin plot of the distribution of call sites to
-\eval and its three variants. For packages that have call sites, the
-median is \NbMedianCallSites call sites. \NbOneCallSite{} packages have a
-single call site, and one package, \emph{\MaxCallSitesPackage}, has
-\MaxCallSites{} call sites. \PackagesNoCallsToEval{} packages have
-no calls sites to \eval. These numbers are under-approximations as \eval may
-be aliased, called reflectively and given as argument to a
+\mypara{Number of eval call sites.}  We observed that the number of
+\eval call sites is small.  Figure~\ref{fig:callSitesPerPackage} shows
+a violin plot of the distribution of call sites to \eval and its three
+variants. For packages that have call sites, the median is
+\NbMedianCallSites call sites. \NbOneCallSite{} packages have a single
+call site, and one package, \emph{\MaxCallSitesPackage}, has
+\MaxCallSites{} call sites. \PackagesNoCallsToEval{} packages have no
+calls sites to \eval. These numbers are under-approximations as \eval
+may be aliased, called reflectively and given as argument to a
 higher-order function such as \c{map}.

### 2020-12-16 analysis/inc/paths.R

@@ -2,72 +2,72 @@
 # PATHS
 ################################################################################
 
-DATA_DIR  <- path(params$base_dir, "revalstudy", "inst", "data")
-PAPER_DIR <- path(params$base_dir, "revalstudy", "inst", "paper")
-RUN_DIR   <- path(params$base_dir, "run")
+DATA_DIR  <- path("Data")
+PAPER_DIR <- path("..", "paper")
+# RUN_DIR   <- path(params$base_dir, "run")
 PLOT_DIR  <- path(PAPER_DIR, "img")
 TAGS_DIR  <- path(PAPER_DIR, "tag")
 
 if (!dir_exists(PLOT_DIR)) dir_create(PLOT_DIR)
 if (!dir_exists(TAGS_DIR)) dir_create(TAGS_DIR)
 
-R_DIR      <- path(params$base_dir, "R-4.0.2")
-R_LIBS_DIR <- path(params$base_dir, "library", "4.0")
+#R_DIR      <- path(params$base_dir, "R-4.0.2")
+#R_LIBS_DIR <- path(params$base_dir, "library", "4.0")
 
 ## INPUTs
 
-PACKAGE_EVALS_DYNAMIC_FILE <- path(DATA_DIR, "evals-dynamic.fst")
-PACKAGE_EVALS_FILE         <- path(DATA_DIR, "evals.fst")
-PACKAGE_COVERAGE_FILE      <- path(DATA_DIR, "coverage.fst")
-PACKAGE_METADATA_FILE      <- path(DATA_DIR, "metadata.fst")
-PACKAGE_SLOC_FILE          <- path(DATA_DIR, "sloc.fst")
-PACKAGE_REVDEPS_FILE       <- path(DATA_DIR, "revdeps.fst")
-PACKAGE_FUNCTIONS_FILE     <- path(DATA_DIR, "functions.fst")
-PACKAGE_RUNNABLE_CODE_FILE <- path(DATA_DIR, "runnable-code-metadata.fst")
-PACKAGE_TRACE_RUNS_FILE    <- path(DATA_DIR, "run-trace.fst")
-PACKAGE_TRACE_LOG_FILE     <- path(DATA_DIR, "parallel-trace.fst")
-CRAN_PROGRAMS_FILE         <- path(DATA_DIR, "cran-programs.fst")
+#PACKAGE_EVALS_DYNAMIC_FILE <- path(DATA_DIR, "evals-dynamic.fst")
[…]

### 2020-12-16 analysis/usage.Rmd
Fix paths

@@
 ---
-
-The setup chunk has a variable `jan` with setting to run the script on
-docker on Jan's machine. This should be removed.
 
@@
 corpus_size <- nrow(P)
-
 overview_table(
@@
 
-The data set has `r nrow(E)`distinct calls to the `eval` and its variants. 
+The data set has `r nrow(E)` distinct calls to the `eval` and its variants. 
 There are `r length(unique(E$ev_package))` corpus packages with `eval` calls
-(note that we may not trigger all `eval`s in a package due to spotty covereage).
+(note that we may not trigger all `eval`s in a package due to spotty coverage).
 
@@
 }
-trig_local <- sites_per_variant %>% filter(ev_variant == "local") %>% .$n
+trig_local <- sites_per_variant %>% filter(ev_variant == "local")
 trig_local <- if(nrow(trig_local) == 0)

### 2020-12-16 analysis/usage.Rmd

@@
 r("static evalq", static_evalq)
-r("static eval.parent", static_eval.parent)
+r("static eval parent", static_eval.parent)
 r("static local", static_local)
@@
 r("triggered evalq", trig_evalq)
-r("triggered eval.parent", trig_eval.parent)
+r("triggered eval parent", trig_eval.parent)
 r("triggered local", trig_local)
@@
 r("evalq percent call site", ratio(trig_evalq, static_evalq))
-r("eval.parent percent call site", ratio(trig_eval.parent, static_eval.parent))
+r("eval parent percent call site", ratio(trig_eval.parent, static_eval.parent))
 r("local percent call site", ratio(trig_local, static_local))

### 2020-12-16 analysis/usage.Rmd

@@
 corpus_size <- nrow(P)
+
+nb_runs <- E %>%
+  pull(file) %>%
+  n_distinct()
 overview_table(
-  r("Corpus",corpus_size)
+  r("Corpus",corpus_size),
+  r("NbRuns", nb_runs)
 ) 
@@
 
-The size of our corpus is `corpus_size` packages.
+The size of our corpus is `r corpus_size` packages.
 
@@
 
+```{r ev-call-precise}
+
+more_1000_calls <- eval_calls_table %>% filter(n > 1000) %>% nrow()
+
+overview_table(
+  r("eval calls less hundred", ev_summary[[1]] + ev_summary[[2]]),
+  r("eval calls more thousand", more_1000_calls)
+)
+```
+
 # Triggered sites
@@
   r("nb one call site", sites_per_package %>% filter(n == 1)),
-  r("at least one call site", sites_per_package),
+  r("at least one call site", sites_per_package$n),
   r("packages no calls to eval", corpus_size - nrow(sites_per_package)),
@@
   count(ev_package, wt = duplicates) %>%
-  left_join(runs_package, by = "ev_package") %>%
-  mutate(n = n / nb_runs)
+  left_join(runs_package, by = "ev_package")
 

### 2020-12-16 analysis/usage.Rmd
Tableof eval variants

@@
 
+```{r exercised-call-site-per-variant}
+static_eval <- P %>% filter(call_fun_name == "eval") %>% nrow()
+static_evalq <- P %>% filter(call_fun_name == "evalq") %>% nrow()
+static_eval.parent <- P %>% filter(call_fun_name == "eval.parent") %>% nrow()
+static_local <- P %>% filter(call_fun_name == "local") %>% nrow()
+
+sites_per_variant <- E %>% # src_ref indicates where the call originates from.
+  group_by(ev_variant) %>% 
+  summarize( n = n_distinct(src_ref)) # This the number of distinct sites
+
+trig_eval <- sites_per_variant %>% filter(ev_variant == "eval")
+trig_eval <- if(nrow(trig_eval) == 0)
+{
+  0
+} else
+{
+  trig_eval$n
+}
+
+trig_evalq <- sites_per_variant %>% filter(ev_variant == "evalq")
+trig_evalq <- if(nrow(trig_evalq) == 0)
+{
+  0
+} else
+{
+  trig_evalq$n
+}
+trig_eval.parent <- sites_per_variant %>% filter(ev_variant == "eval.parent") 
+trig_eval.parent <- if(nrow(trig_eval.parent) == 0)
+{
+  0
+} else
+{
+  trig_eval.parent$n
+}
+trig_local <- sites_per_variant %>% filter(ev_variant == "local")
+trig_local <- if(nrow(trig_local) == 0)
+{
+  0
+} else
+{
+  trig_local$n
+}
+
+overview_table(
+  r("StaticEval", static_eval),
+  r("StaticEvalq", static_evalq),
+  r("StaticEvalParent", static_eval.parent),
+  r("StaticLocal", static_local),
+  r("TriggeredEval", trig_eval),
+  r("TriggeredEvalq", trig_evalq),
+  r("Triggeredvalparent", trig_eval.parent),
+  r("Triggeredlocal", trig_local),
+  r("Triggeredevalpercent", ratio(trig_eval, static_eval)),
+  r("Triggeredevalqpercent", ratio(trig_evalq, static_evalq)),
+  r("Triggeredevalqpercent", ratio(trig_eval.parent, static_eval.parent)),
+  r("Localevalqpercent", ratio(trig_local, static_local))
+) -> TBL
+```
+
+`r TBL`
+
 # Ev-dist
@@
   r("nb one call site", sites_per_package %>% filter(n == 1)),
-  r("at least one call site", sites_per_package$n),
+  r("at least one call site", nrow(sites_per_package)),
   r("packages no calls to eval", corpus_size - nrow(sites_per_package)),
@@
 r("static at least one call site percent", ratio(nrow(sites_per_package %>% filter(n >= 1)), nrow(static_at_least_one_call_site)))
-```
-
-And now, by `eval` variant:
-
-```{r exercised-call-site-per-variant}
-
-static_eval <- P %>% filter(call_fun_name == "eval") %>% nrow()
-static_evalq <- P %>% filter(call_fun_name == "evalq") %>% nrow()
-static_eval.parent <- P %>% filter(call_fun_name == "eval.parent") %>% nrow()
-static_local <- P %>% filter(call_fun_name == "local") %>% nrow()
-
-sites_per_variant <- E %>% # src_ref indicates where the call originates from.
-  group_by(ev_variant) %>% 
-  summarize( n = n_distinct(src_ref)) # This the number of distinct sites
-
-trig_eval <- sites_per_variant %>% filter(ev_variant == "eval")
-trig_eval <- if(nrow(trig_eval) == 0)
-{
-  0
-} else
-{
-  trig_eval$n
-}
-
-trig_evalq <- sites_per_variant %>% filter(ev_variant == "evalq")
-trig_evalq <- if(nrow(trig_evalq) == 0)
-{
-  0
-} else
-{
-  trig_evalq$n
-}
-trig_eval.parent <- sites_per_variant %>% filter(ev_variant == "eval.parent") 
-trig_eval.parent <- if(nrow(trig_eval.parent) == 0)
-{
-  0
-} else
-{
-  trig_eval.parent$n
-}
-trig_local <- sites_per_variant %>% filter(ev_variant == "local")
-trig_local <- if(nrow(trig_local) == 0)
-{
-  0
-} else
-{
-  trig_local$n
-}
-
-r("static eval", static_eval)
-r("static evalq", static_evalq)
-r("static eval parent", static_eval.parent)
-r("static local", static_local)
-
-r("triggered eval", trig_eval)
-r("triggered evalq", trig_evalq)
-r("triggered eval parent", trig_eval.parent)
-r("triggered local", trig_local)
-
-r("eval percent call site", ratio(trig_eval, static_eval))
-r("evalq percent call site", ratio(trig_evalq, static_evalq))
-r("eval parent percent call site", ratio(trig_eval.parent, static_eval.parent))
-r("local percent call site", ratio(trig_local, static_local))
 ```

### 2020-12-16 paper/main.tex

@@
 invocations of \eval and its variants in the CRAN corpus. These calls
-were obtained by XXX runs, with runs that did not trigger \eval
-discarded. Out of the \Corpus CRAN packages with \eval call sites, we
+were obtained by \Nbruns runs, with runs that did not trigger \eval
+discarded. Out of the \NbStaticAtLeastOneCallSite CRAN packages(among the \Corpus packages of our corpus) with \eval call sites, we
 observed calls in \Triggeredpkgs packages, the difference can be chalk
@@
 Figure~\ref{freq} shows the number of \evals generated by each package
-over all runs that invoked functions in that package. Out of the XXX
-packages in the corpus, XXX call have a low frequency of \eval calls,
-with fewer than 100 of them. On the other hand, XXX packages call eval
-more than 1,000 times, with one package (???) calling it XXXX times.
+over all runs that invoked functions in that package. Out of the \Corpus
+packages in the corpus, \EvalCallsLessHundred packages have a low frequency of \eval calls,
+with fewer than 100 of them. On the other hand, \EvalCallsMoreThousand packages call eval
+more than 1,000 times, with one package (\MaxEvalCallPackage) calling it \MaxEvalCallsRnd times.
 

### 2020-12-16 analysis/usage.Rmd
Actual numbers for variants

@@
   r("TriggeredEvalq", trig_evalq),
-  r("Triggeredvalparent", trig_eval.parent),
+  r("Triggeredevalparent", trig_eval.parent),
   r("Triggeredlocal", trig_local),

### 2020-12-16 analysis/usage.Rmd
Correct names

--- 2020-12-16 analysis/usage.Rmd
@@
   r("Triggeredevalqpercent", ratio(trig_evalq, static_evalq)),
-  r("Triggeredevalqpercent", ratio(trig_eval.parent, static_eval.parent)),
-  r("Localevalqpercent", ratio(trig_local, static_local))
+  r("Triggeredevalparentpercent", ratio(trig_eval.parent, static_eval.parent)),
+  r("Triggeredlocalpercent", ratio(trig_local, static_local))
 ) -> TBL

### 2020-12-16 analysis/usage.Rmd
usage

--- 2020-12-16 analysis/usage.Rmd
@@
 ```{r read, message=TRUE, include=F}
+# This test should allow to re-run the whole script without reading the
+# data again from disk. It assumes the data doesn't change.
+if (exists(eval_calls_path) == FALSE) {
   read_fst(eval_calls_path) %>% as_tibble() -> E_raw
@@
   read_csv(eval_static_path) %>% semi_join(C_raw, by = "package") -> P
+}
 ```
@@
 C_raw -> C
-corpus_size <- nrow(P)
+corpus_size <- nrow(C)
 
@@
   r("Corpus",corpus_size),
-  r("NbRuns", nb_runs)
+  r("Nbruns", nb_runs)
 ) 
@@
 
+# Ev-dist
+
+We compute the frequency of evals by package. This shows that some packages
+hardly use eval while other use it intensly. The numbers are "for all runs",
+this means that a popular package may be triggered by more runs and thus
+have higher frequencies. We will later control for this by looking at individual 
+call sites.
+
+```{r ev-dist, include=FALSE}
+# The calls per package
+pack_calls <- E %>% 
+  count(ev_package, wt=duplicates, sort = TRUE)
+# Create bins at every power of ten
+bins <- cut(pack_calls$n, breaks = c(0, 10^seq(1,8)), include.lowest = TRUE)
+sum <- summary(bins)
+# Packages with more than 1K calls
+Kcalls <- pack_calls %>% filter(n > 1000) %>% nrow()
+
+# The number of runs per package
+runs <- E %>%
+  group_by(ev_package) %>%
+  summarise(nb_runs = n_distinct(file))
+
+# Number of eval calls per package
+ecpp <- pack_calls %>% left_join(runs, by = "ev_package") 
+stopifnot( nrow(filter(ecpp, n==0)) == 0)
+# The package with the most calls
+max_calls <- ecpp %>% filter(n == max(n))
+
+overview_table(
+r("bina", sum[[1]]),
+r("binb", sum[[2]]),
+r("binc", sum[[3]]),
+r("bind", sum[[4]]),
+r("bine", sum[[5]]),
+r("binf", sum[[6]]),
+r("bing", sum[[7]]),
+r("binh", sum[[8]]),
+r("Fewcalls", sum[[1]] + sum[[2]]),
+r("Manycalls", Kcalls),
+r("Maxcalls", max_calls$n),
+r("Maxcallspack", max_calls$ev_package)
+) -> TBL
+```
+
+The distribution of eval calls, each bin represents a power of 10, the number 
+in the bin is the count of packages.
+
+`r TBL`
+
+# Triggered sites
+
+```{r triggered-sites, include=F}
+stopifnot(sum(is.na(E$src_ref)) == 0) # No NAs in scr_ref!
+
+sites <- E %>% # src_ref indicates where the call originates from.
+  pull(src_ref) %>%
+  n_distinct() # This the number of distinct sites
+
+sites_per_package <- E %>%
+  count(ev_package, src_ref) %>%
+  count(ev_package) # All packages have at least one site!
+
+max_packages <- sites_per_package %>%
+  filter(n == max(n)) %>%
+  .$n
+
+TBL <- overview_table(
+  # Sites triggered over the whole corpus:
+  r("nb eval call sites", sites),
+  # How many packages in the corpus have a single triggered call site:
+  r("nb one call site", sites_per_package %>% filter(n == 1)),
+  r("at least one call site", sites_per_package$n),
+  r("packages no calls to eval", corpus_size - nrow(sites_per_package)),
+  r("nb median call sites", median(sites_per_package$n)),
+  r("max call sites", max_packages),
+)
+```
+
+`r TBL`
+
+```{r call-sites-violin-plot, include=F}
+E %>%
+  group_by(ev_package) %>%
+  mutate(sites = n_distinct(src_ref)) %>%
+  ggplot(aes(x = ev_variant, y = sites, fill = ev_variant), na.rm = TRUE) +
+  geom_violin(trim = FALSE, width = 1.7) +
+  # geom_jitter(shape=16, position=position_jitter(0.2)) +
+  labs(y = "Call sites per package") +
+  theme(
+    legend.position = c(0.5, 0.85),
+    legend.box.background = element_rect(fill = "white", size = 0.1),
+    legend.box = "horizontal",
+    axis.title.x = element_blank()
+  ) -> PLOT
+
+ggsave(path(PLOT_DIR, "call_sites_per_packages.pdf"))
+```
+
+How many call sites and how many packages did we exercise out of the total number of call sites of the corpus?
+
+```{r exercised-call-sites, include=F}
+
+# TODO :: remove?
+static_sites <- nrow(P)
+
+static_call_sites_per_package <- P %>%
+  count(package, sort = TRUE)
+
+static_at_least_one_call_site <- static_call_sites_per_package %>%
+  filter(n >= 1)
+
+r("nb static call sites", static_sites)
+r("call sites percent", ratio(sites, static_sites))
+r("nb static at least one call site", nrow(static_at_least_one_call_site))
+r("static at least one call site percent", ratio(nrow(sites_per_package %>% filter(n >= 1)), nrow(static_at_least_one_call_site)))
+```
+
+And now, by `eval` variant:
+
 ```{r exercised-call-site-per-variant}
+
 static_eval <- P %>% filter(call_fun_name == "eval") %>% nrow()
@@
 
-overview_table(
-  r("StaticEval", static_eval),
-  r("StaticEvalq", static_evalq),
-  r("StaticEvalParent", static_eval.parent),
-  r("StaticLocal", static_local),
-  r("TriggeredEval", trig_eval),
-  r("TriggeredEvalq", trig_evalq),
-  r("Triggeredvalparent", trig_eval.parent),
-  r("Triggeredlocal", trig_local),
-  r("Triggeredevalpercent", ratio(trig_eval, static_eval)),
-  r("Triggeredevalqpercent", ratio(trig_evalq, static_evalq)),
-  r("Triggeredevalqpercent", ratio(trig_eval.parent, static_eval.parent)),
-  r("Localevalqpercent", ratio(trig_local, static_local))
-) -> TBL
-```
+r("static eval", static_eval)
+r("static evalq", static_evalq)
+r("static eval parent", static_eval.parent)
+r("static local", static_local)
 
-`r TBL`
+r("triggered eval", trig_eval)
+r("triggered evalq", trig_evalq)
+r("triggered eval parent", trig_eval.parent)
+r("triggered local", trig_local)
 
-# Ev-dist
-
-We compute the frequency of evals by package. This shows that some packages
-hardly use eval while other use it intensily. The numbers are "for all runs",
-this means that a popular package may be triggered by more runs and thus
-have higher frequencies. We will later control for this by looking at individual 
-call sites.
-
-```{r ev-dist, include=FALSE}
-eval_calls_table <- E %>% 
-  count(ev_package, wt=duplicates, sort = TRUE)
-breaks <- c(0, 10^seq(1,8))
-group_tags <- cut(eval_calls_table$n,
-                  breaks = breaks,
-                  include.lowest = TRUE)
-ev_summary <- summary(group_tags)
-
-overview_table(
-r("bina", ev_summary[[1]]),
-r("binb", ev_summary[[2]]),
-r("binc", ev_summary[[3]]),
-r("bind", ev_summary[[4]]),
-r("bine", ev_summary[[5]]),
-r("binf", ev_summary[[6]]),
-r("bing", ev_summary[[7]]),
-r("binh", ev_summary[[8]])) -> TBL
-```
-
-The distribution of eval calls, each bin represents a power of 10, the number in the bin is the count of packages.
-
-`r TBL`
-
-```{r ev-call-precise}
-
-more_1000_calls <- eval_calls_table %>% filter(n > 1000) %>% nrow()
-
-overview_table(
-  r("eval calls less hundred", ev_summary[[1]] + ev_summary[[2]]),
-  r("eval calls more thousand", more_1000_calls)
-)
-```
-
-# Triggered sites
-
-```{r triggered-sites, include=F}
-stopifnot(sum(is.na(E$src_ref)) == 0) # No NAs in scr_ref!
-
-sites <- E %>% # src_ref indicates where the call originates from.
-  pull(src_ref) %>%
-  n_distinct() # This the number of distinct sites
-
-sites_per_package <- E %>%
-  count(ev_package, src_ref) %>%
-  count(ev_package) # All packages have at least one site!
-
-max_packages <- sites_per_package %>%
-  filter(n == max(n)) %>%
-  .$n
-
-TBL <- overview_table(
-  # Sites triggered over the whole corpus:
-  r("nb eval call sites", sites),
-  # How many packages in the corpus have a single triggered call site:
-  r("nb one call site", sites_per_package %>% filter(n == 1)),
-  r("at least one call site", nrow(sites_per_package)),
-  r("packages no calls to eval", corpus_size - nrow(sites_per_package)),
-  r("nb median call sites", median(sites_per_package$n)),
-  r("max call sites", max_packages),
-)
-```
-
-`r TBL`
-
-```{r call-sites-violin-plot, include=F}
-E %>%
-  group_by(ev_package) %>%
-  mutate(sites = n_distinct(src_ref)) %>%
-  ggplot(aes(x = ev_variant, y = sites, fill = ev_variant), na.rm = TRUE) +
-  geom_violin(trim = FALSE, width = 1.7) +
-  # geom_jitter(shape=16, position=position_jitter(0.2)) +
-  labs(y = "Call sites per package") +
-  theme(
-    legend.position = c(0.5, 0.85),
-    legend.box.background = element_rect(fill = "white", size = 0.1),
-    legend.box = "horizontal",
-    axis.title.x = element_blank()
-  ) -> PLOT
-
-ggsave(path(PLOT_DIR, "call_sites_per_packages.pdf"))
-```
-
-How many call sites and how many packages did we exercise out of the total number of call sites of the corpus?
-
-```{r exercised-call-sites, include=F}
-static_sites <- nrow(P)
-
-static_call_sites_per_package <- P %>%
-  count(package, sort = TRUE)
-
-static_at_least_one_call_site <- static_call_sites_per_package %>%
-  filter(n >= 1)
-
-r("nb static call sites", static_sites)
-r("call sites percent", ratio(sites, static_sites))
-r("nb static at least one call site", nrow(static_at_least_one_call_site))
-r("static at least one call site percent", ratio(nrow(sites_per_package %>% filter(n >= 1)), nrow(static_at_least_one_call_site)))
+r("eval percent call site", ratio(trig_eval, static_eval))
+r("evalq percent call site", ratio(trig_evalq, static_evalq))
+r("eval parent percent call site", ratio(trig_eval.parent, static_eval.parent))
+r("local percent call site", ratio(trig_local, static_local))
 ```
@@
 # run = package?
+
+### TODO REMOVE
 eval_calls_per_package <- E %>%

### 2020-12-16 paper/main.tex
Table variant of evals
usage

@@
 Figure~\ref{freq} shows the number of \evals generated by each package
-over all runs that invoked functions in that package. Out of the \Corpus
-packages in the corpus, \EvalCallsLessHundred packages have a low frequency of \eval calls,
-with fewer than 100 of them. On the other hand, \EvalCallsMoreThousand packages call eval
-more than 1,000 times, with one package (\MaxEvalCallPackage) calling it \MaxEvalCallsRnd times.
+over all runs that invoked functions in that package. Out of the
+\Corpus packages in the corpus, \Fewcalls packages have a low
+frequency of \eval calls, with fewer than 100 of them. On the other
+hand, \Manycalls packages call eval more than 1,000 times, with one
+package (\Maxcallspack) calling it \Maxcalls times.
 

### 2020-12-16 analysis/usage.Rmd

@@
 
+```{r exercised-call-site-per-variant}
+static_eval <- P %>% filter(call_fun_name == "eval") %>% nrow()
+static_evalq <- P %>% filter(call_fun_name == "evalq") %>% nrow()
+static_eval.parent <- P %>% filter(call_fun_name == "eval.parent") %>% nrow()
+static_local <- P %>% filter(call_fun_name == "local") %>% nrow()
+
+sites_per_variant <- E %>% # src_ref indicates where the call originates from.
+  group_by(ev_variant) %>% 
+  summarize( n = n_distinct(src_ref)) # This the number of distinct sites
+
+trig_eval <- sites_per_variant %>% filter(ev_variant == "eval")
+trig_eval <- if(nrow(trig_eval) == 0)
+{
+  0
+} else
+{
+  trig_eval$n
+}
+
+trig_evalq <- sites_per_variant %>% filter(ev_variant == "evalq")
+trig_evalq <- if(nrow(trig_evalq) == 0)
+{
+  0
+} else
+{
+  trig_evalq$n
+}
+trig_eval.parent <- sites_per_variant %>% filter(ev_variant == "eval.parent") 
+trig_eval.parent <- if(nrow(trig_eval.parent) == 0)
+{
+  0
+} else
+{
+  trig_eval.parent$n
+}
+trig_local <- sites_per_variant %>% filter(ev_variant == "local")
+trig_local <- if(nrow(trig_local) == 0)
+{
+  0
+} else
+{
+  trig_local$n
+}
+
+overview_table(
+  r("StaticEval", static_eval),
+  r("StaticEvalq", static_evalq),
+  r("StaticEvalParent", static_eval.parent),
+  r("StaticLocal", static_local),
+  r("TriggeredEval", trig_eval),
+  r("TriggeredEvalq", trig_evalq),
+  r("Triggeredevalparent", trig_eval.parent),
+  r("Triggeredlocal", trig_local),
+  r("Triggeredevalpercent", ratio(trig_eval, static_eval)),
+  r("Triggeredevalqpercent", ratio(trig_evalq, static_evalq)),
+  r("Triggeredevalparentpercent", ratio(trig_eval.parent, static_eval.parent)),
+  r("Triggeredlocalpercent", ratio(trig_local, static_local))
+) -> TBL
+```
+
+`r TBL`
+
 # Ev-dist

### 2020-12-16 analysis/usage.Rmd
Formatting

--- 2020-12-16 analysis/usage.Rmd
@@
 overview_table(
-  r("Corpus",corpus_size),
+  r("Corpus", corpus_size),
   r("NbRuns", nb_runs)
@@
 ```{r exercised-call-site-per-variant}
-static_eval <- P %>% filter(call_fun_name == "eval") %>% nrow()
-static_evalq <- P %>% filter(call_fun_name == "evalq") %>% nrow()
-static_eval.parent <- P %>% filter(call_fun_name == "eval.parent") %>% nrow()
-static_local <- P %>% filter(call_fun_name == "local") %>% nrow()
+static_eval <- P %>%
+  filter(call_fun_name == "eval") %>%
+  nrow()
+static_evalq <- P %>%
+  filter(call_fun_name == "evalq") %>%
+  nrow()
+static_eval.parent <- P %>%
+  filter(call_fun_name == "eval.parent") %>%
+  nrow()
+static_local <- P %>%
+  filter(call_fun_name == "local") %>%
+  nrow()
 
@@
   group_by(ev_variant) %>%
-  summarize( n = n_distinct(src_ref)) # This the number of distinct sites
+  summarize(n = n_distinct(src_ref)) # This the number of distinct sites
 
 trig_eval <- sites_per_variant %>% filter(ev_variant == "eval")
-trig_eval <- if(nrow(trig_eval) == 0)
-{
+trig_eval <- if (nrow(trig_eval) == 0) {
   0
-} else
-{
+} else {
   trig_eval$n
@@
 trig_evalq <- sites_per_variant %>% filter(ev_variant == "evalq")
-trig_evalq <- if(nrow(trig_evalq) == 0)
-{
+trig_evalq <- if (nrow(trig_evalq) == 0) {
   0
-} else
-{
+} else {
   trig_evalq$n
@@
 trig_eval.parent <- sites_per_variant %>% filter(ev_variant == "eval.parent")
-trig_eval.parent <- if(nrow(trig_eval.parent) == 0)
-{
+trig_eval.parent <- if (nrow(trig_eval.parent) == 0) {
   0
-} else
-{
+} else {
   trig_eval.parent$n
@@
 trig_local <- sites_per_variant %>% filter(ev_variant == "local")
-trig_local <- if(nrow(trig_local) == 0)
-{
+trig_local <- if (nrow(trig_local) == 0) {
   0
-} else
-{
+} else {
   trig_local$n
@@
 eval_calls_table <- E %>%
-  count(ev_package, wt=duplicates, sort = TRUE)
-breaks <- c(0, 10^seq(1,8))
+  count(ev_package, wt = duplicates, sort = TRUE)
+breaks <- c(0, 10^seq(1, 8))
 group_tags <- cut(eval_calls_table$n,
   breaks = breaks,
-                  include.lowest = TRUE)
+  include.lowest = TRUE
+)
 ev_summary <- summary(group_tags)
@@
   r("bing", ev_summary[[7]]),
-r("binh", ev_summary[[8]])) -> TBL
+  r("binh", ev_summary[[8]])
+) -> TBL
 ```
@@
 
-more_1000_calls <- eval_calls_table %>% filter(n > 1000) %>% nrow()
+more_1000_calls <- eval_calls_table %>%
+  filter(n > 1000) %>%
+  nrow()
 

### 2020-12-16 paper/main.tex
Table variants

--- 2020-12-16 paper/main.tex
@@
 
+\begin{figure}[h]
+    \begin{tabular}{|r|r|r|r|r|}
+        \hline
+                       &\eval & \texttt{evalq} & \texttt{eval.parent} & \texttt{local} \\
+       Static    sites  & \Staticeval & \Staticevalq & \Staticevalparent & \Staticlocal \\
+       Exercised sites & \Triggeredeval & \Triggeredevalq & \Triggeredevalparent & \Triggeredlocal \\
+        Invocations & \EvalsRnd & \EvalqsRnd & \EparentsRnd & \LocalsRnd \\
+%        \% triggered & \Triggeredevalpercent & \Triggeredevalqpercent & \Triggeredevalparentpercent & \Triggeredlocalpercent \\
+        \hline
+    \end{tabular}
+    \caption{Variants of \eval} \label{tab:variantseval}
+\end{figure}
+
 \end{document}

### 2020-12-16 analysis/usage.Rmd

--- 2020-12-16 analysis/usage.Rmd
@@
 overview_table(
-  r("Corpus",corpus_size),
-  r("Nbruns", nb_runs)
+  r("Corpus", corpus_size),
+  r("NbRuns", nb_runs)
 ) 

### 2020-12-16 paper/main.tex
Text for table of variant of evals

@@
 
+Figure\ref{tab:variantseval} shows, per variant of \eval, the number of static eval call sites in the corpus, the number of call sites among them that are exercised during tracing, and how many invocations to \eval it results to.
+Most call sites and invocations use the main version of \eval, not its variants.
+
 \end{document}

### 2020-12-16 analysis/usage.Rmd
Plot explo
usage

--- 2020-12-16 analysis/usage.Rmd
@@
 
-```{r exercised-call-site-per-variant}
-static_eval <- P %>% filter(call_fun_name == "eval") %>% nrow()
-static_evalq <- P %>% filter(call_fun_name == "evalq") %>% nrow()
-static_eval.parent <- P %>% filter(call_fun_name == "eval.parent") %>% nrow()
-static_local <- P %>% filter(call_fun_name == "local") %>% nrow()
+# Exercised vs. Static
+
+```{r exercised-vs-static, include=FALSE}
+stat_evl <- P %>% filter(call_fun_name == "eval") %>% nrow()
+stat_evq <- P %>% filter(call_fun_name == "evalq") %>% nrow()
+stat_evp <- P %>% filter(call_fun_name == "eval.parent") %>% nrow()
+stat_loc <- P %>% filter(call_fun_name == "local") %>% nrow()
 
@@
   group_by(ev_variant) %>%
-  summarize( n = n_distinct(src_ref)) # This the number of distinct sites
+  summarize(n = n_distinct(src_ref)) # This the number of distinct sites
 
-trig_eval <- sites_per_variant %>% filter(ev_variant == "eval")
-trig_eval <- if(nrow(trig_eval) == 0)
-{
-  0
-} else
-{
-  trig_eval$n
-}
-
-trig_evalq <- sites_per_variant %>% filter(ev_variant == "evalq")
-trig_evalq <- if(nrow(trig_evalq) == 0)
-{
-  0
-} else
-{
-  trig_evalq$n
-}
-trig_eval.parent <- sites_per_variant %>% filter(ev_variant == "eval.parent") 
-trig_eval.parent <- if(nrow(trig_eval.parent) == 0)
-{
-  0
-} else
-{
-  trig_eval.parent$n
-}
-trig_local <- sites_per_variant %>% filter(ev_variant == "local")
-trig_local <- if(nrow(trig_local) == 0)
-{
-  0
-} else
-{
-  trig_local$n
-}
+tmp <- sites_per_variant %>% filter(ev_variant == "eval")
+dyn_evl <- if (nrow(tmp)==0) { 0 } else { tmp$n }
+tmp <- sites_per_variant %>% filter(ev_variant == "evalq")
+dyn_evq <- if (nrow(tmp) == 0) { 0 } else { tmp$n }
+tmp <- sites_per_variant %>% filter(ev_variant == "eval.parent") 
+dyn_evalp <- if(nrow(tmp) == 0) { 0 } else { tmp$n }
+tmp <- sites_per_variant %>% filter(ev_variant == "local")
+dyn_lol <- if(nrow(tmp) == 0) { 0 } else { tmp$n }
 
 overview_table(
-  r("StaticEval", static_eval),
-  r("StaticEvalq", static_evalq),
-  r("StaticEvalParent", static_eval.parent),
-  r("StaticLocal", static_local),
-  r("TriggeredEval", trig_eval),
-  r("TriggeredEvalq", trig_evalq),
-  r("Triggeredevalparent", trig_eval.parent),
-  r("Triggeredlocal", trig_local),
-  r("Triggeredevalpercent", ratio(trig_eval, static_eval)),
-  r("Triggeredevalqpercent", ratio(trig_evalq, static_evalq)),
-  r("Triggeredevalparentpercent", ratio(trig_eval.parent, static_eval.parent)),
-  r("Triggeredlocalpercent", ratio(trig_local, static_local))
+  r("Staticeval", stat_evl),
+  r("Staticevalq", stat_evq),
+  r("Staticevalparent", stat_evp),
+  r("Staticlocal", stat_loc),
+  r("Triggeredeval", dyn_evl),
+  r("Triggeredevalq", dyn_evq),
+  r("Triggeredevalparent", dyn_evp),
+  r("Triggeredlocal", dyn_loc),
+  r("Triggeredevalpct", ratio(dyn_evl, stat_evl)),
+  r("Triggeredevalqpct", ratio(dyn_evq, stat_evq)),
+  r("Triggeredevalparentpct", ratio(dyn_evp, stat_evp)),
+  r("Triggeredlocalpct", ratio(dyn_loc, stat_loc))
 ) -> TBL

### 2020-12-16 paper/main.tex

--- 2020-12-16 paper/main.tex
@@
         \hline
-                       &\eval & \texttt{evalq} & \texttt{eval.parent} & \texttt{local} \\
+                       &\eval & \c{evalq} & \c{eval} & \c{local} \\[-2mm]
+                       & & & \c{.parent} & \\
+                       \hline
        Static    sites  & \Staticeval & \Staticevalq & \Staticevalparent & \Staticlocal \\
@@
 \end{figure}
+
+Figure~\ref{tab:variantseval} shows, per variant of \eval, the number of static eval call sites in the corpus, the number of call sites among them that are exercised during tracing, and how many invocations to \eval it results to.
+Most call sites and invocations use the main version of \eval, not its variants.
 

### 2020-12-16 analysis/usage.Rmd

@@
 
+Here we compute the number of static call sistes to eval to compare it with
+the number of call sites with triggered dynamically. The former is larger
+than the latter because code coverage is not perfect.
+
 ```{r exercised-vs-static, include=FALSE}
@@
 tmp <- sites_per_variant %>% filter(ev_variant == "eval.parent") 
-dyn_evalp <- if(nrow(tmp) == 0) { 0 } else { tmp$n }
+dyn_evp <- if(nrow(tmp) == 0) { 0 } else { tmp$n }
 tmp <- sites_per_variant %>% filter(ev_variant == "local")
-dyn_lol <- if(nrow(tmp) == 0) { 0 } else { tmp$n }
+dyn_loc <- if(nrow(tmp) == 0) { 0 } else { tmp$n }
 

### 2020-12-16 analysis/usage.Rmd
Calls per run per call sites

--- 2020-12-16 analysis/usage.Rmd
@@
 
+```{r}
+eval_call_triggered %>% 
+  filter(average < 50) %>% 
+  ggplot(aes(x = average)) +
+  geom_histogram(binwidth =  1)
+
+ggsave(path(PLOT_DIR, "calls_per_run_per_call_site.pdf"))
+
+bins <- cut(eval_call_triggered$average, breaks = c(0, 50, 100, 250,500, 1000, 1500,  2000, 3000), include.lowest = TRUE)
+sum <- summary(bins)
+
+overview_table(
+  r("Runbina", sum[[1]]),
+  r("Runbinb", sum[[2]]),
+  r("Runbinc", sum[[3]]),
+  r("Runbind", sum[[4]]),
+  r("Runbine", sum[[5]]),
+  r("Runbinf", sum[[6]]),
+  r("Runbing", sum[[7]]),
+  r("Runbinh", sum[[8]])
+)
+```
+
 # The case of core libraries

### 2020-12-16 paper/main.tex
minor

--- 2020-12-16 paper/main.tex
@@
 
-\begin{figure}[h]
-    \begin{tabular}{|r|r|r|r|r|}
-        \hline
-                       &\eval & \texttt{evalq} & \texttt{eval.parent} & \texttt{local} \\
+\begin{figure}[h]\begin{tabular}{|r|r|r|r|r|}\hline
+   &\eval & \c{evalq} & \c{eval} & \c{local} \\[-2mm]
+           & & & \c{.parent} & \\\hline
 Static sites  & \Staticeval & \Staticevalq & \Staticevalparent & \Staticlocal \\
 Exercised sites & \Triggeredeval & \Triggeredevalq & \Triggeredevalparent & \Triggeredlocal \\
-        Invocations & \EvalsRnd & \EvalqsRnd & \EparentsRnd & \LocalsRnd \\
-%        \% triggered & \Triggeredevalpercent & \Triggeredevalqpercent & \Triggeredevalparentpercent & \Triggeredlocalpercent \\
-        \hline
-    \end{tabular}
-    \caption{Variants of \eval} \label{tab:variantseval}
+Invocations & \EvalsRnd & \EvalqsRnd & \EparentsRnd & \LocalsRnd \\\hline
+\end{tabular}\caption{Variants}\label{tab:variantseval}
 \end{figure}
 
+Figure~\ref{tab:variantseval} shows, per variant of \eval, the number
+of call sites in the corpus (\emph{static}), the number of call sites
+that were observed during tracing (\emph{exercised}), and the total
+number of dynamic calls (\emph{invocations}). The data clearly shows
+that the variants of \eval are rarely used, in particular \c{evalq}
+and \c{local}.
+
 \end{document}
-
-(\EvalsRnd for \eval, \EvalqsRnd for \c{evalq}, \EparentsRnd for
-\c{eval.parent} and \LocalsRnd for \c{local}).
-
-\NbPackageEvalAtLeastOnce packages in our corpus call \eval at least once.
-
-\mypara{Number of eval call sites.}  We observed that the number of
-\eval call sites is small.  Figure~\ref{fig:callSitesPerPackage} shows
-a violin plot of the distribution of call sites to \eval and its three
-variants. For packages that have call sites, the median is
-\NbMedianCallSites call sites. \NbOneCallSite{} packages have a single
-call site, and one package, \emph{\MaxCallSitesPackage}, has
-\MaxCallSites{} call sites. \PackagesNoCallsToEval{} packages have no
-calls sites to \eval. These numbers are under-approximations as \eval
-may be aliased, called reflectively and given as argument to a
-higher-order function such as \c{map}.
-
-\begin{figure}[h]
-	\centering
-	\includegraphics[width=\textwidth]{call_sites_per_packages}
-	\caption{Violin plot of the call sites per packages per eval function. } \label{fig:callSitesPerPackage}
-\end{figure}
 

### 2020-12-16 paper/main.tex
minor

@@
 
-\mypara{Number of calls to eval.} We observed an average \AverageEvalCallsPerPackage{} calls
-to \eval per run. \NOTE{this is in packages that had any call to eval.
-  How do we compute this? Say we are interested in package P. Do we
-  only count runs that call eval from package P. Or do we also count
-  runs that did not have an eval but executed some code from P? The
-  latter may be easier. Which one is right?} Figure~\ref{fig:evalCallsPerPackage} shows a
-violin plot of the distribution of calls to eval per run. The largest
-number of calls per run is \MaxEvalCalls{} and comes from the package \MaxEvalCallPackage.
-
-\begin{figure}[h]
-	\centering
-	\includegraphics[width=\textwidth]{eval_calls_per_packages}
-	\caption{Violin plot of the normalized number of eval calls per package. } \label{fig:evalCallsPerPackage}
-\end{figure}
-
 \mypara{Amount of code loaded by eval.} We use \c{deparse} to turn the

### 2020-12-16 paper/main.tex
alls per call site per run: table and histogram

--- 2020-12-16 paper/main.tex
@@
 
+\begin{figure}[h]
+    \centering
+    \begin{tabular}{|r@{\,}r@{\,}lr|r@{\,}r@{\,}lr|} \hline
+        \tt 0 &--& \tt 50      & \Runbina  & \tt 501 &--&\tt 1000  & \Runbine\\
+        \tt 51 &--& \tt 100    & \Runbinb  & \tt 1001 &--&\tt 1500  & \Runbinf\\
+        \tt 101 &--& \tt 250    & \Runbinc  & \tt 1501 &--&\tt 2000   & \Runbing\\
+        \tt 251 &--& \tt 500    & \Runbind  & \tt 2001 &--& \tt 3000 & \Runbinh\\\hline
+    \end{tabular}
+    \caption{Invocations of \eval per call sites normalized by the number of runs.} \label{tab:callsitesnormalized}
+\end{figure}
+
+Figure~\ref{tab:callsitesnormalized} shows the number of invocation to \eval and its variants per call sites per run exercising that call site.
+If that number is high, it can correspond to a function with an \eval call site that is used widely.
+If that number is very high, more than 1000, we expect it to correspond to an \eval call site in a loop.
+Most of the invocations are less than 50 per run per call site.
+When zooming in, on Figure~\ref{fig:callsitesnormalized}, most call sites have actually only one invocation per run.
+
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=\textwidth]{calls_per_run_per_call_site}
+    \caption{Invocations of \eval per call sites normalized by the number of runs, for less than 50 normalized invocations. Binsize is 1} \label{fig:callsitesnormalized}
+\end{figure}
+
 \end{document}

### 2020-12-16 analysis/usage.Rmd
minor

@@
 # The calls per package
-pack_calls <- E %>% 
-  count(ev_package, wt=duplicates, sort = TRUE)
+pack_calls <- E %>% count(ev_package, wt=duplicates, sort = TRUE)
 # Create bins at every power of ten
@@
 stopifnot(sum(is.na(E$src_ref)) == 0) # No NAs in scr_ref!
-
+# Number of eval call sites
 sites <- E %>% # src_ref indicates where the call originates from.
@@
   n_distinct() # This the number of distinct sites
-
+# Number of call sites per package
 sites_per_package <- E %>%
@@
   count(ev_package) # All packages have at least one site!
-
-max_packages <- sites_per_package %>%
-  filter(n == max(n)) %>%
-  .$n
 
@@
   r("packages no calls to eval", corpus_size - nrow(sites_per_package)),
-  r("nb median call sites", median(sites_per_package$n)),
-  r("max call sites", max_packages),
+  r("nb median call sites", median(sites_per_package$n))
 )
@@
 `r TBL`
-
-```{r call-sites-violin-plot, include=F}
-E %>%
-  group_by(ev_package) %>%
-  mutate(sites = n_distinct(src_ref)) %>%
-  ggplot(aes(x = ev_variant, y = sites, fill = ev_variant), na.rm = TRUE) +
-  geom_violin(trim = FALSE, width = 1.7) +
-  # geom_jitter(shape=16, position=position_jitter(0.2)) +
-  labs(y = "Call sites per package") +
-  theme(
-    legend.position = c(0.5, 0.85),
-    legend.box.background = element_rect(fill = "white", size = 0.1),
-    legend.box = "horizontal",
-    axis.title.x = element_blank()
-  ) -> PLOT
-
-ggsave(path(PLOT_DIR, "call_sites_per_packages.pdf"))
-```
 
@@
 
-This is the number of eval calls which have a call site in each package, divided by the number of runs for the package.
+This is the number of eval calls which have a call site in each package,
+divided by the number of runs for the package.
 
@@
 runs_package %>% datatable()
-```
-
-```{r nb-eval-calls}
-# run = package?
-
-### TODO REMOVE
-eval_calls_per_package <- E %>%
-  count(ev_package, wt = duplicates) %>%
-  left_join(runs_package, by = "ev_package")
-
-average_eval_calls <- eval_calls_per_package %>%
-  filter(n > 0) %>%
-  summarise(average = mean(n))
-max_eval_calls <- eval_calls_per_package %>% filter(n == max(n))
-
-r("average eval calls per package", average_eval_calls$average) # average on packages with at least one eval call
-r("max eval calls", max_eval_calls$n)
-r("max eval call package", max_eval_calls$ev_package)
 ```

### 2020-12-16 analysis/usage.Rmd
minor

@@
 
-```{r variants,echo=F}
+```{r variants, include=FALSE}
 nb_calls <- E %>%
@@
 
-Here we compute the number of static call sistes to eval to compare it with
+Here we compute the number of static call sites to compare it with
 the number of call sites with triggered dynamically. The former is larger
@@
 
-TBL <- overview_table(
-  # Sites triggered over the whole corpus:
-  r("nb eval call sites", sites),
-  # How many packages in the corpus have a single triggered call site:
-  r("nb one call site", sites_per_package %>% filter(n == 1)),
-  r("at least one call site", nrow(sites_per_package)),
-  r("packages no calls to eval", corpus_size - nrow(sites_per_package)),
-  r("nb median call sites", median(sites_per_package$n))
-)
-```
-
-`r TBL`
-
-How many call sites and how many packages did we exercise out of the total number of call sites of the corpus?
-
-```{r exercised-call-sites, include=F}
-
-# TODO :: remove?
 static_sites <- nrow(P)
@@
 
-r("nb static call sites", static_sites)
-r("call sites percent", ratio(sites, static_sites))
-r("nb static at least one call site", nrow(static_at_least_one_call_site))
-r("static at least one call site percent", ratio(nrow(sites_per_package %>% filter(n >= 1)), nrow(static_at_least_one_call_site)))
+overview_table(
+  r("Sites", sites),
+  r("Staticsites", static_sites),
+  r("Sitespercent", ratio(sites, static_sites)),
+  r("Staticatleastonecallsite", nrow(static_at_least_one_call_site))
+)
 ```
@@
 We first compute an approximation of the number of runs per package.
-```{r runs-per-package}
+```{r runs-per-package, include=FALSE}
 runs_package <- E %>%
@@
 weighted.median <- function(x, w, na.rm = FALSE) {
-  if (na.rm) {
-    na.omit(x)
-  }
+  if (na.rm) { na.omit(x) }
   w <- w[order(x)]
   x <- x[order(x)]
-
   prob <- cumsum(w) / sum(w)
@@
   arrange(desc(average))
-```
 
-```{r}
 eval_call_triggered %>% 
@@
 
-# The case of core libraries
-
 ```{r}

### 2020-12-16 paper/main.tex
minor

@@
 were obtained by \Nbruns runs, with runs that did not trigger \eval
-discarded. Out of the \NbStaticAtLeastOneCallSite CRAN packages(among the \Corpus packages of our corpus) with \eval call sites, we
-observed calls in \Triggeredpkgs packages, the difference can be chalk
-down to incomplete code coverage and occasional failures due to our
-pipeline.
+discarded. Out of the \Staticatleastonecallsite CRAN packages(among
+the \Corpus packages of our corpus) with \eval call sites, we observed
+calls in \Triggeredpkgs packages, the difference can be chalk down to
+incomplete code coverage and occasional failures due to our pipeline.
 
@@
 
+Figure~\ref{tab:variantseval} shows, per variant of \eval, the number
+of call sites in the corpus (\emph{static}), the number of call sites
+that were observed during tracing (\emph{exercised}), and the total
+number of dynamic calls (\emph{invocations}). The data clearly shows
+that the variants of \eval are rarely used, in particular \c{evalq}
+and \c{local}.
+
 \begin{figure}[h]\begin{tabular}{|r|r|r|r|r|}\hline
@@
 
-Figure~\ref{tab:variantseval} shows, per variant of \eval, the number
-of call sites in the corpus (\emph{static}), the number of call sites
-that were observed during tracing (\emph{exercised}), and the total
-number of dynamic calls (\emph{invocations}). The data clearly shows
-that the variants of \eval are rarely used, in particular \c{evalq}
-and \c{local}.
-
-\begin{figure}[h]
-    \centering
+\begin{figure}[h] \centering
 \begin{tabular}{|r@{\,}r@{\,}lr|r@{\,}r@{\,}lr|} \hline
@@
 \end{tabular}
-    \caption{Invocations of \eval per call sites normalized by the number of runs.} \label{tab:callsitesnormalized}
+\caption{Normalized invocations} \label{tab:callsitesnormalized}
 \end{figure}
 
-Figure~\ref{tab:callsitesnormalized} shows the number of invocation to \eval and its variants per call sites per run exercising that call site.
-If that number is high, it can correspond to a function with an \eval call site that is used widely.
-If that number is very high, more than 1000, we expect it to correspond to an \eval call site in a loop.
-Most of the invocations are less than 50 per run per call site.
-When zooming in, on Figure~\ref{fig:callsitesnormalized}, most call sites have actually only one invocation per run.
+Figure~\ref{tab:callsitesnormalized} shows the number of invocation to
+\eval and its variants per call sites per run exercising that call
+site. If that number is high, it can correspond to a function with an
+\eval call site that is used widely. If that number is very high, more
+than 1000, we expect it to correspond to an \eval call site in a loop.
+Most of the invocations are less than 50 per run per call site. When
+zooming in, on Figure~\ref{fig:callsitesnormalized}, most call sites
+have actually only one invocation per run.
 

### 2020-12-17 analysis/usage.Rmd
some progress

@@
   ev_package = eval_source, #   Package of the call site
-  arg_size = expr_expression_length, # Size in bytes of eval argument string
+  arg_size = expr_resolved_length, #   Size in bytes of eval argument string
   opcodes = direct_interpreter_eval #  Count of the AST nodes traversed by the
@@
 ```{r eval-calls-violin-plot}
-eval_calls_per_package %>%
+sites_per_package %>%
   ggplot(aes(x = "", y = n, fill = ""), na.rm = TRUE) +
@@
 
-# Amount of code loaded by eval
+# Amount of code
+
+How much code is loaded by eval, we look at argument sizes.
 
@@
 
-r("median size eval", med_size)
-r("max size eval", max_size)
-r("average size eval", average_size)
-r("size nine five", nb_nodes_many$arg_size)
+overview_table(
+  r("median size eval", med_size),
+  r("max size eval", max_size),
+  r("average size eval", average_size),
+  r("size nine five", nb_nodes_many$arg_size),
   r("size nine five exact percent", percent(nb_nodes_many$cumulative))
+) -> TBL
+
+E %>% filter(is.na(arg_size) == FALSE) %>%
+  filter(arg_size > 100) %>%
+  ggplot(aes(x=arg_size)) + geom_histogram(binwidth=1) 
 ```
+
+`r TBL`
 

### 2020-12-17 paper/main.tex
some progress

@@
 
-%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Usage Metrics}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 \begin{figure}[h]\begin{tabular}{|r|r|r|r|r|}\hline
-   &\eval & \c{evalq} & \c{eval} & \c{local} \\[-2mm]
-           & & & \c{.parent} & \\\hline
-Static sites & \Staticeval & \Staticevalq & \Staticevalparent & \Staticlocal \\
-Exercised sites & \Triggeredeval & \Triggeredevalq & \Triggeredevalparent & \Triggeredlocal \\
-Invocations & \EvalsRnd & \EvalqsRnd & \EparentsRnd & \LocalsRnd \\\hline
+  &\eval & \c{evalq} & \c{eval} & \c{local}\\[-2mm]
+           & & & \c{.parent} &\\\hline
+Static sites &\Staticeval&\Staticevalq&\Staticevalparent&\Staticlocal \\
+Exercised sites&\Triggeredeval&\Triggeredevalq&\Triggeredevalparent&\Triggeredlocal\\
+Invocations&\EvalsRnd&\EvalqsRnd&\EparentsRnd&\LocalsRnd\\\hline
 \end{tabular}\caption{Variants}\label{tab:variantseval}
@@
 
-\begin{figure}[h]
-    \centering
+\begin{figure}[h] \centering
     \includegraphics[width=\textwidth]{calls_per_run_per_call_site}
-    \caption{Invocations of \eval per call sites normalized by the number of runs, for less than 50 normalized invocations. Binsize is 1} \label{fig:callsitesnormalized}
+    \caption{Invocations of \eval per call sites normalized by the
+      number of runs, for less than 50 normalized invocations. Binsize
+      is 1} \label{fig:callsitesnormalized}
 \end{figure}
-
-\end{document}
 
@@
 input to \eval into strings and measure the size of that input. The
-largest input is \MaxSizeEval and the median input size is \MedianSizeEval. Most
-inputs are small, less than \SizeNineFiveRnd bytes. The average size is \AverageSizeEval
-bytes.
-
-\mypara{Amount of computation via eval.}
-We compare the number of events intercepted during execution of \eval
-across the whole corpus. For this, we use the number of interpreter
-steps as a proxy for events. We observe \AllEventCountRnd{} \NOTE{Check that number.\quad} events all
-told, and \NbEventsEvalRnd events occur inside eval
-(\EvalEventAllPerc of all events).
-Figure~\ref{fig:eventsPerPackage} show a violin plot of events.
+largest input is \MaxSizeEval and the median input size is
+\MedianSizeEval. Most inputs are small, less than \SizeNineFiveRnd
+bytes. The average size is \AverageSizeEval bytes.
+
+\mypara{Amount of computation via eval.} We compare the number of
+events intercepted during execution of \eval across the whole corpus.
+For this, we use the number of interpreter steps as a proxy for
+events. We observe \AllEventCountRnd{} \NOTE{Check that number.\quad}
+events all told, and \NbEventsEvalRnd events occur inside eval
+(\EvalEventAllPerc of all events). Figure~\ref{fig:eventsPerPackage}
+show a violin plot of events.
 
@@
 \eval calls. The largest number of events generated by an \eval call
-is \MaxiEventsEvalRnd. However, a majority (\SmallEventsEval) of \eval calls performs up
-to 100 events suggesting that most expressions passed to
-\eval are small. It is interesting to note that a higher proportion of
-Core R \eval calls perform side effects in the
+is \MaxiEventsEvalRnd. However, a majority (\SmallEventsEval) of \eval
+calls performs up to 100 events suggesting that most expressions
+passed to \eval are small. It is interesting to note that a higher
+proportion of Core R \eval calls perform side effects in the
 \EventsMinRangeB--\EventsMaxRangeB range. Only \EventsCoreEvalCountC
@@
 \end{figure}
-
-\mypara{The case of CORE libraries} ... \NOTE{Give numbers for the
-  same things as above but for the base/core libraries} Compare.
-\NOTE{Add XXX016 to XXX025.}
 

### 2020-12-17 analysis/usage.Rmd
usage

@@
 # data again from disk. It assumes the data doesn't change.
-if (exists(eval_calls_path) == FALSE) {
+if (exists("E") == FALSE) {
   read_fst(eval_calls_path) %>% as_tibble() -> E_raw
@@
 
-E %>% filter(is.na(arg_size) == FALSE) %>%
-  filter(arg_size > 100) %>%
-  ggplot(aes(x=arg_size)) + geom_histogram(binwidth=1) 
+sizes_prop %>%
+  ggplot(aes(x=arg_size)) + geom_histogram()
+
 ```
@@
 
-# Amount of computations via eval
+# Amount of computation
+
+How much work is involved in each eval as measured by the number of ??? 
+(possible: opcodes, AST nodes).
 
 ```{r amount_computations, include=F}
-nb_events_in_eval <- E %>% summarise(nb_events = sum(opcodes))
+tot_events <- E %>% summarise(n=sum(opcodes)) %>% .$n
+max_events <- E %>% summarize(max = max(opcodes)) %>% .$n
 
-max_events_in_eval <- E %>% summarize(maxi = max(opcodes))
+less_50_events <- E %>% filter(opcodes <= 50) %>%
+  count(wt = duplicates) %>%
+  .$n
 
-less_100_events <- E %>% filter(opcodes <= 100)
-nb_less_100_events <- count(less_100_events, wt = duplicates)$n
+overview_table(
+  r("Events", tot_events),
+  r("Maxevents", max_events),
+  r("Smalleventspct", ratio(less_50_events, nb_calls))
+) -> TBL
 
-r("nb events eval", nb_events_in_eval$nb_events)
-r("maxi events eval", max_events_in_eval$maxi)
-r("small events eval", ratio(nb_less_100_events, nb_calls))
-```
-
-```{r nb-events-violin-plot,include=F}
 E %>%
-  filter(opcodes > 0) %>%
+  filter(opcodes > 50) %>%
   ggplot(aes(x = "", y = opcodes, fill = ""), na.rm = TRUE) +
@@
   scale_y_log10(labels = scales::comma) +
-  labs(y = "Number of events in eval per package") +
+  labs(y = "", x="") +
   theme(legend.position = "none")
 
-ggsave(path(PLOT_DIR, "nb_events_per_packages.pdf"))
+ggsave(path(PLOT_DIR, "events_per_pack_large.pdf"))
+
+E %>%
+  filter(opcodes > 0) %>%
+  filter(opcodes < 50) %>%
+  ggplot(aes(x = "", y = opcodes, fill = ""), na.rm = TRUE) +
+  geom_violin(trim = FALSE, width = 1.7) +
+  labs(y = "Events per package", x="") +
+  theme(legend.position = "none")
+
+ggsave(path(PLOT_DIR, "events_per_pack_small.pdf"))
 ```
+
+`r TBL`
 
@@
 
-Here, we try to have an estimation of how often a call site is exercised per run. If it is often, maybe it is a loop!
+Here, we try to have an estimation of how often a call site is exercised per 
+run. If it is often, maybe it is a loop!
 
-```{r loops}
-eval_calls_exercised <- E %>%
-  group_by(file) %>%
-  count(src_ref, wt = duplicates)
+```{r loops, include=FALSE}
+# Which call sites where exercised and how many times, per run.
+exercised <- E %>%
+  group_by(file) %>%               # group by runs
+  count(src_ref, wt = duplicates)  # weigh by duplicates
 
-eval_calls_exercised %>% ggplot(aes(n)) +
+exercised %>% ggplot(aes(n)) +
   geom_histogram() +
   scale_x_log10(labels = scales::comma) +
-  labs(x = "Number of eval calls per srcref per run")
-```
+  labs(x = "Calls per site per run")
 
-```{r}
-eval_call_triggered <- E %>%
+# For each site, how many runs and within each runs how often do we trigger.
+triggered <- E %>%
   group_by(src_ref) %>%
-  summarize(n_runs = n_distinct(file), n = sum(duplicates), average = n / n_runs) %>%
+  summarize(n_runs = n_distinct(file), 
+            n = sum(duplicates), 
+            average = n / n_runs) %>%
   arrange(desc(average))
 
-eval_call_triggered %>% 
+# Plot the low trigger counts...
+triggered %>% 
   filter(average < 50) %>% 
@@
 
-bins <- cut(eval_call_triggered$average, breaks = c(0, 50, 100, 250,500, 1000, 1500,  2000, 3000), include.lowest = TRUE)
+bins <- cut(triggered$average, breaks = c(0, 50, 100, 250,500, 1000, 1500,  2000, 3000), include.lowest = TRUE)
 sum <- summary(bins)
@@
 
-```{r}
+`r TBL`
+
+```{r include=FALSE}
 duration <- difftime(Sys.time(), now)

### 2020-12-17 paper/main.tex
usage

@@
 
-. We turn off the
-bytecode compiler for this study. The bytecode compiler can also call
-\eval. We do not get source locations for \UndefinedEvalsRnd \eval
-calls. In these cases \eval is either passed as an argument to a
-higher-order functions or is defined in a function returned by a
-higher-order function and the R parser does not retain location
-information for \eval. However, this is a meager \PercentUndefinedEval
-of all \eval calls and is unlikely to affect our analysis. We ignore
-calls to the native \eval function exposed by R. We also ignore the
-\c{rlang::tidy_eval} function which uses native \eval internally
-because \c{rlang} is used to implement a DSL for data analysis in R.
-It introduces a new first-class promise object called \c{quosure} for
-which it implements special evaluation support in \c{tidy_eval}.
+We turn off the bytecode compiler for this study. The bytecode
+compiler can also call \eval. We do not get source locations for
+\UndefinedEvalsRnd \eval calls. In these cases \eval is either passed
+as an argument to a higher-order functions or is defined in a function
+returned by a higher-order function and the R parser does not retain
+location information for \eval. However, this is a meager
+\PercentUndefinedEval of all \eval calls and is unlikely to affect our
+analysis. We ignore calls to the native \eval function exposed by R.
+We also ignore the \c{rlang::tidy_eval} function which uses native
+\eval internally because \c{rlang} is used to implement a DSL for data
+analysis in R. It introduces a new first-class promise object called
+\c{quosure} for which it implements special evaluation support in
+\c{tidy_eval}.
 
@@
 \begin{figure}[h] \centering
-    \includegraphics[width=\textwidth]{calls_per_run_per_call_site}
+    \includegraphics[width=.5\textwidth]{calls_per_run_per_call_site}
     \caption{Invocations of \eval per call sites normalized by the
@@
 For this, we use the number of interpreter steps as a proxy for
-events. We observe \AllEventCountRnd{} \NOTE{Check that number.\quad}
-events all told, and \NbEventsEvalRnd events occur inside eval
+events. We observe \AllEventCountRnd \NOTE{Check that number.}
+events all told, and \EventsRnd events occur inside eval
 (\EvalEventAllPerc of all events). Figure~\ref{fig:eventsPerPackage}
@@
 	\centering
-	\includegraphics[width=\textwidth]{nb_events_per_packages}
-	\caption{Violin plot of the number of events per packages (log scale). } \label{fig:eventsPerPackage}
+	\includegraphics[width=.4\textwidth]{events_per_pack_small}
+	\includegraphics[width=.4\textwidth]{events_per_pack_large}
+	\caption{Events per packages} \label{fig:eventsPerPackage}
 \end{figure}
@@
 \eval calls. The largest number of events generated by an \eval call
-is \MaxiEventsEvalRnd. However, a majority (\SmallEventsEval) of \eval
-calls performs up to 100 events suggesting that most expressions
-passed to \eval are small. It is interesting to note that a higher
-proportion of Core R \eval calls perform side effects in the
+is \MaxeventsRnd. However, a majority (\Smalleventspct) of \eval calls
+performs up to 50 events suggesting that most expressions passed to
+\eval are small. It is interesting to note that a higher proportion of
+Core R \eval calls perform side effects in the
 \EventsMinRangeB--\EventsMaxRangeB range. Only \EventsCoreEvalCountC

### 2020-12-17 analysis/usage.Rmd
progress

@@
 overview_table(
-  r("median size eval", med_size),
-  r("max size eval", max_size),
-  r("average size eval", average_size),
-  r("size nine five", nb_nodes_many$arg_size),
-  r("size nine five exact percent", percent(nb_nodes_many$cumulative))
+  r("Medianszeval", med_size),
+  r("Maxszeval", max_size),
+  r("Avgszeval", average_size),
+  r("Szninefive", nb_nodes_many$arg_size)
   ) -> TBL
@@
 tot_events <- E %>% summarise(n=sum(opcodes)) %>% .$n
-max_events <- E %>% summarize(max = max(opcodes)) %>% .$n
+max_events <- E %>% summarize(n = max(opcodes)) %>% .$n
 

### 2020-12-17 paper/main.tex
progress

@@
 
-To provide a picture of \eval usage in R, we recorded \Allcalls
-invocations of \eval and its variants in the CRAN corpus. These calls
-were obtained by \Nbruns runs, with runs that did not trigger \eval
-discarded. Out of the \Staticatleastonecallsite CRAN packages(among
-the \Corpus packages of our corpus) with \eval call sites, we observed
-calls in \Triggeredpkgs packages, the difference can be chalk down to
-incomplete code coverage and occasional failures due to our pipeline.
+To provide a picture of \eval usage in R, this section focuses on
+library code and the \Corpus packages in the CRAN corpus. Data was
+collected from \Nbruns runs of code extracted from the target packages
+and their reverse dependencies. In total, \Allcalls invocations of
+\eval and its variants were recorded. Runs that did not call \eval
+were discarded.\footnote{The fact that a package contains call sites
+to \eval does not mean every run will execute those call sites.} Out
+of the \Staticatleastonecallsite packages that use \eval,
+\Triggeredpkgs packages we recorded calling the function --- the
+difference can be chalk down to incomplete code coverage and
+occasional failures in our pipeline.
 
@@
 input to \eval into strings and measure the size of that input. The
-largest input is \MaxSizeEval and the median input size is
-\MedianSizeEval. Most inputs are small, less than \SizeNineFiveRnd
-bytes. The average size is \AverageSizeEval bytes.
+largest input is \Maxszeval and the median input size is
+\Medianszeval. Most inputs are small, less than \SzninefiveRnd
+bytes. The average size is \Avgszeval bytes.
 

### 2020-12-17 analysis/usage.Rmd
here

@@
 
-nb_nodes_many <- sizes_prop %>%
+many <- sizes_prop %>%
   filter(cumulative >= 0.95) %>%
@@
   r("Avgszeval", average_size),
-  r("Szninefive", nb_nodes_many$arg_size)
+  r("Szninefive", many$arg_size)
   ) -> TBL

### 2020-12-17 paper/main.tex
main

@@
 To provide a picture of \eval usage in R, this section focuses on
-library code and the \Corpus packages in the CRAN corpus. Data was
-collected from \Nbruns runs of code extracted from the target packages
-and their reverse dependencies. In total, \Allcalls invocations of
-\eval and its variants were recorded. Runs that did not call \eval
-were discarded.\footnote{The fact that a package contains call sites
-to \eval does not mean every run will execute those call sites.} Out
-of the \Staticatleastonecallsite packages that use \eval,
-\Triggeredpkgs packages we recorded calling the function --- the
-difference can be chalk down to incomplete code coverage and
-occasional failures in our pipeline.
+library code and the \Corpus packages in our CRAN corpus. For these
+packages, all runnable code in CRAN was executed, runs that did not
+exercise an \eval in the corpus were discarded. The data presented
+here was collected from \Nbruns runs. In total, \Allcalls invocations
+of \eval and its variants were recorded. Out of the
+\Staticatleastonecallsite packages that use \eval, \Triggeredpkgs
+packages were observed calling it --- the difference can be chalked
+down to incomplete code coverage and occasional infrastructure
+failures.
 
 \begin{wrapfigure}{r}{5.2cm}
-\begin{tabular}{|r@{\,}r@{\,}lr|r@{\,}r@{\,}lr|} \hline
+  \begin{tabular}{|r@{\,}r@{\,}l@{}r|r@{\,}r@{\,}l@{}r|} \hline
+    \multicolumn{3}{|c}{\small\#calls} &\small \#pck
+&     \multicolumn{3}{c}{\small\#calls} &\small\#pck \\\hline
 \tt 0 &--& \tt 10      & \Bina  & \tt 1K &--&\tt 100K  & \Bine\\
@@
 
-Figure~\ref{freq} shows the number of \evals generated by each package
-over all runs that invoked functions in that package. Out of the
-\Corpus packages in the corpus, \Fewcalls packages have a low
-frequency of \eval calls, with fewer than 100 of them. On the other
-hand, \Manycalls packages call eval more than 1,000 times, with one
-package (\Maxcallspack) calling it \Maxcalls times.
-
-Figure~\ref{tab:variantseval} shows, per variant of \eval, the number
-of call sites in the corpus (\emph{static}), the number of call sites
-that were observed during tracing (\emph{exercised}), and the total
-number of dynamic calls (\emph{invocations}). The data clearly shows
-that the variants of \eval are rarely used, in particular \c{evalq}
-and \c{local}.
-
-\begin{figure}[h]\begin{tabular}{|r|r|r|r|r|}\hline
+Figure~\ref{freq} summarizes the frequency of calls to \eval over all
+runs. The left column is the number of observed calls, the right
+column is the number of packages falling in that range. There are
+\Fewcalls packages with low \eval frequency, i.e. fewer than 100
+calls; and a similar number of packages, \Manycalls to be precise, use
+\eval more than 1,000 times. One package, \Maxcallspack, calls it
+\Maxcalls times and thus accounts for over half of the observed
+invocations of \eval.
+
+\begin{wrapfigure}{r}{7cm}
+  \begin{tabular}{|@{\,}r@{\,}|@{\,}rrrr@{\,}|}\hline
   &\eval & \c{evalq} & \c{eval} & \c{local}\\[-2mm]
            & & & \c{.parent} &\\\hline
-Static sites &\Staticeval&\Staticevalq&\Staticevalparent&\Staticlocal \\
-Exercised sites&\Triggeredeval&\Triggeredevalq&\Triggeredevalparent&\Triggeredlocal\\
-Invocations&\EvalsRnd&\EvalqsRnd&\EparentsRnd&\LocalsRnd\\\hline
+\small Static sites &\Staticeval&\Staticevalq&\Staticevalparent&\Staticlocal \\
+\small Exercised sites&\Triggeredeval&\Triggeredevalq&\Triggeredevalparent&\Triggeredlocal\\
+\small Invocations&\EvalsRnd&\EvalqsRnd&\EparentsRnd&\LocalsRnd\\\hline
 \end{tabular}\caption{Variants}\label{tab:variantseval}
-\end{figure}
-
-\begin{figure}[h] \centering
+\end{wrapfigure}
+
+Figure~\ref{tab:variantseval} illustrates the use of variants of
+\eval. For each of the four variants, the first rows shows the number
+of call sites occurring in the corpus (\emph{static)}), the second row
+is the number of call sites that were encountered during tracing
+(\emph{exercised}), the last row is the number of calls
+(\emph{invocations}). The vast majority of call sites, and calls, are
+to \eval itself; \c{evalq} and \c{local} are barely used at all.
+
+\begin{wrapfigure}{r}{5.2cm}
 \begin{tabular}{|r@{\,}r@{\,}lr|r@{\,}r@{\,}lr|} \hline
@@
 \caption{Normalized invocations} \label{tab:callsitesnormalized}
-\end{figure}
-
-Figure~\ref{tab:callsitesnormalized} shows the number of invocation to
-\eval and its variants per call sites per run exercising that call
-site. If that number is high, it can correspond to a function with an
-\eval call site that is used widely. If that number is very high, more
-than 1000, we expect it to correspond to an \eval call site in a loop.
-Most of the invocations are less than 50 per run per call site. When
-zooming in, on Figure~\ref{fig:callsitesnormalized}, most call sites
-have actually only one invocation per run.
-
-\begin{figure}[h] \centering
-    \includegraphics[width=.5\textwidth]{calls_per_run_per_call_site}
-    \caption{Invocations of \eval per call sites normalized by the
-      number of runs, for less than 50 normalized invocations. Binsize
-      is 1} \label{fig:callsitesnormalized}
-\end{figure}
+\end{wrapfigure}
+
+Figure~\ref{tab:callsitesnormalized} shows normalized invocation
+counts. The column on the left is the average number of invocations of
+\eval from a particular call site for any given run. The number on the
+right is number of call sites that fall in that range. For instance,
+there are \Runbinh call sites that are invoked on average over 2,001
+times in any run they are exercised.
+
+\begin{wrapfigure}{r}{5.2cm}
+  \vspace{-8mm}
+    \includegraphics[width=.4\textwidth]{calls_per_run_per_call_site}
+    \caption{Normalized invocations
+      (small)} \label{fig:callsitesnormalized}
+    \vspace{-3mm}
+\end{wrapfigure}
+
+The majority of call sites, \Runbina, are exercised up to 50 times.
+Larger numbers of calls suggest that \eval may be called within a loop
+or from a recursive context -- this seems to be the exception.
+Figure~\ref{fig:callsitesnormalized} zooms in on call sites with an
+average of fewer than 51 invocation. The x-axis shows the number of
+invocations and the y-axis is the number of call sites. Most call
+sites are invoked once, about half as many are invoked twice. After
+that the frequency quickly drops.
 

### 2020-12-17 analysis/usage.Rmd
Density of amount of computations. Ranking per type

@@
   ev_package = eval_source, #   Package of the call site
-  arg_size = expr_resolved_length, #   Size in bytes of eval argument string
+  arg_size = expr_resolved_length, #   Size in characters (unicode) of eval argument string
+  type = expr_resolved_type_tag,
   opcodes = direct_interpreter_eval #  Count of the AST nodes traversed by the
@@
 ```{r exercised-vs-static, include=FALSE}
-stat_evl <- P %>% filter(call_fun_name == "eval") %>% nrow()
-stat_evq <- P %>% filter(call_fun_name == "evalq") %>% nrow()
-stat_evp <- P %>% filter(call_fun_name == "eval.parent") %>% nrow()
-stat_loc <- P %>% filter(call_fun_name == "local") %>% nrow()
+stat_evl <- P %>%
+  filter(call_fun_name == "eval") %>%
+  nrow()
+stat_evq <- P %>%
+  filter(call_fun_name == "evalq") %>%
+  nrow()
+stat_evp <- P %>%
+  filter(call_fun_name == "eval.parent") %>%
+  nrow()
+stat_loc <- P %>%
+  filter(call_fun_name == "local") %>%
+  nrow()
 
@@
 tmp <- sites_per_variant %>% filter(ev_variant == "eval")
-dyn_evl <- if (nrow(tmp)==0) { 0 } else { tmp$n }
+dyn_evl <- if (nrow(tmp) == 0) {
+  0
+} else {
+  tmp$n
+}
 tmp <- sites_per_variant %>% filter(ev_variant == "evalq")
-dyn_evq <- if (nrow(tmp) == 0) { 0 } else { tmp$n }
+dyn_evq <- if (nrow(tmp) == 0) {
+  0
+} else {
+  tmp$n
+}
 tmp <- sites_per_variant %>% filter(ev_variant == "eval.parent")
-dyn_evp <- if(nrow(tmp) == 0) { 0 } else { tmp$n }
+dyn_evp <- if (nrow(tmp) == 0) {
+  0
+} else {
+  tmp$n
+}
 tmp <- sites_per_variant %>% filter(ev_variant == "local")
-dyn_loc <- if(nrow(tmp) == 0) { 0 } else { tmp$n }
+dyn_loc <- if (nrow(tmp) == 0) {
+  0
+} else {
+  tmp$n
+}
 
@@
 # The calls per package
-pack_calls <- E %>% count(ev_package, wt=duplicates, sort = TRUE)
+pack_calls <- E %>% count(ev_package, wt = duplicates, sort = TRUE)
 # Create bins at every power of ten
-bins <- cut(pack_calls$n, breaks = c(0, 10^seq(1,8)), include.lowest = TRUE)
+bins <- cut(pack_calls$n, breaks = c(0, 10^seq(1, 8)), include.lowest = TRUE)
 sum <- summary(bins)
 # Packages with more than 1K calls
-Kcalls <- pack_calls %>% filter(n > 1000) %>% nrow()
+Kcalls <- pack_calls %>%
+  filter(n > 1000) %>%
+  nrow()
 
@@
 ecpp <- pack_calls %>% left_join(runs, by = "ev_package")
-stopifnot( nrow(filter(ecpp, n==0)) == 0)
+stopifnot(nrow(filter(ecpp, n == 0)) == 0)
 # The package with the most calls
@@
 weighted.median <- function(x, w, na.rm = FALSE) {
-  if (na.rm) { na.omit(x) }
+  if (na.rm) {
+    na.omit(x)
+  }
   w <- w[order(x)]
@@
 sizes_prop %>%
-  ggplot(aes(x=arg_size)) + geom_histogram()
+  ggplot(aes(x = arg_size)) +
+  geom_histogram()
 
+E %>%
+  ggplot(aes(x = arg_size)) +
+  stat_density(aes(y = ..count..), color = "black", fill = "blue", alpha = 0.3) +
+  scale_x_continuous(breaks = c(0, 1, 2, 3, 4, 5, 10, 30, 100, 300, 1000), trans = "log1p", expand = c(0, 0)) +
+  labs(x = "Size (characters)")
+
+ggsave(path(PLOT_DIR, "size_loaded_distribution.pdf"))
 ```
@@
 ```{r amount_computations, include=F}
-tot_events <- E %>% summarise(n=sum(opcodes)) %>% .$n
-max_events <- E %>% summarize(n = max(opcodes)) %>% .$n
+tot_events <- E %>%
+  summarise(n = sum(opcodes * duplicates)) %>%
+  .$n
+max_events <- E %>%
+  summarize(n = max(opcodes)) %>%
+  .$n
 
-less_50_events <- E %>% filter(opcodes <= 50) %>%
+less_50_events <- E %>%
+  filter(opcodes <= 50) %>%
   count(wt = duplicates) %>%
@@
   scale_y_log10(labels = scales::comma) +
-  labs(y = "", x="") +
+  labs(y = "", x = "") +
   theme(legend.position = "none")
@@
   geom_violin(trim = FALSE, width = 1.7) +
-  labs(y = "Events per package", x="") +
+  labs(y = "Events per package", x = "") +
   theme(legend.position = "none")
@@
   group_by(src_ref) %>%
-  summarize(n_runs = n_distinct(file), 
+  summarize(
+    n_runs = n_distinct(file),
     n = sum(duplicates),
-            average = n / n_runs) %>%
+    average = n / n_runs
+  ) %>%
   arrange(desc(average))
@@
 
-bins <- cut(triggered$average, breaks = c(0, 50, 100, 250,500, 1000, 1500,  2000, 3000), include.lowest = TRUE)
+bins <- cut(triggered$average, breaks = c(0, 50, 100, 250, 500, 1000, 1500, 2000, 3000), include.lowest = TRUE)
 sum <- summary(bins)
@@
 
+How diverse are the expressions passed to `eval`? We consider the diversity in types in the resolved expression.
+
+```{r type-diversity}
+E %>%
+  ggplot(aes(x = fct_rev(fct_infreq(type)))) +
+  geom_bar() +
+  coord_flip() +
+  labs(x = "type")
+```
+
 ```{r include=FALSE}

### 2020-12-17 paper/main.tex
main

@@
 
-\begin{wrapfigure}{r}{5.2cm}
-\begin{tabular}{|r@{\,}r@{\,}lr|r@{\,}r@{\,}lr|} \hline
-\tt 0 &--& \tt 50    & \Runbina & \tt 501 &--&\tt 1000   & \Runbine\\
-\tt 51 &--& \tt 100  & \Runbinb & \tt 1001 &--&\tt 1500  & \Runbinf\\
-\tt 101 &--& \tt 250 & \Runbinc & \tt 1501 &--&\tt 2000  & \Runbing\\
+\begin{wrapfigure}{r}{5.8cm}
+  \begin{tabular}{|r@{\,}r@{\,}l@{\,}r|r@{\,}r@{\,}l@{}r|} \hline
+\multicolumn{3}{|c}{\small\#calls} &\small\#sites &
+\multicolumn{3}{c}{\small\#calls} &\small\#sites \\\hline
+\tt 0 &--& \tt 50    & \Runbina & \tt 501 &--& \tt 1000   & \Runbine\\
+\tt 51 &--& \tt 100  & \Runbinb & \tt 1001 &--& \tt 1500  & \Runbinf\\
+\tt 101 &--& \tt 250 & \Runbinc & \tt 1501 &--& \tt 2000  & \Runbing\\
 \tt 251 &--& \tt 500 & \Runbind & \tt 2001 &--& \tt 3000 & \Runbinh\\\hline
@@
 
-\begin{wrapfigure}{r}{5.2cm}
+\medskip
+
+\begin{wrapfigure}{r}{5.5cm}
   \vspace{-8mm}
@@
       (small)} \label{fig:callsitesnormalized}
-    \vspace{-3mm}
+   \vspace{-8mm}
 \end{wrapfigure}
@@
 that the frequency quickly drops.
+
+Analysis of the arguments passed to \eval reveals that XXX are of kind
+\c{symbol} (single variables), XXX are \c{language} (an internal R
+type roughly corresponding to a single expression), and XXX are
+\c{expression} (one or more statements). The remaining arguments are R
+values which are returned unchanged.
 
@@
 \NOTE{Talk about the percentage of code executed in eval}
-
-\mypara{Eval Arguments.} Figure~\ref{fig:eval-expression-kind} breaks
-down the expressions passed to \eval by R data type. It separates the
-core packages from CRAN as they have different characteristics. In
-core, XXX036\% of arguments in core are expressions. Whereas, in CRAN
-package XXX037\% are expressions. When the argument is not an
-expression, it is a value. Values are trivial for \eval as they are
-returned unchanged. \PackageEnvironmentInputEvalCallPerc of CRAN
-arguments are environments. This is an anomaly as only four \eval
-call sites dominate this result:
-\begin{compactitem}[$-$]
-\item \c{ggplot2::ggproto} contributes to 99.9\% of package \eval
-  calls that receive environments as input.
-\item \c{R6::generator_funs} function contributes to only 936 \eval
-  calls and implements the same functionality as \c{ggplot2::ggproto}
-  but for the \emph{R6} OOP system.
-\item \c{future::backtrace} function applies \eval to a future object
-  which is implemented as an environment. This is called only once.
-\item \c{RModel::str.RMmodel} function overloads the core \c{str}
-  method for its \c{RModel} objects and maps \eval exactly like
-  \c{str}. This is called only once.
-\end{compactitem}
-
-\begin{figure}[!h]
-  \centering
-  \includegraphics[width=\columnwidth]{eval-expression-kind}
-  \caption{Type of Expression passed to \eval} \label{fig:eval-expression-kind}
-\end{figure}
 

### 2020-12-17 analysis/usage.Rmd
Eval arguments

@@
 ) -> TBL
+```
 
-sizes_prop %>%
-  ggplot(aes(x = arg_size)) +
-  geom_histogram()
+`r TBL`
 
+```{r}
 E %>%
@@
 ```
-
-`r TBL`
 
@@
 ) -> TBL
+```
 
+```{r include=F}
 E %>%
@@
 ggsave(path(PLOT_DIR, "events_per_pack_large.pdf"))
+```
 
+```{r include=F}
 E %>%
@@
   r("Runbinh", sum[[8]])
-)
+) -> TBL
 ```
@@
 E %>%
-  ggplot(aes(x = fct_rev(fct_infreq(type)))) +
+  ggplot(aes(x = fct_rev(fct_infreq(type)), weight=duplicates)) +
   geom_bar() +
@@
   labs(x = "type")
+
+types <- E %>% count(type, wt=duplicates)
+
+symbs <- types %>% filter(type == "symbol") %>% .$n
+languages <- types %>% filter(type == "language") %>% .$n
+expressions <- types %>% filter(type == "expression") %>% .$n
+
+overview_table(
+  r("Symbolpercent", ratio(symbs, nb_calls)),
+  r("Languagepercent", ratio(languages, nb_calls)),
+  r("Expressionpercent", ratio(expressions, nb_calls)),
+  r("Codepercent", ratio(symbs + languages + expressions, nb_calls))
+) -> TBL
 ```
+
+`r TBL`
 

### 2020-12-17 paper/main.tex
Density of sizes and percentages of the argument to eval.

@@
 
-Analysis of the arguments passed to \eval reveals that XXX are of kind
-\c{symbol} (single variables), XXX are \c{language} (an internal R
-type roughly corresponding to a single expression), and XXX are
-\c{expression} (one or more statements). The remaining arguments are R
-values which are returned unchanged.
+Analysis of the arguments passed to \eval reveals that \Symbolpercent are of kind
+\c{symbol} (single variables), \Languagepercent are \c{language} (an internal R
+type roughly corresponding to a single expression), and \Expressionpercent are
+\c{expression} (one or more statements), for a total of \Codepercent of the \eval calls. The remaining arguments are R
+values which are returned unchanged. \NOTE{closures are forced. (but there are nearly none in our dataset)}
 
 \mypara{Amount of code loaded by eval.} We use \c{deparse} to turn the
-input to \eval into strings and measure the size of that input. The
+resolved input to \eval into strings and measure the size of that input. The
 largest input is \Maxszeval and the median input size is
 \Medianszeval. Most inputs are small, less than \SzninefiveRnd
-bytes. The average size is \Avgszeval bytes.
+characters. The average size is \Avgszeval characters. 
+Figure~\ref{fig:sizedistribution} shows the distribution of sizes.
+
+\begin{figure}[h]
+	%\centering 
+	\includegraphics[width=\textwidth]{size_loaded_distribution}
+	\caption{Distribution of the sized of the expressions loaded by \eval.} \label{fig:sizedistribution}
+\end{figure}
 

### 2020-12-17 analysis/usage.Rmd
More

@@
 E %>%
-  filter(opcodes > 0) %>%
-  filter(opcodes < 50) %>%
+  filter(opcodes > 0) %>% # == 0 if from eval.parent, because we do not count eval loops for it
+  filter(opcodes <= 50) %>%
   ggplot(aes(x = "", y = opcodes, fill = ""), na.rm = TRUE) +

### 2020-12-18 paper/main.tex
usage

--- 2020-12-17 paper/main.tex
@@
 To provide a picture of \eval usage in R, this section focuses on
-library code and the \Corpus packages in our CRAN corpus. For these
-packages, all runnable code in CRAN was executed, runs that did not
-exercise an \eval in the corpus were discarded. The data presented
-here was collected from \Nbruns runs. In total, \Allcalls invocations
-of \eval and its variants were recorded. Out of the
+library code in the \Corpus packages of our CRAN corpus. For these,
+all runnable code in CRAN was executed, runs that did not exercise an
+\eval from the corpus were discarded; \Nbruns runs were retained and
+generated the data presented here. In total, \Allcalls invocations of
+\eval and its variants were recorded. Out of the
 \Staticatleastonecallsite packages that use \eval, \Triggeredpkgs
@@
 \Fewcalls packages with low \eval frequency, i.e. fewer than 100
-calls; and a similar number of packages, \Manycalls to be precise, use
-\eval more than 1,000 times. One package, \Maxcallspack, calls it
-\Maxcalls times and thus accounts for over half of the observed
-invocations of \eval.
+calls; and a similar number of packages, \Manycalls to be precise,
+that use \eval more than 1,000 times. One package, \Maxcallspack,
+calls it \Maxcalls times and thus accounts for over half of the
+observed invocations of \eval.
 
@@
 (\emph{invocations}). The vast majority of call sites, and calls, are
-to \eval itself; \c{evalq} and \c{local} are barely used at all.
+to \eval itself; \c{evalq} and \c{local} are barely used at all. The
+difference between static and exercised reflects spotty code coverage.
 
@@
 \caption{Normalized invocations} \label{tab:callsitesnormalized}
+
+\includegraphics[width=.4\textwidth]{calls_per_run_per_call_site}
+    \caption{Normalized invocations
+      (small)} \label{fig:callsitesnormalized}
+   \vspace{-8mm}
 \end{wrapfigure}
@@
 times in any run they are exercised.
-
-\medskip
-
-\begin{wrapfigure}{r}{5.5cm}
-  \vspace{-8mm}
-    \includegraphics[width=.4\textwidth]{calls_per_run_per_call_site}
-    \caption{Normalized invocations
-      (small)} \label{fig:callsitesnormalized}
-   \vspace{-8mm}
-\end{wrapfigure}
-
-The majority of call sites, \Runbina, are exercised up to 50 times.
 Larger numbers of calls suggest that \eval may be called within a loop
-or from a recursive context -- this seems to be the exception.
-Figure~\ref{fig:callsitesnormalized} zooms in on call sites with an
-average of fewer than 51 invocation. The x-axis shows the number of
-invocations and the y-axis is the number of call sites. Most call
-sites are invoked once, about half as many are invoked twice. After
-that the frequency quickly drops.
-
-Analysis of the arguments passed to \eval reveals that XXX are of kind
-\c{symbol} (single variables), XXX are \c{language} (an internal R
-type roughly corresponding to a single expression), and XXX are
-\c{expression} (one or more statements). The remaining arguments are R
-values which are returned unchanged.
-
-\mypara{Amount of code loaded by eval.} We use \c{deparse} to turn the
-input to \eval into strings and measure the size of that input. The
-largest input is \Maxszeval and the median input size is
-\Medianszeval. Most inputs are small, less than \SzninefiveRnd
-bytes. The average size is \Avgszeval bytes.
-
-\mypara{Amount of computation via eval.} We compare the number of
-events intercepted during execution of \eval across the whole corpus.
-For this, we use the number of interpreter steps as a proxy for
-events. We observe \AllEventCountRnd \NOTE{Check that number.}
-events all told, and \EventsRnd events occur inside eval
-(\EvalEventAllPerc of all events). Figure~\ref{fig:eventsPerPackage}
-show a violin plot of events.
-
-\begin{figure}[h]
+or from a recursive context -- this seems to be the exception as the
+majority of call sites, \Runbina, are exercised less than 51 times.
+Figure~\ref{fig:callsitesnormalized} zooms in on low frequency call
+sites. The x-axis shows the normalized invocations and the y-axis is
+the number of call sites for that value. The figure shows that most
+call sites are invoked only once, about half as many are invoked
+twice. After that the frequency of invocation quickly drops.
+
+Analysis of the arguments passed to \eval reveals that \Symbolpercent
+are of kind \c{symbol} (these are single variables such as \c{x}),
+\Languagepercent are \c{language} (an internal R type roughly
+corresponding to a single expression, such as \c{x+1}), and
+\Expressionpercent are \c{expression} (one or more statements). These
+account for \Codepercent of the invocations, the arguments to the
+remaining invocations are R values which are returned unchanged by
+\eval.
+
+To estimate how much executable code is injected into the program
+through \eval, we measure the size of arguments in number of
+characters. This measurement is obtained by de-parsing every argument
+to \eval. Our de-narser does not emit code for values as some values
+can be very large and would overwhelm the code. On average, arguments
+to \eval are \Avgszeval characters long, the median size is
+\Medianszeval. The large input observed is \Maxszeval.
+Figure~\ref{fig:sizedistribution} shows the distribution of sizes. The
+x-axis is the size of arguments in characters, and the y-axis is the
+count of arguments with that size.
+
+\begin{figure}[h!]
+ \includegraphics[width=\textwidth]{size_loaded_distribution}
+\caption{Loaded code} \label{fig:sizedistribution}
+\end{figure}
+
+To measure the work performed in \evals, we instrument the R
+interpreter to count the number of operations performed during
+evaluation of the argument. For the purposes of this experiment, each
+interpreter step is an \emph{event}. For our corpus, \EvalEventAllPerc
+of events (\EventsRnd) are performed within \eval. \Smalleventspct of
+\evals perform 50 or fewer events. Figure~\ref{fig:eventsPerPackage}
+shows two violin plots. The plot on the left is for evals with fewer
+or equal than 50 events, the plot on the right is for more
+work-intensive \evals. The biggest \eval performs \MaxeventsRnd
+events.
+
+\begin{figure}[h!]
 	\centering
@@
 \end{figure}
-
-We observe a very wide spread in the number of events generated by
-\eval calls. The largest number of events generated by an \eval call
-is \MaxeventsRnd. However, a majority (\Smalleventspct) of \eval calls
-performs up to 50 events suggesting that most expressions passed to
-\eval are small. It is interesting to note that a higher proportion of
-Core R \eval calls perform side effects in the
-\EventsMinRangeB--\EventsMaxRangeB range. Only \EventsCoreEvalCountC
-Core R \eval calls and \EventsPackageEvalCountC package \eval call
-generate \EventsMinRangeC to \EventsMaxRangeC events. These \eval
-calls originate from statistical modeling packages such as \mlogit,
-\mboost, \metafor, \lavaan, \mclust and \gamlss.
-
-\NOTE{Talk about the percentage of code executed in eval}
 

### 2020-12-18 paper/main.tex
usage
todo

@@
 
+\newpage
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-\begin{wrapfigure}{r}{5.2cm}
+Figure~\ref{freq} summarizes the frequency of calls to \eval over all
+runs. The left column is the number of observed calls, the right
+column is the number of packages falling in that range. There are
+\Fewcalls packages with low \eval frequency, i.e. fewer than 100
+calls; and a similar number of packages, \Manycalls to be precise,
+that use \eval more than 1,000 times. One package, \Maxcallspack,
+calls it \Maxcalls times and thus accounts for over half of the
+observed invocations of \eval.
+
+\begin{figure}[!h]
+
+\begin{tabular}{@{}l@{\hspace{1.5cm}}l@{}}
+\begin{minipage} {5cm}
   \begin{tabular}{|r@{\,}r@{\,}l@{}r|r@{\,}r@{\,}l@{}r|} \hline
@@
 \caption{Call frequency}\label{freq}
-\end{wrapfigure}
-
-Figure~\ref{freq} summarizes the frequency of calls to \eval over all
-runs. The left column is the number of observed calls, the right
-column is the number of packages falling in that range. There are
-\Fewcalls packages with low \eval frequency, i.e. fewer than 100
-calls; and a similar number of packages, \Manycalls to be precise,
-that use \eval more than 1,000 times. One package, \Maxcallspack,
-calls it \Maxcalls times and thus accounts for over half of the
-observed invocations of \eval.
-
-\begin{wrapfigure}{r}{7cm}
+\end{minipage}
+&
+\begin{minipage}{7cm}
 \begin{tabular}{|@{\,}r@{\,}|@{\,}rrrr@{\,}|}\hline
@@
 \small Invocations&\EvalsRnd&\EvalqsRnd&\EparentsRnd&\LocalsRnd\\\hline
-\end{tabular}\caption{Variants}\label{tab:variantseval}
-\end{wrapfigure}
-
-Figure~\ref{tab:variantseval} illustrates the use of variants of
-\eval. For each of the four variants, the first rows shows the number
-of call sites occurring in the corpus (\emph{static)}), the second row
-is the number of call sites that were encountered during tracing
-(\emph{exercised}), the last row is the number of calls
-(\emph{invocations}). The vast majority of call sites, and calls, are
-to \eval itself; \c{evalq} and \c{local} are barely used at all. The
-difference between static and exercised reflects spotty code coverage.
+\end{tabular}~\\[2mm]\caption{Variants}\label{tab:variantseval}
+\end{minipage}
+\end{tabular}
+\end{figure}
 
 \begin{wrapfigure}{r}{5.8cm}
+\vspace{-4mm}
+
   \begin{tabular}{|r@{\,}r@{\,}l@{\,}r|r@{\,}r@{\,}l@{}r|} \hline
@@
 
-\includegraphics[width=.4\textwidth]{calls_per_run_per_call_site}
+\vspace{-2mm}
+\includegraphics[width=.35\textwidth]{calls_per_run_per_call_site}
+\vspace{-2mm}
     \caption{Normalized invocations
@@
 \end{wrapfigure}
+
+Figure~\ref{tab:variantseval} illustrates the use of variants of
+\eval. For each of the four variants, the first rows shows the number
+of call sites occurring in the corpus (\emph{static)}), the second row
+is the number of call sites that were encountered during tracing
+(\emph{exercised}), the last row is the number of calls
+(\emph{invocations}). The vast majority of call sites, and calls, are
+to \eval itself; \c{evalq} and \c{local} are barely used at all. The
+difference between static and exercised reflects spotty code coverage.
 
@@
 \end{figure}
+
+\subsection{Things todo}
+
+\begin{itemize}
+\item Increase font size on numbers for figures
+\item change colors on figure 8 (don't like the red)
+\item check that on figure 7, the number of elements at 1 is correct
+\item explain the peak at 4, for figure 7
+\item Figure 8 explain why so few 1's
+\item Add a ``Discussion'' paragraph at the end that sums up the
+  observations
+\item Add paragraphs for Kaggle and CORE -- only discuss the
+  differences from CRAN if any.
+\item Question: should we remove ggplot? Would change things
+  significantly?
+\end{itemize}
 

### 2020-12-18 analysis/usage.Rmd
Improve graphs

--- 2020-12-17 analysis/usage.Rmd
@@
 ```{r}
+lbl_colour <- rep("black", 11)
+lbl_colour[[4]] <- "transparent"
 E %>%
@@
   scale_x_continuous(breaks = c(0, 1, 2, 3, 4, 5, 10, 30, 100, 300, 1000), trans = "log1p", expand = c(0, 0)) +
-  labs(x = "Size (characters)")
+  labs(x = "Size (characters)") +
+  theme(text = element_text(size=17),
+        axis.text.x = element_text(color = lbl_colour))
 
@@
 
-```{r include=F}
+```{r}
 E %>%
   filter(opcodes > 50) %>%
-  ggplot(aes(x = "", y = opcodes, fill = ""), na.rm = TRUE) +
-  geom_violin(trim = FALSE, width = 1.7) +
+  ggplot(aes(x = "", y = opcodes), na.rm = TRUE) +
+  geom_violin(trim = FALSE, width = 1.7, fill = "blue", alpha = 0.3) +
   scale_y_log10(labels = scales::comma) +
   labs(y = "", x = "") +
-  theme(legend.position = "none")
+  theme(legend.position = "none", text = element_text(size=17))
 
@@
 
-```{r include=F}
+```{r}
 E %>%
-  filter(opcodes > 0) %>%
-  filter(opcodes < 50) %>%
-  ggplot(aes(x = "", y = opcodes, fill = ""), na.rm = TRUE) +
-  geom_violin(trim = FALSE, width = 1.7) +
-  labs(y = "Events per package", x = "") +
-  theme(legend.position = "none")
+  filter(opcodes > 0) %>% # == 0 if from eval.parent, because we do not count eval loops for it
+  filter(opcodes <= 50) %>%
+  ggplot(aes(x = "", y = opcodes), na.rm = TRUE) +
+  geom_violin(trim = FALSE, width = 1.7, fill = "blue", alpha = 0.3) +
+  labs(y = "Events", x = "") +
+  theme(legend.position = "none", text = element_text(size=17))
 
@@
   arrange(desc(average))
+```
 
+```{r}
 # Plot the low trigger counts...
@@
   ggplot(aes(x = average)) +
-  geom_histogram(binwidth = 1)
+  geom_histogram(binwidth = 1, fill = "blue", alpha = 0.3, colour = "black")
 
 ggsave(path(PLOT_DIR, "calls_per_run_per_call_site.pdf"))
+```
 
+```{r, include = FALSE}
 bins <- cut(triggered$average, breaks = c(0, 50, 100, 250, 500, 1000, 1500, 2000, 3000), include.lowest = TRUE)

### 2020-12-18 analysis/usage.Rmd
Change aspect ratio

@@
   ggplot(aes(x = "", y = opcodes), na.rm = TRUE) +
-  geom_violin(trim = FALSE, width = 1.7, fill = "blue", alpha = 0.3) +
+  geom_violin(trim = FALSE, width = 1, fill = "blue", alpha = 0.3) +
   scale_y_log10(labels = scales::comma) +
@@
   ggplot(aes(x = average)) +
-  geom_histogram(binwidth = 1, fill = "blue", alpha = 0.3, colour = "black")
+  geom_histogram(binwidth = 1, fill = "blue", alpha = 0.3, colour = "black") +
+  theme(text = element_text(size=20), aspect.ratio = 4/3)
 

### 2020-12-18 analysis/usage.Rmd
Using AST nodes

--- 2020-12-18 analysis/usage.Rmd
@@
 library(DT)
+library(scales)
 
@@
   arg_size = expr_resolved_length, #   Size in characters (unicode) of eval argument string
+  ast_nodes = expr_resolved_nodes,
   type = expr_resolved_type_tag,
@@
 
-med_size <- weighted.median(E$arg_size, w = E$duplicates, na.rm = TRUE)
-max_size <- max(E$arg_size, na.rm = TRUE)
-average_size <- weighted.mean(E$arg_size, na.rm = TRUE, w = E$duplicates)
+med_size <- weighted.median(E$ast_nodes, w = E$duplicates, na.rm = TRUE)
+max_size <- max(E$ast_nodes, na.rm = TRUE)
+average_size <- weighted.mean(E$ast_nodes, na.rm = TRUE, w = E$duplicates)
 
 sizes_prop <- E %>%
-  count(arg_size, wt = duplicates) %>%
+  count(ast_nodes, wt = duplicates) %>%
   mutate(proportion = n / sum(n)) %>%
-  arrange(arg_size) %>%
+  arrange(ast_nodes) %>%
   mutate(cumulative = cumsum(proportion))
@@
   r("Avgszeval", average_size),
-  r("Szninefive", many$arg_size)
+  r("Szninefive", many$ast_nodes)
 ) -> TBL
@@
 E %>%
-  ggplot(aes(x = arg_size)) +
+  ggplot(aes(x = ast_nodes, weight = duplicates)) +
   stat_density(aes(y = ..count..), color = "black", fill = "blue", alpha = 0.3) +
   scale_x_continuous(breaks = c(0, 1, 2, 3, 4, 5, 10, 30, 100, 300, 1000), trans = "log1p", expand = c(0, 0)) +
-  labs(x = "Size (characters)") +
+  labs(x = "Size (AST nodes)") +
   theme(text = element_text(size=17),
         axis.text.x = element_text(color = lbl_colour))
+
+E %>% 
+  filter(ast_nodes < 25) %>%
+  ggplot(aes(x = ast_nodes, weight = duplicates)) +
+  geom_histogram(binwidth = 1, color = "black", fill = "blue", alpha = 0.3) +
+  scale_y_continuous(labels = comma)
 
@@
 
-How much work is involved in each eval as measured by the number of ??? 
-(possible: opcodes, AST nodes).
+How much work is involved in each eval as measured by the number of opcodes exectued by the interpreter.
 
@@
   ggplot(aes(x = "", y = opcodes), na.rm = TRUE) +
-  geom_violin(trim = FALSE, width = 1.7, fill = "blue", alpha = 0.3) +
+  geom_violin(trim = FALSE, width = 1, fill = "blue", alpha = 0.3) +
   scale_y_log10(labels = scales::comma) +
   labs(y = "", x = "") +
-  theme(legend.position = "none", text = element_text(size=17))
+  theme(legend.position = "none", text = element_text(size=25))
 
@@
   labs(y = "Events", x = "") +
-  theme(legend.position = "none", text = element_text(size=17))
+  theme(legend.position = "none", text = element_text(size=25))
 
@@
   ggplot(aes(x = average)) +
-  geom_histogram(binwidth = 1, fill = "blue", alpha = 0.3, colour = "black")
+  geom_histogram(binwidth = 1, fill = "blue", alpha = 0.3, colour = "black") +
+  theme(text = element_text(size=16), axis.text = element_text(size=20), aspect.ratio = 4/3)
 

### 2020-12-18 analysis/usage.Rmd
Percent of symbols coming from ggplot

@@
 
+# Diversity of arguments
+
 How diverse are the expressions passed to `eval`? We consider the diversity in types in the resolved expression.
@@
 
+ggplot2_symbs <- E %>% filter(ev_package == "ggplot2") %>% count(type, wt=duplicates) %>% filter(type == "symbol") %>% .$n
+
 overview_table(
@@
   r("Expressionpercent", ratio(expressions, nb_calls)),
-  r("Codepercent", ratio(symbs + languages + expressions, nb_calls))
+  r("Codepercent", ratio(symbs + languages + expressions, nb_calls)),
+  r("Ggplotsymbolpercent", ratio(ggplot2_symbs, symbs))
 ) -> TBL

### 2020-12-18 paper/main.tex
Plots
Fixing figures
ggproto: most of the symbols evaluated

@@
 \vspace{-2mm}
-\includegraphics[width=.35\textwidth]{calls_per_run_per_call_site}
-\vspace{-2mm}
+\includegraphics[width=7.5cm, trim=7.5cm 0 6cm 0, clip]{calls_per_run_per_call_site}
+\vspace{-7mm}
     \caption{Normalized invocations
@@
 remaining invocations are R values which are returned unchanged by
-\eval.
+\eval. Most of the invocations involving a symbol come from ggplot2, \Ggplotsymbolpercent, and from one call site, in \c{ggproto}, evaluating symbol \c{`_inherit`}.
 
@@
 through \eval, we measure the size of arguments in number of
-characters. This measurement is obtained by de-parsing every argument
-to \eval. Our de-narser does not emit code for values as some values
+AST nodes. For instance, a variable name (a symbol) counts for 1 node. Our de-parser does not emit code for values as some values
 can be very large and would overwhelm the code. On average, arguments
-to \eval are \Avgszeval characters long, the median size is
-\Medianszeval. The large input observed is \Maxszeval.
+to \eval are \Avgszeval nodes large, the median size is
+\Medianszeval. The largest input observed is \Maxszeval.
 Figure~\ref{fig:sizedistribution} shows the distribution of sizes. The
-x-axis is the size of arguments in characters, and the y-axis is the
+x-axis is the size of arguments in number of nodes, and the y-axis is the
 count of arguments with that size.
@@
 	\centering
-	\includegraphics[width=.4\textwidth]{events_per_pack_small}
-	\includegraphics[width=.4\textwidth]{events_per_pack_large}
+	\includegraphics[width=.55\textwidth]{events_per_pack_small}
+	\hspace{-10mm}
+	\includegraphics[width=.55\textwidth]{events_per_pack_large}
 	\caption{Events per packages} \label{fig:eventsPerPackage}
@@
 \begin{itemize}
-\item Increase font size on numbers for figures
-\item change colors on figure 8 (don't like the red)
 \item check that on figure 7, the number of elements at 1 is correct

### 2020-12-19 paper/main.tex
usage rewrite

@@
 
-To provide a picture of \eval usage in R, this section focuses on
-library code in the \Corpus packages of our CRAN corpus. For these,
-all runnable code in CRAN was executed, runs that did not exercise an
-\eval from the corpus were discarded; \Nbruns runs were retained and
-generated the data presented here. In total, \Allcalls invocations of
-\eval and its variants were recorded. Out of the
-\Staticatleastonecallsite packages that use \eval, \Triggeredpkgs
-packages were observed calling it --- the difference can be chalked
-down to incomplete code coverage and occasional infrastructure
-failures.
+This section focuses on our corpus of \Corpus CRAN packages that use
+\eval; these represent {\bf XX\%} of the entire repository. To provide
+a picture of \eval usage, all runnable code in CRAN was executed. Runs
+that did not exercise an \eval in the corpus were discarded. We
+retained \Nbruns to generate the data presented here. In total,
+\Allcalls invocations of \eval and its variants were recorded. Out of
+the \Staticatleastonecallsite packages that use \eval, \Triggeredpkgs
+packages were observed calling it in our runs --- the difference can
+be chalked down to incomplete code coverage and occasional
+infrastructure failures.
 
@@
 calls it \Maxcalls times and thus accounts for over half of the
-observed invocations of \eval.
+observed data.
 
@@
 \vspace{-4mm}
-
+\centering
   \begin{tabular}{|r@{\,}r@{\,}l@{\,}r|r@{\,}r@{\,}l@{}r|} \hline
@@
 \end{tabular}
-\caption{Normalized invocations} \label{tab:callsitesnormalized}
+
+\medskip  (a) All
+
+  \includegraphics[width=5.8cm, trim=7.5cm 0 6cm 0, clip]{calls_per_run_per_call_site}
+
+  (b) Small
 
 \vspace{-2mm}
-\includegraphics[width=7.5cm, trim=7.5cm 0 6cm 0, clip]{calls_per_run_per_call_site}
-\vspace{-7mm}
-    \caption{Normalized invocations
-      (small)} \label{fig:callsitesnormalized}
-\vspace{-8mm}
+    \caption{Normalized invocations} \label{cn}
+\vspace{-2mm}
 \end{wrapfigure}
@@
 \eval. For each of the four variants, the first rows shows the number
-of call sites occurring in the corpus (\emph{static)}), the second row
+of call sites occurring in the corpus (\emph{static}), the second row
 is the number of call sites that were encountered during tracing
@@
 (\emph{invocations}). The vast majority of call sites, and calls, are
-to \eval itself; \c{evalq} and \c{local} are barely used at all. The
-difference between static and exercised reflects spotty code coverage.
-
-Figure~\ref{tab:callsitesnormalized} shows normalized invocation
-counts. The column on the left is the average number of invocations of
-\eval from a particular call site for any given run. The number on the
-right is number of call sites that fall in that range. For instance,
-there are \Runbinh call sites that are invoked on average over 2,001
-times in any run they are exercised.
-Larger numbers of calls suggest that \eval may be called within a loop
-or from a recursive context -- this seems to be the exception as the
-majority of call sites, \Runbina, are exercised less than 51 times.
-Figure~\ref{fig:callsitesnormalized} zooms in on low frequency call
-sites. The x-axis shows the normalized invocations and the y-axis is
-the number of call sites for that value. The figure shows that most
-call sites are invoked only once, about half as many are invoked
-twice. After that the frequency of invocation quickly drops.
-
-Analysis of the arguments passed to \eval reveals that \Symbolpercent
-are of kind \c{symbol} (these are single variables such as \c{x}),
-\Languagepercent are \c{language} (an internal R type roughly
+to \eval itself, \c{eval.parent} is rare, as for \c{evalq} and
+\c{local} are barely used at all. The difference between static and
+exercised reflects spotty code coverage and is a limitation
+of dynamic analysis.
+
+Figure~\ref{cn}(a) shows normalized invocation counts. The column on
+the left is the average number of calls from a particular site for any
+given run (\emph{\#calls}). On the right is the count of sites that
+fall in that range (\emph{\#sites}). For instance, \Runbinh sites are
+invoked on average over 2,001 times per run. Larger numbers suggest
+uses from within a loop or a recursive context -- this seems to be the
+exception as most \evals, \Runbina, are exercised 50 times or less.
+Figure~\ref{cn}(b) zooms in on low frequency call sites. The x-axis
+shows normalized invocations and the y-axis is the number sites for
+that value. Most low-frequency \evals are invoked only once, about
+half as many are invoked twice; after that the frequency quickly
+drops.
+
+Arguments to \eval can be either code or values, in the latter case
+\eval has no effect. Code accounts for \Codepercent of arguments in
+our corpus. More specifically, \Symbolpercent of arguments are
+\c{symbol}s (i.e. these are single variables such as \c{x}),
+\Languagepercent are \c{language} object (an internal R type roughly
 corresponding to a single expression, such as \c{x+1}), and
-\Expressionpercent are \c{expression} (one or more statements).  These
-account for \Codepercent of the invocations, the arguments to the
-remaining invocations are R values which are returned unchanged by
-\eval. Most of the invocations involving a symbol come from ggplot2, \Ggplotsymbolpercent, and from one call site, in \c{ggproto}, evaluating symbol \c{`_inherit`}.
-
-To estimate how much executable code is injected into the program
-through \eval, we measure the size of arguments in number of
-AST nodes. For instance, a variable name (a symbol) counts for 1 node. Our de-parser does not emit code for values as some values
-can be very large and would overwhelm the code. On average, arguments
-to \eval are \Avgszeval nodes large, the median size is
-\Medianszeval. The largest input observed is \Maxszeval.
-Figure~\ref{fig:sizedistribution} shows the distribution of sizes. The
-x-axis is the size of arguments in number of nodes, and the y-axis is the
-count of arguments with that size.
+\Expressionpercent are \c{expression} objects (multiple statements).
+Further inspection reveals that most symbols, \Ggplotsymbolpercent to
+be exact, come from a single call in the \c{ggplot2} package and have
+have the value \c{_inherit}. Even if we remove this outlier, the
+majority of \evals are simple expressions.
+
+To estimate how much executable code is dynamically injected into the
+program through \eval, we measure the size of arguments in number of
+abstract syntax tree nodes as returned by R's parser; for example,
+\c{x+1} counts as 3 nodes.\footnote{We experimented with measuring
+code size by de-parsing the expressions to strings, but these
+measurements were dominated by de-parsed data objects, these could
+range in the MBs.} The median argument size is \Medianszeval (recall
+the large number of symbols) and the average is \Avgszeval nodes. The
+largest \eval input observed is \Maxszeval\,-- a significant chunk of
+code. Figure~\ref{fig:sizedistribution} shows the distribution of
+sizes for arguments of fewer or equal to 25 nodes. The x-axis is the
+size of arguments in number of nodes, and the y-axis is the count of
+arguments with that size. The size drops rapidly, with few
+observations larger than 15 nodes. The long tail is omitted from the
+graph for legibility.
 
 \begin{figure}[h!]
- \includegraphics[width=\textwidth]{size_loaded_distribution}
+ \includegraphics[width=1.061\textwidth]{size_loaded_distribution}
 \caption{Loaded code} \label{fig:sizedistribution}
@@
 
-To measure the work performed in \evals, we instrument the R
-interpreter to count the number of operations performed during
-evaluation of the argument. For the purposes of this experiment, each
-interpreter step is an \emph{event}. For our corpus, \EvalEventAllPerc
-of events (\EventsRnd) are performed within \eval. \Smalleventspct of
-\evals perform 50 or fewer events. Figure~\ref{fig:eventsPerPackage}
-shows two violin plots. The plot on the left is for evals with fewer
-or equal than 50 events, the plot on the right is for more
-work-intensive \evals. The biggest \eval performs \MaxeventsRnd
-events.
+To measure the work performed in \evals, we count the instructions
+executed by interpreter during evaluation of the argument. In our
+corpus, \EvalEventAllPerc of all instructions (\EventsRnd) are
+performed within \eval. Most invocations perform relatively little
+work with \Smalleventspct of \evals executing 50 or fewer
+instructions. The violin plot of Figure~\ref{ev}(a) corresponds to
+\evals executing $\leq$ 50 instructions, it is dominated by trivial
+symbol look ups. Figure~\ref{ev}(b) has the work-intensive \evals
+which go all the way to \MaxeventsRnd instructions.
 
 \begin{figure}[h!]
-	\centering
-	\includegraphics[width=.55\textwidth]{events_per_pack_small}
-	\hspace{-10mm}
-	\includegraphics[width=.55\textwidth]{events_per_pack_large}
-	\caption{Events per packages} \label{fig:eventsPerPackage}
+\begin{tabular}{@{}c@{}c@{}}
+\begin{minipage}{7.5cm}
+ \includegraphics[width=\textwidth]{events_per_pack_small}
+\end{minipage}&\begin{minipage}{7.5cm}
+  \includegraphics[width=\textwidth]{events_per_pack_large}
+\end{minipage}\\[-3mm]
+\small (a) Small & \small (b) Large
+\end{tabular}
+ \caption{Events per packages} \label{ev}
 \end{figure}
 
-\subsection{Things todo}
+\subsection{Things to do}
 
 \begin{itemize}
-\item check that on figure 7, the number of elements at 1 is correct
-\item explain the peak at 4, for figure 7
-\item Figure 8 explain why so few 1's
+\item Figure 5(b) -- make `count` and `average` smaller.
+\item Figure 6: (1) Replace `ast-nodes` with `nodes`
+\item Figure 6: (2) Remove gray lines
+\item Figure 6: increase font size
+\item Figure 6: Try to make the figure less wide. Thinner bars/aspect
+  ratio change.
+\item Figure 7: check that element counts are correct
+\item Figure 7: replace events by instructions. Check what we are
+  counting? Are these bytecodes or AST nodes.
+\item Check that the percentage of operations performed within
+  an \eval (5\%) is correct. That feels surprising to me.
+\item I am having second thoughts about keeping the \c{ggplot2} data
+  -- it is a big outlier that skews all the graphs...
 \item Add a ``Discussion'' paragraph at the end that sums up the
@@
 \item Add paragraphs for Kaggle and CORE -- only discuss the
-  differences from CRAN if any.
-\item Question: should we remove ggplot? Would change things
-  significantly?
+  differences from CRAN
 \end{itemize}

### 2020-12-19 analysis/usage.Rmd
Polish usage metrics

@@
   scale_x_continuous(breaks = c(0, 1, 2, 3, 4, 5, 10, 30, 100, 300, 1000), trans = "log1p", expand = c(0, 0)) +
-  labs(x = "Size (AST nodes)") +
+  labs(x = "Size (nodes)") +
   theme(text = element_text(size=17),
@@
   geom_histogram(binwidth = 1, color = "black", fill = "blue", alpha = 0.3) +
-  scale_y_continuous(labels = comma)
+  scale_y_continuous(labels = comma) +
+  labs(x = "nodes") +
+  theme(text = element_text(size=17),  panel.grid.major.x = element_blank(),
+        panel.grid.minor.x = element_blank(), aspect.ratio = 4/3)
 
@@
   geom_violin(trim = FALSE, width = 1.7, fill = "blue", alpha = 0.3) +
-  labs(y = "Events", x = "") +
+  labs(y = "Instructions", x = "") +
   theme(legend.position = "none", text = element_text(size=25))

### 2020-12-20 analysis/usage.Rmd
changes

@@
 ```{r read, message=TRUE, include=F}
-# This test should allow to re-run the whole script without reading the
-# data again from disk. It assumes the data doesn't change.
-if (exists("E") == FALSE) {
-  read_fst(eval_calls_path) %>% as_tibble() -> E_raw
 read_fst(corpus_path) %>% as_tibble() -> C_raw
 read_csv(eval_static_path) %>% semi_join(C_raw, by = "package") -> P
+
+rebuild <- FALSE
+# We only read the orignal evals_calls file if we don't have a summarized
+# version stored on disk.  NOTE: the heuristic to detect changes is file
+# size. If that fails, manually delete `.lock.fst`
+if (file.exists("Data/.lock.fst")) {
+  saved <- read.fst("Data/.lock.fst")[[1,1]]
+  if (saved == file.size(eval_calls_path)) {
+     read.fst("Data/E.fst") %>% as_tibble() -> E
+  } else {
+    rebuild <- TRUE
   }
-```
-
-The two data sets are `E` and `C` for conciseness.  `C` is only used here to 
-compute the corpus size. `E` is used throughout. Recall that `E_raw` has the 
-source data with all columns. `P` corresponds to static information about the
-call sites.
-
-```{r select-columns, message=TRUE, warning=FALSE, include=FALSE}
+}
+if (rebuild) {
+  read_fst(eval_calls_path) %>% as_tibble() -> E_raw
   E_raw %>% select(
@@
     ev_package = eval_source, #   Package of the call site
-  arg_size = expr_resolved_length, #   Size in characters (unicode) of eval argument string
-  ast_nodes = expr_resolved_nodes,
-  type = expr_resolved_type_tag,
+    arg_size = expr_resolved_length, #   Size in characters of argument 
+    ast_nodes = expr_resolved_nodes, #   Size in AST Nodes
+    type = expr_resolved_type_tag,   #   Argument type
     opcodes = direct_interpreter_eval #  Count of the AST nodes traversed by the
-) -> E #                               interpreter during eval (non-recursive).
+  ) %>% #                                interpreter during eval (non-recursive).
+   mutate(across(where(is_character),as_factor)) -> E 
+  write.fst(E, "Data/E.fst")
+  sz <- file.size(eval_calls_path) %>% as_tibble()
+  write.fst(sz,"Data/.lock.fst")
+}
 
@@
 ```
+
+The two data sets are `E` and `C` for conciseness.  `C` is only used here to 
+compute the corpus size. `E` is used throughout. Recall that `E_raw` has the 
+source data with all columns. `P` corresponds to static information about the
+call sites.
 

### 2020-12-20 analysis/usage.Rmd
Download other preprocessed files

@@
   base_dir: Data/
-  calls_packages_path: summarized-packages.fst
+  calls_packages_path: package/summarized.fst
   corpus_path: corpus.fst

### 2020-12-21 analysis/usage.Rmd
usage

--- 2020-12-20 analysis/usage.Rmd
@@
 
-rebuild <- FALSE
+rebuild <- TRUE
 # We only read the orignal evals_calls file if we don't have a summarized
@@
     read.fst("Data/E.fst") %>% as_tibble() -> E
-  } else {
-    rebuild <- TRUE
+    rebuild <- FALSE
   }
@@
   n_distinct()
+
 overview_table(
@@
 ```{r exercised-vs-static, include=FALSE}
-stat_evl <- P %>%
-  filter(call_fun_name == "eval") %>%
-  nrow()
-stat_evq <- P %>%
-  filter(call_fun_name == "evalq") %>%
-  nrow()
-stat_evp <- P %>%
-  filter(call_fun_name == "eval.parent") %>%
-  nrow()
-stat_loc <- P %>%
-  filter(call_fun_name == "local") %>%
-  nrow()
+stat_evl <- P %>% filter(call_fun_name == "eval")        %>% nrow()
+stat_evq <- P %>% filter(call_fun_name == "evalq")       %>% nrow()
+stat_evp <- P %>% filter(call_fun_name == "eval.parent") %>% nrow()
+stat_loc <- P %>% filter(call_fun_name == "local")       %>% nrow()
 
@@
 tmp <- sites_per_variant %>% filter(ev_variant == "eval")
-dyn_evl <- if (nrow(tmp) == 0) {
-  0
-} else {
-  tmp$n
-}
+dyn_evl <- if (nrow(tmp) == 0) { 0 } else { tmp$n }
 tmp <- sites_per_variant %>% filter(ev_variant == "evalq")
-dyn_evq <- if (nrow(tmp) == 0) {
-  0
-} else {
-  tmp$n
-}
+dyn_evq <- if (nrow(tmp) == 0) { 0 } else { tmp$n }
 tmp <- sites_per_variant %>% filter(ev_variant == "eval.parent")
-dyn_evp <- if (nrow(tmp) == 0) {
-  0
-} else {
-  tmp$n
-}
+dyn_evp <- if (nrow(tmp) == 0) { 0 } else { tmp$n }
 tmp <- sites_per_variant %>% filter(ev_variant == "local")
-dyn_loc <- if (nrow(tmp) == 0) {
-  0
-} else {
-  tmp$n
-}
+dyn_loc <- if (nrow(tmp) == 0) { 0 } else { tmp$n }
 
@@
 ```{r ev-dist, include=FALSE}
+
 # The calls per package
@@
 
-```{r eval-calls-violin-plot}
+```{r eval-calls-violin-plot, include=F}
 sites_per_package %>%
@@
   labs(y = "Eval calls per package") +
-  theme(legend.position = "none")
+  theme(legend.position = "none") -> G
 
@@
 
-```{r}
+```{r include=F}
 lbl_colour <- rep("black", 11)
@@
   theme(text = element_text(size=17),
-        axis.text.x = element_text(color = lbl_colour))
+        axis.text.x = element_text(color = lbl_colour)) -> G
 
@@
   theme(text = element_text(size=17),  panel.grid.major.x = element_blank(),
-        panel.grid.minor.x = element_blank(), aspect.ratio = 4/3)
+        panel.grid.minor.x = element_blank(), aspect.ratio = 4/3) -> GG
 
@@
 
-```{r}
+```{r include=F}
 E %>%
@@
   labs(y = "", x = "") +
-  theme(legend.position = "none", text = element_text(size=25))
+  theme(legend.position = "none", text = element_text(size=25)) -> G
 
@@
   labs(y = "Instructions", x = "") +
-  theme(legend.position = "none", text = element_text(size=25))
+  theme(legend.position = "none", text = element_text(size=25)) -> G
 
@@
   arrange(desc(average))
-```
 
-```{r}
 # Plot the low trigger counts...
@@
   geom_histogram(binwidth = 1, fill = "blue", alpha = 0.3, colour = "black") +
-  theme(text = element_text(size=16), axis.text = element_text(size=20), aspect.ratio = 4/3)
+  theme(text = element_text(size=16), axis.text = element_text(size=20), aspect.ratio = 4/3) -> G
 
@@
 
-```{r type-diversity}
+```{r type-diversity, include=F}
 E %>%
@@
   coord_flip() +
-  labs(x = "type")
+  labs(x = "type") -> G
 
@@
 
-ggplot2_symbs <- E %>% filter(ev_package == "ggplot2") %>% count(type, wt=duplicates) %>% filter(type == "symbol") %>% .$n
+ggplot2_symbs <- E %>% filter(ev_package == "ggplot2") %>% 
+  count(type, wt=duplicates) %>% filter(type == "symbol") %>% .$n
 

### 2020-12-21 analysis/usage.Rmd
usage that ignores missing col

@@
   base_dir: Data/
-  calls_packages_path: package/summarized.fst
+  calls_path: base/summarized.fst
   corpus_path: corpus.fst
-  static_evals_path: package-evals-static.csv
+  static_path: package-evals-static.csv
+  force_rebuild: TRUE
 ---
@@
 base_dir <- params$base_dir
-eval_calls_path <- paste0(base_dir, params$calls_packages_path)
+calls_path <- paste0(base_dir, params$calls_path)
 corpus_path <- paste0(base_dir, params$corpus_path)
-eval_static_path <- paste0(base_dir, params$static_evals_path)
+static_path <- paste0(base_dir, params$static_path)
 ```
@@
 read_fst(corpus_path) %>% as_tibble() -> C_raw
-read_csv(eval_static_path) %>% semi_join(C_raw, by = "package") -> P
+read_csv(static_path) %>% semi_join(C_raw, by = "package") -> P
 
@@
   saved <- read.fst("Data/.lock.fst")[[1,1]]
-  if (saved == file.size(eval_calls_path)) {
+  if (saved == file.size(calls_path)) {
     read.fst("Data/E.fst") %>% as_tibble() -> E
@@
 }
-if (rebuild) {
-  read_fst(eval_calls_path) %>% as_tibble() -> E_raw
+if (params$force_rebuild | rebuild) {
+  read_fst(calls_path) %>% as_tibble() -> E_raw
+
+  if (!"direct_interpreter_eval" %in% names(E_raw)) { # We are dealing with
+    E_raw %>% mutate(direct_interpreter_eval=0) -> E_raw # partial data.
+  } #                                                  We create a fake column.
+
   E_raw %>% select(
@@
   write.fst(E, "Data/E.fst")
-  sz <- file.size(eval_calls_path) %>% as_tibble()
+  sz <- file.size(calls_path) %>% as_tibble()
   write.fst(sz,"Data/.lock.fst")

### 2020-12-21 analysis/usage.Rmd
minor

@@
   base_dir: Data/
-  calls_path: base/summarized.fst
+  calls_path: package/summarized.fst
   corpus_path: corpus.fst
   static_path: package-evals-static.csv
-  force_rebuild: TRUE
+  force_rebuild: FALSE
 ---
@@
 
-base_dir <- params$base_dir
-calls_path <- paste0(base_dir, params$calls_path)
-corpus_path <- paste0(base_dir, params$corpus_path)
-static_path <- paste0(base_dir, params$static_path)
+calls_path <- paste0(params$base_dir, params$calls_path)
+corpus_path <- paste0(params$base_dir, params$corpus_path)
+static_path <- paste0(params$base_dir, params$static_path)
 ```
@@
 rebuild <- TRUE
-# We only read the orignal evals_calls file if we don't have a summarized
-# version stored on disk.  NOTE: the heuristic to detect changes is file
-# size. If that fails, manually delete `.lock.fst`
+# We only read the file if we don't have our summarized file on disk.
+# To force reading, set the force_rebuild parameter to true.
 if (file.exists("Data/.lock.fst")) {
@@
 (`r fmt(ratio(evalq_calls,nb_calls))`) calls to `evalq`, `r eval.parent_calls` 
-(`r fmt(ratio(evalq_calls,nb_calls))`) calls to `eval.parent`,  and
-`r local_calls` (`r fmt(ratio(evalq_calls,nb_calls))`) calls to `local`.
+(`r fmt(ratio(eval.parent_calls,nb_calls))`) calls to `eval.parent`,  and
+`r local_calls` (`r fmt(ratio(local_calls,nb_calls))`) calls to `local`.
 
@@
 
+`r TBL`
+
 # Numbers of calls to eval
@@
 
-```{r eval-calls-violin-plot, include=F}
+```{r eval-calls-violin-plot}
 sites_per_package %>%
@@
   labs(y = "Eval calls per package") +
-  theme(legend.position = "none") -> G
+  theme(legend.position = "none")
 
@@
 
-```{r include=F}
+```{r echo=FALSE}
 lbl_colour <- rep("black", 11)
@@
   theme(text = element_text(size=17),
-        axis.text.x = element_text(color = lbl_colour)) -> G
+        axis.text.x = element_text(color = lbl_colour))
 
@@
   theme(text = element_text(size=17),  panel.grid.major.x = element_blank(),
-        panel.grid.minor.x = element_blank(), aspect.ratio = 4/3) -> GG
+        panel.grid.minor.x = element_blank(), aspect.ratio = 4/3)
 
@@
 
-```{r include=F}
+`r TBL`
+
+```{r echo=FALSE}
 E %>%
@@
   labs(y = "", x = "") +
-  theme(legend.position = "none", text = element_text(size=25)) -> G
+  theme(legend.position = "none", text = element_text(size=25))
 
@@
 
-```{r include=F}
+```{r echo=FALSE}
 E %>%
@@
   labs(y = "Instructions", x = "") +
-  theme(legend.position = "none", text = element_text(size=25)) -> G
+  theme(legend.position = "none", text = element_text(size=25))
 
@@
 
-```{r loops, include=FALSE}
+```{r loops, echo=FALSE}
 # Which call sites where exercised and how many times, per run.
@@
   geom_histogram(binwidth = 1, fill = "blue", alpha = 0.3, colour = "black") +
-  theme(text = element_text(size=16), axis.text = element_text(size=20), aspect.ratio = 4/3) -> G
+  theme(text = element_text(size=16), 
+        axis.text = element_text(size=20), 
+        aspect.ratio = 4/3)
 
@@
 
-```{r type-diversity, include=F}
+```{r type-diversity, echo=FALSE}
 E %>%
@@
   coord_flip() +
-  labs(x = "type") -> G
+  labs(x = "type")
 

### 2020-12-21 analysis/usage.Rmd
minor

@@
   base_dir: Data/
-  calls_path: package/summarized.fst
+  calls_path: base/summarized.fst
   corpus_path: corpus.fst
   static_path: package-evals-static.csv
-  force_rebuild: FALSE
+  force_rebuild: TRUE
 ---

### 2020-12-22 analysis/environments.Rmd
Decompose environment class

@@
 params:
-  base_dir: /var/lib/R/project-evalR
-  calls_packages_path: /run/package-evals-traced.10/summarized-packages.fst
+  base_dir: Data
+  calls_packages_path: /package/summarized.fst
 ---
@@
 
-source("insights.R", local = knitr::knit_global())
+source("inc/insights.R", local = knitr::knit_global())
 ```
@@
   as_tibble()
+
+stopifnot(!is.na(eval_calls$eval_call_srcref))
 ```
@@
 
+Now, we can further decompose the environment class to get more insights. It has three components:
+
+- how many times it has been wrapped in `new`
+- how far (in terms of hierarchy of environments) it is from the callee
+- does it correspond to any distinguished environment? ( `global`, `empty`, `base`, `callee`, or some package environment
+
+```{r}
+# new+new+new+new+caller-2-global
+# str_match("caller-2-", "((?:new\\+)*)caller-([[:digit:]]+)-(.*)")
+
+env_calls <- eval_calls_absolute %>%
+  extract(environment_class, regex = "((?:new\\+)*)caller-([[:digit:]]+)-(.*)", into = c("new_env", "hierarchy", "specific_env"), remove = FALSE, convert = TRUE) %>%
+  mutate(new_env = na_if(new_env, ""), specific_env = na_if(specific_env, ""), new_env = str_count(new_env, fixed("new+"))) %>%
+  mutate(specific_env = if_else(specific_env == "emptyR_EmptyEnv", "empty", specific_env))
+```
+
 # `envir` is an integer

### 2020-12-29 analysis/environments.Rmd
Renamed setup.R to functions.R

@@
 source("inc/latextags.R")
-source("inc/setup.R")
+source("inc/functions.R")
 source("inc/paths.R")

### 2020-12-29 analysis/environments.Rmd
Environments and ggplot

--- 2020-12-22 analysis/environments.Rmd
@@
 
-env_calls <- eval_calls_absolute %>%
+env_calls <- eval_calls_absolute %>% ungroup() %>%
   extract(environment_class, regex = "((?:new\\+)*)caller-([[:digit:]]+)-(.*)", into = c("new_env", "hierarchy", "specific_env"), remove = FALSE, convert = TRUE) %>%
@@
   mutate(specific_env = if_else(specific_env == "emptyR_EmptyEnv", "empty", specific_env))
+```
+
+_ggplot2_ again is responsible for this `new.env` 4 times!
+
+```{r}
+env_calls %>% filter(!is.na(new_env), specific_env == "global") %>% count(new_env, wt = nb_ev_calls, sort = TRUE)
+
+env_calls %>% filter(new_env == 4, specific_env == "global") %>% count(eval_source, wt = nb_ev_calls, sort = TRUE)
 ```

### 2020-12-29 analysis/events.Rmd

--- 2020-08-27 inst/analysis/events.Rmd
@@
 source("inc/paths.R")
-source("inc/setup.R")
+source("inc/funtions.R")
 source("inc/latextags.R")
@@
 
-all_data <- rbind(core_data, package_data, kaggle_data) %>% filter(eval_source != "testthat")
+all_data <- rbind(core_data, package_data, kaggle_data)
 ```

### 2020-12-29 analysis/usage.Rmd

@@
 source("inc/latextags.R")
-source("inc/setup.R")
+source("inc/functions.R")
 theme_set(theme_minimal())

### 2021-01-01 analysis/corpus.Rmd
Add 3.1

--- 2020-08-28 inst/analysis/corpus-analysis.Rmd
@@
 ---
-title: "Corpus Analysis"
-output: html_document
-params:
-  base_dir: ../../../
+title: "Usage metrics"
+output:
+    html_document:
+        gallery: false
+        toc: true
+        toc_depth: 3
+        toc_float: true
+        df_print: paged
 editor_options: 
   chunk_output_type: console
+params:
+  base_dir: Data/
 ---
@@
 
-# TODO: move into the package
+source("inc/functions.R")
+source("inc/latextags.R")
 source("inc/paths.R")
-source("inc/setup.R")
-source("inc/latextags.R")
 
-options(repos="https://cran.r-project.org")
+create_tags(path(TAGS_DIR, "corpus.tex"), prefix = "", default = TRUE)
 theme_set(theme_minimal())
-create_tags(path(TAGS_DIR, "corpus.tex"), prefix="Corpus", default=TRUE)
 ```
@@
 ```{r load corpus}
-evals_dynamic <- read_fst(PACKAGE_EVALS_DYNAMIC_FILE) %>% as_tibble() %>% rename(evals_dyn=n)
+CRAN_eval_traces <- 
+  read_fst(PACKAGE_SUM_CALLS_FILE) %>%
+  # remove evals that are inside the scripts,
+  # for CRAN we are only interested in evals in packages
+  filter(!str_starts(eval_call_srcref, "::global::main")) %>% 
+  as_tibble()
 
-corpus <- 
-  read_fst(CORPUS_DETAILS_FILE) %>%
-  as_tibble() %>%
-  left_join(evals_dynamic, by="package") %>%
-  mutate_at(
-    vars(
-      starts_with("runnable_code"),
-      starts_with("runnable_files"),
-      starts_with("package_"),
-      starts_with("funs_"),
-      "evals",
-      "evals_dyn"
-    ), 
-    replace_na, 
-    0
+CRAN_eval_traces_per_package <- 
+  count(CRAN_eval_traces, caller_package, eval_call_srcref) %>%
+  count(caller_package) %>% 
+  rename(package=caller_package, evals_hit=n)
+
+ALL_CRAN <- 
+  read_fst(CORPUS_ALL_FILE) %>%
+  mutate_if(is.numeric, replace_na, 0) %>%
+  mutate(evals=as.integer(evals)) %>% 
+  filter(loadable, package_code > 0) %>% 
+  as_tibble()
+
+CRAN <- 
+  read_fst(CORPUS_FILE) %>%
+  filter(package_code > 0) %>%
+  left_join(CRAN_eval_traces_per_package, by="package") %>% 
+  mutate_if(is.numeric, replace_na, 0) %>%
+  mutate(
+    evals=as.integer(evals),
+    evals_hit=as.integer(evals_hit),
+    # this is the case when there is eval in package loading code
+    # this we will always run
+    evals_hit=if_else(evals_hit > evals, evals, evals_hit),
+    evals_coverage=evals_hit/evals
+  ) %>% 
+  as_tibble()
+
+ALL_CRAN_eval_callsites <- 
+  read_csv(PACKAGE_EVALS_STATIC_FILE) %>% 
+  # TODO: There are very few NA, plus when srcref is NA, the rest of the row in NA
+  # we should look at this
+  filter(!is.na(srcref))
+
+CRAN_eval_callsites <- semi_join(ALL_CRAN_eval_callsites, CRAN, by="package")
+
+ALL_KAGGLE <- local({
+  kernels <- read_csv(KAGGLE_KERNEL_FILE)
+  evals <- read_csv(KAGGLE_EVALS_STATIC_FILE) %>% count(package) %>% rename(evals=n)
+  left_join(kernels, evals, by="package") 
+}) %>%
+  mutate_if(is.numeric, replace_na, 0) %>% 
+  mutate(evals=as.integer(evals)) %>% 
+  as_tibble()
+
+KAGGLE <- filter(ALL_KAGGLE, evals > 0)
+```
+
+## Overview
+
+```{r}
+cran_all_funs <- sum(ALL_CRAN$funs_public) + sum(ALL_CRAN$funs_private)
+cran_funs <- sum(CRAN$funs_public) + sum(CRAN$funs_private)
+cran_funs_with_eval <- sum(CRAN$funs_with_eval)
+cran_code <- sum(CRAN$package_code)
+cran_all_code <- sum(ALL_CRAN$package_code)
+cran_evals <- CRAN$evals
+cran_evals_hit <- CRAN$evals_hit
+
+overview_table(
+  r("CRAN all packages", ALL_CRAN),
+  r("CRAN packages", CRAN),
+  r("CRAN packages ratio", ratio(CRAN, ALL_CRAN)),
+  r("CRAN all code", cran_all_code),
+  r("CRAN all r code", sum(ALL_CRAN$package_r_code)),
+  r("CRAN all native code", sum(ALL_CRAN$package_native_code)),
+  r("CRAN code", cran_code),
+  r("CRAN code ratio", ratio(cran_code, cran_all_code)),
+  r("CRAN all funs", cran_all_funs),
+  r("CRAN funs", cran_funs),
+  r("CRAN funs with eval", cran_funs_with_eval),
+  r("CRAN funs with eval ratio", ratio(cran_funs_with_eval, cran_funs)),
+  r("CRAN all funs with eval ratio", ratio(cran_funs_with_eval, cran_all_funs)),
+  r("CRAN eval call sites", sum(cran_evals)),
+  r("CRAN hit eval call sites", sum(cran_evals_hit)),
+  r("CRAN hit eval call sites avg ratio", percent(mean(cran_evals_hit/cran_evals))),
+  r("CRAN runnable scripts", sum(ALL_CRAN$runnable_files)),
+  r("CRAN runnable code", sum(ALL_CRAN$runnable_code)),
+  r("CRAN runnable code examples", sum(ALL_CRAN$runnable_code_examples)),
+  r("CRAN runnable code tests", sum(ALL_CRAN$runnable_code_tests)),
+  r("CRAN runnable code vignettes", sum(ALL_CRAN$runnable_code_vignettes)),
 )
@@
 
-## CRAN packages
+### Kaggle
 
-```{r filter out inconsistent programs}
-corpus_programs <-
-  read_fst(CRAN_PROGRAMS_FILE) %>%
-  as_tibble()
+```{r}
 ```
 
-```{r summary}
-runnable_code <- 
-  corpus_programs %>%
-  mutate(files=1) %>%
-  pivot_wider(names_from=type, values_from=c(files, code)) %>%
-  select(-file, -runtime, -error, -success, -corpus) %>%
-  group_by(package) %>%
-  summarise_all(sum, na.rm=TRUE) %>%
-  group_by(package) %>%
-  mutate(
-    files=files_examples + files_vignettes,
-    code=code_examples + code_vignettes
-  ) %>%
-  ungroup()
-
-corpus_program_examples <- filter(corpus_programs, type=="examples")
-corpus_programs_tests <- filter(corpus_programs, type=="tests")
-corpus_programs_vignettes <- filter(corpus_programs, type=="vignettes")
+```{r}
+kaggle <- distinct(ALL_KAGGLE, hash, .keep_all=TRUE)
+kaggle_code <- sum(kaggle$code)
+kaggle_evals <- filter(kaggle, evals > 0)
+kaggle_evals_code <- sum(kaggle_evals$code)
 
 overview_table(
-  r("packages",    corpus),
-  rs("revdeps", corpus$revdeps),
-  r("in cranlogs", filter(corpus, !is.na(rank))),
-
-  rs("package programs",         runnable_code$files),
-  rs("package programs code",    runnable_code$code),
-
-  rs("examples programs",  runnable_code$files_examples),
-  rs("vignettes programs", runnable_code$files_vignettes),
-
-  rs("examples code",  corpus_program_examples$code),
-  rs("vignettes code", corpus_programs_vignettes$code),
-
-  r("mean expr coverage",   percent(mean(corpus$coverage_expr))),
-  r("median expr coverage", percent(median(corpus$coverage_expr))),
-
-  r("with R code",      filter(corpus, package_r_code > 0)),
-  r("with native code", filter(corpus, package_native_code > 0)),
-
-  rs("code",        corpus$package_code),
-  rs("R code",      corpus$package_r_code),
-  rs("native code", corpus$package_native_code),
-
-  rs("functions",          corpus$funs_public + corpus$funs_private),
-  rs("function with eval", corpus$funs_with_eval),
-  rs("public functions",   corpus$funs_public),
-  rs("private functions",  corpus$funs_private),
-
-  rs("revdes", corpus$revdeps)
+  r("Kaggle kernels", ALL_KAGGLE),
+  r("Kaggle duplicates", nrow(ALL_KAGGLE)-nrow(kaggle)),
+  r("Kaggle unique", kaggle),
+  r("Kaggle code", kaggle_code),
+  r("Kaggle competitions", count(kaggle, competition)),
+  r("Kaggle with evals", kaggle_evals),
+  r("Kaggle with evals code", ratio(kaggle_evals_code, kaggle_code))
 )
@@
 
-```{r generals}
-overview_table(
-  r("all cran", dim(available.packages())[1]),
-  r("core packages", length(CORE_PACKAGES)),
-  r("most downloaded treshold", MOST_DOWNLOADED),
-  r("large package treshold", PACKAGE_SIZE_OUTLIER),
-  r("evals package treshold", PACKAGE_EVALS_OUTLIER)
-)
+### Matching evals
+
+```{r}
+matched_eval_callsites <-
+  left_join(
+    CRAN_eval_callsites, 
+    count(CRAN_eval_traces, eval_call_srcref), 
+    by=c("srcref"="eval_call_srcref")
+  ) %>%
+  mutate(
+    n=replace_na(n, 0)
+  ) %>%
+  rename(hit=n)
 ```
+
+The dynamic analysis traced `r fmt(ratio(filter(matched_eval_callsites, hit>0), matched_eval_callsites))`
 
@@
 
-```{r plot corpus}
-coverage_mean   <- mean(corpus$coverage_expr)
-coverage_median <- median(corpus$coverage_expr)
-revdeps_mean    <- mean(corpus$revdeps)
-revdeps_median  <- median(corpus$revdeps)
-funs_with_eval_mean    <- mean(corpus$funs_with_eval)
-funs_with_eval_median  <- median(corpus$funs_with_eval)
+```{r}
+eval_call_sites_hist <- function(df, df_all, y_lab, max_evals=15L) {
+  df %>% 
+    mutate(
+      evals=if_else(evals < max_evals, evals, max_evals)
+    ) %>%
+    ggplot(aes(x=evals)) + 
+    geom_histogram(binwidth=1, color="black", fill="blue", alpha=0.3) +
+    geom_vline(aes(xintercept=mean(df$evals)), linetype="dashed", color="red", size=0.2) +
+    scale_x_continuous(
+      limits=c(0, max_evals + 1), 
+      breaks=as.integer(seq(1, 15, length.out=4)), 
+      labels=function(x) c(head(x, -1), str_c(tail(x, 1), "+"))
+    ) +
+    scale_y_continuous(
+      labels=scales::comma_format(accuracy=1),
+      sec.axis=sec_axis(
+        trans=~./nrow(df_all), 
+        labels=scales::percent_format(accuracy=.1), 
+        name=y_lab)
+    ) +
+    labs(x = "Eval call sites", y=y_lab) +
+    theme(text = element_text(size=17),
+          panel.grid.minor.x = element_blank(), aspect.ratio = 4/3) 
+}
 
-size_bins       <- c(0L, 1e3L, 5e3L, 1e4L, 5e4L, 1e5L)
-size_bins_lasb  <- c("< 1000", "1,000 - 4,999", "5,000 - 9,999", "10,000 - 49,999", "> 50,000")
-
-corpus %>%
-  mutate(
-    outlier=evals_dyn > PACKAGE_EVALS_DYN_OUTLIER,# | package_r_code > PACKAGE_SIZE_OUTLIER,
-    label=ifelse(outlier, package, NA),
-    code_size=cut(package_r_code, size_bins, labels=size_bins_lasb, include.lowest=TRUE)
-  ) %>%
-  ggplot(
-    aes(
-      x=coverage_expr, 
-#      y=funs_with_eval,
-      y=evals_dyn,
-      label=label,
-      size=code_size
-    )
-  ) +
-  geom_point(alpha=.4) +
-  geom_text(size=3, check_overlap = T, vjust=-1.5, na.rm = TRUE) + 
-  geom_hline(aes(yintercept=funs_with_eval_mean), linetype="dashed", color="red", size=0.2) +
-  geom_vline(aes(xintercept=coverage_mean), linetype="dashed", color="red", size=0.2) +
-  scale_x_continuous(labels = scales::percent) + 
-  scale_y_continuous(trans=scales::pseudo_log_trans(), breaks=c(0, 1, 2, 5, 10, 20, 50, 100, 200, 500)) +
-  scale_size_discrete(range = c(1, 6)) +
-  theme(
-    legend.position=c(0.2, 0.85),
-    legend.box="horizontal",
-    legend.box.background = element_rect(fill="white", size=0.1)
-  ) +
-  labs(
-    x="Code coverage", 
-    y="Number of eval call sites", 
-    size="Lines of R code",
-    linetype=""
-  ) +
-  guides(
-    size=guide_legend(order=1),
-    color=FALSE,
-    linetype=FALSE
-  )
-
-ggsave(CORPUS_PLOT)
+eval_call_sites_hist(CRAN, ALL_CRAN, "CRAN packages")
+#ggsave(path(PLOT_DIR, "cran-eval-callsites-hist.pdf"))
+eval_call_sites_hist(KAGGLE, ALL_KAGGLE, "KAGGLE programs")
+#ggsave(path(PLOT_DIR, "kaggle-eval-callsites-hist.pdf"))
 ```
-
-### List of corpus packages
-
-```{r list of packages, eval=FALSE}
-datatable(
-  corpus, 
-  filter="top", 
-  options=list(
-    pageLength=50,
-    autoWidth=TRUE
-  )
-)
-```
-
-## Kaggle
-
-```{r kaggle dependencies}
-kaggle_dependencies <- tibble(package=read_lines(KAGGLE_DEPENDENCIES_FILE))
-```
-
-```{r compute kaggle programs}
-kaggle <- read_fst(KAGGLE_KERNELS_FILE) %>% as_tibble()
-kaggle_programs <-
-  read_fst(KAGGLE_PROGRAMS_FILE) %>% as_tibble() %>% filter(success)
-```
-
-```{r kaggle summary}
-kaggle_code     <- kaggle_programs$code
-
-overview_table(
-  r("kaggle",                 kaggle),
-  r("duplicated kaggle",      nrow(kaggle)-nrow(filter(kaggle, runnable))),
-  r("runnable kaggle",        filter(kaggle, runnable)),
-  r("finished kaggle",        kaggle_programs),
-  r("failed kaggle",          nrow(filter(kaggle, runnable))-nrow(kaggle_programs)),
-  rs("finished kaggle code",  kaggle_code)
-)
-```
-
-```{r}
-setdiff(kaggle_dependencies$package, corpus$package)
-```
-
-## Combined
-
-```{r compute programs}
-programs <- bind_rows(
-  corpus_programs %>% select(code, runtime, corpus),
-  kaggle_programs %>% select(code, runtime, corpus)
-)
-```
-
-```{r all summary}
-overview_table(
-  r("all programs",      nrow(kaggle_programs) + nrow(corpus_programs)),
-  r("all runnbale code", sum(kaggle_code) + sum(runnable_code$code)),
-  r("all code",          sum(kaggle_code) + sum(corpus$package_code) + sum(runnable_code$code))
-)
-```
-
-```{r programs-code, eval=F}
-programs %>%
-  ggplot(aes(y=code, x=corpus, fill=corpus)) + 
-  geom_violin() +
-  scale_y_log10(labels = scales::comma) +
-  scale_x_discrete(labels=c("kaggle"="Kaggle", "packages"="Packages")) +
-  scale_fill_viridis(discrete = TRUE) +
-  guides(fill = FALSE) +
-  labs(y="Code size (log)") +
-  theme(
-    legend.box="horizontal",
-    axis.title.x=element_blank()
-  )
-
-ggsave(path(PLOT_DIR, "programs-code.pdf"), width=7, height=5.5)
-```
-
-```{r programs-runtime, eval=F}
-programs %>%
-  ggplot(aes(y=runtime, x=corpus, fill=corpus)) + 
-  geom_flat_violin(trim=FALSE, width=.5) +
-  geom_dotplot(
-    binaxis="y", 
-    dotsize=0.1, 
-    stackdir="down", 
-    binwidth=0.01, 
-    position=position_nudge(-0.025)) +
-  scale_y_log10(labels = function(x, ...) {
-    min <- as.integer(x/60L)
-    sec <- as.integer(x%%60L)
-    sprintf("%.2d:%.2d", min, sec)
-  }) +
-  scale_x_discrete(labels=c("kaggle"="Kaggle", "packages"="Packages")) +
-  scale_fill_viridis(discrete = TRUE) +
-  guides(fill = FALSE) +
-  labs(y="Running time [mm:ss] (log)") +
-  theme(
-    legend.box="horizontal",
-    axis.title.x=element_blank()
-  )
-ggsave(path(PLOT_DIR, "programs-runtime.pdf"), width=3.5, height=5.0)
-```

### 2021-01-01 paper/main.tex
Polish usage metrics
Add 3.1

@@
 
-Our corpus is assembled out of two set of programs, one consisting of
-popular libraries obtained from the Comprehensive R Archive
-Network\footnote{CRAN is the largest repository of R code with over
-\CorpusAllCranRnd packages. It receives about 6 new package
-submissions a day~\cite{Ligges2017}. Unlike other open code
-repositories such as GitHub, CRAN is a curated repository. Each
-submitted package must abide to a number of well-formedness rules that
-are automatically checked asserting certain quality. Most relevant for
-this work is that all of the runnable code is tested and only a
-successfully running package is admitted in the archive.} (we refer to
-it as {\bf CRAN}) and the other from the Kaggle data science
-competition (written {\bf Kaggle}). The intent here is to contrast
-code written by experienced developpers (CRAN) with code authored by
-typical end users of the language (Kaggle).
-
-All the code in our is runnable, i.e. we have scripts that invoke the
-CRAN packages and the Kaggle programs have input data. Clearly code
-coverage will vary; as is the case for any dynamic analysis, the
-quality of results is predicated on how representative the executions
-we can observe are with respect to all possible runs of the corpus. To
-mitigate the threat to validity attached to limited coverage, we
-obtain clients that will generate multiple \emph{runs} of CRAN
-packages. In this context, a run is the execution of a CRAN package or
-Kaggle program with a particular set of input values.
-
-\mypara{CRAN Packages.} The packages that are included
-in this study are the top \CorpusPackages packages based on their
-reverse dependencies. The set of reverse dependencies for some package
-\emph{P} is transitive closure of packages that import \emph{P}. The
-hypothesis underlying that choice is that packages with a large set of
-reverse depencies are likely to be higher in quality and their tests
-will provide better coverage. The packages are implemented in R with
-some C and Fortran code. Using \c{cloc}, we measure a total
-\CorpusRCodeRnd lines of R code and \CorpusNativeCodeRnd lines of
-native code. For each package, we use its runnable code (tests, etc.)
-to compute code coverage. On average, the cde coverage is
-\CorpusMeanExprCoverage which is acceptable without being exhaustive.
-Figure~\ref{fig:corpus} shows these packages, the size of the dots
-reflects the project's size in lines of code. The x-axis indicates
-code coverage in percents and the y-axis gives the number of call
-sites to \eval that were traced, in log scale. Dotted lines indicate
-means. Packages with over \CorpusEvalsPackageTreshold eval call sites
-are named.
-
-\begin{figure}[!h]\centering\includegraphics[width=.7\linewidth]
-  {corpus.pdf}\caption{CRAN packages}\label{fig:corpus}
-\end{figure}
-
-To analyze a package, we need client code that invokes its methods.
-There are three sources of built-in runnable code that come with each
-CRAN package: \emph{tests, examples} and \emph{vignettes}. They are,
-respectively, traditional unit tests, code snippets from the
-documentation, and long-form use-cases written in Rmarkdown. Examples
-and vignettes are automatically extracted and turned into scripts,
-their input is bundled with the package. Most tests had to be discared
-due to a limitation of our pipeline: the \c{testthat} harness uses
-\eval and thus causes the entire test to register as an \eval call.
-The selected packages are bundled with \CorpusPackagePrograms
-programs; \CorpusExamplesProgramsRnd examples and
-\CorpusVignettesProgramsRnd vignettes.
-
-\mypara{Kaggle Scripts.} Kaggle is an online platform
-for data-science and machine-learning. The website allows people to
-submit data-analysis problems, users compete to find the best
-solution. The solutions are uploaded to the platform as either plain
-scripts or notebooks. We chose one of the most popular competition,
-predicting the survival of passengers the
-Titanic\footnote{\url{https://www.kaggle.com/c/titanic}}. Unlike CRAN,
-Kaggle is not curated. After downloading the \CorpusKaggle solutions
-and extracting the R code, we found that \CorpusDuplicatedKaggle were
-duplicates. From the remaining \CorpusRunnableKaggle solutions,
-\CorpusFailedKaggle failed to execute. Next to various runtime
-exceptions, common problems were parse errors and misspelled package
-names. The final set contains \CorpusFinishedKaggle programs
-implemented in \CorpusFinishedKaggleCodeRnd lines of R code.
+Our corpus is assembled out of two set of programs, one consisting of libraries
+obtained from \emph{CRAN} (Comprehensive R Archive Network\footnote{CRAN is the
+  largest repository of R code with over \CorpusAllCranRnd packages. It receives
+  about 6 new package submissions a day~\cite{Ligges2017}}) and the other from
+the \emph{Kaggle} data science competition website\footnote{\cf
+  \url{https://kaggle.com}}. The intent here is to contrast code written by
+experienced developers (CRAN) with code authored by typical end users of the
+language (Kaggle). Another difference is that CRAN is a curated repository. Each
+submitted package must abide to a number of well-formedness rules that are
+automatically checked asserting certain quality. Most relevant for this work is
+that all of the runnable code is tested and only a successfully running package
+is admitted in the archive. Kaggle on the other hand is more like GitHub, anyone
+can submit any code. All the code in our corpus is runnable, \ie we have scripts
+that invoke the CRAN packages and the Kaggle programs have input data.
+
+% Clearly code coverage will vary; as is the case for any dynamic analysis, the
+% quality of results is predicated on how representative the executions we can
+% observe are with respect to all possible runs of the corpus. To mitigate the
+% threat to validity attached to limited coverage, we obtain clients that will
+% generate multiple \emph{runs} of CRAN packages. In this context, a run is the
+% execution of a CRAN package or Kaggle program with a particular set of input
+% values.
+
+% \begin{figure}[!h]\centering\includegraphics[width=.7\linewidth]
+%   {corpus.pdf}\caption{CRAN packages}\label{fig:corpus}
+% \end{figure}
+
+\mypara{CRAN Packages.} We used \CranAllPackages packages. These are almost all
+of the available packages except for a few that we could not install due to
+missing native dependencies. The packages are implemented in R with native C and
+Fortran code. Using \c{cloc}, we measure a total \CranAllRCodeRnd lines of R
+code and \CranAllNativeCodeRnd lines of native code. Out of these packages,
+\CranPackages (\CranPackagesRatio) call eval. These packages represent
+\CranCodeRatio of the code size.
+
+To analyze a package, we need client code that invokes its methods. There are
+three sources of built-in runnable code that come with each CRAN package:
+\emph{tests, examples} and \emph{vignettes}. They are, respectively, traditional
+unit tests, code snippets from the documentation, and long-form use-cases
+written in \LaTeX\xspace or Markdown with executable snippets of R code.
+Examples and vignettes are automatically extracted and turned into scripts,
+their input is bundled with the package. Tests are directly runnable. Together
+we there are \CranRunnableScripts scripts that contain \CranRunnableCode lines
+of code (\CranRunnableCodeExamplesRnd in examples, \CranRunnableCodeVignettesRnd
+in vignettes and \CranRunnableCodeTestsRnd in tests).
+
+\mypara{Kaggle Scripts.} Kaggle is an online platform for data-science and
+machine-learning. The website allows people to submit data-analysis problems,
+users compete to find the best solution. The solutions, called \emph{kernels},
+are uploaded to the platform as either plain scripts or notebooks. We obtained
+\KaggleKernels kernels. Since Kaggle is not a curated repository, we first had
+to remove duplicates. We used SHA-1 hash to identify \KaggleDuplicates
+whole-file copies. The remaining \KaggleUnique kernels represent unique
+solutions to \KaggleCompetitions Kaggle competition. Together, they contain
+\KaggleCode lines of R code. Out of the \KaggleUniqueRnd kernels, only
+\KaggleWithEvals contain calls to eval.
 
@@
 \begin{figure}[h!]
- \includegraphics[width=1.061\textwidth]{size_loaded_distribution}
+ \includegraphics[width=0.5\textwidth, trim=5.5cm 0 6cm 0, clip]{size_loaded_distribution}
 \caption{Loaded code} \label{fig:sizedistribution}
@@
 \end{tabular}
- \caption{Events per packages} \label{ev}
+ \caption{Instructions per packages} \label{ev}
 \end{figure}
@@
 \begin{itemize}
-\item Figure 5(b) -- make `count` and `average` smaller.
-\item Figure 6: (1) Replace `ast-nodes` with `nodes`
-\item Figure 6: (2) Remove gray lines
-\item Figure 6: increase font size
-\item Figure 6: Try to make the figure less wide. Thinner bars/aspect
-  ratio change.
 \item Figure 7: check that element counts are correct
-\item Figure 7: replace events by instructions. Check what we are
-  counting? Are these bytecodes or AST nodes.
 \item Check that the percentage of operations performed within

### 2021-01-01 analysis/inc/paths.R

@@ -2,14 +2,25 @@
 # PATHS
 ################################################################################
 
-DATA_DIR  <- path("Data")
+DATA_DIR  <- params$base_dir
+PACKAGE_DATA_DIR  <- path(DATA_DIR, "package")
+KAGGLE_DATA_DIR  <- path(DATA_DIR, "kaggle")
+
 PAPER_DIR <- path("..", "paper")
-# RUN_DIR   <- path(params$base_dir, "run")
 PLOT_DIR  <- path(PAPER_DIR, "img")
 TAGS_DIR  <- path(PAPER_DIR, "tag")
 
 if (!dir_exists(PLOT_DIR)) dir_create(PLOT_DIR)
 if (!dir_exists(TAGS_DIR)) dir_create(TAGS_DIR)
+
+CORPUS_ALL_FILE <- path(DATA_DIR, "corpus-all.fst")
+CORPUS_FILE <- path(DATA_DIR, "corpus.fst")
+
+PACKAGE_SUM_CALLS_FILE <- path(PACKAGE_DATA_DIR, "summarized.fst")
+PACKAGE_EVALS_STATIC_FILE <-path(PACKAGE_DATA_DIR, "evals-static.csv")
+
+KAGGLE_KERNEL_FILE <- path(KAGGLE_DATA_DIR, "kernel.csv")
+KAGGLE_EVALS_STATIC_FILE <- path(KAGGLE_DATA_DIR, "evals-static.csv")
 
 #R_DIR      <- path(params$base_dir, "R-4.0.2")
 #R_LIBS_DIR <- path(params$base_dir, "library", "4.0")

### 2021-01-01 analysis/corpus.Rmd
Fixes in 3.1

@@
 
+### Cran
+
 ```{r}
@@
 cran_evals_hit <- CRAN$evals_hit
+cran_runnable_code <- sum(ALL_CRAN$runnable_code)
 
 overview_table(
+  r("CRAN available packages", 16000),#length(available.packages()[,1])),
   r("CRAN all packages", ALL_CRAN),
@@
   r("CRAN runnable scripts", sum(ALL_CRAN$runnable_files)),
-  r("CRAN runnable code", sum(ALL_CRAN$runnable_code)),
+  r("CRAN runnable code", cran_runnable_code),
   r("CRAN runnable code examples", sum(ALL_CRAN$runnable_code_examples)),
@@
 ### Kaggle
-
-```{r}
-```
 
@@
   r("Kaggle with evals code", ratio(kaggle_evals_code, kaggle_code))
+)
+```
+
+### All
+
+```{r}
+overview_table(
+  r("Corpus code", kaggle_code + cran_runnable`_code)
 )

### 2021-01-01 paper/main.tex
Fixes in 3.1

@@
   programming in JavaScript. Dynamic analysis of a corpus of
-  \CorpusAllCodeRnd lines of libraries and end-user code confirms that
+  \CorpusCodeRnd lines of libraries and end-user code confirms that
   \eval is indeed in widespread use; R's \eval is more dangerous in
@@
 community. We distinguish between \emph{CRAN packages}
-(\CorpusPackages curated packages that pass stringent quality checks
+(\CranAllPackages curated packages that pass stringent quality checks
 and are equipped with tests and sample data) and \emph{Kaggle scripts}
-(\CorpusFinishedKaggle end-user written programs that performs a
+(\KaggleUnique end-user written programs that performs a
 particular data analysis task). It is reasonable to expect that \eval
@@
 obtained from \emph{CRAN} (Comprehensive R Archive Network\footnote{CRAN is the
-  largest repository of R code with over \CorpusAllCranRnd packages. It receives
-  about 6 new package submissions a day~\cite{Ligges2017}}) and the other from
-the \emph{Kaggle} data science competition website\footnote{\cf
+  largest repository of R code with almost \CranAvailablePackagesRnd packages.
+  It receives about 6 new package submissions a day~\cite{Ligges2017}}) and the
+other from the \emph{Kaggle} data science competition website\footnote{\cf
   \url{https://kaggle.com}}. The intent here is to contrast code written by

### 2021-01-02 analysis/usage.Rmd
direct_interpreter_eval -> interp_eval

@@
 params:
-  base_dir: Data/
-  calls_path: base/summarized.fst
-  corpus_path: corpus.fst
-  static_path: package-evals-static.csv
+  base_dir: Data/package/
+  calls_path: summarized.fst
+  corpus_path: ../corpus.fst
+  static_path: evals-static.csv
   force_rebuild: TRUE
@@
 
-  if (!"direct_interpreter_eval" %in% names(E_raw)) { # We are dealing with
-    E_raw %>% mutate(direct_interpreter_eval=0) -> E_raw # partial data.
-  } #                                                  We create a fake column.
-
   E_raw %>% select(
@@
     type = expr_resolved_type_tag,   #   Argument type
-    opcodes = direct_interpreter_eval #  Count of the AST nodes traversed by the
+    opcodes = interp_eval #  Count of the AST nodes traversed by the
   ) %>% #                                interpreter during eval (non-recursive).

### 2021-01-03 paper/main.tex
Fix number in abstract
Update 3.2

@@
   programming in JavaScript. Dynamic analysis of a corpus of
-  \CorpusCodeRnd lines of libraries and end-user code confirms that
+  \CranAllPackages libraries and \KaggleCode end-user code confirms that
   \eval is indeed in widespread use; R's \eval is more dangerous in
@@
   largest repository of R code with almost \CranAvailablePackagesRnd packages.
-  It receives about 6 new package submissions a day~\cite{Ligges2017}}) and the
-other from the \emph{Kaggle} data science competition website\footnote{\cf
+  It receives about 6 new package submissions a day~\cite{Ligges2017}. \cf
+  \url{https://cran.r-project.org/web/packages/index.html}}) and the other from
+the \emph{Kaggle} data science competition website\footnote{\cf
   \url{https://kaggle.com}}. The intent here is to contrast code written by
@@
 
-The results presented in this paper are the result of an automated
-analysis pipeline that acquires the code of packages, extract
-metadata, executes programs, traces their behavior and summarizes the
-observations. Figure~\ref{fig:pipeline} shows the main steps of the
-pipeline along with approximate time to execute each step, the data
-size, and the number of elements manipulated by the stage. Timings are
-for runs on an Intel Xeon 6140, 2.30GHz with 72 cores and 256GB of
-RAM.
+The results presented in this paper are the result of an automated analysis
+pipeline that acquires the code of packages, extract metadata, executes
+programs, traces their behavior and summarizes the observations.
+Figure~\ref{fig:pipeline} shows the main steps of the pipeline along with
+approximate time to execute each step, the data size, and the number of elements
+manipulated by the stage. Timings are for runs on a cluster of three servers
+with Intel Xeon 6140, 2.30GHz with 72 cores and 256GB of RAM using a shared OCFS
+network storage.
 
@@
   \includegraphics[width=1.05\linewidth]{pipeline.pdf}
-  \caption{Analysis Pipeline}\label{fig:pipeline}
+  \caption{Analysis Pipeline UPDATE!!}\label{fig:pipeline}
 \end{figure}
 
-The first step in the pipeline consist of downloading CRAN packages
-along with their dependencies and acquiring Kaggle programs with the
-help of a web crawler. The second step, is to compute code size and
-coverage metrics for the CRAN packages (we use the \covr package for
-coverage). The third step consist in extracting runnable programs from
-packages: i.e. the tests, examples and vignettes. Each extracted
-program is wrapped into a call to our dynamic analyzer --- the tool is
-called \emph{evil} for \underline{ev}al \underline{i}nspection
-\underline{l}ibrary. This step is needed to ensure that we record
-\eval usage only for the target package. Without this, the data would
-include eval calls from the unit testing frameworks as well as from
-bootstrapping R virtual machine itself. To avoid any interference,
-each program is run in its own R instance.
-The fourth step in the pipeline is to perform dynamic analysis for
-each run of a CRAN package or Kaggle program.
-
-The dynamic analyzer builds upon the dynamic analysis framework,
-\instrumentr that we have implemented to enable us to write dynamic
-analysis logic in R. \instrumentr serves as an intermediary between
-\rdyntrace and \evil, it intercepts the hooks exposed by \rdyntrace
-and attaches R functions exported by \evil as callbacks. The \evil
-callbacks execute on corresponding interpreter events.
-
-The data extracted by \evil from each program is
-concatenated, cleaned and summarized in the post-processing phase by
-custom R scripts. Finally, the summarized data is analyzed in
-RMarkdown notebooks to gather insights. Apart from the figures, the
-data points included in the paper are also generated by RMarkdown
-notebooks as latex macros.
-
-The \emph{evil} framework is implemented as a R package in 2K lines of
-R and 400 lines of C++ code. \instrumentr is an R package implemented
-in 2.5K lines of R and 6K lines of C++ code. It internally uses a
-modified R interpreter, \rdyntrace~\cite{oopsla19a}, that exposes
-hooks from within the interpreter implementation for events of
-interest.
-
-All steps of this pipeline are parallelized using GNU
-parallel~\cite{GNUparallel} and orchestrated by GNU make. To schedule
-and parallelize extraction and analysis of programs, we use the \runr
-package. Furthermore, \runr gracefully handles and reports failures
-across large-scale program runs which greatly aids debugging of the
-analysis pipeline.
-
-\mypara{Limitations.} Dynamic analysis can only observe
-calls that are triggered by the program's input. We believe that
-focusing on R packages with high code coverage does mitigate this to
-some extent. The results we report here were obtained with R's
-bytecode compiler turned off, this should not affect the results as
-the compiler does not optimize \eval.
-
-We turn off the bytecode compiler for this study. The bytecode
-compiler can also call \eval. We do not get source locations for
-\UndefinedEvalsRnd \eval calls. In these cases \eval is either passed
-as an argument to a higher-order functions or is defined in a function
-returned by a higher-order function and the R parser does not retain
-location information for \eval. However, this is a meager
-\PercentUndefinedEval of all \eval calls and is unlikely to affect our
-analysis. We ignore calls to the native \eval function exposed by R.
-We also ignore the \c{rlang::tidy_eval} function which uses native
-\eval internally because \c{rlang} is used to implement a DSL for data
-analysis in R. It introduces a new first-class promise object called
-\c{quosure} for which it implements special evaluation support in
-\c{tidy_eval}.
-
-\newpage
+\noindent The pipeline has the following steps:
+\begin{compactenum}
+  \item \emph{Download.} The first step consists of downloading and installing
+  CRAN packages and acquiring Kaggle corpus. We try to install all available
+  CRAN packages. This is somewhat complicated by the fact that many of packages
+  depend on native libraries, but these dependencies are not explicitly stated
+  in the package description and thus it is hard to automate. For Kaggle we use
+  a web crawler to get the programs and the a Kaggle official command line tool
+  to download the datasets.
+  \item \emph{Runnable code extraction.} Next step is to extract runnabale code.
+  For CRAN package we use \genthat~\cite{issta18}, a tool which converts
+  runnable code snippets from package documentation and vignettes as well as R
+  tests into standalone executable scripts. For Kaggle programs that are not
+  already R scripts, we use \c{rmarkdown} and \c{knitr} R packages to extract
+  the runnable code from the Rmarkdown or the Jupyter notebook formats
+  respectively. Duplicated Kaggle programs are removed. Each of the extracted
+  script is wrapped into a call to our dynamic analyzer so we only record eval
+  usage triggered by running th script. Without this, the data would include
+  eval calls from the unit testing frameworks as well as from bootstrapping R
+  virtual machine itself.
+  \item \emph{Tracing.} With the scripts ready, we run our dynamic analysis that
+  traces the calls to eval and captures other runtime information. For the CRAN
+  packages we run two configurations:
+  \begin{inparaenum}[(1)]
+    \item all of the \CranRunnableScriptsRnd scripts while tracing calls to eval
+    originating in any of the \CranAllPackages installed CRAN package, and
+    \item a random selection of 1/4 of the scripts while tracing only eval calls
+    originating in the core R packages that come bundled with the language.
+  \end{inparaenum}
+  The reason for this split is that the core R packages are intertwined with the
+  language implementation and the usage is different form the user written
+  packages hosted on CRAN. For Kaggle, we only trace the eval calls that
+  originate in the programs themselves. To avoid any interference, each program
+  is run in isolation, in its own R instance.
+  \item \emph{Analysis.} In the final step, the output from the dynamic analysis
+  is merged, cleaned and summarized in the post-processing phase by custom R
+  scripts. The summarized data is then analyzed in RMarkdown notebooks to gather
+  insights. Apart from the figures, the data points included in the paper are
+  also generated by RMarkdown notebooks as \LaTeX macros.
+\end{compactenum}
+
+The code extraction and tracing steps of the pipeline are run in parallel on
+three servers using GNU parallel~\cite{GNUparallel}. To make sure, all three
+servers have identical environment, we use docker images that have all the
+dependencies installed. The pipeline is implemented in 500 lines of Makefile and
+3K lines of auxiliary R and shell scripts. Their main function is to gracefully
+handle and report failures across large-scale program runs which greatly aids
+debugging of the pipeline.
+
+\subsection{Dynamic Analysis}
+
+For this study we need two things: information about calls to eval and related
+functions and low-level data such as object allocation and deallocation and
+side-effects that happen during eval. The former we can get by injecting code to
+the corresponding functions at runtime. For the latter we use
+\rdyntrace~\cite{oopsla19a}. It is a modified R virtual machine based on
+GNU R 4.0.2 that exposes a number of hooks allowing us to get low-level
+information about the R interpreter. \rdyntrace only works with the R
+interpreter and not the byte-code compiler. However, this does not affect the
+result since the compiler does not optimize eval calls.
+
+We use source location to uniquely identify eval call sites. R only retains
+source references at the top-level block expressions. A nested call such as
+\c{try(eval(x))} does not have a source reference for the \c{eval(x)}. To work
+around this, we impute source locations for eval call sites at runtime, by
+traversing functions' AST and attaching source reference attribute manually.
+
+The dynamic analysis is implemented as a R package in 1.5K lines of R code and
+2K lines of C++.
+
+\mypara{Limitations.}
+%
+% We turn off the bytecode compiler for this study. The bytecode
+% compiler can also call \eval.
+%
+We do not get source locations for \UndefinedEvalsRnd UPDATE!! \eval calls. In
+these cases \eval is either passed as an argument to a higher-order functions or
+is defined in a function returned by a higher-order function, or it can be a
+generated called from package native code. However, this is a meager
+\PercentUndefinedEval UPDATE!! of all \eval calls and is unlikely to affect our
+analysis. We ignore calls to the native \eval function exposed by R. We also
+ignore the \c{rlang::tidy_eval} function which uses native \eval internally
+because \c{rlang} is used to implement a DSL for data analysis in R. It
+introduces a new first-class promise object called \c{quosure} for which it
+implements special evaluation support in \c{tidy_eval}.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2021-01-04 analysis/corpus.Rmd
Update undefined calls

--- 2021-01-01 analysis/corpus.Rmd
@@
 KAGGLE <- filter(ALL_KAGGLE, evals > 0)
+
+UNDEFINED <- 
+  read_fst(PACKAGE_UNDEFINED_FILE) %>% 
+  as_tibble()
 ```
@@
 ## Overview
+
+### Cran
 
@@
 cran_evals_hit <- CRAN$evals_hit
+cran_runnable_code <- sum(ALL_CRAN$runnable_code)
+cran_eval_calls <- sum(CRAN_eval_traces$nb_ev_calls)
 
 overview_table(
+  r("CRAN available packages", 15962), #length(available.packages()[,1])), # from the CRAN mirror
   r("CRAN all packages", ALL_CRAN),
@@
   r("CRAN runnable code vignettes", sum(ALL_CRAN$runnable_code_vignettes)),
+  r("CRAN undefined", UNDEFINED),
+  r("CRAN undefined ratio", ratio(nrow(UNDEFINED), cran_eval_calls))
 )
@@
   r("Kaggle with evals code", ratio(kaggle_evals_code, kaggle_code))
+)
+```
+
+### All
+
+```{r}
+overview_table(
+  r("Corpus code", kaggle_code + cran_runnable_code)
 )

### 2021-01-04 analysis/inc/paths.R

@@ -17,6 +17,7 @@ CORPUS_ALL_FILE <- path(DATA_DIR, "corpus-all.fst")
 CORPUS_FILE <- path(DATA_DIR, "corpus.fst")
 
 PACKAGE_SUM_CALLS_FILE <- path(PACKAGE_DATA_DIR, "summarized.fst")
+PACKAGE_UNDEFINED_FILE <- path(PACKAGE_DATA_DIR, "undefined.fst")
 PACKAGE_EVALS_STATIC_FILE <-path(PACKAGE_DATA_DIR, "evals-static.csv")
 
 KAGGLE_KERNEL_FILE <- path(KAGGLE_DATA_DIR, "kernel.csv")

### 2021-01-04 analysis/usage.Rmd
na.rm = TRUE for the opcode computations

@@
 tot_events <- E %>%
-  summarise(n = sum(opcodes * duplicates)) %>%
+  summarise(n = sum(opcodes * duplicates, na.rm = TRUE)) %>%
   .$n
 max_events <- E %>%
-  summarize(n = max(opcodes)) %>%
+  summarize(n = max(opcodes, na.rm = TRUE)) %>%
   .$n

### 2021-01-05 analysis/expressions.Rmd
A notebook to look at eval expressions

--- 2021-01-02 analysis/usage.Rmd
@@
 ---
-title: "Usage metrics"
-author: "Pierre Donat-Bouillud"
+title: "Eval expressions"
 output:
@@
 
-rebuild <- TRUE
-# We only read the file if we don't have our summarized file on disk.
-# To force reading, set the force_rebuild parameter to true.
-if (file.exists("Data/.lock.fst")) {
-  saved <- read.fst("Data/.lock.fst")[[1,1]]
-  if (saved == file.size(calls_path)) {
-    read.fst("Data/E.fst") %>% as_tibble() -> E
-    rebuild <- FALSE
-  }
-}
-if (params$force_rebuild | rebuild) {
-  read_fst(calls_path) %>% as_tibble() -> E_raw
-
-  E_raw %>% select(
-    run_package = package, #      The package which was run to trigger the eval
-    ev_variant = eval_function, # Which eval variant was called
-    duplicates = nb_ev_calls, #   Number of identical eval calls (weight)
-    src_ref = eval_call_srcref, # Source ref for the eval call site
-    file, #                       Script file name (name of this run)
-    ev_package = eval_source, #   Package of the call site
-    arg_size = expr_resolved_length, #   Size in characters of argument 
-    ast_nodes = expr_resolved_nodes, #   Size in AST Nodes
-    type = expr_resolved_type_tag,   #   Argument type
-    opcodes = interp_eval #  Count of the AST nodes traversed by the
-  ) %>% #                                interpreter during eval (non-recursive).
-   mutate(across(where(is_character),as_factor)) -> E 
-  write.fst(E, "Data/E.fst")
-  sz <- file.size(calls_path) %>% as_tibble()
-  write.fst(sz,"Data/.lock.fst")
-}
+E_raw -> E
 
@@
 
-nb_runs <- E %>%
-  pull(file) %>%
-  n_distinct()
-
-overview_table(
-  r("Corpus", corpus_size),
-  r("NbRuns", nb_runs)
-)
 ```
-
-The two data sets are `E` and `C` for conciseness.  `C` is only used here to 
-compute the corpus size. `E` is used throughout. Recall that `E_raw` has the 
-source data with all columns. `P` corresponds to static information about the
-call sites.
-
-The size of our corpus is `r corpus_size` packages.
-
-#  Variants
-
-The data set has `r nrow(E)` distinct calls to the `eval` and its variants. 
-There are `r length(unique(E$ev_package))` corpus packages with `eval` calls
-(note that we may not trigger all `eval`s in a package due to spotty coverage).
-
-```{r variants, include=FALSE}
-nb_calls <- E %>%
-  count(wt = duplicates) %>%
-  .$n
-eval_calls <- E %>%
-  filter(ev_variant == "eval") %>%
-  count(wt = duplicates) %>%
-  .$n
-evalq_calls <- E %>%
-  filter(ev_variant == "evalq") %>%
-  count(wt = duplicates) %>%
-  .$n
-eval.parent_calls <- E %>%
-  filter(ev_variant == "eval.parent") %>%
-  count(wt = duplicates) %>%
-  .$n
-local_calls <- E %>%
-  filter(ev_variant == "local") %>%
-  count(wt = duplicates) %>%
-  .$n
-
-overview_table(
-  r("Allcalls", nb_calls),
-  r("Evals", eval_calls),
-  r("Evalqs", evalq_calls),
-  r("Eparents", eval.parent_calls),
-  r("Locals", local_calls),
-  r("Triggeredpkgs", length(unique(E$ev_package)))
-) -> TBL
-```
-
-We recorded `r nb_calls` to eval, out of these, there were `r evalq_calls`
-(`r fmt(ratio(evalq_calls,nb_calls))`) calls to `evalq`, `r eval.parent_calls` 
-(`r fmt(ratio(eval.parent_calls,nb_calls))`) calls to `eval.parent`,  and
-`r local_calls` (`r fmt(ratio(local_calls,nb_calls))`) calls to `local`.
-
-`r TBL`
-
-# Exercised vs. Static
-
-Here we compute the number of static call sites to compare it with
-the number of call sites with triggered dynamically. The former is larger
-than the latter because code coverage is not perfect.
-
-```{r exercised-vs-static, include=FALSE}
-stat_evl <- P %>% filter(call_fun_name == "eval")        %>% nrow()
-stat_evq <- P %>% filter(call_fun_name == "evalq")       %>% nrow()
-stat_evp <- P %>% filter(call_fun_name == "eval.parent") %>% nrow()
-stat_loc <- P %>% filter(call_fun_name == "local")       %>% nrow()
-
-sites_per_variant <- E %>% # src_ref indicates where the call originates from.
-  group_by(ev_variant) %>%
-  summarize(n = n_distinct(src_ref)) # This the number of distinct sites
-
-tmp <- sites_per_variant %>% filter(ev_variant == "eval")
-dyn_evl <- if (nrow(tmp) == 0) { 0 } else { tmp$n }
-tmp <- sites_per_variant %>% filter(ev_variant == "evalq")
-dyn_evq <- if (nrow(tmp) == 0) { 0 } else { tmp$n }
-tmp <- sites_per_variant %>% filter(ev_variant == "eval.parent")
-dyn_evp <- if (nrow(tmp) == 0) { 0 } else { tmp$n }
-tmp <- sites_per_variant %>% filter(ev_variant == "local")
-dyn_loc <- if (nrow(tmp) == 0) { 0 } else { tmp$n }
-
-overview_table(
-  r("Staticeval", stat_evl),
-  r("Staticevalq", stat_evq),
-  r("Staticevalparent", stat_evp),
-  r("Staticlocal", stat_loc),
-  r("Triggeredeval", dyn_evl),
-  r("Triggeredevalq", dyn_evq),
-  r("Triggeredevalparent", dyn_evp),
-  r("Triggeredlocal", dyn_loc),
-  r("Triggeredevalpct", ratio(dyn_evl, stat_evl)),
-  r("Triggeredevalqpct", ratio(dyn_evq, stat_evq)),
-  r("Triggeredevalparentpct", ratio(dyn_evp, stat_evp)),
-  r("Triggeredlocalpct", ratio(dyn_loc, stat_loc))
-) -> TBL
-```
-
-`r TBL`
-
-# Ev-dist
-
-We compute the frequency of evals by package. This shows that some packages
-hardly use eval while other use it intensly. The numbers are "for all runs",
-this means that a popular package may be triggered by more runs and thus
-have higher frequencies. We will later control for this by looking at individual 
-call sites.
-
-```{r ev-dist, include=FALSE}
-
-# The calls per package
-pack_calls <- E %>% count(ev_package, wt = duplicates, sort = TRUE)
-# Create bins at every power of ten
-bins <- cut(pack_calls$n, breaks = c(0, 10^seq(1, 8)), include.lowest = TRUE)
-sum <- summary(bins)
-# Packages with more than 1K calls
-Kcalls <- pack_calls %>%
-  filter(n > 1000) %>%
-  nrow()
-
-# The number of runs per package
-runs <- E %>%
-  group_by(ev_package) %>%
-  summarise(nb_runs = n_distinct(file))
-
-# Number of eval calls per package
-ecpp <- pack_calls %>% left_join(runs, by = "ev_package")
-stopifnot(nrow(filter(ecpp, n == 0)) == 0)
-# The package with the most calls
-max_calls <- ecpp %>% filter(n == max(n))
-
-overview_table(
-  r("bina", sum[[1]]),
-  r("binb", sum[[2]]),
-  r("binc", sum[[3]]),
-  r("bind", sum[[4]]),
-  r("bine", sum[[5]]),
-  r("binf", sum[[6]]),
-  r("bing", sum[[7]]),
-  r("binh", sum[[8]]),
-  r("Fewcalls", sum[[1]] + sum[[2]]),
-  r("Manycalls", Kcalls),
-  r("Maxcalls", max_calls$n),
-  r("Maxcallspack", max_calls$ev_package)
-) -> TBL
-```
-
-The distribution of eval calls, each bin represents a power of 10, the number 
-in the bin is the count of packages.
-
-`r TBL`
-
-# Triggered sites
-
-```{r triggered-sites, include=F}
-stopifnot(sum(is.na(E$src_ref)) == 0) # No NAs in scr_ref!
-# Number of eval call sites
-sites <- E %>% # src_ref indicates where the call originates from.
-  pull(src_ref) %>%
-  n_distinct() # This the number of distinct sites
-# Number of call sites per package
-sites_per_package <- E %>%
-  count(ev_package, src_ref) %>%
-  count(ev_package) # All packages have at least one site!
-
-static_sites <- nrow(P)
-
-static_call_sites_per_package <- P %>%
-  count(package, sort = TRUE)
-
-static_at_least_one_call_site <- static_call_sites_per_package %>%
-  filter(n >= 1)
-
-overview_table(
-  r("Sites", sites),
-  r("Staticsites", static_sites),
-  r("Sitespercent", ratio(sites, static_sites)),
-  r("Staticatleastonecallsite", nrow(static_at_least_one_call_site))
-)
-```
-
-`r TBL`
-
-# Numbers of calls to eval
-
-This is the number of eval calls which have a call site in each package,
-divided by the number of runs for the package.
-
-We first compute an approximation of the number of runs per package.
-```{r runs-per-package, include=FALSE}
-runs_package <- E %>%
-  group_by(ev_package) %>%
-  summarise(nb_runs = n_distinct(file)) %>%
-  arrange(desc(nb_runs))
-```
-
-What are the packages with the most runs?
-
-```{r show_run-per-package}
-runs_package %>% datatable()
-```
-
-```{r eval-calls-violin-plot}
-sites_per_package %>%
-  ggplot(aes(x = "", y = n, fill = ""), na.rm = TRUE) +
-  geom_violin(trim = FALSE, width = 1.7) +
-  labs(y = "Eval calls per package") +
-  theme(legend.position = "none")
-
-ggsave(path(PLOT_DIR, "eval_calls_per_packages.pdf"))
-```
-
-# Amount of code
-
-How much code is loaded by eval, we look at argument sizes.
-
-```{r amout-of-code, include=F}
-
-weighted.median <- function(x, w, na.rm = FALSE) {
-  if (na.rm) {
-    na.omit(x)
-  }
-  w <- w[order(x)]
-  x <- x[order(x)]
-  prob <- cumsum(w) / sum(w)
-  ps <- which(abs(prob - .5) == min(abs(prob - .5)))
-  return(x[ps])
-}
-
-med_size <- weighted.median(E$ast_nodes, w = E$duplicates, na.rm = TRUE)
-max_size <- max(E$ast_nodes, na.rm = TRUE)
-average_size <- weighted.mean(E$ast_nodes, na.rm = TRUE, w = E$duplicates)
-
-sizes_prop <- E %>%
-  count(ast_nodes, wt = duplicates) %>%
-  mutate(proportion = n / sum(n)) %>%
-  arrange(ast_nodes) %>%
-  mutate(cumulative = cumsum(proportion))
-
-many <- sizes_prop %>%
-  filter(cumulative >= 0.95) %>%
-  .[1, ]
-
-overview_table(
-  r("Medianszeval", med_size),
-  r("Maxszeval", max_size),
-  r("Avgszeval", average_size),
-  r("Szninefive", many$ast_nodes)
-) -> TBL
-```
-
-`r TBL`
-
-```{r echo=FALSE}
-lbl_colour <- rep("black", 11)
-lbl_colour[[4]] <- "transparent"
-E %>%
-  ggplot(aes(x = ast_nodes, weight = duplicates)) +
-  stat_density(aes(y = ..count..), color = "black", fill = "blue", alpha = 0.3) +
-  scale_x_continuous(breaks = c(0, 1, 2, 3, 4, 5, 10, 30, 100, 300, 1000), trans = "log1p", expand = c(0, 0)) +
-  labs(x = "Size (nodes)") +
-  theme(text = element_text(size=17),
-        axis.text.x = element_text(color = lbl_colour))
-
-E %>% 
-  filter(ast_nodes < 25) %>%
-  ggplot(aes(x = ast_nodes, weight = duplicates)) +
-  geom_histogram(binwidth = 1, color = "black", fill = "blue", alpha = 0.3) +
-  scale_y_continuous(labels = comma) +
-  labs(x = "nodes") +
-  theme(text = element_text(size=17),  panel.grid.major.x = element_blank(),
-        panel.grid.minor.x = element_blank(), aspect.ratio = 4/3)
-
-ggsave(path(PLOT_DIR, "size_loaded_distribution.pdf"))
-```
-
-# Amount of computation
-
-How much work is involved in each eval as measured by the number of opcodes exectued by the interpreter.
-
-```{r amount_computations, include=F}
-tot_events <- E %>%
-  summarise(n = sum(opcodes * duplicates)) %>%
-  .$n
-max_events <- E %>%
-  summarize(n = max(opcodes)) %>%
-  .$n
-
-less_50_events <- E %>%
-  filter(opcodes <= 50) %>%
-  count(wt = duplicates) %>%
-  .$n
-
-overview_table(
-  r("Events", tot_events),
-  r("Maxevents", max_events),
-  r("Smalleventspct", ratio(less_50_events, nb_calls))
-) -> TBL
-```
-
-`r TBL`
-
-```{r echo=FALSE}
-E %>%
-  filter(opcodes > 50) %>%
-  ggplot(aes(x = "", y = opcodes), na.rm = TRUE) +
-  geom_violin(trim = FALSE, width = 1, fill = "blue", alpha = 0.3) +
-  scale_y_log10(labels = scales::comma) +
-  labs(y = "", x = "") +
-  theme(legend.position = "none", text = element_text(size=25))
-
-ggsave(path(PLOT_DIR, "events_per_pack_large.pdf"))
-```
-
-```{r echo=FALSE}
-E %>%
-  filter(opcodes > 0) %>% # == 0 if from eval.parent, because we do not count eval loops for it
-  filter(opcodes <= 50) %>%
-  ggplot(aes(x = "", y = opcodes), na.rm = TRUE) +
-  geom_violin(trim = FALSE, width = 1.7, fill = "blue", alpha = 0.3) +
-  labs(y = "Instructions", x = "") +
-  theme(legend.position = "none", text = element_text(size=25))
-
-ggsave(path(PLOT_DIR, "events_per_pack_small.pdf"))
-```
-
-# Loops
-
-Here, we try to have an estimation of how often a call site is exercised per 
-run. If it is often, maybe it is a loop!
-
-```{r loops, echo=FALSE}
-# Which call sites where exercised and how many times, per run.
-exercised <- E %>%
-  group_by(file) %>% # group by runs
-  count(src_ref, wt = duplicates) # weigh by duplicates
-
-exercised %>% ggplot(aes(n)) +
-  geom_histogram() +
-  scale_x_log10(labels = scales::comma) +
-  labs(x = "Calls per site per run")
-
-# For each site, how many runs and within each runs how often do we trigger.
-triggered <- E %>%
-  group_by(src_ref) %>%
-  summarize(
-    n_runs = n_distinct(file),
-    n = sum(duplicates),
-    average = n / n_runs
-  ) %>%
-  arrange(desc(average))
-
-# Plot the low trigger counts...
-triggered %>%
-  filter(average < 50) %>%
-  ggplot(aes(x = average)) +
-  geom_histogram(binwidth = 1, fill = "blue", alpha = 0.3, colour = "black") +
-  theme(text = element_text(size=16), 
-        axis.text = element_text(size=20), 
-        aspect.ratio = 4/3)
-
-ggsave(path(PLOT_DIR, "calls_per_run_per_call_site.pdf"))
-```
-
-```{r, include = FALSE}
-bins <- cut(triggered$average, breaks = c(0, 50, 100, 250, 500, 1000, 1500, 2000, 3000), include.lowest = TRUE)
-sum <- summary(bins)
-
-overview_table(
-  r("Runbina", sum[[1]]),
-  r("Runbinb", sum[[2]]),
-  r("Runbinc", sum[[3]]),
-  r("Runbind", sum[[4]]),
-  r("Runbine", sum[[5]]),
-  r("Runbinf", sum[[6]]),
-  r("Runbing", sum[[7]]),
-  r("Runbinh", sum[[8]])
-) -> TBL
-```
-
-`r TBL`
-
-# Diversity of arguments
-
-How diverse are the expressions passed to `eval`? We consider the diversity in types in the resolved expression.
-
-```{r type-diversity, echo=FALSE}
-E %>%
-  ggplot(aes(x = fct_rev(fct_infreq(type)), weight=duplicates)) +
-  geom_bar() +
-  coord_flip() +
-  labs(x = "type")
-
-types <- E %>% count(type, wt=duplicates)
-
-symbs <- types %>% filter(type == "symbol") %>% .$n
-languages <- types %>% filter(type == "language") %>% .$n
-expressions <- types %>% filter(type == "expression") %>% .$n
-
-ggplot2_symbs <- E %>% filter(ev_package == "ggplot2") %>% 
-  count(type, wt=duplicates) %>% filter(type == "symbol") %>% .$n
-
-overview_table(
-  r("Symbolpercent", ratio(symbs, nb_calls)),
-  r("Languagepercent", ratio(languages, nb_calls)),
-  r("Expressionpercent", ratio(expressions, nb_calls)),
-  r("Codepercent", ratio(symbs + languages + expressions, nb_calls)),
-  r("Ggplotsymbolpercent", ratio(ggplot2_symbs, symbs))
-) -> TBL
-```
-
-`r TBL`
-
-```{r include=FALSE}
-duration <- difftime(Sys.time(), now)
-```
-
-Notebook execution was `r duration` `r units(duration)`.

### 2021-01-05 analysis/expressions.Rmd
Most widespread expressions in terms of call sites/packages

@@
 
-E_raw -> E
+read_fst(calls_path) %>% as_tibble() -> E_raw
+
+ E_raw %>% select(
+    run_package = package, #      The package which was run to trigger the eval
+    ev_variant = eval_function, # Which eval variant was called
+    duplicates = nb_ev_calls, #   Number of identical eval calls (weight)
+    src_ref = eval_call_srcref, # Source ref for the eval call site
+    file, #                       Script file name (name of this run)
+    ev_package = eval_source, #   Package of the call site
+    type = expr_resolved_type_tag,   #   Argument type
+    ev_expr = expr_expression,
+    expr_resolved,
+    parsed = expr_parsed_expression
+  ) -> E
 
@@
 corpus_size <- nrow(C)
+```
 
+Some useful numbers:
+```{r}
+nb_call_sites <- E %>% pull(src_ref) %>% n_distinct()
 ```
+
+# Most frequent expressions
+
+Before resolving, so as the programmer wrote them:
+
+```{r}
+E %>% count(ev_expr, wt = duplicates, sort = TRUE) 
+```
+
+After resolving:
+
+```{r}
+E %>% count(expr_resolved, wt = duplicates, sort = TRUE) 
+```
+
+# Expressions common to several call sites and packages
+
+It emphasizes which expressions are a common pattern across various code bases.
+
+```{r}
+widespread_expr <- E %>% group_by(ev_expr) %>% summarize(n_call_sites = n_distinct(src_ref), n_packages = n_distinct(ev_package))
+```
+
+```{r}
+widespread_expr %>% arrange(desc(n_call_sites)) %>% datatable()
+```
+
+```{r}
+widespread_expr %>% arrange(desc(n_packages)) %>% datatable()
+```
+
+For instance, for the most widespread expression, both in terms of call sites and packages, `mf`, we show here the packages sorted by decreasing number of call sites with this expression name.
+
+```{r}
+E %>% filter(ev_expr == "mf") %>% group_by(ev_package) %>% summarise(n = n_distinct(src_ref)) %>% arrange(desc(n))
+```
+
+Manual inspection seems to show that `mf` is often used as a result of a `match.call`:
+
+- https://github.com/cran/PMCMRplus/blob/4f5234542aa0a60642b230bea7efb668fb16a140/R/mandelkTest.R#L121 
+- https://github.com/fbertran/plsRglm/blob/dfa4e54ea02bca8bf04d29bb65dc7dba611927c9/R/plsRglmmodel.default.R#L4
+
+This is the common pattern when a call is retrieved and then modified, then evaluated.
+
+# Expressions using `parse`, `str2lang`, and `str2expression`
+
+```{r}
+parsed_expr <- E %>% filter(!is.na(parsed))
+
+nb_parsed_call_sites <- parsed_expr %>% pull(src_ref) %>% n_distinct()
+```
+
+There are `r nb_parsed_call_sites` call sites using those functions in their argument, out of the `r nb_call_sites` in total, i.e. `r ratio(nb_parsed_call_sites, nb_call_sites)`%.

### 2021-01-05 analysis/expressions.Rmd
Example

@@
 - https://github.com/fbertran/plsRglm/blob/dfa4e54ea02bca8bf04d29bb65dc7dba611927c9/R/plsRglmmodel.default.R#L4
+- https://github.com/cran/RWeka/blob/55030d5513bf37760d1f6e2f1a49495977113276/R/classifiers.R#L42 with a comment _## The "usual" way of creating a model frame from the call._ (sic!)
 

### 2021-01-06 analysis/expressions.Rmd
Call site with one resolved expressions, to a value type.

@@
 
+# Resolved expressions
+
+Are expressions resolved to very diverse values or expressions?
+If they are uniquely resolved and it is a value, then we can get rid of the `eval`.
+
+```{r}
+
+resolved_variety <- E %>% group_by(ev_expr, src_ref) %>% summarise(n = n_distinct(expr_resolved)) %>% arrange(desc(n)) 
+resolved_variety %>% datatable()
+
+nb_one_resolved <- resolved_variety %>% filter(n == 1) %>% nrow()
+
+```
+
+There are `r nb_one_resolved` call sites that are resolved to only one expression, out of `r nrow(resolved_variety)`.
+
+We can now count how many of these call sites with only one resolved expression resolve to a value type.
+
+```{r}
+one_resolved_type <- E %>% group_by(src_ref) %>% mutate(n_resolved = n_distinct(expr_resolved)) %>% filter(n_resolved == 1) %>% ungroup() %>% select(src_ref, n_resolved, type) %>% distinct()
+
+nb_one_resolved_type <- one_resolved_type %>% nrow()
+```
+
+`r nb_one_resolved_type` call sites resolved to one expression, of only one type.
+
+```{r}
+nb_value_type <- one_resolved_type %>% filter(type %in% c("character", "double", "integer", "logical", "NULL"))
+```
+
+There are `r nb_value_type` such call sites.
+
 # Expressions common to several call sites and packages
@@
 ```{r}
-widespread_expr <- E %>% group_by(ev_expr) %>% summarize(n_call_sites = n_distinct(src_ref), n_packages = n_distinct(ev_package))
+widespread_expr <- E %>% group_by(expr_resolved) %>% summarize(n_call_sites = n_distinct(src_ref), n_packages = n_distinct(ev_package))
 ```

### 2021-01-12 analysis/expressions.Rmd
Constant arithmetic expressions.

@@
   expr_resolved,
+  func = expr_resolved_function, # the function (if any) at the root of the AST of expr_resolved
   parsed = expr_parsed_expression
@@
 ```{r}
-nb_call_sites <- E %>% pull(src_ref) %>% n_distinct()
+nb_call_sites <- E %>%
+  pull(src_ref) %>%
+  n_distinct()
 ```
@@
 
-resolved_variety <- E %>% group_by(ev_expr, src_ref) %>% summarise(n = n_distinct(expr_resolved)) %>% arrange(desc(n)) 
+resolved_variety <- E %>%
+  group_by(ev_expr, src_ref) %>%
+  summarise(n = n_distinct(expr_resolved)) %>%
+  arrange(desc(n))
 resolved_variety %>% datatable()
 
-nb_one_resolved <- resolved_variety %>% filter(n == 1) %>% nrow()
-
+nb_one_resolved <- resolved_variety %>%
+  filter(n == 1) %>%
+  nrow()
 ```
@@
 ```{r}
-one_resolved_type <- E %>% group_by(src_ref) %>% mutate(n_resolved = n_distinct(expr_resolved)) %>% filter(n_resolved == 1) %>% ungroup() %>% select(src_ref, n_resolved, type) %>% distinct()
+one_resolved_type <- E %>%
+  group_by(src_ref) %>%
+  mutate(n_resolved = n_distinct(expr_resolved)) %>%
+  filter(n_resolved == 1) %>%
+  ungroup() %>%
+  select(src_ref, n_resolved, type) %>%
+  distinct()
 
@@
 ```{r}
-widespread_expr <- E %>% group_by(expr_resolved) %>% summarize(n_call_sites = n_distinct(src_ref), n_packages = n_distinct(ev_package))
+widespread_expr <- E %>%
+  group_by(expr_resolved) %>%
+  summarize(n_call_sites = n_distinct(src_ref), n_packages = n_distinct(ev_package))
 ```
@@
 ```{r}
-widespread_expr %>% arrange(desc(n_call_sites)) %>% datatable()
+widespread_expr %>%
+  arrange(desc(n_call_sites)) %>%
+  datatable()
 ```
@@
 ```{r}
-widespread_expr %>% arrange(desc(n_packages)) %>% datatable()
+widespread_expr %>%
+  arrange(desc(n_packages)) %>%
+  datatable()
 ```
@@
 ```{r}
-E %>% filter(ev_expr == "mf") %>% group_by(ev_package) %>% summarise(n = n_distinct(src_ref)) %>% arrange(desc(n))
+E %>%
+  filter(ev_expr == "mf") %>%
+  group_by(ev_package) %>%
+  summarise(n = n_distinct(src_ref)) %>%
+  arrange(desc(n))
 ```
@@
 
+We can also look at the functions called in the expression:
+
+```{r}
+widespread_function <- E %>%
+  filter(!is.na(func)) %>%
+  group_by(func) %>%
+  summarize(n_call_sites = n_distinct(src_ref), n_packages = n_distinct(ev_package))
+```
+
 # Expressions using `parse`, `str2lang`, and `str2expression`
@@
 
-nb_parsed_call_sites <- parsed_expr %>% pull(src_ref) %>% n_distinct()
+nb_parsed_call_sites <- parsed_expr %>%
+  pull(src_ref) %>%
+  n_distinct()
 ```
@@
 There are `r nb_parsed_call_sites` call sites using those functions in their argument, out of the `r nb_call_sites` in total, i.e. `r ratio(nb_parsed_call_sites, nb_call_sites)`%.
+
+# Arithmetic expressions
+
+First, we only keep rows where the root is an arithmetic function, and the rows with a constant number
+
+```{r}
+arith_func <- c("/", "-", "*", "+", "^", "log", "sqrt", "exp", "max", "min", "cos", "sin", "abs", "atan", ":")
+
+arithm <- E %>% filter(func %in% arith_func)
+```
+
+We want to keep only the ones that only use these functions and constants.
+
+```{r}
+is_arithm_expr <- function(exp) {
+  if (is.call(exp)) {
+    function_name <- exp[[1]]
+    function_args <- exp[-1]
+    return(as.character(function_name) %in% arith_func && every(function_args, is_arithm_expr))
+  }
+  else if (is.expression(exp)) {
+    return(every(exp, is_arithm_expr))
+  }
+  else if (typeof(exp) %in% c("integer", "double", "complex")) {
+    return(TRUE)
+  }
+  else {
+    return(FALSE)
+  }
+}
+
+arithm_folded <- arithm %>% filter(map_lgl(expr_resolved, function(exp) {
+  ast <- NA
+  try(ast <- parse(text = exp)[[1]], silent = TRUE)
+  # some expr_resolved have been truncated so we mark them as FALSE (even though they could be true)
+  if(!is.na(ast)) {
+     return(is_arithm_expr(ast))
+  }
+  else {
+    return(FALSE)
+  }
+
+}))
+```
+
+It corresponds to these call sites and packages:
+
+```{r}
+arithm_folded %>% group_by(expr_resolved) %>% summarise(n_call_sites = n_distinct(src_ref), n_packages = n_distinct(ev_package)) %>% datatable()
+```
+
+# Formulas
+
+A formula is defined with the tilde `~` operator.
+
+# Variables

### 2021-01-12 analysis/expressions.Rmd
Expressions with only variable names.

@@
 ```{r}
+n_arithm_call_sites <- arithm_folded %>% pull(src_ref) %>% n_distinct()
+
 arithm_folded %>% group_by(expr_resolved) %>% summarise(n_call_sites = n_distinct(src_ref), n_packages = n_distinct(ev_package)) %>% datatable()
 ```
+
+The total number of call sites with those arithmetic on constants is `r n_arithm_call_sites`.
 
@@
 # Variables
+
+Here, we analyze expressions with only a variable name.
+
+```{r}
+with_var <- E %>% filter(str_detect(expr_resolved, "^(?:\\.[:alpha]|[:alpha:])[[:alpha:][:digit:]_\\.]*$"))
+```
+
+We can categorize them by number of call sites and number of packages:
+```{r}
+n_with_var <- with_var %>% pull(src_ref) %>% n_distinct()
+
+with_var %>% group_by(expr_resolved) %>% summarise(n_call_sites = n_distinct(src_ref), n_packages = n_distinct(ev_package)) %>% datatable()
+```
+
+There are `r n_with_var` such call sites.

### 2021-01-12 analysis/expressions.Rmd
Get rid of NA in the variable names

@@
   # some expr_resolved have been truncated so we mark them as FALSE (even though they could be true)
-  if(!is.na(ast)) {
+  if (!is.na(ast)) {
     return(is_arithm_expr(ast))
@@
   }
-
 }))
@@
 ```{r}
-n_arithm_call_sites <- arithm_folded %>% pull(src_ref) %>% n_distinct()
+n_arithm_call_sites <- arithm_folded %>%
+  pull(src_ref) %>%
+  n_distinct()
 
-arithm_folded %>% group_by(expr_resolved) %>% summarise(n_call_sites = n_distinct(src_ref), n_packages = n_distinct(ev_package)) %>% datatable()
+arithm_folded %>%
+  group_by(expr_resolved) %>%
+  summarise(n_call_sites = n_distinct(src_ref), n_packages = n_distinct(ev_package)) %>%
+  datatable()
 ```
@@
 ```{r}
-with_var <- E %>% filter(str_detect(expr_resolved, "^(?:\\.[:alpha]|[:alpha:])[[:alpha:][:digit:]_\\.]*$"))
+with_var <- E %>%
+  filter(str_detect(expr_resolved, "^(?:\\.[:alpha]|[:alpha:])[[:alpha:][:digit:]_\\.]*$")) %>%
+  filter(expr_resolved == "NA") # "NA" matches the regex of a variable name but is a value so we get rid of it
 ```
@@
 ```{r}
-n_with_var <- with_var %>% pull(src_ref) %>% n_distinct()
+n_with_var <- with_var %>%
+  pull(src_ref) %>%
+  n_distinct()
 
-with_var %>% group_by(expr_resolved) %>% summarise(n_call_sites = n_distinct(src_ref), n_packages = n_distinct(ev_package)) %>% datatable()
+with_var %>%
+  group_by(expr_resolved) %>%
+  summarise(n_call_sites = n_distinct(src_ref), n_packages = n_distinct(ev_package)) %>%
+  datatable()
 ```

### 2021-01-18 analysis/expressions.Rmd
Fix number of call sites evaluating only variables

@@
 
-There are `r nb_value_type` such call sites.
+There are `r nb_value_type %>% nrow()` such call sites.
 
@@
   filter(str_detect(expr_resolved, "^(?:\\.[:alpha]|[:alpha:])[[:alpha:][:digit:]_\\.]*$")) %>%
-  filter(expr_resolved == "NA") # "NA" matches the regex of a variable name but is a value so we get rid of it
+  filter(expr_resolved != "NA") # "NA" matches the regex of a variable name but is a value so we get rid of it
 ```

### 2021-01-19 analysis/expressions.Rmd
Canonical expression

@@
   # some expr_resolved have been truncated so we mark them as FALSE (even though they could be true)
-  if (!is.na(ast)) {
+  if (length(ast) > 1 || !is.na(ast)) {
     return(is_arithm_expr(ast))
@@
 
+```{r}
+with_formula <- E %>%
+  filter(str_detect(expr_resolved, fixed("~")))
+```
+
+```{r}
+n_with_formula <- with_formula %>%
+  pull(src_ref) %>%
+  n_distinct()
+```
+
+The total number of call sites with formulas is `r n_with_formula`.
+
+```{r}
+with_formula %>%
+  group_by(expr_resolved) %>%
+  summarise(n_call_sites = n_distinct(src_ref), n_packages = n_distinct(ev_package)) %>%
+  datatable()
+```
+
+`~` is mostly used within a call to `model.frame`.
+
 # Variables
@@
 There are `r n_with_var` such call sites.
+
+# Better normalization of resolved expressions
+
+We fold all constant arithmetic expressions and replace all variable names by a placeholder.
+
+```{r}
+canonic_expr <- function(exp) {
+  if (is.call(exp)) {
+    function_name <- exp[[1]]
+    function_args <- exp[-1]
+    res <- map(function_args, canonic_expr)
+    # Function_name can be larger than one is a function with a namespace stats::model or another higher-order function
+    # For instance, (function(x, y) x + y)(3, 4)
+    if (length(function_name) == 1 && as.character(function_name) %in% arith_func && every(res, function(v) {
+      v == "NUM"
+    })) {
+      return("NUM")
+    }
+    else if (length(function_name) == 1 && function_name == "::") {
+      str_c(as.character(function_args[[1]]), "::", as.character(function_args[[2]]))
+    }
+    else {
+      function_name <- if (length(function_name) == 1) {
+        as.character(function_name)
+      }
+      else {
+        canonic_expr(function_name)
+      }
+      return(str_c(str_c(function_name, collapse = ", "), "(", str_c(res, collapse = ", "), ")"))
+    }
+  }
+  else if (is.symbol(exp)) {
+    return("VAR")
+  }
+  else if (is.expression(exp)) {
+    return(str_c(map(exp, canonic_expr), collapse = ", "))
+  }
+  else if (typeof(exp) %in% c("integer", "double", "complex")) {
+    return("NUM")
+  }
+  else {
+    return(deparse1(exp))
+  }
+}
+
+canonic_expr_str <- function(exp) {
+  ast <- NA
+  try(ast <- parse(text = exp)[[1]], silent = TRUE)
+  # some expr_resolved have been truncated so we mark them as FALSE (even though they could be true)
+  if (is.symbol(ast) || is.language(ast) || length(ast) > 1 || !is.na(ast)) {
+    return(canonic_expr(ast))
+  }
+  else {
+    return("CANONIZATION ERROR")
+  }
+}
+```
+
+For performance reason, we only apply this function to the unique resolved expressions.
+
+```{r}
+res_exprs <- E %>%
+  select(expr_resolved) %>%
+  unique()
+
+res_exprs <- res_exprs %>% mutate(expr_canonic = map_chr(expr_resolved, canonic_expr_str))
+```

### 2021-01-19 analysis/expressions.Rmd
Rnking of normalized expressions
Formatting

@@
   }
+  else if(typeof(exp) == "logical") {
+    return("BOOL")
+  }
+  else if(typeof(exp) == "character") {
+    return("STR")
+  }
   else {
@@
 res_exprs <- res_exprs %>% mutate(expr_canonic = map_chr(expr_resolved, canonic_expr_str))
+
+E_cano <- E %>% left_join(res_exprs)
 ```
+
+```{r}
+widespread_canonic <- E_cano %>% group_by(expr_canonic) %>% 
+  summarize(n_call_sites = n_distinct(src_ref), n_packages = n_distinct(ev_package))
+
+widespread_canonic %>% arrange(desc(n_call_sites)) %>% datatable()
+```

### 2021-01-20 analysis/expressions.Rmd
Constant folding on strings

@@
 ```{r}
+
+str_op <- c("paste", "paste0", "str_c")
+
 canonic_expr <- function(exp) {
@@
       return("NUM")
+    }
+    else if (length(function_name) == 1 && as.character(function_name) %in% str_op && every(res, function(v) {
+      v == "STR"
+    })) {
+      return("STR")
     }

### 2021-01-21 analysis/expressions.Rmd
Normalize vectors with the same type.

--- 2021-01-19 analysis/expressions.Rmd
@@
 ```{r}
+str_op <- c("paste", "paste0", "str_c")
+comp_op <- c("<", ">", "<=", ">=", "==", "!=")
+bool_op <- c("&", "&&", "|", "||", "!")
+
 canonic_expr <- function(exp) {
@@
     function_args <- exp[-1]
-    res <- map(function_args, canonic_expr)
+    res <- map(function_args, canonic_expr) # TODO: rather directly compute canonic_expr on function_args?
     # Function_name can be larger than one is a function with a namespace stats::model or another higher-order function
     # For instance, (function(x, y) x + y)(3, 4)
-    if (length(function_name) == 1 && as.character(function_name) %in% arith_func && every(res, function(v) {
+    if (length(function_name) == 1 && as.character(function_name) == "(") {
+      return(res[[1]]) # (a) => (a)()
+    }
+    else if (length(function_name) == 1 && as.character(function_name) %in% arith_func && every(res, function(v) {
       v == "NUM"
@@
       return("NUM")
+    }
+    else if (length(function_name) == 1 && as.character(function_name) %in% str_op && every(res, function(v) {
+      v == "STR"
+    })) {
+      return("STR")
+    }
+    else if (length(function_name) == 1 && as.character(function_name) %in% bool_op && every(res, function(v) {
+      v == "BOOL"
+    })) {
+      return("BOOL")
+    }
+    else if(length(function_name) == 1 && as.character(function_name) %in% comp_op && every(res, function(v) {v == "NUM"})) {
+      return("BOOL")
+    }
+    else if(length(function_name) == 1 && as.character(function_name) == "c" && length(res) > 1 && n_distinct(res) == 1 && res[[1]] %in% c("BOOL", "NUM", "STR")) {
+      str_c("c(", res[[1]], ")")
     }
@@
   }
-  else if(typeof(exp) == "logical") {
+  else if (typeof(exp) == "logical") {
     return("BOOL")
   }
-  else if(typeof(exp) == "character") {
+  else if (typeof(exp) == "character") {
     return("STR")
@@
 ```{r}
-widespread_canonic <- E_cano %>% group_by(expr_canonic) %>% 
+widespread_canonic <- E_cano %>%
+  group_by(expr_canonic) %>%
   summarize(n_call_sites = n_distinct(src_ref), n_packages = n_distinct(ev_package))
 
-widespread_canonic %>% arrange(desc(n_call_sites)) %>% datatable()
+widespread_canonic %>%
+  arrange(desc(n_call_sites)) %>%
+  datatable()
 ```

### 2021-01-26 analysis/expressions.Rmd
Collapse list as c, taking account VAR also

--- 2021-01-19 analysis/expressions.Rmd
@@
     }
-    else if (length(function_name) == 1 && function_name == "::") {
-      str_c(as.character(function_args[[1]]), "::", as.character(function_args[[2]]))
+    else if (length(function_name) == 1 && as.character(function_name) %in% str_op && every(res, function(v) {
+      v == "STR"
+    })) {
+      return("STR")
+    }
+    else if (length(function_name) == 1 && as.character(function_name) %in% bool_op && every(res, function(v) {
+      v == "BOOL"
+    })) {
+      return("BOOL")
+    }
+    else if(length(function_name) == 1 && as.character(function_name) %in% comp_op && every(res, function(v) {v == "NUM"})) {
+      return("BOOL")
+    }
+    else if(length(function_name) == 1 && as.character(function_name) %in% c("c", "list") && length(res) > 1 && n_distinct(res) == 1 ) { # && res[[1]] %in% c("BOOL", "NUM", "STR")
+      str_c(as.character(function_name), "(", res[[1]], ")")
+    }
+    else if (length(function_name) == 1 && function_name == "::") { # we drop the namespace
+      as.character(function_args[[2]])
     }
@@
   else {
-    return("CANONIZATION ERROR")
+    return("NORMALIZATION ERROR")
   }

### 2021-01-28 analysis/expressions.Rmd
Update to normalization

--- 2021-01-20 analysis/expressions.Rmd
@@
 ```{r}
+str_op <- c("paste", "paste0", "str_c")
+comp_op <- c("<", ">", "<=", ">=", "==", "!=")
+bool_op <- c("&", "&&", "|", "||", "!")
 
-str_op <- c("paste", "paste0", "str_c")
-
-canonic_expr <- function(exp) {
+canonic_expr <- function(exp, with.names = FALSE) {
     if (is.call(exp)) {
@@
         function_args <- exp[-1]
-    res <- map(function_args, canonic_expr)
+        res <- map(function_args, canonic_expr, with.names) # TODO: rather directly compute canonic_expr on function_args?
         # Function_name can be larger than one is a function with a namespace stats::model or another higher-order function
         # For instance, (function(x, y) x + y)(3, 4)
-    if (length(function_name) == 1 && as.character(function_name) %in% arith_func && every(res, function(v) {
+        if (length(function_name) == 1 && as.character(function_name) == "(") {
+            return(res[[1]]) # (a) => (a)()
+        }
+        else if (length(function_name) == 1 && as.character(function_name) %in% arith_func && every(res, function(v) {
             v == "NUM"
@@
         }
-    else if (length(function_name) == 1 && function_name == "::") {
-      str_c(as.character(function_args[[1]]), "::", as.character(function_args[[2]]))
+        else if (length(function_name) == 1 && as.character(function_name) %in% bool_op && every(res, function(v) {
+            v == "BOOL"
+        })) {
+            return("BOOL")
+        }
+        else if (length(function_name) == 1 && as.character(function_name) %in% comp_op && every(res, function(v) {
+            v == "NUM"
+        })) {
+            return("BOOL")
+        }
+        else if (length(function_name) == 1 && as.character(function_name) %in% c("c", "list") && length(res) > 1 && n_distinct(res) == 1 && res[[1]] %in% c("BOOL", "NUM", "STR", "VAR")) {
+            str_c(as.character(function_name), "(", res[[1]], ")")
+        }
+        else if (length(function_name) == 1 && function_name == "::") { # we drop the namespace
+            as.character(function_args[[2]])
         }
@@
             function_name <- if (length(function_name) == 1) {
-        as.character(function_name)
+                func_name <- as.character(function_name)
+                function_name <- if (func_name %in% arith_func) {
+                    "OP"
+                }
+                else if (func_name %in% comp_op) {
+                    "COMP"
+                }
+                else if (func_name %in% bool_op) {
+                    "LOGI"
                 }
                 else {
-        canonic_expr(function_name)
+                    func_name
                 }
-      return(str_c(str_c(function_name, collapse = ", "), "(", str_c(res, collapse = ", "), ")"))
+            }
+            else {
+                canonic_expr(function_name, with.names)
+            }
+
+            # It might have been transformed by canonic_expr in the recursive call above
+            # it should always have length one
+            res <- if (length(function_name) == 1) {
+                func_name <- as.character(function_name)
+                # VAR is absorbing: if there is at least one, we propagate to all of them
+                res <- if (func_name %in% c("COMP", "LOGI", "OP") && "VAR" %in% res) {
+                    rep.int("VAR", length(res))
+                }
+                else if (func_name == "model.frame") { # special case
+                    # We keep the 1st two arguments (usually formula and data)
+                    # and subset if it is there
+                    subset_arg <- if ("subset" %in% names(res)) {
+                        subset_s <- as.character(res["subset"])
+                        res["subset"] <- NULL
+                        str_c("subset = ", subset_s)
+                    }
+                    else {
+                        "subset = NULL"
+                    }
+                    first_args <- if(length(res) > 1) {
+                          res[1:2]
+                    }
+                    else {
+                      c(res[[1]], "NULL")
+                    }
+                  c(first_args, subset_arg)
+                }
+                else {
+                    res
+                }
+            }
+
+            fun_args <- if(with.names) {
+              str_c(names(res), if_else(names(res) == "", "",  " = "), res, collapse = ", ")
+            } 
+            else
+            {
+              str_c(res, collapse = ", ")
+            }
+
+            return(str_c(str_c(function_name, collapse = ", "), "(", fun_args, ")"))
         }
@@
     else if (is.expression(exp)) {
-    return(str_c(map(exp, canonic_expr), collapse = ", "))
+        return(str_c(map(exp, canonic_expr, with.names), collapse = ", "))
     }
@@
     }
-  else if(typeof(exp) == "logical") {
+    else if (typeof(exp) == "logical") {
         return("BOOL")
     }
-  else if(typeof(exp) == "character") {
+    else if (typeof(exp) == "character") {
         return("STR")
@@
 
-canonic_expr_str <- function(exp) {
+canonic_expr_str <- function(exp, with.names = FALSE) {
     ast <- NA
@@
     if (is.symbol(ast) || is.language(ast) || length(ast) > 1 || !is.na(ast)) {
-    return(canonic_expr(ast))
+        return(canonic_expr(ast, with.names))
     }
     else {
-    return("CANONIZATION ERROR")
+        return("NORMALIZATION ERROR")
     }
@@
 ```{r}
-res_exprs <- E %>%
+res_exprs_uniq <- E %>%
   select(expr_resolved) %>%
@@
 
+res_exprs <- res_exprs %>% mutate(expr_canonic_names = map_chr(expr_resolved, canonic_expr_str), with.names = TRUE)
+
 E_cano <- E %>% left_join(res_exprs)

### 2021-01-28 analysis/expressions.Rmd
fix variable names

@@
 }
+
+canonic_expr_str2 <- function(exp, with.names = FALSE) {
+    ast <- NA
+    try(ast <- parse(text = exp)[[1]], silent = TRUE)
+  return("")
+}
 ```
@@
 
-res_exprs <- res_exprs %>% mutate(expr_canonic = map_chr(expr_resolved, canonic_expr_str))
+res_exprs <- res_exprs_uniq %>% mutate(expr_canonic = map_chr(expr_resolved, canonic_expr_str))
 
-res_exprs <- res_exprs %>% mutate(expr_canonic_names = map_chr(expr_resolved, canonic_expr_str), with.names = TRUE)
+# res_exprs <- res_exprs %>% mutate(expr_canonic_names = map_chr(expr_resolved, canonic_expr_str), with.names = TRUE)
 

### 2021-02-09 analysis/expressions.Rmd
match.call pattern and normalized expressions

--- 2021-01-26 analysis/expressions.Rmd
@@
   calls_path: summarized.fst
+  normalized_path: normalized-expressions.fst
   corpus_path: ../corpus.fst
@@
 static_path <- paste0(params$base_dir, params$static_path)
+normalized_path <- paste0(params$base_dir, params$normalized_path)
 ```
@@
 which package belong to our corpus (`C_raw`). For conciseness, `E` is
-the input data set with only the columns that we use.
+the input data set with only the columns that we use. We also read the file with normalized expressions, `N`.
 
@@
 read_fst(calls_path) %>% as_tibble() -> E_raw
+
+read_fst(normalized_path) %>% as_tibble() -> N
 
@@
   func = expr_resolved_function, # the function (if any) at the root of the AST of expr_resolved
-  parsed = expr_parsed_expression
+  parsed = expr_parsed_expression,
+  hash = expr_resolved_hash,
+  match.call = expr_match_call
 ) -> E
@@
 
-# Better normalization of resolved expressions
+# Match.call pattern
 
-We fold all constant arithmetic expressions and replace all variable names by a placeholder.
+We have traced how many expressions come from a `match.call`
 
 ```{r}
-str_op <- c("paste", "paste0", "str_c")
-comp_op <- c("<", ">", "<=", ">=", "==", "!=")
-bool_op <- c("&", "&&", "|", "||", "!")
-
-canonic_expr <- function(exp) {
-  if (is.call(exp)) {
-    function_name <- exp[[1]]
-    function_args <- exp[-1]
-    res <- map(function_args, canonic_expr) # TODO: rather directly compute canonic_expr on function_args?
-    # Function_name can be larger than one is a function with a namespace stats::model or another higher-order function
-    # For instance, (function(x, y) x + y)(3, 4)
-    if (length(function_name) == 1 && as.character(function_name) == "(") {
-      return(res[[1]]) # (a) => (a)()
-    }
-    else if (length(function_name) == 1 && as.character(function_name) %in% arith_func && every(res, function(v) {
-      v == "NUM"
-    })) {
-      return("NUM")
-    }
-    else if (length(function_name) == 1 && as.character(function_name) %in% str_op && every(res, function(v) {
-      v == "STR"
-    })) {
-      return("STR")
-    }
-    else if (length(function_name) == 1 && as.character(function_name) %in% bool_op && every(res, function(v) {
-      v == "BOOL"
-    })) {
-      return("BOOL")
-    }
-    else if(length(function_name) == 1 && as.character(function_name) %in% comp_op && every(res, function(v) {v == "NUM"})) {
-      return("BOOL")
-    }
-    else if(length(function_name) == 1 && as.character(function_name) %in% c("c", "list") && length(res) > 1 && n_distinct(res) == 1 ) { # && res[[1]] %in% c("BOOL", "NUM", "STR")
-      str_c(as.character(function_name), "(", res[[1]], ")")
-    }
-    else if (length(function_name) == 1 && function_name == "::") { # we drop the namespace
-      as.character(function_args[[2]])
-    }
-    else {
-      function_name <- if (length(function_name) == 1) {
-        as.character(function_name)
-      }
-      else {
-        canonic_expr(function_name)
-      }
-      return(str_c(str_c(function_name, collapse = ", "), "(", str_c(res, collapse = ", "), ")"))
-    }
-  }
-  else if (is.symbol(exp)) {
-    return("VAR")
-  }
-  else if (is.expression(exp)) {
-    return(str_c(map(exp, canonic_expr), collapse = ", "))
-  }
-  else if (typeof(exp) %in% c("integer", "double", "complex")) {
-    return("NUM")
-  }
-  else if (typeof(exp) == "logical") {
-    return("BOOL")
-  }
-  else if (typeof(exp) == "character") {
-    return("STR")
-  }
-  else {
-    return(deparse1(exp))
-  }
-}
-
-canonic_expr_str <- function(exp) {
-  ast <- NA
-  try(ast <- parse(text = exp)[[1]], silent = TRUE)
-  # some expr_resolved have been truncated so we mark them as FALSE (even though they could be true)
-  if (is.symbol(ast) || is.language(ast) || length(ast) > 1 || !is.na(ast)) {
-    return(canonic_expr(ast))
-  }
-  else {
-    return("NORMALIZATION ERROR")
-  }
-}
+E_match.call <- E %>% filter(!is.na(match.call))
 ```
 
-For performance reason, we only apply this function to the unique resolved expressions.
+How many call sites, and different packages, is that?
 
 ```{r}
-res_exprs <- E %>%
-  select(expr_resolved) %>%
-  unique()
+nb_call_sites_match.call <- E_match.call %>% pull(src_ref) %>% n_distinct()
 
-res_exprs <- res_exprs %>% mutate(expr_canonic = map_chr(expr_resolved, canonic_expr_str))
+nb_packages_match.call <- E_match.call %>% pull(ev_package) %>% n_distinct()
 
-E_cano <- E %>% left_join(res_exprs)
+```
+
+What function is evaled in this pattern
+```{r}
+E_match.call %>% group_by(match.call) %>% summarize(n = n_distinct(src_ref)) %>% arrange(desc())
+```
+
+As expected, `model.frame` is the most used one.
+
+# Normalization of resolved expressions
+
+We join the normalized expressions and their statistics with the calls.
+
+```{r}
+E_cano <- E %>% left_join(N, by = c("hash" = "expr_resolved_hash")) # Common column is the hash of the full resolved expression
 ```

### 2021-02-09 analysis/inc/paths.R
Add code.fst

@@ -19,6 +19,7 @@ CORPUS_FILE <- path(DATA_DIR, "corpus.fst")
 PACKAGE_SUM_CALLS_FILE <- path(PACKAGE_DATA_DIR, "summarized.fst")
 PACKAGE_UNDEFINED_FILE <- path(PACKAGE_DATA_DIR, "undefined.fst")
 PACKAGE_EVALS_STATIC_FILE <-path(PACKAGE_DATA_DIR, "evals-static.csv")
+PACKAGE_CODE_FILE <- path(PACKAGE_DATA_DIR, "code.fst")
 
 KAGGLE_KERNEL_FILE <- path(KAGGLE_DATA_DIR, "kernel.csv")
 KAGGLE_EVALS_STATIC_FILE <- path(KAGGLE_DATA_DIR, "evals-static.csv")

### 2021-02-09 analysis/expressions.Rmd
Use the correct function name for the match.call pattern

@@
 ```{r}
-E_match.call %>% group_by(match.call) %>% summarize(n = n_distinct(src_ref)) %>% arrange(desc())
+E_match.call %>% group_by(func) %>% summarize(n_call_sites = n_distinct(src_ref)) %>% arrange(desc(n_call_sites))
 ```

### 2021-02-12 analysis/expressions.Rmd
Correctly counting number of resolved_expr (use hash because of truncation)

@@
 ```
+
+To check whether the normalization makes sense, we have a look at how many normalized expressions we have per call sites.
+```{r}
+by_src_ref <- E_cano %>% group_by(src_ref) %>%
+  summarise(n_resolved = n_distinct(hash), n_canonic = n_distinct(expr_canonic))
+```
+
+There should always be less normalized epxressions than resolved ones. For how many do we have strictly less normalized expressions?
+
+```{r}
+by_src_ref %>% filter(n_resolved > n_canonic) %>% arrange(desc(n_canonic))
+```
+
+```{r}
+by_src_ref %>% filter(n_resolved == 1 | n_resolved > n_canonic)  %>% arrange(desc(n_canonic)) %>% print(n = 100)
+```
+
+```{r}
+exprs_by_srcref <- E_cano %>% group_by(src_ref) %>% summarize(n_canonic = n_distinct(expr_canonic), exprs = paste0(unique(expr_canonic), collapse="    ;     "))
+```

### 2021-02-23 analysis/expressions.Rmd
Port normalized/minimized expression from expressions.Rmd to normalized.Rmd

--- 2021-02-09 analysis/expressions.Rmd
@@
   calls_path: summarized.fst
-  normalized_path: normalized-expressions.fst
+  normalized_path: normalized-expressions.csv
   corpus_path: ../corpus.fst
@@
 read_fst(calls_path) %>% as_tibble() -> E_raw
-
-read_fst(normalized_path) %>% as_tibble() -> N
 
@@
 As expected, `model.frame` is the most used one.
-
-# Normalization of resolved expressions
-
-We join the normalized expressions and their statistics with the calls.
-
-```{r}
-E_cano <- E %>% left_join(N, by = c("hash" = "expr_resolved_hash")) # Common column is the hash of the full resolved expression
-```
-
-```{r}
-widespread_canonic <- E_cano %>%
-  group_by(expr_canonic) %>%
-  summarize(n_call_sites = n_distinct(src_ref), n_packages = n_distinct(ev_package))
-
-widespread_canonic %>%
-  arrange(desc(n_call_sites)) %>%
-  datatable()
-```

### 2021-02-23 analysis/normalized.Rmd
Port normalized/minimized expression from expressions.Rmd to normalized.Rmd

---
title: "Normalized expressions"
output:
    html_document:
        gallery: false
        toc: true
        toc_depth: 3
        toc_float: true
        df_print: paged
date: "`r Sys.Date()`"
editor_options: 
  chunk_output_type: console
params:
  base_dir: Data/package/
  calls_path: summarized.fst
  normalized_path: normalized-expressions.csv
  corpus_path: ../corpus.fst
  static_path: evals-static.csv
  force_rebuild: TRUE
---

```{r setup, include=FALSE}
library(tidyverse)
library(fst)
library(fs)
library(DT)
library(scales)

knitr::knit_hooks$set(time_it = local({
  now <- NULL
  function(before, options) {
    if (before) { # record the current time before each chunk
      now <<- Sys.time()
    } else { # calculate the time difference after a chunk
      res <- difftime(Sys.time(), now)
      # return a character string to show the time
      paste("Time for this code chunk to run:", round(res, 2), units(res))
    }
  }
}))
now <- Sys.time()
knitr::opts_chunk$set(
  echo = TRUE,
  fig.retina = 2,
  fig.width = 10,
  cache.lazy = FALSE,
  time_it = TRUE
)

source("inc/paths.R")
source("inc/latextags.R")
source("inc/functions.R")
theme_set(theme_minimal())
create_tags(path(TAGS_DIR, "usage_metrics.tex"), prefix = "", default = TRUE)

calls_path <- paste0(params$base_dir, params$calls_path)
corpus_path <- paste0(params$base_dir, params$corpus_path)
static_path <- paste0(params$base_dir, params$static_path)
normalized_path <- paste0(params$base_dir, params$normalized_path)
```

# Read

Read in the summarized calls data (`E_raw`) and the file that describes
which package belong to our corpus (`C_raw`). For conciseness, `E` is
the input data set with only the columns that we use. We also read the file with normalized expressions, `N`.

```{r read, message=TRUE, include=F}
read_fst(corpus_path) %>% as_tibble() -> C_raw
read_csv(static_path) %>% semi_join(C_raw, by = "package") -> P

read_fst(calls_path) %>% as_tibble() -> E_raw

read_csv(normalized_path, col_types = cols(
  minimized = col_character(),
  topcall = col_character(),
  is_model = col_logical(),
  has_fundef = col_logical(),
  has_calls = col_integer(),
  has_assigns = col_integer(),
  has_var = col_logical(),
  has_bracket = col_logical(),
  is_assign = col_logical(),
  is_value = col_logical(),
  is_ignore = col_logical(),
  has_dollar = col_logical(),
  has_user_call = col_logical(),
  has_block = col_logical(),
  normalized = col_character(),
  trimmed = col_character(),
  hash = col_character()
)) -> N

E_raw %>% select(
  run_package = package, #      The package which was run to trigger the eval
  ev_variant = eval_function, # Which eval variant was called
  duplicates = nb_ev_calls, #   Number of identical eval calls (weight)
  src_ref = eval_call_srcref, # Source ref for the eval call site
  file, #                       Script file name (name of this run)
  ev_package = eval_source, #   Package of the call site
  type = expr_resolved_type_tag, #   Argument type
  ev_expr = expr_expression,
  expr_resolved,
  func = expr_resolved_function, # the function (if any) at the root of the AST of expr_resolved
  parsed = expr_parsed_expression,
  hash = expr_resolved_hash,
  match.call = expr_match_call
) -> E

C_raw -> C
corpus_size <- nrow(C)
```

Some useful numbers:
```{r}
nb_call_sites <- E %>%
  pull(src_ref) %>%
  n_distinct()
```

There are `r nb_call_sites` in the dataset.

# Full dataset

We join the normalized expressions and their statistics with the calls.

```{r}
E_cano <- E %>% left_join(N, by = "hash") # Common column is the hash of the full resolved expression
```

# Ranking of normalized and minimized expressions

Normalized: 
```{r}
widespread_canonic <- E_cano %>%
  group_by(normalized) %>%
  summarize(n_call_sites = n_distinct(src_ref), n_packages = n_distinct(ev_package))

widespread_canonic %>%
  arrange(desc(n_call_sites)) %>%
  datatable()
```

Minimized:
```{r}
widespread_minimized <- E_cano %>%
  group_by(minimized) %>%
  summarize(n_call_sites = n_distinct(src_ref), n_packages = n_distinct(ev_package))

widespread_minimized %>%
  arrange(desc(n_call_sites)) %>%
  datatable()
```

# How does it evolve per call site

To check whether the normalization makes sense, we have a look at how many normalized expressions we have per call sites.

```{r}
# count hash because resolved expressions are trimmed but hash are on the full expression
by_src_ref <- E_cano %>% group_by(src_ref) %>%
  summarise(n_resolved = n_distinct(hash), n_normalized = n_distinct(normalized), n_minimized = n_distinct(minimized))
```

There should always be less normalized expressions than resolved ones. 
We can see for how many there are strictly less (or one when it was already one).

```{r}
decrease_exprs <- by_src_ref %>% filter(n_resolved == 1 | n_resolved > n_minimized)  %>% arrange(desc(n_minimized))
```

It corresponds to `r nrow(decrease_exprs)` call sites so only `r nb_call_sites - nrow(decrease_exprs)` have non-trivially not decreased the number of call sites.

# Top calls

For the functions at the root of the AST, per call site:

```{r}
widespread_topcall <- E_cano %>%
  group_by(topcall) %>%
  summarize(n_call_sites = n_distinct(src_ref)) %>% 
  arrange(desc(n_call_sites))
```

```{r}
big_topcall <- widespread_topcall %>% filter(n_call_sites > 3) 
```

There are `r nrow(big_topcall)` top calls out of `r nrow(widespread_topcall)` that are present in strictly more than 3 call sites.

```{r}
big_topcall %>% datatable()
```

# Function definitions

```{r}
fundefs <- E_cano %>% filter(has_fundef)

nb_fundef_callsites <- fundefs %>% pull(src_ref) %>% n_distinct()
```

There are `r nb_fundef_callsites` call sites with function definitions.

# Model.frame

```{r}
modelframes <- E_cano %>% filter(is_model)

nb_modelf_call_sites <- modelframes %>% pull(src_ref) %>% n_distinct()
```

There are `r nb_modelf_call_sites` call sites with `mode.frame`.

Is it usually in top call position?

```{r}
modelframes %>% filter(topcall != "model.frame")
```

Yes, always!

Is it correlated with `match.call`?

```{r}
nb_mf_match.call <- modelframes %>% filter(!is.na(match.call)) %>% pull(src_ref) %>% n_distinct()
```

`r nb_mf_match.call` call sites out of `model.frame` `r nb_modelf_call_sites` call sites use the `match.call` pattern.

We can also reciprocally look at how many call sites with `match.call` have `model.frame`:

```{r}
match.call_df <- E_cano %>% filter(!is.na(match.call))

nb_m.c_call_sites <- match.call_df %>% pull(src_ref) %>% n_distinct()

nb_m.c_mf_call_sites <- match.call_df %>% filter(is_model) %>% pull(src_ref) %>% n_distinct()
```

So `r ratio(nb_m.c_mf_call_sites, nb_m.c_call_sites)` of the call sites with the `match.call` pattern use `model.frame`.

# Polymorphism

We have only 15 different kind of minimized expression. Do some minimized expressions often go together?

### 2021-02-23 analysis/normalized.Rmd
Some more stats in the notebook

@@
 
-So `r ratio(nb_m.c_mf_call_sites, nb_m.c_call_sites)` of the call sites with the `match.call` pattern use `model.frame`.
+So `r ratio(nb_m.c_mf_call_sites, nb_m.c_call_sites)`% of the call sites with the `match.call` pattern use `model.frame`.
+
+# Assignments
+
+When do we perform assignments (causing potentially side effects)?
+
+```{r}
+assigns <- E_cano %>% filter(has_assigns > 0)
+
+nb_assigns_call_sites <- assigns %>% pull(src_ref) %>% n_distinct()
+```
+
+There are `r nb_assign_call_sites` call sites with assignments.
 
@@
 We have only 15 different kind of minimized expression. Do some minimized expressions often go together?
+
+# `match.call` pattern
+
+There are `r nb_m.c_call_sites` call sites using the `match.call`pattern.
+
+The top calls for this pattern are:
+
+```{r}
+match.call_df %>% 
+  group_by(topcall) %>%
+  summarize(n_call_sites = n_distinct(src_ref)) %>% 
+  arrange(desc(n_call_sites))
+```
+
+Most are `model.frame`. If we add other model functions, such as `glm` and `plm`:
+
+```{r}
+stat_functions <- c("model.frame", "glm", "plm", "elo.model.frame", "lmer", "randomForest", "betareg", "nlreg", "model.frame.default", "model.matrix", "arima", "glm.nb", "glmer") # There are many more
+
+nb_stat_call_sites <- match.call_df %>% 
+  filter(topcall %in% stat_functions) %>% 
+  pull(src_ref) %>% 
+  n_distinct()
+
+```
+
+it amounts for `r nb_stat_call_sites`.

### 2021-02-24 analysis/normalized.Rmd
More stats

@@
   has_block = col_logical(),
+  has_meta_op = col_logical(),
   normalized = col_character(),
@@
 
-There are `r nb_assign_call_sites` call sites with assignments.
-
-# Polymorphism
-
-We have only 15 different kind of minimized expression. Do some minimized expressions often go together?
+There are `r nb_assigns_call_sites` call sites with assignments.
 
@@
 
-it amounts for `r nb_stat_call_sites`.
+it amounts for `r nb_stat_call_sites` call sites.
+
+# Substitue&co
+
+`substitute`, `quote` and so on are used to do meta-programming.
+
+```{r}
+meta_df <- E_cano %>% filter(has_meta_op)
+
+nb_meta_call_sites <- meta_df %>% pull(src_ref) %>% n_distinct()
+```
+
+There are `r nb_meta_call_sites` call sites with `substitute` and `substitute`-like operators.
+
+Are the `substitute` sometimes not placed as a top call?
+
+```{r}
+meta_not_top <- meta_df %>% filter(!topcall %in% c("substitute", "quote", "enquote", "bquote"))
+
+nb_m_n_top_call_sites <- meta_not_top %>% pull(src_ref) %>% n_distinct()
+```
+
+It happens for `r nb_m_n_top_call_sites` call sites.
+
+# `eval(x[[condition]])` pattern
+
+See https://cran.r-project.org/doc/manuals/r-release/R-lang.html#switch
+
+```{r}
+eval_switch <- E_cano %>% filter(has_var, has_bracket, topcall == "[[")
+
+nb_ev_switch_call_sites <- eval_switch %>% pull(src_ref) %>% n_distinct()
+```
+
+There are `r nb_ev_switch_call_sites` call sites using that pattern.
+
+But this is actually not correct, because we should look at the non-resolved expression to find this pattern.
+
+# Polymorphism in more details
+
+We have only 15 different kind of minimized expression. Do some minimized expressions often go together?

### 2021-02-24 analysis/normalized.Rmd
Precise the preprocessing/postprocessing insertion of blocks

@@
 
-There are `r nb_call_sites` in the dataset.
+There are `r nb_call_sites` call sites in the dataset.
 
@@
   summarise(n_resolved = n_distinct(hash), n_normalized = n_distinct(normalized), n_minimized = n_distinct(minimized))
+```
+
+```{r}
+by_src_ref %>% datatable()
 ```
@@
 
+# Complex blocks
+
+`eval` can evaluate sometimes complex blocks.
+
+```{r}
+blocks <- E_cano %>% filter(has_block)
+
+nb_blocks_call_sites <- blocks %>% pull(src_ref) %>% n_distinct()
+```
+
+There are `r nb_blocks_call_sites` such call sites.
+
+`eval` is used to insert blocks of code inside the function, for instance as a preprocessing and postprocessing step.
+There are some examples in the VGAM package.
+In that case, `{` will be at the root of the AST.
+
+```{r}
+pre_post <- blocks %>% filter(topcall == "{")
+
+nb_pre_post_call_sites <- pre_post %>% pull(src_ref) %>% n_distinct()
+```
+
+It happens in `r nb_pre_post_call_sites` call sites so most of the call sites with blocks.
+
 # `eval(x[[condition]])` pattern

### 2021-02-25 analysis/normalized.Rmd
Nb of interp eval per minimized expression

@@
   hash = expr_resolved_hash,
-  match.call = expr_match_call
+  match.call = expr_match_call,
+  n_op = interp_eval
 ) -> E
@@
   group_by(minimized) %>%
-  summarize(n_call_sites = n_distinct(src_ref), n_packages = n_distinct(ev_package))
+  summarize(n_call_sites = n_distinct(src_ref), n_packages = n_distinct(ev_package), n_operations = mean(n_op, na.rm=TRUE))
 
@@
 ```
+
+`_operations` is the average of R interpreter eval loop per minimized expression.
 

### 2021-02-25 analysis/normalized.Rmd
Environment usage per minimized expression

--- 2021-02-24 analysis/normalized.Rmd
@@
   hash = expr_resolved_hash,
-  match.call = expr_match_call
+  match.call = expr_match_call,
+  n_op = interp_eval,
+  envir = envir_expression
 ) -> E
@@
   group_by(minimized) %>%
-  summarize(n_call_sites = n_distinct(src_ref), n_packages = n_distinct(ev_package))
+  summarize(n_call_sites = n_distinct(src_ref), n_packages = n_distinct(ev_package), n_operations = mean(n_op, na.rm=TRUE), prop_envir = sum(!is.na(envir)) / n())
 
@@
 ```
+
+`n_operations` is the average of R interpreter eval loop per minimized expression.
+`prop_envir` shows how much often an *explicit* environment is passed to `eval`.
+Looking up to a variable and applying a function to a variable, are the two non-trivial categories that most of the time explicitly use the `envir` argument of `eval`.
 

### 2021-02-26 analysis/normalized.Rmd
Generate notebook for each dataset
Declutter normalized.Rmd

@@
 read_fst(corpus_path) %>% as_tibble() -> C_raw
-read_csv(static_path) %>% semi_join(C_raw, by = "package") -> P
+#read_csv(static_path) %>% semi_join(C_raw, by = "package") -> P
 
@@
 ```{r}
-by_src_ref %>% datatable()
+by_src_ref %>% arrange(desc(n_resolved)) %>% datatable()
 ```

### 2021-02-26 analysis/normalized.Rmd
Parse and plot pattern

@@
 
-# Substitue&co
+# Substitute&co
 
@@
 
-# `eval(x[[condition]])` pattern
+# `Parse` and plotting functions
 
-See https://cran.r-project.org/doc/manuals/r-release/R-lang.html#switch
+We approximate the plotting function by just looking for `plot` in their name:
 
 ```{r}
-eval_switch <- E_cano %>% filter(has_var, has_bracket, topcall == "[[")
+plotting <- E_cano %>% filter(str_detect(topcall, fixed("plot")))
 
-nb_ev_switch_call_sites <- eval_switch %>% pull(src_ref) %>% n_distinct()
+nb_plotting_call_sites <- plotting %>% pull(src_ref) %>% n_distinct()
 ```
 
-There are `r nb_ev_switch_call_sites` call sites using that pattern.
+There are `r nb_plotting_call_sites` plotting call sites.
 
-But this is actually not correct, because we should look at the non-resolved expression to find this pattern.
+Are they built using parse? For that, we look at the non-resolved expression (`parse` will disappear after).
+
+Currently, we use a regex. When `parsed` column is fixed, we will get more accurate results.
+
+```{r}
+parse_plot <- plotting %>% filter(str_detect(ev_expr, fixed("parse")))
+
+nb_parse_plot_call_sites <- parse_plot %>% pull(src_ref) %>% n_distinct()
+```
+
+This already amounts to `r nb_parse_plot_call_sites` call sites.
 

### 2021-02-26 analysis/normalized.Rmd
More on parse and plot pattern
Improve

@@
 ```{r}
+widespread_topcall %>% datatable()
+```
+
+```{r}
 big_topcall <- widespread_topcall %>% filter(n_call_sites > 3) 
@@
 There are `r nrow(big_topcall)` top calls out of `r nrow(widespread_topcall)` that are present in strictly more than 3 call sites.
-
-```{r}
-big_topcall %>% datatable()
-```
 
@@
 
+Is the `envir` argument non trivial?
+
+```{r}
+assign_env <- assigns %>% filter(!is.na(envir))
+
+nb_a_env_call_sites <- assign_env %>% pull(src_ref) %>% n_distinct()
+```
+
+There are `r nb_a_env_call_sites` call sites where `envir` is non-trivial, with an assignment.
+
 # `match.call` pattern
@@
 ```{r}
-stat_functions <- c("model.frame", "glm", "plm", "elo.model.frame", "lmer", "randomForest", "betareg", "nlreg", "model.frame.default", "model.matrix", "arima", "glm.nb", "glmer") # There are many more
+stat_functions <- c("model.frame", "glm", "lm", "plm", "elo.model.frame", "lmer", "randomForest", "betareg", "nlreg", "model.frame.default", "model.matrix", "arima", "glm.nb", "glmer") # There are many more
 
@@
 ```{r}
-plotting <- E_cano %>% filter(str_detect(topcall, fixed("plot")))
+plotting <- E_cano %>% filter(str_detect(normalized, fixed("plot"))) # not in topcall because f it results from parse...
 

### 2021-02-27 analysis/normalized.Rmd
Most common minimized exprs for one call site

@@
 
+### Infix operators
+
+```{r}
+infix <- E_cano %>% filter(str_detect(topcall, "%[^%]*%"))
+
+nb_infix_call_sites <- infix %>% pull(src_ref) %>% n_distinct()
+```
+
+It happens in `r nb_infix_call_sites` call sites.
+
+The packages that use or define those infix operators:
+
+```{r}
+infix %>% group_by(ev_package) %>% summarize(n_call_sites = n_distinct(src_ref)) %>% arrange(desc(n_call_sites)) %>% datatable()
+```
+
 # Polymorphism in more details
@@
 We have only 15 different kind of minimized expression. Do some minimized expressions often go together?
+
+```{r}
+minimized_exprs <- E_cano %>% pull(minimized) %>% unique()
+
+create_set <- function(l) {
+  s <- rep.int("F", length(minimized_exprs))
+  names(s) <- minimized_exprs
+
+  for(exp in l) {
+    s[[exp]] = "T"
+  }
+
+  return(paste0(s, collapse = ""))
+}
+
+show_set <- function(s) {
+  l <- strsplit(s, split="")[[1]]
+  names(l) <- minimized_exprs
+
+  l <- l[which(l == "T")]
+  return(paste0(names(l), collapse = "; "))
+}
+
+poly <- E_cano %>% group_by(src_ref) %>%
+    summarize(common_minimized = create_set(minimized), n_exprs = n_distinct(minimized)) 
+```
+
+Now, we can show the mot common minimized expressions when there are more than 2 for one call site:
+
+```{r}
+poly %>% filter(n_exprs > 1) %>%
+  count(common_minimized,  sort = TRUE) %>%
+  mutate(common_minimized = map_chr(common_minimized, show_set)) %>% 
+  datatable()
+```

### 2021-03-01 analysis/usage.Rmd
Coverage per package

@@
 We compute the frequency of evals by package. This shows that some packages
-hardly use eval while other use it intensly. The numbers are "for all runs",
+hardly use eval while other use it intensely. The numbers are "for all runs",
 this means that a popular package may be triggered by more runs and thus
@@
 `r TBL`
+
+```{r coverage-per-package}
+cov_packages <- static_call_sites_per_package %>%
+  left_join(sites_per_package, by = c("package" = "ev_package"), suffix = c("_static", "_tracer")) %>% 
+  mutate(n_tracer = replace_na(n_tracer, 0L))
+
+# if n_static < n_tracer, adjust n_tracer to n_static
+# Probably due to higher-order functions taking some eval
+cov_packages <- cov_packages %>%
+  mutate(n_static = if_else(n_static < n_tracer, n_tracer, n_static))
+
+cov_packages <- cov_packages %>% mutate(coverage = 100 * n_tracer / n_static)
+
+cov_packages %>% ggplot() +
+  geom_histogram(aes(x = coverage)) + 
+  labs(x = "coverage (%)")
+
+ggsave(path(PLOT_DIR, "coverage_per_package.pdf"))
+```
 

### 2021-03-02 analysis/usage.Rmd
Expression coverage

@@
 cov_packages %>% ggplot() +
-  geom_histogram(aes(x = coverage)) + 
+  geom_histogram(aes(x = coverage), binwidth=1) + 
   labs(x = "coverage (%)")
@@
 ```
+
+What is the actual coverage vs the expected coverage?
+```{r}
+expected_cov <- C_raw %>% select(package, coverage_expr) %>% 
+  mutate(coverage_expr = replace_na(coverage_expr, 0)) %>% 
+  mutate(coverage_expr = 100 * coverage_expr)
+
+cov_diff <- expected_cov %>% left_join(cov_packages, by = "package") %>% 
+  mutate(cdiff = pmax(0, coverage_expr - coverage))
+
+cov_diff %>% ggplot() +
+  geom_histogram(aes(x = cdiff), binwidth = 1) + 
+  labs(x = "coverage difference (%)")
+```
+
+Most of the packages are in the first bin, where the difference is less than 1%, so the expression coverage is a good predictor of the actual `eval` coverage.
 

### 2021-03-03 analysis/normalized.Rmd
Percentages

@@
   group_by(minimized) %>%
-  summarize(n_call_sites = n_distinct(src_ref), n_packages = n_distinct(ev_package), n_operations = mean(n_op, na.rm=TRUE), prop_envir = sum(!is.na(envir)) / n())
+  summarize(n_call_sites = n_distinct(src_ref), proportion = ratio(n_call_sites, nb_call_sites), n_packages = n_distinct(ev_package), n_operations = round(mean(n_op, na.rm=TRUE), 2), percent_envir = round(100 * sum(!is.na(envir)) / n(), 2))
 
@@
   group_by(topcall) %>%
-  summarize(n_call_sites = n_distinct(src_ref)) %>% 
+  summarize(n_call_sites = n_distinct(src_ref), percent = round(ratio(n_call_sites, nb_call_sites), 2)) %>% 
   arrange(desc(n_call_sites))
@@
 
-There are `r nb_fundef_callsites` call sites with function definitions.
+There are `r nb_fundef_callsites` call sites with function definitions, i.e. `r ratio(nb_fundef_callsites, nb_call_sites)` %.
 
@@
 
-There are `r nb_modelf_call_sites` call sites with `mode.frame`.
+There are `r nb_modelf_call_sites` call sites with `mode.frame`, i.e. `r ratio(nb_modelf_call_sites, nb_call_sites)`%.
 
@@
 
-`r nb_mf_match.call` call sites out of `model.frame` `r nb_modelf_call_sites` call sites use the `match.call` pattern.
+`r nb_mf_match.call` call sites out of `model.frame` `r nb_modelf_call_sites` call sites use the `match.call` pattern, i.e. `r ratio(nb_mf_match.call, nb_modelf_call_sites)`%.
 
@@
 
-There are `r nb_assigns_call_sites` call sites with assignments.
+There are `r nb_assigns_call_sites` call sites with assignments, i.e. `r ratio(nb_assigns_call_sites, nb_call_sites)`%.
 
@@
 
-There are `r nb_a_env_call_sites` call sites where `envir` is non-trivial, with an assignment.
+There are `r nb_a_env_call_sites` call sites where `envir` is non-trivial, with an assignment, i.e. `r ratio(nb_a_env_call_sites, nb_assigns_call_sites)`% of the call sites with assignments.
 
@@
 
-There are `r nb_m.c_call_sites` call sites using the `match.call`pattern.
+There are `r nb_m.c_call_sites` call sites using the `match.call`pattern, i.e. `r ratio(nb_m.c_call_sites, nb_call_sites)`%.
 
@@
 
-it amounts for `r nb_stat_call_sites` call sites.
+it amounts for `r nb_stat_call_sites` call sites, i.e `r ratio(nb_stat_call_sites, nb_call_sites)`%.
 
@@
 
-It happens for `r nb_m_n_top_call_sites` call sites.
+It happens for `r nb_m_n_top_call_sites` call sites, i.e. `r ratio(nb_m_n_top_call_sites, nb_call_sites)`%.
 
@@
 
-There are `r nb_blocks_call_sites` such call sites.
+There are `r nb_blocks_call_sites` such call sites, i.e. `r ratio(nb_blocks_call_sites, nb_call_sites)`%.
 
@@
 
-It happens in `r nb_pre_post_call_sites` call sites so most of the call sites with blocks.
+It happens in `r nb_pre_post_call_sites` call sites so most of the call sites with blocks (`r ratio(nb_pre_post_call_sites, nb_blocks_call_sites)`%).
 
@@
 
-There are `r nb_plotting_call_sites` plotting call sites.
+There are `r nb_plotting_call_sites` plotting call sites, i.e. `r ratio(nb_plotting_call_sites, nb_call_sites)`%.
 
@@
 
-This already amounts to `r nb_parse_plot_call_sites` call sites.
+This already amounts to `r nb_parse_plot_call_sites` call sites, i.e. `r ratio(nb_parse_plot_call_sites, nb_plotting_call_sites)`% of the plotting call sites.
 
@@
 
-It happens in `r nb_infix_call_sites` call sites.
+It happens in `r nb_infix_call_sites` call sites, i.e. `r ratio(nb_infix_call_sites, nb_call_sites)`%.
 
@@
   mutate(common_minimized = map_chr(common_minimized, show_set)) %>% 
+  mutate(percent = round(100 * n / nb_call_sites, 2)) %>% 
   datatable()

### 2021-03-03 analysis/usage.Rmd
Graph of coverage vs revdep vs nb of static evals

--- 2021-03-01 analysis/usage.Rmd
@@
 cov_packages <- cov_packages %>%
-  mutate(n_static = if_else(n_static < n_tracer, n_tracer, n_static))
+  mutate(n_static_c = if_else(n_static < n_tracer, n_tracer, n_static))
 
-cov_packages <- cov_packages %>% mutate(coverage = 100 * n_tracer / n_static)
+cov_packages <- cov_packages %>% mutate(coverage = 100 * n_tracer / n_static_c)
 
 cov_packages %>% ggplot() +
-  geom_histogram(aes(x = coverage)) + 
+  geom_histogram(aes(x = coverage), binwidth=1) + 
   labs(x = "coverage (%)")
@@
 ggsave(path(PLOT_DIR, "coverage_per_package.pdf"))
+```
+
+What is the actual coverage vs the expected coverage?
+```{r}
+corpus_cov<- C_raw %>% select(package, size, revdeps)
+
+corpus_cov<- corpus_cov %>% left_join(cov_packages, by = "package") %>%
+  mutate(revdeps = replace_na(revdeps, 0))
+
+corpus_cov %>% ggplot() +
+  geom_point(aes(x = coverage, y = revdeps, size = size)) + 
+  scale_y_log10() +
+  scale_size_area() +
+  labs(x = "coverage (%)", y = "Reverse dependencies")
+
+ggsave(path(PLOT_DIR, "coverage_per_package_size.pdf"))
+```
+
+```{r}
+corpus_cov %>% ggplot() +
+  geom_point(aes(x = coverage, y = revdeps, size = n_static)) + 
+  scale_y_log10() +
+  scale_size_area() +
+  labs(x = "coverage (%)", y = "Reverse dependencies", size = "Static call sites")
+
+ggsave(path(PLOT_DIR, "coverage_per_package_static_call_sites.pdf"))
 ```

### 2021-03-04 paper/main.tex
Update undefined calls
Use asmsmall document class (OOPSLA)

@@
-\documentclass[USenglish,cleveref, autoref, thm-restate]{lipics-v2019}
+%\documentclass[USenglish,cleveref, autoref, thm-restate]{lipics-v2019}
+\documentclass[acmsmall]{acmart}
 \usepackage{listings,hyperref,multirow,paralist,xspace,url,wrapfig}
@@
 
+\begin{document}
+
 \title{Why do we Eval?\\[2mm]\Large A large-scale study of Eval usage in R}
-\titlerunning{Why Eval (in R)?}
-
-\begin{document}
-\maketitle
+%\titlerunning{Why Eval (in R)?}
+
+\author{Author }
+\authornote{Both authors contributed equally to this research.}
+\email{a@b.com}
+\orcid{}
+\author{Author}
+\authornotemark[1]
+\email{r@s.com}
+\affiliation{%
+	\institution{Institute for Clarity in Documentation}
+	\streetaddress{P.O. Box 1212}
+	\city{Dublin}
+	\state{Ohio}
+	\country{USA}
+	\postcode{43017-6221}
+}
 
@@
 \end{abstract}
+
+\maketitle
 
@@
 %
-We do not get source locations for \UndefinedEvalsRnd UPDATE!! \eval calls. In
+We do not get source locations for \CranUndefinedRnd \eval calls. In
 these cases \eval is either passed as an argument to a higher-order functions or
@@
 generated called from package native code. However, this is a meager
-\PercentUndefinedEval UPDATE!! of all \eval calls and is unlikely to affect our
+\CranUndefinedRatio of all \eval calls and is unlikely to affect our
 analysis. We ignore calls to the native \eval function exposed by R. We also

### 2021-03-04 paper/main.tex
Conf name
Adjust figures for the acmsmall document class

@@
 
-\title{Why do we Eval?\\[2mm]\Large A large-scale study of Eval usage in R}
+\title{Why do we Eval?\\\Large A large-scale study of Eval usage in R}
 %\titlerunning{Why Eval (in R)?}
@@
   for JavaScript.
-  %% TODO ::: Add a sentance about summarizing why users resort to
+  %% TODO ::: Add a sentence about summarizing why users resort to
   %% eval -- Jan
@@
 
-  \includegraphics[width=5.8cm, trim=7.5cm 0 6cm 0, clip]{calls_per_run_per_call_site}
+  \includegraphics[width=5.4cm, trim=5.5cm 0 5cm 0, clip]{calls_per_run_per_call_site}
 
@@
 \begin{figure}[h!]
- \includegraphics[width=0.5\textwidth, trim=5.5cm 0 6cm 0, clip]{size_loaded_distribution}
+ \includegraphics[width=0.5\textwidth, trim=4.8cm 0 6cm 0, clip]{size_loaded_distribution}
 \caption{Loaded code} \label{fig:sizedistribution}
@@
 \end{figure}
+
+\paragraph{Kaggle}
+
+\paragraph{Base}
 

### 2021-03-04 analysis/expressions.Rmd
Generate separate tag files and names for each dataset

@@
 theme_set(theme_minimal())
-create_tags(path(TAGS_DIR, "usage_metrics.tex"), prefix = "", default = TRUE)
+
+dataset_name <- basename(params$base_dir)
+create_tags(path(TAGS_DIR, paste0(dataset_name, "_expressions.tex")), prefix = dataset_name, default = TRUE)
 

### 2021-03-04 analysis/normalized.Rmd
Generate separate tag files and names for each dataset

@@
 theme_set(theme_minimal())
-create_tags(path(TAGS_DIR, "usage_metrics.tex"), prefix = "", default = TRUE)
+
+dataset_name <- basename(params$base_dir)
+
+# Tags will be prefixed by the dataset name
+create_tags(path(TAGS_DIR, paste0(dataset_name, "_normalized_expr.tex")), prefix = dataset_name, default = TRUE)
 

### 2021-03-04 analysis/usage.Rmd
Generate separate tag files and names for each dataset

@@
 params:
-  base_dir: Data/package/
+  base_dir: Data/kaggle/
   calls_path: summarized.fst
@@
 theme_set(theme_minimal())
-create_tags(path(TAGS_DIR, "usage_metrics.tex"), prefix = "", default = TRUE)
+
+dataset_name <- basename(params$base_dir)
+create_tags(path(TAGS_DIR, paste0(dataset_name, "_usage_metrics.tex")), prefix = dataset_name, default = TRUE)
 
@@
 static_path <- paste0(params$base_dir, params$static_path)
+
+lock_path <- paste0(params$base_dir, ".lock.fst")
+min_path <- paste0(params$base_dir, "E.fst")
 ```
@@
 # To force reading, set the force_rebuild parameter to true.
-if (file.exists("Data/.lock.fst")) {
-  saved <- read.fst("Data/.lock.fst")[[1,1]]
+if (file.exists(lock_path)) {
+  saved <- read.fst(lock_path)[[1,1]]
   if (saved == file.size(calls_path)) {
-    read.fst("Data/E.fst") %>% as_tibble() -> E
+    read.fst(min_path) %>% as_tibble() -> E
     rebuild <- FALSE
@@
    mutate(across(where(is_character),as_factor)) -> E 
-  write.fst(E, "Data/E.fst")
+  write.fst(E, min_path)
   sz <- file.size(calls_path) %>% as_tibble()
-  write.fst(sz,"Data/.lock.fst")
+  write.fst(sz, lock_path)
 }
@@
 ```{r exercised-vs-static, include=FALSE}
-stat_evl <- P %>% filter(call_fun_name == "eval")        %>% nrow()
+# stat_evl <- P %>% filter(call_fun_name == "eval")        %>% nrow()
 stat_evq <- P %>% filter(call_fun_name == "evalq")       %>% nrow()
@@
 
-ggplot2_symbs <- E %>% filter(ev_package == "ggplot2") %>% 
-  count(type, wt=duplicates) %>% filter(type == "symbol") %>% .$n
+ggplot2_symbs <- E %>% filter(ev_package == "ggplot2")
+
+# For Kaggle, there are not eval calls originating from ggplot2
+ggplot2_symbs <- if(nrow(ggplot2_symbs) == 0 ) {
+  0 
+} else {
+  ggplot2_symbs %>% count(type, wt=duplicates) %>% filter(type == "symbol") %>% .$n
+}
 

### 2021-03-04 analysis/usage.Rmd
Improve

@@
 read_fst(corpus_path) %>% as_tibble() -> C_raw
+
+# We do the semi join to only keep the file that are loadable
 read_csv(static_path) %>% semi_join(C_raw, by = "package") -> P
@@
 ```{r exercised-vs-static, include=FALSE}
-# stat_evl <- P %>% filter(call_fun_name == "eval")        %>% nrow()
+ stat_evl <- P %>% filter(call_fun_name == "eval")        %>% nrow()
 stat_evq <- P %>% filter(call_fun_name == "evalq")       %>% nrow()

### 2021-03-08 analysis/usage.Rmd
Fix number of language expression for Kaggle

@@
   calls_path: summarized.fst
-  corpus_path: ../corpus.fst
+  corpus_path: corpus.txt
   static_path: evals-static.csv
@@
 ```{r read, message=TRUE, include=F}
-read_fst(corpus_path) %>% as_tibble() -> C_raw
+tibble(package = read_lines(corpus_path)) -> C_raw
 
@@
 
-```{r coverage-per-package}
-cov_packages <- static_call_sites_per_package %>%
-  left_join(sites_per_package, by = c("package" = "ev_package"), suffix = c("_static", "_tracer")) %>% 
-  mutate(n_tracer = replace_na(n_tracer, 0L))
-
-# if n_static < n_tracer, adjust n_tracer to n_static
-# Probably due to higher-order functions taking some eval
-cov_packages <- cov_packages %>%
-  mutate(n_static_c = if_else(n_static < n_tracer, n_tracer, n_static))
-
-cov_packages <- cov_packages %>% mutate(coverage = 100 * n_tracer / n_static_c)
-
-cov_packages %>% ggplot() +
-  geom_histogram(aes(x = coverage), binwidth=1) + 
-  labs(x = "coverage (%)")
-
-ggsave(path(PLOT_DIR, "coverage_per_package.pdf"))
-```
-
-What is the actual coverage vs the expected coverage?
-```{r}
-corpus_cov<- C_raw %>% select(package, size, revdeps)
-
-corpus_cov<- corpus_cov %>% left_join(cov_packages, by = "package") %>%
-  mutate(revdeps = replace_na(revdeps, 0))
-
-corpus_cov %>% ggplot() +
-  geom_point(aes(x = coverage, y = revdeps, size = size)) + 
-  scale_y_log10() +
-  scale_size_area() +
-  labs(x = "coverage (%)", y = "Reverse dependencies")
-
-ggsave(path(PLOT_DIR, "coverage_per_package_size.pdf"))
-```
-
-```{r}
-corpus_cov %>% ggplot() +
-  geom_point(aes(x = coverage, y = revdeps, size = n_static)) + 
-  scale_y_log10() +
-  scale_size_area() +
-  labs(x = "coverage (%)", y = "Reverse dependencies", size = "Static call sites")
-
-ggsave(path(PLOT_DIR, "coverage_per_package_static_call_sites.pdf"))
-```
-
 # Numbers of calls to eval
@@
 symbs <- types %>% filter(type == "symbol") %>% .$n
-languages <- types %>% filter(type == "language") %>% .$n
+languages <- types %>% filter(type == "language") 
+languages <- if(nrow(languages) == 0) 0 else languages$n
+
 expressions <- types %>% filter(type == "expression") %>% .$n

### 2021-03-08 paper/main.tex
Usage metrics: kaggle

@@
 
-\paragraph{Kaggle}
-
 \paragraph{Base}
+
+\paragraph{Kaggle} In the Kaggle dataset, \kaggleNbruns packages trigger at least one \eval.
+In total, \kaggleAllcalls \eval were recorded.
+Most of them are call to the normal \eval, whereas a few, \kaggleLocals, are calls to \c{local}.
+They are due to the code loading libraries.
+Out of \kaggleStaticeval static \eval sites, we exercise \kaggleTriggeredeval of them.
+Mot \eval are invoked only once.
+Only one \eval amounts for several thousand, \kaggleMaxcalls,  invocations.
+Code amounts for \kaggleCodepercent of arguments in the kaggle corpus.
+However, contrary to the CRAN corpus, they are all expressions, and there are very few symbols (\kaggleSymbolpercent) and no language objects.
+Most expression objects result from a call to the \c{parse} function, suggesting that most \eval in Kaggle parse and evaluate strings.
+The median argument size is \kaggleMedianszeval, which makes sense, as there are very few arguments that are only symbols.
+The largest argument size is \kaggleMaxszeval.
+The distribution of events is similar to the one for CRAN: \kaggleSmalleventspct of \eval calls amount for less than 50 events.
+
+\paragraph{Discussion}
 
@@
   observations
-\item Add paragraphs for Kaggle and CORE -- only discuss the
-  differences from CRAN
 \end{itemize}

### 2021-03-08 analysis/usage.Rmd
Fix problems with notebook generation per dataset

@@
 params:
-  base_dir: Data/kaggle/
+  base_dir: Data/package/
   calls_path: summarized.fst
@@
 create_tags(path(TAGS_DIR, paste0(dataset_name, "_usage_metrics.tex")), prefix = dataset_name, default = TRUE)
+#create_tags(path(TAGS_DIR, "usage_metrics.tex"), default = TRUE)
 

### 2021-03-08 paper/main.tex
Usage metrics: base

@@
 This section focuses on our corpus of \Corpus CRAN packages that use
-\eval; these represent {\bf XX\%} of the entire repository. To provide
+\eval; these represent {\bf 28.9\%} of the entire repository. To provide % \Corpus / \CranAll 
 a picture of \eval usage, all runnable code in CRAN was executed. Runs
@@
 
-\paragraph{Base}
+\paragraph{Base} The base dataset contains 13 packages.
+All of them trigger at least one \eval call.
+In total, \baseAllcalls \eval  were recorded.
+Most of them are calls to he normal \eval.
+There are a few calls to the other variants, less than 4\%.
+Contrary to the CRAN dataset, there are calls to \c{evalq} (\baseEvalqs).
+Out of \baseStaticeval static sites, \baseTriggeredeval are exercised, i.e.
+\baseTriggeredevalpct.
+Most of the arguments, \baseCodepercent,  are code objects.
+However, contrary to CRAN, few, only \baseSymbolpercent, are symbols, but rather language objects, \baseLanguagepercent.
+There are also few expression objects: \baseExpressionpercent.
+The median argument size is \baseMedianszeval, which is more than for CRAN, which we dominated by symbol evaluation, whereas the maximum size is \baseMaxszeval.
+Small event evals (less than 50 events) amount for \baseSmalleventspct of the events in \eval calls.
+
+% Figures for evalq are weird: more triggered than static call sites
 

### 2021-03-09 paper/main.tex
Add table comparing eval for Julia, Javascript, Python and R

@@
 \usepackage{listings,hyperref,multirow,paralist,xspace,url,wrapfig}
+\usepackage{tabularx}
 \input{macro}
@@
 
+\section{Power scale of \eval }
+
+\begin{table}[h]
+	\begin{tabularx}{\textwidth}{|X|X|X|X|X|}
+		\hline
+		& Julia & Javascript & Python & R \tabularnewline
+		\hline % lstlinline does not like tabular environments...
+		{\eval} & \texttt{eval(expr)} & \texttt{eval(expr)} & \texttt{eval(expr[, globals[, locals]])} &  \texttt{eval(expr[, envir[, enclos]])}  \tabularnewline
+		\hline 
+		Environments &Global  & Local & Global or local  & Any \tabularnewline
+		\hline
+	\end{tabularx}
+	\caption{Relative power of \eval throughout languages: in which environment it can evaluate its expression.}
+\end{table}
+
 \section{Conclusion}

### 2021-03-09 paper/main.tex
Discussion for usage metrics

@@
 
-\paragraph{Discussion}
+\paragraph{Discussion} \emph{Base} is more akin to use \eval for metaprogramming, \emph{kaggle} to evaluate parsed strings, and CRAN is dominated by symbol evaluation.
+In terms of number of calls, a few calls dominate, with a call from \emph{ggplot} being prominent in the CRAN corpus.
+When looking at the \eval sites, \eval is more spread out.
+These two measures emphasize two different points of view:
+
+\begin{itemize}
+	\item The number of calls is relevant to an optimizing compiler: the most frequent calls are the ones to be optimized first.
+	\item The number of sites is relevant to the programmer/user: it describes better how programmers use \eval.
+\end{itemize}
+
+% What about a threat model?
 
@@
   an \eval (5\%) is correct. That feels surprising to me.
-\item I am having second thoughts about keeping the \c{ggplot2} data
-  -- it is a big outlier that skews all the graphs...
-\item Add a ``Discussion'' paragraph at the end that sums up the
-  observations
 \end{itemize}

### 2021-03-09 paper/main.tex
Base usage metrics

@@
 Contrary to the CRAN dataset, there are calls to \c{evalq} (\baseEvalqs).
-Out of \baseStaticeval static sites, \baseTriggeredeval are exercised, i.e.
-\baseTriggeredevalpct.
+Out of \baseStaticeval \eval static sites, \baseTriggeredeval are exercised, i.e.
+\baseTriggeredevalpct, out of \baseStaticevalq, \baseTriggeredevalq are exercised, and out of \baseStaticevalparent, \baseTriggeredevalparent are exercised.
 Most of the arguments, \baseCodepercent,  are code objects.
@@
 There are also few expression objects: \baseExpressionpercent.
-The median argument size is \baseMedianszeval, which is more than for CRAN, which we dominated by symbol evaluation, whereas the maximum size is \baseMaxszeval.
+The median argument size is \baseMedianszeval, which is more than for CRAN, which was dominated by symbol evaluation.The maximum size is \baseMaxszeval.
 Small event evals (less than 50 events) amount for \baseSmalleventspct of the events in \eval calls.
@@
 When looking at the \eval sites, \eval is more spread out.
-These two measures emphasize two different points of view:
+Those two measures emphasize two different points of view:
 
@@
 
-\section{Power scale of \eval }
-
 \begin{table}[h]

### 2021-03-09 analysis/inc/latextags.R
Add a way to prefix output of plots

@@ -15,7 +15,7 @@ my_datatable <- function(df, page_size=20, round=TRUE, ...) {
   } else {
     list(paging=TRUE, searching=TRUE, info=TRUE, pageLength=page_size)
   }
-  
+
   table <- datatable(df, options=options, ...)
   #if (round) {
   #  table <- formatRound(table, sapply(df, is.numeric), 2)
@@ -37,7 +37,7 @@ rs <- function(name, value) {
 r <- function(name, value, ...) {
   stopifnot(is.character(name))
   stopifnot(length(name) == 1)
-  UseMethod("r", value)  
+  UseMethod("r", value)
 }
 
 r.numeric <- function(name, value, ...) {
@@ -51,12 +51,12 @@ r.numeric <- function(name, value, ...) {
 
 r.Duration <- function(name, value, ...) {
   units <- c("years", "days", "hours", "minutes", "seconds")
-  
+
   for (u in units) {
     v <- as.numeric(value, units=u)
     class(v) <- "num_with_suffix"
     attr(v, "suffix") <- paste0(" ", u)
-    
+
     if (all(v > 1)) break
   }
 
@@ -74,10 +74,10 @@ r.difftime <- function(name, value, ...) {
 
 r.default <- function(name, value, ...) {
   stopifnot(length(value) == 1)
-  
+
   value <- fmt(value, ...)
   tag_name <- tag(name, value)
-  
+
   tibble(name=str_c(name, " (\\", tag_name, ")"), value=value)
 }
 
@@ -123,7 +123,7 @@ create_tags <- function(filename, prefix="", default=TRUE) {
 }
 
 clear_tags <- function(tags=get_default_tags()) {
-  tags$values <- tibble::tibble(name=character(), 
+  tags$values <- tibble::tibble(name=character(),
                                 value=character(), latex=character())
 }
 
@@ -157,16 +157,16 @@ tag <- function(name, value, tags=get_default_tags()) {
   if (is.null(tags)) {
     return(value)
   }
-  
+
   stopifnot(is.character(name))
   stopifnot(length(name)==1)
-  
+
   value <- as.character(value)
   latex_tag_name <- generate_latex_command_name(name, prefix=tags_prefix(tags))
   latex <- generate_latex_command(latex_tag_name, value)
-  
+
   existing <- which(tags$values$name == name)
-  
+
[…]

### 2021-03-09 analysis/usage.Rmd
Prefix plot pdf by the dataset name

@@
 params:
-  base_dir: Data/package/
+  base_dir: Data/base/
   calls_path: summarized.fst
@@
 
-ggsave(path(PLOT_DIR, "eval_calls_per_packages.pdf"))
+ggsave_prefix("eval_calls_per_packages.pdf")
 ```
@@
 
-ggsave(path(PLOT_DIR, "size_loaded_distribution.pdf"))
+ggsave_prefix("size_loaded_distribution.pdf")
 ```
@@
 
-ggsave(path(PLOT_DIR, "events_per_pack_large.pdf"))
+ggsave_prefix("events_per_pack_large.pdf")
 ```
@@
 
-ggsave(path(PLOT_DIR, "events_per_pack_small.pdf"))
+ggsave_prefix("events_per_pack_small.pdf")
 ```
@@
 
-ggsave(path(PLOT_DIR, "calls_per_run_per_call_site.pdf"))
+ggsave_prefix("calls_per_run_per_call_site.pdf")
 ```

### 2021-03-10 paper/main.tex
Usage metrics

@@
 In total, \baseAllcalls \eval  were recorded.
-Most of them are calls to he normal \eval.
+Most of them are calls to the main \eval.
 There are a few calls to the other variants, less than 4\%.
@@
 Most of the arguments, \baseCodepercent,  are code objects.
-However, contrary to CRAN, few, only \baseSymbolpercent, are symbols, but rather language objects, \baseLanguagepercent.
+However, contrary to CRAN, few arguments, only \baseSymbolpercent, are symbols, but are rather language objects, \baseLanguagepercent.
 There are also few expression objects: \baseExpressionpercent.
-The median argument size is \baseMedianszeval, which is more than for CRAN, which was dominated by symbol evaluation.The maximum size is \baseMaxszeval.
-Small event evals (less than 50 events) amount for \baseSmalleventspct of the events in \eval calls.
+The prevalence of language objects suggests that \emph{base} mainly performs metaprogramming with \eval.
+The median argument size is \baseMedianszeval, which is more than for CRAN, which was dominated by symbol evaluation.
+The maximum size is \baseMaxszeval.
+Small event evals (less than 50 events) amount for \baseSmalleventspct of the events inside \eval calls.
 
@@
 In total, \kaggleAllcalls \eval were recorded.
-Most of them are call to the normal \eval, whereas a few, \kaggleLocals, are calls to \c{local}.
-They are due to the code loading libraries.
+Most of them are calls to the normal \eval, whereas a few, \kaggleLocals, are calls to \c{local}.
+This happens in the machinery that loads library code.
 Out of \kaggleStaticeval static \eval sites, we exercise \kaggleTriggeredeval of them.
@@
 The largest argument size is \kaggleMaxszeval.
-The distribution of events is similar to the one for CRAN: \kaggleSmalleventspct of \eval calls amount for less than 50 events.
-
-\paragraph{Discussion} \emph{Base} is more akin to use \eval for metaprogramming, \emph{kaggle} to evaluate parsed strings, and CRAN is dominated by symbol evaluation.
+The distribution of events is similar to the one for CRAN or \emph{base}: \kaggleSmalleventspct of \eval calls amount for less than 50 events.
+
+\paragraph{Discussion} \emph{Base} is more akin to using \eval for metaprogramming, \emph{kaggle} to evaluating parsed strings, and CRAN is dominated by symbol evaluation.
 In terms of number of calls, a few calls dominate, with a call from \emph{ggplot} being prominent in the CRAN corpus.
@@
   an \eval (5\%) is correct. That feels surprising to me. 
+ \item Add the percentage of operations for base and kaggle
 \end{itemize}

### 2021-03-10 analysis/normalized.Rmd
Fix problems with notebook generation per dataset
Cache mechanism for normalized expression

@@
   normalized_path: normalized-expressions.csv
-  corpus_path: ../corpus.fst
+  corpus_path: corpus.txt
   static_path: evals-static.csv
-  force_rebuild: TRUE
+  force_rebuild: FALSE
 ---
@@
 create_tags(path(TAGS_DIR, paste0(dataset_name, "_normalized_expr.tex")), prefix = dataset_name, default = TRUE)
+# Also use ggsave_prefix instead of ggsave if you need to save a graph
 
@@
 normalized_path <- paste0(params$base_dir, params$normalized_path)
+
+lock_path <- paste0(params$base_dir, ".normalized-lock.fst")
+min_path <- paste0(params$base_dir, "Ecano.fst")
 ```
@@
 ```{r read, message=TRUE, include=F}
-read_fst(corpus_path) %>% as_tibble() -> C_raw
-#read_csv(static_path) %>% semi_join(C_raw, by = "package") -> P
+tibble(package = read_lines(corpus_path)) -> C_raw
+read_csv(static_path) %>% semi_join(C_raw, by = "package") -> P
 
+rebuild <- TRUE
+
+E_cano <- NULL
+
+if (file.exists(lock_path)) {
+  saved <- read.fst(lock_path)[[1,1]]
+  if (saved == file.size(calls_path)) {
+    read_fst(min_path) %>% as_tibble() -> E_cano
+    rebuild <- FALSE
+  }
+}
+
+if(params$force_rebuild || rebuild) {
   read_fst(calls_path) %>% as_tibble() -> E_raw
@@
 
+  E_cano <- E %>% left_join(N, by = "hash") # Common column is the hash of the full resolved expression
+
+  E_cano %>% write_fst(min_path)
+  sz <- file.size(calls_path) %>% as_tibble()
+  write.fst(sz, lock_path)
+}
+
 C_raw -> C
@@
 ```{r}
-nb_call_sites <- E %>%
+nb_call_sites <- E_cano %>%
   pull(src_ref) %>%
@@
 There are `r nb_call_sites` call sites in the dataset.
-
-# Full dataset
-
-We join the normalized expressions and their statistics with the calls.
-
-```{r}
-E_cano <- E %>% left_join(N, by = "hash") # Common column is the hash of the full resolved expression
-```
 

### 2021-03-10 paper/main.tex
Start working on normalized expressions

@@
 \item Check that the percentage of operations performed within
-  an \eval (5\%) is correct. That feels surprising to me. 
- \item Add the percentage of operations for base and kaggle
+  an \eval (5\%) is correct. That feels surprising to me. We should actually remove this number!
 \end{itemize}
@@
 in this section we try to elucidate \emph{what} it does.
+
+\paragraph{Normalized expressions} During tracing, we gather all source expressions and all resolved expression arguments passed to \eval.
+For instance, a non-resolved expression and some of its resolved expressions can be:
+
+\begin{lstlisting}
+parse(text = paste0(\"options(\", i, \"=\", opts[i], \")\"))
+
+options(datatable.alloccol = 1024L)
+options(datatable.allow.cartesian = FALSE)
+options(datatable.print.keys = FALSE)
+\end{lstlisting}
+
+We classify the resolved expressions to take into account the structure of the expression passed to \eval.
+However, we do not want to discriminate expressions based on specific values.
+Hence, the expression goes through several simplification steps: constant folding of arithmetic and string expressions that use commonly used functions (addition, string concatenation), discarding of specific value to a symbolic value holder, variable absorption.
+We keep track of variables, as they are evaluated in the environment specified to \eval, contrary to values that are left unchanged.
+After the simplification steps, we obtain 15 different \emph{minimized} expressions.
+
+\begin{table}[h]
+	\begin{tabular}{|c|c|c|c|c|c|}
+		\hline
+		Minimized & \#sites & \%sites & \#packages & \#operations & \%envir \\
+		\hline
+		\texttt{X} &  &  &  &  &  \\
+		\hline
+		\texttt{F(F())} &  &  &  &  &  \\
+		\hline
+		\texttt{V} &  &  &  &  &  \\
+		\hline
+		\texttt{F(X)}&  &  &  &  &  \\
+		\hline
+		\texttt{model.frame} &  &  &  &  &  \\
+		\hline
+		\texttt{\$}&  &  &  &  &  \\
+		\hline
+		\texttt{F()}&  &  &  &  &  \\
+		\hline
+		\texttt{<-} &  &  &  &  &  \\
+		\hline
+		\texttt{\{BLOCK\}}&  &  &  &  &  \\
+		\hline
+		\texttt{FUN} &  &  &  &  &  \\
+		\hline
+		&  &  &  &  &  \\
+		\hline
+		&  &  &  &  &  \\
+		\hline
+		&  &  &  &  &  \\
+		\hline
+		&  &  &  &  &  \\
+		\hline
+		&  &  &  &  &  \\
+		\hline
+	\end{tabular}
+\end{table}
 

### 2021-03-11 analysis/normalized.Rmd
Numbers for 1st 12 minimzed expressions

@@
   group_by(minimized) %>%
-  summarize(n_call_sites = n_distinct(src_ref), proportion = ratio(n_call_sites, nb_call_sites), n_packages = n_distinct(ev_package), n_operations = round(mean(n_op, na.rm=TRUE), 2), percent_envir = round(100 * sum(!is.na(envir)) / n(), 2))
+  summarize(n_call_sites = n_distinct(src_ref), proportion = ratio(n_call_sites, nb_call_sites), n_packages = n_distinct(ev_package), n_operations = round(mean(n_op, na.rm=TRUE), 2), percent_envir = ratio(sum(!is.na(envir)), n())) %>%
+  arrange(desc(n_call_sites))
 
 widespread_minimized %>%
-  arrange(desc(n_call_sites)) %>%
   datatable()
+```
+
+```{r}
+nrows <- 12
+r_vec("MinimizedCallSites", nrows, pull(widespread_minimized, n_call_sites))
+r_vec("MinimizedPropSites", nrows, pull(widespread_minimized, proportion))
+r_vec("MinimizedPackage", nrows, pull(widespread_minimized, n_packages))
+r_vec("MinimizedOperations", nrows, pull(widespread_minimized, n_operations))
+r_vec("MinimizedPercentEnvir", nrows, pull(widespread_minimized, percent_envir))
 ```

### 2021-03-11 analysis/inc/latextags.R

@@ -409,3 +409,11 @@ ggsave_prefix <- function(name, tags=get_default_tags()) {
   #TODO: add PLOT_DIR as a parameter to the tags when doing create_tags
   ggsave(path(PLOT_DIR, paste0(prefix, "_", name)))
 }
+
+r_vec <- function(name, nrows, v) {
+  stopifnot(length(v) >= nrows)
+  startLetter <- utf8ToInt("A") - 1 # LateX does not like numbers so we suffix with letters of the alphabet
+  for(i in 1:nrows) {
+    r(paste0(name, intToUtf8(startLetter + i)), v[[i]])
+  }
+}

### 2021-03-12 paper/main.tex
Numbers for 1st 12 minimzed expressions
Remove the 5% number
Add total number of sites

@@
 To measure the work performed in \evals, we count the instructions
-executed by interpreter during evaluation of the argument. In our
-corpus, \EvalEventAllPerc of all instructions (\EventsRnd) are
-performed within \eval. Most invocations perform relatively little
+executed by interpreter during evaluation of the argument.  Most invocations perform relatively little
 work with \Smalleventspct of \evals executing 50 or fewer
@@
 
-\subsection{Things to do}
-
-\begin{itemize}
-\item Figure 7: check that element counts are correct
-\item Check that the percentage of operations performed within
-  an \eval (5\%) is correct. That feels surprising to me. We should actually remove this number!
-\end{itemize}
-
 \section{A Taxonomy of Eval}
@@
 We keep track of variables, as they are evaluated in the environment specified to \eval, contrary to values that are left unchanged.
-After the simplification steps, we obtain 15 different \emph{minimized} expressions.
+We also track whether assignment operators are used in the expression or not.
+After the simplification steps, we obtain 15 different \emph{minimized} expressions, of which we show the ten first most frequent in numbers of sites in Table~\ref{tab:minimizedexpressions}.
 
@@
 		\hline
-		\texttt{X} &  &  &  &  &  \\
-		\hline
-		\texttt{F(F())} &  &  &  &  &  \\
-		\hline
-		\texttt{V} &  &  &  &  &  \\
-		\hline
-		\texttt{F(X)}&  &  &  &  &  \\
-		\hline
-		\texttt{model.frame} &  &  &  &  &  \\
-		\hline
-		\texttt{\$}&  &  &  &  &  \\
-		\hline
-		\texttt{F()}&  &  &  &  &  \\
-		\hline
-		\texttt{<-} &  &  &  &  &  \\
-		\hline
-		\texttt{\{BLOCK\}}&  &  &  &  &  \\
-		\hline
-		\texttt{FUN} &  &  &  &  &  \\
-		\hline
-		&  &  &  &  &  \\
-		\hline
-		&  &  &  &  &  \\
-		\hline
-		&  &  &  &  &  \\
-		\hline
-		&  &  &  &  &  \\
-		\hline
-		&  &  &  &  &  \\
+		\texttt{X} & \packageMinimizedcallsitesa  & \packageMinimizedpropsitesa & \packageMinimizedpackagea & \packageMinimizedoperationsaRnd & \packageMinimizedpercentenvira \\
+		\hline
+		\texttt{F(F())} & \packageMinimizedcallsitesb  & \packageMinimizedpropsitesb & \packageMinimizedpackageb  & \packageMinimizedoperationsbRnd & \packageMinimizedpercentenvirb \\
+		\hline
+		\texttt{V} & \packageMinimizedcallsitesc & \packageMinimizedpropsitesc & \packageMinimizedpackagec & \packageMinimizedoperationscRnd & \packageMinimizedpercentenvirc \\
+		\hline
+		\texttt{F(X)}& \packageMinimizedcallsitesd & \packageMinimizedpropsitesd & \packageMinimizedpackaged & \packageMinimizedoperationsdRnd & \packageMinimizedpercentenvird \\
+		\hline
+		\texttt{model.frame} & \packageMinimizedcallsitese & \packageMinimizedpropsitese & \packageMinimizedpackagee & \packageMinimizedoperationseRnd & \packageMinimizedpercentenvire \\
+		\hline
+		\texttt{\$}& \packageMinimizedcallsitesf & \packageMinimizedpropsitesf & \packageMinimizedpackagef & \packageMinimizedoperationsfRnd & \packageMinimizedpercentenvirf \\
+		\hline
+		\texttt{F()}& \packageMinimizedcallsitesg & \packageMinimizedpropsitesg & \packageMinimizedpackageg & \packageMinimizedoperationsgRnd & \packageMinimizedpercentenvirg \\
+		\hline
+		\texttt{<-} & \packageMinimizedcallsitesh & \packageMinimizedpropsitesh & \packageMinimizedpackageh & \packageMinimizedoperationshRnd & \packageMinimizedpercentenvirh \\
+		\hline
+		\texttt{FUN} & \packageMinimizedcallsitesi  & \packageMinimizedpropsitesi & \packageMinimizedpackagei & \packageMinimizedoperationsiRnd & \packageMinimizedpercentenviri \\
+		\hline
+		\texttt{\{BLOCK\}} & \packageMinimizedcallsitesj & \packageMinimizedpropsitesj & \packageMinimizedpackagej & \packageMinimizedoperationsjRnd & \packageMinimizedpercentenvirj \\
 		\hline
 	\end{tabular}
+	\caption{Minimized expressions: ten most frequent in terms of sites.
+\#sites is the number of sites; \%sites the percentage of sites across the total number of sites (\packageNbCallSites sites); \#packages is the number of packages with at least one of the minimized expression and measures how widespread a minimized expression is.
+\#operations is the average number of steps performed by the R interpreter when given that minimized expression, and \%envir is the percentage of call sites for which the \c{envir} argument is explicitly passed.} \label{tab:minimizedexpressions}
 \end{table}
+
+The most common minimized expression is a variable.
+It happens in \packageMinimizedpropsitesa of the sites and is widespread across packages as it is found in \packageMinimizedpackagea packages across our CRAN corpus.
+The \c{envir} argument is explicitly passed in \packageMinimizedpercentenvira\% of the call sites, which suggests that \eval is used here to get the value of the variable in a non-local environment.
+
+Note that a site can exhibit several minimized expressions. The most common combinations are:
 
@@
 
-\section{Case Studies}
-
-\NOTE{5(?) examples from real code?}
-
-\subsection{LEGACY TEXT}
+\section{\eval patterns}
+
+\subsection{Evaluating variables in a specific environment}
+
+\subsection{\c{match.call}}
+
+\subsection{Meta-programming}
+
+\section{LEGACY TEXT}
 

### 2021-03-15 analysis/normalized.Rmd
Add total number of sites
Add percentage of sites with only one minimized expression

@@
   n_distinct()
+
+r("Nb call sites", nb_call_sites)
 ```
@@
 ```{r}
-poly %>% filter(n_exprs > 1) %>%
+poly %>%
   count(common_minimized,  sort = TRUE) %>%
@@
 ```
+
+How many sites have only one minimized expression?
+
+```{r}
+n_one_mini <- poly %>% filter(n_expers == 1) %>% nrow()
+
+r("Nb One Minimized", n_one_mini)
+r("Nb One Minimized Percent", ratio(n_one_mini, nb_call_sites))
+```

### 2021-03-15 paper/main.tex
Minimized expressions: % only one per site

@@
 
-The most common minimized expression is a variable.
-It happens in \packageMinimizedpropsitesa of the sites and is widespread across packages as it is found in \packageMinimizedpackagea packages across our CRAN corpus.
+The most common minimized expression is a variable.\footnote{An arithmetic with some variables and constant will also be normalized to \c{X}, i.e.
+a variable.} It happens in \packageMinimizedpropsitesa of the sites and is widespread across packages as it is found in \packageMinimizedpackagea packages across our CRAN corpus.
 The \c{envir} argument is explicitly passed in \packageMinimizedpercentenvira\% of the call sites, which suggests that \eval is used here to get the value of the variable in a non-local environment.
 
-Note that a site can exhibit several minimized expressions. The most common combinations are:
+Compositions of functions, without variables, is the second most frequent minimized expression.
+
+When a function is evaluated on some variable, the percentage of non default environment is high.
+
+\c{FUN} represents a function definition
+
+\c{{BLOCK}} is a sequence of statements. It is usually large, with in average \packageMinimizedoperationsjRnd operations in it.
+
+The minimized expression classification suggests that the presence of variables in the expression is highly correlated to the specification of a specific environment, and that \eval is used here for its power in precising the evaluation environment.
+
+Note that a site can exhibit several minimized expressions. However, \packageNbOneMinimizedPercent of the sites have only one minimized expression. The most common combinations are:
+
+\NOTE{Maybe we should ditch the table of the minimized expressions and rather use the ones of combinations.
+It has the advantage of showing what is the most frequent, per call site.
+But it's a long tail, and sites that have several minimized expressions are not that numerous, only 13\%}
 
@@
 \subsection{Meta-programming}
+
+\subsection{Logging}
+
+\subsection{Code transform}
+
+\subsection{Plotting}
+
+\section{\eval expressivity and macros}
+
+% The Thomas Lumley's article, Programmer's Niche: Macros in R, (https://www.r-project.org/doc/Rnews/Rnews_2001-3.pdf) describes how to write a `defmacro` function that would like as a macro creator, using `substitute` and `eval`.
+There are no local macro variables in this implementation though.
+
+\section{The dangers of \eval}
 

### 2021-03-16 analysis/normalized.Rmd
Fix
trying to correlate

@@
 library(scales)
+library(corrr)
 
@@
     match.call = expr_match_call,
+    arg_size = expr_resolved_length, #   Size in characters of argument 
+    ast_nodes = expr_resolved_nodes,
     n_op = interp_eval,
-    envir = envir_expression
+    envir = envir_expression,
+    caller_srcref,
+    caller_expression
   ) -> E
@@
   for(exp in l) {
-    s[[exp]] = "T"
+    s[[exp]] <- "T"
   }
@@
   return(paste0(s, collapse = ""))
+}
+
+index_v <- match("V", minimized_exprs)
+
+has_value <- function(l) {
+  substr(l, index_v, index_v) == "T"
+}
+
+count_elements <- function(l) {
+  sum(map_int(strsplit(l, split="")[[1]], function(ch) {if(ch == "T") {1L} else {0L}}))
 }
@@
 ```{r}
-n_one_mini <- poly %>% filter(n_expers == 1) %>% nrow()
+n_one_mini <- poly %>% filter(n_exprs == 1) %>% nrow()
 
@@
 ```
+
+# Values
+
+Many call sites only evaluate a value.
+
+Is it a simple or a complex value?
+
+```{r}
+val_exprs <- E_cano %>% filter(minimized == "V") %>%
+  mutate(caller_srcref = if_else(is.na(caller_srcref), file, caller_srcref)) %>%
+  group_by(src_ref) %>%
+  summarise(avg_ast_nodes = mean(ast_nodes, na.rm = TRUE), avg_arg_size = mean(arg_size, na.rm = TRUE), nb_runs = n_distinct(file), nb_caller_srcref = n_distinct(caller_srcref), nb_calls = sum(duplicates), nb_caller_exprs = n_distinct(caller_expression)) %>%
+  arrange(desc(avg_ast_nodes))
+
+n_val_exprs <- val_exprs %>% nrow()
+n_one_node <- val_exprs %>% filter(avg_ast_nodes == 1) %>% nrow()
+
+r("val one node", n_one_node)
+r("val one node percent", ratio(n_one_node, n_val_exprs))
+```
+
+Are the sites resolving to just a value well covered?
+
+```{r}
+med_runs <- median(val_exprs$nb_runs, na.rm = TRUE)
+```
+
+Is there a correlation between the number of caller srcref and the presence of other minimized expressions for one call site?
+
+There are many `NA` caller srcref, so we count them as 1 currently...
+
+```{r}
+with_value <- poly %>% filter(has_value(common_minimized))
+
+val_exprs_comb <- val_exprs %>% left_join(with_value, by="src_ref")  %>%
+  mutate(calls_per_run = nb_calls / nb_runs) %>%
+  mutate(calls_per_site = nb_calls / nb_caller_srcref) %>%
+  mutate(calls_per_site_per_run = nb_calls / nb_runs / nb_caller_srcref)
+# normalization to be able to see both on the same graph
+
+val_exprs_comb %>% ggplot(aes(x = nb_caller_srcref, y = n_exprs)) + 
+  geom_jitter(alpha = 0.3) +
+  scale_x_log10()
+
+val_exprs_comb %>% ggplot(aes(x = nb_runs, y = n_exprs)) + 
+  geom_jitter(alpha = 0.3) +
+  scale_x_log10()
+
+val_exprs_comb %>% ggplot(aes(x = calls_per_run, y = n_exprs)) + 
+  geom_jitter(alpha = 0.3) +
+  scale_x_log10()
+
+val_exprs_comb %>% ggplot(aes(x = calls_per_site, y = n_exprs)) + 
+  geom_jitter(alpha = 0.3) +
+  scale_x_log10()
+
+val_exprs_comb %>% ggplot(aes(x = calls_per_site_per_run, y = n_exprs)) + 
+  geom_jitter(alpha=0.3) +
+  scale_x_log10()
+
+val_exprs_comb %>% ggplot(aes(x = nb_caller_exprs, y = n_exprs)) + 
+  geom_jitter(alpha=0.3) +
+  scale_x_log10()
+
+df <- val_exprs_comb %>% select(where(is.numeric))
+
+df %>% correlate(method = "spearman") %>% network_plot()
+df %>% correlate(method = "spearman") %>% rplot() + theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=1))
+```

### 2021-03-16 analysis/normalized.Rmd
Percentage of sites with actual values

@@
 
-Many call sites only evaluate a value.
+Many call sites only evaluate a value. It is either a language expression or symbol, or an actual value.
+
+```{r}
+value_minimized <- widespread_minimized %>% filter(minimized == "V")
+```
+
+These are the number of call sites that are actual values.
+```{r}
+main_types <- E_cano %>% group_by(src_ref) %>% summarise(nb_types = n_distinct(type), type = first(type), nb_runs = n_distinct(file))
+
+only_actual_values <- main_types %>% filter(!type %in% c("symbol", "language", "expression", "promise"), nb_types == 1)
+nb_actual_values_sites <- only_actual_values %>% pull(src_ref) %>% n_distinct()
+
+r("nb call sites unique actual value", nb_actual_values_sites)
+r("call sites unique actual value percent", ratio(nb_actual_values_sites, value_minimized$n_call_sites))
+r("median run sites unique actual value", median(only_actual_values$nb_runs, na.rm = TRUE))
+```
 
@@
 
-```{r}
-with_value <- poly %>% filter(has_value(common_minimized))
-
-val_exprs_comb <- val_exprs %>% left_join(with_value, by="src_ref")  %>%
+```{r eval = FALSE}
+val_exprs_comb <- val_exprs %>% left_join(main_types, by="src_ref")  %>%
   mutate(calls_per_run = nb_calls / nb_runs) %>%
@@
 
-val_exprs_comb %>% ggplot(aes(x = nb_caller_srcref, y = n_exprs)) + 
+val_exprs_comb %>% ggplot(aes(x = nb_caller_srcref, y = nb_types)) + 
   geom_jitter(alpha = 0.3) +
@@
 
-val_exprs_comb %>% ggplot(aes(x = nb_runs, y = n_exprs)) + 
+val_exprs_comb %>% ggplot(aes(x = nb_runs, y = nb_types)) + 
   geom_jitter(alpha = 0.3) +
@@
 
-val_exprs_comb %>% ggplot(aes(x = calls_per_run, y = n_exprs)) + 
+val_exprs_comb %>% ggplot(aes(x = calls_per_run, y = nb_types)) + 
   geom_jitter(alpha = 0.3) +
@@
 
-val_exprs_comb %>% ggplot(aes(x = calls_per_site, y = n_exprs)) + 
+val_exprs_comb %>% ggplot(aes(x = calls_per_site, y = nb_types)) + 
   geom_jitter(alpha = 0.3) +
@@
 
-val_exprs_comb %>% ggplot(aes(x = calls_per_site_per_run, y = n_exprs)) + 
+val_exprs_comb %>% ggplot(aes(x = calls_per_site_per_run, y = nb_types)) + 
   geom_jitter(alpha=0.3) +
@@
 
-val_exprs_comb %>% ggplot(aes(x = nb_caller_exprs, y = n_exprs)) + 
+val_exprs_comb %>% ggplot(aes(x = nb_caller_exprs, y = nb_types)) + 
   geom_jitter(alpha=0.3) +
@@
 df %>% correlate(method = "spearman") %>% rplot() + theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=1))
+
 ```

### 2021-03-16 analysis/environments.Rmd
Progressing on explaining values

@@
 params:
-  base_dir: Data
-  calls_packages_path: /package/summarized.fst
+  base_dir: Data/
+  calls_packages_path: package/summarized.fst
 ---

### 2021-03-16 paper/main.tex
Normalized expressions: progress
Progressing on explaining values

@@
 
-\paragraph{Normalized expressions} During tracing, we gather all source expressions and all resolved expression arguments passed to \eval.
+\subsection{Normalized expressions} During tracing, we gather all source expressions and all resolved expression arguments passed to \eval.
 For instance, a non-resolved expression and some of its resolved expressions can be:
@@
 	\caption{Minimized expressions: ten most frequent in terms of sites.
-\#sites is the number of sites; \%sites the percentage of sites across the total number of sites (\packageNbCallSites sites); \#packages is the number of packages with at least one of the minimized expression and measures how widespread a minimized expression is.
+\#sites is the number of sites; \%sites the percentage of sites across the total number of sites (\packageNbCallSites sites) with at least this minimized expression; \#packages is the number of packages with at least one of the minimized expression and measures how widespread a minimized expression is.
 \#operations is the average number of steps performed by the R interpreter when given that minimized expression, and \%envir is the percentage of call sites for which the \c{envir} argument is explicitly passed.} \label{tab:minimizedexpressions}
@@
 
+Note that a site can exhibit several minimized expressions. However, \packageNbOneMinimizedPercent of the sites have only one minimized expression.
+
+\paragraph{Variables.}
 The most common minimized expression is a variable.\footnote{An arithmetic with some variables and constant will also be normalized to \c{X}, i.e.
@@
 
-Compositions of functions, without variables, is the second most frequent minimized expression.
-
-When a function is evaluated on some variable, the percentage of non default environment is high.
-
-\c{FUN} represents a function definition
-
-\c{{BLOCK}} is a sequence of statements. It is usually large, with in average \packageMinimizedoperationsjRnd operations in it.
+\paragraph{Composition of functions. } Compositions of functions, without variables, is the second most frequent minimized expression.
+
+\paragraph{Value.}
+Many sites evaluates to \c{V}.
+\packageValOneNodePercent of those sites are only passed an expression with one node, \ie one value, not a more complex expression that is later constant folded.
+\c{V} corresponds to two main categories of expressions passed to \eval:
+
+\begin{itemize}
+	\item \eval is passed an actual value, a double, an integer, an environment, a closure and so on
+	\item \eval is passed a language expression that trivially evaluates to a value, for instance, \c{quote(1)}, \quote{c("a", "b", "c")}, and does not include variables, so does not depend on a specific environment.
+\end{itemize}
+
+For the first category, \eval is not necessary.
+In \packageNbCallSitesUniqueActualValue sites, \ie \packageCallSitesUniqueActualValuePercent\% sites, it is only passed such actual values.
+The median number of runs for those sites is \packageMedianRunSitesUniqueActualValue, suggesting it happens because there are rarely triggered in our dataset and all paths are not taken.
+
+The percentage of non default environment passed to \eval is \packageMinimizedpercentenvirc. Some sites are passed both a value or a symbol, which explains that a specific environment is precised.
+
+%TODO: find out the percentage of non-default envir for sites which have both a value and a variable
+%TODO: find the same percentage for sites which have at least one variable passed to them?
+
+When a function is evaluated on some variable (\c{F(X)}), the percentage of non default environment is high, \packageMinimizedpercentenvird\%.
+
+\c{FUN} represents a function definition. It entails very few steps of the interpreter, as the body of the new function is just stored for later execution.
+
+\c{\{BLOCK\}} is a sequence of statements. It is usually large, with in average \packageMinimizedoperationsjRnd operations in it.
 
@@
 
-Note that a site can exhibit several minimized expressions. However, \packageNbOneMinimizedPercent of the sites have only one minimized expression. The most common combinations are:
+The most common combinations for sites with several minimized expressions are:
 
@@
 It has the advantage of showing what is the most frequent, per call site.
-But it's a long tail, and sites that have several minimized expressions are not that numerous, only 13\%}
+But it's a long tail, and sites that have several minimized expressions are not that numerous, only 13\%.
+But if include combinations, percentages change.
+For instance, 3314 call sites have only X, i.e.
+22.51, instead of 4353 and 29.6\%.
+}
+
+\NOTE{Two small paragraphs about normalized expressions in Base and Kaggle.}
 

### 2021-03-18 analysis/normalized.Rmd
New stats about values passed to eval

@@
 index_v <- match("V", minimized_exprs)
+index_x <- match("X", minimized_exprs)
 
@@
   substr(l, index_v, index_v) == "T"
+}
+
+is_variable <- function(l) {
+  substr(l, index_x, index_x) == "T"
 }
@@
 r("median run sites unique actual value", median(only_actual_values$nb_runs, na.rm = TRUE))
+r("average run sites unique actual value", mean(only_actual_values$nb_runs, na.rm = TRUE))
+r("sd run sites unique actual value", sd(only_actual_values$nb_runs, na.rm = TRUE))
 ```
@@
 med_runs <- median(val_exprs$nb_runs, na.rm = TRUE)
+avg_runs <- mean(val_exprs$nb_runs, na.rm = TRUE)
+sd_runs <- sd(val_exprs$nb_runs, na.rm = TRUE)
+
+r("one val median runs", med_runs)
+r("one val average runs", avg_runs)
+r("one val sd runs", sd_runs)
 ```
@@
 ```
+
+Manual inspection of the source code shows that there are actually some call sites that take actual values.
+If a site is exercised by many caller sites, or by many different runs, and has only values, then we reckon it is probably a _useless_ `eval`...
+
+```{r}
+val_only <- val_exprs %>% right_join(only_actual_values, by = c("src_ref", "nb_runs"))
+
+max_val_runs <- val_only %>% filter(nb_caller_exprs == max(nb_caller_exprs)) #it happens to also be the one with the most caller expressions
+
+val_only %>% filter(nb_caller_exprs > 5) # those ones are likely to be useless eval
+```
+
+We now want to characterize better the sites that are passed a non-default environment. Are they also the sites that can get a value or something with a variable inside?
+
+We compute the sites which are at least one given a `V` minimized expression. Then we look at whether they are passed teh default envir or not, and if there are variables in
+
+```{r}
+minimized_sites <- E_cano %>% semi_join(val_exprs, by="src_ref") %>% group_by(src_ref) %>% 
+  summarize(has_envir = any(!is.na(envir)), has_var = any(has_var, na.rm = TRUE))
+
+non_default_envir <- minimized_sites %>% filter(has_envir)
+
+non_default_with_var <- minimized_sites %>% filter(has_envir, has_var)
+
+```

### 2021-03-18 analysis/normalized.Rmd
Carrying on on values passed to eval

@@
 
-We compute the sites which are at least one given a `V` minimized expression. Then we look at whether they are passed teh default envir or not, and if there are variables in
+We compute the sites which are at least one given a `V` minimized expression. Then we look at whether they are passed the default envir or not, and if there are variables in
 
@@
 minimized_sites <- E_cano %>% semi_join(val_exprs, by="src_ref") %>% group_by(src_ref) %>% 
-  summarize(has_envir = any(!is.na(envir)), has_var = any(has_var, na.rm = TRUE))
+  summarize(has_envir = all(!is.na(envir)), has_var = any(has_var, na.rm = TRUE))
 
@@
 
+r("non default envir value percent", ratio(nrow(non_default_envir), nrow(minimized_sites)))
+r("non default envir with var percent", ratio(nrow(non_default_with_var), nrow(non_default_envir)))
 ```

### 2021-03-18 paper/main.tex
Carrying on on values passed to eval

@@
 \paragraph{Variables.}
-The most common minimized expression is a variable.\footnote{An arithmetic with some variables and constant will also be normalized to \c{X}, i.e.
+The most common minimized expression is a variable.\footnote{An arithmetic expression with some variables and constant will also be normalized to \c{X}, i.e.
 a variable.} It happens in \packageMinimizedpropsitesa of the sites and is widespread across packages as it is found in \packageMinimizedpackagea packages across our CRAN corpus.
@@
 
+Similarly, when a function is evaluated on some variable (\c{F(X)}), the percentage of non default environment is high, \packageMinimizedpercentenvird\%.
+
 \paragraph{Composition of functions. } Compositions of functions, without variables, is the second most frequent minimized expression.
 
-\paragraph{Value.}
+\paragraph{Values.}
 Many sites evaluates to \c{V}.
@@
 \begin{itemize}
-	\item \eval is passed an actual value, a double, an integer, an environment, a closure and so on
-	\item \eval is passed a language expression that trivially evaluates to a value, for instance, \c{quote(1)}, \quote{c("a", "b", "c")}, and does not include variables, so does not depend on a specific environment.
+	\item \eval is passed an actual value, a double, an integer, an environment, a closure and other simple types,
+	\item \eval is passed a language expression that trivially evaluates to a value, for instance, \c{quote(1)}, \c{quote(1 + 1)}, \quote{c("a", "b", "c")}, and does not include variables, so does not depend on a specific environment.
 \end{itemize}
@@
 For the first category, \eval is not necessary.
-In \packageNbCallSitesUniqueActualValue sites, \ie \packageCallSitesUniqueActualValuePercent\% sites, it is only passed such actual values.
-The median number of runs for those sites is \packageMedianRunSitesUniqueActualValue, suggesting it happens because there are rarely triggered in our dataset and all paths are not taken.
-
-The percentage of non default environment passed to \eval is \packageMinimizedpercentenvirc. Some sites are passed both a value or a symbol, which explains that a specific environment is precised.
-
-%TODO: find out the percentage of non-default envir for sites which have both a value and a variable
-%TODO: find the same percentage for sites which have at least one variable passed to them?
-
-When a function is evaluated on some variable (\c{F(X)}), the percentage of non default environment is high, \packageMinimizedpercentenvird\%.
-
-\c{FUN} represents a function definition. It entails very few steps of the interpreter, as the body of the new function is just stored for later execution.
-
-\c{\{BLOCK\}} is a sequence of statements. It is usually large, with in average \packageMinimizedoperationsjRnd operations in it.
+In \packageNbCallSitesUniqueActualValue sites exhibiting \c{V}, \ie \packageCallSitesUniqueActualValuePercent\% sites, it is only passed such actual values.
+The median number of runs for those sites is \packageMedianRunSitesUniqueActualValue (average \packageAverageRunSitesUniqueActualValue, standard deviation \packageSdRunSitesUniqueActualValue), suggesting it happens because there are rarely triggered in our dataset and all paths are not taken.
+However, manual inspection of the source code of some examples shows that some sites actually pass values.
+For instance, in package \emph{ctmle}, in function \c{evaluate\_candidate\_glmnet}:
+
+\begin{lstlisting}
+f1 <- eval(paste("A ~ ", paste(paste(names(X[,-1]), sep=''), collapse=" + ")))
+f2 <- as.formula(f1)
+\end{lstlisting}
+
+\c{paste} concatenates strings and \eval will just return that string. \c{as.formula} can take a string and convert it to a formula; \eval is not needed here.
+
+The percentage of non default environment passed to \eval is \packageMinimizedpercentenvirc when the minimized expression is a value.
+Some sites which are passed a value, are also passed different minimized expressions.
+For those ones, \packageNonDefaultEnvirValuePercent use a non-default environment, and among the one that use a non-default environment, \packageNonDefaultEnvirWithVarPercent have at least one minimized expression with a variable inside.
+
+\paragraph{Function definition.} \c{FUN} represents a function definition. It entails very few steps of the interpreter, as the body of the new function is just stored for later execution.
+
+\paragraph{Sequence of statements.} \c{\{BLOCK\}} is a sequence of statements. It is usually large, with in average \packageMinimizedoperationsjRnd operations in it.
 

### 2021-03-18 analysis/normalized.Rmd
Minimized expressions and assignments.

@@
 nb_assigns_call_sites <- assigns %>% pull(src_ref) %>% n_distinct()
+
+r("nb assign sites", nb_assigns_call_sites)
+r("assign sites percent", ratio(nb_assigns_call_sites, nb_call_sites))
 ```
@@
 nb_a_env_call_sites <- assign_env %>% pull(src_ref) %>% n_distinct()
+
+r("non default envir assign sites", nb_a_env_call_sites)
+r("non default envir assign sites percent", ratio(nb_a_env_call_sites, nb_assigns_call_sites))
 ```

### 2021-03-18 paper/main.tex
Minimized expressions and assignments.

@@
 		\hline
+		\texttt{F(F(<-))} & \packageMinimizedcallsitesk & \packageMinimizedpropsitesk & \packageMinimizedpackagek & \packageMinimizedoperationskRnd & \packageMinimizedpercentenvirk \\
+		\hline
 	\end{tabular}
@@
 
-\paragraph{Composition of functions. } Compositions of functions, without variables, is the second most frequent minimized expression.
-
 \paragraph{Values.}
@@
 
+\paragraph{\c{model.frame}} This is a function used to create statistical models. We explain it in more details in the section about pattern (see Section~REF).
+
 \paragraph{Function definition.} \c{FUN} represents a function definition. It entails very few steps of the interpreter, as the body of the new function is just stored for later execution.
 
-\paragraph{Sequence of statements.} \c{\{BLOCK\}} is a sequence of statements. It is usually large, with in average \packageMinimizedoperationsjRnd operations in it.
+\paragraph{Assignments.} This amounts to \packageAssignSitesPercent of the sites and corresponds to minimized expressions \c{<-}, where the assignment is the top call, and \c{F(F(<-))}, where the assignment happens inside, and others that we were not enough present to be shown in table~\ref{tab:minimizedexpressions}.
+Assignments are performed with \c{<-}, \c{<<-} or \c{assign}.
+Most often, assignment happens in top call position.
+A non-default environment is specified only in \packageNonDefaultEnvirAssignSitesPercent of the sites with assignments.
+Indeed, most of those assignments occur in the default environment, \ie in the caller of \eval.
+This hints that side effects are not widespread in \eval.\footnote{Side effects can also happen in a function call in the \eval expression, and that function call will hide assignment operators from us.
+We more precisely instrument side-effects to environment in Section~REF.}
+
+\paragraph{Accessing a slot.}
+
+\paragraph{Large expressions }  Compositions of functions, without variables (\c{F(F(X))} minimized expression) is the second most frequent minimized expression.
+\c{\{BLOCK\}} is a sequence of statements.
+It is usually large, with in average \packageMinimizedoperationsjRnd operations in it.
 

### 2021-03-18 analysis/normalized.Rmd
Correct the percentage of non-default environments.

@@
   group_by(minimized) %>%
-  summarize(n_call_sites = n_distinct(src_ref), proportion = ratio(n_call_sites, nb_call_sites), n_packages = n_distinct(ev_package), n_operations = round(mean(n_op, na.rm=TRUE), 2), percent_envir = ratio(sum(!is.na(envir)), n())) %>%
+  summarize(n_call_sites = n_distinct(src_ref), proportion = ratio(n_call_sites, nb_call_sites), n_packages = n_distinct(ev_package), n_operations = round(mean(n_op, na.rm=TRUE), 2), percent_envir = ratio(n_distinct(cur_data()$src_ref[!is.na(cur_data()$envir)]),  n_call_sites)) %>%
   arrange(desc(n_call_sites))
@@
 ```
+
+# Accessing a slot
+
+Accessing a slot is done with `$`, `[]` or `[[]]`.
+
+```{r}
+slot_access <- E_cano %>% filter(minimized == "$")
+
+generalized_slot_access <- E_cano %>% filter(has_dollar | has_bracket)
+
+nb_slot_sites <- slot_access %>% pull(src_ref) %>% n_distinct()
+
+nb_generalized_slot_sites <- generalized_slot_access %>% pull(src_ref) %>% n_distinct()
+
+r("nb slot access", nb_slot_sites)
+r("slot access percent", ratio(nb_slot_sites, nb_call_sites))
+r("generalized slot access", nb_generalized_slot_sites)
+r("generalized slot access percent", ratio(nb_generalized_slot_sites, nb_call_sites))
+```

### 2021-03-18 paper/main.tex
Analyzing expression vs analyzing evaluation of expression
Correct the percentage of non-default environments.

@@
 We also track whether assignment operators are used in the expression or not.
-After the simplification steps, we obtain 15 different \emph{minimized} expressions, of which we show the ten first most frequent in numbers of sites in Table~\ref{tab:minimizedexpressions}.
+After the simplification steps, we obtain 15 different \emph{minimized} expressions, of which we show the eleven first most frequent in number of sites in Table~\ref{tab:minimizedexpressions}.
 
@@
 	\end{tabular}
-	\caption{Minimized expressions: ten most frequent in terms of sites.
+	\caption{Minimized expressions: eleven most frequent in number of sites.
 \#sites is the number of sites; \%sites the percentage of sites across the total number of sites (\packageNbCallSites sites) with at least this minimized expression; \#packages is the number of packages with at least one of the minimized expression and measures how widespread a minimized expression is.
@@
 
-Note that a site can exhibit several minimized expressions. However, \packageNbOneMinimizedPercent of the sites have only one minimized expression.
+Note that a site can exhibit several minimized expressions.
+However, \packageNbOneMinimizedPercent of the sites have only one minimized expression.
+Note also that if a function is called in that expression, we see only the function call but not the body of the function.
+We will trace what happens during the evaluation of the expression by \eval in Section~\ref{sec:dangerseval}, especially about side effects performed in \eval.
 
@@
 
-Similarly, when a function is evaluated on some variable (\c{F(X)}), the percentage of non default environment is high, \packageMinimizedpercentenvird\%.
+Similarly, when a function is evaluated on some variable (\c{F(X)} minimized expression), the percentage of non default environment is high, \packageMinimizedpercentenvird\%.
 
@@
 
-\section{The dangers of \eval}
+\section{The dangers of \eval} \label{sec:dangerseval}
 

### 2021-03-18 analysis/normalized.Rmd
Improvements

@@
   group_by(minimized) %>%
-  summarize(n_call_sites = n_distinct(src_ref), proportion = ratio(n_call_sites, nb_call_sites), n_packages = n_distinct(ev_package), n_operations = round(mean(n_op, na.rm=TRUE), 2), percent_envir = ratio(n_distinct(cur_data()$src_ref[!is.na(cur_data()$envir)]),  n_call_sites)) %>%
+  summarize(n_call_sites = n_distinct(src_ref), proportion = ratio(n_call_sites, nb_call_sites), n_packages = n_distinct(ev_package), n_operations = round(mean(n_op, na.rm=TRUE), 2), percent_envir = ratio(n_distinct(cur_data()$src_ref[!is.na(cur_data()$envir)]),  n_call_sites), percent_envir_calls = ratio(sum(!is.na(envir)), n())) %>%
   arrange(desc(n_call_sites))

### 2021-03-18 paper/main.tex
Improvements

@@
 
-Similarly, when a function is evaluated on some variable (\c{F(X)} minimized expression), the percentage of non default environment is high, \packageMinimizedpercentenvird\%.
+% the percentage of non-default environment is more convincing when looking at the number of calls rather the number of sites
+
+Similarly, when a function is evaluated on some variable (\c{F(X)} minimized expression), the percentage of non default environment is rather high, \packageMinimizedpercentenvird\%.
 

### 2021-03-22 analysis/normalized.Rmd
Add results of manual inspection of 20 sites with only values

@@
 library(corrr)
+library(boot)
 
@@
 There are `r nrow(big_topcall)` top calls out of `r nrow(widespread_topcall)` that are present in strictly more than 3 call sites.
+
+# Variables
+
+The percentage of non-default envir for variables seem to be quite low for the use-case. What if we remove more complex expressions and just keep the symbols?
+
+```{r}
+var_minimized <- E_cano %>% filter(minimized == "X")
+nb_var_min_sites <- var_minimized %>% pull(src_ref) %>% n_distinct()
+
+var_symb <-  E_cano %>% filter(minimized == "X", type == "symbol")
+nb_var_symb_sites <- var_symb %>% pull(src_ref) %>% n_distinct()
+
+var_symb_non_default_envir <- var_symb %>% filter(!is.na(envir))
+nb_var_symb_ndefault_sites <- var_symb_non_default_envir %>% pull(src_ref) %>% n_distinct()
+
+r("nb non default envir variables percent", ratio(nb_var_symb_ndefault_sites, nb_var_symb_sites))
+```
 
@@
 
+And the results of the manual inspection:
+
+```{r, include=FALSE}
+useful_percent <- 0
+```
+
+```{r, eval=dataset_name == "package"}
+manual_value_necessary <- read_csv(path(params$base_dir, "non-necessary-value-evals.csv"), col_types = cols(src_ref = col_character(), necessary = col_logical()))
+
+statistic <- function(data, i) {
+  df <- data[i,] %>% select(necessary) %>% filter(necessary)
+  nrow(df) / nrow(data[i,])
+}
+
+bootstrap <- boot(manual_value_necessary, statistic, R = 1000)
+
+useful_percent <- mean(bootstrap$t, na.rm = TRUE)
+
+r("useful value eval percent", percent(useful_percent))
+```
+
+`r percent(useful_percent)`% of those `eval` are useful.
+
 We now want to characterize better the sites that are passed a non-default environment. Are they also the sites that can get a value or something with a variable inside?
 
-We compute the sites which are at least one given a `V` minimized expression. Then we look at whether they are passed the default envir or not, and if there are variables in
+We compute the sites which are at least one given a `V` minimized expression. Then we look at whether they are passed the default envir or not, and if there are variables in:
 

### 2021-03-23 analysis/normalized.Rmd
Variables

@@
 ```
+
+What happens for the symbol with the default envir?
+
+```{r}
+var_sym_default_envir <- var_symb %>% filter(is.na(envir)) %>% count(src_ref, sort = TRUE)
+```
+
+Seems it was a bug: when the explicit argument was `parent.frame()`, it was counted as being the default argument (but it does not have the same semantics as a default argument or as an explicit one though!).
 
@@
 
-useful_percent <- mean(bootstrap$t, na.rm = TRUE)
+#useful_percent <- mean(bootstrap$t, na.rm = TRUE)
+useful_percent <- bootstrap$t0
 

### 2021-03-23 paper/main.tex
Add results of manual inspection of 20 sites with only values
Actual figure of manual imspection in the paper
Variables

@@
 a variable.} It happens in \packageMinimizedpropsitesa of the sites and is widespread across packages as it is found in \packageMinimizedpackagea packages across our CRAN corpus.
-The \c{envir} argument is explicitly passed in \packageMinimizedpercentenvira\% of the call sites, which suggests that \eval is used here to get the value of the variable in a non-local environment.
+The \c{envir} argument is explicitly passed in \packageMinimizedpercentenvira\% of the call sites, which suggests that \eval is often used here to get the value of the variable in a non-local environment.
+If we look only at the \emph{symbols} passed to \eval, the percentage of non-default environments goes up to \packageNbNonDefaultEnvirVariablesPercent.
 
@@
 
-\c{paste} concatenates strings and \eval will just return that string. \c{as.formula} can take a string and convert it to a formula; \eval is not needed here.
+\c{paste} concatenates strings and \eval will just return that string.
+\c{as.formula} can take a string and convert it to a formula; \eval is not needed here.
+In 20 random sites among the ones with values, a conservative manual inspection showed that \packageUsefulValueEvalPercent of the sites  had a  legitimate reason to use \eval.
 

### 2021-03-24 analysis/patterns2.Rmd
Patterns

--- 2021-03-18 analysis/normalized.Rmd
@@
 ---
-title: "Normalized expressions"
+title: "Patterns and anti-patterns"
 output:
@@
 # Tags will be prefixed by the dataset name
-create_tags(path(TAGS_DIR, paste0(dataset_name, "_normalized_expr.tex")), prefix = dataset_name, default = TRUE)
+create_tags(path(TAGS_DIR, paste0(dataset_name, "_patterns.tex")), prefix = dataset_name, default = TRUE)
 # Also use ggsave_prefix instead of ggsave if you need to save a graph
@@
 
-lock_path <- paste0(params$base_dir, ".normalized-lock.fst")
-min_path <- paste0(params$base_dir, "Ecano.fst")
+lock_path <- paste0(params$base_dir, ".patterns-lock.fst")
+min_path <- paste0(params$base_dir, "Epatterns.fst")
 ```
@@
 
-# Ranking of normalized and minimized expressions
+# Variable lookup in another environment
 
-Normalized: 
+Variables lookup are characterized by:
+
+- symbol as the type of the expression
+- a non-default environment
+
 ```{r}
-widespread_canonic <- E_cano %>%
-  group_by(normalized) %>%
-  summarize(n_call_sites = n_distinct(src_ref), n_packages = n_distinct(ev_package))
+variable_lookup <- E_cano %>% filter(type == "symbol", !is.na(envir))
 
-widespread_canonic %>%
-  arrange(desc(n_call_sites)) %>%
-  datatable()
+nb_variable_lookup <- variable_lookup %>% pull(src_ref) %>% n_distinct()
 ```
 
-Minimized:
-```{r}
-widespread_minimized <- E_cano %>%
-  group_by(minimized) %>%
-  summarize(n_call_sites = n_distinct(src_ref), proportion = ratio(n_call_sites, nb_call_sites), n_packages = n_distinct(ev_package), n_operations = round(mean(n_op, na.rm=TRUE), 2), percent_envir = ratio(n_distinct(cur_data()$src_ref[!is.na(cur_data()$envir)]),  n_call_sites), percent_envir_calls = ratio(sum(!is.na(envir)), n())) %>%
-  arrange(desc(n_call_sites))
+There are `r nb_variable_lookup` sites with variable lookup, i.e `r ratio(nb_variable_lookup, nb_call_sites)`%.
 
-widespread_minimized %>%
-  datatable()
-```
+# `match.call`
 
-```{r}
-nrows <- 12
-r_vec("MinimizedCallSites", nrows, pull(widespread_minimized, n_call_sites))
-r_vec("MinimizedPropSites", nrows, pull(widespread_minimized, proportion))
-r_vec("MinimizedPackage", nrows, pull(widespread_minimized, n_packages))
-r_vec("MinimizedOperations", nrows, pull(widespread_minimized, n_operations))
-r_vec("MinimizedPercentEnvir", nrows, pull(widespread_minimized, percent_envir))
-```
-
-`n_operations` is the average of R interpreter eval loop per minimized expression.
-`prop_envir` shows how much often an *explicit* environment is passed to `eval`.
-Looking up to a variable and applying a function to a variable, are the two non-trivial categories that most of the time explicitly use the `envir` argument of `eval`.
-
-# How does it evolve per call site
-
-To check whether the normalization makes sense, we have a look at how many normalized expressions we have per call sites.
-
-```{r}
-# count hash because resolved expressions are trimmed but hash are on the full expression
-by_src_ref <- E_cano %>% group_by(src_ref) %>%
-  summarise(n_resolved = n_distinct(hash), n_normalized = n_distinct(normalized), n_minimized = n_distinct(minimized))
-```
-
-```{r}
-by_src_ref %>% arrange(desc(n_resolved)) %>% datatable()
-```
-
-There should always be less normalized expressions than resolved ones. 
-We can see for how many there are strictly less (or one when it was already one).
-
-```{r}
-decrease_exprs <- by_src_ref %>% filter(n_resolved == 1 | n_resolved > n_minimized)  %>% arrange(desc(n_minimized))
-```
-
-It corresponds to `r nrow(decrease_exprs)` call sites so only `r nb_call_sites - nrow(decrease_exprs)` have non-trivially not decreased the number of call sites.
-
-# Top calls
-
-For the functions at the root of the AST, per call site:
-
-```{r}
-widespread_topcall <- E_cano %>%
-  group_by(topcall) %>%
-  summarize(n_call_sites = n_distinct(src_ref), percent = round(ratio(n_call_sites, nb_call_sites), 2)) %>% 
-  arrange(desc(n_call_sites))
-```
-
-```{r}
-widespread_topcall %>% datatable()
-```
-
-```{r}
-big_topcall <- widespread_topcall %>% filter(n_call_sites > 3) 
-```
-
-There are `r nrow(big_topcall)` top calls out of `r nrow(widespread_topcall)` that are present in strictly more than 3 call sites.
-
-# Function definitions
-
-```{r}
-fundefs <- E_cano %>% filter(has_fundef)
-
-nb_fundef_callsites <- fundefs %>% pull(src_ref) %>% n_distinct()
-```
-
-There are `r nb_fundef_callsites` call sites with function definitions, i.e. `r ratio(nb_fundef_callsites, nb_call_sites)` %.
-
-# Model.frame
-
-```{r}
-modelframes <- E_cano %>% filter(is_model)
-
-nb_modelf_call_sites <- modelframes %>% pull(src_ref) %>% n_distinct()
-```
-
-There are `r nb_modelf_call_sites` call sites with `mode.frame`, i.e. `r ratio(nb_modelf_call_sites, nb_call_sites)`%.
-
-Is it usually in top call position?
-
-```{r}
-modelframes %>% filter(topcall != "model.frame")
-```
-
-Yes, always!
-
-Is it correlated with `match.call`?
-
-```{r}
-nb_mf_match.call <- modelframes %>% filter(!is.na(match.call)) %>% pull(src_ref) %>% n_distinct()
-```
-
-`r nb_mf_match.call` call sites out of `model.frame` `r nb_modelf_call_sites` call sites use the `match.call` pattern, i.e. `r ratio(nb_mf_match.call, nb_modelf_call_sites)`%.
-
-We can also reciprocally look at how many call sites with `match.call` have `model.frame`:
+The goal is to get a call and modify it or use it.
 
@@
 match.call_df <- E_cano %>% filter(!is.na(match.call))
-
 nb_m.c_call_sites <- match.call_df %>% pull(src_ref) %>% n_distinct()
 
-nb_m.c_mf_call_sites <- match.call_df %>% filter(is_model) %>% pull(src_ref) %>% n_distinct()
 ```
 
-So `r ratio(nb_m.c_mf_call_sites, nb_m.c_call_sites)`% of the call sites with the `match.call` pattern use `model.frame`.
-
-# Assignments
-
-When do we perform assignments (causing potentially side effects)?
-
-```{r}
-assigns <- E_cano %>% filter(has_assigns > 0)
-
-nb_assigns_call_sites <- assigns %>% pull(src_ref) %>% n_distinct()
-
-r("nb assign sites", nb_assigns_call_sites)
-r("assign sites percent", ratio(nb_assigns_call_sites, nb_call_sites))
-```
-
-There are `r nb_assigns_call_sites` call sites with assignments, i.e. `r ratio(nb_assigns_call_sites, nb_call_sites)`%.
-
-Is the `envir` argument non trivial?
-
-```{r}
-assign_env <- assigns %>% filter(!is.na(envir))
-
-nb_a_env_call_sites <- assign_env %>% pull(src_ref) %>% n_distinct()
-
-r("non default envir assign sites", nb_a_env_call_sites)
-r("non default envir assign sites percent", ratio(nb_a_env_call_sites, nb_assigns_call_sites))
-```
-
-There are `r nb_a_env_call_sites` call sites where `envir` is non-trivial, with an assignment, i.e. `r ratio(nb_a_env_call_sites, nb_assigns_call_sites)`% of the call sites with assignments.
-
-# `match.call` pattern
-
-There are `r nb_m.c_call_sites` call sites using the `match.call`pattern, i.e. `r ratio(nb_m.c_call_sites, nb_call_sites)`%.
-
-The top calls for this pattern are:
+There are `r nb_m.c_call_sites` sites using `match.call`, i.e. `r ratio(nb_m.c_call_sites, nb_call_sites)`%.
 
@@
 
-Most are `model.frame`. If we add other model functions, such as `glm` and `plm`:
+## Evaluating some of the call arguments
+
+In that case, `topcall` is `NA`, `$` or `[`.
+
+```{r}
+mc_arg_functions <- c(NA_character_, "$", "[")
+
+nb_arg_mc_sites <- match.call_df %>% 
+  filter(topcall %in% mc_arg_functions) %>% 
+  pull(src_ref) %>% 
+  n_distinct()
+```
+
+it amounts for `r nb_arg_mc_sites` call sites, i.e `r ratio(nb_arg_mc_sites, nb_m.c_call_sites)`%.
+
+## `match.call` and statistical models
+
+Veyr often, `match.call` is used to build statistical models, by replacing the call function by `model.frame` (but also other statistical models, such as `glm`).
 
@@
 
-it amounts for `r nb_stat_call_sites` call sites, i.e `r ratio(nb_stat_call_sites, nb_call_sites)`%.
+it amounts for `r nb_stat_call_sites` call sites, i.e `r ratio(nb_stat_call_sites, nb_m.c_call_sites)`%.
 
-# Substitute&co
+# Parsing strings
 
-`substitute`, `quote` and so on are used to do meta-programming.
+As a first approximation, it corresponds to type `expression` (although `expr` can also build such expressions). However, it is an under-approximation, because the expression could be later used...
 
 ```{r}
-meta_df <- E_cano %>% filter(has_meta_op)
+parse_eval <- E_cano %>% filter(type == "expressions" | str_detect(ev_expr, fixed("parse")))
 
-nb_meta_call_sites <- meta_df %>% pull(src_ref) %>% n_distinct()
+nb_parse_eval <- parse_eval %>% pull(src_ref) %>% n_distinct()
 ```
 
-There are `r nb_meta_call_sites` call sites with `substitute` and `substitute`-like operators.
+There are `r nb_parse_eval` sites using `parse`, i.e. `r ratio(nb_parse_eval, nb_call_sites)`%.
 
-Are the `substitute` sometimes not placed as a top call?
+## Logging
+
+https://github.com/cran/Rcmdr/blob/4b196764a44a047f8c7be3f4c629f52bee2c0c80/R/commander.R#L1301
+
+The goal is to print the command and then to execute it.
+
+## Formulas
 
 ```{r}
-meta_not_top <- meta_df %>% filter(!topcall %in% c("substitute", "quote", "enquote", "bquote"))
+formulas <- E_cano %>% filter(str_detect(ev_expr, fixed("parse")), topcall == "~" | str_detect(ev_expr, fixed("formula")))
 
-nb_m_n_top_call_sites <- meta_not_top %>% pull(src_ref) %>% n_distinct()
+nb_formula_sites <- formulas %>% pull(src_ref) %>% n_distinct()
 ```
 
-It happens for `r nb_m_n_top_call_sites` call sites, i.e. `r ratio(nb_m_n_top_call_sites, nb_call_sites)`%.
+There are `r nb_formula_sites` sites building formulas with `parse`, i.e. `r ratio(nb_formula_sites, nb_call_sites)`%.
+
+# `Parse` and plotting functions
+
+We approximate the plotting function by just looking for `plot` in their name:
+
+```{r}
+plotting <- E_cano %>% filter(str_detect(normalized, fixed("plot"))) # not in topcall because f it results from parse...
+
+nb_plotting_call_sites <- plotting %>% pull(src_ref) %>% n_distinct()
+```
+
+There are `r nb_plotting_call_sites` plotting call sites, i.e. `r ratio(nb_plotting_call_sites, nb_call_sites)`% and `r ratio(nb_plotting_call_sites, nb_parse_eval)`% of the parsing sites.
+
+Are they built using parse? For that, we look at the non-resolved expression (`parse` will disappear after).
+
+Currently, we use a regex. When `parsed` column is fixed, we will get more accurate results.
+
+```{r}
+parse_plot <- plotting %>% filter(str_detect(ev_expr, fixed("parse")))
+
+nb_parse_plot_call_sites <- parse_plot %>% pull(src_ref) %>% n_distinct()
+```
+
+This already amounts to `r nb_parse_plot_call_sites` call sites, i.e. `r ratio(nb_parse_plot_call_sites, nb_plotting_call_sites)`% of the plotting call sites.
 
@@
 
-# `Parse` and plotting functions
+# Symbolic computations
 
-We approximate the plotting function by just looking for `plot` in their name:
+`D` operator to derive an expression symbolically, `Simplify`
 
-```{r}
-plotting <- E_cano %>% filter(str_detect(normalized, fixed("plot"))) # not in topcall because f it results from parse...
+For instance: https://github.com/cran/Deriv/blob/9ca174a959d8eb39b0a1f1a70de26d147a837176/R/Simplify.R#L572
 
-nb_plotting_call_sites <- plotting %>% pull(src_ref) %>% n_distinct()
-```
+Usually on an expression object, not a language object. Environment type seems to be `list` in that case, most often.
 
-There are `r nb_plotting_call_sites` plotting call sites, i.e. `r ratio(nb_plotting_call_sites, nb_call_sites)`%.
-
-Are they built using parse? For that, we look at the non-resolved expression (`parse` will disappear after).
-
-Currently, we use a regex. When `parsed` column is fixed, we will get more accurate results.
-
-```{r}
-parse_plot <- plotting %>% filter(str_detect(ev_expr, fixed("parse")))
-
-nb_parse_plot_call_sites <- parse_plot %>% pull(src_ref) %>% n_distinct()
-```
-
-This already amounts to `r nb_parse_plot_call_sites` call sites, i.e. `r ratio(nb_parse_plot_call_sites, nb_plotting_call_sites)`% of the plotting call sites.
-
-### Infix operators
+# Infix operators
 
@@
 
-# Polymorphism in more details
+# Meta-programming
 
-We have only 15 different kind of minimized expression. Do some minimized expressions often go together?
+As a first approximation, it corresponds to type `language`.
 
 ```{r}
-minimized_exprs <- E_cano %>% pull(minimized) %>% unique()
+parse_eval <- E_cano %>% filter(type == "language")
 
-create_set <- function(l) {
-  s <- rep.int("F", length(minimized_exprs))
-  names(s) <- minimized_exprs
-
-  for(exp in l) {
-    s[[exp]] <- "T"
-  }
-
-  return(paste0(s, collapse = ""))
-}
-
-index_v <- match("V", minimized_exprs)
-index_x <- match("X", minimized_exprs)
-
-has_value <- function(l) {
-  substr(l, index_v, index_v) == "T"
-}
-
-is_variable <- function(l) {
-  substr(l, index_x, index_x) == "T"
-}
-
-count_elements <- function(l) {
-  sum(map_int(strsplit(l, split="")[[1]], function(ch) {if(ch == "T") {1L} else {0L}}))
-}
-
-show_set <- function(s) {
-  l <- strsplit(s, split="")[[1]]
-  names(l) <- minimized_exprs
-
-  l <- l[which(l == "T")]
-  return(paste0(names(l), collapse = "; "))
-}
-
-poly <- E_cano %>% group_by(src_ref) %>%
-    summarize(common_minimized = create_set(minimized), n_exprs = n_distinct(minimized)) 
+nb_parse_eval <- parse_eval %>% pull(src_ref) %>% n_distinct()
 ```
 
-Now, we can show the mot common minimized expressions when there are more than 2 for one call site:
+## selecting columns in a dataframe
 
-```{r}
-poly %>%
-  count(common_minimized,  sort = TRUE) %>%
-  mutate(common_minimized = map_chr(common_minimized, show_set)) %>% 
-  mutate(percent = round(100 * n / nb_call_sites, 2)) %>% 
-  datatable()
+https://github.com/cran/epiDisplay/blob/d3d129a690a84b95c9aec437690227f1d50e3394/R/epiDisplay.R#L5915
+
+```r 
+ selected <- eval(substitute(vars), nl, parent.frame())
 ```
 
-How many sites have only one minimized expression?
+`vars` correspond to the columns in the dataframe we want to use.
 
-```{r}
-n_one_mini <- poly %>% filter(n_exprs == 1) %>% nrow()
+Some examples on how it is called. Columns can be selected by name or by column number.
 
-r("Nb One Minimized", n_one_mini)
-r("Nb One Minimized Percent", ratio(n_one_mini, nb_call_sites))
+```r 
+tableStack(bakedham:fruitsalad, .data)
+
+tableStack(vars=4:25, .data, by=Origin)
+
+tableStack(c(qa1,qa13:qa18,mean.score,total.score), .data, by=sex, test=FALSE)
 ```
 
-# Values
+# Non-necessary `eval`
 
-Many call sites only evaluate a value. It is either a language expression or symbol, or an actual value.
-
-```{r}
-value_minimized <- widespread_minimized %>% filter(minimized == "V")
-```
-
-These are the number of call sites that are actual values.
-```{r}
-main_types <- E_cano %>% group_by(src_ref) %>% summarise(nb_types = n_distinct(type), type = first(type), nb_runs = n_distinct(file))
-
-only_actual_values <- main_types %>% filter(!type %in% c("symbol", "language", "expression", "promise"), nb_types == 1)
-nb_actual_values_sites <- only_actual_values %>% pull(src_ref) %>% n_distinct()
-
-r("nb call sites unique actual value", nb_actual_values_sites)
-r("call sites unique actual value percent", ratio(nb_actual_values_sites, value_minimized$n_call_sites))
-r("median run sites unique actual value", median(only_actual_values$nb_runs, na.rm = TRUE))
-r("average run sites unique actual value", mean(only_actual_values$nb_runs, na.rm = TRUE))
-r("sd run sites unique actual value", sd(only_actual_values$nb_runs, na.rm = TRUE))
-```
-
-Is it a simple or a complex value?
-
-```{r}
-val_exprs <- E_cano %>% filter(minimized == "V") %>%
-  mutate(caller_srcref = if_else(is.na(caller_srcref), file, caller_srcref)) %>%
-  group_by(src_ref) %>%
-  summarise(avg_ast_nodes = mean(ast_nodes, na.rm = TRUE), avg_arg_size = mean(arg_size, na.rm = TRUE), nb_runs = n_distinct(file), nb_caller_srcref = n_distinct(caller_srcref), nb_calls = sum(duplicates), nb_caller_exprs = n_distinct(caller_expression)) %>%
-  arrange(desc(avg_ast_nodes))
-
-n_val_exprs <- val_exprs %>% nrow()
-n_one_node <- val_exprs %>% filter(avg_ast_nodes == 1) %>% nrow()
-
-r("val one node", n_one_node)
-r("val one node percent", ratio(n_one_node, n_val_exprs))
-```
-
-Are the sites resolving to just a value well covered?
-
-```{r}
-med_runs <- median(val_exprs$nb_runs, na.rm = TRUE)
-avg_runs <- mean(val_exprs$nb_runs, na.rm = TRUE)
-sd_runs <- sd(val_exprs$nb_runs, na.rm = TRUE)
-
-r("one val median runs", med_runs)
-r("one val average runs", avg_runs)
-r("one val sd runs", sd_runs)
-```
-
-Is there a correlation between the number of caller srcref and the presence of other minimized expressions for one call site?
-
-There are many `NA` caller srcref, so we count them as 1 currently...
-
-```{r eval = FALSE}
-val_exprs_comb <- val_exprs %>% left_join(main_types, by="src_ref")  %>%
-  mutate(calls_per_run = nb_calls / nb_runs) %>%
-  mutate(calls_per_site = nb_calls / nb_caller_srcref) %>%
-  mutate(calls_per_site_per_run = nb_calls / nb_runs / nb_caller_srcref)
-# normalization to be able to see both on the same graph
-
-val_exprs_comb %>% ggplot(aes(x = nb_caller_srcref, y = nb_types)) + 
-  geom_jitter(alpha = 0.3) +
-  scale_x_log10()
-
-val_exprs_comb %>% ggplot(aes(x = nb_runs, y = nb_types)) + 
-  geom_jitter(alpha = 0.3) +
-  scale_x_log10()
-
-val_exprs_comb %>% ggplot(aes(x = calls_per_run, y = nb_types)) + 
-  geom_jitter(alpha = 0.3) +
-  scale_x_log10()
-
-val_exprs_comb %>% ggplot(aes(x = calls_per_site, y = nb_types)) + 
-  geom_jitter(alpha = 0.3) +
-  scale_x_log10()
-
-val_exprs_comb %>% ggplot(aes(x = calls_per_site_per_run, y = nb_types)) + 
-  geom_jitter(alpha=0.3) +
-  scale_x_log10()
-
-val_exprs_comb %>% ggplot(aes(x = nb_caller_exprs, y = nb_types)) + 
-  geom_jitter(alpha=0.3) +
-  scale_x_log10()
-
-df <- val_exprs_comb %>% select(where(is.numeric))
-
-df %>% correlate(method = "spearman") %>% network_plot()
-df %>% correlate(method = "spearman") %>% rplot() + theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=1))
-
-```
-
-Manual inspection of the source code shows that there are actually some call sites that take actual values.
-If a site is exercised by many caller sites, or by many different runs, and has only values, then we reckon it is probably a _useless_ `eval`...
-
-```{r}
-val_only <- val_exprs %>% right_join(only_actual_values, by = c("src_ref", "nb_runs"))
-
-max_val_runs <- val_only %>% filter(nb_caller_exprs == max(nb_caller_exprs)) #it happens to also be the one with the most caller expressions
-
-val_only %>% filter(nb_caller_exprs > 5) # those ones are likely to be useless eval
-```
-
-We now want to characterize better the sites that are passed a non-default environment. Are they also the sites that can get a value or something with a variable inside?
-
-We compute the sites which are at least one given a `V` minimized expression. Then we look at whether they are passed the default envir or not, and if there are variables in
-
-```{r}
-minimized_sites <- E_cano %>% semi_join(val_exprs, by="src_ref") %>% group_by(src_ref) %>% 
-  summarize(has_envir = all(!is.na(envir)), has_var = any(has_var, na.rm = TRUE))
-
-non_default_envir <- minimized_sites %>% filter(has_envir)
-
-non_default_with_var <- minimized_sites %>% filter(has_envir, has_var)
-
-r("non default envir value percent", ratio(nrow(non_default_envir), nrow(minimized_sites)))
-r("non default envir with var percent", ratio(nrow(non_default_with_var), nrow(non_default_envir)))
-```
-
-# Accessing a slot
-
-Accessing a slot is done with `$`, `[]` or `[[]]`.
-
-```{r}
-slot_access <- E_cano %>% filter(minimized == "$")
-
-generalized_slot_access <- E_cano %>% filter(has_dollar | has_bracket)
-
-nb_slot_sites <- slot_access %>% pull(src_ref) %>% n_distinct()
-
-nb_generalized_slot_sites <- generalized_slot_access %>% pull(src_ref) %>% n_distinct()
-
-r("nb slot access", nb_slot_sites)
-r("slot access percent", ratio(nb_slot_sites, nb_call_sites))
-r("generalized slot access", nb_generalized_slot_sites)
-r("generalized slot access percent", ratio(nb_generalized_slot_sites, nb_call_sites))
-```
+## Values into `eval`

### 2021-03-24 analysis/normalized.Rmd
Function definition and higher-order functions

@@
 ```{r}
-fundefs <- E_cano %>% filter(has_fundef)
 
-nb_fundef_callsites <- fundefs %>% pull(src_ref) %>% n_distinct()
+fundefs <- E_cano %>% filter(minimized == "FUN")
+nb_fundef_sites <- fundefs %>% pull(src_ref) %>% n_distinct()
+ndefault_envir_fundefs <- fundefs %>% filter(!is.na(envir))
+nb_ndefault_envir_fundef_sites <- ndefault_envir_fundefs %>% pull(src_ref) %>% n_distinct()
+
+generalized_fundefs <- E_cano %>% filter(has_fundef)
+nb_generalized_fundef_sites <- generalized_fundefs %>% pull(src_ref) %>% n_distinct()
+ndefault_envir_gen_fundefs <- generalized_fundefs %>% filter(!is.na(envir))
+nb_ndefault_envir_gen_fundef_sites <- ndefault_envir_fundefs %>% pull(src_ref) %>% n_distinct()
+
+r("function definition sites", nb_fundef_sites)
+r("function definition sites percent", ratio(nb_fundef_sites, nb_call_sites))
+r("fundef non default envir sites percent", ratio(nb_ndefault_envir_fundef_sites, nb_fundef_sites))
+r("generalized function definition sites", nb_generalized_fundef_sites)
+r("generalized function definition sites percent", ratio(nb_generalized_fundef_sites, nb_call_sites))
+r("generalized funded non default envir sites percent", ratio(nb_ndefault_envir_gen_fundef_sites, nb_generalized_fundef_sites))
 ```
 
-There are `r nb_fundef_callsites` call sites with function definitions, i.e. `r ratio(nb_fundef_callsites, nb_call_sites)` %.
+There are `r nb_generalized_fundef_sites` call sites with function definitions, i.e. `r ratio(nb_generalized_fundef_sites, nb_call_sites)` %.
 
@@
 
+ndefault_envir_slot <- generalized_slot_access %>% filter(!is.na(envir))
+nb_ndefault_envir_slot_sites <- ndefault_envir_slot %>% pull(src_ref) %>% n_distinct()
+
 r("nb slot access", nb_slot_sites)
@@
 r("generalized slot access percent", ratio(nb_generalized_slot_sites, nb_call_sites))
+r("ndefault envir slot site", nb_ndefault_envir_slot_sites)
+r("ndefault envir slot site percent", ratio(nb_ndefault_envir_slot_sites, nb_generalized_slot_sites))
 ```

### 2021-03-24 paper/main.tex
Function definition and higher-order functions

@@
 		\hline
-		\texttt{\$}& \packageMinimizedcallsitesf & \packageMinimizedpropsitesf & \packageMinimizedpackagef & \packageMinimizedoperationsfRnd & \packageMinimizedpercentenvirf \\
-		\hline
-		\texttt{F()}& \packageMinimizedcallsitesg & \packageMinimizedpropsitesg & \packageMinimizedpackageg & \packageMinimizedoperationsgRnd & \packageMinimizedpercentenvirg \\
-		\hline
-		\texttt{<-} & \packageMinimizedcallsitesh & \packageMinimizedpropsitesh & \packageMinimizedpackageh & \packageMinimizedoperationshRnd & \packageMinimizedpercentenvirh \\
-		\hline
-		\texttt{FUN} & \packageMinimizedcallsitesi  & \packageMinimizedpropsitesi & \packageMinimizedpackagei & \packageMinimizedoperationsiRnd & \packageMinimizedpercentenviri \\
+		\texttt{F()}& \packageMinimizedcallsitesf & \packageMinimizedpropsitesf & \packageMinimizedpackagef & \packageMinimizedoperationsfRnd & \packageMinimizedpercentenvirf \\
+		\hline
+		\texttt{\$}& \packageMinimizedcallsitesg & \packageMinimizedpropsitesg & \packageMinimizedpackageg & \packageMinimizedoperationsgRnd & \packageMinimizedpercentenvirg \\
+		\hline
+		\texttt{FUN} & \packageMinimizedcallsitesh & \packageMinimizedpropsitesh & \packageMinimizedpackageh & \packageMinimizedoperationshRnd & \packageMinimizedpercentenvirh \\
+		\hline
+		\texttt{<-} & \packageMinimizedcallsitesi  & \packageMinimizedpropsitesi & \packageMinimizedpackagei & \packageMinimizedoperationsiRnd & \packageMinimizedpercentenviri \\
 		\hline
@@
 
+% for this one, currently uses any expression with $, [] or [[]] but we should rather use a more restrictive way (take only the ones without interesting calls)
+Minimized expression \texttt{\$} refers to a slot access as the top call in a simple expression, for instance \texttt{a\$slot + 1}.
+It amounts for \packageGeneralizedSlotAccess sites, \ie \packageGeneralizedSlotAccessPercent of the sites.
+
 \paragraph{Values.}
@@
 
-\paragraph{Function definition.} \c{FUN} represents a function definition. It entails very few steps of the interpreter, as the body of the new function is just stored for later execution.
+\paragraph{Function definition.} \c{FUN} represents a function definition at the top level of the expression, and happens in \packageFunctionDefinitionSites sites, \ie \packageFunctionDefinitionSitesPercent of the sites.
+It entails very few steps of the interpreter, as the function is returned by \eval, and the body of the new function is just stored for later execution.
+If we count also function definition inside more complex expressions, it amounts to \packageGeneralizedFunctionDefinitionSites sites, \ie \packageGeneralizedFunctionDefinitionSitesPercent and suggests the widespread use of higher-order functions in R, such as \c{lapply}, the R version of \c{map}.
 
@@
 We more precisely instrument side-effects to environment in Section~REF.}
-
-\paragraph{Accessing a slot.}
 

### 2021-03-24 analysis/normalized.Rmd
Combination of normalized expressions

@@
 ```{r}
-poly %>%
+combination_ranking <- poly %>%
   count(common_minimized,  sort = TRUE) %>%
@@
   mutate(percent = round(100 * n / nb_call_sites, 2)) %>%
+  mutate(cumpercent =cumsum(percent))
+
+combination_ranking %>% 
   datatable()
+```
+
+And represented as a bar plot:
+
+```{r}
+combination_ranking %>%
+  filter(cumpercent < 95) %>%
+  ggplot(aes(x = fct_reorder(common_minimized, n), y = percent)) + 
+  geom_col() + 
+  labs(x = "Combinations", y = "%") +
+  coord_flip()
+  #theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
+
+ggsave_prefix("combination_minimized.pdf")
 ```

### 2021-03-24 paper/main.tex
Combination of normalized expressions

@@
 
-\paragraph{Large expressions }  Compositions of functions, without variables (\c{F(F(X))} minimized expression) is the second most frequent minimized expression.
+\paragraph{Large expressions.}  Compositions of functions, without variables (\c{F(F(X))} minimized expression) is the second most frequent minimized expression.
 \c{\{BLOCK\}} is a sequence of statements.
@@
 
-The most common combinations for sites with several minimized expressions are:
+The most common combinations for sites with several minimized expressions are shown in Figure~\ref{fig:frequentcombinationminimized}.
+
+\begin{figure}
+	\centering 
+	\includegraphics[width=\textwidth]{package_combination_minimized.pdf}
+	\caption{Most frequent combinations of minimized expressions per site, with a cumulative share of 95\% of the sites.} \label{fig:frequentcombinationminimized}
+\end{figure}
 

### 2021-03-27 paper/main.tex
Correct Kaggle numbers

@@
 
-\paragraph{Kaggle} In the Kaggle dataset, \kaggleNbruns packages trigger at least one \eval.
-In total, \kaggleAllcalls \eval were recorded.
-Most of them are calls to the normal \eval, whereas a few, \kaggleLocals, are calls to \c{local}.
-This happens in the machinery that loads library code.
+\paragraph{Kaggle} In the Kaggle dataset, \kaggleNbruns notebooks trigger at least one \eval. In total, \kaggleAllcalls \eval were recorded. All of them are calls to the normal \eval.
 Out of \kaggleStaticeval static \eval sites, we exercise \kaggleTriggeredeval of them.
-Mot \eval are invoked only once.
+Most \eval are invoked only once.
 Only one \eval amounts for several thousand, \kaggleMaxcalls,  invocations.
@@
 
-\subsection{Normalized expressions} During tracing, we gather all source expressions and all resolved expression arguments passed to \eval.
-For instance, a non-resolved expression and some of its resolved expressions can be:
+\subsection{Arguments of \eval}
+
+\eval has three arguments, two of which we study in more details:
+
+\begin{itemize}
+	\item \c{expr}: the expression to evaluate
+	\item \c{envir}: the environment in which performing the evaluation
+	\item \c{enclos}
+\end{itemize}
+
+\subsection{The environment passed to \eval.}
+
+The environment argument can be \c{NULL}, a \c{List}
+or a data frame. This happens in 3.5\% of the cases: \eval copies the
+fields of the list or data frame and creates bindings for them in a
+new environment. This pattern is used to evaluate formulas which can
+directly refer to the fields of the data. The \c{envir} argument can
+also be a number $n$. It means that the environment in which the
+expressions is evaluated will be the result of \c{sys.call(n)} where
+$n$ refers to the $n$-th stack frame.
+
+The top-level environment in R is called the global environment. New
+environments can be created using \c{new.env}. They can be provided a
+parent environment which becomes the enclosing scope of the new
+environment.
+
+We looked at the environments passed to all the \eval calls in our
+corpus. Table~\ref{tab:environments} summarizes the results. A numeric
+environment class \c{n} denotes the environment of the $n$-th call
+stack frame from the current function. \c{global} denotes the
+top-level environment and \c{list} denotes a list passed for
+evaluation of formulas. Environment classes of the form $n+$ denote
+the $n$-th environment extended with a new environment. The new
+environment provides a limited form of sandboxing. All assignments
+using the \c{\<-} function occur inside it and prevent the extended
+environment from mutation. However, it is still possible to mutate the
+extended environment using the \c{\<\<-} or \c{assign} functions; but,
+that happens rarely.
+
+\begin{table}[htbp]{ \centering
+		\begin{tabular}{c|c|c|c}\hline
+			\multicolumn{2}{c|}{Core} & \multicolumn{2}{|c}{Packages}  \\\hline
+			Environment Class & Eval Calls \% & Environment Class & Eval Calls\\\hline
+			\CoreEnvClassA & \CoreProportionA & \PackagesEnvClassA &  \PackagesProportionA\\
+			\CoreEnvClassB & \CoreProportionB & \PackagesEnvClassB &  \PackagesProportionB\\
+			\CoreEnvClassC & \CoreProportionC & \PackagesEnvClassC &  \PackagesProportionC\\
+			\CoreEnvClassD & \CoreProportionD & \PackagesEnvClassD &  \PackagesProportionD\\
+			\CoreEnvClassE & \CoreProportionE & \PackagesEnvClassE &  \PackagesProportionE\\\hline
+	\end{tabular}} \label{tab:environments}
+	\caption{Environments in terms of \eval calls}
+\end{table}
+
+We observe that a disproportionately high number of core R \eval calls
+access the caller's caller's environment. This is because many core R
+functions call functions that pass the result of \c{parent.frame()} to
+an \eval. A disproportionately high number of calls to \eval happens
+in an extended top-level environment. This can be explained by the
+fact that many packages evaluate code passed from the user's workspace
+in the top-level environment to access the bindings.
+
+The \c{imchange} function of package \c{imager} makes it possible to
+modify images using a dedicated formula syntax using
+\c{\~}.\footnote{Inspired by {map} in package \emph{purr}.} Here,
+\eval is evaluated in \c{newenv}, which creates a new environment that
+inherits from \c{parent.frame()} by default (classified as 1+).
+
+\begin{lstlisting}
+	newenv <- new.env()
+	...
+	fo <- parse(text=as.character(fo)[2])
+	im[where] <- eval(fo,envir=newenv,enclos=env)
+\end{lstlisting}
+
+\c{adjCoef} in package \emph{actuar} find the root of an equation
+defined by a function \c{h} whose arguments must be named \c{x} and
+\c{y}. \c{h} is transformed into an auxiliary function \c{h2} that can
+be optimized. Here, the list used for \c{envir} ensures the
+correspondance between the textual arguments of \c{h} and the
+arguments of \c{h2}.
+
+\begin{lstlisting}
+	sh <- substitute(h)
+	fcall <- paste(sh, "(x, y)")
+	...
+	h2 <- function(x, y)
+	eval(parse(text = fcall),
+	envir = list(x = x, y = y),
+	enclos = parent.frame(2))
+\end{lstlisting}
+
+\subsubsection{The expression passed to \eval.}
+
+During tracing, we gather all source expressions and all resolved expression arguments passed to \eval. For instance, a non-resolved expression and some of its resolved expressions can be:
 
@@
 
-Note that a site can exhibit several minimized expressions.
+Note that a site can exhibit several minimized expressions (see Figure~\ref{fig:frequentcombinationminimized}).
 However, \packageNbOneMinimizedPercent of the sites have only one minimized expression.
@@
 
+\begin{figure}[htbp]
+	\centering 
+	\includegraphics[width=\textwidth]{package_combination_minimized.pdf}
+	\caption{Most frequent combinations of minimized expressions per site, with a cumulative share of 95\% of the sites.} \label{fig:frequentcombinationminimized}
+\end{figure}
+
 \paragraph{Variables.}
-The most common minimized expression is a variable.\footnote{An arithmetic expression with some variables and constant will also be normalized to \c{X}, i.e.
-a variable.} It happens in \packageMinimizedpropsitesa of the sites and is widespread across packages as it is found in \packageMinimizedpackagea packages across our CRAN corpus.
-The \c{envir} argument is explicitly passed in \packageMinimizedpercentenvira\% of the call sites, which suggests that \eval is often used here to get the value of the variable in a non-local environment.
-If we look only at the \emph{symbols} passed to \eval, the percentage of non-default environments goes up to \packageNbNonDefaultEnvirVariablesPercent.
-
-% the percentage of non-default environment is more convincing when looking at the number of calls rather the number of sites
+The most common minimized expression is a variable.\footnote{An arithmetic expression with some variables and constant, \eg \c{1 + x}, will also be normalized to \c{X}, i.e.
+a variable.} It happens in \packageMinimizedpropsitesa\% of the sites and is widespread across packages as it is found in \packageMinimizedpackagea packages across our CRAN corpus.
+\packageNbSymbolVarSitePercent are only symbols, \ie not complex expressions.
+The \c{envir} argument is explicitly passed in \packageMinimizedpercentenvira\% of the call sites.
+The call sites without an explicit \c{envir} argument all use a specific \eval variant, \c{eval.parent}.
 
@@
 
-% for this one, currently uses any expression with $, [] or [[]] but we should rather use a more restrictive way (take only the ones without interesting calls)
-Minimized expression \texttt{\$} refers to a slot access as the top call in a simple expression, for instance \texttt{a\$slot + 1}.
-It amounts for \packageGeneralizedSlotAccess sites, \ie \packageGeneralizedSlotAccessPercent of the sites.
+Minimized expression \texttt{\$} refers to a slot access, with \texttt{\$}, \texttt{[]} or \texttt{[[]]}, as the top call in a simple expression, for instance \texttt{a\$slot + 1} and  It amounts for \packageNbSlotAccess sites, \ie \packageSlotAccessPercent of the sites.
+
+As a whole, variable access, either as the top call, or inside a function call, possibly with accessing a slot, represents \packageGeneralizedVarAccessSitePercent of the sites.
 
@@
 In \packageNbCallSitesUniqueActualValue sites exhibiting \c{V}, \ie \packageCallSitesUniqueActualValuePercent\% sites, it is only passed such actual values.
-The median number of runs for those sites is \packageMedianRunSitesUniqueActualValue (average \packageAverageRunSitesUniqueActualValue, standard deviation \packageSdRunSitesUniqueActualValue), suggesting it happens because there are rarely triggered in our dataset and all paths are not taken.
+The median number of runs for those sites is \packageMedianRunSitesUniqueActualValue (average \packageAverageRunSitesUniqueActualValue, standard deviation \packageSdRunSitesUniqueActualValue), suggesting it happens because of a lack of coverage, because there are rarely triggered in our dataset and all paths are not taken.
 However, manual inspection of the source code of some examples shows that some sites actually pass values.
@@
 
-\paragraph{\c{model.frame}} This is a function used to create statistical models. We explain it in more details in the section about pattern (see Section~REF).
+\paragraph{\c{model.frame}} This is a function used to create statistical models. We explain it in more details in the section about the usage of \eval (see Section~REF).
 
@@
 
-The most common combinations for sites with several minimized expressions are shown in Figure~\ref{fig:frequentcombinationminimized}.
-
-\begin{figure}
-	\centering 
-	\includegraphics[width=\textwidth]{package_combination_minimized.pdf}
-	\caption{Most frequent combinations of minimized expressions per site, with a cumulative share of 95\% of the sites.} \label{fig:frequentcombinationminimized}
-\end{figure}
-
 \NOTE{Maybe we should ditch the table of the minimized expressions and rather use the ones of combinations.
@@
 
-\NOTE{Two small paragraphs about normalized expressions in Base and Kaggle.}
+\paragraph{Base.} The ranking of minimized expressions in \emph{base} is similar to the one in CRAN bu the share of \c{V} is much larger, as it is present in \baseMinimizedpropsitesa\% of the sites and \c{X} is a distant second, in \baseMinimizedpropsitesb\% of the sites.
+
+\paragraph{Kaggle.} Kaggle is dominated by sites with minimized expression \c{V}, with \kaggleMinimizedcallsitesa sites, and  \kaggleMinimizedpropsitesa\% of the \kaggleNbCallSites sites in total.
+All of those values are an AST size of one node, suggesting they do not come.
+\NOTE{TODO}
+%TODO: all the src ref here seem to be incorrect and not from kaggle.
 
@@
 
-\mypara{Scope.} The environment argument can be \c{NULL}, a \c{List}
-or a data frame. This happens in 3.5\% of the cases: \eval copies the
-fields of the list or data frame and creates bindings for them in a
-new environment. This pattern is used to evaluate formulas which can
-directly refer to the fields of the data. The \c{envir} argument can
-also be a number $n$. It means that the environment in which the
-expressions is evaluated will be the result of \c{sys.call(n)} where
-$n$ refers to the $n$-th stack frame.
-
-The top-level environment in R is called the global environment. New
-environments can be created using \c{new.env}. They can be provided a
-parent environment which becomes the enclosing scope of the new
-environment.
-
-We looked at the environments passed to all the \eval calls in our
-corpus. Table~\ref{tab:environments} summarizes the results. A numeric
-environment class \c{n} denotes the environment of the $n$-th call
-stack frame from the current function. \c{global} denotes the
-top-level environment and \c{list} denotes a list passed for
-evaluation of formulas. Environment classes of the form $n+$ denote
-the $n$-th environment extended with a new environment. The new
-environment provides a limited form of sandboxing. All assignments
-using the \c{\<-} function occur inside it and prevent the extended
-environment from mutation. However, it is still possible to mutate the
-extended environment using the \c{\<\<-} or \c{assign} functions; but,
-that happens rarely.
-
-\begin{table}[htbp]{ \centering
-\begin{tabular}{c|c|c|c}\hline
-\multicolumn{2}{c|}{Core} & \multicolumn{2}{|c}{Packages}  \\\hline
-Environment Class & Eval Calls \% & Environment Class & Eval Calls\\\hline
-\CoreEnvClassA & \CoreProportionA & \PackagesEnvClassA &  \PackagesProportionA\\
-\CoreEnvClassB & \CoreProportionB & \PackagesEnvClassB &  \PackagesProportionB\\
-\CoreEnvClassC & \CoreProportionC & \PackagesEnvClassC &  \PackagesProportionC\\
-\CoreEnvClassD & \CoreProportionD & \PackagesEnvClassD &  \PackagesProportionD\\
-\CoreEnvClassE & \CoreProportionE & \PackagesEnvClassE &  \PackagesProportionE\\\hline
-\end{tabular}} \label{tab:environments}
-\caption{Environments in terms of \eval calls}
-\end{table}
-
-We observe that a disproportionately high number of core R \eval calls
-access the caller's caller's environment. This is because many core R
-functions call functions that pass the result of \c{parent.frame()} to
-an \eval. A disproportionately high number of calls to \eval happens
-in an extended top-level environment. This can be explained by the
-fact that many packages evaluate code passed from the user's workspace
-in the top-level environment to access the bindings.
-
-The \c{imchange} function of package \c{imager} makes it possible to
-modify images using a dedicated formula syntax using
-\c{\~}.\footnote{Inspired by {map} in package \emph{purr}.} Here,
-\eval is evaluated in \c{newenv}, which creates a new environment that
-inherits from \c{parent.frame()} by default (classified as 1+).
-
-\begin{lstlisting}
-newenv <- new.env()
-...
-fo <- parse(text=as.character(fo)[2])
-im[where] <- eval(fo,envir=newenv,enclos=env)
-\end{lstlisting}
-
-\c{adjCoef} in package \emph{actuar} find the root of an equation
-defined by a function \c{h} whose arguments must be named \c{x} and
-\c{y}. \c{h} is transformed into an auxiliary function \c{h2} that can
-be optimized. Here, the list used for \c{envir} ensures the
-correspondance between the textual arguments of \c{h} and the
-arguments of \c{h2}.
-
-\begin{lstlisting}
-sh <- substitute(h)
-fcall <- paste(sh, "(x, y)")
-...
-h2 <- function(x, y)
-    eval(parse(text = fcall),
-    envir = list(x = x, y = y),
-    enclos = parent.frame(2))
-\end{lstlisting}
-
-\section{\eval patterns}
+\section{\eval patterns and anti-patterns}
 

### 2021-03-29 analysis/normalized.Rmd
Update numbers for package normalized expressions

@@
   static_path: evals-static.csv
-  force_rebuild: FALSE
+  force_rebuild: TRUE
 ---
@@
   read_fst(calls_path) %>% as_tibble() -> E_raw
+
+  E_raw <- E_raw %>% mutate(envir_expression = if_else(envir_type == "NULL", "NULL", envir_expression))
 
@@
   group_by(minimized) %>%
-  summarize(n_call_sites = n_distinct(src_ref), proportion = ratio(n_call_sites, nb_call_sites), n_packages = n_distinct(ev_package), n_operations = round(mean(n_op, na.rm=TRUE), 2), percent_envir = ratio(n_distinct(cur_data()$src_ref[!is.na(cur_data()$envir)]),  n_call_sites), percent_envir_calls = ratio(sum(!is.na(envir)), n())) %>%
+  summarize(n_call_sites = n_distinct(src_ref), 
+            proportion = ratio(n_call_sites, nb_call_sites), 
+            n_packages = n_distinct(ev_package),
+            n_operations = round(mean(n_op, na.rm=TRUE), 2), 
+            percent_envir = ratio(n_distinct(cur_data()$src_ref[!is.na(cur_data()$envir)]),  n_call_sites), 
+            percent_envir_calls = ratio(sum(!is.na(envir)), n()), 
+            env_diversity = n_distinct(envir), 
+            percent_envir_non_parentframe = ratio(n_distinct(cur_data()$src_ref[!is.na(cur_data()$envir) & cur_data()$envir != "parent.frame()"]),  n_call_sites)) %>% 
   arrange(desc(n_call_sites))
@@
 ```{r}
-nrows <- 12
+nrows <- min(12, nrow(widespread_minimized)) # kaggle has less than 12 rows
 r_vec("MinimizedCallSites", nrows, pull(widespread_minimized, n_call_sites))
@@
 r_vec("MinimizedPercentEnvir", nrows, pull(widespread_minimized, percent_envir))
+r_vec("MinimizedPercentEnvirNonParent", nrows, pull(widespread_minimized, percent_envir_non_parentframe))
 ```
@@
 
+overview_table(
+  r("nb symbol var sites", nb_var_symb_sites),
+  r("nb symbol var site percent", ratio(nb_var_symb_sites, nb_var_min_sites)),
   r("nb non default envir variables percent", ratio(nb_var_symb_ndefault_sites, nb_var_symb_sites))
+)
 ```
@@
 ```
-
-Seems it was a bug: when the explicit argument was `parent.frame()`, it was counted as being the default argument (but it does not have the same semantics as a default argument or as an explicit one though!).
 
@@
 
-r("function definition sites", nb_fundef_sites)
-r("function definition sites percent", ratio(nb_fundef_sites, nb_call_sites))
-r("fundef non default envir sites percent", ratio(nb_ndefault_envir_fundef_sites, nb_fundef_sites))
-r("generalized function definition sites", nb_generalized_fundef_sites)
-r("generalized function definition sites percent", ratio(nb_generalized_fundef_sites, nb_call_sites))
+overview_table(
+  r("function definition sites", nb_fundef_sites),
+  r("function definition sites percent", ratio(nb_fundef_sites, nb_call_sites)),
+  r("fundef non default envir sites percent", ratio(nb_ndefault_envir_fundef_sites, nb_fundef_sites)),
+  r("generalized function definition sites", nb_generalized_fundef_sites),
+  r("generalized function definition sites percent", ratio(nb_generalized_fundef_sites, nb_call_sites)),
   r("generalized funded non default envir sites percent", ratio(nb_ndefault_envir_gen_fundef_sites, nb_generalized_fundef_sites))
+)
 ```
@@
 
-r("nb assign sites", nb_assigns_call_sites)
+overview_table(
+  r("nb assign sites", nb_assigns_call_sites),
   r("assign sites percent", ratio(nb_assigns_call_sites, nb_call_sites))
+)
 ```
@@
 
-r("non default envir assign sites", nb_a_env_call_sites)
+overview_table(
+  r("non default envir assign sites", nb_a_env_call_sites),
   r("non default envir assign sites percent", ratio(nb_a_env_call_sites, nb_assigns_call_sites))
+)
 ```
@@
 
-Now, we can show the mot common minimized expressions when there are more than 2 for one call site:
+With these minimized expressions:
+
+```{r}
+minimized_exprs
+```
+
+Now, we can show the mot common minimized expressions per site:
 
@@
 combination_ranking %>%
-  filter(cumpercent < 95) %>%
+  filter(cumpercent < if(nrow(combination_ranking) <= 10) 101 else 95) %>%
   ggplot(aes(x = fct_reorder(common_minimized, n), y = percent)) + 
@@
 
-r("Nb One Minimized", n_one_mini)
+overview_table(
+  r("Nb One Minimized", n_one_mini),
   r("Nb One Minimized Percent", ratio(n_one_mini, nb_call_sites))
+)
 ```
@@
 
-r("nb call sites unique actual value", nb_actual_values_sites)
-r("call sites unique actual value percent", ratio(nb_actual_values_sites, value_minimized$n_call_sites))
-r("median run sites unique actual value", median(only_actual_values$nb_runs, na.rm = TRUE))
-r("average run sites unique actual value", mean(only_actual_values$nb_runs, na.rm = TRUE))
+overview_table(
+r("nb call sites unique actual value", nb_actual_values_sites),
+r("call sites unique actual value percent", ratio(nb_actual_values_sites, value_minimized$n_call_sites)),
+r("median run sites unique actual value", median(only_actual_values$nb_runs, na.rm = TRUE)),
+r("average run sites unique actual value", mean(only_actual_values$nb_runs, na.rm = TRUE)),
 r("sd run sites unique actual value", sd(only_actual_values$nb_runs, na.rm = TRUE))
+)
 ```
@@
 
-r("val one node", n_one_node)
+overview_table(
+  r("val one node", n_one_node),
   r("val one node percent", ratio(n_one_node, n_val_exprs))
+)
 ```
@@
 
-r("one val median runs", med_runs)
-r("one val average runs", avg_runs)
+overview_table(
+  r("one val median runs", med_runs),
+  r("one val average runs", avg_runs),
   r("one val sd runs", sd_runs)
+)
 ```
@@
 
-r("non default envir value percent", ratio(nrow(non_default_envir), nrow(minimized_sites)))
+overview_table(
+  r("non default envir value percent", ratio(nrow(non_default_envir), nrow(minimized_sites))),
   r("non default envir with var percent", ratio(nrow(non_default_with_var), nrow(non_default_envir)))
+)
 ```
@@
 
-r("nb slot access", nb_slot_sites)
-r("slot access percent", ratio(nb_slot_sites, nb_call_sites))
-r("generalized slot access", nb_generalized_slot_sites)
-r("generalized slot access percent", ratio(nb_generalized_slot_sites, nb_call_sites))
-r("ndefault envir slot site", nb_ndefault_envir_slot_sites)
+overview_table(
+  r("nb slot access", nb_slot_sites),
+  r("slot access percent", ratio(nb_slot_sites, nb_call_sites)),
+  r("generalized slot access", nb_generalized_slot_sites),
+  r("generalized slot access percent", ratio(nb_generalized_slot_sites, nb_call_sites)),
+  r("ndefault envir slot site", nb_ndefault_envir_slot_sites),
   r("ndefault envir slot site percent", ratio(nb_ndefault_envir_slot_sites, nb_generalized_slot_sites))
+)
 ```
+
+# Generalized variable access
+
+```{r}
+gen_var <- E_cano %>% filter(minimized %in% c("X", "F(X)", "$"))
+nb_gen_var_sites <- gen_var %>% pull(src_ref) %>% n_distinct()
+
+overview_table(
+  r("generalized var access sites", nb_gen_var_sites),
+  r("generalized var access site percent", ratio(nb_gen_var_sites, nb_call_sites))
+)
+```

### 2021-03-29 analysis/environments.Rmd
Analysis

--- 2020-12-29 analysis/environments.Rmd
@@
 params:
-  base_dir: Data
-  calls_packages_path: /package/summarized.fst
+  base_dir: Data/package/
+  calls_path: summarized.fst
+  corpus_path: corpus.txt
+  static_path: evals-static.csv
+  force_rebuild: FALSE
 ---
@@
 source("inc/latextags.R")
-source("inc/setup.R")
+source("inc/functions.R")
 source("inc/paths.R")
 theme_set(theme_minimal())
-create_tags(path(TAGS_DIR, "environments.tex"), prefix = "", default = TRUE)
 
-source("inc/insights.R", local = knitr::knit_global())
+dataset_name <- basename(params$base_dir)
+
+# Tags will be prefixed by the dataset name
+create_tags(path(TAGS_DIR, paste0(dataset_name, "_environments.tex")), prefix = dataset_name, default = TRUE)
+# Also use ggsave_prefix instead of ggsave if you need to save a graph
+
+calls_path <- paste0(params$base_dir, params$calls_path)
+corpus_path <- paste0(params$base_dir, params$corpus_path)
+static_path <- paste0(params$base_dir, params$static_path)
+
+lock_path <- paste0(params$base_dir, ".environments-lock.fst")
+min_path <- paste0(params$base_dir, "Eenvir.fst")
+
+nb_eval_call_sites <- function(dataset) {
+  dataset %>% pull(src_ref) %>% n_distinct()
+}
+
 ```
@@
 ```{r loading_fst}
-eval_calls_path <- paste0(params$base_dir, params$calls_packages_path)
+tibble(package = read_lines(corpus_path)) -> C_raw
+read_csv(static_path) %>% semi_join(C_raw, by = "package") -> P
 
-eval_calls <- read_fst(eval_calls_path) %>%
-  as_tibble()
+rebuild <- TRUE
 
-stopifnot(!is.na(eval_calls$eval_call_srcref))
+E_envir <- NULL
+
+if (file.exists(lock_path)) {
+  saved <- read.fst(lock_path)[[1,1]]
+  if (saved == file.size(calls_path)) {
+    read_fst(min_path) %>% as_tibble() -> E_envir
+    rebuild <- FALSE
+  }
+}
+
+if(params$force_rebuild || rebuild) {
+  read_fst(calls_path) %>% as_tibble() -> E_raw
+
+  E_raw <- E_raw %>% mutate(envir_expression = if_else(envir_type == "NULL" & is.na(envir_expression), "NULL", envir_expression))
+
+  E_raw %>% select(
+    run_package = package, #      The package which was run to trigger the eval
+    ev_variant = eval_function, # Which eval variant was called
+    duplicates = nb_ev_calls, #   Number of identical eval calls (weight)
+    src_ref = eval_call_srcref, # Source ref for the eval call site
+    file, #                       Script file name (name of this run)
+    ev_package = eval_source, #   Package of the call site
+    type = expr_resolved_type_tag, #   Argument type
+    ev_expr = expr_expression,
+    n_op = interp_eval,
+    envir = envir_expression,
+    envir_type,
+    caller_expression,
+    environment_class
+  ) -> E_envir
+
+  E_envir %>% write_fst(min_path)
+  sz <- file.size(calls_path) %>% as_tibble()
+  write.fst(sz, lock_path)
+}
+
+C_raw -> C
+corpus_size <- nrow(C)
 ```
@@
 ```{r packages}
-nb_eval_calls <- count(eval_calls, wt = nb_ev_calls)
+nb_eval_calls <- count(E_envir, wt = duplicates)
 
-nb_call_sites <- nb_eval_call_sites(eval_calls)
+nb_call_sites <- E_envir %>% pull(src_ref) %>% n_distinct()
 ```
 
-There are `r count(eval_calls, wt=nb_ev_calls)$n` eval calls and `r eval_calls %>% select(eval_call_srcref) %>% n_distinct()` eval call sites in the dataset.
+There are `r nb_eval_calls` eval calls and `r nb_call_sites` eval call sites in the dataset.
 
@@
 ```{r envir_default}
-eval_calls_default <- eval_calls %>% filter(is.na(envir_expression))
+eval_calls_default <- E_envir %>% filter(is.na(envir))
 
-nb_default <- count(eval_calls_default, wt = nb_ev_calls)
+nb_default <- count(eval_calls_default, wt = duplicates)
 nb_default_sites <- nb_eval_call_sites(eval_calls_default)
 
-eval_calls_explicit <- eval_calls %>% filter(!is.na(envir_expression))
-nb_explicit <- count(eval_calls_explicit, wt = nb_ev_calls)
+eval_calls_explicit <- E_envir %>% filter(!is.na(envir))
+nb_explicit <- count(eval_calls_explicit, wt = duplicates)
 nb_explicit_sites <- nb_eval_call_sites(eval_calls_explicit)
+
+overview_table(
+  r("default envir sites", nb_default_sites),
+  r("defaut envir sites percent", ratio(nb_default_sites, nb_call_sites)),
+  r("non default envir sites", nb_explicit),
+  r("non defaut envir sites percent", ratio(nb_explicit, nb_call_sites))
+)
 ```
@@
 
+# Most common envir arguments
+
+We show the most frequent environment expression per site. There is only one environment expression per site, as the envir expression is captured before being resolved.
+
+```{r envir_most_common}
+frequent_envir <- E_envir %>% group_by(src_ref) %>% 
+  summarize(n_envir = n_distinct(envir), envir = paste0(unique(envir), collapse = " ; "))
+
+stopifnot(frequent_envir$n_envir <= 1)
+```
+
 # `envir` with Rcpp
@@
 
-nb_rcpp <- count(eval_calls_rcpp, wt = nb_ev_calls)
+nb_rcpp <- count(eval_calls_rcpp, wt = duplicates)
 nb_rcpp_sites <- nb_eval_call_sites(eval_calls_rcpp)
@@
 
-nb_env_list <- count(eval_calls_env_list, wt = nb_ev_calls)
+nb_env_list <- count(eval_calls_env_list, wt = duplicates)
 nb_env_list_sites <- nb_eval_call_sites(eval_calls_env_list)
@@
 
-nb_arg_caller <- count(eval_calls_arg_caller, wt = nb_ev_calls)
+nb_arg_caller <- count(eval_calls_arg_caller, wt = duplicates)
 nb_arg_caller_sites <- nb_eval_call_sites(eval_calls_arg_caller)
@@
 eval_calls_arg_caller %>%
-  count(envir_from_arg, wt = nb_ev_calls, sort = TRUE) %>%
+  count(envir_from_arg, wt = duplicates, sort = TRUE) %>%
   knitr::kable()
@@
 
-nb_absolute <- count(eval_calls_absolute, wt = nb_ev_calls)
+nb_absolute <- count(eval_calls_absolute, wt = duplicates)
 nb_absolute_sites <- nb_eval_call_sites(eval_calls_absolute)
@@
   filter(envir_distinct == 2) %>%
-  select(eval_call_expression, environment_class, envir_distinct)
+  select(ev_expr, environment_class, envir_distinct)
 ```
@@
 ```{r}
-env_calls %>% filter(!is.na(new_env), specific_env == "global") %>% count(new_env, wt = nb_ev_calls, sort = TRUE)
+env_calls %>% filter(!is.na(new_env), specific_env == "global") %>% count(new_env, wt = duplicates, sort = TRUE)
 
-env_calls %>% filter(new_env == 4, specific_env == "global") %>% count(eval_source, wt = nb_ev_calls, sort = TRUE)
+env_calls %>% filter(new_env == 4, specific_env == "global") %>% count(eval_source, wt = duplicates, sort = TRUE)
 ```
@@
 
-nb_integer <- count(eval_calls_integer, wt = nb_ev_calls)
+nb_integer <- count(eval_calls_integer, wt = duplicates)
 nb_integer_sites <- nb_eval_call_sites(eval_calls_integer)
@@
 
-nb_null <- count(eval_calls_null, wt = nb_ev_calls)
+nb_null <- count(eval_calls_null, wt = duplicates)
 nb_null_sites <- nb_eval_call_sites(eval_calls_null)

### 2021-03-29 analysis/patterns2.Rmd
Analysis

@@
 
+The diversity of non-default environments is higher than for other patterns.
+
 # `match.call`
@@
 ```{r}
-parse_eval <- E_cano %>% filter(type == "expressions" | str_detect(ev_expr, fixed("parse")))
+#parse_eval <- E_cano %>% filter(type == "expressions" | str_detect(ev_expr, fixed("parse")))
+parse_eval <- E_cano %>% filter(!is.na(parsed))
 
@@
 
-As a first approximation, it corresponds to type `language`.
+As a first approximation, it corresponds to type `language` and the usage of `substitute`.
+Indeed, using `quote` or `bquote` to get a language expression is the responsability of the caller of function, not of the function itself.
+On the contrary, with `substitute` and laziness, it is possible to capture the AST of an argument.
 
 ```{r}
-parse_eval <- E_cano %>% filter(type == "language")
+language_eval <- E_cano %>% filter(type == "language")
 
-nb_parse_eval <- parse_eval %>% pull(src_ref) %>% n_distinct()
+nb_language_eval <- language_eval %>% pull(src_ref) %>% n_distinct()
+
 ```
 
-## selecting columns in a dataframe
+There are `r nb_language_eval` sites with language expressions.
+
+We can already detect if substitute is present directly in the `eval` site:
+
+```{r}
+substitute_eval <- language_eval %>% filter(str_detect(ev_expr, fixed("substitute")))
+nb_substitute_eval_sites <- substitute_eval %>% pull(src_ref) %>% n_distinct()
+```
+
+There are `r nb_substitute_eval_sites` sites that have `eval` inside their non-resolved expression.
+
+To also detect the case when `substitute` is used before the call to `eval`, we can approximate it by looking whether a function with `eval` also contains a call to substitute before it.
+
+## Selecting columns in a dataframe
 
@@
 
-# Non-necessary `eval`
+#  `eval` anti-patterns
 
 ## Values into `eval`
+
+See normalized expression notebook.

### 2021-03-29 paper/main.tex
Update numbers for package normalized expressions
Improve description of normalized expr X

@@
 
-\subsection{The environment passed to \eval.}
+\subsubsection{The environment passed to \eval.}
 
@@
 		\hline
-		\texttt{X} & \packageMinimizedcallsitesa  & \packageMinimizedpropsitesa & \packageMinimizedpackagea & \packageMinimizedoperationsaRnd & \packageMinimizedpercentenvira \\
-		\hline
-		\texttt{F(F())} & \packageMinimizedcallsitesb  & \packageMinimizedpropsitesb & \packageMinimizedpackageb  & \packageMinimizedoperationsbRnd & \packageMinimizedpercentenvirb \\
-		\hline
-		\texttt{V} & \packageMinimizedcallsitesc & \packageMinimizedpropsitesc & \packageMinimizedpackagec & \packageMinimizedoperationscRnd & \packageMinimizedpercentenvirc \\
-		\hline
-		\texttt{F(X)}& \packageMinimizedcallsitesd & \packageMinimizedpropsitesd & \packageMinimizedpackaged & \packageMinimizedoperationsdRnd & \packageMinimizedpercentenvird \\
-		\hline
-		\texttt{model.frame} & \packageMinimizedcallsitese & \packageMinimizedpropsitese & \packageMinimizedpackagee & \packageMinimizedoperationseRnd & \packageMinimizedpercentenvire \\
-		\hline
-		\texttt{F()}& \packageMinimizedcallsitesf & \packageMinimizedpropsitesf & \packageMinimizedpackagef & \packageMinimizedoperationsfRnd & \packageMinimizedpercentenvirf \\
-		\hline
-		\texttt{\$}& \packageMinimizedcallsitesg & \packageMinimizedpropsitesg & \packageMinimizedpackageg & \packageMinimizedoperationsgRnd & \packageMinimizedpercentenvirg \\
-		\hline
-		\texttt{FUN} & \packageMinimizedcallsitesh & \packageMinimizedpropsitesh & \packageMinimizedpackageh & \packageMinimizedoperationshRnd & \packageMinimizedpercentenvirh \\
-		\hline
-		\texttt{<-} & \packageMinimizedcallsitesi  & \packageMinimizedpropsitesi & \packageMinimizedpackagei & \packageMinimizedoperationsiRnd & \packageMinimizedpercentenviri \\
-		\hline
-		\texttt{\{BLOCK\}} & \packageMinimizedcallsitesj & \packageMinimizedpropsitesj & \packageMinimizedpackagej & \packageMinimizedoperationsjRnd & \packageMinimizedpercentenvirj \\
-		\hline
-		\texttt{F(F(<-))} & \packageMinimizedcallsitesk & \packageMinimizedpropsitesk & \packageMinimizedpackagek & \packageMinimizedoperationskRnd & \packageMinimizedpercentenvirk \\
+		\texttt{X} & \packageMinimizedcallsitesa  & \packageMinimizedpropsitesa & \packageMinimizedpackagea & \packageMinimizedoperationsaRnd & \packageMinimizedpercentenvirnonparenta \\
+		\hline
+		\texttt{F(F())} & \packageMinimizedcallsitesb  & \packageMinimizedpropsitesb & \packageMinimizedpackageb  & \packageMinimizedoperationsbRnd & \packageMinimizedpercentenvirnonparentb \\
+		\hline
+		\texttt{V} & \packageMinimizedcallsitesc & \packageMinimizedpropsitesc & \packageMinimizedpackagec & \packageMinimizedoperationscRnd & \packageMinimizedpercentenvirnonparentc \\
+		\hline
+		\texttt{F(X)}& \packageMinimizedcallsitesd & \packageMinimizedpropsitesd & \packageMinimizedpackaged & \packageMinimizedoperationsdRnd & \packageMinimizedpercentenvirnonparentd \\
+		\hline
+		\texttt{model.frame} & \packageMinimizedcallsitese & \packageMinimizedpropsitese & \packageMinimizedpackagee & \packageMinimizedoperationseRnd & \packageMinimizedpercentenvirnonparente \\
+		\hline
+		\texttt{F()}& \packageMinimizedcallsitesf & \packageMinimizedpropsitesf & \packageMinimizedpackagef & \packageMinimizedoperationsfRnd & \packageMinimizedpercentenvirnonparentf \\
+		\hline
+		\texttt{\$}& \packageMinimizedcallsitesg & \packageMinimizedpropsitesg & \packageMinimizedpackageg & \packageMinimizedoperationsgRnd & \packageMinimizedpercentenvirnonparentg \\
+		\hline
+		\texttt{FUN} & \packageMinimizedcallsitesh & \packageMinimizedpropsitesh & \packageMinimizedpackageh & \packageMinimizedoperationshRnd & \packageMinimizedpercentenvirnonparenth \\
+		\hline
+		\texttt{<-} & \packageMinimizedcallsitesi  & \packageMinimizedpropsitesi & \packageMinimizedpackagei & \packageMinimizedoperationsiRnd & \packageMinimizedpercentenvirnonparenti \\
+		\hline
+		\texttt{\{BLOCK\}} & \packageMinimizedcallsitesj & \packageMinimizedpropsitesj & \packageMinimizedpackagej & \packageMinimizedoperationsjRnd & \packageMinimizedpercentenvirnonparentj \\
+		\hline
+		\texttt{F(F(<-))} & \packageMinimizedcallsitesk & \packageMinimizedpropsitesk & \packageMinimizedpackagek & \packageMinimizedoperationskRnd & \packageMinimizedpercentenvirnonparentk \\
 		\hline
@@
 \#sites is the number of sites; \%sites the percentage of sites across the total number of sites (\packageNbCallSites sites) with at least this minimized expression; \#packages is the number of packages with at least one of the minimized expression and measures how widespread a minimized expression is.
-\#operations is the average number of steps performed by the R interpreter when given that minimized expression, and \%envir is the percentage of call sites for which the \c{envir} argument is explicitly passed.} \label{tab:minimizedexpressions}
+\#operations is the average number of steps performed by the R interpreter when given that minimized expression, and \%envir is the percentage of call sites for which the \c{envir} argument is explicitly passed, and is not \c{parent.frame()}, which is the most common explicit environment passed to \eval.} \label{tab:minimizedexpressions}
 \end{table}
@@
 The call sites without an explicit \c{envir} argument all use a specific \eval variant, \c{eval.parent}.
-
-Similarly, when a function is evaluated on some variable (\c{F(X)} minimized expression), the percentage of non default environment is rather high, \packageMinimizedpercentenvird\%.
-
-Minimized expression \texttt{\$} refers to a slot access, with \texttt{\$}, \texttt{[]} or \texttt{[[]]}, as the top call in a simple expression, for instance \texttt{a\$slot + 1} and  It amounts for \packageNbSlotAccess sites, \ie \packageSlotAccessPercent of the sites.
+Moreover, the \c{envir} argument is both explicit and different from \c{parent.frame()} in \packageMinimizedpercentenvirnonparenta\% of the sites, which suggests that when given a variable, \eval is often used to get  value in a non-local, non-parent environment.
+
+Similarly, when a function is evaluated on some variable (\c{F(X)} minimized expression), the percentage of non default, non parent environments is rather high, \packageMinimizedpercentenvirnonparentd\%.
+
+Minimized expression \texttt{\$} refers to a slot access, with \texttt{\$}, \texttt{[]} or \texttt{[[]]}, as the top call in a simple expression, for instance \texttt{a\$slot + 1}.
+It amounts for \packageNbSlotAccess sites, \ie \packageSlotAccessPercent of the sites.
+The non-default, non-parent environment percentage is here \packageMinimizedpercentenvirnonparentg.
 

### 2021-03-30 paper/main.tex
Package environment

@@
 We observe that a disproportionately high number of core R \eval calls
-access the caller's caller's environment. This is because many core R
+access the  caller's environment. This is because many core R
 functions call functions that pass the result of \c{parent.frame()} to
@@
 \end{lstlisting}
+
+\paragraph{Touching package environments.} It refers to the \emph{namespace} of a package.
+A call to \c{library(packageName)}, which load libraries in R, for instance, will populate environment \c{package:packageName}.
+It is a \emph{base} function though, so we do not see it in the CRAN dataset.
+However, R still makes it possible to access any package environments and \packageNbPackageEnvSites call sites, \ie \packagePackageEnvSitePercent access a package environment, from \packageNbPackageEnvPackages different packages.
+Those packages perform introspection, to represent and search any environment by name, in package \emph{envnames}, to containerize tests, as with \emph{testthat} and \emph{unitizer}, to create a custom OOP system with \emph{R.oo}, or to create poor man's hidden functions in \emph{USSCXenaTools}.
+
+It is also used to implement a custom caching mechanism in packages \emph{g.data} and \emph{SOAR}.
 

### 2021-03-31 paper/main.tex
wip

--- 2021-03-29 paper/main.tex
@@
-%\documentclass[USenglish,cleveref, autoref, thm-restate]{lipics-v2019}
-\documentclass[acmsmall]{acmart}
+\documentclass[screen,acmsmall]{acmart}
+\settopmatter{printfolios=true,printccs=true,printacmref=true}
+
+\bibliographystyle{ACM-Reference-Format}
+\citestyle{acmauthoryear}   %% For author/year citations
 \usepackage{listings,hyperref,multirow,paralist,xspace,url,wrapfig}
@@
 
-\title{Why do we Eval?\\\Large A large-scale study of Eval usage in R}
-%\titlerunning{Why Eval (in R)?}
+\title{What we {\sf eval} in the shadows}
+\subtitle{A large-scale study of {\sf eval} usage in R programs}
+
+%% Aviral Goel
+%% Pierre Donat-Bouillud
+%% Filip Krikava
+%% Jan Vitek
+
+%% What is the right order?
 
@@
 \begin{lstlisting}
- eval
-   <- function(expr,
-                envir = parent.frame(),
-                enclos = if(is.list(envir)) parent.frame() else baseenv())
-          { ...body... }
+ eval <- function(expr, envir = parent.frame(),
+           enclos = if(is.list(envir)) parent.frame() else baseenv()) {...body...}
 \end{lstlisting}
@@
 \begin{itemize}
-\item \c{env$v} and \c{get("v",envir=env)} read variable \c{v} in
+\item \c{env\$v} and \c{get("v",envir=env)} read variable \c{v} in
   environment \c{env};
-\item \c{env$v<-2} and \c{assign("v",2,envir=env)} write 2 in
+\item \c{env\$v<-2} and \c{assign("v",2,envir=env)} write 2 in
   \c{v}. If not found, \c{v} is created in \c{env}.
@@
 analysis. We ignore calls to the native \eval function exposed by R. We also
-ignore the \c{rlang::tidy_eval} function which uses native \eval internally
+ignore the \c{rlang::tidy\_eval} function which uses native \eval internally
 because \c{rlang} is used to implement a DSL for data analysis in R. It
 introduces a new first-class promise object called \c{quosure} for which it
-implements special evaluation support in \c{tidy_eval}.
+implements special evaluation support in \c{tidy\_eval}.
 
@@
 be exact, come from a single call in the \c{ggplot2} package and have
-have the value \c{_inherit}. Even if we remove this outlier, the
+have the value \c{\_inherit}. Even if we remove this outlier, the
 majority of \evals are simple expressions.
@@
 
-We looked at the environments passed to all the \eval calls in our
-corpus. Table~\ref{tab:environments} summarizes the results. A numeric
-environment class \c{n} denotes the environment of the $n$-th call
-stack frame from the current function. \c{global} denotes the
-top-level environment and \c{list} denotes a list passed for
-evaluation of formulas. Environment classes of the form $n+$ denote
-the $n$-th environment extended with a new environment. The new
-environment provides a limited form of sandboxing. All assignments
-using the \c{\<-} function occur inside it and prevent the extended
-environment from mutation. However, it is still possible to mutate the
-extended environment using the \c{\<\<-} or \c{assign} functions; but,
-that happens rarely.
+We looked at the environments passed to all the \eval calls in our corpus.
+Table~\ref{tab:environments} summarizes the results. A numeric environment class
+\c{n} denotes the environment of the $n$-th call stack frame from the current
+function. \c{global} denotes the top-level environment and \c{list} denotes a
+list passed for evaluation of formulas. Environment classes of the form $n+$
+denote the $n$-th environment extended with a new environment. The new
+environment provides a limited form of sandboxing. All assignments using the
+$\leftarrow$ function occur inside it and prevent the extended environment from
+mutation. However, it is still possible to mutate the extended environment using
+the $\leftarrow$ or \c{assign} functions; but, that happens rarely.
 
@@
 modify images using a dedicated formula syntax using
-\c{\~}.\footnote{Inspired by {map} in package \emph{purr}.} Here,
+\~.\footnote{Inspired by {map} in package \emph{purr}.} Here,
 \eval is evaluated in \c{newenv}, which creates a new environment that
@@
 \c{column[rows] <<- what} &                        55677    & 3.5\%\\
-\c{function(value) freduce(value, `_function_list`)} & 37251& 2.3\%\\
+\c{function(value) freduce(value, `function.list`)} & 37251& 2.3\%\\
 \c{NULL} &                         32005    & 2\%\\
@@
 \c{force(..1)}            &                        20461    & 1.3\%\\
-\c{alist(`_spec`)}       &                         18532    & 1.2\%\\
+\c{alist(`spec`)}       &                         18532    & 1.2\%\\
 \c{inner}               &                          18530    & 1.2\%\\
@@
 The expression \c{Environment} occurs because of the four callsites
-explained above, \c{ggplot2::ggproto}, \c{R6::generator_funs},
+explained above, \c{ggplot2::ggproto}, \c{R6::generator\_funs},
 \c{future::backtrace} and \c{RModel::str.RMmodel}. The next
@@
 contribute to a single side-effect. The expression \c{function(value)
-  freduce(value, `_function_list`)} arises from the
+  freduce(value, `function.list`)} arises from the
 \c{magrittr::\%>\%} function which is a pipe operator that pipes the
@@
 using the \c{\%>\%} function into the next expression. The \c{NULL}
-arises from \c{R6::generator_funs} function when the \eval is passed a
-\c{NULL} argument by the \c{DataMask_generator} package. The
+arises from \c{R6::generator\_funs} function when the \eval is passed a
+\c{NULL} argument by the \c{DataMask\_generator} package. The
 \c{c("default", "default2012", "default2011" ...} pattern arises from
@@
 for its formal parameter and evaluates it. The \c{force(..1)} and
-\c{alist(`_spec`)} patterns occur in \c{glue::glue_data} function
+\c{alist(`spec`)} patterns occur in \c{glue::glue\_data} function
 which concatenates and interpolates strings. The two patterns occur
@@
 forces promises and returns the result of evaluation. The \c{inner}
-pattern arises from \c{glue::identity_transformer} which enables the
+pattern arises from \c{glue::identity\_transformer} which enables the
 creation of custom transformation functions for affecting the

### 2021-03-31 analysis/environments.Rmd

@@
     type = expr_resolved_type_tag, #   Argument type
+    ev_call = eval_call_expression,
     ev_expr = expr_expression,
@@
 frequent_envir <- E_envir %>% group_by(src_ref) %>% 
-  summarize(n_envir = n_distinct(envir), envir = paste0(unique(envir), collapse = " ; "))
+  summarize(n_envir = n_distinct(envir), envir = paste0(unique(envir), collapse = " ; "), envir_type = paste0(unique(envir_type), collapse = " ; "))
 
-stopifnot(frequent_envir$n_envir <= 1)
+#stopifnot(frequent_envir$n_envir <= 1)
 ```
 
-# `envir` with Rcpp
+```{r}
+freq_envir <- frequent_envir %>% count(envir, sort = TRUE) %>% 
+  mutate(percent = round(100 * n / nb_call_sites, 2), cum_percent = cumsum(percent))
 
-```{r envir_rcpp}
-eval_calls_rcpp <- eval_calls_explicit %>% filter(str_detect(expr_expression, "C\\+\\+Constructor|C\\+\\+Field|C\\+\\+OverloadedMethods"))
+nb_parent.frame_sites <- freq_envir %>% filter(envir == "parent.frame()") %>% .$n
+nb_data_sites <- freq_envir %>% filter(envir == "data") %>% .$n
 
-nb_rcpp <- count(eval_calls_rcpp, wt = duplicates)
-nb_rcpp_sites <- nb_eval_call_sites(eval_calls_rcpp)
+overview_table(
+  r("nb parent france sites", nb_parent.frame_sites),
+  r("nb parent france site percent", ratio(nb_parent.frame_sites, nb_call_sites)),
+  r("nb data sites", nb_data_sites),
+  r("nb data sites", ratio(nb_data_sites, nb_call_sites))
+)
+```
 
-eval_calls_explicit <- eval_calls_explicit %>% anti_join(eval_calls_rcpp)
+```{r}
+freq_envir %>% filter(cum_percent <= 90) %>%
+  ggplot() +
+  geom_col(aes(x = fct_reorder(envir, percent), y = percent)) + 
+  coord_flip()
+
+ggsave_prefix("envir_expressions.pdf")
+```
+
+Let's classify the expressions, between calls and variables:
+
+```{r}
+envir_var <- frequent_envir %>% filter(envir != "NA", str_detect(envir, "^(?:\\.[:alpha]|[:alpha:])[[:alpha:][:digit:]_\\.]*$"))
+
+nb_envir_var_sites <- nrow(envir_var)
+
+overview_table(
+  r("envir variable sites", nb_envir_var_sites),
+  r("envir variable site percent", ratio(nb_envir_var_sites, nb_explicit_sites))
+)
+```
+
+What about we classify them by type?
+
+```{r}
+frequent_envir %>% count(envir_type, sort = TRUE) %>% datatable()
 ```
@@
 nb_env_list_sites <- nb_eval_call_sites(eval_calls_env_list)
+
+overview_table(
+  r("nb env list", nb_env_list),
+  r("nb env list sites", nb_env_list_sites),
+  r("nb env list site percent", ratio(nb_env_list_sites, nb_call_sites))
+)
 ```
@@
 
-```{r envir_arg_caller}
+```{r envir_arg_caller, eval = FALSE}
 eval_calls_arg_caller <- eval_calls_explicit %>% filter(!is.na(envir_from_arg))
@@
 
-```{r}
+```{r eval = FALSE}
 eval_calls_arg_caller %>%
@@
 eval_calls_absolute <- eval_calls_explicit %>%
-  anti_join(eval_calls_env_list) %>%
-  anti_join(eval_calls_arg_caller)
+  anti_join(eval_calls_env_list) #%>%
+  #anti_join(eval_calls_arg_caller)
 
@@
 nb_absolute_sites <- nb_eval_call_sites(eval_calls_absolute)
+
+overview_table(
+  r("nb absolute envir", nb_absolute$n),
+  r("nb absolute sites", nb_absolute_sites),
+  r("nb absolute site percent", ratio(nb_absolute_sites, nb_eval_calls))
+)
 ```
@@
 ```{r}
-eval_calls_absolute <- eval_calls_absolute %>%
-  group_by(eval_call_srcref) %>%
-  mutate(envir_distinct = n_distinct(environment_class))
+absolute_diversity <- eval_calls_absolute %>%
+  group_by(src_ref) %>%
+  summarize(envir_distinct = n_distinct(environment_class), envir = paste0(environment_class, collapse = " ; "))
 
-eval_calls_absolute %>%
-  ungroup() %>%
+absolute_diversity %>%
   count(envir_distinct, sort = TRUE)
-```
-
-We can also see a few examples:
-
-```{r}
-eval_calls_absolute %>%
-  filter(envir_distinct == 2) %>%
-  select(ev_expr, environment_class, envir_distinct)
 ```
@@
 
+```{r}
+env_calls %>% group_by(new_env) %>% summarize(n = n_distinct(src_ref), percent = round(100 * n / nb_call_sites, 2), cum_percent = cumsum(percent)) %>% arrange(desc(n))
+
+env_calls %>% group_by(hierarchy) %>% summarize(n = n_distinct(src_ref), percent = round(100 * n / nb_call_sites, 2), cum_percent = cumsum(percent)) %>% arrange(desc(n))
+
+```
+
 _ggplot2_ again is responsible for this `new.env` 4 times!
@@
 
-env_calls %>% filter(new_env == 4, specific_env == "global") %>% count(eval_source, wt = duplicates, sort = TRUE)
+env_calls %>% filter(new_env == 4, specific_env == "global") %>% count(ev_package, wt = duplicates, sort = TRUE)
+```
+
+# Specific environments
+
+```{r}
+specific_env <- env_calls %>% group_by(specific_env) %>% summarize(n = n_distinct(src_ref), n_packages = n_distinct(ev_package), percent = round(100 * n / nb_call_sites, 2), cum_percent = cumsum(percent)) %>% arrange(desc(n))
+
+nb_global_env_sites <- specific_env %>% filter(specific_env == "global") %>% .$n
+nb_callee_env_sites <- specific_env %>% filter(specific_env == "callee") %>% .$n
+nb_empty_env_sites <- specific_env %>% filter(specific_env == "empty") %>% .$n
+nb_base_env_sites <- specific_env %>% filter(specific_env == "base") %>% .$n
+
+overview_table(
+  r("nb global env sites", nb_global_env_sites),
+  r("nb global env site percent", ratio(nb_global_env_sites, nb_call_sites)),
+  r("nb callee env sites", nb_callee_env_sites),
+  r("nb callee env site percent", ratio(nb_callee_env_sites, nb_call_sites)),
+  r("nb empty env sites", nb_empty_env_sites),
+  r("nb empty env site percent", ratio(nb_empty_env_sites, nb_call_sites)),
+  r("nb base env sites", nb_base_env_sites),
+  r("nb base env site percent", ratio(nb_base_env_sites, nb_call_sites))
+)
+```
+
+## Package environments
+
+Our tracer leaks here: 3 srcref are seeing `package:evil`, from `test_that`, `R.oo` and `envnames`.
+
+What are the ones looking into packages?
+
+```{r}
+package_env <- E_envir %>% filter(str_detect(environment_class, fixed("package:"))) %>% count(ev_package, sort = TRUE)
+
+nb_package_env_sites <- E_envir %>% filter(str_detect(environment_class, fixed("package:"))) %>% pull(src_ref) %>% n_distinct()
+
+overview_table(
+  r("nb package env packages", nrow(package_env)),
+  r("nb package env sites", nb_package_env_sites),
+  r("package env site percent", ratio(nb_package_env_sites, nb_call_sites))
+)
+```
+
+They are the following:
+
+- _R.oo_: OOP system
+- _USSCXenaTools_: uses `eval` to create _hidden_ function inside its own namespace (by pasting `".p_"` in front of the function name)
+- _unitizer_: unit test framework that requires to precise the test environment
+- _envnames_: seach env by names, give a name for each environments. Traverse all the call sites with `sys.calls`
+- _testthat_: unit test framework
+
+## Weird packages
+
+There are 3 weird _package_ environments: `.R_Cache`, `.R_Test`, `newdir...`
+
+They all seem to implement a kind of caching mechanism.
+
+They build the environment name with:
+
+```{r, eval = FALSE}
+env <- attach(NULL, pos, basename(directory))
 ```
@@
 
-nb_integer <- count(eval_calls_integer, wt = duplicates)
+nb_integer <- count(eval_calls_integer, wt = duplicates)$n
 nb_integer_sites <- nb_eval_call_sites(eval_calls_integer)
+
+overview_table(
+  r("nb integer envir", nb_integer),
+  r("nb integer envir sites", nb_integer_sites),
+  r("nb integer envir site percent", ratio(nb_integer_sites, nb_call_sites))
+)
 ```
@@
 
-nb_null <- count(eval_calls_null, wt = duplicates)
+nb_null <- count(eval_calls_null, wt = duplicates)$n
 nb_null_sites <- nb_eval_call_sites(eval_calls_null)
+
+overview_table(
+  r("nb null envir", nb_null),
+  r("nb null envir sites", nb_null_sites),
+  r("nb null envir site percent", ratio(nb_null_sites, nb_eval_calls))
+)
 ```
@@
 
-```{r summary}
+```{r summary, eval = FALSE}
 summary <- tribble(

### 2021-03-31 paper/main.tex

@@
 \subsubsection{The expression passed to \eval.}
+\paragraph{Touching package environments.} It refers to the \emph{namespace} of a package.
+A call to \c{library(packageName)}, which load libraries in R, for instance, will populate environment \c{package:packageName}.
+It is a \emph{base} function though, so we do not see it in the CRAN dataset.
+However, R still makes it possible to access any package environments and \packageNbPackageEnvSites call sites, \ie \packagePackageEnvSitePercent access a package environment, from \packageNbPackageEnvPackages different packages.
+Those packages perform introspection, to represent and search any environment by name, in package \emph{envnames}, to containerize tests, as with \emph{testthat} and \emph{unitizer}, to create a custom OOP system with \emph{R.oo}, or to create poor man's hidden functions in \emph{USSCXenaTools}.
+
+It is also used to implement a custom caching mechanism in packages \emph{g.data} and \emph{SOAR}.
 

### 2021-03-31 paper/main.tex
related

@@
 
-\subsection{R in a nutshell}
-
-R is a lazy functional programming language~\cite{ecoop12} with
-dynamic features that allow to write object-oriented code. Most data
-types are vectorized. Values are constructed by the \c{c(...)}
-function, e.g. \c{c("a","bc")} creates a vector of two strings. To
-enable equational reasoning, R copies values accessible through
-multiple variables when they are written to. Values can be tagged by
-user-defined attributes. For instance, one can attach the attribute
-\c{dim} to the value \c{x<-c(1,2,3,4)} with \c{attr(x,"dim")<-c(2,2)}.
-This causes arithmetic functions to treat \c x as a $2 \times 2$
-matrix. Another attribute is \c{class} which can be bound to a list of
-names, e.g., \c{class(x)<-"human"}. This sets the class of \c{x} to
-\c{human}; classes are used for object-oriented dispatch. Every R
-linguistic construct is desugared to a function call, even control
-flow statements, assignments, and bracketing. Furthermore, all
-functions can be redefined in user code. This makes R both flexible
-and challenging to compile~\cite{dls19}. Arguments to user-defined
-functions are bundled into thunks called \emph{promises}. Logically, a
-promise combines an expression's code, its environment, and its value.
-To access the value of a promise, one must force it. Forcing a promise
-triggers evaluation and the computed value is captured for future
-reference.
-
-\subsection{Eval Semantics}
-
-The expressive power of \eval depends on design decisions along two
-axes:
-\begin{itemize}
-\item {\bf Scoping:} What environment does \eval executes its argument
-  in? JavaScript and R evaluate it in the current environment, thus
-  exposing local variables and parameters and breaking the caller's
-  abstraction boundary. Julia is more restrictive, \eval runs at the
-  ``top level'', in the global environment. JavaScript has a global
-  \eval that behaves like Julia, and a ``strict mode'' in which \eval
-  may access local variables, but fobidden from injecting new ones in
+\subsection{R, Briefly}
+
+\citet{ecoop12} gave a programming language-centric overview of the R language.
+They characterized it as a lazy, vectorized, functional language with rich
+dynamic features that allow layering several object systems. Most data types are
+sequences of primitive values constructed by calling the combination function
+\c{c}, so that \c{c(1,2)} evaluates to a vector of two strings, and constants
+such as \c{42} are vectors of length one. To enable equational reasoning, values
+accessible through multiple aliases are copied when written to. Furthermore,
+values can be tagged by user-defined attributes, key-value pairs. For instance,
+the attribute \c{dim}-\c{c(2,2)} can be attached to the value at \c{x} by the
+call \c{attr(x,"dim")<-c(2,2)}. In this case, the addition of this attribute
+ensure that the vector \c x will be treated as a $2 \times 2$ matrix by all
+arithmetic operations. The \c{class} attribute is used to give a 'class', in the
+object-oriented sense, to values. So, \c{class(x)<-"human"} sets the class of
+\c{x} to \c{human}; classes are used for method dispatch. Every linguistic
+construct is desugared to a function call, even control flow statements,
+assignments, and bracketing. All functions can be shadowed or redefined, making
+the language at the same time remarkably flexible and exceeding challenging to
+compile statically as vividly detailed by~\citet{dls19}. R uses a relaxed call-by-need convention for
+passing arguments to functions. Each argument is a thunk composed of an
+expression, its environment, and a slot for the result, these are called
+\emph{promises}. To get the value of an argument, the corresponding promise must
+be forced, once forced the promise's result is cached for future use.
+
+\subsection{The Semantics of Eval and Friends}
+
+While a facility to transform data into code is available in many languages,
+some design choices affect the expressive power of \eval and related functions.
+The key design choices that we review here are the input format, the environment
+in which \eval-generated code is evaluated and the reflective operations that
+are available to that code. Fig.~\ref{comp} summarizes design choices for
+several popular languages.
+
+\begin{figure}[!t]\center\small
+\begin{tabular}{r|l|l|l}\hline
+\tiny\sc Language&\sc\tiny Input&\sc\tiny Scope&\tiny\sc Reflective operations\\\hline
+\bf Julia     & expression& toplevel         & ??\\
+\bf Java      & bytecode  & classloader       & data\\
+\bf JavaScript& text      & current, toplevel& data\\
+\bf R         & expression& programmatic      & data, stack, envir.\\\hline
+\end{tabular}
+\caption{Design space of \eval}\label{comp}
+\end{figure}
+
+The input to \eval can be any format that can be turned into code. Both Julia
+and R mandate that the input format be parsed expressions. JavaScript allows any
+expression as a character string. Java is the most restrictive as a classloader
+only accepts complete classes in the bytecode format recognized by the virtual
+machine.
+
+The environment in which \eval executes the code can be the lexically enclosing
+environment of \eval (current), the global environment (toplevel), a limited
+global environment (classloader), or an environment chosen by the caller of
+\eval (programmatic). The most restrictive variant is that of Java, where the
+scope of newly loaded code is the set of classes exposed by the current class
+loader. Julia is the next most restrictive, as only global definitions can be
+observed and modified by \eval. JavaScript has a variant of \eval that runs in
+the current scope and a most strict mode that is restricted to global variables.
+Finally, R allows the caller to choose which environment to use with \eval.
+
+WORK IN  PROGREESS
+
+JavaScript and R evaluate it in the current environment, thus exposing local
+variables and parameters and breaking the caller's abstraction boundary. Julia
+is more restrictive, \eval runs at the ``top level'', in the global environment.
+JavaScript has a global \eval that behaves like Julia, and a ``strict mode'' in
+which \eval may access local variables, but fobidden from injecting new ones in
 the local environment. In Java, one can implement \eval with
-  Julia-semantics.\footnote{While \eval used to be the purview of
-  interpreted languages, just-in-time compilation lifted this
-  restriction. \Eval takes its input, wraps it in a static method of a
-  new anonymous class, generates bytecode for that class, invokes
-  the class loader to install that code, and finally reflectively
-  calls the method. }
-\item {\bf Reflective API:} What other reflective operations are
+Julia-semantics.\footnote{While \eval used to be the purview of interpreted
+languages, just-in-time compilation lifted this restriction. \Eval takes its
+input, wraps it in a static method of a new anonymous class, generates bytecode
+for that class, invokes the class loader to install that code, and finally
+reflectively calls the method. }
+
+ {\bf Reflective API:} What other reflective operations are
   exposed to user code? JavaScript provides few reflective functions
@@
   \eval.
-\end{itemize}
 

### 2021-04-01 paper/main.tex
=> 2.3

@@
 the attribute \c{dim}-\c{c(2,2)} can be attached to the value at \c{x} by the
-call \c{attr(x,"dim")<-c(2,2)}. In this case, the addition of this attribute
-ensure that the vector \c x will be treated as a $2 \times 2$ matrix by all
-arithmetic operations. The \c{class} attribute is used to give a 'class', in the
-object-oriented sense, to values. So, \c{class(x)<-"human"} sets the class of
-\c{x} to \c{human}; classes are used for method dispatch. Every linguistic
-construct is desugared to a function call, even control flow statements,
-assignments, and bracketing. All functions can be shadowed or redefined, making
-the language at the same time remarkably flexible and exceeding challenging to
-compile statically as vividly detailed by~\citet{dls19}. R uses a relaxed call-by-need convention for
-passing arguments to functions. Each argument is a thunk composed of an
-expression, its environment, and a slot for the result, these are called
-\emph{promises}. To get the value of an argument, the corresponding promise must
-be forced, once forced the promise's result is cached for future use.
-
-\subsection{The Semantics of Eval and Friends}
+call \c{attr(x,"dim")$\leftarrow$c(2,2)}. In this case, the addition of this
+attribute ensure that the vector \c x will be treated as a $2 \times 2$ matrix
+by all arithmetic operations. The \c{class} attribute is used to give a 'class',
+in the object-oriented sense, to values. So, \c{class(x)<-"human"} sets the
+class of \c{x} to \c{human}; classes are used for method dispatch. Every
+linguistic construct is desugared to a function call, even control flow
+statements, assignments, and bracketing. All functions can be shadowed or
+redefined, making the language at the same time remarkably flexible and
+exceeding challenging to compile statically as vividly detailed
+by~\citet{dls19}. R uses a relaxed call-by-need convention for passing arguments
+to functions. Each argument is a thunk composed of an expression, its
+environment, and a slot for the result, these are called \emph{promises}. To get
+the value of an argument, the corresponding promise must be forced, once forced
+the promise's result is cached for future use.
+
+\subsection{On the Expressive Power of Eval}
 
@@
 \begin{figure}[!t]\center\small
-\begin{tabular}{r|l|l|l}\hline
-\tiny\sc Language&\sc\tiny Input&\sc\tiny Scope&\tiny\sc Reflective operations\\\hline
-\bf Julia     & expression& toplevel         & ??\\
-\bf Java      & bytecode  & classloader       & data\\
-\bf JavaScript& text      & current, toplevel& data\\
-\bf R         & expression& programmatic      & data, stack, envir.\\\hline
+\begin{tabular}{r@{~}l|l|l|l}\hline
+\tiny\sc Language&&\sc\tiny Input&\sc\tiny Scope&\tiny\sc Reflective operations\\\hline
+\bf Julia&\cite{julia}     & expression& toplevel         & data\\
+\bf Java&\cite{cl}  & bytecode  & classloader       & data\\
+\bf JavaScript&\cite{ecoop11}& text      & current, toplevel& data\\
+\bf R&\cite{R96}  & expression& programmatic      & data, stack, envir.\\\hline
 \end{tabular}
@@
 
-WORK IN  PROGREESS
-
-JavaScript and R evaluate it in the current environment, thus exposing local
-variables and parameters and breaking the caller's abstraction boundary. Julia
-is more restrictive, \eval runs at the ``top level'', in the global environment.
-JavaScript has a global \eval that behaves like Julia, and a ``strict mode'' in
-which \eval may access local variables, but fobidden from injecting new ones in
-the local environment. In Java, one can implement \eval with
-Julia-semantics.\footnote{While \eval used to be the purview of interpreted
-languages, just-in-time compilation lifted this restriction. \Eval takes its
-input, wraps it in a static method of a new anonymous class, generates bytecode
-for that class, invokes the class loader to install that code, and finally
-reflectively calls the method. }
-
- {\bf Reflective API:} What other reflective operations are
-  exposed to user code? JavaScript provides few reflective functions
-  (other than enumeration of an object's properties and string-based
-  indexing), the impact of \eval is thus limited. R, on the other
-  hand, has a rich reflective API. For example, R allows user code to
-  walk the call stack and arbitrarily add and delete local variables.
-  This means that there are few limits to the side-effects of an
-  \eval.
-
-\noindent
-These design choices impact our ability to reason about code when
-reflection is used. A more expressive \eval translates directly into
-additional restrictions for program analysis and transformation tools.
-The reason Julia restricts \eval to the top level is to preserve the
-compiler's ability to optimize code~\cite{oopsla18a}. In particular,
-\eval had to be prevented from changing the type of local variables as
-this would require recompilation of the method. Julia even adds a
-versoning mechanism, called world age, to ensure that code added
-during an \eval does not invalidate inlining
-optimizations~\cite{oopsla20a}. More permissive languages such as R
-are much harder to deal with from a compiler's perspective.
-
-\mypara{Eval in R.} The reflective interface exposed by
-R is somewhat complex. The core library exports four functions with
-slightly different semantics, \c{eval}, \c{evalq}, \c{eval.parent},
-and \c{local}. \Eval is the more general function, and, unlike in
-JavaScript, it takes three arguments, an expression, its environment
-and the enclosing environment. The following discussion simplify some
-details which are not relevant to this paper. For the interested
-reader we recommend the excellent book by Hadley
-Wickham~\cite{hadley}. The definition of \eval starts as follows:
+The last degree of freedom in the design of \eval is what can be done from
+within the generated code. The main difference between designs lies in what part
+of the system state can be accessed and modified. In R, there is very little
+that cannot be observed from an \eval context and modified. The code can
+override the definition of variables and functions. Also, thanks to a rich
+reflective interface, code can introspect on values, walk and modify environment
+chains, and access the stack frames on the call stacks and their environments.
+
+Given the above, it is reasonable to claim that R's \eval are amongst the
+languages that afford the facility the most expressive power. This comes at the
+cost of the programmer's ability to reason about code in the presence of \eval.
+Expressive power translates directly into limits on the usefulness of program
+analysis.
+
+By contrast, the designers of Julia chose to limit \eval. In Julia, only global
+variables can be side-effected and environments cannot be readily manipulated.
+This is designed to shield optimized code from some of the most pernicious uses
+of the facility~\cite{oopsla18a}. Furthermore, the language provides a
+versioning mechanism, called world age, to ensure that any methods defined
+within an \eval only become visible at well-defined program points and thus that
+recompilation does not have to occur when optimized code is
+running~\cite{oopsla20a}.
+
+\subsection{Eval in R}
+
+The R language exposes a rich reflective interface with four functions with
+different semantics, \c{eval}, \c{evalq}, \c{eval.parent}, and
+\c{local}. The first is the most general, taking three arguments, an expression,
+its environment and an enclosing environment. The following discussion
+simplifies details not relevant to this paper, the interested reader should
+consult \citet{hadley}. It is defined as:
 
 \begin{lstlisting}
- eval <- function(expr, envir = parent.frame(),
-           enclos = if(is.list(envir)) parent.frame() else baseenv()) {...body...}
+ eval <- function(e, envir = parent.frame(),
+                   encl = if(is.list(envir)) parent.frame() else baseenv()) ...
 \end{lstlisting}
-The parameters are \c{expr}, the value to be evaluated, \c{envir}, the
-environment in which evaluation happens, and \c{enclos}, the
-environment to look up objects not found in \c{envir} (by default
-either the top level, \c{baseenv} or the environment of the caller).
-The \c{expr} parameter can take values of different type, for our
-purposes we focus on the most common, {\tt expression}s. It is easiest
-to think of an \c{expression} as an abstract syntax tree as returned
-by R's parser. The simplest way to create an \c{expression} is to call
-\c{quote} passing some R expression:
+Parameter \c e is the expression to evaluate, \c{envir} is the evaluation
+environment, and \c{encl} is used to look up variables not found in \c{envir}.
+The latter is, by default, either the environment of the function that called
+\eval or the top level. Parameter \c e can be of different types, for our
+purposes we focus on {\tt expression}s. These can be thought of as abstract
+syntax trees returned by the parser. An \c{expression} can be obtained by
+calling \c{quote} with an expression or \c{parse} wit a string:
 \begin{lstlisting}
@@
  # a + b
-\end{lstlisting}
-Here the return value is an abstract syntax tree. From a
-string, use  the \c{parse} function:
-\begin{lstlisting}
  > parse(text="a+b")
@@
 \end{lstlisting}
-The most common way to create an expression is to extract it from an
-argument. Each argument comes packaged inside a promise which retains
-the source code of the argument. Consider the following function
-definition, \c{f} is a function with a single parameter \c{x}. When
-the call \c{f(a+b)} is evaluated, \c{a+b} is stored inside a newly
-created promise.
+Even more commonly, expressions are obtained from promises using \c{substitute}.
+Consider a one-parameter function \c{f}, when called with \c{a+b}, the function
+returns \c{a+b} as expression.
 \begin{lstlisting}
-  > f <- function(x) substitute(x), list(a=1)))
+ > f <- function(x) substitute(x)
  > f(a+b)
-  # expression(1+b)
+ # expression(a+b)
  > substitute(a+b,list(a=1,b=2))
@@
 \end{lstlisting}
-The call to \c{substitute(x)} extracts, from the promise bound to
-\c{x}, the expression passed in the call to \c{f}, i.e., \c{a+b}. The
-\c{substitute(exp,env)} function takes two arguments, the second is an
-object that can be used as an environment. This can be either a list
-of named values, a data frame or an actual environment. \c{substitute}
-will look for occurrences of each symbol from \c{env} in \c{exp} and
-replace them with their value taken from \c{env}.
+The call to \c{substitute(x)} extracts the unevaluated expression from the
+promise bound to variable \c x. Function \c{substitute(e,envir)} takes two
+arguments, the second is used as a substitution environment for free variables
+in \c e.
 \begin{lstlisting}
@@
 \end{lstlisting}
-The above shows how to access the current environment, this can belong
-to the current function or, as above, the top level. Environments are
-nested, with each environment having a parent. This nesting is used
-when looking up names. When calling \c{new.env}, the default parent is
-the current environment. The chain of environments can be traversed
-with \c{parent.env}, ultimately one arrives at \c{emptyenv}.
-\c{parent.frame} and \c{sys.frame} grant access to environments
-further up in the calling stack, up to the global environment
-\c{.GlobalEnv}, and packages. One can also directly read, modify or
-create new bindings, given any environment:
+The \c{environment} function returns a variable environment, either from the
+currently executing function or the top level environment. Environments nest,
+each has a parent. When creating a new environment with \c{new.env}, the parent
+is the current environment. Environment chains can be traversed with
+\c{parent.env}, until \c{emptyenv} is reached. The top level environment is
+\c{.GlobalEnv}, it has parents that represent the packages that have been
+loaded. One can also directly read, modify or create new bindings, given any
+environment:
 \begin{itemize}
-\item \c{env\$v} and \c{get("v",envir=env)} read variable \c{v} in
-  environment \c{env};
-\item \c{env\$v<-2} and \c{assign("v",2,envir=env)} write 2 in
-  \c{v}. If not found, \c{v} is created in \c{env}.
+\item \c{envir\$v} and \c{get("v",envir=envir)}: read  \c{v} from \c{envir};
+\item \c{envir\$v<-2} and \c{assign("v",2,envir=envir)}: store 2 in \c{v}, if
+  not found, \c{v} is created in \c{envir}.
 \end{itemize}
-\noindent Environments are often used as hashtables by programmers as
-they have reference semantics (other are copy-on-write) and have a
-built-in string look up. across function calls, or to create package
-namespaces.
+\noindent Programmers sometimes use environments as hash maps as they are the
+only data type with reference semantics and have a built-in string look up.
+Functions \c{parent.frame} and \c{sys.frame} return environments further up the
+call stack.
  \begin{lstlisting}
-  evalq <- function(exp,env,enclos) eval(quote(exp),env,enclos)
-  eval.parent <- function(exp,n) eval(exp,parent.frame(n))
-  local <- function(exp,env,enclos) evalq(exp,new.env())
+  evalq <- function(e,envir,encl) eval(quote(e),envir,encl)
+  eval.parent <- function(e,n) eval(e,parent.frame(n))
+  local <- function(e,envir,encl) evalq(e,new.env())
 \end{lstlisting}
-The three \eval variants can be expressed as calls to \eval. The
-\c{evalq} form quotes the argument to prevent it from being evaluated
-in the current environment. The \c{eval.parent(e,n)} form evaluates
-\c{e} in the environment of the {\tt n}-th caller of this function.
-Finally, \c{local} evaluates an \c{exp} in a new environment to
-avoid polluting the current one.
+The three \eval variants can be expressed with \eval. The \c{evalq} form quotes
+the argument to prevent it from being evaluated in the current environment. The
+\c{eval.parent(e,n)} form evaluates \c{e} in the environment of the {\tt n}-th
+caller of this function. Finally, \c{local} evaluates \c{e} in a new environment
+to avoid polluting the current one.
 

### 2021-04-01 paper/main.tex
minor

@@
 
-\mypara{Intercession.} A common use case for \eval is to be combined
-with \c{match.call}. \c{match.call} walks up the call stack, captures
-the code that invoked the currently executing function, and returns it
-as an unevaluated expression. The pattern is to transform a call to
-some function \c{f} into a call to \c{g} with some arguments retained
-and others modified. As an illustration, consider the \c{vcpart}
-package's function \c{tvcglm} that is translated to a call to \c{tvcm}
-with two modifications to the argument list: argument \c{control}
-can't be missing and \c{fit} is set  \c{"glm"}. The
-function ends with a call to \c{eval.parent} to ensure that the
-rewritten call is evaluted in the same environment the original call
-was.
+\mypara{Intercession.} A common use case for \eval is to be combined with
+\c{match.call}. \c{match.call} walks up the call stack, captures the code that
+invoked the currently executing function, and returns it as an unevaluated
+expression. The pattern is to transform a call to some function \c{f} into a
+call to \c{g} with some arguments retained and others modified. As an
+illustration, consider the \c{vcpart} package's function \c{tvcglm} that is
+translated to a call to \c{tvcm} with two modifications to the argument list:
+argument \c{control} can't be missing and \c{fit} is set \c{"glm"}. The function
+ends with a call to \c{eval.parent} to ensure that the rewritten call is
+evaluted in the same environment the original call was.
 \begin{lstlisting}
- tvcglm <- function(formula, data, family, control=tvcglm_control(), ...) {
-   Call <- match.call()
-   Call[[1L]] <- as.name("tvcm")
-   if (!"control" %in% names(Call)) Call$control<-formals(tvcglm)$control
-   if ("fit" %in% names(list(...))) warning("'fit' is ignored. ")
-   Call$fit <- "glm"
-   return(eval.parent(Call))
+ tvcglm <- function(formula, data, family, ...) {
+   k <- match.call()
+   k[[1]] <- as.name("tvcm")
+   if ("fit" %in% names(list(...))) warning("'fit' ignored.")
+   k$fit <- "glm"
+   eval.parent(k)
  }
@@
 
-\mypara{Code Generation.} The more traditional use of \eval is to
-execute code that was assembled by the programmer into a string. Here
-we show the method \c{plot} for class \c{sback} in package
-\c{wsbackfit}, simplified for explanatory purposes. The function takes
-a long argument list, the names of which are captured in the list
-\c{opt}. The string \c{stub} is composed of a subset of the arguments
-passed to this function; the variable is used to construct a call to
-\c{base::plot} which will draw a plot. \NOTE{This example is a bit
-  messed up. What is var? Does the use of ... in the eval grab the
-  arguments of this call? If yes does it not end up with xlab twice?
-  HELP} The \parse and \eval combination is used by the the \source
-function in the base library to load R code from a file in the current
-workspace.
+\mypara{Code Generation.} The more traditional use of \eval is to execute code
+that was assembled by the programmer into a string. Here we show the method
+\c{plot} for class \c{sback} in package \c{wsbackfit}, simplified for
+explanatory purposes. The function takes a long argument list, the names of
+which are captured in the list \c{opt}. The string \c{stub} is composed of a
+subset of the arguments passed to this function; the variable is used to
+construct a call to \c{base::plot} which will draw a plot. \NOTE{This example is
+  a bit messed up. What is var? Does the use of ... in the eval grab the
+  arguments of this call? If yes does it not end up with xlab twice? HELP} The
+\parse and \eval combination is used by the the \source function in the base
+library to load R code from a file in the current workspace.
 \begin{lstlisting}
@@
     ifelse("xlab" %in% opt,"",paste(",xlab=\"",var,"\"",sep="")),
-    ifelse("main" %in% opt,"",main.aux),
-    ifelse("type" %in% opt,"",",type=\"l\""),
-    sep = "")
+    ifelse("type" %in% opt,"",",type=\"l\""), sep="")
    plot <- paste("plot(x.data,",stub,",...)",sep="")
    eval(parse(text=plot))
-   ...
 \end{lstlisting}
@@
 \begin{lstlisting}
-  opts = c("datatable.verbose"="FALSE", # ... many others
+  opts = c("datatable.verbose"="FALSE", # ...many others
   for (i in names(opts))

### 2021-04-01 paper/main.tex
more

@@
 
-This section explains our methodology for selecting the corpus that
-will be analyzed and how we implemented our dynamic analysis.
+This section explains how we selected our corpus and how we obtained the
+reported results.
 
@@
   It receives about 6 new package submissions a day~\cite{Ligges2017}. \cf
-  \url{https://cran.r-project.org/web/packages/index.html}}) and the other from
+  \url{https://cran.r-project.org}}) and the other from
 the \emph{Kaggle} data science competition website\footnote{\cf
@@
 
-% Clearly code coverage will vary; as is the case for any dynamic analysis, the
-% quality of results is predicated on how representative the executions we can
-% observe are with respect to all possible runs of the corpus. To mitigate the
-% threat to validity attached to limited coverage, we obtain clients that will
-% generate multiple \emph{runs} of CRAN packages. In this context, a run is the
-% execution of a CRAN package or Kaggle program with a particular set of input
-% values.
-
-% \begin{figure}[!h]\centering\includegraphics[width=.7\linewidth]
-%   {corpus.pdf}\caption{CRAN packages}\label{fig:corpus}
-% \end{figure}
-
-\mypara{CRAN Packages.} We used \CranAllPackages packages. These are almost all
-of the available packages except for a few that we could not install due to
-missing native dependencies. The packages are implemented in R with native C and
-Fortran code. Using \c{cloc}, we measure a total \CranAllRCodeRnd lines of R
-code and \CranAllNativeCodeRnd lines of native code. Out of these packages,
-\CranPackages (\CranPackagesRatio) call eval. These packages represent
-\CranCodeRatio of the code size.
-
-To analyze a package, we need client code that invokes its methods. There are
-three sources of built-in runnable code that come with each CRAN package:
-\emph{tests, examples} and \emph{vignettes}. They are, respectively, traditional
-unit tests, code snippets from the documentation, and long-form use-cases
-written in \LaTeX\xspace or Markdown with executable snippets of R code.
-Examples and vignettes are automatically extracted and turned into scripts,
-their input is bundled with the package. Tests are directly runnable. Together
-we there are \CranRunnableScripts scripts that contain \CranRunnableCode lines
-of code (\CranRunnableCodeExamplesRnd in examples, \CranRunnableCodeVignettesRnd
-in vignettes and \CranRunnableCodeTestsRnd in tests).
-
-\mypara{Kaggle Scripts.} Kaggle is an online platform for data-science and
-machine-learning. The website allows people to submit data-analysis problems,
-users compete to find the best solution. The solutions, called \emph{kernels},
-are uploaded to the platform as either plain scripts or notebooks. We obtained
-\KaggleKernels kernels. Since Kaggle is not a curated repository, we first had
-to remove duplicates. We used SHA-1 hash to identify \KaggleDuplicates
-whole-file copies. The remaining \KaggleUnique kernels represent unique
-solutions to \KaggleCompetitions Kaggle competition. Together, they contain
-\KaggleCode lines of R code. Out of the \KaggleUniqueRnd kernels, only
-\KaggleWithEvals contain calls to eval.
-
-\subsection{Analysis Pipeline}
-
-The results presented in this paper are the result of an automated analysis
-pipeline that acquires the code of packages, extract metadata, executes
-programs, traces their behavior and summarizes the observations.
-Figure~\ref{fig:pipeline} shows the main steps of the pipeline along with
-approximate time to execute each step, the data size, and the number of elements
-manipulated by the stage. Timings are for runs on a cluster of three servers
-with Intel Xeon 6140, 2.30GHz with 72 cores and 256GB of RAM using a shared OCFS
-network storage.
+\mypara{Packages.} We obtained \CranAllPackages packages from CRAN, only
+excluding packages that did not install due to missing native dependencies.
+Using \c{cloc}, we measured \CranAllRCodeRnd lines of R and
+\CranAllNativeCodeRnd lines of native code. Out of these packages, \CranPackages
+(\CranPackagesRatio) call eval. These packages represent \CranCodeRatio of the
+code size. To analyze a package, we need client code that invokes its methods.
+There are three sources of built-in runnable code that come with each CRAN
+package: \emph{tests, examples} and \emph{vignettes}. They are, respectively,
+traditional unit tests, code snippets from the documentation, and long-form
+use-cases written in \LaTeX\xspace or Markdown with executable snippets of R
+code. Examples and vignettes are automatically extracted and turned into
+scripts, their input is bundled with the package. Tests are directly runnable.
+Together we there are \CranRunnableScripts scripts that contain
+\CranRunnableCode lines of code (\CranRunnableCodeExamplesRnd in examples,
+\CranRunnableCodeVignettesRnd in vignettes and \CranRunnableCodeTestsRnd in
+tests).
+
+\mypara{Scripts.} Kaggle is an online platform for data science. It allows users
+to submit problems and compete to solve them. Solutions, called \emph{kernels},
+are uploaded as plain scripts or notebooks. We obtained \KaggleKernels kernels.
+Since Kaggle is not curated, used SHA-1 hashes to identify and remove
+\KaggleDuplicates duplicate entries. The remaining \KaggleUnique kernels are
+unique solutions to \KaggleCompetitions competition. They contain \KaggleCode
+lines of R code. Only \KaggleWithEvals kernels  call \eval.
+
+\subsection{Pipeline}
+
+The analysis presented in this has been automated by a pipeline that acquires
+package code, extract metadata, executes programs, traces their behavior and
+summarizes observations. Figure~\ref{fig:pipeline} shows the main steps of
+analysis along with their running time, data size, and number of elements
+manipulated. Timings are from a cluster of three 2.3GHz Intel Xeon 6140 servers,
+with 72 cores and 256GB of RAM, and shared OCFS network storage.
 
@@
   \includegraphics[width=1.05\linewidth]{pipeline.pdf}
-  \caption{Analysis Pipeline UPDATE!!}\label{fig:pipeline}
+  \caption{Pipeline UPDATE!!}\label{fig:pipeline}
 \end{figure}
 
-\noindent The pipeline has the following steps:
+\noindent The pipeline steps are:
 \begin{compactenum}
@@
     CRAN packages and acquiring Kaggle corpus. We try to install all available
-  CRAN packages. This is somewhat complicated by the fact that many of packages
-  depend on native libraries, but these dependencies are not explicitly stated
-  in the package description and thus it is hard to automate. For Kaggle we use
-  a web crawler to get the programs and the a Kaggle official command line tool
-  to download the datasets.
+    CRAN packages. This is somewhat complicated by the fact that many of
+    packages depend on native libraries, but these dependencies are not
+    explicitly stated in the package description and thus it is hard to
+    automate. For Kaggle we use a web crawler to get the programs and the a
+    Kaggle official command line tool to download the datasets.
   \item \emph{Runnable code extraction.} Next step is to extract runnabale code.
@@
   \item \emph{Tracing.} With the scripts ready, we run our dynamic analysis that
-  traces the calls to eval and captures other runtime information. For the CRAN
-  packages we run two configurations:
+    traces the calls to eval and captures other runtime information. For the
+    CRAN packages we run two configurations:
   \begin{inparaenum}[(1)]
@@
   \end{inparaenum}
-  The reason for this split is that the core R packages are intertwined with the
-  language implementation and the usage is different form the user written
+    The reason for this split is that the core R packages are intertwined with
+    the language implementation and the usage is different form the user written
     packages hosted on CRAN. For Kaggle, we only trace the eval calls that
-  originate in the programs themselves. To avoid any interference, each program
-  is run in isolation, in its own R instance.
+    originate in the programs themselves. To avoid any interference, each
+    program is run in isolation, in its own R instance.
   \item \emph{Analysis.} In the final step, the output from the dynamic analysis
     is merged, cleaned and summarized in the post-processing phase by custom R
-  scripts. The summarized data is then analyzed in RMarkdown notebooks to gather
-  insights. Apart from the figures, the data points included in the paper are
-  also generated by RMarkdown notebooks as \LaTeX macros.
+    scripts. The summarized data is then analyzed in RMarkdown notebooks to
+    gather insights. Apart from the figures, the data points included in the
+    paper are also generated by RMarkdown notebooks as \LaTeX macros.
 \end{compactenum}
 
-The code extraction and tracing steps of the pipeline are run in parallel on
-three servers using GNU parallel~\cite{GNUparallel}. To make sure, all three
-servers have identical environment, we use docker images that have all the
-dependencies installed. The pipeline is implemented in 500 lines of Makefile and
-3K lines of auxiliary R and shell scripts. Their main function is to gracefully
-handle and report failures across large-scale program runs which greatly aids
-debugging of the pipeline.
-
-\subsection{Dynamic Analysis}
+\medskip\noindent
+The code extraction and tracing steps of the pipeline are run in
+parallel~\cite{GNUparallel}. Servers have identical environment thanks to docker
+images with all dependencies installed. The pipeline consists of 3.5K lines of
+Make, R and shell scripts.
+
+\subsection{Analysis}
 
@@
 the corresponding functions at runtime. For the latter we use
-\rdyntrace~\cite{oopsla19a}. It is a modified R virtual machine based on
-GNU R 4.0.2 that exposes a number of hooks allowing us to get low-level
-information about the R interpreter. \rdyntrace only works with the R
-interpreter and not the byte-code compiler. However, this does not affect the
-result since the compiler does not optimize eval calls.
+\rdyntrace~\cite{oopsla19a}. It is a modified R virtual machine based on GNU R
+4.0.2 that exposes a number of hooks allowing us to get low-level information
+about the R interpreter. \rdyntrace only works with the R interpreter and not
+the byte-code compiler. However, this does not affect the result since the
+compiler does not optimize eval calls.
 
@@
 
-\mypara{Limitations.}
-%
-% We turn off the bytecode compiler for this study. The bytecode
-% compiler can also call \eval.
-%
-We do not get source locations for \CranUndefinedRnd \eval calls. In
-these cases \eval is either passed as an argument to a higher-order functions or
-is defined in a function returned by a higher-order function, or it can be a
-generated called from package native code. However, this is a meager
-\CranUndefinedRatio of all \eval calls and is unlikely to affect our
-analysis. We ignore calls to the native \eval function exposed by R. We also
-ignore the \c{rlang::tidy\_eval} function which uses native \eval internally
-because \c{rlang} is used to implement a DSL for data analysis in R. It
-introduces a new first-class promise object called \c{quosure} for which it
-implements special evaluation support in \c{tidy\_eval}.
+\mypara{Limitations.} We do not get source locations for \CranUndefinedRnd \eval
+calls. In these cases \eval is either passed as an argument to a higher-order
+functions or is defined in a function returned by a higher-order function, or it
+can be a generated called from package native code. However, this is a meager
+\CranUndefinedRatio of all \eval calls and is unlikely to affect our analysis.
+We ignore calls to the native \eval function exposed by R. We also ignore the
+\c{rlang::tidy\_eval} function which uses native \eval internally because
+\c{rlang} is used to implement a DSL for data analysis in R. It introduces a new
+first-class promise object called \c{quosure} for which it implements special
+evaluation support in \c{tidy\_eval}.
 

### 2021-04-01 paper/main.tex
methodology

@@
 
-Our corpus is assembled out of two set of programs, one consisting of libraries
-obtained from \emph{CRAN} (Comprehensive R Archive Network\footnote{CRAN is the
-  largest repository of R code with almost \CranAvailablePackagesRnd packages.
-  It receives about 6 new package submissions a day~\cite{Ligges2017}. \cf
-  \url{https://cran.r-project.org}}) and the other from
-the \emph{Kaggle} data science competition website\footnote{\cf
-  \url{https://kaggle.com}}. The intent here is to contrast code written by
-experienced developers (CRAN) with code authored by typical end users of the
-language (Kaggle). Another difference is that CRAN is a curated repository. Each
-submitted package must abide to a number of well-formedness rules that are
-automatically checked asserting certain quality. Most relevant for this work is
-that all of the runnable code is tested and only a successfully running package
-is admitted in the archive. Kaggle on the other hand is more like GitHub, anyone
-can submit any code. All the code in our corpus is runnable, \ie we have scripts
-that invoke the CRAN packages and the Kaggle programs have input data.
-
-\mypara{Packages.} We obtained \CranAllPackages packages from CRAN, only
-excluding packages that did not install due to missing native dependencies.
-Using \c{cloc}, we measured \CranAllRCodeRnd lines of R and
-\CranAllNativeCodeRnd lines of native code. Out of these packages, \CranPackages
-(\CranPackagesRatio) call eval. These packages represent \CranCodeRatio of the
-code size. To analyze a package, we need client code that invokes its methods.
-There are three sources of built-in runnable code that come with each CRAN
-package: \emph{tests, examples} and \emph{vignettes}. They are, respectively,
-traditional unit tests, code snippets from the documentation, and long-form
-use-cases written in \LaTeX\xspace or Markdown with executable snippets of R
-code. Examples and vignettes are automatically extracted and turned into
-scripts, their input is bundled with the package. Tests are directly runnable.
-Together we there are \CranRunnableScripts scripts that contain
-\CranRunnableCode lines of code (\CranRunnableCodeExamplesRnd in examples,
-\CranRunnableCodeVignettesRnd in vignettes and \CranRunnableCodeTestsRnd in
-tests).
-
-\mypara{Scripts.} Kaggle is an online platform for data science. It allows users
-to submit problems and compete to solve them. Solutions, called \emph{kernels},
-are uploaded as plain scripts or notebooks. We obtained \KaggleKernels kernels.
+Our corpus is assembled from three sources. Firstly, we start with the
+\emph{Base} libraries that are bundled with the language implementation.
+Secondly, we acquire packages hosted on \emph{CRAN}. Thirdly, we downloaded
+scripts from \emph{Kaggle}.
+
+\mypara{Base.}
+???
+
+\mypara{CRAN.} The Comprehensive R Archive
+Network,\footnote{\url{cran.r-project.org}} or CRAN, is the largest curated
+repository of R packages. It hosts \CranAvailablePackagesRnd packages with 6 new
+packages submitted each day~\cite{Ligges2017}. Packages that are accepted to
+CRAN are authored by experienced developers and they must abide to a number of
+well-formedness rules automatically checked on each commit. Furthermore, each
+package comes with sample data and tests. There are three sources of runnable
+code for each package: \emph{tests, examples} and \emph{vignettes} --
+respectively, unit tests, executable code snippets from the documentation, and
+long-form executable use-cases written in Markdown. Examples and vignettes can
+be turned into scripts by extracting the relevant code. We obtained
+\CranPackages that call \eval. These packages represent \CranPackagesRatio of
+all packages or \CranCodeRatio of the entire code size of CRAN. From these we
+extracted \CranRunnableScripts scripts that contain \CranRunnableCode lines of
+code (\CranRunnableCodeExamplesRnd in examples, \CranRunnableCodeVignettesRnd in
+vignettes and \CranRunnableCodeTestsRnd in tests).
+
+\mypara{Scripts.} Kaggle is an online platform for data
+science.\footnote{\url{kaggle.com}} It allows users to submit problems and
+compete to solve them. Solutions, called \emph{kernels}, are uploaded as plain
+scripts or notebooks. The quality of code on Kaggle is not uniform. Each kernel
+is runnable, and thus there is one script per kernel. Input data is provided.
+We obtained \KaggleKernels kernels.
 Since Kaggle is not curated, used SHA-1 hashes to identify and remove

### 2021-04-01 paper/main.tex
minor

@@
 scripts or notebooks. The quality of code on Kaggle is not uniform. Each kernel
-is runnable, and thus there is one script per kernel. Input data is provided.
-We obtained \KaggleKernels kernels.
-Since Kaggle is not curated, used SHA-1 hashes to identify and remove
-\KaggleDuplicates duplicate entries. The remaining \KaggleUnique kernels are
-unique solutions to \KaggleCompetitions competition. They contain \KaggleCode
-lines of R code. Only \KaggleWithEvals kernels  call \eval.
+is runnable, and thus there is one script per kernel. Input data is provided. We
+obtained \KaggleKernels kernels. Since Kaggle is not curated, used SHA-1 hashes
+to identify and remove \KaggleDuplicates duplicate entries. The remaining
+\KaggleUnique kernels are unique solutions to \KaggleCompetitions competition.
+They contain \KaggleCode lines of R code. Only \KaggleWithEvals kernels call
+\eval.
 
@@
 The analysis presented in this has been automated by a pipeline that acquires
-package code, extract metadata, executes programs, traces their behavior and
-summarizes observations. Figure~\ref{fig:pipeline} shows the main steps of
-analysis along with their running time, data size, and number of elements
-manipulated. Timings are from a cluster of three 2.3GHz Intel Xeon 6140 servers,
-with 72 cores and 256GB of RAM, and shared OCFS network storage.
+packages, extract scripts, executes them, traces their behavior and summarizes
+observations. Figure~\ref{fig:pipeline} shows the main steps along with their
+running time, data size, and number of elements manipulated. Timings are from a
+cluster of three 2.3GHz Intel Xeon 6140 servers, with 72 cores and 256GB of RAM,
+and shared OCFS network storage.
 

### 2021-04-01 paper/main.tex
3.3

@@
 \noindent The pipeline steps are:
+\medskip
 \begin{compactenum}
-  \item \emph{Download.} The first step consists of downloading and installing
-    CRAN packages and acquiring Kaggle corpus. We try to install all available
-    CRAN packages. This is somewhat complicated by the fact that many of
-    packages depend on native libraries, but these dependencies are not
-    explicitly stated in the package description and thus it is hard to
-    automate. For Kaggle we use a web crawler to get the programs and the a
-    Kaggle official command line tool to download the datasets.
-  \item \emph{Runnable code extraction.} Next step is to extract runnabale code.
-    For CRAN package we use \genthat~\cite{issta18}, a tool which converts
-    runnable code snippets from package documentation and vignettes as well as R
-    tests into standalone executable scripts. For Kaggle programs that are not
-    already R scripts, we use \c{rmarkdown} and \c{knitr} R packages to extract
-    the runnable code from the Rmarkdown or the Jupyter notebook formats
-    respectively. Duplicated Kaggle programs are removed. Each of the extracted
-    script is wrapped into a call to our dynamic analyzer so we only record eval
-    usage triggered by running th script. Without this, the data would include
-    eval calls from the unit testing frameworks as well as from bootstrapping R
-    virtual machine itself.
-  \item \emph{Tracing.} With the scripts ready, we run our dynamic analysis that
-    traces the calls to eval and captures other runtime information. For the
-    CRAN packages we run two configurations:
-  \begin{inparaenum}[(1)]
-    \item all of the \CranRunnableScriptsRnd scripts while tracing calls to eval
-      originating in any of the \CranAllPackages installed CRAN package, and
-    \item a random selection of 1/4 of the scripts while tracing only eval calls
-      originating in the core R packages that come bundled with the language.
-  \end{inparaenum}
-    The reason for this split is that the core R packages are intertwined with
-    the language implementation and the usage is different form the user written
-    packages hosted on CRAN. For Kaggle, we only trace the eval calls that
-    originate in the programs themselves. To avoid any interference, each
-    program is run in isolation, in its own R instance.
-  \item \emph{Analysis.} In the final step, the output from the dynamic analysis
-    is merged, cleaned and summarized in the post-processing phase by custom R
-    scripts. The summarized data is then analyzed in RMarkdown notebooks to
-    gather insights. Apart from the figures, the data points included in the
-    paper are also generated by RMarkdown notebooks as \LaTeX macros.
+\item \emph{Download.} Packages are downloaded automatically from CRAN. For
+  scripts, a web crawler retrieves code and the Kaggle command line tool
+  downloads data. Installation is complicated by native dependencies which
+  are not properly documented and thus hard to automatically resolve.
+\item \emph{Extract.} Given installed packages and scripts, the next step is to
+  create runnable programs. The \genthat tool helpfully extracts all runnable
+  code snippets from a package and turns each of these into a self-standing
+  program~\cite{issta18}. Some Kaggle kernels are already scripts, for those
+  nothing more needs be done. Others kernels are packages as notebooks, either
+  as Rmarkdown or Jupyter, for those we use \c{knitr} to extract runnable code.
+  The body of each extracted program is instrumented with calls our dynamic
+  analyzer to ensure that we only record calls to \eval from the code of
+  interest and not from bootstrapping or execution harness operations.
+\item \emph{Trace.} Each program is executed using our dynamic analysis tool
+  which is heavily instrumented interpreter that captures calls to \eval and
+  many other fine-grained runtime events. Packages are run twice, once to
+  capture \eval calls originating from package code, and a second time to
+  capture calls coming from the base libraries. To avoid any interference, each
+  program is run in its own process.
+\item \emph{Analyze.} Finally, analysis output is merged, cleaned and summarized
+  in a post-processing phase driven by series of R scripts. The summarized data
+  is then analyzed in RMarkdown notebooks to gather insights. All figures and
+  numbers appearing in the paper are generated automatically. Figures are
+  produced in PDF by \c{ggplot2}, numbers are exported as \LaTeX macros.
 \end{compactenum}
@@
 
-For this study we need two things: information about calls to eval and related
-functions and low-level data such as object allocation and deallocation and
-side-effects that happen during eval. The former we can get by injecting code to
-the corresponding functions at runtime. For the latter we use
-\rdyntrace~\cite{oopsla19a}. It is a modified R virtual machine based on GNU R
-4.0.2 that exposes a number of hooks allowing us to get low-level information
-about the R interpreter. \rdyntrace only works with the R interpreter and not
-the byte-code compiler. However, this does not affect the result since the
-compiler does not optimize eval calls.
-
-We use source location to uniquely identify eval call sites. R only retains
-source references at the top-level block expressions. A nested call such as
-\c{try(eval(x))} does not have a source reference for the \c{eval(x)}. To work
-around this, we impute source locations for eval call sites at runtime, by
-traversing functions' AST and attaching source reference attribute manually.
-
-The dynamic analysis is implemented as a R package in 1.5K lines of R code and
-2K lines of C++.
-
-\mypara{Limitations.} We do not get source locations for \CranUndefinedRnd \eval
-calls. In these cases \eval is either passed as an argument to a higher-order
-functions or is defined in a function returned by a higher-order function, or it
-can be a generated called from package native code. However, this is a meager
-\CranUndefinedRatio of all \eval calls and is unlikely to affect our analysis.
-We ignore calls to the native \eval function exposed by R. We also ignore the
-\c{rlang::tidy\_eval} function which uses native \eval internally because
-\c{rlang} is used to implement a DSL for data analysis in R. It introduces a new
-first-class promise object called \c{quosure} for which it implements special
-evaluation support in \c{tidy\_eval}.
+The dynamic analysis is performed by \rdyntrace, a modified R virtual machine
+based on GNU R 4.0.2 that exposes low-level callbacks for a variety of runtime
+events~\cite{oopsla19a}. As the GNU R bytecode compiler is written in R, we turn
+it off to avoid recording \eval in its code, and fallback on the interpreter.
+This study requires recording a large number of events, calls to \eval,
+side-effects, function entries and exits, environments, etc. One challenge we
+encountered during the analysis was that R does not produce source references
+for top-level blocks. That is to say, a method whose body would \c{eval(x)}
+would not have debug information that we can use to identify that particular
+\eval. The same expression surrounded by braces would have a line number and
+source reference. This is unfortunately not easily fixed in the R
+implementation. Instead, we extended the dynamic analysis tool to add missing
+source references heuristically.
+
+The analysis has some limitation. Even with the extension described above, there
+are still \CranUndefinedRnd \eval calls without source references. This occurs
+when a reference to the \eval function is passed as argument to a higher-order
+functions or when the \eval call originates from native code. However, these
+missing source references account for a meager \CranUndefinedRatio of all calls;
+they are unlikely to affect our results. Other limitations are that we ignore
+calls to the native \eval function and to the alternate \c{rlang::tidy\_eval}
+function which uses native \eval internally. The \c{rlang} package introduces a
+new kind of promise called \c{quosure} which is evaluated by \c{tidy\_eval}.
+None of these limitation should invalidate our conclusions.
 

### 2021-04-01 paper/main.tex
format

@@
 
-\mypara{Base.}
-???
+\mypara{Base.} The base dataset contains 13 libraries performing basic
+arithmetics, statistics, and operating system related functionalities.
 
@@
 
-This section focuses on our corpus of \Corpus CRAN packages that use
-\eval; these represent {\bf 28.9\%} of the entire repository. To provide % \Corpus / \CranAll
-a picture of \eval usage, all runnable code in CRAN was executed. Runs
-that did not exercise an \eval in the corpus were discarded. We
-retained \Nbruns to generate the data presented here. In total,
-\Allcalls invocations of \eval and its variants were recorded. Out of
-the \Staticatleastonecallsite packages that use \eval, \Triggeredpkgs
-packages were observed calling it in our runs --- the difference can
-be chalked down to incomplete code coverage and occasional
-infrastructure failures.
-
-Figure~\ref{freq} summarizes the frequency of calls to \eval over all
-runs. The left column is the number of observed calls, the right
-column is the number of packages falling in that range. There are
-\Fewcalls packages with low \eval frequency, i.e. fewer than 100
-calls; and a similar number of packages, \Manycalls to be precise,
-that use \eval more than 1,000 times. One package, \Maxcallspack,
-calls it \Maxcalls times and thus accounts for over half of the
-observed data.
+This section focuses on CRAN packages and reports statistics about the usage of
+\eval. Calls to \eval appear in {\bf 28\%} of the packages hosted on CRAN. There
+are \Corpus packages with an \eval, all programs extracted from these packages
+were executed, any run that did not exercise one of the \eval call sites in the
+packages was discarded. This left \Nbruns runs of some of the \Corpus packages
+to generate the data presented here. There were \Allcalls invocations of \eval
+and its variants occurring in \Triggeredpkgs packages.\footnote{Not all packages
+with static call sites of \eval were observed triggering that call site, this
+can be chalked down to incomplete code coverage and occasional infrastructure
+failures of our analysis infrastructure.} Figure~\ref{freq} summarizes the
+frequency of dynamic calls to \eval with the left column being the number of
+calls and the right, the number of packages in that range. There are \Fewcalls
+packages with low \eval frequency, fewer than 100 calls; and a similar number of
+packages, \Manycalls to be precise, that use \eval more than 1,000 times. One
+package, \Maxcallspack, calls it \Maxcalls times and thus accounts for over half
+of the observed calls.
 
@@
 &     \multicolumn{3}{c}{\small\#calls} &\small\#pck \\\hline
-\tt 0 &--& \tt 10      & \Bina  & \tt 1K &--&\tt 100K  & \Bine\\
+\tt 1 &--& \tt 10      & \Bina  & \tt 1K &--&\tt 100K  & \Bine\\
 \tt 11 &--& \tt 100    & \Binb  & \tt 100K &--&\tt 1M  & \Binf\\
@@
 \begin{minipage}{7cm}
-\begin{tabular}{|@{\,}r@{\,}|@{\,}rrrr@{\,}|}\hline
+\begin{tabular}{|@{\,}r|rrrr|}\hline
   &\eval & \c{evalq} & \c{eval} & \c{local}\\[-2mm]
@@
 \end{tabular}~\\[2mm]\caption{Variants}\label{tab:variantseval}
-\end{minipage}
-\end{tabular}
-\end{figure}
+\end{minipage}\end{tabular}\end{figure}
 
@@
   \medskip  (a) All
+  \medskip
+  \medskip
 
@@
 
-\vspace{-2mm}
-    \caption{Normalized invocations} \label{cn}
-\vspace{-2mm}
+\vspace{-2mm} \caption{Normalized invocations} \label{cn}\vspace{-2mm}
 \end{wrapfigure}
 
-Figure~\ref{tab:variantseval} illustrates the use of variants of
-\eval. For each of the four variants, the first rows shows the number
-of call sites occurring in the corpus (\emph{static}), the second row
-is the number of call sites that were encountered during tracing
-(\emph{exercised}), the last row is the number of calls
-(\emph{invocations}). The vast majority of call sites, and calls, are
-to \eval itself, \c{eval.parent} is rare, as for \c{evalq} and
-\c{local} are barely used at all. The difference between static and
-exercised reflects spotty code coverage and is a limitation
-of dynamic analysis.
-
-Figure~\ref{cn}(a) shows normalized invocation counts. The column on
-the left is the average number of calls from a particular site for any
-given run (\emph{\#calls}). On the right is the count of sites that
-fall in that range (\emph{\#sites}). For instance, \Runbinh sites are
-invoked on average over 2,001 times per run. Larger numbers suggest
-uses from within a loop or a recursive context -- this seems to be the
-exception as most \evals, \Runbina, are exercised 50 times or less.
-Figure~\ref{cn}(b) zooms in on low frequency call sites. The x-axis
-shows normalized invocations and the y-axis is the number sites for
-that value. Most low-frequency \evals are invoked only once, about
-half as many are invoked twice; after that the frequency quickly
-drops.
-
-Arguments to \eval can be either code or values, in the latter case
-\eval has no effect. Code accounts for \Codepercent of arguments in
-our corpus. More specifically, \Symbolpercent of arguments are
-\c{symbol}s (i.e. these are single variables such as \c{x}),
-\Languagepercent are \c{language} object (an internal R type roughly
-corresponding to a single expression, such as \c{x+1}), and
-\Expressionpercent are \c{expression} objects (multiple statements).
-Further inspection reveals that most symbols, \Ggplotsymbolpercent to
-be exact, come from a single call in the \c{ggplot2} package and have
-have the value \c{\_inherit}. Even if we remove this outlier, the
-majority of \evals are simple expressions.
-
-To estimate how much executable code is dynamically injected into the
-program through \eval, we measure the size of arguments in number of
-abstract syntax tree nodes as returned by R's parser; for example,
-\c{x+1} counts as 3 nodes.\footnote{We experimented with measuring
-code size by de-parsing the expressions to strings, but these
-measurements were dominated by de-parsed data objects, these could
-range in the MBs.} The median argument size is \Medianszeval (recall
-the large number of symbols) and the average is \Avgszeval nodes. The
-largest \eval input observed is \Maxszeval\,-- a significant chunk of
-code. Figure~\ref{fig:sizedistribution} shows the distribution of
-sizes for arguments of fewer or equal to 25 nodes. The x-axis is the
-size of arguments in number of nodes, and the y-axis is the count of
-arguments with that size. The size drops rapidly, with few
-observations larger than 15 nodes. The long tail is omitted from the
-graph for legibility.
+Figure~\ref{tab:variantseval} enumerate the use of variants of \eval. For each
+of the four variants, the first rows shows the number of call sites in the
+corpus (\emph{static}), the second row is the number of call sites that were
+encountered during analysis (\emph{exercised}), the last row is the number of
+calls (\emph{invocations}). The overwhelming majority of call sites and dynamic
+calls, are to \eval itself, \c{eval.parent} is rare, and both \c{evalq} and
+\c{local} are barely used at all. The difference between sites and exercised
+sites underscores the limitations due to code coverage.
+
+Figure~\ref{cn}(a) shows normalized invocation counts per call site; on the left
+are average number of calls from a given site and given run, on the right are
+counts of sites that fall in that range. For instance, \Runbinh sites are
+invoked on average over 2,000+ times per run. Larger numbers suggest loops or
+recursive contexts -- this seems to be the exception as most \evals, \Runbina,
+are exercised 50 times or less. Figure~\ref{cn}(b) zooms in on low frequency
+call sites. The x-axis shows normalized invocations and the y-axis is the number
+sites for that value. Most low-frequency \evals are invoked only once, about
+half as many are invoked twice; after that the frequency quickly drops.
+
+The \eval can be passed any value, but if the argument is not an expression,
+\eval returns it without doing anything. Expressions account for \Codepercent of
+arguments in our corpus. More specifically, \Symbolpercent of arguments are
+\c{symbol}s (single variables such as \c{x}), \Languagepercent are \c{language}
+object (a single expression, such as \c{x+1}), and \Expressionpercent are
+\c{expression} objects (multiple statements). Further inspection reveals that
+most symbols, \Ggplotsymbolpercent to be exact, come from a single call in the
+\c{ggplot2} package and have have the value \c{\_inherit}. Even if we remove
+this outlier, the majority of \evals are simple expressions.
+
+To estimate how much executable code is injected through \eval, we measure the
+the number of abstract syntax tree nodes as returned by the parser; for example,
+\c{x+1} counts as 3.\footnote{We attempted to measure string lengths after
+de-parsing expressions, but these measurements were dominated by de-parsed data
+objects, which could range in the MBs.} The median argument size is
+\Medianszeval (due to the many symbols) and the average is \Avgszeval nodes. The
+largest \eval input observed is \Maxszeval\,-- a significant chunk of code.
+Figure~\ref{fig:sizedistribution} shows the distribution of sizes for arguments
+of fewer or equal to 25 nodes. The x-axis is the size of arguments in number of
+nodes, and the y-axis is the count of arguments with that size. The size drops
+rapidly, with few observations larger than 15 nodes. The long tail is omitted
+from the graph for legibility.
 
@@
 
-To measure the work performed in \evals, we count the instructions
-executed by interpreter during evaluation of the argument.  Most invocations perform relatively little
-work with \Smalleventspct of \evals executing 50 or fewer
-instructions. The violin plot of Figure~\ref{ev}(a) corresponds to
-\evals executing $\leq$ 50 instructions, it is dominated by trivial
-symbol look ups. Figure~\ref{ev}(b) has the work-intensive \evals
-which go all the way to \MaxeventsRnd instructions.
+To estimate the work performed in \evals, we count the instructions executed by
+interpreter. Most invocations perform relatively little work with
+\Smalleventspct of \evals executing 50 or fewer instructions. The violin plot of
+Figure~\ref{ev}(a) corresponds to \evals executing $\leq$ 50 instructions, it is
+dominated by trivial symbol look ups. Figure~\ref{ev}(b) has the work-intensive
+\evals which go all the way to \MaxeventsRnd instructions.
 
@@
 \end{tabular}
- \caption{Instructions per packages} \label{ev}
+ \caption{Instructions per \eval} \label{ev}
 \end{figure}
 
-\paragraph{Base} The base dataset contains 13 packages.
-All of them trigger at least one \eval call.
-In total, \baseAllcalls \eval  were recorded.
-Most of them are calls to the main \eval.
-There are a few calls to the other variants, less than 4\%.
-Contrary to the CRAN dataset, there are calls to \c{evalq} (\baseEvalqs).
-Out of \baseStaticeval \eval static sites, \baseTriggeredeval are exercised, i.e.
-\baseTriggeredevalpct, out of \baseStaticevalq, \baseTriggeredevalq are exercised, and out of \baseStaticevalparent, \baseTriggeredevalparent are exercised.
-Most of the arguments, \baseCodepercent,  are code objects.
-However, contrary to CRAN, few arguments, only \baseSymbolpercent, are symbols, but are rather language objects, \baseLanguagepercent.
-There are also few expression objects: \baseExpressionpercent.
-The prevalence of language objects suggests that \emph{base} mainly performs metaprogramming with \eval.
-The median argument size is \baseMedianszeval, which is more than for CRAN, which was dominated by symbol evaluation.
-The maximum size is \baseMaxszeval.
-Small event evals (less than 50 events) amount for \baseSmalleventspct of the events inside \eval calls.
-
-% Figures for evalq are weird: more triggered than static call sites
-
-\paragraph{Kaggle} In the Kaggle dataset, \kaggleNbruns notebooks trigger at least one \eval. In total, \kaggleAllcalls \eval were recorded. All of them are calls to the normal \eval.
-Out of \kaggleStaticeval static \eval sites, we exercise \kaggleTriggeredeval of them.
-Most \eval are invoked only once.
-Only one \eval amounts for several thousand, \kaggleMaxcalls,  invocations.
-Code amounts for \kaggleCodepercent of arguments in the kaggle corpus.
-However, contrary to the CRAN corpus, they are all expressions, and there are very few symbols (\kaggleSymbolpercent) and no language objects.
-Most expression objects result from a call to the \c{parse} function, suggesting that most \eval in Kaggle parse and evaluate strings.
-The median argument size is \kaggleMedianszeval, which makes sense, as there are very few arguments that are only symbols.
-The largest argument size is \kaggleMaxszeval.
-The distribution of events is similar to the one for CRAN or \emph{base}: \kaggleSmalleventspct of \eval calls amount for less than 50 events.
-
-\paragraph{Discussion} \emph{Base} is more akin to using \eval for metaprogramming, \emph{kaggle} to evaluating parsed strings, and CRAN is dominated by symbol evaluation.
-In terms of number of calls, a few calls dominate, with a call from \emph{ggplot} being prominent in the CRAN corpus.
-When looking at the \eval sites, \eval is more spread out.
-Those two measures emphasize two different points of view:
-
-\begin{itemize}
-	\item The number of calls is relevant to an optimizing compiler: the most frequent calls are the ones to be optimized first.
-	\item The number of sites is relevant to the programmer/user: it describes better how programmers use \eval.
-\end{itemize}
-
-% What about a threat model?
+\paragraph{Base}
+We recorded \baseAllcalls \eval calls in the base libraries with fewer than 4\%
+of those being to any variants. Unlike CRAN, Base calls \c{evalq} (\baseEvalqs).
+Out of \baseStaticeval call sites, \baseTriggeredevalpct are exercised. Most
+arguments, \baseCodepercent, are expressions. However, unlike CRAN,
+\baseLanguagepercent of the arguments are language objects. The median argument
+size is \baseMedianszeval, which is more than for CRAN. The maximum size is
+\baseMaxszeval. Small instruction counts, less than 50, amount for
+\baseSmalleventspct of calls.
+
+\paragraph{Kaggle}
+Only \kaggleNbruns scripts trigger \eval. In total, \kaggleAllcalls \eval were
+recorded; none are calls to variants. Out of \kaggleStaticeval call sites,
+\kaggleTriggeredeval were exercised. Most call sites are invoked only once. Only
+one site is called \kaggleMaxcalls times. Expressions amount for
+\kaggleCodepercent of arguments. Unlike CRAN corpus, there are very few symbols
+(\kaggleSymbolpercent). Most expressions result from calls to \c{parse}, thus
+most \evals start with strings. The median argument size is \kaggleMedianszeval,
+Which makes sense, as few arguments are only symbols. The largest argument is
+\kaggleMaxszeval. The distribution of instructions per eval is similar to CRAN;
+\kaggleSmalleventspct of \evals execute fewer than 50 instructions.
+
+\paragraph{Discussion}
+These results show that \eval is central to the language implementation as it is
+omnipresent in the relatively small Base library. Packages, which are developed
+by experienced programmers, make regular and varied use of the facility.
+Finally, scripts are written by less sophisticated users, and likely perform
+simpler tasks, and thus have lesser need for \eval, and its uses originate from
+strings which are easier to manipulate for end-users.
 
@@
 
-The previous section gave a quantitative view of the usage of \eval;
-in this section we try to elucidate \emph{what} it does.
+The previous section gave a quantitative view of \eval usage; we now try to
+elucidate \emph{what} it does.
 
@@
 
-The environment argument can be \c{NULL}, a \c{List}
-or a data frame. This happens in 3.5\% of the cases: \eval copies the
-fields of the list or data frame and creates bindings for them in a
-new environment. This pattern is used to evaluate formulas which can
-directly refer to the fields of the data. The \c{envir} argument can
-also be a number $n$. It means that the environment in which the
-expressions is evaluated will be the result of \c{sys.call(n)} where
+The environment argument can be \c{NULL}, a \c{List} or a data frame. This
+happens in 3.5\% of the cases: \eval copies the fields of the list or data frame
+and creates bindings for them in a new environment. This pattern is used to
+evaluate formulas which can directly refer to the fields of the data. The
+\c{envir} argument can also be a number $n$. It means that the environment in
+which the expressions is evaluated will be the result of \c{sys.call(n)} where
 $n$ refers to the $n$-th stack frame.
@@
 The top-level environment in R is called the global environment. New
-environments can be created using \c{new.env}. They can be provided a
-parent environment which becomes the enclosing scope of the new
-environment.
+environments can be created using \c{new.env}. They can be provided a parent
+environment which becomes the enclosing scope of the new environment.
 
@@
 
-We observe that a disproportionately high number of core R \eval calls
-access the  caller's environment. This is because many core R
-functions call functions that pass the result of \c{parent.frame()} to
-an \eval. A disproportionately high number of calls to \eval happens
-in an extended top-level environment. This can be explained by the
-fact that many packages evaluate code passed from the user's workspace
-in the top-level environment to access the bindings.
-
-The \c{imchange} function of package \c{imager} makes it possible to
-modify images using a dedicated formula syntax using
-\~.\footnote{Inspired by {map} in package \emph{purr}.} Here,
-\eval is evaluated in \c{newenv}, which creates a new environment that
-inherits from \c{parent.frame()} by default (classified as 1+).
+We observe that a disproportionately high number of core R \eval calls access
+the caller's environment. This is because many core R functions call functions
+that pass the result of \c{parent.frame()} to an \eval. A disproportionately
+high number of calls to \eval happens in an extended top-level environment. This
+can be explained by the fact that many packages evaluate code passed from the
+user's workspace in the top-level environment to access the bindings.
+
+The \c{imchange} function of package \c{imager} makes it possible to modify
+images using a dedicated formula syntax using \~.\footnote{Inspired by {map} in
+package \emph{purr}.} Here, \eval is evaluated in \c{newenv}, which creates a
+new environment that inherits from \c{parent.frame()} by default (classified as
+1+).
 
@@
 
-\c{adjCoef} in package \emph{actuar} find the root of an equation
-defined by a function \c{h} whose arguments must be named \c{x} and
-\c{y}. \c{h} is transformed into an auxiliary function \c{h2} that can
-be optimized. Here, the list used for \c{envir} ensures the
-correspondance between the textual arguments of \c{h} and the
-arguments of \c{h2}.
+\c{adjCoef} in package \emph{actuar} find the root of an equation defined by a
+function \c{h} whose arguments must be named \c{x} and \c{y}. \c{h} is
+transformed into an auxiliary function \c{h2} that can be optimized. Here, the
+list used for \c{envir} ensures the correspondance between the textual arguments
+of \c{h} and the arguments of \c{h2}.
 
@@
 \subsubsection{The expression passed to \eval.}
-\paragraph{Touching package environments.} It refers to the \emph{namespace} of a package.
-A call to \c{library(packageName)}, which load libraries in R, for instance, will populate environment \c{package:packageName}.
-It is a \emph{base} function though, so we do not see it in the CRAN dataset.
-However, R still makes it possible to access any package environments and \packageNbPackageEnvSites call sites, \ie \packagePackageEnvSitePercent access a package environment, from \packageNbPackageEnvPackages different packages.
-Those packages perform introspection, to represent and search any environment by name, in package \emph{envnames}, to containerize tests, as with \emph{testthat} and \emph{unitizer}, to create a custom OOP system with \emph{R.oo}, or to create poor man's hidden functions in \emph{USSCXenaTools}.
-
-It is also used to implement a custom caching mechanism in packages \emph{g.data} and \emph{SOAR}.
-
-During tracing, we gather all source expressions and all resolved expression arguments passed to \eval. For instance, a non-resolved expression and some of its resolved expressions can be:
+\paragraph{Touching package environments.}
+It refers to the \emph{namespace} of a package. A call to
+\c{library(packageName)}, which load libraries in R, for instance, will populate
+environment \c{package:packageName}. It is a \emph{base} function though, so we
+do not see it in the CRAN dataset. However, R still makes it possible to access
+any package environments and \packageNbPackageEnvSites call sites, \ie
+\packagePackageEnvSitePercent access a package environment, from
+\packageNbPackageEnvPackages different packages. Those packages perform
+introspection, to represent and search any environment by name, in package
+\emph{envnames}, to containerize tests, as with \emph{testthat} and
+\emph{unitizer}, to create a custom OOP system with \emph{R.oo}, or to create
+poor man's hidden functions in \emph{USSCXenaTools}.
+
+It is also used to implement a custom caching mechanism in packages
+\emph{g.data} and \emph{SOAR}.
+
+During tracing, we gather all source expressions and all resolved expression
+arguments passed to \eval. For instance, a non-resolved expression and some of
+its resolved expressions can be:
 
@@
 
-We classify the resolved expressions to take into account the structure of the expression passed to \eval.
-However, we do not want to discriminate expressions based on specific values.
-Hence, the expression goes through several simplification steps: constant folding of arithmetic and string expressions that use commonly used functions (addition, string concatenation), discarding of specific value to a symbolic value holder, variable absorption.
-We keep track of variables, as they are evaluated in the environment specified to \eval, contrary to values that are left unchanged.
-We also track whether assignment operators are used in the expression or not.
-After the simplification steps, we obtain 15 different \emph{minimized} expressions, of which we show the eleven first most frequent in number of sites in Table~\ref{tab:minimizedexpressions}.
-
-\begin{table}[h]
-	\begin{tabular}{|c|c|c|c|c|c|}
-		\hline
-		Minimized & \#sites & \%sites & \#packages & \#operations & \%envir \\
-		\hline
-		\texttt{X} & \packageMinimizedcallsitesa  & \packageMinimizedpropsitesa & \packageMinimizedpackagea & \packageMinimizedoperationsaRnd & \packageMinimizedpercentenvirnonparenta \\
-		\hline
-		\texttt{F(F())} & \packageMinimizedcallsitesb  & \packageMinimizedpropsitesb & \packageMinimizedpackageb  & \packageMinimizedoperationsbRnd & \packageMinimizedpercentenvirnonparentb \\
-		\hline
-		\texttt{V} & \packageMinimizedcallsitesc & \packageMinimizedpropsitesc & \packageMinimizedpackagec & \packageMinimizedoperationscRnd & \packageMinimizedpercentenvirnonparentc \\
-		\hline
-		\texttt{F(X)}& \packageMinimizedcallsitesd & \packageMinimizedpropsitesd & \packageMinimizedpackaged & \packageMinimizedoperationsdRnd & \packageMinimizedpercentenvirnonparentd \\
-		\hline
-		\texttt{model.frame} & \packageMinimizedcallsitese & \packageMinimizedpropsitese & \packageMinimizedpackagee & \packageMinimizedoperationseRnd & \packageMinimizedpercentenvirnonparente \\
-		\hline
-		\texttt{F()}& \packageMinimizedcallsitesf & \packageMinimizedpropsitesf & \packageMinimizedpackagef & \packageMinimizedoperationsfRnd & \packageMinimizedpercentenvirnonparentf \\
-		\hline
-		\texttt{\$}& \packageMinimizedcallsitesg & \packageMinimizedpropsitesg & \packageMinimizedpackageg & \packageMinimizedoperationsgRnd & \packageMinimizedpercentenvirnonparentg \\
-		\hline
-		\texttt{FUN} & \packageMinimizedcallsitesh & \packageMinimizedpropsitesh & \packageMinimizedpackageh & \packageMinimizedoperationshRnd & \packageMinimizedpercentenvirnonparenth \\
-		\hline
-		\texttt{<-} & \packageMinimizedcallsitesi  & \packageMinimizedpropsitesi & \packageMinimizedpackagei & \packageMinimizedoperationsiRnd & \packageMinimizedpercentenvirnonparenti \\
-		\hline
-		\texttt{\{BLOCK\}} & \packageMinimizedcallsitesj & \packageMinimizedpropsitesj & \packageMinimizedpackagej & \packageMinimizedoperationsjRnd & \packageMinimizedpercentenvirnonparentj \\
-		\hline
-		\texttt{F(F(<-))} & \packageMinimizedcallsitesk & \packageMinimizedpropsitesk & \packageMinimizedpackagek & \packageMinimizedoperationskRnd & \packageMinimizedpercentenvirnonparentk \\
-		\hline
+We classify the resolved expressions to take into account the structure of the
+expression passed to \eval. However, we do not want to discriminate expressions
+based on specific values. Hence, the expression goes through several
+simplification steps: constant folding of arithmetic and string expressions that
+use commonly used functions (addition, string concatenation), discarding of
+specific value to a symbolic value holder, variable absorption. We keep track of
+variables, as they are evaluated in the environment specified to \eval, contrary
+to values that are left unchanged. We also track whether assignment operators
+are used in the expression or not. After the simplification steps, we obtain 15
+different \emph{minimized} expressions, of which we show the eleven first most
+frequent in number of sites in Table~\ref{tab:minimizedexpressions}.
+
+\begin{table}[h]\small
+\begin{tabular}{|c|c|c|c|c|c|}\hline
+Minimized & \#sites & \%sites & \#packages & \#operations & \%envir \\	\hline
+\texttt{X} & \packageMinimizedcallsitesa  & \packageMinimizedpropsitesa & \packageMinimizedpackagea & \packageMinimizedoperationsaRnd & \packageMinimizedpercentenvirnonparenta \\\hline
+\texttt{F(F())} & \packageMinimizedcallsitesb  & \packageMinimizedpropsitesb & \packageMinimizedpackageb  & \packageMinimizedoperationsbRnd & \packageMinimizedpercentenvirnonparentb \\\hline
+\texttt{V} & \packageMinimizedcallsitesc & \packageMinimizedpropsitesc & \packageMinimizedpackagec & \packageMinimizedoperationscRnd & \packageMinimizedpercentenvirnonparentc \\\hline
+\texttt{F(X)}& \packageMinimizedcallsitesd & \packageMinimizedpropsitesd & \packageMinimizedpackaged & \packageMinimizedoperationsdRnd & \packageMinimizedpercentenvirnonparentd \\\hline
+\texttt{model.frame} & \packageMinimizedcallsitese & \packageMinimizedpropsitese & \packageMinimizedpackagee & \packageMinimizedoperationseRnd & \packageMinimizedpercentenvirnonparente \\\hline
+\texttt{F()}& \packageMinimizedcallsitesf & \packageMinimizedpropsitesf & \packageMinimizedpackagef & \packageMinimizedoperationsfRnd & \packageMinimizedpercentenvirnonparentf \\\hline
+\texttt{\$}& \packageMinimizedcallsitesg & \packageMinimizedpropsitesg & \packageMinimizedpackageg & \packageMinimizedoperationsgRnd & \packageMinimizedpercentenvirnonparentg \\\hline
+\texttt{FUN} & \packageMinimizedcallsitesh & \packageMinimizedpropsitesh & \packageMinimizedpackageh & \packageMinimizedoperationshRnd & \packageMinimizedpercentenvirnonparenth \\\hline
+\texttt{<-} & \packageMinimizedcallsitesi  & \packageMinimizedpropsitesi & \packageMinimizedpackagei & \packageMinimizedoperationsiRnd & \packageMinimizedpercentenvirnonparenti \\\hline
+\texttt{\{BLOCK\}} & \packageMinimizedcallsitesj & \packageMinimizedpropsitesj & \packageMinimizedpackagej & \packageMinimizedoperationsjRnd & \packageMinimizedpercentenvirnonparentj \\\hline
+\texttt{F(F(<-))} & \packageMinimizedcallsitesk & \packageMinimizedpropsitesk & \packageMinimizedpackagek & \packageMinimizedoperationskRnd & \packageMinimizedpercentenvirnonparentk \\\hline
 \end{tabular}
-	\caption{Minimized expressions: eleven most frequent in number of sites.
-\#sites is the number of sites; \%sites the percentage of sites across the total number of sites (\packageNbCallSites sites) with at least this minimized expression; \#packages is the number of packages with at least one of the minimized expression and measures how widespread a minimized expression is.
-\#operations is the average number of steps performed by the R interpreter when given that minimized expression, and \%envir is the percentage of call sites for which the \c{envir} argument is explicitly passed, and is not \c{parent.frame()}, which is the most common explicit environment passed to \eval.} \label{tab:minimizedexpressions}
+\caption{Minimized expressions: eleven most frequent in number of sites. \#sites
+  is the number of sites; \%sites the percentage of sites across the total
+  number of sites (\packageNbCallSites sites) with at least this minimized
+  expression; \#packages is the number of packages with at least one of the
+  minimized expression and measures how widespread a minimized expression is.
+  \#operations is the average number of steps performed by the R interpreter
+  when given that minimized expression, and \%envir is the percentage of call
+  sites for which the \c{envir} argument is explicitly passed, and is not
+  \c{parent.frame()}, which is the most common explicit environment passed to
+  \eval.} \label{tab:minimizedexpressions}
 \end{table}
 
-Note that a site can exhibit several minimized expressions (see Figure~\ref{fig:frequentcombinationminimized}).
-However, \packageNbOneMinimizedPercent of the sites have only one minimized expression.
-Note also that if a function is called in that expression, we see only the function call but not the body of the function.
-We will trace what happens during the evaluation of the expression by \eval in Section~\ref{sec:dangerseval}, especially about side effects performed in \eval.
+Note that a site can exhibit several minimized expressions (see
+Figure~\ref{fig:frequentcombinationminimized}). However,
+\packageNbOneMinimizedPercent of the sites have only one minimized expression.
+Note also that if a function is called in that expression, we see only the
+function call but not the body of the function. We will trace what happens
+during the evaluation of the expression by \eval in
+Section~\ref{sec:dangerseval}, especially about side effects performed in \eval.
 
@@
 	\includegraphics[width=\textwidth]{package_combination_minimized.pdf}
-	\caption{Most frequent combinations of minimized expressions per site, with a cumulative share of 95\% of the sites.} \label{fig:frequentcombinationminimized}
+	\caption{Most frequent combinations of minimized expressions per site,
+          with a cumulative share of 95\% of the
+          sites.} \label{fig:frequentcombinationminimized}
 \end{figure}
@@
 \paragraph{Variables.}
-The most common minimized expression is a variable.\footnote{An arithmetic expression with some variables and constant, \eg \c{1 + x}, will also be normalized to \c{X}, i.e.
-a variable.} It happens in \packageMinimizedpropsitesa\% of the sites and is widespread across packages as it is found in \packageMinimizedpackagea packages across our CRAN corpus.
+The most common minimized expression is a variable.\footnote{An arithmetic
+expression with some variables and constant, \eg \c{1 + x}, will also be
+normalized to \c{X}, i.e. a variable.} It happens in
+\packageMinimizedpropsitesa\% of the sites and is widespread across packages as
+it is found in \packageMinimizedpackagea packages across our CRAN corpus.
 \packageNbSymbolVarSitePercent are only symbols, \ie not complex expressions.
-The \c{envir} argument is explicitly passed in \packageMinimizedpercentenvira\% of the call sites.
-The call sites without an explicit \c{envir} argument all use a specific \eval variant, \c{eval.parent}.
-Moreover, the \c{envir} argument is both explicit and different from \c{parent.frame()} in \packageMinimizedpercentenvirnonparenta\% of the sites, which suggests that when given a variable, \eval is often used to get  value in a non-local, non-parent environment.
-
-Similarly, when a function is evaluated on some variable (\c{F(X)} minimized expression), the percentage of non default, non parent environments is rather high, \packageMinimizedpercentenvirnonparentd\%.
-
-Minimized expression \texttt{\$} refers to a slot access, with \texttt{\$}, \texttt{[]} or \texttt{[[]]}, as the top call in a simple expression, for instance \texttt{a\$slot + 1}.
-It amounts for \packageNbSlotAccess sites, \ie \packageSlotAccessPercent of the sites.
-The non-default, non-parent environment percentage is here \packageMinimizedpercentenvirnonparentg.
-
-As a whole, variable access, either as the top call, or inside a function call, possibly with accessing a slot, represents \packageGeneralizedVarAccessSitePercent of the sites.
+The \c{envir} argument is explicitly passed in \packageMinimizedpercentenvira\%
+of the call sites. The call sites without an explicit \c{envir} argument all use
+a specific \eval variant, \c{eval.parent}. Moreover, the \c{envir} argument is
+both explicit and different from \c{parent.frame()} in
+\packageMinimizedpercentenvirnonparenta\% of the sites, which suggests that when
+given a variable, \eval is often used to get value in a non-local, non-parent
+environment.
+
+Similarly, when a function is evaluated on some variable (\c{F(X)} minimized
+expression), the percentage of non default, non parent environments is rather
+high, \packageMinimizedpercentenvirnonparentd\%.
+
+Minimized expression \texttt{\$} refers to a slot access, with \texttt{\$},
+\texttt{[]} or \texttt{[[]]}, as the top call in a simple expression, for
+instance \texttt{a\$slot + 1}. It amounts for \packageNbSlotAccess sites, \ie
+\packageSlotAccessPercent of the sites. The non-default, non-parent environment
+percentage is here \packageMinimizedpercentenvirnonparentg.
+
+As a whole, variable access, either as the top call, or inside a function call,
+possibly with accessing a slot, represents
+\packageGeneralizedVarAccessSitePercent of the sites.
 
 \paragraph{Values.}
-Many sites evaluates to \c{V}.
-\packageValOneNodePercent of those sites are only passed an expression with one node, \ie one value, not a more complex expression that is later constant folded.
-\c{V} corresponds to two main categories of expressions passed to \eval:
+Many sites evaluates to \c{V}. \packageValOneNodePercent of those sites are only
+passed an expression with one node, \ie one value, not a more complex expression
+that is later constant folded. \c{V} corresponds to two main categories of
+expressions passed to \eval:
 
@@
 
-For the first category, \eval is not necessary.
-In \packageNbCallSitesUniqueActualValue sites exhibiting \c{V}, \ie \packageCallSitesUniqueActualValuePercent\% sites, it is only passed such actual values.
-The median number of runs for those sites is \packageMedianRunSitesUniqueActualValue (average \packageAverageRunSitesUniqueActualValue, standard deviation \packageSdRunSitesUniqueActualValue), suggesting it happens because of a lack of coverage, because there are rarely triggered in our dataset and all paths are not taken.
-However, manual inspection of the source code of some examples shows that some sites actually pass values.
-For instance, in package \emph{ctmle}, in function \c{evaluate\_candidate\_glmnet}:
+For the first category, \eval is not necessary. In
+\packageNbCallSitesUniqueActualValue sites exhibiting \c{V}, \ie
+\packageCallSitesUniqueActualValuePercent\% sites, it is only passed such actual
+values. The median number of runs for those sites is
+\packageMedianRunSitesUniqueActualValue (average
+\packageAverageRunSitesUniqueActualValue, standard deviation
+\packageSdRunSitesUniqueActualValue), suggesting it happens because of a lack of
+coverage, because there are rarely triggered in our dataset and all paths are
+not taken. However, manual inspection of the source code of some examples shows
+that some sites actually pass values. For instance, in package \emph{ctmle}, in
+function \c{evaluate\_candidate\_glmnet}:
 
@@
 \c{paste} concatenates strings and \eval will just return that string.
-\c{as.formula} can take a string and convert it to a formula; \eval is not needed here.
-In 20 random sites among the ones with values, a conservative manual inspection showed that \packageUsefulValueEvalPercent of the sites  had a  legitimate reason to use \eval.
-
-The percentage of non default environment passed to \eval is \packageMinimizedpercentenvirc when the minimized expression is a value.
-Some sites which are passed a value, are also passed different minimized expressions.
-For those ones, \packageNonDefaultEnvirValuePercent use a non-default environment, and among the one that use a non-default environment, \packageNonDefaultEnvirWithVarPercent have at least one minimized expression with a variable inside.
+\c{as.formula} can take a string and convert it to a formula; \eval is not
+needed here. In 20 random sites among the ones with values, a conservative
+manual inspection showed that \packageUsefulValueEvalPercent of the sites had a
+legitimate reason to use \eval.
+
+The percentage of non default environment passed to \eval is
+\packageMinimizedpercentenvirc when the minimized expression is a value. Some
+sites which are passed a value, are also passed different minimized expressions.
+For those ones, \packageNonDefaultEnvirValuePercent use a non-default
+environment, and among the one that use a non-default environment,
+\packageNonDefaultEnvirWithVarPercent have at least one minimized expression
+with a variable inside.
 
@@
 
-\paragraph{Function definition.} \c{FUN} represents a function definition at the top level of the expression, and happens in \packageFunctionDefinitionSites sites, \ie \packageFunctionDefinitionSitesPercent of the sites.
-It entails very few steps of the interpreter, as the function is returned by \eval, and the body of the new function is just stored for later execution.
-If we count also function definition inside more complex expressions, it amounts to \packageGeneralizedFunctionDefinitionSites sites, \ie \packageGeneralizedFunctionDefinitionSitesPercent and suggests the widespread use of higher-order functions in R, such as \c{lapply}, the R version of \c{map}.
-
-\paragraph{Assignments.} This amounts to \packageAssignSitesPercent of the sites and corresponds to minimized expressions \c{<-}, where the assignment is the top call, and \c{F(F(<-))}, where the assignment happens inside, and others that we were not enough present to be shown in table~\ref{tab:minimizedexpressions}.
-Assignments are performed with \c{<-}, \c{<<-} or \c{assign}.
-Most often, assignment happens in top call position.
-A non-default environment is specified only in \packageNonDefaultEnvirAssignSitesPercent of the sites with assignments.
-Indeed, most of those assignments occur in the default environment, \ie in the caller of \eval.
-This hints that side effects are not widespread in \eval.\footnote{Side effects can also happen in a function call in the \eval expression, and that function call will hide assignment operators from us.
-We more precisely instrument side-effects to environment in Section~REF.}
-
-\paragraph{Large expressions.}  Compositions of functions, without variables (\c{F(F(X))} minimized expression) is the second most frequent minimized expression.
-\c{\{BLOCK\}} is a sequence of statements.
-It is usually large, with in average \packageMinimizedoperationsjRnd operations in it.
-
-The minimized expression classification suggests that the presence of variables in the expression is highly correlated to the specification of a specific environment, and that \eval is used here for its power in precising the evaluation environment.
-
-\NOTE{Maybe we should ditch the table of the minimized expressions and rather use the ones of combinations.
-It has the advantage of showing what is the most frequent, per call site.
-But it's a long tail, and sites that have several minimized expressions are not that numerous, only 13\%.
-But if include combinations, percentages change.
-For instance, 3314 call sites have only X, i.e.
-22.51, instead of 4353 and 29.6\%.
-}
+\paragraph{Function definition.}
+\c{FUN} represents a function definition at the top level of the expression, and
+happens in \packageFunctionDefinitionSites sites, \ie
+\packageFunctionDefinitionSitesPercent of the sites. It entails very few steps
+of the interpreter, as the function is returned by \eval, and the body of the
+new function is just stored for later execution. If we count also function
+definition inside more complex expressions, it amounts to
+\packageGeneralizedFunctionDefinitionSites sites, \ie
+\packageGeneralizedFunctionDefinitionSitesPercent and suggests the widespread
+use of higher-order functions in R, such as \c{lapply}, the R version of
+\c{map}.
+
+\paragraph{Assignments.}
+This amounts to \packageAssignSitesPercent of the sites and corresponds to
+minimized expressions \c{<-}, where the assignment is the top call, and
+\c{F(F(<-))}, where the assignment happens inside, and others that we were not
+enough present to be shown in table~\ref{tab:minimizedexpressions}. Assignments
+are performed with \c{<-}, \c{<<-} or \c{assign}. Most often, assignment happens
+in top call position. A non-default environment is specified only in
+\packageNonDefaultEnvirAssignSitesPercent of the sites with assignments. Indeed,
+most of those assignments occur in the default environment, \ie in the caller of
+\eval. This hints that side effects are not widespread in \eval.\footnote{Side
+effects can also happen in a function call in the \eval expression, and that
+function call will hide assignment operators from us. We more precisely
+instrument side-effects to environment in Section~REF.}
+
+\paragraph{Large expressions.}  Compositions of functions, without variables
+(\c{F(F(X))} minimized expression) is the second most frequent minimized
+expression. \c{\{BLOCK\}} is a sequence of statements. It is usually large, with
+in average \packageMinimizedoperationsjRnd operations in it.
+
+The minimized expression classification suggests that the presence of variables
+in the expression is highly correlated to the specification of a specific
+environment, and that \eval is used here for its power in precising the
+evaluation environment.
+
+\NOTE{Maybe we should ditch the table of the minimized expressions and rather
+  use the ones of combinations. It has the advantage of showing what is the most
+  frequent, per call site. But it's a long tail, and sites that have several
+  minimized expressions are not that numerous, only 13\%. But if include
+  combinations, percentages change. For instance, 3314 call sites have only X,
+  i.e. 22.51, instead of 4353 and 29.6\%. }
 

### 2021-04-01 paper/main.tex
more

@@
 \begin{document}
-
 \title{What we {\sf eval} in the shadows}
@@
 %% Jan Vitek
-
 %% What is the right order?
@@
 \email{r@s.com}
-\affiliation{%
-	\institution{Institute for Clarity in Documentation}
-	\streetaddress{P.O. Box 1212}
-	\city{Dublin}
-	\state{Ohio}
-	\country{USA}
-	\postcode{43017-6221}
-}
+\affiliation{\institution{Institute}}
 
@@
 
-\vspace{2mm}\noindent\emph{Is Past Prologue?} Previous work
-investigated how \eval is used in web programming, specifically in
-websites that use JavaScript~\cite{pldi10a}. In 2010, 17 of the
-largest website used the feature. In 2011, 82\% of the 10,000 most
-accessed sites used \eval~\cite{ecoop11}.  Yet, the strings passed to
-\eval, and their behaviors when executed, are far from random; it was
-shown that when one can observe several calls to \eval, the ``shape''
-of future calls can be predicted with 97\% accuracy~\cite{oopsla12b}.
-Overall, practical usage suggested that most reflective calls were
-relatively harmless. While this backs up the soundiness squad's
-approach, does it generalize to other application domains than web
-programming and to other languages?
-
-\vspace{2mm}\noindent\emph{The Here and Now.} In this study, we
-investigate the usage of \eval in programs written in the R
-programming language. R is language designed by statisticians for
-applications in data science~\cite{r,R96}. What makes looking at R
-after JavaScript interesting is that, while both languages are
-dynamic, they are quite different. While one can program in an
-object-oriented style in R, like in JavaScript, R is primarily a lazy,
-untyped, functional language. JavaScript was designed to run untrusted
-code in browser, while R is used for statistical computing on
-desktops. JavaScript is a general purpose language used by a wide
-community of programmers; while R is used for scientific computing by
-data scientists and domain experts with, often, limited programming
-experience. One can distinguish between library implementers,
-developers with some programming experience and a working knowledge of
-R, and end-users, who are typically not expert programmers and often
-have only a cursory knowledge of the language.\footnote{Consider that
-R evaluates function argument lazily, just like Haskell. We informally
-surveyed end-users, including computer scientists, and did not find a
-single user aware of this fact. Library developers, on the other hand,
-know about laziness and program defensively around it.} Our goal is
-thus to highlight the differences in usage between JavaScript and R,
-and try to explain those differences in terms of language features,
-application domain and programmer experience. Hopefully some of our
+\vspace{2mm}\noindent\emph{Is Past Prologue?} Previous work investigated how
+\eval is used in web programming, specifically in websites that use
+JavaScript~\cite{pldi10a}. In 2010, 17 of the largest website used the feature.
+In 2011, 82\% of the 10,000 most accessed sites used \eval~\cite{ecoop11}. Yet,
+the strings passed to \eval, and their behaviors when executed, are far from
+random; it was shown that when one can observe several calls to \eval, the
+``shape'' of future calls can be predicted with 97\% accuracy~\cite{oopsla12b}.
+Overall, practical usage suggested that most reflective calls were relatively
+harmless. While this backs up the soundiness squad's approach, does it
+generalize to other application domains than web programming and to other
+languages?
+
+\vspace{2mm}\noindent\emph{The Here and Now.} In this study, we investigate the
+usage of \eval in programs written in the R programming language. R is language
+designed by statisticians for applications in data science~\cite{r,R96}. What
+makes looking at R after JavaScript interesting is that, while both languages
+are dynamic, they are quite different. While one can program in an
+object-oriented style in R, like in JavaScript, R is primarily a lazy, untyped,
+functional language. JavaScript was designed to run untrusted code in browser,
+while R is used for statistical computing on desktops. JavaScript is a general
+purpose language used by a wide community of programmers; while R is used for
+scientific computing by data scientists and domain experts with, often, limited
+programming experience. One can distinguish between library implementers,
+developers with some programming experience and a working knowledge of R, and
+end-users, who are typically not expert programmers and often have only a
+cursory knowledge of the language.\footnote{Consider that R evaluates function
+argument lazily, just like Haskell. We informally surveyed end-users, including
+computer scientists, and did not find a single user aware of this fact. Library
+developers, on the other hand, know about laziness and program defensively
+around it.} Our goal is thus to highlight the differences in usage between
+JavaScript and R, and try to explain those differences in terms of language
+features, application domain and programmer experience. Hopefully some of our
 observations will generalize to other languages.
 
-\vspace{2mm}\noindent\emph{The What and How.} One significant benefit
-of choosing R is that every package in the CRAN repository is curated
-and comes with examples of typical usage. This gives us a large code
-base that we can analyze dynamically. To observe \eval we built a
-two-level monitoring infrastructure:\footnote{Our infrastructure is
-open source and publicly available, our anonymized code is at {\tt
-  http://url.com}, a more complete artifact will be submitted to the
-artifact evaluation committee. } we can monitor R programs by
-instrumentation --- this gives us access to many user-visible
-properties of R programs --- but we can also monitor the
-inner-workings of the R interpreter --- this allows us to capture
-details not exposed at the source level. Dynamic analysis is limited,
-it can only observe behaviors triggered by the particular inputs
-passed to a program. Luckily, CRAN libraries come with many tests and
-use-cases. The choice of corpus is crucial. Our corpus has been
-constructed to reflect the levels of sophistication of the R
-community. We distinguish between \emph{CRAN packages}
-(\CranAllPackages curated packages that pass stringent quality checks
-and are equipped with tests and sample data) and \emph{Kaggle scripts}
-(\KaggleUnique end-user written programs that performs a
-particular data analysis task). It is reasonable to expect that \eval
-usage differ between these datasets: the libraries represent a lively
-ecosystem with new libraries added each day, while end user code is
-often thrown together, run once, and never revisited.
+\vspace{2mm}\noindent\emph{The What and How.} One significant benefit of
+choosing R is that every package in the CRAN repository is curated and comes
+with examples of typical usage. This gives us a large code base that we can
+analyze dynamically. To observe \eval we built a two-level monitoring
+infrastructure:\footnote{Our infrastructure is open source and publicly
+available, our anonymized code is at {\tt http://url.com}, a more complete
+artifact will be submitted to the artifact evaluation committee. } we can
+monitor R programs by instrumentation --- this gives us access to many
+user-visible properties of R programs --- but we can also monitor the
+inner-workings of the R interpreter --- this allows us to capture details not
+exposed at the source level. Dynamic analysis is limited, it can only observe
+behaviors triggered by the particular inputs passed to a program. Luckily, CRAN
+libraries come with many tests and use-cases. The choice of corpus is crucial.
+Our corpus has been constructed to reflect the levels of sophistication of the R
+community. We distinguish between \emph{CRAN packages} (\CranAllPackages curated
+packages that pass stringent quality checks and are equipped with tests and
+sample data) and \emph{Kaggle scripts} (\KaggleUnique end-user written programs
+that performs a particular data analysis task). It is reasonable to expect that
+\eval usage differ between these datasets: the libraries represent a lively
+ecosystem with new libraries added each day, while end user code is often thrown
+together, run once, and never revisited.
 
@@
 
-\subsection{Eval Usage in R}
-
-The R language was intended to be extensible, the combination of lazy
-evaluation, \c{substitute} and \c{eval} are the tools given to
-developers to this end. This API is slightly more complex than just
-passing a string, it is conceivable that this may discourage some
-casual users. \Eval is also being used to reduce boilerplate code and
-provide convenience features for programmers. We now give some
-representative examples of its usage.
-
-\mypara{Intercession.} A common use case for \eval is to be combined with
-\c{match.call}. \c{match.call} walks up the call stack, captures the code that
-invoked the currently executing function, and returns it as an unevaluated
-expression. The pattern is to transform a call to some function \c{f} into a
-call to \c{g} with some arguments retained and others modified. As an
-illustration, consider the \c{vcpart} package's function \c{tvcglm} that is
-translated to a call to \c{tvcm} with two modifications to the argument list:
-argument \c{control} can't be missing and \c{fit} is set \c{"glm"}. The function
-ends with a call to \c{eval.parent} to ensure that the rewritten call is
-evaluted in the same environment the original call was.
-\begin{lstlisting}
- tvcglm <- function(formula, data, family, ...) {
-   k <- match.call()
-   k[[1]] <- as.name("tvcm")
-   if ("fit" %in% names(list(...))) warning("'fit' ignored.")
-   k$fit <- "glm"
-   eval.parent(k)
- }
-\end{lstlisting}
-This pattern is recognizable by the fact that the expression is a call
-and the target environment is that of the parent.
-
-\mypara{Code Generation.} The more traditional use of \eval is to execute code
-that was assembled by the programmer into a string. Here we show the method
-\c{plot} for class \c{sback} in package \c{wsbackfit}, simplified for
-explanatory purposes. The function takes a long argument list, the names of
-which are captured in the list \c{opt}. The string \c{stub} is composed of a
-subset of the arguments passed to this function; the variable is used to
-construct a call to \c{base::plot} which will draw a plot. \NOTE{This example is
-  a bit messed up. What is var? Does the use of ... in the eval grab the
-  arguments of this call? If yes does it not end up with xlab twice? HELP} The
-\parse and \eval combination is used by the the \source function in the base
-library to load R code from a file in the current workspace.
-\begin{lstlisting}
- plot.sback <- function(x,...) {
-   opt <- names(list(...))
-   stub <- paste(
-    ifelse("xlab" %in% opt,"",paste(",xlab=\"",var,"\"",sep="")),
-    ifelse("type" %in% opt,"",",type=\"l\""), sep="")
-   plot <- paste("plot(x.data,",stub,",...)",sep="")
-   eval(parse(text=plot))
-\end{lstlisting}
-This pattern is recognizable by its use of \c{parse} to turn a string
-into an expression.
-
-\mypara{Debloating.} \Eval is often used as a means to reduce
-boilerplate code; simple and repetitive code can easily be replaced
-with judcious use of \eval. For example, the \c{data.table} package
-uses \eval to calls the \c{options} function with named arguments
-taken from a vector of strings. While the benefits are limited in this
-example, it is an attractive tool for programmers.
-\begin{lstlisting}
-  opts = c("datatable.verbose"="FALSE", # ...many others
-  for (i in names(opts))
-    eval(parse(text=paste0("options(",i,"=",opts[i],")")))
-\end{lstlisting}
-This pattern is a special case of code generation, recognizable by the
-fact that \eval is executed in a loop.
-
-\mypara{Trivial.} When values are passed to \eval, they are returned
-unchanged. They are an example of trivial uses of \eval. Another
-trivial use is the empty expression, often found in JavaScript, but
-rare in R.
-
 \subsection{Previous Work}
@@
 
+\newpage
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-Our corpus is assembled from three sources. Firstly, we start with the
-\emph{Base} libraries that are bundled with the language implementation.
-Secondly, we acquire packages hosted on \emph{CRAN}. Thirdly, we downloaded
+Our corpus is assembled from three sources: the \emph{Base} libraries that are
+bundled with the language implementation, packages hosted on \emph{CRAN} and
 scripts from \emph{Kaggle}.
 
-\mypara{Base.} The base dataset contains 13 libraries performing basic
-arithmetics, statistics, and operating system related functionalities.
+\mypara{Base.} The base dataset contains 13 core libraries performing basic
+arithmetics, statistics, and operating system related functionalities. These
+libraries are bundled with R and are executed pervasively.
 
 \mypara{CRAN.} The Comprehensive R Archive
-Network,\footnote{\url{cran.r-project.org}} or CRAN, is the largest curated
-repository of R packages. It hosts \CranAvailablePackagesRnd packages with 6 new
-packages submitted each day~\cite{Ligges2017}. Packages that are accepted to
-CRAN are authored by experienced developers and they must abide to a number of
-well-formedness rules automatically checked on each commit. Furthermore, each
-package comes with sample data and tests. There are three sources of runnable
-code for each package: \emph{tests, examples} and \emph{vignettes} --
-respectively, unit tests, executable code snippets from the documentation, and
-long-form executable use-cases written in Markdown. Examples and vignettes can
-be turned into scripts by extracting the relevant code. We obtained
-\CranPackages that call \eval. These packages represent \CranPackagesRatio of
-all packages or \CranCodeRatio of the entire code size of CRAN. From these we
-extracted \CranRunnableScripts scripts that contain \CranRunnableCode lines of
-code (\CranRunnableCodeExamplesRnd in examples, \CranRunnableCodeVignettesRnd in
-vignettes and \CranRunnableCodeTestsRnd in tests).
+Network\footnote{\url{cran.r-project.org}} is the largest curated repository of
+R packages. It hosts \CranAvailablePackagesRnd packages with 6 new one submitted
+each day~\cite{Ligges2017}. Packages accepted to CRAN are authored by
+experienced developers and abide to a number of well-formedness rules
+automatically checked on each commit. Furthermore, each package comes with
+sample data. There are three sources of runnable code in a package: \emph{tests,
+examples} and \emph{vignettes} -- respectively, unit tests, code snippets from
+the documentation, and long-form use-cases written in RMarkdown.
+Examples and vignettes can be turned into scripts by extracting the relevant
+code. We obtained \CranPackages that call \eval. These packages represent
+\CranPackagesRatio of all packages or \CranCodeRatio of the entire code size of
+CRAN. From these we extracted \CranRunnableScripts scripts that contain
+\CranRunnableCode lines of code (\CranRunnableCodeExamplesRnd in examples,
+\CranRunnableCodeVignettesRnd in vignettes and \CranRunnableCodeTestsRnd in
+tests).
 
@@
 cluster of three 2.3GHz Intel Xeon 6140 servers, with 72 cores and 256GB of RAM,
-and shared OCFS network storage.
+and shared OCFS network storage. The pipeline steps are:
 
@@
 
-\noindent The pipeline steps are:
 \medskip
 \begin{compactenum}
-\item \emph{Download.} Packages are downloaded automatically from CRAN. For
-  scripts, a web crawler retrieves code and the Kaggle command line tool
-  downloads data. Installation is complicated by native dependencies which
-  are not properly documented and thus hard to automatically resolve.
+\item \emph{Download.} Packages are downloaded from CRAN. For scripts, a web
+  crawler retrieves code and the Kaggle command line tool gets data.
+  Installation is complicated by native dependencies which are not properly
+  documented and thus hard to automatically resolve.
 \item \emph{Extract.} Given installed packages and scripts, the next step is to
@@
 
+\begin{wrapfigure}{r}{5.8cm}
+ \includegraphics[width=0.5\textwidth, trim=4.8cm 0 6cm 0, clip]{size_loaded_distribution}
+\caption{Loaded code} \label{fig:sizedistribution}
+\end{wrapfigure}
+
 To estimate how much executable code is injected through \eval, we measure the
@@
 from the graph for legibility.
-
-\begin{figure}[h!]
- \includegraphics[width=0.5\textwidth, trim=4.8cm 0 6cm 0, clip]{size_loaded_distribution}
-\caption{Loaded code} \label{fig:sizedistribution}
-\end{figure}
 
@@
 \section{The dangers of \eval} \label{sec:dangerseval}
+
+\subsection{Eval Usage in R}
+
+The R language was intended to be extensible, the combination of lazy
+evaluation, \c{substitute} and \c{eval} are the tools given to
+developers to this end. This API is slightly more complex than just
+passing a string, it is conceivable that this may discourage some
+casual users. \Eval is also being used to reduce boilerplate code and
+provide convenience features for programmers. We now give some
+representative examples of its usage.
+
+\mypara{Intercession.} A common use case for \eval is to be combined with
+\c{match.call}. \c{match.call} walks up the call stack, captures the code that
+invoked the currently executing function, and returns it as an unevaluated
+expression. The pattern is to transform a call to some function \c{f} into a
+call to \c{g} with some arguments retained and others modified. As an
+illustration, consider the \c{vcpart} package's function \c{tvcglm} that is
+translated to a call to \c{tvcm} with two modifications to the argument list:
+argument \c{control} can't be missing and \c{fit} is set \c{"glm"}. The function
+ends with a call to \c{eval.parent} to ensure that the rewritten call is
+evaluted in the same environment the original call was.
+\begin{lstlisting}
+ tvcglm <- function(formula, data, family, ...) {
+   k <- match.call()
+   k[[1]] <- as.name("tvcm")
+   if ("fit" %in% names(list(...))) warning("'fit' ignored.")
+   k$fit <- "glm"
+   eval.parent(k)
+ }
+\end{lstlisting}
+This pattern is recognizable by the fact that the expression is a call
+and the target environment is that of the parent.
+
+\mypara{Code Generation.} The more traditional use of \eval is to execute code
+that was assembled by the programmer into a string. Here we show the method
+\c{plot} for class \c{sback} in package \c{wsbackfit}, simplified for
+explanatory purposes. The function takes a long argument list, the names of
+which are captured in the list \c{opt}. The string \c{stub} is composed of a
+subset of the arguments passed to this function; the variable is used to
+construct a call to \c{base::plot} which will draw a plot. \NOTE{This example is
+  a bit messed up. What is var? Does the use of ... in the eval grab the
+  arguments of this call? If yes does it not end up with xlab twice? HELP} The
+\parse and \eval combination is used by the the \source function in the base
+library to load R code from a file in the current workspace.
+\begin{lstlisting}
+ plot.sback <- function(x,...) {
+   opt <- names(list(...))
+   stub <- paste(
+    ifelse("xlab" %in% opt,"",paste(",xlab=\"",var,"\"",sep="")),
+    ifelse("type" %in% opt,"",",type=\"l\""), sep="")
+   plot <- paste("plot(x.data,",stub,",...)",sep="")
+   eval(parse(text=plot))
+\end{lstlisting}
+This pattern is recognizable by its use of \c{parse} to turn a string
+into an expression.
+
+\mypara{Debloating.} \Eval is often used as a means to reduce
+boilerplate code; simple and repetitive code can easily be replaced
+with judcious use of \eval. For example, the \c{data.table} package
+uses \eval to calls the \c{options} function with named arguments
+taken from a vector of strings. While the benefits are limited in this
+example, it is an attractive tool for programmers.
+\begin{lstlisting}
+  opts = c("datatable.verbose"="FALSE", # ...many others
+  for (i in names(opts))
+    eval(parse(text=paste0("options(",i,"=",opts[i],")")))
+\end{lstlisting}
+This pattern is a special case of code generation, recognizable by the
+fact that \eval is executed in a loop.
+
+\mypara{Trivial.} When values are passed to \eval, they are returned
+unchanged. They are an example of trivial uses of \eval. Another
+trivial use is the empty expression, often found in JavaScript, but
+rare in R.
 

### 2021-04-01 paper/main.tex
5

@@
 
+\newpage
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{A Taxonomy of Eval}
@@
 
-\eval has three arguments, two of which we study in more details:
-
-\begin{itemize}
-	\item \c{expr}: the expression to evaluate
-	\item \c{envir}: the environment in which performing the evaluation
-	\item \c{enclos}
-\end{itemize}
-
-\subsubsection{The environment passed to \eval.}
-
-The environment argument can be \c{NULL}, a \c{List} or a data frame. This
-happens in 3.5\% of the cases: \eval copies the fields of the list or data frame
-and creates bindings for them in a new environment. This pattern is used to
-evaluate formulas which can directly refer to the fields of the data. The
-\c{envir} argument can also be a number $n$. It means that the environment in
-which the expressions is evaluated will be the result of \c{sys.call(n)} where
-$n$ refers to the $n$-th stack frame.
-
-The top-level environment in R is called the global environment. New
-environments can be created using \c{new.env}. They can be provided a parent
-environment which becomes the enclosing scope of the new environment.
-
-We looked at the environments passed to all the \eval calls in our corpus.
-Table~\ref{tab:environments} summarizes the results. A numeric environment class
-\c{n} denotes the environment of the $n$-th call stack frame from the current
-function. \c{global} denotes the top-level environment and \c{list} denotes a
-list passed for evaluation of formulas. Environment classes of the form $n+$
-denote the $n$-th environment extended with a new environment. The new
-environment provides a limited form of sandboxing. All assignments using the
-$\leftarrow$ function occur inside it and prevent the extended environment from
-mutation. However, it is still possible to mutate the extended environment using
-the $\leftarrow$ or \c{assign} functions; but, that happens rarely.
-
-\begin{table}[htbp]{ \centering
-\begin{tabular}{c|c|c|c}\hline
-\multicolumn{2}{c|}{Core} & \multicolumn{2}{|c}{Packages}  \\\hline
-Environment Class & Eval Calls \% & Environment Class & Eval Calls\\\hline
-\CoreEnvClassA & \CoreProportionA & \PackagesEnvClassA &  \PackagesProportionA\\
-\CoreEnvClassB & \CoreProportionB & \PackagesEnvClassB &  \PackagesProportionB\\
-\CoreEnvClassC & \CoreProportionC & \PackagesEnvClassC &  \PackagesProportionC\\
-\CoreEnvClassD & \CoreProportionD & \PackagesEnvClassD &  \PackagesProportionD\\
-\CoreEnvClassE & \CoreProportionE & \PackagesEnvClassE &  \PackagesProportionE\\\hline
-\end{tabular}} \label{tab:environments}
-\caption{Environments in terms of \eval calls}
-\end{table}
-
-We observe that a disproportionately high number of core R \eval calls access
-the caller's environment. This is because many core R functions call functions
-that pass the result of \c{parent.frame()} to an \eval. A disproportionately
-high number of calls to \eval happens in an extended top-level environment. This
-can be explained by the fact that many packages evaluate code passed from the
-user's workspace in the top-level environment to access the bindings.
-
-The \c{imchange} function of package \c{imager} makes it possible to modify
-images using a dedicated formula syntax using \~.\footnote{Inspired by {map} in
-package \emph{purr}.} Here, \eval is evaluated in \c{newenv}, which creates a
-new environment that inherits from \c{parent.frame()} by default (classified as
-1+).
-
-\begin{lstlisting}
-	newenv <- new.env()
-	...
-	fo <- parse(text=as.character(fo)[2])
-	im[where] <- eval(fo,envir=newenv,enclos=env)
-\end{lstlisting}
-
-\c{adjCoef} in package \emph{actuar} find the root of an equation defined by a
-function \c{h} whose arguments must be named \c{x} and \c{y}. \c{h} is
-transformed into an auxiliary function \c{h2} that can be optimized. Here, the
-list used for \c{envir} ensures the correspondance between the textual arguments
-of \c{h} and the arguments of \c{h2}.
-
-\begin{lstlisting}
-	sh <- substitute(h)
-	fcall <- paste(sh, "(x, y)")
-	...
-	h2 <- function(x, y)
-	eval(parse(text = fcall),
-	envir = list(x = x, y = y),
-	enclos = parent.frame(2))
-\end{lstlisting}
-
-\subsubsection{The expression passed to \eval.}
-\paragraph{Touching package environments.}
-It refers to the \emph{namespace} of a package. A call to
-\c{library(packageName)}, which load libraries in R, for instance, will populate
-environment \c{package:packageName}. It is a \emph{base} function though, so we
-do not see it in the CRAN dataset. However, R still makes it possible to access
-any package environments and \packageNbPackageEnvSites call sites, \ie
-\packagePackageEnvSitePercent access a package environment, from
-\packageNbPackageEnvPackages different packages. Those packages perform
-introspection, to represent and search any environment by name, in package
-\emph{envnames}, to containerize tests, as with \emph{testthat} and
-\emph{unitizer}, to create a custom OOP system with \emph{R.oo}, or to create
-poor man's hidden functions in \emph{USSCXenaTools}.
-
-It is also used to implement a custom caching mechanism in packages
-\emph{g.data} and \emph{SOAR}.
-
-During tracing, we gather all source expressions and all resolved expression
-arguments passed to \eval. For instance, a non-resolved expression and some of
-its resolved expressions can be:
-
-\begin{lstlisting}
-parse(text = paste0(\"options(\", i, \"=\", opts[i], \")\"))
-
-options(datatable.alloccol = 1024L)
-options(datatable.allow.cartesian = FALSE)
-options(datatable.print.keys = FALSE)
-\end{lstlisting}
+%%%%%%%% JAN:::: NOT SURE WHAT THIS IS
+%% It refers to the \emph{namespace} of a package. A call to
+%% \c{library(packageName)}, which load libraries in R, for instance, will populate
+%% environment \c{package:packageName}. It is a \emph{base} function though, so we
+%% do not see it in the CRAN dataset. However, R still makes it possible to access
+%% any package environments and \packageNbPackageEnvSites call sites, \ie
+%% \packagePackageEnvSitePercent access a package environment, from
+%% \packageNbPackageEnvPackages different packages. Those packages perform
+%% introspection, to represent and search any environment by name, in package
+%% \emph{envnames}, to containerize tests, as with \emph{testthat} and
+%% \emph{unitizer}, to create a custom OOP system with \emph{R.oo}, or to create
+%% poor man's hidden functions in \emph{USSCXenaTools}.
+
+%% It is also used to implement a custom caching mechanism in packages
+%% \emph{g.data} and \emph{SOAR}.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+The argument passed to a call to \eval determines what the function will do. As
+we have seen in the previous section, there are cases where the argument is a
+value, but these are rare. The common cases are that the argument is some form
+of expression, either a single variable, a simple expression or a statement.
+
+To categorize the expressions, consider a function $min(e)$ which for a given
+expression $e$ returns a normal form that abstracts many incidental details
+allowing the reader to focus on the structure.
 
@@
 \begin{table}[h]\small
-\begin{tabular}{|c|c|c|c|c|c|}\hline
-Minimized & \#sites & \%sites & \#packages & \#operations & \%envir \\	\hline
-\texttt{X} & \packageMinimizedcallsitesa  & \packageMinimizedpropsitesa & \packageMinimizedpackagea & \packageMinimizedoperationsaRnd & \packageMinimizedpercentenvirnonparenta \\\hline
-\texttt{F(F())} & \packageMinimizedcallsitesb  & \packageMinimizedpropsitesb & \packageMinimizedpackageb  & \packageMinimizedoperationsbRnd & \packageMinimizedpercentenvirnonparentb \\\hline
-\texttt{V} & \packageMinimizedcallsitesc & \packageMinimizedpropsitesc & \packageMinimizedpackagec & \packageMinimizedoperationscRnd & \packageMinimizedpercentenvirnonparentc \\\hline
-\texttt{F(X)}& \packageMinimizedcallsitesd & \packageMinimizedpropsitesd & \packageMinimizedpackaged & \packageMinimizedoperationsdRnd & \packageMinimizedpercentenvirnonparentd \\\hline
-\texttt{model.frame} & \packageMinimizedcallsitese & \packageMinimizedpropsitese & \packageMinimizedpackagee & \packageMinimizedoperationseRnd & \packageMinimizedpercentenvirnonparente \\\hline
-\texttt{F()}& \packageMinimizedcallsitesf & \packageMinimizedpropsitesf & \packageMinimizedpackagef & \packageMinimizedoperationsfRnd & \packageMinimizedpercentenvirnonparentf \\\hline
-\texttt{\$}& \packageMinimizedcallsitesg & \packageMinimizedpropsitesg & \packageMinimizedpackageg & \packageMinimizedoperationsgRnd & \packageMinimizedpercentenvirnonparentg \\\hline
-\texttt{FUN} & \packageMinimizedcallsitesh & \packageMinimizedpropsitesh & \packageMinimizedpackageh & \packageMinimizedoperationshRnd & \packageMinimizedpercentenvirnonparenth \\\hline
-\texttt{<-} & \packageMinimizedcallsitesi  & \packageMinimizedpropsitesi & \packageMinimizedpackagei & \packageMinimizedoperationsiRnd & \packageMinimizedpercentenvirnonparenti \\\hline
-\texttt{\{BLOCK\}} & \packageMinimizedcallsitesj & \packageMinimizedpropsitesj & \packageMinimizedpackagej & \packageMinimizedoperationsjRnd & \packageMinimizedpercentenvirnonparentj \\\hline
-\texttt{F(F(<-))} & \packageMinimizedcallsitesk & \packageMinimizedpropsitesk & \packageMinimizedpackagek & \packageMinimizedoperationskRnd & \packageMinimizedpercentenvirnonparentk \\\hline
+\begin{tabular}{|c|r|r|r|r|r|c|}\hline
+$min(e)$& \#sites & \%sites & \#packages & \#operations & \%envir & example\\\hline
+\c X&\packageMinimizedcallsitesa &\packageMinimizedpropsitesa &\packageMinimizedpackagea &\packageMinimizedoperationsaRnd &\packageMinimizedpercentenvirnonparenta & \c{y+1}\\\hline
+\c{F(F())} & \packageMinimizedcallsitesb  & \packageMinimizedpropsitesb & \packageMinimizedpackageb  & \packageMinimizedoperationsbRnd & \packageMinimizedpercentenvirnonparentb & \c{gbov( mean(x), a-1)}\\\hline
+\c{V}&\packageMinimizedcallsitesc &\packageMinimizedpropsitesc &\packageMinimizedpackagec &\packageMinimizedoperationscRnd &\packageMinimizedpercentenvirnonparentc&c(42,21,0)\\\hline
+\c{F(X)}& \packageMinimizedcallsitesd & \packageMinimizedpropsitesd & \packageMinimizedpackaged & \packageMinimizedoperationsdRnd & \packageMinimizedpercentenvirnonparentd \\\hline
+\c{model.frame} & \packageMinimizedcallsitese & \packageMinimizedpropsitese & \packageMinimizedpackagee & \packageMinimizedoperationseRnd & \packageMinimizedpercentenvirnonparente \\\hline
+\c{F()}& \packageMinimizedcallsitesf & \packageMinimizedpropsitesf & \packageMinimizedpackagef & \packageMinimizedoperationsfRnd & \packageMinimizedpercentenvirnonparentf \\\hline
+\c{\$}& \packageMinimizedcallsitesg & \packageMinimizedpropsitesg & \packageMinimizedpackageg & \packageMinimizedoperationsgRnd & \packageMinimizedpercentenvirnonparentg \\\hline
+\c{FUN} & \packageMinimizedcallsitesh & \packageMinimizedpropsitesh & \packageMinimizedpackageh & \packageMinimizedoperationshRnd & \packageMinimizedpercentenvirnonparenth \\\hline
+\c{<-} & \packageMinimizedcallsitesi  & \packageMinimizedpropsitesi & \packageMinimizedpackagei & \packageMinimizedoperationsiRnd & \packageMinimizedpercentenvirnonparenti \\\hline
+\c{\{\}} & \packageMinimizedcallsitesj & \packageMinimizedpropsitesj & \packageMinimizedpackagej & \packageMinimizedoperationsjRnd & \packageMinimizedpercentenvirnonparentj \\\hline
+\c{F(F(<-))} & \packageMinimizedcallsitesk & \packageMinimizedpropsitesk & \packageMinimizedpackagek & \packageMinimizedoperationskRnd & \packageMinimizedpercentenvirnonparentk \\\hline
 \end{tabular}
@@
 \begin{itemize}
-	\item \eval is passed an actual value, a double, an integer, an environment, a closure and other simple types,
-	\item \eval is passed a language expression that trivially evaluates to a value, for instance, \c{quote(1)}, \c{quote(1 + 1)}, \quote{c("a", "b", "c")}, and does not include variables, so does not depend on a specific environment.
+\item \eval is passed an actual value, a double, an integer, an environment, a
+  closure and other simple types,
+\item \eval is passed a language expression that trivially evaluates to a value,
+  for instance, \c{quote(1)}, \c{quote(1 + 1)}, \quote{c("a", "b", "c")}, and
+  does not include variables, so does not depend on a specific environment.
 \end{itemize}
@@
 
-\paragraph{\c{model.frame}} This is a function used to create statistical models. We explain it in more details in the section about the usage of \eval (see Section~REF).
+\paragraph{\c{model.frame}}
+This is a function used to create statistical models. We explain it in more
+details in the section about the usage of \eval (see Section~REF).
 
@@
 
-\paragraph{Base.} The ranking of minimized expressions in \emph{base} is similar to the one in CRAN bu the share of \c{V} is much larger, as it is present in \baseMinimizedpropsitesa\% of the sites and \c{X} is a distant second, in \baseMinimizedpropsitesb\% of the sites.
-
-\paragraph{Kaggle.} Kaggle is dominated by sites with minimized expression \c{V}, with \kaggleMinimizedcallsitesa sites, and  \kaggleMinimizedpropsitesa\% of the \kaggleNbCallSites sites in total.
-All of those values are an AST size of one node, suggesting they do not come.
-\NOTE{TODO}
+\subsubsection{The environment passed to \eval.}
+
+The environment argument can be \c{NULL}, a \c{List} or a data frame. This
+happens in 3.5\% of the cases: \eval copies the fields of the list or data frame
+and creates bindings for them in a new environment. This pattern is used to
+evaluate formulas which can directly refer to the fields of the data. The
+\c{envir} argument can also be a number $n$. It means that the environment in
+which the expressions is evaluated will be the result of \c{sys.call(n)} where
+$n$ refers to the $n$-th stack frame.
+
+The top-level environment in R is called the global environment. New
+environments can be created using \c{new.env}. They can be provided a parent
+environment which becomes the enclosing scope of the new environment.
+
+We looked at the environments passed to all the \eval calls in our corpus.
+Table~\ref{tab:environments} summarizes the results. A numeric environment class
+\c{n} denotes the environment of the $n$-th call stack frame from the current
+function. \c{global} denotes the top-level environment and \c{list} denotes a
+list passed for evaluation of formulas. Environment classes of the form $n+$
+denote the $n$-th environment extended with a new environment. The new
+environment provides a limited form of sandboxing. All assignments using the
+$\leftarrow$ function occur inside it and prevent the extended environment from
+mutation. However, it is still possible to mutate the extended environment using
+the $\leftarrow$ or \c{assign} functions; but, that happens rarely.
+
+\begin{table}[htbp]{ \centering
+\begin{tabular}{c|c|c|c}\hline
+\multicolumn{2}{c|}{Core} & \multicolumn{2}{|c}{Packages}  \\\hline
+Environment Class & Eval Calls \% & Environment Class & Eval Calls\\\hline
+\CoreEnvClassA & \CoreProportionA & \PackagesEnvClassA &  \PackagesProportionA\\
+\CoreEnvClassB & \CoreProportionB & \PackagesEnvClassB &  \PackagesProportionB\\
+\CoreEnvClassC & \CoreProportionC & \PackagesEnvClassC &  \PackagesProportionC\\
+\CoreEnvClassD & \CoreProportionD & \PackagesEnvClassD &  \PackagesProportionD\\
+\CoreEnvClassE & \CoreProportionE & \PackagesEnvClassE &  \PackagesProportionE\\\hline
+\end{tabular}} \label{tab:environments}
+\caption{Environments in terms of \eval calls}
+\end{table}
+
+We observe that a disproportionately high number of core R \eval calls access
+the caller's environment. This is because many core R functions call functions
+that pass the result of \c{parent.frame()} to an \eval. A disproportionately
+high number of calls to \eval happens in an extended top-level environment. This
+can be explained by the fact that many packages evaluate code passed from the
+user's workspace in the top-level environment to access the bindings.
+
+The \c{imchange} function of package \c{imager} makes it possible to modify
+images using a dedicated formula syntax using \~.\footnote{Inspired by {map} in
+package \emph{purr}.} Here, \eval is evaluated in \c{newenv}, which creates a
+new environment that inherits from \c{parent.frame()} by default (classified as
+1+).
+
+\begin{lstlisting}
+ newenv <- new.env()
+ ...
+ fo <- parse(text=as.character(fo)[2])
+ im[where] <- eval(fo,envir=newenv,enclos=env)
+\end{lstlisting}
+
+\c{adjCoef} in package \emph{actuar} find the root of an equation defined by a
+function \c{h} whose arguments must be named \c{x} and \c{y}. \c{h} is
+transformed into an auxiliary function \c{h2} that can be optimized. Here, the
+list used for \c{envir} ensures the correspondance between the textual arguments
+of \c{h} and the arguments of \c{h2}.
+
+\begin{lstlisting}
+ sh <- substitute(h)
+ fcall <- paste(sh, "(x, y)")
+ ...
+ h2 <- function(x, y)
+ eval(parse(text = fcall),
+ envir = list(x = x, y = y),
+ enclos = parent.frame(2))
+\end{lstlisting}
+
+\subsubsection{The expression passed to \eval.}
+\paragraph{Touching package environments.}
+
+\paragraph{Base}
+The ranking of minimized expressions in \emph{base} is similar to the one in
+CRAN bu the share of \c{V} is much larger, as it is present in
+\baseMinimizedpropsitesa\% of the sites and \c{X} is a distant second, in
+\baseMinimizedpropsitesb\% of the sites.
+
+\paragraph{Kaggle}
+Kaggle is dominated by sites with minimized expression \c{V}, with
+\kaggleMinimizedcallsitesa sites, and \kaggleMinimizedpropsitesa\% of the
+\kaggleNbCallSites sites in total. All of those values are an AST size of one
+node, suggesting they do not come. \NOTE{TODO}
 %TODO: all the src ref here seem to be incorrect and not from kaggle.

### 2021-04-01 paper/main.tex
Add examples for normalized expressions

@@
 \c{F(F())} & \packageMinimizedcallsitesb  & \packageMinimizedpropsitesb & \packageMinimizedpackageb  & \packageMinimizedoperationsbRnd & \packageMinimizedpercentenvirnonparentb & \c{gbov( mean(x), a-1)}\\\hline
-\c{V}&\packageMinimizedcallsitesc &\packageMinimizedpropsitesc &\packageMinimizedpackagec &\packageMinimizedoperationscRnd &\packageMinimizedpercentenvirnonparentc&c(42,21,0)\\\hline
-\c{F(X)}& \packageMinimizedcallsitesd & \packageMinimizedpropsitesd & \packageMinimizedpackaged & \packageMinimizedoperationsdRnd & \packageMinimizedpercentenvirnonparentd \\\hline
-\c{model.frame} & \packageMinimizedcallsitese & \packageMinimizedpropsitese & \packageMinimizedpackagee & \packageMinimizedoperationseRnd & \packageMinimizedpercentenvirnonparente \\\hline
-\c{F()}& \packageMinimizedcallsitesf & \packageMinimizedpropsitesf & \packageMinimizedpackagef & \packageMinimizedoperationsfRnd & \packageMinimizedpercentenvirnonparentf \\\hline
-\c{\$}& \packageMinimizedcallsitesg & \packageMinimizedpropsitesg & \packageMinimizedpackageg & \packageMinimizedoperationsgRnd & \packageMinimizedpercentenvirnonparentg \\\hline
-\c{FUN} & \packageMinimizedcallsitesh & \packageMinimizedpropsitesh & \packageMinimizedpackageh & \packageMinimizedoperationshRnd & \packageMinimizedpercentenvirnonparenth \\\hline
-\c{<-} & \packageMinimizedcallsitesi  & \packageMinimizedpropsitesi & \packageMinimizedpackagei & \packageMinimizedoperationsiRnd & \packageMinimizedpercentenvirnonparenti \\\hline
-\c{\{\}} & \packageMinimizedcallsitesj & \packageMinimizedpropsitesj & \packageMinimizedpackagej & \packageMinimizedoperationsjRnd & \packageMinimizedpercentenvirnonparentj \\\hline
-\c{F(F(<-))} & \packageMinimizedcallsitesk & \packageMinimizedpropsitesk & \packageMinimizedpackagek & \packageMinimizedoperationskRnd & \packageMinimizedpercentenvirnonparentk \\\hline
+\c{V}&\packageMinimizedcallsitesc &\packageMinimizedpropsitesc &\packageMinimizedpackagec &\packageMinimizedoperationscRnd &\packageMinimizedpercentenvirnonparentc& \c{c(42,21,0)}\\\hline
+\c{F(X)}& \packageMinimizedcallsitesd & \packageMinimizedpropsitesd & \packageMinimizedpackaged & \packageMinimizedoperationsdRnd & \packageMinimizedpercentenvirnonparentd & \c{seq\_len(iters)} \\\hline
+\c{model.frame} & \packageMinimizedcallsitese & \packageMinimizedpropsitese & \packageMinimizedpackagee & \packageMinimizedoperationseRnd & \packageMinimizedpercentenvirnonparente & \c{model.frame(formula = Z ~ U)} \\\hline
+\c{F()}& \packageMinimizedcallsitesf & \packageMinimizedpropsitesf & \packageMinimizedpackagef & \packageMinimizedoperationsfRnd & \packageMinimizedpercentenvirnonparentf & \c{rgamma(3, 2, n = 10L)} \\\hline
+\c{\$}& \packageMinimizedcallsitesg & \packageMinimizedpropsitesg & \packageMinimizedpackageg & \packageMinimizedoperationsgRnd & \packageMinimizedpercentenvirnonparentg & \c{DF\$B} \\\hline
+\c{FUN} & \packageMinimizedcallsitesh & \packageMinimizedpropsitesh & \packageMinimizedpackageh & \packageMinimizedoperationshRnd & \packageMinimizedpercentenvirnonparenth & \c{function(x, y) x + 3 * y} \\\hline
+\c{<-} & \packageMinimizedcallsitesi  & \packageMinimizedpropsitesi & \packageMinimizedpackagei & \packageMinimizedoperationsiRnd & \packageMinimizedpercentenvirnonparenti & \c{x[1, 2:3, 2:3] <- value}\\\hline
+\c{\{\}} & \packageMinimizedcallsitesj & \packageMinimizedpropsitesj & \packageMinimizedpackagej & \packageMinimizedoperationsjRnd & \packageMinimizedpercentenvirnonparentj & \c{\{ write.csv(iris, tf) \textbackslash n file.size(tf) \}} \\\hline
+\c{F(F(<-))} & \packageMinimizedcallsitesk & \packageMinimizedpropsitesk & \packageMinimizedpackagek & \packageMinimizedoperationskRnd & \packageMinimizedpercentenvirnonparentk & \c{opt\$h.weights <- rep(1, length(x))} \\\hline
 \end{tabular}
@@
 \paragraph{Large expressions.}  Compositions of functions, without variables
-(\c{F(F(X))} minimized expression) is the second most frequent minimized
-expression. \c{\{BLOCK\}} is a sequence of statements. It is usually large, with
+(\c{F(F())} minimized expression) is the second most frequent minimized
+expression. \c{\{\}} is a sequence of statements. It is usually large, with
 in average \packageMinimizedoperationsjRnd operations in it.

### 2021-04-01 paper/main.tex
sttuff

@@
 expression $e$ returns a normal form that abstracts many incidental details
-allowing the reader to focus on the structure.
-
-We classify the resolved expressions to take into account the structure of the
-expression passed to \eval. However, we do not want to discriminate expressions
-based on specific values. Hence, the expression goes through several
-simplification steps: constant folding of arithmetic and string expressions that
-use commonly used functions (addition, string concatenation), discarding of
-specific value to a symbolic value holder, variable absorption. We keep track of
-variables, as they are evaluated in the environment specified to \eval, contrary
-to values that are left unchanged. We also track whether assignment operators
-are used in the expression or not. After the simplification steps, we obtain 15
-different \emph{minimized} expressions, of which we show the eleven first most
-frequent in number of sites in Table~\ref{tab:minimizedexpressions}.
-
-\begin{table}[h]\small
-\begin{tabular}{|c|r|r|r|r|r|c|}\hline
-$min(e)$& \#sites & \%sites & \#packages & \#operations & \%envir & example\\\hline
-\c X&\packageMinimizedcallsitesa &\packageMinimizedpropsitesa &\packageMinimizedpackagea &\packageMinimizedoperationsaRnd &\packageMinimizedpercentenvirnonparenta & \c{y+1}\\\hline
-\c{F(F())} & \packageMinimizedcallsitesb  & \packageMinimizedpropsitesb & \packageMinimizedpackageb  & \packageMinimizedoperationsbRnd & \packageMinimizedpercentenvirnonparentb & \c{gbov( mean(x), a-1)}\\\hline
-\c{V}&\packageMinimizedcallsitesc &\packageMinimizedpropsitesc &\packageMinimizedpackagec &\packageMinimizedoperationscRnd &\packageMinimizedpercentenvirnonparentc& \c{c(42,21,0)}\\\hline
-\c{F(X)}& \packageMinimizedcallsitesd & \packageMinimizedpropsitesd & \packageMinimizedpackaged & \packageMinimizedoperationsdRnd & \packageMinimizedpercentenvirnonparentd & \c{seq\_len(iters)} \\\hline
-\c{model.frame} & \packageMinimizedcallsitese & \packageMinimizedpropsitese & \packageMinimizedpackagee & \packageMinimizedoperationseRnd & \packageMinimizedpercentenvirnonparente & \c{model.frame(formula = Z ~ U)} \\\hline
-\c{F()}& \packageMinimizedcallsitesf & \packageMinimizedpropsitesf & \packageMinimizedpackagef & \packageMinimizedoperationsfRnd & \packageMinimizedpercentenvirnonparentf & \c{rgamma(3, 2, n = 10L)} \\\hline
-\c{\$}& \packageMinimizedcallsitesg & \packageMinimizedpropsitesg & \packageMinimizedpackageg & \packageMinimizedoperationsgRnd & \packageMinimizedpercentenvirnonparentg & \c{DF\$B} \\\hline
-\c{FUN} & \packageMinimizedcallsitesh & \packageMinimizedpropsitesh & \packageMinimizedpackageh & \packageMinimizedoperationshRnd & \packageMinimizedpercentenvirnonparenth & \c{function(x, y) x + 3 * y} \\\hline
-\c{<-} & \packageMinimizedcallsitesi  & \packageMinimizedpropsitesi & \packageMinimizedpackagei & \packageMinimizedoperationsiRnd & \packageMinimizedpercentenvirnonparenti & \c{x[1, 2:3, 2:3] <- value}\\\hline
-\c{\{\}} & \packageMinimizedcallsitesj & \packageMinimizedpropsitesj & \packageMinimizedpackagej & \packageMinimizedoperationsjRnd & \packageMinimizedpercentenvirnonparentj & \c{\{ write.csv(iris, tf) \textbackslash n file.size(tf) \}} \\\hline
-\c{F(F(<-))} & \packageMinimizedcallsitesk & \packageMinimizedpropsitesk & \packageMinimizedpackagek & \packageMinimizedoperationskRnd & \packageMinimizedpercentenvirnonparentk & \c{opt\$h.weights <- rep(1, length(x))} \\\hline
-\end{tabular}
-\caption{Minimized expressions: eleven most frequent in number of sites. \#sites
-  is the number of sites; \%sites the percentage of sites across the total
-  number of sites (\packageNbCallSites sites) with at least this minimized
-  expression; \#packages is the number of packages with at least one of the
-  minimized expression and measures how widespread a minimized expression is.
-  \#operations is the average number of steps performed by the R interpreter
-  when given that minimized expression, and \%envir is the percentage of call
-  sites for which the \c{envir} argument is explicitly passed, and is not
-  \c{parent.frame()}, which is the most common explicit environment passed to
-  \eval.} \label{tab:minimizedexpressions}
-\end{table}
-
-Note that a site can exhibit several minimized expressions (see
-Figure~\ref{fig:frequentcombinationminimized}). However,
-\packageNbOneMinimizedPercent of the sites have only one minimized expression.
-Note also that if a function is called in that expression, we see only the
-function call but not the body of the function. We will trace what happens
-during the evaluation of the expression by \eval in
-Section~\ref{sec:dangerseval}, especially about side effects performed in \eval.
-
-\begin{figure}[htbp]
-	\centering
-	\includegraphics[width=\textwidth]{package_combination_minimized.pdf}
-	\caption{Most frequent combinations of minimized expressions per site,
-          with a cumulative share of 95\% of the
-          sites.} \label{fig:frequentcombinationminimized}
-\end{figure}
-
-\paragraph{Variables.}
-The most common minimized expression is a variable.\footnote{An arithmetic
-expression with some variables and constant, \eg \c{1 + x}, will also be
-normalized to \c{X}, i.e. a variable.} It happens in
-\packageMinimizedpropsitesa\% of the sites and is widespread across packages as
-it is found in \packageMinimizedpackagea packages across our CRAN corpus.
-\packageNbSymbolVarSitePercent are only symbols, \ie not complex expressions.
-The \c{envir} argument is explicitly passed in \packageMinimizedpercentenvira\%
-of the call sites. The call sites without an explicit \c{envir} argument all use
-a specific \eval variant, \c{eval.parent}. Moreover, the \c{envir} argument is
-both explicit and different from \c{parent.frame()} in
-\packageMinimizedpercentenvirnonparenta\% of the sites, which suggests that when
-given a variable, \eval is often used to get value in a non-local, non-parent
-environment.
+allowing the reader to focus on the structure of the evaluated code. The
+minimization function performs several simplifications steps. It performs
+constant folding of arithmetic and string expressions for base operators, e.g.
+$min(\c{1+1})=\c{V}$, value simplification, $min(\c{c(1,2,3+2)})=\c{V}$,
+variable absorption, $min(\c{x+y})=\c{X}$, function absorption,
+$min(\c{g(f(x),h(z))})=\c{F(F(X))}$ and a number of other simplification. There
+are 15 normal forms. We detail the interesting cases next.
+
+\begin{itemize}
+\item \underline{$min(e)=\c{V}$}: Expression that represent values are
+  frequently passed to \eval. The majority of those, \packageValOneNodePercent,
+  are inline constants (integers or doubles). The rest trivially evaluate to a
+  value.\footnote{True as long as base functions such as \c{+} are not
+  redefined. They typically aren't, but this is a limitation of this
+  categorization.} In our corpus, \packageNbCallSitesUniqueActualValue call
+  sites only ever see a simple values; the median number of times they are
+  exercised is \packageMedianRunSitesUniqueActualValue, suggesting a possible
+  lack of coverage. However, manual inspection reveals cases such as:
+\begin{lstlisting}
+ f1 <- eval(paste("A ~ ",paste(paste(names(X[,-1])),collapse="+")))
+\end{lstlisting}
+  The code concatenates strings and \eval will simply return the string as is.
+  From twenty randomly selected \evals, only \packageUsefulValueEvalPercent of
+  them needed to use \eval.
+%% The percentage of non default environment passed to \eval is
+%% \packageMinimizedpercentenvirc when the minimized expression is a value. Some
+%% sites which are passed a value, are also passed different minimized expressions.
+%% For those ones, \packageNonDefaultEnvirValuePercent use a non-default
+%% environment, and among the one that use a non-default environment,
+%% \packageNonDefaultEnvirWithVarPercent have at least one minimized expression
+%% with a variable inside.
+
+\item \underline{$min(e)=\c{X}$}: Variables are the most common normal form.
+  Variables subsume values, so they include arithmetic expressions in which a
+  variable occurs. \packageNbSymbolVarSitePercent of variable expression are
+  symbols, \ie just an \c{x}. A variable may refer to a promise and thus trigger
+  computation the first time it is accessed.
+  %% The
+  %% \c{envir} argument is explicitly passed in \packageMinimizedpercentenvira\% of
+  %% the call sites. The call sites without an explicit \c{envir} argument all use
+  %% a specific \eval variant, \c{eval.parent}. Moreover, the \c{envir} argument is
+  %% both explicit and different from \c{parent.frame()} in
+  %% \packageMinimizedpercentenvirnonparenta\% of the sites, which suggests that
+  %% when given a variable, \eval is often used to get value in a non-local,
+  %% non-parent environment.
+\item \underline{$min(e)=\c{\$}$}: A named lookup, written \c{x\$f}, can be used
+  to look up a name in an environment or a column in a data frame. This category
+  also groups indexing operators, \c{[]} and \c{[[]]}. The presence of these
+  operators subsumes constants and variables.
+%% It amounts for \packageNbSlotAccess sites, \ie
+%% \packageSlotAccessPercent of the sites. The non-default, non-parent environment
+%% percentage is here \packageMinimizedpercentenvirnonparentg.
+
+\item \underline{$min(e)=$~\c{<-}}: Assignments amount to
+  \packageAssignSitesPercent of the sites, they include \c{<-}, \c{<<-} and
+  \c{assign}, and subsume named lookups.
+%% \packageNonDefaultEnvirAssignSitesPercent of the sites with assignments. Indeed,
+%% most of those assignments occur in the default environment, \ie in the caller of
+%% \eval. This hints that side effects are not widespread in \eval.\footnote{Side
+%% effects can also happen in a function call in the \eval expression, and that
+%% function call will hide assignment operators from us. We more precisely
+%% instrument side-effects to environment in Section~REF.}
+
+\item \underline{$min(e)=\c{F()}$}: Function calls that
+
+\item \underline{$min(e)=\c{F(X)}$}: Function calls that
+
+\item \underline{$min(e)=\c{F(F(X))}$}: Function calls that
+
+\item \underline{$min(e)=\c{FUN}$}: Function calls that
+
+\item \underline{$min(e)=\c{\{\,\}}$}: Function calls that
+
+\item \underline{$min(e)=\c{model.frame}$}: Function calls that
+
+\end{itemize}
 
@@
 
-Minimized expression \texttt{\$} refers to a slot access, with \texttt{\$},
-\texttt{[]} or \texttt{[[]]}, as the top call in a simple expression, for
-instance \texttt{a\$slot + 1}. It amounts for \packageNbSlotAccess sites, \ie
-\packageSlotAccessPercent of the sites. The non-default, non-parent environment
-percentage is here \packageMinimizedpercentenvirnonparentg.
-
 As a whole, variable access, either as the top call, or inside a function call,
@@
 \packageGeneralizedVarAccessSitePercent of the sites.
-
-\paragraph{Values.}
-Many sites evaluates to \c{V}. \packageValOneNodePercent of those sites are only
-passed an expression with one node, \ie one value, not a more complex expression
-that is later constant folded. \c{V} corresponds to two main categories of
-expressions passed to \eval:
-
-\begin{itemize}
-\item \eval is passed an actual value, a double, an integer, an environment, a
-  closure and other simple types,
-\item \eval is passed a language expression that trivially evaluates to a value,
-  for instance, \c{quote(1)}, \c{quote(1 + 1)}, \quote{c("a", "b", "c")}, and
-  does not include variables, so does not depend on a specific environment.
-\end{itemize}
-
-For the first category, \eval is not necessary. In
-\packageNbCallSitesUniqueActualValue sites exhibiting \c{V}, \ie
-\packageCallSitesUniqueActualValuePercent\% sites, it is only passed such actual
-values. The median number of runs for those sites is
-\packageMedianRunSitesUniqueActualValue (average
-\packageAverageRunSitesUniqueActualValue, standard deviation
-\packageSdRunSitesUniqueActualValue), suggesting it happens because of a lack of
-coverage, because there are rarely triggered in our dataset and all paths are
-not taken. However, manual inspection of the source code of some examples shows
-that some sites actually pass values. For instance, in package \emph{ctmle}, in
-function \c{evaluate\_candidate\_glmnet}:
-
-\begin{lstlisting}
-f1 <- eval(paste("A ~ ", paste(paste(names(X[,-1]), sep=''), collapse=" + ")))
-f2 <- as.formula(f1)
-\end{lstlisting}
-
-\c{paste} concatenates strings and \eval will just return that string.
-\c{as.formula} can take a string and convert it to a formula; \eval is not
-needed here. In 20 random sites among the ones with values, a conservative
-manual inspection showed that \packageUsefulValueEvalPercent of the sites had a
-legitimate reason to use \eval.
-
-The percentage of non default environment passed to \eval is
-\packageMinimizedpercentenvirc when the minimized expression is a value. Some
-sites which are passed a value, are also passed different minimized expressions.
-For those ones, \packageNonDefaultEnvirValuePercent use a non-default
-environment, and among the one that use a non-default environment,
-\packageNonDefaultEnvirWithVarPercent have at least one minimized expression
-with a variable inside.
 
@@
 
-\paragraph{Assignments.}
-This amounts to \packageAssignSitesPercent of the sites and corresponds to
-minimized expressions \c{<-}, where the assignment is the top call, and
-\c{F(F(<-))}, where the assignment happens inside, and others that we were not
-enough present to be shown in table~\ref{tab:minimizedexpressions}. Assignments
-are performed with \c{<-}, \c{<<-} or \c{assign}. Most often, assignment happens
-in top call position. A non-default environment is specified only in
-\packageNonDefaultEnvirAssignSitesPercent of the sites with assignments. Indeed,
-most of those assignments occur in the default environment, \ie in the caller of
-\eval. This hints that side effects are not widespread in \eval.\footnote{Side
-effects can also happen in a function call in the \eval expression, and that
-function call will hide assignment operators from us. We more precisely
-instrument side-effects to environment in Section~REF.}
-
 \paragraph{Large expressions.}  Compositions of functions, without variables
@@
 evaluation environment.
+
+Table~\ref{tab:minimizedexpressions} gives eleven most frequent in number of
+sites. \#sites is the number of sites; \%sites the percentage of sites across
+the total number of sites (\packageNbCallSites sites) with at least this
+minimized expression; \#packages is the number of packages with at least one of
+the minimized expression and measures how widespread a minimized expression is.
+\#operations is the average number of steps performed by the R interpreter when
+given that minimized expression, and \%envir is the percentage of call sites for
+which the \c{envir} argument is explicitly passed, and is not
+\c{parent.frame()}, which is the most common explicit environment passed to
+\eval.
+
+\begin{table}[h]\small
+\begin{tabular}{|c|r|r|r|r|r|c|}\hline
+$min(e)$& \#sites & \%sites & \#packages & \#operations & \%envir & example\\\hline
+\c X&\packageMinimizedcallsitesa &\packageMinimizedpropsitesa &\packageMinimizedpackagea &\packageMinimizedoperationsaRnd &\packageMinimizedpercentenvirnonparenta & \c{y+1}\\\hline
+\c{F(F(X))} & \packageMinimizedcallsitesb  & \packageMinimizedpropsitesb & \packageMinimizedpackageb  & \packageMinimizedoperationsbRnd & \packageMinimizedpercentenvirnonparentb & \c{gbov( mean(x), a-1)}\\\hline
+\c{V}&\packageMinimizedcallsitesc &\packageMinimizedpropsitesc &\packageMinimizedpackagec &\packageMinimizedoperationscRnd &\packageMinimizedpercentenvirnonparentc& \c{c(42,21,0)}\\\hline
+\c{F(X)}& \packageMinimizedcallsitesd & \packageMinimizedpropsitesd & \packageMinimizedpackaged & \packageMinimizedoperationsdRnd & \packageMinimizedpercentenvirnonparentd & \c{seq\_len(iters)} \\\hline
+\c{model.frame} & \packageMinimizedcallsitese & \packageMinimizedpropsitese & \packageMinimizedpackagee & \packageMinimizedoperationseRnd & \packageMinimizedpercentenvirnonparente & \c{model.frame(formula = Z ~ U)} \\\hline
+\c{F()}& \packageMinimizedcallsitesf & \packageMinimizedpropsitesf & \packageMinimizedpackagef & \packageMinimizedoperationsfRnd & \packageMinimizedpercentenvirnonparentf & \c{rgamma(3, 2, n = 10L)} \\\hline
+\c{\$}& \packageMinimizedcallsitesg & \packageMinimizedpropsitesg & \packageMinimizedpackageg & \packageMinimizedoperationsgRnd & \packageMinimizedpercentenvirnonparentg & \c{DF\$B} \\\hline
+\c{FUN} & \packageMinimizedcallsitesh & \packageMinimizedpropsitesh & \packageMinimizedpackageh & \packageMinimizedoperationshRnd & \packageMinimizedpercentenvirnonparenth & \c{function(x, y) x + 3 * y} \\\hline
+\c{<-} & \packageMinimizedcallsitesi  & \packageMinimizedpropsitesi & \packageMinimizedpackagei & \packageMinimizedoperationsiRnd & \packageMinimizedpercentenvirnonparenti & \c{x[1, 2:3, 2:3] <- value}\\\hline
+\c{\{\}} & \packageMinimizedcallsitesj & \packageMinimizedpropsitesj & \packageMinimizedpackagej & \packageMinimizedoperationsjRnd & \packageMinimizedpercentenvirnonparentj & \c{\{ write.csv(iris, tf) \textbackslash n file.size(tf) \}} \\\hline
+\c{F(F(<-))} & \packageMinimizedcallsitesk & \packageMinimizedpropsitesk & \packageMinimizedpackagek & \packageMinimizedoperationskRnd & \packageMinimizedpercentenvirnonparentk & \c{opt\$h.weights <- rep(1, length(x))} \\\hline
+\end{tabular}
+\caption{Minimized expressions} \label{tab:minimizedexpressions}
+\end{table}
+
+Note that a site can exhibit several minimized expressions (see
+Figure~\ref{fig:frequentcombinationminimized}). However,
+\packageNbOneMinimizedPercent of the sites have only one minimized expression.
+Note also that if a function is called in that expression, we see only the
+function call but not the body of the function. We will trace what happens
+during the evaluation of the expression by \eval in
+Section~\ref{sec:dangerseval}, especially about side effects performed in \eval.
 

### 2021-04-01 paper/main.tex
Environments

@@
 \email{r@s.com}
-\affiliation{\institution{Institute}}
+\affiliation{\institution{Institute} \country{Country}}
 
@@
 %%%%%%%% JAN:::: NOT SURE WHAT THIS IS
-%% It refers to the \emph{namespace} of a package. A call to
-%% \c{library(packageName)}, which load libraries in R, for instance, will populate
-%% environment \c{package:packageName}. It is a \emph{base} function though, so we
-%% do not see it in the CRAN dataset. However, R still makes it possible to access
-%% any package environments and \packageNbPackageEnvSites call sites, \ie
-%% \packagePackageEnvSitePercent access a package environment, from
-%% \packageNbPackageEnvPackages different packages. Those packages perform
-%% introspection, to represent and search any environment by name, in package
-%% \emph{envnames}, to containerize tests, as with \emph{testthat} and
-%% \emph{unitizer}, to create a custom OOP system with \emph{R.oo}, or to create
-%% poor man's hidden functions in \emph{USSCXenaTools}.
-
-%% It is also used to implement a custom caching mechanism in packages
-%% \emph{g.data} and \emph{SOAR}.
+% The namespace stuff of packages was meant for the description of the envir argument.
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
+\paragraph{Base}
+The ranking of minimized expressions in \emph{base} is similar to the one in
+CRAN but the share of \c{V} is much larger, as it is present in
+\baseMinimizedpropsitesa\% of the sites and \c{X} is a distant second, in
+\baseMinimizedpropsitesb\% of the sites.
+
+\paragraph{Kaggle}
+Kaggle is dominated by sites with minimized expression \c{V}, with
+\kaggleMinimizedcallsitesa sites, and \kaggleMinimizedpropsitesa\% of the
+\kaggleNbCallSites sites in total. All of those values are an AST size of one
+node, suggesting they do not come. \NOTE{TODO}
+%TODO: all the src ref here seem to be incorrect and not from kaggle.
+
 \subsubsection{The environment passed to \eval.}
@@
 
-\subsubsection{The expression passed to \eval.}
-\paragraph{Touching package environments.}
-
-\paragraph{Base}
-The ranking of minimized expressions in \emph{base} is similar to the one in
-CRAN bu the share of \c{V} is much larger, as it is present in
-\baseMinimizedpropsitesa\% of the sites and \c{X} is a distant second, in
-\baseMinimizedpropsitesb\% of the sites.
-
-\paragraph{Kaggle}
-Kaggle is dominated by sites with minimized expression \c{V}, with
-\kaggleMinimizedcallsitesa sites, and \kaggleMinimizedpropsitesa\% of the
-\kaggleNbCallSites sites in total. All of those values are an AST size of one
-node, suggesting they do not come. \NOTE{TODO}
-%TODO: all the src ref here seem to be incorrect and not from kaggle.
+\paragraph{Package environments.} Another kind of environments that is passed to \eval is the environment of a package. 
+ It refers to the \emph{namespace} of a package. A call to
+ \c{library(packageName)}, which load libraries in R, for instance, will populate
+ environment \c{package:packageName}. It is a \emph{base} function though, so we
+ do not see it in the CRAN dataset. However, R still makes it possible to access
+ any package environments and \packageNbPackageEnvSites call sites, \ie
+ \packagePackageEnvSitePercent access a package environment, from
+ \packageNbPackageEnvPackages different packages. Those packages perform
+ introspection, to represent and search any environment by name, in package
+ \emph{envnames}, to containerize tests, as with \emph{testthat} and
+ \emph{unitizer}, to create a custom OOP system with \emph{R.oo}, or to create
+ poor man's hidden functions in \emph{USSCXenaTools}.
+
+ It is also used to implement a custom caching mechanism in packages
+ \emph{g.data} and \emph{SOAR}.
 

### 2021-04-01 paper/main.tex
stuff

--- 2021-04-01 paper/main.tex
@@
 \begin{lstlisting}
- f1 <- eval(paste("A ~ ",paste(paste(names(X[,-1])),collapse="+")))
+ f1 <- eval(paste("A~",paste(paste(names(X[,-1])),collapse="+")))
 \end{lstlisting}
-  The code concatenates strings and \eval will simply return the string as is.
+  This concatenates to a string and \eval simply returns it as is.
   From twenty randomly selected \evals, only \packageUsefulValueEvalPercent of
@@
   %% non-parent environment.
-\item \underline{$min(e)=\c{\$}$}: A named lookup, written \c{x\$f}, can be used
+\item \underline{$min(e)=\c{\$}$}: A name lookup, \c{x\$f}, is used
+  can be used
   to look up a name in an environment or a column in a data frame. This category

### 2021-04-01 paper/main.tex
Better description of the possible envir types

--- 2021-04-01 paper/main.tex
@@
 
-The environment argument can be \c{NULL}, a \c{List} or a data frame. This
-happens in 3.5\% of the cases: \eval copies the fields of the list or data frame
+The \c{envir} argument of \eval can be an actual environment, but also \c{Null}, a list, a dataframe, a pairlist or even an integer. % Sometimes also a double, but it is cast to integer anyway.
+
+Most of the time, \ie in \packageEnvirEnvSitePercent of the sites, \eval is passed an actual environment.
+In \packageNbEnvirIncorrectSites sites (\packageEnvirIncorrectSitePercent), \eval is passed an incorrect argument, for instance \c{NA}, a boolean, which makes \eval fail.
+
+\c{NULL}, a \c{list} or a data frame, or an integer, happen in \packageEnvirUnusualSitePercent of the cases: \eval copies the fields of the list or data frame
 and creates bindings for them in a new environment. This pattern is used to
@@
 We observe that a disproportionately high number of core R \eval calls access
-the caller's environment. This is because many core R functions call functions
+the caller's environment. This is because many R functions call functions
 that pass the result of \c{parent.frame()} to an \eval. A disproportionately
@@
  It refers to the \emph{namespace} of a package. A call to
- \c{library(packageName)}, which load libraries in R, for instance, will populate
+ \c{library(packageName)}, which loads libraries in R, for instance, will populate
  environment \c{package:packageName}. It is a \emph{base} function though, so we
@@
  \emph{g.data} and \emph{SOAR}.
+
+\paragraph{\c{enclos} argument.}
 

### 2021-04-01 paper/main.tex
5

--- 2021-04-01 paper/main.tex
@@
 
-\item \underline{$min(e)=\c{X}$}: Variables are the most common normal form.
-  Variables subsume values, so they include arithmetic expressions in which a
-  variable occurs. \packageNbSymbolVarSitePercent of variable expression are
-  symbols, \ie just an \c{x}. A variable may refer to a promise and thus trigger
-  computation the first time it is accessed.
+\item \underline{$min(e)=\c{X}$}: Variables are the most common form. They
+  subsume values, thus \c{X} includes arithmetic expressions in which a variable
+  occurs. \packageNbSymbolVarSitePercent of variable expression are symbols, \ie
+  just an \c{x}. As a variable may refer to a promise, it may trigger
+  computation.
   %% The
@@
   %% non-parent environment.
-\item \underline{$min(e)=\c{\$}$}: A name lookup, \c{x\$f}, is used
-  can be used
-  to look up a name in an environment or a column in a data frame. This category
-  also groups indexing operators, \c{[]} and \c{[[]]}. The presence of these
-  operators subsumes constants and variables.
+\item \underline{$min(e)=\c{\$}$}: A name lookup, \c{x\$f}, can be used to
+  access variable in an environment or a column in a data frame, this also
+  include vector   indexing operators, \c{[]} and \c{[[]]}. A name lookup
+  subsumes variables.
 %% It amounts for \packageNbSlotAccess sites, \ie
@@
 
-\item \underline{$min(e)=$~\c{<-}}: Assignments amount to
-  \packageAssignSitesPercent of the sites, they include \c{<-}, \c{<<-} and
-  \c{assign}, and subsume named lookups.
+\item \underline{$min(e)=$~\c{<-}}: Assignments account for
+  \packageAssignSitesPercent of \eval sites, they are \c{<-}, {\tt <<-} and
+  \c{assign}, and they subsume named lookups.
 %% \packageNonDefaultEnvirAssignSitesPercent of the sites with assignments. Indeed,
@@
 
-\item \underline{$min(e)=\c{F()}$}: Function calls that
-
-\item \underline{$min(e)=\c{F(X)}$}: Function calls that
-
-\item \underline{$min(e)=\c{F(F(X))}$}: Function calls that
-
-\item \underline{$min(e)=\c{FUN}$}: Function calls that
-
-\item \underline{$min(e)=\c{\{\,\}}$}: Function calls that
-
-\item \underline{$min(e)=\c{model.frame}$}: Function calls that
+\item \underline{$min(e)=\c{F()}$}: A plain function call which may have some
+  values as argument but does not include variables. It should be noted that
+  function name require variable lookups -- we distinguish those as they usually
+  do not trigger computation.
+
+\item \underline{$min(e)=\c{F(X)}$}: Function calls whose arguments may include
+  variable references.
+
+\item \underline{$min(e)=\c{F(F(X))}$}: Nested function calls with variables.
+
+\item \underline{$min(e)=\c{FUN}$}: Function definition accounts for
+  \packageFunctionDefinitionSitesPercent of \eval sites. In addition the
+  functions inside other expressions account for
+  \packageGeneralizedFunctionDefinitionSitesPercent of sites, and give an
+  estimate of the use of higher-order functions.
+
+\item \underline{$min(e)=\c{\{\,\}}$}: Arbitrary-length statement blocks. These
+  are larger expressions with, on average, \packageMinimizedoperationsjRnd
+  instructions.
+
+\item \underline{$min(e)=\c{model.frame}$}: Call to the \c{model.frame} function
+  used to create statistical models.
 
@@
 \packageGeneralizedVarAccessSitePercent of the sites.
-
-\paragraph{\c{model.frame}}
-This is a function used to create statistical models. We explain it in more
-details in the section about the usage of \eval (see Section~REF).
-
-\paragraph{Function definition.}
-\c{FUN} represents a function definition at the top level of the expression, and
-happens in \packageFunctionDefinitionSites sites, \ie
-\packageFunctionDefinitionSitesPercent of the sites. It entails very few steps
-of the interpreter, as the function is returned by \eval, and the body of the
-new function is just stored for later execution. If we count also function
-definition inside more complex expressions, it amounts to
-\packageGeneralizedFunctionDefinitionSites sites, \ie
-\packageGeneralizedFunctionDefinitionSitesPercent and suggests the widespread
-use of higher-order functions in R, such as \c{lapply}, the R version of
-\c{map}.
-
-\paragraph{Large expressions.}  Compositions of functions, without variables
-(\c{F(F())} minimized expression) is the second most frequent minimized
-expression. \c{\{\}} is a sequence of statements. It is usually large, with
-in average \packageMinimizedoperationsjRnd operations in it.
 

### 2021-04-01 analysis/environments.Rmd

@@
 
+# Types in `envir`
+
 What about we classify them by type?
@@
 frequent_envir %>% count(envir_type, sort = TRUE) %>% datatable()
+```
+
+And just individual types:
+
+```{r}
+E_envir %>% count(envir_type, sort = TRUE) %>% datatable()
+```
+
+We realize that there are some types that are passed to eval that are not admissible according to the documentation.
+
+```{r}
+envir_env <- E_envir %>% filter(envir_type == "environment") 
+envir_unusual <- E_envir %>% filter(envir_type %in% c("list", "integer", "double", "NULL", "pairlist"))
+envir_incorrect <- E_envir %>% anti_join(envir_env) %>% anti_join(envir_unusual)
+
+nb_envir_env_sites <- envir_env %>% pull(src_ref) %>% n_distinct()
+nb_envir_unusual_sites <- envir_unusual %>% pull(src_ref) %>% n_distinct()
+nb_envir_incorrect_sites <- envir_incorrect %>% pull(src_ref) %>% n_distinct()
+
+overview_table(
+  r("nb envir env sites", nb_envir_env_sites),
+  r("envir env site percent", ratio(nb_envir_env_sites, nb_call_sites)),
+  r("nb envir unusual sites", nb_envir_unusual_sites),
+  r("envir unusual site percent", ratio(nb_envir_unusual_sites, nb_call_sites)),
+  r("nb envir incorrect sites", nb_envir_incorrect_sites),
+  r("envir incorrect site percent", ratio(nb_envir_incorrect_sites, nb_call_sites)),
+)
 ```

### 2021-04-01 paper/main.tex
5.1

@@
 %% with a variable inside.
-
 \item \underline{$min(e)=\c{X}$}: Variables are the most common form. They
@@
 %% percentage is here \packageMinimizedpercentenvirnonparentg.
-
 \item \underline{$min(e)=$~\c{<-}}: Assignments account for
@@
 %% instrument side-effects to environment in Section~REF.}
-
 \item \underline{$min(e)=\c{F()}$}: A plain function call which may have some
@@
   do not trigger computation.
-
 \item \underline{$min(e)=\c{F(X)}$}: Function calls whose arguments may include
   variable references.
-
 \item \underline{$min(e)=\c{F(F(X))}$}: Nested function calls with variables.
-
 \item \underline{$min(e)=\c{FUN}$}: Function definition accounts for
@@
   estimate of the use of higher-order functions.
-
 \item \underline{$min(e)=\c{\{\,\}}$}: Arbitrary-length statement blocks. These
   are larger expressions with, on average, \packageMinimizedoperationsjRnd
-  instructions.
-
+  instructions. The contents of the block are not inspected.
 \item \underline{$min(e)=\c{model.frame}$}: Call to the \c{model.frame} function
-  used to create statistical models.
+  used to create statistical models. This subsume all other categories.
 
@@
 
-Similarly, when a function is evaluated on some variable (\c{F(X)} minimized
-expression), the percentage of non default, non parent environments is rather
-high, \packageMinimizedpercentenvirnonparentd\%.
-
-As a whole, variable access, either as the top call, or inside a function call,
-possibly with accessing a slot, represents
-\packageGeneralizedVarAccessSitePercent of the sites.
-
-The minimized expression classification suggests that the presence of variables
-in the expression is highly correlated to the specification of a specific
-environment, and that \eval is used here for its power in precising the
-evaluation environment.
-
-Table~\ref{tab:minimizedexpressions} gives eleven most frequent in number of
-sites. \#sites is the number of sites; \%sites the percentage of sites across
-the total number of sites (\packageNbCallSites sites) with at least this
-minimized expression; \#packages is the number of packages with at least one of
-the minimized expression and measures how widespread a minimized expression is.
-\#operations is the average number of steps performed by the R interpreter when
-given that minimized expression, and \%envir is the percentage of call sites for
-which the \c{envir} argument is explicitly passed, and is not
-\c{parent.frame()}, which is the most common explicit environment passed to
-\eval.
+%% Similarly, when a function is evaluated on some variable (\c{F(X)} minimized
+%% expression), the percentage of non default, non parent environments is rather
+%% high, \packageMinimizedpercentenvirnonparentd\%.
+
+%% As a whole, variable access, either as the top call, or inside a function call,
+%% possibly with accessing a slot, represents
+%% \packageGeneralizedVarAccessSitePercent of the sites.
+
+%% The minimized expression classification suggests that the presence of variables
+%% in the expression is highly correlated to the specification of a specific
+%% environment, and that \eval is used here for its power in precising the
+%% evaluation environment.
+
+Table~\ref{tab:minimizedexpressions} gives the most frequent forms; \#sites
+(\%sites) is the number (ratio) of call sites receiving arguments of that form;
+\#packages is the number of packages with that form; \#operations is the average
+number of instructions performed by the interpreter when given that minimized
+expression; and \%envir is the percentage of call sites for which the \c{envir}
+argument is explicitly passed, and is not \c{parent.frame()}.
 
@@
 
-Note that a site can exhibit several minimized expressions (see
-Figure~\ref{fig:frequentcombinationminimized}). However,
-\packageNbOneMinimizedPercent of the sites have only one minimized expression.
-Note also that if a function is called in that expression, we see only the
-function call but not the body of the function. We will trace what happens
-during the evaluation of the expression by \eval in
-Section~\ref{sec:dangerseval}, especially about side effects performed in \eval.
-
-\NOTE{Maybe we should ditch the table of the minimized expressions and rather
-  use the ones of combinations. It has the advantage of showing what is the most
-  frequent, per call site. But it's a long tail, and sites that have several
-  minimized expressions are not that numerous, only 13\%. But if include
-  combinations, percentages change. For instance, 3314 call sites have only X,
-  i.e. 22.51, instead of 4353 and 29.6\%. }
+A given \eval call site can be passed different forms at different invocations.
+However, \packageNbOneMinimizedPercent of sites only see a single form.
 
 \paragraph{Base}
-The ranking of minimized expressions in \emph{base} is similar to the one in
-CRAN but  the share of \c{V} is much larger, as it is present in
-\baseMinimizedpropsitesa\% of the sites and \c{X} is a distant second, in
-\baseMinimizedpropsitesb\% of the sites.
+The spread of forms in Base are similar to CRAN with a larger share of \c{V}, as
+it is present in \baseMinimizedpropsitesa\% of the sites and \c{X} is a distant
+second with \baseMinimizedpropsitesb\%.
 
 \paragraph{Kaggle}
-Kaggle is dominated by sites with minimized expression \c{V}, with
-\kaggleMinimizedcallsitesa sites, and \kaggleMinimizedpropsitesa\% of the
-\kaggleNbCallSites sites in total. All of those values are an AST size of one
-node, suggesting they do not come. \NOTE{TODO}
-%TODO: all the src ref here seem to be incorrect and not from kaggle.
-
-\subsubsection{The environment passed to \eval.}
-
-The \c{envir} argument of \eval can be an actual environment, but also \c{Null}, a list, a dataframe, a pairlist or even an integer. % Sometimes also a double, but it is cast to integer anyway.
-
-Most of the time, \ie in \packageEnvirEnvSitePercent of the sites, \eval is passed an actual environment.
-In \packageNbEnvirIncorrectSites sites (\packageEnvirIncorrectSitePercent), \eval is passed an incorrect argument, for instance \c{NA}, a boolean, which makes \eval fail.
-
-\c{NULL}, a \c{list} or a data frame, or an integer, happen in \packageEnvirUnusualSitePercent of the cases: \eval copies the fields of the list or data frame
-and creates bindings for them in a new environment. This pattern is used to
-evaluate formulas which can directly refer to the fields of the data. The
-\c{envir} argument can also be a number $n$. It means that the environment in
-which the expressions is evaluated will be the result of \c{sys.call(n)} where
-$n$ refers to the $n$-th stack frame.
+Kaggle is dominated by \c{V}, with \kaggleMinimizedpropsitesa\% of the
+\kaggleNbCallSites sites.
+
+\subsection{The environment of \eval}
+
+The \c{envir} argument of \eval can be an actual environment, but also \c{Null},
+a list, a dataframe, a pairlist or even an integer.
+
+Most of the time, \ie in \packageEnvirEnvSitePercent of the sites, \eval is
+passed an actual environment. In \packageNbEnvirIncorrectSites sites
+(\packageEnvirIncorrectSitePercent), \eval is passed an incorrect argument, for
+instance \c{NA}, a boolean, which makes \eval fail.
+
+\c{NULL}, a \c{list} or a data frame, or an integer, happen in
+\packageEnvirUnusualSitePercent of the cases: \eval copies the fields of the
+list or data frame and creates bindings for them in a new environment. This
+pattern is used to evaluate formulas which can directly refer to the fields of
+the data. The \c{envir} argument can also be a number $n$. It means that the
+environment in which the expressions is evaluated will be the result of
+\c{sys.call(n)} where $n$ refers to the $n$-th stack frame.
 
@@
 We observe that a disproportionately high number of core R \eval calls access
-the caller's environment. This is because many R functions call functions
-that pass the result of \c{parent.frame()} to an \eval. A disproportionately
-high number of calls to \eval happens in an extended top-level environment. This
-can be explained by the fact that many packages evaluate code passed from the
-user's workspace in the top-level environment to access the bindings.
+the caller's environment. This is because many R functions call functions that
+pass the result of \c{parent.frame()} to an \eval. A disproportionately high
+number of calls to \eval happens in an extended top-level environment. This can
+be explained by the fact that many packages evaluate code passed from the user's
+workspace in the top-level environment to access the bindings.
 
@@
  It refers to the \emph{namespace} of a package. A call to
- \c{library(packageName)}, which loads libraries in R, for instance, will populate
- environment \c{package:packageName}. It is a \emph{base} function though, so we
- do not see it in the CRAN dataset. However, R still makes it possible to access
- any package environments and \packageNbPackageEnvSites call sites, \ie
- \packagePackageEnvSitePercent access a package environment, from
+ \c{library(packageName)}, which loads libraries in R, for instance, will
+ populate environment \c{package:packageName}. It is a \emph{base} function
+ though, so we do not see it in the CRAN dataset. However, R still makes it
+ possible to access any package environments and \packageNbPackageEnvSites call
+ sites, \ie \packagePackageEnvSitePercent access a package environment, from
  \packageNbPackageEnvPackages different packages. Those packages perform

### 2021-04-01 paper/main.tex
more

@@
 
-\subsection{The environment of \eval}
+\subsection{The environments of \eval}
 
 The \c{envir} argument of \eval can be an actual environment, but also \c{Null},
-a list, a dataframe, a pairlist or even an integer.
-
-Most of the time, \ie in \packageEnvirEnvSitePercent of the sites, \eval is
-passed an actual environment. In \packageNbEnvirIncorrectSites sites
-(\packageEnvirIncorrectSitePercent), \eval is passed an incorrect argument, for
-instance \c{NA}, a boolean, which makes \eval fail.
-
-\c{NULL}, a \c{list} or a data frame, or an integer, happen in
+a list, a dataframe, a pairlist or even an integer. Most of the time, \ie in
+\packageEnvirEnvSitePercent of the sites, \eval is passed an actual environment.
+In \packageNbEnvirIncorrectSites sites (\packageEnvirIncorrectSitePercent),
+\eval is passed an incorrect argument, for instance \c{NA}, a boolean, which
+makes \eval fail. \c{NULL}, a \c{list} or a data frame, or an integer, happen in
 \packageEnvirUnusualSitePercent of the cases: \eval copies the fields of the

### 2021-04-01 analysis/environments.Rmd
Envir

@@
 envir_env <- E_envir %>% filter(envir_type == "environment") 
-envir_unusual <- E_envir %>% filter(envir_type %in% c("list", "integer", "double", "NULL", "pairlist"))
+# S4 class: from pim package, uses an environment under the hood
+envir_unusual <- E_envir %>% filter(envir_type %in% c("list", "integer", "double", "NULL", "pairlist", "S4"))
 envir_incorrect <- E_envir %>% anti_join(envir_env) %>% anti_join(envir_unusual)

### 2021-04-01 paper/main.tex
Envir

--- 2021-04-01 paper/main.tex
@@
 
-\item \underline{$min(e)=\c{X}$}: Variables are the most common form. They
-  subsume values, thus \c{X} includes arithmetic expressions in which a variable
-  occurs. \packageNbSymbolVarSitePercent of variable expression are symbols, \ie
-  just an \c{x}. As a variable may refer to a promise, it may trigger
-  computation.
+\item \underline{$min(e)=\c{X}$}: Variables are the most common normal form.
+  Variables subsume values, so they include arithmetic expressions in which a
+  variable occurs. \packageNbSymbolVarSitePercent of variable expression are
+  symbols, \ie just an \c{x}. A variable may refer to a promise and thus trigger
+  computation the first time it is accessed.
   %% The
@@
   %% non-parent environment.
-\item \underline{$min(e)=\c{\$}$}: A name lookup, \c{x\$f}, can be used to
-  access variable in an environment or a column in a data frame, this also
-  include vector   indexing operators, \c{[]} and \c{[[]]}. A name lookup
-  subsumes variables.
+\item \underline{$min(e)=\c{\$}$}: A name lookup, \c{x\$f}, is used
+  can be used
+  to look up a name in an environment or a column in a data frame. This category
+  also groups indexing operators, \c{[]} and \c{[[]]}. The presence of these
+  operators subsumes constants and variables.
 %% It amounts for \packageNbSlotAccess sites, \ie
@@
 
-\item \underline{$min(e)=$~\c{<-}}: Assignments account for
-  \packageAssignSitesPercent of \eval sites, they are \c{<-}, {\tt <<-} and
-  \c{assign}, and they subsume named lookups.
+\item \underline{$min(e)=$~\c{<-}}: Assignments amount to
+  \packageAssignSitesPercent of the sites, they include \c{<-}, \c{<<-} and
+  \c{assign}, and subsume named lookups.
 %% \packageNonDefaultEnvirAssignSitesPercent of the sites with assignments. Indeed,
@@
 
-\item \underline{$min(e)=\c{F()}$}: A plain function call which may have some
-  values as argument but does not include variables. It should be noted that
-  function name require variable lookups -- we distinguish those as they usually
-  do not trigger computation.
-
-\item \underline{$min(e)=\c{F(X)}$}: Function calls whose arguments may include
-  variable references.
-
-\item \underline{$min(e)=\c{F(F(X))}$}: Nested function calls with variables.
-
-\item \underline{$min(e)=\c{FUN}$}: Function definition accounts for
-  \packageFunctionDefinitionSitesPercent of \eval sites. In addition the
-  functions inside other expressions account for
-  \packageGeneralizedFunctionDefinitionSitesPercent of sites, and give an
-  estimate of the use of higher-order functions.
-
-\item \underline{$min(e)=\c{\{\,\}}$}: Arbitrary-length statement blocks. These
-  are larger expressions with, on average, \packageMinimizedoperationsjRnd
-  instructions.
-
-\item \underline{$min(e)=\c{model.frame}$}: Call to the \c{model.frame} function
-  used to create statistical models.
+\item \underline{$min(e)=\c{F()}$}: Function calls that
+
+\item \underline{$min(e)=\c{F(X)}$}: Function calls that
+
+\item \underline{$min(e)=\c{F(F(X))}$}: Function calls that
+
+\item \underline{$min(e)=\c{FUN}$}: Function calls that
+
+\item \underline{$min(e)=\c{\{\,\}}$}: Function calls that
+
+\item \underline{$min(e)=\c{model.frame}$}: Function calls that
 
@@
 \packageGeneralizedVarAccessSitePercent of the sites.
+
+\paragraph{\c{model.frame}}
+This is a function used to create statistical models. We explain it in more
+details in the section about the usage of \eval (see Section~REF).
+
+\paragraph{Function definition.}
+\c{FUN} represents a function definition at the top level of the expression, and
+happens in \packageFunctionDefinitionSites sites, \ie
+\packageFunctionDefinitionSitesPercent of the sites. It entails very few steps
+of the interpreter, as the function is returned by \eval, and the body of the
+new function is just stored for later execution. If we count also function
+definition inside more complex expressions, it amounts to
+\packageGeneralizedFunctionDefinitionSites sites, \ie
+\packageGeneralizedFunctionDefinitionSitesPercent and suggests the widespread
+use of higher-order functions in R, such as \c{lapply}, the R version of
+\c{map}.
+
+\paragraph{Large expressions.}  Compositions of functions, without variables
+(\c{F(F())} minimized expression) is the second most frequent minimized
+expression. \c{\{\}} is a sequence of statements. It is usually large, with
+in average \packageMinimizedoperationsjRnd operations in it.
 
@@
 
+% This table is old and incorrect
+% Working on updating it
 \begin{table}[htbp]{ \centering
@@
 
-\paragraph{\c{enclos} argument.}
+\paragraph{\c{enclos} argument.} \ \\
 

### 2021-04-02 paper/main.tex
various changes

--- 2021-04-01 paper/main.tex
@@
 \begin{document}
-\title{What we {\sf eval} in the shadows}
+\title{What We Eval in the Shadows}
 \subtitle{A large-scale study of {\sf eval} usage in R programs}
@@
 
-%%%%%%%% JAN:::: NOT SURE WHAT THIS IS
-% The namespace stuff of packages was meant for the description of the envir argument.
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 The argument passed to a call to \eval determines what the function will do. As
@@
   them needed to use \eval.
-%% The percentage of non default environment passed to \eval is
-%% \packageMinimizedpercentenvirc when the minimized expression is a value. Some
-%% sites which are passed a value, are also passed different minimized expressions.
-%% For those ones, \packageNonDefaultEnvirValuePercent use a non-default
-%% environment, and among the one that use a non-default environment,
-%% \packageNonDefaultEnvirWithVarPercent have at least one minimized expression
-%% with a variable inside.
 \item \underline{$min(e)=\c{X}$}: Variables are the most common form. They
@@
   computation.
-  %% The
-  %% \c{envir} argument is explicitly passed in \packageMinimizedpercentenvira\% of
-  %% the call sites. The call sites without an explicit \c{envir} argument all use
-  %% a specific \eval variant, \c{eval.parent}. Moreover, the \c{envir} argument is
-  %% both explicit and different from \c{parent.frame()} in
-  %% \packageMinimizedpercentenvirnonparenta\% of the sites, which suggests that
-  %% when given a variable, \eval is often used to get value in a non-local,
-  %% non-parent environment.
 \item \underline{$min(e)=\c{\$}$}: A name lookup, \c{x\$f}, can be used to
@@
   subsumes variables.
-%% It amounts for \packageNbSlotAccess sites, \ie
-%% \packageSlotAccessPercent of the sites. The non-default, non-parent environment
-%% percentage is here \packageMinimizedpercentenvirnonparentg.
 \item \underline{$min(e)=$~\c{<-}}: Assignments account for
@@
   \c{assign}, and they subsume named lookups.
-%% \packageNonDefaultEnvirAssignSitesPercent of the sites with assignments. Indeed,
-%% most of those assignments occur in the default environment, \ie in the caller of
-%% \eval. This hints that side effects are not widespread in \eval.\footnote{Side
-%% effects can also happen in a function call in the \eval expression, and that
-%% function call will hide assignment operators from us. We more precisely
-%% instrument side-effects to environment in Section~REF.}
 \item \underline{$min(e)=\c{F()}$}: A plain function call which may have some
@@
 
-%% Similarly, when a function is evaluated on some variable (\c{F(X)} minimized
-%% expression), the percentage of non default, non parent environments is rather
-%% high, \packageMinimizedpercentenvirnonparentd\%.
-
-%% As a whole, variable access, either as the top call, or inside a function call,
-%% possibly with accessing a slot, represents
-%% \packageGeneralizedVarAccessSitePercent of the sites.
-
-%% The minimized expression classification suggests that the presence of variables
-%% in the expression is highly correlated to the specification of a specific
-%% environment, and that \eval is used here for its power in precising the
-%% evaluation environment.
-
 Table~\ref{tab:minimizedexpressions} gives the most frequent forms; \#sites
@@
 
-The \c{envir} argument of \eval can be an actual environment, but also \c{Null},
-a list, a dataframe, a pairlist or even an integer. Most of the time, \ie in
-\packageEnvirEnvSitePercent of the sites, \eval is passed an actual environment.
-In \packageNbEnvirIncorrectSites sites (\packageEnvirIncorrectSitePercent),
-\eval is passed an incorrect argument, for instance \c{NA}, a boolean, which
-makes \eval fail. \c{NULL}, a \c{list} or a data frame, or an integer, happen in
-\packageEnvirUnusualSitePercent of the cases: \eval copies the fields of the
-list or data frame and creates bindings for them in a new environment. This
-pattern is used to evaluate formulas which can directly refer to the fields of
-the data. The \c{envir} argument can also be a number $n$. It means that the
-environment in which the expressions is evaluated will be the result of
-\c{sys.call(n)} where $n$ refers to the $n$-th stack frame.
+The environment in which evaluation happens determines what will be visible to
+the computation started by \eval, and the potential reach of side-effects. The
+second argument to \eval, \c{envir}, defines the evaluation environment of the
+expression. That argument can be omitted, in which case the default value is
+\c{parent.frame()} which returns the caller of the \eval function.
+
+Other possible values for the environment include, lists, dataframes, pairlists,
+integers and \c{Null}. Lists and dataframe are coerced to new environments,
+integers values are passed to the \c{sys.call} function which returns the
+environment of a stack frame at the requested offset.
 
@@
 environment which becomes the enclosing scope of the new environment.
+
+Another kind of environments that is passed to \eval is the environment of a
+package. It refers to the \emph{namespace} of a package. A call to
+\c{library(packageName)}, which loads libraries in R, for instance, will
+populate environment \c{package:packageName}. It is a \emph{base} function
+though, so we do not see it in the CRAN dataset. However, R still makes it
+possible to access any package environments and \packageNbPackageEnvSites call
+sites, \ie \packagePackageEnvSitePercent access a package environment, from
+\packageNbPackageEnvPackages different packages. Those packages perform
+introspection, to represent and search any environment by name, in package
+\emph{envnames}, to containerize tests, as with \emph{testthat} and
+\emph{unitizer}, to create a custom OOP system with \emph{R.oo}, or to create
+poor man's hidden functions in \emph{USSCXenaTools}. It is also used to
+implement a custom caching mechanism in packages \emph{g.data} and \emph{SOAR}.
+
+\subsubsection{Bits of text, possibly reusable}
+
+Most of the time, \ie in \packageEnvirEnvSitePercent of the sites, \eval is
+passed an actual environment.
+
+In \packageNbEnvirIncorrectSites sites (\packageEnvirIncorrectSitePercent),
+\eval is passed an incorrect argument, for instance \c{NA}, a boolean, which
+makes \eval fail.
+
+\c{NULL}, a \c{list} or a data frame, or an integer, happen in
+\packageEnvirUnusualSitePercent of the cases: \eval copies the fields of the
+list or data frame and creates bindings for them in a new environment.
+
+This pattern is used to evaluate formulas which can directly refer to the fields
+of the data.
 
@@
 The \c{imchange} function of package \c{imager} makes it possible to modify
-images using a dedicated formula syntax using \~.\footnote{Inspired by {map} in
-package \emph{purr}.} Here, \eval is evaluated in \c{newenv}, which creates a
-new environment that inherits from \c{parent.frame()} by default (classified as
-1+).
+images using a dedicated formula syntax.\footnote{Inspired by {map} in package
+\emph{purr}.} Here, \eval is evaluated in \c{newenv}, which creates a new
+environment that inherits from \c{parent.frame()} by default (classified as 1+).
 
@@
 
-\paragraph{Package environments.} Another kind of environments that is passed to \eval is the environment of a package.
- It refers to the \emph{namespace} of a package. A call to
- \c{library(packageName)}, which loads libraries in R, for instance, will
- populate environment \c{package:packageName}. It is a \emph{base} function
- though, so we do not see it in the CRAN dataset. However, R still makes it
- possible to access any package environments and \packageNbPackageEnvSites call
- sites, \ie \packagePackageEnvSitePercent access a package environment, from
- \packageNbPackageEnvPackages different packages. Those packages perform
- introspection, to represent and search any environment by name, in package
- \emph{envnames}, to containerize tests, as with \emph{testthat} and
- \emph{unitizer}, to create a custom OOP system with \emph{R.oo}, or to create
- poor man's hidden functions in \emph{USSCXenaTools}.
-
- It is also used to implement a custom caching mechanism in packages
- \emph{g.data} and \emph{SOAR}.
-
-\paragraph{\c{enclos} argument.}
-
-\mypara{Side-effects.} XXX026 total number of reads performed by
-\eval. \NOTE{What is the definition of a read? An env lookup?} XXX027
-average number of reads. XXX028 total number of writes performed by
-\eval. Average number XXX029.
-
-\NOTE{Aviral wrote''measure number of reads and writes by eval outside
-  of its bubble'' --- what is the bubble here?}
-
-XXX029 reads to caller environment -- where caller is the caller of
-\eval.
-
-XXX030 reads to other caller environment -- i.e. envs that are on the
-call stack.
-
-XXX031 reads to caller environment -- where caller is the caller of
-\eval.
-
-XXX032 reads to other caller environment -- i.e. envs that are on the
-call stack.
+\paragraph{\c{enclos} argument.} \ \\
 
@@
 \EvalWritesRnd writes happen inside \eval. However, all writes are not
-dangerous. Only writes to environments not local to the computation
-spawned by \eval are side-effecting. These writes outlive the
-computation and hence are visible outside it. The remaining writes are
-local to the computation. We observe that \EvalSideEffectingWritesRnd
-writes inside evals are side-effecting. This is only
-\EvalSideEffectingWritesEvalPerc of all variable writes inside \eval
-and \EvalSideEffectingWritesAllPerc of all variable writes in the
-corpus.
-
-An \eval is considered side-effecting if it performs a side-effecting
-write to a variable, directly or indirectly. Only
-\SideEffectingCoreCallPerc \eval calls in Core R are side-effecting
-and \SideEffectingPackageCallPerc \eval calls in CRAN packages are
-side-effecting.
-
-\mypara{New Bindings.} Find number of times, eval introduces new
-bindings. There are many ways -- library, load, attach, source, and
-explicitly introduce bindings using super-assign, assign and define.
-
-\mypara{C Calls.} Find number of times, eval makes call to native code
-using - .Call, .External, etc. Find out if there are native functions
-that are called only from within eval.
-
-\mypara{Non-local returns}
-Eval can do non-local returns effective bypassing evaluation of the
-rest of the function. This can be useful for static analyzers.
-
-\mypara{Purity.}
-Conclude with number of evals that are ``pure'', i.e. evals which
-could be ignored by a static analyzer without any problem.
+dangerous. Only writes to environments not local to the computation spawned by
+\eval are side-effecting. These writes outlive the computation and hence are
+visible outside it. The remaining writes are local to the computation. We
+observe that \EvalSideEffectingWritesRnd writes inside evals are side-effecting.
+This is only \EvalSideEffectingWritesEvalPerc of all variable writes inside
+\eval and \EvalSideEffectingWritesAllPerc of all variable writes in the corpus.
+
+An \eval is considered side-effecting if it performs a side-effecting write to a
+variable, directly or indirectly. Only \SideEffectingCoreCallPerc \eval calls in
+Core R are side-effecting and \SideEffectingPackageCallPerc \eval calls in CRAN
+packages are side-effecting.
+
+\mypara{New Bindings.} Find number of times, eval introduces new bindings. There
+are many ways -- library, load, attach, source, and explicitly introduce
+bindings using super-assign, assign and define.
+
+\mypara{C Calls.} Find number of times, eval makes call to native code using -
+.Call, .External, etc. Find out if there are native functions that are called
+only from within eval.
+
+\mypara{Non-local returns} Eval can do non-local returns effective bypassing
+evaluation of the rest of the function. This can be useful for static analyzers.
+
+\mypara{Purity.} Conclude with number of evals that are ``pure'', i.e. evals
+which could be ignored by a static analyzer without any problem.
 
@@
 
-\mypara{Debloating.} \Eval is often used as a means to reduce
-boilerplate code; simple and repetitive code can easily be replaced
-with judcious use of \eval. For example, the \c{data.table} package
-uses \eval to calls the \c{options} function with named arguments
-taken from a vector of strings. While the benefits are limited in this
-example, it is an attractive tool for programmers.
+\mypara{Debloating.} \Eval is often used as a means to reduce boilerplate code;
+simple and repetitive code can easily be replaced with judcious use of \eval.
+For example, the \c{data.table} package uses \eval to calls the \c{options}
+function with named arguments taken from a vector of strings. While the benefits
+are limited in this example, it is an attractive tool for programmers.
 \begin{lstlisting}
@@
 
-The expression \c{\{info <- loadingNamespaceInfo(...} is added by core
-R to a package directory during installation. To load the package,
-this code is executed. It creates a namespace for the package, injects
-the package bindings, and attaches the namespace to the program search
-path. The \c{NULL} comes from a call to \c{substitute(subset)} in
-\c{stats::model.frame.default} function which has a default value of
-\c{subset} as \c{NULL}. The \c{Symbol} arises from a call to
-\c{as.name} in \c{base::str} function that returns a symbol that is
-looked up by evaluating it in a specific environment. The remaining
-cases arise from calls to \c{match.arg} which is used to look up the
-default choices for a variable and match against the choice passed by
-the caller.
+The expression \c{\{info <- loadingNamespaceInfo(...} is added by core R to a
+package directory during installation. To load the package, this code is
+executed. It creates a namespace for the package, injects the package bindings,
+and attaches the namespace to the program search path. The \c{NULL} comes from a
+call to \c{substitute(subset)} in \c{stats::model.frame.default} function which
+has a default value of \c{subset} as \c{NULL}. The \c{Symbol} arises from a call
+to \c{as.name} in \c{base::str} function that returns a symbol that is looked up
+by evaluating it in a specific environment. The remaining cases arise from calls
+to \c{match.arg} which is used to look up the default choices for a variable and
+match against the choice passed by the caller.
 
@@
 
-The expression \c{Environment} occurs because of the four callsites
-explained above, \c{ggplot2::ggproto}, \c{R6::generator\_funs},
-\c{future::backtrace} and \c{RModel::str.RMmodel}. The next
-expression, \c{column[rows] <<- what}, is used inside the
-\c{plyr::rbind.fill} function to merge data frames by assigning
-concatenated vectors to rows. The \c{<<-} operator is interesting in
-that it skips the current scope and assigns in a parent scope in which
-the variable is already present. In our corpus, all these \eval calls
-contribute to a single side-effect. The expression \c{function(value)
-  freduce(value, `function.list`)} arises from the
-\c{magrittr::\%>\%} function which is a pipe operator that pipes the
-output of previous command to the next one. The expression is
-evaluated in a custom environment to create a function binding for
-evaluating the components of the pipe.\c{String Vector} and \c{List}
-also arise from the same function when a string or a list is piped
-using the \c{\%>\%} function into the next expression. The \c{NULL}
-arises from \c{R6::generator\_funs} function when the \eval is passed a
-\c{NULL} argument by the \c{DataMask\_generator} package. The
+The expression \c{Environment} occurs because of the four callsites explained
+above, \c{ggplot2::ggproto}, \c{R6::generator\_funs}, \c{future::backtrace} and
+\c{RModel::str.RMmodel}. The next expression, \c{column[rows] <<- what}, is used
+inside the \c{plyr::rbind.fill} function to merge data frames by assigning
+concatenated vectors to rows. The \c{<<-} operator is interesting in that it
+skips the current scope and assigns in a parent scope in which the variable is
+already present. In our corpus, all these \eval calls contribute to a single
+side-effect. The expression \c{function(value) freduce(value, `function.list`)}
+arises from the \c{magrittr::\%>\%} function which is a pipe operator that pipes
+the output of previous command to the next one. The expression is evaluated in a
+custom environment to create a function binding for evaluating the components of
+the pipe.\c{String Vector} and \c{List} also arise from the same function when a
+string or a list is piped using the \c{\%>\%} function into the next expression.
+The \c{NULL} arises from \c{R6::generator\_funs} function when the \eval is
+passed a \c{NULL} argument by the \c{DataMask\_generator} package. The
 \c{c("default", "default2012", "default2011" ...} pattern arises from
-\c{copula::polyG} where it reflectively access the default expression
-for its formal parameter and evaluates it. The \c{force(..1)} and
-\c{alist(`spec`)} patterns occur in \c{glue::glue\_data} function
-which concatenates and interpolates strings. The two patterns occur
-because the function captures unevaluated unnamed arguments and maps
-the evaluation of \c{force(..1)} on them. The \c{force} function
-forces promises and returns the result of evaluation. The \c{inner}
-pattern arises from \c{glue::identity\_transformer} which enables the
-creation of custom transformation functions for affecting the
-interpolation and concatenation of input by the \c{glue} package.
+\c{copula::polyG} where it reflectively access the default expression for its
+formal parameter and evaluates it. The \c{force(..1)} and \c{alist(`spec`)}
+patterns occur in \c{glue::glue\_data} function which concatenates and
+interpolates strings. The two patterns occur because the function captures
+unevaluated unnamed arguments and maps the evaluation of \c{force(..1)} on them.
+The \c{force} function forces promises and returns the result of evaluation. The
+\c{inner} pattern arises from \c{glue::identity\_transformer} which enables the
+creation of custom transformation functions for affecting the interpolation and
+concatenation of input by the \c{glue} package.
 
@@
 
-We look at how \emph{consistent} the \c{expr} argument of \eval can
-be, \ie how many different types of the resolved \c{expr} there are
-per call sites. Most of the call sites, \ie \PercentMonomorphic, are
-\emph{consistent}, and this is similar to javascript. However, a few
-ones are highly \emph{polymorphic} (10 different types). They are the
-pipe operators \c{\%>\%}, \c{\%<>\%} and \c{\%\$\%} in package
-\emph{magrittr}. It is effectively used to compose functions on their
-first argument, which can be of any type.
-
-Similarly to JavaScript, there are also unnecessary uses of \eval. For
-example, the \c{PerformanceAnalytics} package contains a function
-\c{chart.QQPlot} that uses \eval to resolve a string into function and
-another to call it and assign its results into a variable:
+We look at how \emph{consistent} the \c{expr} argument of \eval can be, \ie how
+many different types of the resolved \c{expr} there are per call sites. Most of
+the call sites, \ie \PercentMonomorphic, are \emph{consistent}, and this is
+similar to javascript. However, a few ones are highly \emph{polymorphic} (10
+different types). They are the pipe operators \c{\%>\%}, \c{\%<>\%} and
+\c{\%\$\%} in package \emph{magrittr}. It is effectively used to compose
+functions on their first argument, which can be of any type.
+
+Similarly to JavaScript, there are also unnecessary uses of \eval. For example,
+the \c{PerformanceAnalytics} package contains a function \c{chart.QQPlot} that
+uses \eval to resolve a string into function and another to call it and assign
+its results into a variable:
 \begin{lstlisting}
@@
 
-\subsection{Comparison with Javascript}
-
-\begin{table}[h]
-	\begin{tabularx}{\textwidth}{|X|X|X|X|X|}
-		\hline
-		& Julia & Javascript & Python & R \tabularnewline
-		\hline % lstlinline does not like tabular environments...
-		{\eval} & \texttt{eval(expr)} & \texttt{eval(expr)} & \texttt{eval(expr[, globals[, locals]])} &  \texttt{eval(expr[, envir[, enclos]])}  \tabularnewline
-		\hline
-		Environments &Global  & Local & Global or local  & Any \tabularnewline
-		\hline
-	\end{tabularx}
-	\caption{Relative power of \eval throughout languages: in which environment it can evaluate its expression.}
-\end{table}
-
 \section{Conclusion}

### 2021-04-02 paper/main.tex
stuff

@@
 
-The environment in which evaluation happens determines what will be visible to
-the computation started by \eval, and the potential reach of side-effects. The
-second argument to \eval, \c{envir}, defines the evaluation environment of the
-expression. That argument can be omitted, in which case the default value is
-\c{parent.frame()} which returns the caller of the \eval function.
-
-Other possible values for the environment include, lists, dataframes, pairlists,
-integers and \c{Null}. Lists and dataframe are coerced to new environments,
-integers values are passed to the \c{sys.call} function which returns the
+The environment in which evaluation happens determines what is visible to the
+computation started by \eval, and the potential reach of its side-effects. The
+second argument to \eval, \c{envir}, defines that environment. It can be
+omitted, in which case its defaults to \c{parent.frame()} which returns the
+caller's environment. Other possible values for \c{envir} include, lists,
+dataframes, pairlists, integers and \c{Null}. Lists and dataframe are coerced to
+new environments, integer values are passed to \c{sys.call} which returns the
 environment of a stack frame at the requested offset.

### 2021-04-02 analysis/environments.Rmd
Enclos argument

@@
     caller_expression,
-    environment_class
+    environment_class,
+    enclos = enclos_expression
   ) -> E_envir
@@
 
+# Global ranking of environment classes
+
+```{r}
+env_class_ranking <- E_envir %>% group_by(environment_class) %>% 
+  summarize(n = n_distinct(src_ref)) %>% 
+  arrange(desc(n))
+
+env_class_ranking %>% datatable()
+```
+
 # Types in `envir`
@@
   group_by(src_ref) %>%
-  summarize(envir_distinct = n_distinct(environment_class), envir = paste0(environment_class, collapse = " ; "))
+  summarize(envir_distinct = n_distinct(environment_class), envir = paste0(unique(environment_class), collapse = " ; "))
 
@@
   count(envir_distinct, sort = TRUE)
+
+absolute_diversity %>%
+  count(envir, sort = TRUE)
 ```
@@
 
-```
-
-_ggplot2_ again is responsible for this `new.env` 4 times!
-
-```{r}
-env_calls %>% filter(!is.na(new_env), specific_env == "global") %>% count(new_env, wt = duplicates, sort = TRUE)
-
-env_calls %>% filter(new_env == 4, specific_env == "global") %>% count(ev_package, wt = duplicates, sort = TRUE)
 ```
@@
 
-## Summary of the categories
+# `enclos` argument
 
-Total eval calls: `r nb_eval_calls`
+```{r}
+enclos_present <- E_envir %>% filter(!is.na(enclos))
 
-Total eval call sites: `r nb_call_sites`
+nb_enclos_sites <- enclos_present %>% pull(src_ref) %>% n_distinct()
 
-```{r summary, eval = FALSE}
-summary <- tribble(
-  ~category, ~calls, ~percent_calls, ~callsites, ~percent_callsites,
-  "Default", nb_default$n, ratio(nb_default$n, nb_eval_calls$n), nb_default_sites, ratio(nb_default_sites, nb_call_sites),
-  "Rcpp", nb_rcpp$n, ratio(nb_rcpp$n, nb_eval_calls$n), nb_rcpp_sites, ratio(nb_rcpp_sites, nb_call_sites),
-  "List or dataframe", nb_env_list$n, ratio(nb_env_list$n, nb_eval_calls$n), nb_env_list_sites, ratio(nb_env_list_sites, nb_call_sites),
-  "Argument of caller", nb_arg_caller$n, ratio(nb_arg_caller$n, nb_eval_calls$n), nb_arg_caller_sites, ratio(nb_arg_caller_sites, nb_call_sites),
-  "Absolute", nb_absolute$n, ratio(nb_absolute$n, nb_eval_calls$n), nb_absolute_sites, ratio(nb_absolute_sites, nb_call_sites),
-  "Integer", nb_integer$n, ratio(nb_integer$n, nb_eval_calls$n), nb_integer_sites, ratio(nb_integer_sites, nb_call_sites),
-  "Null", nb_null$n, ratio(nb_null$n, nb_eval_calls$n), nb_null_sites, ratio(nb_null_sites, nb_call_sites),
+overview_table(
+  r("nb enclos sites", nb_enclos_sites),
+  r("enclos site percent", ratio(nb_enclos_sites, nb_call_sites))
 )
+```
 
-summary %>% datatable(colnames = c("", "Eval calls", "%", "Call sites", "%"))
+To which type of `envir`is it associated? It should be either pairlists, data frames (list) or NULL (NULL being interpreted as the empty list).
+
+```{r}
+enclos_types <- enclos_present %>% group_by(envir_type) %>% 
+  summarize(n = n_distinct(src_ref)) %>% 
+  arrange(desc(n)) %>% 
+  datatable()
+
+nb_enclos_with_list_sites <- enclos_present %>% 
+  filter(envir_type %in% c("list", "NULL")) %>% 
+  pull(src_ref) %>% 
+  n_distinct()
+
+overview_table(
+  r("nb enclos with list sites", nb_enclos_with_list_sites),
+  r("enclos with list percent", ratio(nb_enclos_with_list_sites, nb_enclos_sites))
+)
 ```

### 2021-04-02 paper/main.tex
Enclos argument

@@
 
-\paragraph{\c{enclos} argument.} \ \\
+\paragraph{\c{enclos} argument.} The \c{enclos} argument of \eval can be provided when \c{envir} is a pairlist or a dataframe.
+It indicates in which environment to look for elements not present in \c{envir}.
+In our dataset, \c{enclos} is passed explicitly to \eval in \packageEnclosSitePercent of the sites.
+% 85% (\packageEnclosWithListPercent) actually have a list \c{(envir)}. The rest of 
+% of them have environment \c{envir}. In that case, \eval ignores \c{enclos} anyway
 
@@
 \mypara{Non-local returns} Eval can do non-local returns effective bypassing
-evaluation of the rest of the function. This can be useful for static analyzers.
+evaluation of the rest of the function.
+For instance, it can inject a \c{break} or \c{next} instructions in a loop.
+It is very rare however, as it happens only twice in our dataset.
+% What kind of non-local returns?
+% eval(quote(return(89))) does not work
 

### 2021-04-04 paper/main.tex
Update pipeline figure
Pass on section 3 and 4

@@
   programming in JavaScript. Dynamic analysis of a corpus of
-  \CranAllPackages libraries and \KaggleCode end-user code confirms that
+  \CranPackages libraries and \KaggleCode end-user code confirms that
   \eval is indeed in widespread use; R's \eval is more dangerous in
@@
 Our corpus has been constructed to reflect the levels of sophistication of the R
-community. We distinguish between \emph{CRAN packages} (\CranAllPackages curated
+community. We distinguish between \emph{CRAN packages} (\CranPackages curated
 packages that pass stringent quality checks and are equipped with tests and
@@
 
-\mypara{Base.} The base dataset contains 13 core libraries performing basic
-arithmetics, statistics, and operating system related functionalities. These
-libraries are bundled with R and are executed pervasively.
+\mypara{Base.} The base dataset contains \BasePackages core libraries
+performing basic arithmetics, statistics, and operating system related
+functionalities. These libraries are bundled with R and are executed
+pervasively. Together they contain \BaseEvalCallSites calls to \eval in
+\BaseFunsWithEvals functions. Some of these functions provide basic R
+functionalities such as package loading and thus there is hardly any R code
+which does not invoke one of these \eval call sites.
 
@@
 Examples and vignettes can be turned into scripts by extracting the relevant
-code. We obtained \CranPackages that call \eval. These packages represent
-\CranPackagesRatio of all packages or \CranCodeRatio of the entire code size of
-CRAN. From these we extracted \CranRunnableScripts scripts that contain
+code. From \CranPackages packages we extracted \CranRunnableScripts scripts that contain
 \CranRunnableCode lines of code (\CranRunnableCodeExamplesRnd in examples,
@@
 
-\mypara{Scripts.} Kaggle is an online platform for data
-science.\footnote{\url{kaggle.com}} It allows users to submit problems and
-compete to solve them. Solutions, called \emph{kernels}, are uploaded as plain
-scripts or notebooks. The quality of code on Kaggle is not uniform. Each kernel
-is runnable, and thus there is one script per kernel. Input data is provided. We
+\mypara{Kaggle.} The Kaggle website\footnote{\url{kaggle.com}} is an online
+platform for data science. It allows users to submit problems and compete to
+solve them. Solutions, called \emph{kernels}, are uploaded as plain scripts or
+notebooks. The quality of code on Kaggle is not uniform. Each kernel is
+runnable, and thus there is one script per kernel. Input data is provided. We
 obtained \KaggleKernels kernels. Since Kaggle is not curated, used SHA-1 hashes
@@
   \includegraphics[width=1.05\linewidth]{pipeline.pdf}
-  \caption{Pipeline UPDATE!!}\label{fig:pipeline}
+  \caption{Pipeline}\label{fig:pipeline}
 \end{figure}
@@
   program is run in its own process.
+  As the GNU R bytecode compiler is written in R, we turn it off
+  to avoid recording \eval in its code, and fallback on the interpreter.
 \item \emph{Analyze.} Finally, analysis output is merged, cleaned and summarized
@@
 
-\medskip\noindent
-The code extraction and tracing steps of the pipeline are run in
-parallel~\cite{GNUparallel}. Servers have identical environment thanks to docker
-images with all dependencies installed. The pipeline consists of 3.5K lines of
-Make, R and shell scripts.
-
-\subsection{Analysis}
+\medskip\noindent The code extraction and tracing steps of the pipeline are run
+in parallel~\cite{GNUparallel} orchestrated by a Makefile. Servers have
+identical environment thanks to docker images with all dependencies installed.
+
+\subsection{Dynamic Analysis}
 
@@
 based on GNU R 4.0.2 that exposes low-level callbacks for a variety of runtime
-events~\cite{oopsla19a}. As the GNU R bytecode compiler is written in R, we turn
-it off to avoid recording \eval in its code, and fallback on the interpreter.
-This study requires recording a large number of events, calls to \eval,
-side-effects, function entries and exits, environments, etc. One challenge we
-encountered during the analysis was that R does not produce source references
-for top-level blocks. That is to say, a method whose body would \c{eval(x)}
-would not have debug information that we can use to identify that particular
-\eval. The same expression surrounded by braces would have a line number and
-source reference. This is unfortunately not easily fixed in the R
-implementation. Instead, we extended the dynamic analysis tool to add missing
-source references heuristically.
-
-The analysis has some limitation. Even with the extension described above, there
-are still \CranUndefinedRnd \eval calls without source references. This occurs
-when a reference to the \eval function is passed as argument to a higher-order
-functions or when the \eval call originates from native code. However, these
-missing source references account for a meager \CranUndefinedRatio of all calls;
-they are unlikely to affect our results. Other limitations are that we ignore
-calls to the native \eval function and to the alternate \c{rlang::tidy\_eval}
-function which uses native \eval internally. The \c{rlang} package introduces a
-new kind of promise called \c{quosure} which is evaluated by \c{tidy\_eval}.
-None of these limitation should invalidate our conclusions.
-
-\newpage
+events~\cite{oopsla19b}.
+
+\mypara{Tracer.} The tracer registers callbacks to all \eval functions as well
+as a few other ones that allow us to better identify where do the expressions
+come from. For example, we taint the results of calls to \c{parse} (turns
+string into expression) and \c{match.call} (reflects on current call). We
+also capture calls to R API for dynamic code loading (\eg\xspace \c{library},
+\c{require} and \c{source}). Next, we subscribe to the events related to a
+variable definition and assignment allowing us to record side effects that
+happen in environments while evaluating code in \eval. One challenge we
+encountered during the analysis was that essentially R only provides source
+references for top-level calls in a function body or in the block surrounded by
+braces. That is to say, a method whose body would \c{eval(x)} would not have
+debug information that we can use to identify that particular \eval. The same
+expression surrounded by braces would have a line number and source reference.
+This is unfortunately not easily fixed in the R implementation. Instead, we
+extended the dynamic analysis tool to attach synthetic source code references
+to all \eval call sites by traversing AST.  The tracer is implemented as an R
+package in 3.2K C++ and 1.3K of R code. For performance reasons, most of the
+tracing is done in C++. While in theory the implementation is rather
+straightforward, what makes it hard in practice is that (1) some of the core R
+functionality such as package loading is implemented using \eval which then
+leaks to the tracer in unexpected places, (2) lazy evaluation makes it harder
+to analyze function arguments while tracing (3) running large amount of
+real-world code triggers exercises a lot of corner cases of highly
+underspecified R behavior.
+
+\mypara{Limitations.} The analysis has some limitation. Even with the extension
+described above, there are still \PkgUndefinedRnd \eval calls without source
+references. This occurs when a reference to the \eval function is passed as
+argument to a higher-order functions or when the \eval call originates from
+native code. However, these missing source references account for a meager
+\PkgUndefinedRatio of all calls; they are unlikely to affect our results. Other
+limitations are that we ignore calls to the native \eval function and to the
+alternate \c{rlang::tidy\_eval} function which uses native \eval internally.
+The \c{rlang} package introduces a new kind of promise called \c{quosure} which
+is evaluated by \c{tidy\_eval}. None of these limitation should invalidate our
+conclusions.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
+\begin{wrapfigure}{r}{5cm} \hspace*{-12mm}
+  \includegraphics[width=75mm]{pkgs-eval-callsites-hist.pdf} \caption{CRAN
+  \eval call sites}%
+  \label{fig:pkgs-eval-callsites-hist}
+\end{wrapfigure}
+%
 This section focuses on CRAN packages and reports statistics about the usage of
-\eval. Calls to \eval appear in {\bf 28\%} of the packages hosted on CRAN. There
-are \Corpus packages with an \eval, all programs extracted from these packages
-were executed, any run that did not exercise one of the \eval call sites in the
-packages was discarded. This left \Nbruns runs of some of the \Corpus packages
-to generate the data presented here. There were \Allcalls invocations of \eval
-and its variants occurring in \Triggeredpkgs packages.\footnote{Not all packages
-with static call sites of \eval were observed triggering that call site, this
-can be chalked down to incomplete code coverage and occasional infrastructure
-failures of our analysis infrastructure.} Figure~\ref{freq} summarizes the
-frequency of dynamic calls to \eval with the left column being the number of
-calls and the right, the number of packages in that range. There are \Fewcalls
-packages with low \eval frequency, fewer than 100 calls; and a similar number of
-packages, \Manycalls to be precise, that use \eval more than 1,000 times. One
-package, \Maxcallspack, calls it \Maxcalls times and thus accounts for over half
-of the observed calls.
+\eval. There are \PkgEvalCallSites calls to \eval in \PkgPackages packages.
+These represent \PkgPackagesRatio of all packages or \PkgCodeRatio of the
+entire code size of CRAN. Over half of the packages have less than 3 \eval
+call sites and except \MaxEvalCallSitesPackage (\MaxEvalCallSitesCount \eval
+calls), all packages contain less than \MaxEvalCallSitesRest \eval call sites
+(\cf Fig~\ref{fig:pkgs-eval-callsites-hist}) These \eval calls appear in
+\PkgFunsWithEval R functions (\CranFunsWithEvalRatio of all R functions in
+CRAN).
+
+For the dynamic analysis, we execute all the \CranRunnableScripts programs
+extracted from \CranPackages packages. Any run that did not exercise one of the
+\eval call sites in the packages was discarded. This left \Nbruns runs of some
+of the \Corpus packages to generate the data presented here. There were
+\Allcalls invocations of \eval and its variants occurring in \Triggeredpkgs
+packages. Out of \PkgEvalCallSitesRnd \eval call sites, \PkgHitEvalCallSites
+were hit at least once\footnote{Not all packages with static call sites of
+\eval were observed triggering that call site, this can be chalked down to
+incomplete code coverage and occasional tracer failure (\PkgFailedProgramsRatio
+of programs do not terminate successfully or within 30 minute timeout).}. On
+average, the analysis covers \PkgHitEvalCallSitesAvgRatio of package \eval call
+sites, which similar to the \PkgCodeCoverage of average package code coverage.
+Figure~\ref{freq} summarizes the frequency of dynamic calls to \eval with the
+left column being the number of calls and the right, the number of packages in
+that range. There are \Fewcalls packages with low \eval frequency, fewer than
+100 calls; and a similar number of packages, \Manycalls to be precise, that use
+\eval more than 1,000 times.  One package, \Maxcallspack, calls it \Maxcalls
+times and thus accounts for over half of the observed calls.
 
@@
 
+Figure~\ref{tab:variantseval} enumerate the use of variants of \eval. For each
+of the four variants, the first rows shows the number of call sites in the
+corpus (\emph{static}), the second row is the number of call sites that were
+encountered during analysis (\emph{exercised}), the last row is the number of
+calls (\emph{invocations}). The overwhelming majority of call sites and dynamic
+calls, are to \eval itself, \c{eval.parent} is rare, and both \c{evalq} and
+\c{local} are barely used at all. The difference between sites and exercised
+sites underscores the limitations due to code coverage.
+
 \begin{wrapfigure}{r}{5.8cm}
@@
 
-Figure~\ref{tab:variantseval} enumerate the use of variants of \eval. For each
-of the four variants, the first rows shows the number of call sites in the
-corpus (\emph{static}), the second row is the number of call sites that were
-encountered during analysis (\emph{exercised}), the last row is the number of
-calls (\emph{invocations}). The overwhelming majority of call sites and dynamic
-calls, are to \eval itself, \c{eval.parent} is rare, and both \c{evalq} and
-\c{local} are barely used at all. The difference between sites and exercised
-sites underscores the limitations due to code coverage.
-
 Figure~\ref{cn}(a) shows normalized invocation counts per call site; on the left
@@
 
-\begin{wrapfigure}{r}{5.8cm}
- \includegraphics[width=0.5\textwidth, trim=4.8cm 0 6cm 0, clip]{size_loaded_distribution}
-\caption{Loaded code} \label{fig:sizedistribution}
-\end{wrapfigure}
-
 To estimate how much executable code is injected through \eval, we measure the
@@
 Figure~\ref{fig:sizedistribution} shows the distribution of sizes for arguments
-of fewer or equal to 25 nodes. The x-axis is the size of arguments in number of
+of fewer or equal to 25 nodes.
+%
+\begin{wrapfigure}{r}{5.8cm}
+ \hspace*{-10mm}
+ \includegraphics[width=0.5\textwidth, trim=4.8cm 0 6cm 0, clip]{size_loaded_distribution}
+ \caption{Loaded code} \label{fig:sizedistribution}
+\end{wrapfigure}
+%
+The x-axis is the size of arguments in number of
 nodes, and the y-axis is the count of arguments with that size. The size drops
@@
 
-\begin{figure}[h!]
+\begin{figure}[tb!]
 \begin{tabular}{@{}c@{}c@{}}

### 2021-04-04 analysis/inc/paths.R

@@ -2,8 +2,9 @@
 # PATHS
 ################################################################################
 
-DATA_DIR  <- params$base_dir
-PACKAGE_DATA_DIR  <- path(DATA_DIR, "package")
+DATA_DIR         <- params$base_dir
+BASE_DATA_DIR  <- path(DATA_DIR, "base")
+PACKAGE_DATA_DIR <- path(DATA_DIR, "package")
 KAGGLE_DATA_DIR  <- path(DATA_DIR, "kaggle")
 
 PAPER_DIR <- path("..", "paper")
@@ -13,13 +14,15 @@ TAGS_DIR  <- path(PAPER_DIR, "tag")
 if (!dir_exists(PLOT_DIR)) dir_create(PLOT_DIR)
 if (!dir_exists(TAGS_DIR)) dir_create(TAGS_DIR)
 
-CORPUS_ALL_FILE <- path(DATA_DIR, "corpus-all.fst")
-CORPUS_FILE <- path(DATA_DIR, "corpus.fst")
+BASE_CORPUS_FILE       <- path(BASE_DATA_DIR, "corpus.txt")
+BASE_EVALS_STATIC_FILE <- path(BASE_DATA_DIR, "evals-static.csv")
 
-PACKAGE_SUM_CALLS_FILE <- path(PACKAGE_DATA_DIR, "summarized.fst")
-PACKAGE_UNDEFINED_FILE <- path(PACKAGE_DATA_DIR, "undefined.fst")
-PACKAGE_EVALS_STATIC_FILE <-path(PACKAGE_DATA_DIR, "evals-static.csv")
-PACKAGE_CODE_FILE <- path(PACKAGE_DATA_DIR, "code.fst")
+PACKAGE_CORPUS_FILE       <- path(PACKAGE_DATA_DIR, "corpus.fst")
+PACKAGE_SUM_CALLS_FILE    <- path(PACKAGE_DATA_DIR, "summarized.fst")
+PACKAGE_UNDEFINED_FILE    <- path(PACKAGE_DATA_DIR, "undefined.fst")
+PACKAGE_EVALS_STATIC_FILE <- path(PACKAGE_DATA_DIR, "evals-static.csv")
+PACKAGE_CODE_FILE         <- path(PACKAGE_DATA_DIR, "code.fst")
+PACKAGE_TRACE_LOG_FILE    <- path(PACKAGE_DATA_DIR, "trace-log.csv")
 
[…]

### 2021-04-04 paper/main.tex
Fix layout in S4

@@
 come from. For example, we taint the results of calls to \c{parse} (turns
-string into expression) and \c{match.call} (reflects on current call). We
-also capture calls to R API for dynamic code loading (\eg\xspace \c{library},
+string into expression) and \c{match.call} (reflects on current call). We also
+capture calls to R API for dynamic code loading (\eg\xspace \c{library},
 \c{require} and \c{source}). Next, we subscribe to the events related to a
@@
 encountered during the analysis was that essentially R only provides source
-references for top-level calls in a function body or in the block surrounded by
-braces. That is to say, a method whose body would \c{eval(x)} would not have
+references for top-level calls in a function body or in the block surrounded
+by braces. That is to say, a method whose body would \c{eval(x)} would not have
 debug information that we can use to identify that particular \eval. The same
@@
 to all \eval call sites by traversing AST.  The tracer is implemented as an R
-package in 3.2K C++ and 1.3K of R code. For performance reasons, most of the
+package in 3.2K C++ and 1.3K of R code. For performance reason, most of the
 tracing is done in C++. While in theory the implementation is rather
@@
 leaks to the tracer in unexpected places, (2) lazy evaluation makes it harder
-to analyze function arguments while tracing (3) running large amount of
-real-world code triggers exercises a lot of corner cases of highly
-underspecified R behavior.
+to analyze function arguments while tracing as prematurely forcing a promise
+might have consequences, and finally (3) running large amount of real-world
+code triggers exercises a lot of corner cases of highly underspecified R
+behavior.
 
@@
 conclusions.
+
+\newpage
 
@@
 
-\begin{figure}[!h]
+\begin{figure}[b]
 
@@
 
-\begin{wrapfigure}{r}{5.8cm}
+\begin{wrapfigure}{4}{5.8cm}
 \vspace{-4mm}

### 2021-04-04 paper/main.tex
Working on provenance

--- 2021-04-02 paper/main.tex
@@
 
-\mypara{Source.}
+\mypara{Source.} % rename to provenance
+
+% Substitute/match.call
+% parse/expression
+% bytecode?
+
 There are various ways to obtain expressions:
@@
 
-\section{\eval patterns and anti-patterns}
-
-\subsection{Evaluating variables in a specific environment}
+\section{Usage of \eval}
+
+We present here the main uses of \eval we have witnessed in R.
+Sometimes, they are design patterns, sometimes, \eval could be replaced by more specific functions.
+For each use of \eval, we discuss whether it is necessary to use.
+
+\subsection{Variable lookup}
+
+\eval is used to look-up the value of a variable in a given environment, different from the current scope so the \c{envir} argument of \eval is explicitly specified.
+
+% TODO: numbers on the prevalence of the pattern
+
+\eval is not necessary in that case and can be replaced by \c{get}, as follows:
+
+\begin{lstlisting}
+ eval(quote(x), envir = env)
+\end{lstlisting}
+$\Downarrow$
+\begin{lstlisting}
+get("x", envir = env)
+\end{lstlisting}
+
+% This is actually linked to NSE, as the symbol given to \eval usually comes
+% from match.call or susbtitute
+
+\subsection{Non-Standard Evaluation} % Or metaprogramming
+
+\subsection{\c{substitute}}
 
@@
 
-\subsection{Meta-programming}
-
 \subsection{Logging}
@@
 \subsection{Code transform}
+
+% Derivation of symbolic expressions
 

### 2021-04-04 analysis/corpus.Rmd

--- 2021-01-01 analysis/corpus.Rmd
@@
 ---
-title: "Usage metrics"
+title: "Corpus"
 output:
@@
 ```{r load corpus}
-CRAN_eval_traces <- 
+PKG_S <- 
   read_fst(PACKAGE_SUM_CALLS_FILE) %>%
   # remove evals that are inside the scripts,
-  # for CRAN we are only interested in evals in packages
+  # for PKG_C we are only interested in evals in packages
   filter(!str_starts(eval_call_srcref, "::global::main")) %>% 
@@
 
-CRAN_eval_traces_per_package <- 
-  count(CRAN_eval_traces, caller_package, eval_call_srcref) %>%
+PKG_evals_per_package <- 
+  count(PKG_S, caller_package, eval_call_srcref) %>%
   count(caller_package) %>% 
-  rename(package=caller_package, evals_hit=n)
+  rename(
+    package=caller_package, 
+    evals_hit=n
+  )
 
-ALL_CRAN <- 
-  read_fst(CORPUS_ALL_FILE) %>%
-  mutate_if(is.numeric, replace_na, 0) %>%
-  mutate(evals=as.integer(evals)) %>% 
-  filter(loadable, package_code > 0) %>% 
-  as_tibble()
-
-CRAN <- 
-  read_fst(CORPUS_FILE) %>%
-  filter(package_code > 0) %>%
-  left_join(CRAN_eval_traces_per_package, by="package") %>% 
+CRAN_C <- 
+  read_fst(PACKAGE_CORPUS_FILE) %>%
   mutate_if(is.numeric, replace_na, 0) %>%
@@
     evals=as.integer(evals), 
+    package_code=package_native_code + package_r_code
+  ) %>%
+  filter(loadable) %>% 
+  as_tibble()
+
+PKG_C <- 
+  CRAN_C %>%
+  filter(in_corpus, package_code > 0) %>%
+  left_join(PKG_evals_per_package, by="package") %>% 
+  mutate(
     evals_hit=as.integer(evals_hit),
@@
   ) %>% 
+  mutate_if(is.numeric, replace_na, 0) %>%
   as_tibble()
 
-ALL_CRAN_eval_callsites <- 
+CRAN_E <- 
   read_csv(PACKAGE_EVALS_STATIC_FILE) %>% 
@@
 
-CRAN_eval_callsites <- semi_join(ALL_CRAN_eval_callsites, CRAN, by="package")
+PKG_E <- semi_join(CRAN_E, PKG_C, by="package")
 
-ALL_KAGGLE <- local({
+KAGGLE_ALL <- local({
   kernels <- read_csv(KAGGLE_KERNEL_FILE)
@@
 
-KAGGLE <- filter(ALL_KAGGLE, evals > 0)
+KAGGLE <- filter(KAGGLE_ALL, evals > 0)
+
+PKG_undefined <- 
+  read_fst(PACKAGE_UNDEFINED_FILE) %>% 
+  as_tibble()
+
+TRACE_LOG <- read_csv(PACKAGE_TRACE_LOG_FILE, col_types = cols(
+  Seq = col_double(),
+  Host = col_character(),
+  Starttime = col_double(),
+  JobRuntime = col_double(),
+  Send = col_double(),
+  Receive = col_double(),
+  Exitval = col_double(),
+  Signal = col_double(),
+  Command = col_character(),
+  V1 = col_character(),
+  Stdout = col_logical(),
+  Stderr = col_logical()
+))
+
+BASE_packages <- read_lines(BASE_CORPUS_FILE)
+BASE_evals_per_package <- read_csv(BASE_EVALS_STATIC_FILE, col_types = cols(
+  package = col_character(),
+  fun_name = col_character(),
+  srcref = col_character(),
+  call_fun_name = col_character(),
+  args = col_character()
+))
+
+BASE_C <- 
+  tibble(package=BASE_packages) %>%
+  left_join(rename(count(BASE_evals_per_package, package), evals=n), by="package") %>%
+  left_join(rename(count(BASE_evals_per_package, package, fun_name) %>% count(package), funs_with_eval=n), by="package")
 ```
@@
 ```{r}
-cran_all_funs <- sum(ALL_CRAN$funs_public) + sum(ALL_CRAN$funs_private)
-cran_funs <- sum(CRAN$funs_public) + sum(CRAN$funs_private)
-cran_funs_with_eval <- sum(CRAN$funs_with_eval)
-cran_code <- sum(CRAN$package_code)
-cran_all_code <- sum(ALL_CRAN$package_code)
-cran_evals <- CRAN$evals
-cran_evals_hit <- CRAN$evals_hit
+overview_table(
+  r("base packages", BASE_C),
+  r("base eval call sites", sum(BASE_C$evals)),
+  r("base funs with evals", sum(BASE_C$funs_with_eval)),
+)
+```
+
+### Packages
+
+```{r}
+cran_funs          <- sum(CRAN_C$funs_public) + sum(CRAN_C$funs_private)
+cran_code          <- sum(CRAN_C$package_code)
+cran_r_code        <- sum(CRAN_C$package_r_code)
+cran_native_code   <- sum(CRAN_C$package_native_code)
+cran_runnable_code <- sum(CRAN_C$runnable_code)
+
+pkg_funs           <- sum(PKG_C$funs_public) + sum(PKG_C$funs_private)
+pkg_funs_with_eval <- sum(PKG_C$funs_with_eval)
+pkg_code           <- sum(PKG_C$package_code)
+pkg_r_code         <- sum(PKG_C$package_r_code)
+pkg_native_code    <- sum(PKG_C$package_native_code)
+pkg_evals          <- PKG_C$evals
+pkg_evals_hit      <- PKG_C$evals_hit
+pkg_runnable_code  <- sum(PKG_C$runnable_code)
+pkg_eval_calls     <- sum(PKG_S$nb_ev_calls)
+
+pkg_failed_programs <- filter(TRACE_LOG, Exitval > 1)
 
 overview_table(
-  r("CRAN all packages", ALL_CRAN),
-  r("CRAN packages", CRAN),
-  r("CRAN packages ratio", ratio(CRAN, ALL_CRAN)),
-  r("CRAN all code", cran_all_code),
-  r("CRAN all r code", sum(ALL_CRAN$package_r_code)),
-  r("CRAN all native code", sum(ALL_CRAN$package_native_code)),
+  r("CRAN available packages", 15962), #length(available.packages()[,1])), # from the PKG_C mirror
+  r("CRAN packages", CRAN_C),
+  r("PKG packages", PKG_C),
+  r("PKG packages ratio", ratio(PKG_C, CRAN_C)),
   r("CRAN code", cran_code),
-  r("CRAN code ratio", ratio(cran_code, cran_all_code)),
-  r("CRAN all funs", cran_all_funs),
+  r("CRAN r code", sum(CRAN_C$package_r_code)),
+  r("CRAN native code", sum(CRAN_C$package_native_code)),
+  r("PKG code", pkg_code),
+  r("PKG code ratio", ratio(pkg_code, cran_code)),
+  r("PKG r code", pkg_r_code),
+  r("PKG r code ratio", ratio(pkg_r_code, cran_r_code)),
+  r("PKG native code", pkg_native_code),
+  r("PKG native code ratio", ratio(pkg_native_code, cran_native_code)),
   r("CRAN funs", cran_funs),
-  r("CRAN funs with eval", cran_funs_with_eval),
-  r("CRAN funs with eval ratio", ratio(cran_funs_with_eval, cran_funs)),
-  r("CRAN all funs with eval ratio", ratio(cran_funs_with_eval, cran_all_funs)),
-  r("CRAN eval call sites", sum(cran_evals)),
-  r("CRAN hit eval call sites", sum(cran_evals_hit)),
-  r("CRAN hit eval call sites avg ratio", percent(mean(cran_evals_hit/cran_evals))),
-  r("CRAN runnable scripts", sum(ALL_CRAN$runnable_files)),
-  r("CRAN runnable code", sum(ALL_CRAN$runnable_code)),
-  r("CRAN runnable code examples", sum(ALL_CRAN$runnable_code_examples)),
-  r("CRAN runnable code tests", sum(ALL_CRAN$runnable_code_tests)),
-  r("CRAN runnable code vignettes", sum(ALL_CRAN$runnable_code_vignettes)),
+  r("PKG funs", pkg_funs),
+  r("PKG funs with eval", pkg_funs_with_eval),
+  r("PKG funs with eval ratio", ratio(pkg_funs_with_eval, pkg_funs)),
+  r("CRAN funs with eval ratio", ratio(pkg_funs_with_eval, cran_funs)),
+  r("PKG eval call sites", sum(pkg_evals)),
+  r("PKG hit eval call sites", sum(pkg_evals_hit)),
+  r("PKG hit eval call sites avg ratio", percent(mean(pkg_evals_hit/pkg_evals))),
+  r("PKG code coverage", percent(mean(PKG_C$coverage)/100)),
+  r("CRAN code coverage", percent(mean(CRAN_C$coverage)/100)),
+  r("CRAN runnable scripts", sum(CRAN_C$runnable_files)),
+  r("CRAN runnable code", cran_runnable_code),
+  r("CRAN runnable code examples", sum(CRAN_C$runnable_code_examples)),
+  r("CRAN runnable code tests", sum(CRAN_C$runnable_code_tests)),
+  r("CRAN runnable code vignettes", sum(CRAN_C$runnable_code_vignettes)),
+  r("PKG undefined", PKG_undefined),
+  r("PKG undefined ratio", ratio(PKG_undefined, pkg_eval_calls)),
+  r("PKG failed programs", pkg_failed_programs),
+  r("PKG failed programs ratio", ratio(pkg_failed_programs, TRACE_LOG))
 )
@@
 ```{r}
-```
-
-```{r}
-kaggle <- distinct(ALL_KAGGLE, hash, .keep_all=TRUE)
+kaggle <- distinct(KAGGLE_ALL, hash, .keep_all=TRUE)
 kaggle_code <- sum(kaggle$code)
@@
 overview_table(
-  r("Kaggle kernels", ALL_KAGGLE),
-  r("Kaggle duplicates", nrow(ALL_KAGGLE)-nrow(kaggle)),
+  r("Kaggle kernels", KAGGLE_ALL),
+  r("Kaggle duplicates", nrow(KAGGLE_ALL)-nrow(kaggle)),
   r("Kaggle unique", kaggle),
@@
   r("Kaggle with evals", kaggle_evals),
-  r("Kaggle with evals code", ratio(kaggle_evals_code, kaggle_code))
+  r("Kaggle with evals code", kaggle_evals_code),
+  r("Kaggle with evals code ratio", ratio(kaggle_evals_code, kaggle_code))
+)
+```
+
+```{r}
+cs_count(PKG_C, evals)
+```
+
+### Details
+
+```{r}
+max_eval_cs_idx <- which.max(PKG_C$evals)
+max_eval_cs_package <- PKG_C$package[max_eval_cs_idx]
+max_eval_cs_count <- PKG_C$evals[max_eval_cs_idx]
+
+overview_table(
+  r("PKG eval call sites median", median(PKG_C$evals)),
+  r("Max eval call sites count", max_eval_cs_count),
+  r("Max eval call sites package", max_eval_cs_package),
+  r("Max eval call sites rest", round(max(PKG_C$evals[-max_eval_cs_idx])/100)*100)
+)
+```
+
+### All
+
+```{r}
+overview_table(
+  r("Corpus code", kaggle_code + pkg_runnable_code)
 )
@@
 ```{r}
-eval_call_sites_hist <- function(df, df_all, y_lab, max_evals=15L) {
+eval_call_sites_hist <- function(df, df_all, y_lab, max_evals=20L) {
+  median_evals <- median(df$evals)
   df %>% 
     mutate(
-      evals=if_else(evals < max_evals, evals, max_evals)
+      evals=ifelse(evals < max_evals, evals, max_evals)
     ) %>%
@@
     geom_histogram(binwidth=1, color="black", fill="blue", alpha=0.3) +
-    geom_vline(aes(xintercept=mean(df$evals)), linetype="dashed", color="red", size=0.2) +
+    geom_vline(aes(xintercept=median_evals), linetype="dashed", color="red", size=0.2) +
     scale_x_continuous(
       limits=c(0, max_evals + 1), 
-      breaks=as.integer(seq(1, 15, length.out=4)), 
+      breaks=as.integer(sort(c(seq(1, max_evals, length.out=4), round(median_evals)))), 
       labels=function(x) c(head(x, -1), str_c(tail(x, 1), "+"))
@@
     ) +
-    labs(x = "Eval call sites", y=y_lab) +
-    theme(text = element_text(size=17),
+    labs(y=y_lab) +
+    theme(text = element_text(size=21),
+          axis.title.x = element_blank(),
           panel.grid.minor.x = element_blank(), aspect.ratio = 4/3) 
@@
 
-eval_call_sites_hist(CRAN, ALL_CRAN, "CRAN packages")
-#ggsave(path(PLOT_DIR, "cran-eval-callsites-hist.pdf"))
-eval_call_sites_hist(KAGGLE, ALL_KAGGLE, "KAGGLE programs")
-#ggsave(path(PLOT_DIR, "kaggle-eval-callsites-hist.pdf"))
+eval_call_sites_hist(PKG_C, CRAN_C, "CRAN packages")
+ggsave(path(PLOT_DIR, "pkgs-eval-callsites-hist.pdf"))
+eval_call_sites_hist(KAGGLE, KAGGLE_ALL, "KAGGLE programs")
+ggsave(path(PLOT_DIR, "kaggle-eval-callsites-hist.pdf"))
 ```

### 2021-04-04 analysis/provenance.Rmd
Analysis of the provenance of an expr: parse, match.call, substitute, comp etc

---
title: "Provenance"
author: "Pierre Donat-Bouillud"
output:
  html_document:
        gallery: false
        toc: true
        toc_depth: 3
        toc_float: true
        df_print: paged
date: "`r Sys.Date()`"
editor_options: 
  chunk_output_type: console
params:
  base_dir: Data/package/
  calls_path: summarized.fst
  corpus_path: corpus.txt
  static_path: evals-static.csv
  force_rebuild: FALSE
---

```{r setup, include=FALSE}
knitr::knit_hooks$set(time_it = local({
  now <- NULL
  function(before, options) {
    if (before) { # record the current time before each chunk
      now <<- Sys.time()
    } else { # calculate the time difference after a chunk
      res <- difftime(Sys.time(), now)
      # return a character string to show the time
      paste("Time for this code chunk to run:", res, units(res))
    }
  }
}))

knitr::opts_chunk$set(
  echo = TRUE,
  fig.align = "center",
  fig.retina = 2,
  fig.width = 10,
  cache.lazy = FALSE
)

now <- Sys.time()

library(tidyverse)
library(fst)
library(fs)
library(DT)

source("inc/latextags.R")
source("inc/functions.R")
source("inc/paths.R")
theme_set(theme_minimal())

dataset_name <- basename(params$base_dir)

# Tags will be prefixed by the dataset name
create_tags(path(TAGS_DIR, paste0(dataset_name, "_provenance.tex")), prefix = dataset_name, default = TRUE)
# Also use ggsave_prefix instead of ggsave if you need to save a graph

calls_path <- paste0(params$base_dir, params$calls_path)
corpus_path <- paste0(params$base_dir, params$corpus_path)
static_path <- paste0(params$base_dir, params$static_path)

lock_path <- paste0(params$base_dir, ".provenance-lock.fst")
min_path <- paste0(params$base_dir, "Eprov.fst")

nb_eval_call_sites <- function(dataset) {
  dataset %>% pull(src_ref) %>% n_distinct()
}

```

# Loading datafiles

```{r loading_fst}
tibble(package = read_lines(corpus_path)) -> C_raw
read_csv(static_path) %>% semi_join(C_raw, by = "package") -> P

rebuild <- TRUE

E_prov <- NULL

if (file.exists(lock_path)) {
  saved <- read.fst(lock_path)[[1,1]]
  if (saved == file.size(calls_path)) {
    read_fst(min_path) %>% as_tibble() -> E_prov
    rebuild <- FALSE
  }
}

if(params$force_rebuild || rebuild) {
  read_fst(calls_path) %>% as_tibble() -> E_raw

  E_raw <- E_raw %>% mutate(envir_expression = if_else(envir_type == "NULL" & is.na(envir_expression), "NULL", envir_expression))

  E_raw %>% select(
    run_package = package, #      The package which was run to trigger the eval
    ev_variant = eval_function, # Which eval variant was called
    duplicates = nb_ev_calls, #   Number of identical eval calls (weight)
    src_ref = eval_call_srcref, # Source ref for the eval call site
    file, #                       Script file name (name of this run)
    ev_package = eval_source, #   Package of the call site
    type = expr_resolved_type_tag, #   Argument type
    ev_call = eval_call_expression,
    ev_expr = expr_expression,
    caller_expression,
    match_call = expr_match_call,
    parsed = expr_parsed_expression,
  ) -> E_prov

  E_prov %>% write_fst(min_path)
  sz <- file.size(calls_path) %>% as_tibble()
  write.fst(sz, lock_path)
}

C_raw -> C
corpus_size <- nrow(C)
```

# Description of the data files

```{r packages}
nb_eval_calls <- count(E_prov, wt = duplicates)

nb_call_sites <- E_prov %>% pull(src_ref) %>% n_distinct()
```

There are `r nb_eval_calls` eval calls and `r nb_call_sites` eval call sites in the dataset.

# Classification of provenance

## Expression tyoes

```{r}
expressions <- E_prov %>% filter(type == "expression")

parsed_exprs <- E_prov %>% filter(!is.na(parsed))

not_parsed_exprs <- expressions %>% anti_join(parsed_exprs)

nb_parsed_exprs_sites <- parsed_exprs %>% pull(src_ref) %>% n_distinct()

nb_not_parsed_exprs_sites <- not_parsed_exprs %>% pull(src_ref) %>% n_distinct()

overview_table(
    r("nb parse exprs sites", nb_parsed_exprs_sites),
    r("parse exprs site percent", ratio(nb_parsed_exprs_sites, nb_call_sites)),
    r("nb not parse exprs sites", nb_not_parsed_exprs_sites),
    r("not parse exprs site percent", ratio(nb_not_parsed_exprs_sites, nb_call_sites)),
)
```

## Language type

```{r}
languages <- E_prov %>% filter(type == "language")

match_call_exprs <- E_prov %>% filter(!is.na(match_call))

str2lang_exprs <- language %>% filter(!is.na(parsed))

# Substitute, quote, bquote etc...
other_langs <- languages %>% anti_join(match_call_exprs) %>% anti_join(str2lang_exprs)

nb_match_call_exprs_sites <- match_call_exprs %>% pull(src_ref) %>% n_distinct()
nb_str2lang_sites <- str2lang_exprs %>% pull(src_ref) %>% n_distinct()
nb_other_langs_sites <- other_langs %>% pull(src_ref) %>% n_distinct()

overview_table(
    r("nb match call exprs sites", nb_match_call_exprs_sites),
    r("match call exprs site percent", ratio(nb_match_call_exprs_sites, nb_call_sites)),
    r("nb strlang sites", nb_str2lang_sites),
    r("strlang site percent", ratio(nb_str2lang_sites, nb_call_sites)),
    r("nb other langs sites", nb_other_langs_sites),
    r("other langs site percent", ratio(nb_other_langs_sites, nb_call_sites))
)
```

## Bytecode

We disable bytecode optimization so if we see a bytecode type, it means that `comp`has been explicitly called to compile to bytecode.

```{r}
bytecode_exprs <- E_prov %>% filter(type == "bytecode")

nb_bytecode_sites <- bytecode_exprs %>% pull(src_ref) %>% n_distinct()

overview_table(
    r("nb bytecode sites", nb_bytecode_sites),
    r("bytecode site percent", ratio(nb_bytecode_sites, nb_call_sites))
)
```

# More details for `parse` and expressions

```{r}
parse_func_exprs <- parsed_exprs %>% filter(str_detect(parsed, fixed("parse")))
str2expression_exprs <- parsed_exprs %>% filter(str_detect(parsed, fixed("str2expression")))

nb_parse_func_sites <- parse_func_exprs %>% pull(src_ref) %>% n_distinct()
nb_str2expression_sites <- str2expression_exprs %>% pull(src_ref) %>% n_distinct()

# File qrgument (but ignore the srcfile one)
from_file_exprs <- parse_func_exprs %>% filter(str_detect(parsed, "\\bfile\\b"))
nb_from_file_sites <- from_file_exprs %>% pull(src_ref) %>% n_distinct()

overview_table(
    r("nb parse func sites", nb_parse_func_sites),
    r("parse func site percent", ratio(nb_parse_func_sites, nb_parsed_exprs_sites)),
    r("nb strexpression sites", nb_str2expression_sites),
    r("strexpression site percent", ratio(nb_str2expression_sites, nb_parsed_exprs_sites)),
    r("nb parse from file sites", nb_from_file_sites),
    r("parse from file site percent", ratio(nb_from_file_sites, nb_parse_func_sites))
)
```

### 2021-04-05 analysis/normalized.Rmd
Add examples for normalized expressions
Provenance of eval arguments

@@
             n_operations = round(mean(n_op, na.rm=TRUE), 2), 
+            med_operations = median(n_op, na.rm = TRUE),
             percent_envir = ratio(n_distinct(cur_data()$src_ref[!is.na(cur_data()$envir)]),  n_call_sites), 
@@
 Looking up to a variable and applying a function to a variable, are the two non-trivial categories that most of the time explicitly use the `envir` argument of `eval`.
+
+# Examples of minimized expression
+
+```{r}
+examples_minimized <- E_cano %>% group_by(minimized) %>% filter(!is.na(expr_resolved) & str_length(expr_resolved) < 50) %>% 
+  summarise(example = nth(unique(expr_resolved), 100))
+
+examples_minimized %>% datatable()
+```
 

### 2021-04-05 analysis/provenance.Rmd
Provenance of eval arguments

@@
 
-## Expression tyoes
+## Expression types
 
@@
 
-parsed_exprs <- E_prov %>% filter(!is.na(parsed))
+parsed_exprs <- expressions %>% filter(!is.na(parsed))
 
@@
 
-str2lang_exprs <- language %>% filter(!is.na(parsed))
+str2lang_exprs <- languages %>% filter(!is.na(parsed))
 
@@
 
+## Sites wwith less signficant provenance
+
+```{r}
+unknown_prov <- E_prov %>% anti_join(expressions) %>% anti_join(languages) %>% anti_join(bytecode_exprs)
+
+nb_unknown_prov_sites <- unknown_prov %>% pull(src_ref) %>% n_distinct()
+
+overview_table(
+    r("nb unknown prov sites", nb_unknown_prov_sites),
+    r("unknown prov site percent", ratio(nb_unknown_prov_sites, nb_call_sites))
+)
+```
+
 # More details for `parse` and expressions

### 2021-04-05 paper/main.tex
Provenance of eval arguments

@@
 
+\mypara{Provenance.}
+
+% Substitute/match.call
+% parse/expression
+% bytecode?
+
+% Explain tha some generate expression objects and other language objects?
+There are various ways to obtain expressions that \eval will evaluate:
+\begin{itemize}
+    \item \c{substitute} synthesizes ASTs from expressions by replacing
+    symbols with their bindings in the specified environment.
+    \item \c{match.call} captures a call and its arguments as expressions.
+    \item \c{quote}, \c{enquote} returns its argument unevaluated, as an R expression.
+    \item \c{expression}  creates a vector of expression
+    objects from text.
+    \item \c{parse}, \c{str2expression} and \c{str2lang} turn strings into
+    expressions.
+    \item \c{comp} compiles R code into R bytecode.
+\end{itemize}
+
+\begin{table}[h]
+  \begin{tabular}{|c|c|c|}
+      \hline
+      & \#sites & \%sites \\
+       \hline
+         \c{substitute}, \c{quote} & \packageNbOtherLangsSites & \packageOtherLangsSitePercent \\
+      \hline
+      \c{parse}, \c{str2expression} & \packageNbParseExprsSites & \packageParseExprsSitePercent \\
+      \hline
+        \c{match.call} &  \packageNbMatchCallExprsSites & \packageMatchCallExprsSitePercent  \\   
+      \hline
+      \c{expression} & \packageNbNotParseExprsSites & \packageNotParseExprsSitePercent  \\
+      \hline
+      \c{str2lang} & \packageNbStrlangSites  & \packageStrlangSitePercent  \\
+      \hline
+      \c{comp} & \packageNbBytecodeSites & \packageBytecodeSitePercent \\
+      \hline
+  \end{tabular}
+\caption{Provenances of \eval expressions. }
+\end{table}
+
+% \packageUnknownProvSitePercent : provenance not relavant (values, symbols etc)
+
+\c{eval(parse(...))} can be used for dynamic code loading. This forms
+the core of \c{source} and \c{sys.source} functions in R that are
+commonly used for loading code in R files in interactive settings. We
+investigated the number of cases in which the output of \c{parse} and
+its variants is passed to \eval, directly or transitively by tainting
+their output.  We observed that very few of the eval call (\packageNbParseFromFileSites in total)
+consume the result of calling \c{parse} on a file. Most of the eval
+calls consume the result of calling \c{parse} on a string. We also
+identified one function in core R, \c{invokeRestartInteractively} that
+prompts the user for input, parses it, and passes it to \eval.
+
 \mypara{New Bindings.} Find number of times, eval introduces new bindings. There
@@
 which could be ignored by a static analyzer without any problem.
-
-\mypara{Source.} % rename to provenance
-
-% Substitute/match.call
-% parse/expression
-% bytecode?
-
-There are various ways to obtain expressions:
-\begin{itemize}
-  \item \c{substitute} synthesizes ASTs from expressions by replacing
-    symbols with their bindings in the specified environment.
-  \item \c{expression}  creates a vector of expression
-    objects from text.
-  \item \c{parse}, \c{str2expression} and \c{str2lang} turn strings into
-    expressions.
-\end{itemize}
-
-We observe 5.2\% cases where \eval directly evaluates the output of
-\c{substitute}, 0.7\% cases where output of \c{parse} is read directly
-and only 704 cases for \c{expression}. Most expressions consumed by
-\eval are generated by other functions.
-
-\c{eval(parse(...))} can be used for dynamic code loading. This forms
-the core of \c{source} and \c{sys.source} functions in R that are
-commonly used for loading code in R files in interactive settings. We
-investigated the number of cases in which the output of \c{parse} and
-its variants is passed to \eval, directly or transitively by tainting
-their output. This corresponds to \PercentParsedCallSites of the total
-\eval call sites and \PercentParsedEvals of the eval calls. We
-observed that very few of the eval call (\NbParseFilesRnd in total)
-consume the result of calling \c{parse} on a file. Most of the eval
-calls consume the result of calling \c{parse} on a string. We also
-identified one function in core R, \c{invokeRestartInteractively} that
-prompts the user for input, parses it, and passes it to \eval.
 

### 2021-04-05 analysis/corpus.Rmd
Add traced evals figure

@@
 ```{r}
-cs_count(PKG_C, evals)
+cs_count(PKG_C, evals) %>% print(n=Inf)
 ```
@@
       limits=c(0, max_evals + 1), 
-      breaks=as.integer(sort(c(seq(1, max_evals, length.out=4), round(median_evals)))), 
+      breaks=as.integer(sort(c(seq(1, max_evals, length.out=4), median_evals))), 
       labels=function(x) c(head(x, -1), str_c(tail(x, 1), "+"))
@@
 ```
+
+### Eval plot
+
+```{r}
+eval_plot_df <- PKG_C %>%
+  filter(package_r_code > 0) %>%
+  select(package, evals, evals_hit, package_r_code)
+
+size_breaks <- c(1e3, 5e3, 10e3, 25e3, 50e3, 100e3)
+# size_labels <- str_c(
+#   c("≤ ", rep("> ", 6)),
+#   scales::comma(accuracy = 1, x=size_breaks[-1])
+# )
+
+eval_plot_df %>% 
+#  filter(evals > 0, evals_hit > 0) %>%
+  mutate(
+    # size=cut(package_r_code, size_breaks, size_labels, include.lowest=T),
+    label=ifelse(evals %in% tail(sort(evals),3), package, NA)
+  ) %>%
+  ggplot(aes(x=evals, y=evals_hit, size=package_r_code, label=label)) + 
+    geom_jitter(alpha=0.3) +
+    geom_text(nudge_y = .15, check_overlap = TRUE, size=4) +
+    scale_x_continuous(
+      trans="log10",
+      breaks=function(x) {
+        f <- scales::breaks_log()
+        sort(c(f(x), max(eval_plot_df$evals)))
+      },
+      labels=scales::comma_format(accuracy = 1)
+    ) +
+    scale_y_continuous(
+      trans="pseudo_log",
+      breaks=function(x) {
+        f <- scales::breaks_log()
+        sort(c(0, 1, f(c(1, x[2])), max(eval_plot_df$evals_hit)))
+      },
+      labels=scales::comma_format(accuracy = 1)
+    ) +
+    scale_size_binned(
+      range=c(1,10),
+      breaks = size_breaks,
+      labels=scales::comma_format(accuracy = 1)
+    ) +
+#    geom_smooth(method = "lm", se = FALSE, show.legend = FALSE, linetype="dashed", color="red", size=0.2) +
+    labs(x="Eval call sites (log)", y="Traced eval call sites (log)", size="Package code size") +
+    theme(legend.position = c(0.15, 0.8))
+ggsave(path(PLOT_DIR, "traced-eval-callsites.pdf"))
+```

### 2021-04-05 paper/main.tex
Eval provenance
Add traced evals figure

@@
 
-\begin{wrapfigure}{r}{5cm} \hspace*{-12mm}
-  \includegraphics[width=75mm]{pkgs-eval-callsites-hist.pdf} \caption{CRAN
-  \eval call sites}%
-  \label{fig:pkgs-eval-callsites-hist}
-\end{wrapfigure}
-%
 This section focuses on CRAN packages and reports statistics about the usage of
@@
 packages. Out of \PkgEvalCallSitesRnd \eval call sites, \PkgHitEvalCallSites
-were hit at least once\footnote{Not all packages with static call sites of
-\eval were observed triggering that call site, this can be chalked down to
-incomplete code coverage and occasional tracer failure (\PkgFailedProgramsRatio
-of programs do not terminate successfully or within 30 minute timeout).}. On
-average, the analysis covers \PkgHitEvalCallSitesAvgRatio of package \eval call
-sites, which similar to the \PkgCodeCoverage of average package code coverage.
-Figure~\ref{freq} summarizes the frequency of dynamic calls to \eval with the
-left column being the number of calls and the right, the number of packages in
-that range. There are \Fewcalls packages with low \eval frequency, fewer than
-100 calls; and a similar number of packages, \Manycalls to be precise, that use
-\eval more than 1,000 times.  One package, \Maxcallspack, calls it \Maxcalls
-times and thus accounts for over half of the observed calls.
+were hit at least once (\cf Fig~\ref{fig:traced-eval-callsites}). On average,
+the analysis covers \PkgHitEvalCallSitesAvgRatio of package \eval call sites,
+which similar to the \PkgCodeCoverage of average package code
+coverage\footnote{Not all packages with static call sites of \eval were
+observed triggering that call site, this can be chalked down to incomplete code
+coverage and occasional tracer failure (\PkgFailedProgramsRatio of programs do
+not terminate successfully or within 30 minute timeout).}.  Figure~\ref{freq}
+summarizes the frequency of dynamic calls to \eval with the left column being
+the number of calls and the right, the number of packages in that range. There
+are \Fewcalls packages with low \eval frequency, fewer than 100 calls; and a
+similar number of packages, \Manycalls to be precise, that use \eval more than
+1,000 times.  One package, \Maxcallspack, calls it \Maxcalls times and thus
+accounts for over half of the observed calls.
+
+\begin{figure}[t]
+\begin{minipage}{.3\textwidth}
+  \vspace*{14mm}
+  \hspace*{-10mm}
+  \includegraphics[width=1.5\textwidth]{pkgs-eval-callsites-hist.pdf} \caption{CRAN
+  \eval call sites}%
+  \label{fig:pkgs-eval-callsites-hist}
+\end{minipage}
+\begin{minipage}{.6\textwidth}
+  \centering
+  \includegraphics[width=\textwidth]{traced-eval-callsites.pdf}
+  \caption{\eval call sites coverage of the \PkgPackages packages.}%
+  \label{fig:traced-eval-callsites}
+\end{minipage}
+\end{figure}
 
@@
 
-\mypara{Provenance.}
-
-% Substitute/match.call
-% parse/expression
-% bytecode?
-
+\paragraph{Provenance.} 
 % Explain tha some generate expression objects and other language objects?
@@
        \hline
-         \c{substitute}, \c{quote} & \packageNbOtherLangsSites & \packageOtherLangsSitePercent \\
+         \c{substitute}, \c{quote}, ... & \packageNbOtherLangsSites & \packageOtherLangsSitePercent \\
       \hline

### 2021-04-05 paper/main.tex
Calls -> call sites

@@
 This section focuses on CRAN packages and reports statistics about the usage of
-\eval. There are \PkgEvalCallSites calls to \eval in \PkgPackages packages.
-These represent \PkgPackagesRatio of all packages or \PkgCodeRatio of the
-entire code size of CRAN. Over half of the packages have less than 3 \eval
-call sites and except \MaxEvalCallSitesPackage (\MaxEvalCallSitesCount \eval
-calls), all packages contain less than \MaxEvalCallSitesRest \eval call sites
-(\cf Fig~\ref{fig:pkgs-eval-callsites-hist}) These \eval calls appear in
+\eval. There are \PkgEvalCallSites call sites to \eval in \PkgPackages
+packages.  These represent \PkgPackagesRatio of all packages or \PkgCodeRatio
+of the entire code size of CRAN. Over half of the packages have less than 3
+\eval call sites and except \MaxEvalCallSitesPackage (\MaxEvalCallSitesCount
+\eval calls), all packages contain less than \MaxEvalCallSitesRest \eval call
+sites (\cf Fig~\ref{fig:pkgs-eval-callsites-hist}) These \eval calls appear in
 \PkgFunsWithEval R functions (\CranFunsWithEvalRatio of all R functions in

### 2021-04-05 paper/main.tex
Fix
Removed some unused text

@@
 \newcommand{\NOTE}[1]{{\it Note: #1}\xspace}
+\newcommand{\authorcomment}[3]{\xspace\textcolor{#1}{{\bf #2} #3}\xspace}
+\newcommand{\todo}[1]{\authorcomment{red}{TODO}{#1}}
 
@@
 
+\todo{Look at Improving dynamic code analysis by code abstraction, VPT'21}
+
 \newpage
@@
 
+\begin{wrapfigure}{r}{5cm} \hspace*{-12mm}
+  \centering
+  \includegraphics[width=75mm]{pkgs-eval-callsites-hist.pdf} \caption{CRAN
+  \eval call sites}%
+  \label{fig:pkgs-eval-callsites-hist}
+\end{wrapfigure}
+%
 This section focuses on CRAN packages and reports statistics about the usage of
@@
 
-\begin{figure}[t]
-\begin{minipage}{.3\textwidth}
-  \vspace*{14mm}
-  \hspace*{-10mm}
-  \includegraphics[width=1.5\textwidth]{pkgs-eval-callsites-hist.pdf} \caption{CRAN
-  \eval call sites}%
-  \label{fig:pkgs-eval-callsites-hist}
-\end{minipage}
-\begin{minipage}{.6\textwidth}
+\begin{figure}[tb]
   \centering
-  \includegraphics[width=\textwidth]{traced-eval-callsites.pdf}
+  \includegraphics[width=.8\textwidth]{traced-eval-callsites.pdf}
+  \centering
   \caption{\eval call sites coverage of the \PkgPackages packages.}%
   \label{fig:traced-eval-callsites}
-\end{minipage}
 \end{figure}
@@
 of fewer or equal to 25 nodes.
+The x-axis is the size of arguments in number of
+nodes, and the y-axis is the count of arguments with that size. The size drops
+rapidly, with few observations larger than 15 nodes. The long tail is omitted
+from the graph for legibility.
 %
@@
 %
-The x-axis is the size of arguments in number of
-nodes, and the y-axis is the count of arguments with that size. The size drops
-rapidly, with few observations larger than 15 nodes. The long tail is omitted
-from the graph for legibility.
 
@@
 \baseSmalleventspct of calls.
+\todo{More info about base and conclude it here.}
 
@@
 \kaggleSmalleventspct of \evals execute fewer than 50 instructions.
+\todo{More info about kaggle and conclude it here.}
 
@@
 
+\todo{Pierre: check the numbers - we want to know when it is an environment, when it is a list / data frame}
 Most of the time, \ie in \packageEnvirEnvSitePercent of the sites, \eval is
 passed an actual environment.
-
-In \packageNbEnvirIncorrectSites sites (\packageEnvirIncorrectSitePercent),
-\eval is passed an incorrect argument, for instance \c{NA}, a boolean, which
-makes \eval fail.
 
@@
 
-\paragraph{Provenance.} 
+\subsection{Provenance.}
 % Explain tha some generate expression objects and other language objects?
@@
 
-\mypara{New Bindings.} Find number of times, eval introduces new bindings. There
-are many ways -- library, load, attach, source, and explicitly introduce
-bindings using super-assign, assign and define.
-
-\mypara{C Calls.} Find number of times, eval makes call to native code using -
-.Call, .External, etc. Find out if there are native functions that are called
-only from within eval.
-
-\mypara{Non-local returns} Eval can do non-local returns effective bypassing
-evaluation of the rest of the function.
-For instance, it can inject a \c{break} or \c{next} instructions in a loop.
-It is very rare however, as it happens only twice in our dataset.
-% What kind of non-local returns?
-% eval(quote(return(89))) does not work
-
-\mypara{Purity.} Conclude with number of evals that are ``pure'', i.e. evals
-which could be ignored by a static analyzer without any problem.
+\subsection{Side effects in \eval}
+
+\todo{Filip: discuss the side-effects}
 
@@
 
-We present here the main uses of \eval we have witnessed in R.
+\todo{Illustrate things that we have discussed in Section 5 - examples}
+\todo{Things to discuss: model.frame, match.call, substitute, parse, plots}
+
+% We present here the main uses of \eval we have witnessed in R.
 Sometimes, they are design patterns, sometimes, \eval could be replaced by more specific functions.
@@
 
-\subsection{Variable lookup}
-
-\eval is used to look-up the value of a variable in a given environment, different from the current scope so the \c{envir} argument of \eval is explicitly specified.
-
-% TODO: numbers on the prevalence of the pattern
-
-\eval is not necessary in that case and can be replaced by \c{get}, as follows:
-
-\begin{lstlisting}
- eval(quote(x), envir = env)
-\end{lstlisting}
-$\Downarrow$
-\begin{lstlisting}
-get("x", envir = env)
-\end{lstlisting}
-
-% This is actually linked to NSE, as the symbol given to \eval usually comes
-% from match.call or susbtitute
-
-\subsection{Non-Standard Evaluation} % Or metaprogramming
-
-\subsection{\c{substitute}}
-
-\subsection{\c{match.call}}
-
-\subsection{Logging}
-
-\subsection{Code transform}
-
-% Derivation of symbolic expressions
-
-\subsection{Plotting}
-
-\section{\eval expressivity and macros}
+% \subsection{Variable lookup}
+
+% \eval is used to look-up the value of a variable in a given environment, different from the current scope so the \c{envir} argument of \eval is explicitly specified.
+
+% \subsection{Metaprogramming}
+
+% \mypara{\c{substitute}}
+
+% \mypara{\c{match.call}}
+
+% \todo{Talk about model.frame}
+
+% \subsection{Logging}
+
+% \subsection{Code transform}
+
+% % Derivation of symbolic expressions
+
+% \subsection{Plotting}
+
+% \section{\eval expressivity and macros}
 
@@
 There are no local macro variables in this implementation though.
-
-\section{The dangers of \eval} \label{sec:dangerseval}
-
-\subsection{Eval Usage in R}
 

### 2021-04-05 paper/main.tex
Removed the legacy text

@@
 \paragraph{Base}
+\todo{I would remove this - we conclude Base in 4}
 The spread of forms in Base are similar to CRAN with a larger share of \c{V}, as
@@
 \paragraph{Kaggle}
+\todo{I would remove this - we conclude Kaggle in 4}
 Kaggle is dominated by \c{V}, with \kaggleMinimizedpropsitesa\% of the
 \kaggleNbCallSites sites.
+
+\todo{The following should be either removed or positioned properly}
+We look at how \emph{consistent} the \c{expr} argument of \eval can be, \ie how
+many different types of the resolved \c{expr} there are per call sites. Most of
+the call sites, \ie \PercentMonomorphic, are \emph{consistent}, and this is
+similar to javascript. However, a few ones are highly \emph{polymorphic} (10
+different types). They are the pipe operators \c{\%>\%}, \c{\%<>\%} and
+\c{\%\$\%} in package \emph{magrittr}. It is effectively used to compose
+functions on their first argument, which can be of any type.
 
@@
 
-\section{LEGACY TEXT}
-
-We looked at the top ten expressions passed to core and package \eval
-calls. The most frequent ten expressions to eval calls from core R
-contribute to 85\% of all \eval calls.
-
-\begin{table}[!h] \centering
-\begin{tabular}{@{}l|rr@{}} \hline
-Expression & Eval Call &  \% \\\hline
-\c{c("auto", "shell", "radix")} & 1,987,105 & 29\%\\
-\c{c("auto", "shell", "quick", "radix")} & 1,593,169  & 23\%\\
-\c{\{info <- loadingNamespaceInfo(...} & 1,008,632 &       14\%\\
-\c{c("onLoad", "attach", "detach", "onUnload")}   & 470,566 &      6.9\%\\
-\c{c("append", "prepend", "replace")} &              261,587&       3.9\% \\
-\c{c("left", "right", "centre", "none")} & 162,086     & 2.4\%\\
-\c{c("no", "ifany", "always")}   &                71,580 &       1.1\%\\
-\c{c("pearson", "kendall", "spearman")}  & 72,962 &      1.1\%\\
-\c{NULL}& 75,330  &      1.1\% \\
-\c{Symbol}&                 66,279&       1\%\\\hline
-\end{tabular}
-\caption{Top ten eval calls in Core}\label{B}
-\end{table}
-
-The expression \c{\{info <- loadingNamespaceInfo(...} is added by core R to a
-package directory during installation. To load the package, this code is
-executed. It creates a namespace for the package, injects the package bindings,
-and attaches the namespace to the program search path. The \c{NULL} comes from a
-call to \c{substitute(subset)} in \c{stats::model.frame.default} function which
-has a default value of \c{subset} as \c{NULL}. The \c{Symbol} arises from a call
-to \c{as.name} in \c{base::str} function that returns a symbol that is looked up
-by evaluating it in a specific environment. The remaining cases arise from calls
-to \c{match.arg} which is used to look up the default choices for a variable and
-match against the choice passed by the caller.
-
-The most frequent ten expressions to eval calls from CRAN packages
-contribute to 77.1\% of all \eval calls.
-
-\begin{table}[!h]  \centering
-\begin{tabular}{@{}l@{~}|@{~}r@{~}r@{}} \hline
-Expression & Eval Call &  \% \\\hline
-\c{Environment} &                                  989302   & 61\%\\
-\c{column[rows] <<- what} &                        55677    & 3.5\%\\
-\c{function(value) freduce(value, `function.list`)} & 37251& 2.3\%\\
-\c{NULL} &                         32005    & 2\%\\
-\c{List} &                         22293    & 1.4\%\\
-\c{c("default", "default2012", "default2011" ...}& 20610    & 1.3\%\\
-\c{force(..1)}            &                        20461    & 1.3\%\\
-\c{alist(`spec`)}       &                         18532    & 1.2\%\\
-\c{inner}               &                          18530    & 1.2\%\\
-\c{String Vector}      &                           17487     & 1.1\%\\
-\end{tabular}\caption{Top ten eval calls in CRAN} \label{C}
-\end{table}
-
-The expression \c{Environment} occurs because of the four callsites explained
-above, \c{ggplot2::ggproto}, \c{R6::generator\_funs}, \c{future::backtrace} and
-\c{RModel::str.RMmodel}. The next expression, \c{column[rows] <<- what}, is used
-inside the \c{plyr::rbind.fill} function to merge data frames by assigning
-concatenated vectors to rows. The \c{<<-} operator is interesting in that it
-skips the current scope and assigns in a parent scope in which the variable is
-already present. In our corpus, all these \eval calls contribute to a single
-side-effect. The expression \c{function(value) freduce(value, `function.list`)}
-arises from the \c{magrittr::\%>\%} function which is a pipe operator that pipes
-the output of previous command to the next one. The expression is evaluated in a
-custom environment to create a function binding for evaluating the components of
-the pipe.\c{String Vector} and \c{List} also arise from the same function when a
-string or a list is piped using the \c{\%>\%} function into the next expression.
-The \c{NULL} arises from \c{R6::generator\_funs} function when the \eval is
-passed a \c{NULL} argument by the \c{DataMask\_generator} package. The
-\c{c("default", "default2012", "default2011" ...} pattern arises from
-\c{copula::polyG} where it reflectively access the default expression for its
-formal parameter and evaluates it. The \c{force(..1)} and \c{alist(`spec`)}
-patterns occur in \c{glue::glue\_data} function which concatenates and
-interpolates strings. The two patterns occur because the function captures
-unevaluated unnamed arguments and maps the evaluation of \c{force(..1)} on them.
-The \c{force} function forces promises and returns the result of evaluation. The
-\c{inner} pattern arises from \c{glue::identity\_transformer} which enables the
-creation of custom transformation functions for affecting the interpolation and
-concatenation of input by the \c{glue} package.
-
-\section{Why do we Eval}
-
-\subsection{Discussion}
-
-Yeah, why?
-
-\subsection{Can we do without?}
-
-We look at how \emph{consistent} the \c{expr} argument of \eval can be, \ie how
-many different types of the resolved \c{expr} there are per call sites. Most of
-the call sites, \ie \PercentMonomorphic, are \emph{consistent}, and this is
-similar to javascript. However, a few ones are highly \emph{polymorphic} (10
-different types). They are the pipe operators \c{\%>\%}, \c{\%<>\%} and
-\c{\%\$\%} in package \emph{magrittr}. It is effectively used to compose
-functions on their first argument, which can be of any type.
+\mypara{Unnecessary use of \eval}
 
@@
 
+\todo{Aviral's story when he contacted a developer about his use of eval}
+
 \section{Conclusion}

### 2021-04-05 paper/main.tex
Add few todos

@@
 
-The R language was intended to be extensible, the combination of lazy
+The R language was intended to be extensible. The combination of lazy
 evaluation, \c{substitute} and \c{eval} are the tools given to
@@
 
-\mypara{Intercession.} A common use case for \eval is to be combined with
-\c{match.call}. \c{match.call} walks up the call stack, captures the code that
-invoked the currently executing function, and returns it as an unevaluated
-expression. The pattern is to transform a call to some function \c{f} into a
-call to \c{g} with some arguments retained and others modified. As an
+\mypara{Intercession.} \todo{Pierre: update wrt the previous section if we want
+to keep it. Also perhaps we could start with a simpler pattern and then show
+this one} A common use case for \eval is to be combined with \c{match.call}.
+\c{match.call} walks up the call stack, captures the code that invoked the
+currently executing function, and returns it as an unevaluated expression. The
+pattern is to transform a call to some function \c{f} into a call to \c{g} with
+some arguments retained and others modified.  \todo{or change the target
+function. Both are described as "patterns" in the R documentation.} As an
 illustration, consider the \c{vcpart} package's function \c{tvcglm} that is
 translated to a call to \c{tvcm} with two modifications to the argument list:
-argument \c{control} can't be missing and \c{fit} is set \c{"glm"}. The function
-ends with a call to \c{eval.parent} to ensure that the rewritten call is
-evaluted in the same environment the original call was.
-\begin{lstlisting}
- tvcglm <- function(formula, data, family, ...) {
-   k <- match.call()
-   k[[1]] <- as.name("tvcm")
-   if ("fit" %in% names(list(...))) warning("'fit' ignored.")
-   k$fit <- "glm"
-   eval.parent(k)
- }
-\end{lstlisting}
-This pattern is recognizable by the fact that the expression is a call
-and the target environment is that of the parent.
+argument \c{control} can't be missing and \c{fit} is set \c{"glm"}. The
+function ends with a call to \c{eval.parent} to ensure that the rewritten call
+is evaluted in the same environment the original call was.  \begin{lstlisting}
+tvcglm <- function(formula, data, family, ...) { k <- match.call() k[[1]] <-
+as.name("tvcm") if ("fit" %in% names(list(...))) warning("'fit' ignored.")
+k$fit <- "glm" eval.parent(k) } \end{lstlisting} This pattern is recognizable
+by the fact that the expression is a call and the target environment is that of
+the parent.
 
@@
 subset of the arguments passed to this function; the variable is used to
-construct a call to \c{base::plot} which will draw a plot. \NOTE{This example is
-  a bit messed up. What is var? Does the use of ... in the eval grab the
-  arguments of this call? If yes does it not end up with xlab twice? HELP} The
-\parse and \eval combination is used by the the \source function in the base
-library to load R code from a file in the current workspace.
-\begin{lstlisting}
- plot.sback <- function(x,...) {
-   opt <- names(list(...))
-   stub <- paste(
-    ifelse("xlab" %in% opt,"",paste(",xlab=\"",var,"\"",sep="")),
-    ifelse("type" %in% opt,"",",type=\"l\""), sep="")
-   plot <- paste("plot(x.data,",stub,",...)",sep="")
-   eval(parse(text=plot))
-\end{lstlisting}
-This pattern is recognizable by its use of \c{parse} to turn a string
-into an expression.
-
-\mypara{Debloating.} \Eval is often used as a means to reduce boilerplate code;
+construct a call to \c{base::plot} which will draw a plot. \NOTE{This example
+  is a bit messed up. What is var? Does the use of ... in the eval grab the
+arguments of this call? If yes does it not end up with xlab twice? HELP}
+\todo{Does this comment still applies?} The \parse and \eval combination is
+used by the the \source function in the base library to load R code from a file
+in the current workspace.  \begin{lstlisting} plot.sback <- function(x,...) {
+opt <- names(list(...)) stub <- paste( ifelse("xlab" %in%
+opt,"",paste(",xlab=\"",var,"\"",sep="")), ifelse("type" %in%
+  opt,"",",type=\"l\""), sep="") plot <-
+  paste("plot(x.data,",stub,",...)",sep="") eval(parse(text=plot))
+  \end{lstlisting} This pattern is recognizable by its use of \c{parse} to turn
+  a string into an expression.
+
+\mypara{Debloating.} \todo{Wouldn't reducing boilerplate be a better name?} \Eval is often used as a means to reduce boilerplate code;
 simple and repetitive code can easily be replaced with judcious use of \eval.
@@
 trivial use is the empty expression, often found in JavaScript, but
-rare in R.
+rare in R. \todo{Given that we do not have any data I would get rid of it}
 

### 2021-04-05 analysis/corpus.Rmd
Update size and labs of the traced evals plot

@@
     geom_jitter(alpha=0.3) +
-    geom_text(nudge_y = .15, check_overlap = TRUE, size=4) +
+    geom_text(nudge_y = .25, check_overlap = TRUE, size=4) +
     scale_x_continuous(
@@
 #    geom_smooth(method = "lm", se = FALSE, show.legend = FALSE, linetype="dashed", color="red", size=0.2) +
-    labs(x="Eval call sites (log)", y="Traced eval call sites (log)", size="Package code size") +
-    theme(legend.position = c(0.15, 0.8))
+    labs(x="Eval call sites (log)", y="Traced eval call sites (log)", size="Package R code size") +
+    theme(
+      text = element_text(size=17),
+      legend.position = c(0.15, 0.8)
+    )
 ggsave(path(PLOT_DIR, "traced-eval-callsites.pdf"))

### 2021-04-06 analysis/environments.Rmd
Changes

@@
 
-  E_raw <- E_raw %>% mutate(envir_expression = if_else(envir_type == "NULL" & is.na(envir_expression), "NULL", envir_expression))
+  E_raw <- E_raw %>% mutate(envir_expression = if_else(envir_type == "NULL" & is.na(envir_expression), "NULL", envir_expression)) %>% 
+    mutate(envir_expression = na_if(envir_expression, "NA"))
 
@@
 ```{r envir_default}
-eval_calls_default <- E_envir %>% filter(is.na(envir))
+eval_calls_default <- E_envir %>% filter(is.na(envir)) # yes :(
+# Or use E_envir %>% filter(ev_variant == "eval", !str_detect(ev_call, "\\benvir\\b|eval\\([^,]+,"))
+# But that regex misses about 1000 sites
 
@@
 
-eval_calls_explicit <- E_envir %>% filter(!is.na(envir))
+eval_calls_explicit <- E_envir %>% filter(is.na(envir))
 nb_explicit <- count(eval_calls_explicit, wt = duplicates)
@@
 ```{r}
-envir_var <- frequent_envir %>% filter(envir != "NA", str_detect(envir, "^(?:\\.[:alpha]|[:alpha:])[[:alpha:][:digit:]_\\.]*$"))
+envir_var <- frequent_envir %>% filter(is.na(envir), str_detect(envir, "^(?:\\.[:alpha]|[:alpha:])[[:alpha:][:digit:]_\\.]*$"))
 
@@
 env_class_ranking <- E_envir %>% group_by(environment_class) %>% 
-  summarize(n = n_distinct(src_ref)) %>% 
+  summarize(n = n_distinct(src_ref), percent = round(100 * n / nb_call_sites, 2), cumpercent = cumsum(percent)) %>% 
   arrange(desc(n))
@@
 ```{r}
-frequent_envir %>% count(envir_type, sort = TRUE) %>% datatable()
+frequent_envir %>% count(envir_type, sort = TRUE) %>%
+  mutate(percent = round(100 * n / nb_call_sites, 2), cumpercent = cumsum(percent)) %>% 
+  datatable()
 ```
@@
 ```{r}
-E_envir %>% count(envir_type, sort = TRUE) %>% datatable()
+E_envir %>% count(envir_type, sort = TRUE)  %>%
+  mutate(percent = round(100 * n / nb_call_sites, 2), cumpercent = cumsum(percent)) %>% 
+ datatable()
 ```
@@
 nb_callee_env_sites <- specific_env %>% filter(specific_env == "callee") %>% .$n
+if(length(nb_callee_env_sites) == 0) nb_callee_env_sites <- 0
 nb_empty_env_sites <- specific_env %>% filter(specific_env == "empty") %>% .$n
+if(length(nb_empty_env_sites) == 0) nb_empty_env_sites <- 0
 nb_base_env_sites <- specific_env %>% filter(specific_env == "base") %>% .$n
+if(length(nb_base_env_sites) == 0) nb_base_env_sites <- 0
 

### 2021-04-06 paper/main.tex
More todo
Changes

--- 2021-04-05 paper/main.tex
@@
 \newcommand{\NOTE}[1]{{\it Note: #1}\xspace}
-\newcommand{\authorcomment}[3]{\xspace\textcolor{#1}{{\bf #2} #3}\xspace}
-\newcommand{\todo}[1]{\authorcomment{red}{TODO}{#1}}
 
@@
 
-\todo{Look at Improving dynamic code analysis by code abstraction, VPT'21}
-
 \newpage
@@
 \begin{wrapfigure}{r}{5cm} \hspace*{-12mm}
-  \centering
   \includegraphics[width=75mm]{pkgs-eval-callsites-hist.pdf} \caption{CRAN
@@
 This section focuses on CRAN packages and reports statistics about the usage of
-\eval. There are \PkgEvalCallSites call sites to \eval in \PkgPackages
-packages.  These represent \PkgPackagesRatio of all packages or \PkgCodeRatio
-of the entire code size of CRAN. Over half of the packages have less than 3
-\eval call sites and except \MaxEvalCallSitesPackage (\MaxEvalCallSitesCount
-\eval calls), all packages contain less than \MaxEvalCallSitesRest \eval call
-sites (\cf Fig~\ref{fig:pkgs-eval-callsites-hist}) These \eval calls appear in
+\eval. There are \PkgEvalCallSites calls to \eval in \PkgPackages packages.
+These represent \PkgPackagesRatio of all packages or \PkgCodeRatio of the
+entire code size of CRAN. Over half of the packages have less than 3 \eval
+call sites and except \MaxEvalCallSitesPackage (\MaxEvalCallSitesCount \eval
+calls), all packages contain less than \MaxEvalCallSitesRest \eval call sites
+(\cf Fig~\ref{fig:pkgs-eval-callsites-hist}) These \eval calls appear in
 \PkgFunsWithEval R functions (\CranFunsWithEvalRatio of all R functions in
@@
 packages. Out of \PkgEvalCallSitesRnd \eval call sites, \PkgHitEvalCallSites
-were hit at least once (\cf Fig~\ref{fig:traced-eval-callsites}). On average,
-the analysis covers \PkgHitEvalCallSitesAvgRatio of package \eval call sites,
-which similar to the \PkgCodeCoverage of average package code
-coverage\footnote{Not all packages with static call sites of \eval were
-observed triggering that call site, this can be chalked down to incomplete code
-coverage and occasional tracer failure (\PkgFailedProgramsRatio of programs do
-not terminate successfully or within 30 minute timeout).}.  Figure~\ref{freq}
-summarizes the frequency of dynamic calls to \eval with the left column being
-the number of calls and the right, the number of packages in that range. There
-are \Fewcalls packages with low \eval frequency, fewer than 100 calls; and a
-similar number of packages, \Manycalls to be precise, that use \eval more than
-1,000 times.  One package, \Maxcallspack, calls it \Maxcalls times and thus
-accounts for over half of the observed calls.
-
-\begin{figure}[tb]
-  \centering
-  \includegraphics[width=.8\textwidth]{traced-eval-callsites.pdf}
-  \centering
-  \caption{\eval call sites coverage of the \PkgPackages packages.}%
-  \label{fig:traced-eval-callsites}
-\end{figure}
+were hit at least once\footnote{Not all packages with static call sites of
+\eval were observed triggering that call site, this can be chalked down to
+incomplete code coverage and occasional tracer failure (\PkgFailedProgramsRatio
+of programs do not terminate successfully or within 30 minute timeout).}. On
+average, the analysis covers \PkgHitEvalCallSitesAvgRatio of package \eval call
+sites, which similar to the \PkgCodeCoverage of average package code coverage.
+Figure~\ref{freq} summarizes the frequency of dynamic calls to \eval with the
+left column being the number of calls and the right, the number of packages in
+that range. There are \Fewcalls packages with low \eval frequency, fewer than
+100 calls; and a similar number of packages, \Manycalls to be precise, that use
+\eval more than 1,000 times.  One package, \Maxcallspack, calls it \Maxcalls
+times and thus accounts for over half of the observed calls.
 
@@
 of fewer or equal to 25 nodes.
-The x-axis is the size of arguments in number of
-nodes, and the y-axis is the count of arguments with that size. The size drops
-rapidly, with few observations larger than 15 nodes. The long tail is omitted
-from the graph for legibility.
 %
@@
 %
+The x-axis is the size of arguments in number of
+nodes, and the y-axis is the count of arguments with that size. The size drops
+rapidly, with few observations larger than 15 nodes. The long tail is omitted
+from the graph for legibility.
 
@@
 \baseSmalleventspct of calls.
-\todo{More info about base and conclude it here.}
 
@@
 \kaggleSmalleventspct of \evals execute fewer than 50 instructions.
-\todo{More info about kaggle and conclude it here.}
 
@@
 
-\todo{Pierre: check the numbers - we want to know when it is an environment, when it is a list / data frame}
 Most of the time, \ie in \packageEnvirEnvSitePercent of the sites, \eval is
 passed an actual environment.
+
+In \packageNbEnvirIncorrectSites sites (\packageEnvirIncorrectSitePercent),
+\eval is passed an incorrect argument, for instance \c{NA}, a boolean, which
+makes \eval fail.
 
@@
 \begin{table}[htbp]{ \centering
-\begin{tabular}{c|c|c|c}\hline
-\multicolumn{2}{c|}{Core} & \multicolumn{2}{|c}{Packages}  \\\hline
-Environment Class & Eval Calls \% & Environment Class & Eval Calls\\\hline
-\CoreEnvClassA & \CoreProportionA & \PackagesEnvClassA &  \PackagesProportionA\\
-\CoreEnvClassB & \CoreProportionB & \PackagesEnvClassB &  \PackagesProportionB\\
-\CoreEnvClassC & \CoreProportionC & \PackagesEnvClassC &  \PackagesProportionC\\
-\CoreEnvClassD & \CoreProportionD & \PackagesEnvClassD &  \PackagesProportionD\\
-\CoreEnvClassE & \CoreProportionE & \PackagesEnvClassE &  \PackagesProportionE\\\hline
+\begin{tabular}{|c|c|}\hline
+ Environment Class & Eval Calls\\\hline
+ \PackagesEnvClassA &  \PackagesProportionA\\
+ \PackagesEnvClassB &  \PackagesProportionB\\
+ \PackagesEnvClassC &  \PackagesProportionC\\
+ \PackagesEnvClassD &  \PackagesProportionD\\
+\PackagesEnvClassE &  \PackagesProportionE\\\hline
 \end{tabular}} \label{tab:environments}
@@
 \subsection{Provenance.} 
-% Explain tha some generate expression objects and other language objects?
+% Explain that some generate expression objects and other language objects?
 There are various ways to obtain expressions that \eval will evaluate:
@@
     \item \c{match.call} captures a call and its arguments as expressions.
-    \item \c{quote}, \c{enquote} returns its argument unevaluated, as an R expression.
+    \item \c{quote}, \c{enquote} returns its argument unevaluated, as an R expression. % Aldd \c{call} ?
     \item \c{expression}  creates a vector of expression
@@
 \end{itemize}
+We show the distribution of the different provenances in Table~\ref{tab:provenance}.
+Note that one site could be passed arguments from different provenances depending on the invokation.
+For some arguments, we could not determine their provenance or it was a value or a symbol, for which the provenance is less significant.
+% Not clear...
+Or maybe add symbols in that table?
+% One can create a symbol with as.name, as.symbol etc. That would be the provenance (but a symbol can also come from quote for instance...)
 
@@
         \c{comp} & \packageNbBytecodeSites & \packageBytecodeSitePercent \\
+        \hline \hline 
+        Other & \packageNbUnknownProvSites & \packageUnknownProvSitePercent \\
         \hline
     \end{tabular}
-\caption{Provenances of \eval expressions. }
+    \caption{Provenances of \eval expressions. } \label{tab:provenance}
 \end{table}
 
-% \packageUnknownProvSitePercent : provenance not relavant (values, symbols etc)
-
-\c{eval(parse(...))} can be used for dynamic code loading. This forms
+We taint the result of a call to \c{parse}, \c{str2lang} and \c{str2expression}, by attaching an R attribute to it.
+R attributes are lost under copy or modification of their attaching object, so we underestimate the number of sites in that case.
+However, we observed that most of the time, the result of \c{parse} is not touched and directly passed to \eval.
+
+We taint the result of \c{match.call} by remembering the addresses of all the elements of the call list.
+As R GC is non-moving, it makes it possible to track them individually, as we observed that calls are modified and individual elements are sometimes evaluated.
+This is also an underestimation of the number of sites though, as we would lose an individual element of the call list that would be modified.
+
+% explain also how we get the numbers of the other categories?
+
+\c{eval(parse(...))} (and \c{str2expression}, and \c{str2lang}) can be used for dynamic code loading. This forms
 the core of \c{source} and \c{sys.source} functions in R that are
@@
 
-\subsection{Side effects in \eval}
-
-\todo{Filip: discuss the side-effects}
+In Kaggle, most of the sites, \ie \kaggleParseExprsSitePercent, are passed  an expressions that result from a call to \c{parse}.
+
+\mypara{New Bindings.} Find number of times, eval introduces new bindings. There
+are many ways -- library, load, attach, source, and explicitly introduce
+bindings using super-assign, assign and define.
+
+\mypara{C Calls.} Find number of times, eval makes call to native code using -
+.Call, .External, etc. Find out if there are native functions that are called
+only from within eval.
+
+\mypara{Non-local returns} Eval can do non-local returns effective bypassing
+evaluation of the rest of the function.
+For instance, it can inject a \c{break} or \c{next} instructions in a loop.
+It is very rare however, as it happens only twice in our dataset.
+% What kind of non-local returns?
+% eval(quote(return(89))) does not work
+
+\mypara{Purity.} Conclude with number of evals that are ``pure'', i.e. evals
+which could be ignored by a static analyzer without any problem.
 
@@
 
-\todo{Illustrate things that we have discussed in Section 5 - examples}
-\todo{Things to discuss: model.frame, match.call, substitute, parse, plots}
-
-% We present here the main uses of \eval we have witnessed in R.
+We present here the main uses of \eval we have witnessed in R.
 Sometimes, they are design patterns, sometimes, \eval could be replaced by more specific functions.
@@
 
-% \subsection{Variable lookup}
-
-% \eval is used to look-up the value of a variable in a given environment, different from the current scope so the \c{envir} argument of \eval is explicitly specified.
-
-% \subsection{Metaprogramming}
-
-% \mypara{\c{substitute}}
-
-% \mypara{\c{match.call}}
-
-% \todo{Talk about model.frame}
-
-% \subsection{Logging}
-
-% \subsection{Code transform}
-
-% % Derivation of symbolic expressions
-
-% \subsection{Plotting}
-
-% \section{\eval expressivity and macros}
+\subsection{Variable lookup}
+
+\eval is used to look-up the value of a variable in a given environment, different from the current scope so the \c{envir} argument of \eval is explicitly specified.
+
+% TODO: numbers on the prevalence of the pattern
+
+\eval is not necessary in that case and can be replaced by \c{get}, as follows:
+
+\begin{lstlisting}
+ eval(quote(x), envir = env)
+\end{lstlisting}
+$\Downarrow$
+\begin{lstlisting}
+get("x", envir = env)
+\end{lstlisting}
+
+% This is actually linked to NSE, as the symbol given to \eval usually comes
+% from match.call or susbtitute
+
+\subsection{Non-Standard Evaluation} % Or metaprogramming
+
+\subsection{\c{substitute}}
+
+\subsection{\c{match.call}}
+
+\subsection{Logging}
+
+\subsection{Code transform}
+
+% Derivation of symbolic expressions
+
+\subsection{Plotting}
+
+\section{\eval expressivity and macros}
 
@@
 There are no local macro variables in this implementation though.
+
+\section{The dangers of \eval} \label{sec:dangerseval}
+
+\subsection{Eval Usage in R}
 

### 2021-04-06 paper/main.tex

--- 2021-04-05 paper/main.tex
@@
 
+In Kaggle, most of the sites, \ie \kaggleParseExprsSitePercent, are passed  an expressions that result from a call to \c{parse}.
+
 \subsection{Side effects in \eval}
@@
 \todo{Illustrate things that we have discussed in Section 5 - examples}
-\todo{Things to discuss: model.frame, match.call, substitute, parse, plots}
+\todo{Things to discuss: model.frame, match.call, subset as example of NSE, substitute, parse, plots}
 

### 2021-04-06 paper/main.tex
Fix tag for package usage metrics

@@
 extracted from \CranPackages packages. Any run that did not exercise one of the
-\eval call sites in the packages was discarded. This left \Nbruns runs of some
-of the \Corpus packages to generate the data presented here. There were
-\Allcalls invocations of \eval and its variants occurring in \Triggeredpkgs
+\eval call sites in the packages was discarded. This left \packageNbruns runs of some
+of the \packageCorpus packages to generate the data presented here. There were
+\packageAllcalls invocations of \eval and its variants occurring in \packageTriggeredpkgs
 packages. Out of \PkgEvalCallSitesRnd \eval call sites, \PkgHitEvalCallSites
@@
 the number of calls and the right, the number of packages in that range. There
-are \Fewcalls packages with low \eval frequency, fewer than 100 calls; and a
-similar number of packages, \Manycalls to be precise, that use \eval more than
-1,000 times.  One package, \Maxcallspack, calls it \Maxcalls times and thus
+are \packageFewcalls packages with low \eval frequency, fewer than 100 calls; and a
+similar number of packages, \packageManycalls to be precise, that use \eval more than
+1,000 times.  One package, \packageMaxcallspack, calls it \packageMaxcalls times and thus
 accounts for over half of the observed calls.
@@
 &     \multicolumn{3}{c}{\small\#calls} &\small\#pck \\\hline
-\tt 1 &--& \tt 10      & \Bina  & \tt 1K &--&\tt 100K  & \Bine\\
-\tt 11 &--& \tt 100    & \Binb  & \tt 100K &--&\tt 1M  & \Binf\\
-\tt 101 &--& \tt 1K    & \Binc  & \tt 1M &--&\tt 10M   & \Bing\\
-\tt 1K &--& \tt 10K    & \Bind  & \tt 10M &--& \tt 100M & \Binh\\\hline
+\tt 1 &--& \tt 10      & \packageBina  & \tt 1K &--&\tt 100K  & \packageBine\\
+\tt 11 &--& \tt 100    & \packageBinb  & \tt 100K &--&\tt 1M  & \packageBinf\\
+\tt 101 &--& \tt 1K    & \packageBinc  & \tt 1M &--&\tt 10M   & \packageBing\\
+\tt 1K &--& \tt 10K    & \packageBind  & \tt 10M &--& \tt 100M & \packageBinh\\\hline
 \end{tabular}
@@
            & & & \c{.parent} &\\\hline
-\small Static sites &\Staticeval&\Staticevalq&\Staticevalparent&\Staticlocal \\
-\small Exercised sites&\Triggeredeval&\Triggeredevalq&\Triggeredevalparent&\Triggeredlocal\\
-\small Invocations&\EvalsRnd&\EvalqsRnd&\EparentsRnd&\LocalsRnd\\\hline
+\small Static sites &\packageStaticeval&\packageStaticevalq&\packageStaticevalparent&\packageStaticlocal \\
+\small Exercised sites&\packageTriggeredeval&\packageTriggeredevalq&\packageTriggeredevalparent&\packageTriggeredlocal\\
+\small Invocations&\packageEvalsRnd&\packageEvalqsRnd&\packageEparentsRnd&\packageLocalsRnd\\\hline
 \end{tabular}~\\[2mm]\caption{Variants}\label{tab:variantseval}
@@
 \multicolumn{3}{c}{\small\#calls} &\small\#sites \\\hline
-\tt 0 &--& \tt 50    & \Runbina & \tt 501 &--& \tt 1000   & \Runbine\\
-\tt 51 &--& \tt 100  & \Runbinb & \tt 1001 &--& \tt 1500  & \Runbinf\\
-\tt 101 &--& \tt 250 & \Runbinc & \tt 1501 &--& \tt 2000  & \Runbing\\
-\tt 251 &--& \tt 500 & \Runbind & \tt 2001 &--& \tt 3000 & \Runbinh\\\hline
+\tt 0 &--& \tt 50    & \packageRunbina & \tt 501 &--& \tt 1000   & \packageRunbine\\
+\tt 51 &--& \tt 100  & \packageRunbinb & \tt 1001 &--& \tt 1500  & \packageRunbinf\\
+\tt 101 &--& \tt 250 & \packageRunbinc & \tt 1501 &--& \tt 2000  & \packageRunbing\\
+\tt 251 &--& \tt 500 & \packageRunbind & \tt 2001 &--& \tt 3000 & \packageRunbinh\\\hline
 \end{tabular}
@@
 are average number of calls from a given site and given run, on the right are
-counts of sites that fall in that range. For instance, \Runbinh sites are
+counts of sites that fall in that range. For instance, \packageRunbinh sites are
 invoked on average over 2,000+ times per run. Larger numbers suggest loops or
-recursive contexts -- this seems to be the exception as most \evals, \Runbina,
+recursive contexts -- this seems to be the exception as most \evals, \packageRunbina,
 are exercised 50 times or less. Figure~\ref{cn}(b) zooms in on low frequency
@@
 The \eval can be passed any value, but if the argument is not an expression,
-\eval returns it without doing anything. Expressions account for \Codepercent of
-arguments in our corpus. More specifically, \Symbolpercent of arguments are
-\c{symbol}s (single variables such as \c{x}), \Languagepercent are \c{language}
-object (a single expression, such as \c{x+1}), and \Expressionpercent are
+\eval returns it without doing anything. Expressions account for \packageCodepercent of
+arguments in our corpus. More specifically, \packageSymbolpercent of arguments are
+\c{symbol}s (single variables such as \c{x}), \packageLanguagepercent are \c{language}
+object (a single expression, such as \c{x+1}), and \packageExpressionpercent are
 \c{expression} objects (multiple statements). Further inspection reveals that
-most symbols, \Ggplotsymbolpercent to be exact, come from a single call in the
+most symbols, \packageGgplotsymbolpercent to be exact, come from a single call in the
 \c{ggplot2} package and have have the value \c{\_inherit}. Even if we remove
@@
 objects, which could range in the MBs.} The median argument size is
-\Medianszeval (due to the many symbols) and the average is \Avgszeval nodes. The
-largest \eval input observed is \Maxszeval\,-- a significant chunk of code.
+\packageMedianszeval (due to the many symbols) and the average is \packageAvgszeval nodes. The
+largest \eval input observed is \packageMaxszeval\,-- a significant chunk of code.
 Figure~\ref{fig:sizedistribution} shows the distribution of sizes for arguments
@@
 interpreter. Most invocations perform relatively little work with
-\Smalleventspct of \evals executing 50 or fewer instructions. The violin plot of
+\packageSmalleventspct of \evals executing 50 or fewer instructions. The violin plot of
 Figure~\ref{ev}(a) corresponds to \evals executing $\leq$ 50 instructions, it is
 dominated by trivial symbol look ups. Figure~\ref{ev}(b) has the work-intensive
-\evals which go all the way to \MaxeventsRnd instructions.
+\evals which go all the way to \packageMaxeventsRnd instructions.
 

### 2021-04-06 analysis/corpus.Rmd
Fix corpus stats

@@
   read_fst(PACKAGE_SUM_CALLS_FILE) %>%
-  # remove evals that are inside the scripts,
-  # for PKG_C we are only interested in evals in packages
-  filter(!str_starts(eval_call_srcref, "::global::main")) %>% 
   as_tibble()
@@
 PKG_evals_per_package <- 
-  count(PKG_S, caller_package, eval_call_srcref) %>%
-  count(caller_package) %>% 
+  PKG_S %>%
+  group_by(eval_source) %>% 
+  summarise(
+    evals_hit=n_distinct(eval_call_srcref)
+  ) %>%
   rename(
-    package=caller_package, 
-    evals_hit=n
+    package=eval_source
   )
@@
   CRAN_C %>%
-  filter(in_corpus, package_code > 0) %>%
+  filter(in_corpus) %>%
   left_join(PKG_evals_per_package, by="package") %>% 
+  mutate_if(is.numeric, replace_na, 0) %>%
   mutate(
     evals_hit=as.integer(evals_hit),
-    # this is the case when there is eval in package loading code
-    # this we will always run
-    evals_hit=if_else(evals_hit > evals, evals, evals_hit),
+    # this is because some evals are defined in functions that are stored in environments (asserthat/base.r)
+    # or lists
+    evals=ifelse(evals < evals_hit, evals_hit, evals),
     evals_coverage=evals_hit/evals
   ) %>% 
-  mutate_if(is.numeric, replace_na, 0) %>%
   as_tibble()
@@
 CRAN_E <- 
-  read_csv(PACKAGE_EVALS_STATIC_FILE) %>% 
+  read_csv(
+    PACKAGE_EVALS_STATIC_FILE,
+    col_types=cols(
+      package = col_character(),
+      fun_name = col_character(),
+      srcref = col_character(),
+      call_fun_name = col_character(),
+      args = col_character()
+    )
+  ) %>% 
   # TODO: There are very few NA, plus when srcref is NA, the rest of the row in NA
@@
 KAGGLE_ALL <- local({
-  kernels <- read_csv(KAGGLE_KERNEL_FILE)
-  evals <- read_csv(KAGGLE_EVALS_STATIC_FILE) %>% count(package) %>% rename(evals=n)
+  kernels <- read_csv(
+    KAGGLE_KERNEL_FILE,
+    col_types = cols(
+      package = col_character(),
+      file = col_character(),
+      id = col_character(),
+      language = col_character(),
+      kernel_type = col_character(),
+      competition = col_character(),
+      blank = col_double(),
+      comment = col_double(),
+      code = col_double(),
+      hash = col_character()
+    )
+  )
+  evals <- read_csv(
+    KAGGLE_EVALS_STATIC_FILE,
+    col_types = cols(
+      package = col_character(),
+      fun_name = col_character(),
+      srcref = col_character(),
+      call_fun_name = col_character(),
+      args = col_character()
+    )
+  ) %>% 
+    count(package) %>% 
+    rename(evals=n)
   left_join(kernels, evals, by="package") 

### 2021-04-06 analysis/normalized.Rmd
Correct nb src refs

@@
 if(params$force_rebuild || rebuild) {
-  read_fst(calls_path) %>% as_tibble() -> E_raw
+  read_fst(calls_path) %>% as_tibble() %>% semi_join(C_raw, by = c("eval_source" = "package")) -> E_raw
 
-  E_raw <- E_raw %>% mutate(envir_expression = if_else(envir_type == "NULL", "NULL", envir_expression))
+  E_raw <- E_raw %>% mutate(envir_expression = if_else(envir_type == "NULL", "NULL", envir_expression)) %>% 
+    mutate(envir_expression = na_if(envir_expression, "NA"))
 

### 2021-04-06 analysis/environments.Rmd
Correct src ref numbers

@@
 if(params$force_rebuild || rebuild) {
-  read_fst(calls_path) %>% as_tibble() -> E_raw
+  read_fst(calls_path) %>% as_tibble() %>% semi_join(C_raw, by = c("eval_source" = "package")) -> E_raw
 
@@
 frequent_envir <- E_envir %>% group_by(src_ref) %>% 
-  summarize(n_envir = n_distinct(envir), envir = paste0(unique(envir), collapse = " ; "), envir_type = paste0(unique(envir_type), collapse = " ; "))
+  summarize(n_envir = n_distinct(envir), envir = paste0(unique(envir), collapse = " ; "), envir_type = paste0(unique(envir_type), collapse = " ; "), n_types = n_distinct(envir_type))
 
 #stopifnot(frequent_envir$n_envir <= 1)
+# 2 sites have 2 envir...
 ```
@@
 
+# TODO: handle scientific notation for caller-1e+00-
+
 env_calls <- eval_calls_absolute %>% ungroup() %>%
@@
 
+# Better-looking env class
+
+`caller-n` wins over any specific env. We add `+` if there are any `new`.
+
+```{r}
+env_calls <- env_calls %>%
+  mutate(env_class = if_else(environment_class == "loop", "loop", paste0(if_else(is.na(hierarchy), specific_env, as.character(hierarchy)), if_else(is.na(new_env), "", "+"))))
+```
+
 # `envir` is an integer

### 2021-04-09 paper/main.tex
background and related

@@
 
-This section provides a short introduction to R as it has a few
-surprising features that impact the use of reflective features of the
-language. We then look at the semantics of \eval in R and discuss some
-design choices. Lastly, we put this paper in context of previous work.
+This section provides a short introduction to R and the reflective features of
+the language; then looks at the semantics of \eval in R and discusses design
+choices; lastly, this paper is put in context of previous work.
 
@@
 \citet{ecoop12} gave a programming language-centric overview of the R language.
-They characterized it as a lazy, vectorized, functional language with rich
-dynamic features that allow layering several object systems. Most data types are
-sequences of primitive values constructed by calling the combination function
-\c{c}, so that \c{c(1,2)} evaluates to a vector of two strings, and constants
-such as \c{42} are vectors of length one. To enable equational reasoning, values
-accessible through multiple aliases are copied when written to. Furthermore,
-values can be tagged by user-defined attributes, key-value pairs. For instance,
-the attribute \c{dim}-\c{c(2,2)} can be attached to the value at \c{x} by the
-call \c{attr(x,"dim")$\leftarrow$c(2,2)}. In this case, the addition of this
-attribute ensure that the vector \c x will be treated as a $2 \times 2$ matrix
-by all arithmetic operations. The \c{class} attribute is used to give a 'class',
-in the object-oriented sense, to values. So, \c{class(x)<-"human"} sets the
-class of \c{x} to \c{human}; classes are used for method dispatch. Every
-linguistic construct is desugared to a function call, even control flow
-statements, assignments, and bracketing. All functions can be shadowed or
-redefined, making the language at the same time remarkably flexible and
-exceeding challenging to compile statically as vividly detailed
-by~\citet{dls19}. R uses a relaxed call-by-need convention for passing arguments
-to functions. Each argument is a thunk composed of an expression, its
-environment, and a slot for the result, these are called \emph{promises}. To get
-the value of an argument, the corresponding promise must be forced, once forced
-the promise's result is cached for future use.
+They characterized it as a lazy, vectorized, functional language with a rich
+complement of dynamic features expressive enough to layer several object systems
+on top of the core language. Most data types are sequences of primitive values
+constructed by calling the combination function \c{c}, for instance
+\c{c("Ha","bye")} evaluates to a vector of two strings, constants such as \c{42}
+are vectors of length one. To enable equational reasoning, values accessible
+through multiple aliases are copied when written to. Furthermore, values can be
+tagged by attributes, these are key-value pairs. For instance, the attribute
+\c{dim}-\c{c(2,2)} can be attached to the value at \c{x} by the call
+\c{attr(x,"dim")$\leftarrow$c(2,2)}. In this case, the addition of this
+attribute turns \c x into a matrix. The \c{class} attribute gives a 'class', in
+the object-oriented sense, to a value. So, \c{class(x)<-"human"} sets the class
+of \c{x} to \c{human}; classes are used for method dispatch. Every linguistic
+construct is desugared to a function call, even control flow statements,
+assignments, and bracketing. All functions can be shadowed or redefined, making
+the language at the same time remarkably flexible and exceeding challenging to
+compile statically as vividly detailed by~\citet{dls19}. R uses a relaxed
+call-by-need convention for passing arguments to functions. Each argument is a
+thunk composed of an expression, its environment, and a slot for the result,
+these are called \emph{promises}. To get the value of an argument, the
+corresponding promise must be forced, once forced the promise's result is cached
+for future use.
 
@@
 While a facility to transform data into code is available in many languages,
-some design choices affect the expressive power of \eval and related functions.
-The key design choices that we review here are the input format, the environment
-in which \eval-generated code is evaluated and the reflective operations that
-are available to that code. Fig.~\ref{comp} summarizes design choices for
-several popular languages.
-
-\begin{figure}[!t]\center\small
-\begin{tabular}{r@{~}l|l|l|l}\hline
+some design choices affect the expressive power of \eval. The key choices are
+the input format, the environment in which generated code is evaluated and the
+reflective operations available to that code. Fig.~\ref{comp} summarizes
+several designs.
+
+\begin{figure}[!t]\center\small\begin{tabular}{r@{~}l|l|l|l}\hline
 \tiny\sc Language&&\sc\tiny Input&\sc\tiny Scope&\tiny\sc Reflective operations\\\hline
@@
 \bf JavaScript&\cite{ecoop11}& text      & current, toplevel& data\\
-\bf R&\cite{R96}  & expression& programmatic      & data, stack, envir.\\\hline
-\end{tabular}
-\caption{Design space of \eval}\label{comp}
+\bf R&\cite{R96}  & expression& programmatic      & data, stack, environment\\\hline
+\end{tabular}\caption{Design space}\label{comp}
 \end{figure}
 
-The input to \eval can be any format that can be turned into code. Both Julia
-and R mandate that the input format be parsed expressions. JavaScript allows any
-expression as a character string. Java is the most restrictive as a classloader
-only accepts complete classes in the bytecode format recognized by the virtual
-machine.
-
-The environment in which \eval executes the code can be the lexically enclosing
-environment of \eval (current), the global environment (toplevel), a limited
-global environment (classloader), or an environment chosen by the caller of
-\eval (programmatic). The most restrictive variant is that of Java, where the
-scope of newly loaded code is the set of classes exposed by the current class
-loader. Julia is the next most restrictive, as only global definitions can be
-observed and modified by \eval. JavaScript has a variant of \eval that runs in
-the current scope and a most strict mode that is restricted to global variables.
-Finally, R allows the caller to choose which environment to use with \eval.
-
-The last degree of freedom in the design of \eval is what can be done from
-within the generated code. The main difference between designs lies in what part
-of the system state can be accessed and modified. In R, there is very little
-that cannot be observed from an \eval context and modified. The code can
-override the definition of variables and functions. Also, thanks to a rich
-reflective interface, code can introspect on values, walk and modify environment
-chains, and access the stack frames on the call stacks and their environments.
-
-Given the above, it is reasonable to claim that R's \eval are amongst the
-languages that afford the facility the most expressive power. This comes at the
-cost of the programmer's ability to reason about code in the presence of \eval.
-Expressive power translates directly into limits on the usefulness of program
-analysis.
-
-By contrast, the designers of Julia chose to limit \eval. In Julia, only global
-variables can be side-effected and environments cannot be readily manipulated.
-This is designed to shield optimized code from some of the most pernicious uses
-of the facility~\cite{oopsla18a}. Furthermore, the language provides a
-versioning mechanism, called world age, to ensure that any methods defined
-within an \eval only become visible at well-defined program points and thus that
-recompilation does not have to occur when optimized code is
-running~\cite{oopsla20a}.
+The input to \eval can be in any format convertible to code. JavaScript allows
+arbitrary strings to be used. Both Julia and R are more restrictive as they
+require expressions (or abstract syntax trees). Finally, Java is the most
+restrictive as its classloader only accepts complete classes in bytecode form.
+
+The choice of the environment of \eval is important as it determines how much of
+a program \eval can observe as well as the reach of potential side-effects
+performed by that operation. The most restrictive semantics is that of Java
+where newly loaded code evaluates in the environment defined by the classes
+visible from the current classloader. Julia and JavaScript's strict mode, limit
+\eval to the global environment. Finally, R is the most flexible as any
+accessible environment can be selected and passed to \eval.
+
+The last degree of freedom is the expressive power of the code executed by
+\eval. The main difference between languages lies in how much of the state of a
+program is accessible through reflective operations. Julia, Java and JavaScript
+all allow some form of introspection on the data. R is more flexible as it is
+possible to inspect the program's call stack. Thus any environment in the
+program can be inspected and modified.
+
+Given the above, the claim that R is amongst the languages with the most
+powerful \eval seems plausible. The rationale for R's design seems to have been
+to expose as much of the language and its internals as possible in order to
+maximize expressivity. In R, \eval is key tool to extend the language and
+implement DSLs, it is also a replacement for macros. By contrast, the designers
+of Julia chose to limit \eval. In Julia, only global variables can be
+side-effected and environments cannot be readily manipulated. This is designed
+to shield optimized code from some of the most pernicious uses of the
+facility~\cite{oopsla18a}. Furthermore, Julia provides a versioning
+mechanism, called world age, to ensure that any methods defined within an \eval
+only become visible at well-defined program points and thus that recompilation
+does not have to occur when optimized code is running~\cite{oopsla20a}.
 
@@
 
-The R language exposes a rich reflective interface with four functions with
-different semantics, \c{eval}, \c{evalq}, \c{eval.parent}, and
-\c{local}. The first is the most general, taking three arguments, an expression,
-its environment and an enclosing environment. The following discussion
-simplifies details not relevant to this paper, the interested reader should
-consult \citet{hadley}. It is defined as:
+The R language exposes a rich reflective interface with four main functions,
+\c{eval}, \c{evalq}, \c{eval.parent}, and \c{local}. The first is the most
+general with three arguments, an expression, its environment and enclosing
+environment. The following simplifies details not relevant to this paper, the
+interested reader should consult \citet{hadley}.
 
@@
 \end{lstlisting}
-Parameter \c e is the expression to evaluate, \c{envir} is the evaluation
+Parameter \c e is the \c{expression} to evaluate, \c{envir} is the evaluation
 environment, and \c{encl} is used to look up variables not found in \c{envir}.
-The latter is, by default, either the environment of the function that called
-\eval or the top level. Parameter \c e can be of different types, for our
-purposes we focus on {\tt expression}s. These can be thought of as abstract
-syntax trees returned by the parser. An \c{expression} can be obtained by
-calling \c{quote} with an expression or \c{parse} wit a string:
+The default value of latter is either the caller's environment or top level.
+\Eval can take several types for \c e, for our purposes on the \c{expression}.
+These can be thought of as the abstract syntax trees returned by the parser. An
+\c{expression} can be obtained by calling \c{quote} or \c{parse} wit a string:
 \begin{lstlisting}
@@
 \end{lstlisting}
-Even more commonly, expressions are obtained from promises using \c{substitute}.
-Consider a one-parameter function \c{f}, when called with \c{a+b}, the function
-returns \c{a+b} as expression.
+Usually, expressions are obtained from promises using \c{substitute}. Consider a
+function called with \c{x = a+b}, then \c f returns an expression object.
 \begin{lstlisting}
@@
 The call to \c{substitute(x)} extracts the unevaluated expression from the
-promise bound to variable \c x. Function \c{substitute(e,envir)} takes two
-arguments, the second is used as a substitution environment for free variables
-in \c e.
+promise bound to \c x. Function \c{substitute(e,envir)} takes two arguments, the
+second is used as a substitution environment for free variables in \c e.
 \begin{lstlisting}
@@
 \end{lstlisting}
-The \c{environment} function returns a variable environment, either from the
-currently executing function or the top level environment. Environments nest,
+The \c{environment} function returns the current environment. Environments nest,
 each has a parent. When creating a new environment with \c{new.env}, the parent
@@
 
-Richards et al.~\cite{ecoop11} provided the first large-scale study of
-the runtime behavior of \eval in JavaScript. They dynamically analyzed
-a corpus of the 10,000 most popular websites with an instrumented web
-browser to gather execution traces. They show that \eval is pervasive
-with 82\% of the most popular websites using it. The reasons for its
-use include the desire to load code on demand, deserialization of JSON
-dataq and lightweight meta-programming to customize web pages. While
-many uses were legitimate, just as many were unnecessary and could be
-replaced with equivalent and safer code. They categorized inputs to
-\eval so as to cover the vast majority of input strings. Restricting
-themselves to \eval in which all named variables refer to the global
-scope, many patterns could be replaced by more disciplined
-code~\cite{oopsla12b, moller12}. The work did not measure code
-coverage, so the numbers presented are a lower bound on the possible
-behaviors. Furthermore, JavaScript usage in 2011 is likely different
-from today, e.g. Node.js was not covered by Richards. More details
-about dynamic analysis of JavaScript can be found in~\cite{liang}.
-
-Wang et al.~\cite{wang} analyzed use of dynamic features in 18 Python
-programs to find if they affect file change-proneness. Files with
-dynamic features are significantly more likely to be the subject of
-changes, than other files. Chen et al. looked at the correlation
-between code changes and dynamic features, including \eval, in 17
-Python programs~\cite{chen}. They did not observe many uses of \eval.
-Callau et al.~\cite{oscar} performed an empirical study of the usage
-of dynamic features in 1,000 Smalltalk projects. While \eval itself is
-not present, Smalltalk has a rich reflective interface. The authors
-found that reflective are used in less than 2\% of methods. The most
-common reflective method is \c{perform:}; it send a message that is
-specified by a string. These features are mostly used in the core
-libraries.
-
-Bodden et al.~\cite{bodden} looked at usage of reflection in the Java
-DaCapo benchmark suite. They found that dynamic loading was triggered
-by the benchmark harness. The harness then executes methods via
-reflection, this caused static analysis tools to generate an incorrect
-call graph for the programs in DaCapo.
-
-Morandat et al.~\cite{ecoop12} had a short section on the usage of
-\eval in R. They found the it widely used in R code with 8500 call
-sites in CRAN and 2.2 million dynamic calls. The 15 most frequent call
-sites account for 88\% of those. The \c{match.arg} function is the
-highest used one with 54\% of all calls. In the other call sites, they
-saw two uses cases. The most common is the evaluation of the source
-code of a promise retrieved by \c{substitute} in a new environment;
-e.g. as done in the \c{with} function. The other use case is the
-invocation of a function whose name or arguments are determined
-dynamically. For this purpose, R provides \c{do.call} and thus \eval
-is overkill.
-
-\todo{Look at Improving dynamic code analysis by code abstraction, VPT'21}
+\citet{ecoop11} provided the first large-scale study of the runtime behavior of
+\eval in JavaScript. They dynamically analyzed a corpus of the 10,000 most
+popular websites with an instrumented web browser to gather execution traces.
+They show that \eval is pervasive with 82\% of the most popular websites using
+it. The reasons for its use include the desire to load code on demand,
+deserialization of JSON dataq and lightweight meta-programming to customize web
+pages. While many uses were legitimate, just as many were unnecessary and could
+be replaced with equivalent and safer code. They categorized inputs to \eval so
+as to cover the vast majority of input strings. Restricting themselves to \eval
+in which all named variables refer to the global scope, many patterns could be
+replaced by more disciplined code~\cite{oopsla12b, moller12}. The work did not
+measure code coverage, so the numbers presented are a lower bound on the
+possible behaviors. Furthermore, JavaScript usage in 2011 is likely different
+from today, e.g. Node.js was not covered by Richards. More details about dynamic
+analysis of JavaScript can be found in~\cite{liang}.
+
+\citet{wang} analyzed use of dynamic features in 18 Python programs to find if
+they affect file change-proneness. Files with dynamic features are significantly
+more likely to be the subject of changes, than other files. Chen et al. looked
+at the correlation between code changes and dynamic features, including \eval,
+in 17 Python programs~\cite{chen}. They did not observe many uses of \eval.
+\citet{oscar} performed an empirical study of the usage of dynamic features in
+1,000 Smalltalk projects. While \eval itself is not present, Smalltalk has a
+rich reflective interface. The authors found that reflective are used in less
+than 2\% of methods. The most common reflective method is \c{perform:}; it send
+a message that is specified by a string. These features are mostly used in the
+core libraries.
+
+\citet{bodden} looked at usage of reflection in the Java DaCapo benchmark suite.
+They found that dynamic loading was triggered by the benchmark harness. The
+harness then executes methods via reflection, this caused static analysis tools
+to generate an incorrect call graph for the programs in DaCapo.
+
+\citet{Arceri21} study \eval in JavaScript from a software security point of
+view. The authors report that 53\% of the malware they studied used \eval as a
+means to obfuscate attack code or to mount attacks. They propose an abstract
+interpretation-based approach to analyzing dynamic languages. One must construct
+a static approximation of the argument to \eval and then analyze possible
+behaviors of the interpreter when evaluating the generated code.
+
+\citet{ecoop12} had a short section on the usage of \eval in R. They found the
+it widely used in R code with 8500 call sites in CRAN and 2.2 million dynamic
+calls. The 15 most frequent call sites account for 88\% of those. The
+\c{match.arg} function is the highest used one with 54\% of all calls. In the
+other call sites, they saw two uses cases. The most common is the evaluation of
+the source code of a promise retrieved by \c{substitute} in a new environment;
+e.g. as done in the \c{with} function. The other use case is the invocation of a
+function whose name or arguments are determined dynamically. For this purpose, R
+provides \c{do.call} and thus \eval is overkill.
 
@@
 
-\begin{figure}[tb]
-  \centering
-  \includegraphics[width=.8\textwidth]{traced-eval-callsites.pdf}
-  \centering
-  \caption{\eval call sites coverage of the \PkgPackages packages.}%
-  \label{fig:traced-eval-callsites}
-\end{figure}
-
 \begin{figure}[b]
@@
 \todo{More info about base and conclude it here.}
+
+\begin{figure}[tb] \centering
+  \includegraphics[width=.8\textwidth]{traced-eval-callsites.pdf} \centering
+  \caption{\eval call sites coverage of the \PkgPackages packages.}%
+  \label{fig:traced-eval-callsites}
+\end{figure}
 

### 2021-04-09 paper/main.tex
methods

@@
 
+\newpage
 \section{Background and Previous work}
@@
 the language; then looks at the semantics of \eval in R and discusses design
-choices; lastly, this paper is put in context of previous work.
+choices; lastly, this work is put in context.
 
@@
 
-While a facility to transform data into code is available in many languages,
-some design choices affect the expressive power of \eval. The key choices are
-the input format, the environment in which generated code is evaluated and the
-reflective operations available to that code. Fig.~\ref{comp} summarizes
-several designs.
-
-\begin{figure}[!t]\center\small\begin{tabular}{r@{~}l|l|l|l}\hline
+While a data-to-code facility is available in many languages, some design
+choices affect the expressive power of \eval. The key choices are the input
+format, the environment in which generated code is evaluated and the reflective
+operations available to that code. Fig.~\ref{comp} summarizes designs.
+
+\begin{figure}[!h]\center\small\begin{tabular}{r@{~}l|l|l|l}\hline
 \tiny\sc Language&&\sc\tiny Input&\sc\tiny Scope&\tiny\sc Reflective operations\\\hline
@@
 
-The R language exposes a rich reflective interface with four main functions,
-\c{eval}, \c{evalq}, \c{eval.parent}, and \c{local}. The first is the most
-general with three arguments, an expression, its environment and enclosing
-environment. The following simplifies details not relevant to this paper, the
-interested reader should consult \citet{hadley}.
+R exposes a rich reflective interface with functions, \c{eval},
+\c{evalq}, \c{eval.parent}, and \c{local}. The following simplifies details not
+relevant to this paper, the interested reader should consult \citet{hadley}.
 
@@
 \end{lstlisting}
-Parameter \c e is the \c{expression} to evaluate, \c{envir} is the evaluation
-environment, and \c{encl} is used to look up variables not found in \c{envir}.
-The default value of latter is either the caller's environment or top level.
-\Eval can take several types for \c e, for our purposes on the \c{expression}.
-These can be thought of as the abstract syntax trees returned by the parser. An
-\c{expression} can be obtained by calling \c{quote} or \c{parse} wit a string:
+The most general is \eval. Parameter \c e is the \c{expression} to evaluate,
+\c{envir} is the evaluation environment, and \c{encl} is used to look up
+variables not found in \c{envir}. The default value of latter is either the
+caller's environment or top level. \Eval can take several types for \c e, for
+our purposes on the \c{expression}. These can be thought of as the abstract
+syntax trees returned by the parser. An \c{expression} can be obtained by
+calling \c{quote} or \c{parse} with a string:
 \begin{lstlisting}
@@
 The call to \c{substitute(x)} extracts the unevaluated expression from the
-promise bound to \c x. Function \c{substitute(e,envir)} takes two arguments, the
-second is used as a substitution environment for free variables in \c e.
+promise \c x. Function \c{substitute(e,envir)} takes two arguments, the second
+is used to substitute free variables in \c e.
 \begin{lstlisting}
@@
 \end{itemize}
-\noindent Programmers sometimes use environments as hash maps as they are the
-only data type with reference semantics and have a built-in string look up.
-Functions \c{parent.frame} and \c{sys.frame} return environments further up the
-call stack.
+\noindent Environments are used as hash maps as they have reference semantics
+and a built-in string look up. Functions \c{parent.frame} and \c{sys.frame}
+return environments further up the call stack.
  \begin{lstlisting}
@@
 \end{lstlisting}
-The three \eval variants can be expressed with \eval. The \c{evalq} form quotes
-the argument to prevent it from being evaluated in the current environment. The
-\c{eval.parent(e,n)} form evaluates \c{e} in the environment of the {\tt n}-th
-caller of this function. Finally, \c{local} evaluates \c{e} in a new environment
+The three variants can be expressed with \eval: \c{evalq} quotes its
+argument to prevent evaluation in the current environment,
+\c{eval.parent(e,n)} evaluates \c{e} in the environment of the {\tt n}-th
+caller of this function; finally, \c{local} evaluates \c{e} in a new environment
 to avoid polluting the current one.
@@
 replaced by more disciplined code~\cite{oopsla12b, moller12}. The work did not
-measure code coverage, so the numbers presented are a lower bound on the
-possible behaviors. Furthermore, JavaScript usage in 2011 is likely different
-from today, e.g. Node.js was not covered by Richards. More details about dynamic
-analysis of JavaScript can be found in~\cite{liang}.
+measure code coverage, so the numbers presented are a lower bound on possible
+behaviors. Furthermore, JavaScript usage in 2011 is likely different from today,
+e.g. Node.js was not covered. More details about dynamic analysis of JavaScript
+can be found in~\cite{liang}.
 
@@
 
-\mypara{Base.} The base dataset contains \BasePackages core libraries
-performing basic arithmetics, statistics, and operating system related
-functionalities. These libraries are bundled with R and are executed
-pervasively. Together they contain \BaseEvalCallSites calls to \eval in
-\BaseFunsWithEvals functions. Some of these functions provide basic R
-functionalities such as package loading and thus there is hardly any R code
-which does not invoke one of these \eval call sites.
-
-\mypara{CRAN.} The Comprehensive R Archive
-Network\footnote{\url{cran.r-project.org}} is the largest curated repository of
-R packages. It hosts \CranAvailablePackagesRnd packages with 6 new one submitted
-each day~\cite{Ligges2017}. Packages accepted to CRAN are authored by
-experienced developers and abide to a number of well-formedness rules
-automatically checked on each commit. Furthermore, each package comes with
-sample data. There are three sources of runnable code in a package: \emph{tests,
-examples} and \emph{vignettes} -- respectively, unit tests, code snippets from
-the documentation, and long-form use-cases written in RMarkdown.
-Examples and vignettes can be turned into scripts by extracting the relevant
-code. From \CranPackages packages we extracted \CranRunnableScripts scripts that contain
-\CranRunnableCode lines of code (\CranRunnableCodeExamplesRnd in examples,
-\CranRunnableCodeVignettesRnd in vignettes and \CranRunnableCodeTestsRnd in
-tests).
-
-\mypara{Kaggle.} The Kaggle website\footnote{\url{kaggle.com}} is an online
+\mypara{Base.} The dataset contains \BasePackages libraries performing basic
+arithmetics, statistics, and operating system functionalities. These libraries
+are bundled with R and executed pervasively. Together they contain
+\BaseEvalCallSites calls to \eval in \BaseFunsWithEvals functions. Some of these
+functions provide basic functionalities such as package loading and thus there
+is hardly any code which does not invoke one of these.
+
+\mypara{CRAN.} The Comprehensive R Archive Network
+(\small{\url{cran.r-project.org}}) is the largest curated repository of R
+packages. It hosts \CranAvailablePackagesRnd packages with 6 new ones submitted
+daily~\cite{Ligges2017}. Packages are authored by experienced developers and
+abide by well-formedness rules automatically checked on each commit.
+Each package comes with sample data. There are three sources of
+runnable code in a package: \emph{tests, examples} and \emph{vignettes} --
+respectively, unit tests, code snippets from the documentation, and long-form
+use-cases written in RMarkdown. Examples and vignettes can be turned into
+scripts by extracting the relevant code. From \CranPackages packages we
+extracted \CranRunnableScripts scripts that contain \CranRunnableCode lines of
+code (\CranRunnableCodeExamplesRnd in examples, \CranRunnableCodeVignettesRnd in
+vignettes and \CranRunnableCodeTestsRnd in tests).
+
+\mypara{Kaggle.} The Kaggle website (\small{\url{kaggle.com}}) is an online
 platform for data science. It allows users to submit problems and compete to
@@
 \begin{figure}[!h]\hspace{-5mm}
-  \includegraphics[width=1.05\linewidth]{pipeline.pdf}
+  \includegraphics[width=.95\linewidth]{pipeline.pdf}
   \caption{Pipeline}\label{fig:pipeline}
@@
 
-\mypara{Tracer.} The tracer registers callbacks to all \eval functions as well
-as a few other ones that allow us to better identify where do the expressions
-come from. For example, we taint the results of calls to \c{parse} (turns
-string into expression) and \c{match.call} (reflects on current call). We also
-capture calls to R API for dynamic code loading (\eg\xspace \c{library},
-\c{require} and \c{source}). Next, we subscribe to the events related to a
-variable definition and assignment allowing us to record side effects that
-happen in environments while evaluating code in \eval. One challenge we
-encountered during the analysis was that essentially R only provides source
-references for top-level calls in a function body or in the block surrounded
-by braces. That is to say, a method whose body would \c{eval(x)} would not have
-debug information that we can use to identify that particular \eval. The same
-expression surrounded by braces would have a line number and source reference.
-This is unfortunately not easily fixed in the R implementation. Instead, we
-extended the dynamic analysis tool to attach synthetic source code references
-to all \eval call sites by traversing AST.  The tracer is implemented as an R
-package in 3.2K C++ and 1.3K of R code. For performance reason, most of the
-tracing is done in C++. While in theory the implementation is rather
-straightforward, what makes it hard in practice is that (1) some of the core R
-functionality such as package loading is implemented using \eval which then
-leaks to the tracer in unexpected places, (2) lazy evaluation makes it harder
-to analyze function arguments while tracing as prematurely forcing a promise
-might have consequences, and finally (3) running large amount of real-world
-code triggers exercises a lot of corner cases of highly underspecified R
-behavior.
+The tracer registers callbacks to all \eval functions as well as a few other
+ones that allow us to better identify where do the expressions come from. For
+example, we taint the results of calls to \c{parse} (turns string into
+expression) and \c{match.call} (reflects on current call). We also capture calls
+to R API for dynamic code loading (\eg\xspace \c{library}, \c{require} and
+\c{source}). Next, we subscribe to the events related to a variable definition
+and assignment allowing us to record side effects that happen in environments
+while evaluating code in \eval. One challenge we encountered during the analysis
+was that essentially R only provides source references for top-level calls in a
+function body or in the block surrounded by braces. That is to say, a method
+whose body would \c{eval(x)} would not have debug information that we can use to
+identify that particular \eval. The same expression surrounded by braces would
+have a line number and source reference. This is unfortunately not easily fixed
+in the R implementation. Instead, we extended the dynamic analysis tool to
+attach synthetic source code references to all \eval call sites by traversing
+AST. The tracer is implemented as an R package in 3.2K C++ and 1.3K of R code.
+For performance reason, most of the tracing is done in C++. While in theory the
+implementation is rather straightforward, what makes it hard in practice is that
+(1) some of the core R functionality such as package loading is implemented
+using \eval which then leaks to the tracer in unexpected places, (2) lazy
+evaluation makes it harder to analyze function arguments while tracing as
+prematurely forcing a promise might have consequences, and finally (3) running
+large amount of real-world code triggers exercises a lot of corner cases of
+highly underspecified R behavior.
 

### 2021-04-09 paper/main.tex
Patterns

--- 2021-04-06 paper/main.tex
@@
 The top-level environment in R is called the global environment. New
-environments can be created using \c{new.env}. They can be provided a parent
+environments can be created using \c{new.env}, or \c{list2env}. They can be provided a parent
 environment which becomes the enclosing scope of the new environment.
+% as.environment also can be used sometimes for that
+% list2env creates an environment only if its argument envir is not provided
 
@@
 
-\subsubsection{Bits of text, possibly reusable}
-
-\todo{Pierre: check the numbers - we want to know when it is an environment, when it is a list / data frame}
 Most of the time, \ie in \packageEnvirEnvSitePercent of the sites, \eval is
@@
 
-\c{NULL}, a \c{list} or a data frame, or an integer, happen in
-\packageEnvirUnusualSitePercent of the cases: \eval copies the fields of the
+\c{NULL}, a \c{list} or a data frame, or an integer: \eval copies the fields of the
 list or data frame and creates bindings for them in a new environment.
+
+\subsubsection{Bits of text, possibly reusable}
 
@@
 
-\mypara{Intercession.} \todo{Pierre: update wrt the previous section if we want
-to keep it. Also perhaps we could start with a simpler pattern and then show
-this one} A common use case for \eval is to be combined with \c{match.call}.
+\mypara{Variable lookup.} % predominant usage of eval in base
+
+\mypara{Symbolic differentiation.} % D and deriv
+
+\mypara{Capturing a call.}  A common use case for \eval is to be combined with \c{match.call}.
 \c{match.call} walks up the call stack, captures the code that invoked the
-currently executing function, and returns it as an unevaluated expression. The
+currently executing function, and returns it as an unevaluated expression that includes the call function and its unevaluated arguments. The
 pattern is to transform a call to some function \c{f} into a call to \c{g} with
@@
 function ends with a call to \c{eval.parent} to ensure that the rewritten call
-is evaluted in the same environment the original call was.  \begin{lstlisting}
+is evaluated in the same environment the original call was.  \begin{lstlisting}
 tvcglm <- function(formula, data, family, ...) { k <- match.call() k[[1]] <-
@@
 the parent.
+
+% We have quantitative data on that if needed...
+This pattern is very often (more than half of the cases) used in relation with statistical models, to build a new model.
+\c{model.frame} to build a new generic model, or a more specialized model like \c{glm} or, \c{lm}.
+% Show example
+
+A related pattern is not to change the call but rather to evaluate the arguments in the parent frame. This can also be achieved with \c{substitute}.
+% Example
+
+In the R community, the call capturing pattern is often referred~\cite{hadley} as \emph{Non-Standard Evaluation}.
 
@@
 \mypara{Debloating.} \todo{Wouldn't reducing boilerplate be a better name?} \Eval is often used as a means to reduce boilerplate code;
-simple and repetitive code can easily be replaced with judcious use of \eval.
+simple and repetitive code can easily be replaced with judicious use of \eval.
 For example, the \c{data.table} package uses \eval to calls the \c{options}
@@
 rare in R. \todo{Given that we do not have any data I would get rid of it}
+\todo{We have the normalized expressions V and the small study I did on values passed to eval. And I did not find any eval() (empty) in the dataset. }
+
+% Creating macros?
 

### 2021-04-09 paper/main.tex
usage

--- 2021-04-09 paper/main.tex
@@
   \centering
-  \includegraphics[width=75mm]{pkgs-eval-callsites-hist.pdf} \caption{CRAN
+  \includegraphics[width=72mm]{pkgs-eval-callsites-hist.pdf} \caption{CRAN
   \eval call sites}%
@@
 This section focuses on CRAN packages and reports statistics about the usage of
-\eval. There are \PkgEvalCallSites call sites to \eval in \PkgPackages
-packages.  These represent \PkgPackagesRatio of all packages or \PkgCodeRatio
-of the entire code size of CRAN. Over half of the packages have less than 3
-\eval call sites and except \MaxEvalCallSitesPackage (\MaxEvalCallSitesCount
-\eval calls), all packages contain less than \MaxEvalCallSitesRest \eval call
-sites (\cf Fig~\ref{fig:pkgs-eval-callsites-hist}) These \eval calls appear in
-\PkgFunsWithEval R functions (\CranFunsWithEvalRatio of all R functions in
-CRAN).
-
-For the dynamic analysis, we execute all the \CranRunnableScripts programs
-extracted from \CranPackages packages. Any run that did not exercise one of the
-\eval call sites in the packages was discarded. This left \packageNbruns runs of some
-of the \packageCorpus packages to generate the data presented here. There were
-\packageAllcalls invocations of \eval and its variants occurring in \packageTriggeredpkgs
-packages. Out of \PkgEvalCallSitesRnd \eval call sites, \PkgHitEvalCallSites
-were hit at least once (\cf Fig~\ref{fig:traced-eval-callsites}). On average,
-the analysis covers \PkgHitEvalCallSitesAvgRatio of package \eval call sites,
-which similar to the \PkgCodeCoverage of average package code
-coverage\footnote{Not all packages with static call sites of \eval were
-observed triggering that call site, this can be chalked down to incomplete code
-coverage and occasional tracer failure (\PkgFailedProgramsRatio of programs do
-not terminate successfully or within 30 minute timeout).}.  Figure~\ref{freq}
-summarizes the frequency of dynamic calls to \eval with the left column being
-the number of calls and the right, the number of packages in that range. There
-are \packageFewcalls packages with low \eval frequency, fewer than 100 calls; and a
-similar number of packages, \packageManycalls to be precise, that use \eval more than
-1,000 times.  One package, \packageMaxcallspack, calls it \packageMaxcalls times and thus
-accounts for over half of the observed calls.
-
-\begin{figure}[b]
-
+\eval. We use the word \emph{site} to an occurrence of a call site to the \eval
+function in the source code, and \emph{call} to denote an observed invocation of
+the \eval function.
+
+\subsection{CRAN}
+
+There are \PkgEvalCallSites \eval sites in \PkgPackages packages. The proportion
+of packages calling \eval is thus \PkgPackagesRatio. Over half of these packages
+have fewer than 3 sites, and with the exception \MaxEvalCallSitesPackage which
+has \MaxEvalCallSitesCount sites, all packages contain fewer than
+\MaxEvalCallSitesRest sites (\cf Fig~\ref{fig:pkgs-eval-callsites-hist} which
+shows a histogram of sites per package). These sites appear in \PkgFunsWithEval
+functions (\CranFunsWithEvalRatio of all functions in CRAN).
+
+\begin{figure}[!b]
+\small
 \begin{tabular}{@{}l@{\hspace{1.5cm}}l@{}}
@@
 \end{tabular}~\\[2mm]\caption{Variants}\label{tab:variantseval}
-\end{minipage}\end{tabular}\end{figure}
-
-Figure~\ref{tab:variantseval} enumerate the use of variants of \eval. For each
-of the four variants, the first rows shows the number of call sites in the
-corpus (\emph{static}), the second row is the number of call sites that were
-encountered during analysis (\emph{exercised}), the last row is the number of
-calls (\emph{invocations}). The overwhelming majority of call sites and dynamic
-calls, are to \eval itself, \c{eval.parent} is rare, and both \c{evalq} and
-\c{local} are barely used at all. The difference between sites and exercised
-sites underscores the limitations due to code coverage.
+\end{minipage}\end{tabular}
+
+%\end{figure}
+
+%\begin{figure}[b] \centering
+  \includegraphics[width=.78\textwidth]{traced-eval-callsites.pdf} \centering
+  \caption{\eval call sites coverage of the \PkgPackages packages.}%
+  \label{fig:traced-eval-callsites}
+\end{figure}
+
+For dynamic analysis, we run \CranRunnableScripts programs extracted from
+\CranPackages packages. Any run that does not exercise \eval is discarded. This
+left \packageNbruns runs from \packageCorpus packages. There were
+\packageAllcalls calls in \packageTriggeredpkgs packages originating from
+\PkgHitEvalCallSites unique sites. In terms of coverage, the data exercised
+\PkgHitEvalCallSitesAvgRatio of sites, a coverage similar to the code coverage
+metric for the packages which is \PkgCodeCoverage. The fact that not all sites
+are exercised can be can be chalked down to incomplete tests occasional analysis
+failures (\PkgFailedProgramsRatio of programs crashed or timed out).
+Fig~\ref{fig:traced-eval-callsites} shows the number of sites that were
+exercised; coverage is unequal. Figure~\ref{freq} summarizes the frequency of
+dynamic calls to \eval with the left column being the number of calls and the
+right, the number of packages in that range. There are \packageFewcalls packages
+with low \eval frequency, fewer than 100 calls; and a smaller, but still
+significant, number of packages, \packageManycalls to be precise, that use \eval
+more than 1,000 times. Package \packageMaxcallspack makes \packageMaxcalls calls
+and thus accounts for over half of the observations.
+Figure~\ref{tab:variantseval} summarizes the use of variants of \eval. For each
+of the four variants, the first rows shows the number of sites in the corpus
+(\emph{static}), the second row is the number of sites that were encountered
+during analysis (\emph{exercised}), the last row is the number of calls
+(\emph{invocations}). The overwhelming majority of sites and calls, are to \eval
+itself, \c{eval.parent} is rare, and both \c{evalq} and \c{local} are barely
+used at all. The difference between sites and exercised sites underscores the
+limitations due to code coverage.
 
 \begin{wrapfigure}{4}{5.8cm}
-\vspace{-4mm}
 \centering
@@
 
-  \medskip  (a) All
+  \medskip  (a) All  \medskip  \medskip
+
+  \includegraphics[width=5.6cm, trim=5.5cm 0 5cm 0, clip]{calls_per_run_per_call_site}
+
+  (b) Small
+
+\caption{Normalized calls} \label{cn}\vspace{-2mm}
+
 \medskip
@@
 
-  \includegraphics[width=5.4cm, trim=5.5cm 0 5cm 0, clip]{calls_per_run_per_call_site}
-
-  (b) Small
-
-\vspace{-2mm} \caption{Normalized invocations} \label{cn}\vspace{-2mm}
+\begin{tabular}{c}
+  {\hspace{-3cm}\includegraphics[width=0.8\textwidth]{size_loaded_distribution}}
+\end{tabular}
+ \caption{Loaded code} \label{fig:sizedistribution}
 \end{wrapfigure}
 
-Figure~\ref{cn}(a) shows normalized invocation counts per call site; on the left
-are average number of calls from a given site and given run, on the right are
-counts of sites that fall in that range. For instance, \packageRunbinh sites are
-invoked on average over 2,000+ times per run. Larger numbers suggest loops or
-recursive contexts -- this seems to be the exception as most \evals, \packageRunbina,
-are exercised 50 times or less. Figure~\ref{cn}(b) zooms in on low frequency
-call sites. The x-axis shows normalized invocations and the y-axis is the number
-sites for that value. Most low-frequency \evals are invoked only once, about
-half as many are invoked twice; after that the frequency quickly drops.
-
-The \eval can be passed any value, but if the argument is not an expression,
-\eval returns it without doing anything. Expressions account for \packageCodepercent of
-arguments in our corpus. More specifically, \packageSymbolpercent of arguments are
-\c{symbol}s (single variables such as \c{x}), \packageLanguagepercent are \c{language}
-object (a single expression, such as \c{x+1}), and \packageExpressionpercent are
+Figure~\ref{cn}(a) shows normalized call counts per site; on the left are
+average number of calls from a given site and given run, on the right are counts
+of sites that fall in that range. For instance, \packageRunbinh sites are
+invoked 2,000+ times per run. Larger numbers suggest loops or recursive contexts
+-- this seems to be the exception as most \evals, \packageRunbina, are exercised
+50 times or less. Figure~\ref{cn}(b) zooms in on low frequency sites. The x-axis
+shows normalized calls and the y-axis is the number sites for that value. Most
+low-frequency \evals are invoked only once, about half as many are invoked
+twice; after that the frequency quickly drops.
+
+\Eval takes any value, but if its argument is not an expression, \eval returns
+it unchanged. Expressions account for \packageCodepercent of arguments in our
+corpus. More specifically, \packageSymbolpercent are \c{symbol}s (single
+variables such as \c{x}), \packageLanguagepercent are \c{language} object (a
+single expression, such as \c{x+1}), and \packageExpressionpercent are
 \c{expression} objects (multiple statements). Further inspection reveals that
-most symbols, \packageGgplotsymbolpercent to be exact, come from a single call in the
-\c{ggplot2} package and have have the value \c{\_inherit}. Even if we remove
-this outlier, the majority of \evals are simple expressions.
+most symbols, \packageGgplotsymbolpercent to be exact, come from a single call
+in the \c{ggplot2} package and have have the value \c{\_inherit}. Even if we
+remove this outlier, the majority of \evals are simple expressions.
 
 To estimate how much executable code is injected through \eval, we measure the
-the number of abstract syntax tree nodes as returned by the parser; for example,
-\c{x+1} counts as 3.\footnote{We attempted to measure string lengths after
-de-parsing expressions, but these measurements were dominated by de-parsed data
-objects, which could range in the MBs.} The median argument size is
-\packageMedianszeval (due to the many symbols) and the average is \packageAvgszeval nodes. The
-largest \eval input observed is \packageMaxszeval\,-- a significant chunk of code.
+the number of nodes in the expressions; for example, \c{x+1} counts as 3. We
+also measured string lengths of unparsed expressions, but these measurements
+were dominated by the size of data objects which could range in the MBs. The
+median argument size is \packageMedianszeval (due to the symbols) and the
+average is \packageAvgszeval nodes. The largest \eval input observed is
+\packageMaxszeval\,-- a significant chunk of code.
 Figure~\ref{fig:sizedistribution} shows the distribution of sizes for arguments
-of fewer or equal to 25 nodes.
-The x-axis is the size of arguments in number of
+of fewer or equal to 25 nodes. The x-axis is the size of arguments in number of
 nodes, and the y-axis is the count of arguments with that size. The size drops
 rapidly, with few observations larger than 15 nodes. The long tail is omitted
-from the graph for legibility.
-%
-\begin{wrapfigure}{r}{5.8cm}
- \hspace*{-10mm}
- \includegraphics[width=0.5\textwidth, trim=4.8cm 0 6cm 0, clip]{size_loaded_distribution}
- \caption{Loaded code} \label{fig:sizedistribution}
-\end{wrapfigure}
-%
-
-To estimate the work performed in \evals, we count the instructions executed by
+for legibility.
+
+To estimate the work performed in \evals, we count instructions executed by
 interpreter. Most invocations perform relatively little work with
-\packageSmalleventspct of \evals executing 50 or fewer instructions. The violin plot of
-Figure~\ref{ev}(a) corresponds to \evals executing $\leq$ 50 instructions, it is
-dominated by trivial symbol look ups. Figure~\ref{ev}(b) has the work-intensive
-\evals which go all the way to \packageMaxeventsRnd instructions.
+\packageSmalleventspct of \evals executing 50 or fewer instructions. The violin
+plot of Figure~\ref{ev}(a) corresponds to \evals executing $\leq$ 50
+instructions, it is dominated by trivial symbol look ups. Figure~\ref{ev}(b) has
+the work-intensive \evals which go all the way to \packageMaxeventsRnd
+instructions.
+
+\newpage
 
@@
 \end{tabular}
- \caption{Instructions per \eval} \label{ev}
+ \caption{Instructions per call} \label{ev}
 \end{figure}
 
-\paragraph{Base}
+\subsection{Base}
 We recorded \baseAllcalls \eval calls in the base libraries with fewer than 4\%
-of those being to any variants. Unlike CRAN, Base calls \c{evalq} (\baseEvalqs).
-Out of \baseStaticeval call sites, \baseTriggeredevalpct are exercised. Most
-arguments, \baseCodepercent, are expressions. However, unlike CRAN,
-\baseLanguagepercent of the arguments are language objects. The median argument
-size is \baseMedianszeval, which is more than for CRAN. The maximum size is
+of those being to any variants (this was done on a small subset of the whole
+corpus). Unlike CRAN, Base calls \c{evalq} (\baseEvalqs). Out of \baseStaticeval
+call sites, \baseTriggeredevalpct are exercised. Most arguments,
+\baseCodepercent, are expressions. However, unlike CRAN, \baseLanguagepercent of
+the arguments are language objects. The median argument size is
+\baseMedianszeval, which is more than for CRAN. The maximum size is
 \baseMaxszeval. Small instruction counts, less than 50, amount for
-\baseSmalleventspct of calls.
-\todo{More info about base and conclude it here.}
-
-\begin{figure}[tb] \centering
-  \includegraphics[width=.8\textwidth]{traced-eval-callsites.pdf} \centering
-  \caption{\eval call sites coverage of the \PkgPackages packages.}%
-  \label{fig:traced-eval-callsites}
-\end{figure}
-
-\paragraph{Kaggle}
+\baseSmalleventspct of calls. The challenge dealing with Base is that every
+program uses it and it can thus generate extreme amounts of data but the data is
+quite predictable. For the rest of the paper, we only mention base when there
+are surprising observations.
+
+\medskip
+
+\subsection{Kaggle}
 Only \kaggleNbruns scripts trigger \eval. In total, \kaggleAllcalls \eval were
@@
 \kaggleMaxszeval. The distribution of instructions per eval is similar to CRAN;
-\kaggleSmalleventspct of \evals execute fewer than 50 instructions.
-\todo{More info about kaggle and conclude it here.}
-
-\paragraph{Discussion}
+\kaggleSmalleventspct of \evals execute fewer than 50 instructions. Manual
+inspection of \eval usage in Kaggle suggest that it is consistent with the data
+obtained for CRAN. We do not discuss it further.
+
+\medskip
+
+\subsection{Discussion}
 These results show that \eval is central to the language implementation as it is
-omnipresent in the relatively small Base library. Packages, which are developed
-by experienced programmers, make regular and varied use of the facility.
-Finally, scripts are written by less sophisticated users, and likely perform
-simpler tasks, and thus have lesser need for \eval, and its uses originate from
-strings which are easier to manipulate for end-users.
+omnipresent in the relatively small Base library. The Base library is, in turn,
+used by every single package. To avoid having Base dominate the result we have
+excluded calls occurring there from the data we report when analyzing packages.
+CRAN Packages, which are typically developed by experienced programmers, make
+regular and varied use of \eval. They represent our most interesting data set as
+these packages are the result of over 20 years of contributions by thousands of
+authors. The code in packages is well maintained and relatively well tested.
+Finally, Kaggle scripts are often written by less sophisticated users, and
+likely perform simpler tasks, and thus have lesser need for \eval, and its uses
+originate from strings as these are easier to manipulate for end-users. Due to
+the relative paucity of data in Kaggle, we do not pursue the data set further.
 
@@
 
-\paragraph{Base}
-\todo{I would remove this - we conclude Base in 4}
-The spread of forms in Base are similar to CRAN with a larger share of \c{V}, as
-it is present in \baseMinimizedpropsitesa\% of the sites and \c{X} is a distant
-second with \baseMinimizedpropsitesb\%.
-
-\paragraph{Kaggle}
-\todo{I would remove this - we conclude Kaggle in 4}
-Kaggle is dominated by \c{V}, with \kaggleMinimizedpropsitesa\% of the
-\kaggleNbCallSites sites.
-
 \todo{The following should be either removed or positioned properly}

### 2021-04-09 paper/main.tex
tax

@@
 
-\begin{itemize}
-\item \underline{$min(e)=\c{V}$}: Expression that represent values are
+\newcommand{\EE}[1]{{{\emph{\framebox{#1}}}}}
+
+\medskip\noindent\EE{$min(e)=\c{V}$} Expression that represent values are
 frequently passed to \eval. The majority of those, \packageValOneNodePercent,
 are inline constants (integers or doubles). The rest trivially evaluate to a
-  value.\footnote{True as long as base functions such as \c{+} are not
-  redefined. They typically aren't, but this is a limitation of this
-  categorization.} In our corpus, \packageNbCallSitesUniqueActualValue call
-  sites only ever see a simple values; the median number of times they are
-  exercised is \packageMedianRunSitesUniqueActualValue, suggesting a possible
-  lack of coverage. However, manual inspection reveals cases such as:
+value.\footnote{True as long as base functions such as \c{+} are not redefined.
+They typically aren't, but this is a limitation of this categorization.} In our
+corpus, \packageNbCallSitesUniqueActualValue call sites only ever see a simple
+values; the median number of times they are exercised is
+\packageMedianRunSitesUniqueActualValue, suggesting a possible lack of coverage.
+However, manual inspection reveals cases such as:
 \begin{lstlisting}
@@
   them needed to use \eval.
-\item \underline{$min(e)=\c{X}$}: Variables are the most common form. They
+
+\medskip\noindent\EE{$min(e)=\c{X}$} Variables are the most common form. They
 subsume values, thus \c{X} includes arithmetic expressions in which a variable
 occurs. \packageNbSymbolVarSitePercent of variable expression are symbols, \ie
-  just an \c{x}. As a variable may refer to a promise, it may trigger
-  computation.
-\item \underline{$min(e)=\c{\$}$}: A name lookup, \c{x\$f}, can be used to
-  access variable in an environment or a column in a data frame, this also
-  include vector   indexing operators, \c{[]} and \c{[[]]}. A name lookup
-  subsumes variables.
-\item \underline{$min(e)=$~\c{<-}}: Assignments account for
+just an \c{x}. As a variable may refer to a promise, it may trigger computation.
+
+\medskip\noindent\EE{$min(e)=\c{\$}$} A name lookup, \c{x\$f}, can be used to
+access variable in an environment or a column in a data frame, this also include
+vector indexing operators, \c{[]} and \c{[[]]}. A name lookup subsumes
+variables.
+
+\medskip\noindent\EE{$min(e)=$~\c{<-}} Assignments account for
 \packageAssignSitesPercent of \eval sites, they are \c{<-}, {\tt <<-} and
 \c{assign}, and they subsume named lookups.
-\item \underline{$min(e)=\c{F()}$}: A plain function call which may have some
+
+\medskip\noindent\EE{$min(e)=\c{F()}$} A plain function call which may have some
 values as argument but does not include variables. It should be noted that
@@
 do not trigger computation.
-\item \underline{$min(e)=\c{F(X)}$}: Function calls whose arguments may include
+
+\medskip\noindent\EE{$min(e)=\c{F(X)}$} Function calls whose arguments may include
 variable references.
-\item \underline{$min(e)=\c{F(F(X))}$}: Nested function calls with variables.
-\item \underline{$min(e)=\c{FUN}$}: Function definition accounts for
+
+\medskip\noindent\EE{$min(e)=\c{F(F(X))}$} Nested function calls with variables.
+
+\medskip\noindent\EE{$min(e)=\c{FUN}$} Function definition accounts for
   \packageFunctionDefinitionSitesPercent of \eval sites. In addition the
@@
   estimate of the use of higher-order functions.
-\item \underline{$min(e)=\c{\{\,\}}$}: Arbitrary-length statement blocks. These
-  are larger expressions with, on average, \packageMinimizedoperationsjRnd
+
+\medskip\noindent\EE{$min(e)=\c{\{\,\}}$} Arbitrary-length statement blocks.
+These are larger expressions with, on average, \packageMinimizedoperationsjRnd
 instructions. The contents of the block are not inspected.
-\item \underline{$min(e)=\c{model.frame}$}: Call to the \c{model.frame} function
-  used to create statistical models. This subsume all other categories.
-
-\end{itemize}
+
+\medskip\noindent\EE{$min(e)=\c{model.frame}$} Call to the \c{model.frame}
+function used to create statistical models. This subsume all other categories.
 

### 2021-04-09 paper/main.tex
tax

@@
 
-\subsection{Arguments of \eval}
-
-The argument passed to a call to \eval determines what the function will do. As
-we have seen in the previous section, there are cases where the argument is a
-value, but these are rare. The common cases are that the argument is some form
-of expression, either a single variable, a simple expression or a statement.
-
-To categorize the expressions, consider a function $min(e)$ which for a given
-expression $e$ returns a normal form that abstracts many incidental details
-allowing the reader to focus on the structure of the evaluated code. The
-minimization function performs several simplifications steps. It performs
-constant folding of arithmetic and string expressions for base operators, e.g.
+\subsection{The arguments of \eval}
+
+The expressions passed to \eval vary widely. In order to categorize them, let us
+use a minimization function $min(e)$ which for a given expression $e$ returns a
+normal form that abstracts incidental details allowing the reader to focus on
+the structure of the evaluated code. The minimization function performs constant
+folding of arithmetic and string expressions for base operators, e.g.
 $min(\c{1+1})=\c{V}$, value simplification, $min(\c{c(1,2,3+2)})=\c{V}$,
 variable absorption, $min(\c{x+y})=\c{X}$, function absorption,
-$min(\c{g(f(x),h(z))})=\c{F(F(X))}$ and a number of other simplification. There
-are 15 normal forms. We detail the interesting cases next.
-
-\newcommand{\EE}[1]{{{\emph{\framebox{#1}}}}}
+$min(\c{g(f(x),h(z))})=\c{F(F(X))}$ and a number of other simplifications.
+Table~\ref{tab:minimizedexpressions} gives the 10 most frequent forms; \#sites
+and \%sites are, respectively, the number and ratio of sites receiving arguments
+of that form; \#packages is the number of packages with that form; \#operations
+is the average number of instructions performed by the interpreter; and \%envir
+is the percentage of sites for which the \c{envir} argument is explicitly passed
+to \eval and is not \c{parent.frame()}.
+
+\begin{table}[h]\small
+\begin{tabular}{|c|r|r|r|r|r|c|}\hline
+  $min(e)$& \#sites & \%sites & \#packages & \#operations & \%envir & example\\\hline
+%% JAN: simplifying osme mumers and adding %
+\c X&\packageMinimizedcallsitesa & 28\% &\packageMinimizedpackagea & 768 &77\% & \c{y+1}\\\hline
+\c{F(F(X))} & \packageMinimizedcallsitesb  & 20\% & \packageMinimizedpackageb  & 466K & 55\% & \c{gbov( mean(x), a-1)}\\\hline
+\c{V}&\packageMinimizedcallsitesc & 17\% &\packageMinimizedpackagec &\packageMinimizedoperationscRnd & 71\%& \c{c(42,21,0)}\\\hline
+\c{F(X)}& \packageMinimizedcallsitesd & 14\% & \packageMinimizedpackaged & \packageMinimizedoperationsdRnd & 75\% & \c{seq\_len(iters)} \\\hline
+\c{model.frame} & \packageMinimizedcallsitese & 7\% & \packageMinimizedpackagee & 3K & 21\% & \c{model.frame(formula = Z ~ U)} \\\hline
+\c{F()}& \packageMinimizedcallsitesf & 6\% & \packageMinimizedpackagef & \packageMinimizedoperationsfRnd & 73\% & \c{rgamma(3, 2, n = 10L)} \\\hline
+\c{\$}& \packageMinimizedcallsitesg & 6\% & \packageMinimizedpackageg & 2K & 78\% & \c{DF\$B} \\\hline
+\c{FUN} & \packageMinimizedcallsitesh & 6\% & \packageMinimizedpackageh & \packageMinimizedoperationshRnd & 23\% & \c{function(x, y) x + 3 * y} \\\hline
+\c{<-} & \packageMinimizedcallsitesi  & 5\% & \packageMinimizedpackagei & 11K & 73\% & \c{x[1, 2:3, 2:3] <- value}\\\hline
+\c{BLOCK} & \packageMinimizedcallsitesj & 4\% & \packageMinimizedpackagej & 440K & 73\% & \c{\{ write.csv(iris, tf) \textbackslash n file.size(tf) \}} \\\hline
+%% \c X&\packageMinimizedcallsitesa &\packageMinimizedpropsitesa\% &\packageMinimizedpackagea &\packageMinimizedoperationsaRnd &\packageMinimizedpercentenvirnonparenta\% & \c{y+1}\\\hline
+%% \c{F(F(X))} & \packageMinimizedcallsitesb  & \packageMinimizedpropsitesb & \packageMinimizedpackageb  & \packageMinimizedoperationsbRnd & \packageMinimizedpercentenvirnonparentb & \c{gbov( mean(x), a-1)}\\\hline
+%% \c{V}&\packageMinimizedcallsitesc &\packageMinimizedpropsitesc &\packageMinimizedpackagec &\packageMinimizedoperationscRnd &\packageMinimizedpercentenvirnonparentc& \c{c(42,21,0)}\\\hline
+%% \c{F(X)}& \packageMinimizedcallsitesd & \packageMinimizedpropsitesd & \packageMinimizedpackaged & \packageMinimizedoperationsdRnd & \packageMinimizedpercentenvirnonparentd & \c{seq\_len(iters)} \\\hline
+%% \c{model.frame} & \packageMinimizedcallsitese & \packageMinimizedpropsitese & \packageMinimizedpackagee & \packageMinimizedoperationseRnd & \packageMinimizedpercentenvirnonparente & \c{model.frame(formula = Z ~ U)} \\\hline
+%% \c{F()}& \packageMinimizedcallsitesf & \packageMinimizedpropsitesf & \packageMinimizedpackagef & \packageMinimizedoperationsfRnd & \packageMinimizedpercentenvirnonparentf & \c{rgamma(3, 2, n = 10L)} \\\hline
+%% \c{\$}& \packageMinimizedcallsitesg & \packageMinimizedpropsitesg & \packageMinimizedpackageg & \packageMinimizedoperationsgRnd & \packageMinimizedpercentenvirnonparentg & \c{DF\$B} \\\hline
+%% \c{FUN} & \packageMinimizedcallsitesh & \packageMinimizedpropsitesh & \packageMinimizedpackageh & \packageMinimizedoperationshRnd & \packageMinimizedpercentenvirnonparenth & \c{function(x, y) x + 3 * y} \\\hline
+%% \c{<-} & \packageMinimizedcallsitesi  & \packageMinimizedpropsitesi & \packageMinimizedpackagei & \packageMinimizedoperationsiRnd & \packageMinimizedpercentenvirnonparenti & \c{x[1, 2:3, 2:3] <- value}\\\hline
+%% \c{\{\}} & \packageMinimizedcallsitesj & \packageMinimizedpropsitesj & \packageMinimizedpackagej & \packageMinimizedoperationsjRnd & \packageMinimizedpercentenvirnonparentj & \c{\{ write.csv(iris, tf) \textbackslash n file.size(tf) \}} \\\hline
+\end{tabular}
+\caption{Minimized expressions} \label{tab:minimizedexpressions}
+\end{table}
+
+15 normal forms. We detail the interesting cases next.
+
+\newcommand{\EE}[1]{{{\emph{\framebox{#1}}}}\\[1mm]}
 
@@
 function used to create statistical models. This subsume all other categories.
-
-Table~\ref{tab:minimizedexpressions} gives the most frequent forms; \#sites
-(\%sites) is the number (ratio) of call sites receiving arguments of that form;
-\#packages is the number of packages with that form; \#operations is the average
-number of instructions performed by the interpreter when given that minimized
-expression; and \%envir is the percentage of call sites for which the \c{envir}
-argument is explicitly passed, and is not \c{parent.frame()}.
-
-\begin{table}[h]\small
-\begin{tabular}{|c|r|r|r|r|r|c|}\hline
-$min(e)$& \#sites & \%sites & \#packages & \#operations & \%envir & example\\\hline
-\c X&\packageMinimizedcallsitesa &\packageMinimizedpropsitesa &\packageMinimizedpackagea &\packageMinimizedoperationsaRnd &\packageMinimizedpercentenvirnonparenta & \c{y+1}\\\hline
-\c{F(F(X))} & \packageMinimizedcallsitesb  & \packageMinimizedpropsitesb & \packageMinimizedpackageb  & \packageMinimizedoperationsbRnd & \packageMinimizedpercentenvirnonparentb & \c{gbov( mean(x), a-1)}\\\hline
-\c{V}&\packageMinimizedcallsitesc &\packageMinimizedpropsitesc &\packageMinimizedpackagec &\packageMinimizedoperationscRnd &\packageMinimizedpercentenvirnonparentc& \c{c(42,21,0)}\\\hline
-\c{F(X)}& \packageMinimizedcallsitesd & \packageMinimizedpropsitesd & \packageMinimizedpackaged & \packageMinimizedoperationsdRnd & \packageMinimizedpercentenvirnonparentd & \c{seq\_len(iters)} \\\hline
-\c{model.frame} & \packageMinimizedcallsitese & \packageMinimizedpropsitese & \packageMinimizedpackagee & \packageMinimizedoperationseRnd & \packageMinimizedpercentenvirnonparente & \c{model.frame(formula = Z ~ U)} \\\hline
-\c{F()}& \packageMinimizedcallsitesf & \packageMinimizedpropsitesf & \packageMinimizedpackagef & \packageMinimizedoperationsfRnd & \packageMinimizedpercentenvirnonparentf & \c{rgamma(3, 2, n = 10L)} \\\hline
-\c{\$}& \packageMinimizedcallsitesg & \packageMinimizedpropsitesg & \packageMinimizedpackageg & \packageMinimizedoperationsgRnd & \packageMinimizedpercentenvirnonparentg & \c{DF\$B} \\\hline
-\c{FUN} & \packageMinimizedcallsitesh & \packageMinimizedpropsitesh & \packageMinimizedpackageh & \packageMinimizedoperationshRnd & \packageMinimizedpercentenvirnonparenth & \c{function(x, y) x + 3 * y} \\\hline
-\c{<-} & \packageMinimizedcallsitesi  & \packageMinimizedpropsitesi & \packageMinimizedpackagei & \packageMinimizedoperationsiRnd & \packageMinimizedpercentenvirnonparenti & \c{x[1, 2:3, 2:3] <- value}\\\hline
-\c{\{\}} & \packageMinimizedcallsitesj & \packageMinimizedpropsitesj & \packageMinimizedpackagej & \packageMinimizedoperationsjRnd & \packageMinimizedpercentenvirnonparentj & \c{\{ write.csv(iris, tf) \textbackslash n file.size(tf) \}} \\\hline
-\c{F(F(<-))} & \packageMinimizedcallsitesk & \packageMinimizedpropsitesk & \packageMinimizedpackagek & \packageMinimizedoperationskRnd & \packageMinimizedpercentenvirnonparentk & \c{opt\$h.weights <- rep(1, length(x))} \\\hline
-\end{tabular}
-\caption{Minimized expressions} \label{tab:minimizedexpressions}
-\end{table}
 

### 2021-04-09 paper/main.tex
tax

@@
 is the percentage of sites for which the \c{envir} argument is explicitly passed
-to \eval and is not \c{parent.frame()}.
+to \eval and is not \c{parent.frame()}. The example column shows one sample
+expression $e$ that is normalized to the particular form.
 
@@
 \c{<-} & \packageMinimizedcallsitesi  & 5\% & \packageMinimizedpackagei & 11K & 73\% & \c{x[1, 2:3, 2:3] <- value}\\\hline
-\c{BLOCK} & \packageMinimizedcallsitesj & 4\% & \packageMinimizedpackagej & 440K & 73\% & \c{\{ write.csv(iris, tf) \textbackslash n file.size(tf) \}} \\\hline
+\c{BLOCK} & \packageMinimizedcallsitesj & 4\% & \packageMinimizedpackagej & 440K & 73\% & \c{\{ write.csv(iris, tf); file.size(tf) \}} \\\hline
 %% \c X&\packageMinimizedcallsitesa &\packageMinimizedpropsitesa\% &\packageMinimizedpackagea &\packageMinimizedoperationsaRnd &\packageMinimizedpercentenvirnonparenta\% & \c{y+1}\\\hline
@@
 
-15 normal forms. We detail the interesting cases next.
+\noindent
+We detail these forms and discuss their implication for the behavior of \eval.
 
@@
 \medskip\noindent\EE{$min(e)=\c{V}$} Expression that represent values are
-frequently passed to \eval. The majority of those, \packageValOneNodePercent,
-are inline constants (integers or doubles). The rest trivially evaluate to a
-value.\footnote{True as long as base functions such as \c{+} are not redefined.
-They typically aren't, but this is a limitation of this categorization.} In our
-corpus, \packageNbCallSitesUniqueActualValue call sites only ever see a simple
-values; the median number of times they are exercised is
-\packageMedianRunSitesUniqueActualValue, suggesting a possible lack of coverage.
-However, manual inspection reveals cases such as:
+frequently passed to \eval -- they represent 17\% of the calls. The majority of
+those, \packageValOneNodePercent, are inline constants (integers or doubles).
+The rest trivially evaluate to a value, \eg~\c{1+1}.\footnote{True as long as
+base functions such as \c{+} are not redefined. They typically aren't, but this
+is a limitation of this categorization.} In our corpus,
+\packageNbCallSitesUniqueActualValue call sites only ever see a simple value.
+Manual inspection reveals cases such as the following
 \begin{lstlisting}
@@
 \end{lstlisting}
-  This concatenates to a string and \eval simply returns it as is.
-  From twenty randomly selected \evals, only \packageUsefulValueEvalPercent of
-  them needed to use \eval.
-
-\medskip\noindent\EE{$min(e)=\c{X}$} Variables are the most common form. They
-subsume values, thus \c{X} includes arithmetic expressions in which a variable
-occurs. \packageNbSymbolVarSitePercent of variable expression are symbols, \ie
-just an \c{x}. As a variable may refer to a promise, it may trigger computation.
-
-\medskip\noindent\EE{$min(e)=\c{\$}$} A name lookup, \c{x\$f}, can be used to
-access variable in an environment or a column in a data frame, this also include
-vector indexing operators, \c{[]} and \c{[[]]}. A name lookup subsumes
-variables.
-
-\medskip\noindent\EE{$min(e)=$~\c{<-}} Assignments account for
-\packageAssignSitesPercent of \eval sites, they are \c{<-}, {\tt <<-} and
-\c{assign}, and they subsume named lookups.
-
-\medskip\noindent\EE{$min(e)=\c{F()}$} A plain function call which may have some
-values as argument but does not include variables. It should be noted that
-function name require variable lookups -- we distinguish those as they usually
-do not trigger computation.
-
-\medskip\noindent\EE{$min(e)=\c{F(X)}$} Function calls whose arguments may include
-variable references.
+where the argument to \eval is a string which \eval simply returns. From twenty
+randomly selected \evals, \packageUsefulValueEvalPercent of them do not need
+\eval. For the other cases, either we are dealing with a value that needs to be
+constructed dynamically or the value is a default case that sometimes is
+replaced by a more interesting expression.
+
+\medskip\noindent\EE{$min(e)=\c{X}$} Variables lookups are the most common form,
+they are found in 28\% of the sites. This form includes simple variable reads,
+\eg~\c{x}, those are \packageNbSymbolVarSitePercent of \c X. The form also
+subsumes \c{V}, so it includes a mixture of arithmetic expressions,
+\eg~\c{x+y+1}. The operations allowed are limited to built-in arithmetics. It is
+noteworthy that, while most \c{X}\!s evaluate in a single step, the variable can
+be bound to a promise, and accessing it may trigger evaluation of that promise,
+thus resulting in an arbitrary amount of computation.
+
+\newpage
+\medskip\noindent\EE{$min(e)=\c{\$}$} This form extends \c X to include lookup
+with the dollar operator, \eg~\c{x\$f}, and vector indexing, \eg~\c{x[42]} or
+\c{x[[24]]}. As with \c X we allow arithmetics and values in this form. Lookup
+occurs in 6\% of the sites. Arbitrary computations can occur if one of the
+variables is bound to a promise.
+
+\medskip\noindent\EE{$min(e)=$~\c{<-}} This form includes both assignments
+operator, the direct assignment {\tt <-}, and assignment to the parent
+environment {\tt <\,\!<-}, and the \c{\$} form. Assignments occur in 4\% of the
+sites. They represent the most obvious source of side-effects.
+
+\medskip\noindent\EE{$min(e)=\c{F()}$} This form captures simple function calls,
+\eg~\c{f(2)}, with neither variables or assignments. Or, more specifically it
+allows for variables in the function position but not in arguments. Usually,
+looking up function names does not trigger computation, but that is not a given
+if the function is returned by a promise or if the function name is shadowed by
+a promise containing a value, then computation will occur. This form occurs in
+6\% of sites.
+
+\medskip\noindent\EE{$min(e)=\c{F(X)}$} This form allows function calls whose
+arguments may include variable references. Assignments are excluded from this
+form. It occurs in 14\% of the sites in the corpus.
 

### 2021-04-09 analysis/inc/latextags.R
Update normalized expressions

@@ -410,10 +410,13 @@ ggsave_prefix <- function(name, tags=get_default_tags()) {
   ggsave(path(PLOT_DIR, paste0(prefix, "_", name)))
 }
 
-r_vec <- function(name, nrows, v) {
+r_vec <- function(name, nrows, v, ...) {
   stopifnot(length(v) >= nrows)
   startLetter <- utf8ToInt("A") - 1 # LateX does not like numbers so we suffix with letters of the alphabet
+  attrs <- attributes(v) # propagate attributes to each of the elements in the vector
   for(i in 1:nrows) {
-    r(paste0(name, intToUtf8(startLetter + i)), v[[i]])
+    elem <- v[[i]]
+    attributes(elem) <- attrs
+    r(paste0(name, intToUtf8(startLetter + i)), elem, ...)
   }
 }

### 2021-04-09 paper/main.tex
tax

@@
 
-\medskip\noindent\EE{$min(e)=\c{F(X)}$} This form allows function calls whose
-arguments may include variable references. Assignments are excluded from this
-form. It occurs in 14\% of the sites in the corpus.
-
-\medskip\noindent\EE{$min(e)=\c{F(F(X))}$} Nested function calls with variables.
-
-\medskip\noindent\EE{$min(e)=\c{FUN}$} Function definition accounts for
-  \packageFunctionDefinitionSitesPercent of \eval sites. In addition the
-  functions inside other expressions account for
-  \packageGeneralizedFunctionDefinitionSitesPercent of sites, and give an
-  estimate of the use of higher-order functions.
-
-\medskip\noindent\EE{$min(e)=\c{\{\,\}}$} Arbitrary-length statement blocks.
-These are larger expressions with, on average, \packageMinimizedoperationsjRnd
-instructions. The contents of the block are not inspected.
-
-\medskip\noindent\EE{$min(e)=\c{model.frame}$} Call to the \c{model.frame}
-function used to create statistical models. This subsume all other categories.
-
-A given \eval call site can be passed different forms at different invocations.
-However, \packageNbOneMinimizedPercent of sites only see a single form.
-
-\todo{The following should be either removed or positioned properly}
-We look at how \emph{consistent} the \c{expr} argument of \eval can be, \ie how
-many different types of the resolved \c{expr} there are per call sites. Most of
-the call sites, \ie \PercentMonomorphic, are \emph{consistent}, and this is
-similar to javascript. However, a few ones are highly \emph{polymorphic} (10
-different types). They are the pipe operators \c{\%>\%}, \c{\%<>\%} and
-\c{\%\$\%} in package \emph{magrittr}. It is effectively used to compose
-functions on their first argument, which can be of any type.
+\medskip\noindent\EE{$min(e)=\c{F(X)}$} This form allows for a simple function
+call whose arguments may include variable references. Assignments are excluded
+from this form. It occurs in 14\% of the sites in the corpus.
+
+\medskip\noindent\EE{$min(e)=\c{F(F(X))}$} This form extends the previous one
+with nested function calls. It is the second most common argument to \eval,
+occurring in 20\% of the sites.
+
+\medskip\noindent\EE{$min(e)=\c{FUN}$} This form captures expressions that are
+function definitions, \eg~\c{function(x)x+1}. \c{FUN} occurs in
+\packageFunctionDefinitionSitesPercent of sites. One data point we have not
+recorded is the side effects that are performed when the generated functions are
+eventually run. In addition to \c{FUN} there
+\packageGeneralizedFunctionDefinitionSitesPercent of sites create functions
+nested in other expressions, this gives an idea of the use of higher-order
+functions.
+
+\medskip\noindent\EE{$min(e)=\c{BLOCK}$} This form captures blocks of code with
+multiple statements. These are larger expressions, we do not inspect contents of
+the blocks. They occur in 4\% of the sites.
+
+\medskip\noindent\EE{$min(e)=\c{model.frame}$} This is a very specific form
+intended to capture the construction of calls to the \c{model.frame} function.
+This function creates statistical models. This subsumes \c{F(F(X))}, \c{FUN} and
+assignments. This particular form occurs in 7\% of the sites.
+
+\paragraph{Consistency} It is interesting to consider how many different
+forms any given site sees. The more forms, the harder it will be to characterize
+the behavior of the program at that site. Luckily, \packageNbOneMinimizedPercent
+of sites only see a single form. There are few sites that are highly polymorphic
+(8 or more different forms), these include the pipe operator of the \c{magrittr}
+package which is used to compose functions.
+
+\paragraph{Discussion} The variety of uses of \eval is evidenced by the number
+of different forms observed in CRAN. In comparison, JavaScript \eval usage was
+simpler and more predictable as reported by \citet{oopsla12b}. Nevertheless,
+simple forms dominate and there are many cases where \eval could be replaced
+by less powerful constructs.
+
+\newpage
 

### 2021-04-09 analysis/environments.Rmd

@@
 
-eval_calls_explicit <- E_envir %>% filter(is.na(envir))
+eval_calls_explicit <- E_envir %>% filter(!is.na(envir))
 nb_explicit <- count(eval_calls_explicit, wt = duplicates)
@@
 ```{r envir_absolute}
-eval_calls_absolute <- eval_calls_explicit %>%
-  anti_join(eval_calls_env_list) #%>%
-  #anti_join(eval_calls_arg_caller)
+eval_calls_absolute <- eval_calls_explicit %>% filter(envir_type == "environment")
 
@@
 nb_global_env_sites <- specific_env %>% filter(specific_env == "global") %>% .$n
-nb_callee_env_sites <- specific_env %>% filter(specific_env == "callee") %>% .$n
+nb_callee_env_sites <- specific_env %>% filter(specific_env == "callee") %>% .$n # Seems it is actually for envir == "new.env()"
 if(length(nb_callee_env_sites) == 0) nb_callee_env_sites <- 0
@@
 env <- attach(NULL, pos, basename(directory))
-```
-
-# Better-looking env class
-
-`caller-n` wins over any specific env. We add `+` if there are any `new`.
-
-```{r}
-env_calls <- env_calls %>%
-  mutate(env_class = if_else(environment_class == "loop", "loop", paste0(if_else(is.na(hierarchy), specific_env, as.character(hierarchy)), if_else(is.na(new_env), "", "+"))))
 ```
@@
 
+# Better-looking env class
+
+`caller-n` wins over any specific env. We add `+` if there are any `new`.
+
+TODO: if there is both new , a caller-n, a specific class, rather ignore the caller-n
+```{r}
+env_classes <- env_calls %>%
+  mutate(env_class = if_else(environment_class == "loop", "loop", paste0(if_else(is.na(hierarchy), specific_env, as.character(hierarchy)), if_else(is.na(new_env), "", "+"))))
+
+env_classes %>% count(env_class, wt = duplicates, sort = TRUE) %>%
+  mutate(pc = round(100 * n / sum(n), 2), cpc = cumsum(pc)) %>%
+  datatable()
+
+env_classes %>% group_by(env_class) %>% summarise(n = n_distinct(src_ref)) %>%
+  arrange(desc(n)) %>%
+  mutate(pc = round(100 * n / sum(n), 2), cpc = cumsum(pc)) %>%
+  datatable()
+```
+
+# Specific envir with their environment class
+
+For the default argument:
+
+```{r}
+eval_calls_default %>% count(environment_class, wt = duplicates, sort = TRUE) %>%
+  mutate(pc = round(100 * n / sum(n), 2), cpc = cumsum(pc))
+
+eval_calls_default %>% group_by(environment_class) %>% summarise(n = n_distinct(src_ref)) %>%
+  arrange(desc(n)) %>%
+   mutate(pc = round(100 * n / sum(n), 2), cpc = cumsum(pc))
+
+# caller-2- is eval.parent
+eval_calls_default %>% filter(environment_class == "caller-2-") %>% count(ev_variant)
+
+```
+
+`eval` is not problematic:
+
+```{r}
+eval_calls_default %>% filter(ev_variant == "eval") %>% count(environment_class, wt = duplicates, sort = TRUE)
+```
+
+But `eval.parent` is:
+
+```{r}
+eval_calls_default %>% filter(ev_variant == "eval.parent") %>% count(environment_class, wt = duplicates, sort = TRUE)
+```
+
+It might be because `parent.frame` is equivalent to  `sys.frame(sys.parent(n))` and `sys.parent(n)` can get surprising result of it is called near the top level...
+
+For `parent.frame()` :
+
+```{r}
+E_envir %>% filter(envir == "parent.frame()") %>% count(environment_class, wt = duplicates, sort = TRUE) %>%
+  mutate(pc = round(100 * n / sum(n), 2), cpc = cumsum(pc))
+
+E_envir %>% filter(envir == "parent.frame()") %>% group_by(environment_class) %>%
+  summarise(n = n_distinct(src_ref)) %>%
+  arrange(desc(n)) %>%
+   mutate(pc = round(100 * n / sum(n), 2), cpc = cumsum(pc))
+
+```
+
+For `new.env()`:
+
+```{r}
+E_envir %>% filter(envir == "new.env()") %>% count(environment_class, wt = duplicates, sort = TRUE) %>%
+  mutate(pc = round(100 * n / sum(n), 2), cpc = cumsum(pc))
+
+E_envir %>% filter(envir == "new.env()") %>% group_by(environment_class) %>%
+  summarise(n = n_distinct(src_ref)) %>%
+  arrange(desc(n)) %>%
+   mutate(pc = round(100 * n / sum(n), 2), cpc = cumsum(pc))
+
+```
+
+For each envir expression, what is the most common environment class? (in calls, in sites)
+
+```{r}
+env_class_envir <- E_envir %>% group_by(envir) %>%
+  group_by(environment_class, .add = TRUE) %>% summarise(n = n_distinct(src_ref)) %>%
+  summarise(env_class = first(environment_class, order_by = order(n, decreasing = TRUE)), n = sum(n)) %>% 
+  arrange(desc(n))
+```
+
 ```{r}

### 2021-04-09 analysis/normalized.Rmd

@@
             n_packages = n_distinct(ev_package),
-            n_operations = round(mean(n_op, na.rm=TRUE), 2), 
+            n_operations = round(mean(n_op, na.rm=TRUE)), 
             med_operations = median(n_op, na.rm = TRUE),
@@
 r_vec("MinimizedCallSites", nrows, pull(widespread_minimized, n_call_sites))
-r_vec("MinimizedPropSites", nrows, pull(widespread_minimized, proportion))
+r_vec("MinimizedPropSites", nrows, pull(widespread_minimized, proportion), digits=0)
 r_vec("MinimizedPackage", nrows, pull(widespread_minimized, n_packages))
 r_vec("MinimizedOperations", nrows, pull(widespread_minimized, n_operations))
-r_vec("MinimizedPercentEnvir", nrows, pull(widespread_minimized, percent_envir))
-r_vec("MinimizedPercentEnvirNonParent", nrows, pull(widespread_minimized, percent_envir_non_parentframe))
+r_vec("MinimizedMedianOperations", nrows, pull(widespread_minimized, med_operations))
+r_vec("MinimizedPercentEnvir", nrows, pull(widespread_minimized, percent_envir), digits=0)
+r_vec("MinimizedPercentEnvirNonParent", nrows, pull(widespread_minimized, percent_envir_non_parentframe), digits=0)
 ```

### 2021-04-09 analysis/normalized.Rmd
Minimized expressions

@@
 r_vec("MinimizedPercentEnvir", nrows, pull(widespread_minimized, percent_envir), digits=0)
+r_vec("MinimizedPercentCallsEnvir", nrows, pull(widespread_minimized, percent_envir_calls), digits=0)
 r_vec("MinimizedPercentEnvirNonParent", nrows, pull(widespread_minimized, percent_envir_non_parentframe), digits=0)

### 2021-04-09 paper/main.tex
Minimized expressions

--- 2021-04-09 paper/main.tex
@@
   $min(e)$& \#sites & \%sites & \#packages & \#operations & \%envir & example\\\hline
-%% JAN: simplifying osme mumers and adding %
-\c X&\packageMinimizedcallsitesa & 28\% &\packageMinimizedpackagea & 768 &77\% & \c{y+1}\\\hline
-\c{F(F(X))} & \packageMinimizedcallsitesb  & 20\% & \packageMinimizedpackageb  & 466K & 55\% & \c{gbov( mean(x), a-1)}\\\hline
-\c{V}&\packageMinimizedcallsitesc & 17\% &\packageMinimizedpackagec &\packageMinimizedoperationscRnd & 71\%& \c{c(42,21,0)}\\\hline
-\c{F(X)}& \packageMinimizedcallsitesd & 14\% & \packageMinimizedpackaged & \packageMinimizedoperationsdRnd & 75\% & \c{seq\_len(iters)} \\\hline
-\c{model.frame} & \packageMinimizedcallsitese & 7\% & \packageMinimizedpackagee & 3K & 21\% & \c{model.frame(formula = Z ~ U)} \\\hline
-\c{F()}& \packageMinimizedcallsitesf & 6\% & \packageMinimizedpackagef & \packageMinimizedoperationsfRnd & 73\% & \c{rgamma(3, 2, n = 10L)} \\\hline
-\c{\$}& \packageMinimizedcallsitesg & 6\% & \packageMinimizedpackageg & 2K & 78\% & \c{DF\$B} \\\hline
-\c{FUN} & \packageMinimizedcallsitesh & 6\% & \packageMinimizedpackageh & \packageMinimizedoperationshRnd & 23\% & \c{function(x, y) x + 3 * y} \\\hline
-\c{<-} & \packageMinimizedcallsitesi  & 5\% & \packageMinimizedpackagei & 11K & 73\% & \c{x[1, 2:3, 2:3] <- value}\\\hline
-\c{BLOCK} & \packageMinimizedcallsitesj & 4\% & \packageMinimizedpackagej & 440K & 73\% & \c{\{ write.csv(iris, tf); file.size(tf) \}} \\\hline
-%% \c X&\packageMinimizedcallsitesa &\packageMinimizedpropsitesa\% &\packageMinimizedpackagea &\packageMinimizedoperationsaRnd &\packageMinimizedpercentenvirnonparenta\% & \c{y+1}\\\hline
-%% \c{F(F(X))} & \packageMinimizedcallsitesb  & \packageMinimizedpropsitesb & \packageMinimizedpackageb  & \packageMinimizedoperationsbRnd & \packageMinimizedpercentenvirnonparentb & \c{gbov( mean(x), a-1)}\\\hline
-%% \c{V}&\packageMinimizedcallsitesc &\packageMinimizedpropsitesc &\packageMinimizedpackagec &\packageMinimizedoperationscRnd &\packageMinimizedpercentenvirnonparentc& \c{c(42,21,0)}\\\hline
-%% \c{F(X)}& \packageMinimizedcallsitesd & \packageMinimizedpropsitesd & \packageMinimizedpackaged & \packageMinimizedoperationsdRnd & \packageMinimizedpercentenvirnonparentd & \c{seq\_len(iters)} \\\hline
-%% \c{model.frame} & \packageMinimizedcallsitese & \packageMinimizedpropsitese & \packageMinimizedpackagee & \packageMinimizedoperationseRnd & \packageMinimizedpercentenvirnonparente & \c{model.frame(formula = Z ~ U)} \\\hline
-%% \c{F()}& \packageMinimizedcallsitesf & \packageMinimizedpropsitesf & \packageMinimizedpackagef & \packageMinimizedoperationsfRnd & \packageMinimizedpercentenvirnonparentf & \c{rgamma(3, 2, n = 10L)} \\\hline
-%% \c{\$}& \packageMinimizedcallsitesg & \packageMinimizedpropsitesg & \packageMinimizedpackageg & \packageMinimizedoperationsgRnd & \packageMinimizedpercentenvirnonparentg & \c{DF\$B} \\\hline
-%% \c{FUN} & \packageMinimizedcallsitesh & \packageMinimizedpropsitesh & \packageMinimizedpackageh & \packageMinimizedoperationshRnd & \packageMinimizedpercentenvirnonparenth & \c{function(x, y) x + 3 * y} \\\hline
-%% \c{<-} & \packageMinimizedcallsitesi  & \packageMinimizedpropsitesi & \packageMinimizedpackagei & \packageMinimizedoperationsiRnd & \packageMinimizedpercentenvirnonparenti & \c{x[1, 2:3, 2:3] <- value}\\\hline
-%% \c{\{\}} & \packageMinimizedcallsitesj & \packageMinimizedpropsitesj & \packageMinimizedpackagej & \packageMinimizedoperationsjRnd & \packageMinimizedpercentenvirnonparentj & \c{\{ write.csv(iris, tf) \textbackslash n file.size(tf) \}} \\\hline
+  % macros are properly rounded and show the percentage sign now
+\c X&\packageMinimizedcallsitesa &\packageMinimizedpropsitesa &\packageMinimizedpackagea &\packageMinimizedoperationsaRnd &\packageMinimizedpercentenvirnonparenta & \c{y+1}\\\hline
+\c{F(F(X))} & \packageMinimizedcallsitesb  & \packageMinimizedpropsitesb & \packageMinimizedpackageb  & \packageMinimizedoperationsbRnd & \packageMinimizedpercentenvirnonparentb & \c{gbov( mean(x), a-1)}\\\hline
+\c{V}&\packageMinimizedcallsitesc &\packageMinimizedpropsitesc &\packageMinimizedpackagec &\packageMinimizedoperationscRnd &\packageMinimizedpercentenvirnonparentc& \c{c(42,21,0)}\\\hline
+\c{F(X)}& \packageMinimizedcallsitesd & \packageMinimizedpropsitesd & \packageMinimizedpackaged & \packageMinimizedoperationsdRnd & \packageMinimizedpercentenvirnonparentd & \c{seq\_len(iters)} \\\hline
+\c{model.frame} & \packageMinimizedcallsitese & \packageMinimizedpropsitese & \packageMinimizedpackagee & \packageMinimizedoperationseRnd & \packageMinimizedpercentenvirnonparente & \c{model.frame(formula = Z ~ U)} \\\hline
+\c{F()}& \packageMinimizedcallsitesf & \packageMinimizedpropsitesf & \packageMinimizedpackagef & \packageMinimizedoperationsfRnd & \packageMinimizedpercentenvirnonparentf & \c{rgamma(3, 2, n = 10L)} \\\hline
+\c{\$}& \packageMinimizedcallsitesg & \packageMinimizedpropsitesg & \packageMinimizedpackageg & \packageMinimizedoperationsgRnd & \packageMinimizedpercentenvirnonparentg & \c{DF\$B} \\\hline
+\c{FUN} & \packageMinimizedcallsitesh & \packageMinimizedpropsitesh & \packageMinimizedpackageh & \packageMinimizedoperationshRnd & \packageMinimizedpercentenvirnonparenth & \c{function(x, y) x + 3 * y} \\\hline
+\c{<-} & \packageMinimizedcallsitesi  & \packageMinimizedpropsitesi & \packageMinimizedpackagei & \packageMinimizedoperationsiRnd & \packageMinimizedpercentenvirnonparenti & \c{x[1, 2:3, 2:3] <- value}\\\hline
+\c{\{\}} & \packageMinimizedcallsitesj & \packageMinimizedpropsitesj & \packageMinimizedpackagej & \packageMinimizedoperationsjRnd & \packageMinimizedpercentenvirnonparentj & \c{\{ write.csv(iris, tf) ; file.size(tf) \}} \\\hline
 \end{tabular}
@@
 \end{table}
+% Other macros that can be used:
+% - \packageMinimizedMedianOperations: median number of operations
+% - \packageMinimizedPercentEnvir: percentages of sites with the default argument
+% - \packageMinimizedPercentCallsEnvir: percentage of calls with the default envir argument
 

### 2021-04-09 analysis/normalized.Rmd
Minimized expressions: better number of operations ( minimized, normalized to eval(1) )

@@
   E_raw <- E_raw %>% mutate(envir_expression = if_else(envir_type == "NULL", "NULL", envir_expression)) %>% 
-    mutate(envir_expression = na_if(envir_expression, "NA"))
+    mutate(envir_expression = na_if(envir_expression, "NA")) %>%
+    mutate(interp_eval = interp_eval - 14)  # 14 is the magic number corresponding to the number of steps of `eval(1)` (or `eval(some value)`)
 

### 2021-04-09 paper/main.tex
Minimized expressions: better number of operations ( minimized, normalized to eval(1) )

@@
 of that form; \#packages is the number of packages with that form; \#operations
-is the average number of instructions performed by the interpreter; and \%envir
+is the median number of instructions performed by the interpreter; and \%envir
 is the percentage of sites for which the \c{envir} argument is explicitly passed
@@
   % macros are properly rounded and show the percentage sign now
-\c X&\packageMinimizedcallsitesa &\packageMinimizedpropsitesa &\packageMinimizedpackagea &\packageMinimizedoperationsaRnd &\packageMinimizedpercentenvirnonparenta & \c{y+1}\\\hline
-\c{F(F(X))} & \packageMinimizedcallsitesb  & \packageMinimizedpropsitesb & \packageMinimizedpackageb  & \packageMinimizedoperationsbRnd & \packageMinimizedpercentenvirnonparentb & \c{gbov( mean(x), a-1)}\\\hline
-\c{V}&\packageMinimizedcallsitesc &\packageMinimizedpropsitesc &\packageMinimizedpackagec &\packageMinimizedoperationscRnd &\packageMinimizedpercentenvirnonparentc& \c{c(42,21,0)}\\\hline
-\c{F(X)}& \packageMinimizedcallsitesd & \packageMinimizedpropsitesd & \packageMinimizedpackaged & \packageMinimizedoperationsdRnd & \packageMinimizedpercentenvirnonparentd & \c{seq\_len(iters)} \\\hline
-\c{model.frame} & \packageMinimizedcallsitese & \packageMinimizedpropsitese & \packageMinimizedpackagee & \packageMinimizedoperationseRnd & \packageMinimizedpercentenvirnonparente & \c{model.frame(formula = Z ~ U)} \\\hline
-\c{F()}& \packageMinimizedcallsitesf & \packageMinimizedpropsitesf & \packageMinimizedpackagef & \packageMinimizedoperationsfRnd & \packageMinimizedpercentenvirnonparentf & \c{rgamma(3, 2, n = 10L)} \\\hline
-\c{\$}& \packageMinimizedcallsitesg & \packageMinimizedpropsitesg & \packageMinimizedpackageg & \packageMinimizedoperationsgRnd & \packageMinimizedpercentenvirnonparentg & \c{DF\$B} \\\hline
-\c{FUN} & \packageMinimizedcallsitesh & \packageMinimizedpropsitesh & \packageMinimizedpackageh & \packageMinimizedoperationshRnd & \packageMinimizedpercentenvirnonparenth & \c{function(x, y) x + 3 * y} \\\hline
-\c{<-} & \packageMinimizedcallsitesi  & \packageMinimizedpropsitesi & \packageMinimizedpackagei & \packageMinimizedoperationsiRnd & \packageMinimizedpercentenvirnonparenti & \c{x[1, 2:3, 2:3] <- value}\\\hline
-\c{\{\}} & \packageMinimizedcallsitesj & \packageMinimizedpropsitesj & \packageMinimizedpackagej & \packageMinimizedoperationsjRnd & \packageMinimizedpercentenvirnonparentj & \c{\{ write.csv(iris, tf) ; file.size(tf) \}} \\\hline
+\c X&\packageMinimizedcallsitesa &\packageMinimizedpropsitesa &\packageMinimizedpackagea &\packageMinimizedmedianoperationsaRnd &\packageMinimizedpercentenvirnonparenta & \c{y+1}\\\hline
+\c{F(F(X))} & \packageMinimizedcallsitesb  & \packageMinimizedpropsitesb & \packageMinimizedpackageb  & \packageMinimizedmedianoperationsbRnd & \packageMinimizedpercentenvirnonparentb & \c{gbov( mean(x), a-1)}\\\hline
+\c{V}&\packageMinimizedcallsitesc &\packageMinimizedpropsitesc &\packageMinimizedpackagec &\packageMinimizedmedianoperationscRnd &\packageMinimizedpercentenvirnonparentc& \c{c(42,21,0)}\\\hline
+\c{F(X)}& \packageMinimizedcallsitesd & \packageMinimizedpropsitesd & \packageMinimizedpackaged & \packageMinimizedmedianoperationsdRnd & \packageMinimizedpercentenvirnonparentd & \c{seq\_len(iters)} \\\hline
+\c{model.frame} & \packageMinimizedcallsitese & \packageMinimizedpropsitese & \packageMinimizedpackagee & \packageMinimizedmedianoperationseRnd & \packageMinimizedpercentenvirnonparente & \c{model.frame(formula = Z ~ U)} \\\hline
+\c{F()}& \packageMinimizedcallsitesf & \packageMinimizedpropsitesf & \packageMinimizedpackagef & \packageMinimizedmedianoperationsfRnd & \packageMinimizedpercentenvirnonparentf & \c{rgamma(3, 2, n = 10L)} \\\hline
+\c{\$}& \packageMinimizedcallsitesg & \packageMinimizedpropsitesg & \packageMinimizedpackageg & \packageMinimizedmedianoperationsgRnd & \packageMinimizedpercentenvirnonparentg & \c{DF\$B} \\\hline
+\c{FUN} & \packageMinimizedcallsitesh & \packageMinimizedpropsitesh & \packageMinimizedpackageh & \packageMinimizedmedianoperationshRnd & \packageMinimizedpercentenvirnonparenth & \c{function(x, y) x + 3 * y} \\\hline
+\c{<-} & \packageMinimizedcallsitesi  & \packageMinimizedpropsitesi & \packageMinimizedpackagei & \packageMinimizedmedianoperationsiRnd & \packageMinimizedpercentenvirnonparenti & \c{x[1, 2:3, 2:3] <- value}\\\hline
+\c{\{\}} & \packageMinimizedcallsitesj & \packageMinimizedpropsitesj & \packageMinimizedpackagej & \packageMinimizedmedianoperationsjRnd & \packageMinimizedpercentenvirnonparentj & \c{\{ write.csv(iris, tf) ; file.size(tf) \}} \\\hline
 \end{tabular}

### 2021-04-10 analysis/usage.Rmd
Most used function

@@
     ast_nodes = expr_resolved_nodes, #   Size in AST Nodes
+    func = caller_function,
     type = expr_resolved_type_tag,   #   Argument type
@@
 
+# Most used function
+
+```{r}
+frequent_funcs <- E %>% count(func, wt = duplicates, sort = TRUE) %>% 
+  mutate(p = n / nb_calls, cp = cumsum(p))
+
+top_func <- frequent_funcs %>% .[1,]
+
+overview_table(
+  r("top func name", top_func$func),
+  r("top func calls", top_func$n),
+  r("top func percent", percent(top_func$p))
+) -> TBL
+```
+
+`r TBL`
+
 ```{r include=FALSE}

### 2021-04-10 paper/main.tex
Add a few more lines on why it is hard

@@
 For performance reason, most of the tracing is done in C++. While in theory the
-implementation is rather straightforward, what makes it hard in practice is that
-(1) some of the core R functionality such as package loading is implemented
-using \eval which then leaks to the tracer in unexpected places, (2) lazy
-evaluation makes it harder to analyze function arguments while tracing as
-prematurely forcing a promise might have consequences, and finally (3) running
-large amount of real-world code triggers exercises a lot of corner cases of
-highly underspecified R behavior.
+implementation is rather straightforward, it is not so in practice:
+%
+\begin{compactitem}[---]
+
+\item The lazy evaluation makes it difficult to analyze function arguments
+  while tracing as prematurely forcing a promise might have dangerous
+  consequences.
+
+\item While the R interpreter is implemented in C, a lot of the core
+  functionality is done in R. For example, package loading is implemented in R
+  using \eval or the S4 and R6 object systems. This makes it hard in the tracer
+  to separate the \eval that are essential to user programs from the
+  accidental ones that are products of the way R implements its basic
+  operations. This is even more so for the side-effects analysis.
+
+\item In the dynamic analysis, we run all the \emph{runnable} code we obtained
+  from CRAN and Kaggle---\ie real code written by a people with highly varying
+  expertise in R and programming in general. This exercise a lot of the corner
+  cases of the highly underspecified R behavior.
+
+\end{compactitem}
 

### 2021-04-11 paper/main.tex
tax/arguments

@@
   $min(e)$& \#sites & \%sites & \#packages & \#operations & \%envir & example\\\hline
-  % macros are properly rounded and show the percentage sign now
 \c X&\packageMinimizedcallsitesa &\packageMinimizedpropsitesa &\packageMinimizedpackagea &\packageMinimizedmedianoperationsaRnd &\packageMinimizedpercentenvirnonparenta & \c{y+1}\\\hline
@@
 
-\medskip\noindent\EE{$min(e)=\c{V}$} Expression that represent values are
-frequently passed to \eval -- they represent 17\% of the calls. The majority of
-those, \packageValOneNodePercent, are inline constants (integers or doubles).
-The rest trivially evaluate to a value, \eg~\c{1+1}.\footnote{True as long as
-base functions such as \c{+} are not redefined. They typically aren't, but this
-is a limitation of this categorization.} In our corpus,
+\medskip\noindent\EE{$min(e)=\c{V}$} Expressions that represent values are
+frequently passed to \eval -- they occur in 17\% of sites. The majority of
+those, \packageValOneNodePercent, are inline constants (integer or double
+vectors). The rest trivially evaluate to a value, \eg~\c{1+1}.\footnote{True as
+long as base functions such as \c{+} are not redefined. They typically aren't,
+but this is a limitation of this categorization.} In our corpus,
 \packageNbCallSitesUniqueActualValue call sites only ever see a simple value.
@@
 constructed dynamically or the value is a default case that sometimes is
-replaced by a more interesting expression.
+replaced by a more interesting expression. This form is usually evaluated in few
+interpreter steps, in fact the median is only
+\packageMinimizedmedianoperationscRnd.
 
@@
 be bound to a promise, and accessing it may trigger evaluation of that promise,
-thus resulting in an arbitrary amount of computation.
+thus resulting in an arbitrary amount of computation. The median number of
+interpreted operations is \packageMinimizedmedianoperationsaRnd, suggesting that
+it is not the common case.
 
@@
 \c{x[[24]]}. As with \c X we allow arithmetics and values in this form. Lookup
-occurs in 6\% of the sites. Arbitrary computations can occur if one of the
-variables is bound to a promise.
+occurs in 6\% of the sites. The interpreter evaluates
+\packageMinimizedmedianoperationsgRnd operations on average; the minimum is 3
+operations.
 
@@
 environment {\tt <\,\!<-}, and the \c{\$} form. Assignments occur in 4\% of the
-sites. They represent the most obvious source of side-effects.
+sites. They represent the most obvious source of side-effects. The median number
+of operations is \packageMinimizedmedianoperationsiRnd; the minimum is 3.
 
@@
 
-\medskip\noindent\EE{$min(e)=\c{F(X)}$} This form allows for a simple function
-call whose arguments may include variable references. Assignments are excluded
-from this form. It occurs in 14\% of the sites in the corpus.
-
-\medskip\noindent\EE{$min(e)=\c{F(F(X))}$} This form extends the previous one
-with nested function calls. It is the second most common argument to \eval,
-occurring in 20\% of the sites.
-
-\medskip\noindent\EE{$min(e)=\c{FUN}$} This form captures expressions that are
-function definitions, \eg~\c{function(x)x+1}. \c{FUN} occurs in
-\packageFunctionDefinitionSitesPercent of sites. One data point we have not
-recorded is the side effects that are performed when the generated functions are
-eventually run. In addition to \c{FUN} there
-\packageGeneralizedFunctionDefinitionSitesPercent of sites create functions
-nested in other expressions, this gives an idea of the use of higher-order
-functions.
-
-\medskip\noindent\EE{$min(e)=\c{BLOCK}$} This form captures blocks of code with
-multiple statements. These are larger expressions, we do not inspect contents of
-the blocks. They occur in 4\% of the sites.
-
-\medskip\noindent\EE{$min(e)=\c{model.frame}$} This is a very specific form
-intended to capture the construction of calls to the \c{model.frame} function.
-This function creates statistical models. This subsumes \c{F(F(X))}, \c{FUN} and
-assignments. This particular form occurs in 7\% of the sites.
+\medskip\noindent\framebox{$min(e)=\c{F(X)}$}~\EE{$min(e)=\c{F(F(X))}$} These
+forms allow for function calls whose arguments may include variable references.
+The latter allows nested calls.Assignments are excluded from this form. They
+occurs in, respectively, 14\% and 20\% of the sites in the corpus. Together they
+are the most frequent forms. The median numbers of interpreter steps are,
+respectively, \packageMinimizedmedianoperationsdRnd and
+\packageMinimizedmedianoperationsbRnd.
+
+\medskip\noindent\EE{$min(e)=\c{FUN}$} This form captures expressions define
+functions, \eg~\c{function(x)x+1}, and do nothing else. \c{FUN} occurs in
+\packageFunctionDefinitionSitesPercent of sites. Evaluating a function
+definition is done in 2 interpreter step; the data does not record the work
+performed by the interpreter when the generated functions are eventually run. In
+addition to \c{FUN}, \packageGeneralizedFunctionDefinitionSitesPercent of sites
+have function definitions nested in other expressions, this gives an idea of the
+use of higher-order functions.
+
+\medskip\noindent\EE{$min(e)=\c{BLOCK}$} This form captures multi-statements
+code blocks, which occur in only 4\% of sites. These are larger expressions, we
+do not inspect contents of the blocks. The median number of executed operations
+is \packageMinimizedmedianoperationsjRnd.
+
+\medskip\noindent\EE{$min(e)=\c{model.frame}$} The \c{model.frame} function
+returns a dataframe that is the result of fitting the model described in a given
+formula. This form subsumes \c{F(F(X))}, \c{FUN} and assignments. It is the
+single most popular function invoked from \eval, it occurs in 7\% of the sites.
 

### 2021-04-11 analysis/normalized.Rmd

@@
     caller_srcref,
-    caller_expression
+    caller_expression,
+    environment_class
   ) -> E
@@
             env_diversity = n_distinct(envir), 
+            percent_envir_parent_frames = ratio(n_distinct(cur_data()$src_ref[str_starts(cur_data()$environment_class, fixed("caller-"))]),  n_call_sites),
             percent_envir_non_parentframe = ratio(n_distinct(cur_data()$src_ref[!is.na(cur_data()$envir) & cur_data()$envir != "parent.frame()"]),  n_call_sites)) %>% 
@@
 r_vec("MinimizedPercentEnvirNonParent", nrows, pull(widespread_minimized, percent_envir_non_parentframe), digits=0)
+r_vec("MinimizedPercentParentFrames", nrows, pull(widespread_minimized, percent_envir_parent_frames), digits=0)
 ```

### 2021-04-11 paper/main.tex
minor
Fix table

@@
 is the median number of instructions performed by the interpreter; and \%envir
-is the percentage of sites for which the \c{envir} argument is explicitly passed
-to \eval and is not \c{parent.frame()}. The example column shows one sample
+is the percentage of sites for which the \c{envir} argument is current frame or one of the parent frames. The example column shows one sample
 expression $e$ that is normalized to the particular form.
@@
   $min(e)$& \#sites & \%sites & \#packages & \#operations & \%envir & example\\\hline
-\c X&\packageMinimizedcallsitesa &\packageMinimizedpropsitesa &\packageMinimizedpackagea &\packageMinimizedmedianoperationsaRnd &\packageMinimizedpercentenvirnonparenta & \c{y+1}\\\hline
-\c{F(F(X))} & \packageMinimizedcallsitesb  & \packageMinimizedpropsitesb & \packageMinimizedpackageb  & \packageMinimizedmedianoperationsbRnd & \packageMinimizedpercentenvirnonparentb & \c{gbov( mean(x), a-1)}\\\hline
-\c{V}&\packageMinimizedcallsitesc &\packageMinimizedpropsitesc &\packageMinimizedpackagec &\packageMinimizedmedianoperationscRnd &\packageMinimizedpercentenvirnonparentc& \c{c(42,21,0)}\\\hline
-\c{F(X)}& \packageMinimizedcallsitesd & \packageMinimizedpropsitesd & \packageMinimizedpackaged & \packageMinimizedmedianoperationsdRnd & \packageMinimizedpercentenvirnonparentd & \c{seq\_len(iters)} \\\hline
-\c{model.frame} & \packageMinimizedcallsitese & \packageMinimizedpropsitese & \packageMinimizedpackagee & \packageMinimizedmedianoperationseRnd & \packageMinimizedpercentenvirnonparente & \c{model.frame(formula = Z ~ U)} \\\hline
-\c{F()}& \packageMinimizedcallsitesf & \packageMinimizedpropsitesf & \packageMinimizedpackagef & \packageMinimizedmedianoperationsfRnd & \packageMinimizedpercentenvirnonparentf & \c{rgamma(3, 2, n = 10L)} \\\hline
-\c{\$}& \packageMinimizedcallsitesg & \packageMinimizedpropsitesg & \packageMinimizedpackageg & \packageMinimizedmedianoperationsgRnd & \packageMinimizedpercentenvirnonparentg & \c{DF\$B} \\\hline
-\c{FUN} & \packageMinimizedcallsitesh & \packageMinimizedpropsitesh & \packageMinimizedpackageh & \packageMinimizedmedianoperationshRnd & \packageMinimizedpercentenvirnonparenth & \c{function(x, y) x + 3 * y} \\\hline
-\c{<-} & \packageMinimizedcallsitesi  & \packageMinimizedpropsitesi & \packageMinimizedpackagei & \packageMinimizedmedianoperationsiRnd & \packageMinimizedpercentenvirnonparenti & \c{x[1, 2:3, 2:3] <- value}\\\hline
-\c{\{\}} & \packageMinimizedcallsitesj & \packageMinimizedpropsitesj & \packageMinimizedpackagej & \packageMinimizedmedianoperationsjRnd & \packageMinimizedpercentenvirnonparentj & \c{\{ write.csv(iris, tf) ; file.size(tf) \}} \\\hline
+\c X&\packageMinimizedcallsitesa &\packageMinimizedpropsitesa &\packageMinimizedpackagea &\packageMinimizedmedianoperationsaRnd &\packageMinimizedpercentparentframesa & \c{y+1}\\\hline
+\c{F(F(X))} & \packageMinimizedcallsitesb  & \packageMinimizedpropsitesb & \packageMinimizedpackageb  & \packageMinimizedmedianoperationsbRnd & \packageMinimizedpercentparentframesb & \c{gbov( mean(x), a-1)}\\\hline
+\c{V}&\packageMinimizedcallsitesc &\packageMinimizedpropsitesc &\packageMinimizedpackagec &\packageMinimizedmedianoperationscRnd &\packageMinimizedpercentparentframesc& \c{c(42,21,0)}\\\hline
+\c{F(X)}& \packageMinimizedcallsitesd & \packageMinimizedpropsitesd & \packageMinimizedpackaged & \packageMinimizedmedianoperationsdRnd & \packageMinimizedpercentparentframesd & \c{seq\_len(iters)} \\\hline
+\c{\$} & \packageMinimizedcallsitese & \packageMinimizedpropsitese & \packageMinimizedpackagee & \packageMinimizedmedianoperationseRnd & \packageMinimizedpercentparentframese & \c{DF\$B}\\\hline
+\c{model.frame}& \packageMinimizedcallsitesf & \packageMinimizedpropsitesf & \packageMinimizedpackagef & \packageMinimizedmedianoperationsfRnd & \packageMinimizedpercentparentframesf &  \c{model.frame(formula = Z $\sim$ U)}   \\\hline
+\c{F()}& \packageMinimizedcallsitesg & \packageMinimizedpropsitesg & \packageMinimizedpackageg & \packageMinimizedmedianoperationsgRnd & \packageMinimizedpercentparentframesg & \c{rgamma(3, 2, n = 10L)} \\\hline
+\c{FUN} & \packageMinimizedcallsitesh & \packageMinimizedpropsitesh & \packageMinimizedpackageh & \packageMinimizedmedianoperationshRnd & \packageMinimizedpercentparentframesh & \c{function(x, y) x + 3 * y} \\\hline
+\c{<-} & \packageMinimizedcallsitesi  & \packageMinimizedpropsitesi & \packageMinimizedpackagei & \packageMinimizedmedianoperationsiRnd & \packageMinimizedpercentparentframesi & \c{x[1, 2:3, 2:3] <- value}\\\hline
+\c{\{\}} & \packageMinimizedcallsitesj & \packageMinimizedpropsitesj & \packageMinimizedpackagej & \packageMinimizedmedianoperationsjRnd & \packageMinimizedpercentparentframesj & \c{\{ write.csv(iris, tf) ; file.size(tf) \}} \\\hline
 \end{tabular}
@@
 The top-level environment in R is called the global environment. New
-environments can be created using \c{new.env}, or \c{list2env}. They can be provided a parent
-environment which becomes the enclosing scope of the new environment.
+environments can be created using \c{new.env}, or \c{list2env}. They can be
+provided a parent environment which becomes the enclosing scope of the new
+environment.
 % as.environment also can be used sometimes for that

### 2021-04-11 analysis/environments.Rmd
Environment

@@
   r("defaut envir sites percent", ratio(nb_default_sites, nb_call_sites)),
-  r("non default envir sites", nb_explicit),
-  r("non defaut envir sites percent", ratio(nb_explicit, nb_call_sites))
+  r("non default envir sites", nb_explicit_sites),
+  r("non defaut envir sites percent", ratio(nb_explicit_sites, nb_call_sites))
 )
@@
 ```{r envir_absolute}
-eval_calls_absolute <- eval_calls_explicit %>% filter(envir_type == "environment")
+eval_calls_absolute <- eval_calls_explicit %>% filter(envir_type %in% c("environment", "S4"))
 
@@
 
-# TODO: handle scientific notation for caller-1e+00-
-
-env_calls <- eval_calls_absolute %>% ungroup() %>%
+env_calls <- E_envir %>% filter(envir_type %in% c("environment", "S4")) %>%
   extract(environment_class, regex = "((?:new\\+)*)caller-([[:digit:]]+)-(.*)", into = c("new_env", "hierarchy", "specific_env"), remove = FALSE, convert = TRUE) %>%
   mutate(new_env = na_if(new_env, ""), specific_env = na_if(specific_env, ""), new_env = str_count(new_env, fixed("new+"))) %>%
-  mutate(specific_env = if_else(specific_env == "emptyR_EmptyEnv", "empty", specific_env))
+  mutate(specific_env = if_else(specific_env == "emptyR_EmptyEnv", "empty", specific_env)) %>% 
+  mutate(hierarchy = hierarchy - 1) # Fix. Remove with the next run
 ```
@@
   r("nb base env site percent", ratio(nb_base_env_sites, nb_call_sites))
+)
+```
+
+Glabal can happen by chance, because `parent.frame()`goes up too much and reaches the global environment. SO when is it explicitly passed?
+
+```{r}
+explicit_global <- E_envir %>% filter(envir %in% c("globalenv()", ".GlobalEnv"))
+
+nb_explicit_global_sites <- explicit_global %>% pull(src_ref) %>% n_distinct()
+
+overview_table(
+  r("nb explicit global sites", nb_explicit_global_sites),
+  r("nb explicit global site percent", ratio(nb_explicit_global_sites, nb_call_sites))
 )
@@
 
-`caller-n` wins over any specific env. We add `+` if there are any `new`.
+A specific env wins over `caller-n`. We add `+` if there are any `new`.
 
-TODO: if there is both new , a caller-n, a specific class, rather ignore the caller-n
 ```{r}
 env_classes <- env_calls %>%
-  mutate(env_class = if_else(environment_class == "loop", "loop", paste0(if_else(is.na(hierarchy), specific_env, as.character(hierarchy)), if_else(is.na(new_env), "", "+"))))
+  mutate(env_class = if_else(environment_class == "loop", "loop", paste0(if_else(is.na(specific_env), as.character(hierarchy), specific_env), if_else(is.na(new_env), "", "+"))))
 
-env_classes %>% count(env_class, wt = duplicates, sort = TRUE) %>%
-  mutate(pc = round(100 * n / sum(n), 2), cpc = cumsum(pc)) %>%
-  datatable()
+not_env_calls <- E_envir %>% anti_join(env_classes) %>% 
+  mutate(env_class = case_when(
+    envir_type == "list" ~ "list",
+    envir_type == "integer" ~ "int",
+    envir_type == "NULL" ~ "NULL",
+    TRUE ~ "other"
+  ))
 
-env_classes %>% group_by(env_class) %>% summarise(n = n_distinct(src_ref)) %>%
+env_classes <- bind_rows(env_classes, not_env_calls)
+
+env_classes_calls <- env_classes %>% count(env_class, wt = duplicates, sort = TRUE) %>%
+  mutate(pc = ratio(n, nb_eval_calls), cpc = cumsum(pc))
+
+env_classes_sites <- env_classes %>% group_by(env_class) %>% summarise(n = n_distinct(src_ref)) %>%
   arrange(desc(n)) %>%
-  mutate(pc = round(100 * n / sum(n), 2), cpc = cumsum(pc)) %>%
-  datatable()
+  mutate(pc = ratio(n, nb_call_sites), cpc = cumsum(pc)) 
+```
+
+```{r}
+env_classes_calls %>% datatable()
+
+env_classes_sites %>% datatable()
+
+nrows <- 8
+r_vec("env class names", nrows, env_classes_sites$env_class)
+r_vec("env class sites", nrows, env_classes_sites$n)
+r_vec("env class site percent", nrows, env_classes_sites$pc)
 ```

### 2021-04-11 paper/main.tex
Environment

@@
 
-\c{NULL}, a \c{list} or a data frame, or an integer: \eval copies the fields of the
+It can also be \c{NULL}, a \c{list} or a data frame, or an integer: \eval copies the fields of the
 list or data frame and creates bindings for them in a new environment.
-
-\subsubsection{Bits of text, possibly reusable}
-
-This pattern is used to evaluate formulas which can directly refer to the fields
-of the data.
 
@@
 \c{n} denotes the environment of the $n$-th call stack frame from the current
-function. \c{global} denotes the top-level environment and \c{list} denotes a
+function: \c{0} is the stack of the current function, \c{1} is the parent frame. \c{global} denotes the top-level environment and \c{list} denotes a
 list passed for evaluation of formulas. Environment classes of the form $n+$
@@
 
-% This table is old and incorrect
-% Working on updating it
+% Updated with the last run
+% I made some changes and we need another run to get the right results
+% The difference should be that there will much more package environments
+% the loop category should disappear from the top results also.
 \begin{table}[htbp]{ \centering
         \begin{tabular}{|c|c|}\hline
- Environment Class & Eval Calls\\\hline
- \PackagesEnvClassA &  \PackagesProportionA\\
- \PackagesEnvClassB &  \PackagesProportionB\\
- \PackagesEnvClassC &  \PackagesProportionC\\
- \PackagesEnvClassD &  \PackagesProportionD\\
-\PackagesEnvClassE &  \PackagesProportionE\\\hline
+            Environment Class & Sites\\\hline
+            \packageEnvClassNamesa &  \packageEnvClassSitePercenta\\
+            \packageEnvClassNamesb &  \packageEnvClassSitePercentb\\
+            \packageEnvClassNamesc &  \packageEnvClassSitePercentc\\
+            \packageEnvClassNamesd &  \packageEnvClassSitePercentd\\
+            \packageEnvClassNamese &  \packageEnvClassSitePercente\\
+            \packageEnvClassNamesf &  \packageEnvClassSitePercentf\\
+            \packageEnvClassNamesg &  \packageEnvClassSitePercentg\\
+            \packageEnvClassNamesh &  \packageEnvClassSitePercenth\\
+            \hline
     \end{tabular}} \label{tab:environments}
-\caption{Environments in terms of \eval calls}
+    \caption{Environments in terms of sites. One site can be passed different kinds of environments.}
 \end{table}
+
+Most of the calls happen in the environment surrounding \eval.
+This corresponds mostly to the default \c{envir} argument of \eval; indeed, \packageDefautEnvirSitesPercent of the sites have that default argument.
+A lot of sites also use \c{parent.frame()} to access their parent frame.
+It corresponds to environment class \c{1} but also to \c{global} when the function is called at the top level.
+Explicitly and directly referring to the global environment using \c{globalenv()} or \c{.GlobalEnv} is rare, as it happens only in \packageNbExplicitGlobalSites sites.
+
+\subsubsection{Bits of text, possibly reusable}
+
+This pattern is used to evaluate formulas which can directly refer to the fields
+of the data.
 
@@
 
+\todo{That should be moved to the side effect section and updated with the last data.}
 Categorize side-effects between env passed to eval and other envs:

### 2021-04-11 analysis/environments.Rmd
Environemnt high-level classification

@@
 
+# Caller environments
+
+```{r}
+caller_envs <- env_calls %>% filter(!is.na(hierarchy), is.na(specific_env))
+nb_caller_env_sites <- caller_envs %>% pull(src_ref) %>% n_distinct()
+
+overview_table(
+  r("nb caller env sites", nb_caller_env_sites),
+  r("caller env site percent", ratio(nb_caller_env_sites, nb_call_sites))
+)
+```
+
+# New environments
+
+```{r}
+new_envs <- env_calls %>% filter(!is.na(new_env))
+nb_new_env_sites <- new_envs %>% pull(src_ref) %>% n_distinct()
+
+overview_table(
+  r("nb new env sites", nb_new_env_sites),
+  r("new env site percent", ratio(nb_new_env_sites, nb_call_sites))
+)
+```
+
 # Specific environments
@@
   r("nb null envir sites", nb_null_sites),
-  r("nb null envir site percent", ratio(nb_null_sites, nb_eval_calls))
+  r("nb null envir site percent", ratio(nb_null_sites, nb_call_sites))
 )

### 2021-04-11 analysis/usage.Rmd
Environemnt high-level classification

@@
 params:
-  base_dir: Data/base/
+  base_dir: Data/package/
   calls_path: summarized.fst

### 2021-04-12 analysis/environments.Rmd
Up-to-date environments

@@
 ```{r}
-envir_var <- frequent_envir %>% filter(is.na(envir), str_detect(envir, "^(?:\\.[:alpha]|[:alpha:])[[:alpha:][:digit:]_\\.]*$"))
+envir_var <- frequent_envir %>% filter(envir != "NA", str_detect(envir, "^(?:\\.[:alpha]|[:alpha:])[[:alpha:][:digit:]_\\.]*$"))
 
@@
 ```{r}
-envir_env <- E_envir %>% filter(envir_type == "environment") 
+envir_env <- E_envir %>% filter(envir_type %in% c("environment", "S4")) 
 # S4 class: from pim package, uses an environment under the hood
@@
 env_calls <- E_envir %>% filter(envir_type %in% c("environment", "S4")) %>%
-  extract(environment_class, regex = "((?:new\\+)*)caller-([[:digit:]]+)-(.*)", into = c("new_env", "hierarchy", "specific_env"), remove = FALSE, convert = TRUE) %>%
+  extract(environment_class, regex = "((?:new\\+)*)(?:caller-(-?[[:digit:]]+)-)?(.*)", into = c("new_env", "hierarchy", "specific_env"), remove = FALSE, convert = TRUE) %>%
   mutate(new_env = na_if(new_env, ""), specific_env = na_if(specific_env, ""), new_env = str_count(new_env, fixed("new+"))) %>%
-  mutate(specific_env = if_else(specific_env == "emptyR_EmptyEnv", "empty", specific_env)) %>% 
-  mutate(hierarchy = hierarchy - 1) # Fix. Remove with the next run
+  mutate(specific_env = if_else(specific_env == "emptyR_EmptyEnv", "empty", specific_env)) 
 ```
@@
 
-Glabal can happen by chance, because `parent.frame()`goes up too much and reaches the global environment. SO when is it explicitly passed?
+Glabal can happen by chance, because `parent.frame()`goes up too much and reaches the global environment. So when is it explicitly passed?
 
@@
 ```
+
+And when is it passed through a variable?
+
+```{r}
+variable_global <- E_envir %>% filter(environment_class == "global") %>% semi_join(envir_var)
+
+nb_variable_global_sites <- variable_global %>% pull(src_ref) %>% n_distinct()
+
+overview_table(
+  r("nb variable global sites", nb_variable_global_sites),
+  r("nb variable global site percent", ratio(nb_variable_global_sites, nb_call_sites))
+)
+```
+
+So we can deduce an approximate (upper-bound) of the number of sites that touch eval not by chance!
 
@@
 env <- attach(NULL, pos, basename(directory))
+```
+
+## Namespaces
+
+In R, namespaces are different from the package environment.
+
+```{r}
+namespace_envs <- E_envir %>% filter(!str_detect(environment_class, fixed("caller-")), !str_detect(environment_class, "global|empty|base|loop"), !str_detect(environment_class, fixed("package:")))
+
+nb_namespace_env_sites <- namespace_envs %>% pull(src_ref) %>% n_distinct()
+
+overview_table(
+  r("nb namespace env sites", nb_namespace_env_sites),
+  r("nb namespace env site percent", ratio(nb_namespace_env_sites, nb_call_sites))
+)
+```
+
+What are the most frequent namespaces?
+
+```{r}
+namespace_envs %>% count(environment_class, wt = duplicates, sort = TRUE) %>%
+  mutate(pc = ratio(n, sum(n)), cpc = cumsum(pc))
+
+namespace_envs %>% group_by(environment_class) %>% 
+  summarise(n = n_distinct(src_ref)) %>% 
+  arrange(desc(n)) %>% 
+   mutate(pc = ratio(n, sum(n)), cpc = cumsum(pc))
 ```

### 2021-04-12 paper/main.tex
Environemnt high-level classification
Up-to-date environments

@@
 
+A more high-level classification: Table~\ref{tab:highlevelenvironments}
+
+\begin{table}[h]
+    \centering
+\begin{tabular}{|c|c|c|}
+    \hline
+    Environment & \#sites & \%sites \\
+    \hline
+    Parent frames &  \packageNbCallerEnvSites &  \packageCallerEnvSitePercent\\
+    Global &  \packageNbGlobalEnvSites & \packageNbGlobalEnvSitePercent \\
+    New & \packageNbNewEnvSites & \packageNewEnvSitePercent \\
+    Base &  \packageNbBaseEnvSites & \packageNbBaseEnvSitePercent \\
+    Package & \packageNbPackageEnvSites & \packagePackageEnvSitePercent \\
+    Namespace & \packageNbNamespaceEnvSites & \packageNbNamespaceEnvSitePercent \\
+    List &  \packageNbEnvListSites & \packageNbEnvListSitePercent \\
+    Integer & \packageNbIntegerEnvirSites  & \packageNbIntegerEnvirSitePercent \\
+    Null &  \packageNbNullEnvirSites & \packageNbNullEnvirSitePercent \\
+    Empty & \packageNbEmptyEnvSites & \packageNbEmptyEnvSitePercent \\
+    \hline
+\end{tabular} \label{tab:highlevelenvironments}
+\caption{Classification of environments per site.}
+\end{table}
+
 \subsubsection{Bits of text, possibly reusable}

### 2021-04-12 paper/main.tex
minor

--- 2021-04-11 paper/main.tex
@@
 is the median number of instructions performed by the interpreter; and \%envir
-is the percentage of sites for which the \c{envir} argument is current frame or one of the parent frames. The example column shows one sample
-expression $e$ that is normalized to the particular form.
+is the ratio of sites for that evaluate in a function environment. The example
+column shows one sample expression $e$ that normalizes to the particular form.
 
@@
 \c{<-} & \packageMinimizedcallsitesi  & \packageMinimizedpropsitesi & \packageMinimizedpackagei & \packageMinimizedmedianoperationsiRnd & \packageMinimizedpercentparentframesi & \c{x[1, 2:3, 2:3] <- value}\\\hline
-\c{\{\}} & \packageMinimizedcallsitesj & \packageMinimizedpropsitesj & \packageMinimizedpackagej & \packageMinimizedmedianoperationsjRnd & \packageMinimizedpercentparentframesj & \c{\{ write.csv(iris, tf) ; file.size(tf) \}} \\\hline
+\c{BLOCK} & \packageMinimizedcallsitesj & \packageMinimizedpropsitesj & \packageMinimizedpackagej & \packageMinimizedmedianoperationsjRnd & \packageMinimizedpercentparentframesj & \c{\{ write.csv(iris, tf) ; file.size(tf) \}} \\\hline
 \end{tabular}
@@
 interpreter steps, in fact the median is only
-\packageMinimizedmedianoperationscRnd.
+\packageMinimizedmedianoperationscRnd. The environment in which they evaluate is
+mostly irrelevant (unless it redefines a built-in operator).
+%% Why then this variety of environments??
 
@@
 Explicitly and directly referring to the global environment using \c{globalenv()} or \c{.GlobalEnv} is rare, as it happens only in \packageNbExplicitGlobalSites sites.
+
+A more high-level classification: Table~\ref{tab:highlevelenvironments}
+
+\begin{table}
+    \centering
+\begin{tabular}{|c|c|c|}
+    \hline
+    Environment & \#sites & \%sites \\
+    \hline
+    Parent frames &  \packageNbCallerEnvSites &  \packageCallerEnvSitePercent\\
+    \hline
+    Global &  \packageNbGlobalEnvSites & \packageNbGlobalEnvSitePercent \\
+    \hline
+    New & \packageNbNewEnvSites & \packageNewEnvSitePercent \\
+    \hline
+    Base &  \packageNbBaseEnvSites & \packageNbBaseEnvSitePercent \\
+    \hline
+    % Should be many more (next run)
+    Package/namespace & \packageNbPackageEnvSites & \packagePackageEnvSitePercent \\
+    \hline
+    List &  \packageNbEnvListSites & \packageNbEnvListSitePercent \\
+    \hline
+    Integer & \packageNbIntegerEnvirSites  & \packageNbIntegerEnvirSitePercent \\
+    \hline
+    Null &  \packageNbNullEnvirSites & \packageNbNullEnvirSitePercent \\
+    \hline
+    Empty & \packageNbEmptyEnvSites & \packageNbEmptyEnvSitePercent \\
+    \hline
+\end{tabular} \label{tab:highlevelenvironments}
+\caption{Classification of environments per site.}
+\end{table}
 

### 2021-04-12 analysis/environments.Rmd
Environments: table for new in progress

@@
 frequent_envir <- E_envir %>% group_by(src_ref) %>% 
-  summarize(n_envir = n_distinct(envir), envir = paste0(unique(envir), collapse = " ; "), envir_type = paste0(unique(envir_type), collapse = " ; "), n_types = n_distinct(envir_type))
+  summarize(n_envir = n_distinct(envir), envir = paste0(sort(unique(envir)), collapse = " ; "), envir_type = paste0(unique(envir_type), collapse = " ; "), n_types = n_distinct(envir_type))
 
@@
 
-```
-
-# Caller environments
-
-```{r}
-caller_envs <- env_calls %>% filter(!is.na(hierarchy), is.na(specific_env))
-nb_caller_env_sites <- caller_envs %>% pull(src_ref) %>% n_distinct()
-
-overview_table(
-  r("nb caller env sites", nb_caller_env_sites),
-  r("caller env site percent", ratio(nb_caller_env_sites, nb_call_sites))
-)
-```
-
-# New environments
-
-```{r}
-new_envs <- env_calls %>% filter(!is.na(new_env))
-nb_new_env_sites <- new_envs %>% pull(src_ref) %>% n_distinct()
-
-overview_table(
-  r("nb new env sites", nb_new_env_sites),
-  r("new env site percent", ratio(nb_new_env_sites, nb_call_sites))
-)
 ```
@@
 
+# Less fine-grained classification
+
+We categorize environments in:
+
+- function environments (parent frames and integer)
+- package/namespace environments
+- list/NULL environments
+- global environment
+- new environment
+- empty environment
+
+## Caller environments
+
+```{r}
+caller_envs <- env_calls %>% filter(!is.na(hierarchy), is.na(specific_env))
+nb_caller_env_sites <- caller_envs %>% pull(src_ref) %>% n_distinct()
+
+overview_table(
+  r("nb caller env sites", nb_caller_env_sites),
+  r("caller env site percent", ratio(nb_caller_env_sites, nb_call_sites))
+)
+```
+
+## New environments
+
+```{r}
+new_envs <- env_calls %>% filter(!is.na(new_env))
+nb_new_env_sites <- new_envs %>% pull(src_ref) %>% n_distinct()
+
+overview_table(
+  r("nb new env sites", nb_new_env_sites),
+  r("new env site percent", ratio(nb_new_env_sites, nb_call_sites))
+)
+```
+
+We classify the most specific environment or caller environment after it.
+```{r}
+env_classes %>% filter(!is.na(new_env)) %>% group_by(env_class) %>% 
+  summarise(n = n_distinct(src_ref)) %>% 
+  arrange(desc(n)) %>% 
+  mutate(pc = ratio(n, nb_call_sites), cpc = cumsum(pc))
+```
+
+And with the higher-level categories:
+
+```{r}
+high_level_new <- env_calls %>% filter(!is.na(new_env)) %>% 
+  mutate(category = case_when(
+    !is.na(hierarchy) ~ "function",
+    specific_env == "global" ~ "global",
+    specific_env == "empty" ~ "empty",
+    str_detect(specific_env, fixed("loop")) ~ "other",
+    !is.na(specific_env) ~ "package"
+  ))
+
+high_level_new_sites <- high_level_new %>%group_by(category) %>%
+  summarise(n = n_distinct(src_ref)) %>% 
+  arrange(desc(n)) %>% 
+  mutate(pc = ratio(n, nb_new_env_sites), cpc = cumsum(pc))
+
+nrows <- 5
+r_vec("new env category name", nrows, high_level_new_sites$category)
+r_vec("new env category sites", nrows, high_level_new_sites$n)
+r_vec("new env category site percent", nrows, high_level_new_sites$pc)
+```
+
+## The final high-lelve classification
+
+```{r}
+nb_dataframe_env_sites <- nb_env_list_sites + nb_null_sites
+
+nb_pack_sites <- nb_package_env_sites + nb_namespace_env_sites + nb_base_env_sites
+
+nb_function_env_sites <- nb_caller_env_sites + nb_integer_sites
+
+overview_table(
+  r("nb dataframe env sites", nb_dataframe_env_sites),
+  r("nb dataframe env site percent", ratio(nb_dataframe_env_sites, nb_call_sites)),
+  r("nb package namespace env sites", nb_pack_sites),
+  r("nb package namespace env site percent", ratio(nb_pack_sites, nb_call_sites)),
+  r("nb function env sites", nb_function_env_sites),
+  r("nb function env site percent", ratio(nb_function_env_sites, nb_call_sites))
+)
+```
+
 # Specific envir with their environment class

### 2021-04-12 paper/main.tex
Environments: table for new in progress

--- 2021-04-11 paper/main.tex
@@
 provided a parent environment which becomes the enclosing scope of the new
-environment.
+environment. The new
+environment provides a limited form of sandboxing. All assignments using the
+$\leftarrow$ function occur inside it and prevent the extended environment from
+mutation. However, it is still possible to mutate the extended environment using
+the $\leftarrow$ or \c{assign} functions; but, that happens rarely.
 % as.environment also can be used sometimes for that
@@
 Another kind of environments that is passed to \eval is the environment of a
-package. It refers to the \emph{namespace} of a package. A call to
+package or its \emph{namespace}. A call to
 \c{library(packageName)}, which loads libraries in R, for instance, will
@@
 
+%TODO: there is a difference between a package environment and a namespace. Shortly explain.
+
 Most of the time, \ie in \packageEnvirEnvSitePercent of the sites, \eval is
@@
 
-We looked at the environments passed to all the \eval calls in our corpus.
-Table~\ref{tab:environments} summarizes the results. A numeric environment class
-\c{n} denotes the environment of the $n$-th call stack frame from the current
-function: \c{0} is the stack of the current function, \c{1} is the parent frame. \c{global} denotes the top-level environment and \c{list} denotes a
-list passed for evaluation of formulas. Environment classes of the form $n+$
-denote the $n$-th environment extended with a new environment. The new
-environment provides a limited form of sandboxing. All assignments using the
-$\leftarrow$ function occur inside it and prevent the extended environment from
-mutation. However, it is still possible to mutate the extended environment using
-the $\leftarrow$ or \c{assign} functions; but, that happens rarely.
-
-% Updated with the last run
-% I made some changes and we need another run to get the right results
-% The difference should be that there will much more package environments
-% the loop category should disappear from the top results also.
+We classify environments to \eval as follows:
+
+\begin{itemize}
+	\item function environments: the frame of the calling function or one of the parent frames (obtained using \c{parent.frame()}, \c{sys.frame()} or with an integer argument for \c{envir} for instance);
+	\item list environments: a list, a data frame, a tibble, or \c{NULL} (seen as the empty list), passed to \eval;
+	\item the global environment;
+	\item package and namespace environments;
+	\item the empty environment (created with \c{emptyenv()})
+\end{itemize}
+We also consider an orthogonal category: environments built with \c{new.env}. They can have as parent an environment of any of the other categories or another environment built with \c{new.env}.
+
+As shown in Table~\ref{tab:highlevelenvironments}, most calls happen in a function environment, with \c{global} as a distant second.
+
+\begin{table}[h]
+	\centering
+	\begin{tabular}{|c|c|c|}
+		\hline
+		Environment & \#sites & \%sites \\
+		\hline
+		Function environments &  \packageNbFunctionEnvSites &  \packageNbFunctionEnvSitePercent\\
+		Global &  \packageNbGlobalEnvSites & \packageNbGlobalEnvSitePercent \\
+		List &  \packageNbEnvListSites & \packageNbEnvListSitePercent \\
+		Package & \packageNbPackageNamespaceEnvSites & \packageNbPackageNamespaceEnvSitePercent \\
+		Empty & \packageNbEmptyEnvSites & \packageNbEmptyEnvSitePercent \\
+		\hline
+		New & \packageNbNewEnvSites & \packageNewEnvSitePercent \\
+		\hline
+	\end{tabular} \label{tab:highlevelenvironments}
+	\caption{Classification of environments per site.}
+\end{table}
+
+The calls happening in a function environment are mostly happening in the environment of the function calling \eval, as shown in Table%TODO
+
+%TODO: add here table of possible frame numbers
+
+Most of the calls happen in the environment surrounding \eval.
+This corresponds mostly to the default \c{envir} argument of \eval; indeed, \packageDefautEnvirSitesPercent of the sites have that default argument.
+A lot of sites also use \c{parent.frame()} to access their parent frame.
+It corresponds to environment class \c{1} but also to \c{global} when the function is called at the top level.
+Explicitly and directly referring to the global environment using \c{globalenv()} or \c{.GlobalEnv} is rare, as it happens only in \packageNbExplicitGlobalSites sites.
+Accessing further-away frames happens much more seldom.
+
+%TODO: add here table of possible environment categories wrapped in new
+
+% Up-to-date
 \begin{table}[htbp]{ \centering
@@
 \end{table}
-
-Most of the calls happen in the environment surrounding \eval.
-This corresponds mostly to the default \c{envir} argument of \eval; indeed, \packageDefautEnvirSitesPercent of the sites have that default argument.
-A lot of sites also use \c{parent.frame()} to access their parent frame.
-It corresponds to environment class \c{1} but also to \c{global} when the function is called at the top level.
-Explicitly and directly referring to the global environment using \c{globalenv()} or \c{.GlobalEnv} is rare, as it happens only in \packageNbExplicitGlobalSites sites.
 

### 2021-04-12 analysis/environments.Rmd
Add environment classification for new

@@
 ```{r}
-caller_envs <- env_calls %>% filter(!is.na(hierarchy), is.na(specific_env))
+caller_envs <- env_calls %>% filter(!is.na(hierarchy), is.na(specific_env) || specific_env == "_callee", is.na(new_env))
+
+loop_envs <- env_calls %>% filter(is.na(new_env), str_detect(specific_env, fixed("loop")))
+
+caller_envs <- bind_rows(caller_envs, loop_envs)
+
 nb_caller_env_sites <- caller_envs %>% pull(src_ref) %>% n_distinct()
@@
 )
+```
+
+```{r}
+# TODO: high-level per caller frame number...
 ```
@@
   mutate(category = case_when(
-    !is.na(hierarchy) ~ "function",
+    !is.na(hierarchy) ~ "function", # Will also include _callee
     specific_env == "global" ~ "global",
     specific_env == "empty" ~ "empty",
-    str_detect(specific_env, fixed("loop")) ~ "other",
+    str_detect(specific_env, fixed("loop")) ~ "function", #if it is loop, it means that we found the eval env in the frames,
     !is.na(specific_env) ~ "package"
@@
 
-nrows <- 5
+nrows <- 4
 r_vec("new env category name", nrows, high_level_new_sites$category)
@@
 
-## The final high-lelve classification
+## The final high-level classification
 

### 2021-04-12 paper/main.tex
Add environment classification for new

@@
 \end{itemize}
-We also consider an orthogonal category: environments built with \c{new.env}. They can have as parent an environment of any of the other categories or another environment built with \c{new.env}.
 
@@
 %TODO: add here table of possible environment categories wrapped in new
+
+\begin{table}[h]
+    \centering 
+    \begin{tabular}{|c|c|c|}
+        \hline
+        Category & \#sites & \%sites \\
+        \hline
+        \packageNewEnvCategoryNamea & \packageNewEnvCategorySitesa & \packageNewEnvCategorySitePercenta \\
+        \hline
+        \packageNewEnvCategoryNameb & \packageNewEnvCategorySitesb &  \packageNewEnvCategorySitePercentb\\
+        \hline
+        \packageNewEnvCategoryNamec & \packageNewEnvCategorySitesc & \packageNewEnvCategorySitePercentc \\
+        \hline
+        \packageNewEnvCategoryNamed & \packageNewEnvCategorySitesd & \packageNewEnvCategorySitePercentd \\
+        \hline
+    \end{tabular}
+\caption{Classification of environments wrapped by one of several \c{new}, in number of sites. Some sites can be passed more than one kind of environments. }
+\end{table}
 

### 2021-04-12 analysis/environments.Rmd
Explain the various environments wrapped with new

@@
 ```{r}
-caller_envs <- env_calls %>% filter(!is.na(hierarchy), is.na(specific_env) || specific_env == "_callee", is.na(new_env))
+caller_envs <- env_calls %>% filter(!is.na(hierarchy), is.na(specific_env) | specific_env == "_callee", is.na(new_env))
 
@@
 ```{r}
-# TODO: high-level per caller frame number...
+caller_env_sites <- caller_envs %>% group_by(hierarchy) %>% 
+  summarise(n = n_distinct(src_ref)) %>% 
+  arrange(desc(n)) %>% 
+   mutate(pc = ratio(n, nb_caller_env_sites), cpc = cumsum(pc))
+
+nrows <- 5
+r_vec("caller env hierarchy name", nrows, caller_env_sites$hierarchy)
+r_vec("caller env hierarchy sites", nrows, caller_env_sites$n)
+r_vec("caller env hierarchy site percent", nrows, caller_env_sites$pc)
+
+nb_far_away_caller_sites <- caller_envs %>% filter(hierarchy > 3 | is.na(hierarchy)) %>% pull(src_ref) %>% n_distinct()
+
+overview_table(
+  r("nb far away caller sites", nb_far_away_caller_sites),
+  r("nb far away caller site percent", ratio(nb_far_away_caller_sites, nb_caller_env_sites))
+)
 ```

### 2021-04-12 paper/main.tex
Explain the various environments wrapped with new

@@
 	\item the empty environment (created with \c{emptyenv()})
+    \item new environments, created with \c{new.env}, \c{list2env} or \c{as.environment}.
 \end{itemize}
@@
 
-The calls happening in a function environment are mostly happening in the environment of the function calling \eval, as shown in Table%TODO
-
-%TODO: add here table of possible frame numbers
-
-Most of the calls happen in the environment surrounding \eval.
+Table~\ref{tab:framenumber} shows the function environments classified according to their position in the call stack: $0$ is the current function, $1$ the parent function and so on.
+
+The calls happening in a function environment are mostly happening in the environment of the function calling \eval.
 This corresponds mostly to the default \c{envir} argument of \eval; indeed, \packageDefautEnvirSitesPercent of the sites have that default argument.
@@
 
-%TODO: add here table of possible environment categories wrapped in new
-
 \begin{table}[h]
@@
         \hline
+        Frame number & \#sites & \%sites \\
+        \hline
+        \packageCallerEnvHierarchyNamea & \packageCallerEnvHierarchySitesaRnd & \packageCallerEnvHierarchySitePercenta \\
+        \packageCallerEnvHierarchyNameb& \packageCallerEnvHierarchySitesbRnd & \packageCallerEnvHierarchySitePercentb \\
+        \packageCallerEnvHierarchyNamec& \packageCallerEnvHierarchySitescRnd & \packageCallerEnvHierarchySitePercentc  \\
+        > 3& \packageNbFarAwayCallerSites &  \packageNbFarAwayCallerSitePercent \\
+        \hline
+    \end{tabular} \label{tab:framenumber}
+\caption{Frame number for the function environments. One site can possibly be passed different kinds of environments.}
+\end{table}
+
+A new environment creation, a parent environment must be specified.
+In Table~\ref{tab:newenvs}, we show what environments the new environment inherits from ultimately.\footnote{It means we directly show the first parent that is different from a new environment.} New environments mostly inherits from function environment: that is actually the default for \c{new.env}, with argument \c{parent = parent.frame()}.
+A large number of new environments are also wrapping package environment or namespaces.
+This is a widespread pattern to store variables to use at the package level.
+When a new environment wraps the empty environment, it can be handily used as a hash table.\footnote{\c{new.env} builds an environment that stores the variable name with a hash table by default.
+}
+% Speak also about new and global?
+
+\begin{table}[h]
+    \centering 
+    \begin{tabular}{|c|c|c|}
+        \hline
         Category & \#sites & \%sites \\
@@
         \packageNewEnvCategoryNamea & \packageNewEnvCategorySitesa & \packageNewEnvCategorySitePercenta \\
-        \hline
         \packageNewEnvCategoryNameb & \packageNewEnvCategorySitesb &  \packageNewEnvCategorySitePercentb\\
-        \hline
         \packageNewEnvCategoryNamec & \packageNewEnvCategorySitesc & \packageNewEnvCategorySitePercentc \\
-        \hline
         \packageNewEnvCategoryNamed & \packageNewEnvCategorySitesd & \packageNewEnvCategorySitePercentd \\
         \hline
-    \end{tabular}
-\caption{Classification of environments wrapped by one of several \c{new}, in number of sites. Some sites can be passed more than one kind of environments. }
+    \end{tabular} \label{tab:newenvs}
+\caption{Classification of environments wrapped by one or several \c{new}, in number of sites. Some sites can be passed more than one kind of environments. }
 \end{table}
@@
 % Up-to-date
+% Probably to remove
 \begin{table}[htbp]{ \centering

### 2021-04-12 analysis/corpus.Rmd
Fixes in S4 and new data

@@
 BASE_packages <- read_lines(BASE_CORPUS_FILE)
-BASE_evals_per_package <- read_csv(BASE_EVALS_STATIC_FILE, col_types = cols(
+BASE_evals_per_package <- read_csv(
+  BASE_EVALS_STATIC_FILE, 
+  col_types = cols(
     package = col_character(),
@@
     args = col_character()
-))
+  )
+)
 
@@
 
-### Matching evals
-
-```{r}
-matched_eval_callsites <-
-  left_join(
-    CRAN_eval_callsites, 
-    count(CRAN_eval_traces, eval_call_srcref), 
-    by=c("srcref"="eval_call_srcref")
-  ) %>%
-  mutate(
-    n=replace_na(n, 0)
-  ) %>%
-  rename(hit=n)
-```
-
-The dynamic analysis traced `r fmt(ratio(filter(matched_eval_callsites, hit>0), matched_eval_callsites))`
-
 ### Plot

### 2021-04-12 analysis/usage.Rmd
Fixes in S4 and new data

@@
   r("Locals", local_calls),
-  r("Triggeredpkgs", length(unique(E$ev_package)))
+  r("Triggeredpkgs", length(unique(E$ev_package))),
+  r("EvalsRatio", ratio(eval_calls, nb_calls)),
+  r("EvalqsRatio", ratio(evalq_calls, nb_calls)),
+  r("EparentsRatio", ratio(eval.parent_calls, nb_calls)),
+  r("LocalsRatio", ratio(local_calls, nb_calls)),
 ) -> TBL

### 2021-04-12 paper/main.tex
Update normalized expressions with last run results

@@
 \c{<-} & \packageMinimizedcallsitesi  & \packageMinimizedpropsitesi & \packageMinimizedpackagei & \packageMinimizedmedianoperationsiRnd & \packageMinimizedpercentparentframesi & \c{x[1, 2:3, 2:3] <- value}\\\hline
-\c{BLOCK} & \packageMinimizedcallsitesj & \packageMinimizedpropsitesj & \packageMinimizedpackagej & \packageMinimizedmedianoperationsjRnd & \packageMinimizedpercentparentframesj & \c{\{ write.csv(iris, tf) ; file.size(tf) \}} \\\hline
+\c{\{\}} & \packageMinimizedcallsitesj & \packageMinimizedpropsitesj & \packageMinimizedpackagej & \packageMinimizedmedianoperationsjRnd & \packageMinimizedpercentparentframesj & \c{\{ write.csv(iris, tf) ; file.size(tf) \}} \\\hline
 \end{tabular}

### 2021-04-12 analysis/corpus.Rmd

@@
       text = element_text(size=17),
-      legend.position = c(0.15, 0.8)
+      legend.position = c(0.2, 0.75)
     )

### 2021-04-12 analysis/side-effects.Rmd

@@
 ---
-title: "Side effects"
-output: html_document
-params:
-  base_dir: /home/aviral/projects/revalstudy2
+title: "Side-effects"
+output:
+    html_document:
+        gallery: false
+        toc: true
+        toc_depth: 3
+        toc_float: true
+        df_print: paged
+date: "`r Sys.Date()`"
 editor_options: 
   chunk_output_type: console
+params:
+  base_dir: Data
 ---
@@
 ```{r setup, include=FALSE}
-for (x in c(
-  "dplyr", 
-  "DT",
-  "fs",
-  "fst",
-  "ggplot2",
-  "ggthemes",
-  "lubridate",
-  "readr",
-  "purrr",
-  "stringr",
-  "tidyr"
-  )) {
-  suppressPackageStartupMessages(library(x, character.only=TRUE))
-}
+knitr::opts_chunk$set(echo = TRUE)
 
-knitr::opts_chunk$set(
-  echo = TRUE,
-  fig.align = "center",
-  fig.retina = 2,
-  fig.width = 10
-)
+library(DT)
+library(digest)
+library(fs)
+library(fst)
+library(purrr)
+library(tidyverse)
 
-# TODO: move into the package
+source("inc/latextags.R")
+source("inc/functions.R")
 source("inc/paths.R")
-source("inc/setup.R")
-source("inc/latextags.R")
 
-options(repos="https://cran.r-project.org")
-theme_set(theme_minimal())
-create_tags(path(TAGS_DIR, "side-effects.tex"), prefix="", default=TRUE)
+create_tags(path(TAGS_DIR, "side-effects.tex"), prefix = "", default = TRUE)
+```
+
+```{r load data}
+SE <- read_fst(PACKAGE_SIDE_EFFECTS_FILE) %>% as_tibble
+```
+
+```{r fix withr::defer callsite}
+
+```
+
+```{r fix S3 methods}
+
 ```
@@
 ```{r}
-core_data <- read_fst(EVALS_SUM_CORE_FILE)
-package_data <- read_fst(EVALS_SUM_PKGS_FILE)
-kaggle_data <- read_fst(EVALS_SUM_KAGGLE_FILE)
-program_data <- read_fst(PROGRAM_FILE)
-
-all_data <- rbind(core_data, package_data, kaggle_data)
+cs_count(SE, eval_call_srcref, wt=N)
+cs_count(SE, variable, wt=N)
 ```
 
-## Side-Effects and Eval
-
-Each entry in `call.fst` contains `direct_writes` and `indirect_writes` (both
-`0` for `eval.parent` and `local`). `direct_writes` are the side-effects
-directly caused by the `eval` in question and `indirect_writes` are the
-side-effects caused by all children evals of the eval. All application
-assignments can be found in `direct_writes` field of `program.fst` and all
-`eval_assignments` can be found in `indirect_writes` field of `program.fst`. All
-non-local assignments can be found by simply summing up the `direct_writes`
-field of all eval calls.
-
-```{r preprocess-write-data}
-eval_writes_data <-
-    all_data %>%
-    filter(eval_function == "eval" | eval_function == "evalq") %>%
-    select(eval_source_type, eval_source, eval_call_srcref, direct_writes, indirect_writes, nb_ev_calls) %>%
-    mutate(all_writes = direct_writes + indirect_writes, side_effecting = as.logical(all_writes))
-
-program_writes_data <-
-    program_data %>%
-    select(direct_writes, indirect_writes)
+```{r}
+overview_table(
+  r("Number of side effects", sum(SE$N))
+)
 ```
 
-```{r compute-side-effecting-write-counts}
-all_writes <- sum(program_writes_data$direct_writes)
-eval_writes <- sum(program_writes_data$indirect_writes)
+## Unknwon srcref
 
-r("All Writes", all_writes)
-r("Eval Writes", eval_writes)
-
-eval_side_effecting_writes <- sum(eval_writes_data$direct_writes * eval_writes_data$nb_ev_calls)
-
-r("Eval Side Effecting Writes", eval_side_effecting_writes)
-
-r("Eval Side Effecting Writes Eval Perc", ratio(eval_side_effecting_writes, eval_writes))
-r("Eval Side Effecting Writes All Perc", ratio(eval_side_effecting_writes, all_writes))
+```{r}
+unknown_srcref <- filter(SE, eval_call_srcref=="???")
 ```
 
-```{r compute-side-effecting-call-counts}
-eval_calls_by_source_writes <-
-    eval_writes_data %>%
-    count(eval_source_type, side_effecting, wt = nb_ev_calls, name = "call_count") %>%
-    group_by(eval_source_type) %>%
-    mutate(total_call_count = sum(call_count)) %>%
-    ungroup() %>%
-    mutate(call_perc = round(100 * call_count/total_call_count, 1)) %>%
-    filter(side_effecting)
+## Where does the side effects happen
 
-datatable(eval_calls_by_source_writes)
+- in the caller - source is `call:fun` where `fun` is the `calle_function` 
+  - most of this should be class that do not specify any parameter to `envir`
+- in the supplied `envir` parameter -- `in_envir` is true
+  - in this case the details are in the `environment_class`
+- in a package - the source is `pkg:`
+- in the `.userHookEnv` - the source is `userHookEnv`
+- in another call
+- is S4, R6
 
-eval_calls_by_source_writes %>%
-pmap(function(eval_source_type, side_effecting, call_count, total_call_count, call_perc) {
-    r(str_c("Side Effecting ", eval_source_type, " call perc"), str_c(call_perc, "%"))
-})
+```{r}
 
-eval_calls_by_package_writes <-
-    eval_writes_data %>%
-    filter(eval_source_type != "core") %>%
-    count(eval_source, side_effecting, wt = nb_ev_calls, name = "call_count") %>%
-    arrange(desc(call_count))
-
-datatable(eval_calls_by_package_writes)
-
-eval_calls_by_source_write_count <-
-    eval_writes_data %>%
-    filter(all_writes != 0) %>%
-    mutate(all_writes = if_else(all_writes <= 5, as.character(all_writes), "> 5")) %>%
-    count(eval_source_type, all_writes, wt = nb_ev_calls, name = "call_count") %>%
-    group_by(eval_source_type) %>%
-    mutate(total_call_count = sum(call_count)) %>%
-    ungroup() %>%
-    mutate(call_perc = round(100 * call_count/total_call_count, 1))
-
-datatable(eval_calls_by_source_write_count)
-
-plot <-
-    eval_calls_by_source_write_count %>%
-    ggplot(aes(factor(all_writes, level = c(1:5, "> 5")), call_perc)) +
-    geom_col(aes(fill = eval_source_type), position = "dodge") +
-    labs(title="", x ="Side-Effects", y = "Side-Effecting Eval Calls %", fill = "") +
-    theme(text = element_text(size=15), axis.text.x = element_text(size=15))
-plot
 ```
-
-The call site computation is incorrect.
-
-```{r compute-side-effecting-site-counts, eval = FALSE}
-eval_sites_by_source_writes <-
-    eval_writes_data %>%
-    group_by(eval_source_type, eval_call_srcref) %>%
-    summarize(side_effecting = any(side_effecting)) %>%
-    ungroup() %>%
-    count(eval_source_type, side_effecting, name = "site_count") %>%
-    group_by(eval_source_type) %>%
-    mutate(total_site_count = sum(site_count)) %>%
-    ungroup() %>%
-    mutate(site_perc = round(100 * site_count / total_site_count, 1)) %>%
-    filter(side_effecting)
-
-datatable(eval_sites_by_source_writes)
-
-eval_sites_by_source_writes %>%
-pmap(function(eval_source_type, side_effecting, site_count, total_site_count, site_perc) {
-    r(str_c("SideEffecting ", eval_source_type, " site perc"), str_c(site_perc, "%"))
-})
-
-eval_sites_by_package_writes <-
-    eval_writes_data %>%
-    filter(eval_source_type != "core") %>%
-    group_by(eval_source, eval_call_srcref) %>%
-    summarize(side_effecting = any(side_effecting)) %>%
-    ungroup() %>%
-    count(eval_source, side_effecting, name = "site_count") %>%
-    group_by(eval_source, side_effecting) %>%
-    mutate(site_perc = round(100 * site_count / sum(site_count), 1)) %>%
-    ungroup() %>%
-    arrange(side_effecting, desc(site_perc))
-
-datatable(eval_sites_by_package_writes)
-```

### 2021-04-12 paper/main.tex

@@
 are bundled with R and executed pervasively. Together they contain
-\BaseEvalCallSites calls to \eval in \BaseFunsWithEvals functions. Some of these
+\BaseEvalCallSites call sites to \eval in \BaseFunsWithEvals functions. Some of these
 functions provide basic functionalities such as package loading and thus there
@@
 \begin{wrapfigure}{4}{5.8cm}
+  \vspace*{-2mm}
 \centering
@@
 
+  \vspace*{-1mm}
   \includegraphics[width=5.6cm, trim=5.5cm 0 5cm 0, clip]{calls_per_run_per_call_site}
@@
 \begin{tabular}{c}
+  \vspace*{-1mm}
   {\hspace{-3cm}\includegraphics[width=0.8\textwidth]{size_loaded_distribution}}
@@
 it unchanged. Expressions account for \packageCodepercent of arguments in our
-corpus. More specifically, \packageSymbolpercent are \c{symbol}s (single
-variables such as \c{x}), \packageLanguagepercent are \c{language} object (a
-single expression, such as \c{x+1}), and \packageExpressionpercent are
-\c{expression} objects (multiple statements). Further inspection reveals that
-most symbols, \packageGgplotsymbolpercent to be exact, come from a single call
-in the \c{ggplot2} package and have have the value \c{\_inherit}. Even if we
-remove this outlier, the majority of \evals are simple expressions.
+corpus. More specifically, \packageSymbolpercent are \texttt{symbol}s (single
+variables such as \c{x}), \packageLanguagepercent are \c{language} objects
+(function calls such as \c{f(x)} or \c{x+1}), and \packageExpressionpercent are
+\c{expression} objects (lists of expressions such as function calls or
+symbols). Further inspection reveals that most symbols,
+\packageGgplotsymbolpercent to be exact, come from a single site in the
+\c{ggplot2} package and have the value \c{\_inherit}.\footnote{This models
+inheritance in \c{ggproto}, one of the many object-oriented systems in
+R that is used exclusively in the \c{ggplot2} graphics library.}
 
@@
 \subsection{Base}
-We recorded \baseAllcalls \eval calls in the base libraries with fewer than 4\%
-of those being to any variants (this was done on a small subset of the whole
-corpus). Unlike CRAN, Base calls \c{evalq} (\baseEvalqs). Out of \baseStaticeval
-call sites, \baseTriggeredevalpct are exercised. Most arguments,
-\baseCodepercent, are expressions. However, unlike CRAN, \baseLanguagepercent of
-the arguments are language objects. The median argument size is
-\baseMedianszeval, which is more than for CRAN. The maximum size is
-\baseMaxszeval. Small instruction counts, less than 50, amount for
-\baseSmalleventspct of calls. The challenge dealing with Base is that every
-program uses it and it can thus generate extreme amounts of data but the data is
-quite predictable. For the rest of the paper, we only mention base when there
-are surprising observations.
+
+We recorded \baseAllcalls \eval calls in the \BasePackages base libraries. This
+comes from running 10\% randomly selected programs from the
+\CranRunnableScriptsRnd extracted programs from CRAN packages, covering
+\baseTriggeredevalpct of the \BaseEvalCallSites sites. Most of the calls
+(\baseEvalsratio) are to the \c{eval} function and most arguments
+(\baseCodepercent) are expressions. However, unlike CRAN, the majority
+(\baseLanguagepercent) of the arguments are language objects. The median
+argument size is \baseMedianszeval, which is more than for CRAN and the maximum
+size is \baseMaxszeval, less than for CRAN. Small instruction counts ($<50$),
+amount for \baseSmalleventspct of calls. A single site in the \c{match.arg}
+function is responsible for \baseTopFuncPercent of all recorded calls. This
+function provides a convenient way for argument verification using partial
+matching. It is heavily relied on in R functions that use string arguments to
+parameterize function behavior, a popular API mechanism in R. For example, in a
+body of \c{center <- function(x, type = c("mean", "median", "trimmed"))} one
+can get the value of \c{type} parameter using \c{type <- match.arg(type)}. A
+call \c{center(x, "trim")} will match \c{type} to \c{"trimmed"}.
+
+The challenge dealing with Base is that every program uses it and it can thus
+generate extreme amounts of data but the data is quite predictable. For the
+rest of the paper, we only mention base when there are surprising
+observations.
 

### 2021-04-13 paper/main.tex
envirnments

@@
 \end{table}
-% Other macros that can be used:
-% - \packageMinimizedMedianOperations: median number of operations
-% - \packageMinimizedPercentEnvir: percentages of sites with the default argument
-% - \packageMinimizedPercentCallsEnvir: percentage of calls with the default envir argument
 
@@
 \packageMinimizedmedianoperationscRnd. The environment in which they evaluate is
-mostly irrelevant (unless it redefines a built-in operator).
-%% Why then this variety of environments??
+mostly irrelevant (unless a built-in operator is redefined).
 
@@
 interpreted operations is \packageMinimizedmedianoperationsaRnd, suggesting that
-it is not the common case.
-
-\newpage
+it is not the common case. Lookups are often evaluated in constructed
+environments, as few as \packageMinimizedpercentparentframesc of these
+expressions are evaluated in a function environment.
+
 \medskip\noindent\EE{$min(e)=\c{\$}$} This form extends \c X to include lookup
@@
 \packageMinimizedmedianoperationsgRnd operations on average; the minimum is 3
-operations.
+operations. This is typically used in a function environment,
+\packageMinimizedpercentparentframese of the time to be precise.
 
@@
 a promise containing a value, then computation will occur. This form occurs in
-6\% of sites.
+6\% of sites and typically does not perform much work in the interpreter.
 
@@
 respectively, \packageMinimizedmedianoperationsdRnd and
-\packageMinimizedmedianoperationsbRnd.
+\packageMinimizedmedianoperationsbRnd. The \packageMinimizedpercentparentframesj
+of these expressions run in function environments.
 
@@
 do not inspect contents of the blocks. The median number of executed operations
-is \packageMinimizedmedianoperationsjRnd.
+is \packageMinimizedmedianoperationsjRnd. They typically run in function
+environments.
 
@@
 single most popular function invoked from \eval, it occurs in 7\% of the sites.
+Each call does quite a lot of work with a 2K instructions median.
 
@@
 computation started by \eval, and the potential reach of its side-effects. The
-second argument to \eval, \c{envir}, defines that environment. It can be
-omitted, in which case its defaults to \c{parent.frame()} which returns the
-caller's environment. Other possible values for \c{envir} include, lists,
-dataframes, pairlists, integers and \c{Null}. Lists and dataframe are coerced to
-new environments, integer values are passed to \c{sys.call} which returns the
-environment of a stack frame at the requested offset.
-
-The top-level environment in R is called the global environment. New
-environments can be created using \c{new.env}, or \c{list2env}. They can be
-provided a parent environment which becomes the enclosing scope of the new
-environment. The new
-environment provides a limited form of sandboxing. All assignments using the
-$\leftarrow$ function occur inside it and prevent the extended environment from
-mutation. However, it is still possible to mutate the extended environment using
-the $\leftarrow$ or \c{assign} functions; but, that happens rarely.
-% as.environment also can be used sometimes for that
-% list2env creates an environment only if its argument envir is not provided
-
-Another kind of environments that is passed to \eval is the environment of a
-package or its \emph{namespace}. A call to
-\c{library(packageName)}, which loads libraries in R, for instance, will
-populate environment \c{package:packageName}. It is a \emph{base} function
-though, so we do not see it in the CRAN dataset. However, R still makes it
-possible to access any package environments and \packageNbPackageEnvSites call
-sites, \ie \packagePackageEnvSitePercent access a package environment, from
-\packageNbPackageEnvPackages different packages. Those packages perform
-introspection, to represent and search any environment by name, in package
-\emph{envnames}, to containerize tests, as with \emph{testthat} and
-\emph{unitizer}, to create a custom OOP system with \emph{R.oo}, or to create
-poor man's hidden functions in \emph{USSCXenaTools}. It is also used to
-implement a custom caching mechanism in packages \emph{g.data} and \emph{SOAR}.
-
-%TODO: there is a difference between a package environment and a namespace. Shortly explain.
-
-Most of the time, \ie in \packageEnvirEnvSitePercent of the sites, \eval is
-passed an actual environment.
-
-It can also be \c{NULL}, a \c{list} or a data frame, or an integer: \eval copies the fields of the
-list or data frame and creates bindings for them in a new environment.
-
-We classify environments to \eval as follows:
+second argument to \eval is used to specify that environment. We propose the
+following high-level classification:
 
 \begin{itemize}
-	\item function environments: the frame of the calling function or one of the parent frames (obtained using \c{parent.frame()}, \c{sys.frame()} or with an integer argument for \c{envir} for instance);
-	\item list environments: a list, a data frame, a tibble, or \c{NULL} (seen as the empty list), passed to \eval;
-	\item the global environment;
-	\item package and namespace environments;
-	\item the empty environment (created with \c{emptyenv()})
-    \item new environments, created with \c{new.env}, \c{list2env} or \c{as.environment}.
+\item \emph{Function:} these environments correspond to some function active on
+  the call stack and are obtained by functions such as \c{parent.frame()} or
+  \c{sys.frame()}.
+\item \emph{Synthetic:} these are environments that built from data structures
+  such as lists, dataframes, etc., or constructed explicitly with \c{new.env},
+  \c{list2env} or \c{as.environment}.
+\item \emph{Global:} the environment in which scripts or interactive commands
+  are evaluated.
+\item \emph{Package:} the environments of a loaded library.
 \end{itemize}
 
-As shown in Table~\ref{tab:highlevelenvironments}, most calls happen in a function environment, with \c{global} as a distant second.
-
-\begin{table}[h]
-	\centering
-	\begin{tabular}{|c|c|c|}
-		\hline
-		Environment & \#sites & \%sites \\
-		\hline
-		Function environments &  \packageNbFunctionEnvSites &  \packageNbFunctionEnvSitePercent\\
+As shown in Table~\ref{tab:highlevelenvironments}, most calls happen in a
+function environment, with \c{global} as a distant second. This means that most
+variable lookups and most side effects will take place amongst the local
+variables of some function. But which function? Table~3 shows the offset from
+the function that directly called \eval, thus 0 is the direct caller and 1 is
+its parent. The data suggests that over 80\% of function environments are the
+direct caller -- this means, from a user's point of view -- the environment of
+the function where \eval occurs textually. It is interesting that some 1.5\% of
+sites evaluate a code in an environment that is three frame or above from the
+caller of \eval. This distance makes it more tricky to reason about the behavior
+of code. As the actions of \eval may happen at a distance.
+
+\begin{table}[h]\centering\small
+\begin{minipage}{4cm}
+  \begin{tabular}{r|r|r}\hline
+ Environment & \#sites & \%sites \\\hline
+ Function & \packageNbFunctionEnvSites &  \packageNbFunctionEnvSitePercent\\
  Global &  \packageNbGlobalEnvSites & \packageNbGlobalEnvSitePercent \\
-		List &  \packageNbEnvListSites & \packageNbEnvListSitePercent \\
-		Package & \packageNbPackageNamespaceEnvSites & \packageNbPackageNamespaceEnvSitePercent \\
-		Empty & \packageNbEmptyEnvSites & \packageNbEmptyEnvSitePercent \\
-		\hline
-		New & \packageNbNewEnvSites & \packageNewEnvSitePercent \\
-		\hline
-	\end{tabular} \label{tab:highlevelenvironments}
-	\caption{Classification of environments per site.}
-\end{table}
-
-Table~\ref{tab:framenumber} shows the function environments classified according to their position in the call stack: $0$ is the current function, $1$ the parent function and so on.
-
-The calls happening in a function environment are mostly happening in the environment of the function calling \eval.
-This corresponds mostly to the default \c{envir} argument of \eval; indeed, \packageDefautEnvirSitesPercent of the sites have that default argument.
-A lot of sites also use \c{parent.frame()} to access their parent frame.
-It corresponds to environment class \c{1} but also to \c{global} when the function is called at the top level.
-Explicitly and directly referring to the global environment using \c{globalenv()} or \c{.GlobalEnv} is rare, as it happens only in \packageNbExplicitGlobalSites sites.
-Accessing further-away frames happens much more seldom.
-
-\begin{table}[h]
-    \centering
-    \begin{tabular}{|c|c|c|}
-        \hline
-        Frame number & \#sites & \%sites \\
-        \hline
+ Synthetic & ? & ? \\
+ Package & \packageNbPackageNamespaceEnvSites & \packageNbPackageNamespaceEnvSitePercent \\\hline
+\end{tabular}
+\caption{Environments per site} \label{tab:highlevelenvironments}
+\end{minipage} \hspace{1cm}
+\begin{minipage}{3.7cm}\centering
+\begin{tabular}{r|r|r}\hline
+ Offset & \#sites & \%sites \\\hline
   \packageCallerEnvHierarchyNamea & \packageCallerEnvHierarchySitesaRnd & \packageCallerEnvHierarchySitePercenta \\
@@
   \packageCallerEnvHierarchyNamec& \packageCallerEnvHierarchySitescRnd & \packageCallerEnvHierarchySitePercentc  \\
-        > 3& \packageNbFarAwayCallerSites &  \packageNbFarAwayCallerSitePercent \\
-        \hline
-    \end{tabular} \label{tab:framenumber}
-\caption{Frame number for the function environments. One site can possibly be passed different kinds of environments.}
+  > 3& \packageNbFarAwayCallerSites &  \packageNbFarAwayCallerSitePercent \\\hline
+ \end{tabular}
+\caption{Function offset}\label{fn}
+\end{minipage}\hspace{1cm}
+\begin{minipage}{3.8cm}
+\begin{tabular}{r|r|r} \hline
+Parent & \#sites & \%sites \\\hline
+Function & \packageNewEnvCategorySitesa & \packageNewEnvCategorySitePercenta \\
+Package & \packageNewEnvCategorySitesb &  \packageNewEnvCategorySitePercentb\\
+Global & \packageNewEnvCategorySitesc & \packageNewEnvCategorySitePercentc \\
+Empty & \packageNewEnvCategorySitesd & \packageNewEnvCategorySitePercentd \\\hline
+\end{tabular} \label{tab:newenvs}
+\caption{Wrapper envs.}
+\end{minipage}
 \end{table}
 
-A new environment creation, a parent environment must be specified.
-In Table~\ref{tab:newenvs}, we show what environments the new environment inherits from ultimately.\footnote{It means we directly show the first parent that is different from a new environment.} New environments mostly inherits from function environment: that is actually the default for \c{new.env}, with argument \c{parent = parent.frame()}.
-A large number of new environments are also wrapping package environment or namespaces.
-This is a widespread pattern to store variables to use at the package level.
-When a new environment wraps the empty environment, it can be handily used as a hash table.\footnote{\c{new.env} builds an environment that stores the variable name with a hash table by default.
-}
-% Speak also about new and global?
-
-\begin{table}[h]
-    \centering 
-    \begin{tabular}{|c|c|c|}
-        \hline
-        Category & \#sites & \%sites \\
-        \hline
-        \packageNewEnvCategoryNamea & \packageNewEnvCategorySitesa & \packageNewEnvCategorySitePercenta \\
-        \packageNewEnvCategoryNameb & \packageNewEnvCategorySitesb &  \packageNewEnvCategorySitePercentb\\
-        \packageNewEnvCategoryNamec & \packageNewEnvCategorySitesc & \packageNewEnvCategorySitePercentc \\
-        \packageNewEnvCategoryNamed & \packageNewEnvCategorySitesd & \packageNewEnvCategorySitePercentd \\
-        \hline
-    \end{tabular} \label{tab:newenvs}
-\caption{Classification of environments wrapped by one or several \c{new}, in number of sites. Some sites can be passed more than one kind of environments. }
-\end{table}
-
-% Up-to-date
-% Probably to remove
-\begin{table}[htbp]{ \centering
-        \begin{tabular}{|c|c|}\hline
-            Environment Class & Sites\\\hline
-            \packageEnvClassNamesa &  \packageEnvClassSitePercenta\\
-            \packageEnvClassNamesb &  \packageEnvClassSitePercentb\\
-            \packageEnvClassNamesc &  \packageEnvClassSitePercentc\\
-            \packageEnvClassNamesd &  \packageEnvClassSitePercentd\\
-            \packageEnvClassNamese &  \packageEnvClassSitePercente\\
-            \packageEnvClassNamesf &  \packageEnvClassSitePercentf\\
-            \packageEnvClassNamesg &  \packageEnvClassSitePercentg\\
-            \packageEnvClassNamesh &  \packageEnvClassSitePercenth\\
-            \hline
-    \end{tabular}} \label{tab:environments}
-    \caption{Environments in terms of sites. One site can be passed different kinds of environments.}
-\end{table}
-
-\subsubsection{Bits of text, possibly reusable}
-
-This pattern is used to evaluate formulas which can directly refer to the fields
-of the data.
-
-We observe that a disproportionately high number of core R \eval calls access
-the caller's environment. This is because many R functions call functions that
-pass the result of \c{parent.frame()} to an \eval. A disproportionately high
-number of calls to \eval happens in an extended top-level environment. This can
-be explained by the fact that many packages evaluate code passed from the user's
-workspace in the top-level environment to access the bindings.
-
-The \c{imchange} function of package \c{imager} makes it possible to modify
-images using a dedicated formula syntax.\footnote{Inspired by {map} in package
-\emph{purr}.} Here, \eval is evaluated in \c{newenv}, which creates a new
-environment that inherits from \c{parent.frame()} by default (classified as 1+).
-
-\begin{lstlisting}
- newenv <- new.env()
- ...
- fo <- parse(text=as.character(fo)[2])
- im[where] <- eval(fo,envir=newenv,enclos=env)
-\end{lstlisting}
-
-\c{adjCoef} in package \emph{actuar} find the root of an equation defined by a
-function \c{h} whose arguments must be named \c{x} and \c{y}. \c{h} is
-transformed into an auxiliary function \c{h2} that can be optimized. Here, the
-list used for \c{envir} ensures the correspondance between the textual arguments
-of \c{h} and the arguments of \c{h2}.
-
-\begin{lstlisting}
- sh <- substitute(h)
- fcall <- paste(sh, "(x, y)")
- ...
- h2 <- function(x, y)
- eval(parse(text = fcall),
- envir = list(x = x, y = y),
- enclos = parent.frame(2))
-\end{lstlisting}
-
-\paragraph{\c{enclos} argument.} The \c{enclos} argument of \eval can be provided when \c{envir} is a pairlist or a dataframe.
-It indicates in which environment to look for elements not present in \c{envir}.
-In our dataset, \c{enclos} is passed explicitly to \eval in \packageEnclosSitePercent of the sites.
-% 85% (\packageEnclosWithListPercent) actually have a list \c{(envir)}. The rest of
-% of them have environment \c{envir}. In that case, \eval ignores \c{enclos} anyway
-
-\todo{That should be moved to the side effect section and updated with the last data.}
-Categorize side-effects between env passed to eval and other envs:
-\begin{itemize}
-\item ephemeral -- read to envs that are created during \eval  XXX033
-\item local -- reads to env of the function in which the call to \eval
-  lexically occurred; XXX034
-\item parent -- reads to parent envs of the local one. XXX035
-\item call-stack -- reads to envs that are on the call stack but not
-  caller. XXX035.
-\end{itemize}
-
-In the corpus we observe \AllWritesRnd writes to variables of which
-\EvalWritesRnd writes happen inside \eval. However, all writes are not
-dangerous. Only writes to environments not local to the computation spawned by
-\eval are side-effecting. These writes outlive the computation and hence are
-visible outside it. The remaining writes are local to the computation. We
-observe that \EvalSideEffectingWritesRnd writes inside evals are side-effecting.
-This is only \EvalSideEffectingWritesEvalPerc of all variable writes inside
-\eval and \EvalSideEffectingWritesAllPerc of all variable writes in the corpus.
-
-An \eval is considered side-effecting if it performs a side-effecting write to a
-variable, directly or indirectly. Only \SideEffectingCoreCallPerc \eval calls in
-Core R are side-effecting and \SideEffectingPackageCallPerc \eval calls in CRAN
-packages are side-effecting.
-
-\subsection{Provenance.}
-% Explain that some generate expression objects and other language objects?
+The use of the global environment is likely split between intentional and
+accidental use. Direct references using \c{globalenv()} or \c{.GlobalEnv} is
+rare, only in \packageNbExplicitGlobalSites sites. We summarize that the
+majority of uses of the global environment come from the fact that most of the
+code in our corpus consists of scripts that are loaded at the top level, thus
+any access of a caller frame may end up executing in the global environment. The
+reason we make this point is that values that are stored in the global
+environment are visible to all functions and are not reclaimed by the garbage
+collector. So accidental uses could result in pollution of that name space.
+
+Synthetic environments must be given a parent. The parent is used to lookup
+variables that are not found the new environment (side effects go to the new
+environment and not to its parent). Table 5 shows the parents of synthetic
+environments. Over half are functions, the second most frequent case is packages.
+??? Say more ????  FIX: so that we get to 100\%
+
+%% The \c{imchange} function of package \c{imager} makes it possible to modify
+%% images using a dedicated formula syntax.\footnote{Inspired by {map} in
+%% package \emph{purr}.} Here, \eval is evaluated in \c{newenv}, which creates a
+%% new environment that inherits from \c{parent.frame()} by default (classified
+%% as 1+).
+%% \begin{lstlisting}
+%%  newenv <- new.env()
+%%  ...
+%%  fo <- parse(text=as.character(fo)[2])
+%%  im[where] <- eval(fo,envir=newenv,enclos=env)
+%% \end{lstlisting}
+
+%% \c{adjCoef} in package \emph{actuar} find the root of an equation defined by
+%% a function \c{h} whose arguments must be named \c{x} and \c{y}. \c{h} is
+%% transformed into an auxiliary function \c{h2} that can be optimized. Here,
+%% the list used for \c{envir} ensures the correspondance between the textual
+%% arguments of \c{h} and the arguments of \c{h2}.
+%% \begin{lstlisting}
+%%  sh <- substitute(h)
+%%  fcall <- paste(sh, "(x, y)")
+%%  ...
+%%  h2 <- function(x, y)
+%%  eval(parse(text = fcall),
+%%  envir = list(x = x, y = y),
+%%  enclos = parent.frame(2))
+%% \end{lstlisting}
+
+\subsection{The origins of \eval}
+
 There are various ways to obtain expressions that \eval will evaluate:
 \begin{itemize}
-    \item \c{substitute} synthesizes ASTs from expressions by replacing
-    symbols with their bindings in the specified environment.
+\item \c{substitute} synthesizes ASTs from expressions by replacing symbols
+    with their bindings in the specified environment.
 \item \c{match.call} captures a call and its arguments as expressions.
-    \item \c{quote}, \c{enquote} returns its argument unevaluated, as an R expression. % Aldd \c{call} ?
-    \item \c{expression}  creates a vector of expression
-    objects from text.
+\item \c{quote}, \c{enquote} returns its argument unevaluated, as an R expression.
+\item \c{expression} creates a vector of expression objects from text.
 \item \c{parse}, \c{str2expression} and \c{str2lang} turn strings into
    expressions.
-    \item \c{comp} compiles R code into R bytecode.
 \end{itemize}
-We show the distribution of the different provenances in Table~\ref{tab:provenance}.
-Note that one site could be passed arguments from different provenances depending on the invokation.
-For some arguments, we could not determine their provenance or it was a value or a symbol, for which the provenance is less significant.
-% Not clear...
-Or maybe add symbols in that table?
-% One can create a symbol with as.name, as.symbol etc. That would be the provenance (but a symbol can also come from quote for instance...)
-
-\begin{table}[h]
-    \begin{tabular}{|c|c|c|}
-        \hline
-        & \#sites & \%sites \\
-        \hline
-        \c{substitute}, \c{quote}, ... & \packageNbOtherLangsSites & \packageOtherLangsSitePercent \\
-        \hline
-        \c{parse}, \c{str2expression} & \packageNbParseExprsSites & \packageParseExprsSitePercent \\
-        \hline
-        \c{match.call} &  \packageNbMatchCallExprsSites & \packageMatchCallExprsSitePercent  \\
-        \hline
-        \c{expression} & \packageNbNotParseExprsSites & \packageNotParseExprsSitePercent  \\
-        \hline
-        \c{str2lang} & \packageNbStrlangSites  & \packageStrlangSitePercent  \\
-        \hline
-        \c{comp} & \packageNbBytecodeSites & \packageBytecodeSitePercent \\
-        \hline \hline
-        Other & \packageNbUnknownProvSites & \packageUnknownProvSitePercent \\
-        \hline
+We show the distribution of the different provenances in
+Table~\ref{tab:provenance}. Note that one site could be passed arguments from
+different provenances depending on the invokation. For some arguments, we could
+not determine their provenance or it was a value or a symbol, for which the
+provenance is less significant.
+
+\begin{table}[h]\small
+\begin{tabular}{r|r|r} \hline
+Origin  & \#sites & \%sites \\\hline
+\c{substitute} & \packageNbOtherLangsSites & \packageOtherLangsSitePercent \\ \hline
+\c{parse} & \packageNbParseExprsSites & \packageParseExprsSitePercent \\ \hline
+\c{match.call} &  \packageNbMatchCallExprsSites & \packageMatchCallExprsSitePercent  \\\hline
+\c{expression} & \packageNbNotParseExprsSites & \packageNotParseExprsSitePercent  \\\hline
+\c{str2lang} & \packageNbStrlangSites  & \packageStrlangSitePercent  \\\hline
+Other & \packageNbUnknownProvSites & \packageUnknownProvSitePercent \\\hline
 \end{tabular}
-    \caption{Provenances of \eval expressions. } \label{tab:provenance}
+\caption{Provenance}\label{tab:provenance}
 \end{table}
 
-We taint the result of a call to \c{parse}, \c{str2lang} and \c{str2expression}, by attaching an R attribute to it.
-R attributes are lost under copy or modification of their attaching object, so we underestimate the number of sites in that case.
-However, we observed that most of the time, the result of \c{parse} is not touched and directly passed to \eval.
-
-We taint the result of \c{match.call} by remembering the addresses of all the elements of the call list.
-As R GC is non-moving, it makes it possible to track them individually, as we observed that calls are modified and individual elements are sometimes evaluated.
-This is also an underestimation of the number of sites though, as we would lose an individual element of the call list that would be modified.
+We taint the result of a call to \c{parse}, \c{str2lang} and \c{str2expression},
+by attaching an R attribute to it. R attributes are lost under copy or
+modification of their attaching object, so we underestimate the number of sites
+in that case. However, we observed that most of the time, the result of
+\c{parse} is not touched and directly passed to \eval.
+
+We taint the result of \c{match.call} by remembering the addresses of all the
+elements of the call list. As R GC is non-moving, it makes it possible to track
+them individually, as we observed that calls are modified and individual
+elements are sometimes evaluated. This is also an underestimation of the number
+of sites though, as we would lose an individual element of the call list that
+would be modified.
 
@@
 
-\c{eval(parse(...))} (and \c{str2expression}, and \c{str2lang}) can be used for dynamic code loading. This forms
-the core of \c{source} and \c{sys.source} functions in R that are
-commonly used for loading code in R files in interactive settings. We
-investigated the number of cases in which the output of \c{parse} and
-its variants is passed to \eval, directly or transitively by tainting
-their output.  We observed that very few of the eval call (\packageNbParseFromFileSites in total)
-consume the result of calling \c{parse} on a file. Most of the eval
-calls consume the result of calling \c{parse} on a string. We also
-identified one function in core R, \c{invokeRestartInteractively} that
-prompts the user for input, parses it, and passes it to \eval.
-
-In Kaggle, most of the sites, \ie \kaggleParseExprsSitePercent, are passed  an expressions that result from a call to \c{parse}.
+\c{eval(parse(...))} (and \c{str2expression}, and \c{str2lang}) can be used for
+dynamic code loading. This forms the core of \c{source} and \c{sys.source}
+functions in R that are commonly used for loading code in R files in interactive
+settings. We investigated the number of cases in which the output of \c{parse}
+and its variants is passed to \eval, directly or transitively by tainting their
+output. We observed that very few of the eval call (\packageNbParseFromFileSites
+in total) consume the result of calling \c{parse} on a file. Most of the eval
+calls consume the result of calling \c{parse} on a string. We also identified
+one function in core R, \c{invokeRestartInteractively} that prompts the user for
+input, parses it, and passes it to \eval.
+
+In Kaggle, most of the sites, \ie \kaggleParseExprsSitePercent, are passed an
+expressions that result from a call to \c{parse}.
 

### 2021-04-13 paper/main.tex
origins

@@
 
-There are various ways to obtain expressions that \eval will evaluate:
+Where does the expression passed to \eval come from? There are various means of
+creating that expression, these are associated with particular use cases.
+
 \begin{itemize}
-\item \c{substitute} synthesizes ASTs from expressions by replacing symbols
-    with their bindings in the specified environment.
-\item \c{match.call} captures a call and its arguments as expressions.
-\item \c{quote}, \c{enquote} returns its argument unevaluated, as an R expression.
-\item \c{expression} creates a vector of expression objects from text.
-\item \c{parse}, \c{str2expression} and \c{str2lang} turn strings into
-   expressions.
+\item {\it Promise:} Function arguments are passed as promises, suspended
+  computations, from a promise one can retrieve the source expression with a
+  call to the \c{substitute} function.
+\item {\it Reflection:} The \c{match.call} function reflectively captures
+  the expression that called the current function.
+\item {\it Constructed:} Expressions can be constructed by invoking the
+  \c{quote}, \c{enquote} and \c{expression} functions.
+\item {\it String:} Finally, expressions can be created from strings by invoking
+  one of \c{parse}, \c{str2expression} and \c{str2lang}.
 \end{itemize}
-We show the distribution of the different provenances in
-Table~\ref{tab:provenance}. Note that one site could be passed arguments from
-different provenances depending on the invokation. For some arguments, we could
-not determine their provenance or it was a value or a symbol, for which the
-provenance is less significant.
+
+Table~\ref{tab:provenance} summarizes the expression provenance in our corpus.
+This data is obtained by dynamically tainting values as they are produced by the
+various sources, due to technical reasons there is some imprecision in the
+results. Manual inspection of numerous examples suggest that ordering is correct.
 
@@
 Origin  & \#sites & \%sites \\\hline
-\c{substitute} & \packageNbOtherLangsSites & \packageOtherLangsSitePercent \\ \hline
-\c{parse} & \packageNbParseExprsSites & \packageParseExprsSitePercent \\ \hline
+Promise & \packageNbOtherLangsSites & \packageOtherLangsSitePercent \\ \hline
+Parse & ?& ?\\ \hline
 \c{match.call} &  \packageNbMatchCallExprsSites & \packageMatchCallExprsSitePercent  \\\hline
-\c{expression} & \packageNbNotParseExprsSites & \packageNotParseExprsSitePercent  \\\hline
-\c{str2lang} & \packageNbStrlangSites  & \packageStrlangSitePercent  \\\hline
-Other & \packageNbUnknownProvSites & \packageUnknownProvSitePercent \\\hline
+Constructed & ? &?  \\\hline
 \end{tabular}
@@
 
-We taint the result of a call to \c{parse}, \c{str2lang} and \c{str2expression},
-by attaching an R attribute to it. R attributes are lost under copy or
-modification of their attaching object, so we underestimate the number of sites
-in that case. However, we observed that most of the time, the result of
-\c{parse} is not touched and directly passed to \eval.
-
-We taint the result of \c{match.call} by remembering the addresses of all the
-elements of the call list. As R GC is non-moving, it makes it possible to track
-them individually, as we observed that calls are modified and individual
-elements are sometimes evaluated. This is also an underestimation of the number
-of sites though, as we would lose an individual element of the call list that
-would be modified.
-
-% explain also how we get the numbers of the other categories?
-
-\c{eval(parse(...))} (and \c{str2expression}, and \c{str2lang}) can be used for
-dynamic code loading. This forms the core of \c{source} and \c{sys.source}
-functions in R that are commonly used for loading code in R files in interactive
-settings. We investigated the number of cases in which the output of \c{parse}
-and its variants is passed to \eval, directly or transitively by tainting their
-output. We observed that very few of the eval call (\packageNbParseFromFileSites
-in total) consume the result of calling \c{parse} on a file. Most of the eval
-calls consume the result of calling \c{parse} on a string. We also identified
-one function in core R, \c{invokeRestartInteractively} that prompts the user for
-input, parses it, and passes it to \eval.
-
-In Kaggle, most of the sites, \ie \kaggleParseExprsSitePercent, are passed an
-expressions that result from a call to \c{parse}.
+The string origin can sometimes be correlated with dynamic code loading. This is
+what the base functions \c{source} and \c{sys.source} do. We observed few calls
+(\packageNbParseFromFileSites in total) that consume the result of calling
+\c{parse} on a file. Most of the calls build strings programmatically. We also
+identified one function \c{invokeRestartInteractively} that prompts the user for
+input, parses it, and passes it to \eval.  The use of strings seems to correlate
+with less sophisticated programmers, as in our Kaggle corpus where
+\kaggleParseExprsSitePercent of site use strings.
 

### 2021-04-13 analysis/normalized.Rmd
Update normalized expressions with last run results
Fix Kaggle usage metrics (top function)

@@
 
+  # We currently merge minimized expressions F(F(X)) and F(F())
+  N <- N %>% mutate(minimized = if_else(minimized == "F(F())", "F(F(X))", minimized))
+
   E_raw %>% select(
@@
 ```{r}
-widespread_minimized <- E_cano %>%
+widespread_minimized <- E_cano %>% filter(type != "bytecode") %>% #Only two sites anyway
   group_by(minimized) %>%

### 2021-04-13 analysis/usage.Rmd
Fix Kaggle usage metrics (top function)

@@
 
-top_func <- frequent_funcs %>% .[1,]
+top_func <- frequent_funcs %>% filter(!is.na(func)) %>% .[1,]
 

### 2021-04-13 analysis/environments.Rmd
Polymorphism in envir

@@
 nb_global_env_sites <- specific_env %>% filter(specific_env == "global") %>% .$n
-nb_callee_env_sites <- specific_env %>% filter(specific_env == "callee") %>% .$n # Seems it is actually for envir == "new.env()"
+nb_callee_env_sites <- specific_env %>% filter(specific_env == "callee") %>% .$n # local, default envir argument
 if(length(nb_callee_env_sites) == 0) nb_callee_env_sites <- 0
@@
 ```{r}
-nb_dataframe_env_sites <- nb_env_list_sites + nb_null_sites
 
-nb_pack_sites <- nb_package_env_sites + nb_namespace_env_sites + nb_base_env_sites
+pack_envs <- env_classes %>% filter(is.na(new_env), is.na(hierarchy), specific_env != "global")
+nb_pack_sites <- n_distinct(pack_envs$src_ref)
 
@@
 
+synthetic_envs <- env_classes %>% filter(envir_type %in% c("list", "NULL") | !is.na(new_env) | specific_env == "empty")
+nb_synthetic_sites <- n_distinct(synthetic_envs$src_ref)
+
+global_envs <- env_classes %>% filter(is.na(new_env), specific_env == "global")
+nb_strict_global_env_sites <- n_distinct(global_envs$src_ref)
+
 overview_table(
-  r("nb dataframe env sites", nb_dataframe_env_sites),
-  r("nb dataframe env site percent", ratio(nb_dataframe_env_sites, nb_call_sites)),
+  r("nb synthetic env sites", nb_synthetic_sites),
+  r("nb synthetic env site percent", ratio(nb_synthetic_sites, nb_call_sites)),
   r("nb package namespace env sites", nb_pack_sites),
@@
   r("nb function env sites", nb_function_env_sites),
-  r("nb function env site percent", ratio(nb_function_env_sites, nb_call_sites))
+  r("nb function env site percent", ratio(nb_function_env_sites, nb_call_sites)),
+  r("nb strict global env sites", nb_strict_global_env_sites),
+  r("nb strict global env site percent", ratio(nb_strict_global_env_sites, nb_call_sites))
 )
+```
+
+Polymorphism? How many sites have envir in several categories?
+
+```{r}
+pack_env_sites <- pack_envs %>% select(src_ref) %>% mutate(category = "package") 
+function_env_sites <- bind_rows(select(caller_envs, src_ref), select(eval_calls_integer, src_ref)) %>% mutate(category = "function")
+synthetic_env_sites <- synthetic_envs %>% select(src_ref) %>% mutate(category = "synthetic")
+global_env_sites <- global_envs %>% select(src_ref)%>% mutate(category = "global")
+
+category_per_site <- bind_rows(pack_env_sites, function_env_sites, synthetic_env_sites, global_env_sites) %>%
+  group_by(src_ref) %>% summarise(n_categories = n_distinct(category))
+
+poly_categories <- category_per_site %>% count(n_categories, sort = TRUE) %>%
+   mutate(pc = ratio(n, nb_call_sites), cpc = cumsum(pc))
+
+one_category <- poly_categories %>% filter(n_categories == 1)
+
+overview_table(
+  r("nb one category envir sites", one_category$n),
+  r("nb one category envir site percent", one_category$pc)
+)
+
 ```

### 2021-04-13 paper/main.tex
Polymorphism in envir

@@
 
-\begin{itemize}
+\begin{compactitem}[---]
 \item \emph{Function:} these environments correspond to some function active on
@@
   such as lists, dataframes, etc., or constructed explicitly with \c{new.env},
-  \c{list2env} or \c{as.environment}.
+  \c{list2env} or \c{as.environment}. It also includes the empty environment.
 \item \emph{Global:} the environment in which scripts or interactive commands
@@
 \item \emph{Package:} the environments of a loaded library.
-\end{itemize}
+\end{compactitem}
 
@@
 variable lookups and most side effects will take place amongst the local
-variables of some function. But which function? Table~3 shows the offset from
+variables of some function. But which function? Table~\ref{tab:funoffset} shows the offset from
 the function that directly called \eval, thus 0 is the direct caller and 1 is
@@
 caller of \eval. This distance makes it more tricky to reason about the behavior
-of code. As the actions of \eval may happen at a distance.
+of code, as the actions of \eval may happen at a distance.
 
@@
  Function & \packageNbFunctionEnvSites &  \packageNbFunctionEnvSitePercent\\
- Global &  \packageNbGlobalEnvSites & \packageNbGlobalEnvSitePercent \\
- Synthetic & ? & ? \\
+ Synthetic & \packageNbSyntheticEnvSites & \packageNbSyntheticEnvSitePercent \\
+ Global &  \packageNbStrictGlobalEnvSites & \packageNbStrictGlobalEnvSitePercent \\
  Package & \packageNbPackageNamespaceEnvSites & \packageNbPackageNamespaceEnvSitePercent \\\hline
@@
  \end{tabular}
-\caption{Function offset}\label{fn}
+\caption{Function offset}\label{tab:funoffset}
 \end{minipage}\hspace{1cm}
@@
 \end{table}
+One site could appear in several categories. However, most sites, \ie \packageNbOneCategoryEnvirSitePercent, have only kind of the above environments passed to them.
 
@@
 
-Synthetic environments must be given a parent. The parent is used to lookup
-variables that are not found the new environment (side effects go to the new
-environment and not to its parent). Table 5 shows the parents of synthetic
+Synthetic environments must be given a parent.
+When the \c{envir} argument is a list or a data frame, \eval also builds a new environment using the \c{enclos} argument as the parent.
+For \c{new.env}, the parent is passed explicitly at construction time.
+The parent is used to lookup
+variables that are not found in the new environment (side effects go to the new
+environment and not to its parent). Table~\ref{tab:newenvs} shows the parents of synthetic
 environments. Over half are functions, the second most frequent case is packages.
 ??? Say more ????  FIX: so that we get to 100\%
+\todo{It is over 100\% because one site can be passed different kinds of environments so it is a percentage over the number of sites in the dataset.
+I could also normalize to 100\% if it is more intuitive.
+In that case, the percentage would be over the sum of the rows (i.e.
+with duplicated sites).}
 
@@
 
-\begin{itemize}
+\begin{compactitem}[---]
 \item {\it Promise:} Function arguments are passed as promises, suspended
@@
   one of \c{parse}, \c{str2expression} and \c{str2lang}.
-\end{itemize}
+\end{compactitem}
 

### 2021-04-13 paper/main.tex
Explain why it is more than 100% for some tables

@@
 
-\begin{table}[h]\centering\small
+\begin{table}[h]
+\centering
+\small
 \begin{minipage}{4cm}
@@
 \end{minipage}
+
+\vspace{-0.3cm} \begin{minipage}{\textwidth} \footnotesize
+One site could appear in several categories. However, most sites, \ie \packageNbOneCategoryEnvirSitePercent, have only kind of the above environments passed to them.
+\end{minipage}
 \end{table}
-One site could appear in several categories. However, most sites, \ie \packageNbOneCategoryEnvirSitePercent, have only kind of the above environments passed to them.
 
@@
 \begin{compactitem}[---]
-\item {\it Promise:} Function arguments are passed as promises, suspended
-  computations, from a promise one can retrieve the source expression with a
-  call to the \c{substitute} function.
 \item {\it Reflection:} The \c{match.call} function reflectively captures
@@
 \item {\it Constructed:} Expressions can be constructed by invoking the
-  \c{quote}, \c{enquote} and \c{expression} functions.
+  \c{quote}, \c{enquote}, \c{expression} and \c{substitute} functions.
+Function arguments are passed as promises, suspended computations, and  \c{substitute} is sometimes used to retrieve the source expression associated to the promise.
 \item {\it String:} Finally, expressions can be created from strings by invoking

### 2021-04-13 analysis/provenance.Rmd
Changes
Fixing environments

@@
 ```
+
+# Another classification
+
+```{r}
+string_exprs <- E_prov %>% filter(!is.na(parsed))
+nb_string_expr_sites <- n_distinct(string_exprs$src_ref)
+
+constructed_exprs <- bind_rows(not_parsed_exprs, other_langs) %>% anti_join(match_call_exprs)
+nb_constructed_expr_sites <- n_distinct(constructed_exprs$src_ref)
+
+unknown_prov_exprs <- E_prov %>% anti_join(string_exprs) %>% anti_join(constructed_exprs) %>% anti_join(match_call_exprs)
+nb_unknown_prov_exprs_sites <- n_distinct(unknown_prov_exprs$src_ref)
+
+overview_table(
+  r("nb constructed sites", nb_constructed_expr_sites),
+  r("nb constructed site percent", ratio(nb_constructed_expr_sites, nb_call_sites)),
+  r("nb string sites", nb_string_expr_sites),
+  r("nb string site percent", ratio(nb_string_expr_sites, nb_call_sites)),
+  r("nb unknown sites", nb_unknown_prov_exprs_sites),
+  r("nb unknown site percent", ratio(nb_unknown_prov_exprs_sites, nb_call_sites)),
+) 
+# last category is match.call but we have it elsewhere
+```
+
+```{r}
+duration <- difftime(Sys.time(), now)
+```
+
+Notebook execution was `r duration` `r units(duration)`.

### 2021-04-13 paper/main.tex
Fixing environments

@@
 Origin  & \#sites & \%sites \\\hline
-Promise & \packageNbOtherLangsSites & \packageOtherLangsSitePercent \\ \hline
-Parse & ?& ?\\ \hline
-\c{match.call} &  \packageNbMatchCallExprsSites & \packageMatchCallExprsSitePercent  \\\hline
-Constructed & ? &?  \\\hline
+Constructed & \packageNbConstructedSites & \packageNbConstructedSitePercent \\ \hline
+String & \packageNbStringSites & \packageNbStringSitePercent \\ \hline
+Reflection &  \packageNbMatchCallExprsSites & \packageMatchCallExprsSitePercent  \\\hline
+Other & \packageNbUnknownSites & \packageNbUnknownSitePercent \\\hline
 \end{tabular}

### 2021-04-13 analysis/environments.Rmd
Environments

@@
 ```{r envir_integer}
-eval_calls_integer <- eval_calls_explicit %>% filter(envir_type == "integer")
+eval_calls_integer <- eval_calls_explicit %>% filter(envir_type %in% c("integer", "double")) 
+# we assume that the doubles are properly casted to integers here
 
@@
 ```{r}
-caller_envs <- env_calls %>% filter(!is.na(hierarchy), is.na(specific_env) | specific_env == "_callee", is.na(new_env))
+caller_envs <- env_calls %>% filter(!is.na(hierarchy), is.na(new_env))
+# Will count also caller-n-specific env
 
@@
 ```{r}
-
-pack_envs <- env_classes %>% filter(is.na(new_env), is.na(hierarchy), specific_env != "global")
+# 4 srcrefs will be counted twice, in package environments, and in caller environments
+pack_envs <- env_classes %>% filter(is.na(new_env),  specific_env != "global")
 nb_pack_sites <- n_distinct(pack_envs$src_ref)
@@
 
-synthetic_envs <- env_classes %>% filter(envir_type %in% c("list", "NULL") | !is.na(new_env) | specific_env == "empty")
+synthetic_envs <- env_classes %>% filter(envir_type %in% c("list", "NULL", "pairlist") | !is.na(new_env) | specific_env == "empty")
 nb_synthetic_sites <- n_distinct(synthetic_envs$src_ref)
@@
 ```{r}
-pack_env_sites <- pack_envs %>% select(src_ref) %>% mutate(category = "package") 
-function_env_sites <- bind_rows(select(caller_envs, src_ref), select(eval_calls_integer, src_ref)) %>% mutate(category = "function")
-synthetic_env_sites <- synthetic_envs %>% select(src_ref) %>% mutate(category = "synthetic")
-global_env_sites <- global_envs %>% select(src_ref)%>% mutate(category = "global")
+pack_env_sites <- pack_envs %>% select(src_ref, duplicates) %>% mutate(category = "package") 
+function_env_sites <- bind_rows(select(caller_envs, src_ref, duplicates), select(eval_calls_integer, src_ref, duplicates)) %>% mutate(category = "function")
+synthetic_env_sites <- synthetic_envs %>% select(src_ref, duplicates) %>% mutate(category = "synthetic")
+global_env_sites <- global_envs %>% select(src_ref,duplicates)%>% mutate(category = "global")
 
-category_per_site <- bind_rows(pack_env_sites, function_env_sites, synthetic_env_sites, global_env_sites) %>%
+env_categories <- bind_rows(pack_env_sites, function_env_sites, synthetic_env_sites, global_env_sites)
+
+category_per_site <- env_categories %>%
   group_by(src_ref) %>% summarise(n_categories = n_distinct(category))
@@
 
-```
-
-# Specific envir with their environment class
-
-For the default argument:
-
-```{r}
-eval_calls_default %>% count(environment_class, wt = duplicates, sort = TRUE) %>%
-  mutate(pc = round(100 * n / sum(n), 2), cpc = cumsum(pc))
-
-eval_calls_default %>% group_by(environment_class) %>% summarise(n = n_distinct(src_ref)) %>%
-  arrange(desc(n)) %>%
-   mutate(pc = round(100 * n / sum(n), 2), cpc = cumsum(pc))
-
-# caller-2- is eval.parent
-eval_calls_default %>% filter(environment_class == "caller-2-") %>% count(ev_variant)
+#TODO: create tags for the poly categories df
 
@@
 
-`eval` is not problematic:
+And what if we classify by using the most present category for site?
+```{r}
+majority_category <- env_categories %>% group_by(src_ref, category) %>% summarise(n_calls = sum(duplicates)) %>%
+  summarize(category = first(category, order_by = order(n_calls, decreasing = TRUE)), n = n(), prop_first = max(n_calls) / sum(n_calls))
 
-```{r}
-eval_calls_default %>% filter(ev_variant == "eval") %>% count(environment_class, wt = duplicates, sort = TRUE)
-```
-
-But `eval.parent` is:
-
-```{r}
-eval_calls_default %>% filter(ev_variant == "eval.parent") %>% count(environment_class, wt = duplicates, sort = TRUE)
-```
-
-It might be because `parent.frame` is equivalent to  `sys.frame(sys.parent(n))` and `sys.parent(n)` can get surprising result of it is called near the top level...
-
-For `parent.frame()` :
-
-```{r}
-E_envir %>% filter(envir == "parent.frame()") %>% count(environment_class, wt = duplicates, sort = TRUE) %>%
-  mutate(pc = round(100 * n / sum(n), 2), cpc = cumsum(pc))
-
-E_envir %>% filter(envir == "parent.frame()") %>% group_by(environment_class) %>%
-  summarise(n = n_distinct(src_ref)) %>%
-  arrange(desc(n)) %>%
-   mutate(pc = round(100 * n / sum(n), 2), cpc = cumsum(pc))
-
-```
-
-For `new.env()`:
-
-```{r}
-E_envir %>% filter(envir == "new.env()") %>% count(environment_class, wt = duplicates, sort = TRUE) %>%
-  mutate(pc = round(100 * n / sum(n), 2), cpc = cumsum(pc))
-
-E_envir %>% filter(envir == "new.env()") %>% group_by(environment_class) %>%
-  summarise(n = n_distinct(src_ref)) %>%
-  arrange(desc(n)) %>%
-   mutate(pc = round(100 * n / sum(n), 2), cpc = cumsum(pc))
-
-```
-
-For each envir expression, what is the most common environment class? (in calls, in sites)
-
-```{r}
-env_class_envir <- E_envir %>% group_by(envir) %>%
-  group_by(environment_class, .add = TRUE) %>% summarise(n = n_distinct(src_ref)) %>%
-  summarise(env_class = first(environment_class, order_by = order(n, decreasing = TRUE)), n = sum(n)) %>% 
-  arrange(desc(n))
+majority_category %>% group_by(category) %>% summarise(n = n_distinct(src_ref)) %>% arrange(desc(n)) %>% mutate(pc = ratio(n, nb_call_sites), cpc = cumsum(pc))
 ```

### 2021-04-13 analysis/environments.Rmd
Number of categories per site

@@
 
-#TODO: create tags for the poly categories df
-
+nrows <- 4
+r_vec("nb category envir", nrows, poly_categories$n_categories)
+r_vec("nb category envir sites", nrows, poly_categories$n)
+r_vec("nb category envir percent", nrows, poly_categories$pc)
+r_vec("nb category envir cumsum percent", nrows, poly_categories$cpc)
 ```
@@
 majority_category <- env_categories %>% group_by(src_ref, category) %>% summarise(n_calls = sum(duplicates)) %>%
-  summarize(category = first(category, order_by = order(n_calls, decreasing = TRUE)), n = n(), prop_first = max(n_calls) / sum(n_calls))
+  summarize(categories = paste(sort(unique(category)), collapse = " ; "), category = first(category, order_by = order(n_calls, decreasing = TRUE)), n = n(), prop_first = max(n_calls) / sum(n_calls), )
 
 majority_category %>% group_by(category) %>% summarise(n = n_distinct(src_ref)) %>% arrange(desc(n)) %>% mutate(pc = ratio(n, nb_call_sites), cpc = cumsum(pc))
+
+majority_category %>% group_by(categories) %>% summarise(n = n_distinct(src_ref)) %>% arrange(desc(n)) %>% mutate(pc = ratio(n, nb_call_sites), cpc = cumsum(pc))
 ```

### 2021-04-13 paper/main.tex
Number of categories per site

@@
 
+One site could appear in several categories.
+However, most sites, \ie \packageNbOneCategoryEnvirSitePercent, have only kind of the above environments passed to them, as shown in Table~\ref{tab:polyenvir}.
+
 \begin{table}[h]
@@
 \end{minipage}
-
-\vspace{-0.3cm} \begin{minipage}{\textwidth} \footnotesize
-One site could appear in several categories. However, most sites, \ie \packageNbOneCategoryEnvirSitePercent, have only kind of the above environments passed to them.
+\begin{minipage}{\textwidth} 
+   \centering
+   \begin{tabular}{c|c|c}
+       \hline
+       \#categories & \#sites &  \%sites \\ \hline
+       \packageNbCategoryEnvira & \packageNbCategoryEnvirSitesaRnd &  \packageNbCategoryEnvirPercenta\\
+       \packageNbCategoryEnvirb &  \packageNbCategoryEnvirSitesbRnd & \packageNbCategoryEnvirPercentb \\
+       \packageNbCategoryEnvirc & \packageNbCategoryEnvirSitescRnd &  \packageNbCategoryEnvirPercentc\\
+       \packageNbCategoryEnvird & \packageNbCategoryEnvirSitesdRnd & \packageNbCategoryEnvirPercentd\\
+       \hline
+   \end{tabular} \label{tab:polyenvir}
+   \caption{Number of categories per site.}
 \end{minipage}

### 2021-04-14 analysis/side-effects.Rmd
Side effects

@@
 library(tidyverse)
+library(xtable)
 
@@
 
-create_tags(path(TAGS_DIR, "side-effects.tex"), prefix = "", default = TRUE)
+create_tags(path(TAGS_DIR, "side-effects.tex"), prefix = "SE", default = TRUE)
 ```
@@
 ```{r load data}
-SE <- read_fst(PACKAGE_SIDE_EFFECTS_FILE) %>% as_tibble
+SE_all <- read_fst(PACKAGE_SIDE_EFFECTS_FILE) %>% 
+  as_tibble %>%
+  filter(transitive==0)
 ```
 
-```{r fix withr::defer callsite}
-
+```{r missing calls}
+SE_missing_eval_call <- filter(SE_all, is.na(eval_function))
+SE_all <- filter(SE_all, !is.na(eval_function))
 ```
 
-```{r fix S3 methods}
+```{r deal with empty srcrefs}
+SE <- mutate(SE_all, eval_call_srcref=case_when(
+  eval_call_srcref == "???" & expr_expression == "handler$expr" & envir_expression == "handler$envir" ~ "::withr::execute_handlers::1",
+  eval_call_srcref == "???" & expr_expression == "quote({\n    old.options <- options(dplyr.summarise.inform = FALSE)\n    o" ~ "::dplyr::tally::1",
+  eval_call_srcref == "???" & expr_expression == "quote({\n    na.action <- attr(mf, \"na.action\")\n    why_omit <- attr(na.a" ~ "::estimatr::cleam_mode_data::1",
+  TRUE ~ eval_call_srcref
+))
 
+SE_missing_sites <- filter(SE, eval_call_srcref=="???")
+SE <- filter(SE, eval_call_srcref!="???")
+```
+
+```{r remove Random.seed}
+SE_random_seed <- filter(SE, variable == ".Random.seed", !str_detect(expr_resolved, "set.seed"))
+SE <- filter(SE, variable != ".Random.seed" | str_detect(expr_resolved, "set.seed"))
+```
+
+```{r remove internal}
+SE_internal <- filter(SE, str_starts(source, fixed("internal:")))
+SE <- filter(SE, !str_starts(source, fixed("internal:")))
+```
+
+```{r load static evals}
+ES <- read_csv(PACKAGE_EVALS_STATIC_FILE) %>% rename(eval_call_srcref=srcref)
+```
+
+```{r load summarize data}
+S <- read_fst(PACKAGE_SUM_CALLS_FILE) %>% as_tibble
+```
+
+```{r}
+CRAN <- read_fst(PACKAGE_CORPUS_FILE) %>% as_tibble
+PKG <- filter(CRAN, in_corpus) 
+```
+
+## Overview
+
+```{r}
+SE_all_N <- sum(SE_all$N)
+SE_all_sites <- count(SE_all, eval_call_srcref)
+SE_all_calls <- count(SE_all, cid)
+
+SE_internal_N <- sum(SE_internal$N)
+SE_internal_sites <- count(SE_internal, eval_call_srcref)
+
+SE_random_seed_N <- sum(SE_random_seed$N)
+SE_random_seed_sites <- count(SE_random_seed, eval_call_srcref)
+
+SE_N <- sum(SE$N)
+SE_calls <- count(SE, cid)
+SE_sites <- count(SE, eval_call_srcref)
+SE_missing_N <- sum(SE_missing_sites$N)
+
+overview_table(
+  r("All", SE_all_N),
+  r("All calls", SE_all_calls),
+  r("All sites", SE_all_sites),
+
+  r("Internal", SE_internal_N),
+  r("Internal ratio", ratio(SE_internal_N, SE_all_N)),
+
+  r("Internal sites", SE_internal_sites),
+  r("Internal sites ratio", ratio(SE_internal_sites, SE_all_sites)),
+
+  r("Random seed", SE_random_seed_N),
+  r("Random seed ratio", ratio(SE_random_seed_N, SE_all_N)),
+
+  r("Random seed sites", SE_random_seed_sites),
+  r("Random seed sites ratio", ratio(count(SE_random_seed, eval_call_srcref), SE_all_sites)),
+
+  r("Missing SE sites", SE_missing_N),
+  r("Missing SE sites ratio", ratio(SE_missing_N, SE_N)),
+
+  r("User", SE_N),
+  r("User ratio", ratio(SE_N, SE_all_N)),
+
+  r("User calls", SE_calls),
+  r("User calls ratio", ratio(SE_calls, SE_all_calls)),
+
+  r("User sites", SE_sites),
+  r("User sites ratio", ratio(SE_sites, SE_all_sites))
+)
+```
+
+## Classification
+
+```{r functions}
+SE <- mutate(
+  SE, 
+  class=case_when(
+    source == str_c("call:", caller_function)           ~ "Local",
+    str_starts(source, fixed("call:"))                  ~ "Function",
+    str_detect(source, "^R6:.*R6.*") | source=="???S4"  ~ "Object",
+    source == "pkg:global"                              ~ "Global",
+    str_starts(source, fixed("pkg:"))                   ~ "Package",
+    TRUE                                                ~ "Synthetic"
+  ),
+  caller=str_replace(eval_call_srcref, "::(.*)::(.*)::.*", "\\1::\\2")
+)
+```
+
+```{r}
+sites_classes <- 
+  SE %>%
+  group_by(eval_call_srcref) %>%
+  summarise(n_class=n_distinct(class), classes=str_c(sort(unique(class)), collapse=", "))
+
+cs_count(sites_classes, n_class)
+cs_count(sites_classes, classes)
+cs_count(SE, class)
+```
+
+```{r}
+funs <- cs_count(SE, caller)
+```
+
+```{r}
+sites_envir <- 
+  SE %>%
+  group_by(eval_call_srcref) %>% 
+  summarize(
+    in_envir=case_when(
+      all(!in_envir) ~ 0,
+      all(in_envir)  ~ 1,
+      TRUE ~ 2
+    )
+  ) %>% 
+  ungroup()
+```
+
+```{r}
+sites_not_in_envir <- filter(sites_envir, in_envir==0)
+sites_in_envir <- filter(sites_envir, in_envir==1)
+sites_both_envir <- filter(sites_envir, in_envir==2)
+
+overview_table(
+  r("Funs nighty", filter(funs, cp <= 90)),
+  r("Funs ten", filter(funs, cp > 90)),
+  r("Sites not in envir", sites_not_in_envir),
+  r("Sites not in envir ratio", ratio(sites_not_in_envir, sites_envir)),
+  r("Sites in envir", sites_in_envir),
+  r("Sites in envir ratio", ratio(sites_in_envir, sites_envir)),
+  r("Sites both envir", sites_both_envir),
+  r("Sites both envir ratio", ratio(sites_both_envir, sites_envir))
+)
+```
+
+```{r}
+top_sites_classes <- cs_count(sites_classes, classes) %>% filter(cp<=91)
+```
+
+```{r}
+print(
+  xtable(
+    mutate(
+      top_sites_classes,
+      p=fmt(p, suffix="%", digit=1),
+      cp=fmt(cp, suffix="%", digit=1),
+    ), 
+    align="llrrr",
+  ),
+  file=file.path(TAGS_DIR, "table-top-sites-classes-se.tex"),
+  comment=FALSE, 
+  include.colnames=FALSE,
+  include.rownames=FALSE,
+  only.contents=TRUE,
+  timestamp=FALSE
+)
+```
+
+## Function classification
+
+```{r}
+single_class_funs <- 
+  left_join(
+    mutate(ES, fun=str_c(package, fun_name)), 
+    sites_classes, 
+    by="eval_call_srcref"
+  ) %>% 
+  group_by(fun) %>% 
+  filter(all(!is.na(n_class))) %>%
+  summarize(n_classes=n_distinct(classes), classes=str_c(unique(classes), collapse=";"))
+
+cs_count(single_class_funs, n_classes)
+cs_count(single_class_funs, classes)
+```
+
+```{r}
+single_class_funs %>%
+  mutate(
+    classes=if_else(str_detect(classes, "[,;]"), "\\emph{multiple}", classes)
+  ) %>%
+  cs_count(classes) %>%
+  mutate(
+    p=fmt(p, suffix="\\%", digit=1),
+    o=if_else(str_starts(classes, fixed("\\")), 0L, n)
+  ) %>% 
+  arrange(desc(o)) %>%
+  select(-cp, -o) %>%
+  xtable(
+    align="llrr",
+  ) %>% 
+  print(
+    file=file.path(TAGS_DIR, "table-top-sites-classes-funs.tex"),
+    comment=FALSE, 
+    include.colnames=FALSE,
+    include.rownames=FALSE,
+    only.contents=TRUE,
+    sanitize.text.function=identity,
+    timestamp=FALSE
+  )
 ```

### 2021-04-14 paper/main.tex
Side effects

@@
 \newcommand{\todo}[1]{\authorcomment{red}{TODO}{#1}}
+
+% cf. https://tex.stackexchange.com/a/144640
+\makeatletter\let\expandableinput\@@input\makeatother
 
@@
 
-\subsection{Eval in R}
+\subsection{Eval in R}\label{sec:eval-in-r}
 
@@
 
-\subsection{The environments of \eval}
+\subsection{The environments of \eval}\label{sec:env}
 
@@
 
-\todo{Filip: discuss the side-effects}
+The code executed by \eval can do side effects. From the compiler perspective,
+we care about the observable side effects in environments---\ie variable
+definitions, updates, and removals that are visible after a call to \eval
+finishes. Knowing where---\ie in which environments do these side-effects
+happen, can help us to determine how much of the compiler knowledge about the
+program will be potentially invalidated.
+
+The \rdyntrace contains low-level hooks capturing the life-cycle of an
+environment and variables defined in it. In the tracer, we register callbacks
+that record each environment creation and removal together with updates to its
+bindings. From the recorded data, we ignore side effects
+coming from \eval sites from unit testing frameworks.\footnote{In the corpus we
+have \c{RUnit, testthat, tinytest} and \c{unitizer} unit testing frameworks.}
+They run the testing code via \eval and thus the results would be biased
+because of the high number of code run in these tests
+(\CranRunnableCodeTestsRnd lines).
+
+From the \packageNbrunsRnd programs we capture \SEAll side effects from
+\SEAllCalls \eval calls in \SEAllSites sites. The challenge is again to remove
+the accidental side-effects that are caused by the R virtual machine
+implementation and are not related to the user code. For example, the
+\c{.Random.seed} variable, which contains the state of the random number
+generator is saved and restored from and to the global environment every time
+user calls one of the base routines for getting random numbers. Removing them
+leaves us with \SEUser side effects (\SEUserRatio) from \SEUserCalls \eval
+calls (\SEUserCallsRatio) in \SEUserSites sites (\SEUserSitesRatio). Half of
+the side-effects comes from three functions: \c{plyr::allocate\_column}
+(allocates space for a new data frame column), \c{withr::execute\_handlers}
+(executes deferred expressions) and \c{foreach::doSEQ} (executes an expression
+on each element in a collection, possibly in parallel). \SEFunsNighty functions
+are responsible for 90\% of side effects. The remaining is spread in \SEFunsTen
+functions.
+
+Most of the \eval sites (\SESitesInEnvirRatio) do side effects in the
+environment specified by the \c{envir} parameter (\cf
+Section~\ref{sec:eval-in-r}). \SESitesNotInEnvirRatio modifies other
+environment and finally \SESitesBothEnvirRatio does both.
+Table~\ref{tab:side-effects-targets} shows the class of the environments where
+\eval side effects happen. It follows the classification described in
+Section~\ref{sec:env} with two extra classes: \emph{Object} which represents a
+side-effect that happens in S4 or R6 object environment, and \emph{Local} that
+corresponds to a side effect in local variables of the \eval caller.
+
+\begin{table}[h]\centering\small
+  \begin{tabular}{l|r|r|r}\hline
+    Environments & \#sites & \%sites & cum. \%sites \\%
+    \expandableinput tag/table-top-sites-classes-se.tex
+  \end{tabular}
+  \caption{Target environments for \eval side-effects} \label{tab:side-effects-targets}
+\end{table}
+
+The good news is that that most \eval sites do all side effects consistently in
+the same environment and that almost half of the side effects happen in either
+\emph{Local} or \emph{Object} environments. These are the cases the compiler
+could possibly deal with.
+\todo{Finish text for Table~\ref{tab:se-funs}}
+
+\begin{table}[h]\centering\small
+  \begin{tabular}{l|r|r}\hline
+    Environment & \#functions & \%functions \\%
+    \expandableinput tag/table-top-sites-classes-funs.tex
+  \end{tabular}
+  \caption{Number of functions in which all \eval side effects happen in the same environment class} \label{tab:se-funs}
+\end{table}
+
+\mypara{Discussion}
+\todo{Is there any comparison with JS?}
 

### 2021-04-14 analysis/inc/functions.R

@@ -39,7 +39,13 @@ classify_exitval <- function(x) {
 }
 
 cs_count <- function(df, ...) {
-  n_rows = nrow(df)
-  count(df, ..., sort=TRUE) %>%
-    mutate(p=n/n_rows*100, cp=cumsum(p))
+  cnt <- count(df, ..., sort=TRUE, name="n")
+  N <- sum(cnt$n)
+  mutate(cnt, p=n/N*100, cp=cumsum(p))
+}
+
+cs_print <- function(df, cp=90) {
+  stopifnot("cp" %in% colnames(df))
+  .N <- cp
+  print(filter(df, cp<=.N), n=Inf)
 }

### 2021-04-14 paper/main.tex
minro

--- 2021-04-13 paper/main.tex
@@
 The environment in which evaluation happens determines what is visible to the
-computation started by \eval, and the potential reach of its side-effects. The
-second argument to \eval is used to specify that environment. We propose the
-following high-level classification:
+computation started by \eval and the potential reach of its side-effects. The
+second argument to \eval specifies that environment. We propose the
+following high-level classification the environments of \eval:
 

### 2021-04-14 paper/main.tex
envirs

@@
 \begin{table}[h]
-\centering
-\small
-\begin{minipage}{4cm}
-  \begin{tabular}{r|r|r}\hline
- Environment & \#sites & \%sites \\\hline
+  \centering\small\hspace{-.5cm}
+\begin{minipage}{3.7cm}
+  \begin{tabular}{@{}r|r|r@{}}\hline
+ Kind & \#sites & \%sites \\\hline
  Function & \packageNbFunctionEnvSites &  \packageNbFunctionEnvSitePercent\\
@@
 \end{tabular}
-\caption{Environments per site} \label{tab:highlevelenvironments}
-\end{minipage} \hspace{1cm}
+\caption{Kinds per site} \label{tab:highlevelenvironments}
+\end{minipage}\hspace{-.2cm}
 \begin{minipage}{3.7cm}\centering
-\begin{tabular}{r|r|r}\hline
+\begin{tabular}{@{}r|r|r@{}}\hline
  Offset & \#sites & \%sites \\\hline
@@
 \caption{Function offset}\label{tab:funoffset}
-\end{minipage}\hspace{1cm}
-\begin{minipage}{3.8cm}
-\begin{tabular}{r|r|r} \hline
+\end{minipage}\hspace{-.2cm}
+\begin{minipage}{3.7cm}
+\begin{tabular}{@{}r|r|r@{}} \hline
 Parent & \#sites & \%sites \\\hline
@@
 \caption{Wrapper envs.}
-\end{minipage}
-\begin{minipage}{\textwidth}
-   \centering
-   \begin{tabular}{c|c|c}
-       \hline
-       \#categories & \#sites &  \%sites \\ \hline
+\end{minipage}\hspace{-.2cm}
+\begin{minipage}{3.7cm}\centering
+   \begin{tabular}{@{}c|c|c@{}} \hline
+       \#kinds & \#sites &  \%sites \\ \hline
        \packageNbCategoryEnvira & \packageNbCategoryEnvirSitesaRnd &  \packageNbCategoryEnvirPercenta\\
@@
    \end{tabular} \label{tab:polyenvir}
-   \caption{Number of categories per site.}
-\end{minipage}
+   \caption{Multiplicities}
+\end{minipage}\hspace{-1cm}
 \end{table}

### 2021-04-14 paper/main.tex
envir

@@
 computation started by \eval and the potential reach of its side-effects. The
-second argument to \eval specifies that environment. We propose the
-following high-level classification the environments of \eval:
+second argument to \eval specifies that environment. Environment of \eval can be
+classified into the following four kinds:
 
 \begin{compactitem}[---]
-\item \emph{Function:} these environments correspond to some function active on
-  the call stack and are obtained by functions such as \c{parent.frame()} or
+\item \emph{Function:} environment for the local variables of some function
+  currently active on the call stack. Obtained by calling \c{parent.frame()} or
   \c{sys.frame()}.
-\item \emph{Synthetic:} these are environments that built from data structures
-  such as lists, dataframes, etc., or constructed explicitly with \c{new.env},
-  \c{list2env} or \c{as.environment}. It also includes the empty environment.
+\item \emph{Synthetic:} environments built from data structures such as lists,
+  dataframes, or constructed explicitly with \c{new.env}, \c{list2env} or
+  \c{as.environment}. Also includes the empty environment.
 \item \emph{Global:} the environment in which scripts or interactive commands
   are evaluated.
-\item \emph{Package:} the environments of a loaded library.
+\item \emph{Package:} the environment of a loaded library.
 \end{compactitem}
 
-As shown in Table~\ref{tab:highlevelenvironments}, most calls happen in a
-function environment, with \c{global} as a distant second. This means that most
-variable lookups and most side effects will take place amongst the local
-variables of some function. But which function? Table~\ref{tab:funoffset} shows the offset from
-the function that directly called \eval, thus 0 is the direct caller and 1 is
-its parent. The data suggests that over 80\% of function environments are the
-direct caller -- this means, from a user's point of view -- the environment of
-the function where \eval occurs textually. It is interesting that some 1.5\% of
-sites evaluate a code in an environment that is three frame or above from the
-caller of \eval. This distance makes it more tricky to reason about the behavior
-of code, as the actions of \eval may happen at a distance.
-
-One site could appear in several categories.
-However, most sites, \ie \packageNbOneCategoryEnvirSitePercent, have only kind of the above environments passed to them, as shown in Table~\ref{tab:polyenvir}.
+\noindent
+As shown by Table~\ref{tab:highlevelenvironments}, most calls evaluate in a
+function environment with the \c{global} kind as a distant second. This means
+that most variable lookups and most side effects either read/update existing
+local variables or introduce new ones. But for which function? From the point of
+view of the function that called \eval, Table~\ref{tab:funoffset} gives an
+offset on the call stack. Thus 0 is the direct caller and 1 is its parent, and
+so on. The data suggests that in 81\% of cases, \eval access its caller's
+environment -- this means, the variables of the function where \eval textually
+occurs are read and written to. It is interesting that some 1.5\% of sites
+evaluate a code in an environment that is three frame or above from \eval. This
+distance implies that, in general, it is modular reasoning is impossible, to
+understand the behavior of any piece of code entails fully understanding the
+behavior of all functions that the code may call as the actions of \eval may
+happen at a distance.
+
+One site could appear in several categories. However, most sites, \ie
+\packageNbOneCategoryEnvirSitePercent, have only kind of the above environments
+passed to them, as shown in Table~\ref{tab:polyenvir}.
 
@@
  \packageNbCategoryEnvirc & \packageNbCategoryEnvirSitescRnd &  \packageNbCategoryEnvirPercentc\\
-       \packageNbCategoryEnvird & \packageNbCategoryEnvirSitesdRnd & \packageNbCategoryEnvirPercentd\\
-       \hline
-   \end{tabular} \label{tab:polyenvir}
-   \caption{Multiplicities}
+ \packageNbCategoryEnvird & \packageNbCategoryEnvirSitesdRnd & \packageNbCategoryEnvirPercentd\\\hline
+\end{tabular}\caption{Multiplicities}\label{tab:polyenvir}
 \end{minipage}\hspace{-1cm}

### 2021-04-14 paper/main.tex
emvirs

@@
 behavior of all functions that the code may call as the actions of \eval may
-happen at a distance.
-
-One site could appear in several categories. However, most sites, \ie
-\packageNbOneCategoryEnvirSitePercent, have only kind of the above environments
-passed to them, as shown in Table~\ref{tab:polyenvir}.
+happen at a distance. Finally note that any particular site may have several
+kind of environments, but in \packageNbOneCategoryEnvirSitePercent of the case a
+site has a single kind.
 
@@
 
-The use of the global environment is likely split between intentional and
-accidental use. Direct references using \c{globalenv()} or \c{.GlobalEnv} is
-rare, only in \packageNbExplicitGlobalSites sites. We summarize that the
-majority of uses of the global environment come from the fact that most of the
-code in our corpus consists of scripts that are loaded at the top level, thus
-any access of a caller frame may end up executing in the global environment. The
-reason we make this point is that values that are stored in the global
-environment are visible to all functions and are not reclaimed by the garbage
-collector. So accidental uses could result in pollution of that name space.
-
-Synthetic environments must be given a parent.
-When the \c{envir} argument is a list or a data frame, \eval also builds a new environment using the \c{enclos} argument as the parent.
-For \c{new.env}, the parent is passed explicitly at construction time.
-The parent is used to lookup
-variables that are not found in the new environment (side effects go to the new
-environment and not to its parent). Table~\ref{tab:newenvs} shows the parents of synthetic
-environments. Over half are functions, the second most frequent case is packages.
-??? Say more ????  FIX: so that we get to 100\%
-\todo{It is over 100\% because one site can be passed different kinds of environments so it is a percentage over the number of sites in the dataset.
-I could also normalize to 100\% if it is more intuitive.
-In that case, the percentage would be over the sum of the rows (i.e.
-with duplicated sites).}
+\noindent
+Sites that evaluate in the global environment are likely split between
+intentional and accidental use. Direct references to the top-level, using
+\c{globalenv()} or \c{.GlobalEnv}, are rare, they occur in only
+\packageNbExplicitGlobalSites sites. We suspect that the majority of uses of the
+global environment come from the fact that our corpus consists of scripts that
+are loaded at the top level, thus that environment may often be the caller of
+\eval or close to it. The reason we make this point is that values stored in the
+global environment are visible to all functions and are not reclaimed by the
+garbage collector. So accidental uses may pollute that name space.
+
+Synthetic environments need a parent. For this, \eval uses its third argument
+(\c{enclos}). When a synthetic environment is created by \eval, it has
+\c{enclos} as a parent. The parent is used to lookup variables not found in its
+child (side effects go to the child only). Table~\ref{tab:newenvs} shows the
+kinds of the parents of synthetic environments. Over half are functions, then
+come packages and global.
 

### 2021-04-14 paper/main.tex
intro/abstract

@@
-\documentclass[screen,acmsmall]{acmart}
+\documentclass[review,nonacm,screen,acmsmall,anonymous=true]{acmart}
 \settopmatter{printfolios=true,printccs=true,printacmref=true}
@@
 \begin{abstract}
-  \noindent Most dynamic languages allow users to turn text into code
-  using various functions, often named \eval, with language-dependent
-  semantics. The widespread use of these reflective functions hinders
-  static analysis and prevents compilers from performing
-  optimizations. This paper aims to provide a better sense of why
-  programmers use \eval. Understanding why \eval is used in practice
-  is key to finding ways to mitigate its negative impact. We have
-  reasons to believe that reflective feature usage is language and
-  application domain specific; we focus on data science code written
-  in R, and compare our results to previous work that analyzed web
-  programming in JavaScript. Dynamic analysis of a corpus of
-  \CranPackages libraries and \KaggleCode end-user code confirms that
-  \eval is indeed in widespread use; R's \eval is more dangerous in
-  some ways, and safer in others, than what was previously reported
-  for JavaScript.
-  %% TODO ::: Add a sentence about summarizing why users resort to
-  %% eval -- Jan
+  \noindent Most dynamic languages allow users to turn text into code using
+  various functions, often named \eval, with language-dependent semantics. The
+  widespread use of these reflective functions hinders static analysis and
+  prevents compilers from performing optimizations. This paper aims to provide a
+  better sense of why programmers use \eval. Understanding why \eval is used in
+  practice is key to finding ways to mitigate its negative impact. We have
+  reasons to believe that reflective feature usage is language and application
+  domain specific; we focus on data science code written in R, and compare our
+  results to previous work that analyzed web programming in JavaScript. This
+  paper studied \CranRunnableScripts scripts extracted from \CranPackages R
+  packages, for a total of \packageAllcalls calls to \eval. We find that \eval
+  is indeed in widespread use; R's \eval is more pervasive and arguably
+  dangerous than what was previously reported for JavaScript.
 \end{abstract}
@@
 end-users, who are typically not expert programmers and often have only a
-cursory knowledge of the language.\footnote{Consider that R evaluates function
-argument lazily, just like Haskell. We informally surveyed end-users, including
-computer scientists, and did not find a single user aware of this fact. Library
-developers, on the other hand, know about laziness and program defensively
-around it.} Our goal is thus to highlight the differences in usage between
-JavaScript and R, and try to explain those differences in terms of language
-features, application domain and programmer experience. Hopefully some of our
-observations will generalize to other languages.
+cursory knowledge of the language.\footnote{Consider that R is lazy like
+Haskell. We informally surveyed end-users and did not find a single user aware
+of this fact. Library developers, on the other hand, know and program
+defensively around laziness.} Our goal is thus to highlight the differences in
+usage between JavaScript and R, and try to explain those differences in terms of
+language features, application domain and programmer experience. Hopefully some
+of our observations will generalize to other languages.
 
@@
 infrastructure:\footnote{Our infrastructure is open source and publicly
-available, our anonymized code is at {\tt http://url.com}, a more complete
-artifact will be submitted to the artifact evaluation committee. } we can
+available and  will be submitted to the artifact evaluation committee. } we can
 monitor R programs by instrumentation --- this gives us access to many

### 2021-04-14 paper/main.tex
origins

@@
 
+\paragraph{Discussion}
+The data presented in this section is what one could expect. ({\it Function})
+The majority of \eval call sites access the environment of some function on the
+call stack. Usually, the current function, but sometimes frames arbitrarily far
+up the call stack. One thing the data does not say is how that environment was
+obtained. The expected case is that the environment was the one obtained from a
+promise; as promises combine code and the environment in which that code
+originates. Likely less frequent are cases where \eval is provided the results
+of programmatically selecting some call stack. ({\it Synthetic}) The relative
+high frequency of synthetic environments corresponds to cases when one wants to
+evaluate an expression in either a restricted environment or use a data
+structure as an environment. ({\it Global}) The fact that global environment
+show up so frequently is likely an artifact of the way the code is run. ({\it
+  Package}) There are only 92 sites that use a package environment in \eval,
+this is probably best as it is bad form to add bindings or modify functions
+of a loaded package.
+
 %% The \c{imchange} function of package \c{imager} makes it possible to modify
@@
 
+\newpage
 \subsection{The origins of \eval}
@@
 Where does the expression passed to \eval come from? There are various means of
-creating that expression, these are associated with particular use cases.
+creating that expression, these are associated with particular use cases, we
+classify them in three categories:
 
 \begin{compactitem}[---]
-\item {\it Reflection:} The \c{match.call} function reflectively captures
-  the expression that called the current function.
+\item {\it Reflection:} This group corresponds to uses of the \c{match.call}
+  function to reflectively capture the expression that invoked the current
+  function.
 \item {\it Constructed:} Expressions can be constructed by invoking the
-  \c{quote}, \c{enquote}, \c{expression} and \c{substitute} functions.
-Function arguments are passed as promises, suspended computations, and  \c{substitute} is sometimes used to retrieve the source expression associated to the promise.
+  \c{quote}, \c{enquote}, \c{expression}. Function arguments are passed as
+  promises and \c{substitute} is used to retrieve the source expression
+  associated to the promise.
 \item {\it String:} Finally, expressions can be created from strings by invoking
-  one of \c{parse}, \c{str2expression} and \c{str2lang}.
+   \c{parse}, \c{str2expression} or \c{str2lang}.
 \end{compactitem}
 
+\noindent
 Table~\ref{tab:provenance} summarizes the expression provenance in our corpus.
@@
 various sources, due to technical reasons there is some imprecision in the
-results. Manual inspection of numerous examples suggest that ordering is correct.
-
-\begin{table}[h]\small
+results. In particular, we are not able to classify all sites. Manual inspection
+of numerous examples suggest that the classified results are accurate.
+
+\begin{wraptable}{r}{5cm}\small\centering
 \begin{tabular}{r|r|r} \hline
 Origin  & \#sites & \%sites \\\hline
-Constructed & \packageNbConstructedSites & \packageNbConstructedSitePercent \\ \hline
-String & \packageNbStringSites & \packageNbStringSitePercent \\ \hline
-Reflection &  \packageNbMatchCallExprsSites & \packageMatchCallExprsSitePercent  \\\hline
-Other & \packageNbUnknownSites & \packageNbUnknownSitePercent \\\hline
+Constructed & \packageNbConstructedSites & \packageNbConstructedSitePercent \\
+String & \packageNbStringSites & \packageNbStringSitePercent \\
+Reflection &  \packageNbMatchCallExprsSites & \packageMatchCallExprsSitePercent\\\hline
 \end{tabular}
 \caption{Provenance}\label{tab:provenance}
-\end{table}
-
-The string origin can sometimes be correlated with dynamic code loading. This is
-what the base functions \c{source} and \c{sys.source} do. We observed few calls
+\end{wraptable}
+
+Strings could correlate with dynamic code loading. This is what the base
+functions \c{source} and \c{sys.source} do. We observed few calls
 (\packageNbParseFromFileSites in total) that consume the result of calling
@@
 input, parses it, and passes it to \eval. The use of strings seems to correlate
-with less sophisticated programmers, as in our Kaggle corpus where
+with less sophisticated programmers; in our Kaggle corpus
 \kaggleParseExprsSitePercent of site use strings.
+
+\paragraph{Discussion}
+The origin data suggests that constructing expressions from strings is a
+minority of the use cases, instead the constructed category shows that the
+majority of \evals comes code that was processed by the compiler, and may be
+slightly modified by the programmer before invoking \eval. Both constructed and
+reflection categories roughly correspond to meta-programming. Some of these use
+cases could likely be replaced by macros if the R designers could be convinced
+to overcome their distaste for those.
 

### 2021-04-14 paper/main.tex
m
Fix typos up to Section 4

@@
 
-%% Aviral Goel
-%% Pierre Donat-Bouillud
-%% Filip Krikava
-%% Jan Vitek
-%% What is the right order?
-
 \author{Author }
 \authornote{Both authors contributed equally to this research.}
-\email{a@b.com}
-\orcid{}
-\author{Author}
-\authornotemark[1]
+\email{a@b.com}\orcid{}\author{Author}\authornotemark[1]
 \email{r@s.com}
@@
 assignments, and bracketing. All functions can be shadowed or redefined, making
-the language at the same time remarkably flexible and exceeding challenging to
+the language at the same time remarkably flexible and exceedingly challenging to
 compile statically as vividly detailed by~\citet{dls19}. R uses a relaxed
@@
 \bf R&\cite{R96}  & expression& programmatic      & data, stack, environment\\\hline
-\end{tabular}\caption{Design space}\label{comp}
+\end{tabular}\caption{Design space of \eval}\label{comp}
 \end{figure}
@@
 to expose as much of the language and its internals as possible in order to
-maximize expressivity. In R, \eval is key tool to extend the language and
+maximize expressivity. In R, \eval is a key tool to extend the language and
 implement DSLs, it is also a replacement for macros. By contrast, the designers
@@
 \c{envir} is the evaluation environment, and \c{encl} is used to look up
-variables not found in \c{envir}. The default value of latter is either the
-caller's environment or top level. \Eval can take several types for \c e, for
+variables not found in \c{envir}. The default value of the latter is either the
+caller's environment or the base environment.\footnote{It contains all the standard operators and functions in R.} \Eval can take several types for \c e, for
 our purposes on the \c{expression}. These can be thought of as the abstract
@@
 is the current environment. Environment chains can be traversed with
-\c{parent.env}, until \c{emptyenv} is reached. The top level environment is
+\c{parent.env}, until \c{emptyenv} is reached. The top-level environment is
 \c{.GlobalEnv}, it has parents that represent the packages that have been
@@
 1,000 Smalltalk projects. While \eval itself is not present, Smalltalk has a
-rich reflective interface. The authors found that reflective are used in less
+rich reflective interface. The authors found that reflective methods are used in less
 than 2\% of methods. The most common reflective method is \c{perform:}; it send
@@
 \mypara{CRAN.} The Comprehensive R Archive Network
-(\small{\url{cran.r-project.org}}) is the largest curated repository of R
+({\small \url{cran.r-project.org}}) is the largest curated repository of R
 packages. It hosts \CranAvailablePackagesRnd packages with 6 new ones submitted
@@
 
-\mypara{Kaggle.} The Kaggle website (\small{\url{kaggle.com}}) is an online
+\mypara{Kaggle.} The Kaggle website ({\small \url{kaggle.com}}) is an online
 platform for data science. It allows users to submit problems and compete to
@@
 runnable, and thus there is one script per kernel. Input data is provided. We
-obtained \KaggleKernels kernels. Since Kaggle is not curated, used SHA-1 hashes
+obtained \KaggleKernels kernels. Since Kaggle is not curated, we used SHA-1 hashes
 to identify and remove \KaggleDuplicates duplicate entries. The remaining
-\KaggleUnique kernels are unique solutions to \KaggleCompetitions competition.
+\KaggleUnique kernels are unique solutions to \KaggleCompetitions competitions.
 They contain \KaggleCode lines of R code. Only \KaggleWithEvals kernels call
@@
   as Rmarkdown or Jupyter, for those we use \c{knitr} to extract runnable code.
-  The body of each extracted program is instrumented with calls our dynamic
+  The body of each extracted program is instrumented with calls to our dynamic
   analyzer to ensure that we only record calls to \eval from the code of
   interest and not from bootstrapping or execution harness operations.
-\item \emph{Trace.} Each program is executed using our dynamic analysis tool
-  which is heavily instrumented interpreter that captures calls to \eval and
+\item \emph{Trace.} Each program is executing using our dynamic analysis tool
+  which is a heavily instrumented interpreter that captures calls to \eval and
   many other fine-grained runtime events. Packages are run twice, once to
@@
 in parallel~\cite{GNUparallel} orchestrated by a Makefile. Servers have
-identical environment thanks to docker images with all dependencies installed.
+identical environments thanks to docker images with all dependencies installed.
 
@@
 The tracer registers callbacks to all \eval functions as well as a few other
-ones that allow us to better identify where do the expressions come from. For
+ones that allow us to better identify where the expressions come from. For
 example, we taint the results of calls to \c{parse} (turns string into
@@
 attach synthetic source code references to all \eval call sites by traversing
-AST. The tracer is implemented as an R package in 3.2K C++ and 1.3K of R code.
+ASTs. The tracer is implemented as an R package in 3.2K C++ and 1.3K of R code.
 For performance reason, most of the tracing is done in C++. While in theory the
@@
 \item In the dynamic analysis, we run all the \emph{runnable} code we obtained
-  from CRAN and Kaggle---\ie real code written by a people with highly varying
-  expertise in R and programming in general. This exercise a lot of the corner
+  from CRAN and Kaggle---\ie real code written by people with highly varying
+  expertise in R and programming in general. This exercises a lot of the corner
   cases of the highly underspecified R behavior.
@@
 The \c{rlang} package introduces a new kind of promise called \c{quosure} which
-is evaluated by \c{tidy\_eval}. None of these limitation should invalidate our
+is evaluated by \c{tidy\_eval}. None of these limitations should invalidate our
 conclusions.
@@
 There are \PkgEvalCallSites \eval sites in \PkgPackages packages. The proportion
-of packages calling \eval is thus \PkgPackagesRatio. Over half of these packages
+of packages calling \eval is \PkgPackagesRatio. Over half of these packages
 have fewer than 3 sites, and with the exception \MaxEvalCallSitesPackage which
@@
 metric for the packages which is \PkgCodeCoverage. The fact that not all sites
-are exercised can be can be chalked down to incomplete tests occasional analysis
+are exercised can be chalked down to incomplete tests and occasional analysis
 failures (\PkgFailedProgramsRatio of programs crashed or timed out).
@@
 50 times or less. Figure~\ref{cn}(b) zooms in on low frequency sites. The x-axis
-shows normalized calls and the y-axis is the number sites for that value. Most
+shows normalized calls and the y-axis is the number of sites for that value. Most
 low-frequency \evals are invoked only once, about half as many are invoked
@@
 
-To estimate how much executable code is injected through \eval, we measure the
-the number of nodes in the expressions; for example, \c{x+1} counts as 3. We
+To estimate how much executable code is injected through \eval, we measure the number of nodes in the expressions; for example, \c{x+1} counts as 3. We
 also measured string lengths of unparsed expressions, but these measurements
@@
 
-To estimate the work performed in \evals, we count instructions executed by
+To estimate the work performed in \evals, we count instructions executed by the
 interpreter. Most invocations perform relatively little work with
@@
 
-The challenge dealing with Base is that every program uses it and it can thus
+The challenge dealing with Base is that every program uses it. It can thus
 generate extreme amounts of data but the data is quite predictable. For the
-rest of the paper, we only mention base when there are surprising
+rest of the paper, we only mention Base when there are surprising
 observations.
@@
 most \evals start with strings. The median argument size is \kaggleMedianszeval,
-Which makes sense, as few arguments are only symbols. The largest argument is
+which makes sense, as few arguments are only symbols. The largest argument is
 \kaggleMaxszeval. The distribution of instructions per eval is similar to CRAN;

### 2021-04-14 paper/main.tex
Lexical scoping vs dynamic scoping

@@
 view of the function that called \eval, Table~\ref{tab:funoffset} gives an
-offset on the call stack. Thus 0 is the direct caller and 1 is its parent, and
-so on. The data suggests that in 81\% of cases, \eval access its caller's
-environment -- this means, the variables of the function where \eval textually
+offset on the call stack. Thus 0 is the direct caller and corresponds to local variables and to \emph{lexical scoping}; 1 is its parent, and
+so on, and correspond to \emph{dynamic scoping}. The data suggests that in 81\% of cases, \eval access its caller's environment -- this means, the variables of the function where \eval textually
 occurs are read and written to. It is interesting that some 1.5\% of sites
 evaluate a code in an environment that is three frame or above from \eval. This
-distance implies that, in general, it is modular reasoning is impossible, to
+distance implies that, in general, modular reasoning is impossible, to
 understand the behavior of any piece of code entails fully understanding the
@@
 happen at a distance. Finally note that any particular site may have several
-kind of environments, but in \packageNbOneCategoryEnvirSitePercent of the case a
+kinds of environments, but in \packageNbOneCategoryEnvirSitePercent of the case a
 site has a single kind.
@@
   \packageCallerEnvHierarchyNamec& \packageCallerEnvHierarchySitescRnd & \packageCallerEnvHierarchySitePercentc  \\
-  > 3& \packageNbFarAwayCallerSites &  \packageNbFarAwayCallerSitePercent \\\hline
+  \ge 3& \packageNbFarAwayCallerSites &  \packageNbFarAwayCallerSitePercent \\\hline
  \end{tabular}
@@
 
-Synthetic environments need a parent. For this, \eval uses its third argument
-(\c{enclos}). When a synthetic environment is created by \eval, it has
-\c{enclos} as a parent. The parent is used to lookup variables not found in its
+Synthetic environments need a parent. It is specified as an argument of the \c{new.env} function, or when \c{envir} is a list or a data frame, \eval uses its third argument
+(\c{enclos}) as the parent.  The parent is used to lookup variables not found in its
 child (side effects go to the child only). Table~\ref{tab:newenvs} shows the
@@
 high frequency of synthetic environments corresponds to cases when one wants to
-evaluate an expression in either a restricted environment or use a data
+evaluate an expression in either a restricted environment, use a data
 structure as an environment. ({\it Global}) The fact that global environment
@@
 this is probably best as it is bad form to add bindings or modify functions
-of a loaded package.
+of a loaded package. Many more sites rather wrap a package environment in a new 
+environment to add variables global the package.
 
@@
 In the R community, the call capturing pattern is often referred~\cite{hadley} as \emph{Non-Standard Evaluation}.
+
+\paragraph{Injecting code in package environments.} R distinguishes two kinds of environments for packages: the \emph{package} environment itself, which contains every publicly accessible functions, and the \emph{namespace} environment, which contains all functions including private ones.
+
+ A call to \c{library(packageName)}, which loads libraries in R, for instance, will
+populate environment \c{package:packageName} using \eval. It is a \emph{base} function
+though, so we do not see it in the CRAN dataset. However, R still makes it
+possible to access any package environments and \packageNbPackageEnvSites call
+sites, \ie \packagePackageEnvSitePercent access a \emph{package} environment, from
+\packageNbPackageEnvPackages different packages. Those packages perform
+introspection, to represent and search any environment by name, in package
+\emph{envnames}, to containerize tests, as with \emph{testthat} and
+\emph{unitizer}, to create a custom OOP system with \emph{R.oo}, or to bypass the visibility of functions in \emph{USSCXenaTools}. It is also used to
+implement a custom caching mechanism in packages \emph{g.data} and \emph{SOAR}.
 

### 2021-04-14 paper/main.tex
Domain Specific Language (usage)

@@
 view of the function that called \eval, Table~\ref{tab:funoffset} gives an
-offset on the call stack. Thus 0 is the direct caller and corresponds to local variables and to \emph{lexical scoping}; 1 is its parent, and
-so on, and correspond to \emph{dynamic scoping}. The data suggests that in 81\% of cases, \eval access its caller's environment -- this means, the variables of the function where \eval textually
+offset on the call stack.
+Thus 0 is the direct caller and corresponds to local variables; 1 is its parent, and so on.
+The data suggests that in 81\% of cases, \eval access its caller's environment -- this means, the variables of the function where \eval textually
 occurs are read and written to. It is interesting that some 1.5\% of sites
@@
 Empty & \packageNewEnvCategorySitesd & \packageNewEnvCategorySitePercentd \\\hline
-\end{tabular} \label{tab:newenvs}
-\caption{Wrapper envs.}
+\end{tabular} 
+\caption{Wrapper envs.} \label{tab:newenvs}
 \end{minipage}\hspace{-.2cm}
@@
 of a loaded package. Many more sites rather wrap a package environment in a new 
-environment to add variables global the package.
+environment to create singletons in a package.
 
@@
 
-\mypara{Symbolic differentiation.} % D and deriv
+\mypara{Domain Specific Language.} R expressions are built and transformed symbolically and then evalled to a concrete value.
+For instance,   base R packages \emph{stats} provides two operators for \emph{symbolic differentiation}, \c{D} and \c{deriv}.
+They support arithmetic operations and functions on real numbers such as \c{sin}.
+
+In \emph{MCMCglmm} (Monte Carl Markov Chain Generalised Linear Mixed Models), a \c{Dtensor} object is created, which contains expressions derived twice (with \c{DD}) and then typically in a list or a data frame.
+\begin{lstlisting}
+ expr<-expression(beta\_1 + time*beta\_2+u)
+ ...
+ mu = data.frame(beta\_1=0.5, beta\_2=1, time=3, u=2.3)
+ ...
+ D[i]<-eval(DD(expr, name[unlist(comb.pos[i,])]), mu)
+\end{lstlisting}
+% See https://github.com/cran/MCMCglmm/blob/cfac9e67ec73a3db2142826faae35e5d2318da31/man/evalDtensor.Rd
 

### 2021-04-14 paper/main.tex
Example about match.call and one argument evaluation

@@
 \begin{lstlisting}
- expr<-expression(beta\_1 + time*beta\_2+u)
+ expr<-expression(beta_1 + time*beta_2+u)
  ...
- mu = data.frame(beta\_1=0.5, beta\_2=1, time=3, u=2.3)
+ mu = data.frame(beta_1=0.5, beta_2=1, time=3, u=2.3)
  ...
@@
 
-A related pattern is not to change the call but rather to evaluate the arguments in the parent frame. This can also be achieved with \c{substitute}.
-% Example
+A related pattern is not to change the call but rather to evaluate the arguments in the parent frame. This could also be achieved with \c{substitute}, but \c{match.call} does it in bulk.
+In package \emph{MedDietCalc},  in function \c{computeMDS95}, \c{match.call} retrieves the argument expressions of the call, which are evaluated in a dataframe \c{data} which contains the numeric data.
+Each argument of function \c{computeMDS95} is also a column in dataframe \c{data}.
+The advantage of using \c{eval} here is to be able to express an argument as a combination of several columns in the dataframe.
+\begin{lstlisting}
+computeMDS95 <- function (data, Vegetables, Legumes, FruitAndNuts, Cereals,
+	 Potatoes = NULL, Meat, Dairy, 	Alcohol, Fats = NULL, MUFA = NULL,
+	  SFA = NULL, Sex, men = "male", women = "female", 	frequency = NULL,
+	  output = "percent", rm.na = FALSE) {
+
+	arguments <- as.list( match.call())
+	Vegetables <- eval(arguments$Vegetables, data)
+	Legumes <- eval(arguments$Legumes, data)
+	FruitAndNuts <- eval(arguments$FruitAndNuts, data)
+	Cereals <- eval(arguments$Cereals, data)
+	Potatoes <- eval(arguments$Potatoes, data)
+	...
+\end{lstlisting}
 

### 2021-04-14 paper/main.tex
Example of match.call and model.frame

@@
 % We have quantitative data on that if needed...
-This pattern is very often (more than half of the cases) used in relation with statistical models, to build a new model.
-\c{model.frame} to build a new generic model, or a more specialized model like \c{glm} or, \c{lm}.
-% Show example
+This pattern is very often (more than half of the cases) used in relation with statistical models, to build a new model, \c{model.frame} to build a new generic model, or a more specialized model like \c{glm} or, \c{lm}.
+The R documentation of \c{match.call} explicitly describes this usage.
+
+In package \emph{survival}\footnote{It is a \emph{recommended package}, so recommended by the CRAN maintainers, thoroughly tested, and following R best practices.}, in function \c{coxph}, the surrounding call is first captured, then \c{model.frame} is injected as the function name in the call, and finally, the new call is evaluated in the parent frame.
+
+\begin{lstlisting}
+Call <- match.call()
+...
+tform <- Call[c(1,indx)]  # only keep the arguments we wanted
+tform[[1L]] <- quote(stats::model.frame)  # change the function called
+...
+mf <- eval(tform, parent.frame())
+\end{lstlisting}
 
@@
 though, so we do not see it in the CRAN dataset. However, R still makes it
-possible to access any package environments and \packageNbPackageEnvSites call
-sites, \ie \packagePackageEnvSitePercent access a \emph{package} environment, from
-\packageNbPackageEnvPackages different packages. Those packages perform
+possible to access any package environments, which we see in
+\packageNbPackageEnvPackages different packages in our corpus. Those packages perform
 introspection, to represent and search any environment by name, in package

### 2021-04-14 paper/main.tex
Fix line numbers in review mode: removing noacm option
minor

--- 2021-04-14 paper/main.tex
@@
 
-\newpage
 \section{Background and Previous work}
@@
 
-\newpage
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-\newpage
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-\newpage
-
 \begin{figure}[tb!]
@@
 
-\newpage
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-\newpage
-
 \subsection{The environments of \eval}\label{sec:env}
@@
 view of the function that called \eval, Table~\ref{tab:funoffset} gives an
-offset on the call stack.
-Thus 0 is the direct caller and corresponds to local variables; 1 is its parent, and so on.
-The data suggests that in 81\% of cases, \eval access its caller's environment -- this means, the variables of the function where \eval textually
+offset on the call stack. Thus 0 is the direct caller and 1 is its parent, and
+so on. The data suggests that in 81\% of cases, \eval access its caller's
+environment -- this means, the variables of the function where \eval textually
 occurs are read and written to. It is interesting that some 1.5\% of sites
 evaluate a code in an environment that is three frame or above from \eval. This
-distance implies that, in general, modular reasoning is impossible, to
+distance implies that, in general, it is modular reasoning is impossible, to
 understand the behavior of any piece of code entails fully understanding the
@@
 happen at a distance. Finally note that any particular site may have several
-kinds of environments, but in \packageNbOneCategoryEnvirSitePercent of the case a
+kind of environments, but in \packageNbOneCategoryEnvirSitePercent of the case a
 site has a single kind.
@@
   \packageCallerEnvHierarchyNamec& \packageCallerEnvHierarchySitescRnd & \packageCallerEnvHierarchySitePercentc  \\
-  \ge 3& \packageNbFarAwayCallerSites &  \packageNbFarAwayCallerSitePercent \\\hline
+  > 3& \packageNbFarAwayCallerSites &  \packageNbFarAwayCallerSitePercent \\\hline
  \end{tabular}
@@
 Empty & \packageNewEnvCategorySitesd & \packageNewEnvCategorySitePercentd \\\hline
-\end{tabular} 
-\caption{Wrapper envs.} \label{tab:newenvs}
+\end{tabular} \label{tab:newenvs}
+\caption{Wrapper envs.}
 \end{minipage}\hspace{-.2cm}
@@
 
-Synthetic environments need a parent. It is specified as an argument of the \c{new.env} function, or when \c{envir} is a list or a data frame, \eval uses its third argument
-(\c{enclos}) as the parent.  The parent is used to lookup variables not found in its
+Synthetic environments need a parent. For this, \eval uses its third argument
+(\c{enclos}). When a synthetic environment is created by \eval, it has
+\c{enclos} as a parent. The parent is used to lookup variables not found in its
 child (side effects go to the child only). Table~\ref{tab:newenvs} shows the
@@
 high frequency of synthetic environments corresponds to cases when one wants to
-evaluate an expression in either a restricted environment, use a data
+evaluate an expression in either a restricted environment or use a data
 structure as an environment. ({\it Global}) The fact that global environment
@@
 this is probably best as it is bad form to add bindings or modify functions
-of a loaded package. Many more sites rather wrap a package environment in a new 
-environment to create singletons in a package.
+of a loaded package.
 
@@
 
-\newpage
 \subsection{The origins of \eval}
@@
 
-\mypara{Domain Specific Language.} R expressions are built and transformed symbolically and then evalled to a concrete value.
-For instance,   base R packages \emph{stats} provides two operators for \emph{symbolic differentiation}, \c{D} and \c{deriv}.
-They support arithmetic operations and functions on real numbers such as \c{sin}.
-
-In \emph{MCMCglmm} (Monte Carl Markov Chain Generalised Linear Mixed Models), a \c{Dtensor} object is created, which contains expressions derived twice (with \c{DD}) and then typically in a list or a data frame.
-\begin{lstlisting}
- expr<-expression(beta\_1 + time*beta\_2+u)
- ...
- mu = data.frame(beta\_1=0.5, beta\_2=1, time=3, u=2.3)
- ...
- D[i]<-eval(DD(expr, name[unlist(comb.pos[i,])]), mu)
-\end{lstlisting}
-% See https://github.com/cran/MCMCglmm/blob/cfac9e67ec73a3db2142826faae35e5d2318da31/man/evalDtensor.Rd
+\mypara{Symbolic differentiation.} % D and deriv
 
@@
 In the R community, the call capturing pattern is often referred~\cite{hadley} as \emph{Non-Standard Evaluation}.
-
-\paragraph{Injecting code in package environments.} R distinguishes two kinds of environments for packages: the \emph{package} environment itself, which contains every publicly accessible functions, and the \emph{namespace} environment, which contains all functions including private ones.
-
- A call to \c{library(packageName)}, which loads libraries in R, for instance, will
-populate environment \c{package:packageName} using \eval. It is a \emph{base} function
-though, so we do not see it in the CRAN dataset. However, R still makes it
-possible to access any package environments and \packageNbPackageEnvSites call
-sites, \ie \packagePackageEnvSitePercent access a \emph{package} environment, from
-\packageNbPackageEnvPackages different packages. Those packages perform
-introspection, to represent and search any environment by name, in package
-\emph{envnames}, to containerize tests, as with \emph{testthat} and
-\emph{unitizer}, to create a custom OOP system with \emph{R.oo}, or to bypass the visibility of functions in \emph{USSCXenaTools}. It is also used to
-implement a custom caching mechanism in packages \emph{g.data} and \emph{SOAR}.
 

### 2021-04-14 paper/main.tex
Explanations about tidyselect

--- 2021-04-14 paper/main.tex
@@
-\documentclass[review,nonacm,screen,acmsmall,anonymous=true]{acmart}
+\documentclass[review,screen,acmsmall,anonymous=true]{acmart} 
+%\documentclass[noacm,review,screen,acmsmall,anonymous=true]{acmart} 
+%Option noacm above messes with the line numbers in review mode
 \settopmatter{printfolios=true,printccs=true,printacmref=true}
@@
 \eval is used in web programming, specifically in websites that use
-JavaScript~\cite{pldi10a}. In 2010, 17 of the largest website used the feature.
+JavaScript~\cite{pldi10a}. In 2010, 17 of the largest websites used the feature.
 In 2011, 82\% of the 10,000 most accessed sites used \eval~\cite{ecoop11}. Yet,
@@
 
-\subsection{The arguments of \eval}
+\subsection{The expression in \eval} \label{sec:minimized}
 
@@
 is the median number of instructions performed by the interpreter; and \%envir
-is the ratio of sites for that evaluate in a function environment. The example
+is the ratio of sites that evaluate in a function environment. The example
 column shows one sample expression $e$ that normalizes to the particular form.
@@
 casual users. \Eval is also being used to reduce boilerplate code and
-provide convenience features for programmers. We now give some
-representative examples of its usage.
-
-\mypara{Variable lookup.} % predominant usage of eval in base
+provide convenience features for programmers. We give examples for the minimized expressions of Section~\ref{sec:minimized} and then dig into high-level \eval design patterns.
+
+\mypara{Variable lookup and values.} 
+This corresponds to minimized expressions \c{X}, \c{V}. We also include here lookup to slots in a list, \ie minimized expression \c{\$}.
+It is the predominant usage of \eval in Base, through function \c{match.args}. In CRAN, it correlates with more exotic environments passed to \eval.
 
@@
 % See https://github.com/cran/MCMCglmm/blob/cfac9e67ec73a3db2142826faae35e5d2318da31/man/evalDtensor.Rd
+
+Package \c{tidyselect} is the backend of several \c{dplyr} and \c{tidyr} functions to select columns.
+It offers a DSL that makes it possible to select a set of columns, over a range of columns, a complement, an intersection, an union, with a regular expression and so on.
+\eval is used to traverse selectors stored as slots of a dataframe.
 

### 2021-04-14 analysis/side-effects.Rmd
Side effects

@@
 
+TODO: fix the srcrefs and caller packages / functions
+
+```{r}
+SE <- mutate(
+  SE,
+  caller=str_replace(eval_call_srcref, "::(.*)::(.*)::.*", "\\1::\\2")
+)
+```
+
 ## Overview
@@
 ```{r}
+S_call <- sum(S$nb_ev_calls)
+S_site <- count(S, eval_call_srcref)
+S_package <- count(S, eval_source)
+S_funs <- sum(PKG$funs_with_eval)
+
 SE_all_N <- sum(SE_all$N)
@@
 SE_sites <- count(SE, eval_call_srcref)
+SE_funs <- count(SE, caller)
+SE_pkgs <- count(SE, caller_package)
 SE_missing_N <- sum(SE_missing_sites$N)
@@
   r("User sites", SE_sites),
-  r("User sites ratio", ratio(SE_sites, SE_all_sites))
+  r("User sites ratio", ratio(SE_sites, SE_all_sites)),
+
+  r("User functions", count(SE, caller)),
+  r("User packages", count(SE, caller_package)),
 )
@@
     TRUE                                                ~ "Synthetic"
-  ),
-  caller=str_replace(eval_call_srcref, "::(.*)::(.*)::.*", "\\1::\\2")
+  )
 )
@@
 ```{r}
-top_sites_classes <- cs_count(sites_classes, classes) %>% filter(cp<=91)
-```
-
-```{r}
-print(
-  xtable(
+sum_sites_classes <- 
+  sites_classes %>%
   mutate(
-      top_sites_classes,
-      p=fmt(p, suffix="%", digit=1),
-      cp=fmt(cp, suffix="%", digit=1),
-    ), 
-    align="llrrr",
-  ),
-  file=file.path(TAGS_DIR, "table-top-sites-classes-se.tex"),
-  comment=FALSE, 
-  include.colnames=FALSE,
-  include.rownames=FALSE,
-  only.contents=TRUE,
-  timestamp=FALSE
-)
-```
-
-## Function classification
-
-```{r}
-single_class_funs <- 
-  left_join(
-    mutate(ES, fun=str_c(package, fun_name)), 
-    sites_classes, 
-    by="eval_call_srcref"
-  ) %>% 
-  group_by(fun) %>% 
-  filter(all(!is.na(n_class))) %>%
-  summarize(n_classes=n_distinct(classes), classes=str_c(unique(classes), collapse=";"))
-
-cs_count(single_class_funs, n_classes)
-cs_count(single_class_funs, classes)
-```
-
-```{r}
-single_class_funs %>%
-  mutate(
-    classes=if_else(str_detect(classes, "[,;]"), "\\emph{multiple}", classes)
+    classes=if_else(str_detect(classes, "[,;]"), "multiple", classes)
   ) %>%
@@
   arrange(desc(o)) %>%
-  select(-cp, -o) %>%
+  select(-cp, -o)
+
+sum_sites_classes %>%
+  mutate(classes=if_else(classes=="multiple", "\\emph{multiple}", classes)) %>%
+  xtable(
+    align="llrr",
+  ) %>%
+  print(
+    file=file.path(TAGS_DIR, "table-top-sites-classes-se.tex"),
+    comment=FALSE, 
+    include.colnames=FALSE,
+    include.rownames=FALSE,
+    only.contents=TRUE,
+    sanitize.text.function=identity,
+    timestamp=FALSE
+  )
+```
+
+```{r}
+sites_class_n <- cs_count(sites_classes, n_class)
+
+overview_table(
+  r("Sites in one class", percent(filter(sites_class_n, n_class==1)$p/100))
+)
+```
+
+## Function classification
+
+```{r}
+funs_srcrefs <- 
+  left_join(
+    mutate(ES, fun=str_c(package, "::", fun_name)), 
+    transmute(count(S, eval_call_srcref), eval_call_srcref, hit=1), 
+    by="eval_call_srcref"
+  ) %>%
+  mutate(hit=replace_na(hit, 0)) %>%
+  select(-args, -call_fun_name) %>%
+  left_join(
+    select(sites_classes, eval_call_srcref, classes),
+    by="eval_call_srcref"
+  )
+
+funs_cov <-
+  funs_srcrefs %>% 
+  group_by(fun) %>%
+  summarize(cov=sum(hit)/n())
+
+funs_classes <- 
+#  semi_join(funs_srcrefs, filter(funs_cov, cov==1), by="fun") %>%
+  semi_join(funs_srcrefs, funs_cov, by="fun") %>%
+  group_by(fun) %>%
+  summarize(
+    cov=sum(hit)/n(),
+    classes=str_c(unique(classes[!is.na(classes)]), collapse=";")
+  )
+
+cs_count(funs_classes, cov)
+cs_count(funs_classes, classes)
+```
+
+```{r}
+sum_funs_classes <-
+funs_classes %>%
+  filter(cov==1, classes != "") %>%
+  mutate(
+    classes=if_else(str_detect(classes, "[,;]"), "multiple", classes)
+  ) %>%
+  cs_count(classes) %>%
+  mutate(
+    p=fmt(p, suffix="\\%", digit=1),
+    o=if_else(str_starts(classes, fixed("\\")), 0L, n)
+  ) %>% 
+  arrange(desc(o)) %>%
+  select(-cp, -o)
+
+sum_funs_classes %>%
+  mutate(classes=if_else(classes=="multiple", "\\emph{multiple}", classes)) %>%
   xtable(
@@
 ```{r}
-cs_count(SE, eval_call_srcref, wt=N)
-cs_count(SE, variable, wt=N)
+left_join(
+  rename(sum_sites_classes, site_n=n, site_p=p),
+  rename(sum_funs_classes, fun_n=n, fun_p=p)
+) %>%
+  mutate(
+    o=if_else(classes=="multiple", 0L, site_n),
+    classes=if_else(classes=="multiple", "\\emph{multiple}", classes),
+    fun_n=replace_na(fun_n, "0"),
+    fun_p=replace_na(fun_p, "0\\%"),
+  ) %>%
+  arrange(desc(o)) %>%
+  select(-o) %>%
+  xtable(
+    align="llrrrr"
+  ) %>%
+  print(
+    file=file.path(TAGS_DIR, "table-se-target-envs.tex"),
+    comment=FALSE, 
+    include.colnames=FALSE,
+    include.rownames=FALSE,
+    only.contents=TRUE,
+    sanitize.text.function=identity,
+    timestamp=FALSE
+  )
 ```
@@
 ```{r}
-overview_table(
-  r("Number of side effects", sum(SE$N))
+event_calls <- cs_count(SE, event, wt=N)
+event_site <- cs_count(SE, eval_call_srcref, event) %>% cs_count(event)
+
+left_join(
+  select(event_calls, event, n_call=n, p_call=p),
+  select(event_site, event, n_site=n, p_site=p),
+  by="event"
+) %>%
+  mutate_if(is.double, ~fmt(., digit=1)) %>%
+  mutate_if(is.integer, ~fmt(.)) %>%
+  xtable(
+    align="llrrrr"
+  ) %>%
+  print(
+    file=file.path(TAGS_DIR, "table-se-types.tex"),
+    comment=FALSE, 
+    include.colnames=FALSE,
+    include.rownames=FALSE,
+    only.contents=TRUE,
+    sanitize.text.function=identity,
+    timestamp=FALSE
   )
+```
+
+```{r}
+filter(SE, event=="rvl") %>% cs_count(class)
 ```

### 2021-04-14 paper/main.tex
typo
Side effects

--- 2021-04-14 paper/main.tex
@@
-\documentclass[review,nonacm,screen,acmsmall,anonymous=true]{acmart}
+\documentclass[review,screen,acmsmall,anonymous=true]{acmart}
+%\documentclass[noacm,review,screen,acmsmall,anonymous=true]{acmart}
+%Option noacm above messes with the line numbers in review mode
 \settopmatter{printfolios=true,printccs=true,printacmref=true}
@@
   \packageCallerEnvHierarchyNamec& \packageCallerEnvHierarchySitescRnd & \packageCallerEnvHierarchySitePercentc  \\
-  \ge 3& \packageNbFarAwayCallerSites &  \packageNbFarAwayCallerSitePercent \\\hline
+$\ge 3$& \packageNbFarAwayCallerSites &  \packageNbFarAwayCallerSitePercent \\\hline
  \end{tabular}
@@
 leaves us with \SEUser side effects (\SEUserRatio) from \SEUserCalls \eval
-calls (\SEUserCallsRatio) in \SEUserSites sites (\SEUserSitesRatio). Half of
-the side-effects comes from three functions: \c{plyr::allocate\_column}
-(allocates space for a new data frame column), \c{withr::execute\_handlers}
-(executes deferred expressions) and \c{foreach::doSEQ} (executes an expression
-on each element in a collection, possibly in parallel). \SEFunsNighty functions
-are responsible for 90\% of side effects. The remaining is spread in \SEFunsTen
-functions.
+calls (\SEUserCallsRatio) in \SEUserSites sites (\SEUserSitesRatio). These
+sites are part of \SEUserFunctions in \SEUserPackages packages. \SEFunsNighty
+functions are responsible for 90\% of side effects. Half of the side-effects
+comes just from three functions: \c{plyr::allocate\_column} (allocates space
+for a new data frame column), \c{withr::execute\_handlers} (executes deferred
+expressions) and \c{foreach::doSEQ} (executes an expression on each element in
+a collection, possibly in parallel).
 
@@
 environment specified by the \c{envir} parameter (\cf
-Section~\ref{sec:eval-in-r}). \SESitesNotInEnvirRatio modifies other
+Section~\ref{sec:eval-in-r}), \SESitesNotInEnvirRatio modifies other
 environment and finally \SESitesBothEnvirRatio does both.
-Table~\ref{tab:side-effects-targets} shows the class of the environments where
-\eval side effects happen. It follows the classification described in
-Section~\ref{sec:env} with two extra classes: \emph{Object} which represents a
-side-effect that happens in S4 or R6 object environment, and \emph{Local} that
-corresponds to a side effect in local variables of the \eval caller.
-
-\begin{table}[h]\centering\small
-  \begin{tabular}{l|r|r|r}\hline
-    Environments & \#sites & \%sites & cum. \%sites \\%
-    \expandableinput tag/table-top-sites-classes-se.tex
+Table~\ref{tab:se-env} shows the class of the environment where \eval side
+effects happen (\emph{target environment}). It follows the classification
+described in Section~\ref{sec:env} with two extra classes: \emph{Object} which
+represents a side-effect that happens in S4 or R6 object environment, and
+\emph{Local} that corresponds to a side effect in local variables of the \eval
+caller. The table shows data for both \eval sites and functions. For a function
+to have a given target environment class means that all of its \eval sites that
+do side effect must have the same target environment class. Majority of the
+\eval sites (\SESitesInOneClass) do all side effects consistently in one
+environment class. The same happens on the function level. Almost half of the
+sites and over a third of the functions do side effects in either \emph{Local}
+or \emph{Object} environments. This gives a ray of hope for the compiler. Even
+though it is possible to do anything anywhere, the data suggest that most
+side-effects are sane.
+
+\begin{table}[h]
+  \small
+  \centering
+  \begin{tabular}{l|r|r|r|r}\hline
+    Environments & \#sites & \%sites & \#funs. & \%funs. \\%
+    \expandableinput tag/table-se-target-envs.tex
   \end{tabular}
-  \caption{Target environments for \eval side-effects} \label{tab:side-effects-targets}
+  \caption{Most used target environments for \eval side-effects} \label{tab:se-env}
 \end{table}
 
-The good news is that that most \eval sites do all side effects consistently in
-the same environment and that almost half of the side effects happen in either
-\emph{Local} or \emph{Object} environments. These are the cases the compiler
-could possibly deal with.
-\todo{Finish text for Table~\ref{tab:se-funs}}
-
-\begin{table}[h]\centering\small
-  \begin{tabular}{l|r|r}\hline
-    Environment & \#functions & \%functions \\%
-    \expandableinput tag/table-top-sites-classes-funs.tex
+Table~\ref{tab:se-types} shows the proportion of the different side effects
+that were recorded. In terms of calls we see mostly assignments and in terms of
+sites definitions. This is expected. A subsequent \eval call will turn a
+definition into an update. Variable removal, while rare, it does happen.
+However, the vast majority comes from the already mentioned
+\c{withr::execute\_handlers} function used to defer evaluation of an expression
+to after the function exit. In our dataset, it is used almost exclusively by
+the \c{tidyselect} package that uses it to remove the reference to current
+quosure environment while interpreting a data frame column
+selectors.\footnote{\cf \url{https://tidyselect.r-lib.org/}}
+
+\begin{table}[h]
+  \small
+  \centering
+  \begin{tabular}{l|r|r|r|r|r|r}\hline
+    Side effect & \#events & \%events & \#calls & \%calls & \#sites & \%sites \\%
+    \expandableinput tag/table-se-types.tex
   \end{tabular}
-  \caption{Number of functions in which all \eval side effects happen in the same environment class} \label{tab:se-funs}
+  \caption{Types of \eval side-effects} \label{tab:se-types}
 \end{table}

### 2021-04-14 paper/main.tex
Changes

@@
 casual users. \Eval is also being used to reduce boilerplate code and
-provide convenience features for programmers. We now give some
-representative examples of its usage.
-
-\mypara{Variable lookup.} % predominant usage of eval in base
+provide convenience features for programmers. We give examples for the minimized expressions of Section~\ref{sec:minimized} and then dig into high-level \eval design patterns.
+
+\mypara{Variable lookup and values.} 
+This corresponds to minimized expressions \c{X}, \c{V}. We also include here lookup to slots in a list or data frame, \ie minimized expression \c{\$}.
+It is the predominant usage of \eval in Base, through function \c{match.args}. In CRAN, it correlates with more exotic environments passed to \eval.
 

### 2021-04-14 paper/main.tex
Example of variable lookups

@@
 
+The dominant site in numbers of calls in CRAN, in the custom object system of \emph{ggplot2}, \c{ggproto}, also evaluates a variable \c{`\_inherit`} in the definition environment of the class.
+It leads to a large diversity of environments, many being package environment or package environments wrapped in \c{new}.
+
+Package \emph{statnet.common} handles formulas, R object that associates an expression of the form \c{x ~ expr} and a carry-on environment where to evaluate the formula.
+It evaluates the left-hand side of the formula, which is always a variable, in \c{eval\_lhs.formula}, as \c{eval(object[[2L]],envir=environment(object))} where \c{object} is the formula and \c{environment} returns the formula associated to the object.
+
+%TODO Add a simple example of a slot lookup (very common with match.call)
+
 \mypara{Domain Specific Language.} R expressions are built and transformed symbolically and then evalled to a concrete value.
@@
 They support arithmetic operations and functions on real numbers such as \c{sin}.
-
 In \emph{MCMCglmm} (Monte Carl Markov Chain Generalised Linear Mixed Models), a \c{Dtensor} object is created, which contains expressions derived twice (with \c{DD}) and then typically in a list or a data frame.

### 2021-04-14 paper/main.tex
Usage

@@
 
+\mypara{Discussion.} % more...
+
+\eval is a powerful tool and can be sometimes with functions with more specialized goals.
+For instance, variable lookup in any environment can be performed with \c{get}.
+Building a call and executing can be done with \c{do.call}.
+Nevertheless, \eval shines when expressions passed to it can be of many different types.
+% And it is shorter to type, and no need to remember all the specialized functions.
+
 \section{Conclusion}

### 2021-04-14 analysis/side-effects.Rmd

@@
 ```{r}
-event_calls <- cs_count(SE, event, wt=N)
+event <- cs_count(SE, event, wt=N)
+event_calls <- cs_count(SE, event)
 event_site <- cs_count(SE, eval_call_srcref, event) %>% cs_count(event)
@@
 left_join(
+  select(event, event, n_event=n, p_event=p),
   select(event_calls, event, n_call=n, p_call=p),
+  by="event"
+) %>%
+  left_join(
     select(event_site, event, n_site=n, p_site=p),
@@
   ) %>% 
-  mutate_if(is.double, ~fmt(., digit=1)) %>%
+  mutate(event=case_when(
+    event=="def" ~ "definition",
+    event=="asn" ~ "update",
+    event=="rvl" ~ "removal"
+  )) %>%
+  mutate_if(is.double, ~fmt(., suffix="%", digit=1)) %>%
   mutate_if(is.integer, ~fmt(.)) %>%
   xtable(
-    align="llrrrr"
+    align="llrrrrrr"
   ) %>%
@@
     only.contents=TRUE,
-    sanitize.text.function=identity,
     timestamp=FALSE
@@
 ```{r}
-filter(SE, event=="rvl") %>% cs_count(class)
+event_rvl_class <- filter(SE, event=="rvl") %>% cs_count(class)
+
+overview_table(
+  r("event rem from synthetic", percent(filter(event_rvl_class, class=="Synthetic")$p/100)),
+  r("event rem from function", percent(filter(event_rvl_class, class=="Function")$p/100)),
+  r("event rem from local", percent(filter(event_rvl_class, class=="Local")$p/100))
+)
 ```
-
-## Unknwon srcref
-
-```{r}
-unknown_srcref <- filter(SE, eval_call_srcref=="???")
-```
-
-## Where does the side effects happen
-
-- in the caller - source is `call:fun` where `fun` is the `calle_function` 
-  - most of this should be class that do not specify any parameter to `envir`
-- in the supplied `envir` parameter -- `in_envir` is true
-  - in this case the details are in the `environment_class`
-- in a package - the source is `pkg:`
-- in the `.userHookEnv` - the source is `userHookEnv`
-- in another call
-- is S4, R6
-
-```{r}
-
-```

### 2021-04-14 analysis/usage.Rmd
Fix figures in Usage metrics

@@
   scale_y_continuous(labels = comma) +
-  labs(x = "nodes") +
+  labs(x = "Nodes", y = "Count") +
   theme(text = element_text(size=17),  panel.grid.major.x = element_blank(),
@@
   geom_histogram(binwidth = 1, fill = "blue", alpha = 0.3, colour = "black") +
+  labs(x = "Average", y = "Count") +
   theme(text = element_text(size=16), 

### 2021-04-14 paper/main.tex
Fix figures in Usage metrics

@@
   \vspace*{-1mm}
-  \includegraphics[width=5.6cm, trim=5.5cm 0 5cm 0, clip]{calls_per_run_per_call_site}
+  \includegraphics[width=5.6cm, trim=5.5cm 0 5cm 0, clip]{package_calls_per_run_per_call_site}
 
@@
   \vspace*{-1mm}
-  {\hspace{-3cm}\includegraphics[width=0.8\textwidth]{size_loaded_distribution}}
+  {\hspace{-3cm}\includegraphics[width=0.8\textwidth]{package_size_loaded_distribution}}
 \end{tabular}
@@
 \begin{minipage}{7.5cm}
- \includegraphics[width=\textwidth]{events_per_pack_small}
+ \includegraphics[width=\textwidth]{package_events_per_pack_small}
 \end{minipage}&\begin{minipage}{7.5cm}
-  \includegraphics[width=\textwidth]{events_per_pack_large}
+  \includegraphics[width=\textwidth]{package_events_per_pack_large}
 \end{minipage}\\[-3mm]

### 2021-04-15 paper/main.tex
Update Kaggle

@@
 \subsection{Kaggle}
-Only \kaggleNbruns scripts trigger \eval. In total, \kaggleAllcalls \eval were
-recorded; none are calls to variants. Out of \kaggleStaticeval call sites,
-\kaggleTriggeredeval were exercised. Most call sites are invoked only once. Only
-one site is called \kaggleMaxcalls times. Expressions amount for
-\kaggleCodepercent of arguments. Unlike CRAN corpus, there are very few symbols
-(\kaggleSymbolpercent). Most expressions result from calls to \c{parse}, thus
-most \evals start with strings. The median argument size is \kaggleMedianszeval,
-which makes sense, as few arguments are only symbols. The largest argument is
-\kaggleMaxszeval. The distribution of instructions per eval is similar to CRAN;
-\kaggleSmalleventspct of \evals execute fewer than 50 instructions. Manual
-inspection of \eval usage in Kaggle suggest that it is consistent with the data
-obtained for CRAN. We do not discuss it further.
+
+In total, \kaggleAllcalls \eval were recorded, all to \eval function. Out of
+\kaggleStaticeval sites from \KaggleWithEvals, only \kaggleTriggeredeval sites
+in \kaggleNbruns scripts were hit. \KaggleFailedScripts scripts failed to run
+and the other did not exercised the \eval sites. This is partially expected
+as Kaggle code does not need to abide to any checks. Upon a manual inspection,
+we observed that indeed the failing scripts were of a poor quality, often not
+finished, using misspelled package names, hard-coded file paths or accessing
+missing files.
+
+Most call sites are invoked only once. Only one site is called \kaggleMaxcalls times.
+Expressions amount for \kaggleCodepercent of arguments. Unlike CRAN corpus,
+there are very few symbols (\kaggleSymbolpercent). Most expressions result from
+calls to \c{parse}, thus most \evals start with strings. The median argument
+size is \kaggleMedianszeval, which makes sense, as few arguments are only
+symbols. The largest argument is \kaggleMaxszeval. The distribution of
+instructions per eval is similar to CRAN; \kaggleSmalleventspct of \evals
+execute fewer than 50 instructions. Manual inspection of \eval usage in Kaggle
+suggest that it is consistent with the data obtained for CRAN. We do not
+discuss it further.
 

### 2021-04-15 analysis/inc/paths.R

@@ -27,6 +27,7 @@ PACKAGE_SIDE_EFFECTS_FILE <- path(PACKAGE_DATA_DIR, "side-effects.fst")
 
 KAGGLE_KERNEL_FILE <- path(KAGGLE_DATA_DIR, "kernel.csv")
 KAGGLE_EVALS_STATIC_FILE <- path(KAGGLE_DATA_DIR, "evals-static.csv")
+KAGGLE_TRACE_LOG_FILE    <- path(KAGGLE_DATA_DIR, "trace-log.csv")
 
 #R_DIR      <- path(params$base_dir, "R-4.0.2")
 #R_LIBS_DIR <- path(params$base_dir, "library", "4.0")

### 2021-04-15 analysis/inc/latextags.R
Fix digits

@@ -308,7 +308,7 @@ fmt <- function(x, prefix="", suffix="", ...) {
 
 # TODO: make this parameter a global option so it can be set globally
 # TODO: test c(0.0123, 1.123, -0.0123, -1.123)
-.fmt.double <- function(x, digits=1, floor=FALSE, ceiling=FALSE) {
+.fmt.double <- function(x, digits=if (x %% 1 == 0) 0 else 1, floor=FALSE, ceiling=FALSE) {
   if (floor) x <- floor(x)
   if (ceiling) x <- ceiling(x)
 
@@ -324,7 +324,7 @@ fmt <- function(x, prefix="", suffix="", ...) {
   formatC(x, big.mark=",", format="f", digits=digits)
 }
 
-.fmt.num_with_suffix <- function(x, floor=FALSE, ceiling=FALSE, digits=1) {
+.fmt.num_with_suffix <- function(x, floor=FALSE, ceiling=FALSE, digits=if (x %% 1 == 0) 0 else 1) {
   suffix <- attr(x, "suffix")
   fmt(as.double(x), suffix=suffix, floor=floor, ceiling=ceiling, digits=digits)
 }

### 2021-04-15 paper/main.tex
format

@@
 \documentclass[review,screen,acmsmall,anonymous=true]{acmart}
-%\documentclass[noacm,review,screen,acmsmall,anonymous=true]{acmart}
-%Option noacm above messes with the line numbers in review mode
-\settopmatter{printfolios=true,printccs=true,printacmref=true}
+\settopmatter{printfolios=false,printccs=false,printacmref=false}
+\setcopyright{none}
+\renewcommand\footnotetextcopyrightpermission[1]{}
+\pagestyle{plain}
 
@@
 variables not found in \c{envir}. The default value of the latter is either the
-caller's environment or the base environment.\footnote{It contains all the standard operators and functions in R.} \Eval can take several types for \c e, for
+caller's environment or the base environment \Eval can take several types for \c e, for
 our purposes on the \c{expression}. These can be thought of as the abstract

### 2021-04-15 paper/main.tex
Example of slot lookup

@@
 %TODO Add a simple example of a slot lookup (very common with match.call)
+Slot lookup happens very often to evaluate some of the elements of a call list returned by \c{match.call}.
+For instance in the \emph{blme} package:
+\begin{lstlisting}
+matchedCall <- match.call()
+...
+df <- eval(matchedCall$df)
+\end{lstlisting}
+
+\mypara{Assigment \c{<-}.}
 
@@
 
-\paragraph{Injecting code in package environments.} R distinguishes two kinds of environments for packages: the \emph{package} environment itself, which contains every publicly accessible functions, and the \emph{namespace} environment, which contains all functions including private ones.
+\mypara{Injecting code in package environments.} R distinguishes two kinds of environments for packages: the \emph{package} environment itself, which contains every publicly accessible functions, and the \emph{namespace} environment, which contains all functions including private ones.
 

### 2021-04-15 paper/main.tex
Explain polymorphism in minimized expressions

@@
 
-%TODO Add a simple example of a slot lookup (very common with match.call)
 Slot lookup happens very often to evaluate some of the elements of a call list returned by \c{match.call}.
-For instance in the \emph{blme} package:
+For instance in the \emph{pwrFDR} package:
 \begin{lstlisting}
-matchedCall <- match.call()
+m <- match.call()
 ...
-df <- eval(matchedCall$df)
+average.power <- m$average.power <- eval(m$average.power, sys.parent())
 \end{lstlisting}
+
+Here, \c{m\$average.power}, depending on the shape of the arguments to the surrounding function,  can be resolved to a variable, a more complex expression, or a number, leading to several possible minimized expressions, including \c{X} and \c{V}.
 

### 2021-04-15 paper/main.tex
Example of assignment: sampling assigments for a MCM run

@@
 
-\mypara{Assigment \c{<-}.}
+\mypara{Assigments.} It corresponds to minimized expression \c{<-} and groups together \c{<-}, \c{<<-} and \c{assign}.
+Package \emph{plyr} has a function combine data frames by row, filling in missing columns.
+During the allocation of the missing columns, an assignment call is built with \c{assignment <- quote(column[rows] <<- what)} and later evaluated on demand in a setter function, with \c{eval(assigmment)}.
+Package \emph{overture} is used to run Markov Chain Monte Carlo.
+It runs an R expression repeatedly and saves any assignments that happen, as samples of the Markov chains.
+The expression is captured with \c{substitute} and a new environment \c{env} wrapping the parent of the caller is created, in which all assignments will happen.
+Regularly, the content of \c{env} is inspected to sample the assigned variables.
+\begin{lstlisting}
+function(expr, overwrite=over.write) {
+	expr_q <- substitute(expr)
+	env <- new.env(parent=parent.frame(1))
+	samps <- list()
+	RunMcmc(samps, 1, expr_q, env, n.save, backing.path, thin, exclude, overwrite)
+}
+
+RunMcmc <- function... {
+	 ...
+	 eval(expr_q, envir=env)
+\end{lstlisting}
 
@@
 \eval is a powerful tool and can be sometimes with functions with more specialized goals.
-For instance, variable lookup in any environment can be performed with \c{get}.
+For instance, variable lookup in any environment can be performed with \c{get}, and assignment in any environment, with \c{assign}.
 Building a call and executing can be done with \c{do.call}.

### 2021-04-15 paper/main.tex
Example of function definition in eval

@@
 
+\mypara{Function definition.} This corresponds to minimized expression \c{FUN}.
+Package \emph{Rcpp} is used to easily interface C++ and R and it uses \eval to bind C++ functions and classes to R objects.
+For instance, \c{.makeCppMethods} create a R function for each C++ method named \c{what} of a given class.
+\begin{lstlisting}
+methods[[what]] <- eval(substitute(
+	function(...) .CppObject$WHAT(...), list(WHAT = as.name(what))),
+	env)
+\end{lstlisting}
+
 \mypara{Domain Specific Language.} R expressions are built and transformed symbolically and then evalled to a concrete value.

### 2021-04-15 paper/main.tex
Correct percentage of V for which eval was actually useful

@@
 where the argument to \eval is a string which \eval simply returns. From twenty
-randomly selected \evals, \packageUsefulValueEvalPercent of them do not need
-\eval. For the other cases, either we are dealing with a value that needs to be
+randomly selected \evals, \packageUsefulValueEvalPercent of them  need % There are a majority which needs eval
+\eval for some inputs. Either we are dealing with a value that needs to be
 constructed dynamically or the value is a default case that sometimes is
@@
 For instance, \c{.makeCppMethods} create a R function for each C++ method named \c{what} of a given class.
+The use of \substitute to inject the name of the method in a language expression entails the use of \eval.
 \begin{lstlisting}
@@
 \end{lstlisting}
+
+\mypara{Block.} %injecting pre nad post processing
 
@@
 
-In the R community, the call capturing pattern is often referred~\cite{hadley} as \emph{Non-Standard Evaluation}.
+In the R community, the call capturing pattern with evaluation in a non-local environment is often referred~\cite{hadley} as \emph{Non-Standard Evaluation}.
 

### 2021-04-15 paper/main.tex
Block
Bypassing CRAN restrictions wiht unlockBinding

@@
 
-\mypara{Block.} %injecting pre nad post processing
+\mypara{Block.} In package \emph{vglm}, \eval is used to inject a pre-processing code at the beginning of a function implementing a statistical model: 
+\begin{lstlisting}
+if (length(slot(family, "first"))) 
+	eval(slot(family, "first"))
+\end{lstlisting}
+
+\mypara{Simple function calls.} This corresponds to minimized expression \c{F()}, \ie calls without nested functions and without variables (but possibly named arguments).
+For instance, package \emph{foreach} provides \c{foreach}-like loops.
+It makes it possible to initialize the iterator variables, with code \c{init}, which is equal \c{quote(list())} by default.
+\c{init} is then used to initialize the \c{foreach} accumulator \c{state\$accum <- eval(combineInfo\$init, envir=extra, enclos=obj\$evalenv)}.
+
+A peculiar use of \eval is to bypass some checks performed to accept code submitted to CRAN.
+For example, bindings can be \emph{locked} in an environment, \ie forbidding its value to be modified.
+The binding can be unlocked though by calling \c{unlockBinding} but packages in CRAN are forbidden to use that function.
+Package \emph{data.table} bypasses this restriction by building a call piece by piece, to unlock binding \c{.N}.
+The call is then evaluated.
+
+\begin{lstlisting}
+if (locked.N) eval(call("unlockBinding", ".N", parent.frame()))  
+\end{lstlisting}
 

### 2021-04-15 paper/main.tex
Update the data.table example

@@
 It is the predominant usage of \eval in Base, through function \c{match.args}. In CRAN, it correlates with more exotic environments passed to \eval.
-
 The dominant site in numbers of calls in CRAN, in the custom object system of \emph{ggplot2}, \c{ggproto}, also evaluates a variable \c{`\_inherit`} in the definition environment of the class.
@@
 The call is then evaluated.
+Now that \c{.N} is unlocked, it can be assigned to.
+Instead of using \eval to assign in the parent frame, they rather use the more specialized function \c{assign}.
 
@@
 if (locked.N) eval(call("unlockBinding", ".N", parent.frame()))  
+assign(".N", nrow(x), envir=parent.frame(), inherits=FALSE)
 \end{lstlisting}

### 2021-04-15 paper/main.tex
Example for F(F(X))

@@
 Now that \c{.N} is unlocked, it can be assigned to.
-Instead of using \eval to assign in the parent frame, they rather use the more specialized function \c{assign}.
+Instead of using \eval to assign in the parent frame, they rather use the more specialized \c{assign} function.
 
@@
 	assign(".N", nrow(x), envir=parent.frame(), inherits=FALSE)
+\end{lstlisting}
+
+\mypara{Complex function calls.} It corresponds to minimized expressions \c{F(X)} and \c{F(F(X))}, \ie function calls with variables, possibly with nested function calls.
+Package \emph{huxtable} is used to create custom tables and export them to various formats.
+To create functions to customize each of the borders of the table, it injects the indices \c{i} and \c{j} of the cells at the border and then evaluate them:
+\begin{lstlisting}
+xxx_border_huxtable <- function(lr_tb, i, j, border_prop) {
+	function (ht) {
+		hickness <- mlist$thickness[eval(i), eval(j), drop = FALSE]
+		...
+	}
+}
+left_border.huxtable <- xxx_border_huxtable("lr_borders",
+i = quote(seq_len(nrow(ht))), j = quote(-last_col))
 \end{lstlisting}

### 2021-04-15 paper/main.tex
Usage of eval

@@
 
+\subsection{Minimized expressions}
+
 \mypara{Variable lookup and values.}
 This corresponds to minimized expressions \c{X}, \c{V}. We also include here lookup to slots in a list or data frame, \ie minimized expression \c{\$}.
-It is the predominant usage of \eval in Base, through function \c{match.args}. In CRAN, it correlates with more exotic environments passed to \eval.
+It is the predominant usage of \eval in Base, through function \c{match.args}. In CRAN, it correlates with more exotic environments passed to \eval. \todo{THis is with a V in practice.}
 The dominant site in numbers of calls in CRAN, in the custom object system of \emph{ggplot2}, \c{ggproto}, also evaluates a variable \c{`\_inherit`} in the definition environment of the class.
-It leads to a large diversity of environments, many being package environment or package environments wrapped in \c{new}.
-
-Package \emph{statnet.common} handles formulas, R object that associates an expression of the form \c{x ~ expr} and a carry-on environment where to evaluate the formula.
+It leads to a large diversity of environments, many being package environments or package environments wrapped in \c{new}.
+\todo{Explain why they use eval and not get... because of \c{::}}
+
+Package \emph{statnet.common} handles formulas, R object that associates an expression of the form \c{x~\textasciitilde~expr} and a carry-on environment where to evaluate the formula.
 It evaluates the left-hand side of the formula, which is always a variable, in \c{eval\_lhs.formula}, as \c{eval(object[[2L]],envir=environment(object))} where \c{object} is the formula and \c{environment} returns the formula associated to the object.
 
+\todo{Show an actual example with \$}
 Slot lookup happens very often to evaluate some of the elements of a call list returned by \c{match.call}.
@@
 \end{lstlisting}
+
+\todo{Add testthat or withr::defer\_handler (fancy on.exit)}
 
@@
 \end{lstlisting}
+
+\subsection{High-level patterns.}
 
@@
 \begin{lstlisting}
-computeMDS95 <- function (data, Vegetables, Legumes, FruitAndNuts, Cereals,
-	 Potatoes = NULL, Meat, Dairy, 	Alcohol, Fats = NULL, MUFA = NULL,
-	  SFA = NULL, Sex, men = "male", women = "female", 	frequency = NULL,
-	  output = "percent", rm.na = FALSE) {
-
+computeMDS95 <- function (data, Vegetables, ...) {
 	arguments <- as.list( match.call())
 	Vegetables <- eval(arguments$Vegetables, data) 
-	Legumes <- eval(arguments$Legumes, data)
-	FruitAndNuts <- eval(arguments$FruitAndNuts, data)
-	Cereals <- eval(arguments$Cereals, data)
-	Potatoes <- eval(arguments$Potatoes, data)
-	...
-\end{lstlisting}
+\end{lstlisting}
+
+\todo{Put a simpler example.}
 
@@
 
-\mypara{Debloating.} \todo{Wouldn't reducing boilerplate be a better name?} \Eval is often used as a means to reduce boilerplate code;
+ Code generation is often used as a means to reduce boilerplate code;
 simple and repetitive code can easily be replaced with judicious use of \eval.
@@
 fact that \eval is executed in a loop.
-
-\mypara{Trivial.} When values are passed to \eval, they are returned
-unchanged. They are an example of trivial uses of \eval. Another
-trivial use is the empty expression, often found in JavaScript, but
-rare in R. \todo{Given that we do not have any data I would get rid of it}
-\todo{We have the normalized expressions V and the small study I did on values passed to eval. And I did not find any eval() (empty) in the dataset. }
-
-% Creating macros?
 

### 2021-04-15 paper/main.tex
Iterate on Usage of eval

@@
 \mypara{Variable lookup and values.}
-This corresponds to minimized expressions \c{X}, \c{V}. We also include here lookup to slots in a list or data frame, \ie minimized expression \c{\$}.
-It is the predominant usage of \eval in Base, through function \c{match.args}. In CRAN, it correlates with more exotic environments passed to \eval. \todo{THis is with a V in practice.}
-The dominant site in numbers of calls in CRAN, in the custom object system of \emph{ggplot2}, \c{ggproto}, also evaluates a variable \c{`\_inherit`} in the definition environment of the class.
-It leads to a large diversity of environments, many being package environments or package environments wrapped in \c{new}.
-\todo{Explain why they use eval and not get... because of \c{::}}
+This corresponds to minimized expressions \c{X}, \c{V}.
+We also include here lookup to slots in a list or data frame, \ie minimized expression \c{\$}.
+Variable lookup leads to a large diversity of environments, many being package environments or package environments wrapped in \c{new}.
+
+It is the predominant usage of \eval in Base, through function \c{match.args} (with \eval on minimized expression \c{V}). In CRAN, it correlates with more exotic environments passed to \eval. 
+The dominant site in numbers of calls in CRAN, is located in \c{ggproto} in the custom object system of \emph{ggplot2}.
+It evaluates a variable \c{`\_inherit`}, which represents the parent of the class, in the definition environment of the class.
+\eval is used here to account for parents scoped in a namespace with \c{::}, which is actually a function call in R.
 
@@
 \todo{Show an actual example with \$}
-Slot lookup happens very often to evaluate some of the elements of a call list returned by \c{match.call}.
-For instance in the \emph{pwrFDR} package:
-\begin{lstlisting}
-m <- match.call()
-...
-average.power <- m$average.power <- eval(m$average.power, sys.parent())
-\end{lstlisting}
-
-Here, \c{m\$average.power}, depending on the shape of the arguments to the surrounding function,  can be resolved to a variable, a more complex expression, or a number, leading to several possible minimized expressions, including \c{X} and \c{V}.
 
@@
 by the fact that the expression is a call and the target environment is that of
-the parent.
+the parent. \todo{Remove that example}
 
@@
 
-\todo{Put a simpler example.}
+\todo{Show also an example with \c{substitute}}
 
@@
 
+  \todo{Put a simpler example, again with building a plot}
+
  Code generation is often used as a means to reduce boilerplate code;
@@
 
+Using \c{parse}, \ie converting strings to R expressions, opens to a more flexible code generation, as the building blocks which are going to be concatenated together, do not need to be valid R expressions.
+On the contrary, combining expression created with \c{substitute} or \c{quote} requires more discipline.
+In Kaggle, which features less experienced programmers,  code generation for plots is widespread, and uses \c{parse}.
+
 \mypara{Unnecessary use of \eval}
@@
 
-\eval is a powerful tool and can be sometimes with functions with more specialized goals.
+\eval is a powerful tool and can be sometimes replaced with functions with more specialized goals.
 For instance, variable lookup in any environment can be performed with \c{get}, and assignment in any environment, with \c{assign}.

### 2021-04-15 paper/main.tex
Work on the parse + plot examples

@@
 
-\subsection{Minimized expressions}
+\subsection{Shapes of \eval}
 
@@
 pattern is to transform a call to some function \c{f} into a call to \c{g} with
-some arguments retained and others modified.  \todo{or change the target
-function. Both are described as "patterns" in the R documentation.} As an
-illustration, consider the \c{vcpart} package's function \c{tvcglm} that is
-translated to a call to \c{tvcm} with two modifications to the argument list:
-argument \c{control} can't be missing and \c{fit} is set \c{"glm"}. The
-function ends with a call to \c{eval.parent} to ensure that the rewritten call
-is evaluated in the same environment the original call was.  \begin{lstlisting}
-tvcglm <- function(formula, data, family, ...) { k <- match.call() k[[1]] <-
-as.name("tvcm") if ("fit" %in% names(list(...))) warning("'fit' ignored.")
-k$fit <- "glm" eval.parent(k) } \end{lstlisting} This pattern is recognizable
+some arguments retained and others modified. This pattern is recognizable
 by the fact that the expression is a call and the target environment is that of
-the parent. \todo{Remove that example}
+the parent.
 
@@
 
-In package \emph{survival}\footnote{It is a \emph{recommended package}, so recommended by the CRAN maintainers, thoroughly tested, and following R best practices.}, in function \c{coxph}, the surrounding call is first captured, then \c{model.frame} is injected as the function name in the call, and finally, the new call is evaluated in the parent frame.
+In package \emph{survival}\footnote{It is a \emph{recommended package}, so recommended by the CRAN maintainers, thoroughly tested, and following R best practices.}, in function \c{coxph}, the surrounding call is first captured, then \c{model.frame} is injected as the function name in the call, and finally, the new call is evaluated in the parent frame, to ensure that the rewritten call
+is evaluated in the same environment the original call was.
 
@@
 \todo{Show also an example with \c{substitute}}
+
+\c{substitute} can be used to capture only a part of the call, \ie one of the arguments.
 
@@
 that was assembled by the programmer into a string. Here we show the method
-\c{plot} for class \c{sback} in package \c{wsbackfit}, simplified for
-explanatory purposes. The function takes a long argument list, the names of
-which are captured in the list \c{opt}. The string \c{stub} is composed of a
-subset of the arguments passed to this function; the variable is used to
-construct a call to \c{base::plot} which will draw a plot. \NOTE{This example
-  is a bit messed up. What is var? Does the use of ... in the eval grab the
-arguments of this call? If yes does it not end up with xlab twice? HELP}
-\todo{Does this comment still applies?} The \parse and \eval combination is
-used by the the \source function in the base library to load R code from a file
-in the current workspace.  \begin{lstlisting} plot.sback <- function(x,...) {
-opt <- names(list(...)) stub <- paste( ifelse("xlab" %in%
-opt,"",paste(",xlab=\"",var,"\"",sep="")), ifelse("type" %in%
-  opt,"",",type=\"l\""), sep="") plot <-
-  paste("plot(x.data,",stub,",...)",sep="") eval(parse(text=plot))
-  \end{lstlisting} This pattern is recognizable by its use of \c{parse} to turn
-  a string into an expression.
-
-  \todo{Put a simpler example, again with building a plot}
+\c{plot} for class \c{gmap} in package \c{gpmaps}, simplified for
+explanatory purposes.   \c{paste} is used to dispatch to a specialized plot function, among \c{plot2\_orig},  \c{plot3\_orig}, \c{plot3\_dec} and so on.
+\begin{lstlisting}
+eval(parse(text=paste('plot',gpmap$nloci,'_orig(gpmap)',sep="")))
+\end{lstlisting}
+
+ \todo{Add an example of a plot with parse from Kaggle}
+
+The \parse and \eval combination is
+used by the \source function in the base library to load R code from a file
+in the current workspace.
 
@@
 
+\todo{There is an example in package comprenhr that does code generation with substitute to create list comprehensions, but it is quite difficult to grasp.}
+
 \mypara{Unnecessary use of \eval}

### 2021-04-15 paper/main.tex
Example from kaggle

@@
 
-\mypara{Variable lookup and values.}
-This corresponds to minimized expressions \c{X}, \c{V}.
-We also include here lookup to slots in a list or data frame, \ie minimized expression \c{\$}.
+\mypara{Variable lookup and values.}  We also include here lookup to slots in a list or data frame.
+This corresponds to minimized expressions \c{X}, \c{V}, and  \c{\$}.
 Variable lookup leads to a large diversity of environments, many being package environments or package environments wrapped in \c{new}.
@@
 It evaluates the left-hand side of the formula, which is always a variable, in \c{eval\_lhs.formula}, as \c{eval(object[[2L]],envir=environment(object))} where \c{object} is the formula and \c{environment} returns the formula associated to the object.
-
-\todo{Show an actual example with \$}
 
@@
 
- \todo{Add an example of a plot with parse from Kaggle}
+In Kaggle, plots are often built with \eval, \c{paste} and \c{ggplot}.
+In this kernel for the \emph{digit-recognizer} challenge, \c{paste0} is used to generate \c{ggplot} commands featuring unique y labels to build 9 different graphs.
+
+\begin{lstlisting}
+p <- lapply(paste0('PC', 1:9), function(pc){
+	eval(parse(text=paste0('ggplot(ld, aes(x=label, y=',pc,')) + geom_boxplot()')))
+})
+\end{lstlisting}
 

### 2021-04-15 analysis/inc/functions.R
Provenance class

@@ -12,9 +12,9 @@ show_url <- Vectorize(function(path, name=basename(path), base_url=params$http_b
 
 read_parallel_task_result <- function(path) {
   file <- if (dir.exists(path)) file.path(path, "parallel.csv") else path
-  
-  read_csv(file) %>% 
-    select(-Seq, -Send, -Receive, -Signal, -Stdout, -Stderr) %>% 
+
+  read_csv(file) %>%
+    select(-Seq, -Send, -Receive, -Signal, -Stdout, -Stderr) %>%
     rename_all(tolower) %>%
     rename(hostname=host) %>%
     mutate(
@@ -49,3 +49,16 @@ cs_print <- function(df, cp=90) {
   .N <- cp
   print(filter(df, cp<=.N), n=Inf)
 }
+
+gen_provenance_class <- function(dataset) {
+  dataset %>%
+    mutate(provenance_class = case_when(
+      !is.na(expr_parsed_expression) ~ "string",
+      !is.na(expr_match_call) ~ "reflection",
+      expr_resolved_type_tag == "language" ~ "constructed",
+      expr_resolved_type_tag == "expression" ~ "constructed",
+      TRUE ~ NA_character_
+    ))
+}
+
+

### 2021-04-15 paper/main.tex
Apply ck's proofread in 5.4

@@
 finishes. Knowing where---\ie in which environments do these side-effects
-happen, can help us to determine how much of the compiler knowledge about the
+happen, can help us determine how much of the compiler knowledge about the
 program will be potentially invalidated.
@@
 The \rdyntrace contains low-level hooks capturing the life-cycle of an
-environment and variables defined in it. In the tracer, we register callbacks
-that record each environment creation and removal together with updates to its
-bindings. From the recorded data, we ignore side effects
-coming from \eval sites from unit testing frameworks.\footnote{In the corpus we
-have \c{RUnit, testthat, tinytest} and \c{unitizer} unit testing frameworks.}
-They run the testing code via \eval and thus the results would be biased
-because of the high number of code run in these tests
-(\CranRunnableCodeTestsRnd lines).
+environment and variables defined in it. We register callbacks in the tracer
+that records each environment creation and removal with updates to its
+bindings. From the recorded data, we ignore side effects coming from \eval
+sites from unit testing frameworks.\footnote{In the corpus we have \c{RUnit,
+testthat, tinytest} and \c{unitizer} unit testing frameworks.} They run the
+testing code via \eval, and thus the results would be biassed because of the
+high number of code in these tests (\CranRunnableCodeTestsRnd lines).
 
 From the \packageNbrunsRnd programs we capture \SEAll side effects from
-\SEAllCalls \eval calls in \SEAllSites sites. The challenge is again to remove
-the accidental side-effects that are caused by the R virtual machine
-implementation and are not related to the user code. For example, the
-\c{.Random.seed} variable, which contains the state of the random number
-generator is saved and restored from and to the global environment every time
-user calls one of the base routines for getting random numbers. Removing them
-leaves us with \SEUser side effects (\SEUserRatio) from \SEUserCalls \eval
-calls (\SEUserCallsRatio) in \SEUserSites sites (\SEUserSitesRatio). These
-sites are part of \SEUserFunctions in \SEUserPackages packages. \SEFunsNighty
-functions are responsible for 90\% of side effects. Half of the side-effects
-comes just from three functions: \c{plyr::allocate\_column} (allocates space
-for a new data frame column), \c{withr::execute\_handlers} (executes deferred
-expressions) and \c{foreach::doSEQ} (executes an expression on each element in
-a collection, possibly in parallel).
+\SEAllCalls \eval calls in \SEAllSites sites. Again, the challenge is to remove
+the accidental side-effects caused by the R virtual machine implementation and
+are not related to the user code. For example, the \c{.Random.seed} variable,
+which contains the state of the random number generator is saved and restored
+from and to the global environment every time the user calls one of the base
+routines that use random numbers. Removing them leaves us with \SEUser side
+effects (\SEUserRatio) from \SEUserCalls \eval calls (\SEUserCallsRatio) in
+\SEUserSites sites (\SEUserSitesRatio). These sites are part of
+\SEUserFunctions in \SEUserPackages packages. \SEFunsNighty functions are
+responsible for 90\% of side effects. Half of the side-effects come just from
+three functions: \c{plyr::allocate\_column} (allocates space for a new data
+frame column), \c{withr::execute\_handlers} (executes deferred expressions), and
+\c{foreach::doSEQ} (executes an expression on each element in a collection,
+possibly in parallel).
 
@@
 Section~\ref{sec:eval-in-r}), \SESitesNotInEnvirRatio modifies other
-environment and finally \SESitesBothEnvirRatio does both.
+environments, and finally \SESitesBothEnvirRatio does both.
 Table~\ref{tab:se-env} shows the class of the environment where \eval side
 effects happen (\emph{target environment}). It follows the classification
-described in Section~\ref{sec:env} with two extra classes: \emph{Object} which
+described in Section~\ref{sec:env} with two extra classes: \emph{Object}, which
 represents a side-effect that happens in S4 or R6 object environment, and
@@
 caller. The table shows data for both \eval sites and functions. For a function
-to have a given target environment class means that all of its \eval sites that
-do side effect must have the same target environment class. Majority of the
-\eval sites (\SESitesInOneClass) do all side effects consistently in one
-environment class. The same happens on the function level. Almost half of the
-sites and over a third of the functions do side effects in either \emph{Local}
-or \emph{Object} environments. This gives a ray of hope for the compiler. Even
-though it is possible to do anything anywhere, the data suggest that most
-side-effects are sane.
+to have a given target environment class, all of its \eval sites that do side
+effects must have the same target environment class The majority of the \eval
+sites (\SESitesInOneClass) do all side effects consistently in one environment
+class. The same happens on the function level. Almost half of the sites (over a
+third of the functions) do side effects in either \emph{Local} or \emph{Object}
+environments. This gives a ray of hope for the compiler. Even though it is
+possible to do anything anywhere, the data suggest that most side-effects are
+sane.
 
@@
 Table~\ref{tab:se-types} shows the proportion of the different side effects
-that were recorded. In terms of calls we see mostly assignments and in terms of
-sites definitions. This is expected. A subsequent \eval call will turn a
-definition into an update. Variable removal, while rare, it does happen.
-However, the vast majority comes from the already mentioned
-\c{withr::execute\_handlers} function used to defer evaluation of an expression
-to after the function exit. In our dataset, it is used almost exclusively by
-the \c{tidyselect} package that uses it to remove the reference to current
-quosure environment while interpreting a data frame column
+that were recorded. In terms of calls, we see primarily assignments and in
+terms of site definitions. This is expected. A subsequent \eval call will turn
+a definition into an update. The most dangerous side effect is variable removal
+as it means that after a call to \eval some binding in some environments will
+disappear. While rare, this happens, but the vast majority comes from a single
+site called by the already mentioned \c{withr::execute\_handlers} function.
+This makes sense as it is used to defer evaluation of an expression to after
+the function exit and thus used for clean up. In the corpus, it is used almost
+exclusively by the \c{tidyselect} package for removing the reference to the
+current quosure environment while interpreting a data frame column
 selectors.\footnote{\cf \url{https://tidyselect.r-lib.org/}}
+
+% TODO type of variables
+% TODO do assignments change types?
 

### 2021-04-15 analysis/inc/functions.R
Environment class

@@ -62,3 +62,25 @@ gen_provenance_class <- function(dataset) {
 }
 
 
+gen_env_class <- function(dataset) {
+  dataset %>%
+    mutate(envir_expression = if_else(envir_type == "NULL" & is.na(envir_expression), "NULL", envir_expression)) %>%
+    mutate(envir_expression = na_if(envir_expression, "NA")) %>%
+    extract(environment_class, regex = "((?:new\\+)*)(?:caller-(-?[[:digit:]]+)-)?(.*)", into = c("new_env", "hierarchy", "specific_env"), remove = FALSE, convert = TRUE) %>%
+    mutate(new_env = na_if(new_env, ""), specific_env = na_if(specific_env, ""), new_env = str_count(new_env, fixed("new+"))) %>%
+    mutate(env_class = case_when(
+      envir_type %in% c("list", "pairlist", "NULL") ~ "synthetic",
+      specific_env == "empty" ~ "synthetic",
+      envir_type %in% c("integer", "double") ~ "function",
+      # You can separate further into local and function here
+      # Local would be hierarchy == 0
+      !is.na(hierarchy) & is.na(new_env) & is.na(specific_env) ~ "function", # package subsumes function
+      is.na(new_env) & str_detect(specific_env, fixed("loop")) ~ "function",
+      !is.na(new_env) ~ "synthetic",
+      is.na(new_env) & specific_env == "global" ~ "global",
+      is.na(new_env) & !is.na(specific_env) ~ "package",
+      TRUE ~ "invalid"
+      # That would be closures or logical, that are not valid to be passed to envir
+    ))
+}
+

### 2021-04-15 analysis/environments.Rmd
Fix environment classification

@@
     n_op = interp_eval,
+    hash = expr_resolved_hash,
     envir = envir_expression,
@@
 ```{r}
-# 4 srcrefs will be counted twice, in package environments, and in caller environments
-pack_envs <- env_classes %>% filter(is.na(new_env),  specific_env != "global")
+# in environment-class such as caller-n-packagename,
+# package category wins
+pack_envs <- env_classes %>% filter(is.na(new_env),  !specific_env %in% c("global", "empty"), !str_detect(specific_env, fixed("loop")))
 nb_pack_sites <- n_distinct(pack_envs$src_ref)
 
+# That might overestimate a bit...
 nb_function_env_sites <- nb_caller_env_sites + nb_integer_sites

### 2021-04-15 paper/main.tex
ck's proofreading for Section 5.1 to 5.3

@@
 
-\medskip\noindent\EE{$min(e)=\c{X}$} Variables lookups are the most common form,
+\medskip\noindent\EE{$min(e)=\c{X}$} Variable lookups are the most common form,
 they are found in 28\% of the sites. This form includes simple variable reads,
@@
 
-\medskip\noindent\EE{$min(e)=$~\c{<-}} This form includes both assignments
-operator, the direct assignment {\tt <-}, and assignment to the parent
-environment {\tt <\,\!<-}, and the \c{\$} form. Assignments occur in 4\% of the
+\medskip\noindent\EE{$min(e)=$~\c{<-}} This form includes both assignment
+operators, the direct assignment {\tt <-}, and assignment to the parent
+environment {\tt <\,\!<-}, the \c{assign} function and the \c{\$} form. Assignments occur in 5\% of the
 sites. They represent the most obvious source of side-effects. The median number
@@
 \medskip\noindent\EE{$min(e)=\c{F()}$} This form captures simple function calls,
-\eg~\c{f(2)}, with neither variables or assignments. Or, more specifically it
+\eg~\c{f(2)}, with neither variables nor assignments. Or, more specifically it
 allows for variables in the function position but not in arguments. Usually,
@@
 forms allow for function calls whose arguments may include variable references.
-The latter allows nested calls.Assignments are excluded from this form. They
-occurs in, respectively, 14\% and 20\% of the sites in the corpus. Together they
+The latter allows nested calls. Assignments are excluded from this form. They
+occur in, respectively, 14\% and 20\% of the sites in the corpus. Together they
 are the most frequent forms. The median numbers of interpreter steps are,
@@
 
-\medskip\noindent\EE{$min(e)=\c{FUN}$} This form captures expressions define
-functions, \eg~\c{function(x)x+1}, and do nothing else. \c{FUN} occurs in
+\medskip\noindent\EE{$min(e)=\c{FUN}$} This form captures expressions that define functions, \eg~\c{function(x) x+1}, and do nothing else. \c{FUN} occurs in
 \packageFunctionDefinitionSitesPercent of sites. Evaluating a function
@@
 
-\medskip\noindent\EE{$min(e)=\c{BLOCK}$} This form captures multi-statements
+\medskip\noindent\EE{$min(e)=\c{BLOCK}$} This form captures multi-statement
 code blocks, which occur in only 4\% of sites. These are larger expressions, we
@@
 computation started by \eval and the potential reach of its side-effects. The
-second argument to \eval specifies that environment. Environment of \eval can be
+second argument to \eval specifies that environment. Environments of \eval can be
 classified into the following four kinds:
@@
 \begin{compactitem}[---]
-\item \emph{Function:} environment for the local variables of some function
+\item \emph{Function:} environments for the local variables of some function
   currently active on the call stack. Obtained by calling \c{parent.frame()} or
@@
   \c{as.environment}. Also includes the empty environment.
-\item \emph{Global:} the environment in which scripts or interactive commands
+\item \emph{Global:}  environments in which scripts or interactive commands
   are evaluated.
-\item \emph{Package:} the environment of a loaded library.
+\item \emph{Package:}  environments of a loaded library.
 \end{compactitem}
@@
 \noindent
-As shown by Table~\ref{tab:highlevelenvironments}, most calls evaluate in a
+As shown in Table~\ref{tab:highlevelenvironments}, most calls evaluate in a
 function environment with the \c{global} kind as a distant second. This means
@@
 Thus 0 is the direct caller and corresponds to local variables; 1 is its parent, and so on.
-The data suggests that in 81\% of cases, \eval access its caller's environment -- this means, the variables of the function where \eval textually
+The data suggests that in 81\% of cases, \eval accesses its caller's environment -- this means, the variables of the function where \eval textually
 occurs are read and written to. It is interesting that some 1.5\% of sites
-evaluate a code in an environment that is three frame or above from \eval. This
+evaluate a code in an environment that is three frames or above from \eval. This
 distance implies that, in general, modular reasoning is impossible, to
@@
 evaluate an expression in either a restricted environment, use a data
-structure as an environment. ({\it Global}) The fact that global environment
-show up so frequently is likely an artifact of the way the code is run. ({\it
+structure as an environment. ({\it Global}) The fact that the global environment
+shows up so frequently is likely an artifact of the way the code is run. ({\it
   Package}) There are only 92 sites that use a package environment in \eval,
@@
   promises and \c{substitute} is used to retrieve the source expression
-  associated to the promise.
+  associated with the promise.
 \item {\it String:} Finally, expressions can be created from strings by invoking
@@
 This data is obtained by dynamically tainting values as they are produced by the
-various sources, due to technical reasons there is some imprecision in the
+various sources, due to technical reasons, there is some imprecision in the
 results. In particular, we are not able to classify all sites. Manual inspection
@@
 minority of the use cases, instead the constructed category shows that the
-majority of \evals comes code that was processed by the compiler, and may be
+majority of \evals comes from code that was processed by the compiler, and may be
 slightly modified by the programmer before invoking \eval. Both constructed and

### 2021-04-15 paper/main.tex
Example of defer in withr
Adding an example of call capture with substitute

@@
 
-\todo{Add testthat or withr::defer\_handler (fancy on.exit)}
+Package \emph{withr} provides facilities to encapsulate side-effecting functions.
+The function \c{defer} is used to attach an expression to be evaluated when exiting any frame in the stack.
+It can be passed arbitrary code, and so blocks.
 
@@
 \c{substitute} can be used to capture only a part of the call, \ie one of the arguments.
+For instance, in package \emph{xyplot}, the expression for argument \c{subset} is captured with \c{substitute} and then evaluated in a dataframe \c{data} and in the definition environment of \c{x}.
+\begin{lstlisting}
+ xyplot.formula <-  function(x,    data = NULL, ..., subset = TRUE)
+ ...
+subset <- eval(substitute(subset), data, environment(x))
+\end{lstlisting}
 

### 2021-04-15 paper/main.tex
A bit of a conclusion

@@
 
+In this paper we provide large-scale study of the use of the \eval function in
+the R programming language. It is based primarily on a corpus of
+\CranRunnableScripts scripts extracted from \CranPackages R packages. As
+expected, \eval is used mainly in libraries and therefore we have found just a
+few \eval call sites in scripts coming from Kaggle that we included to
+represent code written by R practitioners. We used an instrumented version of
+the R virtual machine, \rdyntrace, to capture all \eval calls traces
+including the relevant side effects that happen within these calls. There is a
+number of insights that can be learnt from the data.
+%
+\eval is widely used in the R core packages that come bundled with the
+language. Because of that there is hardly any R code that would not use it.
+\eval is used to implement some basic functionality such as package loading.
+Putting the core packages aside, in CRAN, \eval is used by \PkgPackagesRatio
+of the available \CranPackages packages.
+%
+The code executed by R \eval has unusual expressive power. It can essentially
+modify anything anywhere which is wary for both R developers as well as R
+compiler.
+% TODO expressions passed to eval
+% TODO side effects
+
+% how much can we say about the threat model?
+
 \bibliography{bib/bibliography,bib/jv}

### 2021-04-15 paper/main.tex
Improve

--- 2021-04-15 paper/main.tex
@@
 
-\todo{Show also an example with \c{substitute}}
-
 \c{substitute} can be used to capture only a part of the call, \ie one of the arguments.
@@
 
-\todo{There is an example in package comprenhr that does code generation with substitute to create list comprehensions, but it is quite difficult to grasp.}
+\todo{There is an example in package comprenhr that does code generation with substitute to create list comprehensions, but it is quite difficult to grasp.
+But at the same type, we lack here an example of code generation using quite, enquote or substitute.}
 

### 2021-04-16 analysis/side-effects.Rmd
Data refresh

@@
 )
+
+S <- mutate(
+  S,
+  caller=str_replace(eval_call_srcref, "::(.*)::(.*)::.*", "\\1::\\2"),
+  caller_package=str_replace(eval_call_srcref, "::(.*)::.*::.*", "\\1")
+)
 ```
@@
 ```{r}
-S_call <- sum(S$nb_ev_calls)
-S_site <- count(S, eval_call_srcref)
-S_package <- count(S, eval_source)
-S_funs <- sum(PKG$funs_with_eval)
+S_calls <- sum(S$nb_ev_calls)
+S_sites <- count(S, eval_call_srcref)
+S_pkgs <- count(S, eval_source)
+S_funs <- count(S, caller)
 
@@
 SE_missing_N <- sum(SE_missing_sites$N)
+```
 
+```{r}
 overview_table(
@@
   r("User calls ratio", ratio(SE_calls, SE_all_calls)),
+  r("User calls to all ratio", ratio(SE_calls, S_calls)),
 
@@
   r("User sites ratio", ratio(SE_sites, SE_all_sites)),
+  r("User sites to all ratio", ratio(SE_sites, S_sites)),
 
-  r("User functions", count(SE, caller)),
-  r("User packages", count(SE, caller_package)),
+  r("User functions", SE_funs),
+  r("User functions to all ratio", ratio(SE_funs, S_funs)),
+  r("User packages", SE_pkgs),
+  r("User packages to all ratio", ratio(SE_pkgs, S_pkgs))
 )
@@
   mutate(
-    p=fmt(p, suffix="\\%", digit=1),
+    p=fmt(p, suffix="\\%", digit=0),
     o=if_else(str_starts(classes, fixed("\\")), 0L, n)
@@
 
-sum_sites_classes %>%
-  mutate(classes=if_else(classes=="multiple", "\\emph{multiple}", classes)) %>%
-  xtable(
-    align="llrr",
-  ) %>%
-  print(
-    file=file.path(TAGS_DIR, "table-top-sites-classes-se.tex"),
-    comment=FALSE, 
-    include.colnames=FALSE,
-    include.rownames=FALSE,
-    only.contents=TRUE,
-    sanitize.text.function=identity,
-    timestamp=FALSE
-  )
+# sum_sites_classes %>%
+#   mutate(classes=if_else(classes=="multiple", "\\emph{multiple}", classes)) %>%
+#   xtable(
+#     align="llrr",
+#   ) %>%
+#   print(
+#     file=file.path(TAGS_DIR, "table-top-sites-classes-se.tex"),
+#     comment=FALSE, 
+#     include.colnames=FALSE,
+#     include.rownames=FALSE,
+#     only.contents=TRUE,
+#     sanitize.text.function=identity,
+#     timestamp=FALSE
+#  )
 ```
@@
     mutate(
-    p=fmt(p, suffix="\\%", digit=1),
+      p=fmt(p, suffix="\\%", digit=0),
       o=if_else(str_starts(classes, fixed("\\")), 0L, n)
@@
 
-sum_funs_classes %>%
-  mutate(classes=if_else(classes=="multiple", "\\emph{multiple}", classes)) %>%
-  xtable(
-    align="llrr",
-  ) %>% 
-  print(
-    file=file.path(TAGS_DIR, "table-top-sites-classes-funs.tex"),
-    comment=FALSE, 
-    include.colnames=FALSE,
-    include.rownames=FALSE,
-    only.contents=TRUE,
-    sanitize.text.function=identity,
-    timestamp=FALSE
-  )
+# sum_funs_classes %>%
+#   mutate(classes=if_else(classes=="multiple", "\\emph{multiple}", classes)) %>%
+#   xtable(
+#     align="llrr",
+#   ) %>% 
+#   print(
+#     file=file.path(TAGS_DIR, "table-top-sites-classes-funs.tex"),
+#     comment=FALSE, 
+#     include.colnames=FALSE,
+#     include.rownames=FALSE,
+#     only.contents=TRUE,
+#     sanitize.text.function=identity,
+#     timestamp=FALSE
+#   )
 ```
@@
   )) %>%
-  mutate_if(is.double, ~fmt(., suffix="%", digit=1)) %>%
+  mutate_if(is.double, ~fmt(., suffix="%", digit=0)) %>%
   mutate_if(is.integer, ~fmt(.)) %>%
@@
 ```
+
+## Side effects with norm
+
+```{r}
+NE <- read_csv(
+  PACKAGE_NORMALIZED_EXPRESSION_FILE,
+  col_types = cols(
+    minimized = col_character(),
+    topcall = col_character(),
+    is_model = col_double(),
+    has_fundef = col_double(),
+    has_calls = col_double(),
+    has_assigns = col_double(),
+    has_var = col_double(),
+    has_bracket = col_double(),
+    is_assign = col_double(),
+    is_value = col_double(),
+    is_ignore = col_double(),
+    has_dollar = col_double(),
+    has_user_call = col_double(),
+    has_block = col_double(),
+    has_meta_op = col_double(),
+    normalized = col_character(),
+    trimmed = col_character(),
+    hash = col_character()
+  )
+) %>%
+  select(
+    minimized,
+    topcall,
+    normalized,
+    expr_resolved_hash=hash
+  )
+```
+
+```{r eval=F}
+C <- read_fst(path(PACKAGE_DATA_DIR, "calls-sub.fst")) %>% as_tibble
+CSE <- left_join(
+  C,
+  select(SE, cid, event, class, N),
+  by="cid"
+)
+```

### 2021-04-16 analysis/inc/latextags.R

@@ -308,7 +308,7 @@ fmt <- function(x, prefix="", suffix="", ...) {
 
 # TODO: make this parameter a global option so it can be set globally
 # TODO: test c(0.0123, 1.123, -0.0123, -1.123)
-.fmt.double <- function(x, digits=if (x %% 1 == 0) 0 else 1, floor=FALSE, ceiling=FALSE) {
+.fmt.double <- function(x, digits=1, floor=FALSE, ceiling=FALSE) {
   if (floor) x <- floor(x)
   if (ceiling) x <- ceiling(x)
 
@@ -321,10 +321,10 @@ fmt <- function(x, prefix="", suffix="", ...) {
     else y
   })
 
-  formatC(x, big.mark=",", format="f", digits=digits)
+  prettyNum(x, big.mark=",", scientific=F)
 }
 
-.fmt.num_with_suffix <- function(x, floor=FALSE, ceiling=FALSE, digits=if (x %% 1 == 0) 0 else 1) {
+.fmt.num_with_suffix <- function(x, floor=FALSE, ceiling=FALSE, digits=1) {
   suffix <- attr(x, "suffix")
   fmt(as.double(x), suffix=suffix, floor=floor, ceiling=ceiling, digits=digits)
 }

### 2021-04-16 analysis/inc/paths.R

@@ -24,6 +24,7 @@ PACKAGE_EVALS_STATIC_FILE <- path(PACKAGE_DATA_DIR, "evals-static.csv")
 PACKAGE_CODE_FILE         <- path(PACKAGE_DATA_DIR, "code.fst")
 PACKAGE_TRACE_LOG_FILE    <- path(PACKAGE_DATA_DIR, "trace-log.csv")
 PACKAGE_SIDE_EFFECTS_FILE <- path(PACKAGE_DATA_DIR, "side-effects.fst")
+PACKAGE_NORMALIZED_EXPRESSION_FILE <- path(PACKAGE_DATA_DIR, "normalized-expressions.csv")
 
 KAGGLE_KERNEL_FILE <- path(KAGGLE_DATA_DIR, "kernel.csv")
 KAGGLE_EVALS_STATIC_FILE <- path(KAGGLE_DATA_DIR, "evals-static.csv")

### 2021-04-16 paper/main.tex
Update normalized exprs
Fix format

@@
 \begin{wrapfigure}{4}{5.8cm}
+  \small
   \vspace*{-2mm}
@@
   \begin{tabular}{|r@{\,}r@{\,}l@{\,}r|r@{\,}r@{\,}l@{}r|} \hline
-\multicolumn{3}{|c}{\small\#calls} &\small\#sites &
-\multicolumn{3}{c}{\small\#calls} &\small\#sites \\\hline
+\multicolumn{3}{|c}{\#calls} & \#sites &
+\multicolumn{3}{c}{\#calls} & \#sites \\\hline
 \tt 0 &--& \tt 50    & \packageRunbina & \tt 501 &--& \tt 1000   & \packageRunbine\\
@@
   \vspace*{-1mm}
-  {\hspace{-3cm}\includegraphics[width=0.8\textwidth]{package_size_loaded_distribution}}
+  {\hspace{-25mm}\includegraphics[width=0.7\textwidth]{package_size_loaded_distribution}}
 \end{tabular}
@@
 
+% Add a table with the distribution of types of expr_resolved?
+
 The expressions passed to \eval vary widely. In order to categorize them, let us
@@
 
-\noindent
 We detail these forms and discuss their implication for the behavior of \eval.
@@
 kinds of environments, but in \packageNbOneCategoryEnvirSitePercent of the case a
-site has a single kind.
+site has a single kind. It means that we can predict the \emph{shape} of the expression passed to \eval in \packageNbOneCategoryEnvirSitePercent of the sites.
 

### 2021-04-16 paper/main.tex
Subsubsectionning Section 6

@@
 
+The examples are based on code from CRAN or Kaggle, simplified to fit the limited space of a scientific paper. We indicate the package and the function where the example is located.
+
 \subsection{Shapes of \eval}
 
-\mypara{Variable lookup and values.}  We also include here lookup to slots in a list or data frame.
+We group minimized expressions in larger categories.
+
+\subsubsection{Variable lookup and values.}  We also include here lookup to slots in a list or data frame.
 This corresponds to minimized expressions \c{X}, \c{V}, and  \c{\$}.
@@
 Package \emph{statnet.common} handles formulas, R object that associates an expression of the form \c{x~\textasciitilde~expr} and a carry-on environment where to evaluate the formula.
-It evaluates the left-hand side of the formula, which is always a variable, in \c{eval\_lhs.formula}, as \c{eval(object[[2L]],envir=environment(object))} where \c{object} is the formula and \c{environment} returns the formula associated to the object.
-
-\mypara{Assigments.} It corresponds to minimized expression \c{<-} and groups together \c{<-}, \c{<<-} and \c{assign}.
+It evaluates the left-hand side of the formula, which is always a variable, in function \c{eval\_lhs.formula}, as \c{eval(object[[2L]],envir=environment(object))} where \c{object} is the formula and \c{environment} returns the formula associated to the object.
+
+\subsubsection{Assigments.} It corresponds to minimized expression \c{<-} and groups together \c{<-}, \c{<<-} and \c{assign}.
 Package \emph{plyr} has a function combine data frames by row, filling in missing columns.
@@
 
-\mypara{Function definition.} This corresponds to minimized expression \c{FUN}.
+\subsubsection{Function definition.} This corresponds to minimized expression \c{FUN}.
 Package \emph{Rcpp} is used to easily interface C++ and R and it uses \eval to bind C++ functions and classes to R objects.
@@
 
-\mypara{Simple function calls.} This corresponds to minimized expression \c{F()}, \ie calls without nested functions and without variables (but possibly named arguments).
+\subsubsection{Simple function calls.} This corresponds to minimized expression \c{F()}, \ie calls without nested functions and without variables (but possibly named arguments).
 For instance, package \emph{foreach} provides \c{foreach}-like loops.
@@
 
-\mypara{Complex function calls.} It corresponds to minimized expressions \c{F(X)} and \c{F(F(X))}, \ie function calls with variables, possibly with nested function calls.
+\subsubsection{Complex function calls.} It corresponds to minimized expressions \c{F(X)} and \c{F(F(X))}, \ie function calls with variables, possibly with nested function calls.
 Package \emph{huxtable} is used to create custom tables and export them to various formats.
@@
 
-\mypara{Domain Specific Language.} R expressions are built and transformed symbolically and then evalled to a concrete value.
+We present here high-level patterns that we identified after manual inspections of hundred of \eval sites. Some of the patterns were also described in the R documentation.
+
+\subsubsection{Domain Specific Language.} R expressions are built and transformed symbolically and then evalled to a concrete value.
 For instance,   base R packages \emph{stats} provides two operators for \emph{symbolic differentiation}, \c{D} and \c{deriv}.
@@
 
-\mypara{Capturing a call.}  A common use case for \eval is to be combined with \c{match.call}.
+\subsubsection{Capturing a call.}  A common use case for \eval is to be combined with \c{match.call}.
 \c{match.call} walks up the call stack, captures the code that invoked the
-currently executing function, and returns it as an unevaluated expression that includes the call function and its unevaluated arguments. The
-pattern is to transform a call to some function \c{f} into a call to \c{g} with
-some arguments retained and others modified. This pattern is recognizable
+currently executing function, and returns it as an unevaluated expression that includes the call function and its unevaluated arguments.
+The pattern is to transform a call to some function \c{f} into a call to \c{g} with some arguments retained and others modified.
+This pattern is recognizable
 by the fact that the expression is a call and the target environment is that of
@@
 Call <- match.call()
-...
 tform <- Call[c(1,indx)]  # only keep the arguments we wanted
 tform[[1L]] <- quote(stats::model.frame)  # change the function called
-...
 mf <- eval(tform, parent.frame())
@@
  xyplot.formula <-  function(x,    data = NULL, ..., subset = TRUE)
- ...
 subset <- eval(substitute(subset), data, environment(x))
@@
 In the R community, the call capturing pattern with evaluation in a non-local environment is often referred~\cite{hadley} as \emph{Non-Standard Evaluation}.
-
-\mypara{Injecting code in package environments.} R distinguishes two kinds of environments for packages: the \emph{package} environment itself, which contains every publicly accessible functions, and the \emph{namespace} environment, which contains all functions including private ones.
+This pattern in relation with \c{model.frame} is also described explicitly in the R documentation of \c{match.call}.
+
+\subsubsection{Injecting code in package environments.} R distinguishes two kinds of environments for packages: the \emph{package} environment itself, which contains every publicly accessible functions, and the \emph{namespace} environment, which contains all functions including private ones.
 
@@
 
-\mypara{Code Generation.} The more traditional use of \eval is to execute code
+\subsubsection{Code Generation.} The more traditional use of \eval is to execute code
 that was assembled by the programmer into a string. Here we show the method
@@
 
-\mypara{Unnecessary use of \eval}
+\subsubsection{Unnecessary use of \eval}
 

### 2021-04-16 analysis/environments.Rmd
Enclos class added for synthetic environments

@@
     environment_class,
-    enclos = enclos_expression
+    enclos = enclos_expression,
+    enclosure_class,
+    enclos_type
   ) -> E_envir
@@
 ```{r envir_list}
-eval_calls_env_list <- eval_calls_explicit %>% filter(envir_type == "list")
+eval_calls_env_list <- eval_calls_explicit %>% filter(envir_type %in% c("list", "pairlist"))
 
@@
 
-And with the higher-level categories:
+## Synthetic
+
+This corresponds to `new` and to `list/NULL/pairlist`:
 
 ```{r}
-high_level_new <- env_calls %>% filter(!is.na(new_env)) %>% 
+list_like_envs <- E_envir %>% filter(envir_type %in% c("pairlist", "list", "NULL")) %>% 
+  extract(enclosure_class, regex = "((?:new\\+)*)(?:caller-(-?[[:digit:]]+)-)?(.*)", into = c("new_env", "hierarchy", "specific_env"), remove = FALSE, convert = TRUE) %>%
+    mutate(specific_env = na_if(specific_env, ""), new_env = str_count(new_env, fixed("new+"))) 
+  # We keep new.env not NA because it is used for the classification of new envs...
+
+other_synth_envs <- env_classes %>% filter(!is.na(new_env) | specific_env == "empty")
+
+synthetic_envs <- bind_rows(list_like_envs, other_synth_envs)
+nb_synthetic_sites <- n_distinct(synthetic_envs$src_ref)
+```
+
+```{r}
+high_level_new <- synthetic_envs %>% 
   mutate(category = case_when(
@@
     specific_env == "global" ~ "global",
-    specific_env == "empty" ~ "empty",
+    specific_env == "empty" | enclos_type == "NULL" ~ "empty",
     str_detect(specific_env, fixed("loop")) ~ "function", #if it is loop, it means that we found the eval env in the frames,
@@
   arrange(desc(n)) %>% 
-  mutate(pc = ratio(n, nb_new_env_sites), cpc = cumsum(pc))
+  mutate(pc = ratio(n, nb_synthetic_sites), cpc = cumsum(pc))
 
@@
 nb_function_env_sites <- nb_caller_env_sites + nb_integer_sites
-
-synthetic_envs <- env_classes %>% filter(envir_type %in% c("list", "NULL", "pairlist") | !is.na(new_env) | specific_env == "empty")
-nb_synthetic_sites <- n_distinct(synthetic_envs$src_ref)
 

### 2021-04-16 paper/main.tex
Enclos class added for synthetic environments
Enclosure
Update provenance numbers

@@
 \subsection{The expression in \eval} \label{sec:minimized}
-
-% Add a table with the distribution of types of expr_resolved?
 
@@
 child (side effects go to the child only). Table~\ref{tab:newenvs} shows the
-kinds of the parents of synthetic environments. Over half are functions, then
-come packages and global.
+kinds of the parents of synthetic environments. Most of them are functions, then
+come global and package.
 
@@
 \begin{compactitem}[---]
-\item {\it Reflection:} This group corresponds to uses of the \c{match.call}
-  function to reflectively capture the expression that invoked the current
-  function.
 \item {\it Constructed:} Expressions can be constructed by invoking the
@@
   associated with the promise.
+  \item {\it Reflection:} This group corresponds to uses of the \c{match.call}
+  function to reflectively capture the expression that invoked the current
+  function.
 \item {\it String:} Finally, expressions can be created from strings by invoking
@@
 Constructed & \packageNbConstructedSites & \packageNbConstructedSitePercent \\
-String & \packageNbStringSites & \packageNbStringSitePercent \\
-Reflection &  \packageNbMatchCallExprsSites & \packageMatchCallExprsSitePercent\\\hline
+Reflection &  \packageNbMatchCallExprsSites & \packageMatchCallExprsSitePercent\\
+String & \packageNbStringSites & \packageNbStringSitePercent \\\hline
 \end{tabular}

### 2021-04-16 paper/main.tex
Discussion Section 6

@@
 
-The examples are based on code from CRAN or Kaggle, simplified to fit the limited space of a scientific paper. We indicate the package and the function where the example is located.
+The examples are based on code from CRAN or Kaggle, simplified to fit the limited space of a scientific paper. For each example, we indicate the package and the function where it is located.
 
@@
 
-We group minimized expressions in larger categories.
+We illustrate the minimized expressions of Section~\ref{sec:minimized} on representative examples.
+
+% model.frame is in the next subsection, in the high-level patterns
 
@@
 
-It is the predominant usage of \eval in Base, through function \c{match.args} (with \eval on minimized expression \c{V}). In CRAN, it correlates with more exotic environments passed to \eval. 
+It is the predominant usage of \eval in Base, through function \c{match.args} (with evaluates on  minimized expression \c{V}). In CRAN, it correlates with more exotic environments passed to \eval. 
 The dominant site in numbers of calls in CRAN, is located in \c{ggproto} in the custom object system of \emph{ggplot2}.
@@
 Package \emph{statnet.common} handles formulas, R object that associates an expression of the form \c{x~\textasciitilde~expr} and a carry-on environment where to evaluate the formula.
-It evaluates the left-hand side of the formula, which is always a variable, in function \c{eval\_lhs.formula}, as \c{eval(object[[2L]],envir=environment(object))} where \c{object} is the formula and \c{environment} returns the formula associated to the object.
+It evaluates the left-hand side of the formula, which is always a variable, in function \c{eval\_lhs.formula}, where \c{object} is the formula and \c{environment} returns the formula associated to the object:
+
+\begin{lstlisting}
+eval(object[[2L]],envir=environment(object))
+\end{lstlisting}
 
@@
 RunMcmc <- function... {
-	 ...
 	 eval(expr_q, envir=env)
@@
 
-We present here high-level patterns that we identified after manual inspections of hundred of \eval sites. Some of the patterns were also described in the R documentation.
+We present here high-level patterns that we identified after manual inspections of hundred of \eval sites. Some of the patterns are also described in the R documentation.
 
@@
 On the contrary, combining expression created with \c{substitute} or \c{quote} requires more discipline.
-In Kaggle, which features less experienced programmers,  code generation for plots is widespread, and uses \c{parse}.
+In Kaggle, which features less experienced programmers,  code generation for plots is widespread, and uses \c{parse} on strings with \eval.
 
@@
 But at the same type, we lack here an example of code generation using quite, enquote or substitute.}
+
+\todo{Add a pattern about Library Loading? \c{library(packageName)} in R uses \eval This is one of the patterns in Javascript...}
 
@@
 
-\mypara{Discussion.} % more...
+\mypara{Discussion.} \Eval in R is mostly used for meta-programming purposes and to access environments other than the local one.
+In Javascript, patterns are ad-hoc, such as json loading and parsing, and can often be rewritten without \eval, such as function or method call, or assignment to a local variable or an object property.
+Compared to Javascript, \eval patterns in R we observed are much more legitimate, although this is biased by the fact that our main corpus CRAN features more experienced developers than  websites found at large on the web.
+\todo{can we really claim that?
+SHould we talk about Kaggle there?}
 
@@
 Nevertheless, \eval shines when expressions passed to it can be of many different types.
-% And it is shorter to type, and no need to remember all the specialized functions.
+% And it is shorter to type, and no need to remember all the specialized functions, so it is convenient.
 

### 2021-04-16 paper/main.tex
Updates

--- 2021-04-16 paper/main.tex
@@
 from and to the global environment every time the user calls one of the base
-routines that use random numbers. Removing them leaves us with \SEUser side
-effects (\SEUserRatio) from \SEUserCalls \eval calls (\SEUserCallsRatio) in
-\SEUserSites sites (\SEUserSitesRatio). These sites are part of
-\SEUserFunctions in \SEUserPackages packages. \SEFunsNighty functions are
-responsible for 90\% of side effects. Half of the side-effects come just from
-three functions: \c{plyr::allocate\_column} (allocates space for a new data
-frame column), \c{withr::execute\_handlers} (executes deferred expressions), and
+routines that use random numbers.
+%
+% \begin{wrapfigure}{r}{5cm} \hspace*{-20mm} \centering
+% \includegraphics[width=90mm]{se-types.pdf} \caption{Distribution of R types
+% in target environments for side effects.} \label{fig:se-types}
+% \end{wrapfigure}
+%
+Removing them leaves us with \SEUser side effects (\SEUserRatio) from
+\SEUserCalls \eval calls (\SEUserCallsRatio) in \SEUserSites sites
+(\SEUserSitesRatio). These sites are part of \SEUserFunctions in
+\SEUserPackages packages. \SEFunsNighty functions are responsible for 90\% of
+side effects. Half of the side-effects come just from three functions:
+\c{plyr::allocate\_column} (allocates space for a new data frame column),
+\c{withr::execute\_handlers} (executes deferred expressions), and
 \c{foreach::doSEQ} (executes an expression on each element in a collection,
-possibly in parallel).
-
-Most of the \eval sites (\SESitesInEnvirRatio) do side effects in the
-environment specified by the \c{envir} parameter (\cf
+possibly in parallel). Most of the \eval sites (\SESitesInEnvirRatio) do side
+effects in the environment specified by the \c{envir} parameter (\cf
 Section~\ref{sec:eval-in-r}), \SESitesNotInEnvirRatio modifies other
@@
 
-% TODO type of variables
+Looking closely at the value types in variable updates, we observe that the
+majority of \eval sites involve basic R vectors (\SEBasicTypeRatio) and lists
+(\SEListTypeRatio). From the compiler perspective, we would like to know how
+many sites changes function bindings. In the corpus, this happens in only
+\SEClosureType sites, \SEClosureTypeLocal of them do that in the local
+environment. We have manually inspected a few of these sites, but except for
+manual injection of parameters into a \c{model.frame} execution environment, we
+did not find a common use case.
+
 % TODO do assignments change types?
@@
 
-In this paper we provide large-scale study of the use of the \eval function in
-the R programming language. It is based primarily on a corpus of
+In this paper, we provide a large-scale study of the use of the \eval function
+in the R programming language. It is based primarily on a corpus of
 \CranRunnableScripts scripts extracted from \CranPackages R packages. As
-expected, \eval is used mainly in libraries and therefore we have found just a
-few \eval call sites in scripts coming from Kaggle that we included to
+expected, \eval is used mainly in libraries and therefore we have found just
+a few \eval call sites in scripts coming from Kaggle that we included to
 represent code written by R practitioners. We used an instrumented version of
-the R virtual machine, \rdyntrace, to capture all \eval calls traces
-including the relevant side effects that happen within these calls. There is a
-number of insights that can be learnt from the data.
+the R virtual machine, \rdyntrace, to capture all \eval calls traces including
+the relevant side effects that happen within these calls. Several insights can
+be learnt from the data.
 %
-\eval is widely used in the R core packages that come bundled with the
-language. Because of that there is hardly any R code that would not use it.
+First, \eval is widely used in the R core packages that come bundled with the
+language. Because of that, there is hardly any R code that would not use it.
 \eval is used to implement some basic functionality such as package loading.
-Putting the core packages aside, in CRAN, \eval is used by \PkgPackagesRatio
-of the available \CranPackages packages.
+Putting the core packages aside, in CRAN, \eval is used by \PkgPackagesRatio of
+the available \CranPackages packages.
 %
 The code executed by R \eval has unusual expressive power. It can essentially
-modify anything anywhere which is wary for both R developers as well as R
-compiler.
+modify anything anywhere which is wary for both R developers and R compiler.
 % TODO expressions passed to eval
-% TODO side effects
+Even more so in the case of side effects that can quite literally happen
+anywhere. The data however suggest that this is not the case. The majority of
+eval sites do not do any observable side effects and when they do, it happens
+in a predictable environment.
 
 % how much can we say about the threat model?
+Similarly to \citep{ecoop11}, we started looking into \evals in R with the
+ambition to replace them by other feature that are more amenable to static
+analysis. The data unfortunately suggests that replacing \eval is going to be
+lengthly journey. Even if we disregard core packages as they are part of the
+language and the \eval use is stable, in CRAN \eval is used widely and in more
+sophisticated way than was the case in JavaScript. There the vast majority of
+sites fitted an extremely simple categorization based on simple regular
+expressions. Not so in R. While some usage, especially the one which calls
+\eval with an expression parsed from string, can be categorized and often
+replaced by equivalent yet more disciplined and safer code, the case of \eval
+coming from reflection is much harder.
 

### 2021-04-16 paper/main.tex
SE discussion

@@
 
-\begin{table}[h]
+\begin{wraptable}{r}{7cm}\small\centering
   \small
@@
   \end{tabular}
-  \caption{Most used target environments for \eval side-effects} \label{tab:se-env}
-\end{table}
+  \caption{Target environments for \eval side-effects} \label{tab:se-env}
+\end{wraptable}
 
@@
 
-\mypara{Discussion}
-\todo{Is there any comparison with JS?}
+\mypara{Discussion} In JavaScript assignments in \eval can happen in either
+local scope or less often, when called through an alias, in the global scope.
+In R, given the support for first-class environments, it can happen anywhere.
+The data however suggest that most of side effects from \eval happen in a
+predictable environment.
 

### 2021-04-16 paper/main.tex
Update

@@
 \CranRunnableScripts scripts extracted from \CranPackages R packages. As
-expected, \eval is used mainly in libraries and therefore we have found just
-a few \eval call sites in scripts coming from Kaggle that we included to
-represent code written by R practitioners. We used an instrumented version of
-the R virtual machine, \rdyntrace, to capture all \eval calls traces including
-the relevant side effects that happen within these calls. Several insights can
-be learnt from the data.
+expected, \eval is used mainly in libraries. We have found just a few \eval
+call sites in scripts coming from large corpus of Kaggle that represents code
+written by R practitioners. We used an instrumented version of the R virtual
+machine, \rdyntrace, to capture all \eval calls traces including the relevant
+side effects that happen within these calls. Several insights can be learnt
+from the data.
 %

### 2021-04-16 analysis/side-effects.Rmd

@@
 library(tidyverse)
+library(viridis)
 library(xtable)
@@
 
+## Types
+
+```{r}
+SEXP_TYPES <- tribble(
+  ~type, ~type_name,
+  0, "NILSXP",
+  1, "SYMSXP",
+  2, "LISTSXP",
+  3, "CLOSXP",
+  4, "ENVSXP",
+  5, "PROMSXP",
+  6, "LANGSXP",
+  7, "SPECIALSXP",
+  8, "BUILTINSXP",
+  9, "CHARSXP",
+  10, "LGLSXP",
+  13, "INTSXP",
+  14, "REALSXP",
+  15, "CPLXSXP",
+  16, "STRSXP",
+  17, "DOTSXP",
+  18, "ANYSXP",
+  19, "VECSXP",
+  20, "EXPRSXP",
+  21, "BCODESXP",
+  22, "EXTPTRSXP",
+  23, "WEAKREFSXP",
+  24, "RAWSXP",
+  25, "S4SXP",
+  30, "NEWSXP",
+  31, "FREESXP",
+  99, "FUNSXP"
+) %>%
+  mutate(
+    type_name=str_replace(type_name, "SXP", "")
+  )
+```
+
+```{r}
+SE1 <- left_join(SE, SEXP_TYPES, by="type") %>%
+  select(-type) %>%
+  rename(type=type_name) %>%
+  filter(event != "rvl") %>%
+  mutate(
+    stype=case_when(
+      type == "CLO" ~ "Closure",
+      type %in% c("STR","INT","LGL","RAW","REAL", "CPLX") ~ "Basic",
+      type %in% c("ENV") ~ "Environment",
+      type %in% c("VEC") ~ "List",
+      type %in% c("SYM","LANG","EXPR") ~ "Expression",
+      type %in% c("NIL") ~ "NULL",
+      type %in% c("S4") ~ "S4",
+      TRUE ~ "Other"
+    )
+  )
+
+cs_count(SE1, stype)
+
+class_type <- 
+  SE1 %>%
+  filter(event=="asn") %>%
+  count(eval_call_srcref, stype, class, wt=N) %>% 
+  count(class, stype)
+
+cbp1 <- c("#999999", "#E69F00", "#56B4E9", "#009E73",
+          "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
+
+#class_type %>%
+class_types_sites <-
+  SE1 %>%
+  filter(event=="asn") %>%
+  cs_count(eval_call_srcref, class, stype, wt=N) %>% cs_count(class, stype)
+
+class_types_sites %>%
+  ggplot(aes(x=class, y=n, fill=stype)) + 
+  geom_bar(stat="identity", position="fill", alpha=0.9) +
+  scale_y_continuous(labels=scales::percent) +
+#  scale_fill_viridis(discrete = TRUE, option="E") +
+  scale_fill_manual(values = cbp1) +
+  theme(text = element_text(size=19),
+        axis.title.x = element_blank(),
+        axis.text.x = element_text(angle = 45, hjust = .5),
+        axis.title.y = element_blank(),
+        aspect.ratio = 4/3,
+        panel.grid.minor.x = element_blank(),
+        legend.position="top") +
+  labs(fill="")
+
+ggsave(path(PLOT_DIR, "se-types.pdf"))
+```
+
+```{r}
+types_sites <- 
+  SE1 %>%
+  filter(event=="asn") %>%
+  cs_count(eval_call_srcref, class, stype, wt=N) %>% cs_count(stype)
+
+closure_sites <- 
+  SE1 %>%
+  filter(event=="asn") %>%
+  cs_count(eval_call_srcref, class, stype, wt=N) %>%
+  filter(stype=="Closure") %>% 
+  cs_count(class)
+
+overview_table(
+  r("Basic type ratio", percent(filter(types_sites, stype == "Basic")$p/100)),
+  r("List type ratio", percent(filter(types_sites, stype == "List")$p/100)),
+  r("Closure type ratio", percent(filter(types_sites, stype == "Closure")$p/100)),
+  r("Closure type", filter(types_sites, stype == "Closure")$n),
+  r("Closure type local", filter(closure_sites, class == "Local")$n)
+)
+```
+
 ## Side effects with norm
 
-```{r}
+```{r eval=F}
 NE <- read_csv(

### 2021-04-16 paper/main.tex
Updates

--- 2021-04-16 paper/main.tex
@@
 
-From the \packageNbrunsRnd programs we capture \SEAll side effects from
-\SEAllCalls \eval calls in \SEAllSites sites. Again, the challenge is to remove
-the accidental side-effects caused by the R virtual machine implementation and
-are not related to the user code. For example, the \c{.Random.seed} variable,
-which contains the state of the random number generator is saved and restored
-from and to the global environment every time the user calls one of the base
+From the \packageNbrunsRnd programs we capture \SEAllRnd side effects from
+\SEAllCallsRnd \eval calls (\SEUserCallsToAllRatio) in \SEAllSites sites
+(\SEUserSitesToAllRatio). Again, the challenge is to remove the accidental
+side-effects caused by the R virtual machine implementation and are not
+related to the user code. For example, the \c{.Random.seed} variable, which
+contains the state of the random number generator is saved and restored from
+and to the global environment every time the user calls one of the base
 routines that use random numbers.
@@
 %
-Removing them leaves us with \SEUser side effects (\SEUserRatio) from
-\SEUserCalls \eval calls (\SEUserCallsRatio) in \SEUserSites sites
-(\SEUserSitesRatio). These sites are part of \SEUserFunctions in
-\SEUserPackages packages. \SEFunsNighty functions are responsible for 90\% of
-side effects. Half of the side-effects come just from three functions:
+Removing them leaves us with \SEUserRnd side effects (\SEUserRatio) from
+\SEUserCallsRnd \eval calls (\SEUserCallsRatio) in \SEUserSites sites
+(\SEUserSitesRatio). These sites are part of \SEUserFunctions R functions
+(\SEUserFunctionsToAllRatio of all functions doing \eval) in \SEUserPackages
+packages. From these function, \SEFunsNighty are responsible for 90\% of side
+effects. Half of the side-effects come just from three functions:
 \c{plyr::allocate\_column} (allocates space for a new data frame column),
@@
 effects happen (\emph{target environment}). It follows the classification
-described in Section~\ref{sec:env} with two extra classes: \emph{Object}, which
-represents a side-effect that happens in S4 or R6 object environment, and
+described in Section~\ref{sec:env} with two extra classes: \emph{Object},
+which represents a side-effect that happens in S4 or R6 object environment, and
 \emph{Local} that corresponds to a side effect in local variables of the \eval
-caller. The table shows data for both \eval sites and functions. For a function
-to have a given target environment class, all of its \eval sites that do side
-effects must have the same target environment class The majority of the \eval
-sites (\SESitesInOneClass) do all side effects consistently in one environment
-class. The same happens on the function level. Almost half of the sites (over a
-third of the functions) do side effects in either \emph{Local} or \emph{Object}
-environments. This gives a ray of hope for the compiler. Even though it is
-possible to do anything anywhere, the data suggest that most side-effects are
-sane.
+caller. The table shows data for both \eval sites and functions. For a
+function to have a given target environment class, all of its \eval sites that
+do side effects must have the same target environment class The majority of the
+\eval sites (\SESitesInOneClass) do all side effects consistently in one
+environment class. The same happens on the function level. Almost half of the
+sites (over a third of the functions) do side effects in either \emph{Local} or
+\emph{Object} environments. This gives a ray of hope for the compiler. Even
+though it is possible to do anything anywhere, the data suggest that most
+side-effects are sane.
 
@@
 local scope or less often, when called through an alias, in the global scope.
-In R, given the support for first-class environments, it can happen anywhere.
-The data however suggest that most of side effects from \eval happen in a
-predictable environment.
+In R, given the support for first-class environments, it can happen anywhere
+making it \eval more dangerous than it already is. The data however suggest
+that first, side-effects from \eval is not as wide spread as in the case of
+JavaScript and that over half of them happen in a predictable environment.
 
@@
 
-\todo{Aviral's story when he contacted a developer about his use of eval}
+% \todo{Aviral's story when he contacted a developer about his use of eval}
 

### 2021-04-16 paper/main.tex
Expand conclusion

--- 2021-04-16 paper/main.tex
@@
 the behavior of the program at that site. Luckily, \packageNbOneMinimizedPercent
-of sites only see a single form. There are few sites that are highly polymorphic
-(8 or more different forms), these include the pipe operator of the \c{magrittr}
-package which is used to compose functions.
+of sites only see a single form.
+In other words, a JIT compiler could correctly predict the \emph{shape} of an expression by looking just at the first call in \packageNbOneMinimizedPercent of the sites.
+There are few sites that are highly polymorphic (8 or more different forms), these include the pipe operator of the \c{magrittr} package which is used to compose functions.
 
@@
 kinds of environments, but in \packageNbOneCategoryEnvirSitePercent of the case a
-site has a single kind. It means that we can predict the \emph{shape} of the expression passed to \eval in \packageNbOneCategoryEnvirSitePercent of the sites.
+site has a single kind. It means that we can predict the \emph{kind} of the expression passed to \eval in \packageNbOneCategoryEnvirSitePercent of the sites.
 
@@
 modify anything anywhere which is wary for both R developers and R compiler.
-% TODO expressions passed to eval
-Even more so in the case of side effects that can quite literally happen
-anywhere. The data however suggest that this is not the case. The majority of
-eval sites do not do any observable side effects and when they do, it happens
-in a predictable environment.
+The shapes of expressions passed to \eval are extremely diverse, ranging from simple variable lookups to complex blocks or creation of new functions.
+Luckily, most site only see a single shape.
+\eval unlashes its power and shows up its singularity compared to \eval in other languages with the diversity of environments in which it can evaluate its expressions.
+While most of the time, the \eval environment is the local environment, it is also often the global environment, a synthetic environment, or even a package environment.
+Luckily, as for the expressions passed to \eval, the environment passed to \eval for one given site is easily predictable with the first call to the site.
+
+The expressive power of \eval shines even more  so in the case of side effects that can quite literally happen anywhere.
+The data however suggest that this is not the case.
+The majority of eval sites do not do any observable side effects and when they do, it happens in a predictable environment.
+
+The origins of the expression passed to \eval, either constructed from language expressions, or stemming from reflection, or built by parsing strings.
+Reflection and constructed expressions dominate, which suggest the main use of \eval is meta-programming.
+% macros...
 
@@
 analysis. The data unfortunately suggests that replacing \eval is going to be
-lengthly journey. Even if we disregard core packages as they are part of the
+a lengthy journey. Even if we disregard core packages as they are part of the
 language and the \eval use is stable, in CRAN \eval is used widely and in more

### 2021-04-16 paper/main.tex
Some love for Section 6

@@
 
-\todo{Illustrate things that we have discussed in Section 5 - examples}
-\todo{Things to discuss: model.frame, match.call, subset as example of NSE, substitute, parse, plots}
-
-% We present here the main uses of \eval we have witnessed in R.
-Sometimes, they are design patterns, sometimes, \eval could be replaced by more specific functions.
-For each use of \eval, we discuss whether it is necessary to use.
-
-% \subsection{Variable lookup}
-
-% \eval is used to look-up the value of a variable in a given environment, different from the current scope so the \c{envir} argument of \eval is explicitly specified.
-
-% \subsection{Metaprogramming}
-
-% \mypara{\c{substitute}}
-
-% \mypara{\c{match.call}}
-
-% \todo{Talk about model.frame}
-
-% \subsection{Logging}
-
-% \subsection{Code transform}
-
-% % Derivation of symbolic expressions
-
-% \subsection{Plotting}
-
 % \section{\eval expressivity and macros}
@@
 		hickness <- mlist$thickness[eval(i), eval(j), drop = FALSE]
-		...
 	}
@@
  expr<-expression(beta_1 + time*beta_2+u)
- ...
  mu = data.frame(beta_1=0.5, beta_2=1, time=3, u=2.3)
- ...
  D[i]<-eval(DD(expr, name[unlist(comb.pos[i,])]), mu)

### 2021-04-16 paper/main.tex
Conclusion: advocating for macros in R

@@
 
-% \section{\eval expressivity and macros}
-
-% The Thomas Lumley's article, Programmer's Niche: Macros in R, (https://www.r-project.org/doc/Rnews/Rnews_2001-3.pdf) describes how to write a `defmacro` function that would like as a macro creator, using `substitute` and `eval`.
-There are no local macro variables in this implementation though.
-
 The R language was intended to be extensible. The combination of lazy
@@
 Similarly to \citep{ecoop11}, we started looking into \evals in R with the
-ambition to replace them by other feature that are more amenable to static
+ambition to replace them by other features that are more amenable to static
 analysis. The data unfortunately suggests that replacing \eval is going to be
@@
 
+Even though we reckon that \eval has more expressive power than macros, as it can access runtime information such as its call stack, adding a macro system to R to replace \eval for all meta-programming purposes would help static analysis as macros can be expanded at analysis-time.
+
+% It's dubious whether R programmers would develop with macros though.
+% Macros and R
+% The Thomas Lumley's article, Programmer's Niche: Macros in R, (https://www.r-project.org/doc/Rnews/Rnews_2001-3.pdf) describes how to write a `defmacro` function that would like as a macro creator, using `substitute` and `eval`.
+There are no local macro variables in this implementation though.
+
 \bibliography{bib/bibliography,bib/jv}

### 2021-04-16 paper/main.tex
Fix in intro
Proofreading the introduction with grammarly

@@
   reasons to believe that reflective feature usage is language and application
-  domain specific; we focus on data science code written in R, and compare our
+  domain-specific; we focus on data science code written in R and compare our
   results to previous work that analyzed web programming in JavaScript. This
@@
 boogeyman as it gets. Yet, for McCarthy, \eval was simply the way to
-write down the definition of LISP, he was surprised that someone coded
-it up and offered it to end users. Since then, reflective facilities
+write down the definition of LISP; he was surprised that someone coded
+it up and offered it to end-users. Since then, reflective facilities
 have been used to parameterize programs over code patterns that can be
@@
 binding as the behavior of any particular call to \eval will only be
-known when the program is run and that particular call site is
+known when the program is run, and that particular call site is
 evaluated.
@@
 \vspace{2mm}\noindent\emph{Trouble in Paradise.} Reflective facilities
-hinder most attempts to reason about, or apply meaning-preserving
-transformation to, the code using them. In practice, \eval causes
-static analysis techniques to loose so much precision as to become
-pointless. For compilers, anything but the most trivial, local,
-optimizations are unsound after a use of \eval. Furthermore, the
+hinder most attempts to reason about or apply meaning-preserving
+transformations to the code using them. In practice, \eval causes
+static analysis techniques to lose so much precision as to become
+pointless. For compilers, anything but the most trivial, local
+optimizations are unsound after the use of \eval. Furthermore, the
 addition of arbitrary code --- code that could have been obtained from
@@
 an over-approximation of the set of possible behaviors exhibited by
-the program under study~\cite{cc77}. A reflective may have \emph{any}
-behavior that can be expressed in the target language; i.e. \eval can
-be replaced by any legal sequence of instructions. As dynamic
+the program under study~\cite{cc77}. A reflective facility may have \emph{any}
+behavior that can be expressed in the target language, \ie any legal sequence of instructions can replace \eval. As dynamic
 languages tend to be permissive, the analysis has to, for example,
 assume that many (or all) functions in scope may have been redefined,
-e.g. that \texttt{`+`} now opens a network connection or something
+\eg that \texttt{`+`} now opens a network connection or something
 equally surprising. A single occurrence of \eval causes the static
-analyzer to loose all information about program state and
+analyzer to lose all information about the program state and
 meaning of identifiers. This loss of precision can sometimes be
 mitigated by analyzing the string argument~\cite{moller03} to bound its
-possible behavior but when the string comes from outside the program
-not much can be done. A frustrated group researchers argued giving up
+possible behavior, but when the string comes from outside the program, 
+not much can be done. A frustrated group of researchers argued giving up
 on soundness and, instead, under-approximating dynamic features
-(soundiness)~\cite{soundy}. In their words ``a practical analysis,
-therefore, may pretend that \eval does nothing, unless it can
+(soundiness)~\cite{soundy}. In their words, ``a practical analysis,
+therefore, may pretend that \eval does nothing unless it can
 precisely resolve its string argument at compile time.'' Alas,
-assuming that \eval does not have side-effects, or that side-effects
-will not affect the results of the analysis, may be unduly optimistic.
+assuming that \eval does not have side-effects or that side-effects
+will not affect the results of the analysis may be unduly optimistic.
 
@@
 In 2011, 82\% of the 10,000 most accessed sites used \eval~\cite{ecoop11}. Yet,
-the strings passed to \eval, and their behaviors when executed, are far from
+the strings passed to \eval, and their behaviors, when executed, are far from
 random; it was shown that when one can observe several calls to \eval, the
@@
 \vspace{2mm}\noindent\emph{The Here and Now.} In this study, we investigate the
-usage of \eval in programs written in the R programming language. R is language
+usage of \eval in programs written in the R programming language. R is a language
 designed by statisticians for applications in data science~\cite{r,R96}. What
@@
 are dynamic, they are quite different. While one can program in an
-object-oriented style in R, like in JavaScript, R is primarily a lazy, untyped,
-functional language. JavaScript was designed to run untrusted code in browser,
-while R is used for statistical computing on desktops. JavaScript is a general
-purpose language used by a wide community of programmers; while R is used for
+object-oriented style in R like in JavaScript, R is primarily a lazy, untyped,
+functional language. JavaScript was designed to run untrusted code in a browser,
+while R is used for statistical computing on desktops. JavaScript is a general-purpose language used by a vast community of programmers, while R is used for
 scientific computing by data scientists and domain experts with, often, limited
@@
 of this fact. Library developers, on the other hand, know and program
-defensively around laziness.} Our goal is thus to highlight the differences in
-usage between JavaScript and R, and try to explain those differences in terms of
-language features, application domain and programmer experience. Hopefully some
+defensively around laziness.} Thus, our goal is to highlight the differences in
+usage between JavaScript and R and try to explain those differences in terms of
+language features, application domain and programmer experience. Hopefully, some
 of our observations will generalize to other languages.
@@
 choosing R is that every package in the CRAN repository is curated and comes
-with examples of typical usage. This gives us a large code base that we can
-analyze dynamically. To observe \eval we built a two-level monitoring
+with examples of typical usage. This gives us a large codebase that we can
+analyze dynamically. To observe \eval, we built a two-level monitoring
 infrastructure:\footnote{Our infrastructure is open source and publicly
@@
 user-visible properties of R programs --- but we can also monitor the
-inner-workings of the R interpreter --- this allows us to capture details not
-exposed at the source level. Dynamic analysis is limited, it can only observe
+inner workings of the R interpreter --- this allows us to capture details not
+exposed at the source level. Dynamic analysis is limited; it can only observe
 behaviors triggered by the particular inputs passed to a program. Luckily, CRAN
-libraries come with many tests and use-cases. The choice of corpus is crucial.
+libraries come with many tests and use-cases. The choice of the corpus is crucial.
 Our corpus has been constructed to reflect the levels of sophistication of the R
@@
 sample data) and \emph{Kaggle scripts} (\KaggleUnique end-user written programs
-that performs a particular data analysis task). It is reasonable to expect that
-\eval usage differ between these datasets: the libraries represent a lively
-ecosystem with new libraries added each day, while end user code is often thrown
+that perform a particular data analysis task). It is reasonable to expect that
+\eval usage differs between these datasets: the libraries represent a lively
+ecosystem with new libraries added each day, while end-user code is often thrown
 together, run once, and never revisited.

### 2021-04-16 paper/main.tex
Grammarly's proofreading of the expression of eval

@@
 \eval. We use the word \emph{site} to an occurrence of a call site to the \eval
-function in the source code, and \emph{call} to denote an observed invocation of
+function in the source code and \emph{call} to denote an observed invocation of
 the \eval function.
@@
 vectors). The rest trivially evaluate to a value, \eg~\c{1+1}.\footnote{True as
-long as base functions such as \c{+} are not redefined. They typically aren't,
+long as base functions such as \c{+} are not redefined. They typically are not,
 but this is a limitation of this categorization.} In our corpus,
@@
 \end{lstlisting}
-where the argument to \eval is a string which \eval simply returns. From twenty
+where the argument to \eval is a string which \eval simply returns. Of twenty
 randomly selected \evals, \packageUsefulValueEvalPercent of them  need % There are a majority which needs eval
 \eval for some inputs. Either we are dealing with a value that needs to be
-constructed dynamically or the value is a default case that sometimes is
+constructed dynamically, or the value is a default case that sometimes is
 replaced by a more interesting expression. This form is usually evaluated in few
-interpreter steps, in fact the median is only
+interpreter steps; in fact, the median is only
 \packageMinimizedmedianoperationscRnd. The environment in which they evaluate is
@@
 
-\medskip\noindent\EE{$min(e)=\c{X}$} Variable lookups are the most common form,
-they are found in 28\% of the sites. This form includes simple variable reads,
-\eg~\c{x}, those are \packageNbSymbolVarSitePercent of \c X. The form also
+\medskip\noindent\EE{$min(e)=\c{X}$} Variable lookups are the most common form; they are found in 28\% of the sites. This form includes simple variable reads,
+\eg~\c{x}, which are \packageNbSymbolVarSitePercent of \c X. The form also
 subsumes \c{V}, so it includes a mixture of arithmetic expressions,
@@
 it is not the common case. Lookups are often evaluated in constructed
-environments, as few as \packageMinimizedpercentparentframesc of these
+environments; as few as \packageMinimizedpercentparentframesc of these
 expressions are evaluated in a function environment.
@@
 with the dollar operator, \eg~\c{x\$f}, and vector indexing, \eg~\c{x[42]} or
-\c{x[[24]]}. As with \c X we allow arithmetics and values in this form. Lookup
+\c{x[[24]]}. As with \c X, we allow arithmetics and values in this form. Lookup
 occurs in 6\% of the sites. The interpreter evaluates
@@
 environment {\tt <\,\!<-}, the \c{assign} function and the \c{\$} form. Assignments occur in 5\% of the
-sites. They represent the most obvious source of side-effects. The median number
+sites. They represent the most obvious source of side effects. The median number
 of operations is \packageMinimizedmedianoperationsiRnd; the minimum is 3.
@@
 \medskip\noindent\EE{$min(e)=\c{F()}$} This form captures simple function calls,
-\eg~\c{f(2)}, with neither variables nor assignments. Or, more specifically it
+\eg~\c{f(2)}, with neither variables nor assignments. More specifically, it
 allows for variables in the function position but not in arguments. Usually,
@@
 if the function is returned by a promise or if the function name is shadowed by
-a promise containing a value, then computation will occur. This form occurs in
+a promise containing a value, then the computation will occur. This form occurs in
 6\% of sites and typically does not perform much work in the interpreter.
@@
 respectively, \packageMinimizedmedianoperationsdRnd and
-\packageMinimizedmedianoperationsbRnd. The \packageMinimizedpercentparentframesj
+\packageMinimizedmedianoperationsbRnd. Also,  \packageMinimizedpercentparentframesj
 of these expressions run in function environments.
@@
 \packageFunctionDefinitionSitesPercent of sites. Evaluating a function
-definition is done in 2 interpreter step; the data does not record the work
+definition is done in 2 interpreter steps; the data does not record the work
 performed by the interpreter when the generated functions are eventually run. In
 addition to \c{FUN}, \packageGeneralizedFunctionDefinitionSitesPercent of sites
-have function definitions nested in other expressions, this gives an idea of the
+have function definitions nested in other expressions. This gives an idea of the
 use of higher-order functions.
@@
 \medskip\noindent\EE{$min(e)=\c{BLOCK}$} This form captures multi-statement
-code blocks, which occur in only 4\% of sites. These are larger expressions, we
-do not inspect contents of the blocks. The median number of executed operations
+code blocks, which occur in only 4\% of sites. These are huge expressions; we
+do not inspect the contents of the blocks. The median number of executed operations
 is \packageMinimizedmedianoperationsjRnd. They typically run in function
@@
 \medskip\noindent\EE{$min(e)=\c{model.frame}$} The \c{model.frame} function
-returns a dataframe that is the result of fitting the model described in a given
-formula. This form subsumes \c{F(F(X))}, \c{FUN} and assignments. It is the
-single most popular function invoked from \eval, it occurs in 7\% of the sites.
+returns a dataframe resulting from f fitting the model described in a given
+formula. This form subsumes \c{F(F(X))}, \c{FUN}, and assignments. It is the
+single most popular function invoked from \eval; it occurs in 7\% of the sites.
 Each call does quite a lot of work with a 2K instructions median.
@@
 forms any given site sees. The more forms, the harder it will be to characterize
-the behavior of the program at that site. Luckily, \packageNbOneMinimizedPercent
+the program's behavior at that site. Luckily, \packageNbOneMinimizedPercent
 of sites only see a single form.
 In other words, a JIT compiler could correctly predict the \emph{shape} of an expression by looking just at the first call in \packageNbOneMinimizedPercent of the sites.
-There are few sites that are highly polymorphic (8 or more different forms), these include the pipe operator of the \c{magrittr} package which is used to compose functions.
+Few sites are highly polymorphic (8 or more different forms).
+These include the pipe operator of the \c{magrittr} package, which is used to compose functions.
 
 \paragraph{Discussion} The variety of uses of \eval is evidenced by the number
-of different forms observed in CRAN. In comparison, JavaScript \eval usage was
-simpler and more predictable as reported by \citet{oopsla12b}. Nevertheless,
-simple forms dominate and there are many cases where \eval could be replaced
-by less powerful constructs.
+of different forms observed in CRAN.
+In comparison, JavaScript \eval usage was more straightforward and more predictable as reported by \citet{oopsla12b}.
+Nevertheless, simple forms dominate, and there are many cases where \eval could be replaced by less powerful constructs.
 

### 2021-04-16 paper/main.tex
Grammarly's proofreading of Section V.2 (environments)

@@
 \item \emph{Synthetic:} environments built from data structures such as lists,
-  dataframes, or constructed explicitly with \c{new.env}, \c{list2env} or
-  \c{as.environment}. Also includes the empty environment.
+  dataframes, or constructed explicitly with \c{new.env}, \c{list2env}, or
+  \c{as.environment}. It also includes the empty environment.
 \item \emph{Global:}  environments in which scripts or interactive commands
@@
 Thus 0 is the direct caller and corresponds to local variables; 1 is its parent, and so on.
-The data suggests that in 81\% of cases, \eval accesses its caller's environment -- this means, the variables of the function where \eval textually
+The data suggests that in 81\% of cases, \eval accesses its caller's environment -- this means  the variables of the function where \eval textually
 occurs are read and written to. It is interesting that some 1.5\% of sites
 evaluate a code in an environment that is three frames or above from \eval. This
-distance implies that, in general, modular reasoning is impossible, to
-understand the behavior of any piece of code entails fully understanding the
+distance implies that, in general, modular reasoning is impossible. Understanding the behavior of any piece of code entails fully understanding the
 behavior of all functions that the code may call as the actions of \eval may
-happen at a distance. Finally note that any particular site may have several
-kinds of environments, but in \packageNbOneCategoryEnvirSitePercent of the case a
-site has a single kind. It means that we can predict the \emph{kind} of the expression passed to \eval in \packageNbOneCategoryEnvirSitePercent of the sites.
+happen at a distance. Finally, note that any particular site may have several
+kinds of environments, but in \packageNbOneCategoryEnvirSitePercent of the cases,  a site has a single kind.
+It means that a JIT compiler can predict the \emph{kind} of  expression passed to \eval in \packageNbOneCategoryEnvirSitePercent of the sites by just examining at the first invocation of a site.
 
@@
 intentional and accidental use. Direct references to the top-level, using
-\c{globalenv()} or \c{.GlobalEnv}, are rare, they occur in only
+\c{globalenv()} or \c{.GlobalEnv}, are rare; they occur in only
 \packageNbExplicitGlobalSites sites. We suspect that the majority of uses of the
-global environment come from the fact that our corpus consists of scripts that
-are loaded at the top level, thus that environment may often be the caller of
+global environment come from the fact that our corpus consists of scripts loaded at the top level. Thus, that environment may often be the caller of
 \eval or close to it. The reason we make this point is that values stored in the
 global environment are visible to all functions and are not reclaimed by the
-garbage collector. So accidental uses may pollute that name space.
+garbage collector. So accidental uses may pollute that namespace.
 
@@
 obtained. The expected case is that the environment was the one obtained from a
-promise; as promises combine code and the environment in which that code
+promise, as promises combine code and the environment in which that code
 originates. Likely less frequent are cases where \eval is provided the results
-of programmatically selecting some call stack. ({\it Synthetic}) The relative
+of programmatically selecting some call stack. ({\it Synthetic}) The relatively
 high frequency of synthetic environments corresponds to cases when one wants to
@@
 structure as an environment. ({\it Global}) The fact that the global environment
-shows up so frequently is likely an artifact of the way the code is run. ({\it
-  Package}) There are only 92 sites that use a package environment in \eval,
-this is probably best as it is bad form to add bindings or modify functions
-of a loaded package. Many more sites rather wrap a package environment in a new
+shows up so frequently is likely an artifact of how the code is run. ({\it
+  Package}) There are only \packageNbPackageNamespaceEnvSites sites that use a %77 sites
+ package environment in \eval.
+This is probably best as it is an improper form of adding bindings or modifying functions of a loaded package.
+Many more sites instead wrap a package environment in a new
 environment to create singletons in a package.

### 2021-04-16 paper/main.tex
Grammarly's proofreading of Section V.3 (origins)

@@
 Where does the expression passed to \eval come from? There are various means of
-creating that expression, these are associated with particular use cases, we
-classify them in three categories:
+creating that expression, which are associated with particular use cases. We
+classify them into three categories:
 
@@
   \c{quote}, \c{enquote}, \c{expression}. Function arguments are passed as
-  promises and \c{substitute} is used to retrieve the source expression
+  promises, and \c{substitute} is used to retrieve the source expression
   associated with the promise.
@@
 \item {\it String:} Finally, expressions can be created from strings by invoking
-   \c{parse}, \c{str2expression} or \c{str2lang}.
+   \c{parse}, \c{str2expression}, or \c{str2lang}.
 \end{compactitem}
@@
 This data is obtained by dynamically tainting values as they are produced by the
-various sources, due to technical reasons, there is some imprecision in the
+various sources. Due to technical reasons, there is some imprecision in the
 results. In particular, we are not able to classify all sites. Manual inspection
-of numerous examples suggest that the classified results are accurate.
+of numerous examples suggests that the classified results are accurate.
 
@@
 input, parses it, and passes it to \eval. The use of strings seems to correlate
-with less sophisticated programmers; in our Kaggle corpus
-\kaggleParseExprsSitePercent of site use strings.
+with less sophisticated programmers; in our Kaggle corpus,
+\kaggleParseExprsSitePercent of sites use strings.
 
@@
 The origin data suggests that constructing expressions from strings is a
-minority of the use cases, instead the constructed category shows that the
-majority of \evals comes from code that was processed by the compiler, and may be
+minority of the use cases. Instead, the constructed category shows that most \evals comes from code that was processed by the compiler, and may be
 slightly modified by the programmer before invoking \eval. Both constructed and

### 2021-04-16 paper/main.tex
Grammarly's proofreading of Section 4 (Usage metrics)

@@
 of packages calling \eval is \PkgPackagesRatio. Over half of these packages
-have fewer than 3 sites, and with the exception \MaxEvalCallSitesPackage which
+have fewer than 3 sites, and with the exception \MaxEvalCallSitesPackage, which
 has \MaxEvalCallSitesCount sites, all packages contain fewer than
-\MaxEvalCallSitesRest sites (\cf Fig~\ref{fig:pkgs-eval-callsites-hist} which
+\MaxEvalCallSitesRest sites (\cf Fig~\ref{fig:pkgs-eval-callsites-hist}, which
 shows a histogram of sites per package). These sites appear in \PkgFunsWithEval
@@
 \PkgHitEvalCallSitesAvgRatio of sites, a coverage similar to the code coverage
-metric for the packages which is \PkgCodeCoverage. The fact that not all sites
+metric for the packages, which is \PkgCodeCoverage. The fact that not all sites
 are exercised can be chalked down to incomplete tests and occasional analysis
@@
 exercised; coverage is unequal. Figure~\ref{freq} summarizes the frequency of
-dynamic calls to \eval with the left column being the number of calls and the
+dynamic calls to \eval, with the left column being the number of calls and the
 right, the number of packages in that range. There are \packageFewcalls packages
-with low \eval frequency, fewer than 100 calls; and a smaller, but still
+with low \eval frequency, fewer than 100 calls, and a smaller, but still
 significant, number of packages, \packageManycalls to be precise, that use \eval
@@
 Figure~\ref{tab:variantseval} summarizes the use of variants of \eval. For each
-of the four variants, the first rows shows the number of sites in the corpus
+of the four variants, the first row shows the number of sites in the corpus
 (\emph{static}), the second row is the number of sites that were encountered
 during analysis (\emph{exercised}), the last row is the number of calls
-(\emph{invocations}). The overwhelming majority of sites and calls, are to \eval
+(\emph{invocations}). The overwhelming majority of sites and calls are to \eval
 itself, \c{eval.parent} is rare, and both \c{evalq} and \c{local} are barely
@@
 Figure~\ref{cn}(a) shows normalized call counts per site; on the left are
-average number of calls from a given site and given run, on the right are counts
-of sites that fall in that range. For instance, \packageRunbinh sites are
+the average number of calls from a given site and given run, on the right are the counts of sites that fall in that range. For instance, \packageRunbinh sites are
 invoked 2,000+ times per run. Larger numbers suggest loops or recursive contexts
@@
 50 times or less. Figure~\ref{cn}(b) zooms in on low frequency sites. The x-axis
-shows normalized calls and the y-axis is the number of sites for that value. Most
-low-frequency \evals are invoked only once, about half as many are invoked
-twice; after that the frequency quickly drops.
+shows normalized calls, and the y-axis is the number of sites for that value. Most
+low-frequency \evals are invoked only once; about half as many are invoked
+twice; after that, the frequency quickly drops.
 
@@
 inheritance in \c{ggproto}, one of the many object-oriented systems in
-R that is used exclusively in the \c{ggplot2} graphics library.}
+R, used exclusively in the \c{ggplot2} graphics library.}
 
@@
 were dominated by the size of data objects which could range in the MBs. The
-median argument size is \packageMedianszeval (due to the symbols) and the
+median argument size is \packageMedianszeval (due to the symbols), and the
 average is \packageAvgszeval nodes. The largest \eval input observed is
@@
 To estimate the work performed in \evals, we count instructions executed by the
-interpreter. Most invocations perform relatively little work with
+interpreter. Most invocations perform relatively little work, with
 \packageSmalleventspct of \evals executing 50 or fewer instructions. The violin
 plot of Figure~\ref{ev}(a) corresponds to \evals executing $\leq$ 50
-instructions, it is dominated by trivial symbol look ups. Figure~\ref{ev}(b) has
-the work-intensive \evals which go all the way to \packageMaxeventsRnd
+instructions; it is dominated by trivial symbol lookups. Figure~\ref{ev}(b) has
+the work-intensive \evals, which go all the way to \packageMaxeventsRnd
 instructions.
@@
 \baseTriggeredevalpct of the \BaseEvalCallSites sites. Most of the calls
-(\baseEvalsratio) are to the \c{eval} function and most arguments
+(\baseEvalsratio) are to the \c{eval} function, and most arguments
 (\baseCodepercent) are expressions. However, unlike CRAN, the majority
 (\baseLanguagepercent) of the arguments are language objects. The median
-argument size is \baseMedianszeval, which is more than for CRAN and the maximum
-size is \baseMaxszeval, less than for CRAN. Small instruction counts ($<50$),
-amount for \baseSmalleventspct of calls. A single site in the \c{match.arg}
+argument size is \baseMedianszeval, which is more than for CRAN and the maximum size is \baseMaxszeval, less than for CRAN.
+Small instruction counts ($<50$) amount to \baseSmalleventspct of calls.
+A single site in the \c{match.arg}
 function is responsible for \baseTopFuncPercent of all recorded calls. This
@@
 The challenge dealing with Base is that every program uses it. It can thus
-generate extreme amounts of data but the data is quite predictable. For the
+generate extreme amounts of data, but the data is quite predictable. For the
 rest of the paper, we only mention Base when there are surprising
@@
 \kaggleStaticeval sites from \KaggleWithEvals, only \kaggleTriggeredeval sites
-in \kaggleNbruns scripts were hit. \KaggleFailedScripts scripts failed to run
-and the other did not exercised the \eval sites. This is partially expected
-as Kaggle code does not need to abide to any checks. Upon a manual inspection,
-we observed that indeed the failing scripts were of a poor quality, often not
-finished, using misspelled package names, hard-coded file paths or accessing
+in \kaggleNbruns scripts were hit; \KaggleFailedScripts scripts failed to run,
+and the other did not exercise the \eval sites. This is partially expected
+as Kaggle code does not need to abide by any checks. Upon a manual inspection,
+we observed that indeed the failing scripts were of poor quality, often not
+finished, using misspelled package names, hard-coded file paths, or accessing
 missing files.
@@
 Most call sites are invoked only once. Only one site is called \kaggleMaxcalls times.
-Expressions amount for \kaggleCodepercent of arguments. Unlike CRAN corpus,
+Expressions amount for \kaggleCodepercent of arguments. Unlike the CRAN corpus,
 there are very few symbols (\kaggleSymbolpercent). Most expressions result from
-calls to \c{parse}, thus most \evals start with strings. The median argument
+calls to \c{parse}; thus, most \evals start with strings. The median argument
 size is \kaggleMedianszeval, which makes sense, as few arguments are only
@@
 execute fewer than 50 instructions. Manual inspection of \eval usage in Kaggle
-suggest that it is consistent with the data obtained for CRAN. We do not
+suggests that it is consistent with the data obtained for CRAN. We do not
 discuss it further.
@@
 omnipresent in the relatively small Base library. The Base library is, in turn,
-used by every single package. To avoid having Base dominate the result we have
+used by every single package. To avoid having Base dominate the result, we have
 excluded calls occurring there from the data we report when analyzing packages.
@@
 regular and varied use of \eval. They represent our most interesting data set as
-these packages are the result of over 20 years of contributions by thousands of
+these packages result from over 20 years of contributions by thousands of
 authors. The code in packages is well maintained and relatively well tested.
-Finally, Kaggle scripts are often written by less sophisticated users, and
-likely perform simpler tasks, and thus have lesser need for \eval, and its uses
-originate from strings as these are easier to manipulate for end-users. Due to
-the relative paucity of data in Kaggle, we do not pursue the data set further.
+Finally, Kaggle scripts are often written by less sophisticated users and
+likely perform more straightforward tasks, thus having a lesser need for \eval, and its uses originate from strings as these are easier to manipulate for end-users.
+Due to the relative paucity of data in Kaggle, we do not pursue the data set further.
 

### 2021-04-16 paper/main.tex
Grammarly's proofreading for end of Section V (side effects)

@@
 bindings. From the recorded data, we ignore side effects coming from \eval
-sites from unit testing frameworks.\footnote{In the corpus we have \c{RUnit,
+sites from unit testing frameworks.\footnote{In the corpus, we have \c{RUnit,
 testthat, tinytest} and \c{unitizer} unit testing frameworks.} They run the
-testing code via \eval, and thus the results would be biassed because of the
+testing code via \eval, and thus the results would be biased because of the
 high number of code in these tests (\CranRunnableCodeTestsRnd lines).
@@
 (\SEUserSitesToAllRatio). Again, the challenge is to remove the accidental
-side-effects caused by the R virtual machine implementation and are not
+side effects caused by the R virtual machine implementation and are not
 related to the user code. For example, the \c{.Random.seed} variable, which
-contains the state of the random number generator is saved and restored from
+contains the state of the random number generator, is saved and restored from
 and to the global environment every time the user calls one of the base
@@
 (\SEUserFunctionsToAllRatio of all functions doing \eval) in \SEUserPackages
-packages. From these function, \SEFunsNighty are responsible for 90\% of side
+packages. From these functions, \SEFunsNighty are responsible for 90\% of side
 effects. Half of the side-effects come just from three functions:
@@
 Section~\ref{sec:eval-in-r}), \SESitesNotInEnvirRatio modifies other
-environments, and finally \SESitesBothEnvirRatio does both.
+environments, and finally, \SESitesBothEnvirRatio does both.
 Table~\ref{tab:se-env} shows the class of the environment where \eval side
@@
 Table~\ref{tab:se-types} shows the proportion of the different side effects
-that were recorded. In terms of calls, we see primarily assignments and in
+that were recorded. In terms of calls, we see assignments primarily and in
 terms of site definitions. This is expected. A subsequent \eval call will turn
-a definition into an update. The most dangerous side effect is variable removal
+a definition into an update. The most dangerous side effect is variable removal,
 as it means that after a call to \eval some binding in some environments will
@@
 many sites changes function bindings. In the corpus, this happens in only
-\SEClosureType sites, \SEClosureTypeLocal of them do that in the local
+\SEClosureType sites; \SEClosureTypeLocal of them do that in the local
 environment. We have manually inspected a few of these sites, but except for
@@
 local scope or less often, when called through an alias, in the global scope.
-In R, given the support for first-class environments, it can happen anywhere
-making it \eval more dangerous than it already is. The data however suggest
-that first, side-effects from \eval is not as wide spread as in the case of
+In R, given the support for first-class environments, it can happen anywhere, 
+making it \eval more dangerous than it already is. However, the data  suggest
+that first, side-effects from \eval are not as widespread as in the case of
 JavaScript and that over half of them happen in a predictable environment.
@@
 The R language was intended to be extensible. The combination of lazy
-evaluation, \c{substitute} and \c{eval} are the tools given to
+evaluation, \c{substitute}, and \c{eval} are the tools given to
 developers to this end. This API is slightly more complex than just

### 2021-04-16 paper/main.tex
Expression of eval: comparison with JS

@@
 of different forms observed in CRAN.
-In comparison, JavaScript \eval usage was more straightforward and more predictable as reported by \citet{oopsla12b}.
-Nevertheless, simple forms dominate, and there are many cases where \eval could be replaced by less powerful constructs.
+In comparison, JavaScript \eval usage was more straightforward and more predictable as reported by \citet{oopsla12b}, with 98.6\% of the sites with only one form.
+Nevertheless, simple forms dominate in R also, and there are many cases where \eval could be replaced by less powerful constructs.
 

### 2021-04-16 paper/main.tex
JS and environment of eval

@@
 of different forms observed in CRAN.
-In comparison, JavaScript \eval usage was more straightforward and more predictable as reported by \citet{oopsla12b}, with 98.6\% of the sites with only one form.
+In comparison, JavaScript \eval usage was more straightforward and more predictable as reported by \citet{oopsla12b}, with 98.7\% of the sites with only one form.
 Nevertheless, simple forms dominate in R also, and there are many cases where \eval could be replaced by less powerful constructs.
@@
 
+Contrary to the Javasscript \eval, which does not precise the scope in which to evaluate,the second argument of \eval in R, \c{envir}, specifically is dedicated ot that. 
 The environment in which evaluation happens determines what is visible to the

### 2021-04-16 paper/main.tex
Side effects: numbers to compare with javascript

@@
 \paragraph{Discussion}
+
+% Compare to JS ? Everything comes from a string in JS anyway
+% Composite (concatenation of strings), DOM and constant are the most common
+% ...
 The origin data suggests that constructing expressions from strings is a
@@
 that first, side-effects from \eval are not as widespread as in the case of
-JavaScript and that over half of them happen in a predictable environment.
+JavaScript, \footnote{\citep{ecoop11} shows that in the \emph{Interactive} scenario, \eval in Javascript performs, store events can reach up to 40\% of the events, and 7\% to 8\% of the \eval do side effects in the global scope.
+} and that over half of them happen in a predictable environment.
 

### 2021-04-16 paper/main.tex
arly's proofrading of Section 2 (background)

@@
 on top of the core language. Most data types are sequences of primitive values
-constructed by calling the combination function \c{c}, for instance
+constructed by calling the combination function \c{c}. For instance,
 \c{c("Ha","bye")} evaluates to a vector of two strings, constants such as \c{42}
@@
 through multiple aliases are copied when written to. Furthermore, values can be
-tagged by attributes, these are key-value pairs. For instance, the attribute
+tagged by attributes; these are key-value pairs. For instance, the attribute
 \c{dim}-\c{c(2,2)} can be attached to the value at \c{x} by the call
@@
 call-by-need convention for passing arguments to functions. Each argument is a
-thunk composed of an expression, its environment, and a slot for the result,
+thunk composed of an expression, its environment, and a slot for the result;
 these are called \emph{promises}. To get the value of an argument, the
-corresponding promise must be forced, once forced the promise's result is cached
+corresponding promise must be forced. Once forced, the promise's result is cached
 for future use.
@@
 choices affect the expressive power of \eval. The key choices are the input
-format, the environment in which generated code is evaluated and the reflective
+format, the environment in which generated code is evaluated, and the reflective
 operations available to that code. Fig.~\ref{comp} summarizes designs.
@@
 
-The choice of the environment of \eval is important as it determines how much of
+The choice of the environment of \eval is essential as it determines how much of
 a program \eval can observe as well as the reach of potential side-effects
-performed by that operation. The most restrictive semantics is that of Java
+performed by that operation. The most restrictive semantics is that of Java, 
 where newly loaded code evaluates in the environment defined by the classes
-visible from the current classloader. Julia and JavaScript's strict mode, limit
+visible from the current classloader. Julia and JavaScript's strict mode limit
 \eval to the global environment. Finally, R is the most flexible as any
@@
 \eval. The main difference between languages lies in how much of the state of a
-program is accessible through reflective operations. Julia, Java and JavaScript
+program is accessible through reflective operations. Julia, Java and, JavaScript
 all allow some form of introspection on the data. R is more flexible as it is
-possible to inspect the program's call stack. Thus any environment in the
+possible to inspect the program's call stack. Thus, any environment in the
 program can be inspected and modified.
@@
 to expose as much of the language and its internals as possible in order to
-maximize expressivity. In R, \eval is a key tool to extend the language and
+maximize expressivity. In R; \eval is a key tool to extend the language and
 implement DSLs, it is also a replacement for macros. By contrast, the designers
 of Julia chose to limit \eval. In Julia, only global variables can be
-side-effected and environments cannot be readily manipulated. This is designed
+side-effected, and environments cannot be readily manipulated. This is designed
 to shield optimized code from some of the most pernicious uses of the
@@
 popular websites with an instrumented web browser to gather execution traces.
-They show that \eval is pervasive with 82\% of the most popular websites using
+They show that \eval is pervasive, with 82\% of the most popular websites using
 it. The reasons for its use include the desire to load code on demand,
-deserialization of JSON dataq and lightweight meta-programming to customize web
+deserialization of JSON data, and lightweight meta-programming to customize web
 pages. While many uses were legitimate, just as many were unnecessary and could
@@
 behaviors. Furthermore, JavaScript usage in 2011 is likely different from today,
-e.g. Node.js was not covered. More details about dynamic analysis of JavaScript
+\eg Node.js was not covered. More details about dynamic analysis of JavaScript
 can be found in~\cite{liang}.
@@
 they affect file change-proneness. Files with dynamic features are significantly
-more likely to be the subject of changes, than other files. Chen et al. looked
+more likely to be the subject of changes than other files. Chen et al. looked
 at the correlation between code changes and dynamic features, including \eval,
@@
 than 2\% of methods. The most common reflective method is \c{perform:}; it send
-a message that is specified by a string. These features are mostly used in the
+a message that is specified by a string. These features are primarily used in the
 core libraries.
 
-\citet{bodden} looked at usage of reflection in the Java DaCapo benchmark suite.
+\citet{bodden} looked at the usage of reflection in the Java DaCapo benchmark suite.
 They found that dynamic loading was triggered by the benchmark harness. The
-harness then executes methods via reflection, this caused static analysis tools
+harness then executes methods via reflection. This caused static analysis tools
 to generate an incorrect call graph for the programs in DaCapo.
@@
 view. The authors report that 53\% of the malware they studied used \eval as a
-means to obfuscate attack code or to mount attacks. They propose an abstract
+means to obfuscate attack code or mount attacks. They propose an abstract
 interpretation-based approach to analyzing dynamic languages. One must construct
@@
 \citet{ecoop12} had a short section on the usage of \eval in R. They found the
-it widely used in R code with 8500 call sites in CRAN and 2.2 million dynamic
-calls. The 15 most frequent call sites account for 88\% of those. The
+that \eval  is widely used in R code with 8500 call sites in CRAN and 2.2 million dynamic calls. The 15 most frequent call sites account for 88\% of those. The
 \c{match.arg} function is the highest used one with 54\% of all calls. In the
-other call sites, they saw two uses cases. The most common is the evaluation of
-the source code of a promise retrieved by \c{substitute} in a new environment;
-e.g. as done in the \c{with} function. The other use case is the invocation of a
-function whose name or arguments are determined dynamically. For this purpose, R
-provides \c{do.call} and thus \eval is overkill.
+other call sites, they saw two use cases. The most common is the evaluation of
+the source code of a promise retrieved by \c{substitute} in a new environment,
+\eg as done in the \c{with} function. The other use case is the invocation of a
+function whose name or arguments are determined dynamically. For this purpose, R provides \c{do.call} and thus \eval is overkill.
 
@@
 \end{table}
+
+\medskip
 

### 2021-04-16 paper/main.tex
Updates in 6

@@
 
-\subsection{Base}
+\subsection{Base}\label{sec:usage-base}
 
@@
 The R language was intended to be extensible. The combination of lazy
-evaluation, \c{substitute}, and \c{eval} are the tools given to
-developers to this end. This API is slightly more complex than just
-passing a string, it is conceivable that this may discourage some
-casual users. \Eval is also being used to reduce boilerplate code and
-provide convenience features for programmers. We give examples for the minimized expressions of Section~\ref{sec:minimized} and then dig into high-level \eval design patterns.
-
-The examples are based on code from CRAN or Kaggle, simplified to fit the limited space of a scientific paper. For each example, we indicate the package and the function where it is located.
+evaluation, \c{substitute} and \c{eval} are the tools given to developers to
+this end.
+%This API is slightly more complex than just passing a string, it is
+%conceivable that this may discourage some casual users. \Eval is also being
+%used to reduce boilerplate code and provide convenience features for
+%programmers. 
+In this section we focus on the \emph{why}, giving real-world examples of the
+\eval usage in the wild. We start the discussion with the examples for the
+minimized expressions from Section~\ref{sec:minimized}. They we look into
+high-level \eval design patterns. All of the examples are based on code from
+our corpus often simplified to fit the space and increase clarity. For each
+example, we indicate the package and the function where it is located.
 
@@
 
-We illustrate the minimized expressions of Section~\ref{sec:minimized} on representative examples.
-
-% model.frame is in the next subsection, in the high-level patterns
-
-\subsubsection{Variable lookup and values.}  We also include here lookup to slots in a list or data frame.
-This corresponds to minimized expressions \c{X}, \c{V}, and  \c{\$}.
-Variable lookup leads to a large diversity of environments, many being package environments or package environments wrapped in \c{new}.
-
-It is the predominant usage of \eval in Base, through function \c{match.args} (with evaluates on  minimized expression \c{V}). In CRAN, it correlates with more exotic environments passed to \eval. 
-The dominant site in numbers of calls in CRAN, is located in \c{ggproto} in the custom object system of \emph{ggplot2}.
-It evaluates a variable \c{`\_inherit`}, which represents the parent of the class, in the definition environment of the class.
-\eval is used here to account for parents scoped in a namespace with \c{::}, which is actually a function call in R.
-
-Package \emph{statnet.common} handles formulas, R object that associates an expression of the form \c{x~\textasciitilde~expr} and a carry-on environment where to evaluate the formula.
-It evaluates the left-hand side of the formula, which is always a variable, in function \c{eval\_lhs.formula}, where \c{object} is the formula and \c{environment} returns the formula associated to the object:
-
-\begin{lstlisting}
-eval(object[[2L]],envir=environment(object))
-\end{lstlisting}
-
-\subsubsection{Assigments.} It corresponds to minimized expression \c{<-} and groups together \c{<-}, \c{<<-} and \c{assign}.
-Package \emph{plyr} has a function combine data frames by row, filling in missing columns.
-During the allocation of the missing columns, an assignment call is built with \c{assignment <- quote(column[rows] <<- what)} and later evaluated on demand in a setter function, with \c{eval(assigmment)}.
-Package \emph{overture} is used to run Markov Chain Monte Carlo.
-It runs an R expression repeatedly and saves any assignments that happen, as samples of the Markov chains.
-The expression is captured with \c{substitute} and a new environment \c{env} wrapping the parent of the caller is created, in which all assignments will happen.
-Regularly, the content of \c{env} is inspected to sample the assigned variables.
-\begin{lstlisting}
-function(expr, overwrite=over.write) {
-	expr_q <- substitute(expr)
-	env <- new.env(parent=parent.frame(1))
-	samps <- list()
-	RunMcmc(samps, 1, expr_q, env, n.save, backing.path, thin, exclude, overwrite)
-}
-
-RunMcmc <- function... {
-	 eval(expr_q, envir=env)
-\end{lstlisting}
-
-\subsubsection{Function definition.} This corresponds to minimized expression \c{FUN}.
-Package \emph{Rcpp} is used to easily interface C++ and R and it uses \eval to bind C++ functions and classes to R objects.
-For instance, \c{.makeCppMethods} create a R function for each C++ method named \c{what} of a given class.
-The use of \substitute to inject the name of the method in a language expression entails the use of \eval.
-\begin{lstlisting}
-methods[[what]] <- eval(substitute(
-	function(...) .CppObject$WHAT(...), list(WHAT = as.name(what))),
-	env)
-\end{lstlisting}
-
-\mypara{Block.} In package \emph{vglm}, \eval is used to inject a pre-processing code at the beginning of a function implementing a statistical model: 
-\begin{lstlisting}
-if (length(slot(family, "first"))) 
-	eval(slot(family, "first"))
-\end{lstlisting}
-
-Package \emph{withr} provides facilities to encapsulate side-effecting functions.
-The function \c{defer} is used to attach an expression to be evaluated when exiting any frame in the stack.
-It can be passed arbitrary code, and so blocks.
-
-\subsubsection{Simple function calls.} This corresponds to minimized expression \c{F()}, \ie calls without nested functions and without variables (but possibly named arguments).
-For instance, package \emph{foreach} provides \c{foreach}-like loops.
-It makes it possible to initialize the iterator variables, with code \c{init}, which is equal \c{quote(list())} by default.
-\c{init} is then used to initialize the \c{foreach} accumulator \c{state\$accum <- eval(combineInfo\$init, envir=extra, enclos=obj\$evalenv)}.
-
-A peculiar use of \eval is to bypass some checks performed to accept code submitted to CRAN.
-For example, bindings can be \emph{locked} in an environment, \ie forbidding its value to be modified.
-The binding can be unlocked though by calling \c{unlockBinding} but packages in CRAN are forbidden to use that function.
-Package \emph{data.table} bypasses this restriction by building a call piece by piece, to unlock binding \c{.N}.
-The call is then evaluated.
-Now that \c{.N} is unlocked, it can be assigned to.
-Instead of using \eval to assign in the parent frame, they rather use the more specialized \c{assign} function.
-
-\begin{lstlisting}
-	if (locked.N) eval(call("unlockBinding", ".N", parent.frame()))  
-	assign(".N", nrow(x), envir=parent.frame(), inherits=FALSE)
-\end{lstlisting}
-
-\subsubsection{Complex function calls.} It corresponds to minimized expressions \c{F(X)} and \c{F(F(X))}, \ie function calls with variables, possibly with nested function calls.
-Package \emph{huxtable} is used to create custom tables and export them to various formats.
-To create functions to customize each of the borders of the table, it injects the indices \c{i} and \c{j} of the cells at the border and then evaluate them:
-\begin{lstlisting}
-xxx_border_huxtable <- function(lr_tb, i, j, border_prop) {
-	function (ht) {
-		hickness <- mlist$thickness[eval(i), eval(j), drop = FALSE]
-	}
-}
-left_border.huxtable <- xxx_border_huxtable("lr_borders",
-i = quote(seq_len(nrow(ht))), j = quote(-last_col))
-\end{lstlisting}
+In Table~\ref{tab:minimizedexpressions} we have shown the most common shapes of expressions passed to \eval.
+Here we illustrate them on concrete instances.
+
+\begin{compactitem}[---]
+
+  \item \emph{Variable lookup and values and indexing \c{(X, V, \$)}}
+    constitute one of the most common use case for \eval. Let's consider the
+    \c{subset} function from R base library which allows, among others, to
+    select variables in a data frame. For example, given a data frame \c{df}
+    with three columns \c{A,B,C}, \c{subset(df, sel=A)} returns a subset of
+    \c{df} with just the \c{A} column. The same would happen with \c{subset(df,
+    sel=1)}. However, thanks to \eval it can also be also called with
+    \c{subset(df, sel=c(1,B,x))} returning a subset of the first column, column
+    \c{B} and whatever index the variable \c{x} holds. It is implemented using:
+    \lstinline|eval(substitute(sel), nl, parent.frame())| The first argument
+    returns the expression passed to \c{sel}, the second is the data frame in
+    which this expression will be evaluated and finally the last indicates
+    which environment should be used to look up bindings not found in \c{nl}.
+
+    Another example is the most often called \eval site in CRAN:
+    \lstinline|eval(`_inherit`, env)| defined in the \c{find\_super} function
+    in \c{ggproto} object system. It looks up the variable \c{`\_inherit`} to
+    find the parent class. In the vast majority of cases the variable is symbol
+    and so the \eval seems superfluous. The \eval is however necessary so one
+    can specify parent classes using \c{pkg\_name::class\_name} as the \c{::}
+    operator is a function call in R.
+
+  \item \emph{Assignments \c{(<-)}} appears in the cases where a developer wants
+    to create or update a biding in a specific environment where either the
+    name of the value come from some expression. Most cases come from trivial
+    code generation where the assign expression is assembled using \c{parse} or
+    \c{substitute}, often in a loop: \lstinline|eval(parse(text=paste(capture.output(default_args[[i]]), "<-", i)))|.
+
+    There are also more sophisticated cases. The \c{plyr::allocate\_column}
+    function uses \eval for a deferred assignment to fill missing values in a
+    column. Package \c{overture} repeatedly evaluates R expressions saving any
+    assignments that happen in it, as samples of Markov chains.
+
+  \item \emph{Function definition \c{(FUN)}} is mostly used in conjunction with
+    code generation where new functions are synthesized using either \c{parse}
+    or \c{substitute}. Some FFI frameworks use this to automatically generate R
+    binding to native code. For example, \c{Rcpp}, which bridges R with C++,
+    uses \eval to generate R function for C++ methods:\\
+    \lstinline|eval(substitute(function(...) .CppObj$M(...), list(M=...)), env)|.
+
+  \item \emph{Block \c{(BLOCK)}} can appear anywhere, but it is almost always
+    the case that the block is only passed to \eval rather than being directly
+    part of in as in \c{eval({...})}. Essentially, the block denotes a fragment
+    of a program to be evaluated in a special way. There is a number of use
+    cases: unit testing frameworks (\eg \c{testthat}, \c{testit}), code
+    benchmarking (\eg \c{rbenchmark}, \c{microbenchmark}), running code in
+    parallel (\eg \c{foreach}, \c{doParallel}), or deferring code execution
+    (\eg \c{withr}).
+
+  \item \emph{Simple function calls \c{(F())}}
+
+    A peculiar use of \eval is to bypass some checks performed by the \c{R CMD
+      CHECK} tool allowing one to submit a package to CRAN that would otherwise
+      be refused. For example, the tool checks that packages do not call
+      \c{unlockBinding} function.\footnote{It allows one to open and modify
+      loaded package name space.} Packages such as \c{data.table} circumvent
+      such restrictions by using \eval as the check is only performed
+      statically.
+
+  \item \emph{Complex function calls \c{(F(X), F(F(X)))}}
+
+\end{compactitem}
 

### 2021-04-16 paper/main.tex
grammarly's proofreading for Section 3 (Methodoloy)

--- 2021-04-16 paper/main.tex
@@
 
-Our corpus is assembled from three sources: the \emph{Base} libraries that are
-bundled with the language implementation, packages hosted on \emph{CRAN} and
-scripts from \emph{Kaggle}.
+Our corpus is assembled from three sources: the \emph{Base} libraries 
+bundled with the language implementation, packages hosted on \emph{CRAN}, and scripts from \emph{Kaggle}.
 
@@
 \BaseEvalCallSites call sites to \eval in \BaseFunsWithEvals functions. Some of these
-functions provide basic functionalities such as package loading and thus there
-is hardly any code which does not invoke one of these.
+functions provide basic functionalities such as package loading, and thus, there
+is hardly any code that does not invoke one of these.
 
@@
 Each package comes with sample data. There are three sources of
-runnable code in a package: \emph{tests, examples} and \emph{vignettes} --
+runnable code in a package: \emph{tests, examples}, and \emph{vignettes} --
 respectively, unit tests, code snippets from the documentation, and long-form
@@
 The analysis presented in this has been automated by a pipeline that acquires
-packages, extract scripts, executes them, traces their behavior and summarizes
-observations. Figure~\ref{fig:pipeline} shows the main steps along with their
-running time, data size, and number of elements manipulated. Timings are from a
-cluster of three 2.3GHz Intel Xeon 6140 servers, with 72 cores and 256GB of RAM,
-and shared OCFS network storage. The pipeline steps are:
+packages, extracts scripts, executes them, traces their behavior, and summarizes observations. Figure~\ref{fig:pipeline} shows the main steps along
+ with their running time, data size, and  number of elements manipulated.
+  Timings are from a cluster of three 2.3GHz Intel Xeon 6140 servers, with 72 cores and 256GB of RAM, and shared OCFS network storage. The pipeline steps are:
 
@@
 \item \emph{Download.} Packages are downloaded from CRAN. For scripts, a web
-  crawler retrieves code and the Kaggle command line tool gets data.
-  Installation is complicated by native dependencies which are not properly
-  documented and thus hard to automatically resolve.
+  crawler retrieves code, and the Kaggle command-line tool gets data.
+  Installation is complicated by native dependencies, which are not adequately
+  documented and thus hard to resolve automatically.
 \item \emph{Extract.} Given installed packages and scripts, the next step is to
@@
   code snippets from a package and turns each of these into a self-standing
-  program~\cite{issta18}. Some Kaggle kernels are already scripts, for those
-  nothing more needs be done. Others kernels are packages as notebooks, either
-  as Rmarkdown or Jupyter, for those we use \c{knitr} to extract runnable code.
+  program~\cite{issta18}. Some Kaggle kernels are already scripts. For those,
+  nothing more needs to be done. Others kernels are packages as notebooks, either   as Rmarkdown or Jupyter. For those, we use \c{knitr} to extract
+   runnable code.
   The body of each extracted program is instrumented with calls to our dynamic
@@
   interest and not from bootstrapping or execution harness operations.
-\item \emph{Trace.} Each program is executing using our dynamic analysis tool
-  which is a heavily instrumented interpreter that captures calls to \eval and
+\item \emph{Trace.} Each program executes using our dynamic analysis tool, a heavily instrumented interpreter that captures calls to \eval and
   many other fine-grained runtime events. Packages are run twice, once to
-  capture \eval calls originating from package code, and a second time to
+  capture \eval calls originating from package code and a second time to
   capture calls coming from the base libraries. To avoid any interference, each
@@
   to avoid recording \eval in its code, and fallback on the interpreter.
-\item \emph{Analyze.} Finally, analysis output is merged, cleaned and summarized
+\item \emph{Analyze.} Finally, the analysis output is merged, cleaned, and summarized
   in a post-processing phase driven by series of R scripts. The summarized data
@@
   numbers appearing in the paper are generated automatically. Figures are
-  produced in PDF by \c{ggplot2}, numbers are exported as \LaTeX macros.
+  produced in PDF by \c{ggplot2}; numbers are exported as \LaTeX macros.
 \end{compactenum}
@@
 The tracer registers callbacks to all \eval functions as well as a few other
-ones that allow us to better identify where the expressions come from. For
-example, we taint the results of calls to \c{parse} (turns string into
-expression) and \c{match.call} (reflects on current call). We also capture calls
-to R API for dynamic code loading (\eg\xspace \c{library}, \c{require} and
+ones that allow us to identify better where the expressions come from. For
+example, we taint the results of calls to \c{parse} (turns strings into
+expressions) and \c{match.call} (reflects on current call). We also capture calls
+to R API for dynamic code loading (\eg\xspace \c{library}, \c{require}, and
 \c{source}). Next, we subscribe to the events related to a variable definition
-and assignment allowing us to record side effects that happen in environments
+and assignment, allowing us to record side effects that happen in environments
 while evaluating code in \eval. One challenge we encountered during the analysis
@@
 ASTs. The tracer is implemented as an R package in 3.2K C++ and 1.3K of R code.
-For performance reason, most of the tracing is done in C++. While in theory the
-implementation is rather straightforward, it is not so in practice:
+For performance reasons, most of the tracing is done in C++. While in theory, the
+implementation is relatively straightforward, it is not so in practice:
 %
@@
 
-\item The lazy evaluation makes it difficult to analyze function arguments
+\item The lazy evaluation makes it challenging to analyze function arguments
   while tracing as prematurely forcing a promise might have dangerous
@@
 \item In the dynamic analysis, we run all the \emph{runnable} code we obtained
-  from CRAN and Kaggle---\ie real code written by people with highly varying
+  from CRAN and Kaggle---\ie actual code written by people with highly varying
   expertise in R and programming in general. This exercises a lot of the corner
@@
 
-\mypara{Limitations.} The analysis has some limitation. Even with the extension
+\mypara{Limitations.} The analysis has some limitations. Even with the extension
 described above, there are still \PkgUndefinedRnd \eval calls without source
-references. This occurs when a reference to the \eval function is passed as
-argument to a higher-order functions or when the \eval call originates from
+references. This occurs when a reference to the \eval function is passed as an 
+argument to a higher-order function or when the \eval call originates from
 native code. However, these missing source references account for a meager
@@
 limitations are that we ignore calls to the native \eval function and to the
-alternate \c{rlang::tidy\_eval} function which uses native \eval internally.
+alternate \c{rlang::tidy\_eval} function, which uses native \eval internally.
 The \c{rlang} package introduces a new kind of promise called \c{quosure} which

### 2021-04-16 paper/main.tex
Updates

--- 2021-04-16 paper/main.tex
@@
 
+  \item \emph{Function calls \c{(F(), F(X), F(F(X)))}}
+
   \item \emph{Block \c{(BLOCK)}} can appear anywhere, but it is almost always
     the case that the block is only passed to \eval rather than being directly
-    part of in as in \c{eval({...})}. Essentially, the block denotes a fragment
+    part of in as in \lstinline|eval({...})|. Essentially, the block denotes a fragment
     of a program to be evaluated in a special way. There is a number of use
@@
     (\eg \c{withr}).
-
-  \item \emph{Simple function calls \c{(F())}}
-
-    A peculiar use of \eval is to bypass some checks performed by the \c{R CMD
-      CHECK} tool allowing one to submit a package to CRAN that would otherwise
-      be refused. For example, the tool checks that packages do not call
-      \c{unlockBinding} function.\footnote{It allows one to open and modify
-      loaded package name space.} Packages such as \c{data.table} circumvent
-      such restrictions by using \eval as the check is only performed
-      statically.
-
-  \item \emph{Complex function calls \c{(F(X), F(F(X)))}}
 
@@
 % \todo{Aviral's story when he contacted a developer about his use of eval}
+
+    A peculiar use of \eval is to bypass some checks performed by the \c{R CMD
+      CHECK} tool allowing one to submit a package to CRAN that would otherwise
+      be refused. For example, the tool checks that packages do not call
+      \c{unlockBinding} function.\footnote{It allows one to open and modify
+      loaded package name space.} Packages such as \c{data.table} circumvent
+      such restrictions by using \eval as the check is only performed
+      statically.
 

### 2021-04-16 paper/main.tex
Fixing typo

@@
 \medskip\noindent\EE{$min(e)=\c{model.frame}$} The \c{model.frame} function
-returns a dataframe resulting from f fitting the model described in a given
+returns a dataframe resulting from fitting the model described in a given
 formula. This form subsumes \c{F(F(X))}, \c{FUN}, and assignments. It is the

### 2021-04-16 paper/main.tex
Fix problem with Fig. 4

@@
  > f(a+b)
- # expression(a+b)
+ # a+b
  > substitute(a+b,list(a=1,b=2))
@@
 \noindent Environments are used as hash maps as they have reference semantics
-and a built-in string look up. Functions \c{parent.frame} and \c{sys.frame}
+and a built-in string lookup. Functions \c{parent.frame} and \c{sys.frame}
 return environments further up the call stack.
@@
 &     \multicolumn{3}{c}{\small\#calls} &\small\#pck \\\hline
-\tt 1 &--& \tt 10      & \packageBina  & \tt 1K &--&\tt 100K  & \packageBine\\
+\tt 1 &--& \tt 10      & \packageBina  & \tt 10K &--&\tt 100K  & \packageBine\\
 \tt 11 &--& \tt 100    & \packageBinb  & \tt 100K &--&\tt 1M  & \packageBinf\\
@@
 \begin{tabular}{|@{\,}r|rrrr|}\hline
-  &\eval & \c{evalq} & \c{eval} & \c{local}\\[-2mm]
+  &\eval & \c{evalq} & \c{eval} & \c{local}\\[-1.5mm]
            & & & \c{.parent} &\\\hline

### 2021-04-16 paper/main.tex
Usage metric fixes

@@
 In total, \kaggleAllcalls \eval were recorded, all to \eval function. Out of
-\kaggleStaticeval sites from \KaggleWithEvals, only \kaggleTriggeredeval sites
+\kaggleStaticeval sites from \KaggleWithEvals scripts, only \kaggleTriggeredeval sites
 in \kaggleNbruns scripts were hit; \KaggleFailedScripts scripts failed to run,
-and the other did not exercise the \eval sites. This is partially expected
+and the others did not exercise the \eval sites. This is partially expected
 as Kaggle code does not need to abide by any checks. Upon a manual inspection,
-we observed that indeed the failing scripts were of poor quality, often not
+we observed that indeed, the failing scripts were of poor quality, often not
 finished, using misspelled package names, hard-coded file paths, or accessing

### 2021-04-16 paper/main.tex
Push box on right page

@@
 
-\medskip
-
- We detail these forms and discuss their implication for the behavior of \eval.
+\medskip\noindent  We detail these forms and discuss their implication for the behavior of \eval.
 
@@
 it is not the common case. Lookups are often evaluated in constructed
-environments; as few as \packageMinimizedpercentparentframesc of these
+environments; as few as \packageMinimizedpercentparentframesa of these
 expressions are evaluated in a function environment.
@@
 \c{x[[24]]}. As with \c X, we allow arithmetics and values in this form. Lookup
-occurs in 6\% of the sites. The interpreter evaluates
+occurs in 10\% of the sites. The interpreter evaluates
 \packageMinimizedmedianoperationsgRnd operations on average; the minimum is 3
@@
 
+\ \\[5mm] % To push the box on the next page
 \medskip\noindent\EE{$min(e)=\c{F()}$} This form captures simple function calls,

### 2021-04-16 paper/main.tex
Fix some numbers in the expression of eval
Refactor S6

@@
 respectively, \packageMinimizedmedianoperationsdRnd and
-\packageMinimizedmedianoperationsbRnd. Also,  \packageMinimizedpercentparentframesj
+\packageMinimizedmedianoperationsbRnd. Also,  \packageMinimizedpercentparentframesd and \packageMinimizedpercentparentframesb
 of these expressions run in function environments.
@@
 In this section we focus on the \emph{why}, giving real-world examples of the
-\eval usage in the wild. We start the discussion with the examples for the
-minimized expressions from Section~\ref{sec:minimized}. They we look into
-high-level \eval design patterns. All of the examples are based on code from
-our corpus often simplified to fit the space and increase clarity. For each
-example, we indicate the package and the function where it is located.
+\eval usage in the wild. All of the examples are based on code from our corpus
+often simplified to fit the space and increase clarity. For each example, we
+indicate the package and the function where it is located.
 
@@
 
-In Table~\ref{tab:minimizedexpressions} we have shown the most common shapes of expressions passed to \eval.
-Here we illustrate them on concrete instances.
+% We start the discussion with the examples for the minimized expressions from
+% Section~\ref{sec:minimized}.
+In Table~\ref{tab:minimizedexpressions} we have shown the most common shapes of
+expressions passed to \eval. Here we illustrate them on concrete instances.
 
@@
 
-  \item \emph{Function calls \c{(F(), F(X), F(F(X)))}}
+  \item \emph{Function calls \c{(F(), F(X), F(F(X)))}} represents two very
+    frequent patterns: \emph{code generation} and \emph{call reflection}. There are numerous
+    cases for code generation. One occurring usage is generating ad-hoc calls
+    to libraries such as \c{ggplot2} when one needs to parameterize data or
+    aesthetics selection from a variable. For example \lstinline|ggplot(df, aes(x=A, y=B))|
+    maps X and Y axes in the plot to \c{A} and \c{B} columns in
+    \c{df}. It might not be immediately clear\footnote{\cf
+    \url{https://stackoverflow.com/questions/22309285/how-to-use-a-variable-to-specify-column-name-in-ggplot}}
+    how modify the call so it will pick the actual mapping from \c{A} and \c{B}
+    variables. Another case is the traditional use of \eval as a means to
+    reduce boilerplate code; simple and repetitive code can easily be replaced
+    with judicious use of \eval. For example, the \c{data.table} package uses
+    \eval to calls the \c{options} function with named arguments taken from a
+    vector of strings. While the benefits are limited in this example, it is an
+    attractive tool for programmers.
+    %
+    \begin{lstlisting} 
+    opts = c("datatable.verbose"="FALSE", # ...many others
+    for (i in names(opts)) eval(parse(text=paste0("options(",i,"=",opts[i],")"))) 
+    \end{lstlisting}
+
+    Call reflection refers to the combination of \eval with \c{match.call},
+    function that returns the current call with all its arguments captures as
+    expressions. This is used mostly in the following circumstances: (1) to
+    record a call for a later use in possibly different environment, (2) pass
+    most of the captured call arguments to another function or (3) evaluate
+    arguments of the function in different environments. This is widely used in
+    relation with statistical modeling in combination with a model fitting
+    function or with \c{model.frame} and related functions. For example the
+    \c{survival::coxph} function\footnote{Function fitting a Cox proportional hazards
+    regression model.} contains:
+    %
+    \begin{lstlisting}
+    Call <- match.call()
+    tform <- Call[c(1,indx)]  # only keep the arguments we wanted
+    tform[[1L]] <- quote(stats::model.frame)  # change the function called
+    mf <- eval(tform, parent.frame())
+    \end{lstlisting}
 
@@
     the case that the block is only passed to \eval rather than being directly
-    part of in as in \lstinline|eval({...})|. Essentially, the block denotes a fragment
-    of a program to be evaluated in a special way. There is a number of use
-    cases: unit testing frameworks (\eg \c{testthat}, \c{testit}), code
-    benchmarking (\eg \c{rbenchmark}, \c{microbenchmark}), running code in
-    parallel (\eg \c{foreach}, \c{doParallel}), or deferring code execution
-    (\eg \c{withr}).
+    part of in as in \lstinline|eval({...})|. Essentially, the block denotes a
+    fragment of a program to be evaluated in a special way and in a special
+    environment. The latter is what makes it different from 0-argument closure.
+
+    There is a number of use cases: unit testing frameworks (\eg \c{testthat},
+    \c{testit}), code benchmarking (\eg \c{rbenchmark}, \c{microbenchmark}),
+    running code in parallel (\eg \c{foreach}, \c{doParallel}), or deferring
+    code execution (\eg \c{withr}).
 
@@
 
-\subsection{High-level patterns.}
-
-We present here high-level patterns that we identified after manual inspections of hundred of \eval sites. Some of the patterns are also described in the R documentation.
-
-\subsubsection{Domain Specific Language.} R expressions are built and transformed symbolically and then evalled to a concrete value.
+\subsection{Domain Specific Language.} R expressions are built and transformed symbolically and then evalled to a concrete value.
 For instance,   base R packages \emph{stats} provides two operators for \emph{symbolic differentiation}, \c{D} and \c{deriv}.
@@
 
-\subsubsection{Capturing a call.}  A common use case for \eval is to be combined with \c{match.call}.
-\c{match.call} walks up the call stack, captures the code that invoked the
-currently executing function, and returns it as an unevaluated expression that includes the call function and its unevaluated arguments.
+% \subsubsection{Capturing a call.}  A common use case for \eval is to be combined with \c{match.call}.
+% \c{match.call} walks up the call stack, captures the code that invoked the
+% currently executing function, and returns it as an unevaluated expression that includes the call function and its unevaluated arguments.
 The pattern is to transform a call to some function \c{f} into a call to \c{g} with some arguments retained and others modified.
 This pattern is recognizable
-by the fact that the expression is a call and the target environment is that of
-the parent.
-
-% We have quantitative data on that if needed...
-This pattern is very often (more than half of the cases) used in relation with statistical models, to build a new model, \c{model.frame} to build a new generic model, or a more specialized model like \c{glm} or, \c{lm}.
+% by the fact that the expression is a call and the target environment is that of
+% the parent.
+
+% % We have quantitative data on that if needed...
+% This pattern is very often (more than half of the cases) used in relation with statistical models, to build a new model, \c{model.frame} to build a new generic model, or a more specialized model like \c{glm} or, \c{lm}.
 The R documentation of \c{match.call} explicitly describes this usage.
 
-In package \emph{survival}\footnote{It is a \emph{recommended package}, so recommended by the CRAN maintainers, thoroughly tested, and following R best practices.}, in function \c{coxph}, the surrounding call is first captured, then \c{model.frame} is injected as the function name in the call, and finally, the new call is evaluated in the parent frame, to ensure that the rewritten call
-is evaluated in the same environment the original call was.
-
-\begin{lstlisting}
-Call <- match.call()
-tform <- Call[c(1,indx)]  # only keep the arguments we wanted
-tform[[1L]] <- quote(stats::model.frame)  # change the function called
-mf <- eval(tform, parent.frame())
-\end{lstlisting}
-
-A related pattern is not to change the call but rather to evaluate the arguments in the parent frame. This could also be achieved with \c{substitute}, but \c{match.call} does it in bulk.
-In package \emph{MedDietCalc},  in function \c{computeMDS95}, \c{match.call} retrieves the argument expressions of the call, which are evaluated in a dataframe \c{data} which contains the numeric data.
+% In package \emph{survival}\footnote{It is a \emph{recommended package}, so recommended by the CRAN maintainers, thoroughly tested, and following R best practices.}, in function \c{coxph}, the surrounding call is first captured, then \c{model.frame} is injected as the function name in the call, and finally, the new call is evaluated in the parent frame, to ensure that the rewritten call
+% is evaluated in the same environment the original call was.
+
+% \begin{lstlisting}
+% Call <- match.call()
+% tform <- Call[c(1,indx)]  # only keep the arguments we wanted
+% tform[[1L]] <- quote(stats::model.frame)  # change the function called
+% mf <- eval(tform, parent.frame())
+% \end{lstlisting}
+
+% A related pattern is not to change the call but rather to evaluate the arguments in the parent frame. This could also be achieved with \c{substitute}, but \c{match.call} does it in bulk.
+% In package \emph{MedDietCalc},  in function \c{computeMDS95}, \c{match.call} retrieves the argument expressions of the call, which are evaluated in a dataframe \c{data} which contains the numeric data.
 Each argument of function \c{computeMDS95} is also a column in dataframe \c{data}.
 The advantage of using \c{eval} here is to be able to express an argument as a combination of several columns in the dataframe.
-\begin{lstlisting}
-computeMDS95 <- function (data, Vegetables, ...) {
-	arguments <- as.list( match.call())
-	Vegetables <- eval(arguments$Vegetables, data) 
-\end{lstlisting}
-
-\c{substitute} can be used to capture only a part of the call, \ie one of the arguments.
+% \begin{lstlisting}
+% computeMDS95 <- function (data, Vegetables, ...) {
+% 	arguments <- as.list( match.call())
+% 	Vegetables <- eval(arguments$Vegetables, data) 
+% \end{lstlisting}
+
+% \c{substitute} can be used to capture only a part of the call, \ie one of the arguments.
 For instance, in package \emph{xyplot}, the expression for argument \c{subset} is captured with \c{substitute} and then evaluated in a dataframe \c{data} and in the definition environment of \c{x}.
-\begin{lstlisting}
-xyplot.formula <-  function(x,    data = NULL, ..., subset = TRUE)
-subset <- eval(substitute(subset), data, environment(x))
-\end{lstlisting}
-
-In the R community, the call capturing pattern with evaluation in a non-local environment is often referred~\cite{hadley} as \emph{Non-Standard Evaluation}.
+% \begin{lstlisting}
+% xyplot.formula <-  function(x,    data = NULL, ..., subset = TRUE)
+% subset <- eval(substitute(subset), data, environment(x))
+% \end{lstlisting}
+
+% In the R community, the call capturing pattern with evaluation in a non-local environment is often referred~\cite{hadley} as \emph{Non-Standard Evaluation}.
 This pattern in relation with \c{model.frame} is also described explicitly in the R documentation of \c{match.call}.
 
-\subsubsection{Injecting code in package environments.} R distinguishes two kinds of environments for packages: the \emph{package} environment itself, which contains every publicly accessible functions, and the \emph{namespace} environment, which contains all functions including private ones.
-
- A call to \c{library(packageName)}, which loads libraries in R, for instance, will
-populate environment \c{package:packageName} using \eval. It is a \emph{base} function
-though, so we do not see it in the CRAN dataset. However, R still makes it
-possible to access any package environments, which we see in
-\packageNbPackageEnvPackages different packages in our corpus. Those packages perform
-introspection, to represent and search any environment by name, in package
-\emph{envnames}, to containerize tests, as with \emph{testthat} and
-\emph{unitizer}, to create a custom OOP system with \emph{R.oo}, or to bypass the visibility of functions in \emph{USSCXenaTools}. It is also used to
-implement a custom caching mechanism in packages \emph{g.data} and \emph{SOAR}.
-
-\subsubsection{Code Generation.} The more traditional use of \eval is to execute code
-that was assembled by the programmer into a string. Here we show the method
-\c{plot} for class \c{gmap} in package \c{gpmaps}, simplified for
-explanatory purposes.   \c{paste} is used to dispatch to a specialized plot function, among \c{plot2\_orig},  \c{plot3\_orig}, \c{plot3\_dec} and so on.
-\begin{lstlisting}
-eval(parse(text=paste('plot',gpmap$nloci,'_orig(gpmap)',sep="")))
-\end{lstlisting}
-
-In Kaggle, plots are often built with \eval, \c{paste} and \c{ggplot}.
+% \subsubsection{Injecting code in package environments.} R distinguishes two kinds of environments for packages: the \emph{package} environment itself, which contains every publicly accessible functions, and the \emph{namespace} environment, which contains all functions including private ones.
+
+%  A call to \c{library(packageName)}, which loads libraries in R, for instance, will
+% populate environment \c{package:packageName} using \eval. It is a \emph{base} function
+% though, so we do not see it in the CRAN dataset. However, R still makes it
+% possible to access any package environments, which we see in
+% \packageNbPackageEnvPackages different packages in our corpus. Those packages perform
+% introspection, to represent and search any environment by name, in package
+% \emph{envnames}, to containerize tests, as with \emph{testthat} and
+% \emph{unitizer}, to create a custom OOP system with \emph{R.oo}, or to bypass the visibility of functions in \emph{USSCXenaTools}. It is also used to
+% implement a custom caching mechanism in packages \emph{g.data} and \emph{SOAR}.
+
+% \subsubsection{Code Generation.} The more traditional use of \eval is to execute code
+% that was assembled by the programmer into a string. Here we show the method
+% \c{plot} for class \c{gmap} in package \c{gpmaps}, simplified for
+% explanatory purposes.   \c{paste} is used to dispatch to a specialized plot function, among \c{plot2\_orig},  \c{plot3\_orig}, \c{plot3\_dec} and so on.
+% \begin{lstlisting}
+% eval(parse(text=paste('plot',gpmap$nloci,'_orig(gpmap)',sep="")))
+% \end{lstlisting}
+
+% In Kaggle, plots are often built with \eval, \c{paste} and \c{ggplot}.
 In this kernel for the \emph{digit-recognizer} challenge, \c{paste0} is used to generate \c{ggplot} commands featuring unique y labels to build 9 different graphs.
 
-\begin{lstlisting}
-p <- lapply(paste0('PC', 1:9), function(pc){
-	eval(parse(text=paste0('ggplot(ld, aes(x=label, y=',pc,')) + geom_boxplot()')))
-})
-\end{lstlisting}
-
-The \parse and \eval combination is
-used by the \source function in the base library to load R code from a file
-in the current workspace.
-
- Code generation is often used as a means to reduce boilerplate code;
-simple and repetitive code can easily be replaced with judicious use of \eval.
-For example, the \c{data.table} package uses \eval to calls the \c{options}
-function with named arguments taken from a vector of strings. While the benefits
-are limited in this example, it is an attractive tool for programmers.
-\begin{lstlisting}
-  opts = c("datatable.verbose"="FALSE", # ...many others
-  for (i in names(opts))
-    eval(parse(text=paste0("options(",i,"=",opts[i],")")))
-\end{lstlisting}
-This pattern is a special case of code generation, recognizable by the
-fact that \eval is executed in a loop.
-
-Using \c{parse}, \ie converting strings to R expressions, opens to a more flexible code generation, as the building blocks which are going to be concatenated together, do not need to be valid R expressions.
+% \begin{lstlisting}
+% p <- lapply(paste0('PC', 1:9), function(pc){
+% 	eval(parse(text=paste0('ggplot(ld, aes(x=label, y=',pc,')) + geom_boxplot()')))
+% })
+% \end{lstlisting}
+
+% The \parse and \eval combination is
+% used by the \source function in the base library to load R code from a file
+% in the current workspace.
+
+%  Code generation is often used as a means to reduce boilerplate code;
+% simple and repetitive code can easily be replaced with judicious use of \eval.
+% For example, the \c{data.table} package uses \eval to calls the \c{options}
+% function with named arguments taken from a vector of strings. While the benefits
+% are limited in this example, it is an attractive tool for programmers.
+% \begin{lstlisting}
+%   opts = c("datatable.verbose"="FALSE", # ...many others
+%   for (i in names(opts))
+%     eval(parse(text=paste0("options(",i,"=",opts[i],")")))
+% \end{lstlisting}
+% This pattern is a special case of code generation, recognizable by the
+% fact that \eval is executed in a loop.
+
+% Using \c{parse}, \ie converting strings to R expressions, opens to a more flexible code generation, as the building blocks which are going to be concatenated together, do not need to be valid R expressions.
 On the contrary, combining expression created with \c{substitute} or \c{quote} requires more discipline.
@@
 
-\todo{There is an example in package comprenhr that does code generation with substitute to create list comprehensions, but it is quite difficult to grasp.
+% \todo{There is an example in package comprenhr that does code generation with substitute to create list comprehensions, but it is quite difficult to grasp.
 But at the same type, we lack here an example of code generation using quite, enquote or substitute.}
 
-\todo{Add a pattern about Library Loading? \c{library(packageName)} in R uses \eval This is one of the patterns in Javascript...}
-
-\subsubsection{Unnecessary use of \eval}
+% \todo{Add a pattern about Library Loading? \c{library(packageName)} in R uses \eval This is one of the patterns in Javascript...}
+
+\subsection{Unnecessary use of \eval}
 
@@
   function (R, d="norm", dp, ...) {
-	q.f <- eval(parse(text=paste("q",d,sep="")))
-	z <- NULL
+	  q.f <- eval(parse(text=paste("q",d,sep=""))); z <- NULL
   	eval(parse(text=paste("z<-q.f(",dp,",...)")))
@@
 \end{lstlisting}
-  In both cases, there is no need for \eval:
+  In both cases, there is no need for \eval and the function can be rewritten as
 \begin{lstlisting}
   function (R, d="norm", dp, ...) {
-	q.f <- get(paste0("q",d))
-	z <- q.f(dp, ...)
+	  q.f <- get(paste0("q",d)); z <- q.f(dp, ...)
   }
 \end{lstlisting}
-or even to a oneliner \c{do.call(paste0("q",d), as.list(dp, ...))}.
-
-% \todo{Aviral's story when he contacted a developer about his use of eval}
-
-    A peculiar use of \eval is to bypass some checks performed by the \c{R CMD
-      CHECK} tool allowing one to submit a package to CRAN that would otherwise
-      be refused. For example, the tool checks that packages do not call
-      \c{unlockBinding} function.\footnote{It allows one to open and modify
-      loaded package name space.} Packages such as \c{data.table} circumvent
-      such restrictions by using \eval as the check is only performed
+or even to a one-liner \c{do.call(paste0("q",d), as.list(dp, ...))}.
+
+\subsection{Discussion}
+
+\Eval in R is mostly used for meta-programming purposes and to access
+environments other than the local one. In Javascript, patterns are ad-hoc, such
+as json loading and parsing, and can often be rewritten without \eval, such as
+function or method call, or assignment to a local variable or an object
+property. Compared to Javascript, \eval patterns in R we observed are much more
+legitimate, although this is biased by the fact that our main corpus CRAN
+features more experienced developers than  websites found at large on the web.
+\todo{can we really claim that? SHould we talk about Kaggle there?}
+
+\eval is a powerful tool and can be sometimes replaced with functions with more
+specialized goals. For instance, variable lookup in any environment can be
+performed with \c{get}, and assignment in any environment, with \c{assign}.
+Building a call and executing can be done with \c{do.call}. Nevertheless, \eval
+shines when expressions passed to it can be of many different types. % And it
+is shorter to type, and no need to remember all the specialized functions, so
+it is convenient.
+
+Finally, let us mention an anecdotal use of \eval as a mean to bypass some
+checks performed by the \c{R CMD CHECK} tool allowing one to submit a package
+to CRAN that would otherwise be refused. For example, the tool checks that
+packages do not call \c{unlockBinding} function.\footnote{It allows one to open
+and modify loaded package name space.} Packages such as \c{data.table}
+circumvent such restrictions by using \eval as the check is only performed
 statically.
-
-\mypara{Discussion.} \Eval in R is mostly used for meta-programming purposes and to access environments other than the local one.
-In Javascript, patterns are ad-hoc, such as json loading and parsing, and can often be rewritten without \eval, such as function or method call, or assignment to a local variable or an object property.
-Compared to Javascript, \eval patterns in R we observed are much more legitimate, although this is biased by the fact that our main corpus CRAN features more experienced developers than  websites found at large on the web.
-\todo{can we really claim that?
-SHould we talk about Kaggle there?}
-
-\eval is a powerful tool and can be sometimes replaced with functions with more specialized goals.
-For instance, variable lookup in any environment can be performed with \c{get}, and assignment in any environment, with \c{assign}.
-Building a call and executing can be done with \c{do.call}.
-Nevertheless, \eval shines when expressions passed to it can be of many different types.
-% And it is shorter to type, and no need to remember all the specialized functions, so it is convenient.
 

### 2021-04-16 paper/main.tex
Side effects fixes

--- 2021-04-16 paper/main.tex
@@
 
-Contrary to the Javasscript \eval, which does not precise the scope in which to evaluate,the second argument of \eval in R, \c{envir}, specifically is dedicated ot that. 
+Contrary to the Javascript \eval, which does not precise the scope in which to evaluate,the second argument of \eval in R, \c{envir}, specifically is dedicated to that. 
 The environment in which evaluation happens determines what is visible to the
@@
 \item {\it Constructed:} Expressions can be constructed by invoking the
-  \c{quote}, \c{enquote}, \c{expression}. Function arguments are passed as
+  \c{quote}, \c{enquote}, \c{expression} functions. Function arguments are passed as
   promises, and \c{substitute} is used to retrieve the source expression
@@
 
-From the \packageNbrunsRnd programs we capture \SEAllRnd side effects from
+From the \packageNbrunsRnd programs, we capture \SEAllRnd side effects from
 \SEAllCallsRnd \eval calls (\SEUserCallsToAllRatio) in \SEAllSites sites
 (\SEUserSitesToAllRatio). Again, the challenge is to remove the accidental
-side effects caused by the R virtual machine implementation and are not
+side effects caused by the R virtual machine implementation that are not
 related to the user code. For example, the \c{.Random.seed} variable, which
@@
 described in Section~\ref{sec:env} with two extra classes: \emph{Object},
-which represents a side-effect that happens in S4 or R6 object environment, and
+which represents a side-effect that happens in S4 or R6 object environments, and
 \emph{Local} that corresponds to a side effect in local variables of the \eval
-caller. The table shows data for both \eval sites and functions. For a
+caller (offset 0 in Section~\ref{sec:env}).  The table shows data for both \eval sites and functions. For a
 function to have a given target environment class, all of its \eval sites that
@@
 \eval sites (\SESitesInOneClass) do all side effects consistently in one
-environment class. The same happens on the function level. Almost half of the
+environment class. The same happens at the function level. Almost half of the
 sites (over a third of the functions) do side effects in either \emph{Local} or

### 2021-04-16 paper/main.tex
Some proofreading of Section 6

@@
 (\SEListTypeRatio). From the compiler perspective, we would like to know how
-many sites changes function bindings. In the corpus, this happens in only
+many sites change function bindings. In the corpus, this happens in only
 \SEClosureType sites; \SEClosureTypeLocal of them do that in the local
@@
 that first, side-effects from \eval are not as widespread as in the case of
-JavaScript, \footnote{\citep{ecoop11} shows that in the \emph{Interactive} scenario, \eval in Javascript performs, store events can reach up to 40\% of the events, and 7\% to 8\% of the \eval do side effects in the global scope.
+JavaScript,\footnote{\citep{ecoop11} shows that in the \emph{Interactive} scenario, \eval in Javascript performs, store events can reach up to 40\% of the events, and 7\% to 8\% of the \eval do side effects in the global scope.
 } and that over half of them happen in a predictable environment.
@@
 %programmers. 
-In this section we focus on the \emph{why}, giving real-world examples of the
-\eval usage in the wild. All of the examples are based on code from our corpus
+In this section, we focus on the \emph{why}, giving real-world examples of the
+\eval usage in the wild. All the examples are based on code from our corpus,
 often simplified to fit the space and increase clarity. For each example, we
@@
 In Table~\ref{tab:minimizedexpressions} we have shown the most common shapes of
-expressions passed to \eval. Here we illustrate them on concrete instances.
+expressions passed to \eval. Here, we illustrate them on concrete instances.
 
@@
 
-  \item \emph{Variable lookup and values and indexing \c{(X, V, \$)}}
+  \item \emph{Variable lookup and values, and indexing \c{(X, V, \$)}}
     constitute one of the most common use case for \eval. Let's consider the
-    \c{subset} function from R base library which allows, among others, to
+    \c{subset} function from the R base library, which allows, among others, to
     select variables in a data frame. For example, given a data frame \c{df}
@@
     returns the expression passed to \c{sel}, the second is the data frame in
-    which this expression will be evaluated and finally the last indicates
+    which this expression will be evaluated, and the last indicates
     which environment should be used to look up bindings not found in \c{nl}.
@@
     \lstinline|eval(`_inherit`, env)| defined in the \c{find\_super} function
-    in \c{ggproto} object system. It looks up the variable \c{`\_inherit`} to
-    find the parent class. In the vast majority of cases the variable is symbol
+    in the \c{ggproto} object system. It looks up the variable \c{`\_inherit`} to
+    find the parent class. In the vast majority of cases, the variable is a symbol
     and so the \eval seems superfluous. The \eval is however necessary so one
-    can specify parent classes using \c{pkg\_name::class\_name} as the \c{::}
+    can specify parent classes using \c{pkg\_name::class\_name}, as the \c{::}
     operator is a function call in R.
 
-  \item \emph{Assignments \c{(<-)}} appears in the cases where a developer wants
-    to create or update a biding in a specific environment where either the
-    name of the value come from some expression. Most cases come from trivial
+  \item \emph{Assignments \c{(<-)}} appear in the cases where a developer wants
+    to create or update a binding in a specific environment where either the
+    name or the value come from some expression. Most cases originate from trivial
     code generation where the assign expression is assembled using \c{parse} or
@@
     function uses \eval for a deferred assignment to fill missing values in a
-    column. Package \c{overture} repeatedly evaluates R expressions saving any
-    assignments that happen in it, as samples of Markov chains.
+    column. Package \c{overture} repeatedly evaluates R expressions capturing any  assignments that happen in it, as samples of Markov chains.
 
@@
     binding to native code. For example, \c{Rcpp}, which bridges R with C++,
-    uses \eval to generate R function for C++ methods:\\
+    uses \eval to generate R functions for C++ methods:\\
     \lstinline|eval(substitute(function(...) .CppObj$M(...), list(M=...)), env)|.
 
-  \item \emph{Function calls \c{(F(), F(X), F(F(X)))}} represents two very
+  \item \emph{Function calls \c{(F(), F(X), F(F(X)))}} represent two very
     frequent patterns: \emph{code generation} and \emph{call reflection}. There are numerous
-    cases for code generation. One occurring usage is generating ad-hoc calls
+    cases for code generation. One occurring usage is to generate ad-hoc calls
     to libraries such as \c{ggplot2} when one needs to parameterize data or
@@
     \c{df}. It might not be immediately clear\footnote{\cf
-    \url{https://stackoverflow.com/questions/22309285/how-to-use-a-variable-to-specify-column-name-in-ggplot}}
-    how modify the call so it will pick the actual mapping from \c{A} and \c{B}
+    \url{https://stackoverflow.com/questions/22309285/how-to-use-a-variable-to-specify-column-name-in-ggplot}}  for a ggplot2 user 
+    how to modify the call to pick the actual mapping from \c{A} and \c{B}
     variables. Another case is the traditional use of \eval as a means to
@@
     with judicious use of \eval. For example, the \c{data.table} package uses
-    \eval to calls the \c{options} function with named arguments taken from a
+    \eval to call the \c{options} function with named arguments taken from a
     vector of strings. While the benefits are limited in this example, it is an
@@
 
-    Call reflection refers to the combination of \eval with \c{match.call},
-    function that returns the current call with all its arguments captures as
+    Call reflection refers to the combination of \eval with \c{match.call}, a
+    function that returns the current call with all its arguments captured as
     expressions. This is used mostly in the following circumstances: (1) to
-    record a call for a later use in possibly different environment, (2) pass
+    record a call for a later use in a possibly different environment, (2) pass
     most of the captured call arguments to another function or (3) evaluate
@@
 
+    \todo{Add that this pattern is advised by the R documentation of match.call}
+
   \item \emph{Block \c{(BLOCK)}} can appear anywhere, but it is almost always
@@
 \end{compactitem}
+
+\todo{Say somewhere that the R community commonly calls call/argument capturing (with substitute or match.call), Non-Standard Evaluation.}
 

### 2021-04-17 paper/main.tex
Proofread conclusion

@@
 expected, \eval is used mainly in libraries. We have found just a few \eval
-call sites in scripts coming from large corpus of Kaggle that represents code
+call sites in scripts coming from a large corpus of Kaggle representing code
 written by R practitioners. We used an instrumented version of the R virtual
-machine, \rdyntrace, to capture all \eval calls traces including the relevant
-side effects that happen within these calls. Several insights can be learnt
+machine, \rdyntrace, to capture all \eval calls traces, including the relevant
+side effects that happen within these calls. Several insights can be learned
 from the data.
@@
 The code executed by R \eval has unusual expressive power. It can essentially
-modify anything anywhere which is wary for both R developers and R compiler.
-The shapes of expressions passed to \eval are extremely diverse, ranging from simple variable lookups to complex blocks or creation of new functions.
-Luckily, most site only see a single shape.
+modify anything anywhere, which is wary for both R developers and R compiler.
+The shapes of expressions passed to \eval are extremely diverse, ranging from simple variable lookups to complex blocks or the creation of new functions.
+Luckily, most sites only see a single shape.
 \eval unlashes its power and shows up its singularity compared to \eval in other languages with the diversity of environments in which it can evaluate its expressions.
@@
 The expressive power of \eval shines even more  so in the case of side effects that can quite literally happen anywhere.
-The data however suggest that this is not the case.
-The majority of eval sites do not do any observable side effects and when they do, it happens in a predictable environment.
+The data, however, suggest that this is not the case.
+The majority of eval sites do not do any observable side effects,  and when they do, it happens in a predictable environment.
 
 The origins of the expression passed to \eval, either constructed from language expressions, or stemming from reflection, or built by parsing strings.
-Reflection and constructed expressions dominate, which suggest the main use of \eval is meta-programming.
+Reflection and constructed expressions dominate, which suggests the primary   use of \eval is meta-programming.
 % macros...
@@
 Similarly to \citep{ecoop11}, we started looking into \evals in R with the
-ambition to replace them by other features that are more amenable to static
-analysis. The data unfortunately suggests that replacing \eval is going to be
+ambition to replace them with other features that are more amenable to static
+analysis. The data, unfortunately, suggests that replacing \eval is going to be
 a lengthy journey. Even if we disregard core packages as they are part of the
 language and the \eval use is stable, in CRAN \eval is used widely and in more
-sophisticated way than was the case in JavaScript. There the vast majority of
+sophisticated ways than was the case in JavaScript. There the vast majority of
 sites fitted an extremely simple categorization based on simple regular

### 2021-04-17 paper/main.tex
sec 1 to 5

@@
 
-Our corpus is assembled from three sources: the \emph{Base} libraries 
-bundled with the language implementation, packages hosted on \emph{CRAN}, and scripts from \emph{Kaggle}.
+Our corpus is assembled from three sources: the \emph{Base} libraries bundled
+with the language implementation, packages hosted on \emph{CRAN}, and scripts
+from \emph{Kaggle}.
 
@@
 \mypara{Kaggle.} The Kaggle website ({\small \url{kaggle.com}}) is an online
-platform for data science. It allows users to submit problems and compete to
-solve them. Solutions, called \emph{kernels}, are uploaded as plain scripts or
-notebooks. The quality of code on Kaggle is not uniform. Each kernel is
-runnable, and thus there is one script per kernel. Input data is provided. We
-obtained \KaggleKernels kernels. Since Kaggle is not curated, we used SHA-1 hashes
-to identify and remove \KaggleDuplicates duplicate entries. The remaining
-\KaggleUnique kernels are unique solutions to \KaggleCompetitions competitions.
-They contain \KaggleCode lines of R code. Only \KaggleWithEvals kernels call
-\eval.
+platform for data science. It allows users to submit and compete to solve
+problems. Solutions, called \emph{kernels}, are uploaded as scripts or
+notebooks. While the quality of code is not uniform, each kernel is runnable,
+and maps to ones script. Input data is provided. We obtained \KaggleKernels
+kernels. Since Kaggle is not curated, we used SHA-1 hashes to identify and
+remove \KaggleDuplicates duplicate entries. The remaining \KaggleUnique kernels
+are unique solutions to \KaggleCompetitions competitions. They contain
+665K lines of R code. Only \KaggleWithEvals kernels call \eval.
 
@@
   documented and thus hard to resolve automatically.
-\item \emph{Extract.} Given installed packages and scripts, the next step is to
-  create runnable programs. The \genthat tool helpfully extracts all runnable
-  code snippets from a package and turns each of these into a self-standing
-  program~\cite{issta18}. Some Kaggle kernels are already scripts. For those,
-  nothing more needs to be done. Others kernels are packages as notebooks, either   as Rmarkdown or Jupyter. For those, we use \c{knitr} to extract
-   runnable code.
-  The body of each extracted program is instrumented with calls to our dynamic
-  analyzer to ensure that we only record calls to \eval from the code of
-  interest and not from bootstrapping or execution harness operations.
-\item \emph{Trace.} Each program executes using our dynamic analysis tool, a heavily instrumented interpreter that captures calls to \eval and
-  many other fine-grained runtime events. Packages are run twice, once to
-  capture \eval calls originating from package code and a second time to
-  capture calls coming from the base libraries. To avoid any interference, each
-  program is run in its own process.
-  As the GNU R bytecode compiler is written in R, we turn it off
-  to avoid recording \eval in its code, and fallback on the interpreter.
+\item \emph{Extract.} Given installed packages, the \genthat tool extracts all
+  runnable code snippets and turns each of these into a self-standing
+  program~\cite{issta18}. Some Kaggle kernels are already scripts and nothing
+  more needs to be done. Kernels released as notebooks are processed by
+  \c{knitr} which extracts runnable code. The body of each program is
+  instrumented with calls to our dynamic analyzer to ensure that we only record
+  calls to \eval from the code of interest and not from bootstrapping or
+  execution harness operations.
+\item \emph{Trace.} Each program executes using our dynamic analysis tool, a
+  heavily instrumented interpreter that captures calls to \eval and many other
+  fine-grained runtime events. Packages are run twice, once to capture \eval
+  calls originating from package code and a second time to capture calls coming
+  from the base libraries. To avoid any interference, each program is run in its
+  own process with the GNU R compiler turned off to avoid recording its
+  execution.
 \item \emph{Analyze.} Finally, the analysis output is merged, cleaned, and summarized
@@
 
-\medskip\noindent The code extraction and tracing steps of the pipeline are run
-in parallel~\cite{GNUparallel} orchestrated by a Makefile. Servers have
-identical environments thanks to docker images with all dependencies installed.
+\noindent The pipeline runs in parallel~\cite{GNUparallel} orchestrated
+by a Makefile. Servers have identical environments thanks to docker images with
+all dependencies installed.
 
@@
 based on GNU R 4.0.2 that exposes low-level callbacks for a variety of runtime
-events~\cite{oopsla19b}.
-
-The tracer registers callbacks to all \eval functions as well as a few other
-ones that allow us to identify better where the expressions come from. For
-example, we taint the results of calls to \c{parse} (turns strings into
-expressions) and \c{match.call} (reflects on current call). We also capture calls
-to R API for dynamic code loading (\eg\xspace \c{library}, \c{require}, and
-\c{source}). Next, we subscribe to the events related to a variable definition
-and assignment, allowing us to record side effects that happen in environments
-while evaluating code in \eval. One challenge we encountered during the analysis
-was that essentially R only provides source references for top-level calls in a
-function body or in the block surrounded by braces. That is to say, a method
-whose body would \c{eval(x)} would not have debug information that we can use to
-identify that particular \eval. The same expression surrounded by braces would
-have a line number and source reference. This is unfortunately not easily fixed
-in the R implementation. Instead, we extended the dynamic analysis tool to
-attach synthetic source code references to all \eval call sites by traversing
-ASTs. The tracer is implemented as an R package in 3.2K C++ and 1.3K of R code.
-For performance reasons, most of the tracing is done in C++. While in theory, the
-implementation is relatively straightforward, it is not so in practice:
+events~\cite{oopsla19b}. The tracer registers callbacks to all \eval variants
+and a few additional functions to assist in locating the origin of \eval
+arguments. For example, we taint the results of calls to \c{parse} and
+\c{match.call} (reflects on current call). We also capture calls to R API for
+dynamic code loading. Next, we subscribe to the events related to a variable
+definition and assignment, allowing us to record side effects that happen in
+environments while evaluating code in \eval. One challenge was that R only
+provides source references for block surrounded by braces. Thus, a function
+whose body is \c{eval(x)} would no usable debug information, while the same
+expression surrounded by braces would be fine. This is unfortunately not easily
+fixed. We extended the dynamic analysis tool to attach synthetic source code
+references to all \eval call sites by traversing ASTs. The tracer is implemented
+as an R package in 3.2K C++ and 1.3K of R code. For performance reasons, most of
+the tracing is done in C++. While in theory, the implementation is relatively
+straightforward, not so in practice:
 %
@@
 
-\item While the R interpreter is implemented in C, a lot of the core
-  functionality is done in R. For example, package loading is implemented in R
-  using \eval or the S4 and R6 object systems. This makes it hard in the tracer
-  to separate the \eval that are essential to user programs from the
-  accidental ones that are products of the way R implements its basic
+\item While the R interpreter is implemented in C, most core functionality is in
+  R. For example, package loading is implemented in R using \eval. This makes it
+  hard in the tracer to separate the \eval that are essential to user programs
+  from the accidental ones that are products of the way R implements its basic
   operations. This is even more so for the side-effects analysis.
@@
 
-\mypara{Limitations.} The analysis has some limitations. Even with the extension
-described above, there are still \PkgUndefinedRnd \eval calls without source
-references. This occurs when a reference to the \eval function is passed as an 
-argument to a higher-order function or when the \eval call originates from
-native code. However, these missing source references account for a meager
-\PkgUndefinedRatio of all calls; they are unlikely to affect our results. Other
-limitations are that we ignore calls to the native \eval function and to the
-alternate \c{rlang::tidy\_eval} function, which uses native \eval internally.
-The \c{rlang} package introduces a new kind of promise called \c{quosure} which
-is evaluated by \c{tidy\_eval}. None of these limitations should invalidate our
-conclusions.
+\mypara{Limitations.} Even with the extension described above, there are
+\PkgUndefinedRnd \eval calls without source references. This occurs when \eval
+is passed as an argument to a higher-order function or when the \eval call
+originates from native code. However, these missing source references account
+for a meager \PkgUndefinedRatio of all calls; they are unlikely to affect our
+results. Other limitations are that we ignore calls to the native \eval and to
+the alternate \c{rlang::tidy\_eval} uses native \eval internally. None of these
+limitations should invalidate our conclusions.
 
@@
 
-\begin{wrapfigure}{r}{5cm} \hspace*{-12mm}
+\begin{wrapfigure}{r}{5cm} \vspace{-1cm} \hspace*{-12mm}
   \centering
@@
   \eval call sites}%
-  \label{fig:pkgs-eval-callsites-hist}
+  \label{fig:pkgs-eval-callsites-hist}\vspace{-3mm}
 \end{wrapfigure}
@@
 
-\subsection{CRAN}
-
-There are \PkgEvalCallSites \eval sites in \PkgPackages packages. The proportion
-of packages calling \eval is \PkgPackagesRatio. Over half of these packages
-have fewer than 3 sites, and with the exception \MaxEvalCallSitesPackage, which
-has \MaxEvalCallSitesCount sites, all packages contain fewer than
-\MaxEvalCallSitesRest sites (\cf Fig~\ref{fig:pkgs-eval-callsites-hist}, which
-shows a histogram of sites per package). These sites appear in \PkgFunsWithEval
-functions (\CranFunsWithEvalRatio of all functions in CRAN).
-
 \begin{figure}[!b]
@@
 \begin{tabular}{@{}l@{\hspace{1.5cm}}l@{}}
-\begin{minipage} {5cm}
+\begin{minipage} {5cm}\small
   \begin{tabular}{|r@{\,}r@{\,}l@{}r|r@{\,}r@{\,}l@{}r|} \hline
@@
 &
-\begin{minipage}{7cm}
+\begin{minipage}{7cm}\small
 \begin{tabular}{|@{\,}r|rrrr|}\hline
@@
 
-%\end{figure}
-
-%\begin{figure}[b] \centering
-  \includegraphics[width=.78\textwidth]{traced-eval-callsites.pdf} \centering
+\end{figure}
+
+\begin{figure}[b] \centering
+  \includegraphics[width=.7\textwidth]{traced-eval-callsites.pdf} \centering
   \caption{\eval call sites coverage of the \PkgPackages packages.}%
@@
 
-For dynamic analysis, we run \CranRunnableScripts programs extracted from
-\CranPackages packages. Any run that does not exercise \eval is discarded. This
-left \packageNbruns runs from \packageCorpus packages. There were
-\packageAllcalls calls in \packageTriggeredpkgs packages originating from
-\PkgHitEvalCallSites unique sites. In terms of coverage, the data exercised
-\PkgHitEvalCallSitesAvgRatio of sites, a coverage similar to the code coverage
-metric for the packages, which is \PkgCodeCoverage. The fact that not all sites
-are exercised can be chalked down to incomplete tests and occasional analysis
-failures (\PkgFailedProgramsRatio of programs crashed or timed out).
-Fig~\ref{fig:traced-eval-callsites} shows the number of sites that were
-exercised; coverage is unequal. Figure~\ref{freq} summarizes the frequency of
-dynamic calls to \eval, with the left column being the number of calls and the
-right, the number of packages in that range. There are \packageFewcalls packages
-with low \eval frequency, fewer than 100 calls, and a smaller, but still
-significant, number of packages, \packageManycalls to be precise, that use \eval
-more than 1,000 times. Package \packageMaxcallspack makes \packageMaxcalls calls
-and thus accounts for over half of the observations.
-Figure~\ref{tab:variantseval} summarizes the use of variants of \eval. For each
-of the four variants, the first row shows the number of sites in the corpus
-(\emph{static}), the second row is the number of sites that were encountered
-during analysis (\emph{exercised}), the last row is the number of calls
-(\emph{invocations}). The overwhelming majority of sites and calls are to \eval
-itself, \c{eval.parent} is rare, and both \c{evalq} and \c{local} are barely
-used at all. The difference between sites and exercised sites underscores the
-limitations due to code coverage.
+\paragraph{CRAN}
 
@@
 \end{wrapfigure}
+There are \PkgEvalCallSites \eval sites in \PkgPackages packages. The proportion
+of packages calling \eval is \PkgPackagesRatio. Over half of these packages
+have fewer than 3 sites, and with the exception \MaxEvalCallSitesPackage, which
+has \MaxEvalCallSitesCount sites, all packages contain fewer than
+\MaxEvalCallSitesRest sites (\cf Fig~\ref{fig:pkgs-eval-callsites-hist}, which
+shows a histogram of sites per package). These sites appear in \PkgFunsWithEval
+functions (\CranFunsWithEvalRatio of all functions in CRAN).
+For dynamic analysis, we run \CranRunnableScripts programs extracted from
+\CranPackages packages. Any run that does not exercise \eval is discarded. This
+left \packageNbruns runs from \packageCorpus packages. There were
+\packageAllcalls calls in \packageTriggeredpkgs packages originating from
+\PkgHitEvalCallSites unique sites. In terms of coverage, the data exercised
+\PkgHitEvalCallSitesAvgRatio of sites, a coverage similar to the code coverage
+metric for the packages, which is \PkgCodeCoverage. The fact that not all sites
+are exercised can be chalked down to incomplete tests and occasional analysis
+failures (\PkgFailedProgramsRatio of programs crashed or timed out).
+Fig~\ref{fig:traced-eval-callsites} shows the number of sites that were
+exercised; coverage is unequal. Figure~\ref{freq} summarizes the frequency of
+dynamic calls to \eval, with the left
+column being the number of calls and the right, the number of packages in that
+range. There are \packageFewcalls packages with low \eval frequency, fewer than
+100 calls, and a smaller, but still significant, number of packages,
+\packageManycalls to be precise, that use \eval more than 1,000 times. Package
+\packageMaxcallspack makes \packageMaxcalls calls and thus accounts for over
+half of the observations. Figure~\ref{tab:variantseval} summarizes the use of
+variants of \eval. For each of the four variants, the first row shows the number
+of sites in the corpus (\emph{static}), the second row is the number of sites
+that were encountered during analysis (\emph{exercised}), the last row is the
+number of calls (\emph{invocations}). The overwhelming majority of sites and
+calls are to \eval itself, \c{eval.parent} is rare, and both \c{evalq} and
+\c{local} are barely used at all. The difference between sites and exercised
+sites underscores the limits of code coverage.
 
@@
 
-\Eval takes any value, but if its argument is not an expression, \eval returns
-it unchanged. Expressions account for \packageCodepercent of arguments in our
-corpus. More specifically, \packageSymbolpercent are \texttt{symbol}s (single
-variables such as \c{x}), \packageLanguagepercent are \c{language} objects
-(function calls such as \c{f(x)} or \c{x+1}), and \packageExpressionpercent are
-\c{expression} objects (lists of expressions such as function calls or
-symbols). Further inspection reveals that most symbols,
+\Eval takes any value, but if its argument is not an expression, it is returned
+unchanged. Expressions account for \packageCodepercent of arguments. More
+specifically, \packageSymbolpercent are \texttt{symbol}s (single variables such
+as \c{x}), \packageLanguagepercent are \c{language} objects (function calls),
+  and \packageExpressionpercent are \c{expression} objects (lists of
+  expressions). Further inspection reveals that most symbols,
   \packageGgplotsymbolpercent to be exact, come from a single site in the
   \c{ggplot2} package and have the value \c{\_inherit}.\footnote{This models
-inheritance in \c{ggproto}, one of the many object-oriented systems in
-R, used exclusively in the \c{ggplot2} graphics library.}
+  inheritance in \c{ggproto}, one of the many object-oriented systems in R, used
+  in the \c{ggplot2} graphics library.}
 
@@
 
-\begin{figure}[tb!]
+\begin{figure}[h!]
 \begin{tabular}{@{}c@{}c@{}}
@@
 
-\subsection{Base}\label{sec:usage-base}
+\paragraph{Base}\label{sec:usage-base}
 
@@
 (\baseLanguagepercent) of the arguments are language objects. The median
-argument size is \baseMedianszeval, which is more than for CRAN and the maximum size is \baseMaxszeval, less than for CRAN.
-Small instruction counts ($<50$) amount to \baseSmalleventspct of calls.
-A single site in the \c{match.arg}
+argument size is \baseMedianszeval, which is more than for CRAN and the maximum
+size is \baseMaxszeval, less than for CRAN. Small instruction counts ($<50$)
+amount to \baseSmalleventspct of calls. A single site in the \c{match.arg}
 function is responsible for \baseTopFuncPercent of all recorded calls. This
@@
 matching. It is heavily relied on in R functions that use string arguments to
-parameterize function behavior, a popular API mechanism in R. For example, in a
-body of \c{center <- function(x, type = c("mean", "median", "trimmed"))} one
-can get the value of \c{type} parameter using \c{type <- match.arg(type)}. A
-call \c{center(x, "trim")} will match \c{type} to \c{"trimmed"}.
-
-The challenge dealing with Base is that every program uses it. It can thus
-generate extreme amounts of data, but the data is quite predictable. For the
-rest of the paper, we only mention Base when there are surprising
-observations.
-
-\medskip
-
-\subsection{Kaggle}
+parameterize function behavior, common practice in R. For example, in a body of
+\c{center<-function(x,type=c("mean","median","trimmed"))} one can get the
+value of \c{type} parameter using \c{type<-match.arg(type)}. A call
+\c{center(x,"trim")} will match \c{type} to \c{"trimmed"}. The challenge
+dealing with Base is that every program uses it. It can thus generate extreme
+amounts of data, but the data is quite predictable. From now on, we only mention
+Base when it surprises.
+
+\paragraph{Kaggle}
 
 In total, \kaggleAllcalls \eval were recorded, all to \eval function. Out of
-\kaggleStaticeval sites from \KaggleWithEvals scripts, only \kaggleTriggeredeval sites
-in \kaggleNbruns scripts were hit; \KaggleFailedScripts scripts failed to run,
-and the others did not exercise the \eval sites. This is partially expected
+\kaggleStaticeval sites from \KaggleWithEvals scripts, only \kaggleTriggeredeval
+sites in \kaggleNbruns scripts were hit; \KaggleFailedScripts scripts failed to
+run, and the others did not exercise the \eval sites. This is partially expected
 as Kaggle code does not need to abide by any checks. Upon a manual inspection,
@@
 finished, using misspelled package names, hard-coded file paths, or accessing
-missing files.
-
-Most call sites are invoked only once. Only one site is called \kaggleMaxcalls times.
-Expressions amount for \kaggleCodepercent of arguments. Unlike the CRAN corpus,
+missing files. Most call sites are invoked once. Only one site is called
+\kaggleMaxcalls times. Expressions amount for \kaggleCodepercent of arguments,
 there are very few symbols (\kaggleSymbolpercent). Most expressions result from
@@
 
-\medskip
-
-\subsection{Discussion}
-These results show that \eval is central to the language implementation as it is
-omnipresent in the relatively small Base library. The Base library is, in turn,
-used by every single package. To avoid having Base dominate the result, we have
-excluded calls occurring there from the data we report when analyzing packages.
-CRAN Packages, which are typically developed by experienced programmers, make
-regular and varied use of \eval. They represent our most interesting data set as
-these packages result from over 20 years of contributions by thousands of
-authors. The code in packages is well maintained and relatively well tested.
-Finally, Kaggle scripts are often written by less sophisticated users and
-likely perform more straightforward tasks, thus having a lesser need for \eval, and its uses originate from strings as these are easier to manipulate for end-users.
-Due to the relative paucity of data in Kaggle, we do not pursue the data set further.
+\paragraph{Discussion}
+\Eval is central to R's implementation as it is omnipresent in the Base library.
+That library is, in turn, used by every single package. To avoid Base dominating
+the result, we excluded calls to it from the results. CRAN Packages, which are
+typically developed by experienced programmers, make regular and varied use of
+\eval. They represent our most interesting data set as these packages result
+from over 20 years of contributions by thousands of authors. The code in
+packages is well maintained and relatively well tested. Finally, Kaggle scripts
+are often written by less sophisticated users and likely perform more
+straightforward tasks, thus having a lesser need for \eval, and its uses
+originate from strings as these are easier to manipulate for end-users.
 
@@
 but this is a limitation of this categorization.} In our corpus,
-\packageNbCallSitesUniqueActualValue call sites only ever see a simple value.
-Manual inspection reveals cases such as the following
+\packageNbCallSitesUniqueActualValue call sites only ever see a simple value,
+such as:
 \begin{lstlisting}
@@
 where the argument to \eval is a string which \eval simply returns. Of twenty
-randomly selected \evals, \packageUsefulValueEvalPercent of them  need % There are a majority which needs eval
-\eval for some inputs. Either we are dealing with a value that needs to be
-constructed dynamically, or the value is a default case that sometimes is
-replaced by a more interesting expression. This form is usually evaluated in few
-interpreter steps; in fact, the median is only
-\packageMinimizedmedianoperationscRnd. The environment in which they evaluate is
-mostly irrelevant (unless a built-in operator is redefined).
-
-\medskip\noindent\EE{$min(e)=\c{X}$} Variable lookups are the most common form; they are found in 28\% of the sites. This form includes simple variable reads,
+randomly selected \evals, \packageUsefulValueEvalPercent of them need \eval for
+some inputs. Either we are dealing with a value that needs to be constructed
+dynamically, or the value is a default case that sometimes is replaced by a more
+interesting expression. This form is usually evaluated in few interpreter steps;
+in fact, the median is only \packageMinimizedmedianoperationscRnd. The
+environment in which they evaluate is mostly irrelevant.
+
+\medskip\noindent\EE{$min(e)=\c{X}$} Variable lookups are the most common form;
+they are found in 28\% of the sites. This form includes simple variable reads,
 \eg~\c{x}, which are \packageNbSymbolVarSitePercent of \c X. The form also
@@
 
-\ \\[5mm] % To push the box on the next page
 \medskip\noindent\EE{$min(e)=\c{F()}$} This form captures simple function calls,

### 2021-04-17 paper/main.tex
sec 5

@@
 operators, the direct assignment {\tt <-}, and assignment to the parent
-environment {\tt <\,\!<-}, the \c{assign} function and the \c{\$} form. Assignments occur in 5\% of the
-sites. They represent the most obvious source of side effects. The median number
-of operations is \packageMinimizedmedianoperationsiRnd; the minimum is 3.
+environment {\tt <\,\!<-}, the \c{assign} function and the \c{\$} form.
+Assignments occur in 5\% of the sites. They represent the most obvious source of
+side effects. The median count is \packageMinimizedmedianoperationsiRnd (min=3).
 
@@
 a promise containing a value, then the computation will occur. This form occurs in
-6\% of sites and typically does not perform much work in the interpreter.
+6\% of sites and typically does not perform much work.
 
@@
 respectively, \packageMinimizedmedianoperationsdRnd and
-\packageMinimizedmedianoperationsbRnd. Also,  \packageMinimizedpercentparentframesd and \packageMinimizedpercentparentframesb
+\packageMinimizedmedianoperationsbRnd. Also,
+\packageMinimizedpercentparentframesd and \packageMinimizedpercentparentframesb
 of these expressions run in function environments.
 
-\medskip\noindent\EE{$min(e)=\c{FUN}$} This form captures expressions that define functions, \eg~\c{function(x) x+1}, and do nothing else. \c{FUN} occurs in
-\packageFunctionDefinitionSitesPercent of sites. Evaluating a function
+\medskip\noindent\EE{$min(e)=\c{FUN}$} This form captures expressions that
+define functions, \eg~\c{function(x) x+1}, and do nothing else. \c{FUN} occurs
+in \packageFunctionDefinitionSitesPercent of sites. Evaluating a function
 definition is done in 2 interpreter steps; the data does not record the work
@@
 
-\paragraph{Consistency} It is interesting to consider how many different
-forms any given site sees. The more forms, the harder it will be to characterize
-the program's behavior at that site. Luckily, \packageNbOneMinimizedPercent
-of sites only see a single form.
-In other words, a JIT compiler could correctly predict the \emph{shape} of an expression by looking just at the first call in \packageNbOneMinimizedPercent of the sites.
-Few sites are highly polymorphic (8 or more different forms).
-These include the pipe operator of the \c{magrittr} package, which is used to compose functions.
+\paragraph{Consistency} It is interesting how many different
+forms any given site sees. The more forms, the harder it is to characterize
+behavior at that site. Luckily, \packageNbOneMinimizedPercent of sites only see
+a single form. In other words, a compiler could correctly predict the
+\emph{shape} of an expression by looking just at the first call in
+\packageNbOneMinimizedPercent of the sites. Few sites are highly polymorphic (8
+or more different forms). These include the pipe operator of the \c{magrittr}
+package, used to compose functions.
 
 \paragraph{Discussion} The variety of uses of \eval is evidenced by the number
-of different forms observed in CRAN.
-In comparison, JavaScript \eval usage was more straightforward and more predictable as reported by \citet{oopsla12b}, with 98.7\% of the sites with only one form.
-Nevertheless, simple forms dominate in R also, and there are many cases where \eval could be replaced by less powerful constructs.
+of different forms. In comparison, JavaScript \eval usage was more
+straightforward and more predictable as reported by \citet{oopsla12b}, with
+98.7\% of the sites with only one form. Nevertheless, simple forms dominate in R
+also, and there are many cases where \eval could be replaced by less powerful
+constructs.
 
@@
 
-\subsection{Side effects in \eval}
+\subsection{The effects of \eval}
 
@@
 environments, and finally, \SESitesBothEnvirRatio does both.
-Table~\ref{tab:se-env} shows the class of the environment where \eval side
-effects happen (\emph{target environment}). It follows the classification
-described in Section~\ref{sec:env} with two extra classes: \emph{Object},
-which represents a side-effect that happens in S4 or R6 object environments, and
-\emph{Local} that corresponds to a side effect in local variables of the \eval
-caller (offset 0 in Section~\ref{sec:env}).  The table shows data for both \eval sites and functions. For a
-function to have a given target environment class, all of its \eval sites that
-do side effects must have the same target environment class The majority of the
-\eval sites (\SESitesInOneClass) do all side effects consistently in one
-environment class. The same happens at the function level. Almost half of the
-sites (over a third of the functions) do side effects in either \emph{Local} or
-\emph{Object} environments. This gives a ray of hope for the compiler. Even
-though it is possible to do anything anywhere, the data suggest that most
-side-effects are sane.
 
@@
 \end{wraptable}
+
+Table~\ref{tab:se-env} shows the class of the environment where \eval side
+effects happen (\emph{target environment}). It follows the classification
+described in Section~\ref{sec:env} with two extra classes: \emph{Object}, which
+represents a side-effect that happens in S4 or R6 object environments, and
+\emph{Local} that corresponds to a side effect in local variables of the \eval
+caller (offset 0 in Section~\ref{sec:env}). The table shows data for both \eval
+sites and functions. For a function to have a given target environment class,
+all of its \eval sites that do side effects must have the same target
+environment class The majority of the \eval sites (\SESitesInOneClass) do all
+side effects consistently in one environment class. The same happens at the
+function level. Almost half of the sites (over a third of the functions) do side
+effects in either \emph{Local} or \emph{Object} environments. This gives a ray
+of hope for the compiler. Even though it is possible to do anything anywhere,
+the data suggest that most side-effects are sane.
 

### 2021-04-17 paper/main.tex
5.3

@@
 
-Contrary to the Javascript \eval, which does not precise the scope in which to evaluate,the second argument of \eval in R, \c{envir}, specifically is dedicated to that.
-The environment in which evaluation happens determines what is visible to the
-computation started by \eval and the potential reach of its side-effects. The
-second argument to \eval specifies that environment. Environments of \eval can be
-classified into the following four kinds:
+Contrary to the Javascript \eval, which does not specify the environment in
+which to evaluate, the second argument of \eval in R, \c{envir}, is dedicated to
+that. This argument determines what is visible to the computation started by
+\eval and the potential reach of its side-effects. Environments used by \eval
+can be classified into the following four kinds:
 
@@
 \noindent
-As shown in Table~\ref{tab:highlevelenvironments}, most calls evaluate in a
-function environment with the \c{global} kind as a distant second. This means
-that most variable lookups and most side effects either read/update existing
-local variables or introduce new ones. But for which function? From the point of
-view of the function that called \eval, Table~\ref{tab:funoffset} gives an
-offset on the call stack.
-Thus 0 is the direct caller and corresponds to local variables; 1 is its parent, and so on.
-The data suggests that in 81\% of cases, \eval accesses its caller's environment -- this means  the variables of the function where \eval textually
-occurs are read and written to. It is interesting that some 1.5\% of sites
-evaluate a code in an environment that is three frames or above from \eval. This
-distance implies that, in general, modular reasoning is impossible. Understanding the behavior of any piece of code entails fully understanding the
-behavior of all functions that the code may call as the actions of \eval may
-happen at a distance. Finally, note that any particular site may have several
-kinds of environments, but in \packageNbOneCategoryEnvirSitePercent of the cases,  a site has a single kind.
-It means that a JIT compiler can predict the \emph{kind} of  expression passed to \eval in \packageNbOneCategoryEnvirSitePercent of the sites by just examining at the first invocation of a site.
+As shown in Table~\ref{tab:highlevelenvironments}, most calls evaluate in the
+scope of a function with global as a distant second. This means that most
+variable reads and most side-effects operate on local variables. But for which
+function? From the point of view of the caller of \eval,
+Table~\ref{tab:funoffset} identifies that function by its offset on the call
+stack. Thus 0 is the direct caller, 1 is its parent, and so on. The data
+suggests that in 81\% of cases, \eval accesses the caller -- this means the
+variables of the function where \eval textually occurs are read and written to.
+It is noteworthy that 1.5\% of sites evaluate three frames or above. This means
+that, in general, modular reasoning is impossible in R. To understand what any
+given code snippet may do requires fully understanding all the functions that
+may be called, transitively, from that snippet. The actions of \eval happen at a
+distance. Finally, note that any site may see several kinds, but in
+\packageNbOneCategoryEnvirSitePercent of the cases they have a single kind. It
+means that a compiler could predict the \emph{kind} of environment after the
+first invocation.
 
@@
 \noindent
-Sites that evaluate in the global environment are likely split between
-intentional and accidental use. Direct references to the top-level, using
-\c{globalenv()} or \c{.GlobalEnv}, are rare; they occur in only
-\packageNbExplicitGlobalSites sites. We suspect that the majority of uses of the
-global environment come from the fact that our corpus consists of scripts loaded at the top level. Thus, that environment may often be the caller of
-\eval or close to it. The reason we make this point is that values stored in the
+Global \evals are likely split between intentional and accidental one. Direct
+references to the top-level, using \c{globalenv()} or \c{.GlobalEnv}, are rare;
+they occur in only \packageNbExplicitGlobalSites sites. Thus we suspect most
+uses of global are accidental. They arise from properties of our corpus, most
+scripts are top-level code snippets. Thus, global is often the caller of \eval
+or close to it. The reason we make this point is that values stored in the
 global environment are visible to all functions and are not reclaimed by the
@@
 
-Synthetic environments need a parent. It is specified as an argument of the \c{new.env} function, or when \c{envir} is a list or a data frame, \eval uses its third argument
-(\c{enclos}) as the parent.  The parent is used to lookup variables not found in its
-child (side effects go to the child only). Table~\ref{tab:newenvs} shows the
-kinds of the parents of synthetic environments. Most of them are functions, then
-come global and package.
+Each synthetic environment has parent that is specified when calling
+\c{new.env}. When \c{envir} is a list or a data frame, \eval uses its third
+argument (\c{enclos}). The parent is used to search for variables not found in
+its child (side-effects stay in the child). Table~\ref{tab:newenvs} shows parent
+kinds for synthetic environments. Most of them are functions, then come global
+and package.
 
 \paragraph{Discussion}
-The data presented in this section is what one could expect. ({\it Function})
-The majority of \eval call sites access the environment of some function on the
-call stack. Usually, the current function, but sometimes frames arbitrarily far
-up the call stack. One thing the data does not say is how that environment was
-obtained. The expected case is that the environment was the one obtained from a
-promise, as promises combine code and the environment in which that code
-originates. Likely less frequent are cases where \eval is provided the results
-of programmatically selecting some call stack. ({\it Synthetic}) The relatively
-high frequency of synthetic environments corresponds to cases when one wants to
-evaluate an expression in either a restricted environment, use a data
-structure as an environment. ({\it Global}) The fact that the global environment
-shows up so frequently is likely an artifact of how the code is run. ({\it
-  Package}) There are only \packageNbPackageNamespaceEnvSites sites that use a %77 sites
- package environment in \eval.
-This is probably best as it is an improper form of adding bindings or modifying functions of a loaded package.
-Many more sites instead wrap a package environment in a new
-environment to create singletons in a package.
-
-%% The \c{imchange} function of package \c{imager} makes it possible to modify
-%% images using a dedicated formula syntax.\footnote{Inspired by {map} in
-%% package \emph{purr}.} Here, \eval is evaluated in \c{newenv}, which creates a
-%% new environment that inherits from \c{parent.frame()} by default (classified
-%% as 1+).
-%% \begin{lstlisting}
-%%  newenv <- new.env()
-%%  ...
-%%  fo <- parse(text=as.character(fo)[2])
-%%  im[where] <- eval(fo,envir=newenv,enclos=env)
-%% \end{lstlisting}
-
-%% \c{adjCoef} in package \emph{actuar} find the root of an equation defined by
-%% a function \c{h} whose arguments must be named \c{x} and \c{y}. \c{h} is
-%% transformed into an auxiliary function \c{h2} that can be optimized. Here,
-%% the list used for \c{envir} ensures the correspondance between the textual
-%% arguments of \c{h} and the arguments of \c{h2}.
-%% \begin{lstlisting}
-%%  sh <- substitute(h)
-%%  fcall <- paste(sh, "(x, y)")
-%%  ...
-%%  h2 <- function(x, y)
-%%  eval(parse(text = fcall),
-%%  envir = list(x = x, y = y),
-%%  enclos = parent.frame(2))
-%% \end{lstlisting}
+The data presented in this section is what one could expect. For {\it
+  functions}, they represent the main user case for \eval. Thus \eval is really
+extending the behavior of that function by executing a dynamically selected
+piece of code in the function's scope. Typically, variables are read (mostly),
+written (less), but there are also some cases where new variables are injected
+or existing variables are deleted. Usually, this happens in the current
+function, but also in frames arbitrarily far up the call stack. One data point
+we lack is how the target environment was obtained. The expected case is that
+the body of a promise was modified and the resulting expression was evaluated in
+the same environment the promise originated from. Likely less frequent are cases
+where \eval is provided the results of programmatically selecting some call
+stack. For {\it synthetics}, their relatively high frequency correlates to use
+cases where one wants to evaluate an expression in either a restricted
+environment or to use a data structure as an environment. For {\it global}, its
+relatively high frequency is likely an artifact of how the code of our corpus is
+is run. For {\it packages}, only \packageNbPackageNamespaceEnvSites sites have
+this kind. This is probably for the best as mutating the bindings of
+a loaded package is improper and R tries to make it difficult.
 
@@
 \item {\it Constructed:} Expressions can be constructed by invoking the
-  \c{quote}, \c{enquote}, \c{expression} functions. Function arguments are passed as
-  promises, and \c{substitute} is used to retrieve the source expression
-  associated with the promise.
-  \item {\it Reflection:} This group corresponds to uses of the \c{match.call}
-  function to reflectively capture the expression that invoked the current
-  function.
-\item {\it String:} Finally, expressions can be created from strings by invoking
-   \c{parse}, \c{str2expression}, or \c{str2lang}.
+  \c{quote}, \c{enquote}, \c{expression} functions. Function arguments are
+  passed as promises, and \c{substitute} is used to retrieve the source
+  expression associated with the promise.
+\item {\it Reflection:} This group corresponds to uses of \c{match.call} to
+  reflectively capture the expression that invoked the current function.
+\item {\it String:} Expressions created from strings by invoking \c{parse},
+  \c{str2expression}, or \c{str2lang}.
 \end{compactitem}
 
-\noindent
-Table~\ref{tab:provenance} summarizes the expression provenance in our corpus.
-This data is obtained by dynamically tainting values as they are produced by the
-various sources. Due to technical reasons, there is some imprecision in the
-results. In particular, we are not able to classify all sites. Manual inspection
-of numerous examples suggests that the classified results are accurate.
+\vspace{1mm}\noindent Table~\ref{tab:provenance} summarizes the expression
+provenance in our corpus. This data is obtained by dynamically tainting values
+as they are produced by the various sources. Due to technical reasons, there is
+some imprecision in the results. In particular, we are not able to classify all
+sites. Manual inspection of numerous examples suggests that errors are rare.
 
@@
 input, parses it, and passes it to \eval. The use of strings seems to correlate
-with less sophisticated programmers; in our Kaggle corpus,
+with less sophisticated programmers; in Kaggle,
 \kaggleParseExprsSitePercent of sites use strings.
@@
 \paragraph{Discussion}
-
-% Compare to JS ? Everything comes from a string in JS anyway
-% Composite (concatenation of strings), DOM and constant are the most common
-% ...
 The origin data suggests that constructing expressions from strings is a
-minority of the use cases. Instead, the constructed category shows that most \evals comes from code that was processed by the compiler, and may be
-slightly modified by the programmer before invoking \eval. Both constructed and
+minority of the use cases. Instead, the constructed category shows that most
+\evals comes from code that was processed by the compiler, and may be slightly
+modified by the programmer before invoking \eval. Both constructed and
 reflection categories roughly correspond to meta-programming. Some of these use

### 2021-04-17 paper/main.tex
5.4 wip

@@
 
-The code executed by \eval can do side effects. From the compiler perspective,
-we care about the observable side effects in environments---\ie variable
-definitions, updates, and removals that are visible after a call to \eval
-finishes. Knowing where---\ie in which environments do these side-effects
-happen, can help us determine how much of the compiler knowledge about the
-program will be potentially invalidated.
-
-The \rdyntrace contains low-level hooks capturing the life-cycle of an
-environment and variables defined in it. We register callbacks in the tracer
-that records each environment creation and removal with updates to its
-bindings. From the recorded data, we ignore side effects coming from \eval
-sites from unit testing frameworks.\footnote{In the corpus, we have \c{RUnit,
-testthat, tinytest} and \c{unitizer} unit testing frameworks.} They run the
-testing code via \eval, and thus the results would be biased because of the
-high number of code in these tests (\CranRunnableCodeTestsRnd lines).
-
-From the \packageNbrunsRnd programs, we capture \SEAllRnd side effects from
+\Eval may peform side-effects. From a compiler's perspective, we care about
+effects that can be observed --- \ie variable definitions, updates, and removals
+that are visible after a call finishes. Knowing in which environments these
+side-effects happen, can help us determine how much of the compiler knowledge
+about the program will be invalidated. Our analysis records information about
+every environment. From the recorded data, we discard side-effects coming from
+unit testing frameworks.\footnote{In the corpus, we have \c{RUnit, testthat,
+  tinytest} and \c{unitizer} testing frameworks.} They run their tests via
+\eval, and thus everything becomes a side-effect.
+
+From our 98K
+%\packageNbrunsRnd
+programs, we capture
+14M
+%\SEAllRnd
+side-effects from
 \SEAllCallsRnd \eval calls (\SEUserCallsToAllRatio) in \SEAllSites sites
 (\SEUserSitesToAllRatio). Again, the challenge is to remove the accidental
-side effects caused by the R virtual machine implementation that are not
+side-effects caused by the R virtual machine implementation which are not
 related to the user code. For example, the \c{.Random.seed} variable, which
 contains the state of the random number generator, is saved and restored from
-and to the global environment every time the user calls one of the base
-routines that use random numbers.
+and to the global environment every time  a statistical routine is called.
 %
-% \begin{wrapfigure}{r}{5cm} \hspace*{-20mm} \centering
-% \includegraphics[width=90mm]{se-types.pdf} \caption{Distribution of R types
-% in target environments for side effects.} \label{fig:se-types}
-% \end{wrapfigure}
-%
-Removing them leaves us with \SEUserRnd side effects (\SEUserRatio) from
-\SEUserCallsRnd \eval calls (\SEUserCallsRatio) in \SEUserSites sites
-(\SEUserSitesRatio). These sites are part of \SEUserFunctions R functions
-(\SEUserFunctionsToAllRatio of all functions doing \eval) in \SEUserPackages
-packages. From these functions, \SEFunsNighty are responsible for 90\% of side
-effects. Half of the side-effects come just from three functions:
-\c{plyr::allocate\_column} (allocates space for a new data frame column),
-\c{withr::execute\_handlers} (executes deferred expressions), and
+Removing those leaves us with \SEUserRnd side effects from \SEUserCallsRnd \eval
+calls in \SEUserSites sites. These sites are part of \SEUserFunctions R
+functions (\SEUserFunctionsToAllRatio of all functions doing \eval) in
+\SEUserPackages packages. From these functions, \SEFunsNighty are responsible
+for 90\% of side effects. Half of the side-effects come just from three
+functions: \c{plyr::allocate\_column} (allocates space for a new data frame
+column), \c{withr::execute\_handlers} (executes deferred expressions), and
 \c{foreach::doSEQ} (executes an expression on each element in a collection,

### 2021-04-17 paper/main.tex
Updates in 6

@@
     to create or update a binding in a specific environment where either the
-    name or the value come from some expression. Most cases originate from trivial
-    code generation where the assign expression is assembled using \c{parse} or
-    \c{substitute}, often in a loop: \lstinline|eval(parse(text=paste(capture.output(default_args[[i]]), "<-", i)))|.
+    name or the value come from some expression. Most cases originate from
+    trivial code generation where the assign expression is assembled using
+    \c{parse} or \c{substitute}, often in a loop. For example, the
+    \c{PopED::get\_rse} uses it to assign a number of variables in the current
+    scope: \lstinline|eval(parse(text=paste(default_args[[i]]), "<-", i))|.
 
@@
   \item \emph{Function calls \c{(F(), F(X), F(F(X)))}} represent two very
-    frequent patterns: \emph{code generation} and \emph{call reflection}. There are numerous
-    cases for code generation. One occurring usage is to generate ad-hoc calls
-    to libraries such as \c{ggplot2} when one needs to parameterize data or
-    aesthetics selection from a variable. For example \lstinline|ggplot(df, aes(x=A, y=B))|
-    maps X and Y axes in the plot to \c{A} and \c{B} columns in
-    \c{df}. It might not be immediately clear\footnote{\cf
-    \url{https://stackoverflow.com/questions/22309285/how-to-use-a-variable-to-specify-column-name-in-ggplot}}  for a ggplot2 user
-    how to modify the call to pick the actual mapping from \c{A} and \c{B}
-    variables. Another case is the traditional use of \eval as a means to
-    reduce boilerplate code; simple and repetitive code can easily be replaced
-    with judicious use of \eval. For example, the \c{data.table} package uses
-    \eval to call the \c{options} function with named arguments taken from a
-    vector of strings. While the benefits are limited in this example, it is an
-    attractive tool for programmers.
+    frequent patterns: \emph{code generation} and \emph{call reflection}. There
+    are numerous cases for code generation.
     %
-    \begin{lstlisting}
-    opts = c("datatable.verbose"="FALSE", # ...many others
+    One occurring usage (present also in the Kaggle scripts) is to generate
+    ad-hoc calls to libraries such as \c{ggplot2} when one needs to
+    parameterize data or aesthetics selection from a variable. For example
+    \lstinline|ggplot(df, aes(x=A, y=B))| maps X and Y axes in the plot to
+    \c{A} and \c{B} columns in \c{df}. It might not be immediately
+    clear\footnote{\cf
+    \url{https://stackoverflow.com/questions/22309285/how-to-use-a-variable-to-specify-column-name-in-ggplot}}
+    for a ggplot2 user how to modify the call to assign the axis columns from
+    variables. Another case is the traditional use of \eval as a mean to reduce
+    boilerplate code; simple and repetitive code can easily be replaced with
+    judicious use of \eval. For example, the \c{data.table} package uses \eval
+    to call the \c{options} function with named arguments taken from a vector
+    of strings:
+
+    \begin{lstlisting}[gobble=2]
+    opts = c("datatable.verbose"="FALSE", ... )
     for (i in names(opts)) eval(parse(text=paste0("options(",i,"=",opts[i],")")))
@@
     %
-    \begin{lstlisting}
+    \begin{lstlisting}[gobble=2]
     Call <- match.call()
@@
 
-    \todo{Add that this pattern is advised by the R documentation of match.call}
-
   \item \emph{Block \c{(BLOCK)}} can appear anywhere, but it is almost always
@@
 
-\subsection{Domain Specific Language.} R expressions are built and transformed symbolically and then evalled to a concrete value.
+\subsection{Domain Specific Language}
+
+The ability to capture current call as a language object (\c{match.call()}),
+extract the unevaluated expression from function arguments (\c{substitute()})
+and \eval with first class environment makes the R toolkit for building DSLs.
+There are numerous widely spread DSLs used in R. For example, \c{data.table}
+package defines a DSL for data frame manipulation using the subset operator
+\c{[}. This allowing one to write compact queries such as \c{flights[carrier ==
+  "AA", .N, by = origin]}\footnote{The \c{data.table} subset operation is
+defined as \c{DT[i, j, k]}, where \c{j} expression is used to calculate on a
+subset/reorder of rows from \c{i} grouped by \c{k}.} which will compute the
+number of flights done by American Airlines for each origin airport.
+Essentially, it uses \c{substitute} to capture the operand's expressions,
+transforms then appropriately and finally uses \eval to execute the query.
+
+Another example is the \c{tidyselect} package that defines a DSL for selecting
+columns in a data frame powering number of data manipulation operation in the
+\c{data.table} main competitor, the \c{dply} package. It uses eval to build an
+interpreter for evaluating column selectors such as \c{select(df,
+starts\_with("length") \& !(name:mass))} which returns all columns in \c{df}
+that starts with the string \c{length} and is not between \c{name} and
+\c{mass} columns.
+
+% TODO Pierre
 For instance, base R packages \emph{stats} provides two operators for \emph{symbolic differentiation}, \c{D} and \c{deriv}.
@@
 % See https://github.com/cran/MCMCglmm/blob/cfac9e67ec73a3db2142826faae35e5d2318da31/man/evalDtensor.Rd
-
-Package \c{tidyselect} is the backend of several \c{dplyr} and \c{tidyr} functions to select columns.
-It offers a DSL that makes it possible to select a set of columns, over a range of columns, a complement, an intersection, an union, with a regular expression and so on.
-\eval is used to traverse selectors stored as slots of a dataframe.
 

### 2021-04-17 paper/main.tex
varia

--- 2021-04-17 paper/main.tex
@@
 \begin{wraptable}{r}{7cm}\small\centering
-  \small
-  \centering
   \begin{tabular}{l|r|r|r|r}\hline
@@
   \end{tabular}
-  \caption{Target environments for \eval side-effects} \label{tab:se-env}
+  \caption{Target environments for side-effects} \label{tab:se-env}
 \end{wraptable}
 
-Table~\ref{tab:se-env} shows the class of the environment where \eval side
-effects happen (\emph{target environment}). It follows the classification
-described in Section~\ref{sec:env} with two extra classes: \emph{Object}, which
-represents a side-effect that happens in S4 or R6 object environments, and
-\emph{Local} that corresponds to a side effect in local variables of the \eval
-caller (offset 0 in Section~\ref{sec:env}). The table shows data for both \eval
-sites and functions. For a function to have a given target environment class,
-all of its \eval sites that do side effects must have the same target
-environment class The majority of the \eval sites (\SESitesInOneClass) do all
-side effects consistently in one environment class. The same happens at the
-function level. Almost half of the sites (over a third of the functions) do side
-effects in either \emph{Local} or \emph{Object} environments. This gives a ray
-of hope for the compiler. Even though it is possible to do anything anywhere,
-the data suggest that most side-effects are sane.
-
-Table~\ref{tab:se-types} shows the proportion of the different side effects
-that were recorded. In terms of calls, we see assignments primarily and in
-terms of site definitions. This is expected. A subsequent \eval call will turn
-a definition into an update. The most dangerous side effect is variable removal,
-as it means that after a call to \eval some binding in some environments will
-disappear. While rare, this happens, but the vast majority comes from a single
-site called by the already mentioned \c{withr::execute\_handlers} function.
-This makes sense as it is used to defer evaluation of an expression to after
-the function exit and thus used for clean up. In the corpus, it is used almost
-exclusively by the \c{tidyselect} package for removing the reference to the
-current quosure environment while interpreting a data frame column
-selectors.\footnote{\cf \url{https://tidyselect.r-lib.org/}}
+Table~\ref{tab:se-env} shows environment kinds where side-effects happen.
+Function environments distinguish between \emph{local}, the caller environment
+(offset 0), and \emph{function} (offset >0). \emph{Synthetic} represents
+constructed environments. \emph{Object} is used to denote the environments
+attached to objects and classes in the S4 and R6 object systems. \emph{Multiple}
+denotes cases where side-effects to more than one kind originate from one site.
+The table gives the number of call sites of \eval (and the ratio) performing
+side-effect in a particular environment kind. The table also gives the number of
+functions in which these sites occur (and ratio).
+
+Most sites, \SESitesInOneClass, do all their side-effects consistently in one
+kind of environments. The same happens at the function level. Almost half of the
+sites (over a third of the functions) do side effects in either \emph{Local} or
+\emph{Object} environments. This gives a ray of hope for a hypothetical R
+compiler. Even though \eval can do anything anywhere, the data suggests most
+effects are sane.
+
+Table~\ref{tab:se-types} shows recorded effects. There are over 5M updates. In
+terms of calls, we see assignments primarily and in terms of site definitions.
+This is expected. A subsequent \eval call will turn a definition into an update.
+The most dangerous side effect is variable removal, as it means that after a
+call to \eval some binding in some environments will disappear. While rare, this
+happens, but the vast majority comes from a single site called by the already
+mentioned \c{withr::execute\_handlers} function. This makes sense as it is used
+to defer evaluation of an expression to after the function exit and thus used
+for clean up. In the corpus, it is used almost exclusively by the \c{tidyselect}
+package for removing the reference to the current quosure environment while
+interpreting a data frame column selectors.\footnote{\cf
+\url{https://tidyselect.r-lib.org/}}
 
@@
 \mypara{Discussion} In JavaScript assignments in \eval can happen in either
-local scope or less often, when called through an alias, in the global scope.
-In R, given the support for first-class environments, it can happen anywhere,
+local scope or less often, when called through an alias, in the global scope. In
+R, given the support for first-class environments, it can happen anywhere,
 making it \eval more dangerous than it already is. However, the data suggest
 that first, side-effects from \eval are not as widespread as in the case of
-JavaScript,\footnote{\citep{ecoop11} shows that in the \emph{Interactive} scenario, \eval in Javascript performs, store events can reach up to 40\% of the events, and 7\% to 8\% of the \eval do side effects in the global scope.
-} and that over half of them happen in a predictable environment.
+JavaScript,\footnote{\citep{ecoop11} shows that in the \emph{Interactive}
+scenario, \eval in Javascript performs, store events can reach up to 40\% of the
+events, and 7\% to 8\% of the \eval do side effects in the global scope. } and
+that over half of them happen in a predictable environment.
 
@@
 evaluation, \c{substitute} and \c{eval} are the tools given to developers to
-this end.
-%This API is slightly more complex than just passing a string, it is
-%conceivable that this may discourage some casual users. \Eval is also being
-%used to reduce boilerplate code and provide convenience features for
-%programmers.
-In this section, we focus on the \emph{why}, giving real-world examples of the
-\eval usage in the wild. All the examples are based on code from our corpus,
-often simplified to fit the space and increase clarity. For each example, we
-indicate the package and the function where it is located.
-
-\subsection{Shapes of \eval}
-
-% We start the discussion with the examples for the minimized expressions from
-% Section~\ref{sec:minimized}.
+this end. In this section, we focus on the \emph{why}, giving real-world
+examples of the \eval usage in the wild. All the examples are based on code from
+our corpus, often simplified to fit the space and increase clarity. For each
+example, we indicate the package and the function where it is located.
+
+\subsection{Shapes redux}
+
 In Table~\ref{tab:minimizedexpressions} we have shown the most common shapes of
@@
     returns the expression passed to \c{sel}, the second is the data frame in
-    which this expression will be evaluated, and the last indicates
-    which environment should be used to look up bindings not found in \c{nl}.
+    which this expression will be evaluated, and the last indicates which
+    environment should be used to look up bindings not found in \c{nl}.
 
     Another example is the most often called \eval site in CRAN:
-    \lstinline|eval(`_inherit`, env)| defined in the \c{find\_super} function
-    in the \c{ggproto} object system. It looks up the variable \c{`\_inherit`} to
-    find the parent class. In the vast majority of cases, the variable is a symbol
-    and so the \eval seems superfluous. The \eval is however necessary so one
-    can specify parent classes using \c{pkg\_name::class\_name}, as the \c{::}
-    operator is a function call in R.
+    \lstinline|eval(`_inherit`, env)| defined in the \c{find\_super} function in
+    the \c{ggproto} object system. It looks up the variable \c{`\_inherit`} to
+    find the parent class. In the vast majority of cases, the variable is a
+    symbol and so the \eval seems superfluous. The \eval is however necessary so
+    one can specify parent classes using \c{pkg\_name::class\_name}, as the
+    \c{::} operator is a function call in R.
 
@@
     to create or update a binding in a specific environment where either the
-    name or the value come from some expression. Most cases originate from trivial
-    code generation where the assign expression is assembled using \c{parse} or
-    \c{substitute}, often in a loop: \lstinline|eval(parse(text=paste(capture.output(default_args[[i]]), "<-", i)))|.
+    name or the value come from some expression. Most cases originate from
+    trivial code generation where the assign expression is assembled using
+    \c{parse} or \c{substitute}, often in a loop:
+    \lstinline|eval(parse(text=paste(capture.output(default_args[[i]]), "<-",
+    i)))|.
 
@@
     function uses \eval for a deferred assignment to fill missing values in a
-    column. Package \c{overture} repeatedly evaluates R expressions capturing any  assignments that happen in it, as samples of Markov chains.
+    column. Package \c{overture} repeatedly evaluates R expressions capturing
+    any assignments that happen in it, as samples of Markov chains.
 
@@
   \item \emph{Function calls \c{(F(), F(X), F(F(X)))}} represent two very
-    frequent patterns: \emph{code generation} and \emph{call reflection}. There are numerous
-    cases for code generation. One occurring usage is to generate ad-hoc calls
-    to libraries such as \c{ggplot2} when one needs to parameterize data or
-    aesthetics selection from a variable. For example \lstinline|ggplot(df, aes(x=A, y=B))|
-    maps X and Y axes in the plot to \c{A} and \c{B} columns in
-    \c{df}. It might not be immediately clear\footnote{\cf
-    \url{https://stackoverflow.com/questions/22309285/how-to-use-a-variable-to-specify-column-name-in-ggplot}}  for a ggplot2 user
-    how to modify the call to pick the actual mapping from \c{A} and \c{B}
-    variables. Another case is the traditional use of \eval as a means to
-    reduce boilerplate code; simple and repetitive code can easily be replaced
-    with judicious use of \eval. For example, the \c{data.table} package uses
-    \eval to call the \c{options} function with named arguments taken from a
-    vector of strings. While the benefits are limited in this example, it is an
-    attractive tool for programmers.
+    frequent patterns: \emph{code generation} and \emph{call reflection}. There
+    are numerous cases for code generation. One occurring usage is to generate
+    ad-hoc calls to libraries such as \c{ggplot2} when one needs to parameterize
+    data or aesthetics selection from a variable. For example
+    \lstinline|ggplot(df, aes(x=A, y=B))| maps X and Y axes in the plot to \c{A}
+    and \c{B} columns in \c{df}. It might not be immediately clear\footnote{\cf
+    \url{https://stackoverflow.com/questions/22309285/how-to-use-a-variable-to-specify-column-name-in-ggplot}}
+    for a ggplot2 user how to modify the call to pick the actual mapping from
+    \c{A} and \c{B} variables. Another case is the traditional use of \eval as a
+    means to reduce boilerplate code; simple and repetitive code can easily be
+    replaced with judicious use of \eval. For example, the \c{data.table}
+    package uses \eval to call the \c{options} function with named arguments
+    taken from a vector of strings. While the benefits are limited in this
+    example, it is an attractive tool for programmers.
     %
@@
     function or with \c{model.frame} and related functions. For example the
-    \c{survival::coxph} function\footnote{Function fitting a Cox proportional hazards
-    regression model.} contains:
+    \c{survival::coxph} function\footnote{Function fitting a Cox proportional
+    hazards regression model.} contains:
     %
     \begin{lstlisting}
-    Call <- match.call()
-    tform <- Call[c(1,indx)]  # only keep the arguments we wanted
-    tform[[1L]] <- quote(stats::model.frame)  # change the function called
+    Call <- match.call() tform <- Call[c(1,indx)] # only keep the arguments we
+    wanted tform[[1L]] <- quote(stats::model.frame) # change the function called
     mf <- eval(tform, parent.frame())
@@
 
-\todo{Say somewhere that the R community commonly calls call/argument capturing (with substitute or match.call), Non-Standard Evaluation.}
-
-\subsection{Domain Specific Language.} R expressions are built and transformed symbolically and then evalled to a concrete value.
-For instance,   base R packages \emph{stats} provides two operators for \emph{symbolic differentiation}, \c{D} and \c{deriv}.
-They support arithmetic operations and functions on real numbers such as \c{sin}.
-In \emph{MCMCglmm} (Monte Carl Markov Chain Generalised Linear Mixed Models), a \c{Dtensor} object is created, which contains expressions derived twice (with \c{DD}) and then typically in a list or a data frame.
+\todo{Say somewhere that the R community commonly calls call/argument capturing
+  (with substitute or match.call), Non-Standard Evaluation.}
+
+\subsection{Domain Specific Language.} R expressions are built and transformed
+symbolically and then evalled to a concrete value. For instance, base R packages
+\emph{stats} provides two operators for \emph{symbolic differentiation}, \c{D}
+and \c{deriv}. They support arithmetic operations and functions on real numbers
+such as \c{sin}. In \emph{MCMCglmm} (Monte Carl Markov Chain Generalised Linear
+Mixed Models), a \c{Dtensor} object is created, which contains expressions
+derived twice (with \c{DD}) and then typically in a list or a data frame.
 \begin{lstlisting}
@@
 \end{lstlisting}
-% See https://github.com/cran/MCMCglmm/blob/cfac9e67ec73a3db2142826faae35e5d2318da31/man/evalDtensor.Rd
-
-Package \c{tidyselect} is the backend of several \c{dplyr} and \c{tidyr} functions to select columns.
-It offers a DSL that makes it possible to select a set of columns, over a range of columns, a complement, an intersection, an union, with a regular expression and so on.
-\eval is used to traverse selectors stored as slots of a dataframe.
-
-% \subsubsection{Capturing a call.}  A common use case for \eval is to be combined with \c{match.call}.
-% \c{match.call} walks up the call stack, captures the code that invoked the
-% currently executing function, and returns it as an unevaluated expression that includes the call function and its unevaluated arguments.
-The pattern is to transform a call to some function \c{f} into a call to \c{g} with some arguments retained and others modified.
-This pattern is recognizable
-% by the fact that the expression is a call and the target environment is that of
-% the parent.
-
-% % We have quantitative data on that if needed...
-% This pattern is very often (more than half of the cases) used in relation with statistical models, to build a new model, \c{model.frame} to build a new generic model, or a more specialized model like \c{glm} or, \c{lm}.
-The R documentation of \c{match.call} explicitly describes this usage.
-
-% In package \emph{survival}\footnote{It is a \emph{recommended package}, so recommended by the CRAN maintainers, thoroughly tested, and following R best practices.}, in function \c{coxph}, the surrounding call is first captured, then \c{model.frame} is injected as the function name in the call, and finally, the new call is evaluated in the parent frame, to ensure that the rewritten call
-% is evaluated in the same environment the original call was.
-
-% \begin{lstlisting}
-% Call <- match.call()
-% tform <- Call[c(1,indx)]  # only keep the arguments we wanted
-% tform[[1L]] <- quote(stats::model.frame)  # change the function called
-% mf <- eval(tform, parent.frame())
-% \end{lstlisting}
-
-% A related pattern is not to change the call but rather to evaluate the arguments in the parent frame. This could also be achieved with \c{substitute}, but \c{match.call} does it in bulk.
-% In package \emph{MedDietCalc},  in function \c{computeMDS95}, \c{match.call} retrieves the argument expressions of the call, which are evaluated in a dataframe \c{data} which contains the numeric data.
-Each argument of function \c{computeMDS95} is also a column in dataframe \c{data}.
-The advantage of using \c{eval} here is to be able to express an argument as a combination of several columns in the dataframe.
-% \begin{lstlisting}
-% computeMDS95 <- function (data, Vegetables, ...) {
-% 	arguments <- as.list( match.call())
-% 	Vegetables <- eval(arguments$Vegetables, data)
-% \end{lstlisting}
-
-% \c{substitute} can be used to capture only a part of the call, \ie one of the arguments.
-For instance, in package \emph{xyplot}, the expression for argument \c{subset} is captured with \c{substitute} and then evaluated in a dataframe \c{data} and in the definition environment of \c{x}.
-% \begin{lstlisting}
-% xyplot.formula <-  function(x,    data = NULL, ..., subset = TRUE)
-% subset <- eval(substitute(subset), data, environment(x))
-% \end{lstlisting}
-
-% In the R community, the call capturing pattern with evaluation in a non-local environment is often referred~\cite{hadley} as \emph{Non-Standard Evaluation}.
-This pattern in relation with \c{model.frame} is also described explicitly in the R documentation of \c{match.call}.
-
-% \subsubsection{Injecting code in package environments.} R distinguishes two kinds of environments for packages: the \emph{package} environment itself, which contains every publicly accessible functions, and the \emph{namespace} environment, which contains all functions including private ones.
-
-%  A call to \c{library(packageName)}, which loads libraries in R, for instance, will
-% populate environment \c{package:packageName} using \eval. It is a \emph{base} function
-% though, so we do not see it in the CRAN dataset. However, R still makes it
-% possible to access any package environments, which we see in
-% \packageNbPackageEnvPackages different packages in our corpus. Those packages perform
-% introspection, to represent and search any environment by name, in package
-% \emph{envnames}, to containerize tests, as with \emph{testthat} and
-% \emph{unitizer}, to create a custom OOP system with \emph{R.oo}, or to bypass the visibility of functions in \emph{USSCXenaTools}. It is also used to
-% implement a custom caching mechanism in packages \emph{g.data} and \emph{SOAR}.
-
-% \subsubsection{Code Generation.} The more traditional use of \eval is to execute code
-% that was assembled by the programmer into a string. Here we show the method
-% \c{plot} for class \c{gmap} in package \c{gpmaps}, simplified for
-% explanatory purposes.   \c{paste} is used to dispatch to a specialized plot function, among \c{plot2\_orig},  \c{plot3\_orig}, \c{plot3\_dec} and so on.
-% \begin{lstlisting}
-% eval(parse(text=paste('plot',gpmap$nloci,'_orig(gpmap)',sep="")))
-% \end{lstlisting}
-
-% In Kaggle, plots are often built with \eval, \c{paste} and \c{ggplot}.
-In this kernel for the \emph{digit-recognizer} challenge, \c{paste0} is used to generate \c{ggplot} commands featuring unique y labels to build 9 different graphs.
-
-% \begin{lstlisting}
-% p <- lapply(paste0('PC', 1:9), function(pc){
-% 	eval(parse(text=paste0('ggplot(ld, aes(x=label, y=',pc,')) + geom_boxplot()')))
-% })
-% \end{lstlisting}
-
-% The \parse and \eval combination is
-% used by the \source function in the base library to load R code from a file
-% in the current workspace.
-
-%  Code generation is often used as a means to reduce boilerplate code;
-% simple and repetitive code can easily be replaced with judicious use of \eval.
-% For example, the \c{data.table} package uses \eval to calls the \c{options}
-% function with named arguments taken from a vector of strings. While the benefits
-% are limited in this example, it is an attractive tool for programmers.
-% \begin{lstlisting}
-%   opts = c("datatable.verbose"="FALSE", # ...many others
-%   for (i in names(opts))
-%     eval(parse(text=paste0("options(",i,"=",opts[i],")")))
-% \end{lstlisting}
-% This pattern is a special case of code generation, recognizable by the
-% fact that \eval is executed in a loop.
-
-% Using \c{parse}, \ie converting strings to R expressions, opens to a more flexible code generation, as the building blocks which are going to be concatenated together, do not need to be valid R expressions.
-On the contrary, combining expression created with \c{substitute} or \c{quote} requires more discipline.
-In Kaggle, which features less experienced programmers,  code generation for plots is widespread, and uses \c{parse} on strings with \eval.
-
-% \todo{There is an example in package comprenhr that does code generation with substitute to create list comprehensions, but it is quite difficult to grasp.
-But at the same type, we lack here an example of code generation using quite, enquote or substitute.}
-
-% \todo{Add a pattern about Library Loading? \c{library(packageName)} in R uses \eval This is one of the patterns in Javascript...}
+
+Package \c{tidyselect} is the backend of several \c{dplyr} and \c{tidyr}
+functions to select columns. It offers a DSL that makes it possible to select a
+set of columns, over a range of columns, a complement, an intersection, an
+union, with a regular expression and so on. \eval is used to traverse selectors
+stored as slots of a dataframe.
 
@@
 features more experienced developers than websites found at large on the web.
-\todo{can we really claim that? SHould we talk about Kaggle there?}
 
@@
 Building a call and executing can be done with \c{do.call}. Nevertheless, \eval
-shines when expressions passed to it can be of many different types. % And it
-is shorter to type, and no need to remember all the specialized functions, so
-it is convenient.
+shines when expressions passed to it can be of many different types.
 
@@
 modify anything anywhere, which is wary for both R developers and R compiler.
-The shapes of expressions passed to \eval are extremely diverse, ranging from simple variable lookups to complex blocks or the creation of new functions.
-Luckily, most sites only see a single shape.
-\eval unlashes its power and shows up its singularity compared to \eval in other languages with the diversity of environments in which it can evaluate its expressions.
-While most of the time, the \eval environment is the local environment, it is also often the global environment, a synthetic environment, or even a package environment.
-Luckily, as for the expressions passed to \eval, the environment passed to \eval for one given site is easily predictable with the first call to the site.
-
-The expressive power of \eval shines even more  so in the case of side effects that can quite literally happen anywhere.
-The data, however, suggest that this is not the case.
-The majority of eval sites do not do any observable side effects,  and when they do, it happens in a predictable environment.
-
-The origins of the expression passed to \eval, either constructed from language expressions, or stemming from reflection, or built by parsing strings.
-Reflection and constructed expressions dominate, which suggests the primary   use of \eval is meta-programming.
-% macros...
-
-% how much can we say about the threat model?
+The shapes of expressions passed to \eval are extremely diverse, ranging from
+simple variable lookups to complex blocks or the creation of new functions.
+Luckily, most sites only see a single shape. \eval unlashes its power and shows
+up its singularity compared to \eval in other languages with the diversity of
+environments in which it can evaluate its expressions. While most of the time,
+the \eval environment is the local environment, it is also often the global
+environment, a synthetic environment, or even a package environment. Luckily, as
+for the expressions passed to \eval, the environment passed to \eval for one
+given site is easily predictable with the first call to the site.
+
+The expressive power of \eval shines even more so in the case of side effects
+that can quite literally happen anywhere. The data, however, suggest that this
+is not the case. The majority of eval sites do not do any observable side
+effects, and when they do, it happens in a predictable environment.
+
+The origins of the expression passed to \eval, either constructed from language
+expressions, or stemming from reflection, or built by parsing strings.
+Reflection and constructed expressions dominate, which suggests the primary use
+of \eval is meta-programming.
+
 Similarly to \citep{ecoop11}, we started looking into \evals in R with the
@@
 sites fitted an extremely simple categorization based on simple regular
-expressions. Not so in R. While some usage, especially the one which calls
-\eval with an expression parsed from string, can be categorized and often
-replaced by equivalent yet more disciplined and safer code, the case of \eval
-coming from reflection is much harder.
-
-Even though we reckon that \eval has more expressive power than macros, as it can access runtime information such as its call stack, adding a macro system to R to replace \eval for all meta-programming purposes would help static analysis as macros can be expanded at analysis-time.
+expressions. Not so in R. While some usage, especially the one which calls \eval
+with an expression parsed from string, can be categorized and often replaced by
+equivalent yet more disciplined and safer code, the case of \eval coming from
+reflection is much harder.
+
+Even though we reckon that \eval has more expressive power than macros, as it
+can access runtime information such as its call stack, adding a macro system to
+R to replace \eval for all meta-programming purposes would help static analysis
+as macros can be expanded at analysis-time.
 

### 2021-04-17 paper/main.tex
Update in 6

--- 2021-04-17 paper/main.tex
@@
 Finally, let us mention an anecdotal use of \eval as a mean to bypass some
-checks performed by the \c{R CMD CHECK} tool allowing one to submit a package
-to CRAN that would otherwise be refused. For example, the tool checks that
-packages do not call \c{unlockBinding} function.\footnote{It allows one to open
-and modify loaded package name space.} Packages such as \c{data.table}
-circumvent such restrictions by using \eval as the check is only performed
-statically.
+checks performed by the \c{R CMD CHECK} tool. This tool performs a numer of
+well-formedness rules before accepting a package to the CRAN repository. Many
+checks are performed statically, by looking at syntactic constructs in the R
+code. Using \eval allows one to bypass such checks and thus submit a package to
+CRAN that would otherwise be refused. For example, the it checks that packages
+do not call \c{unlockBinding} function.\footnote{It allows one to open and
+modify loaded package name space.} Packages such as \c{data.table} circumvent
+such restrictions by using \eval as the check is only performed statically.
 

### 2021-04-17 paper/main.tex

--- 2021-04-17 paper/main.tex
@@
     trivial code generation where the assign expression is assembled using
-    \c{parse} or \c{substitute}, often in a loop:
-    \lstinline|eval(parse(text=paste(capture.output(default_args[[i]]), "<-",
-    i)))|.
+    \c{parse} or \c{substitute}, often in a loop. For example, the
+    \c{PopED::get\_rse} uses it to assign a number of variables in the current
+    scope: \lstinline|eval(parse(text=paste(default_args[[i]]), "<-", i))|.
 
@@
     frequent patterns: \emph{code generation} and \emph{call reflection}. There
-    are numerous cases for code generation. One occurring usage is to generate
-    ad-hoc calls to libraries such as \c{ggplot2} when one needs to parameterize
-    data or aesthetics selection from a variable. For example
-    \lstinline|ggplot(df, aes(x=A, y=B))| maps X and Y axes in the plot to \c{A}
-    and \c{B} columns in \c{df}. It might not be immediately clear\footnote{\cf
+    are numerous cases for code generation.
+    %
+    One occurring usage (present also in the Kaggle scripts) is to generate
+    ad-hoc calls to libraries such as \c{ggplot2} when one needs to
+    parameterize data or aesthetics selection from a variable. For example
+    \lstinline|ggplot(df, aes(x=A, y=B))| maps X and Y axes in the plot to
+    \c{A} and \c{B} columns in \c{df}. It might not be immediately
+    clear\footnote{\cf
     \url{https://stackoverflow.com/questions/22309285/how-to-use-a-variable-to-specify-column-name-in-ggplot}}
-    for a ggplot2 user how to modify the call to pick the actual mapping from
-    \c{A} and \c{B} variables. Another case is the traditional use of \eval as a
-    means to reduce boilerplate code; simple and repetitive code can easily be
-    replaced with judicious use of \eval. For example, the \c{data.table}
-    package uses \eval to call the \c{options} function with named arguments
-    taken from a vector of strings. While the benefits are limited in this
-    example, it is an attractive tool for programmers.
-    %
-    \begin{lstlisting}
-    opts = c("datatable.verbose"="FALSE", # ...many others
+    for a ggplot2 user how to modify the call to assign the axis columns from
+    variables. Another case is the traditional use of \eval as a mean to reduce
+    boilerplate code; simple and repetitive code can easily be replaced with
+    judicious use of \eval. For example, the \c{data.table} package uses \eval
+    to call the \c{options} function with named arguments taken from a vector
+    of strings:
+
+    \begin{lstlisting}[gobble=2]
+    opts = c("datatable.verbose"="FALSE", ... )
     for (i in names(opts)) eval(parse(text=paste0("options(",i,"=",opts[i],")")))
@@
     %
-    \begin{lstlisting}
-    Call <- match.call() tform <- Call[c(1,indx)] # only keep the arguments we
-    wanted tform[[1L]] <- quote(stats::model.frame) # change the function called
+    \begin{lstlisting}[gobble=2]
+    Call <- match.call()
+    tform <- Call[c(1,indx)]  # only keep the arguments we wanted
+    tform[[1L]] <- quote(stats::model.frame)  # change the function called
     mf <- eval(tform, parent.frame())
     \end{lstlisting}
-
-    \todo{Add that this pattern is advised by the R documentation of match.call}
 
@@
 
-\subsection{Domain Specific Language.} R expressions are built and transformed
-symbolically and then evalled to a concrete value. For instance, base R packages
-\emph{stats} provides two operators for \emph{symbolic differentiation}, \c{D}
-and \c{deriv}. They support arithmetic operations and functions on real numbers
-such as \c{sin}. In \emph{MCMCglmm} (Monte Carl Markov Chain Generalised Linear
-Mixed Models), a \c{Dtensor} object is created, which contains expressions
-derived twice (with \c{DD}) and then typically in a list or a data frame.
+\subsection{Domain Specific Language}
+
+The ability to capture current call as a language object (\c{match.call()}),
+extract the unevaluated expression from function arguments (\c{substitute()})
+and \eval with first class environment makes the R toolkit for building DSLs.
+There are numerous widely spread DSLs used in R. For example, \c{data.table}
+package defines a DSL for data frame manipulation using the subset operator
+\c{[}. This allowing one to write compact queries such as \c{flights[carrier ==
+  "AA", .N, by = origin]}\footnote{The \c{data.table} subset operation is
+defined as \c{DT[i, j, k]}, where \c{j} expression is used to calculate on a
+subset/reorder of rows from \c{i} grouped by \c{k}.} which will compute the
+number of flights done by American Airlines for each origin airport.
+Essentially, it uses \c{substitute} to capture the operand's expressions,
+transforms then appropriately and finally uses \eval to execute the query.
+
+Another example is the \c{tidyselect} package that defines a DSL for selecting
+columns in a data frame powering number of data manipulation operation in the
+\c{data.table} main competitor, the \c{dply} package. It uses eval to build an
+interpreter for evaluating column selectors such as \c{select(df,
+starts\_with("length") \& !(name:mass))} which returns all columns in \c{df}
+that starts with the string \c{length} and is not between \c{name} and
+\c{mass} columns.
+
+For instance, base R packages \emph{stats} provides two operators for
+\emph{symbolic differentiation}, \c{D} and \c{deriv}. They support arithmetic
+operations and functions on real numbers such as \c{sin}. In \emph{MCMCglmm}
+(Monte Carl Markov Chain Generalised Linear Mixed Models), a \c{Dtensor} object
+is created, which contains expressions derived twice (with \c{DD}) and then
+typically in a list or a data frame.
 \begin{lstlisting}
@@
 \end{lstlisting}
-
-Package \c{tidyselect} is the backend of several \c{dplyr} and \c{tidyr}
-functions to select columns. It offers a DSL that makes it possible to select a
-set of columns, over a range of columns, a complement, an intersection, an
-union, with a regular expression and so on. \eval is used to traverse selectors
-stored as slots of a dataframe.
 

### 2021-04-17 paper/main.tex
Symbolic differentiation

--- 2021-04-17 paper/main.tex
@@
 package defines a DSL for data frame manipulation using the subset operator
-\c{[}. This allowing one to write compact queries such as \c{flights[carrier ==
+\c{[}. This allows one to write compact queries such as \c{flights[carrier ==
   "AA", .N, by = origin]}\footnote{The \c{data.table} subset operation is
@@
 
-% TODO Pierre
-For instance, base R packages \emph{stats} provides two operators for \emph{symbolic differentiation}, \c{D} and \c{deriv}.
-They support arithmetic operations and functions on real numbers such as \c{sin}.
-In \emph{MCMCglmm} (Monte Carl Markov Chain Generalised Linear Mixed Models), a \c{Dtensor} object is created, which contains expressions derived twice (with \c{DD}) and then typically in a list or a data frame.
+Another instance, is the \emph{symbolic differentiation support} provided by base R package \emph{stats}, with two operators \c{D} and \c{deriv}.
+They support arithmetic operations and functions on real numbers such as \c{sin}, on expressions built with \c{expression()} or \c{parse}.
+In \emph{MCMCglmm} (Monte Carl Markov Chain Generalised Linear Mixed Models), a \c{Dtensor} object is created: it contains expressions that are derived twice (with \c{DD}).
+To get a concrete value, it is evaluated in  dataframe \c{mu} where symbolic variables are bound to actual numbers.
 \begin{lstlisting}
@@
 \end{lstlisting}
+
 % See https://github.com/cran/MCMCglmm/blob/cfac9e67ec73a3db2142826faae35e5d2318da31/man/evalDtensor.Rd
@@
 Finally, let us mention an anecdotal use of \eval as a mean to bypass some
-checks performed by the \c{R CMD CHECK} tool. This tool performs a numer of
-well-formedness rules before accepting a package to the CRAN repository. Many
-checks are performed statically, by looking at syntactic constructs in the R
-code. Using \eval allows one to bypass such checks and thus submit a package to
-CRAN that would otherwise be refused. For example, the it checks that packages
-do not call \c{unlockBinding} function.\footnote{It allows one to open and
-modify loaded package name space.} Packages such as \c{data.table} circumvent
-such restrictions by using \eval as the check is only performed statically.
+checks performed by the \c{R CMD CHECK} tool allowing one to submit a package
+to CRAN that would otherwise be refused. For example, the tool checks that
+packages do not call \c{unlockBinding} function.\footnote{It allows one to open
+and modify loaded package name space.} Packages such as \c{data.table}
+circumvent such restrictions by using \eval as the check is only performed
+statically.
 

### 2021-04-17 paper/main.tex
Merge

@@
     function or with \c{model.frame} and related functions. For example the
-    \c{survival::coxph} function\footnote{Function fitting a Cox proportional hazards
-    regression model.} contains:
+    \c{survival::coxph} function\footnote{Function fitting a Cox proportional
+    hazards regression model.} contains:
     %
@@
 
-\todo{Say somewhere that the R community commonly calls call/argument capturing (with substitute or match.call), Non-Standard Evaluation.}
+\todo{Say somewhere that the R community commonly calls call/argument capturing
+  (with substitute or match.call), Non-Standard Evaluation.}
 
@@
 
-Another instance, is the \emph{symbolic differentiation support} provided by base R package \emph{stats}, with two operators \c{D} and \c{deriv}.
+Another instance is the \emph{symbolic differentiation support} provided by base R package \emph{stats}, with two operators \c{D} and \c{deriv}.
 They support arithmetic operations and functions on real numbers such as \c{sin}, on expressions built with \c{expression()} or \c{parse}.
@@
 Building a call and executing can be done with \c{do.call}. Nevertheless, \eval
-shines when expressions passed to it can be of many different types. % And it
-is shorter to type, and no need to remember all the specialized functions, so
-it is convenient.
+shines when expressions passed to it can be of many different types.
 
 Finally, let us mention an anecdotal use of \eval as a mean to bypass some
-checks performed by the \c{R CMD CHECK} tool allowing one to submit a package
-to CRAN that would otherwise be refused. For example, the tool checks that
-packages do not call \c{unlockBinding} function.\footnote{It allows one to open
-and modify loaded package name space.} Packages such as \c{data.table}
-circumvent such restrictions by using \eval as the check is only performed
-statically.
+checks performed by the \c{R CMD CHECK} tool. This tool performs a numer of
+well-formedness rules before accepting a package to the CRAN repository. Many
+checks are performed statically, by looking at syntactic constructs in the R
+code. Using \eval allows one to bypass such checks and thus submit a package to
+CRAN that would otherwise be refused. For example, the it checks that packages
+do not call \c{unlockBinding} function.\footnote{It allows one to open and
+modify loaded package name space.} Packages such as \c{data.table} circumvent
+such restrictions by using \eval as the check is only performed statically.
 

### 2021-04-17 paper/main.tex
more

--- 2021-04-17 paper/main.tex
@@
 call to \eval some binding in some environments will disappear. While rare, this
-happens, but the vast majority comes from a single site called by the already
-mentioned \c{withr::execute\_handlers} function. This makes sense as it is used
-to defer evaluation of an expression to after the function exit and thus used
-for clean up. In the corpus, it is used almost exclusively by the \c{tidyselect}
-package for removing the reference to the current quosure environment while
-interpreting a data frame column selectors.\footnote{\cf
-\url{https://tidyselect.r-lib.org/}}
+happens, but the vast majority comes from a single site
+(\c{withr::execute\_handlers}). This makes sense as it is used to defer
+evaluation of an expression to after the function exit and thus used for clean
+up. It is used almost exclusively by the \c{tidyselect} package for removing the
+reference to the current quosure environment while interpreting a data frame
+column selectors.\footnote{\cf \url{https://tidyselect.r-lib.org/}}
 
@@
 
-\todo{Say somewhere that the R community commonly calls call/argument capturing
-  (with substitute or match.call), Non-Standard Evaluation.}
+%\todo{Say somewhere that the R community commonly calls call/argument capturing
+%(with substitute or match.call), Non-Standard Evaluation.}
 

### 2021-04-17 paper/main.tex
Update 6.4

@@
 function or method call, or assignment to a local variable or an object
-property. Compared to Javascript, \eval patterns in R we observed are much more
-legitimate, although this is biased by the fact that our main corpus CRAN
-features more experienced developers than websites found at large on the web.
-
-\eval is a powerful tool and can be sometimes replaced with functions with more
-specialized goals. For instance, variable lookup in any environment can be
-performed with \c{get}, and assignment in any environment, with \c{assign}.
-Building a call and executing can be done with \c{do.call}. Nevertheless, \eval
-shines when expressions passed to it can be of many different types.
+property. For R, the data suggests that in some cases, the use of \eval in the
+\c{X, V, \$, <-, FUN} and the function calls related to code generation could
+replaced by more specific functions. For instance, variable lookup in any
+environment can be performed with \c{get}, and assignment in any environment,
+with \c{assign}. Building a call and executing can be done with \c{do.call}.
+Nevertheless the many of the observed \eval usage in R, in particular the one
+related to statistical modeling and DSLs are more sophisticated than those
+reported for JavaScript.
 

### 2021-04-17 paper/main.tex
intro

--- 2021-04-17 paper/main.tex
@@
 
-\vspace{2mm}\noindent\emph{Why do we Eval?} {\bf A short summary of the
-results should go here.}
+\vspace{2mm}\noindent\emph{Why do we Eval?} The results of our study suggest
+that \eval is widely used for the implementation of the language, and in many
+libraries. End-user code makes less frequent and less sophisticated use of
+\eval. In many ways, \eval in R is as bad as it gets: it's varied, performs
+side-effects and reaches to many environments. By large, the motivations for
+\eval relate to various forms of language extensions and meta-programming. \Eval
+is used where other languages would provide macros. But, the expressive power of
+\eval is higher as it can reach arbitrarily far back in the call stack.
 

### 2021-04-17 paper/main.tex
Add note to 6.3

@@
 \end{lstlisting}
-or even to a one-liner \c{do.call(paste0("q",d), as.list(dp, ...))}.
+or even to a one-liner \c{do.call(paste0("q",d), c(dp, list(...)))}.
+
+The problem is that these cases are not easy to spot. Because of the
+limits of dynamic analysis we only have a partial coverage and cannot determine
+if for example \eval sites classified as \c{X, V} will not be eventually called
+with an expression in a \c{F(), F(X)} or \c{F(F(X))} category.
 

### 2021-04-17 paper/main.tex
Proofreading Section 6

--- 2021-04-17 paper/main.tex
@@
 In Table~\ref{tab:minimizedexpressions} we have shown the most common shapes of
-expressions passed to \eval. Here, we illustrate them on concrete instances.
+expressions passed to \eval. Here, we illustrate them in concrete instances.
 
@@
     find the parent class. In the vast majority of cases, the variable is a
-    symbol and so the \eval seems superfluous. The \eval is however necessary so
+    symbol, so the \eval seems superfluous. The \eval is however necessary so
     one can specify parent classes using \c{pkg\_name::class\_name}, as the
@@
     column. Package \c{overture} repeatedly evaluates R expressions capturing
-    any assignments that happen in it, as samples of Markov chains.
+    any assignments that happen in it as samples of Markov chains.
 
@@
     expressions. This is used mostly in the following circumstances: (1) to
-    record a call for a later use in a possibly different environment, (2) pass
-    most of the captured call arguments to another function or (3) evaluate
+    record a call for later use in a possibly different environment, (2) pass
+    most of the captured call arguments to another function, or (3) evaluate
     arguments of the function in different environments. This is widely used in
-    relation with statistical modeling in combination with a model fitting
+    relation to statistical modeling in combination with a model fitting
     function or with \c{model.frame} and related functions. For example the
@@
     part of in as in \lstinline|eval({...})|. Essentially, the block denotes a
-    fragment of a program to be evaluated in a special way and in a special
+    fragment of a program to be evaluated in a particular way and in a particular
     environment. The latter is what makes it different from 0-argument closure.
@@
 \end{lstlisting}
-
-% See https://github.com/cran/MCMCglmm/blob/cfac9e67ec73a3db2142826faae35e5d2318da31/man/evalDtensor.Rd
-
-% \subsubsection{Capturing a call.}  A common use case for \eval is to be combined with \c{match.call}.
-% \c{match.call} walks up the call stack, captures the code that invoked the
-% currently executing function, and returns it as an unevaluated expression that includes the call function and its unevaluated arguments.
-The pattern is to transform a call to some function \c{f} into a call to \c{g} with some arguments retained and others modified.
-This pattern is recognizable
-% by the fact that the expression is a call and the target environment is that of
-% the parent.
-
-% % We have quantitative data on that if needed...
-% This pattern is very often (more than half of the cases) used in relation with statistical models, to build a new model, \c{model.frame} to build a new generic model, or a more specialized model like \c{glm} or, \c{lm}.
-The R documentation of \c{match.call} explicitly describes this usage.
-
-% In package \emph{survival}\footnote{It is a \emph{recommended package}, so recommended by the CRAN maintainers, thoroughly tested, and following R best practices.}, in function \c{coxph}, the surrounding call is first captured, then \c{model.frame} is injected as the function name in the call, and finally, the new call is evaluated in the parent frame, to ensure that the rewritten call
-% is evaluated in the same environment the original call was.
-
-% \begin{lstlisting}
-% Call <- match.call()
-% tform <- Call[c(1,indx)]  # only keep the arguments we wanted
-% tform[[1L]] <- quote(stats::model.frame)  # change the function called
-% mf <- eval(tform, parent.frame())
-% \end{lstlisting}
-
-% A related pattern is not to change the call but rather to evaluate the arguments in the parent frame. This could also be achieved with \c{substitute}, but \c{match.call} does it in bulk.
-% In package \emph{MedDietCalc},  in function \c{computeMDS95}, \c{match.call} retrieves the argument expressions of the call, which are evaluated in a dataframe \c{data} which contains the numeric data.
-Each argument of function \c{computeMDS95} is also a column in dataframe \c{data}.
-The advantage of using \c{eval} here is to be able to express an argument as a combination of several columns in the dataframe.
-% \begin{lstlisting}
-% computeMDS95 <- function (data, Vegetables, ...) {
-% 	arguments <- as.list( match.call())
-% 	Vegetables <- eval(arguments$Vegetables, data)
-% \end{lstlisting}
-
-% \c{substitute} can be used to capture only a part of the call, \ie one of the arguments.
-For instance, in package \emph{xyplot}, the expression for argument \c{subset} is captured with \c{substitute} and then evaluated in a dataframe \c{data} and in the definition environment of \c{x}.
-% \begin{lstlisting}
-% xyplot.formula <-  function(x,    data = NULL, ..., subset = TRUE)
-% subset <- eval(substitute(subset), data, environment(x))
-% \end{lstlisting}
-
-% In the R community, the call capturing pattern with evaluation in a non-local environment is often referred~\cite{hadley} as \emph{Non-Standard Evaluation}.
-This pattern in relation with \c{model.frame} is also described explicitly in the R documentation of \c{match.call}.
-
-% \subsubsection{Injecting code in package environments.} R distinguishes two kinds of environments for packages: the \emph{package} environment itself, which contains every publicly accessible functions, and the \emph{namespace} environment, which contains all functions including private ones.
-
-%  A call to \c{library(packageName)}, which loads libraries in R, for instance, will
-% populate environment \c{package:packageName} using \eval. It is a \emph{base} function
-% though, so we do not see it in the CRAN dataset. However, R still makes it
-% possible to access any package environments, which we see in
-% \packageNbPackageEnvPackages different packages in our corpus. Those packages perform
-% introspection, to represent and search any environment by name, in package
-% \emph{envnames}, to containerize tests, as with \emph{testthat} and
-% \emph{unitizer}, to create a custom OOP system with \emph{R.oo}, or to bypass the visibility of functions in \emph{USSCXenaTools}. It is also used to
-% implement a custom caching mechanism in packages \emph{g.data} and \emph{SOAR}.
-
-% \subsubsection{Code Generation.} The more traditional use of \eval is to execute code
-% that was assembled by the programmer into a string. Here we show the method
-% \c{plot} for class \c{gmap} in package \c{gpmaps}, simplified for
-% explanatory purposes.   \c{paste} is used to dispatch to a specialized plot function, among \c{plot2\_orig},  \c{plot3\_orig}, \c{plot3\_dec} and so on.
-% \begin{lstlisting}
-% eval(parse(text=paste('plot',gpmap$nloci,'_orig(gpmap)',sep="")))
-% \end{lstlisting}
-
-% In Kaggle, plots are often built with \eval, \c{paste} and \c{ggplot}.
-In this kernel for the \emph{digit-recognizer} challenge, \c{paste0} is used to generate \c{ggplot} commands featuring unique y labels to build 9 different graphs.
-
-% \begin{lstlisting}
-% p <- lapply(paste0('PC', 1:9), function(pc){
-% 	eval(parse(text=paste0('ggplot(ld, aes(x=label, y=',pc,')) + geom_boxplot()')))
-% })
-% \end{lstlisting}
-
-% The \parse and \eval combination is
-% used by the \source function in the base library to load R code from a file
-% in the current workspace.
-
-%  Code generation is often used as a means to reduce boilerplate code;
-% simple and repetitive code can easily be replaced with judicious use of \eval.
-% For example, the \c{data.table} package uses \eval to calls the \c{options}
-% function with named arguments taken from a vector of strings. While the benefits
-% are limited in this example, it is an attractive tool for programmers.
-% \begin{lstlisting}
-%   opts = c("datatable.verbose"="FALSE", # ...many others
-%   for (i in names(opts))
-%     eval(parse(text=paste0("options(",i,"=",opts[i],")")))
-% \end{lstlisting}
-% This pattern is a special case of code generation, recognizable by the
-% fact that \eval is executed in a loop.
-
-% Using \c{parse}, \ie converting strings to R expressions, opens to a more flexible code generation, as the building blocks which are going to be concatenated together, do not need to be valid R expressions.
-On the contrary, combining expression created with \c{substitute} or \c{quote} requires more discipline.
-In Kaggle, which features less experienced programmers,  code generation for plots is widespread, and uses \c{parse} on strings with \eval.
-
-% \todo{There is an example in package comprenhr that does code generation with substitute to create list comprehensions, but it is quite difficult to grasp.
-But at the same type, we lack here an example of code generation using quite, enquote or substitute.}
-
-% \todo{Add a pattern about Library Loading? \c{library(packageName)} in R uses \eval This is one of the patterns in Javascript...}
 
@@
 \end{lstlisting}
-or even to a one-liner \c{do.call(paste0("q",d), as.list(dp, ...))}.
+or even to a one-liner \c{do.call(paste0("q",d), c(dp, list(...)))}.
+
+The problem is that these cases are not easy to spot. Because of the
+limits of dynamic analysis, we only have a partial coverage and cannot determine
+if for example, \eval sites classified as \c{X, V} will not be eventually called
+with an expression in a \c{F(), F(X)} or \c{F(F(X))} category.
 
@@
 
-\Eval in R is mostly used for meta-programming purposes and to access
+\Eval in R is used chiefly for meta-programming purposes and to access
 environments other than the local one. In Javascript, patterns are ad-hoc, such
@@
 with \c{assign}. Building a call and executing can be done with \c{do.call}.
-Nevertheless the many of the observed \eval usage in R, in particular the one
+Nevertheless, many of the observed \eval usage in R, particularly, the ones
 related to statistical modeling and DSLs are more sophisticated than those
@@
 
-Finally, let us mention an anecdotal use of \eval as a mean to bypass some
-checks performed by the \c{R CMD CHECK} tool. This tool performs a numer of
+Finally, let us mention an anecdotal use of \eval as a means to bypass some
+checks performed by the \c{R CMD CHECK} tool. This tool performs a number of
 well-formedness rules before accepting a package to the CRAN repository. Many
-checks are performed statically, by looking at syntactic constructs in the R
+checks are performed statically by looking at syntactic constructs in the R
 code. Using \eval allows one to bypass such checks and thus submit a package to
-CRAN that would otherwise be refused. For example, the it checks that packages
+CRAN that would otherwise be refused. For example,  it checks that packages
 do not call \c{unlockBinding} function.\footnote{It allows one to open and
-modify loaded package name space.} Packages such as \c{data.table} circumvent
+modify loaded package namespaces.} Packages such as \c{data.table} circumvent
 such restrictions by using \eval as the check is only performed statically.

### 2021-04-17 paper/main.tex
ck's proofreading of Section 6 and conclusion

@@
 
-The ability to capture current call as a language object (\c{match.call()}),
+The ability to capture the current call as a language object (\c{match.call()}),
 extract the unevaluated expression from function arguments (\c{substitute()})
@@
 environments other than the local one. In Javascript, patterns are ad-hoc, such
-as json loading and parsing, and can often be rewritten without \eval, such as
-function or method call, or assignment to a local variable or an object
+as json loading and parsing, and can often be rewritten without \eval, such as a 
+function or method call, or an assignment to a local variable or an object
 property. For R, the data suggests that in some cases, the use of \eval in the
@@
 CRAN that would otherwise be refused. For example,  it checks that packages
-do not call \c{unlockBinding} function.\footnote{It allows one to open and
+do not call the \c{unlockBinding} function.\footnote{It allows one to open and
 modify loaded package namespaces.} Packages such as \c{data.table} circumvent
@@
 written by R practitioners. We used an instrumented version of the R virtual
-machine, \rdyntrace, to capture all \eval calls traces, including the relevant
+machine, \rdyntrace, to capture all \eval calls, including the relevant
 side effects that happen within these calls. Several insights can be learned

### 2021-06-27 paper/main.tex
Low-level writing issues

@@
 write down the definition of LISP; he was surprised that someone coded
-it up and offered it to end-users. Since then, reflective facilities
+it up and offered it to end users. Since then, reflective facilities
 have been used to parameterize programs over code patterns that can be
@@
 developers with some programming experience and a working knowledge of R, and
-end-users, who are typically not expert programmers and often have only a
+end users, who are typically not expert programmers and often have only a
 cursory knowledge of the language.\footnote{Consider that R is lazy like
-Haskell. We informally surveyed end-users and did not find a single user aware
+Haskell. We informally surveyed end users and did not find a single user aware
 of this fact. Library developers, on the other hand, know and program
@@
  with their running time, data size, and  number of elements manipulated.
-  Timings are from a cluster of three 2.3GHz Intel Xeon 6140 servers, with 72 cores and 256GB of RAM, and shared OCFS network storage. The pipeline steps are:
+  Timings are from a cluster of three servers, each with 2.3GHz Intel Xeon 6140 processor with 72 cores and 256GB of RAM, and shared OCFS network storage. The pipeline steps are:
 
@@
 This section focuses on CRAN packages and reports statistics about the usage of
-\eval. We use the word \emph{site} to an occurrence of a call site to the \eval
+\eval. We use the word \emph{site} to refer to an occurrence of a call site to the \eval
 function in the source code and \emph{call} to denote an observed invocation of
@@
 straightforward tasks, thus having a lesser need for \eval, and its uses
-originate from strings as these are easier to manipulate for end-users.
+originate from strings as these are easier to manipulate for end users.
 

### 2021-06-29 paper/main.tex
Update 2.3 section

@@
 
-R exposes a rich reflective interface with functions, \c{eval},
-\c{evalq}, \c{eval.parent}, and \c{local}. The following simplifies details not
-relevant to this paper, the interested reader should consult \citet{hadley}.
-
+The \eval function in R takes three parameters: an \c{expression} to evaluate
+(\c{e}), an environment where to evaluate (\c{envir}), and an enclosure
+(\c{encl}) that is used to look up objects not found in \c{envir}.\footnote{Next to the \eval function R
+  offers three other variants:
+  \begin{inparaenum}[1.]
+    \item \c{evalq} automatically quotes passed expression---a shorthand to
+      \c{eval(quote(...))},
+    \item \c{eval.parent(e, n)} specifies the evaluating
+      environment in terms of number of call frames (\c{n}) to go back---a shorthand
+      to \c{eval(e, parent.frame(n))}, and
+    \item \c{local} evaluates \c{e} in a fresh
+      environment---a shorthand to \c{evalq(e, new.env())}.
+  \end{inparaenum} }
+%
 \begin{lstlisting}
@@
 \end{lstlisting}
-The most general is \eval. Parameter \c e is the \c{expression} to evaluate,
-\c{envir} is the evaluation environment, and \c{encl} is used to look up
-variables not found in \c{envir}. The default value of the latter is either the
-caller's environment or the base environment \Eval can take several types for \c e, for
-our purposes on the \c{expression}. These can be thought of as the abstract
-syntax trees returned by the parser. An \c{expression} can be obtained by
-calling \c{quote} or \c{parse} with a string:
+
+\noindent The expression passed to \eval can be thought of as an abstract
+syntax tree. It can be parsed from a string, created manually or using
+reflection:
+
+\vspace*{-2mm}
+\begin{figure}[htb]
+\begin{minipage}{.49\textwidth}
+\begin{lstlisting}[caption={Examples of calls producing expression \c{a+b}}]
+  parse(text="a+b")
+  quote(a+b)
+  call("+", quote(a), quote(b))
+
+  # reflect promise
+  f <- function(x) substitute(x)
+  f(a+b)
+  \end{lstlisting}
+\end{minipage}
+\begin{minipage}{.49\textwidth}
+  \begin{lstlisting}[caption={Example of a call reflection}]
+  f <- function(x, y) {
+    mc <- match.call() # reflect curr. call
+    mc[[1]] <- as.name("g")
+    mc[["x"]] <- 2
+    mc
+  }
+  f(1,2) # returns a call g(2,y)
+  \end{lstlisting}
+\end{minipage}
+\end{figure}
+
+\vspace*{-2mm}
+\noindent The call to \c{substitute(x)} extracts the unevaluated expression from the
+promise \c{x}. The function \c{match.call} returns an expression representing the current call.
+
+R is generous when it comes to what is allowed to be passed as environment to
+\eval. Besides ordinary environments, it also accepts a list or a data
+frame (using element names or column names for variable resolution) or an
+integer \c{n} (in which case it would use the \c{n.} call frame). If no
+environment is specified, then the default is the environment where the call to
+\eval was made (as returned by \c{parent.frame}).
+%
+Environments nest, each has a parent. When creating a new environment with
+\c{new.env}, the parent is the current environment. Environment chains can be
+traversed with \c{parent.env}, until \c{emptyenv} is reached. The top-level
+environment is \c{.GlobalEnv}, it has parents that represent the packages that
+have been loaded. One can also directly read, modify or create new bindings,
+given any environment:
+%
 \begin{lstlisting}
- > quote(a + b)
- # a + b
- > parse(text="a+b")
- # expression(a+b)
+  # reading                                # writing
+  envir$v                                  envir$v <- 2
+  get("v", envir=envir)                    assign("v", 2, envir=envir)
 \end{lstlisting}
-Usually, expressions are obtained from promises using \c{substitute}. Consider a
-function called with \c{x = a+b}, then \c f returns an expression object.
-\begin{lstlisting}
- > f <- function(x) substitute(x)
- > f(a+b)
- # a+b
- > substitute(a+b,list(a=1,b=2))
- # 1+2
-\end{lstlisting}
-The call to \c{substitute(x)} extracts the unevaluated expression from the
-promise \c x. Function \c{substitute(e,envir)} takes two arguments, the second
-is used to substitute free variables in \c e.
-\begin{lstlisting}
- > environment()
- # <environment: R_GlobalEnv>
-\end{lstlisting}
-The \c{environment} function returns the current environment. Environments nest,
-each has a parent. When creating a new environment with \c{new.env}, the parent
-is the current environment. Environment chains can be traversed with
-\c{parent.env}, until \c{emptyenv} is reached. The top-level environment is
-\c{.GlobalEnv}, it has parents that represent the packages that have been
-loaded. One can also directly read, modify or create new bindings, given any
-environment:
-\begin{itemize}
-\item \c{envir\$v} and \c{get("v",envir=envir)}: read  \c{v} from \c{envir};
-\item \c{envir\$v<-2} and \c{assign("v",2,envir=envir)}: store 2 in \c{v}, if
-  not found, \c{v} is created in \c{envir}.
-\end{itemize}
+%
 \noindent Environments are used as hash maps as they have reference semantics
-and a built-in string lookup. Functions \c{parent.frame} and \c{sys.frame}
-return environments further up the call stack.
- \begin{lstlisting}
-  evalq <- function(e,envir,encl) eval(quote(e),envir,encl)
-  eval.parent <- function(e,n) eval(e,parent.frame(n))
-  local <- function(e,envir,encl) evalq(e,new.env())
-\end{lstlisting}
-The three variants can be expressed with \eval: \c{evalq} quotes its
-argument to prevent evaluation in the current environment,
-\c{eval.parent(e,n)} evaluates \c{e} in the environment of the {\tt n}-th
-caller of this function; finally, \c{local} evaluates \c{e} in a new environment
-to avoid polluting the current one.
+and a built-in string lookup.
 
@@
 arguments. For example, we taint the results of calls to \c{parse} and
-\c{match.call} (reflects on current call). We also capture calls to R API for
-dynamic code loading. Next, we subscribe to the events related to a variable
-definition and assignment, allowing us to record side effects that happen in
-environments while evaluating code in \eval. One challenge was that R only
-provides source references for block surrounded by braces. Thus, a function
-whose body is \c{eval(x)} would no usable debug information, while the same
-expression surrounded by braces would be fine. This is unfortunately not easily
-fixed. We extended the dynamic analysis tool to attach synthetic source code
-references to all \eval call sites by traversing ASTs. The tracer is implemented
-as an R package in 3.2K C++ and 1.3K of R code. For performance reasons, most of
-the tracing is done in C++. While in theory, the implementation is relatively
-straightforward, not so in practice:
+\c{match.call}. We also capture calls to R API for dynamic code loading. Next,
+we subscribe to the events related to a variable definition and assignment,
+allowing us to record side effects that happen in environments while evaluating
+code in \eval. One challenge was that R only provides source references for
+block surrounded by braces. Thus, a function whose body is \c{eval(x)} would no
+usable debug information, while the same expression surrounded by braces would
+be fine. This is unfortunately not easily fixed. We extended the dynamic
+analysis tool to attach synthetic source code references to all \eval call
+sites by traversing ASTs. The tracer is implemented as an R package in 3.2K C++
+and 1.3K of R code. For performance reasons, most of the tracing is done in
+C++. While in theory, the implementation is relatively straightforward, not so
+in practice:
 %

### 2021-06-30 paper/main.tex
Add link to the artifact
Add a placeholder for eval alternatives
Move the History of Lisp reference

--- 2021-06-27 paper/main.tex
@@
 code. We refer to this reflective facility as \eval bowing to its
-origins in LISP, all the way back in 1956~\cite{lisp}. \Eval has been
+origins in LISP, all the way back in 1956. \Eval has been
 much maligned over the years. In computing lore, it is as close to a
@@
 write down the definition of LISP; he was surprised that someone coded
-it up and offered it to end users. Since then, reflective facilities
+it up and offered it to end users~\cite{lisp}. Since then, reflective facilities
 have been used to parameterize programs over code patterns that can be

### 2021-07-08 analysis/preprocess.R
Preprocessing now adds the provenance information

@@ -7,12 +7,15 @@
 # - cut the dataset into smaller ones
 
 
-suppressPackageStartupMessages(library(tidyverse))
 suppressPackageStartupMessages(library(rlang))
+library(purrr)
+library(tidyr)
+library(dplyr)
+library(stringr)
 library(fs)
+library(readr)
 library(fst)
-
-
+library(optparse)
 
 get_expr <- function(eval_call) {
   if (is.na(eval_call)) {
@@ -212,15 +215,6 @@ package_name_from_call_stack <-
   }
 
 
-# we try to find a same combination of caller_package, caller_function and eval_call_expression in both datasets
-# if yes, we inditifed a call site
-# data set undefined should be the packages with missing srcref and which caller_function is not in base
-package_name_from_static <- function(dataset_undefined, static_data) {
-  dataset_undefined %>%
-    semi_join(static_data, by = c("eval_call_expression", "caller_package")) %>%
-    count(eval_call_expression, caller_package, caller_function)
-}
-
 extract_package_name <- function(src_ref, file) {
   # There are 5 possibilities for a srcref:
   # - NA
@@ -235,56 +229,79 @@ extract_package_name <- function(src_ref, file) {
     str_starts(src_ref, fixed("./R/")) ~ "core",
     str_starts(src_ref, fixed("/tmp/")) ~ str_match(src_ref, "/tmp.*/Rtmp[^/]*/R\\.INSTALL[^/]*/([^/]+)/R/.*$")[[2]], # path problem here
     str_starts(src_ref, fixed("/mnt/ocfs_vol")) ~ "base", # depends on where the shared is mounted!
+    str_starts(src_ref, fixed("::")) ~ str_match(src_ref, "::([^:]*)::.*")[[2]],
+    str_starts(src_ref, ".*::") ~ str_match(src_ref, "([^:]*)::.*")[[2]],
     str_starts(src_ref, fixed("test")) ~ str_match(src_ref, "([^/]*)/.*")[[2]],
     str_starts(src_ref, fixed("/:")) ~ str_match(file, "[^/]*/[^/]*/([^/]*)/.*")[[2]],
     TRUE ~ "unknown"
   )
 }
 
+# There are 5 main possibilities for a srcref:
+# - NA
+# - /tmp/Rtmp..../R.INSTALL....../packagename/R/file:linenumbers
+# - mnt/ocfs_vol_00/project-evalR/library/4.0/instrumentr/srcref/packagename/4.0.4/file:linenumbers
[…]

### 2021-07-08 paper/main.tex

--- 2021-06-29 paper/main.tex
@@
 of our observations will generalize to other languages.
+Our data and code are open source and publicly
+available: \url{https://github.com/PRL-PRG/evalr-experiment}.
 
@@
 
+\subsection{Alternatives to \eval}
+
+Talk about lazyeval and tidyeval
+
 \section{Conclusion}

### 2021-07-08 paper/main.tex
Expand Morandat et al comparison

@@
 
-\citet{ecoop12} had a short section on the usage of \eval in R. They found the
-that \eval  is widely used in R code with 8500 call sites in CRAN and 2.2 million dynamic calls. The 15 most frequent call sites account for 88\% of those. The
+\citet{ecoop12} had a few paragraphs on the usage of \eval in R. They found the
+that \eval  is widely used in R code with 8500 call sites in 1238 CRAN packages and 2.2 million dynamic calls.   The 15 most frequent call sites account for 88\% of those. The
 \c{match.arg} function is the highest used one with 54\% of all calls. In the
@@
 \eg as done in the \c{with} function. The other use case is the invocation of a
-function whose name or arguments are determined dynamically. For this purpose, R provides \c{do.call} and thus \eval is overkill.
+function whose name or arguments are determined dynamically.
+For this purpose, R provides \c{do.call} and thus \eval is overkill.
+Our corpus is larger, with \CranPackages packages, \packageTriggeredpkgs packages with at least one \eval, for a total of \PkgHitEvalCallSites sites and \packageAllcallsRnd dynamic calls, although they also looked at the Bioconductor package repository\footnote{\url{https://www.bioconductor.org/}} in addition to CRAN.
+We also track much more runtime information on the \eval calls during tracing.
+It enables us to have a more nuanced and accurate picture, and identify more diverse use cases of \eval.
 

### 2021-07-17 paper/main.tex
Switch pipeline diagram

@@
 \citestyle{acmauthoryear}   %% For author/year citations
-\usepackage{listings,hyperref,multirow,paralist,xspace,url,wrapfig}
+\usepackage{listings,hyperref,multirow,paralist,xspace,url,wrapfig,tikz}
+\usetikzlibrary{positioning,automata,fit,shapes.geometric,backgrounds,calc}
 \usepackage{tabularx}
@@
 
-\begin{figure}[!h]\hspace{-5mm}
-  \includegraphics[width=.95\linewidth]{pipeline.pdf}
-  \caption{Pipeline}\label{fig:pipeline}
+ \begin{figure}[ht!]
+   \centering
+   \scalebox{0.85}{
+     \begin{tikzpicture}
+       \definecolor{red}{HTML}{F8CECC}
+       \definecolor{darkred}{HTML}{B85450}
+       \definecolor{yellow}{HTML}{FFF2CC}
+       \definecolor{darkyellow}{HTML}{D6B656}
+       \definecolor{blue}{HTML}{DAE8FC}
+       \definecolor{darkblue}{HTML}{6C8EBF}
+       \definecolor{green}{HTML}{D5E8D4}
+       \definecolor{darkgreen}{HTML}{82B366}
+
+       \newcommand{\nodesep}[0]{0.035 \textwidth}
+       \newcommand{\textsep}[0]{0.005 \textwidth}
+
+       \newcommand{\nodename}[1]{\begin{tabular}{c}#1\end{tabular}}
+       \newcommand{\nodedesc}[1]{\begin{tabular}{c}#1\end{tabular}}
+
+       \tikzstyle{block}     = [rectangle, rounded corners, minimum width=0.15 \textwidth, minimum height=60pt]
+       \tikzstyle{connector} = [line width=0.25mm, ->]
+
+       \node [block, fill = red, draw = darkred, very thick] (download) {
+         \nodename{
+           \vspace{1mm}\large{\textbf{Download}}\vspace{1mm}\\
+           15K Packages\\
+           8K Kaggle Kernels\\
+           219 Kaggle Datasets
+         }
+       };
+       \node [below = \textsep of download]   (downloaddesc)   {\nodedesc{12 Hours}};
+
+       \node [block, right = \nodesep of download, fill = yellow, draw = darkyellow, very thick] (extract) {
+         \nodename{
+           \vspace{1mm}\large{\textbf{Extract}}\vspace{1mm}\\
+           240K package programs\\%, 4.6M LOC\\
+           74 Kaggle kernels\\%, 13K LOC\\
+           \newline
+         }
+       };
+       \node [below = \textsep of extract]   (extractdesc)   {\nodedesc{2 Hours}};
+
+       \node [block, right = \nodesep of extract, fill = blue, draw = darkblue, very thick] (trace) {
+         \nodename{
+           \vspace{1mm}\large{\textbf{Trace}}\vspace{1mm}\\
+           49M eval calls \\
+           14M side-effects\\
+           \newline
+         }
+       };
+       \node [below = \textsep of trace]     (tracedesc)     {\nodedesc{1 Day 12 Hours}};
+
+       \node [block, right = \nodesep of trace, fill = green, draw = darkgreen, very thick] (analyze) {
+         \nodename{
+           \vspace{1mm}\large{\textbf{Analyze}}\vspace{1mm}\\
+           2.8M unique eval calls\\
+           22\% packages use eval\\
+           \newline
+         }
+       };
+       \node [below = \textsep of analyze]    (analyzedesc)    {\nodedesc{2 Hours}};
+
+       \draw [connector] (download)   edge (extract);
+       \draw [connector] (extract)    edge (trace);
+       \draw [connector] (trace)      edge (analyze);
+     \end{tikzpicture}
+   }
+   \vspace{-2mm}\caption{Pipeline}\label{fig:pipeline}
  \end{figure}
+
+%% \begin{figure}[!h]\hspace{-5mm}
+%%  \includegraphics[width=.95\linewidth]{pipeline.pdf}
+%%  \caption{Pipeline}\label{fig:pipeline}
+%%\end{figure}
 

### 2021-07-26 analysis/provenance2.Rmd
For the new provenance analysis

--- 2021-04-04 analysis/provenance.Rmd
@@
 ---
-title: "Provenance"
+title: "Provenance analysis"
 author: "Pierre Donat-Bouillud"
@@
 static_path <- paste0(params$base_dir, params$static_path)
+
+provenance_path <- paste0(params$base_dir, params$provenance_path)
 
@@
 There are `r nb_eval_calls` eval calls and `r nb_call_sites` eval call sites in the dataset.
-
-# Classification of provenance
-
-## Expression tyoes
-
-```{r}
-expressions <- E_prov %>% filter(type == "expression")
-
-parsed_exprs <- E_prov %>% filter(!is.na(parsed))
-
-not_parsed_exprs <- expressions %>% anti_join(parsed_exprs)
-
-nb_parsed_exprs_sites <- parsed_exprs %>% pull(src_ref) %>% n_distinct()
-
-nb_not_parsed_exprs_sites <- not_parsed_exprs %>% pull(src_ref) %>% n_distinct()
-
-overview_table(
-    r("nb parse exprs sites", nb_parsed_exprs_sites),
-    r("parse exprs site percent", ratio(nb_parsed_exprs_sites, nb_call_sites)),
-    r("nb not parse exprs sites", nb_not_parsed_exprs_sites),
-    r("not parse exprs site percent", ratio(nb_not_parsed_exprs_sites, nb_call_sites)),
-)
-```
-
-## Language type
-
-```{r}
-languages <- E_prov %>% filter(type == "language")
-
-match_call_exprs <- E_prov %>% filter(!is.na(match_call))
-
-str2lang_exprs <- language %>% filter(!is.na(parsed))
-
-# Substitute, quote, bquote etc...
-other_langs <- languages %>% anti_join(match_call_exprs) %>% anti_join(str2lang_exprs)
-
-nb_match_call_exprs_sites <- match_call_exprs %>% pull(src_ref) %>% n_distinct()
-nb_str2lang_sites <- str2lang_exprs %>% pull(src_ref) %>% n_distinct()
-nb_other_langs_sites <- other_langs %>% pull(src_ref) %>% n_distinct()
-
-overview_table(
-    r("nb match call exprs sites", nb_match_call_exprs_sites),
-    r("match call exprs site percent", ratio(nb_match_call_exprs_sites, nb_call_sites)),
-    r("nb strlang sites", nb_str2lang_sites),
-    r("strlang site percent", ratio(nb_str2lang_sites, nb_call_sites)),
-    r("nb other langs sites", nb_other_langs_sites),
-    r("other langs site percent", ratio(nb_other_langs_sites, nb_call_sites))
-)
-```
-
-## Bytecode
-
-We disable bytecode optimization so if we see a bytecode type, it means that `comp`has been explicitly called to compile to bytecode.
-
-```{r}
-bytecode_exprs <- E_prov %>% filter(type == "bytecode")
-
-nb_bytecode_sites <- bytecode_exprs %>% pull(src_ref) %>% n_distinct()
-
-overview_table(
-    r("nb bytecode sites", nb_bytecode_sites),
-    r("bytecode site percent", ratio(nb_bytecode_sites, nb_call_sites))
-)
-```
-
-# More details for `parse` and expressions
-
-```{r}
-parse_func_exprs <- parsed_exprs %>% filter(str_detect(parsed, fixed("parse")))
-str2expression_exprs <- parsed_exprs %>% filter(str_detect(parsed, fixed("str2expression")))
-
-nb_parse_func_sites <- parse_func_exprs %>% pull(src_ref) %>% n_distinct()
-nb_str2expression_sites <- str2expression_exprs %>% pull(src_ref) %>% n_distinct()
-
-# File qrgument (but ignore the srcfile one)
-from_file_exprs <- parse_func_exprs %>% filter(str_detect(parsed, "\\bfile\\b"))
-nb_from_file_sites <- from_file_exprs %>% pull(src_ref) %>% n_distinct()
-
-overview_table(
-    r("nb parse func sites", nb_parse_func_sites),
-    r("parse func site percent", ratio(nb_parse_func_sites, nb_parsed_exprs_sites)),
-    r("nb strexpression sites", nb_str2expression_sites),
-    r("strexpression site percent", ratio(nb_str2expression_sites, nb_parsed_exprs_sites)),
-    r("nb parse from file sites", nb_from_file_sites),
-    r("parse from file site percent", ratio(nb_from_file_sites, nb_parse_func_sites))
-)
-```

### 2021-07-26 analysis/provenance2.Rmd
Provenances in number of calls

@@
 params:
-  base_dir: Data/package/
+  #base_dir: Data/package/
+  base_dir: /mnt/ocfs_vol_00/project-evalr/evalr-experiment/run/preprocess/package/
   calls_path: summarized.fst
@@
 # Tags will be prefixed by the dataset name
-create_tags(path(TAGS_DIR, paste0(dataset_name, "_provenance.tex")), prefix = dataset_name, default = TRUE)
+create_tags(path(TAGS_DIR, paste0(dataset_name, "_provenance2.tex")), prefix = dataset_name, default = TRUE)
 # Also use ggsave_prefix instead of ggsave if you need to save a graph
@@
 
-lock_path <- paste0(params$base_dir, ".provenance-lock.fst")
-min_path <- paste0(params$base_dir, "Eprov.fst")
+lock_path <- paste0(params$base_dir, ".provenance2-lock.fst")
+min_path <- paste0(params$base_dir, "Eprov2.fst")
 
@@
     caller_expression,
-    match_call = expr_match_call,
-    parsed = expr_parsed_expression,
+    provenance, # the main provenance (i.e. climbing up LHSs)
+    provenance_args,
+    nb_provenances,
+    nb_operations,
+    longest_path_size,
+    all_provenances
   ) -> E_prov
 
-  E_prov %>% write_fst(min_path)
+  #E_prov %>% write_fst(min_path)
   sz <- file.size(calls_path) %>% as_tibble()
-  write.fst(sz, lock_path)
+  #write.fst(sz, lock_path)
 }
@@
 There are `r nb_eval_calls` eval calls and `r nb_call_sites` eval call sites in the dataset.
+
+In the dataset, the `provenance` column refers to the main provenance of an expression passed to `eval`. It is obtained by traversing the provenance graph and following the LHSs.
+
+# Provenance in numbers of calls
+
+```{r}
+provenance_calls <- E_prov %>% count(provenance, wt = duplicates, sort = TRUE) %>% 
+  mutate(pc = 100 * n / sum(n), cpc = cumsum(pc))
+
+nrows <- min(10, nrow(provenance_calls))
+r_vec("nb provenance calls", nrows, provenance_calls$provenance)
+r_vec("percent provenance calls", nrows, provenance_calls$pc)
+r_vec("cumpercent provenance calls", nrows, provenance_calls$cpc)
+```
+
+`r datatable(provenance_calls)`
+
+# Provenance in number of sites
+
+# All provenances
+
+## By combinations
+
+## In absolute numbers

### 2021-08-03 paper/main.tex
Add some new uses of eval

@@
 
+\subsection{Simplify Interfaces}
+Many packages use the capability to parse and evaluate expressions in custom
+environments to design simpler interfaces. For example, the \c{AdapSamp} package
+provides functions that implement sampling algorithms. These functions take a
+\c{formula} argument, an expression in \c{x} as a string, that represents the
+kernel of the target distribution. This is turned into a closure using
+\c{function(x){eval(parse(text=formula))}}. While the \c{formula} supplied by
+the user could be a higher-order function, passing \c{"exp(-x\^2/2)"} as an
+argument is simpler for domain experts compared to \c{function(x) exp(-x\^2/2)}.
+The \c{bain::constraint\_to\_row} function has a similar pattern of \eval usage.
+It takes a constraint equation as a string argument, and, parses and evaluates
+it in a custom environment with the expected bindings. The
+\c{ctsem::ctModelTransformsToNum} function similarly evaluates a vector of
+formulae represented as a string vector.
+
+\subsection{Custom Evaluation}
+The \eval function is commonly used to customize the evaluation of expressions.
+An example is the \c{assertthat} package's \c{see\_if} function which is used to
+assert that the sequence of conditions supplied as arguments are true. It
+evaluates the condition expressions one-by-one inside a \c{tryCatch} block and
+produces an error at the first expression that returns \c{FALSE}. If the
+expression itself produces an error, the error message is captured by the
+enclosing \c{tryCatch} block and returned. The remaining expressions in the
+sequence remain unevaluated. A similar pattern is used by the \c{checkmate}
+package.
+
 \subsection{Domain Specific Language}
@@
 extract the unevaluated expression from function arguments (\c{substitute()})
-and \eval with first class environment makes the R toolkit for building DSLs.
+and \eval with first class environment makes R the toolkit for building DSLs.
 There are numerous widely spread DSLs used in R. For example, \c{data.table}
@@
 
-Similarly to JavaScript, there are also unnecessary uses of \eval. For example,
+Similarly to JavaScript, there are also unnecessary uses of \eval. These uses of
+\eval can be replaced with alternate functions yielding the same results.
+For example,
 the \c{PerformanceAnalytics} package contains a function \c{chart.QQPlot} that
@@
 
+  The \c{copyslots} function from the \c{coin} package uses \eval to copy the
+  slots from one \c{S4} object to another. It performs
+  \c{eval(str2lang(paste0("target@", s, " <- source@", s)))} for each slot name
+  \c{s}. This can be replaced by the expression \c{slot(target, s) <- slot(source,
+    s)} which uses the \c{slot<-} and \c{slot} functions to assign and retrieve
+  the slots by name.
+
+  Another examples is in the \c{configr::config.funs.par} function whose
+  simplified definition is shown below.
+
+\begin{lstlisting}
+config.funs.par <- function(fun = "", ...) {
+  args.all <- as.list(match.call()); args.all <- args.all[names(args.all) != ""];
+  parameters <- list()
+  text.1 <- sprintf("parameters <- names(as.list(args(%s)))", fun)
+  text.2 <- "parameters <- parameters[parameters != '']"
+  text.3 <- "parameters <- args.all[names(args.all) %in% parameters]"
+  for (j in c(text.1, text.2, text.3)) {
+    eval(parse(text = j))
+  }
+  return(parameters)
+}
+\end{lstlisting}
+
+  The function appears to use \eval since the function name \c{fun} is received
+  as a string. This string can be resolved to the function object using the
+  \c{get} function and the contents of \c{text.1}, \c{text.2}, and \c{text.3}
+  can be written directly without the need to convert them to text for
+  evaluation using \eval.
+
 The problem is that these cases are not easy to spot. Because of the
@@
 with an expression in a \c{F(), F(X)} or \c{F(F(X))} category.
+
+\subsection{Miscellaneous}
+A common use of \eval is the extraction of unevaluated text of arguments passed
+to \c{...} (vararg). For a non-vararg parameter \c{x}, the expressions
+\c{substitute(x)} can be used to extract the argument text. However, for \c{...}
+arguments, the expression \c{eval(quote(substitute(list(...))))} or its variants
+are used. The \eval call evaluates the quoted expression
+\c{substitute(list(...))}. This returns a \c{list} of unevaluated argument
+expressions. The \c{hasGroupNumber} function from the \c{lattice} package is one
+of the many users of this pattern.
 

### 2021-08-04 paper/main.tex
Add glue and withr package examples

@@
 
+The \c{glue} package provides functions for string interpolation; text enclosed
+by c{\{} and c{\}} is evaluated using \eval as an R expression and replaced by
+its result. The code snippet below shows an example of this.
+
+\begin{lstlisting}
+library(glue)
+greeting <- "Hello"
+glue('{greeting} World!')
+#> Hello World!
+\end{lstlisting}
+
 \subsection{Unnecessary use of \eval}
@@
 of the many users of this pattern.
+
+The \c{withr} package provides the \c{defer} function which defers the
+evaluation of an expression until a stack frame is exited. It attaches this
+expression as an attribute to the frame environment and uses \eval to evaluate
+it on frame exit.
 

### 2021-08-06 paper/main.tex
Fix figure text size and layout of Usage Metrics section

@@
          \nodename{
-           \vspace{1mm}\large{\textbf{Download}}\vspace{1mm}\\
+           \vspace{1mm}\textbf{Download}\vspace{1mm}\\
            15K Packages\\
@@
          \nodename{
-           \vspace{1mm}\large{\textbf{Extract}}\vspace{1mm}\\
+           \vspace{1mm}\textbf{Extract}\vspace{1mm}\\
            240K package programs\\%, 4.6M LOC\\
@@
          \nodename{
-           \vspace{1mm}\large{\textbf{Trace}}\vspace{1mm}\\
+           \vspace{1mm}\textbf{Trace}\vspace{1mm}\\
            49M eval calls \\
@@
          \nodename{
-           \vspace{1mm}\large{\textbf{Analyze}}\vspace{1mm}\\
+           \vspace{1mm}\textbf{Analyze}\vspace{1mm}\\
            2.8M unique eval calls\\
@@
   \centering
-  \includegraphics[width=72mm]{pkgs-eval-callsites-hist.pdf} \caption{CRAN
-  \eval call sites}%
+  \includegraphics[width=72mm]{pkgs-eval-callsites-hist.pdf}
+  \caption{CRAN \eval call sites}%
   \label{fig:pkgs-eval-callsites-hist}\vspace{-3mm}

### 2021-08-06 paper/main.tex
authors

--- 2021-08-04 paper/main.tex
@@
-\documentclass[review,screen,acmsmall,anonymous=true]{acmart}
-\settopmatter{printfolios=false,printccs=false,printacmref=false}
+\documentclass[review,screen,acmsmall]{acmart}
+%\settopmatter{printfolios=false,printccs=false,printacmref=false}
 \setcopyright{none}
@@
 
-\author{Author }
-\authornote{Both authors contributed equally to this research.}
-\email{a@b.com}\orcid{}\author{Author}\authornotemark[1]
-\email{r@s.com}
-\affiliation{\institution{Institute} \country{Country}}
+\author{Pierre Donat-Bouillud}\affiliation{\institution{Czech Technical University in Prague}\country{Czech Republic}}
+\author{Aviral Goel}\affiliation{\institution{Northeastern University}\country{USA}}
+\author{Filip Křikava}\affiliation{\institution{Czech Technical University in Prague}\country{Czech Republic}}
+\author{Christoph Kirsch}\affiliation{\institution{Czech Technical University in Prague}\country{Czech Republic}}
+\author{Jan Vitek}\affiliation{\institution{Northeastern University}\country{USA}}
+\affiliation{\institution{Czech Technical University in Prague}\country{Czech Republic}}
 

### 2021-08-06 analysis/provenance2.Rmd
Provenance tags

@@
     ev_expr = expr_expression,
+    ev_class = expr_resolved_class, # S3/S4 class
     caller_expression,
@@
 
-# Provenance in numbers of calls
+# Provenances using the representative provenance
 
-```{r}
+## Provenance in numbers of calls
+
+```{r prov_number_calls}
 provenance_calls <- E_prov %>% count(provenance, wt = duplicates, sort = TRUE) %>% 
@@
 nrows <- min(10, nrow(provenance_calls))
-r_vec("nb provenance calls", nrows, provenance_calls$provenance)
+r_vec("nb provenance calls", nrows, provenance_calls$n)
 r_vec("percent provenance calls", nrows, provenance_calls$pc)
@@
 
-# Provenance in number of sites
+## Provenance in number of sites
+
+```{r prov_number_sites}
+provenance_sites <- E_prov %>% group_by(src_ref) %>%
+  summarize(provenance = paste0(sort(unique(provenance)), collapse = "; "), n_provs = n_distinct(provenance)) %>% 
+  count(provenance, sort = TRUE) %>% 
+  mutate(pc = 100 * n / sum(n), cpc = cumsum(pc))
+
+# To get to 85 %
+
+nrows <- min(14, nrow(provenance_sites))
+r_vec("provenance name", nrows, if_else(provenance_sites$provenance == "", "NA", provenance_sites$provenance))
+r_vec("nb provenance sites", nrows, provenance_sites$n)
+r_vec("percent provenance sites", nrows, provenance_sites$pc)
+r_vec("cumpercent provenance sites", nrows, provenance_sites$cpc)
+
+```
+
+`r datatable(provenance_sites)`
 
@@
 ## In absolute numbers
+
+# Classes
+
+## In number of calls
+
+```{r class_nb_calls}
+classes_calls <- E_prov %>% count(ev_class, wt = duplicates, sort = TRUE) %>% 
+  mutate(pc = 100 * n / sum(n), cpc = cumsum(pc))
+
+```
+
+## In number of sites
+
+```{r class_nb_sites}
+classes_sites <- E_prov %>% group_by(src_ref) %>%
+  summarize(ev_class = paste0(sort(unique(ev_class)), collapse = "; "), n_classes = n_distinct(ev_class)) %>% 
+  count(ev_class, sort = TRUE) %>% 
+  mutate(pc = 100 * n / sum(n), cpc = cumsum(pc))
+
+```

### 2021-08-06 paper/main.tex
minor
change
Updating provenance

--- 2021-08-06 paper/main.tex
@@
          \nodename{
-           \vspace{1mm}\textbf{Download}\vspace{1mm}\\
+           \vspace{1mm}\large{\textbf{Download}}\vspace{1mm}\\
            15K Packages\\
@@
          \nodename{
-           \vspace{1mm}\textbf{Extract}\vspace{1mm}\\
+           \vspace{1mm}\large{\textbf{Extract}}\vspace{1mm}\\
            240K package programs\\%, 4.6M LOC\\
@@
          \nodename{
-           \vspace{1mm}\textbf{Trace}\vspace{1mm}\\
+           \vspace{1mm}\large{\textbf{Trace}}\vspace{1mm}\\
            49M eval calls \\
@@
          \nodename{
-           \vspace{1mm}\textbf{Analyze}\vspace{1mm}\\
+           \vspace{1mm}\large{\textbf{Analyze}}\vspace{1mm}\\
            2.8M unique eval calls\\
@@
 
-\newpage
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
   \centering
-  \includegraphics[width=72mm]{pkgs-eval-callsites-hist.pdf}
-  \caption{CRAN \eval call sites}%
+  \includegraphics[width=72mm]{pkgs-eval-callsites-hist.pdf} \caption{CRAN
+  \eval call sites}%
   \label{fig:pkgs-eval-callsites-hist}\vspace{-3mm}
@@
 Where does the expression passed to \eval come from? There are various means of
-creating that expression, which are associated with particular use cases. We
+creating that expression, which are associated with particular use cases.
+We track the functions that build expressions or combine expressions during tracing to create a direct acyclic graph of the functions creating the expression passed to \eval.
+Then, we pick a \emph{representative provenance}, \ie following the left-hand side of assignments in the graph.
+
+\NOTE{I can also put the absolute number of provenances, \ie if an expression has multiple provenances, we do not keep only the representative but also the other provenances.
+Also, if one site sometimes gets match.call and sometimes substitute, it is not going to be counted in either of them currently.}
+
+ Table~\ref{tab:detailed_provenances} shows the first provenances in number of sites, for a 85\% cumulative percentage.
+We only track the provenance of language expressions, not of floating point or strings for instance.
+In that case, the provenance is shown as \c{NA}.
+
+\begin{table}
+	\begin{tabular}{|c|c|c|}
+		\hline
+		Provenance & \#sites & \%sites \\
+		\hline
+		\packageProvenanceNamea & \packageNbProvenanceSitesa & \packagePercentProvenanceSitesa \\
+		\packageProvenanceNameb & \packageNbProvenanceSitesb & \packagePercentProvenanceSitesb \\
+		\packageProvenanceNamec & \packageNbProvenanceSitesc & \packagePercentProvenanceSitesc \\
+		\packageProvenanceNamed & \packageNbProvenanceSitesd & \packagePercentProvenanceSitesd \\
+		\packageProvenanceNamee & \packageNbProvenanceSitese & \packagePercentProvenanceSitese \\
+		\packageProvenanceNamef & \packageNbProvenanceSitesf & \packagePercentProvenanceSitesf \\
+		\packageProvenanceNameg & \packageNbProvenanceSitesg & \packagePercentProvenanceSitesg \\
+		\packageProvenanceNameh & \packageNbProvenanceSitesh & \packagePercentProvenanceSitesh \\
+		\packageProvenanceNamei & \packageNbProvenanceSitesi & \packagePercentProvenanceSitesi \\
+		\packageProvenanceNamej & \packageNbProvenanceSitesj & \packagePercentProvenanceSitesj \\
+		\packageProvenanceNamek & \packageNbProvenanceSitesk & \packagePercentProvenanceSitesk \\
+		\packageProvenanceNamel & \packageNbProvenanceSitesl & \packagePercentProvenanceSitesl \\
+		\packageProvenanceNamem & \packageNbProvenanceSitesm & \packagePercentProvenanceSitesm \\
+		\packageProvenanceNamen & \packageNbProvenanceSitesn & \packagePercentProvenanceSitesn \\
+		\hline
+	\end{tabular}
+	\caption{First provenances in number of sites.} \label{tab:detailed_provenances}
+\end{table}
+
+ We
 classify them into three categories:

### 2021-08-06 paper/main.tex

@@
 
-\author{Author }
-\authornote{Both authors contributed equally to this research.}
-\email{a@b.com}\orcid{}\author{Author}\authornotemark[1]
-\email{r@s.com}
-\affiliation{\institution{Institute} \country{Country}}
+\author{Aviral Goel}\affiliation{\institution{Northeastern
+    University}\country{USA}}
+\author{Pierre Donat-Bouillud}\affiliation{\institution{Czech Technical University in Prague}\country{Czech Republic}}
+\author{Filip Křikava}\affiliation{\institution{Czech Technical University in Prague}\country{Czech Republic}}
+\author{Christoph Kirsch}\affiliation{\institution{Czech Technical University in Prague}\country{Czech Republic}}
+\author{Jan Vitek}\affiliation{\institution{Northeastern University}\country{USA}}
+\affiliation{\institution{Czech Technical University in Prague}\country{Czech Republic}}
 
@@
 
-\newpage
+%\newpage
 \section{Background and Previous work}
@@
 
+%\newpage
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2021-08-06 analysis/provenance2.Rmd

@@
 nrows <- min(14, nrow(provenance_sites))
-r_vec("provenance name", nrows, if_else(provenance_sites$provenance == "", "NA", provenance_sites$provenance))
+r_vec("provenance name", nrows, case_when(provenance_sites$provenance == "" ~ "NA", 
+                                          provenance_sites$provenance == "$" ~ "\\$",
+                                          TRUE ~ provenance_sites$provenance))
 r_vec("nb provenance sites", nrows, provenance_sites$n)

### 2021-08-06 paper/main.tex
Classification of the provenances

@@
 In that case, the provenance is shown as \c{NA}.
-
-\begin{table}
+We were also not able to track the provenances of some expressions in some list or data frames.
+In that case, the provenance that we show corresponds to the extraction step from the list, \ie \c{[[} or \c{\$}.
+
+\begin{wraptable}{r}{5cm}\small\centering
 	\begin{tabular}{|c|c|c|}
@@
 	\end{tabular}
-	\caption{First provenances in number of sites.} \label{tab:detailed_provenances}
-\end{table}
-
- We
-classify them into three categories:
+	\caption{Provenances in number of sites (85\% first sites).} \label{tab:detailed_provenances}
+\end{wraptable}
+
+\ \NOTE{I only see 70 sites with formulas (using the class of the expression) so it does not appear here.}
+
+ We classify them into four categories:
 
@@
 \item {\it Constructed:} Expressions can be constructed by invoking the
-  \c{quote}, \c{enquote}, \c{expression} functions. Function arguments are
-  passed as promises, and \c{substitute} is used to retrieve the source
-  expression associated with the promise.
+  \c{quote}, \c{enquote}, \c{expression} functions, or by building a call with \c{call} or \c{as.call}.
 \item {\it Reflection:} This group corresponds to uses of \c{match.call} to
-  reflectively capture the expression that invoked the current function.
+  reflectively capture the expression that invoked the current function. Function arguments are passed as promises, and \c{substitute} is used to retrieve the source
+  expression associated with the promise. \c{formals} access the argument list of a function.
 \item {\it String:} Expressions created from strings by invoking \c{parse},
   \c{str2expression}, or \c{str2lang}.
+\item {\it Symbols: }   a symbol is created with \c{as.name} or \c{as.symbol}, typically to find the value of a symbol in a non-local environment.
 \end{compactitem}

### 2021-08-06 analysis/provenance2.Rmd
Provenance classification

@@
 ```
+
+# Classification
+
+```{r classification_calls}
+
+string_calls <- E_prov %>% filter(provenance %in% c("parse", "str2lang", "str2expression"))
+
+reflection_calls <- E_prov %>% filter(provenance %in% c("match.call", "formals", "substitute") | (provenance == "quote" & type != "symbol"))
+
+constructed_calls <- E_prov %>% filter(provenance %in% c("expression", "quote", "enquote", "call", "as.call", "as.vector"))
+
+symbol_calls <- E_prov %>% filter(provenance %in% c("as.name", "as.symbol") | (provenance == "quote" & type == "symbol"))
+
+```
+
+```{r classification_sites}
+
+string_sites <- string_calls %>% pull(src_ref) %>% n_distinct()
+
+reflection_sites <- reflection_calls %>% pull(src_ref) %>% n_distinct()
+
+constructed_sites <- constructed_calls %>%  pull(src_ref) %>% n_distinct()
+
+symbol_sites <- symbol_calls %>% pull(src_ref) %>% n_distinct()
+
+r("nb string sites", string_sites)
+r("string sites percent", ratio(string_sites, nb_call_sites))
+
+r("nb reflection sites", reflection_sites)
+r("reflection sites percent", ratio(reflection_sites, nb_call_sites))
+
+r("nb constructed sites", constructed_sites)
+r("constructed sites percent", ratio(constructed_sites, nb_call_sites))
+
+r("nb symbol sites", symbol_sites)
+r("symbol sites percent", ratio(symbol_sites, nb_call_sites))
+
+```

### 2021-08-06 paper/main.tex
Improve provenance section

@@
   \c{str2expression}, or \c{str2lang}.
-\item {\it Symbols: }   a symbol is created with \c{as.name} or \c{as.symbol}, typically to find the value of a symbol in a non-local environment.
+\item {\it Environment: }   a symbol is created with \c{as.name} or \c{as.symbol}, typically to find the value of a symbol in a non-local environment.
 \end{compactitem}
@@
 Origin  & \#sites & \%sites \\\hline
-Constructed & \packageNbConstructedSites & \packageNbConstructedSitePercent \\
-Reflection &  \packageNbMatchCallExprsSites & \packageMatchCallExprsSitePercent\\
-String & \packageNbStringSites & \packageNbStringSitePercent \\\hline
+Constructed & \packageNbConstructedSites & \packageConstructedSitesPercent \\
+Reflection &  \packageNbReflectionSites & \packageReflectionSitesPercent\\
+Environment & \packageNbSymbolSites & \packageSymbolSitesPercent \\
+String & \packageNbStringSites & \packageStringSitesPercent \\\hline
 \end{tabular}
@@
 cases could likely be replaced by macros if the R designers could be convinced
-to overcome their distaste for those.
+to overcome their distaste for those. Another use of \eval is to directly build symbols to evaluate them in another environment than the default one.
 

### 2021-08-06 paper/main.tex
Provenance
Move Unnecessary Uses of Eval section

@@
   \c{str2expression}, or \c{str2lang}.
-\item {\it Environment: }   a symbol is created with \c{as.name} or \c{as.symbol}, typically to find the value of a symbol in a non-local environment.
+\item {\it Environment: }   A symbol is created with \c{as.name} or \c{as.symbol}, typically to find the value of a symbol in a non-local environment.
 \end{compactitem}
@@
 
-\subsection{Unnecessary use of \eval}
+\subsection{Miscellaneous}
+A common use of \eval is the extraction of unevaluated text of arguments passed
+to \c{...} (vararg). For a non-vararg parameter \c{x}, the expressions
+\c{substitute(x)} can be used to extract the argument text. However, for \c{...}
+arguments, the expression \c{eval(quote(substitute(list(...))))} or its variants
+are used. The \eval call evaluates the quoted expression
+\c{substitute(list(...))}. This returns a \c{list} of unevaluated argument
+expressions. The \c{hasGroupNumber} function from the \c{lattice} package is one
+of the many users of this pattern.
+
+The \c{withr} package provides the \c{defer} function which defers the
+evaluation of an expression until a stack frame is exited. It attaches this
+expression as an attribute to the frame environment and uses \eval to evaluate
+it on frame exit.
+
+\subsection{Alternatives to \eval}
+Talk about lazyeval and tidyeval
+
+\subsection{Discussion}
+
+\Eval in R is used chiefly for meta-programming purposes and to access
+environments other than the local one. In Javascript, patterns are ad-hoc, such
+as json loading and parsing, and can often be rewritten without \eval, such as a
+function or method call, or an assignment to a local variable or an object
+property. For R, the data suggests that in some cases, the use of \eval in the
+\c{X, V, \$, <-, FUN} and the function calls related to code generation could
+replaced by more specific functions. For instance, variable lookup in any
+environment can be performed with \c{get}, and assignment in any environment,
+with \c{assign}. Building a call and executing can be done with \c{do.call}.
+Nevertheless, many of the observed \eval usage in R, particularly, the ones
+related to statistical modeling and DSLs are more sophisticated than those
+reported for JavaScript.
+
+Finally, let us mention an anecdotal use of \eval as a means to bypass some
+checks performed by the \c{R CMD CHECK} tool. This tool performs a number of
+well-formedness rules before accepting a package to the CRAN repository. Many
+checks are performed statically by looking at syntactic constructs in the R
+code. Using \eval allows one to bypass such checks and thus submit a package to
+CRAN that would otherwise be refused. For example,  it checks that packages
+do not call the \c{unlockBinding} function.\footnote{It allows one to open and
+modify loaded package namespaces.} Packages such as \c{data.table} circumvent
+such restrictions by using \eval as the check is only performed statically.
+
+\subsection{Removing \eval}
 
 Similarly to JavaScript, there are also unnecessary uses of \eval. These uses of
-\eval can be replaced with alternate functions yielding the same results.
-For example,
-the \c{PerformanceAnalytics} package contains a function \c{chart.QQPlot} that
-uses \eval to resolve a string into function and another to call it and assign
-its results into a variable:
+\eval can be replaced with alternate functions yielding the same results. For
+example, the \c{PerformanceAnalytics} package contains a function
+\c{chart.QQPlot} that uses \eval to resolve a string into function and another
+to call it and assign its results into a variable:
 \begin{lstlisting}
@@
 with an expression in a \c{F(), F(X)} or \c{F(F(X))} category.
-
-\subsection{Miscellaneous}
-A common use of \eval is the extraction of unevaluated text of arguments passed
-to \c{...} (vararg). For a non-vararg parameter \c{x}, the expressions
-\c{substitute(x)} can be used to extract the argument text. However, for \c{...}
-arguments, the expression \c{eval(quote(substitute(list(...))))} or its variants
-are used. The \eval call evaluates the quoted expression
-\c{substitute(list(...))}. This returns a \c{list} of unevaluated argument
-expressions. The \c{hasGroupNumber} function from the \c{lattice} package is one
-of the many users of this pattern.
-
-The \c{withr} package provides the \c{defer} function which defers the
-evaluation of an expression until a stack frame is exited. It attaches this
-expression as an attribute to the frame environment and uses \eval to evaluate
-it on frame exit.
-
-\subsection{Discussion}
-
-\Eval in R is used chiefly for meta-programming purposes and to access
-environments other than the local one. In Javascript, patterns are ad-hoc, such
-as json loading and parsing, and can often be rewritten without \eval, such as a
-function or method call, or an assignment to a local variable or an object
-property. For R, the data suggests that in some cases, the use of \eval in the
-\c{X, V, \$, <-, FUN} and the function calls related to code generation could
-replaced by more specific functions. For instance, variable lookup in any
-environment can be performed with \c{get}, and assignment in any environment,
-with \c{assign}. Building a call and executing can be done with \c{do.call}.
-Nevertheless, many of the observed \eval usage in R, particularly, the ones
-related to statistical modeling and DSLs are more sophisticated than those
-reported for JavaScript.
-
-Finally, let us mention an anecdotal use of \eval as a means to bypass some
-checks performed by the \c{R CMD CHECK} tool. This tool performs a number of
-well-formedness rules before accepting a package to the CRAN repository. Many
-checks are performed statically by looking at syntactic constructs in the R
-code. Using \eval allows one to bypass such checks and thus submit a package to
-CRAN that would otherwise be refused. For example,  it checks that packages
-do not call the \c{unlockBinding} function.\footnote{It allows one to open and
-modify loaded package namespaces.} Packages such as \c{data.table} circumvent
-such restrictions by using \eval as the check is only performed statically.
-
-\subsection{Alternatives to \eval}
-
-Talk about lazyeval and tidyeval
 

### 2021-08-06 paper/main.tex
Add alternative to \eval

@@
 
-\subsection{Alternatives to \eval}
-Talk about lazyeval and tidyeval
-
 \subsection{Discussion}
@@
 with an expression in a \c{F(), F(X)} or \c{F(F(X))} category.
+
+\subsection{Alternatives to \eval}
+
+An alternative implementation of \eval, \c{eval\_tidy}, is provided by the
+\c{rlang} package. This forms the base of the tidy evaluation framework used by
+the libraries collectively referred to as the \c{tidyverse}~\cite{tidyverse}. It
+differs from the \c{eval} function in two ways. First, it supports evaluation of
+Quosures, which are also custom objects used by the \c{tidyverse} libraries for
+metaprogramming. Quosures package expression with an evaluation environment. The
+\c{tidy\_eval} function extracts this expression and evaluates it in that
+environment, ignoring the environment passed to it explicitly as an argument.
+Second, the \c{eval\_tidy} function also accepts a data mask argument.This is a
+set of bindings supplied by the user as a data frame or a named vector or list.
+The data mask environment take precedence over the environment passed for
+evaluation. Since the expression is always evaluated in the data mask, it has
+different semantics compared to \c{base::eval}. Effects such as assignment using
+\c{<-} happen in the data mask environment. Expressions such as \c{return()} do
+not work since the data mask environment does not correspond to a frame on the
+call stack.
 

### 2021-08-09 paper/main.tex
Fix author list in the paper heading
methododoly

@@
-\documentclass[review,screen,acmsmall]{acmart}
+\documentclass[screen,acmsmall]{acmart}
 %\settopmatter{printfolios=false,printccs=false,printacmref=false}
@@
 \title{What We Eval in the Shadows}
-\subtitle{A large-scale study of {\sf eval} usage in R programs}
+\subtitle{A Large-scale Study of Eval in R programs}
 
@@
 
-Most dynamic languages provide their users with a facility to
-transform unstructured text into executable code and evaluate that
-code. We refer to this reflective facility as \eval bowing to its
-origins in LISP, all the way back in 1956. \Eval has been
-much maligned over the years. In computing lore, it is as close to a
-boogeyman as it gets. Yet, for McCarthy, \eval was simply the way to
-write down the definition of LISP; he was surprised that someone coded
-it up and offered it to end users~\cite{lisp}. Since then, reflective facilities
-have been used to parameterize programs over code patterns that can be
-provided after the program is written. The presence of such a feature
-in a language is a hallmark of dynamism; it is a form of delayed
-binding as the behavior of any particular call to \eval will only be
-known when the program is run, and that particular call site is
-evaluated.
-
-\vspace{2mm}\noindent\emph{Trouble in Paradise.} Reflective facilities
-hinder most attempts to reason about or apply meaning-preserving
-transformations to the code using them. In practice, \eval causes
-static analysis techniques to lose so much precision as to become
-pointless. For compilers, anything but the most trivial, local
-optimizations are unsound after the use of \eval. Furthermore, the
-addition of arbitrary code --- code that could have been obtained from
-a network connection --- as a program is running is a security
-vulnerability waiting to happen. To illustrate these challenges,
-consider the interaction of a static analysis tool with a dynamic
-language. An abstract interpretation-based program analyzer computes
-an over-approximation of the set of possible behaviors exhibited by
-the program under study~\cite{cc77}. A reflective facility may have \emph{any}
-behavior that can be expressed in the target language, \ie any legal sequence of instructions can replace \eval. As dynamic
-languages tend to be permissive, the analysis has to, for example,
-assume that many (or all) functions in scope may have been redefined,
-\eg that \texttt{`+`} now opens a network connection or something
-equally surprising. A single occurrence of \eval causes the static
-analyzer to lose all information about the program state and
-meaning of identifiers. This loss of precision can sometimes be
-mitigated by analyzing the string argument~\cite{moller03} to bound its
-possible behavior, but when the string comes from outside the program,
-not much can be done. A frustrated group of researchers argued giving up
-on soundness and, instead, under-approximating dynamic features
-(soundiness)~\cite{soundy}. In their words, ``a practical analysis,
-therefore, may pretend that \eval does nothing unless it can
-precisely resolve its string argument at compile time.'' Alas,
-assuming that \eval does not have side-effects or that side-effects
-will not affect the results of the analysis may be unduly optimistic.
+Most dynamic languages provide their users with a facility to transform
+unstructured text into executable code and evaluate that code. We refer to this
+reflective facility as \eval bowing to its origins in LISP, all the way back in
+1956. \Eval has been much maligned over the years. In computing lore, it is as
+close to a boogeyman as it gets. Yet, for McCarthy, \eval was simply the way to
+write down the definition of LISP; he was surprised that someone coded it up and
+offered it to end users~\cite{lisp}. Since then, reflective facilities have been
+used to parameterize programs over code patterns that can be provided after the
+program is written. The presence of such a feature in a language is a hallmark
+of dynamism; it is a form of delayed binding as the behavior of any particular
+call to \eval will only be known when the program is run, and that particular
+call site is evaluated.
+
+\vspace{2mm}\noindent\emph{Trouble in Paradise.} Reflective facilities hinder
+most attempts to reason about or apply meaning-preserving transformations to the
+code using them. In practice, \eval causes static analysis techniques to lose so
+much precision as to become pointless. For compilers, anything but the most
+trivial, local optimizations are unsound after the use of \eval. Furthermore,
+the addition of arbitrary code --- code that could have been obtained from a
+network connection --- as a program is running is a security vulnerability
+waiting to happen. To illustrate these challenges, consider the interaction of a
+static analysis tool with a dynamic language. An abstract interpretation-based
+program analyzer computes an over-approximation of the set of possible behaviors
+exhibited by the program under study~\cite{cc77}. A reflective facility may have
+\emph{any} behavior that can be expressed in the target language, \ie any legal
+sequence of instructions can replace \eval. As dynamic languages tend to be
+permissive, the analysis has to, for example, assume that many (or all)
+functions in scope may have been redefined, \eg that \texttt{`+`} now opens a
+network connection or something equally surprising. A single occurrence of \eval
+causes the static analyzer to lose all information about the program state and
+meaning of identifiers. This loss of precision can sometimes be mitigated by
+analyzing the string argument~\cite{moller03} to bound its possible behavior,
+but when the string comes from outside the program, not much can be done. A
+frustrated group of researchers argued giving up on soundness and, instead,
+under-approximating dynamic features (soundiness)~\cite{soundy}. In their words,
+``a practical analysis, therefore, may pretend that \eval does nothing unless it
+can precisely resolve its string argument at compile time.'' Alas, assuming that
+\eval does not have side-effects or that side-effects will not affect the
+results of the analysis may be unduly optimistic.
 
@@
 For this purpose, R provides \c{do.call} and thus \eval is overkill.
-Our corpus is larger, with \CranPackages packages, \packageTriggeredpkgs packages with at least one \eval, for a total of \PkgHitEvalCallSites sites and \packageAllcallsRnd dynamic calls, although they also looked at the Bioconductor package repository\footnote{\url{https://www.bioconductor.org/}} in addition to CRAN.
+Our corpus is larger, with \CranPackages packages, \packageTriggeredpkgs packages with at least one \eval, for a total of \PkgHitEvalCallSites sites and \packageAllcallsRnd dynamic calls, although they also looked at the Bioconductor package repository in addition to CRAN.
 We also track much more runtime information on the \eval calls during tracing.
@@
 
-This section explains how we selected our corpus and how we obtained the
-reported results.
+This section describes how the corpus of R program was selected and the analysis
+infrastructure.
 
@@
 
-Our corpus is assembled from three sources: the \emph{Base} libraries bundled
+The corpus is assembled from three sources: the \emph{Base} libraries bundled
 with the language implementation, packages hosted on \emph{CRAN}, and scripts
-from \emph{Kaggle}.
-
-\mypara{Base.} The dataset contains \BasePackages libraries performing basic
-arithmetics, statistics, and operating system functionalities. These libraries
-are bundled with R and executed pervasively. Together they contain
-\BaseEvalCallSites call sites to \eval in \BaseFunsWithEvals functions. Some of these
-functions provide basic functionalities such as package loading, and thus, there
-is hardly any code that does not invoke one of these.
-
-\mypara{CRAN.} The Comprehensive R Archive Network
-({\small \url{cran.r-project.org}}) is the largest curated repository of R
-packages. It hosts \CranAvailablePackagesRnd packages with 6 new ones submitted
-daily~\cite{Ligges2017}. Packages are authored by experienced developers and
-abide by well-formedness rules automatically checked on each commit.
-Each package comes with sample data. There are three sources of
-runnable code in a package: \emph{tests, examples}, and \emph{vignettes} --
-respectively, unit tests, code snippets from the documentation, and long-form
-use-cases written in RMarkdown. Examples and vignettes can be turned into
-scripts by extracting the relevant code. From \CranPackages packages we
-extracted \CranRunnableScripts scripts that contain \CranRunnableCode lines of
-code (\CranRunnableCodeExamplesRnd in examples, \CranRunnableCodeVignettesRnd in
-vignettes and \CranRunnableCodeTestsRnd in tests).
-
-\mypara{Kaggle.} The Kaggle website ({\small \url{kaggle.com}}) is an online
-platform for data science. It allows users to submit and compete to solve
-problems. Solutions, called \emph{kernels}, are uploaded as scripts or
-notebooks. While the quality of code is not uniform, each kernel is runnable,
-and maps to ones script. Input data is provided. We obtained \KaggleKernels
-kernels. Since Kaggle is not curated, we used SHA-1 hashes to identify and
-remove \KaggleDuplicates duplicate entries. The remaining \KaggleUnique kernels
-are unique solutions to \KaggleCompetitions competitions. They contain
-665K lines of R code. Only \KaggleWithEvals kernels call \eval.
+from \emph{Kaggle}. Each of these sources represents a different part of
+the user base of the language and thus, potentially, different usage patterns.
+
+\mypara{Base.} The \BasePackages libraries are bundled with R and widely used as
+they provide arithmetics, statistics, and operating system functionalities. Some
+\BaseFunsWithEvals functions contain \BaseEvalCallSites calls to \eval. The
+functions perform such key tasks such as package loading, there is thus hardly
+any R program that does not invoke them. These libraries have been written by
+experts with a deep understanding of the language and its environment.
+
+\mypara{CRAN.} This dataset consists of packages hosted on the Comprehensive R
+Archive Network ({\small \url{cran.r-project.org}}, aka CRAN). There are three
+sources of runnable code in a package: unit \emph{tests} for individual
+functions, code \emph{examples} automatically extracted from the documentation,
+and long-form use-cases called \emph{vignettes}. Package also often have sample
+input files. All of these runnable code snippets can be extract into independent
+files. We downloaded \CranPackages packages and extracted \CranRunnableScripts
+scripts, or \CranRunnableCode lines of code (\CranRunnableCodeExamplesRnd in
+examples, \CranRunnableCodeVignettesRnd in vignettes and
+\CranRunnableCodeTestsRnd in tests). CRAN packages are authored by experienced
+developers.
+
+\mypara{Kaggle.} This dataset contains \KaggleUnique programs from the Kaggle
+website ({\small \url{kaggle.com}}), an online platform that allows users to
+submit and compete to solve problems. Kaggle contains both scripts and notebooks
+as well as input files, but there is some duplication (we used SHA-1 hashes to
+remove \KaggleDuplicates duplicates). The dataset has 665K lines of R code and
+only \KaggleWithEvals calls \eval. Kaggle scripts are written by end-users with
+variable levels of expertise.
 
@@
 
-The analysis presented in this has been automated by a pipeline that acquires
-packages, extracts scripts, executes them, traces their behavior, and summarizes observations. Figure~\ref{fig:pipeline} shows the main steps along
- with their running time, data size, and  number of elements manipulated.
- Timings are from a cluster of three servers, each with 2.3GHz Intel Xeon 6140 processor with 72 cores and 256GB of RAM, and shared OCFS network storage. The pipeline steps are:
+We implemented an automated pipeline to acquires packages, extracts scripts,
+executes them, traces their behavior, and summarizes observations.
+Figure~\ref{fig:pipeline} shows the main steps along with their running time,
+data size, and number of elements manipulated. The pipeline steps are:
 
@@
 
-%% \begin{figure}[!h]\hspace{-5mm}
-%%  \includegraphics[width=.95\linewidth]{pipeline.pdf}
-%%  \caption{Pipeline}\label{fig:pipeline}
-%%\end{figure}
-
-\medskip
-\begin{compactenum}
-\item \emph{Download.} Packages are downloaded from CRAN. For scripts, a web
-  crawler retrieves code, and the Kaggle command-line tool gets data.
-  Installation is complicated by native dependencies, which are not adequately
-  documented and thus hard to resolve automatically.
-\item \emph{Extract.} Given installed packages, the \genthat tool extracts all
-  runnable code snippets and turns each of these into a self-standing
-  program~\cite{issta18}. Some Kaggle kernels are already scripts and nothing
-  more needs to be done. Kernels released as notebooks are processed by
-  \c{knitr} which extracts runnable code. The body of each program is
-  instrumented with calls to our dynamic analyzer to ensure that we only record
-  calls to \eval from the code of interest and not from bootstrapping or
-  execution harness operations.
-\item \emph{Trace.} Each program executes using our dynamic analysis tool, a
-  heavily instrumented interpreter that captures calls to \eval and many other
-  fine-grained runtime events. Packages are run twice, once to capture \eval
-  calls originating from package code and a second time to capture calls coming
-  from the base libraries. To avoid any interference, each program is run in its
-  own process with the GNU R compiler turned off to avoid recording its
-  execution.
-\item \emph{Analyze.} Finally, the analysis output is merged, cleaned, and summarized
-  in a post-processing phase driven by series of R scripts. The summarized data
-  is then analyzed in RMarkdown notebooks to gather insights. All figures and
-  numbers appearing in the paper are generated automatically. Figures are
-  produced in PDF by \c{ggplot2}; numbers are exported as \LaTeX macros.
-\end{compactenum}
-
-\noindent The pipeline runs in parallel~\cite{GNUparallel} orchestrated
-by a Makefile. Servers have identical environments thanks to docker images with
-all dependencies installed.
+\begin{enumerate}
+\item \emph{Download.} Packages are downloaded from CRAN, for Kaggle a web
+  crawler retrieves code, and a command-line tool gets data. Installation is
+  complicated by native dependencies which sometimes have to be resolved
+  manually.
+\item \emph{Extract.} The \genthat tool extracts all runnable code snippets and
+  turns each of these into a self-standing program~\cite{issta18}, \c{knitr}
+  extracts code from notebooks. Each extracted script is instrumented with calls
+  to our dynamic analyzer in a way that avoid recording calls to \eval
+  originating from our execution harness.
+\item \emph{Trace.} Script run with a modified R interpreter to capture
+  execution traces. Each script is run in its own process with the GNU R
+  compiler turned off to avoid recording its execution. CRAN scripts run twice,
+  once to capture \eval calls originating from package code and a second time to
+  capture calls coming from base libraries.
+\item \emph{Analyze.} Analysis outputs are merged, cleaned, and summarized.
+  RMarkdown notebooks process the summarized data to generate all graphs (PDF
+  from \c{ggplot2}) and numbers (exported as \LaTeX macros) appearing in the
+  paper.
+\end{enumerate}
+
+\noindent The pipeline runs in parallel on three servers, each with 2.3GHz Intel
+Xeon 6140 processor with 72 cores and 256GB of RAM, and shared OCFS network
+storage. Servers run identical docker images with all dependencies installed.
 
@@
 
-The dynamic analysis is performed by \rdyntrace, a modified R virtual machine
-based on GNU R 4.0.2 that exposes low-level callbacks for a variety of runtime
-events~\cite{oopsla19b}. The tracer registers callbacks to all \eval variants
-and a few additional functions to assist in locating the origin of \eval
-arguments. For example, we taint the results of calls to \c{parse} and
-\c{match.call}. We also capture calls to R API for dynamic code loading. Next,
-we subscribe to the events related to a variable definition and assignment,
+The tracer that performs dynamic analysis of R scripts is built on top of
+\rdyntrace, an extended R 4.0.2 virtual machine that exposes callbacks for
+various runtime events~\cite{oopsla19b}. The tracer registers callbacks to all
+variants of \eval and a few additional functions to locate the origin of \eval
+arguments (it taints the results of calls to \c{parse} and \c{match.call}). It
+captures dynamic code loading, as well as variable definition and assignment,
 allowing us to record side effects that happen in environments while evaluating
-code in \eval. One challenge was that R only provides source references for
-block surrounded by braces. Thus, a function whose body is \c{eval(x)} would no
-usable debug information, while the same expression surrounded by braces would
-be fine. This is unfortunately not easily fixed. We extended the dynamic
-analysis tool to attach synthetic source code references to all \eval call
-sites by traversing ASTs. The tracer is implemented as an R package in 3.2K C++
-and 1.3K of R code. For performance reasons, most of the tracing is done in
-C++. While in theory, the implementation is relatively straightforward, not so
-in practice:
-%
-\begin{compactitem}[---]
-
-\item The lazy evaluation makes it challenging to analyze function arguments
-  while tracing as prematurely forcing a promise might have dangerous
-  consequences.
-
-\item While the R interpreter is implemented in C, most core functionality is in
-  R. For example, package loading is implemented in R using \eval. This makes it
-  hard in the tracer to separate the \eval that are essential to user programs
-  from the accidental ones that are products of the way R implements its basic
-  operations. This is even more so for the side-effects analysis.
-
-\item In the dynamic analysis, we run all the \emph{runnable} code we obtained
-  from CRAN and Kaggle---\ie actual code written by people with highly varying
-  expertise in R and programming in general. This exercises a lot of the corner
-  cases of the highly underspecified R behavior.
-
-\end{compactitem}
+code in \eval. A challenge was that lack of source code references for
+expression that are not within block surrounded by braces. This is unfortunately
+not easily fixed. We extend our dynamic analysis tool to attach synthetic source
+code references to all \eval call sites, but the approach fails in some edge
+cases. For performance reasons, the tracer is an R package written in C++ (3.2K
+LOC) and R (1.3K LOC). While in theory the implementation should be
+straightforward, not so in practice. Lazy evaluation requires delaying the
+processing of arguments until (and if) they are forced. Accounting for
+side-effects performed in an \eval is complicated by the fact that R is
+implemented in a mixture of R and C, and that the language implementation can
+call \eval. Lastly, large code bases exercise many edge cases of the highly
+underspecified R behavior.
 
 \mypara{Limitations.} Even with the extension described above, there are
-\PkgUndefinedRnd \eval calls without source references. This occurs when \eval
-is passed as an argument to a higher-order function or when the \eval call
-originates from native code. However, these missing source references account
-for a meager \PkgUndefinedRatio of all calls; they are unlikely to affect our
-results. Other limitations are that we ignore calls to the native \eval and to
-the alternate \c{rlang::tidy\_eval} uses native \eval internally. None of these
-limitations should invalidate our conclusions.
-
-%\newpage
+\PkgUndefinedRnd \eval calls without source information (\PkgUndefinedRatio of
+all \eval calls). This happen when \eval is an argument to a higher-order
+function or when it is called from native code. An other limitation is that we
+do not record calls to the native \eval. Neither of these limitations should
+invalidate our conclusions.
+
+\newpage
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-% It's dubious whether R programmers would develop with macros though.
-% Macros and R
-% The Thomas Lumley's article, Programmer's Niche: Macros in R, (https://www.r-project.org/doc/Rnews/Rnews_2001-3.pdf) describes how to write a `defmacro` function that would like as a macro creator, using `substitute` and `eval`.
-There are no local macro variables in this implementation though.
-
 \bibliography{bib/bibliography,bib/jv}

### 2021-08-09 paper/main.tex
3

@@
 \maketitle
+\renewcommand{\shortauthors}{A. Goel, P. Donat-Bouillud, F. Křikava, C. Kirsch and J. Vitek}
 
@@
 
-The corpus is assembled from three sources: the \emph{Base} libraries bundled
-with the language implementation, packages hosted on \emph{CRAN}, and scripts
-from \emph{Kaggle}. Each of these sources represents a different part of
-the user base of the language and thus, potentially, different usage patterns.
-
-\mypara{Base.} The \BasePackages libraries are bundled with R and widely used as
-they provide arithmetics, statistics, and operating system functionalities. Some
-\BaseFunsWithEvals functions contain \BaseEvalCallSites calls to \eval. The
-functions perform such key tasks such as package loading, there is thus hardly
-any R program that does not invoke them. These libraries have been written by
-experts with a deep understanding of the language and its environment.
-
-\mypara{CRAN.} This dataset consists of packages hosted on the Comprehensive R
-Archive Network ({\small \url{cran.r-project.org}}, aka CRAN). There are three
-sources of runnable code in a package: unit \emph{tests} for individual
-functions, code \emph{examples} automatically extracted from the documentation,
-and long-form use-cases called \emph{vignettes}. Package also often have sample
-input files. All of these runnable code snippets can be extract into independent
-files. We downloaded \CranPackages packages and extracted \CranRunnableScripts
-scripts, or \CranRunnableCode lines of code (\CranRunnableCodeExamplesRnd in
-examples, \CranRunnableCodeVignettesRnd in vignettes and
-\CranRunnableCodeTestsRnd in tests). CRAN packages are authored by experienced
-developers.
-
-\mypara{Kaggle.} This dataset contains \KaggleUnique programs from the Kaggle
-website ({\small \url{kaggle.com}}), an online platform that allows users to
-submit and compete to solve problems. Kaggle contains both scripts and notebooks
-as well as input files, but there is some duplication (we used SHA-1 hashes to
-remove \KaggleDuplicates duplicates). The dataset has 665K lines of R code and
-only \KaggleWithEvals calls \eval. Kaggle scripts are written by end-users with
-variable levels of expertise.
+We distinguish three sources for calls to \eval, those originating from
+\emph{Base} libraries bundled with R, packages hosted on \emph{CRAN}, and
+end-user scripts from \emph{Kaggle}. We consider base libraries as part of the
+language implementation and do not include them in our analysis.
+
+\mypara{Base.} The \BasePackages base libraries provide arithmetics, statistics,
+and operating system functionalities. Base has \BaseFunsWithEvals functions with
+\BaseEvalCallSites calls to \eval. These functions perform key tasks such as
+package loading, there is thus hardly any R program that does not invoke them.
+
+\mypara{CRAN.} Packages hosted on the Comprehensive R Archive Network ({\small
+  \url{cran.r-project.org}}, aka CRAN) have three sources of runnable: unit
+\emph{tests} for individual functions, code \emph{examples} embedded in the
+documentation, and use-cases called \emph{vignettes}. All runnable snippets can
+be extracted into independent files. We download \CranPackages packages and
+extract \CranRunnableScripts scripts with \CranRunnableCode lines of code
+(\CranRunnableCodeExamplesRnd in examples, \CranRunnableCodeVignettesRnd in
+vignettes and \CranRunnableCodeTestsRnd in tests). CRAN packages are authored by
+experienced developers.
+
+\mypara{Kaggle.} Kaggle ({\small \url{kaggle.com}}) is an online platform that
+allows users to submit and compete to solve problems. We download \KaggleUnique
+unique scripts and notebooks with their input files (\KaggleDuplicates
+duplicates were removed using SHA-1 hashes); 665K lines of R code making only
+\KaggleWithEvals calls \eval. Authors of these scripts have highly variable
+levels of expertise.
+
+\mypara{Discussion.} When we started this project, our goal was to contrast the
+usage of \eval by package developers and end-users, but the Kaggle dataset have
+so few calls to \eval that there is little that can be said about those. The
+remainder of the paper focuses on CRAN with a few observations about the other
+sources when relevant.
 
@@
        \definecolor{darkgreen}{HTML}{82B366}
-
        \newcommand{\nodesep}[0]{0.035 \textwidth}
        \newcommand{\textsep}[0]{0.005 \textwidth}
-
        \newcommand{\nodename}[1]{\begin{tabular}{c}#1\end{tabular}}
        \newcommand{\nodedesc}[1]{\begin{tabular}{c}#1\end{tabular}}
-
        \tikzstyle{block}     = [rectangle, rounded corners, minimum width=0.15 \textwidth, minimum height=60pt]
@@
        \node [below = \textsep of download]   (downloaddesc)   {\nodedesc{12 Hours}};
-
        \node [block, right = \nodesep of download, fill = yellow, draw = darkyellow, very thick] (extract) {
@@
        \node [below = \textsep of extract]   (extractdesc)   {\nodedesc{2 Hours}};
-
        \node [block, right = \nodesep of extract, fill = blue, draw = darkblue, very thick] (trace) {
@@
        \node [below = \textsep of trace]     (tracedesc)     {\nodedesc{1 Day 12 Hours}};
-
        \node [block, right = \nodesep of trace, fill = green, draw = darkgreen, very thick] (analyze) {
@@
        \node [below = \textsep of analyze]    (analyzedesc)    {\nodedesc{2 Hours}};
-
        \draw [connector] (download)   edge (extract);
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Usage Metrics}
+\section{Measuring Eval across Datasets}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-Figure~\ref{cn}(a) shows normalized call counts per site; on the left are
-the average number of calls from a given site and given run, on the right are the counts of sites that fall in that range. For instance, \packageRunbinh sites are
+Figure~\ref{cn}(a) shows normalized call counts per site; on the left are the
+average number of calls from a given site and given run, on the right are the
+counts of sites that fall in that range. For instance, \packageRunbinh sites are
 invoked 2,000+ times per run. Larger numbers suggest loops or recursive contexts
@@
 50 times or less. Figure~\ref{cn}(b) zooms in on low frequency sites. The x-axis
-shows normalized calls, and the y-axis is the number of sites for that value. Most
-low-frequency \evals are invoked only once; about half as many are invoked
+shows normalized calls, and the y-axis is the number of sites for that value.
+Most low-frequency \evals are invoked only once; about half as many are invoked
 twice; after that, the frequency quickly drops.
@@
 
-To estimate how much executable code is injected through \eval, we measure the number of nodes in the expressions; for example, \c{x+1} counts as 3. We
-also measured string lengths of unparsed expressions, but these measurements
-were dominated by the size of data objects which could range in the MBs. The
-median argument size is \packageMedianszeval (due to the symbols), and the
-average is \packageAvgszeval nodes. The largest \eval input observed is
+To estimate how much executable code is injected through \eval, we measure the
+number of nodes in the expressions; for example, \c{x+1} counts as 3. We also
+measured string lengths of unparsed expressions, but these measurements were
+dominated by the size of data objects which could range in the MBs. The median
+argument size is \packageMedianszeval (due to the symbols), and the average is
+\packageAvgszeval nodes. The largest \eval input observed is
 \packageMaxszeval\,-- a significant chunk of code.
@@
 \paragraph{Base}\label{sec:usage-base}
-
-We recorded \baseAllcalls \eval calls in the \BasePackages base libraries. This
+We recorded \baseAllcalls \eval calls in the base libraries. This
 comes from running 10\% randomly selected programs from the

### 2021-08-09 paper/main.tex
Remove base and kaggle from Section 4

@@
 \emph{tests} for individual functions, code \emph{examples} embedded in the
-documentation, and use-cases called \emph{vignettes}. All runnable snippets can
-be extracted into independent files. We download \CranPackages packages and
+documentation, and use-cases called \emph{vignettes}. All runnable snippets
+can be extracted into independent files. We download \CranPackages packages and
 extract \CranRunnableScripts scripts with \CranRunnableCode lines of code
 (\CranRunnableCodeExamplesRnd in examples, \CranRunnableCodeVignettesRnd in
-vignettes and \CranRunnableCodeTestsRnd in tests). CRAN packages are authored by
-experienced developers.
+vignettes and \CranRunnableCodeTestsRnd in tests). CRAN packages contain 20
+years of contribution from thousands from thousands of authors, often
+experienced R developers.
 
@@
            \vspace{1mm}\textbf{Download}\vspace{1mm}\\
-           15K Packages\\
-           8K Kaggle Kernels\\
-           219 Kaggle Datasets
+           15K packages\\
+           8K Kaggle kernels\\
+           219 Kaggle datasets
          }
        };
-       \node [below = \textsep of download]   (downloaddesc)   {\nodedesc{12 Hours}};
+       \node [below = \textsep of download]   (downloaddesc)   {\nodedesc{12 hours}};
        \node [block, right = \nodesep of download, fill = yellow, draw = darkyellow, very thick] (extract) {
@@
        };
-       \node [below = \textsep of extract]   (extractdesc)   {\nodedesc{2 Hours}};
+       \node [below = \textsep of extract]   (extractdesc)   {\nodedesc{2 hours}};
        \node [block, right = \nodesep of extract, fill = blue, draw = darkblue, very thick] (trace) {
@@
        };
-       \node [below = \textsep of trace]     (tracedesc)     {\nodedesc{1 Day 12 Hours}};
+       \node [below = \textsep of trace]     (tracedesc)     {\nodedesc{36 hours}};
        \node [block, right = \nodesep of trace, fill = green, draw = darkgreen, very thick] (analyze) {
@@
        };
-       \node [below = \textsep of analyze]    (analyzedesc)    {\nodedesc{2 Hours}};
+       \node [below = \textsep of analyze]    (analyzedesc)    {\nodedesc{2 hours}};
        \draw [connector] (download)   edge (extract);
@@
 
-\noindent The pipeline runs in parallel on three servers, each with 2.3GHz Intel
-Xeon 6140 processor with 72 cores and 256GB of RAM, and shared OCFS network
-storage. Servers run identical docker images with all dependencies installed.
+\vspace{1mm} 
+\noindent The pipeline runs in parallel~\cite{GNUparallel}
+orchestrated by a Makefile. Servers have identical environments thanks to a
+docker image with all dependencies installed\footnote{The docker image is part
+  of the accompanied artifact available at\\
+\url{https://github.com/PRL-PRG/evalr-experiment/tree/oopsla21}}.
 
@@
 \end{figure}
-
-\paragraph{CRAN}
 
@@
 
-\paragraph{Base}\label{sec:usage-base}
-We recorded \baseAllcalls \eval calls in the base libraries. This
-comes from running 10\% randomly selected programs from the
-\CranRunnableScriptsRnd extracted programs from CRAN packages, covering
-\baseTriggeredevalpct of the \BaseEvalCallSites sites. Most of the calls
-(\baseEvalsratio) are to the \c{eval} function, and most arguments
-(\baseCodepercent) are expressions. However, unlike CRAN, the majority
-(\baseLanguagepercent) of the arguments are language objects. The median
-argument size is \baseMedianszeval, which is more than for CRAN and the maximum
-size is \baseMaxszeval, less than for CRAN. Small instruction counts ($<50$)
-amount to \baseSmalleventspct of calls. A single site in the \c{match.arg}
-function is responsible for \baseTopFuncPercent of all recorded calls. This
-function provides a convenient way for argument verification using partial
-matching. It is heavily relied on in R functions that use string arguments to
-parameterize function behavior, common practice in R. For example, in a body of
-\c{center<-function(x,type=c("mean","median","trimmed"))} one can get the
-value of \c{type} parameter using \c{type<-match.arg(type)}. A call
-\c{center(x,"trim")} will match \c{type} to \c{"trimmed"}. The challenge
-dealing with Base is that every program uses it. It can thus generate extreme
-amounts of data, but the data is quite predictable. From now on, we only mention
-Base when it surprises.
-
-\paragraph{Kaggle}
-
-In total, \kaggleAllcalls \eval were recorded, all to \eval function. Out of
-\kaggleStaticeval sites from \KaggleWithEvals scripts, only \kaggleTriggeredeval
-sites in \kaggleNbruns scripts were hit; \KaggleFailedScripts scripts failed to
-run, and the others did not exercise the \eval sites. This is partially expected
-as Kaggle code does not need to abide by any checks. Upon a manual inspection,
-we observed that indeed, the failing scripts were of poor quality, often not
-finished, using misspelled package names, hard-coded file paths, or accessing
-missing files. Most call sites are invoked once. Only one site is called
-\kaggleMaxcalls times. Expressions amount for \kaggleCodepercent of arguments,
-there are very few symbols (\kaggleSymbolpercent). Most expressions result from
-calls to \c{parse}; thus, most \evals start with strings. The median argument
-size is \kaggleMedianszeval, which makes sense, as few arguments are only
-symbols. The largest argument is \kaggleMaxszeval. The distribution of
-instructions per eval is similar to CRAN; \kaggleSmalleventspct of \evals
-execute fewer than 50 instructions. Manual inspection of \eval usage in Kaggle
-suggests that it is consistent with the data obtained for CRAN. We do not
-discuss it further.
-
-\paragraph{Discussion}
-\Eval is central to R's implementation as it is omnipresent in the Base library.
-That library is, in turn, used by every single package. To avoid Base dominating
-the result, we excluded calls to it from the results. CRAN Packages, which are
-typically developed by experienced programmers, make regular and varied use of
-\eval. They represent our most interesting data set as these packages result
-from over 20 years of contributions by thousands of authors. The code in
-packages is well maintained and relatively well tested. Finally, Kaggle scripts
-are often written by less sophisticated users and likely perform more
-straightforward tasks, thus having a lesser need for \eval, and its uses
-originate from strings as these are easier to manipulate for end users.
+\paragraph{Summary}
+
+\Eval is used in CRAN packages. Nearly every fourth package contains at least
+one \eval site. However, most packages that use \eval do that modestly. Half of
+them contain less than three \eval sites. Packages that rely heavily on \eval
+are mostly the ones that provide functions for statistical modeling and
+simulation. At runtime, over third of extracted programs triggered calls to
+\eval. Half of the \packageAllcallsRnd recorded calls come from a single \eval
+site, but for the rest, most of the packages have low \eval call frequency.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{A Taxonomy of Eval}
+\section{A Taxonomy of \Eval}
 

### 2021-08-09 paper/main.tex
Update S4

@@
 
-\paragraph{Summary}
+\paragraph{Discussion}
 
@@
 one \eval site. However, most packages that use \eval do that modestly. Half of
-them contain less than three \eval sites. Packages that rely heavily on \eval
-are mostly the ones that provide functions for statistical modeling and
-simulation. At runtime, over third of extracted programs triggered calls to
-\eval. Half of the \packageAllcallsRnd recorded calls come from a single \eval
-site, but for the rest, most of the packages have low \eval call frequency.
+them include less than three \eval sites. Packages that rely heavily on \eval
+are mostly the ones that provide statistical modeling and simulation functions.
+At runtime, over a third of extracted programs triggered calls to \eval. Half
+of the \packageAllcallsRnd recorded calls come from a single \eval site, but
+for the rest, most of the packages have low \eval call frequency.
 

### 2021-08-09 analysis/provenance2.Rmd
Parse from file number
Addin the external line

@@
 
+external_calls <- E_prov %>% filter(provenance == ".External")
+
 ```
@@
 symbol_sites <- symbol_calls %>% pull(src_ref) %>% n_distinct()
+
+external_sites <- external_calls %>% pull(src_ref) %>% n_distinct()
 
@@
 
+r("nb external sites", external_sites)
+r("external sites percent", ratio(symbol_sites, nb_call_sites))
+
 ```
+
+## Parse from a file
+
+```{r parse_from_file}
+
+parsed_from_file_calls <- string_calls %>% filter(str_detect(provenance_args, "\\bfile\\b"))
+
+parsed_from_file_sites <- parsed_from_file_calls %>% pull(src_ref) %>% n_distinct()
+
+r("nb parse from file sites", parsed_from_file_sites)
+
+```

### 2021-08-09 paper/main.tex
Better explain the provenance analysis

--- 2021-08-09 paper/main.tex
@@
 creating that expression, which are associated with particular use cases.
-We track the functions that build expressions or combine expressions during tracing to create a direct acyclic graph of the functions creating the expression passed to \eval.
-Then, we pick a \emph{representative provenance}, \ie following the left-hand side of assignments in the graph.
-
-\NOTE{I can also put the absolute number of provenances, \ie if an expression has multiple provenances, we do not keep only the representative but also the other provenances.
-Also, if one site sometimes gets match.call and sometimes substitute, it is not going to be counted in either of them currently.}
-
- Table~\ref{tab:detailed_provenances} shows the first provenances in number of sites, for a 85\% cumulative percentage.
-We only track the provenance of language expressions, not of floating point or strings for instance.
-In that case, the provenance is shown as \c{NA}.
-We were also not able to track the provenances of some expressions in some list or data frames.
-In that case, the provenance that we show corresponds to the extraction step from the list, \ie \c{[[} or \c{\$}.
-
+During tracing, at each function call, we check if the return value is an expression or a symbol, or if the function is a function of interest.
+If it is, we record it and also look at whether the arguments were recorded before if it is not an expression-building function.
+It makes it possible to build a direct acyclic graph of the functions that build expressions or combine expressions during tracing to create a direct acyclic graph of the functions creating the expression passed to \eval.
+The following example yields the provenance graph with four nodes on Figure~\ref{fig:provgraph} for expression \c{e} passed to \eval.
+
+\begin{lstlisting}
+e <- parse(text = "a ; b")
+e[[2]] <- quote(c)
+eval(e)
+\end{lstlisting}
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=8cm]{provenance-graph}
+\caption{The provenance DAG for the code example above. Assignments and complex assignments are intermediate nodes, whereas expression builders are leaves of the graph.} \label{fig:provgraph}
+\end{figure}
+
+Expression \c{e} has two origins, \c{parse} and \c{quote}, which corresponds to the leaves of the graph.
+We also extract from the graph a representative origin, by traversing the graph from the root, following the left-hand side member of the assignments until a leaf.
+It represents the provenance that is further modified to yield the expression passed to \eval.
+In our example, the representative provenance is \c{parse}.
+
+ Table~\ref{tab:detailed_provenances} shows the first provenances in number of sites.
+In \packagePercentProvenanceSitesd, the expression passed to \eval is not a language expression or a symbol but a floating point, or a string for instance.
+In that case, we did not track the provenance.
+ % Explain also why we do not show [[ or $ in the table.
+ % Explain the functions that we did not explain before.
 \begin{wraptable}{r}{5cm}\small\centering
@@
 		\packageProvenanceNamec & \packageNbProvenanceSitesc & \packagePercentProvenanceSitesc \\
-		\packageProvenanceNamed & \packageNbProvenanceSitesd & \packagePercentProvenanceSitesd \\
+		%\packageProvenanceNamed & \packageNbProvenanceSitesd & %\packagePercentProvenanceSitesd \\
 		\packageProvenanceNamee & \packageNbProvenanceSitese & \packagePercentProvenanceSitese \\
@@
 		\packageProvenanceNamei & \packageNbProvenanceSitesi & \packagePercentProvenanceSitesi \\
-		\packageProvenanceNamej & \packageNbProvenanceSitesj & \packagePercentProvenanceSitesj \\
-		\packageProvenanceNamek & \packageNbProvenanceSitesk & \packagePercentProvenanceSitesk \\
+		%\packageProvenanceNamej & \packageNbProvenanceSitesj & %\packagePercentProvenanceSitesj \\%[[
+		%\packageProvenanceNamek & \packageNbProvenanceSitesk & %\packagePercentProvenanceSitesk \\%$
 		\packageProvenanceNamel & \packageNbProvenanceSitesl & \packagePercentProvenanceSitesl \\
-		\packageProvenanceNamem & \packageNbProvenanceSitesm & \packagePercentProvenanceSitesm \\
+		%\packageProvenanceNamem & \packageNbProvenanceSitesm & %\packagePercentProvenanceSitesm \\
 		\packageProvenanceNamen & \packageNbProvenanceSitesn & \packagePercentProvenanceSitesn \\
@@
 	\end{tabular}
-	\caption{Provenances in number of sites (85\% first sites).} \label{tab:detailed_provenances}
+	\caption{Provenances in number of sites.} \label{tab:detailed_provenances}
 \end{wraptable}
-
-\ \NOTE{I only see 70 sites with formulas (using the class of the expression) so it does not appear here.}
 
@@
 \item {\it Environment: }   A symbol is created with \c{as.name} or \c{as.symbol}, typically to find the value of a symbol in a non-local environment.
+\item {\it External: } a call to a C function. % Put an example
 \end{compactitem}
+
+\NOTE{I only see 70 sites with formulas (using the class of the expression) so it does not appear here.}
+
+\NOTE{I can also put the absolute number of provenances, \ie if an expression has multiple provenances, we do not keep only the representative but also the other provenances.
+Also, if one site sometimes gets match.call and sometimes substitute, it is not going to be counted in either of them currently.}
 
@@
 Environment & \packageNbSymbolSites & \packageSymbolSitesPercent \\
+External & \packageNbExternalSites & \packageExternalSitesPercent \\
 String & \packageNbStringSites & \packageStringSitesPercent \\\hline

### 2021-08-09 paper/main.tex
More Kaggle removal

@@
 
-\paragraph{Consistency} It is interesting how many different
-forms any given site sees. The more forms, the harder it is to characterize
-behavior at that site. Luckily, \packageNbOneMinimizedPercent of sites only see
-a single form. In other words, a compiler could correctly predict the
-\emph{shape} of an expression by looking just at the first call in
-\packageNbOneMinimizedPercent of the sites. Few sites are highly polymorphic (8
-or more different forms). These include the pipe operator of the \c{magrittr}
-package, used to compose functions.
+\paragraph{Consistency} It is interesting how many different forms any given
+site sees. The more forms, the harder it is to characterize behavior at that
+site. Luckily, \packageNbOneMinimizedPercent of sites only see a single form.
+In other words, a compiler could correctly predict the \emph{shape} of an
+expression by looking just at the first call in \packageNbOneMinimizedPercent
+of the sites. Few sites are highly polymorphic (8 or more different forms).
+These include the pipe operator of the \c{magrittr} package, used to compose
+functions, \eg instead of \c{f(g(h(x),y),z)} one can write \c{h(x) \%>\% g(y) \%>\%
+f(z)}.
 
@@
 straightforward and more predictable as reported by \citet{oopsla12b}, with
-98.7\% of the sites with only one form. Nevertheless, simple forms dominate in R
-also, and there are many cases where \eval could be replaced by less powerful
-constructs.
+98.7\% of the sites with only one form. Nevertheless, simple forms dominate
+in R. Among these there are many cases where \eval could be replaced by less
+powerful constructs such as \c{get} to lookup a name in an environment,
+\c{assign} to set a value to a name in an environment or \c{call} to
+reflectively call a function.
+% TODO: how many receive just the simple form
 
@@
 identified one function \c{invokeRestartInteractively} that prompts the user for
-input, parses it, and passes it to \eval. The use of strings seems to correlate
-with less sophisticated programmers; in Kaggle,
-\kaggleParseExprsSitePercent of sites use strings.
+input, parses it, and passes it to \eval.
 
@@
 
-\Eval may peform side-effects. From a compiler's perspective, we care about
+\Eval may perform side-effects. From a compiler's perspective, we care about
 effects that can be observed --- \ie variable definitions, updates, and removals

### 2021-08-09 paper/main.tex
Formatting changes in 3.1

@@
 \emph{Base} libraries bundled with R, packages hosted on \emph{CRAN}, and
-end-user scripts from \emph{Kaggle}. We consider base libraries as part of the
-language implementation and do not include them in our analysis.
-
-\mypara{Base.} The \BasePackages base libraries provide arithmetics, statistics,
-and operating system functionalities. Base has \BaseFunsWithEvals functions with
-\BaseEvalCallSites calls to \eval. These functions perform key tasks such as
-package loading, there is thus hardly any R program that does not invoke them.
-
-\mypara{CRAN.} Packages hosted on the Comprehensive R Archive Network ({\small
-\url{cran.r-project.org}}, aka CRAN) have three sources of runnable: unit
-\emph{tests} for individual functions, code \emph{examples} embedded in the
-documentation, and use-cases called \emph{vignettes}. All runnable snippets
-can be extracted into independent files. We download \CranPackages packages and
-extract \CranRunnableScripts scripts with \CranRunnableCode lines of code
-(\CranRunnableCodeExamplesRnd in examples, \CranRunnableCodeVignettesRnd in
-vignettes and \CranRunnableCodeTestsRnd in tests). CRAN packages contain 20
-years of contribution from thousands from thousands of authors, often
-experienced R developers.
-
-\mypara{Kaggle.} Kaggle ({\small \url{kaggle.com}}) is an online platform that
-allows users to submit and compete to solve problems. We download \KaggleUnique
-unique scripts and notebooks with their input files (\KaggleDuplicates
-duplicates were removed using SHA-1 hashes); 665K lines of R code making only
-\KaggleWithEvals calls \eval. Authors of these scripts have highly variable
-levels of expertise.
+end-user scripts from \emph{Kaggle}:
+
+\begin{itemize}[$-$]
+
+\item \emph{Base.} The \BasePackages base libraries provide arithmetics,
+  statistics, and operating system functionalities. Base has \BaseFunsWithEvals
+  functions with \BaseEvalCallSites calls to \eval. These functions perform key
+  tasks such as package loading, there is thus hardly any R program that does
+  not invoke them. We consider base libraries as part of the language
+  implementation and do not include them in our analysis.
+
+\item \emph{CRAN.} Packages hosted on the Comprehensive R Archive Network
+  ({\small \url{cran.r-project.org}}, aka CRAN) have three sources of runnable:
+  unit \emph{tests} for individual functions, code \emph{examples} embedded in
+  the documentation, and use-cases called \emph{vignettes}. All runnable
+  snippets can be extracted into independent files. We download \CranPackages
+  packages and extract \CranRunnableScripts scripts with \CranRunnableCode
+  lines of code (\CranRunnableCodeExamplesRnd in examples,
+  \CranRunnableCodeVignettesRnd in vignettes and \CranRunnableCodeTestsRnd in
+  tests). CRAN packages contain 20 years of contribution from thousands from
+  thousands of authors, often experienced R developers.
+
+\item \emph{Kaggle.} Kaggle ({\small \url{kaggle.com}}) is an online platform
+  that allows users to submit and compete to solve problems. We download
+  \KaggleUnique unique scripts and notebooks with their input files
+  (\KaggleDuplicates duplicates were removed using SHA-1 hashes); 665K lines of
+  R code making only \KaggleWithEvals calls \eval. Authors of these scripts
+  have highly variable levels of expertise.
+
+\end{itemize}
 
@@
 
-\begin{enumerate}
+\begin{compactenum}[(1)]
 \item \emph{Download.} Packages are downloaded from CRAN, for Kaggle a web
@@
   paper.
-\end{enumerate}
+\end{compactenum}
 

### 2021-08-09 paper/main.tex
Fixes in 5.4
Fix code formatting

@@
 \makeatletter\let\expandableinput\@@input\makeatother
+
+%% https://www.davehofmann.de/defining-custom-language-templates-for-latex-listings/
+% Define Language
+\lstdefinelanguage{smalleR} {
+  % list of keywords
+  morekeywords={
+    for,
+    if,
+    else,
+    function
+  },
+  sensitive=true, % keywords are not case-sensitive
+  morecomment=[l]{\#}, % l is for line comment
+  morestring=[b]{"} % defines that strings are enclosed in double quotes
+}
+
+\lstset{
+  language={smalleR},
+  columns=flexible,
+  captionpos=b,
+  frame=single,
+  framerule=0pt,
+  framexleftmargin=1mm,
+  framexrightmargin=1mm,
+  tabsize=2,
+  belowskip=0pt,
+  basicstyle=\small\ttfamily,
+  backgroundcolor=\color{LightGray},
+  emphstyle=\sffamily,
+  keywordstyle=\bfseries,
+  commentstyle=\color{Gray}\em,
+  stringstyle=\color{Gray},
+  alsoletter={., _, $},
+  breaklines=true
+}
 
@@
 
-From our 98K
-%\packageNbrunsRnd
-programs, we capture
-14M
-%\SEAllRnd
-side-effects from
+From our \packageNbrunsRnd programs, we capture \SEAllRnd side-effects from
 \SEAllCallsRnd \eval calls (\SEUserCallsToAllRatio) in \SEAllSites sites
@@
 %
-Removing those leaves us with \SEUserRnd side effects from \SEUserCallsRnd \eval
-calls in \SEUserSites sites. These sites are part of \SEUserFunctions R
+Removing those leaves us with \SEUserRnd side effects from \SEUserCallsRnd
+\eval calls in \SEUserSites sites. These sites are part of \SEUserFunctions R
 functions (\SEUserFunctionsToAllRatio of all functions doing \eval) in
@@
 that over half of them happen in a predictable environment.
+% TODO connect the minimization group and side-effects
 

### 2021-08-09 paper/main.tex
Move subsections to Discussion section
Fix code macro and improve presentation of Simplify Interfaces section

@@
 \newcommand{\todo}[1]{\authorcomment{red}{TODO}{#1}}
-
+\newcommand{\code}[1]{\lstinline |#1|\xspace}
+%\renewcommand{\c}[1]{\lstinline |#1|\xspace}
 % cf. https://tex.stackexchange.com/a/144640
@@
 %(with substitute or match.call), Non-Standard Evaluation.}
+To understand the how R developers use \eval, we manually inspected top 117 eval
+call-sites which contribute to XX\% of the eval calls. We describe the common
+usage patterns below.
 
 \subsection{Simplify Interfaces}
-Many packages use the capability to parse and evaluate expressions in custom
-environments to design simpler interfaces. For example, the \c{AdapSamp} package
-provides functions that implement sampling algorithms. These functions take a
-\c{formula} argument, an expression in \c{x} as a string, that represents the
-kernel of the target distribution. This is turned into a closure using
-\c{function(x){eval(parse(text=formula))}}. While the \c{formula} supplied by
-the user could be a higher-order function, passing \c{"exp(-x\^2/2)"} as an
-argument is simpler for domain experts compared to \c{function(x) exp(-x\^2/2)}.
-The \c{bain::constraint\_to\_row} function has a similar pattern of \eval usage.
-It takes a constraint equation as a string argument, and, parses and evaluates
-it in a custom environment with the expected bindings. The
-\c{ctsem::ctModelTransformsToNum} function similarly evaluates a vector of
-formulae represented as a string vector.
+
+A common use of \eval is to \emph{simplify interfaces} by letting users pass string
+argument which can be parsed and evaluated in custom a environment. An example
+is the \code{rARS} function from the \code{AdapSamp} package, shown below.
+
+\begin{lstlisting}
+rARS <- function(n, formula, min=-Inf, max=Inf, sp) {
+  ...
+  p <- function(x) { eval(parse(text=formula)) }
+  ...
+}
+
+\end{lstlisting}
+The \code{rARS} function take a \code{formula} argument, an expression in
+\code{x} as a string, that represents the kernel of the target distribution.
+This is turned into the closure \code{function(x) \{ eval(parse(text=formula))
+  \}}. An alternative design would be to supply the closure itself as the
+argument, but passing \code{function(x) exp(-x^2/2)} is probably not as
+convenient for domain-experts as passing \code{"exp(-x^2/2)"}.
 
@@
 The \eval function is commonly used to customize the evaluation of expressions.
-An example is the \c{assertthat} package's \c{see\_if} function which is used to
+An example is the \code{assertthat} package's \code{see\_if} function which is used to
 assert that the sequence of conditions supplied as arguments are true. It
-evaluates the condition expressions one-by-one inside a \c{tryCatch} block and
-produces an error at the first expression that returns \c{FALSE}. If the
+evaluates the condition expressions one-by-one inside a \code{tryCatch} block and
+produces an error at the first expression that returns \code{FALSE}. If the
 expression itself produces an error, the error message is captured by the
-enclosing \c{tryCatch} block and returned. The remaining expressions in the
-sequence remain unevaluated. A similar pattern is used by the \c{checkmate}
+enclosing \code{tryCatch} block and returned. The remaining expressions in the
+sequence remain unevaluated. A similar pattern is used by the \code{checkmate}
 package.
@@
 
-The ability to capture the current call as a language object (\c{match.call()}),
-extract the unevaluated expression from function arguments (\c{substitute()})
+The ability to capture the current call as a language object (\code{match.call()}),
+extract the unevaluated expression from function arguments (\code{substitute()})
 and \eval with first class environment makes R the toolkit for building DSLs.
-There are numerous widely spread DSLs used in R. For example, \c{data.table}
+There are numerous widely spread DSLs used in R. For example, \code{data.table}
 package defines a DSL for data frame manipulation using the subset operator
-\c{[}. This allows one to write compact queries such as \c{flights[carrier ==
-  "AA", .N, by = origin]}\footnote{The \c{data.table} subset operation is
-  defined as \c{DT[i, j, k]}, where \c{j} expression is used to calculate on a
-  subset/reorder of rows from \c{i} grouped by \c{k}.} which will compute the
+\code{[}. This allows one to write compact queries such as \code{flights[carrier ==
+  "AA", .N, by = origin]}\footnote{The \code{data.table} subset operation is
+  defined as \code{DT[i, j, k]}, where \code{j} expression is used to calculate on a
+  subset/reorder of rows from \code{i} grouped by \code{k}.} which will compute the
 number of flights done by American Airlines for each origin airport.
-Essentially, it uses \c{substitute} to capture the operand's expressions,
+Essentially, it uses \code{substitute} to capture the operand's expressions,
 transforms then appropriately and finally uses \eval to execute the query.
 
-Another example is the \c{tidyselect} package that defines a DSL for selecting
+Another example is the \code{tidyselect} package that defines a DSL for selecting
 columns in a data frame powering number of data manipulation operation in the
-\c{data.table} main competitor, the \c{dply} package. It uses eval to build an
-interpreter for evaluating column selectors such as \c{select(df,
-starts\_with("length") \& !(name:mass))} which returns all columns in \c{df}
-that starts with the string \c{length} and is not between \c{name} and
-\c{mass} columns.
-
-Another instance is the \emph{symbolic differentiation support} provided by base R package \emph{stats}, with two operators \c{D} and \c{deriv}.
-They support arithmetic operations and functions on real numbers such as \c{sin}, on expressions built with \c{expression()} or \c{parse}.
-In \emph{MCMCglmm} (Monte Carl Markov Chain Generalised Linear Mixed Models), a \c{Dtensor} object is created: it contains expressions that are derived twice (with \c{DD}).
-To get a concrete value, it is evaluated in  dataframe \c{mu} where symbolic variables are bound to actual numbers.
+\code{data.table} main competitor, the \code{dply} package. It uses eval to build an
+interpreter for evaluating column selectors such as \code{select(df,
+starts\_with("length") \& !(name:mass))} which returns all columns in \code{df}
+that starts with the string \code{length} and is not between \code{name} and
+\code{mass} columns.
+
+Another instance is the \emph{symbolic differentiation support} provided by base R package \emph{stats}, with two operators \code{D} and \code{deriv}.
+They support arithmetic operations and functions on real numbers such as \code{sin}, on expressions built with \code{expression()} or \code{parse}.
+In \emph{MCMCglmm} (Monte Carl Markov Chain Generalised Linear Mixed Models), a \code{Dtensor} object is created: it contains expressions that are derived twice (with \code{DD}).
+To get a concrete value, it is evaluated in  dataframe \code{mu} where symbolic variables are bound to actual numbers.
 \begin{lstlisting}
@@
 
-The \c{glue} package provides functions for string interpolation; text enclosed
+The \code{glue} package provides functions for string interpolation; text enclosed
 by c{\{} and c{\}} is evaluated using \eval as an R expression and replaced by
@@
 A common use of \eval is the extraction of unevaluated text of arguments passed
-to \c{...} (vararg). For a non-vararg parameter \c{x}, the expressions
-\c{substitute(x)} can be used to extract the argument text. However, for \c{...}
-arguments, the expression \c{eval(quote(substitute(list(...))))} or its variants
+to \code{...} (vararg). For a non-vararg parameter \code{x}, the expressions
+\code{substitute(x)} can be used to extract the argument text. However, for \code{...}
+arguments, the expression \code{eval(quote(substitute(list(...))))} or its variants
 are used. The \eval call evaluates the quoted expression
-\c{substitute(list(...))}. This returns a \c{list} of unevaluated argument
-expressions. The \c{hasGroupNumber} function from the \c{lattice} package is one
+\code{substitute(list(...))}. This returns a \code{list} of unevaluated argument
+expressions. The \code{hasGroupNumber} function from the \code{lattice} package is one
 of the many users of this pattern.
 
-The \c{withr} package provides the \c{defer} function which defers the
+The \code{withr} package provides the \code{defer} function which defers the
 evaluation of an expression until a stack frame is exited. It attaches this
@@
 property. For R, the data suggests that in some cases, the use of \eval in the
-\c{X, V, \$, <-, FUN} and the function calls related to code generation could
+\code{X, V, \$, <-, FUN} and the function calls related to code generation could
 replaced by more specific functions. For instance, variable lookup in any
-environment can be performed with \c{get}, and assignment in any environment,
-with \c{assign}. Building a call and executing can be done with \c{do.call}.
+environment can be performed with \code{get}, and assignment in any environment,
+with \code{assign}. Building a call and executing can be done with \code{do.call}.
 Nevertheless, many of the observed \eval usage in R, particularly, the ones
@@
 Finally, let us mention an anecdotal use of \eval as a means to bypass some
-checks performed by the \c{R CMD CHECK} tool. This tool performs a number of
+checks performed by the \code{R CMD CHECK} tool. This tool performs a number of
 well-formedness rules before accepting a package to the CRAN repository. Many
@@
 CRAN that would otherwise be refused. For example,  it checks that packages
-do not call the \c{unlockBinding} function.\footnote{It allows one to open and
-modify loaded package namespaces.} Packages such as \c{data.table} circumvent
+do not call the \code{unlockBinding} function.\footnote{It allows one to open and
+modify loaded package namespaces.} Packages such as \code{data.table} circumvent
 such restrictions by using \eval as the check is only performed statically.
+
+\section{Discussion}
 
@@
 \eval can be replaced with alternate functions yielding the same results. For
-example, the \c{PerformanceAnalytics} package contains a function
-\c{chart.QQPlot} that uses \eval to resolve a string into function and another
+example, the \code{PerformanceAnalytics} package contains a function
+\code{chart.QQPlot} that uses \eval to resolve a string into function and another
 to call it and assign its results into a variable:
@@
 \end{lstlisting}
-  or even to a one-liner \c{do.call(paste0("q",d), c(dp, list(...)))}.
-
-  The \c{copyslots} function from the \c{coin} package uses \eval to copy the
-  slots from one \c{S4} object to another. It performs
-  \c{eval(str2lang(paste0("target@", s, " <- source@", s)))} for each slot name
-  \c{s}. This can be replaced by the expression \c{slot(target, s) <- slot(source,
-    s)} which uses the \c{slot<-} and \c{slot} functions to assign and retrieve
+  or even to a one-liner \code{do.call(paste0("q",d), c(dp, list(...)))}.
+
+  The \code{copyslots} function from the \code{coin} package uses \eval to copy the
+  slots from one \code{S4} object to another. It performs
+  \code{eval(str2lang(paste0("target@", s, " <- source@", s)))} for each slot name
+  \code{s}. This can be replaced by the expression \code{slot(target, s) <- slot(source,
+    s)} which uses the \code{slot<-} and \code{slot} functions to assign and retrieve
   the slots by name.
 
-  Another examples is in the \c{configr::config.funs.par} function whose
+  Another examples is in the \code{configr::config.funs.par} function whose
   simplified definition is shown below.
@@
 
-  The function appears to use \eval since the function name \c{fun} is received
+  The function appears to use \eval since the function name \code{fun} is received
   as a string. This string can be resolved to the function object using the
-  \c{get} function and the contents of \c{text.1}, \c{text.2}, and \c{text.3}
+  \code{get} function and the contents of \code{text.1}, \code{text.2}, and \code{text.3}
   can be written directly without the need to convert them to text for

### 2021-08-09 paper/main.tex
Fix intro
Fix description of assertion packages

@@
 
-A common use of \eval is to \emph{simplify interfaces} by letting users pass string
-argument which can be parsed and evaluated in custom a environment. An example
-is the \code{rARS} function from the \code{AdapSamp} package, shown below.
+A common use of \eval is to \emph{simplify interfaces} by letting users pass
+string argument which can be parsed and evaluated in custom a environment. An
+example is the \code{rARS} function shown below. This function is defined in the
+\code{AdapSamp} package that implements sampling algorithms.
 
@@
 
-\subsection{Custom Evaluation}
-The \eval function is commonly used to customize the evaluation of expressions.
-An example is the \code{assertthat} package's \code{see\_if} function which is used to
-assert that the sequence of conditions supplied as arguments are true. It
-evaluates the condition expressions one-by-one inside a \code{tryCatch} block and
-produces an error at the first expression that returns \code{FALSE}. If the
-expression itself produces an error, the error message is captured by the
-enclosing \code{tryCatch} block and returned. The remaining expressions in the
-sequence remain unevaluated. A similar pattern is used by the \code{checkmate}
-package.
+Assertion packages such as \code{assertthat} provide functions to check pre and
+post conditions and produce friendly error messages on failures. These
+conditions are written as expressions which are evaluated using \eval. Consider
+the \code{see_\if} function shown below. It accepts the conditions as arguments
+to the \code{...} parameter. The code of these conditions is extracted using
+\code{eval(substitute(alist(...)))}. The conditions are then evaluated
+sequentially by the expression \code{eval(assertion, env)}, inside a
+\code{tryCatch} block. If a condition evaluates to \code{FALSE}, the function
+returns immediately with an error message, leaving the remaining conditions
+unevaluated.
+
+\begin{lstlisting}
+see_if <- function(..., env = parent.frame(), msg = NULL) {
+    asserts <- eval(substitute(alist(...)))
+    for (assertion in asserts) {
+      res <- tryCatch({
+        eval(assertion, env)
+      }, assertError = function(e) {
+        structure(FALSE, msg = e$message)
+      })
+    ...
+}
+
+\end{lstlisting}
 

### 2021-08-09 paper/main.tex
Explain ... pattern

@@
 
+An unexpected use of \eval is the extraction of unevaluated text of arguments
+passed to a \code{...} parameter. Consider the definition of the \code{NVL}
+function from the \code{statnet.common} package. This function returns the first
+argument that is non \code{NULL}.
+
+\begin{lstlisting}
+NVL <- function(...) {
+    for(e in eval(substitute(alist(...))))
+        x <- eval(e, parent.frame())
+        if(!is.null(x)) break
+    }
+    x
+}
+\end{lstlisting}
+The expression \code{eval(substitute(alist(...)))} is used to extract the
+unevaluated text of the arguments passed to \code{NVL}.
+\code{substitute(alist(...))} returns an expression representing a call to the
+\code{alist} function with \code{...} replaced by the arguments supplied to
+\code{NVL}. Evaluating this expression returns a \code{list} whose arguments are
+the unevaluated argument expressions which are evaluated sequentially for their
+result.
+
 Assertion packages such as \code{assertthat} provide functions to check pre and
@@
 conditions are written as expressions which are evaluated using \eval. Consider
-the \code{see_\if} function shown below. It accepts the conditions as arguments
+the \code{see\_if} function shown below. It accepts the conditions as arguments
 to the \code{...} parameter. The code of these conditions is extracted using
@@
 \subsection{Miscellaneous}
-A common use of \eval is the extraction of unevaluated text of arguments passed
-to \code{...} (vararg). For a non-vararg parameter \code{x}, the expressions
-\code{substitute(x)} can be used to extract the argument text. However, for \code{...}
-arguments, the expression \code{eval(quote(substitute(list(...))))} or its variants
-are used. The \eval call evaluates the quoted expression
-\code{substitute(list(...))}. This returns a \code{list} of unevaluated argument
-expressions. The \code{hasGroupNumber} function from the \code{lattice} package is one
-of the many users of this pattern.
 

### 2021-08-09 paper/main.tex
Add do.call and %||%

@@
 
+Similar idea is used by the \code{igraph} package to define the \code{\%\|\|\%}
+operator whose code is shown below. This operator returns the result of
+evaluating the first argument (\code{lhs}) if it is not \code{NULL}, otherwise
+it returns the result of evaluating the second argument (\code{rhs}). This is
+similar to the \code{\|\|} operator which performs the logical OR of its
+arguments except it treats \code{NULL} as the \code{FALSE} value.
+
+\begin{lstlisting}
+`%||%` <- function (lhs, rhs) {
+    lres <- withVisible(eval(lhs, envir = parent.frame()))
+    if (is.null(lres$value)) {
+        eval(rhs, envir = parent.frame())
+    } else {
+        if (lres$visible) { lres$value}
+        else { invisible(lres$value) }
+   }
+}
+\end{lstlisting}
+
+Many packages use \eval to implement variations of the \code{base} R functions.
+For example, the \code{igraph} package provides the \code{do_call} function
+shown below. This is a variant of the \code{do.call} function from the
+\code{base} package. This function executes a function call by constructs it
+from a name and a list of arguments and calling \eval.
+\begin{lstlisting}
+do_call <- function(f, ..., .args = list(), .env = parent.frame()) {
+  f <- substitute(f)
+  call <- make_call(f, ..., .args)
+  eval(call, .env)
+}
+\end{lstlisting}
+
 \subsection{Domain Specific Language}
@@
 \end{lstlisting}
-
-\subsection{Miscellaneous}
-
-The \code{withr} package provides the \code{defer} function which defers the
-evaluation of an expression until a stack frame is exited. It attaches this
-expression as an attribute to the frame environment and uses \eval to evaluate
-it on frame exit.
 

### 2021-08-09 analysis/normalized.Rmd
New tags for normalized expressions

@@
     func = expr_resolved_function, # the function (if any) at the root of the AST of expr_resolved
-    parsed = expr_parsed_expression,
+    #parsed = expr_parsed_expression,
     hash = expr_resolved_hash,
-    match.call = expr_match_call,
+    #match.call = expr_match_call,
     arg_size = expr_resolved_length, #   Size in characters of argument 
@@
 
-  E_cano %>% write_fst(min_path)
+  #E_cano %>% write_fst(min_path)
   sz <- file.size(calls_path) %>% as_tibble()
-  write.fst(sz, lock_path)
+  #write.fst(sz, lock_path)
 }
@@
 Looking up to a variable and applying a function to a variable, are the two non-trivial categories that most of the time explicitly use the `envir` argument of `eval`.
+
+```{r}
+simple_minimized <- E_cano %>% filter(type != "bytecode") %>%
+  mutate(minimized = if_else(minimized == "<-", "<", minimized)) %>%
+  group_by(src_ref) %>%
+  filter(all(minimized %in% c("X", "V", "$", "<"))) %>%
+  summarise(minimized = paste0(sort(unique(minimized)), collapse = ""), length = str_length(minimized))
+
+simple_minimized_one <- simple_minimized %>% filter(length == 1) %>% nrow()
+simple_minimized_more <- simple_minimized %>% filter(length > 1) %>% nrow()
+
+r("nb simple minimized one", simple_minimized_one)
+r("nb simple minimized more", simple_minimized_more)
+```
 
@@
 ```{r}
-nb_mf_match.call <- modelframes %>% filter(!is.na(match.call)) %>% pull(src_ref) %>% n_distinct()
+#no more match.call column
+#nb_mf_match.call <- modelframes %>% filter(!is.na(match.call)) %>% pull(src_ref) %>% n_distinct()
+nb_mf_match.call <- modelframes
 ```
@@
 ```{r}
-match.call_df <- E_cano %>% filter(!is.na(match.call))
+# No more match.call column
+#match.call_df <- E_cano %>% filter(!is.na(match.call))
+match.call_df <- E_cano
 
@@
 
-```{r, eval=dataset_name == "package"}
+```{r, eval=dataset_name == "package", eval = FALSE}
 manual_value_necessary <- read_csv(path(params$base_dir, "non-necessary-value-evals.csv"), col_types = cols(src_ref = col_character(), necessary = col_logical()))

### 2021-08-09 paper/main.tex
Add code generation and give names to sections

@@
 To understand the how R developers use \eval, we manually inspected top 117 eval
-call-sites which contribute to XX\% of the eval calls. We describe the common
+call-sites which contribute to 90\% of the eval calls. We describe the common
 usage patterns below.
 
-\subsection{Simplify Interfaces}
-
-A common use of \eval is to \emph{simplify interfaces} by letting users pass
-string argument which can be parsed and evaluated in custom a environment. An
-example is the \code{rARS} function shown below. This function is defined in the
-\code{AdapSamp} package that implements sampling algorithms.
+\emph{Simplify Interfaces}. A common use of \eval is to simplify interfaces by
+letting users pass string argument which can be parsed and evaluated in custom a
+environment. An example is the \code{rARS} function shown below. This function
+is defined in the \code{AdapSamp} package that implements sampling algorithms.
 
@@
 }
-
-\end{lstlisting}
+\end{lstlisting}
+
 The \code{rARS} function take a \code{formula} argument, an expression in
@@
 
-An unexpected use of \eval is the extraction of unevaluated text of arguments
+\emph{Code Generation.} The \c{Rcpp} package which bridges R with C++ is a
+popular user of this pattern. Its \code{.makeCppMethods} function, shown below,
+uses \eval to automatically generate R bindings to native code. The expression
+\code{function(...) .CppObject$WHAT(...), list(WHAT = as.name(what))} is
+evaluated with different values of \code{what} to generate wrapper functions
+which are assigned to the \code{methods} structure keyed by \code{what}.
+
+\begin{lstlisting}
+.makeCppMethods <- function(methods, cppclassinfo, env) {
+    cppMethods <- names(cppclassinfo@methods)
+    newMethods <- names(methods)
+    for(what in cppMethods[! cppMethods %in% newMethods])
+        methods[[what]] <- eval(substitute(
+                  function(...) .CppObject$WHAT(...), list(WHAT = as.name(what))),
+                                env)
+    methods
+}
+\end{lstlisting}
+
+\emph{Extracting ... arguments.} A surprising use of \eval is the extraction of unevaluated text of arguments
 passed to a \code{...} parameter. Consider the definition of the \code{NVL}
@@
 
-Assertion packages such as \code{assertthat} provide functions to check pre and
-post conditions and produce friendly error messages on failures. These
-conditions are written as expressions which are evaluated using \eval. Consider
-the \code{see\_if} function shown below. It accepts the conditions as arguments
-to the \code{...} parameter. The code of these conditions is extracted using
-\code{eval(substitute(alist(...)))}. The conditions are then evaluated
-sequentially by the expression \code{eval(assertion, env)}, inside a
-\code{tryCatch} block. If a condition evaluates to \code{FALSE}, the function
-returns immediately with an error message, leaving the remaining conditions
-unevaluated.
+\emph{Conditional Evaluation.} Developers frequently use \eval to explicitly
+control the evaluation of code. Consider the \code{\%\|\|\%} operator from the
+\code{igraph} package shown below. This operator returns the result of
+evaluating the first argument (\code{lhs}) if it is not \code{NULL}, otherwise
+it returns the result of evaluating the second argument (\code{rhs}). This is
+similar to the \code{\|\|} operator which performs the logical OR of its
+arguments except it treats \code{NULL} as the \code{FALSE} value.
+
+\begin{lstlisting}
+`%||%` <- function (lhs, rhs) {
+    lres <- withVisible(eval(lhs, envir = parent.frame()))
+    if (is.null(lres$value)) {
+        eval(rhs, envir = parent.frame())
+    } else {
+        if (lres$visible) { lres$value}
+        else { invisible(lres$value) }
+   }
+}
+\end{lstlisting}
+
+A similar idea is used by assertion packages such as \code{assertthat} which
+provide functions to check pre and post conditions and produce friendly error
+messages on failures. These conditions are written as expressions which are
+evaluated using \eval. Consider the \code{see\_if} function shown below. It
+accepts the conditions as arguments to the \code{...} parameter. The code of
+these conditions is extracted using \code{eval(substitute(alist(...)))}. The
+conditions are then evaluated sequentially by the expression
+\code{eval(assertion, env)}, inside a \code{tryCatch} block. If a condition
+evaluates to \code{FALSE}, the function returns immediately with an error
+message, leaving the remaining conditions unevaluated.
 
@@
 
-Similar idea is used by the \code{igraph} package to define the \code{\%\|\|\%}
-operator whose code is shown below. This operator returns the result of
-evaluating the first argument (\code{lhs}) if it is not \code{NULL}, otherwise
-it returns the result of evaluating the second argument (\code{rhs}). This is
-similar to the \code{\|\|} operator which performs the logical OR of its
-arguments except it treats \code{NULL} as the \code{FALSE} value.
-
-\begin{lstlisting}
-`%||%` <- function (lhs, rhs) {
-    lres <- withVisible(eval(lhs, envir = parent.frame()))
-    if (is.null(lres$value)) {
-        eval(rhs, envir = parent.frame())
-    } else {
-        if (lres$visible) { lres$value}
-        else { invisible(lres$value) }
-   }
-}
-\end{lstlisting}
-
-Many packages use \eval to implement variations of the \code{base} R functions.
-For example, the \code{igraph} package provides the \code{do_call} function
-shown below. This is a variant of the \code{do.call} function from the
-\code{base} package. This function executes a function call by constructs it
-from a name and a list of arguments and calling \eval.
+\textit{Core function variants.} Many packages use \eval to implement variations of the
+\code{base} R functions. For example, the \code{igraph} package provides the
+\code{do_call} function shown below. This is a variant of the \code{do.call}
+function from the \code{base} package. This function executes a function call by
+constructs it from a name and a list of arguments and calling \eval.
 \begin{lstlisting}

### 2021-08-09 paper/main.tex
Move Shapes Redux to end and add Code Obfuscation section

@@
 The R language was intended to be extensible. The combination of lazy
-evaluation, \c{substitute} and \c{eval} are the tools given to developers to
-this end. In this section, we focus on the \emph{why}, giving real-world
-examples of the \eval usage in the wild. All the examples are based on code from
-our corpus, often simplified to fit the space and increase clarity. For each
-example, we indicate the package and the function where it is located.
+evaluation, \c{substitute} and \c{eval} are some of the tools given to
+developers to this end. To understand \emph{why} developers use \eval, we
+manually inspected top 117 \eval call-sites which contribute to 90\% of the
+\eval calls. In this section, we give examples of the \eval usage from this
+real-world sample. The examples are often simplified to fit the space and
+increase clarity. For each example, we indicate the package and the function
+where it is located.
+
+\emph{Simplify Interfaces}. A common use of \eval is to simplify interfaces by
+letting users pass string argument which can be parsed and evaluated in custom a
+environment. An example is the \code{rARS} function shown below. This function
+is defined in the \code{AdapSamp} package that implements sampling algorithms.
+
+\begin{lstlisting}
+rARS <- function(n, formula, min=-Inf, max=Inf, sp) {
+    ...
+    p <- function(x) { eval(parse(text=formula)) }
+    ...
+}
+\end{lstlisting}
+
+The \code{rARS} function take a \code{formula} argument, an expression in
+\code{x} as a string, that represents the kernel of the target distribution.
+This is turned into the closure \code{function(x) \{ eval(parse(text=formula))
+  \}}. An alternative design would be to supply the closure itself as the
+argument, but passing \code{function(x) exp(-x^2/2)} is probably not as
+convenient for domain-experts as passing \code{"exp(-x^2/2)"}.
+
+\emph{Code Generation.} The \c{Rcpp} package which bridges R with C++ is a
+popular user of this pattern. Its \code{.makeCppMethods} function, shown below,
+uses \eval to automatically generate R bindings to native code. The expression
+\code{function(...) .CppObject$WHAT(...), list(WHAT = as.name(what))} is
+evaluated with different values of \code{what} to generate wrapper functions
+which are assigned to the \code{methods} structure keyed by \code{what}.
+
+\begin{lstlisting}
+.makeCppMethods <- function(methods, cppclassinfo, env) {
+    cppMethods <- names(cppclassinfo@methods)
+    newMethods <- names(methods)
+    for(what in cppMethods[! cppMethods %in% newMethods])
+        methods[[what]] <- eval(substitute(
+                  function(...) .CppObject$WHAT(...), list(WHAT = as.name(what))),
+                                env)
+    methods
+}
+\end{lstlisting}
+
+\emph{Extracting ... arguments.} A surprising use of \eval is the extraction of unevaluated text of arguments
+passed to a \code{...} parameter. Consider the definition of the \code{NVL}
+function from the \code{statnet.common} package. This function returns the first
+argument that is non \code{NULL}.
+
+\begin{lstlisting}
+NVL <- function(...) {
+    for(e in eval(substitute(alist(...))))
+        x <- eval(e, parent.frame())
+        if(!is.null(x)) break
+    }
+    x
+}
+\end{lstlisting}
+The expression \code{eval(substitute(alist(...)))} is used to extract the
+unevaluated text of the arguments passed to \code{NVL}.
+\code{substitute(alist(...))} returns an expression representing a call to the
+\code{alist} function with \code{...} replaced by the arguments supplied to
+\code{NVL}. Evaluating this expression returns a \code{list} whose arguments are
+the unevaluated argument expressions which are evaluated sequentially for their
+result.
+
+\emph{Conditional Evaluation.} Developers frequently use \eval to explicitly
+control the evaluation of code. Consider the \code{\%\|\|\%} operator from the
+\code{igraph} package shown below. This operator returns the result of
+evaluating the first argument (\code{lhs}) if it is not \code{NULL}, otherwise
+it returns the result of evaluating the second argument (\code{rhs}). This is
+similar to the \code{\|\|} operator which performs the logical OR of its
+arguments except it treats \code{NULL} as the \code{FALSE} value.
+
+\begin{lstlisting}
+`%||%` <- function (lhs, rhs) {
+    lres <- withVisible(eval(lhs, envir = parent.frame()))
+    if (is.null(lres$value)) {
+        eval(rhs, envir = parent.frame())
+    } else {
+        if (lres$visible) { lres$value}
+        else { invisible(lres$value) }
+   }
+}
+\end{lstlisting}
+
+A similar idea is used by assertion packages such as \code{assertthat} which
+provide functions to check pre and post conditions and produce friendly error
+messages on failures. These conditions are written as expressions which are
+evaluated using \eval. Consider the \code{see\_if} function shown below. It
+accepts the conditions as arguments to the \code{...} parameter. The code of
+these conditions is extracted using \code{eval(substitute(alist(...)))}. The
+conditions are then evaluated sequentially by the expression
+\code{eval(assertion, env)}, inside a \code{tryCatch} block. If a condition
+evaluates to \code{FALSE}, the function returns immediately with an error
+message, leaving the remaining conditions unevaluated.
+
+\begin{lstlisting}
+see_if <- function(..., env = parent.frame(), msg = NULL) {
+    asserts <- eval(substitute(alist(...)))
+    for (assertion in asserts) {
+      res <- tryCatch({
+        eval(assertion, env)
+      }, assertError = function(e) {
+        structure(FALSE, msg = e$message)
+      })
+    ...
+}
+
+\end{lstlisting}
+
+\emph{Core function variants.} Many packages use \eval to implement variations of the
+\code{base} R functions. For example, the \code{igraph} package provides the
+\code{do_call} function shown below. This is a variant of the \code{do.call}
+function from the \code{base} package. This function executes a function call by
+constructs it from a name and a list of arguments and calling \eval.
+
+\begin{lstlisting}
+do_call <- function(f, ..., .args = list(), .env = parent.frame()) {
+  f <- substitute(f)
+  call <- make_call(f, ..., .args)
+  eval(call, .env)
+}
+\end{lstlisting}
+
+\emph{Domain-Specific Languages.} The combination of laziness, metaprogramming,
+first-class environments and dynamic evaluation make R the perfect toolkit for
+building DSLs. An example of this is the \code{glue} package which provides
+functions for string interpolation. The code snippet below shows an example of
+this.
+
+\begin{lstlisting}
+library(glue)
+greeting <- "Hello"
+glue('{greeting} World!')
+#> Hello World!
+\end{lstlisting}
+
+text enclosed by c{\{} and c{\}} is
+evaluated using \eval as an R expression and replaced by its result.
+
+\emph{Remove Boilerplate}
+
+For example, the PopED::get_rse uses it to assign a number of variables in the current scope: eval(parse(text=paste(default_args[[i]]), "← ", i)).
+There are also more sophisticated cases.
+The plyr::allocate_column function uses eval for a deferred assignment to fill missing values in a column.
+Package overture repeatedly evaluates R expressions capturing any assignments that happen in it as samples of Markov chains.
+
+\emph{Code Obfuscation} Some packages use \eval to bypass the restrictions
+imposed by the \code{R CMD CHECK} tool. This tool enforces some well-formedness
+rules on a submitted package before accepting it for inclusion in CRAN. Some
+static checks ensure that the package code does not use certain restricted
+functions. If package authors cannot avoid using these functions, they use \eval
+to obfuscate the code, bypassing the checks performed by the tool. An example of
+this use is the \code{scalaEnsure} function from the \code{aibd} package. It
+mutates the variable \code{s} in the package environment. However, the package
+environment in R is locked by default, which prevents bindings from being
+mutated. Environment can be unlocked using \code{unlockBinding}, but its use is
+flagged by the well-formedness checks. To work around this limitation, the
+author passes the expression to unlock the package environment as a string to
+\code{parse} which is then evaluated by \code{eval}.
+
+\begin{lstlisting}
+scalaEnsure <- function() {
+    ...
+    env <- parent.env(environment())
+    eval(parse(text=paste0('unlockBinding("s", env)')))
+    assign("s", s, envir=env)
+    lockBinding("s", env)
+}
+\end{lstlisting}
+
+\emph{Miscellaneous}
+
+\subsection{Domain Specific Language}
+
+The ability to capture the current call as a language object (\code{match.call()}),
+extract the unevaluated expression from function arguments (\code{substitute()})
+and \eval with first class environment makes R the toolkit for building DSLs.
+There are numerous widely spread DSLs used in R. For example, \code{data.table}
+package defines a DSL for data frame manipulation using the subset operator
+\code{[}. This allows one to write compact queries such as \code{flights[carrier ==
+  "AA", .N, by = origin]}\footnote{The \code{data.table} subset operation is
+  defined as \code{DT[i, j, k]}, where \code{j} expression is used to calculate on a
+  subset/reorder of rows from \code{i} grouped by \code{k}.} which will compute the
+number of flights done by American Airlines for each origin airport.
+Essentially, it uses \code{substitute} to capture the operand's expressions,
+transforms then appropriately and finally uses \eval to execute the query.
+
+Another example is the \code{tidyselect} package that defines a DSL for selecting
+columns in a data frame powering number of data manipulation operation in the
+\code{data.table} main competitor, the \code{dply} package. It uses eval to build an
+interpreter for evaluating column selectors such as \code{select(df,
+starts\_with("length") \& !(name:mass))} which returns all columns in \code{df}
+that starts with the string \code{length} and is not between \code{name} and
+\code{mass} columns.
+
+Another instance is the \emph{symbolic differentiation support} provided by base R package \emph{stats}, with two operators \code{D} and \code{deriv}.
+They support arithmetic operations and functions on real numbers such as \code{sin}, on expressions built with \code{expression()} or \code{parse}.
+In \emph{MCMCglmm} (Monte Carl Markov Chain Generalised Linear Mixed Models), a \code{Dtensor} object is created: it contains expressions that are derived twice (with \code{DD}).
+To get a concrete value, it is evaluated in  dataframe \code{mu} where symbolic variables are bound to actual numbers.
+\begin{lstlisting}
+ expr<-expression(beta_1 + time*beta_2+u)
+ mu = data.frame(beta_1=0.5, beta_2=1, time=3, u=2.3)
+ D[i]<-eval(DD(expr, name[unlist(comb.pos[i,])]), mu)
+\end{lstlisting}
 
@@
 %(with substitute or match.call), Non-Standard Evaluation.}
-To understand the how R developers use \eval, we manually inspected top 117 eval
-call-sites which contribute to 90\% of the eval calls. We describe the common
-usage patterns below.
-
-\emph{Simplify Interfaces}. A common use of \eval is to simplify interfaces by
-letting users pass string argument which can be parsed and evaluated in custom a
-environment. An example is the \code{rARS} function shown below. This function
-is defined in the \code{AdapSamp} package that implements sampling algorithms.
-
-\begin{lstlisting}
-rARS <- function(n, formula, min=-Inf, max=Inf, sp) {
-    ...
-    p <- function(x) { eval(parse(text=formula)) }
-    ...
-}
-\end{lstlisting}
-
-The \code{rARS} function take a \code{formula} argument, an expression in
-\code{x} as a string, that represents the kernel of the target distribution.
-This is turned into the closure \code{function(x) \{ eval(parse(text=formula))
-  \}}. An alternative design would be to supply the closure itself as the
-argument, but passing \code{function(x) exp(-x^2/2)} is probably not as
-convenient for domain-experts as passing \code{"exp(-x^2/2)"}.
-
-\emph{Code Generation.} The \c{Rcpp} package which bridges R with C++ is a
-popular user of this pattern. Its \code{.makeCppMethods} function, shown below,
-uses \eval to automatically generate R bindings to native code. The expression
-\code{function(...) .CppObject$WHAT(...), list(WHAT = as.name(what))} is
-evaluated with different values of \code{what} to generate wrapper functions
-which are assigned to the \code{methods} structure keyed by \code{what}.
-
-\begin{lstlisting}
-.makeCppMethods <- function(methods, cppclassinfo, env) {
-    cppMethods <- names(cppclassinfo@methods)
-    newMethods <- names(methods)
-    for(what in cppMethods[! cppMethods %in% newMethods])
-        methods[[what]] <- eval(substitute(
-                  function(...) .CppObject$WHAT(...), list(WHAT = as.name(what))),
-                                env)
-    methods
-}
-\end{lstlisting}
-
-\emph{Extracting ... arguments.} A surprising use of \eval is the extraction of unevaluated text of arguments
-passed to a \code{...} parameter. Consider the definition of the \code{NVL}
-function from the \code{statnet.common} package. This function returns the first
-argument that is non \code{NULL}.
-
-\begin{lstlisting}
-NVL <- function(...) {
-    for(e in eval(substitute(alist(...))))
-        x <- eval(e, parent.frame())
-        if(!is.null(x)) break
-    }
-    x
-}
-\end{lstlisting}
-The expression \code{eval(substitute(alist(...)))} is used to extract the
-unevaluated text of the arguments passed to \code{NVL}.
-\code{substitute(alist(...))} returns an expression representing a call to the
-\code{alist} function with \code{...} replaced by the arguments supplied to
-\code{NVL}. Evaluating this expression returns a \code{list} whose arguments are
-the unevaluated argument expressions which are evaluated sequentially for their
-result.
-
-\emph{Conditional Evaluation.} Developers frequently use \eval to explicitly
-control the evaluation of code. Consider the \code{\%\|\|\%} operator from the
-\code{igraph} package shown below. This operator returns the result of
-evaluating the first argument (\code{lhs}) if it is not \code{NULL}, otherwise
-it returns the result of evaluating the second argument (\code{rhs}). This is
-similar to the \code{\|\|} operator which performs the logical OR of its
-arguments except it treats \code{NULL} as the \code{FALSE} value.
-
-\begin{lstlisting}
-`%||%` <- function (lhs, rhs) {
-    lres <- withVisible(eval(lhs, envir = parent.frame()))
-    if (is.null(lres$value)) {
-        eval(rhs, envir = parent.frame())
-    } else {
-        if (lres$visible) { lres$value}
-        else { invisible(lres$value) }
-   }
-}
-\end{lstlisting}
-
-A similar idea is used by assertion packages such as \code{assertthat} which
-provide functions to check pre and post conditions and produce friendly error
-messages on failures. These conditions are written as expressions which are
-evaluated using \eval. Consider the \code{see\_if} function shown below. It
-accepts the conditions as arguments to the \code{...} parameter. The code of
-these conditions is extracted using \code{eval(substitute(alist(...)))}. The
-conditions are then evaluated sequentially by the expression
-\code{eval(assertion, env)}, inside a \code{tryCatch} block. If a condition
-evaluates to \code{FALSE}, the function returns immediately with an error
-message, leaving the remaining conditions unevaluated.
-
-\begin{lstlisting}
-see_if <- function(..., env = parent.frame(), msg = NULL) {
-    asserts <- eval(substitute(alist(...)))
-    for (assertion in asserts) {
-      res <- tryCatch({
-        eval(assertion, env)
-      }, assertError = function(e) {
-        structure(FALSE, msg = e$message)
-      })
-    ...
-}
-
-\end{lstlisting}
-
-\textit{Core function variants.} Many packages use \eval to implement variations of the
-\code{base} R functions. For example, the \code{igraph} package provides the
-\code{do_call} function shown below. This is a variant of the \code{do.call}
-function from the \code{base} package. This function executes a function call by
-constructs it from a name and a list of arguments and calling \eval.
-\begin{lstlisting}
-do_call <- function(f, ..., .args = list(), .env = parent.frame()) {
-  f <- substitute(f)
-  call <- make_call(f, ..., .args)
-  eval(call, .env)
-}
-\end{lstlisting}
-
-\subsection{Domain Specific Language}
-
-The ability to capture the current call as a language object (\code{match.call()}),
-extract the unevaluated expression from function arguments (\code{substitute()})
-and \eval with first class environment makes R the toolkit for building DSLs.
-There are numerous widely spread DSLs used in R. For example, \code{data.table}
-package defines a DSL for data frame manipulation using the subset operator
-\code{[}. This allows one to write compact queries such as \code{flights[carrier ==
-  "AA", .N, by = origin]}\footnote{The \code{data.table} subset operation is
-  defined as \code{DT[i, j, k]}, where \code{j} expression is used to calculate on a
-  subset/reorder of rows from \code{i} grouped by \code{k}.} which will compute the
-number of flights done by American Airlines for each origin airport.
-Essentially, it uses \code{substitute} to capture the operand's expressions,
-transforms then appropriately and finally uses \eval to execute the query.
-
-Another example is the \code{tidyselect} package that defines a DSL for selecting
-columns in a data frame powering number of data manipulation operation in the
-\code{data.table} main competitor, the \code{dply} package. It uses eval to build an
-interpreter for evaluating column selectors such as \code{select(df,
-starts\_with("length") \& !(name:mass))} which returns all columns in \code{df}
-that starts with the string \code{length} and is not between \code{name} and
-\code{mass} columns.
-
-Another instance is the \emph{symbolic differentiation support} provided by base R package \emph{stats}, with two operators \code{D} and \code{deriv}.
-They support arithmetic operations and functions on real numbers such as \code{sin}, on expressions built with \code{expression()} or \code{parse}.
-In \emph{MCMCglmm} (Monte Carl Markov Chain Generalised Linear Mixed Models), a \code{Dtensor} object is created: it contains expressions that are derived twice (with \code{DD}).
-To get a concrete value, it is evaluated in  dataframe \code{mu} where symbolic variables are bound to actual numbers.
-\begin{lstlisting}
- expr<-expression(beta_1 + time*beta_2+u)
- mu = data.frame(beta_1=0.5, beta_2=1, time=3, u=2.3)
- D[i]<-eval(DD(expr, name[unlist(comb.pos[i,])]), mu)
-\end{lstlisting}
-
-The \code{glue} package provides functions for string interpolation; text enclosed
-by c{\{} and c{\}} is evaluated using \eval as an R expression and replaced by
-its result. The code snippet below shows an example of this.
-
-\begin{lstlisting}
-library(glue)
-greeting <- "Hello"
-glue('{greeting} World!')
-#> Hello World!
-\end{lstlisting}
 
@@
 reported for JavaScript.
-
-Finally, let us mention an anecdotal use of \eval as a means to bypass some
-checks performed by the \code{R CMD CHECK} tool. This tool performs a number of
-well-formedness rules before accepting a package to the CRAN repository. Many
-checks are performed statically by looking at syntactic constructs in the R
-code. Using \eval allows one to bypass such checks and thus submit a package to
-CRAN that would otherwise be refused. For example,  it checks that packages
-do not call the \code{unlockBinding} function.\footnote{It allows one to open and
-modify loaded package namespaces.} Packages such as \code{data.table} circumvent
-such restrictions by using \eval as the check is only performed statically.
 

### 2021-08-09 paper/main.tex
Fix errors

@@
 
-\emph{Remove Boilerplate}
-
-For example, the PopED::get_rse uses it to assign a number of variables in the current scope: eval(parse(text=paste(default_args[[i]]), "← ", i)).
-There are also more sophisticated cases.
-The plyr::allocate_column function uses eval for a deferred assignment to fill missing values in a column.
-Package overture repeatedly evaluates R expressions capturing any assignments that happen in it as samples of Markov chains.
-
-\emph{Code Obfuscation} Some packages use \eval to bypass the restrictions
+\emph{Avoid Boilerplate.}
+
+For example, the \code{PopED::get\_rse} uses it to assign a number of variables
+in the current scope: \code{eval(parse(text=paste(default_args[[i]]), "<- ",
+  i))}. There are also more sophisticated cases. The \code{plyr::allocate\_column}
+function uses eval for a deferred assignment to fill missing values in a column.
+Package overture repeatedly evaluates R expressions capturing any assignments
+that happen in it as samples of Markov chains.
+
+\emph{Code Obfuscation.} Some packages use \eval to bypass the restrictions
 imposed by the \code{R CMD CHECK} tool. This tool enforces some well-formedness
@@
 
-\emph{Miscellaneous}
+\emph{Miscellaneous.}
 

### 2021-08-09 paper/main.tex
Add more examples

@@
 
+\begin{figure}[h]
 \begin{lstlisting}
@@
 \end{lstlisting}
+  \caption{AdapSamp::rARS}
+\end{figure}
 
@@
 \end{lstlisting}
+  \caption{Rcpp::.makeCppMethods}
+\end{figure}
 
@@
 \end{lstlisting}
+  \caption{statnet.common::NVL}
+\end{figure}
+
 The expression \code{eval(substitute(alist(...)))} is used to extract the
@@
 \end{lstlisting}
+  \caption{igraph::\%||\%}
+\end{figure}
 
@@
 }
-
-\end{lstlisting}
+\end{lstlisting}
+  \caption{assertthat::see\_if}
+\end{figure}
 
@@
 \end{lstlisting}
+\caption{igraph::do\_call}
+\end{figure}
 
@@
 \end{lstlisting}
+  \caption{glue::glue}
+\end{figure}
 
@@
 
-\emph{Avoid Boilerplate.}
-
-For example, the \code{PopED::get\_rse} uses it to assign a number of variables
-in the current scope: \code{eval(parse(text=paste(default_args[[i]]), "<- ",
-  i))}. There are also more sophisticated cases. The \code{plyr::allocate\_column}
-function uses eval for a deferred assignment to fill missing values in a column.
-Package overture repeatedly evaluates R expressions capturing any assignments
-that happen in it as samples of Markov chains.
+The ability to capture the current call as a language object (\code{match.call()}),
+extract the unevaluated expression from function arguments (\code{substitute()})
+and \eval with first class environment makes R the toolkit for building DSLs.
+There are numerous widely spread DSLs used in R. For example, \code{data.table}
+package defines a DSL for data frame manipulation using the subset operator
+\code{[}. This allows one to write compact queries such as \code{flights[carrier ==
+  "AA", .N, by = origin]}\footnote{The \code{data.table} subset operation is
+  defined as \code{DT[i, j, k]}, where \code{j} expression is used to calculate on a
+  subset/reorder of rows from \code{i} grouped by \code{k}.} which will compute the
+number of flights done by American Airlines for each origin airport.
+Essentially, it uses \code{substitute} to capture the operand's expressions,
+transforms then appropriately and finally uses \eval to execute the query.
+
+Another example is the \code{tidyselect} package that defines a DSL for selecting
+columns in a data frame powering number of data manipulation operation in the
+\code{data.table} main competitor, the \code{dply} package. It uses eval to build an
+interpreter for evaluating column selectors such as \code{select(df,
+starts\_with("length") \& !(name:mass))} which returns all columns in \code{df}
+that starts with the string \code{length} and is not between \code{name} and
+\code{mass} columns.
+
+Another instance is the \emph{symbolic differentiation support} provided by base R package \emph{stats}, with two operators \code{D} and \code{deriv}.
+They support arithmetic operations and functions on real numbers such as \code{sin}, on expressions built with \code{expression()} or \code{parse}.
+In \emph{MCMCglmm} (Monte Carl Markov Chain Generalised Linear Mixed Models), a \code{Dtensor} object is created: it contains expressions that are derived twice (with \code{DD}).
+To get a concrete value, it is evaluated in  dataframe \code{mu} where symbolic variables are bound to actual numbers.
+\begin{lstlisting}
+ expr<-expression(beta_1 + time*beta_2+u)
+ mu = data.frame(beta_1=0.5, beta_2=1, time=3, u=2.3)
+ D[i]<-eval(DD(expr, name[unlist(comb.pos[i,])]), mu)
+\end{lstlisting}
+
+\emph{Easier Maintenance.} The combination of \c{match.call} and \eval is
+frequently used to make it easier to maintain functions with similar interfaces.
+This pattern is very commonly employed by modeling functions. Consider the
+example below of the \code{survfit.formula} function from package
+\code{survival}.
+
+\begin{figure}[h]
+\begin{lstlisting}
+survfit.formula <- function(formula, data, ...) {
+  Call <- match.call()
+  indx <- match(c('formula', 'data', ...), names(Call), nomatch=0)
+  temp <- Call[c(1, indx)]
+  temp[[1L]] <- quote(stats::model.frame)
+  mf <- eval.parent(temp)
+  ...
+}
+\end{lstlisting}
+  \caption{survival::survfit.formula}
+\end{figure}
+
+This function uses \code{match.call} to extract the expression with which it is
+called. Then, it extracts the indices of relevant arguments using the
+\code{match} function. Next, it subsets the \code{Call} variable to extract
+elements at those indices along with the first element which is the function
+name. The function name is then replaced with \code{stats::model.frame}.
+Finally, this expression is evaluated. By using \code{...}, the function accepts
+arbitrary arguments. It filters the arguments of interest using \code{match}
+function. \todo{explain more}
+
+\emph{Remove Boilerplate.} For example, the \code{PopED::get\_rse} uses it to
+assign a number of variables in the current scope:
+\code{eval(parse(text=paste(default_args[[i]]), "<- ", i))}. There are also more
+sophisticated cases. The \code{plyr::allocate\_column} function uses eval for a
+deferred assignment to fill missing values in a column. Package overture
+repeatedly evaluates R expressions capturing any assignments that happen in it
+as samples of Markov chains.
 
@@
 \end{lstlisting}
+\caption{aibd::scalaEnsure}
+\end{figure}
 
@@
 
-\subsection{Domain Specific Language}
-
-The ability to capture the current call as a language object (\code{match.call()}),
-extract the unevaluated expression from function arguments (\code{substitute()})
-and \eval with first class environment makes R the toolkit for building DSLs.
-There are numerous widely spread DSLs used in R. For example, \code{data.table}
-package defines a DSL for data frame manipulation using the subset operator
-\code{[}. This allows one to write compact queries such as \code{flights[carrier ==
-  "AA", .N, by = origin]}\footnote{The \code{data.table} subset operation is
-  defined as \code{DT[i, j, k]}, where \code{j} expression is used to calculate on a
-  subset/reorder of rows from \code{i} grouped by \code{k}.} which will compute the
-number of flights done by American Airlines for each origin airport.
-Essentially, it uses \code{substitute} to capture the operand's expressions,
-transforms then appropriately and finally uses \eval to execute the query.
-
-Another example is the \code{tidyselect} package that defines a DSL for selecting
-columns in a data frame powering number of data manipulation operation in the
-\code{data.table} main competitor, the \code{dply} package. It uses eval to build an
-interpreter for evaluating column selectors such as \code{select(df,
-starts\_with("length") \& !(name:mass))} which returns all columns in \code{df}
-that starts with the string \code{length} and is not between \code{name} and
-\code{mass} columns.
-
-Another instance is the \emph{symbolic differentiation support} provided by base R package \emph{stats}, with two operators \code{D} and \code{deriv}.
-They support arithmetic operations and functions on real numbers such as \code{sin}, on expressions built with \code{expression()} or \code{parse}.
-In \emph{MCMCglmm} (Monte Carl Markov Chain Generalised Linear Mixed Models), a \code{Dtensor} object is created: it contains expressions that are derived twice (with \code{DD}).
-To get a concrete value, it is evaluated in  dataframe \code{mu} where symbolic variables are bound to actual numbers.
-\begin{lstlisting}
- expr<-expression(beta_1 + time*beta_2+u)
- mu = data.frame(beta_1=0.5, beta_2=1, time=3, u=2.3)
- D[i]<-eval(DD(expr, name[unlist(comb.pos[i,])]), mu)
-\end{lstlisting}
+\emph{Lookup and indexing.}
 

### 2021-08-10 paper/main.tex
stuff

@@
-\documentclass[screen,acmsmall]{acmart}
-%\settopmatter{printfolios=false,printccs=false,printacmref=false}
+\documentclass[screen,acmsmall]{acmart}%\settopmatter{printfolios=false,printccs=false,printacmref=false}
 \setcopyright{none}
@@
 \makeatletter\let\expandableinput\@@input\makeatother
-
-%% https://www.davehofmann.de/defining-custom-language-templates-for-latex-listings/
-% Define Language
 \lstdefinelanguage{smalleR} {
-  % list of keywords
   morekeywords={
@@
 }
-
 \lstset{
@@
   framerule=0pt,
-  framexleftmargin=1mm,
-  framexrightmargin=1mm,
+%  framexleftmargin=1mm,
+%  framexrightmargin=1mm,
   tabsize=2,
@@
 \maketitle
-\renewcommand{\shortauthors}{A. Goel, P. Donat-Bouillud, F. Křikava, C. Kirsch and J. Vitek}
+\renewcommand{\shortauthors}{Goel et al.}
+\renewcommand{\shorttitle}{What We Eval in the Shadows}
 
@@
 
-\vspace{2mm}\noindent\emph{Trouble in Paradise.} Reflective facilities hinder
+\paragraph{Trouble in Paradise.} Reflective facilities hinder
 most attempts to reason about or apply meaning-preserving transformations to the
@@
 much precision as to become pointless. For compilers, anything but the most
-trivial, local optimizations are unsound after the use of \eval. Furthermore,
-the addition of arbitrary code --- code that could have been obtained from a
-network connection --- as a program is running is a security vulnerability
-waiting to happen. To illustrate these challenges, consider the interaction of a
-static analysis tool with a dynamic language. An abstract interpretation-based
-program analyzer computes an over-approximation of the set of possible behaviors
-exhibited by the program under study~\cite{cc77}. A reflective facility may have
-\emph{any} behavior that can be expressed in the target language, \ie any legal
-sequence of instructions can replace \eval. As dynamic languages tend to be
-permissive, the analysis has to, for example, assume that many (or all)
-functions in scope may have been redefined, \eg that \texttt{`+`} now opens a
-network connection or something equally surprising. A single occurrence of \eval
-causes the static analyzer to lose all information about the program state and
-meaning of identifiers. This loss of precision can sometimes be mitigated by
-analyzing the string argument~\cite{moller03} to bound its possible behavior,
-but when the string comes from outside the program, not much can be done. A
-frustrated group of researchers argued giving up on soundness and, instead,
-under-approximating dynamic features (soundiness)~\cite{soundy}. In their words,
-``a practical analysis, therefore, may pretend that \eval does nothing unless it
-can precisely resolve its string argument at compile time.'' Alas, assuming that
-\eval does not have side-effects or that side-effects will not affect the
-results of the analysis may be unduly optimistic.
-
-\vspace{2mm}\noindent\emph{Is Past Prologue?} Previous work investigated how
-\eval is used in web programming, specifically in websites that use
-JavaScript~\cite{pldi10a}. In 2010, 17 of the largest websites used the feature.
-In 2011, 82\% of the 10,000 most accessed sites used \eval~\cite{ecoop11}. Yet,
-the strings passed to \eval, and their behaviors, when executed, are far from
-random; it was shown that when one can observe several calls to \eval, the
-``shape'' of future calls can be predicted with 97\% accuracy~\cite{oopsla12b}.
-Overall, practical usage suggested that most reflective calls were relatively
-harmless. While this backs up the soundiness squad's approach, does it
-generalize to other application domains than web programming and to other
-languages?
-
-\vspace{2mm}\noindent\emph{The Here and Now.} In this study, we investigate the
-usage of \eval in programs written in the R programming language. R is a language
-designed by statisticians for applications in data science~\cite{r,R96}. What
-makes looking at R after JavaScript interesting is that, while both languages
-are dynamic, they are quite different. While one can program in an
-object-oriented style in R like in JavaScript, R is primarily a lazy, untyped,
-functional language. JavaScript was designed to run untrusted code in a browser,
-while R is used for statistical computing on desktops. JavaScript is a general-purpose language used by a vast community of programmers, while R is used for
-scientific computing by data scientists and domain experts with, often, limited
-programming experience. One can distinguish between library implementers,
-developers with some programming experience and a working knowledge of R, and
-end users, who are typically not expert programmers and often have only a
-cursory knowledge of the language.\footnote{Consider that R is lazy like
-Haskell. We informally surveyed end users and did not find a single user aware
-of this fact. Library developers, on the other hand, know and program
-defensively around laziness.} Thus, our goal is to highlight the differences in
-usage between JavaScript and R and try to explain those differences in terms of
-language features, application domain and programmer experience. Hopefully, some
-of our observations will generalize to other languages.
-Our data and code are open source and publicly
-available: \url{https://github.com/PRL-PRG/evalr-experiment}.
-
-\vspace{2mm}\noindent\emph{The What and How.} One significant benefit of
-choosing R is that every package in the CRAN repository is curated and comes
-with examples of typical usage. This gives us a large codebase that we can
-analyze dynamically. To observe \eval, we built a two-level monitoring
-infrastructure:\footnote{Our infrastructure is open source and publicly
-available and  will be submitted to the artifact evaluation committee. } we can
-monitor R programs by instrumentation --- this gives us access to many
-user-visible properties of R programs --- but we can also monitor the
-inner workings of the R interpreter --- this allows us to capture details not
-exposed at the source level. Dynamic analysis is limited; it can only observe
-behaviors triggered by the particular inputs passed to a program. Luckily, CRAN
-libraries come with many tests and use-cases. The choice of the corpus is crucial.
-Our corpus has been constructed to reflect the levels of sophistication of the R
-community. We distinguish between \emph{CRAN packages} (\CranPackages curated
-packages that pass stringent quality checks and are equipped with tests and
-sample data) and \emph{Kaggle scripts} (\KaggleUnique end-user written programs
-that perform a particular data analysis task). It is reasonable to expect that
-\eval usage differs between these datasets: the libraries represent a lively
-ecosystem with new libraries added each day, while end-user code is often thrown
-together, run once, and never revisited.
-
-\vspace{2mm}\noindent\emph{Why do we Eval?} The results of our study suggest
+trivial, local optimizations are unsound after use of \eval. Furthermore, the
+addition of arbitrary code --- code that could have been obtained from a network
+connection --- as a program is running is a security vulnerability waiting to
+happen. To illustrate these challenges, consider the interaction of a static
+analysis tool with a dynamic language. A program analyzer computes an
+over-approximation of the set of possible behaviors exhibited by the program
+under study, a reflective facility must be represented by all behaviors that can
+be expressed in the target language. \ie any legal sequence of instructions can
+replace \eval. As dynamic languages are permissive, the tool must assume that
+all functions in scope were redefined, \eg that \texttt{`+`} now opens a network
+connection. A single occurrence of \eval causes the static analyzer to lose all
+information about the program state and meaning of identifiers. This loss of
+precision can sometimes be mitigated by analyzing the string
+argument~\cite{moller03}, but if the string comes from outside the program, not
+much can be done. A frustrated group of researchers argued giving up on
+soundness and, instead, under-approximating dynamic features
+(soundiness)~\cite{soundy}. In their words, ``a practical analysis, may pretend
+that \eval does nothing unless it can precisely resolve its string argument at
+compile time.'' Alas, assuming that \eval does not have side-effects or that
+side-effects will not affect results is unduly optimistic.
+
+\paragraph{Is Past Prologue?} Previous work investigated
+\eval in web programming, specifically JavaScript web pages~\cite{pldi10a}. In
+2011, 82\% of the 10,000 most accessed sites used \eval~\cite{ecoop11}. Yet, the
+strings passed to \eval, and their behaviors, when executed, were far from
+random; it was shown that when one could observe several calls, the ``shape'' of
+future calls could be predicted with 97\% accuracy~\cite{oopsla12b}. Overall,
+practical usage suggested that most reflective calls were relatively harmless.
+While this backed up the soundiness squad's approach, does it generalize to
+other application domains and to other languages?
+
+\paragraph{The Here and Now.} In this study, we investigate the
+usage of \eval in programs written in the R language. R is a language designed
+by statisticians for applications in data science~\cite{r,R96}. What makes
+looking at R after JavaScript interesting is that, while both languages are
+dynamic, they are quite different. While one can program in an object-oriented
+style in R like in JavaScript, R is mostly a lazy functional language.
+JavaScript was designed to run untrusted code in a browser, while R is used for
+statistical computing on desktops. JavaScript is a general-purpose language used
+by a vast community of programmers, while R is used for scientific computing by
+data scientists and domain experts with, often, limited programming experience.
+One can distinguish between library implementers who have programming experience
+and a working knowledge of R, and end-users who are typically not expert
+programmers with a cursory knowledge of the language.\footnote{We surveyed
+end-users and did not find a single one aware that R is lazy.} Our goal is to
+highlight the differences in usage between JavaScript and R and explain them in
+terms of language features, application domain and programmer experience.
+Hopefully, some of our observations generalize to other languages. Our data
+and code are open source and publicly available:
+\url{https://github.com/PRL-PRG/evalr-experiment}.
+
+\paragraph{The What and How.} One benefit of
+R is that every package in the CRAN repository comes with examples of typical
+usage. This gives us a codebase that we can analyze dynamically. To observe
+\eval, we built a two-level monitoring infrastructure: we monitor programs by
+instrumentation and we also monitor the inner workings of the interpreter.
+Dynamic analysis is limited as it can only observe behaviors triggered by the
+particular inputs passed to a program. Luckily, R libraries come with many tests
+and use-cases. Our corpus is constructed to reflect the levels of sophistication
+of the R community. We distinguish between \emph{CRAN packages} (\CranPackages
+curated packages that pass quality checks and have tests and sample data) and
+\emph{Kaggle scripts} (\KaggleUnique end-user written programs) It is reasonable
+to expect \eval usage to differ between these datasets, libraries are part of a
+lively and growing ecosystem, while end-user code is often thrown together, run
+once, and never revisited.
+
+\paragraph{Why do we Eval?} The results of our study suggest
 that \eval is widely used for the implementation of the language, and in many
@@
 
-%\newpage
 \section{Background and Previous work}
@@
 
-\citet{ecoop12} gave a programming language-centric overview of the R language.
+\citet{ecoop12} gives a programming language-centric overview of the R language.
 They characterized it as a lazy, vectorized, functional language with a rich
 complement of dynamic features expressive enough to layer several object systems
-on top of the core language. Most data types are sequences of primitive values
-constructed by calling the combination function \c{c}. For instance,
-\c{c("Ha","bye")} evaluates to a vector of two strings, constants such as \c{42}
-are vectors of length one. To enable equational reasoning, values accessible
-through multiple aliases are copied when written to. Furthermore, values can be
-tagged by attributes; these are key-value pairs. For instance, the attribute
-\c{dim}-\c{c(2,2)} can be attached to the value at \c{x} by the call
-\c{attr(x,"dim")$\leftarrow$c(2,2)}. In this case, the addition of this
-attribute turns \c x into a matrix. The \c{class} attribute gives a 'class', in
-the object-oriented sense, to a value. So, \c{class(x)<-"human"} sets the class
-of \c{x} to \c{human}; classes are used for method dispatch. Every linguistic
-construct is desugared to a function call, even control flow statements,
-assignments, and bracketing. All functions can be shadowed or redefined, making
-the language at the same time remarkably flexible and exceedingly challenging to
-compile statically as vividly detailed by~\citet{dls19}. R uses a relaxed
-call-by-need convention for passing arguments to functions. Each argument is a
-thunk composed of an expression, its environment, and a slot for the result;
-these are called \emph{promises}. To get the value of an argument, the
-corresponding promise must be forced. Once forced, the promise's result is cached
-for future use.
+on top of the core language. Most data types are sequences of primitive values.
+For instance, \c{c("Ha","bye")} evaluates to a vector of two strings. Constants
+such as \c{42} are vectors of length one. To enable equational reasoning, values
+accessible through multiple aliases are copied when written to. Furthermore,
+values can be tagged by attributes; these are key-value pairs. For instance, the
+attribute \c{dim}-\c{c(2,2)} can be attached to the value at \c{x} by
+\c{attr(x,"dim")$\leftarrow$c(2,2)}. Adding this attribute turns \c x into a
+matrix. The \c{class} attribute gives a value a 'class' in the object-oriented
+sense. So, \c{class(x)<-"human"} sets the class of \c{x} to \c{human}; classes
+are used for method dispatch. Every linguistic construct is desugared to a
+function call, even control flow statements, assignments and bracketing. All
+functions can be shadowed and redefined, making R at the same time remarkably
+flexible and exceedingly challenging to compile as vividly detailed
+by~\citet{dls19}. R uses a relaxed call-by-need convention for passing arguments
+to functions. Each argument is a thunk composed of an expression, its
+environment, and a slot for the result; these are called \emph{promises}. To get
+the value of an argument, the corresponding promise must be forced. Once forced,
+the promise's result is cached for future use.
 
@@
 While a data-to-code facility is available in many languages, some design
-choices affect the expressive power of \eval. The key choices are the input
-format, the environment in which generated code is evaluated, and the reflective
-operations available to that code. Fig.~\ref{comp} summarizes designs.
+choices affect its expressive power. The key choices are the input format, the
+environment in which generated code evaluates and the reflective operations
+available to that code. Fig.~\ref{comp} summarizes a few designs.
 
@@
 The input to \eval can be in any format convertible to code. JavaScript allows
-arbitrary strings to be used. Both Julia and R are more restrictive as they
-require expressions (or abstract syntax trees). Finally, Java is the most
-restrictive as its classloader only accepts complete classes in bytecode form.
+arbitrary strings, Julia and R are more restrictive as they require expressions
+(or abstract syntax trees). Finally, Java is the most restrictive as its
+classloader only accepts complete classes in bytecode format. These differences
+mostly affect users who seem more comfortable crafting strings.
 
@@
 a program \eval can observe as well as the reach of potential side-effects
-performed by that operation. The most restrictive semantics is that of Java,
-where newly loaded code evaluates in the environment defined by the classes
-visible from the current classloader. Julia and JavaScript's strict mode limit
-\eval to the global environment. Finally, R is the most flexible as any
-accessible environment can be selected and passed to \eval.
+performed by its execution. The most restrictive semantics is that of Java,
+where newly loaded code evaluates in the environment consisting of the classes
+visible from the current classloader. Julia limits \eval to the symbols visible
+in global environment, so does JavaScript's strict mode. Finally, R is the most
+flexible as any accessible environment can be selected and passed to \eval. The
+choice of the environment is fully under the programmer's control.
 
@@
 program is accessible through reflective operations. Julia, Java and, JavaScript
-all allow some form of introspection on the data. R is more flexible as it is
-possible to inspect the program's call stack. Thus, any environment in the
-program can be inspected and modified.
+allow some form of introspection on the data that is visible in the environment
+in which \eval executes. R is more flexible as lets \eval inspect the program's
+call stack as well as the code of any function. Thus, any environment and any
+binding in contains can be inspected and modified.
 
@@
 to shield optimized code from some of the most pernicious uses of the
-facility~\cite{oopsla18a}. Furthermore, Julia provides a versioning
-mechanism, called world age, to ensure that any methods defined within an \eval
-only become visible at well-defined program points and thus that recompilation
-does not have to occur when optimized code is running~\cite{oopsla20a}.
+facility~\cite{oopsla18a}. Furthermore, Julia provides a versioning mechanism,
+called world age, to ensure that any methods defined within an \eval only become
+visible at well-defined program points and thus that recompilation does not have
+to occur when optimized code is running~\cite{oopsla20a}.
 
@@
 (\c{e}), an environment where to evaluate (\c{envir}), and an enclosure
-(\c{encl}) that is used to look up objects not found in \c{envir}.\footnote{Next to the \eval function R
-  offers three other variants:
+(\c{encl}) that is used to look up objects not found in \c{envir}.\footnote{Next
+to the \eval function R offers three other variants:
   \begin{inparaenum}[1.]
@@
       \c{eval(quote(...))},
-    \item \c{eval.parent(e, n)} specifies the evaluating
-      environment in terms of number of call frames (\c{n}) to go back---a shorthand
-      to \c{eval(e, parent.frame(n))}, and
-    \item \c{local} evaluates \c{e} in a fresh
-      environment---a shorthand to \c{evalq(e, new.env())}.
+    \item \c{eval.parent(e, n)} specifies the evaluating environment in terms of
+      number of call frames (\c{n}) to go back---a shorthand to \c{eval(e,
+        parent.frame(n))}, and
+    \item \c{local} evaluates \c{e} in a fresh environment---a shorthand to
+      \c{evalq(e, new.env())}.
   \end{inparaenum} }
-%
+
 \begin{lstlisting}
@@
 
-\noindent The expression passed to \eval can be thought of as an abstract
-syntax tree. It can be parsed from a string, created manually or using
-reflection:
-
-\vspace*{-2mm}
+\noindent The expression passed to \eval can be thought of as an abstract syntax
+tree. It can be parsed from a string, created manually or using reflection.
+
 \begin{figure}[htb]
@@
 
-\vspace*{-2mm}
-\noindent The call to \c{substitute(x)} extracts the unevaluated expression from the
-promise \c{x}. The function \c{match.call} returns an expression representing the current call.
+\noindent The call to \c{substitute(x)} extracts the unevaluated expression from
+the promise \c{x}. The function \c{match.call} returns an expression
+representing the current call.
 
@@
 
-\citet{ecoop11} provided the first large-scale study of the runtime behavior of
-\eval in JavaScript. They dynamically analyzed a corpus of the 10,000 most
-popular websites with an instrumented web browser to gather execution traces.
-They show that \eval is pervasive, with 82\% of the most popular websites using
-it. The reasons for its use include the desire to load code on demand,
-deserialization of JSON data, and lightweight meta-programming to customize web
-pages. While many uses were legitimate, just as many were unnecessary and could
-be replaced with equivalent and safer code. They categorized inputs to \eval so
-as to cover the vast majority of input strings. Restricting themselves to \eval
-in which all named variables refer to the global scope, many patterns could be
+\citet{ecoop11} provided the first study at scale of the behavior of \eval in
+JavaScript. A corpus of 10,000 popular websites was analyzed with an
+instrumented web browser to gather execution traces. Of those sites, 82\% used
+\eval for purposes such as on-demand code loading, deserialization of JSON data,
+or lightweight meta-programming to customize web pages. While many uses were
+legitimate, just as many were unnecessary and could be replaced with equivalent
+and safer code. The authors categorized inputs to \eval. For inputs in
+which all named variables refer to the global scope, many patterns could be
 replaced by more disciplined code~\cite{oopsla12b, moller12}. The work did not
@@
 \citet{ecoop12} had a few paragraphs on the usage of \eval in R. They found the
-that \eval  is widely used in R code with 8500 call sites in 1238 CRAN packages and 2.2 million dynamic calls.   The 15 most frequent call sites account for 88\% of those. The
-\c{match.arg} function is the highest used one with 54\% of all calls. In the
-other call sites, they saw two use cases. The most common is the evaluation of
-the source code of a promise retrieved by \c{substitute} in a new environment,
-\eg as done in the \c{with} function. The other use case is the invocation of a
-function whose name or arguments are determined dynamically.
-For this purpose, R provides \c{do.call} and thus \eval is overkill.
-Our corpus is larger, with \CranPackages packages, \packageTriggeredpkgs packages with at least one \eval, for a total of \PkgHitEvalCallSites sites and \packageAllcallsRnd dynamic calls, although they also looked at the Bioconductor package repository in addition to CRAN.
-We also track much more runtime information on the \eval calls during tracing.
-It enables us to have a more nuanced and accurate picture, and identify more diverse use cases of \eval.
+that \eval is widely used in R code with 8500 call sites in 1238 CRAN packages
+and 2.2 million dynamic calls. The 15 most frequent call sites account for 88\%
+of those. The \c{match.arg} function is the highest used one with 54\% of all
+calls. In the other call sites, they saw two use cases. The most common is the
+evaluation of the source code of a promise retrieved by \c{substitute} in a new
+environment, \eg as done in the \c{with} function. The other use case is the
+invocation of a function whose name or arguments are determined dynamically. For
+this purpose, R provides \c{do.call} and thus \eval is overkill. Our corpus is
+larger, with \CranPackages packages, \packageTriggeredpkgs packages with at
+least one \eval, for a total of \PkgHitEvalCallSites sites and
+\packageAllcallsRnd dynamic calls, although they also looked at the Bioconductor
+package repository in addition to CRAN. We also track much more runtime
+information on the \eval calls during tracing. It enables us to have a more
+nuanced and accurate picture, and identify more diverse use cases of \eval.
 
@@
 
-\vspace{1mm} 
-\noindent The pipeline runs in parallel~\cite{GNUparallel}
-orchestrated by a Makefile. Servers have identical environments thanks to a
-docker image with all dependencies installed\footnote{The docker image is part
-  of the accompanied artifact available at\\
-\url{https://github.com/PRL-PRG/evalr-experiment/tree/oopsla21}}.
+\noindent The pipeline runs in parallel~\cite{GNUparallel} orchestrated by a
+Makefile. Servers have identical environments thanks to a docker image with all
+dependencies installed.
 
@@
 
-\begin{wrapfigure}{4}{5.8cm}
-  \small
-  \vspace*{-2mm}
-\centering
+\begin{wrapfigure}{4}{5.8cm}  \small  \vspace*{-2mm}\centering
   \begin{tabular}{|r@{\,}r@{\,}l@{\,}r|r@{\,}r@{\,}l@{}r|} \hline
@@
 where the argument to \eval is a string which \eval simply returns. Of twenty
-randomly selected \evals, \packageUsefulValueEvalPercent of them need \eval for
+randomly selected \evals, XXXXX
+%% UNFINED TODO \packageUsefulValueEvalPercent
+of them need \eval for
 some inputs. Either we are dealing with a value that needs to be constructed
@@
 
-\vspace{1mm}\noindent Table~\ref{tab:provenance} summarizes the expression
-provenance in our corpus. This data is obtained by dynamically tainting values
-as they are produced by the various sources. Due to technical reasons, there is
-some imprecision in the results. In particular, we are not able to classify all
-sites. Manual inspection of numerous examples suggests that errors are rare.
+Table~\ref{tab:provenance} summarizes the expression provenance in our corpus.
+This data is obtained by dynamically tainting values as they are produced by the
+various sources. Due to technical reasons, there is some imprecision in the
+results. In particular, we are not able to classify all sites. Manual inspection
+of numerous examples suggests that errors are rare.
 

### 2021-08-10 paper/main.tex
Replace emph with paragraph for usage category headings

@@
 
-\emph{Simplify Interfaces}. A common use of \eval is to simplify interfaces by
+\paragraph{Simplify Interfaces} A common use of \eval is to simplify interfaces by
 letting users pass string argument which can be parsed and evaluated in custom a
@@
 
-\emph{Code Generation.} The \c{Rcpp} package which bridges R with C++ is a
+\paragraph{Code Generation} The \c{Rcpp} package which bridges R with C++ is a
 popular user of this pattern. Its \code{.makeCppMethods} function, shown below,
@@
 
-\emph{Extracting ... arguments.} A surprising use of \eval is the extraction of unevaluated text of arguments
-passed to a \code{...} parameter. Consider the definition of the \code{NVL}
-function from the \code{statnet.common} package. This function returns the first
-argument that is non \code{NULL}.
+\paragraph{Extracting \texttt{...} arguments} A surprising use of \eval is the
+extraction of unevaluated text of arguments passed to a \code{...} parameter.
+Consider the definition of the \code{NVL} function from the
+\code{statnet.common} package. This function returns the first argument that is
+non \code{NULL}.
 
@@
 
-\emph{Conditional Evaluation.} Developers frequently use \eval to explicitly
+\paragraph{Conditional Evaluation} Developers frequently use \eval to explicitly
 control the evaluation of code. Consider the \code{\%\|\|\%} operator from the
@@
 
-\emph{Core function variants.} Many packages use \eval to implement variations of the
+\paragraph{Core function variants} Many packages use \eval to implement variations of the
 \code{base} R functions. For example, the \code{igraph} package provides the
@@
 
-\emph{Domain-Specific Languages.} The combination of laziness, metaprogramming,
+\paragraph{Domain-Specific Languages} The combination of laziness, metaprogramming,
 first-class environments and dynamic evaluation make R the perfect toolkit for
@@
 
-\emph{Easier Maintenance.} The combination of \c{match.call} and \eval is
+\paragraph{Easier Maintenance} The combination of \c{match.call} and \eval is
 frequently used to make it easier to maintain functions with similar interfaces.
@@
 
-\emph{Remove Boilerplate.} For example, the \code{PopED::get\_rse} uses it to
+\paragraph{Remove Boilerplate} For example, the \code{PopED::get\_rse} uses it to
 assign a number of variables in the current scope:
@@
 
-\emph{Code Obfuscation.} Some packages use \eval to bypass the restrictions
+\paragraph{Code Obfuscation} Some packages use \eval to bypass the restrictions
 imposed by the \code{R CMD CHECK} tool. This tool enforces some well-formedness
@@
 
-\emph{Miscellaneous.}
-
-\emph{Lookup and indexing.}
+\paragraph{Miscellaneous}
+
+\paragraph{Lookup and indexing}
 

### 2021-08-10 analysis/normalized.Rmd
Tags with the non necessary evals

@@
 
-```{r, eval=dataset_name == "package", eval = FALSE}
+```{r, eval=dataset_name == "package"}
 manual_value_necessary <- read_csv(path(params$base_dir, "non-necessary-value-evals.csv"), col_types = cols(src_ref = col_character(), necessary = col_logical()))

### 2021-08-10 paper/main.tex
Update discussion in 5.1

@@
 
-\paragraph{Consistency} It is interesting how many different forms any given
-site sees. The more forms, the harder it is to characterize behavior at that
-site. Luckily, \packageNbOneMinimizedPercent of sites only see a single form.
-In other words, a compiler could correctly predict the \emph{shape} of an
-expression by looking just at the first call in \packageNbOneMinimizedPercent
-of the sites. Few sites are highly polymorphic (8 or more different forms).
-These include the pipe operator of the \c{magrittr} package, used to compose
-functions, \eg instead of \c{f(g(h(x),y),z)} one can write \c{h(x) \%>\% g(y) \%>\%
-f(z)}.
-
-\paragraph{Discussion} The variety of uses of \eval is evidenced by the number
-of different forms. In comparison, JavaScript \eval usage was more
-straightforward and more predictable as reported by \citet{oopsla12b}, with
-98.7\% of the sites with only one form. Nevertheless, simple forms dominate
-in R. Among these there are many cases where \eval could be replaced by less
-powerful constructs such as \c{get} to lookup a name in an environment,
-\c{assign} to set a value to a name in an environment or \c{call} to
-reflectively call a function.
-% TODO: how many receive just the simple form
+\paragraph{Discussion}
+
+It is interesting how many different forms any given site sees. The more forms,
+the harder it is to characterize behavior at that site. Luckily,
+\packageNbOneMinimizedPercent of sites only see a single form. In other words,
+a compiler could correctly predict the \emph{shape} of an expression by looking
+just at the first call in \packageNbOneMinimizedPercent of the sites. Few sites
+are highly polymorphic (8 or more different forms). These include the pipe
+operator of the \c{magrittr} package, used to compose functions, \eg instead of
+\c{f(g(h(x),y),z)} one can write \c{h(x) \%>\% g(y) \%>\% f(z)}.
+
+The variety of uses of \eval is evidenced by the number of different forms. In
+comparison, JavaScript \eval usage was more straightforward and more
+predictable as reported by \citet{oopsla12b}, with 98.7\% of the sites with
+only one form. On the other hand, there are \packageNbSimpleMinimizedOne sites
+that only receive one of the simple forms (\ie\xspace \c{X}, \c{V}, \c{\$} or
+\c{<-}). Additional \packageNbSimpleMinimizedMore sites takes multiple forms,
+but all are these simple ones. Among these sites there are many cases where
+\eval could be replaced by less powerful constructs such as \c{get} to lookup a
+name in an environment, \c{assign} to set a value to a name in an environment
+or \c{do.call} to reflectively call a function.
 
@@
   \c{as.environment}. It also includes the empty environment.
-\item \emph{Global:}  environments in which scripts or interactive commands
-  are evaluated.
-\item \emph{Package:}  environments of a loaded library.
+\item \emph{Global:}  environments in which scripts or interactive commands are
+  evaluated. \item \emph{Package:}  environments of a loaded library.
 \end{compactitem}

### 2021-08-10 paper/main.tex
Add more discussion in 5.1

@@
 
-\paragraph{Discussion}
-
+\mypara{Discussion.}
 \Eval is used in CRAN packages. Nearly every fourth package contains at least
@@
 
-\paragraph{Discussion}
-
+\mypara{Discussion.} 
 It is interesting how many different forms any given site sees. The more forms,
@@
 \c{<-}). Additional \packageNbSimpleMinimizedMore sites takes multiple forms,
-but all are these simple ones. Among these sites there are many cases where
-\eval could be replaced by less powerful constructs such as \c{get} to lookup a
+but all of them are these simple ones. Among these sites there are many cases
+where \eval could be replaced by simpler constructs such as \c{get} to lookup a
 name in an environment, \c{assign} to set a value to a name in an environment
@@
 
-\paragraph{Discussion}
+\mypara{Discussion.}
 The data presented in this section is what one could expect. For {\it
@@
 
-\paragraph{Discussion}
+\mypara{Discussion}
 The origin data suggests that constructing expressions from strings is a

### 2021-08-10 paper/main.tex
Update 5.1

@@
 
-The variety of uses of \eval is evidenced by the number of different forms. In
+The number of different forms evidences the variety of uses of \eval. In
 comparison, JavaScript \eval usage was more straightforward and more
@@
 that only receive one of the simple forms (\ie\xspace \c{X}, \c{V}, \c{\$} or
-\c{<-}). Additional \packageNbSimpleMinimizedMore sites takes multiple forms,
-but all of them are these simple ones. Among these sites there are many cases
+\c{<-}). Additional \packageNbSimpleMinimizedMore sites take multiple forms,
+but all of them are these simple ones. Among these sites, there are many cases
 where \eval could be replaced by simpler constructs such as \c{get} to lookup a
-name in an environment, \c{assign} to set a value to a name in an environment
-or \c{do.call} to reflectively call a function.
+name in an environment, \c{assign} to set a value to a name in an environment,
+or \c{do.call} to call a function reflectively.
 

### 2021-08-10 paper/main.tex
Improve provenance

--- 2021-08-10 paper/main.tex
@@
 It makes it possible to build a direct acyclic graph of the functions that build expressions or combine expressions during tracing to create a direct acyclic graph of the functions creating the expression passed to \eval.
-The following example yields the provenance graph with four nodes on Figure~\ref{fig:provgraph} for expression \c{e} passed to \eval.
+The following example yields the provenance graph with five nodes on Figure~\ref{fig:provgraph} for expression \c{e} passed to \eval.
 
@@
 
- Table~\ref{tab:detailed_provenances} shows the first provenances in number of sites.
-In \packagePercentProvenanceSitesd, the expression passed to \eval is not a language expression or a symbol but a floating point, or a string for instance.
-In that case, we did not track the provenance.
- % Explain also why we do not show [[ or $ in the table.
- % Explain the functions that we did not explain before.
+% Wrap tables do not play well with lists (they warn about it in the documentation actually!)
 \begin{wraptable}{r}{5cm}\small\centering
@@
 		\packageProvenanceNamel & \packageNbProvenanceSitesl & \packagePercentProvenanceSitesl \\
-		%\packageProvenanceNamem & \packageNbProvenanceSitesm & %\packagePercentProvenanceSitesm \\
+		%\packageProvenanceNamem & \packageNbProvenanceSitesm & %\packagePercentProvenanceSitesm \\ %as.vector
 		\packageProvenanceNamen & \packageNbProvenanceSitesn & \packagePercentProvenanceSitesn \\
@@
 	\end{tabular}
-	\caption{Provenances in number of sites.} \label{tab:detailed_provenances}
+	\caption{First provenances in number of sites.} \label{tab:detailed_provenances}
 \end{wraptable}
 
- We classify them into four categories:
+Table~\ref{tab:detailed_provenances} shows the first provenances in number of sites.
+In \packagePercentProvenanceSitesd\% of the sites, the expression passed to \eval is not a language expression or a symbol but a floating point, or a string for instance.
+In that case, we did not track the provenance.
+\c{.External} invokes a C function.
+For instance, in the following code, \c{.External(C\_deriv, expr, namevec, function.arg, tag, hessian)} calls a C function called \c{C\_deriv} to derive an expression.
+ Due to technical reasons, there is some imprecision in the
+results.
+In particular, we misclassify some sites but these cases happen rarely.
+For instance,  \packagePercentProvenanceSitesj\% of the sites have provenance \c{[[}, a subsetting operator, as we lose the provenance of the subsetted expression, often when that expression is given attributes.
+Manual inspection of numerous examples suggests that errors are rare.
+
+ We further classify them into four high-level categories:
 
 \begin{compactitem}[---]
-\item {\it Constructed:} Expressions can be constructed by invoking the
-  \c{quote}, \c{enquote}, \c{expression} functions, or by building a call with \c{call} or \c{as.call}.
 \item {\it Reflection:} This group corresponds to uses of \c{match.call} to
@@
 	\c{str2expression}, or \c{str2lang}.
+\item {\it Constructed:} Expressions can be constructed by invoking the
+  \c{quote}, \c{expression} functions, or by building a call with \c{call} or \c{as.call}.
 \item {\it Environment: }   A symbol is created with \c{as.name} or \c{as.symbol}, typically to find the value of a symbol in a non-local environment.
@@
 
-\NOTE{I only see 70 sites with formulas (using the class of the expression) so it does not appear here.}
-
-\NOTE{I can also put the absolute number of provenances, \ie if an expression has multiple provenances, we do not keep only the representative but also the other provenances.
+%\NOTE{I only see 70 sites with formulas (using the class of the expression) so it does not appear here.}
+
+%\NOTE{I can also put the absolute number of provenances, \ie if an expression has multiple provenances, we do not keep only the representative but also the other provenances.
 Also, if one site sometimes gets match.call and sometimes substitute, it is not going to be counted in either of them currently.}
@@
 Table~\ref{tab:provenance} summarizes the expression provenance in our corpus.
-This data is obtained by dynamically tainting values as they are produced by the
-various sources. Due to technical reasons, there is some imprecision in the
-results. In particular, we are not able to classify all sites. Manual inspection
-of numerous examples suggests that errors are rare.
-
-\begin{wraptable}{r}{5cm}\small\centering
+Note that some sites can be invoked several times with different provenances.
+In that case, we have counted them several times in the table.
+
+\begin{wraptable}[8]{r}{5cm}\small\centering
 \begin{tabular}{r|r|r} \hline
 Origin  & \#sites & \%sites \\\hline
+Reflection &  \packageNbReflectionSites & \packageReflectionSitesPercent\\
+String & \packageNbStringSites & \packageStringSitesPercent \\
 Constructed & \packageNbConstructedSites & \packageConstructedSitesPercent \\
-Reflection &  \packageNbReflectionSites & \packageReflectionSitesPercent\\
 Environment & \packageNbSymbolSites & \packageSymbolSitesPercent \\
-External & \packageNbExternalSites & \packageExternalSitesPercent \\
-String & \packageNbStringSites & \packageStringSitesPercent \\\hline
+External & \packageNbExternalSites & \packageExternalSitesPercent \\\hline
 \end{tabular}
@@
 \paragraph{Discussion}
-The origin data suggests that constructing expressions from strings is a
-minority of the use cases. Instead, the constructed category shows that most
+The origin data suggests that reflection is the first use case for \eval, \ie using \c{match.call}, \c{substitute}, or \c{match.call} to access and transform the arguments of a function when calling the function.
+Constructing expressions from strings is also
+a common use case. The constructed category shows that most
 \evals comes from code that was processed by the compiler, and may be slightly

### 2021-08-10 paper/main.tex

@@
 that only receive one of the simple forms (\ie\xspace \c{X}, \c{V}, \c{\$} or
-\c{<-}). Additional \packageNbSimpleMinimizedMore sites takes multiple forms,
-but all are these simple ones. Among these sites there are many cases where
-\eval could be replaced by less powerful constructs such as \c{get} to lookup a
-name in an environment, \c{assign} to set a value to a name in an environment
-or \c{do.call} to reflectively call a function.
+\c{<-}). Additional \packageNbSimpleMinimizedMore sites take multiple forms,
+but all of them are these simple ones. Among these sites, there are many cases
+where \eval could be replaced by simpler constructs such as \c{get} to lookup a
+name in an environment, \c{assign} to set a value to a name in an environment,
+or \c{do.call} to call a function reflectively.
 
@@
 
-\paragraph{Discussion}
+\mypara{Discussion}
 The origin data suggests that reflection is the first use case for \eval, \ie using \c{match.call}, \c{substitute}, or \c{match.call} to access and transform the arguments of a function when calling the function.

### 2021-08-10 paper/main.tex
Add remove boilerplate section

@@
 
-\paragraph{Simplify Interfaces} A common use of \eval is to simplify interfaces by
+\mypara{Simplify Interfaces} A common use of \eval is to simplify interfaces by
 letting users pass string argument which can be parsed and evaluated in custom a
@@
 
-\paragraph{Code Generation} The \c{Rcpp} package which bridges R with C++ is a
+\mypara{Code Generation} The \c{Rcpp} package which bridges R with C++ is a
 popular user of this pattern. Its \code{.makeCppMethods} function, shown below,
@@
 
-\paragraph{Extracting \texttt{...} arguments} A surprising use of \eval is the
+\mypara{Extracting varargs} A surprising use of \eval is the
 extraction of unevaluated text of arguments passed to a \code{...} parameter.
@@
 
-\paragraph{Conditional Evaluation} Developers frequently use \eval to explicitly
+\mypara{Conditional Evaluation} Developers frequently use \eval to explicitly
 control the evaluation of code. Consider the \code{\%\|\|\%} operator from the
@@
 
-\paragraph{Core function variants} Many packages use \eval to implement variations of the
+\mypara{Core function variants} Many packages use \eval to implement variations of the
 \code{base} R functions. For example, the \code{igraph} package provides the
@@
 
-\paragraph{Domain-Specific Languages} The combination of laziness, metaprogramming,
+\mypara{Domain-Specific Languages} The combination of laziness, metaprogramming,
 first-class environments and dynamic evaluation make R the perfect toolkit for
@@
 
-\paragraph{Easier Maintenance} The combination of \c{match.call} and \eval is
-frequently used to make it easier to maintain functions with similar interfaces.
-This pattern is very commonly employed by modeling functions. Consider the
-example below of the \code{survfit.formula} function from package
-\code{survival}.
+\mypara{Boilerplate Removal} The combination of \code{match.call} and \code{eval}
+is frequently used by modeling functions to avoid boilerplate code and ease
+maintenance. Consider the example shown below of function \code{survfit.formula}
+from the \code{survival} package.
 
@@
 \begin{lstlisting}
-survfit.formula <- function(formula, data, ...) {
+survfit.formula <- function(formula, data, weights, ...) {
   Call <- match.call()
-  indx <- match(c('formula', 'data', ...), names(Call), nomatch=0)
+  indx <- match(c('formula', 'data', 'weights'), names(Call), nomatch=0)
   temp <- Call[c(1, indx)]
@@
   mf <- eval.parent(temp)
-  ...
 }
@@
 
-This function uses \code{match.call} to extract the expression with which it is
-called. Then, it extracts the indices of relevant arguments using the
-\code{match} function. Next, it subsets the \code{Call} variable to extract
-elements at those indices along with the first element which is the function
-name. The function name is then replaced with \code{stats::model.frame}.
-Finally, this expression is evaluated. By using \code{...}, the function accepts
-arbitrary arguments. It filters the arguments of interest using \code{match}
-function. \todo{explain more}
-
-\paragraph{Remove Boilerplate} For example, the \code{PopED::get\_rse} uses it to
+Here, \code{match.call} is used to access the expression with which
+\code{survfit.formula} is called. The next three lines construct a call to
+\code{stats::model.frame} with argument text of parameters \code{'formula'},
+\code{'data'}, and \code{'weights'}. Then, the call expression is evaluated in
+the parent environment. The use of \code{match.call}, avoids repeated use of
+\code{substitute} for extracting the argument expressions. This setup is also
+extensible as new parameter names can be easily added to the \code{match} call.
+
+For example, the \code{PopED::get\_rse} uses it to
 assign a number of variables in the current scope:
@@
 
-\paragraph{Code Obfuscation} Some packages use \eval to bypass the restrictions
+\mypara{Code Obfuscation} Some packages use \eval to bypass the restrictions
 imposed by the \code{R CMD CHECK} tool. This tool enforces some well-formedness
@@
 
-\paragraph{Miscellaneous}
-
-\paragraph{Lookup and indexing}
+\mypara{Miscellaneous}
+
+\mypara{Lookup and indexing}
+
+\mypara{Attributes}
+Package \code{nlme} uses \eval to evaluate expressions attached as attributes to
+its objects. Consider the definition of the \code{formula.pdMat} function below.
+It extracts the \code{"formula"} attribute attached to the \code{pdMat} object
+and evaluates it.
+
+\begin{figure}[h]
+\begin{lstlisting}
+formula.pdMat <- function(x, asList, ...) eval(attr(x, "formula"))
+\end{lstlisting}
+  \caption{nlme::formula.pdMat}
+\end{figure}
 

### 2021-08-10 paper/main.tex
Add formula resolution

@@
 
-For example, the \code{PopED::get\_rse} uses it to
-assign a number of variables in the current scope:
-\code{eval(parse(text=paste(default_args[[i]]), "<- ", i))}. There are also more
-sophisticated cases. The \code{plyr::allocate\_column} function uses eval for a
-deferred assignment to fill missing values in a column. Package overture
-repeatedly evaluates R expressions capturing any assignments that happen in it
-as samples of Markov chains.
+\mypara{Formula Resolution} Many packages use \code{eval} to evaluate formula
+expressions. Consider the definition of the \code{formula.pdMat} function from
+\code{nlme} package.
+
+\begin{figure}[h]
+\begin{lstlisting}
+formula.pdMat <- function(x, asList, ...) eval(attr(x, "formula"))
+\end{lstlisting}
+  \caption{nlme::formula.pdMat}
+\end{figure}
+
+It extracts the \code{"formula"} attribute attached to the \code{pdMat} object
+and evaluates it.
 
@@
 to obfuscate the code, bypassing the checks performed by the tool. An example of
-this use is the \code{scalaEnsure} function from the \code{aibd} package. It
-mutates the variable \code{s} in the package environment. However, the package
-environment in R is locked by default, which prevents bindings from being
-mutated. Environment can be unlocked using \code{unlockBinding}, but its use is
-flagged by the well-formedness checks. To work around this limitation, the
-author passes the expression to unlock the package environment as a string to
-\code{parse} which is then evaluated by \code{eval}.
-
+this use is the \code{scalaEnsure} function from the \code{aibd} package.
 \begin{figure}[h]
@@
 
+This function mutates the variable \code{s} in the package environment. However,
+the package environment in R is locked by default, which prevents bindings from
+being mutated. Environment can be unlocked using \code{unlockBinding}, but its
+use is flagged by the well-formedness checks. To work around this limitation,
+the author passes the expression to unlock the package environment as a string
+to \code{parse} which is then evaluated by \code{eval}.
+
 \mypara{Miscellaneous}
@@
 
-\mypara{Attributes}
-Package \code{nlme} uses \eval to evaluate expressions attached as attributes to
-its objects. Consider the definition of the \code{formula.pdMat} function below.
-It extracts the \code{"formula"} attribute attached to the \code{pdMat} object
-and evaluates it.
-
-\begin{figure}[h]
-\begin{lstlisting}
-formula.pdMat <- function(x, asList, ...) eval(attr(x, "formula"))
-\end{lstlisting}
-  \caption{nlme::formula.pdMat}
-\end{figure}
+For example, the \code{PopED::get\_rse} uses it to
+assign a number of variables in the current scope:
+\code{eval(parse(text=paste(default_args[[i]]), "<- ", i))}. There are also more
+sophisticated cases. The \code{plyr::allocate\_column} function uses eval for a
+deferred assignment to fill missing values in a column. Package overture
+repeatedly evaluates R expressions capturing any assignments that happen in it
+as samples of Markov chains.
 

### 2021-08-10 paper/main.tex
Fix DSL description

@@
 \input{macro}
-\newcommand{\mypara}[1]{\medskip\noindent\emph{#1}\xspace}
 \newcommand{\NOTE}[1]{{\it Note: #1}\xspace}
@@
 \begin{abstract}
-  \noindent Most dynamic languages allow users to turn text into code using
+ Most dynamic languages allow users to turn text into code using
   various functions, often named \eval, with language-dependent semantics. The
@@
 
-\noindent The expression passed to \eval can be thought of as an abstract syntax
+The expression passed to \eval can be thought of as an abstract syntax
 tree. It can be parsed from a string, created manually or using reflection.
@@
 
-\noindent The call to \c{substitute(x)} extracts the unevaluated expression from
+ The call to \c{substitute(x)} extracts the unevaluated expression from
 the promise \c{x}. The function \c{match.call} returns an expression
@@
 %
-\noindent Environments are used as hash maps as they have reference semantics
+ Environments are used as hash maps as they have reference semantics
 and a built-in string lookup.
@@
 
-\mypara{Discussion.} When we started this project, our goal was to contrast the
+\paragraph{Discussion.} When we started this project, our goal was to contrast the
 usage of \eval by package developers and end-users, but the Kaggle dataset have
@@
    }
-   \vspace{-2mm}\caption{Pipeline}\label{fig:pipeline}
+   \caption{Pipeline}\label{fig:pipeline}
  \end{figure}
@@
 
-\noindent The pipeline runs in parallel~\cite{GNUparallel} orchestrated by a
+ The pipeline runs in parallel~\cite{GNUparallel} orchestrated by a
 Makefile. Servers have identical environments thanks to a docker image with all
@@
 
-\mypara{Limitations.} Even with the extension described above, there are
+\paragraph{Limitations.} Even with the extension described above, there are
 \PkgUndefinedRnd \eval calls without source information (\PkgUndefinedRatio of
@@
 
-\newpage
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Measuring Eval across Datasets}
+\section{Measuring Eval}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\begin{wrapfigure}{r}{5cm} \vspace{-1cm} \hspace*{-12mm}
+This section reports on the frequency of \eval in CRAN. We use \emph{site} to
+refer to an occurrence of a call site to the \eval function in the source code
+and \emph{call} to denote an observed invocation of the \eval function.
+
+There are \PkgEvalCallSites \eval sites in \PkgPackages packages;
+\PkgPackagesRatio packages use \eval. Over half of these packages have fewer
+than 3 sites, and with the exception \MaxEvalCallSitesPackage which has
+\MaxEvalCallSitesCount sites, all packages contain fewer than
+\MaxEvalCallSitesRest sites. Fig~\ref{fig:pkgs-eval-callsites-hist} shows a
+histogram of sites per package. Sites appear in \PkgFunsWithEval functions;
+\CranFunsWithEvalRatio of all functions use \eval.
+
+\begin{figure}[!b]
 \centering
-  \includegraphics[width=72mm]{pkgs-eval-callsites-hist.pdf}
-  \caption{CRAN \eval call sites}%
-  \label{fig:pkgs-eval-callsites-hist}\vspace{-3mm}
-\end{wrapfigure}
-%
-This section focuses on CRAN packages and reports statistics about the usage of
-\eval. We use the word \emph{site} to refer to an occurrence of a call site to the \eval
-function in the source code and \emph{call} to denote an observed invocation of
-the \eval function.
-
-\begin{figure}[!b]
+  \includegraphics[width=10cm]{pkgs-eval-callsites-hist.pdf}
+  \caption{ \eval call sites}\label{fig:pkgs-eval-callsites-hist}
+
+  \medskip
+
   \small
@@
 
-\end{figure}
-
-\begin{figure}[b] \centering
-  \includegraphics[width=.7\textwidth]{traced-eval-callsites.pdf} \centering
+\medskip
+
+\includegraphics[width=.8\textwidth]{traced-eval-callsites.pdf} \centering
   \caption{\eval call sites coverage of the \PkgPackages packages.}%
@@
 
-  \vspace*{-1mm}
   \includegraphics[width=5.6cm, trim=5.5cm 0 5cm 0, clip]{package_calls_per_run_per_call_site}
@@
 
-\caption{Normalized calls} \label{cn}\vspace{-2mm}
-
-\medskip
-\medskip
-
+\caption{Normalized calls} \label{cn}
+
+\medskip\medskip
 \begin{tabular}{c}
-  \vspace*{-1mm}
 {\hspace{-25mm}\includegraphics[width=0.7\textwidth]{package_size_loaded_distribution}}
@@
 \end{wrapfigure}
-There are \PkgEvalCallSites \eval sites in \PkgPackages packages. The proportion
-of packages calling \eval is \PkgPackagesRatio. Over half of these packages
-have fewer than 3 sites, and with the exception \MaxEvalCallSitesPackage, which
-has \MaxEvalCallSitesCount sites, all packages contain fewer than
-\MaxEvalCallSitesRest sites (\cf Fig~\ref{fig:pkgs-eval-callsites-hist}, which
-shows a histogram of sites per package). These sites appear in \PkgFunsWithEval
-functions (\CranFunsWithEvalRatio of all functions in CRAN).
-For dynamic analysis, we run \CranRunnableScripts programs extracted from
-\CranPackages packages. Any run that does not exercise \eval is discarded. This
-left \packageNbruns runs from \packageCorpus packages. There were
-\packageAllcalls calls in \packageTriggeredpkgs packages originating from
-\PkgHitEvalCallSites unique sites. In terms of coverage, the data exercised
-\PkgHitEvalCallSitesAvgRatio of sites, a coverage similar to the code coverage
-metric for the packages, which is \PkgCodeCoverage. The fact that not all sites
-are exercised can be chalked down to incomplete tests and occasional analysis
-failures (\PkgFailedProgramsRatio of programs crashed or timed out).
-Fig~\ref{fig:traced-eval-callsites} shows the number of sites that were
-exercised; coverage is unequal. Figure~\ref{freq} summarizes the frequency of
-dynamic calls to \eval, with the left
-column being the number of calls and the right, the number of packages in that
-range. There are \packageFewcalls packages with low \eval frequency, fewer than
-100 calls, and a smaller, but still significant, number of packages,
-\packageManycalls to be precise, that use \eval more than 1,000 times. Package
-\packageMaxcallspack makes \packageMaxcalls calls and thus accounts for over
-half of the observations. Figure~\ref{tab:variantseval} summarizes the use of
-variants of \eval. For each of the four variants, the first row shows the number
-of sites in the corpus (\emph{static}), the second row is the number of sites
-that were encountered during analysis (\emph{exercised}), the last row is the
-number of calls (\emph{invocations}). The overwhelming majority of sites and
-calls are to \eval itself, \c{eval.parent} is rare, and both \c{evalq} and
-\c{local} are barely used at all. The difference between sites and exercised
-sites underscores the limits of code coverage.
-
-Figure~\ref{cn}(a) shows normalized call counts per site; on the left are the
-average number of calls from a given site and given run, on the right are the
-counts of sites that fall in that range. For instance, \packageRunbinh sites are
-invoked 2,000+ times per run. Larger numbers suggest loops or recursive contexts
--- this seems to be the exception as most \evals, \packageRunbina, are exercised
-50 times or less. Figure~\ref{cn}(b) zooms in on low frequency sites. The x-axis
-shows normalized calls, and the y-axis is the number of sites for that value.
-Most low-frequency \evals are invoked only once; about half as many are invoked
-twice; after that, the frequency quickly drops.
-
-\Eval takes any value, but if its argument is not an expression, it is returned
-unchanged. Expressions account for \packageCodepercent of arguments. More
-specifically, \packageSymbolpercent are \texttt{symbol}s (single variables such
-as \c{x}), \packageLanguagepercent are \c{language} objects (function calls),
-  and \packageExpressionpercent are \c{expression} objects (lists of
-  expressions). Further inspection reveals that most symbols,
-  \packageGgplotsymbolpercent to be exact, come from a single site in the
-  \c{ggplot2} package and have the value \c{\_inherit}.\footnote{This models
-  inheritance in \c{ggproto}, one of the many object-oriented systems in R, used
-  in the \c{ggplot2} graphics library.}
+
+Our pipeline runs \CranRunnableScripts scripts extracted from \CranPackages
+packages. Any run that does not call \eval is discarded, leaving us with
+\packageNbruns runs. We recorded \packageAllcalls \eval calls originating from
+\PkgHitEvalCallSites sites. This exercised \PkgHitEvalCallSitesAvgRatio of
+sites, the coverage is similar to the package code coverage metric which is
+\PkgCodeCoverage. The reason some sites are not exercised can be chalked down to
+incomplete tests and analysis failures (\PkgFailedProgramsRatio of all scripts
+crashed or timed out). Fig~\ref{fig:traced-eval-callsites} plots exercised site
+with respect to all sites for each program; as can be seen coverage is unequal.
+Figure~\ref{freq} summarizes the frequency of calls, number of calls on the left
+and number of packages on the right. There are \packageFewcalls packages with
+fewer than 100 calls, and \packageManycalls packages with more than 1,000 calls.
+\packageMaxcallspack calls \eval \packageMaxcalls times and thus accounts for
+over half of the observations. Fig.~\ref{tab:variantseval} summarizes use of
+variants, the first row has sites (\emph{static}), the second has sites
+encountered during analysis (\emph{exercised}) and the last gives calls
+(\emph{invocations}). Most sites and calls go to \eval itself, \c{eval.parent}
+is rare, and both \c{evalq} and \c{local} are barely used at all.
+
+Figure~\ref{cn}(a) shows average count of calls from a given site and given run
+and how many sites fall in that range. For instance, \packageRunbinh sites are
+called over 2,000 times per run. Larger numbers suggest loops or recursive
+contexts -- this seems to be the exception; \packageRunbina of sites are called
+fewer than 50 times. Fig.~\ref{cn}(b) zooms in on low frequency sites. The
+x-axis shows normalized calls, and the y-axis counts sites. Most sites are
+invoked once, half as many are invoked twice, the drop off is sharp after that.
+
+\Eval accepts any value as argument, but if the value is not an expression,
+\eval returns it unchanged. Expressions account for \packageCodepercent of
+arguments. More specifically, \packageSymbolpercent are \c{symbol}s (variables
+such as \c{x}), \packageLanguagepercent are \c{language} objects (expressions
+representing a function calls), and \packageExpressionpercent are \c{expression}
+objects (lists of expressions). Further inspection reveals that \c{\_inherit}
+accounts for \packageGgplotsymbolpercent of symbols and comes from one site in
+\c{ggplot2}.\footnote{To model inheritance in \c{ggproto}, one of the many
+object-oriented systems for R.}
 
@@
 
-\mypara{Discussion.}
+\paragraph{Discussion.}
 \Eval is used in CRAN packages. Nearly every fourth package contains at least
@@
 
-\medskip\noindent  We detail these forms and discuss their implication for the behavior of \eval.
+\medskip  We detail these forms and discuss their implication for the behavior of \eval.
 
@@
 
-\medskip\noindent\EE{$min(e)=\c{V}$} Expressions that represent values are
+\medskip\EE{$min(e)=\c{V}$} Expressions that represent values are
 frequently passed to \eval -- they occur in 17\% of sites. The majority of
@@
 
-\medskip\noindent\EE{$min(e)=\c{X}$} Variable lookups are the most common form;
+\medskip\EE{$min(e)=\c{X}$} Variable lookups are the most common form;
 they are found in 28\% of the sites. This form includes simple variable reads,
@@
 
-\medskip\noindent\EE{$min(e)=\c{\$}$} This form extends \c X to include lookup
+\medskip\EE{$min(e)=\c{\$}$} This form extends \c X to include lookup
 with the dollar operator, \eg~\c{x\$f}, and vector indexing, \eg~\c{x[42]} or
@@
 
-\medskip\noindent\EE{$min(e)=$~\c{<-}} This form includes both assignment
+\medskip\EE{$min(e)=$~\c{<-}} This form includes both assignment
 operators, the direct assignment {\tt <-}, and assignment to the parent
@@
 
-\medskip\noindent\EE{$min(e)=\c{F()}$} This form captures simple function calls,
+\medskip\EE{$min(e)=\c{F()}$} This form captures simple function calls,
 \eg~\c{f(2)}, with neither variables nor assignments. More specifically, it
@@
 
-\medskip\noindent\framebox{$min(e)=\c{F(X)}$}~\EE{$min(e)=\c{F(F(X))}$} These
+\medskip\framebox{$min(e)=\c{F(X)}$}~\EE{$min(e)=\c{F(F(X))}$} These
 forms allow for function calls whose arguments may include variable references.
@@
 
-\medskip\noindent\EE{$min(e)=\c{FUN}$} This form captures expressions that
+\medskip\EE{$min(e)=\c{FUN}$} This form captures expressions that
 define functions, \eg~\c{function(x) x+1}, and do nothing else. \c{FUN} occurs
@@
 
-\medskip\noindent\EE{$min(e)=\c{BLOCK}$} This form captures multi-statement
+\medskip\EE{$min(e)=\c{BLOCK}$} This form captures multi-statement
 code blocks, which occur in only 4\% of sites. These are huge expressions; we
@@
 
-\medskip\noindent\EE{$min(e)=\c{model.frame}$} The \c{model.frame} function
+\medskip\EE{$min(e)=\c{model.frame}$} The \c{model.frame} function
 returns a dataframe resulting from fitting the model described in a given
@@
 
-\mypara{Discussion.} 
+\paragraph{Discussion.}
 It is interesting how many different forms any given site sees. The more forms,
@@
 
-\noindent
 As shown in Table~\ref{tab:highlevelenvironments}, most calls evaluate in the
@@
 
-\noindent
 Global \evals are likely split between intentional and accidental one. Direct
@@
 
-\mypara{Discussion.}
+\paragraph{Discussion.}
 The data presented in this section is what one could expect. For {\it
@@
 
-\mypara{Discussion}
+\paragraph{Discussion}
 The origin data suggests that reflection is the first use case for \eval, \ie using \c{match.call}, \c{substitute}, or \c{match.call} to access and transform the arguments of a function when calling the function.
@@
 
-% TODO do assignments change types?
-
 \begin{table}[h]
@@
 
-\mypara{Discussion} In JavaScript assignments in \eval can happen in either
+Out of the minimized expressions, it is \c{BLOCK}, \c{<-} and
+\c{F(F(X)} that contribute to the vast majority of side effects. Concretely,
+92\% of \c{BLOCK}, 87\% of \c{<-} and 15\% of \c{F(F(X))} expressions contain
+side effects. In the case of \c{BLOCK}, most of the side effects happen in the
+environment passed to \eval (which by default is the parent environment, \ie
+the local or function in our classification). For \c{<-} it is 15\%, however,
+that is only because of the \c{plyr::allocate\_column} which does the side
+effect in the environment containing the data frame that is the target for the
+newly allocated column. Without this call, \c{<-} is similarly predictable with
+96\% of side effects in the passed environment. For \c{F(F(X))} minimization
+for, it is less clear as 48\% of side effects happen in a different environment
+to the one that was passed to \eval.
+
+\paragraph{Discussion} In JavaScript assignments in \eval can happen in either
 local scope or less often, when called through an alias, in the global scope. In
@@
 that over half of them happen in a predictable environment.
-% TODO connect the minimization group and side-effects
 
@@
 
-\mypara{Simplify Interfaces} A common use of \eval is to simplify interfaces by
+\paragraph{Simplify Interfaces} A common use of \eval is to simplify interfaces by
 letting users pass string argument which can be parsed and evaluated in custom a
@@
 
-\mypara{Code Generation} The \c{Rcpp} package which bridges R with C++ is a
+\paragraph{Code Generation} The \c{Rcpp} package which bridges R with C++ is a
 popular user of this pattern. Its \code{.makeCppMethods} function, shown below,
@@
 .makeCppMethods <- function(methods, cppclassinfo, env) {
-    cppMethods <- names(cppclassinfo@methods)
-    newMethods <- names(methods)
+    ...
     for(what in cppMethods[! cppMethods %in% newMethods])
         methods[[what]] <- eval(substitute(
-                  function(...) .CppObject$WHAT(...), list(WHAT = as.name(what))),
-                                env)
-    methods
+                  function(...) .CppObject$WHAT(...), list(WHAT = as.name(what))), env)
+    ...
 }
@@
 
-\mypara{Extracting varargs} A surprising use of \eval is the
+\paragraph{Extracting varargs} A surprising use of \eval is the
 extraction of unevaluated text of arguments passed to a \code{...} parameter.
@@
 
-\mypara{Conditional Evaluation} Developers frequently use \eval to explicitly
+\paragraph{Conditional Evaluation} Developers frequently use \eval to explicitly
 control the evaluation of code. Consider the \code{\%\|\|\%} operator from the
@@
 
-\mypara{Core function variants} Many packages use \eval to implement variations of the
+\paragraph{Core function variants} Many packages use \eval to implement variations of the
 \code{base} R functions. For example, the \code{igraph} package provides the
@@
 \mypara{Domain-Specific Languages} The combination of laziness, metaprogramming,
-first-class environments and dynamic evaluation make R the perfect toolkit for
+first-class environments and dynamic evaluation makes R the perfect toolkit for
 building DSLs. An example of this is the \code{glue} package which provides
 functions for string interpolation. The code snippet below shows an example of
-this.
+\code{glue} function in action.
 
@@
 
-text enclosed by c{\{} and c{\}} is
-evaluated using \eval as an R expression and replaced by its result.
-
-The ability to capture the current call as a language object (\code{match.call()}),
-extract the unevaluated expression from function arguments (\code{substitute()})
-and \eval with first class environment makes R the toolkit for building DSLs.
-There are numerous widely spread DSLs used in R. For example, \code{data.table}
-package defines a DSL for data frame manipulation using the subset operator
-\code{[}. This allows one to write compact queries such as \code{flights[carrier ==
-  "AA", .N, by = origin]}\footnote{The \code{data.table} subset operation is
-  defined as \code{DT[i, j, k]}, where \code{j} expression is used to calculate on a
-  subset/reorder of rows from \code{i} grouped by \code{k}.} which will compute the
-number of flights done by American Airlines for each origin airport.
-Essentially, it uses \code{substitute} to capture the operand's expressions,
-transforms then appropriately and finally uses \eval to execute the query.
-
-Another example is the \code{tidyselect} package that defines a DSL for selecting
-columns in a data frame powering number of data manipulation operation in the
-\code{data.table} main competitor, the \code{dply} package. It uses eval to build an
-interpreter for evaluating column selectors such as \code{select(df,
-starts\_with("length") \& !(name:mass))} which returns all columns in \code{df}
-that starts with the string \code{length} and is not between \code{name} and
-\code{mass} columns.
-
-Another instance is the \emph{symbolic differentiation support} provided by base R package \emph{stats}, with two operators \code{D} and \code{deriv}.
-They support arithmetic operations and functions on real numbers such as \code{sin}, on expressions built with \code{expression()} or \code{parse}.
-In \emph{MCMCglmm} (Monte Carl Markov Chain Generalised Linear Mixed Models), a \code{Dtensor} object is created: it contains expressions that are derived twice (with \code{DD}).
-To get a concrete value, it is evaluated in  dataframe \code{mu} where symbolic variables are bound to actual numbers.
-\begin{lstlisting}
- expr<-expression(beta_1 + time*beta_2+u)
- mu = data.frame(beta_1=0.5, beta_2=1, time=3, u=2.3)
- D[i]<-eval(DD(expr, name[unlist(comb.pos[i,])]), mu)
-\end{lstlisting}
+The \code{glue} function extracts snippets of R code enclosed between \c{\{} and
+\c{\}}, evaluates them using \eval, and splices the result to construct the
+final string.
+
+Another example is the \code{data.table} package that defines a DSL for data
+frame manipulation using the subset operator \code{[}. This subset operation is
+defined as \code{DT[i, j, k]}, where expression \code{j} is used to calculate on
+a subset/reorder of rows from \code{i} grouped by \code{k}. This enables compact
+queries on data frames such as \code{flights[carrier == "AA", .N, by = origin]}
+which will compute the number of flights done by American Airlines (\code{"AA"})
+for each origin airport. At its core, \code{[} uses \code{substitute} to capture
+the operand expressions, transforms them, and evaluates them using \eval in an
+environment which has data frame columns available as bindings.
 
@@
 
-\mypara{Formula Resolution} Many packages use \code{eval} to evaluate formula
+\paragraph{Formula Resolution} Many packages use \code{eval} to evaluate formula
 expressions. Consider the definition of the \code{formula.pdMat} function from
@@
 
-\mypara{Code Obfuscation} Some packages use \eval to bypass the restrictions
+\paragraph{Code Obfuscation} Some packages use \eval to bypass the restrictions
 imposed by the \code{R CMD CHECK} tool. This tool enforces some well-formedness
@@
 
-\mypara{Miscellaneous}
-
-\mypara{Lookup and indexing}
+\paragraph{Miscellaneous}
+
+\paragraph{Lookup and indexing}
 

### 2021-08-10 paper/main.tex
Add miscellaneous section

@@
 
-\mypara{Domain-Specific Languages} The combination of laziness, metaprogramming,
+\paragraph{Domain-Specific Languages} The combination of laziness, metaprogramming,
 first-class environments and dynamic evaluation makes R the perfect toolkit for
@@
 
-\mypara{Boilerplate Removal} The combination of \code{match.call} and \code{eval}
+\paragraph{Boilerplate Removal} The combination of \code{match.call} and \code{eval}
 is frequently used by modeling functions to avoid boilerplate code and ease
@@
 
-\paragraph{Miscellaneous}
-
-\paragraph{Lookup and indexing}
-
-For example, the \code{PopED::get\_rse} uses it to
-assign a number of variables in the current scope:
-\code{eval(parse(text=paste(default_args[[i]]), "<- ", i))}. There are also more
-sophisticated cases. The \code{plyr::allocate\_column} function uses eval for a
-deferred assignment to fill missing values in a column. Package overture
-repeatedly evaluates R expressions capturing any assignments that happen in it
-as samples of Markov chains.
+\paragraph{Miscellaneous} Many uses of \eval are diverse and do not fit into a
+category by themselves. For example, packages for unit testing (testthat,
+testit, etc.), code benchmarking (rbenchmark, microbenchmark, etc.), and,
+running code in parallel (foreach, doParallel, etc.), use \eval to evaluate
+program fragments in a particular way in specific environments. Many numerical
+computation and mathematical modeling packages use \eval to evaluate chunks of
+code or bodies of functions in specific environments. Consider the
+\code{findZeros} function from \code{mosaic} package which is used to compute
+the zero of a function.
+
+\begin{figure}[h]
+\begin{lstlisting}
+findZeros <- function(expr, ..., xlim=c(near-within, near+within), near=0, within=Inf) {
+  dots <- list(...)
+
+  pfun <- function(x){  # removed . from name, was .x
+    mydots <- dots
+    mydots[[rhsVars]] <- x
+    eval( lhs(expr), envir=mydots, enclos=parent.frame() )
+  }
+  searchx <- sort(unique(internal.near   + c(0, leftseq, middleseq, rightseq)))
+  y <- sapply( searchx, pfun )
+\end{lstlisting}
+  \caption{mosaic::findZeros}
+\end{figure}
+
+This function can be invoked as: \code{findZeros( A*sin(2*pi*t/P) ~ t,
+  xlim=c(0,100), P=50, A=2)}. It will evaluate \code{A*sin(2*pi*t/P)} inside
+\code{pfun}. It calls \code{pfun} repeatedly with different values of \code{t}
+to find the point at which the expression is zero.
 

### 2021-08-10 paper/main.tex
Add more info about side-effects

--- 2021-08-10 paper/main.tex
@@
 
+Out of the minimized expressions, it is \c{BLOCK}, \c{<-} and
+\c{F(F(X)} that contribute to the vast majority of side effects. Concretely,
+92\% of \c{BLOCK}, 87\% of \c{<-} and 15\% of \c{F(F(X))} expressions contain
+side effects. In the case of \c{BLOCK}, most of the side effects happen in the
+environment passed to \eval (which by default is the parent environment, \ie
+the local or function in our classification). For \c{<-} it is 15\%, however,
+that is only because of the \c{plyr::allocate\_column} which does the side
+effect in the environment containing the data frame that is the target for the
+newly allocated column. Without this call, \c{<-} is similarly predictable with
+96\% of side effects in the passed environment. For \c{F(F(X))} minimization
+for, it is less clear as 48\% of side effects happen in a different environment
+to the one that was passed to \eval.
+
 \mypara{Discussion} In JavaScript assignments in \eval can happen in either

### 2021-08-10 paper/main.tex
4

--- 2021-08-10 paper/main.tex
@@
 
+% TODO do assignments change types?
+
 \begin{table}[h]
@@
 
-Out of the minimized expressions, it is \c{BLOCK}, \c{<-} and
-\c{F(F(X)} that contribute to the vast majority of side effects. Concretely,
-92\% of \c{BLOCK}, 87\% of \c{<-} and 15\% of \c{F(F(X))} expressions contain
-side effects. In the case of \c{BLOCK}, most of the side effects happen in the
-environment passed to \eval (which by default is the parent environment, \ie
-the local or function in our classification). For \c{<-} it is 15\%, however,
-that is only because of the \c{plyr::allocate\_column} which does the side
-effect in the environment containing the data frame that is the target for the
-newly allocated column. Without this call, \c{<-} is similarly predictable with
-96\% of side effects in the passed environment. For \c{F(F(X))} minimization
-for, it is less clear as 48\% of side effects happen in a different environment
-to the one that was passed to \eval.
-
 \paragraph{Discussion} In JavaScript assignments in \eval can happen in either
@@
 .makeCppMethods <- function(methods, cppclassinfo, env) {
-    ...
+    cppMethods <- names(cppclassinfo@methods)
+    newMethods <- names(methods)
     for(what in cppMethods[! cppMethods %in% newMethods])
         methods[[what]] <- eval(substitute(
-                  function(...) .CppObject$WHAT(...), list(WHAT = as.name(what))), env)
-    ...
+                  function(...) .CppObject$WHAT(...), list(WHAT = as.name(what))),
+                                env)
+    methods
 }
@@
 
-\mypara{Domain-Specific Languages} The combination of laziness, metaprogramming,
-first-class environments and dynamic evaluation makes R the perfect toolkit for
+\paragraph{Domain-Specific Languages} The combination of laziness, metaprogramming,
+first-class environments and dynamic evaluation make R the perfect toolkit for
 building DSLs. An example of this is the \code{glue} package which provides
 functions for string interpolation. The code snippet below shows an example of
-\code{glue} function in action.
+this.
 
@@
 
-The \code{glue} function extracts snippets of R code enclosed between \c{\{} and
-\c{\}}, evaluates them using \eval, and splices the result to construct the
-final string.
-
-Another example is the \code{data.table} package that defines a DSL for data
-frame manipulation using the subset operator \code{[}. This subset operation is
-defined as \code{DT[i, j, k]}, where expression \code{j} is used to calculate on
-a subset/reorder of rows from \code{i} grouped by \code{k}. This enables compact
-queries on data frames such as \code{flights[carrier == "AA", .N, by = origin]}
-which will compute the number of flights done by American Airlines (\code{"AA"})
-for each origin airport. At its core, \code{[} uses \code{substitute} to capture
-the operand expressions, transforms them, and evaluates them using \eval in an
-environment which has data frame columns available as bindings.
-
-\mypara{Boilerplate Removal} The combination of \code{match.call} and \code{eval}
+text enclosed by c{\{} and c{\}} is
+evaluated using \eval as an R expression and replaced by its result.
+
+The ability to capture the current call as a language object (\code{match.call()}),
+extract the unevaluated expression from function arguments (\code{substitute()})
+and \eval with first class environment makes R the toolkit for building DSLs.
+There are numerous widely spread DSLs used in R. For example, \code{data.table}
+package defines a DSL for data frame manipulation using the subset operator
+\code{[}. This allows one to write compact queries such as \code{flights[carrier ==
+  "AA", .N, by = origin]}\footnote{The \code{data.table} subset operation is
+  defined as \code{DT[i, j, k]}, where \code{j} expression is used to calculate on a
+  subset/reorder of rows from \code{i} grouped by \code{k}.} which will compute the
+number of flights done by American Airlines for each origin airport.
+Essentially, it uses \code{substitute} to capture the operand's expressions,
+transforms then appropriately and finally uses \eval to execute the query.
+
+Another example is the \code{tidyselect} package that defines a DSL for selecting
+columns in a data frame powering number of data manipulation operation in the
+\code{data.table} main competitor, the \code{dply} package. It uses eval to build an
+interpreter for evaluating column selectors such as \code{select(df,
+starts\_with("length") \& !(name:mass))} which returns all columns in \code{df}
+that starts with the string \code{length} and is not between \code{name} and
+\code{mass} columns.
+
+Another instance is the \emph{symbolic differentiation support} provided by base R package \emph{stats}, with two operators \code{D} and \code{deriv}.
+They support arithmetic operations and functions on real numbers such as \code{sin}, on expressions built with \code{expression()} or \code{parse}.
+In \emph{MCMCglmm} (Monte Carl Markov Chain Generalised Linear Mixed Models), a \code{Dtensor} object is created: it contains expressions that are derived twice (with \code{DD}).
+To get a concrete value, it is evaluated in  dataframe \code{mu} where symbolic variables are bound to actual numbers.
+\begin{lstlisting}
+ expr<-expression(beta_1 + time*beta_2+u)
+ mu = data.frame(beta_1=0.5, beta_2=1, time=3, u=2.3)
+ D[i]<-eval(DD(expr, name[unlist(comb.pos[i,])]), mu)
+\end{lstlisting}
+
+\paragraph{Boilerplate Removal} The combination of \code{match.call} and \code{eval}
 is frequently used by modeling functions to avoid boilerplate code and ease

### 2021-08-10 paper/main.tex

@@
 
+Out of the minimized expressions, it is \c{BLOCK}, \c{<-} and
+\c{F(F(X)} that contribute to the vast majority of side effects. Concretely,
+92\% of \c{BLOCK}, 87\% of \c{<-} and 15\% of \c{F(F(X))} expressions contain
+side effects. In the case of \c{BLOCK}, most of the side effects happen in the
+environment passed to \eval (which by default is the parent environment, \ie
+the local or function in our classification). For \c{<-} it is 15\%, however,
+that is only because of the \c{plyr::allocate\_column} which does the side
+effect in the environment containing the data frame that is the target for the
+newly allocated column. Without this call, \c{<-} is similarly predictable with
+96\% of side effects in the passed environment. For \c{F(F(X))} minimization
+for, it is less clear as 48\% of side effects happen in a different environment
+to the one that was passed to \eval.
+
 \paragraph{Discussion} In JavaScript assignments in \eval can happen in either

### 2021-08-10 paper/main.tex
changes to 4

@@
 \medskip
+  \medskip
 
@@
 
-\begin{wrapfigure}{4}{5.8cm}  \small  \vspace*{-2mm}\centering
+Our pipeline runs \CranRunnableScripts scripts extracted from \CranPackages
+packages. Any run that does not call \eval is discarded, leaving us with
+\packageNbruns runs. In these runs, \packageAllcalls \eval calls were recorded
+originating from \PkgHitEvalCallSites sites. The runs exercised
+\PkgHitEvalCallSitesAvgRatio of sites, a ratio similar to the package code
+coverage metric which is \PkgCodeCoverage. The reasons some sites are not
+exercised can be chalked down to incomplete tests and analysis failures
+(\PkgFailedProgramsRatio of the runs crashed or timed out).
+Fig.~\ref{fig:traced-eval-callsites} plots exercised site with respect to all
+sites for each program; as can be seen coverage is unequal.
+
+Fig.~\ref{freq} summarizes the frequency of calls, call counts on the left and
+number of packages on the right. There are \packageFewcalls packages with fewer
+than 100 calls, and \packageManycalls packages with more than 1,000 calls.
+\packageMaxcallspack calls \eval \packageMaxcalls times and thus accounts for
+over half of our observations.
+
+Fig.~\ref{tab:variantseval} summarizes use of variants, the first row has sites
+(\emph{static}), the second has sites encountered during analysis
+(\emph{exercised}) and the last gives calls (\emph{invocations}). Most sites and
+calls go to \eval itself, \c{eval.parent} is rare, and both \c{evalq} and
+\c{local} are barely used at all.
+
+\begin{wrapfigure}{r}{5cm}  \small  \vspace*{-2mm}\centering
   \begin{tabular}{|r@{\,}r@{\,}l@{\,}r|r@{\,}r@{\,}l@{}r|} \hline
@@
 \end{tabular}
-
-  \medskip  (a) All  \medskip  \medskip
-
-  \includegraphics[width=5.6cm, trim=5.5cm 0 5cm 0, clip]{package_calls_per_run_per_call_site}
-
-  (b) Small
-
 \caption{Normalized calls} \label{cn}
-
-\medskip\medskip
-\begin{tabular}{c}
-{\hspace{-25mm}\includegraphics[width=0.7\textwidth]{package_size_loaded_distribution}}
-\end{tabular}
-\caption{Loaded code} \label{fig:sizedistribution}
 \end{wrapfigure}
 
-Our pipeline runs \CranRunnableScripts scripts extracted from \CranPackages
-packages. Any run that does not call \eval is discarded, leaving us with
-\packageNbruns runs. We recorded \packageAllcalls \eval calls originating from
-\PkgHitEvalCallSites sites. This exercised \PkgHitEvalCallSitesAvgRatio of
-sites, the coverage is similar to the package code coverage metric which is
-\PkgCodeCoverage. The reason some sites are not exercised can be chalked down to
-incomplete tests and analysis failures (\PkgFailedProgramsRatio of all scripts
-crashed or timed out). Fig~\ref{fig:traced-eval-callsites} plots exercised site
-with respect to all sites for each program; as can be seen coverage is unequal.
-Figure~\ref{freq} summarizes the frequency of calls, number of calls on the left
-and number of packages on the right. There are \packageFewcalls packages with
-fewer than 100 calls, and \packageManycalls packages with more than 1,000 calls.
-\packageMaxcallspack calls \eval \packageMaxcalls times and thus accounts for
-over half of the observations. Fig.~\ref{tab:variantseval} summarizes use of
-variants, the first row has sites (\emph{static}), the second has sites
-encountered during analysis (\emph{exercised}) and the last gives calls
-(\emph{invocations}). Most sites and calls go to \eval itself, \c{eval.parent}
-is rare, and both \c{evalq} and \c{local} are barely used at all.
-
-Figure~\ref{cn}(a) shows average count of calls from a given site and given run
-and how many sites fall in that range. For instance, \packageRunbinh sites are
-called over 2,000 times per run. Larger numbers suggest loops or recursive
-contexts -- this seems to be the exception; \packageRunbina of sites are called
-fewer than 50 times. Fig.~\ref{cn}(b) zooms in on low frequency sites. The
-x-axis shows normalized calls, and the y-axis counts sites. Most sites are
-invoked once, half as many are invoked twice, the drop off is sharp after that.
+Fig.~\ref{cn} shows average count of calls from a given site and given run and
+how many sites fall in that range. For instance, \packageRunbinh sites are
+called over 2,000 times per run. Larger call counts suggest the presence of
+loops or recursive functions, but given the data this seems to be the exception;
+\packageRunbina of sites are called fewer than 50 times, most sites are invoked
+once and half as many are invoked twice.
 
@@
 accounts for \packageGgplotsymbolpercent of symbols and comes from one site in
-\c{ggplot2}.\footnote{To model inheritance in \c{ggproto}, one of the many
-object-oriented systems for R.}
+\c{ggplot2}.\footnote{Each object has a field model inheritance, this comes from
+\c{ggproto}, one of the many object-oriented systems for R.}
+
+\begin{wrapfigure}{r}{5.8cm}
+\begin{tabular}{c}
+{\hspace{-25mm}\includegraphics[width=0.7\textwidth]{package_size_loaded_distribution}}
+\end{tabular}
+\caption{Loaded code} \label{fig:sizedistribution}
+\end{wrapfigure}
 
 To estimate how much executable code is injected through \eval, we measure the
-number of nodes in the expressions; for example, \c{x+1} counts as 3. We also
-measured string lengths of unparsed expressions, but these measurements were
-dominated by the size of data objects which could range in the MBs. The median
-argument size is \packageMedianszeval (due to the symbols), and the average is
-\packageAvgszeval nodes. The largest \eval input observed is
-\packageMaxszeval\,-- a significant chunk of code.
-Figure~\ref{fig:sizedistribution} shows the distribution of sizes for arguments
-of fewer or equal to 25 nodes. The x-axis is the size of arguments in number of
-nodes, and the y-axis is the count of arguments with that size. The size drops
-rapidly, with few observations larger than 15 nodes. The long tail is omitted
-for legibility.
-
-To estimate the work performed in \evals, we count instructions executed by the
-interpreter. Most invocations perform relatively little work, with
-\packageSmalleventspct of \evals executing 50 or fewer instructions. The violin
-plot of Figure~\ref{ev}(a) corresponds to \evals executing $\leq$ 50
-instructions; it is dominated by trivial symbol lookups. Figure~\ref{ev}(b) has
-the work-intensive \evals, which go all the way to \packageMaxeventsRnd
-instructions.
+size of the arguments in terms of number of nodes. For example,
+\c{expression(x+1)} has a size of 3. The large number of symbols skew the median
+to \packageMedianszeval but the average size is \packageAvgszeval nodes. The
+largest \eval input we observed was an expression of \packageMaxszeval nodes, a
+significant chunk of code. Fig.~\ref{fig:sizedistribution} shows the
+distribution of sizes for arguments of fewer than 25 nodes. Sizes drop rapidly,
+there are few observations larger than 15 nodes. The long tail is omitted for
+legibility. As an alternative to counting nodes, we tried measuring string
+lengths of expressions after calling a function to deparse values back to
+strings, but these measurements were dominated by massive data objects which
+could range in the MBs.
+
+To assess how much computational work is performed in \evals, we count the
+instructions executed by the interpreter. Most call do relatively little, with
+\packageSmalleventspct of calls executing fewer than 50 instructions. The violin
+plot of Fig.~\ref{ev}(a) shows the distribution of short running \evals; the
+data is dominated by \evals which simply perform symbol lookup. Fig.~\ref{ev}(b)
+shows the distribution of work-intensive \evals which go all the way to
+\packageMaxeventsRnd instructions.
 
@@
 \paragraph{Discussion.}
-\Eval is used in CRAN packages. Nearly every fourth package contains at least
-one \eval site. However, most packages that use \eval do that modestly. Half of
-them include less than three \eval sites. Packages that rely heavily on \eval
-are mostly the ones that provide statistical modeling and simulation functions.
-At runtime, over a third of extracted programs triggered calls to \eval. Half
-of the \packageAllcallsRnd recorded calls come from a single \eval site, but
-for the rest, most of the packages have low \eval call frequency.
+\Eval is widely used in CRAN packages. Nearly every fourth package contains at
+least one \eval call site. However, most packages use \eval modestly. Packages
+that rely heavily on \eval are mostly ones that provide statistical modeling and
+simulation functions. At runtime, over a third of our scripts triggered calls to
+\eval, note that all scripts performed \eval originating from base package but
+these are disregarded here. Half of the calls come from a single site, for the
+rest, most packages have low call frequency. The data is consistent with usage
+of \eval for configuration and meta-programming and not in performance sensitive
+contexts such as loops. While most arguments to \eval are small, often just a
+variable name, and do little work, we have also observed large \eval arguments
+and arguments that perform massive amounts of work.
 
@@
 \paragraph{Domain-Specific Languages} The combination of laziness, metaprogramming,
-first-class environments and dynamic evaluation make R the perfect toolkit for
+first-class environments and dynamic evaluation makes R the perfect toolkit for
 building DSLs. An example of this is the \code{glue} package which provides
 functions for string interpolation. The code snippet below shows an example of
-this.
+\code{glue} function in action.
 

### 2021-08-10 paper/main.tex
Change table presentation

@@
 
-\begin{figure}[!h]\center\small\begin{tabular}{r@{~}l|l|l|l}\hline
-\tiny\sc Language&&\sc\tiny Input&\sc\tiny Scope&\tiny\sc Reflective operations\\\hline
+\begin{figure}[!h]\center\small\begin{tabular}{r@{~}llll}\toprule
+\tiny\sc Language&&\sc\tiny Input&\sc\tiny Scope&\tiny\sc Reflective operations\\\midrule
 \bf Julia&\cite{julia}     & expression& toplevel         & data\\
@@
 \bf JavaScript&\cite{ecoop11}& text      & current, toplevel& data\\
-\bf R&\cite{R96}  & expression& programmatic      & data, stack, environment\\\hline
+\bf R&\cite{R96}  & expression& programmatic      & data, stack, environment\\\bottomrule
 \end{tabular}\caption{Design space of \eval}\label{comp}
@@
 \begin{minipage} {5cm}\small
-  \begin{tabular}{|r@{\,}r@{\,}l@{}r|r@{\,}r@{\,}l@{}r|} \hline
-    \multicolumn{3}{|c}{\small\#calls} &\small \#pck
-&     \multicolumn{3}{c}{\small\#calls} &\small\#pck \\\hline
+  \begin{tabular}{r@{\,}r@{\,}l@{}r|r@{\,}r@{\,}l@{}r} \toprule
+    \multicolumn{3}{c}{\bf \small\#calls} &\bf \small \#pck
+&     \multicolumn{3}{c}{\bf \small\#calls} &\bf \small\#pck \\\midrule
 \tt 1 &--& \tt 10      & \packageBina  & \tt 10K &--&\tt 100K  & \packageBine\\
@@
 \tt 101 &--& \tt 1K    & \packageBinc  & \tt 1M &--&\tt 10M   & \packageBing\\
-\tt 1K &--& \tt 10K    & \packageBind  & \tt 10M &--& \tt 100M & \packageBinh\\\hline
+\tt 1K &--& \tt 10K    & \packageBind  & \tt 10M &--& \tt 100M & \packageBinh\\\bottomrule
 \end{tabular}
@@
 \begin{minipage}{7cm}\small
-\begin{tabular}{|@{\,}r|rrrr|}\hline
-  &\eval & \c{evalq} & \c{eval} & \c{local}\\[-1.5mm]
-           & & & \c{.parent} &\\\hline
+\begin{tabular}{@{\,}r|rrrr}\toprule
+  &\bf \eval & \bf \c{\bf evalq} & \c{\bf eval} & \c{\bf local}\\[-1.5mm]
+           & & & \c{\bf .parent} &\\\midrule
 \small Static sites &\packageStaticeval&\packageStaticevalq&\packageStaticevalparent&\packageStaticlocal \\
 \small Exercised sites&\packageTriggeredeval&\packageTriggeredevalq&\packageTriggeredevalparent&\packageTriggeredlocal\\
-\small Invocations&\packageEvalsRnd&\packageEvalqsRnd&\packageEparentsRnd&\packageLocalsRnd\\\hline
+\small Invocations&\packageEvalsRnd&\packageEvalqsRnd&\packageEparentsRnd&\packageLocalsRnd\\\bottomrule
 \end{tabular}~\\[2mm]\caption{Variants}\label{tab:variantseval}
@@
 sites for each program; as can be seen coverage is unequal.
-
 Fig.~\ref{freq} summarizes the frequency of calls, call counts on the left and
@@
 over half of our observations.
-
 Fig.~\ref{tab:variantseval} summarizes use of variants, the first row has sites
@@
 \begin{wrapfigure}{r}{5cm}  \small  \vspace*{-2mm}\centering
-  \begin{tabular}{|r@{\,}r@{\,}l@{\,}r|r@{\,}r@{\,}l@{}r|} \hline
-\multicolumn{3}{|c}{\#calls} & \#sites &
-\multicolumn{3}{c}{\#calls} & \#sites \\\hline
+  \begin{tabular}{r@{\,}r@{\,}l@{\,}r|r@{\,}r@{\,}l@{}r} \toprule
+    \multicolumn{3}{c}{\bf \#calls} & \bf \#sites &
+     \multicolumn{3}{c}{\bf \#calls} & \bf \#sites \\\midrule
     \tt 0 &--& \tt 50    & \packageRunbina & \tt 501 &--& \tt 1000   & \packageRunbine\\
@@
     \tt 101 &--& \tt 250 & \packageRunbinc & \tt 1501 &--& \tt 2000  & \packageRunbing\\
-\tt 251 &--& \tt 500 & \packageRunbind & \tt 2001 &--& \tt 3000 & \packageRunbinh\\\hline
+    \tt 251 &--& \tt 500 & \packageRunbind & \tt 2001 &--& \tt 3000 & \packageRunbinh\\\bottomrule
   \end{tabular}
@@
 \begin{table}[h]\small
-\begin{tabular}{|c|r|r|r|r|r|c|}\hline
-  $min(e)$& \#sites & \%sites & \#packages & \#operations & \%envir & example\\\hline
-\c X&\packageMinimizedcallsitesa &\packageMinimizedpropsitesa &\packageMinimizedpackagea &\packageMinimizedmedianoperationsaRnd &\packageMinimizedpercentparentframesa & \c{y+1}\\\hline
-\c{F(F(X))} & \packageMinimizedcallsitesb  & \packageMinimizedpropsitesb & \packageMinimizedpackageb  & \packageMinimizedmedianoperationsbRnd & \packageMinimizedpercentparentframesb & \c{gbov( mean(x), a-1)}\\\hline
-\c{V}&\packageMinimizedcallsitesc &\packageMinimizedpropsitesc &\packageMinimizedpackagec &\packageMinimizedmedianoperationscRnd &\packageMinimizedpercentparentframesc& \c{c(42,21,0)}\\\hline
-\c{F(X)}& \packageMinimizedcallsitesd & \packageMinimizedpropsitesd & \packageMinimizedpackaged & \packageMinimizedmedianoperationsdRnd & \packageMinimizedpercentparentframesd & \c{seq\_len(iters)} \\\hline
-\c{\$} & \packageMinimizedcallsitese & \packageMinimizedpropsitese & \packageMinimizedpackagee & \packageMinimizedmedianoperationseRnd & \packageMinimizedpercentparentframese & \c{DF\$B}\\\hline
-\c{model.frame}& \packageMinimizedcallsitesf & \packageMinimizedpropsitesf & \packageMinimizedpackagef & \packageMinimizedmedianoperationsfRnd & \packageMinimizedpercentparentframesf &  \c{model.frame(formula = Z $\sim$ U)}   \\\hline
-\c{F()}& \packageMinimizedcallsitesg & \packageMinimizedpropsitesg & \packageMinimizedpackageg & \packageMinimizedmedianoperationsgRnd & \packageMinimizedpercentparentframesg & \c{rgamma(3, 2, n = 10L)} \\\hline
-\c{FUN} & \packageMinimizedcallsitesh & \packageMinimizedpropsitesh & \packageMinimizedpackageh & \packageMinimizedmedianoperationshRnd & \packageMinimizedpercentparentframesh & \c{function(x, y) x + 3 * y} \\\hline
-\c{<-} & \packageMinimizedcallsitesi  & \packageMinimizedpropsitesi & \packageMinimizedpackagei & \packageMinimizedmedianoperationsiRnd & \packageMinimizedpercentparentframesi & \c{x[1, 2:3, 2:3] <- value}\\\hline
-\c{BLOCK} & \packageMinimizedcallsitesj & \packageMinimizedpropsitesj & \packageMinimizedpackagej & \packageMinimizedmedianoperationsjRnd & \packageMinimizedpercentparentframesj & \c{\{ write.csv(iris, tf) ; file.size(tf) \}} \\\hline
+\begin{tabular}{crrrrrc}\toprule
+  \bf $min(e)$& \bf \#sites & \bf \%sites & \bf \#packages & \bf \#operations & \bf \%envir & \bf example\\\midrule
+\c X&\packageMinimizedcallsitesa &\packageMinimizedpropsitesa &\packageMinimizedpackagea &\packageMinimizedmedianoperationsaRnd &\packageMinimizedpercentparentframesa & \c{y+1}\\
+\c{F(F(X))} & \packageMinimizedcallsitesb  & \packageMinimizedpropsitesb & \packageMinimizedpackageb  & \packageMinimizedmedianoperationsbRnd & \packageMinimizedpercentparentframesb & \c{gbov( mean(x), a-1)}\\
+\c{V}&\packageMinimizedcallsitesc &\packageMinimizedpropsitesc &\packageMinimizedpackagec &\packageMinimizedmedianoperationscRnd &\packageMinimizedpercentparentframesc& \c{c(42,21,0)}\\
+\c{F(X)}& \packageMinimizedcallsitesd & \packageMinimizedpropsitesd & \packageMinimizedpackaged & \packageMinimizedmedianoperationsdRnd & \packageMinimizedpercentparentframesd & \c{seq\_len(iters)} \\
+\c{\$} & \packageMinimizedcallsitese & \packageMinimizedpropsitese & \packageMinimizedpackagee & \packageMinimizedmedianoperationseRnd & \packageMinimizedpercentparentframese & \c{DF\$B}\\
+\c{model.frame}& \packageMinimizedcallsitesf & \packageMinimizedpropsitesf & \packageMinimizedpackagef & \packageMinimizedmedianoperationsfRnd & \packageMinimizedpercentparentframesf &  \c{model.frame(formula = Z $\sim$ U)}   \\
+\c{F()}& \packageMinimizedcallsitesg & \packageMinimizedpropsitesg & \packageMinimizedpackageg & \packageMinimizedmedianoperationsgRnd & \packageMinimizedpercentparentframesg & \c{rgamma(3, 2, n = 10L)} \\
+\c{FUN} & \packageMinimizedcallsitesh & \packageMinimizedpropsitesh & \packageMinimizedpackageh & \packageMinimizedmedianoperationshRnd & \packageMinimizedpercentparentframesh & \c{function(x, y) x + 3 * y} \\
+\c{<-} & \packageMinimizedcallsitesi  & \packageMinimizedpropsitesi & \packageMinimizedpackagei & \packageMinimizedmedianoperationsiRnd & \packageMinimizedpercentparentframesi & \c{x[1, 2:3, 2:3] <- value}\\
+\c{BLOCK} & \packageMinimizedcallsitesj & \packageMinimizedpropsitesj & \packageMinimizedpackagej & \packageMinimizedmedianoperationsjRnd & \packageMinimizedpercentparentframesj & \c{\{ write.csv(iris, tf) ; file.size(tf) \}} \\\bottomrule
 \end{tabular}
@@
 \begin{minipage}{3.7cm}
-  \begin{tabular}{@{}r|r|r@{}}\hline
- Kind & \#sites & \%sites \\\hline
+  \begin{tabular}{@{}rrr@{}}\toprule
+ \bf Kind & \bf \#sites & \bf \%sites \\\midrule
  Function & \packageNbFunctionEnvSites &  \packageNbFunctionEnvSitePercent\\
@@
  Global &  \packageNbStrictGlobalEnvSites & \packageNbStrictGlobalEnvSitePercent \\
- Package & \packageNbPackageNamespaceEnvSites & \packageNbPackageNamespaceEnvSitePercent \\\hline
+ Package & \packageNbPackageNamespaceEnvSites & \packageNbPackageNamespaceEnvSitePercent \\\bottomrule
 \end{tabular}
@@
 \begin{minipage}{3.7cm}\centering
-\begin{tabular}{@{}r|r|r@{}}\hline
- Offset & \#sites & \%sites \\\hline
+  \begin{tabular}{@{}rrr@{}}\toprule
+ \bf Offset & \bf \#sites & \bf \%sites \\\midrule
   \packageCallerEnvHierarchyNamea & \packageCallerEnvHierarchySitesaRnd & \packageCallerEnvHierarchySitePercenta \\
@@
   \packageCallerEnvHierarchyNamec& \packageCallerEnvHierarchySitescRnd & \packageCallerEnvHierarchySitePercentc  \\
-$\ge 3$& \packageNbFarAwayCallerSites &  \packageNbFarAwayCallerSitePercent \\\hline
+$\ge 3$& \packageNbFarAwayCallerSites &  \packageNbFarAwayCallerSitePercent \\\bottomrule
  \end{tabular}
@@
 \begin{minipage}{3.7cm}
-\begin{tabular}{@{}r|r|r@{}} \hline
-Parent & \#sites & \%sites \\\hline
+  \begin{tabular}{@{}rrr@{}} \toprule
+\bf Parent & \bf \#sites & \bf \%sites \\\midrule
 Function & \packageNewEnvCategorySitesa & \packageNewEnvCategorySitePercenta \\
@@
 Global & \packageNewEnvCategorySitesc & \packageNewEnvCategorySitePercentc \\
-Empty & \packageNewEnvCategorySitesd & \packageNewEnvCategorySitePercentd \\\hline
+    Empty & \packageNewEnvCategorySitesd & \packageNewEnvCategorySitePercentd \\\bottomrule
 \end{tabular}
@@
 \begin{minipage}{3.7cm}\centering
- \begin{tabular}{@{}c|c|c@{}} \hline
- \#kinds & \#sites &  \%sites \\ \hline
+  \begin{tabular}{@{}ccc@{}} \toprule
+ \bf \#kinds & \bf \#sites &  \bf \%sites \\\midrule
  \packageNbCategoryEnvira & \packageNbCategoryEnvirSitesaRnd &  \packageNbCategoryEnvirPercenta\\
@@
  \packageNbCategoryEnvirc & \packageNbCategoryEnvirSitescRnd &  \packageNbCategoryEnvirPercentc\\
- \packageNbCategoryEnvird & \packageNbCategoryEnvirSitesdRnd & \packageNbCategoryEnvirPercentd\\\hline
+ \packageNbCategoryEnvird & \packageNbCategoryEnvirSitesdRnd & \packageNbCategoryEnvirPercentd\\\bottomrule
 \end{tabular}\caption{Multiplicities}\label{tab:polyenvir}
@@
 \begin{wraptable}{r}{5cm}\small\centering
-	\begin{tabular}{|c|c|c|}
-		\hline
-		Provenance & \#sites & \%sites \\
-		\hline
+	\begin{tabular}{lcc}
+		\toprule
+		\bf Provenance & \bf \#sites & \bf \%sites \\
+		\midrule
 		\packageProvenanceNamea & \packageNbProvenanceSitesa & \packagePercentProvenanceSitesa \\
@@
 		\packageProvenanceNamen & \packageNbProvenanceSitesn & \packagePercentProvenanceSitesn \\
-		\hline
+		\bottomrule
 	\end{tabular}
@@
 \begin{wraptable}[8]{r}{5cm}\small\centering
-\begin{tabular}{r|r|r} \hline
-Origin  & \#sites & \%sites \\\hline
+\begin{tabular}{rrr} \toprule
+\bf Origin  & \bf \#sites & \bf \%sites \\\midrule
 Reflection &  \packageNbReflectionSites & \packageReflectionSitesPercent\\
@@
 Environment & \packageNbSymbolSites & \packageSymbolSitesPercent \\
-External & \packageNbExternalSites & \packageExternalSitesPercent \\\hline
+External & \packageNbExternalSites & \packageExternalSitesPercent \\\bottomrule
 \end{tabular}
@@
 \begin{wraptable}{r}{7cm}\small\centering
-  \begin{tabular}{l|r|r|r|r}\hline
-    Environments & \#sites & \%sites & \#funs. & \%funs. \\%
+  \begin{tabular}{lrrrr}\toprule
+    \bf Environments & \bf \#sites & \bf \%sites & \bf \#funs. & \bf \%funs. \\%
     \expandableinput tag/table-se-target-envs.tex
@@
   \centering
-  \begin{tabular}{l|r|r|r|r|r|r}\hline
-    Side effect & \#events & \%events & \#calls & \%calls & \#sites & \%sites \\%
+  \begin{tabular}{lrrrrrr}\toprule
+    \bf Side effect & \bf \#events & \bf \%events & \bf \#calls & \bf \%calls & \bf \#sites & \bf \%sites \\%
     \expandableinput tag/table-se-types.tex

### 2021-08-10 paper/main.tex
minor

--- 2021-08-10 paper/main.tex
@@
 
-\begin{wrapfigure}{r}{5.8cm}
+\begin{wrapfigure}{r}{6.2cm}
 \begin{tabular}{c}
-{\hspace{-25mm}\includegraphics[width=0.7\textwidth]{package_size_loaded_distribution}}
+{\hspace{-25mm}\includegraphics[width=0.8\textwidth]{package_size_loaded_distribution}}
 \end{tabular}
@@
 
-The expressions passed to \eval vary widely. In order to categorize them, let us
-use a minimization function $min(e)$ which for a given expression $e$ returns a
-normal form that abstracts incidental details allowing the reader to focus on
-the structure of the evaluated code. The minimization function performs constant
-folding of arithmetic and string expressions for base operators, e.g.
-$min(\c{1+1})=\c{V}$, value simplification, $min(\c{c(1,2,3+2)})=\c{V}$,
-variable absorption, $min(\c{x+y})=\c{X}$, function absorption,
-$min(\c{g(f(x),h(z))})=\c{F(F(X))}$ and a number of other simplifications.
+The expressions passed to \eval vary widely. We categorize expressions using a
+minimization function which abstract some of the incidental details of
+expressions and return a ``shape'' that can be used to group similar \evals. The
+function $min(e)$ which for a given expression $e$ returns a normal form. Normal
+forms use \c V to stand in for values occurring in expression, \c X stands in
+for variables, and \c F stands for functions. The function performs constant
+folding of arithmetic and string expressions over base operators, for instance
+$min(\c{1+1})$ simplifies to $\c{V}$, on the ground that addition of values
+likely returns a value. Value simplification takes $min(\c{c(1,2,3+2)})$ to
+$\c{V}$ as a complex vector containing values is a value. Variable absorption
+has $min(\c{x+y})$ becomes $\c{X}$, this motivated by the fact that addition is
+not an ``interesting'' operation, and that \c X stands in for any number of
+variable lookups. Function absorption simplifies nested functions to keep an
+abstraction of the nesting $min(\c{g(f(x),h(z))})=\c{F(F(X))}$. There are other
+simplifications, the full list is in our artifact. It is worth mentioning that
+these simplifications are heuristics, in R \c{1+1} is not necessary \c 2. The
+addition operator can be redefined, but it typically isn't or at least not in a
+way that would invalidate arithmetic.
+
 Table~\ref{tab:minimizedexpressions} gives the 10 most frequent forms; \#sites

### 2021-08-10 paper/main.tex
5

@@
 
-The expressions passed to \eval vary widely. We categorize expressions using a
-minimization function which abstract some of the incidental details of
-expressions and return a ``shape'' that can be used to group similar \evals. The
-function $min(e)$ which for a given expression $e$ returns a normal form. Normal
-forms use \c V to stand in for values occurring in expression, \c X stands in
-for variables, and \c F stands for functions. The function performs constant
-folding of arithmetic and string expressions over base operators, for instance
-$min(\c{1+1})$ simplifies to $\c{V}$, on the ground that addition of values
-likely returns a value. Value simplification takes $min(\c{c(1,2,3+2)})$ to
-$\c{V}$ as a complex vector containing values is a value. Variable absorption
-has $min(\c{x+y})$ becomes $\c{X}$, this motivated by the fact that addition is
-not an ``interesting'' operation, and that \c X stands in for any number of
-variable lookups. Function absorption simplifies nested functions to keep an
-abstraction of the nesting $min(\c{g(f(x),h(z))})=\c{F(F(X))}$. There are other
-simplifications, the full list is in our artifact. It is worth mentioning that
-these simplifications are heuristics, in R \c{1+1} is not necessary \c 2. The
-addition operator can be redefined, but it typically isn't or at least not in a
-way that would invalidate arithmetic.
-
-Table~\ref{tab:minimizedexpressions} gives the 10 most frequent forms; \#sites
-and \%sites are, respectively, the number and ratio of sites receiving arguments
-of that form; \#packages is the number of packages with that form; \#operations
-is the median number of instructions performed by the interpreter; and \%envir
-is the ratio of sites that evaluate in a function environment. The example
-column shows one sample expression $e$ that normalizes to the particular form.
-
-\begin{table}[h]\small
+\begin{table}[!b]\small
 \begin{tabular}{crrrrrc}\toprule
@@
 
-\medskip  We detail these forms and discuss their implication for the behavior of \eval.
+The expressions passed to \eval vary widely. We categorize expressions using a
+minimization function which abstract some of the incidental details of
+expressions and return a ``shape'' that can be used to group similar \evals. The
+function $min(e)$ which for a given expression $e$ returns a normal form. Normal
+forms use \c V to stand in for values occurring in expression, \c X stands in
+for variables, and \c F stands for functions. The function performs constant
+folding of arithmetic and string expressions over base operators, for instance
+$min(\c{1+1})$ simplifies to $\c{V}$, on the ground that addition of values
+likely returns a value. Value simplification takes $min(\c{c(1,2,3+2)})$ to
+$\c{V}$ as a complex vector containing values is a value. Variable absorption
+has $min(\c{x+y})$ becomes $\c{X}$, this motivated by the fact that addition is
+not an ``interesting'' operation, and that \c X stands in for any number of
+variable lookups. Function absorption simplifies nested functions to keep an
+abstraction of the nesting $min(\c{g(f(x),h(z))})=\c{F(F(X))}$. There are other
+simplifications, the full list can be found in our artifact. It is worth
+mentioning that these simplifications are heuristics, in R \c{1+1} is not
+necessary \c 2. The addition operator can be redefined, but it typically isn't
+or at least not in a way that would invalidate arithmetic.
+
+Table~\ref{tab:minimizedexpressions} gives the 10 most frequent shapes with the
+number and ratio of sites that received arguments of that shape. Operations is
+the median count of instructions performed by the interpreter when evaluating
+argument of the given shape. Envir is the ratio of sites that evaluate in a
+function environment. The last column has a sample expression that normalizes to
+the corresponding shape. We detail these shapes and discuss their implication for
+the behavior of \eval.
 
@@
 
-\medskip\EE{$min(e)=\c{V}$} Expressions that represent values are
-frequently passed to \eval -- they occur in 17\% of sites. The majority of
-those, \packageValOneNodePercent, are inline constants (integer or double
-vectors). The rest trivially evaluate to a value, \eg~\c{1+1}.\footnote{True as
-long as base functions such as \c{+} are not redefined. They typically are not,
-but this is a limitation of this categorization.} In our corpus,
-\packageNbCallSitesUniqueActualValue call sites only ever see a simple value,
-such as:
+\begin{tabular}{@{}p{.97\linewidth}}
+  \medskip\EE{$min(e)=\c{V}$} \\[-2mm]\small Values occur in 16\% of sites, 74\% of
+  those are constants such as integers, the remainder evaluate to a value. \Eval
+  is needed when the expression denoting the values is constructed
+  programmatically. Cases when a value is directly given to \eval, and \eval
+  returns it unchanged, often occur when a computation has a default path and
+  another, more interesting, path that requires evaluation. The following is a
+  call site that always see a string; \packageNbCallSitesUniqueActualValue of
+  sites only ever see a value. This shape usually run in few interpreter steps.
+  The environment in which they evaluate is mostly irrelevant.
 \begin{lstlisting}
- f1 <- eval(paste("A~",paste(paste(names(X[,-1])),collapse="+")))
-\end{lstlisting}
-where the argument to \eval is a string which \eval simply returns. Of twenty
-randomly selected \evals, XXXXX
-%% UNFINED TODO \packageUsefulValueEvalPercent
-of them need \eval for
-some inputs. Either we are dealing with a value that needs to be constructed
-dynamically, or the value is a default case that sometimes is replaced by a more
-interesting expression. This form is usually evaluated in few interpreter steps;
-in fact, the median is only \packageMinimizedmedianoperationscRnd. The
-environment in which they evaluate is mostly irrelevant.
-
-\medskip\EE{$min(e)=\c{X}$} Variable lookups are the most common form;
-they are found in 28\% of the sites. This form includes simple variable reads,
-\eg~\c{x}, which are \packageNbSymbolVarSitePercent of \c X. The form also
-subsumes \c{V}, so it includes a mixture of arithmetic expressions,
-\eg~\c{x+y+1}. The operations allowed are limited to built-in arithmetics. It is
-noteworthy that, while most \c{X}\!s evaluate in a single step, the variable can
-be bound to a promise, and accessing it may trigger evaluation of that promise,
-thus resulting in an arbitrary amount of computation. The median number of
-interpreted operations is \packageMinimizedmedianoperationsaRnd, suggesting that
-it is not the common case. Lookups are often evaluated in constructed
-environments; as few as \packageMinimizedpercentparentframesa of these
+  eval(paste("A~",paste(paste(names(X[,-1])),collapse="+")))
+\end{lstlisting}
+\\
+\medskip\EE{$min(e)=\c{X}$}\\[-2mm]\small Variable lookups are the most common shape,
+\packageNbSymbolVarSitePercent are simple variable names, e.g. \c x. The shapes
+subsumes \c{V} and built-in arithmetic operations, so a mixture of arithmetics
+and variables normalize to \c X. A lookup is one step of execution, if a promise
+is returned an arbitrary number of additional steps may be needed; the median is
+\packageMinimizedmedianoperationsaRnd, suggesting that most variable read do
+little work. Variables are often evaluated in environments that have been
+constructed programmatically, \packageMinimizedpercentparentframesa of
 expressions are evaluated in a function environment.
-
-\medskip\EE{$min(e)=\c{\$}$} This form extends \c X to include lookup
-with the dollar operator, \eg~\c{x\$f}, and vector indexing, \eg~\c{x[42]} or
-\c{x[[24]]}. As with \c X, we allow arithmetics and values in this form. Lookup
-occurs in 10\% of the sites. The interpreter evaluates
-\packageMinimizedmedianoperationsgRnd operations on average; the minimum is 3
-operations. This is typically used in a function environment,
-\packageMinimizedpercentparentframese of the time to be precise.
-
-\medskip\EE{$min(e)=$~\c{<-}} This form includes both assignment
-operators, the direct assignment {\tt <-}, and assignment to the parent
-environment {\tt <\,\!<-}, the \c{assign} function and the \c{\$} form.
-Assignments occur in 5\% of the sites. They represent the most obvious source of
-side effects. The median count is \packageMinimizedmedianoperationsiRnd (min=3).
-
-\medskip\EE{$min(e)=\c{F()}$} This form captures simple function calls,
-\eg~\c{f(2)}, with neither variables nor assignments. More specifically, it
-allows for variables in the function position but not in arguments. Usually,
-looking up function names does not trigger computation, but that is not a given
-if the function is returned by a promise or if the function name is shadowed by
-a promise containing a value, then the computation will occur. This form occurs in
-6\% of sites and typically does not perform much work.
-
-\medskip\framebox{$min(e)=\c{F(X)}$}~\EE{$min(e)=\c{F(F(X))}$} These
-forms allow for function calls whose arguments may include variable references.
-The latter allows nested calls. Assignments are excluded from this form. They
-occur in, respectively, 14\% and 20\% of the sites in the corpus. Together they
-are the most frequent forms. The median numbers of interpreter steps are,
-respectively, \packageMinimizedmedianoperationsdRnd and
-\packageMinimizedmedianoperationsbRnd. Also,
-\packageMinimizedpercentparentframesd and \packageMinimizedpercentparentframesb
-of these expressions run in function environments.
-
-\medskip\EE{$min(e)=\c{FUN}$} This form captures expressions that
-define functions, \eg~\c{function(x) x+1}, and do nothing else. \c{FUN} occurs
-in \packageFunctionDefinitionSitesPercent of sites. Evaluating a function
-definition is done in 2 interpreter steps; the data does not record the work
-performed by the interpreter when the generated functions are eventually run. In
-addition to \c{FUN}, \packageGeneralizedFunctionDefinitionSitesPercent of sites
-have function definitions nested in other expressions. This gives an idea of the
-use of higher-order functions.
-
-\medskip\EE{$min(e)=\c{BLOCK}$} This form captures multi-statement
-code blocks, which occur in only 4\% of sites. These are huge expressions; we
-do not inspect the contents of the blocks. The median number of executed operations
-is \packageMinimizedmedianoperationsjRnd. They typically run in function
+\\
+\medskip\EE{$min(e)=\c{\$}$}\\[-2mm]\small This shape extends \c X to include lookup
+with the dollar operator, \eg~\c{x\$f}, and vector indexing, \eg~\c{x[42]} and
+\c{x[[24]]}. This takes a few more step of evaluation on average, and is
+typically used in a function environment.
+\\
+\medskip\EE{$min(e)=$~\c{<-}}\\[-2mm]\small This includes both direct assignment {\tt
+  <-} and parent environment assignment {\tt <\,\!<-} as well as the \c{assign}
+function. It subsumes \c{\$}. Assignments
+represent the most obvious source of side effects in \eval.
+\\
+\medskip\EE{$min(e)=\c{F()}$}\\[-2mm]\small Function calls without variable lookups or
+assignment. A variable is allowed the function position, looking up
+functions does not trigger computation unless the function's name
+is bound to a promise. This shape typically does not perform much work.
+\\
+\medskip\framebox{$min(e)=\c{F(X)}$}~\EE{$min(e)=\c{F(F(X))}$}\\[-2mm]\small Function
+calls whose arguments may include variable references and nested calls. Together
+they represent the most frequent shape and perform over 16 steps. They are
+frequently evaluated in function environments.
+\end{tabular}
+
+\begin{tabular}{@{}p{.97\linewidth}}
+  \medskip\EE{$min(e)=\c{FUN}$}\\[-2mm]\small Function definitions without any
+  computation, \eg~\c{function(x) x+1}. In addition to \c{FUN},
+  \packageGeneralizedFunctionDefinitionSitesPercent of sites have function
+  definitions nested in other expressions.
+\\
+\medskip\EE{$min(e)=\c{BLOCK}$}\\[-2mm]\small Multi-statement code blocks, as
+they can be large we do not try to normalize their contents. They execute in a
+median \packageMinimizedmedianoperationsjRnd steps, usually, in function
 environments.
-
-\medskip\EE{$min(e)=\c{model.frame}$} The \c{model.frame} function
-returns a dataframe resulting from fitting the model described in a given
-formula. This form subsumes \c{F(F(X))}, \c{FUN}, and assignments. It is the
-single most popular function invoked from \eval; it occurs in 7\% of the sites.
-Each call does quite a lot of work with a 2K instructions median.
-
-\paragraph{Discussion.}
-It is interesting how many different forms any given site sees. The more forms,
-the harder it is to characterize behavior at that site. Luckily,
-\packageNbOneMinimizedPercent of sites only see a single form. In other words,
-a compiler could correctly predict the \emph{shape} of an expression by looking
-just at the first call in \packageNbOneMinimizedPercent of the sites. Few sites
-are highly polymorphic (8 or more different forms). These include the pipe
-operator of the \c{magrittr} package, used to compose functions, \eg instead of
-\c{f(g(h(x),y),z)} one can write \c{h(x) \%>\% g(y) \%>\% f(z)}.
-
-The number of different forms evidences the variety of uses of \eval. In
-comparison, JavaScript \eval usage was more straightforward and more
-predictable as reported by \citet{oopsla12b}, with 98.7\% of the sites with
-only one form. On the other hand, there are \packageNbSimpleMinimizedOne sites
-that only receive one of the simple forms (\ie\xspace \c{X}, \c{V}, \c{\$} or
-\c{<-}). Additional \packageNbSimpleMinimizedMore sites take multiple forms,
-but all of them are these simple ones. Among these sites, there are many cases
-where \eval could be replaced by simpler constructs such as \c{get} to lookup a
-name in an environment, \c{assign} to set a value to a name in an environment,
-or \c{do.call} to call a function reflectively.
+\\
+\medskip\EE{$min(e)=\c{model.frame}$}\\[-2mm]\small The \c{model.frame} function
+returns a dataframe resulting from fitting the model described in a formula.
+This shape subsumes \c{F(F(X))}, \c{FUN}, and assignments. It is the single most
+popular function invoked from \eval. Each call does a median of 2K instructions.
+\end{tabular}
+
+\paragraph{Discussion.} The number of different shapes that any given
+\eval site sees is an indication of the versatility of that site.
+\packageNbOneMinimizedPercent of sites only see a single expression shape. It
+would be encouraging if one could, given a small training set, predict the shape
+of \evals to come. In our corpus, only a few sites are highly polymorphic with
+more than eight shapes. An example of those is the pipe operator of the
+\c{magrittr} package which is used to compose functions, \eg instead of
+\c{f(g(h(x),y),z)} one can write \c{h(x) \%>\% g(y) \%>\% f(z)}. As there many
+different patterns of use, there are also many different shapes. In comparison
+with R, JavaScript's \eval usage was more straightforward and more predictable
+as reported by \citet{oopsla12b}, with 98\% of sites with only one shape. On the
+other hand, there are \packageNbSimpleMinimizedOne sites that only receive one
+of the simple shape (\ie\xspace \c{X}, \c{V}, \c{\$} or \c{<-}), and
+\packageNbSimpleMinimizedMore sites get multiple simple shapes. Among these
+sites, there are many cases where \eval could be replaced by simpler constructs
+such as \c{get} to lookup a name in an environment, \c{assign} to set a value to
+a name in an environment, or \c{do.call} to call a function reflectively.
 

### 2021-08-10 paper/main.tex
tired

@@
 
-Contrary to the Javascript \eval, which does not specify the environment in
-which to evaluate, the second argument of \eval in R, \c{envir}, is dedicated to
-that. This argument determines what is visible to the computation started by
-\eval and the potential reach of its side-effects. Environments used by \eval
-can be classified into the following four kinds:
-
-\begin{compactitem}[---]
+Contrary to JavaScript, in R the environment of \eval can be specifies by its
+\c{env} argument. This gives users control over what is visible to the
+computation started by \eval and the potential reach of its side-effects. We
+distinguish the following kinds of environments:
+
+\begin{itemize}[---]
 \item \emph{Function:} environments for the local variables of some function
@@
 \item \emph{Synthetic:} environments built from data structures such as lists,
-  dataframes, or constructed explicitly with \c{new.env}, \c{list2env}, or
-  \c{as.environment}. It also includes the empty environment.
+  data.frames, or constructed explicitly with \c{new.env}, \c{list2env}, or
+  \c{as.environment} or the empty environment.
 \item \emph{Global:} environments in which scripts or interactive commands are
-  evaluated. \item \emph{Package:}  environments of a loaded library.
-\end{compactitem}
-
-As shown in Table~\ref{tab:highlevelenvironments}, most calls evaluate in the
-scope of a function with global as a distant second. This means that most
-variable reads and most side-effects operate on local variables. But for which
-function? From the point of view of the caller of \eval,
-Table~\ref{tab:funoffset} identifies that function by its offset on the call
-stack. Thus 0 is the direct caller, 1 is its parent, and so on. The data
-suggests that in 81\% of cases, \eval accesses the caller -- this means the
-variables of the function where \eval textually occurs are read and written to.
-It is noteworthy that 1.5\% of sites evaluate three frames or above. This means
-that, in general, modular reasoning is impossible in R. To understand what any
-given code snippet may do requires fully understanding all the functions that
-may be called, transitively, from that snippet. The actions of \eval happen at a
-distance. Finally, note that any site may see several kinds, but in
-\packageNbOneCategoryEnvirSitePercent of the cases they have a single kind. It
-means that a compiler could predict the \emph{kind} of environment after the
-first invocation.
+  evaluated.
+\item \emph{Package:} environments of libraries.
+\end{itemize}
+
+Table~\ref{tab:highlevelenvironments} shows that most calls evaluate in the
+scope of a function, global is a distant second. This means that most reads and
+writes act on local variables. But of which function? Table~\ref{tab:funoffset}
+gives the offset of that function on the call stack: 0 is function that called
+\eval, 1 is that function's caller, and so on. In 81\% of cases, \eval is
+resolved the caller -- the variables of the function where the call to \eval
+textually occurs are read and written. About 1.5\% of sites evaluate their
+argument three frames up the call stack, or above. Modular reasoning is thus
+impossible in R. Since the actions of \eval happen at a distance, understanding
+any given code snippet requires knowing which functions may be called
+transitively from that snippet. In \packageNbOneCategoryEnvirSitePercent of
+sites, only kind of environment is observed.
 
@@
 
-Global \evals are likely split between intentional and accidental one. Direct
-references to the top-level, using \c{globalenv()} or \c{.GlobalEnv}, are rare;
-they occur in only \packageNbExplicitGlobalSites sites. Thus we suspect most
-uses of global are accidental. They arise from properties of our corpus, most
-scripts are top-level code snippets. Thus, global is often the caller of \eval
-or close to it. The reason we make this point is that values stored in the
-global environment are visible to all functions and are not reclaimed by the
-garbage collector. So accidental uses may pollute that namespace.
-
 Each synthetic environment has parent that is specified when calling
@@
 its child (side-effects stay in the child). Table~\ref{tab:newenvs} shows parent
-kinds for synthetic environments. Most of them are functions, then come global
-and package.
+kinds for synthetic environments. Most of them are functions.
+
+Global \evals split between intentional and accidental one. Direct references to
+the top-level, using \c{globalenv()} or \c{.GlobalEnv}, are rare; they occur in
+only \packageNbExplicitGlobalSites sites. Thus we suspect most uses of global
+are accidental. They arise from the fact that our corpus, consists of many code
+snippet that are run at top-level, global is thus the caller of \eval. This
+noteworthy because writes to global variables are visible to all functions and
+are not reclaimed by the garbage collector. So accidental uses may pollute that
+namespace. Table~\ref{tab:polyenvir} gives the number of kinds seen at a given
+call site.
 
 \paragraph{Discussion.}
-The data presented in this section is what one could expect. For {\it
-  functions}, they represent the main user case for \eval. Thus \eval is really
-extending the behavior of that function by executing a dynamically selected
-piece of code in the function's scope. Typically, variables are read (mostly),
-written (less), but there are also some cases where new variables are injected
-or existing variables are deleted. Usually, this happens in the current
-function, but also in frames arbitrarily far up the call stack. One data point
-we lack is how the target environment was obtained. The expected case is that
-the body of a promise was modified and the resulting expression was evaluated in
-the same environment the promise originated from. Likely less frequent are cases
-where \eval is provided the results of programmatically selecting some call
-stack. For {\it synthetics}, their relatively high frequency correlates to use
-cases where one wants to evaluate an expression in either a restricted
-environment or to use a data structure as an environment. For {\it global}, its
-relatively high frequency is likely an artifact of how the code of our corpus is
-is run. For {\it packages}, only \packageNbPackageNamespaceEnvSites sites have
-this kind. This is probably for the best as mutating the bindings of
-a loaded package is improper and R tries to make it difficult.
+The data presented here is not surprising. The main use of \eval is to provide a
+customizable extension mechanism for the behavior of functions. They are a way
+to parameterize the function with any behavior that can be expressed in R. The
+behavior of \eval within its enclosing function is to read and write variables,
+mostly read, and very rarely delete variables. There are also some cases where
+new variables are injected in a function. Usually, this in the direct caller,
+but can sometimes take effect several frames up the call stack. There is
+something brittle about code relying on the position of a caller, many
+refactorings may break code that does that.
+
+One bit of information that we lack is how the environment was obtained. Our
+expectation is that the expression was extracted from a promise using
+\c{substitute}, perhaps modified, and then evaluated with the environment coming
+from the same promise. We believe the case where \eval is provided the results
+of programmatically selecting some call stack to be less frequent.
+
+Synthetic environments are relatively frequent. Common uses-case include
+evaluation of an expression in a sandbox, or using a data structure as
+environment, for example one could evaluate a method of an object and use and
+environment to hold the object's fields.
+
+We have already explained the popularity of global environments, as for package
+environments, they are used in only \packageNbPackageNamespaceEnvSites sites.
+This is probably for the best as mutating the bindings of a loaded package is
+frowned upon and R tries to discourage it.
 
@@
 Where does the expression passed to \eval come from? There are various means of
-creating that expression, which are associated with particular use cases.
-During tracing, at each function call, we check if the return value is an expression or a symbol, or if the function is a function of interest.
-If it is, we record it and also look at whether the arguments were recorded before if it is not an expression-building function.
-It makes it possible to build a direct acyclic graph of the functions that build expressions or combine expressions during tracing to create a direct acyclic graph of the functions creating the expression passed to \eval.
-The following example yields the provenance graph with five nodes on Figure~\ref{fig:provgraph} for expression \c{e} passed to \eval.
+creating an expression, often these means correlate with a particular use case.
+Our analysis records the values returned by some functions of interest as well
+as their arguments. We build a direct acyclic graph that track the provenance of
+the values that are given to \eval. The following example yields the provenance
+graph with five nodes on Figure~\ref{fig:provgraph} for expression \c{e}.
 
 \begin{lstlisting}
-e <- parse(text = "a ; b")
-e[[2]] <- quote(c)
-eval(e)
-\end{lstlisting}
-
-\begin{figure}[h]
-\centering
-\includegraphics[width=8cm]{provenance-graph}
-\caption{The provenance DAG for the code example above. Assignments and complex assignments are intermediate nodes, whereas expression builders are leaves of the graph.} \label{fig:provgraph}
+ > e <- parse(text = "a ; b")
+ > e[[2]] <- quote(c)
+ > eval(e)
+\end{lstlisting}
+
+\begin{figure}[h]\centering
+\includegraphics[width=5cm]{provenance-graph}
+\caption{Provenance} \label{fig:provgraph}
 \end{figure}
 
-Expression \c{e} has two origins, \c{parse} and \c{quote}, which corresponds to the leaves of the graph.
-We also extract from the graph a representative origin, by traversing the graph from the root, following the left-hand side member of the assignments until a leaf.
-It represents the provenance that is further modified to yield the expression passed to \eval.
-In our example, the representative provenance is \c{parse}.
+The provenance of \c{e} includes both \c{parse} and \c{quote}. We pick an origin
+by traversing the graph from the root, following the left-hand side member of
+assignments until a leaf. It represents the provenance that is further modified
+to yield the expression passed to \eval. In our example, the origin is \c{parse}.
 
@@
 \begin{wraptable}{r}{5cm}\small\centering
-	\begin{tabular}{lcc}
-		\toprule
-		\bf Provenance & \bf \#sites & \bf \%sites \\
-		\midrule
+\begin{tabular}{lcc}\toprule
+\bf Provenance & \bf \#sites & \bf \%sites \\\midrule
 		\packageProvenanceNamea & \packageNbProvenanceSitesa & \packagePercentProvenanceSitesa \\
@@
 		\packageProvenanceNamec & \packageNbProvenanceSitesc & \packagePercentProvenanceSitesc \\
-		%\packageProvenanceNamed & \packageNbProvenanceSitesd & %\packagePercentProvenanceSitesd \\
 		\packageProvenanceNamee & \packageNbProvenanceSitese & \packagePercentProvenanceSitese \\
@@
 		\packageProvenanceNamei & \packageNbProvenanceSitesi & \packagePercentProvenanceSitesi \\
-		%\packageProvenanceNamej & \packageNbProvenanceSitesj & %\packagePercentProvenanceSitesj \\%[[
-		%\packageProvenanceNamek & \packageNbProvenanceSitesk & %\packagePercentProvenanceSitesk \\%$
 		\packageProvenanceNamel & \packageNbProvenanceSitesl & \packagePercentProvenanceSitesl \\
-		%\packageProvenanceNamem & \packageNbProvenanceSitesm & %\packagePercentProvenanceSitesm \\ %as.vector
-		\packageProvenanceNamen & \packageNbProvenanceSitesn & \packagePercentProvenanceSitesn \\
-		\bottomrule
+		\packageProvenanceNamen & \packageNbProvenanceSitesn & \packagePercentProvenanceSitesn \\\bottomrule
 \end{tabular}
-	\caption{First provenances in number of sites.} \label{tab:detailed_provenances}
-\end{wraptable}
-
-Table~\ref{tab:detailed_provenances} shows the first provenances in number of sites.
-In \packagePercentProvenanceSitesd\% of the sites, the expression passed to \eval is not a language expression or a symbol but a floating point, or a string for instance.
-In that case, we did not track the provenance.
-\c{.External} invokes a C function.
-For instance, in the following code, \c{.External(C\_deriv, expr, namevec, function.arg, tag, hessian)} calls a C function called \c{C\_deriv} to derive an expression.
- Due to technical reasons, there is some imprecision in the
-results.
-In particular, we misclassify some sites but these cases happen rarely.
-For instance,  \packagePercentProvenanceSitesj\% of the sites have provenance \c{[[}, a subsetting operator, as we lose the provenance of the subsetted expression, often when that expression is given attributes.
-Manual inspection of numerous examples suggests that errors are rare.
-
- We further classify them into four high-level categories:
-
-\begin{compactitem}[---]
-\item {\it Reflection:} This group corresponds to uses of \c{match.call} to
-	reflectively capture the expression that invoked the current function. Function arguments are passed as promises, and \c{substitute} is used to retrieve the source
-	expression associated with the promise. \c{formals} access the argument list of a function.
-\item {\it String:} Expressions created from strings by invoking \c{parse},
-	\c{str2expression}, or \c{str2lang}.
-\item {\it Constructed:} Expressions can be constructed by invoking the
-  \c{quote}, \c{expression} functions, or by building a call with \c{call} or \c{as.call}.
-\item {\it Environment: }   A symbol is created with \c{as.name} or \c{as.symbol}, typically to find the value of a symbol in a non-local environment.
-\item {\it External: } a call to a C function. % Put an example
-\end{compactitem}
-
-%\NOTE{I only see 70 sites with formulas (using the class of the expression) so it does not appear here.}
-
-%\NOTE{I can also put the absolute number of provenances, \ie if an expression has multiple provenances, we do not keep only the representative but also the other provenances.
-Also, if one site sometimes gets match.call and sometimes substitute, it is not going to be counted in either of them currently.}
-
-Table~\ref{tab:provenance} summarizes the expression provenance in our corpus.
-Note that some sites can be invoked several times with different provenances.
-In that case, we have counted them several times in the table.
-
-\begin{wraptable}[8]{r}{5cm}\small\centering
+\caption{Origin} \label{tab:detailed_provenances}
+\medskip
 \begin{tabular}{rrr} \toprule
@@
 
-Strings could correlate with dynamic code loading. This is what the base
-functions \c{source} and \c{sys.source} do. We observed few calls
-(\packageNbParseFromFileSites in total) that consume the result of calling
-\c{parse} on a file. Most of the calls build strings programmatically. We also
-identified one function \c{invokeRestartInteractively} that prompts the user for
-input, parses it, and passes it to \eval.
+Table~\ref{tab:detailed_provenances} shows frequencies of provenance. For
+technical reasons, we misclassify a small number of sites. Manual inspection of
+numerous examples suggests that errors are rare. We further classify them into
+high-level categories:
+
+\begin{itemize}[---]
+\item {\it Reflection:} use of \c{match.call} to reflectively capture the
+  expression that invoked this function, arguments are promises, and
+  \c{substitute} is used to retrieve their source.
+\item {\it String:} create from strings by invoking \c{parse},
+  \c{str2expression}, or \c{str2lang}.
+\item {\it Constructed:} invoked \c{quote} or \c{expression}, or by built with
+  \c{call} or \c{as.call}.
+\item {\it Environment: } created with \c{as.name} or \c{as.symbol}, typically
+  to read a non-local environment.
+\item {\it External: }  call to a C function. % Put an example
+\end{itemize}
+
+Table~\ref{tab:provenance} summarizes provenance. Some sites are invoked with
+different provenances, they are counted multiple times. Strings correlate with
+dynamic code loading as done by \c{source} and \c{sys.source}. Few calls
+(\packageNbParseFromFileSites in total) consume the result of calling \c{parse}
+on a file. Most build strings programmatically. We identified one function
+\c{invokeRestartInteractively} that prompts the user for input and \evals it.
 
 \paragraph{Discussion}
-The origin data suggests that reflection is the first use case for \eval, \ie using \c{match.call}, \c{substitute}, or \c{match.call} to access and transform the arguments of a function when calling the function.
-Constructing expressions from strings is also
-a common use case. The constructed category shows that most
-\evals comes from code that was processed by the compiler, and may be slightly
-modified by the programmer before invoking \eval. Both constructed and
-reflection categories roughly correspond to meta-programming. Some of these use
-cases could likely be replaced by macros if the R designers could be convinced
-to overcome their distaste for those. Another use of \eval is to directly build symbols to evaluate them in another environment than the default one.
+The main use case for \eval involves using \c{match.call}, \c{substitute}, or
+\c{match.call} to access and transform arguments of a function. Constructing
+expressions from strings is also a common use case. The constructed category
+shows that most \evals comes from existing code that was modified by the
+programmer before invoking \eval. Both constructed and reflection categories
+roughly correspond to meta-programming. Some use cases could likely be replaced
+by macros if the R designers could be convinced to overcome their distaste for
+those. Another use of \eval is to directly build symbols to evaluate them in
+another environment.
 
@@
 
-\Eval may perform side-effects. From a compiler's perspective, we care about
-effects that can be observed --- \ie variable definitions, updates, and removals
-that are visible after a call finishes. Knowing in which environments these
-side-effects happen, can help us determine how much of the compiler knowledge
-about the program will be invalidated. Our analysis records information about
+\Eval may perform side-effects. We care about effects that are visible after a
+call finishes, \ie variable definitions, updates, and removals. Knowing in which
+environments these side-effects happen, can help determine the impact of \eval
+on our ability to reason about the code. Our analysis records information about
 every environment. From the recorded data, we discard side-effects coming from
-unit testing frameworks.\footnote{In the corpus, we have \c{RUnit, testthat,
-  tinytest} and \c{unitizer} testing frameworks.} They run their tests via
-\eval, and thus everything becomes a side-effect.
-
-From our \packageNbrunsRnd programs, we capture \SEAllRnd side-effects from
-\SEAllCallsRnd \eval calls (\SEUserCallsToAllRatio) in \SEAllSites sites
-(\SEUserSitesToAllRatio). Again, the challenge is to remove the accidental
-side-effects caused by the R virtual machine implementation which are not
-related to the user code. For example, the \c{.Random.seed} variable, which
-contains the state of the random number generator, is saved and restored from
-and to the global environment every time  a statistical routine is called.
-%
-Removing those leaves us with \SEUserRnd side effects from \SEUserCallsRnd
-\eval calls in \SEUserSites sites. These sites are part of \SEUserFunctions R
-functions (\SEUserFunctionsToAllRatio of all functions doing \eval) in
-\SEUserPackages packages. From these functions, \SEFunsNighty are responsible
-for 90\% of side effects. Half of the side-effects come just from three
+unit testing frameworks as they use \eval to run all their tests and thus
+dominate the side-effect data.\footnote{The corpus has \c{RUnit, testthat,
+  tinytest} and \c{unitizer} testing frameworks.}
+
+We record \SEAllRnd side-effects from \SEAllCallsRnd calls and \SEAllSites
+sites. The challenge is to remove accidental side-effects caused by the R
+virtual machine implementation and not user code. For example, the
+\c{.Random.seed} variable is saved and restored from and to the global
+environment every time a statistical routine is called. Filtering accidental
+side-effects leaves us with \SEUserRnd writes from \SEUserCallsRnd calls (in
+\SEUserSites sites and \SEUserFunctions functions). \SEFunsNighty functions are
+responsible for 90\% of side effects. Half of those come from just three
 functions: \c{plyr::allocate\_column} (allocates space for a new data frame
@@
 \c{foreach::doSEQ} (executes an expression on each element in a collection,
-possibly in parallel). Most of the \eval sites (\SESitesInEnvirRatio) do side
-effects in the environment specified by the \c{envir} parameter (\cf
-Section~\ref{sec:eval-in-r}), \SESitesNotInEnvirRatio modifies other
-environments, and finally, \SESitesBothEnvirRatio does both.
+possibly in parallel). Most of the sites (\SESitesInEnvirRatio) update
+the environment specified by the \c{env} parameter.
 

### 2021-08-11 analysis/side-effects.Rmd
Remove \hline from SE tables

@@
     comment=FALSE, 
+    hline.after=NULL,
     include.colnames=FALSE,

### 2021-08-11 analysis/corpus.Rmd
Update Kaggle
Remove axis label

@@
 
-eval_call_sites_hist(PKG_C, CRAN_C, "CRAN packages")
+eval_call_sites_hist(PKG_C, CRAN_C, "")
 ggsave(path(PLOT_DIR, "pkgs-eval-callsites-hist.pdf"))
-eval_call_sites_hist(KAGGLE, KAGGLE_ALL, "KAGGLE programs")
-ggsave(path(PLOT_DIR, "kaggle-eval-callsites-hist.pdf"))
+
+# eval_call_sites_hist(KAGGLE, KAGGLE_ALL, "KAGGLE programs")
+# ggsave(path(PLOT_DIR, "kaggle-eval-callsites-hist.pdf"))
 ```
@@
     ) +
-#    geom_smooth(method = "lm", se = FALSE, show.legend = FALSE, linetype="dashed", color="red", size=0.2) +
     labs(x="Eval call sites (log)", y="Traced eval call sites (log)", size="Package R code size") +
@@
 ```
+
+## Kaggle extra
+
+The following should be somewhere else
+
+```{r}
+kaggle_log <- read_csv(KAGGLE_TRACE_LOG_FILE)
+```
+
+```{r}
+overview_table(
+  r("Kaggle failed scripts", filter(kaggle_log, Exitval != 0))
+)
+```

### 2021-08-11 paper/main.tex
Fix table 8 and table 9
Fix table
Fix figures so they do not end up at the end
Update 5.4

@@
 \centering
-  \includegraphics[width=10cm]{pkgs-eval-callsites-hist.pdf}
+  \includegraphics[width=9cm]{pkgs-eval-callsites-hist.pdf}
   \caption{ \eval call sites}\label{fig:pkgs-eval-callsites-hist}
@@
 \begin{wraptable}{r}{7cm}\small\centering
-  \begin{tabular}{lrrrr}\toprule
+  \begin{tabular}{lrrrr}
+    \toprule
     \bf Environments & \bf \#sites & \bf \%sites & \bf \#funs. & \bf \%funs. \\%
+    \midrule
     \expandableinput tag/table-se-target-envs.tex
+    \bottomrule
   \end{tabular}
@@
   \centering
-  \begin{tabular}{lrrrrrr}\toprule
+  \begin{tabular}{lrrrrrr}
+    \toprule
     \bf Side effect & \bf \#events & \bf \%events & \bf \#calls & \bf \%calls & \bf \#sites & \bf \%sites \\%
+    \midrule
     \expandableinput tag/table-se-types.tex
+    \bottomrule
   \end{tabular}
@@
 
-Out of the minimized expressions, it is \c{BLOCK}, \c{<-} and
-\c{F(F(X)} that contribute to the vast majority of side effects. Concretely,
-92\% of \c{BLOCK}, 87\% of \c{<-} and 15\% of \c{F(F(X))} expressions contain
-side effects. In the case of \c{BLOCK}, most of the side effects happen in the
-environment passed to \eval (which by default is the parent environment, \ie
-the local or function in our classification). For \c{<-} it is 15\%, however,
-that is only because of the \c{plyr::allocate\_column} which does the side
-effect in the environment containing the data frame that is the target for the
-newly allocated column. Without this call, \c{<-} is similarly predictable with
-96\% of side effects in the passed environment. For \c{F(F(X))} minimization
-for, it is less clear as 48\% of side effects happen in a different environment
-to the one that was passed to \eval.
+Out of the minimized expressions, it is \c{BLOCK}, \c{<-} and \c{F(F(X)} that
+contribute to the vast majority of side effects. Concretely, 92\% of \c{BLOCK},
+87\% of \c{<-} and 15\% of \c{F(F(X))} expressions do side effects. In the case
+of \c{BLOCK}, most of them happen in the environment passed to the \eval call
+(default is the parent environment). For \c{<-} it is 15\%. However, this form
+is dominated by the \c{plyr::allocate\_column} function that side-effects in
+the environment containing the data frame to which it allocates new column.
+Without this call, \c{<-} is similarly predictable with 96\% of side effects
+happening in the passed environment. For \c{F(F(X))}, it is less clear as 58\%
+of side effects happen in a different environment to the one that was passed to
+\eval.
 

### 2021-08-11 paper/main.tex
Better looking provenance graph

--- 2021-08-10 paper/main.tex
@@
 
+%\begin{figure}[h]
+%\centering
+%\includegraphics[width=8cm]{provenance-graph}
+%\caption{The provenance DAG for the code example above. Assignments and complex assignments are intermediate nodes, whereas expression builders are leaves of the graph.} \label{fig:provgraph}
+%\end{figure}
+
 \begin{figure}[h]
 	\centering
-\includegraphics[width=8cm]{provenance-graph}
+	\begin{tikzpicture}[
+		every node/.style={draw, rounded corners, fill=LightGray, text width = 2cm, align = center}
+		]
+		\node (E) at (2,0) {\c{eval(e)}};
+		\node[text width = 3cm] (CA) at (2, 1.5) {\c{[[<-(e, 2, quote(c))}};
+		\node (Q) at (0, 3) {\c{quote(c)}};
+		\node[text width = 4cm] (A) at (4, 3) {\c{e <- parse(text = "a ; b"}};
+		\node[text width = 3cm] (P) at (4, 4.5) {\c{parse(text = "a ; b"}};
+
+		\draw[->, >=latex] (E) -- (CA);
+		\draw[->, >=latex] (CA) -- (Q);
+		\draw[->, >=latex] (CA) -- (A);
+		\draw[->, >=latex] (A) -- (P);
+	\end{tikzpicture}
 	\caption{The provenance DAG for the code example above. Assignments and complex assignments are intermediate nodes, whereas expression builders are leaves of the graph.} \label{fig:provgraph}

### 2021-08-11 paper/main.tex

--- 2021-08-11 paper/main.tex
@@
 
-\begin{figure}[h]\centering
-\includegraphics[width=5cm]{provenance-graph}
+%\begin{figure}[h]
+%\centering
+%\includegraphics[width=8cm]{provenance-graph}
+%\caption{The provenance DAG for the code example above. Assignments and complex assignments are intermediate nodes, whereas expression builders are leaves of the graph.} \label{fig:provgraph}
+%\end{figure}
+
+\begin{figure}[h]
+	\centering
+	\begin{tikzpicture}[
+		every node/.style={draw, rounded corners, fill=LightGray, text width = 2cm, align = center}
+		]
+		\node (E) at (2,0) {\c{eval(e)}};
+		\node[text width = 3cm] (CA) at (2, 1.5) {\c{[[<-(e, 2, quote(c))}};
+		\node (Q) at (0, 3) {\c{quote(c)}};
+		\node[text width = 4cm] (A) at (4, 3) {\c{e <- parse(text = "a ; b"}};
+		\node[text width = 3cm] (P) at (4, 4.5) {\c{parse(text = "a ; b"}};
+
+		\draw[->, >=latex] (E) -- (CA);
+		\draw[->, >=latex] (CA) -- (Q);
+		\draw[->, >=latex] (CA) -- (A);
+		\draw[->, >=latex] (A) -- (P);
+	\end{tikzpicture}
 \caption{Provenance} \label{fig:provgraph}

### 2021-08-11 paper/main.tex
Better show LHS in Figure 10
match.call -> formals

@@
 		\node[text width = 3cm] (CA) at (2, 1.5) {\c{[[<-(e, 2, quote(c))}};
-		\node (Q) at (0, 3) {\c{quote(c)}};
-		\node[text width = 4cm] (A) at (4, 3) {\c{e <- parse(text = "a ; b"}};
-		\node[text width = 3cm] (P) at (4, 4.5) {\c{parse(text = "a ; b"}};
+		\node (Q) at (4, 3) {\c{quote(c)}};
+		\node[text width = 4cm] (A) at (0, 3) {\c{e <- parse(text = "a ; b"}};
+		\node[text width = 3cm] (P) at (0, 4.5) {\c{parse(text = "a ; b"}};
 
@@
 The main use case for \eval involves using \c{match.call}, \c{substitute}, or
-\c{match.call} to access and transform arguments of a function. Constructing
+\c{formals} to access and transform arguments of a function. Constructing
 expressions from strings is also a common use case. The constructed category

### 2021-08-11 paper/main.tex
Fix graph

@@
 \begin{lstlisting}
- > e <- parse(text = "a ; b")
+ > e <- parse(text = "a;b")
  > e[[2]] <- quote(c)
@@
 		\node (E) at (2,0) {\c{eval(e)}};
-		\node[text width = 3cm] (CA) at (2, 1.5) {\c{[[<-(e, 2, quote(c))}};
+		\node[text width = 3cm] (CA) at (2, 1.5) {\c{e[[2]] <-quote(c)}};
 		\node (Q) at (4, 3) {\c{quote(c)}};
-		\node[text width = 4cm] (A) at (0, 3) {\c{e <- parse(text = "a ; b"}};
-		\node[text width = 3cm] (P) at (0, 4.5) {\c{parse(text = "a ; b"}};
-
-		\draw[->, >=latex] (E) -- (CA);
-		\draw[->, >=latex] (CA) -- (Q);
-		\draw[->, >=latex] (CA) -- (A);
-		\draw[->, >=latex] (A) -- (P);
+		\node[text width = 4cm] (A) at (0, 3) {\c{e <- parse(text = "a;b")}};
+		\node[text width = 3cm] (P) at (0, 4.5) {\c{parse(text = "a;b")}};
+
+		\draw[<-, >=latex] (E) -- (CA);
+		\draw[<-, >=latex] (CA) -- (Q);
+		\draw[<-, >=latex] (CA) -- (A);
+		\draw[<-, >=latex] (A) -- (P);
 	\end{tikzpicture}

### 2021-08-11 paper/main.tex
Update the min groups with examples from shapes redux

@@
   \bf $min(e)$& \bf \#sites & \bf \%sites & \bf \#packages & \bf \#operations & \bf \%envir & \bf example\\\midrule
-\c X&\packageMinimizedcallsitesa &\packageMinimizedpropsitesa &\packageMinimizedpackagea &\packageMinimizedmedianoperationsaRnd &\packageMinimizedpercentparentframesa & \c{y+1}\\
-\c{F(F(X))} & \packageMinimizedcallsitesb  & \packageMinimizedpropsitesb & \packageMinimizedpackageb  & \packageMinimizedmedianoperationsbRnd & \packageMinimizedpercentparentframesb & \c{gbov( mean(x), a-1)}\\
-\c{V}&\packageMinimizedcallsitesc &\packageMinimizedpropsitesc &\packageMinimizedpackagec &\packageMinimizedmedianoperationscRnd &\packageMinimizedpercentparentframesc& \c{c(42,21,0)}\\
-\c{F(X)}& \packageMinimizedcallsitesd & \packageMinimizedpropsitesd & \packageMinimizedpackaged & \packageMinimizedmedianoperationsdRnd & \packageMinimizedpercentparentframesd & \c{seq\_len(iters)} \\
-\c{\$} & \packageMinimizedcallsitese & \packageMinimizedpropsitese & \packageMinimizedpackagee & \packageMinimizedmedianoperationseRnd & \packageMinimizedpercentparentframese & \c{DF\$B}\\
-\c{model.frame}& \packageMinimizedcallsitesf & \packageMinimizedpropsitesf & \packageMinimizedpackagef & \packageMinimizedmedianoperationsfRnd & \packageMinimizedpercentparentframesf &  \c{model.frame(formula = Z $\sim$ U)}   \\
-\c{F()}& \packageMinimizedcallsitesg & \packageMinimizedpropsitesg & \packageMinimizedpackageg & \packageMinimizedmedianoperationsgRnd & \packageMinimizedpercentparentframesg & \c{rgamma(3, 2, n = 10L)} \\
-\c{FUN} & \packageMinimizedcallsitesh & \packageMinimizedpropsitesh & \packageMinimizedpackageh & \packageMinimizedmedianoperationshRnd & \packageMinimizedpercentparentframesh & \c{function(x, y) x + 3 * y} \\
-\c{<-} & \packageMinimizedcallsitesi  & \packageMinimizedpropsitesi & \packageMinimizedpackagei & \packageMinimizedmedianoperationsiRnd & \packageMinimizedpercentparentframesi & \c{x[1, 2:3, 2:3] <- value}\\
-\c{BLOCK} & \packageMinimizedcallsitesj & \packageMinimizedpropsitesj & \packageMinimizedpackagej & \packageMinimizedmedianoperationsjRnd & \packageMinimizedpercentparentframesj & \c{\{ write.csv(iris, tf) ; file.size(tf) \}} \\\bottomrule
+\code X&\packageMinimizedcallsitesa &\packageMinimizedpropsitesa &\packageMinimizedpackagea &\packageMinimizedmedianoperationsaRnd &\packageMinimizedpercentparentframesa & \code{y+1}\\
+\code{F(F(X))} & \packageMinimizedcallsitesb  & \packageMinimizedpropsitesb & \packageMinimizedpackageb  & \packageMinimizedmedianoperationsbRnd & \packageMinimizedpercentparentframesb & \code{gbov( mean(x), a-1)}\\
+\code{V}&\packageMinimizedcallsitesc &\packageMinimizedpropsitesc &\packageMinimizedpackagec &\packageMinimizedmedianoperationscRnd &\packageMinimizedpercentparentframesc& \code{c(42,21,0)}\\
+\code{F(X)}& \packageMinimizedcallsitesd & \packageMinimizedpropsitesd & \packageMinimizedpackaged & \packageMinimizedmedianoperationsdRnd & \packageMinimizedpercentparentframesd & \code{seq\_len(iters)} \\
+\code{$} & \packageMinimizedcallsitese & \packageMinimizedpropsitese & \packageMinimizedpackagee & \packageMinimizedmedianoperationseRnd & \packageMinimizedpercentparentframese & \code{DF\$B}\\
+\code{model.frame}& \packageMinimizedcallsitesf & \packageMinimizedpropsitesf & \packageMinimizedpackagef & \packageMinimizedmedianoperationsfRnd & \packageMinimizedpercentparentframesf &  \code{model.frame(formula = Z ~  U)}   \\
+\code{F()}& \packageMinimizedcallsitesg & \packageMinimizedpropsitesg & \packageMinimizedpackageg & \packageMinimizedmedianoperationsgRnd & \packageMinimizedpercentparentframesg & \code{rgamma(3, 2, n = 10L)} \\
+\code{FUN} & \packageMinimizedcallsitesh & \packageMinimizedpropsitesh & \packageMinimizedpackageh & \packageMinimizedmedianoperationshRnd & \packageMinimizedpercentparentframesh & \code{function(x, y) x + 3 * y} \\
+\code{<-} & \packageMinimizedcallsitesi  & \packageMinimizedpropsitesi & \packageMinimizedpackagei & \packageMinimizedmedianoperationsiRnd & \packageMinimizedpercentparentframesi & \code{x[1, 2:3, 2:3] <- value}\\
+\code{BLOCK} & \packageMinimizedcallsitesj & \packageMinimizedpropsitesj & \packageMinimizedpackagej & \packageMinimizedmedianoperationsjRnd & \packageMinimizedpercentparentframesj & \code{{ write.csv(iris, tf) ; file.size(tf) }} \\\bottomrule
 \end{tabular}
@@
 folding of arithmetic and string expressions over base operators, for instance
-$min(\c{1+1})$ simplifies to $\c{V}$, on the ground that addition of values
-likely returns a value. Value simplification takes $min(\c{c(1,2,3+2)})$ to
-$\c{V}$ as a complex vector containing values is a value. Variable absorption
-has $min(\c{x+y})$ becomes $\c{X}$, this motivated by the fact that addition is
-not an ``interesting'' operation, and that \c X stands in for any number of
+$min(\c{1+1})$ simplifies to \code{V}, on the ground that addition of values
+likely returns a value. Value simplification takes $min$(\code{c(1,2,3+2)}) to
+\code{V} as a complex vector containing values is a value. Variable absorption
+has $min$(\code{x+y}) becomes \code{X}, this motivated by the fact that addition is
+not an ``interesting'' operation, and that \code{X} stands in for any number of
 variable lookups. Function absorption simplifies nested functions to keep an
-abstraction of the nesting $min(\c{g(f(x),h(z))})=\c{F(F(X))}$. There are other
+abstraction of the nesting $min$(\code{g(f(x),h(z))})$=$\code{F(F(X))}. There are other
 simplifications, the full list can be found in our artifact. It is worth
-mentioning that these simplifications are heuristics, in R \c{1+1} is not
-necessary \c 2. The addition operator can be redefined, but it typically isn't
+mentioning that these simplifications are heuristics, in R \code{1+1} is not
+necessary \code{2}. The addition operator can be redefined, but it typically isn't
 or at least not in a way that would invalidate arithmetic.
@@
 \begin{tabular}{@{}p{.97\linewidth}}
-  \medskip\EE{$min(e)=\c{V}$} \\[-2mm]\small Values occur in 16\% of sites, 74\% of
+  \vspace{1mm}
+  \EE{$min(e)=\code{V}$} \\[-2mm]\small Values occur in 16\% of sites, 74\% of
   those are constants such as integers, the remainder evaluate to a value. \Eval
@@
 \end{lstlisting}
-\\
-\medskip\EE{$min(e)=\c{X}$}\\[-2mm]\small Variable lookups are the most common shape,
+\vspace{2mm}
+\EE{$min(e)=\code{X}$}\\[-2mm]\small Variable lookups are the most common shape,
 \packageNbSymbolVarSitePercent are simple variable names, e.g. \c x. The shapes
@@
 expressions are evaluated in a function environment.
-\\
-\medskip\EE{$min(e)=\c{\$}$}\\[-2mm]\small This shape extends \c X to include lookup
+\vspace{2mm}
+\EE{$min(e)=\code{\$}$}\\[-2mm]\small This shape extends \c X to include lookup
 with the dollar operator, \eg~\c{x\$f}, and vector indexing, \eg~\c{x[42]} and
@@
 typically used in a function environment.
-\\
-\medskip\EE{$min(e)=$~\c{<-}}\\[-2mm]\small This includes both direct assignment {\tt
-  <-} and parent environment assignment {\tt <\,\!<-} as well as the \c{assign}
-function. It subsumes \c{\$}. Assignments
-represent the most obvious source of side effects in \eval.
-\\
-\medskip\EE{$min(e)=\c{F()}$}\\[-2mm]\small Function calls without variable lookups or
+
+\vspace{2mm}
+\EE{$min(e)=$~\code{<-}}\\[-2mm]\small This includes both direct
+assignment {\tt <-} and parent environment assignment {\tt <\,\!<-} as well as
+the \c{assign} function. It subsumes \c{\$}. Assignments represent the most
+obvious source of side effects in \eval. Most cases originate from trivial code
+generation where the assign expression is assembled using parse or substitute,
+often in a loop. For example (taken from \c{PopED::get\_rse}), the following
+code will populate the current scope with a number of variables:
+%
+\begin{lstlisting}
+for (i in seq_along(args)) eval(parse(text=paste(args[i], "<-", i)))
+\end{lstlisting}
+
+\vspace{2mm}
+\EE{$min(e)=\code{F()}$}\\[-2mm]\small Function calls without variable lookups or
 assignment. A variable is allowed the function position, looking up
@@
 is bound to a promise. This shape typically does not perform much work.
-\\
-\medskip\framebox{$min(e)=\c{F(X)}$}~\EE{$min(e)=\c{F(F(X))}$}\\[-2mm]\small Function
+\vspace{2mm}
+\framebox{$min(e)=\code{F(X)}$}~\EE{$min(e)=\code{F(F(X))}$}\\[-2mm]\small Function
 calls whose arguments may include variable references and nested calls. Together
@@
 \begin{tabular}{@{}p{.97\linewidth}}
-  \medskip\EE{$min(e)=\c{FUN}$}\\[-2mm]\small Function definitions without any
+  \medskip\EE{$min(e)=$ \code{FUN}}\\[-2mm]\small Function definitions without any
   computation, \eg~\c{function(x) x+1}. In addition to \c{FUN},
@@
   definitions nested in other expressions.
-\\
-\medskip\EE{$min(e)=\c{BLOCK}$}\\[-2mm]\small Multi-statement code blocks, as
+
+\vspace{2mm}
+\EE{$min(e)=\code{BLOCK}$}\\[-2mm]\small Multi-statement code blocks, as
 they can be large we do not try to normalize their contents. They execute in a
@@
 environments.
-\\
-\medskip\EE{$min(e)=\c{model.frame}$}\\[-2mm]\small The \c{model.frame} function
+
+\vspace{2mm}
+\EE{$min(e)=\code{model.frame}$}\\[-2mm]\small The \c{model.frame} function
 returns a dataframe resulting from fitting the model described in a formula.
@@
 \c{magrittr} package which is used to compose functions, \eg instead of
-\c{f(g(h(x),y),z)} one can write \c{h(x) \%>\% g(y) \%>\% f(z)}. As there many
+\code{f(g(h(x),y),z)} one can write \code{h(x) \%>\% g(y) \%>\% f(z)}. As there many
 different patterns of use, there are also many different shapes. In comparison
@@
 other hand, there are \packageNbSimpleMinimizedOne sites that only receive one
-of the simple shape (\ie\xspace \c{X}, \c{V}, \c{\$} or \c{<-}), and
+of the simple shape (\ie\xspace \code{X}, \code{V}, \code{\$} or \code{<-}), and
 \packageNbSimpleMinimizedMore sites get multiple simple shapes. Among these
 sites, there are many cases where \eval could be replaced by simpler constructs
-such as \c{get} to lookup a name in an environment, \c{assign} to set a value to
-a name in an environment, or \c{do.call} to call a function reflectively.
+such as \code{get} to lookup a name in an environment, \code{assign} to set a value to
+a name in an environment, or \code{do.call} to call a function reflectively.
 
@@
 \item \emph{Function:} environments for the local variables of some function
-  currently active on the call stack. Obtained by calling \c{parent.frame()} or
-  \c{sys.frame()}.
+  currently active on the call stack. Obtained by calling \code{parent.frame()} or
+  \code{sys.frame()}.
 \item \emph{Synthetic:} environments built from data structures such as lists,
-  data.frames, or constructed explicitly with \c{new.env}, \c{list2env}, or
-  \c{as.environment} or the empty environment.
+  data.frames, or constructed explicitly with \code{new.env}, \code{list2env}, or
+  \code{as.environment} or the empty environment.
 \item \emph{Global:} environments in which scripts or interactive commands are
@@
 Each synthetic environment has parent that is specified when calling
-\c{new.env}. When \c{envir} is a list or a data frame, \eval uses its third
-argument (\c{enclos}). The parent is used to search for variables not found in
+\code{new.env}. When \code{envir} is a list or a data frame, \eval uses its third
+argument (\code{enclos}). The parent is used to search for variables not found in
 its child (side-effects stay in the child). Table~\ref{tab:newenvs} shows parent
@@
 Global \evals split between intentional and accidental one. Direct references to
-the top-level, using \c{globalenv()} or \c{.GlobalEnv}, are rare; they occur in
+the top-level, using \code{globalenv()} or \code{.GlobalEnv}, are rare; they occur in
 only \packageNbExplicitGlobalSites sites. Thus we suspect most uses of global
@@
 expectation is that the expression was extracted from a promise using
-\c{substitute}, perhaps modified, and then evaluated with the environment coming
+\code{substitute}, perhaps modified, and then evaluated with the environment coming
 from the same promise. We believe the case where \eval is provided the results
@@
 the values that are given to \eval. The following example yields the provenance
-graph with five nodes on Figure~\ref{fig:provgraph} for expression \c{e}.
+graph with five nodes on Figure~\ref{fig:provgraph} for expression \code{e}.
 
@@
 	\begin{tikzpicture}[
-		every node/.style={draw, rounded corners, fill=LightGray, text width = 2cm, align = center}
+		every node/.style={draw, rounded corners, fill=LightGray, text width = 2cm, align = center, thick}
 		]
-		\node (E) at (2,0) {\c{eval(e)}};
-		\node[text width = 3cm] (CA) at (2, 1.5) {\c{e[[2]] <-quote(c)}};
-		\node (Q) at (4, 3) {\c{quote(c)}};
-		\node[text width = 4cm] (A) at (0, 3) {\c{e <- parse(text = "a;b")}};
-		\node[text width = 3cm] (P) at (0, 4.5) {\c{parse(text = "a;b")}};
+		\node (E) at (2,0) {\code{eval(e)}};
+		\node[text width = 3cm] (CA) at (2, 1.5) {\code{e[[2]] <-quote(c)}};
+		\node (Q) at (4, 3) {\code{quote(c)}};
+		\node[text width = 4cm] (A) at (0, 3) {\code{e <- parse(text = "a;b")}};
+		\node[text width = 3cm] (P) at (0, 4.5) {\code{parse(text = "a;b")}};
 
@@
 
-The provenance of \c{e} includes both \c{parse} and \c{quote}. We pick an origin
+The provenance of \code{e} includes both \code{parse} and \code{quote}. We pick an origin
 by traversing the graph from the root, following the left-hand side member of
 assignments until a leaf. It represents the provenance that is further modified
-to yield the expression passed to \eval. In our example, the origin is \c{parse}.
+to yield the expression passed to \eval. In our example, the origin is \code{parse}.
 
@@
 different provenances, they are counted multiple times. Strings correlate with
-dynamic code loading as done by \c{source} and \c{sys.source}. Few calls
-(\packageNbParseFromFileSites in total) consume the result of calling \c{parse}
+dynamic code loading as done by \code{source} and \code{sys.source}. Few calls
+(\packageNbParseFromFileSites in total) consume the result of calling \code{parse}
 on a file. Most build strings programmatically. We identified one function
-\c{invokeRestartInteractively} that prompts the user for input and \evals it.
+\code{invokeRestartInteractively} that prompts the user for input and \evals it.
 
 \paragraph{Discussion}
-The main use case for \eval involves using \c{match.call}, \c{substitute}, or
-\c{formals} to access and transform arguments of a function. Constructing
+The main use case for \eval involves using \code{match.call}, \code{substitute}, or
+\code{formals} to access and transform arguments of a function. Constructing
 expressions from strings is also a common use case. The constructed category
@@
 sites (over a third of the functions) do side effects in either \emph{Local} or
-\emph{Object} environments. This gives a ray of hope for a hypothetical R
-compiler. Even though \eval can do anything anywhere, the data suggests most
-effects are sane.
+\emph{Object} environments.
 
@@
 \paragraph{Discussion} In JavaScript assignments in \eval can happen in either
-local scope or less often, when called through an alias, in the global scope. In
-R, given the support for first-class environments, it can happen anywhere,
+local scope or less often, when called through an alias, in the global scope.
+In R, given the support for first-class environments, it can happen anywhere,
 making it \eval more dangerous than it already is. However, the data suggest
@@
 JavaScript,\footnote{\citep{ecoop11} shows that in the \emph{Interactive}
-scenario, \eval in Javascript performs, store events can reach up to 40\% of the
-events, and 7\% to 8\% of the \eval do side effects in the global scope. } and
-that over half of them happen in a predictable environment.
+  scenario, \eval in Javascript performs, store events can reach up to 40\% of
+the events, and 7\% to 8\% of the \eval do side effects in the global scope. }
+and that over half of them happen in a predictable environment. This gives a
+ray of hope for a hypothetical R compiler. Even though \eval can do anything
+anywhere, the data suggests most effects are sane.
 
@@
     the case that the block is only passed to \eval rather than being directly
-    part of in as in \lstinline|eval({...})|. Essentially, the block denotes a
-    fragment of a program to be evaluated in a particular way and in a particular
-    environment. The latter is what makes it different from 0-argument closure.
-
-    There is a number of use cases: unit testing frameworks (\eg \c{testthat},
-    \c{testit}), code benchmarking (\eg \c{rbenchmark}, \c{microbenchmark}),
-    running code in parallel (\eg \c{foreach}, \c{doParallel}), or deferring
-    code execution (\eg \c{withr}).
-
+    part of in as in \lstinline|eval({...})|. 
 \end{compactitem}

### 2021-08-11 paper/main.tex
Migrating \c to \code for normalized expression, environments, and origins of eval

@@
 \begin{tabular}{@{}p{.97\linewidth}}
-  \vspace{1mm}
-  \EE{$min(e)=\code{V}$} \\[-2mm]\small Values occur in 16\% of sites, 74\% of
+  \medskip\EE{$min(e)=\c{V}$} \\[-2mm]\small Values occur in 16\% of sites, 74\% of
   those are constants such as integers, the remainder evaluate to a value. \Eval
@@
 \end{lstlisting}
-\vspace{2mm}
-\EE{$min(e)=\code{X}$}\\[-2mm]\small Variable lookups are the most common shape,
+\\
+\medskip\EE{$min(e)=$\code{X}}\\[-2mm]\small Variable lookups are the most common shape,
 \packageNbSymbolVarSitePercent are simple variable names, e.g. \c x. The shapes
@@
 expressions are evaluated in a function environment.
-\vspace{2mm}
-\EE{$min(e)=\code{\$}$}\\[-2mm]\small This shape extends \c X to include lookup
+\\
+\medskip\EE{$min(e)=$\code{$}}\\[-2mm]\small This shape extends \c X to include lookup
 with the dollar operator, \eg~\c{x\$f}, and vector indexing, \eg~\c{x[42]} and
@@
 typically used in a function environment.
-
-\vspace{2mm}
-\EE{$min(e)=$~\code{<-}}\\[-2mm]\small This includes both direct
-assignment {\tt <-} and parent environment assignment {\tt <\,\!<-} as well as
-the \c{assign} function. It subsumes \c{\$}. Assignments represent the most
-obvious source of side effects in \eval. Most cases originate from trivial code
-generation where the assign expression is assembled using parse or substitute,
-often in a loop. For example (taken from \c{PopED::get\_rse}), the following
-code will populate the current scope with a number of variables:
-%
-\begin{lstlisting}
-for (i in seq_along(args)) eval(parse(text=paste(args[i], "<-", i)))
-\end{lstlisting}
-
-\vspace{2mm}
-\EE{$min(e)=\code{F()}$}\\[-2mm]\small Function calls without variable lookups or
+\\
+\medskip\EE{$min(e)=$~\code{<-}}\\[-2mm]\small This includes both direct assignment {\tt
+  <-} and parent environment assignment {\tt <\,\!<-} as well as the \c{assign}
+function. It subsumes \c{\$}. Assignments
+represent the most obvious source of side effects in \eval.
+\\
+\medskip\EE{$min(e)=$\code{F()}}\\[-2mm]\small Function calls without variable lookups or
 assignment. A variable is allowed the function position, looking up
@@
 is bound to a promise. This shape typically does not perform much work.
-\vspace{2mm}
-\framebox{$min(e)=\code{F(X)}$}~\EE{$min(e)=\code{F(F(X))}$}\\[-2mm]\small Function
+\\
+\medskip\framebox{$min(e)=$\code{F(X)}}~\EE{$min(e)=\c{F(F(X))}$}\\[-2mm]\small Function
 calls whose arguments may include variable references and nested calls. Together
@@
 \begin{tabular}{@{}p{.97\linewidth}}
-  \medskip\EE{$min(e)=$ \code{FUN}}\\[-2mm]\small Function definitions without any
+  \medskip\EE{$min(e)=$\code{FUN}}\\[-2mm]\small Function definitions without any
   computation, \eg~\c{function(x) x+1}. In addition to \c{FUN},
@@
   definitions nested in other expressions.
-
-\vspace{2mm}
-\EE{$min(e)=\code{BLOCK}$}\\[-2mm]\small Multi-statement code blocks, as
+\\
+\medskip\EE{$min(e)=$\code{BLOCK}}\\[-2mm]\small Multi-statement code blocks, as
 they can be large we do not try to normalize their contents. They execute in a
@@
 environments.
-
-\vspace{2mm}
-\EE{$min(e)=\code{model.frame}$}\\[-2mm]\small The \c{model.frame} function
+\\
+\medskip\EE{$min(e)=$\code{model.frame}}\\[-2mm]\small The \c{model.frame} function
 returns a dataframe resulting from fitting the model described in a formula.
@@
 sites (over a third of the functions) do side effects in either \emph{Local} or
-\emph{Object} environments.
+\emph{Object} environments. This gives a ray of hope for a hypothetical R
+compiler. Even though \eval can do anything anywhere, the data suggests most
+effects are sane.
 
@@
 \paragraph{Discussion} In JavaScript assignments in \eval can happen in either
-local scope or less often, when called through an alias, in the global scope.
-In R, given the support for first-class environments, it can happen anywhere,
+local scope or less often, when called through an alias, in the global scope. In
+R, given the support for first-class environments, it can happen anywhere,
 making it \eval more dangerous than it already is. However, the data suggest
@@
 JavaScript,\footnote{\citep{ecoop11} shows that in the \emph{Interactive}
-  scenario, \eval in Javascript performs, store events can reach up to 40\% of
-the events, and 7\% to 8\% of the \eval do side effects in the global scope. }
-and that over half of them happen in a predictable environment. This gives a
-ray of hope for a hypothetical R compiler. Even though \eval can do anything
-anywhere, the data suggests most effects are sane.
+scenario, \eval in Javascript performs, store events can reach up to 40\% of the
+events, and 7\% to 8\% of the \eval do side effects in the global scope. } and
+that over half of them happen in a predictable environment.
 
@@
     the case that the block is only passed to \eval rather than being directly
-    part of in as in \lstinline|eval({...})|. 
+    part of in as in \lstinline|eval({...})|. Essentially, the block denotes a
+    fragment of a program to be evaluated in a particular way and in a particular
+    environment. The latter is what makes it different from 0-argument closure.
+
+    There is a number of use cases: unit testing frameworks (\eg \c{testthat},
+    \c{testit}), code benchmarking (\eg \c{rbenchmark}, \c{microbenchmark}),
+    running code in parallel (\eg \c{foreach}, \c{doParallel}), or deferring
+    code execution (\eg \c{withr}).
+
 \end{compactitem}

### 2021-08-11 paper/main.tex
Better spacing for \code and normalized expressions

@@
 \begin{tabular}{@{}p{.97\linewidth}}
-  \medskip\EE{$min(e)=\c{V}$} \\[-2mm]\small Values occur in 16\% of sites, 74\% of
+  \medskip\EE{$min(e)=$ \code{V}} \\[-2mm]\small Values occur in 16\% of sites, 74\% of
   those are constants such as integers, the remainder evaluate to a value. \Eval
@@
 \\
-\medskip\EE{$min(e)=$\code{X}}\\[-2mm]\small Variable lookups are the most common shape,
+\medskip\EE{$min(e)=$ \code{X}}\\[-2mm]\small Variable lookups are the most common shape,
 \packageNbSymbolVarSitePercent are simple variable names, e.g. \c x. The shapes
@@
 \\
-\medskip\EE{$min(e)=$\code{$}}\\[-2mm]\small This shape extends \c X to include lookup
+\medskip\EE{$min(e)=$ \code{$}}\\[-2mm]\small This shape extends \c X to include lookup
 with the dollar operator, \eg~\c{x\$f}, and vector indexing, \eg~\c{x[42]} and
@@
 \\
-\medskip\EE{$min(e)=$\code{F()}}\\[-2mm]\small Function calls without variable lookups or
+\medskip\EE{$min(e)=$ \code{F()}}\\[-2mm]\small Function calls without variable lookups or
 assignment. A variable is allowed the function position, looking up
@@
 \\
-\medskip\framebox{$min(e)=$\code{F(X)}}~\EE{$min(e)=\c{F(F(X))}$}\\[-2mm]\small Function
+\medskip\framebox{$min(e)=$ \code{F(X)}}~\EE{$min(e)=\c{F(F(X))}$}\\[-2mm]\small Function
 calls whose arguments may include variable references and nested calls. Together
@@
 \begin{tabular}{@{}p{.97\linewidth}}
-  \medskip\EE{$min(e)=$\code{FUN}}\\[-2mm]\small Function definitions without any
+  \medskip\EE{$min(e)=$ \code{FUN}}\\[-2mm]\small Function definitions without any
   computation, \eg~\c{function(x) x+1}. In addition to \c{FUN},
@@
 \\
-\medskip\EE{$min(e)=$\code{BLOCK}}\\[-2mm]\small Multi-statement code blocks, as
+\medskip\EE{$min(e)=$ \code{BLOCK}}\\[-2mm]\small Multi-statement code blocks, as
 they can be large we do not try to normalize their contents. They execute in a
@@
 \\
-\medskip\EE{$min(e)=$\code{model.frame}}\\[-2mm]\small The \c{model.frame} function
+\medskip\EE{$min(e)=$ \code{model.frame}}\\[-2mm]\small The \c{model.frame} function
 returns a dataframe resulting from fitting the model described in a formula.

### 2021-08-11 paper/main.tex
Move the subset example

--- 2021-08-11 paper/main.tex
@@
 
+The use of \eval with \code{substitute} is a general pattern that we often see
+in R packages. It is used to make the interfaces more flexible. For example the
+\c{subset}, which selects columns in a data frame, uses it to allow to specify
+the selected columns using either the column name or index. For example
+\c{subset(df, sel=c(1,B,x))} returns a subset of \code{df} with the first
+column, the column named \code{B} and whatever column is at index the variable
+\code{x} holds.
+
+It is implemented using \code{eval(substitute(sel), nl, parent.frame())}. The
+first argument returns the expression passed to \code{sel}. The second is a
+named list that contains the names of the columns from \code{df} and which will
+be used to evaluate the expression. The last parameter indicates which
+environment should be used to look up bindings not found in \c{nl}.
+
 \paragraph{Boilerplate Removal} The combination of \code{match.call} and \code{eval}

### 2021-08-11 paper/main.tex
Fix
Removed example from min

@@
   \EE{$min(e)=\code{V}$} \\[-2mm]\small Values occur in 16\% of sites, 74\% of
-  those are constants such as integers, the remainder evaluate to a value. \Eval
+  those are constants such as integers, the remainder evaluate to a value (\packageNbCallSitesUniqueActualValue of
+  sites only ever see a value). \Eval
   is needed when the expression denoting the values is constructed
@@
   returns it unchanged, often occur when a computation has a default path and
-  another, more interesting, path that requires evaluation. The following is a
-  call site that always see a string; \packageNbCallSitesUniqueActualValue of
-  sites only ever see a value. This shape usually run in few interpreter steps.
+  another, more interesting, path that requires evaluation.
+  This shape usually run in few interpreter steps.
   The environment in which they evaluate is mostly irrelevant.
-\begin{lstlisting}
-  eval(paste("A~",paste(paste(names(X[,-1])),collapse="+")))
-\end{lstlisting}
+
   \vspace{2mm}
@@
 generation where the assign expression is assembled using parse or substitute,
-often in a loop. For example (taken from \c{PopED::get\_rse}), the following
-code will populate the current scope with a number of variables:
-%
-\begin{lstlisting}
-for (i in seq_along(args)) eval(parse(text=paste(args[i], "<-", i)))
-\end{lstlisting}
+often in a loop.
 
@@
 is bound to a promise. This shape typically does not perform much work.
+
 \vspace{2mm}
@@
 
-Out of the minimized expressions, it is \c{BLOCK}, \c{<-} and \c{F(F(X)} that
+Out of the minimized expressions, it is \c{BLOCK}, \c{<-} and \c{F(F(X))} that
 contribute to the vast majority of side effects. Concretely, 92\% of \c{BLOCK},

### 2021-08-11 paper/main.tex
Fix

@@
 in R packages. It is used to make the interfaces more flexible. For example the
-\c{subset}, which selects columns in a data frame, uses it to allow to specify
+\code{subset}, which selects columns in a data frame, uses it to allow to specify
 the selected columns using either the column name or index. For example
-\c{subset(df, sel=c(1,B,x))} returns a subset of \code{df} with the first
+\code{subset(df, sel=c(1,B,x))} returns a subset of \code{df} with the first
 column, the column named \code{B} and whatever column is at index the variable

### 2021-08-11 paper/main.tex
revert \c to \code

--- 2021-08-11 paper/main.tex
@@
   \bf $min(e)$& \bf \#sites & \bf \%sites & \bf \#packages & \bf \#operations & \bf \%envir & \bf example\\\midrule
-\code X&\packageMinimizedcallsitesa &\packageMinimizedpropsitesa &\packageMinimizedpackagea &\packageMinimizedmedianoperationsaRnd &\packageMinimizedpercentparentframesa & \code{y+1}\\
-\code{F(F(X))} & \packageMinimizedcallsitesb  & \packageMinimizedpropsitesb & \packageMinimizedpackageb  & \packageMinimizedmedianoperationsbRnd & \packageMinimizedpercentparentframesb & \code{gbov( mean(x), a-1)}\\
-\code{V}&\packageMinimizedcallsitesc &\packageMinimizedpropsitesc &\packageMinimizedpackagec &\packageMinimizedmedianoperationscRnd &\packageMinimizedpercentparentframesc& \code{c(42,21,0)}\\
-\code{F(X)}& \packageMinimizedcallsitesd & \packageMinimizedpropsitesd & \packageMinimizedpackaged & \packageMinimizedmedianoperationsdRnd & \packageMinimizedpercentparentframesd & \code{seq\_len(iters)} \\
-\code{$} & \packageMinimizedcallsitese & \packageMinimizedpropsitese & \packageMinimizedpackagee & \packageMinimizedmedianoperationseRnd & \packageMinimizedpercentparentframese & \code{DF\$B}\\
-\code{model.frame}& \packageMinimizedcallsitesf & \packageMinimizedpropsitesf & \packageMinimizedpackagef & \packageMinimizedmedianoperationsfRnd & \packageMinimizedpercentparentframesf &  \code{model.frame(formula = Z ~  U)}   \\
-\code{F()}& \packageMinimizedcallsitesg & \packageMinimizedpropsitesg & \packageMinimizedpackageg & \packageMinimizedmedianoperationsgRnd & \packageMinimizedpercentparentframesg & \code{rgamma(3, 2, n = 10L)} \\
-\code{FUN} & \packageMinimizedcallsitesh & \packageMinimizedpropsitesh & \packageMinimizedpackageh & \packageMinimizedmedianoperationshRnd & \packageMinimizedpercentparentframesh & \code{function(x, y) x + 3 * y} \\
-\code{<-} & \packageMinimizedcallsitesi  & \packageMinimizedpropsitesi & \packageMinimizedpackagei & \packageMinimizedmedianoperationsiRnd & \packageMinimizedpercentparentframesi & \code{x[1, 2:3, 2:3] <- value}\\
-\code{BLOCK} & \packageMinimizedcallsitesj & \packageMinimizedpropsitesj & \packageMinimizedpackagej & \packageMinimizedmedianoperationsjRnd & \packageMinimizedpercentparentframesj & \code{{ write.csv(iris, tf) ; file.size(tf) }} \\\bottomrule
+\c X&\packageMinimizedcallsitesa &\packageMinimizedpropsitesa &\packageMinimizedpackagea &\packageMinimizedmedianoperationsaRnd &\packageMinimizedpercentparentframesa & \c{y+1}\\
+\c{F(F(X))} & \packageMinimizedcallsitesb  & \packageMinimizedpropsitesb & \packageMinimizedpackageb  & \packageMinimizedmedianoperationsbRnd & \packageMinimizedpercentparentframesb & \c{gbov( mean(x), a-1)}\\
+\c{V}&\packageMinimizedcallsitesc &\packageMinimizedpropsitesc &\packageMinimizedpackagec &\packageMinimizedmedianoperationscRnd &\packageMinimizedpercentparentframesc& \c{c(42,21,0)}\\
+\c{F(X)}& \packageMinimizedcallsitesd & \packageMinimizedpropsitesd & \packageMinimizedpackaged & \packageMinimizedmedianoperationsdRnd & \packageMinimizedpercentparentframesd & \c{seq\_len(iters)} \\
+\c{\$} & \packageMinimizedcallsitese & \packageMinimizedpropsitese & \packageMinimizedpackagee & \packageMinimizedmedianoperationseRnd & \packageMinimizedpercentparentframese & \c{DF\$B}\\
+\c{model.frame}& \packageMinimizedcallsitesf & \packageMinimizedpropsitesf & \packageMinimizedpackagef & \packageMinimizedmedianoperationsfRnd & \packageMinimizedpercentparentframesf &  \c{model.frame(formula = Z $\sim$ U)}   \\
+\c{F()}& \packageMinimizedcallsitesg & \packageMinimizedpropsitesg & \packageMinimizedpackageg & \packageMinimizedmedianoperationsgRnd & \packageMinimizedpercentparentframesg & \c{rgamma(3, 2, n = 10L)} \\
+\c{FUN} & \packageMinimizedcallsitesh & \packageMinimizedpropsitesh & \packageMinimizedpackageh & \packageMinimizedmedianoperationshRnd & \packageMinimizedpercentparentframesh & \c{function(x, y) x + 3 * y} \\
+\c{<-} & \packageMinimizedcallsitesi  & \packageMinimizedpropsitesi & \packageMinimizedpackagei & \packageMinimizedmedianoperationsiRnd & \packageMinimizedpercentparentframesi & \c{x[1, 2:3, 2:3] <- value}\\
+\c{BLOCK} & \packageMinimizedcallsitesj & \packageMinimizedpropsitesj & \packageMinimizedpackagej & \packageMinimizedmedianoperationsjRnd & \packageMinimizedpercentparentframesj & \c{\{ write.csv(iris, tf) ; file.size(tf) \}} \\\bottomrule
 \end{tabular}
@@
 folding of arithmetic and string expressions over base operators, for instance
-$min(\c{1+1})$ simplifies to \code{V}, on the ground that addition of values
-likely returns a value. Value simplification takes $min$(\code{c(1,2,3+2)}) to
-\code{V} as a complex vector containing values is a value. Variable absorption
-has $min$(\code{x+y}) becomes \code{X}, this motivated by the fact that addition is
-not an ``interesting'' operation, and that \code{X} stands in for any number of
+$min(\c{1+1})$ simplifies to $\c{V}$, on the ground that addition of values
+likely returns a value. Value simplification takes $min(\c{c(1,2,3+2)})$ to
+$\c{V}$ as a complex vector containing values is a value. Variable absorption
+has $min(\c{x+y})$ becomes $\c{X}$, this motivated by the fact that addition is
+not an ``interesting'' operation, and that \c X stands in for any number of
 variable lookups. Function absorption simplifies nested functions to keep an
-abstraction of the nesting $min$(\code{g(f(x),h(z))})$=$\code{F(F(X))}. There are other
+abstraction of the nesting $min(\c{g(f(x),h(z))})=\c{F(F(X))}$. There are other
 simplifications, the full list can be found in our artifact. It is worth
-mentioning that these simplifications are heuristics, in R \code{1+1} is not
-necessary \code{2}. The addition operator can be redefined, but it typically isn't
+mentioning that these simplifications are heuristics, in R \c{1+1} is not
+necessary \c 2. The addition operator can be redefined, but it typically isn't
 or at least not in a way that would invalidate arithmetic.
@@
 \end{lstlisting}
-\vspace{2mm}
-\EE{$min(e)=\code{X}$}\\[-2mm]\small Variable lookups are the most common shape,
+\\
+\medskip\EE{$min(e)=\c{X}$}\\[-2mm]\small Variable lookups are the most common shape,
 \packageNbSymbolVarSitePercent are simple variable names, e.g. \c x. The shapes
@@
 expressions are evaluated in a function environment.
-\vspace{2mm}
-\EE{$min(e)=\code{\$}$}\\[-2mm]\small This shape extends \c X to include lookup
+\\
+\medskip\EE{$min(e)=\c{\$}$}\\[-2mm]\small This shape extends \c X to include lookup
 with the dollar operator, \eg~\c{x\$f}, and vector indexing, \eg~\c{x[42]} and
@@
 typically used in a function environment.
-
-\vspace{2mm}
-\EE{$min(e)=$~\code{<-}}\\[-2mm]\small This includes both direct
-assignment {\tt <-} and parent environment assignment {\tt <\,\!<-} as well as
-the \c{assign} function. It subsumes \c{\$}. Assignments represent the most
-obvious source of side effects in \eval. Most cases originate from trivial code
-generation where the assign expression is assembled using parse or substitute,
-often in a loop. For example (taken from \c{PopED::get\_rse}), the following
-code will populate the current scope with a number of variables:
-%
-\begin{lstlisting}
-for (i in seq_along(args)) eval(parse(text=paste(args[i], "<-", i)))
-\end{lstlisting}
-
-\vspace{2mm}
-\EE{$min(e)=\code{F()}$}\\[-2mm]\small Function calls without variable lookups or
+\\
+\medskip\EE{$min(e)=$~\c{<-}}\\[-2mm]\small This includes both direct assignment {\tt
+  <-} and parent environment assignment {\tt <\,\!<-} as well as the \c{assign}
+function. It subsumes \c{\$}. Assignments
+represent the most obvious source of side effects in \eval.
+\\
+\medskip\EE{$min(e)=\c{F()}$}\\[-2mm]\small Function calls without variable lookups or
 assignment. A variable is allowed the function position, looking up
@@
 is bound to a promise. This shape typically does not perform much work.
-\vspace{2mm}
-\framebox{$min(e)=\code{F(X)}$}~\EE{$min(e)=\code{F(F(X))}$}\\[-2mm]\small Function
+\\
+\medskip\framebox{$min(e)=\c{F(X)}$}~\EE{$min(e)=\c{F(F(X))}$}\\[-2mm]\small Function
 calls whose arguments may include variable references and nested calls. Together
@@
 \begin{tabular}{@{}p{.97\linewidth}}
-  \medskip\EE{$min(e)=$ \code{FUN}}\\[-2mm]\small Function definitions without any
+  \medskip\EE{$min(e)=\c{FUN}$}\\[-2mm]\small Function definitions without any
   computation, \eg~\c{function(x) x+1}. In addition to \c{FUN},
@@
   definitions nested in other expressions.
-
-\vspace{2mm}
-\EE{$min(e)=\code{BLOCK}$}\\[-2mm]\small Multi-statement code blocks, as
+\\
+\medskip\EE{$min(e)=\c{BLOCK}$}\\[-2mm]\small Multi-statement code blocks, as
 they can be large we do not try to normalize their contents. They execute in a
@@
 environments.
-
-\vspace{2mm}
-\EE{$min(e)=\code{model.frame}$}\\[-2mm]\small The \c{model.frame} function
+\\
+\medskip\EE{$min(e)=\c{model.frame}$}\\[-2mm]\small The \c{model.frame} function
 returns a dataframe resulting from fitting the model described in a formula.
@@
 \c{magrittr} package which is used to compose functions, \eg instead of
-\code{f(g(h(x),y),z)} one can write \code{h(x) \%>\% g(y) \%>\% f(z)}. As there many
+\c{f(g(h(x),y),z)} one can write \c{h(x) \%>\% g(y) \%>\% f(z)}. As there many
 different patterns of use, there are also many different shapes. In comparison
@@
 other hand, there are \packageNbSimpleMinimizedOne sites that only receive one
-of the simple shape (\ie\xspace \code{X}, \code{V}, \code{\$} or \code{<-}), and
+of the simple shape (\ie\xspace \c{X}, \c{V}, \c{\$} or \c{<-}), and
 \packageNbSimpleMinimizedMore sites get multiple simple shapes. Among these
 sites, there are many cases where \eval could be replaced by simpler constructs
-such as \code{get} to lookup a name in an environment, \code{assign} to set a value to
-a name in an environment, or \code{do.call} to call a function reflectively.
+such as \c{get} to lookup a name in an environment, \c{assign} to set a value to
+a name in an environment, or \c{do.call} to call a function reflectively.
 
@@
 \item \emph{Function:} environments for the local variables of some function
-  currently active on the call stack. Obtained by calling \code{parent.frame()} or
-  \code{sys.frame()}.
+  currently active on the call stack. Obtained by calling \c{parent.frame()} or
+  \c{sys.frame()}.
 \item \emph{Synthetic:} environments built from data structures such as lists,
-  data.frames, or constructed explicitly with \code{new.env}, \code{list2env}, or
-  \code{as.environment} or the empty environment.
+  data.frames, or constructed explicitly with \c{new.env}, \c{list2env}, or
+  \c{as.environment} or the empty environment.
 \item \emph{Global:} environments in which scripts or interactive commands are
@@
 Each synthetic environment has parent that is specified when calling
-\code{new.env}. When \code{envir} is a list or a data frame, \eval uses its third
-argument (\code{enclos}). The parent is used to search for variables not found in
+\c{new.env}. When \c{envir} is a list or a data frame, \eval uses its third
+argument (\c{enclos}). The parent is used to search for variables not found in
 its child (side-effects stay in the child). Table~\ref{tab:newenvs} shows parent
@@
 Global \evals split between intentional and accidental one. Direct references to
-the top-level, using \code{globalenv()} or \code{.GlobalEnv}, are rare; they occur in
+the top-level, using \c{globalenv()} or \c{.GlobalEnv}, are rare; they occur in
 only \packageNbExplicitGlobalSites sites. Thus we suspect most uses of global
@@
 expectation is that the expression was extracted from a promise using
-\code{substitute}, perhaps modified, and then evaluated with the environment coming
+\c{substitute}, perhaps modified, and then evaluated with the environment coming
 from the same promise. We believe the case where \eval is provided the results
@@
 the values that are given to \eval. The following example yields the provenance
-graph with five nodes on Figure~\ref{fig:provgraph} for expression \code{e}.
+graph with five nodes on Figure~\ref{fig:provgraph} for expression \c{e}.
 
@@
 
-The provenance of \code{e} includes both \code{parse} and \code{quote}. We pick an origin
+The provenance of \c{e} includes both \c{parse} and \c{quote}. We pick an origin
 by traversing the graph from the root, following the left-hand side member of
 assignments until a leaf. It represents the provenance that is further modified
-to yield the expression passed to \eval. In our example, the origin is \code{parse}.
+to yield the expression passed to \eval. In our example, the origin is \c{parse}.
 
@@
 different provenances, they are counted multiple times. Strings correlate with
-dynamic code loading as done by \code{source} and \code{sys.source}. Few calls
-(\packageNbParseFromFileSites in total) consume the result of calling \code{parse}
+dynamic code loading as done by \c{source} and \c{sys.source}. Few calls
+(\packageNbParseFromFileSites in total) consume the result of calling \c{parse}
 on a file. Most build strings programmatically. We identified one function
-\code{invokeRestartInteractively} that prompts the user for input and \evals it.
+\c{invokeRestartInteractively} that prompts the user for input and \evals it.
 
 \paragraph{Discussion}
-The main use case for \eval involves using \code{match.call}, \code{substitute}, or
-\code{formals} to access and transform arguments of a function. Constructing
+The main use case for \eval involves using \c{match.call}, \c{substitute}, or
+\c{formals} to access and transform arguments of a function. Constructing
 expressions from strings is also a common use case. The constructed category
@@
 environment which has data frame columns available as bindings.
-
-The use of \eval with \code{substitute} is a general pattern that we often see
-in R packages. It is used to make the interfaces more flexible. For example the
-\c{subset}, which selects columns in a data frame, uses it to allow to specify
-the selected columns using either the column name or index. For example
-\c{subset(df, sel=c(1,B,x))} returns a subset of \code{df} with the first
-column, the column named \code{B} and whatever column is at index the variable
-\code{x} holds.
-
-It is implemented using \code{eval(substitute(sel), nl, parent.frame())}. The
-first argument returns the expression passed to \code{sel}. The second is a
-named list that contains the names of the columns from \code{df} and which will
-be used to evaluate the expression. The last parameter indicates which
-environment should be used to look up bindings not found in \c{nl}.
 

### 2021-08-11 paper/main.tex

@@
 \begin{tabular}{@{}p{.97\linewidth}}
-  \medskip\EE{$min(e)=\c{V}$} \\[-2mm]\small Values occur in 16\% of sites, 74\% of
-  those are constants such as integers, the remainder evaluate to a value. \Eval
+  \medskip\EE{$min(e)=\c{V}$}\\[-2mm]\small Values occur in 16\% of sites, 74\% of
+  those are constants such as integers, the remainder evaluate to a value (\packageNbCallSitesUniqueActualValue of
+  sites only ever see a value). \Eval
   is needed when the expression denoting the values is constructed
@@
   returns it unchanged, often occur when a computation has a default path and
-  another, more interesting, path that requires evaluation. The following is a
-  call site that always see a string; \packageNbCallSitesUniqueActualValue of
-  sites only ever see a value. This shape usually run in few interpreter steps.
+  another, more interesting, path that requires evaluation.
+  This shape usually run in few interpreter steps.
   The environment in which they evaluate is mostly irrelevant.
@@
 environment which has data frame columns available as bindings.
+
+The use of \eval with \code{substitute} is a general pattern that we often see
+in R packages. It is used to make the interfaces more flexible. For example the
+\code{subset}, which selects columns in a data frame, uses it to allow to specify
+the selected columns using either the column name or index. For example
+\code{subset(df, sel=c(1,B,x))} returns a subset of \code{df} with the first
+column, the column named \code{B} and whatever column is at index the variable
+\code{x} holds.
+
+It is implemented using \code{eval(substitute(sel), nl, parent.frame())}. The
+first argument returns the expression passed to \code{sel}. The second is a
+named list that contains the names of the columns from \code{df} and which will
+be used to evaluate the expression. The last parameter indicates which
+environment should be used to look up bindings not found in \c{nl}.
 

### 2021-08-11 paper/main.tex
Bring back lost changes

@@
   The environment in which they evaluate is mostly irrelevant.
-\begin{lstlisting}
-  eval(paste("A~",paste(paste(names(X[,-1])),collapse="+")))
-\end{lstlisting}
 \\
@@
   <-} and parent environment assignment {\tt <\,\!<-} as well as the \c{assign}
-function. It subsumes \c{\$}. Assignments
-represent the most obvious source of side effects in \eval.
+function. It subsumes \c{\$}. Assignments represent the most obvious source
+of side effects in \eval. Most cases originate from trivial code generation
+where the assign expression is assembled using parse or substitute, often in
+a loop.
 \\
@@
 environments.
+Essentially, the block denotes a fragment of a program to be evaluated in a
+particular way and in a particular environment. The latter is what makes it
+different from 0-argument closure. There is a number of use cases: unit testing
+frameworks (\eg \c{testthat}, \c{testit}), code benchmarking (\eg
+\c{rbenchmark}, \c{microbenchmark}), running code in parallel (\eg \c{foreach},
+\c{doParallel}), or deferring code execution (\eg \c{withr}).
+
 \\

### 2021-08-11 paper/main.tex
conclusion and discussion

@@
 \newcommand{\code}[1]{\lstinline |#1|\xspace}
-%\renewcommand{\c}[1]{\lstinline |#1|\xspace}
+\renewcommand{\k}[1]{\lstinline |#1|\xspace}
 % cf. https://tex.stackexchange.com/a/144640
@@
 
-\paragraph{Core function variants} Many packages use \eval to implement variations of the
+\paragraph{Core function variants} Many packages use \eval to implement variations of
+the
 \code{base} R functions. For example, the \code{igraph} package provides the
@@
 
-\paragraph{Miscellaneous} Many uses of \eval are diverse and do not fit into a
-category by themselves. For example, packages for unit testing (testthat,
-testit, etc.), code benchmarking (rbenchmark, microbenchmark, etc.), and,
-running code in parallel (foreach, doParallel, etc.), use \eval to evaluate
-program fragments in a particular way in specific environments. Many numerical
-computation and mathematical modeling packages use \eval to evaluate chunks of
-code or bodies of functions in specific environments. Consider the
-\code{findZeros} function from \code{mosaic} package which is used to compute
-the zero of a function.
-
-\begin{figure}[h]
-\begin{lstlisting}
-findZeros <- function(expr, ..., xlim=c(near-within, near+within), near=0, within=Inf) {
-  dots <- list(...)
-
-  pfun <- function(x){  # removed . from name, was .x
-    mydots <- dots
-    mydots[[rhsVars]] <- x
-    eval( lhs(expr), envir=mydots, enclos=parent.frame() )
-  }
-  searchx <- sort(unique(internal.near   + c(0, leftseq, middleseq, rightseq)))
-  y <- sapply( searchx, pfun )
-\end{lstlisting}
-  \caption{mosaic::findZeros}
-\end{figure}
-
-This function can be invoked as: \code{findZeros( A*sin(2*pi*t/P) ~ t,
-  xlim=c(0,100), P=50, A=2)}. It will evaluate \code{A*sin(2*pi*t/P)} inside
-\code{pfun}. It calls \code{pfun} repeatedly with different values of \code{t}
-to find the point at which the expression is zero.
-
 \subsection{Shapes redux}
 
-In Table~\ref{tab:minimizedexpressions} we have shown the most common shapes of
-expressions passed to \eval. Here, we illustrate them in concrete instances.
-
-\begin{compactitem}[---]
-
-  \item \emph{Variable lookup and values, and indexing \c{(X, V, \$)}}
+\emph{Variable lookup and values, and indexing \c{(X, V, \$)}}
     constitute one of the most common use case for \eval. Let's consider the
@@
 
-  \item \emph{Assignments \c{(<-)}} appear in the cases where a developer wants
+ \emph{Assignments \c{(<-)}} appear in the cases where a developer wants
     to create or update a binding in a specific environment where either the
@@
 
-  \item \emph{Function definition \c{(FUN)}} is mostly used in conjunction with
-    code generation where new functions are synthesized using either \c{parse}
-    or \c{substitute}. Some FFI frameworks use this to automatically generate R
-    binding to native code. For example, \c{Rcpp}, which bridges R with C++,
-    uses \eval to generate R functions for C++ methods:\\
-    \lstinline|eval(substitute(function(...) .CppObj$M(...), list(M=...)), env)|.
-
-  \item \emph{Function calls \c{(F(), F(X), F(F(X)))}} represent two very
+ \emph{Function calls \c{(F(), F(X), F(F(X)))}} represent two very
     frequent patterns: \emph{code generation} and \emph{call reflection}. There
@@
 
-  \item \emph{Block \c{(BLOCK)}} can appear anywhere, but it is almost always
-    the case that the block is only passed to \eval rather than being directly
-    part of in as in \lstinline|eval({...})|. 
-\end{compactitem}
-
-%\todo{Say somewhere that the R community commonly calls call/argument capturing
-%(with substitute or match.call), Non-Standard Evaluation.}
-
-\subsection{Discussion}
-
-\Eval in R is used chiefly for meta-programming purposes and to access
-environments other than the local one. In Javascript, patterns are ad-hoc, such
-as json loading and parsing, and can often be rewritten without \eval, such as a
-function or method call, or an assignment to a local variable or an object
-property. For R, the data suggests that in some cases, the use of \eval in the
-\code{X, V, \$, <-, FUN} and the function calls related to code generation could
-replaced by more specific functions. For instance, variable lookup in any
-environment can be performed with \code{get}, and assignment in any environment,
-with \code{assign}. Building a call and executing can be done with \code{do.call}.
-Nevertheless, many of the observed \eval usage in R, particularly, the ones
-related to statistical modeling and DSLs are more sophisticated than those
-reported for JavaScript.
-
 \section{Discussion}
 
-\subsection{Removing \eval}
-
-Similarly to JavaScript, there are also unnecessary uses of \eval. These uses of
-\eval can be replaced with alternate functions yielding the same results. For
-example, the \code{PerformanceAnalytics} package contains a function
-\code{chart.QQPlot} that uses \eval to resolve a string into function and another
-to call it and assign its results into a variable:
+In R, \eval is used chiefly for meta-programming and access remote environments.
+Unlike, JavaScript which had a small set of well-defined patterns that could be
+rewritten without \eval, our results suggest that \eval is an integral part of
+programming in R.
+
+The cases where \eval appears to be overkill include simple expression shapes
+such as \c X, \c V, \$, \c{<-}, and \c{FUN}. Variable lookup in a remote
+environment can be performed by the builtin \code{get} function, a similar case
+can be made for assignment and the \code{assign} function. There is a dedicated
+\c{do.call} function to build function calls. Unnecessary uses of \eval
+often can be uncovered by manual analysis.
+
+\begin{lstlisting}
+PerformanceAnalytics::chart.QQPlot <- function (d="norm",dp,...) {
+   q.f <- eval(parse(text=paste("q",d)))
+   z <- NULL
+   eval(parse(text=paste("z<-q.f(",dp,",...)")))
+  }
+\end{lstlisting}
+
+The above uses \eval to first lookup a target function name from a constructed
+string, the it constructs a call to the target that include arguments to the
+current functions. \Eval is overkill for this code, a semantically equivalent
+function body uses \c{get} to perform the lookup and then calls directly.
+
 \begin{lstlisting}
  function (R, d="norm", dp, ...) {
-	  q.f <- eval(parse(text=paste("q",d,sep=""))); z <- NULL
-  	eval(parse(text=paste("z<-q.f(",dp,",...)")))
-  }
-\end{lstlisting}
-  In both cases, there is no need for \eval and the function can be rewritten as
+   q.f <- get(paste0("q",d))
+   z <- q.f(dp, ...)
+ }
+\end{lstlisting}
+
+Another example, from \c{coin::copyslots}, constructs an expression that access
+fields of an object and copy them. This can be replaced by symbolic field reads
+and writes.
+
 \begin{lstlisting}
-  function (R, d="norm", dp, ...) {
-	  q.f <- get(paste0("q",d)); z <- q.f(dp, ...)
-  }
-\end{lstlisting}
-  or even to a one-liner \code{do.call(paste0("q",d), c(dp, list(...)))}.
-
-  The \code{copyslots} function from the \code{coin} package uses \eval to copy the
-  slots from one \code{S4} object to another. It performs
-  \code{eval(str2lang(paste0("target@", s, " <- source@", s)))} for each slot name
-  \code{s}. This can be replaced by the expression \code{slot(target, s) <- slot(source,
-    s)} which uses the \code{slot<-} and \code{slot} functions to assign and retrieve
-  the slots by name.
-
-  Another examples is in the \code{configr::config.funs.par} function whose
-  simplified definition is shown below.
+  eval(str2lang(paste0("target@", s, " <- source@", s)))
+\end{lstlisting}
+
+As a last example, \code{configr::config.funs.par} construct three expressions
+that are then fed to \eval. Semantically equivalent code would use \c{get}
+to lookup \c{fun}, and then directly execute the three expressions.
 
 \begin{lstlisting}
-config.funs.par <- function(fun = "", ...) {
-  args.all <- as.list(match.call()); args.all <- args.all[names(args.all) != ""];
+ config.funs.par <- function(fun="",...) {
+   args.all <- as.list(match.call())
+   args.all <- args.all[names(args.all) != ""]
    parameters <- list()
-  text.1 <- sprintf("parameters <- names(as.list(args(%s)))", fun)
-  text.2 <- "parameters <- parameters[parameters != '']"
-  text.3 <- "parameters <- args.all[names(args.all) %in% parameters]"
-  for (j in c(text.1, text.2, text.3)) {
+   t1 <- sprintf("parameters <- names(as.list(args(%s)))", fun)
+   t2 <- "parameters <- parameters[parameters != '']"
+   t3 <- "parameters <- args.all[names(args.all) %in% parameters]"
+   for (j in c(t1,t2,t3))
     eval(parse(text = j))
-  }
    return(parameters)
@@
 
-  The function appears to use \eval since the function name \code{fun} is received
-  as a string. This string can be resolved to the function object using the
-  \code{get} function and the contents of \code{text.1}, \code{text.2}, and \code{text.3}
-  can be written directly without the need to convert them to text for
-  evaluation using \eval.
-
-The problem is that these cases are not easy to spot. Because of the
-limits of dynamic analysis, we only have a partial coverage and cannot determine
-if for example, \eval sites classified as \c{X, V} will not be eventually called
-with an expression in a \c{F(), F(X)} or \c{F(F(X))} category.
-
-\subsection{Alternatives to \eval}
-
-An alternative implementation of \eval, \c{eval\_tidy}, is provided by the
-\c{rlang} package. This forms the base of the tidy evaluation framework used by
-the libraries collectively referred to as the \c{tidyverse}~\cite{tidyverse}. It
-differs from the \c{eval} function in two ways. First, it supports evaluation of
-Quosures, which are also custom objects used by the \c{tidyverse} libraries for
-metaprogramming. Quosures package expression with an evaluation environment. The
-\c{tidy\_eval} function extracts this expression and evaluates it in that
-environment, ignoring the environment passed to it explicitly as an argument.
-Second, the \c{eval\_tidy} function also accepts a data mask argument.This is a
-set of bindings supplied by the user as a data frame or a named vector or list.
-The data mask environment take precedence over the environment passed for
-evaluation. Since the expression is always evaluated in the data mask, it has
-different semantics compared to \c{base::eval}. Effects such as assignment using
-\c{<-} happen in the data mask environment. Expressions such as \c{return()} do
-not work since the data mask environment does not correspond to a frame on the
-call stack.
+The challenge for automation is that such examples are not easy to find. Dynamic
+analysis is limited due to coverage issues, and static analysis of R remains an
+open problem.
+
+We have not yet touched on alternative implementations of \eval. The tidyverse
+libraries include their own implementation calls
+\c{rlang::eval\_tidy}~\cite{tidyverse}. This departs from \eval in two ways.
+First, it supports evaluation of quosures, custom objects used by for
+metaprogramming that bundle expressions with an environment. \c{eval\_tidy}
+accepts a data mask argument, a set of bindings such as a dataframe or list,
+which takes precedence over the environment. Effects such as assignments happen
+in the data mask. Expressions such as \c{return()} do not work since the data
+mask does not correspond to a frame on the call stack.
 
@@
 
-In this paper, we provide a large-scale study of the use of the \eval function
-in the R programming language. It is based primarily on a corpus of
-\CranRunnableScripts scripts extracted from \CranPackages R packages. As
-expected, \eval is used mainly in libraries. We have found just a few \eval
-call sites in scripts coming from a large corpus of Kaggle representing code
-written by R practitioners. We used an instrumented version of the R virtual
-machine, \rdyntrace, to capture all \eval calls, including the relevant
-side effects that happen within these calls. Several insights can be learned
-from the data.
-%
-First, \eval is widely used in the R core packages that come bundled with the
-language. Because of that, there is hardly any R code that would not use it.
-\eval is used to implement some basic functionality such as package loading.
-Putting the core packages aside, in CRAN, \eval is used by \PkgPackagesRatio of
-the available \CranPackages packages.
-%
-The code executed by R \eval has unusual expressive power. It can essentially
-modify anything anywhere, which is wary for both R developers and R compiler.
-The shapes of expressions passed to \eval are extremely diverse, ranging from
-simple variable lookups to complex blocks or the creation of new functions.
-Luckily, most sites only see a single shape. \eval unlashes its power and shows
-up its singularity compared to \eval in other languages with the diversity of
-environments in which it can evaluate its expressions. While most of the time,
-the \eval environment is the local environment, it is also often the global
-environment, a synthetic environment, or even a package environment. Luckily, as
-for the expressions passed to \eval, the environment passed to \eval for one
-given site is easily predictable with the first call to the site.
-
-The expressive power of \eval shines even more so in the case of side effects
-that can quite literally happen anywhere. The data, however, suggest that this
-is not the case. The majority of eval sites do not do any observable side
-effects, and when they do, it happens in a predictable environment.
-
-The origins of the expression passed to \eval, either constructed from language
-expressions, or stemming from reflection, or built by parsing strings.
-Reflection and constructed expressions dominate, which suggests the primary use
-of \eval is meta-programming.
-
-Similarly to \citep{ecoop11}, we started looking into \evals in R with the
-ambition to replace them with other features that are more amenable to static
-analysis. The data, unfortunately, suggests that replacing \eval is going to be
-a lengthy journey. Even if we disregard core packages as they are part of the
-language and the \eval use is stable, in CRAN \eval is used widely and in more
-sophisticated ways than was the case in JavaScript. There the vast majority of
-sites fitted an extremely simple categorization based on simple regular
-expressions. Not so in R. While some usage, especially the one which calls \eval
-with an expression parsed from string, can be categorized and often replaced by
-equivalent yet more disciplined and safer code, the case of \eval coming from
-reflection is much harder.
-
-Even though we reckon that \eval has more expressive power than macros, as it
-can access runtime information such as its call stack, adding a macro system to
-R to replace \eval for all meta-programming purposes would help static analysis
-as macros can be expanded at analysis-time.
+The \eval functions used widely and in varied ways in R. The function is an
+essential tool for language implementers. Our analysis observed that the base
+libraries heavily depend on \eval and any R program will end up calling it
+through the core of the language. Independently developed libraries often use
+\eval in subtle ways, mostly to perform macro-programming tasks. Our review of
+code written by less experienced users suggest that \eval is exceedingly rarely
+used. So, it is fair to conclude that majority of R programmers never have to
+write a call to \eval, but that the code they write would not run without it.
+
+\Eval is thus a challenge for automated program understanding. Program analysis
+and transformation tools or compilers must assume the worst when faced with code
+that calls this function. We have observed all sorts of side-effects with
+various degree of visibility.
+
+While our results are not encouraging in general, we have observed many cases
+where \eval is used in a disciplined and predictable ways. While in the general
+case \eval is hell, there are many cases where it is just a function. It does
+not appear that a general purpose replacement for \eval is possible, in future
+work we hope to focus on special cases and propose tools that target particular
+subsets of call sites with some common properties.
 

### 2021-08-11 paper/main.tex
sec 6 clean up

@@
 developers to this end. To understand \emph{why} developers use \eval, we
-manually inspected top 117 \eval call-sites which contribute to 90\% of the
-\eval calls. In this section, we give examples of the \eval usage from this
-real-world sample. The examples are often simplified to fit the space and
-increase clarity. For each example, we indicate the package and the function
-where it is located.
-
-\paragraph{Simplify Interfaces} A common use of \eval is to simplify interfaces by
-letting users pass string argument which can be parsed and evaluated in custom a
-environment. An example is the \code{rARS} function shown below. This function
-is defined in the \code{AdapSamp} package that implements sampling algorithms.
+manually inspected top 117 sites which contribute to 90\% of the calls in the
+corpus. This section gives some real-world examples, we simplify code for
+clarity.
+
+\paragraph{Extended Interfaces}
+Sometimes it is convenient to allow users to extend the behavior of a library
+function. \Eval allows clients to provide a simple text string that shall be
+executed in the context of the function. The example here allows users to
+provide formulas as strings, such as \k{"exp(-x^2/2)"}, the contract being that
+the text can refer to some variable \c x that will be set up by the called
+function. In this case one users could provide an anonymous function.
+
+\begin{figure}[h!]
+\begin{lstlisting}
+ function(n,formula) {
+    p <- function(x) eval(parse(text=formula))
+\end{lstlisting} \caption{AdapSamp::rARS}
+\end{figure}
+
+\paragraph{Code Generation}
+\Eval can be used to generate boiler plate code that would either be tedious to
+write by hand or that is only known after deployment. The example uses \eval to
+construct function objects that are parameterized with string \c{what}.
+This is used  automatically generate R bindings to native code.
 
@@
 \begin{lstlisting}
-rARS <- function(n, formula, min=-Inf, max=Inf, sp) {
-    ...
-    p <- function(x) { eval(parse(text=formula)) }
-    ...
-}
-\end{lstlisting}
-  \caption{AdapSamp::rARS}
+ function(methods,env) {
+   for(what in cppMethods)
+     methods[[what]] <- eval(substitute(function(...)CppObject$WHAT(...,name=what)),env)
+\end{lstlisting}
+  \caption{Rcpp::.makeCppMethods}
 \end{figure}
 
-The \code{rARS} function take a \code{formula} argument, an expression in
-\code{x} as a string, that represents the kernel of the target distribution.
-This is turned into the closure \code{function(x) \{ eval(parse(text=formula))
-  \}}. An alternative design would be to supply the closure itself as the
-argument, but passing \code{function(x) exp(-x^2/2)} is probably not as
-convenient for domain-experts as passing \code{"exp(-x^2/2)"}.
-
-\paragraph{Code Generation} The \c{Rcpp} package which bridges R with C++ is a
-popular user of this pattern. Its \code{.makeCppMethods} function, shown below,
-uses \eval to automatically generate R bindings to native code. The expression
-\code{function(...) .CppObject$WHAT(...), list(WHAT = as.name(what))} is
-evaluated with different values of \code{what} to generate wrapper functions
-which are assigned to the \code{methods} structure keyed by \code{what}.
+\paragraph{Extracting Varargs}
+A somewhat odd use of \eval is to obtain the source code of the arguments passed
+to a vararg parameter. In this example, the function iterates over the contents
+of the vararg until it finds a non-null value. \k{substitute(alist(...))} yields
+an expression that represents a call to \k{alist} but where the \code{...} are
+expended with the arguments supplied to the function. \Eval then returns a
+\code{list} with unevaluated argument expressions which are evaluated
+sequentially for their result.
 
@@
 \begin{lstlisting}
-.makeCppMethods <- function(methods, cppclassinfo, env) {
-    ...
-    for(what in cppMethods[! cppMethods %in% newMethods])
-        methods[[what]] <- eval(substitute(
-                  function(...) .CppObject$WHAT(...), list(WHAT = as.name(what))), env)
-    ...
-}
-\end{lstlisting}
-  \caption{Rcpp::.makeCppMethods}
-\end{figure}
-
-\paragraph{Extracting varargs} A surprising use of \eval is the
-extraction of unevaluated text of arguments passed to a \code{...} parameter.
-Consider the definition of the \code{NVL} function from the
-\code{statnet.common} package. This function returns the first argument that is
-non \code{NULL}.
-
-\begin{figure}[h]
-\begin{lstlisting}
-NVL <- function(...) {
-    for(e in eval(substitute(alist(...))))
+ function(...) {
+   for(e in eval(substitute(alist(...)))) {
      x <- eval(e, parent.frame())
@@
 
-The expression \code{eval(substitute(alist(...)))} is used to extract the
-unevaluated text of the arguments passed to \code{NVL}.
-\code{substitute(alist(...))} returns an expression representing a call to the
-\code{alist} function with \code{...} replaced by the arguments supplied to
-\code{NVL}. Evaluating this expression returns a \code{list} whose arguments are
-the unevaluated argument expressions which are evaluated sequentially for their
-result.
-
-\paragraph{Conditional Evaluation} Developers frequently use \eval to explicitly
-control the evaluation of code. Consider the \code{\%\|\|\%} operator from the
-\code{igraph} package shown below. This operator returns the result of
-evaluating the first argument (\code{lhs}) if it is not \code{NULL}, otherwise
-it returns the result of evaluating the second argument (\code{rhs}). This is
-similar to the \code{\|\|} operator which performs the logical OR of its
-arguments except it treats \code{NULL} as the \code{FALSE} value.
+\paragraph{Controlled Evaluation}
+Developer often want to control when and if expressions are evaluated. For
+example, a package that provides users with the ability to make assertions and,
+if they fail, produce legible error messages. The example here accepts a vararg
+with a list of expressions whose code is extracted using
+\code{eval(substitute(alist(...)))}. The expression are evaluated sequentially
+in the caller's environment. If an expression is false, an error message is
+generated.
 
@@
 \begin{lstlisting}
-`%||%` <- function (lhs, rhs) {
-    lres <- withVisible(eval(lhs, envir = parent.frame()))
-    if (is.null(lres$value)) {
-        eval(rhs, envir = parent.frame())
-    } else {
-        if (lres$visible) { lres$value}
-        else { invisible(lres$value) }
-   }
-}
-\end{lstlisting}
-  \caption{igraph::\%||\%}
-\end{figure}
-
-A similar idea is used by assertion packages such as \code{assertthat} which
-provide functions to check pre and post conditions and produce friendly error
-messages on failures. These conditions are written as expressions which are
-evaluated using \eval. Consider the \code{see\_if} function shown below. It
-accepts the conditions as arguments to the \code{...} parameter. The code of
-these conditions is extracted using \code{eval(substitute(alist(...)))}. The
-conditions are then evaluated sequentially by the expression
-\code{eval(assertion, env)}, inside a \code{tryCatch} block. If a condition
-evaluates to \code{FALSE}, the function returns immediately with an error
-message, leaving the remaining conditions unevaluated.
-
-\begin{figure}[h]
-\begin{lstlisting}
-see_if <- function(..., env = parent.frame(), msg = NULL) {
+ function(..., env=parent.frame(), msg=NULL) {
     asserts <- eval(substitute(alist(...)))
@@
         eval(assertion, env)
-      }, assertError = function(e) {
-        structure(FALSE, msg = e$message)
-      })
-    ...
-}
+      }, assertError=function(e) { structure(FALSE, msg=e$message) })
 \end{lstlisting}
@@
 
-\paragraph{Core function variants} Many packages use \eval to implement variations of
-the
-\code{base} R functions. For example, the \code{igraph} package provides the
-\code{do_call} function shown below. This is a variant of the \code{do.call}
-function from the \code{base} package. This function executes a function call by
-constructs it from a name and a list of arguments and calling \eval.
+\paragraph{Shadowing}
+%%% TODO ?????????????////
+Many packages use \eval to implement variations of the \code{base} R functions.
+For example, the \code{igraph} package provides the \code{do_call} function
+shown below. This is a variant of the \code{do.call} function from the
+\code{base} package. This function executes a function call by constructs it
+from a name and a list of arguments and calling \eval.
 
@@
 \begin{lstlisting}
-do_call <- function(f, ..., .args = list(), .env = parent.frame()) {
+ function(f,...,args=list(),env=parent.frame()) {
    f <- substitute(f)
-  call <- make_call(f, ..., .args)
-  eval(call, .env)
+   eval(make_call(f,...,args), env)
  }
@@
 
-\paragraph{Domain-Specific Languages} The combination of laziness, metaprogramming,
-first-class environments and dynamic evaluation makes R the perfect toolkit for
-building DSLs. An example of this is the \code{glue} package which provides
-functions for string interpolation. The code snippet below shows an example of
-\code{glue} function in action.
+\paragraph{Domain-Specific Language}
+A domain-specific language leverages R's grammar but redefines it semantics in
+some suitable way. The usage example illustrates string interpolation. The
+\k{glue} function extracts snippets of code enclosed between braces and
+evaluates them using \eval, and finally splices their results to construct the
+result.
 
@@
 
-The \code{glue} function extracts snippets of R code enclosed between \c{\{} and
-\c{\}}, evaluates them using \eval, and splices the result to construct the
-final string.
-
-Another example is the \code{data.table} package that defines a DSL for data
-frame manipulation using the subset operator \code{[}. This subset operation is
-defined as \code{DT[i, j, k]}, where expression \code{j} is used to calculate on
-a subset/reorder of rows from \code{i} grouped by \code{k}. This enables compact
-queries on data frames such as \code{flights[carrier == "AA", .N, by = origin]}
-which will compute the number of flights done by American Airlines (\code{"AA"})
-for each origin airport. At its core, \code{[} uses \code{substitute} to capture
-the operand expressions, transforms them, and evaluates them using \eval in an
-environment which has data frame columns available as bindings.
-
-The use of \eval with \code{substitute} is a general pattern that we often see
-in R packages. It is used to make the interfaces more flexible. For example the
-\code{subset}, which selects columns in a data frame, uses it to allow to specify
-the selected columns using either the column name or index. For example
-\code{subset(df, sel=c(1,B,x))} returns a subset of \code{df} with the first
-column, the column named \code{B} and whatever column is at index the variable
-\code{x} holds.
-
-It is implemented using \code{eval(substitute(sel), nl, parent.frame())}. The
-first argument returns the expression passed to \code{sel}. The second is a
-named list that contains the names of the columns from \code{df} and which will
-be used to evaluate the expression. The last parameter indicates which
-environment should be used to look up bindings not found in \c{nl}.
-
-\paragraph{Boilerplate Removal} The combination of \code{match.call} and \code{eval}
-is frequently used by modeling functions to avoid boilerplate code and ease
-maintenance. Consider the example shown below of function \code{survfit.formula}
-from the \code{survival} package.
+\paragraph{Boilerplate Removal}
+A variant of code generation is to let users write compact code that is expended
+via \eval and \code{match.call}. The example uses \code{match.call} to access
+the expression with which the function is called. The next three lines construct
+a call to \code{stats::model.frame} with argument text of parameters
+\code{'formula'}, \code{'data'}, and \code{'weights'}. Then, the call expression
+is evaluated in the parent environment. The use of \code{match.call}, avoids
+repeated use of \code{substitute} for extracting the argument expressions. This
+setup is also extensible as new parameter names can be easily added.
 
@@
 \begin{lstlisting}
-survfit.formula <- function(formula, data, weights, ...) {
+ function(formula, data, weights, ...) {
    Call <- match.call()
@@
    mf <- eval.parent(temp)
-}
 \end{lstlisting}
@@
 
-Here, \code{match.call} is used to access the expression with which
-\code{survfit.formula} is called. The next three lines construct a call to
-\code{stats::model.frame} with argument text of parameters \code{'formula'},
-\code{'data'}, and \code{'weights'}. Then, the call expression is evaluated in
-the parent environment. The use of \code{match.call}, avoids repeated use of
-\code{substitute} for extracting the argument expressions. This setup is also
-extensible as new parameter names can be easily added to the \code{match} call.
-
-\paragraph{Formula Resolution} Many packages use \code{eval} to evaluate formula
-expressions. Consider the definition of the \code{formula.pdMat} function from
-\code{nlme} package.
+\paragraph{Formula Resolution}
+Many packages use \eval to evaluate formulas. The example extracts the
+\code{"formula"} attribute attached to the \code{pdMat} object and evaluates it.
 
@@
 \begin{lstlisting}
-formula.pdMat <- function(x, asList, ...) eval(attr(x, "formula"))
+ function(x, asList, ...) eval(attr(x, "formula"))
 \end{lstlisting}
@@
 
-It extracts the \code{"formula"} attribute attached to the \code{pdMat} object
-and evaluates it.
-
-\paragraph{Code Obfuscation} Some packages use \eval to bypass the restrictions
-imposed by the \code{R CMD CHECK} tool. This tool enforces some well-formedness
-rules on a submitted package before accepting it for inclusion in CRAN. Some
-static checks ensure that the package code does not use certain restricted
-functions. If package authors cannot avoid using these functions, they use \eval
-to obfuscate the code, bypassing the checks performed by the tool. An example of
-this use is the \code{scalaEnsure} function from the \code{aibd} package.
+\paragraph{Obfuscation}
+Some packages use \eval to bypass the restrictions imposed by the \code{R CMD
+  CHECK} tool. This tool enforces some well-formedness rules on a submitted
+package before accepting it for inclusion in CRAN. Some static checks ensure
+that the package code does not use certain restricted functions. Package authors
+use \eval to obfuscate their code. The example mutates the variable \code{s} in
+the package environment. However, that is locked by default and
+\code{unlockBinding}, is a restricted function. To work around this, the call to
+unlock is done through \eval.
+
 \begin{figure}[h]
 \begin{lstlisting}
-scalaEnsure <- function() {
-    ...
+ function() {
    env <- parent.env(environment())
@@
    lockBinding("s", env)
-}
 \end{lstlisting}
@@
 \end{figure}
-
-This function mutates the variable \code{s} in the package environment. However,
-the package environment in R is locked by default, which prevents bindings from
-being mutated. Environment can be unlocked using \code{unlockBinding}, but its
-use is flagged by the well-formedness checks. To work around this limitation,
-the author passes the expression to unlock the package environment as a string
-to \code{parse} which is then evaluated by \code{eval}.
-
-\subsection{Shapes redux}
 
@@
 
-    Another example is the most often called \eval site in CRAN:
-    \lstinline|eval(`_inherit`, env)| defined in the \c{find\_super} function in
-    the \c{ggproto} object system. It looks up the variable \c{`\_inherit`} to
-    find the parent class. In the vast majority of cases, the variable is a
-    symbol, so the \eval seems superfluous. The \eval is however necessary so
-    one can specify parent classes using \c{pkg\_name::class\_name}, as the
-    \c{::} operator is a function call in R.
-
  \emph{Assignments \c{(<-)}} appear in the cases where a developer wants
@@
     scope: \lstinline|eval(parse(text=paste(default_args[[i]]), "<-", i))|.
-
-    There are also more sophisticated cases. The \c{plyr::allocate\_column}
-    function uses \eval for a deferred assignment to fill missing values in a
-    column. Package \c{overture} repeatedly evaluates R expressions capturing
-    any assignments that happen in it as samples of Markov chains.
 
@@
 
-    Call reflection refers to the combination of \eval with \c{match.call}, a
-    function that returns the current call with all its arguments captured as
-    expressions. This is used mostly in the following circumstances: (1) to
-    record a call for later use in a possibly different environment, (2) pass
-    most of the captured call arguments to another function, or (3) evaluate
-    arguments of the function in different environments. This is widely used in
-    relation to statistical modeling in combination with a model fitting
-    function or with \c{model.frame} and related functions. For example the
-    \c{survival::coxph} function\footnote{Function fitting a Cox proportional
-    hazards regression model.} contains:
-    %
-    \begin{lstlisting}[gobble=2]
-    Call <- match.call()
-    tform <- Call[c(1,indx)]  # only keep the arguments we wanted
-    tform[[1L]] <- quote(stats::model.frame)  # change the function called
-    mf <- eval(tform, parent.frame())
-    \end{lstlisting}
-
 \section{Discussion}

### 2021-08-11 paper/main.tex
Fix formatting of code blocks

@@
 on top of the core language. Most data types are sequences of primitive values.
-For instance, \c{c("Ha","bye")} evaluates to a vector of two strings. Constants
-such as \c{42} are vectors of length one. To enable equational reasoning, values
+For instance, \code{c("Ha","bye")} evaluates to a vector of two strings. Constants
+such as \code{42} are vectors of length one. To enable equational reasoning, values
 accessible through multiple aliases are copied when written to. Furthermore,
 values can be tagged by attributes; these are key-value pairs. For instance, the
-attribute \c{dim}-\c{c(2,2)} can be attached to the value at \c{x} by
-\c{attr(x,"dim")$\leftarrow$c(2,2)}. Adding this attribute turns \c x into a
-matrix. The \c{class} attribute gives a value a 'class' in the object-oriented
-sense. So, \c{class(x)<-"human"} sets the class of \c{x} to \c{human}; classes
+attribute \code{dim}-\code{c(2,2)} can be attached to the value at \code{x} by
+\code{attr(x,"dim") <- c(2,2)}. Adding this attribute turns \code{x} into a
+matrix. The \code{class} attribute gives a value a 'class' in the object-oriented
+sense. So, \code{class(x)<-"human"} sets the class of \code{x} to \code{human}; classes
 are used for method dispatch. Every linguistic construct is desugared to a
@@
 
-The \eval function in R takes three parameters: an \c{expression} to evaluate
-(\c{e}), an environment where to evaluate (\c{envir}), and an enclosure
-(\c{encl}) that is used to look up objects not found in \c{envir}.\footnote{Next
+The \eval function in R takes three parameters: an \code{expression} to evaluate
+(\code{e}), an environment where to evaluate (\code{envir}), and an enclosure
+(\code{encl}) that is used to look up objects not found in \code{envir}.\footnote{Next
 to the \eval function R offers three other variants:
   \begin{inparaenum}[1.]
-    \item \c{evalq} automatically quotes passed expression---a shorthand to
-      \c{eval(quote(...))},
-    \item \c{eval.parent(e, n)} specifies the evaluating environment in terms of
-      number of call frames (\c{n}) to go back---a shorthand to \c{eval(e,
+    \item \code{evalq} automatically quotes passed expression---a shorthand to
+      \code{eval(quote(...))},
+    \item \code{eval.parent(e, n)} specifies the evaluating environment in terms of
+      number of call frames (\code{n}) to go back---a shorthand to \code{eval(e,
         parent.frame(n))}, and
-    \item \c{local} evaluates \c{e} in a fresh environment---a shorthand to
-      \c{evalq(e, new.env())}.
+    \item \code{local} evaluates \code{e} in a fresh environment---a shorthand to
+      \code{evalq(e, new.env())}.
   \end{inparaenum} }
@@
 \begin{minipage}{.49\textwidth}
-\begin{lstlisting}[caption={Examples of calls producing expression \c{a+b}}]
+\begin{lstlisting}[caption={Examples of calls producing expression \code{a+b}}]
   parse(text="a+b")
@@
 
- The call to \c{substitute(x)} extracts the unevaluated expression from
-the promise \c{x}. The function \c{match.call} returns an expression
+ The call to \code{substitute(x)} extracts the unevaluated expression from
+the promise \code{x}. The function \code{match.call} returns an expression
 representing the current call.
@@
 frame (using element names or column names for variable resolution) or an
-integer \c{n} (in which case it would use the \c{n.} call frame). If no
+integer \code{n} (in which case it would use the \code{n.} call frame). If no
 environment is specified, then the default is the environment where the call to
-\eval was made (as returned by \c{parent.frame}).
+\eval was made (as returned by \code{parent.frame}).
 %
 Environments nest, each has a parent. When creating a new environment with
-\c{new.env}, the parent is the current environment. Environment chains can be
-traversed with \c{parent.env}, until \c{emptyenv} is reached. The top-level
-environment is \c{.GlobalEnv}, it has parents that represent the packages that
+\code{new.env}, the parent is the current environment. Environment chains can be
+traversed with \code{parent.env}, until \code{emptyenv} is reached. The top-level
+environment is \code{.GlobalEnv}, it has parents that represent the packages that
 have been loaded. One can also directly read, modify or create new bindings,
@@
 rich reflective interface. The authors found that reflective methods are used in less
-than 2\% of methods. The most common reflective method is \c{perform:}; it send
+than 2\% of methods. The most common reflective method is \code{perform:}; it send
 a message that is specified by a string. These features are primarily used in the
@@
 and 2.2 million dynamic calls. The 15 most frequent call sites account for 88\%
-of those. The \c{match.arg} function is the highest used one with 54\% of all
+of those. The \code{match.arg} function is the highest used one with 54\% of all
 calls. In the other call sites, they saw two use cases. The most common is the
-evaluation of the source code of a promise retrieved by \c{substitute} in a new
-environment, \eg as done in the \c{with} function. The other use case is the
+evaluation of the source code of a promise retrieved by \code{substitute} in a new
+environment, \eg as done in the \code{with} function. The other use case is the
 invocation of a function whose name or arguments are determined dynamically. For
-this purpose, R provides \c{do.call} and thus \eval is overkill. Our corpus is
+this purpose, R provides \code{do.call} and thus \eval is overkill. Our corpus is
 larger, with \CranPackages packages, \packageTriggeredpkgs packages with at
@@
 \item \emph{Extract.} The \genthat tool extracts all runnable code snippets and
-  turns each of these into a self-standing program~\cite{issta18}, \c{knitr}
+  turns each of these into a self-standing program~\cite{issta18}, \code{knitr}
   extracts code from notebooks. Each extracted script is instrumented with calls
@@
   RMarkdown notebooks process the summarized data to generate all graphs (PDF
-  from \c{ggplot2}) and numbers (exported as \LaTeX macros) appearing in the
+  from \code{ggplot2}) and numbers (exported as \LaTeX macros) appearing in the
   paper.
@@
 variants of \eval and a few additional functions to locate the origin of \eval
-arguments (it taints the results of calls to \c{parse} and \c{match.call}). It
+arguments (it taints the results of calls to \code{parse} and \code{match.call}). It
 captures dynamic code loading, as well as variable definition and assignment,
@@
 \begin{tabular}{@{\,}r|rrrr}\toprule
-  &\bf \eval & \bf \c{\bf evalq} & \c{\bf eval} & \c{\bf local}\\[-1.5mm]
-           & & & \c{\bf .parent} &\\\midrule
+  &\eval & \code{evalq} & \code{eval} & \code{local}\\[-1.5mm]
+           & & & \code{.parent} &\\\midrule
 \small Static sites &\packageStaticeval&\packageStaticevalq&\packageStaticevalparent&\packageStaticlocal \\
@@
 (\emph{exercised}) and the last gives calls (\emph{invocations}). Most sites and
-calls go to \eval itself, \c{eval.parent} is rare, and both \c{evalq} and
-\c{local} are barely used at all.
+calls go to \eval itself, \code{eval.parent} is rare, and both \code{evalq} and
+\code{local} are barely used at all.
 
@@
 \eval returns it unchanged. Expressions account for \packageCodepercent of
-arguments. More specifically, \packageSymbolpercent are \c{symbol}s (variables
-such as \c{x}), \packageLanguagepercent are \c{language} objects (expressions
-representing a function calls), and \packageExpressionpercent are \c{expression}
-objects (lists of expressions). Further inspection reveals that \c{\_inherit}
+arguments. More specifically, \packageSymbolpercent are \code{symbol}s (variables
+such as \code{x}), \packageLanguagepercent are \code{language} objects (expressions
+representing a function calls), and \packageExpressionpercent are \code{expression}
+objects (lists of expressions). Further inspection reveals that \code{\_inherit}
 accounts for \packageGgplotsymbolpercent of symbols and comes from one site in
-\c{ggplot2}.\footnote{Each object has a field model inheritance, this comes from
-\c{ggproto}, one of the many object-oriented systems for R.}
+\code{ggplot2}.\footnote{Each object has a field model inheritance, this comes from
+\code{ggproto}, one of the many object-oriented systems for R.}
 
@@
 size of the arguments in terms of number of nodes. For example,
-\c{expression(x+1)} has a size of 3. The large number of symbols skew the median
+\code{expression(x+1)} has a size of 3. The large number of symbols skew the median
 to \packageMedianszeval but the average size is \packageAvgszeval nodes. The
@@
   \bf $min(e)$& \bf \#sites & \bf \%sites & \bf \#packages & \bf \#operations & \bf \%envir & \bf example\\\midrule
-\c X&\packageMinimizedcallsitesa &\packageMinimizedpropsitesa &\packageMinimizedpackagea &\packageMinimizedmedianoperationsaRnd &\packageMinimizedpercentparentframesa & \c{y+1}\\
-\c{F(F(X))} & \packageMinimizedcallsitesb  & \packageMinimizedpropsitesb & \packageMinimizedpackageb  & \packageMinimizedmedianoperationsbRnd & \packageMinimizedpercentparentframesb & \c{gbov( mean(x), a-1)}\\
-\c{V}&\packageMinimizedcallsitesc &\packageMinimizedpropsitesc &\packageMinimizedpackagec &\packageMinimizedmedianoperationscRnd &\packageMinimizedpercentparentframesc& \c{c(42,21,0)}\\
-\c{F(X)}& \packageMinimizedcallsitesd & \packageMinimizedpropsitesd & \packageMinimizedpackaged & \packageMinimizedmedianoperationsdRnd & \packageMinimizedpercentparentframesd & \c{seq\_len(iters)} \\
-\c{\$} & \packageMinimizedcallsitese & \packageMinimizedpropsitese & \packageMinimizedpackagee & \packageMinimizedmedianoperationseRnd & \packageMinimizedpercentparentframese & \c{DF\$B}\\
-\c{model.frame}& \packageMinimizedcallsitesf & \packageMinimizedpropsitesf & \packageMinimizedpackagef & \packageMinimizedmedianoperationsfRnd & \packageMinimizedpercentparentframesf &  \c{model.frame(formula = Z $\sim$ U)}   \\
-\c{F()}& \packageMinimizedcallsitesg & \packageMinimizedpropsitesg & \packageMinimizedpackageg & \packageMinimizedmedianoperationsgRnd & \packageMinimizedpercentparentframesg & \c{rgamma(3, 2, n = 10L)} \\
-\c{FUN} & \packageMinimizedcallsitesh & \packageMinimizedpropsitesh & \packageMinimizedpackageh & \packageMinimizedmedianoperationshRnd & \packageMinimizedpercentparentframesh & \c{function(x, y) x + 3 * y} \\
-\c{<-} & \packageMinimizedcallsitesi  & \packageMinimizedpropsitesi & \packageMinimizedpackagei & \packageMinimizedmedianoperationsiRnd & \packageMinimizedpercentparentframesi & \c{x[1, 2:3, 2:3] <- value}\\
-\c{BLOCK} & \packageMinimizedcallsitesj & \packageMinimizedpropsitesj & \packageMinimizedpackagej & \packageMinimizedmedianoperationsjRnd & \packageMinimizedpercentparentframesj & \c{\{ write.csv(iris, tf) ; file.size(tf) \}} \\\bottomrule
+\code{X}&\packageMinimizedcallsitesa &\packageMinimizedpropsitesa &\packageMinimizedpackagea &\packageMinimizedmedianoperationsaRnd &\packageMinimizedpercentparentframesa & \code{y+1}\\
+\code{F(F(X))} & \packageMinimizedcallsitesb  & \packageMinimizedpropsitesb & \packageMinimizedpackageb  & \packageMinimizedmedianoperationsbRnd & \packageMinimizedpercentparentframesb & \code{gbov( mean(x), a-1)}\\
+\code{V}&\packageMinimizedcallsitesc &\packageMinimizedpropsitesc &\packageMinimizedpackagec &\packageMinimizedmedianoperationscRnd &\packageMinimizedpercentparentframesc& \code{c(42,21,0)}\\
+\code{F(X)}& \packageMinimizedcallsitesd & \packageMinimizedpropsitesd & \packageMinimizedpackaged & \packageMinimizedmedianoperationsdRnd & \packageMinimizedpercentparentframesd & \code{seq\_len(iters)} \\
+\code{\$} & \packageMinimizedcallsitese & \packageMinimizedpropsitese & \packageMinimizedpackagee & \packageMinimizedmedianoperationseRnd & \packageMinimizedpercentparentframese & \code{DF\$B}\\
+\code{model.frame}& \packageMinimizedcallsitesf & \packageMinimizedpropsitesf & \packageMinimizedpackagef & \packageMinimizedmedianoperationsfRnd & \packageMinimizedpercentparentframesf &  \code{model.frame(formula = Z ~ U)}   \\
+\code{F()}& \packageMinimizedcallsitesg & \packageMinimizedpropsitesg & \packageMinimizedpackageg & \packageMinimizedmedianoperationsgRnd & \packageMinimizedpercentparentframesg & \code{rgamma(3, 2, n = 10L)} \\
+\code{FUN} & \packageMinimizedcallsitesh & \packageMinimizedpropsitesh & \packageMinimizedpackageh & \packageMinimizedmedianoperationshRnd & \packageMinimizedpercentparentframesh & \code{function(x, y) x + 3 * y} \\
+\code{<-} & \packageMinimizedcallsitesi  & \packageMinimizedpropsitesi & \packageMinimizedpackagei & \packageMinimizedmedianoperationsiRnd & \packageMinimizedpercentparentframesi & \code{x[1, 2:3, 2:3] <- value}\\
+\code{BLOCK} & \packageMinimizedcallsitesj & \packageMinimizedpropsitesj & \packageMinimizedpackagej & \packageMinimizedmedianoperationsjRnd & \packageMinimizedpercentparentframesj & \code{\{write.csv(iris,tf);file.size(tf)\}} \\\bottomrule
 \end{tabular}
@@
 function $min(e)$ which for a given expression $e$ returns a normal form. Normal
-forms use \c V to stand in for values occurring in expression, \c X stands in
-for variables, and \c F stands for functions. The function performs constant
+forms use \code{V} to stand in for values occurring in expression, \code{X} stands in
+for variables, and \code{F} stands for functions. The function performs constant
 folding of arithmetic and string expressions over base operators, for instance
@@
 has $min(\c{x+y})$ becomes $\c{X}$, this motivated by the fact that addition is
-not an ``interesting'' operation, and that \c X stands in for any number of
+not an ``interesting'' operation, and that \code{X} stands in for any number of
 variable lookups. Function absorption simplifies nested functions to keep an
@@
 simplifications, the full list can be found in our artifact. It is worth
-mentioning that these simplifications are heuristics, in R \c{1+1} is not
-necessary \c 2. The addition operator can be redefined, but it typically isn't
+mentioning that these simplifications are heuristics, in R \code{1+1} is not
+necessary \code{2}. The addition operator can be redefined, but it typically isn't
 or at least not in a way that would invalidate arithmetic.
@@
 \medskip\EE{$min(e)=\c{X}$}\\[-2mm]\small Variable lookups are the most common shape,
-\packageNbSymbolVarSitePercent are simple variable names, e.g. \c x. The shapes
-subsumes \c{V} and built-in arithmetic operations, so a mixture of arithmetics
-and variables normalize to \c X. A lookup is one step of execution, if a promise
+\packageNbSymbolVarSitePercent are simple variable names, e.g. \code{x}. The shapes
+subsumes \code{V} and built-in arithmetic operations, so a mixture of arithmetics
+and variables normalize to \code{X}. A lookup is one step of execution, if a promise
 is returned an arbitrary number of additional steps may be needed; the median is
@@
 \\
-\medskip\EE{$min(e)=\c{\$}$}\\[-2mm]\small This shape extends \c X to include lookup
-with the dollar operator, \eg~\c{x\$f}, and vector indexing, \eg~\c{x[42]} and
-\c{x[[24]]}. This takes a few more step of evaluation on average, and is
+\medskip\EE{$min(e)=\c{\$}$}\\[-2mm]\small This shape extends \code{X} to include lookup
+with the dollar operator, \eg~\code{x\$f}, and vector indexing, \eg~\code{x[42]} and
+\code{x[[24]]}. This takes a few more step of evaluation on average, and is
 typically used in a function environment.
 \\
-\medskip\EE{$min(e)=$~\c{<-}}\\[-2mm]\small This includes both direct assignment {\tt
-  <-} and parent environment assignment {\tt <\,\!<-} as well as the \c{assign}
-function. It subsumes \c{\$}. Assignments represent the most obvious source
+\medskip\EE{$min(e)=$~\code{<-}}\\[-2mm]\small This includes both direct assignment {\tt
+  <-} and parent environment assignment {\tt <\,\!<-} as well as the \code{assign}
+function. It subsumes \code{\$}. Assignments represent the most obvious source
 of side effects in \eval. Most cases originate from trivial code generation
@@
   \medskip\EE{$min(e)=\c{FUN}$}\\[-2mm]\small Function definitions without any
-  computation, \eg~\c{function(x) x+1}. In addition to \c{FUN},
+  computation, \eg~\code{function(x) x+1}. In addition to \code{FUN},
   \packageGeneralizedFunctionDefinitionSitesPercent of sites have function
@@
 different from 0-argument closure. There is a number of use cases: unit testing
-frameworks (\eg \c{testthat}, \c{testit}), code benchmarking (\eg
-\c{rbenchmark}, \c{microbenchmark}), running code in parallel (\eg \c{foreach},
-\c{doParallel}), or deferring code execution (\eg \c{withr}).
+frameworks (\eg \code{testthat}, \code{testit}), code benchmarking (\eg
+\code{rbenchmark}, \code{microbenchmark}), running code in parallel (\eg \code{foreach},
+\code{doParallel}), or deferring code execution (\eg \code{withr}).
 
 \\
-\medskip\EE{$min(e)=\c{model.frame}$}\\[-2mm]\small The \c{model.frame} function
+\medskip\EE{$min(e)=\c{model.frame}$}\\[-2mm]\small The \code{model.frame} function
 returns a dataframe resulting from fitting the model described in a formula.
-This shape subsumes \c{F(F(X))}, \c{FUN}, and assignments. It is the single most
+This shape subsumes \code{F(F(X))}, \code{FUN}, and assignments. It is the single most
 popular function invoked from \eval. Each call does a median of 2K instructions.
@@
 more than eight shapes. An example of those is the pipe operator of the
-\c{magrittr} package which is used to compose functions, \eg instead of
-\c{f(g(h(x),y),z)} one can write \c{h(x) \%>\% g(y) \%>\% f(z)}. As there many
+\code{magrittr} package which is used to compose functions, \eg instead of
+\code{f(g(h(x),y),z)} one can write \code{h(x) \%>\% g(y) \%>\% f(z)}. As there many
 different patterns of use, there are also many different shapes. In comparison
@@
 Contrary to JavaScript, in R the environment of \eval can be specifies by its
-\c{env} argument. This gives users control over what is visible to the
+\code{env} argument. This gives users control over what is visible to the
 computation started by \eval and the potential reach of its side-effects. We
@@
 \begin{itemize}[---]
-\item {\it Reflection:} use of \c{match.call} to reflectively capture the
+\item {\it Reflection:} use of \code{match.call} to reflectively capture the
   expression that invoked this function, arguments are promises, and
-  \c{substitute} is used to retrieve their source.
-\item {\it String:} create from strings by invoking \c{parse},
-  \c{str2expression}, or \c{str2lang}.
-\item {\it Constructed:} invoked \c{quote} or \c{expression}, or by built with
-  \c{call} or \c{as.call}.
-\item {\it Environment: } created with \c{as.name} or \c{as.symbol}, typically
+  \code{substitute} is used to retrieve their source.
+\item {\it String:} create from strings by invoking \code{parse},
+  \code{str2expression}, or \code{str2lang}.
+\item {\it Constructed:} invoked \code{quote} or \code{expression}, or by built with
+  \code{call} or \code{as.call}.
+\item {\it Environment: } created with \code{as.name} or \code{as.symbol}, typically
   to read a non-local environment.
@@
 unit testing frameworks as they use \eval to run all their tests and thus
-dominate the side-effect data.\footnote{The corpus has \c{RUnit, testthat,
-  tinytest} and \c{unitizer} testing frameworks.}
+dominate the side-effect data.\footnote{The corpus has \code{RUnit, testthat,
+  tinytest} and \code{unitizer} testing frameworks.}
 
@@
 virtual machine implementation and not user code. For example, the
-\c{.Random.seed} variable is saved and restored from and to the global
+\code{.Random.seed} variable is saved and restored from and to the global
 environment every time a statistical routine is called. Filtering accidental
@@
 responsible for 90\% of side effects. Half of those come from just three
-functions: \c{plyr::allocate\_column} (allocates space for a new data frame
-column), \c{withr::execute\_handlers} (executes deferred expressions), and
-\c{foreach::doSEQ} (executes an expression on each element in a collection,
+functions: \code{plyr::allocate\_column} (allocates space for a new data frame
+column), \code{withr::execute\_handlers} (executes deferred expressions), and
+\code{foreach::doSEQ} (executes an expression on each element in a collection,
 possibly in parallel). Most of the sites (\SESitesInEnvirRatio) update
-the environment specified by the \c{env} parameter.
+the environment specified by the \code{env} parameter.
 
@@
 happens, but the vast majority comes from a single site
-(\c{withr::execute\_handlers}). This makes sense as it is used to defer
+(\code{withr::execute\_handlers}). This makes sense as it is used to defer
 evaluation of an expression to after the function exit and thus used for clean
-up. It is used almost exclusively by the \c{tidyselect} package for removing the
+up. It is used almost exclusively by the \code{tidyselect} package for removing the
 reference to the current quosure environment while interpreting a data frame
@@
 environment. We have manually inspected a few of these sites, but except for
-manual injection of parameters into a \c{model.frame} execution environment, we
+manual injection of parameters into a \code{model.frame} execution environment, we
 did not find a common use case.
@@
 
-Out of the minimized expressions, it is \c{BLOCK}, \c{<-} and \c{F(F(X))} that
-contribute to the vast majority of side effects. Concretely, 92\% of \c{BLOCK},
-87\% of \c{<-} and 15\% of \c{F(F(X))} expressions do side effects. In the case
-of \c{BLOCK}, most of them happen in the environment passed to the \eval call
-(default is the parent environment). For \c{<-} it is 15\%. However, this form
-is dominated by the \c{plyr::allocate\_column} function that side-effects in
+Out of the minimized expressions, it is \code{BLOCK}, \code{<-} and \code{F(F(X))} that
+contribute to the vast majority of side effects. Concretely, 92\% of \code{BLOCK},
+87\% of \code{<-} and 15\% of \code{F(F(X))} expressions do side effects. In the case
+of \code{BLOCK}, most of them happen in the environment passed to the \eval call
+(default is the parent environment). For \code{<-} it is 15\%. However, this form
+is dominated by the \code{plyr::allocate\_column} function that side-effects in
 the environment containing the data frame to which it allocates new column.
-Without this call, \c{<-} is similarly predictable with 96\% of side effects
-happening in the passed environment. For \c{F(F(X))}, it is less clear as 58\%
+Without this call, \code{<-} is similarly predictable with 96\% of side effects
+happening in the passed environment. For \code{F(F(X))}, it is less clear as 58\%
 of side effects happen in a different environment to the one that was passed to
@@
 The R language was intended to be extensible. The combination of lazy
-evaluation, \c{substitute} and \c{eval} are some of the tools given to
+evaluation, \code{substitute} and \code{eval} are some of the tools given to
 developers to this end. To understand \emph{why} developers use \eval, we
@@
 provide formulas as strings, such as \k{"exp(-x^2/2)"}, the contract being that
-the text can refer to some variable \c x that will be set up by the called
+the text can refer to some variable \code{x} that will be set up by the called
 function. In this case one users could provide an anonymous function.
@@
 write by hand or that is only known after deployment. The example uses \eval to
-construct function objects that are parameterized with string \c{what}.
+construct function objects that are parameterized with string \code{what}.
 This is used  automatically generate R bindings to native code.
@@
 
-\emph{Variable lookup and values, and indexing \c{(X, V, \$)}}
+\emph{Variable lookup and values, and indexing \code{(X, V, \$)}}
     constitute one of the most common use case for \eval. Let's consider the
-    \c{subset} function from the R base library, which allows, among others, to
-    select variables in a data frame. For example, given a data frame \c{df}
-    with three columns \c{A,B,C}, \c{subset(df, sel=A)} returns a subset of
-    \c{df} with just the \c{A} column. The same would happen with \c{subset(df,
+    \code{subset} function from the R base library, which allows, among others, to
+    select variables in a data frame. For example, given a data frame \code{df}
+    with three columns \code{A,B,C}, \code{subset(df, sel=A)} returns a subset of
+    \code{df} with just the \code{A} column. The same would happen with \code{subset(df,
       sel=1)}. However, thanks to \eval it can also be also called with
-    \c{subset(df, sel=c(1,B,x))} returning a subset of the first column, column
-    \c{B} and whatever index the variable \c{x} holds. It is implemented using:
+    \code{subset(df, sel=c(1,B,x))} returning a subset of the first column, column
+    \code{B} and whatever index the variable \code{x} holds. It is implemented using:
     \lstinline|eval(substitute(sel), nl, parent.frame())| The first argument
-    returns the expression passed to \c{sel}, the second is the data frame in
+    returns the expression passed to \code{sel}, the second is the data frame in
     which this expression will be evaluated, and the last indicates which
-    environment should be used to look up bindings not found in \c{nl}.
-
- \emph{Assignments \c{(<-)}} appear in the cases where a developer wants
+    environment should be used to look up bindings not found in \code{nl}.
+
+ \emph{Assignments \code{(<-)}} appear in the cases where a developer wants
     to create or update a binding in a specific environment where either the
@@
     trivial code generation where the assign expression is assembled using
-    \c{parse} or \c{substitute}, often in a loop. For example, the
-    \c{PopED::get\_rse} uses it to assign a number of variables in the current
+    \code{parse} or \code{substitute}, often in a loop. For example, the
+    \code{PopED::get\_rse} uses it to assign a number of variables in the current
     scope: \lstinline|eval(parse(text=paste(default_args[[i]]), "<-", i))|.
 
- \emph{Function calls \c{(F(), F(X), F(F(X)))}} represent two very
+ \emph{Function calls \code{(F(), F(X), F(F(X)))}} represent two very
     frequent patterns: \emph{code generation} and \emph{call reflection}. There
@@
     One occurring usage (present also in the Kaggle scripts) is to generate
-    ad-hoc calls to libraries such as \c{ggplot2} when one needs to
+    ad-hoc calls to libraries such as \code{ggplot2} when one needs to
     parameterize data or aesthetics selection from a variable. For example
     \lstinline|ggplot(df, aes(x=A, y=B))| maps X and Y axes in the plot to
-    \c{A} and \c{B} columns in \c{df}. It might not be immediately
+    \code{A} and \code{B} columns in \code{df}. It might not be immediately
     clear\footnote{\cf
@@
     boilerplate code; simple and repetitive code can easily be replaced with
-    judicious use of \eval. For example, the \c{data.table} package uses \eval
-    to call the \c{options} function with named arguments taken from a vector
+    judicious use of \eval. For example, the \code{data.table} package uses \eval
+    to call the \code{options} function with named arguments taken from a vector
     of strings:
@@
 The cases where \eval appears to be overkill include simple expression shapes
-such as \c X, \c V, \$, \c{<-}, and \c{FUN}. Variable lookup in a remote
+such as \code{X}, \code{V}, \code{\$}, \code{<-}, and \code{FUN}. Variable lookup in a remote
 environment can be performed by the builtin \code{get} function, a similar case
 can be made for assignment and the \code{assign} function. There is a dedicated
-\c{do.call} function to build function calls. Unnecessary uses of \eval
+\code{do.call} function to build function calls. Unnecessary uses of \eval
 often can be uncovered by manual analysis.
@@
 current functions. \Eval is overkill for this code, a semantically equivalent
-function body uses \c{get} to perform the lookup and then calls directly.
+function body uses \code{get} to perform the lookup and then calls directly.
 
@@
 
-Another example, from \c{coin::copyslots}, constructs an expression that access
+Another example, from \code{coin::copyslots}, constructs an expression that access
 fields of an object and copy them. This can be replaced by symbolic field reads
@@
 As a last example, \code{configr::config.funs.par} construct three expressions
-that are then fed to \eval. Semantically equivalent code would use \c{get}
-to lookup \c{fun}, and then directly execute the three expressions.
+that are then fed to \eval. Semantically equivalent code would use \code{get}
+to lookup \code{fun}, and then directly execute the three expressions.
 
@@
 libraries include their own implementation calls
-\c{rlang::eval\_tidy}~\cite{tidyverse}. This departs from \eval in two ways.
+\code{rlang::eval\_tidy}~\cite{tidyverse}. This departs from \eval in two ways.
 First, it supports evaluation of quosures, custom objects used by for
-metaprogramming that bundle expressions with an environment. \c{eval\_tidy}
+metaprogramming that bundle expressions with an environment. \code{eval\_tidy}
 accepts a data mask argument, a set of bindings such as a dataframe or list,
 which takes precedence over the environment. Effects such as assignments happen
-in the data mask. Expressions such as \c{return()} do not work since the data
+in the data mask. Expressions such as \code{return()} do not work since the data
 mask does not correspond to a frame on the call stack.

### 2021-08-11 paper/main.tex
Replace \code with \k

--- 2021-08-11 paper/main.tex
@@
 \newcommand{\todo}[1]{\authorcomment{red}{TODO}{#1}}
-\newcommand{\code}[1]{\lstinline |#1|\xspace}
 \renewcommand{\k}[1]{\lstinline |#1|\xspace}
@@
 on top of the core language. Most data types are sequences of primitive values.
-For instance, \c{c("Ha","bye")} evaluates to a vector of two strings. Constants
-such as \c{42} are vectors of length one. To enable equational reasoning, values
+For instance, \k{c("Ha","bye")} evaluates to a vector of two strings. Constants
+such as \k{42} are vectors of length one. To enable equational reasoning, values
 accessible through multiple aliases are copied when written to. Furthermore,
 values can be tagged by attributes; these are key-value pairs. For instance, the
-attribute \c{dim}-\c{c(2,2)} can be attached to the value at \c{x} by
-\c{attr(x,"dim")$\leftarrow$c(2,2)}. Adding this attribute turns \c x into a
-matrix. The \c{class} attribute gives a value a 'class' in the object-oriented
-sense. So, \c{class(x)<-"human"} sets the class of \c{x} to \c{human}; classes
+attribute \k{dim}-\k{c(2,2)} can be attached to the value at \k{x} by
+\k{attr(x,"dim") <- c(2,2)}. Adding this attribute turns \k{x} into a
+matrix. The \k{class} attribute gives a value a 'class' in the object-oriented
+sense. So, \k{class(x)<-"human"} sets the class of \k{x} to \k{human}; classes
 are used for method dispatch. Every linguistic construct is desugared to a
@@
 
-The \eval function in R takes three parameters: an \c{expression} to evaluate
-(\c{e}), an environment where to evaluate (\c{envir}), and an enclosure
-(\c{encl}) that is used to look up objects not found in \c{envir}.\footnote{Next
+The \eval function in R takes three parameters: an \k{expression} to evaluate
+(\k{e}), an environment where to evaluate (\k{envir}), and an enclosure
+(\k{encl}) that is used to look up objects not found in \k{envir}.\footnote{Next
 to the \eval function R offers three other variants:
   \begin{inparaenum}[1.]
-    \item \c{evalq} automatically quotes passed expression---a shorthand to
-      \c{eval(quote(...))},
-    \item \c{eval.parent(e, n)} specifies the evaluating environment in terms of
-      number of call frames (\c{n}) to go back---a shorthand to \c{eval(e,
+    \item \k{evalq} automatically quotes passed expression---a shorthand to
+      \k{eval(quote(...))},
+    \item \k{eval.parent(e, n)} specifies the evaluating environment in terms of
+      number of call frames (\k{n}) to go back---a shorthand to \k{eval(e,
         parent.frame(n))}, and
-    \item \c{local} evaluates \c{e} in a fresh environment---a shorthand to
-      \c{evalq(e, new.env())}.
+    \item \k{local} evaluates \k{e} in a fresh environment---a shorthand to
+      \k{evalq(e, new.env())}.
   \end{inparaenum} }
@@
 \begin{minipage}{.49\textwidth}
-\begin{lstlisting}[caption={Examples of calls producing expression \c{a+b}}]
+\begin{lstlisting}[caption={Examples of calls producing expression \k{a+b}}]
   parse(text="a+b")
@@
 
- The call to \c{substitute(x)} extracts the unevaluated expression from
-the promise \c{x}. The function \c{match.call} returns an expression
+ The call to \k{substitute(x)} extracts the unevaluated expression from
+the promise \k{x}. The function \k{match.call} returns an expression
 representing the current call.
@@
 frame (using element names or column names for variable resolution) or an
-integer \c{n} (in which case it would use the \c{n.} call frame). If no
+integer \k{n} (in which case it would use the \k{n.} call frame). If no
 environment is specified, then the default is the environment where the call to
-\eval was made (as returned by \c{parent.frame}).
+\eval was made (as returned by \k{parent.frame}).
 %
 Environments nest, each has a parent. When creating a new environment with
-\c{new.env}, the parent is the current environment. Environment chains can be
-traversed with \c{parent.env}, until \c{emptyenv} is reached. The top-level
-environment is \c{.GlobalEnv}, it has parents that represent the packages that
+\k{new.env}, the parent is the current environment. Environment chains can be
+traversed with \k{parent.env}, until \k{emptyenv} is reached. The top-level
+environment is \k{.GlobalEnv}, it has parents that represent the packages that
 have been loaded. One can also directly read, modify or create new bindings,
@@
 rich reflective interface. The authors found that reflective methods are used in less
-than 2\% of methods. The most common reflective method is \c{perform:}; it send
+than 2\% of methods. The most common reflective method is \k{perform:}; it send
 a message that is specified by a string. These features are primarily used in the
@@
 and 2.2 million dynamic calls. The 15 most frequent call sites account for 88\%
-of those. The \c{match.arg} function is the highest used one with 54\% of all
+of those. The \k{match.arg} function is the highest used one with 54\% of all
 calls. In the other call sites, they saw two use cases. The most common is the
-evaluation of the source code of a promise retrieved by \c{substitute} in a new
-environment, \eg as done in the \c{with} function. The other use case is the
+evaluation of the source code of a promise retrieved by \k{substitute} in a new
+environment, \eg as done in the \k{with} function. The other use case is the
 invocation of a function whose name or arguments are determined dynamically. For
-this purpose, R provides \c{do.call} and thus \eval is overkill. Our corpus is
+this purpose, R provides \k{do.call} and thus \eval is overkill. Our corpus is
 larger, with \CranPackages packages, \packageTriggeredpkgs packages with at
@@
 \item \emph{Extract.} The \genthat tool extracts all runnable code snippets and
-  turns each of these into a self-standing program~\cite{issta18}, \c{knitr}
+  turns each of these into a self-standing program~\cite{issta18}, \k{knitr}
   extracts code from notebooks. Each extracted script is instrumented with calls
@@
   RMarkdown notebooks process the summarized data to generate all graphs (PDF
-  from \c{ggplot2}) and numbers (exported as \LaTeX macros) appearing in the
+  from \k{ggplot2}) and numbers (exported as \LaTeX macros) appearing in the
   paper.
@@
 variants of \eval and a few additional functions to locate the origin of \eval
-arguments (it taints the results of calls to \c{parse} and \c{match.call}). It
+arguments (it taints the results of calls to \k{parse} and \k{match.call}). It
 captures dynamic code loading, as well as variable definition and assignment,
@@
 \begin{tabular}{@{\,}r|rrrr}\toprule
-  &\bf \eval & \bf \c{\bf evalq} & \c{\bf eval} & \c{\bf local}\\[-1.5mm]
-           & & & \c{\bf .parent} &\\\midrule
+  &\eval & \k{evalq} & \k{eval} & \k{local}\\[-1.5mm]
+           & & & \k{.parent} &\\\midrule
 \small Static sites &\packageStaticeval&\packageStaticevalq&\packageStaticevalparent&\packageStaticlocal \\
@@
 (\emph{exercised}) and the last gives calls (\emph{invocations}). Most sites and
-calls go to \eval itself, \c{eval.parent} is rare, and both \c{evalq} and
-\c{local} are barely used at all.
+calls go to \eval itself, \k{eval.parent} is rare, and both \k{evalq} and
+\k{local} are barely used at all.
 
@@
 \eval returns it unchanged. Expressions account for \packageCodepercent of
-arguments. More specifically, \packageSymbolpercent are \c{symbol}s (variables
-such as \c{x}), \packageLanguagepercent are \c{language} objects (expressions
-representing a function calls), and \packageExpressionpercent are \c{expression}
-objects (lists of expressions). Further inspection reveals that \c{\_inherit}
+arguments. More specifically, \packageSymbolpercent are \k{symbol}s (variables
+such as \k{x}), \packageLanguagepercent are \k{language} objects (expressions
+representing a function calls), and \packageExpressionpercent are \k{expression}
+objects (lists of expressions). Further inspection reveals that \k{\_inherit}
 accounts for \packageGgplotsymbolpercent of symbols and comes from one site in
-\c{ggplot2}.\footnote{Each object has a field model inheritance, this comes from
-\c{ggproto}, one of the many object-oriented systems for R.}
+\k{ggplot2}.\footnote{Each object has a field model inheritance, this comes from
+\k{ggproto}, one of the many object-oriented systems for R.}
 
@@
 size of the arguments in terms of number of nodes. For example,
-\c{expression(x+1)} has a size of 3. The large number of symbols skew the median
+\k{expression(x+1)} has a size of 3. The large number of symbols skew the median
 to \packageMedianszeval but the average size is \packageAvgszeval nodes. The
@@
   \bf $min(e)$& \bf \#sites & \bf \%sites & \bf \#packages & \bf \#operations & \bf \%envir & \bf example\\\midrule
-\c X&\packageMinimizedcallsitesa &\packageMinimizedpropsitesa &\packageMinimizedpackagea &\packageMinimizedmedianoperationsaRnd &\packageMinimizedpercentparentframesa & \c{y+1}\\
-\c{F(F(X))} & \packageMinimizedcallsitesb  & \packageMinimizedpropsitesb & \packageMinimizedpackageb  & \packageMinimizedmedianoperationsbRnd & \packageMinimizedpercentparentframesb & \c{gbov( mean(x), a-1)}\\
-\c{V}&\packageMinimizedcallsitesc &\packageMinimizedpropsitesc &\packageMinimizedpackagec &\packageMinimizedmedianoperationscRnd &\packageMinimizedpercentparentframesc& \c{c(42,21,0)}\\
-\c{F(X)}& \packageMinimizedcallsitesd & \packageMinimizedpropsitesd & \packageMinimizedpackaged & \packageMinimizedmedianoperationsdRnd & \packageMinimizedpercentparentframesd & \c{seq\_len(iters)} \\
-\c{\$} & \packageMinimizedcallsitese & \packageMinimizedpropsitese & \packageMinimizedpackagee & \packageMinimizedmedianoperationseRnd & \packageMinimizedpercentparentframese & \c{DF\$B}\\
-\c{model.frame}& \packageMinimizedcallsitesf & \packageMinimizedpropsitesf & \packageMinimizedpackagef & \packageMinimizedmedianoperationsfRnd & \packageMinimizedpercentparentframesf &  \c{model.frame(formula = Z $\sim$ U)}   \\
-\c{F()}& \packageMinimizedcallsitesg & \packageMinimizedpropsitesg & \packageMinimizedpackageg & \packageMinimizedmedianoperationsgRnd & \packageMinimizedpercentparentframesg & \c{rgamma(3, 2, n = 10L)} \\
-\c{FUN} & \packageMinimizedcallsitesh & \packageMinimizedpropsitesh & \packageMinimizedpackageh & \packageMinimizedmedianoperationshRnd & \packageMinimizedpercentparentframesh & \c{function(x, y) x + 3 * y} \\
-\c{<-} & \packageMinimizedcallsitesi  & \packageMinimizedpropsitesi & \packageMinimizedpackagei & \packageMinimizedmedianoperationsiRnd & \packageMinimizedpercentparentframesi & \c{x[1, 2:3, 2:3] <- value}\\
-\c{BLOCK} & \packageMinimizedcallsitesj & \packageMinimizedpropsitesj & \packageMinimizedpackagej & \packageMinimizedmedianoperationsjRnd & \packageMinimizedpercentparentframesj & \c{\{ write.csv(iris, tf) ; file.size(tf) \}} \\\bottomrule
+\k{X}&\packageMinimizedcallsitesa &\packageMinimizedpropsitesa &\packageMinimizedpackagea &\packageMinimizedmedianoperationsaRnd &\packageMinimizedpercentparentframesa & \k{y+1}\\
+\k{F(F(X))} & \packageMinimizedcallsitesb  & \packageMinimizedpropsitesb & \packageMinimizedpackageb  & \packageMinimizedmedianoperationsbRnd & \packageMinimizedpercentparentframesb & \k{gbov( mean(x), a-1)}\\
+\k{V}&\packageMinimizedcallsitesc &\packageMinimizedpropsitesc &\packageMinimizedpackagec &\packageMinimizedmedianoperationscRnd &\packageMinimizedpercentparentframesc& \k{c(42,21,0)}\\
+\k{F(X)}& \packageMinimizedcallsitesd & \packageMinimizedpropsitesd & \packageMinimizedpackaged & \packageMinimizedmedianoperationsdRnd & \packageMinimizedpercentparentframesd & \k{seq\_len(iters)} \\
+\k{\$} & \packageMinimizedcallsitese & \packageMinimizedpropsitese & \packageMinimizedpackagee & \packageMinimizedmedianoperationseRnd & \packageMinimizedpercentparentframese & \k{DF\$B}\\
+\k{model.frame}& \packageMinimizedcallsitesf & \packageMinimizedpropsitesf & \packageMinimizedpackagef & \packageMinimizedmedianoperationsfRnd & \packageMinimizedpercentparentframesf &  \k{model.frame(formula = Z ~ U)}   \\
+\k{F()}& \packageMinimizedcallsitesg & \packageMinimizedpropsitesg & \packageMinimizedpackageg & \packageMinimizedmedianoperationsgRnd & \packageMinimizedpercentparentframesg & \k{rgamma(3, 2, n = 10L)} \\
+\k{FUN} & \packageMinimizedcallsitesh & \packageMinimizedpropsitesh & \packageMinimizedpackageh & \packageMinimizedmedianoperationshRnd & \packageMinimizedpercentparentframesh & \k{function(x, y) x + 3 * y} \\
+\k{<-} & \packageMinimizedcallsitesi  & \packageMinimizedpropsitesi & \packageMinimizedpackagei & \packageMinimizedmedianoperationsiRnd & \packageMinimizedpercentparentframesi & \k{x[1, 2:3, 2:3] <- value}\\
+\k{BLOCK} & \packageMinimizedcallsitesj & \packageMinimizedpropsitesj & \packageMinimizedpackagej & \packageMinimizedmedianoperationsjRnd & \packageMinimizedpercentparentframesj & \k{\{write.csv(iris,tf);file.size(tf)\}} \\\bottomrule
 \end{tabular}
@@
 function $min(e)$ which for a given expression $e$ returns a normal form. Normal
-forms use \c V to stand in for values occurring in expression, \c X stands in
-for variables, and \c F stands for functions. The function performs constant
+forms use \k{V} to stand in for values occurring in expression, \k{X} stands in
+for variables, and \k{F} stands for functions. The function performs constant
 folding of arithmetic and string expressions over base operators, for instance
@@
 has $min(\c{x+y})$ becomes $\c{X}$, this motivated by the fact that addition is
-not an ``interesting'' operation, and that \c X stands in for any number of
+not an ``interesting'' operation, and that \k{X} stands in for any number of
 variable lookups. Function absorption simplifies nested functions to keep an
@@
 simplifications, the full list can be found in our artifact. It is worth
-mentioning that these simplifications are heuristics, in R \c{1+1} is not
-necessary \c 2. The addition operator can be redefined, but it typically isn't
+mentioning that these simplifications are heuristics, in R \k{1+1} is not
+necessary \k{2}. The addition operator can be redefined, but it typically isn't
 or at least not in a way that would invalidate arithmetic.
@@
 \medskip\EE{$min(e)=\c{X}$}\\[-2mm]\small Variable lookups are the most common shape,
-\packageNbSymbolVarSitePercent are simple variable names, e.g. \c x. The shapes
-subsumes \c{V} and built-in arithmetic operations, so a mixture of arithmetics
-and variables normalize to \c X. A lookup is one step of execution, if a promise
+\packageNbSymbolVarSitePercent are simple variable names, e.g. \k{x}. The shapes
+subsumes \k{V} and built-in arithmetic operations, so a mixture of arithmetics
+and variables normalize to \k{X}. A lookup is one step of execution, if a promise
 is returned an arbitrary number of additional steps may be needed; the median is
@@
 \\
-\medskip\EE{$min(e)=\c{\$}$}\\[-2mm]\small This shape extends \c X to include lookup
-with the dollar operator, \eg~\c{x\$f}, and vector indexing, \eg~\c{x[42]} and
-\c{x[[24]]}. This takes a few more step of evaluation on average, and is
+\medskip\EE{$min(e)=\c{\$}$}\\[-2mm]\small This shape extends \k{X} to include lookup
+with the dollar operator, \eg~\k{x\$f}, and vector indexing, \eg~\k{x[42]} and
+\k{x[[24]]}. This takes a few more step of evaluation on average, and is
 typically used in a function environment.
 \\
-\medskip\EE{$min(e)=$~\c{<-}}\\[-2mm]\small This includes both direct assignment {\tt
-  <-} and parent environment assignment {\tt <\,\!<-} as well as the \c{assign}
-function. It subsumes \c{\$}. Assignments represent the most obvious source
+\medskip\EE{$min(e)=$~\k{<-}}\\[-2mm]\small This includes both direct assignment {\tt
+  <-} and parent environment assignment {\tt <\,\!<-} as well as the \k{assign}
+function. It subsumes \k{\$}. Assignments represent the most obvious source
 of side effects in \eval. Most cases originate from trivial code generation
@@
   \medskip\EE{$min(e)=\c{FUN}$}\\[-2mm]\small Function definitions without any
-  computation, \eg~\c{function(x) x+1}. In addition to \c{FUN},
+  computation, \eg~\k{function(x) x+1}. In addition to \k{FUN},
   \packageGeneralizedFunctionDefinitionSitesPercent of sites have function
@@
 different from 0-argument closure. There is a number of use cases: unit testing
-frameworks (\eg \c{testthat}, \c{testit}), code benchmarking (\eg
-\c{rbenchmark}, \c{microbenchmark}), running code in parallel (\eg \c{foreach},
-\c{doParallel}), or deferring code execution (\eg \c{withr}).
+frameworks (\eg \k{testthat}, \k{testit}), code benchmarking (\eg
+\k{rbenchmark}, \k{microbenchmark}), running code in parallel (\eg \k{foreach},
+\k{doParallel}), or deferring code execution (\eg \k{withr}).
 
 \\
-\medskip\EE{$min(e)=\c{model.frame}$}\\[-2mm]\small The \c{model.frame} function
+\medskip\EE{$min(e)=\c{model.frame}$}\\[-2mm]\small The \k{model.frame} function
 returns a dataframe resulting from fitting the model described in a formula.
-This shape subsumes \c{F(F(X))}, \c{FUN}, and assignments. It is the single most
+This shape subsumes \k{F(F(X))}, \k{FUN}, and assignments. It is the single most
 popular function invoked from \eval. Each call does a median of 2K instructions.
@@
 more than eight shapes. An example of those is the pipe operator of the
-\c{magrittr} package which is used to compose functions, \eg instead of
-\c{f(g(h(x),y),z)} one can write \c{h(x) \%>\% g(y) \%>\% f(z)}. As there many
+\k{magrittr} package which is used to compose functions, \eg instead of
+\k{f(g(h(x),y),z)} one can write \k{h(x) \%>\% g(y) \%>\% f(z)}. As there many
 different patterns of use, there are also many different shapes. In comparison
@@
 other hand, there are \packageNbSimpleMinimizedOne sites that only receive one
-of the simple shape (\ie\xspace \c{X}, \c{V}, \c{\$} or \c{<-}), and
+of the simple shape (\ie\xspace \k{X}, \k{V}, \k{\$} or \k{<-}), and
 \packageNbSimpleMinimizedMore sites get multiple simple shapes. Among these
 sites, there are many cases where \eval could be replaced by simpler constructs
-such as \c{get} to lookup a name in an environment, \c{assign} to set a value to
-a name in an environment, or \c{do.call} to call a function reflectively.
+such as \k{get} to lookup a name in an environment, \k{assign} to set a value to
+a name in an environment, or \k{do.call} to call a function reflectively.
 
@@
 Contrary to JavaScript, in R the environment of \eval can be specifies by its
-\c{env} argument. This gives users control over what is visible to the
+\k{env} argument. This gives users control over what is visible to the
 computation started by \eval and the potential reach of its side-effects. We
@@
 \item \emph{Function:} environments for the local variables of some function
-  currently active on the call stack. Obtained by calling \c{parent.frame()} or
-  \c{sys.frame()}.
+  currently active on the call stack. Obtained by calling \k{parent.frame()} or
+  \k{sys.frame()}.
 \item \emph{Synthetic:} environments built from data structures such as lists,
-  data.frames, or constructed explicitly with \c{new.env}, \c{list2env}, or
-  \c{as.environment} or the empty environment.
+  data.frames, or constructed explicitly with \k{new.env}, \k{list2env}, or
+  \k{as.environment} or the empty environment.
 \item \emph{Global:} environments in which scripts or interactive commands are
@@
 Each synthetic environment has parent that is specified when calling
-\c{new.env}. When \c{envir} is a list or a data frame, \eval uses its third
-argument (\c{enclos}). The parent is used to search for variables not found in
+\k{new.env}. When \k{envir} is a list or a data frame, \eval uses its third
+argument (\k{enclos}). The parent is used to search for variables not found in
 its child (side-effects stay in the child). Table~\ref{tab:newenvs} shows parent
@@
 Global \evals split between intentional and accidental one. Direct references to
-the top-level, using \c{globalenv()} or \c{.GlobalEnv}, are rare; they occur in
+the top-level, using \k{globalenv()} or \k{.GlobalEnv}, are rare; they occur in
 only \packageNbExplicitGlobalSites sites. Thus we suspect most uses of global
@@
 expectation is that the expression was extracted from a promise using
-\c{substitute}, perhaps modified, and then evaluated with the environment coming
+\k{substitute}, perhaps modified, and then evaluated with the environment coming
 from the same promise. We believe the case where \eval is provided the results
@@
 the values that are given to \eval. The following example yields the provenance
-graph with five nodes on Figure~\ref{fig:provgraph} for expression \c{e}.
+graph with five nodes on Figure~\ref{fig:provgraph} for expression \k{e}.
 
@@
 		]
-		\node (E) at (2,0) {\code{eval(e)}};
-		\node[text width = 3cm] (CA) at (2, 1.5) {\code{e[[2]] <-quote(c)}};
-		\node (Q) at (4, 3) {\code{quote(c)}};
-		\node[text width = 4cm] (A) at (0, 3) {\code{e <- parse(text = "a;b")}};
-		\node[text width = 3cm] (P) at (0, 4.5) {\code{parse(text = "a;b")}};
+		\node (E) at (2,0) {\k{eval(e)}};
+		\node[text width = 3cm] (CA) at (2, 1.5) {\k{e[[2]] <-quote(c)}};
+		\node (Q) at (4, 3) {\k{quote(c)}};
+		\node[text width = 4cm] (A) at (0, 3) {\k{e <- parse(text = "a;b")}};
+		\node[text width = 3cm] (P) at (0, 4.5) {\k{parse(text = "a;b")}};
 
@@
 
-The provenance of \c{e} includes both \c{parse} and \c{quote}. We pick an origin
+The provenance of \k{e} includes both \k{parse} and \k{quote}. We pick an origin
 by traversing the graph from the root, following the left-hand side member of
 assignments until a leaf. It represents the provenance that is further modified
-to yield the expression passed to \eval. In our example, the origin is \c{parse}.
+to yield the expression passed to \eval. In our example, the origin is \k{parse}.
 
@@
 \begin{itemize}[---]
-\item {\it Reflection:} use of \c{match.call} to reflectively capture the
+\item {\it Reflection:} use of \k{match.call} to reflectively capture the
   expression that invoked this function, arguments are promises, and
-  \c{substitute} is used to retrieve their source.
-\item {\it String:} create from strings by invoking \c{parse},
-  \c{str2expression}, or \c{str2lang}.
-\item {\it Constructed:} invoked \c{quote} or \c{expression}, or by built with
-  \c{call} or \c{as.call}.
-\item {\it Environment: } created with \c{as.name} or \c{as.symbol}, typically
+  \k{substitute} is used to retrieve their source.
+\item {\it String:} create from strings by invoking \k{parse},
+  \k{str2expression}, or \k{str2lang}.
+\item {\it Constructed:} invoked \k{quote} or \k{expression}, or by built with
+  \k{call} or \k{as.call}.
+\item {\it Environment: } created with \k{as.name} or \k{as.symbol}, typically
   to read a non-local environment.
@@
 different provenances, they are counted multiple times. Strings correlate with
-dynamic code loading as done by \c{source} and \c{sys.source}. Few calls
-(\packageNbParseFromFileSites in total) consume the result of calling \c{parse}
+dynamic code loading as done by \k{source} and \k{sys.source}. Few calls
+(\packageNbParseFromFileSites in total) consume the result of calling \k{parse}
 on a file. Most build strings programmatically. We identified one function
-\c{invokeRestartInteractively} that prompts the user for input and \evals it.
+\k{invokeRestartInteractively} that prompts the user for input and \evals it.
 
 \paragraph{Discussion}
-The main use case for \eval involves using \c{match.call}, \c{substitute}, or
-\c{formals} to access and transform arguments of a function. Constructing
+The main use case for \eval involves using \k{match.call}, \k{substitute}, or
+\k{formals} to access and transform arguments of a function. Constructing
 expressions from strings is also a common use case. The constructed category
@@
 unit testing frameworks as they use \eval to run all their tests and thus
-dominate the side-effect data.\footnote{The corpus has \c{RUnit, testthat,
-  tinytest} and \c{unitizer} testing frameworks.}
+dominate the side-effect data.\footnote{The corpus has \k{RUnit, testthat,
+  tinytest} and \k{unitizer} testing frameworks.}
 
@@
 virtual machine implementation and not user code. For example, the
-\c{.Random.seed} variable is saved and restored from and to the global
+\k{.Random.seed} variable is saved and restored from and to the global
 environment every time a statistical routine is called. Filtering accidental
@@
 responsible for 90\% of side effects. Half of those come from just three
-functions: \c{plyr::allocate\_column} (allocates space for a new data frame
-column), \c{withr::execute\_handlers} (executes deferred expressions), and
-\c{foreach::doSEQ} (executes an expression on each element in a collection,
+functions: \k{plyr::allocate\_column} (allocates space for a new data frame
+column), \k{withr::execute\_handlers} (executes deferred expressions), and
+\k{foreach::doSEQ} (executes an expression on each element in a collection,
 possibly in parallel). Most of the sites (\SESitesInEnvirRatio) update
-the environment specified by the \c{env} parameter.
+the environment specified by the \k{env} parameter.
 
@@
 happens, but the vast majority comes from a single site
-(\c{withr::execute\_handlers}). This makes sense as it is used to defer
+(\k{withr::execute\_handlers}). This makes sense as it is used to defer
 evaluation of an expression to after the function exit and thus used for clean
-up. It is used almost exclusively by the \c{tidyselect} package for removing the
+up. It is used almost exclusively by the \k{tidyselect} package for removing the
 reference to the current quosure environment while interpreting a data frame
@@
 environment. We have manually inspected a few of these sites, but except for
-manual injection of parameters into a \c{model.frame} execution environment, we
+manual injection of parameters into a \k{model.frame} execution environment, we
 did not find a common use case.
@@
 
-Out of the minimized expressions, it is \c{BLOCK}, \c{<-} and \c{F(F(X))} that
-contribute to the vast majority of side effects. Concretely, 92\% of \c{BLOCK},
-87\% of \c{<-} and 15\% of \c{F(F(X))} expressions do side effects. In the case
-of \c{BLOCK}, most of them happen in the environment passed to the \eval call
-(default is the parent environment). For \c{<-} it is 15\%. However, this form
-is dominated by the \c{plyr::allocate\_column} function that side-effects in
+Out of the minimized expressions, it is \k{BLOCK}, \k{<-} and \k{F(F(X))} that
+contribute to the vast majority of side effects. Concretely, 92\% of \k{BLOCK},
+87\% of \k{<-} and 15\% of \k{F(F(X))} expressions do side effects. In the case
+of \k{BLOCK}, most of them happen in the environment passed to the \eval call
+(default is the parent environment). For \k{<-} it is 15\%. However, this form
+is dominated by the \k{plyr::allocate\_column} function that side-effects in
 the environment containing the data frame to which it allocates new column.
-Without this call, \c{<-} is similarly predictable with 96\% of side effects
-happening in the passed environment. For \c{F(F(X))}, it is less clear as 58\%
+Without this call, \k{<-} is similarly predictable with 96\% of side effects
+happening in the passed environment. For \k{F(F(X))}, it is less clear as 58\%
 of side effects happen in a different environment to the one that was passed to
@@
 The R language was intended to be extensible. The combination of lazy
-evaluation, \c{substitute} and \c{eval} are some of the tools given to
+evaluation, \k{substitute} and \k{eval} are some of the tools given to
 developers to this end. To understand \emph{why} developers use \eval, we
@@
 provide formulas as strings, such as \k{"exp(-x^2/2)"}, the contract being that
-the text can refer to some variable \c x that will be set up by the called
+the text can refer to some variable \k{x} that will be set up by the called
 function. In this case one users could provide an anonymous function.
@@
 write by hand or that is only known after deployment. The example uses \eval to
-construct function objects that are parameterized with string \c{what}.
+construct function objects that are parameterized with string \k{what}.
 This is used  automatically generate R bindings to native code.
@@
 of the vararg until it finds a non-null value. \k{substitute(alist(...))} yields
-an expression that represents a call to \k{alist} but where the \code{...} are
+an expression that represents a call to \k{alist} but where the \k{...} are
 expended with the arguments supplied to the function. \Eval then returns a
-\code{list} with unevaluated argument expressions which are evaluated
+\k{list} with unevaluated argument expressions which are evaluated
 sequentially for their result.
@@
 with a list of expressions whose code is extracted using
-\code{eval(substitute(alist(...)))}. The expression are evaluated sequentially
+\k{eval(substitute(alist(...)))}. The expression are evaluated sequentially
 in the caller's environment. If an expression is false, an error message is
@@
 %%% TODO ?????????????////
-Many packages use \eval to implement variations of the \code{base} R functions.
-For example, the \code{igraph} package provides the \code{do_call} function
-shown below. This is a variant of the \code{do.call} function from the
-\code{base} package. This function executes a function call by constructs it
+Many packages use \eval to implement variations of the \k{base} R functions.
+For example, the \k{igraph} package provides the \k{do_call} function
+shown below. This is a variant of the \k{do.call} function from the
+\k{base} package. This function executes a function call by constructs it
 from a name and a list of arguments and calling \eval.
@@
 A variant of code generation is to let users write compact code that is expended
-via \eval and \code{match.call}. The example uses \code{match.call} to access
+via \eval and \k{match.call}. The example uses \k{match.call} to access
 the expression with which the function is called. The next three lines construct
-a call to \code{stats::model.frame} with argument text of parameters
-\code{'formula'}, \code{'data'}, and \code{'weights'}. Then, the call expression
-is evaluated in the parent environment. The use of \code{match.call}, avoids
-repeated use of \code{substitute} for extracting the argument expressions. This
+a call to \k{stats::model.frame} with argument text of parameters
+\k{'formula'}, \k{'data'}, and \k{'weights'}. Then, the call expression
+is evaluated in the parent environment. The use of \k{match.call}, avoids
+repeated use of \k{substitute} for extracting the argument expressions. This
 setup is also extensible as new parameter names can be easily added.
@@
 Many packages use \eval to evaluate formulas. The example extracts the
-\code{"formula"} attribute attached to the \code{pdMat} object and evaluates it.
+\k{"formula"} attribute attached to the \k{pdMat} object and evaluates it.
 
@@
 \paragraph{Obfuscation}
-Some packages use \eval to bypass the restrictions imposed by the \code{R CMD
+Some packages use \eval to bypass the restrictions imposed by the \k{R CMD
   CHECK} tool. This tool enforces some well-formedness rules on a submitted
@@
 that the package code does not use certain restricted functions. Package authors
-use \eval to obfuscate their code. The example mutates the variable \code{s} in
+use \eval to obfuscate their code. The example mutates the variable \k{s} in
 the package environment. However, that is locked by default and
-\code{unlockBinding}, is a restricted function. To work around this, the call to
+\k{unlockBinding}, is a restricted function. To work around this, the call to
 unlock is done through \eval.
@@
 
-\emph{Variable lookup and values, and indexing \c{(X, V, \$)}}
+\emph{Variable lookup and values, and indexing \k{(X, V, \$)}}
     constitute one of the most common use case for \eval. Let's consider the
-    \c{subset} function from the R base library, which allows, among others, to
-    select variables in a data frame. For example, given a data frame \c{df}
-    with three columns \c{A,B,C}, \c{subset(df, sel=A)} returns a subset of
-    \c{df} with just the \c{A} column. The same would happen with \c{subset(df,
+    \k{subset} function from the R base library, which allows, among others, to
+    select variables in a data frame. For example, given a data frame \k{df}
+    with three columns \k{A,B,C}, \k{subset(df, sel=A)} returns a subset of
+    \k{df} with just the \k{A} column. The same would happen with \k{subset(df,
       sel=1)}. However, thanks to \eval it can also be also called with
-    \c{subset(df, sel=c(1,B,x))} returning a subset of the first column, column
-    \c{B} and whatever index the variable \c{x} holds. It is implemented using:
+    \k{subset(df, sel=c(1,B,x))} returning a subset of the first column, column
+    \k{B} and whatever index the variable \k{x} holds. It is implemented using:
     \lstinline|eval(substitute(sel), nl, parent.frame())| The first argument
-    returns the expression passed to \c{sel}, the second is the data frame in
+    returns the expression passed to \k{sel}, the second is the data frame in
     which this expression will be evaluated, and the last indicates which
-    environment should be used to look up bindings not found in \c{nl}.
-
- \emph{Assignments \c{(<-)}} appear in the cases where a developer wants
+    environment should be used to look up bindings not found in \k{nl}.
+
+ \emph{Assignments \k{(<-)}} appear in the cases where a developer wants
     to create or update a binding in a specific environment where either the
@@
     trivial code generation where the assign expression is assembled using
-    \c{parse} or \c{substitute}, often in a loop. For example, the
-    \c{PopED::get\_rse} uses it to assign a number of variables in the current
+    \k{parse} or \k{substitute}, often in a loop. For example, the
+    \k{PopED::get\_rse} uses it to assign a number of variables in the current
     scope: \lstinline|eval(parse(text=paste(default_args[[i]]), "<-", i))|.
 
- \emph{Function calls \c{(F(), F(X), F(F(X)))}} represent two very
+ \emph{Function calls \k{(F(), F(X), F(F(X)))}} represent two very
     frequent patterns: \emph{code generation} and \emph{call reflection}. There
@@
     One occurring usage (present also in the Kaggle scripts) is to generate
-    ad-hoc calls to libraries such as \c{ggplot2} when one needs to
+    ad-hoc calls to libraries such as \k{ggplot2} when one needs to
     parameterize data or aesthetics selection from a variable. For example
     \lstinline|ggplot(df, aes(x=A, y=B))| maps X and Y axes in the plot to
-    \c{A} and \c{B} columns in \c{df}. It might not be immediately
+    \k{A} and \k{B} columns in \k{df}. It might not be immediately
     clear\footnote{\cf
@@
     boilerplate code; simple and repetitive code can easily be replaced with
-    judicious use of \eval. For example, the \c{data.table} package uses \eval
-    to call the \c{options} function with named arguments taken from a vector
+    judicious use of \eval. For example, the \k{data.table} package uses \eval
+    to call the \k{options} function with named arguments taken from a vector
     of strings:
@@
 The cases where \eval appears to be overkill include simple expression shapes
-such as \c X, \c V, \$, \c{<-}, and \c{FUN}. Variable lookup in a remote
-environment can be performed by the builtin \code{get} function, a similar case
-can be made for assignment and the \code{assign} function. There is a dedicated
-\c{do.call} function to build function calls. Unnecessary uses of \eval
+such as \k{X}, \k{V}, \k{\$}, \k{<-}, and \k{FUN}. Variable lookup in a remote
+environment can be performed by the builtin \k{get} function, a similar case
+can be made for assignment and the \k{assign} function. There is a dedicated
+\k{do.call} function to build function calls. Unnecessary uses of \eval
 often can be uncovered by manual analysis.
@@
 current functions. \Eval is overkill for this code, a semantically equivalent
-function body uses \c{get} to perform the lookup and then calls directly.
+function body uses \k{get} to perform the lookup and then calls directly.
 
@@
 
-Another example, from \c{coin::copyslots}, constructs an expression that access
+Another example, from \k{coin::copyslots}, constructs an expression that access
 fields of an object and copy them. This can be replaced by symbolic field reads
@@
 
-As a last example, \code{configr::config.funs.par} construct three expressions
-that are then fed to \eval. Semantically equivalent code would use \c{get}
-to lookup \c{fun}, and then directly execute the three expressions.
+As a last example, \k{configr::config.funs.par} construct three expressions
+that are then fed to \eval. Semantically equivalent code would use \k{get}
+to lookup \k{fun}, and then directly execute the three expressions.
 
@@
 libraries include their own implementation calls
-\c{rlang::eval\_tidy}~\cite{tidyverse}. This departs from \eval in two ways.
+\k{rlang::eval\_tidy}~\cite{tidyverse}. This departs from \eval in two ways.
 First, it supports evaluation of quosures, custom objects used by for
-metaprogramming that bundle expressions with an environment. \c{eval\_tidy}
+metaprogramming that bundle expressions with an environment. \k{eval\_tidy}
 accepts a data mask argument, a set of bindings such as a dataframe or list,
 which takes precedence over the environment. Effects such as assignments happen
-in the data mask. Expressions such as \c{return()} do not work since the data
+in the data mask. Expressions such as \k{return()} do not work since the data
 mask does not correspond to a frame on the call stack.

### 2021-08-12 paper/main.tex
Center more Table 1 (some hspace* ...)
Fig 1 -> Table 1

@@
 environment in which generated code evaluates and the reflective operations
-available to that code. Fig.~\ref{comp} summarizes a few designs.
-
-\begin{figure}[!h]\center\small\begin{tabular}{r@{~}llll}\toprule
+available to that code. Table~\ref{comp} summarizes a few designs.
+
+\begin{table}[!h]\center\small\begin{tabular}{r@{~}llll}\toprule
 \tiny\sc Language&&\sc\tiny Input&\sc\tiny Scope&\tiny\sc Reflective operations\\\midrule
@@
 \end{tabular}\caption{Design space of \eval}\label{comp}
-\end{figure}
+\end{table}
 
@@
 \begin{table}[!b]\small
+\hspace*{-0.5cm}
 \begin{tabular}{crrrrrc}\toprule
@@
 
-\emph{Variable lookup and values, and indexing \k{(X, V, \$)}}
-    constitute one of the most common use case for \eval. Let's consider the
-    \k{subset} function from the R base library, which allows, among others, to
-    select variables in a data frame. For example, given a data frame \k{df}
-    with three columns \k{A,B,C}, \k{subset(df, sel=A)} returns a subset of
-    \k{df} with just the \k{A} column. The same would happen with \k{subset(df,
-      sel=1)}. However, thanks to \eval it can also be also called with
-    \k{subset(df, sel=c(1,B,x))} returning a subset of the first column, column
-    \k{B} and whatever index the variable \k{x} holds. It is implemented using:
-    \lstinline|eval(substitute(sel), nl, parent.frame())| The first argument
-    returns the expression passed to \k{sel}, the second is the data frame in
-    which this expression will be evaluated, and the last indicates which
-    environment should be used to look up bindings not found in \k{nl}.
-
- \emph{Assignments \k{(<-)}} appear in the cases where a developer wants
-    to create or update a binding in a specific environment where either the
-    name or the value come from some expression. Most cases originate from
-    trivial code generation where the assign expression is assembled using
-    \k{parse} or \k{substitute}, often in a loop. For example, the
-    \k{PopED::get\_rse} uses it to assign a number of variables in the current
-    scope: \lstinline|eval(parse(text=paste(default_args[[i]]), "<-", i))|.
-
- \emph{Function calls \k{(F(), F(X), F(F(X)))}} represent two very
-    frequent patterns: \emph{code generation} and \emph{call reflection}. There
-    are numerous cases for code generation.
-    %
-    One occurring usage (present also in the Kaggle scripts) is to generate
-    ad-hoc calls to libraries such as \k{ggplot2} when one needs to
-    parameterize data or aesthetics selection from a variable. For example
-    \lstinline|ggplot(df, aes(x=A, y=B))| maps X and Y axes in the plot to
-    \k{A} and \k{B} columns in \k{df}. It might not be immediately
-    clear\footnote{\cf
-    \url{https://stackoverflow.com/questions/22309285/how-to-use-a-variable-to-specify-column-name-in-ggplot}}
-    for a ggplot2 user how to modify the call to assign the axis columns from
-    variables. Another case is the traditional use of \eval as a mean to reduce
-    boilerplate code; simple and repetitive code can easily be replaced with
-    judicious use of \eval. For example, the \k{data.table} package uses \eval
-    to call the \k{options} function with named arguments taken from a vector
-    of strings:
-
-    \begin{lstlisting}[gobble=2]
-    opts = c("datatable.verbose"="FALSE", ... )
-    for (i in names(opts)) eval(parse(text=paste0("options(",i,"=",opts[i],")")))
-    \end{lstlisting}
-
 \section{Discussion}

### 2021-08-12 paper/main.tex
Proofread of 2.3

@@
 
-The expression passed to \eval can be thought of as an abstract syntax
-tree. It can be parsed from a string, created manually or using reflection.
+The expression passed to \eval can be thought of as an abstract syntax tree.
+Listings~\ref{lst:exprs} and \ref{lst:match.call} show some of the ways of
+creating expressions: parsing from a string, manually or via reflection. The
+call to \k{substitute(x)} extracts the unevaluated expression from the promise
+\k{x}. The call to \k{match.call()} returns an expression representing the
+current call which can then be further modified.
 
@@
 \begin{minipage}{.49\textwidth}
-\begin{lstlisting}[caption={Examples of calls producing expression \k{a+b}}]
+\begin{lstlisting}[caption={Examples of calls producing expression \k{a+b}},label=lst:exprs]
   parse(text="a+b")
@@
 \begin{minipage}{.49\textwidth}
-  \begin{lstlisting}[caption={Example of a call reflection}]
+  \begin{lstlisting}[caption={Example of a call reflection},label=lst:match.call]
   f <- function(x, y) {
@@
     mc[[1]] <- as.name("g")
-    mc[["x"]] <- 2
+    mc[["x"]] <- 3
     mc
   }
-  f(1,2) # returns a call g(2,y)
+  f(1,2) # returns an expr g(x=3,y=2)
   \end{lstlisting}
@@
 \end{figure}
-
- The call to \k{substitute(x)} extracts the unevaluated expression from
-the promise \k{x}. The function \k{match.call} returns an expression
-representing the current call.
 
@@
 %
-\begin{lstlisting}
-  # reading                                # writing
-  envir$v                                  envir$v <- 2
-  get("v", envir=envir)                    assign("v", 2, envir=envir)
-\end{lstlisting}
+\begin{figure}[H]
+\begin{minipage}{.49\textwidth}
+  \begin{lstlisting}
+  # reading
+  envir$v              
+  get("v", envir=envir)
+  \end{lstlisting}
+\end{minipage}
+\begin{minipage}{.49\textwidth}
+  \begin{lstlisting}
+  # writing
+  envir$v <- 2
+  assign("v", 2, envir=envir)  
+  \end{lstlisting}
+\end{minipage}
+\end{figure}
 %
- Environments are used as hash maps as they have reference semantics
+\noindent Environments are used as hash maps as they have reference semantics
 and a built-in string lookup.

### 2021-08-12 paper/main.tex
Remove expression type explanations from Usage of Eval
Update the comparison to Morandat et al

@@
 
-\citet{ecoop12} had a few paragraphs on the usage of \eval in R. They found the
-that \eval is widely used in R code with 8500 call sites in 1238 CRAN packages
-and 2.2 million dynamic calls. The 15 most frequent call sites account for 88\%
-of those. The \k{match.arg} function is the highest used one with 54\% of all
-calls. In the other call sites, they saw two use cases. The most common is the
-evaluation of the source code of a promise retrieved by \k{substitute} in a new
-environment, \eg as done in the \k{with} function. The other use case is the
-invocation of a function whose name or arguments are determined dynamically. For
-this purpose, R provides \k{do.call} and thus \eval is overkill. Our corpus is
-larger, with \CranPackages packages, \packageTriggeredpkgs packages with at
-least one \eval, for a total of \PkgHitEvalCallSites sites and
-\packageAllcallsRnd dynamic calls, although they also looked at the Bioconductor
-package repository in addition to CRAN. We also track much more runtime
-information on the \eval calls during tracing. It enables us to have a more
-nuanced and accurate picture, and identify more diverse use cases of \eval.
+\citet{ecoop12} in their evaluation of the design of the R language, briefly
+looked at the use of \eval in their corpus (1238 CRAN packages, R base
+libraries, and 515 Bioconductor packages). They found 8500 \eval call sites and
+recorded 2.2M \eval calls at runtime. Because they included base, the results
+were dominated by a few functions (15 call sites accounted for 88\% of \eval
+calls) with \c{match.arg}\footnote{A popular base function that matches an
+  argument against a set of candidate values that uses \eval to get the
+candidate values.} being responsible for over half of the recorded \eval
+calls. They discussed two use cases. The first one is the evaluation of an
+expression extracted from a promise using the \k{substitute} call in a new
+environment. The other is the invocation of a function with dynamically
+computed names and arguments.
+
+In this work, we significantly expand that study. First, our corpus covers
+\CranPackages packages (about the whole of CRAN at time of writing) with
+\PkgHitEvalCallSites \eval call sites and \packageAllcallsRnd \eval calls.
+Next, we track the number of information such as the shapes of expressions
+passed to \eval, their provenance, complexity, and side effects. This enables
+us to have an accurate and detailed picture of the use of \eval in the R
+ecosystem. We use this to provide both quantitative and qualitative views on
+\eval, including several use cases to demonstrate the popularity, diversity and
+powerfulness of \eval in R.
 

### 2021-08-12 paper/main.tex
Fix typo
PR section 3

@@
 \PkgHitEvalCallSites \eval call sites and \packageAllcallsRnd \eval calls.
-Next, we track the number of information such as the shapes of expressions
+Next, we track a number of information such as the shapes of expressions
 passed to \eval, their provenance, complexity, and side effects. This enables
@@
   originating from our execution harness.
-\item \emph{Trace.} Script run with a modified R interpreter to capture
-  execution traces. Each script is run in its own process with the GNU R
-  compiler turned off to avoid recording its execution. CRAN scripts run twice,
-  once to capture \eval calls originating from package code and a second time to
-  capture calls coming from base libraries.
+\item \emph{Trace.} Script run with a modified R interpreter to capture calls
+  to \eval in CRAN packages. Each script is run in its own process with the GNU
+  R compiler turned off to avoid recording its execution.
 \item \emph{Analyze.} Analysis outputs are merged, cleaned, and summarized.
@@
 
- The pipeline runs in parallel~\cite{GNUparallel} orchestrated by a
+\medskip The pipeline runs in parallel~\cite{GNUparallel} orchestrated by a
 Makefile. Servers have identical environments thanks to a docker image with all
@@
 variants of \eval and a few additional functions to locate the origin of \eval
-arguments (it taints the results of calls to \k{parse} and \k{match.call}). It
-captures dynamic code loading, as well as variable definition and assignment,
-allowing us to record side effects that happen in environments while evaluating
-code in \eval. A challenge was that lack of source code references for
-expression that are not within block surrounded by braces. This is unfortunately
-not easily fixed. We extend our dynamic analysis tool to attach synthetic source
-code references to all \eval call sites, but the approach fails in some edge
-cases. For performance reasons, the tracer is an R package written in C++ (3.2K
-LOC) and R (1.3K LOC). While in theory the implementation should be
-straightforward, not so in practice. Lazy evaluation requires delaying the
-processing of arguments until (and if) they are forced. Accounting for
-side-effects performed in an \eval is complicated by the fact that R is
-implemented in a mixture of R and C, and that the language implementation can
-call \eval. Lastly, large code bases exercise many edge cases of the highly
-underspecified R behavior.
+arguments. It captures dynamic code loading, as well as variable definition and
+assignment, allowing us to record side effects that happen in environments
+while evaluating code in \eval. A challenge was that lack of source code
+references for expression that are not within block surrounded by braces. This
+is unfortunately not easily fixed. We extend our dynamic analysis tool to
+attach synthetic source code references to all \eval call sites, but the
+approach fails in some edge cases. For performance reasons, the tracer is an R
+package written in C++ (3.2K LOC) and R (1.3K LOC). While in theory the
+implementation should be straightforward, not so in practice. Lazy evaluation
+requires delaying the processing of arguments until (and if) they are forced.
+Accounting for side-effects performed in an \eval is complicated by the fact
+that R is implemented in a mixture of R and C, and that the language
+implementation can (and does) call \eval. Lastly, large code bases exercise
+many edge cases of the highly underspecified R behavior.
 
@@
 all \eval calls). This happen when \eval is an argument to a higher-order
-function or when it is called from native code. An other limitation is that we
+function or when it is called from native code. Another limitation is that we
 do not record calls to the native \eval. Neither of these limitations should

### 2021-08-12 paper/main.tex
PR section 4

@@
 
-\begin{figure}[!b]
-\centering
+\begin{figure}[H]
   \includegraphics[width=9cm]{pkgs-eval-callsites-hist.pdf}
   \caption{ \eval call sites}\label{fig:pkgs-eval-callsites-hist}
-
-\medskip
-  \medskip
-
+\end{figure}
+\begin{table}[H]
+\centering
   \small
@@
 \end{tabular}
-\caption{Call frequency}\label{freq}
+\caption{Call frequency}\label{tab:freq}
 \end{minipage}
@@
 
-\medskip
-
+\end{table}
+\begin{figure}[H]
   \includegraphics[width=.8\textwidth]{traced-eval-callsites.pdf} \centering
@@
 sites for each program; as can be seen coverage is unequal.
-Fig.~\ref{freq} summarizes the frequency of calls, call counts on the left and
+Table~\ref{tab:freq} summarizes the frequency of calls, call counts on the left and
 number of packages on the right. There are \packageFewcalls packages with fewer
@@
 over half of our observations.
-Fig.~\ref{tab:variantseval} summarizes use of variants, the first row has sites
+Table~\ref{tab:variantseval} summarizes use of variants, the first row has sites
 (\emph{static}), the second has sites encountered during analysis
@@
 
-\begin{wrapfigure}{r}{5cm}  \small  \vspace*{-2mm}\centering
+\begin{wraptable}{r}{5cm}  \small  \vspace*{-2mm}\centering
   \begin{tabular}{r@{\,}r@{\,}l@{\,}r|r@{\,}r@{\,}l@{}r} \toprule
@@
   \end{tabular}
-  \caption{Normalized calls} \label{cn}
-\end{wrapfigure}
-
-Fig.~\ref{cn} shows average count of calls from a given site and given run and
+  \caption{Normalized calls} \label{tab:cn}
+\end{wraptable}
+
+Table~\ref{tab:cn} shows average count of calls from a given site and given run and
 how many sites fall in that range. For instance, \packageRunbinh sites are
@@
 \eval returns it unchanged. Expressions account for \packageCodepercent of
-arguments. More specifically, \packageSymbolpercent are \k{symbol}s (variables
-such as \k{x}), \packageLanguagepercent are \k{language} objects (expressions
-representing a function calls), and \packageExpressionpercent are \k{expression}
+arguments. More specifically, \packageSymbolpercent are symbols (variables
+such as \k{x}), \packageLanguagepercent are language objects (expressions
+representing a function calls), and \packageExpressionpercent are expression
 objects (lists of expressions). Further inspection reveals that \k{\_inherit}

### 2021-08-12 paper/main.tex
minor

--- 2021-08-12 paper/main.tex
@@
 to shield optimized code from some of the most pernicious uses of the
-facility~\cite{oopsla18a}. Furthermore, Julia provides a versioning mechanism,
-called world age, to ensure that any methods defined within an \eval only become
-visible at well-defined program points and thus that recompilation does not have
-to occur when optimized code is running~\cite{oopsla20a}.
+facility~\cite{oopsla18a}. Furthermore, Julia provides a versioning mechanism to
+ensure that any method defined within an \eval only become visible at
+well-defined points and thus that optimized code does not have to be
+invalidated~\cite{oopsla20a}.
 
@@
 
-The \eval function in R takes three parameters: an \k{expression} to evaluate
-(\k{e}), an environment where to evaluate (\k{envir}), and an enclosure
-(\k{encl}) that is used to look up objects not found in \k{envir}.\footnote{Next
-to the \eval function R offers three other variants:
-  \begin{inparaenum}[1.]
-    \item \k{evalq} automatically quotes passed expression---a shorthand to
-      \k{eval(quote(...))},
-    \item \k{eval.parent(e, n)} specifies the evaluating environment in terms of
-      number of call frames (\k{n}) to go back---a shorthand to \k{eval(e,
-        parent.frame(n))}, and
-    \item \k{local} evaluates \k{e} in a fresh environment---a shorthand to
-      \k{evalq(e, new.env())}.
-  \end{inparaenum} }
-
-\begin{lstlisting}
- eval <- function(e, envir = parent.frame(),
-                   encl = if(is.list(envir)) parent.frame() else baseenv()) ...
+The \eval function in R takes three parameters: an expression to evaluate
+(\k{e}), an environment where to evaluate (\k{env}), and an enclosure (\k{encl})
+that is used to look up objects not found in \k{env}.\footnote{R offers three
+other variants of {\tt eval}: {\tt evalq} automatically quotes passed
+expression---shorthand for {\tt eval(quote(...))}, {\tt eval.parent(e,n)}
+specifies the evaluating environment in terms of number of call frames ({\tt n})
+to go back---shorthand for {\tt eval(e,parent.frame(n))}, and {\tt local}
+evaluates {\tt e} in a fresh environment---shorthand for {\tt
+  evalq(e,new.env())}.}
+
+\begin{lstlisting}
+ eval <- function(e, env = parent.frame(),
+                   encl = if(is.list(env)) parent.frame() else baseenv()) ...
 \end{lstlisting}
@@
 
-R is generous when it comes to what is allowed to be passed as environment to
-\eval. Besides ordinary environments, it also accepts a list or a data
-frame (using element names or column names for variable resolution) or an
-integer \k{n} (in which case it would use the \k{n.} call frame). If no
-environment is specified, then the default is the environment where the call to
-\eval was made (as returned by \k{parent.frame}).
+R is permissive in terms of what is considered an environment. Besides
+environments, it accepts lists, data frames (using element names or column names
+for variable resolution), or an integer \k{n} (in which case it would use the
+\k{n.} call frame). The default is the environment where the call to \eval was
+made.
 %
-Environments nest, each has a parent. When creating a new environment with
-\k{new.env}, the parent is the current environment. Environment chains can be
-traversed with \k{parent.env}, until \k{emptyenv} is reached. The top-level
-environment is \k{.GlobalEnv}, it has parents that represent the packages that
-have been loaded. One can also directly read, modify or create new bindings,
-given any environment:
+Environments nest, each has a parent. A new environment created with \k{new.env}
+has the current environment as parent. Parent chains canf be traversed with
+\k{parent.env}, until \k{emptyenv} is reached. The top-level environment is
+\k{.GlobalEnv}, its parents are the packages that have been loaded. Environments
+are used as hash maps as they have reference semantics and a built-in string
+lookup. One can also directly read, modify or create new bindings, given any
+environment:
 %
@@
 %
-\noindent Environments are used as hash maps as they have reference semantics
-and a built-in string lookup.
 
@@
 \citet{ecoop12} in their evaluation of the design of the R language, briefly
-looked at the use of \eval in their corpus (1238 CRAN packages, R base
-libraries, and 515 Bioconductor packages). They found 8500 \eval call sites and
-recorded 2.2M \eval calls at runtime. Because they included base, the results
-were dominated by a few functions (15 call sites accounted for 88\% of \eval
-calls) with \c{match.arg}\footnote{A popular base function that matches an
-  argument against a set of candidate values that uses \eval to get the
-candidate values.} being responsible for over half of the recorded \eval
-calls. They discussed two use cases. The first one is the evaluation of an
-expression extracted from a promise using the \k{substitute} call in a new
-environment. The other is the invocation of a function with dynamically
-computed names and arguments.
+looked at the use of \eval in their corpus (1763 packages). They found 8500
+\eval call sites and recorded 2M \eval calls. Because they included the base
+libraries, their results were dominated by a few functions with
+\c{match.arg}\footnote{A base function that matches an argument against a set of
+candidate values, it uses \eval to get the candidate values.} being responsible
+for over half of the recorded \eval calls. They discussed two use cases. The
+first one is the evaluation of an expression extracted from a promise using the
+\k{substitute} call in a new environment. The other is the invocation of a
+function with dynamically computed names and arguments.
 

### 2021-08-12 paper/main.tex
PR 5.1

@@
 a loop.
-\\
+\end{tabular}
+
+\begin{tabular}{@{}p{.97\linewidth}}
 \medskip\EE{$min(e)=\c{F()}$}\\[-2mm]\small Function calls without variable lookups or
@@
 frequently evaluated in function environments.
-\end{tabular}
-
-\begin{tabular}{@{}p{.97\linewidth}}
+\\
   \medskip\EE{$min(e)=\c{FUN}$}\\[-2mm]\small Function definitions without any
@@
 \k{doParallel}), or deferring code execution (\eg \k{withr}).
-
 \\
@@
 \packageNbSimpleMinimizedMore sites get multiple simple shapes. Among these
-sites, there are many cases where \eval could be replaced by simpler constructs
+sites, there are many cases where \eval could be possibly replaced by simpler constructs
 such as \k{get} to lookup a name in an environment, \k{assign} to set a value to

### 2021-08-12 paper/main.tex
Proofread 5.2

--- 2021-08-12 paper/main.tex
@@
 writes act on local variables. But of which function? Table~\ref{tab:funoffset}
-gives the offset of that function on the call stack: 0 is function that called
-\eval, 1 is that function's caller, and so on. In 81\% of cases, \eval is
-resolved the caller -- the variables of the function where the call to \eval
+gives the offset of that function on the call stack: 0 is the function that called
+\eval, 1 is that function's caller, and so on. In 81\% of cases, \eval uses
+its caller's environment -- the variables of the function where the call to \eval
 textually occurs are read and written. About 1.5\% of sites evaluate their
@@
 transitively from that snippet. In \packageNbOneCategoryEnvirSitePercent of
-sites, only kind of environment is observed.
+the sites, only one kind of environment is observed.
 
@@
 
-Each synthetic environment has parent that is specified when calling
+Each synthetic environment has a parent that is specified when calling
 \k{new.env}. When \k{envir} is a list or a data frame, \eval uses its third
@@
 only \packageNbExplicitGlobalSites sites. Thus we suspect most uses of global
-are accidental. They arise from the fact that our corpus, consists of many code
-snippet that are run at top-level, global is thus the caller of \eval. This
+are accidental. They arise from the fact that our corpus consists of many code
+snippets that are run at the top-level, global is thus the caller of \eval. This is 
 noteworthy because writes to global variables are visible to all functions and
@@
 behavior of \eval within its enclosing function is to read and write variables,
-mostly read, and very rarely delete variables. There are also some cases where
+mostly read, and very rarely, delete variables. There are also some cases where
 new variables are injected in a function. Usually, this in the direct caller,
-but can sometimes take effect several frames up the call stack. There is
-something brittle about code relying on the position of a caller, many
-refactorings may break code that does that.
+but it can sometimes take effect several frames up the call stack. There is
+something brittle about code relying on the position of a caller, a lot of
+refactoring may break code that does that.
 
@@
 \k{substitute}, perhaps modified, and then evaluated with the environment coming
-from the same promise. We believe the case where \eval is provided the results
+from the same promise. We believe the case where \eval is provided with the results
 of programmatically selecting some call stack to be less frequent.
@@
 evaluation of an expression in a sandbox, or using a data structure as
-environment, for example one could evaluate a method of an object and use and
+environment. For example one could evaluate a method of an object and use an
 environment to hold the object's fields.
 
-We have already explained the popularity of global environments, as for package
+We have already explained the popularity of global environments. As for package
 environments, they are used in only \packageNbPackageNamespaceEnvSites sites.

### 2021-08-12 paper/main.tex
PR 5.3

@@
 Our analysis records the values returned by some functions of interest as well
-as their arguments. We build a direct acyclic graph that track the provenance of
-the values that are given to \eval. The following example yields the provenance
-graph with five nodes on Figure~\ref{fig:provgraph} for expression \k{e}.
-
+as their arguments. We build a direct acyclic graph that track the provenance
+of the values that are given to \eval. Fig~\ref{fig:provgraph} shows an example
+of the provenance graph for the expression \k{e}.
+The provenance of \k{e} includes both \k{parse} and \k{quote}. We pick an
+origin by traversing the graph from the root, following the left-hand side
+member of assignments until a leaf. It represents the provenance that is
+further modified to yield the expression passed to \eval. In our example, the
+origin is \k{parse}.
+
+\begin{figure}[H]
+  \begin{minipage}{.39\textwidth}
+    \vspace{-35mm}
 \begin{lstlisting}
  > e <- parse(text = "a;b")
- > e[[2]] <- quote(c)
+ > e[[2]] <- quote(c) # e is a;c
  > eval(e)
 \end{lstlisting}
-
-%\begin{figure}[h]
-%\centering
-%\includegraphics[width=8cm]{provenance-graph}
-%\caption{The provenance DAG for the code example above. Assignments and complex assignments are intermediate nodes, whereas expression builders are leaves of the graph.} \label{fig:provgraph}
-%\end{figure}
-
-\begin{figure}[h]
+\end{minipage}
+\begin{minipage}{.59\textwidth}
 	\centering
@@
 	\end{tikzpicture}
-\caption{Provenance} \label{fig:provgraph}
-\end{figure}
-
-The provenance of \k{e} includes both \k{parse} and \k{quote}. We pick an origin
-by traversing the graph from the root, following the left-hand side member of
-assignments until a leaf. It represents the provenance that is further modified
-to yield the expression passed to \eval. In our example, the origin is \k{parse}.
+\end{minipage}
+\caption{Example of a provenance graph} \label{fig:provgraph}
+\end{figure}
 
@@
 \caption{Origin} \label{tab:detailed_provenances}
-\medskip
+\end{wraptable}
+
+Table~\ref{tab:detailed_provenances} shows frequencies of provenance. For
+technical reasons, we misclassify a small number of sites. Manual inspection of
+numerous examples suggests that errors are rare. We further classify them into
+high-level categories:
+
+\begin{itemize}[---]
+\item {\it Reflection:} use of \k{match.call} to reflectively capture the
+  expression that invoked this function, arguments are promises, and
+  \k{substitute} is used to retrieve their source.
+\item {\it String:} create from strings by invoking \k{parse},
+  \k{str2lang}, or \k{str2expression}.
+\item {\it Constructed:} invoked \k{quote} or \k{expression}, or by built with
+  \k{call} or \k{as.call}.
+\item {\it Environment: } created with \k{as.name} or \k{as.symbol}, typically
+  to read a non-local environment.
+\item {\it External: }  call to a C function. % Put an example
+\end{itemize}
+
+\begin{wraptable}{r}{5cm}\small\centering
 \begin{tabular}{rrr} \toprule
@@
 \end{wraptable}
-
-Table~\ref{tab:detailed_provenances} shows frequencies of provenance. For
-technical reasons, we misclassify a small number of sites. Manual inspection of
-numerous examples suggests that errors are rare. We further classify them into
-high-level categories:
-
-\begin{itemize}[---]
-\item {\it Reflection:} use of \k{match.call} to reflectively capture the
-  expression that invoked this function, arguments are promises, and
-  \k{substitute} is used to retrieve their source.
-\item {\it String:} create from strings by invoking \k{parse},
-  \k{str2expression}, or \k{str2lang}.
-\item {\it Constructed:} invoked \k{quote} or \k{expression}, or by built with
-  \k{call} or \k{as.call}.
-\item {\it Environment: } created with \k{as.name} or \k{as.symbol}, typically
-  to read a non-local environment.
-\item {\it External: }  call to a C function. % Put an example
-\end{itemize}
 

### 2021-08-12 paper/main.tex
Fix page 7
Remove invokeRestartInteractively
Remove use of eval from 5.3

@@
 
+Our pipeline runs \CranRunnableScripts scripts extracted from \CranPackages
+packages. Any run that does not call \eval is discarded, leaving us with
+\packageNbruns runs. In these runs, \packageAllcalls \eval calls were recorded
+originating from \PkgHitEvalCallSites sites. The runs exercised
+\PkgHitEvalCallSitesAvgRatio of sites, a ratio similar to the package code
+coverage metric which is \PkgCodeCoverage. The reasons some sites are not
+exercised can be chalked down to incomplete tests and analysis failures
+(\PkgFailedProgramsRatio of the runs crashed or timed out).
+Fig.~\ref{fig:traced-eval-callsites} plots exercised site with respect to all
+sites for each program; as can be seen coverage is unequal.
+Table~\ref{tab:freq} summarizes the frequency of calls, call counts on the left and
+number of packages on the right. There are \packageFewcalls packages with fewer
+than 100 calls, and \packageManycalls packages with more than 1,000 calls.
+\packageMaxcallspack calls \eval \packageMaxcalls times and thus accounts for
+over half of our observations.
+
 \begin{figure}[H]
@@
 
-Our pipeline runs \CranRunnableScripts scripts extracted from \CranPackages
-packages. Any run that does not call \eval is discarded, leaving us with
-\packageNbruns runs. In these runs, \packageAllcalls \eval calls were recorded
-originating from \PkgHitEvalCallSites sites. The runs exercised
-\PkgHitEvalCallSitesAvgRatio of sites, a ratio similar to the package code
-coverage metric which is \PkgCodeCoverage. The reasons some sites are not
-exercised can be chalked down to incomplete tests and analysis failures
-(\PkgFailedProgramsRatio of the runs crashed or timed out).
-Fig.~\ref{fig:traced-eval-callsites} plots exercised site with respect to all
-sites for each program; as can be seen coverage is unequal.
-Table~\ref{tab:freq} summarizes the frequency of calls, call counts on the left and
-number of packages on the right. There are \packageFewcalls packages with fewer
-than 100 calls, and \packageManycalls packages with more than 1,000 calls.
-\packageMaxcallspack calls \eval \packageMaxcalls times and thus accounts for
-over half of our observations.
 Table~\ref{tab:variantseval} summarizes use of variants, the first row has sites
@@
 (\packageNbParseFromFileSites in total) consume the result of calling \k{parse}
-on a file. Most build strings programmatically. We identified one function
-\k{invokeRestartInteractively} that prompts the user for input and \evals it.
+on a file. Most build strings programmatically.
 
@@
 expressions from strings is also a common use case. The constructed category
-shows that most \evals comes from existing code that was modified by the
+shows that most \evals come from existing code that was modified by the
 programmer before invoking \eval. Both constructed and reflection categories
@@
 by macros if the R designers could be convinced to overcome their distaste for
-those. Another use of \eval is to directly build symbols to evaluate them in
-another environment.
+those.
 

### 2021-08-12 paper/main.tex
Fix another page

@@
 a loop.
-\end{tabular}
-
-\begin{tabular}{@{}p{.97\linewidth}}
+\\
 \medskip\EE{$min(e)=\c{F()}$}\\[-2mm]\small Function calls without variable lookups or
@@
 definitions nested in other expressions.
-\\
+\end{tabular}
+
+\begin{tabular}{@{}p{.97\linewidth}}
 \medskip\EE{$min(e)=\c{BLOCK}$}\\[-2mm]\small Multi-statement code blocks, as
@@
 \k{.Random.seed} variable is saved and restored from and to the global
-environment every time a statistical routine is called. Filtering accidental
+environment every time a statistical routine is called. Filtering out accidental
 side-effects leaves us with \SEUserRnd writes from \SEUserCallsRnd calls (in
-\SEUserSites sites and \SEUserFunctions functions). \SEFunsNighty functions are
+\SEUserSites sites and \SEUserFunctions functions). In this set, \SEFunsNighty functions are
 responsible for 90\% of side effects. Half of those come from just three

### 2021-08-12 paper/main.tex
Vertically center code in Figure 6
Fix rot and leaves

@@
 \begin{table}[!b]\small
-\hspace*{-0.5cm}
+\hspace*{-0.2cm}
 \begin{tabular}{crrrrrc}\toprule
-  \bf $min(e)$& \bf \#sites & \bf \%sites & \bf \#packages & \bf \#operations & \bf \%envir & \bf example\\\midrule
+  \bf $min(e)$& \bf \#sites & \bf \%sites & \bf \#packages & \bf \#ops & \bf \%envir & \bf example\\\midrule
 \k{X}&\packageMinimizedcallsitesa &\packageMinimizedpropsitesa &\packageMinimizedpackagea &\packageMinimizedmedianoperationsaRnd &\packageMinimizedpercentparentframesa & \k{y+1}\\
@@
 Our analysis records the values returned by some functions of interest as well
-as their arguments. We build a direct acyclic graph that track the provenance
+as their arguments. We build a direct acyclic graph that track the origin
 of the values that are given to \eval. Fig~\ref{fig:provgraph} shows an example
-of the provenance graph for the expression \k{e}.
-The provenance of \k{e} includes both \k{parse} and \k{quote}. We pick an
-origin by traversing the graph from the root, following the left-hand side
-member of assignments until a leaf. It represents the provenance that is
+of the origin graph for the expression \k{e}.
+The origin of \k{e} includes both \k{parse} and \k{quote}. We pick an
+origin by traversing the graph from the terminal leaf, \ie from the \eval node, following the left-hand side
+member of assignments until reaching a root node. It represents the origin that is
 further modified to yield the expression passed to \eval. In our example, the
@@
   \begin{minipage}{.39\textwidth}
-    \vspace{-35mm}
+    \vspace{-5mm}
 \begin{lstlisting}
@@
 \end{minipage}
-\caption{Example of a provenance graph} \label{fig:provgraph}
+\caption{Example of an origin graph} \label{fig:provgraph}
 \end{figure}
@@
 
-Table~\ref{tab:detailed_provenances} shows frequencies of provenance. For
+Table~\ref{tab:detailed_provenances} shows frequencies of origins. For
 technical reasons, we misclassify a small number of sites. Manual inspection of
@@
 
-Table~\ref{tab:provenance} summarizes provenance. Some sites are invoked with
-different provenances, they are counted multiple times. Strings correlate with
+Table~\ref{tab:provenance} summarizes origins. Some sites are invoked with
+different origins, they are counted multiple times. Strings correlate with
 dynamic code loading as done by \k{source} and \k{sys.source}. Few calls

### 2021-08-12 analysis/side-effects.Rmd
WIP: more SE analysis

@@
 library(digest)
+library(dplyr)
 library(fs)
@@
 library(purrr)
-library(tidyverse)
+library(readr)
+library(stringr)
+library(tidyr)
 library(viridis)
@@
 ```
+
+```{r}
+N <- read_csv(PACKAGE_NORMALIZED_EXPRESSION_FILE) %>%
+  select(minimized, expr_resolved_hash=hash)
+```
+
+```{r}
+E <- read_fst(PACKAGE_SUM_CALLS_FILE) %>% as_tibble()
+EN <- left_join(E, N, by="expr_resolved_hash")
+
+#ENS <- left_join(
+#  select(EN, eval_call_srcref, minimized, N1=nb_ev_calls), 
+#  select(SE, eval_call_srcref, in_envir, class, N2=N), 
+#  by="eval_call_srcref"
+#)
+```
+
+```{r}
+EN_sites <- group_by(EN, eval_call_srcref, minimized) %>% summarise(nb_ev_calls=sum(nb_ev_calls)) %>% ungroup
+cs_count(EN_sites, minimized)
+```
+
+```{r}
+cs_count(EN, wt=nb_ev_calls, minimized)
+```
+
+```{r}
+SEN <- left_join(SE, N, by="expr_resolved_hash")
+```

### 2021-08-12 paper/main.tex
provenance -> origin
origin

@@
 \end{lstlisting}
-\end{minipage}
-\begin{minipage}{.59\textwidth}
-	\centering
+  \end{minipage} \hspace{2cm}
+ \begin{minipage}{.4\textwidth}
+	\centering\scalebox{.7}{
 	\begin{tikzpicture}[
@@
 		\draw[<-, >=latex] (A) -- (P);
-	\end{tikzpicture}
+	\end{tikzpicture}}
 \end{minipage}
@@
 \end{tabular}
-\caption{Origin} \label{tab:detailed_provenances}
+
+\medskip
+
+\begin{tabular}{rrr} \toprule
+\bf Origin  & \bf \#sites & \bf \%sites \\\midrule
+Reflection &  \packageNbReflectionSites & \packageReflectionSitesPercent\\
+String & \packageNbStringSites & \packageStringSitesPercent \\
+Constructed & \packageNbConstructedSites & \packageConstructedSitesPercent \\
+Environment & \packageNbSymbolSites & \packageSymbolSitesPercent \\
+External & \packageNbExternalSites & \packageExternalSitesPercent \\\bottomrule
+\end{tabular}
+\caption{Origins}\label{tab:provenance}\vspace{-1cm}
 \end{wraptable}
 
-Table~\ref{tab:detailed_provenances} shows frequencies of origins. For
-technical reasons, we misclassify a small number of sites. Manual inspection of
-numerous examples suggests that errors are rare. We further classify them into
-high-level categories:
+To have a high-level views of provenance we classify functions into the
+following five categories:
 
@@
 
-\begin{wraptable}{r}{5cm}\small\centering
-\begin{tabular}{rrr} \toprule
-\bf Origin  & \bf \#sites & \bf \%sites \\\midrule
-Reflection &  \packageNbReflectionSites & \packageReflectionSitesPercent\\
-String & \packageNbStringSites & \packageStringSitesPercent \\
-Constructed & \packageNbConstructedSites & \packageConstructedSitesPercent \\
-Environment & \packageNbSymbolSites & \packageSymbolSitesPercent \\
-External & \packageNbExternalSites & \packageExternalSitesPercent \\\bottomrule
-\end{tabular}
-\caption{Provenance}\label{tab:provenance}
-\end{wraptable}
-
-Table~\ref{tab:provenance} summarizes origins. Some sites are invoked with
-different origins, they are counted multiple times. Strings correlate with
-dynamic code loading as done by \k{source} and \k{sys.source}. Few calls
-(\packageNbParseFromFileSites in total) consume the result of calling \k{parse}
-on a file. Most build strings programmatically.
+Table~\ref{tab:provenance} summarizes origins. Is shows frequencies by function
+and by category. For technical reasons, we misclassify a small number of sites.
+Manual inspection of numerous examples suggests that errors are rare. Some sites
+are counted multiple times as they are invoked with different origins. Strings
+correlate with dynamic code loading as done by \k{source} and \k{sys.source}.
+Few calls (\packageNbParseFromFileSites in total) consume the result of calling
+\k{parse} on a file. Most build strings programmatically.
 
 \paragraph{Discussion}
-The main use case for \eval involves using \k{match.call}, \k{substitute}, or
+The main use cases for \eval involve using \k{match.call}, \k{substitute}, or
 \k{formals} to access and transform arguments of a function. Constructing
@@
 programmer before invoking \eval. Both constructed and reflection categories
-roughly correspond to meta-programming. Some use cases could likely be replaced
-by macros if the R designers could be convinced to overcome their distaste for
+roughly correspond to meta-programming. Some uses could be replaced by
+macros if the R designers could be convinced to overcome their distaste for
 those.
@@
 
-\begin{wraptable}{r}{7cm}\small\centering
+\begin{table}[!h]
   \begin{tabular}{lrrrr}
@@
   \caption{Target environments for side-effects} \label{tab:se-env}
-\end{wraptable}
+\end{table}
 

### 2021-08-12 paper/main.tex
use
Update 6.codegen

@@
 
-\begin{figure}[h!]
+\begin{minipage}{.95\textwidth}
+\medskip\underline{\c{AdapSamp::rARS}:}
 \begin{lstlisting}
@@
     p <- function(x) eval(parse(text=formula))
-\end{lstlisting} \caption{AdapSamp::rARS}
-\end{figure}
+\end{lstlisting}\medskip
+\end{minipage}
 
@@
 write by hand or that is only known after deployment. The example uses \eval to
-construct function objects that are parameterized with string \k{what}.
-This is used  automatically generate R bindings to native code.
-
-\begin{figure}[h]
+automatically generate R bindings to C++ methods.
+
+\begin{minipage}{.95\textwidth}
+\medskip\underline{\c{Rcpp::.makeCppMethods}:}
 \begin{lstlisting}
@@
    for(what in cppMethods)
-     methods[[what]] <- eval(substitute(function(...)CppObject$WHAT(...,name=what)),env)
-\end{lstlisting}
-  \caption{Rcpp::.makeCppMethods}
-\end{figure}
+     methods[[what]] <-
+       eval(substitute(function(...)CppObject$what(...)), env)
+\end{lstlisting}\medskip
+\end{minipage}
 
@@
 expended with the arguments supplied to the function. \Eval then returns a
-\k{list} with unevaluated argument expressions which are evaluated
-sequentially for their result.
-
-\begin{figure}[h]
+\k{list} with unevaluated argument expressions which are evaluated sequentially
+for their result.
+
+\begin{minipage}{.95\textwidth}
+\medskip\underline{\c{statnet.common::NVL}:}
 \begin{lstlisting}
@@
  }
-\end{lstlisting}
-  \caption{statnet.common::NVL}
-\end{figure}
+\end{lstlisting}\medskip
+\end{minipage}
 
@@
 
-\begin{figure}[h]
+\begin{minipage}{.95\textwidth}
+  \medskip\underline{\c {assertthat::see\_if}:}
 \begin{lstlisting}
@@
       }, assertError=function(e) { structure(FALSE, msg=e$message) })
-\end{lstlisting}
-  \caption{assertthat::see\_if}
-\end{figure}
+\end{lstlisting}\medskip
+\end{minipage}
 
@@
 
-\begin{figure}[h]
+\begin{minipage}{.95\textwidth}
+  \medskip\underline{\c{igraph::do\_call}:}
 \begin{lstlisting}
@@
  }
-\end{lstlisting}
-\caption{igraph::do\_call}
-\end{figure}
+\end{lstlisting}\medskip
+\end{minipage}
 
@@
 
-\begin{figure}[h]
+\begin{minipage}{.95\textwidth}
+  \medskip\underline{\c{glue::glue}:}
 \begin{lstlisting}
@@
  #> Hello World!
-\end{lstlisting}
-  \caption{glue::glue}
-\end{figure}
+\end{lstlisting}\medskip
+\end{minipage}
 
@@
 
-\begin{figure}[h]
+\begin{minipage}{.95\textwidth}
+  \medskip\underline{\c{survival::survfit.formula}:}
 \begin{lstlisting}
@@
    mf <- eval.parent(temp)
-\end{lstlisting}
-  \caption{survival::survfit.formula}
-\end{figure}
+\end{lstlisting}\medskip
+\end{minipage}
 
@@
 
-\begin{figure}[h]
+\begin{minipage}{.95\textwidth}
+  \medskip\underline{\c{nlme::formula.pdMat}:}
 \begin{lstlisting}
  function(x, asList, ...) eval(attr(x, "formula"))
-\end{lstlisting}
-  \caption{nlme::formula.pdMat}
-\end{figure}
+\end{lstlisting}\medskip
+\end{minipage}
 
@@
 
-\begin{figure}[h]
+\begin{minipage}{.95\textwidth}
+  \medskip\underline{\c{aibd::scalaEnsure}:}
 \begin{lstlisting}
@@
    lockBinding("s", env)
-\end{lstlisting}
-  \caption{aibd::scalaEnsure}
-\end{figure}
+\end{lstlisting}\medskip
+\end{minipage}
 

### 2021-08-12 paper/main.tex
Update 6.extract varargs

@@
 \paragraph{Extracting Varargs}
-A somewhat odd use of \eval is to obtain the source code of the arguments passed
-to a vararg parameter. In this example, the function iterates over the contents
-of the vararg until it finds a non-null value. \k{substitute(alist(...))} yields
-an expression that represents a call to \k{alist} but where the \k{...} are
+\Eval is to obtain the source code of the arguments passed to a vararg
+parameter. One example is a function that iterates over the contents of the
+vararg until it finds a non-null value. \k{substitute(alist(...))} yields an
+expression that represents a call to \k{alist} but where the \k{...} are
 expended with the arguments supplied to the function. \Eval then returns a
 \k{list} with unevaluated argument expressions which are evaluated sequentially
-for their result.
+for their result. The reason \eval is needed is not to force
+any more arguments once a non-null value was found.
 

### 2021-08-12 paper/main.tex
Update 6.control

@@
 if they fail, produce legible error messages. The example here accepts a vararg
-with a list of expressions whose code is extracted using
-\k{eval(substitute(alist(...)))}. The expression are evaluated sequentially
-in the caller's environment. If an expression is false, an error message is
+with a list of expressions. It extracts the associated code in the same way as
+the extracting varags example and evaluates each of the expression sequentially
+in the caller's environment. If an expression does not hold, an error is
 generated.

### 2021-08-12 paper/main.tex
Update 6.dsl

@@
 some suitable way. The usage example illustrates string interpolation. The
-\k{glue} function extracts snippets of code enclosed between braces and
-evaluates them using \eval, and finally splices their results to construct the
-result.
+\k{glue} function from the \k{glue} package extracts snippets of code enclosed
+between braces and evaluates them using \eval, and finally splices their
+results to construct the result.
 
 \begin{minipage}{.95\textwidth}
-  \medskip\underline{\c{glue::glue}:}
-\begin{lstlisting}
- library(glue)
- greeting <- "Hello"
- glue('{greeting} World!')
- #> Hello World!
+\medskip
+\begin{lstlisting}
+> greeting <- "Hello"
+> glue('{greeting} World!') 
+# Hello World!
 \end{lstlisting}\medskip

### 2021-08-12 paper/main.tex
Update 6.obfus
Add 6.nss

@@
 
+\paragraph{Non-standard Scoping}
+
+\Eval allows functions to resolve symbols in different environments that the
+current one. For example, the \k{data.table} redefines the subsetting
+operator\footnote{It defines it as a triplet \texttt{[i, j, k]}, where expression
+  \texttt{j} is used to calculate a subset of rows from \texttt{i} grouped by \texttt{k}.}
+for data frames to enable compact queries. The following  
+computes the average departure delays on 2014 Delta flights for each
+origin airport.
+
+\begin{minipage}{.95\textwidth}
+\medskip
+\begin{lstlisting}
+> flights[carrier=="DL" & year == 2014, mean(dep_delay), by=origin]
+#    origin       V1
+# 1:    LGA 11.35083
+# 2:    EWR 14.96171
+# 3:    JFK 12.45414
+\end{lstlisting}\medskip
+\end{minipage}
+
+At its core, the \k{[} operator uses \k{substitute} to capture its operand
+expressions, transforms them, and evaluates them using \eval in the given data
+frame.\footnote{We omit the code as the implementation spans over 1700 lines,\\
+  \indent \cf
+\url{https://github.com/Rdatatable/data.table/blob/master/R/data.table.R}} The
+symbols \k{carrier}, \k{year}, \k{dep_delay} and \k{origin} are not looked up
+in the current environment, but instead in the \k{flights} data frame.
+
+The non-standard scoping is an important use case in R. It allows ergonomic
+functions for interactive data exploration. The ergonomics come from the
+reduced typing and syntactic noise at the cost of some ambiguity. Since R is
+used primarily interactively, this trade-off is justifiable.
+
 \paragraph{Obfuscation}
-Some packages use \eval to bypass the restrictions imposed by the \k{R CMD
+A few packages use \eval to bypass the restrictions imposed by the \k{R CMD
   CHECK} tool. This tool enforces some well-formedness rules on a submitted

### 2021-08-13 paper/main.tex
Update 6.boilerplate
Update 6.boilderplate

@@
 \paragraph{Boilerplate Removal} 
-A variant of code generation is to let users write compact code that is expended
-via \eval and \k{match.call}. The example uses \k{match.call} to access
-the expression with which the function is called. The next three lines construct
-a call to \k{stats::model.frame} with argument text of parameters
-\k{'formula'}, \k{'data'}, and \k{'weights'}. Then, the call expression
-is evaluated in the parent environment. The use of \k{match.call}, avoids
-repeated use of \k{substitute} for extracting the argument expressions. This
-setup is also extensible as new parameter names can be easily added.
+A variant of code generation is to let users write compact code that is
+expended via \eval and \k{match.call}. The example uses \k{match.call} to
+access the expression with which the function is called. The next three lines
+construct a call to \k{stats::model.frame} with argument text of parameters
+\k{'formula'}, \k{'data'}, and \k{'weights'}. Then, the call expression is
+evaluated in the parent environment. The use of \k{match.call}, avoids repeated
+use of \k{substitute} for extracting the argument expressions. It is extensible
+as new parameter names can be easily added. This example illustrates one of the
+most recurring usage of \eval that is used by the majority of packages that do
+statistical modeling.
 
@@
 \end{minipage}
+
+This combination of \k{match.call} and \eval to pass most of the call to
+another function is also used to created wrappers. For example, the
+\k{write.csv} uses it to forward most of the parameters to the more general
+\k{write.table} function, only setting the CSV-related parameters:
+
+\begin{minipage}{.95\textwidth}
+  \medskip\underline{\c{base::write.csv}:}
+\begin{lstlisting}
+function (...)
+{
+    Call <- match.call(expand.dots = TRUE)
+    Call$sep <- ","
+    Call$dec <- "."
+    Call$qmethod <- "double"
+    Call[[1L]] <- as.name("write.table")
+    eval.parent(Call)
+}
+\end{lstlisting}\medskip
+\end{minipage}
+
+The most probable reason for this is to ease maintenance of such
+wrappers.\footnote{This was speculated by the Brian Ripley, the author of the
+function, in a private conversation.} Wrapping functions in R is somewhat
+harder than in other languages because of the default arguments, the varargs
+and is missing arguments.
 

### 2021-08-13 paper/main.tex
Fix indetantion of code in 6
Update 6.shadowing

@@
 Many packages use \eval to implement variations of the \k{base} R functions.
-For example, the \k{igraph} package provides the \k{do_call} function
-shown below. This is a variant of the \k{do.call} function from the
-\k{base} package. This function executes a function call by constructs it
-from a name and a list of arguments and calling \eval.
+For example, the \k{igraph} package provides the \k{do_call} function shown
+below. This is a variant of the builtin \k{do.call} function with the
+additional support for specifying the environment where the call should happen.
+It constructs a call from a name and a list of arguments, and pass it to \eval.
 

### 2021-08-13 paper/main.tex
Capitalize subsections for consistency
PR 7

@@
 
-\subsection{The expression in \eval} \label{sec:minimized}
+\subsection{The Expression in \Eval} \label{sec:minimized}
 
@@
 
-\subsection{The environments of \eval}\label{sec:env}
+\subsection{The Environments of \Eval}\label{sec:env}
 
@@
 
-\subsection{The origins of \eval}
+\subsection{The Origins of \Eval}
 
@@
 
-\subsection{The effects of \eval}
+\subsection{The Effects of \Eval}
 
@@
 can be made for assignment and the \k{assign} function. There is a dedicated
-\k{do.call} function to build function calls. Unnecessary uses of \eval
-often can be uncovered by manual analysis.
-
-\begin{lstlisting}
+\k{do.call} function to build function calls. However, the unnecessary uses of
+\eval often can only be uncovered by manual analysis.
+
+\begin{lstlisting}[caption={Unnecessary use of \eval},label=lst:unnecessary]
  PerformanceAnalytics::chart.QQPlot <- function (d="norm",dp,...) {
-   q.f <- eval(parse(text=paste("q",d)))
+   q.f <- eval(parse(text=paste0("q",d)))
    z <- NULL
-   eval(parse(text=paste("z<-q.f(",dp,",...)")))
-  }
+   eval(parse(text=paste0("z<-q.f(",dp,",...)")))
 \end{lstlisting}
 
-The above uses \eval to first lookup a target function name from a constructed
-string, the it constructs a call to the target that include arguments to the
-current functions. \Eval is overkill for this code, a semantically equivalent
-function body uses \k{get} to perform the lookup and then calls directly.
-
-\begin{lstlisting}
- function (R, d="norm", dp, ...) {
+For example in Listing~\ref{lst:unnecessary}, the function uses \eval to lookup
+a target function name from a constructed string. Then it constructs a calls it
+with the arguments that are passed to the current function. \Eval is overkill
+in this case, a semantically equivalent function body uses \k{get} to perform
+the lookup and then calls the resulting function directly.
+
+\begin{lstlisting}
  q.f <- get(paste0("q",d))
  z <- q.f(dp, ...)
- }
 \end{lstlisting}
@@
 
+% TODO: show how will it look with symbolic field read and write
+
+% TODO: remove - does not bring anything new or interesting
 As a last example, \k{configr::config.funs.par} construct three expressions
@@
 We have not yet touched on alternative implementations of \eval. The tidyverse
-libraries include their own implementation calls
-\k{rlang::eval\_tidy}~\cite{tidyverse}. This departs from \eval in two ways.
-First, it supports evaluation of quosures, custom objects used by for
-metaprogramming that bundle expressions with an environment. \k{eval\_tidy}
-accepts a data mask argument, a set of bindings such as a dataframe or list,
-which takes precedence over the environment. Effects such as assignments happen
-in the data mask. Expressions such as \k{return()} do not work since the data
-mask does not correspond to a frame on the call stack.
+libraries\footnote{An opinionated collection of R packages designed for data
+science. It consists of some of the most popular packages.} include their own
+implementation calls \k{rlang::eval\_tidy}~\cite{tidyverse}. This departs from
+\eval in two ways. First, it supports evaluation of quosures, custom objects
+used by for metaprogramming that bundle expressions with an environment.
+\k{eval\_tidy} accepts a data mask argument, a set of bindings such as a
+dataframe or list, which takes precedence over the environment. Effects such as
+assignments happen in the data mask. Expressions such as \k{return()} do not
+work since the data mask does not correspond to a frame on the call stack.
 

### 2021-08-13 paper/main.tex
Few fixes in 6

@@
 manually inspected top 117 sites which contribute to 90\% of the calls in the
-corpus. This section gives some real-world examples, we simplify code for
-clarity.
+corpus. In this section we show 10 real-world examples form this batch. The
+goal is to illustrate the variety of use cases that employ \eval in the R
+ecosystem. The code snippets are simplified for clarity.
 
@@
 \paragraph{Extracting Varargs}
-\Eval is to obtain the source code of the arguments passed to a vararg
+\Eval can be used to obtain the source code of the arguments passed to a vararg
 parameter. One example is a function that iterates over the contents of the
@@
 \paragraph{Controlled Evaluation}
-Developer often want to control when and if expressions are evaluated. For
-example, a package that provides users with the ability to make assertions and,
+\Eval can be used to control when and if expressions are evaluated. For
+example, \k{assertthat} provides users with the ability to make assertions and,
 if they fail, produce legible error messages. The example here accepts a vararg
-with a list of expressions. It extracts the associated code in the same way as
-the extracting varags example and evaluates each of the expression sequentially
-in the caller's environment. If an expression does not hold, an error is
-generated.
+with a list of expressions. It extracts the associated code in the same way
+as the extracting varags example and evaluates each of the expression
+sequentially in the caller's environment. If an expression does not hold, an
+error is generated.
 
@@
 \paragraph{Shadowing}
-%%% TODO ?????????????////
+%%% TODO ?????????????//// <- left by Jan
 Many packages use \eval to implement variations of the \k{base} R functions.
@@
  }
-\end{lstlisting}\medskip
-\end{minipage}
-
-\paragraph{Domain-Specific Language}
-A domain-specific language leverages R's grammar but redefines it semantics in
-some suitable way. The usage example illustrates string interpolation. The
-\k{glue} function from the \k{glue} package extracts snippets of code enclosed
-between braces and evaluates them using \eval, and finally splices their
-results to construct the result.
-
-\begin{minipage}{.95\textwidth}
-\medskip
-\begin{lstlisting}
-> greeting <- "Hello"
-> glue('{greeting} World!') 
-# Hello World!
 \end{lstlisting}\medskip
@@
 most recurring usage of \eval that is used by the majority of packages that do
-statistical modeling.
+statistical modeling, \ie the \c{model.frame} shape in Table~\ref{tab:minimizedexpressions}.
 
@@
 
+% TODO: update to explain why the eval is needed or remove
 \paragraph{Formula Resolution}
@@
 
-\Eval allows functions to resolve symbols in different environments that the
+\Eval allows functions to resolve symbols in different environments than the
 current one. For example, the \k{data.table} redefines the subsetting
@@
 used primarily interactively, this trade-off is justifiable.
+
+\paragraph{Domain-Specific Language}
+A domain-specific language leverages R's grammar but redefines it semantics in
+some suitable way. The usage example illustrates string interpolation. The
+\k{glue} function from the \k{glue} package extracts snippets of code enclosed
+between braces and evaluates them using \eval, and finally splices their
+results to construct the result.
+
+\begin{minipage}{.95\textwidth}
+\medskip
+\begin{lstlisting}
+> greeting <- "Hello"
+> glue('{greeting} World!') 
+# Hello World!
+\end{lstlisting}\medskip
+\end{minipage}
+
+The \k{glue} package contains a variant that leverages the non-standard scoping
+as well allowing to resolve interpolation in alternative environments. For
+example, the following snippet pipes the subsetted data frame into the
+\k{glue_data} function that uses it to resolve the symbols in string
+interpolation.
+
+\begin{minipage}{.95\textwidth}
+\medskip
+\begin{lstlisting}
+> flights[carrier=="DL" & year == 2014, mean(dep_delay), by=origin] |> 
+    glue_data("{origin} has average delay of {round(V1)} min")
+# LGA has average delay of 11 min
+# EWR has average delay of 15 min
+# JFK has average delay of 12 min
+\end{lstlisting}\medskip
+\end{minipage}
 

### 2021-08-13 paper/main.tex
Fixes

@@
 manually inspected top 117 sites which contribute to 90\% of the calls in the
-corpus. In this section we show 10 real-world examples form this batch. The
+corpus. In this section we show 10 real-world examples from this batch. The
 goal is to illustrate the variety of use cases that employ \eval in the R
@@
 
-\paragraph{Extended Interfaces}
-Sometimes it is convenient to allow users to extend the behavior of a library
-function. \Eval allows clients to provide a simple text string that shall be
-executed in the context of the function. The example here allows users to
-provide formulas as strings, such as \k{"exp(-x^2/2)"}, the contract being that
-the text can refer to some variable \k{x} that will be set up by the called
-function. In this case one users could provide an anonymous function.
+\paragraph{Extended Interfaces} Sometimes it is convenient to allow users to
+extend the behavior of a library function. \Eval allows clients to provide a
+simple text string that shall be executed in the context of the function. The
+example here allows users to provide formulas as strings, such as
+\k{"exp(-x^2/2)"}, the contract being that the text can refer to some variable
+\k{x} that will be set up by the called function.\footnote{The function is used
+to generates a sequence of random numbers using the adaptive rejection sampling
+algorithm and the string formula is used to describe the target density.} In
+this case users could provide an anonymous function.
 
@@
 The \k{glue} package contains a variant that leverages the non-standard scoping
-as well allowing to resolve interpolation in alternative environments. For
+as well allowing to resolve interpolation in an alternative environment. For
 example, the following snippet pipes the subsetted data frame into the

### 2021-08-13 paper/main.tex
Clarify 5.3

@@
 of the origin graph for the expression \k{e}.
-The origin of \k{e} includes both \k{parse} and \k{quote}. We pick an
-origin by traversing the graph from the terminal leaf, \ie from the \eval node, following the left-hand side
-member of assignments until reaching a root node. It represents the origin that is
+The origin of \k{e} includes both \k{parse} and \k{quote}, the starting nodes. We pick one origin by traversing the graph from the terminal leaf, \ie from the \eval node, following the left-hand side
+member of assignments until reaching a starting node. It represents the origin that is
 further modified to yield the expression passed to \eval. In our example, the
@@
 \item {\it Reflection:} use of \k{match.call} to reflectively capture the
-  expression that invoked this function, arguments are promises, and
+  expression that invoked this function. Function arguments are promises, and
   \k{substitute} is used to retrieve their source. 
@@
   \k{str2lang}, or \k{str2expression}.
-\item {\it Constructed:} invoked \k{quote} or \k{expression}, or by built with
+\item {\it Constructed:} invoked \k{quote} or \k{expression}, or by building with
   \k{call} or \k{as.call}.
@@
   to read a non-local environment.
-\item {\it External: }  call to a C function. % Put an example
+\item {\it External: }  call to a C function. 
 \end{itemize}
 
-Table~\ref{tab:provenance} summarizes origins. Is shows frequencies by function
+Table~\ref{tab:provenance} summarizes origins. It shows frequencies by function
 and by category. For technical reasons, we misclassify a small number of sites.
@@
 \paragraph{Discussion}
-The main use cases for \eval involve using \k{match.call}, \k{substitute}, or
-\k{formals} to access and transform arguments of a function. Constructing
+The main use cases for \eval involve using \k{match.call}, or \k{substitute} to access and transform arguments of a function. Constructing
 expressions from strings is also a common use case. The constructed category

### 2021-08-13 paper/main.tex
Some fixes on Section 7

@@
 can be made for assignment and the \k{assign} function. There is a dedicated
-\k{do.call} function to build function calls. However, the unnecessary uses of
+\k{do.call} function to perform function calls. However, the unnecessary uses of
 \eval often can only be uncovered by manual analysis.
@@
 with the arguments that are passed to the current function. \Eval is overkill
-in this case, a semantically equivalent function body uses \k{get} to perform
-the lookup and then calls the resulting function directly.
+in this case: a semantically equivalent function body can use \k{get} to perform
+the lookup and then call the resulting function directly.
 
@@
 % TODO: remove - does not bring anything new or interesting
-As a last example, \k{configr::config.funs.par} construct three expressions
+As a last example, \k{configr::config.funs.par} constructs three expressions
 that are then fed to \eval. Semantically equivalent code would use \k{get}

### 2021-08-13 paper/main.tex
Section 7: symbolic field reading and writing
Fixes

@@
 manually inspected top 117 sites which contribute to 90\% of the calls in the
-corpus. In this section we show 10 real-world examples from this batch. The
+corpus. In this section we show 9 real-world examples from this batch. The
 goal is to illustrate the variety of use cases that employ \eval in the R
@@
 
-\paragraph{Extended Interfaces} Sometimes it is convenient to allow users to
+\paragraph{Behavioral Extension} Sometimes it is convenient to allow users to
 extend the behavior of a library function. \Eval allows clients to provide a
@@
 algorithm and the string formula is used to describe the target density.} In
-this case users could provide an anonymous function.
+this case users could provide an anonymous function instead.
 
@@
 
-This combination of \k{match.call} and \eval to pass most of the call to
-another function is also used to created wrappers. For example, the
-\k{write.csv} uses it to forward most of the parameters to the more general
-\k{write.table} function, only setting the CSV-related parameters:
+This combination of \k{match.call} and \eval for passing most of the current
+call's argument to another function is also used to created wrappers. For
+example, the \k{write.csv} uses it to forward most of the parameters to the
+more general \k{write.table} function, only setting the CSV-related
+parameters:
 
@@
 function, in a private conversation.} Wrapping functions in R is somewhat
-harder than in other languages because of the default arguments, the varargs
-and is missing arguments.
-
-% TODO: update to explain why the eval is needed or remove
-\paragraph{Formula Resolution}
-Many packages use \eval to evaluate formulas. The example extracts the
-\k{"formula"} attribute attached to the \k{pdMat} object and evaluates it.
-
-\begin{minipage}{.95\textwidth}
-  \medskip\underline{\c{nlme::formula.pdMat}:}
-\begin{lstlisting}
- function(x, asList, ...) eval(attr(x, "formula"))
-\end{lstlisting}\medskip
-\end{minipage}
+harder than in other languages because of the default argument values, the
+varargs and the missing arguments.
+
+% % TODO: update to explain why the eval is needed or remove
+% \paragraph{Formula Resolution}
+% Many packages use \eval to evaluate formulas. The example extracts the
+% \k{"formula"} attribute attached to the \k{pdMat} object and evaluates it.
+
+% \begin{minipage}{.95\textwidth}
+%   \medskip\underline{\c{nlme::formula.pdMat}:}
+% \begin{lstlisting}
+%  function(x, asList, ...) eval(attr(x, "formula"))
+% \end{lstlisting}\medskip
+% \end{minipage}
 
@@
 The \k{glue} package contains a variant that leverages the non-standard scoping
-as well allowing to resolve interpolation in an alternative environment. For
-example, the following snippet pipes the subsetted data frame into the
-\k{glue_data} function that uses it to resolve the symbols in string
-interpolation.
+allowing to resolve interpolation in an alternative environment. For example,
+the following snippet pipes the subsetted data frame into the \k{glue_data}
+function that uses it to resolve the symbols in string interpolation.
 
@@
 can be made for assignment and the \k{assign} function. There is a dedicated
-\k{do.call} function to perform function calls. However, the unnecessary uses of
-\eval often can only be uncovered by manual analysis.
+\k{do.call} function to perform function calls. However, the unnecessary uses
+of \eval are not easy to find and often can only be uncovered by manual
+analysis. Dynamic analysis is limited due to coverage issues, and static
+analysis of R remains an open problem.
 
@@
 Another example, from \k{coin::copyslots}, constructs an expression that access
-fields of an object and copy them. This can be replaced by symbolic field reads
-and writes.
+fields of an object and copy them.
 
@@
 
+This can be replaced by symbolic field reads
+and writes:
+
+% This should be enough:
+\begin{lstlisting}
+ slot(target, s) <- slot(source, s)
+\end{lstlisting}
+
 % TODO: remove - does not bring anything new or interesting
-As a last example, \k{configr::config.funs.par} constructs three expressions
-that are then fed to \eval. Semantically equivalent code would use \k{get}
-to lookup \k{fun}, and then directly execute the three expressions.
-
-\begin{lstlisting}
- config.funs.par <- function(fun="",...) {
-   args.all <- as.list(match.call())
-   args.all <- args.all[names(args.all) != ""]
-   parameters <- list()
-   t1 <- sprintf("parameters <- names(as.list(args(%s)))", fun)
-   t2 <- "parameters <- parameters[parameters != '']"
-   t3 <- "parameters <- args.all[names(args.all) %in% parameters]"
-   for (j in c(t1,t2,t3))
-    eval(parse(text = j))
-   return(parameters)
- }
-\end{lstlisting}
-
-The challenge for automation is that such examples are not easy to find. Dynamic
-analysis is limited due to coverage issues, and static analysis of R remains an
-open problem.
+% As a last example, \k{configr::config.funs.par} constructs three expressions
+% that are then fed to \eval. Semantically equivalent code would use \k{get}
+% to lookup \k{fun}, and then directly execute the three expressions.
+
+% \begin{lstlisting}
+%  config.funs.par <- function(fun="",...) {
+%    args.all <- as.list(match.call())
+%    args.all <- args.all[names(args.all) != ""]
+%    parameters <- list()
+%    t1 <- sprintf("parameters <- names(as.list(args(%s)))", fun)
+%    t2 <- "parameters <- parameters[parameters != '']"
+%    t3 <- "parameters <- args.all[names(args.all) %in% parameters]"
+%    for (j in c(t1,t2,t3))
+%     eval(parse(text = j))
+%    return(parameters)
+%  }
+% \end{lstlisting}
 

### 2021-08-13 paper/main.tex
Typos in conclusion

@@
 
-The \eval functions used widely and in varied ways in R. The function is an
+The \eval functions are used widely and in varied ways in R. The function is an
 essential tool for language implementers. Our analysis observed that the base
@@
 code written by less experienced users suggest that \eval is exceedingly rarely
-used. So, it is fair to conclude that majority of R programmers never have to
+used. So, it is fair to conclude that a majority of R programmers never have to
 write a call to \eval, but that the code they write would not run without it.

### 2021-08-13 paper/main.tex
Fix eval usage

@@
 manually inspected top 117 sites which contribute to 90\% of the calls in the
-corpus. In this section we show 9 real-world examples from this batch. The
-goal is to illustrate the variety of use cases that employ \eval in the R
-ecosystem. The code snippets are simplified for clarity.
+corpus. We present examples from nine classes from this batch. The
+classification is not exhaustive; we only study 117 call sites from the
+\PkgEvalCallSites call sites in our corpus and many uses of \eval do not fit in
+a specific category. The goal here is to demonstrate the wide variety of uses to
+which \eval is put to use. This also speaks to the ubiquity of \eval in the R
+ecosystem. The code snippets have been simplified for clarity.
 
@@
 \paragraph{Code Generation}
-\Eval can be used to generate boiler plate code that would either be tedious to
+\Eval can be used to generate boilerplate code that would either be tedious to
 write by hand or that is only known after deployment. The example uses \eval to
@@
 expended with the arguments supplied to the function. \Eval then returns a
-\k{list} with unevaluated argument expressions which are evaluated sequentially
-for their result. The reason \eval is needed is not to force
-any more arguments once a non-null value was found.
+\k{list} with the unevaluated argument expressions.
 
@@
  function(...) {
-   for(e in eval(substitute(alist(...)))) {
-     x <- eval(e, parent.frame())
-     if(!is.null(x)) break
-   }
-   x
- }
+   exprs <- eval(substitute(alist(...)))
 \end{lstlisting}\medskip
@@
 \paragraph{Boilerplate Removal} 
-A variant of code generation is to let users write compact code that is
-expended via \eval and \k{match.call}. The example uses \k{match.call} to
-access the expression with which the function is called. The next three lines
-construct a call to \k{stats::model.frame} with argument text of parameters
-\k{'formula'}, \k{'data'}, and \k{'weights'}. Then, the call expression is
-evaluated in the parent environment. The use of \k{match.call}, avoids repeated
-use of \k{substitute} for extracting the argument expressions. It is extensible
-as new parameter names can be easily added. This example illustrates one of the
-most recurring usage of \eval that is used by the majority of packages that do
-statistical modeling, \ie the \c{model.frame} shape in Table~\ref{tab:minimizedexpressions}.
+A variant of code generation is to let users write compact code that is expended
+via \eval and \k{match.call}. The example uses \k{match.call} to access the
+expression with which the function is called. The next three lines construct a
+call to \k{stats::model.frame} with argument text of parameters \k{'formula'},
+\k{'data'}, and \k{'weights'}. Then, the call expression is evaluated in the
+parent environment. The use of \k{match.call} avoids repeated use of
+\k{substitute} for extracting the argument expressions. It is extensible as new
+parameter names can be easily added. This example illustrates one of the most
+recurring uses of \eval by packages that do statistical modeling, \ie the
+\c{model.frame} shape in Table~\ref{tab:minimizedexpressions}.
 
@@
 The most probable reason for this is to ease maintenance of such
-wrappers.\footnote{This was speculated by the Brian Ripley, the author of the
-function, in a private conversation.} Wrapping functions in R is somewhat
-harder than in other languages because of the default argument values, the
-varargs and the missing arguments.
+wrappers.\footnote{Private communication with a developer of R.} Wrapping
+functions in R is somewhat harder than in other languages because of default
+arguments, varargs and missing arguments.
 

### 2021-08-13 paper/main.tex
Fix grammar issues and turn on review mode

@@
-\documentclass[screen,acmsmall]{acmart}%\settopmatter{printfolios=false,printccs=false,printacmref=false}
+\documentclass[review,screen,acmsmall]{acmart}%\settopmatter{printfolios=false,printccs=false,printacmref=false}
 \setcopyright{none}
@@
 with a list of expressions. It extracts the associated code in the same way
-as the extracting varags example and evaluates each of the expression
+as the extracting varags example and evaluates the expressions
 sequentially in the caller's environment. If an expression does not hold, an
@@
 %%% TODO ?????????????//// <- left by Jan
-Many packages use \eval to implement variations of the \k{base} R functions.
+Some packages use \eval to implement variations of the \k{base} R functions.
 For example, the \k{igraph} package provides the \k{do_call} function shown
-below. This is a variant of the builtin \k{do.call} function with the
-additional support for specifying the environment where the call should happen.
-It constructs a call from a name and a list of arguments, and pass it to \eval.
+below. This is a variant of the builtin \k{do.call} function with different
+semantics for resolving the function name.
+%with the additional support for specifying the environment where the call
+%should happen. -- wrong, do.call also allows specifying an envir
+It constructs a call from a name and a list of arguments, and passes it to \eval.
 
@@
 parameter names can be easily added. This example illustrates one of the most
-recurring uses of \eval by packages that do statistical modeling, \ie the
-\c{model.frame} shape in Table~\ref{tab:minimizedexpressions}.
+recurring uses of \eval by packages that do statistical modeling, and relates to
+the \c{model.frame} shape in Table~\ref{tab:minimizedexpressions}.
 
@@
 This combination of \k{match.call} and \eval for passing most of the current
-call's argument to another function is also used to created wrappers. For
+call's argument to another function is also used to create wrappers. For
 example, the \k{write.csv} uses it to forward most of the parameters to the
@@
 
-The most probable reason for this is to ease maintenance of such
-wrappers.\footnote{Private communication with a developer of R.} Wrapping
-functions in R is somewhat harder than in other languages because of default
-arguments, varargs and missing arguments.
+The pattern eases maintenance of wrappers by avoiding the need to explicitly
+list all the parameter names.\footnote{Private communication with a developer of
+  R.} Wrapping functions in R is somewhat harder than in other languages because
+of default arguments, varargs and missing arguments.
 
@@
 \Eval allows functions to resolve symbols in different environments than the
-current one. For example, the \k{data.table} redefines the subsetting
+current one. For example, the \k{data.table} package redefines the subsetting
 operator\footnote{It defines it as a triplet \texttt{[i, j, k]}, where expression
@@
 symbols \k{carrier}, \k{year}, \k{dep_delay} and \k{origin} are not looked up
-in the current environment, but instead in the \k{flights} data frame.
-
-The non-standard scoping is an important use case in R. It allows ergonomic
-functions for interactive data exploration. The ergonomics come from the
-reduced typing and syntactic noise at the cost of some ambiguity. Since R is
-used primarily interactively, this trade-off is justifiable.
+in the current environment, but in the \k{flights} data frame.
+
+Non-standard scoping is an important use case of \eval in R. It enables the
+design of functions which require less typing. Since R is used primarily
+interactively, this trade-off is justifiable.
 
 \paragraph{Domain-Specific Language}
-A domain-specific language leverages R's grammar but redefines it semantics in
-some suitable way. The usage example illustrates string interpolation. The
+A domain-specific language leverages R's grammar but redefines its semantics in
+a suitable way. The example below illustrates string interpolation. The
 \k{glue} function from the \k{glue} package extracts snippets of code enclosed
@@
 
-The \k{glue} package contains a variant that leverages the non-standard scoping
-allowing to resolve interpolation in an alternative environment. For example,
-the following snippet pipes the subsetted data frame into the \k{glue_data}
-function that uses it to resolve the symbols in string interpolation.
+The \k{glue} package contains a variant of \k{glue} function that leverages
+non-standard scoping for interpolation in a custom environment. For example, the
+following snippet pipes the subsetted data frame into the \k{glue_data} function
+that uses it to resolve the symbols for string interpolation.
 
@@
 that the package code does not use certain restricted functions. Package authors
-use \eval to obfuscate their code. The example mutates the variable \k{s} in
-the package environment. However, that is locked by default and
-\k{unlockBinding}, is a restricted function. To work around this, the call to
-unlock is done through \eval.
+use \eval to obfuscate their code to get around this limitation. The example
+below mutates the variable \k{s} in the package environment. However, that is
+locked by default and \k{unlockBinding} is a restricted function. To work around
+this, the call to unlock is done through \eval.
 
@@
 
-In R, \eval is used chiefly for meta-programming and access remote environments.
-Unlike, JavaScript which had a small set of well-defined patterns that could be
-rewritten without \eval, our results suggest that \eval is an integral part of
-programming in R.
-
-The cases where \eval appears to be overkill include simple expression shapes
+In R, \eval is used chiefly for meta-programming and accessing remote
+environments. Unlike JavaScript which had a small set of well-defined patterns
+that could be rewritten without \eval, our results suggest that \eval is an
+integral part of programming in R.
+
+The cases where \eval appears to be an overkill include simple expression shapes
 such as \k{X}, \k{V}, \k{\$}, \k{<-}, and \k{FUN}. Variable lookup in a remote
-environment can be performed by the builtin \k{get} function, a similar case
-can be made for assignment and the \k{assign} function. There is a dedicated
-\k{do.call} function to perform function calls. However, the unnecessary uses
-of \eval are not easy to find and often can only be uncovered by manual
-analysis. Dynamic analysis is limited due to coverage issues, and static
-analysis of R remains an open problem.
+environment can be performed by the builtin \k{get} function, a similar case can
+be made for assignment and the \k{assign} function. There is a dedicated
+\k{do.call} function to perform function calls. However, the unnecessary uses of
+\eval are not easy to find and often can only be uncovered by manual analysis.
+Dynamic analysis is limited due to coverage issues, and static analysis of R
+remains an open problem.
 
@@
 
-For example in Listing~\ref{lst:unnecessary}, the function uses \eval to lookup
-a target function name from a constructed string. Then it constructs a calls it
-with the arguments that are passed to the current function. \Eval is overkill
-in this case: a semantically equivalent function body can use \k{get} to perform
-the lookup and then call the resulting function directly.
+For the example in Listing~\ref{lst:unnecessary}, the function uses \eval to
+lookup a target function name from a constructed string. Then it constructs a
+call with the arguments that are passed to the current function. \Eval is
+overkill in this case: a semantically equivalent function body can use \k{get}
+to perform the lookup and then call the resulting function directly.
 
@@
 
-Another example, from \k{coin::copyslots}, constructs an expression that access
-fields of an object and copy them.
+Another example, from \k{coin::copyslots}, constructs an expression that accesses
+fields of an object and copies them.
 
@@
 
-This can be replaced by symbolic field reads
-and writes:
+This can be replaced by symbolic field reads and writes:
 
@@
 \eval in two ways. First, it supports evaluation of quosures, custom objects
-used by for metaprogramming that bundle expressions with an environment.
+used for metaprogramming that bundle expressions with an environment.
 \k{eval\_tidy} accepts a data mask argument, a set of bindings such as a

### 2021-08-14 paper/main.tex
Grammar

@@
 Environments nest, each has a parent. A new environment created with \k{new.env}
-has the current environment as parent. Parent chains canf be traversed with
+has the current environment as parent. Parent chains can be traversed with
 \k{parent.env}, until \k{emptyenv} is reached. The top-level environment is
@@
 
-\citet{wang} analyzed use of dynamic features in 18 Python programs to find if
+\citet{wang} analyzed the use of dynamic features in 18 Python programs to find if
 they affect file change-proneness. Files with dynamic features are significantly
@@
   \CranRunnableCodeVignettesRnd in vignettes and \CranRunnableCodeTestsRnd in
-  tests). CRAN packages contain 20 years of contribution from thousands from
+  tests). CRAN packages contain 20 years of contribution from
   thousands of authors, often experienced R developers.
@@
 \paragraph{Discussion.} When we started this project, our goal was to contrast the
-usage of \eval by package developers and end-users, but the Kaggle dataset have
+usage of \eval by package developers and end-users, but the Kaggle dataset has
 so few calls to \eval that there is little that can be said about those. The
@@
 
-We implemented an automated pipeline to acquires packages, extracts scripts,
+We implemented an automated pipeline that acquires packages, extracts scripts,
 executes them, traces their behavior, and summarizes observations.
@@
 assignment, allowing us to record side effects that happen in environments
-while evaluating code in \eval. A challenge was that lack of source code
+while evaluating code in \eval. A challenge was the lack of source code
 references for expression that are not within block surrounded by braces. This
@@
 (\PkgFailedProgramsRatio of the runs crashed or timed out).
-Fig.~\ref{fig:traced-eval-callsites} plots exercised site with respect to all
-sites for each program; as can be seen coverage is unequal.
+Fig.~\ref{fig:traced-eval-callsites} plots exercised sites with respect to all
+sites for each program; as can be seen, coverage is unequal.
 Table~\ref{tab:freq} summarizes the frequency of calls, call counts on the left and
@@
 
-Table~\ref{tab:variantseval} summarizes use of variants, the first row has sites
+Table~\ref{tab:variantseval} summarizes the use of variants, the first row has sites
 (\emph{static}), the second has sites encountered during analysis
@@
 
-Table~\ref{tab:cn} shows average count of calls from a given site and given run and
+Table~\ref{tab:cn} shows the average count of calls from a given site and given run and
 how many sites fall in that range. For instance, \packageRunbinh sites are
@@
 simulation functions. At runtime, over a third of our scripts triggered calls to
-\eval, note that all scripts performed \eval originating from base package but
+\eval. Note that all scripts performed \evals originating from base package but
 these are disregarded here. Half of the calls come from a single site, for the
-rest, most packages have low call frequency. The data is consistent with usage
-of \eval for configuration and meta-programming and not in performance sensitive
+rest, most packages have low call frequency. The data is consistent with the usage
+of \eval for configuration and meta-programming, not in performance sensitive
 contexts such as loops. While most arguments to \eval are small, often just a
@@
 The expressions passed to \eval vary widely. We categorize expressions using a
-minimization function which abstract some of the incidental details of
+minimization function which abstracts some of the incidental details of
 expressions and return a ``shape'' that can be used to group similar \evals. The

### 2021-08-14 paper/main.tex
Grammer

@@
 simplifications, the full list can be found in our artifact. It is worth
-mentioning that these simplifications are heuristics, in R \k{1+1} is not
+mentioning that these simplifications are heuristics; in R, \k{1+1} is not
 necessary \k{2}. The addition operator can be redefined, but it typically isn't
@@
 Table~\ref{tab:minimizedexpressions} gives the 10 most frequent shapes with the
-number and ratio of sites that received arguments of that shape. Operations is
+number and ratio of sites that received arguments of that shape. Operations (ops) is
 the median count of instructions performed by the interpreter when evaluating
@@
 is returned an arbitrary number of additional steps may be needed; the median is
-\packageMinimizedmedianoperationsaRnd, suggesting that most variable read do
+\packageMinimizedmedianoperationsaRnd, suggesting that most variable reads do
 little work. Variables are often evaluated in environments that have been
@@
 with the dollar operator, \eg~\k{x\$f}, and vector indexing, \eg~\k{x[42]} and
-\k{x[[24]]}. This takes a few more step of evaluation on average, and is
+\k{x[[24]]}. This takes a few more steps of evaluation on average, and is
 typically used in a function environment.
@@
 \medskip\EE{$min(e)=\c{F()}$}\\[-2mm]\small Function calls without variable lookups or
-assignment. A variable is allowed the function position, looking up
+assignment. A variable is allowed in the function position, looking up
 functions does not trigger computation unless the function's name
@@
 \k{magrittr} package which is used to compose functions, \eg instead of
-\k{f(g(h(x),y),z)} one can write \k{h(x) \%>\% g(y) \%>\% f(z)}. As there many
+\k{f(g(h(x),y),z)} one can write \k{h(x) \%>\% g(y) \%>\% f(z)}. As there are many
 different patterns of use, there are also many different shapes. In comparison
@@
 
-Contrary to JavaScript, in R the environment of \eval can be specifies by its
+Contrary to JavaScript, in R the environment of \eval can be specified by its
 \k{env} argument. This gives users control over what is visible to the
@@
 but it can sometimes take effect several frames up the call stack. There is
-something brittle about code relying on the position of a caller, a lot of
+something brittle about code relying on the position of a caller: a lot of
 refactoring may break code that does that.
@@
 
-Synthetic environments are relatively frequent. Common uses-case include
-evaluation of an expression in a sandbox, or using a data structure as
+Synthetic environments are relatively frequent. Common uses-cases include
+the evaluation of an expression in a sandbox, or using a data structure as
 environment. For example one could evaluate a method of an object and use an
@@
 Our analysis records the values returned by some functions of interest as well
-as their arguments. We build a direct acyclic graph that track the origin
+as their arguments. We build a direct acyclic graph that tracks the origin
 of the values that are given to \eval. Fig~\ref{fig:provgraph} shows an example
@@
 member of assignments until reaching a starting node. It represents the origin that is
-further modified to yield the expression passed to \eval. In our example, the
+further modified to yield the expression passed to \eval. In our example, that
 origin is \k{parse}.
@@
 
-To have a high-level views of provenance we classify functions into the
+To have a high-level views of provenance, we classify functions into the
 following five categories:
@@
   \k{substitute} is used to retrieve their source. 
-\item {\it String:} create from strings by invoking \k{parse},
+\item {\it String:} created from strings by invoking \k{parse},
   \k{str2lang}, or \k{str2expression}.

### 2021-08-14 paper/main.tex
Grammar (end)

@@
 The table gives the number of call sites of \eval (and the ratio) performing
-side-effect in a particular environment kind. The table also gives the number of
+side-effects in a particular environment kind. The table also gives the number of
 functions in which these sites occur (and ratio).
@@
 developers to this end. To understand \emph{why} developers use \eval, we
-manually inspected top 117 sites which contribute to 90\% of the calls in the
+manually inspected the top 117 sites which contribute to 90\% of the calls in the
 corpus. We present examples from nine classes from this batch. The
@@
 Non-standard scoping is an important use case of \eval in R. It enables the
-design of functions which require less typing. Since R is used primarily
+design of functions which require less writing. Since R is used primarily
 interactively, this trade-off is justifiable.
@@
 such as \k{X}, \k{V}, \k{\$}, \k{<-}, and \k{FUN}. Variable lookup in a remote
-environment can be performed by the builtin \k{get} function, a similar case can
+environment can be performed by the builtin \k{get} function. A similar case can
 be made for assignment and the \k{assign} function. There is a dedicated

### 2021-09-03 paper/main.tex
Update abstract
Fix grammar and typos in Section 1

@@
 \begin{abstract}
- Most dynamic languages allow users to turn text into code using
-  various functions, often named \eval, with language-dependent semantics. The
+  Most dynamic languages allow users to turn text into code using various
+  functions, often named \eval, with language-dependent semantics. The
   widespread use of these reflective functions hinders static analysis and
@@
   domain-specific; we focus on data science code written in R and compare our
-  results to previous work that analyzed web programming in JavaScript. This
-  paper studied \CranRunnableScripts scripts extracted from \CranPackages R
-  packages, for a total of \packageAllcalls calls to \eval. We find that \eval
-  is indeed in widespread use; R's \eval is more pervasive and arguably
-  dangerous than what was previously reported for JavaScript.
+  results to previous work that analyzed web programming in JavaScript. We
+  analyze \packageAllcalls calls to \eval from \CranRunnableScripts scripts
+  extracted from \CranPackages R packages. We find that \eval is indeed in
+  widespread use; R's \eval is more pervasive and arguably dangerous than what
+  was previously reported for JavaScript.
 \end{abstract}
@@
 write down the definition of LISP; he was surprised that someone coded it up and
-offered it to end users~\cite{lisp}. Since then, reflective facilities have been
+offered it to end-users~\cite{lisp}. Since then, reflective facilities have been
 used to parameterize programs over code patterns that can be provided after the
@@
 much precision as to become pointless. For compilers, anything but the most
-trivial, local optimizations are unsound after use of \eval. Furthermore, the
+trivial, local optimizations are unsound after the use of \eval. Furthermore, the
 addition of arbitrary code --- code that could have been obtained from a network
@@
 
-\paragraph{The Here and Now.} In this study, we investigate the
-usage of \eval in programs written in the R language. R is a language designed
-by statisticians for applications in data science~\cite{r,R96}. What makes
-looking at R after JavaScript interesting is that, while both languages are
-dynamic, they are quite different. While one can program in an object-oriented
-style in R like in JavaScript, R is mostly a lazy functional language.
-JavaScript was designed to run untrusted code in a browser, while R is used for
-statistical computing on desktops. JavaScript is a general-purpose language used
-by a vast community of programmers, while R is used for scientific computing by
-data scientists and domain experts with, often, limited programming experience.
-One can distinguish between library implementers who have programming experience
-and a working knowledge of R, and end-users who are typically not expert
-programmers with a cursory knowledge of the language.\footnote{We surveyed
-end-users and did not find a single one aware that R is lazy.} Our goal is to
-highlight the differences in usage between JavaScript and R and explain them in
-terms of language features, application domain and programmer experience.
-Hopefully, some of our observations generalize to other languages. Our data
-and code are open source and publicly available:
-\url{https://github.com/PRL-PRG/evalr-experiment}.
+\paragraph{The Here and Now.} In this study, we investigate the usage of \eval
+in programs written in the R language. R is a language designed by statisticians
+for applications in data science~\cite{r,R96}. What makes looking at R after
+JavaScript interesting is that, while both languages are dynamic, they are quite
+different. While one can program in an object-oriented style in R like in
+JavaScript, R is mostly a lazy functional language. JavaScript was designed to
+run untrusted code in a browser, while R is used for statistical computing on
+desktops. JavaScript is a general-purpose language used by a vast community of
+programmers, while R is used for scientific computing by data scientists and
+domain experts with, often, limited programming experience. One can distinguish
+between library implementers who have programming experience and a working
+knowledge of R, and end-users who are typically not expert programmers with a
+cursory knowledge of the language.\footnote{We surveyed end-users and did not
+  find a single one aware that R is lazy.} Our goal is to highlight the
+differences in usage between JavaScript and R and explain them in terms of
+language features, application domain, and programmer experience. Hopefully,
+some of our observations generalize to other languages. Our data and code are
+open-source and publicly available at:
+
+\begin{center}
+  \url{https://doi.org/10.5281/zenodo.5415230}
+\end{center}
 
@@
 \eval. In many ways, \eval in R is as bad as it gets: it's varied, performs
-side-effects and reaches to many environments. By large, the motivations for
+side-effects, and reaches many environments. By large, the motivations for
 \eval relate to various forms of language extensions and meta-programming. \Eval

### 2021-09-03 paper/main.tex
Fix typos and grammar in Section 3

@@
 \item \emph{CRAN.} Packages hosted on the Comprehensive R Archive Network
-  ({\small \url{cran.r-project.org}}, aka CRAN) have three sources of runnable:
+  ({\small \url{cran.r-project.org}}, aka CRAN) have three sources of runnable code:
   unit \emph{tests} for individual functions, code \emph{examples} embedded in
@@
   (\KaggleDuplicates duplicates were removed using SHA-1 hashes); 665K lines of
-  R code making only \KaggleWithEvals calls \eval. Authors of these scripts
+  R code making only \KaggleWithEvals calls \eval. The authors of these scripts
   have highly variable levels of expertise.
@@
 Figure~\ref{fig:pipeline} shows the main steps along with their running time,
-data size, and number of elements manipulated. The pipeline steps are:
+data size, and the number of elements manipulated. The pipeline steps are:
 
@@
   extracts code from notebooks. Each extracted script is instrumented with calls
-  to our dynamic analyzer in a way that avoid recording calls to \eval
+  to our dynamic analyzer in a way that avoids recording calls to \eval
   originating from our execution harness.
-\item \emph{Trace.} Script run with a modified R interpreter to capture calls
-  to \eval in CRAN packages. Each script is run in its own process with the GNU
-  R compiler turned off to avoid recording its execution.
+\item \emph{Trace.} Scripts are run with a modified R interpreter to capture
+  calls to \eval in CRAN packages. Each script is run in its own process with
+  the GNU R compiler turned off to avoid recording its execution.
 \item \emph{Analyze.} Analysis outputs are merged, cleaned, and summarized.
-  RMarkdown notebooks process the summarized data to generate all graphs (PDF
-  from \k{ggplot2}) and numbers (exported as \LaTeX macros) appearing in the
-  paper.
+  RMarkdown~\cite{rmarkdown} notebooks process the summarized data to generate all graphs (using
+  \k{ggplot2}~\cite{ggplot}) and numbers (exported as \LaTeX\xspace macros)
+  appearing in the paper.
 \end{compactenum}
@@
 arguments. It captures dynamic code loading, as well as variable definition and
-assignment, allowing us to record side effects that happen in environments
-while evaluating code in \eval. A challenge was the lack of source code
-references for expression that are not within block surrounded by braces. This
-is unfortunately not easily fixed. We extend our dynamic analysis tool to
-attach synthetic source code references to all \eval call sites, but the
-approach fails in some edge cases. For performance reasons, the tracer is an R
-package written in C++ (3.2K LOC) and R (1.3K LOC). While in theory the
-implementation should be straightforward, not so in practice. Lazy evaluation
-requires delaying the processing of arguments until (and if) they are forced.
-Accounting for side-effects performed in an \eval is complicated by the fact
-that R is implemented in a mixture of R and C, and that the language
-implementation can (and does) call \eval. Lastly, large code bases exercise
-many edge cases of the highly underspecified R behavior.
+assignment, allowing us to record side effects that happen in environments while
+evaluating code in \eval. A challenge was the lack of source code references for
+expressions that are not within a block surrounded by braces. This is
+unfortunately not easily fixed. We extend our dynamic analysis tool to attach
+synthetic source code references to all \eval call sites, but the approach fails
+in some edge cases. For performance reasons, the tracer is an R package written
+in C++ (3.2K LOC) and R (1.3K LOC). While in theory, the implementation should
+be straightforward, it is not so in practice. Lazy evaluation requires delaying
+the processing of arguments until (and if) they are forced. Accounting for
+side-effects performed in an \eval is complicated by the fact that R is
+implemented in a mixture of R and C, and that the language implementation can
+(and does) call \eval. Lastly, large codebases exercise many edge cases of the
+highly underspecified R behavior.
 
@@
 \PkgUndefinedRnd \eval calls without source information (\PkgUndefinedRatio of
-all \eval calls). This happen when \eval is an argument to a higher-order
-function or when it is called from native code. Another limitation is that we
-do not record calls to the native \eval. Neither of these limitations should
+all \eval calls). This happens when \eval is an argument to a higher-order
+function or when it is called from native code. Another limitation is that we do
+not record calls to the native \eval. Neither of these limitations should
 invalidate our conclusions.

### 2021-09-03 paper/main.tex
Align figures
Fix typos and grammar in Section 4

@@
 sites for each program; as can be seen, coverage is unequal.
-Table~\ref{tab:freq} summarizes the frequency of calls, call counts on the left and
+Table~\ref{tab:freq} summarizes the frequency of calls, call counts on the left, and
 number of packages on the right. There are \packageFewcalls packages with fewer
@@
 \begin{figure}[H]
-  \includegraphics[width=9cm]{pkgs-eval-callsites-hist.pdf}
+  \includegraphics[width=7cm]{pkgs-eval-callsites-hist.pdf}
   \caption{ \eval call sites}\label{fig:pkgs-eval-callsites-hist}
@@
 \begin{figure}[H]
-  \includegraphics[width=.8\textwidth]{traced-eval-callsites.pdf} \centering
+  \includegraphics[width=.7\textwidth]{traced-eval-callsites.pdf} \centering
   \caption{\eval call sites coverage of the \PkgPackages packages.}%
@@
 called over 2,000 times per run. Larger call counts suggest the presence of
-loops or recursive functions, but given the data this seems to be the exception;
-\packageRunbina of sites are called fewer than 50 times, most sites are invoked
+loops or recursive functions, but given the data, this seems to be the exception;
+\packageRunbina sites are called fewer than 50 times, most sites are invoked
 once and half as many are invoked twice.
 
-\Eval accepts any value as argument, but if the value is not an expression,
+\Eval accepts any value as an argument, but if the value is not an expression,
 \eval returns it unchanged. Expressions account for \packageCodepercent of
-arguments. More specifically, \packageSymbolpercent are symbols (variables
-such as \k{x}), \packageLanguagepercent are language objects (expressions
-representing a function calls), and \packageExpressionpercent are expression
+arguments. More specifically, \packageSymbolpercent are symbols (variables such
+as \k{x}), \packageLanguagepercent are language objects (expressions
+representing function calls), and \packageExpressionpercent are expression
 objects (lists of expressions). Further inspection reveals that \k{\_inherit}
 accounts for \packageGgplotsymbolpercent of symbols and comes from one site in
-\k{ggplot2}.\footnote{Each object has a field model inheritance, this comes from
-\k{ggproto}, one of the many object-oriented systems for R.}
+\k{ggplot2}.\footnote{This field is used to model inheritance in \k{ggproto},
+  one of the many object-oriented systems in R.}
 
@@
 To estimate how much executable code is injected through \eval, we measure the
-size of the arguments in terms of number of nodes. For example,
+size of the arguments in terms of the number of nodes. For example,
 \k{expression(x+1)} has a size of 3. The large number of symbols skew the median
@@
 legibility. As an alternative to counting nodes, we tried measuring string
-lengths of expressions after calling a function to deparse values back to
+lengths of expressions after calling a function to convert values back to
 strings, but these measurements were dominated by massive data objects which
@@
 To assess how much computational work is performed in \evals, we count the
-instructions executed by the interpreter. Most call do relatively little, with
+instructions executed by the interpreter. Most calls do relatively little, with
 \packageSmalleventspct of calls executing fewer than 50 instructions. The violin
 plot of Fig.~\ref{ev}(a) shows the distribution of short running \evals; the
-data is dominated by \evals which simply perform symbol lookup. Fig.~\ref{ev}(b)
+data is dominated by \evals that simply perform symbol lookup. Fig.~\ref{ev}(b)
 shows the distribution of work-intensive \evals which go all the way to
@@
 simulation functions. At runtime, over a third of our scripts triggered calls to
-\eval. Note that all scripts performed \evals originating from base package but
+\eval. Note that all scripts call \eval originating from the base package but
 these are disregarded here. Half of the calls come from a single site, for the
 rest, most packages have low call frequency. The data is consistent with the usage
-of \eval for configuration and meta-programming, not in performance sensitive
+of \eval for configuration and meta-programming, not in performance-sensitive
 contexts such as loops. While most arguments to \eval are small, often just a
-variable name, and do little work, we have also observed large \eval arguments
+variable name, and do little work, we have also observed large arguments
 and arguments that perform massive amounts of work.

### 2021-09-03 paper/main.tex
Fix Section 6

@@
 \lstdefinelanguage{smalleR} {
-  morekeywords={
+  keywords={
     for,
@@
   },
-  sensitive=true, % keywords are not case-sensitive
+  sensitive=true, % keywords are case-sensitive
   morecomment=[l]{\#}, % l is for line comment
@@
   stringstyle=\color{Gray},
-  alsoletter={., _, $},
+  showspaces=false,
+  showstringspaces=false,
+  deletekeywords={see_if, $, _, .},
+  % alsoletter={., _, $},
+  %emph={eval.parent, evalq, eval, local},
+  %emphstyle=\underbar,
   breaklines=true
@@
 libraries, their results were dominated by a few functions with
-\c{match.arg}\footnote{A base function that matches an argument against a set of
+\k{match.arg}\footnote{A base function that matches an argument against a set of
 candidate values, it uses \eval to get the candidate values.} being responsible
@@
 accounts for \packageGgplotsymbolpercent of symbols and comes from one site in
-\k{ggplot2}.\footnote{This field is used to model inheritance in \k{ggproto},
-  one of the many object-oriented systems in R.}
+\k{ggplot2}.\footnote{This field is used to model inheritance in \k{ggproto}, an
+  object-oriented system used by \k{ggplot2}.}
 
@@
 The R language was intended to be extensible. The combination of lazy
-evaluation, \k{substitute} and \k{eval} are some of the tools given to
+evaluation, \k{substitute}, and \k{eval} are some of the tools given to
 developers to this end. To understand \emph{why} developers use \eval, we
-manually inspected the top 117 sites which contribute to 90\% of the calls in the
-corpus. We present examples from nine classes from this batch. The
-classification is not exhaustive; we only study 117 call sites from the
-\PkgEvalCallSites call sites in our corpus and many uses of \eval do not fit in
-a specific category. The goal here is to demonstrate the wide variety of uses to
-which \eval is put to use. This also speaks to the ubiquity of \eval in the R
-ecosystem. The code snippets have been simplified for clarity.
+manually inspected the top 117 sites, which contribute to 90\% of the calls in
+the corpus. We present examples from nine classes from this batch. The
+classification is not exhaustive; we present results from only 117 of the total
+\PkgEvalCallSites call sites in our corpus, and many uses of \eval do not fit in
+a specific category. The \eval usage patterns are described below.
 
@@
 extend the behavior of a library function. \Eval allows clients to provide a
-simple text string that shall be executed in the context of the function. The
+simple text string that is executed in the context of the function. The
 example here allows users to provide formulas as strings, such as
@@
 \k{x} that will be set up by the called function.\footnote{The function is used
-to generates a sequence of random numbers using the adaptive rejection sampling
-algorithm and the string formula is used to describe the target density.} In
-this case users could provide an anonymous function instead.
-
-\begin{minipage}{.95\textwidth}
-\medskip\underline{\c{AdapSamp::rARS}:}
-\begin{lstlisting}
+to generate a sequence of random numbers using the adaptive rejection sampling
+algorithm and the string formula is used to describe the target density.}
+
+\begin{lstlisting}[caption={\c{AdapSamp::rARS}}, captionpos=b]
  function(n,formula) {
    p <- function(x) eval(parse(text=formula))
-\end{lstlisting}\medskip
-\end{minipage}
+\end{lstlisting}
 
@@
 \Eval can be used to generate boilerplate code that would either be tedious to
-write by hand or that is only known after deployment. The example uses \eval to
-automatically generate R bindings to C++ methods.
-
-\begin{minipage}{.95\textwidth}
-\medskip\underline{\c{Rcpp::.makeCppMethods}:}
-\begin{lstlisting}
+write by hand or that is only known after deployment. The following code snippet
+for the \k{Rcpp} package uses \eval to automatically generate R bindings to C++
+methods.
+
+\begin{lstlisting}[caption={\k{Rcpp::.makeCppMethods}}, captionpos=b]
  function(methods,env) {
@@
        eval(substitute(function(...)CppObject$what(...)), env)
-\end{lstlisting}\medskip
-\end{minipage}
+\end{lstlisting}
 
 \paragraph{Extracting Varargs}
-\Eval can be used to obtain the source code of the arguments passed to a vararg
-parameter. One example is a function that iterates over the contents of the
-vararg until it finds a non-null value. \k{substitute(alist(...))} yields an
-expression that represents a call to \k{alist} but where the \k{...} are
-expended with the arguments supplied to the function. \Eval then returns a
-\k{list} with the unevaluated argument expressions.
-
-\begin{minipage}{.95\textwidth}
-\medskip\underline{\c{statnet.common::NVL}:}
-\begin{lstlisting}
+\Eval is used to obtain the source code of the arguments passed to a vararg
+(\k{...}) parameter. In the following example, \k{substitute(alist(...))} yields
+an expression representing a call to \k{alist} with the \k{...} expanded to the
+arguments supplied to the function. \Eval evaluates this expression and returns
+a \k{list} of the unevaluated argument expressions.
+
+\begin{lstlisting}[caption={\k{statnet.common::NVM}}, captionpos=b]
  function(...) {
@@
 \end{lstlisting}\medskip
-\end{minipage}
 
 \paragraph{Controlled Evaluation}
-\Eval can be used to control when and if expressions are evaluated. For
-example, \k{assertthat} provides users with the ability to make assertions and,
-if they fail, produce legible error messages. The example here accepts a vararg
-with a list of expressions. It extracts the associated code in the same way
-as the extracting varags example and evaluates the expressions
-sequentially in the caller's environment. If an expression does not hold, an
-error is generated.
-
-\begin{minipage}{.95\textwidth}
-  \medskip\underline{\c {assertthat::see\_if}:}
-\begin{lstlisting}
+\Eval can be used to control when and if expressions are evaluated. For example,
+\k{assertthat} provides users with the ability to make assertions and, if they
+fail, produce legible error messages. The \k{assertthat::see_if} function
+accepts a vararg with a list of expressions. It extracts these expressions
+(\emph{Extracting Varargs} pattern) and evaluates them sequentially in the
+caller's environment using \k{eval}. If an expression does not hold, an error is
+generated.
+
+\begin{lstlisting}[caption={\k{assertthat::see\_if}}, captionpos=b]
  function(..., env=parent.frame(), msg=NULL) {
@@
 \end{lstlisting}\medskip
-\end{minipage}
 
 \paragraph{Shadowing}
-%%% TODO ?????????????//// <- left by Jan
-Some packages use \eval to implement variations of the \k{base} R functions.
-For example, the \k{igraph} package provides the \k{do_call} function shown
-below. This is a variant of the builtin \k{do.call} function with different
-semantics for resolving the function name.
-%with the additional support for specifying the environment where the call
-%should happen. -- wrong, do.call also allows specifying an envir
-It constructs a call from a name and a list of arguments, and passes it to \eval.
-
-\begin{minipage}{.95\textwidth}
-  \medskip\underline{\c{igraph::do\_call}:}
-\begin{lstlisting}
+Some packages use \eval to implement variations of the \k{base} R functions. For
+example, the \k{igraph} package provides the \k{do_call} function shown below.
+This is a variant of the builtin \k{do.call} function with different semantics
+for resolving the function name. It constructs a call from a name and a list of
+arguments and passes it to \eval.
+
+\begin{lstlisting}[caption={\k{igraph::do\_call}}, captionpos=b]
  function(f,...,args=list(),env=parent.frame()) {
@@
 \end{lstlisting}\medskip
-\end{minipage}
 
@@
 via \eval and \k{match.call}. The example uses \k{match.call} to access the
-expression with which the function is called. The next three lines construct a
+expression with which the function is called. The following three lines construct a
 call to \k{stats::model.frame} with argument text of parameters \k{'formula'},
@@
 recurring uses of \eval by packages that do statistical modeling, and relates to
-the \c{model.frame} shape in Table~\ref{tab:minimizedexpressions}.
-
-\begin{minipage}{.95\textwidth}
-  \medskip\underline{\c{survival::survfit.formula}:}
-\begin{lstlisting}
+the \k{model.frame} shape in Table~\ref{tab:minimizedexpressions}.
+
+\begin{lstlisting}[caption={\k{survival::survfit.formula}}, captionpos=b]
  function(formula, data, weights, ...) {
@@
 \end{lstlisting}\medskip
-\end{minipage}
-
-This combination of \k{match.call} and \eval for passing most of the current
-call's argument to another function is also used to create wrappers. For
-example, the \k{write.csv} uses it to forward most of the parameters to the
-more general \k{write.table} function, only setting the CSV-related
-parameters:
-
-\begin{minipage}{.95\textwidth}
-  \medskip\underline{\c{base::write.csv}:}
-\begin{lstlisting}
- function (...)
- {
+
+Wrapping functions in R is somewhat harder than in other languages because of
+default arguments, varargs, and missing arguments. The combination of
+\k{match.call} and \eval can be used to forward the current call's arguments to
+another function without listing all the argument names. This pattern eases the
+maintenance of wrappers. For example, the \k{base::write.csv} function passes
+most of its arguments unchanged to the more general \k{base::write.table}
+function.
+
+\begin{lstlisting}[caption={\k{base::write.csv}}, captionpos=b]
+ function (...) {
    Call <- match.call(expand.dots = TRUE)
@@
 \end{lstlisting}\medskip
-\end{minipage}
-
-The pattern eases maintenance of wrappers by avoiding the need to explicitly
-list all the parameter names.\footnote{Private communication with a developer of
-  R.} Wrapping functions in R is somewhat harder than in other languages because
-of default arguments, varargs and missing arguments.
-
-% % TODO: update to explain why the eval is needed or remove
-% \paragraph{Formula Resolution}
-% Many packages use \eval to evaluate formulas. The example extracts the
-% \k{"formula"} attribute attached to the \k{pdMat} object and evaluates it.
-
-% \begin{minipage}{.95\textwidth}
-%   \medskip\underline{\c{nlme::formula.pdMat}:}
-% \begin{lstlisting}
-%  function(x, asList, ...) eval(attr(x, "formula"))
-% \end{lstlisting}\medskip
-% \end{minipage}
 
@@
 current one. For example, the \k{data.table} package redefines the subsetting
-operator\footnote{It defines it as a triplet \texttt{[i, j, k]}, where expression
-  \texttt{j} is used to calculate a subset of rows from \texttt{i} grouped by \texttt{k}.}
-for data frames to enable compact queries. The following  
-computes the average departure delays on 2014 Delta flights for each
-origin airport.
-
-\begin{minipage}{.95\textwidth}
-\medskip
-\begin{lstlisting}
+operator, \k{[}\footnote{Defined as a triplet \texttt{[i, j, k]}, where
+  expression \texttt{j} is used to calculate a subset of rows from \texttt{i}
+  grouped by \texttt{k}.}, to enable compact queries on data frames. The example
+below shows the use of this operator on the \k{flights} data frame to compute
+the average departure delays of Delta flights in 2014 by the origin airport. The
+symbols \k{carrier}, \k{year}, \k{dep_delay} and \k{origin} are not looked up in
+the current environment, but in the \k{flights} data frame.
+
+\begin{lstlisting}[caption={\k{data.table::[}}, captionpos=b]
 > flights[carrier=="DL" & year == 2014, mean(dep_delay), by=origin]
@@
 expressions, transforms them, and evaluates them using \eval in the given data
-frame.\footnote{We omit the code as the implementation spans over 1700 lines,\\
-  \indent \cf
-\url{https://github.com/Rdatatable/data.table/blob/master/R/data.table.R}} The
-symbols \k{carrier}, \k{year}, \k{dep_delay} and \k{origin} are not looked up
-in the current environment, but in the \k{flights} data frame.
+frame. We omit the implementation as it spans over 1700 lines at the time of
+this writing\footnote{Its implementation can be found at:
+  \url{https://github.com/Rdatatable/data.table/blob/master/R/data.table.R}}.
 
 Non-standard scoping is an important use case of \eval in R. It enables the
-design of functions which require less writing. Since R is used primarily
+design of functions that require less typing. Since R is used primarily
 interactively, this trade-off is justifiable.
@@
 A domain-specific language leverages R's grammar but redefines its semantics in
-a suitable way. The example below illustrates string interpolation. The
-\k{glue} function from the \k{glue} package extracts snippets of code enclosed
-between braces and evaluates them using \eval, and finally splices their
-results to construct the result.
-
-\begin{minipage}{.95\textwidth}
-\medskip
-\begin{lstlisting}
+a suitable way. The example below illustrates string interpolation. The \k{glue}
+function from the \k{glue} package extracts snippets of code enclosed between
+braces, evaluates them using \eval, and splices their results to construct the
+result.
+
+\begin{lstlisting}[caption={\k{glue::glue}}, captionpos=b]
 > greeting <- "Hello"
@@
 
-\begin{minipage}{.95\textwidth}
-\medskip
-\begin{lstlisting}
+\begin{lstlisting}[caption={\k{glue::glue_data}}, captionpos=b]
 > flights[carrier=="DL" & year == 2014, mean(dep_delay), by=origin] |> 
@@
 below mutates the variable \k{s} in the package environment. However, that is
-locked by default and \k{unlockBinding} is a restricted function. To work around
-this, the call to unlock is done through \eval.
-
-\begin{minipage}{.95\textwidth}
-  \medskip\underline{\c{aibd::scalaEnsure}:}
-\begin{lstlisting}
+locked by default and unlocking it requires \k{unlockBinding} which is a
+restricted function. To work around this, the call to \k{unlockBinding} is done
+through \eval.
+
+\begin{lstlisting}[caption={\k{aibd::scalaEnsure}}, captionpos=b]
  function() {

### 2021-09-03 paper/main.tex
changes

@@
-\documentclass[review,screen,acmsmall]{acmart}%\settopmatter{printfolios=false,printccs=false,printacmref=false}
-\setcopyright{none}
-\renewcommand\footnotetextcopyrightpermission[1]{}
-\pagestyle{plain}
+\documentclass[acmsmall, screen]{acmart}
+\setlength{\pdfpagewidth}{8.5in}
+\setlength{\pdfpageheight}{11in}
+
+%%% The following is specific to OOPSLA '20 and the paper
+%%% 'World Age in Julia: Optimizing Method Dispatch in the Presence of Eval'
+%%% by Julia Belyakova, Benjamin Chung, Jack Gelinas, Jameson Nash, Ross Tate, and Jan Vitek.
+%%%
+\setcopyright{rightsretained}
+\acmPrice{}
+\acmDOI{10.1145/3428275}
+\acmYear{2020}
+\copyrightyear{2020}
+\acmSubmissionID{oopsla20main-p384-p}
+\acmJournal{PACMPL}
+\acmVolume{4}
+\acmNumber{OOPSLA}
+\acmArticle{207}
+\acmMonth{11}
 
@@
 
+%%% TODO CHANGE
+\keywords{eval, dynamic languages}
+\begin{CCSXML}
+	<ccs2012>
+	<concept>
+	<concept_id>10011007.10011006.10011008.10011024</concept_id>
+	<concept_desc>Software and its engineering~Language features</concept_desc>
+	<concept_significance>500</concept_significance>
+	</concept>
+	<concept>
+	<concept_id>10011007.10011006.10011008</concept_id>
+	<concept_desc>Software and its engineering~General programming languages</concept_desc>
+	<concept_significance>300</concept_significance>
+	</concept>
+	<concept>
+	<concept_id>10011007.10011006.10011041.10011044</concept_id>
+	<concept_desc>Software and its engineering~Just-in-time compilers</concept_desc>
+	<concept_significance>300</concept_significance>
+	</concept>
+	<concept>
+	<concept_id>10011007.10011006.10011008.10011009.10011021</concept_id>
+	<concept_desc>Software and its engineering~Multiparadigm languages</concept_desc>
+	<concept_significance>100</concept_significance>
+	</concept>
+	</ccs2012>
+\end{CCSXML}
+
+\ccsdesc[500]{Software and its engineering~Language features}
+\ccsdesc[300]{Software and its engineering~General programming languages}
+
 \author{Aviral Goel}\affiliation{\institution{Northeastern
@@
 1956. \Eval has been much maligned over the years. In computing lore, it is as
-close to a boogeyman as it gets. Yet, for McCarthy, \eval was simply the way to
+close to a boogeyman as it gets. Yet, for \citet{lisp}, \eval was simply the way to
 write down the definition of LISP; he was surprised that someone coded it up and
-offered it to end-users~\cite{lisp}. Since then, reflective facilities have been
+offered it to end-users. Since then, reflective facilities have been
 used to parameterize programs over code patterns that can be provided after the
@@
 much precision as to become pointless. For compilers, anything but the most
-trivial, local optimizations are unsound after the use of \eval. Furthermore, the
-addition of arbitrary code --- code that could have been obtained from a network
-connection --- as a program is running is a security vulnerability waiting to
-happen. To illustrate these challenges, consider the interaction of a static
-analysis tool with a dynamic language. A program analyzer computes an
+trivial, local optimizations are unsound after the use of \eval. Furthermore,
+the addition of arbitrary code --- code that could have been obtained from a
+network connection --- as a program is running is a security vulnerability
+waiting to happen. To illustrate these challenges, consider the interaction of a
+static analysis tool with a dynamic language. A program analyzer computes an
 over-approximation of the set of possible behaviors exhibited by the program
@@
 much can be done. A frustrated group of researchers argued giving up on
-soundness and, instead, under-approximating dynamic features
-(soundiness)~\cite{soundy}. In their words, ``a practical analysis, may pretend
-that \eval does nothing unless it can precisely resolve its string argument at
-compile time.'' Alas, assuming that \eval does not have side-effects or that
-side-effects will not affect results is unduly optimistic.
+soundness and, instead, under-approximating dynamic features~\cite{soundy}. In
+their words, ``a practical analysis, may pretend that \eval does nothing unless
+it can precisely resolve its string argument at compile time.'' Alas, assuming
+that \eval does not have side-effects or that side-effects will not affect
+results is unduly optimistic.
 
@@
 in programs written in the R language. R is a language designed by statisticians
-for applications in data science~\cite{r,R96}. What makes looking at R after
+for applications in data science~\cite{r}. What makes looking at R after
 JavaScript interesting is that, while both languages are dynamic, they are quite
@@
 knowledge of R, and end-users who are typically not expert programmers with a
-cursory knowledge of the language.\footnote{We surveyed end-users and did not
-  find a single one aware that R is lazy.} Our goal is to highlight the
+cursory knowledge of the language. Our goal is to highlight the
 differences in usage between JavaScript and R and explain them in terms of
 language features, application domain, and programmer experience. Hopefully,
-some of our observations generalize to other languages. Our data and code are
-open-source and publicly available at:
-
-\begin{center}
-  \url{https://doi.org/10.5281/zenodo.5415230}
-\end{center}
+some of our observations generalize to other languages.
 
@@
 \eval is higher as it can reach arbitrarily far back in the call stack.
+
+\medskip
+
+Our data and code are
+open-source and publicly available at:
+
+\begin{center}
+  \url{https://doi.org/10.5281/zenodo.5415230}
+\end{center}
 
@@
 
-\subsection{Eval in R}\label{sec:eval-in-r}
-
-The \eval function in R takes three parameters: an expression to evaluate
-(\k{e}), an environment where to evaluate (\k{env}), and an enclosure (\k{encl})
-that is used to look up objects not found in \k{env}.\footnote{R offers three
-other variants of {\tt eval}: {\tt evalq} automatically quotes passed
-expression---shorthand for {\tt eval(quote(...))}, {\tt eval.parent(e,n)}
-specifies the evaluating environment in terms of number of call frames ({\tt n})
-to go back---shorthand for {\tt eval(e,parent.frame(n))}, and {\tt local}
-evaluates {\tt e} in a fresh environment---shorthand for {\tt
-  evalq(e,new.env())}.}
-
-\begin{lstlisting}
- eval <- function(e, env = parent.frame(),
-                   encl = if(is.list(env)) parent.frame() else baseenv()) ...
-\end{lstlisting}
-
-The expression passed to \eval can be thought of as an abstract syntax tree.
-Listings~\ref{lst:exprs} and \ref{lst:match.call} show some of the ways of
-creating expressions: parsing from a string, manually or via reflection. The
-call to \k{substitute(x)} extracts the unevaluated expression from the promise
-\k{x}. The call to \k{match.call()} returns an expression representing the
-current call which can then be further modified.
-
-\begin{figure}[htb]
+\begin{figure}[!t]
 \begin{minipage}{.49\textwidth}
@@
 \end{figure}
+
+\subsection{Eval in R}\label{sec:eval-in-r}
+
+The \eval function in R takes three parameters: an expression to evaluate
+(\k{e}), an environment where to evaluate (\k{env}), and an enclosure (\k{encl})
+that is used to look up objects not found in \k{env}.\footnote{R offers three
+other variants of {\tt eval}: {\tt evalq} automatically quotes passed
+expression---shorthand for {\tt eval(quote(...))}, {\tt eval.parent(e,n)}
+specifies the evaluating environment in terms of number of call frames ({\tt n})
+to go back---shorthand for {\tt eval(e,parent.frame(n))}, and {\tt local}
+evaluates {\tt e} in a fresh environment---shorthand for {\tt
+  evalq(e,new.env())}.}
+
+\begin{lstlisting}
+ eval <- function(e, env = parent.frame(),
+                   encl = if(is.list(env)) parent.frame() else baseenv()) ...
+\end{lstlisting}
+
+The expression passed to \eval can be thought of as an abstract syntax tree.
+Listings~\ref{lst:exprs} and \ref{lst:match.call} show some of the ways of
+creating expressions: parsing from a string, manually or via reflection. The
+call to \k{substitute(x)} extracts the unevaluated expression from the promise
+\k{x}. The call to \k{match.call()} returns an expression representing the
+current call which can then be further modified.
 
@@
 
+\begin{figure}[!h]
+   \includegraphics[width=7cm]{pkgs-eval-callsites-hist.pdf}
+  \caption{ \eval call sites}\label{fig:pkgs-eval-callsites-hist}
+\end{figure}
+
 Our pipeline runs \CranRunnableScripts scripts extracted from \CranPackages
@@
 
-\begin{figure}[H]
-  \includegraphics[width=7cm]{pkgs-eval-callsites-hist.pdf}
-  \caption{ \eval call sites}\label{fig:pkgs-eval-callsites-hist}
-\end{figure}
-\begin{table}[H]
+\begin{table}[!h]
 \centering
@@
 \begin{minipage} {5cm}\small
-  \begin{tabular}{r@{\,}r@{\,}l@{}r|r@{\,}r@{\,}l@{}r} \toprule
+  \begin{tabular}{r@{\,}r@{\,}l@{}r|r@{\,}r@{\,}l@{}r@{}} \toprule
     \multicolumn{3}{c}{\bf \small\#calls} &\bf \small \#pck
@@
 \begin{minipage}{7cm}\small
-\begin{tabular}{@{\,}r|rrrr}\toprule
+\begin{tabular}{@{}r|r@{~}r@{}r@{~}r@{}}\toprule
   &\eval & \k{evalq} & \k{eval} & \k{local}\\[-1.5mm]
@@
 
-\begin{wraptable}{r}{5cm}  \small  \vspace*{-2mm}\centering
+\begin{table}[!h]  \small  \centering
   \begin{tabular}{r@{\,}r@{\,}l@{\,}r|r@{\,}r@{\,}l@{}r} \toprule
@@
   \caption{Normalized calls} \label{tab:cn}
-\end{wraptable}
+\end{table}
 
@@
 
-\begin{wrapfigure}{r}{6.2cm}
+\begin{figure}[!h]
 \begin{tabular}{c}
-{\hspace{-25mm}\includegraphics[width=0.8\textwidth]{package_size_loaded_distribution}}
+{\includegraphics[width=0.8\textwidth]{package_size_loaded_distribution}}
 \end{tabular}
 \caption{Loaded code} \label{fig:sizedistribution}
-\end{wrapfigure}
+\end{figure}
 
@@
   The environment in which they evaluate is mostly irrelevant.
-\\
+\end{tabular}
+
+\begin{tabular}{@{}p{.97\linewidth}}
 \medskip\EE{$min(e)=\c{X}$}\\[-2mm]\small Variable lookups are the most common shape,
@@
 expressions are evaluated in a function environment.
-\\
+\end{tabular}
+
+\begin{tabular}{@{}p{.97\linewidth}}
 \medskip\EE{$min(e)=\c{\$}$}\\[-2mm]\small This shape extends \k{X} to include lookup
@@
 is bound to a promise. This shape typically does not perform much work.
-\\
+\end{tabular}
+
+\begin{tabular}{@{}p{.97\linewidth}}
 \medskip\framebox{$min(e)=\c{F(X)}$}~\EE{$min(e)=\c{F(F(X))}$}\\[-2mm]\small Function
@@
 \k{doParallel}), or deferring code execution (\eg \k{withr}).
-\\
+\end{tabular}
+
+\begin{tabular}{@{}p{.97\linewidth}}
 \medskip\EE{$min(e)=\c{model.frame}$}\\[-2mm]\small The \k{model.frame} function
@@
 
-% Wrap tables do not play well with lists (they warn about it in the documentation actually!)
-\begin{wraptable}{r}{5cm}\small\centering
+\begin{table}[!b]\small\centering
+\begin{tabular}{c@{\hspace{2cm}}c}
   \begin{tabular}{lcc}\toprule
@@
 \end{tabular}
-
-\medskip
-
+&
 \begin{tabular}{rrr} \toprule
@@
 \end{tabular}
-\caption{Origins}\label{tab:provenance}\vspace{-1cm}
-\end{wraptable}
+\end{tabular}
+\medskip
+\caption{Origins}\label{tab:provenance}
+\end{table}
 
@@
 
-\bibliography{bib/bibliography,bib/jv}
+\bibliography{bibexport}
 

### 2021-09-03 paper/main.tex
acks

@@
 
+\begin{acks}
+  This work has received funding from the
+\grantsponsor{NSF}{National Science Foundation}{} awards
+\grantnum{NSF}{1759736}, \grantnum{NSF}{1544542},
+\grantnum{NSF}{1925644}, and \grantnum{NSF}{1910850}, the
+\grantsponsor{BC}{Czech Ministry of Education, Youth and Sports from
+  the Czech Operational Programme Research, Development, and
+  Education}{}, under grant agreement No.
+\grantnum{BC}{CZ.02.1.01/0.0/0.0/\-15\_003/0000421}, and the
+\grantsponsor{ELE}{European Research Council (ERC) under the European
+  Union's Horizon 2020 research and innovation programme}{}, under
+grant agreement No.  \grantnum{ELE}{695412}.
+\end{acks}
+
 \bibliography{bibexport}

### 2021-09-03 paper/main.tex
Fix typos and grammar in Section 8

@@
 
-The \eval functions are used widely and in varied ways in R. The function is an
+The \eval function is used widely, and in varied ways, in R. The function is an
 essential tool for language implementers. Our analysis observed that the base
-libraries heavily depend on \eval and any R program will end up calling it
+libraries heavily depend on \eval, and any R program will end up calling it
 through the core of the language. Independently developed libraries often use
 \eval in subtle ways, mostly to perform macro-programming tasks. Our review of
-code written by less experienced users suggest that \eval is exceedingly rarely
+code written by less experienced users suggests that \eval is exceedingly rarely
 used. So, it is fair to conclude that a majority of R programmers never have to
@@
 that calls this function. We have observed all sorts of side-effects with
-various degree of visibility.
+various degrees of visibility.
 
 While our results are not encouraging in general, we have observed many cases
-where \eval is used in a disciplined and predictable ways. While in the general
+where \eval is used in disciplined and predictable ways. While in the general
 case \eval is hell, there are many cases where it is just a function. It does
-not appear that a general purpose replacement for \eval is possible, in future
-work we hope to focus on special cases and propose tools that target particular
+not appear that a general-purpose replacement for \eval is possible; in future
+work, we hope to focus on special cases and propose tools that target particular
 subsets of call sites with some common properties.

### 2021-09-03 paper/main.tex
Fix typos and grammar in Section 7

@@
 \medskip\EE{$min(e)=\c{model.frame}$}\\[-2mm]\small The \k{model.frame} function
-returns a dataframe resulting from fitting the model described in a formula.
+returns a data frame resulting from fitting the model described in a formula.
 This shape subsumes \k{F(F(X))}, \k{FUN}, and assignments. It is the single most
@@
 In R, \eval is used chiefly for meta-programming and accessing remote
-environments. Unlike JavaScript which had a small set of well-defined patterns
-that could be rewritten without \eval, our results suggest that \eval is an
-integral part of programming in R.
+environments. Unlike JavaScript, which according to ~\citet{ecoop11} had a small
+set of well-defined patterns that could be rewritten without \eval, our results
+suggest that \eval is an integral part of programming in R.
 
@@
    eval(parse(text=paste0("z<-q.f(",dp,",...)")))
-\end{lstlisting}
+\end{lstlisting}\medskip
 
@@
   science. It consists of some of the most popular packages.} include their own
-implementation calls \k{rlang::eval\_tidy}~\cite{tidyverse}. This departs from
-\eval in two ways. First, it supports evaluation of quosures, custom objects
-used for metaprogramming that bundle expressions with an environment.
-\k{eval\_tidy} accepts a data mask argument, a set of bindings such as a
-dataframe or list, which takes precedence over the environment. Effects such as
-assignments happen in the data mask. Expressions such as \k{return()} do not
-work since the data mask does not correspond to a frame on the call stack.
+implementation of \eval called \k{rlang::eval\_tidy}~\cite{tidyverse}, which
+departs from \eval in two ways. First, it supports the evaluation of quosures,
+custom objects used for metaprogramming that bundle expressions with an
+environment. The \k{eval\_tidy} function accepts a data mask argument, a set of
+bindings such as a data frame or list, which takes precedence over the
+environment. Effects such as assignments happen in the data mask, and
+expressions such as \k{return()} do not work since the data mask does not
+correspond to a frame on the call stack.
 

### 2021-09-03 paper/main.tex
Fix typos and grammar in Section 2

@@
 
-\citet{ecoop12} gives a programming language-centric overview of the R language.
+\citet{ecoop12} give a programming language-centric overview of the R language.
 They characterized it as a lazy, vectorized, functional language with a rich
@@
 values can be tagged by attributes; these are key-value pairs. For instance, the
-attribute \k{dim}-\k{c(2,2)} can be attached to the value at \k{x} by
-\k{attr(x,"dim") <- c(2,2)}. Adding this attribute turns \k{x} into a
-matrix. The \k{class} attribute gives a value a 'class' in the object-oriented
-sense. So, \k{class(x)<-"human"} sets the class of \k{x} to \k{human}; classes
-are used for method dispatch. Every linguistic construct is desugared to a
-function call, even control flow statements, assignments and bracketing. All
-functions can be shadowed and redefined, making R at the same time remarkably
-flexible and exceedingly challenging to compile as vividly detailed
-by~\citet{dls19}. R uses a relaxed call-by-need convention for passing arguments
-to functions. Each argument is a thunk composed of an expression, its
-environment, and a slot for the result; these are called \emph{promises}. To get
-the value of an argument, the corresponding promise must be forced. Once forced,
-the promise's result is cached for future use.
+attribute \k{dim}-\k{c(2,2)} can be attached to the value bound to \k{x} by
+\k{attr(x,"dim") <- c(2,2)}. Adding this attribute turns \k{x} into a matrix.
+The \k{class} attribute gives a value a 'class' in the object-oriented sense.
+So, \k{class(x)<-"human"} sets the class of \k{x} to \k{human}; classes are used
+for method dispatch. Every linguistic construct is desugared to a function call,
+even control flow statements, assignments, and bracketing. All functions can be
+shadowed and redefined, making R at the same time remarkably flexible and
+exceedingly challenging to compile as vividly detailed by~\citet{dls19}. R uses
+a relaxed call-by-need convention for passing arguments to functions, studied in
+depth by~\citet{oopsla19b}. Each argument is a thunk composed of an expression,
+its environment, and a slot for the result; these are called \emph{promises}. To
+get the value of an argument, the corresponding promise must be forced. Once
+forced, the promise's result is cached for future use.
 
@@
 choices affect its expressive power. The key choices are the input format, the
-environment in which generated code evaluates and the reflective operations
+environment in which generated code evaluates, and the reflective operations
 available to that code. Table~\ref{comp} summarizes a few designs.
@@
 The input to \eval can be in any format convertible to code. JavaScript allows
-arbitrary strings, Julia and R are more restrictive as they require expressions
+arbitrary strings; Julia and R are more restrictive as they require expressions
 (or abstract syntax trees). Finally, Java is the most restrictive as its
@@
 visible from the current classloader. Julia limits \eval to the symbols visible
-in global environment, so does JavaScript's strict mode. Finally, R is the most
+in the global environment, so does JavaScript's strict mode. Finally, R is the most
 flexible as any accessible environment can be selected and passed to \eval. The
@@
 allow some form of introspection on the data that is visible in the environment
-in which \eval executes. R is more flexible as lets \eval inspect the program's
+in which \eval executes. R is more flexible as it lets \eval inspect the program's
 call stack as well as the code of any function. Thus, any environment and any
@@
 to expose as much of the language and its internals as possible in order to
-maximize expressivity. In R; \eval is a key tool to extend the language and
-implement DSLs, it is also a replacement for macros. By contrast, the designers
+maximize expressivity. In R, \eval is a key tool to extend the language and
+implement DSLs; it is also a replacement for macros. By contrast, the designers
 of Julia chose to limit \eval. In Julia, only global variables can be
@@
 facility~\cite{oopsla18a}. Furthermore, Julia provides a versioning mechanism to
-ensure that any method defined within an \eval only become visible at
+ensure that any method defined within an \eval only becomes visible at
 well-defined points and thus that optimized code does not have to be
@@
 Listings~\ref{lst:exprs} and \ref{lst:match.call} show some of the ways of
-creating expressions: parsing from a string, manually or via reflection. The
+creating expressions: parsing from a string, manually, or via reflection. The
 call to \k{substitute(x)} extracts the unevaluated expression from the promise
@@
 behaviors. Furthermore, JavaScript usage in 2011 is likely different from today,
-\eg Node.js was not covered. More details about dynamic analysis of JavaScript
+\eg Node.js was not covered. More details about the dynamic analysis of JavaScript
 can be found in~\cite{liang}.
@@
 
-\citet{Arceri21} study \eval in JavaScript from a software security point of
-view. The authors report that 53\% of the malware they studied used \eval as a
-means to obfuscate attack code or mount attacks. They propose an abstract
+\citet{Arceri21} studied \eval in JavaScript from a software security point of
+view. The authors reported that 53\% of the malware they studied used \eval as a
+means to obfuscate attack code or mount attacks. They proposed an abstract
 interpretation-based approach to analyzing dynamic languages. One must construct
@@
 In this work, we significantly expand that study. First, our corpus covers
-\CranPackages packages (about the whole of CRAN at time of writing) with
-\PkgHitEvalCallSites \eval call sites and \packageAllcallsRnd \eval calls.
-Next, we track a number of information such as the shapes of expressions
-passed to \eval, their provenance, complexity, and side effects. This enables
-us to have an accurate and detailed picture of the use of \eval in the R
-ecosystem. We use this to provide both quantitative and qualitative views on
-\eval, including several use cases to demonstrate the popularity, diversity and
-powerfulness of \eval in R.
+\CranPackages packages (about the whole of CRAN at the time of writing) with
+\PkgHitEvalCallSites \eval call sites, and \packageAllcallsRnd \eval calls.
+Next, we track a lot more information such as the shapes of expressions passed
+to \eval, their provenance, complexity, and side effects. This enables us to
+have an accurate and detailed picture of the use of \eval in the R ecosystem. We
+use this to provide both quantitative and qualitative views on \eval, including
+several use cases to demonstrate the popularity, diversity and powerfulness of
+\eval in R.
 
@@
 
-% TODO: remove - does not bring anything new or interesting
-% As a last example, \k{configr::config.funs.par} constructs three expressions
-% that are then fed to \eval. Semantically equivalent code would use \k{get}
-% to lookup \k{fun}, and then directly execute the three expressions.
-
-% \begin{lstlisting}
-%  config.funs.par <- function(fun="",...) {
-%    args.all <- as.list(match.call())
-%    args.all <- args.all[names(args.all) != ""]
-%    parameters <- list()
-%    t1 <- sprintf("parameters <- names(as.list(args(%s)))", fun)
-%    t2 <- "parameters <- parameters[parameters != '']"
-%    t3 <- "parameters <- args.all[names(args.all) %in% parameters]"
-%    for (j in c(t1,t2,t3))
-%     eval(parse(text = j))
-%    return(parameters)
-%  }
-% \end{lstlisting}
-
 We have not yet touched on alternative implementations of \eval. The tidyverse

### 2021-09-03 paper/main.tex
Fix typos and grammar in Section 5

@@
 minimization function which abstracts some of the incidental details of
-expressions and return a ``shape'' that can be used to group similar \evals. The
-function $min(e)$ which for a given expression $e$ returns a normal form. Normal
+expressions and returns a ``shape'' that can be used to group similar \evals.
+The function $min(e)$, for a given expression $e$, returns a normal form. Normal
 forms use \k{V} to stand in for values occurring in expression, \k{X} stands in
 for variables, and \k{F} stands for functions. The function performs constant
-folding of arithmetic and string expressions over base operators, for instance
+folding of arithmetic and string expressions over base operators; for instance,
 $min(\c{1+1})$ simplifies to $\c{V}$, on the ground that addition of values
 likely returns a value. Value simplification takes $min(\c{c(1,2,3+2)})$ to
-$\c{V}$ as a complex vector containing values is a value. Variable absorption
-has $min(\c{x+y})$ becomes $\c{X}$, this motivated by the fact that addition is
-not an ``interesting'' operation, and that \k{X} stands in for any number of
+$\c{V}$, as a complex vector containing values is a value. Variable absorption
+has $min(\c{x+y})$ become $\c{X}$; this is motivated by the fact that addition is
+not an ``interesting'' operation and that \k{X} stands in for any number of
 variable lookups. Function absorption simplifies nested functions to keep an
 abstraction of the nesting $min(\c{g(f(x),h(z))})=\c{F(F(X))}$. There are other
-simplifications, the full list can be found in our artifact. It is worth
+simplifications; the full list can be found in our artifact. It is worth
 mentioning that these simplifications are heuristics; in R, \k{1+1} is not
-necessary \k{2}. The addition operator can be redefined, but it typically isn't
+necessary \k{2}. The addition operator can be redefined, but it typically isn't,
 or at least not in a way that would invalidate arithmetic.
 
-Table~\ref{tab:minimizedexpressions} gives the 10 most frequent shapes with the
+Table~\ref{tab:minimizedexpressions} gives the ten most frequent shapes with the
 number and ratio of sites that received arguments of that shape. Operations (ops) is
 the median count of instructions performed by the interpreter when evaluating
-argument of the given shape. Envir is the ratio of sites that evaluate in a
+an argument of the given shape. Envir is the ratio of sites that evaluate in a
 function environment. The last column has a sample expression that normalizes to
@@
   \medskip\EE{$min(e)=\c{V}$}\\[-2mm]\small Values occur in 16\% of sites, 74\% of
-  those are constants such as integers, the remainder evaluate to a value (\packageNbCallSitesUniqueActualValue of
-  sites only ever see a value). \Eval
+  those are constants such as integers, the remainder evaluate to a value (\packageNbCallSitesUniqueActualValue of sites only ever see a value). \Eval
   is needed when the expression denoting the values is constructed
@@
   another, more interesting, path that requires evaluation.
-  This shape usually run in few interpreter steps.
-  The environment in which they evaluate is mostly irrelevant.
+  This shape usually runs in few interpreter steps.
+  The environment in which these expressions evaluate is mostly irrelevant.
 \end{tabular}
@@
   \packageNbSymbolVarSitePercent are simple variable names, e.g. \k{x}. The shapes
-subsumes \k{V} and built-in arithmetic operations, so a mixture of arithmetics
-and variables normalize to \k{X}. A lookup is one step of execution, if a promise
-is returned an arbitrary number of additional steps may be needed; the median is
+  subsume \k{V} and built-in arithmetic operations, so a mixture of arithmetics
+  and variables normalize to \k{X}. A lookup is one step of execution; if a promise
+  is returned, an arbitrary number of additional steps may be needed. The median number of steps is
   \packageMinimizedmedianoperationsaRnd, suggesting that most variable reads do
   little work. Variables are often evaluated in environments that have been
-constructed programmatically, \packageMinimizedpercentparentframesa of
+  constructed programmatically; \packageMinimizedpercentparentframesa of
   expressions are evaluated in a function environment.
@@
   with the dollar operator, \eg~\k{x\$f}, and vector indexing, \eg~\k{x[42]} and
-\k{x[[24]]}. This takes a few more steps of evaluation on average, and is
+  \k{x[[24]]}. This takes a few more steps of evaluation on average and is
   typically used in a function environment.
@@
   \\
-\medskip\EE{$min(e)=\c{F()}$}\\[-2mm]\small Function calls without variable lookups or
-assignment. A variable is allowed in the function position, looking up
+  \medskip\EE{$min(e)=\c{F()}$}\\[-2mm]\small This shape represents function calls without variable lookups or
+  assignment. A variable is allowed in the function position; looking up
   functions does not trigger computation unless the function's name
@@
 \begin{tabular}{@{}p{.97\linewidth}}
-\medskip\framebox{$min(e)=\c{F(X)}$}~\EE{$min(e)=\c{F(F(X))}$}\\[-2mm]\small Function
+  \medskip\framebox{$min(e)=\c{F(X)}$}~\EE{$min(e)=\c{F(F(X))}$}\\[-2mm]\small This shape corresponds to function
   calls whose arguments may include variable references and nested calls. Together
@@
   \\
-\medskip\EE{$min(e)=\c{FUN}$}\\[-2mm]\small Function definitions without any
+  \medskip\EE{$min(e)=\c{FUN}$}\\[-2mm]\small This shape represents function definitions without any
   computation, \eg~\k{function(x) x+1}. In addition to \k{FUN},
-\packageGeneralizedFunctionDefinitionSitesPercent of sites have function
+  \packageGeneralizedFunctionDefinitionSitesPercent sites have function
   definitions nested in other expressions.
@@
 \begin{tabular}{@{}p{.97\linewidth}}
-\medskip\EE{$min(e)=\c{BLOCK}$}\\[-2mm]\small Multi-statement code blocks, as
-they can be large we do not try to normalize their contents. They execute in a
-median \packageMinimizedmedianoperationsjRnd steps, usually, in function
+  \medskip\EE{$min(e)=\c{BLOCK}$}\\[-2mm]\small These are multi-statement code blocks; as
+  they can be large, we do not try to normalize their contents. They execute in a
+  median \packageMinimizedmedianoperationsjRnd steps, usually in function
   environments.
   Essentially, the block denotes a fragment of a program to be evaluated in a
-particular way and in a particular environment. The latter is what makes it
-different from 0-argument closure. There is a number of use cases: unit testing
+  particular way and a particular environment. The latter is what makes it
+  different from a zero-argument closure. There are several use cases: unit testing
   frameworks (\eg \k{testthat}, \k{testit}), code benchmarking (\eg
@@
 different patterns of use, there are also many different shapes. In comparison
-with R, JavaScript's \eval usage was more straightforward and more predictable
+with R, JavaScript's \eval usage was more straightforward and more predictable,
 as reported by \citet{oopsla12b}, with 98\% of sites with only one shape. On the
 other hand, there are \packageNbSimpleMinimizedOne sites that only receive one
-of the simple shape (\ie\xspace \k{X}, \k{V}, \k{\$} or \k{<-}), and
+of the simple shapes (\ie\xspace \k{X}, \k{V}, \k{\$} or \k{<-}), and
 \packageNbSimpleMinimizedMore sites get multiple simple shapes. Among these
@@
 
-Contrary to JavaScript, in R the environment of \eval can be specified by its
+Contrary to JavaScript, in R, the environment of \eval can be specified by its
 \k{env} argument. This gives users control over what is visible to the
@@
 Table~\ref{tab:highlevelenvironments} shows that most calls evaluate in the
-scope of a function, global is a distant second. This means that most reads and
+scope of a function; global is a distant second. This means that most reads and
 writes act on local variables. But of which function? Table~\ref{tab:funoffset}
@@
 textually occurs are read and written. About 1.5\% of sites evaluate their
-argument three frames up the call stack, or above. Modular reasoning is thus
+argument three frames up the call stack or above. Modular reasoning is thus
 impossible in R. Since the actions of \eval happen at a distance, understanding
@@
 are accidental. They arise from the fact that our corpus consists of many code
-snippets that are run at the top-level, global is thus the caller of \eval. This is
+snippets that are run at the top-level; global is thus the caller of \eval. This is
 noteworthy because writes to global variables are visible to all functions and
@@
 mostly read, and very rarely, delete variables. There are also some cases where
-new variables are injected in a function. Usually, this in the direct caller,
+new variables are injected in a function. Usually, this is in the direct caller,
 but it can sometimes take effect several frames up the call stack. There is
@@
 Synthetic environments are relatively frequent. Common uses-cases include
-the evaluation of an expression in a sandbox, or using a data structure as
-environment. For example one could evaluate a method of an object and use an
+the evaluation of an expression in a sandbox or using a data structure as
+environment. For example, one could evaluate a method of an object and use an
 environment to hold the object's fields.
@@
 This is probably for the best as mutating the bindings of a loaded package is
-frowned upon and R tries to discourage it.
+frowned upon, and R tries to discourage it.
 
@@
 Where does the expression passed to \eval come from? There are various means of
-creating an expression, often these means correlate with a particular use case.
+creating an expression; often these means correlate with a particular use case.
 Our analysis records the values returned by some functions of interest as well
@@
 member of assignments until reaching a starting node. It represents the origin that is
-further modified to yield the expression passed to \eval. In our example, that
+further modified to yield the expression passed to \eval. In this example, that
 origin is \k{parse}.
@@
 
-To have a high-level views of provenance, we classify functions into the
+To have a high-level view of provenance, we classify functions into the
 following five categories:
@@
 \paragraph{Discussion}
-The main use cases for \eval involve using \k{match.call}, or \k{substitute} to access and transform arguments of a function. Constructing
-expressions from strings is also a common use case. The constructed category
-shows that most \evals come from existing code that was modified by the
-programmer before invoking \eval. Both constructed and reflection categories
-roughly correspond to meta-programming. Some uses could be replaced by
-macros if the R designers could be convinced to overcome their distaste for
-those.
+The main use cases for \eval involve using \k{match.call}, or \k{substitute} to
+access and transform arguments of a function. Constructing expressions from
+strings is also a common use case. The constructed category shows that most
+\evals come from existing code that was modified by the programmer before
+invoking \eval. Both constructed and reflection categories roughly correspond to
+meta-programming. Some uses could be replaced by macros if the R designers could
+be convinced to overcome their distaste for those.
 
@@
 \Eval may perform side-effects. We care about effects that are visible after a
-call finishes, \ie variable definitions, updates, and removals. Knowing in which
-environments these side-effects happen, can help determine the impact of \eval
-on our ability to reason about the code. Our analysis records information about
-every environment. From the recorded data, we discard side-effects coming from
-unit testing frameworks as they use \eval to run all their tests and thus
+call finishes, \ie variable definitions, updates, and removals. Knowing the
+environments in which these side-effects happen can help determine the impact of
+\eval on our ability to reason about the code. Our analysis records information
+about every environment. From the recorded data, we discard side-effects coming
+from unit testing frameworks as they use \eval to run all their tests and thus
 dominate the side-effect data.\footnote{The corpus has \k{RUnit, testthat,
@@
 Most sites, \SESitesInOneClass, do all their side-effects consistently in one
-kind of environments. The same happens at the function level. Almost half of the
+kind of environment. The same happens at the function level. Almost half of the
 sites (over a third of the functions) do side effects in either \emph{Local} or
@@
 Table~\ref{tab:se-types} shows recorded effects. There are over 5M updates. In
-terms of calls, we see assignments primarily and in terms of site definitions.
-This is expected. A subsequent \eval call will turn a definition into an update.
-The most dangerous side effect is variable removal, as it means that after a
-call to \eval some binding in some environments will disappear. While rare, this
+terms of calls, we primarily see assignments, followed by definitions. This is
+expected. A subsequent \eval call will turn a definition into an update. The
+most dangerous side effect is variable removal, as it means that after a call to
+\eval some binding in some environments will disappear. While rare, this
 happens, but the vast majority comes from a single site
@@
 majority of \eval sites involve basic R vectors (\SEBasicTypeRatio) and lists
-(\SEListTypeRatio). From the compiler perspective, we would like to know how
-many sites change function bindings. In the corpus, this happens in only
+(\SEListTypeRatio). From the perspective of a compiler, we would like to know
+how many sites change function bindings. In our corpus, this happens in only
 \SEClosureType sites; \SEClosureTypeLocal of them do that in the local
-environment. We have manually inspected a few of these sites, but except for
-manual injection of parameters into a \k{model.frame} execution environment, we
-did not find a common use case.
+environment. We manually inspected a few of these sites; except for manual
+injection of parameters into a \k{model.frame} execution environment, we did not
+find a common use case.
 
@@
 Out of the minimized expressions, it is \k{BLOCK}, \k{<-} and \k{F(F(X))} that
-contribute to the vast majority of side effects. Concretely, 92\% of \k{BLOCK},
-87\% of \k{<-} and 15\% of \k{F(F(X))} expressions do side effects. In the case
+contributes to the vast majority of side effects. Concretely, 92\% of \k{BLOCK},
+87\% of \k{<-}, and 15\% of \k{F(F(X))} expressions do side effects. In the case
 of \k{BLOCK}, most of them happen in the environment passed to the \eval call
-(default is the parent environment). For \k{<-} it is 15\%. However, this form
-is dominated by the \k{plyr::allocate\_column} function that side-effects in
-the environment containing the data frame to which it allocates new column.
+(default is the parent environment). For \k{<-} it is 15\%. This form is
+dominated by the \k{plyr::allocate\_column} function that side-effects in the
+environment containing the data frame to which it allocates a new column.
 Without this call, \k{<-} is similarly predictable with 96\% of side effects
@@
 
-\paragraph{Discussion} In JavaScript assignments in \eval can happen in either
+\paragraph{Discussion} In JavaScript, assignments in \eval can happen in either
 local scope or less often, when called through an alias, in the global scope.
 In R, given the support for first-class environments, it can happen anywhere,
-making it \eval more dangerous than it already is. However, the data suggest
+making \eval more dangerous than it already is. However, the data suggest
 that first, side-effects from \eval are not as widespread as in the case of

### 2021-09-04 paper/main.tex
Replace \c with \k
Fixing typo with soundiness

@@
 \author{Filip Křikava}\affiliation{\institution{Czech Technical University in Prague}\country{Czech Republic}}
-\author{Christoph Kirsch}\affiliation{\institution{Czech Technical University in Prague}\country{Czech Republic}}
+\author{Christoph M.~Kirsch}\affiliation{\institution{University of Salzburg}\country{Austria}}
+\affiliation{\institution{Czech Technical University in Prague}\country{Czech Republic}}
 \author{Jan Vitek}\affiliation{\institution{Northeastern University}\country{USA}}
@@
 practical usage suggested that most reflective calls were relatively harmless.
-While this backed up the soundiness squad's approach, does it generalize to
+While this backed up the soundiness of squad's approach, does it generalize to
 other application domains and to other languages?
@@
 
-\begin{lstlisting}[caption={\c{AdapSamp::rARS}}, captionpos=b]
+\begin{lstlisting}[caption={\k{AdapSamp::rARS}}, captionpos=b]
  function(n,formula) {

### 2021-09-08 paper/main.tex
Revert "Fixing typo with soundiness"
This reverts commit 21069d17293f640c9a5db6d96c6e217c3d8d8aef.
ACM-related info. Still to choose: copyright
Remove "Listing i" for all the listings in Section 6
Wrap some figures

@@
 
-%%% The following is specific to OOPSLA '20 and the paper
-%%% 'World Age in Julia: Optimizing Method Dispatch in the Presence of Eval'
-%%% by Julia Belyakova, Benjamin Chung, Jack Gelinas, Jameson Nash, Ross Tate, and Jan Vitek.
+%%% The following is specific to OOPSLA '21 and the paper
+%%% 'Why We Eval in the Shadows'
+%%% by Aviral Goel, Pierre Donat-Bouillud, Filip Křikava, Christoph M. Kirsch, and Jan Vitek.
 %%%
-\setcopyright{rightsretained}
+%\setcopyright{ACMUNKNOWN} %TODO: we still have to choose the copyright
 \acmPrice{}
-\acmDOI{10.1145/3428275}
-\acmYear{2020}
-\copyrightyear{2020}
-\acmSubmissionID{oopsla20main-p384-p}
+\acmDOI{10.1145/3485502}
+\acmYear{2021}
+\copyrightyear{2021}
+\acmSubmissionID{oopsla21main-p121-p}
 \acmJournal{PACMPL}
-\acmVolume{4}
+\acmVolume{5}
 \acmNumber{OOPSLA}
-\acmArticle{207}
-\acmMonth{11}
+\acmArticle{125}
+\acmMonth{10}
 
@@
 
+\usepackage{caption}
+
 \begin{document}
@@
 practical usage suggested that most reflective calls were relatively harmless.
-While this backed up the soundiness of squad's approach, does it generalize to
+While this backed up the soundiness squad's approach, does it generalize to
 other application domains and to other languages?
@@
 
+\begin{wrapfigure}{r}{7cm}
+	\includegraphics[width=7cm]{pkgs-eval-callsites-hist.pdf}
+	\caption{ \eval call sites}\label{fig:pkgs-eval-callsites-hist}
+\end{wrapfigure}
+
 There are \PkgEvalCallSites \eval sites in \PkgPackages packages;
@@
 \CranFunsWithEvalRatio of all functions use \eval.
-
-\begin{figure}[!h]
-   \includegraphics[width=7cm]{pkgs-eval-callsites-hist.pdf}
-  \caption{ \eval call sites}\label{fig:pkgs-eval-callsites-hist}
-\end{figure}
 
@@
 
-\begin{table}[!h]
+\begin{table}[H]
 \centering
@@
 \tt 1K &--& \tt 10K    & \packageBind  & \tt 10M &--& \tt 100M & \packageBinh\\\bottomrule
-\end{tabular}
-\caption{Call frequency}\label{tab:freq}
+\end{tabular}\caption{Call frequency}\label{tab:freq}
 \end{minipage}
@@
 \begin{tabular}{@{}r|r@{~}r@{}r@{~}r@{}}\toprule
-  &\eval & \k{evalq} & \k{eval} & \k{local}\\[-1.5mm]
-           & & & \k{.parent} &\\\midrule
+  &\eval & \k{evalq} & \k{eval} & \k{local}\\[-1.3mm]
+           & & & \k{.parent} &\\[1.3mm]\midrule
 \small Static sites &\packageStaticeval&\packageStaticevalq&\packageStaticevalparent&\packageStaticlocal \\
@@
 \small Invocations&\packageEvalsRnd&\packageEvalqsRnd&\packageEparentsRnd&\packageLocalsRnd\\\bottomrule
-\end{tabular}~\\[2mm]\caption{Variants}\label{tab:variantseval}
+\end{tabular}\caption{Variants}\label{tab:variantseval}
 \end{minipage}\end{tabular}
-
 \end{table}
-\begin{figure}[H]
+
+\begin{figure}[h]
 	\includegraphics[width=.7\textwidth]{traced-eval-callsites.pdf} \centering
@@
 
-\begin{table}[!h]  \small  \centering
+\begin{wraptable}{r}{5cm} \vspace*{-2mm}\ \small  \centering
   \begin{tabular}{r@{\,}r@{\,}l@{\,}r|r@{\,}r@{\,}l@{}r} \toprule
@@
   \caption{Normalized calls} \label{tab:cn}
-\end{table}
+\end{wraptable}
 
@@
 
-\begin{figure}[!h]
-\begin{tabular}{c}
-{\includegraphics[width=0.8\textwidth]{package_size_loaded_distribution}}
-\end{tabular}
+\begin{wrapfigure}{r}{6.2cm}
+\hspace{-25mm} \includegraphics[width=0.8\textwidth]{package_size_loaded_distribution}
 \caption{Loaded code} \label{fig:sizedistribution}
-\end{figure}
+\end{wrapfigure}
 
@@
 
+% To remove "Listing i" in the captions from this point, as we do not refer to them by number
+% We define and refer to listings in 2.2 so we need to put that captionsetup only here
+\captionsetup[lstlisting]{labelformat=empty} 
 \begin{lstlisting}[caption={\k{AdapSamp::rARS}}, captionpos=b]

### 2021-09-08 paper/main.tex
Even better layout

@@
 \begin{wrapfigure}{r}{7cm}
-	\includegraphics[width=7cm]{pkgs-eval-callsites-hist.pdf}
+	\hspace*{-1cm}\includegraphics[width=9cm]{pkgs-eval-callsites-hist.pdf}
 	\caption{ \eval call sites}\label{fig:pkgs-eval-callsites-hist}
@@
 
-\begin{wraptable}{r}{5cm} \vspace*{-2mm}\ \small  \centering
+\begin{wraptable}{r}{5cm}  \small  \centering
   \begin{tabular}{r@{\,}r@{\,}l@{\,}r|r@{\,}r@{\,}l@{}r} \toprule
@@
   \caption{Normalized calls} \label{tab:cn}
+  \vspace{-0.5cm}
 \end{wraptable}
@@
 \begin{wrapfigure}{r}{6.2cm}
-\hspace{-25mm} \includegraphics[width=0.8\textwidth]{package_size_loaded_distribution}
+\hspace{-3.3cm} \includegraphics[width=0.9\textwidth]{package_size_loaded_distribution}
 \caption{Loaded code} \label{fig:sizedistribution}

### 2021-09-09 paper/main.tex
6

@@
     else,
-    function
+    function,
+    break,
+    in
   },
@@
 the corpus. We present examples from nine classes from this batch. The
-classification is not exhaustive; we present results from only 117 of the total
-\PkgEvalCallSites call sites in our corpus, and many uses of \eval do not fit in
-a specific category. The \eval usage patterns are described below.
+classification is not exhaustive as we only study a fraction from the
+\PkgEvalCallSites call sites in our corpus, and many uses of \eval do not fit
+in a specific category. Instead, the goal is to demonstrate the wide variety of
+uses to which \eval is put to use. At the high level, the usage can be grouped
+into two groups: one contains patterns that aim to provide better
+API for R users and the other focuses on simplifying implementation.
+
+\subsection{Better API}
+
+This category contains \eval used to develop a better API in terms of
+ergonomics, extensibility, and ease of use. This is important given
+that R is often being used interactively for data exploration.
+
+\paragraph{Non-standard Scoping} \Eval allows functions to resolve symbols in
+different environments than the current one. For example, the \k{data.table}
+package redefines the subsetting operator, \k{[}\footnote{Defined as a triplet
+  \texttt{[i, j, k]}, where expression \texttt{j} is used to calculate a subset
+of rows from \texttt{i} grouped by \texttt{k}.}, to enable compact queries on
+data frames. The example below shows the use of this operator on the
+\k{flights} data frame to compute the average departure delays of Delta flights
+in 2014 by the origin airport. The symbols \k{carrier}, \k{year},
+\k{dep_delay}, and \k{origin} are not looked up in the current environment but
+in the \k{flights} data frame.
+
+\begin{lstlisting}[caption={\k{data.table::[}}, captionpos=b]
+> flights[carrier=="DL" & year == 2014, mean(dep_delay), by=origin]
+#    origin       V1
+# 1:    LGA 11.35083
+# 2:    EWR 14.96171
+# 3:    JFK 12.45414
+\end{lstlisting}\medskip
+
+At its core, the \k{[} operator uses \k{substitute} to capture its operand
+expressions, transforms them, and evaluates them using \eval in the given data
+frame. We omit the implementation as it spans over 1700 lines at the time of
+this writing\footnote{Its implementation can be found at:
+  \url{https://github.com/Rdatatable/data.table/blob/master/R/data.table.R}}.
+
+Non-standard scoping is an important use case of \eval in R. It enables the
+design of functions that require less typing and syntactic noise at the cost of
+some ambiguity. Since R is used primarily interactively, this trade-off is
+justifiable.
+
+\paragraph{Domain-Specific Language} A domain-specific language leverages R's
+grammar but redefines its semantics in a suitable way. The example below
+illustrates string interpolation. The \k{glue} function from the \k{glue}
+package extracts snippets of code enclosed between braces, evaluates them using
+\eval, and splices their results to construct the result.
+
+\begin{lstlisting}[caption={\k{glue::glue}}, captionpos=b]
+> greeting <- "Hello"
+> glue('{greeting} World!')
+# Hello World!
+\end{lstlisting}\medskip
+
+The \k{glue} package contains a variant of \k{glue} function that leverages
+non-standard scoping for interpolation in a custom environment. For example,
+the following snippet pipes the subsetted data frame into the \k{glue_data}
+function used to resolve the symbols for string interpolation.
+
+\begin{lstlisting}[caption={\k{glue::glue_data}}, captionpos=b]
+> flights[carrier=="DL" & year == 2014, mean(dep_delay), by=origin] |>
+    glue_data("{origin} has average delay of {round(V1)} min")
+# LGA has average delay of 11 min
+# EWR has average delay of 15 min
+# JFK has average delay of 12 min
+\end{lstlisting}
 
@@
 extend the behavior of a library function. \Eval allows clients to provide a
-simple text string that is executed in the context of the function. The
-example here allows users to provide formulas as strings, such as
-\k{"exp(-x^2/2)"}, the contract being that the text can refer to some variable
-\k{x} that will be set up by the called function.\footnote{The function is used
-to generate a sequence of random numbers using the adaptive rejection sampling
-algorithm and the string formula is used to describe the target density.}
+simple text string that is executed in the context of the function. The example
+here allows users to provide formulas as strings, such as \k{"exp(-x^2/2)"},
+the contract being that the text can refer to some variable \k{x} the called
+function will set up.\footnote{The function is used to generate a sequence of
+  random numbers using the adaptive rejection sampling algorithm and the string
+formula is used to describe the target density.}
 
@@
    p <- function(x) eval(parse(text=formula))
-\end{lstlisting}
-
-\paragraph{Code Generation}
-\Eval can be used to generate boilerplate code that would either be tedious to
-write by hand or that is only known after deployment. The following code snippet
-for the \k{Rcpp} package uses \eval to automatically generate R bindings to C++
-methods.
-
-\begin{lstlisting}[caption={\k{Rcpp::.makeCppMethods}}, captionpos=b]
- function(methods,env) {
-   for(what in cppMethods)
-     methods[[what]] <-
-       eval(substitute(function(...)CppObject$what(...)), env)
-\end{lstlisting}
-
-\paragraph{Extracting Varargs}
-\Eval is used to obtain the source code of the arguments passed to a vararg
-(\k{...}) parameter. In the following example, \k{substitute(alist(...))} yields
-an expression representing a call to \k{alist} with the \k{...} expanded to the
-arguments supplied to the function. \Eval evaluates this expression and returns
-a \k{list} of the unevaluated argument expressions.
-
-\begin{lstlisting}[caption={\k{statnet.common::NVM}}, captionpos=b]
+   ...
+ }
+\end{lstlisting}
+
+\paragraph{Extracting Varargs} \Eval is used to obtain the source code of the
+arguments passed to a vararg parameter (\k{...}). Concretely,
+\k{substitute(alist(...))} yields an expression representing a call to
+\k{alist} with the \k{...} expanded to the arguments supplied to the function.
+Passing this to \eval returns a list of the unevaluated argument
+expressions.
+In the following example, it is used to find and return the value of the first expression that evaluates to non-null value.
+The extracting varargs pattern is needed to avoid premature promise forcing.
+
+\begin{lstlisting}[caption={\k{statnet.common::NVL}}, captionpos=b]
  function(...) {
-   exprs <- eval(substitute(alist(...)))
-\end{lstlisting}\medskip
-
-\paragraph{Controlled Evaluation}
-\Eval can be used to control when and if expressions are evaluated. For example,
-\k{assertthat} provides users with the ability to make assertions and, if they
-fail, produce legible error messages. The \k{assertthat::see_if} function
-accepts a vararg with a list of expressions. It extracts these expressions
-(\emph{Extracting Varargs} pattern) and evaluates them sequentially in the
-caller's environment using \k{eval}. If an expression does not hold, an error is
-generated.
-
-\begin{lstlisting}[caption={\k{assertthat::see\_if}}, captionpos=b]
+   for (e in eval(substitute(alist(...)))) {
+     if (!is.null(eval(e, parent.frame()))) break
+   }
+   e
+ }
+\end{lstlisting}
+
+\paragraph{Controlled Evaluation} \Eval can be used to control when and if
+expressions are evaluated. For example, \k{assertthat} provides users with the
+ability to make assertions and, if they fail, produce legible error messages.
+The \k{assertthat::see_if} function accepts a vararg with a list of
+expressions. It extracts these expressions using the extracting varargs
+pattern and evaluates them sequentially in the caller's environment using
+\k{eval}. If an expression does not hold, an error is generated.
+
+\begin{lstlisting}[caption={\k{assertthat::see_if}}, alsoletter={\_}, captionpos=b]
  function(..., env=parent.frame(), msg=NULL) {
@@
      }, assertError=function(e) { structure(FALSE, msg=e$message) })
-\end{lstlisting}\medskip
-
-\paragraph{Shadowing}
-Some packages use \eval to implement variations of the \k{base} R functions. For
-example, the \k{igraph} package provides the \k{do_call} function shown below.
-This is a variant of the builtin \k{do.call} function with different semantics
-for resolving the function name. It constructs a call from a name and a list of
-arguments and passes it to \eval.
+   ...
+ }
+\end{lstlisting}
+
+\paragraph{Shadowing} Some packages use \eval to implement variations of the
+\k{base} R functions. For example, the \k{igraph} package provides the
+\k{do_call} function shown below. This is a variant of the builtin \k{do.call}
+function with different semantics for resolving the function name. It
+constructs a call from a name and a list of arguments and passes it to \eval.
 
@@
  }
-\end{lstlisting}\medskip
-
-\paragraph{Boilerplate Removal}
-A variant of code generation is to let users write compact code that is expended
-via \eval and \k{match.call}. The example uses \k{match.call} to access the
-expression with which the function is called. The following three lines construct a
-call to \k{stats::model.frame} with argument text of parameters \k{'formula'},
-\k{'data'}, and \k{'weights'}. Then, the call expression is evaluated in the
-parent environment. The use of \k{match.call} avoids repeated use of
-\k{substitute} for extracting the argument expressions. It is extensible as new
-parameter names can be easily added. This example illustrates one of the most
-recurring uses of \eval by packages that do statistical modeling, and relates to
-the \k{model.frame} shape in Table~\ref{tab:minimizedexpressions}.
+\end{lstlisting}
+
+\subsection{Implementation simplification}
+
+The patterns in this category use \eval to simplify implementation in terms of
+avoiding repetitive and boilerplate code.
+
+\paragraph{Code Generation} \Eval can be used to generate code that would
+either be tedious to write by hand or that is only known after deployment. The
+following code snippet for the \k{Rcpp} package uses \eval to automatically
+generate R bindings to C++ methods.
+
+\begin{lstlisting}[caption={\k{Rcpp::.makeCppMethods}}, captionpos=b]
+ function(methods, env) {
+   for(what in cppMethods)
+     methods[[what]] <-
+       eval(substitute(function(...)CppObject$what(...)), env)
+   ...
+ }
+\end{lstlisting}
+
+\paragraph{Boilerplate Removal} A variant of code generation is to let users
+write compact code that is expended via \eval and \k{match.call}. The example
+uses \k{match.call} to access the expression with which the function is called.
+The following three lines construct a call to \k{stats::model.frame} with
+argument text of parameters \k{'formula'}, \k{'data'}, and \k{'weights'}. Then,
+the call expression is evaluated in the parent environment. The use of
+\k{match.call} avoids repeated use of \k{substitute} for extracting the
+argument expressions. It is extensible as new parameter names can be easily
+added. This example illustrates one of the most recurring uses of \eval by
+packages that do statistical modeling, and relates to the \k{model.frame} shape
+in Table~\ref{tab:minimizedexpressions}.
 
@@
    mf <- eval.parent(temp)
-\end{lstlisting}\medskip
+   ...
+ }
+\end{lstlisting}
+\medskip
 
@@
  }
-\end{lstlisting}\medskip
-
-\paragraph{Non-standard Scoping}
-
-\Eval allows functions to resolve symbols in different environments than the
-current one. For example, the \k{data.table} package redefines the subsetting
-operator, \k{[}\footnote{Defined as a triplet \texttt{[i, j, k]}, where
-  expression \texttt{j} is used to calculate a subset of rows from \texttt{i}
-  grouped by \texttt{k}.}, to enable compact queries on data frames. The example
-below shows the use of this operator on the \k{flights} data frame to compute
-the average departure delays of Delta flights in 2014 by the origin airport. The
-symbols \k{carrier}, \k{year}, \k{dep_delay} and \k{origin} are not looked up in
-the current environment, but in the \k{flights} data frame.
-
-\begin{lstlisting}[caption={\k{data.table::[}}, captionpos=b]
-> flights[carrier=="DL" & year == 2014, mean(dep_delay), by=origin]
-#    origin       V1
-# 1:    LGA 11.35083
-# 2:    EWR 14.96171
-# 3:    JFK 12.45414
-\end{lstlisting}\medskip
-
-At its core, the \k{[} operator uses \k{substitute} to capture its operand
-expressions, transforms them, and evaluates them using \eval in the given data
-frame. We omit the implementation as it spans over 1700 lines at the time of
-this writing\footnote{Its implementation can be found at:
-  \url{https://github.com/Rdatatable/data.table/blob/master/R/data.table.R}}.
-
-Non-standard scoping is an important use case of \eval in R. It enables the
-design of functions that require less typing. Since R is used primarily
-interactively, this trade-off is justifiable.
-
-\paragraph{Domain-Specific Language}
-A domain-specific language leverages R's grammar but redefines its semantics in
-a suitable way. The example below illustrates string interpolation. The \k{glue}
-function from the \k{glue} package extracts snippets of code enclosed between
-braces, evaluates them using \eval, and splices their results to construct the
-result.
-
-\begin{lstlisting}[caption={\k{glue::glue}}, captionpos=b]
-> greeting <- "Hello"
-> glue('{greeting} World!')
-# Hello World!
-\end{lstlisting}\medskip
-
-The \k{glue} package contains a variant of \k{glue} function that leverages
-non-standard scoping for interpolation in a custom environment. For example, the
-following snippet pipes the subsetted data frame into the \k{glue_data} function
-that uses it to resolve the symbols for string interpolation.
-
-\begin{lstlisting}[caption={\k{glue::glue_data}}, captionpos=b]
-> flights[carrier=="DL" & year == 2014, mean(dep_delay), by=origin] |>
-    glue_data("{origin} has average delay of {round(V1)} min")
-# LGA has average delay of 11 min
-# EWR has average delay of 15 min
-# JFK has average delay of 12 min
-\end{lstlisting}\medskip
-
-\paragraph{Obfuscation}
-A few packages use \eval to bypass the restrictions imposed by the \k{R CMD
-  CHECK} tool. This tool enforces some well-formedness rules on a submitted
-package before accepting it for inclusion in CRAN. Some static checks ensure
-that the package code does not use certain restricted functions. Package authors
-use \eval to obfuscate their code to get around this limitation. The example
-below mutates the variable \k{s} in the package environment. However, that is
-locked by default and unlocking it requires \k{unlockBinding} which is a
-restricted function. To work around this, the call to \k{unlockBinding} is done
-through \eval.
+\end{lstlisting}
+
+\paragraph{Obfuscation}  A few packages use \eval to bypass the restrictions
+imposed by the \k{R CMD CHECK} tool. This tool enforces some well-formedness
+rules on a submitted package before accepting it for inclusion in CRAN. Some
+static checks ensure that the package code does not use certain restricted
+functions. Package authors use \eval to obfuscate their code to get around this
+limitation. The example below mutates the variable \k{s} in the package
+environment. However, that is locked by default and unlocking it requires
+\k{unlockBinding} which is a restricted function. To work around this, the call
+to \k{unlockBinding} is done through \eval.
 
@@
    lockBinding("s", env)
+   ...
+ }
 \end{lstlisting}\medskip
@@
 
-The \eval function is used widely, and in varied ways, in R. The function is an
+The \eval function is used widely, and in varied ways in R. The function is an
 essential tool for language implementers. Our analysis observed that the base
@@
 code written by less experienced users suggests that \eval is exceedingly rarely
-used. So, it is fair to conclude that a majority of R programmers never have to
-write a call to \eval, but that the code they write would not run without it.
+used. So, it is fair to conclude that most R programmers never have to
+write a call to \eval, but the code they write would not run without it.
 

### 2021-09-09 paper/main.tex
Fix in 6

@@
 into two groups: one contains patterns that aim to provide better
-API for R users and the other focuses on simplifying implementation.
+API and the other focuses on simplifying implementation.
 

### 2021-09-09 paper/main.tex
Align better Table 2 and Table 3
7 + minor changes

@@
 \small
-\begin{tabular}{@{}l@{\hspace{1.5cm}}l@{}}
-\begin{minipage} {5cm}\small
+
+\begin{tabular}{@{\hspace{0.5cm}}l@{\hspace{1.2cm}}l}
+\begin{minipage}{5cm}\small
   \begin{tabular}{r@{\,}r@{\,}l@{}r|r@{\,}r@{\,}l@{}r@{}} \toprule
@@
 &
-\begin{minipage}{7cm}\small
-\begin{tabular}{@{}r|r@{~}r@{}r@{~}r@{}}\toprule
+\begin{minipage}{8cm}\small
+\begin{tabular}{@{}r|rrrr}\toprule
   &\eval & \k{evalq} & \k{eval} & \k{local}\\[-1.3mm]
@@
   \medskip\EE{$min(e)=\c{\$}$}\\[-2mm]\small This shape extends \k{X} to include lookup
-  with the dollar operator, \eg~\k{x\$f}, and vector indexing, \eg~\k{x[42]} and
+  with the dollar operator, \eg~\texttt{x\$f}, and vector indexing, \eg~\k{x[42]} and
   \k{x[[24]]}. This takes a few more steps of evaluation on average and is
@@
   <-} and parent environment assignment {\tt <\,\!<-} as well as the \k{assign}
-  function. It subsumes \k{\$}. Assignments represent the most obvious source
+  function. It subsumes \texttt{\$}. Assignments represent the most obvious source
   of side effects in \eval. Most cases originate from trivial code generation
@@
 other hand, there are \packageNbSimpleMinimizedOne sites that only receive one
-of the simple shapes (\ie\xspace \k{X}, \k{V}, \k{\$} or \k{<-}), and
+of the simple shapes (\ie\xspace \k{X}, \k{V}, \texttt{\$} or \k{<-}), and
 \packageNbSimpleMinimizedMore sites get multiple simple shapes. Among these
@@
 from unit testing frameworks as they use \eval to run all their tests and thus
-dominate the side-effect data.\footnote{The corpus has \k{RUnit, testthat,
-    tinytest} and \k{unitizer} testing frameworks.}
+dominate the side-effect data.\footnote{The corpus has \texttt{\scriptsize RUnit, testthat,
+    tinytest} and \texttt{\scriptsize unitizer} testing frameworks.}
 
@@
 
-\begin{lstlisting}[caption={\k{data.table::[}}, captionpos=b]
+\begin{lstlisting}[title={\k{data.table::[}}, captionpos=b]
 > flights[carrier=="DL" & year == 2014, mean(dep_delay), by=origin]
@@
 frame. We omit the implementation as it spans over 1700 lines at the time of
-this writing\footnote{Its implementation can be found at:
-  \url{https://github.com/Rdatatable/data.table/blob/master/R/data.table.R}}.
+this writing.\footnote{Its implementation can be found at:
+  \url{https://github.com/Rdatatable/data.table/blob/master/R/data.table.R}}
 
@@
 
-\begin{lstlisting}[caption={\k{glue::glue}}, captionpos=b]
+\begin{lstlisting}[title={\k{glue::glue}}, captionpos=b]
 > greeting <- "Hello"
-> glue('{greeting} World!')
+> glue("{greeting} World!")
 # Hello World!
@@
 
-\begin{lstlisting}[caption={\k{glue::glue_data}}, captionpos=b]
+\begin{lstlisting}[title={\k{glue::glue_data}}, captionpos=b]
 > flights[carrier=="DL" & year == 2014, mean(dep_delay), by=origin] |>
@@
 % We define and refer to listings in 2.2 so we need to put that captionsetup only here
-\captionsetup[lstlisting]{labelformat=empty} 
-\begin{lstlisting}[caption={\k{AdapSamp::rARS}}, captionpos=b]
+\begin{lstlisting}[title={\k{AdapSamp::rARS}}, captionpos=b]
  function(n,formula) {
@@
 
-\begin{lstlisting}[caption={\k{statnet.common::NVL}}, captionpos=b]
+\begin{lstlisting}[title={\k{statnet.common::NVL}}, captionpos=b]
  function(...) {
@@
 ability to make assertions and, if they fail, produce legible error messages.
-The \k{assertthat::see_if} function accepts a vararg with a list of
+The \texttt{\small assertthat::see\_if} function accepts a vararg with a list of
 expressions. It extracts these expressions using the extracting varargs
@@
 
-\begin{lstlisting}[caption={\k{assertthat::see_if}}, alsoletter={\_}, captionpos=b]
+\begin{lstlisting}[title={\texttt{\small assertthat::see\_if}}, alsoletter={\_}, captionpos=b]
  function(..., env=parent.frame(), msg=NULL) {
@@
 \k{base} R functions. For example, the \k{igraph} package provides the
-\k{do_call} function shown below. This is a variant of the builtin \k{do.call}
+\k{do_call} function shown below. This is a variant of the built-in \k{do.call}
 function with different semantics for resolving the function name. It
@@
 
-\begin{lstlisting}[caption={\k{igraph::do\_call}}, captionpos=b]
+\begin{lstlisting}[title={\k{igraph::do\_call}}, captionpos=b]
  function(f,...,args=list(),env=parent.frame()) {
@@
 
-\begin{lstlisting}[caption={\k{Rcpp::.makeCppMethods}}, captionpos=b]
+\begin{lstlisting}[title={\k{Rcpp::.makeCppMethods}}, captionpos=b]
  function(methods, env) {
@@
 
-\begin{lstlisting}[caption={\k{survival::survfit.formula}}, captionpos=b]
+\begin{lstlisting}[title={\k{survival::survfit.formula}}, captionpos=b]
  function(formula, data, weights, ...) {
    Call <- match.call()
-   indx <- match(c('formula', 'data', 'weights'), names(Call), nomatch=0)
+   indx <- match(c("formula", "data", "weights"), names(Call), nomatch=0)
    temp <- Call[c(1, indx)]
@@
 
-\begin{lstlisting}[caption={\k{base::write.csv}}, captionpos=b]
+\begin{lstlisting}[title={\k{base::write.csv}}, captionpos=b]
  function (...) {
@@
 
-\begin{lstlisting}[caption={\k{aibd::scalaEnsure}}, captionpos=b]
+\begin{lstlisting}[title={\k{aibd::scalaEnsure}}, captionpos=b]
  function() {
@@
 In R, \eval is used chiefly for meta-programming and accessing remote
-environments. Unlike JavaScript, which according to ~\citet{ecoop11} had a small
+environments. Unlike JavaScript, which according to~\citet{ecoop11} had a small
 set of well-defined patterns that could be rewritten without \eval, our results
-suggest that \eval is an integral part of programming in R.
-
-The cases where \eval appears to be an overkill include simple expression shapes
-such as \k{X}, \k{V}, \k{\$}, \k{<-}, and \k{FUN}. Variable lookup in a remote
-environment can be performed by the builtin \k{get} function. A similar case can
-be made for assignment and the \k{assign} function. There is a dedicated
-\k{do.call} function to perform function calls. However, the unnecessary uses of
-\eval are not easy to find and often can only be uncovered by manual analysis.
-Dynamic analysis is limited due to coverage issues, and static analysis of R
-remains an open problem.
-
-\begin{lstlisting}[caption={Unnecessary use of \eval},label=lst:unnecessary]
+suggest that \eval is an integral part of programming in R. However, there are
+cases in which \eval is not needed, and other, more stylized functions could
+have been used. There are also alternatives to \eval that provide
+a more principled way for meta-programming and dealing with lazy evaluation. In
+this section, we briefly discuss these issues.
+
+\paragraph{Unnecessary use of eval} The cases where \eval appears to be an
+overkill include simple expression shapes such as \k{X}, \k{V}, \texttt{\$}, \k{<-},
+and \k{FUN}. Variable lookup in a remote environment can be performed by the
+built-in \k{get} function. A similar case can be made for assignment and the
+\k{assign} function. There is a dedicated \k{do.call} function to perform
+function calls. However, the unnecessary uses of \eval are not easy to find and
+often can only be uncovered by manual analysis. Dynamic analysis is limited due
+to coverage issues, and static analysis of R remains an open problem.
+
+\begin{lstlisting}[H,caption={Unnecessary use of \eval},label=lst:unnecessary]
  PerformanceAnalytics::chart.QQPlot <- function (d="norm",dp,...) {
@@
    eval(parse(text=paste0("z<-q.f(",dp,",...)")))
+   ...
+ }
 \end{lstlisting}\medskip
@@
 
-\begin{lstlisting}
+\begin{lstlisting}[H]
  q.f <- get(paste0("q",d))
@@
 
-Another example, from \k{coin::copyslots}, constructs an expression that accesses
-fields of an object and copies them.
-
-\begin{lstlisting}
-  eval(str2lang(paste0("target@", s, " <- source@", s)))
+\paragraph{Eval alternatives}
+
+There are alternative implementations of \eval. The \k{lazyeval} package
+provides \k{lazy} and \k{lazy_eval} as safer variants for the built-in
+\k{substitute} and \k{eval} functions. The main differences are
+that the \k{lazy} function captures not only the expression, but also the
+environment associated with a function argument, and that it follows promises
+across function invocations. This helps to avoid scoping bugs, as shown in the
+example bellow:
+
+\begin{figure}[H]
+  \begin{minipage}{.48\textwidth}
+    \begin{lstlisting}[title={Use of R \eval},label=lst:lazyeval]
+ filter <- function(df, cond) {
+   conf <- substitute(cond)
+   r <- eval(cond, df, parent.frame())
+   df[r, ]
+ }
+
+ f <- function(...) {
+   x <- 1
+   filter(flights, ...)
+ }
+
+ x <- 2020
+ f(year > x) # uses wrongly 1 for x
     \end{lstlisting}
-
-% TODO: show how will it look with symbolic field read and write
-
-This can be replaced by symbolic field reads and writes:
-
-% This should be enough:
-\begin{lstlisting}
- slot(target, s) <- slot(source, s)
+  \end{minipage}
+  \begin{minipage}{.48\textwidth}
+    \begin{lstlisting}[title={Use of \k{lazyeval}},label=lst:lazyeval]
+ filter <- function(df, cond) {
+   cond <- lazy(cond)
+   r <- lazy_eval(cond, df)
+   df[r, ]
+ }
+
+ f <- function(...) {
+   x <- 1
+   filter(flights, ...)
+ }
+
+ x <- 2020
+ f(year > x) # uses correctly 2020 for x
 \end{lstlisting}
-
-We have not yet touched on alternative implementations of \eval. The tidyverse
-libraries\footnote{An opinionated collection of R packages designed for data
-  science. It consists of some of the most popular packages.} include their own
-implementation of \eval called \k{rlang::eval\_tidy}~\cite{tidyverse}, which
-departs from \eval in two ways. First, it supports the evaluation of quosures,
-custom objects used for metaprogramming that bundle expressions with an
-environment. The \k{eval\_tidy} function accepts a data mask argument, a set of
-bindings such as a data frame or list, which takes precedence over the
-environment. Effects such as assignments happen in the data mask, and
-expressions such as \k{return()} do not work since the data mask does not
-correspond to a frame on the call stack.
+\end{minipage}
+\end{figure}
+
+The \k{filter} function selects rows from \k{df} that match a given condition
+passed in \k{cond}. However, the implementation using the built-in R \eval has a
+scoping issue since it does not correctly capture the proper environment
+associated with the condition.\footnote{This is a simplified version of the
+  built-in \texttt{\scriptsize subset} function which suffers from this exact
+problem.} The \k{lazyeval} version does not manifest the issue as it properly
+captures the correct environment.
+
+The tidyverse libraries\footnote{An opinionated collection of R packages
+designed for data science. It consists of some of the most popular packages.}
+include their own implementation of \eval called
+\k{rlang::eval\_tidy}~\cite{tidyverse}, which departs from \eval in two ways.
+First, it supports the evaluation of quosures, custom objects used for
+metaprogramming that bundle expressions with an environment. The \k{eval\_tidy}
+function accepts a data mask argument, a set of bindings such as a data frame
+or list, which takes precedence over the environment. Effects such as
+assignments happen in the data mask, and expressions such as \k{return()} do
+not work since the data mask does not correspond to a frame on the call stack.
 

### 2021-09-09 paper/main.tex
More fixes

@@
 
+\enlargethispage{2mm}
 \begin{itemize}[---]
@@
 
+\enlargethispage{2mm}
 \begin{lstlisting}[title={\k{glue::glue_data}}, captionpos=b]
@@
 
-\begin{lstlisting}[H,caption={Unnecessary use of \eval},label=lst:unnecessary]
- PerformanceAnalytics::chart.QQPlot <- function (d="norm",dp,...) {
-   q.f <- eval(parse(text=paste0("q",d)))
+\begin{lstlisting}[caption={Unnecessary use of \eval (\k{PerformanceAnalytics::chart.QQPlot})},label=lst:unnecessary]
+function (d="norm", dp, ...) {
+   q.f <- eval(parse(text=paste0("q", d)))
    z <- NULL
-   eval(parse(text=paste0("z<-q.f(",dp,",...)")))
+   eval(parse(text=paste0("z<-q.f(", dp, ", ...)")))
    ...
@@
 
-\begin{lstlisting}[H]
- q.f <- get(paste0("q",d))
+\begin{lstlisting}
+ q.f <- get(paste0("q", d))
  z <- q.f(dp, ...)

### 2021-09-13 paper/main.tex
Fix an ugle page break interfering with a code block
Fix issues in 6

@@
 evaluation, \k{substitute}, and \k{eval} are some of the tools given to
-developers to this end. To understand \emph{why} developers use \eval, we
+developers to this end. To understand \emph{why} developers use eval, we
 manually inspected the top 117 sites, which contribute to 90\% of the calls in
-the corpus. We present examples from nine classes from this batch. The
-classification is not exhaustive as we only study a fraction from the
-\PkgEvalCallSites call sites in our corpus, and many uses of \eval do not fit
-in a specific category. Instead, the goal is to demonstrate the wide variety of
-uses to which \eval is put to use. At the high level, the usage can be grouped
-into two groups: one contains patterns that aim to provide better
-API and the other focuses on simplifying implementation.
-
-\subsection{Better API}
-
-This category contains \eval used to develop a better API in terms of
-ergonomics, extensibility, and ease of use. This is important given
-that R is often being used interactively for data exploration.
+the corpus. We present examples from nine classes from this batch. This
+classification is not exhaustive since there are \PkgEvalCallSites \eval call
+sites in our corpus, and many uses of eval do not fit in a specific category.
+further categorize the nine classes into two groups: those that improve the
+interface for users and those that simplify the implementation for developers.
+
+\subsection{Better Interface}
+
+This group contains \eval uses that enable the development of interfaces for
+quick interactive data exploration.
 
@@
 # 2:    EWR 14.96171
-# 3:    JFK 12.45414
 \end{lstlisting}\medskip
@@
 
-\enlargethispage{2mm}
 \begin{lstlisting}[title={\k{glue::glue_data}}, captionpos=b]
@@
 # EWR has average delay of 15 min
-# JFK has average delay of 12 min
 \end{lstlisting}
@@
 \paragraph{Extracting Varargs} \Eval is used to obtain the source code of the
-arguments passed to a vararg parameter (\k{...}). Concretely,
-\k{substitute(alist(...))} yields an expression representing a call to
-\k{alist} with the \k{...} expanded to the arguments supplied to the function.
-Passing this to \eval returns a list of the unevaluated argument
-expressions.
-In the following example, it is used to find and return the value of the first expression that evaluates to non-null value.
-The extracting varargs pattern is needed to avoid premature promise forcing.
+arguments passed to a vararg parameter (\k{...}). In the example below,
+\k{substitute(alist(...))} yields an expression representing a call to \k{alist}
+with the \k{...} expanded to the arguments supplied to the function. Passing
+this to \eval returns a list of the unevaluated argument expressions. This
+pattern is needed to avoid premature evaluation of vararg arguments.
 
@@
    for (e in eval(substitute(alist(...)))) {
-     if (!is.null(eval(e, parent.frame()))) break
-   }
-   e
- }
 \end{lstlisting}
@@
 \paragraph{Controlled Evaluation} \Eval can be used to control when and if
-expressions are evaluated. For example, \k{assertthat} provides users with the
-ability to make assertions and, if they fail, produce legible error messages.
-The \texttt{\small assertthat::see\_if} function accepts a vararg with a list of
-expressions. It extracts these expressions using the extracting varargs
-pattern and evaluates them sequentially in the caller's environment using
-\k{eval}. If an expression does not hold, an error is generated.
+expressions are evaluated. For example, \k{assertthat} allows users to make
+assertions and, if they fail, produce legible error messages. The \texttt{\small
+  assertthat::see\_if} function accepts a vararg with a list of expressions. It
+extracts these expressions using the extracting varargs pattern and evaluates
+them sequentially in the caller's environment using \k{eval}. If an expression
+does not hold, an error is generated.
 
@@
 
-\paragraph{Shadowing} Some packages use \eval to implement variations of the
-\k{base} R functions. For example, the \k{igraph} package provides the
-\k{do_call} function shown below. This is a variant of the built-in \k{do.call}
-function with different semantics for resolving the function name. It
-constructs a call from a name and a list of arguments and passes it to \eval.
+\paragraph{Shadowing} First-class environment and \eval enable R users to
+implement variations of core language functionality. For example, the \k{igraph}
+package provides the \k{do_call} function shown below. This is a variant of the
+built-in \k{do.call} function with different semantics for resolving the
+function name. \k{do_call} calls a function (\k{f}) in the specified environment
+(\k{env}) with the arguments supplied as a list (\k{args}).
 
@@
 
-The patterns in this category use \eval to simplify implementation in terms of
-avoiding repetitive and boilerplate code.
-
-\paragraph{Code Generation} \Eval can be used to generate code that would
-either be tedious to write by hand or that is only known after deployment. The
-following code snippet for the \k{Rcpp} package uses \eval to automatically
+This group contains \eval uses that simplify implementation for package
+developers.
+
+\paragraph{Code Generation} \Eval can be used to generate code that would either
+be tedious to write by hand or that is only known after deployment. The
+following code snippet from the \k{Rcpp} package uses \eval to automatically
 generate R bindings to C++ methods.
@@
 the call expression is evaluated in the parent environment. The use of
-\k{match.call} avoids repeated use of \k{substitute} for extracting the
-argument expressions. It is extensible as new parameter names can be easily
-added. This example illustrates one of the most recurring uses of \eval by
-packages that do statistical modeling, and relates to the \k{model.frame} shape
-in Table~\ref{tab:minimizedexpressions}.
+\k{match.call} avoids repeated use of \k{substitute} for extracting the argument
+expressions. It is extensible as new parameter names can be easily added. This
+example illustrates one of the most recurring uses of \eval by packages that do
+statistical modeling. It is a typical example of the \k{model.frame} shape in
+Table~\ref{tab:minimizedexpressions}.
 
@@
 limitation. The example below mutates the variable \k{s} in the package
-environment. However, that is locked by default and unlocking it requires
+environment. However, it is locked by default, and unlocking it requires
 \k{unlockBinding} which is a restricted function. To work around this, the call
@@
 suggest that \eval is an integral part of programming in R. However, there are
-cases in which \eval is not needed, and other, more stylized functions could
-have been used. There are also alternatives to \eval that provide
-a more principled way for meta-programming and dealing with lazy evaluation. In
-this section, we briefly discuss these issues.
+cases in which \eval is not needed, and other, more stylized functions can be
+used. There are also alternatives to \eval that provide a more principled way
+for meta-programming and dealing with lazy evaluation. In this section, we
+briefly discuss these issues.
 
 \paragraph{Unnecessary use of eval} The cases where \eval appears to be an
-overkill include simple expression shapes such as \k{X}, \k{V}, \texttt{\$}, \k{<-},
-and \k{FUN}. Variable lookup in a remote environment can be performed by the
-built-in \k{get} function. A similar case can be made for assignment and the
+overkill include simple expression shapes such as \k{X}, \k{V}, \texttt{\$},
+\k{<-}, and \k{FUN}. Variable lookup in a remote environment can be performed by
+the built-in \k{get} function. A similar case can be made for assignment and the
 \k{assign} function. There is a dedicated \k{do.call} function to perform
-function calls. However, the unnecessary uses of \eval are not easy to find and
+function calls. However, unnecessary uses of \eval are not easy to find and
 often can only be uncovered by manual analysis. Dynamic analysis is limited due
@@
 
-For the example in Listing~\ref{lst:unnecessary}, the function uses \eval to
+In the example in Listing~\ref{lst:unnecessary}, the function uses \eval to
 lookup a target function name from a constructed string. Then it constructs a
@@
 \paragraph{Eval alternatives}
-
-There are alternative implementations of \eval. The \k{lazyeval} package
-provides \k{lazy} and \k{lazy_eval} as safer variants for the built-in
-\k{substitute} and \k{eval} functions. The main differences are
-that the \k{lazy} function captures not only the expression, but also the
-environment associated with a function argument, and that it follows promises
-across function invocations. This helps to avoid scoping bugs, as shown in the
-example bellow:
+Alternative implementations of \eval are also available from R packages. The
+\k{lazyeval} package provides \k{lazy} and \k{lazy_eval} as safer variants for
+the built-in \k{substitute} and \k{eval} functions, respectively. These variants
+differ from their counterparts in that they capture both the argument
+expressions and their environments, and they follow promises across function
+invocations. This helps to avoid scoping bugs, as shown in the example below.
 
@@
  x <- 2020
- f(year > x) # uses wrongly 1 for x
+ f(year > x) # incorrectly uses 1 for x
     \end{lstlisting}
@@
  x <- 2020
- f(year > x) # uses correctly 2020 for x
+ f(year > x) # correctly uses 2020 for x
 \end{lstlisting}
@@
 The \k{filter} function selects rows from \k{df} that match a given condition
-passed in \k{cond}. However, the implementation using the built-in R \eval has a
-scoping issue since it does not correctly capture the proper environment
-associated with the condition.\footnote{This is a simplified version of the
-  built-in \texttt{\scriptsize subset} function which suffers from this exact
-problem.} The \k{lazyeval} version does not manifest the issue as it properly
-captures the correct environment.
+passed to \k{cond}. The implementation using the built-in \eval suffers from a
+scoping bug since it always evaluates \k{cond} in the caller's environment
+irrespective of where it is passed from.\footnote{This is a simplified version
+  of the built-in \texttt{\scriptsize subset} function which suffers from this
+  exact problem.} The \k{lazyeval} version avoids this problem since it
+correctly captures the environment associated with \k{cond}.
 
@@
   designed for data science. It consists of some of the most popular packages.}
-include their own implementation of \eval called
-\k{rlang::eval\_tidy}~\cite{tidyverse}, which departs from \eval in two ways.
-First, it supports the evaluation of quosures, custom objects used for
-metaprogramming that bundle expressions with an environment. The \k{eval\_tidy}
-function accepts a data mask argument, a set of bindings such as a data frame
-or list, which takes precedence over the environment. Effects such as
-assignments happen in the data mask, and expressions such as \k{return()} do
-not work since the data mask does not correspond to a frame on the call stack.
+include their own implementation of \eval called \k{eval\_tidy}~\cite{tidyverse}
+from the \k{rlang} package. This function which departs from the built-in \eval
+in two ways. First, it supports the evaluation of quosures, custom objects used
+for metaprogramming that bundle expressions with an environment. Second, the
+\k{eval\_tidy} function accepts a data mask argument, a set of bindings such as
+a data frame or list, which takes precedence over the environment. Consequently,
+effects such as assignments happen in the data mask, and expressions such as
+\k{return()} do not work since the data mask does not correspond to a frame on
+the call stack.
 

### 2021-09-13 paper/main.tex
Remove vspace and do minor fixes

@@
          \nodename{
-           \vspace{1mm}\textbf{Download}\vspace{1mm}\\
+           \textbf{Download}\\
            15K packages\\
@@
          \nodename{
-           \vspace{1mm}\textbf{Extract}\vspace{1mm}\\
+           \textbf{Extract}\\
            240K package programs\\%, 4.6M LOC\\
@@
          \nodename{
-           \vspace{1mm}\textbf{Trace}\vspace{1mm}\\
+           \textbf{Trace}\\
            49M eval calls \\
@@
          \nodename{
-           \vspace{1mm}\textbf{Analyze}\vspace{1mm}\\
+           \textbf{Analyze}\\
            2.8M unique eval calls\\
@@
   \caption{Normalized calls} \label{tab:cn}
-  \vspace{-0.5cm}
+  %\vspace{-0.5cm}
 \end{wraptable}
@@
   \begin{minipage}{.39\textwidth}
-    \vspace{-5mm}
+    %\vspace{-5mm}
 \begin{lstlisting}
@@
 sites in our corpus, and many uses of eval do not fit in a specific category.
-further categorize the nine classes into two groups: those that improve the
+We further categorize the nine classes into two groups: those that improve the
 interface for users and those that simplify the implementation for developers.
@@
 
-\paragraph{Shadowing} First-class environment and \eval enable R users to
+\paragraph{Shadowing} First-class environments and \eval enable R users to
 implement variations of core language functionality. For example, the \k{igraph}
-package provides the \k{do_call} function shown below. This is a variant of the
-built-in \k{do.call} function with different semantics for resolving the
-function name. \k{do_call} calls a function (\k{f}) in the specified environment
-(\k{env}) with the arguments supplied as a list (\k{args}).
+package provides the \k{do_call} function shown below. \k{do_call} calls a
+function (\k{f}) in the specified environment (\k{env}) with the arguments
+supplied to \k{...}, and \k{args}. This is a variant of the built-in \k{do.call}
+function with two differences: it resolves the function name differently, and it
+also allows arguments to be passed as varargs.
 
 \begin{lstlisting}[title={\k{igraph::do\_call}}, captionpos=b]
- function(f,...,args=list(),env=parent.frame()) {
-   f <- substitute(f)
-   eval(make_call(f,...,args), env)
- }
+ function(f,...,args=list(),env=parent.frame())
+   eval(make_call(substitute(f),...,args), env)
 \end{lstlisting}
 
-\subsection{Implementation simplification}
+\subsection{Implementation Simplification}
 

### 2021-09-13 paper/main.tex
Fix CCS classification

@@
 
-%%% TODO CHANGE
 \keywords{eval, dynamic languages}
@@
 <concept>
-	<concept_id>10011007.10011006.10011008.10011024</concept_id>
-	<concept_desc>Software and its engineering~Language features</concept_desc>
+<concept_id>10002944.10011123.10010912</concept_id>
+<concept_desc>General and reference~Empirical studies</concept_desc>
 <concept_significance>500</concept_significance>
@@
 <concept_desc>Software and its engineering~General programming languages</concept_desc>
-	<concept_significance>300</concept_significance>
+<concept_significance>500</concept_significance>
 </concept>
 <concept>
-	<concept_id>10011007.10011006.10011041.10011044</concept_id>
-	<concept_desc>Software and its engineering~Just-in-time compilers</concept_desc>
-	<concept_significance>300</concept_significance>
-	</concept>
-	<concept>
-	<concept_id>10011007.10011006.10011008.10011009.10011021</concept_id>
-	<concept_desc>Software and its engineering~Multiparadigm languages</concept_desc>
-	<concept_significance>100</concept_significance>
+<concept_id>10011007.10011006.10011050.10010517</concept_id>
+<concept_desc>Software and its engineering~Scripting languages</concept_desc>
+<concept_significance>500</concept_significance>
 </concept>
@@
 
-\ccsdesc[500]{Software and its engineering~Language features}
-\ccsdesc[300]{Software and its engineering~General programming languages}
-
-\author{Aviral Goel}\affiliation{\institution{Northeastern
-    University}\country{USA}}
+\ccsdesc[500]{General and reference~Empirical studies}
+\ccsdesc[500]{Software and its engineering~General programming languages}
+\ccsdesc[500]{Software and its engineering~Scripting languages}
+
+\author{Aviral Goel}\affiliation{\institution{Northeastern University}\country{USA}}
 \author{Pierre Donat-Bouillud}\affiliation{\institution{Czech Technical University in Prague}\country{Czech Republic}}

### 2021-09-24 paper/main.tex
Add copyright
Submission related changes

@@
 \setlength{\pdfpageheight}{11in}
-
-%%% The following is specific to OOPSLA '21 and the paper
-%%% 'Why We Eval in the Shadows'
-%%% by Aviral Goel, Pierre Donat-Bouillud, Filip Křikava, Christoph M. Kirsch, and Jan Vitek.
-%%%
-%\setcopyright{ACMUNKNOWN} %TODO: we still have to choose the copyright
-\acmPrice{}
-\acmDOI{10.1145/3485502}
-\acmYear{2021}
-\copyrightyear{2021}
-\acmSubmissionID{oopsla21main-p121-p}
-\acmJournal{PACMPL}
-\acmVolume{5}
-\acmNumber{OOPSLA}
-\acmArticle{125}
-\acmMonth{10}
 
@@
 }
-
 \usepackage{caption}
 
+\setcopyright{rightsretained}
+
 \begin{document}
+%%% The following is specific to OOPSLA '21 and the paper
+%%% 'Why We Eval in the Shadows'
+%%% by Aviral Goel, Pierre Donat-Bouillud, Filip Křikava, Christoph M. Kirsch, and Jan Vitek.
+%%% 
+\acmPrice{}
+\acmDOI{10.1145/3485502}
+\acmYear{2021}
+\copyrightyear{2021}
+\acmSubmissionID{oopsla21main-p121-p}
+\acmJournal{PACMPL}
+\acmVolume{5}
+\acmNumber{OOPSLA}
+\acmArticle{125}
+\acmMonth{10}
+
 \title{What We Eval in the Shadows}
-\subtitle{A Large-scale Study of Eval in R programs}
+\subtitle{A Large-Scale Study of Eval in R Programs}
 
@@
 \author{Jan Vitek}\affiliation{\institution{Northeastern University}\country{USA}}
-\affiliation{\institution{Czech Technical University in Prague}\country{Czech Republic}}
+\affiliation{\institution{Czech Technical University in Prague}\country{Czech
+    Republic}}
+\authorsaddresses{} 
+\renewcommand{\shortauthors}{A. Goel, P. Donat-Bouillud, F. Křikava, C. M.
+  Kirsch, J. Vitek}
 
@@
 \maketitle
-\renewcommand{\shortauthors}{Goel et al.}
+%\renewcommand{\shortauthors}{Goel et al.}
 \renewcommand{\shorttitle}{What We Eval in the Shadows}
@@
 
-\begin{wrapfigure}{r}{7cm}
-	\hspace*{-1cm}\includegraphics[width=9cm]{pkgs-eval-callsites-hist.pdf}
-	\caption{ \eval call sites}\label{fig:pkgs-eval-callsites-hist}
-\end{wrapfigure}
-
 There are \PkgEvalCallSites \eval sites in \PkgPackages packages;
@@
 \CranFunsWithEvalRatio of all functions use \eval.
+
+\begin{wrapfigure}{r}{7cm}
+	\hspace*{-1cm}\includegraphics[width=9cm]{pkgs-eval-callsites-hist.pdf}
+	\caption{ \eval call sites}\label{fig:pkgs-eval-callsites-hist}
+\end{wrapfigure}
 
@@
 \begin{tabular}{@{}c@{}c@{}}
-\begin{minipage}{7.5cm}
+\begin{minipage}{6cm}
  \includegraphics[width=\textwidth]{package_events_per_pack_small}
-\end{minipage}&\begin{minipage}{7.5cm}
+\end{minipage}&\begin{minipage}{6cm}
   \includegraphics[width=\textwidth]{package_events_per_pack_large}
@@
 \k{<-} & \packageMinimizedcallsitesi  & \packageMinimizedpropsitesi & \packageMinimizedpackagei & \packageMinimizedmedianoperationsiRnd & \packageMinimizedpercentparentframesi & \k{x[1, 2:3, 2:3] <- value}\\
-\k{BLOCK} & \packageMinimizedcallsitesj & \packageMinimizedpropsitesj & \packageMinimizedpackagej & \packageMinimizedmedianoperationsjRnd & \packageMinimizedpercentparentframesj & \k{\{write.csv(iris,tf);file.size(tf)\}} \\\bottomrule
+\k{BLOCK} & \packageMinimizedcallsitesj & \packageMinimizedpropsitesj & \packageMinimizedpackagej & \packageMinimizedmedianoperationsjRnd & \packageMinimizedpercentparentframesj & \k{\{write(iris,tf);file.size(tf)\}} \\\bottomrule
 \end{tabular}
@@
 \begin{table}[h]
-  \centering\small\hspace{-.5cm}
-\begin{minipage}{3.7cm}
+  \centering\small%\hspace{-.5cm}
+\begin{minipage}{3.5cm}
   \begin{tabular}{@{}rrr@{}}\toprule
@@
 \caption{Kinds per site} \label{tab:highlevelenvironments}
-\end{minipage}\hspace{-.2cm}
-\begin{minipage}{3.7cm}\centering
+\end{minipage}%\hspace{-.2cm}
+\begin{minipage}{3.5cm}\centering
   \begin{tabular}{@{}rrr@{}}\toprule
@@
 \caption{Function offset}\label{tab:funoffset}
-\end{minipage}\hspace{-.2cm}
-\begin{minipage}{3.7cm}
+\end{minipage}%\hspace{-.2cm}
+\begin{minipage}{3.5cm}
   \begin{tabular}{@{}rrr@{}} \toprule
@@
 \caption{Wrapper envs.} \label{tab:newenvs}
-\end{minipage}\hspace{-.2cm}
-\begin{minipage}{3.7cm}\centering
+\end{minipage}%\hspace{-.2cm}
+\begin{minipage}{3.45cm}\centering
   \begin{tabular}{@{}ccc@{}} \toprule

### 2021-09-29 paper/main.tex
Add authorsaddresses
Move Table captions to top, change table positions, fix minor issues

@@
     Republic}}
-\authorsaddresses{} 
-\renewcommand{\shortauthors}{A. Goel, P. Donat-Bouillud, F. Křikava, C. M.
-  Kirsch, J. Vitek}
+\renewcommand{\shortauthors}{Aviral Goel, Pierre Donat-Bouillud, Filip Křikava, Christoph M.
+  Kirsch, Jan Vitek}
 
@@
 
-\begin{table}[!h]\center\small\begin{tabular}{r@{~}llll}\toprule
+\begin{table}[!h]\center\small
+  \caption{Design space of \eval}\label{comp}
+  \begin{tabular}{r@{~}llll}\toprule
 \tiny\sc Language&&\sc\tiny Input&\sc\tiny Scope&\tiny\sc Reflective operations\\\midrule
@@
 \bf R&\cite{R96}  & expression& programmatic      & data, stack, environment\\\bottomrule
-\end{tabular}\caption{Design space of \eval}\label{comp}
+\end{tabular}
 \end{table}
@@
 
+\subsection{Eval in R}\label{sec:eval-in-r}
+
+The \eval function in R takes three parameters: an expression to evaluate
+(\k{e}), an environment where to evaluate (\k{env}), and an enclosure (\k{encl})
+that is used to look up objects not found in \k{env}.\footnote{R offers three
+other variants of {\tt eval}: {\tt evalq} automatically quotes passed
+expression---shorthand for {\tt eval(quote(...))}, {\tt eval.parent(e,n)}
+specifies the evaluating environment in terms of number of call frames ({\tt n})
+to go back---shorthand for {\tt eval(e,parent.frame(n))}, and {\tt local}
+evaluates {\tt e} in a fresh environment---shorthand for {\tt
+  evalq(e,new.env())}.}
+
+\begin{lstlisting}
+ eval <- function(e, env = parent.frame(),
+                   encl = if(is.list(env)) parent.frame() else baseenv()) ...
+\end{lstlisting}
+
+The expression passed to \eval can be thought of as an abstract syntax tree.
+Listings~\ref{lst:exprs} and \ref{lst:match.call} show some of the ways of
+creating expressions: parsing from a string, manually, or via reflection. The
+call to \k{substitute(x)} extracts the unevaluated expression from the promise
+\k{x}. The call to \k{match.call()} returns an expression representing the
+current call which can then be further modified.
+
 \begin{figure}[!t]
@@
 \end{figure}
-
-\subsection{Eval in R}\label{sec:eval-in-r}
-
-The \eval function in R takes three parameters: an expression to evaluate
-(\k{e}), an environment where to evaluate (\k{env}), and an enclosure (\k{encl})
-that is used to look up objects not found in \k{env}.\footnote{R offers three
-other variants of {\tt eval}: {\tt evalq} automatically quotes passed
-expression---shorthand for {\tt eval(quote(...))}, {\tt eval.parent(e,n)}
-specifies the evaluating environment in terms of number of call frames ({\tt n})
-to go back---shorthand for {\tt eval(e,parent.frame(n))}, and {\tt local}
-evaluates {\tt e} in a fresh environment---shorthand for {\tt
-  evalq(e,new.env())}.}
-
-\begin{lstlisting}
- eval <- function(e, env = parent.frame(),
-                   encl = if(is.list(env)) parent.frame() else baseenv()) ...
-\end{lstlisting}
-
-The expression passed to \eval can be thought of as an abstract syntax tree.
-Listings~\ref{lst:exprs} and \ref{lst:match.call} show some of the ways of
-creating expressions: parsing from a string, manually, or via reflection. The
-call to \k{substitute(x)} extracts the unevaluated expression from the promise
-\k{x}. The call to \k{match.call()} returns an expression representing the
-current call which can then be further modified.
 
@@
   \begin{minipage}{5cm}\small
+    \caption{Call frequency}\label{tab:freq}
   \begin{tabular}{r@{\,}r@{\,}l@{}r|r@{\,}r@{\,}l@{}r@{}} \toprule
@@
 \tt 1K &--& \tt 10K    & \packageBind  & \tt 10M &--& \tt 100M & \packageBinh\\\bottomrule
-\end{tabular}\caption{Call frequency}\label{tab:freq}
+\end{tabular}
 \end{minipage}
@@
 \begin{minipage}{8cm}\small
+\caption{Variants}\label{tab:variantseval}
 \begin{tabular}{@{}r|rrrr}\toprule
@@
 \small Invocations&\packageEvalsRnd&\packageEvalqsRnd&\packageEparentsRnd&\packageLocalsRnd\\\bottomrule
-\end{tabular}\caption{Variants}\label{tab:variantseval}
+\end{tabular}
 \end{minipage}\end{tabular}
@@
 
-Table~\ref{tab:variantseval} summarizes the use of variants, the first row has sites
-(\emph{static}), the second has sites encountered during analysis
-(\emph{exercised}) and the last gives calls (\emph{invocations}). Most sites and
-calls go to \eval itself, \k{eval.parent} is rare, and both \k{evalq} and
-\k{local} are barely used at all.
-
 \begin{wraptable}{r}{5cm}  \small  \centering
+  \caption{Normalized calls} \label{tab:cn}
   \begin{tabular}{r@{\,}r@{\,}l@{\,}r|r@{\,}r@{\,}l@{}r} \toprule
@@
   \end{tabular}
-  \caption{Normalized calls} \label{tab:cn}
-  %\vspace{-0.5cm}
+  % \vspace{-0.5cm}
 \end{wraptable}
+
+Table~\ref{tab:variantseval} summarizes the use of variants, the first row has sites
+(\emph{static}), the second has sites encountered during analysis
+(\emph{exercised}) and the last gives calls (\emph{invocations}).
+Most sites and
+calls go to \eval itself, \k{eval.parent} is rare, and both \k{evalq} and
+\k{local} are barely used at all.
 
@@
   \hspace*{-0.2cm}
+  \caption{Minimized expressions} \label{tab:minimizedexpressions}
 \begin{tabular}{crrrrrc}\toprule
@@
 \k{X}&\packageMinimizedcallsitesa &\packageMinimizedpropsitesa &\packageMinimizedpackagea &\packageMinimizedmedianoperationsaRnd &\packageMinimizedpercentparentframesa & \k{y+1}\\
-\k{F(F(X))} & \packageMinimizedcallsitesb  & \packageMinimizedpropsitesb & \packageMinimizedpackageb  & \packageMinimizedmedianoperationsbRnd & \packageMinimizedpercentparentframesb & \k{gbov( mean(x), a-1)}\\
+\k{F(F(X))} & \packageMinimizedcallsitesb  & \packageMinimizedpropsitesb & \packageMinimizedpackageb  & \packageMinimizedmedianoperationsbRnd & \packageMinimizedpercentparentframesb & \k{gbov(mean(x), a-1)}\\
 \k{V}&\packageMinimizedcallsitesc &\packageMinimizedpropsitesc &\packageMinimizedpackagec &\packageMinimizedmedianoperationscRnd &\packageMinimizedpercentparentframesc& \k{c(42,21,0)}\\
@@
 \k{\$} & \packageMinimizedcallsitese & \packageMinimizedpropsitese & \packageMinimizedpackagee & \packageMinimizedmedianoperationseRnd & \packageMinimizedpercentparentframese & \k{DF\$B}\\
-\k{model.frame}& \packageMinimizedcallsitesf & \packageMinimizedpropsitesf & \packageMinimizedpackagef & \packageMinimizedmedianoperationsfRnd & \packageMinimizedpercentparentframesf &  \k{model.frame(formula = Z ~ U)}   \\
+\k{model.frame}& \packageMinimizedcallsitesf & \packageMinimizedpropsitesf & \packageMinimizedpackagef & \packageMinimizedmedianoperationsfRnd & \packageMinimizedpercentparentframesf &  \k{model.frame(formula = Z~U)}   \\
 \k{F()}& \packageMinimizedcallsitesg & \packageMinimizedpropsitesg & \packageMinimizedpackageg & \packageMinimizedmedianoperationsgRnd & \packageMinimizedpercentparentframesg & \k{rgamma(3, 2, n = 10L)} \\
-\k{FUN} & \packageMinimizedcallsitesh & \packageMinimizedpropsitesh & \packageMinimizedpackageh & \packageMinimizedmedianoperationshRnd & \packageMinimizedpercentparentframesh & \k{function(x, y) x + 3 * y} \\
+\k{FUN} & \packageMinimizedcallsitesh & \packageMinimizedpropsitesh & \packageMinimizedpackageh & \packageMinimizedmedianoperationshRnd & \packageMinimizedpercentparentframesh & \k{function(x)x+1} \\
 \k{<-} & \packageMinimizedcallsitesi  & \packageMinimizedpropsitesi & \packageMinimizedpackagei & \packageMinimizedmedianoperationsiRnd & \packageMinimizedpercentparentframesi & \k{x[1, 2:3, 2:3] <- value}\\
@@
 \end{tabular}
-\caption{Minimized expressions} \label{tab:minimizedexpressions}
 \end{table}
@@
 \begin{tabular}{@{}p{.97\linewidth}}
-  \medskip\EE{$min(e)=\c{V}$}\\[-2mm]\small Values occur in 16\% of sites, 74\% of
-  those are constants such as integers, the remainder evaluate to a value (\packageNbCallSitesUniqueActualValue of sites only ever see a value). \Eval
-  is needed when the expression denoting the values is constructed
-  programmatically. Cases when a value is directly given to \eval, and \eval
-  returns it unchanged, often occur when a computation has a default path and
-  another, more interesting, path that requires evaluation.
-  This shape usually runs in few interpreter steps.
-  The environment in which these expressions evaluate is mostly irrelevant.
-\end{tabular}
-
-\begin{tabular}{@{}p{.97\linewidth}}
   \medskip\EE{$min(e)=\c{X}$}\\[-2mm]\small Variable lookups are the most common shape,
@@
   expressions are evaluated in a function environment.
-\end{tabular}
-
-\begin{tabular}{@{}p{.97\linewidth}}
+  \\
+  \medskip\framebox{$min(e)=\c{F(X)}$}~\EE{$min(e)=\c{F(F(X))}$}\\[-2mm]\small This shape corresponds to function
+  calls whose arguments may include variable references and nested calls. Together
+  they represent the most frequent shape and perform over 16 steps. They are
+  frequently evaluated in function environments.
+  \\
+  \medskip\EE{$min(e)=\c{V}$}\\[-2mm]\small Values occur in 16\% of sites, 74\% of
+  those are constants such as integers, the remainder evaluate to a value (\packageNbCallSitesUniqueActualValue of sites only ever see a value). \Eval
+  is needed when the expression denoting the values is constructed
+  programmatically. Cases when a value is directly given to \eval, and \eval
+  returns it unchanged, often occur when a computation has a default path and
+  another, more interesting, path that requires evaluation.
+  This shape usually runs in few interpreter steps.  
+  \\
   \medskip\EE{$min(e)=\c{\$}$}\\[-2mm]\small This shape extends \k{X} to include lookup
@@
   typically used in a function environment.
+  \\
+  \medskip\EE{$min(e)=\c{model.frame}$}\\[-2mm]\small The \k{model.frame} function
+  returns a data frame resulting from fitting the model described in a formula.
+  This shape subsumes \k{F(F(X))}, \k{FUN}, and assignments. It is the single most
+  popular function invoked from \eval. Each call does a median of 2K instructions.
+\end{tabular}
+
+\begin{tabular}{@{}p{.97\linewidth}}
+  \medskip\EE{$min(e)=\c{F()}$}\\[-2mm]\small This shape represents function calls without variable lookups or
+  assignment. A variable is allowed in the function position; looking up
+  functions does not trigger computation unless the function's name
+  is bound to a promise. This shape typically does not perform much work.
+  \\
+  \medskip\EE{$min(e)=\c{FUN}$}\\[-2mm]\small This shape represents function definitions without any
+  computation, \eg~\k{function(x) x+1}. In addition to \k{FUN},
+  \packageGeneralizedFunctionDefinitionSitesPercent sites have function
+  definitions nested in other expressions.
   \\
@@
   \\
-  \medskip\EE{$min(e)=\c{F()}$}\\[-2mm]\small This shape represents function calls without variable lookups or
-  assignment. A variable is allowed in the function position; looking up
-  functions does not trigger computation unless the function's name
-  is bound to a promise. This shape typically does not perform much work.
-\end{tabular}
-
-\begin{tabular}{@{}p{.97\linewidth}}
-  \medskip\framebox{$min(e)=\c{F(X)}$}~\EE{$min(e)=\c{F(F(X))}$}\\[-2mm]\small This shape corresponds to function
-  calls whose arguments may include variable references and nested calls. Together
-  they represent the most frequent shape and perform over 16 steps. They are
-  frequently evaluated in function environments.
-  \\
-  \medskip\EE{$min(e)=\c{FUN}$}\\[-2mm]\small This shape represents function definitions without any
-  computation, \eg~\k{function(x) x+1}. In addition to \k{FUN},
-  \packageGeneralizedFunctionDefinitionSitesPercent sites have function
-  definitions nested in other expressions.
-\end{tabular}
-
-\begin{tabular}{@{}p{.97\linewidth}}
   \medskip\EE{$min(e)=\c{BLOCK}$}\\[-2mm]\small These are multi-statement code blocks; as
@@
   \k{doParallel}), or deferring code execution (\eg \k{withr}).
-\end{tabular}
-
-\begin{tabular}{@{}p{.97\linewidth}}
-  \medskip\EE{$min(e)=\c{model.frame}$}\\[-2mm]\small The \k{model.frame} function
-  returns a data frame resulting from fitting the model described in a formula.
-  This shape subsumes \k{F(F(X))}, \k{FUN}, and assignments. It is the single most
-  popular function invoked from \eval. Each call does a median of 2K instructions.
 \end{tabular}
@@
   \begin{minipage}{3.5cm}
+    \caption{Kinds per site} \label{tab:highlevelenvironments}
   \begin{tabular}{@{}rrr@{}}\toprule
@@
 \end{tabular}
-\caption{Kinds per site} \label{tab:highlevelenvironments}
 \end{minipage}%\hspace{-.2cm}
 \begin{minipage}{3.5cm}\centering
+  \caption{Function offsets}\label{tab:funoffset}
   \begin{tabular}{@{}rrr@{}}\toprule
@@
  \end{tabular}
-\caption{Function offset}\label{tab:funoffset}
 \end{minipage}%\hspace{-.2cm}
 \begin{minipage}{3.5cm}
+  \caption{Wrapper envs.} \label{tab:newenvs}
   \begin{tabular}{@{}rrr@{}} \toprule
@@
 \end{tabular}
-\caption{Wrapper envs.} \label{tab:newenvs}
 \end{minipage}%\hspace{-.2cm}
 \begin{minipage}{3.45cm}\centering
+  \caption{Multiplicities}\label{tab:polyenvir}
   \begin{tabular}{@{}ccc@{}} \toprule
@@
  \packageNbCategoryEnvird & \packageNbCategoryEnvirSitesdRnd & \packageNbCategoryEnvirPercentd\\\bottomrule
-\end{tabular}\caption{Multiplicities}\label{tab:polyenvir}
+\end{tabular}
 \end{minipage}\hspace{-1cm}
@@
 \begin{table}[!b]\small\centering
+  \caption{Origins}\label{tab:provenance}
 \begin{tabular}{c@{\hspace{2cm}}c}
-  \begin{tabular}{lcc}\toprule
+  \begin{tabular}{lrr}\toprule
 \bf Provenance & \bf \#sites & \bf \%sites \\\midrule
-		\packageProvenanceNamea & \packageNbProvenanceSitesa & \packagePercentProvenanceSitesa \\
-		\packageProvenanceNameb & \packageNbProvenanceSitesb & \packagePercentProvenanceSitesb \\
-		\packageProvenanceNamec & \packageNbProvenanceSitesc & \packagePercentProvenanceSitesc \\
-		\packageProvenanceNamee & \packageNbProvenanceSitese & \packagePercentProvenanceSitese \\
-		\packageProvenanceNamef & \packageNbProvenanceSitesf & \packagePercentProvenanceSitesf \\
-		\packageProvenanceNameg & \packageNbProvenanceSitesg & \packagePercentProvenanceSitesg \\
-		\packageProvenanceNameh & \packageNbProvenanceSitesh & \packagePercentProvenanceSitesh \\
-		\packageProvenanceNamei & \packageNbProvenanceSitesi & \packagePercentProvenanceSitesi \\
-		\packageProvenanceNamel & \packageNbProvenanceSitesl & \packagePercentProvenanceSitesl \\
-		\packageProvenanceNamen & \packageNbProvenanceSitesn & \packagePercentProvenanceSitesn \\\bottomrule
+		\packageProvenanceNamea & \packageNbProvenanceSitesa & \packagePercentProvenanceSitesa\unskip\% \\
+		\packageProvenanceNameb & \packageNbProvenanceSitesb & \packagePercentProvenanceSitesb\unskip\% \\
+		\packageProvenanceNamec & \packageNbProvenanceSitesc & \packagePercentProvenanceSitesc\unskip\% \\
+		\packageProvenanceNamee & \packageNbProvenanceSitese & \packagePercentProvenanceSitese\unskip\% \\
+		\packageProvenanceNamef & \packageNbProvenanceSitesf & \packagePercentProvenanceSitesf\unskip\% \\
+		\packageProvenanceNameg & \packageNbProvenanceSitesg & \packagePercentProvenanceSitesg\unskip\% \\
+		\packageProvenanceNameh & \packageNbProvenanceSitesh & \packagePercentProvenanceSitesh\unskip\% \\
+		\packageProvenanceNamei & \packageNbProvenanceSitesi & \packagePercentProvenanceSitesi\unskip\% \\
+		\packageProvenanceNamel & \packageNbProvenanceSitesl & \packagePercentProvenanceSitesl\unskip\% \\
+		\packageProvenanceNamen & \packageNbProvenanceSitesn & \packagePercentProvenanceSitesn\unskip\% \\\bottomrule
 \end{tabular}
@@
 \medskip
-\caption{Origins}\label{tab:provenance}
 \end{table}
@@
 \begin{table}[!h]
+  \caption{Target environments for side-effects} \label{tab:se-env}
   \begin{tabular}{lrrrr}
@@
   \end{tabular}
-  \caption{Target environments for side-effects} \label{tab:se-env}
 \end{table}
@@
 Function environments distinguish between \emph{local}, the caller environment
-(offset 0), and \emph{function} (offset >0). \emph{Synthetic} represents
+(offset 0), and \emph{function} (offset > 0). \emph{Synthetic} represents
 constructed environments. \emph{Object} is used to denote the environments
@@
   \centering
+  \caption{Types of \eval side-effects} \label{tab:se-types}
   \begin{tabular}{lrrrrrr}
@@
   \end{tabular}
-  \caption{Types of \eval side-effects} \label{tab:se-types}
 \end{table}
@@
 Out of the minimized expressions, it is \k{BLOCK}, \k{<-} and \k{F(F(X))} that
-contributes to the vast majority of side effects. Concretely, 92\% of \k{BLOCK},
+contribute to the vast majority of side effects. Concretely, 92\% of \k{BLOCK},
 87\% of \k{<-}, and 15\% of \k{F(F(X))} expressions do side effects. In the case
